%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiastes
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 1459
%%% o-1347
\startChapter[
  title={Chapter 06}
  ]

\Summary{Riches make not men happie, becauſe manie dye shortly: 3.~and
  manie rich men wil not vſe their riches. 8.~Likewiſe ſtudie to know al
  ſecrete thinges is vanitie, not felicitie.}


There is alſo an other euil, which I haue ſene vnder the ſunne, and that
frequent with men: \V A man to whom God hath geuen riches, and
ſubſtance, and honour, and nothing is lacking to his ſoule of al
thinges, which he deſireth: neither doth God geue him powre to eate
therof:
%%% 1460
but a ſtrange man shal eate it vp. This is
\SNote{Al this ſhevveth euidently that felicitie conſiſteth not in
riches.}
vanitie and great miſerie. \V If a man shal begette an hundred children, and
shal liue manie yeares, and haue manie dayes of age, and his ſoule vſe
not the goods of his ſubſtance, and he lacke burial: of this man I
pronounce,
\Fix{shat}{that}{obvious typo, fixed in other}
the vntimely borne is better then he. \V For he came in vaine, and
paſſeth to darknes, and his name shal be cleane forgotten. \V He hath
not ſene the ſunne, nor knowen the diſtance of good and euil: \V
although he liued two thouſand yeares, and hath not enioyed good
thinges: do not al thinges haſten to one place? \V Al the labour of man
is in his mouth: but his ſoule shal not be filled. \V What hath the
wiſeman more then the foole? and what the poore man, but to paſſe
thither, where life is? \V Better it is to ſee that, which thou maiſt
couete, then to deſire that, which thou canſt not know. But this alſo is
vanitie, and preſumption of ſpirite. \V
\SNote{It is plainly preached (ſaith S.~Iero.) of the coming of Chriſt,
vvhoſe name was geuen, and knovven before he vvas borne in fleſh.}
He that shal be, his name is already called: and it is knowne, that he
is a man, and can not contend in iudgement againſt a ſtronger then
himſelf. \V There be manie wordes, that haue much vanitie in diſputing.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
