%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiastes
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 1458
%%% o-1346
\startChapter[
  title={Chapter 05}
  ]

\Summary{An exhortation to ſpeake diſcretly, and reuerently of God; 3.~to
  performe vowes; 6.~not to be trubled with imaginations, nor preſent
  oppreſſions of the poore. 9.~Auarice is neuer ſatiate, 11.~riches
  ſometimes cauſeth ſicknes, ruine of the bodie, 18.~and obliuion of
  God.}


Speake
\SNote{Becauſe no man is able to attaine perfect knowlege of God, it
behoueth al to ſpeake and thinge ſoberly of him.}
not anie thing rashly, neither let thy hart be ſwift to vtter a word
before God. For God is in heauen, and thou vpon the earth: therfore let
thy wordes be few. \V Dreames do folow manie cares, and in manie wordes
follie wil be found. \V 
\CNote{\XRef{Deut.~23.}}
\SNote{Vow or promiſe once made muſt be fulfilled.}
If thou haſt vowed anie thing to God, differre not to pay it: for an
vnfaithful and foolish promiſe diſpleaſeth him. But what ſoeuer thou
haſt vowed, pay it: \V and it is much better not to vow, then after a
vow not to performe the thinges promiſed. \V Geue not thy mouth to make
thy flesh to ſinne: neither ſay thou before
\SNote{The proper Angel which aſſociateth euerie man.}
the Angel: There is no prouidence: leſt perhaps God being wrath againſt
thy wordes, diſſipate al the workes of thy handes. \V Where manie
dreames are, there are manie vanities, and wordes innumerable: but do
thou feare God. \V If thou shalt ſee the
%%% 1459
oppreſſions of the poore, and violent iudgements, and iuſtice to be
ſubuerted in the prouince, meruel not at this matter: becauſe there is
an other higher then the high, and ouer theſe alſo there are others more
eminent: \V and beſides the king of al the earth reigneth ouer his
ſeruant. \V A couetous man shal not be filled with money: and he that
loueth riches, shal take no fruite of them: and this therfore is
vanitie. \V Where great riches are, there are alſo manie that eate
them. And what doth it profite the owner, but that he ſeeth the riches
with his eyes? \V
\CNote{\XRef{Iob.~20.}}
Sleepe is ſwete to him that worketh, whether he eate
much or litle: but the ſatietie of the rich doth not ſuffer him to
ſleepe. \V There is alſo an other very il infirmitie, which I haue ſene
vnder the ſunne: riches kept to the hurt of the owner. \V For they
perish in very euil affliction: he
\SNote{As temporal riches are often the occaſion of their owners ruine:
ſo vaine philoſophie, and hereſie auaile not heretikes, nor their
folowers, but as they came naked from their mothers wombe, the peruerſe
Church, ſo shal both ſuch maſters, and ſcolars depart without comforth
into the wrath which they prepare to themſelues.
\Cite{S.~Ierom. in hunc locum.}}
hath begotten a ſonne, which shal be in great pouertie. \V As he came
forth naked from his mothers wombe, ſo shal he returne, and shal take
nothing away with him of his labour. \V An infirmitie vtterly
miſerable: as he came, ſo shal he returne. What doth it then profite
him, that he hath labored into the winde? \V Al the dayes of his life
he eateth in darknes, and in miſerie, and in heauines. \V This therfore
hath ſemed good to me, that a man eate, and drinke, and take ioy of his
labour, wherwith he hath labored vnder the ſunne, the number of the
dayes of his life, which God hath geuen him, and this is his portion. \V
And to euerie man, vnto whom God hath geuen riches, and ſubſtance, and
hath geuen him powre to eate of them, and to enioy his portion, and to
reioyce of his labour: this is the gift of God. \V For he
%%% o-1347
shal not greatly remember the dayes of his life, becauſe God doth
occupie his hart with delightes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
