%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiastes
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 1452
%%% o-1340
\startChapter[
  title={Chapter 01}
  ]

\Summary{Al
\MNote{The firſt part of this Sermon ſheweth, that Felicitie conſiſteth
  not in anie temporal thing.}
temporal thinges (in compariſon of true felicitie) are vaine,
4.~becauſe they are mutable, 8.~neither can anie man attaine perfect
knowlege, to his ſatisfaction: 12.~as appeareth by Salomons owne
experience.}


The wordes of Ecclesiastes, the ſonne of Dauid, king of Ieruſalem. \V
Vanitie of vanities, ſayd Ecclesiastes: vanitie of vanities, & al
thinges vanitie. \V What hath a man more of al his labour, wherby he
laboreth vnder the ſunne? \V Generation paſſeth,
%%% o-1341
and generation cometh: but
\SNote{In the end of this world the earth ſhal be purified, & ſo remaine
for euer.}
the earth ſtandeth for euer.
%%% 1453
\V The ſunne riſeth, and goeth downe, and returneth to his place: and
there riſing againe, \V compaſſeth by the South, and bendeth to the
North: compaſſing al thinges, goeth forward in circuite, & returneth
vnto his circles. \V Al riuers enter into the ſea, and the ſea
ouerfloweth not: to the place, whence the riuers iſſue forth, they do
returne, that they may flow againe. \V Al thinges are hard: man can not
explicate them in word. The eye is not filled with ſeing, neither is the
eare filled with hearing. \V What is that hath bene? the ſame thing that
shal be. What is that hath bene done? the ſame that is to be done. \V
\SNote{Mens ſoules dayly created are of the ſame kind as Adams ſoule:
other creatures either were in their kindes from the beginning of the
vvorld, or are procreated of diuers kindes preexiſting.
%%% !!! Part of this Cite is unreadable!
\Cite{S.~Tho. q.~73. a.~1. ad.~3.}}
Nothing vnder the ſunne is new, neither is anie man able to ſay: Behold
this is new: for it hath already gone before in the ages, that were
before vs. \V There is no memorie of former thinges: but neither of
thoſe thinges verily, which hereafter are to come, shal there be
remembrance with them, that shal be in the later end. \V I Ecclesiastes
haue bene king of Iſrael in Ieruſalem, \V and haue propoſed in my mind,
to ſeke and ſearch wiſely of al thinges, that are done vnder the
ſunne. This very euil occupation hath God geuen to the children of men,
that they might be occupied in it. \V I haue ſene al thinges, that are
done vnder the ſunne, and behold
\SNote{Al natural thinges are inſufficient meanes to attaine felicitie.}
al are vanitie, & affliction of ſpirit. \V The peruerſe are hardly
corrected, and the number of fooles is infinite. \V I haue ſpoken in my
hart, ſaying: Loe I haue bene made great, and haue gone beyond al in
wiſdom, that were before me in Ieruſalem: and my minde hath contemplated
manie thinges wiſely, and I haue lerned. \V And I haue geuen my hart to
know prudence, and doctrine, and errors and follie: and I haue perceiued
that in theſe alſo there was labour, and affliction of ſpirite, \V for
that
\LNote{In much vviſdom is much indignation.}{How
\MNote{VViſdom maketh men to be angrie with ſinne, to feare God, and to
proceede in vertue.}
much more anie man profiteth in wiſdom, ſo much more he is angrie with
himſelf, for the euils which he hath done: ſo much better he knovveth
the ſtrict iudgement of God, which he muſt paſſe: ſo much more he
vnderſtandeth the vvant of perfect vviſdom: ſo much more he ſeeth that
labour is required to procede in vertue: and to conſerue that ſmal
portion vvhich he hath gotten. Neither is anie man ordinarily aſſured
that he hath gotte anie part of true vviſdom, for he knovveth not,
vvhether he be vvorthie of loue or hate.
\XRef{Eccle.~9.}}
in much
\SNote{VViſdom is the beſt thing in this vvorld, yet is not perfect
felicitie, but the meanes to attaine it.}
wiſdom there is much indignation: and he that addeth knowlege, addeth
alſo labour.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
