%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiastes
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 1457
%%% o-1344
\startChapter[
  title={Chapter 04}
  ]

\Summary{In this world manie innocents are oppreſſed, 4.~the potent
  enuied, ſometimes contemned, 15.~and forſaken by their ſubiectes,
  17.~eſpecially when the ſuperiors obey not God.}


I turned myſelf to other thinges, and I ſaw the oppreſſions, that are done
vnder the ſunne, and the
\SNote{God ſuffereth the innocent to be afflicted for a time, of his
ſpecial prouidence: becauſe they therby merite a great reward.
\XRef{Pſal.~71.}}
teares of the innocents, and no comforter: & that they can not reſiſt
their violence, being deſtitute of al mens helpe. \V And I praiſed
rather the dead, then the liuing: \V and happier then both haue I iudged
him, that is not 
%%% !!! SNote not marked in this.
\SNote{It is in dede better not to be at al then to be in eternal
miſerie,
\XRef{(Mat.~26. v.~24.)}
but temporal affliction which the iuſt ſuffer is not miſerie, as
worldlie men eſteme it, but a ſpecial meane to attaine eternal
felicitie.}
yet borne, nor hath ſene the euils that are done vnder
the ſunne. \V Againe I haue contemplated al the labours of men, and
their induſtries I haue perceiued to lie open to the enuie of their
neighbour: and in this therfore there is vanitie, and ſuperfluous
care. \V A foole foldeth his handes together, and eateth his owne flesh,
ſaying: \V Better is an handful with reſt, then both handes ful with
labour, and affliction of mind. \V Conſidering I found alſo an other
vanitie vnder the ſunne: \V There is one, and he hath not a ſecond, not
a ſonne, nor a brother, and yet he ceaſeth not to labour, neither are
his eyes ſatisfied with riches, neither doth he recount,
%%% o-1345
ſaying: For whom do I labour, and defraud my ſoule of good thinges? in
this alſo is vanitie, and very il affliction. \V
\LNote{It is better that tvvo be together.}{Beſides
\MNote{Man without Chriſts helpe can neither reſiſt tentations, nor riſe
from ſinne.}
the commendation of charitie, and freindſhipe amongſt men: which is one
proper ſenſe of this place: S.~Ierom expoundeth it alſo of the
neceſſitie of Chriſts dwelling in mans ſoule, & of his continual
aſſiſting grace: that man alone lie not open to the deceiptes of the
aduerſarie. The benefite of feloshippe (ſayth he) is ſtreightwayes
ſhewed in the profite of ſocietie. For if the one fal (as
\CNote{\XRef{Prou.~24. v.~16.}}
the iuſt falleth often) Chriſt raiſeth vp his partner: for vvoe to him,
vvho falling hath not Chriſt in him, to raiſe him vp. If one alſo ſlepe,
that is, be diſſolued by death, and haue Chriſt vvith him, being vvarmed,
and quickned, he ſooner reuiueth. And if the diuel be ſtronger in
impugning againſt a man, the man shal ſtand, and Chriſt vvil ſtand
(\L{pro homine ſuo, pro ſodali ſuo}) for his man, for his companion.
\MNote{And without mans free conſent Chriſt wil not ſtay him, nor raiſe
him vp.}
Not that Chriſts povvre alone is vveake againſt the diuel, but that
freewil is leift to man, and we doing our endeuour he becometh ſtronger
in feighting. And if the Father, the Sonne, and the Holie Ghoſt come
vvithal, this ſodalitie is not ſoone broken. Yet that vvhich is not
ſoone broken, may ſometimes be broken. For this triple coard, vvas in
Iudas the Apoſtle, but becauſe
\CNote{\XRef{Ioan.~13. v.~27.}}
after the morſel the diuel entered into him, this coard was broken. Thus
\Fix{S.~Irom}{S.~Ierom}{obvious typo, fixed in other}
teacheth, that neither can man without Chriſt reſiſt tentations, nor
riſe from anie ſinne in this life, or
%%% !!! Not really a TNote
\TNote{In purgatorie.}
in the next; neither vvil Chriſt vvithout mans conſent and endeuour,
ſtay him from falling, nor raiſe him vp being fallen.}
It is better therfore that two be together, then one: for they haue
profite of their ſocietie: \V if one fal, he shal be ſtayed vp of the
other. Woe to him that is alone: becauſe when he falleth, he hath none
to lift him vp. \V And if two ſleepe together, they shal warme eche
other: one how shal he be warmed? \V And if a man preuaile againſt one,
two reſiſt him: a
\SNote{True charitie not contaminate with enuie, by how much it is
increaſed in number, ſo much it increaſeth in ſtrength.
\Cite{S.~Ierom.}}
triple coard is hardly broken. \V Better is a child that is poore and
wiſe, then a king old and foolish, that knoweth not to foreſee for
hereafter. \V Becauſe out of the priſon, and cheynes ſometime there
cometh one forth to a kingdom: and an other borne in his kingdom, is
conſumed with pouertie. \V I ſaw al men aliue, that walke vnder the
ſunne, with the ſecond yongman, which shal riſe vp for him. \V The
number of the people, of al that haue bene before him is infinite: and
they that shal be afterward, shal not reioyce in him. But this alſo is
vanitie, and affliction of ſpirite. \V Take heede to kepe thy foote,
when thou entreſt into the houſe of
%%% 1458
God, and approch thou to heare. \V
\CNote{\XRef{1.~Reg.~15.}
\XRef{Oſee.~6.}}
For much better is obedience, then the victimes of fooles, who know not
what euil they doe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
