%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiastes
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 1466
%%% o-1353
\startChapter[
  title={Chapter 11}
  ]

\Summary{Workes
\MNote{The 3.~part.

An exhortation to beginne quickly, and perſeuere in Gods ſeruice.}
of mercie are neceſſarie, whiles we haue time, 3.~becauſe after death
none can merite: 4.~neither muſt we differ to beginne, nor ceaſe from
good dedes, 8.~but ſtil mindful of death and iudgement: 10.~auoiding
wrath and malice.}


Caſt thy bread vpon the paſſing waters:
\SNote{Of al vertues, the workes of mercie, corporal and ſpiritual, moſt
auaile for obtaining eternal felicitie:
\XRef{Mat.~25.}}
becauſe after much time thou shalt finde it. \V Geue a portion
\SNote{So the ſame be grounded in true faith, beleuing al that is
written in the old and new teſtament ſignified by ſeuen & eight.}
to ſeuen, and alſo to eight: becauſe thou knoweſt not what euil shal be
vpon the earth. \V If the cloudes be ful, they wil powre out raine vpon
the earth. If the tree shal fal to the South, or to the North, in what
place ſoeuer it shal fal:
\SNote{After death none can either merite or demerite.}
there shal it be. \V He that obſerueth the winde, ſoweth not: and he
%%% 1467
that conſidereth the cloudes, shal neuer reape. \V As thou art ignorant
which is the way of the ſpirite, & how the bones are framed together in
the wombe of her that conceiueth childe: ſo thou knoweſt not the workes
of God, who is the maker of al. \V In
\SNote{Both in youth and old age do good workes.}
the morning ſow thy ſeede, and in the euening let not thy hand ceaſe:
for thou knoweſt not which may rather ſpring, this or that: and if both
together, it shal be the better. \V The light is ſweete: and it is
delectable for the eyes to ſee the ſunne. \V If a man shal liue manie
yeares, and shal haue reioyced in them al, he muſt remember the
darkeſome time, and manie dayes: which when they shal come, the thinges
paſt shal be reproued of vanitie. \V Reioyce therfore yongman in thy
youth, and let thy hart be in good, in the dayes of thy youth, and walke
in the wayes of thy hart, and in the ſight of thyne 
eyes: and know that for al theſe God wil bring thee into iudgement. \V
Take away anger from thy hart, and remoue malice from thy flesh. For
youth and pleaſure are vaine.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
