%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiastes
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 1467
%%% o-1354
\startChapter[
  title={Chapter 12}
  ]

\Summary{In youth is fitteſt time, and moſt meritorious to ſerue God. In
  age the ſame is more and more neceſſarie, but harder to beginne, and
  leſſe gratful. 8.~In this booke the preacher hath shewed, that al
  worldlie thinges are vanitie, 13.~and that true felicitie is only
  procured by wiſdom, which 
  \Fix{conſteth}{conſiſteth}{obvious typo, fixed in other}
  in the feare of God; and obſeruation of his commandments.}


Remember thy Creator in the dayes
\SNote{An admonition to al in general to liue wel in this world,
remembring the day of general iudgement before which ſuch ſignes ſhal
come as are deſcribed here, and by our Sauiour.
\XRef{Mat.~24.}
And likewiſe euerie one is admoniſhed in particular to ſerue God diligently whiles
he hath time, before death come, when al his ſenſes & former helpes shal
faile.}
of thy youth, before the time of affliction come, & the yeares approch,
of which thou maiſt ſay: They pleaſe me not, \V before the ſunne, and
light, and moone, and ſtarres be darke, and the cloudes returne after
the raine: \V when the kepers of the houſe shal be moued, and the
ſtrongeſt men shal ſtagger, and the grinders shal be idle in a ſmal
number: and they shal waxe darke that looke through the holes: \V and
they shal shut the doores in the ſtreate, at the baſenes of the grinders
voice, and they shal riſe vp at the voice of the birde, and al the daughters
of ſong shal be deafe. \V The high thinges alſo shal feare, and they
shal be afrayd in the way, the almondtree shal florish, the locuſt
shal be fatted, and the capertree shal be deſtroyed: becauſe man shal goe
into the houſe of his eternitie, and the mourners shal goe round about
in the ſtreate. \V Before the ſiluer coard be broken, and the golden
headband recurre,
%%% 1468
and the water pot be broken vpon the fountaine, and the wheele be broken
vpon the ceſterne, \V and the duſt returne into his earth, from whence
it was, and the ſpirite returne to God, who gaue it. \V
Vanitie of vanities, ſayd
\TNote{The preacher.}
Ecclesiastes, and al thinges vanitie. \V And wheras Ecclesiastes was
moſt wiſe, he taught the people, and declared the thinges that he had
done: and ſearching forth made manie parables. \V He ſought profitable
wordes, and wrote wordes moſt right, and ful of truth. \V The wordes of wiſemen
are as prickes, and as nailes deeply ſtricken in, which by the counſel
of maiſters are geuen of one paſtour. \V More then theſe my ſonne
require not. Of making manie bookes there is no end: and often
meditation is affliction of the flesh. \V
\SNote{This is the
\Fix{brife}{brief}{likely typo, fixed in other}
ſumme of al profitable doctrine: Feare God, & kepe his commandments.}
Let vs al heare together the end of ſpeaking. Feare God, and obſerue his
commandments: for this is euerie man: \V and al thinges that are done,
God wil bring into iudgement for euerie
\TNote{Hidden, or obſcure thing.}
errour, whether it be good, or euil.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
