%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiastes
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 1463
%%% o-1350
\startChapter[
  title={Chapter 09}
  ]

\Summary{None knoweth (certainly and ordinarily) whether they be in Gods
  grace or no. 4.~The euil are in worſe caſe dead then aliue,
  11.~neither can we know the euent of temporal thinges, nor the terme
  of our life, nor how gratful others wil be towards vs. 16.~Sure it is,
  that wiſdom is better then ſtreingth.}


Al theſe thinges haue I diſcourſed in my hart, that I might curiouſly
vnderſtand them: there are iuſt men and wiſe, and their workes are in
the hand of God: and yet
\SNote{Mortal men ſuffering calamities knovv not vvhether the ſame be
inflicted for their proofe and merite as in Iob and Tobie, or for their
ſinnes only, as in Pharao, & the Ægyptians: but shal know in the
iudgement after their death.}
man knoweth not, whether he be worthie of loue, or hatred: \V but al
thinges are reſerued vncertaine for the time to come, becauſe al thinges
do equally chance to the iuſt and impious, to the good and the euil, to
the cleane and vncleane, to him that
\Fix{immoleth}{immolateth}{obvious typo, fixed in other}
victimes, and him that contemneth ſacrifices. As the good ſo alſo is the
ſinner: as the periured, ſo he alſo that ſweareth truth. \V This is a
very euil thing among al, which are done vnder the ſunne, that the ſame
thinges chance to al men. Wherby alſo the hartes of the children of men
are filled with malice, and with contempt in their life, and after that
they shal be brought downe to hel. \V There is no man that may liue
alwayes, and that can haue confidence of this thing: better is
\SNote{A ſinner in this life may amend if he wil, & become iuſt: but
after death he can not repent.
\XRef{chap.~11. v.~3.}
Myſtically, the Gentiles were iudged by Salomon better then the Iewes.
\Cite{S.~Ierom.}}
a dog liuing then a lion dead. \V For the liuing know that they shal
dye, but the dead know nothing more, neither haue they reward anie more:
becauſe the memorie of them is forgotten. \V Loue alſo, and hatred, and
enuies haue perished together, neither haue they part in this world, and
in the worke, that is done vnder the ſunne.
%%% 1464
\V Goe therfore and eate thy bread in ioy, & drinke thy wine with
gladnes: becauſe thy workes pleaſe God. \V At al time let thy garments
be white, and let not oyle fal from of thy head. \V Enioy life, with thy
wife whom thou loueſt, al the dayes of the life of thy inſtabilitie,
which are geuen to thee vnder the ſunne, al the time of thy vanitie: for
this is the portion in life, and in thy labour, wherwith thou laboreſt
vnder the ſunne. \V Whatſoeuer thy hand is able to doe, worke it
inſtantly: for neither worke, nor reaſon, nor wiſdom nor knowlege shal
be in hel, whither thou doſt haſten. \V I turned me to an other thing,
and I ſaw vnder the ſunne, that neither running is of the ſwift, nor
warre of the ſtrong, nor bread of the wiſe, nor riches of the lerned,
nor grace of the artificers: but time and chance in al. \V Man knoweth
not his owne end: but as fishes are taken with the hooke, and as birdes
are caught with the ſnare: ſo men are taken in the euil time, when it
shal ſudenly come vpon them. \V This wiſdom alſo I haue ſene vnder the
ſunne, and haue proued it to be very great: \V A litle citie, and few
men in it: there came againſt it a great king, and compaſſed it, and
builded fortes
%%% o-1351
round about, and the ſiege was perfited. \V And there was found in it a
man poore and wiſe, and he deliuered the citie by his wiſdom, and no man
afterward remembred that pooreman. \V And I ſayd, that wiſdom is better
then ſtreingth: how then was the wiſdom of the pooreman contemned, & his
wordes were not heard? \V The wordes of the wiſe are heard in ſilence,
more then the crie of a prince among fooles. \V Better is wiſdom, then
weapons of warre: and he that shal offend in one point, shal loſe manie
good thinges.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
