%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiastes
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 1460
%%% o-1347
\startChapter[
  title={Chapter 07}
  ]

\Summary{It is in vaine to ſeke, and vnpoſſible to know al natural
  thinges. 2.~It importeth to leade this shorte life in mortification,
  4.~pœnance, 8.~and patience: 12.~ſeeking wiſdom, with competent
  temporal meanes; 15.~prouiding for the next world; 24.~not yelding to
  concupiſcence.}


VVhat
\SNote{None can perfectly knovv the nature of al thinges preſent, or to
come.}
nedeth a man to ſeke thinges greater then himſelf, wheras he is
ignorant, what is profitable for him in his life, in the number of the
dayes of his peregrination, and the time that paſſeth as a shadow? Or
who can tel him what shal be after him vnder
%%% o-1348
the ſunne?

\V
\MNote{The 2.~part.

True felicitie is procured by good life, and conſiſteth in the eternal
ſight of God.}
Better is a good name then precious ointments: and the day of death,
then the day of natiuitie. \V 
\CNote{\XRef{Prou.~22.}}
It is better to goe to the houſe of mourning, then to the houſe of
banketing: for in that the end of al men is ſignified, and he that
liueth thinketh what shal be. \V
\SNote{Anger rightly vſed, is commendable & beneficial to correcte our
owne faultes and others.}
Anger is better then laughter: becauſe by ſadnes of the countenance, the
mind of the offender is corrected. \V The hart of wiſemen where ſadnes
is, and the hart of fooles where mirth. \V It is better to be rebuked of
a wiſeman, then to be deceiued with the flaterie of fooles. \V Becauſe
as the ſound of thornes burning vnder a potte, ſo
%%% 1461
the laughter of a foole: but this alſo is vanitie. \V Oppreſſion
trubleth the wiſe, and shal deſtroy the ſtreingth of his hart. \V Better
is the end of a ſpeach, then the beginning.
\SNote{But haſtie, or immoderate anger is ſinful and hurtful.}
Better is the patient man then the arrogant. \V Be not quickly angrie,
becauſe anger reſteth in the boſom of a foole. \V Say not: What is the
cauſe thinkeſt thou that the former times were better then they are now?
for this maner of queſtion is foolish. \V Wiſdom with
\SNote{Likewiſe moderate riches profite the ſeruants of God, ſo that
they ſette not their mind vpon them.}
riches is more profitable, and doth more profite them that ſee the
ſunne. \V For as wiſdom protecteth, ſo money protecteth. But lerning and
wiſdom haue this much more, that they geue life to their owner. \V
Conſider the workes of God, that no man can correct whom he hath
deſpiſed. \V In the good day enioy good thinges, and beware before of
the euil day. For as this, ſo that alſo hath God made, that man finde
not againſt him iuſt
\Fix{complants.}{complaints.}{likely typo, fixed in other}
\V Theſe thinges alſo I ſaw in the dayes of my vanitie: The iuſt man
perisheth in his iuſtice, and the impious liueth a long time in his
malice. \V Be not iuſt too much: neither be more wiſe, then is
neceſſarie, leſt thou become more dul. \V Doe not impiouſly much: and be
not foolish, leſt thou dye not in thy time. \V It is good that thou hold
vp the iuſt; yea and from him withdraw not thy hand: becauſe he that
feareth God, neglecteth nothing. \V Wiſdom hath ſtreingthned the wiſe
aboue tenne princes of the citie. \V
\CNote{\XRef{3.~Reg.~8.}
\XRef{2.~Par.~6.}
\XRef{Pro.~20.}
\XRef{1.~Ioan.~1.}}
For there is no iuſt man in the earth, that doth good, and ſinneth
not. \V But to al wordes alſo, that are ſpoken, do not applie thy hart:
leſt perhaps thou heare thy ſeruant curſing thee. \V For thy conſcience
knoweth, that thou alſo hath curſed others. \V I haue proued al thinges
in wiſdom. I haue ſayd: I wil become wiſe, & it departed farder from me
\V much more then it was: and a depe profunditie, who shal finde it? \V
I haue vewed al thinges with my minde, that I might know, and conſider,
and might ſeke wiſdom, and reaſon: and that I might know the impietie of
the foole, and the errour of the imprudent: \V and I haue found that a
woman is more bitter then death, who is the ſnare of hunters, and her
hart a nette, her handes are bandes. He that pleaſeth God, wil auoide
her: but he that is a ſinner, wil be caught of her. \V Loe this haue I
found, ſayd
\TNote{The preacher.}
Eccleſiaſtes, one thing and an other, that I might finde
reaſon, \V which yet my ſoule ſeketh, and I haue not found it.
\SNote{Mans reaſon which is the ſuperior part of his ſoule, rarely
thinketh good:}
A man of a thouſand I haue found one,
\SNote{Senſualitie ſtil inclineth to euil.}
a woman of al I haue not
%%% 1462
found. \V Only this I haue found, that
\SNote{God made man right, & he of his owne wil fel from God ſee
\Cite{S.~Aug. li.~14. c.~11. ciuit.}}
God made man right, and he hath intangled himſelf with infinite
%%% o-1349
queſtions. Who is ſuch a one as the wiſe? and who hath knowne the
reſolution of
\TNote{The obſcure thing.}
the word.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
