%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiastes
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 1455
%%% o-1343
\startChapter[
  title={Chapter 03}
  ]

\Summary{Contrarie thinges ſuccede in their ſeaſons, and paſſe away,
  9.~wherof man getteth no perfect knowlege, how long ſoeuer he liueth,
  and laboreth to know them: 16.~neither wil there be equitie where it
  ought to be in this world: 21.~but in the next, good and euil shal be
  ſeparated, and iudged according to their deſertes.}


Al thinges haue a time, and in their ſpaces al thinges paſſe vnder
\SNote{Generation, Corruption, & Alteration perteine only to thinges of
this world vnder heauen, often here deſcribed by the terme, Vnder the
ſunne. For ſpiritual ſubſtances are not comprehended in time, nor place.
\Cite{S.~Ierom in hunc locum. to.~7.}}
heauen. \V A time to be borne, & a time to dye. A time to plant, & a
time to pluck vp that which was planted. \V A time to kil, and a time to
heale. A time to deſtroy, and a time to builde.

%%% 1456
\V A time to wepe, & a time to laugh. A time to mourne, and a time to
dance.

\V A time to diſperſe ſtones, and a time to gather.

A time to embrace, and a time to be farre from embracings.

\V A time to gette, and a time to loſe.

A time to kepe, and a time to caſt away.

\V A time to rent, and a time to ſow together.

A time to kepe ſilence, & a time to ſpeake.

\V A time of loue, and a time of hatred. A time of warre, and a time of
peace. \V
\SNote{Al this ſheweth that nothing of this world is perpetual, but to be
vſed interchangeably, in due time, place, maner, and order. For diſorder
maketh confuſion.}
What hath man more of his labour? \V I haue ſene the affliction, which
God hath geuen to the children of men, that they may be diſtracted in
it. \V He hath made al thinges good in their time, and hath deliuered
the world to their diſputation, and that man can not finde the worke,
which God hath wrought from the beginning vnto the end.
%%% o-1344
\V And I haue knowne that there was no better thing then to reioyce, and
to do wel in
\Fix{this}{his}{likely typo, same in both}
life. \V For euerie man, that eateth and drinketh, and
ſeeth good of his labour, this is the gift of God. \V I haue lerned that
al the workes, which God hath made, perſeuere for euer: we can not adde
anie thing, nor take away from thoſe thinges, which God hath made that
he may be feared. \V That which hath bene made, the ſame is permanent:
the thinges that shal be, haue already bene: and God reſtoreth that
which is paſt. \V I ſaw vnder the ſunne in the place of iudgement
impietie, and in the place of iuſtice iniquitie. \V And I ſayde in my
hart: The iuſt and the impious God wil iudge, and then shal be the time
of euerie thing. \V I ſayd in my hart of the children of men, that God
would proue them, and shew them to be like beaſtes. \V Therfore there is
one death of man, and beaſtes, and the condition of both equal: as man
dieth, ſo they alſo dye: al thinges breath alike, and man hath nothing
more then beaſt: al thinges are ſubiect to vanitie, \V and al thinges
paſſe to one place: of earth they were made, and into earth they returne
together. \V Who knoweth if the ſpirit of the children of Adam aſcend
vpward, and if the ſpirite of beaſtes deſcend downward? \V And I haue
found that nothing is better then for a man to reioyce in his worke, and
that this is his portion. For who shal bring him, to know the thinges
that shal be after him?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
