%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiastes
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 1454
%%% o-1342
\startChapter[
  title={Chapter 02}
  ]

\Summary{Humane
\Fix{deligthes}{delightes}{obvious typo, fixed in other}
are al vaine: 4.~as gorgious buildinges, fruitful vinyards, plentie of
fish, cattle, ſeruantes, ſiluer, gold, muſike: 11.~not ſatisfying mans
deſire. 18.~Neither can anie man know, how his heyre wil behaue
himſelf.}


I ſayde
\SNote{A
\Fix{vvordlie}{vvorldlie}{obvious typo, same in both}
man may obiect, that ſeing vviſdom bringeth not felicitie in his life,
it ſemeth beſt to take his pleaſure & eaſe, & not to labour for it.}
therfore in my hart: I wil goe, & flow in delightes, and enioy good
thinges.
\SNote{But the wiſeman cõfuteth this conceipt. Becauſe
\Fix{vvordlie}{vvorldlie}{obvious typo, fixed in other}
ioy is ſhort & vncertaine, neither is anie ioy that men haue in this
life true ioy of the bleſſed, but only a conſolation in miſeries.}
And I ſaw that this alſo was vanitie. \V Laughter I haue reputed errour:
and to ioy I haue ſaide: Why art thou deceiued in vaine? \V I haue
thought in my hart, to withdraw my flesh from wine, that I might
transferre my minde to wiſdom, and might auoid follie, til I might ſee
what should be profitable for the children of men: what is nedeful to
be done vnder the ſunne, in the number of the dayes of their life. \V I
haue magnified my workes, I haue built me houſes, & planted
vineyards, \V I haue made gardens, and orchards, and ſet them with trees
of al kindes, \V and I haue made me ponds of waters, to watter the wood
of ſpringing trees, \V I haue poſſeſſed menſeruants and wemenſeruants,
and haue had a great familie: heardes alſo, and great flockes of shepe,
aboue al that were before me in Ieruſalem: \V I haue heaped together to
myſelf ſiluer, and gold, and the ſubſtance of kinges, and prouinces: I
made me ſingingmen, & ſingingwemen, and the delightes of the
children of men: cuppes, and goblets to ſerue to powre out wines: \V and
I ſurpaſſed in riches al, that were before me in Ieruſalem: wiſdom alſo
hath perſeuered with me. \V And al thinges, that myne eies deſired, I
haue not denied to them: neither haue I ſtayed my hart, but that it
enioyed al pleaſure, & delighted itſelf in theſe thinges, which I had
prepared: and this I eſtemed my portion, if I did vſe my labour. \V And
when I had turned myſelf to al the workes, which my handes had done, &
to the laboures, wherin I had ſwette in vaine, I ſaw in al thinges
vanitie, and affliction of minde, & nothing to be permanent vnder the
ſunne. \V I paſſed further to contemplate wiſdom, and errors, and follie
(what is man, quoth I, that he can folow the king his Maker?) \V and I
ſaw that wiſdom ſo much excelled follie, as light differeth from
darknes. \V The
\SNote{Conſideration is the directorie of al good workes.}
eyes of a wiſeman are in his head: the foole walketh in darknes: and I
haue lerned that there was one death of both. \V And I ſayd in my hart:
If the fal of the foole & myne shal
%%% 1455
be one,
\SNote{The conceit of worldlie men not conſidering the life to come.}
what doth it profite me, that I haue beſtowed greater labour for wiſdom? And
ſpeaking with my minde, I perceiued that this alſo was vanitie. \V For
there shal be no memorie of the wiſe in like maner as of the foole for
euer, and the times to come shal couer al thinges together with obliuion:
the lerned dieth in like maner as the vnlerned. \V And therfore I haue
bene wearie of my life, ſeing al thinges vnder the ſunne to be euil, and
al thinges vanitie and affliction of ſpirite. \V Againe I deteſted al
myne induſtrie, wherwith I haue laboured vnder the ſunne moſt
ſtudiouſly,
%%% o-1343
being like to haue an heyre after me: \V whom I know not, whether he wil
be a wiſeman or a foole, and he shal rule in my labours, wherewith I
haue ſwette and haue bene careful: and is there anie thing ſo vaine? \V
Wherfore I ceaſed, and my hart hath renounced to labour anie more vnder
the ſunne. \V For whereas one laboreth in wiſdom, and doctrine, and
carefulnes, he leaueth the thinges gotten to an idle man: and this
therfore is vanitie, and great euil. \V For what profite shal be to a
man of al his labour, and affliction of ſpirite, wherwith he is vexed
vnder the ſunne? \V Al his daies are ful of ſorowes and miſeries,
neither by night doth he reſt in minde, and is not this vanitie? \V Is
it not better to eate and drinke, and shew vnto his ſoule good thinges
of his laboures?
\SNote{It is better to vſe that is honeſtly gotten for our neceſſitie,
then to be ſtil ſolicitous to get more.}
& this is of the hand of God. \V Who shal ſo deuour, and flow with
delightes as I? \V To a man good in his ſight, God hath geuen wiſdom,
and knowlege, and ioy: but to the ſinner he hath geuen affliction, and
ſuperfluous care, to adde, and to gather together, and deliuer it to him
that hath pleaſed God: but this alſo is vanitie, & vaine carefulnes of
the minde.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
