%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiastes
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 1462
%%% o-1349
\startChapter[
  title={Chapter 08}
  ]

\Summary{A ſigne of true wiſdom appereth in obſeruing Gods commandments,
  6.~in this shorte time of meriting eternal reward. 9.~Rule of others,
  11.~and want of feare hurt manie. 14.~Why God ſuffereth the wicked to
  proſper, and the iuſt to be afflicted in this life, no mortal man can
  know.}


The wiſdom of a man
\SNote{As probable coniecture of a mans inward diſpoſition is made by
his exteriour countenance: ſo his good workes ſhining before men are
good ſigne of internal vertues; which in dede are right and meritorious,
when the intention is ſincere, referring al to Gods glorie, &
edification of others, without deſire of vaine praiſe in the world.}
shineth in his countenance, and the moſt mightie wil change his face. \V
I obſerue the mouth of the king, and the precepts of the oath of God. \V
Haſten not to depart from his face, nor continew thou in an euil worke:
becauſe al that he pleaſeth, he wil doe, \V and his word is ful of
powre: neither can anie man ſay to him: Why doſt thou ſo? \V He that
kepeth the precept, shal finde no euil. The hart of a wiſeman
vnderſtandeth time and anſwer. \V There is a time for al buſines, and
opportunitie, and much affliction of man: \V becauſe he is ignorant of
thinges paſt, and thinges to come he can know by no meſſenger. \V It is
not in mans powre to prohibite the ſpirite, neither hath he powre in the
day of death, neither is he ſuffered to reſt when warre is at hand,
neither shal impietie ſaue the impious. \V Al theſe thinges I haue
conſidered, and gaue my hart on al the workes, that are done vnder the
ſunne. Sometime man ruleth ouer man to his owne hurt. \V I ſaw the
impious buried: who alſo when they yet liued, were in holie place, and
were praiſed in the citie as men of iuſt workes. But this alſo is
vanitie. \V Becauſe ſentence is not ſpeedely pronounced againſt the
euil, the children of men committe euils without anie feare. \V But yet
a ſinner by this that he doth euil an hundred times, & by patience is
borne withal, I know that it shal be good to them that feare God, which
dread his face. \V Let there be no good to the impious, neither let his
dayes be prolonged, but as a shadow let them paſſe, that feare not the
face of our Lord. \V There is alſo an other
\SNote{By this terme vanitie is ſtil vnderſtood that felicitie is not in
proſperous thinges of this world: neither are al men miſerable, that
ſuffer aduerſitie.}
vanitie, which is done vpon the earth. There are iuſt men, to whom euils
happen, as though they had done the workes of the impious: and there are
impious men, which are ſo
%%% 1463
ſecure, as though they had the dedes of the iuſt. But this alſo I iudge
moſt vaine. \V I therfore haue praiſed mirth that there was no good
thing for a man vnder the ſunne, but that he should eate, and drinke,
and be glad: and this only he should take away with him of his labour in
the dayes of his life, which God hath geuen him vnder the ſunne. \V And
I haue ſet my hart to know wiſdom, and to vnderſtand the
\TNote{Buſie ſearching.}
diſtraction that is in the earth: There is a man that dayes and nightes
taketh no ſlepe with his eyes. \V And I vnderſtood that man can finde no
reaſon of al thoſe workes of God, that are done vnder the ſunne: and the
more he shal labour to
%%% o-1350
ſeke, ſo much the leſſe he can finde: yea if the wiſman shal ſay, that
he knoweth, he is not able to finde it.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
