%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% One Thessalonians
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 2787
%%% o-2647
\startChapter[
  title={Chapter 2}
  ]

\Summary{He calleth euen themſelues to witnes, that his preaching vnto
  them was as he ſaid, in moſt commendable manner. 13.~And againe on the
  other ſide he thanketh God for their manner of receiuing it: that is,
  with al ioy, notwithſtanding the perſecution of their owne citizens.}

For your ſelues know, Brethren, our entrance vnto you, that it was not
vaine: \V but
\SNote{A notable example for Catholike Preachers, and paſſing
comfortable, when in the middes of perſecutions and reproches they
preach ſincerely, to pleaſe God & not men.}
hauing ſuffered before and
\CNote{\XRef{Act.~16,~12.~23.}}
been abuſed with contumelies (as you know) at Philippi, we had
confidence in our God, to ſpeake vnto you the Ghoſpel of God in much
carefulnes. \V For our exhortation was not of errour, nor of
vncleanneſſe, nor in deceit: \V but as we were approued of God that the
Ghoſpel should be committed to vs, ſo we ſpeake: not as pleaſing men,
but God, who proueth our harts. \V For neither haue we been, at any time
in the word of adulation, as you know; nor in occaſion of auarice, God
is witnes: \V nor ſeeking glorie of men, neither of you, nor of
others. \V Wheras we might haue been a burden to you, as the Apoſtles
of Chriſt; but we became
\Var{children}{milde}
in the middes of you, as if a nource should cherish her children: \V ſo
hauing a deſire to you, we would gladly deliuer vnto you not only the
Ghoſpel of God, but alſo our owne ſoules: becauſe you are become moſt
deare vnto vs. \V For you are mindful, Brethren, of
\CNote{\XRef{Act.~20,~34.}
\XRef{1.~Cor.~4,~12.}
\XRef{2.~Theſſ.~3.}}
our labour and toile. Day and night working, leſt we should charge any
of you, we preached among you the Ghoſpel of God. \V You
%%% o-2648
are witneſſes and God, how holily, and iuſtly & without blame, we haue
been to you that did beleeue. \V As you know in what manner we deſiring
and
\Fix{conforting}{comforting}{possible typo, fixed in other}
you, haue adiured euery one of you (as a father his children) that you
would walke worthie of God, who hath called you into his Kingdome and
glorie.

\V Therfore we alſo giue thankes to God without intermiſſion: becauſe
that when you had receiued of vs the word of the hearing of God, you
receiued it not as the word of men, but (as it is indeed)
\LNote{The word of God.}{The
\MNote{Not only the written word is the word of God.}
Aduerſaries wil haue no word of God but that which is written and
conteined in the Scripture: but here they might learne that al Paules
preaching before he wrote to them, was the very word of God. They might
alſo learne that whatſoeuer the lawful Apoſtles, Paſtours, and Prieſtes
of God's Church preach in the vnitie of the ſame Church, is to be taken
for God's owne word, & ought not to be reputed of them for doctrines of
men or Phariſaical traditions, as they falſly cal canons, precepts, and
decrees of holy Church.}
the word of God, who worketh in you that haue beleeued. \V For you,
Brethren, are become followers of the Churches of God that be in Iewrie,
in Chriſt \Sc{Iesvs}: for you alſo haue ſuffered the ſame things of your
owne lineage, as they alſo of the Iewes, \V who both killed our Lord
\Fix{\Sc{Iessvs,}}{\Sc{Iesvs},}{obvious typo, fixed in other}
and the Prophets, and haue perſecuted vs, and pleaſe not God, and are
aduerſaries to al men, \V prohibiting vs to ſpeake to the Gentils that
they may be ſaued, to make vp their ſinnes alwaies. For the wrath of God
is come vpon them euen to the end. \V But we, Brethren,
\TNote{\G{ἀπορφανισθέντες}}
depriued of you for a ſhort time, in ſight, not in hart; haue haſtned
the more aboundantly to ſee your face with much deſire. \V For we would
haue come to you, I Paul certes, once and againe: but Satan hath
hindred vs. \V For what is our hope, or ioy, or crowne of glorie?
\SNote{If the Apoſtle without iniurie to God, in right good ſenſe cal
his ſcholers the Theſſalonians, his hope, ioy, glorie; why blaſpheme the
Proteſtants the Cath. Church and her children for terming our B.~Ladie
or other Saints, their hope, for the ſpecial confidence they haue in
their praiers?}
Are not you, before our Lord \Sc{Iesvs} Chriſt in his comming? \V For
you are our glorie and ioy.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

