%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% One Thessalonians
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 2790
%%% o-2651
\startChapter[
  title={Chapter 5}
  ]

\Summary{To talke of the time of the Reſurrection is not neceſſarie, but
  to prepare our ſelues againſt that time ſo ſodaine, and ſo terrible to
  the vnprepared. 12.~He beſeecheth the layetie to be obedient, 14.~and
  the Clergie to be vigilant, with many short precepts moe.}

And of the times and momentes, Brethren, you need not that we write to
you. \V For your ſelues know perfectly that the day of our Lord ſhal
ſo come as
\CNote{\XRef{Mt.~24,~44.}
\XRef{2.~Pet.~3,~10.}
\XRef{Apoc.~3,~3.}
\XRef{16,~15.}}
a theefe in the night. \V For when they ſhal ſay, peace & ſecuritie;
then ſhal ſodaine deſtruction come vpon them, as the paines to her that
is with child, and they ſhal not eſcape. \V But you, Brethren, are not
in darkeneſſe; that the ſame day may as a theefe ouer-take you.

\V For you are the children of light, and children of the day: we are
not of the night nor of darkneſſe. \V Therfore let vs not ſleep as alſo
others: but let vs watch & be ſober. \V For they that ſleep, ſleep in
the night; & they that be drunke, be drunke in the night. \V But we that
are of the day, are ſober,
\CNote{\XRef{Eſa.~59,~27.}
\XRef{Eph.~6,~17.}}
hauing on, the breaſt-plate of faith and
\SNote{A Chriſtian man's whole armour is not faith only, but al the
three vertues here named.}
charitie, and a helmet, the hope of ſaluation. \V For God hath not
appointed vs vnto wrath, but vnto the purchacing of ſaluation by our
Lord \Sc{Iesvs} Chriſt, \V who died for vs: that whether we watch, or
ſleep, we may liue together with him. \V For the which cauſe comfort one
another: and edifie one another, as alſo you doe.

\V And we beſeech you, Brethren, that you wil know them that labour
among you, and that gouerne you in our Lord and admoniſh you: \V that
you haue them more aboundantly in charitie for their worke. Haue peace
with them. \V And we beſeech you, Brethren, admoniſh the vnquiet,
comfort the weake-minded, beare vp the weake, be patient to al. \V See
that
\CNote{\XRef{Pro.~17,~13.}
\XRef{Ro.~12.}
\XRef{1.~Pet.~3,~9.}}
none render euil for euil to any man: but alwaies that which is good
purſue toward each other, & towards al. \V Alwaies reioyce. \V
\SNote{To deſire eternal life of him that only can giue it, is to pray
without intermiſſion: but becauſe that deſire is oftẽ by worldly cares
cooled, certaine houres & times of vocal praier were appointed. See
\Cite{S.~Aug. ep.~121. ad Probam.}}
Pray
\CNote{\XRef{Luc.~18,~1.}}
without intermiſſion. \V In al things giue thankes. For this is the wil
of God in Chriſt \Sc{Iesvs} in al you. \V The Spirit extinguiſh not. \V
Prophecies deſpiſe not. \V But
\LNote{But proue.}{Though
\MNote{Not rashly to credit euery ſpirit.}
we may not extinguiſh the ſpirit, nor contemne the Prophets, yet we muſt
beware we be not deceiued by giuing too light credit to euery one that
vanteth himſelf of the ſpirit, as Arch-heretikes euer did. We muſt trie
them by the doctrine of the Apoſtles and the Spirit of the Catholike
Church, which can not beguile vs.}
proue al things hold that which is good. \V From al appearance of euil
refraine your ſelues.

\V And the God of peace himſelf ſanctifie you in al things:
%%% o-2652
that your whole ſpirit, and ſoule and body without blame may be
preſerued in the comming of our Lord \Sc{Iesvs} Chriſt. \V He is
faithful, that hath called you, who alſo wil doe it. \V Brethren pray
for vs. \V Salute al the Brethren in a holy kiſſe. \V I adiure you by
our Lord that this epiſtle be read to al the holy Brethren. \V The grace
of our Lord
\Sc{Iesvs} Chriſt be with you. Amen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

