%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% One Thessalonians
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 2788
%%% o-2648
\startChapter[
  title={Chapter 3}
  ]

\Summary{Becauſe he could not come himſelf, as he deſired, he ſent
  Timothee. 6.~At whoſe returne now vnderſtanding that they ſtand ſtil
  ſteadfaſt, notwithſtanding al thoſe perſecutions, he reioyceth
  exceedingly: 10.~praying that he may ſee them againe, 12.~and for
  their increaſe in charitie.}

%%% o-2649
For the which cauſe forbearing no longer, it pleaſed vs to remaine at
Athens, alone. \V And we ſent Timothee our Brother, & the Miniſter of
God in the Ghoſpel of Chriſt, to confirme you and exhort you for your
faith, \V that no man be moued in theſe tribulations: for your ſelues
know, that we are appointed to this. \V For euen when we were with you,
we fore-told you that we ſhould ſuffer tribulations, as alſo it is come
to paſſe, and you know. \V Therfore I alſo forbearing no longer, ſent to
know your faith: leſt perhaps he that tempteth, hath tempted you, and
our labour be made vaine. \V But now
\CNote{\XRef{Act.~18,~5.}}
Timothee comming vnto vs from you, and reporting to vs your faith and
charitie, and that you haue a good remembrance of vs alwaies, deſiring
to ſee vs, as we alſo you: \V therfore we are comforted, Brethren, in
you, in al our neceſſitie, & tribulation, by your faith, \V becauſe now
we liue, if you ſtand in our Lord. \V For what thankes-giuing can we
render to God for you, in al ioy wherwith we reioyce for you before our
God, \V night and day more aboundantly praying that we may
\SNote{Though letters or epiſtles in abſence giue great comfort &
confirmation in faith, yet it is preaching in preſence by which the
faith of Chriſt and true religion is alwaies both begun and
accompliſhed.}
ſee your face, and may accompliſh thoſe things that want of your faith?

\V And God himſelf and our Father, & our Lord \Sc{Iesvs} Chriſt direct
our way to you. \V And our Lord multiplie you, & make your charitie
abound one to another, and toward al men: as we alſo in you, \V to
confirme your harts without blame, in holineſſe, before God and our
Father, in the comming of our Lord \Sc{Iesvs} Chriſt with al his
Saints. Amen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

