%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% One Thessalonians
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 2786
%%% o-2646
\startChapter[
  title={Chapter 1}
  ]

\Summary{He thanketh God for them, 4.~and gathereth that they are elect,
  becauſe his preaching at their firſt conuerſion was with diuine power,
  and they on the other ſide receiued it with al ioy, not withſtanding the
  great perſecution that was raiſed againſt them.}

Paul and Syluanus and Timothee to the Church of the Theſſalonians in God
the Father, and our Lord \Sc{Iesvs} Chriſt. Grace to you and peace.

\V We giue thankes to God alwaies for al you; making a memorie of you in
our praiers without intermiſſion, \V mindful of the worke of your faith
and labour, and of the charitie, & of the enduring of the hope of our
Lord \Sc{Iesvs} Chriſt, before God and our Father: \V knowing, Brethren
beloued of God, your election: \V that our Ghoſpel hath not been to you
in word only, but in power & the holy Ghoſt, & in much fulneſſe, as you
know what manner of men we haue been among you for your ſakes. \V And
you became
\LNote{Followers of vs.}{S.~Paul
\MNote{Religious perſons imitation of diuers holy men is the imitation
of Chriſt himſelf.}
is bold to commend them for imitation of him, yea and to ioyne himſelf
in that point with Chriſt, to be their paterne to walke after. Where
without curioſitie he nameth himſelf firſt, and our Lord afterward,
becauſe he was a more neer and ready obiect then Chriſt, who was not nor
could not be followed but through the preaching and conuerſation of the
Apoſtle, who was in their ſight or hearing. And this imitation of ſome
holy man or other, hath made ſo many Religious men of diuers Orders and
Rules, al tending to the better imitation of Chriſt our Lord. See the
like words of the Apoſtle,
\XRef{1.~cor.~11,~1.}
and
\XRef{Philip.~3,~17.}}
followers of vs, & of our Lord; receiuing the word in much tribulation,
with ioy of the Holy Ghoſt: \V ſo that you were made a paterne to al
that beleeue in Macedonia & in Achaia. \V For from you was bruited the
word of our Lord: not only in Macedonia and in Achaia, but alſo in euery
place, your 
faith which is to God-ward, is proceeded, ſo that it is not neceſſarie
for vs to ſpeake any thing. \V For they themſelues report of vs what
manner of entring we had to you; and how you are turned to God
\SNote{In this & the like places the Heretikes maliciouſly and moſt
falſely tranſlate, conſtrue, and apply al things meant of the Heathen
idols, to the memories and images of Chriſt and his Saints, namely the
English Bibles of the yeares
\Cite{1562.}
\Cite{1577.}
See the
\XRef{Annotation 1.~Io.~5,~21.}}
from Idols, to ſerue the liuing and true God, \V and to expect his Sonne
from Heauen (whom
%%% o-2647
he raiſed vp from the dead) \Sc{Iesvs}, who hath deliuered vs from the
wrath to come.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

