%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% One Thessalonians
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 2789
%%% o-2649
\startChapter[
  title={Chapter 4}
  ]

\Summary{He exhorteth them to liue as he taught them: and namely to
  abſteine from al fornication, 9.~to loue one another, 11.~to meddle
  only with their owne matters, 12.~to behaue themſelues wel toward the
  Infidels. 13.~Touching their freinds departed he comforteth them,
  shewing that they shal meet againe at the Reſurrection, and be with
  Chriſt for euer.}

For the reſt therfore, Brethren, we deſire and beſeech you in our Lord
\Sc{Iesvs}, that as you haue receiued of vs how you ought to walke, and
to pleaſe God, as alſo you doe walke, that you abound more. \V For you
know what precepts I haue
%%% o-2650
giuen to you by our Lord \Sc{Iesvs}. \V For this is the wil of God, your
ſanctification: that you abſtaine from fornication, \V that euery one
may know to poſſeſſe his veſſel in ſanctification and honour: \V not in
the paſſion of luſt, as alſo the Gentils that know not God, \V and that
no man ouer-goe, nor circumuent his brother in buſineſſe: becauſe our
Lord is reuenger of al theſe things, as we haue fore-told you, and haue
teſtified. \V For God hath not called vs vnto vncleanneſſe, but into
ſanctification. \V Therfore he that deſpiſeth theſe things, deſpiſeth
\LNote{Not man but God.}{He
\MNote{The precepts of the Church.}
that deſpiſeth the Churches or her lawful Paſtours precepts, offendeth
no leſſe then if he contemned God's expreſſe commandements. For they be
of the Holy Ghoſt, and are not to be counted among the commandements of
men only.}
not man but God, who alſo hath giuen his holy Spirit in vs.

\V But concerning the charitie of the
\SNote{Al Catholike Chriſtians make one Fraternitie or Brotherhood.}
Fraternitie, we haue no need to write to you: For
\CNote{\XRef{Io.~13,25.}
\XRef{15,~17.}
\XRef{Heb.~13,~1.}}
your ſelues haue learned of God to loue one another. \V Yea and you doe
it toward al the Brethren in al Macedonia. But we deſire you, Brethren,
that you
\SNote{Chriſtian men ought to proceed and profit continually in good
workes and iuſtification.}
abound more: \V and that you employ your endeauour to be quiet, and that
you doe your owne buſineſſe, and worke with your owne hands, as we haue
commanded you: \V and that you walke honeſtly toward them that are
without; and need nothing of any man's.

\V And we wil not haue you ignorant, Brethren, concerning them that
\LNote{Sleep.}{Some Heretikes peruerſly inferred of this that the ſoules
did ſleep til the day of iudgement: where it is meant of the bodies
only.}
ſleep, that you be not ſorrowful, as alſo others that haue no hope. \V
For if we beleeue that \Sc{Iesvs} died and roſe againe, ſo alſo God them
that haue ſlept by \Sc{Iesvs} wil bring with him. \V For this we ſay to
you in the word of our Lord,
\CNote{\XRef{1.~Cor.~15,~23.}}
that
\SNote{He ſpeaketh in the perſon of thoſe that ſhal be aliue when our
Sauiour returneth to iudgement.}
we which liue, which are remaining in the aduent of our Lord, shal not
preuent them that haue ſlept. \V For our Lord himſelf in commandement,
and in the voice of an
\CNote{\XRef{Mt.~24,~31.}
\XRef{1.~Cor.~13,~52.}}
Archangel, & in the trompet of God wil deſcend from heauen: and the dead
that are in Chriſt, shal riſe againe firſt. \V Then we that liue, that
are left, withal shal be taken vp with them in the clouds to meet
Chriſt, into the aire, and ſo alwaies we shal be with our Lord. \V
Therfore comfort ye one another in theſe words.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

