%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 1088
%%% o-0982
\startChapter[
  title={Chapter 6}
  ]

\Summary{Iob anſwereth the obiections of Eliphaz, shewing that in dede
  the calamitie which he ſuffereth is much greater then his ſinnes
  deſerue; and therfore his lamentation is excuſable, 8.~wisheth (if it
  ſo pleaſe God) that he may dye, 13.~complaneth that his freindes are
  become his aduerſaries, 16.~grauely expoſtulateth that they reprehend
  him, 22.~and helpe him not.}

But Iob anſwering, ſayd: \V
\SNote{A man of ſincere conſcience confeſſeth the ſinnes wherof he is
guiltie, yet acknowledgeth not al wherwith others may vniuſtly charge
him.}
Would God my ſinnes were weyed, wherby I haue deſerued wrath, and the
calamitie, which I ſuffer in a balance. \V
\LNote{As the ſand of the ſea.}{Scarſe
\MNote{The vſe of \Emph{Hyperbole} in holie Scripture.}
anie figure is more common in holie Scripture
then \Emph{Hyperbole}. Wherby our vnderſtanding is drawne to cõceiue the
greatnes of thinges that otherwiſe ſurpaſſe vulgar capacitie. So Iob
ſignifieth here that \Emph{his calamitie being vveighed vvith his ſinnes in
balance, vvould appeare heauier}, not preciſely in proportion of the
number of ſandes in the ſea, but exceding much in true compariſon.
\MNote{Treſure of ſatisfactorie workes in the Church.}
Of which ſuperabundance of paines patiently ſuffered by holie Iob, and
other Sainctes, more then their ſinnes deſerued, eſpecially of our
B.~Sauiour, who could not ſinne, and of our B.~Ladie, who neuer ſinned,
remayneth an infinite treaſure of ſatisfactorie workes, appliable by the
ſupreme ſtuard of Gods Church, Chriſts Vicar in earth, for ſatisfaction
of others which haue nede, and are in the communion of Sainctes, and
performe the conditions appointed.
\MNote{Pardons do applie the ſuffering of ſome for the ſatisfaction of
others.}
And ſo this high ſtuard may in due maner by his authoritie, geue
limited, or plenarie pardones to penitentes, of the paines which
otherwiſe they ſhould ſuffer in this life, or in purgatorie, for their
ſinnes firſt remitted. Other Biſhops can alſo geue or diſpence ſo much
as the Supreme Biſhop allotteth to their powre, al to edification and
neceſſitie of Gods ſeruantes, as diſpencers, not diſſipators of ſo
holie treaſure.}
As the ſand of the ſea this would appeare heauier, wherfore my wordes
alſo are ful of ſorrow. \V Becauſe the arrowes of our Lord be in me, the
indignation whereof drinketh vp my ſpirit, and the terrours of our Lord
warre agaynſt me. \V Wil the wilde aſſe roare when he hath graſſe? or
the oxe loweth when he shal ſtand before the ful manger? \V Or
can an vnſauerie thing be eaten, that is not ſeaſoned with ſalt? or can a
man taſt that which being taſted bringeth death? \V The thinges which
before my ſoule would not touch, now
%%% o-0983
for anguish are my meates. \V Who wil graunt that my petition may come:
and that God would geue me that which I expect? \V And he that hath
begune, the ſame would looſe his hand, and cut me of? \V And this might
be my comfort, that afflicting me with ſorrow, he
%%% 1089
ſpare not, nor gayneſay the wordes of the Holie one. \V For what is my
ſtrength, that I can ſuſteyne it? or what is mine end, that I should doe
patiently? \V Neither is my ſtrength the ſtrength of ſtones, neither is my flesh of
braſſe. \V Behold there is no help for me in my ſelf, and my familiar
freindes alſo are departed from me. \V He that taketh away mercie from
his frend, forſaketh the feare of our Lord. \V My brethren haue paſſed
by me, as the torrent that paſſeth ſwiftly in the valleys. \V They that
feare the hoarefroſt, ſnow shal fal vpon them. \V At the time, when they
shal be diſſipated they shal perish: and after they waxe hote they shal
be diſſolued out of their place. \V The pathes of their ſteppes are
intangled: they shal walke in vayne, and shal perish. \V Conſider ye the
pathes
\SNote{Sandie wayes without water.}
of Thema, the wayes of Saba, & expect a litle while. \V They are
confounded, becauſe I haue hoped: they are come alſo euen vnto me, and
are couered with shame. \V Now you are come: and euen now ſeing my
plague you are afrayd. \V Haue I ſayd: Bring ye to me, and of your
ſubſtance geue to me? \V Or deliuer me from the hand of the enemie, and
out of the hand of the ſtrong deliuer me? \V
\SNote{Iobs aduerſaries preſumed to teach him, but becauſe they erred in
their applying of true aſſertions in his caſe, which himſelf vnderſtood
and not they, he
\Fix{acconted}{accounted}{obvious typo, fixed in other}
not their diſcourſes for good doctrine.}
Teach ye me, and I wil hold my peace: and if I perhaps haue beene
ignorant in anie thing, inſtruct ye me. \V Why haue you detracted from
the wordes of truth, whereas there is none of you that can controwle
me? \V To rebuke only you frame ſpeaches, and you vtter wordes in the
wind. \V You rush in vpon a pupil, and you endeuoure to ouerthrow your
frend. \V Notwithſtanding accomplish that which you haue begune: geue
eare, and ſee whether I lie. \V Anſwer I beſech you without contention:
and ſpeaking that which is iuſt, iudge ye. \V And you shal not finde
iniquitie in my tongue, neither shal folie ſound in my iawes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
