%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 29
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-29


\project douay-rheims


%%% 1112
%%% o-1005
\startChapter[
  title={Chapter 29}
  ]

\Summary{Againe Iob recounteth Gods former benefites, as wel his grace,
  wherby he did good workes, 5.~as temporal proſperitie, 9.~and wiſdome
  aboue other princes.}

Iob alſo added, taking
\SNote{Parables are not only ſimilitudes of thinges but alſo pithie and
profound ſentences ſuch as Iob, Salomon, and other wiſeſt men vttered.}
his parable, and ſayd: \V Who wil grant me, that I may be according to
the former monethes, according to the daies in which God kept me? \V
When his lampe ſhined ouer my head, & I walked by his light in
darknes? \V As I was in the daies of my youth, when God was ſecretly in
my tabernacle? \V When the Omnipotent was with me: and my ſeruantes
round about me? \V When I waſhed my feete with butter, and the rocke
powred me riuers of oile? \V When I went forth to the
\Fix{gatte}{gate}{likely typo, fixed in other}
of the citie, and in the ſtreate they prepared me a chaire? \V Yong men
ſawe me, and hid them ſelues: and old men riſing vp ſtoode. \V The
princes ceaſed to ſpeake, and did put the finger vpon their mouth. \V
Dukes held in their voice, and their tongue cleaued to their
%%% 1113
throte. \V The eare hearing counted me bleſſed, and the eie ſeing gaue
teſtimonie to me. \V For that I had deliuered the poore man crying out,
and the pupil, that had no helper. \V The bleſſing of him that was
readie to periſh came vpon me, and I conforted the hart of the widow. \V
I was clothed with iuſtice: and I reueſted me with my iudgement, as with
a garment and crowne. \V I was an eie to the blind, and a foote to the
lame. \V I was the father of the poore: and the cauſe which I knew not,
I ſearched moſt diligently. \V I brake the iawes of the wicked man, and
out of his teeth I tooke away the praye. \V And I ſaid: I wil die in my
litle neſt, & as a palmetree wil multiplie daies. \V My roote is opened
beſide the waters, and dewe ſhal continue in my harueſt. \V My glorie
shal alwaies be renewed, and my bow in my hand shal be repayred. \V They
that heard me, expected ſentence, and attent held their peace at my
counſel. \V To my wordes they durſt adde nothing, and my ſpeach
diſtilled vpon them. \V They expected me as rayne, and they opened their
mouth as it were to a lateward shower. \V If at anie time I laughed on
them, they beleued not, and the light of my countenance fel not on the
earth. \V If I would haue gone to them, I ſate firſt, and when I ſate
\SNote{This particle (as) importeth not here a ſimilitude but rather
that he was a very king or ſupreme prince as hauing ſupreme authoritie
\XRef{v.~7.}
royal vertue and crowne.
\XRef{v.~14.}
\Cite{Iſidorus li.~de vita Sanct.~Beda & alij.}}
as a king, with his armie ſtanding about him, yet was I a conforter of
them that mourned.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
