%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 37
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-37


\project douay-rheims


%%% 1122
%%% o-1014
\startChapter[
  title={Chapter 37}
  ]

\Summary{Eliu continueth his diſcourſe, shewing Gods wiſdom, powre, and
  iuſtice, by his meruelous workes of Meteors, 14.~and vſe therof to
  mans commoditie: 18.~which the wiſeſt man ſufficiently vnderſtand not,
  much leſſe may preſume (as he vniuſtly chargeth Iob) to contend with
  God.}

Vpon
\SNote{Conſideration of heauenly rewardes mentioned in the end of the
former chapter, comforteth the afflicted: but thunder and other meteors
being figures of Gods iudgement, ſtrike the hart with terrour.}
this my hart is ſore afrayd, and is moued out of his place. \V Heare ye
his ſpeach in the terrour of his voice, and the ſound proceding out of
his mouth. \V Vnder al the heauens he conſidereth, and his light is vpon
the endes of the earth. \V After him ſhal ſounding roare, he ſhal
thunder with the voice of his greatnes, & ſhal not be found out when his
voice shal be heard. \V God shal thunder in his voice meruelouſly, he
that doeth great & vnſearcheable thinges. \V He that commandeth the ſnow
to deſcend vpon the earth, and the winter raines, and the shower of his
ſtrength. \V He that ſigneth in the hand of al men, that euerie one may
know his workes. \V The beaſt shal enter into his couert, and shal abide
in his denne. \V From the inner partes shal tempeſt come forth, and cold
from
\SNote{North wind, or north pole.}
Arcturus. \V When God bloweth froſt congeleth, and againe waters are
powred moſt largely. \V Corne deſireth clowdes, and the clowdes ſpred
their light. \V Which goe round about, whither ſoeuer the wil of
\SNote{God directeth the cloudes in the ayre as a maſter mariner
gouerneth a ſhippe.}
the gouerner shal lead them, to al that he shal cõmand them vpon the
face of the whole earth. \V Whether in one tribe or in his land, or in
what place ſoeuer of his mercy he shal command them to be found. \V
Harken to theſe things Iob: ſtand, and conſider the maruels of God. \V
Doeſt thou know when God commanded
%%% 1123
the raines, that they shew the light of his clowdes? \V Knoweſt thou the
great pathes of the clowdes, and the perfect knowledges? \V Are not thy
garments hote, when the earth shal be blowen with the South winde? \V
Thou perhaps madſt the heauens with him, which are moſt ſound, caſt as
it were of braſſe. \V Shew vs what we may ſay to him: for we are wrapped
in darkenes. \V Who shal tel him the things that I ſpeake? yea if man
shal ſpeake, he shal be deuoured. \V But now they ſee not the light:
ſodenly the ayre ſhal be thickned into clowdes, and the wind paſſing by
ſhal driue them away. \V From the North gold cometh, & toward God
\SNote{Man not able to praiſe God ſufficiently, prayſeth him with
feare.}
fearful prayſing. \V We can not find him worthely: great of ſtrength,
and
%%% o-1015
iudgement, and iuſtice, and he can not be vttered. \V Therfore shal men
feare him, and al that ſeme to themſelues to be wiſe, shal not dare to
behold
\SNote{Iob conuinced the former three with ſound anſwers, this laſt and
moſt arrogant with ſilence.}
him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
