%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 1105
%%% o-0998
\startChapter[
  title={Chapter 21}
  ]

\Summary{Iob requiring his freindes to heare him, 7.~diſcourſeth of the
  cauſe, why ſome euil men proſper al this life.}

But Iob anſwering, ſayd: \V Heare I beſech you my wordes, and doe
penance. \V Beare with me, that I alſo may ſpeake, and after my wordes,
if it shal ſeme good, laugh ye. \V Is my diſputation
\SNote{Though he diſputed with three men, yet it was concerning diuine
thinges not humaine, but of Gods prouidence, and iuſtice, of the
reſurrection of eternal life and puniſhment.}
agaynſt man, that I ought not worthely to be ſorie? \V Harken to me, and
be aſtonied, and put the finger vpon your mouth. \V And I when I shal
remember, am afrayd, and trembling shaketh my flesh. \V Why then doe the
impious liue, are they aduanced, and ſtrengthened with riches? \V Their
ſeede contineweth before them, a multitude of kinſemen, and of nephewes
in their ſight. \V Their houſes be ſecure and peaceable, & the rod of
God is not vpon them. \V Their bullock hath conceiued, and hath not made
abortion: their cow hath calued, and is not depriued of her calfe. \V
Their litle ones goe forth as flockes, and their infantes reioyſe with
paſtimes. \V They hold the timbrel, & the harpe, & reioyſe at the ſound
of the organe. \V They
\SNote{See
\XRef{ch.~20. v.~5.}}
lead their daies in wealth, and in a moment they goe downe to
\SNote{The ſame word is in Hebrew Greke and Latin for hel as in the
\XRef{7}
and
\XRef{17.~chap.}
& other places, which ſheweth that beſides hel of the dãned the reſting
place alſo of holie fathers in the old Teſtament was called by the
general name of hel.}
hel. \V Who ſayd to God: Depart from vs, we wil not the knowlege of thy
waies. \V Who is the Omnipotent, that we should ſerue him? and what doth
it profite vs if we shal pray him? \V But yet
becauſe their good things are not in their hand, be the counſel of the
impious far from me. \V How often shal the candel of the impious be
extinguished, and inundation come vpon them, and shal he deuide the
ſorowes of his furie? \V They shal be as chaffe before the face of the
winde, and as ashes, which the whirlewinde ſcattereth. \V God shal
reſerue the ſorow of the father to his children: and when he shal haue
rendred it, then shal he know. \V His eies shal ſee his owne ſlaughter,
and he shal drincke the furie of the Omnipotent. \V For what doeth it
pertayne to him concerning his houſe after him: although the number of
his monethes be diminished the halfe? \V Shal anie man teach God
knowledge, who iudgeth the high ones. \V This man dieth ſtrong and in
health, rich and happie. \V His
\Fix{vowels}{bowels}{obvious typo, fixed in other}
be ful of fatte, and his bones be embrewed with marrow. \V But an other
dieth in bitterneſſe of ſoule without anie riches. \V And yet they shal
ſleepe together in the duſt, and wormes shal couer them. \V Surely
%%% 1106
I know your cogitations, and vniuſt ſentences agaynſt me. \V For you
ſay: Where is the houſe of the prince?
%%% o-0999
and where are the tabernacles of the impious? \V Aſke anie of the
wayfaring men, and you ſhal vnderſtand that he knoweth theſe ſelf ſame
thinges. \V Becauſe the euil man is kept vnto the day of perdition, and
he ſhal be led to the day of furie. \V Who ſhal reproue his way before
him? and who shal repay him the thinges that he hath done? \V He ſhal be
brought to the graues, and shal watch in the heade of the dead. \V He
hath beene ſweete to the grauel of
\TNote{A riuer of hel.}
Cocytus, & after him he shal drawe euerie man, and before him
innumerable. \V How therfore doe ye comforth me in vayne, whereas your
anſwer is shewed to be repugnant to the truth?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
