%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 1092
%%% o-0986
\startChapter[
  title={Chapter 9}
  ]

\Summary{Iob approueth that no man auouching his owne iuſtice before God
  is iuſtified. 22.~Teacheth that affliction of the innocent ſtandeth
  wil with Gods iuſtice, wiſdom, and powre.}

And Iob anſwering, ſaid: \V
\SNote{Iob here granteth that which was truly ſaid by his aduerſarie, &
ſheweth how he did wrong applie true doctrin againſt him, and ſo ſtil
defendeth his owne innocencie, and iuſt quarel.}
In deede I know it is ſo, & that man can not be iuſtified compared with
God. \V If he wil contend with him, he can not anſwer him one for a
thouſand. \V He is wiſe of hart, and ſtrong of force: who hath reſiſted
him, & hath had peace? \V He that tranſported mountaynes, and they whom
he ſubuerted in his furie, knew not. \V He that remoueth the earth out
of her place, and the pillers therof are ſhaken. \V He that commandeth
the ſunne, & it riſeth not: and shutteth vp the ſtarres as it were vnder
a ſeale: \V He that alone ſpreadeth the heauens, and goeth vpon the
waues of the ſea. \V He that maketh Arcturus, and Orion, and Hyades, and
the inner partes of the ſouth. \V He that doth great thinges, and
incomprehenſible, and meruelous of the which there is no number. \V If
he come to me, I shal not ſee him: if he depart, I shal not
vnderſtand. \V If ſodenly he aske, who shal anſwer him? or who can ſay:
Why doeſt thou ſo? \V God whoſe wrath no man can reſiſt, and vnder whom
they ſtoope that
\SNote{Angels moue the ſpheres of heauen.}
carie the world. \V How great am I then, that I may anſwer him, and
ſpeake in my wordes with him? \V Who although I haue anie iuſt thing,
wil not anſwer,
\SNote{Euen ſo S.~Paul, though he was not guiltie in conſcience of anie
crime, yet he would not therin iuſtifie himſelf.
\XRef{1.~Cor.~4.}}
but wil beſech my iudge. \V And when he shal heare me inuocating, I doe
not beleue that he hath heard my voice. \V For in a hurle winde shal he
breake me, and shal multiplie my woundes yea without cauſe. \V He
graunteth not my ſpirit to reſt, and he filleth me with bitterneſſe. \V
If ſtrength be demaunded, he is moſt ſtrong: if equitie of iudgement,
\Fix{not}{no}{obvious typo, fixed in other}
man dare geue teſtimonie for me. \V If I wil iuſtifie my ſelf, mine owne
mouth shal condemne me, if I wil shew my ſelf innocent, he shal proue me
wicked. \V Although I shal be ſimple, the ſelf ſame shal my ſoul be
ignorant of, and I shal be wearie of my life. \V
\SNote{This is an aſſured true aſſertion, that God afflicteth both good
& euil in this life.}
One thing there is that I haue ſpoken, both the innocent and the impious
he conſumeth. \V If he ſcourge, let him kil at once, and not laugh at
the paynes
%%% 1093
of innocentes. \V The earth is geuen into the handes of the impious, he
couereth the face of the iudges therof: and if it be not he, who is it
then? \V My dayes haue bene ſwifter then a poſte: they haue fled and
haue not ſene good. \V They haue paſſed by as shippes carying fruites,
as an eagle flying to meate. \V When I shal ſay: I wil not ſpeake ſo, I
change my face, and am tormented with ſorow. \V I
\SNote{Feare on mans part, and hope in God, do wel conſiſt together. So
both preſumption and deſperation are auoided.}
feared al my workes, knowing that thou
%%% o-0987
didſt not ſpare the offender. \V But if ſo alſo I am impious, why haue I
laboured in vayne? \V If I be waſhed as it were with ſnow waters, and my
handes shal ſhine as moſt cleane. \V Yet ſhal thou dippe me in filth,
and my garmentes ſhal abhorre me. \V For neither I wil anſwer a man that
is like my ſelf: nor that may be heard with me equally in iudgement. \V
There is none that may be able to reproue both, and to put his hand
betwen both. \V Let him take his rod from me, and let not his dread
terrifie me. \V I wil ſpeake, and wil not feare him: for I can not
anſwer fearing.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
