%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 1093
%%% o-0987
\startChapter[
  title={Chapter 10}
  ]

\Summary{Iob ſcarſe able to ſpeake yet sheweth that there is no
  iniuſtice nor ignorance in God, neither is his ſinne the cauſe of ſo
  great afflictions. 9.~Acknowledgeth Gods loue and benefites towards
  himſelf, 15.~and dreadeth his ſtrict iudgement.}

My ſoule is wearie of my life, I wil let my ſpeach paſſe agaynſt my
ſelf. I wil ſpeake in the bitternes of my ſoule. \V I wil ſay to God:
Condemne me not: tel me why thou iudgeſt me ſo. \V Doth it ſeeme good to
thee, if thou
\SNote{Holie Iob knowing it to be vnpoſſible that God calumniateth anie
man, inquireth what is the cauſe why his goodnes afflicteth the iuſt?}
calumniate me, and oppreſſe me the worke of thy handes, and helpe the
counſel of the impious? \V Haſt thou eies of flesh: or as a man ſeeth,
ſhalt thou alſo ſee? \V Are thy daies as the daies of man, and are thy
yeares as the times of men? \V That thou ſekeſt my iniquitie, and
ſearcheſt my ſinne? \V And thou mayſt knowe that I haue done no impious
thing, whereas there is no man that can deliuer out of thy hand. \V
\SNote{In that I am thy creature thou maiſt iuſtly deſtroy me: but in
deede becauſe I am thy creature thou wilt ſpare me through thy mercy,
geuing me thy grace which if I vſe wel thou wilt alſo geue me eternal
reſt.}
Thy handes haue made me, and framed me wholly round about, and doſt thou
ſo ſodenlie caſt me downe headlong? \V Remember I beſech thee that as
clay thou madeſt me, and into duſt thou wilt bring me agayne. \V Haſt
thou not as milke milked me, and curded me as cheeſe? \V With skinne and
fleſh thou haſt clothed me. With bones & ſinowes thou haſt compacted
me. \V Life and mercie thou haſt geuen to me, and
%%% 1094
thy viſitation hath kept my ſpirit. \V Although thou conceale theſe
thinges in thy hart, yet I know that thou remembreſt al thinges. \V If I
haue ſinned and thou haſt ſpared me for an houre: why doeſt thou not
ſuffer me to be cleane from mine iniquitie? \V And if I ſhal be impious,
woe is to me: and if iuſt, I shal not lift vp my head, filled with
affliction and miſerie. \V And for pride as a lioneſſe thou wilt take
me, and returning thou doeſt meruelouſly torment me. \V Thou reneweſt
thy witneſſes agaynſt me, and multiplied thy wrath toward me, and paynes
doe warre vpon me. \V Why didſt thou bring
me forth out of the matrice? Who would God, I had beene
%%% o-0988
conſumed that eye might not ſee me. \V I had beene as if I were not,
caried from the wombe to the graue. \V Shal not the fewnes of my daies
be ended ſhortly? Suffer me
\SNote{Repentance is alwayes neceſſarie, but moſt eſpecially at the
houre of death.}
therfore, that I may a litle lament my ſorow. \V Before I goe, and
returne not, vnto the darke land, that is couered with the miſt of
death. \V A land of miſerie and darkeneſſe, where is the ſhadow of
death, and no order, but euerlaſting horrour inhabiteth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
