%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 1085
%%% o-0980
\startChapter[
  title={Chapter 4}
  ]

\Summary{Eliphaz
\MNote{The ſecond part.

Diuers diſcourſes and diſputes about the cauſe of Iobs afflictions.}
blameth Iob as guiltie of impatience, arguing thereupon that he was not
ſo perfect in vertue as he ſemed, 7.~and therfore is now punished by
God, who (as Eliphaz falſly ſuppoſeth) afflicteth not innocent men,
12.~alleaging for proofe an imaginarie viſion.}

%%% 1086
But
\MNote{The firſt conflict, betwen Eliphaz and Iob.}
Eliphaz the Themanite anſwering, ſayd: \V If we shal begin to ſpeake to
thee, perhaps thou wilt take it greuouſly, but the word conceiued who
can hold? \V Behold thou haſt taught manie, & wearie handes thou haſt
ſtrengthned. \V Them that wauered thy wordes haue confirmed, and
trembling knees thou haſt ſtrengthened. \V But now a plague is come vpon
thee, and thou haſt faynted: hath touched thee, and art trubled. \V
Where is thy feare, thy ſtrength, thy patience, and the perfection of
thy wayes? \V Remember I beſech thee, who euer being innocent hath
\SNote{No innocent euer periſhed eternally: but innocent Abel was ſlaine
temporaly, and innumerable others ſuffer calamities for their greater
merite.}
perished? or when haue the iuſt bene deſtroyed? \V Yea rather I haue
ſene them, that worke iniquitie, and ſow ſorrowes, & reape them, \V to
haue perished by the blaſt of God, and with the ſpirit of his wrath to
haue bene conſumed. \V The roaring of the lion, and the voice of the
lioneſſe, & the teeth of the whelpes of lions are bruiſed. \V The tigre
hath perished, becauſe he had no praye, and the lions whelpes are
deſtroyed. \V Moreouer
\SNote{Heretikes pretend ſuch obſcure viſions more to gette credite then
to edifie others.
\Cite{S.~Greg. li.~5 c.~18.}}
to me there was ſpoken a ſecret word, and as it were by ſtealth hath
mine eare receiued the vaines of the whiſpering therof. \V  In the
horrour of a viſion by night, when deepe ſleepe is wont to hold men, \V
feare held me, and trembling, and al my bones were made ſore afrayd. \V
And when the ſpirit paſſed in my preſence, the heares of my flesh ſtood
vpright. \V There ſtood one, whoſe countenance I knew not, an image
before mine eies, and I heard the voyce as it were of a gentle winde. \V
What,
\SNote{Iob eaſely granteth that man may not compare nor contend with
God.
\XRef{ch.~9. v.~1.}
Yet men may be innocent & free from greuous ſinnes.}
shal man be iuſtified in compariſon of God, or shal a man be more pure
then his maker? \V Behold they that ſerue him, are not ſtable, and in
his Angels he
found wickednes? \V How much more they that inhabite houſes of clay,
which haue an earthly foundation, shal be conſumed as it were of the
moth? \V From morning vntil euening they shal be cut downe: and becauſe
none vnderſtandeth, they shal perish foreuer. \V And they that shal be
leaft, shal be taken away from them: they shal die, and not in wiſedom.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
