%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 30
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-30


\project douay-rheims


%%% 1113
%%% o-1006
\startChapter[
  title={Chapter 30}
  ]

\Summary{Holie Iob sheweth the great change of his temporal eſtate, from
  welfare into great calamitie.}

But now they of yonger time ſcorne me, whoſe fathers I vouchſafed not to
put
\SNote{Men ſcarſe fitte to haue care of dogges derided Iob in his
affliction ſo he was contemned of the moſt contemptible.}
with the dogs of my flocke. \V The force of whoſe handes was to me as
nothing, and they were thought vnworthie of life itſelf. \V Barren with
pouertie and famine, who gnawed in the wildernes, il fauoured by
calamitie and miſerie. \V And they did eate graſſe, and the barkes of
trees, and the roote of iunipers was their meat. \V Who taking theſe
thinges violently out of the valies, when they had found euerie thing,
they ranne to them with a crie. \V They dwelt in the deſerts of
torrentes, and in caues of the earth, or vpon grauel. \V Who reioyſed
among theſe kind of thinges, and counted it delicacies to be vnder the
briars. \V The children of fooliſh and baſe men, and in the earth not
appearing at al. \V Now am I turned into their ſong, and
%%% 1114
become a prouerb with them. \V They abhorre me, and flee far from me,
and are not afrayd
\SNote{Our Sauiour alſo ſuffered this reproch.
\XRef{Mat.~26.}}
to ſpit in my face. \V For he hath opened his quiuer, and hath afflicted
me, and
\SNote{In the Hebrew \Emph{hath put}, in the plural number, importing
pluralitie of Diuine Perſons.}
hath put a bridle into my mouth. \V At the right hand of me riſing, my
calamities forthwith aroſe: they haue ouerthrowen my feete, and as with
waues haue oppreſſed with their pathes. \V They haue diſſipated my
waies, they haue lyen in wayte againſt me, and they haue preuailed, and
there was not that would helpe. \V As when a wal is broken, and the gate
opened, they haue broken violently vpon me, and are come trambling downe
to my miſeries. \V
\Fix{I}{I am}{obvious typo, fixed in other}
brought to nothing: as a wind he hath taken away my
\Fix{deſire mand}{deſire: and}{obvious typo, fixed in other}
my proſperitie hath paſſed away as a clowde. \V And now my ſoule
withereth in my ſelf, and the daies of affliction poſſeſſe me. \V In the
night my bone is pearſed with ſorrowes: and they that eate me, ſleepe
not. \V In the multitude of them my garment is conſumed, and they haue
girded me about, as it were with the
\Fix{coler of a wate.}{coller of a coate.}{likely typo, fixed in other}
\V I am compared to durt, and am reſembled to imbers and aſhes. \V I
crie to thee, and thou heareſt me not: I ſtand, and thou doeſt not
reſpect me. \V Thou art changed to be cruel toward me, and in the
hardeneſſe of thy hand thou art againſt me. \V Thou didſt lift me vp,
and ſetting me as it were vpon the wind thou haſt mightely dashed me. \V
I know
\SNote{Death is a comforth to a iuſt man in tribulation.}
that thou wilt deliuer me to death, where a houſe is appointed for euery
one that liueth. \V But yet not to my conſumption doeſt thou ſend forth
thy hand: and if they shal fal, thou wilt ſaue. \V I wept ſometime vpon
him that was afflicted, and my ſoule had compaſſion on the poore. \V I
\Fix{expect}{expected}{likely typo, fixed in other}
%%% o-1007
good thinges, and euils are come vpon me: I taried for light, and
darkeneſſe brake forth. \V My inner partes haue boyled without anie reſt,
the dayes of affliction haue preuented me. \V I went mourning without
furie, riſing vp, I cried in the multitude. \V I was the
\SNote{Like to ſuch beaſtes as ſeke ſolitary places to lament in.}
brother of dragons, and felow of Oſtriches. \V My skinne is made blacke
vpon me, and my bones are dried with heate. \V My harpe is turned into
mourning, and my inſtrument into the voice of weepers.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
