%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 40
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-40


\project douay-rheims


%%% 1124
%%% o-1018
\startChapter[
  title={Chapter 40}
  ]

\Summary{Our Lord further sheweth that mans powre, nor iuſtice, is not
  comparable to Gods, 10.~as appeareth in ouercoming Behemoth, 20.~and
  Leuiathan.}

And our Lord anſwering Iob out of the whirlewinde, ſayd: \V Gird thy
loynes as a man: I wil aske thee, and doe thou tel me. \V Shalt thou
make my
\SNote{Though Iob had truly auouched that his ſinnes deſerued not ſo
great afflictions, yet he ought to haue acknowleged that God afflicted
him iuſtly, for ſome other cauſe knowne to God, but vnknowẽ to him,
which he not confeſſing ſemed to make Gods iudgement vniuſt: or,
\Emph{of none effect}.
\Cite{S.~Greg. li.~32. c.~3.}}
iudgement of none effect: and condemne me, that thou mayſt be
iuſtified? \V And haſt thou an arme as God, and doſt thou thunder with
like voice? \V Put beautie about thee, and ſet vp thy ſelfe aloft, and
be glorious, and put on goodlie garmentes. \V Diſperſe the prowde in thy
furie, and beholding euerie arrogant man, humble him. \V Behold al the
prowde, and confound them, and deſtroy the impious in their place. \V
Hide them in the duſt together, and plunge their faces in the pit. \V
And I wil confeſſe, that thy right hand is able to ſaue thee. \V Behold,
\SNote{An Elephant the greateſt of al beaſtes, of long life, ſtrong,
meke, temperate, chaiſt, ouercome by the Vnicorne, or Dragon, or taken by
the noſe & ledde away. How much more doth Gods prouidence geue man powre
to ouercome the diuel?}
Behemoth whom I made with thee, shal eate hay as it were an oxe. \V His
ſtrength is in his loynes, and his powre in the nauil of his bellie. \V
He gathereth together his taile as the ceder tree, the ſinewes of his
ſtones are perplexe. \V His bones are as pipes of braſſe, his griſtle as
it were plates of iron. \V He is the beginning of the wayes of God,
which made him, he shal applie his ſword. \V To him the mountaines beare
graſſe: al the beaſtes of the filde shal play there. \V He ſleepeth
vnder the shadow, in the ſecrete of the reede, and in moyſt places.
%%% 1125
\V Shadowes doe protect his shadow, the willowes of the torrent shal
compaſſe him. \V Loe, he shal ſup vp the riuer, and shal not meruaile:
and he hath confidence that Iordan may runne into his mouth. \V In his
eies as with a hooke he shal take him, and with ſtakes he shal boare
through his noſthrels. \V Canſt thou drawe out the
\SNote{An huge great fiſh, perhaps, the whale, exceding mans powre to be
managed, yet is ſubiect (as alſo the diuel ſignified therby) to Gods
powre, and prouidence.}
Leuiathan with a hooke, and with a rope shalt thou tye his tongue? \V
Shalt thou put a ring in his noſthrels, or bore through his iaw with a
buckle? \V Wil he multiplie prayers to thee, or ſpeake to thee gentle
wordes? \V Wil he make a couenant with thee, and shalt thou take him to
be a ſeruant for euer? \V Shalt
\Fix{shou}{thou }{obvious typo, fixed in other}
delude him as a birde, or tye him
for thy handmaydes? \V Shal freindes cut him, merchantes diuide him? \V
Shalt thou fil nettes with his skinne, and the cabbin of fishes with his
heade? \V Lay thy hand vpon him: remember battel, and adde to ſpeake no
more. \V Behold his hope shal fruſtrate him, and in the ſight of al he
shal be caſt downe headlong.

 
\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
