%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 1096
%%% o-0990
\startChapter[
  title={Chapter 13}
  ]

\Summary{Of their owne wordes Iob confuteth his aduerſaries, that they
  haue ſpoken that which they know not, 13.~defendeth his owne
  innocencie, 22.~deſiring of God, if he be afflicted for ſecrete
  ſinnes, that he may know them.}

Behold mine eie hath ſeene al theſe thinges, and mine eare hath heard, &
I haue vnderſtood euerie thing. \V According to your knowledge I alſo do
know: neither am I inferiour to you. \V But yet I wil ſpeake to the
Omnipotent, and I couet to diſpute with God. \V Firſt ſhewing you to be
forgers of lying, and mainteyners of peruerſe opinions. \V And would God
ye would hold your peace, that you might be thought to be wiſe men. \V
Heare ye therfore my correptions, and attend the iudgement of my
lippes. \V Hath God neede of your lye, that for him you ſpeake
guiles? \V Doe you take his perſon, and doe you endeuour
\SNote{Iobs ſtate of ſinne or innocencie was beſt knowen to God, next to
his owne conſcience, not at al to his aduerſaries, that preſumed to iudge
therof.}
to iudge for God? \V Or ſhal it pleaſe him from whom nothing can be
concealed, or ſhal he be deceiued as a man, with your fraudulent
dealings? \V He ſhal reproue you, becauſe in ſecrete you take his
perſon. \V Forthwith as he ſhal moue himſelf, he ſhal truble you: and
his terrour ſhal come violently vpon you. \V Your memorie ſhal be
compared to ashes, and your neckes ſhal be brought into clay. \V Hold
your peace a litle while,
%%% 1097
that I may ſpeake, whatſoeuer my minde shal prompt me. \V
\SNote{It ſemed to his aduerſaries that of deſperation he would tear his
flesh, and ſo kil himſelf, and to be ſo nere death as if one held his
ſoule in his hand readie to
\Fix{bet}{let}{obvious typo, fixed in other}
it fal from him.}
Why doe I teare my flesh with my teeth, & carie my ſoule in my
handes? \V Although he shal kil me, I wil
\SNote{He was in extreme anguiſh, but ſtil truſted in God.}
truſt in him: but yet I wil reproue my waies in his ſight. \V And he
ſhal be my ſauiour: for no hypocrite shal come in his ſight. \V Heare ye
my word, and receiue the obſcure ſayings with your eares. \V If I shal
be iudged,
%%% !!! The above SNote is repeated here?
%%% \SNote{}
I know that I shal be found iuſt. \V What is he that wil be iudged with
me? let him come: why am I conſumed holding my peace? \V Two things only
do not to me, and then shal I not be hid from thy face. \V Make thy hand
far from me, and let not thy feare terrifie me. \V Cal me, and I wil
anſwer thee: or els I wil ſpeake, and doe thou anſwer me. \V How great
\Fix{iniquites}{iniquities}{obvious typo, fixed in other}
and ſinnes I haue, my wicked deedes and my offences
\SNote{He denieth that he is guiltie in conſcience but deſireth to know
of God, if he haue anie hid ſinnes which himſelf knoweth not.}
shewe thou me. \V Why hideſt thou thy face, and thinkeſt me thine
enemy? \V Againſt the leafe, that is violently taken with the wind, thou
sheweſt thy might, and perſecuteſt drie ſtuble. \V For thou writeſt
bitternes againſt me, and wilt conſume me with the ſinnes of my
youth. \V Thou haſt put my feete in the ſtockes, and haſt obſerued al my
pathes,
%%% o-0991
and haſt conſidered the ſteppes of my feete. \V Who as rottenes am to be
conſumed, and as a garment, that is eaten of the moth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
