%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 36
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-36


\project douay-rheims


%%% 1121
%%% o-1013
\startChapter[
  title={Chapter 36}
  ]

\Summary{Yet further Eliu sheweth, that God by his powre and wiſdome
  geueth to euerie one that is iuſt. 16.~Inferring (falſly) that al Iobs
  affliction is for his ſinnes, 22.~diſcourſing ſtil of Gods powre,
  wiſdom, and prouidence.}

Eliu alſo adding
\Fix{ſpeake}{ſpake}{likely typo, fixed in other}
theſe wordes: \V Beare with me a litle, and I wil shew to thee: for as
yet I haue what to ſpeake for God. \V I wil repete my knowledge from the
beginning, and I wil proue my maker iuſt. \V For in deede my wordes are
\SNote{No orator is ſo vaine but he promiſeth al truth, & ſpeaketh ſome
to get credite with his auditorie.}
without lye, and perfect knowledge shal be proued to thee. \V God doth
not caſt away the mightie, wheras himſelf alſo is mightie. \V But he
ſaueth not the impious, and he giueth iudgement to the poore. \V He shal
not take away his eyes from the iuſt man, and he placeth kinges in the
throne
\SNote{If kinges reigne wel their praiſe remaineth for euer.
\XRef{v.~11.}}
for euer, and there they are extolled. \V And if they shal be in
cheines, and be bound with the ropes of pouertie: \V He shal shew them
their workes, and their wicked deedes, becauſe they haue bene
violent. \V He alſo shal reuele their eare, to chaſtiſe them: and shal
ſpeake, that they may returne from iniquitie. \V If they shal heare and
obſerue, they shal accomplish their daies in good, and their yeares in
glorie. \V But if they heare not they shal paſſe by the ſworde, and shal
be conſumed in folie. \V Diſſemblers and craftie men prouoke the wrath
of God, neither shal they crie when they are bound. \V Their ſoule shal
dye in tempeſt, & their life among the effeminates. \V He shal deliuer
the poore out of his diſtreſſe, and shal reuele his eare in
tribulation. \V Therfore he shal ſaue thee moſt largely out of the
narrow mouth, and not hauing foundation vnder it: and the quietneſſe of
thy table shal be ful of fatneſſe. \V Thy cauſe is iudged as an impious
mans, cauſe and iudgement thou shalt receiue. \V Let not therfore anger
ouercome thee, that thou oppreſſe anie man: neither let multitude of
giftes incline thee. \V Lay downe thy greatnes without tribulation, and
al the puiſſant of ſtrength. \V
\SNote{The night is drawne long, when tentations are not ſpedely
reſiſted,}
Protract not the night, that
\SNote{wherby ſtil worſe and worſe cogitations ſuccede in place of the
firſt.
\Cite{S.~Greg. li.~26. c.~38.}}
peoples may come vp for them. \V Beware thou decline not to iniquitie:
for thou haſt begunne to folow it after miſerie. \V Behold, God is high
in his ſtrength, and none is like to him
\SNote{Aboue al other Lawgeuers God is moſt able to punish
tranſgreſſors, but moſt willing and moſt able to reward the
obſeruers. S.~Gregorie expoundeth this to be a prophecie of \Emph{Chriſt
our ſingular lavvgeuer}.
\Cite{li.~27. c.~1.}}
among the lawgeuers. \V Who can ſearch his waies? or who can ſay to him:
Thou
%%% 1122
haſt wrought iniquitie? \V Remember that thou knoweſt not his worke,
wherof men haue ſong. \V Al men ſee him, euerie one beholdeth far of. \V
Behold, God is great ſurmounting our knowledge: the number of his yeares
is ineſtimable. \V Who taketh away the droppes of raine, and powreth out
showers as it were gulfes of water. \V Which flow out from the clowdes,
that couer al thinges from aboue. \V If he wil
%%% o-1014
ſtretch forth clowdes as his tent, \V and lighten with his light from
aboue, he shal couer alſo the endes of the ſea. \V For by theſe he
iudgeth peoples, and geueth victuals to manie mortal men. \V In his
handes
\SNote{Chriſt wil geue the glorious light of heauen which now is hidden
to men that loſt terreſtrial paradiſe.
\Cite{ibidẽ. c.~12.}}
he hideth the light, and commandeth it that it come agayne. \V He
ſheweth his freind therof, that it is his poſſeſſion, and that he may
aſcend to it.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
