%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 1102
%%% o-0996
\startChapter[
  title={Chapter 19}
  ]

\Summary{Iob lamenteth of his freindes crueltie, 6.~affirmeth that his
  ſo great affliction is not for his ſinnes, 25.~and comforteth himſelf
  with his vndoubted beliefe of the Reſurrection.}

But Iob anſwering, ſayd: \V How long doe you afflict my ſoule, and weare
me with wordes? \V Behold, ten times you
\Fix{counfound}{confound}{possible typo, fixed in other}
me, and are not ashamed oppreſſing me. \V For although I haue beene
ignorant, mine ignorance shal be with me. \V But you are ſet vp againſt
me, and reproue me with my reproches. \V At the leaſt now vnderſtand
ye, that God hath afflicted me
\SNote{Bleſſed Iob (ſaith S.~Gregorie
\Cite{li.~14. c.~16,~17.)}
looking ſincerly on his owne life, ſaw that his affliction was greater
then his ſinnes deſerued and in that reſpect was not equal: yet it was
iuſt, for God, being iuſt geueth a iuſt reward, \Emph{a crowne of
iuſtice} as S.~Paul ſpeaketh of himſelf.
\XRef{(2.~Tim.~4.)}}
not with equal iudgement, and hath compaſſed me with his ſcourges. \V
Behold I shal crie ſuffering violence, and no man wil heare: I shal crie
out, and there is none to iudge. \V He hath hedged my path round about,
and I can not paſſe, and in my way hath put darkeneſſe. \V He hath
ſpoiled me of my glorie, and hath taken the crowne from my head. \V He
hath deſtroyed me on euerie ſide, and I perish, and as it were from a
tree plucked hath he taken away my hope. \V His
%%% 1103
furie is wrath againſt me, and he hath ſo accounted me as his enemie. \V
His ſeriantes haue come together, and haue made them ſelues a way by me,
and haue beſieged my tabernacle round about. \V He hath made my brethren
far from me, and my acquaintance as ſtrangers haue departed from me. \V
My kinſemen haue forſaken me, and they that knew me haue forgotten
me. \V The gheſtes of my houſe, and my maydſeruantes haue counted me an
aliene, and I haue bene as it were a ſtranger in their eies. \V I called
my ſeruant, and he did not anſwer me, with mine owne mouth I beſought
him. \V My wife hath abhorred my breath, and I prayed the children of my
wombe. \V Fooles alſo deſpiſed me, and when I was departed from them,
they backbited me. \V My counſelers ſometime haue abhorred me: and he
\SNote{Al refuſed holie Iob in his affliction, euen thoſe whom he had
loued moſt: who therfore ought moſt to haue loued him againe.}
whom I loued moſt hath turned againſt me. \V The flesh being conſumed my
bone hath cleaued to my skinne, and there are left onlie lippes about my
teeth. \V Haue mercie vpon me, haue mercie vpon me, at the leaſt you my
frendes, becauſe the hand of our Lord hath touched me. \V Why doe you as
God perſecute me, and are filled with my flesh. \V Who wil grant me that
my wordes may be writen? who wil geue me that they may be drawen in a
booke, \V with yron penne, and in plate of leade, or els with ſtile
might be grauen in flintſtone? \V For
\SNote{An expreſſe profeſſion of his faith of the Reſurrection.}
I know that my Redemer liueth, and in the laſt day I shal riſe out of
the earth. \V And I shal be compaſſed agayne with my skinne, and in my
flesh I shal ſee God. \V Whom I my ſelf shal ſee, and mine eies shal
behold, and
\SNote{VVe ſhal riſe againe, not as one tree riſeth in place of an
other, but the ſelf ſame perſones, changed in qualities, not in
ſubſtance.}
none other: this my hope
%%% o-0997
is layd vp in my boſome. \V Why then doe you ſay now: Let vs perſecute
him, and let vs finde roote of word agaynſt him? \V Flee therfore from
the face of the ſword, becauſe the ſword is the reuenger of iniquities:
and know ye that there is iudgement.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
