%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 35
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-35


\project douay-rheims


%%% 1120
%%% o-1012
\startChapter[
  title={Chapter 35}
  ]

\Summary{Eliu pretending that Iob had ſayde God to be vniuſt, sheweth
  that mans pietie nor impietie neither profiteth nor diſprofiteth God:
  13.~and that he iudgeth al thinges rightly.}

Therfore Eliu againe ſpake theſe wordes: \V Doth thy cogitation ſeme
iuſt to thee, that thou ſaydſt:
\SNote{Eliu inſiſteth much in this calumniation, for Iob neuer ſaid,
that he \Emph{was iuſt compared vvith God}, nor \Emph{iuſter then
God}. But that his affliction was greater then his ſinne.
\XRef{ch.~6. v.~3.}
\XRef{c.~23. v.~7.~&c.}}
I am iuſter then God? \V For thou ſaydſt: That which is right doth not
pleaſe thee: or what wil it profite thee if I ſinne? \V Therfore wil I
anſwer to thy wordes, & to thy frendes with thee. \V
\SNote{Theſe are ſtrong ſentences (ſaith S.~Gregorie) but they agree not
(or be euil applied) to the bleſſed perſon of Iob.
\Cite{li.~26. c.~7.}}
Looke vp to heauen and ſee, and behold the skie, that it is higher then
thou. \V If thou ſinne, what shalt thou hurt him? and if thine
iniquities be multiplied, what shalt thou doe againſt him? \V More ouer
if thou doe iuſtly, what shalt thou geue him, or what shal he receiue of
thy hand? \V Man that is like to thee, thy impietie shal hurt: and thy
iuſtice shal helpe the ſonne of man. \V Becauſe of the multitude of
calumniatours they shal cry: and shal waile for the force of the arme of
tyrantes. \V And he hath not ſayd: Where is God, that made me, that hath
geuen ſongs in the night? \V Who teacheth vs aboue the beaſtes of the
earth, and inſtructeth vs aboue the foules of the ayre. \V There shal
they crie, and he wil not heare, becauſe of the pride of the euil. \V
God therfore wil not heare without cauſe, and the Omnipotent wil behold
the cauſes of euerie one. \V Yea when thou shalt ſay: He conſidereth not:
be iudged before him, & expect him. \V For he doth not now inferre his
furie, neither doth he reuenge wickedneſſe
%%% 1121
excedingly. \V Therfore Iob in vaine openeth his mouth, and without
knowledge multiplieth wordes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
