%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-23


\project douay-rheims


%%% 1107
%%% o-1000
\startChapter[
  title={Chapter 23}
  ]

\Summary{Iob expecteth helpe and ſentence of God, 6.~with iuſt feare,
  yet with good conſcience maintaineth his owne innocencie.}

But Iob anſwering, ſayd: \V
\SNote{VVhiles he expected ſome comforth of his freindes, they ſtil
afflicted him more and more, chargeing him with falſe crimes and ſo
aggrauating his grief both of bodie and mind.}
Now alſo my talke is in bitterneſſe, and the hand of my plague is
aggrauated vpon my mourning. \V Who wil grant me that I may know and
find him and come euen to
\SNote{Therfore he appealeth to Gods iudgement ſeate for ſentence.}
his throne? \V I wil ſet iudgement before him, and wil fil my mouth with
accuſations. \V That I may know the wordes, that he wil anſwer me, and
vnderſtand what he wil ſpeake to me. \V I wil not that he contend with
me with much ſtrength, nor that he oppreſſe me with the weight of his
greatnes. \V Let him propoſe equitie againſt me, and my iudgement shal
come to victorie. \V If I shal goe to the Eaſt, he appeareth not: if to
the Weſt, I shal not vnderſtand him. \V If to the left hand, what shal I
doe? I shal not apprehend him: if I turne myſelf to the right hand, I
shal not ſee him. \V But he knoweth my way, & hath proued
%%% 1108
me as gold that paſſeth through the fyre. \V My foote hath folowed his
ſteppes, I haue kept his way, & haue not declined out
\Fix{if}{of}{obvious typo, same in both}
it. \V From the commandementes of his lippes I haue not departed, and I
haue hid the wordes of his mouth in my boſome. \V For he is alone, and
no man can turne away his cogitation: and whatſoeuer his ſoule would,
that hath he done. \V And when he shal haue fulfilled his wil in me,
manie other things alſo are at hand
\Fix{wit}{with}{obvious typo, fixed in other}
him. \V And therfore I am trubled at his face, and conſidering him I am
made penſife with feare. \V God hath mollified my hart, and the
Omnipotent hath
%%% o-1001
trubled me. \V For I haue not periſhed becauſe of the imminent
darkeneſſe, neither hath the miſt couered my face.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
