%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 1101
%%% o-0994
\startChapter[
  title={Chapter 17}
  ]

\Summary{For the greuouſnes of his paine Iob expecteth ſpeedie death,
  4.~chargeth his freindes with folie for holding only remuneration in
  this life, 6.~himſelf hopeth happie reſt in the other world.}

My ſpirit ſhal be extenuated, my daies ſhal be ſhortened, and the graue
only remaineth for me. \V I haue
\SNote{Not hauing committed ſuch ſinnes as he was charged withal, in
this duble bitternes of corporal paine and calumniation, his eye was
ſtil vpon God, expecting to be deliuered.}
not ſinned, and mine eie abideth in bitterneſſe. \V Deliuer me, and ſet
me beſide thee, and let anie mans hand fight againſt me. \V Thou haſt
made their hart far from diſcipline, and therfore they shal not be
exalted. \V He promiſeth a praye to his felowes, and the eies of his
children ſhal faile. \V He hath ſet me as it were for a prouerbe of the
comon people, and I am an example before them. \V Mine eie is dimne for
indignation, and my members are brought as it were to nothing. \V The
iuſt ſhal be aſtonied vpon this, and the innocent ſhal be raiſed vp
againſt the hypocrite. \V And the iuſt ſhal hold his way, and with
cleane handes ſhal adde ſtrength. \V Be al you therfore conuerted, and
come, and I ſhal not finde among you anie wiſeman. \V My daies haue
paſſed, my cogitations are diſſipated, tormenting my hart. \V Night they
haue turned into day, and againe after darkeneſſe I hope for light. \V
If I ſhal expect,
\SNote{This text ſheweth euidently that there was a place of reſt called
hel.}
\TNote{{\HH{Seol}}

\G{Ἀδὴς}

\L{Infernus}}
hel is my houſe, and in darkeneſſe I haue made my bed. \V I haue
ſayd to rottenes: Thou art my father, my mother, and my ſiſter, to
wormes. \V Where is now then my expectation, and my patience who
conſidereth. \V Al my thinges ſhal deſcend into moſt deepe hel: there at
the leaſt, ſhal I haue reſt thinkeſt thou?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
