%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 1094
%%% o-0988
\startChapter[
  title={Chapter 11}
  ]

\Summary{Sophar
\MNote{The third conflict.}
imputeth Iobs diſcourſe, about the cauſe of his ſo great afflictions, to
inſolencie of mind, and loquaſitie of tongue, perſwading him to
acknowlege greuous ſinnes, that ſo he may haue the reward of a iuſt man.}

But Sophar the Naamathite anſwering, ſaid: \V Why, ſhal he that ſpeaketh
manie thinges, not heare alſo? or
\SNote{Sophar might haue applied the vice of much ſpeaking to himſelf,
and his felowes, alleaging manie thinges, which proued not their opinion,
wheras Iobs allegations proued directly that which he affirmed.}
shal a man ful of wordes be iuſtified? \V To thee onlie shal men hold
their peace? and when thou haſt mocked others, shalt thou be confuted of
none? \V For thou haſt ſayd: My word is pure, and I am cleane in thy
ſight. \V And I would wish that God would ſpeake with thee, and would
open his lippes to thee. \V That he might shew thee the ſecretes of
wiſdom, and that his law is manifold, and thou mighteſt vnderſtand that
thou art exacted much leſſer thinges of him,
\SNote{Iobs owne conſcience affirmed the contrarie.
\XRef{c.~6. v.~3.}}
then thy iniquitie deſerueth. \V Peraduenture thou wilt comprehend the
ſteppes of God, and wil find out the Omnipotent perfectly? \V He is
higher then heauen, and what wilt thou doe: deeper then hel, and how
wilt thou know? \V The meaſure of him is longer then the earth, and
broder then the ſea. \V If he ſhal ouerthrow al things, or ſhal ſtrayten
them into one, who shal ſay againſt him? \V For he knoweth the vanitie
of men, & ſeing
%%% 1095
iniquity doth he not cõſider? \V A vaine man is extolled into pride, and
thinketh him ſelf borne free as a wilde aſſes colt. \V But thou haſt confirmed
thy hart, & haſt ſpred thy handes to him. \V If thou ſhalt take away
from thee the iniquitie
\SNote{He could not iuſtly confeſſe iniquitie which he had not
cõmitted.}
that is in thy hand, and iniuſtice remaine not in thy tabernacle. \V
Then mayſt thou lift vp thy face
\Fix{wthout}{without}{obvious typo, fixed in other}
ſpotte, and thou shalt be ſtable, and ſhalt not feare. \V Thou shalt
alſo forget miſerie, and shalt remember it as waters that are paſſed. \V
And the brightnes as it were of noone daies, shal ariſe to thee at
euening: and when thou shalt think thyſelf conſumed, thou shalt riſe as
the day ſtarre. \V And thou shalt haue confidence, hope being ſet before
thee, and buried thou shalt ſleepe ſecure. \V Thou shalt reſt, and there
shal be none to terrifie thee: and verie manie shal beſech thy face. \V
But the eies of
%%% o-0989
the impious shal decay, and eſcape shal faile them, and their hope the
abomination of the ſoule.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
