%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 34
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-34


\project douay-rheims


%%% 1118
%%% o-1011
\startChapter[
  title={Chapter 34}
  ]

\Summary{Againe Eliu chargeth Iob with blaſphemie, and other crimes,
  10.~sheweth the equitie of Gods iudgement: 20.~and that al thinges are
  ſubiect to Gods powre and knowlege.}

Eliu therfore pronouncing, ſpake theſe wordes alſo: \V Heare ye wiſe men
my wordes, and ye learned harken
%%% 1119
to me: \V For the eare proueth wordes, and the throate diſcerneth meates
by the taſt. \V Let vs chooſe vs iudgement, and among vs let vs ſee what
is the better. \V Becauſe Iob ſayd: I am iuſt, and God hath
\SNote{Iob ſaid not that God \Emph{ſubuerted iudgement} (or wrongfully
iudged) but God \Emph{hath taken avvay my iudgement}.
\XRef{ch.~27. v.~2.}
that is, differred to iudge my cauſe.}
ſubuerted my iudgement. \V For in iudging me there is a lie: mine arrow
is violent without anie ſinne. \V What man is there as is Iob, that
drinketh ſkorning as it were water? \V That goeth with them that worke
iniquitie, and walketh with impious men? \V For he hath ſayd: Man shal
not pleaſe God
\SNote{Neither did Iob ſay this
\XRef{(c.~9. v.~22.)}
but Eliu wreſted his meaning that he might haue ſomething to reprehend.
\Cite{S.~Greg. li.~24. c.~25.}}
although he runne with him. \V Therfore ye diſcrete men heare me, far
from God be impietie, and iniquitie from the Omnipotent. \V For he
wil render a man his worke, and according to the waies of euerie one he
wil recompence them. \V For in verie deed God wil not condemne without
cauſe, neither wil the Omnipotent ſubuert iudgement. \V What other hath
he appointed ouer the earth? or whom hath he ſette ouer the world, which
he made? \V If he direct his hart to him, he shal drawe his ſpirit and
breath vnto him. \V Al flesh shal faile together, and man shal returne
into ashes. \V If then thou haue vnderſtanding, heare that is ſayd, and
harken to the voice of my ſpeach. \V Can he that loueth not iudgement,
be healed? and how doeſt thou ſo far condemne him, that is iuſt? \V That
ſayth to the king, Apoſtata: that calleth dukes impious: \V Who
accepteth not the perſones of princes: nor hath knowen the tyrant, when
he contended againſt the poore man: for al are the worke of his
handes. \V They shal ſodenly die, and at midnight peoples shal be
trubled, and shal paſſe, and take away the violent without hand. \V For
his eies are vpon the waies of men, and he conſidereth al their
ſteppes. \V There is not darkeneſſe, and there is not shadow of death,
that they may be hid there which worke iniquitie. \V For it is no more
in mans powre, to come to God into iudgement. \V He shal deſtroy manie,
& innumerable, & shal make other to ſtand for them. \V For he knoweth
their workes: and therfore he shal bring night, and they shal be
deſtroyed. \V As impious men he hath ſtroken them in the place of them
that ſee. \V Who as it were of purpoſe haue reuolted from him, and would
not vnderſtand al his waies: \V That they cauſed the crie of the needie
man to come to him, and he heard the voice of the poore. \V For he
granting
%%% o-1012
peace, who is there that can condemne? After he ſhal hide his
countenance, who is there that may behold him,
%%% 1120
both vpon nations, and vpon al men? \V Who maketh a man that is
\SNote{Eliu applieth this to Iob, as though he had made falſe ſhew of
vertue which he had not, and that God had ſuffered him hitherto to
reigne, or rather to tyranize for iuſt puniſhment of his peoples
ſinnes. But God at laſt declared that Iob was not ſuch a one.
\XRef{ch.~42. v.~7.}}
an hypocrite reigne for the ſinnes of the people. \V Therfore becauſe I
haue ſpoken to God, thee alſo I wil not prohibite. \V If I haue erred,
teach thou me: if I haue ſpoken iniquitie, I wil adde no more. \V Doth
God require it of thee, becauſe it hath diſpleaſed thee? for thou
beganſt to ſpeake, and not I: but if thou know anie better thing,
ſpeake. \V Let men of vnderſtanding ſpeake to me, and let a wiſeman
heare me. \V But Iob hath ſpoken folishly, and his wordes found not
diſcipline. \V O my father, let Iob be proued euen to the end: ceaſe not
from the man of iniquity. \V Who addeth blaſphemie vpon his ſinnes, let
him be reſtrayned in the meane time among vs: and then let him prouoke
God to iudgement with his ſpeaches.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
