%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 1098
%%% o-0992
\startChapter[
  title={Chapter 15}
  ]

\Summary{Eliphaz
\MNote{The fourt conflict.}
againe chargeth Iob to haue ſpoken preſumptuouſly & blaſphemouſly,
14.~auoucheth that no man is innocent nor iuſt, 20.~deſcribing the
malediction of impious and hypochrites.}

But Eliphaz the Themanite anſwering, ſayd: \V Wil a wiſe man anſwer as
it were ſpeaking into the wind, and fil his ſtomacke with burning? \V
Thou
\SNote{Eliphaz not able to anſwer Iobs ſolide reaſons, raileth againſt
him, as if he were iniurious to God, or taught others to caſt away
feare.}
reproueſt him in wordes, that is not equal to thee, and ſpeakeſt that
which is not expedient for thee. \V As much as is in thee, thou haſt
euacuated feare, and haſt taken away prayers before God. \V For thine
iniquitie hath taught thy mouth, and thou doeſt imitate the tongue of
blaſphemers. \V Thine owne mouth shal condemne thee, and not I: and thy
lippes shal anſwer thee. \V Waſt thou the firſt man borne, and formed
before the litle hilles? \V Haſt thou heard Gods counſel, and shal his
wiſedome be inferiour to thee? \V What doeſt thou know, that we are
ignorant of? what doeſt thou vnderſtand that we know not? \V There are
%%% 1099
both oldmen and ancientes among vs, much elder then thy fathers. \V Is
it a great matter that God should comfort thee: but thy naughtie wordes
hinder it. \V Why doth thy hart eleuate thee, and as thinking great
thinges, haſt thou eſtonied eies. \V Why doeth thy ſpirit ſwel agaynſt God,
to vtter ſuch wordes out of thy mouth. \V What is man, that he should be
\SNote{It is a very rare priuilege to be without ſpot.}
without ſpot, and that the borne of a woman should appeare
\SNote{But ſpot of venial ſinnes may be in a iuſt man.}
iuſt? \V Behold among his ſainctes none is immutable, and the heauens
are not cleane in his ſight. \V How much more is man abominable, and
vnprofitable, who drinketh iniquitie as it were water? \V I wil shew
thee, heare me: that which I haue ſeene I wil tel thee. \V Wiſemen
confeſſe and hide not their fathers. \V To whom onlie the earth was
geuen, and ſtranger hath not paſſed by them. \V The impious is proud al
his daies, and
\SNote{Al theſe miſeries are incident to the wicked, but are falſly
applied to holie Iob, who in dede was iuſt.}
the number of the yeares of his tyrannie is vncertaine. \V The ſound of
terrour is alwaies in his eares: and when there is peace, he alwaies
ſuſpecteth treaſon. \V He beleueth not that he may returne from
darkeneſſe to light, looking round about for the ſword on euerie
ſide. \V When he shal moue himſelf to ſeeke bread, he knoweth that the
day of darkeneſſe is prepared in his hand. \V Tribulation shal terrifie
him, and diſtreſſe shal compaſſe him, as a king that is prepared to
battel. \V For he hath ſtretched his hand againſt God, and is
ſtrengthened againſt the Omnipotent. \V He hath runne againſt him with
necke ſet vp right, and is armed with a fatte necke. \V Fatneſſe hath
couered his face, and from his ſides there hangeth tallow. \V He hath
dwelt in deſolate cities, and in deſert houſes, that are brought into
hillockes.
%%% o-0993
\V He shal not be enriched, neither shal his ſubſtance continew, neither
shal he put his roote in the earth. \V He shal not depart out of
darkenes, the flame shal drie his boughes, and he shal be taken away
with the ſpirit of his owne mouth. \V He shal not beleeue vainely
deceiued with errour, that he may be redemed with anie price. \V Before
his daies be accomplished, he shal perish: and his handes shal
wither. \V His cluſter in the firſt flower shal be hurt as a vine, and
as the oliue tree caſting his flower. \V For the congregation of the
hypocrite is barren, and fire shal deuoure their tabernacles, which
gladly take giftes. \V He conceiued ſorow, and hath brought forth
iniquitie, and his wombe
\Fix{preprareth}{prepareth}{obvious typo, fixed in other}
guiles.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
