%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 1081
%%% o-0975
\startChapter[
  title={Chapter 1}
  ]

\Summary{Holie
\MNote{The firſt part.

The change of Iobs proſperous ſtate into affliction.}
Iob offereth ſacrifice for euerie one of his children, 6.~whoſe good
eſtate Satan enuying, by Gods permiſsion ſpoyleth him of al his goodes,
and children, 20.~for which he being penſiue offendeth not, but thanketh
God for al.}

There was a man in the Land of Hus, named Iob, & that man was
\LNote{Simple, right, and fearing God.}{Moyſes,
\MNote{As wel Iob as other prophetes writ their owne actes and prayſes.}
Samuel, Eſdras, & other Prophetes writ their owne actes among others,
alſo their owne prayſes, ſpeaking of themſelues in the third perſon, &
ſo did holie Iob, humbly, truly, and ſimply, without vayne glorie, or
arrogancie, as S.~Gregorie noteth, ſaying:
\CNote{\Cite{Præfat. in Iob.}}
Bleſſed Iob, inſpired with the Holie Ghoſt, might well write his owne
actes, which were the giftes of ſupernal inſpiration.
\MNote{Good workes are both Gods workes and mens workes.}
God being the principal efficient cauſe of al good mens workes, and men
the ſecondarie cauſe of the ſame.

Moreouer Iobs ſingular patience and other vertues are likewiſe commended
in other holie Scriptures & by ancient Doctores.
\XRef{Tobie.~2.}
God permitted tentation to happen vnto Tobias, that example might be
geuen to his folowers of his patience, \Emph{as of holie Iob}. S.~Iames
\XRef{(c.~5.)}
exhorting to patience, ſayeth:
\MNote{Iobs patience inuincible in al ſortes of affliction.}
\Emph{You haue heard the ſuffering of Iob, and you haue ſene the end of
our Lord.} Tertullian
\Cite{(li. de patientia.)}
admiring Iobs patience exclameth thus: Ô moſt happie man whom neither
the driuing away of his flockes of catle, nor conſuming of the reſt with
fire, nor the loſſe of his children, nor torments of his bodie, could
driue from patience, but he ſtood immoueable in the ſeruice of God, for
example to vs, that we fall not for anie worldlie damage, loſſe of
deareſt friendes or corporal afflictions. And bleſſed be God, by whoſe
bleſſing we may now ſay:
\MNote{English Catholiques loſe their goodes and children, and are
afflicted in bodie.}
Ô happie Engliſh Catholiques, that patiently ſuffer the very ſame kindes
of tribulation, (though not in ſo great a degree) in our time.
S.~Cyprian
\Cite{(li. de bono patientiæ.)}
Iob examined & proued by the vertue of patience was aduanced to the very
height of prayſe, a rich lord in poſſeſſion, & a more rich father in
children, ſudainly was neither lord, nor father: was alſo moſt greuouſly
afflicted in his fleſh: and that no tentation might be wanting the diuel
armed his wife againſt him, yet was he not moued, but by victorious
patience thanked God for al. In like ſorte other holie fathers, for
encoregement and conſolation of the afflicted, write much of holie Iobs
inuincible patience. Aboue al S.~Gregorie, our Apoſtle, dilateth moſt
excellently in thirtie and fiue whole bookes, deſcribing & propoſing his
ſo great vertues to  be imitated by al Chriſtianes, firſt of al
\Cite{(li.~1. c.~5.)}
obſeruing
\MNote{Great commendation to be good amongſt the bad.}
how great a prayſe it is to be good in ſuch a place,
\CNote{\XRef{Philip.~2.}}
\Emph{in the middes of a crooked and peruerſe nation, shining as a light
in the vvorld},
\CNote{\XRef{Apoc.~2.}}
dwelling in the gentiles,
\CNote{Cantic.}
vvhere vvas the ſeate of Satan, a lillie among thornes.}
ſimple, & right, and fearing God, and departing from euil. \V And there
were borne to him ſeuen ſonnes and three daughters. \V And his
poſſeſſion was ſeuen thouſand ſheepe, and three thouſand camels, alſo
fiue hundred yoke of oxen and fiue hundred ſhe aſſes, and a familie
exceding great: and that man was great among al them of the Eaſt. \V And
his ſonnes went, and made a feaſt by houſes, euerie one in his day. And
ſending they called their three ſiſters, to eate and drinke with
them. \V And when the dayes of feaſting had paſſed about in courſe, Iob
ſent to them, and ſanctified them: and riſing vp early
\LNote{Offered holocauſtes for euerie one.}{Albeit
\MNote{Sacrifice being of infinite vertue, yet the value in application
is limited.}
the vertue of Sacrifice as wel of Holocauſt in the old Teſtament, as
eſpecially of Chriſts Sacrifice in the New, is of infinite value in it
ſelfe, yet the application therof to particulare perſons, and purpoſes
is limited, and therfore holie Iob offered not only once for al his
children but manie times, ſeuerally for euerie one. Wherof ſee
\Cite{Cardinal Allen li.~2. de Euchariſt. Sacrific. c.~35.}}
offered holocauſtes for euerie one. For he ſayd: Leſt perhaps my ſonnes
haue ſinned, and haue bleſſed God in their hartes. So did Iob al the
dayes. \V But on a certaine day when the ſonnes of God were come to
aſſiſt before our Lord,
\SNote{Diuels appeare not in Gods ſight, but ſometime in preſence of
Angels which repreſent God.
\Cite{S.~Athan. q.~8. ad Antiochum.}}
Satan alſo was preſent amongſt them. \V To whom our Lord ſayd: From
whence comeſt thou? Who anſwering, ſayd: I haue gone round about the
earth, and walked through it. \V And our Lord ſayd to him: Haſt thou
conſidered my ſeruant Iob, that
%%% o-0976
there is not the like to him in the earth, a man ſimple, and right and
fearing God, and departing from euil? \V To whom Satan anſwering, ſaid:
Why, doth Iob feare God in vayne? \V haſt not thou fenſed him, and his
houſe, and al his ſubſtance round about, bleſſed the workes of his
handes, and his poſſeſſion hath increaſed on the earth? \V But ſtretch
forth thy hand a little, and touch al thinges that he poſſeſſeth,
vnleſſe he bleſſe thee in the face. \V Our Lord therfore ſayd to Satan:
\SNote{God doth not directly ſend euils but permitteth them to happen to
his ſeruantes.}
Behold, al things that he hath, are in thy hand, onlie vpon him extend
not thy hand. And Satan went forth from the
%%% 1082
face of our Lord. \V And when vpon a certeine day his ſonnes and
daughters did eate, and drinke wyne in the houſe of their eldeſt
brother, \V there came a meſſenger to Iob, which ſayd: The oxen
plowghed, and the ſhe aſſes fed beſide them, \V and the Sabeians came in
violently, & haue taken al things, and haue ſtroken the ſeruantes with
the ſword, and I onlie haue eſcaped to tel thee. \V And when he yet
ſpake, an other came, and ſayd: The fire of God fel from heauen, and
ſtriking the sheepe and the ſeruantes hath conſumed them, and I only
haue eſcaped to tel thee. \V But whiles he alſo was yet ſpeaking, there
came an other and ſaid: The Chaldees made three troupes, and haue
inuaded the camels, and taken them, moreouer the ſeruantes alſo they
haue ſtrooken with the ſword, and I alone am fled to tel thee. \V He yet
ſpake, and behold an other came in, and ſaid: Thy ſonnes and daughters
eating, and drinking wine in the houſe of their eldeſt brother, \V
ſodenly a vehement winde came violently from the countrie of the deſert,
and shooke the foure corners of the houſe, which falling oppreſſed thy
children and they are dead, and I alone haue eſcaped to tel thee. \V
Then Iob roſe vp, and rent his garmentes, and with powled head falling
on the ground, adored, \V and ſaid:
\SNote{By this holie example ancient fathers condeme the ſenſles Stoikes
who are neuer moued.
\Cite{S.~Aug. li.~1. c.~9. ciuit.}
S.~Paul alſo reproueth \Emph{men vvithout affection}.
\XRef{Rom.~1. v.~31.}}
Naked came I out of my mothers wombe, and naked shal I returne thither:
Our Lord gaue, and our Lord hath taken away: as it hath pleaſed our
Lord, ſo is it done: the name of our Lord be bleſſed. \V In al theſe
thinges Iob ſinned not with his lippes, neither ſpake he anie foolish
thing againſt God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
