%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 26
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-26


\project douay-rheims


%%% 1109
%%% o-1002
\startChapter[
  title={Chapter 26}
  ]

\Summary{Iob refuteth his aduerſaries needles and common argumentes, by
  more ſound diſcourſing of Gods powre, and wiſdome.}

But Iob anſwering, ſayd: \V
\SNote{Iob sheweth that neither God nedeth mans helpe,}
Whoſe helper art thou? his that is weake? and doeſt thou hold vp the
arme of him, that is not ſtrong? \V To whom haſt thou geuen counſel?
%%% 1110
perhaps to him, that hath not wiſdom, and
\SNote{neither is Baldads prudence able to helpe if there were nede.}
thy prudence haſt thou ſhewed very great. \V Whom wouldeſt thou teach?
not him that made breath. \V Behold the
\SNote{Giantes were not able to wade in Noes floud, but were drowned
with the reſt.}
gyantes grone vnder the waters, and they that dwel with them. \V Hel is
naked before him, and there is no couert to perdition. \V Who ſtretcheth
out the northwind ouer the vacant, and hangeth the earth vpon
nothing. \V Who bindeth the waters in his cloudes, that they breake not
forth together downeward. \V Who holdeth the countenance of his throne,
and ſpreddeth his clowde ouer it. \V He hath made a limite about the
waters, til light & darkenes be ended. \V The pillers of heauen tremble,
and dread at his beck. \V In his ſtrength ſodenly the
%%% o-1003
ſeas are gathered together, and with his wiſdom he ſtroke the proud
man. \V His ſpirite hath adorned the heauens, and his hand being the
midwife, the
\SNote{Not only great thinges before recited but alſo the very leaſt are
made by God, and depend vpon his prouidence.}
winding ſerpent is brought forth. \V Loe, theſe things are ſayd in part
of his waies: and where as we haue heard ſcarce a litle droppe of his
word, who ſhal be able to behold the thunder of his greatnes?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
