%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 31
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-31


\project douay-rheims


%%% 1114
%%% o-1007
\startChapter[
  title={Chapter 31}
  ]

\Summary{Holie Iob reciteth ſincerly his owne vertues, shewing therby
  that he is not punished ſo greuouſly for his ſinnes, but by Gods
  prouidence for ſome other cauſe.}

%%% 1115
I haue
\SNote{Wheras there is a continual warre betwen a chaiſt mind and
rebellious fleſh, holie Iob made this condition of truce betwene theſe
enimies: that his eye ſhould neuer geue occaſion to carnal
concupiſcence.}
made a couenant with mine eyes, that I would not ſo much as
\SNote{By which meanes, he was alſo ſafe from carnal cogitations.
\Cite{S.~Greg. li.~21. c.~2.}}
thinke of a virgin. \V For what part ſhould God from aboue haue in me, &
inheritance the Omnipotent from on high? \V Is there not perdition to
the wicked man, and alienation to them that worke iniuſtice? \V Doth not
he conſider my waies, and number al my ſteppes? \V If I haue walked in
vanitie, and my foote hath haſtened in guile: \V let him wey me in a
iuſt balance, and let God know my ſimplicitie. \V If my ſteppe haue
declined out of the way, and if mine eie hath folowed my hart, and if
ſpotte hath cleaued to my handes: \V let me
\Fix{ſawe,}{ſow,}{obvious typo, fixed in other}
and let an other eate it: and let my progenie be plucked vp by the
rootes. \V If my hart hath bene deceiued vpon a woman, and if I haue
lyene in waite at my freinds doore: \V let my wife be the harlot of an
other man, and let other men lye with her. \V For this is a hainous
thing, and moſt great iniquitie. \V It is a fire deuoring euen to
perdition, and rooting vp al thinges that ſpring. \V If I haue contemned
to abide iudgement with my man ſeruant, and my mayd ſeruant, when they
had anie controuerſie againſt me. \V For what ſhal I doe when God shal
riſe to iudge? and when he shal aske, what shal I anſwer him? \V Did not
he make me in the wombe that made him alſo: and did not one forme me in
the matrice? \V If I haue denied to the poore, that which they would,
and haue made the eyes of the widow to expect: \V if I haue eaten my
morſel alone, and the pulpil hath not eaten therof with me. (\V Becauſe
from mine infancie mercy that growen with me: and from my mothers wombe
it came forth with me.) \V If I haue diſpiſed him that perisheth, for
that he had not clothing, and the poore man without wherwithal to couer
him: \V if his ſides haue not bleſſed me, & he was not warmed with the
fliſes of my ſheepe: \V if I haue lifted vp my hand ouer the pupil, yea
when I ſaw my ſelf in the gate the ſuperior: \V let my ſhoulder fal from
his iuncture, and let my arme
%%% o-1008
with his bones be broken. \V For I haue alwaies feared God as waues
ſwelling vpon me, and his weight I could not beare. \V If I haue thought
gold my ſtrength, and haue ſaid to fine gold: My confidence. \V If I
haue reioyſed vpon my great riches, and becauſe my hand found manie
thinges. \V If I ſaw the ſunne when it ſhined, and the moone going
cleerly: \V and my hart in ſecret reioyſed, and I kiſſed my hand with my
%%% 1116
mouth. \V Which is
\SNote{By this demand he prouoked his aduerſaries to produce what they
could to conuince him of idolatrie or denying God wherwith they
indirectly charged him.}
moſt great iniquitie, and a denial againſt God the moſt high? \V If I
haue beene glad of his fal, that hated me, & haue reioyſed that euil had
found him. \V For I haue not geuen my throte to ſinne, that curſing I
wiſhed his ſoule. \V If the men of my tabernacle haue not ſayd: Who wil
giue of his fleſh that we may be filled? \V The ſtranger taried not
without, my doore was open to the wayfaring man. \V If as man I haue hid
my ſinne, and haue concealed my iniquitie in my boſome. \V If I haue
bene afrayd at a verie great multitude, & the contempt of kinſmen hath
terrified me: and I haue not rather held my peace, & not gone out of the
doore. \V Who wil grant me an hearer, that the Omnipotent would heare my
deſire: and that himſelf that iudgeth would write a booke. \V That I may
carie it on my ſhoulder, and put it about me as a crowne? \V At euerie
ſteppe of mine I wil pronounce it, and as to the prince I wil offer
it. \V If my Land cry againſt me, and with it the furrowes therof
lament: \V If I haue eaten the fruites therof without money, and haue
afflicted the ſoule of the tillers therof. \V For wheate let the bryar
grow to me, and for barlie the thorne.

The
\MNote{Iob diſputed no more with his freindes: but afterward ſubmitted
him ſelf to God, acknowledging ſome vnaduiſed ſpeach.
\XRef{c.~39. v.~37.}
\XRef{c.~42. v.~3.}}
wordes of Iob are ended.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
