%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 1091
%%% o-0985
\startChapter[
  title={Chapter 8}
  ]

\Summary{Baldad
\MNote{The ſecond conflict.}
chargeth Iob to haue ſpoken, vniuſtly exhorting him to
  turne to God, and ſo he shal proſper better then
  heretofore. 13.~Shewing that hypocrites shal not proſper,
  20.~inferreth (falſly) that God afflicteth not the ſincere, nor
  helpeth the malignant.}

But Baldad the Suhite anſwering, ſayd: \V
\SNote{This ſecond diſputer charged Iob to be obſtinate who in dede was
conſtant in a true ſetled iudgement.}
How long wilt thou ſpeake ſuch thinges, and shal the ſpirit of the word
of thy mouth be multiplied? \V Why doth God ſupplant iudgement? or doth
the Omnipotent ſubuert that which is iuſt? \V Although thy children haue
ſinned to him, and he hath leaft them in the hand of their iniquitie: \V
Yet if thou wilt ariſe early to God, and wilt beſech the Omnipotent: \V
If thou wilt walke cleane & vpright, he wil forthwith awake vnto thee,
and wil make the habitation of thy iuſtice peaceable: \V In ſo much,
that if thy former thinges haue bene litle, thy later thinges may be
multiplied excedingly. \V For aske the old generation, and ſearch
diligently the memorie of the fathers. (\V For we are but as yeſterday,
and are ignorant
\SNote{Euen thus heretikes imagine Catholikes to
\Fix{the}{be}{obvious typo, fixed in other}
ignorant, and
therfore fil their mouthes and bookes with thinges that are not denied,
and yet inferre much falſhood ſophiſtically applying one truth againſt
an other being them ſelues ignorãt how to reconcile difficulties.}
that our daies vpon the earth are as a shadow.) \V And they shal teach
thee: they shal ſpeake to thee, and from their hart shal vtter
wordes. \V Can the rush be greene without moyſture? or a feggie place
grow without water? \V When it is yet in his flowre, and is not plucked
with the hand, it
\Fix{witereth}{withereth}{obvious typo, fixed in other}
before al hearbes: \V Euen ſo the waies of al, that forget God, and the
hope of the hypocrite shal perish: \V His folie shal not pleaſe him, and
his confidence as the ſpiders webbe. \V He shal leane vpon his houſe,
and it shal not ſtand: He shal ſtay it vp, and it shal not riſe: \V He
ſemeth moyſtened before the ſunne come, and in his riſing his bloſſome
shal goe forth. \V Vpon a heape of rockes his rootes shal be thicke, and
among ſtones he shal abide. \V If he ſwallow him vp out of his place, he
wil denie him, & wil ſay: I know thee not. \V For this is the ioy
\Fix{of of}{of}{obvious typo, fixed in other}
his way, that others may ſpring againe of the earth. \V God wil not
reiect the ſimple, nor reach his hand to the malignant. \V Vntil thy
mouth be filled with laughter, and
%%% 1092
thy lippes with iubilation. \V They that hate thee, shal be clothed with
confuſion: and the tabernacle of the impious shal not ſtand.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
