%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 1090
%%% o-0984
\startChapter[
  title={Chapter 7}
  ]

\Summary{Iob explicateth diuers calamities of mans life, and namely of
  his owne. 6.~Suppoſing it not likelie that he shal returne to former
  proſperous ſtate, 15.~deſireth to dye.}

The life of man vpon earth is a
\SNote{A ſouldiar muſt be alwayes readie to indure trauel, to be
promptly obedient, content to be beaten by his ſuperior with out al
reſiſtance, vpon paine of his life, he is alwayes ſubiect to cares, and
to danger of death, and therfore muſt euer be readie to dye.}
warfare, & his daies, as the daies of an hyred man. \V As a ſeruant
deſireth the shadow, & as the hired man taryeth for the end of his
worke: \V So I alſo haue had vayne monethes, and haue numbred to myſelf
laborious nightes. \V If I ſleepe, I shal ſay: When shal I ariſe? And
agayne I shal expect the euening, and shal be replenished with ſorowes
euen vntil
\Fix{darkenens.}{darkenes.}{obvious typo, fixed in other}
\V My flesh is clothed with rotteneſſe and filth of duſt, my skinne is
withered, & drawen together. \V My daies haue paſſed more ſwiftly, then
the webbe is cut of the weauer, and are conſumed without anie hope. \V
Remember that my life is a winde, and myne eie shal not returne to ſee
good thinges. \V Neither shal the ſight of man behold me: thine eies
vpon me, and I shal not ſtand. \V As a clowde is conſumed, and paſſeth
away: ſo he that shal deſcend to
\TNote{\HH{Seol}.}
hel shal not aſcend. \V Neither shal he returne anie more into his houſe,
neither shal his place know him anie more. \V Wherfore I alſo wil not
ſpare my mouth, I wil ſpeake in the tribulation of my ſpirit: I wil
talke with the bitterneſſe of my ſoule. \V Why, am I a ſea, or a whale,
that thou haſt compaſſed me with a priſon? \V If I ſay: My litle bed
shal confort me, and I shal be releeued ſpeaking with myſelf in my
couch: \V Thou wilt terrefie me by dreames, and by viſions shake me
with horrour. \V For the which thing my ſoule hath choſen hanging, and
my bones death. \V I haue deſpayred. I shal now liue no longer:
\SNote{Iuſt Iob ſuppoſing he was at the point of death, prayed God to
ſpare or \Emph{ceaſe to} puniſh him more, and to accept of that
affliction which he had already ſuffered. So the Church in behalf of
ſoules departed in ſtate of grace prayeth God to ſpare and ceaſe from
further puniſhing them, and to geue them eternal reſt.}
Spare me, for my daies are nothing. \V What is man that thou magnifieſt
him? or why ſetteſt thou thy hart toward him? \V Thou doeſt viſite him
early, and ſodenly thou proueſt him. \V How long doeſt thou not ſpare
me, nor ſuffer me that I ſwallow my ſpitle? \V I haue ſinned, what shal
I doe to thee ô keeper
%%% 1091
of men? why haſt thou ſette me contrarie to thee, and I am become
burdenous to my ſelf? \V Why doeſt thou not take away my
%%% o-0985
ſinne, and why doeſt thou not take away mine iniquitie? Behold now I
shal ſleepe in the duſt: and if thou ſeeke me in the morning, I shal not
be.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
