%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 1100
%%% o-0993
\startChapter[
  title={Chapter 16}
  ]

\Summary{Iob moued by his importune freindes, 4.~expoſtulateth their
  ſeueritie, 12.~further deſcribeth his afflictions, and appealeth to
  Gods iudgement, that he ſuffereth more then his ſinnes deſerue.}

But Iob anſwering ſayd: \V I haue heard often times ſuch things,
\SNote{True and freindlie comforters ought to heare the afflicted with
patience, and not vnmercifully charge him with crimes which they neither
know, nor his conſcience is guiltie of.}
heauie conforters you are al. \V Shal wordes ful of winde haue an end?
or is anie thing trubleſome to thee, if thou ſpeake? \V I alſo could
ſpeake thinges like to you: and would God your ſoule were for my
ſoule. \V I alſo would confort you with wordes, and would wag my head
vpon you. \V I would ſtrengthen you with my mouth, and would moue my
lippes, as ſparing you. \V But what shal I doe? If I ſpeake, my paine
wil not reſt: and if I hold my peace, it wil not depart from me. \V But
now my ſorow hath oppreſſed me, and al my limmes are brought to
nothing. \V My wrinkles giue teſtimonie againſt me, and
\SNote{A great affliction, when one ful of paine and diſtres is alſo
forced to defend his owne innocencie againſt calumniators.}
a falſe ſpeaker is rayſed vp againſt my face contradicting me. \V He
hath gathered his furie vpon me, and threatening me hath gnaſhed againſt
me with his teeth, mine enemy hath beheld me with terrible eies. \V They
haue opened their mouthes vpon me, and exprobating haue ſtrooken my
cheke, they are filled with my paines. \V God hath ſhut me vp with the
wicked man, and hath deliuered me to the hands of the impious. \V I
ſometime that welthie one ſodenly am broken: he hath held my necke,
broken me, and ſet me to himſelf as it were a marke. \V He hath
compaſſed me with his ſpeares, he hath wounded my loynes, he hath not
ſpared, and hath powred out on the earth my bowels. \V He hath cut me
with wound vpon wound, he hath come violently vpon me as it were a
giant. \V I haue ſowed ſackcloth vpon my ſkinne, and haue couered my
flesh with ashes. \V My face is ſwollen with weeping, and my eyeliddes
are dimme.
%%% o-0994
\V Theſe thinges haue I ſuffred
\SNote{As the aduerſaries ſtil obiect great iniquitie to him ſo he
yeldeth them the ſame true anſwer.}
without the iniquitie of my hand, wheras I had cleane prayers to God. \V
Earth couer not my bloud, neither let my crie find place in thee to be
hid. \V For behold my witneſſe is in heauen, and he that knoweth my
conſcience on high. \V My freindes ful of wordes: mine eie diſtilleth
vnto God. \V And would God a man might ſo be iudged with God, as the
ſonne of man is iudged with his companion. \V For behold the short
yeares paſſe away, and I walke the path, by the which I shal not
returne.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
