%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 1084
%%% o-0979
\startChapter[
  title={Chapter 3}
  ]

\Summary{Iob lamenteth, deſcribing his owne, and the general calamities
  of man, 13.~and shewing how they eſcape manie miſeries which either are
  \Fix{neſter}{neuer}{obvious typo, fixed in other}
  borne or dye preſently after their birth.}

After theſe things Iob opened his mouth, and
\SNote{After ſo long ſilence at laſt Iob expreſſeth his grief before his
freindes to be ſo great, that in reſpect therof he iuſtly curſeth al
euil thinges as not made by God but adioyned to mans life, being the
effectes of ſinne, euen from his conception and birth: wiſhing that what
ſoeuer concurred to his longer afflictions in this life, and hindered
his more ſpedie deliuerie from dangers and calamities had not bene, for
ſo he had benne ſowner at reſt as Gods goodnes ſhould diſpoſe of him. Al
which was a lawful deſire and no ſinne at al. See
\Cite{Pineda in c.~2. Iob.}}
curſed his day, \V and ſpake. \V Perish may the day wherin I was
%%% 1085
borne, and the night wherin it was ſayd: A man is conceiued. \V Be that
day turned into darkeneſſe. God require it not from aboue, and let it
not be lightened with light. \V Let darkenes, and the ſhadow of death
obſcure it, let a miſt poſſeſſe it, and be it wrapped in bitterneſſe. \V
A darkeſome hurle wind poſſeſſe that night, be it not counted in the
daies of the yeare, nor numbred in the monethes. \V Be that night
ſolitarie, nor prayſe worthie. \V Let them curſe it which curſe the day,
which are readie to raiſe vp Leuiathan. \V Let the ſtarres be darkened
with the miſt therof: let it expect light and let it not ſee, neither
the ryſing of the appearing morning. \V Becauſe it ſhut not vp the
doores of the wombe, that bare me, nor tooke away euils from myne
eies. \V Why died I not in the matrice, periſhed not forthwith being
come forth of the wombe? \V Why receiued vpon the knees? why nurced with
the breaſtes? \V For now ſleping I should be quiet, and should reſt in
my ſleepe. \V With kinges and conſules of the earth, which build
themſelues ſolitarie places. \V Or with princes, that poſſeſſe gold, and
replenish their houſes with ſiluer. \V Or as a thing vntimely borne that
is hid I should not be, or as they that being conceiued haue not ſeene the
light. \V There the impious haue ceaſed from tumult, & there the wearied
with ſtrength haue reſted. \V And they ſometime bound together without
griefe, haue not heard the voyce of the exactor. \V Litle and great are
there, and the ſeruant free from his maſter. \V Why is there light geuen
to a miſerable man, and life to them, that are in bitterneſſe of
ſoule? \V That expect death, and it cometh not, as they that dig vp
treaſure. \V And they reioyce excedingly when they haue found the
graue. \V To a man whoſe life is hid, and God hath compaſſed him with
darkenes? \V Before I eate, I ſigh: and as it were ouerflowing waters to
my roaring. \V Becauſe the feare which I feared, hath chanced to me: and
that which I was afrayd of, hath happened. \V Haue I not diſſembled?
haue I not kept ſilence? haue I not beene at eaſe? and indignation is
come vpon me.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
