%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 33
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-33


\project douay-rheims


%%% 1117
%%% o-1009
\startChapter[
  title={Chapter 33}
  ]

\Summary{Eliu endeuoreth to proue by Iobs ſpeach that he is vniuſt:
  13.~arguing that God (by afflicting him) hath
\Fix{alreadly}{already}{obvious typo, fixed in other}
ſo iudged. 23.~But if by an Angels admonition he repent, al shal be
remitted.}

Heere therfore Iob my ſayings, and harken to al my wordes. \V Behold I
haue opened my mouth, let my tongue ſpeake within my iawes. \V My wordes
are of my ſimple hart, and my lippes shal ſpeake a pure ſentence. \V The
Spirit of God made me, and the breath of the Omnipotent gaue me life. \V
If thou canſt, anſwer me, and ſtand againſt my face. \V Behold God hath
made me alſo euen as thee, and of the ſame clay I alſo was formed. \V
But yet let not my
\SNote{Arrogant men imagine their owne conceiptes, & vtterance to be
more meruelous thẽ other mens, not knowing how fooliſh their owne pride
is.
\Cite{S.~Greg. li.~23. c.~16.}}
miracle terrifie thee, and let not my eloquence be burdenous to thee. \V
Thou therfore haſt ſayd in my eares, and I haue heard the voice of thy
wordes: \V I am cleane, and without ſinne: vnſpotted, and there is no
iniquitie in me. \V Becauſe he hath found quarrels in me, therfore hath
he thought me his
%%% o-1010
enemy. \V He hath put my feete in the ſtockes, he hath obſerued al my
%%% 1118
waies. \V This therfore is it, wherein thou art not iuſtified: I wil
anſwer thee, that God is greater then man. \V Doeſt thou contend againſt
him becauſe he hath not anſwered thee to al wordes? \V God
\SNote{It is true that Gods wil once vttered ought to ſuffice al men,
for he anſwereth not to ech one by a particular ſpeach, but by common
ſpeach (or fact) ſatisfieth al mens queſtions. Wherof S.~Gregorie noteth
this general rule, \L{vita præcedentium ſit forma ſequentium.} \Emph{The
life of them that goe before is made a forme} (or rule) \Emph{of them
that folovv.}
\Cite{li.~23. c.~18.~&.~19.}
But Eliu falſely ſuppoſed that God by Iobs affliction had declared that
he was a wicked man. For in dede God declared the contrarie.
\XRef{c.~1. v.~1.~&~8.}
&
\XRef{c.~2. v.~3.}}
ſpeaketh once, & repeateth not the ſelf ſame the ſecond time. \V By a
dreame in a viſion by night, when heauie ſleepe falleth vpon men, and
they ſleepe in their bed. \V Thẽ doth he open the eares of men, &
teaching inſtructeth them with diſcipline, \V that he may turne a man
from theſe things, which he doth, & may deliuer him from pride. \V
Deliuering his ſoule from corruption: and his life, that it paſſe not
vnto the ſword. \V He rebuketh alſo by ſorow in the bed, and he maketh
al his bones to wither. \V Bread is become abominable to him in his
life, and to his ſoule the meate before deſired. \V His flesh ſhal
conſume, and the bones that had beene couered, ſhal be made naked. \V
His ſoule hath approched to corruption, and his life to things cauſing
death. \V If there ſhal be an Angel ſpeaking for him, one of thouſandes,
to declare mans equitie. \V He ſhal haue mercie on him, and ſhal ſay:
Deliuer him, that he deſcend not into corruption: I haue found wherein I
may be propitious to him. \V His fleſh is conſumed with punishments, let
it returne to the daies of his youth. \V He ſhal beſeche God, and he wil
be pacified towards him: and he ſhal ſee his face in iubilation, and he
wil render to a man his iuſtice. \V He ſhal behold men, and ſhal ſay: I
haue ſinned, and in deede I haue offended, and, as I was worthie, I haue
not receiued. \V He hath deliuered his ſoule that it ſhould not goe into
death, but liuing ſhould ſee the light. \V Behold, al theſe things doth
God worke three times in euerie one. \V That he may reclame their ſoules
from corruption, and illuminate them with the light of the liuing. \V
Attend Iob, and heare me: and hold thy peace, whiles I ſpeake. \V But if
thou haſt what to ſpeake, anſwer me, ſpeake: for I would thee to appeare
iuſt. \V And if thou haue not, heare me: hold thy peace, and I wil teach
thee wiſedom.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
