%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 27
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-27


\project douay-rheims


%%% 1110
%%% o-1003
\startChapter[
  title={Chapter 27}
  ]

\Summary{More and more Iob confirmeth his innocencie, auouching that God
  not preſently iudging, 11.~wil in time condemne the wicked.}

Iob alſo added, taking his parable, and ſayd: \V God liueth, who hath
\SNote{God would not as yet haue Iobs cauſe iudged, but reſerued the
ſentence, for his greater trial in patience.}
taken away my iudgement, and the Omnipotent, which hath brought my ſoule
to bitterneſſe. \V That as long as breath remaineth in me, and the
ſpirit of God in my noſthrels, \V my lippes shal not ſpeake iniquitie,
neither shal my tongue meditate
\SNote{It were a lie to acknowlege ſuch ſinnes as he had not committed.}
lying. \V God forbid that I ſhould iudge you to be iuſt: til I faile, I
wil not depart from mine innocencie. \V My iuſtification which I haue
begune to hold, I wil not forſake: for my hart doth not reprehend me in
al my life. \V Let mine enemie be as the impious, and mine aduerſarie as
the wicked one. \V For what is the hypocrites hope if couetouſely he take
by violence, and God deliuer not his ſoule? \V Wil God heare his crie,
when diſtreſſe shal come vpon him? \V Or can he be delighted in the
Omnipotent, and inuocate God at al time? \V I wil teach you by the hand
of God, what the Omnipotent hath, neither wil I hide it. \V Loe, you doe
al knowe, and why ſpeake you vaine thinges without cauſe? \V This is the
portion of an impious man with God, and inheritance of the violent,
which they shal receiue of the
%%% 1111
Omnipotent. \V If his children be multiplied, they shal be in the ſword,
& nephewes shal not be filled with bread. \V They that shal be leaft of
him, shal be buried in death, and his widowes shal not weepe. \V If he
shal heape together ſiluer as earth, and as clay shal prepare
garmentes: \V he shal prepare in deede, but the iuſt man shal be clothed
with them: and the innocent shal diuide the ſiluer. \V He hath built his
houſe as a moth, and as a keeper hath he made a bowre. \V The rich man
\SNote{Some part of Gods iudgement falleth on the wicked in this life,
but eſpecially at their death.
\XRef{Pſal.~75.}}
when he shal ſleepe shal take away nothing with him: he shal open his
eies, & finde nothing. \V Pouertie as water shal apprehend him, in the
night a tempeſt shal oppreſſe him. \V The burning wind shal take him vp,
and cary him away, and as a whirlewinde shal pul him violently out of
his place. \V And he shal caſt vpon him, and shal not ſpare: out of his
hand fleing he shal flee. \V He shal claſpe his handes vpon him, and
shal hiſſe vpon him, beholding his place.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
