%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 1097
%%% o-0991
\startChapter[
  title={Chapter 14}
  ]

\Summary{Againe Iob deſcribeth the miſeries of mans life. 3.~Neuertheles
  Gods great prouidence towards him: 7.~profeſſeth his beleefe of the
  Reſurrection.}

Man borne of
\SNote{Al children taking their ſubſtance from the mother, and hauing
only a temporal life, continually tending to death, can not but be
ſubiect to manie miſeries.}
woman, liuing a short time, is replenished with many
\Fix{mſeiries.}{miſeries.}{obvious typo, fixed in other}
\V Who as
a flowre cometh forth and is deſtroyed, & fleeth as a shadow, & neuer
abideth in the ſame ſtate. \V And doeſt thou counte it a worthy thing to
open thine eies vpon ſuch an one, and to bring him with thee into
iudgement? \V Who can make cleane him that is conceiued of vncleane
ſeede? is it not thou which onlie art? \V The daies of man are short, &
the number of his monethes is with thee, thou haſt appointed his
limittes
%%% !!! This SNote is run into the one above?
\SNote{For time itſelfe though it may ſeme long, is
a limited thing, and is very ſhorte, yea hath no proportion in reſpect
of eternitie, which is infinite: yet God hath care to bring this weake
creature to life eternal.}
which can not be paſſed. \V Depart a litle from him, that he may reſt,
vntil his day wished for, come, euen as the hyred man. \V A tree hath
hope: if it be cut, it waxeth greene againe, and the boughes thereof
ſpring. \V If his roote be old in the earth, and the truncke therof be
dead in the duſt, \V at the ſent of water it shal ſpring, and bring
forth leaues, as when it was
%%% 1098
firſt planted. \V But when man shal be dead, and naked and conſumed,
where is he I pray? \V As if the waters should depart out of the ſea,
and a riuer made emptie should be dried vp. \V So man when he is a
ſleepe shal not riſe agayne, til heauen perish, he shal not awake, nor
riſe vp out of his ſleepe. \V Who wil grant me this, that in hel thou
protect me, and hide me, til thy furie paſſe, and appoynt me a time,
wherin thou wilt remember me? \V Shal man that is dead, thinkeſt thou,
liue agayne? al the daies, in which I am now in warfare, I expect vntil
my change do come. \V Thou shalt cal me, and I shal anſwer thee: to the
worke of thy handes thou shalt reach thy right hand. \V Thou in dede
haſt numbred my ſteppes: but thou wilt ſpare my ſinnes. \V Thou haſt
ſealed my offences as it were in a bag, but haſt cured mine
iniquitie. \V A mountaine falling ſlideth downe, and a rock is remoued
out of his place. \V Waters make ſtones holow, and with inundation the
earth by litle and litle is conſumed: and men therfore thou shalt
deſtroy in like maner. \V Thou haſt ſtrengthened him a litle that he
might paſſe away foreuer: thou shalt chãge his face, and shalt ſend him
forth. \V Whether his children shal be noble, or vnnoble, he
\SNote{The parentes after death are not afflicted with the ſtate of
their children, as they be in this life.}
shal not vnderſtand. \V But yet his flesh whiles he shal liue shal haue
ſorow, & his ſoule shal mourne vpon himſelf.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
