%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 1101
%%% o-0995
\startChapter[
  title={Chapter 18}
  ]

\Summary{Baldad
\MNote{The fifth conflict.}
ſetteth vpon Iob againe, chargeing him with preſent
  impatience, and former impietie, 6.~and that therfore he ſuffereth
  worthie punishment.}

But Baldad the Suhite anſwering, ſayd: \V How long wil
\SNote{Baldad perceiuing Iob to ſpeake confidently as the common doctrin
of manie oppoſeth him ſelf neuertheles againſt him and al that thincke as
he doth, and ſo ſpeaketh as to manie in the plural number, \Emph{vvil ye
ſpeake}, and of himſelfe & his felowes: \Emph{are vve reputed &c.} So
holie Iob a figure of the Church defended the common cauſe, his
aduerſaries a figure of heretikes ſpeaking ſome truthes mixed falſe
thinges therwith.
\Cite{S.~Greg. li.~14. c.~1.}}
ye ſpeake vaunting wordes? vnderſtand ye firſt, and ſo let vs ſpeake. \V
Why are we reputed as beaſtes, and accounted vile before you? \V Which
deſtroyeſt thy ſoule in thy furie, shal the earth be forſaken for thee,
and ſhal rockes be tranſported out of their place? \V Shal not the light
of the impious be extinguiſhed, and the flame of his fire not shine? \V
The light shal be darke in his tabernacle, and the candel that is
%%% 1102
ouer him, shal be extinguished. \V The ſteppes of his power shal be
ſtraytened, and his owne counſel shal caſt him downe headlong. \V For he
hath thruſt his feete into a nette, and walketh in the mashes therof. \V
The ſole of his foote shal be held in a ſnare, and thirſt shal burne
againſt him. \V His ginne is hid in the earth, and his ſpring vpon the
path. \V Feares shal terrifie him on euerie ſide, and his feete shal
entangle him. \V Let his ſtrength be extenuated with famine, and let
hunger inuade his ribbes. \V Let it deuoure the beautie of his skinne,
let death the firſtborne conſume his armes. \V Let his confidence be
plucked away out of his tabernacle, and let deſtruction as a king tread
vpon him. \V Let the companions of him, that is not, dwel in his
tabernacle, let
\Fix{brinſton}{brimſton}{likely typo, fixed in other}
be ſprinkled in his tent. \V Let his rootes be dried downward, and his
harueſt deſtroyed vpward. \V Let the memorie of him perish from the
earth, and let not his name be renowmed in the ſtreates. \V He shal
expel him out of light into darkeneſſe, and shal tranſport him out of
the world. \V His ſeed shal not be, nor his progenie in
his people, nor anie remnantes in his countries. \V In his daie the laſt
shal be aſtonied, and horrour shal inuade the firſt. \V Theſe are then
the tabernacles of the wicked man, and this is the place of him that
knoweth not God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
