%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 41
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-41


\project douay-rheims


%%% 1125
%%% o-1019
\startChapter[
  title={Chapter 41}
  ]

\Summary{Leuiathan is further deſcribed by the peculiar partes of his
  bodie, and terrible compoſition of al his members.}

Not as
\SNote{God ruleth al his creatures, not with crueltie as a
\Fix{tyranne,}{tyrant,}{obvious typo, fixed in other}
but with iuſtice, eaſe, and powre.}
cruel wil I raiſe him: for who can reſiſt my
\Fix{continence?}{countinence?}{possible typo, fixed in other}
\V Who hath geuen me before, that I may render vnto him? Al thinges that
are vnder heauen be myne. \V I wil not ſpare him, and his mightie
wordes, and framed to beſech. \V Who shal reuele the face of his
garment: and who shal enter in to the middes of his mouth? \V Who shal
open the gate of his countenance? dread is round about his teeth. \V His
bodie as shildes that are caſt, compact with skales faſt cleauing
together. \V One is ioyned to an other, and not ſo much as anie ayre
entereth betwen them. \V One shal ſticke to an other, & holding eche
other, they shal not be ſeperated. \V His ſneeſing is as the shining of
fire, & his eies as the twinklings of the morning. \V Out of his mouth
procede lampes, as it were torches of lighted fire. \V Out of his
noſthrels procedeth ſmoke, as it were of a pot heated and boyling. \V
His breath maketh coales to burne, & a flame cometh forth out of his
mouth. \V In his necke ſhal ſtrength abide, & needines goeth before his
face. \V The members of his flesh cleaue together one to an other:
\SNote{God at laſt deſtroyeth him whom man can not ouercome.}
he shal ſend lightnings againſt him, and he shal not be caried to an
other place. \V His hart shal be
%%% 1126
hardened as a ſtone, and ſhal be ſtifly compact as the ſmithes
ſtithie. \V When he shal be taken away, the
\SNote{Angels with reuerent feare doe honour Gods powre. And valient
mariners and other ſoldiars are terrified when they ſee this ſo huge a
fish. Myſtically, Gods preachers and perfecteſt ſeruantes ſhal naturally
feare the terrour of Gods iudgement.}
Angels shal feare, and being feared shal be purged. \V When the ſword
shal apprehend him, neither ſpeare, nor breſtplate shal be able to
abide. \V For he shal eſteme yron as chaffe, and braſſe, as rotten
wood. \V The bowman shal not put him to flight, the ſtones of the ſling,
to him are turned into ſtubble. \V As ſtubble wil he eſteme the hammer,
and he wil laugh him to skorne that shaketh the ſpeare. \V The beames of
the ſunne shal be vnder him, and he shal ſtraw gold vnder him as
durt. \V He shal make the deepe ſea to boyle as a pot, and shal put it
as when ointmentes boyle. \V A path shal shine after him, he shal
eſteme the depth as waxing old. \V There is no power vpon the earth,
that may be compared with him, who is made to feare no man. \V He
ſeeth euerie high thing, he is
\SNote{And the diuel reigneth ouer proude men.
\Cite{S.~Greg. li.~34. c.~4.~&.~17.}}
king ouer al the children of pryde.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
