%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 1095
%%% o-0989
\startChapter[
  title={Chapter 12}
  ]

\Summary{Iob sheweth the knowlege, which his freinds much boaſt of, to
  be the common knowne doctrin of Gods ſeruantes. He more truly, and
  more profoundly diſcourſeth of Gods powre and wiſdome, ſtil defending
  his owne innocencie in reſpect of great ſinnes.}

But Iob anſwering, ſayd: \V Are you then men alone, & shal wiſedome die
with you? \V I alſo haue a hart euen as you, neither am I inferiour to
you: for who is ignorãt of theſe thinges, which you know? \V He that is
mocked of his frend as I, shal inuocate God & he wil heare him. For the
\SNote{God ſuffereth his ſimple true meaning ſeruantes to be ſcorned for the
time, but afterward the wicked shal be forced to confeſſe, that thoſe
whom they derided are worthily in honour before God.
\XRef{Sap.~5. v.~3.}}
ſimplicitie of the iuſt man is ſcorned. \V The lampe contemned in the
cogitations of the riche, is prepared to the time appointed. \V The
tabernacles of robbers abound, & they prouoke God bouldly, wheras he
hath geuen al thinges into their handes. \V For aske the beaſtes, and
they shal teach thee: and the foules of the ayre, and they shal tel
thee. \V Speake to the earth, and it shal anſwer thee, and the fishes of
the ſea shal tel. \V Who is ignorant that the hand of our Lord hath made
al theſe things? \V In whoſe hand is the ſoule of euerie liuing thing,
and the ſpirit of al the flesh of man. \V Doth not the eare diſcerne
wordes, and the iawes of him that eateth, the taſt? \V In the ancientes
is wiſedom, and in long time prudence. \V With him is wiſedom and
ſtrength, he hath counſel and vnderſtanding. \V If he shal deſtroy,
there is no man that can build: if he shut vp a
%%% 1096
man, there is none that can open. \V If he hold in the waters, al things
shal be dried: and if he ſend them forth, they shal ouerthrow the
earth. \V With him is ſtrength and wiſedom: he knoweth both the
deceiuer, & him that is deceiued. \V He bringeth
\SNote{Craftie plotters of deuiſes at laſt by Gods iuſt puniſhement commit
ſo great
\Fix{abſurdites,}{abſurdities,}{obvious typo, fixed in other}
that the meaneſt may ſee their follie.}
counſelers to a folish end, and iudges to aſtonishment. \V He looſeth
the belt of kings, and girdeth their reines with a corde. \V He leadeth
away Prieſtes without glorie, and ſupplanteth the great men. \V Changing
the lippe of the true, and taking away the doctrine of the ancientes. \V
He powreth out contempt vpon princes, releeuing them that had bene
oppreſſed. \V Who reueleth profound things out of darkeneſſe, and
bringeth forth the shadow of death into light. \V Who multiplieth
nations, and deſtroyeth them, and reſtoreth the deſtroyed whole
agayne. \V Who changeth the hart of the princes of the people of the
earth, and deceiueth them, that they may goe in vayne where is no
paſſage. \V They shal grope as in the darke, and not in the light, and
he shal make them goe amis as druncken men.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
