%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-22


\project douay-rheims


%%% 1106
%%% o-0999
\startChapter[
  title={Chapter 22}
  ]

\Summary{Eliphaz
\MNote{The ſeuenth conflict.}
contendeth that God is not pleaſed with a iuſt mans afflictions,
5.~falſly imputeth enormious crimes to holie Iob, 12.~and groſſe
errors. 21.~Wisheth him therfore to repent, that ſo he may proſper.}

But Eliphaz the Themanite anſwering, ſayd: \V Can man be compared with
God, yea though he be of perfect knowlege. \V What doth it
\SNote{In dede whẽ a iuſt man hath donne his dutie he is vnprofitable to
God: but he is profitable to himſelf, which greatly pleaſeth God, who
deſireth mans good, and it redoundeth to Gods glorie that he hath ſuch
ſeruantes.
\XRef{Mat.~5. v.~17.}}
profite God if thou be iuſt? or what doeſt thou aduantage him if thy way
be vnſpotted. \V Shal he be afrayde to reproue thee, and come with thee
into iudgement? \V And not for thy very great malice, and thine infinite
iniquities? \V For thou haſt taken away the pledge of thy brethren
without cauſe, and the naked thou haſt ſpoyled of clothes. \V Water to
the wearie thou haſt not geuen, and from the hungrie thou haſt
withdrawen bread. \V In the ſtrength of thine arme thou didſt poſſeſſe
the earth, and being the mightieſt thou didſt obteyne it. \V Widowes
thou haſt ſent away emptie, and the armes of pupilles thou haſt broken
in peeces. \V Therfore art thou compaſſed with ſnares, and ſoden feare
trubleth thee. \V And thoughteſt thou that thou shouldeſt not ſee
darkenes, and that thou shouldeſt not be oppreſſed with the violence of
ouerflowing waters? \V Doeſt thou not thinke that God is higher then
heauen, & is exalted aboue the toppe of the ſtarres? \V And thou ſayeſt:
For what knoweth God? and he iudgeth as it were by a miſt. \V The
cloudes are his couert,
\SNote{After imputation of falſe crimes, this diſputer chargeth holie
Iob alſo with heathniſh error of the Ægyptians, that God hath no
prouidence of men in this world.
\Cite{Ariſtotle li. de mundo. textu.~84.}
So ſome heretikes in their phrenſie accuſe Catholiques of condemned
hereſies.}
neither doth he conſider our thinges, and he walketh about the poles of
heauen. \V Doeſt thou couet to keepe the path of worldes, which wicked
men haue troden? \V Who
%%% 1107
were taken away before their time, and a floud hath ouerthrowen their
foundation. \V Who ſayd to God: Depart from vs: and as though the
Omnipotent could do nothing, they eſtemed him. \V Whereas he had filled
their houſes with good things, whoſe ſentence be far from me. \V The
iuſt shal ſee, and shal reioyſe, and the innocent man shal ſkorne
them. \V Is
%%% o-1000
not their exaltation cut downe, and hath not fire deuoured the remnantes
of them? \V Agree thou therfore to him, and haue peace: and by theſe
thinges thou shalt haue the beſt fruites. \V Receiue the law of his
mouth, and put his wordes in thy hart. \V If thou wilt returne to the
Omnipotent, thou shalt be builded vp, and ſhalt make iniquitie far from
thy tabernacle. \V He shal geue for earth flint, and for flint torrentes
of gold. \V And the Omnipotent shal be agaynſt thine enemies, & ſiluer
shal be heaped together vnto thee. \V Then shalt thou abound in
delightes vpon the
\Fix{Omnipoten,}{Omnipotent,}{obvious typo, fixed in other}
and shalt lift vp thy face to God. \V Thou shalt aske him, and he wil
heare thee, and thou shal pay thy vowes. \V Thou shalt decree a thing,
and it shal come to thee, and light shal shine in thy waies. \V For who
wil be humbled, shal be in glorie: and he that wil bow downe his eies,
he shal be ſaued. \V The innocent shal be ſaued, but he shal be ſaued in
the cleanneſſe of his handes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
