%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 42
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-42


\project douay-rheims


%%% 1126
%%% o-1020
\startChapter[
  title={Chapter 42}
  ]

\Summary{Holie Iob inſtructed & comforted by Gods diſcourſe,
  acknowlegeth his fault, and craueth pardon for his ouerſight in
  ſpeach, or cogitation. 7.~God iuſtifieth his cauſe againſt his
  freindes. 9.~They offer ſacrifice for their offence. 10.~Al thinges
  proſper with Iob, duble to that he had before, 16.~and he dieth
  happely.}

And Iob anſwering our Lord, ſayd: \V I know that thou canſt doe al
thinges, and no cogitation is hid from thee. \V Who is this, that
concealeth counſel without knowledge? Therfore haue I ſpoken
\SNote{Iob here ſimply
\Fix{ackowledged}{acknowledged}{obvious typo, fixed in other}
his error in ſpeaking ſo much in defence of his owne innocencie, and ſo
litle of Gods prouidence in afflicting him, for his more merite, and
Gods more glorie.}
vnwiſely, and the thinges that did excede my knowledge beyond meſure. \V
Heare (I beſech thee) and I wil ſpeake: I wil aske thee, and do thou tel
me. \V With the hearing of the eare I haue heard thee, but now my eye
ſeeth thee. \V Therfore
\SNote{Before he defended a truth againſt his oppoſite freindes, now
with more reſignation he contented himſelf with his affliction.}
I reprehend myſelf, and I do penance in imbers and ashes. \V And after
our Lord ſpake theſe wordes to Iob, he ſayd to Eliphaz the Themanite: My
furie is wrath againſt thee, and againſt thy two frendes, becauſe
\LNote{You haue not ſpoken right, as my ſeruant Iob.}{Holie
\MNote{God by his ſentence condemned the error of Iobs freindes, and
iuſtified his aſſertion.}
Iob being throughly tried in the fornace of tribulation, and by diuine
inſtruction confirmed in perfect patience, and other vertues, God at
laſt gaue ſentence, condemning the guiltie, and iuſtifying the innocent:
in plaine termes
\Fix{prononcing}{pronouncing}{likely typo, fixed in other}
that Eliphaz, Baldad, and Sophar had not ſpoken right before him, as his
ſeruant Iob. And ſo theſe three being conuinced, that notwithſtanding
their former pretence of defending Gods cauſe, they had erred, and Iob
had maintained the truth, they ſubmitted themſelues as faultie, and
humbly did penance, bringing their oblations for ſacrifice to Iob, as
they were commanded.
\XRef{v.~8.~&.~9.}

As for Eliu the laſt diſputer, perſiſting more vehemẽt in his errour
when others ceaſſed from contention, he was ſufficiently condemned in
his felowes. And the rather for that true point of his owne doctrine
\XRef{(ch.~33. v.~14.)}
\Emph{that God ſpeaketh once, and repeteth not the ſame, the ſecond
time}.
\MNote{Errors ought not to be holden ſtil, being once condemned.}
For it was euer a general rule, that when God once reueiled anie thing
by publique fact, or vnto competent witneſſes, it ſufficed for euer, to
al reaſonable men, and ſo Eliu might applie the ſentence to himſelfe,
which God had denounced to his three freindes, in the ſame cauſe.
\CNote{\Cite{S.~Aug. de fide & oper.}
\Cite{S.~Cyprian. li.~4. ep.~2. }
\Cite{Apud. Euſeb. li.~6. c.~35.}}
\MNote{Much leſſe raiſed againe being hertofore buried.}
Much leſſe are Proteſtantes excuſable, which not only perſiſt in errors
cõdemned in their owne felowes, as in Luther, Caluin, Beza & others, but
alſo in other old hereſies: as their doctrine of iuſtification by only
faith condemned in the Apoſtles time: their denying the cerimonies of
Baptiſme, & the Sacrament of Confirmation: & of Penance, condemned in
the Nouations: or that patient ſuſtaining of worldlie loſſes, and other
afflictions, ſuch as Iob ſuffered, are not ſatisfactorie, or meritorious
workes.}
you haue not ſpoken right before me, as my ſeruant Iob. \V Take therfore
vnto you
\LNote{Take ſeuen oxen, and ſeuen rammes.}{In
\MNote{Numbers myſtical.}
the number of ſeuen is a myſterie of perfection and fulnes, often vſed
in holie Scripture. 
\CNote{\XRef{Gen.~2.}}
For God creating the world and al thinges therin, in ſix dayes, reſted
the ſeuenth.
\CNote{\XRef{4.}}
Seuenfold puniſhment was required for Cain.
\CNote{\XRef{7.}}
Seuen payres of cleane beaſtes were conſerued
\Fix{in in}{in}{obvious typo, fixed in other}
Noes arke.
\CNote{\XRef{Exod.~23.}
\XRef{Deut.~15.}}
And in the law the ground reſted from tillage the ſeuenth yeare, with
innumerable the like, as
\Cite{S.~Gregorie in this place,}
and other Fathers note in their workes.
\MNote{Great or manie ſacrifices for great offences.}
Alſo twiſe ſeuen victimes offered in ſacrifice, import the greatnes of
theſe mens offence in accuſing Iob, and in auouching their falſe
opinion.}
ſeuen oxen, and ſeuen rammes, and goe to my ſeruant Iob, and offer
holocauſte for your ſelues: and my ſeruant
\LNote{Iob shal pray for you.}{Neither
\MNote{Deuotion of him that offereth ſacrifice increaſeth the effect.}
had ſo manie ſacrifices ſufficed, as S.~Chriſoſtom obſerueth,
\Cite{Orat.~5. in Iudæos.}
vnles Iob alſo had prayed for the offenders. VVhere we ſee that both
Sacrifice, & the deuotion of him that offereth it, haue their effectes:
whereof commeth the diſtinction vſed in Scholes, of \L{Opus operatum}
and \L{Opus operantis}.
\MNote{Prayers of holie men or Sainctes derogate not from Chriſt.}
As it is manifeſt likewiſe, that mediation of one man for an other, did
not derogate from Gods benigne mercie, in the law of nature, wherein
theſe men liued. Nor now from Chriſtes mediation, in the law of Grace.
\XRef{2.~Cor.~1. v.~11.}}
Iob shal pray for you: his face I wil receiue, that the follie be not
imputed to you: for you haue not ſpoken right thinges before me, as my
ſeruant Iob. \V Eliphaz therfore the Themanite, and Baldad the Suhite,
and Sophar the
%%% 1127
Naamathite went, and did as our Lord had ſpoken to them, and our Lord
receiued the face of Iob. \V Our Lord alſo was turned at the
\SNote{Iob did penance both for himſelf and others.}
penance of Iob, when he prayed for his freindes. And our Lord added al
thinges whatſoeuer had bene Iobes, duble. \V And al his brethren came to
him, and al his ſiſters, and al that knew him before, and they did eate
bread with him in his houſe: and wagged the head vpon him, and comforted
him vpon al the euil that God had brought in vpon him. And euerie man
gaue him one ewe, and one earelet of gold. \V And our Lord bleſſed the
laſt daies of Iob more then his beginning. And he had fourtene thouſand
sheepe, and ſix thouſand camels, & a thouſand yoke of oxen, and a
thouſand she aſſes. \V And he had
\SNote{In that he had al other thinges duble, and children in the ſame
number as before, it is a ſigne that the former periſhed not but died in
good ſtate.}
ſeuen ſonnes, and three daughters. \V And he called the name of one
Dies, and the name of the ſecond Caſſia, & the name of the third
Cornuſtibij. \V And there were not found in al the earth wemen ſo
beautiful as the daughters of Iob: and their father gaue them
inheritance among their brethren. \V And Iob liued after theſe thinges,
an hundred fourtie yeares, and he ſaw his children, and his childrens
children, vnto the fourth generation, and he died an old man, and ful of
daies.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
