%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 1083
%%% o-0977
\startChapter[
  title={Chapter 2}
  ]

\Summary{Satan by Gods permiſsion, 7.~ſtriketh Iob with ſores from the
  ſole of his foote to the toppe of his head. 9.~His wife alſo inſulteth
  againſt him, but he ſinneth not. 11.~Three freindes coming to viſite
  and conforte him, ſitte ſilent by him ſeuen dayes.}

And it came to paſſe when on a certaine day the ſonnes of God were come,
and ſtood before our Lord, and Satan came among them, and ſtood in his
ſight, \V that our Lord ſayd to Satan: From whence comeſt thou? Who
anſwering ſayd: I haue gone round about the earth, and walked through
it. \V And our Lord ſayd to Satan: Haſt thou conſidered my ſeruant Iob,
that there is not the like to him in the earth, a man
\SNote{A man of plaine and true dealing mixed with meeknes without al
guile is called a ſimple right man.
\Cite{S.~Greg. li.~1. c.~2. &~10.}}
ſimple, and right, and fearing God, and departing from euil, and yet
reteyning innocencie? But
%%% o-0978
thou haſt moued me againſt him, that I should afflict him in vaine. \V
To whom Satan anſwering, ſaid: Skinne for ſkinne, & al thinges that a
man hath, he wil geue for his life: \V otherwiſe put thy hand, and touch
his bone and flesh, and then shalt thou ſee that he wil bleſſe thee in
the face. \V Our Lord therfore ſayd to Satan: Behold he is in thy hand,
but yet ſaue his life. \V Satan
%%% 1084
therfore going out from the face of our Lord, ſtroke Iob with a verie
ſore boile, from the ſole of the foote euen to the toppe of his head:
\V who with a shel ſcraped the corruption, ſitting on a dunghil. \V And
\SNote{His wife perſwading him to deſperation and blaſphemie, ſignified
carnal
\Fix{cogitions}{cogitations}{obvious type, fixed in other}
which corrupt the ſoule inwardly, as afflictions do aſſault the flesh
outwardly.
\Cite{S.~Greg. li.~3. c.~24.}}
his wife ſayd to him: Doeſt thou yet continue in the ſimplicitie? Bleſſe
God and die. \V Who ſayd to her: Thou haſt ſpoken like one of the
foolish wemen, if we haue receiued good things of the hand of God, euil
thinges why should we not receiue? In al theſe things Iob ſinned not
with his lippes. \V Therfore
\LNote{Iobs three freindes.}{For
\MNote{Theſe viſiters of Iob were in dede his freindes, and profeſſed true
religion.}
better intelligence of theſe conflictes betwen holie Iob and his
freindes, it may here be obſerued, that they were in dede his freindes,
as the text ſimply calleth them: that they beleued rightly in God
Almightie, and were not idolaters: that they came of freindly good
affection \Emph{to comforth him though they fel into vvordes of
reprouing him}, (as S.~Gregorie
\CNote{\Cite{li.~3. c.~24.}}
teacheth) they alleaged alſo manie excellent diuine ſentences very
truly, which therfore Iob reproued not.
\MNote{They erred in Iobs particlar caſe.}
But they erred in their ??? againſt Iob: and that of ignorance rather
then of ſette malice, concluding that Iob was guiltie of ſome enormious
ſinnes, becauſe they ſaw him ſo greuouſly puniſhed, and heard him
complaine therof; his owne conſcience knowen to him and hid to them
(wherof they raſhly iudged) teſtifying that he was innocent, in reſpect
of ſo great crimes.
\MNote{And prefigured heretikes.}
And in this their particular error, though they were not heretikes,
being not obſtinate after that the truth was ſufficiently reueled vnto
them, yet they prefigured heretikes, endeuouring by one truth to deſtroy
an other, and by arrogating knowlege which they had not:
\CNote{\Cite{li.~5. c.~18.}}
\Emph{promiſing alſo nevv thinges & vnheard of, rather to dravv others
to admire them, then to edification}.}
Iobs three frendes hearing al the euil, that had
\Fix{changed}{chanced}{obvious type, fixed in other}
to him, came
euerie one out of their place, Elephaz a Themanite, and Baldad a Suhite,
and Sophar a Naamathite. For they had appointed, that coming together
they would viſite him, and confort him. \V And when they had lifted vp
their eies a far of, they knew him not, and crying out they wept, and
renting their clothes ſprinkled duſt ouer their head toward heauen. \V
And they ſate with him on the ground
\SNote{Seuen dayes together euerie day and night ſome good part.}
ſeuen daies and ſeuen nightes, and no man ſpake to him a word: for they
ſaw the payne to be vehement.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
