%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 24
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-24


\project douay-rheims


%%% 1108
%%% o-1001
\startChapter[
  title={Chapter 24}
  ]

\Summary{God in his prouidence knoweth when he wil punish the wicked,
  which his true ſeruantes know not, much leſſe the impious.}

Times are not hid from the Omnipotent: but they
\SNote{Gods ſeruantes know that he wil puniſh wickednes, but know not
when: the wicked preſume that he wil neuer puniſh them.}
that know him, know not his daies. \V Some haue transferred boundes,
ſpoiled flockes, & fed them. \V They haue driuen away the aſſe of
pupilles, & haue taken away the widowes oxe for a pledge. \V They haue
ſubuerted the way of the poore, and haue oppreſſed together the meeke of
the earth. \V Others as wilde aſſes in the deſerte goe forth to their
worke: watching to the praye, doe prepare bread for their children. \V
They reape the filde that is not theirs, and gather the grapes of his
vineyard, whom by violence they haue oppreſſed. \V They ſend men away
naked, taking away their clothes, which haue no couering in the cold. \V
Whom the ſhowers of the moutaynes doe waſh, and not hauing a couert,
they embrace ſtones. \V They did violence ſpoyling the pupilles, and the
common poore people they ſpoyled. \V From the naked and them that goe
without clothing, and the hungrie they haue taken away the eares of
corne. \V They haue reſted the noonetide among their heapes, which
hauing troden the wine preſſes are athirſt. \V Out of the cities they
haue made men to mourne, and the ſoule of the wounded hath cryed, and
God doth not ſuffer it to paſſe vnreuenged. \V They haue bene
\SNote{Heretikes doing and teaching againſt their owne knowlege, are
afterward ſtriken with blindnes, that they can not ſee the truth.
\Cite{S.~Greg/ li.~16. c.~26.}}
rebellious to the light, they haue not knowen his wayes, neither did
they returne by his pathes. \V At the verie breake of day the murderer
ryſeth, he killeth the needie, and the poore
%%% 1109
man: but by night he wil be as a theefe. \V The eie of the aduouterer
obſerueth darkeneſſe, ſaying: Eie ſhal not ſee me: and he wil couer his
face. \V He diggeth through houſes in the darke, as in the day they had
appoynted with them ſelues, and they haue not knowen the light. \V If
ſodenly the morning shal appeare, they thinke it the ſhadow of death:
and they walke ſo in darkeneſſe as it were in light. \V He is light
aboue the face of the water: curſed be his portion in the earth, neither
walke he by the way of the vineyardes. \V Let him paſſe
\SNote{Sinners running into both extremes of defect and exceſſe, are
likewiſe puniſhed with contrarie tormentes.}
from ſnowe waters to exceding heate, and his ſinne euen vnto hel. \V Let
mercie forget him: wormes his ſweetnes: be he not in remembrance, but be
he broken in peeces as an vnfruitful tree. \V For he hath fedd the
barren, and her that bareth not, and to the widow he hath not done
good. \V He hath pulled downe the ſtrong in his ſtrength, and when he
shal ſtand, he wil not credit his life. \V God hath geuen
%%% o-1002
him place for penance, and he
\SNote{Man by powre of free wil oftẽ preſumeth to ſpend the time in
ſinning which God granteth him to do penance, for former ſinnes.
\XRef{Rom.~2.}}
abuſeth it vnto pride: but his eies be vpon his waies. \V They are
eleuated for a litle, and shal not ſtand, and shal be humbled as al
thinges, and shal be taken away, and as the toppes of the eares of corne
they ſhal be broken. \V And if it be not ſo, who can reproue me that I haue
lied, and ſet my wordes before God?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
