%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 32
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-32


\project douay-rheims


%%% 1116
%%% o-1008
\startChapter[
  title={Chapter 32}
  ]

\Summary{Eliu
\MNote{The ninth conflict.}
a young man being angrie that Iob perſiſted in his opinion, and that his
  three freindes could not conuince him, 8.~taketh vpon him to confute
  Iob which they could not do.}

But theſe three men omitted to anſwer Iob, for that he ſeemed iuſt to
himſelf. \V And
\SNote{This youngman wittie & lerned, but proud withal, was a figure of
the hote and arrogant diſputers who wil ſeme to know more then their
elders.
\Cite{S.~Greg. li.~23. c.~2.}}
Eliu the ſonne of Barachel a Buzite, of the kinred of Ram, was angrie and
tooke indignation: and he was angrie againſt Iob, for that he ſayd
himſelf to be iuſt before God. \V Moreouer againſt his freindes he had
indignation, for that they had not found a reaſonable anſwer, but onlie
had condemned Iob. \V Therfore Eliu expected Iob ſpeaking, becauſe they
were his elders that ſpake. \V But when he ſaw that the three were not
able to anſwer, he was wrath excedingly. \V And Eliu the ſonne of
Barachel a Buzite anſwering, ſayd: I am yonger in time, and you more
%%% o-1009
ancient, therfore caſting downe my head, I was afrayd to shewe you my
ſentence. \V For I hoped that longer age would ſpeake, and that a
multitude of yeares would teach wiſdom. \V But as I ſee, there is a
Spirite in men, and the
%%% 1117
inſpiration of the Omnipotent geueth vnderſtanding. \V They of many
yeares are not the wiſe men, neither doe the ancientes vnderſtand
iudgement. \V Therfore wil I ſpeake: Heare ye me, I alſo wil shew you my
wiſedom. \V For I haue expected your wordes, I haue heard your wiſdom,
as long as you contended in wordes. \V And as long as I thought you ſaid
ſomewhat, I conſidered: but as I ſee,
\SNote{A notorious arrogancie to hold himſelfe wiſer then any man, either
of his owne ſect, or of his aduerſaries.}
here is none of you that can reproue Iob, and anſwer his wordes. \V Leſt
perhaps you may ſay: We haue found wiſedom, God hath reiected him, not
man. \V He hath ſpoken nothing
\SNote{Thoſe that neither credite Catholique Doctors, nor yet relie vpon
their owne elders, but euery one vpon his owne priuate ſpirit, denie
that to perteyne to them which is ſpoken to others in the ſame errors.
\Cite{S.~Greg. li.~23. c.~8.}}
to me, and I wil not anſwer him according to your wordes. \V They were
afrayd, and anſwered no more, & they haue taken away talke from
themſelues. \V Therfore becauſe I haue expected, and they ſpake not:
they ſtoode, & anſwered no more. \V I alſo wil anſwer my part, and wil
shew my knowledge. \V For I am ful of wordes, and the ſpirit of my
bellie ſtreyneth me. \V Behold, my bellie is as new wine without a vent,
which breaketh new veſſels. \V I wil ſpeake, and take breath a litle: I
wil open my lippes, and wil anſwer. \V I wil not accept the perſon of a
man, and I wil not make God equal to man. \V For I know not how long I
shal continewe, and whether after a while my maker wil take me away.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
