%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 38
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-38


\project douay-rheims


%%% 1123
%%% o-1015
\startChapter[
  title={Chapter 38}
  ]

\Summary{God
\MNote{The third part.

For the tenth and laſt diſpute God diſcuſſeth the controuerſie, and
geueth ſentence for Iob.}
after terrour of a whirlewind, by way of examining his client Iob of
diuers creatures about their nature, sheweth that no man hath perfect
knowlege of them, much leſſe of Gods immenſitie.}

But our Lord anſwering Iob out of a whirlewind, ſayd: \V Who is this
that wrappeth in ſentences with vnskilful wordes? \V Gird thy loynes as
a man: I wil aske thee, and
\SNote{Onlie God the Creator hath abſolute and perfect knowlege of al
creatures. As may appeare by induction, or example:}
anſwer thou me. \V Where waſt thou when I layd
\SNote{Of the creation of the earth,}
the foundations of the earth? tel me if thou haſt vnderſtanding. \V Who
ſet the meaſures therof, if thou know? or who ſtretched out the line
vpon it? \V Vpon what are the foundations therof grounded? or who let
downe the corner ſtone therof, \V when the morning ſtarres praiſed me
together, and al the ſonnes of God made iubilation? \V Who ſhut in the
\SNote{Of the ſea,}
ſea with doores when it brake forth proceding as it were out of a
matrice? \V When I made a clowde the garment therof, and wrapped it in
darkenes as in cloutes of infancie. \V I compaſſed it with my boundes,
and put barre and doores. \V And I ſayd: Hitherto thou ſhalt come, and
ſhalt not procede farder, & here thou ſhalt breake thy ſwelling
waues. \V Didſt thou after thy birth command
\SNote{Diſtinction of lightes,}
the morning, and ſhew the dawning his place? \V And didſt thou hold the
extremities of the earth ſhaking them, and haſt thou ſhaken the impious
out of it? \V The ſeale shal be reſtored as clay and shal ſtand as a
garment: \V From the impious their light shal be taken away, and the
high arme
%%% !!! Here we are missing a couple of pages. So I follow other for a
%%% bit.
ſhal be broken. \V Haſt thou entred into
\SNote{The depth of the ſea,}
the depths of the ſea, and walked in the loweſt parts of the great
depth? \V Haue the gates of death been opẽ to thee, & haſt thou ſeen the
darkeſome doores? \V Haſt thou cõſidered the bredth of the earth? tel me
if thou know al things, \V in what way the light dwelleth, & what is the
place of darkneſſe. \V That thou cãſt bring euerie thing to his
borders, & vnderſtãd the pathes of the houſe therof. \V
\SNote{Mans owne natiuitie, and what shal happen after his death,}
Didſt thou know then that thou ſhouldeſt be borne? & didſt thou know the
number of thy dayes? \V Haſt thou entred into the treaſures of the ſnow,
or haſt thou beheld treaſures of haile? \V Which I haue prepared for the
time of the enemie, for the day of fight & battel? \V What way is the
\SNote{The ſunnes light and heat,}
light ſpred, is heate diuided vpõ the earth? \V Who
\SNote{Diuers Meteors.}
gaue courſe to the moſt vehemẽt ſhower,
%%% o-1016
and the way of the ſounding thunder: \V That it ſhould rayne vpon the
earth, without man in the deſert, where no mortal man abideth: \V That
it ſhould fil the deſert and deſolate ground, and ſhould bring forth
greene graſſe? \V Who is the father of rayne? or who begot the drops of
dew? \V Out of whoſe womb came forth yce? and froſt from heauen who
ingendred? \V Waters are hardned like ſtone, and the face of the depth
is congealed. \V Shalt thou be able to ioyne together the ſhining
\SNote{Starres, and planets.}
ſtarres Pleiades, or canſt thou diſſipate the circuite of Arcturus? \V
Doeſt thou bring forth the day ſtarre in his time, & make the euening
ſtarre to riſe vpon the childrẽ of the earth? \V Doeſt thou know the
order of heauen, and ſhalt thou put downe the reaſon therof on the
earth? \V Shalt thou eleuate the voice in the cloud, & the violence of
the waters couer thee? \V Shalt thou ſend lightnings, and wil they goe,
and returning ſhal they ſay to thee: Here we are? \V Who put
\SNote{Man hath his knowledge from God,}
wiſedom in the hart of man? or who gaue the
\SNote{And the cock hath skil which man wanteth.}
cocke vnderſtãding? \V Who ſhal declare the manner of the heauens, and
the harmonie of heauen who ſhal make to
\TNote{\Emph{or reſt from motion.}}
ſleep? \V When was the duſt powred on the earth, and the clods cõpact
together? \V
\SNote{In ſome Editions the 39.~chap. beginneth here, ſhewing by
induction of ſenſible things (as before of inſenſible) that only God
knoweth the nature of al creatures.}
Shalt thou take a prey for the lioneſſe and fil the appetite of her
whelps? \V When they lie in the dennes, & in holes ſit in wayte? \V
Who prepareth for the rauen her meate, when her yong ones crie to God,
wandring about, becauſe they haue not meate?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
