%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 28
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-28


\project douay-rheims


%%% 1111
%%% o-1004
\startChapter[
  title={Chapter 28}
  ]

\Summary{The maruelous workes of God, the author of nature, shew his
  powre and wiſdome, 12.~and that true riches conſiſt not in temporal
  creatures, but in wiſdome, 28.~and feare of God.}

Siluer hath
\SNote{By theſe more precious and rare creatures, men ought to conſider
the creator and ſo not ſet their reſt in them, but in him, which is true
wiſdome.}
beginnings of her vaines, and gold hath a place, where in it is
molten. \V Yron is taken out of the earth, and ſtone reſolued with heate,
is turned into braſſe. \V He hath ſet a time for darkeneſſe, and the end
of al thinges he conſidereth, the ſtone alſo of darkeneſſe, and the
shadow of death. \V The
\SNote{Sudaine headie waters burſting out do change the wayes and
paſſages of men.}
torrent diuideth from the pilgrime people, them whom the foote of the
needie man hath forgotten, and to whom there is no way. \V The land, out
of which bread grew in his place, is deſtroyed with fire. \V The place
of the ſapphire the ſtones thereof, and the cloddes of it gold. \V The
bird hath not knowen the path, neither hath the eie of the vulter beheld
it. \V The children of merchantes haue not troden it, neither hath the
lioneſſe paſſed by it. \V He hath ſtretched forth his hand to the flint,
he hath ouerthrowen mountaines from the rootes. \V In the rockes he hath
cut out riuers, and his eie hath ſeene euerie pretious thing. \V The
depthes alſo of riuers he hath ſearched, & hid things he hath brought
forth to light. \V But where is wiſdome to be found, and what is the
place of vnderſtanding? \V Man knoweth not the price therof,
\SNote{True wiſdom is not in natural thinges, but in ſupernatural
vertues.}
neither is it found in the land of them
%%% 1112
that liue pleaſently. \V The depth ſayth: It is not in me. And the ſea
ſpeaketh: It is not with me. \V The fineſt gold ſhal not be geuen for
it, neither ſhal ſiluer be weyed in the change therof. \V It ſhal not be
compared with the died colours of India, nor with Sardonyx the moſt
pretious ſtone, or with the Sapphire. \V Gold or glaſſe ſhal not be
equal to it, neither ſhal veſſels of gold be changed for it. \V High and
eminent thinges shal not be mentioned in compariſon of it: and wiſedom
is drawen out of ſecrete places. \V The topazius of Æthiopia shal not be
equal to it, neither shal it be compared to the cleaneſt
\Fix{diyng.}{dying.}{obvious typo, fixed in other}
\V From whence then cometh wiſedom? and what is the place of
vnderſtanding? \V It is hid from the eies of al liuing, the foules of
the ayre alſo know it not. \V Perdition and death haue ſayd: With our
eares we haue heard the fame therof. \V God vnderſtandeth the way of it,
and he knoweth the place therof. \V For he beholdeth the endes of the
world: & looketh on al thinges that are vnder heauen. \V Who made a
poiſe to the windes, and weyed the waters in a meaſure. \V When he gaue
a lawe to the raynes, and a way to the
%%% o-1005
ſounding ſtormes. \V Then he ſaw it, and declared, and prepared and
ſearched it. \V And he ſayd to man: Behold
\SNote{VVhen man hath conſidered Gods workes, his dutie is to feare God,}
the feare of our Lord, that is wiſedom: and to
\SNote{then to depart from euil and do good.}
depart from euil, vnderſtanding.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
