%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Job
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 1103
%%% o-0997
\startChapter[
  title={Chapter 20}
  ]

\Summary{Sophar
\MNote{The ſixth conflict.}
auoucheth that ſome wicked men florish for a time, but are afterwards
iuſtly plaged. 19.~Therupon condemneth Iob as an hypocrite.}

But Sophar the Naamathite anſwering, ſayd: \V Therfore my diuerſe
cogitations ſuccede one an other, and my minde is rapt into ſundrie
things. \V The doctrine, wherwith thou reproueſt me, I wil heare, and
the ſpirit of my vnderſtanding shal anſwer me. \V This I know from the
beginning,
%%% 1104
ſince man was placed vpon the earth. \V That the praiſe of the impious
is
\SNote{Al mans life is ſhort and as a moment in compariſon of eternitie;
but it is not alwayes true that the
\Fix{wiked}{wicked}{obvious typo, fixed in other}
are ſhortly puniſhed in reſpect of this life, as this diſputer applieth
it to proue Iob to
\Fix{impious.}{be impious.}{obvious typo, fixed in other}
Therfore Iob anſwereth in the
\XRef{next chapter v.~13.}
That ordinarily the wicked, \Emph{lead their dayes} (to witt their whole
life) \Emph{in vvealth and} (then) \Emph{in a moment goe downe to hel.} So
the reſt of theſe mens aſſertions are for moſt part true in ſome ſenſe
but il applied.}
short, and the ioy of the hypocrite as it were for a moment. \V If his
pride riſe vp euen to the heauen, and his head touch the cloudes: \V as
a dunghil in the end he shal be deſtroyed, and they that had ſene him,
shal ſay: Where is he? \V As a dreame that fleeth away he shal not be
found, he shal paſſe as a viſion by night. \V The eie that had ſene him,
shal not ſee him, neither shal his place behold him anie more. \V His
children shal come to naught with pouertie, and his handes shal render
him his ſorow. \V His bones shal be filled with the vices of his youth,
and they shal ſleepe with him in the duſt. \V For when euil shal be
ſweete in his mouth, he wil hide it vnder his tongue. \V He wil ſpare it,
and not leaue it, and wil hide it in his throte. \V His bread in his
belly shal be turned into the gaule of
%%% !!! missing TNote?
%%% \TNote{}
aſpes within him. \V The riches which he hath deuoured, he shal vomite
out, and God shal draw them forth out of his belly. \V He shal ſucke the
head of aſpes, and the vipers tongue shal kil him. (\V Let him not ſee
the ſtreames of the riuer, the torrentes of honie, & of butter.) \V He
shal be punished for al thinges that he did, & yet not be cõſumed:
according to the multitude of his inuentions ſo alſo shal he ſuffer. \V
Becauſe breaking in he hath made the poore naked: he hath violently
taken houſe, & built it not. \V Neither is his bellie filled: and when
he shal haue the thinges he coueted, he can not poſſeſſe them. \V There
remayned not of his meate, & therfore nothing shal continewe of his
goodes. \V When he shal be filled, he shal be ſtraytened, he shal burne,
and al ſorow shal fal vpon him. \V Would God his belly were filled, that
he may ſend forth the wrath of his furie vpon him, and rayne his battel
vpon him. \V He shal flee wepons of yron, and shal fal vpon a bowe of
braſſe. \V The ſword plucked out, and coming forth of his ſcabbard, and
gliſtering in his bitterneſſe: the horrible shal goe and come vpon
him. \V Al darkeneſſe is hid in his ſecretes: fyre that is not kindled
shal deuoure him, he shal be
%%% o-0998
afflicted leaſt in his tabernacle. \V The heauens shal reuele his
iniquitie, and the earth shal riſe againſt him. \V The bloſſome of his
houſe shal be opened, he shal be plucked downe in the day of Gods
furie. \V This the portion of an impious man from God, & the
inheritance of his wordes from our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
