%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Philemon
%%%% Chapter
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter


\project douay-rheims


%%% 2840
%%% o-2699
\startChapter[
  title={}
  ]

\Summary{Hearing of Philemons vertue, who was a Coloſsian, he writeth a
  familiar letter from Rome (being priſoner there) about his fugitiue
  ſeruant Oneſimus: not doubting but that he might command him, yea
  rather requeſting that he wil forgiue him, yea and receiue him as he
  would Paul himſelf, who alſo hopeth to come vnto him.}

%%% o-2700
Paul the priſoner of Chriſt \Sc{Iesvs}, and brother Timothee: to
Philemon the beloued and our Coadiutour, \V and to Appia our deareſt
ſiſter, and to Archippus our fellow-ſouldiar and to the Church which is
in thy houſe. \V Grace to you and peace from God our Father, and our
Lord \Sc{Iesvs} Chriſt.

\V I giue thankes to my God, alwaies making a memorie of thee in my
praiers, \V hearing thy
\SNote{Faith and charitie commended alwaies together, both neceſſarie to
make a complete Chriſtian man & to iuſtification  & ſaluation.}
charitie & faith which thou haſt in our Lord \Sc{Iesvs}, and
\LNote{Toward al the Saints.}{The
\MNote{Faith and beleefe in Saints.}
Apoſtle ſticketh not to ſay, Charitie and faith in Chriſt and al his
Saints, which our captious Aduerſaries count in Catholike mens
\Fix{ſpeahes}{ſpeaches}{obvious typo, fixed in other}
and writings, very abſurd, feining that in al ſuch we make no difference
betwixt the loue we beare to Chriſt, and the loue we owe to our
neighbours: betwixt the truſt or beleefe we haue in God, and that which
we haue in his holy Saints. Malice and contention doth ſo blind al
Heretikes.}
toward al the Saints: \V that the communication of thy faith may be made
euident in the agnition of al
\Var{good}{good worke}
that is in you in Chriſt \Sc{Iesvs}: \V For I haue had great ioy and
conſolation in thy charitie, becauſe the bowels of the Saints
\SNote{The dueties of charitie and mercie done to Chriſtes priſoners,
are exceeding acceptable to God and al good men.}
haue reſted by thee brother.

\V For the which thing hauing great confidence in Chriſt \Sc{Iesvs} to
command thee that which pertaineth to the purpoſe: \V for charitie
rather I beſeech, whereas thou art ſuch an one, as Paul being old and
now priſoner alſo of \Sc{Iesvs} Chriſt. \V I beſeech thee for my ſonne
whom I haue begotten in bands,
\CNote{\XRef{Col.~4,~9.}}
Oneſimus, \V who hath been ſometime vnprofitable to thee, but now
profitable both to me and thee, \V whom I haue ſent backe to thee.
And
\SNote{Al Spiritual men ought to be exceeding propenſe & ready to
procure mens pardon and reconciliation to al penitents.}
doe thou receiue him as mine owne bowels. \V Whom I would haue reteined
with me, that for thee he might miniſter to me in the bands of the
Ghoſpel: \V but without thy counſel I would doe nothing: that thy good
might be not as it were of neceſſitie, but voluntarie. \V For perhaps
therfore he departed for a ſeaſon from thee, that thou mighteſt take him
againe for euer. \V Now not as a ſeruant, but for a ſeruant, a moſt
deare brother, eſpecially to me, but how much more to thee both in the
fleſh and in our Lord? \V If therfore thou take me for thy fellow;
receiue him as my ſelf. \V And if he hath hurt thee any thing or is in thy
debt, that impute to me. \V I Paul haue written with mine owne hand: I
wil repay it: not to ſay to thee,
\SNote{The great debt and duetie that we owe to ſuch as be our ſpiritual
parents in Chriſt.}
that thou oweſt me thine owne ſelf alſo. \V Yea brother.
\TNote{\G{ὀναίμην}}
God grant I may enioy thee in our Lord. Refreſh my bowels in our
Lord. \V Truſting in thy obedience I haue written to thee, knowing that
thou
%%% o-2701
wilt doe aboue that alſo which I doe ſay. \V And withal prouide me alſo a
lodging. For I hope by your praiers that I ſhal be giuẽ to you.

\V There ſalut thee Epaphras my fellow-priſoner in Chriſt \Sc{Iesvs}, \V
Marke, Ariſtarchus, Demas, and Luke my Coadiutours. \V The grace of our
Lord \Sc{Iesvs} Chriſt be with your ſpirit. Amen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

