%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Jude
%%%% Chapter
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter


\project douay-rheims


%%% 2931
%%% o-2793
\startChapter[
  title={}
  ]

\Summary{He exhorteth them to ſtand to their old faith, shewing them by
  examples, that it is damnable not to continue and be conſtant:
  8.~inueighing againſt the lecherie, blaſphemie, apoſtaſie, banketing
  of the heretikes, 14.~and that their dãnation was long
  foretold. 19.~Catholikes therfore to be vnmoueable, to reproue the
  obſtinate, to recouer al not deſperate, to confirme the weake, and to
  liue themſelues vertuouſly and without mortal ſinne, which by God's
  grace they may doe.}

Iude the ſeruant of \Sc{Iesvs} Chriſt, and brother of Iames: to them
that are in God the Father beloued, and in \Sc{Iesvs} Chriſt preſerued,
and called. \V Mercie to you, and peace and charitie be accompliſhed.
\V My Deareſt, taking al care to write vnto you of your common
ſaluation, I thought it neceſſarie to write vnto you: beſeeching you to
contend for the faith once deliuered to the Saints. \V For there are
certaine men ſecretly entred in (which were long agoe preſcribed vnto
this iudgement) impious, transferring the grace of our God
\SNote{Diuers Heretikes abuſe the libertie of Chriſtes grace and
Ghoſpel, to the fulfilling of their carnal luſtes and concupiſcences.}
into riotouſnes, and denying the only Dominatour, & our Lord \Sc{Iesvs}
Chriſt. \V
\CNote{\XRef{2.~Pet.~2.}}
But I wil admonish you, that once know al things, that
\SNote{This is our Sauiour, not Ioſue, as S.~Hierom noteth.
\Cite{ep.~17.}
See
\XRef{Abac. c.~3. v.~18.}}
\Sc{Iesvs}, ſauing the people out of the land of Ægipt,
\CNote{\XRef{Nu.~14,~37.}}
ſecondly deſtroied them which beleeued not. \V But the Angels which kept
not their principalitie, but forſooke their owne habitation, he hath
reſerued vnder darkeneſſe in eternal bonds vnto the iudgement of the
great day. \V As
\CNote{\XRef{Gen.~19.}}
Sodom and Gomorrhe, and the cities adioyning in like manner hauing
\TNote{\L{exfornicatæ}}
fornicated, and going after other flesh, were made an example,
ſuſtaining the paine of eternal fire. \V In like manner theſe alſo
defile the fleſh, and
\SNote{Such heretikes, that wil not be ſubiect to any ſuperiour, or that
refuſe to obey the lawes either of Spiritual or Temporal Rulers. In
which kind (ſpecially in blaſpheming the ſupreme Spiritual Magiſtrate)
the Proteſtants do paſſe.}
deſpiſe dominion, & blaſpheme maieſtie. \V When Michael the Archangel,
diſputing with the Diuel, made altercation
%%% o-2794
\LNote{For the body of Moyſes.}{When,
\MNote{Truths vnwritten, and knowen by tradition.}
why, or how this altercation or combat was between S.~Michael and the
Diuel about Moyſes body, no man can declare. Only this we ſee that many
truthes and ſtories were kept in the mouthes and harts of the faithful,
that were not written in Scriptures canonical, as this was among the
Iewes.}
for the body of Moyſes, he durſt not inferre iudgement of blaſphemie,
but ſaid, Our Lord
\Var{command}{rebuke}
thee. \V But theſe, what things ſoeuer certes they are ignorant of,
\LNote{They blaſpheme.}{He
\MNote{Ignorance maketh Heretikes blaſpheme.}
ſpeaketh of Heretikes, who being ignorant in God's myſteries and the
diuine doctrine of his Church, when they can not reproue the things,
then they fal to execrations, irriſions, and blaſphemies againſt the
Prieſts, Church, and Sacraments, and whatſoeuer is godly.}
they blaſpheme: and what things ſoeuer naturally, as dumme beaſts, they
know, in thoſe they are corrupted.

%%% 2932
\V Woe vnto them,
\Var{which}{becauſe they}
haue gone in the way of
\CNote{\XRef{Gen.~4,~8.}}
\LNote{Cain, Balaam, Core.}{The
\MNote{Heretikes reſembled to Cain, Balaam, and Core.}
Apoſtle would haue Heretikes ſpecially to be knowen, by the reſemblance
they haue, firſt to Cain, in that for enuy that his brothers ſeruice and
Sacrifice was accepted and his reiected, ſlew his ſaid brother, and was
a fugitiue from the face and citie of God, which is the
Church. Secondly, by their reſemblance to Balaam, who for money was
induced to curſe God's people, as couetouſnes is cõmonly the cauſe that
firſt maketh Heretikes & falſe Prophets. Whereupõ S.~Aug. ſaith:
\Emph{He is an Heretike that for temporal commodities ſake either
coineth or followeth new opinions.}
\Cite{S.~Auguſt. li. de vtil. cred. cap.~1.}
And laſtly, by the reſemblance they haue with the ancient and notorious
Schiſmatike Core, and his companions, who forſooke the ordinarie
Prieſthood appointed by God, and would needs doe Sacrifice themſelues
without lawful calling.

Such indeed be al Heretikes, and ſuch be al their ſacraments, ſeruice,
and offices in their Church, as Cores were in his ſchiſmatical
tabernacles. And as pride was the cauſe of his reuolting from the
obedience of Moyſes and Aaron his Prieſts and true Gouernours; ſo is
intolerable pride the cauſe of al Heretikes forſaking their lawful
Paſtours and Rulers, and namely of forſaking Chriſtes owne Vicar in
earth, our true Aaron, as S.~Bernard calleth him.
\Cite{De conſid. li.~2. cap.~8.}
To al ſuch forſakers the Apoſtle here giueth the curſe and Væ due to
the ſaid three, Cain, Balaam, and Core, and telleth them that the ſtorme
of darkenes and eternal damnation is prouided for them: moſt liuely
deſcribing al Heretikes (as in ſome we to our woe haue experience by
their mãners in our daies) in al this paſſage euen to the end of the
Epiſtle.}
Cain: and with the errour of
\CNote{\XRef{Nu.~12.}}
Balaam, haue for reward powred out themſelues, and haue periſhed in the
%%% !!! not marked in either
\CNote{\XRef{Nu.~16.}}
contradiction of Core. \V Theſe are in their bankets, ſpots, feaſting
together without feare, feeding themſelues, clouds without water which
are carried about of winds, trees of autumne, vnfruitful, twiſe dead,
plucked vp by the roots, \V raging waues of the ſea, foming out their
owne confuſions, wandering ſtarres: to whom the ſtorme of darkneſſe is
reſerued for euer. \V And of theſe prophecied Enoch, the ſeuenth from
Adam, ſaying:
%%% !!! XRef?
Behold our Lord is come in his holy thouſands, \V to doe iudgement
againſt al, and to reproue al the impious, of al the workes of their
impietie whereby they haue done impiouſly, and of al the hard things
which impious ſinners haue ſpoken againſt him. \V Theſe are murmurers,
ful of complaints, walking according to their owne deſires, and their
mouth ſpeaketh pride, admiring perſons for gaine ſake.

\V But you, my Deareſt, be mindful of the words which haue been ſpoken
before by the Apoſtles of our Lord \Sc{Iesvs} Chriſt, \V who told you,
\CNote{\XRef{1.~Tim.~4.}
\XRef{2.~Tim.~3.}
\XRef{2.~Pet.~2.}}
that in the laſt time ſhal come mockers, according to their owne deſires
walking in impieties. \V
\LNote{Theſe are they which ſegregate themſelues.}{The
\MNote{Al Heretikes ſegregate themſelues.}
conditions of Heretikes in the later daies, that is, euer ſince Chriſt's
time, not of theſe only of our Age. For there were many that forſooke
Gods Church and \Emph{ſegregated themſelues} from the fellowship of the
faithful euen in the primitiue Church: that we may the leſſe maruel at
theſe mens ſegregating themſelues, and going out from the reſt, into
ſeueral Sects, which
%%% !!! Cite?
S.~Auguſtin therfore calleth \Emph{Segregations.}}
Theſe are they which ſegregate themſelues, ſenſual, hauing not the
Spirit. \V But you, my Deareſt, building your ſelues vpon
\Var{our}{your}
moſt holy faith, in the Holy Ghoſt, praying, \V keep your ſelues in the
loue of God, expecting the mercie of our Lord \Sc{Iesvs} Chriſt vnto
life euerlaſting. \V And theſe certes reproue being iudged: \V but them
ſaue, pulling our of the fire. And on other haue mercie in feare: hating
alſo that which is carnal, the ſpotted cote.

\V And to him that is able to preſerue you without ſinne, and to ſet you
immaculate before the ſight of his glorie in exultation in the comming
of our Lord \Sc{Iesvs} Chriſt, \V to the only God our Sauiour by
\Sc{Iesvs} Chriſt our Lord be glorie and magnificence, empire and power
before al worlds, and now and for al worlds euermore. Amen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

