%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% One John
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 2913
%%% o-2775
\startChapter[
  title={Chapter 01}
  ]

\Summary{Good cauſe there is to beleeue the Apoſtles preaching. 5.~And
  this is one point of their preaching, that to haue participation with
  God, we muſt not only beleeue, but alſo abſtaine from al mortal ſinne,
  8.~though we al ſinne venially.}

That which was from the beginning, which we haue heard, which we haue
ſeen with our eyes, which we haue looked vpon, and our hands haue
handled, of the Word of life: (\V and the life was manifeſted: and we
haue ſeen, and doe teſtifie, and declare vnto you the life eternal which
was with the Father, and hath appeared to vs) \V that which we haue ſeen
and haue heard, we declare vnto you, that
\LNote{You may haue ſocietie.}{S.~Iohn
\MNote{No ſaluation but in the Societie of the Church.}
sheweth manifeſtly, that whoſoeuer deſire to be partakers with God, muſt
firſt be vnited to the Churches ſocietie, learne that faith, and receiue
thoſe Sacraments, which the Diſciples receiued of the Truth it-ſelf,
conuerſant with them in flesh. So ſaith Venerable Bede
\Cite{vpon this place.}
Whereby we ſee there is no Societie with God in Sectes or ſchiſmes, nor
any-where but in the vnitie, fellowship, & communion of that Church
which can proue it-ſelf to deſcend from the Apoſtles.}
you alſo may haue ſocietie with vs,
%%% Extra LNote mark? Not in other.
%%% \LNote{}{}
and our ſocietie may be with the Father and with his Sonne \Sc{Iesvs}
Chriſt. \V And theſe things we write to you, that you may reioyce, and
your ioy may be ful.

\V And this is the annuntiation which we haue heard of him, and declare
vnto you, That
\CNote{\XRef{Io.~8,~11.}}
God is light, and in him there is no darkeneſſe: \V If we ſhal ſay that
we haue ſocietie with him, and walke in darkeneſſe, we lie, and doe not
the truth. \V But if we walke in the light, as he alſo is in the light:
we haue ſocietie one toward another, and
\CNote{\XRef{Heb.~9.}
\XRef{1.~Pet.~1.}
\XRef{Apo.~1.}}
\LNote{The bloud of Ieſus.}{Whether
\MNote{Many meanes & inſtruments of remitting ſinne, but al by the force
& merits of Chriſt's bloud applied by them.}
ſinnes be remitted by praiers, by faſting, by almes, by faith, by
charitie, by Sacrifice, by Sacraments, & by the Prieſts, (as the holy
Scriptures doe plainely attribute remiſsion to euery of theſe) yet none
of al theſe doe otherwiſe remit, but in the force, by the merit & vertue
of Chriſts bloud: theſe being but the appointed meanes & inſtruments by
which Chriſt wil haue his holy bloud to worke effectually in vs. Which
meanes whoſoeuer contemneth, depriueth himſelf of the commoditie of
Chriſt's owne bloud & continueth ſtil in ſinne and vncleaneſſe, vaunt he
himſelf neuer ſo much of Chriſt's death. Which point let the Proteſtãts
marke wel, and ceaſe to beguile their poore deceiued followers,
perſuading them, that the Catholikes derogate from Chriſt's bloud, or
ſeeke remiſsion otherwiſe then by it, for that they vſe humbly the
meanes appointed by Chriſt to apply the benefit of his holy bloud vnto
them.}
the bloud of \Sc{Iesvs} Chriſt his Sonne cleanſeth vs
\LNote{From al ſinne.}{From
\MNote{Al remiſsion of ſinnes is by the Paſsion of Chriſt though by
ſecundarie meanes alſo.}
original and actual, venial and mortal, \L{a culpa & pœna}, that is from
the fault and the paine due for the ſame. V.~Bede ſaith, that Chriſt's
Paſsion doth not only remit in Baptiſme the ſinnes before committed, but al
other afterward alſo done by frailtie: yet ſo, if we vſe for the
remiſsion of them, ſuch meanes as be requiſit and as Chriſt hath
appointed, whereof he reckneth ſome.
\Cite{Bede vpon this place.}
See S.~Auguſtin alſo
\Cite{vpon this place to.~9.}
and S.~Hierome
\Cite{li.~2. con. Pelag. c.~3.}}
from al ſinne. \V
\CNote{\XRef{3.~Reg.~8,~46.}
\XRef{2.~Par.~6,~36.}}
If we ſhal ſay
\LNote{That we haue no ſinne.}{We
\MNote{Some ſinnes venial.}
gather by theſe wordes and the former, that there be two ſortes of
ſinnes: one mortal, excluding vs from light & the ſocietie of God:
another venial, which is found euen in thoſe that walke in the light,
and are in the ſocietie of God.
\MNote{A man may be truely iuſt, notwithſtanding venial ſinnes.}
Alſo we note againſt the Pelagians, that we be truely called the ſonnes
of God, and ſo iuſt indeed, though we be not without al ſinnes, euery
one of vs, as wel iuſt as vniuſt, being taught and bound to confeſſe our
offenſes, and to aske pardon daily of God, by this petition of the
\L{Pater noſter}, \Emph{Forgiue vs our debts}. Therfore S.~Auguſtin
\Cite{li. de natura & grat. c.~36.}
reckneth vp al the holy Patriarches, Prophets, and renowmed iuſt
perſons, to haue been ſinners, euen when they were in grace, and
iuſtice:
\MNote{S.~Auguſtin excepteth our B.~Ladie from ſinnes.}
excepting alwaies our B.~Ladie, \L{de qua propter honorem Domini, nullam
prorſus, cum de peccati agitur, habere volo quæſtionem}; \Emph{of whom},
ſaith he, \Emph{for the honour of our Lord, when we talke of ſinnes, I
wil haue no queſtion}. And Pelagius asking what ſinnes Abel and ſuch iuſt
men did commit,
\CNote{c.~38.}
S.~Auguſtin anſwereth,
\MNote{Examples of venial ſinnes.}
that they might laugh ſometime immoderately, or ieſt too much, or couet
ſome-what intemperately, or plucke fruit ouer greedily, or in eating
take ſome-what more then afterward was wel digeſted, or haue their
intention in time of praier ſome what diſtracted, and ſuch like. Thus in
ſenſe S.~Auguſtin. Whereby we may learne which be venial ſinnes, that
cõſiſt with true iuſtice &
\CNote{\Cite{de dono perſeuerantis c.~2.}}
can not alwaies be auoided euen of holy men in this life. In the booke
\Cite{de fide ad Petram c.~41.}
are excepted from this common rule of ſinners, the children which be
newely baptized and haue not yet vſe of reaſon to ſinne either mortally
or venially.}
that we haue no ſinne, we ſeduce our ſelues, and the truth is not in
vs. \V If we confeſſe our ſinnes, he is faithful & iuſt, for to forgiue
vs our ſinnes, and to cleanſe vs
%%% o-2776
from al iniquitie. \V If we ſhal ſay that we haue not ſinned, we make
him a lier, and his word is not in vs.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

