%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Argument
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent argument


\project douay-rheims


%%% 0837
%%% o-0750
\startArgument[
  title={\Sc{The Argvment of Paralipomenon.}},
  marking={Argument of Paralipomenon.}
  ]

\Emph{Paralipomenon},
\MNote{The ſignification of the name; and the contents of this booke.}
that is, \Emph{A ſupplement of thinges ommitted}, called by the Hebrewes
\HH{Dibre haiamim}, \Emph{The wordes of the dayes}, or Chronicle, is an
Abridgement briefly shewing, beſides diuers other genealogies, from the
beginning of the world, the faith and religion both of the progenitors,
and ofspring of the Patriarch Iacob, whoſe progenie God choſe and made
his peculiar people; and in that nation more particularly recounting the
Actes of king Dauid, and other kinges of his line, til they were led
captiue into Babylon. \Emph{A booke of ſuch and ſo great importance}
(ſayeth
\CNote{\Cite{Epiſt. ad Paulin.}}
S.~Hierom.) \Emph{as whoſoeuer without it arrogateth the knowledge of
Scriptures, may mocke himſelfe.} The author is vncertaine, but probably
it ſemeth to be gathered by \Emph{Eſdras}, out of other bookes and
traditions, for the perfecting of the old Teſtament: and is vndoubtedly
canonical Scripture.
\MNote{Diuided into two bookes.}
In Hebrew it is al one booke, but being large, is with the Greekes and
Latines parted into two.
\MNote{The firſt booke into three partes.}
And the firſt booke may be diuided into three principal partes. The
firſt nine chapters conteyne diuers genealogies; firſt by the only right
line from Adam to Noe, then by diuers lines of Noes progenie, but moſt
ſpecially of Iacobs twelue ſonnes. In the tenth chapter the reiection,
and death of king Saul is repeted. The other ninetene chapters are al of
Dauid; to wit, of his election to be king, and inunction, his vertues,
his faultes alſo, and his ſpecial actes, concerning Gods ſeruice,
gouernment of the people; and prouiſion made for building Gods Temple.

\stopArgument


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
