%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 0846
%%% o-0759
\startChapter[
  title={Chapter 5}
  ]

\Summary{Genealogies of Ruben, whoſe birthright, concerning duble
  portion, is tranſlated to Ioſeph, the principalitie to Iuda, with
  their ſpecial actes. 11.~Alſo of Gad. 18.~Who with Ruben, and half
  tribe of Manaſſes, ſubdue the Agarenes: 25.~but for their ſinnes are
  led captiue into Aſſyria.}

Alſo the ſonnes of Ruben the firſt begotten of Iſrael (for he was his
firſt begotten:
\SNote{See
\XRef{annotations Gen.~49.}
\XRef{Nu.~4.}}
but when he had violated his fathers bed, his birthright was giuen to
the ſonnes of Ioſeph the ſonne of Iſrael, and he was not reputed for the
firſtbegotten. \V Moreouer Iudas, which was the ſtrongeſt among his
brethren, of his ſtocke ſprang the princes: but the firſtbirthright was
reputed to Ioſeph.) \V The ſonnes then of Ruben the firſtbegotten of
Iſrael: Enoch, & Phallu, Eſron, and Charmi. \V The ſonnes of Ioel: Samia
his ſonne, Gog his ſonne, Semei his ſonne, \V Micha his ſonne, Reia his
ſonne, Baal his ſonne, \V Beera his ſonne, whom Theglathphalnaſar the
king of the Aſſyrians led away captiue, and he was prince in the tribe
of Ruben. \V And his brethren, and al his kinred, when they were numbred
by their families, had theſe princes Iehiel, and Zacharias. \V Moreouer
Bala the ſonne of Azaz, the ſonne of Samma, the ſonne of Ioel, he dwelt
in Aroer as far as Nebo, and Beelmeon. \V Agaynſt the
%%% 0847
eaſt quarter alſo he dwelt vnto the entrance of the deſert, and the
riuer Euphrates. For they poſſeſſed a great number of cattel in the land
of Galaad. \V And in the daies of Saul they fought agaynſt the Agarenes,
and ſlewe them, and dwelt for them in their tabernacles, in al the
quarter, that looketh to the Eaſt of Galaad. \V But the children of Gad
dwelt ouer agaynſt them in the land of Baſan, as far as Selcha: \V Iohel
the head, and Sapham the ſecond: and Ianai, and Saphar in Baſan. \V And
their brethren according to the houſes of their kinredes, Michael, and
Moſollam, and Sebe, and Iorai, and Iacan, and Zie, and Heber, ſeuen. \V
Theſe are the ſonnes of Abihail, the ſonnes
of Huri, the ſonne of Iara, the ſonne of Galaad, the ſonne of Michael,
the ſonne of Ieſeſi, the ſonne of Ieddo, the ſonne of Buz. \V Alſo the
brethren of the ſonne of Abdiel, the ſonne of Guni, prince of the houſe
in their families. \V And they dwelt in Galaad, and in Baſan, and in the
townes therof, and in al the ſuburbes of Saron, vnto the borders. \V Al
theſe were numbred in the daies of Ioatham the king of Iuda, and in the
daies of Ieroboam the king of Iſrael. \V The children of Ruben, and of
Gad, and of halfe the tribe of Manaſſes, men of warre, carying ſheildes,
& ſwordes, and bending the bow, and taught to battels, four and fourtie
%%% o-0760
thouſand, and ſeuen hundred threeſcore going forth to fight. \V They
fought agaynſt the Agarenes: but the Itureians, and Naphis, and Nodab \V
gaue them ayde. And the Agarenes were deliuered into their handes, and
al that were with them, becauſe they called vpon God when they fought:
and he heard them,
\SNote{Of theſe and the like, S.~Paul ſayth: \Emph{By faith they
ouercame kingdomes.}
\XRef{Heb.~11.}}
becauſe they beleued in him. \V And they tooke al that they poſſeſſed,
camels fiftie thouſand, and ſheepe two hundred fiftie thouſand, and
aſſes two thouſand, and of men an hundred thouſand ſoules. \V And manie
fel downe wounded: for it was the battel of our Lord. And they dwelt for
them vntil the tranſmigration. \V Alſo the children of the halfe tribe
of Manaſſes poſſeſſed the land, from the coſtes of Baſan vnto Baal,
Hermon, and Sanir, and mount Hermon, for the number was great. \V And
theſe were the princes of the houſe of their kinred, Epher, and Ieſi,
and Eliel, and Eſriel, and Ieremia, and Odoia, and Iediel, moſt valiant
men and mightie, and renowned princes in their families. \V But they
forſooke the God of their fathers, and fornicated after the goddes of
the peoples of the land, whom
%%% 0848
God tooke away before them. \V And the God of Iſrael rayſed vp the
ſpirit of Phul king of the Aſſyrians, and the ſpirit of
Thelgathphalnaſar king of Aſſur: and he tranſported Ruben, and Gad, and
the halfe tribe of Manaſſes, and brought them into Lahela, and into
Habor, and to Ara, and to the riuer of Gozan, vntil this day.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
