%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 28
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-28


\project douay-rheims


%%% 0885
%%% o-0794
\startChapter[
  title={Chapter 28}
  ]

\Summary{Agayne king Dauid calling al his ſtates together, with
  commemoration of Gods ſingular fauour in electing him, and Salomon,
  8.~exhorteth them al, and namely his ſonne to ſerue God ſincerly:
  10.~and to build the temple, deſcribing in what forme to make it.}

Dauid
\CNote{\XRef{3.~Reg.~1.}}
therfore called together al the princes of Iſrael, the dukes of the
tribes, and the heades of the compaynies, which miniſtred to the king:
the tribunes alſo and the centurions, and them that were chiefe ouer the
ſubſtance and poſſeſſions of the king, and his ſonnes with the eunuches,
and the mightie, and al the moſt
\Fix{viliant}{valiant}{obvious typo, fixed in other}
in the hoſt of Ieruſalem. \V And when the king had riſen vp, and ſtoode,
he ſayd: Heare me my brethren, and my people: I meant to haue built a
houſe, wherein the Arke of our Lord might reſt, and the foote ſtoole of
the feete of our God: and to build it I prepared al thinges. \V But God ſayd
to me:
\CNote{\XRef{2.~Reg.~7.}}
Thou ſhalt not build a houſe to my name: becauſe thou art a man of
warre, and haſt ſhed bloud. \V Howbeit our Lord the God of Iſrael choſe
me of al the houſe of my father, to be king ouer Iſrael for euer: for
\SNote{So Iacob prophecied, that the royal ſcepter ſhould come to the
tribe of Iuda, and therein remaine til Chriſt.
\XRef{Gen.~49.}}
of Iuda, he choſe princes: moreouer of the houſe of Iuda, my fathers
houſe: and of the ſonnes of my father, it pleaſed 
him to chooſe me king ouer al Iſrael. \V Yea and of my ſonnes (for our
Lord hath geuen me manie ſonnes) he hath choſen Salomon my ſonne, to
ſitte in the throne of the kingdome of our Lord ouer Iſrael, \V and he
ſayd to me: Salomon thy ſonne ſhal build my houſe, and my courtes: for
him haue I choſen to me for a ſonne, and I wil be to him for a
father. \V And I wil confirme his kingdome for euer, if he shal
perſeuere to doe my preceptes, and my iudgementes,
%%% 0886
as
\SNote{It is here euident, that Salomon was ſome time the true ſeruant
and childe of God, keeping his preceptes and as certaine, that he fel
into great ſinnes, and loſt Gods grace, and finally it is doubtful in
what ſtate he died,
\XRef{3.~Reg.~11.}
wherby is conuinced their vaine phancie that thinck, he who is once the
child of God can neuer fal, nor become a wicked man.}
at this day. \V Now then before al the aſſemblie of Iſrael in the
hearing of our God, keepe ye, & ſearch al the commandmentes of our Lord
God: that you may poſſeſſe the good land, and may leaue it to your
children after you for euer. \V And thou Salomon my ſonne, know the God
of thy father, and ſerue him with a perfect hart, and a voluntarie
minde: for our Lord ſearcheth al hartes, and vnderſtandeth al
cogitations of mindes. If thou ſeeke him, thou ſhalt finde: but if thou
forſake him he wil reiect thee for euer. \V Now therfore becauſe our
Lord hath choſen thee to build the houſe of the Sanctuarie, take
courage, and perfitte it. \V And Dauid gaue to Salomon his ſonne a
deſcription of the porch, and of the temple,
%%% o-0795
and of the cellars, and of the vpper loft, and of the chambers in the
inner roomes, and of the houſe of the propitiation, \V moreouer alſo of al
the courtes, which he had thought vpon, and of the chambers round about,
for the treaſures of the houſe of our Lord, and for the treaſures of the
holie thinges, \V and of the diuiſions of the Prieſtes and the Leuites,
for al the workes of the houſe of our Lord, and for al the veſſels of
the miniſterie of the temple of our Lord. \V Gold in weight for euerie
veſſel of the miniſterie. A weight of ſiluer alſo according to the
diuerſitie of the veſſels and workes. \V And for golden
candleſtickes alſo, and their lampes, gold according to the meaſure of
euerie candleſticke, and of the lampes. In like maner alſo for the
ſiluer candleſtickes, and for their lampes, according to the diuerſitie
of the meaſure, he deliuered a weight of ſiluer. \V He gaue alſo gold for
tables of propoſition, according to the diuerſitie of the tables: in
like maner alſo ſiluer for other ſiluer tables. \V For fleſh hookes
alſo, and phials, and cenſars of moſt pure gold, and for little lions of
gold, according to the qualitie of the meaſure he diſtributed a weight,
for lyon and lyon. In like maner alſo for lyons of ſiluer he ſeperated a
diuerſe weight of ſiluer. \V And for the Altar whereupon incenſe is
burnt, he gaue moſt pure gold: that of it might be made a ſimilitude of
the chariote of the Cherubs ſpredding their winges, and couering the
Arke of the couenant of our Lord. \V Al thinges, quoth he, came writen
with the hand of our Lord vnto me: that I might vnderſtand al the workes
of the paterne. \V Dauid alſo ſayd to Salomon his ſonne: Deale manfully,
and take courage, and doe it: feare not, and be not diſmayed: for our
Lord my God
%%% 0887
wil be with thee, and wil not leaue thee, nor forſake thee, til thou
perfite al the worke of the miniſterie of the houſe of our Lord. \V
Behold the diuiſions of the Prieſtes and the Leuites, for euerie
miniſterie of the houſe of our Lord, aſſiſt thee, and are readie, and as
wel the princes know, as the people, to doe al thy preceptes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
