%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 0851
%%% o-0763
\startChapter[
  title={Chapter 7}
  ]

\Summary{Genealogies of Iſſachar, 6.~Beniamin, 13.~Nephthali,
  14.~Manaſſes, 20.~Ephraim, 30.~and Aſer.}

Moreouer the ſonnes of Iſſachar: Thola, and Phua, Iaſub and Simeron,
foure. \V The ſonnes of Thola: Ozi and Raphaia, and Ieriel, and Iemai,
and Iebſem, and Samuel, princes by the houſes of their kinredes. Of the
ſtocke of Thola were numbred moſt valiant men in the daies of Dauid, two
and twentie thouſand ſix hundred. \V The ſonnes of Ozi: Izrahia, of
whom were borne Michael, and Obadia, and Ioel, and Ieſia, fiue, al
princes. \V And with them by their families and peoples, readie girded
to battel, moſt valiant men, ſix and thirtie thouſand: for they had
many wiues, and children. \V Their brethren alſo through al the kinred
of Iſſachar, moſt ſtrong to fight, were numbred foure ſcore and ſeuen
thouſand. \V The ſonnes of Beniamin: Bela, and Bechor, &
\SNote{\XRef{Gen.~46.}
This third ſonne is called Aſbel: & ſeuen more are there recited. And ſo
in the reſt is much difference: but al may be reconciled by ſuch rules
as are noted.
\XRef{chap.~1.}}
Iadihel, three. \V The ſonnes of Bela: Eſbon, and Oei, and Oziel, and
Ierimoth, and Vrai, fiue princes of their families, and moſt ſtrong to
fight, and their number was twentie two thouſand and thirtie foure. \V
Moreouer the ſonnes of Bechor: Zamira, and Ioas, and Eliezer, and
Elioenai, and Amri, and Ierimoth, and Abia, and Anathoth, and Almath: al
theſe the ſonnes of Bechor. \V And there were numbred by their families
princes of their kinredes moſt
%%% 0852
valiant vnto battel, twentie thouſand and two hundred. \V Moreouer the
ſonnes of Iadihel: Balon. And the ſonnes of Balan: Iehus, and Beniamin,
and Aod, and Chanana, and Zerhan, and Tharſis, Ahiſahar. \V Al theſe the
ſonnes of Iadihel, princes of their kinredes, moſt valiant men,
ſeuentene thouſand, and two hundred going forth to battel. \V Sepham
alſo, and Hapham the ſonnes of Hir: and Haſim the ſonnes of Aher. \V And
the ſonnes of Nephthali: Iaſiel, and Guni, and Iezer, and Sellum, the
ſonnes of Bala. \V Moreouer the ſonne of Manaſſes, Ezriel: and his
concubine the Syrian bare Machir the father of Galaad. \V And Machir
tooke wiues for his ſonnes Happhim, and Saphan: and he had a ſiſter
named Maacha: the name of the ſecond Salphaad, and to Salphaad were
borne daughters. \V And Maacha the wife of Machir
%%% o-0764
bare a ſonne, and ſhe called his name Phares: moreouer the name of his
brother, was Sares: and his ſonnes, Vlam, and Recen. \V And the ſonne of
Vlam, Badan. Theſe are the children of Galaad, the ſonne of Machir, the
ſonne of Manaſſes. \V And his ſiſter
\SNote{\HH{Moleocheth}: \L{Regina} Queene.}
Queene bare
\SNote{\HH{Ishod}, \L{Virum decorum}, that is, A comelie perſonable or
goodlie man. So we leaue the hebrew names in this place becauſe
S.~Ierom, and the whole Church doth ſo in the latin text, which we
tranſlate.}
Goodlieman, and Abiezer, and Mohola. \V And the ſonnes of Semida were,
Ahin, and Sechem, and Leci, and Anion. \V And the ſonnes of Ephraim:
Suthala, Bared his ſonne, Thahath his ſonne, Elada his ſonne, Thahath
his ſonne, and this mans ſonne Zabad, \V and this mans ſonne Suthala,
and this mans ſonne Ezer, and Elad. And the men of Geth borne in the
land ſlewe them, becauſe they came downe to inuade their poſſeſſions. \V
Eghraim therfore their father mourned many daies, and his brethren came
to comfort him. \V And he went in vnto his wife: who conceiued and bare
a ſonne, and he called his name Beria, for that he was borne in the
euils of his houſe. \V And his daughter was Sara, who built Bethhoron,
the nether and the vpper, and Ozenſara. \V Moreouer his ſonne Rapha, and
Reſeph, and Thale, of whom was borne Thaan, \V who begat Laadan: this
mans ſonne alſo was Ammiud, who begat Eliſama, \V of whom was borne Nun,
who had Ioſue his ſonne. \V And their poſſeſſion and habitation, was
Bethel with her daughters, and agaynſt the eaſt of Noran, and on the
weſt quarter of Gazer and her daughters, Sichem alſo with her daughters,
as farre as Aſa with her daughters. \V Alſo neere the children of
Manaſſes Bethſan and her daughters, Thanach & her daughters, Mageddo and
her daughters: Dor and her daughters: in
%%% 0853
theſe dwelt the children of Ioſeph, the ſonne of Iſrael. \V The children
of Aſer: Iemna, and Ieſua, and Ieſſui, and Baria, and Sara their
ſiſter. \V And the ſonnes of Baria: Heber, and Melchiel: he is the
father of Barſaith. \V And Heber begat Iephlat, and Somer, and Hotham,
and Suaa their ſiſter. \V The ſonnes of Iephlat: Phoſech, and Chamaal,
and Aſoth: theſe be the ſonnes of Iephlat. \V Moreouer the ſonnes of
Somer: Ahi, and Roaga, and Haba, and Aram. \V And the ſonnes of Helem
his brother: Supha, and Iemna, and Selles, and Amal. \V The ſonnes of
Supha: Sue Harnapher, and Sual, and Beri, and Iamra, \V Boſor, and Hod,
and Samma, and Saluſa, and Iethran, and Bera. \V The ſonnes of Iether:
Iephone, and Phaſpha, and Ara. \V And the ſonnes of Olla: Aree, and
Haniel, and Reſia. \V Al theſe be the ſonnes of Aſer, princes of their
kinredes, the choſen and moſt valiant dukes of dukes: and their number
of the age that was fitte for battel, was ſix and twentie thouſand.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
