%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 0868
%%% o-0778
\startChapter[
  title={Chapter 17}
  ]

\Summary{Dauid determining to build a Temple, 11.~is admonished by
  Nathan the prophet, that not he, but his ſonne, shal build it; and be
  eſtablished in the kingdom. 16.~Dauid extolleth Gods benignitie
  towards him, and the people.}

And when Dauid dwelt in his houſe, he ſayd to Nathan the prophete:
Behold I dwel in a houſe of cedar: and the Arke of the couenant of our
Lord is vnder skinnes. \V And Nathan ſayd to Dauid: Al thinges, that are
in thy hart doe: for God is with thee. \V Therfore that night the word
of God came to Nathan, ſaying: \V Goe, and ſpeake to Dauid my ſeruant:
Thus ſayth our Lord: Thou ſhalt not build me a
%%% o-0779
houſe to dwel in. \V For neither haue I remayned in houſe from the time,
that I brought out Iſrael, vntil this day: but I haue bene
alwaies changing places of tabernacle, and in tent \V abyding with al
Iſrael. Did I ſpeake to one, at the leaſt, of al the iudges of Iſrael,
whom I commanded to feede my people, and did I ſay: Why haue you not
built me a houſe of cedar? \V Now therfore ſo ſhalt thou ſpeake to my
ſeruant Dauid: Thus ſayeth the Lord of hoſtes: I tooke thee, when in the
paſtures thou dideſt folow the flocke, that thou ſhouldeſt be prince of
my people Iſrael. \V And I haue beene with thee whither ſoeuer thou
wenteſt: and I haue ſlayne al thine enemies before thee, and haue made
thee a name as of one of the great ones, that are renowmed in the
earth. \V And I haue
%%% 0869
geuen a place to my people Iſrael: it ſhal be planted, and ſhal dwel
therin, and ſhal be moued no more, neither ſhal the children of
iniquitie conſume them, as from the beginning, \V ſince the dayes that I
gaue Iudges to my people Iſrael, and humbled al thine enimies. I
therfore tel thee, that our Lord wil build thee
\SNote{That is continew and preſerue thy ſeede, and familie as we ſee it
performed euen to the B.~virgin Marie, and Chriſt of the houſe of Dauid.
\XRef{Mat.~1.}
\XRef{Luc.~2.~3.}}
a houſe. \V And when thou ſhalt haue accompliſhed thy daies to goe to
thy fathers, I wil rayſe vp thy ſeede after thee, which ſhal be of thy
children: and I wil eſtabliſh his kingdom. \V He shal build me a houſe,
and I wil confirme his throne for euer. \V I wil be to him for a father,
and he shal be to me for a ſonne: and my mercie I wil not take from him,
as I tooke from him, that was before thee. \V And I wil eſtablish him in
my houſe, and in my kingdom for euer: and his throne shal be moſt firme
for euer. \V According to al theſe wordes, and according to al this
viſion, ſo ſpake Nathan to Dauid. \V And when king Dauid came, and ſate
before our Lord, he ſayd: Who am I Lord God, and what is my houſe, that
thou shouldeſt geue me ſuch thinges? \V But this alſo hath ſemed little
in thy ſight, and therfore thou haſt ſpoken concerning the houſe of thy
ſeruant for time to come alſo: and haſt made me renowned aboue al men
Lord God. \V What can Dauid adde farther, wheras thou haſt ſoe glorified
thy ſeruant, and knowen him? \V Lord for thy ſeruant according to thy
hart thou haſt done al this magnificence, and would haue al thy great
wonders to be knowen. \V Lord, there is not the like to thee: and there
is none other beſide thee, of al whom we haue heard with our eares. \V
For what other is there, as thy people Iſrael, one nation in the
\Fix{earrh,}{earth,}{obvious typo, fixed in other}
to the which God went, to deliuer it and make it his people, and with
his greateneſſe and terrours caſt out the nations before the face of it,
which he deliuered out of Ægypt? \V And thou haſt made thy people Iſrael
to be thy people
\Fix{euer,}{for euer,}{obvious typo, fixed in other}
and thou Lord art made the God therof. \V Now therfore Lord, the word,
which thou haſt ſpoken to thy ſeruant, and concerning his houſe, be in
confirmed for euer, and do as thou haſt ſpoken. \V And let thy name
remayne and be magnified for euer: and let it be ſayd: The Lord of
hoſtes is God of Iſrael, and the houſe of Dauid his ſeruant permanent
before him. \V For thou Lord my God haſt reueled the eare of thy
ſeruant, to build him a houſe: and therfore thy ſeruant hath
%%% o-0780
found confidence, to pray before thee. \V Now therfore
%%% 0870
Lord thou art God: and thou haſt ſpoken to thy ſeruant ſo great
benefittes. \V And thou haſt begunne to bleſſe the houſe of thy ſeruant,
that it be alwaies before thee: for thee ô Lord bleſſing it, it ſhal be
bleſſed for euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
