%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 0848
%%% o-0760
\startChapter[
  title={Chapter 6}
  ]

\Summary{The genealogies of Leui, 4.~with the right line of Aaron by
  Eleazar to Ioſadech, high prieſt in the captiuitie of Babylon,
  16.~other progenies of his three ſonnes; Gerſon, Caath, and Merari,
  31.~with their offices in the temple: 49.~only Aarons ſonnes admitted
  to prieſthood. 54.~Particular poſſeſsions of the Leuites dwelling
  amongſt the other tribes.}

The ſonnes of Leui: Gerſon, Caath, and Merari. \V The ſonnes of Caath:
Amram, Iſaar, Hebron, and Oziel. \V The children of Amram: Aaron,
Moyſes, and Maria. The ſonnes of Aaron: Nadab and Abiu, Eleazar, and
Ithamar. \V
\SNote{The lineal ſucceſſion of Hiegh prieſtes from Aaron to the
captiuitie in Babylon. Nicephorus counteth ſome others among theſe.
\Cite{li.~2. c.~4.}
Ioſephus alſo differeth from this catalogue.
\Cite{li.~10. c.~11.}}
Eleazar begat Phinees, and Phinees begat Abiſue, \V and Abiſue begat
Bocci, & Bocci begat Ozi. \V Ozi begat Zaraias, and Zaraias begat
Maraioth. \V Moreouer Meraioth begat Amarias, and Amarias begat
Achitob. \V Achitob begat Sadoc, and Sadoc begat Achimaas, \V Achimaas
begat Azarias, Azarias begat Iohanan, \V Iohanan begat Azarias.
\CNote{\XRef{2.~Paral.~26.}}
The ſame is he that executed the prieſtlie office in the houſe, which
Salomon built in Ieruſalem. \V And Azarias begat Amarias, and Amarias
begat Achitob, \V and Achitob
%%% o-0761
begat Sadoc, and Sadoc begat Sellum, \V Sellum begat Helcias, and
Helcias begat Azarias, \V Azarias begat Saraias, and Saraias begat
Ioſedec. \V Moreouer Ioſedec went forth, when our Lord tranſported Iuda,
and Ieruſalem by the handes of Nabuchodonoſor. \V The ſonnes then of
Leui: Gerſon, Caath, and Merari. \V And theſe be the names of the ſonnes
of Gerſon: Lobni and Semei. \V The ſonnes of Caath: Amram, and Iſaar,
and Hebron, and Oziel. \V The ſonnes of Merari: Moholi and Muſi. And
theſe are the kinredes of Leui according to their families. \V Gerſon,
Lobni his ſonne, Iahath his ſonne, Zamma his ſonne, \V Ioah his ſonne,
Addo his ſonne, Zara his ſonne, Iethrai his ſonne. \V The ſonnes of
Caath, Aminadab his ſonne, Core his ſonne, Aſir his ſonne, \V Elcana his
ſonne, Abiaſaph his ſonne, Aſir his ſonne, \V Thahath his ſonne, Vriel
his ſonne, Ozias his ſonne, Saul his ſonne. \V The ſonnes of Elcana:
%%% 0849
Amaſai, and Achimoth, \V and Elcana. The ſonnes of Elcana: Sophai his
ſonne, Nahath his ſonne, \V Eliab his ſonne, Ieroham his ſonne, Elcana
his ſonne. \V The ſonnes of Samuel: the firſt begotten
\Fix{Vaſleni,}{Vaſſeni,}{obvious typo, fixed in other}
and Abia. \V And the ſonnes of Merari: Moholi, Lobni his ſonne, Semei
his ſonne, Oza his ſonne, \V Sammaa his ſonne, Haggia his ſonne, Aſaia
his ſonne. \V Theſe are they, whom Dauid appointed ouer the ſinging men
of the houſe of our Lord, ſince the Arke was placed: \V and they
miniſtred before the tabernacle of teſtimonie, ſinging vntil Salomon
built the houſe of our Lord in Ieruſalem: and they ſtood according to
their order in the miniſterie. \V And theſe are they, which aſſiſted
with their ſonnes, of the ſonnes of Caath, Hemam ſinging man, the ſonne
of Ioel, the ſonne of Samuel, \V the ſonne of Elcana, the ſonne of
Ieroham, the ſonne of Eliel, the ſonne of Thohu, \V the ſonne of Suph,
the ſonne of Elcana, the ſonne of Mahath, the ſonne of
Amaſai, \V the ſonne of Elcana, the ſonne of Iohel, the ſonne of
\Fix{Azaries,}{Azarias,}{obvious typo, fixed in other}
the ſonne of Sophonias, \V the ſonne of Thahath, the ſonne of Aſir, the
ſonne of Abiaſaph, the ſonne of Core, \V the ſonne of Iſaar, the ſonne
of Caath, the ſonne of Leui, the ſonne of Iſrael. \V And his brother
Aſaph, who ſtood on his right hand, Aſaph the ſonne of Barachias, the
ſonne of Samaa, \V the ſonne of Michael, the ſonne of Baſaia, the ſonne
of Melchia, the ſonne of Athanai, the ſonne of Zara, the ſonne of
Adaia, \V the ſonne of Ethan, the ſonne of Zamma, the ſonne of Semei, \V
the ſonne of Ieth, the ſonne of Gerſon, the ſonne of Leui. \V And the
children of Merari their brethren, on the left hand, Ethan the ſonne of
Cuſi, the ſonne of Abdi, the ſonne of Maloch, \V the ſonne of Haſabia,
the ſonne of Amaſia, the ſonne of Helcias, \V the ſonne of Amaſai, the
ſonne of Boni, the ſonne of Somer, \V the ſonne of Moholi, the ſonne of
Moſi, the ſonne of Merari, the ſonne of Leui. \V Their brethren alſo the
Leuites, which were ordained for al the miniſterie of the tabernacle of
the houſe of our Lord. \V But Aaron, and his ſonnes burnt incenſe vpon
the altar of holocauſt, and vpon the altar of incenſe, for
euerie worke of \L{Sancta Sanctorum}: and to pray for
%%% o-0762
Iſrael according to al thinges, which Moyſes the ſeruant of God had
commanded. \V And theſe are the ſonnes of Aaron: Eleazar his ſonne,
Phinees his ſonne, Abiſue his ſonne, \V Bocci his ſonne, Ozi his ſonne,
Zarahia his ſonne, \V Meraioth his ſonne, Amarias
%%% 0850
his ſonne, Achitob his ſonne, \V
\SNote{In Sadoc
\XRef{(3.~Reg.~2.)}
the high prieſthood was reduced to the line of Eleazar which by Gods
ordinance was tranſlated to Heli of Ithamars line but ſtil continued in
the line of Aaron. The reſt of Sadochs line by Achimaas &c. to Ioſedech
in the captiuitie appeareth before.
\XRef{v.~9. ad~15.}}
Sadoc his ſonne, Achimaas his ſonne. \V And theſe are their habitations
by the townes and confines, to witte, of the ſonnes of Aaron, according
to the kinredes of the Caathites: for they were fallen to them by
lotte. \V They gaue therfore to them Hebron in the Land of Iuda, and the
ſuburbes therof round about: \V but the fieldes of the citie, and the
townes to Caleb the ſonne of Iephone. \V Moreouer to the ſonnes of Aaron
they gaue cities, to flee vnto, Hebron, and Lobna, and the ſuburbes
therof, \V Iether alſo and Eſthemo with the ſuburbes therof, yea &
Helon, and Dabir with their ſuburbes: \V Aſan alſo & Bethſemes, & their
ſuburbes. \V And of the tribe of Beniamin: Gabee and the ſuburbes
therof, Almath with the ſuburbes therof, Anothoth alſo with the ſuburbes
therof. Al the cities, thirtene, by their kinredes. \V And to the
children of Caath the reſidue of their kinred they gaue of the halfe
tribe of Manaſſes in poſſeſſion ten cities. \V Moreouer to the children
of Gerſon by their kinredes of the tribe of Iſſachar, & of the tribe of
Aſer, and of the tribe of Nephthali, and of the tribe of Manaſſes in
Baſan, thirtene cities. \V And to the ſonnes of Merari by their kinredes
of the tribe of Ruben, and of the tribe of Gad, and of the tribe of
Zabulon, they gaue by lot twelue cities. \V Alſo the children of Iſrael
gaue to the Leuites cities, and their ſuburbes: \V and they gaue by lot,
of the tribe of the children of Iuda, & of the tribe of the children of
Simeon, and of the tribe of the children of Beniamin, theſe cities,
which they called by their names, \V and to them, that were of the kinred
of the ſonnes of Caath, and the cities in their borders were of the
tribe of Ephraim. \V They therfore gaue them cities, to flee vnto, Sichem
with the ſuburbes therof in mount Ephraim, and Gazer with the ſuburbes
therof, \V Iecmaan alſo with the ſuburbes therof, and Bethhoron in like
maner. \V Moreouer Helon alſo with the ſuburbes therof, and Gethremmon
in like maner. \V Moreouer of the halfe tribe of Manaſſes, Aner & the
ſuburbes therof, Balaam and the ſuburbes therof: to witte, to them,
which were leaft of the kinred of the ſonnes of Caath. \V And to the
sonnes of Gerſom of the kinred of the halfe tribe of Manaſſes, Gaulon in
Baſan, and the ſuburbes therof, and Aſtaroth with the ſuburbes
therof. \V Of the tribe of Iſſachar, Cedes and the ſuburbes therof, and
Dabereth with the ſuburbes therof, \V Ramoth alſo and the ſuburbes
therof, and Anem
%%% 0851
with the ſuburbes therof. \V And of the tribe of Aſer: Maſal with the
ſuburbes therof, & Abdon in like
\Fix{mamer,}{maner,}{obvious typo, fixed in other}
\V Hucac alſo and the ſuburbes therof, and Rohob with the ſuburbes
therof. \V Moreouer of the tribe of Nephthali, Cedes in Galilee and the
ſuburbes therof, Hamon with the ſuburbes therof, and Cariathaim, and the
ſuburbes therof. \V And to the reſt of the ſonnes of Merari: of the
tribe of
%%% o-0763
Zabulon, Remmone and the ſuburbes therof, and Thabor with the ſuburbes
therof: \V beyond Iordan alſo ouer agaynſt Iericho, agaynſt the Eaſt of
Iordan, of the tribe of Ruben, Boſor in the wilderneſſe with the ſuburbes
therof, and Iaſſa with the ſuburbes therof, \V Cademoth alſo and the
ſuburbes therof, and Mephaath with the ſuburbes thereof. \V Moreouer
alſo of the tribe of Gad, Ramoth in Galaad and the ſuburbes therof, and
Manaim with the ſuburbes therof, \V yea and Heſebon with the ſuburbes
therof, and Iezer with the ſuburbes therof.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
