%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 0860
%%% o-0771
\startChapter[
  title={Chapter 12}
  ]

\Summary{VVho folowed Dauid when he fled from Saul. 23.~And who came
  into Hebron to make him king.}

Theſe alſo came to Dauid into Siceleg, when as yet he fled from Saul the
ſonne of Cis, the which were moſt valiant and excellent warriers, \V
drawing the bowe, and hurling with both handes ſtones in ſlinges, and
ſhooting arrowes directly: of the brethren of Saul of Beniamin. \V The
prince Ahiezer, and Ioas the ſonnes of Sammaa a Gabaathite, and Iaziel,
and Phallet the ſonnes of Azmoth, and Beracha, and Iehu an
Anothothite. \V Samaias alſo a Gabaonite the moſt valiant amongſt the
thirtie and aboue the thirtie. Ieremias, and Ieheziel, and Iohanan, and
Iezabad a Gaderothite. \V And Eluzai, and Ierimuth, and Baalia, and
Samaria, and Saphatia an Haruphite. \V Elcana, and Ieſia, and Azareel,
and Ioezer, & Ieſbaam of Carehim: \V Ioela alſo, and Zabadia the ſonnes
of Ieroham of Gedor. \V Yea and of Gaddi alſo there fled to Dauid, when
he lay hid in the deſert moſt valiant men, and the beſt warriers,
holding ſheild and ſpeare: their faces as the faces of a lion, and ſwift
as the roebuckes on the mountaynes: \V Ezer the prince, Obdias the
ſecond, Eliab the third, \V Maſmana the fourth, Ieremias the fifth, \V
Ethi the ſixth, Eliel the ſeuenth, \V Iohanan the eight, Elzebad the
ninth, \V Ierenias the tenth, Machbani the eleuenth. \V Theſe of the
children of Gad were the princes of his armie: the
%%% 0861
meaneſt was captayne ouer an hundred ſouldiars, and the greateſt, ouer a
thouſand. \V Theſe are they which paſſed Iordan the firſt moneth, when
it vſed to flow ouer his bankes: and they chaſed away al that dwelt in
the valleis toward the eaſt quarter and the weſt. \V And there came alſo
of Beniamin, and of Iuda to the hold, wherin Dauid abode. \V And Dauid
went out to mete them, and ſayd: If you be come peacably to me for to
helpe me, my hart be ioyned to you, but if you lye in wayte againſt me
for my aduerſaries, whereas I haue no iniquitie in my handes, the God of
our fathers ſee, and iudge. \V But the ſpirit came on Amaſai the prince
among thirtie, and he ſayd: We are thine ô Dauid, and with thee ô ſonne
of Iſai: peace, peace be to thee, & peace to thy helpers. For thy God
helpeth thee. Dauid therfore receiued them, and appoynted them princes
of the band. \V Moreouer of Manaſſes there fled to Dauid, when he came
with the Philiſthijms agaynſt Saul, to fight and he fought not with
them: becauſe the princes of the Philiſthimes taking counſel ſent him
backe, ſaying: with the peril of our head wil he returne to his lord
%%% o-0772
Saul. \V Therfore when he returned into Siceleg, there fled to him of
Manaſſes, Ednas, and Iozabab, and Iedihel, and Michael, and Ednas, &
Iozabad, and Eliu, and Salathi, the princes of a thouſand in
Manaſſes. \V Theſe did ayde Dauid agaynſt the rouers: for they were al
moſt valiant men, and were made commanders in the armie. \V Yea and
there came euerie day to Dauid to helpe him, til it became a great
number, as it were
\SNote{An armie of manie good men, wel diſpoſed in order and concord, is
called \Emph{the armie of God}, like the great number of Angels, which
to vs \Emph{are innumerable}.
\XRef{Iob.~25.}}
the armie of God. \V This alſo is the number of the commanders of the
armie, which came to Dauid, when he was in Hebron, to transferre the
kingdom of Saul to him, according to the word of our Lord. \V The
children of Iuda bearing ſhield and ſpeare, ſix thouſand eight hundred
wel appoynted to 
battel. \V Of the children of Simeon valiant men to fight, ſeuen
thouſand one hundred. \V Of the children of Leui, foure thouſand ſix
hundred. \V Ioiada alſo prince of the ſtocke of Aaron, and with him
three thouſand ſeuen hundred. \V Sadoc alſo a young man of goodlie
towardenes, and the houſe of his father, princes twentie two. \V And of
the children of Beniamin the brethren of Saul, three thouſand: for a
great part of them as yet folowed the houſe of Saul. \V Moreouer of the
children of Ephraim twentie thouſand eight hundred, valiant of ſtrength,
men renowmed in
%%% 0862
their kinredes. \V And of the halfe tribe of Manaſſes, eightene
thouſand, euerie one by their names, came to make Dauid king. \V Alſo of
the children of Iſſachar men of vnderſtanding, that knewe al times to
command what Iſrael ſhould doe, princes two hundred: and al the reſt of
the tribe did folow their counſel. \V Moreouer of Zabulon ſuch as went
forth to battel, & ſtood in aray wel appoynted with armour of warre,
there came fiftie thouſand to ayde, not in a duble hart. \V And of
Nephthali, commanders a thouſand: and with them furniſhed with ſhield
and ſpeare, ſeuen and thirtie thouſand. \V Of Dan alſo prepared to
battel, twentie eight thouſand ſix hundred. \V And of Aſer going forth
to fight, and prouoking in battel, fourtie thouſand. \V And beyond
Iordan of the children of Ruben, and of Gad, and the halfe part of the
tribe of Manaſſes furniſhed with armour of warre an hundred twentie
thouſand. \V Al theſe men of warre wel appoynted to fight, with perfect
hart came into Hebron, to make Dauid King ouer al Iſrael: yea and al the
reſt of Iſrael, were of one hart, that Dauid ſhould be made king. \V And
they were there with Dauid three daies eating and drinking: for their
brethren had prepared for them. \V Yea and they that were neere them as
farre as Iſſachar, & Zabulon, and Nephthali, brought loaues on Aſſes,
and on camels, and on mules, and vpon oxen, to eate: meale, figges,
reyſens, wine, oile, beeues, muttons, in al abundance, for there was ioy
in Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
