%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 0863
%%% o-0774
\startChapter[
  title={Chapter 14}
  ]

\Summary{King Dauid prouideth timber, and workmen to build his owne
  houſe; 3.~marieth more wiues, and hath manie children; 8.~ouerthroweth
  the Philiſtians, 13.~twiſe.}

Hiram alſo the king of Tyre ſent meſſengers to Dauid, and cedar trees,
and artificers for walles, and timber: to build him a houſe. \V And
Dauid knew that our Lord had eſtabliſhed him to be king ouer Iſrael, and
that his kingdom was exalted ouer his people Iſrael. \V Dauid alſo tooke
other wiues in Ieruſalem: and he begat ſonnes, and daughters. \V And
theſe be their names, that were borne to him in Ieruſalem: Samua, and
Sobad, Nathan, and Salomon, \V Iebahar, and Eliſua, and Eliphaler, \V
Noga alſo, and Napheg, and Iaphia, \V Eliſama, and Baalaida, and
Eliphaler. \V And the Philiſthijms hearing that Dauid was annoynted for
king ouer al Iſrael, they went vp al to ſeeke him: which when Dauid had
heard,
%%% 0864
he went out to meete them. \V Moreouer the Philiſthijms coming, were
ſpred abrode in the Vale Raphaim. \V And Dauid conſulted our Lord,
ſaying: Shal I goe vp to the Philiſteans, and wilt thou deliuer them
into my hand? And our Lord ſayd to him: Goe vp, and I wil deliuer them
into thy hand. \V And when they were come vp into Baalpharaſim, Dauid
ſtroke them there, and ſayd: God hath diuided myn enemies by my hand, as
waters are diuided: and therfore the name of that place was called Baal
Pharaſim. \V And they leaft there their goddes, which Dauid commanded to
be burnt. \V An other time alſo the Philiſthijms
\Fix{invaded,}{inuaded,}{obvious typo, fixed in other}
and were diſperſed in the Vale. \V And Dauid conſulted God agayne, and
God ſayd to him: Goe not vp after them, retyre from them, and thou ſhalt
come agaynſt them ouer agaynſt the pearetrees. \V And when thou ſhalt
heare the ſound of one going in the toppe of the pearetrees, then ſhalt
thou yſſue forth to battel. For God is gone forth before thee, to ſtrike
the campe of the Philiſthijms. \V Dauid therfore did as God had
commanded him, and ſtroke the campe of the Philiſthianes from Gabaon
vnto Gazera. \V And the name of Dauid was bruited in al countries, and
our Lord gaue the dreade of him ouer al nations.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
