%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 24
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-24


\project douay-rheims


%%% 0878
%%% o-0788
\startChapter[
  title={Chapter 24}
  ]

\Summary{King Dauid diſpoſeth ſixtene families of Eleazar, and eight of
  Ithamar, 7.~by lottes, 19.~to ſerue in the Temple, according to their
  prieſtlie function,
  \Fix{20.~likwiſe}{20.~likewiſe}{likely typo, fixed in other}
  principal Leuites in their offices.}

Moreouer to the ſonnes of Aaron theſe were the partitions: The ſonnes of
Aaron: Nadab, and Abiu, and Eleazar, and Ithamar. \V But Nadab and Abiu
died before their father without children: and Eleazar and Ithamar did
the function of prieſthood. \V And Dauid diuided them, that is, Sadoc of
the ſonnes of Eleazar, and Ahimelech of the ſonnes of Ithamar,
according to their courſes and miniſterie. \V And there were found manie
more children of Eleazar among the
%%% 0879
principal men, then children of Ithamar. And he diuided to them, that is, to the
children of Eleazar princes by their families ſixtene: and to the
children of Ithamar by their families and houſes eight. \V Moreouer he
diuided both families betwen themſelues by lottes: for there were
\SNote{They were called princes of the Sanctuarie in reſpect of
Sacrifices, & other ſacred functions.}
princes of the Sanctuarie, and princes
\SNote{And princes of God in reguard of their ſpiritual iuriſdiction in
the Church, or houſe of God.}
of God, as wel of the children of Eleazar, as of the children of
Ithamar. \V And Semeias wrote them, the ſonne of Nathanael the Scribe a
Leuite, before the king and princes, and Sadoc the Prieſt, and Ahimelech
the ſonne of Abiathar, the princes alſo of the Prieſtlie and Leuitical
families: one houſe, which was ouer the reſt, Eleazars: and an other
houſe, which had the reſt vnder it, Ithamars. \V And the firſt lot came
forth to Ioiarib, the ſecond to Iedei, \V the third to Harim, the fourth
to Seorim, \V the fifth to Melchia, the ſixt to Maiman, \V the ſeuenth
to Accos, the eight to Abia, \V the ninth to Ieſua, the tenth to
Sechenia, \V the eleuenth to Eliaſib, the twelfth to Iaeim, \V the
thirtenth to Hoppha, the fourtenth to Iſbaab, \V the fiftenth to Belga,
the ſixtenth to Emmer, \V the ſeuententh to Hezir, the eightenth to
Aphſes, \V the ninetenth to Pheteia, the twenteth to Hezechiel, \V the
one and twenteth to iachin, the two and twenteth to Gamul, \V the three
and twenteth to Dalaiau, the foure and twenteth to Maaziau. \V Theſe be
their courſes according to their miniſteries, to enter into the houſe of
our Lord, and according to their rite vnder the hand of Aaron their
father: as our Lord the God of Iſrael had commanded. \V Moreouer of the
children of Leui, which were remayning, there was Subael of the children
of Amram: and of the children of Subael, Iehedeia. \V Alſo of the
children of Rohobia the prince of Ieſias. \V And the ſonne of Iſaari
Salemoth, and the ſonne of Salemoth Iahath: \V and his ſonne Ieriau the
firſt, Amarias the ſecond, Iahaziel the
%%% o-0789
third, Iecmaan the fourth. \V The ſonne of Oziel, Micha: the ſonne of
Micha, Samir. \V The brother of Micha, Ieſia: and the ſonne of Ieſia,
Zacharias. \V The ſonnes of Merari: Moholi and Muſi. The ſonne of Oziau:
Benno. \V The ſonne alſo of Merari: Oziau, and Soam, and Zacchur, and
Hebri. \V Moreouer the ſonne of Moholi: Eleazar, who had no children. \V
And the ſonne of Cis, Ieramael. \V The ſonnes of Muſi: Moholi, Eder, and
Ierimoth. Theſe be the ſonnes
\Fix{of of}{of}{obvious typo, fixed in other}
Leui according to the houſes of their families. \V And they
%%% 0880
alſo did caſt lottes agaynſt their brethren the ſonnes of Aaron before
Dauid the king and Sadoc, and Ahimelech, and the princes of the
Prieſtlie and Leuitical families, as wel the elder as the yonger. Lotte
diuided al equally.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
