%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 0862
%%% o-0773
\startChapter[
  title={Chapter 13}
  ]

\Summary{The Arke is brought from Abinadabs houſe, 8.~Dauid and others
  dancing before it. 9.~Oza for touching it is ſtrooken
  dead. 13.~VVherupon Dauid fearing to bring it to Ieruſalem, it
  remayneth three monethes in the houſe of Obededom.}

And Dauid tooke counſel with the tribunes, and centurions, and al the
commanders, \V and he ſayd to al the aſſemblie of Iſrael: If it pleaſe
you: and if the word which I ſpeake procede from our Lord God, let vs
ſend to the reſt of our brethren into al the countries of Iſrael, & to
the Prieſtes, and Leuites, that dwel in the ſuburbes of the cities, that
they may be gathered together vnto vs, \V and we may bring agayne vnto
vs the Arke of our God: for we ſought it not in the daies of Saul. \V
And the whole multitude anſwered that it ſhould be ſoe done: for the
word had pleaſed al the people. \V Dauid
%%% 0863
therfore aſſembled al Iſrael from Sihor of Ægypt, til thou enter into
Emath, to bring the Arke of God from Caria Thiarim. \V And Dauid went
vp, and euerie man of Iſrael to the hil of Caria Thiarim which is in
Iuda, to fetch thence the Arke of our Lord God ſitting vpon the
Cherubim, where his name is inuocated. \V And they layd the Arke of God
vpon a new wayne, out of the houſe of Abinadab. And Oza and his brother
did driue the wayne. \V Moreouer Dauid, and al Iſrael playd before our
Lord with al their might in ſongues, and on harpes, and pſalteries, and
timbrels, and cymbals, and trumpettes. \V And when they were come to the
Floore of Chidon, Oza ſtretched forth his hand, to hold vp the Arke: for
the oxe being wantone had made it leane a litle a ſide. \V Our Lord
therfore was angrie agaynſt Oza, and ſtroke him, for that he had touched
the Arke: and he died there before our Lord. \V And Dauid was ſtrooken
ſad, becauſe our Lord had diuided Oza: and he called that place, the
Diuiſion of Oza vntil this preſent day. \V And he feared God at that
time, ſaying: How may I bring in the Arke of God vnto me? \V And for
this cauſe he brought it not vnto himſelfe, that is, into the citie of
Dauid, but turned it away into the houſe of
\SNote{Obededom was a Leuite,
\XRef{(c.~15. v.~18.)}
and therfore more mete to kepe the arke.}
Obededom the
\SNote{Called a Getheite becauſe he had dwelt in the towne of Geth.}
Getheite. \V Therfore the Arke of God remayned in the houſe of Obededom
three monethes: & our Lord bleſſed his houſe, and al thinges that he
had.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
