%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 0853
%%% o-0765
\startChapter[
  title={Chapter 8}
  ]

\Summary{The progenie of Beniamin is further recited vnto Saul: 33.~and
  his iſſue.}

And Beniamin begat Bale his firſt begotten, Aſbel the ſecond, Ahara the
third, \V Nohaa the fourth, and Rapha the fifth. \V And the ſonnes of
Bale were: Addar, and Gera, and Abiud, \V Abiſue alſo, and Naaman, and
Ahoe, \V and alſo Gera, and Sephuphan, and Huram. \V Theſe are the
ſonnes of Ahod, princes of their kinredes that dwelt in Gabaa, which
were tranſported into Manahath. \V And Naaman, and Achia, and Gera he
tranſported them, and begat Oza, and Ahiud. \V Moreouer he begat
Saharaim in the countrie of Moab, after he diſmiſſed Huſim and Bara his
wiues. \V And of Hodes his wife he begat Iobab, and Sebia, and Moſa, and
Molehom, \V Iehus alſo, and Sechia, and Marma. Theſe are his ſonnes
princes in their families. \V And Mehuſim begat Abitob, and Elphaal. \V
Moreouer the ſonnes of Elphaal: Heber, and Miſaam, and Samad. This man
built Ono, and Lod, and her daughters. \V And Baria, and Sama princes of
their kinredes that dwelt in Aialon: theſe droue away the inhabitantes
of Geth. \V And Ahio, and Seſac, and Ierimoth, \V and Sabadia, and Arod,
and Heder, \V Michael alſo, and Ieſpha, and Ioha the ſonnes of Baria. \V
And Zabadia, and Moſollam, and Hezeci, and Heber, \V and Ieſamari, and
Iezlia, and Iobab
%%% 0854
ſonnes of Elphaal, \V and Iacim, and Zechri, and Zabdi, \V and Elioenai,
and Selethai, and Eliel, \V and Adaia, and Baraia, and Samarath the ſonnes
of Semei. \V And Ieſpham, and Heber, and Eliel, \V and Abdon, and
Zechri, and Hanan, \V and Hanania, and Ælam, and Anathothia, \V and
Iephdaia, and Phanuel the ſonnes of Seſac. \V And Samſari, and Sohoria
and Otholia, \V and Ierſia, and Elia, and Zechri, the ſonnes of
Ieroham. \V Theſe be the
\Fix{patriaches,}{patriarches,}{obvious typo, fixed in other}
and princes of their kinredes,
which dwelt in Ieruſalem. \V And in Gabaon dwelt Abigabaon, and the name
of his wife Maacha. \V And his firſt begotten ſonne Abdon, and Sur, and
Cis, and Baal, and Nadab. \V Gedor alſo, and Ahio, and Zacher, and
Macelloth. \V And Macelloth begat Samaa: and they dwelt ouer agaynſt
their brethren in Ieruſalem with their brethren. \V And Ner begat Cis,
and Cis begat Saul. Moreouer Saul begat Ionathas, and Melchiſua, and
Abinadab, and Eſbaal. \V And the ſonne of Ionathas, Meribbaal: and
Meribbaal begat Micha. \V The ſonnes of Micha: Phithon, and Melech, &
Tharaa, and Ahaz. \V And Ahaz begat Ioada: and Ioada begat Alamath, and
Azmoth, and Zamri: moreouer Zamri begat Moſa, \V and Moſa begat Banaa,
whoſe ſonne was Rapha, of whom was borne Elaſa, who begat Aſel. \V
Moreouer Aſel had ſix ſonnes of theſe names: Ezricam, Bochru, Iſmahel,
Saria, Obdia,
%%% o-0766
and Hanan. Al theſe the ſonnes of Aſel. \V And the ſonnes of Eſec his
brother: Vlam the firſtbegotten, and Iehus the ſecond, and Eliphalet the
third. \V And the ſonnes of Vlam were moſt ſtrong men, and archers of
great force: and hauing many ſonnes and nephewes, vnto an hundred
fiftie. Al theſe the children of Beniamin.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
