%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 0872
%%% o-0782
\startChapter[
  title={Chapter 20}
  ]

\Summary{King Dauid proſpereth in warre agaynſt the Ammonites, 4.~and
  Philiſthims, 6.~among whom Ionathan Dauids nephew by his brother,
  killeth a monſtruous giant, which had twelue fingers and twelue toes.}

And it came to paſſe after the compaſſe of a yeare,
\CNote{\XRef{2.~Reg.~21.}}
at ſuch time when kinges are wont to goe forth to battel, Ioab gathered
together the armie and force of warre, and ſpoyled the land of the
children of Ammon: and went on and beſieged Rabba. Moreouer Dauid taried
in Ieruſalem, when Ioab ſtroke Rabba, and deſtroyed it. \V And Dauid
tooke the crowne of Melchom from his head, and found in it a talent
weight of gold, and moſt precious pearles, and he made himſelf therof a
diademe: he tooke alſo the ſpoiles of the citie very much. \V And the
people that was therin, he
%%% 0873
brought forth: and he made harrowes, and ſleddes, and chariotes ſhod
with Iron to paſſe ouer them, ſoe that they were cut in ſunder, & broken
in peces: ſo did Dauid to al the cities of the children of Ammon: & he
returned with al his people into Ieruſalem. \V After theſe thinges there
was warre begunne in Gazer agaynſt the Philiſthians: in which Sobachai
the Huſathite ſtroke Saphai of the kinred of Raphaim, & humbled them. \V
An other battel alſo was fought agaynſt the Philiſtheans, wherein
Adeodarus the ſonne of Saltus a Bethlehemite ſtroke the brother of
Goliath the Getheite, the ſtaffe of whoſe ſpeare was as it were a
weauers beame. \V But an other battel alſo happened in Geth, wherin
there was a verie long man, hauing fingers and toes by ſix and ſix, that
is, together foure and twentie: who alſo was borne
%%% o-0783
of the ſtocke of Rapha. \V This man blaſphemed Iſrael: and Ionathan the
ſonne of Samaa the brother of Dauid ſtroke him. Theſe be the children of
Rapha in Geth, which fel by the hand of Dauid and of his ſeruantes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
