%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 0858
%%% o-0769
\startChapter[
  title={Chapter 11}
  ]

\Summary{Dauid
\MNote{The third part.

King Dauids reigne and his ſpecial actes.}
is elected and annointed king. 5.~He ouerthroweth the
  Iebuſites taking the towre of Sion in Ieruſalem, and
  proſpereth. 10.~His valiant men, and their heroical actes are
  recited. 17.~He deſiring water from the ciſterne of Bethleem, wil not
  drinke it, but offereth it in ſacrifice, becauſe it is brought with
  danger of his valiant mens liues. 20.~Other valiant men of the ſecond
  order, are likewiſe recited.}

Al Iſrael therfore was gathered together to Dauid in Hebron, ſaying: We
are thy bone, and thy fleſh. \V Yeſterday alſo, and the day before when
Saul as yet reigned, thou waſt he that didſt leade out and leade in
Iſrael: for to thee our Lord thy God ſayd: Thou ſhalt feede my people
Iſrael, and thou ſhalt be prince ouer it. \V Al the ancientes therfore
of Iſrael came to the king into Hebron, and Dauid entred into a league
with them before our Lord: and they annoynted him king ouer Iſrael,
according to the word of our Lord, which he ſpake in the hand of
Samuel. \V Dauid alſo went, and al Iſrael into Ieruſalem. This is Iebus,
where the Iebuſeites were the inhabiters of the land. \V And they that
dwelt in Iebus, ſayd to Dauid: Thou ſhalt not come in here. Moreouer
Dauid tooke the towre of Sion, which is the Citie of Dauid, \V and he
ſayd: Euerie one that ſhal among the firſt ſtrike the Iebuſeite, ſhal be
the prince and chiefe captayne. Ioab therfore the ſonne of Saruia went
vp firſt, and was made the prince. \V And Dauid dwelt in the towre, and
therfore it was called the Citie of Dauid. \V And he built the citie
round about from Mello vnto a round compaſſe, and Ioab built the reſt of
the citie. \V And Dauid proſpered going and increaſing, and the Lord of
hoſtes was with him. \V Theſe are the princes of the valiant men of
Dauid, which holpe him to be made king ouer al Iſrael according to the
word of our Lord, which he ſpake to Iſrael. \V And this is the number of
Dauids ſtrong ones: Ieſbaam the ſonne of Hachamoni prince among thirtie:
this man lifted vp his ſpeare vpon three hundred wounded at one time. \V
And after him Eleazar his vncles ſonne an Ahohite, who was among the
three mighties. \V This was with Dauid in Pheſdomim, when the
Philiſthijms were gathered to that place into battel: and the field of
that countrie was ful of barley, and the people was fled from the face
of the Philiſthimes. \V Theſe ſtood in the middes of the fielde, and
%%% 0859
defended him: and when they had ſtriken the Philiſtheans, our Lord gaue
great health to his people. \V And there went downe three of the thirtie
princes to a rocke, wherin Dauid was, to the caue of Odollam, when
%%% o-0770
the Philiſthijms had camped in the Vale raphaim. \V Moreouer Dauid was
in a hold, and the ward of the Pheliſthimes in Bethlehem. \V Dauid
therfore deſired and ſayd: O that ſome man would giue me water of the
ceſterne of Bethlehem, which is in the gate. \V Theſe three therfore
went forward through the middes of the campe of the Philiſthimes, and
drewe water of the ceſterne of Bethlehem, which was in the gate, and
brought it to Dauid to drinke: who would not, but rather offered it to
our Lord, \V ſaying: God forbid that in the ſight of my God I ſhould doe
this, and ſhould drinke the bloud of theſe men: becauſe in the peril of
their liues they haue brought me the water. And for this cauſe he would
not drinke. Theſe thinges did the three moſt valiantes. \V Abiſai alſo
the brother of Ioab he was prince of three, and he lifted vp his ſpeare
agaynſt three hundred wounded, and he was among three moſt renowned, \V
and among the ſecond three he the noble one, and prince of them: but yet
vnto the three firſt he raught not. \V Banaias the ſonne of Ioiada a
moſt valiant man, of Cabſeel: who had done manie factes, he ſtroke the
\SNote{Two ſtout men of Moab, as if they had benne lions.
\XRef{2.~Reg.~23.}}
two ariel of Moab: and he went downe, and ſlewe the lion in the middes
of the ciſterne in the time of ſnow. \V And he
\Fix{ſtorke}{ſtroke}{obvious typo, fixed in other}
the Ægyptian, whoſe ſtature was of fiue cubites, and which had a ſpeare
as the weauers beame: he therfore went downe to him with a rod, and by
force tooke away the ſpeare, that he held in his hand, and ſlewe him
with his owne ſpeare. \V Theſe thinges did Banaias the ſonne of Ioiada,
who was among the three valiantes moſt renowned, \V the firſt among
thirtie, but yet the three he raught not: and Dauid made him of his
counſel. \V Moreouer the moſt valiant men in the armie, Aſahel the
brother of Ioab, and Elchanan the ſonne of his vncle of Bethlehem, \V
Sammoth an Arorite, Helles a Phalonite, \V Ira
\Fix{the the}{the}{obvious typo, fixed in other}
ſonne of Acces a Thecaite, Abiezer an Anathothite, \V Sobbochai an
Huſathite, Ilai an Ahohite, \V Maharai a Netophathite, Heled the ſonne
of Baana a Netophathite, \V Ethai the ſonne of Ribai of Gabaath the
children of Beniamin, Banaia a Pharatonite, \V Hurai of the Torrent
Gaas, Abiel an Arbathite, Azmoth a Bauramite, Eliaba a Salabonite.
%%% 0860
\V The ſonnes of Aſſem a Gezonite, Ionathan the ſonne of Sage an
Ararite, \V Ahiam the ſonne of Sachar an Ararite, \V Eliphal the ſonne
of Vr, \V Hepher a Mecherathite, Ahia Phelonite, \V Heſro a Carmelite,
Naarai the ſonne of Azbai, \V Ioel the brother of Nathan, Mibahar the
ſonne of Agarai. \V Selet an Ammonite, Naharai a Berothite the armour
bearer of Ioab the ſonne of Saruia, \V Ira a Iethreite, Gareb a
Iethreite, \V Vrias the Hetheite, Zabad the ſonne of Oholi, \V Adina the
ſonne of Sizaa Rubenite the prince of the Rubenites, & with him
thirtie: \V Hanan the ſonne of Maacha, and Ioſaphat a Mathanite, \V Ozia
an Aſtarothite, Samma, and Iehiel the ſonnes of Hotham an Arorite, \V
Iedihel the ſonne of Zamri, and Ioha his brother a Thoſaite, \V Eliel a
Mahumite, and Ieribai, and Ioſaia the ſonnes of Elnaim, and
%%% o-0771
Iethma a Moabite, Eliel, and Obed, and Iaſiel of Maſobia.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
