%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 27
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-27


\project douay-rheims


%%% 0883
%%% o-0792
\startChapter[
  title={Chapter 27}
  ]

\Summary{Twelue militarie tribunes, with foure and twentie thouſand
  vnder euerie one, by courſe of monethes ſerue about the
  king. 16.~Other twelue tribunes of the people gouerne in ſeueral
  tribes. 25.~Prefectes alſo are made ouer the kinges treaſures,
  cattles, and other commodities in the court, campe, and countrie.}

And the children of Iſrael according to their number, the princes of
families, tribune, and centurions, and ouerſeers, that miniſtred to the
king according to their companies, coming in and going out euerie moneth
in the yeare, euerie one was chiefe ouer foure and twentie thouſand. \V
Ouer the firſt companie the firſt moneth Ieſboam was chiefe the ſonne of
Zabdiel, and vnder him foure and twentie thouſand. \V Of the ſonnes of
Phares, the prince of al the princes in the hoſte the firſt moneth. \V
The companie of the ſecond moneth had Dudia an Ahohite, and after him an
other named Macelloth, which gouerned part of foure and twentie
thouſand. \V Captayne alſo of the third companie in the third moneth,
was Banaias the ſonne of Ioiada the Prieſt: and in his diuiſion foure
and twentie thouſand. \V The ſame is Banaias the moſt valiant among
thirtie, and aboue the thirtie, & ouer his companie Amizabad his ſonne
was chiefe. \V The fourth, in the fourth moneth, Aſahel the brother of
Ioab, and Zabadias his ſonne after him: and in his companie foure and
twentie thouſand. \V The fifth, in the fifth moneth, prince Samaoth a
Iezerite: and his companie foure and twentie thouſand. \V The ſixth, in
the ſixth moneth, Hira the ſonne of Acces a Thecuite: and in his
companie foure and twentie thouſand. \V The ſeuenth, in the ſeuenth
moneth, Helles a Phallonite of the children of Ephraim: and in his
companie
%%% 0884
foure and twentie thouſand. \V The eight, in the eight moneth, Sobochai
an Huſathite of the ſtocke of Sarahi: and in his companie foure and
twentie thouſand. \V The ninth, in the ninth moneth, Abiezer an
Anathothite of the children of Iemini: and in his companie foure and
twentie thouſand. \V The tenth, in the tenth moneth, Marai, and he a
Netophathite of the ſtocke of Zarahi: and in his companie foure and
twentie thouſand. \V The eleuenth, in the eleuenth moneth, Banaias, a
Pharathonite of the children of Ephraim: and in his companie foure and
twentie thouſand. \V The twelfth, in the twelfth moneth, Holdai a
Netophathite, of the ſtocke of Gothoniel: and in his companie foure and
twentie thouſand. \V Furthermore there were chiefe ouer the tribes of
Iſrael, ouer the Rubenites, Duke Eliezer the ſonne of Zechri: ouer the
Simeonites, Duke Sapharias the ſonne of Maacha: \V ouer the Leuites,
Haſabias
%%% o-0793
the ſonne of Camuel: ouer the Aaronites, Sadoc: \V ouer Iuda, Eliu the
brother of Dauid: ouer Iſſachar, Amri the ſonne of Michael: \V ouer the
Zabulonites, Ieſmaias the ſonne of Abdias: ouer the Nepthalites,
Ierimoth the ſonne of Ozriel: \V ouer the children of Ephraim, Oſee the
ſonne of Ozaziu: ouer the halfe tribe of Manaſſes, Ioel the ſonne of
Phadaia: \V and ouer the halfe tribe of Manaſſes in Galaad, Iaddo the ſonne
of Zacharias: and ouer Beniamin, Iaſiel the ſonne of Abner. \V But ouer
Dan, Ezrihel the ſonne of Ieroham: theſe be the princes of the children
of Iſrael. \V And Dauid
\SNote{King Dauid hauing hertofore offended in commanding to number the
whole people,
\XRef{2.~Reg.~24.}
would not now attempt to number al (which were in deede innumerable) but
only thoſe that were to be diſpoſed of in ſpecial ſeruices.}
would not number them from twentie yeares downeward: becauſe our Lord
had ſayd that he would multiplie Iſrael
%%% !!! LNote marked in both, but missing in both
%%% \LNote{}{}
as the ſtarres of heauen. \V Ioab the ſonne of Saruia beganne to number,
and finiſhed not: becauſe vpon this there fel wrath vpon Iſrael: and
therfore the number of them that were numbered, was not regeſtred in
the chronicles of king Dauid. \V And ouer the kinges treaſures was
Azmoth the ſonne of Adiel, and ouer thoſe treaſures, which were in the
cities, and in the townes, and in the towers, Ionathan the ſonne of
Ozias was preſident. \V And ouer huſbandrie, and the huſbandmen, which
tilled the ground, Ezri the ſonne of Chelub was ouerſeer: \V and ouer
the dreſſers of vineyardes, Semeias a Romathite: and ouer the wine
cellars, Zabdias an Aphonite. \V For ouer the Oliuetes and the
figgegroues, which were in the champayne, was Balanam a Gederite: and
ouer the oile cellars, Ioas. \V Furthermore ouer
%%% 0885
the heardes that fed in Saron, the ouerſeer was Setrai a Saronite: and
ouer the oxen in the valleis Saphat the ſonne of Adli: \V but ouer the
camels, Vbil an Iſmaelite: and ouer the aſſes, Iadias a Meronathite: \V
ouer the ſheepe alſo Iaziz an Agarene. Al theſe, princes of the
ſubſtance of king Dauid. \V And Ionathan Dauids vncle, a counſeler, a
man wiſe and learned: He and Iahiel the ſonne of Hachamoni were with the
kinges ſonnes. \V Achitophel alſo the kinges counſeler, and Chuſai an
Arachite the kinges freinde. \V After Achitophel was Ioiada the ſonne of
Banaias, and Abiathar. And the prince of the kinges armie was Ioab.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
