%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-23


\project douay-rheims


%%% 0877
%%% o-0786
\startChapter[
  title={Chapter 23}
  ]

\Summary{King Dauid being old conſtituteth Salomon king, 3.~diſpoſeth
  the offices of Leuites, 7.~to wit, the families of Gerſon, 12.~of
  Caath, 21.~and of Merari: (26.~ceaſsing to carrie the tabernacle.)
  27.~to ſerue in the temple.}

Dauid therfore being old & ful of daies, made Salomon his ſonne king
ouer Iſrael. \V And he gathered al the princes of Iſrael, and the
Prieſtes and Leuites. \V And the Leuites were numbred from thirtie
yeares, & vpward: and there were found thirtie eight thouſand men. \V
Of theſe were choſen, and diſtributed into the miniſterie of the houſe
of our Lord foure and twentie thouſand: and of the ouerſeers and iudges,
ſix thouſand. \V Moreouer foure thouſand porters: and as manie ſinging
to our Lord on inſtrumentes, which he had made to ſing on. \V And
\SNote{Aarons ſonnes being conſecrated Prieſtes according to Gods
ordinance,
\XRef{Leuit.~8.}
the Leuites to do other offices about the tabernacle.
\XRef{Num.~3. &.~4.}
King Dauid (with Sadoc the high prieſt, and other chief men
\XRef{ch.~24. v.~6.)}
diſpoſed them by lottes which ſhould ſerue by courſes, to ſing and play
on inſtrumentes, as wel in the tabernacle now reſting in Ieruſalem,
\XRef{v.~26.}
as in the Temple when it ſhould be built.}
Dauid diſtributed them by the courſes of the children of Leui, to witte,
of Gerſon, and Caath, and Merari. \V The ſonnes of Gerſon: Leedan, and
Semei. \V The ſonnes of Leedan: the prince Iahiel, & Zethan, and Ioel,
three. \V The ſonnes of Semei: Salomith, and Hoſiel, and Aram,
three. Theſe be the princes of the families of Leedan. \V Moreouer the
ſonnes of Semei: Leheth, and Ziza, and Iaus, and Baria: theſe be the
ſonnes of Semei, foure. \V And Leheth was the firſt, Ziza the ſecond:
moreouer Iaus and Baria had not manie children, and therfore they were
counted in one familie, and in one houſe. \V The children of Caath:
Amram, and Iſaar, Hebron, and Oziel, foure. \V The ſonnes of Amram:
Aaron, and Moyſes. And Aaron was ſeperated to miniſter in \L{Sancta
Sanctorum}, he and his
%%% o-0787
ſonnes for euer, and to burne incenſe to our Lord according to his rite,
and to bleſſe his name for euer. \V The ſonnes of Moyſes alſo the man of
God were numbred in the tribe of Leui. \V The ſonnes of Moyſes: Gerſom
and
\Fix{Eliezari.}{Eliezer.}{obvious typo, fixed in other}
The ſonnes of Gerſom: Subuel the firſt. \V And the ſonnes of Eliezer
were: Rohobia the firſt. And
\Fix{Eleezer}{Eliezer}{obvious typo, fixed in other}
had noe moe ſonnes. Moreouer the children of Rohobia were multiplied
excedingly. \V The ſonnes of Iſaar: Salomith the firſt. \V The ſonnes of
Hebron: Ieriau the firſt, Amarias the ſecond, Iahaziel the third,
Iecmaan the fourth. \V The ſonnes of Oziel: Micha the firſt, Ieſia the
ſecond. \V The ſonnes of Merari: Moholi, & Muſi. The ſonnes of Moholi:
Eleazar and Cis. \V And Eleazar died, and had no ſonnes but daughters,
%%% 0878
and the ſonnes of Cis their brethren tooke them. \V The ſonnes of Muſi:
Moholi, and Eder, and Ierimoth, three. \V Theſe be the children of Leui
in their kinredes and families, princes by courſes, and number of euerie
head, that did the workes of miniſterie of the houſe of our Lord from
twentie yeares, and vpward. \V For Dauid ſayd: Our Lord the God of
Iſrael hath geuen reſt to his people, and the habitation of Ieruſalem
for euer. \V Neither ſhal it be the office of the Leuites to carie any
more the tabernacle, and al the veſſels therof to miniſter. \V According
to the laſt preceptes alſo of Dauid the number of the children of Leui
ſhal be numbred from twentie yeares and vpward. \V And they ſhal be
vnder the hand of the ſonnes of Aaron for the ſeruice of the houſe of
our Lord, in the entrances, and in the chambers, and in the place of
purification, and in the Sanctuarie, and in al the workes of the
miniſterie of the temple of our Lord. \V And the Prieſtes, ouer the
loaues of propoſition, and for the ſacrifice of floure, and for cakes
and azimes, and the frying panne, and to roſt, and ouer al weight and
meaſure. \V But the Leuites to ſtand in the morning to confeſſe, and
ſing to our Lord: and in like maner at euening, \V as wel in the
oblation of the holocauſtes of our Lord, as in the Sabbathes and
Calendes, and the reſt of the ſolemnities, according to the number, and
ceremonies of euerie thing, continually before our Lord. \V And let them
keepe the obſeruations of the tabernacle of couenant, and the rite of
the Sanctuarie, and the obſeruance of the children of Aaron their
brethren, that they miniſter in the houſe of our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
