%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 0866
%%% o-0776
\startChapter[
  title={Chapter 16}
  ]

\Summary{The Arke is placed in a tabernacle. Sacrifice is offered. Dauid
  bleſſeth the people, 4.~diſpoſeth the offices of Leuites, 8.~and
  maketh a Pſalme of praiſe to God.}

They therfore brought the Arke of God, and ſet it in the middes of the
tabernacle, which Dauid had pitched for it: and they offered
holocauſtes, and pacifiques before God. \V And when Dauid had finiſhed
offering holocauſtes, and pacifiques, he
\SNote{Not only king Dauid being a holie Prophet, but anie other
Superior might bleſſe his ſubiectes.}
bleſſed the people in the name of our Lord. \V And he
\Fix{duided}{diuided}{obvious typo, fixed in other}
to al through out euerie one, from man vnto woman a loafe of bread, and
a peece of roſted beefe, and
\Fix{floore}{floure}{likely typo, fixed in other}
fryed with oile. \V And he
\SNote{That he alſo diſpoſed certaine offices of Leuites was by ſpecial
priuilege, which was no preiudice to the hiegh prieſts authoritie, for
ſuperior powre is proued by Gods inſtitution, rather then by factes,
either of good men, which do manie thinges by way of diſpenſation, or of
euil vſurping without warrant, that to them perteyneth not. For it is
clere that God inſtituted ſupreme ſpiritual powre in the high prieſt.
\XRef{Deut.~17.}
And al kinges and temporal princes are to receiue the law at the
prieſtes hand.
\XRef{ibidem. v.~18.}
Eleazar the high prieſt was oppointed to conſult our Lord for Ioſue.
\XRef{Nu.~27.}
Finally by Gods ordinance, \Emph{the lavv of truth vvas in the mouth of
prieſtes}.
\XRef{Malac.~2.}}
appoynted before the Arke of our Lord of the Leuites, that ſhould
miniſter, and ſhould remember his workes, & glorifie, and prayſe our
Lord the God of Iſrael: \V Aſaph the prince, and the ſecond after him
Zacharias: moreouer Iahiel, and Semiramoth, and Iehiel, and Mathathias,
and Eliab, and Banaias, & Obededom: and
%%% o-0777
Iehiel ouer the inſtrumentes of pſalterie, and the harpes: & Aſaph to
ſound vpon the cymbals: \V but Banaias, and Iaziel prieſtes, to ſound
the trumpet continually before the Arke of the couenant of our Lord. \V
In that day Dauid made Aſaph, prince to confeſſe to our Lord and his
brethren.

\V Confeſſe ye to our Lord, and inuocate his name: make his inuentions
knowen among the peoples.

\CNote{\XRef{Pſal.~104.}}
\V Chaunt to him, and ſing to him: and tel ye al his meruelous thinges.

\V Prayſe ye his holie name: let the hart of them that reioyſe, ſeeke
our Lord.

\V Seeke ye our Lord, and his powre:
%%% 0867
ſeeke ye his face alwayes.

\V Remember his meruelous thinges which he hath done: his ſignes, and
the iudgementes of his mouth.

\V The ſeede of Iſrael his ſeruant: the children of Iacob his elect.

\V He is the Lord our God: in al the earth are his iudgementes.

\V Remember for euer his couenant: the word, which he commanded vnto a
thouſand generations.

\V Which he couenanted with Abraham: and his othe with Iſaac.

\V And he appoynted it to Iacob for a precept: and to Iſrael for an euer
laſting couenant:

\V Saying: To thee wil I geue the Land of Chanaan: the corde of your
inheritance.

\V When they were few in number: ſmal and ſoiourners therof.

\V And they paſſed from nation into nation: & from kingdom to an other
people.

\V He ſuffered not any man to calumniate them: but rebuked kinges for
their ſake.

\V Touch not my annoynted: and vnto my prophetes be not malicious.

\CNote{\XRef{Pſal.~95.}}
\V Sing ye to our Lord al the earth: ſhew forth from day to day his
ſaluation.

\V Tel his glorie among the gentiles: among al
\Fix{peoles}{peoples}{obvious typo, fixed in other}
his meruelous workes.

\V Becauſe our Lord is great, and laudable excedingly: and terrible ouer
al goddes.

\V For al the goddes of the peoples, be idols: but our Lord made the
heauens.

\V Confeſſion and magnificence before him: ſtrength and ioy in his
place.

\V Bring to our Lord ye families of peoples: bring to our Lord glorie
and impire.

\V Geue our Lord glorie, to his name, eleuate ſacrifice, and come ye in
his ſight: and adore our Lord in holie honour.

\V Let al the earth be moued before his face: for he founded the world
vnmoueable.

%%% o-0778
\V Let the heauens be glad, & the earth reioyce: and let them ſay among
the nations, Our Lord hath reigned.

\V Let the ſea thunder, and the fulneſſe therof: let the fieldes
reioyſe, and al thinges that are in them.

\V Then ſhal the trees of the foreſt prayſe before our Lord: becauſe he
is come to iudge the earth.

\CNote{\XRef{Pſal.~105.}}
\V Confeſſe ye to our Lord, becauſe he is good: becauſe his mercie is
for euer.

\V And ſay ye: Saue vs ô God our ſauiour: and gather vs together, &
deliuer vs out of the nations, that we may confeſſe to thy holie name,
and may reioyſe in thy ſongues.

\V Bleſſed be our Lord the God of Iſrael from eternitie vnto eternitie:
and let al the people ſay: Amen, and hymne to God.

\V He therfore leaft Aſaph there before the Arke of the couenant of our
Lord, and his brethren to miniſter in the preſence
%%% 0868
of the Arke continually day by day, and in their courſes. \V Moreouer
Obededom, and his brethren ſixtie eight: and Obededom the ſonne of
Idithun, and Hoſa he appoynted for porters. \V And Sadoc the prieſt, and
his brethren prieſtes, before the tabernacle of our Lord in the excelſe,
which was in Gabaon, \V that they ſhould offer holocauſtes to our Lord
vpon the altar of holocauſt continually, morning and euening, according
to al thinges that are writen in the law of our Lord, which he commanded
Iſrael. \V And after him Heman, and Idithun, and the reſt of the choſen
men, euerie one by his name to confeſſe vnto our Lord: Becauſe his
mercie is for euer. \V Heman alſo and Idithun ſounding the trumpet, and
quauering on the cymbals, and al muſical inſtrumentes to ſing vnto God:
and the ſonnes of Idithun he made porters. \V And al the people returned
into their houſe: and Dauid, to bleſſe alſo his houſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
