%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-22


\project douay-rheims


%%% 0875
%%% o-0785
\startChapter[
  title={Chapter 22}
  ]

\Summary{VVorkemen and al neceſſaries being prepared, 6.~Dauid commandeth
  Salomon to build the Temple, for ſo God hath appoynted. 13.~Exhorteth
  him to ſerue God, 17.~and other principal men to aſsiſt him.}

And Dauid ſayd: This is the houſe of God, and this is an altar for
holocauſte to Iſrael. \V
\CNote{\XRef{2.~Reg.~7.}
\XRef{3.~Reg.~5.}}
And he commanded that the proſelytes of the land of Iſrael ſhould be
gathered together, and he appoynted of them maſons to hewe ſtones and
poliſh them, that the houſe of God might be built. \V Dauid prepared
alſo verie much yron for the nayles of the gates, and for the ioyninges
and ioynctures: and of braſſe an innumerable weight. \V The cedar trees
alſo could not be eſtemed, which the Sidonians, and Tyrians brought
downe to Dauid. \V And Dauid ſayd: Salomon my ſonne is yet a litle child
and delicate, and the houſe which I would haue to be builded to our
Lord, muſt be ſuch as may be renowned in al countries: I therfore wil
prepare him neceſſaries. And for this cauſe before his death he prepared
al the expenſes. \V And he
%%% 0876
called Salomon his ſonne: and commanded him that he ſhould build a houſe
to our Lord the God of Iſrael. \V And Dauid ſayd to Salomon: My ſonne,
it was my wil to haue built a houſe to the name of our Lord my God. \V
But the word of our Lord was made to me, ſaying: Thou haſt ſhed much
bloud, and fought verie manie battels, thou canſt not build a houſe
to my name, ſo much bloud being ſhed before me: \V the ſonne, which ſhal
be borne to thee, ſhal be a moſt quiet man: for I wil make him reſt from
al his enemies round about: and for this cauſe, he ſhal be called
\SNote{Salomon ſignifieth Peaceable, and therin, as in manie other
thinges, prefigured Chriſt who amongſt other names is called,
\Emph{Prince of peace}.
\XRef{Iſai.~9.}}
Peaceable: and I wil geue peace and quietneſſe in Iſrael al his
dayes. \V He ſhal build a houſe to my name, & he ſhal be to me for a
ſonne, and I wil be to him for a father: and I wil
\Fix{eſtablih}{eſtabliſh}{obvious typo, fixed in other}
the throne of his kingdom ouer Iſrael for euer. \V Now therfore my
ſonne, Our Lord be with thee, and doe thou proſper, and build the houſe
to our Lord thy God, as he hath ſpoken of thee. \V Our Lord alſo giue
thee wiſdome and vnderſtanding, that thou mayſt be able to rule Iſrael,
and to kepe the law of our Lord thy God. \V For then thou ſhalt be able
to proſper, if thou ſhalt keepe the commandementes, and iudgementes,
which our Lord commanded Moyſes to teach Iſrael: take courage and play
the man, feare not, neither be diſmayde. \V Behold I in my pouertie haue
prepared the charges of the houſe of our Lord, of gold an hundred
thouſand talentes, and of ſiluer a thouſand thouſand talentes: but of
braſſe, and of yron there is noe weight, for the number is ſurpaſſed
with the greatneſſe: timber and ſtones I haue
%%% o-0786
prepared to al the charges. \V Thou haſt alſo verie manie artificers,
hewers of ſtones, and maſons, & carpenters, and of al occupations moſt
skilful to make worke, \V in gold and ſiluer and braſſe and yron, wherof
there is no number. Ryſe therfore and do it, and our Lord wil be with
thee. \V Dauid alſo commanded al the princes of Iſrael, that they ſhould
help Salomon his ſonne. \V You ſee, quoth he, that our Lord your God is
with you, and hath geuen you reſt round about, and hath deliuered al
your enemies into your handes, and the land is ſubdewed before our Lord,
and before his people. \V Geue therfore your hart and your ſoules, to
ſeeke our Lord your God: and ariſe, and build a ſanctuarie to our Lord
God, that the Arke of the couenant of our Lord, and the veſſels
conſecrated to our Lord, may be brought into the houſe, which is built
to the name of our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
