%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 0864
%%% o-0775
\startChapter[
  title={Chapter 15}
  ]

\Summary{VVith ſolemnitie the Arke is brought into Ieruſalem, caried by
  the Prieſtes and Leuites, 16.~with muſick of diuers
  ſortes. 26.~Sacrifice of thankeſgeuing is offered. 29.~Michol derideth
  Dauids deuotion.}

He made alſo houſes for him ſelfe in the citie of Dauid: and built a
place for the Arke of God, & pitcht a tabernacle for it. \V Then ſayd
Dauid: It is vnlawful that the Arke of God be caried of any man, but of
the Leuites: whom our Lord choſe to carie it, and to miniſter vnto
himſelf for euer. \V And he gathered together al Iſrael into Ieruſalem,
that the Arke of God might be brought into his place, which he had
prepared for it. \V Moreouer alſo the ſonnes of Aaron, and the
Leuites. \V Of the children of Caath, Vriel was the prince: and his
brethren an hundred twentie. \V Of the ſonnes of Merari, Aſaia the
prince: and his brethren two hundred twentie. \V Of the ſonnes of
Gerſom, Ioel the prince: and his brethren an hundred thirtie. \V Of the
ſonnes of Eliſaphan, Semeias the prince: and his brethren two hundred. \V
Of the
%%% 0865
ſonnes of Hebron, Eliel the prince: and his brethren eightie. \V Of the
ſonnes of Oziel, Aminadab the prince: and his brethren
\Fix{and}{an}{obvious typo, fixed in other}
hundred twelue. \V And Dauid called Sadoc, and Abiathar the Prieſtes,
and the Leuites, Vriel, Aiaia, Ioel, Semeia, Eliel, and Aminadab: \V and
he ſayd to them: You that are the princes of the Leuitical families, be
ſanctified with your brethren, and fetch the Arke of our Lord the God of
Iſrael to the place, which is prepared for it: \V leſt as from the
beginning, becauſe
\Fix{yow}{you}{obvious typo, fixed in other}
were not preſent, our Lord ſtrike vs: ſo now alſo it come to paſſe, we
doing ſome vnlawful thing. \V The Prieſtes therfore, and the Leuites
were ſanctified, to carie the Arke of our Lord the God of Iſrael. \V And
the ſonnes of Leui tooke the Arke of God, as Moyſes had commanded,
according to the word of our Lord, vpon their ſhoulders, on barres. \V
And Dauid ſayd to the princes of the Leuites, that they ſhould appoynt
of their brethren ſinging men on muſical inſtrumentes, to witte, on
nables, & harpes, and cymbals, that the ſound of ioy might reſound on
high. \V And they appoynted Leuites: Hemam the ſonne of Ioel, and of his
brethren Aſaph the ſonne of Barachias: and of the children of Merari,
their brethren: Ethan the ſonne of Caſaia. \V And with them their
brethren: in the ſecond order, Zacharias, and Ben, and Iaziel, and
Semiramoth, and Iahiel, and Ani, Eliab, and Banaias, and Maaſias, and
Mathathias, and Eliphalu, and Macenias, and Obededom, and Iehiel,
porters. \V Moreouer them that ſang, Heman, Aſaph, and Ethan:
%%% o-0776
ſounding on braſen cymbals. \V And Zacharias, and Oziel, and Semiramoth,
and Iahiel, & Ani, and Eliab, and Maaſias, and Banaias vpon nables ſang
myſteries. \V Moreouer Mathathias, and Eliphalu, and Macenias, and
Obededom, and Iehiel, and Ozaziu, vpon harpes for the octaue ſang a
triumphant ſong. \V And Chonenias the prince of the Leuites, was chiefe
ouer prophecie, to beginne the melodie: for he was verie coning. \V And
Barachias, and Elcana: doore keepers of the Arke. \V Moreouer Sebenias,
and Ioſaphat, and Nathanael, and  Amaſai, and Zacharias, and Banaias,
and Eliezer prieſtes, ſounded with trumpettes, before the Arke of God:
and Obededom, and Iehias kept the doore of the Arke. \V Therfore Dauid
and al the ancientes of Iſrael, and the tribunes, went to fetch the Arke
of the couenant of our Lord, out of the houſe of Obededom with ioy. \V
And when God
%%% 0866
had holpen the Leuites, which caried the Arke of the couenant of our
Lord, there were immolated ſeuen oxen, & ſeuen rammes. \V Moreouer Dauid
was clothed with a robe of fine linen, and al the Leuites that caried
the Arke, and the ſinging men, and Chonenias the prince of prophecie
among the ſingers: and Dauid alſo was clothed with an
\SNote{An other linen garment vſed by prophetes, ſuch as Samuel did
weare being a child.
\XRef{1.~Reg.~2.}}
Ephod of linen. \V And al Iſrael brought the Arke of the couenant of our
Lord in iubiley, and ſounding with the ſound of ſhaulme, and with
trumpettes, and cymbals, and nables, and harpes. \V And when the Arke of
the couenant of our Lord was come vnto the citie of Dauid, Michol the
daughter of Saul looking forth through a window, ſaw Dauid the king
dancing and playing, and ſhe deſpiſed him in her hart.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
