%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 26
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-26


\project douay-rheims


%%% 0881
%%% o-0790
\startChapter[
  title={Chapter 26}
  ]

\Summary{Porters are deſigned by lottes to watch at foure gates of the
  temple. 20.~Others made kepers of the holie treaſure, and
  veſſel. 30.~Officers alſo appointed in the two tribes and half ouer
  Iordan, for Gods ſeruice, and the kinges.}

And the diuiſions of the porters: of the Corites Meſelemia, the ſonnes of
Core, of the ſonnes of Aſaph. \V The ſonnes of Meſelemia: Zacharias the
firſt begotten, Iadihel the ſecond, Zabadias the third, Iathanael the
fourth, \V Aelam the fifth, Iohanan the ſixth, Elioenai the ſeuenth. \V
And the ſonnes of Obededom: Semeias the firſtbegotten, Iozabad the
ſecond, Ioaha the third, Sachar the fourth, Nathanael the fifth, \V
Ammiel the ſixth, Iſſachar the ſeuenth, Phollathi the eight: becauſe our
Lord bleſſed him. \V And to Semias his ſonne were borne ſonnes, the
chiefe of their families: for they were moſt valiant men. \V The ſonnes
then of Semeias: Othni, and Raphael, and Obed, Elizabad, and his
brethren moſt valiant men: Eliu alſo, and Samachias. \V Al theſe of the
ſonnes of Obededom: they, and their ſonnes, and their brethren moſt able
to miniſter, ſixtie two of Obededom. \V Moreouer the ſonnes of
Meſelemia, and their brethren moſt ſtrong, eightene. \V And of Hoſa,
that is of the
%%% 0882
ſonnes of Merari: Semri the prince
%%% o-0791
(%
\SNote{Either his firſt ſonne was dead, or was not fitte to be chief
ouer the reſt.}
for he had not a firſtbegotten, and therfore his father made him
chief) \V Helcias the ſecond, Tabelias the third, Zacharias the fourth:
al theſe the ſonnes, and brethren to Hoſa, thirtene. \V Theſe were
diuided into porters, that the princes alſo of the watches, euen as
their brethren, might miniſter in the houſe of our Lord. \V Lottes
therfore were caſt equally, both to litle, and great, by their families
for euerie one of the gates. \V The lotte then of the Eaſt fel to
Selemia. Moreouer to Zacharias his ſonne, a moſt wiſe man, and skilful,
the North quarter chanced by lot. \V But to Obededom and his ſonnes,
toward the South: in which part of the houſe was the councel of the
ancientes. \V Sephim, and Hoſa toward the Weſt, by the gate, which
leadeth to the way of aſcent: watch againſt watch. \V And toward the
Eaſt ſix Leuites: and toward the North foure a day: and toward the South
likewiſe in a day foure: and where the councel was two and two. \V In
the celles alſo of the porters toward the Weſt foure in the way: and two
at euerie celle. \V Theſe be the diuiſions of the porters of the ſonnes
of Core, and of Merari. \V Moreouer Achias was ouer the treaſures of the
houſe of God, and the veſſels of the holie places. \V The ſonnes alſo of
Ledan, the ſonnes of Gerſonni: of Ledan princes of the families, Ledan,
and Gerſonni, Iehieli. \V The ſonnes of Iehieli: Zathan, and Ioel his
brethren ouer the treaſures of the houſe of our Lord, \V to the
Amramites, and Iſaarites, and Hebronites, and Ozielites. \V And Subael
the ſonne of Gerſom, the ſonne of Moyſes, chiefe ouer the treaſures. \V
His brethren alſo Eliezer, whoſe ſonne Rahabia, and his ſonne Iſaias,
and his ſonne Ioram, his ſonne alſo Zechri, and his ſonne Selemith. \V
Selemith himſelf, and his
\Fix{brerbren}{brethren}{obvious typo, fixed in other}
ouer the treaſures of the ſanctified holie thinges, which Dauid the king
\SNote{Thinges are ſanctified by deſignation to holie vſe.}
ſanctified, and the princes of families, and the centurions, and the
captaynes of the hoſt \V of the warres, and of the ſpoyles of the
battels, which they had conſecrated to the maynteynance and furniture of
the temple of our Lord. \V And al theſe thinges did Samuel the Seer
ſanctifie, and Saul the ſonne of Cis, and Abner the ſonne of Ner, and
Ioab the ſonne of Saruia: and al that had ſanctified them by the hand of
Selemith, and of his brethren. \V But the chiefe ouer the Iſaarites was
Chonenias, and his ſonnes to the workes abrode ouer Iſrael to teach and
iudge them. \V Moreouer the Hebronites
%%% 0883
Haſabias, and his brethren, moſt able men, a thouſand ſeuen hundred were
chiefe ouer Iſrael beyond Iordan agaynſt the Weſt, in al the
\Fix{workes}{workes of}{obvious typo, fixed in other}
our Lord, and for the miniſterie of the king. \V And the prince of the
Hebronites was Ieria according to their families and kinredes. In the
fourteth yeare of king Dauid they were numbred, and there were found
moſt valiant men in Iazer Galaad, \V and his brethren of ſtronger age,
two thouſand ſeuen hundred princes of families. And Dauid the king made
them ouerſeers ouer the Rubenites and the Gaddites, and the halfe tribe
of Manaſſes, for al the miniſterie of God, and the king.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
