%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 0873
%%% o-0783
\startChapter[
  title={Chapter 21}
  ]

\Summary{Dauid ſinneth in numbring his people, 8.~repenteth, and
  prayeth, yet is punished, many dying of the plague, 15.~til God
  shewing mercie ſpareth the reſt. 16.~Dauid accuſing himſelfe, and
  excuſing the people is commanded by the Angel to offer ſacrifice,
  which he (22.~bying ground for an altar) 26.~performeth.}

But
\CNote{\XRef{2.~Reg.~24.}}
\SNote{By Gods permiſſion Dauid was tempted and ouercome.
\XRef{2.~Reg.~24.}}
Satan roſe agaynſt Iſrael: and moued Dauid to number Iſrael. \V And
Dauid ſayd to Ioab, and to the princes of the people: Goe, and number
Iſrael from Berſabee vnto Dan, and bring me the number that I may
know. \V And Ioab, anſwered: Our Lord increaſe his people an hundred
fold more then they are: are they not my lord king al thy ſeruantes?
why doth my lord ſeeke this, which may be reputed for a ſinne to
Iſrael? \V But the kinges word preuailed more: and Ioab went forth, and
went about al Iſrael: and returned to Ieruſalem. \V And he gaue Dauid
the number of them, whom he had ſurueyed: and al the number of Iſrael
was found a thouſand thouſand and an hundred thouſand men that drew
ſword: and of Iuda foure hundred ſeuentie thouſand men of warre. \V For
Leui and Beniamin he numbred not: becauſe Ioab vnwillingly executed the
kings commandement. \V And that which was commanded diſpleaſed God: and
he ſtroke
%%% 0874
Iſrael. \V And Dauid ſayd to God: I haue ſinned excedingly in that I
would doe this: I beſech thee take away
\SNote{King Dauid was not without faith, nor hope, and yet was
contaminate with iniquitie, from which he prayed to be deliuered.}
the iniquitie of thy ſeruant, becauſe I haue done fooliſhly. \V And our
Lord ſpake to Gad the Seer of Dauid, ſaying: \V Goe, and ſpeake to Dauid,
and tel him: Thus ſayth our Lord: I geue thee the choyſe of three
thinges, chooſe one which thou wilt, and I wil doe it to thee. \V And
when Gad was come to Dauid, he ſayd to him: Thus ſayth our Lord: Chooſe
which thou wilt: \V either three yeares famine: or three monethes to flee
from thine enemies, and not to be able to eſcape their ſword: or three
dayes the ſword of our Lord, and peſtilence to be in the land, and the
Angel of our Lord to kil in al the coſtes of Iſrael: now therfore ſee
what I ſhal anſwer him, that ſent me. \V And Dauid ſayd to Gad: I am on
euerie ſide in great diſtreſſe: but it is better, that I fal into the
handes of our Lord, becauſe his mercies be manie, then into the handes
%%% o-0784
of men. \V Our Lord therfore ſent the peſtilence in Iſrael, and there
fel of Iſrael ſeuentie thouſand men. \V He ſent alſo an Angel into
Ieruſalem, to ſtrike it: and when it was ſtriken, our Lord ſaw, and had
compaſſion vpon the greateneſſe of the euil: and commanded the Angel,
that ſmote: It is ſufficient, now let thy hand ceaſe. Moreouer the Angel
of our Lord ſtoode beſide the floore of Ornan the Iebuſeite. \V And
Dauid lifting vp his eies, ſaw the Angel of our Lord ſtanding betwen
heauen and earth, and a ſword drawen in his hand, turned agaynſt
Ieruſalem: & they fel as wel he as the ancientes clothed in heare
clothes, flatte on the earth. \V And Dauid ſayd to God: Am not I he,
that commanded the people to be numbred? It is I that haue ſinned: it is
I that haue done the euil: this flocke what hath it deſerued? Lord my
God let thy hand be turned I beſech thee vpon me, and vpon my fathers
houſe: and let not thy people be ſtriken. \V And the Angel of our Lord
commanded Gad to tel Dauid, that he ſhould goe vp, and build an altar to
our Lord God in the floore of Ornan the Iebuſeite. \V Dauid therfore
went vp according to the word of Gad, which he had ſpoken to him in the
name of our Lord. \V Moreouer Ornan when he had looked vp, and ſene the
Angel, and his foure ſonnes with him, they hid themſelues: for at that
time he threſhed wheat in the floore. \V Therfore when Dauid came to
Ornan, Ornan beheld him, & went forth to mete him out of the floore, and
adored him flatte on the
%%% 0875
ground. \V And Dauid ſayd to him: Geue me the place of thy floore, that
I may build therein an altar to our Lord: ſo that thou take as much
ſiluer as it is worth, and the plague may ceaſe from the people. \V And
Ornan ſayd to Dauid: Take it, and let my lord the king doe whatſoeuer
pleaſeth him: yea the oxen alſo I geue for holocauſt, and
\TNote{little cartes.}
the dreyes for wood, and wheat for ſacrifice: I wil giue al thinges
willingly. \V And king Dauid ſayd to him: It ſhal not be ſo, but I wil
geue thee ſiluer as much as it is worth: for I may not take it from
thee, and ſo offer to our Lord holocauſtes geuen gratis. \V Dauid
therfore gaue Ornan for the place, ſix hundred ſicles of gold of moſt
iuſt weight. \V And he built there an altar to our Lord: and he offered
holocauſtes, and pacifiques, and he inuocated our Lord, & he heard him
\SNote{God ſhewed by ſending fire miraculouſly to burne the ſacrifice,
that he had heard Dauids prayer.}
in fyre from heauen vpon the altar of holocauſte. \V And our Lord
commanded the Angel: and he turned his ſword into the ſcabbard. \V Dauid
therfore forthwith, ſeing that our Lord had heard him in the floore of
Ornan the Iebuſeite, immolated victimes there. \V But the tabernacle of
our Lord, which Moyſes made in the deſert, and the altar of holocauſtes,
was at that time in the excelſe of Gabaon. \V And Dauid could not goe to
the altar to pray God there: for he had bene frighted with exceding
feare, ſeing the ſword of the Angel of our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
