%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 0854
%%% o-0766
\startChapter[
  title={Chapter 9}
  ]

\Summary{VVho of Iſrael (after the captiuitie) firſt inhabited
  Ieruſalem. 10.~VVho exerciſed the offices of Prieſtes, 14.~and
  Leuites, 35.~with repetition of part of Sauls progenie.}

Al Iſrael therfore
\SNote{The genealogies of al Iſrael being hitherto recited before their
captiuitie, others are now added which firſt returned to Ieruſalem after
their releaſe.}
was numbred: and the ſumme of them was written in the Booke of the
kinges of Iſrael, and Iuda: and they were tranſported into Babylon for
their ſinne. \V And they that dwelt firſt in their poſſeſſions and in
their cities: Iſrael, and the Prieſtes, and the Leuites, and the
Nathineans. \V There dwelt in Ieruſalem of the children of Iuda, and of
the children of Beniamin, alſo of the children of
%%% 0855
Ephraim, and Manaſſes. \V Othei the ſonne of Ammiud, the ſonne of Amri,
the ſonne of Omrai, the ſonne of Bonni, of the ſonnes of Phares the
ſonne of Iuda. \V And of Siloni: Aſaia the firſtbegotten, and his
ſonnes. \V And of the ſonnes of Zara: Iehuel, and their brethren, ſix
hundred ninetie. \V Moreouer of the ſonnes of Beniamin: Salo the ſonne
of Moſollam, the ſonne of Odiua the ſonne of Aſana: \V and Iobania the
ſonne of Ieroham: and Ela the ſonne of Ozi, the ſonne of Mochori: and
Moſollam the ſonne of Saphatias, the ſonne of Rahuel, the ſonne of
Iebania, \V and their brethren by their families, nine hundred fiftie
ſix. Al theſe princes of their kinredes by the houſes of their
fathers. \V And of the prieſtes: Iedaia, Ioiarib, and Iachin: \V Azarias
alſo the ſonne of Helcias, the ſonne of Moſollam, the ſonne of Sadoc,
the ſonne of Maraioth, the ſonne of Achitob, high prieſt of the houſe of
God. \V Moreouer Adaias the ſonne of Ieroham, the ſonne of Phaſſur, the
ſonne of Melchias: and Maaſai the ſonne of Adiel, the ſonne of Iezra,
the ſonne of Moſollam, the ſonne of Moſollamith, the ſonne of Emmer. \V
Their brethren alſo princes by their families a thouſand ſeuen hundred
three ſcore, moſt able men of ſtrength to the worke of the miniſterie in
the houſe of God. \V And of the Leuites: Semeia the ſonne of Haſſub, the
ſonne of Ezricam, the ſonne of Haſebia of the ſonnes of Merari. \V
Bacbacar alſo a carpenter, and Galal, and Mathania the ſonne of Micha,
the ſonne of Zechri, the ſonne of Aſaph: \V and Obdia the ſonne of
Semeia, the ſonne of Galal, the ſonne of Idithun: and Barachia the ſonne
of Aſa, the ſonne of Elcana, who dwelt in the courtes of Netophati. \V
And the
%%% o-0767
porters: Sellum, and Accub, and Telmon, and Ahimam: and their brother
Sellum the prince, \V vntil that time, in the kinges gate toward the
eaſt, wayted by their courſes of the children of Leui. \V But Sellum the
ſonne of Core the ſonne of Abiaſaph, the ſonne of Core, with his
brethren, and his fathers houſe, theſe are the Corites ouer the workes
of the miniſterie, keepers of the entrances of the tabernacle: and their
families in courſe keeping the entrance of the campe of our Lord. \V And
Phinees the ſonne of Eleaſar, was their prince before our Lord. \V
Moreouer Zacharias the ſonne of Moſollamia, porter of the gate of the
tabernacle of teſtimonie. \V Al theſe choſen men for porters, at euerie
gate, two hundred twelue: and appointed out in their proper
%%% 0856
townes. Whom Dauid and Samuel the Seer appointed, vpon their
fidelitie. \V As wel them as their ſonnes, in the doores of the houſe of
our Lord, and in the tabernacle by their courſes. \V By the foure windes
were the porters: that is to ſay, toward the Eaſt, and toward the Weſt,
and toward the North, and toward the South. \V And their brethren dwelt
in villages, and came vpon their Sabbathes from time to time. \V To
theſe four Leuites was committed al the number of porters, and they were
ouer the chambers, and treaſures of the houſe of our Lord. \V Alſo round
about the temple of our Lord they abode in their watches: that when it
was time, they in the morning might open the doores. \V Of theſe mens
ſtocke there were alſo ouer the veſſels of the miniſterie: for by number
the veſſels were both brought in and caried out. \V Of them alſo they that
had the implements of the ſanctuarie committed vnto them, did ouerſee
the floure, and wine, and oile, and frankincenſe, and ſpices. \V And the
ſonnes of prieſtes made oyntementes of the ſpices. \V And Mathathias a
Leuite the firſtbegotten of Sellum the Corite, was ouerſeer of thoſe
thinges, which were fryed in the frying panne. \V Moreouer of the
children of Caath their brethren, there were ouer the loaues of
propoſition, to prepare alwaies new euerie Sabbath. \V Theſe are the
chief of the ſinging men by the families of the Leuites, which abode in
the chambers, that they might day and night continually ſerue in their
miniſterie. \V The heades of the Leuites, princes in their families,
taried in Ieruſalem. \V And in Gabaon abode Iehiel the father of Gabaon,
and the name of his wife Maacha. \V His firſt begotten ſonne Abdon, and
Sur, and Cis, and Baal, and Ner, and Nadab, \V Gedor alſo, and Ahio, and
Zacharias, and Macelloth. \V Moreouer Macelloth begat Samaan: theſe
dwelt ouer agaynſt their brethren in Ieruſalem, with their brethren. \V
And Ner begat Cis: and Cis begat Saul: and Saul begat Ionathas, and
Melchiſua, and Abinadab, and Eſbaal. \V And the ſonne of Ionathas,
Meribbaal: and Meribbaal begat Micha. \V Moreouer the ſonnes of Micha:
Phithon, and Melech, and Tharaa, and Ahaz. \V And Ahaz begat Iara, and
Iara begat Alamath, and Azmoth, and Zamri. And Zamri begat Moſa. \V And
Moſa begat Banaa: whoſe ſonne Raphaia
\Fix{bepat}{begat}{obvious typo, fixed in other}
Elaſa: of whom was borne Aſel. \V Moreouer Aſel had
%%% o-0768
ſix ſonnes of theſe names: Ezricam, Bochru, Iſmael, Saria, Obdia, Hanan,
theſe are the ſonnes of Aſel.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
