%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 25
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-25


\project douay-rheims


%%% 0880
%%% o-0789
\startChapter[
  title={Chapter 25}
  ]

\Summary{Foure ſonnes of Aſaph, ſix of Idithun, and fourtene of Heman,
  chiefe muſicians, 7.~with their brethren in al two hundred and
  fourſcore, 8.~are diſtributed by Lottes, in foure and twentie
  companies, to ſerue in the temple.}

Therfore Dauid, and the officers of the hoſte ſeperated for the
miniſterie the ſonnes of Aſaph, and Heman, and
\SNote{Otherwiſe called Ethan.
\XRef{ch.~6. v.~44.}
\XRef{c.~15. v.~19.}}
Idithun: which ſhould
\SNote{They played on inſtrumentes, Pſalmes and Canticles made by
Prophetes.}
prophecie on harpes, & pſaltries, & cymbals according to their number
ſeruing the office dedicated to them. \V Of the ſonnes of Aſaph:
Zacchur, and Ioſeph, and Nathania, & Aſarela, the ſonnes of Aſaph: vnder
the hand of Aſaph prophecying neere the king. \V Moreouer Idithun: the
ſonnes of Idithun: Godolias, Sori, Ieſeias, and Haſabias, and
Mathathias, ſix, vnder the hand of their father Idithun, who prophecied
on harpe ouer them that confeſſed and prayſed our Lord. \V Of Heman
alſo: the ſonnes of Heman: Bocciau, Mathaniau, Oziel, Subuel, and
Ierimoth, Hananias, Hanani, Eliatha, Geddelthi, and Romemthiezer, and
Ieſbacaſſa, Mellothi, Othir, Mahazioth. \V Al theſe the ſonnes of Heman
the Seer of the king in the wordes of God, that he might exalt the
horne: & God gaue to Heman fourtene ſonnes and three daughters. \V Al
vnder their fathers were diſtributed to ſing in the temple of our Lord,
on cymbals, and pſalteries and harpes, for the miniſteries of the houſe
of our Lord neere the king: to witte, Aſaph, and Idithun, and Heman. \V
And the number of them with their brethren, that taught the ſongue of
our Lord, al the teachers two hundred eightie eight. \V And they did
caſt lottes by their courſes, equally as wel the elder as the yonger,
the learned and the vnlearned together. \V And the firſt lotte came
forth to Ioſeph, which was of Aſaph. the ſecond to Godolias, to him and
his ſonnes, and his brethren twelue. \V The third to Zachur, to his
ſonnes and his brethren twelue. \V The fourth to Iſari, to his ſonnes
and his brethren twelue.
%%% o-0790
\V The fifth to Nathanias, to his ſonnes & his brethren twelue. \V The
ſixth to Bocciau, to his ſonnes and his brethren twelue. \V The ſeuenth
to Iſreela, to his ſonnes and his brethren twelue. \V The eight to
Ieſaia, to his ſonnes and his brethren 
%%% 0881
twelue. \V The ninth to Mathanias, to his ſonnes and his brethren
twelue. \V The tenth to Semeias, to his ſonnes and his brethren
twelue. \V The eleuenth to Azareel, to his ſonnes and his brethren
twelue. \V The twelfth to Haſabia, to his ſonnes and his brethren
twelue. \V The thirtenth to Subael, to his ſonnes and his brethren
twelue. \V The fourtenth to Mathathias, to his ſonnes and his brethren
twelue. \V The fiftenth to Ierimoth, to his ſonnes and his brethren
twelue. \V The ſixtenth to Hananias, to his ſonnes and his brethren
twelue. \V The ſeuententh to Ieſbacaſſa, to his ſonnes and his brethren
twelue. \V The eightenth to Hanani, to his ſonnes and his brethren
twelue. \V The ninetenth to Mellothi, to his ſonnes and his brethren
twelue. \V The twenteth to Eliatha, to his ſonnes and his brethren
twelue. \V The one and twenteth to Othir, to his ſonnes and his brethren
twelue. \V The two and twenteth to Geddelthi, to his ſonnes and his
brethren twelue. \V The three and twenteth to Mahazioth, to his ſonnes
and his brethren twelue. \V The foure and twenteth to Romemthiezer, to
his ſonnes and his brethren twelue.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
