%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 0857
%%% o-0768
\startChapter[
  title={Chapter 10}
  ]

\Summary{King
\MNote{The ſecond part.

King Saul and his familie
\Fix{ouerthrwone}{ouerthrowne}{obvious typo, fixed in other}}
Saul with his three ſonnes are ſlaine by the Philiſtians. 8.~Who ſpoile
the armie and carie away Sauls head. 11.~The men of Iabes Galaad burie
his bodie and his ſonnes, faſting for them ſeuen daies. And the kingdom
is tranſlated to Dauid.}

\Fix{Amd}{And}{obvious typo, fixed in other}
the Philiſtijms fought agaynſt Iſrael, and the men of Iſrael fled from
the Paleſthines, and they fel wounded in mount Gelboe. \V And when the
Philiſtians were come nere purſewing Saul, and his ſonnes, they ſtroke
Ionathas, and Abinadab, and Melchiſua the ſonnes of Saul. \V And the
battel grewe ſore agaynſt Saul, and the archers found him, and wounded
him with arrowes. \V And Saul ſayd to his eſquyer: Draw thy
\Fix{ſwoord,}{ſword,}{likely typo, fixed in other}
and kil me: leſt perhaps theſe vncircumciſed come, and deride me. But
his harnes bearer would not, being frighted with feare: Saul therfore
caught his ſword, and fel vpon it. \V Which when his harnes bearer had
ſeene, to witte, that Saul was dead, him ſelf alſo fel vpon his ſword,
and died. \V Saul therfore died, and his three ſonnes, and al his houſe
fel together. \V Which when the men of Iſrael had ſene, that dwelt in
the
\Fix{chainpayne,}{champayne,}{obvious typo, fixed in other}
they fled: and Saul and his ſonnes being dead, they forſooke their
cities, and were diſperſed hither and thither: and the Philiſthijms
came, and dwelt in them. \V The next day therfore the Philiſthijms
taking away the ſpoiles of them that were ſlayne, found Saul and his
ſonnes lying on mount Gelboe. \V And when they had ſpoyled him, and cut
of his head, and ſtripped him of his armour, they ſent into their land,
that it ſhould be caried about, and ſhould be ſhewed in the temples of
the Idols, and to the people: \V and his armour they dedicated in the
temple of their god, and the head they nailed vp in the temple of
Dagon. \V When the men of Iabes Galaad had heard this, to witte, al
thinges that the Philiſtijms had done vpon Saul, \V euerie one of the
valiant men aroſe, and tooke the bodies of Saul and of his ſonnes, and
brought them into Iabes, and buried their bones vnder an oke, that was
in Iabes, and they faſted ſeuen daies. \V Saul therfore died for his
iniquities, for that he
\SNote{He offered
\Fix{Sancrifice}{Sacrifice}{obvious typo, fixed in other}
on an Altar without warrant.
\XRef{1.~Reg.~13.}
And deſtroyed not the Amalacites, as he was commanded.
\XRef{1.~Reg.~15.}}
tranſgreſſed the commandement of our Lord which he had commanded, and
kept it not: yea and beſides alſo conſulted the Pythoneſſe, \V and
truſted not in our Lord: for the which he ſlewe him, and transferred his
kingdome to Dauid the ſonne of Iſai.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
