%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 29
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-29


\project douay-rheims


%%% 0887
%%% o-0796
\startChapter[
  title={Chapter 29}
  ]

\Summary{King Dauid once more by word, and example of his owne
  bountifulnes, inuiteth others to aſsiſt in building the temple,
  6.~wherto the princes and people contribute much. 10.~So after
  praiſes, 18.~prayeres, 21.~and ſacrifices offered to God, 22.~Salomon
  is annointed king. 26.~And king Dauid dieth bleſſedly, hauing reigned
  fourtie yeares.}

And Dauid the king ſpake to al the aſſemblie: Salomon my ſonne alone
hath God choſen, as yet a child and a litle tender one: for the worke is
great, neither is the habitation prepared for man, but for God. \V And I
with al my abilitie haue prepared the expenſes of the houſe of my
God. Gold for veſſels of gold, and ſiluer for them of ſiluer, braſſe for
them of braſſe, yron for them of yron, wood for them of wood: and onyx
ſtones, and as it were
\TNote{a kind of fine white ſtone.}
ſtibians, & of diuerſe colours, and euerie precious ſtone, and marble of
Paros moſt abundantly: \V and aboue theſe thinges, which I haue offered
into the houſe of my God, I geue
\SNote{That which he had already vowed and prepared he counted not his
owne, and now geueth more to the building and furniſhing of the Temple.}
of myne owne peculiar goods, gold and ſiluer vnto the temple of my God,
beſide thoſe thinges, which I haue prepared for the holie houſe. \V
Three thouſand talentes of gold of the gold of Ophir: and ſeuen thouſand
talentes of moſt approued ſiluer, to guilte the walles of the temple. \V
And, whereſoeuer there needeth gold, of gold: and wherſoeuer there
needeth ſiluer, of ſiluer: let the workes be made by the handes of the
artificers: and if any man offer voluntarily, let him fil his hand to
day, and offer what he wil to our Lord. \V The princes therfore of the
families promiſed, and the nobles of the tribes of Iſrael, the tribunes
alſo and the centurions, and the princes of the kinges poſſeſſions. \V
And they gaue vnto the workes of the houſe of our Lord, of gold, fiue
thouſand talentes, and of braſſe eightene thouſand talentes: of yron alſo
an hundred thouſand talentes. \V And with whomſoeuer were found ſtones,
they gaue them into the treaſures of the houſe
%%% 0888
of our Lord, by the hand of Iahiel the Gerſonite. \V And the people
reioyſed, when they promiſed vowes of their owne accord: becauſe they
did offer them to our Lord with al their hart: yea and Dauid the king
reioyſed with great ioy. \V And he bleſſed our Lord before al the
multitude, and he ſayd: Bleſſed art thou ô Lord the God of Iſrael our
father from eternitie vnto eternitie. \V Thine ô Lord is the magnificence,
and might, and glorie, and victorie: & to thee is the prayſe: for al
thinges that be in heauen, and in the earth, are thine: thine ô Lord, is the
kingdom, and thou art ouer al princes. \V Thine are
%%% o-0797
riches and thine is glorie: thou haſt dominion ouer al, in thy hand is
power & might: in thy hand greateneſſe, and the empire of al thinges. \V
Now therfore our God we confeſſe to thee, and we prayſe thy glorious
name. \V Who am I, and what is my people, that we can promiſe thee al
theſe thinges? al are thine: and thinges that we receiued of thy hand,
we haue geuen thee. \V For we are pilgrimes before thee, and ſtrangers,
\SNote{As Abraham Iſaac Iacob and others were ſtrãgers in reſpect of
Chanaan: ſo both they and alſo Dauid Salomon and al men in this world
are pilgrimes in reſpect of heauen.
\XRef{Heb.~13.}}
as al our fathers. Our daies are as a ſhadow vpon the earth, and there
is noe abyding. \V Lord our God, al this plentie, which we haue prepared
that a houſe might be built to thy holie name, is of thy hand, and al
thinges are thine. \V I know my God that thou proueſt the hartes, and
loueſt ſimplicitie, wherfore I alſo in the ſimplicitie of my hart,
gladly haue offered al theſe thinges: and thy people, which is here
found, I haue ſene with great ioy offer thee donaries. \V Lord God of
Abraham, and Iſaac, and Iſrael our fathers, keepe for euer this wil of
their hart, and let this mind
\Fix{reinayne}{remayne}{obvious typo, fixed in other}
alwaies to the honour of thee. \V
To Salomon alſo my ſonne geue a perfect hart, that he keepe thy
commandementes, thy teſtimonies, and thy ceremonies, and doe al thinges:
and build the houſe, the expenſes whereof I haue prepared. \V And Dauid
commanded the whole aſſemblie: Bleſſe ye our Lord God. And al the
aſſemblie bleſſed our Lord the God of their fathers: and they
\SNote{Albeit they did the ſame exterior act of honour to God and the
king: yet in their mind & intention they gaue diuine honour to God, and
ciuil to the king. See
\XRef{Exod.~20.}}
bowed themſelues, and adored God, and then the king. \V And they
immolated victime to our Lord: and they offered holocauſtes the day
folowing, oxen a thouſand, rammes a thouſand, lambes a thouſand with
their libamentes, & with al rite moſt abundantly for al Iſrael. \V And
they did eate, and drinke before our Lord in that day with great ioy.
\CNote{\XRef{3.~Reg.~1. v.~34.}}
And they anoynted the ſecond time Salomon the ſonne of Dauid. And they
anoynted him to our Lord for the
%%% 0889
prince, and Sadoc for the high prieſt. \V And Salomon ſate vpon the
throne of our Lord as king for Dauid his father, and he pleaſed al: & al
Iſrael obeyed him. \V Yea and al the princes, and men of might, and al
the ſonnes of king Dauid gaue their hand, and were ſubiect to Salomon
the king. \V Our Lord therfore magnified Salomon ouer al Iſrael: and
gaue him the glorie of a kingdom, ſuch as no king of Iſrael had before
him. \V Dauid therfore the ſonne of Iſai reigned ouer al Iſrael. \V And
the daies, that he reigned ouer Iſrael, were fourtie yeares: in Hebron
he reigned ſeuen yeares, and in Ieruſalem three and thirtie yeares. \V
And he died in a good age, ful of daies, and riches, and glorie. And
Salomon his ſonne reigned for him. \V But the actes of king Dauid the
firſt and the laſt are writen in the Booke of Samuel the Seer, & in the
Booke of
\SNote{Either Nathan and Gad writ the later part of the firſt booke, and
the ſecond booke of kinges, or els their bookes are not now extant.}
Nathan the prophete, & in the Volume of Gad the Seer: \V and of al his
kingdom, and ſtrength, and of the times, that paſſed vnder him, either
in Iſrael, or in al the kingdomes of the earth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
