%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 0870
%%% o-0780
\startChapter[
  title={Chapter 18}
  ]

\Summary{King Dauid hath great victories, making manie nations
  tributarie. 15.~His chiefe officeres are recounted.}

And it came to paſſe after theſe thinges, that
\CNote{\XRef{2.~Reg.~8.}}
Dauid ſtroke the Philiſthijms, and humbled them, and tooke away Geth,
and her
\SNote{The leſſer townes and
\Fix{villagies}{villages}{likely typo, fixed in other}
are commonly called the
\Fix{dauhgters}{daughters}{obvious typo, fixed in other}
of ſome great towne or citie to which they perteyne.}
daughters out of the hand of the Philiſthijms, \V and ſtroke
\Fix{Moob,}{Moab,}{obvious typo, fixed in other}
and the Moabites were made Dauids ſeruantes, offering him giftes. \V At
that time Dauid ſtroke Adazezer alſo the king of Soba of the countrie of
Hemath, when he went on to dilate his empire as farre as the riuer
Euphrates. \V Dauid therfore tooke a thouſand chariotes of his, & ſeuen
thouſand horſmen, & twentie thouſand footemen, & he hoghſinewed al the
chariot horſes, ſauing an hundred chariotes, which he reſerued to
himſelf. \V And the Syrian alſo of Damaſcus came moreouer, to giue ayde
to Adarezer the king of Soba but Dauid ſtroke alſo of his two & twentie
thouſand men. \V And he put ſouldiars in Damaſcus, that Syria alſo
ſhould ſerue him, & bring giftes. And our Lord holpe him in al thinges,
to the which he went. \V Dauid alſo tooke the golden quyuers, which the
ſeruantes of Adarezer had, and he brought them into Ieruſalem. \V
Moreouer of Thebath, and Chun the cities of Adarezer verie much braſſe,
of which Salomon made the braſen Sea, and pillers, and braſen veſſels. \V
Which when Tou the king of Hemath had heard, to witte, that Dauid had
ſtriken al the armie of Adarezer the king of Soba, \V he ſent Adoram his
ſonne to king Dauid, to deſire peace of him, & to congratulate him that
he had ſtriken, and had ouerthrowen Adarezer: for Tou was aduerſarie to
Adarezer. \V But al the veſſel alſo of gold, and ſiluer, and braſſe king
Dauid
\SNote{That which is dedicated to ſacred vſe, is conſecrated to God.}
conſecrated to our Lord, with the ſiluer and gold, which he had taken
out of al the nations, as wel of Idumea, and Moab, & the children of
Ammon, as of the Philiſthijms and Amalec. \V And Abiſai the ſonne of
Saruia ſtroke Edom in the Vale of ſalt pittes, eightene thouſand: \V and
he appoynted a garryſon in Edom, that Idumea ſhould ſerue Dauid: and our
Lord ſaued Dauid in al thinges, to which he went. \V Dauid therfore
reigned ouer al Iſrael, & did iudgement and
%%% 0871
iuſtice to al his people. \V Moreouer Ioab the ſonne of Saruia was ouer
the armie, and Ioſaphat the ſonne of Ahilud commenter.
%%% o-0781
\V And Sadoc the ſonne of Achitob, & Ahimelech the ſonne of Abiathar,
Prieſtes: and Suſa, Scribe. \V Banaias alſo the ſonne of Ioiada ouer the
legions Cerethi, and Phelethi: moreouer the ſonnes of Dauid the chief at
the kinges hand.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
