%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 0843
%%% o-0756
\startChapter[
  title={Chapter 3}
  ]

\Summary{The ſonnes of King Dauid. 10.~The line of the Kinges of Iuda
  from Salomon to Ioſias. 15.~With diuers generations of the ſame
  Ioſias.}

But Dauid had theſe ſonnes, which were borne to him in Hebron: the
firſtbegotten Ammon of Achinoam the Ieſraelite, the ſecond Daniel of
Abigail the Carmelite, \V the third Abſolom the ſonne of Maacha the
daughter of Tolmai the king of Geſſur, the fourth Adonias the ſonne of
Aggith, \V the fift Saphatias of Abital, the ſixth Iethraham of Egla his
wife. \V Six ſonnes therfore were borne to him in Hebron, where he
reigned ſeuen yeares and ſix monethes. And in Ieruſalem he reigned three
and thirtie yeares. \V Moreouer in Ieruſalem ſonnes were borne to him,
Simmaa, and Sobab, and Nathan, & Salomon, foure of Bethſabea the
daughter of Ammiel, \V Iebaar alſo and Eliſama, \V and Eliphaleth, and
Noge, and Nepheg, and Iaphia, \V moreouer Eliſanna, and Eliada, and
Elipheleth, nine: \V al theſe the ſonnes of Dauid, beſide the ſonnes of
his concubines: and they had a ſiſter Thamar. \V And the ſonne of
Salomon, Roboam: whoſe ſonne Abia begat Aſa. Of this alſo was borne
Ioſaphat, \V the father of Ioram: which Ioram begat Ochozias, of whom
roſe Ioas: \V and his ſonne Amaſias begat Azarias. Moreouer Azarias the
ſonne of Ioathan \V begat Achaz, the father of Ezechias, of whom was
borne Manaſſes. \V But Manaſſes alſo begat Amon the father of Ioſias. \V
And the ſonnes of Ioſias were: the firſtbegotten Iohanam, the ſecond
\SNote{S.~Matthew omitteth this Ioakim, and counteth Iechonias as the
ſonne of Ioſias. The ſame Ieconias was alſo otherwiſe called Ioachin.
\XRef{4.~Reg.~24. v.~6.}
&
\XRef{25. v.~27.}
\Cite{S.~Hierom. li.~1. in Matth.}}
Ioakim, the third Sedecias, the fourth Sellum. \V Of Ioakim was borne
%%% 0844
Iechonias, and Sedecias. \V The ſonnes of Iechonias were Aſir,
Salathiel, \V Melchiram, Phadaia, Senneſer & Iecemia, Sama, and
Nadabia. \V Of Phadaia were borne Zorobabel and Semei. Zorobabel begat
Moſollom, Hananias, and Salomith their ſiſter: \V Haſaba alſo, and Ohol,
and Barachias, and Haſadias, Iaſabheſed, fiue. \V And the ſonne of
Hananias, Phaltias the father of Ieſeias, whoſe ſonne was Raphaia. This
mans ſonne alſo Arnan, of whom was borne Obdia, whoſe ſonne was
Sechenias. \V The ſonne of Sechenias: Semeia, whoſe ſonnes were Hattus,
and Iegaal, and Baria, and Naaria, and Saphat,
\SNote{Semeia with his fiue ſonnes are counted ſix ſonnes of Sechenias,
though Semeia only was his proper ſonne, the other his nephewes. See
\XRef{ch.~1. num.~3.}}
ſix in number. \V The ſonnes of Naaria: Elioenai, & Ezechias, and
Ezricam, three. \V The ſonnes of Elioenai: Oduia, and Eliaſub, and
Pheleia, and Accub, and Iohanan, and Dalaia, and Anani, ſeuen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
