%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 0841
%%% o-0754
\startChapter[
  title={Chapter 2}
  ]

\Summary{The names of Iſraels twelue ſonnes. 3.~The geneologie of Iuda,
  firſt in the right line to Dauid, the ſeuenth ſonne of Iſai: 16.~then
  other genealogies of the ſame Iuda.}

And the children of
\SNote{This patriarch firſt called \Emph{Iacob} ſignifying
\Fix{\Emph{ſupplater}}{\Emph{ſupplanter}}{obvious typo, fixed in other}
was afterward called \Emph{Iſrael}, that is, \Emph{Seing God}, or
\Emph{valient vvith God}.
\XRef{Gen.~35.}}
Iſrael: Ruben, Simeon, Leui, Iuda, Iſſachar, and Zabulon, \V Dan,
Ioſeph, Beniamin, Nephthali, Gad, and Aſer. \V The ſonnes of Iuda: Her,
Onan, and Sela. Theſe three were borne to him of the Chananite the
daughter of Sue. And Her the firſtbegotten of Iuda, was euil before our
Lord, and he ſlewe him. \V And Thamar his daughter in law bare him
Phares and Zara. \V Therfore al the ſonnes of Iuda, were fiue. \V And
the ſonnes of Phares: Heſron and Hamul. \V The ſonnes alſo of Zara: Zamri,
and Ethan, and Eman, Chalcal alſo, and Dara, together fiue. \V And the
ſonnes of Charmi:
\SNote{Either this man had two names, or there is error in the laſt
letter, here, or
\XRef{Ioſue.~7.}}
Achar, who trubled Iſrael, & ſinned in the theft of the anathema. \V The
ſonnes
of Ethan: Azarias. \V And the ſonnes of Heſron that were borne to him:
Ierameel, and Ram, and Calubi. \V Moreouer Ram begat Aminadab, and
Aminadab begat Nahaſſon, the prince of the children of Iuda. \V Nahaſſon
alſo begat Salma, of whom was borne Booz. \V But Booz begat Obed, who
alſo begat Iſai. \V And Iſai begat the firſtbegotten Eliab, the ſecond
Abinadab, the third Simmaa, \V the fourth Nathanael, the fifth Raddai, \V
the ſixt Aſom, the ſeuenth Dauid. \V Whoſe ſiſters were Sarnia, and
Abigail. The ſonnes of Saruia: Abiſai, Ioab, and Aſael, three. \V And
Abigail beare Amaſa, whoſe father was Iether the Iſmaelite. \V But
\SNote{Otherwiſe called Calubi.
\XRef{v.~9.}}
Caleb the ſonne of Heſron tooke a wife named Azuba, of whom he begat
Ierioth: and her ſonnes were Iaſer, and Sobab, and Ardon. \V  And when
Azuba was dead, Caleb tooke to wife Ephratha: who bare him Hur. \V
Moreouer Hur begat Vri: and Vri begat Bezeleel. \V After theſe thinges
Heſron went in to the daughter of Machir the father of Galaad, and tooke
her, when he was three ſcore yeares old: who bare him Segub. \V But
Segub alſo
%%% 0842
begat Iair, & poſſeſſed three and twentie cities in the Land of
Galaad. \V And he tooke Geſſur, and Aram the townes of Iair, and Canath,
and the villages therof, threeſcore cities: al theſe are
\SNote{By \Emph{ſonnes} as often elſewhere, are vnderſtood nephewes and
other ofspring.}
the ſonnes of Machir the father of Galaad. \V And when Heſron was dead:
Caleb went in to Ephrata. Heſron alſo had to wife Abia who bare him
Aſhur the father of Thecua. \V And there were borne ſonnes to Ierameel
the firſtbegotten Heſron: Ram his firſtborne, and Buna, and Aram, and
Aſom, and Achia. \V Ierameel maried alſo an other wife, named Atara,
which was the mother of Onam. \V But the ſonnes alſo of Ram the
firſtbegotten of Ierameel,
%%% o-0755
were Moos, Iamin, and Acar. And Onam had ſonnes Semei, and Iada. And the
ſonnes of Semei: Nadab, and Abiſur. \V But the name of Abiſurs wife, was
Abihail, who bare him Ahobban, and Molid. \V And the ſonnes
of Nadab were Saled, & Aphaim. And Saled died without children. \V But
the ſonne of Apphaim, Ieſi: which Ieſi begat Sefan. Moreouer Sefan begat
Oholai. \V And the ſonnes of Iada the brother of Semei: Iether, and
Ionathan. But Iether alſo died without children. \V Moreouer Ionathan
begat Phaleth, and Ziza. Theſe were the ſonnes of Ierameel. \V And Sefan
had noe ſonnes, but daughters: and a ſeruant an Ægyptian, named
Ieraa. \V And he gaue his daughter to wife: who bare him Ethei. \V And
Ethei begat Nathan, and Nathan begat Zabad. \V Zabad alſo begat Ophlal,
& Ophlal begat Obed, \V Obed begat Iehu, Iehu begat Azarias, \V Azarias
begat Helles, and Helles begat Elaſa, \V Elaſa begat Siſamoi, Siſamoi begat
Sellum, \V Sellum begat
\Fix{Icamiam,}{Icamia,}{obvious typo, fixed in other}
and Icamia begat Eliſama. \V And the ſonnes of Caleb the brother of
Ierameel: Meſa his firſtbegotten, he is the father of Ziph: and the
ſonnes of Mareſa the father of Hebron. \V Moreouer the ſonnes of Hebron: Core, and
Thaphua, and Recem, and Samma. \V And Samma begat Raham, the father of
Iercaam, and Recem begat Sammai. \V The ſonne of Sammai, Maon: and Maon
the father of Bethſur. \V And Epha the concubine of Caleb bare Haran,
and Moſa, and Gezez. Moreouer Haran begat Gezez. \V And the ſonnes of
Iahaddai: Regom, and Ioathan, and Geſan, and Phalet, and Epha, and
Saaph. \V The concubine of Caleb Maacha bare Saber, and Tharana. \V And
Saaph the father of Madmena begat Sue the father of Machbena, and the
father of Gabaa. But the daughter of Caleb, was Achſa. \V Theſe
%%% 0843
were the ſonnes of Caleb, the ſonne of Hur the firſtbegotten of Ephrata,
Sobal the father of Cariathiarim. \V Salma the father of Bethlehem,
Hariph the father of Bethgader. \V And there were ſonnes of Sobal the
father of Chariathiarim, he that ſaw the halfe of the
\TNote{dvvellings, or reſting places.}
reſtinges. \V And of the kinred of Cariatharim, the Iethreites, and
Aphutheites, and Sematheites, and Maſeretes. Out of theſe iſſued the
Saraites, and Eſthaolites. \V The ſonnes of Salma: Bethlehem, and
Netophathi, the
\SNote{Valient men by whoſe help Ioab got victories and triumphant
crownes.}
Crownes of the houſe of Ioab, and the Halfe of the reſting of Sarai. \V
The kinredes alſo of the ſcribes dwelling in Iabes, ſinging and
ſounding, and abyding in tabernacles. Theſe are the Cineites, which came
from heate of the father of the houſe of Rechab.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
