%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Paralipomenon
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 0844
%%% o-0757
\startChapter[
  title={Chapter 4}
  ]

\Summary{Other genealogies of Iuda: 24.~and of Simeon, 29.~by whom the
  progenie of Cham, 42.~and reliques of Amelacites are ſubdued.}

The ſonnes of Iuda: Phares, Heſron, and Charmi, and Hur, and Sobal. \V
But Raia the ſonne of Sobal begat Iahath, of whom were borne Ahumai, and
Laad. Theſe be the kinredes of Sarathi. \V This
\Fix{allo}{alſo}{obvious typo, fixed in other}
is the ſtocke of Etam: Iezrahel, and Ieſema, and Iedebos. And the name
of their ſiſter Aſalelphuni. \V And Phanuel the father of Gedor, and
Ezar the father of Hoſa, theſe are the ſonnes of Hur the firſt begotten
of Ephratha the father of Bethlehem. \V But Aſſur the father of Thecua
had two wiues, Halaa, & Naara. \V And Naara bare him Oozam, and Hepher,
and Themani, and Ahaſthari. Theſe are the ſonnes of Naara. \V Moreouer
the ſonnes of Halaa: Sereth, Iſaar, and Ethnan. \V And Cos begat Anob,
and Soboba, and the kinred of Aharehel the ſonne of Arum. \V And Iabes
was honorable aboue his brethren, and his mother called his name Iabes,
ſaying: Becauſe I bare him in ſorow. \V But
\SNote{In ioyning a vow to his prayer he imitated holie Iacob.
\XRef{Gen.~28.}
And they both deſired temporal thinges for the better ſeruing of God,
& aduancing his glorie: eſpecially that they might be aſſiſted with
grace not to yeld to tentations, nor ſinne of malice.}
Iabes inuocated the God of Iſrael, ſaying: If bleſſing thou wilt bleſſe
me, and wilt enlarge my borders, and thy hand be with me, and thou wilt
make that I be not oppreſſed by malice. And God granted the thinges that
he prayed for. \V And Caleb the brother of Sua begat Mahir, who was the
father of Eſthon. \V Moreouer Eſthon begat Bethrapha, and Pheſſe, and
Tehinna the father of the citie of Naas: theſe are the men of Recha. \V
And the ſonnes of Cenez: Othoniel, and Saraia. Moreouer the ſonnes
%%% 0845
of Othoniel: Hathath, and Maonathi. \V Maonathi begat Ophra, & Saraia
begat Ioab
\SNote{Chief lord of the valley; where the artificers dwelt that made
the Temple.}
the father of the Vale of Artificers: for there were artificers. \V And
the ſonnes of Caleb the ſonne of Iephone: Hir, and Ela, and Naham. The
ſonnes
alſo of Ela: Cenez. \V The ſonnes alſo of Ialeleel: Ziph, and Zipha,
Thiria, and Aſrael. \V And the ſonnes of Eſra: Iether, and Mered, and
Epher, and Ialon, and he begat Mariam,
\Fix{and and}{and}{obvious typo, same in both}
Sammai, and Ieſba the father
of Eſthamo. \V Alſo his wife Iudaia, bare Iared the father of Gedor, and
Heber the father of Socho, and Ieuthiel the father of Zanoe. And theſe
are the ſonnes of Bethia the daughter of Pharao, whom Mered tooke. \V
And the ſonnes of the wife of Odaia the ſiſter of Naham the father of
Ceila, Garmi, and Eſthamo, which was of Macathi. \V The ſonnes alſo
Simon, Amnon, and Rinna the ſonne of Hanan, and Thilon. And the ſonnes
of Ieſi: Zoheth, and Benzoheth. \V The ſonnes of Sela the ſonne of Iuda:
Her the father of Lecha, and Laada the father of Mareſa, and the
%%% o-0758
kinredes of their houſe that worke ſilke in the Houſe of oath. \V And he
that made the Sunne to ſtand, and the men of Lying, and Secure, and
Burning, which were princes in Moab, and which returned into Lahem. And
theſe are old wordes. \V Theſe are potters, dwelling in Plantinges, and
in Hedges, with the king in his workes, and they abode there. \V The
ſonnes of Simeon: Namuel, and Iamin, Iarib, Zara, Saul: \V Sellum his
ſonne, Mapſam his ſonne, Maſma his ſonne. \V The ſonnes of Maſma: Hamuel
his ſonne, Zachur his ſonne, Semei his ſonne. \V The ſonnes of Semei
ſixtene, and ſix daughters: but his brethren had not manie ſonnes, and
the whole kinred could not reach to the ſumme of the children of
Iuda. \V And they dwelt in Berſabee, and Molada, and Haſarſuhal, \V and
in Bala, and in Aſom, and in Tholad, \V and in Bathuel, and in Hormi,
and in Siceleg, \V and in Bethmarcaboth, and in Haſarſuſim, and in
Bethberai, and in Saatim. Theſe were their cities vntil king Dauid. \V
Their townes alſo: Etam, and Aen, Rhemmon, and Thochen, and Aſan, fiue
cities. \V And al their villages round about theſe cities vnto
Baal. This is their habitation, and the diſtribution of their
dwellinges. \V Moſabab alſo and Iemlech, and Ioſa the ſonne of
Amaſias, \V and Ioel, and Iehu the ſonne of Ioſabia the ſonne of Saraia,
the ſonne of Aſiel, \V and Elioenai, and Iacoba, and Iſuhaia, and Aſaia,
and Adiel, and Iſmiel, and
%%% 0846
Banaia, \V Ziza alſo the ſonne of Zephei the ſonne of Allon the ſonne of
Idaia the ſonne of Semri the ſonne of Samaia. \V Theſe renowmed princes
in their kinredes, & in the houſe of their affinities they were
multiplied excedingly. \V And they went forth to enter into Gador as far
as the Eaſt ſide of the valley, and to ſeeke paſtures for their
flockes. \V And they found fatte paſtures, and very good, and a countrie
very large and quiet and fruiteful, in the which before had dwelt the
ſtocke of Cham. \V Theſe therfore, whom before we deſcribed by name,
came in the dayes of Ezechias the king of Iuda: and they ſtroke their
\Fix{tabernales,}{tabernacles,}{obvious typo, fixed in other}
and the inhabitantes that were found there, & cleane
deſtroyed them vntil this preſent day: and they dwelt for them, becauſe
they found there moſt fatte paſtures. \V Alſo of the children of Simeon
there went into mount Seir fiue hundred men, hauing their princes
Phaltias and Naaria and Raphaia and Oziel the ſonnes of Ieſi: \V and they
ſtroke the remnant of the Amalecites, which were able to eſcape, and
they dwelt there for them vntil this day.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
