%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 0323
%%% o-0292
\startChapter[
  title={Chapter 21}
  ]

\Summary{At what funerals Prieſts may not be preſent. 7.~VVhat wemen
  they may not marie, 9.~a prieſts daughter committing fornication muſt
  be burned. 10.~The high Prieſt shal not vncouer his head, nor rent his
  garment, nor be preſent at anie funeral, nor at al goe forth of the
  holie place. 13.~When he marieth he muſt take a virgin. 16.~None that
  hath a blemish in his bodie (though he be of Aarons ſtock) shal
  miniſter in the Sanctuarie, nor approch to the Altar.}

%%% 0324
Ovr Lord ſaid alſo to Moyſes: Speake to the prieſtes the ſonnes of
Aaron, and thou ſhalt ſay to them: Let not a prieſt be contaminated in
the deathes of his citizens, \V but onlie in his kinne, and nigh of
bloud, that is to ſay, vpon his father and mother, and ſonne, and
daughter, brother alſo, \V and ſiſter being a virgin which hath not bene
maried to a huſband: \V but neither in
\SNote{By the prĩce S.~Paul vnderſtood the high Prieſt.
\XRef{Act.~23.}}
the prince of his people ſhal he be contaminated. \V Neither ſhal they
ſhaue their head, nor beard, not make inciſions in their fleſh. \V They
ſhal be holie to their God, and shal not pollute his name: for the burnt
ſacrifice of the Lord, and breades of their God doe they offer, and
therfore they shal be holie. \V A whore and a vile ſtrumpette he shal
not take to wife, nor her, that is put away from her huſband: becauſe
they are cõſecrated to their God, \V and offer the breades of
propoſition. Be they holie therfore, becauſe I alſo am holie, the Lord,
that ſanctifie them. \V The daughter of a prieſt if she be taken in
whordome, & diſhonour the name of hir father, shal be burnt with
fire. \V The grand bishoppe, that is to ſay the prieſt that is greateſt
among his brethren, vpon whoſe head hath bene poured the oyle of
vnction, and whoſe handes were conſecrated in prieſthood, and who was
reueſted with the holie veſtimentes, shal not vncouer his head, he shal
\SNote{Caiphas rẽt his garments
\XRef{(Mat.~26.)}
cõtrarie to this law, malice making him neglect his owne dignitie.}
not rent his garments: \V and to no dead perſon shal
%%% o-0293
he enter in at al. Vpon his father alſo, and mother shal he not be
contaminated. \V Neither shal he goe forth out of the holie places, leſt
he pollute the Sãctuarie of the Lord, becauſe the oyle of the holie
vnction of his God is vpon him. I the Lord. \V He shal take a virgin
vnto his wife: \V but a widow and her that is put away, and a filth, and
a whore he shal not take, but a maide of his owne people: \V that he
mingle not the ſtocke of his kinred with the common people of his
nation: becauſe I am the Lord that ſanctifie him. \V And our Lord ſpake
to Moyſes, ſaying: \V Speake to Aaron: The man of thy ſeede throughout
their families, that hath
\SNote{If ſuch deformities and defectes made men irregular, and vnmete
to exerciſe prieſtlie function in the old teſtament, how
%%% Rest of this note missing in Heretikes, present in other.
much more in the Church of Chriſt?}
a blemish, shal not offer breades to his God, \V neither shal he approch
to his miniſterie: If he be blinde, if lame, if he haue a litle, or a
great, or a crooked noſe, \V if his foote be broken, if his hand, \V if
he be crooke backed, or blere eyed, or haue a pearle in his eye, or a
continual ſcabbe, or drie ſcurffe in his bodie, or be burnt. \V Euerie
one that hath a
%%% 0325
blemish of the ſeede of Aaron the prieſt, shal not approch to offer the
hoſtes to the Lord, nor the breades to his God. \V He shal eate
notwithſtanding of the breades, that are offered in the Sanctuarie, \V
yet ſo that he enter not within the veilen, or approch to the altar,
becauſe he hath a blemish, and he muſt not contaminate my Sanctuarie. I
the Lord that ſanctifie them. \V Moyſes therfore ſpake to Aaron, and to
his ſonnes and to al Iſrael, al thinges that had bene commanded him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
