%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 0285
%%% o-0258
\startChapter[
  title={Chapter 2}
  ]

\Summary{How to offer flovvre, 4.~loaues, wafers, with oile and incenſe,
  without leauen or honie, 12.~alſo firſt fruictes. 13.~And ſalt in
  euerie oblation.}

VVhen a ſoule ſhal offer an oblation of ſacrifice to our Lord,
\SNote{Theſe oblations were ſpecially for the pooreſt ſorte. So God
tempered his law to the abilitie of euerie one.
\Cite{Theod. q.~1. in Leuit.}}
fine flowre ſhal be his oblation. And he ſhal poure oyle vpon it, and
put franckincenſe, \V and ſhal carie it to the ſonnes of Aaron the
prieſts: of whom one ſhal take a handful of the flowre, and the oile,
and al the franckincenſe, and shal put it a memorial vpon the Altar for
a moſt ſweete ſauour to our Lord. \V And that which shal be left of the
ſacrifice, ſhal be Aarons, and his ſonnes, Holie of holies among the
oblations of our Lord. \V But when thou offereſt a ſacrifice baked in
the ouen: of flowre, to wit, loaues without leauen, tempered with oyle,
and wafers vnleauened layd ouer with oyle. \V If thine oblation be of
the frying panne, of flowre tempered with oyle, and without leauen, \V
thou shalt diuide it in litle peeces, and ſhalt poure oyle vpon it. \V
And if the ſacrifice be from the gridiron, in like maner the flowre ſhal
be tempered with oyle, \V which offering to our Lord, thou ſhalt deliuer
to the handes of the prieſt. \V Who hauing offered it, ſhal take a
memorie of the ſacrifice, and burne vpon the altar for a ſwete ſauour to
our Lord, \V and whatſoeuer is left, shal be Aarons, and his ſonnes,
Holie of holies among the oblations of our Lord. \V Euerie oblation,
that is offered to our Lord, shal be made
\SNote{As literally no leauen, nor honey might be offered in ſacrifice:
ſo al ſinne and carnal delectatiõ muſt be excluded in chriſtian life.}
without leauen, neyther ſhal any leauen and
%%% !!! This SNote is the same as the above.
%%% \SNote{}
honie be burned in the ſacrifice of our Lord. \V The firſt fruites only
of them and the giftes you shal offer: but vpon the altar they shal not
be put, for a ſauour of ſwetnes. \V What ſacrifice ſoeuer thou offereſt,
thou ſhalt ſeaſon it with ſalt neither ſhalt thou take away the ſalt of
the couenant of thy God from thy ſacrifice. In euerie oblation thou
shalt offer
\SNote{That ſalt ſignifieth diſcretion, appeareth by S.~Paul exhorting
to ſeaſon al our talke with ſalt.
\XRef{Colloſ.~4. v.~6.}}
ſalt. \V But if thou offer a gyfte of the firſt fruites of thy corne to
our Lord, of the eares being yet greene, thou shalt drie it at the fire,
and bruiſe it in maner of meale, and ſo ſhalt thou offer thy firſt
fruites to our Lord, \V powring oyle vpon it, and putting on
franckincenſe, becauſe it is the oblation of our Lord. \V Wherof the
prieſt shal burne for memorie of the gift, part of the meale bruiſed,
and of the oyle, and al the frankincenſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
