%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 0290
%%% o-0263
\startChapter[
  title={Chapter 6}
  ]

\Summary{Oblation for ſinne vvittingly committed. 8.~The maner of
  offering holocauſt. 12.~Continual fire to be kept in the
  Altar. 14.~The ſacrifices vvhich Prieſts shal offer at their
  Conſecration. 24.~In general of hoſtes for ſinne, and vvho shal eate
  of the ſame and vvhere.}

Ovr Lord ſpake to Moyſes, ſaying: \V The ſoule that shal ſinne, and
contemning the Lord, shal denie vnto his neighbour the thing deliuered
to his cuſtodie, which was committed to his credite, or shal by force
extort anie thing, or do oppreſſion, \V or shal finde a thing loſt, and
denying it, be alſo foreſworne, and shal doe anie other thing of manie,
wherin men are wount to ſinne, \V being conuicted of the offence, \V he
shal render al thinges which by
%%% 0291
fraude, he would haue obteyned, whole, and the fift part beſides to the
owner vnto whom he did the damage. \V But
\SNote{Such examples of penãce or ſatisfaction for ſinne beſides
reſtitution of that was wrongfully taken, are frequent in Moyſes Law.}
for his ſinne he shal offer a ramme without ſpott out of the flocke, and
shal geue it to the prieſt, according to the eſtimation, and meaſure of
the offence: \V who shal pray for him before the Lord, and he shal haue
forgeuenes for euerie thing that in doing he ſinned. \V And Our Lord
ſpake to Moyſes, ſaying: \V Command Aaron and his ſonnes: This is the
Law of an holocauſt: It shal be burnt vpon the altar, al night vntil
morning: the fire shal be vpon the ſame altar. \V The prieſt shal be
reueſted with the tunike and the linnen femoralles: and he shal take vp
the ashes, which the deuouring fire burned, and putting them beſides the
altar, \V shal be vnueſted of his former veſtments, and being clothed
with others, shal carie them forth without the campe, and in a moſt
cleane place shal cauſe them to be conſumed vnto duſt. \V And the fire
on the altar
\SNote{This fire was firſt ſent miraculouſly from God.
\XRef{(infra. c.~9. v.~24.)}
and according to this cõmandment, was
\Fix{perperpetually}{perpetually}{obvious typo, fixed in other}
conſerued: from which al fire was to be taken that was vſed in the
tabernacle.
\Cite{S.~Aug. q.~31.}
&
\Cite{Theod. q.~9. in Leuiticum.}
Myſtically it ſignified, that the fire of charitie being firſt kindled
in mans hart by Gods grace, muſt be continually nouriſhed and kept
burning, from which al other good workes are deriued.}
shal alwaies burne, which the prieſt shal nourish, putting wood
vnderneth, in the morning euerie
%%% o-0264
day, and laying on the holocauſt, therupon shal burne the fatte of the
pacifiques. \V This fire is continual which shal neuer faile on the
altar. \V This is the Law of the ſacrifice and libamentes, which the
children of Aaron shal offer before the Lord, and before the altar. \V
The prieſt shal take vp a handful of the floure, that is tempered with
oyle, and al the frankincenſe, that is put vpon the floure: and he shal
burne it on the altar for a moniment of moſt ſweete odour to the
Lord: \V and the part of the floure that is left, shal Aaron eate with
his ſonnes, without leauen: and he shal eate it in the Holie place of
the court of the tabernacle. \V And therefore it shal not be leauened,
becauſe part therof is offered for the burnt ſacrifice of the Lord. It
shal be moſt Holie, as that for ſinne, and for offence. \V The males
onlie of Aarons ſtocke shal eate it. \V It shal be an ordinance and
euerlaſting in your generations of the ſacrifices of the Lord. Euerie
one that toucheth them, shal be ſanctified. \V And our Lord ſpake to
Moyſes, ſaying: \V This is the oblation of Aaron, and of his ſonnes,
which they muſt offer to the Lord, in the day of their vnction. The
tenth part of an ephi of floure shal they offer in a ſacrifice for euer,
halfe therof in the morning, and halfe therof at euen: \V which being
tempered with oile shal be fried in a frying panne. \V And the prieſt
%%% 0292
that by right ſucceedeth his father, shal offer it hote, for a moſt
ſwete odour to the Lord, and it shal wholy be burnt on the altar. \V For
euerie ſacrifice of the prieſt shal be conſumed with fire, neither shal
anie man eate therof. \V And our Lord ſpake to Moyſes, ſaying: \V Speake
to Aaron and his ſonnes: This is the law of the hoſte for ſinne. In the
place where the holocauſt is offered, shal it be immolated before our
Lord. It is Holie of holies. \V The prieſt that doth offer it, shal eate
it in a holie place, in the courte of the tabernacle. \V Whatſoeuer shal
touch
\SNote{By fleſh of penance is vnderſtood, faſting, watchĩg, haircloth,
teares, prayers, almes which whoſoeuer duly toucheth ſhal be ſanctified.
\Cite{Heſychius Hieroſol. li.~2. in Leuit. c.~6.}}
the flesh therof, shal be ſanctified. If of the bloud therof a garment
be ſprinkled, it shal be washed in a holie place. \V And the earthen
veſſel, wherin it was ſodden, shal be broken, but if the veſſel be of
braſſe, it shal be ſcoured, and washed with water. \V Euerie male of the
prieſtlie race shal eate of the flesh therof, becauſe it is Holie of
holies. \V For the hoſte that is ſlaine for ſinne, whoſe bloud is caried
into the tabernacle of teſtimonie to make expiation in the Sanctuarie,
shal not be eaten, but shal be burnt with fire.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
