%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 0299
%%% o-0271
\startChapter[
  title={Chapter 10}
  ]

\Summary{Nadab and Abiu the ſonnes of Aaron, for offering ſtrange fire,
  are burnt to death, and caſt out of the campe, 6.~for whom the people
  mourne, but not the Prieſts. 8.~Prieſts are forbid to drinke wine,
  when they enter into the tabernacle, 12.~and are commanded to eat the
  reſidew of oblations in the holie place. 16.~Which this time in part
  they omitted, and are excuſed, being ſorowful for that vvhich happened
  to Nadab and Abiu.}

%%% 0300
And Nadab, and Abiu the ſonnes of Aaron catching cenſors, did put in
fire, and incenſe therupon, offering before our Lord ſtrange fire: which
was not commanded them. \V And fire comming forth from our Lord,
\SNote{Such as receiued more at Gods hand, are more ſeuerly puniſhed if
they tranſgreſſe.
\Cite{S.~Aug. q.~21. in Leuit.}
By this alſo al are warned to be content with the doctrine of the Holie
Ghoſt, to abhorre hereſies, the fautors wherof adde falſehood to Gods
word, & preferre their owne wicked inuentions before the true ſenſe of
holie Scripture.
\Cite{Theod. q.~9. in Leuit.}}
deuoured them, and they dyed before our Lord. \V And Moyſes ſaid to
Aaron: This is it which our Lord hath ſpoken: I wil be ſanctified in
them, that approch to me, and in the ſight of al the people I wil be
glorified. Which Aaron hearing held his peace. \V And Moyſes calling
Miſael, and Eliſaphan the ſonnes of Oziel, the vncle of Aaron, ſaid to
them: Goe and take away your brethren from the ſight of the Sanctuarie, and
carie them without the campe. \V And going forthwith they tooke them as
they lay, reueſted with linnen tunikes, & did caſt them forth, as it had
bene commanded them. \V And Moyſes ſpake to Aaron, & to Eleazar, and
Ithamar his ſonnes:
%%% o-0272
Vncouer not your heades, and rent not your
veſtiments, leſt perhaps you die, and indignation come vpon al the
aſſemblie. Let your brethren, and al the houſe of Iſrael, lament the
burning that our Lord hath raiſed, \V and your ſelues ſhal not goe out
of the dore of the tabernacle, otherwiſe you shal perish: for the oyle
of holie vnction is vpon you. Who did al thinges according to the
precept of Moyſes. \V Our Lord alſo ſaid to Aaron: \V Wine, and anie
thing that may make drunke, you shal not drinke, thou and thy ſonnes,
\SNote{Abſtinence from wine cõmanded to Prieſts when they ſerued in the
tabernacle not at other times: for they ſerued at certaine times by
turnes.
\Cite{Theod. q.~10. in Leuit.}
As for drunkennes it is forbid to al men, and at al times.}
when you enter into the tabernacle of teſtimonie, leſt you die: becauſe
it is an euerlaſting precept through your generations. \V And that you
may haue knowledge to diſcerne betwen the holie and prophane, betwen the
polluted and cleane: \V and may teach the children of Iſrael al my
ordinances, which the Lord hath ſpoken to them by the hand of Moyſes. \V
And Moyſes ſpake to Aaron and to Eleazar, and Ithamar his ſonnes, that
were left: Take the ſacrifice, that is remayning of the oblation of our
Lord, and eate it without leauen beſide the altar, becauſe it is Holie
of holies. \V And you ſhal eate it in a holie place: which is giuen to
thee and thy ſonnes of the oblations of our Lord, as it hath bene
commanded me. \V The breſt alſo that is offered, and the ſhoulder that
is ſeperated, you shal eate in a moſt cleane place thou and thy ſonnes,
and thy daughters with thee. For they are laid apart for thee and thy
children, of the healthful hoſtes of the children of Iſrael: \V becauſe
the ſhoulder and the
%%% 0301
\Fix{the breſt,}{breſt,}{obvious typo, fixed in other}
and the fatte that is burnt on the altar, they haue eleuated before our
Lord, and they pertaine to thee, and to thy ſonnes by a perpetual lawe,
as our Lord hath commanded. \V Among theſe thinges, when Moyſes ſought
for the bucke goate, that had bene offered for ſinne, he found it burnt:
and being angrie againſt Eleazar, and Ithamar the ſonnes of Aaron that
remained, he ſaid: \V Why did you not eate the hoſte for ſinne, in a
holie place, which is moſt Holie, and geuen to you, that you may beare
the iniquitie of the multitude, and may pray for it in the ſight of our
Lord, \V eſpecially wheras of the bloud therof, there hath not bene
caried within the holie places, and you ought to haue eaten it in the
Sanctuarie, as it was commanded me? \V Aaron anſwered: This day hath
bene offered the victime for ſinne, and the holocauſt before our Lord:
and to me that is chanced which thou ſeeſt. How could I eate it, or
pleaſe our Lord in ceremonies hauing
\SNote{Natural grife of mind made Aaron both vnwilling to eate & leſſe
apt to complete al the ceremonies: ſo without ſinne he omitted that
pertained to his commoditie, offering it to God.}
a ſorowful hart? \V Which when Moyſes had heard, he was ſatisfied with
his anſwere.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
