%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 0316
%%% o-0286
\startChapter[
  title={Chapter 18}
  ]

\Summary{Mariage prohibited in certaine degrees of conſanguinitie and
  affinitie. 18.~And diuers carnal, and execrable ſinnes committed in
  other nations, are ſtrictly forbidden.}

And our Lord ſpake to Moyſes, ſaying: \V Speake to the children of
Iſrael, and thou shalt ſay to them: I the
%%% 0317
Lord your God, \V according to the cuſtome of the Land of Ægypt, wherin
you haue dwelt, you ſhal not doe: and according to the maner of the
Countrie of Chanaan, into the which I wil bring you, you ſhal not doe,
nor walke in their ordinãces. \V You ſhal doe my iudgements, and shal
obſerue my precepts, and shal walke in them. I the Lord your God. \V
Keepe my lawes and iudgmentes, which a man doing, shal liue in them. I
the Lord. \V No man shal approch to her that is
\LNote{Next of his bloud.}{Mariage
\MNote{Mariage forbid in al degrees in the right line, by the law of
nature.}
is forbid firſt and moſt ſtrictly by the law of nature, in al degrees in
the right line aſcending and deſcending, both in conſanguinitie and
affinitie. S.~Paul teſtifying that
\CNote{1.~Cor.~5. v.~1.}
\Emph{among the heathen, no man could haue his fathers vvife.} And in
the right line God him ſelfe (who onlie can) neuer diſpenſed.
\MNote{Secondarily in the firſt collateral degree of cõſanguinitie.}
Secondarily, the firſt collateral degree in conſanguinitie, that is,
betwen brother and ſiſter, by one parent, or by both, is alſo vnlawful by
the law of nature, except in the beginning of the world, when Adams
children muſt nedes marie together,
\CNote{\XRef{Act.~17. v.~26.}
\Cite{S.~Aug. de bono cõiug. c.~1.}}
God ſo ordayning that al mankind ſhould be propagated by one man (for of
him alſo the firſt woman was made) but after this beginning it was neuer
allowed, nor perhaps can be diſpenſed withal, at leaſt neuer was by anie
man.
\MNote{Beza belieth Pope Martin.}
Though Beza
\Cite{(li. de repudijs & diuortijs.)}
and ſome English Bezites charge Pope Martin the fifth, to haue diſpenſed
with one, that had maried his owne natural ſiſter: which is a falſe
reporte. For it was with one, who hauing committed fornication with one
ſiſter, afterwardes maried the other, from whom he could not be
ſeparated without great ſcandal, the pretended mariage being publike,
and the impediment ſecrete: as S.~Antoninus writeth.
\Cite{par.~3. ſum. Theol. tit.~1. c.~11.}
\MNote{Al other degrees depend on poſitiue lawes: which haue bene & may
be altered.}
But beſides the right line, and the firſt collateral degree in
conſanguinitie, no other collateral degrees are prohibited by the law
of nature, but by poſitiue only. So this preſent law, written by Moyſes,
forbade to marie in the firſt collateral degree of affinitie, but the
ſame law commanded
\XRef{(Deut.~25.)}
that in caſe a maried man died without iſſue, his brother ſhould marie
the widow. VVherby is clere that this degree, and others more remote,
were  not prohibited by the law of nature.
\MNote{Proued by Scriptures, and reaſons.}
For then God would not haue
made a contrarie general law, in anie caſe, for the whole nation of the
Iewes, his people; and that vnder penaltie to be obſerued, which is
contrarie to the qualitie of indulgence or diſpenſation, and no ſuch
neceſſitie, as in the beginning of the world.
\MNote{Firſt proofe.}
VVherfore al proteſtants that ſay, the whole law written by Moyſes
concerning degrees of conſanguinitie and affinitie, is the law of
nature, and ſo 
pertaineth to Chriſtians, muſt neceſſarily ſay alſo, that if now a
maried man die without iſſue, his brother muſt marie his wife. VVhich
ſpecially they denie.
\MNote{2.~proofe.}
It is alſo proued that this and ſome other degrees expreſſed in this
place, were not againſt the law of nature 
\CNote{\Cite{Ariſtot. li.~2. Pol.}}
(which is common to al
nations, commonly or eaſely knowne to al men by diſcourſe of reaſon)
becauſe no common wealth among the Gentiles did puniſh, nor modeſt men
forbeare, or reprehend ſuch mariages: as appeareth by Laban, who after
he had deceiued Iacob by geuing him one ſiſter for an other, offered him
alſo the former promiſed, whom without difficultie of conſcience he
accepted
\XRef{(Gen.~29.)}
neither did that holie Patriarch thinke it
\Fix{vnlauful}{vnlawful}{possible typo, fixed in other}
to keepe them both. And when Iudas matched his ſecond ſonne, and
promiſed the third to the wife of his firſt ſonne, he did it according
to the cuſtome of that place & time.
\XRef{Gen.~38.}
And Noemi ſpoke according to the ſame cuſtome.
\XRef{Ruth.~1. v.~11.}
\MNote{3.~proofe.}
Againe where this law forbiddeth a man to marie, or companie with his
wiues ſiſter, it addeth, \Emph{vvhiles she is liuing}, not prohibiting
mariage, when his firſt wife is dead. Yet his wiues ſiſter is as nere
in affinitie, as his brothers wife.
\MNote{4.~proofe.}
Likewiſe the diuerſitie of puniſhments
\XRef{(chap.~20.)}
for tranſgreſſion of this law, either in the right line, or in the firſt
collateral degree of conſanguinitie, who were puniſhed by death; and for
tranſgreſſing in the firſt collateral degree of affinitie, or in the
ſecond either of conſanguinitie or affinitie, who had leſſe puniſhments,
ſheweth that the former degrees are prohibited by the law of nature, and
not the other: for then the violation ſhould be like ſinne, and puniſhed
alike.
\MNote{5.~proofe.}
Finally it is euident, that certaine of theſe degrees are not againſt
the law of nature, by the example of holie Abraham, who in, and
according, to the law of nature, maried his brothers daughter called
Sarai, otherwiſe Ieſcha,
\XRef{Gen.~11.}
\CNote{S.~Hiero. queſt. Heb. in Gen.}
which mariage God approued by manie bleſſings. Alſo Iacob maried two
ſiſters together. Two ſonnes of Iudas maried the ſame woman
ſucceſſiuely. And Amram (Moyſes father) maried his aunt, his fathers
ſiſter.
\XRef{Exod.~6. v.~20.}
\XRef{Num.~26. v.~59.}
VVherfore ſeing neither the firſt collateral degree in affinitie, nor
the ſecond collateral in conſanguinitie or affinitie, is forbid by the
law of nature but by poſitiue only, and that 
\MNote{Ceremonial & iudicial lawes of Moyſes are abrogated by Chriſt.}
both ceremonial and
iudicial lawes of the old Teſtament ceaſſed in the New, and are
abrogated by Chriſt, it reſteth proued that the ſame bind not
Chriſtians, but as they are renewed and eſtabliſhed by the Church, or
Chriſtian commonwelthes. And as this is donne in temporal cauſes by
temporal States, partly by renewing and eſtabliſhing the ſame, which was
in the law of Moyſes, as by puniſhing wilful murder by death;
\XRef{Exod.~21. v.~12.}
partly with alteration, as by puniſhing theift in ſome countries with
death, but not adultrie, which were contrarie in the old Teſtament,
\XRef{Gen.~38. v.~24.}
\XRef{44.~17.}
\XRef{Exod.~22. v.~1.}
\XRef{Leuit.~20. v.~10.}:
\MNote{And new are eſtabliſhed.}
in like ſorte the Church of Chriſt ordaineth lawes, altereth, & vpon
iuſt occaſions diſpenceth, in al degrees of conſanguinitie and
affinitie, not forbid by the law of nature.}
next of his bloud, to reueale her
\SNote{It is then turpitude when the act is vnlawful. But honeſt in
lawful Mariage.
\Cite{S.~Aug. li.~3. Locutionum.}
&
\Cite{li. de bono coniugali.}}
turpitude. I the Lord. \V The turpitude of thy father, and the turpitude
of thy mother thou shalt not diſcouer: ſhe is thy mother thou shalt not
reueale her turpitude. \V The turpitude of thy fathers wife thou shalt
not diſcouer: for it is the turpitude of thy father. \V The turpitude of
thy ſiſter by father, or by mother, which was borne at home or abroad,
thou shalt not reueale. \V The turpitude of thy ſonnes daughter or of
thy neece by thy daughter, thou shalt not reueale: becauſe it is thy
turpitude. \V The turpitude of thy fathers wiues daughter, which she
bare to thy father, and is thy ſiſter, thou shal not reueale. \V The
turpitude of
\SNote{See
\XRef{chap.~20.}
the difference of puniſhmẽts for violating theſe lawes in the firſt and
ſecond degree. Alſo betwen conſanguinitie & affinitie, in the ſame
collateral degree.}
thy fathers ſiſter thou shalt not diſcouer: becauſe she is the flesh of
thy father. \V The turpitude of thy mothers ſiſter thou shalt not
reueale, becauſe she is the flesh of thy mother. \V The turpitude of thy
fathers brother thou shalt not reueale, neither shalt thou approch to
his wife, who is ioyned to thee by affinitie. \V The turpitude of thy
daughter in law thou shalt not reueale, becauſe she is thy ſonnes wife,
neither shalt thou diſcouer her ignominie. \V The turpitude of thy
brothers wife thou shalt not reueale: becauſe it is the turpitude of thy
brother. \V The turpitude of thy wife, and her daughter thou shalt not
reueale. Her ſonnes daughter, and her daughters daughter, thou shalt not
take, to reueale her ignominie: becauſe they are her flesh, and ſuch
copulation is inceſt. \V Thou shalt not take thy wiues ſiſter for an
harlote, to vexe her withal, neither shalt thou reueale her
%%% o-0287
turpitude, whiles she is yet liuing. \V To a woman, hauing her flowers,
thou shalt not approch, neither shalt thou reueale her turpitude. \V
With thy neighbours wife thou shalt not companie, nor be polluted with
commixtion of ſeede. \V Of thy ſeede thou shalt not geue to be
conſecrated to the idol Moloch, nor pollute the name of thy God: I the
Lord. \V Companie not with mankind, as with womankind,
%%% 0318
becauſe it is abomination. \V With no beaſt shalt thou companie, neither
shalt thou be polluted with it. A woman shal not lie downe to a beaſt,
nor companie with it: becauſe it is an hainous fact. \V Neither be ye
polluted in anie of the thinges wherwith al the nations haue bene
contaminated, which I wil caſt out before your ſight, \V and wherwith
the land is polluted: whoſe abominations I wil viſite, that it vomite
out the inhabitants therof. \V Keepe my ordinances and iudgements, and
doe not any of theſe abominations, as wel the ſame countrieman as the
ſtranger, that ſeiourneth with you. \V For al theſe execrable thinges
did the inhabitants of the land, that haue bene before you, and haue
polluted it. \V Beware therfore leſt in like maner it vomite out you
alſo, when you shal doe the like thinges, as it vomited out the nation
that was before you. \V Euerie ſoule, that ſhal doe anie of theſe
abominations, shal perish from the middes of his people. \V Keepe my
commandements. Doe not the thinges which they haue done, that haue bene
before you, and be not polluted in them: I the Lord your God.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
