%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 25
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-25


\project douay-rheims


%%% 0331
%%% o-0299
\startChapter[
  title={Chapter 25}
  ]

\Summary{The law of the ſeuenth yeare, 8.~and fiftith yeare, which
  is the Iubilie, 13.~when al enheritance ſold returneth to the former
  owner (24.~as alſo it may in the meane time be redeemed.) 35.~Vſurie
  prohibited, 39.~and ſeruitude among the Iſraelites, only they may be
  hyred til the Iubilee yeare, 47.~and may be redeemed from ſeruitude of
  ſtrangers: 54.~at leaſt they shal be free in the yeare of Iubilee.}

And our Lord ſpake to Moyſes in the mount Sinai, ſaying: \V Speake to
the children of Iſrael, and thou shalt ſay to them: When you shal be
entred to the land which
%%% 0332
I wil geue you, thou shalt ſabbatize the ſabbath to the Lord. \V Six
yeares thou shalt ſowe thy field, and ſix yeares thou ſhalt cut thy
vineyard, and ſhalt gather the fruites therof: \V but in the ſeuenth
yeare a ſabbath shal be to the earth, of the reſting of the Lord: the
field thou shalt not ſowe, and the vineyard thou shalt not cut. \V The
thinges that the ground shal bring forth of it ſelf, thou shalt not
reape: and the grapes of thy firſt fruites thou shalt not gather as a
vintage: for it is a yeare of the reſting of the earth: \V but they ſhal
be vnto you for meate, to thee and thy manſeruant, to thy woman ſeruant
and hyreling, and to the ſtranger that ſeiourneth with thee: \V to thy
beaſtes and cattel al thinges that grow shal giue meate. \V Thou shalt
number thee alſo ſeuen weekes of yeares, that is to ſay, ſeuen times
ſeuen, which together make fourtie nine, yeares: \V and thou
\SNote{Of ſounding with trũpets, which is pleaſant & ioyful, cometh the
name of Iubilee: the effect of it is remiſſion of al bondes,
reſtauration of former libertie, and recouerie of enheritãce. In the old
Teſtament of tẽporal things, in the new of ſpiritual, præfigured therby: as
remiſſiõ of ſinne; deliuerie frõ bondage therof; recouerie of grace; and
preparation to eternal glorie.}
shalt ſound with the trumpet the ſeuenth moneth, the tenth day of the
moneth, in the time of propitiation in al your land. \V Thou shalt
ſanctifie the fifteth yeare, and shalt proclaime remiſſion to al the
inhabitantes of thy land: for it is the yeare of Iubilee. Euerie man
shal returne to his poſſeſſion, and euerie one shal goe backe to his old
familie: \V becauſe it is the Iubilee and the fifteth yeare. You shal
not ſowe nor reape the thinges that growe in the field of their owne
accord, and the firſt fruites of vintage you shal not gather, \V becauſe
of the ſanctification of the Iubilee, but forthwith as they grow you
shal eate them. \V In the yeare of Iubilee al shal returne to their
poſſeſſions. \V When thou ſhalt ſel anie thing to thy neighbour, or
shalt buy of him, preſſe not thy brother, but according to the number of
the yeares of
%%% o-0300
Iubilee thou shalt buy of him, \V and according to the ſupputation of
the fruites he shal ſel to thee. \V The moe yeares remaine after the
Iubilee, ſo much more shal the price increaſe: and the leſſe time that
thou shal account, ſo much the leſſe shal the purchaſe be valued,
for the time of the fruites he shal ſel to thee. \V Doe not afflict
your contrimen, but let euerie one feare his God, becauſe I the Lord
your God. \V Do my preceptes, and keepe my iudgementes, and fulfill them:
that you may dwel in the land without anie feare, \V and the ground may
yeld you her fruites, which you may eate vnto your fil, fearing no mans
inuaſion. \V But if you ſay: What shal we
%%% 0333
eate the ſeuenth yeare, if we ſowe not, nor gather our fruites? \V I wil
geue you my benediction the ſixt yeare, and it shal yeld the fruites of
three yeares: \V and the eight yeare you shal ſowe, and shal eate of the
old fruites, vntil the ninth yeare: til new be growne, you shal eate the
old. \V The land alſo shal not be ſould for euer: becauſe it is mine,
and you are my ſtrangers and ſeiourners. \V For the which cauſe al the
countrie of you poſſeſſion shal be ſould vnder the condition of
redemption. \V If thy brother impouerished ſel his litle poſſeſſion, and
his kinſeman wil, he may redeme that which he had ſould. \V But if he
haue no kinſeman, and him ſelf can finde the price to redeme it: \V the
fruites shal be accounted from that time when he ſould it: and the
reſidue he shal reſtore to the buier, and ſo he shal receiue his
poſſeſſion againe. \V But if his hand finde not to repay the price, the buier
shal haue that he bought, vntil the yeare of Iubilee. For in it al ſale
shal returne to the owner, and to the old poſſeſſour. \V He that ſelleth
a houſe within the walles of a citie, shal haue licence to redeme it,
vntil one yeare be expired, \V if he redeme it not, and the compaſſe of
the yeare be fully out, the buier shal poſſeſſe it, and his poſteritie
for euer, and it can not be redemed, no not in the Iubilee. \V But if
the houſe be in a village, that hath not walles, it shal be ſould
according to the law of fieldes. If it be not redemed before, in the
Iubilee it ſhal returne to the owner. \V The houſes of Leuites, which
are in cities, may alwaies be redemed: \V if they be not redemed, in the
Iubilee they shal returne to the owners, becauſe the houſes of the
cities of the Leuites are for poſſeſſions among the children of
Iſrael. \V But let not their ſuburbes be ſould, becauſe it is a
perpetual poſſeſſion. \V If thy brother be empouerished, and weake of
hand, and thou receiue him as a ſtranger and ſeiourner, and he liue with
thee, \V take not vſuries
\SNote{Iewes for their aduãtage hold it lawful to take vſurie of
ſtrangers, not obſeruing that it is alſo commanded often in
ſcripture, \Emph{not to afflict, but to loue ſtrangers}.
\XRef{Exod.~22.~23.}
\XRef{Leuit.~19.}}
of him, nor more then thou gaueſt. Feare thy God, that thy brother may
liue with thee. \V Thou ſhalt not geue him thy money to vſurie, and an
ouer plus of the fruites thou shalt not exact of him. \V I the Lord your
God, that brought you out of the Land of Ægypt, that I might geue you
the Land of Chanaan, and might be your God. \V If thy brother
conſtrained by pouertie, ſel him ſelf to thee, thou shalt not oppreſſe
him with the ſeruitude of ſeruants, \V but he shal be as an hireling,
and a ſeiourner:
%%% 0334
vntil the yeare of Iubilee he shal worke with thee, \V and afterward he
shal goe out
%%% o-0301
with his children, and shal returne to his kinred and to the poſſeſſion
of his fathers, \V for they are my ſeruantes, and I brought them out of
the Land of Ægypt. Let them not be ſould by the condition of
ſeruantes: \V afflict him not by might, but feare thy God. \V Let your
man ſeruant, and woman ſeruant, be of the nations that are round about
you. \V And of the ſtrangers, that ſeiourne with you, or that were borne of
them in your land, theſe you ſhal haue for ſeruantes: \V and by right of
inheritance shal leaue them to your poſteritie, and shal poſſeſſe
them for euer. But your brethren the children of Iſrael doe ye not
oppreſſe by might. \V If the hand of a ſtranger or ſeiourner grow ſtrong
among you, and thy brother empouerished ſel him ſelf to him, or to any
of his ſtocke: \V after the ſale he may be redeemed. He that wil of his
brethren shal redeme him, \V both the vncle by father, and the vncles
ſonne, and the kinſman, and the allied. But and if him ſelf be able alſo, he
shal redeme him ſelfe, \V accounting onlie the yeares from the time of
his ſelling vnto the yeare of Iubilee: and accounting the money, that he
was ſould for, according to the number of the yeares and the reckning of
an hyreling. \V If they be more yeares that remaine vntil
the Iubilee, according to theſe alſo shal he repay the price. \V If few,
he shal make the reckning with him according to the number of the
yeares, and shal repay to the buyer for that which remaineth of the
yeares, \V his wages being allowed for the which he ſerued before: he
shal not afflict him violently in thy ſight. \V And if by theſe meanes
he can not be redemed, in the yeare of Iubilee he shal goe out with his
children. \V For the children of Iſrael are my ſeruantes, whom I brought
forth out of the Land of Ægypt.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
