%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 0289
%%% o-0262
\startChapter[
  title={Chapter 5}
  ]

\Summary{Of hoſtes, for the ſinne of concealing an others periurie,
  2.~for vncleanes, 4.~for vaine ſvvearing, 14.~for errour in exerciſing
  holie rites, 17.~& for anie ſinne committed by ignorance.}

If a ſoule ſinne, and heare the voice of one ſwearing, and be witnes
becauſe either he him ſelfe ſawe, or is priuie to it:
\SNote{VVhen periurie doth preiudice an others cauſe, he that knoweth
the truth is bond to reueil it to the Iudges; but with diſcretion to
auoide ſcandal.}
vleſſe he vtter it, he ſhal beare his iniquitie. \V The ſoule that
toucheth anie vncleane thing, either that which was killed of a beaſt,
or died of it ſelfe, or anie other thing that creepeth: and forgeteth
his vncleannes is guiltie, and hath offended: \V and if he touch anie
thing of the vncleanneſſe of man, according to anie impuritie wherwith
he is wount to be polluted, & hauing forgotten, doe knowe it afterward,
he shal be guiltie of an offence. \V The ſoule that ſweareth, and
vttereth with his lippes, that he would doe
\SNote{Swearing to do that is euil, or not doing that is lawfully
ſworne, is ſinne.}
either il; or wel, and bindeth the ſame with an oathe, and his word, &
hauing forgotten afterward vnderſtandeth his offence, \V let him do
penance for his ſinne, \V and offer of the flockes an ewe lambe, or a
shee goate, and the prieſt shal pray for him, and for his
ſinne: \V but if he be not able to offer a beaſt, let him offer two
turtles, or two young pigions to our Lord, one for ſinne, and the other
for an holocauſt, \V and he shal geue them to the prieſt: who offering
the firſt for ſinne, shal wryth backe the heade therof to the litle
pinions, ſo that it ſticke to the necke, and be not altogeather broken
of. \V And of the bloud therof he shal ſprinckle the wall of the altar,
and whatſoeuer is left, he shal make it diſtil to the bottome therof,
becauſe it is for ſinne.
%%% 0290
\V And the other he shal burne for an holocauſte, as is wount to be
done: and the prieſt shal pray for him, and for his ſinne, and it shal
be forgeuen him. \V And if his hand be not able to offer two turtles, or
two young pigions, he shal offer for his ſinne, of floure the tenth part
of an ephi. He shal not put oyle vpon it, nor caſt anie frankincenſe
theron, becauſe it is for ſinne, \V and he shal deliuer it to the
prieſt: who taking therof a ful handful, shal burne it vpon the altar
for a moniment, of him that did offer it: \V praying for him and making
expiation, but the part that is left, him ſelfe shal haue for a
gifte. \V And our Lord ſpake to Moyſes, ſaying: \V If a ſoule
tranſgreſſing the ceremonies, by errour shal ſinne in thoſe thinges that are
ſanctified to our Lord, he shal offer for his offence a
ramme, without ſpotte out of the flockes, that may be bought for two
ſicles, according to the weight of the Sanctuarie:
%%% o-0263
\V and
\SNote{For remiſſion of ſinne reſtitution is firſt required if iniurie
was done.}
the damage it ſelfe which he did, he shal reſtore, and the fift part he
shal adde beſides, deliuering it to the prieſt, who shal pray for him,
\SNote{Beſides reſtitution ſatisfaction is alſo neceſſarie for the
offence to God.
\Cite{Theodoret. q.~2. in Leuitic.}}
offering the ramme, and it ſhal be forgeuen him. \V If a ſoule ſinne by
ignorance, and do one of thoſe thinges which by the lawe of the Lord are
forbidden, and being guiltie of ſinne, vnderſtand his iniquitie, \V he
shal offer a ramme without ſpotte of the flockes to the prieſt,
according to the meaſure, and eſtimation of the ſinne: who shal pray for
him, becauſe he did it vnwitting: and it shal be forgiuen him, \V
becauſe by errour he offended againſt the Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
