%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 0313
%%% o-0283
\startChapter[
  title={Chapter 16}
  ]

\Summary{VVhen and how the high Prieſt muſt enter into the
  Sanctuarie. 14.~How he shal expiate (or reconcile) the ſame, 16.~and
  the Tabernacle, 18.~and the Altar. 20.~How he shal offer a liue goate,
  and ſend him into the wildernes. 29.~And al muſt celebrate the feaſt
  of expiation, or clenſing from ſinnes.}

And our Lord ſpake to Moyſes, after the death of the two ſonnes of
Aaron, when they were ſlaine offering ſtrange fire: \V and he commanded
him, ſaying: Speake to Aaron thy brother, that he
\SNote{Only once in the yeare the high prieſt, and no other, entred into
\L{Sancta Sanctorum}.
\XRef{v.~34.}
ſignifying that heauen was not open to anie Sainct, before Chriſts
paſſion:
\XRef{Heb.~9. v.~8.}}
enter not at al times into the Sanctuarie, that is within the veile
before the propitiatorie, wherwith the arke is couered, leſt he die (for
in a cloude wil I appeare ouer the oracle) \V vnleſſe he doe theſe
thinges before: He shal offer a calfe for ſinne, and a ramme for an
holocauſt. \V He shal be reueſted with a linnen tunike, he shal hide his
priuities with linnen femoralles: he shal be girded with a linnen
girdle, a linnen mitre shal be put vpon his head: for theſe are holie
veſtmentes: withal which, when he is washed, he shal be reueſted. \V And
he shal receiue of the whole multitude of the children of Iſrael two
bucke goates for ſinne, and one ramme for an holocauſt. \V And when he
hath offered the calfe, and praied for him ſelf, and for his owne
houſe, \V he shal make the two bucke goates to ſtand before the Lord in
the dore of the tabernacle of teſtimonie: \V and caſting lottes vpon
both, one for the Lord, & an other for the goate of diſmiſſion: \V that,
whoſe lotte fel to the Lord, he shal offer for ſinne: \V but that, whoſe
lotte was to be the goate of diſmiſſion, he shal ſet aliue before
the Lord, that he may
\SNote{Praying that al their ſinnes may be remitted.}
poure out prayers vpon him, and diſmiſſe him into the wilderneſſe. \V
After that theſe thinges be
%%% 0314
duly celebrated, he shal offer the calfe, and praying for him ſelf and
for his houſe, he shal immolate it: \V and taking the thurible, which he
hath filled of the burning coales of the altar, and taking vp with his
hand of the compounded perfume for incenſe, he shal goe in beyond the
veile into the Holie place: \V that when the incenſe is put vpon the
fire, the cloude therof and the vapour may couer the oracle, which is
ouer the teſtimonie, and he die not. \V He shal take alſo of the bloud
of the calfe, and ſprinckle with his finger ſeuen times againſt the
propitiatorie to the eaſt. \V And when he hath killed the bucke goate
for the ſinne of the people, he shal carie in the bloud therof within
the veile, as hath bene
%%% o-0284
commanded of the bloud of the calfe, that he may ſprinckle it againſt
the oracle, \V and
\SNote{Sinnes do ſo defile the ſoule that the very holieſt of al places
is accounted as contaminate therby.
\Cite{Theodo. q.~22. in Leuit. in fine.}}
expiate the Sanctuarie from the vncleanneſſe of the children of Iſrael,
and from their preuarications, and al their ſinnes. According to this
rite shal he doe to the tabernacle of teſtimonie, which is fixed among
them in the middeſt of the filth of their habitation. \V Let no man be
in the tabernacle when the high prieſt goeth into the Sanctuarie, to
pray for him ſelf and for his houſe, & for the whole aſſemblie of
Iſrael, vntil he come forth. \V And when he is come forth to the altar
that is before the Lord, let him pray for him ſelf, and taking the bloud
of the calfe, and of the bucke goate, let him poure it vpon the hornes
therof round about: \V and ſprinckling with his finger ſeuen times, let
him expiate, and ſanctifie it from the vncleanneſſe of the children of
Iſrael. \V After he hath clenſed the Sanctuarie, and the tabernacle, and
the altar, then let him offer the liue goate: \V and putting both handes
vpon his head, let him confeſſe al the iniquities of the children of
Iſrael, and al their offences and ſinnes: which praying to light on his
head, he shal ſend him forth by a man ready therto, into the deſert. \V
And
\SNote{God ſo remitteth ſinnes to thoſe that are truly 
\Fix{peninitent,}{penitent,}{obvious typo, fixed in other}
as that
which is caried into a wildernes and neuer returneth.}
when the goat hath caried al their iniquities into the ſolitarie ground,
and shal be let goe into the deſert, \V Aaron shal returne into the
tabernacle of teſtimonie, and putting of the veſtiments, which he had on
him before when he entred into the Sanctuarie, and leauing them
there, \V he shal wash his flesh in a holie place, and shal be clothed
with his owne garments. And after that he hath gone forth and offered
his owne holocauſt, and the peoples, he shal pray as wel for him ſelf,
as for the people: \V and the
%%% 0315
fatte, that is offered for ſinnes, he shal burne vpon the altar, \V but
he, that hath let goe the goate of diſmiſſion, shal wash his clothes,
and bodie with water, and ſo shal enter into the campe. \V But the calfe
& the bucke goate, that were immolated for ſinne, and whoſe bloud was
caried into the Sanctuarie, to accomplish the expiation, they shal carie
forth without the campe, and shal burne with fire aſwel the skinnes as
their flesh, and the dung: \V and whoſoeuer burneth them, shal wash his
clothes, and his flesh with water, and ſo shal enter into the campe. \V
And this shal be to you an euerlaſting ordinance: The
\SNote{Beſides particular ſacrifices for euerie ones ſinnes, once in the
yeare was inſtituted a general expiatiõ of al.}
ſeuenth moneth, the tenth day of the moneth you shal afflict your
ſoules, and no worke shal you doe, whether he be of the ſame countrie, or
a ſtranger that ſoiourneth among you. \V Vpon this day shal be the
expiation of you, and clenſing from al your ſinnes: before the Lord you
shal be clenſen. \V For it is a ſabath of reſt, and you shal afflict
your ſoules by a perpetual religion. \V And the prieſt shal expiate, that
is annoynted, and whoſe handes are conſecrated to do the function of
prieſthood for his father: and he shal be reueſted with the linnen ſtole
and the holie veſtments, \V and he shal expiate the Sanctuarie and the
tabernacle of teſtimonie and the altar, the prieſtes alſo and al the
people. \V And this shal be an ordinance for euer, that you pray for the
children of Iſrael, and for al their ſinnes once in a yeare. He did
therfore as our Lord had commanded Moyſes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
