%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 0321
%%% o-0290
\startChapter[
  title={Chapter 20}
  ]

\Summary{VVhoſoeuer geueth of his ſeede to Moloch muſt be ſtoned to
  death. 6.~Al that decline to Magicke, 9.~curſe their parents,
  10.~commit adultrie, certaine inceſt, or beſtialitie shal
  die. 19.~Other inceſt is depriued of children. 22.~The Iſraelites alſo
  shal be caſt out of the Land, if they committe ſuch ſinnes.}

%%% 0322
And our Lord ſpake to Moyſes, ſaying: \V Theſe thinges thou shalt ſpeake
to the children of Iſrael: If anie man of the children of Iſrael, and of
the ſtrangers, that dwel in Iſrael, geue of his ſeede to the idol
Moloch, dying let him die: the people of the land ſhal ſtone him. \V And
I wil ſet my face againſt him: and wil cut him of from the middes of his
%%% o-0291
people, becauſe he hath geuen of his ſeede to Moloch, and hath
contaminated my Sanctuarie, and polluted my holy name. \V And if the
people of the land neglecting, and as it were litle eſteming my
commandement, let alone the man that hath geuen of his ſeede to Moloch,
and wil not kil him: \V I wil ſette my face vpon that man, and his
kinred, and wil cut of both him, and al that conſented with him, to
committe fornication with Moloch, out of the middes of their people. \V
The ſoule, that ſhal decline to Magitians, and ſouth ſayers, and shal
committe fornication with them, I wil ſette my face againſt it, and
deſtroy it out of the middes of his people. \V Sanctifie your ſelues,
and be holie, becauſe I am the Lord your God. \V Keepe my precepts, and
doe them. I the Lord that ſanctifie you. \V He that curſeth his father,
or mother, dying let him die: he hath curſed father, and mother, his
bloud be vpon him. \V If any man commit adulterie with an other mans
wife, and commit
\Fix{aduontrie}{aduoutrie}{obvious typo, fixed in other}
with his neighbours wife, dying let them die, both the adulterer and the
aduoutreſſe. \V He that lieth with his ſtepmother, and reuealeth the
ignominie of his father, dying let both die: their bloud be vpon
them. \V If anie man lie with his daughter in law, let both die, becauſe
they haue done an heinous fact: their bloud be vpon them. \V He that
lieth with man as if he should companie with woman, both haue committed
abomination dying let them die: their bloud be vpon them. \V He that
beſides his wife the daughter,
\SNote{Violating this law in anie degree in the right line, either of
conſanguinitie, or affinitie, or in the firſt collateral degree of
cõſanguinitie, was puniſhed with death; but in the firſt collateral of
affinitie, and in the ſecond collaterſl degree as wel of cõſanguinitie,
as affinitie, with leſſe puniſhment. VVhich ſheweth greatter obligation,
& greatter ſinne, cõcerning the right line, then the collateral; alſo in
the firſt degree of conſanguinitie then in the ſecond; and more in
conſanguinitie then in affinitie.}
marieth her mother, hath done wickednes: he shal burne aliue with them,
neither shal there ſo great abomination remaine in the middes of you. \V
He that shal companie with beaſt and cattel, dying let him die: the
beaſt alſo doe ye kil. \V The woman that shal lie vnder anie beaſt, shal
be killed together with the ſame: their bloud be vpon them. \V He that
taketh his ſiſter the daughter of his father, or the daughter of his
mother, and ſeeth her turpitude, and she beholdeth her brothers
ignominie: they haue committed a ſhameful thing: they shal be
%%% 0323
ſlaine; in the ſight of their people, becauſe they haue reuealed one an
others turpitude, and they shal beare their iniquitie. \V He that
compaineth with a woman in her menſtrual fluxe, and reuealeth her
turpitude, and she openeth the fountaine of her bloud, both shal be
deſtroyed out of the middes of their people. \V The turpitude of
%%% Same as above SNote? Not marked in other.
%%% \SNote{}
thy aunt by thy mother, and of thy aunt by thy father, thou shalt not
diſcouer: he that doeth this, hath diſcloſed the ignominie of his flesh,
both shal beare their iniquitie. \V He that compaineth with the wife of
his vncle by the father, or of his vncle by the mother, and reuealeth
the ignominie of his kinred, both shal beare their ſinne: without
children they shal die. \V He that marieth his brothers wife, doth an
vnlawful thing, he hath reuealed his brothers turpitude: they shal be
without children. \V Keepe my lawes, and iudgementes, and doe them: leſt
the land which you shal enter into and inhabite, vomite out you alſo. \V
Walke not in the ordinances of the nations, which I wil expel before
you. For
\SNote{Not for euerie kind of ſinne (though euerie one is puniſhable)
but for the more hainous, & for al together the Chananites were caſt out
of their land.}
al theſe thinges haue they done, and I haue abhorred them. \V But to you
I ſpeake:
%%% o-0292
Poſſeſſe their land, which I wil geue you for an inheritance, a land
flowing with milke and honie. I the Lord you God, that haue ſeperated
you from other peoples. \V Therfore doe you alſo ſeperate the cleane
beaſt, from the vncleane, and the cleane foule from the vncleane:
pollute not your ſoules, in beaſtes, and birdes, and al thinges that
moue on the earth, and which I haue shewed vnto you, to be polluted. \V
You shal be holie vnto me, becauſe I the Lord am holie, and I haue
ſeparated you from other peoples, that you should be mine. \V Man, or
woman, in whom is a pithonical or diuining ſpirite, dying let them die,
they shal ſtone them: their bloud be vpon them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
