%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 0286
%%% o-0259
\startChapter[
  title={Chapter 3}
  ]

\Summary{How the pacifique hoſtes muſt be offered of beeues, 6.~shepe,
  7.~lambes, 12.~and goates.}

And if his oblation be an hoſte of pacifiques, and he wil offer of
beues, male or
\SNote{In holocauſt onlie the male was offered,
\XRef{chap.~1.}
in other ſacrifices both ſexes were acceptable.}
female, without ſpot shal he offer them before our Lord. \V And he shal
lay his hand vpon the head of his victime, which shal be immolated in
the entrie of the tabernacle of teſtimonie, and the ſonnes of Aaron the
prieſt shal poure the bloud in the circuite of the altar. \V And they
shal offer of the hoſt of pacifiques, for an oblation to our Lord, the
fatt that couereth the entralles, and whatſoeuer fatt is within: \V the
two kidneys with the fatt wherwith the guttes are couered, and the caule
of the liuer with the two little kydneys. \V And they shal burne them
vpon the altar, for an
\SNote{Theſe partes and the fatte were burned as an Holocauſt, the reſt
of the oblation being pacifique.}
holocauſt, putting fire vnder the wood: for an oblation of moſt ſwete
ſauour to our Lord. \V But if his oblation, and the hoſte of pacifiques
be of flocks, whether he offer male, or female, they shal be without
ſpot. \V If he offer a lambe before our Lord, \V he shal put his hand
vpon the head of his victime; which shal be immolated in the entrie of
the tabernacle of teſtimonie: and the ſonnes of Aaron shal powre the
bloud therof in the circuite of the altar. \V And they shal offer of the
hoſte of pacifiques, a ſacrifice to our Lord: the fatt and the whole
rumpe, \V with the kidneys, and the fatt that couereth the bellie and al
the vital partes, and both little kydneys, with the fatt that is about
the guttes, and the caule of the liuer with the two little kidneys. \V
And the prieſt shal burne them vpon the altar, to the foode of the fire,
and of the oblation of our Lord. \V If his oblation be a goate, and he
offer it to our Lord, \V he shal put his hand vpon the head therof: and
shal immolate it in the entrie of the tabernacle of teſtimonie. And the
ſonnes of Aaron shal poure the bloude therof in the circuite of the
altar. \V And they shal take of it to the foode of our Lords fire, the
fatt that couereth the bellie, and that couereth al the vital partes: \V
the two little kydneys with the caule, that is vpon them about the
guttes, and the tallowe of the liuer with the little kidneys: \V and
the prieſt shal
%%% 0287
burne them vpon the altar, to the food of the fire, and of a moſt ſweete
ſauour. And the fatt shal be our Lordes \V by a perpetual right in your
generations, and in al your habitations: you shal eate no bloude nor fat
at al.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
