%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 0304
%%% o-0276
\startChapter[
  title={Chapter 12}
  ]

\Summary{The law of wemens purification that beare children.}

And our Lord ſpake to Moyſes, ſaying: \V Speake to the children of
Iſrael, and thou ſhalt ſay to them: A woman, if
\SNote{The moſt pure virgin mother, was not ſubiect to this law. For she
conceiued not by the ſeede of man, yet obſerued the cuſtome of other
wemen
\XRef{Luc.~2.}
as Chriſt alſo would be baptized by S.~Iohn Baptiſt
\XRef{Mat.~3.}
to geue example of humilitie.
\Cite{S.~Cyril. li.~8. in Leuit.}
\Cite{S.~Bernard. Ser.~3. de Purific.}}
receauing ſeede ſhe beare a man child, shal be vncleane ſeuen daies,
according to the daies of the ſeparation of her floures. \V And the
eight day the litle infant shal be circumciſed: \V but her ſelf shal
remaine three & thirtie
%%% 0305
daies in the bloud of her purification: No holie thing shal she touch,
neither shal she enter into the Sanctuarie, vntil the daies of her
purification be expired. \V But if she beare a woman child, she shal be
vncleane two weekes, according to the cuſtome of the fluxe of her
floweres, and ſixtie ſix daies ſhe shal remaine in the bloud of her
purification. \V And when the daies of her purification are expired, for
a ſonne, or for a daughter, she shal bring a lambe of a yeare old for an
holocauſt, and a young pigeon or a turtle for ſinne, to the dore of the
tabernacle of teſtimonie, and shal deliuer them to the prieſt, \V who
shal offer them before our Lord, and shal pray for her, and ſo she shal
be cleanſed from the fluxe of her bloud. This is the lawe for her that
beareth a man child or a woman child. \V And if her hand finde not,
neither is able to offer a lambe, she shal take two turtles, or two
young pigeons, one for an holocauſt, and an other for ſinne, and the
prieſt shal pray for her, and ſo she shal be cleanſed.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
