%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 0282
%%% o-0255
\startChapter[
  title={Chapter 1}
  ]

\Summary{Diuers rites in
\Fix{oering}{offering}{obvious typo, fixed in other}
holocauſtes, as wel of cattle, 14.~as of birdes.}

And
\MNote{The firſt part of this booke.

Of diuers ſortes of Sacrifices.}
our Lord called Moyſes, and ſpake to him out of the tabernacle of
teſtimony, ſaying: \V Speake to the children of Iſrael; & thou shalt ſay to
them:
\LNote{The man that shal offer.}{Sacrifice
\MNote{Sacrifice preſuppoſed to be neceſſary, God preſcribeth the rites
to be obſerued therin.}
being the moſt ſpecial external ſeruice, wherby man acknowledgeth the
ſupreme dominion of God, and his owne ſubiection and homage to his
diuine Maieſtie, was ſo wel knowen to be neceſſarie (as being in moſt
frequent vſe in the law of nature, and in al nations) that here neded
not anie new precept in general, that the people of God should offer
ſacrifice, though for ſpecial purpoſes, certaine particular ſacrifices
were ſome times appointed, but this dutie & obligation preſuppoſed, our
Lord firſt admoniſhing to offer the beſt, and perfecteſt things in
euerie kind, preſcribeth with what rites, and ceremonies it shal be
done. As in offering an holocauſt of the heard, it muſt be \Emph{a male
vvithout ſpotte}; and be offered \Emph{at the dore of the tabernacle},
the offerer \Emph{putting his handes vpon the head of the hoſte; the
prieſts muſt offer the bloud, povvring it in the circuite of the altar;
plucke of the skinne; cut the ioyntes in peeces; lay them in order; the
entrales and feete being vvashed, burne al vpon the Altar}: And the like
in other ſacrifices, al for iuſt and reaſonable cauſes, without which
the wiſdome of God doth nothing.
\XRef{Sap.~7.}
&
\XRef{Pſal.~103. v.~24.}}
The man of you, that shal offer an hoſte to our Lord, of beaſtes, that
is of oxen & sheepe, offering victimes \V if his oblation be
\LNote{An holocauſt.}{In
\MNote{Three kindes of Sacrifice.}
reſpect of diuers things offered, the diuers maner, and cauſes of
offering, there were manie ſortes of Sacrifices: but al are reduced to
three kindes.
\MNote{Holocauſte.}
The firſt was Holocauſt, in which al was burned in the honour of God,
and reſolued into vapour, which aſcendeth vpwards in ſigne that al we
haue is of God.
\MNote{Sacrifice for ſinne.}
The ſecond was Sacrifice for ſinne, & that of diuers ſortes, for the
varietie of ſinnes and perſons; and part of this ſacrifice was burned,
the other part belonged to the Prieſts.
\MNote{Pacifique Sacrifice.}
The third was Pacifique ſacrifice, wherof one part was burnt, an other
pertayned to the Prieſts, and an other to them, that gaue the
oblation.
\MNote{For benefites receiued: or deſired.}
And of this kinde there were two ſortes, one of thankſgeuing
for benefites receiued: the other to procure fauoure in anie good
enterpriſe, or deſire.
\MNote{Al ſacrifices of the old Teſtament prefigured Chriſts Sacrifice
on the Croſſe, and in the Euchariſt.}
Al the which did prefigure and forshew one only
Sacrifice of Chriſts bodie and bloud, offered by him in two maners;
bloudie on the Croſſe once for euer, wherof S.~Paul expreſſly ſpeaketh
\XRef{(Heb.~9.)}
vnbloudie in formes of bread and wine, wherof the ſame S.~Paul ſpeaketh
\XRef{(Heb.~13. v.~10.)}
shewing that Chriſtians \Emph{haue an Altar}, and conſequently a
Sacrifice farre excelling thoſe of the Tabernacle; and our Sauiour him
ſelfe
\XRef{(Math.~26. v.~25.)}
ſpeaking of the contents in the chalice, ſaid: it was \Emph{his bloud of
the nevv Teſtament}, which he then inſtituted and dedicated, as is there
noted. And the ancient Fathers (by Caluins confeſſion
\Cite{in Heb.~9.})
generally vſe this diſtinction of the ſame Sacrifice offered in bloudie,
& in vnbloudie maner. They likewiſe teach that al lawful Sacrifices of
the Law of nature, and of Moyſes did end, and were complete in this one,
which is our daylie Sacrifice, our immaculate lambe, our manna, our
libament, our holocauſt, our Sacrifice for ſinne, our Pacifique
Sacrifice for al purpoſes, and in ſteede of al old Sacrifices. So
\Cite{S.~Auguſtin lib.~8. c.~27.}
\Cite{lib.~17. c.~20. de ciuit.}
\Cite{lib.~3. de Baptiſ. c.~19.}
\Cite{lib.~1. cont. aduerſ. leg & prophet. c.~18. & ~20.}
\Cite{S.~Chryſoſt. in Pſal.~95.}
\Cite{S.~Leo. ſer.~8. de Paſſ.}
and other fathers teach.}
an holocauſte, and of the heard; he shal offer
\SNote{The beſt and perfecteſt of euerie kind is to be offered to God,
not the blind, lame, or weake.
\XRef{Gen.~4.}
\XRef{Malach.~1.}}
a male, without ſpotte, at the dore of the tabernacle of teſtimonie, to
propitiate our Lord vnto him: \V and he shal put his handes vpon the
heade of the hoſte, and it shal be acceptable, and profitable to his
expiation. \V And he shal immolate the calfe before our Lord, and the
children of Aaron the prieſtes shal offer the bloud therof, powring it
in the circuite of the altar, which is before the dore of the
tabernacle. \V And the skinne of the hoſte being plucked of, the ioyntes
they shal cut into peeces, \V and shal put fire vnderneth in the altar,
hauing before laid a pyle of wood in order: \V and the ioyntes that are
cut out, laying in order thereupon, to wit, the head, & al thinges that
cleaue to the liuer, \V the entralles and feete being washed with water,
and the prieſt shal burne them vpon the altar for an holocauſte, and
\LNote{Svvete ſauour.}{Not
\MNote{Not the external worke for it ſelfe but deuotion pleaſeth God.}
that the ſauour of corporal things (though it were ſweter then of burnt
fleſh and bones) delighteth Gods moſt pure ſubſtance: but for that mans
frailtie in ſome good ſorte performing his dutie, is very acceptable to
his diuine goodnes. For otherwiſe he required not theſe Sacrifices, nor
other external Rites for him ſelf, but
\MNote{External ſacrifices were ordained:}
he would haue his people for
their owne good to be exerciſed therein: eſpecially for three cauſes.
\MNote{1.~to keepe the people from Idolatrie.}
Firſt to kepe them from Idolatrie, wherto they were very prone, as
appeareth by their often falling, notwithſtanding continual admonitions
to the contrarie. For being as it were burdened with manie ceremonies,
pertaining to Gods true ſeruice, they might haue leſſe mind, leyſure,
and occaſion to ſerue Idols.
\MNote{2.~To induce them to internal vertues.}
Secondly, for ſo much as man conſiſteth of ſoule and bodie, as the ſoule
muſt interiorly
\CNote{\XRef{Io.~4. v.~24.}}
\Emph{vvorship God in ſpirite and veritie}; ſo the bodie muſt alſo
honour him exteriorly,
\CNote{\XRef{Rom.~6. v.~19.}}
\Emph{ſeruing iuſtice vnto ſanctification}: that is by external good
workes to increaſe iuſtice, and ſanctitie, when by them the mind is
inſtructed and inuited to know and honoure God. For otherwiſe ſaieth
S.~Dionyſe
\Cite{(c.~1. cæleſt. Hierer.)}
vnles mans vnderſtanding vſe the helpe of corporal things, diuine
veritie can not be attained. And S.~Auguſtin
\Cite{(lib.~10. c.~5. ciuit.)}
teacheth that God commanded external Sacrifices, thereby to lead his
ſeruants vnto mortified ſpirites, contrite and humbled harts,
\CNote{\XRef{Pſal.~50.}}
to mercie and compaſſion towards others. In briefe
\Cite{(c.~3. Enchir.)}
to the true and perfect ſeruing of his Diuine powre by faith, hope, and
charitie.
\MNote{3.~To ſignifie greater Myſteries of the new Teſtament.}
Thirdly, that theſe external Sacrifices and Rites might prefigure and
ſignifie greater, more excellent, and more effectual Myſteries of the
new Teſtament. For as S.~Paul ſpeaketh
\XRef{(Heb.~10.)}
\Emph{the lavv} (of Moyſes) \Emph{hauing a shadovv of good things to
come, not the verie image of the things}, brought not to
perfection: \Emph{nor tooke avvay ſinnes by the bloud of oxen or
goates}, but being (as is ſaid) \Emph{a shadovv}, rather ſhaded then
perfectly ſhewed the great benefites, which the new law as a perfect
image liuely repreſenteth: eſpecially Chriſts paſſion, which is the
verie fountaine of grace and mercie. And wheras the old law could not
iuſtifie
\XRef{(Gal.~3.)}
the law of Chriſt doth in dede iuſtifie, as the Goſpel witneſſeth,
ſaying:
\XRef{(Ioan.~1. v.~17.)}
\Emph{the lavv vvas geuen by Moyſes, grace and veritie vvas made by
Ieſus Chriſt.}}
ſweete ſauoure to our Lord. \V And if the oblation be of flockes, an
holocauſte of sheepe or of goates, a lambe of a yeare old without ſpot
shal he offer: \V and he shal immolate it at the ſide of the altar that
looketh
%%% 0283
to the North, before our Lord: but the bloud therof the
%%% o-0256
ſonnes of Aaron shal poure vpon the altar round about: \V and they shal
diuide the ioyntes, the head, and al that cleaue to the lyuer: and shal
lay them vpon the wood, vnder which the fire is to be put: \V but the
entrales and the feete they shal wash with water. And the whole the
prieſt shal offer, and burne vpon the altar for an holocauſt, and moſt
ſweete ſauoure to our Lord. \V But if the oblation of holocauſt to our
Lord be of birdes, of turtles and young pigions, \V the prieſt shal
offer it at the altar: and writhing the head to the necke, and breaking
the place of the wound, he shal make the bloud to runne downe vpon the
brimme of the altar: \V but the croppe of the throate, and the fethers
he shal caſt nigh to the altar at the eaſt ſide, in the place where the
ashes are wount to be powred out, \V and he shal breake the pinnions
therof, and shal not cut, nor diuide it with a knife, and shal burne it
vpon the altar, putting fire vnder the wood. It is an holocauſt and
oblation of moſt ſweete ſauoure to our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
