%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 0308
%%% o-0279
\startChapter[
  title={Chapter 14}
  ]

\Summary{Sacrifices for clenſing leproſie in men, 33.~the maner of
  viewing whether leproſie be in a houſe or no, 40.~and of clenſing it.}

And our Lord ſpake to Moyſes, ſaying: \V This is the rite of a lepre,
when he is to be cleanſed. \V He shal be brought to the prieſt: who
going out of the campe, when he shal finde that the leproſie is
cleanſed, \V he shal command him, that is purified, that he
\SNote{This gift & ſacrifice were commanded (ſaith S.~Auguſtin) becauſe
the Sacrifice of Chriſts bodie was not yet ordained, which now ſerueth
for al other ſacrifices.
\Cite{li.~1. ca.~19. et~20. cont. aduerſar. leg. et prophet.}}
offer for him ſelfe two liue ſparowes, which it is lawful to eate, and
cedar wood, and ſcarlet & hyſſope, \V and he shal command that one of
the ſparowes be immolated in an earthen veſſel ouer
\TNote{Of vvel or riuer, not of a ceſterne poole, or marrish.}
liuing water: \V but the other being aliue with the cedar wood, and
ſcarlet and the hyſſope he shal dippe in the bloud of the ſparowe that
is immolated, \V wherwith he shal ſprinckle him, that is to be cleanſed,
ſeuen times, that he may be rightly purged: and he shal let goe the liue
ſparow, that it flie into the field. \V And when the man hath washed his
clothes, he shal ſhaue al the heare of his bodie, and shal be washed with
water: and being purified he shal enter into the campe, yet ſo for al
that, that he tarie without his owne tent ſeuen daies, \V and the
ſeuenth day he shal shaue the heare of his head, and
%%% 0309
his beard and eye browes, and the heare of his whole bodie. And hauing
washed againe his clothes, and his bodie, \V the eight day he ſhal take
two lambes without ſpotte, and an ewe of a yeare old without ſpotte, and
three tenths of floure tempered with oyle for a ſacrifice, and a ſextarie
of oyle apart. \V And when the prieſt that purifieth the man, hath ſet
him, and al theſe things before the Lord in the dore of the tabernacle
of teſtimonie, \V he ſhal take a lambe, and offer it for offence, & the
ſextarie of oyle, and hauing offered al before the Lord, \V he shal
immolate the lambe, where the hoſte for ſinne is wont to be immolated,
and the holocauſt, that is, in a holie place. For as for ſinne, ſo alſo
the hoſte for offence pertaineth to the prieſt: it is Holie of
holies. \V And the prieſt taking of the bloud of the hoſte, that was
immolated for offence, ſhal put it vpon the tippe of the right eare of
him that is cleanſed, and vpon the thumbes of his right hand and foote:
%%% o-0280
\V and of the ſextarie of oile he shal poure into his owne left hand, \V
and shal dippe his right finger in it and shal ſprinckle before the Lord
ſeuen times. \V And the reſidue of the oile in his left hand, he shal
poure vpon the tippe of the right eare of him that is cleanſed, and vpon
the thumbes of his right hand and foote, and vpon the bloud that was
ſhed for offence, \V and vpon his head. \V And he ſhal pray for him
before the Lord, and shal make the ſacrifice for ſinne. Then shal he
immolate the holocauſt, \V and put it on the altar with the libaments
therof, and the man shal orderly be cleanſed. \V But if he be poore, and
his hand can not finde the thinges aforeſaid, for offence he shal take a
lambe for an oblation, that the prieſt may pray for him, and a tenth
part of floure tempered with oile for a ſacrifice, and of oile a
ſextarie, \V and two turtles or two young pigeons, of the which let one
be for ſinne, and the other for an holocauſt: \V and he shal offer them
the eight day of his purification to the prieſt, at the dore of the
tabernacle of teſtimonie before the Lord: \V who receiuing the lambe for
offence, and the ſextarie of oile, shal eleuate them together: \V and
the lambe being immolated, of the bloud therof he shal put vpon the tippe
of the right eare of him that is cleanſed, and vpon the thumbes of his
right hand and foote: \V but part of the oile he shal poure into his
owne left hand, \V wherin dipping the finger of his right hand, he shal
%%% 0310
ſprinckle it ſeuen times before the Lord: \V and he shal touch the tippe
of the right eare of him that is cleanſed, and the thumbes of his right
hand & foote, in the place of the bloud that was shed for offence: \V
and the other part of the oyle, that is in his left hand, he shal poure
vpon the head of the purified perſon, that he may propitiate the Lord
for him: \V and a turtle, or yong pigeon he shal offer, \V one for
offence, and the other for an holocauſt, with their libaments. \V This
is the ſacrifice of a lepre, that is not able to haue al thinges for the
cleanſing of him ſelf. \V And our Lord ſpake to Moyſes and Aaron,
ſaying: \V When you shal be entred into the Land of Chanaan, which I wil
geue you in poſſeſſion, if there be the plague of leproſie in a houſe, \V
he whoſe houſe it is, shal goe & tel the prieſt, ſaying: It ſemeth to
me, that there is as it were the plague of leproſie in my houſe. \V But
he shal command, that they carie forth al thinges out of the houſe,
before he goe into it, and ſee whether it haue the leproſie, leſt al
thinges become vncleane that are in the houſe. And afterward he shal goe
in to conſider the leproſie of the houſe: \V & if he ſee in the walles
therof as it were litle dintes, diſfigured with paleneſſe or redneſſe,
and lower then al the reſt, \V he shal goe out of the dore of the houſe,
and forthwith shut it ſeuen daies. \V And returning the ſeuenth day, he
shal conſider it. If he finde that the leproſie is growne, \V he shal
command, the ſtones wherein the leproſie is, to be plucked out, and to
be throwen without the citie into an vncleane place: \V & the houſe it
ſelf to be ſcraped on the inſide round about, and the duſt of the
ſcraping to be diſperſed without the citie into
%%% o-0281
an vncleane place, \V and other ſtones to be laid in their places, that
were taken away, and the houſe to be playſtered with other morter. \V
But if after the ſtones be plucked out, & the duſt ſcraped of, and it
playſtered with other earth, \V the prieſt going in perceiue that the
leproſie is returned, and the walles ful of ſpottes, it is a laſting
leproſie, and the houſe vncleane: \V the which forthwith they shal
deſtroy, and the ſtones and timber therof, and al the duſt they shal
caſt without the towne into an vncleane place. \V He that entreth into
the houſe when it is shut, shal be vncleane vntil euen: \V and he that
ſleepeth in it, and eateth any thing, shal wash his clothes. \V But if
the prieſt going in perceiue that the leproſie is not growne farder in
the houſe, after it was
%%% 0311
playſtred againe, he shal purifie it being made whole againe: \V and for
the purification therof he shal take two ſparowes, and cedrewood, and
ſcarlet and hyſſope: \V and when one ſparow is immolated in an earthen
veſſel ouer liuing waters, \V he shal take the ceder wood, and the
hyſſope, and the ſcarlet, and the liue ſparowe, and shal dippe al in the
bloud of the ſparowe that is immolated, and in the liuing water, and he
ſhal ſprinckle the houſe ſeuen times, \V and shal purifie it as wel with
the bloud of the ſparow, as with the liuing water and with the liue
ſparow, and with the cedrewoode and the hyſſope and the ſcarlet. \V And
when he hath let goe the ſparow to flie freely away into the field, he
shal pray for the houſe, and it shal be rightly cleanſed. \V This is the
law of al leproſie and percuſſion, \V of the leproſie of garments and
houſes, \V of a ſcarre and of bliſters breaking forth, of a shining
ſpotte, and when the colours are changed into diuers kindes, \V that it
may be knowen at what time anie thing is cleane, or vncleane.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
