%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 0315
%%% o-0285
\startChapter[
  title={Chapter 17}
  ]

\Summary{Al Sacrifices muſt be offered at the doore of the Tabernacle,
  7.~with ſpecial prohibitiõ of Idolatrie. 10.~None muſt eate
  bloud. 15.~Whoſoeuer eateth caraine flesh is contaminate, and muſt be
  washed.}

And our Lord ſpake to Moyſes, ſaying: \V Speake to Aaron and his ſonnes,
and to al the children of Iſrael, ſaying to them: This is the word,
which our Lord hath cõmanded, ſaying: \V Anie man whoſoeuer of the houſe
of Iſrael, if he
\SNote{If anie killed for ſacrifice he muſt offer it at the dore of the
tabernacle, that a prieſt might offer it on the Altar, for no other man,
nor place was allowed, without ſpecial diſpenſation of God. And ſo
Samuel offered ſacrifice in an other place.
\XRef{1.~Reg.~7.}
Elias
\XRef{3.~Reg.~18.}
and Dauid alſo being no Prieſt.
\XRef{2.~Reg.~24.}
ſo S.~Aug explicateth, this text.
\Cite{q.~56. in Leuit.}}
kil an oxe or a sheepe, or a goate in the campe or without the campe,
and offer it not at the dore of the tabernacle an oblation to the Lord,
shal be guiltie of bloud: as if he had shed bloud, ſo shal he perish out
of the middes of his people. \V Therfore shal the children of Iſrael
bring to the prieſt their hoſtes, which they kil in the filde, that they
%%% 0316
may be ſanctified to our Lord before the dore of the tabernacle of
teſtimonie, & they may immolate them pacifique hoſtes to our Lord. \V
And the prieſt shal poure the bloud vpon the altar of our Lord, at the
dore of the tabernacle of teſtimonie, and shal burne the fatte for a
ſwete odour to our Lord: \V and they shal no more immolate their hoſtes
to diuels, with whom they haue committed fornication. It shal be an
ordinance for euer to them and to their poſteritie. \V And to them thou
shalt ſay: The man of the houſe of Iſrael, and of the ſtrangers, which
ſeiourne with you, that offereth an holocauſt or victime, \V and
bringeth it not to the dore of the tabernacle of teſtimonie, that it may
be offered to our Lord, ſhal perish out of his people. \V Anie man
whoſoeuer of the houſe of Iſrael, and of the ſtrangers, that ſeiourne
among them, if he eate bloud, I wil ſette my face againſt his life, and
wil deſtroy it out of his people, \V becauſe the life of the fleſh is in
the bloud: and I haue geuen it to you, that vpon the altar you may make
expiation with it for your ſoules, and the bloud may be for an expiation
of the ſoule. \V Therfore haue I ſaid to the children of Iſrael: No
ſoule of you shal eate bloud, nor of the ſtrangers, that ſeiourne with
you. \V Anie man whoſoeuer of the children of Iſrael, and of the
ſtrangers that ſeiourne with you, if by hunting or fowling, he take
wild beaſt or foule, which it is lawful to eate, let him poure out the
bloud therof, and couer it with earth. \V For the life of al flesh is in
the bloud: wherupon I ſaid to the children of Iſrael: The bloud of no
flesh ſhal you eate, becauſe the life of the flesh is in the bloud: and
whoſoeuer eateth it, ſhal die. \V The ſoule that eateth carraine, or
that which is taken of a beaſt, aſwel of them of the ſame countrie as of
ſtrangers, ſhal wash his clothes, and him ſelf with water, and
%%% o-0286
ſhal be contaminated vntil euen: and in this order he ſhal be made
cleane. \V And if he doe not wash his clothes, and his bodie, he ſhal
beare his iniquitie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
