%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 27
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-27


\project douay-rheims


%%% 0337
%%% o-0303
\startChapter[
  title={Chapter 27}
  ]

\Summary{How ſome vowes of diuers perſons may be redemed, but ſome may
  not be changed. 28.~Al, one way or other, muſt be
  diſcharged. 30.~Tithes alſo muſt be payed, either the ſame that are
  due or more.}

And
\MNote{The fifth part.

Of vowes and Tithes.}
our Lord ſpake to Moyſes, ſaying: \V Speake to the children of Iſrael,
and thou ſhalt ſay to them: The man that ſhal haue made a vow, and
\SNote{Becauſe no other but the tribe of Leui could ſerue about the
tabernacle, and yet others might deſire to ſerue there, they might in
ſteed therof geue a price & haue the reward of their good wil.}
bound his ſoule to God, by eſtimation he ſhal geue the price. \V If it
be a man
%%% o-0304
from the twenteth yeare vntil three ſcore, he shal geue fiftie ſicles of
ſiluer, after the meaſure of the Sanctuarie: \V if a woman, thirtie. \V
But from the fift yeare vntil the twenteth, a man ſhal geue twentie
ſicles: a woman tenne. \V From one moneth vntil the fift yeare, for a
man ſhal be geuen fiue ſicles: for a woman three. \V One that is three
ſcore and aboue a man shal geue fiftene ſicles: a woman tenne. \V If he
be poore, and not able to pay the eſtimation he ſhal ſtand before the
prieſt: and as much as he ſhal eſteme, and ſee him able to pay, ſo much
ſhal he geue. \V But the beaſt, that may be immolated to the Lord, if a
man doe vow it, shal be holie, \V and
%%% 0338
can not be changed, that is to ſay,
\SNote{The thing that is vowed, if it may be performed, pleaſeth God
better, then a change.}
neither a better for a bad, nor a worſe for a good. And if he change it:
both it ſelf that was changed, and that for the which it was changed,
ſhal be conſecrated to the Lord. \V The vncleane beaſt, which can not be
immolated to the Lord, if anie man vow it shal be brought before the
prieſt. \V Who iudgeing whether it be god or euel, shal ſet the
price. \V Which if he that offereth wil geue, he shal adde aboue the
eſtimation the fift part. \V If a man vow his houſe, and ſanctifie it to
the Lord, the prieſt shal conſider it, whether it be good or bad, and
according to the price, which he shal appoint, it shal be ſold. \V But
if he that had vowed, wil redeme it, he shal geue the fift part of the
eſtimation beſides, and shal haue the houſe. \V And if he vow the field
of his poſſeſſion, and conſecrate it to the Lord, the price shal be
eſtemed according to the meaſure of the ſeede. If the ground be ſowed
with thirtie bushels of barly, let it be ſolde for fiftie ſicles of
ſiluer. \V If he vow he field immediatly from the yeare of Iubilee, that
is beginning, how much it can be worth, at ſo much it shal be
eſtemed: \V but if ſometime after: the prieſt shal account the money
according to the number of yeares, that remaine vntil the Iubilee, and
there shal be diminished of the price. \V And if he that had vowed, wil
redeme his field, he shal adde the fift part of the eſtemed money, and
shal poſſeſſe it. \V And if he wil not redeme it, but it be ſould to any
other man, he that had vowed it, can redeme it no more: \V for when the
day of Iubilee commeth, it ſhal be ſanctified to the Lord, and the
poſſeſſion conſecrated pertaineth to the right of the prieſtes. \V If
the field be bought, and being not of his
\Fix{auncetors}{aunceſtors}{likely typo, fixed in other}
poſſeſſion be ſanctified to the Lord, \V the prieſt ſhal account the
price according to the number of yeares, vnto the iubilee: and he that
had vowed, ſhal geue that to the Lord. \V But in the Iubilee, it shal
returne to the former owner, that ſould it, and had it in the lotte of
his poſſeſſion. \V Al eſtimation shal be weighed by the
\SNote{A ſicle was about 15.~d. obolus three farthings.}
ſicle of the ſanctuarie. A ſicle hath twentie oboles. \V The
\LNote{The firſt borne.}{God
\MNote{Vowes are properly of things not commanded.}
here forbiddeth to vow the firſt borne, and geueth the reaſon, for that
\Emph{they are the Lordes}, ſhewing that thoſe things, wherto we are
already bound, are not properly matter of vow. But a vowe properly is a
religious promiſe voluntarily made to God, of a good thing, vnto which
we were not bound.
\MNote{And are gratful to God.}
And that the ſame is verie gratful to God, appeareth not only in this
chapter, and in
\CNote{\XRef{Num.~6.~30.}
\XRef{Deut.~23.}}
manie other places of Moyſes law, but alſo in the law of nature.
\XRef{Gen.~28.}
Iacob vowed, and God accepted therof.
\XRef{Gen.~31. v.~13.}
And the royal Prophet in
\CNote{\XRef{Pſal.~21.}
\XRef{49.}
\XRef{75.}
\XRef{115.}
\XRef{131.}}
diuers Pſalmes pertaining to the new Teſtament commendeth vowes.
\MNote{Alſo in the new Teſtament.}
It is certaine alſo & manifeſt
\XRef{1.~Tim.~5.}
that widowes did lawfully vowe chaſtitie in the primitiue Church: and
ſuch as did afterward breake the ſame, did violate their promiſe to
God. Innumerable alſo moſt lerned and moſt godlie fathers, haue
euer from Chriſts time both taught and practiced religious vowes, of
obedience to ſuperiors, who otherwiſe had no authoritie ouer them, and
of perpetual chaſtitie, and voluntarie pouertie. It is likewiſe, and
continually hath benne, a moſt common practiſe in the Church, to vow
other good workes of pietie, as to viſite holie places, to build
Churches, Collegies, Hoſpitals, and the like, being no way bound therto
but of mere deuotion. See
\XRef{Annotations. 1.~Tim.~5.}}
firſt borne, which pertaine to the Lord, no man may ſanctifie and vow:
whether it be oxe, or sheepe, they are the Lordes. \V And if it be an
vncleane beaſt, he that offereth it shal redeme it, according to thy
eſtimation, and shal adde the fift part of the price. If he wil not
redeme it, it shal be ſould to an other
%%% 0339
for how much ſoeuer it was eſtemed by
%%% o-0305
thee. \V Anie thing that is
\SNote{A vow made approued and conſecrated to God, can not be changed by
anie man.
\Cite{Theodoret. q.~vlt. in Leuit.}}
conſecrated to the Lord, whether it be man, or beaſt, or field, shal not
be ſould, 
neither can it be redemed. Whatſoeuer is once conſecrated, shal be holie
of holies to the Lord. \V And any conſecration, that is offered of a
man, shal not be redemed, but dying shal die. \V Al tithes of the land,
whether of corne, or of the fruites of trees, are the Lordes, and are
ſanctified to him. \V And if anie man wil redeme his tithes, he shal
adde the fift part of them. \V Of al the tithes of oxen, and sheepe and
goates, that paſſe vnder the sheepheardes rodde, euerie tenth that
commeth shal be ſanctified to the Lord. \V It shal not be choſen neither
good nor bad, neither shal it be changed for an other. If anie man
change it: both that which was changed, and that for which it was
changed, shal be ſanctified to the Lord, and shal not be redeemed. \V
Theſe are the precepts, which our Lord commanded Moyſes vnto the
children of Iſrael in the mount Sinai.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
