%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 0319
%%% o-0289
\startChapter[
  title={Chapter 19}
  ]

\Summary{Diuers Moral, Ceremonial, and Iudicial precepts are briefly
  recited.}

%%% 0320
Ovr Lord ſpake to Moyſes, ſaying: \V Speake to al the aſſemblie of the
children of Iſrael, and thou shalt ſay to them: Be ye holie, becauſe I
the Lord your God am holie. \V Let euerie one feare his father, and
mother. Keepe my Sabbathes. I the Lord your God. \V Turne not your
ſelues to idols, neither make you to your ſelues molten goddes. I the
Lord your God. \V If ye immolate an hoſte of pacifiques to the Lord,
that it may be placable, \V that day wherin it is immolated, shal you
eate it, and the next day: and whatſoeuer ſhal be left vntil the third
day, you shal burne with fire. \V If after two daies anie man eate
therof, he shal be profane, and guiltie of impietie: \V and shal beare
his iniquitie, becauſe he hath polluted the holie of the Lord, and that
ſoule shal perish out of his people. \V And when thou reapeſt the corne
of thy lande, thou shalt not sheare the face of the earth to the verie
ground: neither shalt thou gather the eares that remaine. \V Neither in
thy vineyard shalt thou gather the cluſters, and grapes that fal downe,
but shalt leaue them to the poore, and the ſtrangers to take. I the Lord
your God. \V You shal not committe theft. You shal not lie, neither shal
anie man deceaue his neighbour. \V Thou shalt not forſweare thy ſelf in
my name, nor pollute the name of thy God. I the Lord. \V Thou shalt not
calumniate thy neighbour, nor oppreſſe him by violence. The worke of thy
hireling shal not abide with thee vntil morning. \V Thou shalt not
ſpeake euil of the deafe man, nor put a ſtumbling blocke before the
blinde: but thou shalt feare the Lord thy God, becauſe I am the Lord. \V
Thou shalt not doe that which is vniuſt, not iudge vniuſtly. Conſider
not the perſon of the poore man, neither honour thou the countenance of
him that is mightie. Iudge iuſtly to thy neighbour. \V Thou shalt not be
a criminatour, nor a whiſperer among the people. Thou shalt not ſtand
againſt the bloud of thy neighbour. I the Lord. \V Thou shalt not hate
thy brother in thy hart, but controwle him openly, leſt thou incurre
ſinne through him. \V Seeke not reuenge, nor be mindeful of the iniurie
of thy citizens. Thou shalt loue thy freind as thy ſelf. I the Lord. \V
Keepe ye my lawes. Thou shalt not make thy cattel to gender with the
beaſts of an other kinde. Thy field thou shalt not ſowe with diuers
ſeede. A garment,
\SNote{Theſe diuerſities are not prohibited for them ſelues, for the
ornaments of the tabernacle and of Prieſts were made of diuers things;
but ſchiſme, and al participation with heretikes, and other infidels is
forbid.
\XRef{2.~Cor.~6.}
\Emph{Dravv not in yoke vvith infidels.}
\Cite{Theodoret. q.~27. in Leuit.}}
that is wouen of two ſortes, thou shalt not put on. \V If a man lie with
a woman by carnal
%%% 0321
copulation, that is a bondwoman alſo mariable, and yet not redemed with
a price, nor made
%%% o-0290
free, both shal be beaten, and they shal not die, becauſe she was not
free. \V And for his offence he shal offer to the Lord, at the dore of
the tabernacle of teſtimonie a ramme: \V and the prieſt shal pray for
him, and for his ſinne before the Lord, and he shal become propitious to
him againe, and the ſinne shal be forgeuen. \V When you shal be entred
into the land, and haue planted in it fruite trees, you shal take away
the prepuces of them: the fruite, that come forth shal be vncleane to
you, neither shal you eate of them. \V But in the fourth yeare, al their
fruite shal be ſanctified, laudable to the Lord. \V And the fifth yeare
you shal eate the fruites, gathering the offspring, that they bring
forth. I the Lord your God. \V You shal not eate with bloud. You shal
not diuine,
\SNote{See
\XRef{Gen.~40. v.~8.}}
nor obſerue dreames. \V Neither shal you cut your heare roundwiſe: nor
shaue your beard. \V And for the dead you shal not cut your flesh,
neither shal you make in your ſelues any figures or markes, I the
Lord. \V Make not thy daughter a common ſtrumpet, leſt the land be
contaminated, and filled with wickedneſſe. \V Keepe ye my Sabbathes, and
feare my Sanctuarie. I the Lord. \V Decline not to magicians, neither
aske any thing of ſoothſayers, to be polluted by them. I the Lord your
God. \V Before the hoare head riſe vp, and honour the perſon of an old
man: and feare the Lord thy God. I am the Lord. \V If a ſtranger dwel in
your land, and abide among you, doe not vpbraid him: \V but let him be
among you as the ſame countrie man: and you shal loue him as your
ſelues: for you alſo haue bene ſtrangers in the Land of Ægypt. I the
Lord your God. \V Do not anie vniuſt thing in iudgement, in rule, in
weight, or meaſure. \V Let the balance be iuſt, and the weightes equal,
the bushel iuſt, and the ſextarie equal. I the Lord your God, that
brought you out of the Land of Ægypt. \V Keepe al my preceptes, and al
my iudgements, & doe them. I the Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
