%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 0305
%%% o-0276
\startChapter[
  title={Chapter 13}
  ]

\Summary{The law concerning leproſie in men, 47.~and in garments.}

And our Lord ſpake to Moyſes, and Aaron, ſaying: \V The man, in whoſe
skinne and flesh shal ariſe a diuers colour or a bliſter, or any thing
as it were shyning, that is to ſay the plague of the leproſie, shal be
brought
\SNote{It pertained to the Prieſts to diſcerne of leproſie, in figure of
Prieſtes authoritie to bind and looſe ſinnes in the new Teſtament.
\Cite{S.~Chryſoſt. li.~3. de Sacerdotio.}}
to Aaron the prieſt, or any one of his ſonnes. \V Who ſeing the leproſie
in his skinne, & the heare turned into a white colour, and the forme it
ſelf of the leproſie lower then the skinne,
%%% o-0277
and the other flesh: it is the plague of the leproſie, and at his
arbitrement he shal be ſeperated. \V But if there be a shining
whiteneſſe in the
skinne, and not lower then the other flesh, and the heare of the old
colour, the prieſt shal shut him vp ſeuen daies, \V and the ſeuenth day
he shal vew him: and if the leproſie be growne no farder, nor hath
paſſed the former limites in the skinne, againe he shal shut him vp
other ſeuen daies. \V And the ſeuenth day he shal behold: if the
leproſie be ſomewhat obſcure, and not growne in the skinne, he ſhal
cleanſe him, becauſe it is a ſcabbe: and the man shal wash his clothes,
and shal be cleane. \V And if the leproſie grow againe, after that he
was ſene of the prieſt and reſtored to cleanneſſe; he ſhal be brought
vnto him, \V and shal be condemned of vncleanneſſe. \V If the plague of
the
%%% 0306
leproſie be in a man, he ſhal be brought to the prieſt, \V and he shal
view him. And when there is
\SNote{Leproſie making ſpottes in the skinne of an other colour,
ſignifieth hereſie, that mixeth falſhood with truth.
\Cite{S.~Aug. li.~2. queſt. Euangel. c.~40.}}
a white colour in the skinne, and hath changed the looke of the heare,
and the flesh alſo it ſelfe appeare quicke: \V it shal be iudged a verie
old leproſie, and growne into the skinne. The prieſt therfore shal
contaminate him, and shal not shut him vp, becauſe he is vncleane
euidently. \V But if the leproſie ſpring forth running about in the
skinne, and couer al the flesh from the head to the feete, whatſoeuer
falleth vnder the ſight of the eies, \V the prieſt shal vew him, and
shal iudge that he is taken with a
\SNote{Sometimes that ſemeth leproſie, or hereſie, which is not: wherof
the prieſt is to iudge.
\XRef{Deut.~17.}}
moſt cleane leproſie: for that it is al turned into whitneſſe, and
therfore the man shal be cleane. \V But when there ſhal appeare in him
quicke flesh, \V they by the iudgement of the prieſt he shal be
polluted, and shal be reputed among the vncleane, for quick fleſh if it
be ſpotted with leproſie, is vncleane. \V And if againe it be turned
into whiteneſſe, and couer al the man, \V the prieſt shal vew him, and
shal iudge him to be cleane. \V And the fleſh and the skinne wherin a
boile is riſen, and healed, \V and in the place of the boile, there
appeare a white ſcarre, or ſomewhat redde, the man shal be brought to
the prieſt: \V and when he ſhal ſee the place of the leproſie lower then
the other flesh, and the heare turned into whitenes, he shal contaminate
him: for the plague of leproſie is riſen in the boile. \V But if the
heare be of the old colour, and the ſcarre ſomewhat obſcure, & be not
lower then the next flesh, he shal shut him vp ſeuẽ dayes. \V And if it
be growne farder, he shal iudge him to haue the leproſie, \V but if it
ſtay in his place, it is the ſcarre of a boile, and the man shal be
cleane. \V And the fleſh and skinne, that fire hath burnt, and being
healed hath a white or reade ſcarre, \V the prieſt shal conſider it, and
loe it is turned into whiteneſſe, and the place therof is lower then the
other skinne: he shal contaminate him, becauſe the plague of leproſie is
riſen in the ſcarre. \V But if the colour of the heare be not changed,
nor the blemish lower then the other fleſh, and the forme it ſelf of
the leproſie be ſomewhat obſcure, he shal shut him vp ſeuen dais, \V and
the 
\Fix{ſeuente}{ſeuenth}{obvious typo, fixed in other}
day he shal behold him: if the leproſie be growne farder in the skinne,
he shal contaminate him, \V but if the
%%% o-0278
whitnes ſtay in his place, not very cleare, it is the plague of a
burning, and therfore he shal be cleanſed, becauſe it is the ſcarre of a
burning.
%%% 0307
\V Man, or woman, in whoſe head or beard the leproſie riſeth, the prieſt
shal ſee them, \V and if the place be lower then the other flesh, and
the heare yelow, and thinner then it was wont, he shal contaminate them,
becauſe it is the leproſie of the heade and the beard. \V But if he
perceiue the place of the ſpotte equal with the flesh neere vnto it, and
the heare blacke: he shal shut him vp ſeuen daies, \V and in the ſeuenth
day he shal looke vpon it. If the ſpotte be not growne, and the heare be
of his owne colour, and the place of the plague euen with the other
flesh: \V the man shal be shauen ſauing the place of the ſpotte, and
shal be shut vp other ſeuen daies. \V If the ſeuenth day the plague ſeeme
to haue ſtaid in his place, & not lower then the other flesh, he shal
cleanſe him, and his clothes being washed he shal be cleane. \V But if
after his clenſing the ſpotte be growne againe in the skinne, \V he
shal no more ſeeke whether the heare be changed into a yelow colour,
becauſe he is euidently vncleane. \V Moreouer if the ſpotte be ſtaid, &
the heare be blacke, let him know that the man is healed, and let him
boldly pronounce him cleane. \V Man, or woman, in whoſe skinne
appeareth whiteneſſe, \V the prieſt shal vew them. If he find that
whitneſſe ſomewhat obſcure shineth in the skinne, let him know that it
is not the leproſie, but a ſpotte of white colour, and that the man is
cleane. \V The man whoſe heare falleth of from his head, is bald and
cleane: \V and if the heare fal from his forehead, he is bald before and
cleane. \V But if in the baldneſſe or in the baldneſſe before there be
riſen a white or reddish colour, \V and the prieſt perceiue this, he
shal condemne him vndoubtedly of leproſie, which is riſen in the
baldneſſe. \V Whoſoeuer therfore shal be defiled with leproſie & is
ſeparated at the arbitrement of the prieſt, \V ſhal haue his clothes
hanging looſe, his head bare, his mouth couered with a cloth, he shal
crie him ſelf polluted and vncleane. \V Al the time that he is a lepre &
vncleane, he shal dwel alone without the campe. \V A wollen or linnen
garment, that shal haue the leproſie, \V in the warpe, and the woufe, or
els a skinne, or whatſoeuer is made of a skinne, \V if it be a white or
reddish ſpott, it shal be reputed infected with leproſie, and shal be
shewed to the prieſt. \V Who hauing conſidered it shal shut it vp ſeuen
daies: \V and the ſeuenth day againe beholding it, if he find that it is
growne, it is a leproſie
%%% 0308
that continueth: he shal iudge the garment polluted, and euerie thing
wherin it shal be found, \V and therfore it shal be burnt with fire. \V
But if he ſee that it is not growne, \V he shal command, and they shal
wash that, wherin the leproſie is, and he shal shut it vp other ſeuen
daies. \V And when he shal ſee that the old shape is not returned,
neither yet that the leproſie is growen furder, he shal iudge it
vncleane, and burne it with fire, for that the leproſie is ſpred in the
outſide of the garment or through the whole. \V But if the place of the
leproſie be
%%% o-0279
ſomewhat obſcure, after that the garment is washed, he shal breake it
of, and diuide it from that which is ſounde. \V And if beſides in thoſe
places, that before were without ſpotte there appeare a flying and
wandring leproſie: it muſt be burnt with fire: \V if it be ceaſſed, he
shal wash with water the partes, that be pure, the ſecond time, & they
shal be cleane. \V This is the law of the leproſie of a wollen and
linnen garment, of the warpe and the woufe, and of al ſtuffe of skinnes,
how it ought to be clenſed, or contaminated.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
