%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 0298
%%% o-0270
\startChapter[
  title={Chapter 9}
  ]

\Summary{Sacrifices for ſinne, 12.~and of holocauſte, 18.~and pacifiques
  are offered: 22.~and Aaron ſtretching forth his hand bleſseth the
  people.}

And when the eight day was come, Moyſes called Aaron and his ſonnes, and
the ancients of Iſrael, and ſaid to Aaron: \V Take of the heard
\SNote{The people before worſhipped a calfe for God.
\XRef{Exod.~32.}
Now therfore they offer a calfe in ſacrifice to God, for their ſinne,
and to kepe them from idolatrie.
\Cite{S.~Hieron. in Hierem.~7.}}
a calfe for ſinne, and a ramme for an holocauſt, both without ſpot, and
offer them before our Lord. \V And to the children of Iſrael thou ſhalt
ſpeake: Take ye a bucke goate for ſinne, and a calfe, and a lambe, of a
yeare old, & without ſpot for an holocauſt, \V an oxe and a ramme for
pacifiques: and immolate them before our Lord, offering in the ſacrifice
of euerie one, flowre tempered with oile; for to day our Lord wil appeare
to you. \V They tooke therfore al thinges that Moyſes had cõmanded
before the dore of the tabernacle: where when al the multitude ſtood, \V
Moyſes ſaid: This is the word, which our Lord hath commanded: doe it,
\SNote{God appeared in his worke by ſending fire to burne the ſacrifice,
without mans induſtrie.
\XRef{v.~24.}}
and his glorie wil appeare to you. \V And he ſaid to Aaron: Approch to
the altar, & immolate for thy ſinne: offer the holocauſt, and pray for
thy ſelf and for the people, as our Lord hath cõmanded. \V And forthwith
Aaron approching to the altar, immolated the calfe for his ſinne: \V the
bloud wherof his ſonnes brought to him: wherin dipping
%%% 0299
his finger, he touched the hornes of the altar, and poured the reſt at
the foote therof. \V And the fatte, and the litle kidneis, and the caule
of the liuer, which are for ſinne, he burnt vpon the altar, as our Lord
had commanded Moyſes: \V but the flesh and skinne therof he burnt with
fire without the campe. \V He immolated alſo the victime of holocauſt:
and his ſonnes brought him the bloud therof, which he poured in the
circuite of the altar. \V The hoſte alſo it ſelfe being cut into peeces,
they brought with the head and euerie member. Al which he burnt with
fire vpon the altar, \V hauing firſt washed the entralles and the feete
with water. \V And offering for the ſinne of the people, he ſlew the
bucke goat: and
%%% o-0271
expiating the altar: \V he made the holocauſt, \V adding in the
ſacrifice the libaments, which are offered withal, and burning them vpon
the altar, beſide the ceremonies of the morning holocauſt. \V He
immolated alſo the oxe, and the ramme, the pacifique hoſtes of the
people: and his ſonnes brought him the bloud, which he poured vpon the
altar round about. \V The fatte alſo of the oxe, and the rump of the
ramme, and the two little
\Fix{kindneis}{kidneis}{obvious typo, same in other}
with their fatte, and the caule of the liuer \V they put vpon the
breſts, and after the fatte was burnt vpon the altar, \V their breſts,
and the right shoulders, Aaron did ſeperate, eleuating them before our
Lord, as Moyſes had commanded. \V And
\SNote{This did ſignifie that Chriſt, in whõ al nations are bleſſed,
ſhould be ſtretched on the Croſſe where he redemed vs, in memorie wherof
we now make the ſigne of the Croſſe.}
ſtretching forth his hand to the people, he bleſſed them. And ſo the
hoſtes for ſinne, and the holocauſtes, and the pacifiques being
finiſhed, he deſcended. \V And ſo the hoſtes for ſinne, and the
holocauſtes, and the pacifiques being finiſhed, he deſcended. \V And
Moyſes and Aaron going into the
tabernacle of teſtimonie, and afterward comming forth, bleſſed the
people. And the glorie of our Lord appeared to al the multitude: \V and
behold
\SNote{The Prieſts were commãded to nouriſh and kepe this fire
perpetually, that it should not be extinguiſhed.
\XRef{Cha.~6. v.~12.}}
a fire coming forth from our Lord, deuoured the holocauſt, and the
fatte that was vpon the altar: Which thing, when the multitude had ſenne,
they praiſed our Lord, falling on their faces.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
