%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 26
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-26


\project douay-rheims


%%% 0334
%%% o-0301
\startChapter[
  title={Chapter 26}
  ]

\Summary{VVith new prohibition of Idolatrie, and commandment to kepe the
  Sabbath. 3.~Rewardes are promiſed to al that obſerue Gods
  precepts. 14.~And manie miſerable punishments are threatned to al
  tranſgreſsors.}

I the Lord your God: you shal not make to your ſelues an idol and
\SNote{Heretikes holding their corrupt courſe, wil nedes haue an image of
Chriſt, or Sainct, to be the grauen thing, which is forbidden in holie
Scriptures: & therfore falſly tranſlate, \HH{Peſel} a \Emph{grauen image}
where in dede it ſignifieth an image, picture or purtrature of an idol,
that is, 
a \Emph{grauen idol}. So here as in other places, it is forbid to make
an idol, or ſimilitude of anie idol.}
thing grauen, neither shal you erect titles, nor ſet a notorious ſtone
in your land, for to adore it,
%%% 0335
for I am the Lord your God. \V Keepe my ſabbathes, and dread my
Sanctuarie. I the Lord. \V If you walke in my preceptes, and keepe my
commandementes, and doe them, I wil geue you raine in their ſeaſons, \V
and the earth shal bring forth her ſpring, and the trees ſhal be
repleniſhed with fruites. \V The threshing of your harueſt shal reach
vnto vintage, and the vintage shal reach vnto ſowing time: and you shal
eate your bread to your fil, and without feare shal you dwel in your
land. \V I wil geue peace in your coaſtes: you shal ſleepe, and there
shal be none to make you
%%% o-0302
afraid. I wil take away euil beaſtes: and the ſword shal not paſſe
through your quarters. \V You shal purſue your enemies, and they shal
fal before you. \V Fiue of yours shal purſue an hundred ſtrangers, and
an hundred of you tenne thouſand: your enemies shal fal by the ſeord in
your ſight. \V I wil reſpect you, and make you encreaſe: you shal be
multiplied, and I wil eſtablish my couenant with you. \V You shal eate
of the eldeſt of the old ſtore, and new coming vpon it you shal caſt forth
the old. \V I wil ſette my tabernacle in the middes of you, and my ſoule
shal not caſt you of. \V I wil walke among you, and wil be your God, and
you shal be my people. \V I the Lord your God: that haue brought you out
of the Land of the Ægyptians, that you should not ſerue them, and that haue
broken the chaines of your neckes, that you might goe vpright. \V But if
you wil not heare me, nor doe al my commandements, \V if you diſpiſe my
lawes, and contemne my iudgementes that you do not thoſe thinges which
are appointed by me, and bring my couenant to nothing worth: \V I alſo
wil do theſe thinges to you: I shal quickly viſite you with pouertie,
and burning heat, which shal waiſt your eies, and conſume your
liues. You shal ſowe your ſeede in vaine, which ſhal be deuoured of the
enemies. \V I wil ſette my face againſt you, and you shal fal downe
before your enemies, and shal be made ſubiect to them that hate you. You
shal flee, when no man purſueth you. \V But if you wil not obey me ſo
neither, I wil increaſe your chaſtiſementes ſeuen fould for your
ſinnes, \V and wil breake the pride of your ſtubburneſſe. And I wil make
to you the heauen, from aboue as iron, and the earth as braſſe. \V Your
labour ſhal be ſpẽt in vaine, the earth ſhal not bring forth her ſpring,
nor the trees yeld their fruites. \V If you walke contrarie
%%% 0336
to me, and wil not heare me, I wil increaſe your plagues vntil ſeuen
fould for your ſinnes: \V and I wil ſend in vpon you the beaſtes of the
field, which may conſume you, and your cattaile, and may bring al
thinges to a ſmal number, and that your waies may be made deſert. \V And
if you wil neither ſo receiue diſcipline, but walke rather contrarie to
me: \V I alſo wil goe oppoſite againſt you, and wil ſtrike you ſeuen
times for your ſinnes. \V And I wil bring in vpon you the ſword a
reuenger of my couenant. And when you ſhal flee into the cities, I wil
ſend the peſtilence in the middeſt of you, and you shal be deliuered in
the handes of the enemies, \V after I shal haue broken the ſtaffe of
your bread: ſo that tenne wemen shal bake your breades in one ouen, and
ſhal render them by weight: and you shal eate, and shal not be
filled. \V But if you wil neither by theſe meanes heare me, but walke
againſt me: \V I alſo wil goe againſt you in contrarie furie, and wil
chaſtice you with ſeuen plagues for your ſinnes, \V ſo that you shal
\SNote{This extreme famine fel vpon ſome of them in Samaria,
\XRef{4.~Reg.~6.}
vpon others in Hieruſalem,
\XRef{4.~Reg.~25.}
moſt ſpecially whẽ they were beſiged by Titus.
\Cite{Ioſephus li.~7. c.~6. de bella Iudaico.}}
eate the flesh of your ſonnes and of your daughters. \V I wil deſtroy your
excelſes, and breake your idols. You ſhal fal among the ruines of your
idols and my ſoule ſhal abhorre you, \V in ſo much that I wil bring your
cities into a wilderneſſe, and I wil make your Sanctuaries deſert,
neither wil I receiue any more the moſt
%%% o-0303
ſweete odour. \V And I wil diſtroy your land, and your enemies ſhal be
aſtonied vpon it, when they ſhal be inhabitants therof. \V And you I wil
diſperſe into the Gentiles, and wil draw out the ſword after you, and
your land shal be deſert, and your cities deſtroied. \V Then shal the
land take pleaſure in her ſabbathes al the daies of her deſolation: when
you shal be \V in the enemies land, she shal ſabbatize, and reſt in the
ſabbathes of her deſolation, becauſe ſhe did not reſt in your ſabbathes
when you dwelt in it. \V And they that shal remaine of you, I wil put
feare in their hartes in the countries of their enemies, the ſound of a
flying leafe shal terrifie them, and they shal flie it as it were a
ſword: they shal fal, when no man purſueth, \V and they shal euerie one
fal vpon their brethren, as flying from warres, none of you shal be ſo
hardie as to reſiſt your enemies. \V You shal perish among the Gentiles,
and the enemies land shal conſume you. \V And if of them alſo ſome
remaine, they shal pine away in their iniquities, in the land of their
enemies, and for the ſinnes of
%%% 0337
their fathers, and their owne they shal be afflicted: \V vntil they
confeſſe their owne and their
\Fix{auncetours}{aunceſtours}{likely typo, fixed in other}
iniquities, wherby they haue preuaricated againſt me, and walked
contrarie vnto me. \V I alſo therfore wil walke againſt them, and bring
them into their enemies land, vntil their vncircumciſed mind be ashamed:
then shal they pray for their impieties. \V And I wil remember my
couenant, that I made with
\SNote{Iacob is firſt here named, becauſe he had no other children but
this people, for Iſaac was alſo father of the Idumeans, and Abrahã
moreouer of the Iſmaelites and Madianites, and becauſe the great
promiſes made to Abraham and Iſaac pertained only to the Iſraelites.
\Cite{Theodoret. q.~36. in Leuit.}}
Iacob, and Iſaac, and Abraham. Of the land alſo I wil be mindful: \V
which when it shal be left of them, shal take pleaſure in her ſabbathes,
being deſolate for them. But they shal pray for their ſinnes, for that
they reiected my iudgementes, and diſpiſed my lawes. \V Howbeit euen when
they were in the land of their enemies,
\SNote{The church neuer wholly decayeth.}
I did not caſt them of altogether, neither did I ſo diſpiſe them, that
they should be conſumed, and I should make my couenant with them
fruſtrate. For I am the Lord their God, \V and I wil remember mine old
couenant, when I brought them out of the Land of Ægypt, in the ſight of
the Gentiles, for to be their God. I the Lord. Theſe are the
iudgementes, and precepts, and lawes, which our Lord gaue betwen him
ſelf and the children of Iſrael in Mount Sinai by the hand of Moyſes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
