%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 0292
%%% o-0265
\startChapter[
  title={Chapter 7}
  ]

\Summary{The maner of offering hoſtes for offences, 11.~and of pacifique
  victimes for thankſgeuing. 22.~No fatte, 26.~nor bloud is to be eaten.}

This alſo is the lawe of an hoſte for an offence, the moſt Holie: \V
therfore where the holocauſt is immolated, the victime alſo for an
offence shal be ſlaine: the bloud therof shal be poured round about the
altar. \V They shal offer therof the rumpe and the fatte that couereth
the entralles: \V the two little kidneys, and the fatte that is beſide
the guttes, and the caule of the lyuer with the two litle kidneys. \V
And the prieſt shal burne them vpon the altar: it is the burnt ſacrifice
of our Lord for an offence. \V Euerie male of the prieſts ſtocke, in a
holie place ſhal eate this fleſh, becauſe it is moſt Holie. \V As the
hoſte for
\SNote{This text ſheweth there is difference in the nature of a
fault \Emph{committed}, commonly called \Emph{ſinne}, & of
dutie \Emph{omitted} here called \Emph{offence}. In Latin \L{peccatũ}
and \L{delictum}. Yet both are alike offencife to God, in matter of
equal importance. As appeareth for that the ſame ſacrifice was offered
for both.
\Cite{S.~Aug. q.~20. in Leuit.}}
ſinne is offered, ſo alſo that for an
%%% !!! This is the same SNote as above.
%%% \SNote{}
offence: the lawe of both hoſtes ſhal be one: to the prieſt that
offereth it, it ſhal pertaine. \V The prieſt that offereth the victime
of holocauſt, ſhal haue the skinne therof. \V And euerie ſacrifice of
flowre, that is baked in the ouen, and whatſoeuer is prepared vpon the
grediron, or in the frying panne, it shal be that prieſtes by
%%% 0293
whom it is offered. \V Whether they be tempered with oyle, or drie, to
al the ſonnes of Aaron equal meaſure shal be diuided to euerie one. \V
This is the law of the hoſte of pacifiques that is offered to our
Lord. \V If the oblation be for thankes geuing, they shal offer loaues
without leauen tempered with oyle, and wafers vnleauened laid ouer with
oyle, and fryed floure, and manchets tempered with the mingling of
oyle: \V loaues alſo leauened with the hoſte of thankes, which is
offered for pacifiques: \V wherof one for firſt fruites shal be
\SNote{Geuen or preſented to our Lord, not offered vpon the Altar, for
no leauen could be offered in ſacrifice.
\XRef{cap.~2. v.~11.}}
offered to our Lord, and shal be the prieſts that shal poure out the
bloud of the hoſte. \V The flesh wherof shal be eaten the ſame day,
neither shal anie of it remaine vntil morning. \V If anie man by vowe,
or of his owne accorde offer an hoſte, it shal in like maner be eaten
the ſame day: but if ought remaine vntil the morrow, it is lawful to
eate it: \V but whatſoeuer the third day shal finde, fire shal conſume
it. \V If anie man eate of the flesh of the victime of pacifiques the
third day, the oblation shal be of none effect, neither shal it profite
the offerer: yea rather whatſoeuer ſoule shal defile it ſelfe with ſuch
meate, shal be guiltie of preuarication. \V The flesh that hath touched
any vncleane thing, shal not be eaten, but shal be burnt with fire: he
that is cleane, shal eate therof. \V A ſoule being polluted that eateth
of the flesh of the hoſte of pacifiques, which is offered to our Lord,
shal perish from his people. \V And
%%% o-0266
that which hath touched the vncleannes of man, or of beaſt, or of anie
thing that can pollute, and eateth of ſuch kind of flesh, shal perish
from his people. \V And our Lord ſpake to Moyſes, ſaying: \V Speake to
the children of Iſrael: The fatte of a sheepe, and of an oxe, and of a
goate you shal not eate. \V The fatte of the carcaſſe of carraine, and
of the beaſt, that was caught of an other beaſt, you shal haue for
diuers vſes. \V If anie man eate the fatte, that should be offered for
the burnt ſacrifice of our Lord, he shal perish out of his people. \V
The bloud alſo of whatſoeuer beaſt you shal not take in meate, aſwel of
birdes as of cattel. \V Euerie ſoule that eateth bloud, shal perish out
of his people. \V And our Lord ſpake to Moyſes, ſaying: \V Speake to
the children of Iſrael, ſaing: He that offereth a victime of pacifiques
to our Lord, let him offer therwith a ſacrifice alſo, that is, the
libaments therof. \V He shal hold in his handes the fatte of the hoſte,
%%% 0294
and the breſt: and when he hath offered an conſecrated both to our Lord,
he shal deliuer them to the prieſt, \V who shal burne the fatte vpon the
altar, but the breaſt shal be Aarons, and his ſonnes. \V The right
shoulder alſo of the pacifique hoſtes shal fal for firſt fruites of the
prieſt. \V He of the ſonnes of Aaron, that offereth the bloud, and the
fatte, him ſelfe shal haue the right shoulder alſo for his portion. \V
For the breſt of eleuation and the shoulder of ſeperation I haue taken
of the children of Iſrael, from their pacifique hoſtes, and haue geuen
them to Aaron the prieſt, and to his ſonnes, by a lawe for euer, of al
the people of Iſrael. \V This is the anoynting of Aaron and his ſonnes,
in the ceremonies of our Lord, in the day when Moyſes offered them, that
they might doe the function of prieſthood, \V and the thinges that our
Lord commanded to be geuen them of the children of Iſrael, by a
perpetual religion in their generations. \V This is the lawe of
holocauſte, and of the ſacrifice for ſinne, and for an offence, and for
conſecration, and the victimes of pacifiques: \V Which our Lord apointed
to Moyſes in mount Synai when he commanded the children of Iſrael, that
they should offer their oblation to our Lord in the deſert of Synai.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
