%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-22


\project douay-rheims


%%% 0325
%%% o-0293
\startChapter[
  title={Chapter 22}
  ]

\Summary{VVho may eate of ſanctified things. 17.~And what things may be
  offered.}

Ovr Lord alſo ſpake to Moyſes, ſaying: \V Speake to Aaron and to his
ſonnes, that they beware of thoſe that are the conſecrated thinges of the
children of Iſrael, and
\Fix{contaninate}{contaminate}{possible typo, fixed in other}
not the name of the thinges ſanctified to me, which they offer. I the
Lord. \V Say to them, and to their poſteritie: Euerie man of your
ſtocke, that approcheth to thoſe thinges that are conſecrated, and which
the children of Iſrael haue offered to the Lord, in whom there is
vncleaneſſe, shal perish before the Lord. I am the Lord. \V The man of
the ſeede of Aaron, that is a leper, or hath a fluxe of ſeede,
\SNote{By theſe accidental vncleannes, was prefigured the cenſure of
ſuſpenſion in clergie men.}
shal not eate of thoſe thinges that are ſanctified to me vntil he be
healed. He that toucheth a thing vncleane by occaſion of that is dead,
and he from whom iſſueth ſeede as it were of copulation, \V and he that
toucheth a creeping beaſt, and whatſoeuer vncleane thing, the touching
wherof is filthie, \V shal be vncleane vntil euen, and shal not eate
thoſe thinges that are ſanctified: but when he hath washed his flesh
with water, \V and the ſunne is downe, then being cleanſed he shal eate
of
%%% o-0294
the ſanctified thinges, becauſe it is his meate. \V Carrien and that
which was taken of a beaſt, they shal not eate, nor be polluted in
them. I am the Lord. \V Let them keepe my preceptes, that they be not
ſubiect to ſinne, and die in the Sanctuarie, when they shal haue
polluted it. I the Lord that ſanctifie them. \V No ſtranger shal eate of
the ſanctified thinges, the prieſtes gueſt, and hireling shal not eate
of them. \V But whom the prieſt hath bought, and he that is his ſeruant
borne in his houſe, theſe shal eate of them. \V If the daughter of a
prieſt be maried to anie of the people: of thoſe thinges that are
ſanctified, and of the
%%% 0326
firſt fruites ſhe ſhal not eate. \V But if ſhe be a widow, or put away,
and without children returne to her fathers houſe: as ſhe was wont being
a maide, ſhe ſhal be fedde with her fathers meates. No ſtranger hath
power to eate of them. \V He that eateth of the ſanctified thinges by
ignorance, shal adde the fifth part with that which he did eate, and
shal geue it to the prieſt into the Sanctuarie. \V Neither shal they
contaminate the ſanctified thinges of the children of Iſrael, which they
offer to the Lord: \V leſt perhappes they ſuſtaine the iniquitie of
their offence, when they shal haue eaten the ſanctified thinges. I the
Lord that ſanctifie them. \V And our Lord ſpake to Moyſes, ſaying: \V
Speake to Aaron, and to his ſonnes, and to al the children of Iſrael,
and thou shalt ſay to them: The man of the houſe of Iſrael, and of the
ſtrangers which dwel with you, that offereth his oblation, either paying
his vowes, or offering of his owne accord, whatſoeuer that be which he
preſenteth for an holocauſt of the Lord, \V to be offered by you, it
shal be a male without ſpotte of beefes, and muttons, & of goates. \V If
it haue a blemish, you shal not offer it, neither shal it be
acceptable. \V The man that offereth a victime of pacifiques to the
Lord, either paying his vowes, or offering of his owne accorde, aſwel of
beefes as of muttons, shal offer it without blemish that it may be
acceptable: there shal be no blemish in it. \V If it be blind, if it be
broken, if it haue a ſcarre, if bliſters, or a ſcabbe, or a drie ſcurfe:
you shal not offer them to the Lord, nor burne of them vpon the Lordes
altar. \V An oxe and a sheepe, hauing the eare and the taile cut of,
thou maieſt offer voluntarily but a vow can not be paied of them. \V No
beaſt that hath the ſtones bruiſed, or crushed, or cut and taken away,
shal you offer to the Lord, and in your land make not this at al. \V Of
the hand of a ſtranger you shal not offer breades to your God, and what
other thing ſoeuer he would geue: becauſe they are al corrupted, and
blemished: you shal not receiue them. \V And our Lord ſpake to Moyſes,
ſaying: \V An oxe, a sheepe, and a goate, when they are brought forth,
shal be ſeuen daies vnder the vdder of their damme: but the eight day,
and ſo forward they may be offered to the Lord. \V Whether it be a
beefe, or a sheepe, they shal not be immolated in one day with their
young ones. \V If you immolate an hoſte for thankes giuing to the Lord,
that he may be placable, \V the
%%% 0327
ſame day you shal eate it, there shal not ought remaine vntil the
morning of the next day. I the Lord. \V Keepe my
%%% o-0295
commandementes, and doe them. I the Lord. \V Pollute not my holie name,
that I may be ſanctified in the middes of the children of Iſrael. I the
Lord that ſanctifie you, \V and brought you out of the Land of Ægypt,
that I might be your God. I the Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
