%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 24
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-24


\project douay-rheims


%%% 0330
%%% o-0298
\startChapter[
  title={Chapter 24}
  ]

\Summary{Prouiſion of oile for lampes in the Tabernacle. 5.~The making,
  and diſpoſing the loaues of propoſition. 10.~The punishment of
  blaſphemie, and man ſlaughter. 18.~And the paine of equal reuenge.}

And our Lord ſpake to Moyſes, ſaying: \V Command the children of Iſrael,
that they bring vnto thee: oyle of oliues moſt pure, and cleare, to
furnish the lampes continually, \V without the veile of the teſtimonie
in the tabernacle of couenant. And Aaron shal ſette them from euen vntil
morning before the Lord, by a perpetual ſeruice and rite in your
generations. \V Vpon the candleſticke moſt cleane shal they be putte
alwaies in the ſight of the Lord. \V Thou shalt take alſo floure, and
shalt bake therof twelue breades, which shal haue euerie one
\SNote{Two tenthes of an ephi, that is, two gomors. A gomor of Manna,
which is the tenth part of an ephi,
\Fix{ſufficied}{ſufficed}{obvious typo, fixed in other}
one man for a day
\XRef{Exod.~16. v.~16. &~36.}
ſo that one of theſe loaues was as much as al the meate which two do
ordinarily eate in one day.}
two tenthes: \V which thou shalt ſette ſix one againſt an other vpon the
moſt cleane table before the Lord, \V and thou shalt put vpon them the
cleareſt franckincenſe, that the bread may be for a moniment of the
oblation of the Lord. \V Euerie ſabbath they shal be changed before the
Lord, receiued of the children of Iſrael by an euerlaſting couenant: \V
and they shal be Aarons and his ſonnes, that they may eate them in a
holie place: becauſe it is moſt Holie of the ſacrifices of the Lord
%%% 0331
by a perpetual right. \V And behold there went forth the ſonne of a
woman of Iſrael, whom she had borne of an Ægyptian among the children of
Iſrael, and fel at wordes in the campe with a man of Iſrael. \V And when
he had blaſphemed the name, and had curſed it, he was brought  to
Moyſes: (And his mother was called Salumith, the daughter of Dabri of
the tribe of Dan.) \V And they did caſt him into priſon, til they might
know what our Lord would command. \V Who ſpake to Moyſes, \V ſaying:
Bring forth the blaſphemer without the campe, and let al that heard him,
put their handes vpon his head, and let al the people ſtone him. \V And
to the children of Iſrael thou ſhalt ſpeake: The man that curſeth his
God, shal beare his ſinne: \V and he that blaſphemeth the name of the
Lord, dying let him dye: al the multitude of the people shal ſtone him,
whether he be a natural, or ſtranger. He that blaſphemeth the name of
the Lord, dying let him dye. \V He that ſtriketh, and killeth a man,
dying let him dye. \V He that ſtriketh a beaſt, shal render one for it,
that is to ſay, ſoule for ſoule. \V He that giueth anie of his
neighbours a blemish,
\SNote{This Law deſigning equalitie, was to put a limite, not to enforce
to reuenge, for the partie damaged, if he would, might remitte al or
part.
\Cite{S.~Aug. li.~19. c.~25. cont. Fauſtum.}}
as he did, ſo shal it be done to him: \V fracture for fracture, eye for
eye, tooth for tooth he shal reſtore. What blemish he gaue, the like
shal he
%%% o-0299
be compelled to ſuſtaine. \V He that ſtriketh a beaſt, ſhal render an
other. He that ſtriketh a man, shal be puniſhed. \V Let there be equal
iudgement among you, whether a ſtranger, or a natural ſinne: becauſe I
am the Lord, your God. \V And Moyſes ſpake to the children of Iſrael:
and they brought him forth that had blaſphemed, without the campe, and
they ſtoned him. And the children of Iſrael did as our Lord had
commanded Moyſes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
