%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 0311
%%% o-0281
\startChapter[
  title={Chapter 15}
  ]

\Summary{The law of vncleane iſſues in men, 19.~and in wemen.}

And our Lord ſpake to Moyſes and Aaron, ſaying: \V Speake to the
children of Iſrael, and ſay to them: The man that hath a fluxe of ſede,
shal be
\SNote{If natural infirmities brought vncleãnes much more laſciuiouſnes
of the mind.
\Cite{Theod. q.~15. &~20. in Leuit.}}
vncleane. \V And then he shal be iudged ſubiect to this fault, when the
filthie humour at euerie monent, cleaueth to his flesh, and is
congealed. \V Euerie bedde wheron he ſleepeth, shal be vncleane, and
where ſoeuer he ſitteth. \V If anie man touch his bedde, he shal waſh
his clothes: and him ſelf being washed with water, shal be vncleane
vntil euen. \V If he ſitte where that man had ſitten, he alſo shal wash
his clothes: and being washed with water, shal be vncleane vntil
euen. \V He that toucheth his flesh, shal wash his clothes: and him ſelf
being washed with water shal be vncleane vntil euen. \V If ſuch a man
caſt his ſpittle vpon him that is cleane, he shal wash his clothes: and
being
%%% o-0282
washed with water he shal be vncleane vntil euen. \V The padde wherupon
he ſitteth shal be vncleane: \V and whatſoeuer hath bene vnder him, that
hath the fluxe of ſeede, shal be polluted vntil euen. He that carieth
any of theſe thinges, shal wash his clothes: and him ſelf being washed
with water, shal be vncleane vntil euen.
%%% 0312
\V Euerie one, whom he toucheth that is ſuch an one, hauing not waſhed
his handes before, ſhal wash his clothes, and being washed with water,
ſhal be vncleane vntil euen. \V The earthen veſſel that he toucheth ſhal
be broken: but the wodden veſſel ſhal be washed with water. \V If he be
healed that hath ſuch a diſeaſe, he shal number ſeuen daies after his
cleanſing, and hauing washed his clothes, and al his bodie, in liuing
water, he shal be cleane. \V And the eight day he ſhal take two turtles,
or two yong pigeons, and he shal come into the ſight of our Lord, to the
dore
\Fix{fo}{of}{obvious typo, fixed in other}
the tabernacle of teſtimonie, and ſhal geue them to the prieſt, \V who
ſhal
\SNote{To make, offer, and ſacrifice an hoſte, is al one.}
make one for ſinne, and the other for an holocauſt, and he shal pray for
him before our Lord, that he may be cleanſed of the fluxe of his
ſeede. \V The man from whom iſſueth the ſeede of copulation, shal wash
with water al his bodie: and he shal be vncleane vntil euen. \V The
garment and skinne, that he weareth, he shal wash with water, and it
shal be vncleane vntil euen. \V The woman with whom he compaineth shal
be washed with water and shal be vncleane vntil euen. \V The woman that
monthly hath the fluxe of bloud, shal be ſeparated ſeuen daies. \V
Euerie one that toucheth her, shal be vncleane vntil euen: \V and that
whereon she ſleepeth or ſitteth in the daies of her ſeparation shal be
polluted. \V He that toucheth her bedde shal wash his clothes: and him
ſelf being washed with water, shal be vncleane vntil euen. \V Whoſoeuer
shal touch any veſſel wherupon she ſitteth, he shal wash his clothes:
and him ſelf being washed with water shal be polluted vntil euen. \V If
a man companie with her in the time of her menſtrual bloud, he shal be
vncleane ſeuen daies: and euerie bedde wheron he ſleepeth shal be
polluted. \V The woman that hath a fluxe of bloud many daies not in her
menſtrual time, or that ceaſeth not to haue a fluxe, after the menſtrual
bloud, as long as she is ſubiect to this diſeaſe, she shal be vncleane,
as if she were in her menſtrual time. \V Euerie bedde, whereupon she
ſleepeth, and veſſel wheron she ſitteth, shal be polluted. \V Whoſoeuer
toucheth them shal wash his clothes: and him ſelf being washed with
water, shal be vncleane, vntil euen. \V If the bloud ſtay and ceaſe to
runne, she shal count the ſeuen daies of her purification: \V and the
eight day she shal offer for her ſelfe to the prieſt, two turtles, or
two young pigeons, at the dore of the
%%% 0313
tabernacle of teſtimonie: \V who shal make one for ſinne, and the other
for an holocauſt, and he shal pray for her before our Lord, and for the
fluxe of her vncleanneſſe. \V You shal teach therfore the children of
Iſrael, that they take heede of
\SNote{S.~Hierom. (in
\Cite{cap.~5. ad Galat.})
vnderſtandeth this place of the abominable ſĩnes that may not be named.}
vncleanneſſe, and die not in their filthineſſe, when they shal pollute
my tabernacle that is among them. \V This is the law of him
%%% o-0283
that hath the fluxe of ſede and that is polluted by copulation, \V and
the woman that is ſeparated in her menſtrual times, or that hath a
continual fluxe of bloud, and of the man, that ſleepeth with her.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
