%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 0287
%%% o-0260
\startChapter[
  title={Chapter 4}
  ]

\Summary{Hovv a Prieſt, 13.~the
\Fix{multitud,}{multitude,}{possible typo, fixed in other}
22.~a Prince, 27.~or anie one of the people, ſinning of ignorance, muſt
offer hoſtes.}

And our Lord ſpake to Moyſes, ſaying: \V Speake to the children of
Iſrael: The ſoule that ſinneth by
\SNote{Ignorance of that we are bond to know is ſinne: and more in
Prieſts then in others.}
ignorance, and doth anie thing of al the commandmentes of our Lord,
which he commanded not to be done. \V If the prieſt that is anointed
ſinne, making the people to offend, he ſhal offer for his ſinne, a calfe
without ſpott to our Lord: \V and he shal bring it to the dore of
teſtimonie before our Lord, and shal put his hand vpon the head therof,
and shal immolate it to our Lord. \V He shal drawe alſo of the bloud of
the calfe, carying it into the tabernacle of teſtimonie. \V And when he
hath dipped his finger in the bloud he shal ſprinckle it
\SNote{VVe ſee here and in manie places that numbers are myſtical, not
alwayes ſuperſticious.}
ſeuen times before our Lord, againſt the veile of the Sanctuarie. \V And
of the ſame bloud he shal put vpon the hornes of the altar of incenſe
moſt acceptable to our Lord, which is in the tabernacle of
teſtimonie. And al the reſt of the bloud he shal powre at the foote of
the altar of holocauſte in the entrie of the tabernacle. \V And the fatt
of the calfe he shal take away for the ſinne, aſwel that which couereth
the entralles, as al the partes that are within. \V The two little
kidneys, and the caule that is vpon them beſide the guttes, and the fatt
of the liuer with the two litle kidneis, \V as is taken away from the
calfe which is an hoſt of pacifiques, and he shal burne them vpon the
altar of holocauſte. \V But the skinne and al the flesh with the heade
and feete and bowles and dung, \V and the reſt of the bodie he shal
carie forth without the campe into a cleane place, where the ashes are
wount to be powred out: and he shal burne them vpon a pyle of wood,
which shal be burnt in the place where the ashes are powred out. \V And
if al the multitude of Iſrael be ignorant, and through ignorance do that
which is againſt the commandement of our Lord, \V and
%%% 0288
afterward vnderſtand their ſinne, they ſhal offer for their ſinne
\SNote{A Prieſt, and the whole multitude offered the ſame ſacrifice of a
calfe, for their ſĩnes: the Prince offered a male goate, a priuate
perſon a femal. See
\Cite{Theodoret. q.~1. in Leuit. verſus finem.}}
a calfe, and ſhal bring it to the dore of the tabernacle. \V And the
ancientes of the people shal put their handes vpon the head therof before
our Lord. And the calfe being immolated in the ſight of our Lord, \V the
prieſt that is anoynted, shal carie of the bloud into the tabernacle of
teſtimonie, \V dipping his finger, and ſprinkling ſeuen times againſt
the veile. \V And he shal put of the ſame bloud on the hornes of the altar,
that is before our Lord, in the tabernacle of teſtimonie: and the reſt
of the bloud, he shal poure at the foote of the altar of holocauſte,
which is at the dore of the tabernacle of
%%% o-0261
teſtimonie. \V And al the fatte therof he shal take vp, & shal burne it
vpon the altar: \V doing ſo with this calfe, as he did alſo before: and
the prieſt praying for them, our Lord wil be propitious vnto them. \V
But the calfe it ſelfe he shal carie forth without the campe, and shal
burne it like as the former calfe:
\SNote{Though in ordinarie ſacrifices for ſinne, one part was
alotted to the Prieſts (by whoſe miniſterie God remitted ſinnes) yet of
the ſacrifice for a prieſts ſinnes, or for the ſinnes of the multitude,
the prieſts had no part, leſt they ſhould ſaue coſt, or reape commoditie
by ſacrifice for their owne or the whole peoples ſinnes. But al was
burned as in a holocauſt.
\Cite{Theodoret. q.~3. in Leuit.}
\Cite{S.~Tho. l.~2. q.~102. a.~3. ad.~8.}}
becauſe it is for the ſinne of the multitude. \V If a Prince ſinne, and
by ignorance do of manie thinges one, that by the law of our Lord is
forbidden: \V and afterward vnderſtandeth his ſinne, he shal offer an
hoſte to our Lord, a bucke of the goates without ſpotte. \V And he shal
put his hand vpon the head therof: and when he hath immolated it in the
place where holocauſt is wont to be ſlaine before our Lord, becauſe it
is for ſinne, \V the prieſt shal dippe his finger in the bloud of the
hoſte for ſinne, touching the hornes of the altar of holocauſte, and the
reſt powring at the foote therof. \V But the fatt he shal burne vpon it,
as is wont to be done in the victimes of pacifiques: and the prieſt shal
pray for him, and for his ſinne, and it shal be forgiuen him. \V And if
a ſoule of the people of the land shal ſinne through ignorance, doing
anie of thoſe thinges, that by the law of our Lord are forbidden, and
offending, \V and knoweth his ſinne, he shal offer a she goate without
ſpotte. \V And he shal put his hand vpon the head of the hoſt that is
for ſinne, and shal immolate it in the place of holocauſt. \V And the
prieſt shal take vp of the bloud with his finger: and touching the
hornes of the altar of holocauſt, the reſt he ſhal powre out at the
foote therof. \V But taking away al the fatt, as is wont to be taken
away of the victimes of pacifiques, he shal burne it vpon the altar, for
a ſwete ſauour to our Lord; and he shal pray for him, and it shal be
forgeuen him. \V But if he offer of the flocke
%%% 0289
a victime, for his ſinne, to wit, an ewe without ſpotte; \V he ſhal put
his hand vpon the head therof, and shal immolate it in the place where
the hoſtes of holocauſts are wont to be ſlayne. \V And the prieſt shal
take of the bloud therof with his finger, and touching the hornes of the
altar of holocauſt, the reſt he shal powre at the foote therof. \V Al
the fatte alſo he shal take away, as the fatte of the ramme, that is
offered for pacifiques, is wont to be taken away: and shal burne it vpon
the altar, a burnt ſacrifice of our Lord: and he shal pray for him, and
for his ſinne, and it shal be forgiuen him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
