%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Leviticus
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-23


\project douay-rheims


%%% 0327
%%% o-0295
\startChapter[
  title={Chapter 23}
  ]

\Summary{The ſolemnities of the Sabbath, 5.~of Paſch and firſt fruites,
  15.~of Pentecoſt, 23.~of Trumpettes, 26.~of Expiations, 33.~of
  Tabernacles; and with what rites the ſame shal be celebrated.}

And
\MNote{The fourth part.

Of feaſtes, times of reſt, & Iubilie, with
\Fix{priuilegies,}{priuileges,}{likely typo, fixed in other}
rewardes & puniſhments.}
our Lord ſpake to Moyſes, ſaying: \V Speake to the children of Iſrael,
and thou shalt ſay to them: Theſe are
\LNote{The feſtiuities.}{As
\MNote{Feſtiual daies perteine to the ſeruice of God.}
other lawes written by Moyſes concerning
Sacrifices, Sacraments, Degrees hindering mariage, puniſhments of
ſinnes, and the like, are partly moral, pertaining to the law of nature,
partly ceremonial, or iudicial, which may be altered: ſo this law of
feſtiuities is partly moral, for that al men are bound to kepe ſome
feſtiual dayes in honour of God, partly ceremonial, and ſo the Sabbath
day was kept holie in the old Teſtament, the ſeuenth day of the weke,
and other feaſtes, the dayes here preſcribed. But theſe particular
feaſtes and times, are abrogated by Chriſt, whom they prefigured.
\MNote{It is hereſie to kepe the Sabbath holie day.}
In ſo much that now it is not lawful to kepe them, for it would ſignifie
that Chriſt were not come, as S.~Paul teacheth
(\XRef{Rom.~14.}
\XRef{Galat.~4.}
\XRef{Colloſſ.~2.}
and in other places) and it were plaine Iudaiſme, and Hereſie condemned
by the Councel of Laodicia
\Cite{cap.~29.}
\Emph{accurſing them that Iudaize abſtayning that day from
vvorkes}. S.~Gregorie alſo refuteth this hereſie,
\Cite{li.~11. Epiſt.~3.}
ſhewing that Antichriſt wil embrace it fauouring the Iewes:
\MNote{In place therof we kepe Sunday.}
In place wherof the next day (which we cal ſunday) is made a perpetual
holie day, by authoritie of the Church, and called \L{dies
Dominica} \Emph{our Lords day}
\XRef{(Apoc.~1.)}
And this change the Proteſtants confeſſe to be lawful and
\Fix{neceſſaire,}{neceſſarie,}{obvious typo, fixed in other}
though we haue no other expreſſe ſcripture, when, or by whom it was
donne, but only that S.~Iohn had his reuelation \Emph{in our Lords day},
but by perpetual tradition al Chriſtians know, that the day after the
ſabbath is our wekelie holie day, in memorie of Chriſts Reſurrection the
ſame day, and in figure of the general reſurrection of al men, and of
life euerlaſting to the bleſſed.
\Cite{S.~Aug. li.~22. c.~30. ciuit.}
&
\Cite{Epiſt.~119. c.~15.}
and
\Cite{S.~Hierome Epiſt. ad Hedib.}
\MNote{Other feaſtes alſo changed, and new inſtituted by the ſame
authoritie.}
The ſame reaſon and authoritie do alſo warrant the change of other
feaſtes, and inſtitution of new, in honour of God, our Sauiour Ieſus
Chriſt, his Mother, and other Sainctes, and in memorie of benefites
receiued, as here we ſee in the old Teſtament diuers were commanded by
God, ſome alſo inſtituted long after Moyſes, as by Mardocheus and other
Iewes,
\XRef{Eſter.~9.}
and the reſtauration with new dedication of the altar
\XRef{1.~Machab.~4.}
obſerued by our ſauiour
\XRef{Ioan.~10. v.~22.}}
%%% !!! I moved the SNotes a bit to make more sense.
\SNote{There were eight ſeueral, feaſtes commãded by this law (beſides
the dailie ſacrifice) mentioned
\XRef{Num.~28. &~29.}
of which onlie ſeuen are here expreſſed.}
the feſtiuities of our Lord, which you shal cal holie. \V Six daies ye
shal doe worke:
\SNote{1.~The Sabbath in memorie that God created al things in ſix daies
& reſted the ſeuenth.}
the ſeuenth day, becauſe it is the reſt of the ſabbath, shal be called
holie. No worke shal you doe in it: it is the Sabbath of the Lord in al
your habitations. \V Theſe therfore are the holie feſtiuities of the
Lord, which you muſt celebrate in their times. \V The firſt moneth, the
fourtenth day of the moneth at euen, is the
\SNote{2.~Paſch in memorie of their deliuerie from Ægypt.}
Phaſe of the Lord: \V and the fiftenth day of this moneth is the
ſolemnitie of the Azymes of the Lord. Seuen daies shal you eate
azymes. \V The firſt day shal be moſt ſolemne vnto you, and holie: no
ſeruile worke shal you doe in it: \V but you shal offer ſacrifice in
fire to the Lord ſeuen daies. But the ſeuenth day shal be more ſolemne,
and more holie: and you shal doe no ſeruile worke in it. \V And our Lord
ſpake to Moyſes, ſaying: \V Speake to the children of Iſrael, and thou
shalt ſay to them: When you shal be entred into the land, which I wil
geue you, and shal reape your corne, you shal bring sheaues of eares,
the firſt fruites of your harueſt to the prieſt: \V who shal eleuate the
bundle before the Lord, the next day after the Sabbath, that it may be
acceptable for you, and shal ſanctifie it. \V And in the ſelf ſame day
that the sheafe is conſecrated, shal be killed a lambe without ſpotte of
a yeare old for an holocauſt of the Lord. \V And the libaments shal be
offered with it, two tenthes of floure tempered with oile, for a burnt
ſacrifice of the Lord, and a moſt ſweete odour: libaments alſo of wine,
the fourth part of an hin. \V Bread, and fried barlie, and frumentie,
you shal not
%%% 0328
eate of the corne, vntil the day that you offer therof to your God. It
is a precept for euer in your generations, and al your habitations. \V
You shal number therfore from the morow after the Sabbath, wherin you
did offer the ſheafe of the firſt fruites, ſeuen ful weekes, \V vnto the
morow after the ſeuenth weeke be expired, that is to ſay
\SNote{3.~Pentecoſt in remembrance of receiuing the law.}
fiftie daies, and ſo you ſhal offer a new ſacrifice to the Lord, \V out
of al your
%%% o-0296
habitations, two loaues of firſt fruites, of two tenthes of floure
\SNote{See
\XRef{chap.~7. v.~14.}}
leauened, the which you shal bake for the firſt fruites of the Lord. \V
And you shal offer with the breades ſeuen lambes without ſpotte of a
yeare old, and one calfe from the heard, and two rammes, and they shal
be for an holocauſt with their libamentes, for a moſt ſweete odour to
the Lord. \V You shal make a bucke goate alſo for ſinne, and two lambes
of a yeare old for hoſtes of pacifiques. \V And when the prieſt hath
eleuated them with the breades of the firſt fruites before the Lord,
they shal turne to his vſe. \V And you shal cal this day moſt ſolemne,
and moſt holie: no ſeruile worke shal you doe in it. It shal be an
euerlaſting ordinance in al your habitations, and generations. \V And
after you reape the corne of your land, you shal not cut it to the verie
ground: neither shal you gather the eares that remaine, but you shal let
them alone for the poore and for ſtrangers. I am the Lord your God. \V
And our Lord ſpake to Moyſes, ſaying: \V Speake to the children of
Iſrael: The ſeuenth moneth, on
\SNote{4.~Feaſt of trumpets, in memorie that a rãme ſticking by the
hornes was offered by Abraham in ſtead of Iſaac.}
the firſt day of the moneth, shal be a Sabbath, a memorial, by ſounding
of trumpettes, and shal be called holie: \V no ſeruile worke shal you
doe in it, and you shal offer holocauſt to the Lord. \V And our Lord
ſpake to Moyſes, ſaying: \V vpon the tenth day of this ſeuenth moneth
shal be the day of
\SNote{5.~Feaſt of Expiation in memorie of the ſinne in worſhipping the
calfe, and for al ſinnes forgotten, or vnknowne.}
expiations moſt ſolemne, and it shal be called holie: and you shal
afflict your ſoules in it, and shal offer holocauſt to the Lord. \V No
ſeruile worke shal you doe the time of this day: becauſe it is a day of
propitiation, that the Lord your God may become propitious vnto you. \V
Euerie ſoule, that is not afflicted this day, shal perish out of his
people: \V and which shal doe anie worke, the ſame wil I deſtroy out of
his people. \V No worke therfore shal you doe in it: it shal be an
euerlaſting ordinance vnto you in al your generations, and
habitations. \V It is a Sabbath of reſting, and you shal afflict your
ſoules the ninth day of the
%%% 0329
moneth: from euen vntil euen you shal celebrate your ſabbathes. \V And
our Lord ſpake to Moyſes, ſaing: \V Speake to the children of Iſrael:
From the fiftenth day of this ſeuenth moneth, shal be the feſtiuitie of
\SNote{6.~Feaſt of Tabernacles, to remẽber Gods protection in the
wildernes, where they dwelled in tabernacles 40.~yeares.}
tabernacles ſeuen daies to the Lord. \V The firſt day shal be called
moſt ſolemne and moſt holie: no ſeruile worke shal you doe in it. And
ſeuen daies you shal offer holocauſtes to the Lord. \V The eight day
alſo shal be moſt ſolemne and moſt holie, and you shal offer
holocauſtes to the Lord: for it is of
\SNote{7.~Feaſt of Aſſemblie and collection in memorie of peace geuen in
the land of promiſe.}
aſſemblie and collection: no ſeruile worke shal you doe in it. \V Theſe
are the feſtiuities of the Lord, which you shal cal moſt ſolemne and
moſt holie, and shal offer in them oblations to the Lord, holocauſts and
libaments according to the rite of euerie day: \V beſide the ſabbathes
of the Lord, and your giftes, and thoſe that you shal offer by vow, or
which you shal giue to the Lord voluntarily. \V Therfore from the
fiftenth day of the ſeuenth moneth, when you haue gathered al the
fruites of you land, you shal celebrate the feſtiuitie of
the Lord ſeuen daies, on the firſt day and the eight shal be a ſabbath,
that is reſt. \V And you shal take to you the
%%% o-0297
firſt day the fruites of a moſt faire tree, and the branches of palmes,
and boughes of the tree with thicke leaues, and willowes of the brooke,
and you shal reioice before the Lord your God. \V And you shal celebrate
the ſolemnitie therof ſeuen daies in the yeare. It shal be an
\SNote{Theſe feaſts were \Emph{euerlaſtĩg} to the Iewes \Emph{in their
generations}, that is, neuer to be altered by them, nor during their
ſtate.
\Cite{S.~Aug. q.~43. in Exod.}}
euerlaſting ordinance in your generations. The ſeuenth moneth shal you
celebrate the feſtiuitie, \V and shal dwel in bowres ſeuen daies, euerie
one, that is of the ſtocke of Iſrael, shal abide in tabernacles: \V that
your poſteritie may learne that I made the children of Iſrael, to dwel
in tabernacles, when I brought them out of the Land of Ægypt. I the Lord
your God. \V And Moyſes ſpake cõcerning the ſolemnities of our Lord to
the children of Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
