%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Ephesians
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-01


\project douay-rheims


%%% 2756
%%% o-2613
\startChapter[
  title={Chapter 1}
  ]

\Summary{He magnifieth the grace of Gods eternal predeſtination and
  temporal vocation, 11.~both of the Iewes, 13.~and alſo of the
  Epheſians being Gentils. 15.~For whoſe excellẽt faith and charitie he
  reioyceth, and continually praieth for their increaſe, that they may
  ſee more clearely the greatnes both of the inheritance in heauen, &
  alſo of God's might which helpeth them therunto: 20.~an example of
  which might they may behold in the ſupereminent exalting of Chriſt.}

Paul an Apoſtle of \Sc{Iesvs} Chriſt by the wil of God: to al the Saints
that are at Epheſus; & to the faithful in Chriſt \Sc{Iesvs}. \V Grace to
you and peace from God our Father, and our Lord \Sc{Iesvs} Chriſt.

\V Bleſſed be God and the Father of our Lord \Sc{Iesvs} Chriſt, which
hath bleſſed vs in al ſpiritual bleſſing in celeſtials, in Chriſt: \V as
he choſe vs in him before the cõſtitution of the world, that we ſhould
be holy and immaculate
\SNote{We learne here that by God's grace men be holy and immaculate, not
only in the ſight of men, not by imputation, but truely and before God:
contrarie to the Doctrine of the Caluiniſtes.}
in his ſight in charitie. \V Who hath predeſtinated vs vnto the adoption
of ſonnes, by \Sc{Iesvs} Chriſt, vnto himſelf; according to the purpoſe
of his wil: \V vnto the praiſe of the glorie of his grace, wherin he
hath
\TNote{\G{ἐχαρίτωσεν}}
gratified vs in his beloued Sonne. \V In whom we haue redemption by his
bloud (the remiſſion of ſinnes) according to the riches of his grace. \V
Which hath ſuperabounded in vs in al wiſedom and prudence, \V that he
might make knowen vnto vs the ſacrament of his wil, according to his
good pleaſure, which he purpoſed in himſelf, \V in the diſpenſation of
the fulnes of times, to
\TNote{\G{ἀνακεφαλαιώσασθαι}}
perfit al things in Chriſt, that are in heauen and in earth, in him. \V
In whom we alſo are called by lot:
%%% o-2614
predeſtinate according to the purpoſe of him that worketh al things,
according to the counſel of his wil: \V that we may be vnto the praiſe
of his glorie, which before haue hoped in Chriſt: \V in whom you alſo,
when you had heard the word of truth (the Ghoſpel of your ſaluation:) in
which alſo beleeuing you were
\SNote{Some referre this to the grace of Baptiſme: but to many learned it
ſeemeth that the Apoſtle alludeth to the giuing of the Holy Ghoſt in the
Sacrament of Confirmation, by ſigning the baptized with the ſigne of the
Croſſe & holy Chriſme. For that was the vſe in the Apoſtles time, as
els-where we haue proued.
\XRef{Annot. Act.~8.}}
ſigned with the holy Spirit of promiſe, \V which is the pledge of our
inheritance, to the redemption of acquiſition, vnto the praiſe of his
glorie.

%%% 2757
\V Therfore I alſo hearing your faith that is in our Lord \Sc{Iesvs},
and loue toward al the Saints, \V ceaſe not to giue thankes for you,
making a memorie of you in my praiers, \V that God of our Lord
\Sc{Iesvs} Chriſt, the Father of glorie, giue you the ſpirit of wiſedom
and of reuelation, in the knowledge of him, \V the eyes of your hart
illuminated, that you may know what the hope is of his vocation, and
what are the riches of the glorie of his inheritance in the Saints, \V
and what is the paſſing greatnes of his power toward vs that beleeue:
according to the operation of the might of his power, \V which he
wrought in Chriſt, raiſing him vp from the dead, and ſetting him on his
right hand in celeſtials \V aboue
\LNote{Al Principalitie.}{The
\MNote{Nine orders of Angels.}
Fathers vpon this, and other places of the old and new Teſtament, where
they find the orders of holy Angels or Spirits named, agree that there
be nine orders of them. Of which ſome be here counted and called, as we
ſee: in the
\XRef{Epiſtle to the Coloſſians,}
the order of Thrones is ſpecified, which maketh fiue: to which if we
adde theſe foure Cherubim, Seraphim, Angels, and Archangels, which are
commonly named in holy writ, in al there be nine.
\Cite{S.~Denys cal Hier. c.~7.~8.~9.}
&
\Cite{Ec. Hier. c.~1.}
\Cite{S.~Athan. li. de Communi iſſent. in fine.}
\Cite{Gregor. Moral. li.~32. c.~18.}
Therfore good Reader, make no account of
\CNote{\Cite{Cal. vpon this place.}}
Caluins and others infidelitie, which blaſphemouſly blame and condemne
the holy Doctours diligence in this point, of curioſitie and
impietie. The whole endeauour of theſe Heretikes is, to bring al into
doubt, and to corrupt euery Article of our Religion.}
al Principalitie &
\Fix{Proteſtate}{Poteſtate}{obvious typo, fixed in other}
& Power, and Domination, & euery name that is named not only in this
world, but alſo in that to come. \V And he hath
\CNote{\XRef{Pſ.~8,~8.}}
ſubdued al things vnder his feet: and hath made him
\LNote{Head.}{It
\MNote{As Chriſt is King, and yet men are Kings alſo: ſo Chriſt  is Head
of the Church, and yet man may be Head therof alſo.}
maketh a high proofe among the Proteſtants, that no man can be Head of
the Church, becauſe it is a calling and dignitie proper to Chriſt. But
in truth by as good reaſon there should be no King nor Lord, becauſe
\CNote{\XRef{Apo.~19.}}
\Emph{He is King and Lord}: neither should there be Bishop or Paſtour,
becauſe he is
\CNote{\XRef{1.~Pet.~2,~25.}}
\Emph{the Bishop and Paſtour of our ſoules}: nor Pontifex, nor Apoſtle,
for by thoſe titles S.~Paul termeth him
\XRef{Heb.~3.}
none should be piller, foundation, rock, light, or Maiſter of the Church
or truth, becauſe Chriſt is properly al theſe. And yet our new Doctours
(though they be exceeding ſeditious and would for the aduantage of their
Sect be gladly ridde of Kings and al other Superiours temporal, if they
feared not the ſword more then God, and would find as good Scriptures to
be deliuered of them, as they now find to diſcharge themſelues of
obedience to Popes.) Yet (I ſay) they wil not deny, al the former titles
and dignities (notwithſtanding Chriſtes ſoueraigne right in the ſame) to
be giuen and communicated to the Princes and Magiſtrates of the earth
both ſpiritual and temporal. Though Chriſt in a more diuine, ample,
abſolute, excellent, and tranſcendent ſort, haue al theſe things
attributed or appropriated to himſelf.
\MNote{Chriſt is Head of his Church in a farre more excellent ſort, then
any man can be.}
So then, though he be the Head of the Church, and the only Head in ſuch
ſoueraigne and Principal manner, as no earthly man or mere creature euer
is or can be, and is ioyned to the Church in a more excellent ſort of
coniunction, then any King is to his ſubiects or Countrie, or any Pope
or Prelate to the Church wherof he is Gouernour, euen ſo farre that it
is called his body Myſtical: life, motion, ſpirit, grace iſſuing downe
from him to it and the members of the ſame, as from the head to the
natural body: Though in this ſort (we ſay) no man can be Head but
Chriſt, nor the Church be body to any but to Chriſt; yet the Pope may be
the miniſterial Head, that is to ſay, the cheefe Gouernour, Paſtour, and
Prelate of the ſame, and may be his Vicar or Viceregent in the regiment
of that part which is in earth: as S.~Hierom calleth Damaſus the Pope,
\CNote{\Cite{Hier. ep.~123.}}
\L{Summum Sacerdotum}, \Emph{the cheefe and higheſt Prieſt}: and the
Apoſtle ſaith of this miniſterial Head,
\CNote{\XRef{1.~Cor.~12,~21.}}
\Emph{The head can not ſay to the feet, you are not neceſſarie for me.}
For therin alſo is a great difference between Chriſt and euery mortal
Prelate, that (as the Apoſtle here ſaith) he is Head of the whole
Church, meaning of the triumphant (& of al Angels alſo, though in
another ſort) no leſſe then of the Church militant. So Peter was not,
nor any Pope, nor any man can be. Where you muſt obſerue, that for this
ſoueraigne preeminence of Chriſt in this caſe, the Church is not called
the body myſtical of any Gouernour, Peter, Paul, or what Prelate or Pope
ſoeuer.}
Head ouer al the Church, \V which is his body, the
\SNote{Chriſt is not ful, whole, and perfect without the Church no more
then the head without the body.}
fulnes of him, which is filled al in al.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

