%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Ephesians
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 2763
%%% o-2620
\startChapter[
  title={Chapter 5}
  ]

\Summary{He continueth his exhortation to good life, 5.~aſſuring them
  againſt al deceiuers, that no committer of mortal ſinne shal be ſaued:
  conſidering that for ſuch ſinnes it is that the Heathens shal be
  damned: 8.~& that Chriſtians muſt rather be the light of al
  others. 22.~Then he commeth in particular and exhorteth husbands and
  wiues to doe their duty one towards the other, by the example of
  Chriſt and his obedient and beloued Spouſe the Church.}

%%% o-2621
Be ye therfore followers of God, as moſt deare children: \V and
\CNote{\XRef{Io.~13,~34.}}
walke in loue, as Chriſt alſo loued vs, and deliuered himſelf for vs an
oblation and hoſt to God in an odour of ſweetnes. \V But
\CNote{\XRef{Col.~3,~5.}}
fornication and al vncleannes, or auarice let it not ſo much as be named
among you as it becõmeth Saints: \V or filthines, or fooliſh talke, or
ſcurrilitie, being to no purpoſe: but rather giuing of thankes. \V For
vnderſtanding know you this that no fornicatour, or vncleane, or
couetous perſon (which is
\SNote{See the heretical corruptiõ of this place in the
\XRef{Annotatiõ Col.~3. v.~5.}}
the ſeruice of Idols) hath inheritance in the Kingdom of Chriſt and of God.

\V Let no man ſeduce you with vaine words. For, for theſe things commeth
the anger of God vpon the children of diffidence. \V Become not therfore
partakers with them. \V For you were ſometime darknes, but now light in
our Lord. Walke as children of the light, (\V for the fruit of the light
is in al goodnes, and iuſtice, and veritie) \V prouing what is wel
pleaſing to God: \V and communicate not with the vnfruitful workes of
darknes, but rather reproue them. \V For the things that are done of thẽ
in ſecret, it is ſhame euen to ſpeake. \V But al things that are
reproued, are manifeſted by the light. For al that is manifeſted, is
light. \V For the which cauſe he ſaith:
\CNote{\XRef{Eſ.~9,~60.}}
\Emph{Riſe thou that ſleepeſt, & ariſe from the dead: and Chriſt wil
illuminate thee.} \V See therfore, Brethren, how you walke warily. Not
as vnwiſe, but
\CNote{\XRef{Col.~4,~5.}}
as wiſe: \V redeeming the time, becauſe the daies are euil. \V Therfore
become not vnwiſe, but
\CNote{\XRef{Ro.~12,~2.}}
vnderſtanding what is the wil of God. \V And be not drunke with wine
wherin is riotouſnes, but be filled with the Spirit, \V ſpeaking to your
ſelues in Pſalmes & hymnes, and ſpiritual canticles, chanting and
ſinging in your harts to our Lord: \V giuing thankes alwaies for al
things, in the name of our Lord \Sc{Iesvs} Chriſt to God and the
Father. \V Subiect one to another in the feare of Chriſt.

\V Let
\CNote{\XRef{Col.~3,~18.}
\XRef{Tit.~2,~5.}
\XRef{1.~Pet.~3,~1.}}
women be ſubiect to their husbands, as to our Lord: \V becauſe
\CNote{\XRef{Gen.~3,~16.}
\XRef{1.~Cor.~11,~3.}}
the man is the Head of the woman: as Chriſt is the Head of the
\SNote{It is much to be noted, that in the firſt English Bibles there is
not once the name of \Sc{Chvrch} in al the Bible, but inſteed therof,
\Emph{Congregatiõ}. Which is ſo notorious a corruption, that thẽſelues
in the later Bibles correct it for shame, but yet ſuffer the other to be
read and vſed ſtil. See the
\Cite{Bible printed in the yeare.~1562.}}
\Sc{Chvrch}. Himſelf,
\LNote{Sauiour of his body.}{None
\MNote{No ſaluation out of the Cat. Church.}
hath ſaluation or benefit by Chriſt, that is not of his body the Church:
and what Church that is, S.~Auguſtin expreſſeth in theſe words:
\Emph{The Catholike Church only is the body of Chriſt, wherof he is
Head. Out of this body the Holy Ghoſt quickneth no man.} And a litle
after: \Emph{He that wil haue the Spirit, let him beware he remaine not
out of the} \Sc{Chvrch}. \Emph{Let him beware he enter not into it
feinedly.}
\Cite{Auguſt. epiſt.~50. ad Bonifacium Comitem in fine.}}
the Sauiour of his body. \V But as the \Sc{Chvrch} is
\LNote{Subiect to Chriſt.}{The
\MNote{The Church neuer erreth.}
\Sc{Chvrch} is alwaies ſubiect to Chriſt, that is, not only vnder him,
but euer obedient to his words and commandement. Which is an euident and
inuincible demonſtration that she neuer rebelleth againſt Chriſt, neuer
falleth from him by errour, Idolatrie, or falſe worship, as the
Heretikes now, and the Donatiſts of old did teach.}
ſubiect to
%%% o-2622
Chriſt, ſo alſo the women to their husbands in al things. \V Husbands,
loue your wiues, as Chriſt alſo
\LNote{Loued the Church.}{Loe
\MNote{Chriſt's loue toward his Church.}
Chriſtes ſingular loue of the \Sc{Chvrch}, for which only & the members
therof he effectually ſuffered his Paſſion, and for whoſe continual
cleanſing & purifying in this life, he inſtituted holy Baptiſme and
other Sacraments:
\MNote{The Church triumphant without ſpot or wrinkle.}
that at length in the next life it may become without al ſpot, wrinkle,
or blemish. For in this world by reaſon of the manifold infirmities of
diuers her members, she can not be wholy without ſinne, but muſt ſay
alwaies: \L{Dimitte nobis debita noſtra}, \Emph{Forgiue vs our debts.}
\Cite{Aug. li.~2. Retract. c.~18.}}
loued the \Sc{Chvrch}, and deliuered himſelf for it: \V that he might
%%% 2764
ſanctifie it, cleanſing it by the lauer of water in the Word, \V that he
might preſent to himſelf a glorious \Sc{Chvrch}, not hauing ſpot or
wrinkle, or any ſuch thing, but that it may be holy and vnſpotted. \V So
alſo men ought to loue their wiues as their owne bodies. He that loueth
his wife, loueth himſelf. \V For no man euer hated his owne fleſh: but
he nouriſheth and cheriſheth it,
\LNote{As Chriſt the Church.}{It
\MNote{The manifold dignitie of the Church.}
is an vnſpeakeable dignitie of the \Sc{Chvrch}, which the Apoſtle
expreſſeth oftẽ els-where, but ſpecially in this whole paſſage, to be
that creature only for which Chriſt effectually ſuffered, to be washed &
embrued with water & bloud iſſuing out of his holy ſide, to be nourished
with his owne body (for ſo doth S.~Ireneus expound
\Cite{lib.~5. in principio)}
to be his members,
\CNote{\Cite{Aug. li.~1. de Sym. and Catech. c.~9.}}
to be ſo ioyned vnto him, as the body and members of the ſame flesh,
bone and ſubſtance, to the head: to be loued and cherished of him as
wife of husband, yea to be his wife and moſt deare Spouſe, taken and
formed (as S.~Auguſtin often ſaith) out of his owne ſide vpon the
Croſſe, as Eue our firſt father Adam's ſpouſe was made of his ribbe. 
\Cite{In Pſal.~126.}
&
\Cite{in Pſal.~127.}
&
\Cite{tract.~9. in Ioan.}
&
\Cite{tract.~120.}
\MNote{The Church is the principal creature.}
In reſpect of which great dignitie and excellencie, the ſame holy Father
affirmeth the \Sc{Chvrch} to be the principal creature, and therfore
named in the Creed next after the Holy Ghoſt. And he proueth againſt the
Macedonians, the Holy Ghoſt to be God, becauſe he is named before the
Church in the confeſſion of our faith. Of which incomparable excellencie
of the Church, ſo beloued of Chriſt, and ſo inſeparately ioyned in
marriage with him, if the Heretikes of our time had any ſenſe or
conſideration, they would neither thinke their contemptible companie or
congregation to be the glorious Spouſe of our Lord,
\MNote{The Church can not erre.}
nor teach that the Church may erre, that is to ſay, may be diuorced from
her Spouſe for Idolatrie, Superſtition, Hereſie, or other abominations:
\MNote{Abſurdities that follow, if the Church may erre.}
Wherupon one of theſe abſurdities would enſue, that either Chriſt may
ſometimes be without a Church & Spouſe in earth (as he was al the while
there were no Caluiniſts, if their Church be the ſpouſe of Chriſt) or
els if the Catholike Church only is and hath been his wife, and the ſame
haue ſuch errours as the Heretikes falſely pretend, that his wife ſo
deare and ſo praiſed here, is notwithſtanding a very whoore. Which
horrible abſurdities proue and conuince to any man of common ſenſe, both
that the Catholike Church alwaies is, and that it teacheth truth
alwaies, and to honour God truely and ſincerely alwaies; whatſoeuer the
adulterous Generation of Heretikes thinke or blaſpheme.}
as alſo Chriſt the \Sc{Chvrch}: \V becauſe we be the members of his
body, of his fleſh, and of his bones. \V
\CNote{\XRef{Geneſ.~2,~24.}
\XRef{Mat.~19,~5.}}
\Emph{For this cauſe, shal man leaue his father and mother: and shal
cleaue to his wife, and they shal be two in one flesh.} \V
\LNote{This is a great Sacrament.}{Marriage
\MNote{\Sc{Matrimonie} a Sacrament.}
a great Sacrament of Chriſt and his Church prefigured in the firſt
parents. \Emph{Adam} (ſaith S.~Auguſtin
\Cite{tract.~25. in Io.)}
\Emph{who was a forme of figure of him that was to come, yea rather God
in him, gaue vs a great token of a Sacrament. For both he deſerued
ſleeping to take a wife, and of his ribbe his wife was made vnto him:
becauſe of Chriſt ſleeping on the Croſſe the Church was to be made out
of his ſide.} In another place he maketh matrimonie a Sacrament of
Chriſt and his Church: in that, that as the married man muſt forſake
father and mother and cleaue vnto his wife, ſo Chriſt as it were left
his father, exinaniting himſelf by his incarnation, and left the
Synagogue his mother & ioyned himſelf to the Church.
\Cite{li.~12. c.~8. cont. Fauſt.}
In diuers other places he maketh it alſo a Sacrament, ſpecially in that
it is an inſeparable bond betwixt two, & that can neuer be diſſolued but
by death: ſignifying Chriſt's perpetual and indiſſoluble coniunction
with the Church his one only Spouſe.
\Cite{de Gen. ad lit. li.~9. c.~7. cont. Pelag.}
\Cite{de pec. ori. li.~2. c.~34.}
\Cite{De fid. & op. c.~7.}
\Cite{De bono coniug. c.~7. &~18.}
And in another place, \Emph{The good of Marriage} (ſaith he) \Emph{among
the people of God is in the holines of a Sacrament.}
\Cite{De bono coniugali. c.~24.}

Who would haue thought ſuch myſteries & Sacraments to be in Marriage,
that the ioyning of man & wife together should repreſent ſo great a
myſterie, if the Apoſtle himſelf, & after him this holy Father and
others, had not noted it? or who can maruel that the holy Church taketh
this to be a Sacrament,
\MNote{The grace giuen by this Sacrament.}
and to giue grace of ſanctification to the
parties married, that they may liue together in mutual fidelitie, bring
vp their children in faith and feare of God, and poſſeſſe their veſſel
(as the Apoſtle
\CNote{\XRef{1.~Theſ.~4.}}
ſpeaketh) in ſanctification and honour,
\MNote{The Proteſtãts fleshly eſtimation of marriage.}
and not in paſſion of luſt and
ignominie, as the Heathen doe which know not God, and as our brutish new
Maiſters ſeeme to doe that commend marriage aboue al things ſo farre as
it feedeth their concupiſcences, but for grace, Sacrament, myſterie, or
ſanctification thereby, they care no more then the Heathen or brute
beaſts doe? And thus we gather that Matrimonie is a Sacrament, and not
of the Greek word
\MNote{\G{μυϛήριον}.}
\Emph{Myſterie} only as Caluin falſely ſaith, nor of the Latin word
\MNote{\L{Sacramentum.}}
\Emph{Sacrament}, both which we know haue of their nature a more general
ſignification, and that in the Scriptures alſo: but whereas theſe names
are here giuen to Matrimonie by
the Apoſtle, & are not giuen in the Scriptures to
\Fix{Biptiſme}{Baptiſme}{obvious typo, fixed in other}
and the Euchariſt, let them tel vs why they alſo apply theſe words from
their general ſignification to ſignifie ſpecially & peculiarly thoſe two
Sacraments neuer ſo named expreſly in Scripture, and doe not likewiſe
follow the Catholike Church in calling matrimonie by the ſame name,
which is here ſo called of the Apoſtle, ſpecially whereas the
ſignification in it, is as great as in any other of the Sacraments, and
rather greater.}
This is a great Sacrament. But I ſpeake in Chriſt & in
the \Sc{Chvrch}. \V Neuertheles you alſo euery one, let each loue his
wife as himſelf: and let the wife feare her husband.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

