%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Ephesians
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 2758
%%% o-2615
\startChapter[
  title={Chapter 2}
  ]

\Summary{He putteth them in mind of their vnworthines before they were
  Chriſtians; that al the praiſe may be giuen to the grace of God:
  11.~and of the enmitie that was then between the Iew and the Gentil,
  15.~vntil now that Chriſt by his Croſſe hath made both one, taking
  away the ceremonies of the Law, and making one body, and building one
  holy Temple of al in his Catholike Church.}

And you when you were dead by your offenſes and ſinnes, \V wherin
ſometime you walked according to the courſe of this world, according to
the
\CNote{\XRef{Eph.~6,~22.}}
Prince of the power of this aire, of the ſpirit that now worketh on the
children of diffidence, \V in whom alſo we al conuerſed ſometime in the
deſires of our fleſh, doing the wil of the fleſh and of thoughts, and
were by nature the children of wrath as alſo the reſt: \V but God (which
is rich in mercie) for his exceeding charitie wherwith he loued vs, \V
euen when we were dead by ſinnes, quickned vs together in Chriſt, (by
whoſe grace you are ſaued,) \V and raiſed vs vp with him, and hath made
vs ſit with him in the celeſtials in Chriſt \Sc{Iesvs}, \V that he might
ſhew in the worlds ſucceeding, the aboundant riches of his grace, in
bountie vpon vs in Chriſt \Sc{Iesvs}. \V For by
\LNote{By grace you are ſaued through faith.}{Our
\MNote{Our firſt iuſtificatiõ of mere grace, & faith the foundation
therof.}
firſt iuſtification is of God's grace, and not of our deſeruing: becauſe
none of al our actions that were before our iuſtification, could merit
or iuſtly procure the grace of iuſtification. Againe, he
ſaith, \Emph{through faith}: for that faith is the beginning,
foundation, and root of al iuſtification, and the firſt of al other
vertues, without which it is impoſſible to pleaſe God.}
grace you are ſaued through faith (and that not of your ſelues,
%%% o-2616
for it is the guift of God) \V
\SNote{It is ſaid, not of workes, as thine, of thy-ſelf being vnto thee,
but as thoſe in which God hath made, formed, and created thee.
\Cite{Aug. de gr. & lib. arbit. c.~8. &~ſeq.}}
not of workes, that no man glorie. \V For we are his worke, created in
Chriſt \Sc{Iesvs} in good workes, which God hath prepared that we should
walke in them.

\V For the which cauſe be mindful that ſometime you were Gentils in the
fleſh, who were
\CNote{\XRef{1.~Reg.~17,~26.}
\XRef{Ezec.~44,~7.}}
called prepuce of that which is called circumciſion in the flesh, made
with hand: \V who were at that time without Chriſt alienated from the
conuerſation of Iſrael, and
\CNote{\XRef{Ro.~9,~4.}}
ſtrangers of the Teſtaments, hauing no hope of the promiſe, and without
God in this world. \V But now in Chriſt \Sc{Iesvs}, you that ſometime
were farre off, are made nigh in the bloud of Chriſt. \V For he is our
peace, who hath made both one, and diſſoluing the middle wal of the
partition, the enmities in his flesh: \V euacuating the law of
commandements
\TNote{\G{ἐν δόγμασι}}
in decrees: that he may create the two in himſelf into one new man,
making peace, \V and may reconcile both in one body to God by the
croſſe, killing the enmities in himſelf.

%%% 2759
\V And comming he euangelized peace to you that were farre off, and
peace to them that were nigh. \V For
\CNote{\XRef{Ro.~5,~2.}}
by him we haue acceſſe both in one Spirit to the Father. \V Now then you
are not ſtrangers and forreiners: but you are
\TNote{\G{συμπολῖται}}
citizens of the Saints, and the domeſticals of God, \V
\LNote{Built vpon the fundation.}{Note
\MNote{The Church builded vpon Chriſt, and yet vpon the Apoſtles alſo.}
againſt the Heretikes that thinke it diſhonourable to Chriſt, to
attribute his titles or callings to mortal men, that the faithful
(though builded firſt, principally, and properly vpon Chriſt) yet are
ſaid here to be built alſo vpon the Apoſtles and Prophets. Why may not
the Church then be builded vpon Peter?}
built vpon the foundation of the Apoſtles and Prophets, \Sc{Iesvs}
Chriſt himſelf being the higheſt corner-ſtone: \V in whom al building
framed together, groweth into an holy Temple in our Lord, \V in whom you
alſo are built together into an habitation of God in the holy Ghoſt.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

