%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Ephesians
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 2759
%%% o-2617
\startChapter[
  title={Chapter 3}
  ]

\Summary{For witneſsing the vocation of the Gentils, as being the
  Apoſtle of the Gentils, he is in priſon: 13.~wherin the Gentils
  therfore haue cauſe to reioyce, rather then to shrinke. So he ſaith,
  14.~and alſo praieth to God (who is Almightie) to confirme their
  inward man, though the outward be infirmed by perſecution.}

For this cauſe, I Paul the priſoner of \Sc{Iesvs} Chriſt, for you
Gentils: \V if yet you haue heard the diſpenſation of the grace of God,
which is giuen me toward you, \V becauſe according to reuelation the
ſacrament was made knowen to me, as I haue writen before in briefe:
\V according as you reading may vnderſtand my wiſedom in the myſterie of
Chriſt, \V which vnto other Generations was not knowen to the ſonnes of
men, as now it is reuealed to his holy Apoſtles, and Prophets in the
Spirit. \V The Gentils to be coheires and concorporate and
comparticipant of his promiſe in Chriſt \Sc{Iesvs} by the Ghoſpel: \V
wherof I am made a Miniſter according to the guift of the grace of God,
which is giuẽ me according to the operatiõ of his power. \V To me
\CNote{\XRef{1.~Cor.~15,~9.}}
the leaſt of al the Saints is giuẽ this grace, amõg the Gẽtils to
euangelize the vnſearcheable riches of Chriſt, \V & to illuminate al men
%%% !!! placed wrong in this
what is the diſpenſation of the ſacrament hidden
\SNote{That is, \Emph{for euer before}.}
\TNote{\G{ἀπὸ τῶν αἰώνων}}
from worlds in God,
%%% 2760
who created al things: \V that the manifold wiſedom of God, may be
notified to the Princes & Poteſtates in the Celeſtials by the Church, \V
according to the 
%%% !!! TNote in both?
\SNote{That is \Emph{the eternal
\Fix{præfinitnio}{præfinition}{obvious typo, fixed in other}}}
prefinition of worlds, which he made in Chriſt \Sc{Iesvs} our Lord. \V
In whom we haue affiance and acceſſe in confidence by the faith of
him. \V For the which cauſe I deſire that you faint not in my
tribulations for you, which is your glorie.

\V For this cauſe I bow my knees to the Father of our Lord \Sc{Iesvs}
Chriſt, \V of whom al paternitie in the Heauens and in earth is
named, \V that he giue you according to the riches of his glorie, power
to be fortified by his Spirit in the inner man. \V Chriſt
\SNote{Chriſt dwelleth in vs by his guifts, and we be iuſt by thoſe
his guifts remaining and reſident in vs, & not by Chriſtes proper iuſtice
only, as the Heretikes affirme.}
to dwel by faith in your harts rooted and founded
\SNote{Not faith only muſt be in vs, but charitie which accomplisheth al
vertues.}
in charitie, \V that you may be able to comprehẽd with al the Saints,
what is the bredth, and length, and height, and depth, \V to know alſo
the charitie of
%%% o-2618
Chriſt, ſurpaſſing knowledge, that you may be filled vnto al the fulnes
of God. \V And
\CNote{\XRef{Ro.~16,~23.}}
to him that is able to doe al things more aboundantly then we deſire or
vnderſtand, according to the power that worketh in vs: \V to him be
glorie in the \Sc{Chvrch}, and in Chriſt \Sc{Iesvs}, vnto al Generations
world without end. Amen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

