%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Ephesians
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 2765
%%% o-2623
\startChapter[
  title={Chapter 6}
  ]

\Summary{Likewiſe children and parents he exhorteth, 5.~item ſeruants
  and maiſters. 10.~Then that al take courage in the might of God, but
  ſo, that withal they arme themſelues (conſidering what mightie enemies
  they haue) with al peeces of ſpiritual armour, 18.~praying alwaies
  feruently, and for him alſo.}

Children, obey your parents in our Lord. For this is iuſt. \V
\CNote{\XRef{Col.~3,~20.}
\XRef{Exo.~20,~12.}
\XRef{Deut.~5,~6.}}
\Emph{Honour thy father & thy mother} (which is the firſt cõmandment in
the promiſe) \V \Emph{that it may be vvel with thee & thou maieſt be
long-liued vpon the earth.} \V And you fathers, prouoke not your children
to anger: but bring them vp in the diſcipline & correption of our Lord.

%%% 2766
\V
\CNote{\XRef{Col.~3,~22.}
\XRef{Tit.~2,~9.}
\XRef{1.~Pet.~2,~8.}}
Seruants, be obedient to your Lords according to
%%% o-2624
the fleſh, with feare and trembling, in the ſimplicitie of your hart as
to Chriſt: \V not ſeruing to the eye, as it were pleaſing men, but as
the ſeruants of Chriſt doing the wil of God from the hart, \V with a
good wil ſeruing, as to our Lord and not to men. \V Knowing that euery
one what good ſoeuer he ſhal doe, that ſhal he
\SNote{God leaueth no good worke vnrewarded.}
receiue of our Lord, whether he be bond, or free. \V And you Maiſters,
doe the ſame things to them,
\TNote{\G{ἀνιέντες}}
remitting threatnings: knowing that both their Lord and yours is in
Heauen: and
\CNote{\XRef{Deu.~10,~17.}}
acceptation of Perſons is not with him.

\V Hence-forth, Brethren, be ſtrengthned in our Lord, and in the might
of his power. \V Put you on the
\TNote{\G{πανοπλίαν}}
armour of God, that you may ſtãd againſt the deceits of the Diuel. \V
For our wreſtling is not againſt fleſh and bloud: but againſt Princes
and Poteſtates, againſt the
\CNote{\XRef{Eph.~2,~2.}}
Rectours of the world of this darknes, againſt the ſpirituals of
wickednes in the celeſtials. \V Therfore take the armour of God, that
you may reſiſt in the euil day, and ſtand in al things perfect. \V Stand
therfore hauing your loines girded in truth, and
\SNote{If man could not be truely iuſt or haue iuſtice in himſelf, how
could he be clothed with iuſtice?}
clothed with the breaſt-plate of iuſtice, \V & hauing your feet ſhod
to the preparation of the Ghoſpel of peace: \V in al things taking the
ſhield of faith, wherwith you may extinguiſh al the firie darts of the
moſt wicked one. \V And take vnto you the
\CNote{\XRef{Eſ.~59,~17.}
\XRef{1.~Theſ.~5,~8.}}
helmet of ſaluation: and the ſword of the ſpirit (which is the word of
God) \V in al praier and ſupplication praying at al time in ſpirit: and
in the ſame watching in al inſtance and ſupplication for al the
Saints: \V and for me, that ſpeach may be giuen me in the opening of my
mouth with confidence, to make knowen the myſterie of the Ghoſpel, \V
for the which I am a Legate in this chaine, ſo that in it I may be bold
according as I ought, to ſpeake.

\V And that you alſo may know the things about me, what I doe, Tychicus
my deareſt Brother and faithful Miniſter in our Lord, wil make you
vnderſtand al things: \V whom I haue ſent to you for this ſame purpoſe,
that you may know the things about vs, and he may comfort your harts. \V
Peace to the Brethren and
\SNote{S.~Auguſtin noteth in ſundrie places vpon this ſame text, that
faith without charitie ſerueth not to ſaluatiõ.
\Cite{li.~50. hom.~7.}}
charitie with faith from God the Father, and our Lord \Sc{Iesvs}
Chriſt. \V Grace with al that loue our Lord \Sc{Iesvs} Chriſt in
incorruption. Amen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

