%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 2955
%%% o-2816
\startChapter[
  title={Chapter 10}
  ]

\Summary{Another ſtrong Angel crying out, 3.~ſeuen thunders doe
  ſpeake. 6.~The Angel ſweareth that there shal be time no more, but at
  the voice of the ſeuenth Angel the myſterie shal be fully
  accomplished. 9.~He giueth Iohn a book to deuoure.}

%%% o-2817
And
\SNote{\Sc{Christ} the valiant Angel is here deſcribed.}
I ſaw another Angel, ſtrong, deſcending from heauen, clothed with a
cloud, and a raine-bow on his head, and his face was as the ſunne, and
his feet as
\Var{a piller}{pillers}
of fire. \V And he had in his hand a litle booke opened: and he put his
right foot vpon the ſea, and his left vpon the land. \V And he cried
with a loud voice, as when a Lion roareth. And when he had cried, the
ſeuen thunders ſpake their voices. \V And when the ſeuen thunders had
ſpoken their voices, I was about to write: and I heard a voice from
heauen ſaying to me: Signe the things which the ſeuen thunders haue
ſpoken; and
\SNote{Many great myſteries and truths are to be preſerued in the
Church, which for cauſes knowen to Gods prouidence are not to be written
in the booke of holy Scripture.}
write them not.

\V And
\CNote{\XRef{Dan.~12,~7.}}
the Angel which I ſaw ſtanding vpon the ſea and vpon the land,
\SNote{This was the manner of taking an oth by the true God, as
\XRef{Deut.~32.}}
lifted vp his hand to heauen, \V and he ſware by him that liueth for
euer and euer, that created heauen and thoſe things which are in it: and
the earth, and thoſe things which are in it: and the ſea, and thoſe
things which are in it: That there ſhal be time no more: \V but in the
daies of the voice of the ſeuenth Angel, when the trompet ſhal begin to
ſound, the myſterie of God ſhal be conſummate, as he hath euangelized by
his ſeruants the Prophets.

\V And I heard a voice from heauen againe ſpeaking with me, and ſaying:
Goe, and take the booke that is opened, of the hand of the Angel
ſtanding vpon the ſea and vpon the land. \V And I went to the Angel,
ſaying vnto him, that he should giue me the booke. And he ſaid to me:
\CNote{\XRef{Ezec.~3,~1.}}
Take the booke, and
\SNote{By earneſt ſtudie and meditation.}
deuoure it: and it ſhal make thy belly to be bitter, but in thy mouth it
ſhal be
\SNote{Sweet in the reading, but in fulfilling, ſome-what bitter,
becauſe it commandeth workes of penance and ſuffering of tribulations.}
ſweet as it were honie. \V And I tooke the booke of the hand of the
Angel, and deuoured it: and it was in my mouth as it were honie,
ſweet. And when I had deuoured it, my bellie was made bitter, \V and he
ſaid to me: Thou muſt againe prophecie to Nations, and peoples, and
tongues, and many Kings.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

