%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 2963
%%% o-2824
\startChapter[
  title={Chapter 14}
  ]

\Summary{1.~Virgins follow the Lamb whitherſoeuer, ſinging a new
  canticle. 6.~One Angel euangelizeth the Ghoſpel: 8.~another Angel
  telleth the fal of Babylon: 9.~the third declareth their torments that
  haue adored the beaſt. Moreouer two hauing ſickles, 15.~one of them is
  commanded to reap downe the corne, 18.~the other to gather the grapes
  as in vintage, which are troden in the lake of Gods wrath.}

And I looked, & behold
\SNote{Chriſt, and the ſame number of elect that were ſigned
\XRef{chap.~7.}}
a Lamb ſtood vpõ mount Sion, & with him an hundred fourtie foure
thouſand hauing his name, and the name of his Father written in their
foreheads. \V And I heard a voice from heauen, as the voice of many
waters, and as the voice of great thunder: and the voice which I heard,
as of harps harping on their harps. \V And they ſang as it were a new
ſong before the ſeat and before the foure beaſts, and Seniours, & no
%%% o-2825
man could
\TNote{learne, \G{μαθεῖν}}
ſay the ſong, but thoſe hundred fourtie foure thouſand, that were
bought from the earth. \V Theſe are they which were not defiled with
women. For they are
\SNote{One ſtate of life more excellent then another. And virgins for
their puritie paſſing the reſt, & alwaies accompanying Chriſt according
to the Churches hymnes out of this place, \L{Quocunque pergis, virgines
ſequunter &c.}}
virgins. Theſe follow the Lamb whitherſoeuer he ſhal goe. Theſe were
bought from among men,
\SNote{This the Church applieth to the holy Innocents that died firſt
for Chriſt.}
the firſt fruits to God and the Lamb: \V and in their mouth there was
found no lie. For they are without ſpot before the throne of God.

\V And I ſaw another Angel flying through the middes of heauen, hauing
the eternal Ghoſpel, to euangelize vnto them that ſit vpon the earth,
and vpon euery Nation, and Tribe, and tongue, and people; \V ſaying
with a loud voice: Feare our Lord, and giue him honour, becauſe the
houre of his iudgement is come: and adore ye him
\CNote{\XRef{Pſal.~145.}
\XRef{Act.~14.}}
that made heauen and earth, the ſea and al things that are in them, and
the fountaines of waters.

\V And another Angel followed, ſaying:
\CNote{\XRef{Eſa.~21.}
\XRef{Ier.~51.}
\XRef{Apo.~18.}}
Fallen fallen is that great
\SNote{The citie of the diuel, which is the vniuerſal ſocietie of the
wicked miſbelieuers and il liuers in the world.}
Babylon, which of the wine of the wrath of her fornication made al
Nations to drinke.

\V And the third Angel followed them, ſaying with a loud voice: If any
man adore the beaſt and his image, and receiue the character in his
forehead, or in his hand; \V
\SNote{The great damnatiõ that shal follow thẽ that forſake Chriſt and
the Church & worship Antichriſt or his image.}
he alſo ſhal drinke of the wine of the wrath of God, which is mingled
with pure wine in the cup of his wrath, and ſhal be tormented with fire
& brimſtone in the ſight of the holy Angels and before the ſight of the
Lamb. \V And the ſmoke of their torments ſhal aſcend for euer and euer:
neither haue they reſt day and night which haue adored the beaſt, and
his image, and if any man take the character of his name. \V Here is the
patience of Saints, which
\SNote{Faith is not enough to ſaluatiõ, without fulfilling of Gods
cõmandements.}
keep the commandements of God and the faith of \Sc{Iesvs}.

\V And I heard a voice from heauen, ſaying to me: Write, Bleſſed are the
head which die in our Lord.
%%% !!! LNote marked only in other
\LNote{From hence-forth now.}{This being ſpecially ſpoken of Martyrs (as
not only S.~Auguſtin ſeemeth to take it, but the
\CNote{Beza.}
Caluiniſts themſelues, tranſlating, \L{in domino}, \Emph{for our Lords
cauſe}) the Proteſtants haue no reaſon to vſe the place againſt
Purgatorie or praier for the departed: ſeeing the Catholike Church and
al her children confeſſe, that al Martyrs are ſtraight after their
death, in bliſſe, and need no praiers. Whereof this is S.~Auguſtines
knowen ſentence:
\MNote{Praying for the dead, and vnto Saints, at the altar.}
\Emph{He doeth iniurie to the Martyr, that praieth for the Martyr.}
\Cite{Ser.~17. de verb. Apoſt. c.~1.}
and againe to this purpoſe he writeth thus moſt excellently
\Cite{tract.~84. in Ioan.}
\Emph{We keep not a memorie of Martyrs at our Lords table, as we doe of
other that reſt in peace, that is, for the intent to pray for them, but
rather that they may pray for vs &c.}

But
\MNote{The place abuſed againſt Purgatorie, anſwered.}
if we take the words generally for al deceaſed in ſtate of grace, as it
may be alſo, then we ſay that euen ſuch, though they be in Purgatorie
and Gods chaſtiſement in the next life, & need our praiers, yet
(according to the foreſaid wordes of S.~Auguſtin) \Emph{doe reſt in
peace}, being diſcharged from the labours, afflictions, and perſecutions
of this world, and (which is more) from the daily dangers of ſinne and
damnation, and put into infallible ſecuritie of eternal ioy with
vnſpeakable comfort of conſcience. And ſuch indeed are more happie &
bleſſed then any liuing, who yet are vſually in the Scriptures called
bleſſed, euen in the middes of the tribulations of this life. Whereby we
ſee that theſe wordes, \Emph{from hence forth they shal reſt from their
labours}, may truely agree to them alſo that are in Purgatorie, and ſo
here is nothing proued againſt Purgatorie. Laſtly, this
aduerb, \L{Amodo} in Latin, as in the Greek \G{ἀπ' ἄρτι} doth not
properly ſignifie, from this preſent time foreward, as though the
Apoſtle had ſaid, that after their death and ſo forward they are happie:
\TNote{\G{ἀπ' ἄρτι} \L{Amodo}
\Cite{Photius in Lexico.}}
but it noteth and ioyneth the time paſt together with the time preſent,
in this ſenſe, that ſuch as haue died ſince Chriſts Aſcenſion, when he
firſt entring into heauen opened it for others, goe not to \L{Limbus
Patrum}, as they were wont before Chriſts time, but are in caſe to goe
ſtraight to bliſſe, except the impediment be in themſelues. Therfore
they are here called bleſſed, that die now in this ſtate of grace and of
the new Teſtament, in compariſon of the old faithful and good perſons.}
From hence-forth now, ſaith the
%%% 2964
Spirit, that they reſt from their labours. For their workes follow them.

\V And I ſaw, and behold a white cloud: and vpon the cloud one ſitting
like to the Sonne of man, hauing on his head a crowne of gold, and in
his hand a ſharp ſickle.

\V And another Angel came forth frõ the temple, crying with a loud voice
to him that ſate vpon the cloud:
\CNote{\XRef{Ioel.~3.}
\XRef{Mat.~13.}}
Thruſt in thy ſickle, and reape, becauſe the houre is come to reape,
%%% o-2826
for the harueſt of the earth is drie. \V And he that ſate vpon the
cloud, thruſt his ſickle into the earth, and the earth was reaped. \V
And another Angel came forth from the temple which is in heauen, himſelf
alſo hauing a sharp ſickle. \V And another Angel came forth from the
altar, which had power ouer the fire: and he cried with a loud voice to
him that had the sharp ſickle, ſaying: Thruſt in thy sharp ſickle, and
gather the cluſters of the vineyard of the earth: becauſe the grapes
thereof be ripe. \V And the Angel thruſt his sharp ſickle into the
earth, and gathered the vineyard of the earth, & caſt it into the great
\TNote{\G{ληνὸν} \L{lacum} \Emph{fat, trough, lake}.}
preſſe of the wrath of God: \V and the preſſe was troden without the
citie, and bloud came forth out of the preſſe, vp to the horſe bridles,
for a thouſand ſixe hundred furlongs.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

