%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-22


\project douay-rheims


%%% 2981
%%% o-2843
\startChapter[
  title={Chapter 22}
  ]

\Summary{The tree of life being watered with liuing water, yealdeth
  fruits euery moneth. 3.~There is neither curſe nor night in the
  citie. 9.~The Angel that shewed Iohn al theſe things, refuſeth to be
  adored of him. 14.~He telleth him that the iuſt shal enter into the
  citie, but the reſt shal be caſt forth. 18.~Laſtly, he proteſteth and
  threatneth againſt them that shal preſume to adde to this prophecie,
  or take away from the ſame.}

And he ſhewed me a riuer of
\Var{liuing water,}{water of life,}
cleere as cryſtal, proceeding from the ſeat of God and of the Lamb. \V
In the middes of the ſtreat thereof, & on both ſides of the riuer,
\SNote{Chriſt is our tree of life; in the Church, by the B.~Sacrament; &
in heauen, by his viſible preſence and influence of life euerlaſting
both to our bodies & ſoules: of whõ Salomon ſaith, \Emph{The tree of
life to al that apprehend him.}
\XRef{Prou.~3.}}
the tree of life, yealding twelue fruits, rendring his fruit euery
moneth, & the leaues of the tree for the curing of the Gentils. \V And
no curſe ſhal be any more: & the ſeat of God & of the Lamb ſhal be in
it, & his ſeruants ſhal ſerue him. \V And they ſhal ſee his face: and
his name in their foreheads. \V And
\CNote{\XRef{Apo.~21.}
\XRef{Eſa.~60.}}
night ſhal be
%%% o-2844
no more: and they ſhal not need the light of lamp, nor the light of the
ſunne, becauſe our Lord God doth illuminate them, and they ſhal reigne
for euer and euer.

\V
\MNote{\Sc{The Conclvsion}.}
And he ſaid to me: Theſe words are moſt faithful & true. And our Lord
the God of the ſpirits of the Prophets, ſent his Angel to ſhew his
ſeruants thoſe things which muſt be done quickly. \V And behold I come
quickly. Bleſſed is he that keepeth the words of the prophecie of this
booke.

\V And I Iohn which haue heard, and ſeen theſe things. And
\CNote{\XRef{Apo.~19,~10.}}
after I had heard and ſeen, I fel downe
\SNote{You ſee it is al one to adore before the feet of the Angel, & to
adore the Angel: though, to adore him, be not expreſſed as in the
\XRef{19.~chap.}
See the
\XRef{annotatiõ there v.~10.}}
to adore before the feet of the Angel which ſhewed me theſe things; \V
and he ſaid to me: See thou doe not, for I am thy fellow-ſeruant,
and of thy brethren the Prophets, and of them that keep the words of
this booke. Adore God. \V And he ſaith to me: Seale not the words of the
prophecie of this booke. For the time is neer. \V
\LNote{He that hurteth.}{It is not an exhortation, but a commination or
threatning, that how farre ſoeuer the wicked increaſe in naughtines, God
hath prouided anſwerable punishment for them.}
He that hurteth, let him hurt yet: and he that is in filth, let him be
filthie yet: and he that is iuſt,
\SNote{Man by Gods grace & doing good workes, doth increaſe his
iuſtice.}
let him be iuſtified yet: and let the holy be ſanctified yet. \V Behold
I come quickly. And my reward is with me,
\CNote{\XRef{Ro.~2,~6.}}
to render to euery man
\SNote{Heauen is the reward, hire, & repaiment for good workes, in al
the Scriptures, yet the aduerſaries wil not ſee it.}
according to his workes. \V I am
\CNote{\XRef{Apo.~1,~8.}
\XRef{21,~6.}}
Alpha and Omega, the firſt and the laſt, the beginning and the end. \V
Bleſſed are they that waſh their ſtoles: that their power may be in the
tree of life, and they may enter by the gates into the citie. \V Without
are dogges and ſorcerers, and the vnchaſt, and murderers, and ſeruers of
Idols, & euery one that loueth and maketh a lie.

\V I \Sc{Iesvs} haue ſent mine Angel, to teſtifie to you theſe things in
the Churches. I am the root and ſtocke of Dauid, the bright and morning
ſtarre. \V And the Spirit & the bride ſay, Come. And he that heareth,
let him ſay, Come. And
\CNote{\XRef{Eſ.~55,~1.}}
he that thirſteth, let him come: and he that wil, let him take the water
of life, gratis.

\V For I teſtifie to euery one hearing the words of the prophecie
%%% 2982
of this booke,
\LNote{If any man shal adde.}{The
\MNote{The curſe againſt adding or diminishing is againſt Heretikes, not
Catholike expoſitours.}
Authour of the
\Cite{commentaries vpon this booke, bearing the name of S.~Ambroſe,}
ſaith thus of this point. \Emph{He maketh not this proteſtation againſt
the expoſitours of his prophecie, but againſt Heretikes. For the
expoſitour doth adde or diminish nothing, but openeth the obſcuritie of
the narration, or sheweth the moral or ſpiritual ſenſe. He curſeth
therfore Heretikes, that vſed to adde ſome-what of their owne that was
falſe, and to take away other things that were contrarie to their
hereſies.} So ſaith this ancient Writer. And this was the propertie of
them in al Ages, & ſo is it of ours now, as we haue noted through the
whole Bible, & as we haue in ſundrie places ſet forth to the ſight of al
indifferent Readers, in the new Teſtament: that al the world may ſee
that the Apoſtles curſe is fallen vpon them, and may beware of them.}
If any man ſhal adde to theſe things, God ſhal adde vpon him the plagues
written in this booke. \V And if any man ſhal diminiſh of the words of
the booke of this prophecie: God ſhal take away his part out of the
booke of life, and out of the holy citie, and of theſe things that be
written in this booke, \V ſaith he that giueth teſtimonie of theſe
things. Yea I come quickely: Amen.
\LNote{Come Lord Ieſus.}{And
\MNote{A breefe petition vnto \Sc{Iesvs} Chriſt, to come quickly, as
S.~Iohn here ſpeaketh, and to iudge the cauſe of Catholikes &
Proteſtants.}
now ô Lord Chriſt, moſt iuſt and merciful, we thy poore creatures that
are ſo afflicted for confeſſion and defenſe of the holy, Catholike, and
Apoſtolike truth, conteined in this thy ſacred booke, and in the
infallible doctrine of thy deare ſpouſe our mother the Church, we crie
alſo vnto thy Maieſtie with tenderneſſe of our harts vnſpeakable:
\Sc{Come Lord Iesvs
\Fix{Qvikcly,}{Quickly}{obvious typo, fixed in other}}
and iudge betwixt vs and our Aduerſaries, and in the meane time giue
patience, comfort, and conſtancie to al that ſuffer for thy name, and
truſt in thee, ô Lord God our only helper and protectour, tarie not
long. \Sc{Amen}.}
Come
%%% o-2845
Lord \Sc{Iesvs}. \V The grace of our Lord \Sc{Iesvs} Chriſt be with
you al. Amen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

