%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 2958
%%% o-2819
\startChapter[
  title={Chapter 12}
  ]

\Summary{4.~The
\MNote{\Sc{The 3.~Part}.

The Dragons incredulous & perſecuting multitude, and Antichriſt the
cheefe head thereof.}
great dragon (the Diuel) watching the woman that brought
  forth a man child, to deuoure it, God tooke away the child to himſelf,
  and fed the woman in the deſert. 7.~Michael fighting with the dragon
  ouercommeth him. 13.~Who being throwen downe to earth, perſecuteth the
  woman and her ſeed.}

%%% o-2820
And a great ſigne appeared in heauen,
\SNote{This is properly and principally ſpoken of the Church: and by
alluſiõ of our B.~Lady alſo.}
a woman clothed with the ſunne, and the moone vnder her feet, & on her
head a crowne of twelue ſtarres: \V and being with child, ſhe cried alſo
traueling, and is in anguish to be deliuered. \V And there was ſeen
another ſigne in heauen, and behold
\MNote{The great Diuel Lucifer.}
a great red dragon hauing ſeuen heads, & ten hornes: and on his heads
ſeuen diadems, \V and his taile drew
\SNote{The ſpirits that fal from their firſt ſtate into Apoſtaſie with
him and by his meanes.}
the third part of the ſtarres of heauen, and caſt them to the earth: and
the dragon ſtood before the woman which was ready to be
deliuered; that when she should be deliuered, he might
\SNote{The Diuels endeauour againſt the Churches children, and ſpecially
our B.~Ladies only Sonne the head of the reſt.}
deuoure her ſonne. \V And ſhe brought forth a man child, who was
\CNote{\XRef{Pſ.~2,~9.}
\XRef{Apoc.~2,~27.}}
to gouerne al Nations in an yron rodde; & her ſonne was taken vp to God
and to his throne, \V &
\LNote{The woman fled.}{This
\MNote{The Church shal fly as to a deſert in Antichriſts time, but not
decay or be vnknowen, no not for ſo short a time.}
great perſecution that the Church shal fly from, is in the time of
Antichriſt, and shal endure but three yeares and a halfe, as is noted
\XRef{v.~14. in the margent.}
In which time for al that, she shal not want our Lordes protection, nor
true Paſtours, nor be ſo ſecret, but al faithful men shal know and
follow her: much leſſe shal she decay, erre in faith, or degenerate and
follow Antichriſt, as Heretikes doe wickedly feine. As the Church
Catholike now in England in this time of perſecution, becauſe it hath
not publike ſtate of regiment nor open free exerciſe of holy functions,
may be ſaid to be fled into the deſert, yet it is neither vnknowen to
the faithful that follow it, nor the enemies that perſecute it: as the
hid company that the proteſtants talke of, was for ſome worlds
together, neither knowen to their freinds nor foes, becauſe there was
indeed none ſuch for many Ages together. And this is true, if we take
this flight for a very corporal retiring into wildernes. Where indeed it may
be, and is of moſt expounded, to be a ſpiritual flight, by forſaking the
ioyes and ſolaces of the world, & giuing herſelf to contemplation and
penance, during the time of perſecution vnder Antichriſt. And by
enlarging the ſenſe, it may alſo very wel ſignifie the deſolation and
affliction that the Church ſuffereth and hath ſuffered from time to time
in this wildernes of the world, by al the fore-runners and Miniſters of
Antichriſt, Tyrants and Heretikes.}
the woman fled into the wilderneſſe where ſhe had a place prepared of
God, that there they might feed her a thouſand two hundred ſixtie daies.

\V And there was made
\LNote{A great battel.}{In
\MNote{S.~Michael fighting with the dragon.}
the Church there is a perpetual combat betwixt S.~Michael (Protectour of
the Church militant as he was ſometime of the Iewes Synagogue
\XRef{Dan.~10,~21.)}
and his Angels, and the Diuel and his Miniſters. The perfect victorie
ouer whom, shal be at the iudgement. Marke here alſo the cauſe why
S.~Michael is commonly painted fighting with a dragon.}
a great battel in heauen, Michael and his Angels fought with the dragon,
and the dragon fought and his Angels: \V and they preuailed not, neither
was their place found any more in heauen. \V And that great dragon was
caſt forth, the old ſerpent, which is called the Diuel and Satan, which
ſeduceth the whole world: and he was caſt into the earth, & his Angels
were throwen downe with him. \V And I heard a great voice in heauen
ſaying: Now is there made ſaluation and force, and the Kingdom of our
God, and the power of his Chriſt: becauſe the accuſer of our Brethren is
caſt forth, who accuſed them before the ſight of our God day and
night. \V And
\SNote{When the Angels or we haue the victorie, we muſt know it is by
the bloud of Chriſt, and ſo al is referred alwaies to him.}
they ouercame him by the bloud of the Lamb, and by the word of their
teſtimonie, and they loued not their liues euen vnto death. \V Therfore
reioyce, ô heauens, and you that dwel therein. Woe to the earth & to the
ſea, becauſe the Diuel is deſcended to you, hauing great wrath, knowing
that he hath a litle time.

\V And after the dragon ſaw that he was throwen into the earth, he
perſecuted the woman which brought forth the man-child: \V and there
were giuen to the woman two wings of a great Egle, that ſhe might flie
into the deſert
%%% o-2821
vnto her place, where ſhe is nouriſhed for
\SNote{This often inſinuation that Antichriſts reigne shal be but three
yeares & a halfe
\XRef{(Dã.~7,~25.}
\XRef{Apoc.~11,~2.~3.}
& in
\XRef{this chap. v.~6.}
\XRef{c.~13,~5.)}
proueth that the heretikes be exceedingly blinded with malice, that hold
the Pope to be Antichriſt, who hath ruled ſo
\Fix{manes}{manie}{obvious typo, fixed in other}
Ages.}
a time & times, & halfe a time, from the face of the ſerpent. \V And the
ſerpent caſt out of his mouth after the woman, water as it were a floud:
that he might make her
\LNote{To be caried away.}{By
\MNote{Antichriſts attempts to draw from the true faith.}
great perſecution he would draw her, that is, her children from the true
faith: but euery one of the faithful elect, gladly bearing their part
thereof, ouercome his tyrannie. At whoſe conſtancie he being the more
offended, worketh malicious attempts in aſſaulting the frailer ſort, who
are here ſignified by \Emph{the reſt of her ſeed that keep the
commandements}, but are not ſo perfect as the former.}
to be caried away with the floud. \V And the earth holpe the woman, and
the earth opened her mouth,
%%% 2959
and ſwallowed vp the floud which the dragon caſt out of his mouth. \V
And the dragon was angrie againſt the woman: and went to make battel
with the reſt of her ſeed, which keep the commandements of God, and haue
the teſtimonie of \Sc{Iesvs} Chriſt. \V And he ſtood vpon the ſand of
the ſea.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

