%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 2941
%%% o-2801
\startChapter[
  title={Chapter 2}
  ]

\Summary{He is commanded to write diuers things to the Churches of
  Epheſus, Smyrna, Pergamus, and Thyatira: praiſing them that had not
  admitted the doctrine of the Heretikes called Nicolaita, 22.~and calling
  others by threats vnto penance: 26.~and promiſing reward to him that
  manfully ouercometh.}

And
\SNote{That which before he willed  him to write to the church he now
willeth to be written to the Angels or Bishops of the ſame only. Where
we ſee, it is al one, to the Church, & to the Head or Gouernour thereof.}
to the Angel of the Church of Epheſus write: Thus ſaith he which
\LNote{Holdeth the ſeuen.}{Much
\MNote{Chriſts care of his Church.}
to be obſerued, that Chriſt hath ſuch care ouer the Church and the
Bishops thereof, that he is ſaid here to beare them vp in his right
hand, & to walk in the middes of them: no doubt to vphold and preſerue
them and to guide them in al truth.}
holdeth the ſeuen ſtarres in his right hand, which walketh in the middes
of the ſeuen Candleſticks of gold: \V I know
\LNote{Thy workes, labour, patience, &c.}{Things
\MNote{Special vertues required in a Bishop.}
required in a Bishop. Firſt, good workes, and great patience in
tribulation. Next zeale and sharp diſcipline toward offenders is here
commended in them. Thirdly, wiſedom and diligence in trial of falſe
Apoſtles & Preachers comming in sheep-skinnes: where is ſignified the
watchful prouidence that ought to be in them, that Heretikes enter not
into their flocks.}
thy workes and labour, and thy patience: & that thou canſt not beare
euil men & haſt tried them which ſay themſelues to be Apoſtles, and are
not, and haſt found them
%%% 2942
liars: \V and thou haſt patience, and haſt borne for my
%%% o-2802
name, and haſt not fainted. \V But I haue againſt thee a few things,
becauſe
\SNote{By this we ſee is plainely refuted that which ſome heretikes
hold, that a man once in grace or charitie cã neuer fal from it.}
thou haſt left thy firſt charitie. \V Be mindful therfore from whence
thou art fallẽ: and doe penance, and doe the firſt workes. But if not: I
come to thee, &
\LNote{Wil moue.}{Note
\MNote{Sinne is the cauſe that God taketh the Catho. faith from
countries.}
that the cauſe why God taketh the truth from certaine countries, and
remoueth their Bishops or Churches into captiuitie or deſolation, is the
ſinne of the Prelates and people. And that is the cauſe (no doubt) that
Chriſt hath taken away our golden candleſticke, that is, our Church in
England. God grant vs to remember our fal, to doe penance and the former
workes of charitie which our firſt Bishops and Church were notable and
renowmed for.}
wil moue thy candleſticke out of this place, vnleſſe thou doe penance.
\V But this thou haſt,
\LNote{Becauſe thou hateſt.}{We
\MNote{Zeale againſt Heretikes.}
ſee here that of al things, Chriſtian people (ſpecially Bishops) should
haue great zeale againſt Heretikes and hate them, that is their wicked
doctrine & conditions, euen as God hateth them. For which only zeale, our
Lord ſaith here that he beareth with ſome Churches and Prelates, and
ſaueth them from perishing.}
becauſe thou hateſt the facts
\LNote{Of the Nicolaites.}{Heretikes
\MNote{\Emph{Nicolaites} the firſt Heretikes ſo called, as a paterne of
Arians, Lutherãs, and the like peculiar callings.}
haue their callings of certaine perſons, as is noted at large
\XRef{Act.~11,~26.}
Theſe had their name of Nicolas, one of the 7.~firſt Deacons that were
choſen
\XRef{Act.~6.}
Who is thought to haue taught cõmunitie of women or wiues, & that it was
lawful to eate of meates offered to idols. Which later point is ſuch a
thing, as if one should hold it lawful to receiue the bread or wine of
the new Communion, which is a kind of \L{Idolothyta}, that is,
\Emph{idolatrous meates}. For though ſuch creatures be good by creation,
yet they be made execrable by profane bleſſings of Heretikes or
Idolaters. And concerning the name of Nicolaites giuen here by our Lord
himſelf to thoſe Heretikes, it is a very paterne & marke vnto the
faithful for euer, what kind of men they should be that should be called after
the like ſort, Arians, Macedonians, Neſtorians, Lutherans, Zwinglians,
&c. See S.~Hierom
\Cite{cont. Lu. in fine.}}
of the Nicolaites, which I alſo hate. \V He that hath an eare, let him
heare what the Spirit ſaith to the Churches: To him that ouercometh, I
wil giue to eate of the tree of life, which is in the Paradiſe of my
God.

\V And to the Angel of the Church of Smyrna write: Thus ſaith
\CNote{\XRef{Apo.~1,~17.}}
the Firſt and the Laſt, who was dead, and liueth: \V I know thy
tribulation and thy
\SNote{This Church repreſenteth the ſtate of thẽ that are ſpoiled of
their goods, empriſoned, & manifoldly afflicted for the Catho. faith.}
pouertie, but thou art rich: and thou art blaſphemed of them that ſay
themſelues to be Iewes and are not, but are the Synagogue of Satan.
\V Feare none of theſe things which thou ſhalt ſuffer. Behold the Diuel
wil ſend ſome of you into priſon that you may be tried: and you ſhal
haue tribulation ten daies. Be thou faithful vntil death: and I wil giue
thee
\SNote{The ſingular reward of Martyrdom.}
the crowne of life. \V He that hath an eare, let him heare what the
Spirit ſaith to the Churches: He that ſhal ouercome, ſhal not be hurt of
the
\SNote{The death of the body is the firſt death: the death of the ſoule,
the ſecõd which Martyrs are ſureſt to eſcape of al men.}
ſecond death.

\V And to the Angel of the Church of Pergamus write: Thus ſaith he that
hath the ſharp two-edged ſword: \V I know where thou dwelleſt, where the
ſeate of Satan is: and thou holdeſt my name, and haſt not denied my
faith. And in thoſe daies Antipas my faithful witneſſe, who was ſlaine
among you,
\SNote{The ſpecial reſidẽce of Satan is where the faithful are
perſecuted for Chriſtes truth. Where not to deny the Cath. faith for
feare, is much here commended.}
where Satan dwelleth. \V But I haue againſt thee a few things: becauſe
thou haſt there, them that hold the doctrine of Balaam, 
\CNote{\XRef{Nu.~24,~14.}
\XRef{25,~2.}}
who taught Balac
\LNote{To caſt a ſcandal.}{Ioſephus
\CNote{\Cite{li.~4. Antiq. c.~6.}}
\MNote{Balaam ouercõming God's people by perſuaſion of lecherie and
belli-cheere, was a type of Heretikes.}
writeth that when Balaam could not curſe God's people, nor otherwiſe
anoy them, he taught Balac a way how to ouerthrow them: to wit, by
preſenting vnto them their Heathen women very beautiful, and delicate
dishes of meate offered to Belphegor: that ſo being tempted they might
fal to heathenish manners and diſpleaſe God. To which craftie counſel of
Balaam the Apoſtle reſembleth Heretikes fraud, 
\CNote{\XRef{2.~Pet.~2,~15.}
\XRef{Iuda v.~11.}}
who by offering of libertie of meate, women, Church goods, breach of vowes,
and ſuch other licentious allurements cauſe many moe to fal, then by
their preaching.}
to caſt a ſcandal before the children of Iſrael, to eate and commit
fornication: \V ſo haſt thou alſo them that hold the doctrine of the
Nicolaites. \V In like manner doe penance. If not: I wil come to thee
quickly, and wil fight againſt them with the ſword of my mouth. \V He
that hath an eare, let him heare what the Spirit ſaith to the Churches:
To him that ouer commeth I wil giue the hidden manna, and wil giue him a
white
\TNote{\G{ψῆφον} \L{calculũ}.}
counter, and in the counter, a new name written, which no man knoweth,
but he that receiueth it.

\V And to the Angel of the Church of Thyatira write: Thus ſaith the
Sonne of God, which hath eyes as a flame of
%%% o-2803
fire, and his feet like to latten. \V I know thy
\SNote{None of theſe are any thing worth without the other.}
workes, and faith, and thy charitie, and miniſterie, and thy patience
and thy laſt workes moe then the former. \V But I haue againſt thee a
few things: becauſe thou permitteſt
\LNote{The woman Iezabel.}{He
\MNote{Zeale againſt Heretikes.}
warneth Bishops to be zelous and ſtout againſt falſe Prophets and
Heretikes of what ſort ſoeuer, by alluding couertly to the example of
holy Elias,
\CNote{\XRef{3.~Reg.~18.}}
that in zeale killed 450.~falſe prophets of Iezabel, and ſpared not
\MNote{Achab and Iezabel.}
Achab nor Iezabel themſelues, but told them to their faces that they
troubled Iſrael, that is, the faithful people of God. And whether there
were any ſuch great woman then, a furtherer and promotour of the
Nicolaites, whom the Prophet should here meane, it is hard to ſay.}
the woman
\CNote{\XRef{3.~Reg.~18.}}
Iezabel, who calleth her ſelf a Propheteſſe, to teach, and to ſeduce my
ſeruants to fornicate, and to eate of things ſacrificed to idols. \V And
I gaue her a time that ſhe might doe penãce &
\LNote{She wil not repent.}{See
\MNote{Free-wil.}
free-wil here moſt plainely, and that God is not the proper cauſe of
obduration or impenitence, but man himſelf only.
\MNote{God is not authour of euil.}
Our Lord giueth ſinners ſo long life, ſpecially to expect their
amendement: but Iezabel (to whom the Apoſtle here alludeth) would neuer
repent.}
ſhe wil not repẽt frõ her fornication. \V Behold I wil caſt her into a
bed: and
\LNote{They that commit aduoutrie with her.}{Such
\MNote{They that communicate with Heretikes, shal be damned with them.}
as communicate with Heretikes, shal be damned (alas) with them. For, not
only ſuch as were in their harts of Iezabels religion, or inwardly
beleeued in Baal, but ſuch as externally for feare worshipped him (which
the Scriptures cal,
\CNote{\XRef{3.~Reg.~19. v.~18.}
\XRef{Ro.~12.}}
bowing of their knees to Baal) are culpable. As now many bow their knees
to the Communion, that bow not their harts.}
they that cõmit aduoutrie with her, ſhal be in very great tribulation,
vnleſſe they doe penãce frõ their workes: \V and her childrẽ I wil kil
with death, & al the Churches ſhal know
\CNote{\XRef{1.~Reg.~16.}
\XRef{Pſ.~7,~10.}
\XRef{Ier.~11,~20.}}
that I am he that ſearcheth the reines & harts & I wil
%%% 2943
giue to euery one of you
\SNote{Who ſeeth not here that good workes deſerue ſaluation, as il
workes damnation: and that it is not faith alone that God rewardeth, but
that faith which worketh by charitie.}
according to his workes. \V But I ſay to you and the reſt which are at
Thyatira, whoſoeuer haue not this doctrine, which haue not knowen the
%%% !!! Var only in other
\Var{depth}{depthes}
of Satan, as they ſay, I wil not caſt vpon you another weight. \V
Yet that which you haue, hold til I come. \V And he that ſhal ouercome
and keep my workes vnto the end,
\LNote{I wil giue him power.}{Obſerue
\MNote{Saints alſo are Patrones, not only Angels.}
that not only Angels haue power and regiment ouer Countries vnder God,
but now for the honour of Chriſts humane nature, and for his miniſterie
in the world, the Saints deceaſed alſo, being in heauen, haue
gouernement ouer men and Prouinces, and therfore haue to doe with our
affaires in the world. Which is againſt the Heretikes of theſe daies,
that to take away our praiers to Saints, would ſpoile them of many
ſoueraigne dignities, wherein the Scriptures make them equal with
Angels.}
I wil giue him power ouer the Nations, \V and
\CNote{\XRef{Pſ.~2,~9.}}
he ſhal rule them with a rod of yron, and as the veſſel of a potter ſhal
they be broken, \V
\SNote{This great priuiledge of Saints riſeth of the power and
preeminence of Chriſt, which his Father gaue him according to his
humanitie; and therfore to deny it to Saints, is to deny it to Chriſt
himſelf.}
as I alſo haue receiued of my Father: and I wil giue him the morning
ſtarre. \V He that hath an eare, let him heare what the Spirit ſaith to
the Churches.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

