%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-11


\project douay-rheims


%%% 2956
%%% o-2817
\startChapter[
  title={Chapter 11}
  ]

\Summary{S.~Iohn meaſuring the Temple, 3.~heareth of two witneſſes that
  shal preach: 7.~whom the beaſt comming vp from the ſea shal
  kil. 11.~But they riſing againe aſcend into heauen, 13.~and ſeuen
  thouſand perſons are ſlaine with an earthquake: 15.~and at the ſound
  of the ſeuenth Angel, the foure and twentie Seniours giue praiſe and
  thankes to God.}

%%% o-2818
And there was giuen me a reed like vnto a rod: and it was ſaid to me:
Ariſe, and meaſure the Temple of God, and the altar, and them that adore
in it. \V But the court which is without the Temple, caſt forth, &
meaſure not that: becauſe it is giuen to the Gentils, and they ſhal
tread vnder-foot the holie citie
\SNote{Three yeares and a halfe, which is the time of Antichriſts reigne
& perſecutions.}
two & fourtie months: \V and I wil giue to
\LNote{My two witneſſes.}{Enoch
\MNote{Enoch & Elias yet aliue, shal preach in the time of Antichriſt.}
and Elias, as it is commonly expounded. For, that Elias shal come againe
before the later day, \Emph{it is a moſt notorious knowen thing} (to vſe
S.~Auguſtines words) \Emph{in the mouthes and harts of faithful men.}
See
\Cite{li.~20. de Ciuit. Dei c.~29.}
\Cite{Tract.~4. in Ioan.}
and both of Enoch and Elias,
\Cite{Lib.~1. de pec. merit. c.~3.}
So the reſt of the Latin Doctours: as, S.~Hierom
\Cite{ad Pammach. ep.~61. c.~11.}
\Cite{in Pſal.~20.}
S.~Ambroſe
\Cite{in Pſal.~45.}
S.~Hilarie
\Cite{20.~can. in Mat.}
Proſper
\Cite{li. vltimo de Promiſsionibus c.~13.}
S.~Gregorie
\Cite{li.~14. Moral c.~11.}
\Cite{ho.~12. in Ezech.}
Beda
\Cite{in 9.~Marci.}
The Greek Fathers alſo, as S.~Chryſoſtom
\Cite{ho.~58. in Mat.}
and
\Cite{ho.~4. in 2.~Theſſal.}
\Cite{ho.~21. in Geneſ.}
and
\Cite{ho.~22. in ep. ad Hebr.}
Theophylact and Oecumenius in
\Cite{17.~Matthæi}
S.~Damaſcene
\Cite{li.~4. de Orthodoxa fide~27.}

Furthermore, that they liue alſo in Paradiſe, it is partly gathered out
of the Scripture
\XRef{Eccleſiaſtici~44,~16.}
where it is plainely ſaid of Enoch, that \Emph{he is tranſlated into
Paradiſe}, as al our Latin exemplars doe read: and of Elias, that he was
taken vp aliue, it is euident
\XRef{4.~Reg.~2.}
And S.~Irenæus ſaith, it is the tradition of the Apoſtles, that they be
both there.
\Cite{li.~5. in initio.}
\L{Dicunt Preſbyteri} (ſaith he) \L{qui ſunt Apoſtolorum Diſcipuli},
\Emph{So ſay the Prieſts or Ancients that are the ſcholers of the
Apoſtles}. See S.~Iuſtine
\Cite{q.~85. ad Orthodoxos.}
Finally, that they shal returne into the companie of men in the end of
the world, to preach againſt Antichriſt, and to inuite both Iewes and
Gentils to penance, and ſo be martyred, as this place of the Apocalypſe
ſeemeth plaine, ſo we haue in part other teſtimonies hereof
\XRef{Malac.~4.}
\XRef{Eccleſiaſtici~44,~16.}
\XRef{48,~10.}
\XRef{Mat.~17,~11.}
See alſo Hypolitus booke of
\Cite{Antichriſt and the end of the world.}
Al which being wel conſidered, the Heretikes are too contentious and
incredulous, to diſcredit the ſame, as they commonly doe.}
my two witneſſes, and they ſhal prophecie a thouſand two hundred ſixtie
daies, clothed with ſacke-clothes. \V Theſe are the two oliue trees &
the two candleſticks that ſtand in the ſight of the Lord of the
earth. \V And if any man wil hurt them, fire ſhal come forth out of
their mouthes, and ſhal deuoure their enemies. And if any man wil hurt
them, ſo muſt he be ſlaine. \V Theſe haue power to ſhut heauen, that it
raine not in the daies of their prophecie: and they haue power ouer the
waters to turne them into bloud, and to ſtrike the earth with al plague
as often as they wil.

\V And when they ſhal haue finiſhed their teſtimonie, the
\MNote{The great Antichriſt.}
beaſt which aſcendeth from the depth, ſhal make warre againſt them, and
ſhal ouercome them, and kil them. \V And their bodies ſhal lie in
the ſtreets of the
\SNote{He meaneth Hieruſalem, named Sodom and Ægypt for the imitation of
them in wickednes. So that we ſee his cheefe reigne shal be there,
though his tyrannie may extend to al places of the world.}
great citie, which is called ſpiritually Sodom and Ægypt, where their
Lord alſo was crucified. \V And there shal of Tribes, and peoples, and
tongues, and Gentils, ſee their bodies for three daies, and a halfe: and
they ſhal not ſuffer their bodies to be laid in monuments. \V And the
inhabitants of the earth
\SNote{The wicked reioyce, when holy men are executed by the tyrants of
the world, becauſe their life and doctrine are burdenous vnto them.}
ſhal be glad vpon them, and make merie: and ſhal ſend guifts one to
another, becauſe theſe two Prophets tormented them that dwelt vpon the
earth. \V And after three daies and a halfe, the ſpirit of life from God
entred into them. And they ſtood vpon their feet, and great feare fel
vpon them that ſaw them. \V And they heard a loud voice frõ heauen
ſaying to them: Come vp hither. And they went vp into heauen in a cloud:
and their enemies ſaw them. \V And in that houre there was made a great
earth-quake: and the tenth part of the citie fel: and there were ſlaine
in the earth-quake names of mẽ ſeuen thouſand: and the reſt were caſt
into a feare, and gaue glorie to the God of heauen.

\V The ſecond woe is gone: and behold the third woe wil come
quickly. \V And the ſeuenth Angel ſounded with
%%% o-2819
a trompet: and there were made loud voices in heauen ſaying,
\SNote{The Kingdõ of this world vſurped before by Satan & Antichriſt,
shal afterward be Chriſtes for euer.}
The kingdom of this world is made our Lords & his Chriſtes, and he ſhal
reigne for euer and euer. Amen.

%%% 2957
\V And the foure and twentie Seniours which ſit on their ſeats in the
ſight of God, fel on their faces, and adored God, \V ſaying: We thanke
thee Lord God omnipotent, which art, and which waſt, and which ſhalt
come: becauſe thou haſt receiued thy great power, and haſt reigned. \V
And the Gentils were angrie, and thy wrath is come, and the time of the
dead, to be iudged, and
\SNote{To repay the hire or wages (for ſo both the Greek word and the
Latin ſignifie) due to holy men, proueth againſt the Proteſtants, that
they did truely merit the ſame in this life.}
to render reward to thy Seruants the Prophets and Saints, and to them
that feare thy name,
%%% !!! Missing LNote in both
%%% \LNote{}{}
litle and great, and to deſtroy them that haue corrupted the earth.

\V And the Temple of God was opened in heauen: and the arke of his
Teſtament was ſeen in his Temple, and there were made lightnings, and
voices, and an earth-quake and great haile.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

