%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 2951
%%% o-2812
\startChapter[
  title={Chapter 7}
  ]

\Summary{The earth being to be punished, 3.~they are commanded to ſaue
  them that are ſigned in their fore-heads: 4.~which are deſcribed and
  numbred both of the Iewes and Gentiles, bleſsing God. 13.~Of them that
  were clothed in white ſtoles or long robes.}

After theſe things I ſaw foure Angels ſtanding vpon the foure corners of
the earth, holding the foure winds of the earth that they ſhould not
blow vpon the land, nor vpon the ſea, nor on any tree. \V And I ſaw
another Angel aſcending from the riſing of the ſunne, hauing the ſigne
of the liuing God; & he cried with a loud voice to the foure Angels, to
whom it was giuen to hurt the earth and the ſea, \V ſaying: Hurt not the
earth and the ſea, nor the trees, til
\SNote{It is an alluſiõ to the ſigne of the Croſſe which the faithful
beare in their foreheads to shew they be not ashamed of Chriſt.
\Cite{S.~Aug. tract.~43. in Io.}}
we ſigne the ſeruants of our God in their foreheads.

\V And I heard the number of them that were ſigned, an hundred fourtie
foure thouſand were ſigned,
\SNote{Of al the Tribes put together, ſo many, 144,000.}
of euery Tribe of the children
%%% 2952
of Iſrael. \V
\SNote{He ſignifieth by theſe thouſands and the multitude following, al
the elect: but the elect of the Iewes, to be in a certaine number: the
elect of the Gentils to be innumerable.}
Of the Tribe of Iuda, twelue thouſand ſigned. Of the Tribe of Ruben,
twelue thouſand ſigned. Of the Tribe of Gad, twelue thouſand ſigned. \V
Of the Tribe of Aſer, twelue thouſand ſigned. Of the Tribe of Nephali,
twelue thouſand ſigned. Of the Tribe of Manaſſes,
%%% o-2813
twelue thouſand ſigned. \V Of the Tribe of Simeon, twelue thouſand
ſigned. Of the Tribe of Leui, twelue thouſand ſigned. Of the Tribe of
Iſſachar, twelue thouſand ſigned. \V Of the Tribe of Zabulon, twelue
thouſand ſigned. Of the Tribe of Ioſeph, twelue thouſand ſigned. Of the
Tribe of Beniamin, twelue thouſand ſigned.

\V After theſe things I ſaw
\MNote{The elect of the Gentils.}
a great multitude which no man could number, of al Nations, and Tribes,
and peoples, & tongues: ſtanding before the throne, and in the ſight of
the Lamb, clothed in white robes, and
\SNote{Boughes of the palme tree be tokens of triumph & victorie.}
palmes in their hands: \V And they cried with a lowd voice, ſaying:
Saluation to our God which ſitteth vpon the throne, and to the Lamb. \V
And al the Angels ſtood in the circuit of the throne, and of the
Seniours, and of the foure beaſts: and they fel in the ſight of the
throne vpon their faces, and adored God, \V ſaying, Amen. Benediction,
and glorie, and wiſedom, and thankes-giuing, honour and power and
ſtrength to our God for euer and euer. Amen.

\V And one of the Seniours anſwered, and ſaid to me: Theſe that are
clothed in the white robes, who be they? and whence came they? \V And I ſaid
to him: My Lord thou knoweſt. And he ſaid to me: Theſe are they which
are come out of great tribulation, and haue waſhed their robes, and made
them white in the bloud of the Lamb. \V
\MNote{The glorie of Martyrs.}
Therfore they are before the throne of God, and they ſerue him day and
night in his Temple: and he that ſitteth in the throne, ſhal dwel ouer
them. \V
\CNote{\XRef{Eſa.~49,~10.}}
They ſhal no more hunger nor thirſt, neither shal the ſunne fal vpon
them, nor any heat. \V Becauſe the Lamb which is in the middes of the
throne, ſhal rule them, and shal conduct them to the liuing fountaines
of waters, and
\CNote{\XRef{Eſ.~25,~8.}
\XRef{Apo.~21,~4.}}
God wil wipe away al teares from their eyes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

