%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 2976
%%% o-2838
\startChapter[
  title={Chapter 20}
  ]

\Summary{An Angel caſteth the dragon (or diuel) bound, into the depth
  for a thouſand yeares, in which the ſoules of Martyrs in the firſt
  reſurrection shal reigne with Chriſt. 7.~After which yeares, Satan
  being let looſe, shal raiſe Gog & Magog, an innumerable armie, againſt
  the beloued citie: 9.~but a fire from heauen shal deſtroy
  them. 12.~Then bookes are opened, and he that ſitteth vpon the throne,
  iudgeth al the dead according to their workes.}

And
\SNote{See in S.~Auguſtin
\Cite{(li.~20. de Ciuit. c.~7.~8. &~ſeq.)}
the expoſition of this chapter.}
I ſaw an Angel deſcending from heauen, hauing the key of the bottomles
depth, and a great chaine in his hand. \V And he apprehended the dragon
the old ſerpent, which is the Diuel and Satan, and
\LNote{Bound him.}{Chriſt by his Paſsion hath abridged the power of the
Diuel for a thouſand yeares, that is, the whole time of the new
Teſtament, vntil Antichriſts time, when he shal be looſed againe, that
is, be permitted to deceiue the world, but for a short time only, to
wit, three yeares and a halfe.}
bound him for a thouſand yeares. \V And he threw him into the depth, and
shut him vp, and ſealed ouer him, that he ſeduce no more the Nations,
til the thouſand yeares be conſummate. And after theſe things he muſt be
looſed a litle time.

\V And
\LNote{I ſaw ſeats.}{S.~Auguſtin
\MNote{Bishops Cõſiſtories & iudicial power.}
\Cite{(lib.~20. de Ciuit. Dei. c.~9.)}
taketh this to be ſpoken, \Emph{not of the laſt iudgement, but of the
Sees or Conſiſtories of Bishops and Prelates, and of the Prelates
themſelues, by whom the Church is now gouerned. As the iudgement here
giuen, can be taken no otherwiſe better, then of that which was ſaid by
our Sauiour
\XRef{Mat.~18.}
whatſoeuer you bind in earth, shal be bound in heauen: and therfore
the Apoſtle ſaith,
\CNote{\XRef{1.~Cor.~5.}}
what haue I to doe, to iudge of them that are without?}}
I ſaw ſeats: and they ſate vpon them, and iudgement was giuen them,
\LNote{And the ſoules.}{\Emph{He
\MNote{During a thouſand yeares (that is the time of this militant
Church) Saints reigne with Chriſt in ſoule only.}
meaneth} (ſaith S.~Auguſtin in the
\Cite{place alleaged)}
\Emph{the ſoules of Martyrs, that they shal in the meane time, during
theſe thouſand yeares, which is the time of the Church militant, be in
heauen without their bodies, and reigne with Chriſt: for, the ſoules}
(ſaith he) \Emph{of the godly departed, are not ſeparated from the
Church which is euen now the Kingdom of Chriſt, for els there should be
kept no memorie of them at the altar of God in the communicating of the
body of Chriſt: neither should it auaile to haſten to Baptiſme in the
perils of death, for feare of ending our life without it: nor to haſten
to be reconciled, if we fortune for penance or of il conſcience to be
ſeparated from the ſame body. And why are al theſe things done, but for
that the faithful departed alſo be members of the Church?} And though
for an example the Martyrs be only named here, yet it is meant of others
alſo that die in the ſtate of grace.}
and the ſoules of the beheaded for the teſtimonie of \Sc{Iesvs}, and for
the word of God, and that adored not the beaſt nor his image, nor
receiued his character in their foreheads or in their hãds, haue liued
& reigned with Chriſt
\SNote{\L{Quid in millenario numero niſi ad proferendam nouam ſobolem perfecta
vniueritas præſcitæ generationis exprimitur? hinc per Ioannem dicitur;
Et regnabunt cũ illo mille annis, quia regnum ſanctæ Eccleſiaæ,
vniuerſitatis perfectione ſolidatur.}
\Cite{D.~Gregor. li.~9. Moral. c.~1.}}
a thouſãd yeares. \V
\LNote{The reſt liued not.}{The
\MNote{The reſt are dead and damned in ſoule, during the ſame time.}
reſt which are not of the happie number aforeſaid, but liued and died in
ſinne, reigne not with Chriſt in their ſoules during this time of the
new Teſtament, but are dead in ſoule ſpiritually and in body naturally,
til the day of iudgement.
\Cite{S.~Auguſt. ibidem.}}
The reſt of the dead liued not, til the thouſand yeares be conſummate.
\LNote{This is the firſt reſurrection.}{As
\MNote{The firſt reſurrection, of the ſoule only.}
there be two regenerations, one by faith, which is now in Baptiſme; and
another according to the flesh, when at the later day the body shal
be made immortal and incorruptible: ſo there are two reſurrections, the
one now of the ſoules to ſaluation when they die in grace, which is
called, the firſt, the other of the bodies at the later day. S.~Auguſtin
\Cite{li.~20. de Ciuit. c.~6.}}
This is
%%% o-2839
the firſt reſurrection. \V Bleſſed and holy is he that hath part
in the firſt reſurrection. In theſe the ſecond death hath not power: but
\LNote{They shal be Prieſts.}{\Emph{It
\MNote{Prieſts, ſome properly ſo called, ſome improperly.}
is not ſpoken} (ſaith S.~Auguſtin
\Cite{li.~20. de Ciuit. c.~10.)}
\Emph{of Bishops and Prieſts only, which are properly now in the Church
called Prieſts: but as we cal al Chriſtians, for the myſtical Chryſme or
ointment, ſo al Prieſts, becauſe they are the members of one Prieſt, of
whom the Apoſtle Peter ſaith,
\CNote{\XRef{1.~Pet.~2.}}
A holy people, a Kingly Prieſthood.} Which words be notable for their
learning that thinke there be none properly called Prieſts now in the
new Teſtament, no otherwiſe then al Chriſtian men and women, and a
confuſion to them that therfore haue turned the name \Emph{Prieſts} into
\Emph{Miniſters.}}
they ſhal be Prieſts of God and of Chriſt: and shal reigne with him a
thouſand yeares.

\V And when the thouſand yeares shal be conſummate,
\LNote{Satan shal be looſed.}{In
\MNote{The binding and looſing of Satan, explicated by S.~Auguſtin.}
the whole
\Cite{8.~chapter of the ſaid 20.~booke de Ciuitate Dei}
in S.~Auguſtin, is a notable commentarie of theſe words. Where firſt he
declareth, that neither this binding nor looſing of Satan is in reſpect
of ſeducing or not ſeducing the Church of God: prouing that whether he
be bound or looſe, he can neuer ſeduce the ſame. \Emph{The ſame}, ſaith
he, \Emph{shal be the ſtate of the Church at that time when the Diuel is
to be looſed, euen as ſince it was inſtituted, the ſame hath it been &
shal be at al time in her children that ſucceed each other by birth &
death.} And a litle after: \Emph{This I thought was therfore to be
mentioned, leſt any man should thinke, that during the litle time
wherein the Diuel shal be looſed, the Church shal not be vpon the earth,
he either not finding it here when he shal be let looſe, or conſuming
it when he shal by al meanes perſecute the ſame.} Secondly he declareth,
that the Diuel to be bound, is nothing els but not to be permitted by
God to exerciſe al his force or fraud in tentations: as to be looſed, is
to be ſuffered by God for a ſmal time, that is, for three yeares and a
halfe, to practiſe and proue al his power and arts of tentations againſt
the Church and her children, and yet not to preuaile againſt
them. Thirdly this Doctour sheweth by what great mercie our Lord hath
tied Satan and abridged his power during the whole millenarie or
thouſand yeares, which is al the time of the new Teſtament vntil then:
\MNote{The short reigne of Antichriſt.}
and with what wiſedom he permitteth him to breake looſe that litle time
of three yeares & ſixe moneths, toward the later day, which shal be the
reigne of Antichriſt. Laſtly he sheweth what kind of men shal be moſt
ſubiect to the Diuels ſeduction, (euen ſuch as now by tentation of
Heretikes goe out of the Church) and who shal auoid it.

By al which we may confute diuers falſe expoſitions of old & late
Heretikes.
\MNote{\L{Millenarii} or \L{Chiliaſtæ}.}
Firſt, the ancient Sect of the Millenaries, that grounded vpon theſe
thouſand yeares named by the Prophet, this hereſie, that there should be
ſo many yeares after the reſurrection of our bodies, in which we should
reigne with Chriſt in this world, in our bodies, in al delights and
pleaſures corporal of meats, drinkes, and ſuch like, which they called
the firſt reſurrection. Of which hereſie Cerinthus was the Authour. 
\Cite{Epiph. hær.~77. in fine.}
\Cite{Hiero. Comment. in c.~19. Mat.}
\Cite{Auguſt. hær.~8. ad Quod vult. Deum.}
Euſebius alſo
\Cite{(li.~3. hiſtorie c.~33.)}
sheweth that ſome principal men were in part (though after a more honeſt
manner concerning thoſe corporal delicacies) of the ſame opinion by
miſconſtruction of theſe words of S.~Iohn.
\MNote{The Scriptures hard.}
Whereby we learne and al the world may perceiue, the holy Scriptures to
be hard, when ſo great Clerkes did erre, and that there is no ſecuritie
but in that ſenſe which the Church alloweth of.

The
\MNote{By S.~Auguſtins foreſaid explication, is euidently deduced
againſt the Proteſtãts, that the Church can not erre, and that the Pope
can not poſſibly be Antichriſt.}
late Heretikes alſo by the ſaid S.~Auguſtines words are fully refuted,
affirming not only that the Church may be ſeduced in that great
perſecution of the Diuels looſing; but that it hath been ſeduced euen a
great peece of this time when the Diuel is bound: holding that the very
true Church may erre or fal from truth to errour and idolatrie, yea
(which is more blaſphemie) that the cheefe Gouernour of the Church is
Antichriſt himſelf, and the very Church vnder him, the whore of
Babylon: and that this Antichriſt, (which the Scriptures in ſo many
places, and here plainely by S.~Auguſtines expoſition, teſtifie, shal
reigne but a ſmal time, and that toward the laſt iudgement,) hath been
reuealed long ſithence, to be the Pope himſelf, Chriſts owne Vicar, &
that he hath perſecuted the Saints of their ſect for theſe thouſand
yeares at the leaſt. Which is no more but to make the Diuel to be looſe,
& Antichriſt to reigne the whole thouſand yeares, or the moſt part
thereof, that is, almoſt the whole time of the Churches ſtate in the new
Teſtament: (which is againſt this & other Scriptures euidently,
appointing that to be the time of the Diuels binding:) Yea it is to make
Antichriſt & the Diuel weaker toward the day of iudgemẽt then before,
and the truth better to be knowen, and the faith more common, the neerer
we come to the ſame iudgement: which is expreſly
\CNote{\XRef{Mt.~24.}
\XRef{Lu.~18,~8.}}
againſt the Ghoſpel and this prophecie of S.~Iohn.

We
\MNote{An inuincible demonſtratiõ.}
ſee that the Sects of Luther, Caluin, and other, be more ſpred through
the world then they were euer before, and conſequently the Pope and his
religion leſſened, and his power of punishing (or, as they cal it)
perſecuting the ſaid Sectaries, through the multitude of his aduerſaries
diminished. How then is the Pope Antichriſt, whoſe force shal be greater
at the later end of the world, then before? Or how can it be otherwiſe,
but theſe Sect Maiſters should be Antichriſts neere precurſours, that
make Chriſts cheefe Miniſters & the Churches cheefe Gouernours that haue
been theſe thouſand yeares and more, to be Antichriſts; & themſelues and
their Sects to be true, that come ſo neere the time of the Diuels
looſing and ſeduction, and of the perſonal reigne of Antichriſt.}
Satan shal be looſed out of his priſon, and shal goe forth, and ſeduce
\SNote{S.~Auguſtin thinketh that theſe doe not ſignifie any certaine
Nations, but al that shal then be ioyned with the Diuel and Antichriſt
againſt the Church.
\Cite{li.~20. de Ciuit. c.~11.}
See S.~Hierom
\Cite{in Ezech. li.~11.}}
the Nations that are vpon the foure corners of the earth,
\CNote{\XRef{Ezec.~38,~14.}
\XRef{39,~1.}}
Gog and Magog, and shal gather them into battel, the number of whom is
as the ſand of the ſea. \V And they aſcended vpon the bredth of the
earth, and compaſſed
\LNote{The camp of the Saints.}{S.~Auguſtin in the ſaid
\Cite{20.~booke de Ciuit. Dei cap.~11.}
\MNote{The camp of Saints is the Catho. Church through the world.}
\Emph{It is not} ſaith he, \Emph{to be taken that the perſecutours shal
gather to any place, as though the camp of the Saints or the beloued
citie should be but in one place, which indeed are no other thing but
the Church of Chriſt ſpred through the whole world. And therfore
whereſoeuer the Church shal then be (which shal be in al Nations euen
then, for ſo much is inſinuated by this latitude of the earth here
ſpecified) there shal the tents of Saints be, and the beloued citie of
God, and there shal she be beſieged by al her enemies, which shal be in
euery countrie where she is, in moſt cruel and forcible ſort.} So
writeth this profound holy Doctour. Whereby we ſee,
\MNote{As now Heretikes in particular coũtries, ſo Antichriſt shal
perſecute the Churches of al Nations.}
that, as now the particular Churches of Englãd, Scotland, Flanders, &
ſuch like, be perſecuted by their enemies in thoſe countries, ſo in the
time of Antichriſt, the Churches of al Nations, as of Italie, Spaine,
France, and al other which now be quiet, shal be aſſaulted as now the
foreſaid are, and much more, for that the general perſecution of the
whole, shal be greater then the particular perſecution of any Churches
in the world.}
the camp of the Saints, and the beloued citie. \V And
\LNote{There came downe a fire.}{It
\MNote{What is meant by fire from heauen.}
is not meant of the fire of Hel (ſaith S.~Auguſtin
\Cite{ib. c.~12.)}
into which the wicked shal be caſt after the reſurrection of their
bodies, but of an extraordinarie help that God wil ſend from heauen, to
giue ſuccourſe to the Saints of the Church that then shal fight againſt
the wicked: or, the very feruent & burning zeale of religion & Gods
honour, which God wil kindle in the harts of the faithful, to be
conſtant againſt al the forces of that great perſecution.}
there came downe fire from God out of heauen, and deuoured them: \V and
the Diuel which ſeduced them, was caſt into the poole of fire and
brimſtone, where both the beaſt & the Falſe-prophet shal be tormented
day and night for euer and euer.

\V
\MNote{\Sc{The 6.~Vision}.}
And I ſaw a great white throne, and one ſitting vpon it, from whoſe
ſight
\SNote{They shal then be new, not the ſubſtãce, but the shape chãged.
\XRef{2.~Pet.~3.}
See S.~Auguſtin
\Cite{li.~20. de Ciuit. c.~14.}}
earth and heauen fled, and there was no place found for them. \V And I
ſaw the dead, great and litle, ſtanding in the ſight of the throne, and
\SNote{The bookes of mens conſciences, where it shal plainely be read
what euery mans life hath been.}
books were opened: and
\CNote{\XRef{Apoc.~3,~5.}
\XRef{21,~27.}}
\LNote{Another booke.}{This
\MNote{The booke of euery mans workes, opened in the day of iudgement.}
is the booke of God's knowledge or predeſtinatiõ, wherein that which
before was hid to the world, shal be opened, & wherein the true record
of euery mans workes shal be conteined, and they haue their iudgement
diuerſly according to their workes, and not according to faith only, or
lacke of faith only. For, al infidels (as Turkes, obſtinate Iewes, and
Heretikes) shal neuer come to that examination, being otherwiſe
condemned.}
another booke was opened, which is of life: and the dead were iudged of
thoſe things which were written in the books according to their
works. \V And the ſea gaue the dead that were in it, and death and hel
gaue their dead that were in them, and it was iudged of euery one
according to their works.
%%% 2977
\V And hel and death were caſt into the poole of fire. This is the
ſecond death. \V And
\SNote{Such as doe no good workes, if they haue age and time to doe
them, are not found in the booke of life.}
he that was not found written in the booke of life, was caſt into the
poole of fire.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

