%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-05


\project douay-rheims


%%% 2947
%%% o-2808
\startChapter[
  title={Chapter 5}
  ]

\Summary{4.~S.~Iohn weeping, becauſe no man could open the booke ſealed
  with ſeuen ſeales; 6.~the Lamb that was ſlaine, opened it: which being
  done, 8.~the foure beaſts and foure and twentie Seniours, with an
  innumerable multitude of Angels and al creatures, did glorifie him
  exceedingly.}

And
\MNote{\Sc{The 3.~Vision}.}
I ſaw in the right hand of him that ſate vpon the throne, a
\SNote{S.~Gregorie taketh it to be the booke of holy Scripture.
\Cite{li.~4. Dialog. c.~42.}}
booke written within and without, ſealed with ſeuen ſeales. \V And I ſaw
a ſtrong Angel, preaching with a loud voice: Who is worthie to open the
booke, & to looſe the ſeales thereof? \V And no man was able neither in
heauen nor in earth, nor
\SNote{He ſpeaketh not of the damned in Hel, of whom there could be no
queſtion: but of the faithful in Abrahams boſome, & in Purgatorie.}
vnder the earth, to open the booke, nor looke on it. \V And I wept
much becauſe no man was found worthie to open the booke, nor to ſee
it. \V And one of the Seniours ſaid to me: Weep not; behold
\SNote{So did Iacob
\XRef{(Gen.~49.)}
cal Chriſt, for his kingly fortitude in ſubduing the world vnto him.}
the
\CNote{\XRef{Gen.~49,~9.}}
Lion of the Tribe of Iuda, the root of Dauid, hath won, to open the
booke, and to looſe the ſeuen ſeales thereof.

\V And I ſaw, and behold in the middes of the throne and of the foure
beaſts and in the middes of the Seniours,
\SNote{So Chriſt is called for that he is the immaculate Hoſt or
Sacrifice for our ſinnes.}
a Lamb ſtanding as it were ſlaine, hauing ſeuen hornes & ſeuen eyes:
which are the ſeuen Spirits of God, ſent into al the earth. \V And he
came, and receiued the booke out of the right hand of him that ſate in
the throne. \V And when he had opened the booke, the foure beaſts and
the foure and twentie Seniours fel before the Lamb, hauing euery one
harps, and golden vials ful of odours, which are
\LNote{The praiers of Saints.}{Hereby
\MNote{The Saints in heauen offer our praiers to God.}
it is plaine that the Saints in heauen offer vp the praiers of faithful
and holy perſons in earth (called here Saints, and in Scripture often)
vnto Chriſt. And among ſo many diuine & vnſearchable myſteries ſet downe
without expoſition, it pleaſed God yet, that the Apoſtle himſelf should
open this one point vnto vs, that theſe odours be the lauds and praiers
of the faithful, aſcending and offered vp to God as incenſe, by the
Saints in heauen: that ſo the Proteſtants may haue no excuſe of their
errour, That the Saints haue no knowledge of our affaires or deſires.}
the praiers of
Saints: \V and they ſang a new canticle, ſaying: Thou art worthie ô Lord
to take the booke, and to
%%% 2948
open the ſeales thereof:
\SNote{This maketh againſt the Caluiniſts who are not content to ſay
that we merit not, but that Chriſt merited not for himſelf.
\Cite{Calu. Philip.~2. v.~9.}}
becauſe thou waſt ſlaine, and haſt redeemed vs to God in thy bloud out
of euery tribe and tongue and people and Nation, \V and
\CNote{\XRef{1.~Pet.~2.}}
haſt made vs to our God
\LNote{A Kingdom and Prieſts.}{To
\MNote{Spiritual Kings and Prieſts.}
ſerue God and ſubdue vices and ſinnes, is to reigne or to be a King
ſpiritually. Likewiſe to offer vnto him the Sacrifices of good workes,
is to be a Prieſt after a ſort: though neither the one nor the other in
proper ſpeach. See the
\XRef{Annotation before Chap.~1. v.~6.}}
%%% !!! Var only in other
\Var{a Kingdom}{kings}
and Prieſts, and we ſhal reigne vpon the earth.

\V And I looked, and heard the voice of many Angels round about the
throne, and of the beaſts & of the Seniours: and the number of them was
\CNote{\XRef{Dan.~7,~10.}}
thouſands of thouſands,
%%% o-2809
\V ſaying with a loud voice: The Lamb that was ſlaine, is worthie to
receiue power, and
%%% !!! Var only in other
\Var{diuinitie,}{riches}
and wiſedom, and ſtrength, and honour, and glorie, and benediction. \V
And
\LNote{Euery creature.}{He
\MNote{\L{Limbus Patrum} and Purgatorie.}
meaneth the creatures in heauen, as Angels and Saints: the holy perſons
in earth, & thoſe that were in Limbo, or be in Purgatorie (for of the
damned in hel he can not ſpeake in this caſe:) laſtly, of the peoples in
Ilands (here called the ſea) which the Prophets vſe often to name
ſeuerally, when they foretel the ſpreading of Chriſts glorie through the
world, as
\XRef{Eſa. c.~49.}
\Emph{Heare ye Ilands and you people a farre off, &c.}}
euery creature that is in heauen, and vpon the earth, and vnder the
earth, and that are in the ſea, and that are therein: al did I heare
ſaying:
\CNote{\XRef{Apoc.~4,~11.}}
To him that ſitteth in the throne, and
\SNote{Al the ſaid creatures are bound to giue honour, not only to God,
but to Chriſt as man, and our Redeemer: & ſo they here doe.}
to the Lamb, benediction and honour and glorie and power for euer and
euer. \V And the foure beaſts ſaid, Amen. And the foure and twentie
Seniours fel on their faces: and adored him that liueth for euer and
euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

