%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 2965
%%% o-2827
\startChapter[
  title={Chapter 16}
  ]

\Summary{Vpon the pouring out of the ſeuen cups of Gods wrath, on the
  land, the ſea, the fountaines, the ſeat of the beaſt, Euphrates, and
  the aire; there ariſe ſundrie plagues in the world.}

And I heard a great voice out of the temple, ſaying to the ſeuen Angels:
Goe, and poure out the ſeuen vials of the wrath of God vpon the
earth. \V And the firſt went, and poured out his vial vpon the earth, &
there was made a cruel & very ſore wound vpon men that had the character
of the beaſt: and vpon them that adored the image thereof.

\V And the ſecond Angel poured out his vial vpon the ſea, and there was
made bloud as it were of one dead: and
%%% o-2828
euery liuing ſoule died in the ſea.

\V And the third poured out his vial vpon the riuers and the
fountaines of waters: and there was made bloud. \V And I heard the Angel
%%% 2966
of the waters, ſaying: Thou art iuſt, ô Lord, which art, and which waſt,
the holy one, becauſe thou haſt iudged theſe things: \V
\SNote{The great reuenge that God wil doe at the later day vpon the
perſecutours of his Saints.}
becauſe they haue ſhed the bloud of the Saints and Prophets, & thou haſt
giuen them bloud to drinke. For they are worthie. \V And I heard
another, ſaying: Yea Lord God omnipotent, true and iuſt are thy
iudgements.

\V And the fourth Angel poured out his vial vpon the ſunne, and it was
giuen vnto him to afflict men with heat and fire: \V and men boiled with
great heat, and
\SNote{The deſperate and damned perſons shal blaſpheme God
perpetually. Which shal be ſuch only as doe not repent in this life.}
blaſphemed the name of God hauing power ouer theſe plagues, neither did
they penance to giue him glorie.

\V And the fift Angel poured out his vial vpon the ſeat of the beaſt:
and his Kingdom was made darke, and they together did eate their tongues
for paine: \V and they blaſphemed the God of heauen becauſe of their
paines and wounds, and
\SNote{See
\XRef{chap.~9. v.~20. in the margent.}}
did not penance from their workes.

\V And the ſixt Angel poured out his vial vpon that great riuer
Euphrates: and dried vp the water thereof that a way might be prepared
to the Kings from the riſing of the ſunne.

\V And I ſaw from the mouth
\SNote{The dragon is the Diuel: the beaſt, Antichriſt, or the ſocietie
whereof he is head: the falſe-prophet, either Antichriſt himſelf, or the
companie of Heretikes and ſeducers that follow him.}
of the dragon, and from the mouth of the beaſt, and from the mouth of
the falſe-prophet
\Var{three}{iſſue forth three}
vncleane ſpirits in manner of frogs. \V For they are the ſpirits of
Diuels working ſignes, and they goe forth to the Kings of the whole
earth to gather them into battel at the great day of the omnipotent
God. \V Behold
\CNote{\XRef{Apoc.~3.}}
I come as a theefe: Bleſſed is he that watcheth, & keepeth his garments,
that he
\CNote{\XRef{2.~Cor.~5,~3.}}
walke not naked, and they ſee his turpitude. \V And he ſhal gather them
into a place which in Hebrew is called
\TNote{\Emph{The hil of theeues}, by
%%% !!! Cite?
S.~Hierõ interpretation.}
\Emph{Armagedon}.

\V And the ſeuenth Angel poured out his vial vpon the aire, and there
came forth a loud voice out of the temple from the throne, ſaying: It is
done. \V And there were made lightnings, and voices, and thunders, and a
great earth-quake was made, ſuch an one as neuer hath been ſince men
were vpõ the earth, ſuch an earth-quake, ſo great. \V And
\SNote{The citie or cõmon-wealth of the wicked deuided into three parts:
into infidels, Heretikes, and euil Catholikes. This citie is here called
Babylon, whereof ſee the
\XRef{Annot. vpon the next chapter, v.~5.}}
the great citie was made into three parts: and the cities of the
%%% o-2829
Gentils
fel. And Babylon the great came into memorie before God,
\CNote{\XRef{Ier.~25,~15.}}
to giue her the cup of wine of the indignation of his wrath. \V And
euery Iland fled, and mountaines were not found. \V And great haile like
a talent came downe from heauen vpon men: and men blaſphemed God for the
plague of the haile: becauſe it was made exceeding great.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

