%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 2960
%%% o-2821
\startChapter[
  title={Chapter 13}
  ]

\Summary{1.~A beaſt riſing vp out of the ſea, hauing ſeuen heads and ten
  hornes & ten diadems, 5.~blaſphemeth God, 7.~and warreth againſt the
  Saints and deſtroieth them. 11.~And another beaſt riſing out of the
  earth with two hornes, was altogether for the fore-ſaid beaſt,
  conſtraining men to make and adore the image thereof, and to haue the
  character of his name.}

%%% o-2822
And I ſaw
\LNote{A beaſt comming vp.}{This
\MNote{Many myſteries expounded.}
beaſt is the vniuerſal companie of the wicked, whoſe head is Antichriſt;
and the ſame is called
\XRef{(Apoc.~17.)}
the whore of Babylon. The 7.~heads be expounded
\XRef{(Apoc.~17.)}
ſeuen Kings: fiue before Chriſt, one preſent, and one to come. The
10.~hornes be alſo there expounded to be 10.~Kings that shal reigne a
short while after Antichriſt. This dragon is the Diuel, by whoſe power
the whore or beaſt or Antichriſt worketh. For in the words following
\XRef{(v.~3. &~4.)}
Antichriſt is called the beaſt, to whom the dragon, that is, the diuel
giueth that power of feined miracles. And as we adore God for giuing
power to Chriſt and his followers, ſo they shal adore the Diuel for
aſſiſting Antichriſt and giuing him power.}
a beaſt comming vp from the ſea, hauing ſeuen heads, and ten hornes, &
vpon his hornes ten diadems, and vpon his heads names of blaſphemie. \V
And the beaſt which I ſaw, was like to a Libard, and his feet as of a
Beare, and his mouth, as the mouth of a Lion. And the dragon gaue him
his owne force and great power. \V And I ſaw one of his heads as it were
ſlaine to death: and the wound of his death was cured. And al the earth
was
\SNote{They that now follow the ſimpleſt & groſſeſt heretikes that euer
were without ſeeing miracles, would then much more follow this great
ſeducer working miracles.}
in admiration after the beaſt. \V And they adored the dragon which gaue
power to the beaſt: and they adored the beaſt, ſaying: Who is like
to the beaſt? and who ſhal be able to fight with it? \V And there was
giuen to it a mouth ſpeaking great things and blaſphemies: and power was
giuen to it to worke two and fourtie months. \V And he opened his mouth
vnto blaſphemies toward God,
\SNote{No heretikes euer liker Antichriſt, then theſe in our daies,
ſpecially in blaſphemies againſt Gods Church, Sacraments, Saints,
Miniſters, and al ſacred things.}
to blaſpheme his name, & his tabernacle, & thoſe that dwel in heauen. \V
And it was giuen vnto him
\LNote{To make battel with the Saints.}{He
\MNote{Great perſecution by Antichriſt and his Miniſters.}
shal kil the Saints then liuing, Elias and Enoch, and infinit moe that
profeſſe Chriſt. Whereby we muſt learne, not to maruel when we ſee the
wicked perſecute and preuaile againſt the iuſt, in this life. Then shal
his great perſecutiõ & crueltie trie the Saints patience, as his
wonderful meanes to ſeduce shal trie the ſtedfaſtnes of their faith,
which is ſignified by theſe words following,
\MNote{Their bleſſednes that continue conſtant.}
\Emph{Here is the patience and the faith of Saints}. And when it is
ſaid, \Emph{They adored the beaſt, whoſe names are not written in the
booke of life of the Lamb}, it giueth great ſolace and hope to al them
that shal not yeald to ſuch perſecutions, that they are of Gods elect,
and their names written in the booke of life.}
to make battail with the Saints, & to ouercome them. And power was giuen
him vpon euery Tribe and people, and tongue, and Nation, \V and al that
inhabit the earth, adored it,
\CNote{\XRef{Apoc.~3,~5.}}
whoſe names be not written in the booke of life of the Lamb, which was
ſlaine from the beginning of the world.

\V If any man haue an eare, let him heare. \V He that ſhal lead into
captiuitie, goeth into captiuitie:
\CNote{\XRef{Gen.~9,~6.}
\XRef{Mt.~26,~52.}}
he that ſhal kil in the ſword, he muſt be killed with the ſword. Here is
the patience and the faith of Saints.

\V And I ſaw
\SNote{Another falſe Prophet inferiour to Antichriſt shal worke wonders
alſo, but al referred to the honour of his Maiſter Antichriſt. So doth
Caluin, & other Arch-heretikes peruert the world to the honour of
Antichriſt, and ſo doe their ſcholers alſo for the honour of them.}
another beaſt comming vp from the earth: and he had two hornes, like to
a lamb, & he ſpake as a dragon. \V And al the power of the former beaſt
he did in his ſight: and he made the earth and the inhabitants therein,
to adore the firſt beaſt, whoſe wound of death was cured. \V And he did
great ſignes, ſo that he made alſo fire to come downe from heauen vnto
the earth in the ſight of men. \V And he ſeduceth the inhabitants on the
earth through the ſignes which were giuen him to doe in the ſight of the
beaſt, ſaying to them that dwel on the earth, that they ſhould make
\LNote{The image of the beaſt.}{They
\MNote{The honour of Chriſts image is for the honour of Chriſt.}
that now refuſe to worship Chriſts image, would then worship
Antichriſts. And we may note here, that as the making or honouring of
this image was not againſt the honour of Antichriſt, but wholy for it,
as alſo the image erected of Nabuchodonoſor and the worship thereof was
altogether for the honour of him, ſo is the worship of Chriſts image,
the honour of Chriſt himſelf, and not againſt him, as Proteſtants madly
imagin.}
the image of the beaſt which hath the ſtroke of the ſword, and liued. \V
And it was giuen him to giue ſpirit to the image of the beaſt, and
that the image of the beaſt ſhould ſpeake:
%%% o-2823
and ſhould make, that whoſoeuer ſhal not adore the image of the beaſt,
be ſlaine. \V And he ſhal make al, litle & great, and rich and poore,
and free-men and bond-men, to haue a character in their right hand, or
in their foreheads. \V And that no man may buie or ſel, but he
%%% 2961
that hath
\LNote{The character or the name.}{As
\MNote{Antichriſts triple honour againſt the honour of Chriſt.}
belike for the peruerſe imitation of Chriſt, whoſe image (ſpecially as
on the Rood or Crucifix) he ſeeth honoured and exalted in euery Church,
he wil haue his image adored (for that is Antichriſt, in emulation of
like honour, aduerſarie to Chriſt) ſo for that he ſeeth al true
Chriſtian men to beare the badge of his Croſſe in their foreheads, he
likewiſe wil force al his to haue another marke, to abolish the ſigne of
Chriſt. By the like emulation alſo and wicked oppoſition he wil haue his
name and the letters thereof to be ſacred, and to be worne in mens
cappes, or written in ſolemne places, and to be worshipped, as the name
of \Sc{Iesvs} is and ought to be among Chriſtian men. And as the
ineffable name of God was among the Iewes expreſſed by a certaine number
of
\TNote{\H{יהוה}}
4.~characters (therfore called \Emph{Tetragrammaton}) ſo it ſeemeth
the Apoſtle alludeth here to the number of Antichriſts name.

And
\MNote{The Proteſtãts by abolishing of Chriſtes image, & croſſe &
irreuerence to the name \Sc{Iesvs}, make a ready way to the honour of
Antichriſt.}
here it is much to be noted, that the Proteſtants plucking downe the
image of Chriſt out of al Churches, & his ſigne of the croſſe from mens
foreheads, & taking away the honour and reuerence of the
name \Sc{Iesvs}, doe make roome for Antichriſts image, & marke, and
name. And when Chriſts images and enſignes or armes shal be abolished,
and the Idol of Antichriſt ſet vp inſteed thereof, as it is already
begun; then is the abomination of deſolation which was foretold by
Daniel and our Sauiour.}
the character, or the name of the beaſt, or the number of his name. \V
Here is wiſedom. He that hath vnderſtanding,
\LNote{Let him count.}{Though
\MNote{Antichriſts name ſecret.}
God would not haue it manifeſt before-hand
to the world, who in particular this Antichriſt should be: yet it
pleaſed him to giue ſuch tokens of him, that when he commeth, the
faithful may eaſily take notice of him, according as it is written of
the euent of other prophecies concerning our Sauiour,
\CNote{\XRef{Io.~14,~19.}}
\Emph{That when it is come to paſſe you may beleeue}. In the meane time
we muſt take heed that we iudge not ouer rashly of Gods ſecrets: the
holy Writer here ſignifying, that it is a point of high vnderſtanding,
illuminated extraordinarily by Gods ſpirit, to recken right and decipher
truely before hand, Antichriſts name and perſon.}
let him count the number of the beaſt. For
\LNote{It is the number of a man.}{A
\MNote{Antichriſt shal be one ſpecial man, and of a peculiar name.}
man he muſt be, and not a Diuel or ſpirit, as here it is cleere, & by
S.~Paul
\XRef{2.~Theſſal.~2.}
where he is called, \Emph{the man of ſinne}. Againe, he muſt be one
particular perſõ, & not a number, ſucceſſiõ, or whole order of any
degree of men: becauſe his proper name & the peculiar number, & the
characters thereof be (though obſcurely) inſinuated. Which reproueth the
wicked vanitie of Heretikes, that would haue Chriſts owne Vicars, the
Succeſſours of his cheefe Apoſtle, yea the whole order of them for many
Ages together, to be this Antichriſt. Who by his deſcription here and in
the ſaid Epiſtle to the Theſſalonians, muſt be one ſpecial man, and of a
particular proper name, as our Lord \Sc{Iesvs} is. And whoſoeuer he be,
theſe Proteſtants vndoubtedly are his Precurſours. For as they make his
way by ridding away Chriſts images, croſſe, and name, ſo they
exceedingly promote the matter by taking away Chriſts cheefe Miniſter,
that al may be plaine for Antichriſt.

If
\MNote{The Pope can not be Antichriſt.}
the Pope had been Antichriſt, and had been reuealed now  a good many
yeares ſithence, as theſe fellowes ſay he is to them, then the number of
this name would agree to him, and the prophecie being now fulfilled, it
would euidently appeare that he bare the name and number here noted. For
(no doubt) when he commeth, this count of the letters or number of his
name which before is ſo hard to know, wil be eaſie. For he wil ſet vp
his name in euery place, euen as we faithful men doe now aduance
\Sc{Iesvs}. And what name proper or appellatiue of al or any of the
Popes doe they find to agree with this number, notwithſtanding they
boaſt that they haue found the whole order and euery of them theſe
thouſand yeares to be Antichriſt, and the reſt before euen from
S.~Peter, fore-workers toward his Kingdom?}
it is the number of a man; and
\LNote{The number~666.}{Foraſmuch
\MNote{Al framing of letters to expreſſe Antichriſts name, is
vncertaine.}
as the ancient Expoſitours & other doe thinke (for certaine
knowledge thereof no mortal man can haue without an expreſſe reuelation)
that his name conſiſteth of ſo many, & ſuch letters in Greek, as
according to their manner of numbring by the Alphabet make 666. and
foraſmuch as the letters making that number, may be found in diuers
names both proper and common; (as
\CNote{\Cite{Iren. l.~5. in fine.}}
S.~Irenæus findeth them in \Emph{Latinos} and \Emph{Teitan}, Hippolytus
in \G{αρνιυμασ}, Aretas in \Emph{Lampetis}, and ſome in this Age in
\Emph{Luderus}, which was Luthers name in the Alman tongue:) therfore we
ſee there can be no certaintie, and euery one frameth and applieth the
letters to his owne purpoſe. And moſt abſurd folly it is of the
Heretikes, to applie the word \Emph{Latinos}, to the Pope: neither the
whole order in common, nor euer any particular Pope being ſo called. And
S.~Irenæus the firſt that obſerued it in that word, applied it to the
Empire and ſtate of the Romane Emperour, which then was Heathen, and not
to the Pope of his daies or after him: and yet preferred the
word, \Emph{Teitan}, as more agreable, with this admonition, that it
were a very perilous and preſumptuous thing to define any certaintie
before-hand, of that number and name. And truely whatſoeuer the
Proteſtants preſume herein of the Pope, we may boldly diſcharge Luther
of that dignitie. He is vndoubtedly one of Antichriſts Precurſours, but
not Antichriſt himſelf.}
the number of him is ſixe hundred ſixtie ſixe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

