%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 2953
%%% o-2815
\startChapter[
  title={Chapter 9}
  ]

\Summary{The
\SNote{Moſt vnderſtãd al this of Heretikes.}
fifth Angel ſounding the trompet, a ſtarre falleth. 3.~The iſſuing forth
of locuſtes from the ſmoke of the deep pitte to vexe men, 7.~and the
deſcription of them. 13.~The ſixt Angel ſounding, foure Angels are let
looſe, 18.~which with a great troup of horſmen doe murder the third part
of men.}

And the fifth Angel ſounded with the trompet, & I ſaw
\SNote{The fal of an Arch-heretike, as Arius, Luther, Caluin, out of the
Church of God which haue the key of Hel to open & bring forth al the old
cõdemned hereſies buried before in the depth.}
a ſtarre to haue fallen from heauen vpon the earth, and there was giuen
to him the key of the pit of bottomles depth. \V And he openeth the pit
of the bottomles depth: and the ſmoke of the pit aſcended, as the ſmoke
of a great fornace: and the ſunne was darkned and the aier with the
ſmoke of the pit. \V And from the ſmoke of the pit there iſſued forth
\SNote{Innumerable pety heretikes following their Maiſters after the
opening & the ſmoke of the bottomleſſe pit.}
locuſts into the earth. And power was giuen to them, as the ſcorpions of
the earth haue power: \V and it was commanded them that they ſhould not
hurt the graſſe of the earth
\LNote{Nor any green thing.}{The
\MNote{Who are ſeduced by Heretikes.}
Heretikes neuer hurt or ſeduce the green tree, that is, ſuch as haue a
liuing faith working by charitie, but commonly they corrupt him in faith
who should otherwiſe haue perished for il life, and him that is
reprobate, that hath neither the ſigne of the Croſſe (which is Gods
marke) in the forehead of his body, nor the note of election in his
ſoule.}
nor any green thing, nor any tree: but only men which haue not the ſigne
of God
%%% 2954
in their foreheads. \V And it was giuen vnto thẽ that they ſhould not
kil them: but that they ſhould be tormented fiue moneths: and their
torments as the torments of a ſcorpion when he ſtriketh a man. \V And
\CNote{\XRef{Apoc.~6,~16.}}
in thoſe daies men ſhal ſeeke for death, and ſhal not find it: and they
ſhal deſire to die, & death ſhal fly from them.

\V And the ſimilitudes of the locuſts, like to horſes
\LNote{Prepared into battel.}{Heretikes
\MNote{The manifold hypocriſie of Heretikes.}
being euer ready to contend, doe pretend victorie, and counterfeit gold:
in shape as men, as ſmothe and delicate as women, their tongues and
pennes ful of gal and venim: their harts obdurate: ful of noiſe and
shuffling: their doctrine as peſtiferous and ful of poiſon, as the taile
and ſting of a ſcorpion; but they endure for a litle ſeaſon.}
prepared into battel: and vpon their heads as it were crownes like to
gold: and their faces as the faces of men. \V And they had haire as the
haire of women; & their teeth were as of Liõs. \V And they had
\TNote{coate of male, or other coate of offence.}
habbergions as habbergiõs of yron, and the voice of their wings as the
voice of the chariots of many horſes running into battel. \V And they had
tailes like to ſcorpions, and ſtings were in their tailes: and their
power was to hurt men fiue months. \V And they had ouer thẽ a King,
\SNote{The cheefe Maiſter of heretikes.}
the Angel of the bottomles depth, whoſe name in Hebrew is
\TNote{\H{אֲבַדּוֹן} \G{Ἀπολλύων}, in English \Emph{Deſtroyer}.}
\HH{Abaddon}, and in Greek \GG{Apollyon}: in Latin hauing the
name \L{Exterminans}. \V One woe is gone, & behold two woes come yet
after theſe.

\V And the ſixt Angel ſounded with the trompet: and I
%%% o-2816
heard one voice
from the foure hornes of the golden altar, which is before the eyes of
God, \V ſaying to the ſixt Angel which had the trompet: Looſe the foure
Angels which are bound in the great riuer Euphrates. \V And the foure
Angels were looſed, who were prepared for an houre, & a day, and a
moneth, and a yeare: that they might kil the third part of men. \V And
the number of the armie of horſe-men was twentie thouſand times ten
thouſand. And I heard the number of them. \V And ſo I ſaw the horſes in
the viſion: and they that ſate vpon them, had habbergions of fire and of
hyacinth and brimſtone, and the heads of the horſes were as it were the
heads of Lions: and from their mouth proceedeth fire, and ſmoke, and
brimſtone.

\V And by theſe three plagues was ſlaine the third part of men, of the
fire and of the ſmoke and of the brimſtone, which proceeded from their
mouth. \V For the power of the horſes is in their mouth, and in their
tailes. For, their tailes be like to ſerpents, hauing heads: and in
theſe they hurt.

\V
\SNote{Pagans, Infidels, and ſinful impenitent Catholikes muſt be
condemned alſo.}
And the reſt of men which were not ſlaine with theſe plagues, neither
\SNote{This phraſe being the like both in Greek and Latin, ſignifieth ſuch
ſorowful & penal repentance as cauſeth a man to forſake his former
ſinnes and depart from them. \G{οὐ μετενόησαν ἐκ τῶν ἔργων}. See the
ſame phraſe
\XRef{c.~2,~21,~22.}
&
\XRef{Act.~8. v.~22.}}
haue done penance from the workes of their hands, not to adore Diuels
and
\LNote{Idols of gold.}{Here
\MNote{Heret. tranſlation.}
againe the new Tranſlatours abuſe the people, for \Emph{idols} ſaying
\Emph{images}: this place being plainely againſt the pourtraits of
Heathen Gods, which are here and in the
\XRef{Pſalme~95.}
called, \L{dæmonia}, \Emph{Diuels}.}
Idols of gold and ſiluer and braſſe and ſtone and wood, which neither
can ſee, nor heare, nor walke, \V and haue not done penance from their
murders, nor from their ſorceries, nor from their fornication, nor from
their thefts.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

