%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 2965
%%% o-2826
\startChapter[
  title={Chapter 15}
  ]

\Summary{2.~They
\MNote{\Sc{The 4.~Part}.

Of the 7.~laſt plagues and final damnation of the wicked.}
that had now ouercome the beaſt and his image and the number of his
name, doe glorifie God. 6.~To ſeuen Angels hauing the ſeuen laſt
plagues, are giuen ſeuen cups ful of the wrath of God.}

%%% o-2827
And
\MNote{The tribulations about the day of iudgement.}
I ſaw another ſigne in heauen great & maruelous: ſeuen Angels hauing
the ſeuen laſt plagues. Becauſe in them the wrath of God is conſummate. \V
And I ſaw as it were
\SNote{Baptiſme.}
a ſea of glaſſe mingled with fire, & them that ouercame the beaſt and
his image and the number of his name, ſtanding vpon the ſea of glaſſe,
hauing the harps of God: \V and ſinging
\SNote{The ſong of Moyſes and Chriſt, is the new Teſtament and the old.}
the ſong of Moyſes the ſeruant of God, and the ſong of the Lamb, ſaying:
Great and maruelous are thy workes Lord God omnipotent: iuſt and true
are thy waies King of the
\Var{worlds.}{Saints.}
\V Who ſhal not feare thee, ô Lord, and magnifie thy name? Becauſe thou
only art holy, becauſe al Nations ſhal come, and adore in thy ſight,
becauſe thy iudgements be manifeſt.

\V
\MNote{\Sc{The
\Fix{First}{Fifth}{obvious typo, fixed in other}
Vision}.}
And after theſe things I looked, and behold the temple of the tabernacle
of teſtimonie was opened in heauen: \V and there iſſued forth the ſeuen
Angels, hauing the ſeuen plagues, from the temple: reueſted with cleane
and white
\Var{ſtone,}{linen,}
\TNote{\G{λίνον λίθον}}
and girded about the breaſts with girdles of gold. \V And one of the
foure beaſts, gaue to the ſeuen Angels ſeuen vials of gold ful of the
wrath of the God that liueth for euer and euer. \V And the temple was
filled with ſmoke at the maieſtie of God, and at his power: and no man
could enter into the temple, til the ſeuen plagues of the ſeuen Angels
were conſummate.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

