%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 2952
%%% o-2813
\startChapter[
  title={Chapter 8}
  ]

\Summary{1.~The ſeuenth ſeale being opened, there appeare Angels with
  trompets: 6.~and when an other Angel powred out fire taken from the
  altar, vpon the earth, there follow diuers tempeſtes. 7.~In like
  manner, whiles foure Angels of the ſeuen ſound their trompets, there
  fal ſundrie plagues.}

%%% o-2814
And
\MNote{\Sc{The 4.~Vision}}
when he had opened the ſeuenth ſeale, there was made ſilence in heauen,
as it were halfe an houre. \V And I ſaw ſeuen Angels ſtanding in the
ſight of God: and there were giuen to them ſeuen trompets. \V And
another Angel came and ſtood
\SNote{The Prieſt ſtanding at the altar praying & offering for the
people in the time of the high myſteries, Chriſt himſelf alſo being
preſent vpon the altar, is a figure of this thing, & thereunto he
alludeth.}
before the altar, hauing a golden cenſar: and there were giuen to him
many incenſes that he ſhould giue of the praiers of al Saints vpon the
altar of gold,
%%% 2953
which is before the throne of God. \V And the ſmoke of the incenſes
%%% !!! Originally a SNote
\LNote{The praiers of the Saints}{If this be S.~Michael or any Angel, and
not Chriſt himſelf, as ſome take it, Angels offer vp the praiers of the
faithful, as the 24.~Elders did
\XRef{c.~5.}
for this word, \Emph{Saints}, is taken here for the holy perſõs on
earth, as oftẽ in the Scripture: though it be not againſt the
Scriptures, that the inferiour Saint or Angel in heauen should offer
their praiers to God by their ſuperiours there. But hereby we conclude
agaĩſt the Proteſtants, that it derogateth not from Chriſt, that Angels
or Saints offer our praiers to God, as alſo it is plaine of Raphael.
\XRef{Tob.~12,~12.}}
of the praiers of the Saints aſcended from the hand of the Angel before
God. \V And the Angel tooke the cenſar, & filled it of the fire of the
altar, and caſt it on the earth, and there were made thunders & voices &
lightnings, and a great earth-quake. \V And the ſeuen Angels which had
the ſeuen trompets, prepared themſelues to ſound with the trompet.

\V And the firſt Angel ſounded with the trompet, and there was made
haile and fire, mingled in bloud, and it was caſt on the earth, and the
third part of the earth was burnt, & the third part of trees was burnt,
and al green graſſe was burnt.

\V And the ſecond Angel ſounded with the trumpet: and as it were a great
moũtaine burning with fire, was caſt into the ſea, & the third part of
the ſea was made bloud: \V and the third part of thoſe creatures died,
which had liues in the ſea, and the third part of the ships perished.

\V And the third Angel ſounded with the trompet, and a great ſtarre fel
from heauen, burning as it were a torch, and it fel on the third part of
the flouds, and on the fountaines of waters: \V and the name of the
ſtarre is called worme-wood. And the third part of the waters was made
into worme-wood: and many men died of the waters, becauſe they were made
bitter.

\V And the fourth Angel ſounded with the trumpet, and the third part of
the ſunne was ſmitten, and the third part of the moon, and the third
part of the ſtarres, ſo that the third part of them was darkned, and of
the day there shined not the third part, and of the night in like
manner. \V And I looked, and heard the voice of one egle flying through
the middes of heauẽ, ſaying with a loud voice: Woe, woe, woe to the
%%% o-2815
inhabiters on the earth: becauſe of the reſt of the voices of the three
Angels which were to ſound with the trompet.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

