%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 2946
%%% o-2806
\startChapter[
  title={Chapter 4}
  ]

\Summary{1.~A
\MNote{\Sc{The 2.~Part.}

Firſt, the booke with 7.~ſeales. Secondly, 7.~Angels with trompets.}
dore being open in heauen he ſaw one ſitting in a throne, 4.~and round
about him foure and twentie Seniours ſitting, 6.~and the foure beaſts
here deſcribed, 9.~which with the 24.~Seniours continually glorified him
that ſate in the throne.}

%%% o-2807
After
\MNote{\Sc{The 2.~Vision}.

In which is repreſented vnto vs the glorie and maieſtie of God in
heauen, and the inceſſant honour & praiſes of al Angels and Saints
aſſiſting him. Which is reſembled in the daily honour done to him by al
orders and ſorts of holy men in the Church militant alſo.}
theſe things I looked, & behold a doore open in heauen, and the firſt
voice which I heard, was as it were of a trompet ſpeaking with me,
ſaying: Come vp hither, and I wil ſhew thee the
\Fix{thigs}{thinges}{obvious typo, fixed in other}
which muſt be done
quickly after theſe. \V And immediately I was in ſpirit: and behold
there was a ſeat ſet in heauen, and vpon the ſeat one ſitting. \V And he
that ſate, was like in ſight to the Iaſper ſtone, and the Sardine: and
there was a raine-bow round about the ſeat, like to the ſight of an
Emeraud. \V And round about the ſeat, foure and twentie ſeats: and vpon
the thrones, foure and twentie Seniours ſitting, clothed about in white
garments, and on their heads crownes of gold. \V And from the throne
proceeded lightnings, and voices, and thunders: and ſeuen lamps burning
before the throne, which are the ſeuen Spirits of God. \V And in the
ſight of the ſeat, as it were a ſea of glaſſe like to chryſtal: and in
the middes of the ſeat and round about the ſeat
\SNote{Theſe foure beaſts, and the like deſcribed
\XRef{Ezech.~1.}
by the iudgement of the holy Doctours ſignifie the 4.~Euangeliſts, and
in them al true Preachers.

The man Mathew: the Lion, Marke: the Calfe, Luke: the Egle, Iohn. See
the cauſes hereof in the
\Cite{Summe of the 4.~Eunageliſts. Pag.~1. S.~Grego. in 1.~Ezech.}}
foure beaſts ful of eyes before and behind. \V And the firſt beaſt, like
to a Lion: and the ſecond beaſt, like to a Calfe: and the third beaſt,
hauing the face as it were of a man: and the fourth beaſt, like to an
Egle flying. \V And the foure beaſts, euery one of them had ſixe wings round
about: and within they are ful of eyes. And they had no reſt day and
night, ſaying:
\CNote{\XRef{Eſ.~6,~3.}}
\LNote{Holy, holy, holy.}{This
\MNote{The \L{Sanctus} thriſe repeated.}
word is thriſe repeated here, and
\XRef{Eſa.~6:}
and to the imitation therof, in the Seruice of the holy Church, at \L{Te
Deum}, and at Maſſe, ſpecially in the Preface next before the great
myſteries, for the honour of the three Perſons in the B.~Trinitie, and
that the Church militant may ioyne with the triumphant, & with al the
Orders of Angels, who alſo are preſent at the conſecration, and doe
ſeruice there to our common Lord and Maiſter, as S.~Chryſoſtom writeth,
\Cite{li.~6. de Sacerdotio,}
and
\Cite{ho.~1. de verb. Eſa. to.~1.}
The Greeks cal it, the hymne \GG{Triſagios}, that is, \Emph{Thriſe holy}.}
\Emph{Holy, Holy, Holy, Lord God omnipotent}, which was, and which is,
and which ſhal come. \V And when thoſe beaſts gaue glorie and honour and
benediction to him that ſitteth
%%% 2947
vpon the throne, that liueth for euer and euer: \V the foure and twentie
Seniours fel downe before him that ſitteth in the throne, & adored him
that liueth for euer and euer, and caſt their crownes before the throne,
ſaying: \V Thou art worthie O Lord our God to receiue glorie and honour
and power: becauſe thou haſt created al things, and for they wil they
were and haue been created.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

