%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 2949
%%% o-2809
\startChapter[
  title={Chapter 6}
  ]

\Summary{1.~Foure ſeales of the ſeuen being opened, there follow diuerſe
  effects againſt the earth. 9.~When the fifth ſeale was opened, the
  ſoules of martyrs deſire that the iudgement may be haſtned: 12.~and at
  the opening of the ſixt, there are ſignes shewed of the iudgement to
  come.}

And I ſaw that the Lamb had opened one of the ſeuen ſeales, and I heard
one of the foure beaſts, ſaying, as it were the voice of thunder: Come,
and ſee. \V And I ſaw: And behold a white horſe, and he that ſate vpon
him had a bow, and there was a crowne giuen him, and he went forth
conquering that he might conquer.

%%% o-2810
\V And when he had opened the ſecond ſeale, I heard the ſecond beaſt,
ſaying: Come, & ſee. \V And there went forth an other horſe, redde: and
he that ſate thereon, to him it was giuen that he should take peace from
the earth, and that they should kil one another, and a great ſword was
giuen to him.

\V And when he had opened the third ſeale, I heard the third beaſt,
ſaying: Come, and ſee. And behold a black horſe, and he that ſate vpon
him, had a balance in his hand. \V And I heard as it were a voice in the
middes of the foure beaſts ſaying: Two pounds of wheat for a penie, and
thriſe two pounds of barley for a penie, and wine and oile hurt thou
not.

\V And when he had opened the fourth ſeale, I heard a voice of the fourth
beaſt, ſaying: Come, & ſee. \V And behold a pale horſe: and he that ſate
vpon him, his name was death, and hel followed him. And power was giuen
to him ouer the foure parts of the earth, to kil with ſword, with
famine, and with death, and with beaſts of the earth.

\V And when he had opened the fifth ſeale: I ſaw
\LNote{Vnder the altar.}{Chriſt
\MNote{Conſecration of altars with Saints relikes.}
as man (no doubt) is this altar, vnder which the ſoules of al Martyrs
liue in heauen expecting their bodies, as Chriſt their Head hath his
body there already. And for correſpondence to their place or ſtate in
heauen, the Church laieth cõmonly their bodies alſo or relikes neer or
vnder the altars, where our Sauiours body is offered in the holy Maſſe:
and hath a ſpecial prouiſo that no altars be erected or conſecrated
without ſome part of a Saints body or relikes.
\Cite{Conc. African. can.~50.}
\Cite{Carthag.~5. can.~14.}
See S.~Hierom
\Cite{cont. Vigilant. c.~3.}
S.~Auguſtin.
\Cite{de ciuit li.~8. c.~27.}
S.~Gregorie
\Cite{li.~5. ep.~50.}
\Cite{li.~1. ep.~52.}
\Cite{li.~2. ep.~58,}
Whereunto the Prophet ſeemeth here to allude, making their ſoules alſo
to haue their being in heauen, as it were vnder the altar. But for this
purpoſe note wel the words of S.~Auguſtin (or what other ancient Writer
ſoeuer was the Authour thereof)
\Cite{Ser.~11. de Sanctis.}
\Emph{Vnder the altar} (ſaith he) \Emph{of God I ſaw the ſoules of the
ſlaine. What is more reuerent or honourable, then to reſt vnder that
altar on which Sacrifice is done to God, and in which our Lord is the
Prieſt: as it is written, Thou art a Prieſt according to the Order of
Melchiſedech? Rightly doe the ſoules of the iuſt reſt vnder the altar,
becauſe vpon the altar our Lords body is offered. Neither without cauſe
doe the iuſt there cal for reuenge of their bloud, where alſo the bloud
of Chriſt is shed for ſinners.} And many other goodly words to that
purpoſe.

This
\MNote{Saints be preſent at their tombs and relikes.}
place alſo the wicked heretike Vigilantius (as S.~Hierom writing againſt
him witneſſeth
\Cite{c.~2.)}
abuſed, to proue that the ſoules of Martyrs and other Saints were
included in ſome certaine place, that they could not be preſent at their
bodies and monuments (where Chriſtian people vſed in the primitiue
Church to pray vnto them, as Catholike men doe yet) nor be where they
liſt, or where men pray vnto them. To which the holy doctour anſwereth
at large, that they be whereſoeuer Chriſt is according to his humanitie:
for vnder that altar they be.
\MNote{The Caluiniſts hereſie concerning the Saints confuted by
S.~Hierom long agoe.}
Part of his words be theſe, that you may ſee how this Bleſſed Father
refuted in that Heretike the Caluiniſtes ſo long before they were borne.
\Emph{Doeſt thou} (ſaith he) \Emph{preſcribe lawes to God? Doeſt thou
fetter the Apoſtles, that they may be kept in priſon til the day of
iudgement, and be kept from their Lord, of whom it is written,
\CNote{\XRef{Apoc.~14.}}
They follow the Lamb whitherſoeuer he goeth? If the Lamb be in euery
place, then they that be with the Lamb, muſt be euery-where. And if the
diuel and wicked ſpirits gadding abrode in the world with paſsing
celeritie, be preſent euery-where; shal holy Martyrs after the sheading
of their bloud, be kept cloſe vnder an altar that they can not ſturre
out from thence?} So anſwereth this learned Doctour.

Which
\MNote{They vnlearnedly accuſe S.~Hierom as an Vbiquiſt.}
miſliketh our Caluiniſts ſo much, that they charge him of great errour,
in that he ſaith, Chriſt according to his humanitie is euery-where, as
though he were an Vbiquetarie Proteſtant. Where if they had any
iudgement, they might perceiue that he meaneth not, that Chriſt or his
Saints should be perſonally preſent at once in euery place alike, as God
is:
\MNote{How S.~Hierõ ſaith, Chriſt & his Saints are
\Fix{euery.}{euery-where.}{obvious typo, fixed in other}}
but that their motion, ſpeed, and agilitie to be where they liſt, is
incomparable, and that their power and operation is accordingly. Which
they may learne to be the holy Doctours meaning, by the words that
follow of the Diuel and his miniſters: whom he affirmeth to be
euery-where no otherwiſe but by their exceeding celeritie of being and
working miſcheefe now in one place, now in another, and that in a
moment. For though they be ſpirits, yet are they not euery-where at once
according to their eſſence. And for our new Diuines it were a hard thing
to determine, how long Satan (that
\CNote{\XRef{Iob.~1.}}
told our Lord he had circuited the earth) was in his iourney, and in the
particular conſideration and tentation of Iob: and how many men he
aſſaulted in that his one circuit. No, no, ſuch curious companions know
nothing, nor beleeue nothing, but that they ſee with corporal eyes, and
teach nothing but the way to infidelitie.}
vnder the altar the ſoules of them that were ſlaine for the word of God,
and for the teſtimonie which they had. \V
\LNote{And they cried.}{S.~Hierom
\MNote{That Saints pray for vs, S.~Hierom proueth againſt the Heretike
Vigilantius.}
alſo againſt the ſaid Vigilantius reporteth that he vſed an argument
againſt the praiers of Saints out of this place, for that theſe Martyrs
cried for reuenge & could not obtaine. But we wil report his words, that
you may ſee how like one Heretike is to another, theſe of our daies to
thoſe of old. \Emph{Thou ſaieſt in my booke} (ſaith S.~Hierom
\Cite{c.~3.)}
\Emph{that whiles we be aliue, one of vs may pray for another: but after
we be dead, no mans praier shal be heard for another: ſpecially ſeeing
the Martyrs asking reuenge of their bloud, could not obtaine.} So ſaid
the Heretike. Againſt which the holy Doctour maketh a long refutation,
prouing that they pray much more after they be in heauen, then they did
here in earth: and that they shal be much ſooner heard of God, then when
they were in the world.

But for the Heretikes argument framed out of theſe words of the
Apocalypſe thus, \Emph{Theſe Martyrs did not obtaine, ergo Saints doe
not pray for vs}; it was ſo friuolous, and the antecedent ſo manifeſtly
falſe that he vouchſafed not to ſtand about it. For it is plaine that
the Martyrs here were heard, and that their petition should be fulfilled
in time appointed by God (whereunto they did and doe alwaies conforme
themſelues:) for it was ſaid vnto them, \Emph{That they should reſt yet
a litle time til, &c.} And that Martyrs praiers be heard in this caſe,
our Sauiour teſtifieth,
\XRef{Luc.~18.}
ſaying, \Emph{And wil not God reuenge his elect that crie to him day and
night? I ſay to you, he wil quickly reuenge them.} And if God doe not
heare the Saints ſometime nor grant their requeſts, is it therfore
conſequent that they doe not or may not pray? Then Chriſt himſelf should
not haue praied his Father to remoue the bitter cup of death from him,
becauſe that petition was not granted.}
And they cried with a loud voice, ſaying: How long Lord (holy and true)
iudgeſt thou not and
\LNote{Reuengeſt thou not.}{They
\MNote{How Martyrs crie for reuenge.}
doe not deſire reuenge vpon their enemies for hatred, but of charitie
and zeale of Gods honour, praying that his enemies & the perſecutours of
his Church and Saints, that wil not repent, may be confounded: and that
our Lord would accelerate his general iudgement, that ſo they might
attaine the perfect crowne of glorie promiſed vnto them, both in body
and ſoule: which is to deſire the reſurrectiõ of their bodies, which
then shal triumph perfectly and fully ouer the perſecutours that ſo
cruelly handled the bodies of the elect, which shal then appeare
glorious, to the enemies confuſion.}
reuengeſt thou not our bloud of them that dwel on the earth? \V And
white ſtoles were giuen, to euery one of them
\SNote{This one ſtole ſignifieth the glorie or bliſſe of the ſoule only:
but at the day of iudgemẽt they shal haue it doubled by adding the
glorie of their body alſo.}
one; and it was ſaid to them, that they should reſt yet a litle time,
\LNote{Til their fellow ſeruants be complete.}{There is a certaine
number that God hath ordained to die for the teſtimonie of truth and the
Catholike faith, for conformitie of the members to the
Head \Sc{Christ} our cheefe Martyr. And til that number be accomplished,
the general condemnation of the wicked perſecutours shal not come, nor
the general reward of the elect.}
til their fellow-ſeruants be complete, and their Brethren, that are to
be ſlaine euen as they.

\V And I ſaw, when he had opened the ſixt ſeale, and
\SNote{The tribulation that shal fal in the time of Antichriſt.}
behold there was made a great earth-quake, and the ſunne became black as
it were ſack-cloth of haire: and the whole moon became as bloud: \V and
the ſtarres from heauen fel vpon the earth, as the figge-tree caſteth
her green figges when it is shaken of a great wind: \V and heauen
departed as a booke folded together: and euery hil, and Ilands were
moued out of their places. \V And the Kings of the earth, and Princes,
and Tribunes, and the rich, and the ſtrong, and euery bond-man, and
%%% 2950
free-man
\CNote{\XRef{Oſee.~10.}}
hid themſelues in the dennes and the rocks of mountaines. \V And they
ſay to the
%%% o-2811
mountaines and the rocks:
\CNote{\XRef{Lu.~23,~50.}}
Fal vpon vs, and hide vs from the face of him that ſitteth vpon the
throne, and from the wrath of the Lamb: \V becauſe the great day of
their wrath is come, and who ſhal be able to ſtand?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

