%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 2970
%%% o-2832
\startChapter[
  title={Chapter 18}
  ]

\Summary{The fal of Babylon, her iudgement, plagues and reuenges: for
  the which, 9.~the Kings, 16.~and marchants of the earth that ſometime
  did cleaue vnto her, shal mourne bitterly: 20.~but heauen, and the
  Apoſtles and Prophets shal reioyce.}

And after theſe things I ſaw another Angel comming downe from heauen,
hauing great power: and the earth was illuminated of his glorie. \V And
he cried out in force, ſaying:
\CNote{\XRef{Apoc.~14,~8.}}
Fallen fallen is Babylon the great: and it is become the habitation of
Diuels, and
\TNote{\G{φυλακὴ}}
the cuſtodie of euery vncleane Spirit, & the cuſtodie of euery vncleane
& hateful bird; \V becauſe al Nations haue drunke of the wine of the
wrath of her fornication: and the Kings of the earth haue fornicated
\Fix{whith}{with}{likely typo, fixed in other}
her: and the marchants of the earth were made rich by the vertue of her
delicacies.

\V And I heard another voice from heauen, ſaying: Goe out from her my
people: that you be not partakers of her ſinnes, and receiue not of her
plagues. \V Becauſe her ſinnes are come euen to heauen, and God hath
remembred her iniquities. \V Render to her as she alſo hath rendred to
you: and double ye double according to her workes: In the cup wherein
ſhe hath mingled, mingle ye double vnto her. \V As much as ſhe hath
glorified her ſelf, & hath been in delicacies,
\SNote{The meaſure of paines and damnation, according to the wicked
pleaſures, or vnlawful delights of this life. Which is a ſore ſentence
for ſuch people as turne their whole life to luſt and riot.}
ſo much giue her torment and mourning: becauſe ſhe ſaith in her hart,
\CNote{\XRef{Eſ.~47,~8.}}
I ſit a Queen, and widow I am not, and mourning I ſhal not ſee. \V
Therfore in one day ſhal her plagues come, death, and mourning, and
famine, & with fire ſhe ſhal be burnt: becauſe God is ſtrong that shal
iudge her.

%%% o-2833
\V And
\SNote{Kings and Marchants are moſt encombered, dãgered & drowned in
the pleaſures of this world: whoſe whole life & traficke is (if they be
not exceeding vertuous) to find varietie of earthly pleaſures. Who
ſeeing once the extreme end of their ioyes and of al that made their
heauẽ here, to be turned into paines & damnation eternal, then shal
houle & weep too late.}
the Kings of the earth, which haue fornicated with her, & haue liued in
delicacies, ſhal weep, and bewaile themſelues vpon her, when they ſhal
ſee the ſmoke of her burning: \V ſtanding farre off for the feare of her
torments, ſaying: Woe, woe, that great citie Babylon, that ſtrong citie:
becauſe in one houre is thy iudgement come.

\V And the marchants of the earth ſhal weep, & morne vpon her: becauſe
no man ſhal buy their merchandiſe any more, \V merchandiſe of gold and
ſiluer and precious ſtone: and of pearle, and fine linnen, and purple,
and ſilke, & ſcarlet and al Thyne wood, and al veſſels of yuorie, and al
veſſels of precious ſtone and of braſſe and yron and marble, \V and
cynamon, and of odours, and ointment, and frankincenſe, and wine, and
oile, and floure, & wheat, and beaſts, and ſheep, and horſes, and
chariots, & ſlaues, and ſoules of men. \V And the apples of the deſire
of thy ſoule are departed from thee, and al fat and goodly things are
periſhed from thee, and they ſhal no more find them. \V The marchants of
theſe things which are made rich, ſhal ſtand farre from her for feare of
her tormentes, weeping and mourning, \V & ſaying: Woe, woe, that great
citie, which was clothed with ſilke, and purple, and ſcarlet, and was
gilted with gold, and pretious ſtone, and pearls: \V becauſe
%%% 2971
in one houre are ſo great riches made deſolate: and euery gouernour, &
euery one that ſaileth into the lake, and the ſhip-men, and they that
worke in the ſea, ſtoode a farre off, \V and cried ſeeing the place of
her burning, ſaying: What other is like to this great citie? \V And they
threw duſt vpon their heads, and cried weeping and mourning, ſaying:
Woe, woe, that great citie, in the which al were made rich that had
ſhips in the ſea, of her prices: becauſe in one houre she is deſolate.

\V
\SNote{The Angels and al Saints shal reioyce & laud God to ſee the
wicked confunded, & Gods iuſtice executed vpon their oppreſſours &
perſecutours, & this is that which the Martyrs praied for,
\XRef{c.~6.}}
Reioyce ouer her, heauen, and ye holy Apoſtles and Prophets: becauſe God
hath iudged your iudgement of her. \V And one ſtrong Angel tooke vp as
it were a great milſtone, and threw it into the ſea, ſaying:
\CNote{\XRef{Ier.~51,~63.}}
With this violence ſhal
\SNote{By this it ſeemeth cleere that the Apoſtle meaneth not any one
citie, but the vniuerſal companie of the reprobate, which shal perish in
the day of iudgemẽt. The old Prophets alſo naming the whole number of
Gods enemies myſtically, Babylon, as
\XRef{Ierem. c.~52.}}
Babylon that great citie be throwen, and ſhal now be found no more. \V
And the voice of harpers, and of Muſicians, and of them that ſing on
ſhalme and trompet, ſhal no more be heard in thee, & euery artificer of
euery art ſhal be found no more in thee, and the noiſe of the mil ſhal
no more be heard in thee, \V and the light of the lamp ſhal no more
%%% o-2834
ſhine in thee, & the voice of the bride-grome and the bride ſhal no more
be heard in thee: becauſe thy marchants were the Princes of the earth,
becauſe al Nations haue erred in thine inchantments. \V And in her is
found the bloud of the Prophets and Saints, and of al that were ſlaine
in the earth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

