%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 2979
%%% o-2841
\startChapter[
  title={Chapter 21}
  ]

\Summary{Heauen
\MNote{\Sc{The 5.~Part.}

The final glorification of the Church.}
and earth being made new, S.~Iohn ſeeth the new citie Hieruſalem
prepared and adorned for the ſpouſe of the Lamb. 6.~The iuſt are
glorified, 7.~and the wicked thruſt into the poole of fire. 12.~The wal
and gates and foundations of the citie are deſcribed and meaſured:
18.~al which are gold and cryſtal, pretious ſtones and pearles.}

And I ſaw a new heauen and a new earth. For
\CNote{\XRef{Eſa.~63,~17.}
\XRef{66,~21.}
\XRef{2.~Pet.~3,~13.}}
the firſt heauen, and the firſt earth was gone, & the ſea now is not. \V
And I Iohn ſaw
\SNote{The Church triumphant.}
the holy citie Hieruſalem new deſcending from heauen, prepared of God,
as a bride adorned for her husband. \V And I heard a loud voice from the
throne ſaying: Behold
\SNote{This tabernacle is Chriſt according to his humanitie.}
the tabernacle of God with men, and he wil dwel with them. And they ſhal
be his people: and he God with them ſhal be their God. \V And
\CNote{\XRef{Eſa.~25,~8.}
\XRef{Apoc.~7,~17.}}
\SNote{This happie day shal make an end of al the miſeries of this
mortalitie.}
God ſhal wipe away al teares from their eyes: and death ſhal be no
more. Nor mourning, nor crying, neither ſhal there be ſorow any more,
\Var{which}{becauſe the}
firſt things, are gone.

%%% o-2842
\V And he that ſate in the throne, ſaid:
\CNote{\XRef{Eſa.~43,~19.}}
Behold I make al things new. And he ſaid to me: Write, becauſe theſe
wordes be moſt faithful and true. \V And he ſaid to me: It is done,
\CNote{\XRef{Apoc.~1,~8.}
\XRef{22,~13.}}
I am Alpha and Omega: the beginning and the end. To him that thirſteth I
wil giue of the fountaine of the water of life, gratis. \V He that ſhal
ouercome,
\SNote{He that hath the victorie againſt ſinne in the Church militant,
shal haue his reward in the triumphant.}
ſhal poſſeſſe theſe things, and I wil be his God: and he ſhal be my
ſonne. \V But
\SNote{Al that commit mortal ſinnes and repent not, shal be damned.}
to the feareful, and incredulous, and execrable, and murderers, and
fornicatours, and ſorcerers, and Idolaters, and al liers, their part
ſhal be in the poole burning with fire and brimſtone, which is the ſecond
death.

\V
\MNote{\Sc{The 7.~and Last Vision}.}
And there came one of the ſeuen Angels that had the vials of the ſeuen
laſt plagues, and ſpake with me, ſaying: Come, and I wil ſhew thee the
bride, the wife of the Lamb. \V And he tooke me vp in ſpirit
%%% 2980
vnto a mountaine great and high: and he ſhewed me the holy citie
Hieruſalem deſcending out of heauen from God, \V
\SNote{The glorie of the Church triumphant.}
hauing the glorie of God, and the light thereof like to a pretious
ſtone, as it were to the iaſper ſtone, euen as cryſtal. \V And it had a
wal great and high, hauing twelue gates, and in the gates twelue Angels,
& names written theron, which are
\SNote{The names of the Patriarches and Apoſtles honourable and glorious
in the triumphant Church.}
the names of the twelue Tribes of the children of Iſrael. \V On the
Eaſt, three gates: and on the North, three gates: and on the South,
three gates: and on the Weſt, three gates. \V And the wal of the citie
hauing twelue foundations: and in them, twelue names of the twelue
Apoſtles of the Lamb.

\V And he that ſpake with me, had a meaſure of a reed, of gold, to
meaſure the citie and the gates thereof, and the wal. \V And
\SNote{See S.~Hierom
\Cite{ep.~17.}
touching this deſcription of the heauenly Hieruſalem, which is the
Church triumphant, teaching that theſe thĩgs muſt be taken ſpiritually,
not carnally.}
the citie is ſituated quadrangle-wiſe, and the length thereof is as
great as alſo the bredth: and he meaſured the citie with the reed for
twelue thouſand furlongs, & the length and height and bredth thereof be
equal. \V And he meaſured the wal thereof of an hundred fourtie foure
cubits, the meaſure of a man which is of an Angel. \V And the building
of the wal thereof was of iaſper ſtone: but the citie it-ſelf
\LNote{Pure gold.}{S.~Gregorie
\MNote{The ſtate of glorified bodies.}
\Cite{(li.~18. Moral. c.~28.)}
ſaith, the heauenly ſtate is reſembled to gold, pretious ſtone, cryſtal,
glaſſe, and the like, for the puritie, claritie, glittering of the
glorious bodies: where one mans body, conſcience, and cogitations are
repreſented to another, as corporal things in this life are ſeen through
cryſtal or glaſſe.}
pure gold, like to pure glaſſe. \V And the foundations of the wal of the
citie, were adorned with al pretious ſtone. The firſt foundation, the
iaſper: the ſecond, the ſaphire: the third, the calcedonius: the fourth,
the emerauld: \V the fifth, the Sardonix: the ſixt, the ſardius: the
ſeuenth, the chryſolithus: the eight, the beryllus: the ninth, the
topazius: the tenth, the
%%% o-2843
chryſopraſus: the eleuenth, the hyacinth: the twelfth, the amethyſt. \V
And the twelue gates: there are twelue pearles, one to euery one: &
euery gate was of one ſeueral pearle. And the ſtreet of the citie pure
gold, as it were tranſparent glaſſe. \V And
\SNote{Al external Sacrifice which now is neceſſarie dutie of the
faithful, shal then ceaſe, and therfore there shal need no material
tẽple.}
temple I ſaw not therein. For our Lord the God omnipotent is the temple
thereof and the Lamb. \V And
\CNote{\XRef{Iſa.~60,~19.}}
the citie needeth not ſunne nor moone, to ſhine in it. For the glorie of
God hath illuminated it, and the Lamb is the lamp thereof. \V And
\CNote{\XRef{Iſa.~60,~3.}}
the Gentils ſhal walke in the light of it: and the Kings of the earth
ſhal bring their glorie and honour into it. \V And
\CNote{\XRef{Eſa.~60,~21.}}
the gates thereof ſhal not be ſhut by day: for there ſhal be no night
there. \V And they ſhal bring the glorie and honour of Nations into
it. \V There ſhal
\SNote{None not perfectly cleanſed of their ſinnes, can enter into this
heauenly Hieruſalem.}
not enter into it any polluted thing, nor that doeth abomination and
maketh lie, but
\CNote{\XRef{Apo.~20,~12.}}
they that are written in the booke of life of the Lamb.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

