%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Apocalypse
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-03


\project douay-rheims


%%% 2944
%%% o-2804
\startChapter[
  title={Chapter 3}
  ]

\Summary{He is commanded to write to the Churches of Sardis,
  Philadelphia, and Laodicia: recalling them that erre, to penance by
  threatning, but praiſing the reſt, and promiſing reward to him that
  ouercometh: 15.~deteſting alſo the cold indifferent Chriſtian. 20.~He
  ſaith that God knocketh at the doore of mens harts by offering his
  grace, for to enter in to him that wil open vnto him by conſent of
  free-wil.}

And to the Angel of the Church of Sardis, write: Thus ſaith he that hath
the ſeuen 
Spirits of God, and the ſeuen ſtarres: I know thy workes, that thou haſt
the name that thou liueſt, and thou art dead. \V Be vigilant, and
confirme the reſt of the things which were to die. For I find not thy
workes ful before my God. \V Haue in mind therfore in what manner thou
haſt receiued and heard:
%%% o-2805
and keep, and doe penance. If therfore thou watch not,
\CNote{\XRef{1.~Theſ.~5,~2.}
\XRef{2.~Pet.~3,~10.}
\XRef{Apo.~16,~15.}}
I wil come to thee as a theefe, and thou ſhalt not know what houre I wil
come to thee. \V But thou haſt a few names in Sardis,
\SNote{Such as haue not committed deadly ſinne after Baptiſme.}
which haue not defiled their garments: and they ſhal walke with me in
whites, becauſe they
\SNote{Note that there is in man a worthineſſe of the ioyes of heauen,
by holy life: and this is a common ſpeach in holy Scripture, that man is
worthy of God, of heauen, of ſaluation.}
are worthy. \V
\LNote{He that shal ouercome.}{In
\MNote{Doing wel in reſpect of reward.}
al theſe ſpeaches to diuers Bishops and their Churches he continually
encourageth them to conſtancie in faith and good life, by ſetting before
their eyes the reward of the next life. And yet the Caluiniſts would
haue no man doe good in reſpect of ſuch reward.}
He that ſhal ouercome, ſhal thus be veſted in white garments, and I wil
not put his name out of the booke of life, and I wil confeſſe his name
before my Father, and before his Angels. \V He that hath an eare, let
him heare what the Spirit ſaith to the Churches.

\V And to the Angel of the Church of Philadelphia, write: Thus ſaith the
Holy one and the True one, he that hath the
\CNote{\XRef{Eſa.~22,~22.}}
key of Dauid; he that openeth, and no man ſhutteth, ſhutteth, and no man
openeth:
%%% 2945
\V I know thy workes. Behold I haue giuen before thee a doore opened
which no man can ſhut: becauſe thou haſt a litle power, and haſt kept my
word, and haſt not denied my name. \V Behold I wil giue of the Synagogue
of Satan, which ſay they be Iewes, and are not, but doe lie. Behold I
wil make them come and
\LNote{Adore before thy feet.}{You
\MNote{Adoration of creatures, called \L{Dulia}.}
ſee this word of adoration is in Scriptures vſed for worship of
creatures alſo, and that to fal before the feet of holy men or Angels
for duety and reuerence, is not idolatrie, except the proper honour due
to God, be giuen vnto them. See the
\XRef{Annotations vpon the 19.~and 22.~Chapter}
concerning the Apoſtles proſtration before the Angel. And the
Aduerſaries euaſion, ſaying that the adoration was of God only: and
that, \Emph{before the feet} of the partie, ſignifieth nothing els but,
\Emph{in his preſence}, is falſe and againſt the phraſe of Scriptures:
as
\XRef{4.~Reg.~4.}
where the Sunamite adored Eliſæus, falling downe before his feet, and
\XRef{4.~Reg.~2.}
the Sonnes of the Prophets adored him in the ſame ſort. And here this
adoration can not be meant but of the Bishop or Angel of Philadelphia,
becauſe he promiſeth this honour as a reward, and as an effect of his
loue towards him, ſaying: \Emph{And they shal know that I haue loued
thee.} And that which he ſaith in the
\XRef{22.~chapter,}
\Emph{I fel downe to adore before the feet of the Angel}; the very ſame
he expreſſeth thus in the
\XRef{19.~chapter.}
\Emph{I fel before his feet to adore him}: making it al one, to adore
before his feet, and to adore him.}
adore before thy feet. And they ſhal know that I haue loued thee. \V
Becauſe thou haſt kept the word of my patience, and I wil keep thee from
the houre of tentation, which ſhal come vpon the whole world to tempt
the inhabitants on the earth. \V Behold I come quickly: hold that which
thou haſt,
\LNote{That no man take thy crowne.}{That
\MNote{Perſeuerance in good, and continuing to the end.}
is, his crowne of euerlaſting life and glorie, if he perſeuer not to the
end in faith & good workes: otherwiſe another shal enter into his place,
as Matthias did both to the dignitie of the Apoſtleship, & to the
heauenly crowne due for the wel vſing and executing of the ſame
function, which Iudas might and should haue had, if he had perſeuered to
the end: and as the Gentils came into the grace and place of the
Iewes. Other difficulties concerning this kind of ſpeach are reſolued in
School-men, and are not here to be ſtood vpon.}
that no man take thy crowne. \V He that ſhal ouercome, I wil make him a
piller in the temple of my God: and he ſhal goe out no more: and I wil
write vpon him the name of my God, and the name of the citie of my God,
new Hieruſalem, which deſcendeth out of Heauen from my God, and my new
name. \V He that hath an eare, let him heare what the Spirit ſaith to the
Churches.

\V And to the Angel of the Church of Laodicia write: Thus ſaith
\TNote{\G{ἀμήν}}
Amen, the faithful and true witneſſe,
\CNote{\XRef{Eccleſiaſtici~24,~9.~14.}
\XRef{Col.~1,~15.}}
which is the beginning of the creature of God: \V I know thy workes,
that thou art neither cold, nor hot. I would thou wert cold, or hot. \V
But becauſe thou art
\LNote{Luke-warme.}{Zeale
\MNote{Neuters or indifferents in religion.}
and feruour is commendable, ſpecially in Gods cauſe: and the Neuters
that be neither hot nor cold, are to Chriſt and his Church burdenous and
lothſome, as luke-warm water is to a mans ſtomake, prouoking him to
vomit: and therfore he threatneth to void vp ſuch Neuters out of his
mouth.}
luke-warme, and neither cold nor hot, I wil begin to vomit thee out of my
mouth. \V Becauſe thou ſaieſt, That I am rich, and enriched, and lacke
nothing; and knoweſt not that thou art a miſer, and miſerable, and
poore, and blind, and naked. \V I counſel thee to buy of me gold
fire-tried, that thou maieſt
%%% o-2806
be made rich: and maieſt be clothed in white garmentes, that the
confuſion of thy nakednes appeare not: and with eye-ſalue anoint thine
eyes, that thou maieſt ſee. \V I,
\CNote{\XRef{Prou.~3,~12.}
\XRef{Heb.~12,~6.}}
whom I loue, doe rebuke and chaſtiſe. Be zelous therfore and doe
penance. \V Behold I ſtand at the doore and
\SNote{God firſt calleth vpon man and knocketh at the doore of his hart,
that is to ſay, offereth his grace. And it lieth in man to giue conſent
by free-wil holpen alſo by his grace.}
knock. If any man ſhal heare my voice, and open to me the gate, I wil
enter in to him, and wil ſup with him, and he with me. \V He that ſhal
ouercome, I wil giue vnto him to ſit with me in my throne: as I alſo
haue ouercome, and haue ſitten with my Father in his throne. \V He that
hath an eare, let him heare what the Spirit ſaith to the Churches.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

