%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Two Peter
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-02


\project douay-rheims


%%% 2908
%%% o-2769
\startChapter[
  title={Chapter 02}
  ]

\Summary{As not only Prophets, but alſo Falſe-prophets were in the old
  Teſtamẽt, ſo now likewiſe there shal be Maiſters of Hereſie, to the
  damnation of themſelues, and of their followers. 4.~And of their
  damnation he pronounceth by examples (as he comforteth the vertuous
  Catholikes or true beleeuers with the example of Lot) becauſe of their
  railing at their Superiours and Prelates, their blaſpheming of
  Catholike doctrine, their voluptuous liuing, their lecherie, their
  couetouſnes, their manner of ſeducing, and the perſons ſeduced,
  20.~for whom it had been leſſe damnable, if they had neuer been
  Chriſtians.}

But there were alſo Falſe-prophets in the people, as alſo in you there
\CNote{\XRef{1.~Tim.~4.}
\XRef{Iude.}}
ſhal be lying maiſters which ſhal bring in Sects of perdition, and denie
him that hath bought them,
\TNote{\G{δεσπότην}}
the Lord: bringing vpon themſelues ſpeedie perditiõ. \V And
\SNote{Heretikes (of whom he prophecieth here) doe gaine ſcholers, by
preaching libertie, and by their owne licentious life, which is
ſpecially ioyned to the hereſie of theſe daies.}
many ſhal follow their riotouſneſſes, by whom the way of truth ſhal be
blaſphemed. \V And in auarice ſhal they
\SNote{Al the ſweet words of heretikes, ſpeaking much of the word of the
Lord, the Ghoſpel, \Sc{Iesvs Christ} &c. are but termes of art to buie
and ſel poore mens ſoules.}
with feined words make merchãdiſe of you. Vnto whom the iudgemẽt now
long ſince ceaſeth not: and their perdition ſlumbereth not. \V For if
God ſpared not Angels ſinning: but with the ropes of Hel being drawen
downe into Hel deliuered them to be tormented, that they ſhould be
reſerued vnto iudgement: \V and he ſpared not the original world, but
\CNote{\XRef{Gen.~7.}}
kept the eight, Noe, the Preacher of iuſtice, bringing in the deluge
vpon the world of the impious. \V And
\CNote{\XRef{Gen.~19.}}
bringing the cities of the Sodomites & of the Gomorrheites into aſhes,
he damned them with ſubuerſion, putting an example of them that ſhal doe
impiouſly: \V and
\CNote{\XRef{Gen.~19.}}
deliuered iuſt Lot oppreſſed by the iniurie and luxurious conuerſation
of the
%%% o-2770
abominable men. \V For in ſight and
\CNote{\XRef{Gen.~16.}}
hearing he was iuſt: dwelling with them who from day to day vexed the
iuſt ſoule with vniuſt workes.

\V Our Lord knoweth to deliuer the godly from tentation, but to reſerue the
vniuſt vnto the day of iudgement to be tormented: \V
\SNote{The ſpecial properties of Heretikes.}
and eſpecially them which walke after the fleſh in concupiſcence of
vncleãneſſe, and contemne dominion, bold, ſelf-pleaſers: they
%%% !!! LNote ? (this) or CNote ? (other) marked, but missing in both
feare not to bring in Sects, blaſpheming. \V Whereas Angels being
greater in ſtrength and power, beare not the execrable iudgement
%%% SNote only in other
\SNote{From our Lord.}
againſt them. \V But theſe men as vnreaſonable beaſts, naturally tending
to the ſnare and into deſtruction,
\SNote{So heretikes blaſpheme the higheſt myſteries of our faith through
ignorance.}
in thoſe things which they know not, blaſpheming, ſhal periſh in their
corruption, \V receiuing the reward of iniuſtice, eſteeming for a
pleaſure the delights of a day:
\TNote{\L{coinquinationes & maculæ}}
coinquinations and ſpots, flowing in delicacies, in their feaſtings
rioting with you, \V hauing eyes ful of adulterie and inceſſant ſinne:
alluring vnſtable ſoules, hauing their hart exerciſed with auarice, the
children of malediction:
%%% 2909
\V leauing the right way they haue erred, hauing followed the way of
Balaam of Boſor, which loued the reward of iniquitie, \V but
\CNote{\XRef{Nu.~22.~23.}}
had a check of his madneſſe, the dumme beaſt vnder the yoke, ſpeaking
with man's voice, prohibited the fooliſhnes of the Prophet.

\V Theſe are fountaines without water, and clouds, toſſed with
whirle-winds, to whom the miſt of darkeneſſe is reſerued. \V For
ſpeaking the proud things of vanitie, they allure in the deſires of
fleſhly riotouſnes, thoſe that eſcape a litle, which conuerſe in
errour, \V
\SNote{Who euer promiſed more libertie to their followers then Luther,
Caluin, and the like, taking away penance, faſting, continencie or
chaſtitie, keeping of vowes, neceſſitie of good workes (becauſe faith
doeth al) obedience to Eccleſiaſtical Paſtours and Councels, and ſuch
like?}
promiſing them libertie, whereas themſelues are the ſlaues of
corruption. For
\CNote{\XRef{Io.~8,~34.}
\XRef{Ro.~6,~16.}}
wherewith a man is ouercome, of that he is the ſlaue alſo. \V For if
flying from the coinquinations of the world in the knowledge of our Lord
and Sauiour \Sc{Iesvs} Chriſt, they againe intangled with the ſame be
ouercome:
\CNote{\XRef{Mt.~12,~45.}}
the later things are become vnto them worſe then the former. \V For it
was better for them not to know the way of iuſtice, then after the
knowledge, to turne backe from that holy commandement which was
deliuered to them. \V For, that of the true prouerb is chanced to them,
\CNote{\XRef{Pro.~26,~11.}}
The dogge returned to his vomit: and, The ſow washed
\TNote{\G{εἰς κυλισμὸν}}
into her wallowing in the mire.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

