%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 0759
%%% o-0681
\startChapter[
  title={Chapter 15}
  ]

\Summary{Abias reigneth wickedly in Iuda three yeares. 8.~After him his
  ſonne Aſa ſucceding deſtroyeth idolatrie, reigning fourtie one
  yeares. 16.~VVho hauing
  %%% 0760
  warres with the king of Iſrael, maketh league with the king of
  Syria. 24.~Aſa dying Ioſaphat ſuccedeth. 25.~Nadab reigneth wickedly
  two yeares in Iſrael, is then ſlaine by Baaſa of the tribe of
  Iſſachar, 29.~and his whole familie is deſtroyed. 33.~Baaſa alſo
  reigneth wickedly twentie foure yeares.}

Therfore in the eightenth yeare of king Ieroboam the ſonne of Nabat,
Abias reigned ouer Iuda. \V Three yeares he reigned in Ieruſalem: the
name of his mother was Maacha the daughter of Abeſſalom. \V And he
walked in al the ſinnes of his father, which he had done before him:
neither was his hart perfect with our Lord his God, as the hart of Dauid
his father. \V But for Dauids ſake our Lord his God gaue him a lampe in
Ieruſalem, that he might rayſe vp his ſonne after him, and eſtablish
Ieruſalem: \V
\SNote{Dauids poſteritie conſerued for his ſake.}
becauſe Dauid had done right in the eies of our Lord, and had not
declined from al thinges, which he commanded him, al the daies of his
life, except the matter of Vrias the Hetheite. \V But there was
%%% o-0682
warre betwen Roboam and Ieroboam al the time of his life. \V And the
reſt of the wordes of Abias, and al that he did, are they not written in
the
%%% !!! XRef?
Booke of the wordes of the kinges of Iuda? And there was warre betwen
Abias and Ieroboam. \V And Abias ſlept with his fathers, and they buried
him in the citie of Dauid: and Aſa his ſonne reigned for him. \V In the
twenteth yeare therfore of Ieroboam the king of Iſrael reigned Aſa the
king of Iuda. \V And he reigned one & fourtie yeares in Ieruſalem. His
mothers name was Maaca, the daughter of Abeſſalom. \V And Aſa did right
before the ſight of our Lord, as Dauid his father: \V and he tooke away
the effeminate out of the land, and he purged al the filth of the idols,
which his father had made. \V Moreouer he remoued alſo Maaca his
mother, that ſhe ſhould not be princeſſe in the ſacrifices of Priapus,
and in the groue which she had conſecrated: and he deſtroyed her denne,
and brake the moſt filthie idol, and burnt it in the torrent cedron: \V
but
\SNote{Thoſe altares which Salomon had made for his wiues, that were
idolaters Aſa deſtroyed not, but al which Roboã and Abias had made, or
ſuffered to be made for their owne people he pulled downe. Ioſias
afterward deſtroyed alſo thoſe which Salomon had made.
\XRef{2.~Paral.~34.}}
the excelſes he did not take away. Otherwiſe the hart of Aſa was perfect
with our Lord al his daies: \V and he caried in thoſe thinges, which his
father had ſanctified, and vowed into the houſe of our Lord, ſiluer and
gold, and veſſels. \V And there was warre betwen Aſa, and Baaſa the king
of Iſrael al their daies. \V Baaſa alſo the king of Iſrael went vp into
Iuda, and built Rama, that no man might go out or come in of Aſaes
%%% 0761
ſide the king of Iuda. \V Aſa therfore taking al the ſiluer, and gold
that remained in the treaſures of the houſe of our Lord, and in the
treaſures of the kinges houſe, gaue it into the handes of his ſeruantes:
and he ſent to Benadad the ſonne of Tabremon the ſonne of Hezion, the
king of Syria, which dwelt in Damaſcus, ſaying: \V There is a league
betwen me and thee, & betwixt my father and thy father: therfore I haue
ſent thee giftes, ſiluer and gold: and I deſire thee that thou come, and
make void the league, that thou haſt with Baaſa the king of Iſrael, and
he may retire from me. \V Benadad agreing to king Aſa, ſent the princes
of his armie into the
\Fix{citiees}{cities}{obvious typo, fixed in other}
of Iſrael, and they ſtroke Ahion, and Dan, and Abeldomum of Maacha, and
al Cenneroth, to witte, al the Land of Nephthali. \V Which when Baaſa
had heard, he intermitted to build Rama, and returned into Therſa. \V
But king Aſa ſent word into al Iuda, ſaying: Let no man be excuſed, and
they tooke ſtones from Rama, and the timber therof, wherwith Baaſa had
built, and Aſa of it built GabaaBeniamin, and Maſpha. \V But the reſt of
al the wordes of Aſa, and al his forces, and al that he did, & the
cities that he built, are not theſe written in the
%%% !!! XRef?
Booke of the words of the dayes of the kinges of Iuda? Howbeit in the
time of his old age he was diſeaſed in his feete. \V And he ſlept with
his fathers, & was buried with them in the citie of Dauid his
father. And Ioſaphat his ſonne reigned for him. \V But Nadab the ſonne
of Ieroboam reigned ouer Iſrael the ſecond yeare of Aſa the king of
Iuda: and he reigned ouer Iſrael two yeares. \V And he did that which is
euil in the ſight of our Lord, and walked in the waies of his
%%% o-0683
father, and in his ſinnes, wherwith he made Iſrael to ſinne. \V And Baaſa
the ſonne of Ahias of the houſe of Iſſachar, lay in wayte againſt him,
and ſtroke him in Gebbethon, which is a citie of the Philiſtimes: for
Nadab and al Iſrael beſieged Gebbethon. \V Baaſa therfore ſlew him in
the third yeare of Aſa the king of Iuda, and reigned for him. \V And
when he reigned, he ſtoke
\SNote{The authour of ſchiſme puniſhed in his poſteritie.}
al the houſe of Ieroboam: he leaft not ſo much as one ſoule of his
ſeede, til he deſtroied him according to the word of our Lord, which he
had ſpoken in the hand of Ahias the Silonite; \V for the ſinnes of
Ieroboam, which he had ſinned, and wherwith he had cauſed Iſrael to
ſinne, and for the offence, wherwith he prouoked our Lord the God of
Iſrael. \V But the reſt of the wordes of Nadab, and al
%%% 0762
that he wrought, are not theſe thinges written in the
%%% !!! XRef?
Booke of the wordes of the daies of the kinges of Iſrael? \V And there
was warre betwen Aſa and Baaſa the king of Iſrael al their daies. \V In
the third yeare of Aſa the king of Iuda, reigned Baaſa the ſonne of
Ahias, ouer al Iſrael in Therſa foure and twentie yeares. \V And he did
euil before our Lord, & walked in the waies of Ieroboam, and in his
ſinnes, wherwith he made Iſrael to ſinne.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
