%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 0713
%%% o-0641
\startChapter[
  title={Chapter 3}
  ]

\Summary{King Salomon marieth Pharaos ſiſter. 3.~Offereth victims in
  high places. 5.~Admonished by God in his ſleepe to demand what he wil,
  he asketh wiſdom
  %%% 0714
  to gouerne his people. 10.~Which God granteth him, with much richeſse
  alſo and glorie. 16.~He decideth a controuerſie betwen two wemen
  contending about a liuing child and a dead.}

The kingdom therfore was eſtablished in the hand of Salomon, and he was
ioyned in affinitie to Pharao the king of Ægypt: for he tooke his
daughter, and brought her into the citie of Dauid, vntil he accomplished
building his owne houſe, and the houſe of our Lord, and the wal of
Ieruſalem round about. \V But yet the people immolated in the excelſes:
for there was no temple built to the name of our Lord vntil that day. \V
And Salomon loued our Lord, walking in the preceptes of Dauid his
father, ſauing that he immolated in the excelſes, and burnt incenſe. \V
He went therfore into Gabaon, to immolate there: for that was a verie
great excelſe: a thouſand hoſtes for holocauſt did Salomon offer vpon
that altar in Gabaon. \V And our Lord appeared to Salomon in a dreame by
night, ſaying: Aske what thou wilt that I may geue it thee. \V And
Salomon ſaid: Thou haſt done great mercie with thy ſeruant Dauid my
father, euen as he walked in thy ſight in truth, and iuſtice, and a
right hart with thee: for thou haſt kept thy great mercie, and haſt
geuen him a ſonne ſitting vpon his throne, as it is this day. \V And now
Lord God, thou haſt made thy ſeruant to reigne for Dauid my father: but
I am a litle childe, and ignorant of my going out and coming in. \V And
thy ſeruant is in the middes of the people, which thou haſt choſen, a
people infinite, which can not be numbred and counted for the
multitude. \V Thou shalt therfore geue to thy ſeruant a docible hart,
that he may iudge the people, & diſcerne betwen good and euil. For who
ſhal be able to iudge this people, this thy people great in number? \V
The word therfore was liked before our Lord, that Salomon had asked ſuch
a thing. \V And our Lord ſayd to Salomon: Becauſe thou haſt asked this
thing, and haſt not deſired for thy ſelf manie dayes, nor riches, nor
the liues of thine enemies, but haſt deſired wiſedom for thy ſelf to
diſcerne iudgement: \V behold I haue done vnto thee according to thy
wordes, & haue geuen thee a wiſe hart and intelligent, inſo much that
none before thee hath beene like thee, nor shal ariſe after thee. \V Yea
and theſe thinges, which thou did not aske, I haue geuen thee: to witte,
riches, and glorie, ſo that none
%%% 0715
hath beene like thee among the kinges al dayes hertofore. \V And if thou
wilt walke in
%%% o-0642
my wayes, and keepe my preceptes, and my commandementes, as thy father
walked, I wil make thy dayes long. \V Therfore Salomon awaked, and
perceiued that it was a dreame: and when he was come to Ieruſalem, he
ſtood before the arke of couenant of our Lord, and offered holocauſtes,
and made pacifique victimes, and a great feaſt to al his ſeruantes. \V
Then came there two wemen harlottes to the king, and ſtood before
him: \V of the which one ſaid: I beſech thee, my lord, I and this woman
dwelt in one houſe, and I was deliuered of a childe beſide her in the
chamber. \V And the third day, after that I was deliuered, she alſo
was deliuered, and we were together, and no other perſon with vs in the
houſe, except we two. \V And this womans childe died in the night. For
ſleping she oppreſſed him. \V And ryſing in the dead tyme of the night,
she tooke my childe from the ſide of me thy handmayd being aſlepe, and
layed it in her boſome: and her childe that was dead, she put in my
boſome. \V And when I was ryſen in the morning to geue my childe milke,
he appeared dead: whom more diligently beholding when it was cleere day,
I found that it was not mine which I bare. \V And the other woman
anſwered: It is not ſo as thou ſayſt, but thy childe is dead, and mine
liueth. On the contrarie part she ſayd: Thou lieſt: for my childe
liueth, and thy childe is dead. And in this maner they ſtroue before the
king. \V Then ſayd the king: This woman ſaith: My childe liueth, and thy
childe is dead. And this hath anſwered: No, but thy childe is dead, and
mine liueth. \V The king therfore ſaid: Bring me a ſword. And when they
had brought a ſword before the King: \V Diuide, quoth he, the liuing
child into two partes, and geue the halfe part to one, and halfe to the
other. \V But the woman, whoſe childe was aliue, ſaid to the king (for
her bowels were moued vpon her childe) I beſech thee my lord, geue her
the childe aliue, and kil it not. On the contrarie part she ſayd:
\SNote{So heretikes not being able to proue that their ſynagogue is the
true & permanent Church, would deſtroy the Catholique and ſo haue none
at al.}
Be it neither mine, nor thine, but let it be diuided. \V The king
anſwered, and ſaid: Geue vnto this woman the infant aliue, and let it
not be killed for this is the mother therof. \V Al Iſrael therfore heard
the iudgement that the king had iudged, and they feared the king, ſeing
the wiſedom of God to be in him to doe iudgement.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
