%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 0747
%%% o-0671
\startChapter[
  title={Chapter 10}
  ]

\Summary{The quene of Saba coming to king Salomon, admireth his vviſdom,
  magnificence, and order of gouernment. 10.~She geueth and receiueth
  giftes. 14.~Salomon receiueth much gold diuers vvayes: 16.~maketh
  golden targets, 18.~a magnificent throne, 21.~and much golden
  veſsel. 25.~Manie bring him preſentes. 26.~He hath manie chariottes,
  horſemen, 27.~abundance of ſiluer. 28.~Marchantes of diuers kingdomes
  ſel him horſes.}

But the Queene of
\SNote{Part of Arabia is called Saba, nere to Iurie, but this Saba is
beyond Arabia, as S.~Hierom teſtifieth in
\Cite{Eſais.~60. li.~17.}
it ſemeth to be in Æthiopia, for our Sauiour ſaith
\XRef{Mat.~12.}
\Emph{The quene of the South came frõ the endes of the earth, to heare
the vviſdom of Salomon.}}
Saba alſo hauing heard the fame of Salomon, in the name of our Lord came
to proue him, in hard propoſitions. \V And entring into Ieruſalem with a
great trayne, and riches, and camels carying ſpices, and gold exceding
infinite, and pretious ſtones, she came to king Salomon, and ſpake to
him al thinges that she had in her hart. \V And Salomon interpreted  to
her al the wordes, that she propoſed: there was not a word, that the
king could be ignorant of, and could not anſwer her. \V And the queene
of Saba ſeing al the wiſedom of Salomon, and the houſe, which he had
built, \V and the meates of his table, and the habitations of his
ſeruantes, and the orders of them that ſerued, and their garmentes,
cupbearers, and the holocauſtes
%%% 0748
which he offered in the houſe of our Lord:
\SNote{As this quene had no ſpirite, when ſhe ſaw Salomons wiſdom, ſo
the Church gathered of gentiles knowing Chriſts grace, & finding the
maſters of Euangelical doctrin, caſting away the ſpirite of pride, and
laying of al hautineſſe of mind, lerned to diſtruſt in her ſelf, and to
truſt in the great mercie of her king.
\Cite{S.~Greg. in Pſal.~7.}}
ſhe had no longer ſpirit, \V and she ſaid to the king: The report is
true, which I haue heard in my countrie, \V concerning thy wordes, and
concerning thy wiſedom, and I did not beleue them that told me, til my
ſelfe came, and ſawe with myne eies, and haue proued that the half not
been told me: greater is thy wiſdom, and thy workes, then the rumour,
which I haue heard. \V Bleſſed are thy men, and bleſſed are thy
ſeruantes, which ſtand before thee alwaies, and heare thy wiſedom. \V Be
the Lord thy God bleſſed, whom thou haſt pleaſed, and that hath ſette
thee vpon the throne of Iſrael, for that the Lord hath loued Iſrael for
euer, and hath appointed thee king, to do iudgement and iuſtice. \V She
therfore gaue to the king an hundred and twentie talentes of gold, and
ſpices exceding much, and pretious ſtones: There was no more brought ſo
much ſpice, as that which the Quene of Saba gaue to king Salomon.
(\V But the naue alſo of Hiram, which caried gold out of Ophir, brought
from Ophir thyine trees exceding manie, and pretious ſtones. \V And the
king made of the thyine trees the porches of the houſe of our Lord, and
of the kinges houſe and harpes & vials for the fingers: there were not
ſuch thyine trees brought, nor ſeen vntil this preſent daye.) \V And king
%%% o-0672
Salomon gaue to the queene of Saba al that she would, and asked of him:
beſide thoſe thinges, which of him ſelfe he offered her for a royal
gift. Who returned, and went into her countrie with her ſeruantes. \V
And the weight of the gold, that was brought to Salomon euerie yeare,
was of ſix hundred ſixtie ſix talentes of gold: \V beſide that, which
the men brought, that were ouer the tributes, and merchantes, and al
that ſold light wares, and al the kinges of Arabia, and the dukes of the
land. \V Salomon alſo made two hundred ſhieldes of moſt pure gold, ſix
hundred ſicles of gold did he allow for the plates of one shield. \V And
three hundred tergattes of tried gold: and three hundred poundes of gold
garnished one terget: and the king put them in the houſe of the foreſt
of Libanus. \V King Salomon alſo made a great throne of yuorie: and
couered it with gold exceding yellow, \V which had ſixe ſteppes: and the
toppe of the throne was round in the hinder part: and the two handes on
either ſide holding the ſeate: and two lyons ſtood at euery hand. \V
And twelue litle lyons ſtanding vpon the ſixe ſteppes on either ſide:
there
%%% 0749
was not ſuch a worke made in al kingdomes. \V Yea and al the veſſels,
out of the which the king Salomon drunke, were of gold: and al the
furniture of the houſe of the foreſt of Libanus of moſt pure gold: there
was no ſiluer, neither was it thought of any price in the daies of
Salomon, \V becauſe the kinges nauie, once in three yeares, went with
the nauie of Hiram on the ſea into Tharſis, bringing thence gold, and
ſiluer, and the teeth of elephantes, and apes, and pecockes. \V King
Salomon therfore was magnified aboue al the kinges of the earth in
riches, and wiſedom. \V And al the earth deſired to ſee Salomons face,
that they might heare his wiſedom, which God had geuen in his hart. \V
And euerie one preſented him giftes, veſſel of ſiluer and gold,
garmentes and inſtrumentes for warre, ſpices alſo, and horſes and mules
euerie yeare. \V And Salomon gathered together the chariotes and
horſemen, and there amounted to him a thouſand foure hundred chariotes,
and twelue thouſand horſemen: and he diſpoſed them in fenſed cities, and
with the king in Ieruſalem. \V And he made that there was as great
abundance of ſiluer in Ieruſalem, as of ſtones: and of ceder trees he
cauſed ſuch a multitude, as if it were ſycomore trees, which grow in the
playnes. \V And there were horſes brought for Salomon out of Ægypt, and
Coa, for the kinges merchantes brought them out of Coa, and brought them
at a ſette price. \V And a chariote of foure horſes came out of Ægypt,
for ſixe hundred ſicles of ſiluer, and one horſe for an hundred and
fiftie. And after this maner did al the kinges of the Hetheites, and of
Syria ſel horſes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
