%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 0749
%%% o-0673
\startChapter[
  title={Chapter 11}
  ]

\Summary{Salomon louing and marying manie wemen of diuers nations, is
  drawen by them to idolatrie. 9.~God therfore offended ſuffereth Adad
  an Idumean, 23.~Razon king of Damaſcus, 26.~and Hieroboam his owne
  ſeruant to make warre againſt him. 29.~Ahias the prophet fortelleth
  Hieroboam, that he shal reigne ouer tenne tribes, leauing but tvvo to
  Salomons heyres, 38.~vvith promiſe to proſper, if he ſerue
  God. 42.~Salomon dieth.}

And king Salomon loued manie wemen ſtrangers, the daughter alſo of
Pharao, and Moabites, and Ammonites, Idumeians, and Sidonians, and
Hetheians: \V of the nations, wherof our Lord ſayd to the children of
Iſrael: You ſhal not goe in vnto them, neither shal anie of them come in
%%% 0750
vnto yours: for they wil moſt certainly turne away your hartes to folow
their goddes. To theſe therfore was Salomon copled in moſt feruent
loue. \V And he had wiues as it were queenes
\SNote{Though pluralitie of wiues was then alowed, yet it was forbid to
multiplie manie.
\XRef{Deut.~17.}}
ſeuen hundred, and concubines three hundred: and the wemen turned away
his hart. \V And when he was now old, his hart was depraued by wemen,
that he folowed ſtrange goddes: neither was his hart perfect with our
Lord his God, as the hart of Dauid his father. \V But Salomon worshipped
Aſtarthee the goddeſſe of the Sidonians, and Moloch the idol of the
Ammonites. \V And Salomon did that which was not liked before our Lord,
and he accomplished not to folow our Lord, as Dauid his father. \V Then
built Salomon a temple to Camos the idol of Moab, in the mount that is
agaynſt Ieruſalem, and to Moloch the idol of the children of Ammon. \V
And in this maner did he to al his wiues that were ſtrangers, which
burnt frankencenſe, and immolated to their goddes. \V Therfore our Lord
was wrath with Salomon, becauſe his minde was turned away from our Lord
the God of Iſrael, who had appeared vnto him the ſecond tyme, \V and had
commanded him concerning this word, that he should not folow ſtrange
goddes, & he kept not the thinges which our Lord commanded him. \V Our
Lord therfore ſayd to Salomon: Becauſe thou haſt done this, and haſt not
kept my couenant, and my preceptes, which I haue commanded thee,
breaking I wil rent aſunder thy kingdom, and wil geue it to thy
ſeruant. \V Neuertheleſſe in thy dayes I wil not doe it, becauſe of
Dauid thy father: out of the hand of thy ſonne I wil rent it, \V neither
wil I take away the whole kingdom, but
\SNote{The tribe of Iuda.}
one tribe I wil geue to thy ſonne for Dauid my ſeruant, and
\SNote{By Ieruſalem is vnderſtood the tribe of Beniamin, wherin it
ſtood. So there remained two tribes to Salomons heyres.}
Ieruſalem, which I haue choſen. \V And our Lord rayſed vp an aduerſarie
to Salomon, Adad an Idumeite of the kinges ſeede, who was in Edom. \V
For when Dauid was in Idumea, and Ioab the general of the warfare was
gone vp to burie them that
%%% o-0674
were ſlayne, and had ſlayne al malekind in Idumea,
\CNote{\XRef{2.~Reg.~8.}}
(\V for Ioab taried there ſix monethes and al Iſrael, til he ſlew al
malekind in Idumea,) \V Adad him ſelf fled, and men of Idumea of his
fathers ſeruantes with him, to goe into Ægypt: and Adad was a litle
boy. \V And when they roſe out of Madian, they came into Pharan, and
they tooke with them men of Pharan, and entered into Ægypt to Pharao the
king of Ægypt: who gaue him a houſe, and appoynted him meates, and
%%% 0751
aſſigned him land. \V And Adad found grace before Pharao excedingly, in
ſo much that he gaue him to wife, the germane ſiſter of his wife Taphnes
the queene. \V And the ſiſter of Taphnes bare him a ſonne Genubath, and
Taphnes brought him vp in the houſe of Pharao: and Genubath was dwelling
at Pharaoes houſe with his children. \V And when Adad in Ægypt had
heard, that Dauid ſlept with his fathers, and that Ioab the general of
the warefare was dead, he ſayd to Pharao: Diſmiſſe me, that I may goe
into my countrie. \V And Pharao ſayd to him: For what lackeſt thou with
me, that thou ſeekeſt to goe into thyne owne countrie? But he anſwered:
Nothing: yet I beſech thee that thou diſmiſſe me. \V God alſo rayſed vp
to him an aduerſarie, Razon, the ſonne of Eliada,
\CNote{\XRef{2.~Reg.~10.}}
who had fled Adarezer the king of Soba his lord: \V and he gathered men
agaynſt him, and he became the captayne of theues, when Dauid killed
them: and they went to Damaſcus, and dwelt there, and they made him king
in Damaſcus, \V and he was an aduerſarie to Iſrael
\SNote{From the time that Salomon fel to idolatrie, he was more impugned
by three perpetual aduerſaries, Adad, Razon, and Hieroboam; myſtically
ſignifying the fleſh, the world, and the diuel.}
al the dayes of Salomon: and this is the euil of Adad, and hatred
agaynſt Iſrael, and he reigned in Syria. \V Ieroboam alſo the ſonne of
Nabath, an Ephratheite of Sareda, the ſeruant of Salomon, whoſe mother
was called Serua, a woman widow lifted vp his hand agaynſt the king. \V
And this is the cauſe of his rebellion agaynſt him, becauſe Salomon
built Mello, and filled vp the breache of the citie of Dauid his
father. \V And Ieroboam was a ſtrong man and mightie: and Salomon ſeing
the youngman of a good witte & induſtrious, had made him chief ouer the
tributes of al the houſe of Ioſeph. \V It came to paſſe therfore at that
tyme, that Ieroboam went out of Ieruſalem, and the prophete Ahias the
Silonite found him in the way, couered with a new cloke: and they two
onlie were in the field. \V And Ahias taking his new cloke, wherwith he
was couered,
\SNote{This fact cõfirmed his wordes, that he ſpoke ſeriouſly & fained
not.}
cut it into twelue partes. \V And he ſayd to Ieroboam: Take vnto thee
ten pieces: for thus ſayth our Lord the God of Iſrael: Behold I wil rent
the kingdom out of the hand of Salomon, and wil geue thee ten tribes. \V
But one tribe ſhal remayne to him for my ſeruant Dauid, and Ieruſalem
the citie, which I haue choſen of al the tribes of Iſrael: \V becauſe he
hath forſaken me, and hath adored Aſtarthee the goddeſſe of the
Sidonians, & Chamos the god of Moab, and Moloch the god of the children
of Ammon: and hath
%%% 0752
not walked in my waies, to doe iuſtice before me, and my preceptes, and
iudgementes as Dauid his father. \V Neither wil I take away al the
kingdom out of his hand, but I wil make him
%%% o-0675
prince al the daies of his life, for Dauid my ſeruant, whom I choſe, who
kept my commandmentes and my preceptes. \V But I wil take away the kingdom
out of his ſonnes hand, and wil geue thee ten tribes: \V and to his
ſonne I wil geue one tribe, that there may remayne a lampe to Dauid my
ſeruant at al times before me in Ieruſalem, the citie which I haue
choſen, that my name might be there. \V And thee wil I take, and thou
shalt reigne ouer al thinges, that thy ſoule deſireth, and thou shalt be
king ouer Iſrael. \V If therfore thou wilt heare al thinges, that I
shal command thee, and wilt walke in my waies, and doe that which is
right before me, keeping my commandmentes and my preceptes, as Dauid my
ſeruant did: I wil be with thee, and wil build thee a faythful houſe, as
I built a houſe to Dauid, and I wil deliuer Iſrael to thee: \V and I
wil afflict the ſeede of Dauid vpon this, but yet not alwaies. \V
Salomon therfore would haue killed Ieroboam: who aroſe, and fled into
Ægypt to Sefac the king of Ægypt, and was in Ægypt vntil the death of
Salomon. \V And the reſt of the wordes of Salomon, and al that he did,
and his wiſedom: behold they are al written in the
%%% !!! XRef? Does this still exist?
Booke of the wordes of the daies of Salomon. \V And the daies, that
Salomon reigned in Ieruſalem ouer al Iſrael, are fourtie yeares. \V And
Salomon
\SNote{VVhether he repented and was ſaued or no, is vncertaine.}
ſlept with his fathers, and was buried in the citie of Dauid his father,
and Roboam his ſonne reigned for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
