%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 0757
%%% o-0679
\startChapter[
  title={Chapter 14}
  ]

\Summary{Ahias the prophet forsheweth the ruine of Hieroboams familie:
  12.~namely the death of his ſonne, for whom being ſick, the mother
  conſulteth the prophet. 20.~Hieroboam dieth, and his ſonne Nadab
  reigneth. 21.~Some alſo of the people of Iuda, committing idolatrie and
  other ſinnes, 25.~the King of Ægypt inuadeth and ſacketh
  Hieruſalem. 31.~Roboam dieth and his ſonne Abias reigneth.}

At that time Abia the ſonne of Ieroboam was ſicke. \V And Ieroboam ſayd
to his wife: Ariſe, and change thy habite, that thou be not knowen to be
the wife of Ieroboam, and goe into Silo, where Ahias the prophete is,
which ſpake to me, that I should reigne ouer this people. \V Take alſo
in thy hand ten loaues, and crackneles, and a veſſel of honie, and goe
to him: for he wil shew thee what shal happen to this childe. \V The
wife of Ieroboam did as he had
%%% 0758
ſpoken: and riſing vp went into Silo, and came into the houſe of Ahias:
but he could not ſee, becauſe his eies were dimme
%%% o-0680
for age. \V And our Lord ſayd to Ahias: Behold the wife of Ieroboam
cometh in, to conſult thee concerning her ſonne that is ſicke: thus and
thus shalt thou ſpeake to her. When ſhe therfore entered in, and
diſſembled to be that she was, \V Ahias heard the ſound of her feete
entring in at the doore, and ſayd: Come in Ieroboams wife: Why doeſt
thou fayne thy ſelf to be an other woman? But I am ſent to thee a heauie
meſſenger. \V Goe, and tel Ieroboam: Thus ſayth our Lord the God of
Iſrael: Becauſe I haue exalted thee out of the middes of the people, and
made thee prince ouer my people Iſrael: \V and haue rent the kingdom of
the houſe of Dauid, and geuen it to thee, and thou haſt not beene as my
ſeruant Dauid, who kept my commandementes, and folowed me in al his
hart, doing that which was wel liked in my ſight: \V but haſt wrought
euil aboue al, that haue beene before thee, and haſt made thee ſtrange
and molten goddes,
\SNote{Ieroboam did not wittingly and of purpoſe ſet vp falſe goddes, to
the end he might prouoke God to anger: for his intention only was to
kepe the people frõ going to Ieruſalem, leſt by that occaſion \Emph{they
should returne to Roboam their Lord, king of Iuda}.
\XRef{ch.~12. v.~27.}
But by ſettĩg vp idols he did prouoke God conſequently to anger. So
here and in other places this phraſe: \Emph{that he might prouoke: that
it might be fulfilled}, and the like, ſignifieth not the final cauſe,
but the ſequele of other factes, without direct intention.}
that thou mighteſt prouoke me to anger, and haſt reiected me behind thy
backe: \V therfore behold I wil bring in euils vpon the houſe of
Ieroboam, and wil ſtrike of Ieroboam him that pyſſeth to the wal, and
the incloſed, and the vileſt in Iſrael: and I wil cleanſe the remaynes
of the houſe of Ieroboam, as dung is wont to be cleanſed til al be
pure. \V They that shal die of Ieroboam in the citie, them the dogges
ſhal eate: and they that shal die in the field, them the foules of the
ayre shal deuoure: becauſe our Lord hath ſpoken. \V Thou therfore ariſe,
and goe into thy houſe: and in the verie entrance of thy feete into the
citie, the childe shal die, \V and al Iſrael shal mourne for him, and
shal burie him: for this onlie of Ieroboam shal be brought into the
ſepulchre, becauſe vpon him hath beene found a good word from our Lord
the God of Iſrael, in the houſe of Ieroboam. \V And our Lord wil appoynt
to him ſelf a king ouer Iſrael, that shal ſtrike the houſe of Ieroboam
in this day, and in this time: \V and our Lord the God of Iſrael shal
ſtrike it, as a reede is wont to be moued in the water: and he shal
plucke out Iſrael from this good countrie, which he gaue to their
fathers, and shal ſcatter them ouer the Riuer: becauſe they haue made to
them ſelues groues, to prouoke our Lord. \V And our Lord shal deliuer
Iſrael for the ſinnes of Ieroboam, who hath ſinned, & made Iſrael to
ſinne. \V The wife therfore
%%% 0759
of Ieroboam aroſe, and departed, and came into Therſa: and when she
entered the threshold of the houſe, the childe died, \V and they buried
him. And al Iſrael mourned for him according to the word of our Lord,
which he ſpake in the hand of his ſeruant Ahias the prophete. \V But the
reſt of the wordes of Ieroboam, how he fought, and how he reigned,
behold they are written in the
%%% !!! XRef?
Booke of the wordes of the daies of the kinges of Iſrael. \V And the
daies, that Ieroboam reigned, are two and twentie yeares: and he ſlept
with his fathers: and Nadab his ſonne reigned for him. \V Moreouer
Roboam the ſonne of Salomon reigned in Iuda: one and fourtie yeares old
was Roboam when he began to reigne: ſeuentene yeares reigned he in
Ieruſalem the citie, which our Lord choſe to put his name there,
%%% o-0681
of al the tribes of Iſrael. And his mothers name was Naama an
Ammonite. \V And
\Fix{Iudas}{the men of Iuda}{not really a typo, same in both}
did euil before our Lord, and prouoked him aboue al thinges, that their
fathers had done, in their ſinnes which they ſinned. \V For they alſo
built them altars, and ſtatues, and groues vpon euerie high hil, and
vnder euerie tree ful of grene leaues: \V yea and effeminates were in
the land, and they did al the abominations of the gentiles, which our
Lord deſtroyed before the face of the children of Iſrael. \V And in the
fifth yeare of the reigne of Roboam, Seſac the king of Ægypt came vp
into Ieruſalem, \V and tooke the treaſures of the houſe of our Lord,
and the kinges treaſures, and al thinges he ſpoyled: the shieldes alſo
of gold, which Salomon had made: \V for the which Roboam made braſen
shieldes, & deliuered them into the hand of the captaynes of shield
bearers, and of them that kept watch before the doore of the kinges
houſe. \V And when the king went into the houſe of our Lord, they that
had the office to goe before, caried them: and afterward they recaried
them to the armourie of the shieldbearers. \V And the reſt of the wordes
of Roboam, & al that he did, behold they are written in the
%%% !!! XRef?
Booke of the wordes of the daies of the kinges of Iuda. \V And there was
warre betwen Roboam and Ieroboam alwaies. \V And Roboam ſlept with his
fathers, and was buried with them in the citie of Dauid: and his mothers
name was Naama an Ammonite: and Abias his ſonne reigned for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
