%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 0710
%%% o-0638
\startChapter[
  title={Chapter 2}
  ]

\Summary{Dauid geueth godlie preceptes to Salomon. 5.~Willeth him to
  punish certaine offenders: 10.~and dieth. 12.~Salomon
  reigneth. 13.~Adonias, by interceſsion of Bethſabee, requeſteth to
  haue Abiſag to wife: 22.~but is put to death for demanding
  her. 26.~Abiathar the prieſt is banished, and depoſed, for conſpiring
  with Adonias. 28.~Ioab alſo for the ſame cauſe, and former crimes is
  ſlaine. 36.~Semei is confined in Ieruſalem, 40.~& for tranſgreſsing
  his limites together with old faultes is likewiſe ſlaine.}

And the dayes of Dauid approched that he should die, and he commanded
his ſonne Salomon, ſaying: \V I enter into the way of al flesh: take
courage, and play the man. \V And obſerue the watches of our Lord thy
God, that thou walke in his waies, and keepe his ceremonies, and his
preceptes, and iudgementes, and teſtimonies, as is written in the law of
Moyſes: that thou mayſt vnderſtand al thinges which thou doeſt, and
whitherſoeuer thou shalt turne thy ſelfe:
%%% 0711
\V that our Lord may confirme his wordes, which he hath ſpoken of me,
ſaying: If thy children ſhal keepe their waies, and ſhal walke before me
in truth, in al their hart, and in al their ſoul, there shal not
altogether be taken from thee a man out of the throne of Iſrael. \V Thou
knoweſt alſo what Ioab the ſonne of Saruia hath done to me, what he did
to the two princes of the armie of Iſrael, to Abner the ſonne of Ner,
and Amaſa the ſonne of Iether: whom he ſlewe, & shed the bloud of warre
in peace, & put the bloud of battel in his belt, which was about his
loynes, and in his shoe, which was on his feete. \V Thou shalt doe
therfore according to thy wiſdom, and shalt not bring his hoare head
peaceably vnto hel. \V But to the ſonnes alſo of Berzellai the Galaadite
thou shalt render kindnes, and they shal eate on thy table: for they
mette me when I fled from the face of Abſalom thy brother. \V Thou
haſt alſo with thee Semei the ſonne of Gera the ſonne of Iemini of
Bahurim, who curſed me with a wicked curſe, when I went to the Campe;
but becauſe he came downe to meete me when I paſſed Iordan, and I ſware
to him by our Lord, ſaying: I wil not kil thee with the ſword: \V doe
not thou ſuffer
%%% o-0639
him to be guiltles. But thou art a wiſe man, ſo that thou knoweſt what
thou ſhalt doe to him. And thou shalt bring his hoare heare with bloud
vnto hel. \V Dauid therefore ſlept with his fathers, and was buried in
the citie of Dauid. \V And the daies that Dauid reigned in Iſrael, are
fourtie yeares: in Hebron he reigned ſeuen yeares, in Ieruſalem thirtie
three.

\V
\MNote{The ſecond part.

Of Salomons reigne, and actes good and bad.}
And Salomon ſate vpon the throne of Dauid his father, and his kingdome
was confirmed excedingly. \V And Adonias the ſonne of Haggith entered in
to Bethſabee the mother of Salomon. Who ſaid to him: Is thy coming
peaceable? Who anſwered: Peaceable. \V And he added: I haue a word to
ſpeake with thee. To whom she ſaid: Speake. And he: \V Thou knoweſt,
quoth he, that the kingdom was mine, and al Iſrael had purpoſed to make
me ouer them to be their king: but the kingdom is tranſpoſed, & is made
my brothers: for it was appointed him of our Lord. \V Now therefore I
deſire one petition of thee; Confound not my face. Who ſaid to him:
Speake. \V And he ſaid: I pray ſpeake to Salomon the king (for he can
not denie thee any thing) that he geue me Abiſag the Sunamite to
wife. \V And Bethſabee ſaied: Wel, I wil ſpeake for thee to the king. \V
Bethſabee therefore came to king Salomon, to
%%% 0712
ſpeake vnto him for Adonias: and the king aroſe to meete her, and adored
her, and ſate downe vpon his throne: and a throne was ſette for the
kinges mother, who ſate on his right hand. \V And ſhe ſaid to him: One
litle petition I deſire of thee, confound not my face. And the king ſaid
to her: My mother aske: for
\SNote{In al conuenient ſuites it behoueth the ſonne to heare his
mother.}
it behoueth not that I turne away thy face. \V Who ſaid: Let Abiſag the
Sunamite be geuen to Adonias thy brother to wife. \V And king Salomon
anſwered, and ſaid to his mother: Why doeſt thou aske Abiſag the
Sunamite for Adonias? aske for him alſo the kingdom: for he is my
brother elder then I, & hath Abiathar the prieſt, & Ioab the ſonne of
Saruia. \V Therfore king Salomon ſware by our Lord, ſaying: Theſe
thinges doe God to me, and theſe adde he, becauſe Adonias hath ſpoken
this word againſt his life. \V And now, our Lord liueth which hath
eſtabliſhed me, and placed me vpon the throne of Dauid my father, and
which hath made me a houſe, as he ſpake, this day ſhal Adonias be
ſlaine. \V And king Salomon ſent by the hand of Banaias the ſonne of
Ioiada, who ſlewe him, and he died. \V To Abiathar alſo the prieſt the
king ſaid: Goe into Anathoth to thy field, and thou in deede art a man
of death: but to day I wil not kil thee, becauſe thou didſt carie the
arke of our Lord God before Dauid my father, & haſt ſuſteyned labour in
al thinges, wherein my father laboured. \V Salomon therefore
\SNote{By ſpecial inſtinct Salomon did this extraordinarie fact as a
prophet & miniſter of God, executing his ſentence geuen before againſt
the houſe of Heli for the ſinnes of his children,
\XRef{1.~Reg.~2. v.~31.}
and for Abiathars proper fault ioyning with Adonias againſt Salomon.
\XRef{3.~Reg.~1.}}
caſt out Abiathar, that he ſhould not be the prieſt of our Lord, that
the word of our Lord might be fulfilled, which he ſpake concerning the
houſe of Heli in Silo. \V And then a meſſenger came to Ioab for that
Ioab had turned after Adonias, and after Salomon had not turned: Ioab
therfore fled into the tabernacle of our Lord, & caught the horne of the
altar. \V And it was told king Salomon, that Ioab was fled into the
tabernacle of our Lord, and was beſide the altar: & Salomon ſent Banaias
the ſonne of Ioiada, ſaying:
%%% o-0640
Goe, kil him. \V And Banaias came to the tabernacle of our Lord, and
ſaid to him: Thus ſayeth the king: Come forth. Who ſaid: I wil not come
forth, but here wil I die. Banaias reported his word to the king,
ſaying: Thus ſpake Ioab, and thus he anſwered me. \V And the king ſaid
to him: Doe as he hath ſpoken: and kil him, & burie him, and thou ſhalt
remoue the innocent bloud, which hath beene ſhed of Ioab, from me, and
from the houſe of my father. \V And our Lord ſhal render his bloud vpon
his head, becauſe he murdered two iuſt men, &
%%% 0713
better then him ſelf: and ſlew them with his ſword, my father Dauid not
knowing, Abner the ſonne of Ner general of the warfare of Iſrael, and
Amaſa the ſonne of Iether general of the armie of Iuda: \V and their
bloud ſhal returne vpon the head of Ioab, and vpon the head of his ſeede
for euer. But to Dauid and his ſeede and his houſe, and to his throne be
peace for euer from our Lord. \V Banaias therfore the ſonne of Ioiada
went vp, and ſetting vpon him ſlewe him: and he was buried in his houſe
in the deſert. \V And the king appoynted Banaias the ſonne of Ioiada
for him ouer the armie, and Sadoc the prieſt he
\SNote{Salomon was not only a kĩg but alſo a prophet. Moreouer ſome
ſecular princes doe nominate ſpiritual ſuperiours, and inſtal them in their
ſeates, yet their iuriſdiction is not depending on the prince: but the
prince to be directed by them.
\XRef{Num.~27. v.~21.}}
placed for Abiathar. \V The king alſo ſent, and called Semei, and ſaid
to him: Build thee a houſe in Ieruſalem, and dwel there: and thou shalt
not goe out thence hither and thither. \V But what day ſoeuer thou shalt
goe out, and shalt paſſe the Torrent Cedron know that thou art to be
ſlaine: thy bloud shal be vpon thy head. \V And Semei ſaid to the king:
The ſaying is good: as my lord the king hath ſpoken, ſo wil thy ſeruant
doe. Semei therfore dwelt in Ieruſalem, manie dayes. \V And it came to
paſſe after three yeares, that the ſeruantes of Semei fled to Achis the
ſonne of Maacha the king of Geth: and it was told Semei that his
ſeruantes were gone into Geth. \V And Semei aroſe, and ſadled his aſſe,
and went to Achis into Geth to require his ſeruantes, and he brought
them out of Geth. \V And it was told Salomon that Semei went into Geth
out of Ieruſalem, and was returned. \V And ſending he called him, and
ſaid to him: Did I not teſtifie to thee by our Lord, and told thee
before: What day ſoeuer thou going out shalt paſſe hither & thither,
know that thou shalt die? And thou didſt anſwer me: The ſaying is good,
which I haue heard. \V Why then haſt thou not kept the oath of our
Lord, and the precept that I commanded thee? \V And the king ſaid to
Semei: Thou knoweſt al the euil, wherof thy hart is priuy to thy ſelfe,
which thou didſt to Dauid my father: our Lord hath rendred thy malice
vpon thy head: \V And king Salomon be bleſſed, and the throne of Dauid
shal be ſtable before our Lord for euer. \V The king therfore commanded
Banaias the ſonne of Ioiada: who going out, ſtroke him, and he died.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
