%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 0707
%%% o-0635
\startChapter[
  title={Chapter 1}
  ]

\Summary{King Dauid waxing old, Abiſag a Sunamite is brought to
  him. 5.~Adonias pretending to reigne, 11.~Nathan and Bethſabee
  obtaine, 28.~that Salomon is declared and annointed
  King. 41.~VVherupon Adonias (his folowers parting to their houſes)
  50.~fleeth to the altar in the tabernacle, but vpon promiſe of ſafetie
  doth homage to Salomon.}

And
\MNote{The firſt part.

King Dauides admonitions to his ſonne: and his death.}
king Dauid was old, and had manie daies of age: and when he was couered
with clothes, he was not warmed. \V His ſeruantes therfore ſayd to him:
Let vs ſeeke for our lord the king a yong woman a virgin, and let her
ſtand before the king, and cherishe him, and ſleepe in his boſome, and
warme our lord the king. \V They ſought therfore a beautiful yong woman
in al the coſtes of Iſrael, and they found Abiſag a Sunamite, and
brought her to the king. \V And the damſel was exceding beautiful, and
she ſlept with the king, and ſerued him, but the king did not know
her. \V And Adonias the ſonne of Haggith was eleuated, ſaying: I wil
reigne. And he made him ſelf chariotes and horſemen, and fiftie men,
that should runne before him. \V Neither did his father controwle him at
any time, ſaying: Why didſt thou this? And he alſo was very beautiful,
the ſecond borne after Abſalom. \V And he had talke with Ioab the ſonne
of Saruia, & with
\SNote{For this conſpiracie Abiathar was depoſed.
\XRef{ch.~2. v.~27.}}
Abiathar the prieſt, who furthered
%%% o-0636
Adonias ſide. \V But Sadoc the
prieſt, and Banaias the ſonne of Ioiada, and Nathan the prophet, and
Semei, and Rei, and the ſtrength of Dauids armie was not with
Adonias. \V Adonias therfore hauing immolated rammes and calues, and al
fatte beaſtes beſide the Stone zoheleth, which was nigh to the Fountaine
Rogel, called al his brethren the ſonnes of the
%%% 0708
king, and al the men of Iuda the ſeruantes of the king: \V But Nathan
the prophet, and Banaias, and al the ſtrong ones, and Salomon his
brother he called not. \V Nathan therfore ſayd to Bethſabee the mother
of Salomon: Haſt thou not heard that Adonias the ſonne of Haggith hath
reigned, and our lord Dauid is ignorant therof? \V Now therfore come,
take counſel of me, and ſaue thy life, and thy ſonne Salomons. \V Goe,
and enter in to king Dauid, and ſay to him: Didſt not thou my lord king
ſweare to me thy handmayd, ſaying: Salomon thy ſonne ſhal reigne after
me, and he ſhal ſitte in my throne? Why then reigneth Adonias? \V And
whiles thou art yet ſpeaking there with the king, I wil come after thee,
and make vp thy wordes. \V Bethſabee therfore went in to the king in the
chamber: and the king was exceding old, and Abiſag the Sunamite
miniſtred to him. \V Bethſabee bowed her ſelf, and adored the king. To
whom the king ſayd: What is thy wil? quoth he. \V Who anſwering ſayd: My
lord, thou didſt ſweare to thy handmayd by our Lord thy God, that
Salomon thy ſonne ſhal reigne after me, and he ſhal ſitte in my
throne. \V And behold now Adonias reigneth, thou my lord the king not
knowing therof. \V He hath killed oxen, & al fatte thinges, and manie
rammes, and called al the kinges ſonnes, Abiathar alſo the prieſt, and
Ioab the General of the warfarre: but Salomon thy ſeruant he called
not. \V Notwithſtanding my lord king, the eyes of al Iſrael looke vpon
thee, that thou wouldeſt ſhew them, who ſhal ſitte in thy throne my lord
king after thee. \V And it ſhal be when my lord king ſleepeth with his
fathers, I and my ſonne Salomon shal be
\SNote{Baſely eſtemed, or puniſhed as offenders vpon ſuſpition or
ieloſie.}
ſinners. \V As she was yet ſpeaking with the king, Nathan the prophet
came. \V And they told the king, ſaying: Nathan the prophete is
here. And when he was gone in to the king, and had adored bowing to the
earth, \V Nathan ſayd: My lord king, didſt thou ſay: Let Adonias reigne
after me, and let him ſitte vpon my throne? \V Becauſe he is gone downe
to day, and hath immolated oxen, and fattelinges, and manie rammes, and
called al the kinges ſonnes, and the captaynes of the armie, Abiathar
alſo the prieſt: and they eating and drinking before him, and ſaying:
God ſaue the king Adonias: \V me thy ſeruant, and Sadoc the prieſt, and
Banaias the ſonne of Ioiada, and Salomon thy ſeruant he called not. \V
Is this word proceded from my lord the king, and haſt thou not told me
thy ſeruant who
%%% 0709
ſhould ſitte vpon the throne of my lord the king after him? \V And king
Dauid anſwered, ſaying: Cal vnto me Bethſabee. Who when ſhe was entered
in to the king, and ſtood before him, \V the king ſware, and ſayd: Our
Lord liueth, which hath deliuered my ſoule from al diſtreſſe, \V that as
I ſware
%%% o-0637
to thee by our Lord the God of Iſrael, ſaying: Salomon thy ſonne ſhal
reigne after me, and he ſhal ſitte vpon my throne for me, ſo wil I doe
this day. \V And Bethſabee bowing her contenance vnto the earth adored
the king, ſaying: God ſaue my lord for euer. \V King Dauid alſo ſaid:
Cal me Sadoc the prieſt, and Nathan the prophet, & Banaias the ſonne of
Ioiada. Who when they were entred in before the king, \V he ſaid to
them: Take with you the ſeruantes of your lord, and ſette Salomon my
ſonne vpon my mule: and bring him into Gihon. \V And let Sadoc the
prieſt anoint him there, and Nathan the prophet to be king
ouer Iſrael: and you ſhal ſound the trumpet, and ſhal ſay: God ſaue king
Salomon. \V And you ſhal goe vp after him, and he ſhal come, and ſhal
ſitte vpon my throne, and he shal reigne for me: and I wil ordaine him
that he be prince ouer Iſrael, and ouer Iuda. \V And Banaias the ſonne
of Ioiada anſwered the king, ſaying: Amen: ſo ſpeake our Lord the God of
my lord the king. \V As our Lord hath beene with my lord the king, ſo be
he with Salomon, and make his throne higher then the throne of my lord
king Dauid. \V Sadoc therfore the prieſt, and Nathan the prophet went
downe, and Banaias the ſonne of Ioiada, and Cherethi, and Phelethi: and
they ſet Salomon vpon the mule of king Dauid, and brought him into
Gihon. \V And Sadoc the prieſt tooke a horne of oile out of the
tabernacle, and annointed Salomon: and they ſounded the trumpet, and al
the people ſaid: God ſaue king Salomon. \V And al the multitude went vp
after him, and the people ſinging on ſhaulmes, and reioyſing with great
gladnes, and the earth ſounded of their crie. \V And Adonias heard, and
al that were inuited of him, and the feaſt was ended: yea and Ioab
hearing the voice of the trumpet, ſaid: What meaneth the crie of the
citie making a tumult? \V As he yet ſpake, came Ionathas the ſonne of
Abiathar the prieſt: to whom Adonias ſaid: Come in, becauſe thou art a
ſtout man, and bringeſt good newes. \V And Ionathas anſwered Adonias:
Not ſo: for our lord king Dauid hath appointed Salomon king, \V and hath
ſent with
%%% 0710
him Sadoc the prieſt, and Nathan the prophete, and Banaias the ſonne of
Ioiada, and Cerethi, and Phelethi, and they haue ſet him vpon the kinges
mule. \V And Sadoc the prieſt, and Nathan the prophete haue annointed
him king in Gihon: & they are gone vp thence reioyſing, and the citie
ſounded: this is the voice that you heard. \V Yea and Salomon ſitteth
vpon the throne of the kingdom. \V And the kinges ſeruantes entring in,
haue bleſſed our lord king Dauid, ſaying: God amplify the name of
Salomon aboue thy name, and magnifie his throne aboue thy throne. And
the king
\SNote{King Dauid did not adore his ſonne as a ſubiect adoreth his
prince, but adored God, geuing thankes for this benefite of a ſucceſſor,
as it foloweth in the next verſe.}
adored in his bed: \V and he hath thus ſpoken: Bleſſed be our Lord the
God of Iſrael, who hath geuen this day one ſitting in my throne, mine
eies ſeing it. \V They therfore were terrified, and they al aroſe, that
had beene inuited of Adonias, and euery man went his way. \V And Adonias
fearing Salomon aroſe, and went, and held the horne of the altar. \V And
they told Salomon,
%%% o-0638
ſaying: Behold Adonias fearing king Salomon, hath taken hold of the
horne of the altar, ſaying: Let king Salomon ſweare to me this day, that
he wil not kil his ſeruant with the ſword. \V And Salomon ſayd: If he be
a good man, there ſhal not ſo much as one heare of his fal vpon the
ground: but if euil ſhal be found in him, he ſhal die. \V King Salomon
therfore ſent, and brought him out from the altar: and going in he
adored king Salomon: and Salomon ſayd to him: Goe to thy houſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
