%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 0719
%%% o-0645
\startChapter[
  title={Chapter 6}
  ]

\Summary{In the yeare foure hundred and foure ſcore after the Iſraelites
  came from Ægypt, Salomon, the fourth yeare of his reigne, begineth to
  build the Temple. 2.~The principal partes with the greatnes, forme,
  and ornamentes therof are deſcribed. 38.~It is in building ſeuen
  yeares.}

And it came to paſſe in the foure hundred and foure ſcore yeare of the
coming forth of the children of Iſrael out of the Land of Ægypt, in the
fourth yeare, the moneth Zio (that is the ſecond moneth) of the reigne
of Salomon ouer Iſrael, he began to build a houſe to our Lord. \V And
the houſe, which king
%%% o-0646
Salomon built to our Lord, had three ſcore cubites in length, and
twentie cubites in bredth, and thirtie cubites in height. \V And there
was a porche before the temple of twentie cubites of length, according
to the meaſure of the bredth of the temple: and it had ten cubites of
bredth before the face of the temple. \V And he made in the temple
\TNote{Narovv vvithout & broad vvithin.}
oblique windowes. \V And he built vpon the wal of the temple loftes
round about, \V in the walles of the houſe round about the temple, and
the oracle, and he made ſides round about. \V The loft that was
vnderneth, had fiue cubites of bredth, & the middle loft was of ſix
cubites in bredth, and the third loft had ſeuen cubites of bredth. And
he put beames in the houſe round about on the outſide, that they might
not cleaue to the walles of the temple. \V And the houſe when it was
built, was built of ſtones hewed and perfected: and hammer, and hatchet,
and al the toole of yron were
\SNote{In perfect people is neither noiſe of murmur in aduerſitie, nor
of boaſting in proſperitie, but patience and humilitie with ſilence.}
not heard in the houſe when it was built. \V The doore of the middle
ſide was in the wal of the houſe on the right hand: and by wynding
ſtaires they went vp into the middle rowme, and from the middle into the
third. \V And he built the houſe, and finished it: he couered alſo the
houſe with ſeelings of cedre trees. \V And he built a loft ouer al the
houſe fiue cubites of height, and he couered the houſe with cedre
timber. \V And the word of our Lord came to Salomon, ſaying: \V This
houſe, which thou buildeſt, if thou wilt walke in my preceptes, and doe
my iudgementes, and keepe al my commandementes, going in them, I wil
eſtablish my word to thee, which I ſpake to Dauid thy father. \V And I
%%% 0720
wil dwel in the middes of the children of Iſrael, and wil not forſake my
people Iſrael. \V Salomon therfore built the houſe and finished it. \V
And he built the walles of the houſe on the inſide, with cedre loftes, from
the pauement of the houſe to the toppe of the walles, and to the roofes,
he couered it with cedre trees on the inſide: and he couered the floore
of the houſe with boordes of firre. \V And he built loftes of cedre
timber of twentie cubites at the hinder part of the temple, from the
pauement to the higher partes: and he made the inner houſe of the oracle
to be
\SNote{\Emph{Holie of holies} or moſt holie place.}
\L{Sanctum Sanctorum}. \V Moreouer the temple it ſelf was fourtie
cubites before the doores of the oracle. \V And al the houſe was couered
within with ceder, hauing roundels, and the ioyntes therof coningly
wrought and the engrauinges ſtanding out: al thinges were couered with bordes:
\SNote{Stones in buildinges, and bones in liuing bodies repreſent the
ſtate of religious perſones: who being hidde in their Monaſteries and
celles, ſo much the more fortifie the Church, by how much leſſe they
appeare abrode, becauſe \Emph{their office is not to teach, but to
mourne}.
\Cite{S.~Bernard. Ser.~64.}}
neither could there a ſtone appeare in the wal at al. \V And he made the
oracle in the middes of the houſe, in the inner part, that he might put
the arke of couenant of our Lord there. \V Moreouer the oracle had
twentie cubites in length, and twentie cubites of bredth, and twentie
cubites in height. And he couered and ſeeled it with moſt pure gold. And
the altar alſo he decked with ceder. \V The houſe alſo before the oracle
he couered with moſt pure gold, and faſtened on plates with nailes of
gold. \V And there was nothing in the temple that was not couered with
gold: yea and al the altar of the oracle he couered with gold. \V And he
made in the oracle
%%% o-0647
two cherubs of oliue trees, of ten cubites in height. \V One wing of a
cherub of fiue cubites, and the other wing of a cherub fiue cubites:
that is, hauing tenne cubites, from the end of one wing vnto the end of
the other wing. \V Of ten cubites alſo was the ſecond cherub: in like
meaſure, and the worke was one in both cherubs, \V that is to ſay, one
cherub had the height of ten cubites, and in like maner the ſecond
cherub. \V And he put the cherubs in the middes of the inner temple: and
the cherubs extended their winges, and the one wing touched the wal, and
the wing of the ſecond cherub touched the other wal: and the other
winges in the middle part of the temple touched ech other. \V He couered
alſo the cherubs with gold. \V And al the walles of the temple round
about he graued with diuerſe engrauinges and caruing: & he made in them
cherubs, and palme trees, and
\SNote{It is a clere caſe, that al carued & grauen pictures, or images
were not vnlawful, but were religiouſly made & ſette in the holie
Temple, for the more honour of God.}
diuerſe pictures, as it were ſtanding
%%% 0721
out of the wal, and coming forth. \V Yea the pauement alſo of the houſe
he couered with gold within and without. \V And in the entrance of the
oracle he made litle doores of the timber of oliuetrees, and fiue corner
poſtes. \V And two doores of oliuetimber: and he graued in them pictures
of Cherubs, and figures of Palme trees, and grauen workes ſtanding out
very much; and he couered them with gold: and he couered as wel the
cherubs as the palmetrees, and the other thinges with gold. \V And he
made in the entrance of the temple poſtes of oliuetimber foure
ſquare: \V and two doores of firre trees, one agaynſt an other: and
either doore was duble, and ſo opened with folding leaues. \V And he
graued cherubs, and palmetrees, and engrauinges appearing very much: and
he couered al with golden plates in ſquare worke by rule. \V And he
built the inner court with three rowes of ſtones polished, and one rowe
of ceder timber. \V In the fourth yeare was the houſe of our Lord
founded in the moneth of Zio: \V
\MNote{The end of the fourth age.}
and in the eleuenth yeare in the moneth Bul (that is the eight moneth)
the houſe was perfected in al the workes therof, and in al the
implementes therof: and he was building it ſeuen yeares.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
