%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 0740
%%% o-0665
\startChapter[
  title={Chapter 8}
  ]

\Summary{The arke is brought in, and the temple dedicated. 10.~A
  glorious clovvde replenisheth it. 14.~Salomon prayeth long to God,
  55.~bleſseth the people, 62.~and manie victimes are offered in this
  ſolemne feſtiuitie.}

Then were gathered together al the ancientes of Iſrael with the princes
of the tribes, and the heades of the families of the children of Iſrael
to king Salomon into Ieruſalem: that they might carrie the Arke of the
couenant of our Lord out of the citie of Dauid, that is, out of Sion. \V
And al Iſrael aſſembled to king Salomon in the moneth of Ethanim, on a
ſolemne day, that is the ſeuenth moneth. \V And al the ancientes of
Iſrael came, and the prieſtes tooke the arke, \V and caried the arke of
our Lord, and the tabernacle of couenant, and al the veſſels of the
Sanctuarie, that were in the tabernacle: and the Prieſtes and the
Leuites caried them.
%%% 0741
\V And king Salomon, and al the multitude of Iſrael, which was aſſembled
vnto him, went with him before the arke, and they immolated ſheepe and
oxen without eſtimation & number. \V And the prieſtes brought in the
arke of the couenant of our Lord into his place, into the oracle of the
temple, into \L{Sanctum ſanctorum} vnder the winges of the cherubs. \V
For the cherubs ſpred their winges ouer the place of the arke, and
couered the arke, and the barres therof aboue. \V And wheras the barres
ſtood out, and the endes of them appeared without in the Sanctuarie
before the oracle, they appeared no farder outward, which alſo were there
vntil this preſent day. \V And in the arke there was
\SNote{There was no more within the arke,
\XRef{Deut.~10.}
but on the outſide was the rodde of Aaron,
\XRef{Nu.~17.}
\XRef{Heb.~9.}
the golden potte with Manna,
\XRef{Exod.~16.}
\XRef{Heb.~9.}
and the booke of the law repeted by Moyſes.
\XRef{Deut.~31.}}
nothing els but two tables of ſtone, which Moyſes put in it in Horeb,
when our Lord made the couenant with the children of Iſrael, when they
came out of the Land of Ægypt. \V And it came to paſſe, when the
prieſtes were gone out of the Sanctuarie, a clowde filled the houſe of
our Lord, \V and the prieſtes could not ſtand and miniſter for the
clowde: for the glorie of our Lord had filled the houſe of our Lord. \V
Then ſayd Salomon: Our Lord ſayd that he would dwel in a clowde. \V
Building I haue built a houſe for thy habitation, thy moſt firme throne
for euer. \V And the king turned his face, and
\SNote{Prĩces bleſſe their people & parentes their children.}
bleſſed al the church of Iſrael: for al the church
\Fix{if}{of}{obvious typo, fixed in other}
Iſrael ſtood. \V And
%%% o-0666
Salomon ſayd: Bleſſed be our Lord the God of Iſrael, who ſpake by his
mouth to Dauid my father, & in his owne handes hath perfected it,
ſaying: \V Since the day that I brought my people Iſrael out of Ægypt, I
choſe no citie of al the tribes of Iſrael, that a houſe might be built,
and my name might be there: but I choſe Dauid to be ouer my people
Iſrael. \V And Dauid my father would haue built a houſe to the name of
our Lord the God of Iſrael: \V and our Lord ſayd to Dauid my father: In
that thou haſt thought in thy hart to build a houſe to my name, thou
haſt done wel, caſting this ſame thing in thy mynd. \V Neuertheleſſe
thou ſhalt not build me a houſe, but thy ſonne, that shal come forth of
thy reynes, he shal build a houſe to my name. \V Our Lord hath confirmed
his word, which he ſpake: and I ſtand for Dauid my father, and ſitte
vpon the throne of Iſrael, as our Lord hath ſpoken: and I haue built a
houſe to the name of our Lord the God of Iſrael. \V And I haue appoynted
there a place for the arke, wherin the couenant of our Lord is, which he
made with our fathers,
%%% 0742
when they came out of the Land of Ægypt. \V And Salomon ſtood before the
altar of our Lord in the ſight of the aſſemblie of Iſrael, and extended
his handes toward heauen, \V and ſayd: Lord God of Iſrael, there is not
God like to thee in heauen aboue, and vpon the earth beneth: which
keepeſt couenant and mercie with thy ſeruantes, that walke before thee
in al their hart. \V Which haſt kept to thy ſeruant Dauid my father, the
thinges that thou haſt ſpoken to him: by mouth thou didſt ſpeake, and
with thy handes thou haſt accomplished, as this day proueth. \V Now
therfore Lord God of Iſrael, keepe vnto thy ſeruant Dauid my father the
thinges which thou haſt ſpoken to him, ſaying: There shal not be taken
away of thee a man before me, which ſitteth vpon the throne of Iſrael:
\SNote{Salomon knew wel Gods conditional promiſe, but perſeuered not in
keping his cõmandments, and therfore a great part of the kingdom was
takẽ from his children, yet the right of the kingdom of Iuda remayned to
his ſeede euen to Chriſt our Sauiour.}
yet ſo if thy children shal keepe their way, that they walke before me
as thou haſt walked in my ſight. \V And now Lord God of Iſrael, let thy
wordes be eſtablished, which thou haſt ſpoken to thy ſeruant Dauid my
father. \V Is it then to be thought that in deede God dwelleth vpon the
earth? for if heauen, and the heauens of heauens can not conteyne thee,
how much more this houſe, which I haue built? \V But looke toward the
prayer of thy ſeruant, and to his petitions ô Lord my God: heare the
hymne and the prayer, which thy ſeruant prayeth before thee this day: \V
that thy eies be opened vpon this houſe night and day: vpon the houſe,
wherof thou ſayeſt: My name ſhal be there: that thou heare the prayer,
which thy ſeruant prayeth in this place to thee. \V That thou heare the
requeſt of thy ſeruant and of thy people Iſrael, whatſoeuer they shal
pray for in this place, and thou shalt heare in the place of thy
habitation in heauen: and when thou haſt heard, thou shalt be
merciful. \V If a man shal ſinne agaynſt his neighbour, and shal haue
any oath, wherwith he is held faſt bound: and shal come becauſe of the
oath before thine altar into thy houſe, \V thou shalt heare in heauen:
and shalt doe, and iudge thy ſeruantes, condemning the impious, and
rendring his way vpon his head, and iuſtifying the iuſt, and
%%% o-0667
\SNote{Reward of good workes.}
rewarding him according to his iuſtice. \V If thy people Iſrael shal
flee their enemies (becauſe they wil ſinne agaynſt thee) and doing
penance, and confeſſing to thy name, shal come, and pray, and beſech
thee in this houſe; \V heare in heauen, and forgeue the ſinne of thy
people Iſrael, and thou shalt reduce them vnto the land, which thou
gaueſt to their
%%% 0743
fathers. \V If the heauen shal be shut, and it rayne not, becauſe of
their ſinnes, and praying in this place, they doe penance to thy name,
and shal be conuerted from their ſinnes through their affliction: \V
heare them in heauen, and forgeue the ſinnes of thy ſeruantes, and of
thy people Iſrael: and shew them a good way wherin they may walke, and
geue rayne vpon thy land, which thou haſt geuen to thy people in
poſſeſſion. \V If famine aryſe in the land, or peſtilence, or corrupt
ayre, or blaſting, or locuſt, or ruſt, and their enemie afflict them
beſieging the gates, al plague, al infirmity, \V al curſing, and
banning, that shal chance to any man of thy people Iſrael: if any man
shal know the wound of his hart, and shal ſpred forth his handes in this
houſe, \V thou ſhalt heare in heauen, in the place of thy habitation,
and shalt be merciful agayne, and shalt ſo doe that thou geue to euerie
one according to his wayes, as thou shalt ſee his hart (for thou onlie
knoweſt the hart of al the children of men) \V that they feare thee al
the dayes, which they liue vpon the face of the land, which thou haſt
geuen our fathers. \V Moreouer alſo the ſtranger, which is not of thy
people Iſrael, when he shal come from a farre countrie for thy name (for
thy great name shal be heard of, and thy ſtrong hand, \V and thy
ſtretched out arme euerie where) when therfore he shal come, and shal
pray in this place, \V thou shalt heare in heauen, in the firmament of
thy habitation, & thou shalt doe al thinges, for the which the ſtranger
shal inuocate thee: that al the peoples of the earth may lerne to feare
thy name, as thy people Iſrael, and may proue that thy name is inuocated
vpon this houſe, which I haue built. \V If thy people shal goe forth to
warre agaynſt their enemies, by the way, whitherſoeuer thou shalt ſend
them, they shal pray to thee agaynſt the way of the citie, which thou
haſt choſen, and agaynſt the houſe, which I haue built to thy name, \V
and thou shalt heare in heauen their prayers, and their petitions, and
shalt doe iudgement for them. \V But if they shal ſinne to thee (for
there is no man which ſinneth not) and thou being wrath shalt deliuer
them to their enemies, and they shal be led captiue into the land of
their enemies farre or neere, \V and shal doe penance in their
\SNote{External workes of penance, except they proceed from the hart,
ſuffice not for remiſſion of ſinne.}
hart in the place of captiuitie, and conuerted shal beſech thee in their
captiuitie, ſaying: We haue ſinned, we haue done wickedly, we haue dealt
impiouſly: \V and
%%% 0744
shal returne to thee in al their hart, and al their ſoule, in the land
of their enemies, to the which they shal be led captiue: & shal pray to
thee agaynſt the way of their land, which thou gaueſt to their fathers,
and of the citie which thou haſt choſen, & of the temple which I haue
built to thy name: \V thou shalt heare in heauen, in the firmament of
thy throne their
%%% o-0668
prayers, and their petitions, and shalt doe their iudgement for them: \V
and shalt be merciful to thy people, which ſinned to thee, and to al
their iniquities, wherewith they haue tranſgreſſed agaynſt thee: and
thou shalt geue mercie before them, that shal haue them captiues, that
they may haue compaſſion on them. \V For they are thy people, and thine
inheritance, whom thou haſt brought out of the Land of Ægypt, from the
middes of the yron fornace. \V That thy eies be open to the petition of
thy ſeruant, and of thy people Iſrael, & thou heare them in al thinges
for which they shal inuocate thee. \V For thou haſt ſeparated them to
thee for an inheritance from al the peoples of the earth, as thou haſt
ſpoken by Moyſes thy ſeruant, when thou didſt bring our fathers out of
Ægypt, Lord God. \V And it came to paſſe, when Salomon had accomplished
praying to our Lord al this prayer and petition, he roſe from the ſight
of the altar of our Lord: for he had faſtened both knees on the ground,
and had ſpred his handes toward heauen. \V He ſtood therfore and bleſſed
al the aſſemblie of Iſrael with a lowde voyce, ſaying: \V Bleſſed be our
Lord, which hath geuen reſt to his people Iſrael, according to al
thinges that he hath ſpoken: there hath not fayled ſo much as one word
of al the good thinges, that he ſpake by Moyſes
\Fix{this}{his}{obvious typo, fixed in other}
ſeruant. \V Be our Lord God with vs, as he hath beene with our fathers,
not forſaking, nor reiecting vs. \V But incline he our hartes to him,
that we may walke in al his waies, and keepe his commandementes, and his
ceremonies, and iudgementes whatſoeuer he commanded our fathers. \V And
be theſe my wordes, wherewith I haue prayed before our Lord, approching
to our Lord God day and night, that he may doe iudgement for his
ſeruant, and for his people Iſrael day by day: \V that al the peoples of
the earth may knowe, that our Lord he is God, and there is none other
beſides him. \V Let our hart alſo be perfect with our Lord God, that we
walke in his decrees, and keepe his cõmandementes, as alſo this day. \V
Therfore the king, and al Iſrael with him,
%%% 0745
did immolate victimes before our Lord. \V And Salomon killed pacifique
hoſtes, which he immolated to our Lord, of oxen two and twentie
thouſand, and of sheepe an hundred twentie thouſand: and they dedicated
the temple of our Lord, the king, and the children of Iſrael. \V In that
day the king ſanctified the middes of the court, that was before the
houſe of our Lord: for he made the holocauſte there, and ſacrifice,
and fatte of the pacifiques: becauſe the braſen altar, that was before
our Lord, was too litle, and could not take the holocauſte, and
ſacrifice, and fatte of the pacifiques. \V Salomon therfore made in that
time a ſolemne feſtiuitie, and al Iſrael with him, a great multitude
from the entrance of Emath vnto the Ryuer of Ægypt, before our Lord God,
ſeuen daies and ſeuen daies, that is, fourtene daies. \V And in the
eight day he diſmiſſed the people: Who
%%% !!! SNote marked only in this, text missing in both 
%%% \SNote{}
bleſſing the king, went into their tabernacles reioycing, and with a
ioyful hart for al the good thinges, that our Lord had done to Dauid his
ſeruant, and to Iſrael his people.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
