%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 0762
%%% o-0683
\startChapter[
  title={Chapter 16}
  ]

\Summary{Iehu for prophecying the deſtruction of Baaſa and his houſe,
  7.~is ſlaine. 8.~Yet his ſonne Ela reigneth tvvo yeares. 9.~Then
  Zambri rebelleth, killeth Ela, and reigneth. 16.~Part of the people
  chooſing Amri (prince of the armie) their king, 18.~Zambri deſporatly
  burneth himſelfe and the kings palace. 21.~An other part folow Thebni
  as king til his death. 23.~Amri reigneth tvvelue yeares
  vvickedly. 29.~His ſonne Achab ſuccedeth, marieth Iezabel, and ſerueth
  Baal. 34.~In the meane time Hiel repaireth Iericho.}

And the word of our Lord came to
\Fix{Iohu}{Iehu}{obvious typo, fixed in other}
the ſonne of Hanani agaynſt Baaſa, ſaying: \V For ſo much as I haue
exalted thee out of the duſt, & ſette thee duke ouer my people Iſrael,
but thou haſt walked in the way of Ieroboam, and haſt made my people
Iſrael to ſinne, that thou mighteſt anger me with their ſinnes: \V
behold, I wil cut downe the poſteritie of Baaſa, and the poſteritie of
his houſe, and I wil make thy houſe as the houſe of Ieroboam the ſonne
of Nabat. \V Whoſoeuer of Baaſa shal die in the citie, him shal the
dogges eate: and whoſoeuer of his shal die in the countrie, him shal the
fowles of the ayre deuoure. \V But the reſt of the wordes of Baaſa, and
whatſoeuer he did, and his battels, are not theſe things written in the
%%% !!! XRef?
Booke of the wordes of the daies of the kinges of Iſrael? \V Baaſa
therfore ſlept with his fathers, and was buried in Therſa: and Ela his
ſonne reigned for him. \V And when the word of our Lord came in the hand
of Iehu
%%% o-0684
the ſonne of Hanani the prophete agaynſt Baaſa, and agaynſt his houſe,
and agaynſt al the euil, that he had done before our Lord, to anger him
in the workes of his handes, that it should be made as the houſe of
Ieroboam: for this cauſe he ſlew him, that is to ſay, iehu the ſonne of
Hanani, the prophete. \V In the ſixe and twenteth yeare of Aſa the king
of Iuda, reigned Ela the ſonne of Baaſa ouer Iſrael in Therſa two
yeares. \V And his ſeruant Zambri rebelled agaynſt him, the captayne of
the
%%% 0763
halfe part of the horſemen: and Ela was in Therſa drinking, and dronken
in the houſe of Arſa the gouernour of Therſa. \V Zambri therfore rushing
in, ſtroke and ſlew him in the ſeuen and twenteth yeare of Aſa the king
of Iuda, & he reigned for him. \V And when he reigned, and ſate vpon his
throne, he ſtroke al the houſe of Baaſa, and he leaft not of it one that
could pyſſe agaynſt a wal, & his kinsfolke and frendes. \V And Zambri
deſtroyed al the houſe of Baaſa, according to the word of our Lord, that
he had ſpoken to Baaſa in the hand of Iehu the prophet, \V for al the
ſinnes of Baaſa, and the ſinnes of Ela his ſonne, who ſinned, and made
Iſrael to ſinne, prouoking our Lord the God of Iſrael in their
vanities. \V But the reſt of the wordes of Ela, and al that he did, are
not theſe writen in the
%%% !!! XRef?
Booke of the wordes of the daies of the kinges of Iſrael? \V In the
ſeuen and twenteth yeare of Aſa the king of Iuda, reigned Zambri ſeuen
daies in Therſa: moreouer the armie beſieged Gebbethon a citie of the
Philiſthines. \V And when they heard that Zambri had rebelled, and
ſlayne the king,
\SNote{Al thoſe that were in the campe choſe their general to be their
king and preuailed therin: though an other half of Iſrael choſe and
folowed an other for a time.}
al Iſrael made Amri their king, who was General of the warfare ouer
Iſrael that day, in the campe. \V Amri therfore went vp, and al Iſrael
with him from Gebbethon, and they beſieged Therſa. \V And Zambri ſeing
that the citie should be taken, he went into the palace, and burnt him
ſelf with the kinges houſe: and he died \V in his ſinnes, which he had
ſinned doing euil before our Lord, and walking in the way of Ieroboam,
and in his ſinne; wherwith he made Iſrael to ſinne. \V But the reſt of
the wordes of Zambri, and of his treaſon, and tyrannie, are not theſe
thinges writen in the
%%% !!! XRef?
Booke of the wordes of the dayes of the kinges of Iſrael? \V Then was
the people of Iſrael diuided into two partes: the halfe part of the
people folowed Thebni the ſonne of Gineth, to make him king: and the
halfe part Amri. \V But the people that was with Amri, preuayled ouer
the people that folowed Thebni the ſonne of Gineth: and Thebni died, and
Amri reigned. \V In
\SNote{Thebni being then dead he reigned peacably for he began his
reigne the 17.~yeare of Aſa
\XRef{v.~15. &~16.}
and reigned in al 12.~yeares.}
the one and thirteth yeare of Aſa the king of Iuda Amri reigned ouer
Iſrael, twelue yeares: in Therſa he reigned ſix yeares. \V And he bought
the mount of Samaria of Somer for two talentes of ſiluer: and he built
it, and he called the citie which he had built, by the name of Semer the
lord of the mount of Samaria. \V And Amri did euil in the ſight of our
Lord, and wrought wickedly aboue al, that were
%%% 0764
before him. \V And he walked in al the way of Ieroboam the ſonne of
Nabat, and in his ſinnes wherwith he made Iſrael to ſinne: that they
might anger our Lord the God of Iſrael in their vanities.
%%% o-0685
\V But the reſt of the wordes of Amri, and the battels he made, are not
theſe thinges writen in the
%%% !!! XRef?
Booke of the wordes of the daies of the kinges of Iſrael? \V And Amri
ſlept with his fathers, and was buried in Samaria and Achab his ſonne
reigned for him. \V But Achab the ſonne of Amri reigned ouer Iſrael the
eight and thirteth yeare of Aſa the king of Iuda. And Achab the ſonne of
Amri reigned ouer Iſrael in Samaria two and twentie yeares. \V And Achab
the ſonne of Amri did euil in the ſight of our Lord aboue al, that were
before him. \V Neither did it ſuffice him that he walked in the ſinnes
of Ieroboam the ſonne of Nabat: beſides he tooke to wife Iezabel the
daughter of Ethbaul the king of the Sidonians. And he went, and ſerued
Baal, and adored him. \V And he ſette an altar to Baal in the temple of
Baal, which he had built in Samaria, \V and he planted a groue: and
Achab added in his worke, prouoking our Lord the God of Iſrael aboue al
the kinges of Iſrael, that were before him. \V In his daies Hiel of
Bethel built Iericho:
\SNote{VVhen Hiel began to build Iericho, his eldeſt ſonne died, ſo the
reſt ſucceſſiuely, that the laſt died when he finiſhed the building:
becauſe God by the mouth of Ioſue had forbid the building therof.}
in Abiram his firſt borne he founded it, and in Segub his laſt he ſette
vp the gates therof: according to the word of our Lord, which
\CNote{\XRef{Ioſue.~8.}}
he ſpake in the hand of Ioſue the ſonne of Nun.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
