%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 0771
%%% o-0691
\startChapter[
  title={Chapter 20}
  ]

\Summary{The Syrians threatning and beſieging Samaria, 13.~God
  ſignifieth by a prophet to Achab, that he shal haue the
  victorie. 20.~Which he obtaineth. 23.~Alſo the ſecond yeare fighting
  in the champaine. 31.~But ſauing the king of
  \Fix{Syrians}{Syrias}{possible typo, same in both}
  life, and making league with him, 35.~one of the children of the
  prophetes being ſlaine, for not ſtryking when he was ſo commanded,
  37.~an other denounceth reuenge to Achab, for not killing the Syrian
  King.}

Moreouer Benadad the king of Syria, gathered together al his hoſt, & two
& thirtie kinges with him, & horſes, and chariotes: and going vp fought
agaynſt Samaria, and beſieged it. \V And ſending meſſengers to Achab the
king of Iſrael into the citie, \V he ſaid: Thus ſaith Benadad: Thy
ſiluer, and thy gold is mine: and thy wiues, and thy principal children
be mine. \V And the king of Iſrael anſwered: According to thy word my
lord king, I am thine, and al that is mine. \V And the meſſengers
returning, ſaid: Thus ſaith Benadad, which ſent vs vnto thee: Thy
ſiluer, and thy gold, and thy wiues, and thy children thou shalt geue
me. \V To morow therfore this verie houre I wil ſend my ſeruantes to
thee, and they shal ſearch thy houſe, and the houſe of thy ſeruantes:
and al that pleaſeth them, they shal put in their handes, and take
away. \V And the king of Iſrael called al the ancientes of the land, and
ſaid: Marke, and ſee that he ſeeketh to intrappe vs. For he ſent to me
for my wiues, and children, and for the ſiluer and gold: and I ſaid not
nay. \V And al the ancientes, and al the people ſaid to him: Heare not,
neither agree vnto him. \V He therfore anſwered the meſſengers of
Benadad: Tel my lord the king: Al thinges for the which thou didſt ſend
to me thy ſeruant in the beginning I wil doe: but this thing I can not
doe. \V And the meſſengers returning made report vnto him, who ſent
againe, and ſayd: Theſe thinges doe the goddes to me, and theſe adde
they, if the duſt of Samaria shal ſuffice for the
\Fix{hanfulles}{handfulles}{obvious typo, fixed in other}
of al the people, that foloweth me. \V And the king of Iſrael anſwering,
ſayd: Tel him:
\SNote{Not at the putting on of armour, is time to boaſt of victorie,
but rather at putting it of. For thoſe that are girded, and ſtrongly
armed do vnwiſely in truſting their owne ſtrength, and contemning their
aduerſaries: for by ſo doing they are often ouerthrowne. And thoſe that
diſtruſting them ſelues truſt in God, one way or other alwaies
preuaile. A neceſſarie leſſon, both in temporal and ſpiritual warfayre.}
Let not the girded glorie as the vngirded. \V And it came to paſſe, when
Benadad had heard this word, himſelf and the kinges dranke in pauilions,
and he ſayd to his ſeruantes: Beſette the citie, and they did beſette
it. \V And behold a prophete coming to Achab the king of Iſrael, ſaid to
him: Thus ſayth our Lord: Haſt thou in dede ſene al this exceding great
multitude?
%%% 0772
behold, I wil deliuer them into thy hand this day: that thou mayſt know,
that I am the Lord. \V And Achab ſaid: By whom?
%%% o-0692
And he ſaid to him: Thus ſaith our Lord: By the ſeruantes of the princes
of the prouinces. And he ſaid: Who ſhal begin to fight? And he ſaid:
Thou. \V He therfore muſtered the ſeruantes of the princes of the
prouinces, and he found the number of two hundred thirtie two: and he
muſtered after them the people, al the children of Iſrael, ſeuen
thouſand. \V And they went forth at noone. But Benadad dranke al dronken
in his tent, and two and thirtie kinges with him, which were come to ayde
him. \V The ſeruantes therfore of the princes of the prouinces iſſued
forth in the forefront. Benadad therfore ſent. Who told him, ſaying: Men
are come forth out of Samaria. \V And he ſaid: Whether they come for
peace, take them aliue: or els to fight, aliue take ye them. \V The
ſeruantes therfore of the princes of the prouinces iſſued forth, and the
reſt of the armie folowed: \V and euerie one ſtroke the man, that came
againſt him: and the Syrians fled, and Iſrael purſued them. Benadad alſo
the king of Syria fled on horſebacke with his horſemen. \V Moreouer the
king of Iſrael iſſuing forth ſtroke the horſes and chariotes, and he
ſtroke Syria with a great ſlaughter. (\V And a prophet coming to the king
of Iſrael, ſaid to him: Goe, and take courage, and know, and ſee what
thou doeſt: for the yeare folowing the king of Syria wil come vp againſt
thee.) \V But the ſeruantes of the king of Syria ſaid to him: The goddes of
the mountaynes be their goddes, therfore haue they ouercome vs: but it
is better that we fight againſt them in the champaine, & we ſhal
ouercome them. \V Thou therfore do this word: Remoue al the kinges from
thine armie, and put captaines for them: \V and repaire the number of
ſouldiars, that are ſlaine of thine, and horſes according to the old
horſes, & chariotes according to the chariotes, which thou hadſt before:
and we wil fight againſt them in the champaine, and thou ſhalt ſee that
we shal ouercome them. He beleued their counſel, and did ſo. \V Therfore
after a yeare was paſſed, Benadad muſtered the Syrians, and went vp into
Aphec, to fight againſt Iſrael. \V Moreouer the children of Iſrael were
muſtered, and taking victuals they went forth on the contrarie ſide, and
camped againſt them, as it were two litle flockes of goates: but the
Syrians filled the land. (\V And a man of God coming, ſaid to the king
of Iſrael:
%%% 0773
Thus ſaith our Lord: Becauſe the Syrians haue ſaid: The Lord is God of
the mountaines, and is not God of the Valleis: I wil geue al this great
multitude into thy hand, and
\SNote{Manie victories and other benefites were beſtowed vpon Achab, to
make him knowe God: but he cõtemning them al remained in his impietie,
and finally was ſlaine.
\XRef{ch.~22. v.~38.}}
you ſhal know that I am the Lord.) \V And ſeuen dayes did theſe, and
they direct their armies one againſt the other, and in the ſeuenth day
was the battel fought: and the children of Iſrael ſtroke the Syrians an
hundred thouſand footemen in one day. \V And they that remained in
Aphec, fled into the citie: and the wal fel vpon ſeuen and twentie
thouſand men, that were leaft. Moreouer Benadad fleing entered the citie,
into a chamber that was within a chamber, \V and his ſeruantes ſaid to
him: Behold, we haue heard that the kinges of the houſe of Iſrael are
merciful: Let vs therfore put ſackeclothes on our
%%% o-0693
loynes, and cordes on our heades, and goe forth to the king of Iſrael:
perhaps he wil ſaue our liues. \V They girded their loines with
ſackclothes, and put cordes on their heades, and came to the king of
Iſrael, and ſaid to him: Thy ſeruant ſaith: Let my ſoule liue, I beſech
thee. And he ſaid: If he be yet aliue he is my brother. \V Which the men
tooke for good lucke: and in haſt caught the word of his mouth, and
ſaid: Thy brother Benadad. And he ſaid to them: Goe and bring him to
me. Benadad therfore came out to him, and he lifted him vp into his
chariote. \V Who ſaid to him: The cities which my father tooke from thy
father, I wil render: doe thou make thee ſtretes in Damaſcus, as my
father made in Samaria, and I confederate wil depart from thee. He
therfore made a league, and dimiſſed him. \V Then a certaine man of the
children of the prophetes ſayd to his felow in the word of our Lord:
Strike me. But he would not ſtrike. \V To whom he ſaid: Becauſe thou
wouldeſt not heare the word of our Lord, behold thou ſhalt depart from
me, and a lion shal ſtrike thee. And when he was departed a litle from
him, a lyon found him, and ſlew him. \V But finding alſo an other man,
he ſaid to him: Strike me. Who ſtroke him, and wounded him. \V The
prophete therfore went, and mette the king in the way, and with
ſprinkling of duſt changed his face and his eies. \V And when the king
paſſed by, he cried to the king, and ſaid: Thy ſeruant went forth to
fight hand ſtrokes: and when a certaine man was fled, one brought him to
me, and ſaid: Keepe this man: who if he shal ſlippe away, thy life shal
be for his life, or thou shalt pay a talent of ſiluer. \V And
%%% 0774
whiles I being trubled turned hither and thither, ſodenly he appeared
not. And the king of Iſrael ſayd to him: This is thy iudgement, which
thy ſelf haſt decreed. \V But he forthwith wyped of the duſt from his
face, and the king of Iſrael knew him, that he was of the prophetes. \V
Who ſayd to him: Thus ſayth our Lord:
\SNote{Fooliſh pittie in ſparing a dangerous & common enimie is
offenſiffe to God, & ſeuerely puniſhed by his iuſtice.}
Becauſe thou haſt let goe out of thy hand a man worthie to die, thy life
ſhal be for his life, and thy people for his people. \V The king of
Iſrael therfore returned into his houſe, contemning to heare, & raging
came into Samaria.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
