%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 0764
%%% o-0685
\startChapter[
  title={Chapter 17}
  ]

%%% !!! Original verse numbers were wrong. Silently fixed.
\Summary{Elias by his prayer shutteth the heauen from raining. 6.~Is fed
  by a crovv, 15.~and by a vvidovv of Sareptha, 16.~vvhoſe potte of
  meale, and barrel of oyle diminisheth not. 17.~Her ſonne dieth, and is
  rayſed to life.}

And Elias the Theſbite of the inhabiters of Galaad ſaid to Achab: Our
Lord liueth the God of Iſrael, in whoſe ſight I ſtand, if there ſhal be
theſe
\Fix{yearer}{yeares}{obvious typo, fixed in other}
dew and rayne, but according to the wordes of my mouth. \V And the word
of our Lord came to him, ſaying: \V Depart from hence, and goe agaynſt
the Eaſt, and be hidde in the Torrent carith, which is agaynſt
Iordan, \V and there thou shalt drinke of the torrent: and I haue
commanded the rauens that they feede thee there. \V He therfore went,
and did according to the word of our Lord: and when he was gone, he ſate
in the Torrent carith, which is againſt Iordan. \V The rauens alſo
brought him bread
%%% 0765
and flesh in the morning, in like maner bread and flesh in the euening,
and he dranke of the torrent. \V But after certayne daies the torrent
was dried: for it had not rayned vpon the earth. \V Therfore the word of
our Lord came to him, ſaying: \V Ariſe, and goe into Sareptha of the
Sidonians, and thou shalt tarie there: for I haue commanded a wydow
woman there to feede thee. \V He aroſe,
%%% o-0686
and went into Sareptha. And when he was come to the gate of the citie,
the wydow woman appeared to him gathering ſtickes, and he called her,
and ſayd to her: Geue me a litle water in a veſſel, that I may
drinke. \V And when she went to fetch it, he cried after her ſaying:
Bring me alſo, I beſeche thee, a morſel of bread in thy hand. \V Who
anſwered: Our Lord thy God liueth, I haue no bread, but ſo much meale in
a potte as a hand can hold, and a litle oile in a veſſel: behold I
gather two ſtickes, that I may goe in, and dreſſe it for me and my
ſonne, that we may eate, and die. \V To whom Elias ſayd: Feare not, but
goe, and doe as thou haſt ſayd: but firſt make for me of the ſame meale
a litle harth cake, and bring it to me: and for thy ſelf and thy ſonne
thou shalt make afterward. \V For thus ſayth our Lord the God of Iſrael:
The potte of meale shal not fayle, nor the veſſel of oile be diminished
vntil the day, wherin our Lord wil geue rayne vpon the face of the
earth. \V Who went and did according to the word of Elias: and he did
eate, and she, and her houſe: and from that day \V the pot of meale
fayled not, and the veſſel of oyle was not diminished, according to the
word of our Lord, which he ſpake in the hand of Elias. \V And it came to
paſſe after theſe thinges, the ſonne of the woman, the goodwife of the
houſe, fel ſicke, and the ſickeneſſe was verie vehement ſo that there
remayned no breath in him. \V She therfore ſayd to Elias: What is to me
and thee thou man of God? comeſt thou vnto me, that myne iniquities
might be remembred, and thou mighteſt kil my ſonne? \V And Elias ſayd to
her: Geue me thy ſonne. And he tooke him from her boſome, and caried him
into the vpper chamber where him ſelf abode, and layd him vpon his
bed. \V And he cried to our Lord, and ſayd: O Lord my God,
\SNote{To this queſtion (ſayth S.~Auguſtin) the prophet anſwered in
ſpirite: No. For God killed not this childe to afflict ſo good a mother,
but to confirme her in true religion & comforte her by rayſing him from
death. So Lazarus died not to remaine dead, but to be rayſed to life,
for Gods more glorie.
\XRef{Ioan.~11. v.~4.}
\Cite{S.~Aug. li.~2. q.~5. ad Simplicia.}}
what, the widow alſo with whom I am after a ſort ſuſteyned, haſt thou
afflicted, that thou wouldeſt kil her ſonne? \V And he ſtretched forth,
& meaſured him ſelfe vpon the childe three tymes, & he cried to our
Lord, and ſayd: O Lord my God, let the ſoule of this childe, I beſeche
%%% 0766
thee, returne into his bodie. \V And our Lord heard the voice of Elias:
and the ſoule of the child returned into him, and he reuiued. \V And
Elias tooke the childe, and brought him downe from the vpper chamber
into the lower houſe, and deliuered him to his mother, and ſayd to her:
Behold thy ſonne liueth. \V And the woman ſayd to Elias: Now, in this I
haue knowen that thou art a man of God, and the word of our Lord in
\Fix{the}{thy}{obvious typo, fixed in other}
mouth is true.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
