%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 0745
%%% o-0669
\startChapter[
  title={Chapter 9}
  ]

\Summary{Our Lord appearing againe to Salomon, 4.~admonisheth him and
  his people to keepe the precepts, 6.~threatning punishment if they do
  not. 10.~The king of Tyre receiueth tvventie cities of Salomon, but
  liketh them not. 14.~Salomon buildeth more cities and
  tovvnes. 20.~Maketh diuers nations tributarie. 24.~The Quene repayreth
  to her houſe. 25.~The king offereth victimes thriſe euerie yeare,
  26.~and fetcheth gold from Ophir.}

And it came to paſſe when Salomon had perfected the building of the
houſe, of our Lord, & the kinges houſe, and al that he wiſhed and would
haue done, \V our Lord appeared to him the ſecond time, as he had
appeared to him in Gabaon. \V And our Lord ſaid to him: I haue heard thy
prayer & thy petition, which thou haſt prayed before me: I haue
ſanctified this houſe, which thou haſt built, that I might put my name
there for euer, and myne eies and my hart shal be there alwaies. \V Thou
alſo if thou wilt walke before me, as thy father walked,
\SNote{External worſhip is not acceptable to God, except it procede from
internal ſinceritie and deuotion. VVherfore S.~Auguſtin ſayth: \Emph{God
is worshipped in faith, hope, and charitie.}
\Cite{Enchirid. c.~3.}}
in ſimplicitie of hart, and in equitie: and wilt doe al thinges,
which I haue commanded thee, and wilt keepe my ordinances and my
iudgementes, \V I wil ſette the throne of thy kingdom ouer Iſrael for
euer, as I haue ſpoken to Dauid thy father, ſaying: There ſhal not be
taken away a man of thy ſtocke from the throne of Iſrael. \V But if by
reuolting you
%%% 0746
and your children ſhal turne away, not folowing me, nor keeping my
commandementes, and my ceremonies, which I haue propoſed to you, but
shal goe and worshippe ſtrange goddes, and adore them: \V I wil take
away Iſrael from the face of the land, which I haue geuen them, and the
temple which I haue ſanctified to my name, I wil caſt away from my
ſight, and Iſrael shal be for a prouerbe, and for a fable to al
peoples. \V And this houſe shal be for an example: euerie one that shal
paſſe by it, shal wonder, and hiſſe, and ſay: Why hath the Lord done
thus to this land, and to this houſe? \V And they shal anſwere: Becauſe
they haue forſaken the Lord their God, which brought their fathers out
of the Land of Ægypt, and haue folowed ſtrange goddes, & adored them,
and worshipped them: therfore hath the Lord brought vpon them al this
euil. \V And twentie yeares being complete, after that Salomon had built
the two houſes, that is, the houſe of our Lord, and the houſe of the
king, \V (Hiram the king of Tyre miniſtring to Salomon ceder trees &
firre trees, and gold according to al that he had neede) then Salomon
\SNote{Salomon did not ſel theſe cities, for he could not alienate them,
but let the king of Tyre haue the vſe and reuenewes in payment for
timber, & for the gold which he ſent.}
gaue to Hiram twentie townes in the Land of Galilee. \V And Hiram went
from Tyre, to ſee the townes which Salomon had geuen him, and they
pleaſed him not, \V and he ſayd: Are theſe the cities, which thou haſt
geuen me, brother? And he called them
%%% o-0670
the land
\TNote{dirtie, or diſpleaſing.}
Chabul, vntil this day. \V Hiram alſo ſent to king Salomon an hundred
and twentie talentes of gold. \V This is the ſumme of the expences,
which king Salomon offered to build the houſe of our Lord, and his owne
houſe, and Mello, and the wal of Ieruſalem, and Heſer, and Mageddo, and
Gazer. \V Pharao the king of Ægypt came vp and tooke Gazer, and burnt it
with fire: and the Chananite, that dwelt in the citie, he ſlewe, and
gaue it for a dowrie to his daughter the wife of Salomon. \V Salomon
therfore built Gazer, and Bethhoron the lower, \V and Baalath, and
Palmira in the Land of the wildernes. \V And al the villages, that
perteyned to him, and were without wal, he fenſed, and the cities of the
chariotes, and the cities of the horſemen, and whatſoeuer pleaſed him to
build in Ieruſalem, and in Libanus, and in al the land of his
dominion. \V Al the people, that was remayning of the Amorrheites, and
Hetheites, and Pherezeites, and Heueites, and Iebuſeites, that are not
of the children of Iſrael: \V their children, that were remayning in the
land, to witte, thoſe whom
%%% 0747
the children of Iſrael could not abolish: Salomon made tributaries,
vntil this day. \V But of the children of Iſrael Salomon appoynted not
any man to ſerue, but they were men of warre, and his ſeruantes, and
princes, and captaynes, and ouerſeers of the chariotes and horſes. \V
And there were princes ouer al the workes of Salomon, made ouerſeers,
fiue hundred fiftie, which had the people ſubiect, and commanded ouer
their appoynted workes. \V And the daughter of Pharao went vp out of the
citie of Dauid into her houſe, which Salomon had built her: then did he
build Mello. \V Salomon alſo offered three tymes euerie yeare
holocauſtes, and pacifique victimes vpon the altar, which he had built
to our Lord, and he burnt incenſe before our Lord: and the temple was
perfected. \V King Salomon alſo made
\SNote{A monument.}
a name in Aſiongaber, which is beſide Ailath in the ſhore of the Readſea
in the Land of Idumea. \V And Hiram ſent in that nauie his men, that
were mariners & skilful of the ſea, with the ſeruantes of Salomon. \V
Who when they were come into Ophir, the gold taken thence of foure
hundred and twentie talentes, they brought to king Salomon.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
