%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 0766
%%% o-0687
\startChapter[
  title={Chapter 18}
  ]

\Summary{The third yeare of famine, Elias meeting the chief gouernour of
  Achabs houſe, hardly perſwadeth him to tel Achab, that he is
  preſent. 17.~Achab blameth Elias; but Elias freely auoucheth that not
  he, but Achab trubleth Iſrael. 19.~By a miracle foure hundred and
  fiftie falſprophetes are conuinced, 40.~and are ſlaine. 41.~Elias
  prayeth and it rayneth.}

After manie daies the word of our Lord came to Elias, the third yeare,
ſaying: Goe, and ſhew thy ſelf to Achab, that I may geue rayne vpon the
face of the earth. \V Elias therfore went to ſhew himſelf to Achab: and
there was ſore famine in Samaria. \V And Achab called Abdias the
gouernour of his houſe: and Abdias did feare our Lord very much. \V For
when Iezabel killed the prophetes of our Lord, he tooke an hundred
prophetes, and hid them by fiftie and fiftie in caues, and fed them with
bread and water. \V Achab therfore ſayd to Abdias: Goe into the land to
al the fountaynes of waters, and into al valleys, if perhaps we may
finde graſſe, and ſaue the horſes and mules, and the beaſtes may not
vtterly perish. \V And they diuided the countries betwen them, that they
might goe circuite about them: Achab went one way, and Abdias an other
way ſeuerally. \V And when Abdias was in the way, Elias mette him: who when
he knew him,
\SNote{Abdias adored Elias as the prophete of God, and a holie man not
with ciuil honour, for in the world Abdias was the greater perſon, nor
with diuine honour, for that had benne idolatrie. It was therfore
religious honour, called \L{dulia}, due to ſpiritual excellencie of Gods
ſeruantes.}
fel on his face, and ſayd: My lord, art not thou Elias? \V To whom he
anſwered: I am. Goe, and tel thy lord: Elias is here. \V And he ſayd:
What haue I ſinned, that thou deliuereſt me thy ſeruant into the hand of
Achab, that he may kil me? \V Our Lord thy God liueth, there is no
nation or kingdom, whither my lord hath not ſent to ſeeke thee: and al
anſwering: He is not here: he adiured al kingdomes and nations, for that
thou waſt not found. \V And now thou ſayeſt to me: Goe, and tel thy lord:
Elias is here. \V And when I am departed from thee, the Spirit of our
Lord wil carie thee into a place, that I know
%%% 0767
not: and I entering in shal tel Achab, and not finding thee, he wil kil
me: and thy ſeruant feareth our Lord from his infancie. \V Hath it not
beene told thee my lord, what I did when Iezabel killed the prophetes of
our Lord, that I hid of the prophetes of our Lord an hundred men, by
fiftie and fiftie in caues, and fed them with bread and water? \V And
now thou ſayeſt: Goe, and tel thy lord: Elias is here: that he may kil
me? \V And Elias ſayd: The Lord of hoſtes liueth, before whoſe face I
ſtand, this day wil I appeare to him. \V Abdias therfore went to meete
Achab, and told him: and Achab came to meete Elias. \V And when he had
ſeene him, he ſayd: Art thou he that doeſt truble Iſrael? \V And he
ſayd: Not I haue trubled Iſrael, but thou, and the houſe of thy father,
%%% o-0688
who haue forſaken the commandmentes of our Lord, and haue folowed
Baalim. \V Neuertheleſſe ſend now, and gather vnto me al Iſrael in the
mount of Carmel, and the prophetes of Baal foure hundred fiftie, and the
prophetes of the groues, foure hundred, which eate of Iezabels table. \V
Achab ſent to al the children of Iſrael, and gathered together the
prophetes in the mount of Carmel. \V And Elias coming to al the people,
ſayd:
\SNote{Such zelous expoſtulation is neceſſarie to al Neutrals in
religion who are \Emph{neither hott nor cold, but lukevvarme}, ſuch as
Angels deteſt.
\XRef{Apoc.~3.}}
How long halt you on two ſides? If our Lord be God, folow him: but if
Baal, folow him. And the people did not anſwer him a word. \V And Elias
ſayd agayne to the people: I onlie remayne a prophete of our Lord: and
the prophetes of Baal are foure hundred and fiftie men. \V Let two oxen
be geuen vs, and let them chooſe to them ſelues one oxe, and cutting it
into peeces let them lay it vpon wood, but put no fyre vnder: and I wil
dreſſe the other oxe, and wil lay it vpon wood, and put no fire
vnder. \V Inuocate ye the names of your goddes, & I wil inuocate the
name of my Lord: and the God that shal heare by fyre, let the ſame be
God. And al the people anſwering, ſayd: A very good propoſition. \V Elias
therfore ſayd to the prophetes of Baal: Chooſe you one oxe, and make it
firſt, becauſe you are manie: & inuocate the names of your goddes, and
put no fire vnder. \V Who when they had taken the oxe, which he gaue
them, they dreſſed it: and they inuocated the name of Baal from morning
vntil midday, ſaying: Baal heare vs. And there was no voice, nor any
that anſwered: and they leaped ouer the altar, that they had made. \V
And when it was now midday, Elias ieſted at them, ſaying: Crie with a
lowder voice: for he is God, and perhaps he ſpeaketh,
%%% 0768
or is in his inne, or in the way, or at the leaſt he ſlepeth, that he
muſt be waked. \V They cried therfore with a lowd voice, and cut them
ſelues after their rite with kniues and lancers, til they were al
embrewed with bloud. \V And after the midday was paſt, and whiles they
prophecied, the time was come, when they vſed to offer ſacrifice,
neither voice was heard, nor any did anſwer, nor attend them praying: \V
Elias ſayd to al the people: Come vnto me. And the people coming to him,
he repayred the altar of our Lord, that was deſtroyed. \V And he tooke
twelue ſtones according to the number of the tribes of the children of Iacob, to whom
the word of our Lord came, ſaying: Iſrael ſhal be thy name. \V And he
built of the ſtones an altar in the name of our Lord: and he made a water
gutter, as it were by two furrowes round about the altar, \V and he layed
the wood in order, and diuided the oxe in iointes, and layd it vpon the
wood, \V and ſaid: Fil foure buckettes with water, and powre vpon the
holocauſte, and vpon the wood. And againe he ſayd: Doe it alſo the
ſecond time. Who hauing donne it the ſecond time, he ſaid: The third
time alſo doe the ſame. And they did ſo the third time, \V and the
waters ranne about the altar, and the trough of the conduite was
filled. \V And when it was now time that the holocauſte ſhould be
offered, Elias the prophete coming ſaid: Lord God of Abraham,
%%% o-0689
and Iſaac, & Iſrael, ſhew this day that thou art the God of Iſrael, and
I thy ſeruant, & that according to thy commandment I haue done al theſe
thinges. \V Heare me Lord, heare me:
\SNote{VVhen miracles are attempted for trial of the truth, the diuels
powre is reſtrayned, & only the truth is teſtified: \Emph{Our Lord
vvorking vvithal, & confirming the doctrin vvith ſignes folovving}.
\XRef{Marc. vlt.}}
that this people may learne, that thou art our Lord God, & thou haſt
conuerted their hart againe. \V And the fire of our Lord fel, & deuoured
the holocauſte, & the wood, and the ſtones, licking alſo the duſt, and
the water, that was in the water gutter. \V Which when al the people had
ſeene, they fel on their face, and ſaid: Our Lord he is God, our Lord he
is God. \V And Elias ſaid to them: Apprehend the prophetes of Baal, and
let not one eſcape of them. Whom when they had taken, Elias brought them
to the Torrent ciſon, and killed them there. \V And Elias ſayd to Achab:
Goe vp, eate, and drinke: becauſe there is ſound of much raine. \V
Achab went vp to eate and drinke: and Elias went vp into the toppe of
Carmel, and flatte on the earth put his face betwen his knees, \V and he
ſaid to his ſeruant: Goe vp, and looke toward the ſea. Who when he was
gone vp, and had looked, he ſaid: There
%%% 0769
is nothing. And againe he ſaid to him: Returne ſeuen tymes. \V And in
the ſeuenth time: Behold a litle cloud as it were a mans foote, came vp
from the ſea. Who ſaid: Goe vp and ſay to Achab: Yoke thy chariote and
goe downe, leſt the raine preuent thee. \V And when he turned him ſelfe
hither and thither, behold the heauens were darkened, and cloudes, and
winde, and there fel great raine. Achab therfore going vp went into
Iezrahel: \V and the hand of our Lord was made vpon Elias, and his
loynes girded he ranne before Achab, vntil he came into Iezrahel.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
