%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 0752
%%% o-0675
\startChapter[
  title={Chapter 12}
  ]

\Summary{Roboam
\MNote{The third part.

The diuiſion of the Kingdom. Seueral reignes of certaine kinges: and
preaching of ſpecial prophetes.}
folowing youngmens counſel, 16.~Ieroboam poſseſſeth ten tribes of his
Kingdom. 21.~VVhich he endouoring to recouer by warre, is admonished by a
prophet to ceaſe. 26.~Hieroboam ſetteth vp golden calues to be adored,
making temples, altares, and prieſtes fitte for his purpoſe.}

And Roboam came into Sichem: for thither was al Iſrael gathered together
to make him king. \V But Ieroboam the ſonne of Nabat, when he was yet in
Ægypt fugitiue from the face of king Salomon, hearing of his death,
returned out of Ægypt. \V And they ſent and called him: Ieroboam
therfore came, and al the multitude of Iſrael, and they ſpake to Roboam,
ſaying: \V Thy father layd a moſt hard
%%% 0753
yoke vpon vs: thou therfore diminish now a litle of thy fathers moſt
hard empire, and of the moſt heauie yoke, that he layd vpon vs, and we
wil ſerue thee. \V Who ſayd to them: Goe vntil the third day, and
returne to me. And when the people was gone, \V king Roboam tooke
counſel with the ancientes, that aſſiſted before Salomon his father,
whiles he yet liued, and he ſayd: What counſel doe you geue me, that I
may anſwer this people? \V Who ſayd to him: If this day thou wilt yeld
to this people, and condeſcend to them,
%%% o-0676
and graunt to their petition, and wilt ſpeake to them gentle wordes,
they wil be thy ſeruantes alwaies. \V Who leaft the counſel of the
ancientes, which they had geuen him, and admitted yongmen, that had bene
brought vp with him, and wayted on him, \V and he ſayd to them: What
counſel geue you me, that I may anſwer this people, which haue ſayd to
me: Make the yoke lighter which thy father hath put vpon vs? \V And the
yongmen, that had bene brought vp with him, ſayd: Thus ſpeake to this
people, which haue ſpoken to thee, ſaying: Thy father aggrauated our
yoke, doe thou eaſe it. Thus ſhalt thou ſpeake to them: My leaſt finger
is groſſer then the backe of my father. \V And now my father layd vpon
you a heauie yoke, but I wil adde vpon your yoke: my father
\Fix{bette}{beate}{possible typo, fixed in other}
you with ſcourges, but I wil beate you with ſcorpions. \V Ieroboam
therfore came, and al the people to Roboam the third day, as the king
had ſpoken, ſaying: Returne to me the third day. \V And the king
anſwered the people rough wordes, leauing the counſel of the ancientes,
which they had geuen him, \V and he ſpake to them according to the
counſel of the youngmen, ſaying: My father made your yoke heauie, but I
wil adde to your yoke: my father
\Fix{bette}{beate}{possible typo, fixed in other}
you with whippes, but I wil beate you with ſcorpions. \V And the king
condeſcended not to the people: becauſe our Lord was turned away from
him,
\SNote{This phraſe noteth the ſequel, not the final cauſe. As
\XRef{chap.~14. v.~9.}}
that he might rayſe vp his word, which he had ſpoken in the hand of
Ahias the Silonite, to Ieroboam the ſonne of Nabat. \V The people
therfore ſeing that the king would not heare them, anſwered him, ſaying:
What part haue we in Dauid? or what inheritance in the ſonne of Iſai?
Goe into thy tabernacles Iſrael, now ſee to thy houſe Dauid. And Iſrael
went into their tabernacles. \V But ouer the children of Iſrael,
whoſoeuer dwelt in the cities of Iuda, Roboam reigned. \V King Roboam
%%% 0754
therfore ſent Aduram, who was ouer the tributes: and al Iſrael ſtoned
him, and he died, moreouer King Roboam in haſt went vp into his
chariote, and fled into Ieruſalem: \V and Iſrael reuolted from the houſe
of Dauid, vntil this preſent day. \V And it came to paſſe when al Iſrael
had heard, that Ieroboam was returned, they ſent, and called him, an
aſſemblie being gathered, and they made him king ouer al Iſrael, neither
did any man folow the houſe of Dauid beſide the tribe of Iuda onlie. \V
And Roboam came to Ieruſalem, and gathered together al the houſe of
Iuda, and the tribe of Beniamin, an hundred foureſcore thouſand choſen
men warriers, to fight agaynſt the houſe of Iſrael, and to reduce the
kingdom to Roboam the ſonne of Salomon. \V But the word of our Lord came
to Semeias the man of God, ſaying: \V Speake to Roboam the ſonne of
Salomon, the king of Iuda, and to al the houſe of Iuda, and Beniamin,
and the reſt of the people, ſaying: \V Thus ſayth our Lord: You ſhal not
goe vp, neither ſhal you fight agaynſt your brethren the children of
Iſrael: let euerie man returne into his houſe, for this word is done by
me. They heard the word of our Lord, and returned from their iourney as
our Lord had commanded them. \V And
%%% o-0677
Ieroboam built Sichem in mount Ephraim, and dwelt there: and departing
thence he built Phanuel. \V And Ieroboam ſayd in his hart: Now wil the
kingdom returne to the houſe of Dauid, \V if this people shal goe vp to
make ſacrifices in the houſe of our Lord into Ieruſalem: and the hart of
this people wil be turned to their lord Roboam the king of Iuda, and
they wil kil me, and returne to him. \V And finding out
\SNote{A diueliſh policie to make a religion conformable to the temporal
ſtate.}
a deuiſe he made two golden calues, and ſayd to them: Goe vp no more
into Ieruſalem: Behold thy goddes Iſrael, which brought thee out of the
Land of Ægypt. \V And he put one in Bethel, and the other in Dan: \V and
this thing was an occaſion of ſinne: for the people went to adore the
calfe, as farre as Dan. \V And he made temples in the excelſes, and
prieſtes
\SNote{For ſuch a religion ſuch prieſtes were fitteſt.}
of the abiectes of the people, which were not of the children of
Leui. \V And he appoynted a ſolemne day in the eight moneth, the
fiftenth day of the moneth, after the ſimilitude of the ſolemnitie, that
was celebrated in Iuda. And going vp he made in like maner an altar in
Bethel, to immolate to the calues, which he had framed: and he ordayned
in Bethel prieſtes of the
\SNote{Places on hilles, where they ſacrificed calues, and other thinges
to the images of calues.}
excelſes, which he had made. \V And he
%%% 0755
went vpon the altar, which he had built in Bethel, the fiftenth day of
the eight moneth, which he had forged out of his owne hart: and he made
a ſolemnitie to the children of Iſrael, and went vp vpon the altar, to
burne incenſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
