%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 0774
%%% o-0694
\startChapter[
  title={Chapter 21}
  ]

\Summary{Naboth for denying his vineyard to King Achab, is by quene
  Iezabels commandment, falſly accuſed, and ſtoned to death. 10.~Achab
  haſtening to poſſeſse the vineyard, Elias the prophet threatneth him
  manie euiles, 23.~and no leſse to Iezabel: both being obſtinate in
  ſinne. 27.~Yet Achab for feare of punishment, doth external workes of
  penance, and therby eſcapeth part of the temporal plagues.}

And after theſe wordes, at that tyme Naboth the Iezrahelite had a
vineyard, which was in Iezrahel, beſide the palace of Achab the king of
Samaria. \V Achab therfore ſpake to Naboth, ſaying: Geue me thy
vineyard, that I may make me a garden of herbes, becauſe it is nigh, and
ioyning to my houſe, and
\SNote{If no ſubiect were lord of anie land, but only at the kings
pleaſure & kinges were proper lordes of al the landes in their
kingdomes, then Achab might haue taken Naboths vinyard; eſpecially
geuing him a better, or money for it. Neither was it a vaine ſcruple in
Naboth, to conſerue his anceſters inheritance. For both his denial is
here iuſtified, and Achabs extortion condemned. S.~Ambroſe
\Cite{li.~3. offic. c.~9.}
counteth Naboth a Martyr. See
\XRef{Annot. 1.~Reg.~8.}}
I wil geue thee for it a better vineyard: or if thou thinke it more
commodious for thee, the price of ſiluer, ſo much as it is worth. \V To
whom Naboth anſwered: Our Lord be merciful to me, that I geue not the
inheritance of my fathers to thee. \V Achab therfore came into his houſe
with indignation, and fretting vpon the word, that Naboth the
Iezrahelite had ſpoken to him, ſaying: I wil not geue thee the
inheritance of my fathers. And caſting him ſelf vpon his bed, he turned
away his face to the wal, and did not eate bread. \V And Iezabel his
wife went in vnto him, and ſayd to him: What is this matter, whereupon
thy ſoule is greued? and why eateſt thou not bread? \V Who anſwered her:
I ſpake to Naboth the Iezrahelite, and ſayd to him: Geue me thy
vineyard, taking money for it: or if it pleaſe thee, I wil geue thee a
better vineyard for it. And he ſayd: I wil not geue thee my vineyard. \V
Iezabel therfore his wife ſayd to him: Thou art of great authoritie, and
doeſt wel gouerne the kingdome of Iſrael: Ariſe, and eate bread, and be
of good cheere, I wil geue thee the vineyard of Naboth the Iezrahelite.
%%% 0775
\V She therfore wrote letters in the name of Achab, and ſigned them with
his ring, and ſent to the ancientes, and the chiefe men that were in his
citie, and dwelt with Naboth. \V And this was the tenor of the letters:
Proclame a faſt, and make Naboth ſitte among the chiefe of the
people, \V and ſuborne two men the children of Belial agaynſt him, and
let them beare falſe teſtimony: that he hath
\SNote{To auoide horrors of blaſphemie holie ſcripture often vſeth the
terme \Emph{bleſsing} for \Emph{curſing}.}
bleſſed God and the king: and bring him forth, and ſtone him, and ſo let
him die. \V His citizens therfore the ancientes and chiefe men, that
dwelt with him in the citie, did as Iezabel had commanded them, and as
it was written in the letters which she ſent to them: \V they proclamed
a faſt, and made Naboth ſitte among the chiefe of the people. \V And two
men the children of the diuel being brought forth, they
%%% o-0695
made them ſitte agaynſt him: but they, as deuelish men, gaue teſtimonie
agaynſt him before the multitude: Naboth hath bleſſed God and the king:
for the which thing they brought him forth without the citie, and killed
him with ſtones. \V And they ſent to Iezabel, ſaying: Naboth is ſtoned,
and is dead. \V And it came to paſſe, when Iezabel had heard that Naboth
was ſtoned, and dead, she ſpake to Achab: Ariſe, and poſſeſſe the
vineyard of Naboth the Iezrahelite, who would not agree vnto thee, and
geue it taking money: for Naboth liueth not, but is dead. \V Which when
Achab had heard, to witte, that Naboth was dead, he aroſe, & went downe
into the vineyard of Naboth the Iezrahelite, to poſſeſſe it. \V The word
of our Lord therfore came to Elias the Theſbite, ſaying: \V Ariſe, and
goe downe to meete Achab the king of Iſrael, who is in Samaria: behold
he goeth downe to the vineyard of Naboth, to poſſeſſe it: \V and thou
shalt ſpeake to him, ſaying: Thus ſayth our Lord: Thou haſt ſlayne,
moreouer alſo thou haſt poſſeſt. And after theſe wordes thou shalt adde:
Thus ſayth our Lord: In this place, wherein the dogges haue licked the
bloud of Naboth, they
\CNote{\XRef{chap.~22. v.~38.}}
shal licke thy bloud alſo. \V And Achab ſayd to Elias: Haſt thou found
me thyne enemie? Who ſayd: I haue found thee, for that thou art ſold, to
doe euil in the ſight of our Lord. \V Behold I wil bring euil vpon thee,
and wil cut downe thy poſteritie, and wil kil of Achab him that piſſeth
agaynſt the wal, and the incloſed, and the laſt in Iſrael. \V And I wil
make thy houſe, as the houſe of Ieroboam the ſonne of Nabat, and as the
houſe of Baaſa the ſonne of
%%% 0776
Ahias: becauſe thou haſt done, to prouoke me to anger, and haſt made
Iſrael to ſinne. \V But of Iezabel alſo our Lord ſpake, ſaying: The
dogges shal eate Iezabel in the field of Iezrahel. \V If Achab die in
the citie, the dogges shal eate him: but if he die in the field, the
foules of the ayre shal eate him. \V Therfore there was not ſuch an
other as Achab, who
\SNote{So addicted to wickednes, as if he had ſolde him ſelfe to the
diuel for ſome temporal profite or pleaſure. \Emph{In your iniquities
you are ſolde.}
\XRef{Iſaie.~50.}
\Cite{S.~Aug. q.~102. ex vtro. teſt.}
\Cite{S.~Greg. ho.~10. in Ezech.}}
was ſold to doe euil in the ſight of our Lord: for his wife Iezabel
ſette him on, \V and he became abominable, in ſo much that he folowed
the idols, which the Amorrheites had made, whom our Lord conſumed before
the face of the children of Iſrael. \V Therfore when Achab had heard
theſe wordes, he rent his garmentes, and couered his flesh with
haercloth, and faſted and ſlept in ſackcloth, and walked caſting downe
his head. \V And the word of our Lord came to Elias the Theſbite,
ſaying: \V Haſt thou not ſeene Achab humbled before me? therfore, becauſe
he hath humbled him ſelf for my ſake, I wil not bring in the euil in his
dayes, but in his ſonnes daies wil I bring the euil vpon his houſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
