%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 0769
%%% o-0689
\startChapter[
  title={Chapter 19}
  ]

\Summary{Elias fleing Iezabel, in the deſert eateth a harth cake and
  drinketh water, brought by an Angel, and ſo goeth fourtie dayes and
  nightes vnto mount Horeb. 9.~Lamenting that he alone is leift of the
  prophetes of God. 15.~God commandeth him to returne to Damaſcus, and
  anoint Hazael king of Syria, Iehu king of Iſrael, and Eliſeus a
  prophet: 18.~and telleth him, there remaine ſeuen thouſand in Iſrael,
  which haue not bowed to Baal.}

And Achab told Iezabel al thinges that Elias had done, and how he had
killed al the prophetes with the ſword. \V And Iezabel ſent a meſſenger
to Elias, ſaying: Theſe thinges do the goddes to me, and theſe adde
they, if this houre to morow I make not thy ſoule as the ſoule of one
of them. \V Elias therfore was afraid, & ryſing he went whither ſoeuer
his wil caried him: and he came into Berſabee of Iuda, and leaft his
ſeruant there, \V and went forward into the deſert, one dayes
iourney. And when he was come, and ſate vnder a iuniper tree, he deſired
%%% o-0690
for his ſoule to die, and ſayd: It ſufficeth me Lord, take my ſoul: for
I am not better then my fathers. \V And he caſt him ſelf downe, and
ſlept in the ſhadow of the iuniper tree: and behold an Angel of our Lord
touched him, and ſayd to him: Ariſe, and eate. \V He looked, and behold
at his head harthbaked bread, and a veſſel of water: he therfore did
eate, and drinke, and he ſlept againe. \V And the Angel of our Lord
returned the ſecond time, and touched him, and ſaid to him: Ariſe, eate:
for thou haſt yet a great way to goe. \V Who when he was riſen: did eate
and drinke, & walked in the ſtrength of that meate fourtie daies, and
fourtie nightes, vnto the mount of God, Horeb. \V And when he was come
thither, he taried in a caue: and behold the word of our Lord
\Fix{vnto}{came vnto}{likely typo, same in both}
him, & he ſaid to him: What doeſt thou here Elias? \V But he anſwered:
With zele haue I beene zelous
%%% 0770
for our Lord the God of hoſtes becauſe the children of Iſrael haue
forſaken thy couenant: thy altars haue they deſtroyed, & thy prophetes
they haue ſlaine with the ſword, &
\SNote{Elias knew no other at that inſtant in the tenne tribes, that
were not partakers in idolatrie; but there were in deede ſeuen thouſand
\XRef{(v.~18.)}
that bowed not their knees to Baal. And the whole kingdom of Iuda
obſerued true religion, where Aſa then reigned, and after him Ioſaphat:
both good kinges,
\XRef{chap.~15. v.~11.}
\XRef{22.~v.~43.}
\XRef{2.~Par.~15.}
&.
\XRef{17.}}
I alone am leaft, & they ſeeke my life to take it away. \V And he ſaid
to him: Come forth, and ſtand in the mount before our Lord: and behold
our Lord paſſeth, and a great winde, and ſtrong ouerthrowing mountaynes,
and breaking rockes before our Lord: not in the winde is our Lord, and
after the winde an earth quake: not in the earth quake is our Lord, \V
and after the earth quake fire: not in the fire is our Lord, and after
the fire a wiſtling of a gentle winde. \V Which when Elias had heard, he
couered his face with his mantel, and coming forth ſtood in the doore of
the caue, and behold a voyce vnto him, ſaying: What doeſt thou here
Elias? \V And he anſwered: With zele haue I beene zelous for our Lord
the God of hoſtes: becauſe the children of Iſrael haue forſaken thy
couenant: thyn altars they haue deſtroyed, and thy prophetes they haue
ſlaine with the ſword, and I alone am leaft, and they ſeeke my life to
take it away. \V And our Lord ſayd to him: Goe, and returne into thy way
by the deſert of Damaſcus: and when thou art come thither thou
\SNote{Elias prophecied that theſe two ſhould be kinges, and caſt his
cloke ouer Elizeus.
\XRef{v.~19.}
Elizeus declared to Hazael that he ſhould be king.
\XRef{4.~Reg.~8.}
An other prophet annointed Iehu.
\XRef{4.~Reg.~9.}}
ſhalt anoint Hazael king ouer Syria, \V and Iehu the ſonne of Namſi thou
ſhalt anoynt king ouer Iſrael: and Eliſeus the ſonne of Saphat, which is
of Abelmeula, thou shalt anoynt prophet for thee. \V And it shal be,
whoſoeuer shal eſcape the ſword of Hazael, him Iehu shal kil: and
whoſoeuer shal eſcape the ſword of Iehu, him shal Eliſeus kil. \V And I
wil leaue me in Iſrael ſeuen thouſand men, whoſe knees haue not beene
bowed before Baal, and euerie mouth, that hath not adored him kyſſing
his handes. \V Elias therfore departing thence, found Eliſeus the ſonne
of Saphat, ploughing with twelue yoke of oxen, and he was one of them
that ploughed with twelue yoke of oxen: and when Elias came to him, he
caſt his mantel vpon him. \V Who forthwith leauing the oxen ranne after
Elias, and ſaid: Let me kiſſe, I pray thee, my father, and my mother,
and ſo I wil folow thee. And he ſayd to him: Goe, and returne: for that
which was my part, I haue donne to thee. \V And returning from him, he
tooke a
%%% o-0691
yoke of oxen, and killed them, and ſod the flesh with the plough of the
oxen, and gaue to the people, and they did eate: and ryſing
vp he departed, and folowed Elias, & miniſtred to him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
