%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 0716
%%% o-0643
\startChapter[
  title={Chapter 4}
  ]

\Summary{Chiefe men of Salomons kingdom are recited by their names, and
  offices. 22.~Likewiſe the prouiſion of victuals for his houſe, 26.~the
  number of his horſes. 29.~His wiſedom excelleth al others. 32.~He
  writte manie parables and verſes, and lernedly diſcourſed of al
  thinges.}

And king Salomon was reigning ouer al Iſrael: \V and theſe were the
princes which he had: Azarias the ſonne of Sadoc the prieſt: \V
Elihoreph, and Ahia the ſonnes of Siſa Scribes: Ioſaphat the ſonne of
Ahilud, regiſter: \V Banaias the ſonne of Ioiada, ouer the armie: and
Sadoc, and Abiathar prieſtes. \V Azarias the ſonne of Nathan, ouer them
that aſſiſted the king: Zabud the ſonne of Nathan prieſt the kinges
frend: \V and Ahizar gouernour of the houſe: and Adoniram the ſonne of
Abda ouer the tributes. \V And Salomon had twelue gouernours ouer al
Iſrael, which ſerued out victuals for the king and for his houſe: for euerie
one miniſtred neceſſaries, eche man his moneth in the yeare. \V And
theſe are their names: Benhur in mount Ephraim. \V Bendecar, in Macces,
and in Salebim, and in Bethſames, and in Elon, and in Bethhanan. \V
Benheſed in Aruboth: his was Socho, and al the land Epher. \V
Benabinadab, whoſe was al Nepha Dor, had Tapheth the daughter of Salomon
to wife. \V Bana the ſonne of Ahilud gouerned Thanach and Mageddo, and
al Bethſan, which is beſide Sarthana vnder Iezrahel, from Bethſan vnto
Abelmehula ouer againſt Iecmaan. \V Bengaber in Ramoth galaad: had
Auothiair the ſonne of Manaſſes in Galaad, he was chiefe in al the
countrie of Argob, which is in Baſan, three ſcore cities great and
walled, which had braſen lockes. \V Ahinadab the ſonne of Addo was
chiefe in Manaim. \V Achimaas in Nephthali: yea he alſo had Baſemath the
daughter of Salomon in mariage. \V Baana the ſonne of Huſi, in Aſer, and
in Baloth. \V Ioſaphat the ſonne of Pharue, in Iſacar. \V Semei the
ſonne of Ela, in Beniamin. \V Gaber the ſonne of Vri, in the land of
Galaad in the land of Sehon the king of the Ammorrheite, & of Og the
king of Baſan, ouer al thinges that were in that land. \V Iuda and
Iſrael innumerable, as the ſand of the ſea in multitude: eating, and
drinking, and reioyſing. \V And Salomon was in his dominion, hauing al
the kingdomes with him from the riuer of the land of the Philiſthijms
vnto the border of Ægypt: of them that offered him
%%% 0717
preſents, and ſerued him al the dayes of his life. \V And the prouiſion
of Salomon was euerie day thirtie meaſures of floure, & three ſcore
meaſures of meale, \V tenne fat oxen and twentie paſture fed, & a
hundred rammes, beſide the veniſon of hartes, roes, and buſtles, &
fatted foule. \V For his poſſeſſed
%%% o-0644
al the countrie, which was beyond the riuer, from Thapſa vnto Goza, and
al the kinges of thoſe countries: and he had peace on euerie ſide round
about. \V And Iuda and Iſrael dwelt without anie feare, euerie one vnder
his vine, and vnder his figtree, from Dan vnto Berſabe al the dayes of
Salomon. \V And Salomon had fourtie thouſand ſtalles of chariot-horſes,
and twelue thouſand for the ſadle. \V And the foreſaide gouernours of
the king fed them: yea and the neceſſaries of king Salomons table they
gaue forth with great care in their time. \V Barley alſo and ſtrow for
the horſes, and beaſtes, they brought to the place, where the king was,
according as it was appointed them. \V God alſo gaue wiſdom to Salomon
and prudence exceding much, and latitude of hart as the ſand that is in
the ſea ſhore. \V And the wiſdom of Salomon paſſed the wiſdom of al them
in the eaſt, and of the Ægyptians, \V and he was wiſer then al men:
wiſer then Ethan the Ezralite, and Heman, and Chalcol, and Dorda the
ſonnes of Mahol and he was renowmed in al nations round about. \V
Salomon alſo ſpake
\SNote{Theſe bookes are not extant.}
three thouſand parables: and his ſongues were a thouſand & fiue. \V And
he diſputed of trees from the ceder, that is in Libanus, vnto the hyſſop
which cometh out of the wal: and he diſcourſed of beaſtes, and foules,
and creeping wormes, and fiſhes. \V And there came from al people to
heare the wiſdom of Salomon, and from al the kinges of the earth, which
heard his wiſdom.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
