%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 0737
%%% o-0662
\startChapter[
  title={Chapter 7}
  ]

\Summary{Salomons
\MNote{The beginning of the fifth age.}
palace, 2.~his houſe in the foreſt, 8.~and the quenes houſe is
built. 13.~Two great braſen pillers: 23.~a ſea (or lauer) 27.~tenne
braſen feete, 38.~tenne leſse lauatories, and other veſsels, and
implementes pertaining to the Temple, adorned vvith images of Angels, and
other creatures are further deſcribed.}

And his owne houſe Salomon built in thirtene  yeares, and
\Fix{brough}{brought}{obvious typo, fixed in other}
it to perfection. \V He built alſo the houſe of the foreſt of Libanus of
an hundred cubites in length, and fiftie cubites in bredth, and thirtie
cubites in height: and foure ſcore galleries betwen pillers of ceder:
for he had cut ceder trees into pillers. \V And he decked the whole vaut
with bordes of ceder, which was held vp with fiue and fourtie
pillers. And one order had fiften pillers, \V ſet one againſt an
other, \V and looked one ouer againſt an other, with equal ſpace betwen
the pillers, and ouer the pillers ſquare beames in al equal. \V And the
porche of the pillers he made of fiftie cubites in length, and thirtie
cubites in bredth: and an other porche before the greater porche: and
pillers, and toppes vpon the pillers. \V He made alſo the porche of the
throne, wherein the ſeat of iudgement is; and couered it with ceder wood
from the pauement vnto the toppe. \V And the litle houſe, where they
ſate in iudgement, was in the middes of the porche of like worke. He
made alſo a houſe for the daughter of Pharao (which Salomon had taken to
wife) of ſuch worke, as alſo this porche. \V Al of choſen ſtones, which
were ſawed by a certain rule & meaſure both within & without: from the
%%% o-0663
fundation to the toppe of the walles, & without vnto the greater
courte. \V And the fundation of choſen ſtones, great ſtones of ten or
eight cubites. \V And aboue there were hewed choſen ſtones of equal
meaſure, and in like maner of ceder. \V And the greater court round with
three rewes of hewed ſtones,
%%% 0738
and one rew of planed ceder, moreouer alſo in the inner court of the
houſe of our Lord, and in the porche of the houſe. \V King Salomon alſo
ſent, and tooke Hiram from Tyre, \V the ſonne of a widow woman of the
tribe of Nepthali, his father a Tyrian, an artificer in braſſe, and ful
of wiſdom, and intelligence, and skil to make al worke of braſſe. Who
when he was come to king Salomon, made al his worke. \V And he caſt two
braſen pillers, of eightene cubites in height one piller: and a line of
twelue cubites compaſſed both pillers. \V He made alſo two litle heades,
which should be put vpon the heades of the pillers, caſt of braſſe: fiue
cubites high one litle head, and fiue cubites the other litle head: \V
and as it were in maner of a nette, and of cheynes knitte one to the
other with maruelous worke. Both litle heades of the pillers were caſt:
ſeuen rewes of litle nettes in one litle head, & ſeuen litle nettes in
the other litle head. \V And finiſhed the pillers, and two rewes round
about euerie nette, that they might couer the litle heades, which were
ouer the toppe of the pomegranates: in like maner did he alſo to the
ſecond litle head. \V And the litle heades, that were vpon the heades of
the pillers, were made as it were with lilie worke, in the porche, of
foure cubites. \V And againe other litle heades in the toppe of the
pillers aboue, according to the meaſure of the piller againſt the litle
nettes: and of the pomegranates were two hundred rewes round about the
ſecond litle head. \V And he ſette two pillers in the porch of the
temple: and when he had erected the piller on the right hand, he
called the name therof,
\TNote{Firmnes.}
Iachin: in like maner he erected the ſecond piller, and called the name
therof
\TNote{In ſtrength.}
Booz. \V And vpon the heades of the pillers he put a worke in maner of a
lilie: and the worke of the pyllers was perfected. \V He made alſo
\SNote{A veſſel ſo called for the bigneſſe being a great lauatorie.}
a ſea of founders worke of ten cubites from brimme to brimme, round in
cõpaſſe, the height therof was fiue cubites, and a corde of thirtie
cubites did compaſſe it, round about. \V And the grauing vnder the brĩme
compaſſed it, ten cubites going about the lauatorie: there were two rewes of
chamfered forowed grauinges caſt. \V And it ſtood vpon twelue oxen, of
which three looked to the North, and three to the Weſt, and three to the
South, and three to the Eaſt, and the ſea was ouer them: whoſe hinder
partes were al hid inward. \V And the thickenes of the lauatorie was of
three ounces: and the brimme therof as it were
%%% 0739
the brimme of a chalice, and the leafe of criſped lilie: it contayned
two thouſand
\SNote{\L{Batus} conteyneth about fiue galons: ſo this veſſel conteyned
tenne thouſand galons.}
bates. \V And he made ten braſen feete, of foure cubites in length
euerie foote, and foure cubites in bredth, and three cubites in
height. \V And the verie worke it ſelfe of the feete, was entergrauen:
and entergrauinges betwen the ioyntures. \V And betwen the litle crownes
and the
%%% o-0664
playtes, lions, and oxen, and cherubs: and in the ioyntures likewiſe
aboue: and vnder the lions, and oxen as it were bandes of braſſe hanging
downe. \V And foure wheeles at euerie foote, and axeltrees of braſſe:
and at foure ſides as it were litle ſhoulders vnder the lauatorie caſt,
looking one againſt an other. \V The mouth alſo of the lauatorie was
inward in the toppe of the head: and that which appeared outward, was of
one cubite al round, and together it had one cubite & a halfe: and in
the corners of the pillers were diuers engrauinges: and the middle
enterpillers ſquare not round. \V The foure wheeles alſo, which were at
the foure corners of a foote, ioyned one to an other vnder the foote:
one wheele had in height a cubite and a halfe. \V And they were ſuch
wheeles as are accuſtomed to be made in a chariote: and their axeltrees
and ſpokes, and ſtrakes, and naues, al caſt. \V For thoſe foure litle
ſhoulders alſo at euerie corner of one foote, were caſt out of the foote
and ioyned together. \V And in the toppe of the foote was a certayne
roundnes of halfe a cubite, ſo wrought, that the lauatorie might be put
thereon, hauing the engrauinges therof, and diuerſe caruinges of it
ſelf. \V He graued alſo in thoſe ſielinges, which were of braſſe, and in
the corners, cherubs, and lions, and palmetrees, as it were in the
ſimiltitude of a man ſtanding, that they ſemed not to be engrauen, but
put to round about. \V After this maner made he ten feete, of one caſting
and meaſure, & like grauing. \V He made alſo ten lauatories of braſſe:
one lauatorie conteyned fourtie bates, and it was of foure cubites: alſo
at euerie foote, that is ten, he put ſo manie lauatories. \V And he
ſette the ten feete, fiue on the right ſide of the temple, and fiue on
the leaft: and the ſea he put on the right ſide of the temple againſt
the Eaſt toward the South. \V Hiram therfore made cauldrons, and
ſhoueles, and litle pottes, and perfected al the worke of king Salomon
in the temple of our Lord. \V Two pyllers, and two cordes of the litle
heades, vpon the litle heades of the pyllers: and two litle nettes, to
couer the two cordes, that were
%%% 0740
ouer the heades of the pyllers. \V And foure hundred pomegranates in the
two nettes: two rewes of
\Fix{pomegramates}{pomegranates}{likely typo, fixed in other}
in euerie nette, to couer the cordes of the litle heades, which were
vpon the heades of the pyllers. \V And tenne feete, and tenne lauatories
vpon the feete. \V And one ſea, and twelue oxen vnder the ſea. \V And
cauldrons, and ſhouels, and litle pottes. Al the veſſels that Hiram
made to king Salomon in the houſe of our Lord, were of bright latten. \V
In the champayne countrie of Iordan did the king caſt thoſe thinges in a
clay ground, betwen Sacoth and Sarthan. \V And Salomon placed al
the veſſel: but for the exceding great multitude the braſſe could not be
weyed. \V And Salomon made al the veſſels in the houſe of our Lord: an
altar of gold, and a table, wherupon the loaues of propoſition ſhould be
put, of gold: \V and candleſtickes of gold, fiue on the right hand, and
fiue on the leaft againſt the oracle, of pure gold: and as it were litle
floures, and lampes aboue of gold: and golden ſnuffers, \V and water
pottes, and fleshehookes, and
%%% o-0665
phiales, and morters, and cenſars, of moſt pure gold: and the hindges of
doores of the inner \L{Sanctum ſanctorum}, and of the doores of the
houſe of the temple, were of gold. \V And Salomon perfected al the worke
that he did in the houſe of our Lord, and brought in the thinges that
Dauid his father
\SNote{Had deſigned and dedicated to holie vſes.}
had ſanctified, ſiluer and gold, and the veſſel, and layed them in the
treaſures of the houſe of our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
