%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 0717
%%% o-0644
\startChapter[
  title={Chapter 5}
  ]

\Summary{Hiram king of Tyre granteth timber and workmen for building the
  Temple: Salomon allowing victuals, and paying wages. 13.~The number of
  workmen and ouerſeers.}

Hiram alſo the king of Tyre ſent his ſeruantes to Salomon: for he heard
that they had anointed him king for his father: becauſe Hiram had bene
Dauids frend at al time. \V And Salomon ſent to Hiram, ſaying: \V Thou
knoweſt the wil of Dauid my father, and that he could not build a houſe
to the name of our Lord his God, becauſe of warres
%%% 0718
imminent round about vntil our Lord put them vnder the ſole of his
feete. \V But now our Lord my God hath geuen me reſt round about: and
there is no ſatan, nor il rencounter. \V Wherfore I purpoſe to build a
temple to the name of our Lord my God, as our Lord hath ſpoken to Dauid
my father, ſaying: Thy ſonne, whom I wil geue for thee vpon thy throne,
he shal build a houſe to my name. \V Command therfore that thy
%%% o-0645
ſeruantes cut me downe cedres out of Libanus, and let my ſeruantes be
with thy ſeruantes: and I wil geue thee the hire of thy ſeruantes
whatſoeuer thou wilt aske, for thou knoweſt how there is not in my
people a man that hath skil to hew wood as the Sidonians. \V When Hiram
therfore had heard the wordes of Salomon, he reioyſed excedingly, and
ſaid: Bleſſed be the Lord God this day, who hath geuen vnto Dauid a
ſonne moſt wiſe ouer this people ſo great in number. \V And Hiram ſent to
Salomon, ſaying: I haue heard whatſoeuer thou haſt willed me: I wil doe
al thy wil in cedre trees, and firre trees. \V My ſeruantes shal bring
them downe from Libanus to the ſea: and I wil put them in boates in the
ſea, vnto the place, which thou ſhalt ſignifie to me; and wil land them
there, and thou shalt take them: and thou ſhalt allow me neceſſaries,
that there be meate geuen for my houſe. \V Therfore Hiram gaue Salomon
cedre trees, and firre trees, according to al his wil. \V And Salomon
allowed Hiram twentie thouſand cores of wheate, for prouiſion for his
houſe, and twentie cores of moſt pure oile: theſe thinges did Salomon
geue to Hiram euerie yeare. \V Our Lord alſo gaue wiſedom to Salomon, as
he ſpake to him: & there was peace betwen Hiram & Salomon, and both
made a league. \V And king Salomon choſe workmen out of al Iſrael, and
the taxed number was of thirtie thouſand men. \V And he ſent them into
Libanus, ten thouſand euerie moneth by courſe, ſo that two monethes they
were in their houſes: and Adoniram was ouer this taxing. \V And Salomon
had ſeuentie thouſand of them that caried burdens, and eightie thouſand
hewers of ſtones in the mountaine: \V beſides the ouerſeers which were
ouer euerie worke, in number three thouſand, and three hundred that
commanded the people, and them that did the worke. \V And the king
commanded, that they should take great ſtones, choſen ſtones for the
foundation of the temple, and should ſquare them: \V which the maſons of
Salomon, and maſons of
%%% 0719
Hiram hewed: moreouer the Giblians prepared timber and ſtones, to build
the houſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
