%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 0755
%%% o-0677
\startChapter[
  title={Chapter 13}
  ]

\Summary{A prophet ſent from Iuda to Bethel fortelleth the birth of
  Ioſias, and deſtruction of Ieroboams altar, 4.~whoſe hand being
  ſodenly withered, 6.~is reſtored by the prophets prayer. 11.~The ſame
  prophet is deceiued by an other prophet, and ſlaine by a
  lion. 33.~Hieroboam proceedeth in impietie.}

And behold a man of God came out of Iuda, in the word of our Lord into
Bethel, Ieroboam
%%% !!! Same SNote as below. That is, two refs to same note.
%%% \SNote{}
ſtanding vpon the altar, and cenſing. \V And he cried out againſt the
altar in the word of our Lord, and ſayd: Altar, altar, thus ſaith our
Lord: Behold a child ſhal be borne to the houſe of Dauid, named
\SNote{This foreſhewing long before the name of a childe that ſhould be
borne, importeth that he ſhould do great thinges. See
\XRef{4.~Reg.~22.}}
Ioſias, and he ſhal immolate vpon thee prieſtes of the excelſes, which
now doe burne frankencenſe on thee, and he ſhal burne mens bones vpon
thee. \V And he gaue a ſigne in that day, ſaying: This ſhal be the
ſigne, that our Lord hath ſpoken: Behold the altar shal be clouen, and
the ashes shal be powred out in it. \V And when the king had heard the
word of the man of God, which he cried out againſt the altar in Bethel,
he ſtretched forth his hand from the altar, ſaying: Take him. And his
hand withered, which he ſtretched forth agaynſt him: neither was he able
to draw it backe vnto him. \V The altar alſo was clouen, and the ashes
were powred out of the altar, according to
%%% o-0678
the ſigne which the man of God had told before in the word of our
Lord. \V And the king ſaid to the man of God: Beſech the face of our
Lord thy God, and pray for me, that my hand may be reſtored me. And the
man of God beſought the face of our Lord, and the kinges hand was
reſtored to him, and it became as it was before. \V And the king ſpake
to the man of God: Come home with me, that thou mayſt dyne, and I wil
geue
\Fix{the}{thee}{obvious typo, fixed in other}
giftes. \V And the man of God anſwered the king: If thou wouldeſt geue
me the halfe part of thy houſe, I wil not come with thee, nor eate
bread, nor drinke water in this place: \V for ſo was it enioyned in the
word of our Lord cõmanding: Thou ſhalt not eate bread nor drinke water,
nor returne by the way that thou cameſt. \V He departed therfore by an
other way, and returned not by the way, that
\Fix{the}{he}{obvious typo, fixed in other}
came into Bethel.
%%% 0756
\V And a certaine prophete being old dwelt in Bethel, to whom his ſonnes
came and told him al the workes, that the man of God had done that day
in Bethel: and the wordes which he had ſpoken to the king, they told
their father. \V And their father ſayd to them: What way went he? His
ſonnes ſhewed him the way, by which the man of God was gone, which came
out of Iuda. \V And he ſaid to his ſonnes: Sadle me
\Fix{and}{an}{obvious typo, same in both}
aſſe. Who when they had ſadled it, he got vp, \V and went after the man
of God, and found him ſitting vnder a terebinth: and he ſaid to him: Art
thou the man of God that cameſt out of Iuda? He anſwered: I am he. \V
And he ſayd to him: Come home with me, that thou mayſt eate bread. \V
Who ſayd: I can not returne, nor come with thee, neither wil I eate
bread, nor drinke water in this place: \V becauſe our Lord ſpake to me
in the word of our Lord, ſaying: Thou ſhalt not eate bread, and thou
ſhalt not drinke water there, nor returne by the way thou wenteſt. \V
Who ſayd to him: I alſo am a prophet like to thee: and
\SNote{This man of Bethel was in deede a prophet of God, but in this
lied wickedly, and ſo deceiuing the other prophet, made him to breake
Gods commandment, for which he was ſlaine. VVherupon Hieroboam (whom the
wiked prophet ſought to pleaſe) was leſſe afeard to procede in
idolatrie.}
an Angel hath ſpoken to me in the word of our Lord, ſaying: Bring him
backe with thee into thy houſe, that he may eate bread, and drinke
water. He deceiued him, \V and brought him backe with him: he did eate
therfore bread in his houſe, and drunke water. \V And when they ſate at
the table, the word of our Lord came to the prophete, that brought him
backe. \V And he cried out to the man of God, which came out of Iuda,
ſaying: Thus ſayth our Lord: Becauſe thou haſt not beene obedient to the
mouth of our Lord, and haſt not kept the commandment, which our Lord thy
God commanded thee, \V and haſt returned, and eaten bread, & drunke
water in the place wherin he commanded thee that thou ſhouldeſt not eate
bread, nor drinke water, thy dead bodie ſhal not be brought into the
ſepulchre of thy fathers. \V And when he had eaten & drunke, he ſadled
his aſſe for the prophet, whom he brought backe. \V Who when he was
gone, a lion found him in the way, and
\SNote{Not only the deceiuer, but alſo he that is deceiued, is guiltie
and puniſhable for breakĩg Gods cõmandment.}
killed him, and his bodie was caſt forth in the way: and the aſſe ſtood
by him, and the lion ſtood by the dead bodie. \V And behold, men paſſing
by ſaw the dead bodie caſt in the way, and the lion ſtanding beſide the
bodie.
%%% o-0679
And they came and diuulged it in the citie, wherein that old prophet
dwelt. \V Which when that prophet heard, which had brought him backe out
of the way, he ſayd: It is the
%%% 0757
man of God, that was diſobedient to the mouth of our Lord, and our Lord
hath deliuered him to the lion, & he hath torne him, and killed him
according to the word of our Lord, that he ſpake to him. \V And he ſayd
to his ſonnes: Sadle me an aſſe. Who when they had ſadled, \V and he was
gone, he found his dead bodie caſt forth in the way, and the aſſe and
the lion ſtanding by the corſe: the lion
\SNote{By this it appeareth to be Gods worke and puniſhment.}
did not eate of the dead bodie, nor hurt the aſſe. \V The prophet
therfore tooke the corſe of the man of God, and layd it vpon the aſſe,
and returning brought it into the citie of the old prophete, that they
might mourne for him. \V And he layd his corſe in his owne ſepulchre:
and they mourned for him: Alas, alas my brother. \V And when they had
mourned for him, he ſayd to his ſonnes: When I ſhal be dead, burie me in
the ſepulchre, wherein the man of God is buried: beſide his bones lay my
bones. \V For aſſuredly the word ſhal come to paſſe, which he hath
foretold in the word of our Lord agaynſt the altar that is in Bethel:
and agaynſt al the temples of the excelſes, that are in the cities of
Samaria. \V After theſe wordes Ieroboam returned not from his wicked
way: but on the contrarie part he made of the moſt abiect of the people
prieſtes of the excelſes: Whoſoeuer would, he filled his hand, and he was
made a prieſt of the excelſes. \V And for this cauſe did the houſe of
Ieroboam ſinne, and was ouerthrowen, and deſtroyed from the face of the
earth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
