%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Three Kings
%%%% Chapter 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-22


\project douay-rheims


%%% 0776
%%% o-0696
\startChapter[
  title={Chapter 22}
  ]

\Summary{Achab King of Iſrael conſulting and beleuing falſe prophetes
  rather then Micheas a true prophet, accompanied with Ioſaphat king of
  Iuda, reſolueth to fight againſt the Syrians for Ramoth
  Galaad. 26.~Committeth Micheas to priſon, 29.~procedeth to the warre,
  34.~is ſlaine, and Ochozias ſuccedeth. 41.~Ioſophat refuſeth to
  traffike longer with Ochozias; dieth, and his ſonne Ioram
  reigneth. 52.~Ochozias foloweth the euil ſteppes of his parents.}

There paſſed therfore three yeares without warre betwen Syria and
Iſrael. \V And in the third yeare, Ioſaphat the king of Iuda went downe
to the king of Iſrael. (\V And the king of Iſrael ſayd to his ſeruantes:
Are you ignorant that Ramoth Galaad is ours, and we neglect to take it
out of the hand of the king of Syria?) \V And he ſayd to Ioſaphat: Wilt
thou come with me to fight into Ramoth Galaad? \V And Ioſaphat ſayd to
the king of Iſrael: I as I am, ſo thou alſo: my people and thy people are
one: and my horſemen thy horſemen. And Ioſaphat ſayd to the king of
Iſrael: Aske, I beſech thee, this day the word of our Lord. \V The king
of Iſrael therfore aſſembled the prophetes, about foure hundred men, and
he ſayd to them: Shal I goe into Ramoth Galaad to fight, or ſitte ſtil?
Who anſwered: Goe vp, and our Lord wil geue it into the kinges hand. \V
And Ioſaphat ſayd:
\SNote{The godlie King Ioſaphat iuſtly ſuſpecting the ſchiſmatical falſe
prophetes, aduiſed the other King to cõſult a true prophet of God.}
Is there not
%%% 0777
here ſome prophete of our Lord, that we may aske by him? \V And the king
of Iſrael ſayd to Ioſaphat: There is one man leaft, by whom we may aske
our Lord: but I hate him, becauſe he doeth not prophecie vnto me good,
but euil, Micheas the ſonne of Iemla. To whom Ioſaphat ſayd: Speake not
ſo ô king. \V The king of Iſrael therfore called a certaine eunuch, and
ſayd to him: Make haſt, and bring hither Micheas the ſonne of Iemla. \V
And the king of Iſrael, and Ioſaphat the king of Iuda ſate eche in his
throne clothed with royal attyre, in a court beſide the doore of the
gate of Samaria, and al the prophetes prophecied before them. \V
Sedecias the ſonne of Chanaana made him ſelf
\SNote{Falſprophetes imitate true prophetes in ſome exterior thinges, to
make their prophecies ſeme more authentical. Such hornes appeared in a
prophetical viſion to Zacharie the prophet.
\XRef{Zach.~1. v.~18.}}
hornes of yron, and ſayd: Thus ſayth our Lord: With theſe shalt thou
ſtrike Syria, til thou deſtroy it. \V And al the prophetes in like maner
prophecied, ſaying: Aſcend into Ramoth Galaad, and goe proſperouſly, &
our Lord wil deliuer it into the Kinges handes. \V But the meſſenger,
that went to cal Micheas, ſpake to him, ſaying: Behold the wordes of the
prophetes with one mouth preach good thinges to the king: let thy word
therfore be like to theirs, and ſpeake good thinges. \V To whom Micheas
ſayd: Our Lord liueth, whatſoeuer our Lord shal tel me, that wil I
ſpeake. \V He therfore came to the
%%% o-0697
king, and the king ſayd to him: Micheas, shal we goe into Ramoth Galaad
to fight, or ſitte ſtil? To whom he anſwered:
\SNote{The prophet geueth not this for a reſolute anſwer, but ſeing the
king wil goe, he prayeth he may go proſperouſly. And the king conceiueth
no otherwiſe of his anſwer, and therfore vrgeth him to anſwer reſolutly
in the next wordes, \Emph{I adiure thee &c.}}
Aſcend, and goe proſperouſely, and our Lord wil deliuer it into the
kinges handes. \V But the king ſayd to him: I adiure thee agayne and
agayne, that thou ſpeake not to me but that which is true in the name of
our Lord. \V And he ſayd: I ſaw al Iſrael diſperſed in the mountaynes,
as ſheepe not hauing a ſhepeheard, and our Lord ſayd: Theſe haue no
maiſter: let euerie man returne into his houſe in peace. (\V Therfore
the king of Iſrael ſayd to Ioſaphat: Did I not tel thee, that he doeth
not prophecie me good, but alwaies euil?) \V But he adding, ſayd:
Therfore heare the word of our Lord: I ſaw our Lord ſitting vpon his
throne, and al the hoſte of heauen aſſiſting him on the right hand and
on the left: \V and our Lord ſayd: Who shal deceiue Achab the king of
Iſrael, that he may goe vp, and fal in Ramoth Galaad? And one ſayd theſe
maner of wordes, and an other otherwiſe. \V And there came forth a
ſpirit, and ſtood before our Lord, and ſayd: I wil deceiue him. To whom
our Lord ſpake: Wherein? \V And he ſayd: I
%%% 0778
wil goe forth, and be a lying ſpirit in the mouth of al his
prophetes. And our Lord ſayd: Thou shalt deceiue, and shalt preuayle:
\SNote{The coherence of the text sheweth that God only permitted, but
commanded not the diuel to deceiue Achab. So S.~Auguſtin
\Cite{li.~2. q.~vlt. ad Simplic.}
S.~Greg.
\Cite{li.~2, c.~21. Moral.}
and other fathers explicate this & like places.}
goe forth, and doe ſo. \V Now therfore behold our Lord hath geuen the
ſpirite of lying in the mouth of al thy prophetes, that are here, and
our Lord hath ſpoken euil agaynſt thee. \V And Sedecias the ſonne of Chanaana
came, and ſmote Micheas on the cheeke, and ſayd: Hath the Spirit of our
Lord leaft me, and hath it ſpoken to thee? \V And Micheas ſayd: Thou
ſhalt ſee in that day, when thou shalt enter into thy chamber,
\SNote{VVhen this falſe prophet heard that the kĩg was ſlaine he hidde
him ſelfe, fearing the kings ſonnes, that they would kil him.
\Cite{Ioſephus li.~8. c.~14. Antiq.}}
within the chamber to be hid. \V And the king of Iſrael ſayd: Take
Micheas, and let him tarie with Amon the gouernour of the citie, and
with Ioab the ſonne of Amelech, \V and tel them: Thus ſayth the king:
Caſt this man into priſon, and feede him with bread of tribulation, and
water of diſtreſſe, til I returne in peace. \V And Micheas ſayd: If thou
returne in peace, our Lord hath not ſpoken in me. And he ſayd: Heare al
ye peoples. \V Therfore the king of Iſrael went vp, and Ioſaphat the
king of Iuda into Ramoth Galaad. \V The king of Iſrael therfore ſayd to
Ioſaphat: Take armour, and goe into the battel, and put on thyne owne
garmentes. Moreouer the king of Iſrael changed his habite, and went into
the battel. \V And the king of Syria had commanded the princes of his
chariotes thirtie and two, ſaying: You shal not fight agaynſt any
leſſer, or greater, but agaynſt the king of Iſrael onlie. \V When
therfore the princes of the chariotes had ſeene Ioſaphat, they ſuſpected
that he was the king of Iſrael, and making a violent aſſault they fought
agaynſt him: & Ioſaphat cried out. \V And the princes of the chariotes
perceiued that he was not the king of Iſrael, and they ceaſed from
him. \V And a certaine man bent his bow, directing the arrow at al
aduenture, and
\SNote{It happened by chance in reſpect of the archers intentiõ, but
otherwiſe by Gods prouidence directing his hand. So Achabs craftie
perſwading Ioſaphat to put on his kinglie attyre,
\XRef{(v.~30.)}
him ſelf fighting in vulgar armour, ſaued him not from iuſt reuenge.}
by chance he ſtroke the king of Iſrael betwen the lunges and the
ſtomacke. But he ſayd to his cochere: Turne thy hand, and
%%% o-0698
carrie me out of the armie, becauſe I am greuouſely wounded. \V The
battel therfore was fought that day, and the king of Iſrael ſtood in his
chariote agaynſt the Syrians, and he died in the euening: & the bloud of
the wound ranne into the middes of the chariote, \V and the herauld
ſounded in al the armie before the ſunne ſette, ſaying: Euerie man
returne into his citie, and into his countrie. \V And the king died, and
was carried into Samaria: and they buried the king
%%% 0779
in Samaria, \V and washed his chariote in the poole of Samaria, and the
dogges licked his bloud, and they washed the raynes of the bridle,
according to the word of our Lord which he had ſpoken. \V But the reſt
of the wordes of Achab, and al that he did, and the houſe of yuorie,
that he built, and of al the cities that he built, are not theſe thinges
written in the
%%% !!! XRef ?
Booke of the wordes of the daies of the kinges of Iſrael? \V Achab
therfore ſlept with his fathers, and Ochozias his ſonne reigned for
him. \V But Ioſaphat the ſonne of Aſa began to reigne ouer Iuda the
fourth yeare of Achab the king of Iſrael. \V Fiue and thirtie yeares old
was he when he began to reigne, and fiue and twentie yeares he reigned
in Ieruſalem: the name of his mother was Azuba the daughter of Salai. \V
And he walked in al the way of Aſa his father, and he declined not from
it: and he did that which was right in the ſight of our Lord. \V But yet
he tooke not away the excelſes: for as yet the people did ſacrifice, and
burnt incenſe in the excelſes. \V And Ioſaphat had peace with the king
of Iſrael. \V But the reſt of the wordes of Ioſaphat, and his workes,
which he did, and his battels, are not theſe thinges written in the
%%% !!! XRef ?
Booke of the wordes of the daies of the kinges of Iuda? \V Yea and the
remnant of the effeminate, which remayned in the daies of Aſa his
father, he tooke out of the land. \V Neither was there then a king
appoynted in Edom. \V But king Ioſaphat had made nauies on the ſea,
which should not ſaile into Ophir for gold: and they could not goe,
becauſe they were broken in Aſiongaber. \V Then ſayd Ochozias the ſonne
of Achab to Ioſaphat: Let my ſeruantes goe with thy ſeruantes in the
ſhippes. And Ioſaphat would not. \V And Ioſaphat ſlept with his fathers,
and was buried with them in the Citie of Dauid his father: and Ioram his
ſonne reigned for him. \V And Ochozias the ſonne of Achab began to
reigne ouer Iſrael in Samaria, in the ſeuententh yeare of Ioſaphat the
king of Iuda, and he reigned ouer Iſrael two yeares. \V And he did euil
in the ſight of our Lord, and walked in the way of his father and his
mother, and in the way of Ieroboam the ſonne of Nabat, who made Iſrael
to ſinne. \V He ſerued alſo Baal, and adored him, and prouoked our Lord
the God of Iſrael, according to al thinges which his father had done.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
