%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 1502
%%% o-1386
\startChapter[
  title={Chapter 17}
  ]

\Summary{Horrible darkenes falling in Ægypt, 19.~the reſt of the world
  had ordinarie light.}

%%% 1503
For thy iudgements ô Lord are great, & thy wordes inexplicable, for this
cauſe the ſoules lacking diſcipline haue erred. \V For whiles the wicked
are perſwaded that they can rule ouer the holie nation:
\CNote{\XRef{Exo.~1.}
&
\XRef{10.}}
fettered with the bands of darknes, and
\SNote{Literally the Ægyptians had darknes three dayes together,
\XRef{Exo.~10. v.~22.}
Morally they & other gentiles vvere in darknes vvithout faith in God,
til Chriſts Reſurrection the third day.}
long night, shut vp vnder roofes, they haue lyen fugitiues from the
euerlaſting prouidence. \V And whiles they thincke that they lie hid in
obſcure
%%% o-1387
ſinnes, they were diſperſed by the darke couert of obliuion, being
horribly afrayd and diſturbed with exceding admiration. \V For neither
did the denne that conteyned them, keepe them without feare: becauſe the
ſound coming downe trubled them, and ſorowful viſions appearing to them,
put them in feare. \V And no force certes of the fyre could geue them
light, neither could the clere flames of the ſtarres lighten that
horrible night. \V But there appeared to them ſoden fyre, ful of feare:
and being ſtroken with the feare of that face, which was not ſene, they
eſtemed the thinges that were ſene to be worſe: \V and there were added
deriſions of the magical art, and contumelious rebuke of the glorie of
their wiſdom. \V For they which promiſed that they would expel feares
and pertubations from the languishing ſoule, theſe with deriſion
languished ful of feare. \V For although none of the monſters diſturbed
them: being moued with the paſſing by of beaſts, and hiſſing of
ſerpents, they perished trembling: and denying that they ſaw the ayre,
which by no meanes any man could auoyde. \V For wheras wickednes is
fearful, it geueth teſtimonie of condemnation: for
\SNote{A trubled conſcience is a great torment.}
a trubled conſcience doth alwayes preſume cruel thinges. \V For feare is
nothing els but a bewraying of the aydes of cogitation. \V And whiles
inwardly there is leſſe expectation, the greater doth he count the
ignorance of that cauſe which maketh the torment. \V But they that
during the night in deede impotent, and coming vpon them from the loweſt
and higheſt hel, ſlept the ſame ſleepe, \V were ſometime moleſted with
the feare of monſters, ſometime fayled by paſſing away of the ſoule: for
ſoden feare and vnlooked for came vpon them. \V Moreouer if any of them
had fallen downe, he was kept shut vp in priſon without yron. \V For if
one were a husbandman, or if a shepheard, or worker of the labours in
the filde were ſodenly taken, he ſuſteyned neceſſitie ineuitable. \V For
with one chayne of darkenes they were al tyed together. Whether it were
the hiſſing winde, or among
%%% 1504
the thicke boughes of trees the ſweete ſound of birdes, or the force of
water running downward exceedingly, \V or the mightie ſound of rockes
tumbled headlong, or the running of playing beaſts, that were not ſene,
or the mightie noyſe of roaring beaſtes, or an Echo reſounding from the
higheſt mountaynes: they made them ſwoone for feare. \V For
\SNote{This ſignified the conuerſion of al nations to Chriſt.}
al the world was illuminated with a cleare light, & none was hindered in
their workes. \V But ouer them onlie was layd an heauie night, the image
of darkenes, which was to come vpon them. They therfore were vnto
themſelues more heauie then the darknes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
