%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 1504
%%% o-1388
\startChapter[
  title={Chapter 18}
  ]

\Summary{In the Ægyptian darknes, the Iſraelites ſaw clerly, and were
  not ſene of their enimies. 5.~For the Ægyptians crueltie againſt the
  Hebrewes infants, al their owne firſt borne were ſlaine, and their
  whole armie drowned in the redſea. 20.~But fire deuouring the
  rebellions in Chores ſchiſme, was quenched by Aarons interceſſion.}

But to thy
\SNote{The Church is called holie becauſe it profeſſeth holines and hath
alwayes ſome holie men, & without the Church there is no ſanctitie.}
ſaincts there was very great light, and their voyce in deede they heard,
but figure they ſaw not. And becauſe themſelues alſo did not ſuffer by
the ſame thinges, they magnified thee: \V and they that before had bene
hurt, becauſe they were not hurt, gaue thankes: and that there might be
a difference, they asked a gift. \V For the which cauſe they had a
burning pillar of fyre for a guide of the vnknowen way, and thou gaueſt
them the ſunne without hurt of a good harbour. \V They in deede worthie
to lacke light, and to ſuffer the priſon of darkenes, which kept thy
children shut vp, by whom the vncorrupt light of the law began to be
geuen to the world. \V When they decreed to kil the infantes of the
iuſt: and
\SNote{VVhen the Ægyptians drowned the Hebrews children, Moyſes was
ſaued, and reſerued to guide the Iſraelites, when the Ægyptians were
drowned.}
one child being layd forth, and deliuered, thou to the reprouing of
them, didſt take away a multitude of children, and deſtroyeſt them
together in the mightie water. \V For that night was knowen before of
our fathers, that they knowing in deede what othes they had credited,
might be of better comfort. \V And by thy people in deede the health of
the iuſt, was receiued, but deſtruction by the vniuſt. \V For as thou
didſt hurt the aduerſaries: ſo vs alſo thou didſt magnifie, prouoking
vs. \V For the iuſt children of the good ſacrificed ſecretly, and
diſpoſed the law of iuſtice in concorde: that the iuſt should receiue
both good & euil alike, ſinging now the prayſes of the fathers. \V But
there ſounded a
%%% 1505
diſagreing voyce of the enemies, and a lamentable moorning was heard for
the bewayled infants. \V And the ſeruant with the maſter was afflicted
with like punishment, and a man of the vulgar ſort ſuffered the like
thinges to the king. \V Al therfore alike by one name of death had dead
ones innumerable. For neither did the liuing ſuffice to burie them:
becauſe in one moment, that which was the nobler nation of them, was
deſtroyed. \V For concerning al thinges being incredulous becauſe of the
inchantments, but then firſt when there was deſtruction of the firſt
begotten, they promiſed to be the people of God. \V
\SNote{A prophecie of Chriſt, comming into this world when there was
temporal peace, but extreme darkenes of ignorance.}
For when quiet ſilence conteyned al thinges, and the night was in the
midde way of her courſe, \V thy omnipotent word ſalying out of heauen
from the royal ſeates, lighted as a ſeuere conquerour vpon the
%%% o-1389
middes of the land of deſtruction, \V a sharpe ſword carying thy vnfeyned
commandment, and ſtanding filled al with death, and reached euen to
heauen ſtanding on the earth. \V Then incontinent the viſions of
naughtie dreames trubled them, and feares vnlooked for came vpon
them. \V And one here an other there, caſt forth halfe aliue, shewed for
what cauſe of death he died. \V For the viſions, that trubled them,
forewarned theſe thinges, that they might not perish as ignorant why
they ſuffered euils. \V But then
\SNote{An other example of difference in Gods punishing his people for
their amendment, and of the obſtinate vnto their ruine.
\XRef{Exo.~14. v.~28.}
\XRef{Num.~16, v.~46.}}
there touched the iuſt alſo a tentation of death, and
\CNote{\XRef{Num.~16.}}
a diſturbance of the multitude was made in the wildernes: but thy wrath
did not long continew. \V For a man without blame haſting to pray for
the people, bringing forth the shilde of his miniſterie prayer, and by
incenſe alleaging ſupplication, reſiſted the wrath, and made an end of
the neceſſitie, shewing that he is thy ſeruant. \V And he ouercame the
multitudes, not in ſtrength of bodie, nor with might of armour, but with
a word ſubdewed he him, that vexed him, rehearſing the oathes of the
parents, and the teſtament. \V For when they were now falled dead by
heapes one vpon an other, he ſtood betwen, and cut of the violence, and
ſeperated that way which leadeth to the liuing. \V For in the veſture
downe to the foote, which he had, was al the world: and the glorious
thinges of the fathers were grauen in the foure iewels of ſtones & thy
magnifence was written in the diademe of his head. \V And to theſe he
that deſtroyed gaue place, & theſe did he feare: for the proofe onlie of
wrath was ſufficient.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
