%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 1483
%%% o-1369
\startChapter[
  title={Chapter 04}
  ]

\Summary{Great difference betwen chaſte, and adulterous
  generations. 7.~Speedier death of the iuſt is recompenſed by Gods
  prouidence, 19.~but the wicked incurre greater damnation by liuing
  long.}

O how beautiful is
\SNote{Chaſtitie of the bodie is a ſingular great vertue, & ſpiritual
chaſtitie of true faith and religion is greatter, and more generally
commended, as the roote and fundation of al vertues. For vvithout
faith it is vnpoſſible to pleaſe God.}
the chaſte generation with glorie: for the memorie therof is immortal:
becauſe it is knowen both with God, and with men. \V When it is preſent,
they imitate it, and they deſire it when it hath withdrawen itſelf, and
it triumpheth crowned for euer, winning the reward of vndefiled
conflictes. \V But the multitude of the impious, that hath manie
children shal not be profitable, and baſtarde plants shal not take deepe
roote, nor lay ſure fundation. \V And if in the boughes for a time they
shal ſpring being weakly ſet, they shal be moued of the winde, and by
the vehemencie of the windes they shal be rooted out. \V For the
vnperfect boughes shal be broken, and their fruites shal be
vnprofitable, and ſowre to eate, and meete for nothing. \V For the
children that be borne of wicked ſleepes, are witneſſes of wickednes
againſt the parents in their examination. \V But
\SNote{VVhen ſoeuer the iuſt dieth it is profitable for him & dying yong
his immaculate life is more commendable then old age in the wicked.
\XRef{v.~16.}}
the iuſt if he be preuented with death, shal be in a place of
refreshing. \V For venerable old age is not that of long time, nor
accounted by the number of yeares: but the vnderſtanding of a man are
grey heares: \V and an immaculate life is old age. \V Pleaſing God he is
made beloued, and liuing among ſinners he was tranſlated. \V
\LNote{He vvas taken avvay.}{By this place S.~Auguſtin proueth that ſuch
as died in good ſtate, might haue fallen into wickednes, if they had
liued longer: and therfore it was a benefite to them, to dye ſooner.
\MNote{Gods preſcience doth not preiudice mans free wil.}
And that it is neuertheles certaine, that God both knew the poſſibilitie
that ſuch might, yea would haue ſinned, if they had liued longer, and
alſo knew that they ſhould die ſooner, and ſo eſcape that danger. VVhich
aſſured foreknowlege of al thinges that shal be, or may be, ſtandeth wel
with mans free wil: againſt certaine that inclined to Pelagianiſme,
attributing too much to mans free wil; and detracting from Gods
foreknowlege, and prouidence. VVhich he prouing by this place, his
aduerſaries excepted againſt the auctoritie of this booke, and therfore
he alſo proueth that it is Canonical Scripture.
\CNote{et
\Cite{l.~17. c.~20. ciuit.}}
\Cite{li. de Prædeſt. Sanct. c.~14.}}
He was taken away leſt malice should change his vnderſtanding, or leſt
anie guile might deceiue his ſoule. \V For the bewitching of vanitie
obſcureth good thinges, and the inconſtancie of concupiſcence peruerteth
the vnderſtanding that is without malice. \V Being conſummate in short
ſpace he fulfilled much time: \V for his ſoule pleaſed God: for this
cauſe he haſtened to bring him out of the middes of iniquities: but the
peoples that are ſeing, and not vnderſtanding, nor putteth ſuch thinges
in their hartes: \V that the grace of God, and mercie is toward his
ſaintes, and reſpect toward his elect. \V But the iuſt dead condemneth
the
%%% 1484
impious aliue, and youth ſoone ended, the long life of the vniuſt. \V
For they shal ſee the end of the wiſe, and shal not vnderſtand what God
hath thought of him, and why our Lord
%%% o-1370
hath fenſed him. \V For they shal ſee and shal contemne him: but our
Lord shal laugh them to ſcorne, \V and they shal fal after this without
honour, and in contumelie among the dead for euer: becauſe he shal
breake them puffed vp
\SNote{The damned ſhal be vvithout al excuſe vtterly confounded in their
owne conſciences.}
without voice, and shal remoue them from the fundations, and they shal
be made deſolate vnto the higheſt degree: and shal be
\Fix{moorning,}{mourning,}{obvious typo, fixed in other}
and the memorie of them shal perish. \V They shal come feareful in
cogitation of their ſinnes, and their iniquities on the contrarie shal
conuince them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
