%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 1484
%%% o-1370
\startChapter[
  title={Chapter 05}
  ]

\Summary{In the general iudgement, the wicked ſeing the iuſt, whom they
  had contemned, to be in great honour, shal bewaile their owne miſerie,
  9.~conſidering that their pleaſure was short, 16.~and the ioy of the
  bleſſed shal be for euer. 18.~God wil arme himſelf, and al creatures,
  to punish the impious.}

Then shal the iuſt ſtand in great conſtancie againſt thoſe that haue
afflicted them, and
\SNote{VVicked men in their falſe conceipt iudge the trauels of the iuſt
to be vaine & fructles.}
taken away their labours. \V They ſeing shal be trubled with horrible
feare, and shal meruel at the ſodennes of vnexpected ſaluation, \V
ſaying within themſelues,
\SNote{Repentance of the damned is only for the loſſe & paine whereinto
they are fallen not of loue towards God, or vertue, nor of hate towards
ſinne, & therfore is fructles bringing no comforte nor helpe at al, but
euerlaſting torment and anguiſh of mind.}
repenting, and ſighing for anguish of ſpirit: Theſe are they, whom we
had ſometime in deriſion, and in a parable of reproch. \V
\CNote{\XRef{Chap.~3. v.~2.}}
We ſenſleſſe eſtemed their life madnes, and their end without honour. \V
Behold how they are counted among the children of God, and their lot is
among the ſaints. \V We therfore haue erred from the way of truth, and
%%% 1485
the light of iuſtice hath not shined to vs, and the ſunne of
vnderſtanding roſe not to vs. \V We are weried in the way of iniquitie
and perdition,
%%% o-1371
and haue walked hard wayes, but the way of our Lord we haue not
knowen. \V What hath pride profited vs? or what commoditie hath the
vaunting of riches brought to vs? \V Al thoſe thinges are paſſed away as
a shadow, and as a meſſenger running before, \V
\CNote{Pro.~3. v.~19.}
and as a shippe, that
paſſeth through the ſurging waters: wherof, when it is paſt, the trace
can not be found, nor the path of that shippes keele in the waues: \V
or as a bird, that flyeth through in the ayre, of which there is no
token can be found of her paſſage, but only a ſound of the winges
beating the light winde: and by vehemence of going cutting the ayre,
mouing the winges she is flowen through, and afterward there is no ſigne
found of her way: \V or as when an arrow is shotte forth to a ſette
marke, the diuided ayre is forthwith cloſed in itſelf againe, ſo that
the paſſage therof is not knowen: \V ſo we alſo being borne forthwith
ceaſed to be: and of vertue certes haue bene able to shew no ſigne: but
in our naughtines we are conſumed. \V Such thinges
\SNote{For the certaintie of thinges that shal be. Prophetes do very
often ſpeake in the pretertence, of thinges to come as if they vvere
already paſt.}
ſayd they in hel, which ſinned: \V becauſe the hope of the impious is as
duſt, which is taken away with the winde: and as a thinne froth, which
is diſperſed by the ſtorme: and as ſmoke that is ſcatered abrode by the
winde: and as the memorie of a gheſt of one day that paſſeth. \V But the
iuſt shal liue for euer, and their reward is with our Lord, and
cogitation of them with the Higheſt. \V Therfore shal they receiue a
kingdom of honour, & a crowne of beautie at the hand of our Lord:
becauſe with
his right hand he wil couer them, and
\Fix{with in}{with}{obvious typo, fixed in other}
his holie arme defend
them. \V And
\SNote{As the ioy of the bleſſed, ſo contrariewiſe the miſerie of the
damned is meruelous great, & for euer vnchangeable.}
his zele wil take armour, and he wil arme the creature to the reuenge of
the enemies. \V He wil put on iuſtice for a breſtplate, & wil take
ſincere iudgement for an helmet: \V He wil take equitie for an
\Fix{inuicible}{inuincible}{obvious typo, fixed in other}
ſhilde: \V and He wil ſharpen fierce wrath for a ſpeare, and the round
world shal fight with Him againſt the ſenſleſſe. \V The shottes of
lightenings shal goe directly, & as it were from a bow of the clouds wel
bent they shal be caſt forth, and shal light on a certaine place. \V And
from rocked wrath shal thicke haile ſtones be caſt, the water of the ſea
shal rage againſt them, and the riuers shal runne together roughly. \V A
ſpirit of powre shal ſtand againſt them, and as a hurlewinde shal diuide
them: and their
%%% 1486
iniquitie shal bring al the land to a deſert, and naughtines shal
ouerthrow the ſeates of the mightie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
