%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 1482
%%% o-1368
\startChapter[
  title={Chapter 03}
  ]

\Summary{The iuſt contemned by the wicked, and proued by tentations are
  happie, 10.~and the wicked vnhappie. 12.~Chaſtitie shal be rewarded, &
  adulterous generations shal not proſper.}

But
\SNote{Temporal death of the iuſt, is the way to eternal life. VVhere
damnation (called here the tormẽt of death) ſhal not touch them.}
the ſoules of the iuſt are in the hand of God, and the torment of death
shal not touch them. \V
\CNote{\XRef{Chap.~5. v.~4.}}
\SNote{For albeit Martyrs ſeme in the eyes of the vnwiſe to dye, or to
be extinguiſhed, they paſſe in dede into eternal, and vnſpeakable
glorie.}
They ſemed in the eies of the vnwiſe to die: and their deceaſe was
counted affliction: \V and that which with vs is the way, is
deſtruction: but they are in peace. \V And though before men they
ſuffered torments, their hope is ful of immortalitie. \V Vexed in few
thinges, in many they shal be wel diſpoſed of: becauſe God hath tempted
them, and hath found them worthie of him ſelfe. \V As gold in the
furnace he hath proued them, and as an hoſt of holocauſt he hath
receiued them, and in time there shal be reſpect of them. \V
\CNote{\XRef{Mat.~13. v.~43.}}
The iuſt shal shine, and as ſparkes in a place of reedes they shal runne
abrode. \V They
\SNote{Al the iuſt ſhal approue Gods iudgmẽt, condemning the wicked.}
shal iudge nations, & haue dominion ouer peoples, and their Lord shal
reigne for euer. \V They that truſt in him, shal vnderſtand truth: and
the faithful in liue shal reſt in him: becauſe reſt and peace is to his
elect. \V But the impious according to the thinges which they haue
thought, shal haue correption: which haue neglected the iuſt, & haue
reuolted from our Lord. \V For he that reiecteth wiſdom, and diſcipline,
is vnhappie: and their hope is vaine, and labours without fruite and
their workes vnprofitable. \V Their
\SNote{Literally is vnderſtood, that the wiues of adulterers, often become
adultreſſes, & their children wicked. Morally, their ſenſualitie & al
their vvorkes are wholly corrupted.}
wiues are ſenſles, and their children moſt wicked. \V Curſed is their
creature: becauſe happie is the barren woman: and the vndefiled, which
hath not knowen bed in ſinne, she shal haue fruite in viſitation of
holie ſoules: \V and
\CNote{See S.~Ierom
\Cite{in Iſa.~56. v.~4.}}
the eunuch, that hath not wrought iniquitie with his hands, nor thought
moſt wicked things againſt God: for the choſen gift of fayth shal be
geuen to him, and a moſt acceptable lot in the temple of God. \V For
%%% o-1369
of good labour there is glorious fruite, and the roote of wiſdom which
falleth not. \V But the children of adulterers shal be in conſummation,
and the ſede of the vnlawful bed
%%% 1483
shal be deſtroyed. \V And if certes they be of long life, they shal be
reputed for nothing, & their laſt old age shal be without honour. \V And
if they dye quickly they shal haue no hope, nor ſpeach of comfort in the
day of acknowledging. \V For of a wicked nation the endes are cruel.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
