%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 1491
%%% o-1376
\startChapter[
  title={Chapter 10}
  ]

\Summary{The
\MNote{The 3.~part.

The excellent effectes of wiſdom & iuſtice.}
benefites of wiſdom are declared by examples, in Adam, 4.~Noe,
5.~Abraham, 6.~Lot, 10.~Iacob, 13.~Ioſeph, 15.~and the people of
Iſrael.}

%%% !!! Make the SNotes with the names referenced above into MNotes?
She 
\CNote{\XRef{Gen.~1. v.~27.}}
kept him,
\SNote{Adam.}
that was firſt made of God father of the world, when he was created
alone, \V and she
\SNote{By this it is certaine that our firſt parẽt Adam truly repented,
and had remiſſion of his ſinne.}
brought him out of his ſinne, and gaue him powre to conteyne al
thinges. \V
\CNote{\XRef{Gen.~4. v.~8.}}
After the vniuſt departed in his anger from her, by the furie of
brothers manſlaughter perished. \V
\CNote{\XRef{Gen.~6.}}
For whoſe cauſe, when water deſtroyed the earth, wiſdom healed it
againe, gouerning the
\SNote{Noe.}
iuſt by contemptible wood. \V
\CNote{\XRef{Gen.~19.}}
She euen in the conſent of wickednes, when the nations had confederated
themſelues, knewe
\SNote{Abraham.}
the iuſt, and preſerued him without blame to God, and in his ſonnes
mercie kept the ſtrong. \V She deliuered
\SNote{Lot.}
the iuſt, fleing from the impious that perished, when the fyre came
downe vpon Pentapolis: \V to whom for a witnes of their wickednes the
deſert land ſtandeth ſmoking, and
\SNote{In al trees about Sodom there is only ſhew of fruite, which when
it is touched falleth into duſt.}
trees haue fruites
%%% o-1377
at vncertaine ſeaſon, and the memorie of an incredulous ſoule a ſtanding
\SNote{Lots wife an example of inconſtancie.}
piller of ſalt. \V For pretermitting wiſdom they did not only ſlippe in
this, that they were ignorant of good thinges, but they left alſo vnto
men a memorie of their foolishnes, that in thoſe thinges, in which they
ſinned in, they could not be hid neither. \V But wiſdom hath deliuered
them, that obſerue her from ſorowes. \V 
\CNote{\XRef{Gen.~28.}}
And
\SNote{Iacob.}
the iuſt fleing his brothers wrath, she conducted by the right wayes,
and shewed him the kingdom of God, and gaue him the knowlege of the
holie, did honeſt
%%% 1492
him in labours, and
\Fix{accomplised}{accomplished}{obvious typo, fixed in other}
his labours. \V In the fraude of the circumuenters of him she was
preſent with him, and made him honorable. \V She kept him from the
enemies, and from ſeducers she defended him, and from ſeducers she
defended him, and gaue him a ſtrong fight,
\Fix{thas}{that}{obvious typo, fixed in other}
he might ouercome, and know that wiſdom is mightier then al. \V 
\CNote{\XRef{Gen.~37.}}
She forſooke not
\SNote{Ioſeph.}
the iuſt being ſold, but deliuered him from ſinners: and she went downe
with him into the pitte, \V
\CNote{\XRef{Gen.~41.}}
and in bands leaft him not, til she brought him the ſcepter of a
kingdome, and might againſt them, that oppreſſed him: and shewed them to
be lyers, that ſpotted him, and gaue him eternal glorie. \V
\CNote{\XRef{Exo.~1.}}
The
\SNote{Gods peculiar people.}
iuſt people, and ſeede without blame she deliuered from the nations that
oppreſſed them. \V
\CNote{\XRef{Exo.~3.}}
She entered into the ſoule of the ſeruant of God, and ſtood againſt
dreadful kinges in wonders and ſignes. \V And she rendred to the iuſt the
hope of their labours, and conducted them in a meruelous way: and she
was vnto them for a 
\Fix{couerr}{couer}{obvious typo, fixed in other}
in the day, and for the light of ſtarres by night: \V
\CNote{\XRef{Exo.~14.}}
and she tranſported them through the Redſea, and caried them ouer
through a great water. \V But their enemies she drowned in the ſea, and
from the depth of hel she brought them out. Therfore the iuſt tooke the
ſpoyles of the impious, \V
\CNote{\XRef{Exo.~12.}
\XRef{Exo.~15.}}
and they ſang thy holie name ô Lord, and thy victorious hand they
praiſed together: \V becauſe wiſdom hath opened the mouth of the dumme,
and the tongues of infants she hath made eloquent.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
