%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 1486
%%% o-1372
\startChapter[
  title={Chapter 06}
  ]

\Summary{Kinges and al Magiſtrates are againe admonished to exerciſe
  iuſtice: 7.~otherwiſe they shal be more greuouſly punished. 13.~Wiſdom
  may eaſely be found, 18.~by thoſe that ſincerely deſire it. 22.~And is
  very profitable (25.~excepting the enuious, or il diſpoſed) 26.~both
  to prince and people.}

VViſedom
\CNote{\XRef{Eccle.~9. v.~18.}}
is better then ſtrength: and a wiſeman then a ſtrong. \V Heare therfore
ye kinges, & vnderſtand, lerne ye iudges of the endes of the earth. \V
Geue eare ye, that rule multitudes, and that pleaſe yourſelues in
multitudes of nations: \V
\CNote{\XRef{Rom.~13. v.~1.}}
becauſe the
\SNote{Al powre is from God: & therfore to be reſpected, though
the magiſtrates ſometimes abuſe their auctoritie.}
powre is geuen you of our Lord, and ſtrength by the Higheſt, who wil
examine your workes, and ſearch your cogitations: \V becauſe when you
were the miniſters of his kingdom, you iudged not rightly, nor kept the
law of iuſtice, nor haue walked according to the wil of God. \V Horribly
and quickly wil he appeare to you: becauſe moſt ſeuere iudgement shal be
done on them, that beare rule. \V For to the litle one mercie is
granted: but
\SNote{As euerie ones charge is more or leſſe, ſo his account is eaſier
or harder, and the punishment (if he offend) ſmaller or greater.
\Cite{S.~Greg. ho.~9. in Euang.}}
the mightie shal mightely ſuffer torments. \V For God wil not except any
mans perſon, neither wil feare the greatnes of any man: becauſe he made
the litle and the great, & he hath equally care of al. \V But to the
ſtronger more ſtrong torment is imminent. \V To you therfore ô kings are
theſe my wordes, that you may lerne wiſdom, and not fal. \V For they
that haue kept iuſt thinges iuſtly, shal be iuſtified: and they that
haue lerned theſe thinges, shal find what they may anſwer. \V Couet ye
therfore my wordes, and loue them, and you shal haue diſcipline. \V
Wiſdom is cleere, and ſuch as neuer fadeth, and is eaſely ſene of them
that loue her, and is found of them that ſeeke her. \V She preuenteth
them that couete her, that she firſt may shew herſelf vnto them. \V He
that awaketh early to her, shal not labour: for he shal find her
ſitting at her doores. \V To thinke therfore of her, is perfect
vnderſtanding: and he that watcheth for her, shal quickly be ſecure. \V
Becauſe she goeth about ſeeking them that be worthie of her, and in the
wayes she wil shew her ſelf to them cheerefully, and in al prouidence
she wil meete them. \V For
\SNote{VViſdom is attained by this gradation and ſo from firſt to laſt,
a reſolute deſire ioyneth faithful ſoules to God.
\XRef{v.~21.}}
the beginning of her is the moſt true deſire of diſcipline. \V The care
therfore of diſcipline, is loue:
%%% 1487
and loue is the keeping of her lawes: and the keeping of the lawes, is
the conſummation of incorruption: \V and incorruption maketh to be next
to God. \V Therfore the deſire of wiſdom leadeth to the euerlaſting
kingdom. \V If therfore you be delighted with thrones, and with ſcepters
ô ye kinges of the people, loue wiſdom, that you may reigne for euer. \V
Loue the light of wiſdom al ye that beare rule ouer peoples. \V But what
wiſdom is, and how she was made, I wil declare: and I wil not
%%% o-1373
hide from you the myſteries of God, but from the beginning of her
natiuitie I wil ſearch out, and ſette the knowlege of her into light,
and wil not let paſſe the truth: \V neither wil I goe with pyning enuie:
becauſe ſuch a man shal not be partaker of wiſdom. \V But the multitude
of the wiſe is the health of the round world: and a wiſe king is the
ſtabilitie of the people. \V Therfore take ye diſcipline by my wordes,
and it shal profite you.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
