%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 1480
%%% o-1367
\startChapter[
  title={Chapter 02}
  ]

\Summary{Such as hope not of life to come, 6.~addict themſelues to
  preſent pleaſures: 10.~and perſecute the iuſt, eſpecially our Sauiour
  Chriſt, as contrarie to their wickednes. 23.~Death came vpon man by
  the diuels enuie.}

For they haue ſaid thinking with themſelues not wel:
\SNote{Carnal men thincking the ſoule to be mortal, and ſo neither paine
nor reward to be after death, fal to theſe prophane thoughts and
ſpeaches of infidelitie.}
Little, and with tediouſnes is the time of our life: and in the end of a
man there is no recouerie, and there is none knowne that hath returned
from hel: \V becauſe of nothing were we borne, and after this, we shal
be as if we had not bene: becauſe the breath is a ſmoke in our
noſthrels: & ſpeach
%%% 1481
a ſparke to moue our hart. \V Which being extinguished, our bodie shal
be ashes, and the ſpirit shal be powred abrode as ſoft ayre, and our
life shal paſſe as the trace of a cloude, and shal be diſſolued as a
miſt, which is driuen away by the beames of the ſunne, and oppreſſed
with the heate therof: \V and our name in time shal be forgotten, and no
man shal haue remembrance of our workes. \V For our time is the paſſing
of a shadow, and there is no returne of our end: becauſe it is ſealed,
and no man returneth. \V
\SNote{Of infidelitie touching paine or reward after death, procedeth
the Epicures life.}
Come therfore, and let vs enioy the good thinges that are, and let vs
quickly vſe the creature as in youth. \V Let vs fil ourſelues with
precious wine, and oyntments: and let not the flowre of the time paſſe
vs. \V Let vs crowne ourſelues with roſes, before they wither: let there
be no medow, which our riote shal not paſſe through. \V Let none of vs
be exempted from our rioteouſnes: euerie where let vs leaue ſignes of
ioy: becauſe this is our portion, and this our lot. \V Let vs
\SNote{Infidels are not content to liue in riote, but doe alſo enuie and
perſecute the iuſt, vvhoſe good examples vexe their mindes, ſturred
therto by the diuel.}
oppreſſe the poore iuſt man, and not ſpare the widow, nor reuerence the
oldmans grey head of long time. \V But let our ſtrength be the law of
iuſtice: for that which is weake, is found vnprofitable. \V
\SNote{An euident prophecie of the Iewes malice perſecuting our
Sauiour. Fulfilled by the chief Prieſtes, Scribes, and Ancientes,
recorded by the Euangeliſts.
\XRef{Mat.~27. v.~41.}
\XRef{Mat.~14. v.~53.}}
Let vs therfore circumuent the iuſt, becauſe he is vnprofitable to vs,
and he is contrarie to our workes, and reprochfully obiecteth vnto vs
the ſinnes of the law, and defameth in vs the ſinnes of our
diſcipline. \V He boaſteth that he hath the knowlege of God, and nameth
himſelfe the ſonne of God. \V 
\CNote{\XRef{Mat.~27. v.~43.}}
He is made vnto vs to the defaming of our cogitations. \V He is greuous
vnto vs euen to behold, becauſe his life is vnlike to others, and his
wayes are changed. \V We are eſtemed of him as triflers, and he
abſteyneth from our wayes as from vncleannes, and he preferreth the
later ends of the iuſt, and glorieth that he hath God for his father. \V
Let vs ſee therfore if his wordes be true, and let vs proue what thinges
shal come to him, and we
%%% o-1368
shal know what shal be his later ends. \V For if he be the true ſonne of
God, he wil defend him, & wil deliuer him from the hands of the
aduerſaries. \V By contumelie and torment let vs examine him, that we may
know his reuerence, and proue his patience. \V To a moſt shameful death
let vs condemne him: for there shal be reſpect had vnto him by his
wordes. \V Theſe thinges haue they thought, and haue erred: for their
malice hath blinded them. \V And they haue not knowen
\SNote{VVant of beleuing diuine Myſteries, namely the reward of the
iuſt, & puniſhmẽt of the wicked, is cauſe of diſſolute life, and of
hatred againſt the good.}
the ſacraments of God, nor hoped for the reward of iuſtice, nor
%%% 1482
eſtemed the honour of holie ſoules. \V For God created man
incorruptible, and to the image of his owne likenes he made him. \V But
by the enuie of the diuel, death entred into the world: \V and they
folow him that are of his part.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
