%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 1479
%%% o-1366
\startChapter[
  title={Chapter 01}
  ]

\Summary{Superiors
\MNote{The 1.~part.

An admonition to loue and practiſe iuſtice.}
are admonished to do iuſtice, ſincerely ſeking God. 7.~Who being euery
  where ſeeth al thinges. 11.~Murmuration, detraction, and lying bring
  to perdition. 13.~God created men to liue, but they brought death vpon
  themſelues.}

Loue
\CNote{\XRef{3.~Reg.~3.}
\XRef{Iſa.~56.}
\XRef{2.~Par.~15.}}
iuſtice, you that iudge the earth. Thincke of our Lord in goodnes, and
in ſimplicitie of hart ſeeke him: \V becauſe he is found of them, that
tempt him not: and he appeareth to them, that haue faith in him. \V For
\SNote{Mortal ſinnes are not only committed in dedes & vvordes, but alſo
in thoughtes.}
peruerſe cogitations ſeperate from God: and proued powre chaſteneth the
vnwiſe: \V becauſe
%%% 1480
wiſdom wil not enter into a malicious ſoule, nor dwel in a bodie
ſubiect to ſinnes. \V For the Holie Ghoſt of diſcipline, wil flie from
him that feyneth, and wil withdraw himſelfe from the cogitations, that
are without vnderſtanding, and he shal be chaſtened of iniquitie
enſewing. \V
\CNote{\XRef{Gal.~5. v.~22.}}
For the ſpirite of wiſdom is gentle, and wil not deliuer
\SNote{He that maliciouſly curſeth is curſed of God.}
the curſer from his lippes: becauſe God is witnes of his reynes, and he
is a true ſearcher of his hart, and an hearer of his tongue. \V Becauſe
the Spirite of our Lord hath replenished the whole world: and that which
contayneth al thinges, hath the knowledge of voice. \V For this cauſe he
that ſpeaketh vniuſt thinges, can not be hid, neither shal the
chaſtiſing iudgment paſſe him. \V For in the cogitations of the impious
there shal be examination: and the hearing of his workes shal come to
God, to the chaſtiſing of his iniquities. \V Becauſe the eare of
ielouſie heareth al thinges, and the tumult of murmurings shal not be
hid. \V Kepe your ſelues therfore from murmuring, which profiteth
nothing, and refraine your tongue from detraction, becauſe an obſcure
ſpeache shal not paſſe in vaine: and the mouth that lyeth, killeth the
ſoule. \V
\SNote{Be not cauſe of your owne eternal death by euil life.}
Zeale not death in the errour of your life, neither procure ye perdition
by the workes of your handes. \V Becauſe God made not death, neither
doth he reioyce in the perdition of the liuing. \V For he created al
thinges to be: and he made the nations of the earth to health: and
\SNote{Deſperate death deliuereth not the wicked from calamities.}
there is no medicine of deſtruction in them,
%%% o-1367
\SNote{Neither are anie damned vvhiles they are in this life.}
nor kingdome of hel in the earth. (\V For iuſtice is perpetual and
immortal.) \V But the impious with handes & wordes haue
\SNote{But ſinners not repenting bring eternal death to them ſelues.}
prouoked it: and eſteming it a freind, haue fallen to decay, and haue
made couenantes with it: becauſe they are worthie to be the part
therof.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
