%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 1489
%%% o-1374
\startChapter[
  title={Chapter 08}
  ]

\Summary{VViſdom excelling al thinges, that can be deſired, 9.~is
  worthely preferred as the cauſe of much eſtimation, 13.~and of
  immortal glorie, 16.~without moleſtation. 21.~Al which is Gods gift.}

She
\SNote{God, the increated wiſdom is infinite and wiſdom created is alſo
moſt excellent amongſt Gods giftes.}
reacheth therfore from end vnto end mightely, and diſpoſeth al thinges
ſwetely. \V Her haue I loued, and haue ſought her out from my youth, and
haue ſought to take her for my ſpouſe, and I was made a louer of her
beautie. \V
\CNote{\XRef{3.~Reg.~3.}}
She glorifieth her nobilitie, hauing conſociation with God: yea and the
Lord of al hath loued her. \V For she is the miſtreſſe of the diſcipline
of God, & the chooſer of his workes. \V And if riches be deſired in
life, what is richer then wiſdom, which worketh al thinges? \V And if
vnderſtanding doe worke: who is the worker of thoſe things that are,
more then she? \V And if a man loue iuſtice: her labours haue great
vertues: for she teacheth ſobrietie, and prudence, and iuſtice, and
ſtrength, then the which nothing is more profitable in life to men. \V
And if a man deſire multitude of
%%% o-1375
knowlege:
\CNote{\XRef{Prou.~3.}}
ſhe knoweth thinges paſt, & coniectureth of thinges to come: she knoweth
the ſubtilities of wordes, and the ſolution of arguments: she knoweth
ſignes and wonders before they be done, and the euentes of times and
ages. \V
\SNote{This is alſo the ſpeach of Salomon, recited by the writer of this
booke.}
I purpoſed therfore to bring her to me to liue together: knowing that
she wil communicate vnto me of good thinges, and wil be a comfort of
my cogitation & tediouſnes. \V I shal haue for her ſake
\SNote{Of Salomons wiſdom, riches, glorie & renowmed fame not only the
bookes of Kinges & Paralipomenon, but alſo our Sauiour doth witnes.
\XRef{Mat.~6.}
&
\XRef{12.}}
glorie with the multitudes, and honour with the ancient being yong: \V
and I shal be found sharpe in iudgement, and in the ſight of the mightie
I shal be meruelous, and the faces of princes wil meruel at me. \V
Holding my peace they shal expect me, and whiles I ſpeake manie wordes,
they shal lay their hands on their mouth. \V Moreouer by her I shal haue
\SNote{It is not certaine that Salomon hath immortal glorie, but rather
by \Emph{immortalitie} is here vnderſtood that his glorious fame
remaineth to the end of this world.}
immortalitie: and I shal leaue an eternal memorie to them, that shal be
after me. \V I shal diſpoſe peoples: and nations shal be ſubiect to
me. \V Horrible kings hearing shal feare me: in the multitude I shal
ſeme good, and in battel ſtrong. \V Entring into my houſe, I shal reſt
with her: for her conuerſation hath no bitternes, nor her companie
tediouſnes, but ioy and gladneſſe. \V Thinking theſe thinges with
myſelfe, and recording in my hart, that
\SNote{Neuertheles wiſdom of her part, geueth life and glorie
euerlaſting, to al that perſeuere to the end of this life.}
immortalitie is in the kindred of wiſedom, \V and good delectation in
her frendship, and in the workes of her
%%% 1490
handes honeſtie without defect, and wiſdom in the diſputation of her
talke, and glorie in the communication of her wordes: I went about
ſeeking, that I might take her to me. \V And I was a wittie childe, and
had gotten a good ſoule. \V And wheras I was more good, I came to
\SNote{It is certaine that Salomon was ſometime innocent and holie, but
was peruerted by wemen.
\XRef{3.~Reg.~11.}}
a bodie vndefiled. \V And as I knew that I could not otherwiſe be
continent, vnleſſe God gaue it, this verie thing alſo was wiſdom, to
know whoſe this gift was: I went to our Lord, and beſought him, and ſaid
from my whole hart:


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
