%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 1487
%%% o-1373
\startChapter[
  title={Chapter 07}
  ]

\Summary{Wheras
\MNote{The ſecond part.

VViſdom procedeth from God, and is procured by prayer.}
al men haue the like birth and death, 7.~Wiſdom maketh great difference,
bringing al goodnes, 13.~and knowlege, 17.~as wel of natural thinges,
22.~as moral. 25.~Which heauenlie gift is a ſparckle, and participation
of wiſdom increated, God himſelf.}

I alſo certes am a mortal man, like to al, and of the earthlie kinred of
him, that was made firſt, and in the wombe of my mother was I fashioned
flesh, \V the time of
\SNote{The perfecteſt children are borne in the beginning of the tenth
moneth.}
ten monethes was I brought together in bloud, of the ſeede of man, and
the delectation of
\SNote{Children in the mothers wombe are as in ſlepe.}
ſleepe concurring. \V And I being borne receiued the common ayre, and
fel vpon the earth, that is made alike, and the firſt voice like to al
men did I put forth weeping. \V I was nourished in ſwadling clothes, and
great cares. \V For none of the kinges had other begynning of
natuitie. \V There is one entrance therfore into life to al men, and
like departure. \V For this cauſe
\SNote{Salomon (whoſe ſayinges are here recited) praied for wiſdom, &
obtained it.
\XRef{3.~Reg.~3.}}
I wished, and vnderſtanding was geuen me: and I inuocated, and the
ſpirit of wiſdom came vpon me: \V and I preferred her before kingdoms and
thrones, and riches I counted to be nothing in compariſon of her. \V
Neither did I compare the precious ſtone to her: becauſe al gold in
compariſon of her, is a litle ſand, and ſiluer in the ſight of her shal
be eſtemed as clay. \V Aboue health and beautie did I loue her, and
purpoſed to haue her for light: becauſe her light can not be
extinguished. \V
\CNote{\XRef{3.~Reg.~3. v.~9.}}
And al good
%%% 1488
thinges came to me together with her, and very much honeſtie by her
handes, \V and I reioyced in al: becauſe
\SNote{God firſt gaue him grace to deſire wiſdom before al other
thinges, as he explicateth plainly.
\XRef{cap.~8.v.~21.}}
this wiſdom went before me, and I was ignorant that she is the mother of
al theſe. \V Which I lerned without fiction, and doe communicate without
enuie, and her honeſtie I hid not. \V For she is an infinite treaſure to
men: which who ſo haue, are made partakers of the
\Fix{frenship}{frendship}{obvious typo, same in both}
of God, commended for the gifts of diſcipline. \V And to me God hath
geuen to ſpeake according to my minde, and to preſume thinges worthie of
thoſe, that are geuen me: becauſe he is the guide of wiſdom, and the
creator of the wiſe: \V for in his hand are both we, & our wordes, and
wiſdom, and the knowlege and diſcipline of workes. \V For he gaue me
\SNote{Salomon was a moſt excellent Philoſopher.}
the true knowlege of thoſe
%%% o-1374
thinges, which are: that I may know the
diſpoſition of the round world, and the vertues of the elements, \V the
beginning, & end, & middes of times, the permutations of changeable
ſeaſons, and conſummations of times, \V the courſes of the yeare, and
diſpoſitions of the ſtarres, \V the natures of beaſtes, and furies of
wilde beaſtes, the force of windes, and the cogitations of men, the
differences of plantes, and vertues of rootes, \V and whatſoeuer are hid
thinges and not forſene, I haue lerned: for wiſdom the worker of al
taught me. \V For in her is the ſpirite of vnderſtanding,
\SNote{Proper Epithetons of the ſpirite of wiſdom.}
holie, onlie, manifold, ſubtil, eloquent, moueable, vndefiled, ſure,
ſwete, louing good, sharpe, who nothing hindereth wel doing, \V gentle,
benigne, ſtable, certaine, ſecure, hauing al powre, forſeing al thinges,
and that conteyneth al ſpirites: intelligible, cleane, ſubtile. \V For
wiſdom is more moueable then al moueable thinges: and reacheth euerie
where becauſe of her cleannes. \V For she
\SNote{See the
\XRef{Annotation, Prouerb.~1. v.~2.}}
is a vapour of the powre of God, & a certaine ſincere emanation of the
glorie of God omnipotent: and therfore no defiled thing cometh vnto
her. \V For she is the brightnes of eternal light, & the vnſpotted
glaſſe of Gods maieſtie, and
\CNote{\XRef{Heb.~1. v.~3.}}
the image of his goodnes. \V And wheras she is one, she can doe al
thinges: and permanent in herſelfe she reneweth al thinges, and by
nations tranſporteth herſelf into holie ſoules, she maketh the frendes
of God, and Prophetes. \V For God loueth none, but him, that dwelleth
with wiſdom. \V For she is more beautiful then the ſunne, and aboue al
diſpoſition of the ſtarres, being compared to light, she is found the
firſt. \V For night ſuccedeth to it, but malice ouercometh not wiſdom.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
