%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 1501
%%% o-1385
\startChapter[
  title={Chapter 16}
  ]

\Summary{God plaguing the Ægyptians for idolatrie, and crueltie,
  deliuered the Iſraelites. 5.~Chaſticed them alſo, but againe shewed
  them mercie, 20.~and fedde them with Manna.}

For
\SNote{The Ægyptians were plagued for their idolatrie.}
theſe thinges, and
\SNote{And that by beaſtes, becauſe they worshipped beaſtes for goddes:
and by death of their firſt begotten, for their crueltie againſt Gods
people.}
by the like to theſe, they haue worthely ſuffered torments, and were
deſtroyed by a multitude of beaſts. \V For the which torments thou didſt
wel diſpoſe of thy people, to whom thou gaueſt the deſire of their
delectation a new taſte, preparing them the quaile for meate: \V that
they in deede coueting meate, becauſe of thoſe thinges which were shewed
and ſent them, might be turned away euen from neceſſarie
concupiſcence. But they in short time being made needie, taſted a new
meate. \V For it behoued that without excuſe deſtruction should come
vpon them exerciſing tyrannie:
\SNote{God puniſhed his owne people as a father, for their amendment.}
but to theſe onlie to shew how their enemies were deſtroyed. \V For when
the cruel wrath of beaſtes came vpon them, they were deſtroyed with the
bytings of peruerſe ſerpents. \V Howbeit thy wrath endured not for euer,
but for chaſtiſement they were trubled a short time, hauing a ſigne of
ſaluation for the remembrance of the commandment of thy law. \V For he
that turned to it,
\SNote{The braſen ſerpent not by anie vertue inherent, but as a ſigne of
Gods fauoure, vvas the meanes of curing the people.
\XRef{Num.~21.}}
was not healed by that which he ſaw, but by thee the ſauiour of al: \V
and in this thou didſt shew to our enemies, that thou art he which
deliuereſt from al euil. \V For the bitings of locuſts, and flies killed
them, and there was found no remedie for their life: becauſe they were
worthie to be deſtroyed by ſuch thinges. \V But neither the teeth  of
venemous dragons ouercame thy children: for thy mercie coming healed
them. \V For in memorie of thy wordes they were examined, & were quickly
ſaued, leſt falling into deepe obliuion, they might not vſe thy
helpe. \V For neither herbe, nor playſter healed them, but thy word, ô
Lord, which healeth al thinges. \V For it is thou, ô Lord, that haſt
powre
%%% o-1386
of life and death, and bringeſt downe to
%%% 1502
the gates of death, and ſercheſt agayne: \V but man certes killeth by
malice, and when the ſpirit is gone forth, it shal not returne, neither
shal he cal backe the ſoule that is receiued: \V but it is vnpoſſible to
eſcape thy hand. \V For the impious denying to knowe thee, haue bene
ſcourged by the ſtrength of thine arme, ſuffering perſecution by ſtrange
waters, and haile, and rayne, and conſumed by fyre. \V For that which
was meruelous in water, which extinguisheth al thinges,
\SNote{VVith the plague of haile there vvas alſo fire mixed.
\XRef{Exo.~9. v.~24.}
VVhich deſtroyed the profitable cattle,}
fyre more preuayled: for the world is reuenger of the iuſt. \V For a
certayne time, the fyre was mitigated, that
\SNote{but burnt not other beaſtes, that plagued the Ægyptians.}
the beaſts which were ſent to the impious might not be burnt, but that they
ſeing might know that by Gods iudgement they ſuffer punishment. \V And
at a certaine time the fyre aboue his powre burnt
\SNote{Haile did not extinguiſh the fire, by Gods povvre aboue nature.}
in water on euerie ſide, that it might deſtroy the nation of a wicked
land. \V For the which thinges thou didſt nourish thy people with the
meate of Angels, and bread prepared thou gaueſt them from heauen without
labour,
\SNote{See the miracles of Manna.
\XRef{Annot. Exod.~16.}}
hauing in it al delectation, and the ſweetnes of al taſte. \V For thy
ſubſtance did shew thy ſweetnes which thou haſt toward thy children, and
ſeruing euerie mans wil, it was turned to that, that euerie man
would. \V Yea ſnow and yce ſuſteyned the force of fyre, and melted not:
that they might know that fyre burning in hayle, and lightening in rayne
deſtroyed the fruites of the enemies. \V And this againe, that the iuſt
might be nourished, it forgat alſo his owne ſtrength. \V For the
creature ſeruing thee the Creatour, is fierce into torment againſt the
vniuſt: and is made more gentle to doe good for them, that truſt in
thee. \V For this cauſe euen then being transformed into al thinges they
ſerued thy grace the nource of al, at their wil that deſired thee: \V
that thy children might know, whom thou loueſt ̂ô Lord, that not the
fruites of natiuitie doe feede men, but thy word preſerueth them, that
beleue in thee. \V For that which could not be deſtroyed by fyre,
forthwith being
\SNote{He ſpeaketh againe of Manna.}
heated with a litle beame of the ſunne did melt: \V that it might be
knowen to al men, that we ought to preuent the ſunne to bleſſe thee, and
at the ryſing of light to adore thee. \V For the
\SNote{The vaine imagination of the vvicked that himſelf shal be ſaued
vvil faile him.}
fayth of the vngratful shal melt as winter yce, & shal perish as
vnprofitable water.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
