%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 1494
%%% o-1379
\startChapter[
  title={Chapter 12}
  ]

\Summary{Gods wiſdom and mercie in deſtroying the wicked inhabitants of
  Chanaan, by parts (10.~that they might haue amended,) whom he could
  haue ſlaine ſudenly. 15.~In that God neuer condemneth the iuſt,
  19.~his people are inſtructed to confide in him, 25.~and ſinners to
  turne vnto him.}

O how good, and ſweete is thy ſpirit ô Lord in al! \V And therfore
thoſe, that erre, by partes thou doeſt chaſtiſe: and doeſt admonish, and
ſpeake to them, concerning the thinges wherin they ſinne: that leauing
naughtines, they may beleue in thee ô Lord. \V For, thoſe old
inhabitantes of thy holie land, whom thou didſt abhorre, \V becauſe they
did workes odious to thee by ſorceries, and vniuſt ſacrifices, \V and
the murderers of their owne children without mercie, and eaters of mens
bowels, and deuourers of bloud from the middes
\SNote{From the land of Iurie called ſacred becauſe God was there rightly
ſerued in the old teſtament, and mans redemption vvas wrought there by
Chriſt.}
of thy ſacrament, \V and the parents authors of aydeleſſe ſoules, thou
wouldſt deſtroy by the handes of our parents, \V that they might receiue
a peregrination worthie of the children of God, which is a land of al
moſt deare to thee. \V But them alſo as men thou didſt ſpare, and didſt
ſend forerunners of thine hoſt, waſpes, that by litle and litle they
might deſtroy them. \V Not becauſe thou waſt vnable in battel to ſubdewe
the impious to the iuſt, or with cruel beaſtes, or with a sharpe word to
deſtroy them together: \V
\CNote{\XRef{Exo.~23.}
\XRef{Deut.~7.}}
but iudging by partes thou gaueſt place of repentance, being not
ignorant, that the nation of them is wicked, and their malice
\SNote{By cuſtom malice became as it were natural, after that nature was
corrupted.}
natural, & that their cogitation could not be changed for euer. \V For
it was a curſed ſeede from the begynning: neither fearing any, didſt
thou geue pardon to their ſinnes. \V For who shal ſay to thee: what haſt
thou done? or who shal ſtand againſt thy iudgement? or who in thy ſight
shal come reuenger of the wicked men? or who shal impute it to thee, if
the nations perish, which thou haſt made? \V For there is no other God
but thou, who haſt care of al, that thou mayſt shew that thou doeſt not
geue iudgement vniuſtly. \V Neither king, nor tyrant in thy ſight shal
enquire of them, whom thou haſt deſtroyed. \V For ſo much then as thou
art iuſt, thou
%%% 1495
doeſt diſpoſe al thinges iuſtly: thou alſo eſtemeſt it diſagreable from
thy powre, to condemne him, who ought not to be punished. \V For thy
powre is the begynning of iuſtice: and for this that thou art Lord of
al, thou makeſt thyſelf to ſpare al. \V For thou sheweſt powre, which
art not thought to be abſolute in powre, and thou conuinceſt the boldnes
of them, that know thee not. \V But thou dominatour of powre, iudgeſt
with tranquilitie, and with great reuerence diſpoſeſt of vs: for
\SNote{Gods powre being almightie is only limited by his vvil.}
it is in thy powre when thou wilt, to be able. \V And thou haſt taught
thy people by ſuch workes, that they muſt be iuſt and gentle, and haſt
%%% o-1380
made thy children of good hope: becauſe iudging thou geueſt in ſinnes
place of repentance. \V For if thou didſt punish the enemies of thy
ſeruants, and that deſerued to die, with ſo great conſideration, geuing
time and place, whereby they might be changed from their wickednes. \V
With what diligence haſt thou iudged thy children, to whoſe parents thou
gaueſt othes and couenants of good promiſes? \V Therfore when thou
geueſt vs diſcipline, thou ſcourgeſt our enemies very manie wayes, that
iudging we may thinke vpon thy goodnes: and when we are iudged, we may
hope for thy mercie. \V Wherfore to them alſo, which in their life haue
liued foolishly & vniuſtly, thou haſt geuen great torments by the ſame
thinges, which they did worshipe. \V 
\CNote{\XRef{Rom.~1. v.~23.}}
For they wandered long in the way of errour, eſteming for goddes thoſe
thinges, that in beaſts
\SNote{Serpents, battes, moles, & like beaſtes, which ſeme not only
ſuperfluous in the world, but alſo hurtful, yet were eſtemed as goddes.}
are ſuperfluous, liuing after the maner of ſenſles infants. \V For this
cauſe thou haſt geuen iudgement on them as on ſenſles children to be in
deriſion. \V But they that were not amended by ſcornes and
reprehenſions, haue tried the worthie iudgement of God. \V For in what
thinges they ſuffering tooke indignation, by thoſe whom they thought
goddes, when ſeing they were deſtroyed in them, him, whom in time paſt
they denied that they knewe, they acknowleged the true God:
\SNote{Knowing him to be the only true God, by vvhom they ſaw their
falſe goddes deſtroyed, yet they did not ſerue him as God.}
for the which cauſe the end alſo of their condemnation shal come vpon
them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
