%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 1506
%%% o-1389
\startChapter[
  title={Chapter 19}
  ]

\Summary{Ægyptians perſecuting the Hebrewes were drowned, 10.~hauing
  bene plagued before with flies, and frogges. 11.~Quailes were geuen to
  the people of Iſrael. 13.~The barbarous not receiuing, or euil
  intreating Gods people, were ſtrooken with blindnes. 17.~And al
  creatures ſerue God in punishing the impious, and rewarding the
  godlie.}

But vpon the impious euen to the later end there came wrath without
mercie. For
\SNote{God foreſeing the Ægyptians malice, permitted them to perſecute
his people, but was no way the cauſe nor auctor of their ſinne.}
he foreknew alſo the thinges that should come vnto them: \V
\CNote{\XRef{Exo.~14.}}
becauſe when they had permitted that they should depart, and had ſent
them away with great diligence, they repenting purſued them.
%%% Verse marker missing in this, but number appears in margin.
\V For hauing as yet moorning betwen their hands, and lamenting at the
graues of the dead, they tooke to themſelues an other cogitation of
follie: and whom by intreating they had caſt forth, them they
purſued as fugitiues: \V for worthie neceſſitie brought them to this end:
and they loſt the remembrance of thoſe thinges, which had chanced, that
punishment might fulfil the thinges that wanted to the torments: \V and
that thy people certes might paſſe through meruelouſly, but they might
finde a new death. \V
%%% o-1390
For euerie creature according to his kind was fashioned agayne from the
begyning, ſeruing thy precepts, that thy children might be kept without
hurt. \V For a clowde ouer shadowed their campe, and out of the water
which was before, there appeared drie land, and in the redſea a way
without impediment, and of the great depth a ſpringing filde: \V through
the which al the nation paſſed, which was protected with thy hand, ſeing
thy meruelous thinges and wonders. \V For euen as horſes they fed on
meate, and as lambes they reioyced, magnifying thee ô Lord, which didſt
deliuer them. \V For they were mindful of thoſe thinges, which had bene
done in their ſeiourning, how for the nation of beaſts the earth
brought forth flies, and for fishes the riuer yelded a multitude of
frogges. \V And laſt of al they ſaw a new creature of birdes, when
allured by concupiſcence they deſired meates of deliciouſnes. \V
\CNote{\XRef{Exo.~16.}}
For in comfort of their deſire, there came vp to them the quaile from the
ſea: and vexations came vpon the ſinners, not without thoſe arguments,
which were made before by the force of lightninges: for they
ſuffered iuſtly 
\Fix{occording}{according}{obvious typo, fixed in other}
to their wickednes.
%%% 1507
For they inſtituted a more deteſtable inhoſpitalitie:
\SNote{The Amorrheites refuſed to grant them paſſage.
\XRef{Num.~21. v.~21.}
The Ægyptians brought them into ſeruitude.
\XRef{Exo.~1.}}
ſome certes receiued not the vnknowen ſtrangers, and other ſome brought
the good ſtrangers into ſeruitude. \V And not onlie theſe thinges, but
in deede there was an other reſpect alſo of them: for they againſt their
wil receiued the ſtrangers. \V But they that receiued them with gladnes,
did afflict them with moſt cruel ſorowes, that vſed the ſame rightes: \V
but they were ſtriken with blindnes: as they in the dores of the iuſt,
when they were couered with ſodaine darkenes, euerie man ſought the
paſſage of his doore. \V For whiles the
\SNote{God changing the natural properties of elementes, by them wrought
iuſtice on ſinners. S.~Greg.
\Cite{ho.~55. in Euang.}}
elements are turned in themſelues, as in an inſtrument the ſound of the
qualitie is changes, and al keepe their ſound: wherfore it may be
certainly iudged by the very ſight. \V For the thinges of the fild were
turned into thinges of the water: and what ſoeuer were ſwimming thinges
paſſed into the land. \V
\CNote{\XRef{Exo.~9. v.~24.}}
The fyre had force in water aboue his powre, and the water forgot her
quenching nature. \V On the contrarie the flames vexed not the flesh of
corruptible beaſts walking therwith, 
\CNote{\XRef{Exo.~19. v.~21.}}
neither did they melt that good meate, which was eaſely diſſolued euen
as yce. For in al thinges thou didſt magnifie thy people ô Lord, and
didſt honour them, and didſt not deſpiſe them, at al time, and in
euerie place aſſiſting them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
