%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 1495
%%% o-1380
\startChapter[
  title={Chapter 13}
  ]

\Summary{Men folowing their phantaſies knew not God by his creatures,
  but honored the creatures for goddes. 10.~Moſt ſottishly alſo
  worshipped thinges fashioned by mens handes, as goddes.}

%%% 1496
But al men be vaine, in whom there is not the knowlege of God:
\CNote{\XRef{Rom.~1.}}
and of theſe good thinges, which are ſene
%%% !!! Marked as an SNote in this, unmarked in other
\LNote{They could not vnderſtand Him, that is.}{Philoſophers
\MNote{God is the beginning of al thinges, abſolute and independing.}
diſcuſſing the nature of manie creatures, ſaw that euerie creature
proceded of ſome other thing, & ſo there muſt nedes be one beginning of
al, abſolute of itſelfe, neither proceding nor depending of an other,
nor a limited ſubſtance. As the thing that is a man, is not a beaſt; an
oxe is not a horſe: &c. and telling what anie thing is, we exclude
therby that it is not other thinges, but ſaying without addition,
\Sc{He which is}, we shew the beginning of al, in no ſorte limited: and
this is God. VVhom ſome Philoſophers, by ſuch diſcourſe found, & knew,
in general, & ſometimes confeſſed, but did not honour him as God, and
therfore were inexcuſable, as S.~Paul concludeth againſt them.
\XRef{Rom.~1.}}
they could not vnderſtand Him
\SNote{Gods moſt proper name is \Sc{He which is}.
\XRef{Exo.~3. v.~14.}}
that is, neither attending to the workes haue they agniſed who was the
workeman: \V
\CNote{\XRef{Deut.~4.}}
but either the fyre, or the wind, or the ſwift ayre, or a circle of
ſtarres, or exceding much water, or the ſunne and the moone, they
thought to be goddes rulers of the world. \V With whoſe beautie if being
delighted, they thought them goddes: let them know how much the Lord of
them is more beautiful then they. For the author of beautie made al
thoſe thinges. \V Or if they merueled at their vertue, & operations, let
them vnderſtand by them, that he which made theſe, is ſtronger then
they: \V for by the greatnes of the beautie, and of the creature the
creator of them may be ſene, to be knowen therby. \V But notwithſtanding
there is yet in theſe leſſe complainte. For they alſo perhaps erre,
ſeeking God, and deſirous to finde him. \V For whereas they conuerſe in
his workes, they
%%% o-1381
enquire: & they are perſwaded that the thinges be good which are
ſeene. \V But againe neither ought theſe to be pardoned. \V For if they
could know ſo much, that they were able to eſtimate the world: how did
they not more eaſely find the Lord therof? \V But they are vnhappie, &
their hope is among the dead, who haue called
\SNote{Seing no creature how excellent ſoeuer, is or may be eſtemed a
god, it is more fooliſh to thinke, an Image, or \L{ſtatua}, or anie thing
framed by mans handes can be God.}
the workes of mens handes goddes, gold & ſiluer, the inuention of art,
and the ſimilitudes of beaſtes, or an vnprofitable ſtone the worke of an
old hand. \V 
\CNote{\XRef{Iſa.~4.}}
Or if an artificer a carpenter, cut ſtreight timber out of the wood, &
pare of al the barke therof cunningly, and vſing his art,
\CNote{\XRef{Ier.~10.}}
diligently frameth a veſſel profitable for the common vſe of this
life, \V and vſeth the chippes of that worke to dreſſe his meate: \V and
maketh that which is left therof, which is for no vſes, but being a
crooked peece of wood, and ful of knobes, carueth it diligently in the
holownes therof, and by the skil of his art fashioneth it, and maketh it
like to the image of a man, \V or compareth it to ſome beaſt, ſtraking it
ouer with redde, and with paynting making the colour therof ruddie, and
layeth  a colour ouer euerie ſpot that is in it: \V and maketh a worthie
habitation for it, and ſetting it in a wal, and faſtening it with
yron, \V leſt perhaps it fal, prouiding for it, knowing that it can not
helpe it ſelfe: for it is an image, and it nedeth helpe. \V And
concerning his ſubſtance, & his children, & for mariage, making a vow he
ſeeketh to it. He is not ashamed to
%%% 1497
ſpeake with him, that is without ſoule: \V and for health certes he
beſecheth the weake, and for life asketh the dead, and for helpe
inuocateth him that is vnprofitable: \V and for a iourney asketh him,
that can not walke: and for getting, and for working, and for the euent
of al thinges he asketh him, that in al is vnprofitable.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
