%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 1499
%%% o-1384
\startChapter[
  title={Chapter 15}
  ]

\Summary{The wiſe gratfully praiſe the ſwetenes, and mercie of God, by
  whom they are deliuered from idolatrie: 6.~deteſting the makers, &
  worshippers of idols.}

%%% 1500
But ô thou our God, art ſweete, and true, patient, and diſpoſing al
thinges in mercie. \V For if we ſinne, we are thine, knowing thy
greatnes: and if we ſinne not we know that we are counted with thee. \V
For to know thee, is abſolute iuſtice: and to know iuſtice, and thy
powre, is the roote of immortalitie. \V For mens inuention of euil art
hath not brought vs into errour, nor the shadow of a picture being
labour without fruite, a shape grauen by diuerſe colours, \V the ſight
wherof geueth concupiſcence to the ſenſles,
\Fix{and and}{and}{obvious typo, fixed in other}
he loueth the shape without life of a deade image. \V The louers of
euils, are worthie to haue their hope in ſuch thinges, both they that
make them, and that loue, and that worshippe them. \V Yea and the potter
preſſing ſofte earth, with labour fashioneth euerie veſſel to our vſes,
and of the ſame clay maketh the veſſels, that are cleane to vſe, and in
like maner them, that are contrarie to theſe: but what the vſe of theſe
veſſels is, the potter is iudge. \V And with vaine labour he
\SNote{Of the diuers ſortes of idols and idolaters ſee our brife
\XRef{Annotation vpon the 113.~Pſal.}}
fashioneth a god of the ſame clay: he which a litle before was made of
earth, and a litle after returneth backe, whence he was taken, being
exacted the debte of the life which he had. \V But his care is, not
becauſe he shal labour, nor becauſe he hath a short life, but he
contendeth with gold ſmithes, and ſiluer ſmithes: yea and he imitateth
the copper ſmithes, and counteth it a glorie, becauſe he maketh vaine
thinges. \V For his hart is ashes, and his hope vaine earth, and his
life viler then clay: \V becauſe he was ignorant who made him, and
who inſpired into him the ſoule which worketh, and who breathed into
him the vital ſpirite. \V Yea and
\SNote{Idolaters hauing forſaken and forgot the onlie true God, become
as Atheiſts, making their temporal gaine of falſe goddes.}
they eſtemed our life to be a paſtime, and the conuerſation of life made
for a gayne; & that we muſt get euerie way euen of euil. \V For he
knoweth that he offendeth aboue al men, which of the matter of earth
fashioneth frayle veſſels, and ſculptils. \V For al the vnwiſe, and
vnhappie aboue meaſure of the foole, proude
\SNote{And ſo waxing inſolent, contemne and perſecute the ſeruants of
God.}
are the enemies of thy people, and rule ouer them: \V becauſe they haue
eſtemed al the idols
%%% o-1385
of the nations for goddes, which neither haue vſe of eies to ſee, nor
noſthrels to take breath, nor eares to heare, nor fingers of the hands
to handle, yea and their feete are ſlow to walke. \V For a man made
them: and he that borowed breath, the ſame fashioned them. For no man
can make God like to himſelf. \V For wheras himſelf is mortal, he
maketh a dead thing with his wicked handes. For
%%% 1501
he is better then they, whom he worshippeth, becauſe he in deede liued,
though he were mortal, but they neuer. \V But
\SNote{Some idolaters worſhipped brute beaſtes for goddes, as being
better then ſenſles images: but al are abominable.}
they worship alſo moſt miſerable beaſts: for the ſenſleſſe thinges
compared to theſe, are worſe then they. \V Yea neither by ſight can any
man ſee good of theſe beaſts. But they haue fled from the prayſe of God,
and from his bleſſing.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
