%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 1492
%%% o-1377
\startChapter[
  title={Chapter 11}
  ]

\Summary{Other benefites of wiſdom, protecting the Iſraelites in the
  deſert, 3.~ouerthrowing their enimies, 4.~geuing them water out of a
  rocke, 8.~plaguing the Ægyptians, 21.~yet not al ſudenly, but by often
  admonitions, that they might haue repented if they would.}

She
\CNote{\XRef{Exo.~16.}}
directed their workes in the handes of
\SNote{Moyſes.}
a holy prophet. \V
\CNote{\XRef{Exo.~27.}}
They made a iourney through the deſerts, that were not inhabited: and in
deſert places they pitched cottages. \V
%%% !!! This CNote was originally part of the next SNote.
\CNote{\XRef{Exo.~17.}}
They ſtood againſt
\SNote{The Amalechites.}
the aduerſaries, and reuenged themſelues of the enemies. \V
\CNote{\XRef{Num.~20.}}
They thirſted, and inuocated thee, and water was geuen them out of a
moſt high
\Fix{rorcke}{rocke}{obvious typo, fixed in other}
For by the thinges wherby their enemies
%%% o-1378
ſuffered punishment, for defect of their drinke, and therein, when the
children of Iſrael abunded, they did reioyce; \V by theſe thinges, when
others lacked the ſame, it
%%% 1493
went wel with them. \V For in ſteede of the fountaine of an euerlaſting
riuer, thou gaueſt mans bloud to the vniuſt. \V Who when they were
diminished in the deſtruction of the murdered infants, thou gaueſt them
\Fix{abundant}{abundant water}{obvious typo, fixed in other}
vnlooked for: \V shewing by the thirſt, that then was, how thou didſt
exalt thine, & didſt kil their aduerſaries. \V For when
\SNote{VVhen the Iſraelites wanted water God gaue them abundance out of
rockes,}
they were tempted, and in deede with mercie taking diſcipline, they knew
how
\SNote{but turned the Ægyptians water into blood.}
the impious being iudged with wrath did ſuffer torments. \V Theſe certes
admonishing as a father thou didſt proue: but them examining as a hard
king thou didſt condemne. \V For the abſent and the preſent were
tormented alike. \V For duble tediouſnes had taken them, and ſighing
with the memorie of good thinges paſt. \V For when they vnderſtood
\SNote{After affliction the benefite of peace is more gratful.}
by their punishement that it went wel with them, they remembred our
Lord, merueling at the end of the euent. \V  For
\SNote{Moyſes was reiected when he iudged betwen his bretheren
\XRef{Exo.~2. v.~14.}
but was afterwards the deliuer of the whole people.
\XRef{Act.~7. v.~45.}}
whom before they derided, being caſt forth in that wicked laying out to
perish, him they merueled at in the end of the euent: not thirſting in
like maner to the iuſt: \V but for ſenſeles cogitations of their
iniquitie, for that
\SNote{Ægyptians ſeruing beaſts for goddes, were plaged by frogges,
ſiniphes, flees, and locuſtes.}
ſome erring did worshippe dumme ſerpents, and ſuperfluous beaſts, thou
didſt ſend vpon them a multitude of dumme beaſts for reuenge: \V that
they might know that by what thinges a man ſinneth, by the ſame alſo he
is tormented. \V 
\CNote{\XRef{Leuit.~26. v.~22.}}
For thine omnipotent hand, which made the world of inuiſible matter, was
not vnable to ſend vpon them a multitude of beares, or fierce lyons, \V
\CNote{\XRef{Iere.~8. v.~17.}}
or vnknowen beaſts ful of anger of a new kind; or breathing the vapour
of fires, or caſting forth the ſauour of ſmoke, or shooting horrible
ſparkes from their eies: \V of which not onlie their hurt was able to
deſtroy them, but alſo their ſight to kil them for feare. \V Yea and
without theſe with one ſpirit, they might haue beene ſlaine ſuffering
perſecution of their owne factes, and diſperſed by the ſpirit of thy
powre: but thou haſt diſpoſed al thinges in meaſure, and number, and
weight. \V For, to be of great force reſted alwaies in thee onlie: & who
shal reſiſt the powre of thyne arme? \V Becauſe as the leaſt weight of
the balance, ſo is the round world before thee, and as a droppe of the
dewe before day, that falleth vpon the earth. \V But thou haſt mercie on
al, becauſe thou canſt do al thinges, and diſſembleſt the ſinnes of men
for repentance. \V For thou
\SNote{God made no creature euil as the Manichees fooliſhly imagined,
neither is there any God but one who alone created al thinges.}
loueſt al thinges that are, and hateſt nothing of thoſe which thou haſt
made: for thou didſt not ordaine, or
%%% 1494
make any thing hating it. \V And how could any thing continew, vnles
thou wouldeſt? or be preſerued which was not called of thee? \V But thou
ſpareſt al: becauſe they are thine ô Lord, which loueſt ſoules.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
