%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Wisdom
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 1490
%%% o-1375
\startChapter[
  title={Chapter 09}
  ]

\Summary{A prayer (made by Salomon) for wiſdom, 9.~wherby Superiors are
  able to gouerne: 13.~which by only humane wiſdom, they can not rightly
  performe.}

God of my fathers, and Lord of mercie, which madeſt al thinges with thy
word, \V
\CNote{\XRef{3.~Reg.~3. v.~9.~12.}}
and by thy wiſdom didſt appoint man, that he should haue dominion of the
creature, that was made by thee, \V that he should diſpoſe the round
world in equitie and iuſtice, and execute iudgement in direction of
hart: \V geue me wiſdom the aſſiſtant of thy ſeates, and repel me not
from thy children: \V becauſe I am thy ſeruant, and the ſonne of thy
handmaide, a weake man, and of ſmal time, and leſſe to the vnderſtanding
of iudgement and lawes.
%%% o-1376
\V And if one be perfect among the children of men, and thy wiſdom be
abſent from him, he shal be counted for nothing. \V
\CNote{\XRef{1.~Paral.~28. v.~5.}}
\SNote{Here againe it is euident, that the Auctor reporteth Salomons
ſpeaches.}
Thou haſt choſen me king to thy people, and iudge of thy ſonnes and
daughters: \V 
\CNote{\XRef{2.~Par.~1. v.~9.}}
and badſt me build a temple in thy holie mount, and an altar in the
citie of thy habitation, a ſimilitude of thy holie tabernacle, which
thou didſt prepare from the beginning: \V 
\CNote{\XRef{Prou.~8. v.~22.}}
and
\SNote{VViſdom increated is with God, yea is God him ſelfe.}
thy wiſdom with thee, which knew thy workes, which then alſo was preſent
when thou madeſt the round world, and knew what was pleaſing to thyne
eyes, and what was direct in thy precepts. \V
\SNote{VViſdom vvhich is geuen to men procedeth from God as a gift
created.}
Send her from thy holy heauens, and from the ſeate of thy greatnes, that
she may be with me, and may labour with me, that I may know what is
acceptable with thee: \V for she knoweth al thinges, & vnderſtandeth,
and shal conduct me in my workes ſoberly, & shal keepe me with her
might. \V And my workes shal be acceptable, and I shal gouerne thy
people iuſtly, and shal be worthie of the ſeates of my father. \V For
\SNote{Mans wiſdom vvithout ſpecial wiſdom from God is not ſufficient to
gouerne ourſelues, much leſſe others.}
who of
%%% 1491
men is able to know the counſel of God? or who can thinke what God
wil? \V For the cogitations of mortal men be fearful, and our
prouidences vncertaine. \V For the bodie, that is corrupted burdeneth
the ſoule, and the earthlie habitation preſſeth downe the vnderſtanding
that thinketh manie thinges. \V And we doe hardly coniecture the thinges
that are in the earth: and the thinges that are in ſight, we finde with
labour. But the thinges that are in the heauens who shal ſearch out? \V
And thy ſenſe who shal know, vnles thou geue wiſdom, and ſend thy holie
ſpirit from on high: \V and ſo the pathes of them, that are on the earth
may be corrected, and men lerne the thinges that pleaſe thee? \V For by
wiſdom they were healed,
\Fix{whoſeouer}{whoſoeuer}{obvious typo, fixed in other}
haue pleaſed thee ô Lord from the begynning.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
