%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Isaia
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 1596
%%% o-1470
\startChapter[
  title={Chapter 07}
  ]

\Summary{Ieruſalem being beſeeged, 3.~the prophet telleth the king, that
  the enemies shal not preuaile (8.~but the kingdome of Iſrael shal be
  deſtroyed) 10.~geueth for a ſigne that a Virgin shal conceiue and bare
  a Sonne. 17.~Prophecieth alſo the captiuitie, and deſolation of the
  kingdom of Iuda.}


And
\CNote{\Xref{4.~Reg.~16.}}
it came to paſſe
\SNote{Before this, the kinges of Syria and of Iſrael had taken king
Achaz in battel and caried avvay great ſpoyles.
\XRef{2.~Paral.~28.}
But preſuming, to do the like againe, God ſuffered them not to preuaile.

Myſtically this ſignified that heretikes of diuers ſectes conſpire
together, to impugne the Catholike Church. VVhich they do much afflict,
and terrifie, but can neuer ouerthrow it.
\Cite{S.~Ierom in hunc locum.}}
in the dayes of Achaz the ſonne of Ioathan, the ſonne of Ozias king of
Iuda, there came vp Raſin the king of Syria, and Phacee the ſonne of
Romelia the king of Iſrael to Ieruſalem, to fight againſt it: and they
could not ouercome it. \V And they told the houſe of Dauid, ſaying:
Syria hath reſted vpon Ephraim, & his hart was moued, and the hart of
his people, as the trees of the
\Fix{wooddes}{woodes}{obvious typo, fixed in other}
are moued at the face of the winde. \V And our Lord ſayd to Iſaie: Goe
forth to meete Achaz, thou, and Iaſub thy ſonne that is leaft, to the
conduite of the vpper poole, in the way of the Fullers filde. \V And
thou shalt ſay to him: See thou be ſtil: feare not, and let not thine
hart be afrayd of the two tailes of theſe ſmoking fyrebrands, in the
wrath of the furie of Raſin the king of Syria, and of the ſonne of
Romelia.
%%% o-1471
\V For that Syria hath taken counſel againſt thee, vnto the euil of
Ephraim, and the ſonne of Romelia, ſaying: \V Let vs goe vp to Iuda, and
rayſe it vp, and plucke it away to vs, and make the ſonne of Tabeel king
in the middes therof. \V Thus ſayth our Lord God: It shal not ſtand, &
this shal not be. \V But the head of Syria Damaſcus, & the head of
Damaſcus Raſin: and yet threeſcore and fiue yeares, and Ephraim shal
ceaſe to be a people. \V And the head of Ephraim Samaria, and the head
of Samaria the ſonne of Romelia. If you wil not beleue, you shal not be
permanent. \V And our Lord added to ſpeake to Achaz, ſaying: \V Aske a
ſigne for thee of the Lord thy God vnto the depth of hel, or vnto the
height aboue. \V And Achaz
%%% 1597
ſaid: I wil not aske, and
\SNote{Though Achaz vvas very vvicked and committed idolatrie,
\XRef{4.~Reg.~16.}
\XRef{2.~Par.~28.}
yet he beleued in God Almightie, knovving that he ought not to tempt
him.}
I wil not tempte our Lord. \V And he ſaid: Heare ye therfore ô houſe of
Dauid: Is it a ſmal thing for you, to be greuous to men, that your are
greuous to my God alſo? \V Therfore shal our Lord himſelfe
\SNote{Vpon occaſion of Gods mercie promiſed vvithout mans deſert, which
king Achaz hardly beleued, to confirme the ſame with a farre greatter
example, God inſpired the Prophet alſo to forſhevv the greater myſterie
of Chriſts Incarnation, his conception & birth of a virgin, for the
redemption of al mankind.}
geue you a ſigne. Behold a virgin shal conceiue, and beare a ſonne, &
his name shal be called Emmanuel. \V He shal eate butter and honie, that
he may know to refuſe euil, and chooſe the good. \V For before the
childe know to refuſe euil, and chooſe good, the land which thou doeſt
deteſt shal be forſaken of the face of her two kinges. \V Our Lord wil
bring vpon thee, & vpon thy people, and vpon the houſe of thy father,
daies that came not ſince the daies of the ſeparation of Ephraim from
Iuda, with the king of the Aſſirians. \V And it shal be in that day: our
Lord shal hiſſe to the flie, that is in the vtter moſt part of the
riuers of Ægypt, and to the bee, that is in the land of Aſſur. \V And
they shal come, and shal reſt al in the torrentes of vallies, and in the
caues of rockes, and in al shrubbe places, and in al holes. \V 
\CNote{\XRef{4.~Reg.~19.}}
In that day our Lord shal shaue with a raſer, hyred by them, that are
beyond the riuer, by the king of the Aſſirians, the head, and the haires
of the feete, and the whole beard. \V And it shal be in that day: a man
shal nourish a young cowe, and two ewes. \V And for the abundance of
milke he shal eate butter: for butter and honie shal euerie one eate,
that shal be leaft in the middes of the land. \V And it shal be in that
day: euerie place where there shal be a thouſand vines, for a thouſand
peeces of ſiluer, they shal be into thornes and bryers. \V With arrowes
and bow they shal goe in thither: for bryers and thornes shal be in al
the land. \V And al mountaines, that shal be weeded with a weeding
hooke, the terrour of thornes and bryers shal not come thither, and it
shal be for the oxe to feede on, and cattle to treade vpon.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
