%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Isaia
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 1591
%%% o-1466
\startChapter[
  title={Chapter 03}
  ]

\Summary{The Iewes shal be depriued of wiſe men, 4.~and be ſubiect to
  childish, and effeminate gouerners, 8.~for their greuous
  ſinnes. 16.~The proud, curious, and laſciuious attyre of their wemen,
  24.~shal be turned into ignominie, and ſorow.}


For behold the dominatour the Lord of hoſtes
\SNote{This was fulfilled firſt in the captiuitie of Babylon and more
notoriouſly after Chriſts paſſion, in the deſtruction of Ieruſalem, and
diſperſion of the Iewes: euen to this day: and yet forward, til nere the
end of this world.}
shal take away from Ieruſalem, and from Iuda the valiant and the ſtrong,
al ſtrength of bread, and al ſtrength of water. \V The ſtrong, and the
man of warre, the iudge, and the prophete, and ſouthſayer, and the
ancient. \V The prince ouer fiftie, and the honorable of countenance,
and the counſeler, and the wiſe of workemaſters, and the skilful of
myſtical ſpeach. \V And I wil geue children to be their princes, and
the effeminate shal rule ouer them. \V And the people shal rush
violently, man againſt man, and euerie one againſt his neighbour: the
childe shal make tumult againſt the ancient, and the baſe againſt the
noble. \V For a man shal take hold of his brother, one of the houſe of
his father: Thou haſt a garment, be thou our prince, and let this ruine
be vnder thy hand. \V He shal anſwer in that day, ſaying: I am no
phyſicion, & in my houſe there is no bread, nor garment: do not appoint
me prince of the people. \V For Ieruſalem is gone to ruine, and Iuda is
fallen: becauſe their tongue, & their inuentions were againſt our Lord,
to prouoke the eyes of his maieſtie. \V The knowlege of their face hath
anſwered them: and they haue proclaimed their ſinne as Sodom, neither
haue they hid it: woe to their
%%% 1592
ſoule, becauſe euils are rendered to them. \V Say to the iuſt that it is
wel, becauſe he shal eate the fruite of his inuentions. \V Woe to the
impious vnto euil: for the reward of his handes shal be made to him. \V
My people their exactours haue ſpoyled, & wemen haue ruled ouer them. My
people,
\SNote{An Eccleſiaſtical preacher muſt not flatter the people. He muſt
moue teares (ſayth S.~Ierom) not laughter.}
they that cal thee bleſſed, the ſame deceiue thee, and diſſipate the way
of thy ſteppes. \V Our Lord ſtandeth to iudge, and he ſtandeth to iudge
peoples. \V
\CNote{\XRef{Apoc.~1.}}
Our Lord shal come to iudgement with the ancients of his people, and his
princes: for you haue deuoured the vineyard, and the ſpoile of the poore
is in your houſe. \V Why do you conſume my people, and grinde the faces
of the poore, ſayth our Lord the God of hoſtes? \V And our Lord ſayd:
For that
\SNote{By the metaphor of wemen, S.~Ierom vnderſtandeth the cities of
Iurie, of which Ieruſalem was the head, and Sion the chiefe place
thereof, al which were defaced, by the Babylonians, but more fully
deſtroyed by Titus and Veſpaſian, fourtie yeares after Chriſts Paſſion.}
the daughters of Sion are haughtie, and haue walked with ſtretched out
necke, and went with twinglings
%%% o-1467
of eies, and clapped their handes, walked on their feete, and ietted in
a ſet pace. \V Our Lord shal make balde the crowne of the daughters of
Sion, and our Lord shal diſcouer their haire. \V In that day shal our
Lord take away the ornament of shoes, and litle moones. \V And cheynes,
and ouches, and bracelettes, and bonnettes. \V And the sheading combes,
and ſloppes, and tablettes, and ſweete balles, and earlets. \V And
ringes, and pearles hanging on the forehead. \V And changes of apparel,
and shorte clokes, and the fine linen, and nedles, \V and
\Fix{loking}{looking}{obvious typo, fixed in other}
glaſſes, and launes, and headbands, and bonegraces. \V And for ſwete
ſauour there shal be ſtinke, and for a girdle a corde, and for frilled
haire baldnes, and for ſtomacher hairecloth. \V Thy fayreſt men alſo
shal fal by the ſworde, and thy ſtrong ones in battle. \V And her gates
shal lament and moorne, and she shal ſit deſolate on the ground.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
