%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Isaia
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 1593
%%% o-1468
\startChapter[
  title={Chapter 05}
  ]

\Summary{Vnder the figure of a barren vineyard, is prophecied the
  reiection of the Iewes, 7.~for their ſinnes of auarice,
  11.~intemperance, & other wickednes, 18.~one iniquitie drawing an
  other, 20.~iudging good to be euil, and euil good. 25.~For which the
  Gentiles of diuers nations shal afflict them.}


I wil ſing to my beloued
\SNote{Iſaie of the tribe of Iuda here prophecieth the doleful ſongue
which Chriſt vttered weeping ouer Ieruſalem, foreſeeing & foretelling
their deſtruction.
\XRef{Luc.~19. v.~41.}}
the canticle of my coſin concerning his vinyard.
\CNote{\XRef{Mat.~21.}}
A vineyard was made to my beloued in horne the ſonne of oile. \V And he
hedged it, and choſe ſtones out of it, and planted it elect, and built a
towre in the middes thereof, and ſet vp a preſſe therein: and looked
that it should yeld grapes, and it yelded wilde grapes. \V Now therfore
ye inhabitants of Ieruſalem, and men of Iuda, iudge betwen me and my
vineyard. \V What is there that I ought to doe more to my vineyard, and
haue not done to it? Whether that I looked, it should yeld grapes, and
it hath yelded wilde grapes? \V And now I wil shew you what I wil doe to
my vineyard.
\SNote{Al this ſheweth that God only ſubtracting his protection, no man
nor people is able to ſtand; of whoſe ruine God is not the auctor, but
only permitteth that they fal into ſinnes, and ſo into other miſeries.}
I wil take away the hedge therof, and it shal be into ſpoile: I wil
throw downe the wal thereof, and it shal be to be troden vpon. \V And I
wil lay it waiſt: it shal not be pruned, and it shal not be digged: and
bryers and thornes shal ouergrowe it: and I wil command the cloudes that
they rayne no shower vpon it. \V For the vineyarde of the Lord of
hoſtes, is the houſe of Iſrael: and the man of Iuda, his delectable bud:
& I looked that it should doe iudgement, and behold iniquitie: and
iuſtice, and behold clamour. \V Woe to you that ioyne houſe to houſe,
and lay filde to filde, euen to the end of the place: why shal you alone
dwel in the middes of the earth? \V Theſe thinges are in my eares, ſayth
the Lord
%%% 1594
of hoſtes: vnles manie great and fayre houſes become deſolate, without an
inhabiter. \V For ten acres of the vineyards shal yeld one litle flagon, and
thirtie busheles of ſeede shal yeld three busheles. \V Woe to you that
riſe vp earely to folow drunkennes, and to drinke euen vntil euening,
that you may be inflamed with wine. \V Harpe, and viole, & timbrel, and
shalme, and wine in your feaſtes: and
\SNote{An admonition to celebrate feſtiual dayes with holy religious
exerciſes, and not to folow drunkennes nor other wicked or vaine
thinges.}
the worke of our Lord you regard not, nor conſider the workes of his
handes. \V Therfore is my people led away captiue, becauſe they had not
knowlege, and their nobles died with famine, and the multitude thereof
dried away with thirſt. \V Therfore hath hel dilated his ſoule, and
opened his mouth without anie limite, and their ſtrong ones, and their
people, and their high and glorious ones shal deſcend into it. \V And
man shal be bowed, and man shal be humbled, and the eies of the loftie
shal be brought
\Fix{law.}{low.}{obvious typo, fixed in other}
\V And the Lord of hoſtes shal be exalted in iudgement, and the holie
God shal be ſanctified in iuſtice. \V And the lambes shal feede
according to their order, and
%%% o-1469
ſtrangers shal eate the deſerts turned into frutefulnes. \V Woe to you
that draw iniquitie in cordes of vanitie, & ſinne as the linke of a
wayne. \V Which ſay: Let him make haſt, & let his worke come quickly,
that we may ſee it: & let the counſel of the holie one of Iſrael come,
and we shal know it. \V Woe vnto you that cal euil good, and good euil:
putting darknes light, and light darknes: putting bitter for ſwete, &
ſwete for bitter. \V 
\CNote{\XRef{Rom.~12.}}
Woe to you that are wiſe in your owne eies, and prudent before
yourſelues. \V Woe to you that are mightie to drinke wine, & ſtout men
in drunkennes. \V Which iuſtifie the impious for giftes, and take away
the iuſtice of the iuſt from them. \V For this, euen as the tongue of
fire deuoureth ſtuble, and the heate of the flame burneth it vp: ſo shal
their roote be as iſles, and their bud shal riſe vp as duſt, for they
haue caſt away the law of the Lord of hoſtes, and haue blaſphemed the
word of the holie one of Iſrael. \V Therfore is the furie of our Lord
wrath againſt his people, and he hath ſtretched out his hand vpon them,
and ſtriken them: and the mountaines were trubled, and their carcaſſes
were made as dung in the middes of the ſtreates. In al theſe thinges his
furie is not turned away,
\SNote{Greuous ſinnes muſt be greuouſly puniſhed. Such as was the ſinne
of the Ievves perſecuting Chriſt.}
but yet his hand ſtretched forth. \V And he shal lift vp a ſigne in the
nations a farre, and shal whiſtle to him of the ends of the earth: and
behold he shal come in haſt ſpedely. \V There
%%% 1595
is none that faynteth, nor that laboreth in them, he wil not ſlumber,
nor ſleepe, neither shal the girdle of his reines be looſed, neither
shal the latchet of his shoe be broken. \V His arrowes sharpe, and al
his bowes bent. The hoofes of his horſes as the flint, & his wheeles as
the violence of a tempeſt. \V His roaring as a lions, he shal roare as
lions whelpes: and he wil gnash, and hold the praye, and claſpe it, and
there shal be none to plucke it away. \V And it shal ſound vpon him in
that day, as the ſound of the ſea: we shal looke towards the earth, and
behold darkenes of tribulation, and the light is darkened in the miſt
therof.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
