%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Isaia
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 1599
%%% o-1473
\startChapter[
  title={Chapter 09}
  ]

\Summary{Theglath Phalaſar carieth ſome Iſraelites captiue, and Salman
  afar manie more, in figure of a few diſciples conuerted to Chriſt in
  Galilee, and al Iewrie, but manie more in the whole world. 7.~Whoſe
  Impyre shal be great, and durable: 8.~but the Iewes glorie, eſpecially
  of the tenne tribes, shal be obſcured, for their pride, hypochriſie,
  and other ſinnes.}


At
\CNote{\Xref{Mat.~4.}}
\SNote{S.~Mathew expoundeth this prophecie of Chriſt, firſt preaching in
Galilee.

VVhere his diſciples beleued in him & folowed him.}
the firſt time was the land of Zabulon alleuiated: and the land of
Nepthali: and at
\SNote{But after his paſſion few Iewes beleued in him, in compariſon of
the Gentiles.}
the laſt was aggrauated the way of the ſea beyond Iordan of Galilee of
the Gentiles. \V The people that walked in darknes, hath ſene great
light: to them that dwelt in the countrie of the shadow of death, light
is riſen. \V Thou haſt multiplied the nation, and not magnified the
ioy. They shal reioyce before thee, as they that reioyce in harueſt, as
conquerors reioyce after a pray is taken, when they diuide the
ſpoiles. \V For the yoke of their burden, and the rod of their shoulder,
and the ſcepter of their exactour thou haſt ouercome, as in the day of
Madian. \V
\CNote{\XRef{Iudic.~7.}}
Becauſe al violent taking of pray with tumult, and garment mingled with
bloud, shal be to be burnt, and foode for the fyre. \V
\CNote{\XRef{Luc.~2.}}
For
\SNote{He that is great, yea omnipotent God, is borne a litle one in
this vvorld, and vvithout violence conquereth & ruleth al the vvorld.}
\Sc{a litle child is borne to vs}, and a ſonne is geuen to vs, and
principalitie is made vpon his shoulder: and his name shal be called,
Meruelous, Counſeler, God, Strong, Father of the world to come, the
Prince of peace. \V His empire shal be multiplied, and there shal be no
end of peace: he shal ſit vpon the throne of Dauid, and vpon his
kingdom: that he may confirme it, and ſtrengthen it in iudgement and
iuſtice, from this time & for euer: the zeale of the Lord of hoſtes shal
doe this. \V Our Lord hath ſent a word into Iacob, and it is fallen in
Iſrael. \V And al the people of Ephraim shal know, & the
\Fix{inhabitans}{inhabitants}{obvious typo, fixed in other}
of Samaria, ſaying in pride & greatnes of hart: \V Brickes are fallen,
but we wil build with ſquare ſtones: they haue cut downe ſycomores, but
we wil change them into ceders. \V And our Lord
%%% o-1474
shal lifte vp the enemies
%%% 1600
of Raſin ouer him, and shal turne his enemies into tumult. \V Syria from
the Eaſt, and the Philiſthims from the Weſt: and they shal deuoure
Iſrael with ful mouth. In al theſe thinges his
\SNote{God puniſhing ſinners, and they not repenting, his iuſt furie
ſtil increaſeth: puniſhing eternally al thoſe that neuer repent.}
furie is not turned away, but his hand is yet ſtretched forth. \V And
the people is not returned to him, who hath ſtrooken them, and haue not
ſought after the Lord of hoſtes. \V And our Lord shal deſtroy from
Iſrael the head & the tayle, the peruerter and reſtrayner in one day. \V
The aged and honorable, he is the head, & the prophet that teacheth a
lie, he is the tayle. \V And they that cal this people bleſſed, ſeducing
them: and that are called bleſſed, shal be throwen headlong. \V For this
cauſe our Lord shal not reioyce vpon their yong men: and on their
pupilles, and widowes he shal not haue mercie: becauſe euerie one is an
hypocrite & wicked, and euerie mouth hath ſpoken follie. In al theſe
thinges his furie is not turned away, but his hand is yet ſtretched
forth. \V For impietie is kindled as a fyre, it shal deuoure bryer and
thorne: and it shal be kindled in the thicket of the foreſt, and it shal
be wrapped vp together in the pride of ſmoke. \V In the wrath of the
Lord of hoſtes the earth is trubled, and the people shal be foode for
the fyre: man shal not ſpare his brother. \V And he shal decline to the
right hand, and shal be hungrie: and shal eate on the left hand, and
shal not be filled: euerie one shal eate the flesh of his arme: Manaſſes
Ephraim, and Ephraim Manaſſes, they together againſt Iuda. \V
\SNote{VVhere is no repentance, there can be no remiſſion. As
\XRef{v.~12.~17.}
&
\XRef{ch.~10. v.~4.}
&c.}
In al theſe thinges his furie is not turned away, but his hand is yet
ſtretched forth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
