%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Isaia
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 1597
%%% o-1472
\startChapter[
  title={Chapter 08}
  ]

\Summary{Vnder the figure of a new name, Chriſts birth of a virgin is
  againe prophecied, 4.~but firſt the kingdomes of Syria, and Iſrael
  shal be deſtroyed: and Iuda ſore afflicted: 8.~yet conſerued with
  loſſe of manie. 16.~Which is a myſterie hidden from the
  Iewes. 21.~Great euils hang ouer them, that depart from the law.}


And our Lord ſayd to me: Take thee
\SNote{The myſterie here prophecied is of ſo great importance, as would
require a very great booke for ful explication therof.}
a great booke, & write in it with the pen of man. Take away the ſpoiles
ſpedely, quickly take prayes. \V And I tooke to me faithful
%%% 1598
witneſſes, Vrias the prieſt, & Zacharias the ſonne of Barachias. \V And
I went to the propheteſſe, and she conceiued, and bare a ſonne. And our
Lord ſayd to me: Cal his name,
\SNote{Chriſt the Sonne of God and virgins child quickly taketh the pray
from the diuel, who before poſſeſſed almoſt al the world.}
Haſten to take away the ſpoiles: make haſt to take prayes. \V For before
the childe know to cal his father and his mother, the ſtrength of
Damaſcus shal be taken away, and the ſpoiles of Samaria before the king
of the Aſſirians. \V And our Lord added yet to ſpeake vnto me, ſaying: \V
For that
\SNote{The prophet ſpeaketh of the tenne tribes, vvhich ioyned forces
with the king of Syria againſt Ieruſalem, but themſelues vvere firſt
brought into captiuitie by the Aſſyrians, God protecting Ieruſalem for
that time, and long after.}
this people hath caſt away the waters of Siloe, that runne with ſilence,
and rather taken Raſin, and the ſonne of Romelia: \V for this cauſe
behold our Lord wil bring vpon them the waters of the riuer ſtrong and
manie, the king of the Aſſirians, and al his glorie: and he wil aſcend
ouer al their riuers, and wil flowe ouer al their bankes. \V And wil goe
through Iuda, ouerflowing, and paſſing through shal come euen to the
necke. And the ſtretching out of his winges shal fil the bredth of thy
land ô Emmanuel. \V Gather ye together ô peoples, and be ouercome, and
heare al ye landes far of: Take courege, and be ouercome, gird
yourſelues, & be ouercome. \V Take counſel and it shal be defeated:
ſpeake a word, and it shal not be done: becauſe God is with vs. \V For
thus ſayth our Lord to me: As in a ſtrong arme he hath taught me, that I
should not walke in the way of this people, ſaying: \V Say not:
Conſpiracie: for al thinges that this people ſpeaketh, is conſpiracie:
and feare ye not their feare, neither dread ye. \V The Lord of hoſtes
him ſanctifie ye: be he your dread, and he your terrour. \V
\CNote{\XRef{Luc.~2.}
\XRef{Rom.~9.}
\XRef{2.~Pet.~2.}}
And he shal be a ſanctification to you. But for a ſtone of offence, and
for a rocke of ſcandal to the two houſes of Iſrael, for a ſnare and a
ruine to the inhabitants of Ieruſalem. \V And verie manie of them shal
ſtumble and fal, and shal be broken in peeces, and shal be ſnared, and
taken. \V Binde the teſtimonie, ſeale the law in my diſciples. \V And I
wil expect our Lord, who hath hid his face from the houſe of Iacob, and
I wil wayte for him. \V Behold I and my children, whom our Lord hath
geuen me for a ſigne, and for a wonder in Iſrael from the Lord of
hoſtes, which dwelleth in mount Sion. \V And when they shal ſay to you:
Aske of Pythones, and of diuiners, which whiſper in their inchantments:
shal not
%%% o-1473
the people aske viſion of their God, for the liuing of the dead? \V To
the law rather, and to the teſtimonie. And if they ſpeake not according
to this word, they shal not haue the morning
%%% 1599
light. \V And he shal paſſe by it, shal fal, and be hungrie: and when he
shal be hungrie, he wil be angrie, and curſe his king, and his God, and
\SNote{\Fix{VVheter}{Whether}{obvious typo, fixed in other}
they ſeke to God in their extreme diſtreſſe, not ſincerely, but
coacted,}
wil looke vpwards. \V And he wil looke
\SNote{or ſeke worldlie helpe, they ſhal not eſcape miſerie.}
to the earth, and behold tribulation and darknes, diſſolution and
diſtreſſe, and miſt perſecuting, and he can not flie away from his
diſtreſſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
