%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Isaia
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 1588
%%% o-1465
\startChapter[
  title={Chapter 02}
  ]

\Summary{Al nations shal come to the Church of Chriſt; which shal
  beginne in Ieruſalem. 6.~And the Iewes shal be reiected for their
  idolatrie, auarice, and other ſinnes. 11.~Proud men shal be humbled, &
  Gods glorie shal increaſe. 18.~Idolatrie shal be deſtroyed.}


The word, that Iſaie the ſonne of Amos ſaw vpon Iuda and Ieruſalem. \V
And in
\SNote{It is not only certaine, but alſo euident that the prophet
ſpeaketh here, and in innumerable other places, of the Church of Chriſt,
which is the citie ſet vpon a mountaine,
\XRef{Mat.~5.}
vnto which al nations are gathered, and al the time of the new Teſtament
is called the laſt houre,
\XRef{1.~Ioan.~2.}
becauſe no time shal folow after this, but al eternitie.}
the later dayes the montaine of the houſe of our Lord shal be prepared,
in the toppe of montaines, and it shal be eleuated aboue the little
hilles: and al nations shal flowe vnto it. \V And manie peoples shal
goe, & shal ſay: Come and let vs goe vp to the mount of our Lord, and to
the houſe of the God of Iacob, and he wil teach vs his wayes, and we
shal walke in his pathes: becauſe the law shal come forth from Sion, and
the word of our Lord from Ieruſalem. \V And he shal iudge the Gentiles,
and rebuke manie peoples: and they shal turne their ſwordes into
culters, and their ſpeares into ſiethes: nation shal not lift vp ſword
againſt nation, neither shal they be exerciſed any more to battel. \V
Houſe of Iacob come ye, and let vs walke in the light of our Lord. \V
For thou haſt
\SNote{The Iewes were reiected after Chriſts death: before which they
were ſtil conſerued: though often ſeuerely punished. And ſo now the
Church of Chriſt ſhal neuer be
\Fix{reiected, nor}{reiected.}{obvious typo, fixed in other}}
reiected thy people, the houſe of Iacob: becauſe they are filled as in
times paſt, and haue had ſouthſayers as the Philiſthijms, and haue
ſtucke faſt to ſtrange children. \V The land is replenished with ſiluer
and gold: and there is no end of their treaſures. \V And their land is
replenished with horſes: and their chariotes are innumerable. And their
land is ful of idoles: they haue adored the worke of their handes, which
their fingers made. \V And man bowed himſelf, and man was humbled:
therfore forgeue them not. \V Enter thou into the rocke, and be hid in a
pitte, in the ground from the face of the feare of our Lord, & from the
glorie of his maieſtie. \V The loftie eies of man are humbled, and the
height of men shal be made to ſtoupe: & our Lord onlie shal be exalted
in that day. \V Becauſe the day of the Lord of hoſtes shal be vpon al
the proude and loftie, and vpon euerie one that is arrogant, and he shal
be humbled. \V And vpon al the ceders of Libanus high, & eleuated, &
vpon al the okes of Baſan. \V And vpon al the high mountaines, and vpon
al little hilles eleuated. \V And vpon euerie high towre, and euerie
fenſed wal. \V And vpon al the shippes of Tharſis, and vpon al, that is
fayre to
%%% 1591
behold. \V And the loftines of men shal be bowed, and the height of men
shal be humbled, and our Lord onlie shal be exalted in that day. \V And
idols shal
\SNote{It is moſt abſurde, and contrarie to this and other Scriptures,
that Proteſtants feyne of great idolatrie in the Chriſtian world, for a
thouſand or more yeares together, profeſſing Chriſts name & Religion,
and yet continually committing (as theſe new maſters imagine) groſſe
idolatrie.}
vtterly be deſtroyed. \V And they shal enter into the caues of rockes,
and into the pittes of the earth from the face of the feare of our Lord,
and from the glorie of his maieſtie, when he shal riſe vp to ſtrike the
earth. \V In that day shal a man caſtaway the idols of his ſiluer,
%%% o-1466
and the idols of his gold, which he had made him to adore, mowles and
battes. \V And he shal goe into the clefts of rockes, and into the caues
of ſtones from the face of the feare of our Lord, and from the glorie of
his maieſtie, when he shal riſe vp to ſtrike the earth. \V Ceaſe
therfore from the man, whoſe ſpirit is in his noſthrels, becauſe he is
reputed high.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
