%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Isaia
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 1595
%%% o-1469
\startChapter[
  title={Chapter 06}
  ]

\Summary{The prophet after a glorious viſion, 5.~lamenteth his former
  ſilence; 6.~his lippes being purified by an Angel, with a hote coal,
  8.~he is willingly ſent & ſo prophecieth that the peoples hart wil be
  hardned, their cities shal be deſtroyed, but good reliques shal be
  conſerued.}


In
\CNote{\XRef{4.~Reg.~15.}
\XRef{2.~Par.~26.}}
the yeare that king Ozias died, I ſaw our Lord ſitting vpon an high
throne and eleuated: and thoſe thinges that were vnder him filled the
temple. \V Seraphims ſtoode vpon the ſame: ſix winges to one, and ſix
winges to the other: with two
\SNote{Neither Iſaie, nor Moyſes, nor anie other mortal man did euer
ſee God in himſelfe but only ſhadowed. Yet the wicked calumniouſly
accuſed, condemned, and put Iſaie to cruel death, vpon pretence of
blaſphemie, for ſaying that he ſaw God. VVhich he otherwiſe ſaid not,
but \Emph{couered by the vvinges of the Seraphimes}.
\Cite{Origen in hunc locum.}
&
\Cite{S.~Ierom. Tradi. Hebrraicis in Paral.}}
they couered his face, and with two they couered his feete, and with two
they flewe. \V And they cried one to an other, and ſayd:
\CNote{\XRef{Apoc.~4.}}
Holie, holie, holie, the Lord God of
%%% o-1470
hoſtes, al the earth is ful of his glorie. \V 
\CNote{\XRef{Rom.~11.}}
And the lintels of the dores were moued at the voice of him that cried,
and the houſe was filled with ſmoke. \V And I ſaid: Woe is me becauſe I
haue held my peace, becauſe I am a man of polluted lippes, and I dwel in
the middes of a people that hath polluted lippes, and the king the Lord
of hoſtes I haue ſene with mine eies. \V And one of the Seraphims flewe
to me, and in his hand an hote cole, which he had taken with tonges from
the altar. \V And he touched my mouth, and ſaid: Behold this hath
touched thy lippes, and thine iniquitie shal be taken away, & thy ſinne
shal be cleanſed. \V And I heard the voice of our Lord ſaying: Whom
\SNote{Iſaie was not only an Euangelical but alſo an Apoſtolical
prophet, with whom God here treateth and procedeth, as with an
\Emph{Apoſtle}, ſaying: \Emph{VVhom shal I ſend?} and the prophet
anſwering: \Emph{Send me.} God ſent him, ſaying: \Emph{Goe.} &c.
\Cite{S.~Ierom in Proæmio in Iſaie.}}
shal I ſend? and who shal goe for vs? And I ſayd: Loe I am here, ſend
me. \V And he ſayd: Goe, and thou shalt ſay to this people: Heare ye
that heare, and vnderſtand not: and ſee viſion, and knowe it not. \V
\CNote{\XRef{Mat.~2.}}
Blinde the hart of this people, and make their eares heauie, and shut
their eies: leſt perhaps they may ſee with their eies, and heare with
their eares, and
%%% 1596
vnderſtand with their hart, and be conuerted and I heale them. \V And I
ſaid: How long Lord? And he ſaid: Vntil the cities be made deſolate
without inhabiter, and the houſes without man, and the land shal be
leaft deſert. \V And our Lord shal make men goe farre of, and he shal be
multiplied that was leaft in the middes of the earth. \V And there shal
be tithing in it, and she shal be conuerted, and shal be to the shew as
a terebinth, and as an oke, that ſpreadeth his boughes: that which shal
ſtand in it, shal be an holie ſeede.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
