%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Isaia
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 1592
%%% o-1467
\startChapter[
  title={Chapter 04}
  ]

\Summary{After the deſtruction of the Iewes, manie wemen shal ſeeke to
  marie with one man, 2.~but the reliques, repenting of their ſinnes,
  shal returne to God, 5.~and florish vnder his protection.}


And
\SNote{After the reduction of heathniſh, or heretical people to
catholique religion, there wil be great want of ſpiritual paſtors.}
ſeuen wemen shal take hold of one man in that day, ſaying: We wil eate
our owne bread, and be couered with our garments: only let thy name be
called vpon vs, take away our reproch. \V In that day the bud of our
Lord shal be in magnificence, and glorie, and the fruite of the earth
high, and exultation to them, that shal be ſaued of Iſrael. \V And it
shal be: Euerie one that shal be leaft in Sion,
%%% 1593
and
\SNote{Not al the Iewes that eſcaped temporal death in the deſtruction
of Ieruſalem, but thoſe only shal be eternally ſaued, that beleuing shal
be baptized and liue wel.}
shal remaine in Ieruſalem, shal be called holie, euerie one that is
written in life in Ieruſalem. \V If our Lord shal cleanſe the filth of
the daughters of Sion, and shal wash the bloud of Ieruſalem out of the
middes thereof in the ſpirit of iudgement, and ſpirit of heate. \V And
our Lord shal create vpon euerie place of mount Sion, and where he is
inuocated, a clowde by day, and ſmoke, and the brightnes of flaming fyre
in the night: for vpon al glorie protection. \V And there shal be a
tabernacle for a place of shadow in the day from the heate, and for
ſecuritie, and couert from the whyrlewind, and from rayne.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
