%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Isaia
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 1600
%%% o-1474
\startChapter[
  title={Chapter 10}
  ]

\Summary{Makers of wicked lawes are curſed. 3.~For which the Iſraelites
  shal be afflicted by the Aſſirians. 5.~The Aſſirians ouerthrowne by
  extraordinarie meanes ſent from God: 21.~and the Iewes deliuered from
  imminent danger: with diuers myſteries of Chriſt intermixed.}


VVoe to them that make
\SNote{VVheras good lawes are the ſtabilitie of the cõmonwealth, wicked
are the ruine therof. Such as Ieroboam made forbidding to goe to
Ieruſalem and ſetting vp golden calues in Bethel and Dan: cauſing the
people to ſerue them as the goddes of Iſrael.
\XRef{3.~Reg.~12. v.~16.}
Such alſo as the Scribes and
\Fix{Phariſes}{Phariſees}{likely typo, fixed in other}
made deuiſing wicked traditions, contrarie to Gods commandments.
\XRef{Mat.~15. v.~5.}}
wicked lawes: and writing, haue written iniuſtice. \V That they might
oppreſſe the poore in iudgement, & doe violence to the cauſe of the
humble of my people: that widowes might be their praye, and they might
ſpoile pupilles. \V What wil you doe in the day of viſitation, and of
calamitie coming from farre? to whoſe helpe wil ye flee? and where wil
ye leaue your glorie? \V That you be not bowed vnder the bond, and fal
with the ſlaine? In al theſe thinges his furie is not turned away, but
his
%%% 1601
hand is yet ſtretched forth. \V Woe to Aſſur, he is the rod of my furie,
and the ſtaffe, myne indignation is in
%%% o-1475
their handes. \V I wil ſend him to
\SNote{The Iewes are called a deceiptful nation becauſe they broke their
promiſe made to God, that they would ſerue him and kepe his
commandments.
\XRef{Exo.~19. v.~8.}}
a deceitful nation, & I wil geue him commandment againſt the people of
my furie, that he take away ſpoiles, and catche the praye, and put them
to be troden vpon, as the mire of the ſtreates. \V But he shal not ſo
thinke, and his hart shal not eſteme it ſo: but his hart shal be ſet to
deſtroy, and to the deſtruction of no few nations. \V For he shal
ſay: \V Are not my princes with al kinges? Is not as Charcamis, ſo
Calano: and as Arphad, ſo Emath? Is not as Damaſcus, ſo Samaria? \V Euen
as my hand hath found the kingdomes of the idol, ſo alſo their idols of
Ieruſalem, & of Samaria. \V Shal I not as I haue done to Samaria and
her idols, ſo do to Ieruſalem and her idols? \V
\SNote{\XRef{4.~Reg.~9.}}
And it shal be: when the Lord shal haue accomplished al his workes in
mount Sion, and in Ieruſalem, I wil viſite ouer the fruite of the
magnifical hart of the king of Aſſur, and ouer the glorie of the
hautines of his eyes. \V For he hath ſaid: In the ſtrength of mine owne
hand haue I done it, and in mine owne wiſdome haue I vnderſtood: and I
haue taken away the borders of peoples, and haue ſpoiled their princes,
and haue pulled downe as a mightie man, them that ſate on high. \V And
my hand hath found the ſtrength of peoples as a neſt: and as egges be
gathered, that are leaft, ſo haue I gathered together al the earth: and
there was none that moued wing, and opened mouth, and once muttered. \V
Shal
\SNote{Senacharib not by his owne powre but as Gods inſtrument &
miniſter afflicted the Iſraelites. Neuertheles he perſecuted them of his
owne free wil, which God vſed for the puniſhment of his people. In
general therfore euil men are like to inſtruments without ſenſe, but
differ in that mens actions are voluntarie, & vnreaſonable, and ſenſles
creatures haue no wil at al, but only natural aptnes, and inclination.}
the axe glorie againſt him, that cutteth with it? or shal the ſaw exalt
itſelfe againſt him, by whom it is drawen? As if a rod should lift vp
itſelf agaynſt him, that lifteth it vp, and a ſtaffe exalt itſelf which
is certes but wood. \V For this cauſe the Dominatour the Lord of hoſtes
shal ſend leannes in his fat ones: and vnder his glorie shal burne as it
were the burning of fyre kindled. \V And the light of Iſrael shal be in
fyre, and the Holie one therof in flame: and his thorne shal be kindled,
and be deuoured, and the briars in one day. \V And the glorie of his
foreſt, and of his carmelus shal be conſumed, from the ſoule euen to the
flesh, and he shal be a fugitiue for feare. \V And the remaynes of the
woode of his foreſt, for the fewnes shal be numbred, and a child shal
write them. \V And it shal be in that day, the reſidue of Iſrael, and
they that shal eſcape of the houſe of Iacob, shal not adde to leane vpon
him, that ſtriketh them: but they shal leane vpon our Lord the
%%% 1602
holie one of Iſrael in truth. \V The remnant shal be conuerted, the
remnant I ſay of Iacob, to the ſtrong God. \V For if thy people ô Iſrael
shal be as the ſand of the ſea, the remnant therof shal be conuerted,
\Fix{conſumnation}{conſummation}{likely typo, fixed in other}
abbridged shal make iuſtice ouerflow. \V For our Lord the God of hoſtes
shal make conſummation, and abbridgement in the middes of al the
earth. \V For this cauſe, thus ſayth our Lord the God of hoſtes: O my
people
\Fix{inhatiter}{inhabiter}{obvious typo, fixed in other}
of Sion, be not afrayd of Aſſur: he shal ſtrike thee with his rod, and
shal lift vp his ſtaffe ouer thee in the way of Ægypt. \V For yet a
litle and a very litle, and mine indignation and furie vpon their
wickednes shal be conſummate. \V And the Lord of hoſtes shal rayſe vp a
ſcourge vpon
%%% o-1476
him,
\CNote{\XRef{Iudi.~7.}}
according to the plague of Madian in the Rocke Oreb, and his rod vpon
the ſea, and he shal lift it vp in the way of Ægypt. \V And it shal be
in that day: his burden shal be taken away from of thy shoulder, and his
yoke from of thy necke, and the yoke shal putrifie at the face of
oile. \V He shal come into
\SNote{By theſe places Senacherib paſſed with his armie from Ægypt to
Ieruſalem.}
Aiath, he shal paſſe into Magron: at Machmas he shal commend his
veſſels. \V They haue paſſed in haſt, Gaba is our ſeate: Rama was
aſtonied, Gabaath of Saul fled. \V Neay with thy voice ô daughter of
Gallim, attend Laiſa, feelie poore Anathoth. \V Medemena is remoued: ye
inhabitants of Gabin take courege. \V Yet there is day, to ſtand in
Nobe: he shal shake his hand ouer the mountaine of the daughter of Sion,
the litle hil of Ieruſalem. \V Behold the dominatour the Lord of hoſtes
shal breake the litle flagon in terrour, and the high of ſtature shal be
cut downe, and the loftie shal be humbled. \V And the thicke places of
the foreſt shal be ouerthrowen with iron, and Libanus with the high ones
shal fal.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
