%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Isaia
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 1588
%%% o-1463
\startChapter[
  title={Chapter 01}
  ]

\Summary{Iſaie
\MNote{The firſt part.

The kingdom of Iuda shal be captiue in Babylon, for their ingratitude
towards
\Fix{Gods,}{God,}{likely typo, same in both}
and other ſinnes.}
prophecying in the dayes of foure kinges of Iuda, 2.~admonisheth both
princes and people of their ingratitude, and other ſinnes againſt God:
7.~for which they shal be led captiue. 11.~Neither shal ſacrifices, nor
prayers ſaue them, 16.~except they cleanſe their ſoules from ſinnes,
20.~which they not doing shal be ſeuerely punished. 26.~Wherby the
reliques shal be purged, and the Church shal flourish.}


The viſion of Iſaie the ſonne of
\SNote{Not Amos the third of the leſſe prophetes, for this name is
written in other letters in Hebrew: but one of the royal bloud as
S.~Ierom teſtifieth.}
Amos, which he ſawe concerning Iuda and Ieruſalem in the dayes of Ozias,
Ioathan, Achaz, and Ezechias kinges of Iuda. \V Heare ye heauens, & geue
eare ô earth, becauſe our Lord hath ſpoken. I haue brought vp children,
and exalted them: but they haue deſpiſed me. \V The oxe hath knowen his
owner, and the aſſe his maſters crib: but Iſrael hath not knowen me, and
my people hath not vnderſtood. \V Woe to the ſinful nation, the people
loden with greeuous iniquitie, the wicked ſeede, vngracious children:
they haue forſaken our Lord, they haue
blaſphemed the holie one of Iſrael, they are reuolted backewards. \V For
what shal I ſtrike you anie more, which adde preuarication?
\SNote{Prince, Prieſt, and people are al ſicke of ingratitude againſt
God, & other iniquities.}
euerie head is ſicke, & euerie hart in heauines. \V From the ſole of the
foote vnto the toppe of the head, there is no health therein: wound, and
wayle, and ſwelling ſtroke: it is not bound vp, nor cured with
medicine, nor mollified with oile. \V Your land is deſolate, your cities
burnt with fire: your countrie ſtrangers deuoure before your face, and
it shal be made deſolate as in the ſpoile of enemies. \V And
\SNote{Ieruſalem defaced and deſtroyed.}
the daughter of Sion shal be left as a vineyard, and as a cottage in a
place of cucumbers, and as a citie that is waſted. \V
\CNote{\XRef{Rom.~9.}}
\SNote{God continually preſerueth ſome holie ſeede, that his Church
neuer faileth.}
Vnleſſe the Lord of hoſtes had lefte vs ſeed, we had beene as Sodom, and
we should be like to Gomorrha. \V Heare the word of our Lord ye princes
of Sodom, geue eare to the law of our God ye people of Gomorrha. \V To
what purpoſe do you offer me the multitude of your victimes, ſaith our
Lord? I am ful, the holocauſt of rammes, and the fatte of fatlings, and
the bloud
%%% o-1464
of calues, and lambes, and buck goates I haue not deſired. \V When you
should haue come before my ſight, who ſought
%%% 1589
for theſe thinges at your handes, that you should walke in my courtes? \V
Offer ſacrifice no more in vaine: incenſe is abomination to me. The Newe
moone, and the Sabbath, and other feſtiuities I wil not abide, your
aſſemblies are wicked. \V My ſoule hateth your Calendes, and your
ſolemnities: they are become tedious to me, I haue laboured in
ſuſteyning. \V And when you shal ſtreach forth your hands, I wil turne
away mine eies from you: & when you shal multiplie prayer, I wil not
heare: for your handes are ful of bloud. \V Wash you, be cleane, take
away the euil of your cogitations from mine eies: ceaſe to doe
peruerſely. \V Lerne to doe good: ſeeke iudgement, ſuccour the
oppreſſed, iudge for the pupil, defend the widow. \V And come, and
accuſe me, ſayth our Lord: If your ſinnes shal be as ſcarlet, they shal
be made white as ſnow: and if they be red as vermelon, they shal be
white as wooll. \V If you be willing, and wil heare me, you shal eate
the good thinges of the earth. \V But if you wil not, and wil prouoke me
to wrath: the ſword shal deuoure you, becauſe the mouth of our Lord hath
ſpoken. \V How is the faythful citie, ful of iudgement,
\SNote{Much wickednes reigned in Ieruſalem before their captiuitie in
Babylon: but much more at Chriſts paſſion, when they perſecuted him euen
to the Croſſe: and his diſciples, and al chriſtians, til their citie was
taken by the Romanes and the whole nation diſperſed.}
become an harlot? iuſtice hath dwelled in it, but now mankillers. \V Thy
ſiluer is turned into droſſe: thy wine is mingled with water. \V Thy
princes are vnfaithful, companions of theues: al loue giftes, folow
rewardes. They iudge not for the pupil: and the widowes cauſe goeth not
in to them. \V For this cauſe ſayth our Lord the God of hoſtes the
mightie one of Iſrael: Alas, I wil comfort myſelfe vpon mine
aduerſaries: and wil be reuenged of mine enemies. \V And I wil turne
mine hand to thee, and I wil boyle out thy droſſe til it be pure, & wil
take away al thy tinne. \V And I wil reſtore
\SNote{The Rabbins vnderſtand thoſe Iudges and Prieſtes that gouerned
the people after their deliuerie from captiuitie: but S.~Ierom
expoundeth this and the like places of the Apoſtles and their
ſucceſſors.}
thy iudges as they haue beene before, and thy counſelers as of
old. After theſe thinges thou shalt be called the iuſt, a faithful
citie. \V Sion shal be redemed in iudgement, and they shal bring her
backe in iuſtice. \V And he shal deſtroy the wicked, and the ſinners
together: and they that haue forſaken our Lord, shal be conſumed. \V For
they shal be confounded for the idols, to which they haue ſacrificed:
and you shal be ashamed of the gardens, which you choſe. \V When you
shal be as an oke the leaues falling of, and as a garden without
water. \V And your ſtrength shal be, as the iſles of towe, and your worke
as a ſparke: and both shal be ſet on fire together, and there shal be
none to quench it.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
