%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Philippians
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 2768
%%% o-2626
\startChapter[
  title={Chapter 1}
  ]

\Summary{Hauing ſignified that he vſeth to thanke God for their vertue,
  9.~and alſo to pray for their increaſe, 12.~he certifieth them (for
  their confirmation & comfort) what good was come through his trouble
  at Rome, 24.~& that he doubteth not (though he rather deſire
  martyrdom) but to come againe vnto them, 27.~exhorting them to liue as
  they ought to doe, 28.~and namely not to shrinke for perſecution.}

Paul and Timothee the ſeruants of \Sc{Iesvs} Chriſt; to al the Saints in
Chriſt \Sc{Iesvs} that are at Philippi, with the
\LNote{Bishops and Deacons.}{Wicleffe
\MNote{Bishops and Prieſts alwaies diſtinct functions.}
and other Heretikes would proue by this that Prieſts are not here named;
& for that there could not be many Bishops of this one towne, that there is
no difference betwixt a Bishop and a Prieſt, which was the old hereſie
of Aerius, of which matter,
\CNote{See
\XRef{Annot. Tit.~1. v.~5.}}
in other places. For this preſent it is enough to know that in the
Apoſtles time there were not obſerued alwaies proper diſtinct names of
either function, as they were quickly afterward, though they were
alwaies diuers degrees & diſtinct functions. See
%%% !!! Bad Citing.
\Cite{S.~Chryſoſtom,}
\Cite{Occumenius,}
\Cite{Theophylactus,}
and the reſt of the Græcians vpon this place.}
Biſhops and Deacons. \V Grace to you and peace from God our Father, and
our Lord \Sc{Iesvs} Chriſt.

\V I giue thankes to my God in al memorie of you (\V alwaies in al my
praiers for al you, with ioy making petition) \V for your communicating
in the Ghoſpel of Chriſt from the firſt day vntil now. \V Truſting this
ſame thing, that he which hath begun in you a good worke, wil perfit it
vnto the day of Chriſt \Sc{Iesvs}. \V As it is reaſon for me, this to
thinke for al you, for that I haue you in hart, & in my bands, and in
the defenſe, and the confirmation of the Ghoſpel, al you to be partakers
of my ioy. \V For God is my witnes, how I couet you al in the bowels of
\Sc{Iesvs} Chriſt. \V And this I pray, that your charitie may more and
more abound in knowledge and in al vnderſtanding: \V that you may
approue the better things, that you may be ſincere and without offence
vnto the day of Chriſt, \V repleniſhed with the fruit of iuſtice by
\Sc{Iesvs} Chriſt, vnto the glorie and praiſe of God.

\V And I wil haue you know, Brethren, that the things about me are come
to the more furtherance of the Ghoſpel:
%%% o-2627
\V ſo that my bands were made manifeſt in Chriſt in al the court, and in
al the reſt, \V that many of our Brethren in our Lord, hauing confidence
in my bands, were bold more aboundantly without feare to ſpeake the
%%% 2769
word of God. \V Some indeed euen for enuie and contention: but ſome alſo
for good wil preach Chriſt. \V Some of charitie: knowing that I am ſet
vnto the defence of the Ghoſpel. \V And ſome of contention preach Chriſt
not ſincerely: ſuppoſing that they raiſe affliction to my bands. \V But
what? So that by al meanes, whether by occaſion, or by truth, Chriſt be
preached: in this alſo I reioyce, yea & wil reioyce. \V For I know that
this ſhal fal our to me vnto ſaluation by your praier and the
ſubminiſtration of the Spirit of \Sc{Iesvs} Chriſt, \V according to my
expectation & hope; becauſe in nothing ſhal I be confounded, but in al
confidence as alwaies, not alſo ſhal Chriſt be magnified in my body,
whether it be by life, or by death. \V For vnto me, to liue is Chriſt:
and to die is gaine. \V And if to liue in the fleſh, this vnto me be the
fruit of the worke, and what I ſhal chooſe I know not. \V And I am
ſtraitned of the two: hauing deſire to be diſſolued & to be with Chriſt,
a thing much more better. \V But to abide in the fleſh,
\Var{neceſſarie}{more neceſſarie}
for you. \V And truſting this, I know that I ſhal abide and continue
with you al, vnto your furtherance and ioy of the faith: \V that your
gratulation may abound in Chriſt \Sc{Iesvs} in me, by my comming againe
to you.

\V Only
\CNote{\XRef{Eph.~4,~1.}}
conuerſe ye worthie of the Ghoſpel of Chriſt: that whether when I come
and ſee you, or els be abſent, I may heare of you that you ſtand in one
Spirit, of one mind labouring together to the faith of the Ghoſpel. \V
And in nothing be ye terrified of the Aduerſaries, which to them is
\SNote{\G{ἔνδειξις}, A manifeſt proofe and euidence.}
cauſe of perdition: but to you of ſaluation, & this of God: \V for to
you
\TNote{\G{ἐχαρίσθη}}
it is giuen for Chriſt, not only that you beleeue in him, but alſo that
you ſuffer for him, \V hauing the ſame combat like as you haue ſeen in
me, and now
\Var{haue heard}{doe heare}
of me.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

