%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Philippians
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.




\startcomponent chapter-02


\project douay-rheims


%%% 2770
%%% o-2628
\startChapter[
  title={Chapter 2}
  ]

\Summary{He exhorteth them moſt inſtantly to keep the vnitie of the
  Church, and to humble themſelues for that purpoſe one to another,
  5.~by the example of the maruelous humilitie of Chriſt, 9.~ſpecially
  ſeeing how maruelouſly he is not exalted for it. 12.~Item to
  obedience, feare, and perſeuerance. 17.~Inſinuating (leſt it should
  afterwards trouble them) that he may be martyred at this
  time. 19.~Timothee he hopeth to ſend, whom he highly commendeth:
  25.~as alſo Epaphroditus, whom he preſently ſendeth.}

If therfore there be any conſolation in Chriſt, if any ſolace of
charitie, if any ſocietie of ſpirit, if any bowels of commiſeration; \V
fulfil my ioy, that you be of one meaning, hauing the ſame charitie, of
one mind, agreeing in one. \V Nothing by contention, neither by vaine
glorie: but in humilitie,
\CNote{\XRef{Ro.~12,~10.}}
each counting other better then themſelues: \V
\CNote{\XRef{1.~Cor.~10,~24.}}
euery one not conſidering the things that are their owne, but thoſe that
are other mens. \V For this thinke in your ſelues, which alſo in Chriſt
\Sc{Iesvs}; \V who when he was in the forme of God, thought it no
robberie, himſelf to be equal to God: \V but he exinanited himſelf,
taking the forme of a ſeruant, made into the ſimilitude of men, and in
ſhape found as a man. \V He
\CNote{\XRef{Heb.~1,~9.}}
humbled himſelf, made obedient vnto death: euen the death of the
croſſe. \V
\LNote{For the which.}{Caluin
\MNote{Caluin's blaſphemie againſt Chriſt's owne merits.}
doth ſo abhorre the name of merit in Chriſtian men toward their owne
ſaluation, that he wickedly and vnlearnedly denieth Chriſt himſelf to
haue deſerued or merited any thing for himſelf: though theſe words
(which he shamefully writheth from the proper and plaine ſenſe, to
ſignifie a ſequele and not a cauſe of his exaltation) and diuers other
in holy writ, proue that he merited for himſelf according to al learned
mens iudgement. As
\XRef{Apoc.~5.}
\Emph{The Lamb that was ſlaine, is worthie to receiue power and
Diuinitie.} And
\XRef{Heb.~2.}
\Emph{We ſee} \Sc{Iesvs} \Emph{for the paſsion of death, crowned with
glorie and honour.} See 
\Cite{S.~Auguſtin vpon theſe words of the Pſalme~109.}
\L{propterea exaltabit caput}.}
For the which thing God alſo hath exalted him, and hath giuen him a name
which is aboue al names: \V that
\CNote{\XRef{Eſ.~45,~14.}
\XRef{Ro.~14,~11.}}
in the
\LNote{Name of \Sc{Iesvs}.}{By
\MNote{The Proteſtants wil haue no reuerence done at the name
of \Sc{Iesvs}.}
the like wickednes they charge the faithful people for capping or
kneeling when they heare the name of \Sc{Iesvs}. As though they
worshipped not our Lord God therin, but the ſyllables or letters or
other material elements wherof the word written or ſpoken conſiſteth;
and al this, by ſophiſtication to draw the people from due honour and
deuotion toward \Sc{Christ Iesvs}, which is Satans drift by putting
ſcruples into poore ſimple mens minds about his Sacraments, his Saints,
his Croſſe, his name, his image, & ſuch like, to abolish al true religion
out of the world, and to make them plaine Atheiſts.
\MNote{How Catholikes honour the name of \Sc{Iesvs}, and other things
pertaining to him.}
But the Church knoweth Satans cogitations, and therfore by the
Scriptures and reaſon, warranteth and teacheth al her children to doe
reuerence when ſo-euer \Sc{Iesvs} is named. Becauſe Catholikes doe not
honour theſe things nor count them holy, for their matter, colour,
ſound, and ſyllables, but for the reſpect and relation they haue to our
Sauiour, bringing vs to the remembrance and apprehenſion of Chriſt, by
ſight, hearing, or vſe of the ſame ſignes: els why make we not reuerence
at the name of Ieſus the ſonne of Sirach, as wel as of \Sc{Iesvs Christ}?
And it is a pitieful caſe to ſee theſe prophane ſubtelties of Heretikes
to take place in religiõ, which were ridiculous in al other trade of
life. When we heare our Prince or Soueraigne named, we may without theſe
ſcruples doe obeiſance, but towards Chriſt it muſt be ſuperſtitious.}
name of \Sc{Iesvs} euery knee bow of the celeſtials, terreſtrials, and
infernals: \V and euery tongue confeſſe that our Lord \Sc{Iesvs} Chriſt
is in the glorie of God the Father.

\V Therfore, my Deareſt, (as you haue alwaies obeied) not as in the
preſence of me only, but much more now in my abſence,
\LNote{With feare and trembling.}{Againſt
\MNote{Veine ſecuritie of ſaluation.}
the vaine preſumption of Heretikes that make men ſecure of their
predeſtination and ſaluation, he willeth the Philippians to worke their
ſaluation with feare and trembling, according to that other Scripture,
\Emph{Bleſſed is the man that alwaies is fearful.}
\XRef{Prouerb.~28,~14.}}
with feare and trembling worke your ſaluation. \V For it is God that
\LNote{Worketh in you.}{Of
\MNote{S.~Auguſtin anſwereth the obiection againſt free-wil.}
this thus ſaith S.~Auguſtin:
\Emph{Not becauſe the Apoſtle ſaith, it is God that worketh in you both
to wil and worke, muſt we thinke he taketh away our free-wil. For if it
were ſo, then would he not a litle before haue willed them to worke
their owne ſaluation with feare and trembling. For when they be
commanded to worke, their free-wil is called vpon: but, with trembling
and feare, is added, leſt by attributing their wel-working to
themſelues, they might be proud of their good deeds as though they were
of themſelues.}
\Cite{Auguſt. de gratia & lib. arbit. c.~9.}}
worketh in you both to wil and to accompliſh, according to his good
wil. \V And doe ye al things without murmurings and ſtaggerings: \V that
you may be without blame, and the ſimple children of God,
%%% o-2629
without reprehenſion in the middes of a crooked and peruerſe
Generation. Among whom you ſhine as lights in the world: \V conteining
the word of life
\SNote{Such as haue by their preaching gained any to Chriſt, shal ioy
and glorie therin exceedingly at the day of our Lord.}
to my glorie in the daie of Chriſt, becauſe I haue not runne in vaine,
nor in vaine laboured. \V But and if I be
\SNote{Paſtours ought to be ſo zelous of the ſaluation of their flock,
that with S.~Paul they should offer themſelues to death for the ſame.}
immolated, vpon the
\LNote{The Sacrifice.}{The
\MNote{Martyrdom.}
obedience of faith and Martyrdom be ſo acceptable actes to God, when
they be voluntarily referred to his honour, that by a metaphore they be
called Sacrifice & pleaſant Hoſts to God.}
Sacrifice and ſeruice of you faith, I reioyce and congratulate with you
al. \V And the ſelf-ſame thing doe you alſo reioyce, and congratulate
with me.

\V And I hope in our Lord \Sc{Iesvs} to ſend Timothee vnto you quickly,
that I alſo may be of good comfort, when I know the things pertaining to
you. \V For I haue no man ſo of one mind that with ſincere affection is
careful for you. \V For
\SNote{Many forſake their Teachers when they ſee them in bands and
priſon for their faith, becauſe moſt men preferre the world before
Chriſt's glorie.}
al ſeeke the things that are their owne; not the things that are
\Sc{Iesvs} Chriſts. \V And know ye an experiment of him, that as a ſonne
the Father, ſo hath he ſerued
%%% 2771
with me in the Ghoſpel. \V This man therfore I hope to ſend vnto you,
immediately as I ſhal ſee the things that concerne me. \V And I truſt in
our Lord that my ſelf alſo ſhal come to you quickly. \V But I haue
thought it neceſſarie to ſend to you Epaphroditus my brother and
coadiutour and fellow-ſouldiar, but your Apoſtle, and miniſter of my
neceſſitie. \V Becauſe indeed he had a deſire toward you al: and was
penſiue, for that you had heard that he was ſicke. \V For indeed he was
ſicke euen to death: but God had mercie on him: and not only on him, but
on me alſo, leſt I ſhould haue ſorrow vpon ſorrow. \V Therfore I ſent
him the more ſpeedily: that ſeeing him, you may reioyce againe, and I
may be without ſorrow. \V Receiue him therfore with al ioy in our Lord:
and ſuch intreat with honour, \V becauſe for the worke of Chriſt, he
came to the point of death: yealding his life, that he might fulfil that
which on your part wanted toward my ſeruice.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

