%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Gradual Psalmes note
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent gradual-psalmes-note


\project douay-rheims


%%% 1369
%%% o-1257
\startChapter[
  title={\Sc{A Briefe Note Concerning the Gradual Psalmes.}}
  ]

Here folow in order fifetene Pſalmes intitled Gradual Canticles.
\MNote{VVhy the 15.~Pſalmes folowing are called Gradual Canticles.}
The Hebrew word \HH{Mahaloth} ſignifieth Steppes, or Aſcenſions. The
reaſon wherof Aadias, and ſome other Rabbins yeld, for that they were
ſongue with hiegheſt eleuated notes, that can be in Muſike. The Talmud
ſaith they are ſo called, becauſe they were ſongue in the fifetene
ſteppes going vp into the Temple.
\MNote{The hiſtorical ſenſe.}
But S.~Auguſtin, S.~Baſil, and other Chriſtian Fathers expound them
according to the hiſtorie, and immediate prophetical ſenſe, of the
deliuerie of the Iewes from captiuitie of Babylon, aſcending into
Ieruſalem: which is ſo ſituated on mountaines, that the way from al
partes was by aſcending vnto it.
\MNote{The myſtical ſenſe.}
According to the Myſtical ſenſe, of aſcending ſpiritually by vertues to
perfection, and to eternal felicitie. For the way tending to vertue
(ſaith S.~Baſil) is like to certaine ſteppes (or degrees) by litle and
litle bringing the man that loueth wiſdome vnto heauen. Theſe Canticles
therfore are prayers, mixed with conſolations, for the ioyful deliuerie
of Gods people, from that great captiuitie in Babylon, which the
Pſalmiſt King Dauid ſaw in prophetical ſpirite, and which his poſteritie
felt, and ſometime indured. VVhich againe as a figure ſignifieth the
returne and aſcending of mankind from ſinne to grace, and from the
miſerable ſtate of this world into heauen. VVherupon S.~Auguſtin
interpreteth this prophecie of the aſcenſion, or eleuation of the hart,
from the vaile of teares.
\MNote{Theſe Pſalmes are conſolatorie prayers and prophecies.}
In the meane time, whiles we are in this world, theſe Pſalmes are
conſolatorie prayers, and prophetical aſſurance, that Gods people,
Catholique Chriſtians ſhal be deliuered from thraldom, and perſecution
of Paganes, Turkes, and Heretikes. As partly we ſee by the deliuerie
from the Romane perſecuting Emperors, from the Vandals, Gothes, and
Hunnes: & therfore with aſſured confidence we hope and expect the like
deliuerie from Turkes, and al Heretikes of Luthers broode.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
