%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Interpretation
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent interpretation


\project douay-rheims


%%% 1149
%%% o-1041
\startArgument[
  title={\Sc{Concerning Interpretation of Holie Scriptures.}},
  marking={Concerning Interpretation of Holie Scriptures.}
  ]

As
\MNote{Scriptures are to be expounded by the cõmon ſpirit of the Church,
not by priuate men.}
Prophecie (\Emph{or other holie Scripture}) was not at anie time by mans
wil, but the (\Emph{Prophetes}) holie men of God ſpake, inſpired by the
Holie Ghoſt: ſo no prophecie (\Emph{nor explication}) of Scripture, is
made by priuate interpretation,
\XRef{2.~Pet.~1.}
but by the ſame Spirite wherwith it was written, which our Sauiour gaue
to the Church, to abide for euer, the Spirite of truth; to teach al
truth.
\XRef{Ioan.~14,~16.}
\Emph{Neither} perteyneth it to euerie one, to diſcerne the true
ſpirite, but to ſome.
\XRef{1.~Cor.~12.}

Holie
\MNote{They conſiſt in vnderſtanding.}
Scriptures conſiſt not in reading, but in vnderſtanding.
\Cite{S.~Ierom Dialogo aduerſ. Luciferianos.}

The
\MNote{Holy Fathers do beſt expound them.}
wordes of holie Scripture are ſo to be vnderſtood, as holie men, the
Sainctes of God, haue vnderſtood them,
\Cite{S.~Aug. Ser.~18. de verb. Domini.}
Men muſt lerne of men, not expect knowlege immediatly of God, nor only
by Angels.
\Cite{Idem, in prologo Doct. Chriſt.}

There
\MNote{Some Myſteries are hidden.}
be ſome thinges, mentioned in holie Scripture, which God wil haue
hidden, and thoſe are not to be curiouſly ſearched.
\Cite{S.~Amb. li.~1. c.~7. de vocat. Gentium.}

By
\MNote{They proue our humilitie.}
thoſe thinges, which to vs are hidden in holie Scripture, our humilitie
is proued,
\Cite{S.~Greg. ho.~17. ſuper Ezech.}


\stopArgument


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
