%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 083
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-083


\project douay-rheims


%%% 1291
%%% o-1181
\startChapter[
  title={Psalme 83}
  ]

\PSummary{Deuout
\MNote{Eternal glorie.

The 10.~key.}
perſons feruently deſire eternal glorie, 6.~accounting it, in the meane
time, a happie ſtate to be in the militant Church, 12.~where God firſt
geuing grace, wil geue glorie in the triumphant.}

\PTitle{Vnto the end,
\SNote{For men afflicted in this vale of miſerie.}
for wine preſſes,
\Fix{t}{to}{obvious typo, fixed in other}
\SNote{By the children of Core not being muſitians (but potters in the
temple,
\XRef{1.~Par.~26.)}
%%% !!! Cite? upon this place?
S.~Auguſtin here and in other titles of Pſalmes, vnderſtandeth the
faithful children of Chriſt.}
the children of Core, a Pſalme.}

%%% o-1182
\VV How beloued are
\SNote{The glorious manſions in heauen, which God hath prepared for the
iuſt.}
thy tabernacles ô Lord of hoaſtes! \V my ſoule coueteth, and
\SNote{Vehement deſires do ſometimes depriue vs of external ſenſe.}
fainteth vnto the courtes of our Lord.

My hart, and
\SNote{The mind reioycing in hope, the bodie is alſo recreated,
releeued, and reuiued, which before was dulle and heauie.}
my flesh haue reioyced toward the liuing God.

%%% 1292
\V For
\SNote{As ſparowes by natural inſtinct ſeeking habitations, finde houſes
to dwel in,}
the ſparow alſo hath found her an houſe: and
\SNote{and turtles haue neſtes, wherin to lay their young ones: ſo
faithful ſoules ſeeke to dwel in heauen, and in the meane time to lay vp
good workes within the Catholique Church; out of which (ſayth
\Cite{S.~Auguſtin in this place)}
how good ſoeuer workes do ſeme (as when paganes, and heretikes feede the
hungrie, cloth the naked, receiue ſtrangers into their houſes, viſite
the ſicke, comforte priſoners) being not laid in the neſt,
\L{conculcabuntur, & conterentur: non ſeruabuntur, non cuſtodientur}:
they ſhal be trodde vnder foote, they ſhal be bruiſed in peeces: they
ſhal not be conſerued, they ſhal not be kept: but (that ſuch workes may
be profitable and be conſerued) they muſt be donne in true faith, in the
Catholique faith, in ſocietie of the vnitie of the Church.}
the turtledoue a neſt for her ſelfe, where she may lay her young ones.

Thine
\SNote{VVorkes are good, and rightly laide vp when they are donne in
vnitie, and participation of Gods Altares, the moſt proper places of
Diuine Seruice, of external Sacrifice in this life, and ſpiritual
ſacrifice of perfect praiſes in eternal glorie; where al Sainctes
without ceaſing ſing: Holie, holie, holie, Lord God of hoaſtes.
\XRef{Iſaiæ.~6.}
\XRef{Apoc.~4.}}
altars ô Lord of hoaſtes: my King, and my God.

\V Bleſſed are they, that dwel in
\SNote{The Catholique Church.}
thy houſe ô Lord: for euer and euer they shal praiſe thee.

\V Bleſſed is the man, whoſe helpe is from thee: he
\SNote{The iuſt by Gods grace and helpe, may reſolue to aſcend by
ſteppes and degrees, from vertue to vertue
\XRef{(v.~8.)}
euen to heauen,}
hath diſpoſed aſcenſion in his hart, \V in
\SNote{though he be now in this vaile of teares, by reaſon of mans
ſinne, who otherwiſe was before ſinne in paradiſe, a place of delight.}
the vaile of teares, in the place, which he hath appointed.

\V For
\SNote{Chriſt our lawgeuer,}
the lawgeuer shal geue
\SNote{geueth abundance of graces,}
bleſſing, they shal goe
\SNote{with continual increaſe.}
from vertue into vertue:
\SNote{But our only omnipotent God, is to be
\Fix{ſenne}{ſeen}{obvious typo, fixed in other}
by this effect of his grace, in the Church and not elſwhere.}
the God of goddes shal be ſeene in Sion.

\V Lord God of hoaſtes heare my prayer: receiue with thine eare ô God of
Iacob.

\V Behold ô God our protectour: and
\SNote{Agreably to this the Church maketh al her petitions, concluding
al prayers, By Chriſt our Lord.}
looke vpon the face of thy Chriſt.

\V Becauſe
\SNote{In reſpect of the future retribution, which euerie one ſhal
receiue according to their deſertes, one day in Gods Church is better
then thouſands out of it.}
better is one day in thy courtes, aboue thouſands.

I haue choſen to be
\SNote{And better to be in the pooreſt ſtate of Catholique Chriſtians,}
an abiect in the houſe of my God: rather then to dwel
\SNote{then in greateſt palaces or hiegheſt dignities amongſt ſinners.}
in the tabernacles of ſinners.

%%% 1293
%%% o-1183
\V Becauſe God
\SNote{The Diuine wiſdom ſo vſeth mercie, and veritie, that neither may
preiudice the other:}
loueth mercie and truth: our Lord wil geue
\SNote{and ſo geueth grace in this life,}
grace,
\SNote{and glorie in the next.}
and glorie.

\V He wil not depriue them of good thinges, that
\SNote{Beſides innocencie conſerued without ſinne, there is alſo
innocencie after remiſſion of ſinne, of which the prophet here
ſpeaketh.}
walke in innocencie: ô Lord of hoaſtes bleſſed is the man, that hopeth
in thee.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
