%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 125
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-125


\project douay-rheims


%%% 1373
%%% o-1263
\startChapter[
  title={Psalme 125}
  ]

\PSummary{Iſraelites
\MNote{The deliuered from captiuitie reioyce.

The 7.~key.}
releaſed from captiuitie of Babylon, much more the bleſſed both of the
old and new teſtament deliuered by Chriſt from captiuitie of ſinne, do
reioyce: 5.~who neuertheles muſt paſſe through tribulation to eternal
felicitie.}

\PTitle{A gradual Canticle.}

\NV When our Lord turned the captiuitie of Sion: we were made as
\SNote{The Ievves releaſed from captiuitie, vvere excedingly, and almoſt
incredibly comforted, as men for great, and vnexpected ſudaine ioy
thinke it rather a dreame, then a truth that they are deliuered from
miſeries. So
\CNote{\XRef{Act.~22.}}
S.~Peter vvhen he vvas deliuered out of priſon by an Angel, thought it
rather a viſion, then a true deliuerie. Such ſpiritual ioy deuoute
ſoules haue vvhen they are deliuered from ſinne.}
men comforted.

%%% 1374
\V Then was our mouth replenished with ioy: and our tongue with
exultation.

Then
\SNote{VVheras in the tvvo former verſes (and very commonly) the Prophet
ſpeaketh in the preter tenſe, for the aſſurance of that he foreſhevveth,
as if it vvere already donne, yet here he vttereth his prophecie in the
future tenſe, that the Gentiles wil confeſſe that God dealeth
magnifically vvith his people.}
shal they ſay among the Gentiles: Our Lord hath done magnifically with
them.

\V
\SNote{The people alſo themſelues gratfully confeſſe that God dealeth
magnifically vvith them.}
Our Lord hath done magnifically with vs: we are made ioyful.

\V
\SNote{The Prophet forſeing al this in ſpirite, prayeth for the
performance hereof.}
Turne our captiuitie ô Lord,
\SNote{And that it may ſpedely be done, as a torrent that runneth in the
ſouth part of the vvorld is commonly very great, much deſired, but
ſcarſe expected.}
as a torrent in the South.

\V
\SNote{This is the ordinarie diſpoſition of God, that his ſeruants ſhal
make their ſeeding, vvhich is, doe good vvorkes (ſaith S.~Auguſtin)
vvith teares, in tribulation vpon earth:}
They that ſow in teares,
\SNote{and reape a plentiful harueſt, the revvard of their ſuffering and
vvel vvorking, in the next life. In aſſured hope vvherof the Pſalmiſt,
and the vvhole Church ioyfully conclude this Pſalme vvith the tvvo
verſes folovving.}
shal reape in ioyfulneſſe.

\V Going they went and wept, caſting their ſeedes.

\V But coming they shal come with exultation, carying their sheaues.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
