%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 049
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-049


\project douay-rheims


%%% 1231
%%% o-1123
\startChapter[
  title={Psalme 49}
  ]

\PSummary{Chriſt
\MNote{General Iudgement.

The 9.~key.}
in his firſt coming calleth al Nations. 3.~In his ſecond wil iudge the
world. 7.~In the meane time God exhorteth al men to ſerue him in puritie
of virtue, which he much preferreth before external ſacrifice of the old
law, 17.~reprehending ſuch as profeſſe or teach the right way, and liue
wickedly.}

%%% 1232
\PTitle{A Pſalme
\SNote{To be ſongue or tuned by Aſaph a maiſter of muſike.}
to Aſaph.}

%%% !!! This and the next verse are marked as v. 2. in both versions.
\NV The
\SNote{God almightie, who is greater then are al falſly ſuppoſed goddes,
or holie perſons, that participating of his goodnes are called goddes
(as Kinges, Prieſtes, Iudges) coming into this world in mans nature,
calleth al men to ſaluation.}
God of goddes our Lord hath ſpoken: and he hath called the earth, from
the ryſing of the ſunne euen to the going downe.

\V Out
\SNote{The Church of Chriſt began in Sion.}
of Syon the beauty of his comelines.

\V God wil come
\SNote{Chriſt that came in humilitie, and more obſcurely to ſuffer, and
to redeme vs, vvil come in maieſtie, and manifeſtly to iudge.}
manifeſtly: our God and he wil not kepe ſilence.
\SNote{Immediately before the general iudgement, fire ſhal burne al
tranſitorie thinges.}
Fire shal burne forth in his ſight: and round about him a mighty
tempeſt.

\V He shal
\SNote{Geue ſignes in the firmament,}
cal the heauen from aboue: and
\SNote{and in earth.}
the earth to diſcerne his people.

\V Gather ye together his ſaincts vnto him: which ordaine his teſtament
\SNote{VVhich know that to keepe Gods commandments in folowing vertues,
is aboue the oblation of external ſacrifice.}
aboue ſacrifices.

\V And the heauens shal shew forth his iuſtice: becauſe God is Iudge.

\V
\SNote{God inſtructeth his people.}
Heare ô my people, and I wil ſpeake: Iſrael, and I wil teſtifie to thee:
God thy God am I.

\V I wil not rebuke thee in thy ſacrifices: and thy holocauſtes
\SNote{Sacrifices are gratful to God.}
are in my ſight alwaies.

\V I wil
\SNote{But in regard that God needeth not theſe earthly thinges, he
rather requireth a gratful mind. For otherwiſe man in dede can geue
nothing to God: ſeing al that is in the whole world is Gods owne in
proprietie.}
not take calues out of thy houſe: nor buckegoats out of thy flockes.

\V Becauſe al the wilde beaſts of the woods be myne, the cattle in the
%%% o-1124
mountaines and oxen.

\V I haue knowne al the foules of the ayer: and the beauty of the fielde
is with me.

\V If I shal be hungrie, I wil not tel thee: for the round earth is
myne, and the fulnes therof.

\V Wil I eate the flesh of oxen? or wil I drinke the blood of bucke
goats?

%%% 1233
\V
\SNote{Spiritual ſacrifice of prayſe,}
Immolate to God
\LNote{The ſacrifice of praiſe.}{For
\MNote{Sacrifice of praiſe diſpoſeth men to the fruit of external
ſacrifice.}
better and more due performing of external ſacrifice, it is requiſite,
that thoſe which offer it, or deſire to participate, do bring with them
neceſſarie internal vertues, or diſpoſition; as ſorow and repentãce for
their ſinnes, which is a kind of improper ſacrifice (mentioned in the
\XRef{next Pſalme)}
the ſacrifice of iuſtice, which rendereth to euerie one that is due
\XRef{(Pſal.~4.)}
and ſacrifice of praiſe, or thankes geuing, for al Gods benefites
receiued or expected; which kindes of internal and improper ſacrifices,
do nothing preiudice, but rightly prepare men to the fruict of external
ſacrifice, euer vſed in the law of nature, the law of Moyſes, and of
Chriſt. This
place alſo hath an other higher and prophetical ſenſe of the Sacrifice
of Chriſts bodie in the Euchariſt, which is both propitiatorie, and
Sacrifice of praiſe, and thankes geuing. So S.~Auguſtin
\Cite{(orat. aduerſus Iudeos. c.~6.)}
teacheth, that here certainly is a plaine change of the old ſacrifices.
\MNote{The Sacrifice of the Euchariſt prophecied.}
The ſame he affirmeth
\Cite{Ep.~120. c.~18.}
God foreshewing that the old ſacrifices should be changed, which were
offered in shadow of a ſacrifice to come. I wil not take (ſaith God to
Iſrael) calues nor goates at thy hand, &c. but appointeth that al Iſrael
(al nations from the ryſing of the ſunne to the ſetting) shal immolate
the ſacrifice of praiſe, the ſame Chriſt, whom old Simeon knew an
infant, whom he receiued into his handes. Likewiſe,
\Cite{li. contra aduerſ. legis & prophet. c.~20.}
The Church offereth to God in the bodie of Chriſt the ſacrifice of
praiſe.}
the ſacrifice of praiſe, and
\SNote{& due payment of voluntarie vowes made in honour of God,}
pay thy vowes to the Higheſt.

\V And
\SNote{and praying to him for helpe in tribulation are moſt grateful.}
inuocate me in the day of tribulation: I wil deliuer thee, and thou
shalt glorifie me.

\V But to the ſinner God hath ſayde:
\SNote{He that wil teach others, muſt eſpecially flee from ſinne, &
ſerue God ſincerly.}
Why doeſt thou declare my iuſtices, and takeſt my teſtament by thy
mouth?

\V But thou haſt hated diſcipline: & caſt my words behind thee.

\V If thou didſt ſee a theefe, thou didſt
\Fix{rune}{runne}{obvious typo, fixed in other}
with him: and with adulterers thou didſt put thy portion.

\V Thy mouth hath abounded with malice: and thy tongue fourged guiles.

\V Sitting thou ſpakeſt againſt thy brother, and againſt thy mothers
ſonne thou didſt put a ſcandal: \V theſe things haſt thou done, and I
haue held my peace.

\V Thou haſt thought vniuſtly that I wil be like thee: I wil reproue
thee, and ſet it againſt thy face.

\V Vnderſtand theſe things you that forget God: leſt ſometime he take
you violently and there be none to deliuer you.

\V The
%%% !!! This LNote should be part of the above.
%%% \LNote{}{}
ſacrifice of prayſe
\SNote{God is honored by mans gratitude, and other good workes.}
shal glorifie me: and there is the way, by which I wil shew him the
ſaluation of God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
