%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 111
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-111


\project douay-rheims


%%% 1340
%%% o-1230
\startChapter[
  title={Psalme 111}
  ]

\PSummary{True
\MNote{The meanes to be happie.

The 7.~key.}
happines conſiſteth in ſeruing God, keping his commandments, 5.~and in
doing workes of mercie. 10.~The contrarie bringeth to miſerie.}

\PTitle{Alleluia,
\SNote{The Septuagint Interpreters added this mention of Aggeus and
Zacharie, returning from captiuitie, to ſignifie that this Pſalme was
very proper, & meete to be cõmended to the people at that time, wherby
they might lerne, that their ſinnes were the cauſe of their captiuitie,
and of al their miſeries: and if they deſired temporal, or ſpiritual
proſperitie, they muſt obſerue the meanes here preſcribed to obtayne the
ſame.}
Of the returne of Aggeus, and Zacharie.}

\NV Bleſſed 
%%% !!! SNote should go before 'Bleſſed'
\SNote{This Pſalme is alſo compoſed by the Alphabet, as the next
before.}
is the man that feareth our Lord:
\CNote{\XRef{Pſal.~1.}}
\SNote{He that ſincerly feareth God, wil take great delight in keping
his commandments.}
he shal haue great delight in his commandmentes.

%%% 1341
\V
\SNote{So doing he and his shal proſper.}
His ſeede shal be mightie in the earth: the generation of the righteous
shal be bleſſed.

\V Glorie, and riches in his houſe: and
\SNote{The iuſt shal not only proſper in this world but alſo in the
next.}
his iuſtice abideth for euer and euer.

\V
\SNote{God wil alſo comfort the iuſt in tribulations.}
Light is riſen vp in darkenes to the righteous: he is merciful, and
pitiful, and iuſt.

%%% o-1231
\V Acceptable is the man, that is merciful and lendeth, that
\SNote{That shal geue diſcrete and wholſome counſel to the afflicted.}
shal diſpoſe his wordes in iudgement: \V Becauſe he shal not be moued
for euer.

\V The iuſt shal be in eternal memorie: he shal not feare at the hearing
of euil.

\V His hart is readie to hope in our Lord, his hart is confirmed:

\V He shal not be moued til he looke ouer his enemies.

\V He diſtributed, he gaue to the poore:
\SNote{Workes of mercie are alſo called iuſtice, becauſe they concurre
to mans iuſtification,}
his iuſtice remaineth for euer and euer
\SNote{and to his ſaluation.}
his horne shal be exalted in glorie.

\V The ſinner shal ſee, and wil be angrie, he shal gnash his teeth and
pine away: the deſire of ſinners shal perish.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
