%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 62
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-62


\project douay-rheims


%%% 1250
%%% o-1141
\startChapter[
  title={Psalme 62}
  ]

\PSummary{Dauid
\MNote{Dauids deuotion in baniſhment.

The 8.~key.}
in banishment with great affection deſireth to vnite himſelfe with God in
meditation, 4.~purpoſing and promiſing euer to praiſe him:
10.~prophicieth the vaine endeuoures, and condemnation of his enimies;
and his owne aduancement.}

\PTitle{A Pſalme
\SNote{Holie Dauid made this deuout meditation when he was in the foreſt
of Haret, or deſert of Ziph
\XRef{1.~Reg.~22. &~23.}
and could not come to the tabernacle of God, nor to Ieruſalem, where he
eſpecially deſired to be, in the inheritance of our Lord, which was to
him a great affliction. As the like is now to Catholiques, when they are
put in cloſe priſon for their faith, or otherwiſe hindered, that they
can not be preſent at the moſt holie and daylie Sacrifice. In which caſe
we muſt ſupplie as we may, this great loſſe, and comfort ourſelues with
this, or like Pſalme, or prayer, ſaying: O God my God, to thee I watch.}
of Dauid when he was in the deſert of Iuda.
\XRef{(1.~Reg.~22.)}}

\VV O God my God to thee I watch,
\SNote{Euen from the firſt
\Fix{downing}{dawning}{possible typo, fixed in other}
of the morning,}
from the morning light.
\SNote{my ſoule thirſteth after thee,}
My ſoule hath thirſted to thee,
\SNote{yea alſo my very flesh, and whole bodie feeleth great paines, by
this affliction of mind, and deſireth releefe and reſt.}
my flesh to thee very manie wayes.

%%% o-1142
\V In a deſert land, and inacceſſible, and without water,
\SNote{Being now in caſe that I can not ſerue thee, ô God, as I would,
yet I exhibite myſelfe preſent in ſpirite, before thy holie place,}
ſo in the holie houſe haue I appeared to thee,
\SNote{meditating thy powre, and thy glorie.}
that I might ſee thy ſtrength, and thy glorie.

\V Becauſe
\SNote{This conſolation in banishment from thy diuine Seruice, is ſweeter
to me then manie temporal liues, or anie worldlie proſperitie.}
thy mercie is better then liues, my lippes shal prayſe thee.

\V
\SNote{For as the Paſſions of Chriſt abound in vs (ſaith S.~Paul
\XRef{2.~Cor.~1.)}
ſo alſo by Chriſt our comfort aboundeth.}
So wil I bleſſe thee in my life: and in thy name I wil lifte vp my
hands.

%%% 1251
\V As with
\SNote{Repleniſh my ſoul ô God, with the aboundance of thy grace,}
marrow and fatnes let my ſoule be filled:
\SNote{ſo ſhal I be more able to praiſe thee.}
and my mouth shal praiſe with lippes of exultation.

\V
\SNote{Seing in the night, alſo in my bed I meditate of thee;}
If I haue bene mindful of thee vpon my bedde,
\SNote{I wil more diligently do the ſame in the morning.}
in the morning I wil meditate on thee: \V becauſe thou haſt bene my
helper.

\V And in the couert of thy winges I wil reioice, my ſoule hath cleaued
after thee: thy right hand hath receiued me.

\V But
\SNote{My temporal and ſpiritual enimies:}
they in vaine haue ſought my ſoule,
\SNote{and they ſhal be damned for their ſinnes.}
they shal enter into the inferiour partes of the earth.

\V They
\SNote{It happened literally to Saul, that he was ſlaine in battle,
which he made againſt his enimies;}
shal be deliuered into the handes of the ſworde, they shal be
\SNote{and his dead bodie was hung on a wall
\XRef{(1.~Reg.~31.)}
expoſed to wilde beaſtes, or birdes, though it was afterwards burnt and
buried.}
the portions of foxes.

\V But
\SNote{Dauid was preſently after Sauls death exalted to the kingdome, in
figure of Chriſt, whoſe name and glorie was exalted, after the
deſtruction of the Iewes by Pagane Emperours.}
the king shal reioice in God, al shal be praiſed that ſweare by him:
becauſe the mouth is ſtopped of thoſe that ſpeake wicked thinges.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
