%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 129
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-129


\project douay-rheims


%%% 1376
%%% o-1266
\startChapter[
  title={Psalme 129}
  ]

\PSummary{The
\MNote{The ſixth penitential Pſalme.

The 7.~key.}
Iewes, or other people in tribulation for ſinne, or temporal captiuitie,
crie to God to be deliuered: 4.~truſting, and encoreging ech other in
Gods accuſtomed mercie, aſſuredly hoping that he wil redeme & deliuer
them.}

\PTitle{A gradual Canticle.}

\NV From
\SNote{This prayer agreeth to al true penitents, crying to God for
helpe, being either in depth of ſorovv for ſinne, and ſo it is one of
the Penitential Pſalmes: or the depth of feruent deſire to aſcend
tovvards perfection in vertue, and from this vaile of miſerie into
heauen, and ſo it is a Gradual Pſalme: or in the depth of temporal
paines, and ſo it is a ſpecial prayer for ſoules in Purgatorie, offered
by the Church in their behalfe.}
the depthes I haue cried to thee ô Lord: \V Lord heare my voice:

%%% 1377
Let thine eares be intent, to the voice of my petition.

\V If thou shalt obſerue iniquities ô Lord: Lord
\SNote{None is able to abide the rigour of Gods iuſtice.}
who shal ſuſteyne it?

\V Becauſe with thee there is
\SNote{But al muſt relie vpon his mercie.}
propiciation: and for
\SNote{For thy promiſes made in the law, that thou wilt remitte ſinnes
to the penitient, geue more grace to them that ſeeke it, and mitigate
alſo the paines due for ſinnes.}
thy law I haue expected thee ô Lord.

%%% o-1267
My ſoule hath expected in his word: \V my ſoule hath hoped in our Lord.

\V From
\SNote{The hope of penitents is like to the watches of the day time,
from morning vntil night, vvhich are more comfortable, then vvatches of
the night.}
the morning watch euen vntil night: let Iſrael hope in our Lord.

\V Becauſe with our Lord there is mercie: and with him
\SNote{The greateſt comforth is in Chriſt our Redemer, vvhoſe plentiful
Redemption bringeth more abundance of grace.}
plenteous redemption.

\V And he shal redeme
\SNote{Chriſts Redemption being ſufficient for al the vvorld, is
effectual only to true liuing members of the Catholique Church.}
Iſrael, from al his iniquities.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
