%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 015
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-015


\project douay-rheims


%%% 1170
%%% o-1062
\startChapter[
  title={Psalme 15}
  ]

\PSummary{Chriſt,
\MNote{Of Chriſts victorie. The 5.~key.}
by the mouth of Dauid, declareth his future victory,
  and triumph ouer the world, 9.~and death.}

\PTitle{The
\SNote{Stylographia ſignifieth a thing moſt worthie to be noted, to
  witte, Chriſt crucified,}
inſcription of the title
\SNote{and moſt worthie of the Prophets conſideration.}
to Dauid him ſelf.}

\VV Preſerue
\SNote{Chriſt as man did often pray, as appeareth in the Goſpel.}
me Ô Lord, becauſe I haue hoped in thee.

\V I haue ſaid to our Lord: Thou art my God, becauſe thou
\SNote{Chriſts paſſion was not needful nor profitable to God, but to
man.}
needeſt not my goods.

\V
\SNote{God ſpeaketh shewing that Chriſt should make his meruelous
charity knowen to his Apoſtles, and other ſeruantes.}
To the ſainctes, that are in his land, he hath made al my willes
meruelous in them.

\V Their infirmities were
\SNote{Men feeling their infirmities and miſeries,}
multiplied: afterward they
\SNote{make haſt in ſeeking remedies.}
made haſt.

I wil not aſſemble their
\SNote{Sacrifices to idols shal ceaſe in Gentiles.}
conuenticles of bloud: neither wil I be mindful of their
\SNote{Their names shal be changed from heathen to be called
Chriſtians.}
names by my lippes.

%%% o-1063
\V
\SNote{Eternal glorie cõſiſteth in ſeing God.}
Our Lord
\LNote{Our Lord the portion of myn inheritance.}{Chriſt
\MNote{Chriſt a King ſometimes exerciſed temporal iuriſdiction.}
whom the Iewes expected as an earthlie conquerour, that ſhould aduance
himſelf and them temporally in this world, was in dede, as the children
and multitude called him, king of Iſrael.
\XRef{(Ioan.~12.)}
At which time (as alſo before) he exerciſed temporal Iuriſdiction, in
correcting abuſes in the Temple
\XRef{(Mat.~21.}
\XRef{Ioan.~2.)}
And when Pilate demanded of him, if he were a king
\XRef{(Ioan.~18. v.~37.)}
he anſwered: \Emph{Thou ſaiſt, that I am a king.} For this I was borne,
and for this came I into the world, that I ſhould geue teſtimonie to the
truth. And though he anſwered withal, that his kingdom (to witte the
poſſeſſion, and vſe therof) was not of this world, yet Pilate by Gods
prouidence, writte the title, and would not alter it, \Sc{Iesvs} of
Nazareth King of the Iewes.
\MNote{God the proper inheritãce of Chriſt, and Chriſtians.}
But Chriſts chief inheritance, and reward of his merites is God himſelf,
as here he profeſſeth by his prophet Dauid: which is alſo the only true
& perfect inheritance of al Chriſts ſeruãtes, vvherfore
\MNote{Clergie men profeſſe expreſly to ſerue God, for God him ſelfe not
for temporal profite.}
Clergy men more particularly profeſſe the ſame, when they
firſt enter into their ſpiritual ſtate, addicting and dedication them
ſelues to ſerue God in Eccleſiaſtical function, not for temporal
inheritance, but for a better lotte, God himſelf, who is al Good, and
moſt perfect goodnes, true riches, and eternal inheritance. In which
election of ſtate to liue and ſerue God in, euerie Clergie man ſayth:
\Emph{Our Lord is the portion of myn inheritance, and of my cuppe: Thou
art he that vvil reſtore myn inheritance vnto me.} Man calleth
it \Emph{his} inheritance, becauſe he was created to ſerue God, and for
his ſeruice to inherite God: which reward though he loſt by ſinne, yet
euerie one returning to Gods ſeruice, and perſeuering therin, recouereth
by Chriſt, new right and title to the ſame inheritance, performing their
duties in their ſeueral vocations. Some traueling in the world, but not
louing it: others ſequeſtered from ſecular affayres, duly adminiſtring
ſacred offices, more peculiarly called Diuine ſeruice.}
the portion of myne inheritance, and of my
\SNote{God is the reward of ſuffering paines for Chriſt.}
cuppe: thou art he, that wil reſtore myne inheritance vnto me.

\V
\SNote{In diuiſion of temporal inheritance, land is meaſured by cordes:
as
\XRef{Ioſue.~10.}
ſo portions in heauen are geuen with large meaſure.}
Cordes are fallen to me in goodly places: for
\SNote{Chriſt alſo receiued al nations for his inheritance.}
mine inheritance is goodlie vnto me.

\V I wil bleſſe our Lord, who hath
\SNote{Wiſdome to make good election of ſpiritual thinges rather then
temporal.}
geuen me vnderſtanding:
%%% 1171
moreouer alſo euen til
\SNote{Not only by day, but alſo by night.}
night, my
\SNote{Alſo my corporal paines geue me inſtruction.}
veines haue rebuked me.

\V 
\CNote{\XRef{Act.~2. v.~25.}}
I
\SNote{Chriſt had God continually before his eyes: euerie man ought to
thincke frequently of God,}
forſaw our Lord in my ſight alwaies: becauſe he is
\SNote{for God ſtil protecteth the iuſt.}
at my right hand, that I be not moued.

\V For this thing my hart hath beene glad, and my tongue hath reioyced:
moreouer alſo my flesh shal reſt in hope.

\V
\CNote{\XRef{Act.~2.}
&
\XRef{13.}}
Becauſe thou wilt
\LNote{Not leaue my ſoule in hel.}{How
\CNote{\XRef{Pſal.~61. v.~11.}
\XRef{1.~Tim.~2. v.~4.}}
\MNote{Proteſtantes denying that Chriſt deſcended into \L{limbus}
tranſlate \Emph{graue} for \Emph{hel}.}
Caluin and Beza ſometimes corrupt this text, alwayes peruert the ſenſe,
and moſt abſurdly oppoſe them ſelues againſt al ancient holie Fathers,
concerning the Article of Chriſts deſcending in ſoule into that part of
hel called \L{Limbus patrũ}, is largely noted
\XRef{Gen.~37.}
\XRef{Act.~2.}
&
\XRef{1.~Pet.~3.}
Only here we may not omitte to aduertiſe the reader, that ſome
%%% !!! Here or with 'latter edition' below?
\CNote{\Cite{1552.}
\Cite{1577.}
\Cite{1603.}}
Proteſtants Bibles permitting the word \Emph{hel} to remaine in the
text, a latter Edition for \Emph{hel}, putteth \Emph{graue}, with this
only note in the former place, that \Emph{this is chiefly meant of
Chriſt by whoſe Reſurrection al his members haue immortality.} And
\XRef{Act.~2.}
they repete their new text by this paraphraſis: \Emph{Thou shalt not
leaue} me \Emph{in the graue.} VVreſting that which perteineth to the
bodie, riſing from the graue, to the ſoule, which was not at al in the
graue, al the time the bodie lay there.}
not leaue my ſoule in
\SNote{In \L{limbo patrum}.}
hel: neither wilt geue
\SNote{Chriſts body corrupted not in the graue.}
thy holie one to ſee corruption.

Thou haſt made
\SNote{Death and reſurrection is the way to life.}
the waies of life knowen to me, thou shalt make me ful of ioy with
\SNote{Perfect glorie conſiſteth in ſeing God,}
thy countenance: delectations on thy right hand, euen to
\SNote{in eternity.}
the end.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
