%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 020
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-020


\project douay-rheims


%%% 1180
%%% o-1072
\startChapter[
  title={Psalme 20}
  ]

\PSummary{Praiſe
\MNote{Chriſts exaltation.

The 5.~key.}
to God for Chriſts exaltation after his paſſion: 9.~and
  depreſſion of his enemies.}

\PTitle{Vnto
\SNote{Perteyning to the nevv Teſtament, principally to
\Fix{Chrriſt,}{Chriſt,}{obviious typo, fixed in other}
partly to godly and victorious kinges, and generally to al the bleſſed,
vvhich ouercome ſpiritual enemies.}
the end, the Pſalme of Dauid.}

\VV Lord in
\SNote{Chriſt our king as man ouercoming his enemies by his diuine
povvre, reioyceth in victorie.}
thy power the king shal be glad: and vpon thy ſaluation he shal reioyce
excedingly.

\V The
\SNote{Chriſts moſt ſpecial deſire was the ſaluation of his people.}
deſire of his hart thou haſt geuen him: and of the wil of his lippes
thou haſt not defrauded him.

\V Becauſe thou haſt
\SNote{This moſt principally verified in Chriſt, is alſo applied to
Martyres, vvhich ſuffer or are readie in preparation of mind to ſuffer
death for the truth.}
preuented him in bleſſinges of ſweetneſſe: thou haſt put on his head a
crowne of pretious ſtone.

\V He asked life of thee: and thou gaueſt him length of daies for euer,
and for euer and euer.

\V Great is his glorie in thy ſaluation: glorie and great beautie thou
shalt put vpon him.

\V Becauſe thou shalt geue him
\SNote{Al Sainctes receiue bleſſing of glorie, but only Chriſt imperteth
ſuch grace to others: for in him al are bleſſed that are eternally
glorified.
\XRef{Gen.~21.}
\XRef{Ioan.~1.}}
to be a bleſſing for euer and euer: thou shalt make him ioyful in
gladneſſe with thy countenance.

\V Becauſe the king hopeth in our Lord: and in the mercie of the Higheſt
he shal not be moued.

\V Let
\SNote{The iuſt cõforming their deſires to Gods vvil, do pray that the
vvicked may be punished. It is alſo a prophecie, that ſo it shal come to
paſſe.}
thy hand be found of al thy enemies: let thy righthand find al, that
hate thee.

\V Thou
\SNote{And ſo it folovveth in the next verſe: \Emph{Thou shalt put them
&c.}}
shalt put them as an ouen of fyre
\SNote{This alſo can not be vnderſtood of anie but of Chriſt, vvho in
the general or particular iudgement ſhal geue ſentence vpon the
vvicked,}
in the time of thy countenance: our Lord in his wrath
\SNote{and immediatly hel fire shal deuoure them.}
shal truble them, and fyre shal deuoure them.

\V Their
\SNote{Al the wicked, and namely the Iewes who perſecuted Chriſt, faile
in their deuiſes.}
fruite thou shalt deſtroy from the land: and their ſeede from the
children of men.

\V Becauſe thy haue turned the euils vpon thee: they haue
%%% 1181
deuiſed
%%% o-1073
counſels, which they could not eſtablish.

\V Becauſe thou shalt put them backe: in thy remnantes thou shalt
prepare their countenance.

%%% !!! Verse 13 begins here in other, not above. Which is right.
Be exalted Lord in thy powre:
\SNote{Whiles the wicked perish, the iuſt reioyce and praiſe God in
ſongue and Pſalme.}
we wil chaunte and ſing thy powers.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
