%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 050
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-050


\project douay-rheims


%%% 1234
%%% o-1125
\startChapter[
  title={Psalme 50}
  ]

\PSummary{King
\MNote{The fourth penitential Pſalme.

The 7.~key.}
Dauid in great ſorow for his ſinnes of adultrie and murder, moſt
ſeriouſly prayeth God of his manifold mercies to remitte and purge al
his offences, and paines doe for them: 12.~to reſtore vnto him the grace
of the Holie Ghoſt, loſt by his ſinnes; 15.~that he may teach others (as
in deede his ſingular example may teach the whole world true penance)
19.~contrition of hart, worthely to offer ſacrifice, for the whole
Church.}

%%% !!! Second verse in middle of title. Will this work?
\PTitle{Vnto
\SNote{Pertayning not only to Dauid, but alſo to al penitentes,
eſpecially of the new teſtament.}
the end, a Pſalme of Dauid, \V
\LNote{VVhen Nathan came to Dauid.}{As
\MNote{Temporal puniſhment is due after remiſſion of ſinnes.}
Nathan denouncing to Dauid that our Lord had (vpon his repentance and
confeſſion) taken away his ſinne, added neuertheles that becauſe he had
made the enimies of God to blaſpheme, his ſonne should dye: ſo Dauid
knowing that more was required then only confeſſion, for that the bond
of ſatisfaction remained after his ſinnes were remitted, perſiſted in
penance, praying, lamenting, and beſeching God according to his great
and manifold mercies, to take away his iniquitie, albeit the prophet
Nathan had now told him, that our Lord had taken away his ſinne, becauſe
there yet remained temporal paine due for the ſame. He prayeth alſo
\XRef{v.~4.}
that God wil, \Emph{vvash him more amply, from his iniquitie, and
cleanſe him from his ſinne}.
\MNote{Cuſtome of ſinne maketh more
\Fix{pronnes}{pronenes}{possible typo, fixed in other}
to fal againe.}
For albeit the guilt of mortal ſinne be washed and taken away, yet
beſides temporal punishment that is due, the ſoule that was ſo polluted,
nedeth to be washed, and cleanſed from the euil habite, or
\Fix{pronnes}{pronenes}{possible typo, fixed in other}
to fal againe, gotten by the former cuſtome, or delectation in ſinne.}
when Nathan the Prophet came to him, after that he had ſinned with
Bethſabee.
\XRef{(2.~Reg.~12.)}}

\VV Haue mercie on me ô God,
\SNote{My ſinnes being very great, nede thy great mercie.}
according to thy great mercie.

And according to
\SNote{Yea manie ſortes of thy mercies: not only remiſſion of the
crimes, but alſo mitigation of the paines doe for the ſame. Thy merciful
grace to be truly ſorie, to make ſome part of ſatisfaction, to beware
hereafter not to fal againe, to geue better example of penance, and of
vertuous life, and to perſeuer to the end.}
the multitude of thy commiſerations, take away myne iniquitie.

\V
%%% Marked in both, but missing
%%% \LNote{}{}
Wash me
\SNote{O God thou haſt forgeuen me, and taken away my ſinnes, as thy
prophet hath told me
\XRef{(2.~Reg.~12. v.~13.)}
but my ſoule ſo fouly polluted, nedeth yet more waſhing.}
more amply from mine iniquitie: &
\SNote{Cleanſe alſo the dregges that remaine, and al habites and
inclinations to ſinne. So our Sauiour afterwards taught.
\XRef{(Ioan.~13. v.~10.)}
He that is waſhed nedeth not ſauing to waſh his feete (il affections and
reliques of former ſinnes) but is cleane wholy.}
cleanſe me from my ſinne.

\V
\SNote{VVhiles I did not know; nor conſider nor acknowledge my ſinnes, I
could not be forgeuen, but now I know and acknowledge them:}
Becauſe I do know myne iniquitie: and my ſinne is
\SNote{and I ceaſe not to conſider of them with ſorow.}
before me alwaies.

%%% o-1126
\V To thee
\SNote{Principally (for ſo this particle [only] here ſignifieth) the
enormities of my ſinnes conſiſt, in that I haue offended thy Diuine
Goodnes and Maieſtie, the King of the worldes, immortal, inuiſible,
onlie God, to whom is due al honour and glorie for euer and euer.
\XRef{1.~Tim.~1. v.~17.}}
onely haue I ſinned, and haue done euil before thee: that thou mayſt
\SNote{Thou which haſt promiſed forgeuenes to al ſinners that truly
conuert, ſhal herein be iuſtified by receiuing me againe to grace:}
be iuſtified in thy words, and mayſt
\SNote{and ouerthrow thy calumniators, that iudge wickedly of thy
proceedings, as if either thy iuſtice or mercie were peruerted.}
ouercome when thou art iudged.

\V For behold
\LNote{I vvas conceiued in iniquitie.}{An
\MNote{Cõcupiſcence remaineth after original ſinne.}
other reaſon why ſinners after remiſſion of al mortal ſinnes, neede to
be washed, and cleanſed, is, becauſe being borne in original ſinne,
after remiſſion therof, there remaneth concupiſcence, that ſtriueth
againſt vertue, and inclineth to ſinne, from which we muſt pray, and
labour to be more and more washed and cleanſed.}
I
\SNote{I and al are borne in original ſinne, the reliques wherof,
concupiſcence and weakenes incline vs to other ſinnes, which we haue
added. In regard of which our infirmitie, thy mercie is readie to recal
vs, and help vs.}
was conceiued in iniquities: & my mother conceiued me in ſinnes.

%%% 1235
\V For behold thou
\SNote{Beſides thou haſt alſo geuen me knowledge of true faith, and
right doctrine, which thou euer loueſt, and art accuſtomed to reduce,
and direct ſuch into the true way of penance.}
haſt loued truth:
\SNote{Yea thou haſt moreouer ſhewed to me thinges vncertaine, or
vnknowen to manie others, geuen me the gift and ſpirit of prophecie, to
know hidde myſteries, and to euerie one God geueth ſome particular
benefites, which he loueth in him, and is ready of his part to confirme
and maintaine the ſame, that they be not loſt.}
the vncertaine, and hidden thinges of thy wiſdome thou haſt made
manifeſt to me.

\V Thou shalt ſprinkle me with
\SNote{Moſt merciful Lord thou wilt (as I ſee in the ſpirit of
prophecie) ſprinkle me, and al men with thy bloud, from the Croſſe,
where they ſhal geue thee vinegre about hyſſoppe to drinke,
\XRef{(Ioan.~19.)}}
hyſſope, and I shal be clenſed: thou shalt wash me, and I shal be made
\SNote{by which washing I shal be cleane from ſinne, and become in time
pure, yea whiter then ſnow. A figure of this hyſſope was obſerued in
Moyſes Law.
\XRef{Num.~19.}
ſignifying the liuelie heat of Chriſts infinite charitie.}
whiter then ſnow.

\V To
\SNote{When myn affections shal be cleane purged, I shal take ſingular
great delight to heare of thee,}
my hearing thou shalt geue ioy and gladnes, and
\SNote{and al my powres of mind and bodie, which are now afflicted, shal
reioyce.}
the bones humbled shal reioyce.

\V
\SNote{Leaue of thy cogitation of punishing, to which purpoſe firſt take
away myn iniquities, for otherwiſe if they remaine, Gods iuſtice can not
but punish them.}
Turne away thy face from my ſinnes: and wipe away al mine iniquities.

\V
\SNote{Create in me new grace, wherby my hart shal be pure. So S.~Paul
calleth a iuſt ſoul a new creature.
\XRef{Galat.~6. v.~15.}}
Create a cleane hart in me ô God: and renew a right ſpirit in my
\SNote{In my invvard thoughtes.}
bowels.

\V Caſt me not away from thy face: and thy Holie ſpirit
\SNote{Suffer me not ſo to fal againe, that thy grace depart from me.}
take not from me.

\V Render vnto me
\SNote{Which I had before my fal, of Chriſt promiſed of my ſeede, and
alter not the ſame for my ſinnes. Dauid alſo and other penitents pray
here, that God wil reſtore vnto them the ioy, which they had in the
ſtate of grace, of eternal ſaluation promiſed;}
the ioy of thy ſaluation, and
\SNote{confirme & conſerue in me hereafter, a ſtrong, conſtant, and
willing ſpirite to perſeuere.}
confirme me with the principal ſpirit.

%%% o-1127
\V I
\SNote{No way can a penitent better ſhew him ſelf gratful to God, for
remiſſion of his ſinnes, then by inſtructing, exhorting, and perſwading
other ſinners to repentance, to leaue their former il wayes, and turne
to God.}
wil teach the vniuſt thy waies: and the impious shal be conuerted to
thee.

\V Deliuer me
\SNote{From the guilt and punishment of murder, cauſing Vrias and others
with him to be ſlaine. Other penitents pray to be deliuered from what
ſinnes ſoeuer they haue committed, by ſheding bloud, or other wrongs and
iniuries; promiſing to praiſe Gods iuſtice, in offering and geuing
grace, according to his promiſe to ſinners, that they may repent.}
from bloudes ô God, the God of my ſaluation: and my tongue shal exult
[for] thy iuſtice.

%%% 1236
\V Lord, thou
\SNote{Thou ô God firſt ſturring me vp, opening my lippes, which of my
ſelfe I can not do, then my tongue and mouth wil praiſe thee.}
wilt open my lippes: & my mouth shal shew forth thy prayſe.

\V Becauſe if thou
\SNote{If thou wouldeſt eſpecially legal ſacrifice, I would eaſily haue
offered great ſtore:}
wouldeſt haue had ſacrifice, I had verily giuen it: with holocauſtes
thou wilt
\SNote{but the beſt of that kind is not ſufficient:}
not be delighted.

\V A
\LNote{Sacrifice.}{Holie
\MNote{Spiritual ſacrifice preferred before external.}
Scriptures make often compariſon betwen two kindes of ſacrifices,
preferring internal before external, as more gratful to God. And of
ſpiritual ſacrifices, this of a contrite ſpirite is firſt in order, and
maketh the way to the ſacrifice of iuſtice, becauſe iuſtice preſuppoſeth
repentance, and finally ſuccedeth ſacrifice of praiſe, and
thankeſgeuing.} 
ſacrifice to God is
\SNote{true contrition of hart pleaſeth thee farre better.}
an afflicted ſpirit: a contrite, and humbled hart, ô God thou wilt not
deſpiſe.

\V Deale fauourably ô Lord in thy good wil
\SNote{After a penitent hath remiſſion of his owne ſinnes, he muſt pray
for the whole Church.}
with Sion: that the walles of Ieruſalem may be built vp.

\V
\SNote{The Church proſpering, her faithful children shal offer}
Then shalt thou accept ſacrifice of
\SNote{the ſacrifice of iuſtice, rendering to euerie one that is due;}
iuſtice,
\SNote{alſo free offeringes without obligation,}
oblations,
&
\SNote{yea holocauſtes, which is the chiefeſt,}
holocauſtes:
\SNote{calues, and like hoſtes vpon the altar, according to the ſtate of
the old law: but in the law of Chriſt, the moſt B.~Sacrifice by him
inſtituted.}
then shal they lay calues vpon thyne altar.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
