%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 105
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% !!! The CNotes in this Pſalme are likely placed incorrectly and
%%% confused.



\startcomponent psalme-105


\project douay-rheims


%%% 1328
%%% o-1219
\startChapter[
  title={Psalme 105}
  ]

\PSummary{The
\MNote{The Iſraelites often ſinned, and were mercifully puniſhed.

The 4.~key.}
prophet exhorteth the people to render thankes and praiſes to God,
6.~for remitting their manifold ſinnes, in the deſert, 34.~and in the
conquered land: 38.~foreshewing like ſinnes to come, Gods wrath and
punishment for the ſame, 44.~and that he wil geue grace of repentance:
to ſome 47.~for which he prayeth, and praiſeth God.}

%%% 1329
\PTitle{Alleluia.}

\NV
\CNote{\XRef{Iudith.~13. v.~21.}}
Confeſſe ye to our Lord
\SNote{God is of himſelfe and eſſentially good. Al other goodnes is
participated of him.}
becauſe he is good becauſe his mercie is for euer.

\V
\SNote{No creature can fully expreſſe Gods perfections.}
Who shal ſpeake the powers of our Lord, shal make al his praiſes to be
heard?

\V
\SNote{It is a happie ſtate in this life, either to kepe Gods law:}
Bleſſed are they, that keepe iudgement, and
\SNote{or to repent and do worthie penance for tranſgreſſing.}
doe iuſtice at al time.

\V Remember vs ô Lord
\SNote{According to thy wonted beneuolence, promiſed to thy people.}
in the good pleaſure of thy people: viſite vs in thy ſaluation:

\V To
\SNote{That we may ſee, and enioy the benefites promiſed to thyne
elect.}
ſee in the goodnes of thyne elect, to reioyce in the ioy of thy nation:
that thou maiſt be prayſed with thine inheritance.

\V We haue
\SNote{VVe haue erred and hurt ourſelues:}
ſinned with our fathers: we haue
\SNote{we haue wronged our neighboures:}
dealt vniuſtly, we haue
\SNote{we haue offended againſt God.}
done iniquitie.

\V Our fathers in Ægypt did not vnderſtand thy meruelous workes: they
were not mindeful of the multitude of thy mercie.

And
\CNote{\XRef{Exod.~14. v.~11.}}
they prouoked thee to wrath going vp vnto the ſea, the Read ſea.

\V And he ſaued them
\SNote{Albeit the people by their murmuring deſerued more puniſhment, yet God
for the glorie of his owne name ſaued them from vtter deſtruction.}
for his name ſake; that he might make his power knowen.

\V And he rebuked the Read ſea, and it was made drie: and he led them in
the depths as in a deſert.

\V And he ſaued them from the hand of them that hated them: and
\LNote{He redeemed them.}{VVhat
\MNote{The read ſea a figure of Baptiſme.}
price (or ranſom)
%%% !!! Cite? On this place?
ſaith S.~Auguſtin, was geuen in this redemption? Or is it a prophecie,
that this was done in figure of Baptiſme, where we are redemed from the
hand of the diuel, by a great price, which is the bloud of Chriſt?
VVherupon it was more conueniently figured, not by what ſea ſoeuer, but by
the read ſea. For bloud hath redde coulor. And touching the effect of
Baptiſme deſtroying al former ſinnes, he teacheth in the
\Cite{expoſition of the next Pſalme}
(and either the ſame holie father, or ſome other good author,
\Cite{Ser.~42. de temp.)}
that as the Iſraelites paſſed ſafely through the read ſea, and al the
Ægyptians going in with them were drowned:
\MNote{Al former ſinnes deſtroyed in Baptiſme.}
ſo the baptiſed are ſaued in the water of Baptiſme, and al their ſinnes
are deſtroyed.}
he redemed them out of the hand of the enemie.

\V And water ouerwhelmed thoſe that afflicted them: there did not one of
them remaine.

\V And they beleued his wordes: and they ſang his praiſe.

%%% o-1220
\V
\CNote{\XRef{Exo.~15.}}
They had
\SNote{They perſeuered not long in their dutie towards God, ſeing his
omnipotent powre by his meruelous workes;}
quickly donne, they forgot his workes: and they
\SNote{nor were content with his prouidence, but carnally coueted
thinges, not neceſſarie.}
expected not his counſel.

%%% 1330
\V
\CNote{\XRef{Exo.~16. &~17.}}
And they coueted concupiſcence in the deſert: and tempted God in the
place without water.

\V He gaue them their petition: and ſent ſaturitie into their
\SNote{According to their carnal deſires.}
ſoules.

\V
\CNote{\XRef{Num.~16.}}
And they prouoked Moyſes in the campe: Aaron the
\SNote{Holie by his function.}
holie of our Lord.

\V The earth was opened, and ſwalowed Dathan: and ouerwhelmed the
congregation of Abiron.

\V And a fire flamed vp in their ſinagogue: the flame burnt the ſinners.

\V
\CNote{\XRef{Exo.~23.}}
And they made a calfe in Horeb: and they adored
\SNote{They adored the image that repreſented a calf, not God.}
the ſculptil.

\V And they
\SNote{God being their true glorie, they changed him, for a falſe god of
the Ægyptians (who eſpecially honoured a calfe called Apis) making an
image therof, and attributed their deliuerie from Ægypt to this imagined
god.
\XRef{Exo.~32. v.~48.}
Of which and the like fooliſh, and abominable idolatrie S.~Paul writeth,
\XRef{Rom.~1. v.~23.}
They changed the glorie of the incorruptible God, into a ſimilitude of
the image of a corruptible man, and of foules, and of foure footed
beaſtes, and of them that crepe: where we ſee what maner of imagies
holie Scripture condemne, and not the imagies of Chriſt and his
Sainctes.}
changed their glorie into the ſimilitude of a calfe that eateth graſſe.

\V They forgat God, which ſaued them, which did great thinges in Ægypt,
\V meruelous thinges in the land of Cham, terrible thinges in the Read
ſea.

\V
\CNote{\XRef{Exo.~32.}}
And
\SNote{He ſayd, he would deſtroy them, but for Moyſes prayer ſpared
them.}
he ſayd to deſtroy them: if Moyſes his elect had not ſtood in the way
before him:

To turne away his wrath that he should not deſtroy them: \V
\CNote{\XRef{Num.~14.}}
and they eſtemed for naught the land that was to be deſired.

They did not beleue his word, \V and they murmured in their
tabernacles: they heard not the voice of our Lord.

\V
\CNote{\XRef{Num.~14. v.~21.~22.}}
And he lifted vp his hand ouer them: to ouerthrowe them in the deſert:

\V And to caſt doune their ſeede among the Nations: and to diſperſe them
in the countries.

\V
\CNote{\XRef{Num.~25. v.~3.}}
And they were profeſſed to
\SNote{The Idol of Moabites, and
\Fix{Madianitees.}{Madianites.}{obvious typo, fixed in other}}
Beelphegor: and they did eate the ſacrifices
\SNote{As God is in deede the liuing God, that liueth of himſelfe, and
geueth life to others: ſo falſe goddes are called dead goddes, that can
not geue life to anie, but doe kil al that ſerue them: at leaſt
ſpiritually, and often corporally.}
of the dead.

\V And they prouoked him in their inuentions: and ruine was multiplied
on them.

%%% 1331
%%% o-1221
\V And Phinees ſtood, and
\SNote{Phinees moued by the zele of God (as the holie text witneſſeth.
\XRef{Nu.~25. v.~11.)}
in killing the adulterers pleaſed God, and merited reward.}
pacified: and the ſlaughter ceaſed.

\V And it was reputed to him vnto iuſtice, in generation and generation
euen for euer.

\V
\CNote{\XRef{Num.~20. v.~2.~12.}}
And they prouoked him at the waters of contradiction: and
\SNote{Moyſes afflicted in ſpirite, by the enormious murmuring of the
people, doubted whether God would geue them water out of the rocke or
no, not doubting of his powre, but of his wil: and ſo when he ſhould
haue ſpoken to the rocke,
\XRef{Num.~20. v.~8.}
he ſpoke to the incredulous people,
\XRef{v.~10.}
and therin offended God. For which he was temporally puniſhed.
\XRef{v.~12.}
\XRef{Deut.~1. v.~37.}
&
\XRef{c.~3. v.~26.}
\XRef{c.~4. v.~21.}}
Moyſes was vexed for them: \V becauſe they exaſperated his ſpirit.

And he playnely affirmed in his lippes: \V they deſtroyed not the
nations, of which our Lord ſpake to them.

\V
\CNote{\XRef{Deut.~2. v.~2.}
&
\XRef{12. v.~2. &~3.}}
And they were mingled among the nations, and learned their workes: \V
and they ſerued their ſculptils: and it became a ſcandal to them.

\V
\CNote{\XRef{Iudic.~2. v.~11.~12.}
\XRef{Iudic.~3. v.~5.~6. &.~c.}}
And they immolated their ſonnes, and their daughters to diuels.

\V
\CNote{\XRef{Iere.~19. v.~5.}}
And
\SNote{Some Iewes offered theſe moſt cruel, vnnatural, and abominable
ſacrifices: perhaps in the times of Iudges, when they were mingled with
idolatrous people, and ſerued their goddes: 
\XRef{Iud.~2. v.~12.}
\XRef{c.~3. v.~6.}
But it is more expreſſe after Dauids time, wherof he here prophecieth,
and was veryfied by Achaz.
\XRef{4.~Reg.~16. v.~3.}
and by Manaſſes
\XRef{4.~Reg.~21. v.~6.}
VVhich with other idolatrie king Ioſias deſtroyed.
\XRef{4.~Reg.~23. v.~1.}}
they shed innocent bloud: the bloud of their ſonnes and of their
daughters, which they ſacrificed to the ſculptils of Chanaan.

And the land was infected with bloud, \V and was contaminated in their
workes: and they did fornicate in their inuentions.

\V And our Lord was wrath with furie vpon his people: and he abhorred
his inheritance.

\V And he deliuered them into the handes of the nations: and they that
hated them, had the dominion of them.

\V And their enimies afflicted them: and they were humbled vnder their
handes: \V he did often deliuer them.

But they exaſperated him in their counſel: and they were humbled in
their iniquities.

\V And he
\SNote{God reſpected them with his merciful eye, and gaue them grace to
repent.}
ſaw when they were afflicted: and he heard their prayer.

\V And he was mindeful of his teſtament: and it repented him according
to the multitude of his mercie.

%%% 1332
\V And he gaue them into mercies in the ſight of al, that had taken
them.

\V
%%% !!! This first SNote belongs before \V
\SNote{Here the Pſalmiſt concludeth both the hiſtorie and prophecie of
this Pſalme, with prayer and praiſe, as foloweth:}
\SNote{A verie fitte prayer in time of ſchiſme.}
Saue vs ô Lord our God: and geather vs out of the Nations:

That we may confeſſe to thy holie name: & may glorie in thy prayſe.

\V Bleſſed be our Lord the God of Iſrael from euerlaſting vnto
euerlaſting: and al the people shal ſay: Be it, be it.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
