%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 038
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-038


\project douay-rheims


%%% 1213
%%% o-1105
\startChapter[
  title={Psalme 38}
  ]

\PSummary{A
\MNote{Gods prouidence.

The 3.~key.}
iuſt man in remediles perſecution reſolueth to ſuffer al with peace and
ſilence: 5.~praying God to take him from this world, confeſſing the
vanitie therof, 8.~and relying on Gods prouidence (11.~who punisheth man
for his ſinnes) prayeth for releaſe.}

%%% 1214
\PTitle{Vnto
\SNote{Some expound this Pſalme of the Iewes in captiuitie in Babylon,
but this title, and the matter conteyned shew, that it rather 
\Fix{peteineth}{perteineth}{possible typo, same in both}
to the new Teſtament.}
the end, to
\SNote{To be ſongue by Idithun and his ſcholars & ſucceſſors, or rather
by Chriſtianes.}
Idithun him ſelfe, a canticle of Dauid.}

\VV I
\SNote{Weake men in affliction not able to deliuer them ſelues, ſeeke
reuenge, by murmuring and other euil ſpeaches, but the perfect reſolue
to rule their tongues:}
haue ſaid: I wil keepe my waies: that I offend not in my tongue.

I haue ſet a gard to my mouth, when the ſinner ſtood againſt me.

\V I was dumme and humbled, and kept ſilence
\SNote{euen to forbeare ſometimes from their owne iuſt defence:}
from good thinges: and
\SNote{though therby they indure more perſecution.}
my ſorrow was renewed.

\V My
\SNote{Sorow ſuppreſſed maketh the hart to burne with zele, and
indignation.}
hart waxed hote within me: and in my meditation a fyre shal burne.

%%% o-1106
\V I haue ſpoken in my tongue: Lord
\SNote{If it may pleaſe thee let me know how long I shal liue, deſiring
to dye, as Elias deſired.
\XRef{3.~Reg.~19.}}
make mine end knowne to me.

And the number of my daies what it is: that I may know what is lacking to
me.

\V Behold thou haſt put my daies meaſurable, and my
\SNote{My life and al that I haue is as nothing compared to thy
eternitie.}
ſubſtance is as nothing before thee.

Doubtles al things are vanitie, euerie man liuing.

\V Surely man paſſeth as
\SNote{As a shadow or image appearing in a glaſſe, which is quickly
forgotte.}
an image; yea and he is trubled
\SNote{Therfore there is no cauſe man should be trubled in mind for
temporal miſeries.}
in vayne.

He gathereth treaſure; and knoweth not to whom he shal gather them.

\V And now what is my expectation? is not our Lord? and my ſubſtance is
with thee.

\V From al mine iniquities deliuer me:
\SNote{Thou haſt ſuffered me to be reproched by the foolish that proſper
in this world.}
a reproch to the foolish thou haſt geuen me.

\V I was dumme, and opened not my mouth,
\SNote{I know my tribulation is by thy prouidence.}
becauſe thou didſt it: \V Remoue thy ſcourges from me.

\V By the ſtrength of thy hand I haue faynted in reprehenſions: for
iniquitie thou haſt chaſtiſed man.

And thou haſt made his
\SNote{My life decayeth as a ſpider hauing ſpent al her moyſture.}
ſoule pyne away as a ſpider: but vaynly is euerie man trubled.

%%% 1215
\V Heare my prayer ô Lord, and my petition; with thyne eares receiue my
teares. Keepe not ſilence: becauſe I am a
\SNote{Al men are ſtrangers in this life, heauen being our home.}
ſtranger with thee, and a pilgrime, as my fathers.

\V Forgeue me, that I may
\SNote{That I may recouer ſpiritual ſtreingth in this life:}
be refreshed before I depart: and
\SNote{after which I shal not be in ſtate to do free workes of
ſatisfaction nor merite.}
shal be no more.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
