%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 57
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-57


\project douay-rheims


%%% 1244
%%% o-1135
\startChapter[
  title={Psalme 57}
  ]

\PSummary{Holie
\MNote{Gods prouidence in ſuffering euil.

The 3.~key.}
Dauid inueigheth againſt diſſembling wicked men, 7.~deſcribeth their
manifold punishment, 11.~wherin the iuſt shal be comforted.}

\PTitle{Vnto
\SNote{This Pſalme was made vpon the ſame occaſion, and to the ſame
purpoſe as the former,}
the end,
\SNote{to exhorte the iuſt and innocent to patience,}
deſtroy not, to Dauid in
\SNote{by Dauids memorable example.}
the inſcription.}

\VV If
\SNote{Few are ſo wicked, but they ſpeake and pretend iuſt thinges:}
in very dede you ſpeake iuſtice: iudge right thinges ye ſonnes of men.

\V For in
\SNote{but neither thincke wel,}
the hart you worke iniquities: in the earth your
\SNote{nor do wel, but both contrarie, which feaned ſanctitie is duble
iniquitie.}
handes forge iniuſtice.

\V
%%% !!! SNote not marked in either
\SNote{Theſe wicked ſinners that flatter and incite king Saul, ſeme to
haue ſpent al their life from their infancie in malice.}
Sinners are alienated from the matrice, they haue erred from the
wombe: they haue ſpoken falſe thinges.

%%% o-1136
\V They haue furie according to the ſimilitude of
\SNote{Their furie is vnquiet, til they may wound the innocent with
their poiſonful ſting,}
a ſerpent: as of the aſpe that is deafe, and
%%% !!! SNote not marked in either
\SNote{neither wil they harken to good admonitions, but ſtoppe their
eares like an aſpe, that layeth one eare cloſe to the ground, and
ſtoppeth the other with his taile.}
ſtoppeth his eares.

%%% 1245
\V Which wil not heare the voice of the inchanters, and of the ſorcerer
inchanting wiſely.

\V God shal breake their
\SNote{But God wil breake their cruel force,}
teeth in their mouth, the
\SNote{though it ſemeth moſt ſtrong, and inſuperable.}
cheeke tooth of the lions, our Lord wil breake in peeces.

\V They shal come to nothing as water running downe: he hath
\SNote{Gods iuſt determination of punishing the wicked ſtil remaineth
bent and readie, though execution be ſome while differred.}
bent his bow til they be weakened.

\V As
\SNote{That force and powre which is now inuincible, hard and ſtrong
like a lions ſtrongeſt teeth, shal then be as impotent and ſoft as waxe:}
waxe that melteth, shal they be taken away:
\SNote{Gods wrath, like fire, the moſt forcible element, shal fal vpon
them, and they shal be caſt into vtter darkenes, depriued of the ſunne
and al comfortable light.}
fyre hath falne on them, and they haue not ſeene the ſunne.

\V Before your
\SNote{Before their malice can bring to effect, the great miſchieffes
which they plotte and purpoſe, God ſuddainly cutteth them of before they
fully vnderſtand of, either ſicknes or death, caſteth them as it were
aliue into hel.}
thornes did vnderſtand the old bryar: as liuing ſo in wrath he
swalloweth them.

\V The iuſt
\SNote{The iuſt reioyce in the punishment of the wicked for three
cauſes, firſt in zele of iuſtice, conforming his wil and mind to Gods
iudgement; ſecondly, for that himſelfe through Gods mercie hath eſcaped
that terrible damnation; thirdly, for that he is now deliuered from
moleſtation, and continual tribulation.}
shal reioyce when he shal ſee reuenge: he shal wash his handes in the
bloud of a ſinner.

\V And man shal ſay: If certes there be fruite to the iuſt:
\SNote{The iuſt ſeing, or by faith knowing what punishment remaineth for
the wicked, is therby aſſured that the good shal reape fruict for his
wel doing, and that in the meane time God ruleth and iudgeth on the
earth, though as yet it appeareth not ſo euidently.}
there is a God certes iudging them on the earth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
