%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 134
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-134


\project douay-rheims


%%% 1381
%%% o-1271
\startChapter[
  title={Psalme 134}
  ]

\PSummary{God
\MNote{Gods perfection and goodnes.

The 1.~key.}
who only is omnipotent, and fountaine of al goodnes, is of al to be
praiſed. 8.~He is the ſpecial protector of his elected
people. 15.~Contrariwiſe falſe goddes are impotent, vaine, and can not
helpe thoſe that ſerue them. 19.~Only the Church doth rightly praiſe
him.}

\PTitle{Alleluia.}

\NV Prayſe ye the name of our Lord,
\SNote{Gods ſeruantes.}
ye ſeruantes prayſe our Lord.

\V Ye
\SNote{Clergie that ſerue in the Temple,}
that ſtand in the houſe of our Lord, in
\SNote{and ye Laitie that attend to his ſeruice, muſt praiſe him for
theſe cauſes.}
the courtes of the houſe of our God.

\V Prayſe ye our Lord, becauſe our Lord is
\SNote{His goodnes.}
good: ſing ye to his name, becauſe it is
\SNote{His benignitie.}
ſweete.

\V Becauſe our Lord
\SNote{He hath adopted vs his children.}
hath choſen Iacob to himſelf, Iſrael for his owne poſſeſſion.

\V Becauſe I haue knowne that our Lord is
\SNote{He is omnipotent, as appeareth by his vvorkes, here and elſvvhere
recited: and manie wayes knovvne to the vvorld.}
great, and our God aboue al goddes.

\V Al thinges whatſoeuer our Lord would he hath done, in heauen, in
earth, in the ſea, and in al the depthes.

\V Bringing forth clowdes from the vttermoſt of the earth: lightenings
he hath turned into rayne.

Who bringeth forth the windes out of his treaſures: \V who
\CNote{\XRef{Exo.~11.}}
ſtroke the firſt begotten of Ægypt from man euen to beaſt.

\V He ſent forth ſignes, and wonders in the middes of thee ô Ægypt: vpon
Pharao, and vpon al his ſeruantes.

\V Who ſtroke manie nations: and ſlew ſtrong kinges:

\V
\CNote{\XRef{Ioſ.~12.}}
Schon the king of the Amorrheites, &
\CNote{\XRef{Num.~2.}}
Og the king of Baſan, and
\SNote{\XRef{Ios.~13.}}
al the kingdomes of Chanaan.

\V And he gaue their land for inheritance, for an inheritance to Iſrael
his people.

%%% o-1272
\V Lord thy name is for euer: Lord thy memorial is vnto generation and
generation.

\V Becauſe our Lord shal iudge his people: and wil be intreated toward
his ſeruantes.

\V
\CNote{\XRef{Pſal.~11.}}
The idoles of the Gentiles are ſiluer, and gold, the workes of mens
handes.

\V They haue mouth, and shal not ſpeake: they haue eies, and shal not
ſee.

\V They haue eares, and shal not heare: for neither is there breath in
their mouth.

%%% 1382
\V Let them that make them become like to them: and al that haue
confidence in them.

\V Ye houſe of Iſrael bleſſe our Lord: ye houſe of Aaron bleſſe our
Lord.

\V Ye houſe of Leui bleſſe our Lord: you that feare our Lord, bleſſe ye
our Lord.

\V Bleſſed be our Lord out of Sion, who dwelleth in Ieruſalem.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
