%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 084
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-084


\project douay-rheims


%%% 1293
%%% o-1183
\startChapter[
  title={Psalme 84}
  ]

\PSummary{With
\MNote{Incarnation of Chriſt.

The 5.~key.}
commemoration of Gods former benefites, 5.~Chriſts Incarnation is
prophecied, 9.~bringing peace and ſaluation, 11.~mercie and iuſtice
concurring together.}

\PTitle{Vnto the end, to the children of Core, a Pſalme.}

\VV Ô Lord thou
\SNote{God beſtowed manie great benefites vpon the people of Iſrael:}
haſt bleſſed thy land: thou
\SNote{he brought them out of the bondage of Ægypt.}
haſt turned away the captiuitie of Iacob.

\V Thou
\SNote{Remitted their manifold ſinnes:}
haſt forgeuen the iniquitie of thy people: thou
\SNote{pardoned alſo a great part of due punishment.}
haſt couered al their ſinnes.

\V Thou haſt mitigated al thy wrath: thou haſt turned away from the
wrath of thine indignation.

\V
\SNote{As thou haſt ſpared thy peculiar people, ſo we beſech thee ô God
creator and general Sauiour of al mankind,}
Conuert vs ô God our ſauiour: and
\SNote{mitigate thy wrath towards vs al.}
auert thy wrath from vs.

\V Wilt thou be wrath with vs for euer? or wilt thou extend thy wrath
from generation vnto generation?

\V Ô God thou
\SNote{Til God firſt ſhew his mercie, ſinners lye dead in guilt of
ſinne, but by his grace they are ſturred vp, and quickned:}
being turned shalt quicken vs: and thy people
\SNote{and ioyfully returne to God.}
shal reioyce in thee.

\V Shew vs ô Lord thy mercie and giue vs thy ſaluation.

\V I wil
\SNote{The wordes of the prophet,}
heare what our Lord God wil ſpeake in me:
\SNote{ſignifying that God had reueled vnto him the redemption of
mankind.}
becauſe he wil ſpeake peace vpon his people.

And vpon
\SNote{Not al men are iuſtified, and ſaued, but thoſe that are hartely
and ſincerely conuerted.}
his ſaincts: and vpon them, that are conuerted to the hart.

%%% 1294
\V But yet his ſaluation is nigh to them that
\SNote{Though al be not ſaued (becauſe manie wil not cooperate to Gods
grace) yet very manie hauing the feare of God, which is the beginning of
godlie wiſdom, freely accept of Gods mercie, and ſo the Church is
gloriouſly propogated.}
feare him: that glorie may inhabite in our land.

%%% o-1184
\V
\SNote{VVheras Gods mercie would ſaue al, and his truth, or iuſtice
requireth that ſinnes be duly puniſhed, by Chriſts Paſſion and death,
ſufficient ſatisfaction is offered for al ſinnes, and thoſe that wil be
partakers by penance, and conformitie to Gods law, may haue remiſſion,}
Mercie and truth haue met each other:
\SNote{and ſo iuſtice is obſerued, and peace made betwen God and his
ſubiects.}
iuſtice and peace haue kiſſed.

\V
\SNote{Integritie of conſcience reigneth in good men,}
Truth is riſen out of the earth:
\SNote{God ſending iuſt meanes from heauen to ſaue them.}
and iuſtice hath looked downe from heauen.

\V For
\SNote{God geueth grace,}
our Lord certes wil geue benignitie: and
\SNote{and ſo men yeld fruict.}
our land shal giue her fruite.

\V
\SNote{Yea they walke in iuſtice, and right path of Gods law.}
Iuſtice shal walke before him: and shal ſet his ſteppes in the way.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
