%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 073
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-073


\project douay-rheims


%%% 1270
%%% o-1161
\startChapter[
  title={Psalme 73}
  ]

\PSummary{Faithful
\MNote{A prayer in long affliction.

The 7.~key.}
people preſſed with perſecution lamentably complayning, beſecheth God to
reſpect his owne inheritance, cruelly afflicted; 10.~and leift long
without helpe, 12.~wheras heretofore he releeued his people in like
diſtreſſes. 18.~And therfore confidently hopeth he wil reuenge the
blaſphemers of his name.}

\PTitle{Vnderſtanding
\SNote{Inſtruction for the congregation of the faithful in diſtreſſe,
not to ſeeke temporal rewardes and proſperitie, but to expect ſpiritual
and eternal.
%%% !!! Fix Cite
\Cite{S.~Aug.}}
to Aſaph.}

\NV Why haſt thou ô God
\SNote{In long perſecution the weake beginne to feare, or ſuſpect, that
God hath vtterly abandoned them.}
repelled for euer: is thy furie wrath vpon the sheepe of thy paſture?

\V Be
\SNote{But the ſtronger pray with confidence,}
mindful of thy congregation,
\SNote{being aſſured that God who hath conſerued his Church hitherto,
from the beginning of the world, wil conſerue it ſtil to the end.}
which thou haſt poſſeſſed from the beginning.

%%% 1271
Thou haſt
\SNote{An other reaſon why God wil conſerue his Church is, becauſe he
hath deliuered it often in a ſtrong arme; as from bondage in Ægypt, from
perſecution and oppreſſion in the time of Iudges, and kinges, and laſtly
redemed it from the tyrannie of the diuel by Chriſts death.}
redemed the rod of thine
\Fix{inheritancce:}{inheritance:}{Obvious typo, fixed in other}
mount Sion, in which thou haſt dwelt.

\V Lift vp thy handes vpon
\SNote{Thirdly the pride of the enimies moueth God to reuenge their
wickednes, and ſo to deliuer his owne people.}
their prides for euer: how great thinges hath the enimie done
malignantly in the holy place:

\V And they that hate thee haue gloried in the middes of
\SNote{They pretend to kepe ſolemne feaſtes like to thyne.}
thy ſolemnitie.

They haue ſette
\SNote{They ſet vp altars againſt thyn altares, or ſome new deuiſed
external ſhew, againſt the holie Rites of the Church:}
their ſignes, for ſignes: \V and haue not knowne,
\SNote{and that in publike places, in the endes of high wayes, or portes,
and pinnacles.}
as in the iſſue on high.

As
\SNote{The prophet deſcribeth the maner how perſecutors deſtroy temples,
and al ſacred thinges.}
in a wood of trees they haue with axes \V cut out the gates therof
together: in hatchet, and chippeaxe they haue caſt it downe.

\V 
\CNote{\XRef{4.~Reg.~25.}}
They haue burnt thy ſanctuarie with fire: they haue polluted the
tabernacle of thy name in the earth.

\V Their
\SNote{The whole crew, or band of perſecutors.}
kinred together haue ſaide in their hart: Let vs make al the feſtiual
daies of God to ceaſe from the earth.

%%% o-1162
\V Our
\SNote{The weake complaine that God ſheweth not wonted ſignes, nor
ſendeth ſuccourſe to his people, by raiſing ſome prophet, or other
meanes to helpe them.}
ſignes we haue not ſeene, there is now no prophet: and he wil know vs no
more.

\V How long ô God shal the enimie vpbraide: the aduerſarie prouoke thy
name for euer?

\V Why doeſt thou turne away thy hand, and thy right hand, out of the
middes of thy boſome for euer?

\V
\CNote{\XRef{Luc.~1. v.~70}}
But
\SNote{The perfect do anſwer, that God both hath, and wil releeue his
people.}
God our king before the worldes: he hath wrought ſaluation in the middes
of the earth.

\V Thou in thy ſtrength haſt confirmed
\SNote{Made the redde ſea like walles.}
the ſea: thou haſt
\SNote{Drowned Pharao and his hoſte.}
crushed the head of Dragons in the waters.

%%% 1272
\V Thou haſt broken the heads of the dragon: thou haſt giuen him for
meate to the peoples
\SNote{Of blacke diuels.}
of the Æthiopians.

\V Thou haſt
\SNote{Geuen water out of rockes.}
broken vp fountanes, and torrentes: thou haſt dried the riuers
\SNote{Made paſſage ouer Iordan.}
of Ethan.

\V The day is thine, and the night is thine: thou haſt made the morning
and the ſunne.

\V Thou haſt made al the coaſts of the earth: the ſummer and the ſpring
thou haſt formed them.

\V Be mindeful of this, the enimie hath vpbraided our Lord: and a
foolish people hath prouoked thy name.

\V Deliuer not to beaſts the ſoules that confeſſe to thee: and the
ſoules of thy poore forget not for euer.

\V Haue reſpect vnto thy teſtament: becauſe they that are
\SNote{Ignorant, blind, and barbarous infidels, lodge al kindes of
iniquitie in their conſciences;}
obſcure of the earth, are filled with houſes of iniquities.

\V
\SNote{and therfore it is vnmeete, that faithful people should be in
bondage vnder them.}
Let not the humble be turned away being confounded: the poore and needy
shal praiſe thy name.

\V Ariſe God, iudge thy cauſe: be mindful of thoſe thy reproches, that
are
\SNote{Theſe foolish audacious men do continually blaſpheme al holie
thinges.}
from the foolish man al the day.

\V
\SNote{Omite not to punish them,}
Forget not the voices of thine enimies: the
\SNote{ſeing they are obſtinate and obdurate in pride, and in hatred.}
pride of them that hate thee, hath aſcended alwaies.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
