%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-18


\project douay-rheims


%%% 1177
%%% o-1069
\startChapter[
  title={Psalme 18}
  ]

\PSummary{Gods
\MNote{Propagation of the Catholique faith.

The 6.~key.}
perfect goodnes and glorie is shewed by his great workes, and by his
  Apoſtles ſent with heauenlie commiſſion, to preach in al tongues to al
  nations. 6.~Chriſt coming into the world, and returne vnto
  heauen. 8.~His immaculate Law: 13.~wherin notwithſtanding the iuſt
  shal haue nede to pray for remiſſion of ſmaller and daylie ſinnes.}

\PTitle{Vnto
\SNote{Perteyning to the beloued of the new Teſtament.}
the end, the pſalme of Dauid.}

\VV The
\SNote{The ſilent workes of God declare his Maieſtie to them that
conſider therof; his preachers declare the ſame by wordes, to al that
wil heare.}
heauens shew forth the glorie of God, and the firmament declareth the
workes of his handes.

\V
\SNote{The cõſtant courſe of times sheweth that the ſame was diſpoſed by
Gods powre, and dayly propagation of faith, eſpecially of Chriſtian
doctrine sheweth Chriſts powre, & aſſured perpetuitie of his Church.}
Day vnto day vttereth word: and night vnto night sheweth knowledge.

\V There are
\SNote{Some of euery language or natiõ haue beleued in Chriſt, receiuing
the Catholique Religion.}
no languages, nor ſpeaches, whoſe voyces are not heard.

%%% o-1070
\V
\SNote{S.~Paul affirmeth that this hath bene fulfilled by the Apoſtles
preaching in al the world.
\XRef{Rom.~10.}}
Their ſound hath gone forth into al the earth; and vnto the endes of the
round world the wordes of them.

\V He put his tabernacle in
\SNote{By the ſunne a moſt excellent and ſuperelemental creature, the
Prophet deſcribeth Chriſt, making his courſe through this world,
illuminating, comforting, and ſtreingthning the Church his tabernacle,
wherein he perpetually dwelleth.}
the ſunne: & himſelf as a
\SNote{Chriſt the bridgrome, and the Church his bride are neuer
diuorſed; his loue, wiſdome, and powre, euer conſeruing her by his
immaculate law.}
bridgrome coming forth of his bridechamber.

%%% 1178
He hath reioyced as a giant to runne the way, \V
\CNote{\XRef{Luc.~24.}}
his comming forth from the toppe of heauen: And his recourſe euen to the
toppe therof: neither is there that can hide him ſelfe from his heate.

\V
\CNote{\XRef{Mat.~17.}
\XRef{Mar.~15.}}
\LNote{The lavv of our Lord is immaculate, conuerting ſoules.}{Gods
\MNote{God conuerteth and iuſtifieth ſoules inſtructing thẽ by his law,
and ſweetly drawing their free cooperation by grace.}
law in it ſelfe being moſt pure, and immaculate, is the proper meanes,
wherby the Holie Ghoſt conuerteth ſoules from ſinnes to iuſtice. Not
that euerie one is iuſtified, vvhich readeth, heareth, or knoweth the
lavv, but by keping it, through grace of the Holie Ghoſt. VVho is the
author and vvorker of iuſtice, by diſpoſing the ſoule to cooperate in
maner, partly here deſcribed: \Emph{Gods} faithful and moſt true
\Emph{teſtimonie}, by his lavv, \Emph{geueth vviſdome}, faith and
knovvledge to the humble; his \Emph{right iuſtices, comfort the hartes}
of the faithful, his \Emph{clere precept teacheth them} their dutie;
\Emph{his holie feare perſeuereth} vvith them; \Emph{his true
iudgements} are moſt delectable, and \Emph{deſirable} aboue al vvorldlie
riches or pleaſures. So by theſe and like ſpiritual motiues the ſoule
freely deſireth, accepteth, vndertaketh, endureth, and by continual
aſſiſtance of grace, obſerueth Gods lavv; and ſo meriteth the great
revvard, vvhich is promiſed for keping it.
\XRef{v.~12.}
\XRef{2.~Tim.~4. v.~6.}}
The law of our Lord is immaculate
\SNote{The old law was likewiſe pure in it ſelfe and holy, but the new
alſo maketh the obſeruers immaculate.}
conuerting ſoules: the teſtimonie of our Lord is faithful, geuing
wiſedome to litle ones.

\V The iuſtices of our Lord be right, making hartes ioyful: the precept
of our Lord lightſome; illuminating eies.

\V The feare of our Lord is holie, permanent for euer and euer: the
iudgementes of our Lord be true, iuſtified in themſelues.

\V To be deſired aboue gold and much pretious ſtone: and more ſweete
aboue honie and the honie combe.

\V 
\CNote{\XRef{Mat.~17.}
\XRef{Mar.~15.}}
For thy ſeruant
\SNote{How ſwete the law of God is, his ſeruantes finde not by reading,
or by hearing only, but by keping it.}
keepeth them, in keeping them is
\SNote{Conformably to this text the prophete profeſſeth
\XRef{(Pſal.~118. v.~12.)}
that he kept them for reward, in which place the heretikes tranſlation
is corrupted.}
much reward.

\V Sinnes
\SNote{None in this life knoweth perfectly his owne eſtate,
\Emph{vvhether he be vvorthie of loue or hate},
\XRef{Eccle.~9.}
but hopeth and feareth.}
who vnderſtandeth? from my ſecrete ſinnes cleanſe me: \V and from other
mens ſpare thy ſeruant.

If
\SNote{If mortal ſinnes haue not dominion in the ſoule it is iuſt: and
shal be in time immaculate from al ſinne.}
they shal not haue dominion ouer me, then shal I be immaculate; and shal
be cleanſed from the greateſt ſinne.

\V And the wordes of my mouth shal be ſuch as may pleaſe: and the
meditation of my hart in thy ſight alwayes.

Ô Lord
\SNote{Gods helping grace is ſtil neceſſarie to perſeuer,}
my helper, and my
\SNote{as his firſt grace redeeming man is
\Fix{neſſarie}{neceſſarie}{obvious typo, fixed in other}
for our firſt conuerſion.}
redemer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
