%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 042
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-042


\project douay-rheims


%%% 1220
%%% o-1111
\startChapter[
  title={Psalme 42}
  ]

\PSummary{The
\MNote{One God the B.~Trinitie.

The 1.~key.}
iuſt inuocateth Gods ſentence againſt the deceiptful, that ſeeke his
ſpiritual ouerthrow, 4.~acknowledgeth his helpe, from almightie God the
B.~Trinitie, in whoſe viſion glorie conſiſteth.}

\PTitle{A Pſalme
\SNote{Holie Dauid often prefiguring Chriſt, here repreſenteth euerie
faithful ſeruant of God, and particularly when they beginne a great and
holie worke; as when Prieſtes celebrate the diuine Sacrifice, they with
their aſſiſtantes recite by interchangeable verſes this Pſalme.}
of Dauid.}

\NV Iudge
\SNote{After that we haue examined, and prepared our ſelues to the moſt
holie Sacrifice and Sacrament, according to S.~Paules admonition (let a
man proue himſelfe, and ſo eate this bread, and drinke this chalice,
\XRef{1.~Cor.~11.)}
we pray God, to iudge betwen our true ſincere intention, and the vniuſt
\Fix{deceipful}{deceitful}{likely typo, fixed in other}
endeuoures of our enimie:}
me ô God, & diſcerne my cauſe from the nation not holie, from the vniuſt
and deceitful man
\SNote{and ſo to deliuer and protect vs from ſutle malice.}
deliuer me.

%%% o-1112
\V Becauſe thou are God
\SNote{VVith thee I can doe anie thing, without thee nothing.}
my ſtrength:
\SNote{Thou ſemeſt ſometimes not to regard me,}
why haſt thou repelled me? and why goe I ſorowful,
\SNote{whiles tentations are more ſenſible then thy grace.}
whiles the enemie afflicteth me?

\V Send forth
\SNote{As thou haſt ſent Chriſt the light and truth into this world,
grant vs the ſame now in particular.}
thy light and thy truth:
\SNote{Theſe two giftes of God, the light of knowing our duties and
truth, with ſincere intention to performe the ſame, haue brought vs into
thy Church, and vnto thy Altar.}
they haue conducted me, and haue brought me into thy holy hil, and into
thy tabernacles.

\V And
\SNote{Accompanied with light of truth, and ſincere intention, we
confidently approch to thyn Altar ô God,}
I wil goe in to the altar of God: to God, which
\SNote{vvho changeſt our old corruption into newnes of life.}
maketh my youth ioyful.

%%% 1221
\V I wil confeſſe to thee
\SNote{But to this purpoſe we praiſe God on the harpe, mortifying our
affections.}
on the harpe
%%% !!! This SNote includes the next.
\SNote{The former word is of the plural number in hebrevv, \HH{Eloim},
the other of the ſingular, ſignifying the Bleſſed Trinitie, one God.}
ô God
%%% \SNote{}
my God:
\SNote{Thou needeſt not therfore my ſoule, be penſiue, or deſolate.}
why art thou ſorowful ô my ſoule? and doſt thou truble me?

\V
\SNote{But truſt in God,}
Hope in God, becauſe yet
\SNote{praiſe him,}
wil I cõfeſſe to him: the ſaluation
\SNote{vvhom I hope to ſee face to face,}
of my countenance, and
\SNote{the true eternal God.}
my God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
