%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 075
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-075


\project douay-rheims


%%% 1274
%%% o-1164
\startChapter[
  title={Psalme 75}
  ]

\PSummary{The
\MNote{Gods prouidence towards his people.

The 3.~key.}
royal prophet ſingeth Gods praiſes, for his particular prouidence
towards the Iewes: 10.~further to be extended to al the meeke of the
whole earth.}

\PTitle{Vnto the end, in prayſes, a Pſalme
\SNote{For the congregation of faithful and godlie people.}
to Aſaph, a Canticle to
\SNote{Amongſt others, one notorious example of Gods prouidence is
recorded of an hundred foureſcore & fiue thouſand Aſſirians, ſlaine in
one night by an Angel.
\XRef{4.~Reg.~19.}}
the Aſſirians.}

\VV God is knowne
\SNote{God was not only knowen in general, as to pagane Philoſophers,
and ſome others, but more particularly to the Iewes, the iſſue of
Abraham, Iſaac, and Iacob, by his ſpecial benefites towards them.}
in Iewrie: in Iſrael his name is great.

\V And
\SNote{God ſuffering al other nations to erre for their ſinnes, in their
fond phantaſies of falſe goddes, reſerued the Iſraelites for his Church,}
his place is made in peace: and
\SNote{eſtabliſhing the principalitie therof in Sion.}
his habitation in Sion.

\V
\SNote{For obtaining and conſeruing of Sion from wicked Infidels, God
ouerthrewe al ſortes of contrarie forces.}
There he brake the powres of bowes, the shilde, the ſword, and the
battle.

\V Thou
\SNote{God not only gaue his people temporal victories ouer their
enimies, but alſo illuminated their mindes with knowlege of true
religion:}
doeſt illuminate meruelouſely from the eternal mountaynes: \V al the
foolish of hart
\SNote{others are often trubled in mind, hearing the truth, but are not
conuerted through their obſtinate follie.}
were trubled.

%%% !!! This line marked as a verse (but not in other) but this does not
%%% follow the verse numbers.
%%% \V
They
\SNote{VVorldlie men in ſupine carleſnes as in a ſleepe, paſſe ouer this
life, and afterwards find themſelues excluded from heauen, for lacke of
merites, and good workes, with the fooliſh virgins.}
ſlept their ſleepe: and al the men of riches found nothing in their
handes.

%%% o-1165
\V At
\SNote{God vndertaking the defence of his people, and threatning the
aduerſe part,}
thy reprehenſion ô God of Iacob, they
\SNote{they failed in courage, as men ouercome with drowſines of
ſleepe.}
haue al ſlumbered that mounted on horſes.

\V Thou art terrible, and who shal reſiſt thee?
\SNote{Euen from the firſt notice of thy wil, the aduerſaries were
deiected, fearing thy potent wrath.}
from that time thy wrath.

%%% 1275
\V From
\SNote{Terrible ſignes from the firmament appearing before the day of
Iudgement,}
heauen thou
\MNote{The Prophetes do often ſpeake in the pretertence, for the
aſſurance of the thinges to come.}
haſt made thy iudgement hearde:
\SNote{perſecutors and others being terrified ſhal be aſtonied and
ſilent.}
the earth trembled and was quiet.

When God aroſe vnto iudgement,
\SNote{God vvil come to iudge the vvorld, more eſpecially for the iuſts
ſake.}
that he might ſaue al the meeke of the earth.

\V Becauſe
\SNote{Men that ſhal ſeriouſly thinke and meditate vpon theſe thinges,
vvil praiſe and thanke God for them:}
the cogitation of man shal confeſſe to thee: and
\SNote{and the effect, and ſvvete repaſt of ſuch meditations, ſhal make
as it vvere a great feſtiual day, in the deuout ſoule ſpiritually ioyned
vvith God.}
the remanes of the cogitation shal keepe feſtiual day to thee.

\V
\SNote{The ſoule thus inflamed with Gods loue, is then apt of gratitude,
for his goodnes tovvardes man, to make vovves of thinges vvherto vve are
not obliged,}
Vowe ye, and
\SNote{but moſt neceſſarie it is to be maturely aduiſed, and not raſhly,
nor lightly to vovv; for being once vovved, vve are ſtrictly bound to
vvhatſoeuer vve haue lavvfully promiſed. And it is great ſinne to vovv
vndiſcretly.}
render to our Lord your God: al ye that round about him bring giftes.

To
\SNote{Remembring that for vovves, and al other vvorkes vve muſt anſvver
to God, vvho is a terrible iudge, readie to punish in bodie and ſoule,
ſparing none for their greatnes, not princes, nor kinges, nor vulgar
ſorte, for al are to him alike.}
the terrible, \V and him that taketh away the ſpirite of princes,
terrible to the kinges of the earth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
