%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 056
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-056


\project douay-rheims


%%% 1243
%%% o-1134
\startChapter[
  title={Psalme 56}
  ]

\PSummary{The
\MNote{Dauids great patience.

The 8.~key.}
Prophet prayeth in tribulation, 4.~teſtifieth Gods helpe,
  6.~praiſeth his greatnes: 8.~promiſing, and inuiting al nations to praiſe
  him.}

\PTitle{Vnto
\SNote{The heroical facts of Dauid are for examples to al Chriſtians.}
the end,
\SNote{Innocent Dauid hauing opportunitie to kil his vniuſt perſecutor,
obeyed the motion of God, ſuggeſting vnto him, not to deſtroy his
enimie; contrarie to the counſel of his freindes:}
deſtroy not, to Dauid in
\SNote{a thing moſt worthie to be recorded for perpetual memorie,}
the inſcription of the title;
\SNote{being in ſo great and vniuſt truble, as to lye in the caue of a
mountaine, yet ſpared to kil, or hurt him, that driue him into ſuch
ſtraictes. See the hiſtorie
\XRef{1.~Reg.~24.}}
when
\CNote{\XRef{1.~Reg.~22,~24.}}
he fled from the face of Saul into the caue.}

\VV Haue mercie on me ô God, haue mercie on me: becauſe my ſoule hath
truſted in thee.

And I wil hope in the shadow of thy winges, vntil iniquitie paſſe.

\V I wil crie to God the higheſt: God that hath done me good.

\V He ſent
\SNote{Extraordinarie diuine helpe, paſſing mans powre.}
from heauen, and deliuered me: he hath geuen into
\SNote{It fel reprochfully to Saul, that Dauid might haue ſlaine him if
he would, yet did neither hurt him, nor inſult vpon him, but meekely and
piouſly admonished him of his error, and iniurious perſecution.}
reproche them that trode vpon me.

%%% o-1135
God hath ſent his mercie, and his truth, \V and hath deliuered my
\SNote{My life,}
ſoule out of the middes of
\SNote{from moſt mightie and rauenous perſecuters.}
Lions whelpes: I ſlept trubled.

The ſonnes of men, their
\SNote{Though they haue not lions natural teeth, yet they exerciſe
crueltie by artificial weapons,}
teeth are weapons and arrowes: and their
\SNote{and with their cruel tongues incite their folowers to the ſame
furie.
\XRef{1.~Reg.~22. v.~16.}}
tongue a sharpe ſword.

%%% 1244
\V Be exalted aboue the heauens ô God: and thy glorie vpon al the earth.

\V They prepared a ſnare for my feete: and bowed downe my ſoule.

They
\SNote{Saul endeuored manie wayes to ouerthrow Dauid, amongſt other
meanes prouoked him to ſette vpon the Philiſtims, thincking they should
haue ſlaine him,
\XRef{1.~Reg.~18. v.~17.}
but the ſame Philiſtims ouerthrew Saul.
\XRef{1.~Reg.~31.}}
digged a pit before my face: and they are falne into it.

\V My hart is readie ô God, my hart is readie: I wil ſing, and ſay
\SNote{Pſalmes more in vſe with Chriſtian Gentils, then they were with
the Iewes. See
%%% !!! Fix me
\XRef{page.~12.}}
a Pſalme.

\V Ariſe my glorie, ariſe pſalter and harpe: I wil ariſe early.

\V I wil confeſſe to thee among
\SNote{Gods benefites beſtowed vpon Dauid, and vpon faithful Chriſtians
prefigured by him, are for euer to be praiſed by al peoples and
nations.}
peoples ô Lord: and I wil ſay
%%% !!! This Snote refers to the above about 'a Pſalme'
%%% \SNote{}
a Pſalme to thee among
%%% !!! This Snote refers to the above about 'peoples'
%%% \SNote{}
the Gentiles.

\V Becauſe thy mercie is magnified euen to the heauens, and thy truth
euen to the cloudes.

\V Be \Fix{axalted}{exalted}{obvious typo, fixed in other}
aboue the heauens ô God: and thy glorie vpon al the earth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
