%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 127
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-127


\project douay-rheims


%%% 1375
%%% o-1265
\startChapter[
  title={Psalme 127}
  ]

\PSummary{Happines
\MNote{Feare of God the vvay to happines.

The 7.~key.}
both of this life and of the next, is obtayned by fearing, and ſincerely
ſeruing God.}

\PTitle{A gradual Canticle.}

\NV Bleſſed are al that feare our Lord, that walke in his wayes.

\V Becauſe thou shalt eate the labours of thy handes: bleſſed art thou,
and it shal be wel with thee.

\V Thy
\SNote{If ſuch ſeruants of God be maried, they ſhal ordinarily haue
iſſue, and ſucceſſion in their familie; but eſpecially the ſoules of
ſuch ſhal bring forth manie meritorious vvorkes,}
wife as a fruitful vine, in
\SNote{in the boſome of the Catholique Church, vvhich vvas founded in
Chriſts ſide.}
the ſides of thy houſe.

Thy
%%% !!! SNote should go before 'Thy'
\SNote{Children alſo ſignifie good workes.}
children as young plantes of oliuetrees, round about thy table.

\V Behold ſo shal the man be bleſſed, that feareth our Lord.

\V Our Lord out of Sion bleſſe thee: and that thou mayſt ſee the good
thinges of Ieruſalem al the dayes of thy life.

And that thou mayſt ſee thy
\SNote{Revvard in heauen for good vvorkes in earth.}
childrens children, peace vpon Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
