%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 019
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-019


\project douay-rheims


%%% 1179
%%% o-1071
\startChapter[
  title={Psalme 19}
  ]

\PSummary{Faithful
\MNote{The ſubiects prayer for their ſuperior.

The 7.~key.}
people ioyne their prayer, for their Prince or
  Prelate, 4.~with ſacrifice offered for his preſeruation, and
  proſperous ſucceſſe.}

\PTitle{Vnto
\SNote{Though ſuch a prayer was very fitly made for Dauid, Ezechias, or
other kinges of Iuda, yet it more properly ſerueth for Chriſtian Kinges
and Prelates.}
the end, the Pſalme of Dauid.}

\VV Our Lord heare
\SNote{The king, or other ſuperior praying for him ſelfe, his ſubiects
pray with him, and for him. It may alſo be applied to Chriſt, praying
whiles he vvas mortal, or novv praying for his myſtical body the
Church, but in both theſe caſes, our prayers are only neceſſary for his
ſeruantes not for him.}
thee in the day of tribulation: the name of the God of Iacob protect
thee.

\V Send he ayde to thee from the holie place: and from Sion defend he
thee.

\V Be he mindeful of al thy
\SNote{The Hebrevv vvord \HH{Minca} ſignifieth ſacrifice of floure, and
vnbloudy, ſo perteyneth to the Euchariſt, in forme of bread, and vvine.}
ſacrifice: and be thy holocauſt made
\SNote{Be acceptable for him, for vvhom it is offered.}
fatte.

\V Geue he vnto thee according to thy hart: and confirme he al thy
counſel.

\V We shal  reioyce
\SNote{In the proſperous ſtate of thee our ſuperiour.}
in thy ſaluation: and in the name of our God we shal be magnified.

\V Our Lord accomplish al thy petitions: now haue I knowen that our Lord
hath ſaued
\SNote{His anointed king, or Prieſt. Myſtically faithful people
acknovvlege the victory of Chriſt our Sauiour ouer death and al
enemies.}
his \Sc{Christ}.

He shal heare him from his holie heauen: the ſaluation of his righthand
is,
\SNote{In great ſtreingth, or heauenlie forces.}
in powers.

\V Theſe in chariotes, and theſe in horſes: but we wil inuocate in the
name of the Lord our God.

\V
\SNote{Men truſting in humane and worldly powre fal into captiuitie.}
They are bound, and haue fallen: but we haue riſen and are ſet vpright.

\V Lord
\SNote{By protection of the head the body is alſo conſerued.}
ſaue the king, and heare vs in the day, that we shal inuocate thee.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
