%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 51
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-51


\project douay-rheims


%%% 1237
%%% o-1128
\startChapter[
  title={Psalme 51}
  ]

\PSummary{Holie
\MNote{Dauids inuectiue againſt Doeg.

The 8.~key.}
Dauid inueigheth againſt wicked Doeg a traitor, 7.~prophecieth his
ruine, 10.~and his owne exaltation.}

\PTitle{Vnto the end, vnderſtanding to Dauid, \V when Doeg
\SNote{Of the race of Eſau, half a Ievv, but either an Infidel, or
fautor of Infidels, a ſpie for Saul, a perſecutor of Dauid, & a murderer
of Innocents.
\XRef{2.~Reg.~22. v.~9.~18.}}
the Idumeite came and told Saul: Dauid is come into the houſe of
\SNote{High prieſt, ſlaine with 84.~more Prieſtes and others, becauſe
they were ſuppoſed to fauour Dauid.
\XRef{ibidem.}}
Achimelech.
\XRef{(1.~Reg.~22.)}}

\VV Why doeſt thou
\SNote{Thou perſecutor Doeg, why art thou ſo malicious, to abuſe thy
credite with king Saul, to the murdering of innocents?}
glorie in malice, which art mightie in iniquitie?

\V Al the day hath thy tongue thought iniuſtice: as a sharp raſor thou
\SNote{Playing the part of a ſpie, in betraying to Saul, that I was with
Achimelech?}
haſt done guile.

\V Thou haſt loued malice more then benignitie:
\SNote{Though he told a truth, yet it was iniquitie to betray
innocents.}
iniquitie rather then to ſpeake equitie.

\V Thou haſt loued al wordes of precipitation, a deceitful tongue.

\V Therfore wil God deſtroy thee for euer, he wil
\SNote{Thou shalt vtterly be deſtroyed,}
plucke thee out, & remoue thee out of thy tabernacle: &
\SNote{& al thy race.}
thy roote out of the land of the liuing.

\V The iuſt shal ſee, and feare, and shal laugh at him, and they shal
ſay:
%%% o-1129
%%% !!! Verse not properly marked in either
\V Behold the man, that hath not put God for his helper;

But hath hoped in the multitude of his riches, and hath
\SNote{For a short time in this world.}
preualed in his vanitie.

\V But I as
\SNote{Dauid prophecieth his owne exaltation, and conſeruation of his
ſeede in the kingdom of Iſrael.}
a fruitful oliue tree in the houſe of God, haue hoped in the mercie of
God for euer: and for euer and euer.

\V I wil
\SNote{Sing praiſe, and thankes to thee.}
confeſſe to thee for euer, becauſe thou haſt done it: and I wil expect
thy
\SNote{Thy Goodnes, which agreeth to thy name.}
name, becauſe it is good in the ſight of thy ſaints.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
