%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 039
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-039


\project douay-rheims


%%% 1215
%%% o-1106
\startChapter[
  title={Psalme 39}
  ]

\PSummary{Chriſts
\MNote{Chriſts comming and redeming of mankind.

The 5.~key.}
faithful members after long expectation congratulate his coming in
flesh. 6.~He directing his ſpeach to his Father, profeſſeth to performe
the Redemption of mankind, and to denounce the ſame in the whole world:
12.~prayeth for his ſeruantes, vndertaking to ſatisfie for their
ſinnes.}

\PTitle{Vnto
\SNote{Perteyning vnto the new Teſtament.}
the end, a Pſalme to Dauid himſelfe.}

\VV Expecting
\SNote{The faithful of the old and new Teſtament reioyce in the coming
of Chriſt.}
I expected our Lord, and he hath attended to me.

%%% o-1107
\V And he heard my prayers, and brought me out of the lake of miſerie,
and from the myre of dregges.

And hath ſet my feete vpon a rocke: and hath directed my ſteppes.

\V And he hath put a new canticle into my mouth: a ſong to our God.

Manie shal ſee, and shal feare: and they shal hope in our Lord.

\V Bleſſed is the man, whoſe hope is in the name of our Lord: and hath
not had regard to vanities and falſe madnes.

\V Thou haſt done manie thy meruelous thinges ô Lord my God: & in thy
cogitations there is none that may be like to thee.

I
\SNote{Chriſt by him ſelfe and by others preached the Goſpel of
ſaluation.}
haue declared and haue ſpoken: they
\SNote{The multiplication of Chriſtians therby.}
multiplied aboue number.

\V 
\CNote{\XRef{Heb.~10.}}
\SNote{No ſacrifice of the old Teſtament ſufficed to ſatisfie Gods
iuſtice for the ſinne of man.}
Sacrifice and oblation thou wouldeſt not: but
\SNote{Chriſt by the eare of obedience performed the redemption of man
by his death, as was determined from eternitie. S.~Paul for \Emph{eares}
ſaith \Emph{bodie}. See
\XRef{Annotations Heb.~10.}}
eares thou haſt perfited to me.

Holocauſt and for ſinne thou didſt not require: \V then ſaid I: Behold I
come.

%%% 1216
In the
\SNote{The ſumme of holie Scripture is of Chriſts Incarnation & death
for redemption of man.}
head of the booke it is written of me, \V that I should doe thy wil: my
God I would, and thy law in the middes of my hart.

\V I haue
\SNote{Againe Chriſt inculcateth the preaching and receiuing of his
Goſpel in the whole world.}
declared thy iuſtice in the great Church, loe I wil not ſtay my lippes:
Lord thou haſt knowen it.

\V Thy iuſtice I haue not hid in my hart: thy truth and thy ſaluation I
haue ſpoken. I haue not hid thy mercie, and thy truth from the
\SNote{In the greateſt and wiſeſt congregations of this world, Chriſt
concealeth not his mercie and truth. So himſelfe profeſſed before Annas,
Caiphas, Pilate, and their councels. S.~Paul preached Chriſt at Athens,
and in manie nations, and ſo the other Apoſtles. For their voice went
into al the coaſtes of the earth.}
great councel.

\V
\SNote{The prophet now ſpeaketh in the name of Chriſts myſtical bodie
the Church: praying to be made partaker of mercie, and to be deliuered
from euiles.}
But thou ô Lord make not thy commiſerations farre from me: thy mercie
and thy truth haue alwayes receiued me.

\V Becauſe euils haue compaſſed me, which haue no number: mine
iniquities haue ouertaken me, & I was
\SNote{The ſinnes alſo of thoſe which beleue in Chriſt are ſo manie,
that they can not be fully ſene in particular.}
not able to ſee.

They are multiplied aboue the heares of my head: and
\SNote{I almoſt faint in conſidering ſo manie and ſo great iniquities
amongſt thoſe that profeſſe Chriſt.}
my hart hath forſaken me.

%%% o-1108
\V It may pleaſe thee ô Lord to deliuer
\SNote{The whole Church prayeth in the name of al for the infirme
members.}
me: Lord haue reſpect to helpe me.

\V Let
\SNote{The prophet foresheweth that the reprobate for their obſtinate
malice, ſeeking to hurt others shal be confounded.}
them be confounded and ashamed together, that ſeeke my ſoule, to take it
away.

Let them be turned backward, and be ashamed that wil me euils.

\V Let them forthwith receiue their confuſion, that
\SNote{That ſcornfully ſay: Wel, wel: wishing al euil to good men.}
ſay to me: Wel, wel.

\V Let al
\SNote{Which not only in mouth and outward profeſſion, but alſo in
ſinceritie of hart ſeeke thee, may with confidence reioyce, and praiſe
God.}
that ſeeke thee, reioyce and be glad vpon thee: and let them that loue
thy ſaluation, ſay alwayes: Our Lord be magnified:

%%% 1217
\V But
\SNote{Chriſt ſpeaketh in the name of ſinners truly repenting, whoſe
ſinnes he vndertaketh to redeme, and waſh away by his paſſion.}
I am a begger, and poore: Our Lord is careful of me.

Thou art
\SNote{The faithful of the old Teſtament pray for Chriſts firſt coming
into this world, and the faithful now pray for his ſecond coming to
purge his Church, and to reward the good.}
my helper, & my protector: my God be not ſlacke.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
