%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 110
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-110


\project douay-rheims


%%% 1339
%%% o-1229
\startChapter[
  title={Psalme 110}
  ]

\PSummary{Praiſe
\MNote{Graces geuen to the Church.

The 6.~key.}
of God for benefites, 4.~eſpecially for the B.~Sacrament of the
Euchariſt, 6.~with other graces imparted to the Catholique Church.}

\PTitle{Alleluia.}

\NV I
%%% !!! SNote should go before 'I'
\SNote{This Pſalme in the Hebrew is compoſed with euerie verſe and
middle of verſe, beginning with a diſtinct letter, in order of the
Alphabet.}
wil confeſſe to thee ô Lord with al my hart: in
\SNote{I wil praiſe God both in ſecret for diſcharge of myn owne
conſcience,}
the counſel of the iuſt, and
\SNote{and in publique for edification of others.}
the congregation.

%%% 1340
\V The workes of our Lord are great: exquiſite according
\SNote{Gods wil is the whole cauſe of al his workes.}
to al his willes.

\V
\SNote{Euerie worke of his is praiſe worthie, and magnifical.}
Confeſſion and magnificence his worke: and his iuſtice contineweth for
euer and euer.

\V He hath made
\SNote{God hath leift one moſt ſpecial and beneficial memorie of al
other benefites, his owne bodie and bloud, in memorie of his Paſſion,
and our redemption,}
a memorie of his meruelous workes: a merciful and pittiful Lord: \V he
hath geuen
\SNote{the ſpiritual foode and ſuſtinance of al the ſoules that rightly
feare him.}
meate to them that feare him.

%%% o-1230
He wil be mindful for euer
\SNote{Of his promiſe to conſerue his Church perpetually.}
of his teſtament: \V
\SNote{The powrable operation of his death, and of al his myſteries.}
the force of his workes he wil shew forth to his people:

\V To geue them the inheritance of the gentiles: the workes of his
handes truth, and iudgement.

\V Al his commandmentes
\SNote{Gods commandments do iuſtifie al that kepe them.}
are faithful: confirmed for euer and euer, made in truth and equitie.

\V He ſent
\SNote{He alſo of his mercie redemed man; that he might be able to kepe
his precepts.}
redemption to his people: he commanded his teſtament for euer.

Holie, and terrible is his name: \V
\SNote{Beginning with feare of God, bringeth at laſt by other degrees to
true wiſedom, which two are the firſt and laſt of the ſeuen giftes of
the Holie Ghoſt.}
the feare of our Lord is the beginning of wiſedom.

\V Vnderſtanding is good to al that doe it: his prayſe remaineth for
euer and euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
