%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 046
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-046


\project douay-rheims


%%% 1227
%%% o-1118
\startChapter[
  title={Psalme 46}
  ]

\PSummary{Gentiles
\MNote{Vocation of Gentiles.

The 6.~key.}
are called, and inuited to praiſe God for his magnificence: 6.~for
Chriſts Aſcenſion, and powre.}

\PTitle{Vnto the end, for
\SNote{For Chriſtians that leaue the ſinnes of their fathers, and
reioyce in Chriſt crucified: See
\XRef{Annotation. Pſal.~41.}}
the ſonnes of Core.}

\VV Al ye Nations
\SNote{True ioy of the hart ſheweth it ſelfe both in voice of
exultation, and alſo in geſture of body, by clapping of handes, dancing
(as king Dauid did before the Arke.
\XRef{2.~Reg.~4.)}
likevviſe vvith inſtruments.}
clappe handes: make iubilation to God in the voyce of exultation.

\V Becauſe our Lord is high,
\SNote{To al the wicked.}
terrible; a great king ouer
\SNote{Not only of one or few kingdoms, but of al the earth.}
al the earth.

%%% o-1119
\V He hath made peoples ſubiect
\SNote{VVhen kinges, and countries become Chriſtians, they are made
ſubiectes to the Church that vvas before, not heades and rulers
therfore.}
to vs: & gentiles vnder our feete.

\V He hath choſen his inheritance in vs: the beautie of Iacob which he
loued.

\V
\SNote{Chriſt, God & man, after his Paſſion, roſe from death and
aſcended:}
God is aſcended in
\SNote{not leauing his Church deſolate, but making her ioyful by an
other comforter the Holie Ghoſt.}
iubilation: and our Lord in the voice of trumpet.

%%% 1228
\V Sing ye to our
\SNote{The ſame Chriſt is our God, by his Diuinitie:}
God, ſing ye: Sing ye to our
\SNote{and our king by his Humanitie.}
king, ſing ye.

\V Becauſe God is king of al the earth: ſing ye
\SNote{Doe your endeuour to vnderſtand vvhat you ſing, read, or heare in
Gods word. At leaſt to know the principal Myſteries, and pointes of
Chriſtian doctrin, euerie one according to their capacitie and ſtate of
profeſſion.}
wiſely.

\V God shal reigne ouer the gentiles: God ſitteth vpon his holie ſeate.

\V Princes of peoples are gathered together with
\SNote{The faithful of the old and nevv Teſtament are vnited in the
ſeruice of one, and the ſame eternal God.}
the God of Abraham: becauſe the ſtrong
\SNote{In reſpect of the Bleſſed Trinitie, holie Scripture here, and in
manie places vſeth names of the plural number as \HH{Eloim} Goddes,
not diuiding Gods ſubſtance,
vvhich is one, but inſinuating diſtinction of Diuine Perſons. The
Father, the Sonne, and the Holie Ghoſt. VVhich Myſterie is more expreſly
mentioned in Baptiſme, and profeſſed by Chriſtian gentils, then it was
by the people of the Ievves.}
goddes of the earth, are excedingly aduanced.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
