%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 081
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-081


\project douay-rheims


%%% 1289
%%% o-1179
\startChapter[
  title={Psalme 81}
  ]

\PSummary{The
\MNote{Admonition to Magiſtrates.

The 7.~key.}
Prophet declareth, 2.~that God reprehendeth vniuſt Iudges and
Magiſtrates: premonisheth them of his ſeuere & eternal punishment:
8.~wherto the prophet conforming his wil prayeth for the ſame.}

\PTitle{A Pſalme to Aſaph.}

\NV God ſtood in the aſſemblie
\SNote{The wordes of the prophet, admoniſhing al magiſtrates, that when
they ſitte in iudgement, or determine anie cauſe, God who is there, and
euerie where preſent,}
of goddes, and in the middes he
\SNote{attendeth their proceſſe: and therfore it behoueth them to be
aduiſed what they doe: euen as if they heard God ſpeaking as here
foloweth:}
iudgeth goddes.

\V
\SNote{The wordes of God, though not vttered ſenſibly, yet in effect
intimated by his law, according wherto he wil procede in iudgement
againſt vniuſt Iudges.}
How long iudge ye iniquitie: and accept ye the perſons of ſinners.

\V Iudge ye for the needie and the pupil: iuſtifie ye the humble, and
the poore.

\V Take away the poore: and deliuer the needie from the hand of the
ſinner.

%%% o-1180
\V They
\SNote{Such iudges procede in groſſe ignorance, not caring to
vnderſtand, but content to walke in darknes.}
knew not, neither did they vnderſtand, they walke in darkeneſſe:
\SNote{Ye are ſo euil diſpoſed, that you would turne al vpſide downe,
iuſtifying the wicked, and condemning the iuſt.}
al the fundations of the earth shal be moued.

\V I ſaide:
\Fix{Yow}{You}{likely typo, same in both}
are
\SNote{For your office which you participate of me, you are certaine
goddes vpon earth:}
goddes, and the ſonnes of the higheſt al.

\V
\SNote{but when you die you shal find that you are men, ſubiect to Gods
iudgement:}
But you shal die as men:
\CNote{\XRef{Sap.~6.}}
and fal
\SNote{yea your punishment wil be greuous, and importable: for the
mightie shal mightely ſuffer torments.}
as one of the princes.

\V
\SNote{Againe the wordes of the prophet praying God,}
Ariſe ô God, iudge the earth:
\SNote{that
\Fix{eing}{ſeeing}{obvious typo, fixed in other}
he is Lord of al he wil iudge al.}
becauſe thou shalt inherite in al the Gentiles.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
