%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 141
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-141


\project douay-rheims


%%% 1390
%%% o-1280
\startChapter[
  title={Psalme 141}
  ]

\PSummary{Holie
\MNote{Dauids prayer in extreme diſtreſſe.

The 8.~key.}
Dauid being fled into a caue, and beſeeged round about by Sauls armie,
explicating his diſtreſſe, 6.~prayeth to be deliuered.}

\PTitle{Of
\SNote{This Pſalme in forme of a prayer, ſhevveth vvhat cogitations
Dauid had in extreme danger.}
vnderſtanding to Dauid, when he was in the caue, a prayer.
\XRef{1.~Reg.~24.}}

\VV VVith my voice I
\SNote{Not with exterior voice, for ſo he should haue detected himſelfe,
but with feruoure of ſpirite.}
haue cried to our Lord: with my voice I haue prayed to our Lord:

%%% o-1281
\V I powre out my prayer in his ſight, and I pronounce my tribulation
before him.

\V
\SNote{VVhen by reaſon of extremitie, I vvas not able to thinke hovv to
helpe my ſelfe,}
When my ſpirit faileth of myſelf, and
\SNote{thou ô God knovving my actions and
\Fix{demaneur,}{demeanour,}{obvious typo, fixed in other}
didſt deliuer me.}
thou haſt knowne my pathes.

In this way, which I walked, they hid a ſnare for me.

\V I looked toward
\SNote{I looked for helpe,}
the right hand, and
\SNote{and diligently looked about me,}
ſay: and
\SNote{but none vvould ſeme to knovv me, vvhen I required their helpe.}
there was none that would know me.

Flight
%%% !!! SNote should go before 'Flight'
\SNote{Endeuoring to ſaue my ſelf by flight, I found no ſecure place,
for being in a caue or hole of a montaine, the vvhole armie beſeeged me;
al ſeeke to take avvay my life, none to ſaue it.}
hath failed me: and there is none to require my ſoule.

%%% 1391
\V
\SNote{Thus leift deſolate of al mans helpe, and deſtitute of al
vvorldlie ſhift, I cried to thee ô Lord,}
I haue cried to thee ô Lord, I haue ſayd: Thou art
\SNote{my only hopeful refuge.}
my hope,
\SNote{Neither do I deſire to liue for anie vvorldlie reſpect, but
hauing choſen thee ô God for my portion, and inheritance,}
my portion,
\SNote{I deſire to be out of this deſert place, and to be in the land,
vvhere is right vſe of religious diuine ſeruice.}
in the land of the liuing.

\V Attend to my petition: becauſe I am
\SNote{Afflicted.}
humbled excedingly.

Deliuer me from them that perſecute me: becauſe they are made ſtrong
ouer me.

\V Bring forth my ſoule out of priſon,
\SNote{Dauids deſire of libertie, was eſpecially to this end, that he
might haue conuenient place, and other meanes to ſerue and praiſe God:}
to confeſſe vnto thy name:
\SNote{the good, and vvel diſpoſed people of Iſrael, can not now ſerue
thee as they deſire, but expect me,}
the iuſt expect me,
\SNote{whom thou, ô Lord, vvilt aduance to the kingdom, that then vve may
ſerue thee, more freely and more commodiouſly.

Al
\MNote{The prophetical ſenſe of this Pſalme.}
this happened in figure of Chriſt, of vvhom prophetically
%%% !!! Cites?
S.~Auguſtin, S.~Hilarie, Caſſianus, Caſſiodorus, and others expound the
vvhole Pſalme. Amongſt others
%%% !!! Cite?
S.~Beda briefly in theſe vvordes. VVheras in the title, Vnderſtanding is
premiſed to Prayer, therby is ſignified, that Dauid in his deſtreſſes,
and in the denne whither he fled, vnderſtood vvhat our Lord ſhould
ſuffer of the Ievves, and hovv he vvould pray to his Father. In the
firſt part our Lord crieth to his Father, complaining of the deteſtable
deceiptes of Iudas the perſecutor. In the ſecond he prayeth to be
deliuered from the priſon of hel (Limbus, vvhere he vvas free) becauſe
the faith of al the Sainctes depended on his Reſurrection.}
til thou reward me.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
