%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 028
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-028


\project douay-rheims


%%% 1193
%%% o-1086
\startChapter[
  title={Psalme 28}
  ]

\PSummary{The
\MNote{The Church of Chriſt endowed with excellent myſteries.

The 6.~key.}
royal prophet ſeing in ſpirite the moſt ſacred Myſteries,
  brought by Chriſt into this world, inuiteth al to offer their beſt
  thinges, euen themſelues wholly, as ſacrifice of thankes, for ſo
  excellent benefites, preached with magnificence. 5.~VVherby innumerable
  are gethered into his Church here replenished with grace, and in
  heauen with glorie.}

\PTitle{The Pſalme of Dauid,
\LNote{In conſummation of the tabernacle.}{The
ſeuentie Interpreters teſtifie by adding this title, that king Dauid
made this Pſalme (as he did alſo
\CNote{\XRef{104.}
\XRef{105.}
&c.}
ſome others) when the Arke of God was brought into the tabernacle, which
he had pitched for it in Sion.
\XRef{2.~Reg.~6.}
\XRef{1.~Paral.~16.}
\MNote{Myſteries of the Catholique Church prophecied in this Pſalme.}
VVherin he ſaw by prophetical ſpirite, and here vttereth other farre
greater myſteries, & more excellent benefites, brought into the world by
Chriſt, and preached by him, and his Apoſtles, then agreed to the Arke,
or the time of the old Teſtament; but are verified in the admirable
fruite of innumerable people of al nations, and of manie great
Potentates conuerted to Chriſtianitie.}
in the conſummation
\SNote{\XRef{2.~Reg.~6. v.~17.}
\XRef{1.~Par.~16. v.~1.}}
of the tabernacle.}
%%% !!! The annotations appear here rather than at the end of the Psalme
%%% as is usual. And the SNotes mostly appeare at the end of the Psalme.
%%% We do not follow this.

%%% 1194
\NV Bring
\SNote{Offer ſacrifice of thankes for the ſingular benefites after
recounted in this Pſalme.}
to our Lord ye children of God:  bring to our Lord
\SNote{Rammes were of the more principal thinges that were offered in
the law of Moyſes. But the ſonnes of rammes importe in myſtical ſenſe
better hoſtes then rammes.}
the ſonnes of rammes.

\V Bring to our Lord
\SNote{The firſt thing in ſacrifice is to glorifie, honour, and adore
God in ſinceritie of ſpirite,}
glorie and honour, bring to our Lord glorie vnto his name: adore ye our
Lord
\SNote{in his holie Catholique Church.}
in his holie court.
\V The voice of our Lord
\SNote{Here is a greater matter intimated then happened in the bringing
of the Arke into a tabernacle prepared in Sion, when Dauid danced, and
offered hoſtes for ſacrifice, others ioyning with him in that
ſolemnitie. But this \Emph{voice of our Lord vpon vvaters} is rather
verified of our Bleſſed Sauiours owne preaching with}
vpon waters, the God of
\SNote{\Emph{maieſtie},}
maieſtie hath
\SNote{\Emph{thundering}, & by his Apoſtles vpon}
thundered: Our Lord vpon
\SNote{\Emph{manie vvaters}, manie nations:}
manie waters.

\V The voice of our Lord
\SNote{\Emph{in povvre} of miracles:}
in powre: the voice of our Lord
\SNote{\Emph{in magnificence} preaching as \Emph{hauing} in
dede \Emph{powre, not as the Scribes and Phariſes}.
\XRef{Mat.~7. v.~29.}}
in magnificence.

\V The voice of our Lord
\SNote{\Emph{breaking cedars} among innumerable others, conuerting
higheſt Potentates,}
breaking ceders: and our Lord shal breake the
\SNote{\Emph{of Libanus}, Emperoures, kinges, and greateſt Princes of
the world:}
ceders of Libanus:

\V And he shal breake them in peeces as a
\SNote{\Emph{as a calfe of Libanus} ſo meekly ſubmitting them ſelues to
Chriſts yoke, and ſpiritual obedience of his Church.}
calfe of Libanus.
\SNote{Al which is done by Chriſt our Lord \Emph{the beloued} of God,}
And the
\Fix{heloued}{beloued}{obvious typo, fixed in other}
as
\SNote{\Emph{as the ſonne of vnicornes}, is moſt tenderly beloued by the
parentes.}
the ſonne of vnicornes.

\V The voice of our Lord
\SNote{This \Emph{voice of our Lord diuiding the flame of fire}, the
Holie Ghoſt proceding from the Father and the Sonne, came vpon the
Apoſtles, as in diuided tongues of fire.}
diuiding the flame of fire:

%%% o-1087
\V The voice of our Lord
\SNote{Wherwith \Emph{the deſert}, the Gentiles of the wide and wild
world, \Emph{vvere shaken} and moued,}
shaking the deſert; and our Lord shal moue
\SNote{\Emph{the deſert of Cades}, ſome of the Iewes alſo compunct in
hart with remorſe of conſcience hearing the voice of S.~Peter, and other
Apoſtles.}
the deſert of Cades.

\V The voice of our Lord
\SNote{The ſame \Emph{voice of our Lord preparing hartes}, inſpiring the
mindes of men with ſpede, like hyndes and does, to aſcend the high
hilles of free and perfect life, in contemplatiue vertues.}
preparing
\TNote{\L{cervos}}
hartes, and he shal
%%% !!! not marked in either
\SNote{So our Sauiour shal \Emph{diſcouer the thicke vvodes}, reueile
the hidden Myſteries of the old Law, by preaching Chriſtian doctrine,
and vſe of Chriſtian Religion,}
diſcouer thicke woodes:
\Fix{and in}{and}{obvious typo, same in both}
\SNote{\Emph{in his holie Temple}, the Catholique Church, wherin al true
Chriſtians shal \Emph{glorifie God},}
in his temple al shal ſay glorie.

\V Our Lord maketh
\SNote{making the great abundance of people, who are like the \Emph{ſea
vvhen it ouerflovveth the land}, to dwel in the ſame Church.}
to inhabite the floud: and our Lord
\SNote{Chriſt \Emph{our Lord ſitting} & ruling \Emph{king} ouer
al \Emph{for euer}:}
shal ſit king for euer.

Our Lord
\SNote{by his grace \Emph{geuing ſtreingth to his people}, to paſſe
through the tentations of this life;}
wil geue ſtrength to his people: our Lord
\SNote{and \Emph{bleſſe} the iuſt \Emph{vvith} eternal \Emph{peace} in
heauen.}
wil bleſſe his people in peace.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
