%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 098
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-098


\project douay-rheims


%%% 1315
%%% o-1206
\startChapter[
  title={Psalme 98}
  ]

\PSummary{Chriſt
\MNote{Chriſt our Meſſias.

The 5.~key.}
reigneth, notwithſtanding his enimies repine, is adored (5.~alſo his
footeſtoole) 6.~whom ancient Prophetes did inuocate.}

\PTitle{A Pſalme to Dauid himſelfe.}

\NV Our Lord hath reigned,
\SNote{Though manie enimies do rage, and impugne Chriſt:}
let peoples be angrie: he that ſitteth vpon the Cherubs, let
\SNote{though the whole earth be trubled thervvith, yet Chriſt vvho
ſitteth ouer the higheſt Angels, Cherubins and Seraphins, obteyneth the
victorie, reigneth, and doth his vvil in al the earth.}
the earth be moued.

\V Our Lord great in Sion: and high aboue al peoples.

%%% 1316
\V Let them confeſſe to thy great name: becauſe it is terrible, and
holie.

\V And the honour of the king
%%% !!! SNote marked, but missing in other
\SNote{Requireth diſcretion.}
loueth iudgement.

Thou haſt prepared directions: thou haſt done iudgement and iuſtice
%%% !!! SNote marked, but missing in other
\SNote{In fauour of thyn elect people.}
in Iacob.

\V Exalt ye the Lord our God, and
\LNote{Adore his footeſtoole.}{For
\MNote{Chriſts humanitie is his foote ſtoole, adored in the Euchariſt.}
ſo much as al Expoſitors, alſo the Hebrevv Rabbins, affirme that the
Pſalmiſt here prophecieth of Chriſt the promiſed Meſſias, that ſhould
redeme mankind; and ſeing the Arke of couenant perteyneth not to the
ſeruice of Chriſt; but vvas only a figure of him, the footeſtoole of
Meſſias here mentioned, muſt nedes be ſomething perteyning to him: and
therfore moſt ancient Fathers expound it of Chriſts humanitie. And
becauſe the Prophet ſpeaketh of perpetual adoration, not only of the
ſhorte time, he conuerſed vvith men in this life, vvhen very fevv adored
him, the ſame fathers vnderſtand here the adoration of Chriſt in the
bleſſed Sacrament of the Euchariſt.
\MNote{S.~Ambroſe.}
Which S.~Ambroſe teacheth
\Cite{(lib.~3. de Spiritu Sancto. c.~12.)}
in theſe plaine vvordes:
%%% !!! A quote, so Emph ?
By the footſtoole muſt be vnderſtood the earth, by the earth the fleſh
of Chriſt. VVhich vve alſo at this day adore in the Myſteries, and
vvhich the Apoſtles adored in our Lord Ieſus.
\MNote{S.~Auguſtin.}
S.~Auguſtin more largely
\Cite{vpon this Pſalme,}
%%% !!! A quote, so Emph ?
I am made doubtful (ſaith he) I feare to adore the earth, leſt he
condeme me that made heauen and earth. Againe I feare not to adore the
footſtoole of my Lord, becauſe the Pſalme ſaith to me: Adore his
footſtoole. I ſeeke vvhat is his footſtoole, and the Scripture
\XRef{(Iſaiæ.~66.)}
telleth me, the earth is his footſtoole. Doubtful I turne myſelf vnto
Chriſt, becauſe I ſeeke him here, and I finde hovv vvithout impietie the
earth may be adored, vvithout impietie his footſtoole may be adored. For
he tooke earth of earth, becauſe fleſh is of earth, and he tooke fleſh
of the fleſh of (the B.~virgin) Marie.
\MNote{The receiuers of the B.~Sacrament do ſinne if they do not adore
it.}
And becauſe he vvalked here in the ſame fleſh, and gaue the very fleſh
to vs to eate, vnto ſaluation, and no man eateth that fleſh, onles he
firſt adore it: it is found hovv ſuch a footſtoole of our Lord may be
adored: and not only vve doe not ſinne in adoring, but vve ſhould ſinne
in not adoring. Thus farre S.~Auguſtin. Further inſtructing, not to
cõceiue of Chriſts fleſh, as the Capharnaites did, that he would cute it
in peeces from his bodie, and geue them portions therof. His very fleſh
is geuen and eaten, not in fleſhlie maner, but in ſacramental. See
\XRef{Annotations. Ioan.~6.}}
adore
\SNote{Hebrevv Doctors expound this of the Arke in the old teſtament,
but the Doctors of the Church vnderſtand Chriſts humanitie, in the holie
Euchariſt.}
his footſtoole: becauſe it is holie.

\V Moyſes, and Aaron in his
\SNote{Here it is euident (and
%%% !!! Does this work? CNote in SNote
\CNote{\Cite{q.~23. in leuit.}}
S.~Auguſtin ſayth this place taketh avvay al doubt) that Moyſes vvas a
Prieſt, againſt thoſe that for maintaining the hereſie of
%%% !!! Laiheadſhippe, in this, but I follow Other here.
Lay-head-ship, denie it.}
prieſtes: and Samuel among them, that inuocate his name:

\V They inuocated our Lord
\SNote{By example of their praying and obtaining, the Pſalmiſt
confirmeth his prophecie, that Prieſtes of the new Teſtament ſhal pray,
and obtaine mercie of Chriſt for the Church.}
and he heard them: \V in a piller of a cloud he ſpake to them.

They kept his teſtimonies, & the precept which he gaue them.

\V O Lord our God thou heardeſt them: God thou waſt propitious to them,
and taking vengeance vpon al
\SNote{God reuenged the machinations made againſt them, puniſhing the
rebellion of
\Fix{Chore,}{Core,}{obvious typo, same in both}
Dathan and Abyron.
\XRef{Num.~16.}}
their inuentions.

Exalt ye the Lord our God, and adore ye in his holie mount: becauſe the
Lord our God is holie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
