%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 132
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-132


\project douay-rheims


%%% 1380
%%% o-1270
\startChapter[
  title={Psalme 132}
  ]

\PSummary{Fraternal
\MNote{Fraternal concord.

The 7.~key.}
concord is commended to al in the Church, as neceſſarie and delectable
in it ſelf, and bleſſed of God.}

\PTitle{A gradual Canticle
\SNote{Dauid being a moſt peaceable man, euen tovvard his enimies, is
added by Eſdras in this title, as an example for others to imitate.}
of Dauid.}

\NV Behold
\SNote{That concord is both good and pleaſant, nedeth no other proofe,
but only to conſider, & to behold the fruict therof in euerie
communitie, and eſpecially in the Church of God.}
how good, and how pleaſant a thing it is, for brethren to dwel in one:

\V
\SNote{The ointment, vvhervvith Aaron, and other High Prieſtes vvere
conſecrated, vvas precious, and moſt odoriferous, ſignifying grace
bought by Chriſts blood, and producing vertues, vvhich make ſvvete
ſauoure:}
As oyntement on
\SNote{vvas povvred vpon the High Prieſts head, vvho being head of the
Church,}
the head, which ranne downe vpon the beard,
\SNote{vnitie and concord deſcended from him to other Prieſtes,}
the beard of Aaron, which ranne downe vnto
\SNote{and ſo to the people, euen to the lovveſt, and meaneſt in Gods
Church.}
the hemme of his garment:

\V
\SNote{An other ſimilitude to declare the excellencie of concord: As the
devv of mount Hermon, vvhich is perpetually couered vvith ſnovv, that
falleth from the heauen, thence deſcending refresheth & fruictifieth the
hil of Sion: ſo mutual concord, and fraternal charitie amongſt the
faithful, nouriſheth each other, making them fructful in al good
vvorkes.}
As the dew of Hermon, which runneth downe vpon mount Sion.

Becauſe
\SNote{VVhere there is vnitie God geueth abundance of grace, as vvhere
the Apoſtles vvith other faithful vvere geathered, and continued in
prayer, the Holie Ghoſt came vpon them al. And the multitude of beleuers
(then increaſing) had one hart and one ſoule.
\XRef{Act.~2.}
&
\XRef{4.~v.~32.}}
there hath our Lord commanded bleſſing, and life euen for euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
