%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 096
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-096


\project douay-rheims


%%% 1313
%%% o-1204
\startChapter[
  title={Psalme 96}
  ]

\PSummary{Al
\MNote{The laſt iudgement.

The 9.~key.}
the earth is inuited to reioyce in Chriſts kingdom, 3.~with deſcription
of the ſignes coming before the day of Iudgement. 7.~Idolaters shal be
confounded. 8.~Holie Angels and iuſt men shal adore Chriſt, and
reioyce.}

\PTitle{To
\SNote{In figure of Chriſt,}
this Dauid,
\SNote{whoſe bodie roſe the third day after his death: to whom manie
returned beleuing in him after his reſurrection, which fel from him in
his paſſion: and to whom al thinges ſhal be ſubdued, as to their true
Lord, in the day of iudgement.}
when his land was reſtored agane to him.}

\NV Our Lord hath reigned, let the earth reioyce: let
\SNote{Holie Dauid, and other Prophetes hauing great ioy to ſee long
before in ſpirite only, Chriſts kingdom  extended in the whole earth,
yea to the Ilandes, we Ilanders haue great cauſe to be gladde, that God
hath not only ſo bleſſed vs long ſince, but as yet conſerueth ſeede,
wherby we truſt the whole Iland ſhal be againe reſtored vnto him.}
manie Ilands be glad.

%%% 1314
\V
\SNote{As in a cloud with terror God gaue his law to the Iewes: ſo in a
cloud with greater terror and maieſtie, he wil iudge the world;}
Cloude, and miſt round about him: iuſtice, and iudgement
\SNote{not as manie corrupted ſeates of iudgement, in this world, but as
a corrected tribunal, where iuſtice and right iudgement ſhal be practiſed.}
the correction of his ſeat.

\V Fire shal goe before him, and shal inflame his enimies round about.

\V His lightninges
\SNote{Theſe thinges are denounced as if they were alredy donne, for
the aſſured certaintie therof.}
shined to the round world, the earth ſawe, and was moued.

\V The mountaines melted as waxe, before the face of our Lord: before
the face of our Lord al the earth.

\V The heauens haue shewed forth his iuſtice: and al peoples haue ſeene
his glorie.

\V Let them al be confounded, that adore
\SNote{As wel the worſhippers of grauen, or painted images of Iupiter,
Mars, Bacchus and the like, as the worſhippers of the ſame imagined
falſe goddes ſhal be confounded.}
ſculptils: and that glorie in their idoles.

%%% o-1205
Adore him al ye his Angels: \V
\SNote{The Catholique Church.}
Sion heard, and was glad.

And
\SNote{And al particular Churches, members of the vniuerſal.}
the daughters of Iuda reioyced, becauſe of thy iudgements ô Lord.

\V Becauſe thou Lord moſt high ouer al the earth: thou art exalted
excedingly aboue al goddes.

\V You that loue our Lord, hate ye euil: our Lord keepeth the ſoules of
his ſaintes, out of the hand of the ſinner he wil deliuer them.

\V Light is riſen to the iuſt, and ioy to the right of hart.

\V Be glad ye iuſt in our Lord: and confeſſe ye to the memorie
\SNote{Praiſe our Lord Chriſt who is ſanctitie it ſelfe, and ſanctifieth
others.}
of his ſanctification.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
