%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 117
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-117


\project douay-rheims


%%% 1347
%%% o-1237
\startChapter[
  title={Psalme 117}
  ]

\PSummary{Faithful
\MNote{Chriſt beneficial Myſteries are celebrated by his Church.

The 6.~key.}
people collected in the Church of Chriſt, exhorte each other to render
thankes to God, for their deliuerie from ſpiritual and temporal
tribulations. 16.~The Laitie demand participation of Chriſts Myſteries,
promiſing to ſerue him duly: 25.~which the Paſtors freely impart, and
together with the people, ſolemnely celebrate  Gods praiſe.}

\PTitle{Alleluia.}

\NV Confeſſe
%%% !!! CNote should go before 'Confeſſe'
\CNote{\XRef{Pſal.~105.}
\XRef{106.}
\XRef{135.}}
ye to our Lord becauſe he is
\SNote{Let vs praiſe God, for his goodnes, in making vs of nothing,
geuing vs manie benefites,}
good: becauſe
\SNote{and remitting our ſinnes.}
his mercie is for euer.

\V Let Iſrael
\SNote{Let the Church of the nevv teſtament eſpecially confeſſe his
goodnes, vvhich hath receiued more mercie and grace.}
now ſay that he is good: that his mercie is for euer.

\V Let the houſe of Aaron
\SNote{Let al the Clergie praiſe God novv in the time of more grace, and
of greater ſpiritual functions.}
now ſay: that his mercie is for euer.

\V Let them now ſay
\SNote{Yea let the whole bodie of the Church, al that feare, and ſerue
God praiſe his mercie.}
which feare our Lord: that his mercie is for euer.

\V From
\SNote{As wel ſpiritual as temporal.}
tribulation I inuocated our Lord: and our Lord heard me in largenes.

\V Our Lord is my helper: I wil not feare what man can doe to me.

%%% o-1238
\V Our Lord is my helper: and I wil looke ouer mine enemies.

\V It is good to hope in our Lord, rather then to hope in man.

\V It is good to hope in our Lord, rather then to hope in princes.

\V
\SNote{Though innumerable oppoſe, and endeuour to hurt me (ſaith the
Church, or anie iuſt perſon)}
Al nations haue compaſſed me: and in the name of our Lord am I
\SNote{yet by Gods povvre, not by myn ovvne, I am defended, and they
puniſhed, and ſo the iuſt hath the victorie, and triumpheth.}
reuenged on them.

\V Compaſſing they haue compaſſed me: and in the name of our Lord I was
reuenged on them.

%%% 1348
\V They compaſſed me as
\SNote{In great troopes, and furie,}
bees, and were inflamed as
\SNote{vvith ſharpe though ſhorte force, and vvith ſpecial noiſe, to
terrifie me, but in God I ouercame al.}
fyre in thornes: and in the name of our Lord I was reuenged on them.

\V Being thruſt
\SNote{I vvas ſometimes by vehemencie of tentation, declining to ſinne:}
I was ouerturned to fal: and our Lord
\SNote{but Gods grace aſſiſted and ſtrengthned me.}
receiued me.

\V Our Lord is my ſtrength, and my prayſe: and he is made my ſaluation.

\V The voice of exultation, and of ſaluation in the tabernacles of the
iuſt.

\V The right hand of our Lord hath wrought ſtrength: the right hand of
our Lord hath exalted me,
\SNote{The ſame vvord (right hand) thriſe mentioned, ſignifieth the
B.~Trinitie. Alſo (Our Lord) ſignifying Chriſt in his humanitie the
chief inſtrument of God, is here often repeted, to ſignifie the ſingular
efficacie therof.}
the right hand of our Lord hath wrought ſtrength.

\V I shal not die, but shal liue: and I wil tel the workes of our Lord.

\V Our Lord
\SNote{God chaſtiſeth his children,}
chaſtiſing hath chaſtiſed me: and
\SNote{becauſe he vvould not that they ſhould dye eternally. So he
puniſheth as a father, not as an enimie.}
to death he hath not deliuered me.

\V
\SNote{The Prophet novv ſpeaketh in the perſon of iuſt ſoules, requiring
ſpiritual doctrin, and foode,}
Open ye the gates of iuſtice to me, being entred into them
%%% !!! SNote not marked in either
\SNote{and promiſing to ſerue God.}
I wil
confeſſe to our Lord: \V this is the gate of our Lord, the iuſt shal
enter into it.

\V I wil confeſſe to thee becauſe thou haſt heard me: and art become my
ſaluation.

\V
\SNote{An euident prophecie of Chriſt vttered by the Royal Pſalmiſt, and
novv confeſſed by euerie Chriſtian, that our Sauiour reiected by the
Ievves, is neuertheles the builder of his Church, by ioyning the tvvo
peoples of Ievves and Gentiles, as tvvo vvalles into one houſe.}
The ſtone, which the builders reiected: the ſame is made into the head of
the corner.

\V This was done by our Lord: and it is meruelous
\Fix{id}{in}{obvious typo, fixed in other}
our eies.

\V
\CNote{\XRef{Mat.~21.}
\XRef{Luc.~20.}
\XRef{Act.~4.}
\XRef{Rom.~9.}
\XRef{1.~Pet.~2.}}
This
\SNote{God ordayned this acceptable time of grace.}
is the day, which our Lord made: let vs reioice, and be glad therein.

%%% o-1239
\V
\SNote{The ſongue of the Hebrevv children, vvhen Chriſt entered
Ieruſalem vvith palmes of triumph, and acclamations.}
O Lord ſaue me, ô Lord geue good ſucceſſe: Bleſſed be he that cometh in
the name of our Lord.

%%% 1349
\V
\SNote{The voice of Chriſt, and his Apoſtles and other Clergie bleſſing
the people as they deſire.}
We haue bleſſed you of the houſe of our Lord: \V our Lord is God, and he
hath geuen light to vs.

Appoint a ſolemne day with
\SNote{This was fulfilled when Chriſt was brought with bowes of palme,
and other ſignes of triumph, from Bethania,}
thicke bowes, euen to
\SNote{through the whole citie, euen into the Temple and vnto the
Altar:
\XRef{Mat.~21.}}
the horne of the altar.

\V Thou art my God, and I wil confeſſe to thee: thou art my God, and I
wil exalt thee.

I wil confeſſe to thee becauſe thou haſt heard me: and art become my
ſaluation.

\V
\SNote{Our firſt, chife, and final dutie is to praiſe God. v.~1. &.~vlt.}
Confeſſe ye to our Lord becauſe his is good: becauſe his mercie is for
euer.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
