%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 131
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-131


\project douay-rheims


%%% 1378
%%% o-1268
\startChapter[
  title={Psalme 131}
  ]

\PSummary{The
\MNote{Chriſts coming to reſtore man.

The 5.~key.}
Pſalmiſt earneſtly prayeth God to geue him leaue, & to shew him where,
to build a Temple. But more eſpecially prayeth for, and prophecieth the
coming of Chriſt, the promiſed Sonne of Dauid, 14.~ſignifying Gods
promiſe therof, and of eſtablishing his Church.}

\PTitle{A gradual Canticle.}

\NV Remember
\SNote{It is an vſual thing that Dauid, Moyſes, & other Prophetes ſpeake
of themſelues in the third perſon.}
Dauid ô Lord, and al his meekenes:

\V As he ſware to our Lord,
\SNote{King Dauid deſiring, and ſo farre as lay in him, promiſing to
build a Temple to God, vvith great inſtance prayed that he might
performe the ſame. But God diſpoſing othervviſe, that not he, but his
ſonne ſhould build it, he neuertheles prepared the matter, vvorkemen,
and money, ſhevved the forme, and diſpoſed the Leuites hovv to ſerue
therin.}
vowed a vowe to the God of Iacob.

\V
\SNote{Moreouer by vovv depriued himſelf of entering into his ovvne houſe,
or taking his ordinarie reſt, til he might (if it ſo pleaſed God) knovv
the place vvhere it ſhould be built.}
If I shal enter into the tabernacle of my houſe, if I shal aſcend into
the bed of my couch.

\V If I shal geue ſleepe to mine eies, and ſlumbering to mine eie
liddes:

\V And reſt to my temples: vntil I finde a place for our Lord, a
tabernacle for the God of Iacob.

\V Behold we haue heard of it
\SNote{It vvas reueled to Dauid, that the Temple ſhould be built, in
that part of Ieruſalem, vvhich looketh tovvards Bethleem (othervviſe
called Ephrata) vvhere our Sauiour vvas borne.}
in Ephrata: we haue found it in
\SNote{VVithin Ieruſalem, vvhich is compaſſed vvith vvoodes. In this
viſion alſo the vvhole forme of the Temple vvas reueled vnto him, as he
teſtifieth.
\XRef{1.~Par.~28. v.~19.}
Al thinges, quoth he, came vvritten vvith the hand of our Lord vnto me:
that I might vnderſtand al the vvorkes of the paterne.}
the fildes of the wood.

\V
\SNote{Holie Dauid moued vvith exceding deuoution, repared to the place,
vvhere Gods Temple ſhould be built;}
We wil enter into his tabernacle: we wil
\SNote{& adored God, vvhere the Propitiatorie, as a footſtoole
repreſenting Gods preſence, ſhould ſtand. VVhat meruel then if deuotion
moue Chriſtians, to viſite the holie places, vvhere our Sauiour, God and
Man vvas Incarnate, vvas borne, ſuffered death, vvas buried, aſcended
into heauen, or anie other place vvhere his feete ſtood.}
adore in the place where his feete ſtood.

\V
\SNote{O God, leauing Silo, Gabaon, and the like places, come into thy
holie Temple.}
Ariſe Lord into thy reſt, thou and
\SNote{VVith the Arke of couenant, vvhere thou ſanctifieſt thy
people. And here againe the Prophet illuminated vvith a higher Myſterie,
and inflamed vvith more deuotion, prayeth for Chriſtes coming into the
vvorld, and that after his Paſſion he vvil riſe, not only in glorie of
ſoule, but alſo of bodie, prefigured by the Arke of teſtimonie, vvhich
vvas in the tabernacle, and after in the Temple.}
the arke of thy ſanctification.

%%% 1379
%%% o-1269
\V Let thy Prieſtes be clothed with
\SNote{Grant therfore that thy Prieſtes, vvhich muſt offer ſacrifice in
this ſacred place, be indued vvith vertues, and good life vvorthie of
their degree:}
iuſtice: & let thy
\SNote{and the Leuites vvho are ordained to ſerue there, be likevviſe
made fitte for their diuers functions, both in the old and nevv
Teſtament.}
ſainctes reioyce.

\V
\SNote{And ſeing thou haſt geuen ſuch meeknes, deuotion, zele,
ſinceritie, and other vertues making him a man according to thyne ovvne
hart, and therupon promiſed to eſtabliſh his ſede,}
For Dauid thy ſeruants ſake,
\SNote{differ not to ſend thy promiſed Meſſias, Chriſt our Redemer.}
turne not away the face of thy Chriſt.

\V Our Lord
\SNote{Henceforth to the end of this Pſalme, the Prophet relateth Gods
reuelation to him: That he hath truly promiſed vvith an oath,}
hath ſworne truth to Dauid, and
\SNote{and vvil performe,}
he wil not diſapoint it:
\SNote{to ſette one of Dauids ſonnes vpon his Throne: vvhich vvas not
only fulfilled in Salomon, vvho reigned in great peace, and builded the
Temple, but eſpecially in Chriſt the Sonne of Dauid, to vvhom our Lord
God gaue the ſeate of Dauid his father, & he ſhal reigne in the houſe of
Iacob for euer, and of his kingdom there ſhal be no end.
\XRef{Luc.~1. v.~32.~33.}
By this promiſe S.~Peter alſo proueth Chriſts Reſurrection.
\XRef{Act.~2. v.~30.}}
Of the fruite of thy wombe I wil ſet vpon thy ſeate.

\V
\SNote{Concerning the children of Chriſt, members of his kingdom the
Church, the promiſe is conditional, if they obſerue Gods commandments,
they ſhal ſit vpon his ſeate: be coheyres of his kingdom.
\XRef{Rom.~8. v.~17.}}
If thy children shal keepe my teſtament, and theſe my teſtimonies which
I wil teach them:

Their children alſo euen for euer, shal ſit vpon thy ſeate.

\V Becauſe our Lord hath choſen Sion: he hath choſen it for an
habitation to himſelf.

\V This is my
\SNote{Sion, the Catholique Church, is a perpetual place where God
dwelleth.}
reſt for euer and euer: here wil I dwel becauſe I haue choſen it.

\V Bleſſing I wil bleſſe her
\SNote{The Church vvanting Chriſts viſible preſence, is repleniſhed vvith
manie bleſſinges,}
widow: her
\SNote{and al her humble children are releeued vvith holie Sacraments.}
poore I wil fil with breades.

\V Her
\SNote{The Church hath alwayes ſome holie Prieſtes,}
Prieſtes I wil clothe with ſaluation: and her
\SNote{and ſome holie people.}
ſainctes shal reioyce with ioyfulnes.

\V
\SNote{Of his Church, elected by God, Chriſt is the Protector, and
ſtrong defence.}
Thither wil I bring forth a horne to Dauid, I haue prepared
\SNote{And the ſame Church ſhal be a perpetual lampe, vvherby others may
come to him.}
a lampe to my Chriſt.

\V His enemies I wil clothe with confuſion: but vpon him shal my
ſanctification florish.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
