%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 055
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-055


\project douay-rheims


%%% 1242
%%% o-1133
\startChapter[
  title={Psalme 55}
  ]

\PSummary{Dauid
\MNote{Dauids prayer in danger.

The 8.~key.}
being in danger before Achis king of Geth, confidently imploreth Gods
helpe, againſt the great malice and powre of his enimies; 8.~foresheweth
their ruine, his owne exaltation, 12.~and offereth praiſes and thankes.}

\PTitle{Vnto
\SNote{This Pſalme perteyneth alſo to future times,}
the end,
\SNote{for the vſe of anie iuſt perſons, or people, that are againſt
their wil ſeparated from the publike diuine Seruice of holie Church:}
for a people, that is made far from the Sainctes, Dauid in
\SNote{moſt worthie to be noted with title, for perpetual memorie,}
the inſcription of the title,
\SNote{made by Dauid when the Philiſtims detected him to their king in
Geth.}
when the foreners held him in Geth.
\XRef{(1.~Reg.~12. v.~12.)}}

\VV Haue mercie on me ô God, becauſe
\SNote{Now one ſort of il diſpoſed men, now an other,}
man hath troden vpon me:
\SNote{neuer ceaſe to ſeeke my deſtruction.}
al the day impugning he hath afflicted me.

\V Myne enimies haue troden vpon me al the day: becauſe they are
\SNote{Saul with his great armie, the Philiſtimes, and other ſtrangers,
ſome in manifeſt hoſtilitie, others detecting and betraing me to myn
aduerſaries. So al that liue godly in Chriſt haue manie enimies viſible
and inuiſible.}
manie that warre againſt me.

\V From
\SNote{Of theſe moſt eminent great dangers I am in deede afeard,}
the height of the day I shal feare:
\SNote{but ſo that my truſt and aſſured confidence is in thee ô God.}
but I wil truſt in thee.

\V In God I wil praiſe
\SNote{Wordes and promiſes made to me; or the good which I ſpeake or do
by Gods grace.}
my wordes, in God haue I hoped: I wil not feare what flesh may do to me.

\V Al the day did they
\SNote{They calumniate whatſoeuer I ſay,}
deteſt my wordes; againſt me,
\SNote{wreſting al my wordes to euil ſenſe.}
al their
%%% o-1134
cogitations are vnto euil.

\V They wil
\SNote{They meete together, and ſecretly conſpire to intrappe me or
catch me tripping.}
inhabite and keepe ſecret: they wil obſerue my heele.

\V As they haue expected
\SNote{To take my life:}
my ſoule, \V
\SNote{for this their vaine purpoſes to deſtroy me, thou wilt ſaue them,
as they deſerue, that is,}
for nothing shalt thou ſaue them:
\SNote{thou wilt breake them in peeces.}
in wrath thou wilt breake peoples.

O God, \V I haue shewed my life to thee: thou haſt ſet my teares in
thy ſight.

As alſo in thy promiſe: \V then shal mine enemies be turned backeward.

%%% 1243
In what day ſoeuer I shal inuocate thee: Loe I haue knowne that thou art
my God.

\V In God I wil praiſe
\SNote{I wil alwayes gratfully acknowlege thy promiſes, and ſayings, for
they are aſſured.}
the word, in our Lord wil I praiſe thee
\Fix{ſayng:}{ſaying:}{obvious typo, fixed in other}
I haue hoped in God, I wil not feare what man can do to me.

\V In me, ô God,
\SNote{I haue purpoſed and vowed to offer ſacrifice of praiſe, and by
thy helpe wil performe it.}
are thy vowes, which I wil render, praiſes to thee.

\V Becauſe thou haſt deliuered my ſoule from death, and my feete from
falling: that I may
\SNote{Do that pleaſeth God,}
pleaſe before God, in
\SNote{in true faith and pious workes.}
the light of the liuing.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
