%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 100
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-100


\project douay-rheims


%%% 1317
%%% o-1208
\startChapter[
  title={Psalme 100}
  ]

\PSummary{King
\MNote{Inſtruction to gouerne.

The 7.~key.}
Dauid gratfully celebrateth the two general diuine vertues, Mercie and
Iuſtice: 2.~by his owne example exhorteth al, eſpecially Superiors, to
direct their wayes in ſinceritie, 4.~and to ſeperate the wicked from
conuerſation of the good.}

\PTitle{A Pſalme to Dauid himſelfe.}

\NV Mercie
\SNote{Theſe tvvo capital diuine vertues are euer ioyned in al Gods vvorkes,
for both vvhich experienced tovvards him ſelfe, the Pſalmiſt rendereth
thankes and praiſes.}
and iudgement I wil ſing to thee ô Lord: I wil ſing, \V and I
\SNote{I wil do myn endeuoure to knovv the immaculate vvay,}
shal vnderſtand in the immaculate way,
\SNote{vvhich I can not do, but by thy grace coming vnto me. For by
helpe therof I did as folovveth.}
when thou shalt come to me.

%%% 1318
I walked through in the inocencie of my hart, in the middes of my houſe.

\V I did not propoſe before mine eies any vniuſt thing: I hated them
that do preuarication.

\V A peruerſe hart hath not cleaued to me: the malignant declining from
me I knew not.

\V
\SNote{That is, al and euerie one thus wickedly diſpoſed I abhorred.}
One ſecretly detracting from his neighbour, him did I perſecute. One of
a proud eye, and vnſatiable hart, with him I did not eate.

\V Mine eies are towards the faithful of the earth that they may ſit
with me. A man that walketh in the immaculate way,
\SNote{I kept ſuch vnder, as a ſeruant or ſlaue.}
he did miniſter to me.

\V He that doth proudly shal not dwel in the middes of my houſe: he that
ſpeaketh vniuſt thinges, hath
\SNote{Proſpered not, gotte no benefite by me.}
not directed in the ſight of mine eies.

\V In
\SNote{Speedely and without delay I cutte of al diſordered people:}
the morning did I kil al the ſinners of the earth: that I might deſtroy
\SNote{that others might not be corrupted by them.}
out of the citie of our Lord, al thoſe that worke iniquitie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
