%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-14


\project douay-rheims


%%% 1169
%%% o-1061
\startChapter[
  title={Psalme 14}
  ]

\PSummary{For
\MNote{Of eternal Beatitude. The 10.~key.}
attayning eternal glorie in heauen, it is neceſſarie to
  flee from ſinnes, and do good workes.}

\PTitle{The
\LNote{The Pſalme of Dauid.}{As
\MNote{VVhy this and certaine others are called the Pſalmes of Dauid.}
the appropriating of the general name of Pſalme vnto ſome, doth not
preiudice, but that the reſt are alſo Pſalmes, though they be called
Prayers, Canticles, Teſtimonies, and the like: ſo the application of
Dauids name to certaine Pſalmes, proueth not other authores of the
reſt. But the name of Pſalme ſheweth a ſpiritual ſongue, apt for muſical
inſtrument; and the name of \Emph{Dauid} by interpretation ſignifieth,
that it particularly perteyneth to the \Emph{beloued}.}
Pſalme of Dauid.}

\VV Lord who shal dwel in
\SNote{In heauẽ, as appeareth by the laſt verſe.}
thy tabernacle? or who shal reſt in thy holie hil?

\V He that walketh
\SNote{One requiſite thing is to be free, or cleanſed from ſinne,}
without ſpot, and
\SNote{the ſecond is to do good.}
worketh iuſtice.

\V He that ſpeaketh truth in his
\SNote{Sincerely in thought,}
hart, that hath not done guile in his
\SNote{word, and}
tongue. Nor hath
\SNote{dede:}
done euil to his neighbour, and hath not taken
\SNote{nor harkened to detraction.}
reproch againſt his neighbour.

The malignant is brought to nothing in his ſight; but them that feare
our Lord, he
\SNote{Glorie is the reward of good workes.}
glorifieth: he that ſweareth to his neighbour, and deceiueth not,

\V that hath not geuen his money to
\SNote{Vſurie excludeth from heauen.}
vſurie, and hath not taken
\SNote{Likewiſe doing wrong for bribes.}
giftes vpon the innocent:

He that doeth
\LNote{He that doth theſe thinges.}{Wheras
\MNote{Both faith and good workes neceſſarie to ſaluation.}
this, or anie other place of holie Scripture, attributeth ſaluation to
certaine good workes, neither faith, nor other workes are therby
excluded, but preſuppoſed as no leſſe neceſſarie, then thoſe which are
mentioned. 
\CNote{\XRef{Heb.~11.}}
Eſpecially faith is alwayes requiſite, without which it is
impoſſible to pleaſe God, and other vertues either in practice, or in
purpoſe, and preparation of mind, when and where occaſion requireth.}
theſe thinges, shal
\LNote{Shal not be moued for euer.}{Al
\MNote{Only the ſtate of glorie is immutable.}
ſtates of this world are mutable, and only eternal felicitie in heauen
ſhal continew for euer. Therfore this Pſalme can not be vnderſtood of
the Tabernacle, nor Temple of the old Teſtament, which were but figures
of eternal glorie. But if ſo much puritie was then requiſite, much more
al ſinceritie, and great ſanctitie are neceſſarie for entrance into
heauen.}
not be moued for euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
