%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-12


\project douay-rheims


%%% 1167
%%% o-1059
\startChapter[
  title={Psalme 12}
  ]

\PSummary{A
\MNote{A prayer in tribulation. The 7.~key.}
general prayer of the Church, in tribulation, either
  temporal or ſpiritual.}

\PTitle{Vnto
\SNote{It more perteyneth to the new Teſtament then to the old.}
the end, the Pſalme of Dauid.}

\VV How long Ô Lord wilt thou forget me vnto the end? How long doeſt
thou turne away thy face from me?

\V How long shal I put
\SNote{
\Fix{Carful}{Careful}{likely typo, fixed in other}
and almoſt perplex cogitatiõs by reaſon of long
perſecution and mans weaknes.}
counſels in my ſoule, ſorrow in my hart
\SNote{Very often euerie day.}
by day?

\V How long shal mine enemies be exalted ouer me?

\V Regard and heare me Ô Lord my God.

Illuminate mine eies that I ſleepe not
\SNote{Fal not into mortal ſinne.}
in death at any time:

%%% !!! Heretikes (which I am following) has verse 5 start here:
\V leſt ſometime mine enemie ſay: I haue preuailed againſt him.

%%% !!! Other has verse 5 start here.
They that truble me, wil reioyce if I be moued: 
%%% !!! Heretikes (which I am following) has verse 6 start here:
\V but I haue hoped in thy mercie. 

%%% !!! Other has verse 6 start here.
My hart shal reioyce in thy ſaluation: I wil ſing to our Lord which
geueth me
\SNote{Patience in tribulation, and reward for victory.}
good thinges: and I wil ſing to the name of our Lord moſt high.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
