%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 121
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-121


\project douay-rheims


%%% 1370
%%% o-1260
\startChapter[
  title={Psalme 121}
  ]

\PSummary{Vnder
\MNote{Ioyes of heauen promiſed to the iuſt.

The 10.~key.}
the figure of returne to Ieruſalem, vehemently deſired by the Iewes in
captiuitie, the Pſalmiſt deſcribeth the deſire, and hope of the iuſt to
poſſeſſe heauen.}

%%% 1371
\PTitle{A gradual Canticle.}

\NV I reioyced in theſe thinges, which
\SNote{Diuers Prophetes told the Iewes in captiuitie, that they ſhould
returne to Ieruſalem. Al Prophetes, Chriſt alſo, and his Apoſtles, and
Prieſtes preach the entrance, and ioyes of heauen to the iuſt. Al vvhich
the Pſalmiſt ſavv in prophetical ſpirite, and reioyced.}
were ſayd to me: We shal goe into the houſe of our Lord.

\V
\SNote{The Ievves cõſider that ſometimes they vvere ioyful in the Temple
of Ieruſalem: Chriſtians reioyce in the comforte they haue in the
militants Church.}
Our feete were ſtanding, in thy courtes ô Ieruſalem.

\V Ieruſalem, which is built as a citie: whoſe
\SNote{Communitie and participation of ſpiritual graces, is a great ioy
to Catholiques, vvherof the ſame Prophet ſpeaketh,
\XRef{Pſal.~118. v.~61.}
and often elſvvhere.}
participation is together in it ſelf.

\V For
\SNote{Al the tvvelue Tribes frequented Ieruſalem,}
thither did the tribes aſcend,
\SNote{and al nations of the vvorld do come to the Catholique Church.}
the tribes of our Lord: the teſtimonie of Iſrael to confeſſe vnto the
name of our Lord.

\V Becauſe ſeates
\SNote{Seates of Iudgement vvere placed in Ieruſalem,}
ſate there in iudgement,
\SNote{and ſeates of Iudgement in the Catholique Church}
ſeates vpon the houſe
\SNote{of Chriſt.}
of Dauid.

\V
\SNote{Chriſt exhorteth to aske,}
Aske ye the thinges that are for the peace of Ieruſalem: and
\CNote{\XRef{Matt.~21. v.~22.}}
\SNote{and promiſeth to geue that is rightly asked; Chriſtians alſo
inuite each other to pray for the Church.}
abundance to them that loue thee.

\V Peace be made in thy ſtrength: and abundance in thy towers.

\V
\SNote{Chriſt prayeth for his Church,}
For my bretheren, and my neighbours ſakes,
\SNote{and gaue his peace to the Apoſtles, and in them to their
ſucceſſors.}
I ſpake peace of thee:

\V For the houſe of our Lord God, I haue ſought
\SNote{For in heauenlie Ieruſalem, al good thinges are prepared, & are
geuen to Sainctes reigning there for euer.}
good thinges to thee.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
