%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 120
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-120


\project douay-rheims


%%% 1370
%%% o-1259
\startChapter[
  title={Psalme 120}
  ]

\PSummary{The
\MNote{Gods protection.

The 3.~key.}
iuſt reioyce that their prayer is heard, and that God continually
protecteth them.}

\PTitle{A gradual Canticle.}

\NV I haue lifted vp mine eies vnto
\SNote{Towardes Ieruſalem, and towards heauen.}
the mountaines,
\SNote{Al helpe cometh from heauen, that is, from God, who of his diuine
ordinance eſpecially heareth prayers made in holie places.
\XRef{3.~Reg.~7.}}
from whence helpe shal come to me.

\V My helpe is from our Lord, who made heauen and earth.

\V Geue he not
\SNote{The iuſt ſpeaketh and wiſheth wel to his owne ſoule. Eſpecially
the vvhole Church reioyceth in Gods aſſured protection.}
thy foote to be moued: neither doe he ſlumber that keepeth thee.

\V Loe he shal not ſlumber nor ſleepe, that kepeth
\SNote{The militant Church.}
Iſrael.

\V Our Lord kepeth thee, our Lord is thy protection, vpon thy right
hand.

\V By day
\SNote{Proſperitie,}
the ſunne shal not burne thee: nor
\SNote{nor aduerſitie can ouerthrow the Church.}
the moone by night.

\V Our Lord doth kepe thee from al euil: our Lord kepe
\SNote{Spiritual life.}
thy ſoule.

\V Our Lord kepe thy coming in, and thy going out: from hence forth now,
and for euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
