%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 145
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-145


\project douay-rheims


%%% 1395
%%% o-1286
\startChapter[
  title={Psalme 145}
  ]

\PSummary{The
\MNote{Al are exhorted to praiſe God and truſt in his aſſured
  prouidence.

The 3.~key.}
Pſalmiſt exciteth himſelf, and al others to praiſe God for his ſingular
regard, and prouidence of al that truſt in him: 3.~shewing that neither
princes, nor other men are able to helpe, 5.~but God can and doth
continually releue al ſortes of neceſſities.}

%%% 1396
\PTitle{Alleluia,
\SNote{The Septuagint added the names of theſe two Prophetes, for the
like reaſon as they added Ieremie,
\XRef{Pſal.~136.}
becauſe Aggæus and Zacharias prophecying in the reduction of the people
from captiuitie, exhorted them, as here the Pſalmiſt doth, to truſt in
Gods prouidence, and preferre his ſeruice before vvorldlie cares.}
of Aggæus, and Zacharie.}

\VV My ſoule prayſe thou our Lord, I wil praiſe our Lord in my life: I
wil ſing to my God as long as I shal be.

Put not confidence in princes: \V in
\SNote{In one Sonne of Man (Chriſt) is ſaluation (ſaith
%%% !!! Cite?
S.~Auguſtin) and in him, not becauſe he is the ſonne of man, but becauſe
he is the Sonne of God.}
the ſonnes of men, in whom there is no ſaluation.

\V His ſpirit shal goe forth, and he shal returne into
\SNote{He ſaith not, that the ſpirit or ſoule ſhal turne into the earth,
but the ſoule ſhal depart from the bodie, and ſo euerie one (in whom
worldlie men put their truſt) ſhal according to the bodie, returne into
his earth.}
his earth: in that day al their cogitations shal perish.

\V Bleſſed is he, whoſe helper is the God of Iacob, his hope in our Lord
his God: \V who made heauen and earth, the ſea, and al thinges, that are
in them.

\V Which keepeth truth for euer, doth iudgement for them that ſuffer
wrong: geueth foode to the hungrie.

Our Lord looſeth the fettered: \V our Lord illuminateth the blind.

Our Lord lifteth vp the bruiſed, our Lord loueth the iuſt.

\V Our Lord keepeth ſtrangers, the pupil, and widow he wil receiue: and
the wayes of ſinners he wil deſtroy.

\V Our Lord wil reigne for euer thy God ô Sion, in generation and
generation.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
