%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 033
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-033


\project douay-rheims


%%% 1202
%%% o-1095
\startChapter[
  title={Psalme 33}
  ]

\PSummary{King
\MNote{Gods prouidence.

The 3.~key.}
Dauid by his owne example being deliuered from danger, exhorteth al men
to render thankes for Gods benefites, 12.~shewing wherin iuſtice
conſiſteth, 16.~and Gods ſpecial prouidence towards the iuſt.}

\PTitle{To
\MNote{This Pſalme is alſo compoſed in order of the Alphabet.}
Dauid, when
\LNote{He changed his countenance.}{S.~Auguſtin
\MNote{A ſecret great miſterie in the title of this Pſalme.}
by holie Dauids changing of his countenance, and by changing the king of
Geth his name, who in the booke of kinges (where the hiſtorie is
recorded) is called Achis, and here Abimelech, gethereth that here is an
hidden, and great Myſterie. VVhich he explicateth partly by
interpretation of the Hebrew names, but more eſpecially by Dauids
changing of his countenance, which prefigured Chriſt, eternal God
becoming alſo man, and ſo making great changes in the world. For as
Dauid killed Goliath, and for his good act gotte enuie, ſo Chriſt killing
the diuel, and humilitie in Chriſts mẽbers killing pride, are perſecuted
by the wicked.
\CNote{\XRef{Luc.~2.}}
\MNote{Some become vvorſe, ſome better by Chriſt.}
For Chriſt was both to the ruine, and Reſurrection of manie.
\MNote{Sacrifice and Prieſthood changed.}
He changed Sacrifice and Prieſthood. The Iewes had ſacrifice according
to the order of Aaron, in victims of cattle, and this was in
myſterie. For there was not then the Sacrifice of the bodie and bloud of
our Lord: which the faithful, and thoſe that haue read the Goſpel do
know, which Sacrifice is now ſpread in al the round earth. A litle after
the Sacrifice of Aaron is taken away, and the Sacrifice according to the
order of Melchiſedech begane to be. He therfore, \Emph{I knovv not
vvho}, changed his countenance, Let it not be, \Emph{I knovv not vvho},
for our Lord \Sc{Iesvs} Chriſt is knowen. He would haue our health to be
in his bodie and bloud. From whence did he commend his bodie and bloud?
from his humilitie.
\MNote{God moſt high is become lovve.}
For vnles he were humble, he would neither be eaten nor druncke. Behold
his highnes: In the beginning was \Emph{the vvord}, and \Emph{the vvord}
was with God, and God \Emph{the vvord}. Loe the euerlaſting meate, and
Angels eate it, ſupernal powres eate it, celeſtial ſpirites eate it, and
they eate, and are fatted, and the thing remaineth whole, which ſatiateth
and reioyceth them.
\MNote{He is our ſacramental meate.}
How then hath the vviſdome of God fedde vs vvith the ſame bread, the word
made fleſh and dwelt in vs: It were too long to recite this great
Doctors vvhole diſcourſe.
\MNote{Chriſt leaft the Ievves, & receiued the Gentiles.}
He further ſheweth that Chriſt diſmiſſed the
Ievves, and vvent from them to the Gentiles. Thou ſeekeſt novv Chriſt
(ſaieth he) among the Ievves, and findeſt him not: becauſe he hath
changed his countenance. For they ſticking to the ſacrifice according to
the order of Aaron, held not the Sacrifice according to the order of
Melchiſedech, and haue loſt Chriſt: and the Gentiles haue begunne to
haue him. Againe this holie father vvilleth vs to remember the Goſpel:
\MNote{The real preſence of Chriſt in the Sacrament.}
VVhen our Lord \Sc{Iesvs} Chriſt ſpake of his bodie, he ſaid: Vnles you
eate the fleſh of the Sonne of man, and drinke his bloud, you ſhal not
haue life in you: becauſe he had changed his countenance, this ſemed as
furie, and madnes vnto them, to geue his fleſh to be eaten of men, & his
bloud to be drunke, therfore Dauid vvas reputed madde before Achis,
vvhen he ſaid: You haue brought this madde man vnto me. Doth it not ſeme
madnes: Eate ye my fleſh, and drinke my bloud? He ſemed to be madde,
thus S.~Auguſtin. Neuer imagining the figuratiue interpretation of our
nevv Sacramentaries: vvho ſay Chriſt gaue no more but a figure of his
bodie & bloud, for then it had bene eaſily vnderſtood by the
Capharnaites, and no ſuch contradiction, nor murmuring had happened. Yet
S.~Auguſtin ſaith more plainly, if more plaine may be:
\CNote{\Cite{Catech. Ro. p.~2. c.~3. q.~24.}}
\Emph{Chriſt caried himſelfe in his ovvne handes.} And hovv this can be
done (bretheren) in man, vvho can vnderſtand? For vvho is caried in his
ovvne handes? A man may be caried in the handes of others, no man is
caried in his ovvne handes. VVe find not hovv it can be vnderſtood in
Dauid, according to the letter: but in Chriſt vve find it. For Chriſt
vvas carried in his ovvne handes, vvhen geuing his verie bodie, he ſaid:
This is my bodie. For he caried his bodie in his ovvne handes.}
he changed his countenance before
\SNote{His proper name was Achis
\XRef{1.~Reg.~21.}
but al kinges of Paleſtina were called Abimelech, as Pharao in Ægypt,
Nabuchodonoſor in Babylon.}
Abimelech, and he diſmiſt him, and he went away.
\XRef{(1.~Reg.~21.)}}

\VV I
\SNote{I prayſe God}
wil bleſſe our Lord at
\SNote{both in proſperitie and aduerſitie.}
al time: his prayſe alwayes in my mouth.

\V In our Lord
\SNote{When I ſerue our Lord, my ſoule shal be praiſed in his ſeruice.}
my ſoule shal be praiſed: let the milde heare, and reioyce.

\V Magnifie ye our Lord with me: and let vs exalt his name for euer.

\V I haue ſought out our Lord, and he hath heard me: and from al my
tribulations he hath deliuered me.

%%% 1203
\V Come ye to him, and be illuminated: and your faces shal not be
confounded.

\V This
\SNote{Euerie man be he how poore ſoeuer when he prayeth shal be heard.}
poore man hath cried, and our Lord hath heard him: and from al his
tribulations he hath ſaued him.

\V The
\SNote{The proper guardian Angel of euerie one.}
Angel of our Lord shal put in him ſelfe about them that feare him: and
shal deliuer them.

\V Taſt ye, and ſee that our Lord is ſweete: bleſſed is the man that
hopeth in him.

\V Feare
\SNote{With filial feare.}
ye our Lord al ye his ſainctes: becauſe there is no lacke to them that
feare him.

\V The
\SNote{The rich of this world ſetting their mind vpon their wealth, are
poore in ſpiritual giftes.}
rich haue wanted, and haue bene hungrie: but they that ſeeke after our
Lord shal not be diminished of any good.

\V Come children, heare me: I wil teach you the feare of our Lord.

\V Who is the man
\SNote{Euerie one deſireth to be happie, but he in dede shal be happie,
that fleeth from euil, and doth good.}
that wil haue life: loueth to ſee good daies.

\V
\LNote{Stay thy tongue, &c.}{Both
\MNote{Not only faith but good vvorkes with faith iuſtifie.}
theſe verſes, and frequent other places in the Pſalmes, ſhevv plainly
that iuſtice conſiſteth not only in faith, but in abſtayning from euil
and doing good: yet requiring and preſuppoſing true faith, vvithout which
no workes are aualable to iuſtice, nor to euerlaſting life.}
Stay thy tongue from euil: and thy lippes that they ſpeake not guile.

\V Turne away from euil, and do good: ſeeke after peace, and purſewe it.

\V The eies of our Lord vpon the iuſt: and his eares vnto their
prayers.

\V But the
\SNote{God ſeing al mens actiõs & intentions, wil render as they
deſerue.}
countenance of our Lord is vpon them that doe euil things:
%%% o-1096
to deſtroy their memorie out of the earth.

\V The  iuſt haue cried, and our Lord hath heard them: and out of al
their tribulations he hath deliuered them.

\V Our Lord is nigh to them, that are of a contrite hart: and the humble
of ſpirit he wil ſaue.

\V Manie are the tribulations of the iuſt: and out of al theſe our Lord
wil deliuer them.

Our Lord keepeth al their
\SNote{Though the iuſt ſeme for a time to be forſaken, yet God that
geueth them internal ſtreingth, wil at laſt reward and crowne in them his
owne giftes.}
bones: there shal not one of them be broken.

The death of ſinners is verie il: and they that hate the iuſt
\SNote{For their ſinne they are ſuffered to fal into more ſinne.}
shal offend.

\V Our Lord wil redeme the ſoules of his ſeruantes: and al that hope in
him
\SNote{Contrariwiſe thoſe that accept of his grace, shal finally not
offend.}
shal not offend.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
