%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 119
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-119


\project douay-rheims


%%% 1369
%%% o-1258
\startChapter[
  title={Psalme 119}
  ]

\PSummary{The
\MNote{Prayer in tribulation.

The 7.~key.}
Iewes in captiuitie of Babylon, Chriſtians in perſecution, or other
great tribulation, pray with confidence to be deliuered from danger, and
ſclander of wicked tongues, 5.~lamenting their long indurance.}

\PTitle{A gradual Canticle.}

\NV VVhen I was in tribulation I cried to our Lord: and
\SNote{Former experience of Gods mercie in hearing the prayers of thoſe
that inuocated him, geueth hope that he wil heare in like caſe.}
he heard me.

\V O Lord deliuer my ſoule from
\SNote{Nothing is more dangerous then vntrue and deceiptful tongues,
nothing more damage then to diminiſh, and detract from the good fame of
the iuſt: and therfore this is a moſt neceſſarie prayer, that God wil
deliuer vs, from the wicked tongues of Turkes, Heretikes, and other
wicked men.}
vniuſt lippes, and from a deceiptful tongue.

\V
\SNote{What punishment is great ynough for wicked tongues?}
What may be geuen thee, or what may be added vnto thee to a deceiptful
tongue?

%%% 1370
\V
\SNote{Surely the malice of wicked tongues deſerueth sharpeſt
punishments to be afflicted by ſtrong handes.}
The sharpe arrowes of the mightie, with coales of deſolation.

\V
\SNote{Iewes lamented their long abſence from the holie land,
eſpecially from Ieruſalem, and the Temple. Chriſtians mourne for their
reſtraint from Gods Seruice, and long perigrination from heauen.}
Woe is to me, that my ſeiourning is prolonged: I haue dwelte with the
\SNote{Of Cedar the ſonne of Iſmael, came Mahomet, the Turkes falſe
prophet, whoſe tyrannie is great and long. The name Cedar is
interpreted, blacknes, and obſcuritie, which ſignifieth darkenes of
error, and ſinne.}
inhabitantes of Cedar: \V My ſoule hath bene long a ſeiourner.

\V With them, that hated peace, I was peaceable: when I ſpake to them,
they impugned me
\SNote{The wicked afflict thoſe that geue no cauſe of offence.}
without cauſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
