%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 78
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-78


\project douay-rheims


%%% 1284
%%% o-1174
\startChapter[
  title={Psalme 78}
  ]

\PSummary{The
\MNote{The Church ſuffereth verie great perſecutions.

The 6.~key.}
Prophet, in perſon of the Church, lamenteth the crueltie of perſecutors
(both in the old and new teſtament) 5.~prayeth for releaſe, with iuſt
reuenge againſt Gods enimies, that blaſpheme his name, 13.~and promiſeth
to be gratful in diuine praiſes.}

\PTitle{A Pſalme
\SNote{To be ſungue by Aſaph and his companie, in the behalfe of people
vnited in faith.}
to Aſaph.}

\NV Ô God
\SNote{Cruel infidels haue inuaded thoſe thinges, which perteyne to thy
Church,}
the Gentiles are come into thine inheritance, they
\SNote{euen poſſeſſed and prophaned the holie temple of the Iewes, and
Churches of Catholique Chriſtians. Fulfilled by Antiochus, in Ieruſalem;
by the Vandals in Afrike; by Proteſtants, and Puritanes, in diuers
partes of Europe, and wil be more vniuerſally by Antichriſt in al
Chriſtendom.}
haue polluted thy holie temple: they haue made Ieruſalem
\SNote{In ſteed of great and fare Churches, Gods ſeruantes are fane to
vſe meane houſes, yea poore cotages.}
as a watch toure of fruits.

\V They haue made
\SNote{Some perſecutors ſuffer not Martyrs bodies to be buried, but hang
them on polles, and pinnacles, where birdes may eate them.}
the carcaſes of thy ſeruants, meats for the foules of the aire: the
flesh of thy ſainctes for the beaſtes of the land.

\V They haue poured out their bloude as water round about Ieruſalem: and
there was none to burie them.

\V We are
\SNote{Chriſtianes Gallileans, Catholiques Homouſians, and now Papiſtes,
in ſpite and reproch.}
become a reproch to our neighboures: a ſcorne and mocke to them that are
round about vs.

\V How long ô Lord wilt thou
\SNote{Sinnes prouoking Gods wrath, are one cauſe why he ſuffereth his
people to be perſecuted.}
be angrie for euer: shal thy zele be kindled as a fire?

\V Poure out thy wrath vpon the Gentiles,
\SNote{Though the faithful committe ſome ſinnes, yet in reſpect that
they beleue rightly, haue zele in Gods cauſe, and denie not God, they
are nerer to grace and fauour.}
that haue not knowne thee: & vpon the kingdomes, that haue not inuocated
thy name.

%%% 1285
\V Becauſe they haue deuoured Iacob: and his place they haue made deſolate.

%%% o-1175
\V
\SNote{Eſpecially when they repent, and confeſſe their ſinnes, then thoſe
that neither know nor wil know God; but obſtinatly impugne and reſiſt
the truth.}
Remember not our old iniquities, let thy mercies quickly preuent vs:
becauſe we are become exceding
\SNote{So much afflicted, and deſpiſed, that none wil regard the truth
which we profeſſe.}
poore.

\V Help vs ô God our ſauiour: and for the glorie of thy name ô Lord
deliuer vs: and be propicious to our ſinnes for thy names ſake.

\V Leſt they ſay perhaps amongſt the Gentiles:
\SNote{Albeit ô Lord, thou ſuffer vs to be puniſhed, yet ſuffer not thyn
enimies to inſult, as though thou were not our God; and conſequently
thou haddeſt no people at al in the world.}
Where is their God? and
\SNote{Therfore we pray, that the reuenge of thy ſeruantes bloud may be
ſo notified, that it can not be denied, nor doubted of. Which is alſo
here prophecied, that it wil ſo come to paſſe in the end.}
be notified in the nations before our eies.

The reuenge of the bloud of thy ſeruants, which hath benne shed: \V let
the groning of the fettered enter in thy ſight.

According to the greatnes of thine arme, poſſeſſe thou
\SNote{Preſerue alſo, ô God, the reliques of thy ſeruantes, the
ſucceſſors of thy Martyrs.}
the children of them that are put to death.

\V And
\SNote{This alſo is a prophecie, that God wil moſt ſeuerely reuenge the
blaſphemie of perſecutors.}
render to our neighboures ſeuenfold in their boſome: their reproch
wherewith they haue reproched thee ô Lord.

\V But
\SNote{Gods people deliuered from perſecution, and perpetually
conſerued, wil alwayes praiſe God for the ſame.}
we thy people, and the sheepe of thy paſture, wil confeſſe to thee for
euer.

Vnto generation and generation we wil shew forth thy praiſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
