%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 043
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-043


\project douay-rheims


%%% 1221
%%% o-1112
\startChapter[
  title={Psalme 43}
  ]

\PSummary{The
\MNote{The ſtate of the Ievves.

The 4.~key.}
prophet deſcribeth the firſt calling, and difficulte ſtate of the Iewish
nation, 6.~their proſperitie at other times. 10.~Againe their
afflictions in captiuitie, and perſecutions.}

\PTitle{Vnto
\SNote{Though this Pſalme doth firſt and literally perteyne to the
people of Iſrael, yet al thinges happening to them, vvere in figure of
the Chriſtian Catholique Church, vvhich begane vvith difficulties,
aftervvards proſpered, and againe ſuffereth much perſecution.}
the end, for the ſonnes of
\SNote{Core ſignifieth \L{caluus}, bauld; alſo \L{caluaria} a ſcul, or
place of ſculles: the name of the place vvhere our Sauiour vvas
crucified, ſo the children of Core ſignifie the children of Chriſt.
\Cite{S.~Aug.}}
Core to vnderſtanding.}

\VV Ô God we haue heard with our eares: our fathers haue declared to
vs.

The
\SNote{The particular calling of Abraham out of Chaldea, protection of
him, and Iſaac, and Iacob, the deliuerie of al Iſrael out of Ægypt, and
eſtablishing them in the promiſed land of Chanaan, with innumerable,
great, and ſtrange thinges donne for them.}
worke, that thou haſt wrought in their dayes: and in the dayes of old.

\V Thy hand deſtroyed the nations, and thou didſt plant them: thou didſt
%%% o-1113
afflict the peoples, and expel them:

\V For
\SNote{The Iſraelites conquered not by ordinarie povvre,  but by the
miraculous hand of God. See
\XRef{Ioſue.~23.}}
not by their owne ſword did they poſſeſſe the land, and their owne arme
did not ſaue them:

But thy right hand, and thyne arme, and the illumination of thy
countenance: becauſe
\SNote{Not that this people deſerued of themſelues, but of Gods free
election, al the vvorld being vvicked, he gaue peculiar grace to
Abraham, Iſaac, Iacob, and ſome others, and then for their ſakes
protected the vvhole people, in them conſeruing a viſible Church.}
thou waſt pleaſed in them.

\V Thou art the ſame my king and my God: which commandeſt the ſaluations
of Iacob.

\V In
\SNote{As in former examples, ſo in Dauids time, not mans ſtreingth, but
Gods hand gaue them great victories.}
thee we shal turne out our enemies with
\SNote{As an oxe vvith his horne caſteth a ſmal thing into the vvind.}
the horne, & in thy name we shal contemne them that ryſe vp againſt vs.

%%% 1222
\V 
\CNote{\XRef{Ios.~24.}
\XRef{1.~Re.~17.}
\XRef{2.~Reg.~8.}}
For I wil not hope in my bowe: and my ſword wil not ſaue me.

\V For thou haſt ſaued vs from them that afflict vs: and them that hate
vs thou haſt confounded.

\V In God we shal be prayſed al the day: and in thy name we wil confeſſe
for euer.

\V But
\SNote{The prophet fortelleth that after proſperitie God vvould ſuffer
the Ievves to falle into captiuitie, & manie afflictions, which alſo
ſignified allegorically diuers ſtates of Chriſts Church.}
now thou haſt repelled and confounded vs: and thou wilt not goe forth ô
God in our hoſtes.

\V Thou haſt turned vs backe behind our enemies: and they that hated vs,
ſpoyled for themſelues.

\V Thou haſt geuen vs as sheepe that are to be eaten: and thou
\SNote{The Iewes are now meruelouſly diſperſed, and depreſſed.}
haſt diſperſed vs among the nations.

\V Thou haſt ſold thy people
\SNote{In the deſtruction of Ieruſalem the remnant of the people were
ſold for ſmal, as it
\Fix{wree}{were}{likely typo, fixed in other}
for no price. They had ſold Chriſt for thirtie pence:}
without price: and there was
\SNote{and now no multitude, nor number of money at al was geuen for
them, but thirtie of them were ſold for one pennie,
\Cite{Ioſephus de bello Iudaico.}}
no multitude in the exchanges of them.

\V Thou haſt made vs a reproche to our neighbours, a ſcorne and mocking
ſtocke to them, that are round about vs.

\V Thou haſt made vs for a parable to the Gentiles: a wagging of the
head among the peoples.

\V Al the day my shame is againſt me, and the confuſion of my face hath
couered me.

\V At the voyce of the vpbrayder, and the reprocher: at the face of the
enemie and perſecutor.

\V Al theſe thinges haue come vpon vs,
\SNote{Vntil Chriſts paſſion the Iewish people did not wholly fal from
God, and true religion. And of them were choſen the Apoſtles, and manie
others, that founded and propagated the Church of Chriſt.}
neither haue we forgotten
%%% o-1114
thee: and we haue not done wickedly in thy teſtament.

\V And our hart hath not reuolted backward: and
\SNote{The negatiue particle is here vnderſtood by zeugma, according
to the hebrew thus: Our hart hath not reuolted backward, neither haſt
thou ſuffered our pathes to decline frõ thy way.}
thou haſt declined our pathes from thy way:

\V Becauſe thou haſt humbled vs in the place of affliction, and the
shadow of death hath couered vs.

\V
\SNote{An other hebrew phraſe, If we haue, for, We haue not.}
If we haue forgotten the name of our God, and if we haue ſpred forth our
handes to a ſtrange God:

%%% 1223
\V Wil not God enquire of theſe thinges? For he knoweth the ſecretes of
the hart.

Becauſe
\SNote{The Prophetes and others perſecuted partly before Chriſt, much
more the Apoſtles, and other Chriſtians in the new Teſtament.}
for thee we are killed al the day:
\CNote{\XRef{Rom.~8.}}
we are eſtemed as sheepe of ſlaughter.

\V
\SNote{A prayer in affliction.}
Ariſe why ſleepeſt thou ô Lord? Ariſe, and expel vs not to the end.

\V Why doſt thou turne away thy face, forgetteſt our pouertie and our
tribulation?

\V Becauſe
\SNote{We are at deaths dore, readie to become duſt.}
our ſoule is humbled in the duſt: our
\SNote{Lying as groueling ſorowing on the earth:}
bellie is glewed in the earth.

\V
\SNote{til thou deliuer vs from theſe tribulations.}
Ariſe Lord, helpe vs: and redeme vs for thy name.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
