%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 044
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-044


\project douay-rheims


%%% 1223
%%% o-1114
\startChapter[
  title={Psalme 44}
  ]

\PSummary{Dauid
\MNote{Chriſt moſt excellently indowing his Church.

The 6.~key.}
ſingularly moued in hart and tongue, 3.~prophecieth Chriſts excellencie,
indowing his Church with moſt worthie dowries. 11.~By way of exhortation
forshewing her internal and external beautie: 17.~with perpetual
ſucceſſion of Paſtors feeding the flock euen to the worlds end.}

\PTitle{Vnto
\SNote{Perteyning to the new Teſtament.}
the end, for them,
\SNote{Gentiles conuerted from paganiſme to Chriſtianitie:}
that shal be changed,
\SNote{and al others returning from ſchiſme, or other ſinnes,}
to the ſonnes of Core,
\SNote{for their inſtruction,}
for vnderſtanding,
\SNote{this Pſalme is a marriage ſongue of the beloued bridgrome and
bride: Chriſt and his Church.}
a Canticle for the beloued.}

\VV My
\SNote{I haue receiued by diuine inſpiration in my hart and cogitation,}
hart hath vttered
\SNote{a moſt high Myſterie.}
a good word: I tel my workes
\SNote{To the honour therfore and glorie of this king (vvhom I ſecretly
ſee in my hart) I vtter and referre al my vvorkes, and this particular
Canticle.}
to the king.

%%% o-1115
My
\SNote{From the abundance of my hart, my tongue alſo ſpeaketh,}
tongue is the penne of a ſcribe, that
\SNote{& that preſently without delay.}
writeth ſwiftly.

\V
\SNote{Deſcription of Chriſt, moſt excellent in al internal and external
giftes.}
Goodly of beautie aboue the ſonnes of men, grace is powred abrode in
thy lippes: therfore hath God bleſſed thee for euer.

\V Be
\SNote{The prophet ſeing in ſpirite the perfections which he wisheth in
Chriſt, in maner of congratulating, deſcribeth his fortitude, fighting
againſt the diuel for the Church.}
girded with thy ſword vpon thy thigh, ô moſt mightie.

%%% 1224
\V With thy beautie and fayrneſſe
\SNote{Purpoſing,}
intend,
\SNote{proſecuting,}
procede proſperouſely, and
\SNote{and perfecting the conqueſt, and ſo eſtablishing thy ſpiritual
kingdome.}
reigne.

Becauſe of
\SNote{Not vvith warlike armour of this world, but by aſſaulting the
aduerſarie with truth:}
truth, and
\SNote{
\Fix{defeading}{defending}{obvious typo, fixed in other}
thyſelfe and thy ſouldiers with the ſhield of mildnes,}
mildeneſſe, and
\SNote{and ſtriking the enemie with the ſword of iuſtice. VVhich right
force of ſpiritual fight hath meruelous good ſucceſſe.}
iuſtice: and thy right hand shal conduct thee meruelouſly.

\V Thy
\SNote{Preaching of Chriſts Goſpel, his grace mouing the hartes of the
hearers, is liuelie and forcible, more pearcing then anie two edged
ſword.}
sharpe arrowes, the
\SNote{The example of people conuerted, shal moue the hartes of the
aduerſaries to come alſo vnto the truth.}
peoples vnderneth thee shal fal into the hartes of the kings enemies.

\V
\LNote{Thy ſeate ô God for euer and euer.}{Seing
\MNote{Caluin expoundeth this Pſalme contrarie to S.~Paul.}
S.~Paul
\XRef{(Heb.~1. v.~8.)}
affirmeth expreſly that theſe wordes are ſpoken of the Sonne of God,
Chriſt our Sauiour, and therby proueth his excellencie aboue Angels:
Iohn Caluin is wonderful bold to auouch that in the ſimple & proper
ſenſe, Dauid ſpake of his ſonne Salomon, and the daughter of Pharao, as
if that were the literal ſenſe, and S.~Paul only expounded it
myſtically. But firſt the ſolemne preface in the
\XRef{two firſt verſes}
importeth farre greater thinges, then agree to anie terreſtrial
king. Secondly, this excellent beautie deſcribed
\XRef{(v.~3.)}
aboue the ſonnes of men, can not be verified of Salomon, for Abſalom
\XRef{(2.~Reg.~14. v.~25.)}
and Adonias were alſo very beautiful.
\XRef{(3.~Reg.~1. v.~6.)}
As for Salomons wiſdom, or other vertues, he perſeuered not therin, and
ſo he was not bleſſed for euer. Thirdly, the prophet here calleth the
perſon of whom, and to whom he ſpeaketh, God.
\XRef{v.~7.}
&
\XRef{12.}
Fourthly, not only the ancient Fathers, and Doctors of the Church, but
alſo the Hebrew Rabbins, and the Chaldee paraphraſis, expound this
Pſalme literally of the promiſed Meſſias, and his kingdom the Church.}
Thy ſeate ô God
\SNote{Chriſts kingdom shal haue no end.
\XRef{Luc.~1. v.~33.}}
for euer and euer: a rod of direction the rod of thy kingdom.

\V Thou
\SNote{Thou defendeſt and rewardeſt the good, finally forſakeſt and
puniſheſt the wicked.}
haſt loued iuſtice, and haſt hated iniquitie: therfore God,
\SNote{More peculiarly the God of Chriſt, by hypoſtatical vnion.}
thy God, hath annoynted thee with the oile of gladnes
\SNote{Diuers kinges (as Dauid him ſelfe, Ioſaphat, Ezechias, and Ioſias)
were as godlie as Salomon, and perſeuered good to the end, which is
doubted Salomon did not: but Chriſt incomparably was annointed, & indued
with al graces aboue al kinges.}
aboue thy felowes.

\V
\SNote{Mortification which conſerueth from putrifying,}
Myrrhe, and
\SNote{humilitie aſwaging pride,}
Aloes, and
\SNote{being ſmal in the firſt ſpring grovveth great,}
Caſſia from thy
\SNote{humanitie aſſumpted; and ſanctified perſons, in vvhom Chriſt
dwelleth as in cleane, ſhining, odoriferous houſes.}
garmentes, from houſes of
\Fix{yuorie;}{iuorie}{not a typo, same in both}
out of the which \V
\SNote{Sincere faithful ſoules more deare to their ſpouſe Chriſt, then
daughters of temporal kinges.}
the daughters of kinges haue delighted thee in thy honour.

\V The
\SNote{The Catholique Church, in faith purified as gold:}
Queene ſtood on thy right hand in golden rayment: compaſſed with
\SNote{vvith varietie of ſtates, as Clergie, Laity, and diuers ſortes of
religious Orders, and other profeſſions, al vnited in the ſame faith,
hope, and charitie.}
varietie.

%%% o-1116
Heare
%%% !!! This SNote should go before the 'Heare'
\SNote{
\Fix{Carifully}{Carefully}{ovious typo, fixed in other}
al that Chriſt thy ſpouſe ſpeaketh to thee by his Spirite.}
daughter, and
\SNote{Diligently put the ſame in practice:}
ſee, and
\SNote{vvith al obedience and readines, and returne not to former
infidelitie, nor to corrupt life.}
incline thyne eare: and forget thy people, and the houſe of thy father.

\V And the king
\SNote{Chriſt loueth the Church adorned with his giftes,}
wil couer thy beautie: becauſe he is the Lord thy God, and
\SNote{and mutually his true children loue and ſerue him.}
they shal adore him.

%%% 1225
\V And the daughters
\SNote{Manie of al nations ſubmitte themſelues, and al that they haue to
Chriſt.}
of Tyre with giftes, al the rich of the people shal beſech thy
countenance.

\V Al the glorie of that daughter of the king is
\SNote{Internal vertues are moſt eſpecial ornaments:}
within, in
\SNote{exterior are required to edifie others in diuers ſortes of
vertues.}
borders of gold \V clothed round about with varieties.

Virgins
%%% !!! This SNote should go before the 'Virgins'
\SNote{By this meanes manie more are conuerted to chriſtianitie:}
shal be brought to the king after her: her
\SNote{and one countrie inuiteth and draweth an other.}
neighbours shal be brought to thee.

\V They shal be brought in ioy and exultation:
\LNote{They shal be brought into the temple of the king.}{The
\MNote{No ſaluation out of the Church.}
temple of the king, ſaith
\Cite{S.~Auguſtin,}
is the Church, the temple of the
king is in vnity, the temple of the king is not ruinous, not cut in
ſunder, not diuided: the ioyning of liuing ſtones is charitie. Nothing
is more euident. Attend now the verie temple of the king, for from
thence he ſpeaketh, becauſe of the vnitie ſpread in the round earth. For
thoſe that vvould be virgins (faithful ſoules) vnles they be brought
in to the temple of the king (the Catholique Church) they can not pleaſe
the bridgrome.}
they shal be brought into the temple of the king.

\V
\LNote{For thy fathers there are borne ſonnes to thee.}{The
\MNote{Perpetual ſucceſſion of Byſhops in place of the Apoſtles.}
Apoſtles begotte thee (ô Chriſt an Church) they were ſent, they
preached, they are the fathers. But could they be alwaies corporally
with vs? Could anie of them tarie here til this time? Could they tarie
to the time yet to come? But vvas therfore the Church leaft deſolate by
their departure? God forbid. For thy fathers, ſonnes are borne to
thee. VVhat is this for thy fathers, ſonnes are borne to thee? The
Apoſtles were ſent fathers, in place of the Apoſtles ſonnes are borne to
thee; Byſhops are appointed. For whence were the Biſhops borne, that are
at this day through the vvorld? The Church herſelfe calleth them
fathers, ſhe begate them, and appointed them in the ſeates of the
fathers. Do not therfore thinke thy ſelf deſolate (ô chriſtian Church)
becauſe 
thou ſeeſt not Peter, ſeeſt not Paul: for thou ſeeſt not them by vvhom
thou waſt borne, but of thyne iſſue fatherhood is ſprong to thee. For
thy fathers, ſonnes are borne to thee, thou ſhalt make them princes ouer
al the earth. This is the Catholique Church. Her children
are made princes ouer al the earth: her ſonnes are conſtituted for
fathers. Let them
acknowlege this that are cut of: let them come to the vnitie, be they
brought into the temple of the king. Thus
\Cite{S.~Auguſtin.}}
For
\SNote{As Apoſtles came in place of Patriarches and Prophetes: ſo ſtil
Biſhops, and Prieſtes ſuccede in the Church, paſtors, and gouernours
therof.}
thy fathers there are borne ſonnes to thee: thou shalt make them princes
ouer al the earth.

\V They shal
\SNote{Theſe paſtores shal ſtil teach the true Chriſtian doctrin.}
be mindeful of thy name in al generation and generation.

Therfore shal
\SNote{And ſtil there shal be Chriſtian people that wil folow and
profeſſe the ſame.}
peoples confeſſe to thee for euer: and for euer and euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
