%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 095
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-095


\project douay-rheims


%%% 1312
%%% o-1203
\startChapter[
  title={Psalme 95}
  ]

\PSummary{Al
\MNote{Chriſts diuine powre.

The 5.~key.}
peoples & nations are inuited to praiſe the bleſſed Trinitie, 3.~for
Chriſts Incarnation, and ſpiritual kingdom in al the world. 10.~Euen
ſenſles creatures acknowledging his maieſtie, 13.~and iudicial powre.}

\PTitle{A Canticle
\SNote{Inſpired to Dauid, and written by him:}
to Dauid
\CNote{\XRef{1.~Par.~16. v.~23.}}
himſelfe,
\SNote{prophecying the reſtauration of the temple, after the future
captiuitie. And that in figure of the vniuerſal redemption of mankind by
Chriſt, from the captiuitie of the diuel.}
when the houſe was built after the captiuitie.}

\NV Sing ye to our Lord
\SNote{For a new benefite, farre greater then the deliuerie of Iſrael
from Ægypt.}
a new ſong: ſing to our Lord al the earth.

\V
\SNote{The ſame wordes (Sing to our Lord) thriſe repeted, ſignifie the
Bleſſed Trinitie, as ſome Fathers note. Likewiſe
\XRef{v.~7. and~8.}
Bring ye to our Lord, &c. in both places concluding in the ſingular
number, bleſſe his name, bring to his name, importing one God.}
Sing ye to our Lord, and bleſſe his name: shew forth his ſaluation from
day to day.

\V Shew forth  his glorie among the Gentiles, his meruelous workes in al
peoples.

\V Becauſe our Lord is great, and exceeding laudable: he is terrible
aboue al goddes.

\V Becauſe al the goddes of the Gentiles are
\SNote{VVhat creatures ſoeuer ſpiritual or corporal, viſible or
inuiſible, the paganes ſerue for goddes, ſtil they be diuels that
deceiue them, and diuers wayes vſurpe diuine honour, making ſuch
idolaters to thinke, that there is diuine powre, where none is.}
diuels, but our Lord
\SNote{He only is true God, who is Creator of heauen, and of al
creatures. For no creature can create anie thing at al, that is, make
anie thing of nothing but only God.}
made the heauens.

\V Confeſſion, and beauty in his ſight: holineſſe, and magnificence in
his ſanctification.

%%% 1313
\V Bring to our Lord ye families of Gentiles, bring ye to our Lord
glorie and honour: \V bring to our Lord glorie vnto his name.

Take vp hoaſtes, and enter into his courtes: \V adore ye our Lord in his
holie court.

Let al the earth be moued before his face: \V ſay ye among the Gentiles
that our Lord
\SNote{Diuers ancient Doctors read more in this place: Our Lord hath
reigned from the wood, to witte, Chriſt by his death on the croſſe
conquered the diuel, ſinne, and death, and thence begane to reigne.
\Cite{S.~Iuſtinus Martyr dialogo aduers. Triphonem.}
\Cite{Tertullian li. aduerſ. Iudæos. c.~9. &.~13.}
&
\Cite{aduerſ. Marcionem. li.~3. c.~19. &.~23.}
\Cite{S.~Auguſtin in this place,}
according to the old Roman Pſalter. Before him Arnobius, and after him
Caſſiadorus and others, wherby it is probable, that it was ſometimes in
the Hebrew text, and blotted out by the Iewes.}
hath reigned.

%%% o-1204
For he hath corrected the round world which shal not be moued: he wil
iudge peoples in equitie.

\V
\SNote{The Pſalmiſt in abundance of ſpirite inuiteth al creatures to
praiſe God, as Daniel in his Canticle.
\XRef{c.~3.}}
Let the heauens be glad, and the earth reioyce, the ſea be moued, and
the fulneſſe therof: \V the fieldes shal be glad, and al things, that
are in them.

The shal the trees of the woodes reioyce \V before the face of our Lord,
becauſe he cometh: becauſe he cometh to iudge the earth.

He
\SNote{Chriſt iudgeth now in the world by his miniſters diſcerning and
deciding cauſes, rewarding and puniſhing, but eſpecially he wil iudge al
in the laſt day.}
wil iudge the round world in equitie, and peoples in his truth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
