%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 091
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-091


\project douay-rheims


%%% 1306
%%% o-1197
\startChapter[
  title={Psalme 91}
  ]

\PSummary{God
\MNote{The vvorkes of God admirable.

The 2.~key.}
is by al maner of voices and inſtruments to be praiſed in his admirable
workes: 7.~which the foolish not doing are punished; 11.~and the wiſe
are rewarded.}

\PTitle{A Pſalme
\SNote{Voices beginning inſtrumentes proſecute this ſong.}
of Canticle,
\SNote{When we reſt from worke then eſpecially vve ought to thincke vpon
Gods vvorkes, praiſe and thanke him for the ſame.}
in the ſabbath day.}

\VV It is good
\SNote{To geue thankes.}
to confeſſe to our Lord; and to ſing to thy name ô Higheſt.

\V To shew forth thy mercie in
\SNote{In proſperitie,}
the morning: and thy truth in
\SNote{in aduerſitie.}
the night.

\V In
\SNote{On euerie inſtrument of tenne ſtringes, ſignifying the
obſeruation of the tenne commandments:}
the inſtrument of tenſtrings, in
\SNote{namely on the Pſalter.}
Pſalter: with Canticle, on
\SNote{Alſo on the harpe, which ſignifieth mortification.}
the Harpe.

\V Becauſe thou haſt delighted me ô Lord in thy workemanship: and in the
workes of thy handes I wil reioyce.

%%% 1307
\V How are thy workes magnified ô Lord! thy cogitations are made very
profounde.

\V The
\SNote{Carnal and ſenſual man,}
vnwiſe man wil not know; and the
\SNote{he that thinketh only of preſent thinges, not of future.}
\Fix{ſoule}{foole}{likely typo, fixed in other}
wil not vnderſtand theſe thinges.

\V When ſinners shal ſpring vp as graſſe: and al that worke iniquitie
shal appeare.

That they may perish for euer: \V but thou the Higheſt foreuer ô Lord.

\V Becauſe loe thine enimies ô Lord, becauſe loe thine enimies shal
perish: and al that worke iniquitie shal be diſperſed.

\V And
\SNote{The iuſt in confidence of a good conſcience, expect exaltation of
their powre:}
my horne shal be exalted, as the vnicorns:
\SNote{and great conſolation in the end of their life.}
and my old age in plentiful mercie.

\V And
\SNote{The ſhal the iuſt ſee their enimies depreſſed: and themſelues
floriſh, like the palme, and ceder trees as folovveth.}
mine eie hath looked vpon mine enimies: and the malignant ryſing vp
againſt me, mine ears shal heare.

\V The iuſt, shal florish as a palme tree: as the ceder of Libanus shal
he be multiplied.

\V They that are planted in
\SNote{Militant Church:}
the houſe of our Lord, shal florish in
\SNote{triumphant.}
the courtes of the houſe of our God.

\V As yet shal they be multiplied in plentiful old age: and they shal be
wel affected, \V that they
\SNote{Publikly profeſſe Gods praiſes, as in the wordes folowing:}
may shew forth:

%%% o-1198
That the Lord our God is righteous, and there is no iniquitie in him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
