%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 74
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-74


\project douay-rheims


%%% 1272
%%% o-1163
\startChapter[
  title={Psalme 74}
  ]

\PSummary{Chriſt
\MNote{General Iudgement.

The 9.~key.}
(with his Aſſeſſors) wil iudge the whole world, at the laſt day: in the
meane time exhorteth ſinners to amend their life; 7.~for none shal
eſcape iuſt iudgement. 11.~The wicked shal be punished, and the good
rewarded.}

\PTitle{Vnto the end,
\SNote{Either this was the beginning of a ſongue, to the tune wherof
this Pſalme was ſong; as ſome Hebrew Rabbins teſtifie; or, as
S.~Auguſtin and other fathers explicate, the Septuagint do admoniſh vs,
by occaſion of this Pſalme, not to faile in our hope, for that God wil
in no caſe violate his promiſe nor purpoſe, but wil render to euerie one
in the day of Iudgement, as they deſerue.}
Corrupt not, a Pſalme of Canticle to Aſaph.}

%%% 1273
\VV We
\SNote{Chriſt with his Apoſtles, and other aſſeſſorie iudges, wil praiſe
and thanke God, for his admirable good prouidence, and gouernement of
this world.}
wil confeſſe to thee ô God:
\SNote{This duplication of the ſame word, confirmeth vs of the aſſured
performance of that, which is here prophecied.}
we wil confeſſe, and wil inuocate thy name.

We wil tel thy meruelouſe workes: \V when
\SNote{Chriſt our ſingular, principal, and proper Iudge only, and no
mere creature, knoweth the determinate time of general iudgement.}
I shal take a time, I wil iudge iuſtices.

\V
\SNote{The earth at that time ſhal be purged with flaming fire:}
The earth is melted, and al that dwel in it:
\SNote{yet ſhal not be deſtroyed, but changed in qualities.}
I haue confirmed the pillers thereof.

\V I ſaid to the wicked:
\SNote{An epitome, or briefe ſumme of Chriſts doctrine, to flee from
ſinne.}
Doe not wickedly: and to them that offend:
\SNote{Be not proud, abuſe not the powre you haue, to do what you
pleaſe.}
Exalt not the horne.

\V Exalt not your horne on high: ſpeake not iniquitie aganſt God.

\V For neither from the Eaſt, nor from the Weſt, nor from the deſert
mountanes: \V
\SNote{VVhen other Iudges ſitte on tribunales, ſome may be abſent, or
eſcape from their ſentences: but al without exception ſhal be brought in
perſons, when Chriſt God and Man ſhal iudge.}
becauſe God is Iudge.

This
%%% !!! Snote goes before 'This'
\SNote{Some ſhal haue ſentence of eternal paine in hel; ſome of eternal
reward in glorie,}
man he humbleth, and him he exalteth, \V
\SNote{though (partly in this world, partly in particular iudgement of
euerie ſoule) God hath
\Fix{alreadry,}{already,}{obvious typo, fixed in other}
and wil before the laſt day puniſh ſinners, yet there reſteth more
puniſhment, and no impenitent ſinners ſhal eſcape, but al ſhal drinke of
the bitter cuppe of eternal damnation.}
becauſe there is a cuppe in the hand of our Lord
\SNote{Strong wine not delayde with water, but mingled with bitter
ſharpe thinges, as gal, vinegre, brimſtone &c. and ſo powred into
feſtered woundes, ſo they ſhal ſuffer vntolerable endles paine, mixed
with the worme of conſcience, loſſe of glorie, rancor of hart, and
eternal deſperation.}
of mere wine ful of mixture.

%%% o-1164
And he hath powred it out
\SNote{Varietie of helliſh torments, from one extreme to an other: as
from ſnowwaters to exceding heate.
\XRef{Iob.~24.}}
of this into that, but yet the dregges therof are not emptied: al the
ſinners of the earth shal drinke.

\V But I wil shew forth for euer: I wil ſing to the God of Iacob.

\V And I wil breake al the hornes of ſinners: and
\SNote{The iuſt for wel vſing their powre of freewil, and of al powre
geuen them in this life, are rewarded in heauen.}
the hornes of the iuſt shal be exalted.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
