%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 080
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-080


\project douay-rheims


%%% 1287
%%% o-1178
\startChapter[
  title={Psalme 80}
  ]

\PSummary{Al
\MNote{Inuitation to celebrate feſtiual dayes deuoutly.

The 7.~key.}
men are inuited to celebrate feſtiual dayes, 6.~which were inſtituted in
memorie of benefites receiued. 9.~The deuout shal be protected, and the
negligent leift in diſtreſſe.}

\PTitle{Vnto
\SNote{This Pſalme perteyneth not only to the old teſtament, but alſo to
the new.}
the end, for
\SNote{Gods ſeruants oppreſſed with tribulation.}
wine preſſes, a Pſalme to
\SNote{To be ſongue by Aſaph a chief maſter of muſike.}
Aſaph him ſelfe.}

\VV Reioice to God our helper: make Iubilation to the God of Iacob.

\V Take ye Pſalme, and
\SNote{Make readie al theſe muſical inſtruments.}
giue timbrel: pleaſant Pſalter with the harpe.

%%% 1288
\V Sound ye with trumpet in
\SNote{In the Calendes, or firſt day of euerie moneth, in remembrance of
Gods prouidence, and perpetual gouernment of al creatures:}
the
\MNote{The feaſt of Neomenia.}
new moone, in
\SNote{and moſt eſpecial ſolemnitie in the firſt day of the ſeuenth
moneth, in memorie of Iſaac conſerued from death, in whom God promiſed
Abraham, to multiplie his ſeede, and to bleſſe al nations.
\XRef{Gen.~7. v.~21.}
\XRef{c.~21. v.~12.}
\XRef{c.~22. v.~18.}}
the
\MNote{Feaſt of trumpetes.}
notable day of your ſolemnitie.

\V Becauſe it is a precept in Iſrael, and iudgement to the God of Iacob.

\V He put it for a teſtimonie in
\SNote{The people of Iſrael ſignified by Ioſeph, as
\XRef{Pſal.~79. v.~2.}}
Ioſeph, when he came out of the Land of Ægypt: he heard a tongue which
he knew not.

\V He
\MNote{In memorie of this benefite Paſch was inſtituted.}
turned away his backe
\SNote{God deliuered the ſame people from their vntolerable bondage of
carying burdens in baskettes, from geathering ſtraw, making brickes, and
other ſeruitude.
\XRef{Exo.~1.~14.}
\XRef{c.~5. v.~7.}}
from burdens: his handes ſerued in baskettes.

\V In
\SNote{The admonition of God to the people:}
tribulation thou dideſt inuocate me, and I heard thee: I heard thee in
the ſecret of the tempeſt: I
\SNote{yet after ſo great benefites thou didſt murmure, and contradict
me.
\XRef{Exo.~17.}
\XRef{Num.~20.}}
proued thee at the water of contradiction.

\V Heare ô my people: and I
\SNote{Seriouſly admonish thee.}
wil conteſt thee: Iſrael
\SNote{Man by free wil may chooſe whether he wil obey or no.}
if thou wilt heare me.

\V
\MNote{The feaſt of Pentecoſt in memorie of the lawe.}
There shal be
\SNote{This was an other great benefite, to geue an expreſſe law, for
their inſtruction.}
no new God in thee, neither shalt thou adore a ſtrange God.

%%% o-1179
\V For I am the Lord thy God, which brought thee out of the land of
Ægypt:
\SNote{Obſeruing my commandments, aske what thou wilt, and I wil geue it
thee.}
dilate thy mouth and I wil fil it.

\V And my people heard not my voice: and Iſrael attended not to me.

\V
\CNote{\XRef{Rom.~1. v.~24.}}
And I let them alone, according to the deſires of their hart, they shal
goe in their owne inuentions.

\V If my people had heard me: if Iſrael had walked in my wayes:

\V I had
\SNote{Very eaſely.}
for nothing,
\SNote{Of my freewil and liberalitie, without neceſſitie or obligation.}
perhaps humbled their enimies: and had laide my hand vpon thoſe that
afflict them.

%%% 1289
\V The enimies of our Lord haue lied to him: and
\SNote{The wicked that promiſe to ſerue God, and do it not shal be in
eternal torments.}
their time shal be for euer.

\V And
\SNote{In the meane time God beſtowed theſe benefites vpon them, for the
iuſts ſake.}
he fedde them of the fatte of corne: and out of the rocke, with honie he
filled them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
