%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 064
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-064


\project douay-rheims


%%% 1252
%%% o-1143
\startChapter[
  title={Psalme 64}
  ]

\PSummary{God
\MNote{Conuerſion of Gentiles.

The 6.~key.}
is rightly praiſed in Sion and Ieruſalem (in his Church only) for his
benefites beſtowed and promiſed. 8.~Vnto which alſo in the time of
grace, al nations shal be called.}

\PTitle{To the end, a Pſalme of Dauid,
\SNote{The ſeuentie Interpreters ſeing Dauid here prophecie of the
peoples returne from Babylon, added the names of Ieremie and Ezechiel,
who being in that captiuitie, prophecied the ſame more largely. As
likewiſe theſe and other Prophetes, foreſavv in ſpirite, and more
eſpecially prophecied the going forth of al nations from Babylon, that
is, forſaking Idolatrie, and embracing true Religion in the Church of
Chriſt, ſo
%%% !!! Cites?
S.~Auguſtin, Eutymius, and others.}
the Canticle of Ieremie, and Ezechiel, to the people of the
tranſmigration, when they begane to goe forth.}

\VV An hymne ô God becometh thee
\SNote{Not in Babylon, nor els vvhere, but only in the Church, praiſes
and vovves are gratful to God.}
in Sion: & a vow shal be rendered to thee in Ieruſalem.

\V Heare my prayer:
\SNote{Al nations shal know thee.}
al flesh shal come to thee.

\V The wordes of the wicked
\SNote{The wicked are inſolent in threatning,}
haue preuailed vpon vs: and thou wilt
%%% o-1144
be
\SNote{but thou mercifully pardoning our ſinnes, they ſhal not hurt vs.}
propitious to our impieties.

\V
\SNote{They are happie to whom thou haſt prepared grace, and glorie.}
Bleſſed is he, whom thou haſt choſen, and taken: he shal dwel in thy
courtes.

\V
\SNote{The voice of the faithful, reioycing in the hope of eternal
glorie.
\XRef{Rom.~5.}}
We shal be replenished in the goods of thy houſe: holie is thy temple,
\SNote{Nothing polluted ſhal enter into heauen.
\XRef{Apo.~21.}}
meruelous in equitie.

%%% 1253
Heare vs ô God our ſauiour, the hope of al the ends of the earth; and in
the ſea farre.

\V Preparing
\SNote{Thou which art al powreful, as appeareth by the huge montaines,}
mountanes in thy ſtrength, girded with might: which trubleſt
\SNote{ſeas, and other thy workes.}
the depths of the ſea, the ſound of the waues therof.

\V The Gentiles shal
\SNote{Thou wilt by thy omnipotent powre moue the hartes of obdurate
men, and ſo conuert innumerable of al nations to thee.}
be trubled, and they that inhabite the borders shal be affraide of thy
ſignes:
\SNote{Thou wilt draw manie to thee, with ioy and gladnes, from the
vttermoſt coaſtes of the eaſt and weaſt.}
the outegoinges of the morning and euening thou shalt delight.

\V Thou haſt viſited the earth, and haſt inebriated it: thou haſt
multiplied to inrich it.

The
\SNote{God wrought diuers miracles in waters
\XRef{(Gen.~7.}
\XRef{Exo.~7.}
\XRef{14.}
\XRef{15.}
\XRef{Ioſue.~3.}
\XRef{4.~Reg.~5.}
\XRef{6.}
&c.)}
riuer of God is replenished with waters, thou haſt prepared their
\SNote{likewiſe in prouiding meate for his people
\XRef{(Exo.~16.}
\XRef{3.~Reg.~17.}
\XRef{4.~Reg.~4.}
\XRef{17.}
&c.) which were figures of Baptiſme, Euchariſt, and other Sacraments of
Chriſt, waſhing from ſinnes, and augmenting grace,}
meat: becauſe
\SNote{ſo repleniſhing the Church with moſt ſacred Myſteries.}
ſo is the preparation therof.

\V Inebriate
\SNote{Endewing the Apoſtles and other preachers with ſpiritual grace
and lerning,}
her riuers,
\SNote{continuing the ſucceſſion of paſtores to watter and feede the
faithful people.}
multiplie her fruites: in her dropps she shal reioyce ſpringing.

\V Thou
\LNote{God vvil bleſſe the crovvne.}{Vnder
\CNote{\Cite{S.~Aug. Eutym.}}
\MNote{Perpetuitie of the Church.}
the Allegorie of the land of Iurie, vvherunto the people of God vvere to
be reſtored, after their captiuitie in Babylon, the Pſalmiſt here
prophecieth greater thinges, then can be verified of the temporal ſtate
of the Iewes, that the militant Church ſhal ſtil be bleſſed from the
begining to the end, yelding expected fruict;  and al the iuſt, that
perſeuere to the end of their liues, ſhal receiue moſt happie and
glorious revvardes of their laboures.
\MNote{Reward of the iuſt.}
As S.~Paul after his meritorious trauels
confidently expected his glorious revvard, vvhen he ſaide
\XRef{(2.~Tim.~4.)}
I haue fought a good fight, I haue conſummate my courſe, I haue kept my
faith. Concerning the reſt, there is layde vp for me, a Crovvne of
iuſtice, vvhich our Lord vvil render to me in that day, a iuſt
Iudge. And not only to me, but to them alſo that loue his coming.
\MNote{Sainctes crownes are of Gods benignitie.}
And this is called the crovvne of the yeare of Gods benignitie, becauſe
God of his ovvne benignitie, vvithout mans former deſert geueth grace,
and in the end for merite folovving, geueth a crovvne of glorie. So our
Bleſſed Sauiour according to his fulnes of grace, vvhich vvas in his
ſoule, and infinite merite, receiued a crovvne of glorie, in the
conſumation of his temporal life, after the ſpace of thirtie three
yeares. And our Bleſſed Ladie the mother of God, receiued an anſvverable
crovvne to her excellent grace and merites, in the conſummation of her
life, at the end of ſixtie three yeares.
\MNote{The Corones of our Lord, and our Ladie.}
In memorie of vvhich numbers of yeares, deuout men haue piouſly
inſtituted certaine formes of prayers, called the Crovvnes, or Corones
of our Sauiour, and of our Ladie.}
wilt
\SNote{God bleſſeth the whole courſe, or circle of time, of the Church
militant in this world,}
bleſſe the crowne of the yeare of thy goodnes:
\SNote{and the crowne, or happie end of euerie iuſt perſons life.}
and thy fildes shal be replenished with plentie.

\V The
\SNote{Euen thoſe which before had only a ſhew of beautie, but in dede
vvere barren, ſhal yelde abundant fruict.}
beautiful places of the deſert shal be fat: and the
\SNote{Thoſe vvhich are more eminent ſhal particularly reioyce in their
ovvne, and others ſpiritual progreſſe in vertue.}
litle hilles shal be girded aboute with exultation.

\V The
\SNote{The principal paſtores ſhal in proportion reioyce aboue the reſt,
for the grace and glorie of al their flocke.}
rammes of the shepe are clothed, and
\SNote{The ſubiectes alſo and inferiour people ſhal be ſatiate vvith
their happie lotte.}
the vallies shal abounde with corne:
\SNote{Al together, prelates and people, higher and lovver ſhal vvith
vniforme voice ſing praiſes to God, and perpetual hymnes.}
they wil crie, yea they wil ſay an hyme.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
