%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 052
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-052


\project douay-rheims


%%% 1238
%%% o-1129
\startChapter[
  title={Psalme 52}
  ]

\PSummary{As
\MNote{The general Iudgement.

The 9.~key.}
in the thirtenth Pſalme, Chriſts Incarnation is prophecied, after that
ſinne abunded in the world: ſo here is foreshewed that after general
wickednes, 5.~Chriſt wil come to iudge the bad, 7.~and deliuer the
good.}

\PTitle{Vnto the end, for
\SNote{Weakenes, or mourning.}
Maeleth;
\SNote{S.~Auguſtin expoundeth this Pſalme as an inſtruction to thoſe
that ſuffer perſecution and iniuries, eſpecially nere the end of the
world.}
vnderſtandings of Dauid.}

\NV The foole hath ſaid in his hart: There is no God.

\V They are corrupte, and become abominable in iniquities: there is not
that doth good.

\V God hath looked forth from heauen, vpon the children of men: to ſee
if there be that vnderſtãdeth, or ſeeketh after God.

\V Al haue declined, they are become vnprofitable together: there is not
that doth good, no there is not one.

\V Shal they not al know that worke iniquitie, that deuoure my people as
food of bread?

\V God they haue not inuocated: there haue they trembled for feare,
where no feare was.

Becauſe God hath
\SNote{God wil ouerthrow al the counſels, and forces}
diſſipated the bones of them
\SNote{of worldlie politiques.}
that pleaſe men: they are confounded, becauſe God hath deſpiſed them.

\V
\SNote{The true Church afflicted deſireth Chriſts coming to deliuer the
oppreſſed.}
Who wil geue out of Sion the ſaluation of Iſrael? when God shal conuert
the captiuitie of his people: Iacob shal reioyce, and Iſrael shal be
glad.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
