%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 53
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-53


\project douay-rheims


%%% 1238
%%% o-1130
\startChapter[
  title={Psalme 53}
  ]

\PSummary{Dauid
\MNote{A praier in diſtreſſe.

The 7.~key.}
in diſtreſſe crieth to God for helpe, 6.~confidently truſting therin,
8.~and promiſing ſacrifice of thankeſgeuing.}

\PTitle{Vnto
\SNote{Though hiſtorically this Pſalme}
the end,
\SNote{was ſong by Dauid the author therof, shewing how he prayd in
danger, and rendered thankes for his deliuerie,}
in ſongs vnderſtanding for Dauid \V when the Zipheites were come, and
ſaid to Saul:
\SNote{when vpon notice geuen that he abode in the mountaines, Saul
ſtraictly beſeged him, with a great armie, but the Philiſtims inuading
the countrie, Saul was forced to leaue Dauid, and to turne his forces
againſt them,
\XRef{1.~Reg.~23.}
yet it perteyneth alſo literally to al iuſt men in diſtreſſe, eſpecially
to the Church of Chriſt, praying in like dangers, and God by his like
prouidence, deliuering his ſeruantes in extremities.}
Is not Dauid hid with vs?
\XRef{(1.~Reg.~1.~23.}
&
\XRef{26.)}}

%%% 1239
\VV O God ſaue me
\SNote{For the glorie of thy name.}
in thy name: and in thy ſtrength
\SNote{For the iuſtnes of my cauſe defend me.}
iudge me.

\V O God heare my prayer with thine eares: receiue the words of my
mouth.

\V Becauſe
\SNote{Barbarous highland men haue betraide the place of myne abode to
the perſecuters.}
ſtrãgers haue riſen vp againſt me, & the ſtrong haue ſought my ſoule:
and they haue not ſet God before their eies.

\V For
\SNote{But I feare them not, becauſe I am in Gods protection.}
behold God helpeth me: and our Lord is the receiuer of my ſoule.

\V
\SNote{A iuſt prayer, that God wil turne intended miſchief, vpon the
deuiſers heades,}
Turne away the euils to mine enimies: and in
\SNote{according to his promiſe, that he wil defend the innocent.}
thy truth deſtroy them.

\V
\SNote{Offering voluntarie ſacrifice, more then is commanded:}
I wil voluntarily ſacrifice to thee, and
\SNote{and praiſe thee ô God,}
wil confeſſe to thy name ô Lord, becauſe it is good:

\V
\SNote{as I am bond.}
Becauſe thou haſt deliuered me out of al tribulation: and
\SNote{I reioyce in thy iuſt iudgements againſt the wicked.}
mine eie hath looked downe vpon mine enimies.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
