%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-22


\project douay-rheims


%%% 1186
%%% o-1078
\startChapter[
  title={Psalme 22}
  ]

\PSummary{A
\MNote{Thãkſgeuing for Gods protection.

The 7.~key.}
forme of thankeſgeuing for al ſpiritual benefites (deſcribed vnder the
metaphor of temporal proſperitie) euen from a ſinners firſt conuerſion,
to final perſeuerance, and eternal beatitude.}

\PTitle{The Pſalme of Dauid.}

\NV
\CNote{\XRef{Iſa.~40.}
\XRef{Iere.~23.}
\XRef{Ezech.~34.}
\XRef{Ioan.~10.}
\XRef{1.~Pet.~2.}
\XRef{&~5.}}
Our
\SNote{Chriſt the good paſtor, gouerneth, protecteth,}
Lord ruleth me, and nothing shal be wanting to me:
\V in place
\SNote{and feedeth his faithful flocke.}
of paſture there he hath placed me.

Vpon
\SNote{Baptiſme of regeneration,}
the water of refection he hath brought me vp: \V He hath
\SNote{which is the firſt iuſtification.}
conuerted my ſoule.

He hath conducted me vpon
\SNote{Gods precepts which the baptized muſt obſerue.
\XRef{Mat.~28. v.~20.}}
the pathes of iuſtice,
\SNote{Saluation is in the name and powre of Chriſt, not in mans owne
merites.}
for his name.

\V For, although I shal walke in
\SNote{In great dangers of tentations to mortal ſinne,}
the middes of the shadow of death, I wil
\SNote{yet by Gods grace we may reſiſt.}
not feare euils: becauſe thou art with me.

Thy
\SNote{Gods direction and law is ſtreight,}
rod and thy
\SNote{and ſtrong.}
ſtaffe: they haue comforted me.

\V Thou haſt prepared in my ſight
\SNote{Chriſt hath prepared for our ſpiritual foode the B.~Sacrament of
the Euchariſt,
\Cite{S.~Cyprian. Epiſt.~63.}
\Cite{Eutim. in hunc Pſal.}}
a table,
\SNote{againſt al ſpiritual enemies, the world, the flesh, and the diuel.}
againſt them, that truble me.

Thou
\SNote{Chriſtian ſoules are alſo ſtreingthned by the Sacraments of
Confirmation, Penance, holie Orders, Matrimonie, and Extreme Vnction.}
haſt fatted my head with oyle: and my
\SNote{The B.~Sacrament and Sacrifice of Chriſts bodie and bloud.}
chalice inebriating how goodlie is it!

\V And thy mercie shal folow me
\SNote{Continual and final perſeuerance is by Gods ſpecial grace.}
al the dayes of my life:

And that I may dwel in the houſe of our Lord,
\SNote{In eternal life.}
in longitude of dayes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
