%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 128
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-128


\project douay-rheims


%%% 1375
%%% o-1265
\startChapter[
  title={Psalme 128}
  ]

\PSummary{The
\MNote{The Church ſtil firme in perſecution.

The 6.~key.}
Church often (3.~and much) impugned, is not ouercome. 4.~Her perſecuters
are deſtroyed, 6.~contemned, 8.~and curſed.}

\PTitle{A gradual Canticle.}

%%% 1376
\NV Often
%%% !!! SNote should go before 'Often'
\SNote{Iſrael vvhich is the Church of God, reioycing ſaith: that enimies
haue often,}
haue they impugned me
\SNote{euen from the beginning of the world, perſecuted me, as when Cain
perſecuted Abel, other wicked perſecuted Seth, Enoch, Noe: the Chalders perſecuted
Abraham: the Ægyptians perſecuted the Iſraelites: and ſo in other generations.}
from my youth, let Iſrael now ſay:

\V Often haue they impugned me from my youth:
\SNote{But they haue neuer ouercome me. So the Pſalmiſt teſtifieth for
al times paſt, & prophecieth the ſame for times to come.}
but they haue not preuailed againſt me.

\V Sinners
\SNote{Perſecuters not being able to ouerthrow, or ſuppreſſe the Church,
haue laide great weightie burdens of tribulations vpon her backe, which
ſhe hath patiently and ſtrongly borne,}
haue builded vpon my backe: they haue
\SNote{they haue ſtil perſiſted, one ſorte after an other, but with
longanimitie the Church hath ſtood faſt, and conſtantly paſſed through
al diſtreſſes. In moral ſenſe ſinners build iniquitie vpon the back of
the Church, yea and vpon Gods back, when they preſume to ſinne, truſting
in the end to be abſolued by vertue of holie Sacraments leift in the
Church. Likewiſe when they excuſe their ſinnes, imputing the cauſe to
other creatures of God, vvherby they are allured. VVhich is in effect
(ſaith S.~Auguſtin) to accuſe God, and to build iniquities on Gods back,
vvho made thoſe creatures.}
prolonged their iniquitie.

%%% o-1266
\V
\SNote{God therfore who is iuſt, vvil at laſt caſt ſuch preſumptuous
ſinners from his back, and breake their ſtiffe neckes:}
Our iuſt Lord wil cut the neckes of ſinners: \V let them al
\SNote{then ſhal they be confounded,}
be confounded and
\SNote{ſeparated eternally from God, become like fruitles, and vvithered
graſſe, caſt avvay, deſpiſed, yea curſed of al, and bleſſed by none, as
the Prophet denounceth in the next verſes.}
turned backward, that hate Sion.

\V Let them be made as graſſe in the toppes of houſes: which is withered
before it be plucked vp.

\V Wherof the reaper hath not filled his hand, and he that gathereth the
sheaues his beſome.

\V And they ſayd not that paſſed by: The bleſſing of our Lord be vpon
you: we haue bleſſed you in the name of our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
