%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 86
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-86


\project douay-rheims


%%% 1295
%%% o-1186
\startChapter[
  title={Psalme 86}
  ]

\PSummary{The
\MNote{The Catholique Church glorious.

The 6.~key.}
Church of Chriſt beginning in Ieruſalem, 3.~is extended to al Nations,
5.~glorious, 6.~and permanent, 7.~in holie ioy.}

%%% 1296
\PTitle{To the children of Core, a Pſalme of Canticle.}

\NV The fundations thereof in the holie
\SNote{Chriſts Church was firſt founded in Ieruſalem on Whitſunday,
\XRef{Act.~2.}
in mount Sion, which hath two toppes, in one of which the Temple ſtood,
in the other Dauids towre or palace.}
mountaynes: \V our Lord loueth the gates of Sion, aboue al the
tabernacles of Iacob.

\V Glorious thinges are ſaide of thee, ô citie of God.

\V
\SNote{The Prophet in the perſon of Chriſt, ſaith he wil commend vnto
his Apoſtles, and other Apoſtolical men, that they conuert al nations
(as our Sauiour gaue expreſſe commiſſion and commandment
\XRef{Math.~28.}
\XRef{Luc.~24. v.~47.}
\XRef{Act.~1. v.~8.)}}
I wil be mindeful of
\SNote{naming here Raab (which is Ægypt) and Babylon:}
Raab, and Babylon knowing me.

Behold
\SNote{The Philiſtims, Tyrians, Æthiopians,}
the foreners, and Tyre, & the people of the Æthiopians, theſe
\SNote{& the reſt, ſhal be regenerate in this Church, which for the
aſſured certaintie therof (after the prophets maner of ſpeaking) is
affirmed in the pretertence, as if it were then donne.}
were there.

\V Shal it not be ſaid of Sion
\SNote{It ſhal be reported, or one ſhal ſay to an other: Loe this and
that man, al theſe and al theſe men are regenerate by Baptiſme in the
Church of Chriſt.}
Man and man, is borne in her; and
\SNote{God himſelfe, Chriſt God and Man, founded this Church.}
the Higheſt himſelfe founded her?

\V Our Lord wil declare
\SNote{The multitude of the elect is ſo great, that only God knoweth the
number,}
in ſcriptures of peoples, and
\SNote{and the qualities of al ſortes, of Princes, Prelates, and Peoples.}
of princes: of thoſe that haue bene in her.

\V The
\SNote{Great ſpiritual ioy with peace of conſcience, is in true
Chriſtian Catholiques in the militant Church; but the bleſſed haue the
moſt abſolute & ſecure ioy of al, in the Church triumphant.}
habitation in thee, is as it were of al reioycing.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
