%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 007
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-007


\project douay-rheims


%%% 1159
%%% o-1052
\startChapter[
  title={Psalme 7}
  ]

\PSummary{Dauid
\MNote{Dauids prayer in tribulation. The 8.~key.}
in confidence of his iuſt cauſe, and vniuſt perſecution,
  prayeth for Gods helpe, 7.~and iuſt reuenge of his enemies,
  15.~deſcribing their malitious intention, and ruine.}

\PTitle{The Pſalme of Dauid which he ſong to our Lord, for the wordes of
wordes of
\SNote{This Chuſi defeated the counſel of Athitophel as S.~Aug. S.~Baſil
& S.~Chriſ. expound it.}
Chuſi the ſonne of Iemini.
\XRef{(2.~Reg.~16.)}}

%%% 1160
\VV O Lord my God I haue hoped in thee: ſaue me from al that perſecute
me, and deliuer me.

\V Leſt ſometime
\SNote{Abſalom, or anie other enemie, if God doe not reſiſt and hinder
him.}
he as a Lyon violently take
\SNote{My life.}
my ſoule, whiles there is none to redeme, nor to ſaue.

\V Ô Lord my God if I haue done
\SNote{Such euil as myne enemies fayne & obiect againſt me.}
this, if there be iniquitie in my handes;

\V If I haue rendred to them that repayd me euils, let me worthely fal
\SNote{Let me not haue the victorie of them.}
emptie from myne enemies.

\V Let the enemie perſecute my ſoule, and take it, and
\SNote{Let me dye with ignominie.}
treade downe my life in the earth, and bring downe my glorie into the
duſt.

\V Ariſe Lord in thy wrath: and
\SNote{Shew thy powre.}
be exalted in the coaſtes of myne enemies.

%%% !!! Other has verse 8 start here:
And ariſe Ô Lord my God in the
\SNote{Seing thou diddeſt command to make me king, it behoueth thee to
protect me.}
precept which thou haſt cõmanded:
%%% !!! Heretikes (which I am following) has verse 8 start here:
\V and a
\SNote{Manie wil ſerue thee the only true God.}
ſinagogue of peoples shal compaſſe thee.

%%% !!! Other has verse 9 start here:
And
\SNote{For this increaſe of faithful people.}
for it returne on high:
%%% !!! Heretikes (which I am following) has verse 9 start here:
\V our Lord iudgeth peoples.

Iudge me Ô Lord according to
\SNote{My iuſt cauſe againſt my particular enemies.}
my iuſtice, and according to my innocencie
\SNote{Let fal vpon me.}
vpon me.

\V
\CNote{\XRef{1.~Par.~28.}}
The wickedneſſe of ſinners shal be conſumed, and thou shalt direct the
iuſt, which ſearcheſt the hart and raynes Ô God.

\V
\CNote{\XRef{Iere.~11,~17.}}
My iuſt helpe is from our Lord, who ſaueth thoſe that be right of hart.

\V God is a iuſt iudge, ſtrong, & patient: is he angrie
\SNote{God doth not punish al ſinners preſently, but often differreth,}
euerie day?

\V
\SNote{expecting if they wil amẽd.}
Vnleſſe you wil be conuerted, he shal shake his ſword, he hath bent his
bow, and prepared it.

%%% o-1053
\V
\CNote{\XRef{Iſa.~59.}
\XRef{Iob.~15.}}
And in it he hath prepared the veſſels of death: he hath made his
arrowes for them that burne.



\V Behold
\SNote{The iniurious perſecutor purpoſing iniuſtice, conceiueth enuie,
and bringeth forth iniquitie.}
he hath bredde with iniuſtice: he hath conceiued ſorow, and brought
forth iniquitie.

\V He hath opened a pit, and digged it vp: and he is fallen into the
diche, which he made.

\V His
\SNote{Enuie, & deſire to hurt others turneth to the ruine of the
enuiers.}
ſorrow shal be turned vpon his head: and his iniquitie shal deſcend vpon
his crowne.

\V I wil confeſſe to our Lord according to his iuſtice: and wil ſing to
the name of our Lord moſt high.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
