%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 017
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-017


\project douay-rheims


%%% 1173
%%% o-1066
\startChapter[
  title={Psalme 17}
  ]

\PSummary{King
\MNote{Dauid ſingularly protected by God.

The 8.~key.}
Dauids thankes to God for his often deliuerie from great dangers, firſt
in general, 9.~then more particularly deſcribeth Gods terrible maner of
fighting for him, 18.~againſt his cruel, and otherwiſe potent enimies:
22.~attributing the ſame to Gods good pleaſure, and iuſtice of his cauſe,
31.~praiſeth God, 33.~his only protector, 41.~and depreſſer of his
enemies.}

\PTitle{Vnto
\SNote{Though literally this Pſalme perteyned to Dauid, yet in figure of
Chriſt: and of the Church, or euerie iuſt ſoule.}
the end,
\SNote{The Holie Ghoſt inſpired Dauid to render theſe thankes for his
often deliuerie from dangers.}
to the ſeruant of our Lord Dauid, who ſpake to our Lord the wordes of
this canticle, in the day, that our Lord deliuered him out of the hand
of al his enemies, and out of the hand of
\SNote{Saul is ſpecially named becauſe he was his moſt potent worldlie
enimie.}
Saul, and he ſaid:
\XRef{(2.~Reg.~22.)}}

%%% 1174
\VV I wil
\SNote{Theſe firſt wordes (as alſo diuers others) are added, and many
changed in this and other Pſalmes by the Septuagint, who often leauing
the Hebrew text render the ſenſe, and ſo this agreeth in ſubſtance with
the ſame Pſalme recorded
\XRef{2.~Reg.~22.}}
loue thee ô Lord
\SNote{By whom I am ſtrong.}
my ſtrength: \V Our Lord is my firmament, and my refuge, and my
deliuerie.

%%% !!! Verse marker missing here. Present in other
\V
\CNote{\XRef{Heb.~2.}}
My God is my helper, and I wil hope in him. My protectour and the
\SNote{High & firme ſaluation.}
horne of my ſaluation, and my receiuer.

\V Prayſing I wil inuocate our Lord: and I shal be ſaued from mine
enemies.

\V The
\SNote{This is aptly applied to al mankind after his fal, declaring our
ſtate in ſinne, and inducing to penance, in the office of Maſſe on
Septuageſima ſunday.}
ſorrowes of
\SNote{Mortal flesh ſubiect to death.}
death haue compaſſed me: and
\SNote{Violent incurſions of
\Fix{tentions}{tentations}{obvious typo, fixed in other}
to ſinne.}
torrentes of iniquitie haue trubled me.

\V The ſorrowes of
\SNote{Exceeding great afflictiõs of mind, like to torments of hel,
which I alſo feare.}
hel haue compaſſed me: the
\SNote{Secrete tentations haue deceiued me.}
ſnares of death haue preuented me.

\V In my  tribulation I haue inuocated our Lord, and haue
\SNote{Earneſt prayer is the beſt remedie in al tribulations.}
cried to my God:

And
\SNote{As it is certaine the God heard Dauids prayers; ſo he aſſuredly
heareth al that ſincerly flee vnto him.}
he hath heard my voice from his holie temple: and my crie in his ſight,
hath entered into his eares.

\V The earth was shaken & trembled: the fundations of mountaines were
trubled, and were moued,
\SNote{Gods anger againſt ſinne maketh high and loftie thinges to shake,
euen the moſt obſtinate preſumptuous ſinners.}
becauſe he was wrath with them.

\V
\SNote{Diuine wrath is like to ſmoke of the noſethrles, or flaming fire,
and burning coles.}
Smoke aroſe in his wrath: and fire flamed vp from his face: coles
were kindled from him.

%%% o-1067
\V He
\SNote{Gods punishment ſometimes cometh ſo ſwiftly, as if the heauens
bowed towards the earth.}
bowed the heauens, and deſcended: and
\SNote{Gods furie is as a darke deſolate night, or horrible miſt.}
darkeneſſe vnder his feete.

%%% 1175
\V And he
\SNote{Yet whẽ ſinners repẽt God moſt ſpeedely, as fleing with winges of
mercie, comforteth & protecteth them.}
aſcended vpon the cherubs, and flew: he flew vpon the wings of windes.

\V And he put darkeneſſe his
\SNote{God being in him ſelfe incõprehenſible, is alſo ſecret in his
determinations, and couert in his procedinges or actes.}
couert, his tabernacle is round about him: darkeſome water in the
cloudes of the aire.

%%% Verse not marked here, preſent in other.
\V Becauſe of the
\SNote{Gods ſplẽdor oppreſſing mãs ſenſe, yet inſtructeth him by his
meruelous vvorkes. VVhich myſtically ſignifieth, that Chriſt
illuminateth the vvorld by his Apoſtles, and other preachers denouncing
his iuſtice, peace, and his vvil in al thinges perteyning to man.}
brightneſſe in his ſight the cloudes paſſed, hayle and coles of fire.

\V And our Lord thundered from heauen, and the Higheſt goue his voice:
haile and coles of fire.

\V And he shot his arrowes, and diſſipated them: he multiplied
lightnings, and trubled them.

\V And the fountaynes of waters appeared, and the fundations of the
world were reueled.

At thy rebuke ô Lord, at the blaſt of the ſpirit of thy wrath.

\V He ſent from on high, and tooke me: and he receiued me out of manie
\SNote{From tribulations.}
waters.

\V He deliuered me
\SNote{From Saul, Abſolõ, Moabites, Ammonites, and al temporal and
ſpiritual enemies.}
from my moſt ſtrong enemies, and from them that hated me: becauſe they
were made ſtrong ouer me.

\V 
\MNote{So in the reſt of this Pſalme the Prophet ſpeaketh for moſt part
in proper termes, vvithout Metaphores or other figures. Yet in the
myſtical ſenſe of Chriſt, and Chriſtians.}
They preuented me in the day of mine affliction: and our Lord was made
my protectour.

\V And he brought me out into largeneſſe: he ſaued me, becauſe he
\SNote{Of his good pleaſure without my deſertes.}
would me.

\V And our Lord wil reward me according to my iuſtice, and according to
the purity of my handes he wil reward me.

\V Becauſe I haue kept the waies of our Lord, neither haue I done
impiouſly from my God.

\V Becauſe al his iudgementes are in my ſight: and his iuſtices I haue
not repelled from me.

\V And I shal be immaculate
\SNote{By his grace.}
with him; and shal keepe me from mine iniquitie.

%%% o-1068
\V
\SNote{He repeteth the 21.~verſe, inculcating that God wil render to
euerie one as they deſerue.}
And our Lord wil reward me according to my iuſtice: and according to the
puritie of my handes in the ſight of his eies.

\V With the holie thou shalt be holie, and with the innocent man thou
shalt be innocent.

%%% 1176
\V And with the elect thou shalt be elect: and with the peruerſe thou
shalt be peruerted.

\V Becauſe thou wilt ſaue the humble people: and the eies of the proud
thou wilt humble.

\V Becauſe thou doſt illuminate my lampe ô Lord: my God illuminate my
darkeneſſe.

\V Becauſe in thee I shal be deliuered from tentation, and in my God I
shal
\SNote{Paſſe ouer al difficulties.}
goe ouer the wal.

\V My God his
\SNote{Gods pecepts.}
way is vnpolluted: the wordes of our Lord are examined by fire: he is
protector of al that hope in him.

\V For
\SNote{One only God, Creator and Sauiour of al.}
who is God but our Lord? or who is God but our God?

\V God that girded me with ſtrength: and made my way immaculate.

\V That perfited my feete as it were of hartes: and ſetting me vpon high
thinges.

\V That teacheth my handes to battel: and haſt put mine armes,
%%% !!! CNote placement? Goes with SNote? Can we do this?
\CNote{\XRef{2.~Re.~22.}}
\SNote{Amongſt other actes, Dauid killed a lion, and a beare, & Goliath.
\XRef{1.~Reg.~17.}}
as a bow of braſſe.

\V And haſt geuen me the protection of thy ſaluation: and thy right hand
hath receiued me:

And thy diſcipline hath corrected me vnto the end: and thy diſcipline
the ſame shal teach me.

\V Thou haſt enlarged my paſes vnder me: and my ſteppes are not
weakened:

\V I wil purſew myne enemies, and ouertake them: and wil not returne til
they faile.

\V I wil breake them, neither shal they be able to ſtand: they shal fal
vnder my feete.

\V And thou haſt girded me with ſtrength to battel: and haſt ſupplanted
them that riſe againſt me vnder me.

\V And
\SNote{As God geueth ſtrength to his ſeruãts, ſo he diminisheth the natural
ſtreingth and corage of his enemies.}
myne enemies thou haſt geuen me their backe, and them that hate me thou
haſt deſtroyed.

\V  They cried, neither was there that would ſaue them, to our Lord;
neither did he heare them.

\V And I wil breake them to powder, as the duſt before the face of
winde: as the duſt of the ſtreates I wil deſtroy them.

\V Thou wilt deliuer me from the contradictions of the people: thou
%%% o-1069
wilt appoynte me to be head of the Gentiles.

\V A people,
\SNote{Conuerſion of Gentiles to Chriſt, as the Moabites, Idumeans, and
others were ſubdued by Dauid.
\XRef{1.~Par.~11.}
\XRef{14.}
\XRef{18.}
\XRef{19.}
\XRef{20.}}
which I knew not, hath ſerued me: in the hearing of the eare it hath
obeyed me.

%%% 1177
\V The
\SNote{The reuolting and reprobation of the Iewes prefigured by Abſoloms
rebellion, and others.
\XRef{2.~Reg.~15.~16.}}
children being alienes haue lyed to me, the children alienes are
inueterated, and haue halted from their pathes.

\V Our Lord liueth, and bleſſed be my God, and the God of my ſaluation
be exalted.

\V Ô God
\SNote{God ſtil protecteth the Church of Chriſt, as he preſerued Dauid.}
which geueſt me reuenges, & ſubdeweſt peoples vnder me, my deliuerer
from mine angrie enemies.

\V And from them that riſe vp againſt me thou wilt exalt me: from the
vniuſt man thou wilt deliuer me.

\V
\CNote{\XRef{2.~Reg.~22.}
\XRef{Rom.~15.}}
Therfore wil I confeſſe to thee among nations ô Lord: and wil ſay
\SNote{Vſe of Pſalmes is moſt frequent in the Church of Gentiles. See
the
\XRef{Proemial Annotations. Page.~???}}
a pſalme to thy name,

\V Magnifying the ſaluations of his king, and doing mercie to his Chriſt
Dauid, and to his ſeede for euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
