%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 58
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-58


\project douay-rheims


%%% 1245
%%% o-1137
\startChapter[
  title={Psalme 58}
  ]

\PSummary{Holie
\MNote{An other prayer of Dauid in danger.

The 8.~key.}
Dauid (being beſeeged in his owne houſe by men ſent to kil him)
confidently prayeth God to deliuer him, 6.~and al faithful nations in
like danger: 7.~and praiſeth God.}

\PTitle{Vnto the end, deſtroy not, to Dauid in the inſcription of the
title,
\SNote{King Saul hauing thriſe attempted in vaine to kil Dauid,
\XRef{(1.~Reg.~18. v.~11.}
&
\XRef{c.~19. v.~9.)}
ſent ſome of his guard to fetch him, from his owne houſe, that he might
be ſlaine: but God moued the mind of Michol, to admonish him of the
danger, and to helpe him away in ſaftie, though Saul thought she would
haue bene a ſcandal vnto him (or cauſe of ruine) by the handes of the
Philiſtians.
\XRef{1.~Reg.~18. v.~21.}
Vpon which occaſion Dauid made this Pſalme. As he alſo made others, for
perpetual memorie of Gods like benefites, in deliuering him in iminent
dangers, vvhen Saul ſent three troupes of ſerieants to kil him, and
folowed them himſelfe;
\XRef{1.~Reg.~19. v.~20.}
likewiſe vvhen he vvas knovven and bevvrayed before Achis king of Geth;
\XRef{1.~Reg.~21.}
alſo in Ceila, in the deſerts of Ziph, and of Maon,
\XRef{c.~23.}
in Engaddi,
\XRef{c.~24.}
in Hachila,
\XRef{c.~26.}
and againe amongſt the Philiſtians.
\XRef{c.~27.}
and
\XRef{30.}}
when Saul ſent, and watched his houſe to kil him.
\XRef{1.~Reg.~19.}}

\VV Deliuer me from mine enimies ô my God: and from them that riſe vp
againſt me defend me.

\V Deliuer me from them that worke iniquitie: and from bloudy men ſaue
me.

\V Becauſe loe they
\SNote{They haue ſo ſtraictly beſeged me, that it is now in their
handes, to take away my life.}
haue taken my ſoule: the ſtrong haue falne violently vpon me.

\V Neyther is it mine iniquitie, nor my ſinne ô Lord:
\SNote{Of my part I haue committed no fault againſt myn enimies, for
which they can haue anie iuſt cauſe to perſecute me.}
without iniquitie haue I runne, and gone directly.

\V Ryſe vp to meete me, and ſee: and thou ô Lord the God of powres, God
of Iſrael attend to viſite
\SNote{The prophet foreſeing in ſpirite, that the Catholique Church shal
be vniuſtly perſecuted, prayeth, and teacheth others to pray, that God
wil mercifully viſite his faithful people of al nations:}
al nations: haue
\SNote{and not ſpare obſtinate perſecuters.}
no mercie on al that worke iniquitie.

\V They wil returne
\SNote{Perſecuters laboring how much, or how long ſoeuer, shal at night,
that is, in the end of al their wicked endeuoures be vnſatisfied in
their deſires,}
at euening: and they shal ſuffer famine
\SNote{as hungrie dogges that runne hunting al the day, & night alſo, ſtil
ſeeking & not finding wherwith to fil their rauenous mouthes and
deuouring bellies.}
as dogges, and shal compaſſe the citie.

\V Behold they
\SNote{They threaten and determine to vſe al crueltie,}
wil ſpeake in their mouth, and a ſworde in their lippes:
\SNote{as if there were no God, that heareth, and wil punish it.}
becauſe who hath heard?

%%% o-1138
\V And thou ô Lord wilt ſcorne them: thou wilt bring to naught al the
nations.

\V I wil keepe my ſtrength to thee:
\SNote{Through Gods grace the Church is ſtil ſtrong and the vertuous do
perſeuere.}
becauſe thou art my receiuer: \V my God, thy mercie shal preuent.

\V God wil shew vnto me concerning mine enimies, kil them not
\SNote{God ſuffereth afflictions to fal vpon his ſeruants to kepe them
exerciſed, leſt in proſperitie they forgete their duties to him.}
leſt ſometime my peoples forgete.

Diſperſe them in thy ſtrength: and
\SNote{Depriue them of powre, that they may not do ſo much euil as they
deſire.}
depoſe them my protector ô Lord.

\V The ſinne of their mouth, the word of their lippes: and let them be
taken in their pride.

%%% 1247
And for
\SNote{After that their iniquitie is complete,}
curſing and lying they shal be
\SNote{they shal be accurſed and punished for their blaſphemies and
lies.}
talked of \V in conſummation: in wrath of
\Fix{comſummation}{conſummation}{likely typo, fixed in other}
and they shal not be.

And they shal know that God wil rule ouer Iacob: and ouer the ends of
the earth.

\V They
\SNote{As
\XRef{v.~7.}}
shal be turned at euening, and shal ſuffer famine as dogges: and shal
compaſſe the citie.

\V They shal be
\SNote{They shal in vaine ſeeke oyle for their lampes with the foolish
virgins, repent with Iudas, and finding no helpe,}
diſperſed to eate: and if they be not filled, they
\SNote{ſhal continually blaſpheme in hel.}
wil murmur alſo.

\V But I wil ſing thy ſtrength: and wil exalt thy mercie in
\SNote{In the reſurrection.}
the morning.

Becauſe thou art become my receiuer, and my refuge, in the day of
my tribulation.

My helper, I wil ſing to thee, becauſe thou art God my receiuer: my God,
my mercie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
