%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 093
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-093


\project douay-rheims


%%% 1308
%%% o-1199
\startChapter[
  title={Psalme 93}
  ]

\PSummary{The
\MNote{Eternal ſaluation, and damnation.

The 10.~key.}
faithful ſeruant of God aſſuredly profeſſeth, that al the pride,
5.~crueltie, 7.~foolish imaginations, and ſecret thoughts of the wicked
are manifeſt to God. 12.~Acknowlegeth himſelf happie, that he is better
inſtructed of God: wheras he had otherwiſe bene damned. 20.~Sharply
reprehendeth thoſe that conſider not of Gods iudgements: concluding that
the iuſt shal be glorified, and the wicked damned.}

\PTitle{To
\SNote{The Hebrevv letter \HH{Lamed}, vvhich ordinarily is prefixed to
the datiue caſe, or ſignifieth to, being ſet before proper names, is a
ſigne of the genetiue caſe. Yet the Septuagint, expreſſe it by the
datiue, and ſo doth the latin, \L{ipſi Dauid}, and conſequently our
Engliſh hath, to Dauid himſelfe, to ſhevv a difference betvven ſacred
and profane vvriters. For in humane bookes the vvriter and auctor is al
one; but in diuine, the Holie Ghoſt is the proper auctor, and a man is
the vvriter. To ſignifie therfore the principal auctor, Dauid is
ſometimes named as the inſtrumental cauſe, to vvhom the Holie Ghoſt
inſpired this, and other Pſalmes, and by vvhom they vvere vvritten. And
vvhen the titles expreſſe othervviſe: A Pſalme of Dauid, yet it is ſo to
be vnderſtood, that the Holie Ghoſt is alvvayes the principal auctor,
and Dauid the inſtrumental, miniſterial, or ſecondarie auctor. But vvhen
other names are expreſſed, either in the genetiue or datiue caſe, or
hovvſoeuer, it proueth not that thoſe men vvere the vvriters of the ſame
Pſalmes, but importeth ſome other thing, as by S.~Auguſtins iudgement,
vve noted in the
%%% !!! fix XRef
\XRef{proemial Annotations page.~3. &.~4.}
vvherby is proued that this Pſalme vvas not written, nor compoſed by
Moyſes, as Hebrevv Rabbins ſuppoſe, but by the Royal Pſalmiſt Dauid.}
Dauid himſelfe, in
\SNote{Made and ordinarily ſongue in the fourth day of the vveke, our
vveneſday, in vvhich day Iudas the traitor ſold our Sauiour Chriſt to
his enimies. The reuenge of vvhich vvickednes, and of al other ſinnes,
is here prophecied.}
the fourth of the ſabbath.}

%%% !!! The first verse?
\NV Our Lord God
\SNote{God more commonly called the God of mercie (vvhich vertue in him
is aboue al his vvorkes,
\XRef{Pſal.~144.)}
is alſo the God of reuenges, according to his iuſtice.}
of reuenges: the God of reuenges hath done
\SNote{He procedeth in iudgement reſolutly, not depending, nor fearing,
nor reſpecting anie perſon, povvre, dignitie, vviſdome, or other like
qualitie, but their iuſt merites.}
freely.

\V Be
\SNote{A prayer of iuſt zele.}
exalted thou that iudgeſt the earth: render retribution to the proude.

\V How long shal ſinners ô Lord: how long shal ſinners glorie?

\V Shal they vtter, and ſpeake iniquitie? shal al they ſpeake,
\SNote{Shal moſt wicked men ſtil be ſuffered to ſpeake ſo inſolently?}
that worke iniuſtice?

\V
\SNote{A deſcription of heathniſh and heretical crueltie.}
Thy people ô Lord they haue humbled: and thine inheritance they haue
vexed.

\V The widow, and the ſtranger they haue ſlaine: and the pupilles they
haue killed.

%%% o-1200
\V And
\SNote{Scarſe anie Atheiſtes are ſo blind, as thus to thinke but manie
ſinners ſo behaue themſelues, as if God ſaw not, knew not, or at leaſt
cared not vvhat they do.}
they haue ſaide: The Lord shal not ſee, neither shal the God of Iacob
vnderſtand.

\V Vnderſtand ye foolish in the people: and ye fooles be wiſe at
ſometime.

\V He that planted the eare, shal he not heare? Or he that made the eie
doth he not conſider?

\V He that chaſtiſeth nations, shal he not rebuke: he that teacheth man
knowledge?

\V Our Lord knoweth
\SNote{So vnpoſſible is it that God ſhould be ignorant, or careles vvhat
men do, that he alſo knovveth and obſerueth moſt ſecret thoughtes.}
the cogitations of men: that they be vaine.

\V Bleſſed is the man, whom thou shalt inſtruct ô Lord: and shalt teach
out of thy lawe.

\V That thou maiſt
\SNote{Mitigate and temper his afflictions, that by patience and
fortitude, the iuſt may perſeuere, and not be ouerwhelmed.}
geue him quietnes from the euil daies: til a pitte be digged for the
ſinner.

%%% 1310
\V Becauſe our Lord
\SNote{The whole Church shal neuer be reiected, nor forſaken.}
wil not reiect his people: and his inheritance he wil not forſake.

\V Vntil iuſtice
\SNote{Iuſtice is conuerted into iudgement, vvhen iuſt meaning is put in
vvorke and practiſe, that it may appeare in iudgement. Alſo God vvho
doth & ſuffereth al iuſtly, vvil conſerue his inheritance the Church,
euen vnto the day of iudgement.}
be turned into iudgement: and
\SNote{The ſenſe is eaſie by tranſpoſing the vvordes: al that are right
of hart, are nere it, that is, ſhal like and approue Gods iuſtice, vvhen
the vvicked ſhal repine, and blaſpheme it.}
they who are neere it, are al that are right of hart.

\V Who shal riſe for me againſt the malignant? or who shal ſtand with me
againſt them that worke iniquitie?

\V But that our Lord hath holpen me: within very litle my ſoule had
dwelt in hel.

\V If I ſaid:
\SNote{When I felt, and complained that I was in danger, thou didſt
aſſiſt me.}
My foote is moued: thy mercie ô Lord did help me.

\V According to the multitude of my ſorrowes in my hart: thy
conſolations haue made my ſoule ioyful.

\V Doth the ſeat of iniquitie cleaue to thee: which makeſt
\SNote{Onlie faith ſufficeth not, but careful laboure, in keping Gods
\Fix{commandmnts}{commandments}{obvious typo, fixed in other}
is required.}
labour in precept?

\V They wil hunt after the ſoule of the iuſt: and wil condemne innocent
bloud.

\V And our Lord became my refuge: and my God the helpe
\SNote{The iuſt do hope for eternal ſaluation, to which God wil bring
them.}
of my hope.

\V And he
\SNote{And God, the reuenger of wronges, wil at laſt caſt the wicked
into eternal torments.}
wil repay them their iniquitie: and in their malice he wil deſtroy
them: the Lord our God wil deſtroy them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
