%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 123
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-123


\project douay-rheims


%%% 1372
%%% o-1261
\startChapter[
  title={Psalme 123}
  ]

\PSummary{The
\MNote{Gods protection.

The 3.~key.}
whole Church, and euerie iuſt perſon, gratfully confeſſeth, that by Gods
protection they are ſecure, without which none could eſcape ruine.}

\PTitle{A gradual Canticle.}

\NV But that our Lord
\SNote{If God had not ſent his helpe, and defence to vs,}
was in vs, let Iſrael now ſay: \V but that our Lord was in vs,

When
%%% !!! SNote should go before 'When'
\SNote{we could not haue eſcaped the force of our enimies.}
men roſe vp againſt vs, \V
\SNote{The vvord (perhaps) doth not here import a doubt, or
\Fix{vncertanitie,}{vncertaintie,}{obvious typo, fixed in other}
but in modeſt maner of aſſeueration, leaueth the iudgement of the euent
that ſhould hapen (if God did not protect his ſeruantes) to their ovvne
conſideration: vvhich is an vſual phraſe in moſt languages.}
perhaps they had ſwalowed vs
\SNote{So ſudainly ſhould the vveake, vvithout Gods protection, be
deſtroyed, as men are deuoured & ſvvalovved vp by rauening vvild beaſts,
euen before they be throughly dead. So vvas Ionas ſvvalovved into the
vvhales bellie.}
aliue:

%%% o-1262
When their furie was angrie againſt vs, \V perhaps
\SNote{Furious perſecution.}
water had ſwalowed vs.

\V Our ſoule hath paſſed through
\SNote{Suddaine great trubles.}
a torrent: perhaps our ſoule had paſſed through an intolerable water.

\V Bleſſed be our Lord which hath not geuen vs, for a pray to their
teeth.

\V Our ſoule as a ſparow is deliuered from the ſnare of the fowlers:

The ſnare is
\SNote{Mans ſubteltie often deceiueth an other man, but there is no
counſel able to circumuent God.}
broken, and we are deliuered.

\V Our helpe is in the name of our Lord, who made heauen and earth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
