%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 114
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-114


\project douay-rheims


%%% 1345
%%% o-1235
\startChapter[
  title={Psalme 114}
  ]

\PSummary{The
\MNote{A prayer, with praiſe of God.

The 7.~key.}
prayer of a iuſt man in tribulation, with confidence and gratitude
towards God.}

\PTitle{Alleluia.}

\NV I
\SNote{I am induced to loue God,}
haue loued, becauſe our Lord
\SNote{becauſe he alwayes heareth my prayers.}
wil heare the voice of my prayer.

\V Becauſe he hath inclined his eare to me: and
\SNote{So long as I ſhal liue.}
in my daies I wil inuocate.

\V The ſorowes
\SNote{VVhen I erred like a ſtray ſheepe from thee, the paines of death,}
of death haue compaſſed me: and
\SNote{and the danger of hel-torments, both due for ſinne, inuironed
me, and I was not ware therof:}
the perils of hel haue found me.

I
%%% !!! SNote should be before 'I'
\SNote{But by tribulation falling vpon me, I came to knowe my dangerous
eſtate:}
haue found tribulation and ſorow: \V and I
\SNote{turned to thee, and prayed as foloweth:}
inuocated the name of our Lord.

O Lord deliuer my ſoule: \V our Lord is merciful, and iuſt, and our God
doth mercie.

\V Our Lord kepeth litle ones:
\SNote{Afflicted with tribulations.}
I was humbled, and he hath deliuered me.

\V Turne ô my ſoule into thy reſt: becauſe our Lord hath done good to
thee.

\V Becauſe he hath deliuered my ſoule from death: mine eies from teares,
my feete from ſliding.

\V I
\SNote{I wil endeuoure to pleaſe God, in the congregation of thoſe that
liue here in grace, and in heauen in glorie.}
wil pleaſe our Lord in the countrie of the liuing.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
