%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 106
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-106


\project douay-rheims


%%% 1332
%%% o-1222
\startChapter[
  title={Psalme 106}
  ]

\PSummary{Againe
\MNote{Gods perpetual prouidence towards al men.

The 3.~key.}
the Pſalmiſt inuiteth al men to render thankes to God for their
deliuerie from dangers, or euils in general: 4.~particularly from
dangers in iorney, 10.~in priſon, or captiuitie, 17.~in ſicknes
ſpiritual and corporal, 23.~in nauigation, 33.~deſcribing the
changeable courſe of thinges in this world, 38.~eſpecially of mens
ſtates; 42.~for al which the iuſt wil praiſe God.}

\PTitle{Alleluia.}

\NV Confeſſe
\SNote{Praiſe God by confeſſing his mercie, prouidence, and goodnes.}
ye to our Lord becauſe he is good:
\CNote{\XRef{1.~Par.~16. v.~34.}}
becauſe his mercie is for euer.

\V Let them ſay that are
\SNote{God of his mercie promiſed the Redemer of mankind ſtreight after
Adams fal:}
redemed of our Lord, whom he redemed out of the hand of the enemie: and
out of the countries he gathered them:

\V From
\SNote{VVhich redemption was intended for al, and faileth not of Gods
part in anie, but of mens owne wilful refuſing to be duly penitent, and
to kepe Gods precepts.}
the riſing of the ſunne, and the going downe: from the north, and the
ſea.

%%% 1333
\V They
\SNote{Literally of ſuch as wander in this world, hauing no ſetled place
to dwel in: ſpiritually of al mankind after his fal.}
wandered in the wildernes, in a place without water: the way of the
citie for habitation they found not.

\V Hungrie and thirſtie: their ſoule fainted in them.

\V And
\SNote{Whenſoeuer they cal vpon God, he helpeth them, as is beſt for
their ſpiritual health.}
they cried to our Lord when they were in tribulation: and he
%%% o-1223
deliuered them out of their neceſſities.

\V And he conducted them in to the right way: to goe into a citie of
habitation.

\V Let
\SNote{Al Gods benefites, which are of his mercie, not of mans deſerte,
are iuſt matter of praiſing God.}
the mercies of our Lord confeſſe to him: and his meruelous workes to the
children of men.

\V Becauſe he hath filled the emptie ſoule: and the hungrie ſoule he
hath filled with good thinges.

\V Them that ſate in darkenes, and in the shadow of death: bound in
needines, and yron.

\V Becauſe they
\SNote{Calamities in this world are commonly inflicted for ſinnes.}
exaſperated the wordes of God: and they prouoked the counſel of the
Higheſt.

\V And their hart was humbled in labours: they were weakened, neither
was there anie to helpe.

\V And they cried to our Lord when they were in tribulation: and he
deliuered them out of their neceſſities.

\V And he brought them out of darkenes, and the shadow of death: and
brake their bondes aſunder.

\V Let the mercies of our Lord confeſſe to him: and his meruelous workes
to the children of men.

\V Becauſe he hath deſtroyed the gates of braſſe: and the barres of yron
he hath broken.

\V He hath receiued them out of the way of their iniquitie: for they
were humbled for their iniuſtices.

\V Their ſoule did abhorre al meate: & they approched euen to the gates
of death.

\V And they cried to our Lord when they were in tribulation: and he
deliuered them out of their neceſſities.

\V He ſent his word, and healed them: and deliuered them out of their
deſtructions.

\V Let the mercies of our Lord confeſſe to him: and his meruelous workes
to the children of men.

%%% 1334
\V And let them ſacrifice the ſacrifice of praiſe: and shew forth his
workes in exultation.

\V They that goe downe into the ſea in shippes, making trafike in the
great waters.

\V They haue ſene the workes of our Lord, and his meruelous thinges in
the depth.

\V He ſayd, and the blaſt of the ſtorme ſtood: and the waues therof were
exalted.

\V They aſcend euen to the heauens, and they deſcend euen to the
depthes: their ſoule pyned away in euils.

%%% o-1224
\V They were trubled, and were moued as a drunken man: and al their
wiſedom was deuoured.

\V And
\SNote{As before in the 6,~13, and~19. verſes.}
they cried to our Lord when they were in tribulation, and he brought
them out of their neceſſities.

\V And he turned his ſtorme into calme: and the waues therof were quiet.

\V And they reioyced becauſe they were quiet: and he conducted them into
the hauen of their wil.

\V Let
\SNote{This verſe alſo is foure times in this Pſalme v.~8, 15, 21,
and~31. to admoniſh vs, that as there is one meanes to eſcape from al
dangers by crying to God, as v.~6, 13. 19. and~28. with mourning and
penance: ſo there is one cauſe of praiſe and thankes for our deliuerie,
which is Gods  mercie and grace.}
the mercies of our Lord confeſſe to him: and his meruelous workes to the
children of men.

\V And let them exalt him in the church of the people: and the chayre of
the ancientes let them praiſe him.

\V
\SNote{God to ſhew ſometimes his powre, alſo to benefite ſome, and to
puniſh others, changeth the accuſtomed courſe of thinges, and ſtates of
men, at his diuine pleaſure; as here, the Royal prophet reciteth ſome
examples. And ſome others are recorded in diuers times and places.}
He turned
\SNote{No doubt much change was made in the earth by Noes flood. And
manie thincke that the land of Chanaan, was made more fruictful in the
time of the Iewes inhabiting, and now is more barrane againe.}
the riuers into a deſert: and the iſſues of waters into drineſſe.

\V The
\SNote{He alludeth to the countrie about Sodome, and Gomorre, which was
moſt fruictful, and moſt pleaſant,
\XRef{Gen.~13. v.~10.}
but ſhortly after vvas burnt vvith fire and brimſtone,
\XRef{Gen.~19. v.~24.}
ſubuerted, and turned into a dead and ſalt ſea.}
fruiteful land into a ſalt ground, for the malice of them that inhabite
it.

\V He turned the deſert into pooles of waters: and the land without
water into iſſues of waters.

\V And he placed the hungrie there, and they built a citie of
habitation.

%%% 1335
\V And they ſowed fildes, and planted vineyardes: and they made fruicte
\SNote{Made abundance of fruict to grow.}
of natiuitie.

\V And he bleſſed them, and they were multiplied excedingly: and their
beaſtes he leſſened not.

\V And
\SNote{Againe ſome countries puniſhed for ſinnes.}
they were made fewe: and were vexed by the tribulation of euiles, and
with ſorow.

\V Contempt was powred out vpon princes: and he made them wander where
was no way, and not in the way.

\V And he
\SNote{An other change in releeuing the poore being humbled.}
did helpe the poore out of pouertie: and made families as sheepe.

\V The iuſt shal ſee, and shal reioyce: & al iniquitie shal ſtoppe her
mouth.

\V Who is wiſe and wil keepe theſe thinges? and wil vnderſtand the
mercies of our Lord?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
