%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 30
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-30


\project douay-rheims


%%% 1196
%%% o-1089
\startChapter[
  title={Psalme 30}
  ]

\PSummary{A
\MNote{How to pray in affliction.

The 7.~key.}
prayer of a iuſt man excedingly afflicted ſtil confident in God,
11.~deſcribing his manie calamities (in reſpect of his enemies)
vndeſerued, 18.~prayeth for his owne deliuerie, and their iuſt
  punishment. 20.~Praiſeth and thanketh God for his
goodnes. 24.~Exhorteth al others to do the ſame.}

\PTitle{Vnto
\SNote{Perteyning to the new Teſtament:}
the end, the Pſalme of Dauid, for
\SNote{eſpecially to the iuſt trubled and almoſt diſtracted in mind in
great affliction. See
\XRef{v.~23.}}
exceſſe of minde.}

\VV In thee ô Lord
\SNote{How greuouſly ſoeuer I am afflicted yet I truſt in thee,}
haue I hoped,
\SNote{therfore I pray thus.}
let me not be confounded for euer:
\CNote{\XRef{Pſal.~70.}}
in thy iuſtice deliuer me.

\V Incline thine eare to me, make haſt to deliuer me. Be vnto me for a
God protector: and for a houſe of refuge, that thou mayſt ſaue me.

\V Becauſe thou art my ſtrength, and my refuge: and for thy name thou
wilt conduct me, and wilt nourish me.

\V Thou wilt bring me out of this ſnare, which they haue hid for me:
becauſe thou art my protector.

\V
\CNote{Luc.~23.}
Into
\SNote{I offer and reſigne my ſelfe to thee.}
thy handes I commend my ſpirit: thou haſt
\SNote{Thou haſt often deliuered me. This verſe is moſt aptly applied in
al reſignatiõs of our ſoules to Gods wil, namely whenſoeuer we receiue
the B.~Sacrament; likewiſe in the houre of death, and at manie other
times.}
redeemed me ô Lord God of truth.

%%% 1197
\V Thou haſt hated them that obſerue vanities, vnprofitably.

But I haue hoped in our Lord: \V I shal reioyce and be ioyful in thy
mercie.

Becauſe thou haſt reſpected my humilitie, thou haſt ſaued my ſoule out
of neceſſities.

\V Neither
\SNote{Not ſuffered me to be shut vp.}
haſt thou shut me vp in the handes of the enemie: thou haſt ſet
\Fix{me}{my}{probable typo, same in both}
feete in a large place.

\V Haue mercie vpon me ô Lord becauſe I am in tribulation: myne
\Fix{eieis}{eie is}{obvious typo, fixed in other}
trubled for wrath, my ſoule, and my bellie:

\V Becauſe
\SNote{Al my partes external and internal, body and mind are trubled.}
my life is decayed for ſorrowe: and my yeares for gronings.

My ſtrength is weakened for pouertie; and
\Fix{by}{my}{obvious typo, fixed in other}
bones are trubled.

\V Aboue al myne enemies I am made a reproch both to my neighbours
excedingly: and
\SNote{My freindes dare not conuerſe with me, leſt they incurre
diſpleaſure for my ſake.}
a feare to my acquantance.

They that ſaw me, fled forth from me: \V I am forgotten, from the hart
as one dead.

%%% o-1090
I am made as a veſſel deſtroyed: \V becauſe I haue heard the
reprehenſion of manie that abide round about:

In that whiles, they aſſembled together againſt me, they conſulted to
take my ſoule.

\V But I haue hoped in thee ô Lord: I ſayd: Thou art my God: \V my
lottes are in thy handes.

Deliuer me out of the handes of my enemies, and from them that perſecute
me.

\V
\SNote{Make thy light ſo ſhine in my ſoule, that I may vnderſtand that
is right,}
Illuſtrate thy face vpon thy ſeruant,
\SNote{and through thy mercie deliuer me from the force of myne
aduerſaries.}
ſaue me in thy mercie:

\V Lord let me not be confounded, becauſe I haue inuocated thee.

Let the impious be ashamed, and brought downe into hel: \V let the
deceitful lippes be made mute.

Which ſpeake iniquitie againſt the iuſt,
\SNote{So proudly behauing them ſelues, as if they had no ſuperior,
neither in earth nor in heauen, to whom they ſhal at laſt render
account:}
in pride, and
\SNote{and abuſing their preſent powre and authoritie, which they haue
of God.}
abuſe.

%%% 1198
\V How great is the multitude of thy ſweetneſſe ô Lord; which thou
\SNote{As yet in this preſent life the reward of the iuſt is hidde:}
haſt hid for them that feare thee.

Thou haſt perfited it for them that hope in thee,
\SNote{but shal be made manifeſt in ſight of al men.}
in the ſight of the children of men.

\V Thou
\SNote{In the meane time the iuſt is in great eſtimation in the ſecrete
knowledge of God.}
shalt hide them in the ſecrete of
\SNote{Title of honoure, as we ſpeake to a king, \Emph{your maieſtie}:
or to a noble man, \Emph{your Lordshippe}.}
thy face from the diſturbance of men.

Thou shalt protect them in thy tabernacle from the contradiction of
tongues.

\V Bleſſed be our Lord: becauſe he hath made his mercie meruelous to me
in the fenſed citie.

\V But I haue ſaid in
\SNote{In myn extreme affliction being almoſt diſtracted in my mind, I
ſaid that in reaſon I would not haue ſayd. Holie Iob ſpake ſome thinges
in ſuch ſtate of affliction.
\XRef{ch.~3.}
&
\XRef{42.}}
the exceſſe of my minde: I am caſt away from the ſight of thine eies.

Therfore thou haſt heard the voice of my praier, whiles I cried to thee.

\V
\SNote{The prophet, or other iuſt perſon exhorteth al the ſeruants of
God,}
Loue our Lord al ye his ſainctes: becauſe our Lord wil require truth,
& wil repay them abundantly that doe proudly.

\V Do ye
\SNote{to conſtancie,}
manfully, and let your
\SNote{longanimitie,}
hart take courage, al ye that
\SNote{and final perſeuerance to the end.}
hope in our Lord.


\stopChapter

\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
