%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 54
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-54


\project douay-rheims


%%% 1239
%%% o-1131
\startChapter[
  title={Psalme 54}
  ]

\PSummary{The
\MNote{Gods prouidence towards the good and bad.

The 3.~key.}
prophet (as wel in his owne, as other iuſt mens perſon) deſcribeth great
calamities ſuffered, 10.~prayeth againſt the wicked, 13.~lamenting
eſpecially that thoſe which profeſſe frendshipe, are aduerſaries,
17.~and declareth Gods prouidence in protecting the good, and deſtroying
the bad.}

\PTitle{Vnto
\SNote{A ſong as wel for king Dauid himſelfe, as others of al times;}
the end,
\SNote{to ſing,}
in ſongs,
\SNote{and conſider Gods prouidence, in ſuffering one man to afflict an
other in this life.}
vnderſtanding to Dauid.}

\VV Heare my prayer ô God, deſpiſe not my petition.

\V Attend to me; and heare me.

\V I am made ſorowful in my
\SNote{This life is a warefare, and a continual combate.}
exerciſe: and am trubled at the voice of the enimie: and at the
tribulation of the ſinner.

Becauſe they haue
\SNote{Calumniated me,}
wreſted iniquities vpon me: &
\SNote{and perſecuted me in great furie.}
in anger they were trubleſome to me.

\V My hart is trubled in me: and
\SNote{So inwardly afflicted, as if death were at hand.}
the feare of death is falne vpon me.

%%% 1240
\V Feare and trembling are come vpon me: and
\SNote{I haue ſcarſe ſenſe, or diſcourſe of reaſon, being almoſt
ouerwhelmed with trubles.}
darkenes hath couered me.

\V And I ſaid:
\SNote{Would God I could flie, that in the ſimplicitie of a doue, I
might ſpedely part away from theſe afflictions.}
Who wil geue me wings as of a doue, and I wil fly: and reſt?

\V
\SNote{I haue fled ſo farre as I could from trubles.}
Loe I haue gone far flying away: and I abode in the wildernes.

\V I
\SNote{For the reſt I remitted to Gods wil and good pleaſure,}
expected him, that
\SNote{and he ſuffered me not to be ouer throwne,}
ſaued me from
\SNote{though I am weake,}
puſillanimitie of ſpirit, and
\SNote{and the tentations are great.}
tempeſt.

\V
\SNote{O God abate the pride of arrogant perſecuters,}
Precipitate ô Lord, and
\SNote{ſuffer them not to agree amongſt them ſelues.}
diuide their tongues: becauſe I haue ſene
\SNote{They are ful of al iniquitie,}
iniquitie, and
\SNote{they haue alſo contentions among themſelues, turne the ſame to
our good.}
contradiction in the citie.

\V Day and night shal iniquitie compaſſe it vpon the walles therof; and
\SNote{VVith their continual great iniquitie, they haue their trubles,}
labour in the middeſt therof, and
\SNote{but leaue not their iniuſtice.}
iniuſtice.

\V And there hath not ceaſed out of the ſtreetes therof
\SNote{They are ſtil vſurers, and deceiptful oppreſſors of the poore.}
vſurie, and guile.

%%% o-1132
\V For
\SNote{It is a greater griefe to ſuffer iniuries of thoſe that ſeme to
be freindes.}
if myne enimie had ſpoken euil to me, I would haue borne it.

And if he that hated me had ſpoken great thinges vpon me: I would
perhaps haue hid myſelfe from him.

\V But
\SNote{A man that was, or ſemed of the ſame mind, faith, and religion,}
thou a man of the ſame minde: my
\SNote{whom I ſo truſted, that I would haue gone, whitherſoeuer he
should haue ledde me.}
guide, and my familiar.

\V Which diddeſt
\SNote{Thou that didſt participate the ſame holie ſacraments with me,}
take ſwete meats together with me, in the
\SNote{in the Catholique Church.}
houſe of God we walked with conſent.

\V Let
\SNote{As Core & his complices: ſpoken of iuſt zele, not of deſire to
reuenge: verified in thoſe that ſinne wittingly and knowing, for they
deſcend, as it were, aliue into hel.}
death come vpon them: and let them goe downe quicke into hel.

Becauſe there is wickednes in their habitations, in
\SNote{The whole crew of the wicked conſpire in iniquitie.}
the middes of them.

%%% 1241
\V But I haue cried to God, and our Lord wil ſaue me.

\V In
\SNote{The prophet alludeth to three more ſpecially appointed houres of
Diuine ſeruice, the daylie ſacrifice at morning, and euening, and other
ſacrifices commonly about midday. VVhich alſo are the three principal
times of Diuine Seruice in the Church of Chriſt. Mattins, Euenſong, and
the Sacrifice of Maſſe. VVhich Eutymius and other Grecians cal
\L{Lyturgiam}. S.~Clement alſo
\Cite{(li.~7. c.~25. Apoſt. Inſtit.)}
teſtifieth that the Apoſtles ordained three ſet houres of common prayer
euerie day.}
the euening, and morning, & at midday, I wil
\TNote{or pray.}
ſpeake, and declare, and he wil heare my voice.

\V He wil redeme my ſoule in peace from them, that approch to me:
becauſe among manie
\SNote{Manie enimies combined together approched vnto me, to ouerthrow
me.}
they were with me.

\V God wil heare, and he
\SNote{Eternal God.}
which is before the worldes wil humble them.

\V For there is
\SNote{They wil neuer repent of their wickednes.}
no change with them, & they feared not God: he hath ſtreached forth his
hand in repaying.

\V They haue contaminated his teſtament, they are
\SNote{They harden their hartes againſt his threatned wrath:}
diuided by the wrath of his countenance; and
\SNote{but Gods
\Fix{prouidente}{prouidence}{obvious typo, fixed in other}
illuminateth others to know and teach the truth, when it is impugned, or
contemned.}
his hart hath approched.

His
%%% !!! SNote goes before 'His'
\SNote{Gods wordes, which in them ſelues are meke and ſwete,}
wordes are made ſofter then oile: and
\SNote{are hard to the incredulous, & as dartes that wound them. Chriſt
ſaid
\XRef{(Ioan.~6.)}
Vnles one eate my flesh, & drinke my bloud, he shal not haue life in him,
which the Capharnaites not vnderſtanding, ſaid one to an other: This is
a hard ſpeach, who can abide it? which S.~Auguſtin here ſaith was the
firſt hereſie againſt our Sauiours preaching. It was not hard to
S.~Peter, who in the name of the reſt, anſwered, that Chriſt had the
wordes of eternal life. He yet vnderſtood not the ſecrete of our Lords
ſpeach, but he piouſly beleued that the wordes vvere good, vvhich he
vnderſtood not.}
the ſame are dartes.

%%% o-1133
\V
\SNote{Therfore in al doubtes of doctrin, in al diſtreſſes of
perſecution, and other difficulties vvhich ſurpaſſe thy vveakenes, caſt
thy care vpon our Lord, and he vvil nourish thee.}
Caſt thy care vpon our Lord: and he wil nourish thee: he wil not geue
\SNote{He vvil not ſuffer the iuſt to remaine alvvayes in fluctuation,
that is, in doubtful, dangerous, and vvauering thoughtes or
perplexities, as vvhen a shippe is toſſed in the vvaues of the ſea, but
vvil geue quiet repoſe of mind, as in a ſure hauen vvithout danger of
drovvning.}
fluctuation to the iuſt for euer.

\V But thou ô God wilt bring
\SNote{Contrariwiſe, the wicked and obſtinate ſhal fal into
deſtruction.}
them downe into the pitte of deſtruction.

Bloudy
%%% !!! SNote goes before 'Bloudy'
\SNote{Often or for moſt part, bloud-ſuckers dye before the courſe
of nature requireth, as Saul, Abſolon, Achitophel, Achab, Iezabel, and
the like.}
and deceitful men shal not liue halfe their daies, but I wil hope
in thee ô Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
