%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 089
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-089


\project douay-rheims


%%% 1302
%%% o-1192
\startChapter[
  title={Psalme 89}
  ]

\PSummary{Vnder
\MNote{Man rightly created fel by ſinne into miſeries.

The 2.~key.}
the forme of prayer, the Pſalmiſt deſcribeth the shortnes of mans life
and other calamities: 7.~Gods ſtrict iudgement, 13.~but firſt his
comfortable mercie: 16.~and perpetual regard of his owne worke.}

\PTitle{A prayer
\SNote{Some Expoſitors thincke Moyſes was the author of this Pſalme, and
of the tenne next folowing. But others hold that Dauid vvas author of
al, and that Moyſes his name is here put in the title by Eſdras, becauſe
this Pſalme is like to the prayer of Moyſes, vvhen the people prouoked
Gods vvrath by their ſinnes in the deſert. And becauſe mans creation,
fal, puniſhmẽt, and Gods mercie tovvards him, are here deſcribed: which
Moyſes firſt vvritte, as going before the vvritten lavv. And that Moyſes
made not this Pſalme is probably gethered by the
\XRef{10.~verſe,}
where the ordinarie age of men is deſcribed to be (in ſtreingth and
vigore) ſeuentie yeares, or of ſome fourſcore: and the greater part (of the
one or the other) is in labour and ſorovv. And it is euident
\XRef{(Deut.~34.)}
that Moyſes liued in al an hundred and tvventie yeares, and his eye was
not dimme, neither vvere his teeth moued. So Aaron, Ioſue, and others
commonly liued longer then is here mentioned. But Dauid vvas old and
impotent at ſeuentie yeares.
\XRef{3.~Reg.~1.}
S.~Hilarion, liuing neere ſeuentie yeares in his hermitage, S.~Remigius
gouerning the Church of Rhemes, ſeuentie yeares, and the like
are accounted to 
haue bene ful of dayes, and ſuch as liued longer are reputed
extraordinarie. Againe it is more
\Fix{euidenly}{euidently}{obvious typo, fixed in other}
proued Moyſes vvas not author of the
\XRef{94.}
and
\XRef{95.~Pſalmes.}}
of Moyſes the man of God.}

\NV Lord, thou art made a refuge for vs:
\SNote{Alwayes from the beginning of the world to the end.}
from generation vnto generation.

\V Before
\SNote{The Prophet sheweth that the world was created in and with time,
not eternal.}
the mountaines were made, or the earth and the world formed:
\SNote{And that only God is eternal.}
from euerlaſting euen vnto euerlaſting thou art God.

%%% o-1193
\V Turne not away man into humiliation: thou ſaidſt:
%%% !!! not marked in either
\SNote{God hath often ſaide, that he vvould not the death of ſinners,
but rather that they be conuerted and liue for euer.}
Be conuerted ye children of men.

%%% 1303
\V Becauſe
\SNote{Though ſome liued long (none for al that did reach to a
thouſand yeares) yet it is nothing before God, and in reſpect of
eternitie.}
a thouſand years before thine eies, are as yeſterday, that is paſt.

And as a watch in the night, \V thinges that are counted nothing shal
their years be.

\V In
\SNote{The youth of man quickly paſſeth:}
the morning as an herbe he shal paſſe, in the morning he shal florish,
and paſſe: in
\SNote{old age can not laſt long: vvherof cometh our Engliſh prouerb: A
young man may dye ſovvne, an old man can not liue long.}
the euening he shal fal, be hardened, and withered.

\V Becauſe we haue faynted in thy wrath, and in thy furie we are
trubled.

\V Thou haſt put
\SNote{Sinne the cauſe of shortnes of mans life.}
our iniquities in thy ſight: our
\TNote{\L{Sæculum.}}
age in the light of thy countinance.

\V Becauſe al our daies haue failed: and in thy wrath we haue failed.

Our yeares shal be conſidered
\SNote{Mans life as brickle as a ſpiders vveb: or mans life vvaſteth
continually, as a ſpider vvaſteth her ſelf by ſpinning, and conſuming
her ovvne ſubſtance.}
as a ſpyder: \V the daies of our yeares in them, are
\SNote{Theſe numbers literarly ſhew the ſhortnes of the longer ſorte of
mens liues. Myſtically, ſeuen ſignifie the reſt after laboures of this
vvorld, and perteyn to the old teſtament: eight ſignifie the revvard in
the reſurrection, perteyning to the nevv teſtament. VVhich multiplied by
tenne, a perfect number, make ſeuentie and eightie. VVhich ioyned
together make an hundred and fiftie. The number of al theſe Pſalmes.}
ſeuentie yeares.

And if in ſtrong ones
%%% !!! This Snote and the previous one are the same note.
%%% \SNote{}
eightie years: and the more of them, labour and ſorrow.

Becauſe
\SNote{It is of Gods milde prouidence, that mans life is ſhort, for that
manie if they vvere ſure, or had probabilitie to liue long, vvould
preſume to ſinne more.}
mildnes is come vpon vs: and we shal be chaſtiſed.

\V
\SNote{Seing God of his iuſtice puniſhed al mankind, for one ſinne of
our firſt parente, his vvrath muſt nedes be very great to euerie ſinner,
for his ovvne proper ſinnes.}
Who knoweth the powre of thy wrath: and for feare \V to number thy
wrath?

So make thy righthand knowne: and men learned in hart, in wiſedome.

\V Turne ô Lord, how long? and be intreated for thy ſeruants.

\V
\SNote{The hope of glorious reſurrection turneth our calamities into
ſpiritual ioy.}
We are replenished in the morning with thy mercie: and we haue
reioyced, and are delighted al our daies.

%%% 1304
%%% o-1194
\V
\SNote{Yea the more we ſuffer in this life for the truth, the greater is
our comforth in hope of reward.}
We haue reioyced for the daies wherin thou haſt humbled vs: the yeares,
wherin we haue ſeene euils.

\V Looke vpon
\SNote{Not only in that we are thy creatures, but alſo in that we are
thy ſeruants, we are thy proper worke, therfore in both theſe reſpectes,
ô God looke vpon vs with clemencie.}
thy ſeruants, and vpon thy workes: and
\SNote{Lead alſo our poſteritie into the right way, and make them thy
ſeruantes.}
direct their children.

\V And
\SNote{Ô God illuminate our vnderſtanding,}
let the brightnes of our Lord God be vpon vs, and
\SNote{make our actions by thy grace profitable to vs:}
direct thou the workes of our handes ouer vs: and
\SNote{and make perfect in vs the worke of charitie. In which one worke,
al good workes are included and to which al other are directed. For then
workes are right (ſayth
%%% !!! Cite ?
S.~Auguſtin) when they are directed to this one end.}
the worke of our handes doe thou direct.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
