%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 115
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-115


\project douay-rheims


%%% 1345
%%% o-1236
\startChapter[
  title={Psalme 115}
  ]

\PSummary{A
\MNote{Thankes-geuing for our Redemer.

The 5.~key.}
iuſt man acknowledging that ſpiritual life beginneth by faith, and by
publique profeſſion therof, 4.~gratfully accepteth of Chriſts
Redemption, 5.~dedicateth his life, and al he hath to Gods ſeruice.}

\PTitle{Alleluia.}

\NV I
%%% !!! First SNote should go before 'I'
\SNote{This in the Hebrew is ioyned to the next Pſalme before.}
\SNote{I beleued that God would helpe me:}
beleued,
\SNote{therfore I freely profeſſed that I truſted in him. For then in
dede faith is perfect, when we confeſſe with mouth, that which we beleue
in hart.}
for which cauſe I ſpake: but I
\SNote{I was vehemently afflicted in tribulations.}
was humbled excedingly.

%%% 1346
\V I ſaid
\SNote{In the middes of my great affliction I profeſſed, that al mans
helpe is vaine, falſe,
\Fix{deceipful,}{deceiptful,}{likely typo, fixed in other}
and defectiue, and therfore our truſt muſt be in God only.}
in mine exceſſe: Euerie man is a lyer.

\V What
\SNote{Conſidering that God hath not only geuen, and beſtowed manie
great benefites vpon me, and al mankind, but alſo hath rendered good for
euil, mercie for our ſinnes, we hauing rendered euil for good: what now
ſhal I render, ſayth a true penitent, for al that he hath thus rendered
to me, deſeruing ſo euil?}
shal I render to our Lord, for al thinges that he
%%% !!! Same as previous SNote.
%%% \SNote{}
hath rendred to me?

\V I
\SNote{Seing I am not able to render anie thing worthie of Gods fauoure
to me, yet I wil do that I can: I wil gratfully accept his great
benefite, the cuppe of Chriſts paſſion, which he
\Fix{druunke}{drunke}{obvious typo, fixed in other}
for mankind,}
wil take the chalice of ſaluation: and
\SNote{and wil praiſe, and cal vpon his name.}
I wil inuocate the name of our Lord.

\V I wil
\SNote{I wil pay voluntarie vowes,}
render my vowes to our Lord
\SNote{for Gods glorie, and edification of others.}
before al his people: \V
\SNote{Yea I wil offer my life, and ſuffer death, when Gods glorie ſhal
require it, in whoſe ſight, the death of Sainctes is precious, and moſt
highly eſteemed.}
precious in the ſight of our Lord is the death of his Sainctes.

\V O Lord becauſe
\SNote{Alvvayes vnderſtood, that ſuch as ſuffer perſecution, be in good
ſtate of their ſoules, the true ſeruantes of God,}
I am thy ſeruant: I am thy ſeruant, and
\SNote{the children of the Church his handmaide.}
the ſonne of thy handmaide.

Thou haſt
\SNote{Deliuered me from captiuitie of ſinne.}
broken my bondes: \V I wil ſacrifice to thee the hoaſt of praiſe, and I
wil inuocate the name of our Lord.

\V I wil render my vowes to our Lord in the ſight of al his people: \V
in the courtes of the houſe of our Lord, in the middes of
\SNote{In the Church of the faithful.}
thee ô Ieruſalem.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
