%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 109
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-109


\project douay-rheims


%%% 1338
%%% o-1228
\startChapter[
  title={Psalme 109}
  ]

\PSummary{Chriſt
\MNote{Chriſts exaltation.

The 5.~key.}
riſing and aſcending into heauen ſitteth on the right hand of God:
2.~beginning in Ieruſalem reigneth in the Church of the whole earth:
4.~vſeth the Prieſthood of Melchiſedechs order to the end of the
world. 6.~And shal iudge the world.}

\PTitle{A pſalme of Dauid.}

\NV Our
%%% !!! CNote should go before 'Our'
\CNote{\XRef{Mat.~22.}
\XRef{Act.~2.}
\XRef{1.~Cor.~15.}
\XRef{Heb.~1.}
&
\XRef{10.}}
\SNote{God the Father}
Lord ſayd
\SNote{to God the Sonne, the Lord of Dauid, and of al mankind, yet the
ſonne of Dauid, according to his humanitie.}
to my Lord: Sitte on my right hand:
\SNote{He limiteth not the time, but excludeth al time, wherin the
enimie might imagine, that Chriſts kingdom ſhould ceaſe: ſignifying that
Chriſt ſhal reigne, til al his enimies be ſubdued, much more afterwards
in al eternitie.}
til I make thine enemies, thy footeſtoole of thy feete.

\V Our Lord wil ſend forth the rod of thy ſtrength
\SNote{The Church of Chriſt beginning in Ieruſalem on whitſunday, the
fiftith day from his Reſurrection, continueth euer more.}
from Sion: rule thou in the middes of thine enemies.

\V With thee
\SNote{Thou ſhalt haue principalitie,}
the beginning in the day
\SNote{in the day of thy powrful conqueſt, and riſing from death,}
of thy ſtrength, in
\SNote{in excellencie of al holie ſpiritual myſteries and graces:}
the brightnes of holie thinges: from the wombe
\SNote{becauſe, I God the Father of my ſubſtance begate thee, God the
Sonne in eternitie. The ſame which Micheas ſaith
\XRef{(c.~5. v.~2.)}
His coming forth from the beginning, from the dayes of eternitie.}
before the day ſtarre I begat thee.

\V Our Lord
\SNote{God moſt firmly, and vnchangeably affirmed: that thou (Chriſt our
Meſſias) art not only a King, but alſo a Prieſt:}
ſware, and it shal not repent him:
\CNote{\XRef{Heb.~5. v.~7.}}
Thou art
\LNote{A Prieſt for euer.}{In
\MNote{Chriſts Prieſthood for euer both in function and in effect.}
two reſpectes Chriſt is a Prieſt for euer: in that from the firſt
inſtant of his Incarnation he was, and remaineth a Prieſt, now alſo in
heauen: and al other Prieſtes are his miniſterial vicares, not
ſucceſſors. So that al prieſtlie functions, which they doe, he by them
doth the ſame, as the principal Prieſt. VVherupon ſaith S.~Paul
\XRef{(1.~Cor.~4.)}
So let a man thincke of vs, as of the miniſters of Chriſt, and
diſpenſers of the myſteries of God. Secondly Chriſt dayly offering
Sacrifice by the handes of his Prieſtes, doth continually pacifie Gods
wrath, in behalf of thoſe ſinners, for whom it is duly applied, euen to
the end of the world. VVheras the Prieſthood of Aaron, and of al others
in the old Teſtament, ceaſſed by their deathes, both in the office, and
in the effect.}
a Prieſt
\SNote{not for a time, as Aaron was, but for euer,}
for euer
\LNote{According to the order of Melchiſedech.}{As
\MNote{The reſemblance of Chriſts and Melchiſedecs Prieſthood.}
Melchiſedech, king of peace and iuſtice, without father, mother, or
genealogie, expreſſed in holie Scriptures, or otherwiſe knowen to the
world, was Prieſt of the Hiegheſt, offered bread and wine, an vnbloudie
ſacrifice; communicating with both Chananeites and Hebrewes, bleſſed
Abraham, and tooke tithes of him, and his ſubiectes: ſo Chriſt the true
King of peace & iuſtice, without father of his humanity, without mother
of his Diuinitie, the Sonne of God, of ineffable genealogie, borne of a
virgin in his humanitie, the Prieſt of God, offereth Sacrifice (not only
bloudie on the Croſſe, but alſo) vnbloudie in the formes of bread and
wine, continueth the ſame by the miniſterie of other Prieſtes, maketh al
nations partakers therof, bleſſeth them and receiueth of them al dutiful
and religious ſeruice, as of his ſubiectes.}
according to
\SNote{neither of Aarons order, but according to the Order of
Melchiſedec.}
the order of Melchiſedech.

%%% 1339
\V Our Lord on thy righthand,
\SNote{Kinges that ſometimes perſecute Chriſtians, are ſubdued with
other people to Chriſt.}
hath broken kinges in the day of his wrath.

\V He
\SNote{He ſhal iudge and puniſh the incredulous people,}
shal iudge in nations, he
\SNote{make great ſlaughters amongſt thoſe that reſiſt:}
shal fil ruines, he
\SNote{and bring princes with their populous kingdoms to nothing.}
shal crush the heads in the land of manie.

\V
\SNote{He ſhal in the meane time (and alſo his beſt ſeruants) ſuffer
much tribulation in this life:}
Of the torrent in the way he shal drinke:
\SNote{and for the ſame be highly exalted in life euerlaſting.}
therfore shal he exalt the head.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
