%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 147
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-147


\project douay-rheims


%%% 1397
%%% o-1288
\startChapter[
  title={Psalme 147}
  ]

\PSummary{Againe
\MNote{Gods prouidence eſpecially tovvards the Church.

The 6.~key.}
God is to be praiſed for his goodnes towards his peculiar people, 4.~yea
towards al the world: 7.~but moſt abundantly towards his Church.}

\PTitle{Alleluia.}

\NV O
\MNote{The Hebrews ioyne this Pſalme vvith the precedent.}
\SNote{Ieruſalem in the latter part of Dauids time, al the time of
Salomon, and part of other kings reignes til the captiuitie, had peace
& proſpered. Againe after the captiuitie, the Citie was repared, the
Temple reedified, and the whole land receiued and enioyed manie
bleſſinges. But al this was no more then a figure of the excellent
benefites here prophecied, and more euidently verified in Chriſts
Catholique Church: partly here militant in the whole world, and
eſpecially in the glorious Ieruſalem, and Sion, the perfect viſion of
peace, and contemplation of God, in eternal life.}
Ieruſalem praiſe our Lord: praiſe thy God ô Sion.

%%% 1398
\V Becauſe he
\SNote{In compariſon of other cities, and peoples of the world, the
gates of Ieruſalem were ſtrongly fenſed,}
hath ſtrengthened the lockes of thy gates: he
\SNote{and the citizens bleſſed: much more the Church of Chriſt is built
vpon a ſure rock, her faithful children indued with al ſpiritual graces;
and moſt of al, heauen it ſelfe is free from al danger of calamitie, and
the Sainctes are moſt ſecure & moſt happie, enioying eternal fruition of
God.}
hath bleſſed thy children in thee.

\V Who
\SNote{Hath geuen peace in thy borders,}
hath ſet thy borders peace: and filleth thee with
\SNote{and the very beſt corne, and al other fruictes: wine, oyle,
milke, honey, and the reſt. Allegorically in the Church, reconciliation
with God, by remiſſion of ſinnes, and peace of conſcience, in the
Sacraments of Baptiſme & Penance; vvith the moſt ſpiritual food of
Chriſts Bodie and Bloud in the Euchariſt, and graces of other
Sacraments. In heauen moſt aſſured peace and ioy without end.}
the fatte of corne.

\V Who ſendeth forth
\SNote{This perteyneth moſt ſpecially to Euangelical doctrine, preached}
his ſpeach to the earth: his word
\SNote{and quickely receiued in al the world.
\XRef{Rom.~10. v.~18.}}
runneth ſwiftly.

\V Who geueth
\SNote{Snow nouriſheth the earth, making it vvarme by Antiperiſtaſis, as
is euident in natural Philoſophie,}
ſnow as wool: ſcatereth
\SNote{and noyſome ayre is changed into clere vveather. In the Church
by penance & auſtere life men are purged from ſinnes and vices, euil
ſpirites are alſo driuen avvay. If your ſinnes ſhal be as ſcarlet, they
ſhal be made vvhite as ſnovv, and if they be redde as vermilion, they
ſhal be vvhite as vvool.
\XRef{Iſaie.~1. v.~18.}}
miſt as ashes.

\V He caſteth
\SNote{Yea ſome that are hardned in ſinne, as yſe, or chryſtal, ſhal be
melted, broken, or made fitte to be ingraſſed in Gods Church. So
S.~Peter vvas admoniſhed by a viſion
\XRef{(Act.~10. v.~13.)}
to kil and eate.}
his cryſtal as morſeles: before the face of his cold
\SNote{Othervviſe vvithout Gods grace geuing remorſe and
ſorovv, no man can ouercome his ovvne vices.}
who shal endure?

\V He
\SNote{But Gods vvord preached,}
shal ſend forth his word, and shal melt them:
\SNote{and his grace touching mens hartes,}
his ſpirit shal blowe, and
\SNote{innumerable are conuerted.}
waters shal flowe.

%%% o-1289
\V Who declareth his word to
\SNote{The Church only enioyeth theſe ſpiritual benefites.}
Iacob: his iuſtices, and iudgements to Iſrael.

\V He hath
\SNote{Conſidering that al mankind vvas in the maſſe of ſinne, and that
God letteth manie iuſtly periſh, thoſe to whom he geueth his grace to
iuſtification, are ſpecially bond to praiſe him.}
not done in like maner to anie nation: and his iudgements he hath not
made manifeſt to them.
\SNote{And therfore the Prophet concludeth this Pſalme, and the reſt
folovving, vvith Alleluia.}
Alleluia.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
