%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 88
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-88


\project douay-rheims


%%% 1298
%%% o-1188
\startChapter[
  title={Psalme 88}
  ]

\PSummary{Gods
\MNote{The Church of Chriſt neuer faileth.

The 6.~key.}
mercie and truth, with his great promiſes to Dauid, 6.~his powre in the
whole world, and iuſt iudgements, are the true ioy of his
ſeruantes. 20.~Chriſts kingdom shal remaine for euer: 31.~yea manie
offending yet al shal not perish, 39.~but after great affliction,
47.~God wil reſpect mans infirmitie, 50.~his owne promiſe, and the
enimies reproching his ſeruantes and himſelfe: 53.~who is bleſſed for
euer.}

\PTitle{Of vnderſtanding
\SNote{Otherwiſe called Idithun
\XRef{1.~Paral.~25.}
or rather Ethan, who was very wiſe, mentioned with others, to whom
Salomon is preferred for wiſdom.
\XRef{3.~Reg.~4. v.~31.}
and ſignifieth ſtrong, applied here to thoſe that are ſtrong in aſſured
hope of Chriſts promiſes, notwithſtanding it ſemeth ſometimes to the
weake, that his promiſes are not performed.}
to Ethan the Ezrahite.}

%%% !!! There are several 'regular' SNotes in this chapter, as well as
%%% !!! the normal Pſalme SNotes marked by letters. Do I want to
%%% !!! distinguish between these?

\VV The mercies of our Lord I wil ſing for euer.

In generation
\SNote{In al generations.}
and generation I wil shewe forth thy truth, in my mouth.

%%% o-1189
\V Becauſe thou ſaidſt: Mercie shal be built vp for euer
\SNote{The heauens ſhal rather fal, then Gods truth faile. Myſtically,
in the Apoſtles, and by their preaching, the Church of Chriſt is built
for euer.}
in the heauens: thy truth shal be prepared in them.

\V I haue ordained a teſtament with mine elect, I haue ſworne to Dauid
my ſeruant: \V
\SNote{Dauids ſeede conſerued til Chriſt was borne of his virgin mother:
and in his ſpiritual ſeede, his kingdom the Church is for euer
conſerued. Otherwiſe not verified of Dauids temporal kingdom, which
decayed in the captiuitie of Babylon, and is now wholly deſtroyed.}
for
\Fix{ever}{euer}{obvious typo, fixed in other}
wil I prepare thy ſeede. And I wil build thy ſeat vnto
generation and generation.

\V The heauens shal confeſſe thy meruelous workes ô Lord: yea and thy
truth in the church of ſaintes.

%%% 1299
\V For who in the cloudes shal be equal to our Lord: shal be like to God
among
\SNote{The Angels.}
the ſonnes of God?

\V God, who is glorified in the counſel of ſaintes: great and terrible
ouer al that are round about him.

\V Ô Lord God of hoaſtes who is like to thee? thou art mightie ô Lord,
and thy truth round about thee.

\V Thou ruleſt ouer the powre of the ſea: and the mouing of the waues
therof thou doeſt mitigate.

\V Thou
\SNote{The prophet aludeth to the plagues and miracles in Ægypt, and in
other enimies.}
humbledſt the proud one, as one wounded: in the arme of thy ſtrength
thou haſt diſperſed thine enimies.

\V The heauens are thine, and the earth is thine, the round earth, and
the fulnes therof thou haſt founded: \V the north, and the ſea thou haſt
created.

Thabor
%%% !!! SNote belongs before 'Thabor'
\SNote{Conuerſion of Gentiles.}
and Hermon shal reioice in thy name: \V thy arme is with might.

Let
\SNote{VVhether God puniſh, as with the left hand,}
thy hand be confirmed, and thy
\SNote{or beſtow benefites, as with the right hand, al is to his glorie,
and according to mercie, and truth.}
righthand exalted: \V iuſtice and iudgement is the preparation of thy
ſeat.

Mercie and truth shal goe before thy face: \V
\SNote{They are ſpiritually happie, that do thus conſider of Gods
meruelous procedings, praiſe the ſame, and reioyce therin.}
bleſſed is the people that knoweth iubilation.

Lord they shal walke in the light of thy countinance, \V and in thy name
they shal reioyce al the day: and in thy iuſtice they shal be exalted.

\V Becauſe thou art the glorie of their ſtreingth: and in thy good
pleaſure shal our
\SNote{Powre, and kingdom.}
horne be exalted.

\V Becauſe our protection is of our Lord: and of the holie one of Iſrael
our king.

\V Then didſt thou ſpeake in viſion to thy ſaintes, and ſaidſt
\SNote{Thus God promiſed to eſtabliſh the kingdom of the Iewes in Dauid
and his familie.
\XRef{1.~Reg.~16.}
\XRef{2.~Reg.~5.}
and other places. Which was performed as in a figure but more fully in
Chriſt.
\XRef{Act.~13. v.~22.}}
I haue put
%%% o-1190
helpe on the mightie one: and haue exalted an elect one of my people.

\V I haue found Dauid my ſeruant: with myne holie oyle haue I anointed
him.

\V For mine hand shal helpe him: and myne arme shal ſtrengthen him.

\V The enimie shal nothing preuale in him: and the ſonne of iniquitie
shal not adde to hurt him.

\V And I wil cut downe his enimies before his face: and them that hate
him I wil put to flight.

%%% 1300
\V And my truth, and my mercie with him: and in my name shal his horne
be exalted.

\V And I wil put his hand in the ſea: and his righthand in the riuers.

\V He shal inuocate me: Thou art my Father: my God, and the protector of
my ſaluation.

\V And I wil put him the firſtbegotten, high aboue the kings of the
earth. 

\V I wil kepe my mercie vnto him for euer: and my teſtament faithful to
him.

\V I wil put his ſeed for euer and euer: and his throne as the daies of
heauen.

\V But if his children shal forſake my lawe: and wil not walke in my
Iudgementes:

\V If they shal profane my iuſtices: and not kepe my commandmentes:

\V I wil viſite their iniquities with a rod: and their ſinnes with
ſtripes:

\V But
\LNote{My mercie I vvil not take avvay from him.}{Although,
\MNote{Though Chriſtians do ſinne, yet Chriſt loſeth not his Church.}
Chriſtians ſignified by the childrẽ or ſucceſſors of Dauid, ſinne moſt
grieuouſly, yea ſuppoſe they wil ſinne with deſperation (ſaith
%%% !!! Cite ? upon this place?
S.~Auguſtin) and obſtinatly perſiſt in ſinne, that they offend the eyes
of their Father, & deſerue to be diſenherited: &c. Yet for theſe Chriſt
ſhal not remaine without inheritance, the corne ſhal not alſo periſh for
the chafe, ſome fiſhes ſhal be geathered out of the nette into veſſels,
notwithſtanding the euil fiſhes are caſt away. And a litle after, the
ſame Doctor diſcourſing of eternal glorie both in bodie and ſoule, of
thoſe that dye in Gods fauour, ſayth:
%%% !!! Quotation? \Emph ?
Theſe thinges are promiſed concerning Chriſt, very certaine, very firme,
very plaine, and vndoubted.
\MNote{Hard places explicated by the cleare.}
For albeit ſome thinges are couered in myſteries, yet ſome thinges are ſo
manifeſt, that by them the obſcure thinges may moſt eaſily be cleared.}
my mercie I wil not take away from him: neither wil I hurt in my truth:

\V Neither wil I profane my teſtament: and the words that procede from
my mouth I wil not make fruſtrate.

\V Once I haue ſworne in my holie,
%%% !!! TNote ?
\TNote{I vvil not lie.}
if I lie to Dauid: \V his ſeede shal continewe for euer.

\V And
\SNote{Chriſtian iuſt ſoules as the ſunne,}
his throne as the Sunne in my ſight, and
%%% !!! Not marked in either.
\SNote{and as the perfect, or full moone. See
%%% !!! ???
\XRef{the firſt Tome. page~716.}
%%% !!! Cite ?
S.~Auguſtin alſo expoundeth this verſe in the Anagogical ſenſe, of the
iuſt after the Reſurrection in glorie, where the ſoule ſhal be like the
ſunne, and the bodie, which now is mutable, ſhal be like the moone, not
as now alwayes changing, but as the ful moone, alwayes perfect.}
as the Moone perfect for euer: and a faithful witneſſe in heauen.

\V
\LNote{But thou haſt repelled &c.}{Againe
\MNote{Gods promiſes to Dauid were not fulfilled in Salomon, but in
Chriſt.}
S.~Auguſtin, addeth
\Cite{vpon the next verſes folowing:}
%%% !!! Quotation? \Emph ?
God performed not theſe promiſes in Dauid, that when thou ſeeſt they
were not
fulfilled in Dauid, which neceſſarily muſt be fulfilled, thou maiſt ſeke
an other, in whom it may be ſhewed that they were fulfilled. God
promiſed ſome thing (a kingdom) for euer of Dauids ſeede: and Salomon
was borne: and became of ſo great wiſdom, and ſo great prudence, that
Gods promiſe concerning Dauids ſeede, ſeemed to be fulfilled in him. But
Salomon fell, and gaue place of expecting Chriſt; that becauſe God
neither can be deceiued, nor deceiue, he put not his promiſe in him,
whom he knew would fall, but thou ſhouldeſt relie vpon God, and exact
his promiſe. A litle after:
\MNote{Defectes in the Iewes ſupplied in the Gentiles.}
%%% !!! Quotation? \Emph ?
Thou ſeekeſt the kingdom of the Iewes, it is not: thou ſeekeſt the altar
of the Iewes, it is not: thou ſeekeſt the ſacrifice of the Iewes, it is
not: thou ſeekeſt the prieſthood of the Ievves, it is not. VVherupon he
concludeth:
%%% !!! Quotation? \Emph ?
Al theſe defectes came to the Ievves: yet vvas not Chriſt taken from
them, but differred. Some Ievves beleued in him, and manie Gentiles. As
the Pſalmiſt prophecieth from the
\XRef{47.~verſe}
to the end of this Pſalme.}
But
\SNote{God hauing promiſed al the aforeſaide the prophet in the perſon of
the weake, lamenteth that the contrarie ſhal happen as wel in the
temporal kingdom oppreſſed by the Aſſirians, Babylonians, Perſians,
Grecians, and Romanes: as in the Church impugned by innumerable ſortes
of Heretikes and other Infidels.}
thou haſt repelled and diſpiſed: thou
\SNote{Amongſt manie penſiue thinges, this one word doth comforte vs,
thy promiſe remaineth, thou haſt not denied to ſend Chriſt, but
differred him.}
haſt differred thy Chriſt.

\V Thou haſt ouerthrowne the teſtament of thy ſeruant: thou haſt
profaned his ſanctuarie on the earth.

\V Thou haſt deſtroyed al the hedges therof: thou haſt made the
firmament therof feare.

\V Al that paſſe by the way haue ſpoiled him: he is become a reproch to
his neighbours.

%%% 1301
%%% o-1191
\V Thou haſt exalted the righthand of them that oppreſſe him: thou haſt
made al his enimies ioyful.

\V Thou haſt turned away the helpe of his ſword: and haſt not holpen him
in battel.

\V Thou haſt deſtroied him from
\SNote{From the vſe of Sacrifice and Sacraments, wherby ſinners were
wount to be cleanſed.}
emundation: and his ſeat thou haſt broken downe to the ground.

\V Thou haſt leſſened the daies of his time: thou haſt ouerwhelmed him
with confuſion.

\V
\SNote{The Pſalmiſt prayeth and prophecieth that God wil reſpect the
weaknes of man, maintaine his Church in manie natiõs, and ſaue manie
ſoules.}
How long ô Lord doeſt thou turne away for euer: shal thy wrath burne as
a fire?

\V Remember what my ſubſtance is: for haſt thou made al the children of
men in vaine?

\V Who is the man that shal liue, and shal not ſee death: shal deliuer
his ſoule from the hand of hel?

\V Where are thyne old mercies ô Lord, as thou ſwareſt to Dauid in thy
truth?

\V Be mindeful ô Lord of the aeproch of thy ſeruantes (which I haue held
in my boſome) of manie nations.

\V Which thine enimies haue reproched ô Lord, which they haue reproched
\SNote{As though Chriſt were changed and turned from vs.}
the commutation of thy Chriſt.

\V Bleſſed be our Lord for euer:
\SNote{So we wiſh and pray that al may bleſſe and praiſe thee. Amen.}
Be it, be it.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
