%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 032
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-032


\project douay-rheims


%%% 1201
%%% o-1093
\startChapter[
  title={Psalme 32}
  ]

\PSummary{The
\MNote{The perfectiõ of Gods workes deſcribed.

The 2.~key.}
prophet exhorteth to praiſe God, 4.~deſcribing his powre, prouidence,
mercie, and wiſdom. 16.~No ſaluation but by him, 20.~and therfore
prayeth for his helpe.}

\PTitle{The Pſalme of Dauid.}

\NV Reioyce ye iuſt
\SNote{Not in your ſelues.}
in our Lord:
\SNote{Praiſe is not comelie in the mouth of a ſinner.
\XRef{Eccli.~15.}}
prayſing becometh the righteous.

\V Confeſſe ye to our Lord on
\SNote{In mortification offering your bodies a liuing hoſte.
\XRef{Rom.~12.}}
the harpe: on
\SNote{Of this inſtrument this booke is called the Pſalter and it
ſignifieth the obſeruation of the tenne commandments, without which no
praiſe pleaſeth God.}
a pſalter of ten ſtrings ſing to him.

\V Sing ye to him
\SNote{Praiſing God for the grace of Chriſt in the new Teſtament.}
a new ſong: ſing wel to him in iubilation.

\V Becauſe the
\SNote{Gods rules and precepts are moſt iuſt and right.}
word of our Lord is right, and al his workes are
\SNote{He performeth whatſoeuer he promiſeth.}
in faith.

\V He loueth
\SNote{God euer ioyneth theſe vertues together.}
mercie and iudgement: the earth is ful of the mercie of our Lord.

\V By
\SNote{Gods word is omnipotent.}
the word of our Lord the heauens are eſtablished: and by the ſpirit of
his mouth al the power of them.

%%% o-1094
\V Gathering together the waters of the ſea, as it were in a bottel:
putting the depthes
\SNote{Vnknowen, as riches hid in ſecrete places.}
in treaſures.

\V Let al the earth feare our Lord: and let al the inhabitantes of the
world be moued at him.

\V Becauſe he ſaid, and they were made: he commanded and they were
created.

\V Our Lord
\SNote{As in Babylon he confounded their tongues.}
diſſipateth the counſels of nations: and he reproueth the cogitations of
people, and he reproueth the counſels of princes.

\V But the counſel of our Lord abydeth for euer: the cogitations
\SNote{His abſolute wil is alwayes fulfilled.}
of his hart in generation and generation.

\V Bleſſed is the nation, whoſe God is our Lord: the people
%%% 1202
\SNote{The people of Iſrael in the old Teſtament: and Chriſtians of al
nations in the
new.}
whom he hath choſen for his inheritance.

\V Our
\SNote{Gods perpetual prouidence.}
Lord hath looked from heauen: he hath ſene al the children of men.

\V From his prepared habitation he hath looked vpon al, that inhabite
the earth.

\V Who
\SNote{Gods wiſdom infinite.}
made their hartes ſeuerally: who vnderſtandeth al their workes.

\V The
\SNote{No powre in earth is of anie force without God.}
king is not ſaued by much powre: and the gyant shal not be ſaued in the
multitude of his ſtrength.

\V The horſe fayleth
\SNote{Of his maſter,}
to ſafetie,
\SNote{or of him ſelfe.}
and in the abundance of his force he shal not be ſaued.

\V Behold the eies of our Lord be vpon them that feare him: and on them,
that hope vpon his mercie.

\V That he may deliuer their ſoules from death: and nourish them in
famine.

\V Our ſoule expecteth our Lord: becauſe he is our helper and
protector.

\V Becauſe in him our hart shal reioyce: and we haue truſted in his
holie name.

Let
%%% !!! really should go before 'Let'
\SNote{O God which haſt al perfection, shew thy mercie in protecting, and
ſauing al that truſt in thee.}
thy mercie ô Lord be made vpon vs: as we haue hoped in thee.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
