%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 82
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-82


\project douay-rheims


%%% 1290
%%% o-1180
\startChapter[
  title={Psalme 82}
  ]

\PSummary{The
\MNote{Perſecuters of the Church confounded or conuerted.

The 6.~key.}
Church impugned by al ſortes of enimies, 10.~prayeth God to confound
them, as
\Fix{the}{she}{obvious typo, same in both}
hath donne diuers the like, 17.~wherby ſome shal be conuerted.}

\PTitle{A Canticle of Pſalme to Aſaph.}

\VV Ô God
\SNote{Seing none is like to thee ô God,}
who shal be like to thee?
\SNote{shew thy powre and maieſtie,}
hold not thy peace,
\SNote{be not ſilent.}
neither be thou appeaſed ô God:

\V For behold
\SNote{The cruel perſecuters are moſt inſolent,}
thine enimies haue made a ſound: they that hate thee,
\SNote{and proud.}
haue lifted vp the head.

\V They haue taken malignant counſel vpon thy people: and they haue
deuiſed againſt thy ſaincts.

\V They haue ſaide: Come, and let vs deſtroy them out of
\SNote{That there be no more anie faithful people,}
the nation: and let
\SNote{anie Catholiques leift aliue.}
the name of Iſrael be remembred no more.

\V Becauſe they haue deuiſed with one conſent: they haue together made a
couenant againſt thee.

\V The tabernacles of
\SNote{The progenie of Eſau,}
the Idumeians, and
\SNote{the ſeede of Iſmael,}
the Iſmahelites,
\SNote{deſcending from the elder ſonne of Lot.}
Moab, and
\SNote{The iſſue of Abraham by Agar, who falſly cal them ſelues
Saraſcens, as if they were of Sara.}
the Agarens.

\V
\SNote{People of Gebal a citie of Syria,}
Gebal, and
\SNote{of the other ſonne of Lot.}
Ammon, and
\SNote{Thoſe that firſt oppugned the Iſraelites, after they were parted
from Ægypt.
\XRef{Exo.~17.}}
Amalec:
\SNote{The Philiſtians,}
the foreners, with
\SNote{and Tyrians, al nere neighboures, and ſome of them nere akine to
the Iſraelites, were their great enimies.}
the inhabitantes of Tyre.

%%% o-1181
\V Yea and
\SNote{Others alſo coming further of, ioyned againſt Gods people, in
figure that al heretikes, and other infidels conſpire together againſt
Catholiques.}
Aſſur alſo is come with them: they are made an aide to the children of
Lot.

\V Doe to them as
\SNote{The Pſalmiſt therfore prayeth (and withal prophecieth) that God
wil at laſt deſtroy them, as he did Madian.
\XRef{Num.~31.}
\XRef{Iudic,~6. &~7.}}
to Madian, and
\SNote{Siſara captaine general,}
Siſara: as to
\SNote{for Iabin king of Aſor nere Ciſſon.
\XRef{Iudic.~4. v.~7. &~23.}}
Iabin in the torrent Ciſſon.

\V They perished
\SNote{Within the territorie of Manaſſes
\XRef{(Ioſ.~17.)}
which they inuaded.}
in Endor: they were made as
\SNote{Slaine and not buried.}
the dung of the earth.

%%% 1291
\V Put their princes as
\SNote{Theſe foure princes of the Madianites were ſlaine by Gedeons
forces.
\XRef{Iudic.~7. &~8.}}
Oreb, and Zeb, and Zebee, and Salmana.

Al their princes: \V which haue ſaide: Let vs poſſeſſe the Sanctuarie of
God for an inheritance.

\V My God put them
\SNote{By foure ſimilitudes the prophet deſcribeth the punishment that
shal fal vpon perſecuters.}
as a wheele: and as ſtubble before the face of the winde.

\V Euen as fire that burneth a wood & as a flame that burneth the
mountaines:

\V So shalt thou purſew them in thy tempeſt: and in thy wrath thou shalt
truble them.

\V Fil their faces with ignominie: and
\SNote{God by punishing ſeeketh the conuerſion of ſinners, not their
eternal death.}
they wil ſeeke thy name, ô Lord.

\V Let
\SNote{But ſuch as be ſtil obſtinate, and finally impenitent do perish
for euer.}
them be ashamed, and trubled for euer and euer: and let them be
confounded, and perish.

\V And let them know that
\SNote{God only the creator of heauen and earth is properly called
\Sc{Lord}: VVhoſe eſſential, and incommunicable name is \Sc{VVhich is}.
\XRef{Exo.~3. v.~14.}
&
\XRef{6. v.~3.}}
Lord is thy name, thou onlie the Higheſt in al the earth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
