%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 149
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-149


\project douay-rheims


%%% 1399
%%% o-1290
\startChapter[
  title={Psalme 149}
  ]

\PSummary{The
\MNote{The Church muſt euer praiſe God.

The 6.~key.}
Church is moſt ſingularly bond to praiſe God, 4.~for the grace,
ſanctitie, victorie, glorie, 7.~and iudicial powre, which he geueth to
his Sainctes.}

\PTitle{Alleluia.}

\NV Sing ye to our Lord a new ſong: let his prayſe be in
\SNote{God our Lord whom al creatures are bond to praiſe, only accepteth
thoſe mens praiſes, that liue in his holie Church.}
the Church of ſainctes.

%%% 1400
\V Let Iſrael be ioyful in him, that made him: and let the children of
Sion reioyce in their king.

\V Let them praiſe his name in quire: on tymbrel, and pſalter let them
ſing to him:

\V Becauſe our Lord is wel pleaſed in his people: and he wil exalt the
meeke vnto ſaluation.

\V The ſainctes shal reioyce in glorie: they shal be ioyful in
\SNote{Eternal reſt.}
their beddes.

\V The
\SNote{Hiegheſt praiſes of God ſhal be continually vttered by Sainctes
in glorie: for ſtil as praiſes paſſe by their tongues and mouthes, more
like praiſes ſhal ſuccede from their throte, and hartie affection, ſo
out of the abundance of the hart their mouth ſhal ſtil ſpeake Gods praiſe.}
exaltation of God in their throte: and
\SNote{Glorified Sainctes ſhal alſo haue iudiciarie povvre. Firſt, al in
general ſhal like and approue Gods iuſtice in puniſhing the
vvicked. Secondly the vvicked ſhal be iuſtly condemned in compariſon of
the bleſſed, vvho paſſed through, and ouercame the like, yea and greater
tribulations and tentations then thoſe, by vvhich the damned vvere
ouercome. Thirdly, certaine more excellent Sainctes, namely the
Apoſtles, and al thoſe that forſaking proprietie of temporal goodes,
geue that they haue to the poore, as ſome religious Orders doe; or into
a Communitie as the Apoſtles, and manie primitiue Chriſtians did
\XRef{(Mat.~4. v.~20.}
\XRef{Cap.~19. v.~17.}
\XRef{Act.~4. v.~34.}
\XRef{1.~Cor.~6. v.~3.)}
ſhal ſitte in iudgement ſeates, aſſeſſorie iudges with Chriſt, and iudge
thoſe that render account, of wel or euil ſpending the temporal landes
or goodes, which they poſſeſſed in this vvorld. So teach
\CNote{\Cite{In Pſal.~121. v.~5.}}
S.~Auguſtin,
\CNote{\Cite{in Mat.~19.}}
S.~Beda and
\CNote{\Cite{ho. in Nat. S.~Bened.}}
others.}
two edged ſwordes in their handes.

\V To doe reuenge in the nations: chaſtiſements among the peoples.

\V To binde their kinges in fetters: and their nobles in yron manicles.

\V That they may doe in them the iudgement that is
\SNote{Decreed by God.}
written:
\SNote{This iudicial povvre is part of Sainctes glorie.}
this glorie is to al his ſainctes. Alleluia.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
