%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 122
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-122


\project douay-rheims


%%% 1371
%%% o-1261
\startChapter[
  title={Psalme 122}
  ]

\PSummary{A
\MNote{A prayer in affliction.

The 7.~key.}
feruent and attentiue prayer, to be deliuered from captiuitie, or anie
other affliction.}

\PTitle{A gradual Canticle.}

\NV To thee haue I lifted vp mine eies, which dwelleſt in the heauens.

%%% 1372
\V Behold as the eies
\SNote{Seruantes expecting neceſſaries at their maſters handes, are
commonly very attentiue to receiue that vvhich they hope for: ſo muſt
the faithful praying God, be very attentiue, and not diſtracted in their
prayers.}
of ſeruantes, are on the handes of their maiſters,

As the eies
\SNote{An other example of handmaides, who generally are more diligent
then men.}
of the handmaide on the handes of her miſtreſſe: ſo are our eies to our
Lord God vntil he haue mercie on vs.

\V Haue mercie on vs ô Lord, haue mercie on vs: becauſe we are
\SNote{Though God ſuffer his ſeruantes to ſuſtaine ſome affliction and
reproch, yet he heareth and deliuereth them before they be oppreſſed.}
much replenished with contempt.

\V Becauſe our ſoule is
\SNote{Seing perſecution ſtil increaſe, or continevv long, the faithful
are then ſpecially to conceiue hope of ſpeedie reliefe.}
much replenished: reproch to them that
\Fix{abund,}{abound,}{likely typo, fixed in other}
and comtempt to the prowde.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
