%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 148
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-148


\project douay-rheims


%%% 1399
%%% o-1289
\startChapter[
  title={Psalme 148}
  ]

\PSummary{Al
\MNote{Our Creator to be praiſed by al creatures.

The 2.~key.}
creatures ſpiritual and corporal, are inuited to praiſe God, their
Creator and Conſeruer, 13.~as incomparably excellent.}

\PTitle{Alleluia.}

\NV Prayſe ye our Lord
\SNote{Al ye heauenlie ſpirites praiſe God for the excellencie of your
nature.}
from the heauens: praiſe ye him in the high places.

\V Prayſe ye him al his Angels: prayſe him
\SNote{And for your innumerable multitude.}
al his hoſtes:

\V Prayſe ye him
\SNote{Al creatures, wanting ſenſe or reaſon, ſhew forth the Maieſtie
and excellencie of their Creator.}
ſunne and moone: prayſe him al ye ſtarres, and light.

\V Prayſe him ye heauens of heauens: and the waters that are aboue the
heauens, \V let them praiſe the name of our Lord.

Becauſe he ſayd, and they were made: he commanded, and they were
created.

\V He eſtablished them for euer, and for euer and euer: he put a
precept, and it shal not paſſe.

\V Prayſe our Lord from the earth: ye dragons, and al depthes.

\V Fyre, haile, ſnow, yſe, ſpirit of ſtormes: which doe his worde:

\V Mountaines, and al litle hilles: trees that beare fruite, and al
ceders.

\V  Beaſtes, and al cattel: ſerpentes, and fethered fowles:

\V
\SNote{Againe God is to be praiſed for the diuerſitie of ſtates in men,
wherby the whole communitie, is conſerued & gouerned.}
Kinges of the earth, and al peoples: princes, and al iudges of the
earth.

\V Yongmen, and virgins: old with yong let them prayſe the name of our
Lord: \V becauſe the name of him
\SNote{God only, & no falſe imagined god made & diſpoſed al thinges in
order.}
alone is exalted.

\V The confeſſion of him aboue heauen, and earth: and he hath exalted
\SNote{More eſpecially for that God hath ſo fortified his Church.}
the horne of his people.

An hymne to al his
\SNote{Sanctified children,}
ſainctes: to the children of Iſrael, a people
\SNote{that by grace and free wil, which he geueth them, approch vnto him.}
approching vnto him.
\SNote{Al this conſidered the Pſalmiſt concluding with Alleluia,
inuiteth al to praiſe our Lord.}
Alleluia.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
