%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 70
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-70


\project douay-rheims


%%% 1264
%%% o-1155
\startChapter[
  title={Psalme 70}
  ]

\PSummary{King
\MNote{A prayer for perſeuerance in vertue.

The 7.~key.}
Dauid, or anie other iuſt perſon, prayeth God to continew his grace, and
protection againſt the malice of the enimie. 5.~Recounteth Gods mercie
in deliuering him from falling into tentations, 6.~prayeth for the ſame
in old age, or weakenes of bodie and ſpirite, 20.~and promiſeth perpetual
gratitude and praiſes.}

\PTitle{A Pſalme
\SNote{Though this Pſalme (as alſo diuers others) is intitled to, or for
Dauid, it proueth not that ſome other was the author therof, but the
Seuentie Interpreters inſinuate hereby, that it perteyneth in more
particular ſorte to Dauid, growing old;}
for Dauid,
\SNote{they adde alſo the ſonnes of Ionadab a moſt holie familie,}
of the ſonnes of Ionadab, and
\SNote{who for their ſingular pietie, were ſuffered to remaine in
Ieruſalem, in the firſt captiuitie.
\XRef{Ierem.~35.}}
the former captiues.}

\NV In
\CNote{\XRef{Pſal.~30.}}
thee ô Lord
\SNote{The wordes of Dauid, or anie faithful iuſt perſon.}
I haue hoped, let me not be confounded for euer: \V in
\SNote{God of his iuſtice reuengeth the iniuries done to his ſeruantes.}
thy iuſtice deliuer me, and receiue me.

Incline thine eare to me, and ſaue me.

\V Be vnto me for a God protector, and for a fenſed place: that thou
maiſt ſaue me:

Becauſe thou art my firmament, and my refuge.

\V My God deliuer me out of the hand of the ſinner, and out of the hand
of him that doth aganſt the law, and of the vniuſt.

\V Becauſe thou art
\SNote{Through thy grace I haue patience to expect thy pleaſure, when I
shal be releaued.}
my patience ô Lord: ô Lord my hope from my youth.

%%% 1265
\V Vpon thee haue I bene confirmed from the wombe: from my mothers
bellie thou art my protector.

In thee is my ſinging alwaies: \V I was made to manie as
\SNote{It ſemeth ſtrange to moſt men, why the iuſt haue more trubles
then the wicked.}
a wonder: and thou art a ſtrong helper.

\V Let my mouth be filled with praiſe, that I may ſing thy glorie:
\SNote{A neceſſarie prayer for perſeuerance, which none can haue without
ſpecial and continual helpe of grace.}
al the day thy greatnes.

\V Reiect me not in the time of olde age:
\SNote{Againſt this ſpecial danger, and laſt aſſaultes of the enimie,
Chriſt hath inſtituted the Sacrament of Extreme Vnction.}
when my ſtrength shal faile, forſake me not.

\V Becauſe mine enimies haue ſaid to me: and they that watched my ſoule,
conſulted together,

%%% o-1156
\V Saying: God hath forſaken him, purſew, and take him: becauſe there is
none to deliuer.

\V O God be not farre from me: my God haue reſpect to mine ayde.

\V Let them be confounded, and fayle that detract from my ſoule: let
them be couered with confuſion, and shame, that ſeeke euils to me.

\V But I wil alwaies hope: and
\SNote{The beſt meanes to conſerue vertue, is to exerciſe the ſame,
adding good workes toward our neighboures, and praiſes to God.}
wil adde vpon al thy praiſe.

\V My mouth shal shew forth thy iuſtice: al the day thy
ſaluation. Becauſe
\SNote{I profeſſe not exact knowlege of natural thinges;}
I haue not knowne lerning,
 \SNote{but I conſider and contemplate the infinite powre of God, which
 appeareth in his workes,}
 I wil enter into the
 \SNote{and his iuſt iudgements, as partly haue appeared in puniſhing,
 and rewarding ſome in this life, but eſpecially wil be manifeſt in the
 end of this world.}
 powres of our Lord: \V ô Lord I wil be mindful of thy iuſtice onely.

\V O God thou haſt taught me from my youth: and vntil now I wil
 pronounce thy meruelous workes.

\V And vnto ancient age, and olde age: ô God forſake me not. Vntil I shew
forth thy arme to al the generation, that is to come.

Thy might, \V and thy iuſtice ô God euen to the higheſt, great meruailes
which thou haſt done: ô God who may be like to thee?

%%% 1266
\V How great tribulations haſt thou shewed me, manie and euil: and
turning thou
\SNote{After tribulations recreated me.}
haſt quickened me: and from the depthes of the earth thou haſt brought
me backe againe.

\V Thou haſt multiplied thy magnificence: and being turned thou haſt
comforted me.

\V For I alſo wil confeſſe to thee in the inſtruments of Pſalme thy
truth: ô God I wil ſing to thee on the harpe, holie one of Iſrael.

\V My
\SNote{I wil praiſe thee with mouth and hart.}
lippes shal reioyce when I shal ſing to thee; and my ſoule, which thou
haſt redemed.

\V Yea and my tongue al the day shal meditate thy iuſtice: when they
shal be confounded and ashamed that ſeeke euils to me.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
