%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 066
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-066


\project douay-rheims


%%% 1256
%%% o-1147
\startChapter[
  title={Psalme 66}
  ]

\PSummary{The
\MNote{Propagation of the Church.

The 6.~key.}
prophet prayeth for (and withal foresheweth) the propagation of the
Church of Chriſt.}

\PTitle{Vnto the end, in hymes, a Pſalme
\SNote{This Pſalme beginning to be ſongue by voices, inſtruments were
adioyned.}
of Canticle to Dauid.}

\VV God
\SNote{God firſt remitte our ſinnes:}
haue mercie vpon vs, and
\SNote{then geue vs thy manifold graces,}
bleſſe vs:
\SNote{grant faith and repentance,}
illuminate his countenance vpon vs, and
\SNote{and ſo forgeuenes of ſinnes.}
haue mercie on vs.

\V That we may know thy way vpon earth: in al nations thy ſaluation.

\V Let peoples ô God, confeſſe to thee:
\SNote{Al nations shal be conuerted.}
let al peoples confeſſe to thee.

\V Let nations be glad & reioice: becauſe thou iudgeſt peoples in
equitie, and the nations in earth thou doeſt direct.

\V Let peoples ô God confeſſe to thee, let al peoples confeſſe to
thee: \V the earth hath yelded her fruite.

God,
%%% !!! The first SNote should be before 'God'
\SNote{God the
\Fix{Fater,}{Father,}{obvious typo, fixed in other}}
\SNote{God the Sonne,}
our God bleſſe vs, \V
\SNote{God the Holie Ghoſt, ſaue the peoples of al nations, by
Euangelical preaching of thee, the moſt Bleſſed Trinitie.}
God bleſſe vs: and let al the endes of the earth feare him.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
