%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 047
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-047


\project douay-rheims


%%% 1228
%%% o-1119
\startChapter[
  title={Psalme 47}
  ]

\PSummary{God
\MNote{The Church founded and protected by God.

The 6.~key.}
moſt, and euerie where laudable, is eſpecially praiſed in the Church of
Chriſt (prefigured by Sion, and there begunne.) 9.~Al thinges being
fulfilled in the Church, euen as they were prophecied, and promiſed,
12.~the faithful are exhorted to conſider and congratulate the ſame.}

\PTitle{A Pſalme
\SNote{Voices beginning the muſike inſtruments proſecuted.}
of Canticle to the ſonnes of Core, the
\SNote{Eſpecially for the ſecond day of the weke, the day after the
ſabbath, which is our Sunday, called Dominica, our Lords day.}
ſecond of the Sabbath.}

\VV Great is our Lord, and to be prayſed excedingly in
\SNote{Ieruſalem, and mount Sion were moſt obliged to praiſe God, for
greateſt benefites receiued, ſo the Catholique Church therby prefigured,
and hauing receiued farre greater, is moſt of al bonden to be gratful.}
the citie of our God, in his holie mount.

%%% o-1120
\V Mount Sion is founded with the exultation of
\SNote{This can not be affirmed of Sion, or Ieruſalem, but is only
verified of the Catholique Chriſtian Church:}
the whole earth,
\SNote{whoſe coaſtes do extend to the North, and to al quarters of the
round earth.}
the ſides of the North, the citie of the great king.

\V God shal be knowen in
\SNote{The ſame one God, one Chriſt, one Faith, and one Religion in al
particular Churches of the vvhole militant Church.}
the houſes therof,
\SNote{And this Vniuerſalitie, and Vnitie ſhal be, after that Chriſt
taking mans nature ſhal be aſcended, and ſhal ſend the Holie Ghoſt, to
found & beginne this Church.}
when he shal receiue it.

\V For behold the kings of the earth
\SNote{For the aſſured certaintie of that is foreſhewed, the Prophet
ſpeaketh in the pretertence, as if it vvere already done in his time,
which he then ſavv in ſpirite.}
were gathered together: they aſſembled in one.

%%% 1229
\V They ſeing it ſo, were in admiration, were trubled, were moued: \V
trembling tooke them.

Their ſorowes
\SNote{Nothing more moueth the hart, & affecteth al the bodie and ſoule,
then ſpiritual cogitations of faith and religion, and therfore it is
compared to a woman traueling with child, who hath moſt careful and
greuous paines.}
as a woman traueling. \V In a vehement ſpirit
\SNote{In vvhich great
\Fix{confflict}{conflict}{possible typo, same in both}
of mans ſpirite, God by his grace geueth force, to breake through the
contrarie aſſaultes of our enimie, to remoue al impediments, and to
ouercome the difficulties.}
thou shalt breake the shippes of Tharſis.

\V
\SNote{This conſideration that al is now done, that was of old
prophecied, is a meruelous confirmation and conſolation to Chriſtians.}
As we haue heard, ſo haue we ſeene in the citie of the Lord of hoſtes,
in the citie of our God: God hath founded it for euer.

\V We haue receiued thy mercie, ô God,
\SNote{Grace and mercie is only granted to thoſe that are vvithin, or
come vnto the Catholique Church.}
in the middes of thy temple.

\V According to thy name ô God, ſo alſo is thy prayſe vnto the endes of
the earth: thy right hand is ful of
\SNote{As God is praiſed for his mercie, ſo alſo for his iuſtice; which
do neuer preiudice the one the other.}
iuſtice.

\V Let mount Sion be glad, and the daughters of Iuda reioyce, becauſe of
thy iudgementes ô Lord.

\V Compaſſe Sion, and embrace ye her:
\SNote{Conſider the fortreſſes of the Church, which are the holie
Fathers, and Doctors, that watch and defend her vvalles.}
tel ye in her towers.

\V Set your hartes
\SNote{So reſt you aſſured for al matters of faith in this pillar of
truth.}
on her ſtrength: and
\SNote{Obſerue and marke diligently how manie particular Churches were
ſpedely founded in the world,}
diſtribute ye her houſes, that you
\SNote{and declare this to other generations, that they may alſo hold
faſt the ſame faith, or returne vnto it, if they be relapſed; or at laſt
embrace it, if ſowner they haue not.}
may declare it in an other generation.

\V Becauſe
\SNote{Chriſt God incarnate that vvorketh al this, is our very God and
Sauiour, not for a few yeares, an hundred, ſix hundred, or a thouſand,
but for euer and euer.}
this is God, our God for euer, and for euer and euer: he
\SNote{He ſhal rule as a king, and conſequently haue a kingdom his
militant Church euermore, to the very end of this vvorld. As he ſhal
likevviſe haue his triumphant Church in eternitie.}
shal rule vs euermore.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
