%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 97
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-97


\project douay-rheims


%%% 1314
%%% o-1205
\startChapter[
  title={Psalme 97}
  ]

\PSummary{Al
\MNote{The Church in al nations.

The 6.~key.}
men are againe inuited ioyfully to celebrate the meruelous conqueſt of
Chriſt in al nations, 4.~with hart, voice, and inſtruments: 8.~al
creatures acknowledging his coming to iudge the world.}

\PTitle{A pſalme
\SNote{Prefiguring Chriſt, who hath made his ſaluation knowen in al
nations.}
to Dauid himſelfe.}

%%% 1315
\NV Sing ye to our Lord
\SNote{A new benefite of grace, making men new in ſpirite, requireth a
new ſongue of gratitude.}
a new ſong: becauſe he hath done meruelous thinges.

His righthand hath wrought ſaluation
\SNote{Raiſed vp himſelfe from death.}
to himſelfe: and his arme is holie.

\V Our Lord hath
\SNote{Made his grace effectually knowen by raiſing men from ſinne, and
deliuering them from the powre of the diuel.}
made knowne his ſaluation: in the ſight of the Gentiles he hath reueled
his iuſtice.

\V He hath remembred his mercie, and his truth to
\SNote{Some of the Iewes conuerted to Chriſtianitie.
\XRef{Rom.~11.}}
the houſe of Iſrael.

Al the ends of the earth haue ſeene the ſaluation of our God.

\V Make ye iubilation to God al the earth:
\SNote{In voice.}
\TNote{\L{cantate}}
chaunt, and
\SNote{In hart.}
\TNote{\L{exsultate}}
reioyce, and
\SNote{In inſtruments.}
\TNote{\L{psallite}}
ſing.

\V Sing to our Lord on harpe, on harpe and voice of pſalme: \V on long
drawen trumpets, and voice of cornet of horne.

Make iubilation in the ſight of the king our Lord: \V let the ſea be
%%% o-1206
moued, and the fulnes therof: the round world, and they that dwel
therin:

\V The riuers shal clappe with hand, the mountaynes together shal
reioyce \V at the ſight of our Lord:
\SNote{Chriſt directeth and diſpoſeth al thinges rightly in this world.}
becauſe he cometh to iudge the earth.

He
%%% !!! SNote should be before 'He'
\SNote{And wil accordingly geue iuſt ſentence in the end.}
wil iudge the round earth in iuſtice, and the peoples in equitie.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
