%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 139
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-139


\project douay-rheims


%%% 1388
%%% o-1278
\startChapter[
  title={Psalme 139}
  ]

\PSummary{The
\MNote{Eternal paine of the wicked, and ioy of the bleſſed.

The 10.~key.}
iuſt diuerſly afflicted by the wicked, pray to be defended, 7.~repoſe their
confidence in God, 10.~who wil adiudge the reprobate to eternal
punishment, 13.~and reward the good with the fruition of himſelf.}

\PTitle{Vnto the end, a Pſalme of Dauid.}

\NV Deliuer me ô Lord from the euil man: from the vniuſt man reſcue me.

\V Which
\SNote{Vnquiet euil diſpoſed men ſtil deuiſe wicked plottes}
haue deuiſed iniquitie in their hart:
\SNote{and neuer ceaſe to make diſcord, and debates.}
al the day they did appoint battels.

They haue whet their tongues as that of a ſerpent: \V the venome of
aſpes is vnder their lippes.

\V Kepe me ô Lord from the hand of the ſinner: and from vniuſt men
deliuer me.

\V Who haue deuiſed to ſupplant my ſteppes:
%%% !!! Verse marked in both, but not counted.
the proude haue hid a ſnare for me:

And they haue ſtreched out ropes for a ſnare: they haue layd
\TNote{\L{scandalum}}
a ſtumbling blocke for me nere the way.

\V
\SNote{In time of tentation prayer is moſt neceſſarie.}
I ſayd to our Lord: Thou art my God: heare ô Lord the voice of my
petition.

\V O Lord, Lord
\SNote{Man is not able to reſiſt tentations,}
the ſtrength of my ſaluation:
\SNote{vnleſſe God, by his grace, as by a helmet, defend him from
yelding conſent.}
thou haſt ouershadowed my head in the day of battel.

\V Yeld me not ô Lord from
\SNote{Suffer me not to fal from that which I now deſire (which is to be
conſtant in vertue) not to conſent to ſinners perſwaſions.}
my deſire, to the ſinner: they haue deuiſed againſt me, forſake me not,
\SNote{VVho then would triumph ouer me.}
leſt they perhaps be proude.

%%% o-1279
\V
\SNote{The ſumme of their miſcheuous deuiſes,}
The head of their compaſe:
\SNote{conſiſting in their ſuttle perſuaſions with ſwete wordes,}
the labour of their lippes
\SNote{ſhal ouerwhelme, and bring themſelues to perdition.}
shal couer them.

\V
\SNote{Hel fire ſhal be their lote,}
Coales shal fal vpon them,
\SNote{into which they ſhal fal,}
thou shalt caſt them downe into fyre: in miſeries
\SNote{& euer be in intolerable torments.}
they shal not ſtand vp.

\V A man ful of tongue shal not be directed in the earth: euils shal
take the vniuſt man into deſtruction.

\V
\SNote{I haue lerned by good inſtructions, and by experience, that in
the end thou ô God wilt comfort the iuſt, who are now afflicted, and
puniſh the vniuſt, that liue in temporal delightes, as it happened to
Lazarus, and the glutton.}
I haue knowne that our Lord wil do the iudgement of the needie: and the
reuenge of the poore.

%%% 1389
\V But as for the iuſt, they shal confeſſe to thy name: and the
righteous shal dwel with
\SNote{Eternal glorie conſiſteth in ſeeing God.}
thy countenance.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
