%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-10


\project douay-rheims


%%% 1165
%%% o-1057
\startChapter[
  title={Psalme 10}
  ]

\PSummary{Dauids
\MNote{Gods prouident care of the iuſt. The 3.~key.}
freindes aduiſing him to flee from the perſecution of Saul, he
anſwereth, that his truſt is in Gods protection. 2.~Though the
perſecutor be very malitious, 4.~yet God wil ouer throw him, 5.~and
deliuer the iuſt.}

\PTitle{Vnto
\SNote{S.~Auguſtin applieth it to heretikes, perſwading Catholiques to
repare vnto their ſeparate congregation, falſly calling it the
montayne.}
the end the Pſalme of Dauid.}

\VV I truſt in our Lord: how ſay
\SNote{Ye my freindes ſay thus.}
ye to my ſoule: Paſſe ouer vnto the mountayne as a ſparrow?

\V For behold ſinners
\SNote{Perſecuters vſing al rigor and force,}
haue bent the bow, they haue prepared their arrowes in the quiuer, that
they may shoote in
\SNote{falſly pretẽd other cauſes againſt the innocent to deſtroy them.}
the darke, at them that be right of hart.

\V For they haue
\SNote{It is the maner of perſecuters, and eſpecially of heretikes, to
deſtroy and pul downe that others haue built.}
deſtroyed the thinges, which thou didſt perfite: but the iuſt what hath
he done?

\V
\CNote{\XRef{Abac.~2.}}
Our Lord is in his holie temple, our Lord his ſeate is in heauen.

\V His eies haue reſpect vnto the poore: his
\SNote{Though God ſeme to winke or ſleepe, yet his prouidence ſtil
watcheth, and obſerueth al mens actions.}
eieliddes examine the ſonnes of men.

\V Our Lord
\SNote{Proueth by tribulations.}
examineth the iuſt, and the impious: but he that
\SNote{Continuãce in ſinne bringeth damnation to the ſoule.}
loueth iniquity, hateth his owne ſoule.

%%% o-1058
\V
\SNote{God ſparing for the time at laſt muſt needes of iuſtice punish
ſeuerely.}
He shal rayne ſnares vpon ſinners: fyre and brimſtone, and blaſt of
ſtormes the portion of their cuppe.

%%% 1166
\V Becauſe our Lord is iuſt and hath loued iuſtice: his countenance hath
ſeene equitie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
