%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 102
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-102


\project douay-rheims


%%% 1321
%%% o-1211
\startChapter[
  title={Psalme 102}
  ]

\PSummary{Thankes
\MNote{Gratitude for Gods benefits.

The 7.~key.}
to God for priuate, 6.~and publike benefites. 17.~His mercie, iuſtice,
and other proprieties are immutable. 20.~Angels, and al other creatures
are inuited to praiſe him.}

\PTitle{To
\SNote{Inſpired to Dauid, and written by him.}
Dauid himſelf.}

\NV My ſoule
\SNote{Shew forth praiſe, and thankes:}
bleſſe thou our Lord: and
\SNote{al my cogitations, affections, ſenſes, and powres.}
al thinges, that are within me, his holie name.

\V My ſoule bleſſe thou our Lord: and forget not al his retributions.

\V Who is
\SNote{The firſt benefite of grace is remiſſion of ſinnes:}
propitious to al thine iniquities: who
\SNote{the ſecond, is curing euil habites, or diſpoſitions.}
healeth al thine infirmities.

\V Who
\SNote{The third, to conſerue from falling againe:}
redemeth thy life from deadly falling: who
\SNote{the fourth, to geue victorie and reward in abundant meaſure.}
crowneth thee in mercie and commiſerations.

\V Who
\SNote{The fifth, to grant al lawful petitions temporal and ſpiritual,
which are good for the ſoule,}
replenisheth thy deſire in good thinges:
\LNote{Thy youth shal be renevved, as the Eagles.}{Ariſtotel
\MNote{An Eagle freſh in old age as in youth.}
and Plinie write, that an Eagle decayeth not, nor euer dieth by old age,
but the vpper part of her beake ſtil growing, at laſt hindereth her from
eating, and ſo ſhe dieth of famine. Saadias, and other Hebrew Rabbins,
reporte that an Eagle euerie tenne yeares waſheth herſelfe in the ſea as
in a bath, & then flying very hiegh burneth her fethers in the elemental
fire, & new fethers growing ſhe becometh freſh, as in her firſt youth,
til at laſt about an hundred yeares old, ſhe is not able to riſe from
the water and ſo is drowned. S.~Auguſtin more probably
%%% !!! Cite?
affirmeth that in long time her beake growing long, and ſtopping her
mouth, that ſhe can not eate, ſhe breaketh the vpper hooked part therof
againſt a ſtone, and ſo receiueth meate, and recouereth ſtrength, as in
her youth. But whatſoeuer is the natural propertie of this kingly birde,
the Royal Prophet here inſtructeth vs, by the ſimilitude of her long
life, or by the renouation of her ſtreingth, that
\MNote{Renouation of a ſinner by grace.}
iuſt men, Gods
ſeruantes are ſpiritually renouated in Chriſt, the principal rocke, on
whom the Church, & al the faithful are built, either by receiuing new
ſtreingth by his grace in their ſoules, after they are weakened by
ſinne; as S.~Ierom and Euthymius
%%% !!! Really two cites.
\Cite{expound this place:}
or by reſtauration of their bodies glorified in the reſurrection; as
%%% !!! Cite (upon this place)?
S.~Auguſtin teacheth: or by both, as moſt Catholique Doctors vnderſtand
it.
\MNote{Diuerſe ſenſes of the ſame Scripture.}
For one ſenſe of holie Scripture excludeth not an other. Eſpecially when
one is ſubordinate to the other. As here theſe two ſenſes do very wel
concurre, ſeing the ſtate of the bodie after reſurrection, dependeth
vpon the ſtate of the ſoule, at the time of death.}
thy youth
\SNote{the ſixth, reſurrection of flesh in glorie.}
shal be
\Fix{rewed}{renewed}{obvious typo, fixed in other}
as the eagles.

%%% o-1212
\V Our Lord
\SNote{Our Lord, whoſe ſpecial propertie is to ſhew mercie,}
doth mercies: and
\SNote{when the ſame is neglected, he reuengeth the wrongs, deliuering
the oppreſſed, and puniſhing the oppreſſors.}
iudgement to al that ſuffer wrong.

\V He made his waies
\SNote{By geuing them a written law.}
knowne to Moyſes, his willes to the children of Iſrael.

%%% 1322
\V Our Lord is
\SNote{God is naturally pittiful, to releeue the afflicted,}
pitieful, and
\SNote{merciful towards ſinners,}
merciful:
\SNote{loath to be angrie, or to puniſh.}
long ſuffering, and very merciful.

\V
%%% !!! LNote not marked in either
\LNote{He vvil not be angrie alvvayes.}{Origen
\MNote{Origens hereſie that al ſhal be ſaued.}
miſunderſtood this place, and ſome other like, holding an erronius
opinion, that al, euen the moſt wicked ſinners, both men and diuels,
ſhal at laſt be ſaued, and not eternally damned: which is a condemned
hereſie, contrarie to euident places of holie Scriptures
\XRef{Pſal.~9. v.~6.}
The impious hath periſhed: their name thou haſt deſtroyed for euer, &
foreuer & euer.
\XRef{Mat.~25.}
The wicked ſhal goe into fire euerlaſting, into euerlaſting puniſhment.
\XRef{Apoc.~20.}
The beaſt and the falſe prophet (and the ſame reaſon is for al the
vvicked) ſhal be tormented day and night, for euer and euer. Neither are
theſe vvordes (God vvil not be angrie alvvayes) ſpoken vniuerſally,
touching al ſinners vvhoſoeuer, but are limited,
\XRef{v.~13.}
&
\XRef{17.}
to thoſe that feare him, and kepe his teſtament:
\MNote{The damned can neuer repent.}
vvheras al thoſe that
dye in mortal ſinne, are ſtil obſtinate in malice, and can neuer rightly
repent, nor rightly feare God, nor kepe his commandments.}
He wil
\SNote{God puniſheth not penitents with eternal paine, but with
temporal: for though (as in the next verſe) our ſinnes of their owne
nature, deſerue eternal puniſhment, yet Chriſt paying our ranſome, true
penitents are only puniſhed temporally.}
not be angrie alwayes: neither wil he threaten for euer.

\V He hath not done to vs according to our ſinnes: neither according to
our iniquities hath he rewarded vs.

\V For according to the height of heauen from the earth: hath he
ſtrengthned his mercie vpon them that feare him.

\V As far
\SNote{See here the perfect washing away, and ſeparation of ſinnes.}
as the Eaſt is diſtant from the Weſt: hath he made our iniquities far
from vs.

\V As a father hath compaſſion of his children, ſo hath our Lord
compaſſion on them that feare him: \V becauſe he hath knowen
\SNote{In regard of our frailtie he hath compaſſion: yet his mercie only
auaileth to them that feare him: as in the next wordes before, and
v.~17.}
our making.

He remembred that we are duſt: \V man, his daies are as graſſe, as the
floure of the filde ſo shal he florish.

\V Becauſe the ſpirit shal paſſe in him, and he shal not ſtand: and he
shal know his place no more.

\V But the mercie of our Lord from euerlaſting, and vnto euerlaſting
vpon them that feare him.

And his iuſtice is vpon the childrens children, to them that keepe his
teſtament.

\V And are mindful of his commandmentes, to
\SNote{A iuſt man not only knoweth, and remembreth, but alſo doth the
commandments.}
doe them.

\V Our Lord hath prepared his ſeate in heauen: and his
\Fix{kindom}{Kingdom}{obvious typo, fixed in other}
shal haue dominion ouer al.

\V Bleſſe our Lord al ye his Angels: mightie in powre, doing his word,
that feare the voice of his wordes.

\V Bleſſe our Lord al ye his hoaſtes: you his miniſters, that doe his
wil.

\V Bleſſe ye our Lord
\SNote{Al creatures, though naturally ſenſeles, yet praiſe God, becauſe
they are his worke,}
al his workes: in euerie place
\SNote{and becauſe they are in his dominion.}
of his dominion, my ſoule bleſſe thou our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
