%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 108
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-108


\project douay-rheims


%%% 1336
%%% o-1226
\startChapter[
  title={Psalme 108}
  ]

\PSummary{Chriſt
\MNote{Chriſt perſecuted & his enimies puniſhed.

The 5.~key.}
(by the mouth of Dauid) requeſteth of God to be iuſtly declared
innocent, and his enimies punished, 6.~particularly deſcribing Iudas the
traitors malice, 21.~and his owne temporal afflictions, 26.~prayeth,
30.~and praiſeth God for his deliuerie.}

\PTitle{Vnto the end, a Pſalme of Dauid.}

\NV O
\SNote{The wordes of Chriſt.}
God conceale not my prayſe: becauſe the mouth of the ſinner, and the
mouth of
\SNote{The Phariſees and Herodians
\XRef{(Mat.~22.)}
with their mouth acknowleged Chriſt a true ſpeaker, and a teacher of the
way of God in truth, therby to draw him into danger, and to ſheede his
bloud.}
the deceitful man is open vpon me.

\V They
\SNote{At other times they accuſed him of great crimes, laſtly of
treaſon againſt Cæſar.}
haue ſpoken againſt me with deceitful tongue, and with wordes of hatred
they haue compaſſed me: and they haue impugned me without cauſe.

\V For that they should loue me, they backbited me: but I prayed.

\V And they ſet againſt me euil thinges for good: and hatred for my
loue.

\V Appoint
\SNote{A prediction that Iudas would not make recourſe to anie good
counſeller, but complaine of his miſerable tormented conſcience to the
wicked, who gaue him no comfort at al,}
a ſinner ouer him: and
\SNote{and ſo deſparing, the diuel perſwaded him to hang himſelfe.}
let the diuel ſtand on his righthand.

%%% 1337
\V When he is iudged, let him come forth condemned: and let his prayer
be turned into ſinne.

\V Let his dayes be made fewe: and
\CNote{\XRef{Act.~2. v.~16.}}
let an other take his
\SNote{The office of Apoſtleſhippe.}
bishopricke.

\V Let
\SNote{The poſteritie, or ſucceſſors of wicked perſecuters proſper not
long in this world.}
his children be made orphans: and his wife a widow.

\V Let his children be tranſported wandering, and let them begge: and
let them be caſt out of their habitations.

\V Let the vſurer ſearch al his ſubſtance: and let ſtrangers ſpoile his
labours.

\V Let there be none to helpe him: neither let there be anie to haue
pittie on his pupilles.

%%% o-1227
\V Let his children come to deſtruction: in
\SNote{Arch heritikes that deuiſe newe opinions, are ſhortly forſaken,
their folowers ſtil coining new hereſies of their owne, differing from
their falſe maſters.}
one generation let his name be cleane put out.

\V Let the iniquitie of his fathers returne to memorie in the ſight of
our Lord: and let not the ſinne of his mother be blotted out.

\V Let them be before our Lord alwayes, and let the memorie of them
perish out of the earth: \V For that he remembred not to doe mercie.

\V And he perſecuted the poore, and needie man, and the compunct in hart
to kil him.

\V And he
\SNote{Let them obſerue this, that vſe more ſwearing and blaſpheming,
then praying or meditating.}
loued curſing, and it shal come to him: and he would not bleſſing, and
it shal be far from him.

And he put on curſing as a garment; and it entred as water into his
inner partes, and as oile in his bones.

\V Be it to him as a garment, wherwith he is couered: and as a girdle,
wherwith he is alwayes girded.

\V This is the worke of them, that detract from me before our Lord: and
that ſpeake euils againſt my ſoule.

\V And thou, Lord, Lord, doe with me for thy names ſake: becauſe thy
mercie is ſwete.

Deliuer me \V becauſe I am needie, and poore: and
\SNote{Chriſts ſoule was penſiue when he prayed in the garden, and he
did workes of penance for our ſinnes al his life.}
my hart is trubled within me.

\V As a shadow when it declineth, am I taken away: and I am shaken as
locuſtes.

\V My knees are weakened with faſting: and my flesh is changed by reaſon
of oile.

\V And I am made a reproch to them: they ſaw me, and wagged their
heades.

%%% 1338
\V Helpe me ô Lord my God:
\SNote{He prayed alſo for his reſurrection, and glorification.}
ſaue me according to thy mercie.

\V And let them know that this is thy hand: and thou ô Lord haſt done
it. 

\V They wil curſe, and thou shalt bleſſe: let them that riſe vp againſt
me, be confounded: but thy ſeruant shal reioyce.

\V Let them that detract from me, be clothed with shame: and let them be
couered with their confuſion, as with a
\SNote{VVith al poſſible confuſion.}
duble patched cloke.

\V I wil confeſſe to our Lord excedingly with my mouth: and in the
middes of manie I wil prayſe him:

\V Becauſe he hath ſtood on the righthand of the poore, that he might
ſaue my ſoule from the perſecutors.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
