%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 79
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-79


\project douay-rheims


%%% 1285
%%% o-1176
\startChapter[
  title={Psalme 79}
  ]

\PSummary{The
\MNote{Chriſt our Redemer from ſinne and captiuitie.

The 5.~key.}
Prophet prayeth for the releaſe of Iſrael in great tribulation,
15.~forsheweth the coming of Chriſt to redeme man from ſinne, and from
thraldome of the diuel.}

\PTitle{Vnto
\SNote{Perteyning to the new Teſtament,}
the end, for
\SNote{eſpecially to Gentiles, that shal be conuerted to Chriſt.}
them, that shal be changed, teſtimonie to
\SNote{For perpetual memorie to the congregation of faithful.}
Aſaph, a Pſalme.}

%%% 1286
\VV Thou that ruleſt Iſrael, attend: that
\SNote{By Ioſeph the prophet vnderſtandeth al Iſrael: becauſe the firſt
birthright being taken from Ruben was geuen to him, to witte duble
portion of enheritance, two tribes of twelue.}
conducteſt Ioſeph as a sheepe.

Which ſitteſt vpon the cherubs,
\SNote{Appeare, and ſhew thy mightie hand before thy people:}
be manifeſt \V before
\SNote{wherof Ephraim, Beniamin, and Manaſſes folowing the Arke, when
they marched or camped, ſhould moſt directly ſee what appeared therin,
the other nine tribes being placed before it, and on both ſides.
\XRef{Num.~2.}}
Ephraim, Beniamin, and Manaſses.

Raiſe vp
\SNote{Thou which alwayes can helpe vs, now vſe thy powre in deliuering
vs from this temporal miſerie.}
thy might, and come, to ſaue vs.

\V Ô God
\SNote{Set vs in a better ſtate.}
conuert vs: and shew
\SNote{Shew thy benigne countenance and fauoure, Myſtically, ſend the
promiſed Meſſias, Chriſt, the eſſential Image of God.
\XRef{2.~Cor.~4. v.~4.}
\XRef{Colloſ.~1. v.~15.}}
thy face, & we shal be ſaued.

\V Ô Lord the God of hoſtes, how long wilt thou be angrie
\SNote{How long wilt thou differre to heare our prayer?}
vpon the prayer of thy ſeruant?

\V Thou wilt feede vs with the bread of teares: and giue vs drinke with
teares
\SNote{Thou doſt iuſtly punish vs, but thou tempereſt the ſame with
meaſure, that it exceede not to our ruine.}
in meſure.

\V Thou haſt made vs to be a contradiction to our neighboures and our
enimies haue ſcorned vs.

\V Ô God of hoſts
\SNote{God firſt preuenting vvith his grace,}
conuert vs: and shew thy face, and we
\SNote{man may cooperate therwith to his iuſtification, and ſaluation.}
shal be ſaued.

\V Thou didſt tranſport
\SNote{Thy Church and people.}
the vineyeard out of Ægypt: thou didſt caſt out
\SNote{The Chananites.}
the Gentiles, and planted it.

\V Thou waſt the guide of the way
\SNote{The cloude, and piller of fire were viſible guides.}
in the ſight therof: thou didſt
%%% o-1177
plant the rootes therof, and it
\SNote{And it multiplied mightely.}
filled the earth.

\V
\SNote{Hyperbolical ſpeach to ſignifie the great multiplication of the
Iſraelites.}
The shadow of it couered the mountanes: and the bowghes of it the ceders
of God.

\V It extended her branches euen to
\SNote{From the mediterranean ſea of Paleſtin,}
the ſea; and her boughes vnto
\SNote{to the riuer Euphrates.
\XRef{Exo.~23. v.~31.}
\XRef{Deut.~11. v.~24.}}
the riuer.

\V Why haſt thou deſtroyed the hedge therof: and al that paſſe by the
way doe plucke it.

%%% 1287
\V The
\SNote{Cruel enimies more like to moſt cruel beaſtes then to men.}
boare of the wood hath deſtroyed it: and
\SNote{Yea ſo cruel, as none els in the world are like vnto them.}
the ſingular wilde beeſt hath eaten it.

\V Ô God of hoaſts returne: regard
\SNote{The Prophet now prayeth for the coming of Chriſt, which he ſaw in
ſpirite.}
from heauen, and ſee, and viſite this vineyeard.

\V And
\SNote{The Church of the old teſtament, in her beſt ſtate, wanted the
perfection, which the Church of Chriſt hath.}
perfite it, which thy right hand hath planted: and vpon
\SNote{Chriſt our Meſſias moſt commonly
\Fix{calleth}{called}{obvious typo, fixed in other}
by this title: The Sonne of man.}
the ſonne of man whom thou haſt confirmed to thee.

\V
\SNote{So thy vinyard can not indure if it be ſtil afflicted, and
trodden downe.}
Thinges burnt with fyre, and digged doune at the increpation of thy
countenance shal perish.

\V Let
\SNote{Chriſt working by Gods powre redreſſeth al miſeries.}
thy hand be vpon the man of thy right hand: and vpon the ſonne of man,
whom thou haſt confirmed to thee.

\V And we depart not from thee, thou wilt quicken vs: and we wil
inuocate thy name.

\V
\SNote{The ſame is the 8.~and the 15.~verſe (with litle alteration) and
here repeted the third time. In which we alſo pray for three thinges:
firſt to be purged, and conuerted from ſinne: ſecond to be illuminated
by Chriſt, the Image of God: thirdly to be ſanctified, and ſaued in
eternal glorie to haue the fruition of the moſt bleſſed Trinitie.}
Ô Lord the God of hoaſtes conuert vs: and shew thy face, and we shal be
ſaued.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
