%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 71
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-71


\project douay-rheims


%%% 1266
%%% o-1157
\startChapter[
  title={Psalme 71}
  ]

\PSummary{By
\MNote{Chriſt our King & Iudge.

The 5.~key.}
way of prayer the Pſalmiſt prophecieth Chriſts coming, our King and
Iudge: 4.~the deliuerer of mankind from the thraldom of the
diuel. 8.~The greatnes of his ſpiritual kingdom in the Gentiles, 16.~and
his continual glorie and praiſe.}

\PTitle{Concerning
%%% !!! SNote should be before 'Concerning'
\SNote{This Pſalme is of Chriſt, perfigured by Salomon, whoſe kingdom
was moſt glorious of al the kinges of Gods people; for of king Salomon
himſelfe manie thinges in this Pſalme can not be truly vnderſtood.
%%% !!! Fix Cite
\Cite{S.~Aug.}}
Salomon.}

\VV O God giue thy iudgement to
\SNote{O God moſt bleſſed Trinitie, geue powre and authoritie, to the
Sonne of Man, God incarnate, King of al kinges,}
the king: and thy iuſtice to
\SNote{the Sonne of king Dauid, to iudge for mankind againſt the diuel.}
the ſonne of the king.

To iudge thy people in
\SNote{Chriſt paying ranſom for al mankind, and ſo man renouncing the
diuel, and ſeruing God is iuſtly, not iniuriouſly deliuered from
captiuitie of ſinne, and of the diuel.}
iuſtice, and thy poore in iudgement.

\V Let
\SNote{A prophecie of the Apoſtles receiuing powre to preach Chriſts
Goſpel, of peace and reconciliation of men to God by penance;}
the mountains receiue peace for the people: and
\SNote{and of other Apoſtolical men that folow their ſteppes.}
the litle hilles iuſtice.

\V He
\SNote{Salomon in figure of Chriſt was for a time a iuſt and good king.}
shal iudge the poore of the people, and shal ſaue the children of the
poore: and he shal humble the calumniator.

%%% 1267
\V And he shal
\SNote{But only Chriſt, not Salomon, nor anie other king of that people,
contineweth, or reigneth for euer.}
continew with the ſunne, and before the moone in generation and
generation.

\V He
\SNote{The maner of Chriſts Incarnation moſt ſilent, ſwete, and
gratful.}
shal deſcend as rayne vpon a fleece: and as droppes diſtilling vpon the
earth.

\V There shal riſe in his dayes iuſtice, and aboundance of peace,
\SNote{Agane the prophet inculcateth that Chriſts kingdom, his Church,
ſhal continew for euer.}
vntil the moone be taken away.

\V And he shal rule
\SNote{The Church is not only vniuerſal in al times, but alſo in
places.}
from ſea vnto ſea: and from the riuer euen to the ends of the round
world.

\V
\CNote{\XRef{Iſaia.~60.}}
Before him shal the Æthiopians fal downe: and his enimies shal lick the
earth.

\V
\CNote{\XRef{Mat.~2.}}
The
\SNote{The three Sages or Kinges, which adored our Sauiour, and offered
gold, frankencenſe, and myrrh, were the firſt that fulfilled this
prophecie: and afterwards Conſtantin the great, and other Emperors,
Kinges, and Princes.}
kinges of Tharſis, and
\SNote{Amongſt other Ilandes great Bryttanie (the greateſt of Europe) was
conuerted to Chriſt, according to this prophecie: firſt ſome few in the
Apoſtles time
\Cite{(Metaphraſtes apud Surium,}
\Cite{Theodoret, epiſt. ad Timoth.}
\Cite{Sophronius, Ser. de Nat. Apoſt.}
& \L{alij.}) More in the time of Eleutherius. Laſtly our
\Fix{Engliſe}{Engliſh}{obvious typo, fixed in other}
nation by S.~Auguſtin and others ſent by S.~Gregorie.}
the Ilands shal offer preſentes: the kings of the Arabians, and of Saba
shal bring giftes.

%%% o-1158
\V And al kinges of the earth shal adore him: al nations shal ſerue
him.

\V Becauſe he shal deliuer the poore from the mightie: and the poore
which had no helper.

\V He shal ſpare the poore and needy: and he shal ſaue the ſoules of the
poore.

\V From vſuries and iniquitie he shal redeme their ſoules: and their
name shal be honorable before him.

\V And
\SNote{How is this prophecie verified, except the Church be alwaies
viſible?}
he shal liue, and there shal be giuen him of the gold of Arabia, and
they shal adore it alwaies: al the day they shal bleſſe him.

\V And there shal be a firmament in the earth, in the toppes of
mountanes, the fruite therof shal be extolled farre aboue Libanus: and
they shal florish of the citie, as the graſſe of the earth.

\V Be his name bleſſed for euer: before the ſunne his name is permanent.

And al the tribes of the earth shal be bleſſed in him: al nations shal
magnifie him.

%%% 1268
\V Bleſſed be our Lord the God of Iſrael,
\SNote{No miracles can be donne but by Gods powre.}
who only doth meruelous thinges.

\V And bleſſed by the name of his maieſtie for euer: and al the earth
shal be filled with his maieſtie.
\SNote{Aboue al other deſires the holie prophet wiſhed Gods glorie, and
praiſes in al the earth, as it is in heauen.}
Be it, be it.

\V
\SNote{It ſemeth by this appendix added by Eſdras, that this Pſalme was
laſt compoſed, though not put in the laſt place, but as it came to his
handes.}
The praiſes of Dauid,
\SNote{Dauid the natural ſonne of Ieſſe, made no more Hymnes, that is,
no more muſical praiſes of God after this Pſalme, but the prefigured
Dauid Chriſt our Sauiour, the Sonne of God, in his bodie the Church,
ſingeth ſtil new praiſes.
\Cite{S.~Aug. in enarrat. Pſal. ſequintis.~72.}}
the ſonne of Ieſſie are ended.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
