%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 016
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-016


\project douay-rheims


%%% 1172
%%% o-1064
\startChapter[
  title={Psalme 16}
  ]

\PSummary{A
\MNote{Gods prouidence protecting the iuſt.

The 3.~key.}
iuſt mans prayer in tribulation, 10.~deſcribing his enemies
  cruelty, 13.~by way of imprecation foresheweth their deſtruction,
  15.~and declareth that the iuſt shal be ſatisfied in glorie.}

\PTitle{The
\LNote{The prayer of Dauid.}{This
\MNote{This Pſalme called a Praier is both a ſword & buckler in
afflictiõ.}
Pſalme of the matter conteyned is called a prayer. VVhich holie Dauid ſo
compoſed, as was  both conuenient for himſelfe, being moleſted with
vniuſt afflictions by the wicked, and for anie other iuſt perſon, or the
whole Church in perſecution, ſeruing as a ſpiritual ſword to ſtrike the
enimies, and as a ſhield to beare of with patience and fortitude al
their forces.}
prayer of Dauid.}

\VV Heare ô Lord my
\SNote{In my iuſt cauſe heare my petition,}
iuſtice: attend my petition:

With thine eares heare my prayer,
\SNote{ſeing I pray ſincerly, not in feaned affection.}
not in deceitful lippes.

\V From
\SNote{Thou that ſeeſt al things declare my right againſt mine
aduerſaries.}
thy countenance let my iudgement procede: let thine eies ſee equities.

\V Thou haſt proued
\SNote{My intẽtion.}
my hart, and viſited it
\SNote{Moſt ſecret cogitations.}
by night:
\SNote{By tribulations.}
by fire thou haſt examined me,
\SNote{Whoſe conſcience is pure from greuous ſinne, may pray with this
confidence, otherwiſe repentance is firſt neceſſarie. But the whole
Church may euer pray in this maner, becauſe there be alwayes ſome iuſt
and holy, in reſpect of whom it is truly called holy.}
and there is no iniquitie found in me.

\V That my mouth ſpeake not the workes of men: for the
\SNote{For thy precepts.}
wordes of thy lippes I haue kept the
\SNote{The narrow way of vertue.}
hard wayes.

\V
\SNote{None can of themſelues walke rightly, but by Gods helpe.}
Perfite my paſes in thy pathes: that my ſteppes be not moued.

\V I haue cried, becauſe thou haſt heard me ô God: incline thyne eare
%%% o-1065
to me, and heare my wordes.

\V Make thy mercies meruelous, which ſaueſt them that hope in thee.

\V From them that reſiſt
\SNote{Againſt thy omnipotent powre.}
thy right hand keepe me, as the apple of the eie.

\V Vnder the shadowe of thy winges protect me: \V from the
\SNote{From their cruel and furious countenance.}
face of the impious, that haue afflicted me.

%%% 1173
Mine enemies haue compaſſed my ſoule, \V they haue shut vp their
\SNote{They haue shut out al pitie or commiſeration.}
fatte; their mouths hath ſpoken pride.

\V Caſting me forth now haue they compaſſed me: they haue ſette their
eies to bend them
\SNote{They intend vtterly to deſtroy me euen to the ground.}
vnto the earth.

\V They haue taken me as a lion readie to the pray: and as a lions
whelpe dwelling in hid places.

\V Ariſe Lord,
\SNote{Except God preuent, mãs induſtry is not ſufficiẽt.}
preuent him, and ſupplant him: deliuer my ſoule from the impious,
\SNote{Reſtraine their powre, which they haue by thy permiſſion, that
they may not perſecute ſo much as they intend.}
thy ſword, \V from the enemies of thy hand.

Lord from
\SNote{This is a prophecy, that the wicked which are many, shal at the
day of iudgement be ſeparated from the elect,}
a few out of the land diuide them,
\SNote{which iudgement beginneth ſometimes in this life.}
in their life: their bellie is filled of
\SNote{The pleaſures of this world, which God approueth not, nor
acknowledgeth amongſt good thinges.}
thy ſecretes.

They are filled with children: and they haue leaſt their remnantes to
their litle ones.

\V But
\SNote{The iuſt shal be approued.}
I in iuſtice shal appeare to thy ſight: I shal be
\SNote{Nothing doth ſatiate mans mind, but the ſight of God in eternal
glorie.}
filled when thy glorie shal appeare?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
