%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 113
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-113


\project douay-rheims


%%% 1342
%%% o-1232
\startChapter[
  title={Psalme 113}
  ]

\PSummary{For
\MNote{The meruelous paſſage of Iſrael from Ægypt.

The 4.~key.}
the meruelous paſſage of Iſrael out of Ægypt, 3.~the red ſea, the riuer
of Iordan, 7.~and the hilles geuing them place, 8.~the rockes yelding
them water, 9.~God, not themſelues, is to be praiſed. 12.~Idols and
Idolaters are vaine, and shal be confounded: 17.~the faithful truſt in
God, 20.~are bleſſed, and for euer praiſe God.}

\PTitle{Alleluia.}

\NV In the comming forth of Iſrael out of Ægypt, of the houſe of Iacob
from
\SNote{People of falſe religion counted barbarous, eſpecially ſuch as
alſo perſecute the true Religion; for otherwiſe the Ægyptians were both
ciuil in maners, and lerned in manie ſciences.}
the barbarous people.

\V
\SNote{The people of the Iewes were more notoriouſly renowmed in the
world from the time of their deliuerie out of Ægypt, for the peculiar
people, whom God ſanctified, and in whom, as in his elected enheritance
or dominion, he dwelled and reigned.}
Iewrie was made his ſanctification, Iſrael his dominion.

\V The
\SNote{The Pſalmiſt writing in verſe doth often deſcribe thinges in
poetical maner, but more truly then prophane poetes: for that in very
dede, al creatures otherwiſe ſenſles, as the ſea, do in a ſorte feele
the powre of their Creator, & obey his wil.}
ſea ſaw, and
\CNote{\XRef{Exo.~14.}}
\SNote{VVhen the Iſraelites went forth of Ægypt,}
fled: Iordan
\CNote{\XRef{Ioſue.~3.}}
\SNote{when they entred into Chanaan.}
was turned backeward.

\V
\SNote{Either there was an earthquake, or ſome other mouing of hilles
not mentioned by Moyſes, or els the Pſalmiſt ſpeaketh of the rockes of
the torrentes which bowed, that the Iſraelites might reſt in Ar, and lie
in the borders of the Moabites,
\XRef{Num.~21. v.~15.}}
The mountaines leaped as rammes: and the litle hilles as the lambes of
sheepe.

\V
\SNote{By the figure Apoſtrophe he ſpeaketh to the ſea, riuer, and
hilles, vſing alſo Proſopopœia: as if ſenſles thinges, vnderſtood, and
ſhould anſwer.}
What ayleth thee ô ſea that thou didſt flee: and thou ô Iordan, that
thou waſt turned backeward?

\V Ye mountaines leaped as rammes, and ye litle hilles as the lambes of
shepe.

\V At the face of our Lord was the earth moued, at the face of the God
of Iacob.

\V Who turned
\SNote{As other miraculous benefite, that the rocke yelded them water in
their neceſſitie.}
the rocke into pooles of waters, and ſtonie hil into fountaines of
waters.

%%% 1343
\V
\SNote{Here ſome Hebrewe Rabbins beginne an other Pſalme, but by the
coherence of the matter, S.~Auguſtin proueth that it is but one Pſalme,
where is ſhewed that the true inuiſible God is knowen by ſuch workes as
are here recited: and contrariwiſe, that the Gẽtils idoles are not
goddes, becauſe they are made of ſiluer, gold, or other matter, by mens
handes, hauing reſemblance of liuing thinges are altogether ſenſles.}
\Sc{Not to vs o Lord, not to vs:} but to thy name geue the glorie.

\V For
\SNote{Thou didſt al this ô God, of mere mercie towards thy people:}
thy mercie, and
\SNote{for thy truths ſake, ſeing thou didſt promiſe to protect them:}
thy truth: leſt at any time
\SNote{that the Gentiles ſhould not take occaſion to blaſpheme.}
the Gentiles ſay: Where is their God?

%%% o-1233
\V But our God is in heauen: he hath done al thinges what ſoeuer he
would.

\V
\LNote{The Idols of the Gentiles are ſiluer and gold.}{Al
\MNote{The definition of Idolatrie.}
Catholique Diuines agree in this authentical definition, of Idolatrie,
that it is diuine honour geuen to anie creature, as to a god.
\MNote{Diuers ſortes of Idolatrie.}
Of the diuers ſortes alſo of Idolatrie the ancient lerned Doctors haue
vvritten much. Namely Iuſtinus Martyr in his
\Cite{Orations againſt the Gentiles,}
Tertullian in
\Cite{Apologetico,}
Arnobius
\Cite{Orat. ad Gentes.}
Lactantius,
\Cite{li.~2. c.~7. Diuinar. Inſtit.}
and manie others. But moſt copiouſly and profoundly S.~Auguſtin,
expecially in his
\Cite{tenne firſt bookes de Ciuitate Dei.}
\MNote{Angels honored as goddes.}
Into vvhich error & crime the Platoniſtes fel, holding that ſpiritual
inuiſible created ſubſtances, to vvitte, Angels good and euil, vvhom
they called \L{Intelligentias ſeparatas}, had diuine povvre, & ſo gaue
to them diuine honour.
\MNote{Men liuing or dead.}
Others honour dead men, and ſome before their death as goddes, for their
notable actes atchiued in this life, as Saturne, Iuppiter, Hercules, and
the like.
\MNote{Corporal creatures, ſenſible and without ſenſe.}
Some yeld diuine honour to mere corporal creatures, liuing or vvithout
life, as to beaſtes and ſerpentes, the ſunne, the moone, fire, vvater,
earth, the vvhole machin of the vvorld, as if it vvere animate, and that
vvith diuine ſpirite or ſoule.
\MNote{Imagees of falſe goddes.}
Againe al theſe haue bene vvorſhipped as gods, not only in them ſelues
but alſo in their imagees.
\MNote{Imagees them ſelues reputed goddes.}
But to omite other diuerſities, the moſt groſſe ſorte of al, is the
Idolatrie vvhich the Pſalmiſt here deſcribeth, vvhen Imagees made by
mens handes are immediatly honored as goddes in themſelues. For ſuch
goddes the Gentiles had, and of them the Prophet here
ſpeaketh. Comparing theſe viſible ſenſles imagined goddes, vvith the one
eternal inuiſible God; vvho is made notorious by his diuine conſpicuous
vvorkes, vvheras theſe idols, by hovv much they are more viſible, ſo
much more they
are contemptible: becauſe being faſhioned vvith mouth, eyes, eares, noſe,
handes, feete, throte, and other members, they are altogether ſenſles,
not able to ſpeake vvith their mouth, and therin more baſe in nature,
then the men that made them; not able to ſee, heare, ſmel, taiſt, feele,
to vvalke, to moue, nor to crie, al vvhich beaſtes can doe.
\MNote{Idolaters are voide of reaſon.}
And therfore
thoſe that make them, or haue confidence in them (as the Prophet here
ſignifieth that ſome haue
\XRef{v.~16.)}
are moſt abſurde: becoming like to the ſame idols, in their
vnderſtanding and internal cogitations, as theſe idols are voide of
external ſenſe.
\MNote{And ſeruants of diuels.}
And al Idolaters are moſt vvicked, geuing diuine honour
to anie creature, & are therin ſeruantes of diuels: VVhether they
immediatly honour diuels, as vvhen ſorcerers and vvitches, making pact
vvith the diuel, adore him, and he for the ſame doth ſome thing vvhich
they demand: or that they honour ſome other creature, vvherin by the
diuels falſe perſvvaſion, they thincke there is diuine
\Fix{poore.}{power.}{obvious typo, fixed in other}
Both vvhich vvaies diuels vſurpe Gods honour and therfore the ſame Prophet
ſaide in an
\CNote{\XRef{Pſal.~95. v.~5.}}
other Pſalme: Al the goddes of Gentiles are diuels.}
The idols of the gentiles are ſiluer, and gold, the workes of mens
handes.

\V They haue mouth, and shal not ſpeake: they haue eies, and shal not
ſee.

\V They haue eares, and shal not heare: they haue noſthrels and shal not
ſmel.

\V They haue handes, and shal not handle: they haue feete, and shal not
walke: they shal not crie in their throte.

\V
\SNote{This is a iuſt prayer of the zelous, conforming their deſires to
Gods wil. But if God geue idolaters grace to amend, then al the iuſt wil
alſo reioice in their conuerſion.}
Let them that make them become like to them: and al that haue confidence
in them.

\V
\SNote{Though manie Iewes fel to idolatrie, yet there alwayes remained
ſo manie in Gods true ſeruice, that it might ſtil be truly ſaide: The
houſe of Iſrael hath hoped in our Lord, as is here auerred.}
The houſe of Iſrael hath hoped in our Lord: he is their helper and their
protector.

\V The houſe of Aaron hath hoped in our Lord: he is their helper and
their protector.

\V They that feare our Lord, haue hoped in our Lord: he is their helper
and their protector.

\V Our Lord hath beene mindful of vs: and hath bleſſed vs:

He hath bleſſed the houſe of Iſrael: he hath bleſſed the houſe of Aaron.

\V He hath bleſſed al, that feare our Lord, the litle with the great.

\V Our Lord adde vpon you: vpon you, & vpon your children.

\V Bleſſed be you of our Lord, which made heauen, and earth.

\V
\SNote{This in effect al worldlie politikes ſay in their hartes: as it
were  quitting their intereſt of heauen to God,}
The heauen of heauen is to our Lord: but
\SNote{and contenting themſelues with earthlie poſſeſſions.}
the earth he hath geuen to the children of men.

%%% 1344
\V
\SNote{But when ſuch prophane men are dead, they make no ſhew at al of
praiſing God:}
The dead shal not prayſe thee ô Lord: nor al they
\SNote{for parting from the earth, they deſcend into hel, and there
eternally blaſpheme God.}
that goe downe
\Fix{hel.}{into hel.}{obvious typo, fixed in other}

\V But we
\SNote{Contrarivviſe the iuſt, aſpiring to heauen, vvhich is the proper
kingdom of God, & vſing this vvorld as they ought to do, for a meanes to
aſcend into heauen, ſhal bleſſe and praiſe God for euermore.}
that liue, doe bleſſe our Lord, from this time, and for euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
