%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 143
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-143


\project douay-rheims


%%% 1392
%%% o-1283
\startChapter[
  title={Psalme 143}
  ]

\PSummary{The
\MNote{King Dauid praiſeth God for his victories.

The 8.~key.}
royal Prophet thanketh God, for al his victories, and poſſeſſion of the
kingdom. 3.~Admiring Gods benignitie towards man, 5.~prayeth to be ſtil
defended from al enimies, 9.~promiſeth a new ſongue of prayſe,
11.~deſcribeth the vanitie of worldlie men, 15.~concluding that true
felicitie is in ſeruing God.}

\PTitle{A Pſalme of Dauid,
\SNote{Becauſe this vvas Dauids firſt, and a very notable victorie, the
Septuagint Interpreters make mention of Goliath in this title.}
againſt Goliath.}

\NV Bleſſed be our Lord my God, who
\SNote{God made Dauid a warier, and victorer againſt Goliath, without
anie former training in armes.}
teacheth my handes to battel; and my fingers to warre.

%%% 1393
\V My mercie, and my refuge: my defender, and my deliuerer.

My protectour, and I haue hoped in him, who
\SNote{Made him afterwards king of a great people.}
ſubdeweth my people vnder me.

\V Lord
\SNote{Al mankind was vnvvorthie before Chriſt,}
what is man, that thou art
\SNote{that God should be reueled vnto them,}
made knowne to him? or the ſonne of man,
\SNote{eſpecially that he ſhould haue care of the progenie of men, after
their ſinne.}
that thou eſtemeſt him?

\V
\SNote{In dede man in himſelf, in his ovvne nature and frailtie, is but
a vaine and tranſitorie creature, paſſing from life to death, as a
ſhadovv that can not conſiſt of it ſelfe, neither can man vvithout God.}
Man is made like to vanitie: his dayes paſſe as a shadow.

\V
\SNote{By a poetical deſcription he prayeth for Gods helpe, as if God
ſhould make the heauens to bovv, and ſo deſcend; or make the montaines
ſmoke, as vvhen Moyſes receiued the lavv; or declare himſelf by Meteors,
as folovveth:}
Lord incline thy heauens, and deſcend: touch the mountaynes, and they
wil ſmoke.

\V Lighten lightening, and thou shalt diſperſe them: shoote out thine
arrowes, and thou shalt deſtroy them.

\V Send forth thy hand from on high, take me out, and deliuer me from
manie
\SNote{Tribulations or tentations.}
waters: from the hand
\SNote{Children of the Church,}
of children
\SNote{but of euil life.}
ſtrangers.

%%% o-1284
\V Whoſe mouth hath ſpoken
\SNote{They ſpeake in vanitie that promiſe to kepe Gods law, and
performe it not.}
vanitie: and their right hand is the right hand of iniquitie.

\V O God I wil ſing to thee a new ſong: in
\SNote{In that principal inſtrument, apt for a new ſongue, and for
extraordinarie benefites.}
the pſalter of ten ſtringes, I wil ſing to thee.

\V Who geueſt ſaluation to kinges: who haſt redemed Dauid thy ſeruant
from the malignant ſword: \V deliuer me.

And reſcue me out of the hand of
\SNote{Both Ievves & Chriſtians that liue not vvel, are as ſtrangers
that frame to themſelues ſuch a falſe felicitie, as is here deſcribed:
making riches or vvorldlie pleaſures their God.}
children ſtrangers, whoſe mouth hath ſpoken vanitie: and their right
hand, is the right hand of iniquitie.

\V Whoſe ſonnes, are as new plantes in their youth.

Their daughters comly trimmed: decked about after the ſimilitude of a
temple.

\V Their ſtorehouſes ful, flowing out of this into that.

Their ewes ful of yong, abunding in their going forth: \V their oxen are
fatte.

There is no ruine of wal, nor
\TNote{tranſmigration.}
paſſage, nor crie in their ſtreates.

\V They haue ſaid, that it is a happie people, which hath theſe things:
\SNote{True happines conſiſteth not in vvorldlie thinges,}
bleſſed is the people, whoſe
\SNote{but in preferring God before al.}
God is our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
