%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 067
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-067


\project douay-rheims


%%% 1256
%%% o-1147
\startChapter[
  title={Psalme 67}
  ]

\PSummary{Notwithſtanding
\MNote{The Church ſtil conſerued.

The 6.~key.}
great perſecutions the Church proſpereth. 10.~Eſpecially in the new
Teſtament, by Apoſtolical function, 18.~miniſterie of Angels, Chriſts
Aſcenſion, coming of the Holie Ghoſt, 31.~confirming the faithful,
repreſſing the inſolent, and conuerting manie. 35.~For al which the
prophet inuiteth al men to praiſe God.}

\PTitle{Vnto the end, a Pſalme of Canticle, to Dauid himſelf.}

\VV Let
\SNote{In maner of praying that God wil vouchſafe to defend the Church,
the Pſalmiſt prophecieth that God wil ariſe,}
God ariſe, and let his enimies be diſperſed, and let them that hate him
\SNote{and the enimies ſhal flee away, not daring to abide the combate.}
flee from his face.

%%% o-1148
\V As ſmoke vanisheth, let them vanish away: as waxe melteth at the
preſence of fire, ſo let ſinners perish at the preſence of God.

\V And
\SNote{As God is terrible to the wicked: ſo he is comfortable to the
iuſt.}
let the iuſt make merrie, and reioyce in the ſight of God: and let them
be delighted in mirth.

%%% 1257
\V Sing to God, ſay a Pſalme to his name:
\SNote{Reſiſt not Gods inſpiration, but receiue it with ioy and thankes.}
make way to him, who
\SNote{Who triumpheth ouer death.}
mounteth vpon the weſt,
\SNote{God is Lord, not only of theſe or thoſe nations, countries or
other creatures, but abſolutly and vniuerſally of al.}
Lord is his name.

Reioyce ye in his ſyght, they shal be trubled at the preſence of him: \V
the father of orphanes, and iudge of widowes.

God in his
\SNote{That is the true holie Church, which hath}
holie place: \V God that maketh men to inhabite
\SNote{vnitie in doctrin, touching faith and maners.
\Cite{S.~Cypr. ep.~76.}}
of one manner in a houſe.

That bringeth forth them,
\SNote{That be bond in ſinne.}
that be bound, in
\Fix{ſtrengh,}{ſtrength,}{obvious typo, fixed in other}
likewiſe them, that
\SNote{Euen rebellious willes are altered by Gods mercie, and freely
embrace his law.}
exaſperate, that dwel
\SNote{Alſo the dead and drie hartes that cared not for ſpiritual
thinges, are ſoftened, and quickned with new grace.}
in ſepulchers.

\V O God
\SNote{The benefites beſtowed on the Iſraelites, are written in the
bookes of Moyſes, Ioſue, and Iudges.}
when thou wenteſt forth in the ſight of thy people, when thou didſt
paſſe through the deſert.

\V The earth was moued, and the heauens alſo diſtilled, at the preſence
of the God of Sina, at the face of the God of Iſrael.

\V
\SNote{Not mans deſeruing, but Gods mere good wil, and free grace cauſed
Chriſt to come, and by himſelfe and his Apoſtles to preach the
Euangelical doctrin, which watereth the whole world.}
Voluntarie rayne shalt thou ſeperat ô God to thine inheritance: and it
was
\SNote{God choſe the weake, but made them ſtrong.}
weakned, but thou haſt perfited it.

\V Thy
\SNote{Thoſe whom thou haſt choſen, and ſo made thyn owne peculiar
people, ſhal enioy this grace.}
liuing creatures shal dwel in it: thou haſt prepared
\SNote{Thou gaueſt Manna in the deſert, the B.~Sacrament in the new
Teſtament.}
in thy ſwetnes for the poore, ô God.

\V Our Lord shal geue
\SNote{God geueth to the preacher what to ſpeake,}
the word to them that euangelize,
\SNote{and to ſome he geueth alſo powre to worke miracles, in
confirmation of their doctrin.
\XRef{Mar.~16.}}
with great powre.

\V The
\SNote{Some potent king, or (as it is in the Hebrew) kinges being
beloued, of the beloued of God, the only Sonne of God, ſhal yeld
themſelues to the ſame beloued Sonne of God:}
king of hoaſtes the beloued of the beloued: and
\SNote{which ſhal redound to the glorie and beautie of his Church, gayning
ſuch ſpiritual prayes from the diuel.}
to the beautie of the houſe, to diuide the ſpoyles.

\V If ye
\SNote{If you be in ſuch danger, that the aduerſaries caſt dice, or
lottes for your perſons, and goodes, yet you ſhal be deliuered, as if a
doue, with her gliſtering fethers, like ſiluer and gold, flie away into
a ſecure place, without loſſe or diminution, but rather with increaſe of
vertues.}
ſleepe among the middeſt of the lottes, the winges of a doue layde ouer
with ſiluer, and the hinder parts of her backe in the palenes of gold.

%%% 1258
%%% o-1149
\V Whiles
\SNote{VVhen the heauenlie king determineth thus of earthlie kinges,}
the heauenlie diſcerneth kings ouer her,
\SNote{they ſhal be purged from their ſinnes, and made white like ſnow,
that falleth in mount Selmon, which is a ſhadowed hil, thicke with
trees, in mount Ephraim nere to Iordan.}
with ſnow they shal be made white in Selmon.

\V
\LNote{The mountaine of God.}{For
\MNote{Markes of the Church:}
better diſcerning the true Church from other congregations, the Prophet
here deſcribeth certaine properties therof, for he calleth it a
mountane,
\MNote{Viſibilitie.}
becauſe it is moſt viſible to al men. Secondly a fatte mountaine, that
is,
\MNote{Sanctitie.}
repleniſhed vvith al vertues, and giftes of the Holie Ghoſt, vvherof it
is called Holie. Thirdly it is crudded, or
\MNote{Vnitie.}
conſolidated in vnitie of
faith and Religion, vvhich conioyne the vvhole bodie, making it ſolide
and firme, as the ruen turneth the liquide milke into curde, and ſo into
cheeſe. Fourtly, it is the Congregation,
\MNote{Perpetuitie.}
vvherin God alvvayes remaneth,
euen to the end, for euer; vvhich ſhevveth tvvo other properties; that
\MNote{Aſſured veritie.}
the Church neuer faileth; nor erreth in doctrin: God ſtil dvveling
therin, and conſequently conſerueth it from error in doctrin.}
The
\SNote{The Church of God is viſible, and durable like to a mountane.}
mountane of God a fat mountane.

\V A mountane
\SNote{Combined, or ioyned together, as when milke is turned into curde,
and ſo into cheeſe.}
crudded as cheeſe, a
\SNote{Fruictful, enriched by ſpiritual giftes of the Holie Ghoſt.}
fatte mountane: \V
\LNote{VVhy ſuppoſe you crudded montaines?}{As
\MNote{No other pretended Church hath the marke of vnitie or the reſt.}
for other congregations, it is certaine and euident, that they are not
the Church of God, becauſe they are not crudded, that is, not vnited, in
the ſame pointes of faith, but only in negatiue pointes, and in general
oppoſition againſt the Catholique Church, and among themſelues
notoriouſly diſagreing and diuided. As they alſo vvant the other markes
of the true Church.}
why
\SNote{Ye that are not of this Church, do in vaine and erroniouſly
imagine, that anie other mountaines are vnited.}
ſuppoſe you crudded mountanes?

A mountane, in which it hath wel pleaſed God to dwel therin: for in dede
our Lord wil dwel euen to the end.

\V The
\SNote{Innumerable Angels miniſters of Gods wil, do continually attend
vpon his Diuine Maieſtie, as if he (who otherwiſe nedeth no ſeruice) were
caried by them, as in a chariotte of infinite magnificence.
\XRef{Dan.~7.}}
chariote of God is ten thouſand folde, thouſands of them that reioyce,
our Lord in them,
\SNote{So God appeared in Maieſtie, when he gaue his law in mount Sinai.}
in Sina in the holie place.

\V
\CNote{\XRef{Epheſ.~4.}}
Thou art
\SNote{Chriſt aſcended with innumerable Angels attending vpon him,}
aſcended on hygh, thou
\SNote{caried with him the fathers of the old Teſtament, that had benne
captiue,}
haſt taken captiuitie: thou
\SNote{as man he receiued giftes of God, in and for men, his faithful
ſeruantes,}
haſt receiued gifts in men: for euen thoſe
\SNote{yea alſo he receiued for his merite, that innumerable, which
before were incredulous, were conuerted, and God dwelt in their ſoules.}
that do not beleue, our Lord God to inhabite.

\V Bleſſed be our Lord day by day: the God of our ſaluations wil make vs
a proſperous iourney.

\V Our God is the God of ſauing: and
\SNote{Our Lord, I ſay, our Lord, and none but he could ouercome death
by dying.}
the iſſues of death are of our Lord, our Lord.

\V But yet God
\SNote{Though Chriſt died to deliuer al men from death, yet he wil geue
capital ſentence of eternal death to al that obſtinatly remaine his
enimies, and multiplie ſinnes vpon ſinnes, to the end of their temporal
life.}
ſtil breake the heads of his enimies: the hearie croune of them, that
walke in their ſinnes.

\V Our Lord ſaid: Out
\SNote{Euen of the rudeſt barbarous nations, manie ſhal be conuerted to
\Fix{Chriſtiantie.}{Chriſtianitie.}{possible typo, same in both}}
of Baſan I wil conuert, I wil conuert
\SNote{Namely Gods grace is extended into the Ilandes of the Ocean, and
other ſeas.}
into the depth of the ſea.

\V That thy foote
\SNote{But ſuch ſeuere ſlaughter ſhal fal vpon the obſtinate contemners
of this grace, that mens feete ſhal be defiled in their bloud, and
dogges ſhal lappe it.}
may be dipped in bloude: the tongue of thy dogges
%%% o-1150
[\Emph{made redde}] with
\TNote{\L{ex inimicis ab ipso}}
the ſame [\Emph{bloud}] of the enimies.

%%% 1259
\V They
\SNote{Manie haue ſene, or knowen in general, but the faithful more
exactly know how Chriſt came into this world, his conuerſation therin,
and his going forth:}
haue ſeene thy entrings in ô God, the entrings of my God: of my King Who
is
\SNote{his reigning now in heauen our Mediatour, by whom al other
interceſſors haue acceſſe to God.}
in the holie place.

\V
\SNote{The Apoſtles ſowing the firſt ſeede of Euangelical doctrin,}
Princes came before
\SNote{with whom other Apoſtolical men,}
ioyned with them that ſang, in the middes of
\SNote{and other ſoules of al nations conuerted by their preaching, moſt
ioyfully ſing together in hart, voice, and inſtrument: eſpecially in
good workes, ſhew their gratful affections to our Redeemer.}
young wemen plaing on tymbrels.

\V In
\SNote{And al this in the particular Churches of diuers Kingdomes, and
partes of the world:}
churches bleſſe ye God our Lord, of
\SNote{beginning with the Iſraelites in Ieruſalem, and ſo proceding into
al Iurie, and Samaria, and to the vtmoſt of the earth.
\XRef{Act.~1.}}
the fountains of Iſrael.

\V There
\SNote{S.~Paul of Iacobs yongeſt ſonne Beniamin, laſt called to
Apoſtleſhip, was chiefly ſent to the Gentiles.}
Beniamin a youngman, in exceſſe of minde.

The
%%% !!! SNote belongs befoe 'The'
\SNote{Other Apoſtles of diuers tribes ſent firſt to the Iewes,
ſecondarily to Gentiles.}
Princes of Iuda: their leaders: the Princes of Zabulon, the Princes of
Nephthali.

\V Command thy ſtrength ô God:
\SNote{As the Church begane by the omnipotent powre of God; ſo by the
ſame only powre it is conſerued.}
confirme this ô God, which thou haſt wrought in vs.

\V From thy temple in Ieruſalem, kinges shal offer giftes to thee.

\V Rebuke
\SNote{Chaſtice therfore ô God, al perſecuters of thy Church, who are
but as weake wauering reedes, in compariſon of thy powre:}
the wilde beaſts of the reede, the congregation
\SNote{no better then bulles, with kyne, that is, captaines and popular
people,}
of bulles in the kine of thy peoples:
\SNote{endeuoring to alienate the conſtant proued confeſſors from their
faith.}
that they may exclude them, which are tried with ſiluer.

Diſſipate the nations that wil warres: \V Legates shal come
\SNote{A prophecie that manie ſhould be conuerted to Chriſt in Ægypt,
and Æthiopia: as appeareth by the innumerable multitude of religious
Monkes, & Nunnes in thoſe countries, ſhortly after the Apoſtles dayes.}
out of Ægypt: Æthiopia shal
%%% !!! ??? 'v' in middle of a word: Latin is præveniet
prevent his handes to God.

\V Ye kingdomes
\SNote{The like afterwards in al other nations, whom therfore the
prophet inuiteth to praiſe God, for ſo ineſtimable benefites in the
whole world.}
of the earth ſing to God: ſing to our Lord:

\V Sing ye to God, that mounteth vpon the heauen of heauen, to the Eaſt.

%%% 1260
%%% o-1151
Behold he wil giue to his voice
\SNote{Chriſt wil come to iudge, in terrour of voice, and vvith
magnificence, accompanied vvith holie Angels and other Sainctes.}
the voice of ſtrength, \V giue you glorie to God vpon Iſrael, his
magnificence, and his powre in the cloudes.

\V God is meruelous in his Saintes, the God of Iſrael he wil giue powre,
and ſtrength to his people, God be bleſſed.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
