%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 76
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-76


\project douay-rheims


%%% 1275
%%% o-1166
\startChapter[
  title={Psalme 76}
  ]

\PSummary{Anie
\MNote{Gods ſpecial protection of the Ievves.

The 4.~key.}
faithful deuout perſon meditating Gods benefites, 7.~examineth his
cõſcience, that nothing be in his ſoule, that may offend God. Who is
alwayes readie to forgeue: 11.~and therfore he ſtil reneweth his purpoſe
to ſerue God ſincerly, 15.~particularly remembring the deliuerie of
Iſrael from Ægypt.}

\PTitle{Vnto the end, for
\SNote{For Idithun to ſing, or to make tune for it.}
Idithun, a pſalme
\SNote{For the faithful congregation to conſider Gods benefites.}
to Aſaph.}

\VV With
\CNote{\XRef{Pſal.~141.}}
\SNote{Hauing heretofore prayed,}
my voice I haue cried to our Lord: with my voice to God
\SNote{I haue obtayned.}
and he attended to me.

\V In the day of my tribulation I ſought God, with my handes
\SNote{Eſpecially being in tribulation, and praying vvith hart and
handes lifted vp, as vvel in the night as day:}
in the night before him: and I
\SNote{I vvas not fruſtrate of my prayer.}
was not deceiued.

%%% 1276
My
\SNote{I vvas ſometimes in ſuch anguiſh that nothing ſemed comfortable,}
ſoule refuſed to be comforted, \V I
\SNote{but I relied vpon God ſo firmly,}
was mindeful of God, and was delighted, and was exerciſed: and
\SNote{that my ſpirite came into an extaſie, or traunce.}
my ſpirite fainted.

\V Myne eies
\SNote{I aroſe early before the ordinarie time of avvayking,}
preuented the watches: I was
\SNote{my hart being attentiue invvardly, I vttered nothing vvith my
tongue.}
trubled, & ſpake not.

\V I thought vpon old daies: and the eternal yeares I had in minde.

\V And I meditated in the night with my hart, and I was exerciſed, and I
\SNote{I diligently examined my conſcience.}
ſwept my ſpirite.

\V Why,
\SNote{Aſſuredly God vvil not reiect for euer, but he vvil be pleaſed
with his Church.}
wil God reiect for euer: or wil he not adde to be better pleaſed as yet?

\V Or wil he cutte of his mercie for euer, from generation vnto
generation?

\V Or wil God forget to haue mercie? or wil he in his wrath keepe in his
mercies?

\V And I ſayde:
\SNote{VVhiles I thus thought, I erred greatly, novv I ſee and confeſſe
that God ſuffereth al calamities for the good of his ſeruantes:}
Now haue I begune:
\SNote{and this I knevv not by my ſelfe, but by the inſpiration of God,
making this change in me by his gracious hand.}
this is the change of the right hand of the Higheſt.

\V I haue bene mindful of the workes of our Lord: becauſe I wil be
mindful, from the beginning of thy meruelous workes.

\V And I wil meditate in al thy workes: and in thy inuentions I wil be
exerciſed.

%%% o-1167
\V Ô God in the holie is thy way: what God is great as our God? \V thou
art the God that doeſt meruelous thinges.

Thou haſt made thy powre knowne among peoples: \V thou haſt with thine
arme redemed thy people, the children of
\SNote{The progenie of Iacob, receiued and nouriſhed in Ægypt for
Ioſephs ſake, as his adopted children.}
Iacob, and Ioſeph.

\V The waters
\SNote{The read ſea, and Iordan felt thy diuine powre, and obeyed thy
wil.}
ſaw thee ô God, the waters ſaw thee: and they were afrayd, and the
depthes were trubled.

%%% 1277
\V A
\SNote{Noiſe of vvaters meeting after the Iſraelites vvere paſſed,
thunders and lightninges alſo hapened, to the terror of the perſecutors,
though not mentioned in Exodus.}
multitude of the ſounding of waters: the cloudes gaue a voice.

For in deede arrowes doe paſſe: \V the voice of thy thunder in a
wheele.

Thy lightenings shined to the round world: the earth was moued and
trubled.

\V Thy way in the ſea, and thy pathes in many waters: and thy ſteppes
shal not be knowne.

\V
\CNote{\XRef{Exo.~14.}}
Thou haſt conducted thy people as sheepe:
\SNote{By the miniſterie of Moyſes and Aaron.}
in the hand of Moyſes and Aaron.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
