%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 27
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-27


\project douay-rheims


%%% 1192
%%% o-1085
\startChapter[
  title={Psalme 27}
  ]

\PSummary{Dauid
\MNote{An other prayer of Dauid for deliuerie from euils.

The 8.~key.}
prayeth to be defended from the eternal deſtruction of the wicked,
4.~which (by way of imprecation, or conformitie to Gods iuſtice) he
prophecieth. 6.~Feeling by inſpiration that his prayer is heard,
rendereth thankes to God, 9.~and prayeth for al the people.}

\PTitle{A Pſalme to Dauid him ſelfe.}

\NV To thee ô Lord I wil crie, my God
\SNote{Omitte not to comfort me.}
keepe not ſilence from me: leſt at any time thou hold thy peace from me,
%%% 1193
and I shal be like to them that goe downe into the lake.

\V Heare ô Lord the voice of my petition whiles I pray to thee: whiles I
lift vp my handes to thy holie temple.

\V
\SNote{Suffer me not to be ouercome: for God tempteth none to euil.
\XRef{Iac.~1.}}
Draw me not together with ſinners: and with them that worke iniquitie
deſtroy me not.

Which ſpeake peace with their neighbour, but euils in their hartes.

\V
\SNote{The iuſt in zele of iuſtice pray that ſinne may be punished.}
Geue them according to their workes, and according to the wickedneſſe of
their inuentions.

According to the workes of their handes geue vnto them: render them
their retribution.

\V Becauſe they haue
\SNote{Ignorance doth not excuſe when men may and wil not vnderſtand.}
not vnderſtood the workes of our Lord, and in the workes of his handes
thou shalt deſtroy them, and
\SNote{God ſaueth not without our cooperation with his grace.}
not build them vp.

\V Bleſſed be our Lord: becauſe he hath heard the voice of my petition.

\V Our Lord is my helper, and my protectour: in him my hart hath hoped,
and I was holpen.

And
\SNote{Being comforted in ſpirite my bodie is as it were refreshed.}
my flesh florished againe: and
\SNote{Freely and gladly.}
with my wil I wil confeſſe to him.

\V Our Lord is the ſtrength of his people: and he is
\SNote{God protecteth and proſpereth the kings good endeuoures for his
people.}
the protector of
%%% o-1086
the ſaluations of his annoynted.

\V Saue
\SNote{As
\XRef{Pſal.~19.}
and often elſwhere the ſubiectes pray for their Superior: ſo mutually
the ſuperior prayeth for the ſubiectes.}
thy people ô Lord, and bleſſe thine inheritance: and rule them, and
extol them for euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
