%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 135
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-135


\project douay-rheims


%%% 1382
%%% o-1272
\startChapter[
  title={Psalme 135}
  ]

\PSummary{God
\MNote{Gods meruelous workes.

The 2.~key.}
being meruelous in himſelf, 4.~hath shewed his powre and goodnes in his
workes, as wel in general to al the world, 10.~as in particular towards
his elected people.}

\PTitle{Alleluia.}

\NV Confeſſe ye to our Lord
\SNote{VVhoſe goodnes is incomparable.}
becauſe he is good:
\SNote{This ſecond part of euerie verſe, firſt pronounced by the
Prieſtes, confeſſing & praiſing Gods mercie aboue al his workes, was
ſtil repeted by muſicians, or other aſſiſtants, in maner as now in
Litanies (ô Lord deliuer vs) and (VVe beſech thee to heare vs) And as
\L{Gloria Patri &c.} is added in the end of Pſalmes, and our Lords
prayer, and the Angelical ſalutation are often repeted in diuers
Offices, and in the Roſarie.}
becauſe his mercie is for euer.

\V Confeſſe ye to
\SNote{VVhoſe diuine powre created al other thinges.}
the God of goddes: becauſe his mercie is for euer.

\V
\SNote{By this inuitation to praiſe our Lord God, thriſe repeted, we
profeſſe the Bleſſed Trinitie, who is one God in ſubſtance, and three
Diuine Perſons.}
Confeſſe ye to
\SNote{VVhoſe dominion is ouer al the world.}
the Lord of lordes: becauſe his mercie is for euer.

\V Who
\SNote{Only God can do true miracles; to wit, workes aboue the courſe of
al nature.}
onlie doth great meruels: becauſe his mercie is for euer.

\V
\CNote{\XRef{Gen.~1.}}
Who made the heauen
\SNote{In admirable wiſdom.}
in vnderſtanding: becauſe his mercie is for euer.

\V Who eſtablished
\SNote{Made the waters aboue nature to remaine vvithin their bondes, &
not to couer the earth.}
the earth ouer the waters: becauſe his mercie is for euer.

%%% o-1273
\V Who made the great lightes: becauſe his mercie is for euer.

\V The ſunne to rule the day: becauſe his mercie is for euer.

\V The moone, and ſtarres to rule the night: becauſe his mercie is for
euer.

\V
\CNote{\XRef{Exo.~11.}}
Who ſtroke Ægypt with their firſtbegotten: becauſe his mercie is for
euer.

%%% 1383
\V
\CNote{\XRef{Exo.~13.}}
Who brought forth Iſrael out of the middes of them: becauſe his mercie
is for euer.

\V In a mightie hand, and loftie arme: becauſe his mercie is for euer.

\V Who diuided the Read ſea into diuiſions: becauſe his mercie is for
euer.

\V And brought forth Iſrael through the middes therof: becauſe his
mercie is for euer.

\V
\CNote{\XRef{Exo.~14.}}
And he ouerthrew Pharao, and his hoſt in the Readſea: becauſe his mercie
is for euer.

\V
\CNote{\XRef{Num.~21.}}
Who led his people through the deſert: becauſe his mercie is for euer.

\V Who ſtroke great kinges: becauſe his mercie is for euer.

\V And ſlewe ſtrong kinges: becauſe his mercie is for euer.

\V Schon the king of the Amorrheites: becauſe his mercie is for euer.

\V And Og the king of Baſan: becauſe his mercie is for euer.

\V And he gaue their land for an inheritance: becauſe his mercie is for
euer.

\V For an inheritance to Iſrael his ſeruant: becauſe his mercie is for
euer.

\V For in our humiliation he was mindful of vs: becauſe his mercie is
for euer.

\V And he redemed vs from our enemies: becauſe his mercie is for euer.

\V Who geueth foode to
\SNote{Al Gods benignitie, as wel of particular benefites towards his
people, as general to them and al others procedeth from his mercie.}
al fleshe: becauſe his mercie is for euer.

\V Confeſſe ye to the God of heauen: becauſe his mercie is for euer.

Confeſſe ye to the Lord of lordes: becauſe his mercie is for euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
