%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 087
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-087


\project douay-rheims


%%% 1296
%%% o-1187
\startChapter[
  title={Psalme 87}
  ]

\PSummary{A
\MNote{A prayer in long affliction.

The 7.~key.}
faithful perſon ſore and long afflicted, lamentably complaineth, praying
God, 15.~not ſtil to repel him, being leift deſolate, 19.~without al
conſolation of freindes.}

\PTitle{A Canticle of a Pſalme, to the children of Core, vnto the end,
for
\SNote{An inſtrument of muſike apt for lamentable ſongues. Not expreſſed
in the title of anie other Pſalme. Perhaps becauſe this Pſalme
\Fix{mixteth}{mixeth}{possible typo, fixed in other}
not anie conſolation with mourning, as other Pſalmes do, which are alſo
prayers in affliction. As the
\XRef{30.}
\XRef{53.}
\XRef{63.}
\XRef{73.}}
Maheleth
\SNote{Neither is this word in anie other title. It is added here to
admoniſh vs, that as this Pſalme, and ſome others were ſongue by two
quires, one anſwering the other, ſo we muſt in anſwer and imitation of
Chriſt, ſuffer long and great afflictions with patience, ſeing he in his
paſſion was leift without ordinarie cõſolation.}
to anſwer,
\SNote{By ſome interpreted (his bretheren) For Chriſt (ſaith
%%% Cite ?
S.~Auguſtin)
\Fix{vouchſaffeth}{vouchſafeth}{possible typo, fixed in other}
to make them his bretheren, which vnderſtand the myſterie of his Croſſe,
and not only are not aſhamed therof, but alſo faithfully glorie therin.}
of vnderſtanding to Eman the Ezrahite.}

\VV Ô Lord the God of my ſaluation:
\SNote{I haue cried to thee very often both by day, and by night.}
in the day haue I cried, and in the night before thee.

\V Let my prayer enter in thy ſight: incline thine eare to my petition.

\V Becauſe my ſoule is replenished with euils: and
\SNote{I am almoſt dead.}
my life hath approched to hel.

\V I
\SNote{Accounted as dead, and readie to be buried.}
am accounted with them that deſcend into the lake.

I am become as a man without helpe, \V
\SNote{If I were dead, I ſhould be free from theſe
afflictions. Eſpecially it agreeth to Chriſt, who was free, yea of
infinite powre amongſt the dead.}
free among the dead, as the wounded ſleeping in the ſepulchres, of whom
thou art mindful no more: and they are caſt of from thy hand.

\V They haue put me in
\SNote{Myn enimies haue endeuoured not only to bereue me of temporal
life, wherby I ſhould goe into limbus, but alſo to kil my ſoule
ſpiritually, wherby I ſhould deſcend into the lower hel of the damned.}
the lower lake: in the darke places, and in the shadowe of death.

\V Thy
\SNote{Thy iuſt wrath alſo ô God hath excedingly afflicted me.}
furie is confirmed vpon me: and al thy waues thou haſt brought in vpon
me.

\V Thou haſt made my familiars far from me: they haue put me abomination
to themſelues.

I was deliuered and came not forth: \V myne eies languished for
pouertie.

%%% o-1188
I cried to thee ô Lord al the day: I ſtretched out my handes to thee.

\V Wilt thou
\SNote{Ô God deliuer me whiles I am yet liuing, for I may not looke for
extraordinarie, and miraculous helpe, as to be raiſed againe after
death:}
doe meruels to the dead: or shal phiſicians raiſe to life, and they
confeſſe to thee?

\V Shal any
\SNote{when I shal be dead & buried, I can not denounce thy praiſes as
now I can to mortal men.}
in the ſepulchre declare thy mercie, and thy truth in perdition?

\V Shal thy meruelous workes be knowne
\SNote{Much leſſe shal the damned praiſe thee in eternal perdition.}
in darkenes: and thy iuſtice in the land of obliuion.

\V And I ô Lord haue cried to thee: and in the morning shal my praier
preuent thee.

%%% 1298
\V Why doeſt thou ô Lord reiect my prayer: turneſt away thy face from
me?

\V I am poore, and in labours
\SNote{As wel young,}
from my youth: and being
\SNote{as waxing elder I haue bene ſtil afflicted.}
exalted, humbled, and troubled.

\V Thy wrathes haue paſſed vpon me: and thy terrours haue trubled me.

\V They haue compaſſed me as water al the day: they compaſſed me
together.

\V Thou
\SNote{My miſerable eſtate hath alienated al freindes, neighboures, &
acquantances from me.}
haſt made frend, and neighboure far from me: and my familiars becauſe of
miſerie.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
