%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 34
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-34


\project douay-rheims


%%% 1205
%%% o-1097
\startChapter[
  title={Psalme 34}
  ]

\PSummary{Dauid,
\MNote{Chriſts perſecution.

The 5.~key.}
in figure of Chriſt, prophetically by way of inuocating Gods helpe,
forsheweth his perſecution, and the iuſt reuenge vpon his perſecutors,
9.~with praiſe to God, 13.~his charitie towards his cruel aduerſaries,
17.~whom neuertheles God punisheth, 20.~for pretending peace in wordes,
and in fact perſecuting, 23.~rendering to al as they deſerue.}

\PTitle{To
\SNote{Dauid ſignifieth beloued,
\Fix{deſiderable,}{deſirable,}{obvious typo, fixed in other}
or ſtrong of hand, that is Chriſt, aboue al beloued of God, deſired of man,
the ſtrong conquerour of death and hel.
\Cite{S.~Auguſtin in hunc locum.}}
Dauid himſelf.}

\NV Iudge
\SNote{By way of imprecation he prophecieth that God wil ouerthrow the
perſecutors of Chriſt, and of Chriſtians.}
ô Lord them that hurt me: ouerthrow them that impugne me.

\V Take
\SNote{Offenſiue.}
armour and
\SNote{Defenſiue.}
shield: and riſe vp to helpe me.

\V Bring forth the ſword, and
\SNote{Preoccupate and preuent the malice of the perſecutor.}
shut vp againſt them, that perſecute me: ſay to my ſoule: I am thy
ſaluation.

\V Let them
\SNote{Such a punishment and confuſion shal fal in the end vpon al the
malicious, after that the iuſt shal haue ouercome tribulations.}
be confounded & ashamed, that ſeeke my ſoule.

Let them be turned backward, and be confounded that thinke euil
%%% o-1098
againſt me.

\V Be they made as duſt before the face of the winde; and the angel of
our Lord ſtraictning them.

\V Let their way be made darkeneſſe and ſlippernes: and the angel of our
Lord purſewing them.

\V Becauſe they haue hid the deſtruction of their ſnare for me without
cauſe: in vaine haue they vpbrayded my ſoule.

\V Let the ſnare which he knoweth not, come on him; and the net, which
he hath hid, catch him: and let him fal into the verie ſame ſnare.

\V But
\SNote{In the time of truble, in hope;}
my ſoule shal reioyce in our Lord: and shal be delighted vpon
\SNote{after deliuerie, in eternal ſaluation.}
his ſaluation.

\V Al my bones shal ſay: Lord, who is like to thee?

%%% 1206
Deliuering the needie from the hand of them that are ſtronger then he:
the needie and poore from them that ſpoile him.

\V
\SNote{Such falſe witneſſes did riſe againſt Chriſt.
\XRef{Mat.~26.}}
Vniuſt witneſſes ryſing vp, asked me things that
\SNote{Thinges that were not; for God himſelfe, that knoweth al thinges
knoweth not that which neither was, is, nor can be.}
I knew not.

\V They repayed me euil things for good,
\SNote{They made priuation of my life, verified in Chriſt, not in Dauid,
for he was killed by his enemies.}
ſterilitie to my ſoule.

\V But I when they were trubleſome to me,
\SNote{Al our Sauiours life was penance for others, needing none for
himſelfe.}
did put on cloth of heare.

I humbled my ſoule in faſting: and my prayer shal be turned into my
boſome.

\V As a
\SNote{Chriſt the good Samaritane that releeued the wounded man.
\XRef{Luc.~10.}}
neighbour, as our brother, ſo did I pleaſe: as mourning and ſorowful ſo
was I humbled.

\V And they
%%% !!! This SNote encompasses the next two also.
\SNote{Al this was fulfilled according to the letter in our Sauiours
paſſion.}
reioyced againſt me, and
%%% \SNote{}
came together:
%%% \SNote{}
ſcourges were gathered together vpon me, and I was
\SNote{Our Lord knoweth not anie iuſt cauſe, why the Iewes ſo perſecuted
him, for they had no iuſt cauſe but meere malice.}
ignorant.

\V They were diſſipated, and not compunct, they tempted me, they ſcorned
me with ſcorning: they gnashed vpon me with their teeth.

\V Lord
\SNote{As
\XRef{Pſal.~21.}
and
\XRef{Mat.~27.}
God why haſt thou forſaken me? not deliuered me from temporal death, nor
yelded me ſuch conſolation, as thou gaueſt other Sainctes in their
agonies?}
when wilt thou regard?
\SNote{A prophecie of Chriſts reſurrection.}
reſtore thou my ſoule from their malignitie, myne only one from the
lions.

\V I wil confeſſe to thee in
\SNote{Prophecie of the Catholique Church. as
\XRef{Pſal.~21.}}
the great Church, in a graue people I wil prayſe thee.

\V Let them not reioyce ouer me that are my aduerſaries vniuſtly:
\SNote{This place is applied by our Sauiour to himſelfe.
\XRef{Ioan.~15.}}
that hate me without cauſe, and twinckle with the eies.

%%% o-1099
\V Becauſe they ſpake in deede
\SNote{The Phariſes and Herodians ſaid: Maſter we know that thou art a
true ſpeaker &c. meaning to intrappe him with treaſon.
\XRef{Mat.~22.}}
peaceably to me: and in the anger of the earth ſpeaking they meant
guiles.

\V And they
\SNote{The ſame Phariſes and Prieſtes iudged him worthy of death, and
procured the people to crie: Crucifie him, crucifie him.}
opened their mouth awide vpon me: they ſaid: Wel, wel, our eies haue
ſeene.

%%% 1207
\V Thou haſt ſene ô Lord, keepe not ſilence:
\SNote{Againe his Reſurrection is prophecied.}
Lord depart not from me.

\V Ariſe and attend to my iudgement: my God, and my Lord vnto my cauſe.

\V Iudge me according to thy iuſtice ô Lord my God, and let them not
reioyce ouer me.

\V Let them not ſay in their hartes: Wel, wel, to our ſoule: neither let
them ſay: We haue deuoured him.

\V Let
\SNote{At the day of iudgement the wicked shal receiue ſentance of
damnation,}
them blush and be ashamed together, that reioyce at my euils.

Let them be clothed with confuſion and shame; that ſpeake great things
vpon me.

\V Let
\SNote{the bleſſed of eternal glorie.}
them reioyce and be glad, that wil my iuſtice: and let them ſay alwayes:
Our Lord be magnified, that wil the peace of his ſeruant.

\V And my tongue shal meditate thy iuſtice, thy prayſe al the day.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
