%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 023
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-023


\project douay-rheims


%%% 1186
%%% o-1079
\startChapter[
  title={Psalme 23}
  ]

\PSummary{Chriſt
\MNote{Chriſt Lord of al the world.

The 5.~key.}
is Lord of the whole earth, being Creatour, and Redemer of man. 3.~Good
life (with faith in him) is the way to heauen, 7.~whither Chriſt
aſcending with triumph Angels admire him.}

\PTitle{The
\SNote{Chriſt riſing from death the firſt day of the weeke, had al powre
geuen him in heauen and in earth.
\XRef{Mat.~28.}}
firſt of the Sabbath, the Pſalme of Dauid.}

%%% 1187
\NV The earth is our Lordes, and
\SNote{Not only the ſoile it ſelfe but al the fruict, and al that dwel
therin are Gods.}
the fulnes therof: the round world, and al that dwel therein.

\V Becauſe he hath founded it vpon the ſeas: and vpon the riuers hath
prepared it.

\V Who
\SNote{Though Chriſt created and redeemed al, yet only the iuſt shal
inherite heauen.}
shal aſcend into the mount of our Lord? or who shal ſtand in his holie
place?

\V The innocent of handes, and of cleane hart, that hath not taken his
ſoule
\SNote{Not occupied himſelfe in vaine and vnprofitable thinges, but in
commendable workes.}
in vayne, nor ſworne to his neighbour in guile.

\V He shal receiue bleſſing of our Lord: and
\SNote{Gods mercy goeth before iuſtification: iuſt workes folow, and ſo
glorie is the reward of al.}
mercie of God his Sauiour.

\V This is
\SNote{This ſorte of people thus ſeruing God, shal receiue euerlaſting
bliſſe.}
the generation of them that ſeeke him, of them that ſeeke the face of
the God of Iacob.

\V Lift vp your gates
\SNote{The prophet contemplating in ſpirite Chriſts Aſcenſion, inuiteth
Angels to receiue him, and (by proſopopeia) ſpeaketh alſo to the gates
of heauen, by which he is to enter.}
ye princes, and be ye lifted vp ô eternal gates: and the king of glorie
shal enter in.

\V
\SNote{Angels anſwer admiring, & demanding as in a dialogue, how Chriſt
is become ſo glorious!}
Who is this king of glorie? Our Lord
\SNote{The Prophet anſwereth that Chriſt by his powre hath ouercome al
enemies in battel.}
ſtrong & mightie, our Lord mightie in battel.

\V
\SNote{Againe he willeth Angels to open the gates, and biddeth the gates
to enlarge them ſelues.}
Lift vp your gates ye princes, and be ye lifted vp ô eternal gates: and
the king of glorie shal enter in.

%%% !!! SNote should go before 'Who'
Who
\SNote{The Angels demand as before:}
is this king of glorie?
\SNote{the Prophet anſwereth: that Chriſt is Lord alſo of Angels, and al
heauenlie powres vnder God.}
The Lord of powers he is the king of glorie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
