%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-02


\project douay-rheims


%%% 1151
%%% o-1043
\startChapter[
  title={Psalme 2}
  ]

\Summary{Chriſts
\MNote{Of Chriſt, the 5.~key. Alſo of his Church, the 6.~key.}
glorie, the world repining therat, 4.~shal be
  propagated in al the world. 7.~His diuine powre as wel ſpiritual in
  conuerting mens hartes, as external, in ſeuere iuſtice, is
  prophecied.}

VVhy
\CNote{\XRef{Act.~4.}}
did the
\SNote{Both Gẽtiles,}
Gentiles rage, and
\SNote{and Iewes ſtriue in vaine againſt Chriſt.}
peoples meditate vaine things?

\V The
%%% !!! This LNote not marked in either.
\LNote{Kinges and Princes againſt Chriſt.}{VVhen
\MNote{Perſecution can not hinder the glorie of Chriſt.}
Chriſt, and his Apoſtles preached the Goſpel, both Iewes and Gentiles,
with their Princes, Kinges, and Emperors moſt furiouſly reſiſted, but al
in vaine. For they could not hinder the wil and powre of God. But the
more they perſecuted, the more was increaſed the zele, and number of
Chriſtians.}
\SNote{Pilate and Herod.}
kings of the earth ſtood vp, and the
\SNote{Annas and Caiphas.}
princes came together in one againſt our Lord, and againſt his Chriſt.

\V Let
\SNote{The voice of wicked men,}
vs breake their bondes a ſunder: and let vs caſt away their
\SNote{& eſpecially libertines ſtriuing to shake of al diſcipline.}
yoke from vs.

\V He that dwelleth in the heauens, shal
\SNote{God for al this wil turne the hartes of manie.}
laugh at them: and our Lord shal ſcorne them.

\V Then shal he ſpeake to them in his
\SNote{ſeuerly reprehend,}
wrath, & in his
\SNote{and iuſtly punish the obſtinate.}
furie he shal truble them.

\V But
\SNote{Chriſt shal reigne in his kingdome the Church.}
I am appoynted king by him ouer Sion his holie hil, preaching his
precept.

\V
\CNote{\XRef{Act.~13.}
\XRef{Heb.~1.}
&
\XRef{5.}}
The
\SNote{God the Father ſpeaketh to}
Lord ſaid to me: Thou art my
\SNote{God the Sonne.}
Sonne, I this day haue begotten thee.

%%% o-1044
\V Aske of me, and I wil geue
\SNote{Chriſt as man hath the Church for his inheritance.}
thee 
%%% !!! This LNote not marked in either.
\LNote{The gentiles thyn inheritance.}{By
\MNote{The Church neuer faileth.}
this promiſe of God to Chriſt,
\CNote{\Cite{in hunc Pſal.}}
S.~Auguſtin conuinced the Donatiſtes; &
in them the Proteſtantes, that ſay, the Church of Chriſt failed, and
became ſmal, or inuiſible: as though Chriſt the Sonne of God, could
ſometimes loſe his inheritance, which is the Catholique Church, gethered
of the Gentiles, and \Emph{has poſſeſsion} extended \Emph{to the endes
of the earth}.}
the Gentiles, for thyne inheritance, and thy poſſeſſion
\SNote{Spred through the whole world.}
the endes of the earth.

%%% 1152
\V Thou 
\CNote{\XRef{Apoc.~2.}
&
\XRef{19.}}
shalt rule them
%%% Only marked in Heretikes, and not used.
%%% \LNote{}{}
in
\SNote{Gods inflexible wil & powre.}
a rod of yron, and
\LNote{As a potters veſſel.}{If
\MNote{Cuſtome in ſinne more hardly cured.}
a potters veſſel (ſaith S.~Ierom
\Cite{in hunc Pſal. tom.~8.})
be broken whiles it is ſoft, it may eaſily be repared, but after it is
hard, it can not be made whole againe. So ſinners are more eaſily
reſtored to grace, shortly repenting, then long obdurate: yet that which
is vnpoſſible to man, is poſſible to God.
\XRef{Mat.~10.}
\Emph{For as clay in the potters hand, ſo are you in my hand, ſaith our
Lord.}
\XRef{Iere.~18.}}
as a potters veſſel thou shalt breake them in peeces.

\V And
\SNote{A prophecie that kinges shal be conuerted, and ſubmit themſelues
to Chriſts diſcipline.}
now
\LNote{Ye kinges vnderſtand.}{Not
\MNote{Kinges conuerted to Chriſtianitie.}
onlie innumerable other people of al nations, but alſo after a while,
Kinges and Emperors beleued in Chriſt. And ſuch as at firſt perſecuted,
became moſt Chriſtian, Catholique, Defenders of the faith.}
ye kings vnderſtand:
\LNote{Take inſtruction you that iudge the earth.}{Petilianus,
\MNote{Defend Catholiques and puniſh heretikes.}
Gaudentius, & other Donatiſtes inueyghing againſt Chriſtian Kinges, for
punishing heretikes, moſt falſly auoched, that Chriſtianitie neuer found
kinges but inuious, enimies, and perſecuters. To whom 
\CNote{\Cite{Ad Petil. li.~2. c.~91.}
\Cite{Ad Gaud. li.~2. c.~26.}}
S.~Auguſtin anſwereth in ſeueral bookes, that Chriſtian Kinges, and
Princes are not enemies to Chriſtianitie: but are enemies to heretikes,
the rebelles of Chriſt, and his Church. For according to this prophecie
of king Dauid, Chriſtian \Emph{kinges are inſtructed}, and know it is
their dutie, in the ſeruice of God, to defend the Church againſt
Heretikes, and other Infidelles.
\MNote{Apoſtataes fauoure heretikes and ſchiſmatikes, becauſe they hate
al Chriſtians.}
And it is the propertie of Apoſtataes to fauour heretikes. So good
Conſtantin the great maintained Catholique vnitie; and Iulian the
Apoſtata to make greater diuiſion, tooke Churches from Catholiques, and
gaue them to Donatiſtes, to nourish diſſention, and ſo to ouerthrow al
Chriſtians. But God ſtil protecteth the true Church, againſt al ſuch
ſuttle, and malicious deuiles: \Emph{becauſe it is Chriſts inheritance.}}
take inſtruction you that iudge the earth.

\V Serue our Lord in
\SNote{None is ſecure before death.}
feare: and
\LNote{Reioyce vvith trembling.}{Gods
\MNote{Ioy and feare.}
ſeruice is tempered with two affections: with ioy, in conſideration of
his goodnes, mercie, & meeknes; and with feare, in reſpect of his
iuſtice, and ſeuere iudgement. The one is a remedie againſt deſperation:
the other againſt preſumption.}
reioyce to him with trembling.

\V Apprehend diſcipline
\SNote{Some fal from the way of ſaluation.}
leſt ſometime our Lord be wrath, and you perish out of the iuſt way.

\V When his wrath
\SNote{God wil iudge iuſtly in the end of this short life both the euil
and good.

So this Pſalme concludeth with the ninth key.}
shal burne in short time, bleſſed are al, that truſt in him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
