%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 26
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-26


\project douay-rheims


%%% 1191
%%% o-1083
\startChapter[
  title={Psalme 26}
  ]

\PSummary{Dauid
\MNote{An other confident prayer of Dauid in tribulation.

The 3.~key.}
being in great diſtreſſe through perſecution, and hauing aſſured
confidence in God, deſcribeth the great ſecuritie of Gods protection,
7.~sheweth the ſame experienced in him ſelf, 12.~prayeth for continuance
therof, 13.~and incorageth his owne ſoule, in hope of life euerlaſting,
to perſeuer in vertue.}

\PTitle{The Pſalme of Dauid
\SNote{Before his ſecond annointing, as is probable.
\XRef{2.~Reg.~2.}}
before he was annoynted.}

\NV Our Lord is
\SNote{Againſt ignorance God illuminateth his ſeruantes.}
my illumination, and
\SNote{Againſt infirmitie he geueth ſtreingth.}
my ſaluation,
\SNote{So he nedeth not to feare anie mans malice, ſuteltie, nor force.
\XRef{Luc.~21. v.~15.}}
whom
%%% o-1084
shal I feare?

Our Lord is the protectour of my life, of whom shal I be afrayd?

\V Whiles the harmeful approch vpon me, to eate my flesh. Mine enemies
that truble me, them ſelues are weakened and are fallen.

\V If campes ſtand together againſt me, my hart shal not feare.

If battel riſe vp againſt me, in this wil I hope.

\V One thing I haue asked of our Lord, this wil I ſeeke for, that I
\SNote{How ſpecial a benefite Dauid eſtemed it to be in the Catholique
Church the only true houſe of God!}
may dwel in the houſe of our Lord al the dayes of my life.

That I may ſee the pleaſentnes of our Lord, and viſite his temple.

\V Becauſe he hath
\SNote{Albeit the ſpiritual or carnal enemie ſeke to ouerthrow me, yet I
am ſecure in the Catholique Church.}
hid me in his tabernacle: in the day of euils he hath protected me, in
\SNote{God either ſuffereth not the enemie to find his ſeruant,}
the ſecrete of his tabernacle.

\V In
\SNote{or not to be able to hurt him ſpritually.}
a rocke he hath exalted me: and now he hath
\SNote{Whẽ a martyr of confeſſour dieth, then he getteth the victorie
againſt the perſecutors.}
exalted my head ouer mine enemies.

%%% 1192
I
\SNote{Diligently recounting al thy benefites, I render thankes by
ſacrifice and praiſe,}
haue gone round about, and haue immolated in his tabernacle an hoſt of
iubilation: I
\SNote{not only in hart, but alſo ſinging with loud voice and
inſtrument.}
wil ſing, and ſay a Pſalme to our Lord.

\V Heare ô Lord my voice, wherwith I haue cried to thee: haue mercie on
me, and heare me.

\V My
\SNote{In my inward ſincere cogitation I deſire, and ſeeke that I may
ſee thee,}
hart hath ſayd to thee, my face hath ſought thee out:
\SNote{face to face.
\XRef{1.~Cor.~13. v.~12.}}
thy face ô Lord I wil ſeeke.

\V Turne
\SNote{In the meane time ô Lord, grant me thy fauour,}
not away thy face from me:
\SNote{leaue me not though thou be angrie with me.}
decline not in wrath from thy ſeruant.

Be
\SNote{He ſpeaketh in the perſon of orphanes.}
thou my helper: forſake me not, neither deſpiſe me ô God my Sauiour.

\V Becauſe
\SNote{Though carnal parentes forſake the iuſt man in tribulation, yet
God hath then moſt ſpecial care of him.}
my father and my mother haue forſaken me: but our Lord hath taken me.

\V Geue me
\SNote{Eſtabliſh my hart in thy law.}
a law ô Lord in thy way:
\SNote{Conſerue me in the right way, which thou haſt already taught me,
and it is the more neceſſarie, becauſe myne enemies labour to peruert
me.}
and direct me in the right path, becauſe of mine enemies.

\V Deliuer me not into
\SNote{The willes.}
the ſoules of them that truble me: becauſe
\SNote{Falſe witneſſes accuſed Dauid, others accuſed Chriſt:
\XRef{Mat.~16.}
others do ſtil accuſe the iuſt.
\XRef{Mat.~5.}}
vniuſt witneſſes haue riſen vp againſt me, and iniquity hath lyed
\SNote{The wicked pleaſe themſelues in lying, but the chief hurt finally
turneth vpon themſelues.}
%%% o-1085
to itſelfe.

\V I
\SNote{The prophet and al iuſt men are comforted by God, and hope of
reward in heauen.}
beleue to ſee the good things of our Lord, in the land of the liuing.

\V
\SNote{The iuſt exhorteth his owne ſoule to patience,}
Expect our Lord, doe manfully: and
\CNote{\XRef{Pſal.~30.}}
\SNote{fortitude,}
let thy hart take courage, and
\SNote{and longanimitie.}
expect thou our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
