%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 45
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-45


\project douay-rheims


%%% 1226
%%% o-1117
\startChapter[
  title={Psalme 45}
  ]

\PSummary{The
\MNote{The Church proſpereth alſo in perſecution.

The 6.~key.}
Church in perſecution acknowledgeth Gods perpetual defence, 5.~making
her therby more glorious, 10.~ſometimes granting reſt (11.~God himſelf
checking the perſecuters) and euer protecting her.}

\PTitle{Vnto
\SNote{Belonging to the Church of Chriſt.}
the end, to the ſonnes of Core, for
\SNote{As wel the cauſe, vvhy God ſuffereth his Church to be perſecuted,
as his aſſured protection in difficulties, are hidden ſecretes to the
world.}
the ſecretes.}

\VV Our God is a refuge
\SNote{Al refuge is not ſecure for one man is not able alwayes to
defend an other: but God is a ſure and ſtrong refuge,}
and ſtrength: an
\SNote{euer able and in conuenient time vviling to helpe.}
helper in tribulations, which
\SNote{This whole vvorld is ful of tribulations, but the Church ſuffered
the greateſt in the firſt perſecutions, & ſhal ſuffer as great in the
time of Antichriſt. Engliſh Catholiques ſuffer moſt of al nations in
this age, and can not be ſuppreſſed, but ſtil increaſe in number and
fortitude.}
haue found vs excedingly.

\V Therfore wil we
\SNote{Therfore al Catholiques may aſſuredly know, that the whole Church
can not faile}
not feare when
\SNote{though very manie, as now in England,}
the earth shal be trubled: and
\SNote{and very eminent perſons, as ſome noblemen, and ſome Prieſtes
haue reuolted, yet al vvil not.}
mountaines tranſported into the hart of the ſea.

%%% o-1118
\V Their waters haue ſounded, and were trubled: the mountaines were
trubled in his ſtrength.

\V The violence of the riuer
\SNote{Such bad examples make the good to recollect themſelues more
diligently, and to reioyce in Gods grace, by which they ſtand faſt.}
maketh the citie of God ioyful: the Higheſt hath ſanctified his
tabernacle.

\V God is in the middes therof, it shal not be moued: God wil helpe it
\SNote{Before the heate of perſecution ſhal inuade al, for the elect the
dayes of tribulation are ſhortned.}
in the morning early.

%%% 1227
\V Nations are trubled, and
\SNote{Sometimes one nation of kingdome rebelleth againſt the Church,
but can not deſtroy it.}
kingdomes are inclined: he gaue
\SNote{By the ſpirite of Chriſt, Antichriſt, and al his members ſhal be
deſtroyed.}
his voice, the earth was moued.

\V The Lord of hoſtes is with vs: the God of Iacob is our defender.

\V Come ye, and ſee the workes of our Lord, what wonders he hath put
vpon the earth: \V
\SNote{The Church ſometimes hath great peace, and tranquilitie.}
taking away warres euen vnto the end of the earth.

He shal deſtroy bow, & breake weapons: and shields he shal burne with
fire.

\V
\SNote{God himſelfe reſtrayneth the wicked, ſuddainly abating their
furie, or cutting of their forces.}
Be quiet, and ſee that I am God: I shal be exalted among the gentiles,
and I shal be exalted in the earth.

\V The Lord of hoſtes is with vs: the God of Iacob is our defender.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
