%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 085
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-085


\project douay-rheims


%%% 1294
%%% o-1184
\startChapter[
  title={Psalme 85}
  ]

\PSummary{In
\MNote{A prayer for continual grace.

The 7.~key.}
conſideration of his owne imperfections, the royal prophet, or other
faithful perſon prayeth God, 5.~according to his mercie and goodnes,
9.~shewed in conuerting Gentiles, 13.~and in deliuering the ſupplicant
him ſelf from the ſtate of damnation, 16.~that he wil ſtil direct and
defend him againſt al enimies.}

\PTitle{A praier
\SNote{A forme of prayer for king Dauid, and for anie faithful perſon.}
to Dauid him ſelfe.}

\NV Incline thine eare ô Lord, and
\SNote{There be ſundrie iuſt cauſes which moue God to heare our prayers:}
heare me:
\SNote{firſt our neceſſitie requireth Gods helpe.}
becauſe I am needie and poore.

\V Keepe my ſoule, becauſe
\SNote{Secondly becauſe we profeſſe, and promiſe to lead a holie life.}
I am holie: ſaue thy ſeruant my God, that
\SNote{Thirdly, becauſe we truſt and hope in God.}
hopeth in thee.

\V Haue mercie on me ô Lord, becauſe I haue cried to thee
\SNote{Fourtly, becauſe we perſeuere in prayer.}
al the day: \V make ioyful the ſoule of thy ſeruant, becauſe to thee ô
Lord haue I
\SNote{Fiftly, if we pray with attention of mind.}
lifted vp my ſoule.

%%% 1295
\V Becauſe thou ô Lord art
\SNote{Sixtly becauſe God of his owne nature is benigne, readie to
beſtow benefites.}
ſwete, and
\SNote{Seuently, he is meeke to remitte offences.}
milde: and
\SNote{Eightly, he is merciful to mitigate the puniſhment, to thoſe that
make recourſe vnto him.}
of much mercie to al that inuocate thee.

%%% o-1185
\V
%%% !!! This SNote realy at end of previous verse.
\SNote{For theſe cauſes we pray as foloweth.}
Receiue ô Lord my praier with thine ears: and attend to the voice of my
petition.

\V In the day of my tribulation I haue called to thee: becauſe thou haſt
heard me.

\V There is not the like to thee amongſt goddes ô Lord: and there is not
according to thy workes.

\V
\SNote{Vocation of Gentiles.}
Al nations whatſoeuer thou haſt made, shal
\SNote{They ſhal come by faith,}
come, and shal adore before thee ô Lord: and they shal
\SNote{and glorifie God by good workes.
\XRef{Mat.~5. v.~17.}}
glorifie thy name.

\V Becauſe thou art great and doing meruelous thinges: thou onlie art
God.

\V Conduct me ô Lord in thy way, and I wil walke in thy truth: let my
hart reioyce that it may feare thy name.

\V I wil
\SNote{Confeſſion of praiſe.}
confeſſe to thee ô Lord my God with al my hart, and wil glorifie thy
name for euer:

\V Becauſe thy mercie is great vpon me: and thou haſt deliuered my ſoule
out of
\SNote{From the ſtate of eternal damnation.}
the lower hel.

\V Ô God the wicked are riſen vp vpon me, and the ſynagogue of the
mightie haue ſought my ſoule: and they haue not ſet thee in their ſight.

\V And thou ô Lord the God of compaſſion and merciful, patient, and of
much mercie, and
\SNote{In performing al promiſes.}
true.

\V Haue reſpect to me, and haue mercie on me,
\SNote{A digreſſion (vſual to prophetes) of Chriſts Empyre and Kingdom
the Church, geuen to him, being the ſonne of an immaculate virgin, the
handmaide of God.}
giue thine empire to thy ſeruant: and ſaue the ſonne of thy handmaide.

\V Make with me
\SNote{The chief and principal ſigne of Chriſts, and his Churches
glorie, is his Reſurrection, præfigured in Ionas.}
a ſigne vnto God, that they may ſee which hate me, and may be
confounded:
\SNote{VVherby al enimies are confounded, either to their conuerſion, or
to eternal damnation. See
%%% !!! vpon this place?
\Cite{S.~Auguſtin.}}
becauſe thou ô Lord haſt holpen me.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
