%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 061
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-061


\project douay-rheims


%%% 1249
%%% o-1140
\startChapter[
  title={Psalme 61}
  ]

\PSummary{A
\MNote{Exhortation to good life, in reſpect of reward, or punishment.

The 7.~key.}
iuſt man encoregeth his owne ſoule to ſerue God in ſincere
  humilitie, 9.~exhorteth alſo al others to truſt in God, not in falſe
  and worldlie pollicie, or wealth, becauſe Gods powre and mercie wil
  render to euerie one as they deſerue.}

\PTitle{Vnto the end, for
\SNote{Directed to Idithun, one of the maſters of muſike, to ſing it, or
to make tune for it.}
\Fix{Iduthun}{Idithun}{obvious typo, fixed in other}
a Pſalme of Dauid.}

\VV Shal
\SNote{The wicked
\Fix{threating}{treating}{possible typo, fixed in other}
to ruinate others, Dauid, or anie iuſt man, feareth them not, becauſe
his ſoule is ſubiect to God.}
not my ſoule be ſubiect to God? for of him is my ſaluation.

\V For he is my God, and my ſauiour: my receiuer
\SNote{Therfore I firmely purpoſe neuer to be moued from God.}
I shal be moued no more.

\V
\SNote{In vaine do you myn aduerſaries ſtil aſſault me,}
How long ſet you violently vpon a man:
\SNote{though ye be al confederate to kil me;}
you al doe kil: as it were vpon
\SNote{ſuppoſing me to be like a ruinous, or shaken wal, that is eaſily
throwne downe.}
a wal, that is leaning, and a wal shaken.

\V But
\SNote{They thinck ſtil to depriue me of my reward, the price of my
laboures and merites,}
yet they thought to repel my price,
\SNote{but I runne ſo much more diligently, as thirſting after
righteouſnes in this life, and glorie in the next, to finish my courſe.}
I ranne in thirſt: they
\SNote{A moſt dangerous tentation, when after threates and crueltie,
perſecutors endeuour, by ſwete wordes, and promiſes to
\Fix{perſwaed}{perſwade}{likely typo, fixed in other}
the iuſt to fal into ſinne.}
bleſſed with their mouth, and curſed with their hart.

%%% o-1141
\V But yet my ſoule be thou ſubiect to God: becauſe my patience is from
him.

\V Becauſe he is my God, and my ſauiour: my helper
\SNote{I reſolutly purpoſe not to yeld to anie tentations.}
I shal not remoue.

\V In God is my ſaluation, and my glorie: the God of my helpe, and my
hope is in God.

\V Hope in him
\SNote{Gods faithful ſeruantes are not only conſtant themſelues, but
alſo exhorte and perſwade al others, as much as in them lieth, to ſerue
God and truſt in him.}
al ye the congregation of people: Powre out your harts before him, God
is our helper for euer.

%%% 1250
\V But yet the children of men are vaine, the children of men are
\SNote{Vſing falſe weightes they defraud one an other.}
liers in balances: that they may deceiue by vanitie together.

\V Hope not in iniquitie, and couete not robberies: if riches abound ſet
not your hart vpon them.

\V
\SNote{God hauing
\Fix{nce}{once}{obvious typo, fixed in other}
ſpoken it is moſt aſſured.}
Once hath God ſpoken, theſe
\SNote{Two eſpecial attributes of God.}
two things haue I heard:

\V 
\CNote{\XRef{Mat.~16.}
\XRef{Rom.~2.}
\XRef{1.~Cor.~3.}
\XRef{Gal.~6.}}
That
\SNote{God is Omnipotent, ſo that he can both reward, and puniſh
infinitly;}
powre is Gods, and
\SNote{and Merciful, that he is readie to receiue al ſinners into his
fauour, if they wil repent and turne vnto him.}
mercie ô Lord is to thee: becauſe thou wilt render to euery one
according to his workes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
