%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 137
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-137


\project douay-rheims


%%% 1385
%%% o-1275
\startChapter[
  title={Psalme 137}
  ]

\PSummary{The
\MNote{Thankes to God for benefites.

The 7.~key.}
whole Church, or anie iuſt perſon, rendreth thankes to God for his
benefites, 4.~praying that al kinges and kingdomes may do the ſame,
\Fix{4.~becauſe}{6.~becauſe}{obvious typo, same in both}
God being high reſpecteth, and aduanceth the humble.}

\PTitle{To
\SNote{This Pſalme is a fitte forme of thankes for Dauid, or anie other
ſeruant of God.}
Dauid him ſelf.}

\NV I wil
\SNote{I vvil render thankes and praiſe.}
confeſſe to thee ô Lord in my whole hart:
\SNote{God euer heareth the prayer that is rightly made, for vvhich the
ſupplicant is therfore to geue thankes.}
becauſe thou haſt heard the wordes of my mouth.

In
\SNote{Angels are preſent vvhere the faithful pray,
\CNote{\XRef{1.~Cor.~11. v.~10.}}
obſerue our prayers,
and offer them to God, if they be ſincere, as the prayers of
\CNote{\XRef{Tob.~12.}}
Tobias, and
\CNote{\XRef{Act.~10.}}
Cornelius.}
the ſight of Angels I wil ſing to thee: \V I wil adore
\SNote{VVhen there vvas not acceſſe to the Temple, yet
\CNote{Pſal.~24. v.~10.}
the Ievves praying in captiuitie, turned themſelues tovvards the
Temple.}
toward thy holie temple, and wil confeſſe to thy name.

For thy mercie, and thy truth: becauſe thou haſt magnified aboue euerie
thing
\SNote{The Name and Maieſtie of \Sc{God}, The \Sc{VVord} (which is the
Sonne of God) alſo the name of \Sc{Iesvs}, is magnified aboue al names,
or temporal thinges.}
thy holie name.

\V In what day ſoeuer I shal inuocate thee, heare me: thou wilt
multiplie ſtrength in my ſoule.

\V Let
\SNote{Sovvner or later Kinges and Princes of al kingdomes and nations,
haue bene or ſhal be conuerted to Chriſt.}
al the kinges of the earth ô Lord confeſſe to thee: becauſe they haue
heard al the wordes of thy mouth:

\V And let them ſing in the wayes of our Lord: becauſe great is the
glorie of our Lord.

\V Becauſe our Lord is high, and he beholdeth low thinges: and high
thinges he knoweth
\SNote{God knovveth proud men, not as his freindes or ſeruants, but
farre of, as ſtrangers and enimies.}
far of.

\V If I shal walke in the middes of tribulation, thou wilt quicken me:
and vpon the wrath of mine enimies thou haſt extended thy hand, and thy
right hand hath ſaued me.

%%% o-1276
\V Our Lord wil repay for me: ô Lord thy mercie is for euer: deſpiſe not
the workes of thy handes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
