%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 140
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-140


\project douay-rheims


%%% 1389
%%% o-1279
\startChapter[
  title={Psalme 140}
  ]

\PSummary{The
\MNote{The Church prayeth and preuaileth.

The 6.~key.}
Church prayeth that her children may auoide ſinful wordes, 4.~not make
excuſes of ſinnes committed, not communicate with others in ſinne, nor
to harken to flatterers: 6.~but to pray that they may amend (the
Pſalmiſt by the way prophecieth that manie shal be conuerted) 8.~though
ſometimes perſecution be great, the Church faileth not.}

\PTitle{A Pſalme of Dauid.}

\NV Lord I haue cried to thee, heare me: attend to my voice, when I shal
crie to thee.

\V 
\SNote{In theſe wordes the Church prayeth, offering incenſe in her
ſolemne Offices.}
Let my prayer be directed as
\SNote{As the fume of incenſe is ſwete and aſcendeth vpwards: ſo the
Church prayeth that her petitions may be gratful, and aſcend to God.}
incenſe in thy ſight: the eleuation of my handes as
\SNote{In the old teſtament morning ſacrifice was rather more ſolemne,
and more frequent, but the Prophet ſemeth to allude vnto Chriſts
Sacrifice, which he was to offer towards euening on the Croſſe: and the
ſame alſo in vnbloudie maner, the euening before his Paſſion, in the
Euchariſt.}
euening ſacrifice.

\V Set ô Lord
\SNote{A iuſt care and conſideration, what to ſpeake, before the mouth
be opened.}
a watch to my mouth: and
\SNote{Lippes muſt not be alwayes ſtopt, for it is a ſinne ſometimes not
to ſpeake, but as a dore that is to be opened, and ſhut at due ſeaſons:
for example, opened to confeſſe our ſinnes, and Gods truth:}
a doore round about to my lippes.

%%% o-1280
\V Decline not my hart
\SNote{shutte from wordes of malice, and from excuſing ſinnes committed.}
into wordes of malice, to make excuſes in ſinnes.

With men that worke iniquitie: and I wil not communicate with the chiefe
of them.

\V The iuſt shal rebuke me in mercie, and
\SNote{Admonition of the iuſt is a profitable reprehenſion,}
shal reprehend me: but let not the oyle of a ſinner
\SNote{but the fawning of flatterers is pernicious.}
fatte my head.

Becauſe
\SNote{The Church ceaſeth not to pray for ſinners,}
yet alſo my prayer is in their
\SNote{though as yet they take pleaſure in their ſinnes,}
good pleaſures: \V
\SNote{but the chiefeſt of them, which ſeme moſt potent, if they remaine
obſtinate to their death, ſhal then periſh, as men throwne againſt
rockes.}
their iudges are ſwalowed vp ioyned to the rocke.

They
%%% !!! SNote should go before 'They'
\SNote{They ſhal then heare and vnderſtand, that the prayers of the
Church}
shal heare my wordes
\SNote{are effectual, obtaining grace of conſtancie to her children, not
to feare anie perſecution, nor anie kind of death; obtayning alſo
iuſtificatiõ of their cauſe, vvhen the perſecuters ſhal ſee that the
wordes, and doctrin of the Church are true, and preuaile in true
iudgement, as true, modeſt, peaceable, not ſedicious, turbulent, nor
againſt the commonwealth.}
becauſe they haue preuailed: \V
\SNote{As much earth ſticking together is made fructful, by breaking it
into ſmal mould, ſo the children of the Church by perſecution, bring
forth more fruict then before.
\Cite{S.~Aug.}}
as the groſſenes of the earth is broken out vpon the earth.

%%% 1390
Our
\SNote{Some perſecuters are ſo cruel as to rage againſt the bones, and
other reliques of Martyrs,}
bones are diſſipated
\SNote{caſting them into the vileſt places they can, willing if they
could, to throw them into hel,}
nere to hel: \V
\SNote{but the Church and al her members repoſe confidence in God.}
for to thee ô Lord, Lord are mine eies; in thee haue I hoped, take not
away my
\Var{ſoule.}{life.}

\V Keepe me from
\SNote{Therfore ſhe prayeth that her children be not entrapped by
guilful deceiptes,}
the ſnare, which they haue ſet for me: and from
\SNote{nor ouerthrowne by anie ſtumbling blocke caſt in their way.}
the ſcandals of them that worke iniquitie.

\V
\SNote{In the end, al perſecuters and other wicked ſhal be caught in
their ovvne nette of perdition,}
Sinners shal fal in his net:
\SNote{the Church is ſingularly protected euen to the end.}
I am alone vntil I paſſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
