%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 35
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-35


\project douay-rheims


%%% 1207
%%% o-1099
\startChapter[
  title={Psalme 35}
  ]

\PSummary{The
\MNote{Gods prouidence.

The 3.~key.}
prophet deſcribeth the wicked malice of obſtinate ſinners. 6.~Againſt
which he oppoſeth Gods infinite goodnes, 9.~with his prouident mercie
towards the worſt, and iuſt reward of the good; 12.~praying to eſcape
the dangerous gulfe of pride.}

\PTitle{Vnto
\SNote{More ſpecially deſcribing the ſtate of men in the new Teſtament,
then in the old.}
the end, to
\SNote{Inſtruction for Dauid not as a king, or a prophet, but as the
poore ſeruant of God.}
the ſeruant of our Lord Dauid him ſelfe.}

\VV The vniuſt hath ſaid within him ſelfe, that he
\SNote{Wittingly and reſolutely preferring wicked life before vertuous.}
would ſinne: there is no feare of God before his eies.

\V Becauſe he hath done deceitfully in his ſight:
\SNote{God ſo hateth ſinne committed of meere malice, that he commonly
reiecteth ſuch ſinners, and more often offereth new grace to thoſe, that
ſinne of frailitie, or ignorance.}
that his iniquitie may be found vnto hatred.

\V The wordes of his mouth are iniquitie, and guile: he
\SNote{Some ignorance is inuincible whẽ one hath a good wil to lerne,
doing his endeuour to knovv the truth in doctrin, & his dutie in
manners, but can not get knowledge therof, and then he is excuſed before
God, though he erre in opinion, or in fact: others are negligent to
lerne, and their error is groſſe ignorance and is a ſinne, greater of
leſſe, according to the importance of the thing, which they
\Fix{ouhgt}{ought}{obvious typo, fixed in other}
to knovv. Others are more vvilful, deſiring to be ignorant; that they
may ſinne vvith the leſſe remorſe, or repining of their ovvne
conſcience, and that is affected ignorance, and moſt hainous, and odious
ſinne. For which God often leauing them deſtitute of ordinarie grace,
which he geueth to others, they fal into reprobate ſenſe, and into more
horrible ſinnes.}
would not
%%% o-1100
vnderſtand that he might doe wel.

%%% 1208
\V He hath meditated iniquitie in his bed: he hath ſet himſelfe on euery
way not good, and malice he hath not hated.

\V Lord
\SNote{God doth not vtterly ſhut vp his mercie from the moſt wilful &
wicked ſinners, but geueth them ſometimes good motions, and ſufficient
helpe, that they may repent, be iuſtified & ſaued, if they do not
wittingly harden their owne hartes, and ſtil wilfully repel Gods grace.}
thy mercie is in heauen: and
\SNote{For ſo God
\Fix{pomiſeth}{promiſeth}{likely typo, same in both}
(vvho is moſt faithful) that he vvil forgeue ſinners, and receiue them
into his fauoure againe, whenſoeuer they reſoluing to ſerue him, repent
and ceaſe to ſinne.}
thy truth euen to the clowdes.

\V Thy iuſtice as the hilles of God: thy iudgementes are great depth.

Men &
\SNote{Thou ô God that haſt care of al creatures, not only of men but
alſo of brute beaſtes, art euer readie of thy part to ſaue both moderate
men, in vvhom the light of reaſon remaineth, and alſo groſſe ſenſeles
perſons, vvhich are become brutiſh like horſe and mule or other beaſtes.}
beaſtes thou wilt ſaue ô Lord: \V as thou
\SNote{For ſo our Lord multiplieth his mercie.}
haſt multiplied thy mercie ô God.

But
\SNote{Yet with condition that ſenſles or brutiſh men, muſt become
reaſonable men, the children of men, not coltes, whelpes, pigges, &c.}
the children of men,
\SNote{Sinners thus conuerted ſhal not only haue al neceſſaries in this
life, as al liuing creatures haue in this world, but alſo ſhal hope of
ſpiritual, heauenly, & eternal glorie prepared for Angels, and children
of men: as in the verſes folovving.}
shal hope in the couert of thy winges.

\V They shal be inebriated with the plentie of thy houſe: and with the
torrent of thy pleaſure thou shalt make them drinke.

\V Becauſe with thee is the fountaine of life: and in thy light we shal
ſee light.

\V Extend thy mercie to them that know thee, and thy iuſtice to them,
that are of a right hart.

\V Let
\SNote{Leſt anie impediment hinder the obtayning and poſſeſſion of
eternal revvard, the iuſt muſt ſpecially pray not to be infected vvith
pride:}
not the foote of pride come to me: and let not
\SNote{nor be ouercome by the forcible tentations of other ſinners, by
perſvvaſion, nor euil example.}
the hand of a ſinner moue me.

\V There
\SNote{The firſt ſinne, tovvit of diuels, vvas pride, and mans ſinne
vvas by perſvvaſion of the diuel,}
haue they fallen that worke iniquitie: they were expelled,
\SNote{neither of which could eſcape puniſhment.}
neither could they ſtand.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
