%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 101
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-101


\project douay-rheims


%%% 1318
%%% o-1209
\startChapter[
  title={Psalme 101}
  ]

\PSummary{A
\MNote{The firſt penitential Pſalme.

The 7.~key.}
ſinner in affliction of mind prayeth God to deliuer him, 10.~deſolate of
al other helpe. 13.~Conceiueth comforth in Gods eternal goodnes, and
ſingular mercie, in redeming mankind, and propagating the Church.
24.~Prayeth to be made mature in vertue before he dye, that he may liue
with God: 26.~who only and wholly being immutable, eſtablisheth his
ſeruantes for euer.}

\PTitle{The prayer of the poore, when he shal be anxious, and shal make
his petition before our Lord.}

\VV Lord heare my
\SNote{Euerie petition is a prayer,}
prayer: and let my
\SNote{and that which procedeth from more feruent affection, is called a
crie, though it burſt not out into clamoure, nor perhaps into anie voice
at al. For God ſaide to Moyſes, praying in mere ſilence, but vvith
vehemencie of ſpirite
\XRef{(Exod.~14.)}
VVhy crieſt thou to me?}
crie come to thee.

%%% 1319
\V Turne
\SNote{Though ſinne prouoke Gods wrath, becauſe we by ſinning turne from
him, and not he firſt from vs: yet we pray God not ſo to leaue vs, but
to geue vs new grace, that by humilitie, and penance we may returne to
him, and not dye in ſinne.}
not away thy face from me: in what day ſoeuer I am in tribulation,
incline thine eare to me.

In what day ſoeuer I shal inuocate thee, heare me ſpeedely.

\V Becauſe my dayes haue vanished as
\SNote{Mans dayes, and al his workes are nothing worth, but vaniſh like
ſmoke, ſo long as he is in mortal ſinne:}
ſmoke: and my
\SNote{yea his beſt workes, as if he geue
\Fix{almoſe}{almeſe}{obvious typo, fixed in other}
faſt, pray, and dye for the truth, yet al thoſe auaile nothing
\XRef{(1.~Cor.~13.)}
but are}
bones are withered as
\SNote{like dryed ſtickes, or chippes, fitte to kindle the fire.}
a drie burnt firebrand.

\V I
\SNote{My ſoule ſeparated by ſinne from God, withereth as graſſe, that
is cutte from the roote:}
am ſtriken as graſſe, and my hart is withered: becauſe
\SNote{becauſe I haue loſt al ſauour and appetite to ſpiritual meate.}
I haue forgotten to eate my bread.

\V For
\SNote{In this miſerable ſtate,}
the voyce of my groning, my
\SNote{I am as bones and fleſh cleauing together without moyſture, or
radical humour.}
bone hath cleaued to my flesh.

\V I am become like
\SNote{I fled from conuerſation of men for ſorow, and ſhame of my
ſinnes:}
a pellicane of the wildernes: I am become as
\SNote{as a crow that only flieth by night: or as an owle, or batte.}
a nightcrow in the houſe.

\V I haue watched, and am become as
\SNote{Alſo as a ſparow hauing loſt her mate, remaineth mourning and
ſolitarie in the accuſtomed neſt, or nere vnto it.}
a ſparow ſolitarie in the houſetoppe.

\V Al the day did mine enemies vpbrayde me: and they
\SNote{Thoſe that were wont to praiſe or flatter me, now are as ſworne
enimies againſt me.}
that prayſed me, ſware againſt me.

%%% o-1210
\V Becauſe I did
\SNote{Bread ſauoured to me no better then ashes,}
eate ashes as bread, & mingled
\SNote{and drinke gaue me no comfort, but ſtil I wept.}
my drinke with weeping.

\V At
\SNote{I am moſt eſpecially aflicted, becauſe thou art angrie.}
the face of thy wrath and indignation: becauſe
\SNote{In that thou didſt ſometime aduance me in proſperitie,}
lifting me vp thou haſt
%%% !!! SNote not marked in either
\SNote{my fall is ſo much greater, and more grieuous.}
throwne me downe.

\V My daies haue declined
\SNote{As a ſhadow declineth to nothing, and al becometh darknes, when
the ſunne, and other light departeth, euen ſo I, that am but a ſhadow,
decline to mere darknes, when thy fauoure parteth from me;}
as a shadow: and I am withered
\SNote{and I loſe my beautie, as graſſe cutte from the ground
withereth.}
as graſſe.

%%% 1320
\V
\SNote{But I am meruelouſly comforted, conſidering that thou our
Meſſias, the Sonne of God, art immutable for euer:}
But thou ô Lord endureſt for euer: and
\SNote{and thy memorable promiſe of redeeming mankind, wil haue effect
in al generations.}
thy memorial in generation and generation.

\V Thou
\SNote{Thou riſing to helpe, who ſemedſt to haue forgote, wilt protect
the Church, and euerie faithful ſoule,}
ryſing vp shal haue mercie on Sion: becauſe it is
\SNote{becauſe thou haſt differred long:}
time to haue mercie on it, becauſe
\SNote{and becauſe the time by thee diſigned ſemeth to approch.}
the time cometh.

\V Becauſe
\SNote{Men that ſhal heare thyn Apoſtles preach, ſhal proue good and
fitte matter, for the building of thy Church:}
the ſtones therof haue pleaſed thy ſeruantes: and they shal haue pittie
on
\SNote{and the ſimpleſt poore people, as it were, the earth or duſt,}
the earth
\SNote{ſhal participate of this mercie.}
therof.

\V And
\SNote{Beſides thoſe Iewes that ſhal beleue in Chriſt, much more the
Gentiles ſhal feare and ſerue him.}
the Gentiles shal feare thy name ô Lord, and al
\SNote{His glorie is ſo euident that al kinges know it, though al be not
conuerted.}
the kinges of the earth thy glorie.

\V Becauſe our Lord hath built Sion: and he shal be ſeene in his glorie.

\V He hath had reſpect to the prayer
\SNote{Of holie Patriarches, Prieſtes, Prophetes, and of al true
penitents.}
of the humble: and he hath not deſpiſed their petition.

\V Let theſe thinges be written vnto an other generation: and the
people, that shal
\SNote{That shal be made a new creature in Chriſt.}
be created, shal praiſe our Lord.

\V Becauſe he hath looked forth from his high holie place: our Lord from
heauen hath looked vpon the earth.

\V That he might heare the gronings of the fettered: that he might
looſe the children of them that are ſlayne:

\V That they may
\SNote{The faithful people of the Church, according to their habilitie
endeuour to ſerue Chriſt.}
shew forth the name of our Lord in Sion: and his praiſe in Ieruſalem.

%%% o-1211
\V In the aſſembling of the people together in one, and kinges to ſerue
our Lord.

\V He anſwered him in the way of his ſtrength: Shew me the fewnes of my
daies.

\V Cal me not backe
\SNote{Grant me time and meanes to be mature in vertue in this life.}
in the halfe of my daies: thy yeares are vnto generation and generation.

\V In the beginning ô Lord thou didſt found the earth: and the heauens
are the workes of thy hands.

%%% 1321
\V
\CNote{\XRef{Heb.~1.}}
They
\SNote{Be changed in qualitie.}
shal perish, but thou art permanent: and they shal al waxe old as a
garment.

And as a veſture thou shalt change them, and they shal be changed: \V
but thou art the ſelf ſame, and thy yeares shal not faile.

\V The children of thy ſeruantes shal inhabite: and
\SNote{The Church of Chriſt perpetual.}
their ſeede shal be directed for euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
