%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 60
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-60


\project douay-rheims


%%% 1248
%%% o-1139
\startChapter[
  title={Psalme 60}
  ]

\PSummary{Faithful
\MNote{A confident prayer for Chriſts Incarnation.

The 5.~key.}
people of the whole earth pray, and acknowledge that God mercifully
heareth their prayer, 6.~expect the eternal kingdom of Chriſt, in which
they shal praiſe him for euermore.}

\PTitle{Vnto the end,
\SNote{In ſongues of praiſe and thankes to God.}
in hyms to Dauid.}

\VV Heare ô God my petition: attend to my praier.

\V From
\SNote{From al coaſtes of the earth faithful people pray to God,}
the ends of the earth I haue cried to thee: whiles
%%% o-1140
my hart was in anguish, thou
\SNote{the Church builded vpon an aſſured fundation, is exalted to great
powre and dignitie.}
didſt exalt me on a rocke.

\V Thou haſt
\SNote{God conducteth, defendeth, and deliuereth thoſe that confidently
truſt in him.}
conducted me, becauſe thou art made my hope: a toure of ſtrength from the
face of the enimie.

\V I shal inhabite in
\SNote{In the Church, a place of aſſured protection.}
thy tabernacle for euer: I shal be protected in the couert of thy
winges.

\V Becauſe thou my God haſt heard my prayer: thou haſt giuen inheritance
to thoſe that feare thy name.

\V Thou wilt adde
\SNote{Chriſts kingdome the Church perpetual to the end of this world,
and eternal after the general Reſurrection.}
dayes vpon the dayes of the king: his years euen vnto the day of generation
and generation.

%%% 1249
\V He is permanent for euer in the ſight of God: his
\SNote{Who is able to vnderſtand, or explicate how great Chriſts mercie
is in redeming vs,}
mercie and
\SNote{and his truth in performing his promiſed rewardes?}
truth who shal require?

\V So
\SNote{For ſo imeſtimable benefites, I wil alwayes praiſe thee with
Pſalmes, Canticles, or other thankes in this life:}
wil I ſay a Pſalme to thy name for euer and euer: that I may render my
vowes
\SNote{and eternally in the life to come.}
from day to day.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
