%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 092
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-092


\project douay-rheims


%%% 1307
%%% o-1198
\startChapter[
  title={Psalme 92}
  ]

\PSummary{Chriſt
\MNote{Perpetuitie of the Church.

The 6.~key.}
reigneth for euer in his Church: 3.~notwithſtanding manie and great
perſecution againſt the faithful.}

\PTitle{Prayſe
\SNote{Praiſe to be ſongue vvith voice:}
of Canticle
\SNote{compoſed by Dauid:}
to Dauid himſelfe, in
\SNote{the ſixth day of the weeke, vvhich is our friday,}
the day before the ſabbath,
\SNote{in vvhich day the Church of Chriſt vvas founded, by his bloud
ſhed on the croſſe.}
when the earth was founded.}

%%% !!! This is the first verse?
\NV Our Lord
\SNote{Our Sauiour founding his Church by his death begane then to
reigne therin:}
hath reigned, he
\SNote{gloriouſly
\Fix{eſcending}{deſcending}{obvious typo, fixed in other}
in ſoule into limbus, and in bodie to his graue:}
hath put on beutie: our Lord hath
\SNote{he then put on al armour of ſtrength to reforme the world, and to
inlarge his kingdom, according to his owne prediction where he ſaid: If
I be exalted from the earth, I wil draw al thinges vnto myſelfe.
\XRef{Ioan.~12. v.~31.}}
put on ſtrength, and hath girded him ſelfe.

%%% 1308
For he hath eſtablished
\SNote{Not only in Iurie and Samaria, but the whole earth:}
the round world, which
\SNote{and the ſame Church ſhal not be deſtroyed.}
shal not be moued.

\V Thy ſeat is prepared from that time: thou
\SNote{Chriſt being eternal, hath an euerlaſting Church.}
art from euerlaſting.

\V The
\SNote{Al ſortes of perſecuters, the High prieſtes (who ſometimes
vvatered the ſpiritual land, like riuers) vvith Scribes, Phariſes, and
other incredulous Ievves; alſo Paganes, Turkes, and Heretikes haue
oppugned the Church.}
riuers ô Lord haue lifted vp: the riuers haue lifted vp their voice.

\V The riuers haue lifted vp their waues, \V
\SNote{With more force then anie perſecutions in the old Teſtament.}
aboue the voices of manie waters.

The ſurges of the ſea are meruelous;
\SNote{But though al theſe aſſaultes be great, and meruelous, yet Chriſt
in protecting his Church is more meruelous.}
meruelous is our Lord on high.

\V Thy teſtimonies are made
\SNote{Articles of faith are not euidently apparent to knovvlege, but
euident to credibilitie, to thoſe that are diſpoſed by Gods grace,
illuminating their vnderſtanding, and mouing their free vvil, to geue
conſent of beleefe if they vvil.}
credible exceedingly:
\SNote{It behoueth therfore al members of the Church to conuerſe
piouſly, and religiouſly in this life, ſeing ſhe hath ſo excellent a
ſpouſe, protector, and inſtructor,}
holines becometh thy houſe ô Lord
\SNote{euen to the end of the world.}
for length of daies.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
