%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 048
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-048


\project douay-rheims


%%% 1229
%%% o-1121
\startChapter[
  title={Psalme 48}
  ]

\PSummary{The
\MNote{Exhortation to flee from ſinne for feare of hel.

The 7.~key.}
royal prophet inuiting al ſtates and ſortes of men, to heare him
attentiuely, 6.~sheweth that al ought to feare eternal damnation, that
liue wickedly, 9.~vainly and foolishly ſeeking (13.~euen like brute
beaſtes) carnal pleaſures, which they can not long enioy, nor long
eſcape hel: 16.~confidently animating him ſelfe, and al good men, that
truſt not in this world.}

%%% 1230
\PTitle{Vnto the end,
\SNote{In this and diuers other titles, both before and yet enſuing, is
ſaid, To the ſonnes, or, for the ſonnes of Core, a Pſalme, or Canticle,
or vnderſtanding & the like; but in no place, a Pſalme, Canticle, &c. of
the ſonnes of Core, vvhich no way proueth that they vvere the authores
of ſuch Pſalmes, but rather the contrarie.}
to the ſonnes of Core a Pſalme.}

\VV Heare theſe thinges
\SNote{Al ye nations and ſortes of people,}
al ye Gentiles: receiue with your eares al ye, that
\SNote{that dwel vpon the earth, lerne this leſſon vvhich I wil teach
you.}
inhabite the earth.

\V Al ye earthly perſons, and children of men: together in one the rich
and the poore.

\V My mouth shal ſpeake wiſedom, and the meditation of my hart prudence.

\V I wil
\SNote{Holie Dauid harkened to God inſpiring him,}
incline mine eare vnto a parable: I wil
\SNote{and declared to others that vvhich he receiued from God,}
open my propoſition
\SNote{not only by his penne or tongue, but alſo for better inſtilling
it into their mindes he ſounded it vpon the inſtrument called the
Pſalter, vvhich had tenne ſtringes, ſignifying the obſeruation of the
tenne commandments.}
on a Pſalter.

\V
\SNote{What eſpecial thing is there in this life, vvhy or for vvhich I
or anie haue cauſe to feare the dreadful day of iudgement?}
Why shal I feare in the euil day?
\SNote{Marry this we muſt feare, iniquitie, by which any ſupplanteth,
defraudeth, oppreſſeth, or anie vvay wrongeth others, for that vvil
inuolue the offender in the ſentence of eternal damnation.}
the iniquitie of my heele shal compaſſe me.

\V They
\SNote{Such be they that truſt in their preſent powre, riches, or other
worldlie thing.}
that truſt in their ſtrength: and glorie in the multitude of their
riches.

\V A
\SNote{A mans owne brother can not helpe a ſinner in that day,}
brother doth not redeme,
\SNote{much leſſe anie other man; ſo the Hebrew phraiſe by zeugma,
vnderſtandeth an other negatiue particle.}
man shal redeme: he shal not geue vnto God his reconciliation.

%%% o-1122
\V And the price of the redemption of his owne ſoule: and he shal
\SNote{Stil ſuffer paine,}
labour for euer, \V and
\SNote{and not dye, but liue in eternal torments.}
shal liue yet vnto the end.

\V He shal not ſee death, when he shal ſee
\SNote{Al both wiſe and fooliſh do dye temporally: but the wiſe liuing
in eternal ioy, the foolish liue in eternal paine.}
the wiſe dying;
\SNote{Thoſe that beleue not anie other life after this,}
the vnwiſe, and
\SNote{and thoſe that beleuing an other life, yet liue badly in this,
shal perish in eternal damnation.}
the foole shal perish together.

And they shal leaue their riches to ſtrangers: \V and their
\SNote{They shal neuer returne from their ſepulchers,}
ſepulchers their
\SNote{to enioy againe their houſes and earthlie poſſeſſions,}
houſes for euer.

Their Tabernacles in generation and generation: they haue renowmed their
\SNote{which vainely they labour to eſtablish in their poſteritie.}
names in their landes.

%%% 1231
\V And
\SNote{A moſt pitie and brief conſideration, for man to thinke, how
abſurdly, he being endewed with reaſon, vnderſtanding, & free wil, like
vnto Angels, and capable of eternal glorie, ſetteth his vvhole ſtudie,
and care vpon corporal and temporal thinges, ſo making himſelfe like
vnto brute beaſtes.}
man, when he was in honour, did not vnderſtand: he was compared to
beaſts without vnderſtanding, and became like to them.

\V This their way is
\SNote{This care of worldlie thinges is the ſtumbling block, and cauſe
of eternal ruine:}
a ſcandal to them: and
\SNote{yet they ſhal be obſtinate, and praiſe their owne deſires, ſtil
\Fix{peſiſting}{perſiſting}{obvious typo, fixed in other}
therin.}
afterward in their mouth they shal take pleaſure.

\V As
\SNote{Amongſt other creatures a ſheepe can leaſt helpe her ſelfe in
miſerie: euen ſo the damned in hel are altogether vnable to deliuer
themſelues from thence, or to get any relief.}
sheepe they are put in hel: death shal feede vpon them. And the iuſt
shal rule ouer them
\SNote{In the general reſurrection they ſhal be moſt of al in miſerie,
as euer dying and neuer dead: the iuſt vvhom they vvronged, shal be
their iudges, al freindes shal faile them, after they haue paſſed their
glorie, and pleaſure in this vvorld.}
in the morning: and their aide shal waxe old in hel from their glorie.

\V Neuertheleſſe
\SNote{The confidence of the iuſt.}
God wil redeme my ſoule out of the hand of hel, when he shal take me.

\V Feare not when a man shal be made rich: and when the glory of his
houſe shal be multiplied,

\V Becauſe when he shal dye, he shal not take
\SNote{He shal leaue al worldlie thinges and take nothing with him.}
al thinges: neyther shal his glorie goe downe with him.

\V Becauſe his ſoule in his life shal be
\SNote{Temporally.}
bleſſed: he wil confeſſe to thee
\SNote{So long as he enioyeth worldlie profites he wil ſeme gratful to God.}
when thou shalt do him good.

\V He shal enter in, euen to the progenies of his fathers: and he shal
\SNote{But they shal not ſee the true light of heauen.}
not ſee light for euer.

%%% o-1123
\V
\SNote{Remember and conſider ô worldlie man, that God made thee an
excellent creature: which thou neglecting makeſt thyſelf like to a
beaſt. As,
\XRef{v.~13.}}
Man, when he was in honour, did not vnderſtand: he was compared to
beaſts without vnderſtanding, and became like to them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
