%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 036
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-036


\project douay-rheims


%%% 1208
%%% o-1101
\startChapter[
  title={Psalme 36}
  ]

\PSummary{An
\MNote{An exhortatiõ to contempt of this world.

The 7.~key.}
exhortation not to enuie nor imitate the euil, who for moſt part proſper
in this world, and are damned eternally: but to flee euil and doe good,
duly conſidering that God diuerſly permitteth and punisheth the wicked,
and likewiſe comforteth and afflicteth the iuſt, al for their good.}

%%% 1209
\PTitle{A
\SNote{This Pſalme is compoſed in order of the Alphabet, euerie diſtick
beginning vvith a diuers letter, to moue the reader to diligent
attention, which may ſerue in place of a larger cõmentarie.}
Pſalme
\SNote{For Dauid and euery iuſt mans inſtruction.}
to Dauid him ſelf.}

\NV Haue
\SNote{Neither be thou offended that the wicked do proſper in this
world, nor imitate them that thou maiſt alſo proſper.}
no emulation toward the malignant: neither enuie them that doe
iniquitie.

\V Becauſe
\SNote{For al this life, and conſequently the proſperitie therof is
shorte, and vncertaine.}
they shal quickely wither as graſſe: and as the bloſſomes of herbes they
shal ſoone fal.

\V
\SNote{Put thy truſt in God, liuing content in this world:}
Hope in our Lord and doe good: and inhabite the land, and thou
\SNote{& he wil geue thee that is neceſſarie.}
shalt be fed in the riches therof.

\V Be delighted in our Lord: and he wil geue thee the petitions of thy
hart.

\V
\SNote{Commend al thyn affaires to God.}
Reuele thy way to our Lord, and hope in him: and he wil doe it.

\V And he wil
\SNote{Partly making vertue appeare to the comfort of the vertuous, and
example of others in this life, but eſpecially in the next world.}
bring forth thy iuſtice as light: and thy iudgement as midday: \V be
ſubiect to our Lord, and pray him.

Haue no emulation in him, that proſpereth in his way: in a man that doth
iniuſtices.

\V Ceaſe from wrath, and leaue furie: haue not emulation that thou be
malignant.

\V Becauſe they that are malignant, shal be caſt out: but they that
expect our Lord, the ſame shal inherite
\SNote{The land of the liuing.}
the land.

\V And yet a litle while, and the ſinner shal not be: and thou shalt
ſeeke his place, and shal not find it.

\V But the meeke shal inherite the land, and shal be delighted in
multitude of peace.

\V The ſinner shal obſerue the iuſt: and shal gnash vpon him with his
teeth.

\V But our Lord shal ſcorne him: becauſe he foreſeeth that his day shal
come.

\V Sinners haue drawen out the ſword; they haue bent their bowe;

%%% 1210
%%% o-1102
That they may deceiue the poore and needie: that they may murder the
right of hart.

\V Let
\SNote{By way of imprecation (as in manie other places) the prophet
forsheweth that wicked men shal fal into the euils, which they prepare
for others.}
their owne ſwordes enter into their hartes: and let their bowe be
broken.

\V Better is a litle to the iuſt, aboue much riches of ſinners.

\V Becauſe the armes of ſinners shal be broken in pieces: but our Lord
comfirmeth the iuſt.

\V Our Lord knoweth the daies of the immaculate: and their inheritance
shal be for euer.

\V They shal not be confounded in the euil time, and in the dayes of
famine they shal be filled: \V becauſe the ſinners shal perish.

But the enemies of our Lord forthwith as they shal be honoured and
exalted, vanishing shal vanish as ſmoke.

\V The ſinner shal borrow, and not pay: but the iuſt is merciful and wil
geue.

\V Becauſe they that bleſſe him shal inherite the land: but they that
curſe him shal perish.

\V With our Lord the ſteppe of man shal be directed: and he shal like
wel of his way.

\V When
\SNote{Though the iuſt fal of frailtie or ignorance into venial ſinne,
yet Gods grace shal ſtay him that he fal not into mortal. The iuſt
falleth ſeuen times in the day, & riſeth.
\XRef{Prou.~24.}}
he shal fal, he shal not be bruſed: becauſe our Lord putteth his hand
vnder.

\V I haue bene yong, for I am old: and I haue not ſene the iuſt
forſaken,
\SNote{So king Dauid obſerued: and it very rarely happeneth, that the
iuſt or their children are deſtitute of neceſſarie ſuſtenance in this
vvorld.
\CNote{\XRef{Eccle.~2.}}
If it chance in ſome, it is to their greater merite, and is manifeſtly
recompenſed in ſpiritual giftes. In which ſenſe S.~Baſil expoundeth,
that it is alwayes verified. For God euer rewardeth good workes either
temporally, or ſpiritually, or both wayes. S.~Auguſtin alſo
\Cite{(conc.~3. in hunc Pſal.)}
exemplifieth in Abraham, Iſaac, and Iacob with al his familie, who were
forced to goe into other countries by reaſon of famine, and by Gods
prouidence were there ſuſtained,
\XRef{Gen.~12.}
\XRef{26.}
\XRef{46.}
and S.~Paul among his other tribulations mentioneth famine and thirſt.
\XRef{2.~Cor.~11. v.~27.}
Seing therfore theſe ſo iuſt perſons ſought their bread in neceſſitie,
he expoundeth this holie Scripture in the Allegorical ſenſe, that the
Church from her beginning in Ieruſalem to the end of the world, neuer
wanteth the true word of God, true faith and doctrin, which is the
ſpiritual bread vvhervvith the ſoule is nouriſhed.}
nor his ſeede ſeeking bread.

\V Al the day he is merciful and lendeth: and his ſeede shal be in
bleſſing.

%%% 1211
\V
\SNote{In theſe tvvo prĩciples, declining frõ euil and doing good, true
iuſtice conſiſteth,}
Decline from euil, and doe good: and
\SNote{and he that finally obſerueth theſe tvvo pointes, meriteth and
ſhal poſſeſſe heauen.}
inhabite for euer and euer.

\V Becauſe our Lord loueth iudgement, and he wil not forſake his
ſainctes: they shal be preſerued for euer.

%%% o-1103
The vniuſt shal be punished: and the ſeede of the impious shal perish.

\V But the iuſt shal inherite the land: and shal inhabite for euer and
euer vpon it.

\V
\CNote{\XRef{Pro.~31.}}
The mouth of the iuſt shal meditate wiſedom,
\CNote{\XRef{Iſa.~51.}}
and his tong shal ſpeake iudgement.

The law of his God in his hart: and his ſteppes shal not be ſupplanted.

\V The ſinner conſidereth the iuſt: and ſeeketh to
\SNote{To draw him to mortal ſinne, which is death of the ſoule.}
murder him.

\V But our Lord wil not leaue him in his handes: neither wil he condemne
him, when iudgement shal be geuen of him.

\V Expect our Lord, and keepe his way: and he wil exalt thee, that thou
mayeſt inherite the land: when the ſinners shal perish thou shalt ſee.

\V I haue ſeene the impious highly exalted, and aduanced as the cerers
of Libanus.

\V And I paſſed by, and behold he was not: and I ſought him, and his
place was not found.

\V Keepe innocencie, and ſee equitie: becauſe there are
\SNote{Rewardes.}
remaynes for the peaceable man.

\V But the vniuſt shal perish together: the remaines of the impious shal
perish.

\V But the ſaluation of the iuſt is
\SNote{Mans iuſtice and wel doing is not of his owne powre but of Gods
grace.}
of our Lord: and he is their 
\Fix{potector}{protectour}{obvious typo, fixed in other}
in the time of tribulation.

\V And our Lord wil helpe them, and deliuer them: and he wil take them
away from ſinners, and ſaue them: becauſe they haue hoped in him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
