%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 025
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-025


\project douay-rheims


%%% 1189
%%% o-1082
\startChapter[
  title={Psalme 25}
  ]

\PSummary{Dauid
\MNote{Dauids prayer diſtreſſed in perſecution.

The 8.~key.}
in banishment among the
\Fix{Philiſtimes,}{Philiſtines,}{likely typo, fixed in other}
truſteth in the iuſtice of his cauſe, 9.~and prayeth God earneſtly to
deliuer him, that he may with more freedom and commodity ſerue him as he
deſireth.}

\PTitle{Vnto
\SNote{This Pſalme is alſo a conuenient prayer for anie Chriſtian in
tribulation.}
the end the Pſalme of Dauid.}

\NV Iudge
\SNote{Be thou ô God arbiter of the cauſe betwen Saul and me, thou
knoweſt myn innocencie in this behalfe, though I am vniuſtly charged by
Saul, and his freindes.}
me ô Lord becauſe I haue walked in my innocencie: and hoping in our Lord
I shal not be weakened.

\V 
\SNote{Leſt perhaps I be not ſo innocent as I deſire, and as in reſpect
of Saul I hope that I am, do thou Ô God proue me as thou wilt by
tribulations.}
Proue me Lord, and tempt me: burne my reynes and my hart.

\V Becauſe thy mercie is before mine eies: and I am wel pleaſed in thy
truth.

%%% 1190
\V I
\SNote{Dauid in confidence of a good conſcience, and zele againſt the
wicked, alleageth his ſincere proceding, more then ordinarie men may do,
God ſo inſpiring him extraordinarily.}
haue not ſitten with the councel of vanitie: and with them that doe
vniuſt thinges I wil not enter in.

\V I
\LNote{I haue hated the Church of the malignant.}{Holie
\MNote{A ſingular great afflictiõ to be hindered from Gods true
ſeruice.}
Dauid forced by reaſon of perſecution to dwel amongſt Infidels the
Philiſtians: after he had twiſe ſpared king Saules life
\XRef{(1.~Reg.~24. v.~5.}
et
\XRef{c.~26. v.~9.)}
lamented
\XRef{(v.~19.)}
how great affliction it was to him, \Emph{to be caſt out, that he could
not dvvel in the inheritance of our Lord} (where God was rightly
ſerued) and that his enemies had done ſo much as in them lay, to make
him fal into idolatrie, by their fact, as it were ſaying: \Emph{Goe
ſerue ſtrange goddes}.
\MNote{Chriſtians muſt abhorre, and abſtaine from al conuenticles of
Heretikes and other Infidels.}
Neuertheles his zele was ſuch that (as he here profeſſeth) \Emph{he
hated the Church of the malignant}: that is, the congregations of al
miſcreants: & his immaculate religious puritie was ſo perfect, that he
would not ſo much as in external ſhew, conforme his actions to theirs,
in matters of religion, nor yeld his bodelie preſence in their
conuenticles: but ſaid: \Emph{VVith the impious I vvil not
ſitte}. Inſtructing vs Chriſtians (for the word \Emph{to the end}, in
the title ſheweth that this Pſalme perteyneth alſo to vs) that we muſt
both \Emph{hate the Church or congregations of the malignant}, to witte,
of Painims, Iewes, Turkes, and Heretikes, and \Emph{not ſitte}, nor
perſonally be preſent with them in the exerciſe of their falſe pretended
religions.}
haue hated the Church of the malignant: and with the impious I wil not
ſitte.

\V I
\SNote{The reſt of this Pſalme euerie Prieſt reciteth in Maſſe, before
he offer the holie Sacrifice, profeſſing, & putting him ſelfe in
memorie, that he muſt 
only communicate with the innocent, or of pure conſcience;}
wil wash my handes among innocentes:
\SNote{and ſo approch to the Altar. 
\CNote{Epi.~120. c.~11.}
prefigured
\XRef{Leuit.~16. v.~4.}}
and wil compaſſe thy altar ô Lord:

\V That I
\SNote{Shutting the eares of my hart from euil and vaine thoughtes I wil
attend to godlie inſpirations,}
may heare the voice of praiſe, and
\SNote{and ſo with mental prayer, and external voice (as the holie order
of this ſacred office requireth) praiſe thee ô God in thy meruelous
workes.}
shew forth al thy meruelous workes.

\V Lord I haue loued
\SNote{I can not but ſingularly loue the excellencie of this place
dedicated to thy ſeruice: where is true faith, vnitie, and charitie of
thy people, the guard of holie Angels, the adminiſtratiõ of ſacred
myſteries, aſſiſtance of the Holie Ghoſt, real preſence of Chriſt our
Lord, al replenished with Diuine maieſtie.}
the beautie of thy houſe, and the place of
\SNote{This repreſenteth vnto me the glorious heauenlie kingdom of God
and al Sainctes.}
the habitation of thy glorie.

\V
\SNote{Suffer me not therfore to be contaminate by the wicked, nor to be
deuoured with them.}
Deſtroy not ô God my ſoule with the impious, and my life with bloudie
men.

%%% o-1083
\V In whoſe
\SNote{They are ſtil readie to committe more and more iniquities.}
handes are iniquities: their
\SNote{Themſelues being corrupted, endeuour by giftes of worldly
commodities to corrupt others.}
righthand is replenished with giftes.

\V But
\SNote{Euerie one ought ſo to purge his conſcience, that he may be
innocent, or free from great ſinne.}
I haue walked in mine innocencie:
\SNote{Deliuer me from this neceſſitie of dwelling among the wicked.}
redeme me, and haue mercie on me.

\V My
\SNote{I intend to walke right.}
foote hath ſtood in the direct way: in
\SNote{I deſire to praiſe thee amongſt thy true faithful ſeruantes.}
the Churches I wil bleſſe thee ô Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
