%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 68
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-68


\project douay-rheims


%%% 1260
%%% o-1151
\startChapter[
  title={Psalme 68}
  ]

\PSummary{Chriſt
\MNote{Chriſts afflictions and victorie.

The 5.~key.}
in middes of afflictions (as one in dangerous waters) 5.~deſcribing the
malice of perſecutors, and his owne true zele, 14.~prayeth his heauenlie
Father for helpe. 23.~By way of iuſt imprecation, forsheweth the ſeuere
punishment of his aduerſaries, 30.~his owne glorious Reſurrection, and
proſperous building of his Church. For which he inuiteth al creatures to
praiſe God.}

\PTitle{Vnto
\SNote{Perteyning to the nevv Teſtament,}
the end,
\SNote{for Gentiles conuerted to Chriſtianitie, and from vice to vertue,}
for them that shal be changed,
\SNote{prefigured in Dauid.}
to Dauid.}

\VV Saue me ô God: becauſe
\SNote{Vehement afflictions inuiron my hart.}
waters are entered into my ſoule.

\V I
\SNote{I am as one intangled vvith quickeſand, or quadmyre in the bottom
of a great vvater.}
ſticke faſt in the myre of the depth: and there is no ſure ſtanding.

%%% 1261
I am come into the depth of the ſea: and a tempeſt hath ouerwhelmed me.

\V I haue
\SNote{Our Lord ſvveat bloud for anguiſh in his prayer, and vvas not
deliuered from his Paſſion, neither are his ſeruantes preſently
deliuered from tribulations, but as is moſt to Gods honour and their
ovvne good.}
labored crying, my iawes are made hoarſe: my eies haue
%%% o-1152
failed, whiles I hope in my God.

\V
\CNote{\XRef{Ioan.~15.}}
They are multiplied aboue the heares of my head, that hate me without
cauſe.

Mine enimies are made ſtrong, that haue perſecuted me vniuſtly:
\SNote{Our Sauiour who had no ſinne, payde the ranſom for al ſinnes.}
then did I pay the thinges that I tooke not.

\V O God thou knoweſt my
\SNote{O God thou knovveſt, that this vvhich ſemeth follie to vvorldlie
men, is true vviſdom;}
foolishnes: and
\SNote{and though men charge me vvith offences, thou knovveſt that I am
innocent.}
mine offences are not hide from thee.

\V
\SNote{Suffer not the weake to be ſcandalized in my paſſions.}
Let them not be ashamed vpon me, which expect thee ô Lord, Lord of
hoſtes.

Let them not be confounded vpon me that ſeeke thee, ô God of Iſrael.

\V Becauſe for thee haue I ſuſtained reproch, confuſion hath couered my
face.

\V  I am become a forener to my brethren, and a ſtranger to the ſonnes of
my mother.

\V
\CNote{\XRef{Ioan.~2.}
\XRef{Rom.~15.}}
Becauſe
\SNote{The zele of ſeeking Gods honour, in propagating and aduancing his
Church, is the cauſe of perſecution. As vve ſee thoſe are leſſe
perſecuted, vvhich haue leſſe godlie zele.}
the zele of thy houſe hath eaten me: and the reproches of them that
reproched thee, fel vpon me.

\V And I couered my ſoule in faſting: and it
\SNote{The vvicked do reproch thoſe that mortifie themſelues.}
was made a reproch to me.

And I put heare cloth aganſt my garment, &  became a parable to them.

\V They ſpake aganſt me
\SNote{The great men and iudges, alſo the drunkards and raſcalitie of
the people.}
that ſat in the gate; and they ſoong aganſt me that dranke wine.

\V But I,
\SNote{But I direct my prayer to thee,}
my prayer to thee ô Lord: a time of thy good pleaſure ô God.

In
%%% !!! SNote goes before 'In'
\SNote{expecting the time of thy good pleaſure.}
the multitude of thy mercie heare me, in the truth of thy ſaluation:

\V Deliuer me out of the myre, that I ſtick not faſt: deliuer me from
them that hate me, and from the depthes of
\SNote{Tribulations.}
waters.

%%% 1262
\V Let not the tempeſt of water drowne me, nor the depth ſwallowe me:
\SNote{Though Chriſt died and vvas buried, and in ſoule deſcended into
hel, yet he could not be holden in his ſepulchre, nor in limbo, but roſe
againe.}
neyther let the pit shut his mouth vpon me.

\V Heare me ô Lord becauſe thy mercie is benigne: according to the
multitude of thy commiſerations haue reſpect to me.

%%% o-1153
\V And turne not away thy face from thy ſeruant: becauſe I am in
tribulation, heare me ſpeedily.

\V Attend to my ſoule and deliuer it:
\SNote{That they may either be conuerted, or confounded, and ſo do no
more hurt.}
becauſe of mine enimies deliuer me.

\V Thou knoweſt my reproch, & my confuſion, & my shame.

\V In thy ſight are al they that afflict me, my hart hath looked for
reproch and miſerie.

And I expected ſome bodie that would be ſorie together with me, and
there
\SNote{Not anie that could mitigate our Sauiours affliction, vvould
ſhevv compaſſion tovvards him.}
was none: and that would comfort me, and I founde not.

\V
\CNote{\XRef{Mat.~27.}
\XRef{Ioan.~19.}}
And they gaue
\SNote{But contrarivviſe vvhen he complained of thirſt, they gaue him
gal and vinegre to drinke.}
gal for my meate: & in my thirſt they gaue me vinegre to drinke.

\V 
\CNote{\XRef{Rom.~11.}}
\SNote{A prophecie of the deſtruction of the Ievves, at the time of
Paſch, vvhen they ſhould thinke to eate their paſchal lamb vvith ioy.}
Let their table be made a ſnare before them, & for retributions, and for
a ſcandal.

\V Let
\SNote{They are alſo blinde in hart, that they vvil not ſee the truth of
Chriſts doctrine:}
their eies be darkned that they ſee not: and make
\SNote{but bovv themſelues to earthlie thinges, and vvorldlie gaines
euen to this day.}
their backe crooked alwaies.

\V Poure out thy wrath vpon them: and let the furie of thy wrath
ouertake them.

\V
\CNote{\XRef{Act.~1.}}
Let their habitation be made deſert: and in their tabernacles let
there be none to dwel.

\V Becauſe
\SNote{Chriſt, vvhom God of this mercie deſigned to ſuffer death, for
redemption of mankind, the Ievves of mere enuie and malice perſecuted to
death.}
whom thou haſt ſtriken, they haue perſecuted: and vpon the ſorrow of my
wounds they haue added.

\V Adde thou
\SNote{A prophecie that God would ſuffer them, to fal from one iniquitie
to an other.}
iniquitie vpon their iniquitie: and let them not enter into thy
iuſtice.

%%% 1263
\V
\SNote{In the end of their liues, they ſhal not be found in the booke of
life, vvhere they ſuppoſe
\Fix{themſeelues}{themſelues}{obvious typo, fixed in other}
to be vvritten.}
Let them be put out of the booke of the liuing: and with the iuſt let
them not be written.

I
%%% !!! SNote goes befoe 'I'
\SNote{The voice of Chriſt, humbling himſelfe to death, euen to death of
the Croſſe:}
am poore and ſorowful:
\SNote{from which he roſe againe.}
thy ſaluation ô God hath receiued me.

\V I wil praiſe the name of God with canticle: and wil magnifie him in
prayſe.

\V And
\SNote{Deuout praiſe and thankeſgeuing pleaſe God more then ſacrifices
of the moſt tender calues, which vvere othervviſe alſo gratful
ſacrifices.}
it shal pleaſe God more then a young calfe: that bringeth forth hornes and
hoofes.

%%% o-1154
\V Let the poore ſee and reioyce: ſeeke ye God, and your ſoule shal
liue.

\V Becauſe our Lord hath heard the poore: and he hath not deſpiſed
\SNote{God doth aſſuredly comfort al ſuch, as are impriſoned for
profeſſing the truth.}
his priſoners.

\V Let
\SNote{Al the creatures of God.}
the heauens and earth praiſe him, the ſea, and al the creeping beaſtes
in them.

\V
\SNote{God wil alvvayes eſtabliſh and protect the Catholique Church:}
Becauſe God wil ſaue Sion: and
\SNote{and particular Churches, members of the vniuerſal ſhal alſo
proſper.}
the cities of Iuda shal be built vp.

And they shal inhabite there, & by inheritance they shal get it.

\V And
\SNote{Perpetual ſucceſſion of the Catholique Church.}
the ſeede of his ſeruants shal poſſeſſe it, and they that loue his name
shal dwel in it.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
