%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 072
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-072


\project douay-rheims


%%% 1268
%%% o-1159
\startChapter[
  title={Psalme 72}
  ]

\PSummary{The
\MNote{Affliction in this world is recompenſed in the end.

The 9.~key.}
royal prophet, firſt profeſſing the vnſpeakable goodnes of God, 2.~in
the perſon of the weake complaineth, that the wicked proſper, and the
iuſt are afflicted: 15.~reprehendeth the murmuration, for though we can
not comprehend the ſecrete iudgements of God, 18.~yet they are moſt
iuſt. 25.~So with deſire to reſt in God, he promiſeth euer to praiſe
him.}

\PTitle{A Pſalme
\SNote{To be ſongue by Aſaph the maſter of muſike, and his companie.}
to Aſaph.}

\NV How
\CNote{\XRef{Pſal.~39.}}
\SNote{God is ſo excellent good, as no tongue can expreſſe, no hart can
conceiue.}
good is God to Iſrael, to them, that are of a right hart!

\V But
\SNote{The peace and proſperitie of the wicked tempted me, to leaue the
right narrow way, and to goe the brode eaſie way with them.}
my feete were almoſt moued, my ſteppes almoſt ſlipped.

\V Becauſe I haue had zele vpon the wicked, ſeing the peace of ſinners.

\V Becauſe there is
\SNote{They neuer thincke of death;}
no reſpect to their death: and
\SNote{and if they ſuffer ſome affliction, they quickly remoue it from them.}
ſtabilitie in
\Fix{ther}{their}{obvious typo, fixed in other}
plague.

\V In
\SNote{They trauel not ſo much as good men do, neither are they ſo often
punished.}
the labours of men they are not, and with men they shal not be ſcourged.

\V Therfore
\SNote{Which maketh them proud, and ſo they runne into al ſortes of
iniquitie.}
hath pride helde them, they are couered with their iniquitie and
impietie.

%%% 1269
\V Their iniquitie hath proceded as it were of fatte: they haue paſſed
into
\SNote{Whatſoeuer miſchief came into their mind, they feared not to
committe it in fact.}
the affection of the hart.

\V They haue thought, and haue ſpoken wickednes they haue ſpoken
iniquitie
\SNote{Boldly extolling their owne wordes and workes, and deſpiſing
others.}
on high.

\V They haue ſette their mouth vnto heauen:
\Fix{ad}{and}{obvious typo, fixed in other}
their tongue hath paſſed in the earth.

\V Therfore wil my people
\SNote{For this proſperitie of the wicked, ſome of Gods people wil fall
from the right way,}
returne here: and
\SNote{and their dayes wil abound with calamities.}
ful daies shal be found in them.

\V And they haue ſaide:
\SNote{The voice of the weake, inclining to diffidence, and doubt of
Gods knowlege, and prouidence.}
How doth God know, and is there knowledge in the higheſt?

\V Loe the ſinners them ſelues, and they that abounde in the world, haue
obtained riches.

%%% o-1160
\V And I ſaide: Then haue I iuſtified my hart without cauſe, and haue
washed my handes amongſt innocentes:

\V And haue bene ſcourged al the day, and my chaſtiſing in the
morninges.

\V If I ſaide: I wil ſpeake this: Behold
\SNote{The prophet in his owne perſon anſwereth, that if he ſhould ſay
ſuch thinges, he ſhould reproue Gods children, as faultie, vnworthie of
his protection, and to be neglected; wheras it is Gods prouidence to
chaſtice his children whom he loueth.}
I reproued the nation of thy children.

\V I
\SNote{In this life we can not know the particular cauſes, why the iuſt
are afflicted, and manie wicked proſper in this world.}
thought to know this thing, it is labour before me:

\V Vntil I may enter into the ſanctuarie of God: and may vnderſtand
concerning their latter endes.

\V But yet
\SNote{But only in general we are taught, that euils are prepared for
the guiles, or deceptes,}
for guiles thou haſt put it to them: thou haſt caſt them downe whiles
they
\SNote{pride, and other ſinnes of the wicked.}
were eleuated.

\V How are they brought into deſolation, they haue failed ſodanely: they
haue perished for their iniquitie.

\V As the dreame of them that riſe ô Lord, in thy citie thou shalt bring
their
\SNote{Not real but imaginarie felicitie.}
image to nothing.

%%% 1270
\V Becauſe my hart is
\SNote{Almoſt burned vp with afflictions.}
inflamed, and my reynes are changed:

\V And I am brought to nothing, and knew not.

\V As
\SNote{Charged with heauie burdens, without knowlege how long, or to
what end,}
a beaſt am I become with thee: and I
\SNote{yet not deſtitute of thy protection.}
alwaies with thee.

\V Thou haſt helde my right hand: and in thy wil thou haſt conducted me,
and with glorie thou haſt receiued me.

\V For what is to me
\SNote{Nothing to be deſired in heauen, nor in earth but God.}
in heauen? and beſides thee what would I vpon the earth?

\V My
\SNote{In bodie and mind I thirſt after thee, my true inheritance.}
flesh hath fainted, and my hart: God of my hart, and God my portion for
euer.

\V For behold, they
\SNote{Men by their freewil make themſelues enimies to God,}
that make them ſelues farre from thee, shal perish: thou
\SNote{determined their damnation, forſeing their final ſinnes.}
haſt deſtroyed al, that fornicate from thee.

\V But it is good for me to cleaue to God: to put my hope in our Lord
God:

That I may shew forth al thy prayſes, in the gates of the daughter of
Sion.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
