%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 59
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-59


\project douay-rheims


%%% 1247
%%% o-1138
\startChapter[
  title={Psalme 59}
  ]

\PSummary{King
\MNote{King Dauids thankes for victories.

The 8.~key.}
Dauid after his owne and the peoples manie tribulations, 8.~rendereth
thankes for their renowmed victories, 11.~atchiued by Gods only powre.}

\PTitle{Vnto the end, for them,
\SNote{The change of ſtate from aduerſitie to proſperitie in the people
of Iſrael, was a figure of the like change in the Church of Chriſt,}
that shal be changed, in
\SNote{worthie to be remembred,}
the inſcription of the title
\SNote{for the inſtruction,}
to Dauid him ſelfe,
\SNote{of Gods beloued,}
for doctrine, \V
\SNote{as the ſame are more largely recorded in the bookes of kinges.}
when he ſet fire on Meſopotamia of Syria, and in Sobal, and Ioab
returned, and ſtroke Idumæa in the valley of ſalt-pittes twelue
thouſand.
\XRef{(2.~Reg.~8.}
&
\XRef{10.}
&
\XRef{1.~Paril.~18.}}

%%% o-1139
\VV O God thou
\SNote{God ſuffereth his people to be afflicted, as wel for their
ſinnes, as for exerciſe in vertue,}
haſt repelled vs, & haſt deſtroyed vs: thou waſt angrie
\SNote{after ſheweth his mercie in pardoning, and fauour in aduancing
them.}
and haſt had mercie on vs.

\V Thou haſt moued the earth, and haſt trubled it: heale the breaches
thereof, becauſe it is moued.

\V Thou haſt shewed vnto thy people
\SNote{By punishĩg ſinners.}
hard thinges: thou haſt made vs drinke the wine of compunction.

\V Thou haſt
\SNote{VVarning them to amẽd,}
geuen a ſignification to them that feare thee: they flee from the face
of the bow.

%%% !!! Only marked in other
\V That thy beloued
\SNote{and then reſtoreth them to former good ſtate.}
may be deliuered, ſaue me with thy right hand; and heare me.

%%% 1248
\V God hath ſpoken in
\SNote{God alſo as he hath promiſed by his holie oracle,}
his holie: I shal reioyce, and
\SNote{hath aduanced king Dauid, in his temporal kingdom, and doth much
more aduance him, and other elect in euerlaſting life.}
shal diuide Sichem: and shal meſure the valley of tabernacles.

\V Galaad is mine, and Manaſſes is mine: and Ephraim the ſtrength of my
heade.

\V Iuda my king: Moab
\SNote{As a veſſel for meaner vſes.}
the pot of my hope.

Into Edom wil I
\SNote{Bring it vnder my dominion.}
ſtretch out my shoe: the foreners are ſubiect to me.

\V Who shal conduct me into a fenſed citie? who shal conduct me euen
into Idumea?

\V Shal not thou ô God,
\SNote{As God doth ſometimes punish:}
that haſt expelled vs: and
\SNote{ſo he alſo rewardeth.}
wilt not thou ô God goe forth in our hoſtes?

\V Giue vs aide from tribulation: becauſe mans ſaluation is vaine.

\V In God we shal doe
\SNote{Strongly, with fortitude.}
ſtrength: and he shal bring to nothing them that afflict vs.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
