%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 65
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-65


\project douay-rheims


%%% 1254
%%% o-1145
\startChapter[
  title={Psalme 65}
  ]

\PSummary{The
\MNote{Gentils ſuccede the Iewes.

The 6.~key.}
prophet inuiteth al men to praiſe God for his meruelous workes, and
benefites donne to the Iewes. 7.~Who being vngratful, 8.~Gentiles are
called, 16.~and bring forth better fruict.}

\PTitle{Vnto the end, a Canticle
\SNote{Myſtical reſurrection, Gentiles
\Fix{ſucceding}{ſucceeding}{likely typo, fixed in other}
in place of the Iewes.}
of reſurrection.}

\NV Make ye
\SNote{Shew your internal ioy by external wordes and deedes.}
iubilation to God al the earth, \V ſay a pſalme to his name: geue glorie
to his praiſe.

\V Say ye to God:
\SNote{In drovvning the vvorld, in confounding the tongues in Babel, in
burning Sodom and Gomorrha with brimſton, in plaging the Ægyptians, in
drovvning Pharao and his vvhole armie in the read ſea, in deſtroying the
Chananites and other infidels, in puniſhing the tenne tribes, and
aftervvards the other tvvo by captiuitie, and innumerable other
puniſhments, al for ſinnes,}
How terrible are thy workes ô Lord! in the multitude of thy ſtrength
thine enimies
\SNote{for vvhich euen the vvicked, though not ſincerly conuerted, yet
of ſeruile feare, feaned and falſly promiſed to amend, but performed it
not: as Pharao afflicted vvith plagues, vvas forced to promiſe libertie
to the children of Iſrael, vvhich he aftervvards denied.}
shal lie to thee.

%%% o-1146
\V Let al the earth adore thee, and ſing to thee: let it ſing a pſalme
to thy name.

\V Come ye, and ſee the workes of God: terrible in counſels ouer the
children of men.

%%% 1255
\V Who turneth
\SNote{The read ſea,}
the ſea into drie land, in
\SNote{when Ioſue brought the people ouer Iordan,}
the riuer they shal paſſe on foote,
\SNote{in remembring and reciting theſe ſingular benefites.}
there we shal reioyce in him.

\V Who ruleth in his ſtrength for euer his eyes looke vpon the gentiles:
\SNote{Gods choſen people the Iewes did often exaſperate God by their
ingratitude, murmuring, and other ſinnes: whom the prophet therfore
admoniſheth,}
they that exaſperate him let them
\SNote{not to be proud, leſt they be
\Fix{ſuboued,}{ſubdued,}{obvious typo, fixed in other}
and brought lowe.}
not be exalted in themſelues.

\V Ye
\SNote{By way of inuitation the Pſalmiſt prophecieth the conuerſion of
Gentiles.}
Gentiles bleſſe our God: and make the voice of his prayſe hearde.

\V Who hath put
\SNote{The voice of the whole Church: confeſſing Gods prouidence and
protection, that ſhe neuer faileth; for the Iewes falling from Chriſt
the Gentiles beleued in him; and ſome nations, or countries falling from
Religion, others are conuerted.}
my ſoule in life: and hath not geuen my feete to be moued.

\V Becauſe thou
\SNote{God ſuffereth his Church to be perſecuted with al kindes of
tribulation, as ſome are here recited.}
haſt proued vs ô God: by fire thou haſt tried vs, as ſiluer is tried.

\V Thou haſt brought vs into a ſnare, thou haſt laide tribulations on
our backe: thou haſt ſet men vpon our heades.

\V
\SNote{But through Gods aſſiſtance his ſeruants paſſe through, and
ouercome al tentations.}
We haue paſſed through fire and water: and thou haſt brought vs out into
refreshing.

\V I wil goe into thy houſe with
\SNote{Sacrifice of thankes,}
holocauſtes: I wil render thee
\SNote{and voluntarie vowes.}
my vowes: \V which my lippes haue diſtinguished. And
\SNote{That which anie promiſeth to God in tribulation, they muſt
performe accordingly.}
my mouth hath ſpoken in my tribulation.

\V
\SNote{Theſe were the beſt external ſacrifices of the old law.}
Holocauſtes with marrow wil I offer to thee, with incenſe of rammes: I
wil offer to thee oxen with bucke goates.

\V
\SNote{But both then and now the internal ſacrifices of contrite hart,
of iuſtice, and diuine praiſe beſt pleaſe God.}
Come ye, heare, and I wil tel al ye that feare God, what great things he
hath done for my ſoule.

\V To him haue I cried with my mouth, and haue exulted
\SNote{From the hart, which is vnder the tongue, and directeth the tongue
what to ſpeake.}
vnder my tongue.

\V
\SNote{VVhoſoeuer wil be heard in prayer, muſt repent of his ſinnes.}
If I haue beheld iniquitie in my hart, our Lord wil not heare.

%%% 1256
%%% o-1147
\V Therfore hath God heard, and hath attended to the voice of my
petition.

\V Bleſſed be God who hath not remoued my prayer, and his mercie from
me.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
