%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 008
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-008


\project douay-rheims


%%% 1161
%%% o-1053
\startChapter[
  title={Psalme 8}
  ]

\PSummary{God
\MNote{Chriſts Incarnation. The 5.~key.}
is magnified & praiſed for his meruelous worke of
  creatures, 5.~but eſpecially of mankind, ſingularly exalted by the
  Incarnation of Chriſt.}

\PTitle{Vnto
\SNote{Belonging to the newe Teſtament.}
the end
\SNote{Suffering of Chriſt, and of Chriſtians.}
for
\LNote{Preſſes.}{Moſt
\MNote{Preſſes ſignify Chriſts Paſſiõ.}
Hebrew Doctors ſay the word \HH{Gittith}, may either ſignifie the place,
where this Pſalme was made, or the muſical inſtrument, on which it was
ſong. But moſt Chriſtian Doctors expound it literally of Chriſts
Paſſion, who was ſtretched on the Croſſe, and al his ſacred bloud
preſſed, and drawne out of his bodie. VVhich Metaphor Iſaias alſo vſeth,
demanding of Chriſt:
\CNote{\XRef{Iſa.~63.}}
VVhy is thy clothing redde, and thy garments as theirs, that tread in
the \Emph{vine preſſe}? and anſwereth in Chriſts perſon: I haue
troden \Emph{the preſſe} alone.
\MNote{Morally it ſignifieth the trauels of the Church militant.}
S.~Auguſtin alſo applieth it morally to
the Church, where Chriſt is the vine, the Apoſtles are the branches, &
ſpreaders (that is preachers) of the Ghoſpel, Chriſtians are the grapes,
Chriſtian vertues are the wine. Namely patience, and fortitude in
afflictions. VVherby the good are purified, and ſeuered from amiddes the
reprobate, as wine is preſſed out of the grapes, barreled, and laid vp
in ſellers, and
\TNote{\L{vinacia.}}
the huskes and carnels caſt to hogges, or other beaſtes.}
preſſes, the Pſalme of Dauid.}

\VV O Lord
\SNote{God the lord of al creation, is our ſingular lord, that beleue
and truſt in him.}
our Lord, how meruelous is thy name in the whole earth! Becauſe thy
magnificence is eleuated, aboue
\SNote{God more excelent then the heauens, he being the creator, they a
creature.}
the heauens.

\V
\CNote{\XRef{Mat.~21.}}
Out of the mouth of
\SNote{Fulfilled whẽ Chriſt coming into Ieruſalem, children ſang
\HH{Ozanna}
\XRef{Mat.~21.}}
infantes and ſucklinges, thou haſt perfected praiſe becauſe of thine
enemies, that thou mayeſt deſtroy the enemie and reuenger.

\V Becauſe I shal ſee thy heauens, the workes of thy fingers: the moone
and the ſtarres, which thou haſt founded.

\V What is man, that thou art mindful of him? or the ſonne of man, that
thou viſiteſt him?

\V 
\CNote{\XRef{Heb.~2.}}
Thou haſt
\SNote{The Sonne in aſſumpted humãe nature became leſſe then Angels.}
minished him a litle leſſe then Angels; with
\SNote{But in him mans nature is exalted aboue Angels.}
glorie and honour thou haſt crowned him:

\V and haſt appointed him
\SNote{Chriſt the Lord of al creatures,}
ouer the worke of thy handes.

\V
\CNote{\XRef{Gen.~1. v.~17.}
\XRef{1.~Cor.~15.}}
Thou haſt ſubiected
\SNote{yea of Angels.
\XRef{Heb.~2.}}
al thinges vnder his feete, al
\SNote{Not only al reaſonable creatures, but al beaſtes, and other
thinges obey him. The ſea and the windes obey him.
\XRef{Mat.~8.}}
sheepe and oxen: moreouer alſo the beaſtes of the field.

\V The birdes of the ayre, and fishes of the ſea; that walke the pathes
of the ſea.

\V
\SNote{The ſame end & beginning ſignifie that as God was meruelous in
creating man in ſo happie ſtate, that if he would, he might haue auoyded
both ſinne & death: ſo he is meruelous in that he ſo rewardeth the
bleſſed in the reſurrection, that they can neither ſinne nor dye.}
Ô Lord our Lord, how meruelous is thy name in the whole earth!


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
