%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 142
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-142


\project douay-rheims


%%% 1391
%%% o-1282
\startChapter[
  title={Psalme 142}
  ]

\PSummary{King
\MNote{The ſeuenth penitential Pſalme.

The 7.~key.}
Dauid (or anie other) in ſpiritual or temporal tribulation, not truſting
in his owne iuſtice, layeth open his calamitie, 5.~conſidering Gods
benignitie, 6.~prayeth to be ſpedely deliuered; 11.~and confidently
aſſureth himſelfe therof.}

\PTitle{A Pſalme of Dauid, when
\CNote{\XRef{2.~Reg.~17.}}
Abſalom his ſonne perſecuted him.}

\NV Lord heare my prayer: with thine eares receiue my petition in
\SNote{God hauing ſo promiſed is bond by his truth,}
thy truth: heare me in
\SNote{and his iuſtice, to heare penitents praying for remiſſion of
ſinnes.}
thy iuſtice.

\V And
\SNote{Deale not vvith me in rigour of iuſtice,}
enter not into iudgement with thy ſeruant:
\SNote{for no mortal man is able of himſelfe to be iuſtified,
abſtracting from Gods mercie.}
becauſe no man liuing shal be iuſtified in thy ſight.

\V Becauſe the enemie
\SNote{Sought my life,}
hath perſecuted my ſoule: he
\SNote{and brought it into great danger. Spiritually, the diuel as a
roring lion ſtil ſeeking vvhom he may deuour, hath tempted me
vehemently.}
hath humbled my life in the earth.

He
%%% !!! SNote should go before 'He'
\SNote{I am preſſed with great calamities, temporal or ſpiritual.}
hath ſet me in obſcure places as the dead of the world: \V and my ſpirit
is in anguish vpon me, within me my hart is trubled.

%%% 1392
\V \SNote{In this caſe I conſider, how God hath hertofore shewed his
goodnes towards me, and others.}
I was mindful of old dayes, I haue meditated in al thy workes: in the
factes of thy handes did I meditate.

\V I haue
\SNote{Stretching forth handes a ceremonie in prayer, wherby the
ſupplicant is made more attentiue, and alſo indureth ſome paine for part
of ſatisfaction.}
ſtretched forth my handes to thee: my ſoule is
\SNote{Mans mind vvithout Gods illumination is drie and barren.}
as earth without water vnto thee.

\V Heare me
\SNote{In great tentations Gods grace and helpe is more preſently
nedeful to preuent our weaknes, leſt vve yeld conſent.}
quickly ô Lord: my ſpirite hath faynted.

Turne not away thy face from me:
\SNote{If God leaue man vvithout ſpecial and continual grace, he vvil fal,}
and I shal be like to them that deſcend into
\SNote{into ſinne, as into a deepe lake, from vvhence vvithout helpe he
can not riſe vp againe.}
the lake.

\V Make me heare thy mercie
\SNote{In the firſt aſſault of tentation.}
in the morning: becauſe I haue hoped in thee.

Make the way knowen to me, wherein I may walke:
\SNote{Seing by thy grace I haue begunne to pray vnto thee.}
becauſe I haue lifted vp my ſoule to thee.

\V Deliuer me from mine enemies ô Lord, to thee I haue fled: \V teach me
to doe thy wil, becauſe thou art my God.

%%% o-1283
Thy
\SNote{The penitent thus humbling himſelfe and praying may aſſuredly
truſt that God doth remitte his ſinnes by the holie Sacraments, and
iuſtifieth him, protecteth him, and wil bring him out of al dangers of
ſpiritual or temporal enemies.}
good ſpirite wil conduct me into the right way: \V for thy name ſake ô
Lord thou wilt quicken me, in thine equitie.

Thou wilt bring forth my ſoule out of tribulation: \V and in thy mercie
thou wilt deſtroy mine enemies.

And thou wilt deſtroy al, that afflict my ſoule: becauſe I am thy
ſeruant.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
