%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 41
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-41


\project douay-rheims


%%% 1218
%%% o-1110
\startChapter[
  title={Psalme 41}
  ]

\PSummary{The
\MNote{Eternal glory.

The 10.~key.}
feruent deſire of the iuſt, 6.~much afflicted in this life, 12.~and
aſſured hope of eternal ioy.}

\PTitle{Vnto the end, vnderſtanding to the ſonnes of
\SNote{The ſonnes of Core repented, and departed from their fathers
ſchiſme, and ſo eſcaped miraculoſly the horrible pitte of damnation,
into which their father and his complices fel.
\XRef{Num.~26. v.~10.}
By which example al ſeduced, and deceiued Chriſtians are admoniſhed, not
to perſiſt in ſchiſme or other ſinnes. And worldlie men, ambicious of
honour be warned to deſire & ſeke God aboue al thinges, firſt of al the
kingdome of heauen (to be liuing members of the Catholique Church) and
the iuſtice therof: to ſeke thinges which are aboue, not which are vpon
the earth; leſt hel deuoure them, as it deuoured the complices of Core.
\XRef{Num.~16. v.~31.}}
Core.}

\VV Euen as the harte
\SNote{A harte waxing old, and burdened with much heare, and great
hornes, draweth a ſerpent into his noſethrels, ſo being infected with
poyſon, deſireth moſt ardently to drinke, and afterwards caſteth his
hornes, and heare, and becometh as it were yong againe.}
deſireth after the fountaines of waters:
\SNote{With ſuch feruent deſire a true penitent, feeling him ſelfe
infected with
\Fix{poyſou}{poyſon}{obvious typo, fixed in other}
of ſinnes, ſeeketh the water of Gods grace.}
ſo doth my ſoule deſire after thee ô God.

%%% 1219
\V My ſoule hath thirſted after God
%%% !!! This SNote includes the next.
\SNote{God is omnipotent, and in dede the only true liuing God: diuels
who are honored in idols, can do no more then God permitteth, and ſo
they can kil the ſoules, that conſent vnto their tentations, but can not
reſtore ſpiritual life againe.}
the ſtrong
%%% \SNote{}
liuing:
\SNote{The ſoule being iuſtified, and ſtil aſſaulted with new tentations
deſireth to be with God.}
when shal I come and appeare before the face of God?

\V My
\SNote{I haue had no other refection, but to lenifie my ſorow with weping.}
teares haue beene breades vnto me day and night: whiles it is ſaid to me
dayly:
\SNote{The wicked exprobate the iuſt, as though God would neuer helpe
them, becauſe he ſuffereth them to be ſometimes long in tribulation.}
Where is thy God?

\V Theſe thinges haue I remembred, and haue powred out my ſoule in me,
becauſe I shal paſſe into the place
\SNote{King Dauid was not permitted to build the temple, much leſſe did
he enter into anie ſuch meruelous tabernacle in his mortal life, but
muſt nedes be vnderſtood, to ſpeake here of the heauenlie tabernacle,
prepared by Chriſt for his ſeruants.}
of a meruelous tabernacle, euen to the houſe of God.

In the
\Fix{yoyce}{voice}{obvious typo, fixed in other}
of exultation, and confeſſion: the ſound of one feaſting.

\V Why
\SNote{The prophet
\Fix{conforteth}{comforteth}{obvious typo, fixed in other}
him ſelfe, or anie iuſt ſoule, in the hope of euerlaſting ioy.}
art thou ſorowful my ſoule? and why doſt thou truble me?

Hope in God, becauſe yet I
\SNote{I render thankes and praiſes.}
wil confeſſe to him: the ſaluation of my countenance, \V and my God.

My ſoule is trubled toward my ſelfe; therfore wil I be mindful of
%%% o-1111
thee from the land
\SNote{Al this life is like to the ſmal ſtraict place betwene Iordan and
a litle hil called Hermoniim, but from this ſtraictnes the hope of the
iuſt is, to be placed in heauen.}
of Iordan, and Hermoniim from the litle mountaine.

\V Depth
\SNote{One tentation ſtil ſuccedeth an other:}
calleth on depth, in the voice of
\SNote{and the ſame ſo great, as if God opened the gates, and ſuffered
them to ouerflow like fluddes of water.}
thy flould-gates. Al thy high thinges, and thy waues haue paſſed ouer
me.

\V In
\SNote{But God helpeth in opportunitie, not ſuffering his ſeruantes to
be tempted aboue their ſtreingth, geuing them fruit with tentations:}
the day our Lord hath commanded his mercie: and
\SNote{yea in the greateſt tribulation, he geueth ordinarily moſt
comfort, making them ſing ſpiritually in hart, if not alſo in voice.}
in the night a ſong of him.

%%% 1220
With me
\SNote{One ſpecial meanes to procure diuine conſolation is prayer in
diſtreſſe.}
is prayer to the God of my life: \V I wil ſay to God: Thou art my
defender.

Why haſt thou forgotten me? and why goe I ſorowful, whiles mine enemie
afflicteth me?

\V Whiles my bones are broken, mine enemies that truble me haue
vpbrayded me:

Whiles they ſay to me day by day: Where is thy God?

\V Why art thou heauie ô my ſoule? & why doſt thou truble me?
\SNote{Stil the iuſt ſoule taketh comforth in aſſured hope of ſaluation,
the eternal viſion of God.}
Hope in God, becauſe yet I wil confeſſe to him: the ſaluation of my
countenance, and my God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
