%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 124
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-124


\project douay-rheims


%%% 1373
%%% o-1262
\startChapter[
  title={Psalme 124}
  ]

\PSummary{The
\MNote{Gods protection.

The 3.~key.}
Church, and holie members therof, are alwayes protected by God: 5.~The
wicked being ſuffered for a while, are at laſt punished.}

\PTitle{A gradual Canticle.}

\NV They
\SNote{Thoſe that confidently truſt in God, are in ſuch ſecuritie as is
mount Sion, vvhich is a hil, defended alſo vvith other hilles round
about.}
that truſt in our Lord, as mount Sion: he shal not be moued for euer,
\SNote{An other thing required to this ſecuritie, is to dvvel vvithin
Ieruſalem, not the terreſtrial citie, for (as S.~Auguſtin obſerueth) the
Ievves that dwelt therin are deſtroyed or made captiues, and hitherto
reiected of God, but in the Catholique Church.}
that dwelleth \V in Ieruſalem.

Mountaines round about it: and our Lord round about his people, from
hence forth now and for euer.

\V Becauſe our Lord
\SNote{God wil not alwayes, nor finally leaue his ſeruants in
tribulation, but only a vvhile for their good.}
wil not leaue the rod of ſinners vpon the lote of the iuſt: that the
iuſt reach not their handes to iniquitie.

\V
\SNote{This prayer is alſo an aſſeueration, for it is certaine that God
wil defend, and reward the good, and right of hart:}
Doe wel ô Lord to the good, and right of hart.

\V But
\SNote{And no leſſe aſſuredly God wil puniſh not only the notorious
wicked, and principal auctors of wickednes, but alſo al thoſe that for
feare or for commoditie, or for anie other cauſe, decline into
obligations, bondes, couenants, or anie vvay conſent, in external ſhew
with the wicked againſt God: as in outward profeſſion of Hereſie, or
Schiſme, though ſuch temporizers do not thincke in their hart, that the
pretended religion is true, wherto they are drawen to yeld external
conformitie. For, as the Pſalmiſt here teacheth, our Lord wil bring al
ſuch acceſſarie offenders, to the ſame iudgement and puniſhment, vvith
the principal vvorkers of iniquitie.}
thoſe that decline into obligations, our Lord wil bring with them that
worke iniquitie:
\SNote{Al vvhich being puniſhed, then Ieruſalem, the Catholique Church,
ſhal haue peace.}
peace vpon Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
