%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 063
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-063


\project douay-rheims


%%% 1251
%%% o-1142
\startChapter[
  title={Psalme 63}
  ]

\PSummary{A
\MNote{A confident prayer in tribulation.

The 7.~key.}
prayer of the iuſt repoſing their whole truſt in God: 7.~and reioycing
that the enimies machinations are fruſtrate.}

\PTitle{Vnto the end, a Pſalme of Dauid.}

\VV Heare ô God my prayer when I make petition: from the feare of my
enimie deliuer my ſoule.

\V Thou
\SNote{By example of thy former protection,}
haſt protected me from the
\SNote{from the conſpiracie of wicked men, I truſt moſt aſſuredly in thy
helpe.}
aſſemblie of the malignant:
%%% o-1143
from the multitude of them that worke iniquitie.

\V Becauſe they haue sharpned their tongues as a ſworde: they haue bent
the bow a bitter thing, \V that they may shoote in ſecretes at the
immaculate.

\V Sodanely they wil shoote at him, and wil not feare: they
\SNote{They are reſolued to intrappe me.}
haue confirmed to themſelues a wicked worde.

They haue talked to hide ſnares: they haue ſaide, who shal ſee them?

%%% 1252
\V They haue ſearched iniquities: they
\SNote{But as they haue failed: ſo they ſhal ſtil faile, and be
ouerreached in their bad counſels, as Achitophel.
\XRef{2.~Reg.~17.}}
haue failed ſearching with ſcrutanie.

\V Man shal come to a deepe hart: and God shal be exalted.

Childrens
%%% !!! should go before 'childrens'
\SNote{God hath choſen the weake of this world to confound the ſtrong.}
arrowes are made their woundes: \V and their tongues are weakned againſt
them.

Al that ſaw them
\SNote{Much merueled, ſeing the wicked ſo punished.}
were trubled: \V and euerie man feared.

And they shewed forth the workes of God: and they vnderſtood his
doinges.

\V The iuſt shal reioice in our Lord, and shal hope in him, and al the
right of hart
\SNote{The iuſt shal be praiſed, for rightly ſeruing God.}
shal be praiſed.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
