%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 130
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-130


\project douay-rheims


%%% 1377
%%% o-1267
\startChapter[
  title={Psalme 130}
  ]

\PSummary{Anie
\MNote{Confidence of innocencie.

The 7.~key.}
iuſt ſoule in humble confidence offereth his innocencie, as a ſpiritual
and gratful ſacrifice to God, 3.~exhorting al Gods ſeruants euer to hope
in him.}

\PTitle{A gradual Canticle of
\SNote{Dauid by Gods ſpecial grace hauing a ſincere minde tovvards al
men, euen tovvards his enimies, and an humble hart, not deſiring anie
thing ambiciouſly, but al to the honour of God, propoſeth his ovvne
example, for others to imitate: that they may vvith him offer the
ſacrifice of humilitie, and innocencie vnto God, from vvhom al good
thinges procede.}
Dauid.}

\NV Lord my hart is not exalted:
\SNote{Al this vvith a thankful mind to God vvho gaue this grace.}
neither are mine eies loftie. Neither haue I walked in great matters:
nor in meruelous thinges aboue me.

\V If I was not humbly mynded: but exalted my ſoule: As
\SNote{As children after they are vveaned come ſtil vvillingly to their
mother, ſo doth the childe of God relie vpon Gods helpe, though he
alvvayes feele not the ſame ſvvetnes:}
the weaned childe is toward his mother,
\SNote{& according to his demaneur herein, he expecteth revvard.}
ſo retribution in my ſoule.

\V
\SNote{King Dauid, or anie other being for his vertue aduanced, and
revvarded by God, is a good example to moue others to do the like, ſo
shal they receiue like revvard.}
Let Iſrael hope in our Lord, from henceforth now and for euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
