%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-21


\project douay-rheims


%%% 1181
%%% o-1073
\startChapter[
  title={Psalme 21}
  ]

\PSummary{Chriſt
\MNote{Chriſts Paſſion & effectes therof.

The 5.~key.}
prayeth in his Paſſion, 7.~deſcribeth the acerbitie therof;
20.~foresheweth by way of prayer his Reſurection: 23.~more clerly the
fundation and propagation of his Church (27.~&~30.~interpoſing the
ſingular foode of the moſt bleſſed Sacrament) euen to the endes of the
earth in al nations.}

\PTitle{Vnto the end
\LNote{For the morning enterpriſe.}{In
\MNote{Chriſts Reſurrection.}
reſpect of the end for which Chriſt ſuffered, this Pſalme is intitled:
\Emph{for the morning enterpriſe}: that is, for Chriſts glorious
Reſurrection, and other effectes of his Paſſion.
\MNote{The Paſſion of Chriſt according to Dauid.}
VVhich holie Dauid by the ſpirite of prophecy ſo deſcribeth here long
before with diuers particular circumſtances as the Euangeliſtes haue
ſince hiſtorically recorded, that it may not vnfitly be called,
\Emph{The Paſsion of Ieſus Chriſt according to Dauid}.}
for the
\SNote{Redemptiõ of mankind vndertakẽ by Chriſt, and performed by his
death, beganne to be shewed by his Reſurrection in the morning of the
third day.
\Cite{S.~Aug.}}
morning enterpriſe, the Pſalme
\SNote{In figure of Chriſt the \Emph{beloued of God.}}
of Dauid.}

\VV God
\SNote{God is God of al his creatures, but in more peculiar ſorte God of
Chriſt by perſonal vnion.}
my God haue reſpect to me:
\CNote{\XRef{Mat.~27.}}
\SNote{Thou neither deliuereſt me frõ dying: nor yeldeſt me ſuch
comforth, as thou voutſafeſt to other Sainctes.
\XRef{v.~6. Mat.~27.}}
why haſt thou forſaken me?
\SNote{I am neither deliuered from dying, nor comforted in my paſſion,
hauing vndertaken to dye for the ſinnes of mankind, & reputed them as
myn owne wordes or factes.}
far from my ſaluation are the wordes of my ſinnes.

\V My God I shal crie
\SNote{Albeit I crie by day,}
by day,
\SNote{and by night, on the croſſe and in the garden, to haue the
chalice of death remoued from me, and shal not be heard,}
and
\LNote{Thou vvilt not heare.}{Our
\MNote{Chriſts conditional prayer was not heard.}
B.~Sauiour ſeing his moſt terrible death imminent, prayde conditionally,
\Emph{if it pleaſed his heauenlie Father}, to haue the ſame remoued from
him; and was not heard, as the Pſalmiſt here prophecieth. The principal
reaſon was, becauſe God of his diuine charitie had decreed, that mankind
ſhould be redemed by this death of his Sonne. Chriſt alſo him ſelfe of
his excellent charitie, conſented here vnto & therfore perſiſted not in
his conditional prayer, but added and abſolutly prayed, that not his owne
wil, but his Fathers might be fulfilled.
\MNote{His abſolute prayers were alwaies heard.}
And in this he was heard, to his owne more glorie, and other infinite
benefites of innumerable ſoules, as it foloweth,
\XRef{v.~25.}
\Emph{vvhen I cried to him he heard me}. S.~Paul alſo witneſſeth
\XRef{(Heb.~5. v.~7.)}
that Chriſt \Emph{offering prayers and ſupplications to him that could
ſaue him from death, vvas heard for his reuerence}: that is, in reſpect
of his ineſtimable merite in humane nature vnited in perſon to God.
\MNote{Chriſt ſuffered for our example.}
An other cauſe, why Chriſt was not deliuered from violent death, as
manie holie perſons were, when they cried to God in diſtreſſe, as
S.~Auguſtin ſheweth
\Cite{(Epiſt.~120. c.~11.)}
was for example to Chriſtians, whom God wil haue to ſuffer temporal
afflictions, and death, for the glorie of life euerlaſting, according to
S.~Peters doctrin,
\CNote{\XRef{1.~Pet.~2.}}
\Emph{Chriſt ſuffered for vs, leauing an example that you may folovv his
ſteppes}.}
thou wilt not heare: and by night, and
\SNote{yet there is no follie, nor fault in this petition, which is with
ſubmiſſion of mans wil to Gods wil.}
not for follie vnto me.

\V But
\SNote{Thou from heauen wilt heare when it is conuenient.}
thou dwelleſt in the holie place,
\SNote{Whoſe wiſdome and goodnes the Church worthely praiſeth.}
the prayſe of Iſrael.

\V In thee
\SNote{Patriarkes, Prophets, and other holy mẽ praying in their
diſtreſſes obtained their requeſtes.}
our fathers haue hoped: they hoped, and thou didſt deliuer them.

\V They cried to thee, and were
\SNote{VVere temporally deliuered by thy mightie hand from their
perſecuters,}
ſaued: they hoped in thee, and were
\SNote{not fruſtrate of their humble prayers.}
not confounded.

\V
\SNote{God that comforted his ſeruantes in their tribulations, leift
Chriſt without his ordinarie conſolation, to ſuffer more then euer any
other did.}
But I am a
\SNote{Wicked perſecuters reſpected not Chriſt as a man, but contemned
him, as a very worme,}
worme and no man: a
\SNote{as moſt reprochful of al men,}
reproch of men and
\SNote{as the baſeſt of al the people.}
outcaſt of the people.

%%% 1182
%%% o-1074
\V
\SNote{Almoſt al became Chriſts cruel enimies at the time of his
paſſion, afflicting, blaſpheming, and ſcorning him, as the Euangeliſtes
record.}
Al that ſee me haue ſcorned me: they haue ſpoken with the lippes, and
wagged the head.

\V
\SNote{The Euangeliſtes write how al theſe thinges were fulfilled by
wicked men ſpeaking theſe blaſphemies, and reproches.}
He hoped in the Lord, let him deliuer him: let him ſaue him, becauſe he
\SNote{God ſemed to be wel pleaſed with Chriſt, as with his owne Sonne:
if it be ſo, let him deliuer him from theſe afflictions, ſay theſe
blaſphemers.}
wil him.

\V Becauſe thou art he that
\SNote{Diuine powre without man formed me in the wombe of my mother, a
virgin.}
haſt drawen me out of the wombe: my hope from the breſtes of my mother.

\V Vpon thee I
\SNote{As I haue no father but thee Ô God: ſo without intermiſſion from
myn incarnation to this time, I haue had thee my protector:}
haue bene caſt from the matrice:
\SNote{leaue me not now without comforte ſeing I muſt dye as thou haſt
determined, and I freely conſented: yet leaue me not in death but raiſe
me againe to life.
\XRef{Pſal.~15. v.~9.~10.}}
from my mothers wombe thou art my God, \V depart not from me.

Becauſe tribulation is verie nigh: becauſe there is
\SNote{Almoſt al are become myn enemies, and thoſe few that would, can
not helpe me.}
not that wil helpe.

\V Manie
\SNote{Delicate laſciuious yongmen,}
calues haue compaſſed me:
\SNote{and the ſcribes, Phariſes, and elders of the people haue al
conſpired againſt me.}
fatte bulles haue beſieged me.

\V They haue
\SNote{Condemning me and perſwading the people to crie: \Emph{Crucifie,
crucifie him}.}
opened their mouth vpon me, as a lion rauening and roaring.

\V As
\SNote{So weakned with paines of torments, as fluide water not able to
conſiſt.}
water I am powred out: and al
\SNote{My bones and ſtrongeſt partes of my bodie are weakned. Verified,
when our Sauiour fel downe vnder his croſſe.}
my bones are diſperſed.

My
\SNote{The part that firſt and laſt liueth is weakened as ſoft waxe by
heat of the fire, and ready to faile.}
hart is made as waxe melting in the middes of my bellie.

\V My ſtrength is withered as a
\SNote{Al my powres, and radical humiditie is dried vp, as a potters
veſſel is baked in the furnace.}
pot-shard, and my tongue
\SNote{Through exceding great 
\Fix{drught,}{drought,}{possible typo, fixed in other}
which our Sauiour profeſſed on the croſſe ſaying: \Emph{I thirſt.}}
cleaued to my iawes: and thou haſt
\SNote{Thus thou Ô God haſt ſuffered me to come to the laſt breath of
life, next to death. Yet finally our Sauiour gaue vp his ſpirite, before
he ſhould haue died.
\XRef{v.~21.}}
brought me downe into the duſt of death.

\V Becauſe manie
\SNote{Agane this royal Prophet recounteth by whom, and how our
B.~Sauiour ſhould ſuffer, euen as clere as the Euangeliſtes afterwards
haue written the hiſtorie.}
dogges haue compaſſed me: the counſel of the malignant hath beſieged me.

%%% o-1075
\V
\CNote{\XRef{Ioan.~19.}}
They
\LNote{They haue digged.}{Of
\MNote{The Hebrew text corrupted by the Iewes.}
obſtinate malice the Iewes haue corrupted this place (and God knoweth
how manie others) in the Hebrew text of ſome editions,
reading \HH{caari}, which ſignifieth \Emph{as a lion}, without al
coherence of the ſenſe for \HH{caaru}, \Emph{they digged}
or \Emph{pearced}, to auoid ſo plaine a prophecie of nailing Chriſts
handes and feete to the croſſe.}
haue digged my handes and my feete: \V they haue
%%% 1183
\SNote{Our Sauiours body was ſo racked on the croſſe, that his bones
might be ſeene and counted.}
numbered al my bones.

But themſelues haue
\SNote{The perſecuters vvitingly determined al this crueltie: beheld it
vvith their eyes, and vvithout al compaſſion perſiſted in malice,
reioyced, and blaſphemed.}
conſidered and beheld me: \V they haue
\SNote{The ſouldiars that crucified our Sauiour, taking his garmentes
for their paye:}
deuided my garmentes among them, and vpon my
\SNote{yet in myſterie of his Church, diuided not his coate.}
veſture they haue caſt lot.

\V But thou Lord
\SNote{He prophecieth Chriſts ſpeedy reſurrection.}
prolong not thy helpe from me: looke toward my defenſe.

\V Deliuer, ô God,
\SNote{Chriſts ſoul vvas not ſeperated from his bodie by force of the
torments, but he preuenting death freely yelded vp his ſpirite.
\XRef{Ioan.~10. v.~9. &~10.}}
my ſoule from the ſword: and myne
\SNote{The moſt pure, and ſanctified ſoule, of vvhoſe fulnes al other
iuſt ſoules are ſanctified.}
onlie one from
\SNote{That it ſtay not in hel, vvhich deuoured al other ſoules in the
old Teſtament.}
the hand of the dog.

\V Saue me out of the lions mouth: and my humilitie from the hornes of
vnicornes.

\V
\LNote{I vvil declare thy name to my brethren.}{Here
\MNote{This Pſalme is of Chriſt.}
it is euident that this Pſalme is of Chriſt, not of Dauid, by S.~Pauls
allegation
\XRef{(Heb.~2. v.~11.~12.)}
ſaying: He that ſanctifieth (to witt Chriſt) diſdained not to cal the
ſanctified his bretheren.}
I wil
\SNote{The propagation of the Church of Chriſt in al nations.}
declare thy name to my brethren:
\LNote{In the middes of the Church I vvil praiſe thee.}{After Chriſts
Paſſion and Reſurrection, in the reſt of this Pſalme, other two
principal pointes of Chriſtian Religion are likewiſe prophecied: His
perpetual viſible Church, and the B.~Sacrament of his bodie.
\MNote{Prophecie of the viſible and vniuerſal Church.}
The former is here prophecied by way of inuiting \Emph{al the ſeede of
Iacob to glorifie God}
\XRef{(v.~24.)}
\Emph{al the ſeede of Iſrael to feare him}
\XRef{(v.~25.)}
to wit, innumerable Chriſtians the true Iſraelites, the vniuerſal Church
in the whole world. As for heretical partes, or parcels in the world,
ſuch as the Donatiſtes, which going forth from the Catholique Church ſay:
Chriſt hath loſt his great Church, the diuel hath taken the whole world
from him, and he remaineth only in a part of Africa, they do not praiſe
God (ſaith
\CNote{\Cite{in hunc Pſal.}}
S.~Auguſtin) but diſhonour God and Chriſt, as if God were not faithful
in his promiſe, as if Chriſt were diſpoſſeſſed of his kingdome the
Catholique Church. Leſt anie ſhould replie, that Chriſt is praiſed
though the Church be decaied, or be very ſmal, the Holie Ghoſt hath
preuented ſuch arguments, ſaying:
\XRef{v.~26.}
His praiſe is \Emph{in the great Church}. VVhich could neither be
verified in the part of Donatiſtes in Afrike, nor now in the part of
Proteſtantes ſince Luther in Europe.
\MNote{S.~Auguſtin proueth the Church to be alwaies viſible and great by
this Pſalme.}
Further S.~Auguſtin explicateth, & vrgeth the verſes folowing in this
Pſalme, againſt the ſame blind, deafe, and obſtinate Donatiſtes, who did
not, or would not ſee, nor heare, that \Emph{al the endes of the earth
shal remember, and be conuerted to our Lord}. The holie Scripture ſaith
not, \Emph{the endes of the earth}, but \Emph{al the endes}: wel goe too
(ſaith this great Doctor) peraduenture there is but one verſe, thou
thoughteſt vpon ſome thing els, thou talkeſt with thy brother, when one
read this; marke, he repeteth and knocketh vpon the deaf: \Emph{Al the
families of the Gentiles shal adore in his ſight}: Yet the heretike is
deaf, he heareth not, let one knocke againe: \Emph{Becauſe the kingdom
is our Lords, and he shal haue dominion ouer the Gentiles}. Hold theſe
three verſes bretheren. Thus and more S.~Auguſtin againſt thoſe that
thinke the true Church may faile, or become inuiſible or obſcure. And
though it be not in like proſperous ſtate at al times, and in al places,
yet it is alwayes conſpicuous, and more general then anie other
congregation profeſſing whatſoeuer pretenſed religion.}
in the middes of the Church I wil prayſe thee.

\V Ye that feare our Lord praiſe him: al the
\SNote{Not the carnal but ſpiritual children of Iacob, Iſaac, and
Abraham.
\XRef{Rom.~9. v.~8.}}
ſeede of Iacob glorifie ye him:

\V Let al the ſeede of Iſrael feare him: becauſe he hath not contemned,
nor deſpiſed the petition of the poore. Neither hath he turned away his
face from me: and when I cried to him he heard me.

\V With thee is my praiſe in
\SNote{The Church gathered both of Ievves and Gentiles is very great and
vniuerſal.}
the great Church: I
\SNote{Our Sauiour promiſed to geue his ovvne bodie, the bread of life
\XRef{Ioan.~6.}
and performed the ſame at his laſt ſupper.}
wil render my vowes in the ſight of them that feare him.

\V
\LNote{The poore shal eate.}{Seing
\MNote{The Euchariſt prophecied in this place.}
this Pſalme is of Chriſt, as is proued by S.~Pauls allegation of
23.~verſe, and by the concordance therof with the Euangeliſts, it is
neceſſarily deduced, that \Emph{the vovves} mentioned in the former
verſe, and theſe wordes, \Emph{the poore shal eate and be filled}, can
not be referred to the ſacrifices of the old Teſtament, but to the
bleſſed Sacrifice and Sacrament of the Euchariſt, which our Sauiour
promiſed, after he had repleniſhed the people with fiue loaues, and
which he inſtituted at his laſt ſupper in preſence of his Apoſtles. So
S.~Auguſtin doubteth not to vnderſtand it, and to teach, as wel in his
duble
\Cite{expoſitiõ of this Pſalme}
as in his
\Cite{120.~Epiſtle. c.~27.}
The poore, that is the humble and poore in ſpirite shal eate & be
filled, the fatte ones, or the rich being proud, do alſo adore and eate,
but are not filled. They alſo are brought to the \Emph{table of
Chriſt}, and participate his bodie & bloud, but they \Emph{adore} only,
are not alſo filled, becauſe they do not imitate (Chriſts humilitie)
they diſdaine to be humble.
\MNote{Real preſence of Chriſt in the Euchariſt.}
VVhere it is clere, this holie father by Chriſts bodie and bloud meaneth
not bread and wine as ſignes of his bodie and bloud, for bread and wine
can not be lawfully \Emph{adored}: neither doth he meane our Lords bodie
as it was on the croſſe, or is in heauen, for ſo it is not eaten, but as
it is in formes of bread and wine on \Emph{Chriſts table}, the Altar.}
The
\SNote{Thoſe that be faithful, humble and poore in ſpirit participat the
fruict of this moſt excellent Sacrament.}
poore shal eate, and shal be filled: and they shal praiſe our Lord that
ſeeke after him: their hartes
\SNote{The effect of this B.~Sacrament is the reſurrection in glorie,
and life euerlaſting.}
shal liue for euer and euer.

\V Al the endes of the earth
\SNote{Gentiles which haue bene idolaters shal recollect themſelues,
when they heare Chriſt preached, and shal turne to true Religion.}
shal remember, and be conuerted to our Lord:

\V And al the families of the Gentiles shal adore in his ſight.

%%% 1184
\V
\SNote{Although men can neither deſerue to be conuerted, nor to
perſeuere in iuſtice; yet Chriſt meriteth to haue a continual kingdom;
which is the perpetual viſible Catholique Church.}
Becauſe the kingdome is our Lords: and he shal haue dominion
%%% o-1076
ouer the Gentiles.

\V Al the
\SNote{Not only the poore ſorte, but alſo the mightie ones of the world
shal be conuerted to Chriſt, participate his B.~Bodie in the Sacrament:}
fatte ones of the earth haue eaten, and
\SNote{and religiouſly adore the ſame.}
adored: in his ſight shal al fal,
\SNote{Al that adore God,
\Fix{ſhad}{ſhal}{obvious typo, fixed in other}
adore him in this Sacraments.}
that deſcend into the earth.

\V And
\SNote{Death being once ouercome, it shal haue no more powre.}
my ſoule shal liue to him: and my
\SNote{Againe the prophet inculcateth the continuance of the Catholique
Church.}
ſeede shal ſerue him.

\V The generation to come shal be shewed to our Lord: and the
\SNote{Apoſtles and other preachers of Chriſt.}
heauens shal shew forth his iuſtice to the people that shal be borne,
whom our Lord hath made.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
