%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 126
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\startcomponent psalme-126


\project douay-rheims


%%% 1374
%%% o-1264
\startChapter[
  title={Psalme 126}
  ]

\PSummary{Neither
\MNote{Gods helpe in al good vvorkes.

The 3.~key.}
houſe nor citie can be built, or kept without Gods ſpecial prouidence
and helpe, 3.~thoſe that truſt in him shal proſper in their endeuours.}

\PTitle{A gradual Canticle
\SNote{God not permitting Dauid to build the Temple, promiſed that his
ſonne ſhould build it; and therfore beſides other good admonitions,
geuen to his ſonne Salomon, he directeth this Pſalme to him, to be
ſongue vvith others in the dedication of the Temple.
\XRef{3.~Reg.~8,~2.}
\XRef{Par.~5.}}
of Salomon.}

\NV Vnleſſe
\SNote{Vnleſſe God be the principal Agent, no vvorke can proſper.}
our Lord build the houſe, they haue laboured in vayne that build it.

Vnleſſe our Lord kepe the citie, he watcheth in vayne that kepeth it.

\V It is vaine for you to riſe
\SNote{It is vaine to atempt anie thing vvithout Gods grace aſſiſting:}
before light:
\SNote{they that ſo doing, & thincking they haue done ſomething, reſt}
riſe ye after ye haue ſitten, which eate
\SNote{after their painful trauel, muſt riſe againe & beginne anevv,
becauſe that vvhich they ſeme to haue done vvel, is nothing vvorth, nor
ſhal haue revvard.}
the bread of ſorow.

%%% 1375
\V
\SNote{Contrarivviſe vvhere God geuing grace, thoſe that truly loue him,
do good vvorkes,}
When he shal giue
\SNote{vvith great eaſe, and delight, as they take their ſleepe,}
ſleepe to his beloued: behold
\SNote{they merite inheritance in heauen,}
the inheritance of our Lord
\SNote{for their good vvorkes.}
are children:
\SNote{The revvard is promiſed to the true children of God, borne to him
in the vvombe of the Catholique Church.}
the reward, the fruite of the wombe.

\V As
\SNote{Moreouer as a ſtrong archer ſtriketh deepe with his arrovves; ſo
they that patiently ſuffer much in this vvorld,}
arrowes in the hand of the mightie: ſo are the
\SNote{do multiplie good vvorkes.}
children of them that are shaken.

\V
\SNote{Such ſhal be very happie,}
Bleſſed is the man that hath filled his deſire of them: he shal
\SNote{& very eaſily anſvver al that can be obiected againſt them}
not be confounded when he shal ſpeake to his enemies
\SNote{in the day of Iudgement.}
in the gate.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
