%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Psalmes
%%%% Psalme 103
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent psalme-103


\project douay-rheims


%%% 1323
%%% o-1214
\startChapter[
  title={Psalme 103}
  ]

\PSummary{The
\MNote{Gods workes meruelous.

The 2.~key.}
Pſalmiſt inuiteth himſelf and others to praiſe God, for his meruelous
workes in the heauens, 5.~the earth, and water, 9.~limiting their
bondes, producing al thinges neceſſarie for al liuing creatures, in
conuenient ſeaſons, 27.~with continual prouidence of al.}

\PTitle{To Dauid him ſelf.}

\NV My ſoule bleſſe thou our Lord: ô Lord my God thou art magnified
excedingly.

%%% 1324
\V Thou haſt put on
\SNote{Thou poſſeſſeſt al maieſtie, and matter of praiſe.}
confeſſion and beautie: being clothed with light as with a garment:

Stretching out the heauen as a skinne: \V which couereſt the higher
partes therof
\SNote{Thou haſt compaſſed the ſphere of the fixed ſtarres, with a
ſphere of chriſtalline ſubſtance; which is as water congeled.}
with waters.

Which
\SNote{According to our capacitie the prophet deſcribeth the ſpeedie
coming, or vvorking of God as if he came in a ſvvift cloud, or vvith
vvinges of the vvind, to ſignifie that he vvorketh vvhat and vvhen he
pleaſeth vvithout delay: He ſaide, and thinges vvere made: he commanded
and they vvere created.
\XRef{Pſal.~32.}}
makeſt the cloude for thee to aſcend on: which walkeſt vpon the winges
of windes.

\V Which makeſt ſpirites thine
\SNote{Thy meſſengers to execute thy vvil:}
Angels: and thy
\SNote{& the ſame Angels are as a burning fire in operation, yea they
diſpatch more eaſily, and more ſpedely then vve can conceiue.}
miniſters a burning fyre.

\V Which haſt founded the earth vpon
\SNote{Moſt firmly eſtablished by natural weight, in the center of the
world.}
the ſtabilitie therof: it shal not be inclined for euer and euer:

\V The
\SNote{The water}
depth,
\SNote{should naturally couer al the earth:}
as a garment, is his clothing: vpon the mountaines shal waters ſtand.

\V At
\SNote{But by thy commandment the waters are conteined in their limited
places.}
thy reprehention they shal flee: at the voice of thy thunder they shal
feare.

\V The
\SNote{The waters being conteyned in their appointed chanels, both
hilles and fieldes appeare, which otherwiſe would be couered.}
mountaines aſcend: and the plaine fildes deſcend into the place, which
thou haſt founded for them.

\V Thou haſt ſet a bound, which
\SNote{The waters.}
they shal not paſſe ouer: neither shal they returne to couer the earth.

\V Which ſendeſt forth fontaines in the valles: betwen the middeſt of
mountaines shal waters paſſe.

%%% o-1215
\V Al the beaſtes of the filde shal drinke: the wilde aſſes shal
\SNote{Hope for and receiue.}
expect in their thirſt.

\V Ouer them shal the foules of the ayre inhabite: out of the middes of
rockes they shal geue forth voices.

\V Watering the mountaines from his higher places: of the fruite of thy
worke shal the earth be filled:

\V Bringing forth graſſe for beaſtes, and herbe for the ſeruice of men.

%%% 1325
That thou mayeſt bring forth
\SNote{By theſe three principal kindes of foode, breade, wine, and oyle,
al ſortes of nutriment are vnderſtood.}
bread out of the earth: \V and
%%% !!! This Snote is part of the above.
%%% \SNote{}
wine may make the hart of man ioyful:

That he may make the face chereful with
%%% !!! This Snote is part of the above.
%%% \SNote{}
oile: and bread may confirme the hart of man.

\V The trees of the filde shal be filled, and the ceders of Libanus,
which he hath planted: \V there ſparowes shal make their neſt.

The houſe of the hearne is the leader of them: \V the high mountaines
for hartes: the rocke a refuge for the Irchins.

\V He made the moone for ſeaſons: the ſunne knoweth his going downe.

\V Thou didſt appoint darkenes, and night was made: in it shal al the
beaſtes of the wood paſſe.

\V The whelpes of lions roaring, to rauen, and to ſeeke of God meate for
themſelues.

\V The ſunne is riſen, and they are gathered together: and in their
couches they shal be placed.

\V Man shal goe forth to his worke: and to his working vntil euening.

\V How magnified are thy workes ô Lord! thou haſt made al thinges in
wiſedom: the earth is filled with
\SNote{VVith thy creatures.}
thy poſſeſſion.

\V This great ſea, and very large, there are
\SNote{Fiſhes, ſerpents, wormes, and al liuing creatures that lacke feete.}
creeping beaſtes, wherof
\SNote{No kind of liuing creatures multiplieth ſo much as fiſhes.
\Cite{Ariſtotel li.~9. Animal. c.~17.}}
there is no number.

Litle beaſtes with great: \V there shippes shal paſſe.

This
\SNote{A moſt huge fiſh called Leuiathan.
\XRef{Iob.~40. v.~20.}}
dragon, whom thou madeſt to
\SNote{Albeit in the water he paſſeth mans ſtreingth, yet depriued of
water he is not able to defend himſelfe.
\XRef{v.~19.}}
delude: \V al expect of thee that thou geue them meate in ſeaſon.

\V Thou geuing vnto them, they shal gather it: thou opening thy hand, al
shal be filled with bountie.

\V But thou turning away the face, they shal be trubled: thou shalt take
away their ſpirite, and they shal faile, and shal returne into their
duſt.

\V Thou shalt ſend forth thy ſpirit, and they shal be created: and thou
shalt renewe the face of the earth.

%%% o-1216
\V Be the glorie of our Lord for euer: our Lord wil reioice in his
workes:

\V Who looketh vpon the earth, & maketh it to tremble: who toucheth the
mountaines, and they ſmoke.

\V I wil chaunte to our Lord in my life: I wil ſing to my God as long as
I am.

\V Let my ſpeach be acceptable to him: but I wil take delight in our
Lord.

%%% 1326
\V
\SNote{A prediction that impenitent ſinners ſhal be damned, wherin the
Prophet conforming his wil to Gods, vttereth it in forme of a prayer.}
Let ſinners faile from the earth, and the vniuſt, ſo that they be not:
my ſoule bleſſe thou our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
