%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 0185
%%% o-0169
\startChapter[
  title={Chapter 5}
  ]

\Summary{Moyſes and Aaron require of Pharao in the behalfe of God, to
  let his people the Hebrewes goe and ſacrifice in the deſert. VVhich he
  contemning, 5.~oppreſſeth them more, denying them ſtravv, and yet
  exacting the accuſtomed number of brickes. 20.~The people oppreſsed
  impute their miſerie to Moyſes and Aaron. 22.~But Moyſes prayeth to
  God for them.}

After theſe things Moyſes and Aaron went in, and ſaid to Pharao: This
ſaith the Lord God of Iſrael: diſmiſſe my people that they may ſacrifice
to me in the deſert. \V But he anſwered: Who is the Lord, that I ſhould
heare his voice, and diſmiſſe Iſrael? I know not the Lord, and Iſrael I
wil not diſmiſſe. \V And they ſaid: The God of the Hebrewes hath called
vs, to goe three daies iourney into the wilderneſſe, and to ſacrifice to
the Lord our God: leſt perhappes there chance to vs peſtilence or
ſword. \V The king of Ægypt ſaid to them: Why do you Moyſes and Aaron
ſolicite the people from their workes? Goe you to your burdens. \V And
Pharao ſaid: The people of the land is much: you ſee that the multitude
is ſecretly increaſed:
\SNote{VVordlie men thinke Gods people encreaſe moſt by reſt, but in
dede they multiplie more, when they are oppreſſed.
\Cite{S.~Cyprian. de exhort. Mart. c.~10.}}
how much more if you geue them reſt from their workes? \V Therfore he
commanded in that day the ouerſeers of the workes and the exactores of
the people, ſaying: \V You ſhal no more geue ſtraw to the people for to
make brickes, as before: but let them ſelues goe and geather ſtraw. \V
And the taske of brickes, which they did before, you ſhal put vpon them,
neither ſhal you diminiſh any thing: for they are idle, and therfore
they crie, ſaying: Let vs goe, and ſacrifice to our God. \V Let them be
oppreſſed with workes, and let them accompliſh them: that they hearken
not to lying wordes. \V Therfore the ouerſeers of the workes and the
exactors going forth ſaid vnto the people: Thus ſaith Pharao: I allow
you no ſtraw: \V goe, and geather if you can find any where:
%%% 0186
neither ſhal anie thing of your worke be diminiſhed. \V And the people
was diſperſed through al the Land of Ægypt to geather ſtraw. \V And the
%%% o-0170
ouerſeers of the workes were inſtant, ſaying: Finiſh your worke euerie
day, as before you were wont to doe when ſtraw was geuen vnto you. \V
And the ouerſeers of the workes of the children of Iſrael were ſcourged
of Pharaos exactors, ſaying: Why do you not make vp the taske of brickes
as before, neither yeſterday, not to day? \V And the ouerſeers of the
children of Iſrael came, and cried out to Pharao, ſaying: Why dealeſt
thou ſo againſt thy ſeruantes? \V Straw is not geuen vs, and brickes are
commanded vs in like ſorte: behold we thy ſeruantes are beaten with
whippes, and thy people is vniuſtly dealt withal. \V Who ſaid: You are
idle, and therfore you ſay: Let vs goe and ſacrifice to our Lord. \V Goe
therfore and worke: ſtraw ſhal not be geuen you, and you ſhal geue vp
the accuſtomed number of brickes. \V And the ouerſeers of the children
of Iſrael ſaw them ſelues in a hard caſe, becauſe it was ſaid vnto them:
There ſhal not a whitte be diminiſhed of the brickes for euerie day. \V
And they mette Moyſes and Aaron, who ſtood ouer againſt them, coming
forth from Pharao: \V and they ſaid to them: Our Lord ſee and iudge,
becauſe
\SNote{The craftie diuel knowĩg that weake men afflicted are eaſely
moued to murmur, ſturred this people againſt their owne leaders.
\Cite{S.~Greg. li.~29. c.~14. Moral.}}
you haue made our ſauour to ſtinke before Pharao and his ſeruantes, and
you haue geuen him a ſword, for to kil vs. \V And Moyſes returned to our
Lord, and ſaid: Lord
\SNote{Gods prouidẽce ſuffereth his children, to be moſt afflicted, when
relife is nere at hand.
\Cite{Theod. q.~13. in Exod.}}
why haſt thou afflicted this people? wherfore haſt thou ſent me? \V For
ſince the time that I entered in to Pharao to ſpeake in thy name, he
hath afflicted thy people: and thou haſt not deliuered them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
