%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 30
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-30


\project douay-rheims


%%% 0259
%%% o-0235
\startChapter[
  title={Chapter 30}
  ]

\Summary{How, and of what matter, the Altar of incenſe shal be made:
  12.~VVhat money shal be gathered for the vſe of the Tabernacle. 18.~A
  braſen lauer is alſo to be made, 25.~and holie oile of vnction.}

Thou shalt make alſo an Altar to burne incenſe, of the wood ſetim, \V
hauing a cubite of length, and an other of bredth, that is, foure
ſquare, and two cubites in height. The hornes shal procede out of the
ſame. \V And thou shalt plate it with the pureſt gold, as wel the grate
therof, as the walles round about, and the hornes. And thou shalt make
to it a 
crowne of gold round about, \V and two golden ringes vnder the crowne on
either ſide, that the barres may
%%% 0260
be put into them, and the Altar may be caried. \V The barres alſo them
ſelues thou ſhalt make of the wood ſetim, and shalt plate them with
gold. \V And thou shalt ſet the Altar againſt the veile, that hangeth
before the arke of teſtimonie before the propitiatorie wherwith the
teſtimonie is couered, where I wil ſpeake to thee. \V And Aaron ſhal
burne incenſe vpon it, ſwetely fragrant, in the morning. When he ſhal
dreſſe the lampes, he ſhal burne it: \V and when he ſhal place them at
euen, he ſhal burne incenſe euerlaſting before the Lord through your
generations. \V You shal not offer vpon it incenſe of an other
compoſition, nor oblation, and victime, neitheir ſhal you offer
libamentes. \V And Aaron shal pray vpon the hornes therof once a yeare,
with the bloud of that which was offered for ſinne, and shal pacifie
vpon it in your generations. It shal be moſt Holie to the Lord. \V And
our Lord ſpake to Moyſes, ſaying: \V When thou shalt take the ſumme of
the children of Iſrael according to their number, euerie one of them
shal giue a price for their ſoules to the Lord, and there shal be no
ſcourge among them, when they shal be reckened. \V And this shal euerie
one giue that paſſeth to the naming,
\SNote{That is, 7.~d. ob. Engliſh. For a ſicle of the Sanctuarie is
about 15.~d.}
halfe a ſicle according to the meaſure of the temple. A ſicle hath
twentee
\SNote{\L{Obolos}, 3.~farthings.}
\Fix{aboles.}{oboles.}{likely typo, fixed in other}
The halfe part of a ſicle shal be offered to the Lord. \V He
that is accounted in the number, for twentie yeares and vpward, shal
giue price. \V The rich man shal not adde to halfe a ſicle, and the
poore man ſhal diminiſh nothing. \V And the money being receiued, which
was contributed of the children of Iſrael, thou shalt deliuer vnto the
vſes of the tabernacle of teſtimonie, that it may be a monument of them
before the Lord, and he may be propitious to their ſoules. \V And our
Lord ſpake to Moyſes, ſaying: \V Thou ſhalt make alſo a lauer with his
foote of braſſe, to wash in: and thou shalt ſet it betwen the tabernacle
of the teſtimonie and the Altar.
%%% o-0236
And water being put into it, \V Aaron and his ſonnes ſhal wash therin
their handes and feete, \V when they are going into the tabernacle of
teſtimonie, and when they are to come vnto the Altar, to offer on it
incenſe to the Lord, \V leſt perhappes they die. It shal be an
euerlaſting law to him, and to his ſeede by ſucceſſions. \V And our Lord
ſpake to Moyſes, \V ſaying: Take ſpices, of principal and choſen myrrh
fiue hundred ſicles, and of cinnamon halfe ſo much, that is, two
%%% 0261
hundred fiftie ſicles, of calamus in like maner two hundred fiftie, \V
and of caſia fiue hundred ſicles after the weight of the Sanctuarie, of
oile of oliues the meaſure hin: \V and thou shalt make the holie oile of
vnction, an ointment compounded by the art of an vnguentarie, \V and
therof thou shalt anoynt the tabernacle of teſtimonie, and the arke of
the teſtament, \V and the table with the veſſel therof, the
candleſticke, and the furniture therof, the Altars of incenſe, \V and of
holocauſte, and al the furniture that perteyneth to the ſeruice of
them. \V And thou shalt ſanctifie al, and they shal be moſt Holie: he
that shal touch them, shal be ſanctified. \V Thou shalt anoynt Aaron and
his ſonnes, and shalt ſanctifie them, that they may doe the function of
prieſthood vnto me. \V To the children of Iſrael alſo thou shalt ſay:
This oile of vnction shal be holie vnto me through your generations. \V
The flesh of man shal not be anoynted therwith, and you shal make none
after the compoſition of it, becauſe it is ſanctified, and shal be holie
vnto you. \V What man ſoeuer shal compound ſuch, and shal giue therof to
a ſtranger, shal be abandoned out of his people. \V And our Lord ſaid to
Moyſes: Take vnto thee ſpices, ſtactee, and onycha, galbanum of ſwete
ſauour, and the cleareſt frankincenſe, al shal be of equal weight: \V
and thou shalt make incenſe compounded by the worke of an vnguentarie,
exactly tempered, and pure, and moſt worthie of ſanctification. \V And
when thou haſt beaten al into verie ſmal pouder, thou shalt ſet of it
before the tabernacle of teſtimonie, in the place where I wil appeare to
thee. Moſt Holie shal the incenſe be vnto you. \V Such confection you
shal not make vnto your owne vſes, becauſe it is holie to the Lord. \V
What man ſoeuer shal make the like, to enioy the ſmel therof, shal
perish out of his people.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
