%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 0233
%%% o-0213
\startChapter[
  title={Chapter 19}
  ]

\Summary{Nere to mount Sinai, with commemoration of their deliuerie from
  Ægypt, the people are commanded to be ſanctified: 16.~and ſo our Lord
  coming in thunders and lightnings ſpeaketh with Moyſes.}

%%% 0234
In the third moneth of the departure of Iſrael out of the Land of Ægypt,
\LNote{This day.}{The
\MNote{Agreement of old and new myſteries.}
firſt day of the third moneth the children of Iſrael came into the
deſert of Sinai. So counting 16.~dayes remaining of the firſt moneth
when they parted from Ægypt, al the ſecond moneth of 30.~dayes, this
firſt day of the third moneth, and three dayes more, in which they were
ſanctified by waſhing and other ceremonies
\XRef{(v.~10.)}
the Law was geuen the fifteth day, in figure of the Law of Chriſt,
promulgated on whitſunday, the fifteth day after our Redemption. VVherby
we ſee meruelous correſpondence of diuine Myſteries, in the old and new
Teſtament.
\Cite{S.~Auguſtin Epiſt.~119. c.~16.}}
this day they came into
\SNote{To this place (which was their 12.~manſion) they came the 47.~day
after they parted from Ægypt. And the third day folowĩg which was the
50, the law was geuen in mount Sinay.
\Cite{S.~Hierom. Epiſt.~1. ad Fabiolam.}}
the wilderneſſe of Sinai. \V For departing out of Raphidim, and coming
to the deſert of Sinai, they camped in the ſame place, and there Iſrael
pitched their tentes ouer againſt the mountaine. \V And Moyſes went vp to
God: and our Lord called him from the mountaine, and ſaid:
\SNote{God would haue their free conſent, els it were not a perfect
couenant.
\Cite{Theodoret q.~35. in Exod.}}
This ſhalt thou ſay to the houſe of Iacob, and ſhalt tel the children of
Iſrael: \V Your ſelues haue ſene what I haue done to the Ægyptians, how
I haue carried you vpon the winges of eagles, and haue taken you vnto
me. \V If therfore you wil heare my voice, and keepe my couenant, you
ſhal be
\SNote{In this couenant God promiſeth particular loue; Prieſtly
function, wherby they might better ſerue him; and effectual grace and
ſanctitie.}
my peculiar of al peoples: for al the earth is myne. \V And you ſhal be
vnto me
%%% !!! another referance to the SNote 'my peculiar of al peoples.
%%% how do we handle this?
%%% \SNote{}
a prieſtlie kingdome, and
%%% !!! another referance to the SNote 'my peculiar of al peoples.
%%% how do we handle this?
%%% \SNote{}
a holie nation: theſe are the wordes that thou ſhalt ſpeake to the
children of Iſrael. \V Moyſes came: and calling together the nations of
the people, he declared al the wordes which our Lord had commanded
him. \V And al the people anſwered together:
\SNote{The people promiſe loyaltie to God, and to keepe his
commandements.}
Al thinges that our Lord hath ſpoken, we wil doe. And when Moyſes had
reported the peoples wordes to our Lord, \V our Lord ſaid to him: Now
preſently wil I come to thee in the darkeneſſe of a cloude, that the
people may heare me ſpeaking to thee, and may beleue thee for
euer. Moyſes therfore
\SNote{So Angels & Saincts offer our prayers & other good workes to God,
though he know al things before hand.}
told the peoples wordes to our Lord. \V Who ſaid to him: Goe to the
people, and ſanctifie them to day, and to morow, and let them waſh their
garmentes. \V And let them be readie againſt the third day: for in the
third day the Lord wil deſcend in the ſight of al the people vpon the
mount Sinai. \V And thou ſhalt appoynt certaine limites to the people in
circuite, and ſhalt ſay to them: Beware ye aſcend not into the mount,
and that you touch not the endes therof: euerie one that toucheth the
mount, dying shal dye. \V Handes shal not touch him, but he shal be
ſtoned to death, or shal be shot through with arrowes: whether it be
beaſt, or man, it shal not liue. When the trumpet shal beginne to ſound,
then let them aſcend into the mount. \V And Moyſes came downe from the
mount to the people, and ſanctified them. And when they had washed their
garments, \V he ſaid to them: Be readie againſt the third day, and come
not neere your wiues. \V And
%%% o-0214
now the third day was come, and the morning appeared: and behold
thunders
\Fix{deganne}{beganne}{obvious typo, fixed in other}
to be heard,
%%% 0235
and lightenings to flash, and a verie thicke cloude to couer the mount,
and the noyſe of the trumpet ſounded exceedingly: and the people, that
was in the campe, feared. \V And when Moyſes had brought them forth to
mete with God from the place of the campe, they ſtoode at the botome of
the mount. \V And al the mount Sinai ſmoked: for becauſe our Lord was
deſcended vpon it in fyre, and the ſmoke aroſe from it as out of a
fornace: and al the mount was terrible. \V And the ſound of the trumpet
grew lowder by litle and litle, and was drawen out a length: Moyſes
ſpake, and God anſwered him. \V And our Lord deſcended vpon the mount
Sinai in the very toppe of the mount, and he called Moyſes into the
toppe therof. Whither when he was aſcended, \V he ſaid
vnto him: Goe downe, and charge the people: leſt perhaps they wil paſſe
their limittes to ſee the Lord, and a very great multitude of them
perish. \V The prieſtes alſo that come to the Lord, let them be
ſanctified, leſt he ſtrike them. \V And Moyſes ſaid to our Lord: The
common people can not aſcend into the mount Sinai: for thou didſt charge,
and command, ſaying: Put limittes about the mount, and ſanctifie it. \V
To whom our Lord ſaid: Goe, gette thee downe and thou shalt come vp, &
Aaron with thee: but
\SNote{The people and al inferior clergie alſo, are to kepe their
limites and to lerne Gods wil of their ſuperiors.}
the prieſts and the people let them not paſſe the limittes, nor aſcend
to the Lord, leſt perhappes he kil them. \V And Moyſes went downe to the
people, and told them al.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
