%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-23


\project douay-rheims


%%% 0244
%%% o-0222
\startChapter[
  title={Chapter 23}
  ]

\Summary{Lawes are appointed to Iudges, (the enemies oxe, or aſse to be
  ſaued) 8.~namely not to take bribes. 10.~The ſeuenth yeare, and day al
  muſt reſt. 14.~Three principal feaſtes muſt be ſolemnized euerie
  yeare. 20.~Conduction and protection of an Angel is promiſed. 24.~The
  people is agane commanded to deſtroy Idols. 29.~VVhy their enemies
  shal be deſtroyed by litle and litle.}

Thou shalt not admitte a lying voice: neither shalt thou ioyne thy hand
to ſay falſe teſtimonie for a wicked perſon. \V Thou shalt not folow the
multitude to doe euil: neither shalt thou in iudgement, agree to the
ſentence of the moſt part, to ſtray from the truth. \V The
\SNote{Al vertues being foũded in iuſtice, ceaſe to be true vertues,
when iuſtice is not firſt obſerued.
\Cite{S.~Hierom in Pſal.~32.}
\L{et}
\Cite{Prou.~31.}}
poore man
%%% 0245
alſo thou ſhalt
%%% o-0223
not pitie in iudgement. \V If thou meete thy enemies
oxe, or aſſe going aſtray, bring it backe to him. \V If thou ſee the
aſſe of him that hateth thee lye vnderneth his burden, thou ſhalt not
paſſe by, but shalt lift him vp with the ſame. \V Thou shalt not decline
the poore mans iudgement. \V A lye thou shalt auoide. The innocent and
iuſt perſon thou shal not put to death: becauſe I abhorre the impious
man. \V Neither ſhalt thou take bribes, which doe blind alſo the wiſe,
and peruert the wordes of the iuſt. \V The ſtranger thou ſhalt not
moleſt, for you know the hartes of ſtrangers: becauſe your ſelues alſo
were ſtrangers in the Land of Ægypt. \V Six yeares thou shalt ſow thy
ground, and ſhalt gather the corne therof. \V But the ſeuenth yeare thou
shalt let it alone, and make it to reſt, that the poore of thy people
may eate, and whatſoeuer shal be leift, let the beates of the field eate
it: ſo ſhalt thou doe in thy vineyard and thy oliuete. \V Six dayes thou
shalt worke: the ſeuenth day thou shalt ceaſe, that thy oxe may reſt and
thine aſſe: and the ſonne of thy handmaide may be refreshed, and the
ſtranger. \V Kepe al thinges that I haue ſaid to you. And by the name of
foren goddes you shal not ſweare, neither shal it be heard out of your
mouth. \V
\MNote{Three principal feaſtes beſides the Sabbath, & ſome others.}
Three times euerie yeare you shal celebrate feaſtes to me. \V Thou shalt
keepe
\SNote{Paſch in memorie of their deliuerie from Ægypt.}
the ſolemnitie of Azymes. Seuen dayes shalt thou eate azymes, as I
commanded thee, in the time of the moneth of new corne, when thou didſt
come forth out of Ægypt: thou ſhalt not appeare in my ſight emptie. \V
And the ſolemnitie of the harueſt
\SNote{Pentecoſt, when they receiued the Law.}
of the firſt fruites of thy worke, whatſoeuer thou didſt ſow in the
field. The ſolemnitie alſo in the end of the yeare,
\SNote{Tabernacles in memorie of Gods protection fourtie yeares in the
deſert.}
when thou haſt gathered al thy corne out of the field. \V Thriſe a yeare
shal al thy male ſexe appeare before the Lord thy God. \V Thou shalt not
ſacrifice the bloud of my victime vpon leuen, neither shal the fatte of
my ſolemnitie remaine vntil the morning. \V The firſt fruites of the
corne of thy ground thou shalt carrie into the houſe of the Lord thy
God. Thou shalt not boyle a kidde in the milke of his damme. \V Behold I
wil ſend myne Angel, which shal goe before thee, and keepe thee in thy
iourney, and bring thee into the place that I haue prepared. \V Obſerue
him, and heare his voice, neither doe thou thinke him one to be
contemned: for he wil not forgeue when thou haſt ſinned, and
%%% 0246
my name is in him. \V But if thou wilt heare his voice, and doe al that
I ſpeake, I wil be enemie to thyne enemies, & wil afflict them that
afflict thee. \V And myne Angel ſhal goe before thee, and ſhal bring
thee in vnto the Amorrheite, and Hetheite, and Pherezeite, and
Chananeite, and Heueite, and Iebuzeite, whom I wil deſtroy. \V Thou
ſhalt not adore their goddes, nor ſerue them. Thou shalt not doe their
workes, but shalt deſtroy them, and breake their ſtatues. \V And you
shal ſerue the Lord your God, that I may bleſſe your bread & waters, and
may take away infirmitie from the middes of thee. \V There shal not be a
fruitleſſe nor barren bodie in thy land: I wil fil the number of thy
dayes. \V I wil ſend my terrour to runne before thee,
%%% o-0224
and wil kil al
people, to whom thou shalt enter: and wil turne the backes of al thyne
enemies before thee: \V ſending forth hornets before, that shal chaſe
away the
\Fix{Herueite,}{Heueite,}{obvious typo, fixed in other}
and Chananeite, and Hetheite,
before thou enter. \V I wil not caſt them out from thy face in one
yeare: leſt the land be brought into a wilderneſſe, and beaſtes encreaſe
againſt thee. \V By litle and litle I wil expel them from thy ſight, til
thou be increaſed, and doſt poſſeſſe the Land. \V And I wil ſette thy
boundes from the Redde ſea vnto the ſea of the Paleſtines, and from the
deſert vnto the riuer: I wil deliuer the inhabitantes of the Land in
your handes, and wil caſt them out from your ſight. \V Thou shalt
\SNote{Peace with infidels forbidden to Gods people.}
not enter league with them, nor with their goddes. \V Let them not dwel
in thy land, leſt perhapes they make thee to ſinne againſt me, if thou
ſerue their goddes: which vndoubtedly wil be a ſcandal to thee.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
