%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 34
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-34


\project douay-rheims


%%% 0267
%%% o-0242
\startChapter[
  title={Chapter 34}
  ]

\Summary{Moyſes goeth againe into Mount Sinai, vvith new tables, praying
  for the people, 10.~to whom God promiſeth to giue poſſeſsion of the
  Land. 12.~Prohibiteth al aſsociation vvith the Gentiles, for feare of
  Idolatrie, 18.~geueth precepts concerning the firſt borne, the
  Sabbath, and other feaſtes. 28.~After fourtie dayes faſt, Moyſes
  returneth to the people with the commandements, and his face appearing
  horned, he couereth it, whenſoeuer he ſpeaketh to the people.}

And after this he ſaid:
\SNote{The firſt tables being broken, yet others are made: ſo though the
firſt grace geuen in Baptiſme be loſt, yet there remaineth penance, as the
ſecond table of ſaftie after ſhipwrake.
\Cite{S.~Hiero. Epiſ. ad Demedriad.}}
Cut thee two tables of ſtone like vnto the former, and I wil write vpon them
the wordes, which the tables had, which thou haſt broken. \V Be readie
in the morning, that thou maieſt forwith go vp into the mount
Sinai, and thou ſhalt ſtand with me vpon the toppe of the
mount. \V Let no man go vp with thee, neither let anie man be ſene
throughout the whole mount: the oxen
%%% 0268
alſo and the sheepe let them not feede ouer againſt. \V He cut out
therfore two tables of ſtone, ſuch as had bene before: and riſing very
early he went vp into mount Sinai, as our Lord had commanded him,
carying with him the tables. \V And when our Lord was deſcended in a
cloude, Moyſes ſtoode with him, calling vpon the name of our Lord. \V Who
paſſing before his, he ſaid:
\TNote{Chief ruler.}
Dominatour Lord God, merciful and clement, patient and of much
compaſſion, and true, \V which keepeſt mercie vnto thouſandes: which
takeſt away iniquitie, and wicked factes, and ſinnes, and no man of him
ſelfe is innocent before thee. Which doeſt render the iniquitie of the
fathers to the children, and to the nephewes vnto the third and
%%% o-0243
fourth generation. \V And Moyſes making haſt, bowed flatte vnto the
earth, and adoring \V he ſaid: If I haue found grace in thy ſight o
Lord, I beſeech thee that thou wilt goe with vs (for it is a ſtiffe
necked people) and take away our iniquities and ſinnes, and poſſeſſe
vs. \V Our Lord anſwered:
\SNote{Notwithſtanding his former commination
\XRef{chap.~33. v.~3.}
God here promiſeth new benefites.}
I wil make a couenant in the ſight of al, I wil do ſignes that were
neuer ſene vpon the earth, nor in anie nations: that this people may
ſee, in the middes of whom thou art, the terrible worke of the Lord
which I wil doe. \V Obſerue al thinges which this day I command thee: I
my ſelf wil caſt out before thy face the Amorrheite, and Chananeite, and
Hetheite, the Pherezeite alſo, and Heueite, and Iebuſeite. \V Beware
thou neuer ioyne amitie with the inhabitants of that land, which may be
thy ruine: \V but deſtroy their altars, breake their ſtatues, and cut
downe their groues: \V adore not a ſtrange God. The Lord his name is
Ielouſe, God is an emulatour. \V Enter no traffick with the men of thoſe
regions: leſt, when they haue fornicated with their goddes, and haue
adored their idols, ſome man cal thee to eate of the thinges
immmolated. \V Neither ſhalt thou take a wife for thy ſonnes of their
daughters: leſt after them ſelues haue fornicated, they make thy ſonnes
alſo to fornicate with their goddes. \V Molten goddes thou shalt not
make to thee. \V Thou ſhalt keepe the ſolemnitie of the azymes. Seuen
dayes ſhalt thou eate azymes, as I haue commanded thee, in the time of
the moneth of new corne: for in the moneth of ſpring time thou didſt goe
out of Ægypt. \V Al of the male kind, that openeth the matrice, shal be
mine. Of al beaſtes as wel oxen as of ſheepe, it ſhal be mine. \V The
firſtborne of an aſſe
%%% 0269
thou ſhalt redeeme with a sheepe: but if thou wilt not giue a price for
it, it ſhal be ſlaine. The firſtborne of thy ſonnes thou shalt redeeme:
neither shalt thou appeare emptie in my ſight. \V Six dayes ſhalt thou
worke, the ſeuenth day thou ſhalt ceaſe to eate, and reape. \V The
ſolemnitie of weekes thou shalt make to thee, in the firſt fruites of
corne of thy wheate harueſt, and the ſolemnitie, when the time of the
yeare returneth that al thinges are laid vp. \V Three tymes of the yeare
al thy male shal appeare in the ſight of the omnipotent Lord God of
Iſrael. \V For when I ſhal haue taken away the nations from thy face,
and shal haue dilated thy borders, no man shal lie in wayte againſt thy
land, when thou doeſt goe vp, and appeare in the ſight of the Lord thy
God thriſe in a yeare. \V Thou shalt not immolate the bloud of my hoſte
vpon leauen: neither ſhal there remaine in the morning of the victime of
the ſolemnitie of the Phaſe. \V The firſt of the fruites of thy ground
thou shalt offer in the houſe of the Lord thy God. Thou shalt not boile
a kidde in the milke of his damme. \V And our Lord ſaid to Moyſes: Write
thee theſe wordes, in which I haue made a couenant both with thee and
with Iſrael. \V Therfore he was there with our Lord fourtie dayes and
fourtie nightes: he did not eate bread, and he dronke no water, and
\SNote{God by an Angel not Moyſes. \L{ſupra}
\XRef{v.~1.}
&
\XRef{Deut.~10. v.~2. &~4.}}
he wrote in the tables the wordes of the couenant,
\SNote{Howſoeuer the commandments are diuided in both tables, here it is
certaine, that there be no more nor fewer then tenne in al.}
tenne. \V And when Moyſes came downe from the mount Sinai, he held the
two tables of teſtimonie,
%%% o-0244
and he knew not that his face was
\SNote{So his face appeared to the beholders, by reaſon of the
gliſtering beames of his countenance ſhining gloriouſly, after his
conuerſation with God fourtie dayes: which ſignifieth that \Emph{much
more that which abideth} (in al eternitie) \Emph{is in glorie}.
\XRef{2.~Cor.~3.}}
horned by the conuerſation of the talke of our Lord. \V And Aaron and
the children of Iſrael ſeing the face of Moyſes horned, they were afraid
to come neere. \V And being called of him, they returned as wel Aaron as
the princes of the ſynagogue. And after that he ſpake to them, \V al the
children of Iſrael alſo came to him: whom he commanded al thinges that
he had heard of our Lord in mount Sinai. \V And hauing ended his talke,
he put
\SNote{The ſame veile (ſaieth S.~Paul) remaneth vpon the hart of the
Iewes, that they can not ſee Chriſt, til by his ſpecial grace they ſhal
be illuminated:
\XRef{2.~Cor.~3.}
The like is vpon Heretikes that can not ſee the Church.
\Cite{S.~Aug. in Pſal.~30. con.~2.}}
a veile vpon his face. \V Which going in to our Lord, and ſpeaking with
him, he toke away vntil he went forth, and then he ſpake to the children
of Iſrael al thinges that had bene commanded him. \V Who ſay that the
face of Moyſes coming forth was horned, but he couered his face againe,
if at anie time he ſpake to them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
