%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 26
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-26


\project douay-rheims


%%% 0250
%%% o-0227
\startChapter[
  title={Chapter 26}
  ]

\Summary{The forme of the Tabernacle, with the appertinances, and of
  vvhat matter, number, and qualities al things shal be.}

And the tabernacle thou shalt make thus: Tenne curtines shalt thou make
of twiſted ſilke, and hyacinth, and purple, and ſcarlet twiſe died,
varied with imbrodered worke. \V The length of one curtine shal haue
twentie eight cubites: the bredth, shal be of foure cubites. Al the
curtines shal be of one meaſure. \V Fiue curtines shal be
\SNote{Chriſts members by their vnion & communitie aſſiſt ech other, and
adorne his tabernacle, the Church.}
ioyned one to an other, and the other fiue shal hang together with the
connexion. \V Loupes of hyacinth thou shalt make in the ſides and toppes
of the
\Fix{certines,}{curtines,}{possible typo, fixed in other}
that they may be compled one to an other. \V Fiftie loupes shal euerie
curtine haue on both ſides, ſo ſet on, that one loupe may be againſt
another loupe, and one may be fitted to the other. \V Thou shalt make
alſo fiftie circles of gold wherwith the veiles of the curtines are to
be ioyned, that it may be made one tabernacle. \V Thou shalt make alſo
eleuen curtines of haire, to couer the toppe of the tabernacle. \V The
length of one haire curtine shal haue thirtie cubites: and the bredth,
foure: the meaſure of al the curtines shal be equal. \V Of the which,
fiue thou shalt ioyne apart, and the ſix thou shalt couple one to an
other, ſo that the ſixt curtine in the front of the roofe thou shalt
duble. \V Thou shalt make alſo fiftie loupes in the edge of one
%%% 0251
curtine, that it may be ioyned with the other: and fiftie loupes in the
edge of the other curtine, that it may be coupled with his felow. \V
Thou shalt make alſo fiftie buckles of braſſe, wherwith the loups may be
ioyned, that of al there
%%% o-0228
may be made one couering. \V And that which shal remaine in the
curtines, that are prepared for the roofe, to witte, one curtine that is
ouerplus, with the halfe therof thou ſhalt couer the backeſide of the
tabernacle. \V And there ſhal hang downe a cubite on the one ſide, and
an other on the other ſide, which is the ouerplus in the length of the
curtines, ſenſing both ſides of the tabernacle. \V Thou shalt make alſo
an other couer to the roofe of rammes skinnes died redde: and ouer that
againe an other couer of ianthine skinnes. \V Thou shalt make alſo the
bordes of the tabernacle ſtanding vpright of the wood ſetim, \V of the
which let euerie one haue ten cubites in length, and in bredth one and
an halfe a peece. \V In the ſides of the borde, shal be made two
morteſes, wherby one borde may be ioyned to an other borde: and after
this maner shal al the bordes be prepared. \V Of the which twentie ſhal
be in the ſouth ſide that tendeth Southward. \V For the which thou shalt
caſt fourtie feete of ſiluer, that there may two feete be put vnder
euerie borde at the two corners. \V In the ſecond ſide alſo of the
tabernacle that looketh to the North, there shal be twentie bordes, \V
hauing fourtie feete of ſiluer, two feete shal be put vnder euerie
borde. \V But on the weſt quarter of the tabernacle thou ſhalt make ſix
bordes, \V and againe other two which ſhal be erected in the corners at
the backe of the tabernacle. \V And they ſhal be ioyned together from
beneth vnto the toppe, and one ioynture shal hold them al. The like
ioynture shal be kept for the two bordes alſo that are to be put in the
corners. \V And they shal be in al eight bordes, their ſiluer feete
ſixtene, two feete accounted for euerie borde. \V Thou shalt make alſo
fiue barres of the wood ſetim, to hold together the bordes on the one
ſide of the tabernacle, \V and fiue others on the other ſide, and as
manie at the weſt ſide: \V which shal be put along by the middes of the
bordes from one end to the other. \V The bordes alſo them ſelues thou
shalt plate with gold, and shalt caſt ringes of gold to be ſette vpon
them, through which the barres may hold together the bordeworke: the
which thou ſhalt couer with plates of
%%% 0252
gold. \V And thou ſhalt erect the tabernacle according to the paterne
that was ſhewed thee in the Mount. \V Thou ſhalt make alſo a veile of
hyacinth, and purple, and ſcarlet twiſe died, and twiſted ſilke, wrought
with imbrodered worke and goodlie varietie: \V which thou ſhalt hang
before foure pillers of the wood ſetim, the which themſelues alſo ſhal
be plated with gold, and ſhal haue foure heades of gold, but feete of
ſiluer. \V And the veile shal be hanged on with ringes, within the which
thou ſhalt put the arke of teſtimonie, with the which alſo the
Sanctuarie, and the ſanctuaries of the Sanctuarie, shal be diuided. \V
Thou shalt ſet alſo the Propitiatorie vpon the arke of teſtimonie in the
\SNote{The chieffeſt part of the Tabernacle, called \L{Sancta ſanctorum}:
\Emph{Holie of holies}.}
\L{Sancta ſanctorum}: \V and the table without the veile: and ouer
againſt the table the candleſticke in the ſouth ſide of the tabernacle:
for the table shal ſtand in the north ſide. \V Thou shalt make alſo a
hanging in the entring of the tabernacle of hyacinth, and purple, and
ſcarlet twiſe died, and twiſted 
%%% o-0229
ſilke, with imbrodered worke. \V And
fiue pillers of the wood ſetim thou shalt plate with gold, before the
which the hanging shal be drawen: whoſe heades shal be of gold, and
feete of braſſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
