%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 32
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-32


\project douay-rheims


%%% 0262
%%% o-0238
\startChapter[
  title={Chapter 32}
  ]

\Summary{The people (Aaron conſenting) make & adore the image of a
  calfe. 7.~VVhich God reueiling to Moyſes, 11.~he praieth our Lord, for
  Abraham, Iſaac, and Iacobs ſake to ſpare the people, and performe his
  promiſe.
%%% 0263
14.~VVherwith God is pacified. 15.~Yet Moyſes coming from the Mount, and
ſeeing the calfe, and idolatrie, throweth downe the tables and breaketh
them, 20.~deſtroyeth the idol, 21.~blameth Aaron, 27.~cauſeth manie
Idolaters to be ſlaine, 31.~and againe prayeth for the people.}

And the people ſeeing that Moyſes made tariance ere he came downe from
the mount, being aſſembled againſt Aaron, they ſayd: Ariſe, make vs
\SNote{Aaron knew what goddes they ment, to wit, ſuch as they had ſenne
worſhipped in Ægypt, and therfore he made them a molten calfe.
\XRef{v.~4.}}
goddes, that may goe before vs: for what hath chanced to this Moyſes the
man that brought vs out of the Land of Ægypt, we know not. \V And Aaron
ſayd to them: Take the golden earlettes from the eares of your wiues,
and ſonnes and daughters, & bring them to me. \V And the people did that
he had cõmanded, bringing the earlettes to Aaron. \V Which when he had
receiued, he formed them by founders worke, and made of them
\LNote{A molten calfe.}{No
\MNote{This people thought the calfe to be the true God.}
other reaſon can be imagined, why the people falling to Idolatrie,
required the image of a calfe, rather then of anie other thing, but for
that
\CNote{\Cite{S.~Aug. li.~18. c.~5. ciuit.}}
they thought the blacke calfe with white ſpottes, called \Emph{Apu},
or \Emph{Serapis}, whom they ſawe the Ægyptians  eſtemed moſt of al their
goddes, to be the chiefe, or onlie God. And therfore to this famous
Idol, they aſcribed the benefite of their deliuerie from bondage,
ſaying: \Emph{Theſe are thy goddes, o Iſrael, that haue brought thee out
of the Land of Ægypt.}
\MNote{They adored that which the image repreſented.}
So they ment not to worſhip our Lord, the true God, in that image, as
\CNote{\Cite{li.~1. c.~11. para.~9. Inſtit.}}
Caluin would haue it, but the very calfe whoſe image it was, for adoring
immediatly the calues image, \Emph{and immolating hoſtes to it},
\XRef{(v.~8.)}
they proteſted that to be their God, which the image repreſented. This
appeareth alſo,
\XRef{Deut.~32. v.~18.}
\Emph{God that begat thee thou haſt forſaken: and haſt forgotten our
Lord thy creatour.} And
\XRef{Pſal.~105. v.~21.}
\Emph{They forgot God, vvhich ſaued them.}}
a molten calfe. And they ſayd: Theſe are thy goddes Iſrael, that haue
brought thee out of the land of of Ægypt. \V Which when Aaron had ſenne, he
builded an altar before it, and by a cryers voice proclaimed ſaying: To
morow is the ſolemnitie of the Lord. \V And ryſing in the morning, they
offered holocauſtes, and pacifique hoſtes, and the people ſate downe to
eate, and to drinke, and they roſe vp
\SNote{Exceſſe in play called fooliſh mirth, is the daughter of
gluttonie, and mother of Idolatrie.
\Cite{S.~Greg. li.~31. c.~31. Moral.}}
to play. \V And our Lord ſpake to Moyſes, ſaying: Goe, get thee downe:
thy people, which thou haſt brought out of the Land of Ægypt, hath
ſinned. \V They haue quickly reuolted from the way, that thou didſt ſhew
them: and they haue made to them ſelues a molten calfe, and haue adored,
and immolating hoſtes vnto
\SNote{To the molten calfe, which they had made.}
it, haue ſayd: Theſe are thy goddes Iſrael, that haue brought thee out
of the Land of Ægypt. \V And againe our Lord ſaid to Moyſes: I ſee that
this people is ſtiffenecked: \V
\SNote{God ſaying, \Emph{Suffer me}, ſignifieth that he could be
hindered.
\Cite{S.~Hierom. in Ione.~1.}}
ſuffer me, that my furie may be angrie againſt them, and that I may
deſtroy them, and I wil make thee into a great nation. \V But
\LNote{Moyſes beſought.}{Albeit
\MNote{Caluin chargeth Moyſes with arrogancie.}
Moyſes with moſt humble ſubmiſſion prayed for the people, which God ſo
accepted, that he was therby pacified,
\XRef{(v.~14.)}
yet
\CNote{\Cite{In hunc locum.}}
Caluin here condemneth him of arrogancie, and pride, as though he
imperiouſly preſcribed law to God, ſpoyling him of his iuſtice.
\MNote{Moyſes charitie concurred with Gods prouidence.}
Much otherwiſe S.~Hierom
\Cite{(Epiſt.~12. ad Gaudent.)}
commending his feruent charitie, doubteth not to ſay, \L{Dei potentiam
ſerui preces impediebant.} \Emph{The ſeruantes prayers hindered Gods
povver}, becauſe God himſelf ſaying: \Emph{ſuffer me, that my furie may
be angrie againſt them, and that I may deſtroy them}: ſhewed his diuine
prouidence to be ſuch, as \Emph{he might be ſtaied, from doing that
vvhich he threatned}.}
Moyſes beſought the Lord his God, ſaying: Why Lord, is thy furie angrie
againſt thy people, whom thou haſt brought forth of the Land of Ægypt,
in great power, and in a ſtrong hand? \V Let not the Ægyptians ſay I
beſeech thee: He hath craftely brought them forth, that he might kil
them in the mountaynes, and deſtroy them from the earth: let thyne anger
ceaſe, and be pacified vpon the wickednes of thy people. \V
\LNote{Remember Abraham.}{It
\MNote{God ſheweth mercie for the merites of his ſeruants.}
much trubled Caluin, that for obtaining pardon for the people, the
Patriarches are mentioned, for whoſe ſake and merites, mercie, and
protection was promiſed by God,
\XRef{(Gen.~18.}
\XRef{22.}
\XRef{26.)}
prophecied by Iacob,
\XRef{(Gen.~48.)}
performed here, and in manie other places. And it is a vaine euaſion to
ſay: God ſhewed his mercie for his promiſe only; 
%%% Do theſe go here?
\CNote{\Cite{S.~Aug. q.~149. in Exod.}
\Cite{S.~Chryſ. ho.~42. in Gen.}
\Cite{Theod. q.~67. in Exo.}}
for he promiſed the ſame for their merites; as appeareth in the places
alleaged.
\MNote{Grace goeth before merites.}
Though al merites procede from Gods grace, firſt geuen without merites.
\XRef{1.~Cor.~4. v.~7.}
\Cite{S.~Aug. de grat. & lib. arb. c.~6.}}
Remember Abraham, Iſaac, and Iſrael
\SNote{Not only Gods promiſe, but alſo his ſeruants merites are here
propoſed for procuring mercie to the people.
\XRef{See the Annotation.}}
thy ſeruantes,
%%% 0264
to whom thou ſwareſt by thine owne ſelf, ſaying: I wil multiplie your
ſeed as the ſtarres of heauen: and this whole land, wherof I haue
ſpoken, I wil giue to your ſeed,
%%% o-0239
and you ſhal poſſeſſe it alwayes. \V And our Lord was pacified from
doing the euil which he had ſpoken againſt his people. \V And Moyſes
returned from the mount, carying the two tables of teſtimonie in his
hand, written on both ſides, \V and made by the worke of God: the
writing alſo of God was grauen in the tables. \V And Ioſue hearing the
tumult of the people crying out, ſaid to Moyſes: The noyſe of battaile
is heard in the campe. \V Who anſwered: It is not the crie of men
encouraging of fight, nor the ſhoute of men compelling to flee: but I
doe heare the voice of ſingers. \V And when he approched to the campe,
he ſaw the calfe, and the daunces: and being
\SNote{Moyſes the meekeſt man on earth
\XRef{(Nu.~12.)}
in Gods cauſe was moſt zelous againſt ſinne.
\Cite{S.~Aug. q.~144. in Exod.}}
very wrath, he threw the tables out of his hand, and brake them at the
foote of the mount. \V And catching the calfe which they had made, he
burnt it, and bette it into powder, which he ſtrawed into water, and
gaue therof drinke to the children of Iſrael. \V And he ſaid to Aaron:
What hath this people done to thee, that thou ſhouldeſt bring vpon them
an heinous ſinne? \V To whom he anſwered: Let not my lord be offended:
for thou knoweſt this people, that it is prone to euil: \V they ſaid to
me: Make vs goddes, that may goe before vs: for vnto this ſame Moyſes,
that brought vs forth out of the Land of Ægypt, we know not what is
chanced. \V To whom I ſaid: Which of you hath gold? They tooke, and
brought to me: and I caſt it into the fire, and
\SNote{Aaron confeſſed the fault briefly, not intending a friuolous
excuſe: for he could not thinke, but Moyſes knew the truth.
\Cite{S.~Aug. q.~145. in Exod.}}
this calfe came forth. \V Moyſes therfore ſeeing the people that they
were naked (for Aaron had ſpoiled them for the ignominie of filth, and
had ſet them naked among their enemies) \V and ſtanding in the gate of
the campe, he ſaid: If any man be our Lords, let him ioyne to me. And
there gathered vnto him al the ſonnes of Leui: \V to whom he ſaid: This
ſaith the Lord God of Iſrael: Put euerie man his ſword vpon his thigh:
goe, & returne from gate to gate through the middes of the campe, and
euerie man kil his brother, and frend, and neighbour. \V And
\SNote{Their zeale vſed with authoritie and order here rewarded: which
otherwiſe wanting, when Simeon and Leui ſlew the Sichemites, was blamed
by Iacob.
\XRef{Gen.~34.}
&
\XRef{49.}}
the ſonnes of Leui did according to the ſaying of Moyſes, and there were
ſlaine in that day about three thouſand men. \V And Moyſes ſaid: You
haue conſecrated your handes this day to our Lord, euerie man in his
ſonne & in his brother, that bleſſing may be giuen to you.
%%% 0265
\V And when the next day was come, Moyſes ſpake to the people: You haue
ſinned a verie great ſinne: I wil goe vp to our Lord, if by anie meanes
I may be able to intreate him for your ſinful fact. \V And returning to
our Lord, he ſaid: I beſeech thee: this people hath ſinned a heinous
ſinne, and they haue made to them ſelues goddes of gold: either forgiue
them this treſpaſſe, \V or if thou doe not,
\SNote{Moyſes not content with his owne ſaluation, would rather periſh
with the people, then they ſhould al be deſtroied: and therfore at his
inſtance God pardoned them.
\Cite{S.~Hiero. Ep.~12. ad Gaud.}
&
\Cite{in Ione.~1.}
\Cite{S.~Aug. q.~147. in Exod.}}
ſtrike me out of the booke that thou haſt written. \V To whom our Lord
anſwered: He that hath ſinned to me, him wil I ſtrike out of my
booke: \V but goe thou, and leade this people whither I haue told thee:
myne Angel ſhal goe before thee. And I in the day of reuenge wil viſite
this ſinne alſo of theirs. \V Our Lord therfore ſmote the people for the
fault concerning the calfe, which Aaron had made.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
