%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 37
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-37


\project douay-rheims


%%% 0273
%%% o-0247
\startChapter[
  title={Chapter 37}
  ]

\Summary{Beſeleel maketh the Arke, 6.~the Propitiatorie, with
  Cherubimes, 10.~the Table, vvith veſsel belonging therto, 17.~the
  Candleſticke vvith bowles and branches, 23.~ſeuen lampes with
  ſnuffers, 25.~the Altar of incenſe, 29.~and compoundeth the incenſe.}

%%% 0274
And Beſeleel made alſo the arke of the wood ſetim, hauing two cubites and
an halfe in length, and a cubite and an halfe in bredth, the height alſo
was of one cubite and an halfe: and he plated it with the pureſt gold
within and without. \V And he made to it a crowne of gold round
about, \V caſting foure ringes of gold at the foure corners
%%% o-0248
thereof: two ringes in the one ſide, and two in the other. \V Barres
alſo he made of the wood ſetim, which he plated with gold, \V and which
he put into the ringes, that were at the ſides of the arke to carie
it. \V He made alſo the
\Fix{Propititorie,}{Propitiatorie,}{obvious typo, fixed in other}
that is, the Oracle, of the pureſt gold, two cubites and an halfe in
length, and a cubite and an halfe in bredth. \V Two Cherubins alſo of
beaten gold, which he ſette on either ſide of the Propitiatorie: \V One
Cherub in the toppe of one ſide, and the other Cherub in the toppe of
the other ſide: two Cherubins in each toppe of the
\Fix{Propitirtorie,}{Propitiatorie,}{obvious typo, fixed in other}
\V ſpreading their winges, and
\SNote{The Cherubins couering al vpon and within the arke ſignifie
(ſaith S.~Gregorie Nyſſen) that the Scriptures haue a hiegher ſenſe then
the literal.
\Cite{de vita Moyſeos poſt medium.}}
couering the Propitiatorie, and looking one toward the other and toward
it. \V He made alſo the table of the wood ſetim in length two cubites,
and in bredth one cubite, which had in height a cubite & an halfe. \V
And he did compaſſe it with the fineſt gold, and he made to it a
golden ledge round about, \V and to the ledge it ſelfe a golden crowne
enterpolished of foure fingers, and vpon the ſame an other golden
crowne. \V And he caſt foure ringes of gold, which he put in the foure
corners at euerie foote of the table \V againſt the crowne: and he put
the barres into them, that the table might be caried. \V The barres alſo
them ſelues he made of the wood ſetim, and compaſſed them with gold. \V
And the veſſel for the diuers vſes of the table, ſawcers, phiales, and
goblettes, and cenſars, of pure gold, wherein the libamentes are to
be offered. \V He made alſo the candleſticke beaten of the fineſt
gold. From the shaft wherof the branches, cuppes, and boules and lilies
did proceede: \V ſix on both ſides, three branches on one ſide, and
three on the other: \V three cuppes in maner of a nutte on euerie
branch, and boules withal and lilies: and three cuppes of the faſhion of
a nutte in an other branch, and boules withal and lilies. The worke of
the ſix branches, that proceded from the ſhaft of the candleſticke,
equal. \V And in the shaft it ſelfe were foure cuppes after the maner of
a nutte, and boules withal at euerie one and lilies: \V and boules vnder
two branches in
%%% 0275
three places, which together make ſix branches proceding from one
ſhaft. \V Both the boules therfore, & the branches were out of it, al
beaten of the pureſt gold. \V He made alſo the ſeuen lampes with their
ſnuffers, and the veſſel, where the ſnuffings ſhould be put out, of moſt
pure gold. \V The candleſticke withal the veſſel therof did weigh a
talent of gold. \V He made alſo the altar of incenſe of the wood ſetim,
hauing a cubite euerie way foure ſquare, and in height two: from the
corners wherof the hornes did procede. \V And he plated it with the
pureſt gold, with the grate and the walles and the hornes. \V And he
made to it a crowne of gold round about, and two golden ringes vnder the
crowne at either ſide, that the barres may be put into them, and the
altar may be caried. \V And the barres them ſelues he made alſo of the
wood ſetim, and couered them with plates of gold. \V He compounded alſo
oile for the ointment of ſanctification, and incenſe of the pureſt
ſpices with the worke of a pigmentarie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
