%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 0203
%%% o-0186
\startChapter[
  title={Chapter 11}
  ]

\Summary{God biddeth Moyſes cauſe the people of Iſrael to borow ſiluer
  and gold veſſels of the Ægyptians. 4.~Fortelleth one other plague, the
  death of the firſt-borne, 9.~and that Pharao wil ſtil be obdurate.}

And our Lord ſaid to Moyſes: Yet with one plague more wil I touch Pharao
& Ægypt, and after this he ſhal diſmiſſe you, and compel you to goe
forth. \V Thou shalt ſay therfore to al the people that euerie man aske
of his frend, & euery woman of her neighbour veſſels of ſiluer, & of
gold. \V And the Lord wil geue grace to his people in the ſight of the
Ægyptians. And Moyſes was a very great man in the Land of
%%% 0204
Ægypt, in the ſight of Pharaoes ſeruantes, & of al the people. \V And he
ſaid: This ſaith our Lord: At midnight I wil enter into Ægypt: \V and
\MNote{The 10.~plague. Death of the firſtborne in men & beaſtes of the
Ægyptians.}
euerie firſt-begotten in the Land of the Ægyptians ſhal dye, from the
firſt-begotten of Pharao who ſitteth in his throne, euen to the
firſt-begotten of the handmaid that is at the mil, & al the
firſt-begotten of beaſtes. \V And there shal be a great crie in the
whole Land of Ægypt, ſuch as neither hath bene before, nor ſhal be
afterward. \V But with al the children of Iſrael there shal not a dogge
mutter, from man euen to beaſt: that you may know with how great a
miracle our Lord doth diuide the Ægyptians & Iſrael. \V And al theſe thy
ſeruantes shal come downe to me, and shal adore me, ſaying: Goe forth
thou, & al the people that is vnder thee: after this we shal goe
forth. \V And he departed from Pharao exceeding angrie. And our Lord
ſaid to Moyſes: Pharao wil not heare you that manie ſignes may be done
in the Land of Ægypt. \V And Moyſes and Aaron did al the wonders that
are written, before Pharao. And our Lord
\SNote{As before.
\XRef{c.~7. v.~3.}
\XRef{c.~9. v.~16.}
\XRef{c.~10. v.~1.}}
hardned Pharaoes hart, neither did he diſmiſſe the children of Iſrael
out of his Land.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
