%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 0232
%%% o-0211
\startChapter[
  title={Chapter 18}
  ]

\Summary{Iethro Moyſes father in law bringeth to him his wife and
  childrens. 8.~And hearing the great workes of God, 12.~offereth
  Sacrifice: 13.~and wiſely aduiſed Moyſes to appoint ſubordinate
  officers to iudge leſse cauſes, reſeruing the greater to him ſelfe.}

And when Iethro the
\SNote{\HH{Cohen} in Hebrew ſignifieth \Emph{Prince} or \Emph{Prieſt},
which offices in the law of nature were often ioyned in one perſon.}
prieſt of Madian, the allied of Moyſes, had heard al the thinges, that
God had done to Moyſes, and to Iſrael his people, and that our Lord had
brought forth Iſrael out of Ægypt: \V he tooke Sephora the wife of
Moyſes whom he had ſent backe: \V and her two ſonnes, of which one was
called Gerſam, his father ſaying: I haue bene a ſtranger in a forren
countrie. \V And the other Eliezer: for the God of my father, quoth he,
is my helper, and hath deliuered me from Pharaoes ſword. \V Iethro
therfore the allied of Moyſes came and his ſonnes, and his wife to
Moyſes into the deſert, where he was camped beſide the mountayne of
God. \V And he ſent word to Moyſes, ſaying:
%%% o-0212
I Iethro thy allied come to thee, and thy wife, and thy two children
with her. \V Who going forth to mete his allied, adored, and kiſſed him:
and they ſaluted
\Fix{on}{one}{likely typo, fixed in other}
an other with wordes of peace. And when he was entred into the tent, \V
Moyſes told his allied al thinges that our Lord had done to Pharao, and
the Ægyptians for Iſrael: and the whole trauaile which had chanced to
them in the iourney, and that our Lord had deliuered them. \V And Iethro
reioyced for al the good thinges, that our Lord had done to Iſrael,
becauſe he had deliuered them out of the handes of the Ægyptians, \V and
he ſaid: Bleſſed is the Lord, that hath deliuered you out of the hand of
the Ægyptians, and out of the hand of Pharao, that hath deliuered his
people out of the hand of Ægypt. \V Now doe I know, that the Lord is
great aboue al goddes: for becauſe they dealt proudely againſt them. \V
Iethro therfore the allied of Moyſes offered holocauſtes and hoſtes to
God: and Aaron and al the ancientes of Iſrael came, to eate bread with
him before God. \V And the next day Moyſes ſate to iudge the people, who
ſtoode by Moyſes from morning
%%% 0233
vntil night. \V Which thing when his allied had ſeene, to witte, al
thinges that he did in the people, he ſaid: What is this that thou doeſt
in the people? Why ſitteſt thou alone, and al the people attendeth from
morning vntil night? \V To whom Moyſes anſwered: The people cometh to me
ſeeking the ſentence of God. \V And when anie controuerſie chanceth
among them, they come vnto me to iudge betwene them, and to ſhew the
preceptes of God, and his lawes. \V But he ſaid: Thou doeſt not wel: \V
thou art tyred with fooliſh labour, both thou, and this people that is
with thee, the busines is aboue thy ſtrength, thou alone canſt not
ſuſteyne it. \V But
\SNote{Manifold wiſdome, wherof Daniel prophecieth
\XRef{(c.~12. v.~4.)}
in Chriſtian gentils, was here prefigured in Iethro a gentil.}
heare my wordes and counſeils, and God ſhal be with thee. Be thou to the
people in thoſe thinges that pertaine to God, to report their wordes vnto
him: \V and to ſhew to the people the ceremonies and rite of
worſhipping, and the way wherin they ought to walke, and the worke that
they ought to doe. \V And prouide out of al the people men that are
wiſe, and doe feare God, in whom there is truth, and that doe hate
auarice, and appointe of them tribunes, and centurions, and
quinquagenarians, and deanes, \V which may iudge the people at al tymes:
and what great matter ſoeuer ſhal fal out, let them referre it to thee
and let them iudge the leſſe matters only: and ſo it may be lighter for
thee, the burden being imparted vnto others. \V If thou doeſt this, thou
ſhalt fulfil the commandment of God, and ſhalt be able to beare his
preceptes: and al this people ſhal returne to their places with
peace. \V Which thinges when Moyſes heard
\SNote{To whom Moyſes willingly yelded.
\Cite{Origen in hunc locum.}
Morally Superiors are admoniſhed by Moyſes example to lerne of anie man,
that which is good. S.~Chryſoſtom
\Cite{ho. de ferendis reprehenſionibus.}}
he did al thinges that he had ſuggeſted vnto him. \V And chooſing
ſubſtantial men
out of al Iſrael, he appointed them princes of the people, tribunes, and
centurions, and quinquagenarians, and deanes. \V Who iudged the people
at al time: and whatſoeuer was of greater difficultie they referred to
him, themſelues iudging the eaſier caſes only. \V And he diſmiſſed his
allied: who returning went into his countrie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
