%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-22


\project douay-rheims


%%% 0242
%%% o-0221
\startChapter[
  title={Chapter 22}
  ]

\Summary{The punishment of theift, 5.~and other treſpaſses, 7.~if a
  thing committed to cuſtodie or lent doth perish, 16.~of deflovvring a
  virgin, 18.~of inchanting, beſtialitie, and idolatrie, 21.~of hurting
  ſtrangers, vvidovves, and orphanes. 25.~The lavv of lending vvithout
  vſurie, 26.~of taking pleadge, 28.~of reuerence to ſuperiors, and of
  paying tithes.}

%%% 0243
If any man ſteale an oxe or a ſheepe, and kil or ſel it: he ſhal
reſtore
\SNote{VVhere great faults are cõmitted, puniſhment is inflicted
according to the enormitie of the ſinne, aboue the proportion of the
iniurie.
\Cite{Theodoret. q.~50. in Exod.}
\XRef{Deut.~25.}
Myſtically, He that taketh from the Church a daylie laborer in Gods
field, ſinneth more greuouſly and deſerueth more puniſhment, then he
that taketh a priuate man of Chriſts flock.
\Cite{Rabanus.}}
fiue oxen for one oxe, and
%%% !!! second ref to SNote above, VVhere great faults are ...
%%% \SNote{}
foure sheepe for one ſheepe. \V If the theefe be found breaking vp the
houſe or vndermining it, and taking a wound die: the ſtriker shal not be
guiltie of bloud. \V But if he doe this when the ſunne is riſen, he hath
cõmitted manſlaughter, and himſelf shal die. If he haue not wherwith to
make reſtitution for the theft, himſelfe shal be ſold. \V If that which
he ſtole, be found with him, aliue, either oxe, or aſſe, or sheepe: he
shal reſtore duble. \V If anie man hurt a field or a vineyard, and let
goe his beaſt to feede vpon that which is other mens: the beſt of
whatſoeuer he hath in his owne field, or in his vineyard, he shal
reſtore according to the eſtimation of the damage. \V If fire breaking
forth light vpon the thornes, and catch ſtackes of corne, or corne
ſtanding in the fieldes, he shal render the damage that kindled the
fire. \V If a man committe money, or veſſel vnto his frend to keepe, and
they be ſtolen away from him, that receaued them: if the theefe be
found, he shal reſtore duble: \V if the theefe be not knowen, the
maiſter of the houſe shal be brought to
\SNote{Iudges called goddes for their eminent authoritie.
\XRef{Exo.~7. v.~1.}}
the goddes, and shal ſweare that he did not extend his hand vpon his
neighbours good, \V to doe any fraud, as wel in oxe as in aſſe, and
sheepe and rayment, and whatſoeuer may bring damage: the cauſe of both
parties shal come to the goddes: and if they geue iudgement, he shal
reſtore duble to his neighbour. \V If a man committe aſſe, oxe, sheepe,
or any beaſt, to his neighbours cuſtodie, and it die, or be hurt, or be
taken of enemies, and no man ſaw: \V there shal be an oath betwen them,
that he did not put forth his hand to his neighbours good: and the owner
shal admitte the oath, and he shal not be compelled to make
%%% o-0222
reſtitution. \V But if it were taken away by ſtelth, he shal reſtore the
damage to the owner. \V If it were eaten of a beaſt, let him bring vnto
him that which was ſlaine, and he shal not make reſtitution. \V He that
asketh of his neighbour to borow any of theſe thinges, and it be hurt or
dead the owner not being preſent, he shal be compelled to make
reſtitution. \V But if the owner be preſent, he shal not make
reſtitution, eſpecially if it were hired and came for the hire of the
ſame. \V If a man ſeduce a virgin being not yet deſpouſed, and lie with
her: he shal endowe her, and haue her to wife. \V If the virgins father
%%% 0244
wil not geue her, he ſhal geue money according to the maner of the
dowrie, which virgins are wont to receaue. \V Inchanters thou ſhalt not
ſuffer to liue. \V He that lieth with a brute beaſt, dying let him
die. \V He that ſacrificeth to goddes, ſhal be put to death, but to the
Lord only. \V Thou shalt not moleſt a ſtranger, nor afflict him:
\SNote{The law of nature requireth to do to others as we would they
ſhould do to vs. For which cauſe (beſides others) God ſuffered his
people to be ſtrangers in Ægypt, to moue them to compaſſion towards
others in like caſe.
\Cite{Rabanus.}}
for your ſelues alſo were ſtrangers in the Land of Ægypt. \V A widow and
an orphane you shal not hurt. \V If you hurt them, they wil crie out to
me, and I wil heare their crie: \V and my furie ſhal take indignation,
and I wil ſtrike you with the ſword, and your wiues shal be widowes, and
your children orphanes. \V If thou lend money to my people being poore,
that dwelleth with thee, thou shalt not vrge them as an exactour, nor
oppreſſe them with vſuries. \V If thou take of thy neighbour a garment
in pledge, thou shalt geue it him againe before ſunne ſette. \V for that
ſame is the onlie thing, wherwith he is couered, the clothing of his
bodie, neither hath he other to ſleepe in: if he
\SNote{Oppreſſion of the poore crieth to God for reuenge.}
crie to me, I wil heare him, becauſe I am merciful. \V Thou shalt not
detract from
\SNote{Prieſts called goddes for their ſacred function. See
\XRef{the commentary on Chap.~7. v.~1.}}
the goddes, and the prince of thy people thou shalt not curſe. \V Thy
tithes and thy firſt fruites thou shalt not ſlacke to pay, the
firſtborne of thy ſonnes thou shalt geue me. \V Of thy oxen alſo &
sheepe thou shalt doe in like maner: ſeuen dayes let it be with the
damme, the eight day thou shalt render it to me. \V Holie men you shal
be to me: the flesh that beaſtes haue taſted of before, you shal not
eate, but shal caſt it to the dogges.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
