%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 33
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-33


\project douay-rheims


%%% 0266
%%% o-0241
\startChapter[
  title={Chapter 33}
  ]

\Summary{Gods wrath being mitigated by Moyſes, the people mourne for
  their ſinne. 7.~Moyſes pitcheth the tabernacle vvithout the campe, and
  therin conuerſeth familiarly vvith God, 18.~deſiring to ſee his glorie.}

And our Lord ſpake to Moyſes, ſaying: Goe, get thee vp from this place,
thou and thy people which thou haſt brought out of the Land of Ægypt,
into the land wherof I ſware to Abraham, Iſaac, and Iacob, ſaying: To
thy ſeede I wil giue it: \V and I wil ſend an Angel thy precurſor, that
I may caſt out the Chananeite, and Amorrheite, and Hetheite, and
Pherezeite, and Heueite, and Iebuſeite, \V and thou maieſt enter into
the land that floweth with milke and honey: for
\SNote{God would not in this paſſage worke ſuch miracles, as he did,
bringing them forth of Ægypt. So it is a comination becauſe they were
ſtubborne and ſtiffe necked.}
I wil not goe vp with thee, becauſe thou art a ſtiffenecked people: leſt
perhappes I deſtroie thee in the way. \V And the people hearing this
verie il ſaying, mourned: and no man put on his ornamentes after the
cuſtome. \V And our Lord ſayd to Moyſes: Speake to the children of
Iſrael: Thou art a ſtiffenecked people, once I ſhal goe vp in the middes
of thee, and ſhal deſtroy thee. Now preſently lay away thy ornamentes,
that I may know what to doe vnto thee. \V Therfore the children of
Iſrael layd away their ornamentes from mount Horeb.

\V Moyſes alſo taking the tabernacle, pitched it without the campe a
farre of, and called the name thereof, The Tabernacle of couenant. And
al the people, that had anie queſtion, went forth to the Tabernacle of
couenant, without the campe. \V And when Moyſes went forth to the
tabernacle, al the people roſe vp, and euerie one ſtoode in the dore of
his pauilion, and they beheld the backe of Moyſes, til he entred into
the tabernacle. \V And when he was entred into the Tabernacle of
couenant, the piller of the cloude came downe, and ſtoode at the doore,
and he ſpake with Moyſes, \V al they beholding that the piller of the cloud
ſtoode at the doore of the Tabernacle. And they ſtoode, and adored at
the doores of their tabernacles. \V And our Lord ſpake vnto Moyſes face
to face, as a man is wont to ſpeake to his frend. And when he returned
into the campe, his miniſter Ioſue the ſõne of Nun, a yong man, departed
not from the Tabernacle. \V And Moyſes ſaid to our Lord: Thou commandeſt
me to
%%% 0267
leade forth this people: and doeſt not ſhew me whom thou wilt ſend with
me, eſpecially wheras thou haſt ſaid: I know thee by name, and thou haſt
found grace in my ſight. \V If therfore I haue found grace in thy ſight,
shew me thy face, that I may know thee, and may find grace before thine
eyes, looke vpon thy people this nation. \V And our Lord ſaid: My face
ſhal goe before thee, and I wil giue thee
%%% o-0242
reſt. \V And Moyſes ſayd: If thy ſelfe doeſt not goe before, bring vs
not out of this place. \V For whereby shal we be able to know I and thy
people, that we haue found grace in thy ſight, vnles thou walke with vs,
that we may be glorified of al peoples, that dwel vpon the earth? \V
And our Lord ſaid to Moyſes: This word alſo, which thou haſt ſpoken, wil
I doe: for thou haſt found grace before me, and thy ſelfe I haue knowen
by name. \V Who ſaid: Shew me thy glorie. \V He anſwered: I wil shew
thee
\SNote{The viſion of God in glorie, is \Emph{al good}.}
al good, and
\SNote{God by his grace maketh his ſeruants to cal vpon his name.
\Cite{S.~Aug. q.~154. in Exod.}}
wil cal in the name of the Lord before thee: and I wil haue mercie on
whom I wil, and I wil be merciful to whom it shal pleaſe me. \V And
againe he ſayd: Thou canſt not ſee my face:
\SNote{None in this life can ſee God as Saincts do in glorie.
\XRef{1.~Ioan.~3.}}
for man shal not ſee me, and liue. \V And againe: Behold, quoth he,
there is a place with me, and thou shalt ſtand vpon the rocke. \V And
when my glorie shal paſſe,
\CNote{\XRef{Rom.~9.}}
I wil ſette thee in a hole of the rocke, and protect thee with my right
hand, vntil I paſſe: \V and I wil take away my hand, and thou shalt ſee
\SNote{Moyſes ſaw more glorious workes & effectes of God, then other Prophets,
yet not his ſubſtance and diuine nature.
\Cite{Theodoret. q.~68. in Exod.}
\Cite{S.~Hier. de verb. Iſa. vidi. Dom.}
\Cite{S.~Chryſoſt. ho.~4. de in copit. Dei natura.}}
my backe-partes: but my face thou canſt not ſee.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
