%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 27
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-27


\project douay-rheims


%%% 0252
%%% o-0229
\startChapter[
  title={Chapter 27}
  ]

\Summary{An Altar muſt be made vvith things belonging therto. 9.~Alſo
  the court of the tabernacle vvith hangings and pillers. 20.~And
  prouiſion of oyle for lampes.}

Thou shalt make alſo an Altar of the wood ſetim, which shal haue fiue
cubites in length, and as manie in bredth, that is, foure ſquare, and
three cubites in height. \V And 
there shal be at the foure corners hornes of the ſame: and thou shalt
couer it with braſſe. \V And thou shalt make for the vſes therof pannes
for to take the ashes, and tongues and fleshhookes, and fire pannes: al
the veſſel thou shalt make of braſſe. \V And a grate in maner of a nette
of braſſe: at the foure corners wherof shal be foure ringes of
braſſe, \V which thou shalt put vnder the hearth of the Altar: and the
grate shal be vnto the middes of the Altar. \V Thou shalt make alſo two
barres for the Altar of the wood ſetim, which thou shalt couer with
plates of braſſe: \V and thou shalt drawe them through ringes, and they
shal be on both ſides of the Altar to carrie it. \V Not maſſie, but
emptie and
%%% 0253
hollow in the inſide ſhalt thou make it, as it was ſhewed thee in the
Mount. \V Thou ſhalt make alſo the court of the tabernacle, in the ſouth
part wherof againſt the ſouth there shal be hanginges of twiſted ſilke:
one ſide shal hold in length an hundred cubites. \V And twentie pillers
with as manie feete of braſſe, which shal haue heades with their
engrauinges of ſiluer. \V In like maner alſo on the north ſide there
shal be in length hanginges of an hundred cubites, twentie pillers, and
feete of braſſe as manie, and their heades with their engrauinges of
ſiluer. \V But in the bredth of the court, that looketh to the weſt,
there shal be hanginges of fiftie cubites, and ten pillers, and as manie
feete. \V In that bredth alſo of the court, which looketh to the eaſt,
there shal be fiftie cubites. \V In the which there shal be deputed to
one ſide hanginges of fiftene cubites, and three pillers, and as manie
feete. \V And in the other ſide there shal be hanginges conteyning
fiftene cubites, three pillers, and as manie feete. \V And in the
entring of the court there shal be made an hanging
of twentie cubites of hyacinth and purple, and ſcarlet twiſe died, and
twiſted ſilke, with embrodered worke: it shal haue foure pillers, with
as manie feete. \V Al the pillers of the court round about shal be
garnished with plates of ſiluer, ſiluer heades, and feete of braſſe. \V
In length the court
%%% o-0230
shal occupie an hundred cubites, in bredth fiftie, the height shal be of
fiue cubites: and it shal be made of twiſted ſilke, and shal haue feete
of braſſe. \V Al the veſſel of the tabernacle for al vſes and
ceremonies, the pinnes as wel of it as of the court, thou shalt make of
braſſe. \V Command the children of Iſrael that they bring thee oyle of
the oliuetrees the pureſt, and beaten with a peſtil: that a lampe may
burne
\SNote{God would not haue darknes in his tabernacle by day nor night,
ſignifying that his people ought alwayes to ſhine in good workes.
\Cite{S.~Beda. li.~3. c.~2. de tabernac.}}
alwayes \V in the tabernacle of the teſtimonie, without the veile that
is drawen before the teſtimonie. And Aaron and his ſonnes shal place it,
that it may geue light before the Lord vntil the morning. \V It shal be
a perpetual obſeruance through out their ſucceſſions before the children
of Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
