%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 0212
%%% o-0193
\startChapter[
  title={Chapter 14}
  ]

\Summary{Pharao perſecuting the children of Iſrael with a great armie,
  10.~they murmur againſt Moyſes, 13.~but are encoureged by him, and
  paſſe through the red ſea drie foote. 23.~Pharao and his hoſte
  wilfully following are drowned.}

And our Lord ſpake to Moyſes, ſaying: \V Speake to the children of
Iſrael: Let them returne and campe ouer againſt Phihahiroth which is
betwene Magdal and the ſea againſt Beelſephon: in the ſight therof you
ſhal campe vpon the ſea. \V And Pharao wil ſay concerning the children
of Iſrael: They are ſtraitened in the land, the deſert hath shute them
in. \V And I wil indurate his hart,
\SNote{Although the Hebrew Greke and Latin haue, \Emph{And he}, yet
Proteſtants corruptly thruſt in the text, \Emph{that he shal}, to make
it ſound to their ſenſe, that God did not only permit, but worke Pharaoes
induration.}
and he wil purſew you: and I wil be glorified in Pharao, and in al his
armie: and the Ægyptians shal know that I am the Lord. And they did
ſo. \V And it was told the king of Ægyptians that the people was fled:
and the hart of Pharao and of his ſeruantes was changed toward the
people, and they ſaid: What meant we to doe, that we diſmiſſed Iſrael
from ſeruing vs? \V Therfore he made readie his chariotte, and tooke al
his people with him. \V And he tooke ſix hundred choſen chariottes, and
al the chariottes that were in Ægypt: and captaines of the whole
armie. \V And our Lord hardned Pharaoes hart the king of Ægypt, and he
purſewed the children of Iſrael: but they went forth in a mightie
hand. \V And when the Ægyptians
%%% o-0194
purſewed their ſteppes going before, they found them encamped at the ſea
ſide: al Pharaoes horſe and chariottes, and the whole armie were in
Phihahiroth againſt Beelſephon. \V And when Pharao approched, the
children of Iſrael lifting vp their eies, ſaw the Ægyptians behind them:
and they feared exceedingly, and cried to our Lord, \V and ſaid to
Moyſes: Perhappes there were no graues in Ægypt, therfore thou haſt
taken vs thence to die in the wildernes: why wouldeſt thou die this, in
bringing vs out of Ægypt? \V Is not this the word that we ſpake to thee
in Ægypt, ſaying: Depart from vs, that we may ſerue the Ægyptians? for
it was much better to ſerue them, then to die in the wildernes. \V And
Moyſes ſaied to the people: Feare not: ſtand, and ſee the great wonders
of our Lord that he wil doe this day: for the Ægyptians, whom now you
ſee, you shal no
%%% 0213
more ſee for euer. \V Our Lord wil fight for you, and you ſhal hold your
peace. \V And our Lord ſaid to Moyſes:
\SNote{A ſorowful hart, lamentably mourning for the people is called
crying to God.
\Cite{S.~Hierom. in Gal.~4.}}
Why crieſt thou to me? ſpeake to the children of Iſrael that the goe
forward. \V But thou lift vp thy rodde, and ſtretch forth thy hand vpon
the ſea, & diuide it: that the children of Iſrael may goe in the middes
of the ſea by drie ground. \V
\CNote{\XRef{1.~Reg.~6.}}
And I wil indurate the hart of the Ægyptians to purſew you: and I wil be
glorified in Pharao, and in al his hoſte, and in his chariottes and in
his horſemen. \V And the Ægyptians ſhal know that I am the Lord when I
ſhal be glorified in Pharao, and in his chariottes & in his horſemen. \V
And
\SNote{Protection of Angels.}
the Angel of God, that went before the campe of Iſrael, remouing him
ſelfe, went behind them: and together with him the piller of the cloude,
leauing the foreward, \V ſtrode behind, betwen the Ægyptians campe and
the campe of Iſrael: and it was a darke cloud, and lightening the night,
ſo that they could not come to ech other the whole night time. \V And
when Moyſes had ſtretched forth his hand vpon the ſea, our Lord tooke it
away, a vehement and burning winde blowing al the night, and turned it
into drie ground: and the water was diuided. \V And the children of
Iſrael went through the middes of the drie ſea: for the water was as it
were a wal on their right hand & their left. \V And the Ægyptians
purſewing went in after them, and al Pharaoes horſes, his chariottes and
horſemen through the middes of the ſea. \V And now the morning watch was
come, and behold our Lord looking vpon the Ægyptians campe through the
piller of fire & the cloude, ſlew their armie: \V and ouerthrew the
wheeles of the chariottes, and they were borne into the depth. The
Ægyptians therfore ſaid: Let vs flee from Iſrael: for the Lord fighteth
for them againſt vs. \V And our Lord ſaid to Moyſes: Stretch forth thy
hand vpon the ſea, that the waters may returne to the Ægyptians vpon
their chariottes and horſemen. \V And when Moyſes had ſtretched forth
his hand againſt the ſea, it returned in the firſt breake of day to the
former place: and the Ægyptians fleeing away, the waters came vpon them,
and our Lord enwrapt them in the middes of the waues. \V And the waters
returned, and ouerwhelmed the chariottes and
%%% o-0195
the horſemen of al Pharaoes armie, who folowing were entred into the
ſea,
\SNote{So in Baptiſme al ſinnes are deſtroyed.
\Cite{S.~Cyp. Ep.~76. in fine.}
\Cite{S.~Aug. Tract.~12. &~13. in Ioan.}}
neither did there ſo much as one of them remaine. \V But the children of
Iſrael marched through the middes of the drie ſea,
%%% 0214
& the waters were vnto them as in ſtede of a wal on the right hand and
on the left: \V and our Lord deliuered Iſrael in that day out of the
hand of the Ægyptians. \V And they ſaw the Ægyptians dead vpon the ſea
ſhore, and the mightie hand that our Lord had exerciſed againſt them:
and the people feared our Lord, & they beleued our Lord,
\SNote{The ſame credite is geuen to God ſpeaking by Moyſes, as if he had
ſpoken immediatly by himſelf.
\Cite{S.~Hiero. in Epiſt. as Philem.}}
& Moyſes his ſeruant.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
