%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 28
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-28


\project douay-rheims


%%% 0253
%%% o-0230
\startChapter[
  title={Chapter 28}
  ]

\Summary{God commandeth Moyſes to make diuers ſortes of veſtures for
  Aaron and his ſonnes, preſcribing the matter, maner, and ornaments
  therof.}

%%% 0254
Take vnto thee alſo Aaron thy brother with his ſonnes, from among the
children of Iſrael,
\SNote{Vocation neceſſarie to ſpiritual function.
\XRef{Heb.~5.}}
that they may doe the function of prieſthoode vnto me: Aaron, Nadab, and
Abiu, Eleazar, and Ithamar. \V And thou ſhalt make an holie veſture to
Aaron thy brother for glorie and bewtie. \V And thou ſhalt ſpeake to al
the wiſe of hart, whom I haue repleniſhed with the ſpirit of wiſdome,
that they make Aarons veſtures, wherin he being ſanctified may miniſter
to me. \V And
\SNote{Theſe veſtments ſignifie that Biſhopes and Prieſts muſt haue
ſpecial vertues, diſcretion, puritie of life, ſincere intention,
contemplatiõ of God, ſupportation of the peoples infirmitie, ſolicitude
of their good, examplar life, ſound doctrin, and band of vnion.
\Cite{S.~Hiero. ad Fabiol. de veſtitu Sacerdotum. to.~3.}}
theſe shal be the veſtments that they shal make: Rationale and an Ephod,
a tunike and a ſtraite linnen garment, a mitre and a girdle. They shal
make the holie veſtments for thy brother Aaron and his ſonnes, that they
may doe the function of prieſthood vnto me. \V And they shal take gold,
and hyacinth, and purple, and ſcarlet twiſe died, and ſilke. \V
And they 
shal make the Ephod of gold and hyacinth and purple, and ſcarlet twiſe
died, and twiſted ſilke, embrodered with diuers colours. \V It shal haue
two edges ioyned in the toppe on both ſides, that they may be cloſed
together. \V The verie workemanship alſo and al the varietie of the
worke shal be of gold and hyacinth, and purple, and ſcarlet twiſe died,
and twiſted ſilke. \V And thou shalt take two Onyx ſtones, and shalt
graue in them the names of the children of Iſrael: \V ſix names in one
ſtone, and the other ſix in the other, according to the order of their
natiuitie. \V After the worke of a grauer and the grauing of a
lapidarie, thou shalt graue them with the names of the children of
Iſrael, ſette in gold and compaſſed about: \V and thou shalt put them in
both ſides of the Ephod, a memorial for the children of Iſrael. And
Aaron shal beare their names before the Lord vpon both shoulders, for a
remembrance. \V
%%% o-0231
Thou shalt make alſo hookes of gold, \V and two litle cheynes of moſt
pure gold linked one to an other, which thou shalt put into the
hookes. \V The Rationale of iudgement alſo thou shalt make with
embrodered worke of diuers colours, according to the workmanship of the
Ephod of gold, hyacinth, and purple, and ſcarlet twiſe died, and twiſted
ſilke. \V It shal be foure ſquare and duble: it shal haue the meaſure of
a palme aſwel in length as in bredth. \V And thou shalt ſet in it foure
rewes of ſtones: In the firſt rew shal be the ſtone ſardius, and
topazius, and the emeraud: \V in the ſecond the carbuncle, the ſapphire,
and the iaſper:
%%% 0255
\V in the third a ligurius, an achates, an amethyſt: \V in the fourth a
chryſolith, an onyx, and beryllus. They shal be ſet in gold by their
rewes. \V And they shal haue the names of the children of Iſrael: with
twelue names shal they be grauen, euerie ſtone with the names of euerie
one according to the twelue tribes. \V Thou shalt make in the Rationale
cheynes linked one to an other of the pureſt gold: \V and two ringes of
gold, which thou shalt put in both the toppes of the Rationale: \V and
the golden cheynes thou shalt ioyne to the ringes, that are in the edges
therof: \V and the endes of the cheynes them ſelues thou shalt couple
with two hookes on both ſides of the Ephod, which is toward the
Rationale. \V Thou shalt make alſo two ringes of gold which thou shalt
put in the toppes of the Rationale, in the brimmes, that are ouer
againſt the Ephod, & looke toward the backe partes therof. \V Moreouer
alſo other two ringes of gold, which are to be ſet on both ſides of the
Ephod beneth, that looketh toward the nether ioyning, that the
Rationale may be fitted with the Ephod, \V and may be faſtened by the
ringes therof vnto the ringes of the Ephod with a lace of hyacinth, that
the ioyning artificially wrought may continew, and the Rationale and
Ephod may 
not be ſeparated one from the other. \V And Aaron shal beare the names
of the children of Iſrael in the Rationale of iudgement vpon his breſt,
when he shal enter into the Sanctuarie, a memorial before the Lord for
euer. \V And thou shalt put in the Rationale of iudgement
\SNote{Knowledge of the cauſe, and ſincere proceding therin, are the two
keyes of right iudgement.}
Doctrine, and Veritie, which shal be on Aarons breſt, when he shal goe
in before the Lord: and he shal beare the iudgement of the children of
Iſrael on his breſt, in the ſight of the Lord alwayes. \V And thou shalt
make the tunike of the Ephod al of hyacinth, \V in the middes wherof
aboue shal be a hole for the head, and a bordeer round about it wouen,
as is wont to be made in the vtmoſt partes of garments, that it may not
eaſily be broken. \V And beneth at the feete of the ſame tunike, round
about, thou shalt make as it were pomegranates, of hyacinth, and purple,
and ſcarlet twiſe died, litle belles interpoſed betwen, \V ſo that there
be a bel of gold and a pomegranate: and againe an other bel of gold and
a pomegranate. \V And Aaron shal be veſted with it in the office of his
miniſterie, that the ſound may be heard, when he goeth in and cometh out
of the Sanctuarie, in the ſight
%%% 0256
of the Lord, and that he die not. \V Thou ſhalt make alſo a plate of the
pureſt gold: wherin thou shalt graue after the
%%% o-0232
worke of a grauer, Holie to the Lord. \V And thou ſhalt tie it with a
lace of hyacinth, and it ſhal be vpon the mitre, \V hanging ouer the
forehead of the high Prieſt. And Aaron ſhal carie the iniquities of
thoſe thinges, which the children of Iſrael haue offered and
ſanctified, in al their giftes and donaries. And the plate shal be
alwayes in his forehead, that the Lord may be wel pleaſed with them. \V
And thou shalt gird the tunike with ſilke, and thou ſhalt make a ſilken
mitre, and a bawdrike of embrodered worke. \V Moreouer for the ſonnes of
Aaron thou shalt prepare linnen tunikes, and bawdrikes and mitres for
glorie and bewtie: \V And with al theſe thinges thou shalt veſt Aaron
thy brother, and his ſonnes with him. And thou shalt conſecrate the
handes of them al, and shalt ſanctifie them, that they may doe the
function of prieſthood vnto me. \V Thou shalt make alſo linnen breches,
to couer the flesh of their turpitude from the reynes vnto the
thighes: \V and Aaron and his ſonnes ſhal vſe them when they shal enter
into the tabernacle of teſtimonie, or when they approch to the Altar to
miniſter in the Sanctuarie, leſt guiltie of iniquitie they die. It shal
be a law for euer to Aaron, and to his ſeede after him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
