%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 35
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-35


\project douay-rheims


%%% 0270
%%% o-0244
\startChapter[
  title={Chapter 35}
  ]

\Summary{The precept of the Sabbath is yet renewed. 4.~Firſt fruites,
  and other giftes are required, and duly offered, for the making of the
  tabernacle and other thinges therto belonging, which are here
  recited. 30.~Beſeleel and Ooliab are appointed workmen for this
  purpoſe.}

Therfore
\SNote{After the fall of the people to Idolatrie, their punishment, and
repentãce, their reconciliation to God and new tables of the
commandments made and written, Moyſes repeteth the former precept, of
keping the Sabbath, and prouideth al neceſſaries to the making of the
Tabernacle, wherto the Princes and people moſt promptly and liberally
contribute.
\Cite{S.~Aug. q.~172. in Exo.}}
al the multitude of the children of Iſrael being gathered together, he
ſaid to them: Theſe are the thinges which our Lord hath commanded to be
done. \V Six dayes you ſhal do worke: the ſeuenth day shal be holie vnto
you, the ſabbath, and reſt of our Lord: he that shal do anie worke in
it, shal be ſlaine. \V You shal not kindle fire in al your habitations
on the ſabbath day. \V And Moyſes ſaid to al the aſſemblie of the
children of Iſrael: This is the word that our Lord hath commanded,
ſaying: \V Separate with you firſt fruites to the Lord. Let euerie one
that is willing, and hath a readie hart, offer them to the Lord: gold and
ſiluer, and braſſe, \V hyacinth and purple, and ſcarlet twiſe died, and
ſilke, the haire of goates, \V and rammes skinnes died redde, and
ianthin skinnes, the wood ſetim, \V and oile to maintaine lightes, and
to make ointment, and moſt ſweete incenſe, \V Onys ſtones, and pretious
ſtones, for the adorning of the Ephod and the Rationale. \V Whoſoeuer of
you is wiſe, let him come, and make that which our Lord hath
commanded: \V to wit, the Tabernacle, and the roofe therof, and the
couer, the ringes, and the bordeworke with the barres, the pinnes and the
feete: \V the
%%% o-0245
Arke and the ſtaues, the propitiatorie, and the veile, that is drawen
before it: \V the Table with the barres and the veſſel, and the loaues
of propoſition: \V the Candleſticke to beare vp the lightes, the veſſel
therof and the lampes, and the oile to the nouriſhing of fires: \V the
Altar of incenſe, and the barres, and the oile of vnction and the
incenſe of ſpices: the Hanging at the doore of the tabernacle: \V the
Altar of holocauſte, and his grate of braſſe, with the barres and veſſel
therof: the Lauer and his feete: \V the Curtines of the court with the
pillers and the feete, the hanging in the doores of the entrie, \V the
pinnes of the tabernacle and of the court with their litle cordes: \V
the Veſtimentes, that are to be vſed in the miniſterie of the
ſanctuarie, the veſture of Aaron the high
%%% 0271
Prieſt, and of his ſonnes, to do the function of Prieſthood vnto me. \V
And al the multitude of the children of Iſrael going forth from the
ſight of Moyſes, \V offered firſt fruites to our Lord with a moſt prompt
and deuoute minde, to make the worke of the tabernacle of the
teſtimonie. Whatſoeuer was neceſſarie to the ſeruice therof and to the
holie veſtiments, \V both men and wemen did giue, tablettes and
earlettes, ringes and bracelettes: euerie veſſel of gold was ſeparated
for the donaries of our Lord. \V If anie man had hyacinth, and purple,
and ſcarlet twiſe died redde, and ianthin skinnes, \V metal of ſiluer
and braſſe, they offered to our Lord, and the wood ſetim for diuers
vſes. \V But the skilful wemen alſo gaue ſuch thinges as they had
ſponne, hyacinth, purple, and ſcarlet, and ſilke, \V and goates haire,
geuing al of their owne accord. \V But the princes offered onyx ſtones,
and pretious ſtones, for the Ephod and the Rationale, \V and ſpices and
oile to maintaine the lightes, and for the preparing of ointment, and to
make the incenſe of moſt ſweete ſauour. \V Al men and wemen with deuoute
mind offered donaries, that the workes might be made which our Lord had
commanded by the hand of Moyſes. Al the children of Iſrael did dedicate
voluntarie thinges to our Lord.

\V And Moyſes ſaid to the children of Iſrael: Behold, our Lord hath
\SNote{As matter alone is not ſufficient for a building without
artificers, to whom God geueth ſpecial skil: ſo for expounding holie Scripture
God geueth particular knowledge to \Emph{Paſtors and Doctors, to the
conſummation of Sainctes, to the vvorke of the miniſterie, to the
edifying of the bodie of Chriſt.} (the Church.)
\XRef{Epheſ.~4.}}
called by name Beſeleel the ſonne of Vri the ſonne of Hur of the tribe
of Iuda. \V And hath filled him with the ſpirit of God, with, wiſdome
and intelligence, and ſcience and al learning \V to deuiſe and to make
worke in gold and ſiluer, and braſſe, \V and in grauing ſtones, and in
carpenters worke. Whatſoeuer can be deuiſed artificially, \V he hath
giuen in his hart: Ooliab alſo the ſonne of Achiſamech of the tribe of
Dan: \V both hath he inſtructed with wiſedome, to make the workes of a
carpenter, a tapeſter, an embroderer of hyacinth and purple, and ſcarlet
twiſe died, and ſilke, and to weaue al thinges, and to inuent al new
things.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
