%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 31
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-31


\project douay-rheims


%%% 0261
%%% o-0237
\startChapter[
  title={Chapter 31}
  ]

\Summary{Beſeleel and Ooliab are deputed by our Lord to make the
  Tabernacle, and the things belonging therto. 12.~The obſeruation of
  the ſabbath day is againe commanded. 18.~And our Lord deliuereth to
  Moyſes tvvo tables vvritten with the finger of God.}

And our Lord ſpake to Moyſes, ſaying: \V Behold, I haue called by name
Beſeleel the ſonne of Vri the
%%% 0262
ſonne of Hur of the tribe of Iuda, \V and I haue replenished him with
the ſpirit of God, with wiſdome, & vnderſtanding, and knowledge in al
worke, \V to deuiſe whatſoeuer may be artificially made of gold, and
ſiluer, and braſſe, \V of marble, and precious ſtones, and diuerſitie of
wood. \V And I haue geuen him for his felow Ooliab the ſonne of
Achiſamech of the tribe of Dan. And in the hart of euerie skilful man
haue I put wiſdome: that they may make al things which I haue commanded
thee, \V the tabernacle of couenant, and the arke of teſtimonie, and the
propitiatorie, that is ouer it, and al the veſſel of the tabernacle, \V
and the table and the veſſel therof, the candleſticke moſt pure with the
veſſel therof, and the Altares of incenſe, \V and of holocauſte, and al
their veſſel, the lauer with his foote, \V the holie veſtments in
the miniſterie for Aaron the prieſt, and for his ſonnes, that they may
execute their office, about the ſacred things: \V the oile of vnction,
and the incenſe of ſpices in the Sanctuarie, al thinges which I haue
commanded thee, shal they make. \V And our Lord ſpake to Moyſes,
ſaying: \V Speake to the children of Iſrael, and thou shalt ſay to them:
See that you keepe my ſabbath: becauſe it is a ſigne betwen me and you
in your generations: that you may know that I am the Lord, which
ſanctifie you. \V Keepe you my ſabbath: for it is holie vnto you: he
that shal pollute it, dying shal die: he that shal doe worke in it, his
ſoule shal perish out of the middes of his people. \V Six dayes shal you
doe worke: in the ſeuenth day is the ſabbath, the holie reſt to the
Lord. Euerie one that shal doe any worke in this day, shal die. \V Let
the children of Iſrael keepe the Sabbath, and celebrate it in their
generations. It is an euerlaſting couenant \V betwen me and the children
of Iſrael, and a ſigne perpetual: for in ſix dayes the Lord made heauen
and earth, and in the ſeuenth he ceaſed from worke. \V And our Lord,
when he had ended ſuch ſpeaches in mount Sinai, gaue vnto Moyſes two
ſtone tables of teſtimonie, written
\SNote{Not by Moyſes, but by an Angel, at Gods appointment.
\XRef{Gal.~3. v.~19.}}
with the finger of God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
