%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 0199
%%% o-0182
\startChapter[
  title={Chapter 9}
  ]

\Summary{The fifth plague is peſtilence among the Ægyptians
  cattle. 8.~The ſixt boyles in men and beaſtes. 18.~The ſeuenth,
  haile. 27.~Pharao confeſsing God to be iuſt, and him ſelfe and his
  people impious, promiſeth againe to diſmiſſe the people, 34.~but
  faileth to do it.}

And our Lord ſaid to Moyſes: Goe in to Pharao, and ſpeake to him: This
ſaith our Lord, the God of the Hebrewes: diſmiſſe my people to ſacrifice
vnto me. \V And if thou refuſe, and holdeſt them: \V behold my hand shal
be vpon thy fieldes: and vpon thy horſes, and aſſes, and camels, and
oxen, and sheepe,
\MNote{The 5.~plague. Peſtilence amongſt cattel.}
a verie ſore peſtilence. \V And our Lord wil make a merueile betwene the
poſſeſſions of Iſrael & the poſſeſſions of the Ægyptians, that nothing
at al perish of thoſe thinges that pertaine to the children of
Iſrael. \V And our Lord hath appointed a time, ſaying: To morow wil our
Lord doe this thing in the land. \V Our Lord therfore did this thing the
next day: and
\SNote{Not al the beaſtes died, for ſome died in the 7.~& 10.~plagues
but al that died pertained to the Ægyptians.}
al the beaſtes of the Ægyptians dyed, but of the beaſtes of the children
of Iſrael nothing at al perished. \V And Pharao ſent to ſee: neither was
there any thing dead of that which Iſrael poſſeſſed. And Pharaoes hart
\SNote{In Hebr. \HH{Vaijchbad leb Parhaoh.} \Emph{Pharaoes hart hardned
it ſelfe.}}
was hardned, and he did not diſmiſſe the people. \V And our Lord ſaid to
Moyſes, & Aaron: Take your handes ful of ashes out of the chimney, and
let Moyſes ſprinkle it into the ayre before Pharao. \V And be there
%%% 0200
duſt vpon al the Land of Ægypt: for there shal be in men, & beaſtes
\MNote{The 6.~plague. Boyles in men and beaſtes.}
boyles, and ſwelling bladders in the whole land of Ægypt. \V And they
tooke ashes out of the chimney, and ſtoode before Pharao, and Moyſes
ſprinkled it into the ayre: and there were made boyles of ſwelling
bladders in men and beaſtes. \V
\SNote{Poore Enchanters, that could neither eſcape, nor cure this
plague.}
Neither could the enchanters ſtand before Moyſes for boyles that were
vpon them, and in al the Land of Ægypt. \V And our Lord did indurate
Pharaoes hart, & he heard them not, as our Lord ſpake to Moyſes. \V And
our Lord ſaid to Moyſes: In the morning ariſe, and ſtand before Pharao,
& thou shalt ſay to him: This ſaith the Lord, the God of the Hebrewes:
Diſmiſſe my people to ſacrifice vnto me. \V Becauſe this time I wil ſend
al my plagues vpon thy hart, and vpon thy ſeruantes, and vpon thy
people: that thou mayeſt know there is not the like to me in al the
earth. \V For now ſtretching forth my hand I wil ſtrike thee, and thy
people with peſtilence, and thou shalt perish from the earth. \V And
\CNote{\XRef{Rom.~9.}}
therfore
\SNote{In Hebrew \Emph{I haue made thee ſtand}, in the 70 and chalde
paraphraſis \Emph{I haue kept thee aliue}. In the
Latin \Emph{I haue put or ſet thee}, that in thee, through thyn
owne malice indurate, I may make knowen my powre to mãkind.
\Cite{S.~Aug. de Prædeſt. et Grat. c.~6.}}
haue I ſet thee, that in thee I may shew my might, and my name may be
told in al the earth. \V Doeſt thou yet hold backe my people: and wilt
thou not diſmiſſe them? \V Behold I wil raine to morow this very houre
\MNote{The 7.~plague. Terrible haile, thunders and lightnings.}
haile exceeding
%%% 0-0183
much: ſuch as was not in Ægypt from the day that it was
founded, vntil this preſent time. \V Send therfore now preſently, and
gather together thy cattle, and al thinges that thou haſt in the field:
for men & beaſtes, and al thinges that ſhal be found abroad, and not
gathered together out of the fieldes, and the haile fal vpon them, ſhal
die. \V He that feared the word of our Lord of Pharaoes ſeruantes, made
his ſeruantes to flie, and his beaſtes into houſes: \V but he that
neglected the word of our Lord, let alone his ſeruantes, and his beaſtes
in the fieldes. \V And our Lord ſaid to Moyſes: Stretch forth thy hand
towards heauen, that there may be haile in the whole Land of Ægypt vpon
men, and vpon beaſtes, & vpon euerie herbe of the field in the Land of
Ægypt. \V And Moyſes ſtretched forth his rodde toward heauen, and our
Lord gaue thunders, and haile and running lightenings on the land: and
our Lord rained haile vpon the Land of Ægypt. \V And the haile and fire
mixt together did driue: and it was of ſo great bignes, as neuer before
appeared in the whole Land of Ægypt ſince that nation was made. \V And
the haile ſmote in al the Land of Ægypt al
%%% 0201
thinges that were in the fieldes, from man euen vnto beaſt: and euerie
herbe of the field did the haile ſtrike, and euerie tree of the countrie
it did breake. \V Only in the Land of Geſſen, where the children of
Iſrael were, the haile fel not. \V And Pharao ſent, and called Moyſes
and Aaron, ſaying to them: I haue ſinned now alſo, the Lord is iuſt: I
and my people, impious. \V Pray ye the Lord that the thunders may ceaſe,
and the haile: that I may diſmiſſe you, and ye tarie not here any
longer. \V Moyſes ſaid: When I ſhal be gone forth out of the citie, I wil
ſtretch forth my handes to our Lord, and the thunders ſhal ceaſe, and
the haile ſhal not be: that thou maiſt know that the earth is our
Lords: \V but I know that neither thou, nor thy ſeruantes do yet feare
the Lord God. \V The flaxe therfore, and the barley were hurt, becauſe
the barley came vp grene, and the flaxe now was boulled: \V but the
wheate, and other winter corne were not hurt, becauſe they were
lateward. \V And Moyſes going forth from Pharao out of the citie,
ſtretched forth his handes to our Lord: and the thunders & haile ceaſed,
neither did there droppe raine any more vpon the earth. \V And Pharao
ſeing that the raine, and the haile and thunders were ceaſed, he
increaſed his ſinne: \V
\SNote{In Hebrew, \HH{Vaiach bedlibbe hu vahabadaf}. \Emph{And he hardned
his ovvne hart, he and his ſeruants.}}
and his hart was aggrauated, and the hart of his ſeruantes, and indurate
exceedingly: neither did he diſmiſſe the children of Iſrael, as our Lord
had commanded by the hand of Moyſes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
