%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 0230
%%% o-0210
\startChapter[
  title={Chapter 17}
  ]

\Summary{The people murmuring againe in Raphidim for want of drinck, our
  Lord giueth them water out of a rock. 8.~Amalech fighteth vvith
  them. And Moyſes lifting vp his hand in prayer, Iſrael ouercometh,
  otherwiſe Amalech preuaileth.}

Therfore al the multitude of the children of Iſrael ſetting forward from
the deſert Sin, by their manſions,
%%% 0231
according to the word of our Lord, camped in Raphidim, where there was
no water for the people to drinke. \V Who chiding againſt Moyſes, ſaid:
Geue vs water, that we may drinke. To whom Moyſes anſwered: Why chide
you againſt me? Wherfore doe you tempt our Lord? \V The people therfore
was thirſtie there for lacke of water, and murmured againſt Moyſes,
ſaying: Why didſt thou make vs goe forth out of Ægypt, to kil vs, and
our children, and our beaſtes with thirſt. \V And Moyſes cried to our
Lord, ſaying: What ſhal I doe to this people? Yet a litle while, and
they wil ſtone me. \V And our Lord ſaid to Moyſes: Goe before the
people, and take with thee of the ancients of Iſrael: and the rodde
wherwith thou didſt
%%% o-0211
ſtrike the riuer, take in thy hand, and goe. \V
Behold I wil ſtand there before thee, vpon the rocke Horeb: and thou
ſhalt ſtrike the rocke, and water ſhal goe out therof, that the people
may drinke. Moyſes did ſo before the ancientes of Iſrael: \V and he
called the name of that place, Temptation, becauſe of the chiding of the
children of Iſrael, and for that they tempted our Lord, ſaying: Is the
Lord amongſt vs or not? \V And Amalec came, and fought againſt Iſrael in
Raphidim. \V And Moyſes ſayd to Ioſue: Chooſe out men: and goe forth and
fight againſt Amalec: to morow I wil ſtand in the toppe of the hil,
hauing the rodde of God in my hand. \V Ioſue did as Moyſes had ſpoken,
and he fought againſt Amalec: but Moyſes and Aaron and Hur went vp vpon
the toppe of the hil. \V And when Moyſes lifted vp his hands, Iſrael
ouercame: but
\SNote{If this ceremonie \Emph{of holding vp his handes} was of ſuch
importance in the law of nature, why do Heretikes deride the ſame, and
the like in the Catholique Church? VVheras alſo our
Sauiour \Emph{lifting vp his hands} bleſſed his diſciples.
\XRef{Luc.~24.}
S.~Damaſcen alſo teacheth
\Cite{(li.~4. c.~12. Ortho.)}
that this extenſion of his handes prefigured the Croſſe of Chriſt. And
now it repreſenteth the ſame.}
if he did lette them downe a little, Amalec ouercame. \V And the handes
of Moyſes were heauie: therfore they tooke a ſtone, and putte vnder him,
wherupon he ſate: and Aaron and Hur ſtaied vp his handes on both
ſides. And it came to paſſe that his handes were not wearie vntil ſunne
ſette. \V And Ioſue put Amalec to flight, & his people by the edge of
the ſword. \V And our Lord ſaid to Moyſes: Write this for a monument in
a booke, & deliuer it to the eares of Ioſue: for I wil deſtroy the
memorie of Amalec from vnder heauen. \V And Moyſes builded an Altar:
and called the name therof, Our Lord my exaltation, ſaying: \V Becauſe
the hand of our Lords throne, and the warre of our Lord ſhal be againſt
Amalec, from generation vnto generation.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
