%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 25
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-25


\project douay-rheims


%%% 0248
%%% o-0225
\startChapter[
  title={Chapter 25}
  ]

\Summary{Oblations of firſt fruictes, and freegiftes for making the
  Tabernacle, and things perteyning therto. 10.~The Arck. 17.~The
  Propitiatorie, and Cherubims. 23.~A table, and theron the Loaues of
  propoſition. 31.~A candleſtick, 37.~and ſeuen lampes, with ſnuffers of
  gold.}

And our Lord ſpake to Moyſes, ſaying: \V Speake to the children of
Iſrael, that they take firſt fruites for me of euerie man that
offereth of his owne accord, you shal take them. \V And theſe are the
thinges which you muſt take:
\SNote{As the Iſraelites were prompt to offer theſe external things in
the old law, ſo Chriſtians muſt offer the like for Gods ſeruice: but
ſpecially al ſortes of vertues, Faith, hope, charitie, penance,
deuotion, prayer, almes, faſting, &c.}
Gold, and ſiluer, and braſſe, \V hyacinth and purple, and ſcarlet twiſe
died, and ſilke, and the haire of goates, \V and
%%% o-0226
rammes skinnes died redde, and ianthin skinnes, and the wood ſetim: \V
oyle to make lightes: ſpices for oyntement, and for incenſe of good
ſauour: \V Onyx ſtones, and pretious ſtones to adorne the ephod, and
rationale. \V And they shal make me a ſanctuarie, and I wil dwel in the
middes of them: \V according to al the ſimilitude of the tabernacle
which I wil ſhew thee, & of al the veſſel to the ſeruice therof: & thus
you ſhal make it: \V frame an arke of the wood ſetim, the length wherof
ſhal haue two cubites & an halfe: the bredth, a cubite and an halfe: the
height, likewiſe a cubite and an halfe. \V And thou shalt plate it with
moſt pure gold within and without: and ouer it thou shalt make a golden
crowne round about: \V and foure golden ringes, which thou shalt put at
the foure corners of the arke: let two ringes be on the one ſide, and
two on the other. \V Thou shalt make barres alſo of the wood
%%% 0249
ſetim, and shalt couer them with gold. \V And thou shalt put them in
through the ringes that are in the ſides of the arke, that it may be
caried on them: \V the which shal be alwayes in the ringes, neither shal
they at anie time be drawen out of them. \V And thou shalt put in the
arke the teſtification which I wil geue thee. \V Thou shalt make a
Propitiatorie of moſt pure gold: the length therof shal hold two cubites
and an halfe, and the bredth a cubite & an halfe. \V Two
\SNote{If Images were vnlawful God would not haue commanded to make
Cherubims.
\Cite{cõ. Nic.~2.}}
Cherubims alſo thou shalt make of beaten gold, on both ſides of the
oracle. \V Let one Cherub be on the one ſide, and the other on the
other. \V Let them couer both ſides of the Propitiatorie ſpreading
their winges, and couering the oracle, and let them looke one towards
the other, their faces turned vnto the Propitiatorie wherwith the arke
is to be couered, \V wherin thou shalt put the teſtimony that I wil geue
thee. \V Thence wil I command, and wil ſpeake to thee ouer the
propitiatorie & from the middes of the two Cherubims which shal be vpon
the arke of teſtimonie, al thinges which I wil command the children of
Iſrael by thee. \V Thou shalt make a table alſo of the wood ſetim,
hauing two cubites in length, and in bredth a cubite, and in height a
cubite and an halfe. \V And thou shalt plate it with moſt pure gold: &
thou shalt make to it a golden ledge round about, \V and to the ledge
it ſelf a crowne enterpolished, four fingers high: and vpon the ſame, an
other golden crowne. \V Thou ſhalt prepare alſo foure golden ringes, and
shalt put them in the foure corners of the ſame table at euerie
foote. \V Vnder the crowne shal the golden ringes be, that the barres
may be put through them, and the table may be carried. \V The barres
alſo them ſelues thou shalt make of the wood ſetim, and shalt compaſſe
them with gold to beare vp the table. \V Thou shalt prepare alſo
ſawcers, and phials, cenſers, and goblettes, wherein the libamentes are
to be offered, of moſt pure gold. \V And thou shalt ſette vpon the table
\SNote{For the perpetual vſe, and ſanctitie of theſe loaues, which none
might eate but ſuch as were pure
\XRef{(1.~Reg.~21.)}
they prefigured the holie Euchariſt.
\Cite{S.~Hier. in Tit.~1.}
\Cite{S.~Damaſcen. de ortho. li.~4. c.~14.}
\Cite{S.~Cyril cathec.~4.}

And conſequently Chriſt is really preſent in the B.~Sacrament. For if
there were bread in ſubſtance, it ſhould not excel the figure; which is
required in eierie thing prefigured.
\XRef{Colloſſ.~2.}}
loaues of propoſition in my ſight alwaies. \V Thou shalt make alſo a
candleſticke beaten of moſt pure gold, the shaft therof, and branches,
cuppes, and boules, and lilies proceeding from the ſame. \V Six branches
shal goe forth of the ſides, three out of one ſide, and three out of
%%% o-0227
the other. \V Three cuppes as it were in the maner of a nutte on euerie
branch, and a boule withal, and a lilie: and three cuppes likewiſe of
the fashion of a nutte in an other branch, and
%%% 0250
a boule withal, and a lilie. This shal be the worke of the ſix branches,
that are to be drawen forth from the shaft: \V and in the candleſticke
it ſelfe shal be foure cuppes in maner of a nutte, and at euerie one
boules and lilies. \V Boules vnder two branches in three places, which
together make ſix
\Fix{comnig}{comming}{obvious typo, fixed in other}
forth out of one shaft. \V Both the boules therfore and the branches
shal be out of it, al the whole beaten of moſt pure gold. \V Thou
\Fix{thalt}{shalt}{obvious typo, fixed in other}
make alſo ſeuen lampes, and shalt ſette them vpon the candleſticke, for
to geue light ouer againſt. \V The ſnuffers alſo and where the
ſnuffinges shal be put out, let them be made of moſt pure gold. \V The
whole weight of the candleſticke with al the furniture therof shal haue
a talent of moſt pure gold. \V Looke, and make it according to the
paterne, that was shewed thee in the mount.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
