%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 0210
%%% o-0192
\startChapter[
  title={Chapter 13}
  ]

\Summary{God commandeth to remember their deliuerie from Ægypt, by the
  ſolemnitie of Paſch, 2.~and by conſecrating to him the
  firſt-borne. 17.~And ſo leadeth them through the deſert towards the
  red ſea (Moyſes taking with him Ioſephs bones) by a piller of fire in
  the night, and a cloud in the day.}

And our Lord ſpake to Moyſes, ſaing: \V Sanctifie vnto me euerie firſt
borne that openeth the matrice in the children of Iſrael, as wel of men
as of beaſtes: for they are al mine. \V And Moyſes ſaid to the people:
Remember this day in the which you went forth out of Ægypt, and out of
the houſe of ſeruitude, becauſe with a ſtrong hand hath our Lord brought
you forth out of this place: that you eate not leauened bread. \V This
day you goe forth in the moneth of new corne. \V And when our Lord ſhal
haue brought thee into the Land of Chananeite and Hetheite and
Amorrheite and Heueite and Iebuſeite, which he ſware to thy fathers that
he would geue thee, a land that
\SNote{The old Teſtament propoſed cõmonly temporal rewardes.
\Cite{S.~Hierom. Ep. ad Dardanum.}}
floweth with milke and honie, thou ſhalt celebrate this maner of ſacred
rites in this moneth. \V Seuen daies ſhalt thou eate azimes: and in the
ſeuenth day shal be the ſolemnitie of our Lord. \V Azimes shal you eate
ſeuen dayes: there shal not be ſeene anie leauened thing with thee, nor
in al thy coaſtes. \V And thou ſhalt tel thy ſonne in that day, ſaying:
This is that which our Lord did to
%%% 0211
me when I came forth out of Ægypt. \V And it ſhal be as a ſigne in thy
hand, and as a moniment before thine eyes: and that the law of our Lord
be alwayes in the mouth, for in a ſtrong hand our Lord hath brought thee
out of Ægypt. \V Thou ſhalt keepe this obſeruation at the ſette time
from dayes to dayes. \V And when our Lord ſhal haue brought thee into
the Land of the Chananeite, as he ſware to thee and thy fathers, and
ſhal geue it thee: \V thou ſhalt ſeparate al that openeth the matrice
vnto our Lord, and al that is brought forth in thy cattel: whatſoeuer
thou ſhalt haue of male ſexe thou ſhalt conſecrate to our Lord. \V The
firſt borne of an aſſe thou shal change for a sheepe: and if thou doe
not redeeme it, thou shalt kil it. And euerie firſt borne of men among
thy children, thou shalt 
\Fix{redemne}{redeeme}{obvious type, fixed in other}
with a price. \V And when thy ſonne shal
aske thee to morow, ſaying: What is this? thou shalt anſwere him: With a
ſtrong hand did our Lord bring vs forth out of the land of Ægypt, out of
the houſe of ſeruitude. \V For
\SNote{In the Hebrew, \Emph{vvhen Pharao had indurated him ſelfe.}}
when Pharaoes hart was indurate, and would not diſmiſſe vs, our Lord
ſlew euerie firſt-borne in the Land of Ægypt, from the firſt-borne of
man to the firſt-borne of beaſtes: therfore I ſacrifice to our Lord al
that
%%% o-0193
openeth the matrice of the male
\Fix{ſixe,}{ſexe,}{obvious typo, fixed in other}
and al the firſt-borne of my ſonnes I doe redeme. \V It shal be therfore
as a ſigne in thy hand, and as a thing hanged before thine eyes, for a
remembrance: becauſe our Lord by a ſtrong hand hath brought vs forth out
of Ægypt. \V Therfore when Pharao had ſent forth the people, our Lord
ledde them not by the way of the Philiſthijns countrie which is neere:
thinking
\SNote{Gods preuention to auoid tentations ſheweth freewil in man.}
leſt perhappes it would repent them, if they should ſee warres ariſe
againſt them, and would returne into Ægypt. \V But he ledde them about
by the way of the deſert, which is beſides the Red ſea: and the children
of Iſrael went vp out of the Land of Ægypt armed. \V Moyſes alſo
\SNote{By this appeareth how much Moyſes eſtemed Ioſephs charge
concerning tranſlation of his bones. Alſo S.~Paul commendeth it.
\XRef{Heb.~11.}}
tooke Ioſephes bones with him: becauſe he had adiured the children of
Iſrael, ſaying: God shal viſite you, carrie out my bones from hence with
you. \V And marching from Socoth they camped in Etham in the vtmoſt
coaſtes of the wilderneſſe. \V And our Lord went before them to shew the
way by day in a piller of a cloude, and by night in a piller of fire:
that he might be the guide of their iourney both times. \V There neuer
failed the piller of the cloude by day, nor the piller of fire by night,
before the people.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
