%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 39
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-39


\project douay-rheims


%%% 0277
%%% o-0250
\startChapter[
  title={Chapter 39}
  ]

\Summary{Al the ornamentes of Aaron and his ſonnes are made, 31.~and the
  whole vvorke of the Tabernacle is perfited.}

Moreouer of hyacinth and purple, ſcarlet and ſilke he made the veſtures,
that Aaron should weare when he miniſtred in the holie places, as our
\Fix{lord}{Lord}{obvious typo, fixed in other}
commanded Moyſes. \V He made therfore an Ephod of gold, hyacinth, and
purple, and ſcarlet twiſe died, and twiſted ſilke, \V with embrodered
worke, and he did cut thinne plates of gold, and drew them ſmal into
threedes, that they might be twiſted with the woufe of the former
colours, \V and two edges coupled one to the other in the toppe on
either ſide, \V and
%%% o-0251
a bawdrike of the ſame colours, as our Lord had commanded Moyſes. \V He
prepared alſo two Onyx ſtones, faſt ſette and cloſed in gold, and grauen
by the art of a lapidarie, with the names of the children of Iſrael: \V
and he ſet them in the ſides of the Ephod for a moniment of the children
of Iſrael, as our Lord had commanded Moyſes. \V He made alſo a Rationale
with embrodered worke according to the worke of the Ephod, of gold,
hyacinth, purple, and ſcarlet twiſe died, and twiſted ſilke: \V foure
ſquare, duble, of the meaſure of a palme. \V And he ſet foure rewes of
precious ſtones. In the firſt rewe was ſardius, topazius, an emeraud. \V
In the ſecond, a carbuncle, a ſaphire, and a iaſper. \V In the third, a
ligurius, an achates, and an amethiſt. \V In the fourth a chryſolith, an
onyx, and beryllus, compaſſed and encloſed in gold by their
%%% 0278
rewes. \V And the twelue ſtones them ſelues, were grauen with the names
of the twelue tribes of Iſrael, euerie one with his ſeueral name. \V
They made alſo in the rationale litle cheynes linked one to an other of
the pureſt gold, \V and two hookes, and as manie ringes of
gold. Moreouer the ringes they ſet on either ſide of the Rationale, \V
on the which the two golden cheynes should hang, which they put into the
hookes, that ſtoode out in the corners of the Ephod. \V Theſe both
before and behind did ſo agree with them ſelues, that the Ephod and the
Rationale might be knit one to the other, \V tyed to the bawdrike and
with ringes ſtrongly coupled, which a lace of hyacinth ioyned, leſt they
should flagge looſely, and be moued one from the other, as our Lord
commanded Moyſes. \V They made alſo the tunike of the Ephod al of
hyacinth, \V and a hole for the head in the vpper part againſt the
middes, and the border of the hole round about wouen: \V and beneth at
the feete pomegranates of hyacinth, purple, ſcarlet, and twiſted
ſilke: \V and litle belles of the pureſt gold, which they did put betwen
the pomegranates in the vtmoſt part of the tunike round about: \V to
witte, a bel of gold, and a pomegranate, wherwith the high prieſt went
adorned, when he executed his miniſterie, as our Lord had commanded
Moyſes. \V They made alſo ſilken tunikes with wouen worke for Aaron and
his ſonnes: \V and mitres with their litle crownes of ſilke: \V linnen
breeches alſo, of fine line: \V and a girdle of twiſted ſilke, hyacinth,
purple, & ſcarlet twiſe died, with the art of embrodering, as our Lord
had commanded Moyſes. \V They made alſo the plate of
\SNote{Alexander the great ſeeing Iaddus the hiegh Prieſt, bearing this
venerable plate on his forehead, with great reuerence went vnto him, and
adored the name of God written in the plate.
\Cite{Ioſephus. li.~11. c.~8. Antiq.}}
ſacred veneration of moſt pure gold, and they wrote in it with the worke
of a lapidarie: The Holie of our Lord: \V and they tyed it to the mitre
with a lace of hyacinth, as our Lord had commanded Moyſes. \V Therfore
al the worke of the tabernacle & of the roofe of teſtimonie was
perfited: and the children of Iſrael did al thinges which our Lord had
commanded Moyſes. \V And they offered the tabernacle and the roofe and
the whole furniture, ringes, bordes, barres, pillers and their feete, \V
the couer of rammes skinnes died redde, and the other couer of ianthin
skinnes, \V the veile, the arke, the barres, the propitiatorie, \V the
table with the veſſel therof and the loaues of propoſition: \V the
%%% o-0252
candleſticke, the lampes, and the furniture of them with the oyle:
%%% 0279
\V the altar of gold, and the ointment, and the incenſe of ſpices: \V
and the hanging in the entrie of the tabernacle: \V the altar of braſſe,
the grate, the barres, and al the veſſel therof: the lauer with the
foote therof: the hanginges of the court, and the pillers with their
feete: \V the hanging in the entrie of the court, and the litle cordes,
and the pinnes therof. Nothing wanted of the veſſel, that was commanded
to be made for the miniſterie of the tabernacle, and for the roofe of
couenant. \V The veſtimentes alſo, which the prieſtes vſe in the
Sanctuarie, to witte, Aaron and his ſonnes, \V the children of Iſrael
offered, as our Lord had commanded. \V Which thinges after that Moyſes
ſaw al finiſhed, he bleſſed them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
