%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 0182
%%% o-0167
\startChapter[
  title={Chapter 4}
  ]

\Summary{Moyſes receiuing power to vvorke miracles in confirmation of
  his miſsion, 14.~and his brother Aaron being deſigned to aſsiſt him,
  20.~goeth vvith vvife and children towards Ægypt, 25.~is in danger to
  be ſlaine for not ſooner circumciſing his ſonne. 27.~Aaron meeteth
  him, 29.~ſo they goe together, and declare to the people, that God wil
  deliuer them.} 

Moyſes anſwering ſaid:
\LNote{They vvil not beleue me.}{Moyſes wiſely conſidering that the
children of Iſrael, much leſſe Pharao, would hardly beleue his bare
\Fix{woord,}{word,}{obvious typo, fixed in other}
affirming that he was ſent to them by God, propoſed this difficultie
before he tooke the Embacie vpon him. For without good proofe both the
Iſraelites, and Ægyptians might haue reiected him, as ſeming to come of
his owne priuate ſpirite, being no ordinarie ſuperior, neither of the
whole people, not of his owne tribe, nor firſt of his familie; for Aaron
was his elder brother.
\MNote{Miracles neceſſarie and ſufficient to proue extraordinary
vocation of new preachers.}
Therfore God gaue him powre of working miracles, to proue his
extraordinarie miſſion true and lawful. VVhich ſufficed to make euen
Pharao him ſelfe to know, that he was ſent from God almightie, though it
mollified not his ſtubborne hart, to obey Gods commandement: and it
fully ſatisfied the children of Iſrael touching al things which he
denounced, beleuing him that God mercifully looked vpon their affliction
& would deliuer them. VVherupon they adored proſtrate as the laſt
wordes of this chapter teſtifie. VVhere we ſee both the neceſſitie, and
ſufficiencie of miracles to proue the extraordinarie vocation of ſuch as
preach otherwiſe then was taught before. For this cauſe our Sauiour
himſelfe confirming his doctrin by miracles, ſaied to the Iewes:
\CNote{\XRef{Ioan.~10.}}
\Emph{If you vvil not beleue me, beleue my vvorkes.} Againe he ſaid of
them:
\CNote{\XRef{Ioan.~15.}}
\Emph{If I had not done among them vvorkes that no other man hath done,
they should not haue ſinne.}
\CNote{\XRef{Mat.~9.}
\XRef{Mat.~16.}}
And cõformably ſending his Apoſtles to preach the Goſpel, gaue them
powre to worke miracles in his name. So did S.~Peter and S.~Iohn heale
the lame.
\XRef{Act.~3.}
And S.~Paul auouched miracles for the ſignes of his Apoſtleſhip.
\XRef{2.~Cor.~12.}}
They wil not beleue me, nor heare my voice, but they wil ſay: Our Lord
hath not appeared to thee. \V Therfore he ſayd to him: What is that thou
holdeſt in thy hand? He anſwered: A rodde. \V And our Lord ſayd: Caſt it
vpon the ground. He did caſt it, and it was turned into a ſerpent, ſo
that Moyſes fled. \V And
%%% 0183
our Lord ſayd:
\Fix{Sretch}{Stretch}{obvious typo, fixed in other}
thy hand, and catch the tayle therof. He ſtretched it forth, & tooke
hold of it, and it was turned into a rodde. \V That they may beleue,
quoth he, that the Lord God of their fathers hath appeared to thee, the
God of Abraham, the God of Iſaac, & the God of Iacob. \V And our Lord
ſayd agayne: Put thy hand into thy boſome. Which when he had put into
his boſome, he brought it forth ful of leproſie like ſnow. \V Draw
backe, quoth he, thy hand into thy boſome. He drew it backe, and brought
it forth agayne, & it was like the other fleſh. \V If they wil not
beleue thee, quoth he, nor heare the word of the former ſigne, they wil
beleue the word of the ſigne folowing. \V And if ſo be they wil beleue
neither of theſe two ſignes, nor heare thy voice: take water of the
riuer, & powre it out vpon the drie land, and whatſoeuer thou draweſt of
the riuer, ſhal be turned into bloud. \V Moyſes ſayd: I beſeech thee,
Lord, I am not eloquent from yeſterday and the day before: and ſince
thou haſt ſpoken to thy ſeruant, I haue more impediment & ſlownes of
tongue. \V Our Lord ſayd to him: Who made the mouth of man? or who
framed the dumme and deafe, the ſeeing and the blinde? did not I? \V Goe
on therfore, and I wil be in thy mouth: & wil teach thee what thou ſhalt
ſpeake. \V But he ſayd: I beſech thee, Lord, ſend whom thou wilt
ſend. \V Our Lord being angrie at Moyſes, ſayd: Aaron thy brother the
\Fix{leuite,}{Leuite,}{likely typo, fixed in other}
I know that he is eloquẽt: behold he cometh forth to meete thee, &
ſeeing thee ſhal be glad at the hart. \V Speake to him, and put my wordes in
his mouth: & I wil be in thy mouth, and in his mouth, and wil ſhew you
what yee muſt doe. \V He ſhal ſpeake in thy ſteed to the people, and
shal be thy mouth: but thou shalt be to him in thoſe thinges that
perteine to God. \V
\SNote{God deſigned a rodde for an inſtrument to work miracles.}
This rodde alſo take in thy hand, wherwith thou shalt doe the ſignes. \V
Moyſes went his way, & returned to Iethro his
%%% o-0168
father in law, and ſayd to him: I wil goe and returne to my brethren
into Ægypt, that I may ſee if they be yet aliue. To whom Iethro ſayd:
Goe in peace. \V Therfore our Lord ſayd to Moyſes in Madian: Goe, and
returne into Ægypt: for they are al dead that ſought thy life. \V Moyſes
therfore tooke his wife, & his children, and ſet them vpon an aſſe: and
returned into Ægypt, carying the rodde of God in his hand. \V And our
Lord ſaid to him returning into Ægypt:
%%% 0184
See that thou doe al the wonders, which I haue put in thy hand, before
Pharao:
\SNote{See the
\XRef{Annotations cha.~7. v.~3.}}
I wil indurate his hart, and he wil not diſmiſſe the people. \V And thou
shalt ſay to him: This ſayeth the Lord: My firſt begotten ſonne is
Iſrael. \V I ſayd to thee: diſmiſſe my ſonne that he may ſerue me, &
thou ſouldeſt not diſmiſſe him: behold I wil kil thy
firſt-begotten-ſonne. \V And when he was in his iourney, in the Inne,
our Lord mette him, and would haue killed him. \V Sephora by & by tooke
a very sharp ſtone, and
\Fix{circuncided}{circumciſed}{obvious typo, fixed in other}
the prepuce of her ſonne, &
\SNote{Sephora caſt the prepuce at Moyſes feete and ſaid:
%%% (a)
I had loſt thee my ſpouſe except I had redeemed thee with the bloud of
my child. And
%%% (b)
the Angel let Moyſes goe. 
\Cite{S.~Aug. q.~11. in Exod. iuxta~70.}}
touched his feete, and ſayd:
%%% !!! This SNote refers to the above at (a).  What do we want to do
%%% here?
%%% \SNote{}
A blouddie ſpouſe thou art to me. \V And
%%% !!! This SNote refers to the above at (b).  What do we want to do
%%% here?
%%% \SNote{}
he let him goe after ſhe had ſaid, A blouddie ſpouſe thou art to me,
becauſe of the circumciſion. \V And our Lord ſaid to Aaron: Goe to
Moyſes into the deſert. Who went forth to mete him vnto the Mountaine of
God, and kiſſed him. \V And Moyſes told Aaron al the wordes of our Lord,
by which he had ſent him, & the ſignes that he had commanded. \V And
they came together, and gathered together al the ancientes of the
children of Iſrael. \V And Aaron ſpake al the wordes which our Lord had
ſaid to Moyſes: and he wrought the ſignes before the people, \V and
\SNote{Miracles a motiue to true beleefe.}
the people beleued. And they heard that our Lord had viſited the
children of Iſrael, and that he had looked vpon their affliction: & they
adored proſtrate.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
