%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 0179
%%% o-0164
\startChapter[
  title={Chapter 3}
  ]

\Summary{God appeareth to Moyſes in a bush burning but not conſuming,
  7.~deſigneth him the Gouernour of the children of Iſrael, 10.~with
  commiſsion to tel them, that they shal be deliuered from Ægypt:
  21.~and shal ſpoile the Ægyptians.}

And Moyſes fed the ſheepe of Iethro his father in law the prieſt of
Madian: and hauing driuen the flock to the inner partes of the deſert, he
came to the mountaine of God, Horeb. \V And
\LNote{Our Lord appeared.}{S.~Steuen
\CNote{\XRef{Act.~7.}}
reciting this viſion ſaieth, an Angel appeared to Moyſes: and ſo it is
in the Hebrew text, in the Chaldee Paraphraſis, and in the Septuagint
Interpreters.
\MNote{Al apparitions to the Patriarches and Prophetes were made by
Angels, though ſometimes attributed to God.}
Neither is the latin Edition (reading \Emph{Lord})
contrarie to the other which reade \Emph{Angel}, no more then one place
of holie Scripture, is contrarie to an other in the ſame language, but
very conſonant in ſenſe, ſometimes attributing the ſame apparitions and
other workes to God, as the author and principal Agent, and ſometimes to
Angels, the next and immediate miniſters of God. For ſo not only
S.~Steuen in the place alleaged, but alſo S.~Paul ſaith plainly
\XRef{(Gal.~3.)}
that \Emph{the Lavv vvas deliuered by Angels}. And in his
\XRef{Epiſtle to the Hebrewes,}
proueth the excellencie of Chriſts Law aboue the old law, by
the difference of the perſons, by whom both were geuen: affirming that
the former
\CNote{\XRef{Heb.~2.}}
\Emph{vvas ſpoken by Angels}, the \Emph{other declared by our Lord}
Ieſus Chriſt.
\MNote{Proued by holie Scriptures and Fathers.}
VVherof S.~Cyril of Alexandria diſcourſeth largely
\Cite{(li.~8. c.~2. Theſau.)}
ſhewing that in deede Angels deliuered the law, yet not by their owne
authoritie, but as ſeruants and legates of God. And before him
S.~Dionyſe of Ariopagite
\Cite{(li. cœleſt. Hierar. c.~4.)}
taught the very ſame, \Emph{the lavv} (ſayeth he) \Emph{as holie
vvrittes teſtifie, vvas geuen to vs by Angels}: yea al apparitions, made
to the ancient fathers before the law, and after it, were made by
Angels. A litle after obiecting to himſelfe, that diuine Scriptures alſo
teſtifie, that
\CNote{\XRef{Exod.~19.~20.}}
the law was geuen and granted to Moyſes by God, to teach
vs that in deede it hath the forme of ſacred and diuine law,
anſwereth, \L{eam Angelorum ad nos opera perueniſſe}, that it came to vs
(from God) by the meanes of Angels. In like maner S.~Iuſtinus Martyr
\Cite{(in explic. qq. neceſt. q.~142.)}
ſaieth, al thoſe Angels, which haue appeared in Gods place, or haue
ſpoken with men, haue alſo benne called by the name of God, as he that
ſpoke with Iob, and with Moyſes. S.~Auguſtin after a large diſcourſe of
this matter, in his
\Cite{ſecond, third, and fourth bookes de Trinitate,}
hath theſe wordes:
\Cite{(li.~4. c.~vlti.)}
If it be demanded of me, how either the voices, or ſenſible formes, and
ſhewes were made before the incarnation of the \Emph{vvord of God},
which prefigured the ſame, I anſwer that God wrought them by Angels,
which alſo I ſuppoſe I haue ſufficiently ſhewed by teſtimonies of holie
Scriptures. Likewiſe S.~Gregorie
\Cite{(Prefat. in Iob. c.~2.)}
ſaith plainly, that an Angel appeared to Moyſes in the firie buſh, yet
is called God, becauſe he was the legate of God, and therfore ſpoke, as
if God himſelfe had ſpoke in diuine Perſon, explicating the ſame by two
examples,
\MNote{Examples.}
Dauid ſaied:
\CNote{\XRef{Pſal.~77.}}
\Emph{My people attend my lavv}, yet neither the people, nor law was
Dauids, but Gods. And the reader dayly amiddes the people proclameth:
\CNote{\XRef{Exo.~3.}}
\Emph{I am the God of Abraham, the God of Iſaac, and the God of Iacob.}
Neither doth he truly ſay, that he is God, nor by that he ſayeth, doth
he goe from the rule of truth. He alſo confirmeth the ſame doctrin,
\Cite{li.~28. Moral. ca.~5.}
\MNote{God executeth his wil by Angels.}
And further teacheth that Angels protect men, and prouinces, and execute
Gods wil in this inferior world. And ſo do the other Doctors of the
Church. S.~Gregorie Nazianzen,
\Cite{orat. ad 150.~Epiſc.}
&
\Cite{orat.~2. de Theologia. in fine vtriuſque}
S.~Baſil,
\Cite{li.~3. cont. Ennom. Alexan. in fine}
S.~Athanaſius,
\Cite{ſer.~4. cont Arian, longius a princ.}
&
\Cite{Epiſt. de ſenten. Dominij Alexan. in fine.}
S.~Ambroſe,
\Cite{ſer.~1. in Pſal.~118.}
S.~Chryſoſt.
\Cite{ho.~6.}
and
S.~Hierom,
\Cite{li.~3. comment. in Mat.~18.}}
our Lord appeared to him in a flame of fire out of the middes of a buſh:
and he ſaw that the bush was on fire, and was not burnt. \V Moyſes
therfore ſaid: I wil goe, and ſee this great viſion, why the buſh is not
burnt. \V And our Lord ſeeing that he went forward to
%%% 0180
ſee, he called him out of the middes of the buſh, and ſaid: Moyſes,
Moyſes. Who anſwered: Here I am. \V But he ſaid: Approch not hither,
\SNote{See what maner of reuerence and deuotion is preſcribed, to goe
bare foote to holie places.}
looſe of thy ſhoe from thy feet: for the place, wherin thou ſtandeſt, is
\SNote{Of holie places, and of chriſtian deuotion in going to them,
S.~Hierom writeth largely.
\Cite{Epiſt.~17. 18.~&~27.}}
holie ground. \V And he ſaid: I am the God of thy father, the God of
Abraham, the God of Iſaac, and the God of Iacob. Moyſes hid his face:
for he durſt not looke againſt God. \V To whom our Lord ſaid: I haue
ſene the affliction of my people in Ægypt, and I haue heard their crye
becauſe of their rigour that ouerſee the workes: \V and knowing their
ſorow, I am deſcended to deliuer them out of the handes of the
Ægyptians, and to bring them out of that land into a land good, and
large, into a land that floweth with milke and honie, to the places of
the Chananeite, and Hetheite, and Amorrheite, and Pherezeite, and
Heueite, and Iebuſeite. \V Therfore the crye of the children of Iſrael
is come vnto me: and I haue
%%% o-0165
ſene their affliction, wherwith they are oppreſſed by the Ægyptians. \V
But come, and I wil ſend thee to Pharao, that thou mayeſt
\Fix{being}{bring}{obvious typo, fixed in other}
forth my people, the children of Iſrael out of Ægypt. \V And
\Fix{moyſes}{Moyſes}{obvious typo, fixed in other}
ſaid to God: Who am I that I ſhal goe to Pharao, and bring forth the
children of Iſrael out of Ægypt? \V Who ſaid to him: I wil be with thee:
and this thou ſhalt haue for a ſigne, that I haue ſent thee: When thou
ſhalt haue brought my people out of Ægypt, thou shalt ſacrifice to God
vpon this mountaine. \V Moyſes ſaid to God: Loe I shal goe to the
children of Iſrael, and ſay to them: The God of your fathers hath ſent me
to you. If they ſhal ſay to me: What is his name? What shal I ſay to
them? \V God ſaid to Moyſes:
\LNote{I am vvhich am.}{Al other things, beſides God, once were not; and
being limited in nature, neither could perſiſt vnles God conſerued them;
manie things alſo haue loſt, or ſhal loſe their proper eſſence and
being, and whiles he remaine haue continual alterations. Onlie God
eternally is without beginning, ending, limitation, dependence, or
mutation, cõſiſting only of himſelfe, and al other things are of him.
\MNote{The moſt proper name of God is, \Sc{He which is.}}
Therfore this name, \L{\Sc{Qvi est}}, \Sc{He which is}, is moſt proper
to God, not determining anie maner, but indeterminatly ſignifying al
maners of being, for ſo it importeth the very infinite immenſitie of
Gods ſubſtance. S.~Damaſcen.
\Cite{li.~1. c.~12. Orthodoxæ fidei.}
S.~Tho.
\Cite{p.~1. q.~13. a.~11.}}
\Sc{I am which am.} He ſaid: Thus shalt thou ſay to the children of
Iſrael: \Sc{He which is}, hath ſent me to you. \V And God ſaid againe to
Moyſes: Theſe thinges shalt thou ſay to the children of Iſrael: The Lord
God of your fathers, the God of Abraham, the God of Iſaac, and the God
of Iacob hath ſent me to you:
\SNote{This is the moſt proper name, but the moſt common is \Sc{God},
deriued in manie languages of \Emph{Good}.
\XRef{Mat.~19. v.~17.}}
this is my name for euer, and this is my memorial into generation and to
generation. \V Goe, and geather together the ancients of Iſrael, and
thou ſhalt ſay to them: The Lord God of your fathers hath appeared to me,
the God of Abraham, the God of Iſaac, and the God of Iacob, ſaying:
Viſiting I haue viſited you: and I haue ſene al thinges that haue
chanced to you in Ægypt: \V and I haue ſaid the word to bring you
%%% 0181
forth out of the affliction of Ægypt, into the land of the Chananeite,
and Hetheite, and Amorrheite, and Pherezeite, and Heueite, and Iebuſeite,
to a Land that floweth with milke & honie. \V And they ſhal heare thy
voice: and thou ſhalt enter in, thou and the ancientes of Iſrael to the
king of Ægypt, and thou ſhalt ſay to him: The Lord God of the Hebrewes
hath called vs: We wil goe three dayes iourney into the wildernes, to
ſacrifice vnto the Lord our God.

\V But I know that the king of Ægypt wil not diſmiſſe you to goe but by
mightie hand. \V For I wil ſtretch forth my hand, and wil ſtrike Ægypt
in al my marueils, which I wil doe in the middes of them: after theſe he
wil diſmiſſe you. \V And I wil geue grace to this people, in the ſight
of the Ægyptians: and when you ſhal goe forth, you ſhal not depart
emptie: \V but ech woman ſhal aske of her neighbour and of her that is
in houſe with her, veſſels of ſiluer and of gold, and rayment: and you
ſhal lay it vpon your ſonnes and daughters, and
\SNote{Al that anie man poſſeſſeth in this world, is but lent by
God. And therfore he iuſtly taketh away, and lendeth to others;
diſpoſing of al as pleaſeth him.}
ſhal ſpoyle Ægypt.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
