%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 36
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-36


\project douay-rheims


%%% 0271
%%% o-0246
\startChapter[
  title={Chapter 36}
  ]

\Summary{More being geuen then was needeful, 6.~Moyſes made to be
  proclamed that no more should be offered. 8.~So the curtines,
  13.~ringes, 18.~buckles, 19.~the couer, 20.~barres, 35.~a veile,
  36.~pillers, and a hanging are made readie.}

%%% 0272
Beſeleel therfore, and Ooliab, and euerie wiſe man, to whom our Lord
gaue wiſedome and vnderſtanding, to know how to worke artificially, made
the thinges that are neceſſarie for the vſes of the Sanctuarie, and
which our Lord did command. \V And when Moyſes had called them, and
euerie cunning man, to whom our Lord had geuen wiſedome, and ſuch as of
their owne accord had offered them ſelues to the making of the worke, \V
he deliuered al the donaries of the children of Iſrael vnto them. Who
being earneſt about the worke, the people daily in the morning did offer
their vowes. \V Whereupon
\SNote{As the people abounded in deuotion, ſo the workmen of modeſtie
and religion would haue no more then neceſſarie.
\Cite{S.~Auguſtin q.~171. in Exod.}}
the artificers being conſtrained to come, \V ſaid to Moyſes: The people
offereth more then is neceſſarie. \V Moyſes therfore commanded
proclamation to be made by the criers voice: Let neither man nor woman
offer anie more in the worke of the Sanctuarie. And ſo they ceaſed from
offering giftes, \V becauſe the thinges that were offered did ſuffice
and were ouer much. \V And al the wiſe harted men, to accompliſh the
worke of the tabernacle, made ten curtines of twiſted ſilke, and
hyacinth, and purple, and ſcarlet twiſe died, with varied worke, and the
art of embrodering: \V of which one had in length twentie eight cubites,
and in bredth foure: there was one meaſure of al the curtines. \V And he
ioyned fiue curtines, one to an other, and the other fiue he coupled to
themſelues one with an other. \V He made alſo loupes of
hyacinth in the edge of one curtine on either ſide, and in the edge of
the other curtine in like maner, \V that the loupes might meete one
againſt an other, and might be ioyned eech with other. \V Whereupon alſo
he did caſt fiftie ringes of gold, that might catch the loupes of the
curtines, and might be made one tabernacle. \V He made alſo eleuen
curtines of goates haire to couer the roofe of the tabernacle: \V one
curtine in length had thirtie cubites, & in bredth foure cubites: al the
curtines were of one meaſure: \V of which fiue he ioyned apart, & the
other ſix apart. \V And he made fiftie loupes in the edge of one
curtine, and fiftie in the edge of an other curtine, that they might be
ioyned one to an other. \V And fiftie buckles of braſſe wherwith the
roofe might be knitte together, that of al the curtines there might be
made one couering. \V He made
%%% o-0247
alſo a couer for the tabernacle of rammes skinnes died redde: & an other
couer ouer that of ianthin skinnes. \V He made
%%% 0273
alſo the bordes of the tabernacle of the wood ſetim ſtanding. \V The
length of one borde was ten cubites: and the bredth contained one cubite
and an halfe. \V There were two morteſes throughout euerie borde, that
one might be ioyned to the other. So made he in al the bordes of the
tabernacle. \V Of the which twentie were at the ſouth ſide againſt the
South, \V with fourtie feete of ſiluer. Two feete were put vnder one
borde on either ſide of the corners, where the morteſes of the ſides end
in the corners. \V At that ſide alſo of the tabernacle, that looked
toward the North, he made twentie bordes, \V with fourty feete of
ſiluer, two feete for euery borde. \V But againſt the weſt, to witte, at
that ſide of the tabernacle, which looketh to the ſea, he made ſix
bordes, \V and two other at ech corner of the tabernacle behind: \V
which were alſo ioyned from beneth vnto the toppe, & they grew together
into one connexion. So he made on either ſide at the corners \V that
there were in al eight bordes, and had ſixteene feete of ſiluer, to
witte, two feete vnder euerie borde. \V He made alſo barres of the wood
ſetim, fiue to hold together the bordes of one ſide of the
tabernacle, \V and fiue other to ioyne together the bordes of the other
ſide: and beſides theſe, fiue other barres at the weſt ſide of the
tabernacle againſt the ſea. \V He made alſo an other barre, that might
come by the middes of the bordes from corner vnto corner. \V And the
bordeworke it ſelfe he plated with gold. And their ringes he made of
gold, through which the barres might be drawen: the which alſo
themſelues he couered with plates of gold. \V He made alſo a veile of
hiacinth, and purple, ſcarlette, and twiſted ſilke, with embrodered
worke, varied and diſtinguiſhed: \V and foure pillers of the wood ſetim,
which with their heades he plated with gold, caſting their feete of
ſiluer. \V He made alſo a hanging in the entrie of the tabernacle of
hyacinth, purple, ſcarlet, and twiſted ſilke, with the worke of an
embroderer: \V and fiue pillers with their heades, which he couered with
gold, and their feete he did caſt of braſſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
