%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 0204
%%% o-0187
\startChapter[
  title={Chapter 12}
  ]

\Summary{The maner of preparing, and eating the Paſchal lambe,
  ſprinckling the dore-poſtes with bloud therof: 15.~eating no leuened bread
  ſeuen dayes together. 29.~The firſt borne of men and beaſtes among the
  Ægyptians are ſlaine. 35.~The Iſraelites goe away ſpoyling
  Ægypt. 43.~Incircumciſed men may not eate the Phaſe.}

And our Lord ſaid to Moyſes, and Aaron in the Land of Ægypt: \V This
moneth, shal be to you the beginning of moneths: it shal be the firſt in
the monethes of the yeare. \V Speake yee to
the whole aſſemblie of the children of Iſrael, & ſay to them:
\LNote{The tenth day.}{Our
\MNote{Chriſts action ſheweth that the Paſchal lambe was a figure of the
Euchariſt.}
Sauiour Chriſt inſtituting the Sacrament of the Euchariſt, after the
celebration of the Paſchal lambe, whiles they were at ſupper, the night
before his death, therby ſufficiently declared, that this old Paſch was
a figure, not only of his Paſſion and Sacrifice on the Croſſe, but alſo
of that he then did ſo ſolemnly with his Apoſtles; whom alſo in that
action he made Prieſts, commanding them, and their ſucceſſors, to do the
ſame in commemoration of him, til the end of the world. Other
circumſtances likewiſe, and conference of the one with the other make it
more clere, that as in ſome reſpectes it more reſembled Chriſts Paſſion,
and Sacrifice on the Croſſe, ſo in others it more expreſſed the
Euchariſt, and myſtical commemoration of his death, 
\MNote{Some things in the Paſchal lambe prefigured Chriſt both on the
Croſſe and at his laſt ſupper.}
though alſo in manie it prefigured Chriſt in both places. For example,
The preparing of the lambe \Emph{the tenth day} ſignified our Sauiours
coming into Hieruſalem, the ſame tenth day of the firſt moone, now
repreſented in the Church on Palmeſunday. Alſo the choiſe qualities of
the lambe, \Emph{vvithout ſpotte, a male, of the firſt yeare}, foreſhewed
in general the puritie, fortitude, meeknes, and al perfection of the
true
\CNote{\XRef{Ioan.~1.}}
\Emph{Lambe of God, that taketh avvay the ſinnes of the vvorlde}.
\MNote{Some more expreſly ſignified his Paſſion.}
More particularly \Emph{the killing} and bereuing the Paſchal Lambe of
natural life, \Emph{the ſprinkling of his bloud on the dore-poſtes, the
roſting at the fire}, and
\CNote{\XRef{Ioa.~19.}}
\Emph{not breaking anie bone therof}, moſt ſpecially expreſſed Chriſts
death on the Croſſe.
\MNote{Others immediatly the Euchariſt.}
But \Emph{the fourtenth day}, & \Emph{the euening} agree only with the
Euchariſt, inſtituted the night before our Lords Paſſion, which he
ſuffered the fiftenth (being the ful moone) and at midday, as ancient
S.~Dionyſe of Ariopagite (in two Epiſtles,
\Cite{to Policarpus,}
and
\Cite{to Appolophanes)}
teſtifieth, admiring the miracle of the ſunnes Eclipſe, that hapned the
ſame time. Neither did the \Emph{eating of the Lambe} directly
prefigurate the oblation on the Croſſe, for Chriſt was not crucified to
be eaten but the Sacrament \Emph{in formes
\Fix{and}{of}{obvious typo, fixed in other}
bread and vvine} was expreſly figured by eating the lambe
with \Emph{vnleauened bread}, and drinking \Emph{the cuppe therto
adioyned}.
\XRef{(Luc.~22. v.~17.)}
In like ſorte the Lambe immolated \Emph{in commemoration} of the
deliuerie of Iſrael from death, and from ſeruitude, when the firſt-borne
of Ægypt were ſlaine, moſt aptly prefigured the Euchariſt, which
is \Emph{a perpetual commemoration} of mans redemption, and deliuerie
from eternal death, and from bondage of the diuel and ſinne, by Chriſtes
death on the Croſſe, which death in dede was the very redemption and
deliuerie of mankind, and not a commemoration therof. Finally the
immolating of the Lambe \Emph{vvithin the houſe} with preciſe
commandment to \Emph{carie nothing therof forth}, perteined particularly
to the Euchariſt, which our Lord celebrated \Emph{vvithin the houſe},
wherby S.~Cyprian
\Cite{(lib. de vnit. Eccleſ.)}
proueth, that the B.~Sacrament muſt not be giuen to anie out of
the \Emph{Catholique Church}, though Chriſts Paſſion be extended to al
the world, as wel to bring ſuch as are without, into the Church, as to
ſaue thoſe that are already entred in.
\MNote{Ancient writers expound this figure of the Euchariſt.}
In this ſorte the moſt ancient and beſt expoſitors of holie Scripture,
explicate this ſpecial figure of the Paſchal Lambe. As we ſhal here
produce ſome witneſſes in confirmation of this truth.

Tertulian
\MNote{Tertullian  proueth, by this figure fulfilled in the Euchariſt,
that Chriſt hath a true and not a phantaſtical bodie.}
\Cite{lib.~4. contra Marcionem,}
expounding our Sauiours wordes:
\CNote{\XRef{Luc.~22.}}
\Emph{VVith deſire I haue deſired to eate this Paſch vvith you before I
ſuffer} ſaieth, Chriſt coueted not \L{veruecinam Iudæorum}, the mutton of
the Iewes, but profeſſing that with deſire he deſired to eate the Paſch,
as his owne (for it was vnmete that God ſhould couete anie thing not his
owne) the bread which he tooke, and gaue to his diſciples, he made his
owne bodie, ſaying: \Emph{This is my bodie}, that is, a figure of my
bodie. \L{Figura autem non fuiſſet, niſi veritatis eſſet
corpus.} \Emph{But it had not bene a figure} (ſaith he) \Emph{vnles it
vvere a bodie of veritie}, or \Emph{a verie bodie}, to wit, not
phantaſtical as the heretike Marcion imagined; becauſe the figures in
the old Teſtament were not figures, except a true bodie anſwered vnto
them. So the Sacramentaries ſenſe, that Tertullian ſhould cal the
Euchariſt a figure, is quite againſt his meaning, and maketh him
conclude nothing againſt Marcion; wheras his whole drift is, by the
figures of the old Teſtament to proue, that in the Euchariſt is the true
& real bodie of Chriſt, and that conſequently Chriſt hath a true and
real bodie. Origen
\Cite{(in 26.~Mat.)}
teacheth that in the great parlar (where Chriſt did eate the Paſchal
Lambe) 
%%% !!! Jeremiah. Where does this go?
%%% !!! Other has Hiere.~11. v.~15.
\CNote{\XRef{Hiere.~12. v.~19.}}
he alſo made his new Paſch.

S.~Cyprian
\Cite{(de Cæna Dom.)}
ſaieth: In the ſupper of ſacramental banquets, old and new Inſtitutions
met together. The lambe being \Emph{conſumed}, which old tradition
propoſed, the Maſter ſetteth \Emph{inconſumptible} meate to his
diſciples. S.~Gregorie Nazianzen
\Cite{(Orat.~2. de Paſcha.)}
ſaieth, God commanded the Paſchal Lambe ſhould be eaten in the euening,
becauſe Chriſt in the euening gaue the
\Fix{Sacrment}{Sacrament}{obvious typo, fixed in other}
of his owne bodie to his diſciples. S.~Hierom
\Cite{(in 26.~Mat.)}
After that the figuratiue Paſch was complete, and Chriſt had eaten the
fleſh of the lambe with his Apoſtles, he taketh bread,
\CNote{\XRef{Pſal.~103.}}
\Emph{vvhich confirmeth the hart of man}, and paſſeth ouer to the true
Sacrament of Paſch. Likewiſe S.~Chriſoſtom
\Cite{(Ho. de prodit. Iudæ)}
ſaieth, In the ſame table both the Paſches, of the figure, and of the
veritie were celebrated.
\MNote{The ſame Sacrifice offered by Prieſts.}
S.~Ambroſe
\Cite{(in Lucae.~1.)}
expreſly applieth this figuratiue lambe to the Euchariſt, as it is
celebrated in the Church, by him ſelfe and other Prieſts, ſaying: VVhen
we ſacrifice, Chriſt is preſent, Chriſt is ſacrificed: for
\CNote{\XRef{1.~Cor.~5.}}
\Emph{Chriſt our Paſch is immolated}. The like affirmeth S.~Auguſtin
\Cite{(li.~2. cont. lit. Petal. c.~27.)}
It is an other Paſch that the Iewes celebrated of a ſhepe, an other
which we receiue in the bodie and bloud of our Lord. S.~Leo
\Cite{(ſer.~7. de Paſſ.)}
To the end ſhadowes might geue place to the bodie, and figures might
ceaſe in preſence of the veritie, the old obſeruation is taken away by
the new Sacrament, hoſte paſſeth into hoſte, bloud excludeth bloud, and
when the legal feſtiuitie is changed, it is fulfilled.

S.~Gregorie
\MNote{S.~Gregories moralization of this figure, applied to the
\Fix{B.~Sracrament.}{B.~Sacrament.}{obvious typo, fixed in other}}
\Cite{(ho.~22. in Euang.)}
proueth by theſe wordes, \Emph{You shal not eate therof anie thing
ravv}, that beſides the letter there is a ſpiritual ſenſe. Behold,
(ſaieth he) the verie wordes of the hiſtorie driue vs from the
hiſtorical vnderſtanding. For did the Iſraelitical people in Ægypt vſe
to eate a lambe raw, that the law ſhould nede to ſay: \Emph{you shal not
eate it ravv}? And ſo in that homilie this great Doctor explicateth how
we ought to celebrate, and receiue the Sacrament of the Euchariſt, by
the figure of this Paſchal lambe. This \Emph{bloud} (ſaieth he)
is \Emph{ſprinkled on both poſtes}, when the Sacrament of his Paſſion is
\Emph{receiued vvith mouth}, to redemption, and mediated
with \Emph{intentiue mind} to imitation, and in \Emph{the tranſome} ouer
the dore, when pure intention directeth the exterior act, alſo when we
carie \Emph{the Croſſe} of his paſſion \Emph{in our forhead}. The fleſh
of the lambe is eaten \Emph{at night}, becauſe we now receiue our Lords
bodie in the Sacrament, when yet we \Emph{ſee not ech others
conſciences}; \Emph{roſted at the fire}, when we ioyne to our
beleefe \Emph{good vvorkes of feruent charitie}; with \Emph{vnleauened
bread}, and \Emph{vvith lettice}, that is, in ſinceritie, \Emph{vvithout
corruption of vaineglorie}, and with \Emph{bitter penance for ſinnes};
not \Emph{ravv}, nor \Emph{ſodde in vvater}, to wit, neither eſteeming
Chriſt \Emph{a mere man}, nor conſidering of him, with \Emph{humane
vviſdome} or priuate ſpirite of heretikes, called \Emph{ſtollen vvater}.
\XRef{(Prou.~9.)}
\Emph{To deuoure the head vvith the feete and entrails}, is by faith to
beleue \Emph{the Diuinitie of Chriſt}, and to imitate by loue
the \Emph{ſteppes of his humanitie}, and greedely to \Emph{learne al
Chriſtian myſteries}. Nothing is \Emph{left til morning}, when we
endeuoure in this life before \Emph{the reſurrection} to know euerie
point of Chriſtian doctrin, ſo farre as to vs perteineth. But if anie
thing \Emph{be left, it muſt be burned in the fire}, becauſe thoſe hard
and hiegheſt myſteries, which we \Emph{can not vnderſtand}, we
muſt \Emph{remitte to the Holie Ghoſt}, leſt anie proudly preſume either
to contemne, or to proclaime that he vnderſtandeth not.
\MNote{VVhat perſons are to receiue the B.~Sacrament.}
He further deſcribeth alſo what maner of perſons are to eate this new
Paſch. Their \Emph{loines} muſt be \Emph{gyrded}, that is,
al \Emph{carnal pleaſures tamed}. They muſt haue \Emph{shoes on their
feete}, by the good examples of former Sainctes dead before,
muſt \Emph{ſtrengthen their ſteppes}, to flie from vice, and follow
vertue: holding \Emph{ſtaues in their handes}, to rule & ſtay themſelues
and others \Emph{from ſliding}, by the ſtaffe of authoritie. They muſt
eate the Paſch \Emph{ſpeedely}, that is without delay or procraſtination
muſt learne the myſteries of mans redemption, and heauenlie life, and ſo
performe Gods wil and precepts, in this life \Emph{vvith ſpeede}. To
this effect S.~Gregorie diſcourſeth at large in the moral ſenſe, which
we haue abridged, and otherwiſe (though holie Scripture be ful herof)
ſeldome touch.

Returning therfore to our particular purpoſe, in al theſe teſtimonies we
ſpecially vrge, that the paſchal lambe was a figure, not only of Chriſts
Paſſion, but alſo of the Euchariſt. VVhereupon, beſides the often
expreſſe mention of our B.~Sauiours bodie and bloud in the ſame, which
Proteſtants would wreſt (as they do alſo the ſame termes in holie
Scripture) to figuratiue ſenſe, it neceſſarily followeth, that there be
farre more excellent contents in the Sacrament of the Euchariſt, then
natural bread and wine.
\MNote{The thing figured farre excelleth the figure.}
For S.~Paul teacheth
\XRef{(Colloſs.~2.)}
that \Emph{as the bodie excelleth the shadovv}, ſo the veritie, or thing
figured excelleth the figure. VVheras the ſubſtance of bread and wine
doth not excel, much leſſe ſo farre excel the Paſchal Lambe, as by
S.~Pauls doctrine is required.
\MNote{The Euchariſt is alſo a Sacrifice.}
Againe ſeing the Paſchal lambe was a Sacrifice, as appeareth in
\XRef{this Chap. v.~6.}
\XRef{&~27.}
alſo
\XRef{Num.~9. v.~7.}
\XRef{&~13.}
and
\XRef{Mar.~14. v.~12.}
and as it was immolated was a figure of the Euchariſt, as before
appeareth by conference of the one with the other, in reſpect of the
time, place, maner of offering, and eating it, and by teſtimonie of the
Doctors aboue cited, it followeth alſo that the Holie Euchariſt is a
Sacrifice farre excelling the figure.}
The tenth day of this moneth let euerie man take a lambe by their
families and houſes. \V But if the number be leſſe then may ſuffice to
eate the lambe, he shal take vnto him his neighbour that ioyneth to his
houſe, according to the number of ſoules which may ſuffice to the eating
of the lambe. \V And it shal be a lambe without ſpotte, a male, of a
yeare old: according alſo to which rite you shal take
\SNote{Such as had not meanes to take a lambe, tooke a kidde vſing the
ſame Rites.}
a kidde. \V And you shal kepe him vntil
%%% 0205
the fourtenth day of this moneth: and the whole multitude of the
children of Iſrael shal
\SNote{\HH{Shachatu}, \L{immolabunt}, \Emph{shal offer}
or \Emph{ſacrifice}: not only \Emph{kil}, as proteſtants tranſlate.}
ſacrifice him at euen. \V And they shal take of the bloud therof, and
put vpon both the poſtes, and on the vpper dore-poſtes of the houſes,
wherin they shal eate him. \V And they shal eate the flesh that night
roſted at the fire, and vnleauened bread with wilde lettice. \V You shal
not eate therof any thing raw, nor boyled in water, but only roſted at
the fire: the head with the feete and entralles therof you shal
deuoure. \V Neither shal there remaine any thing of him vntil
morning. If there be any thing left, you shal burne it with fire. \V And
thus you shal eate him: you shal gird your reynes, and you shal haue
shoes on your feete, holding ſtaues in your handes, and you shal eate
ſpeedely: for it is the
\SNote{Paſſage in killing the firſt-borne of Ægypt, and not of Iſrael.
\Cite{S.~Hiero. in Mat.~26.}}
Phaſe (that is the Paſſage) of the Lord. \V And I wil paſſe through the
Land of Ægypt that night, and wil ſtrike euery firſt begotten in the
Land of Ægypt from man euen vnto beaſt: and
\SNote{The idols of Ægypt were ouerthrowne, as Dagon was in Azotum.
\XRef{1.~Reg.~5.}
\Cite{S.~Hierom. Epiſt. ad Fabiol. exiradit. Hebr.}}
in al the goddes of Ægypt I wil doe iudgements, I the Lord. \V And the
bloud shal be vnto you for a ſigne in the houſes where you shal be: and
I shal ſee the bloud, and shal paſſe ouer you: neither shal there be
among you a deſtroying plague when I shal ſtrike the Land of Ægypt. \V
And you shal haue this day for a moniment: and you shal celebrate it
ſolemne to the Lord in your generations with an euerlaſting
obſeruation. \V Seuen dayes shal you eate azimes: in the firſt day there
shal be noe leauen in your houſes: whoſoeuer shal eate leauen, that
ſoule shal perish out
%%% o-0188
of Iſrael, from the firſt day vntil the ſeuenth day. \V The firſt day
shal be holie and ſolemne, and the ſeuenth day with the like feſtiuitie
shal be venerable: no worke shal you doe in them, except thoſe thinges,
that pertaine to eating. \V And you ſhal obſerue the azymes: for in the
ſelfe ſame day I wil bring forth your armie out of the Land of Ægypt,
and you ſhal keepe this day vnto your generations with a perpetual
rite. \V The firſt moneth, the fourtenth day of the moneth at euen you
ſhal eate
\SNote{Chriſt obſeruing this precept, had no leuened bread at his laſt
ſupper: and ſo inſtituted the Euchariſt in vnleuened.}
azymes vntil the one and twentith day of the ſame moneth at euen. \V
Seuen dayes there ſhal not be found leauened in your houſes: he that
ſhal eate leauened, his ſoule ſhal periſh out of the aſſemblie of
Iſrael, as wel of ſtrangers as of them that are borne in the land. \V
Nothing leauened ſhal you eate: in al your habitations you shal eate
azymes. \V And Moyſes
%%% 0206
called al the Ancients of the children of
\Fix{Iſael,}{Iſrael,}{obvious typo, fixed in other}
and ſaid to them: Goe take a lambe by your families, and ſacrifice the
Phaſe. \V And
\SNote{Sprinckling of bloud with hyſſop here &
\XRef{Leuit.~14.}
\XRef{Num.~19.}
preſcribed ſignifieth mãs deliuerie by Chriſts bloud working in Baptiſme
and other Sacraments.
\XRef{Heb.~9.}}
dippe a bunche of hyſſope in the bloud that is at the doore, and
ſprinkle the vppertranſome of the doore therwith, and both the doore
cheekes: let none of you goe out of the doore of his houſe til
morning. \V For our Lord wil paſſe ſtriking the Ægyptians: and when he
ſhal ſee the bloud on the vpperſil, and on both the poſtes, he wil paſſe
ouer the doore of the houſe, and not ſuffer the ſtriker to enter your
houſes and to hurt. \V Keepe this thing as a law to thee and thy
children for euer. \V And when you are entred into the Land, which our
Lord wil geue you as he hath promiſed, you ſhal obſerue theſe
ceremonies. \V And when your children ſhal ſay to you: What is this
religion? \V you ſhal ſay to them: It is the victime of our Lords
paſſage, when he paſſed ouer the houſes of the
\Fix{chidlren}{children}{obvious typo, fixed in other}
of Iſrael in Ægypt ſtriking the Ægyptians, and deliuering our houſes. And
the people bowing them ſelues adored. \V And the children of Iſrael
going forth did as our Lord had commanded Moyſes and Aaron. \V And it
came to paſſe at midnight, our Lord ſtroke
\SNote{Puniſhment conforme to their ſinne, for perſecuting
Gods \Emph{firſt begotten ſonne Iſrael.}
\XRef{Exod.~4. v.~22.}
\Cite{Theodor. q.~22. in Exod.}}
euerie firſt-begotten in the Land of Ægypt, from the firſt-begotten of
Pharao, who ſate in his throne, vnto the firſt-begotten of the captiue
woman that was in the priſon, and euerie firſt-begotten of
beaſtes. \V And Pharao aroſe in the night, and al his ſeruantes, and al
Ægypt: and there aroſe a great crie in Ægypt: for neither was there a
houſe wherin there lay not a dead one. \V And Pharao calling Moyſes and
Aaron, in the night, ſaid: Ariſe and goe forth from my people, you and
the children of Iſrael: goe, ſacrifice to the Lord as you ſay. \V Your
ſheepe and heardes take you as you demanded, and departing bleſſe me. \V
And the Ægyptians vrged the people to goe forth out of the land quickly,
ſaying: We ſhal al die. \V The people therfore tooke dough before it was
leauened: and tying it in their clokes, put it vpon their
ſhoulders. \V And the children of Iſrael did as Moyſes had commanded:
and they asked of the Ægyptians veſſels of ſiluer and gold, and very
much rayment. \V And our Lord gaue grace to the people before the
Ægyptians that they did lend them: and
\SNote{Lawful ſpoile by the warrant of God, Lord of al.}
they ſpoyled the Ægyptians. \V And the children
%%% o-0189
of Iſrael ſette forward from Rameſſe into Socoth, almoſt ſix hundred
thouſand of foote men, beſide litle ones.
%%% 0207
\V But alſo the common people of al ſortes innumerable went vp with
them, sheepe and heardes and beaſtes of diuerſe kindes exceding
manie. \V And they baked the meale, which a litle before they had taken
out of Ægypt tempered: and made hearth cakes vnleauened: for it could
not be leauened the Ægyptians vrging them to depart, & not ſuffering
them to make any tarriance: neither did they thinke vpon preparing any
meate. \V And the dwelling of the children of Iſrael that they abode
\SNote{From the promiſe made to Abraham
\XRef{(Gen.~12. v.~7.)}
and his firſt going into Ægypt
\XRef{(v.~10.)}
to this time were 430.~yeares,
\XRef{Gal.~3.}
of which they were in great perſecution aboue 80.~yeares, before that in
ſeruitude about 60.~more, before that alſo they were ſtrangers partly in
Ægypt, partly in Chanaan the reſt of the time. See
\XRef{Gen.~15. v.~13.}

The 70. read in Ægypt and in \Emph{Chanaan}, for explication, as
S.~Auguſtin noteth.
\Cite{li.~16. c.~10. ciuit.}}
in Ægypt, was foure hundred thirty yeares. \V The which being expired,
the ſame day al the armie of our Lord went forth out of the Land of
Ægypt. \V This is the obſeruable night of our Lord, when he brought them
forth out of the Land of Ægypt: this night al the children of Iſrael
muſt obſerue in their generations. \V And our Lord ſaid to Moyſes and
Aaron: This is the religion of the Phaſe: No aliene shal eate of it. \V
And euerie bought ſeruant shal be circumciſed, and ſo shal eate. \V The
ſtranger and the hireling shal not eate therof. \V In one houſe shal it
be eaten, neither shal you carrie forth of the flesh therof out of the
houſe, neither shal you breake a bone therof. \V Al the aſſemblie of the
children of Iſrael shal make it. \V And if any of the ſoiourners be
willing to dwel among you, and make the Phaſe of the Lord, firſt al the
male that he hath shal be circumciſed, and then shal he celebrate it
according to the rite: & he shal be as he that is borne in the land: but
if there be any man vncircumciſed, he shal not eate therof. \V Al one
law shal be to him that is borne in the land and to the proſelyte that
ſoiourneth with you. \V And al the children of Iſrael did as our Lord
had commanded Moyſes and Aaron. \V And the ſame day our Lord brought
forth the children of Iſrael out of the Land of Ægypt by their troupes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
