%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 24
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-24


\project douay-rheims


%%% 0246
%%% o-0224
\startChapter[
  title={Chapter 24}
  ]

\Summary{Moyſes with others are commanded to aſcend, he to the Lord, the
  reſt a farre of. 4.~They offer Sacrifice. 8.~Moyſes ſprinckleth the
  bloud of the Teſtament vpon the people. 15.~Then aſcending to the
  mountaine, God couereth it with a firie cloude.}

To Moyſes alſo he ſaid: Goe vp to the Lord, thou, and Aaron, Nadab and
Abiu, and ſeuentie Ancientes of Iſrael, and you shal adore a farre
of. \V And Moyſes onlie shal aſcend to the Lord, and they shal not
approach: neither shal the people aſcend with him. \V Moyſes therfore
came
%%% 0247
and told the people al the wordes of our Lord, and the iudgementes: and
al the people anſwered with one voice: Al the wordes of our Lord, which
he hath ſpoken we wil doe. \V And Moyſes wrote al the wordes of our
Lord: and riſing in the morning he
\SNote{As when Moyſes had brought the Iſraelites from bondage, and
receiued the law for them he built an Altar for Sacrifice: ſo Chriſt
hauing redemed vs, and geuẽ vs a Law, for applicatiõ of the fruict therof
Altares are erected, & Sacrifice offered.}
builded an Altar at the foote of the mount, & twelue titles according
the twelue tribes of Iſrael. \V And he ſent young men of the children of
Iſrael, and they offered holocauſtes, and ſacrificed pacifique victimes
to our Lord, calues. \V Moyſes therfore tooke the halfe part of the
bloud, and put it into bowles: and the reſidue he powred vpon the
Altar. \V And taking the volume of the couenant, he reade the people
hearing it: Who ſaid: Al thinges that our Lord hath ſpoken, we wil doe,
and we wil be obedient. \V And he tooke the bloud, and
\SNote{This was donne corporally to the Iewes.

In Chriſtians Chriſts bloud applied by Sacrifice and Sacraments
ſanctifieth their ſoules.
\XRef{Heb.~9.}}
ſprinkled it vpon the people, and ſaid:
\LNote{This is the bloud of the couenant.}{Our
\MNote{A figure of Chriſts bloud in the B.~Sacrament.}
Sauiour in the inſtitution of the Euchariſt, by vſing the ſame wordes,
applying them to him ſelfe,
\CNote{\XRef{Mat.~26.}}
\Emph{This is my bloud of the nevv Teſtament}, ſignifieth that he
fulfilled this figure at his laſt ſupper. VVhich proueth both a
Sacrifice of bloud then offered by him, as this bloud of the old
Teſtament was already ſhed, when Moyſes pronounced thoſe wordes; and the
real preſence of Chriſts bloud. For els, if it were but wine, it were
not better in ſubſtance then the figure, which was real bloud.
\Cite{Iſychius. li.~1. c.~4. in Leuit.}}
This is the bloud of the Couenant which our Lord hath made with you vpon
al theſe wordes. \V And there went vp Moyſes and Aaron, Nadab and Abiu, and
ſeuentie of the ancientes of Iſrael: \V and they ſaw the God of Iſrael:
and vnder his feete as it were a worke of ſapphirſtone, and as the
heauen, when it is clere. \V Neither did he ſet his hand vpon thoſe
%%% o-0225
of the children of Iſrael, that retired farre of, and they ſaw God, and
did eate, and drinke. \V And our Lord ſaid to Moyſes: Come vp to me into
the mount, and be there: and I wil geue thee tables of ſtone, and the
law, and the commandementes which I haue written: that thou maiſt teach
them. \V Moyſes roſe vp, and his miniſter Ioſue: and Moyſes aſcending
into the mount of God, \V ſaid to the Ancientes: Expect here til we
returne to you, you haue Aaron and Hur with you: If anie queſtion shal
riſe, you shal referre it to them. \V And when Moyſes was aſcended, a
cloud couered the mount, \V and the glorie of our Lord dwelt vpon Sinai,
couering it with a cloud ſix dayes, and the ſeuenth day he called him
out of the middes of the darkeneſſe. \V And the forme of the glorie of
our Lord, was as it were fire burning vpon the toppe of the mount, in
the ſight of the children of Iſrael. \V And Moyſes entring into the
middes of the cloude, aſcended into the mount: and he was there fourtie
dayes, and fourtie nightes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
