%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 40
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-40


\project douay-rheims


%%% 0279
%%% o-0252
\startChapter[
  title={Chapter 40}
  ]

\Summary{According to Gods commandment Moyſes erecteth the Tabernacle,
  vvith al thinges appertayning, the firſt day of the ſecond yeare after
  their deliuerie from Ægypt. 32.~God replenisheth the ſame vvith his
  Maieſtie, a cloude remayning ouer it by day, and a piller of fire by
  night, but vvhen they shal march, the ſame paſseth before them.}

And our Lord ſpake to Moyſes, ſaying: \V The
\SNote{The Tabernacle, prepared in the firſt yeare and erected the firſt
day of the ſecond yeare, ſignifieth the Church of Chriſt prepared in the
old Teſtament, & eſtabliſhed, exalted, and confirmed in the new.}
firſt moneth, the firſt day of the moneth, thou ſhalt erect the
tabernacle of the teſtimonie, \V and ſhalt put in it the arke, and ſhalt
let downe before it the veile: \V and bringing in the table, thou ſhalt
ſet vpon it the thinges that are commanded after the rite. The
candleſticke ſhal ſtand with the lampes therof, \V and the altar of gold
whereon the incenſe is burned, before the arke of teſtimonie. Thou ſhalt
put the hanging in the entrie of the tabernacle, \V and before it the
altar of holocauſte: \V the lauer betwen the altar and the tabernacle,
which thou ſhalt fil with water. \V And thou ſhalt compaſſe about the
court with hanginges, and the entrie therof. \V And taking the oile of
vnction thou shalt anoynte the tabernacle with the veſſel therof, that
they may be ſanctified: \V the altar of holocauſte and al the veſſel
therof: \V the lauer with the foote therof; al ſhalt thou conſecrate
with the oile of vnction, that they may be
\SNote{More holie then anie thĩg without the Sanctuarie, but
the \L{Sancta Sanctorum} it ſelfe was then moſt holie of al places in
this world.}
moſt Holie. \V And thou ſhalt bring Aaron and his ſonnes to the doore of
the tabernacle of teſtimonie, and hauing waſhed them with water, \V
thou ſhalt reueſt them with the ſacred
%%% 0280
veſtimentes, that they may miniſter to me, and the vnction of them may
proſper to an euerlaſting prieſthood. \V And Moyſes did al thinges which
our Lord had commanded. \V
%%% o-0253
Therfore the firſt moneth of the ſecond yeare, the firſt day of
the moneth, the tabernacle was placed. \V And Moyſes erected it, and put
the bordes and feete and barres, and reared the pillers, \V and ſpred
the roofe ouer the tabernacle, putting ouer it a couer, as our Lord had
commanded. \V He put alſo the teſtimonie in
\SNote{A gomor of Manna was now put in the arke mentioned before.
\XRef{chap.~16.}}
the arke, thruſting barres vnderneth, and the oracle aboue. \V And when
he had brought the arke into the tabernacle, he drew before it the veile
to fulfil the commandement of our Lord. \V He ſette the table alſo in
the tabernacle of teſtimonie at the north ſide without the veile, \V
ordering the bread of propoſition before it, as our Lord had commanded
Moyſes. \V He ſette the candleſticke alſo in the tabernacle of
teſtimonie ouer againſt the table on the ſouth ſide, \V placing the
lampes in order, according to the precept of our Lord. \V He ſet alſo
the altar of gold vnder the roofe of teſtimonie againſt the veile, \V
and burned vpon it the incenſe of ſpices, as our Lord had commanded
Moyſes. \V He put alſo the hanging in the entrie of the tabernacle of
teſtimonie, \V and the altar of holocauſte in the entrie of the
teſtimonie, offering on it the holocauſte, and the ſacrifices, as our
Lord had commanded. \V The lauer alſo he ſet betwen the tabernacle of
teſtimonie and the altar, filling it with water. \V And Moyſes and
Aaron, and his ſonnes waſhed their handes and feete, \V when they entred
the roofe of couenant, and went to the altar, as our Lord had commanded
Moyſes. \V He erected alſo the court round about the tabernacle and the
altar, drawing the hanging in the entrie therof. After al thinges were
perfited, \V the cloude couered the tabernacle of teſtimonie, and the
glorie of our Lord filled it. \V
\SNote{VVithout al doubt (ſaith
\Cite{S.~Auguſtin. q.~173. in Exod.)}
Moyſes prefigured other perſons when he entred into the cloud on mount
Sinai, and others now when he could not enter into the tabernacle
repleniſhed with the glorie of God. In Sinai he ſignified thoſe that
penetrate the profound myſteries of Chriſt, here the Iewes who vnderſtãd
not the ſame.}
Neither could Moyſes enter the roofe of couenant, the cloude couering al
thinges, and the maieſtie of our Lord shining, becauſe the cloude had
couered al thinges. \V If at anie time the cloud did leaue the
tabernacle, the children of Iſrael went forward by their troupes: \V If
it hong ouer, they remained in the ſame place. \V For the cloude of our
Lord honge ouer the tabernacle by day, and a fire by night, in the ſight
of al the children of Iſrael throughout al their manſions.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
