%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 0177
%%% o-0162
\startChapter[
  title={Chapter 2}
  ]

\Summary{A child of the Hebrewes, and Tribe of Leui, being expoſed to
  the water, 5.~is taken from thence by Pharaos daughter, 8.~who
  committeth him to be nurſed, vnwitting to his owne mother, adopteth
  him and calleth him Moyſes. 11.~He afterwardes viſiting his brethren,
  killeth an Ægyptian; 15.~flieth into Madian; 21.~marrieth a wife, and
  hath two ſonnes.}

After theſe thinges there came forth a man of the houſe of Leui: and he
tooke a wife of his owne ſtocke. \V Who conceaued, and bare a ſonne: and
ſeing him a goodlie one, hid him three monethes. \V And
\LNote{VVhen she could not conceal him.}{Theſe
\MNote{Moyſes parẽts did prudently expoſe him to ſome danger, to auoide
greater.}
godlie and prudent parents, conſidering that when the Ægyptians ſhould
perceiue ſuch an infant to be borne, and not drowned according to the
Kings Edict, they would deſtroy both the child, and whole familie:
to auoid the greater danger, choſe the leſſe. To bring him to the water
ſide, not omitting their owne induſtrie, as wel by cloſing him in a
basket, that would draw no water, as by ſetting his ſiſter to watch what
became of him: that if better ſucceſſe happened not the firſt day, the
mother might at euening geue him ſuck, and miniſter other neceſſaries;
and ſo expect an other day, or manie dayes Gods prouidence, til his
diuine pleaſure ſhould more appeare.

Ioſephus
\MNote{Reuelations and Gods determinations do not exclude but include
mans endeuour.}
\CNote{\Cite{li.~2. Antiq.}}
writeth that Amran Moyſes father, being ſolicitous, when his wife was
great, how to ſaue the infant, if it were a man child, God reueled to
him, that ſhe had conceiued a ſonne, who ſhould not only be ſaued from
Pharaos furie, but alſo be the deliuerer of the whole Hebrew nation from
thraldome, and ſeruitude of the Ægyptians. VVherupon they aſſuredly
truſted that God would protect and proſper him, yet ſo, if they did
their owne endeuour, which
\CNote{\Cite{li.~16. c.~19. de ciuit.}}
S.~Auguſtin teacheth to be alwayes neceſſarie.}
when now ſhe could not conceale him, ſhe tooke a basket made of
bulruſhes, and dawbed it with
\TNote{a kind of glevv, ſo called.}
bitume and pitch: and put with in it the litle infant, and laid him in a
ſedgie place by the riuers brinke, \V his ſiſter ſtanding a farre of,
and conſidering the euent of the thing. \V And behold the daughter of
Pharao came downe to be washed in the riuer: and her maides walked by
the riuers brinke. Who when she ſaw the basket in the ſedges, she ſent
one of her handmaides: and when it was brought \V opening it, and ſeeing
within it an infant crying, hauing pitty on it,
ſaid: This is one of the infantes of the Hebrewes. \V To whom the
childes ſiſter ſaid: Wilt thou that I goe, & cal to thee an Hebrew woman,
that may nurſe the litle infant? \V She anſwered: Goe. The maid went and
called her mother. \V To whom Pharaos daughter ſpeaking: Take, quoth ſhe,
this child, and nurſe him for me:
%%% 0178
I wil geue thee thy hyre. The woman tooke, and nurſed the child: and
when he was growen, deliuered him to Pharaos daughter. \V Whom ſhe
adopted into the place of a ſonne, and called him
\SNote{\Emph{Mos}, in the Ægyptian tongue ſignifieth \Emph{vvater},
and \Emph{Iſes}, \Emph{ſaued}.
\Cite{Ioſeph. li.~2. Antiq.}
&
\Cite{Clemens. Alexan. li.~1. Stromat.}}
Moyſes, ſaying: Becauſe from the water I did take him. \V In thoſe dayes
after that Moyſes was growen, he went forth to his brethren: and he ſaw
their affliction, and a man that was an Ægyptian ſtriking one of the
%%% o-0163
Hebrewes his brethren. \V And when he had looked about hither & thither,
and ſaw no man preſent,
\LNote{He ſtroke the Ægyptian.}{Moyſes
\MNote{Though Moyſes iuſtly killed the Ægyptian, yet others may not
imitate his example.}
not of carnal loue towards his brethren, nor of priuate paſſion, but by
diuine inſpiration killed the Ægyptian, as S.~Auguſtin proueth
\Cite{(li.~qq. in Exod. q.~2.)}
by the teſtimonie of S.~Steuen ſaying:
\CNote{\XRef{Act.~7.}}
Moyſes \Emph{thought his brethren had vnderſtood, that God by his hand,
vvould ſaue them.} VVherby appeareth that Moyſes himſelfe knew it was
Gods pleaſure, he ſhould kil that Ægyptian inuading an Hebrew. Yet
others may not imitate ſuch particular examples.
\Cite{Catech. Rom. p.~3. c.~6. q.~5.}}
he ſtroke the Ægyptian, and hid him in the ſand. \V And going forth an
other day, he ſaw two Hebrewes brawling: and he ſaid to him that did the
wrong: Why ſtrikeſt thou thy neighbour? \V Who anſwered:
\SNote{The guiltie perſõ reiected Moyſes for lack of knowẽ authoritie,
but God cõfirmed his cõmiſſion.
\XRef{Act.~7.}
So the Iewes reiected Chriſt, Iudge of the world.}
Who hath appointed thee prince & iudge ouer vs? Wilt thou kil me, as
yeſterday thou dideſt the Ægyptian? Moyſes
\SNote{He feared to tempt God by ſtaying, but \Emph{not the fiercenes of
the king.}
\XRef{Heb.~11.}}
feared, and ſaid: How is this thing come abroad? \V And Pharao heard of
this talke, and ſought to kil Moyſes: who fleeing from his ſight, abode
in the Land of Madian, and ſate beſide a wel. \V And the prieſt of
Madian had ſeuen daughters, which were come to draw water: and when the
troughes were filled, they deſired to water their fathers flockes. \V
The ſhepeheardes came vpon them, and droue them away: and Moyſes aroſe,
and defending the maides, watered their ſheepe. \V Who being returned to
Raguel their father, he ſaid to them: Why are you come ſooner then you
were
\Fix{woont?}{wont?}{possible typo, fixed in other}
\V They anſwered: A certaine man an Ægyptian deliuered vs from the hand
of the ſhepheardes: moreouer alſo he drew water with vs, and gaue the
sheepe to drinke. \V But he ſaid: Where is he? Why haue you let the man
goe? cal him that he may eate bread. \V Therfore Moyſes ſware that he
would dwel with him. And he tooke Sephora his daughter to wife: \V who
bare him a ſonne, whom he called Gerſam, ſaying: I haue bene a ſtranger
in a forren countrey. And she bare him an other, whom he called Eliezer,
ſaying: for the God of my father my helper hath deliuered me out of the
hand of Pharao. \V But after much time the king of Ægypt died: and the
children of Iſrael groning, cried out becauſe of the workes: and
\SNote{Oppreſſion of innocents crieth to heauen for reuenge.}
their crie aſcended vnto God from the workes. \V And he heard their
groning, & remembred the couenant which he made with Abraham, Iſaac, and
Iacob. \V And our Lord looked vpon the children of Iſrael and knew
them. 


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
