%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 38
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-38


\project douay-rheims


%%% 0275
%%% o-0249
\startChapter[
  title={Chapter 38}
  ]

\Summary{The ſame Beſeleel maketh the Altar of Holocauſte, 8.~the braſen
  lauer, 9.~the court with pillers and hanginges. 21.~The giftes that
  were offered are recited.}

He made alſo the Altar of Holocauſte of the wood ſetim, fiue cubites
foure ſquare, and three in height: \V the hornes wherof did procede from
the corners, and he couered it with plates of braſſe. \V And for the
vſes therof he prepared of braſſe diuers veſſels, cauldrons, tonges,
fleſh-hookes, pothookes, & firepannes. \V And the grate therof in maner
of nette he made of braſſe, and vnder it in the middes of the altar an
hearth, \V caſting foure ringes at as manie toppes of the nette, to put
in barres to carie it: \V the which themſelues alſo he made of the wood
ſetim, and couered them with plates of braſſe: \V and he drew them
through the ringes, that ſtoode out in the ſides of the altar. And the
altar it ſelfe was not maſſie, but holow of bordes, and within
emptie. \V He made alſo the lauer of braſſe, with the foote therof, of
wemens glaſſes,
\SNote{Theſe wemen watched there for deuotion, and it ſemeth the ſame
cuſtome continued til Chriſts time. For Anna the widow obſerued this
ſtate of life.
\XRef{Luc.~1.}}
that watched in the doore of the tabernacle. \V He made alſo the court,
in the ſouth ſide wherof were hanginges of twiſted ſilke, of an hundred
cubites, \V twentie pillers of braſſe with their feete,
%%% 0276
the heades of the pillers, & the whole grauing of the worke, of
ſiluer. \V In like maner at the north ſide the hanginges, pillers, and
feete and the heades of the pillers were of the ſame meaſure, and worke
and metal. \V But on that ſide that looketh to the Weſt, there were
hanginges of fiftie cubites, ten braſen pillers with their feete, and
the heades of the pillers, and al the grauing of the worke, of
ſiluer. \V Moreouer againſt the Eaſt he prepared
\Fix{haginges}{hanginges}{obvious typo, same in other}
of fiftie cubites: \V of the which, one ſide conteyned fiftene cubites
of three pillers, with their feete: \V and on the other ſide (becauſe
betwen both he made the entrie of the tabernacle) there were hanginges
equally of fiftene cubites, and three pillers, and feete as manie. \V Al
the hanginges of the court were wouen of twiſted ſilke. \V The feete of
the pillers were of braſſe, and their heades with al their grauinges of
ſiluer: but the pillers alſo of the court them ſelues he plated with
ſiluer. \V And in the entrie therof he made with embrodered worke a
hanging of hyacinth, purple, ſcarlet, and twiſted ſilke, that had
twentie cubites in length, but the height was fiue cubites according to
the meaſure, which al the hanginges of the court had. \V And the pillers
in the entrie were foure with feete of braſſe, and their heades and
%%% o-0250
grauinges of ſiluer. \V The pinnes alſo of the tabernacle and of
\Fix{tpe}{the}{obvious typo, fixed in other}
court round about he made of braſſe. \V Theſe are the inſtrumentes of the

\Fix{taberbacle}{tabernacle}{obvious typo, fixed in other}
of teſtimonie, which were numbered according to the precept
of Moyſes, in the ceremonies of the Leuites by the hand of Ithamar the
ſonne of Aaron the prieſt: \V which Beſeleel the ſonne of Vri, the ſonne
of Hur, of the tribe of Iuda had accomplished, as our Lord commanded by
Moyſes, \V hauing ioyned to him ſelfe for his compagnion Ooliab the
ſonne of Achiſamech of the tribe of Dan: who was himſelf alſo an
egregious artificer in wood, and a tapiſter and embroderer of hyacinth,
purple, ſcarlet, and ſilke. \V Al the gold that was ſpent in the worke
of the Sanctuarie, and that was offered in donaries, was nine and
twentie talentes, and ſeuen hundred thirtie ſicles according to the
meaſure of the Sanctuarie. \V And it was offered of them that paſſed to
the number, from twentie yeares and vpward, of ſix hundred three
thouſand, and fiue hundred fiftie, able men to beare armes. \V There
were moreouer an hundred talentes of ſiluer, wherof were caſt the feete
of the Sanctuarie, and of the entrie where
%%% 0277
the veile hangeth. \V An hundred feete were made of an hundred talentes,
one talent being accoũted for euerie foote. \V And of the thouſand ſeuen
hundred, and ſeuentie fiue he made the heades of the pillers, which them
ſelues he alſo plated with ſiluer. \V Of braſſe alſo there were offered
ſeuentie two thouſand talentes, and foure hundred ſicles beſides, \V of
the which were caſt the feete in the entrie of the tabernacle of
teſtimonie, and the altar of braſſe with the grate therof, and al the
veſſel, that pertayne to the vſe therof, \V and the feete of the court
as wel in the circuite as in the entrie therof, and the pinnes of the
tabernacle and of the court, round about.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
