%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 0214
%%% o-0195
\startChapter[
  title={Chapter 15}
  ]

\Summary{Moyſes with the people ſing a Canticle of thanks-geuing, for
  their deliuerie. 22.~The people being three daies in the deſert
  without water, then finding that is bitter, do murmure. 25.~It is made
  ſwete. 27.~Coming to Elim they finde twelue fountaines, and ſeuentie
  palmetrees.}

Then
\SNote{The firſt of al Canticles, ſacred or prophane.
\Cite{Origen. ho.~6. in Exod.}}
ſang Moyſes and the children of Iſrael this ſong to our Lord, and ſaid:

Let vs ſing to our Lord: for he is glouriouſly magnified, the horſe and
the rider he hath throwen into the ſea.

\V My ſtrength, and my praiſe is our Lord, and he is made vnto me a
ſaluation: this is my God, and I wil glorifie him: the God of my father,
and I wil exalt him.

\V Our Lord is a man of warre, omnipotent is his name.

\V Pharaoes chariottes and his armie
\SNote{God only ſuffered them to goe into the ſea. For they went of
their owne accord ſuppoſing they might folowe where the Iſraelites went
before.
\Cite{S.~Aug. ſer.~89. de temp.}}
he hath caſt into the ſea: his choſen princes are drowned in the red
ſea.

\V The depthes haue ouerwhelmed them, they are ſonke into the botome
like a ſtone.

\V Thy right hand ô Lord is magnified in ſtrength: thy right hand, ô
Lord, hath ſtriken the enimie.

\V And in the multitude of thy glorie thou haſt put downe thy
aduerſaries: thou haſt ſent thy wrath, which hath deuoured them like
ſtuble.

\V And in the ſpirite of thy furie were the waters gathered together:
the flowing water ſtoode, the depthes were gathered together in the
middes of the ſea.

\V The enimie ſaid: I wil purſew and ouertake, I wil diuide the ſpoiles,
my ſoule ſhal haue his fil: I wil draw forth my ſword, my hand ſhal kil
them.

\V The ſpirite blewe and the ſea ouerwhelmed them: they ſanke as lead in
the vehement waters.

\V Who is like to thee, among the ſtrong ô Lord? who is like to thee,
magnifical in ſanctitie, terrible and laudable, doing meruailes?

\V Thou didſt ſtretch forth thy hand, and the earth deuoured them.

\V Thou haſt in thy mercie bene a guide to the people which thou haſt
redemed: and in thy ſtrength thou haſt
%%% 0215
caried them vnto thy holie habitation.

\V Nations roſe vp, and were angrie: ſorowes poſſeſſed the inhabiters of
Philiſthijm.

%%% o-0196
\V Then were the princes of Edom trubled, trembling ceazed on the
ſturdie of Moab: al the inhabiters of Chanaan were ſtarke.

\V Let feare and dread fal vpon them, in the greatnes of thy arme: let
them become vnmoueable as a ſtone, vntil thy people ô Lord ſhal paſſe,
vntil thy people ſhal paſſe, this which thou haſt poſſeſſed.

\V Thou ſhalt bring them in, and plant them in the mountaine of thy
inheritance, in thy moſt firme habitation, which thou haſt wrought ô
Lord: thy ſanctuarie Lord, which thy handes haue confirmed.

\V Our Lord ſhal reigne for euer and euermore.

\V For Pharao on horſebake entred in with his chariottes and horſemen
into the ſea: and our Lord brought backe vpon them the waters of the
ſea: but the children of Iſrael walked on drie ground in the
middes therof.

\V Marie therfore the propheteſſe, Aarons ſiſter, tooke
\SNote{Muſical inſtrumẽts vſed before the law of Moyſes in the ſeruice
of God.}
a tymbrel in her hand: and al the wemen went forth after her with
tymbrels and daunces, \V to whom ſhe beganne the ſong, ſaying: Let vs
ſing to our Lord, for he is gloriouſly magnified, the horſe and his
rider he hath caſt into the ſea.

\V And Moyſes remoued Iſrael from
\SNote{Theſe things chanced to them in figure.
\XRef{1.~Cor.~10.}}
the red ſea, and they went forth into the deſert Sur: and they walked
three dayes through the wilderneſſe, and found not water. \V And they
came into Mara, neither could they drinke the waters of Mara, becauſe
they were bitter: wherupon he gaue a name alſo agreable to the place,
calling it Mara, that is, bitterneſſe. \V And the people murmured
againſt Moyſes, ſaying: What ſhal we drinke? \V But he cried to our
Lord. Who did ſhew him
\SNote{The wholſome wood of the Croſſe made the bitter ſea of gentiles,
ſwete.
\Cite{Theodoret. q.~26. in Exod.}}
a peece of woode: which when he had caſt into the waters, they were turned
into ſweteneſſe. There he appointed him precepts, and iudgements, and
there he proued him, \V ſaying: If thou wilt heare the voice of the Lord
thy God, and doe that is right before him, and obey his commandementes,
and keepe al his preceptes, none of the maladies, that I layd vpon Ægypt,
wil I bring vpon thee: for I am the Lord God thy curer. \V And the
children of Iſrael came into Elim, where there were twelue fountaines of
water, & ſeuentie palme trees: and they camped byſide the waters.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
