%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 0197
%%% o-0180
\startChapter[
  title={Chapter 8}
  ]

\Summary{The ſecond plague is of frogges. 7.~The enchanters make the
  like. 8.~Pharao promiſeth to let the Iſraelites goe and ſacrifce, ſo
  the frogges be taken away. 13.~Which being donne he breaketh
  promiſe. 16.~The third plague is of ſciniphes, 18.~which the
  enchanters can not make. 21.~The fourth is of flies. 29.~Pharao againe
  promiſeth to diſmiſse the people of God, but doth it not.}

Our Lord alſo ſaid to Moyſes: Goe in to Pharao, and thou shalt ſay vnto
him: This ſaith the Lord: Diſmiſſe my people, for to ſacrifice vnto
me. \V But
\SNote{If Pharao had not freewil threatning of puniſhmẽt were vniuſt.
\Cite{Origen. li.~3. Periar. c. de lib. arbrit.}
He that can not do otherwiſe doth not ſinne, as both lerned and vnlerned
cõfeſſe.
\Cite{S.~Aug. de vera Relig. c.~14.}}
if thou wilt not diſmiſſe them, behold I wil ſtrike al thy coaſtes with
frogges. \V And the riuer shal bubble with frogges: which shal come vp,
and enter into thy houſe, and thy bed chamber, and vpon thy bedde, and
into the houſes of thy ſeruantes, and vnto thy people, and into thy
ouens, and into the remaines of thy meates: \V and vnto thee, & to thy
people, and to al thy ſeruantes shal the frogges enter. \V And our Lord
ſaid to Moyſes: Say vnto Aaron: Stretch forth thy hand vpon the floudes,
and vpon the riuers and the pooles, and bring forth
\MNote{The 2.~plague. Multitude of frogges.}
frogges vpon the Land of Ægypt. \V And Aaron ſtretched forth his hand
vpon the waters of Ægypt, and the frogges came vp, and couered the Land
of Ægypt. \V And the enchanters alſo by their enchantments did in like
maner, and they brought forth frogges vpon the Land of Ægypt. \V And
Pharao called Moyſes & Aaron, and ſaid to them:
\SNote{The Enchãters could bring more frogges, but not take theſe away.}
Pray ye to the Lord to take away the frogges from me & from my people:
and I wil diſmiſſe the people to ſacrifice vnto the Lord. \V And Moyſes
ſaid to Pharao: Appoint me when I shal pray for thee, and for thy
ſeruantes, and for thy people, that the frogges may be driuen away from
thee and from thy houſe, and from thy ſeruantes, and from thy people:
and may remaine only in the riuer. \V Who anſwered: To morow. But he
ſaid: According to thy word wil I doe: that thou maiſt know that there
is not the like to the Lord our God. \V And the frogges ſhal depart from
thee, and from thy houſe, and from thy ſeruantes, and from thy people:
and ſhal remaine only in the riuer. \V And Moyſes and Aaron went forth
from Pharao: and Moyſes cried to our Lord for
%%% 0198
the promiſe, concerning the frogges, which he had agreed to Pharao. \V
And our Lord did according to the word of Moyſes: and the frogges dyed
out of the houſes, and out of the villages, and out of the fieldes: \V
and they gathered them together into huge heapes, and the earth did
rotte. \V And Pharao ſeing that reſt was geuen
\SNote{Pharaos induration aſcribed to himſelf.}
he hardned his owne hart; and heard them not, as
%%% o-0181
our Lord had
commanded. \V And our Lord ſaid to Moyſes: Speake to Aaron: Stretch
forth thy rodde, and ſtrike the duſt of the earth: and be there
\MNote{The 3.~plague. Sciniphes, ſmale flying beaſtes, eſpecially
moleſting mens eyes.
\Cite{Philo. l.~1. de vita Moyſi.}}
Sciniphes in the whole
\Fix{Lord}{Land}{obvious typo, fixed in other}
of Ægypt. \V And they did ſo. And Aaron ſtretched forth his hand,
holding the rodde: and he ſtroke the duſt of the earth, and there were
made ſciniphes on men and on beaſtes: al the duſt of the earth was
turned into ſciniphes through the whole Land of Ægypt. \V And the
enchaunters with their enchauntmentes practiſed in like maner, to bring
forth ſciniphes, and
\SNote{The diuels powre limited by God.
\XRef{Iob.~1.~2.}}
they could not: and there were ſciniphes as wel on men as on beaſtes. \V
And the enchaunters ſaid to Pharao:
\SNote{The enchanters conuinced in their vnderſtanding, confeſſed the
powre of God, but not changed in affection, perſiſted in malice againſt
the truth.}
This is the finger of God. And Pharaoes hart was indurate, and he heard
them not as our Lord had commanded. \V Our Lord alſo ſaid to Moyſes:
Ariſe early, and ſtand before Pharao: for he wil goe forth to the waters:
and thou ſhalt ſay to him: This ſaith our Lord: Diſmiſſe my people to
ſacrifice vnto me. \V And if thou wilt not diſmiſſe them, behold I wil
ſend in vpon thee, and vpon thy ſeruantes, and vpon thy people, and vpon
thy houſes al kind of
\MNote{The 4.~plague. Abundance of al ſortes of flies.}
flies: and the houſes of Ægypt ſhal be filled with flies of diuers
kindes, and the whole land wherein they ſhal be. \V And I wil make the
Land of Geſſen merueilous in that day, wherin my people is, ſo that
flies ſhal not be there: and thou ſhalt know that I am the Lord in the
middes of the earth. \V And I wil put a diuiſion betwene my people & thy
people: to morow ſhal this ſigne be. \V And our Lord did ſo. And there
came a very greuious flie into the houſes of Pharao and of his
ſeruantes, and into al the Land of Ægypt: and the Land was corrupted by
ſuch kind of flies. \V And Pharao called Moyſes and Aaron, and ſaid
to them: Goe and ſacrifice to your God in this land. \V And Moyſes ſaid:
It can not ſo be done: for if we ſhal offer the abominations of the
Ægyptians to the Lord our God: and
\SNote{Ægyptians worſhipping beaſts thought it intolerable abomination
to kil, or eate, or burne them in ſacrifice.
\XRef{Gen.~43. v.~32.}
\XRef{46.~v.~34.}}
we kil thoſe thinges which the Ægyptians doe worſhippe before them: they
wil beate vs downe with ſtones. \V We wil goe forth three dayes
%%% 0199
iourney into the wildernes: and we wil ſacrifice vnto the Lord our God,
as he hath commanded vs. \V And Pharao ſaid: I wil diſmiſſe you to
ſacrifice to the Lord your God in the deſert: but goe noe farder: pray
for me. \V And Moyſes ſaid: Being gone forth from thee, I wil pray to
our Lord: and the flie shal depart from Pharao, and from his ſeruantes,
and from his people to morow: but deceiue no more ſo, that thou wilt not
diſmiſſe the people to ſacrifice vnto our Lord. \V And Moyſes being gone
forth from Pharao, prayed our Lord. \V Who did according to his word: and
he tooke away the flies from Pharao, and from his ſeruants, and from his
people: there was left not ſo much as one. \V And Pharaoes hart
\SNote{In the Hebrew: \Emph{Pharao hardned his ovvne hart, alſo this
time.}}
was hardned, ſo that neither this time would he diſmiſſe the people.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
