%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 0226
%%% o-0207
\startChapter[
  title={Chapter 16}
  ]

\Summary{The
\MNote{The beginning of the fourth age.}
people murmuring for meate, and that they had left the flesh pottes of
Ægypt, 4.~God geueth them quailes, and Manna. 16.~VVherof they are
commanded to gather for euerie day, 22.~but the ſixt day duble for the
Sabbaoth, 32.~and to kepe a meaſure of it in the tabernacle for a
memorie.}

And
\MNote{The ſecond parte of this booke. How the Iſraelites were ſuſtained
in the deſert, & prepared to receiue the Law.}
they ſette forward from Elim, and al the multitude of the children of
Iſrael came into the deſert Sin,
%%% 0227
which is betwene Elim & Sinay: the fiftenth day of the ſecond moneth,
after they came forth out of the Land of Ægypt. \V And al the aſſemblie
of the children of Iſrael murmured againſt Moyſes and Aaron in the
wilderneſſe. \V And the children of Iſrael ſaid to them: Would to God we
had died by the hand of our Lord in the Land of Ægypt, when we ſate ouer
the fleſh pottes, and ded eate bread our fill: why haue you brought vs
into this deſert, that you might kil al the multitude with famine? \V
And our Lord ſaid to Moyſes: Behold I wil raine you bread from heauen:
let the people goe forth, and gather that ſufficeth for euerie day:
\SNote{God leaft it in their wil to be content with ynough, or to couete
more, yet ſuffered them not to haue more, when it came to meaſuring.
\XRef{v.~18.}
&
\XRef{2.~Cor.~8.}}
that I may proue them whether they wil walke in my law, or no. \V But
the ſixt day let them prouide for to bring in: and let it be duble to
that they were wont to gather euerie daie. \V And Moyſes and Aaron ſaid
to al the children of Iſrael: At euen you ſhal know that our Lord hath
brought you forth out of the land of Ægypt: \V and in the morning you
ſhal ſee the glorie of our Lord: for he hath heard your murmuring
againſt our Lord: but as for vs, what are we, that you mutter againſt
vs? \V And Moyſes ſaid: At euen our Lord wil geue you flesh to eate, and
in the morning bread your fill: for he hath heard your murmuringes,
which you haue murmured againſt him, for what are we? neither is your
murmuring againſt vs, but againſt our Lord. \V Moyſes alſo ſaid to
Aaron: Say to the whole aſſemblie of the children of Iſrael: Approch you
before our Lord: for he hath heard your murmuring. \V And when Aaron
ſpake to al the aſſemblie of the children of Iſrael, they looked toward
the wilderneſſe: and behold the glorie of our Lord appeared in a
cloud. \V And our Lord ſpake to Moyſes, ſaying: \V I haue heard the
murmuringes of the children of Iſrael, ſay to them: At euen you shal
eate fleſh, and in the morning you ſhal haue your fil of bread: and you
ſhal know that I am the Lord your God. \V Therfore it came to paſſe at
euen, and
\SNote{Theſe birdes by Gods prouidence came from other places to the
children of Iſrael.
\XRef{Nu.~11. v.~31.}}
the quaile roſe, and couered the campe: in the morning alſo a dew lay
round about the campe. \V And when it had couered the face of the earth,
it appeared in the wilderneſſe ſmal, and as it were beaten with a peſtil
like vnto the hoare froſt on the ground. \V Which when the children of
Iſrael had ſeene, they ſaid one to an
%%% o-0208
other:
\LNote{\HH{Manhu}! vvhat is this!}{VVhen
\MNote{Manna ſo called of \HH{Man-hu}.}
the people of Iſrael in the deſert had ſpent their prouiſion of meate
brought from Ægypt, and according to Gods promiſe had receiued ſtore of
quailes; going forth in the morning they ſawe a ſtrange thing lye vpon
the ground like to hoare froſt, wherat merueling they ſaid one to an
other: \Emph{VVhat is this}! in their
language \HH{Manhu}! \Emph{VVherupon} ſaith Theodoret
\Cite{(q.~30. in Exod.)}
\Emph{their demand vvas turned into the name, and it vvas called Manna.}
VVhich as the ſame and other ancient Doctors gather by the holie
Scriptures, was a wonderful and miraculous meate,
\MNote{It was a figure of the Euchariſt.}
and withal a figure of a more excellent thing, long after promiſed, and
geuen by our B.~Sauiour, in the holie Sacrament of the Euchariſt. As
witneſſe S.~Gregorie Nyſſen
\Cite{(Enar. vitæ Moyſi, circa medium)}
S.~Ambroſe
\Cite{(li. de ijs qui Myſt. initiant  cap.~8.)}
%%% !!! How do theſse line up?
\CNote{\Cite{li.~3. c.~37.}}
S.~Cyril Alexandrinus,
\CNote{\Cite{Ho.~45.}}
S.~Chryſoſtom, S.~Auguſtin, 
\CNote{\Cite{Tract.~26.}}
Theophilact and others, vpon the
%%% Maybe these could be attached to the proper quotes below?
\CNote{\XRef{Ioan.~6. v.~25. 41. 49. 51. 55.}}
\XRef{ſixt of S.~Iohn.}
VVhere alſo the text of our Sauiours long diſcourſe with the
Capharnaites, ſheweth euidently that he promiſed to geue a farre better
meate then Manna, to thoſe that beleued in him. \Emph{I am} ſaieth our
Lord, \Emph{the bread of life, vvhich deſcended from heauen; your
fathers did eate Manna in the deſert, and died. The bread vvhich I vvil
geue is my flesh, for the life of the vvorld. My flesh is meate in
deede, and my bloud is drinck in deede &c.} S.~Paul likewiſe teacheth
\XRef{(1.~Cor.~10.)}
that this \Emph{Manna}, and \Emph{the vvater} iſſuing out \Emph{of the
rocke}, were figures of the ſame B.~Sacrament, as is noted in thoſe
places.

Here
\MNote{Twelue miracles in Manna.}
only we commend to the Chriſtian readers remẽbrance that the thing
figured, doth euer excede the figure (according to S.~Pauls doctrin,
\XRef{Colloſ.~2.)}
wiſhing him therfore to conſider, that in \Emph{Manna} were at leaſt
twelue clere miracles.
%%% !!! Where do these go?
\CNote{\XRef{Pſal.~77. v.~25.}}
\CNote{\Cite{Roſſin. li.~1. c.~12. cont. Oecolamp.}}
\CNote{\Cite{D.~Heſk. li.~3. c.~12. parlam.}}
\MNote{1.}
Firſt, it was made by Angels, wherof it is called, \Emph{the bread of
Angels}.
\MNote{2.}
Secondly, it was not produced from the earth, nor water, as ordinarie
meates are, but came from the ayre.
\MNote{3.}
Thirdly, how faſt or ſlowly ſoeuer anie man did gather it, in the end
ech one had the ſame meaſure ful, called a gomor, and no more nor
leſſe.
\MNote{4.}
Fourthly, the ſixth day (which was next before the Sabbath) that which
they gathered, was found to be duble portions to other dayes, that is,
two gomors for euerie one.
\MNote{5.}
Fiftly, there fel none at al on the Sabbath day.
\MNote{6.}
Sixtly, if in the reſt of the weke anie part was left al night, it
putrified, and was corrupt in the morning, but the night before the
Sabbath day, it remained ſound and good.
\MNote{7.}
Seuently, notwithſtanding diuerſitie of ſtomakes, in ſo great a
multitude, the ſame meaſure was ſufficient and no more, to euerie one
young and old, and of middle age.
\MNote{8.}
Eightly the heate of the ſunne melted and conſumed that which remained
in the field, though otherwiſe it indured heate of the fire, ſeething in
water, grinding in milles, and beating in morters.
\MNote{9.}
Ninthly, it taſted to euerie one, what they deſired.
\MNote{10.}
Tenthly, it ſeemed neuertheleſſe to the euil minded, loathſome and light
meate, but pleaſant to the good.
\MNote{11.}
Eleuently, part of it was kept in the Arke by Gods commandment, and was
not corrupted in manie hundreth
\Fix{peares.}{yeares.}{obvious typo, fixed in other}
\MNote{12.}
Twelftly, this ſtrange and extraordinarie prouiſion, continued fourtie
yeares, that is, til the children of Iſrael came to the promiſed land
and then ceaſed.

You
\MNote{No miracle in Proteſtants Communion.}
ſee then ſo miraculous a figure farre excelled Zuinglius, or Caluins
communion bread, which containeth no miracle at al, only ſignifying
Chriſts bodie. But, as our Sauiours owne wordes importe, and ancient
fathers teach vs, by Manna was prefigured Chriſts verie bodie and bloud,
with his ſoule and Diuinitie vnder the forme of bread.

For
\MNote{Al the ſaid miracles are more eminent in the B.~Sacrament.}
this indeede infinitly excelleth Manna, containing al the foreſaid, or
rather much more eminent miracles. For
\MNote{1.}
(firſt) it was conſecrated by the maruelous power of Chriſts word, and
euer ſince the ſame is done, by the like power communicated to Prieſts,
\MNote{2.}
(2) in his Church militant,
\MNote{3.}
(3) one and the verie ſame, and not manie, in innumerable places, and in
euerie leſſe or greater forme, yea in the leaſt particle of the
accidents that may be, Chriſt is whole and entire.
\MNote{4.}
(4) It geueth abundance of grace in this life, ſignified by the day
before the Sabbath, for the glorie of the next life in eternal reſt,
\MNote{5.}
(5) where is no more vſe of Sacraments, but euerlaſting fruition of
glorie.
\MNote{6.}
(6) VVhoſoeuer therfore would make temporal commoditie of this heauenlie
foode, as it were reſeruing Manna for other dayes of the weke, it
periſheth to him, and turneth to his ignominie, but being reſerued in
the faithful ſoule, for the life to come, which is the true Sabbath, it
remaineth an heauenlie treaſure.
\MNote{7.}
(7) And ſo it auaileth to euerie one, as their ſoule, which is the
ſpiritual ſtomake of ſupernatural meate, is leſſe or more diſpoſed.
\MNote{8.}
(8) Though heate of perſecution, and other aduerſe power take away this
Sacrament and Sacrifice, abrode in the field of this world, yet no power
extinguiſheth it within the Church, where it is in due maner prepared
and miniſtred to the children of God,
\MNote{9.}
(9) where it yeldeth al comfort, ſtrength and contentment to good
ſpiritual deſires,
\MNote{10.}
(10) but to the incredulous Capharnaites ſemeth vnpoſſible, and to
carnal conceipts loathſome.
\MNote{11.}
(11) Being worthely receiued into our mortal bodies, our arke or
temporal tabernacle, it remaineth in incorruptible effect, wherby the
bodie ſhal be raiſed againe from death, and together with the ſoule be
eternally glorified.
\MNote{12.}
(12) In the meane time of this
\Fix{pilagrimage}{pilgrimage}{likely typo, fixed in other}
of mankind, it is our way-faring ſpecial prouiſion, dailie and
ſuperſubſtantial bread, til we ſhal poſſeſſe the promiſed land, the
kingdome of heauen in eternal bliſſe.}
\HH{Man-hu}! which ſignifieth: What is this! for they knew not what it
was. To whom Moyſes ſaid: This is
%%% 0228
the bread, which our Lord hath geuen you to eate. \V This is the word,
that our Lord hath commanded: Let euerie one gather of it ſo much as
ſufficeth to eate: a gomor euerie man, according to the number of your
ſoules that dwel in a tent ſo shal you take vp. \V And the children of
Iſrael did ſo: and they gathered, one more, an other leſſe. \V And they
meaſured by the meaſure of a gomor: neither he that gathered more, had
aboue: nor he that prouided leſſe, found vnder: but euerie one gathered,
according to that which they were able to eate. \V And Moyſes ſaid to
them: Let no man leaue therof til the morning. \V Who heard him not, but
certaine of them left vntil the morning, and it beganne to be ful of
wormes, and it putrified. And Moyſes was angrie againſt them. \V And
euerie one of them gathered in the morning, ſo much as might ſuffice to
eate: and after the ſunne waxed hotte, it melted. \V But in the ſixt day
they gathered duble portions, that is, two gomors euerie man: and al the
princes of the multitude
\SNote{By their woũdering at the duble quãtitie, it appeareth they
intended not to gather ſo much.}
came, and told Moyſes. \V Who ſaid to them: This is it which our Lord hath
ſpoken: The Sabbathes reſt is ſanctified vnto our Lord to
morow. Whatſoeuer is to be wrought, doe it: and the meates that are to
be made readie, make them readie: and whatſoeuer shal remaine, lay it vp
vntil the morning. \V And they did ſo as Moyſes had commanded, and it
putrified not, neither was there worme found in it. \V And Moyſes ſaid:
Eate it to day, becauſe it is the Sabbath of our Lord: to day it shal
not be found in the field. \V Gather it ſix dayes: but in the ſeuenth
day is the Sabbath of our Lord, therfore it shal not be found. \V And
the ſeuenth day came: and ſome of the people going forth to gather,
found not. \V And our Lord ſaid to Moyſes: How long wil you not keepe my
commandementes, and my law? \V See that our Lord hath geuen you a
Sabbath, and for this cauſe on the ſixt day he geueth you duble
portions: let each man tarie with himſelfe, and let none goe forth out
of his place the ſeuenth day. \V And the houſe of Iſrael called the name
therof Manna: which was as it were coriander ſeede white, and the taiſt
therof like to flowre with honie. \V And Moyſes ſayd: This is the word,
which our Lord hath commanded:
\SNote{By anticipation Moyſes writeth here the commãdment geuen when the
Tabernacle and Arck were finiſhed.
\XRef{Exo.~vlt.}}
Fil a gomor of it, and let it be kept vnto the generations to come
hereafter: that they may know the bread, wherwith I
%%% 0229
fed you in the wildernes, when you were brought forth out of the Land of
\Fix{Æhypt.}{Ægypt.}{obvious typo, fixed in other}
\V And Moyſes ſayd to Aaron: Take
\SNote{This Relique was put in a golden veſſel
\XRef{Heb.~9.}
though it was infinitly inferior to Chriſts fleſh
\XRef{Io.~6.}
yea inferior to the fleſh of anie glorified Sainct.}
one veſſel, and put Manna into it, ſo much as a gomor can hold: and lay
it vp before our Lord to keepe vnto your generations: \V as our Lord
commanded Moyſes. And Aaron put it in the tabernacle to be reſerued. \V
And the children of Iſrael did eate Manna fourtie yeares, til they came
into the habitable land: with this meate were they fed, vntil they
touched the borders of the land of Chanaan. \V And a gomor is the tenth
part of an ephi.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
