%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 29
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-29


\project douay-rheims


%%% 0256
%%% o-0232
\startChapter[
  title={Chapter 29}
  ]

\Summary{The maner of conſecrating Aaron and other Prieſts: vvith burnt
  offerings, 26.~and pacifiques, wherof Aaron and his ſonnes shal
  participate. 38.~The inſtitution of the daylie Sacrifice of tvvo
  lambes, one in the morning, the other at euen.}

But this alſo shalt thou doe,
\SNote{Special preparation before Biſhops and Prieſts be conſecrated.}
that they may be conſecrated to me in prieſthood. Take a calfe from the
heard, and two rammes without ſpotte, \V and vnleuened bread, and a cake
without leuen, tempered with oile, wafers alſo vnleauened anoynted with
oile: of wheaten flowre thou shalt make al. \V And being put in a basket
thou shalt offer them: and the calfe and the two rammes. \V And thou
shalt bring Aaron and his ſonnes to the doore of the tabernacle of
teſtimonie. And
\SNote{The firſt preparation in the partie to be conſecrated is
cleanſing from ſinne: then to be adorned with the vertues aboue
mentioned.
\XRef{Exod.~18. v.~4.}}
when thou haſt washed the father with his
%%% 0257
ſonnes in water, \V thou shalt veſt Aaron with his veſtments, that is,
with the linnen garment and the tunicke, and the Ephod and the
Rationale, which thou shalt gird with the baudrike. \V And thou shalt
put the 
mitre vpon his head, and the holie plate vpon the mitre, \V and thou
shalt powre the oile of vnction vpon his head: and by this rite shal he
%%% o-0233
be conſecrated. \V His ſonnes alſo thou shalt bring, and shalt inueſt
them with the linnen tunickes, and gird them with a bawdrike, \V to
witte, Aaron
and his children, and thou shalt put mitres vpon them: and they shal be
prieſtes to me by a perpetual religion. After that thou shalt haue
conſecrated their handes, \V thou shalt preſent alſo the calfe before
the tabernacle of teſtimonie. And Aaron and his ſonnes ſhal lay their
handes vpon his head, \V and thou shalt kil him in the ſight of the Lord,
beſide the doore of the tabernacle of teſtimonie. \V And that which thou
takeſt of the bloud of the calfe, thou shalt put vpon the hornes of the
Altar with thy finger, and the reſt of the bloud thou shalt powre at the
botome therof. \V Thou shalt take alſo the whole fatte that couereth the
entralles, and the caule of the liuer, and the two kidneys, and the
fatte that is vpon them, and shalt offer a burnt ſacrifice vpon the
Altar: \V but the flesh of the calfe and the hide and the dung, thou
shalt burne abrode without the campe, becauſe it is for ſinne. \V Thou
shalt take alſo one ramme, vpon the head wherof Aaron & his ſonnes shal
lay their handes. \V Which when thou haſt killed, thou shalt take of the
bloud therof, and powre round about the Altar. \V And the ramme it ſelfe
thou shalt cut into peeces, and his entralles and feete being waſhed,
thou shalt put vpon the flesh cut in peeces, and vpon his head. \V And
thou shalt offer the whole ramme for a burnt ſacrifice vpon the Altar:
it is an oblation to the Lord, a moſt ſweete ſauoure of the victime of
the Lord. \V Thou shalt take alſo the other ramme, vpon whoſe head Aaron
and his ſonnes shal lay their handes. \V Which when thou haſt immolated,
thou shalt take of his bloud, and put vpon the tippe of the right eare
of Aaron and of his ſonnes, and vpon the thumbes and great toes of their
right hand and foote, and thou shalt powre the bloud vpon the Altar
round about. \V And
%%% 0258
when thou haſt taken of the bloud that is vpon the Altar, and of the
oile of vnction, thou shalt ſprinkle Aaron and his veſture, his ſonnes &
their veſtmentes. And after they and their veſtmentes are
conſecrated, \V thou shalt take the fatte of the ramme, and the tayle &
the talow, that couereth the lunges, and the caule of the liuer, and the
two kidneies, and the fatte, that is vpon them, and the right shoulder,
becauſe it is the ramme of conſecration: \V and a peece of one loafe, a
cake tempered with oile, a wafer out of the basket of azymes, which is
ſette in the ſight of the Lord: \V and thou shalt put al vpon the handes
of Aaron and of his ſonnes, and shalt ſanctifie them eleuation before
the Lord. \V And thou shalt take al from their handes: and shalt burne
them vpon the Altar for an holocauſte, a moſt ſweete ſauour in the ſight
of the Lord, becauſe it is his oblation. \V Thou shalt take alſo the
breſt of the ramme, wherwith Aaron was conſecrated, and eleuating it
thou shalt ſanctifie it before the Lord, and it shal fal to thy part. \V
And thou shalt ſanctifie both the conſecrated breſt, and the shoulder
%%% o-0234
that thou didſt ſeparate of the ramme, \V wherwith Aaron was conſecrated
and his ſonnes, and they shal fal to Aarons part and his ſonnes by a
perpetual right from the children of Iſrael: becauſe they are the
primitiues and beginninges of their pacifique victimes which they offer
to the Lord. \V And the holie veſture, which Aaron shal vſe, his ſonnes
shal haue after him, that they may be anoynted, and their handes
conſecrated in it. \V He of his ſonnes that shal be appoynted high
prieſt in his ſteede, & that shal enter into the tabernacle of teſtimonie
to miniſter in the Sanctuarie, shal weare it ſeuen dayes. \V And thou
shalt take the ramme of the conſecration, and shalt boyle the flesh
therof in a holie place; \V which Aaron shal eate and his ſonnes. The
loaues alſo, that are in the basket, they shal eate in the entrie of the
tabernacle of teſtimonie, \V that it may be a placable ſacrifice, and
the handes of the offerers may be ſanctified. A ſtranger shal not eate
of them, becauſe they are holie. \V And if there remaine of the
conſecrated flesh, or of the bread til the morning, thou shalt burne the
remaynes with fire: they shal not be eaten, becauſe they are
ſanctified. \V Al, that I haue commanded thee, thou shalt doe vpon Aaron
and his ſonnes. Seuen dayes shalt thou conſecrate
%%% 0259
their handes: \V and thou shalt offer a calfe for ſinne euerie day for
expiation. And thou shalt cleanſe the Altar when thou haſt offered the
hoſte of expiation, and shalt anoynt it vnto ſanctification. \V Seuen
dayes shalt thou expiate the
\Fix{altar}{Altar}{possible typo same in other}
& ſanctifie it, and it shal be moſt Holie. Euerie one, that ſhal touch
it, shal be ſanctified.

\V This is it which thou ſhalt doe vpon the Altar: Two lambes of a yeare
old
\SNote{Diuers things were offered at diuers times, and al ſignified
Chriſts Sacrifice in his Church.
\Cite{S.~Aug. li.~1. c.~18. cont. aduerſ. leg. & prophet.}
yet none daylie but a lambe: more particularly
\Fix{ſignifiing}{ſignifying}{likely typo, fixed in other}
the \Emph{daylie} offering of the lambe of God and \Emph{perpetual}
effect therof.
\Cite{Origen. in Ioan.~1.}}
euerie day continually, \V one lambe in the morning, & an other at
euen, \V the tenth part of flowre tempered with oile beaten, which ſhal
haue in meaſure the fourth part of an hin, and wine for libation of the
ſame meaſure to one lambe. \V And the other lambe thou ſhalt offer at
euen, according to the rite of the morning oblation, and according to
that which we haue ſaid, for a ſauour of ſweetneſſe: \V it is a
ſacrifice to the Lord, by perpetual oblation vnto your generations, at
the doore of the tabernacle of teſtimonie before the Lord, where I wil
appoint to ſpeake vnto thee. \V And there wil I command the children of
Iſrael, and the Altar shal be ſanctified in my glorie. \V I wil
ſanctifie alſo the tabernacle of teſtimonie with the Altar, and Aaron
with his ſonnes, to doe the function of prieſthood vnto me. \V And I wil
dwel in the middes of the children of Iſrael, and wil be their God, \V
and they shal know that I am the Lord their God, that haue brought them
out of the Land of Ægypt, that I might abide among them, I the Lord
their God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
