%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Exodus
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 0201
%%% o-0184
\startChapter[
  title={Chapter 10}
  ]

\Summary{The eight plague, of locuſtes. 21.~The ninth darknes. Pharao
  yeldeth that al men and children should goe to the deſert, but not the
  cattle. 28.~At laſt commandeth Moyſes to come no more in his ſight,
  which Moyſes foretelleth shal ſo be.}

And our Lord ſaid to Moyſes: Goe in to Pharao: for
\SNote{By Gods patience ouer Pharao & his ſeruants, in not deſtroying
them, their wicked mind became more obſtinate.
\Cite{S.~Aug. q.~30. &~36. in Exod.}}
I haue indurate his hart, and the hart of his ſeruantes: that I may
worke theſe my ſignes in him, \V and thou maiſt tel in the eares of thy
ſonne, and of thy nephewes, how often I haue broken the Ægyptians, &
wrought my ſignes in them: and you may know that I am the Lord. \V
Moyſes therfore and Aaron went in to Pharao, and ſaid to him: Thus ſaith
the Lord the God of the Hebrewes: Til when wilt thou not be
%%% 0202
ſubiect to me? diſmiſſe my people, to ſacrifice vnto me. \V But if thou
reſiſt, and wilt not diſmiſſe them: behold I wil bring in to morow
\MNote{The 8.~plague. Innumerable locuſtes, litle flying beaſtes with
long hinder legges that deſtroy graine, graſſe & fruict.
\Cite{Plinius. li.~11. c.~29.}
\Cite{S.~Greg. li.~31. c.~20. Moral.}}
the locuſt into thy coaſtes: \V which may couer the face of the earth,
that nothing therof appeare, but that which the haile hath left may be
eaten: for it ſhal gnawe al trees that ſpring in the fieldes. \V And
they ſhal fil thy houſes, and the houſes of thy ſeruantes, and of al the
Ægyptians: ſuch a number as thy fathers haue not ſeene, nor
grand-fathers, ſince they aroſe vpon the earth, vntil this preſent
day. And he turned him ſelfe away, and went forth from Pharao. \V And
Pharaoes ſeruantes ſaid to him: How long ſhal we endure this ſcandal?
Diſmiſſe the men, to ſacrifice to the Lord their God. Doeſt thou not ſee
that Ægypt is vndone? \V And they called back Moyſes, and Aaron vnto
Pharao: who ſaid to them: Goe, ſacrifice to the Lord your God: who are
they that shal goe? \V Moyſes ſaid: With our young and old we wil goe,
with our ſonnes and daughters, with our sheepe and heardes: for it is
the ſomemnitie of the Lord our God. \V And Pharao anſwered: So be the
Lord with you, as I shal diſmiſſe you, and your litle ones: who doubteth
but that
\SNote{Becauſe Gods ſeruants may not temporize in religion, politiques
vniuſtly charge them to haue bad intentions.}
you intend very wickedly? \V It shal not ſo be: but goe ye men only, and
ſacrifice to the Lord: for this your ſelues alſo deſired. And immediatly
they were caſt out from Pharaoes ſight. \V And our Lord ſaid to Moyſes:
Strech forth thy hand vpon the Land of Ægypt vnto the locuſt, that it
come vpon it, and deuoure euerie herbe that remained after the haile. \V
And Moyſes ſtretched forth his rodde vpon the Land of Ægypt: and our
Lord brought in a burning wind al that day, & night: and when it was
morning, the burning winde raiſed the locuſtes:
%%% o-0185
\V which came vp ouer
the whole Land of Ægypt: and ſate in al the coaſtes of the Ægyptians
innumerable, the like as had not bene before that time, nor shal
be afterward. \V And they couered the whole face of the earth,
waſting al thinges. Therfore the graſſe of the earth was deuoured, and
what fruites ſoeuer on the trees, which the haile had left: there was
alſo nothing at al left that was greene in the trees, and in the herbes
of the earth, in al Ægypt. \V For the which cauſe Pharao in haſt called
Moyſes and Aaron, and ſaid to them: I haue ſinned againſt the Lord your
God, and againſt you. \V But now forgeue me my ſinne this time alſo, and
pray to the Lord your God, that he take away from me this death.
%%% 0203
\V And Moyſes going forth from Pharaoes ſight, prayed to our Lord: \V
who made a very vehement wind to blow from the weſt, and taking the
locuſtes it threw them into the Red ſea: there remained not ſo much as
one in al the coaſtes of Ægypt. \V And our Lord did indurate Pharaoes
hart, neither did he diſmiſſe the children of Iſrael. \V And our Lord
ſaid to Moyſes: Stretch
\Fix{for}{forth}{possible typo, same in other}
thy hand toward heauen: and be there
\MNote{The 9.~plague. Horrible darknes three dayes together.}
darkeneſſe vpon the Land of Ægypt ſo thicke, that it be palpable. \V And
Moyſes ſtretched forth his hand toward heauen: and there was made
horrible darkeneſſe in the whole Land of Ægypt three dayes. \V No man
ſaw his brother, nor moued himſelfe out of the place where he was: but
whereſoeuer the children of Iſrael dwelt, there was light. \V And Pharao
called Moyſes and Aaron, and ſaid to them: Goe ſacrifice to the Lord:
let your ſheepe only, and heardes remaine, let your litle ones goe with
you. \V Moyſes ſaid: Hoſtes alſo & holocauſtes thou shalt geue to vs, which
we may offer to the Lord our God. \V Al the flockes ſhal goe with vs:
there ſhal
\SNote{Gods people muſt be reſolute in Religion.}
not a hoofe remaine of them: the which are neceſſarie vnto the ſeruice
of the Lord our God: eſpecially wheras we know not what muſt be offered,
til we come to the very place. \V And our Lord did indurate Pharaoes
hart, and he would not diſmiſſe them. \V And 
\Fix{Pharro}{Pharao}{obvious typo, fixed in other}
ſaid to Moyſes: Gette thee from me, and beware thou ſee not my face any
more: in what day ſoeuer thou ſhalt come in my ſight, thou ſhalt dye. \V
Moyſes anſwered: So ſhal it be as thou haſt ſpoken, I wil not ſee thy
face any more. 

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
