%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Two Corinthians
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 2735
%%% o-2592
\startChapter[
  title={Chapter 13}
  ]

\Summary{He driueth into them the feare of excommunication: to the end
  that they doing penance before-hand, he may not be compelled to vſe
  his authoritie when he commeth, and as he hath threatned. 11.~And ſo
  with a general exhortation he endeth.}

%%% o-2593
Loe this is the third time I come vnto you:
\CNote{\XRef{Deut.~19,~15.}}
In the mouth of two or three witneſſes ſhal euery word ſtãd. \V I
foretold and doe fore-tel as preſent, and now abſent, to them that
ſinned before, and al the reſt, that if I come againe, I wil not
ſpare. \V Seeke you an experiment of him that ſpeaketh in me, Chriſt;
who in you is not weake, but is Mightie in you? \V For although he was
crucified of infirmitie; yet he liueth by the power of God. For we alſo
are weake in him: but we ſhal liue with him by the power of God on
you. \V
\LNote{Trie your ſelues.}{The
\MNote{We may know that we haue faith, but not that we are in grace.}
Heretikes argue hereupon, that euery man may know himſelf certainely to
be in grace: where the Apoſtle ſpeaketh expreſly and only of faith: the
act whereof a man may know and feele to be in himſelf, becauſe it is an
act of vnderſtanding, though he can not be aſſured that he hath his
ſinnes remitted, and that he is in al points in ſtate of grace and
ſaluation: becauſe euery man that is of the Catholike faith, is not
alwaies of good life agreable thereunto, nor the acts of our wil ſo
ſubiect to vnderſtanding, that we can know certainely whether we be good
or euil. See S.~Auguſtin
\Cite{to.~7. perfect. iuſtitia c.~15.}
\Cite{Li. de Cor. et grat. c.~13.}
&
\Cite{S.~Thomas 1.~2. q.~12,~2. art.~5.}}
Trie your owne ſelues if you be in the faith, proue ye your ſelues. Know
you not your ſelues that Chriſt \Sc{Iesvs} is in you, vnleſſe perhaps
you be reprobates. \V But I hope you
\Var{know}{shal know}
that we are not reprobates. \V And we pray God, that you doe no euil,
not that we may appeare approued, but that you may doe that which is
good, and we be as reprobates. \V For we can not any thing againſt the
truth; but for the truth. \V For we reioyce, for that we are weake, and
you are mightie. This alſo we pray for, your conſummation. \V Therfore
theſe things I write abſent: That being preſent I may not deale hardly
according to the
%%% 2736
\SNote{Eccleſiaſtical power to punish offenders by the cenſures of the
Church.}
power which our Lord hath giuen me vnto edification and not vnto
deſtruction.

\V For the reſt, Brethren, reioyce, be perfect,
\TNote{\G{παρακαλεῖσθε}}
take exhortation, be of one mind, haue peace; and the God of peace & of
loue ſhal be with you. \V Salute one another in a
\CNote{\XRef{No.~16,~16.}
\XRef{1.~Cor.~16,~20.}}
holy kiſſe. Al the Saints ſalute you. \V The grace of our
Lord \Sc{Iesvs} Chriſt, and the charitie of God, and the communication
of the Holy Ghoſt be with you al. Amen.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

