%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Two Corinthians
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 2727
%%% o-2584
\startChapter[
  title={Chapter 8}
  ]

\Summary{By the example of the poore Macedonians he exhorteth them to
  contribute largely vnto the Church of Hieruſalem, 7.~& by praiſing of
  them, 9.~and by the example of Chriſt, 14.~and by their owne ſpiritual
  profit in being partakers of that Churches merits, 16.~and by
  commending the Collectours that he ſendeth.}

And we doe you to vnderſtand, Brethren, the grace of God, that is giuen
in the Churches of Macedonia, \V that in much experience of tribulation
they had aboundance of ioy, & their very deep
\Fix{prouertie}{pouertie}{obvious typo, fixed in other}
abounded vnto the riches of their ſimplicitie, \V for according to their
power (I giue them teſtimonie) and aboue their power 
they were willing, \V with much exhortation requeſting vs the grace and
communication of the miniſterie that is done toward the Saints. \V And
not as we hoped, but their owne ſelues they gaue, firſt to our Lord,
\SNote{The principal reſpect next after God is to be had of our Maiſters
in religion, in al temporal and ſpiritual dueties.}
then to vs by the wil of God: \V in ſo much that we deſired Titus, that
as he began, ſo alſo he would perfit in you this grace alſo.
%%% 2728
\V But as in al things you abound in faith, and word, & knowledge, & al
carefulnes, moreouer alſo in your charitie toward vs, that in this grace
alſo you may abound, \V I ſpeake not as cõmãding:
but by the carefulnes of others, approuing alſo the good diſpoſitiõ of
your charitie. \V For you know the grace of our Lord \Sc{Iesvs} Chriſt,
that for you he was made poore, wheras he was rich; that by
\Fix{is}{his}{obvious typo, fixed in other}
pouertie you might be rich. \V And in this point I giue counſel: for
this is profitable for you, which haue begun not only to doe, but alſo
to be willing, from the yeare paſt: \V but now performe ye it alſo in
deed: that as your mind is prompt to be willing, ſo it may be alſo to
performe, of that which you haue. \V For if the wil be prompt, it is
accepted according to that which it hath, not according to that which it
hath not. \V For not that other ſhould haue eaſe, and you tribulation:
but by an equalitie. \V Let in this preſent time your
\LNote{Aboundance ſupply.}{He
\MNote{Temporal benefits vpon ſpiritual perſons.}
meaneth that ſuch as abound in worldly riches, ſhould communicate for
ſupply of other their Brethrens neceſſities, whatſoeuer they may: that
on the other ſide they whom they help in temporals, may impart to them
againe ſome of their ſpiritual riches, as praiers, and other holy workes
and graces, which is a happie change and entercourſe for the wealthy men,
if they could ſee it.
\MNote{One may ſatisfie and ſupererogate for another.}
And this place proueth plainely that the faſtings and ſatisfactorie deeds
of one man, be auailable to others, yea and that holy Saints or other
vertuous perſons may in meaſure and proportion of other mens neceſſities
and deſeruings, allot vnto them, as wel the ſupererogation of their
ſpiritual workes, as theſe that abound in worldly goods, may giue almes
of their ſuperfluities, to them which are in neceſſitie. Which
interchange and proportion of things the Apoſtle doth euidently ſet
downe.}
aboundance ſupplie their want: that their aboundance alſo may ſupplie
your want, that there be an equalitie \V as it is written:
\CNote{\XRef{Exo.~16,~28.}}
\Emph{He that had much abounded not: and he that had litle, wanted not.}

\V And thankes be to God, that hath giuen the ſelf-ſame carefulnes for
you in the hart of Titus, \V for that he admitted indeed exhortation:
but being more careful, of his owne
%%% o-2585
wil he went vnto you. \V We haue ſent alſo with him the Brother, whoſe
praiſe is in the Ghoſpel through al Churches: \V & not only that, but
alſo he was ordeined of the churches fellow of our peregrination, for
this grace which is miniſtred of vs to the glorie of our Lord, and our
determined wil: \V auoiding this, leſt any man might reprehend vs in
this fulnes that is miniſtred of vs. \V For we prouide good things
\CNote{\XRef{Ro.~12,~27.}}
not only before God, but alſo before men. \V And we haue ſent with them
our Brother alſo, whom we haue proued in many things often to be
careful: but now much more careful, for the great confidence in you, \V
either for Titus which is my fellow and Coadiutor toward you, or our
Brethren Apoſtles of the Churches, the glorie of Chriſt. \V The
declaration therfore which is of your charitie and our glorying for you,
declare ye toward them in the face of the churches.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

