%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Two Corinthians
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 2732
%%% o-2588
\startChapter[
  title={Chapter 11}
  ]

\Summary{He reaſoneth the matter with the Corinthians, why they should
  preferre the falſe Apoſtles before him. And becauſe they giue them leaue
  to bragge and commend themſelues and to abuſe them ſo miſerably, he
  truſteth they wil alſo giue him the hearing: 21.~and ſo he beginneth,
  and firſt shewing himſelf in al Iudaical reſpectes (wherin only ſtood
  al their boaſting) to be as they are, he addeth afterward ſuch a long
  roll of his ſuffering for Chriſt, as is incomparable.}

%%% o-2589
Would God you could beare ſome litle of my folly: but doe ye alſo
ſupport me: \V for I emulate you with the emulation of God. For I haue
\SNote{The Apoſtles and their Succeſſors did deſpouſe the people whom
they conuerted, to Chriſt, ĩ al puritie & chaſtitie of truth and wholy
vndefiled and void of errour and hereſie.}
deſpouſed you to one man, to preſent you a chaſt virgin vnto Chriſt. \V
But I feare leſt, as
\CNote{\XRef{Gen.~3,~4.}}
the ſerpent ſeduced Eue by his ſubteltie, ſo your
ſenſes may be corrupted, & fal
\LNote{From the ſimplicitie.}{People
\MNote{As Eue by the Serpent, ſo the people are ſeduced by Heretikes.}
fal from their firſt faith, virginitie, and ſimplicitie in Chriſt, not
by ſodain reuolt, but by litle & litle, in giuing eare to the ſubtil
perſuaſion of the Serpent, ſpeaking to them by the ſweet mouths &
allurements of Heretikes. Of which kind of ſeduction he giueth Eue for
an example, who was by her greedy deſire of knowledge and the Diuels
promiſe of the ſame, drawen from the natiue ſimplicitie and obedience to
God. As at this day, promiſe and pretenſe of knowledge driueth many a
poore ſoul from the ſure, true, ſincere, and only beleefe of God's
Church.}
from the ſimplicitie that is in Chriſt. \V For if he that
\SNote{The note of a falſe Teacher, \Emph{to come}: that is without
lawful calling or ſending to thruſt and intrude himſelf into another
mans charge.}
commeth, preach another Chriſt whom we haue not preached, or you receiue
another ſpirit whom you haue not receiued; or another Ghoſpel which you
haue not receiued, you might wel ſuffer it. \V For I ſuppoſe that I haue
done nothing leſſe then the great Apoſtles. \V For although
\LNote{Rude in ſpeach.}{Hereby
\MNote{Heretikes ſometime eloquent.}
we ſee that the ſeditious and falſe Teachers haue often the guift of
eloquence wherby the ſimple be eaſily beguiled. Such were Core and
Dathan, as Ioſephus writeth
\Cite{Ant. li.~4. c.~2.}
for the ſame, S.~Auguſtin
\Cite{(li.~5. Confeſ. c.~3. &~13.)}
calleth the Heretike Fauſtus Manichæus, \L{magnum laqueum Diaboli},
\Emph{a great ſnare of the Diuel}, ſaying that he paſſed the glorious
Doctor S.~Ambroſe in ſhew of words, but farre inferiour to him (without
al compariſon) in ſubſtance and matter.
\MNote{Knowledge better thẽ gay words.}
In which ſort the Apoſtle here is glad to compare himſelf with the falſe
Apoſtles, whom the Corinthians did follow and extol farre aboue him by
reaſon of their eloquence; granting to them that guift, but chalenging
to himſelf ſuperioritie in knowledge, which al wiſe men preferre before
vaine words.
\MNote{Yong Oratours among Heretikes preferred before the ancient
Doctours.}
And it is the bane of our poore countrie, that the people now a-daies
giue credit rather to new Oratours and fooliſh yonkers, for their ſweet
ſpeaches; then to the glorious Doctours of Chriſtes Church, for their
ſingular knowledge and more graue eloquence.}
rude in ſpeach, yet not in knowledge. But in al things we are made
manifeſt to you. \V Or did I commit a ſinne, humbling my ſelf, that you
might be exalted? becauſe I euangelized vnto you the Ghoſpel of God
gratis? \V Other Churches I ſpoiled, taking a ſtipend, for your
miniſterie. \V And when I was with you, and had need, I was burdenous to
none: for that which I wanted, the Brethren ſupplied that came from
Macedonia: and in al things I haue kept my ſelf without burden to you,
and wil keep. \V The truth of Chriſt is in me, that this glorying ſhal
not be infringed toward me in the countries of Achaia. \V Wherfore?
becauſe I loue you not? God doth know. \V But that which I doe, I wil
alſo doe, that I may cut away the occaſion of them that deſire occaſion:
that, in that which they glorie, they may be found euẽ like vs. \V For
ſuch falſe Apoſtles are
\SNote{A proper terme for Heretikes that ſhape themſelues into the habit
of true Teachers,
\Fix{ſpacially}{ſpecially}{obvious typo, fixed in other}
by often allegation and commendatiõ of the Scriptures. Read the notable
admonition of the ancient writer Vincentius Lirinenſis in his golden
booke
\Cite{Againſt the Prophane nouelties of al hereſies.}}
craftie workers, trãsfiguring themſelues into Apoſtles of Chriſt. \V And
no maruel: for Satã himſelf transfigureth himſelf into an Angel of
light. \V It is no great matter therfore if his Miniſters be
transfigured as the Miniſters of iuſtice: whoſe end ſhal be according to
their workes.

\V Againe I ſay, (let no man thinke me to be fooliſh: otherwiſe take me
as fooliſh, that I alſo may glorie a litle,) \V that which I ſpeake, I
ſpeake not according to God, but as it were in fooliſhnes, in this
ſubſtance of glorying. \V Becauſe many glorie according to the fleſh, I
alſo wil glorie. \V For you doe gladly ſuffer the fooliſh: whereas your
ſelues are wiſe. \V For you ſuffer if a mã bring you into ſeruitude, if
a man deuoure, if a mã take, if a mã be extolled, if a mã ſtrike you on
the face. \V I ſpeake according
%%% o-2590
to diſhonour, as though we had been weake in this part. Wherin any man
dare (I ſpeake
\Fix{foliſhly)}{fooliſhly)}{likely typo, same in both}
I dare alſo. \V
\CNote{\XRef{Phil.~3,~5.}}
They
%%% 2733
are Hebrewes: and I. They are Iſraelites: and I. They are the ſeed of
Abraham: and I. \V They are the Miniſters of Chriſt: and I. (I ſpeake as
one ſcarſe wiſe) more I: in many moe labours, in priſons more
aboundantly, in ſtripes aboue meaſure, in deaths often. \V Of the Iewes
fiue times did I receiue
\CNote{\XRef{Deu.~25,~3.}}
fortie ſauing one. \V Thriſe was I beaten
\CNote{\XRef{Act.~16,~23.}}
with rods,
\CNote{\XRef{14,~18.}}
once I was ſtoned, thriſe I ſuffred
\CNote{\XRef{27,~15.}}
ſhipwrack; night and day haue I been in the depth of the ſea, \V in
iourneying often, perils of waters, perils of theeues, perils of my
Nation, perils of Gentils, perils in the citie, perils in the wildernes,
perils in the ſea, perils among falſe Brethren, \V in labour and
miſerie, in much watchings, in hunger and thirſt, in faſting often, in
cold and nakednes, \V beſide thoſe things which are outwardly: my daily
%%% !!! All one SNote in original
\SNote{S.~Chryſoſtom and Theophylact interpret it of daily conſpiracie
againſt him: others, of multitude of cares inſtant & vrgent vpon him.}
\TNote{\G{ἐπίστασίς}}
inſtance, the carefulnes of al Churches. \V Who is weake, and I am not
weake? Who is ſcandalized, and I
\TNote{\L{non vror}? \G{πυροῦμαι}?}
am not burnt? \V If I muſt glorie: I wil glorie of the things that
concerne my infirmitie. \V The God and Father of our Lord \Sc{Iesvs}
Chriſt, who is bleſſed for euer, knoweth that I lie not. \V
\CNote{\XRef{Act.~9,~14.}}
At Damaſcus the Gouernour of the Nation vnder Aretas the King, kept the
citie of the Damaſcenes for to apprehend me: \V and through a window in
a basket was I let downe by the wal, and ſo eſcaped his hands.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

