%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Two Corinthians
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 2721
%%% o-2577
\startChapter[
  title={Chapter 4}
  ]

\Summary{That according as ſo glorious a miniſterie requireth, he
  liueth and preacheth ſincerely, 7.~the which glorie his Aduerſaries
  can not count vaine, conſidering his perſecutions: becauſe perſecution
  is to God's glorie, and to our humilitie and hope, and meritorious of
  increaſe of grace in this life, and of moſt glorious bodies and ſoules
  afterward.}

Therfore hauing this miniſtration; according as we haue obteined mercie,
we faile not, \V but we renounce the ſecret things of diſhoneſtie, not
walking in craftines, nor
\LNote{Adulterating.}{He
\CNote{See
\Cite{Ire. li.~1. c.~1.}}
\MNote{Heretikes corrupters of God's word; Catholike Doctours, right
handlers therof.}
giueth often warning of falſe Teachers, whoſe ſpecial and proper ſtudie
is to falſifie and adulterate by deceitful conſtructions,
interpretations, and applications, the word of God: hauing no other end
but to make their aduantage of the Scriptures, and to
\Fix{againe}{gaine}{obvious typo, fixed in other}
glorie and eſtimation among the ſinful and ſimple, by new deuiſed
expoſitions. Wherin the Proteſtants doe excel the ancient Heretikes,
none euer more impurely handling the word of God then they doe. Origen
\CNote{\Cite{in 2.~ad.~Rom.}}
calleth ſuch \L{Scripturarum fures & adulteros}, theeues and adulterers
of the Scriptures. S.~Cyprian
\Cite{(de vnit. Ec. nu.~7.)}
calleth them, corrupters of the Ghoſpel, falſe interpreters, artificers
and crafts-maſters in corrupting the truth. On the other ſide, for
ſpecial reuerence and ſinceritie of dealing
\Fix{it}{in}{obvious typo, fixed in other}
thoſe matters, the Fathers and al Catholike Preachers or Expoſitours
were of old called according to
\CNote{\XRef{2.~Tim.~2.}}
S.~Paules words to Timothee, \L{recte tractantes verbum Dei}, right
handlers of the word of God.}
adulterating the word of God, but in manifeſtation of the truth
commending our-ſelues to euery conſcience of men
%%% o-2578
before God. \V And if our Ghoſpel be alſo hid, in them that periſh it is
hid, \V in whom the God of this world hath blinded the minds of the
infidels, that the illumination of the Ghoſpel of the glorie of Chriſt,
who is the Image of God, might not ſhine to them. \V For we preach not
our-ſelues, but \Sc{Iesvs} Chriſt our Lord: and vs, your ſeruants by
\Sc{Iesvs}: \V becauſe God that commanded light to ſhine of darkenes, he
hath shined in our harts to the illumination of the knowledge of the
glorie of God, in the face of Chriſt \Sc{Iesvs}. \V But we haue this
treaſure in earthen veſſels, that the excellencie may be of the power of
God, and not of vs. \V In al things we ſuffer tribulation, but are not
in diſtreſſe:
\TNote{\L{aporiamur} See
\Cite{S.~Amb. Theoph.}}
we want, but are not deſtitute: \V we ſuffer perſecution, but are not
forſakẽ: we are caſt downe, but we perish not: \V alwaies bearing about
in our body the mortification of \Sc{Iesvs}, that the life alſo of
\Sc{Iesvs} may be manifeſted in our bodies. \V For we that liue are
alwaies deliuered vnto death for \Sc{Iesvs}: that the life alſo of
\Sc{Iesvs} may be manifeſted in our mortal fleſh. \V Death thẽ worketh
in vs, but life in you. \V And hauing the
%%% 2722
ſame ſpirit of faith, as it is written:
\CNote{\XRef{Pſ.~115,~10.}}
\Emph{I beleeued, for the which cauſe I haue ſpoken}, we alſo beleeue,
for the which cauſe we ſpeake alſo: \V knowing that he which raiſed vp
\Sc{Iesvs}, wil raiſe vp vs alſo with \Sc{Iesvs} and ſet vs with you. \V
For al things are for you: that the grace abounding by many in giuing
of thanks, may abound vnto the glorie of God. \V For which cauſe we faile
not: but although that our man which is without,
\TNote{\L{corrumpatur}}
corrupt: yet that which is within, is renewed from day to day. \V For
that our tribulation which preſently is momentarie & light,
\LNote{Worketh.}{The
\MNote{Tribulation meritorious of glorie.}
temporal and short tribulations which we patiently and willingly ſuffer
for Chriſt, doe winne vs euerlaſting ioy and glorie. And it is here to
be noted againſt the Heretikes, that tribulations doe worke or cauſe the
ſaid ſaluation, which they deny to be giuen for ſuch things, but for or
by faith only. S.~Auguſtin
\CNote{Aug. in Pſ.~93. prope finem.}
maketh ſuch tribulations for Chriſt ſo much
the meritorious cauſe of euerlaſting life and reſt, that he ſaith it is
ſalable and bought thereby. And it is written
\XRef{Sap.~10.}
\Emph{God rendreth or repaieth to iuſt men the hire of their labours.}}
\SNote{The English Bible~1577 doth falſely tranſlate, \Emph{prepareth}.}
\TNote{\G{κατεργάζεται}}
worketh aboue meaſure exceedingly an eternal weight of glorie in vs, \V
we not conſidering the things that are ſeen, but that are not ſeen. For
the things that be ſeen, are temporal: but thoſe that be not ſeen, are
eternal.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

