%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Two Corinthians
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 2730
%%% o-2587
\startChapter[
  title={Chapter 10}
  ]

\Summary{Againſt the falſe Apoſtles granting the infirmitie of his
  perſon, he doth notwithſtanding ſet out the power of his Apoſtleship,
  12.~reprehending them alſo for chalenging to themſelues the praiſe of
  other mens labours.}

And I Paul my ſelf beſeech you by the mildenes and modeſtie of Chriſt,
who in preſence indeed am humble among you, but abſent am bold on
you. \V But I beſeech you, that being preſent I need not be bold by that
confidence wherwith I am thought to be bold againſt ſome: which thinke
vs as though we walke according to the fleſh. \V For walking in the
fleſh, we warre not according to the fleſh. \V For the
\LNote{Weapons.}{He
\MNote{Puniſhing of Heretikes.}
meaneth the ample ſpiritual and Apoſtolical power giuen by Chriſt for
the puniſhment of falſe Apoſtles, Heretikes, and rebelles to God's
Church, who are here noted ſpecially by pride, and inſolence (which is
the proper marke of ſuch fellowes)
\MNote{Their pride.}
to extol themſelues aboue the meaſure
of the ſcience of God, which conſiſteth in humble obedience to the faith
and the Preachers of the ſame.}
weapons of our warfare are not carnal: but mightie to God vnto the
deſtruction of munitions, deſtroying counſels, \V and al loftineſſe
extolling itſelf againſt the knowledge of God, and bringing into
captiuitie al vnderſtanding vnto the obedience of Chriſt, \V and hauing
in a readineſſe
\LNote{To reuenge.}{You
\MNote{The ſpiritual power of Biſhops againſt Heretikes.}
may ſee hereby, that the ſpiritual power of Biſhops is not only in
preaching the Ghoſpel, and ſo by perſuaſion and exhortation only (as
ſome Heretikes hold) to remit or retaine ſinnes, but that it hath
authoritie to puniſh, iudge, and condemne Heretikes and other like
rebelles: which power
\CNote{\Cite{Caluin vpon this place.}}
one of the principal rebelles of this time being conuinced by the
euidence of the place, acknowledgeth to be grounded vpon Chriſtes word,
\Emph{Whatſoeuer you bind in earth, shal be bound in Heauen}:
\XRef{Mat.~18,~18.}
applying alſo the words ſpoken to Hieremie
\XRef{c.~1,~10.}
\Emph{Behold I appoint thee ouer Nations and Kingdoms, that thou plant,
plucke vp, build, and deſtroy}, to confirme & explicate the power
Apoſtolike here alleaged by S.~Paul.
\MNote{Heretical Conſiſtories.}
Mary they would gladly draw this power from the lawful Succeſſours of
the Apoſtles, to themſelues, their Miniſters, and Conſiſtories, which
are nothing els but the ſhops and Councels of ſedition and al the
conſpiracies of this time, againſt the lawful Princes of the world.}
to reuenge al diſobedience, when your obediẽce ſhal be fulfilled. \V See
the things that are according to appearance. If any man haue affiance in
himſelf, that he is Chriſt's: let him thinke this againe with himſelf,
that as he is Chriſt's, ſo we alſo. \V For and if I ſhould glorie
ſome-what more of our power, which our Lord hath giuen vs
\LNote{Vnto edification.}{This
\MNote{Eccleſiaſtical cenſures (namely Excommunication) when & where to
be executed.}
great power of the Churches cenſures, ſpecially of excommunication, as
it was giuen for the good and ſaluation of the people, ſo it muſt not be
vſed againſt the innocent, no nor yet vpon Heretikes or other
offenders, but where & when it may by likelyhood benefit either the
parties, or the people, or may be executed without the hurt of
pertubation of the whole Church, as oftentimes it can not be, by reaſon
of the multitude of offenders. Which cauſed the Apoſtle here to ſignifie
that he would not vſe his vttermoſt authoritie againſt the falſe
Apoſtles which diſturbed him; til themſelues were in perfect obedience
vnto him, leſt by puniſhing the principal offenders, a greater
diſturbance & reuolt might fal among the people, if they were not before
in perfect obedience.}
vnto edification and not to your deſtruction; I ſhal not be aſhamed. \V But
that I may not be thought as it were to terrifie you by epiſtles (\V for
his epiſtles indeed, ſay they, are ſore and vehement; but his bodily
preſence weake & his ſpeach contemptible) \V let him this thinke that
is ſuch a one, that ſuch as we are in word by epiſtles, abſent; ſuch
alſo we are indeed, preſent. \V For we dare not match or compare our
ſelues with 
certaine, that cõmend thẽſelues: but we meaſure our ſelues in our
ſelues, & cõpare our ſelues to our ſelues. \V But we wil not glorie
aboue our meaſure: but according to the meaſure of the rule, which God
hath meaſured to vs, a meaſure to reach euẽ vnto you.
%%% o-2588
\V For not, as though we reached
%%% 2731
not vnto you, doe we extend ourſelues beyond. For we are come as farre
as to you in the Ghoſpel of Chriſt. \V Not glorying aboue meaſure in
other mens labours: but hauing
\TNote{\G{ἐλπίδα αὐξανομένης τῆς πίστεως ὑμῶν}}
hope of your faith increaſing, to be magnified in you according to our
Rule aboundãtly, \V yea vnto thoſe places that are beyond you, to
euãgelize, not in another man's Rule, to glorie in thoſe things that are
prepared before. \V But he that glorieth, let him glorie in our Lord. \V
\CNote{\XRef{Ier.~9,~23.}}
For not he that commendeth himſelf, the ſame is approued; but whom God
commendeth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

