%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Two Corinthians
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 2729
%%% o-2585
\startChapter[
  title={Chapter 9}
  ]

\Summary{He proceedeth exhorting them to the foreſaid contribution,
  3.~to verifie his commending of them. 6.~And to doe it liberally, that
  ſo they may merit the more, and God be the more praiſed.}

For concerning the miniſterie that is done
\LNote{Toward the Saints.}{By
\MNote{Proctours for Catholike Priſoners.}
the Apoſtles earneſt and often calling vpon the Corinthians to giue
almes for relieuing the faithful in diſtreſſe, the Paſtours of Gods
Church may learne that it ſpecially pertaineth to their office to be
Proctours for holy men in priſon, pouertie, & al other neceſſitie,
ſpecially when their want commeth for confeſſion of their faith.}
toward the Saints, it is ſuperfluous for me to write vnto you. \V For I
know your prompt mind: for the which I glorie of you to the Macedonians:
That Achaia alſo is ready from the yeare paſt, and your emulation hath
prouoked very many. \V But I haue ſent the Brethren, that the thing
which we glorie of you, be not made void in this behalfe, that (as I
%%% o-2586
haue ſaid) you may be ready: \V leſt when the Macedonians ſhal come with
me, and find you vnready, we (that we ſay not, ye) may be aſhamed
\SNote{That is, in this matter of almes.
\Cite{Chryſoſ. Theoph.}}
in this ſubſtance. \V Therfore I thought it neceſſarie to deſire the
Brethren that they would come to you, and prepare this bleſſing before
promiſed, to be ready ſo, as a bleſſing,
\LNote{Not as auarice.}{The
\MNote{Cheerful giuing.}
couetous man that parteth with
\Fix{is}{his}{obvious typo, fixed in other}
peny painefully and with ſorrow, as though he loſt a limme of his body,
is noted; and cheerful, ready, voluntarie, and large contribution is
commended.}
not as auarice. \V And this I ſay, he that
\LNote{Soweth ſparingly.}{Almes
\MNote{The greater almes, the greater merit and reward.}
is compared to ſeed. For as the ſeed throwen into the ground, though it
ſeeme to be caſt away, yet is not loſt, but is laid vp in certaine hope
of great encreaſe: ſo that which men giue in almes, though it ſeeme to be
caſt away & to periſh in reſpect of the giuer, yet indeed it is moſt
fruitful; the benefit therof manifoldly returning to him
againe. Whereupon the Apoſtles concluſion is cleere, that according to
the meaſure of the almes or ſeeding (which is more or leſſe in reſpect
of the wil & abilitie of the giuer) the encreaſe & aboundance of
harueſt, that is, of grace and glorie ſhal enſue. See S.~Auguſtin
\Cite{in Pſal.~49. circa med.}
&
\Cite{q.~4. ad Dulcitium}}
ſoweth ſparingly, ſparingly alſo ſhal reap: and he that ſoweth in
bleſſings, of bleſſings alſo ſhal reap. \V Euery one as he hath
determined in his hart, not of ſadnes or of neceſſitie. \V For
\CNote{\XRef{Eccleſ.~35,~11.}}
\Emph{God loueth a cheerful giuer.} \V And God is able to make al grace
abound in you: that in al things alwaies hauing al ſufficiencie, you may
abound vnto al good workes, \V as it is written:
\CNote{\XRef{Pſ.~11,~9.}}
\Emph{He diſtributed, he gaue to the poore:
\SNote{The fruit of almes is the encreaſe of grace in al iuſtice and
good workes to life euerlaſtĩg: God giuing theſe things for reward &
recompenſe of charitable workes, which therfore be called the ſeed or
meritorious cauſe of theſe ſpiritual fruits.}
his iuſtice remaineth for euer.} \V And he that miniſtreth ſeed to the
ſower, wil giue bread alſo for to eate: and wil multiply your ſeed and
wil augment the increaſes of the fruits of your iuſtice: \V that being
enriched in al things, you may abound vnto al ſimplicitie, which worketh
by vs thankes-giuing to God. \V Becauſe the miniſterie of this office
\LNote{Doth not only ſupply.}{When
\MNote{Almes redoũd to God's honour.}
almes are giuen, ſpecially to holy men, not only the giuers obtaine
great benefits thereby, and the wants of others be ſupplied, but God
alſo by the receiuers continual praiers and thankes-giuing therfore, is
exceedingly honoured: ſo that charitie beſtowed in this ſort, is an acte
of Gods worſhip and of religion.}
doth not only ſupplie thoſe things that the Saints want, but aboundeth
alſo by many thankes-giuings in our Lord, \V by the proofe of this
miniſterie, glorifying God in the obediẽce of your confeſſiõ vnto the
Ghoſpel of Chriſt, and in the ſimplicitie of communicating vnto them,
and vnto al, \V and in their praying for you, being deſirous of you
becauſe of the excellent grace of God in you. \V Thankes be to God for
his vnſpeakable guift.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

