%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Two Corinthians
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 2734
%%% o-2591
\startChapter[
  title={Chapter 12}
  ]

\Summary{He telleth of his incomparable viſions, 5.~but for humilitie
  liketh better to talke of his infirmities: 11.~putting the fault in
  the Corinthians for that he is faine thus to rehearſe his owne
  commendations. 13.~Where againe he reaſoneth the matter with them like
  a father, why they should preferre thoſe falſe Apoſtles before
  him. 20.~And feareth leſt at his comming he shal be compelled to
  excommunicate many of them.}

If I muſt glorie (it is not
\Fix{expedẽt}{expedient}{obvous typo, fixed in other}
indeed) but I wil come to the
\LNote{Viſions.}{S.~Cyprian
\MNote{Viſions haue no credit with heretikes.}
\Cite{(ep.~69. nu.~4.)}
complaineth that the Aduerſaries of Gods Church and Prieſts, giue no
credit to viſions. But their incredulitie is much more in our daies that
condemne al ſuch reuelations, though they be reported and recorded for
moſt certaine, of holy S.~Gregorie, S.~Bede, or who-els ſoeuer. Yea they
are ſo wicked in this caſe, that the viſion which the holy Author  of
the Booke of Machabees
\CNote{\XRef{2.~Mac.~15,~11.}}
calleth \L{fide dignum}, worthy of credit, is one cauſe why they deny
the whole booke to be Canonical: and as wel might they for this viſion
deny al S.~Paules Epiſtles, and for the like, the Actes of the Apoſtles
\XRef{Act.~9.}
\XRef{10.}
\XRef{11.}
\XRef{12.}
\XRef{27.}
And the Ghoſpel it-ſelf,
\XRef{Matth.~1,~10}
\XRef{2,~13,~19.}}
viſions & reuelatiõs of our Lord. \V I know a man in Chriſt aboue
fourteen yeares agoe (whether in the body, I know not, or out of the
body, I know not: God doth know) ſuch a one
\SNote{By this we may proue that it is neither impoſſible, incredible,
nor vndecent, that is reported by the ancient Fathers of ſome that haue
been rauiſhed or rapt (whether in body or out of body God knoweth) &
brought to 
ſee the ſtate of the next life, as wel of the ſaued as damned.}
rapt euen to the third Heauẽ. \V And I know ſuch a man (whether in the
body, or out of the body, I know not; God doth know) \V that he was rapt
into Paradiſe; & heard ſecret words, which it is not lawful for a man to
ſpeake. \V For ſuch an one I wil glorie: but for my ſelf I wil glorie
nothing, ſauing in my infirmities. \V For and if I wil glorie, I ſhal
not be fooliſh: for I ſhal ſay truth. But I ſpare, leſt any man ſhould
eſteem me aboue that which he ſeeth in me, or heareth any thing of
me. \V And leſt the greatnes of the reuelatiõs might extol me, there was
giuẽ me a prick of my fleſh, an Angel of Satã, to buffet me. \V For the
which thing thriſe I beſought our Lord, that it might depart frõ me. \V
And he ſaid to me: My grace ſufficeth thee, for power is perfited in
infirmitie. Gladly therfore wil I glorie in mine infirmitie, that the
power of Chriſt may dwel in me. \V For the which cauſe I pleaſe myſelf
in infirmities, in cõtumelies, in neceſſities, in perſecutions, in
diſtreſſes for Chriſt. For when I am weake, then am I mightie.

\V I am become fooliſh: you haue cõpelled me. For I ought to haue been
cõmended of you: for I haue been nothing leſſe then they that are
\LNote{Aboue meaſure Apoſtles.}{Though
\MNote{The Apoſtles ſome greater then others.}
al were in that they were Apoſtles, of one and the ſame order, yet we
may ſee that ſome had maruelous great preeminence and priuilege aboue
others in the ſame office: ſpecially S.~Peter and S.~Iohn, whom S.~Paul
often calleth \Emph{great Apoſtles, aboue meaſure or paſsing Apoſtles,
the pillars, &c.}
\XRef{2.~Cor.~11,~5.}
\XRef{12,~11.}
\XRef{Gal.~2,~9.}}
aboue meaſure Apoſtles: although I am nothing. \V Yet the ſignes of my
Apoſtleſhip haue beẽ done vpõ you in al patiẽce,
\LNote{In ſignes.}{Miracles
\MNote{We muſt ſtick to the faith firſt planted by miracles.}
be neceſſarie, and be great ſignes of truth, when it is firſt newly
taught. And therfore let al Catholike men hold faſt that faith which was
firſt preached & cõfirmed by miracles. As in England by S.~Auguſtin, &
in other Nations by holy Apoſtolike men. And let the Heretikes that
preach extraordinarily, newly and otherwiſe then we receiued at our firſt
conuerſion, ſhew their calling and doctrine by miracles, or els let them
be taken for falſe Apoſtles as they be.}
in ſignes & wõders & mighty deeds. \V For what is there that you haue
had leſſe thẽ the other Churches; but that I my ſelf haue not burdned you?
Pardõ me this iniurie. \V Behold, now the third time I am ready to come
to you; & I wil not be burdenous vnto you. For I ſeeke not the things
that are
%%% o-2592
yours, but you. For neither ought the children lay vp treaſures for the
parents, but the parents for the childrẽ. \V But I moſt gladly wil
beſtow & wil my ſelf moreouer be beſtowed for your ſoules: although
louing you more, I am loued leſſe.

\V But be it ſo: I haue not burdened you: but being craftie, I tooke you
by guile. \V Haue I circumuented you by any of them whom I ſent to
you? \V I requeſted Titus, and I ſent with him a brother. Did Titus
circumuẽt you? walked we not with one ſpirit? not in the ſelf-ſame
ſteps? \V
\Var{Of old}{Againe}
thinke you that we excuſe ourſelues to you? Before God in Chriſt we
ſpeake: but al things (my Deareſt) for your edifying. \V For I feare
leſt perhaps when I come, I find you not ſuch as I would: and I be found
of you ſuch an one as you would not. Leſt perhaps cõtentions,
emulations, ſtomakings, diſſenſions, detractiõs, whiſperings, ſwellings,
ſeditions be among you. \V Leſt againe when I come, God humble me among
you: & I mourne many of them that ſinned before, &
%%% !!! Originally only a SNote.
\SNote{Which S.~Auguſtin ſaith
\Cite{(Ep.~108.)}
is ſpoken here of doing great penance for heinous ſinnes,
as \L{Pœnitentes} did in the primitiue Church. So that it is not only
to repent or to amend their liues, as the Proteſtants tranſlate it.}
\TNote{\G{τῶν μὴ μετανοησάντων}}
haue not done penance for the vncleannes & fornication and
incontinencie that they haue committed.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

