%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Two Corinthians
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 2726
%%% o-2582
\startChapter[
  title={Chapter 7}
  ]

\Summary{He proceedeth to exhort them to puritie, and to receiue him
  into their charitie. 3.~Which leſt they should thinke he ſpeaketh to
  accuſe them, he commendeth them highly, both for their behauiour
  toward Titus, and for their penance which they had done vpon his other
  epiſtle.}

Hauing therfore theſe promiſes, my Deareſt, let vs cleanſe our ſelues
from al inquination of the fleſh and ſpirit, perfiting ſanctification in
the feare of God. \V Receiue vs. We haue hurt no man, we haue corrupted
no man, we haue circumuented no man. \V I ſpeake not to your
condemnation. For I ſaid before that you are in our harts to die
together and to liue together. \V Much is my confidence with you, much
is my glorying for you. I am repleniſhed with
%%% o-2583
conſolation; I doe exceedingly abound in ioy in al our tribulation. \V
For alſo when we were come into Macedonia, our fleſh had no reſt, but we
ſuffered al tribulation: without, combats; within, feares. \V But God
that comforteth the humble, did confort vs, in the comming of Titus. \V
And not only in his comming, but alſo in the conſolation, wherwith he
was comforted among you, reporting to vs your deſire, your weeping, your
emulation for me, ſo that I reioyced the more. \V For although I made
you ſorie in an epiſtle, it repenteth me not: albeit it repented me,
ſeeing that the ſame epiſtle (although but for a time) did make you
ſorie. \V Now I am glad: not becauſe you were made ſorie, but becauſe
you were made
\LNote{Sorie to penance.}{The
\MNote{Contrition for a man's ſinne worketh ſaluation.}
ſorrow which a man taketh for worldly loſſes or any temporal aduerſitie,
is not here commended, but that which is & ought to be in al men for
their ſinnes paſt, which is called here, Sorrow towards God & for
penance, otherwiſe called Contrition, & is a thing exceedingly requiſit
& much praiſed, the fruits thereof are theſe that the Apoſtle reckneth,
working ſaluation. Which doctrine is farre diſtant from
\CNote{\Cite{To.~3. in aſſet. art.~6. Leone damnat.}}
Luther's and Caluinn's, and ſuch wicked Libertines, that teach
contrition to be altogether a meanes to make ſinners either hypocrites,
or to put them in diſpaire.}
ſorie to penance. For you were made ſorie according to God, that in
nothing you ſhould ſuffer detriment by vs. \V For
\SNote{Contrition or ſorrowful lamenting of our offenſes, is the cauſe
of ſaluation. Not only faith then ſaueth, as the Heretikes affirme.}
the ſorrow that is according to God, worketh penance vnto ſaluation that
is ſtable: but the ſorrow of the world worketh death. \V For
%%% 2727
behold this very thing, that you were made ſorie according to God, how
great carefulnes it worketh in you: yea defẽſe, yea indignatiõ, yea
feare, yea deſire, yea emulatiõ, yea reuenge, in al things you haue
ſhewed your ſelues to be vndefiled in the matter. \V Therfore although I
wrote to you, not for him that did the iniurie, nor for him that
ſuffered: but to manifeſt our carefulnes that we haue for you before
God, \V therfore we are comforted. But in our conſolation, we did the
more aboundantly reioyce vpon the ioy of Titus, becauſe his ſpirit was
refreſhed of al you. \V And if to him I gloried any thing of you, I am
not confounded: but as we ſpake al things to you in truth, ſo alſo our
glorying that was to Titus, is made a truth, \V and his bowels are more
aboundantly toward you: remembring the obedience of you al, how with
feare and trembling you receiued him. \V I reioyce that in al things I
haue confidence in you.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

