%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% One Corinthians
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 2701
%%% o-2556
\startChapter[
  title={Chapter 13}
  ]

\Summary{That aboue al other Guifts they should ſeek after Charitie: as
  that without which nothing profiteth, 4.~and which doth al as is to be
  done, and remaineth alſo in Heauen.}

If I ſpeake with the tongues of men, and of Angels, and haue
\LNote{Not Charitie.}{Without
\MNote{Charitie.}
charitie, both toward euery particular perſon, and ſpecially toward the
common body of the Church, none of the guifts and graces of God be
profitable.}
not charitie, I am become as ſounding braſſe, or a tinkling cymbal. \V
And if I ſhould haue prophecie, and knew al myſteries, and al knowledge,
& if I ſhould
\SNote{This proueth that faith is nothing worth to ſaluation without
workes, and that there may be true faith without charitie.}
haue al faith ſo that I could remoue mountaines, and haue not charitie,
I am nothing. \V And if I ſhould diſtribute al my goods to be meate for
the poore, and if I ſhould
\LNote{Deliuer my body.}{\Emph{Beleeue}
\MNote{Falſe Martyrs.}
(ſaith S.~Auguſtin) \Emph{aſſuredly and hold for certaine, that no
Heretike and ſchiſmatike that vniteth not himſelf to the Catholike
Church againe, how great almes ſoeuer he giue, yea or shead his bloud
for Chriſtes name, can poſsibly be ſaued. For many Heretikes, by the
cloke of Chriſtes cauſe deceiuing the ſimple ſuffer much, but where true
faith is not, there is no iuſtice, becauſe the iuſt liueth by faith. So
it is alſo of Schiſmatikes, becauſe where charitie is not, iuſtice can
there be none: which if they had, 
\Fix{the}{they}{obvious typo, fixed in other}
would neuer plucke in peeces the body of Chriſt which is the Church.}
\Cite{Aug. feu. Fulg. de fid. ad Pet. c.~39.}
So ſaith S.~aug. in diuers places, not only of Heretikes that died
directly for defenſe of their hereſie, as the Anabaptiſtes and Caluiniſts
now adaies doe (for that is more damnable:) but of ſome Heretikes and
Schiſmatikes that may die among the Heathen or Turkes for defenſe of
truth or ſome Article of Chriſtes religion.
\Cite{Aug. de verb. Do. ſer.~50. c.~2.}
&
\Cite{in Pſ.~34. conc.~2. prope finem.}
\Cite{Cyp. de vnit. Ec. nu.~8.}}
deliuer my body ſo that I burne, and haue not charitie, it doth profit
me nothing.

\V Charitie is patient, is benigne: Charitie enuieth not, dealeth not
peruerſly: is not puffed vp, \V is not ambitious, ſeeketh not her owne,
is not prouoked to anger, thinketh not euil, \V reioyceth not vpõ
iniquitie, but reioyceth
%%% o-2557
with the truth: \V ſuffereth al things, beleeueth al things, hopeth al
things, beareth al things. \V Charitie neuer falleth away: whether
prophecies ſhal be made void, or tongues ſhal ceaſe, or knowledge ſhal
be deſtroied. \V For in part we know, & in part we prophecie. \V But
\SNote{By this text S.~Auguſtin
\Cite{lib.~22. ciu. c.~29.}
proueth that the Saints in Heauen haue more perfect knowledge of our
affaires here, then they had when they liued here.}
when that ſhal come that is perfect, that ſhal be made void that is in
part. \V When I was a litle one, I ſpeake as a litle one, I vnderſtood
as a litle one, I thought as a litle one. But when I was made a man, I
did away the things that belonged to a litle one. \V We ſee now by a
glaſſe in a darke ſort: but then face to face. Now I know in part: but
then I ſhal know as alſo I am knowen. \V And now there remaine, faith,
hope, charitie,
%%% !!! only marked in other.
\LNote{Theſe three.}{Theſe
\MNote{The 3.~vertues Theological.}
are the three vertues Theological, each one by nature and definition
diſtinct from another: and faith is by nature the firſt, and may be and
often is before, and without Charity:
\MNote{Charitie loſt by mortal ſinne, not faith.}
and truely remaineth in diuers
after they haue by deadly ſinne loſt charitie. Beware therfore of the
Heretikes opinion, which is, that by euery mortal ſinne faith is loſt no
leſſe then charitie.}
theſe three: but the
\SNote{Charitie is of al the three the greateſt. How then doth only
faith, being inferiour to it ſaue & iuſtifie and not charitie?}
greater of theſe is charitie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

