%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% One Corinthians
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 2708
%%% o-2564
\startChapter[
  title={Chapter 15}
  ]

\Summary{He
\MNote{The 7.~part.

Of the reſurrection of the dead.}
proueth the reſurrection of the dead by the reſurrection of Chriſt, and
with many other arguments: and 31.~anſwereth alſo obiections made
againſt it. 49.~And then exhorteth in reſpect of it, vnto good life.}

And I doe you to vnderſtand, Brethren, the Ghoſpel which I preached to
you, which alſo you receiued, in the which alſo you ſtand, \V by the
which alſo you are ſaued, after what manner I preached vnto you if you
keep it, vnleſſe you haue beleeued in vaine. \V For
\SNote{This deliuerie in the Latin & Greek
importeth \Emph{tradition}. So by Tradition did the Apoſtles plant the
Church in al truth, before they wrote any thing.}
\TNote{\L{Tradidi} \G{παρέδωκα}}
I deliuered vnto you firſt of al which I alſo receiued: that Chriſt died
for our ſinnes
\CNote{\XRef{Eſ.~53,~8.}}
according to the Scriptures: \V and that he was buried, and that he roſe
againe the third day,
\CNote{\XRef{Dan.~9,~26.}
\XRef{Pſ.~15,~10.}}
according to the Scriptures: \V and that he was
\CNote{\XRef{Ion.~2,~2.}
\XRef{Lu.~24.}}
ſeen of Cephas; and after that of the eleuen. \V Then
%%% 2709
was he ſeen of moe then fiue hundred Brethren together: of which many
remaine vntil this preſent, and ſome are a-ſleep. \V Moreouer he was
ſeen of Iames, then of al the Apoſtles. \V And laſt of al, as it were of
\TNote{one borne out of time.}
and abortiue
\CNote{\XRef{Act.~9,~3.}}
he was ſeen alſo of me. \V For I am the leaſt of the Apoſtles, who am
not worthy to be called an Apoſtle, becauſe I perſecuted the Church of
God. \V But by the grace of God I am that which I am; & his grace in me
hath not been
\SNote{In him God's grace is not void, that worketh by his free wil
according to the motion and direction of the ſame grace.}
void, but I haue laboured more aboundantly then al they: yet not I, but
the grace of God
\LNote{With me.}{God
\MNote{Free-wil with grace.}
vſeth not man as a brute beaſt or a block: but ſo worketh in him and by
him that free-wil may concurre in euery action with his grace, which is
alwaies the principal.
\MNote{Heret. tranſlation.}
The heretikes to auoid this concurrence in working and labouring,
tranſlate, \Emph{which is with me}: where the Apoſtle rather ſaith,
\Emph{which laboureth with me}.}
with me. \V For whether I, or they, ſo we preach, and ſo you haue
beleeued.

\V But if Chriſt be preached, that he is riſen againe from the dead, how
doe certaine among you ſay, that there is no reſurrection of the
dead? \V And if there be no reſurrection of the dead, neither is Chriſt
riſen againe. \V And if Chriſt be not riſen againe, then vaine is our
preaching, vaine alſo is your faith. \V And we are found alſo
\SNote{So may we ſay: If the Cath. faith in al points be not true, then
our firſt Apoſtles were falſe witneſſes, then hath our Countrie beleeued
in vaine al this while, then are al our Forefathers dead in their ſinnes
and perished. Which (preſuppoſing Chriſt to be God) were the greateſt
abſurditie in the world.}
falſe witneſſes of God: becauſe we haue giuen teſtimonie againſt God,
that he hath raiſed vp Chriſt, whom he hath not raiſed vp, if the dead
riſe not againe. \V For if the dead riſe not againe, neither is Chriſt
riſen againe. \V And if Chriſt be not riſen againe, vaine is your faith,
for yet you are in your ſinnes. \V Then they alſo that are a-ſleep in
Chriſt, are periſhed. \V If in this life onely we be hoping in Chriſt, we
are more miſerable then al men.

%%% o-2565
\V But now Chriſt is riſen againe from the dead, the
\CNote{\XRef{Col.~1,~18.}
\XRef{Ap.~1,~5.}}
firſt fruits of them that ſleep: \V by a man death; and by a man the
reſurrection of the dead. \V And as in Adam al die, ſo alſo in Chriſt al
ſhal be made aliue. \V But
\CNote{\XRef{Ro.~5,~12.}
\XRef{1.~Theſ.~4,~15.}}
euery one in his owne order: the firſt fruits Chriſt, then they that
are of Chriſt, that beleeued in his comming. \V Then the end, when he
ſhal haue deliuered the Kingdom to God and the Father, when he ſhal haue
aboliſhed al principalitie & authoritie & power. \V And he muſt reigne,
\CNote{\XRef{Pſ.~109,~1.}}
\Emph{Vntil he put al his enemies vnder his feet.} \V And the enemie
death ſhal be deſtroied laſt. \Emph{For he hath ſubdued al things vnder
his feet.} And wheras he ſaith, \V
\CNote{\XRef{Pſ.~8,~8.}}
\Emph{Al things are ſubdued to him}; vndoubtedly, except him that
ſubdued al things vnto him. \V And when al things ſhal be ſubdued to
him; then the Sonne alſo himſelf ſhal be ſubiect to him that ſubdued al
things vnto him, that God may be Al in al.

\V Otherwiſe what ſhal they doe that are baptized for the dead, if the
dead riſe not againe at al? \V Why alſo are they baptized for them? Why
alſo are we in danger euery houre? \V I die daily by your glorie,
Brethren, which I haue in Chriſt \Sc{Iesvs} our Lord. \V If (according
to man) I fought with beaſts at Epheſus, what doth it profit me, if the
dead riſe not againe?
\CNote{\XRef{Eſa.~22,~13.}}
\LNote{Let vs eate and drinke.}{S.~Ambroſe
\MNote{Faſting is meritorious.}
applieth theſe words to our Chriſtian Epicurians that take away faſting,
and deny the merit therof: \Emph{How can we be ſaued} (ſaith he)
\Emph{if we wash not away our ſinnes by faſting, ſeeing the ſcriptures
ſay, faſting and almes deliuer from ſinne? What are theſe new Maiſters
then that exclude al merit of faſting? is not this the very voice of
the Heathen, ſaying: Let vs eate and drinke, to morow we shal die?}
\Cite{li.~10. epiſt. ep.~81.}}
\Emph{Let vs eate and drinke, for to morow we shal die.} \V Be not
ſeduced,
\CNote{\Cite{Menander.}}
\Emph{Euil communications corrupt good manners.} \V Awake ye iuſt, and
ſinne not. For ſome haue not the knowledge of God, I ſpeake to your
ſhame.

\V But ſome man ſaith: How doe the dead riſe againe? and with what
manner of body ſhal they come? \V Foole, that which thou ſoweſt is not
quickned, vnleſſe it die firſt. \V And that which thou ſoweſt, not the
body that ſhal be, doeſt thou ſow; but bare graine, to wit, of wheat, or
%%% 2710
\Fix{or of}{of}{obvious typo, fixed in other}
ſome of the reſt. \V And God giueth it a body as he wil: & to euery ſeed
his proper body. \V Not al fleſh, is the ſame fleſh: but one of men,
another of beaſts, another of
\Fix{birbs,}{birds,}{obvious typo, fixed in other}
another of fiſhes. \V And bodies
celeſtial, & bodies terreſtrial: but, one glorie of the celeſtial, and
another of the terreſtrial. \V One indeed glorie of the ſunne, another
glorie of the moone, and another glorie of the ſtarres. For
\SNote{The glorie of the bodies of Saints shal not be al alike, but
different in Heauen according to mens merits.}
ſtarre differeth from ſtarre in glorie: \V ſo alſo the reſurrection of
the dead. It is ſowen in corruption, it shal riſe in incorruption. \V It
is ſowen in diſhonour, it shal riſe in glorie. It is
%%% o-2566
ſowen in infirmitie, it ſhal riſe in power. \V It is ſowen a natural
body; it ſhal riſe a
\SNote{As to become ſpiritual doth not take away the ſubſtance of the
body glorified: no more when Chriſtes body is ſaid to be in ſpiritual
ſort in the Sacrament, doth it import the abſence of his true body and
ſubſtance.}
ſpiritual body. If there be a natural body, there is alſo a
ſpiritual, \V as it is written:
\CNote{\XRef{Gen.~2,~7.}}
\Emph{The firſt man Adam was made into a liuing ſoule}: the laſt Adam
into a quickning ſpirit. \V Yet that is not firſt which is ſpiritual,
but that which is natural: afterward that which is ſpiritual. \V The
firſt man of earth, earthly: the ſecond man from Heauen, heauenly. \V
Such as is the earthly, ſuch alſo are the earthly: and ſuch as the
heauenly, ſuch alſo are the heauenly. \V Therfore as we haue borne the
image of the earthly, let vs beare alſo the image of the heauenly. \V
This I ſay, Brethren, that
\SNote{Flesh and bloud ſignifie not here the ſubſtance of thoſe things,
but the corrupt qualitie incident to them in this life by the fal of
Adam.}
fleſh and bloud can not poſſeſſe the Kingdom of God: neither ſhal
corruption poſſeſſe incorruption.

\V Behold I tel you a myſterie. We ſhal al indeed riſe againe: but we
ſhal not al be changed. \V In a moment, in the twinkling of an eye, at
the
\CNote{\XRef{Ap.~8,~2.}}
laſt trompet (for
\CNote{\XRef{11,~15.}}
the trompet ſhal ſound) and the dead ſhal riſe againe incorruptible: and
we ſhal be changed. \V For this corruptible muſt doe-on
incorruption; & this mortal doe-on immortalitie. \V And when this
mortal doe-on immortalitie, then ſhal come to paſſe the ſaying that is
written:
\CNote{\XRef{Oſ.~13,~14.}}
\Emph{Death is ſwallowed vp in victorie. \V Death where is thy victorie?
Death where is thy ſting?} \V And the 
ſting of death, is ſinne: and the power of ſinne is the Law. \V But
thankes be to God that hath giuen vs the victorie by our Lord \Sc{Iesvs}
Chriſt. \V Therfore, my beloued Brethren, be ſtable and vnmoueable;
abounding in the worke of our Lord alwaies, knowing that your labour is
not vaine in our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

