%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% One Corinthians
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-12


\project douay-rheims


%%% 2699
%%% o-2554
\startChapter[
  title={Chapter 12}
  ]

%%% !!! At least one of the '12' below is wrong.
\Summary{They
\MNote{The 6.~part.

As touching the Guifts of the Holy Ghoſt.}
muſt not make their diuerſitie of Guifts an occaſion of Schiſme,
conſidering that al are of one Holy Ghoſt, and for the profit of the one
body of Chriſt which is the Church. 12.~Which alſo could not be a body,
without ſuch varietie of members. 12.~Therfore neither they that haue
the inferiour guifts, muſt be diſcõtent, ſeeing it is God's
diſtribution: nor they that haue the greater, contemne the other,
conſidering they are no leſſe neceſſarie: 25.~but al in al ioyne
together, 28.~and euery one know
\Fix{is}{his}{obvious typo, fixed in other}
owne place.}

And concerning ſpiritual things, I wil not haue you ignorant,
Brethren. \V You know that when you were Heathen, you went to dumme
Idols according as you were led. \V Therfore I doe you to vnderſtand
that no man ſpeaking in the Spirit of God,
%%% 2700
ſaith anathema to \Sc{Iesvs}. And no man can ſay: Our Lord \Sc{Iesvs},
but in the Holy Ghoſt.

%%% o-2555
\V And there are diuiſions of graces, but one Spirit. \V And there are
diuiſions of miniſtrations; but one Lord. \V And there are diuiſions of
operations, but one God, which worketh al in al. \V And the manifeſtatiõ
of the Spirit is giuẽ vnto euery one to profit. \V To one certes by
the Spirit is giuen
\SNote{Al theſe Guifts be thoſe which the learned cal \L{Gratias gratis
datas}: which be beſtowed often euen vpon il liuers, which haue not the
other graces of God whereby their perſons ſhould be grateful, iuſt, &
holy in his ſight.}
the word of wiſedom: and to another, the word of knowledge according to
the ſame Spirit: \V
\LNote{Faith in the ſame.}{This
\MNote{Zealous faith.}
faith is not another in ſubſtance then the common faith in Chriſt, but
is of another accidental qualitie only, that is, of more feruour,
deuotion, zeale and confident truſt, ſpecially for doing of miracles.}
to another faith in the ſame Spirit: to another, the grace of doing
cures in one Spirit: \V to another, the working of miracles: to another,
prophecie: to another, diſcerning of Spirits: to another, kinds of
tongues: to another, interpretation of languages. \V And al theſe things
worketh one and the ſame Spirit, diuiding to euery one according as he wil.

\V For
\CNote{\XRef{Ro.~12,~4.}
\XRef{Eph.~4,~7.}}
as the body is one, and hath many members, and al the members of the body
wheras they be many, yet are
\SNote{A maruelous vnion betwixt Chriſt & his Church, & a great comfort
to al Catholikes being members therof, that the Church and he, the head
& the body, make & be called one Chriſt.
\Cite{Augu. de vnit Eccl.}}
one body; ſo alſo Chriſt. \V For in one Spirit, were we al baptized into
\Var{one,}{one body,}
whether Iewes, or Gentils, or bondmen, or free: & in one Spirit we were
al made to drinke. \V For the body alſo is not one member, but many. \V
\LNote{If the foot.}{The
\MNote{Vnitie.}
Church is of exceeding great diſtinction of members, guifts, orders, and
offices: yet of great concord, concurrence, mutual communion and
participation, in al actions of her members among themſelues, and with
Chriſt the head of the body.}
If the foot ſhould ſay, becauſe I am not the hand, I am not of the body:
is it therfore not of the body? \V And if the eare ſhould ſay, becauſe I
am not the eye, I am not of the body: is he therfore not of the body? \V
If the whole body were the eye: 
where is the hearing? If the whole were the hearing: where is the
ſmelling? \V But now God hath ſet the members, euery one of them in the
body as he would. \V And if al were one member, where were the body? \V
But now there are many members indeed, yet one body. \V And the eye can
not ſay to the hand: I need not thy help; or againe the head to the
feet: You are not neceſſarie for me. \V But much more thoſe that ſeeme
to be the more weak members of the body, are more neceſſarie: \V and
ſuch as we thinke to be the baſer mẽbers of the body, vpon them we put
more aboundant honour: and thoſe that are our vnhoneſt
parts, haue more aboundant honeſtie. \V And our honeſt parts need
nothing: but God 
hath tempered the body, giuing to it that wanted, the more aboundant
honour, \V that there might be no
\LNote{Schiſme in the body.}{As
\MNote{Schiſme.}
Charitie and vnitie of ſpirit, is the proper bond and weale of the
common Body: ſo is diuiſion or ſchiſme, which is the interruption of
peace and mutual Societie between the parts of the ſame, the ſpecial
plague of the Church, and as odious to God as rebellion to the temporal
Soueraigne.}
ſchiſme in the body, but the members together might be careful one for
another. \V And if one member ſuffer any thing, al the members ſuffer
with it. Or if one member doe glorie, al the members reioyce with it.
%%% o-2556
\V And you are the body of Chriſt, and members of member.

\V And
\CNote{\XRef{Eph.~4,~1.}}
ſome verily God hath ſet in the Church firſt Apoſtles, ſecondly
Prophets, thirdly Doctours, next miracles, then the graces of doing
cures, helps, gouernements, kinds of tongues. \V Are al Apoſtles? are al
Prophets? are al Doctours? \V
\SNote{S.~Auguſtine
\Cite{ep.~137.}
giueth the ſame reaſon, why miracles & cures be done at the memories or
bodies of ſome Saints more then at others: & by the ſame Saints in one
place of their memories rather then at other places.}
are al miracles? haue al the grace of doing cures? doe al ſpeake with
tongues? doe al interpret? \V But purſue the better guifts. And yet I
ſhew you a more excellent way.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

