%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% One Corinthians
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 2711
%%% o-2567
\startChapter[
  title={Chapter 16}
  ]

\Summary{He
\MNote{The 8.~part.

Of the contributions.}
preſcribeth an order for their contributing to the Chriſtians at
Hieruſalem, 3.~promiſing to come vnto them. 10.~Of Timothee, and of
Apollo's comming thither. 13.~And ſo with exhortation, and diuers
commendations, he endeth.}

And concerning the collections that are made for the Saints, as I haue
ordeined to the Churches of Galatia, ſo doe ye alſo. \V In
\SNote{That is Sunday.
\Cite{Hiero. q.~4. Hedibiæ.}
So quickly did the Chriſtians keepe Sunday, holiday, and aſſembled to
Diuine Seruice on the ſame.}
the firſt of the Sabboth let euery one of you put a part with himſelf,
laying vp what ſhal wel like him: that not when I come, then collections
be made. \V And when I ſhal be preſent; whom you ſhal approue by letters,
them wil I ſend to carie your grace into Hieruſalem. \V And if it be
worthie that I alſo goe, they ſhal goe with me.

\V And I wil come to you, when I ſhal haue paſſed through Macedonia. For
I wil paſſe through Macedonia. \V And with you perhaps I wil abide, or
wil winter alſo: that you may bring me on my way whither-ſoeuer I
goe. \V For I wil not now ſee you by the way, for I hope that I ſhal
abide with you ſometime, if our Lord wil permit. \V But I wil tarie at
Epheſus vntil
\SNote{The Heretikes & other new-fangled ſtriue among themſelues,
whether Pentecoſt ſignifie here the terme of fiftie daies, or els the
Iewes holy-day ſo called. But it cõmeth not to their minds that it is
moſt like to be the feaſt of whitſontide, kept & inſtituted euen then by
the Apoſtles, as appeareth by the Fathers. See
\Cite{S.~Aug. ep.~119. c.~15. &~16.}
\Cite{Ambr. in c.~17. Lucæ.}}
Pentecoſt. \V For a great doore and euident is opened vnto me: and many
aduerſaries. \V And if Timothee come, ſee that he be without feare with
you, for he worketh the worke of our Lord, as alſo I. \V Let no man
therfore deſpiſe him, but conduct ye him in peace: that he may come to
me. For I expect him with the Brethren. \V And of Brother Apollo I doe
you to vnderſtand, that I much intreated him to come vnto you with the
Brethren: and
\TNote{\G{πάντως}}
at al it was not his mind to come now. But he wil come when he ſhal haue
leiſure.

\V Watch ye, ſtand in the faith, doe manfully, & be ſtrengthned. \V Let
al your things be done in charitie. \V And I beſeech you, Brethren, you
know the houſe of Stephanas, and of Fortunatus, that they are the firſt
fruits of Achaia, & haue ordeined themſelues to the miniſterie of the
Saints: \V that you alſo be ſubiect to ſuch, and to euery one that
helpeth & laboureth with vs. \V And I reioyce in the preſence of
Stephanas and Fortunatus and Achaicus, becauſe
\TNote{\G{τὸ ὑμῶν ὑστέρημα}}
that which you wanted, they haue ſupplied. \V For they haue refreſhed
both my ſpirit and yours. Know them therfore that are ſuch.

%%% o-2568
\V The Churches of Aſia ſalute you. Aquila and Priſcilla with their
domeſtical Church ſalute you much in our Lord. \V Al the Brethren ſalute
you. Salute one another in a
\CNote{See
\XRef{Ro.~16,~16.}}
holy kiſſe. \V The ſalutation with mine owne hand Paules. \V If any man
loue not our Lord \Sc{Iesvs} Chriſt, be he \GG{anathema} 
\TNote{\H{אתא מרן}}
\HH{Maranatha}. \V The grace of our Lord \Sc{Iesvs} Chriſt be with
you. \V My charitie be with you al in Chriſt \Sc{Iesvs}. Amen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

