%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% One Corinthians
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 2681
%%% o-2535
\startChapter[
  title={Chapter 6}
  ]

\Summary{He rebuketh them for going to law before Iudges that were not
  Chriſtians, 9.~telling that extorſion (as many other offenſes
  likewiſe) is a mortal ſinne. 12.~And with diuers reaſons he inueigheth
  againſt fornication, bidding alſo to fly al occaſion therof.}

Dare any of you hauing a matter againſt another, to be iudged before the
vniuſt, & not before the Saints? \V Or know you not that
\SNote{The faithful iudge and giue ſentence with God at the latter day,
ſpecially the Apoſtles and the perfect Chriſtians that haue forſaken al
for Chriſt's ſake.}
the Saints ſhal iudge of the world? And if the world ſhal be iudged by
you: are you vnworthie to iudge of the leaſt things? \V Know you not
that we ſhal iudge Angels? how much more ſecular things? \V If therfore
you haue ſecular iudgements; the contemptible that are in the Church,
ſet them to iudge. \V I ſpeake to your ſhame. So is there not among you
any wiſe man, that can iudge between his brother? \V But brother with
brother
\LNote{Contendeth in iudgement.}{To
\MNote{Going to law before heathen or heretical iudges.}
be giuen to brabling and litigiouſnes for euery trifle, to ſpend a pound
rather then loſe a peny, the Apoſtle much reprehendeth in Chriſtian
men. For a Chriſtian man to draw another to the iudgements, ſeats, and
courts of Heathen Princes (which then only reigned) and not to ſuffer
their controuerſies and quarels to be taken vp among themſelues
brotherly and peaceably, was a great fault: as, for one Catholike to
draw another for mere trifles before ſecular or heretical Officers, is a
very vnchriſtian part.}
contendeth in iudgement: & that before infidels? \V Now certes there is
plainely
\LNote{A fault.}{He
\MNote{Going to law not forbidden but to agree otherwiſe better.}
forbad not al iudgements of controuerſies, but only ſignified that it
was a fault, and that it proceeded of ſome iniuries done one to another,
& imperfectiõs, that they ſo moleſted one another: and that it had been
nor agreable to Chriſtian perfection and charitie, rather to tolerate
and ſuffer a ſmal iniurie, then to draw his fellow to iudgement ſeats.}
a fault in you, that you haue iudgements among you. Why doe you not
rather take wrong? why doe you not rather ſuffer fraud? \V But your
ſelues doe wrong and defraud: and that to Brethren. \V Know you not that
the vniuſt ſhal not poſſeſſe the Kingdom of God? Doe not erre: Neither
fornicatours, nor
\SNote{For this, the
\Cite{English Bible~1562.}
falſely tranſlateth, \Emph{worshippers of images}.}
\TNote{\G{εἰδωλολάτραι}}
ſeruers of Idols, nor aduouterers, nor the effeminate, nor the liers
with mankind, \V nor theeues, nor the couetous, nor drunkards, nor
railers, nor extorſioners ſhal poſſeſſe the Kingdom of God. \V And theſe
things certes you were, but you are waſhed, but you are ſanctified, but
you are iuſtified in the name of our Lord \Sc{Iesvs} Chriſt, and in the
Spirit of our God.

\V Al things are lawful for me, but al things are not expedient. Al
things are lawful for me, but I wil be brought vnder the power of
none. \V The meat to the belly, &
%%% o-2536
the belly to the meats: but God wil deſtroy both it and them: and the
body not to fornication, but to our Lord, & our Lord to the body. \V But
God both hath raiſed vp our Lord, and wil raiſe vp vs alſo by his
power. \V Know you not that your bodies are the members of Chriſt?
Taking therfore the members of Chriſt, ſhal I make them the members of
an harlot? God forbid. \V Or know you not, that he which cleaueth to an
harlot, is made one body?
\CNote{\XRef{Gen.~2,~24.}}
\Emph{For they shal be}, ſaith he, \Emph{two in one flesh}. \V But he
that cleaueth to our Lord, is one ſpirit. \V Fly fornication. Euery
ſinne whatſoeuer a man doeth, is without the body: but he that doth
fornicate, ſinneth
\SNote{Fornication is not only enemy to the ſoule but waſteth,
weakeneth, corrupteth and defileth the body more properly and directly
then any other ſinnes doe.}
againſt his
%%% 2682
owne body. \V Or know you not that your members are the temple of the
holy Ghoſt which is in you, whome you haue of God, and you are not your
owne. \V For you are bought with a great price. Glorifie and beare God
in your body.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

