%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% One Corinthians
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-08


\project douay-rheims


%%% 2686
%%% o-2541
\startChapter[
  title={Chapter 8}
  ]

\Summary{He
\MNote{The 4.~part.

Of meats ſacrificed to Idols.}
rebuketh the learned who in pride of their knowledge did eate
\L{Idolothyta}, that is things offered to Idols, vſing (as they ſaid) their
libertie; but not conſidering that the ignorant tooke their doing as an
example for them to frequent ſuch meats ſo, as they did before in their
Paganiſme, with opinion that they did ſanctifie the eaters.}

And concerning thoſe things that are ſacrificed to Idols, we know that
\LNote{We al haue knowledge.}{The
\MNote{No meats vncleane.}
ſpiritual and perfectly inſtructed Chriſtians knew no meats now to be
vncleane, neither for ſignification, as in the Law of Moyſes; nor
alwaies by nature and creation, as the Manichees thought; nor by any
other pollution, as in that they were offred to Idols: and therefore
they did eate boldly of ſuch meats as were ſacrificed, contemning &
condemning their Idols as mere nothing, and the worship of them as the
honour of things imaginarie.
\MNote{Giuing of ſcandal reprehended.}
Which their fact, for their want of diſcretion and charitie, and for the
vſe of that their libertie to the offenſe & ſcandal of the weake, the
Apoſtle doth here reprehend.}
we al haue knowledge.
\SNote{Knowledge without charitie puffeth vp in pride, and profiteth
nothing at al: when it is ioyned with charitie, then it edifieth.
\Cite{Aug. li.~9. ciu. Dei. c.~20.}}
Knowledge puffeth vp; but charitie edifieth. \V And if any man thinke
that he knoweth ſomething, he hath not yet knowen, as he ought to
know. \V But if any man loue God, the ſame is knowen of him. \V But as
for the meats that are immolated to Idols, we know that an Idol is
nothing in the world, and that there is no God, but one. \V For although
there be that are called gods, either in Heauen, or in earth, (for there
are many gods, and many lords) \V yet to vs there is one God, the
Father, of whom al things, and we vnto him: and one Lord \Sc{Iesvs
Christ}, by whom al things, and we by him. \V But there is not knowledge
in al. For
\LNote{Some with a conſcience.}{The perfecter mens fault was, that they
gaue offenſe by their eating, to the weaker Chriſtians. Who ſeeing them
whom they reputed wiſe & learned, to eate the meats, offered to Idols,
conceiued that there was ſome vertue and ſanctification in thoſe meats,
from the Idol to which they were offered: and thought that ſuch things
were or might be eaten with the ſame conſcience and deuotion as before
their conuerſion.

Therfore
\MNote{The Heretikes ridiculouſly apply S.~Paules words againſt the
Churches faſts and abſtinence.}
the caſe ſtanding thus, and the Apoſtles diſcourſe of eating or not
eating meats being ſo as is declared (a thing ſo euidẽt that it
admitteth no other interpretatiõ) if the Proteſtants apply any of this
admonition againſt our faſts in the Cath. Church, they be too
ridiculous.}
ſome vntil this preſent with a conſcience of the Idol, eate as a thing
ſacrificed to Idols: and their conſcience being weak, is polluted. \V
But meate doth not commend vs to God. \V For neither if we eate, ſhal we
abound: nor if we eate not, ſhal we lack. \V But take heed leſt perhaps
this your libertie become an offenſe to the weake. \V For if a man ſee
him that hath knowledge, ſit at table
\LNote{In the Idol's Temple.}{Like
\MNote{Going to the Communion, what a ſinne in Catholikes.}
as now, ſome Catholikes haue ſaid, they know that Caluin's Communion is
but as other bread and wine. But yet the ignorant ſeeing ſuch goe to the
Communion, thinke that it is a good acte of Religion. Yea whatſoeuer
they pretend, it muſt needs ſeeme an honour to Caluin's Communion, when
they are ſeen in the Idol's Temple ſolemnly ſitting or communication at
the abominable table.}
in the Idol's Temple; ſhal not his conſcience, being weake, be edified,
to eate things ſacrificed to Idols? \V And through thy knowledge ſhal
the
\CNote{\XRef{Ro.~14,~15.}}
weake Brother periſh, for whom Chriſt hath died? \V But ſinning thus
againſt the Brethren, and ſtriking their weake conſcience; you ſinne
againſt Chriſt. \V Wherfore if
\CNote{\XRef{Ro.~14,~21.}}
meate ſcandalize my Brother, I wil neuer eate fleſh, leſt I ſcandalize
my Brother.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

