%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% One Corinthians
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 2678
%%% o-2531
\startChapter[
  title={Chapter 4}
  ]

\Summary{He requireth to be eſteemed for his office, but regardeth not
  to be praiſed of man for his vertue: conſidering that neither his owne
  conſcience is a ſufficient iudge therof, but only God who ſeeth
  al. 8.~He toucheth them for contemning in their pride, the Apoſtles
  themſelues as miſerable: 18.~threatning to come to thoſe proud
  Falſe-apoſtles who were the Authours of al theſe ſchiſmes.}

%%% o-2532
So let a man eſteeme vs as the Miniſters of Chriſt, and the diſpenſers
of the myſteries of God. \V Here now is required among the diſpenſers
that a man be found faithful. \V But to me it is a thing of leaſt
account, to be iudged of you, or of man's day: But I iudge not my-ſelf
neither. \V For I am not guilty in conſcience of anything,
\LNote{But not iuſtified.}{The
\MNote{No man ſure of grace or iuſtification.}
Heretikes are certaine that they be in God's grace, but S.~Paul though
guiltie of no crime in his conſcience, durſt not aſſure himſelf that he
was iuſtified, neither could take vpon him to be iudge of his owne hart
and cogitations, whether they were pure or no: but the trail therof he
left only to God's iudging day.}
but I am not iuſtified herein: but he that iudgeth me, is our Lord. \V
Therfore iudge not before the time; vntil our Lord doe come, who alſo wil
lighten the hidden things of darknes, & wil manifeſt the counſels of
the harts: & then the praiſe ſhal be to euery man of God.

\V But theſe things, Brethren,
\SNote{Loe when he named himſelfe, & Apollo, & Cephas he meant other
ſeditious and factious Preachers whoſe names he ſpared.}
I haue transfigured into my ſelf and Apollo, for you; that in vs you may
learne, one not to be puffed vp for one againſt another, aboue that is
written. \V For who diſcerneth thee? Or what haſt thou that thou haſt
not receiued? And if thou haſt receiued what doeſt thou  glorie as
though thou haſt not receiued? \V Now you are filled, now are you become
rich: without vs you reigne; & I would to God you did reigne, that we
alſo might reigne with you. \V For I thinke that God hath ſhewed vs
Apoſtles the laſt, as it were deputed to death: becauſe we are made a
ſpectacle to the world, and to Angels and men. \V We are fooles for
Chriſt; but you wiſe in Chriſt. We weak; but you ſtrong. You noble, but
we baſe. \V Vntil this houre we doe both hunger, and thirſt, and are
naked, and are beaten with buffets, and are wanderers, \V and labour
working with our owne hands. We are curſed; and doe bleſſe. We are
perſecuted; and ſuſtaine it. \V We are blaſphemed; and we beſeech. We
are made the refuſe of this world, the droſſe of al euen vntil not. \V
Not to confound you, doe I write theſe things; but as my deereſt
children I admoniſh you. \V For
\SNote{So may S.~Auguſtin our Apoſtle ſay to vs Englishmẽ.}
if you haue ten thouſand Pædagogues in Chriſt; yet not many Fathers. For
in Chriſt \Sc{Iesvs} by the Ghoſpel I begat you. \V I beſeech you
therfore be followers of me. \V Therfore haue I ſent to you Timothee,
who is my deereſt Sonne and faithful in our Lord; who
%%% 2679
wil put you in mind of my waies that are in Chriſt \Sc{Iesvs}, as
euerywhere in euery Church I teach. \V As though I would not come to
you, ſo certaine are puffed vp. \V But I wil come to you quickly, if our
Lord  wil: and wil know not the words
%%% o-2533
of them that be puffed vp, but the power. \V For the Kingdom of God is
not in words, but in power. \V What wil you?
\LNote{In rod.}{The
\MNote{Spiritual power to punish or pardon.}
Apoſtles haue power of diſcipline and cenſures againſt offenders, and
power of gentlenes, meeknes, and indulgence alſo; to vſe either
punishing or pardoning, according to their wiſedom, and according to the
occaſions of time and place.}
in rod that I come to you; or in charitie, and the ſpirit of mildnes?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

