%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 2588
%%% o-2433
\startChapter[
  title={Chapter 15}
  ]

\Summary{Some of thoſe Iewes alſo that were Chriſtians, doe fal and are
  authours of the Hereſie of Iudaizing. 2.~They referre the matter to
  Councel: 7.~Wherein after great diſputation, Peter ſtriking the
  ſtroke, 12.~and other confirming his ſentence with miracles, 13.~and
  with Scriptures: 22.~the Apoſtles and Prieſts doe write and command in
  the name of the Holy Ghoſt what is to be done. 30.~And the faithful
  thereby are ſtraightwaies quieted in mind. 36.~After which, Paul and
  Barnabas thinking to goe againe their aboueſaid circuit together, are
  by occaſion of Marke parted, to the greater increaſe of the Church.}

And certaine comming downe from Iewrie, taught the Brethren: That
\CNote{\XRef{Gal.~5,~2.}}
vnles you be circumciſed according to the manner of Moyſes, you can not
be ſaued. \V No litle ſedition therfore being riſen to Paul and Barnabas
againſt them, they
\LNote{Appointed.}{We learne by this example, what is to be done when
any controuerſie ariſeth in religion between the teachers or other
Chriſtian people. We ſee it is not enough to contend by allegations of
Scriptures or other proofes ſeeming to make for either part: for ſo of
contentious part-taking there should be no end, but the more writing,
wreſtling, ſtriuing there were, euery one for his owne fanſie, cloking
it with the title of Gods word and Scripture, the more Schiſmes, Sectes,
and diuiſions would fal: as we ſee ſpecially in the reſtles Hereſies of
our time. Whoſe fautors admitting no iudges, ſtand to no trial of mortal
men, to no tribunal of Pope, Councels, Bishops, Synodes, but ech man to
his owne phantaſtical ſpirit, his owne ſenſe of Scriptures, and his owne
wilful obdurate rebelliõ againſt Gods Church and his Superiours in the
ſame.
\MNote{The way to end diſſenſion in religion, is to commit it to a
Councel.}
But here we ſee S.~Paul and Barnabas, men that were Apoſtles and
ful of the Spirit of God, & the other parties, though neuer ſo much
partial to the ceremonies of their Law by their former long vſe and
education therein, yet not to ſtand ſtifly to their owne opiniõ on
either ſide, but to condeſcend to referre the whole controuerſie and the
determination thereof to the Apoſtles, Prieſts or Ancients of
Hieruſalem, that is to ſay, to commit the matter to be tried by the
Heads and Bishops, and their determination in Councel. This is Gods holy
and wiſe prouidence among other iudgements in his Church, to keepe the
Chriſtian people in truth & vnitie, and to condemne ſectes, and falſe
teachers and troublers of the Church. By which iudgements and order,
whoſoeuer wil not or dare not be tried in al their doctrine and doings,
they shew themſelues to miſtruſt their owne cauſe, and to flye from the
light, & ordinance of God. Without which order of appeaſing al
differences in faith and conſtructions of the Scriptures, the Church had
beene more defectual and inſufficient, then any Commonwealth or Societie
of men in the world: none of which euer wanteth good meanes to decide al
diſcordes and diſſenſion ariſing among the ſubiects & citizens of the
ſame.}
appointed that Paul and Barnabas ſhould goe vp, & certaine others of
\Var{the reſt,}{them}
to the Apoſtles and
\TNote{\G{πρεσβυτέρους}}
Prieſts vnto Hieruſalem, vpon this queſtion.

%%% o-2434
\V They therfore being brought on their way by the Church, paſſed
through Phœnice & Samaria, reporting the conuerſion of the Gentiles: and
they made great ioy to al the Brethren.

\V And when they were come to Hieruſalem, they were receiued of the
Church and of the Apoſtles and
\TNote{\G{πρεσβυτέρων}}
\SNote{Anciẽts here, & often in this chapter are the ſame that Prieſtes
\XRef{verſ.~2.}
as S.~Hiero. taketh it alſo
\XRef{1.~Pet.~5.}
& the greeke approueth, being alwaies one, \G{πρεσβυτέροι}, Prieſts
\Cite{Hiero. in 1.~ad Tit.}
&
\Cite{4.~ad Galat.}}
Ancients, declaring whatſoeuer God had done with them. \V And there
aroſe certaine of the hereſie of the Phariſees that beleeued, ſaying,
That they muſt be circumciſed; cõmanded alſo to keepe the law of
Moyſes. \V And the
\LNote{Apoſtles and Ancients.}{The Hereſies of our Proteſtants which
would haue al men to giue voice, or to be preſent in Councels, and of
others that would haue none but the holy or elect to be admitted, are
refuted by this example, where we ſee none but Apoſtles & Prieſts or
Ancients aſſembled to diſpute of the matter though many deuout people
were in the citie the ſame time.
\MNote{Of what perſons a Councel conſiſteth.}
Neither did euer any other in the Ancient Councels of the Church
aſſemble to debate and define the matter, but ſuch, though many other
for other cauſes be euer preſent. Secular men or women, be their guifts
neuer ſo great, can not be iudges in cauſes of faith and religion.
\CNote{\XRef{Deut.~17.}}
\Emph{If any thing}, ſaith God, \Emph{be hard and doubtful, thou shalt
come to the Prieſts of the Leuitical ſtocke, and thou shalt follow their
ſentence.} Againe,
\CNote{\XRef{Mal.~2,~7.}}
\Emph{The lippes of the Prieſt shal keepe knowledge, and the Law thou
shalt require of his mouth.} Againe,
\CNote{\XRef{Agga.~2,~12.}}
\Emph{Aske the Law of the Prieſt.} Much more muſt we referre al to our
Bishops and Paſtours, whom God hath placed in the regiment of the Church
with much larger priuiledge, then euer he did the old Prieſts ouer the
Synagogue: to whom it is ſaid,
\CNote{\XRef{Luc.~10,~16.}}
\Emph{He that deſpiſeth you deſpiſeth me.}
\MNote{A general Councel repreſenteth the whole Church.}
And it is to be noted that the Bishops ſo gathered in Councel, repreſẽt
the whole Church, haue the authouritie of the whole Church, & the Spirit
of God to protect them frõ errour, as the whole Church: SS.~Paul and
Barnabas come hither for the definition of the whole Church. \Emph{The
ſentence of a plenarie or general Councel} (ſaith
\CNote{\Cite{li.~1. c.~18. de bapt.}}
S.~Auguſtine) \Emph{is the conſent of the whole Church.} And ſo it muſt
needes be in the Church, becauſe the Magiſtrates, Senate, Coũcel or
deputies of al commonwealthes, repreſent the whole body: and to haue it
otherwiſe (as the Churches Rebels wish) were to bring al to hel and
horrour, and themſelues to be perpetually, by the ſeditious and popular
perſons vpholden againſt Law, reaſon, and religion, in their wickednes.}
Apoſtles and Ancients
\LNote{Aſſembled.}{A
\MNote{The firſt Councel at Hieruſalem.}
Councel was called to diſcuſſe the matter which Councel was the more
eaſily gathered, becauſe the Chriſtian Bishops and countries were not
yet ſo many but that the principal Gouernours of the Church being not
farre diſperſed, and as many learned men as were neceſſarie, might be in
Hieruſalem or eaſily called thither. And it was not a Prouincial Councel
or Synode only, but a general Councel, conſiſting of the cheefe Apoſtles
& Bishops 
that then were, though the number was nothing ſo great as afterward vſed
to aſſemble, when the Church was ſpred into al Nations.}
aſſembled to conſider of this word.

\V And when there was made a
\SNote{See the
\XRef{Annot. vers.~28. toward the end.}}
great diſputation,
\LNote{Peter riſing vp.}{S.~Peter
\MNote{Peter Head of the Councel, & his Succeſſours after him.}
as the Head of the Church ſpeaketh firſt, as his Succeſſours haue euer
had, not only in their perſonal preſence, but in their abſence by their
Legates and Subſtitutes, the cheefe voice in al Councels general, none
euer receiued into authoritie & credite in the Church without their
Confirmation. And therfore the Councels of the Arians and of other
Heretikes, were they neuer ſo great, wanting the Popes aſſent,
aſſiſtance, or confirmation, did shamfully erre, as Ariminenſe for the
Arians, and Epheſinum ſecundum for the Neſtorians, & ſuch like condemned
Aſſemblies.}
Peter riſing vp ſaid to them: Men Brethren, you know that
\CNote{\XRef{Act.~10,~20.}}
of old daies God among vs
\LNote{Choſe that by my mouth.}{Though Paul were called and appointed
ſpecially to be the Apoſtle of the Gentils,
\MNote{Peters preeminence both toward Iewes & Gentils.}
yet that was S.~Peters ſpecial priuiledge by Gods owne choiſe, that the
firſt Gentils should be called by his mouth, and that he firſt should
vtter to the Church that truth of the admiſſion of the Gentils himſelf,
for that he was Chriſtes Vicar, being notwithſtanding (as his Maiſter
was)
\CNote{\XRef{Ro.~15,~8.}
\XRef{Gal.~1,~7.}}
\L{Miniſter Circumciſionis}, that is, Apoſtle of the Iewes; Chriſt
deferring al preeminence vnto him in that point alſo.}
choſe, that by my mouth the Gentils ſhould heare the word of the Ghoſpel,
and beleeue. \V And God which knoweth the hartes, gaue teſtimonie,
\CNote{\XRef{Act.~10,~45.}}
giuing vnto them the Holy Ghoſt as wel as to vs, \V and hath put no
difference between vs and them,
\SNote{By that faith which worketh by charitie. For a dead faith cã not
purifie the hart of mã. See
\XRef{chap.~16,~31.}}
by faith purifying their hartes. \V Now therfore why tempt you God, to
put a yoke vpon the necks of the Diſciples, which neither our Fathers
nor we haue been able to beare? \V But by the grace of our
Lord \Sc{Iesvs Christ} we beleeue to be ſaued, in like manner as they
alſo.

\V And al the multitude held their peace: and they heard Barnabas and
Paul telling what great ſignes and wonders God had done among the
Gentils by them.

\V And after they held their peace,
\LNote{Iames.}{S.~Iames becauſe he was an Apoſtle and alſo Bishop of
Hieruſalem, gaue his ſentence next. For the ſpeach interpoſed of
SS.~Paul & Barnabas, was but for their better information in the
deciſion of the matter, & for confirmation of S.~Peters ſentence:
though they being Apoſtles, & Bishops, had voices in the councel alſo;
as many moe had, though their ſentences be not here reported.
\MNote{S.~Iames and the reſt follow S.~Peters ſentence.}
And where
S.~Iames in his ſpeach ſaith, \Emph{I iudge}, it is not meant that he
gaue the principal definitiue ſentence: for he (as al the reſt) followed
and allowed the ſentence of S.~Peter, as it is plaine in the text, the
whole aſſembly for reuerence of his perſon & approbation of his ſentence,
holding their peace. \Emph{Al the multitude}
(ſaith
\CNote{\Cite{Hier. to.~2. ep.~89. ad Aug. c.~2.}}
S.~Hierom) \Emph{held their peace, and into his ſentence Iames the
Apoſtle & al the Prieſts did paſſe together.} For though S.~Iames did
particulariſe certaine points incident to the queſtion debated, as of
eating ſtrangled meates &c. yet the proper controuerſie for which the
Councel aſſembled, was,
\MNote{The principal queſtion.}
Whether the Gentils conuerted were bound to obſerue the Law of Moyſes: &
it was concluded that they were not bound, nor ought to be charged with
Moyſes Law or the Sacraments & ceremonies of the ſame. This is the
ſubſtance and principal purpoſe of this Councels decree, which doth bind
for euer; and \Emph{Peter} (ſaith S.~Hierom
\Cite{in the ſame place})
was \Emph{Prince} or \Emph{Authour of this decree}
\MNote{Incident queſtions.}
the matter of
fornication & Idolothytes being but incident to the queſtion, or
reſolution; & the forbidding of eating ſtrangled & bloud, but a temporal
prohibition, which by the conſent of the Church or otherwiſe afterward was
abrogated; the Church of God hauing the true ſenſe of difference of
times, places, & perſons, when and how farre ſuch things are to be
obſerued, and when not.
\MNote{How later Councels alter the former.}
And in ſuch things as theſe, and in other like which according to
circumſtances require alteration, it is, that S.~Auguſtine ſaith,
\Cite{li.~2. de Bapt. c.~1. to.~7.}
\Emph{The former general or plenarie Councels may be amended by the
later.}}
Iames anſwered, ſaying: Men Brethren, heare me. \V Simon hath told how
God firſt viſited to take of the Gentils a people to his name. \V And to
this accord the wordes of the Prophets, as it is written: \V 
%%% !!! This first CNote appears only in Heretikes, and seems out of
%%% place. The second seems wrong (in other it is Amos.~11?). 
%%% \CNote{\XRef{Tim.~3,~5.}}
\CNote{\XRef{Amos.~7,~11.}}
\Emph{After theſe things I wil returne, and wil reedifie the tabernacle
of Dauid which was fallen, and the ruines thereof I wil reedifie, and
ſet it vp: \V that the reſidue of men may ſeeke after the Lord, and al
%%% 2589
Nations vpon whom my name is inuocated, ſaith the Lord that doth theſe
things.}

\V To our Lord was his owne worke knowen from the beginning of the
world. \V For the which cauſe
%%% !!! missing note? Marked in both.
%%% \LNote{}{}
I iudge, that they which of the Gentils are conuerted to God, are not to
be diſquieted, \V but to write vnto them that they refraine themſelues
from the contaminations of Idols, and
\LNote{Fornication.}{Fornication and contamination with Idols, are of
themſelues mortal ſinnes, and therfore can neuer be lawful: yet becauſe
the Gentils by cuſtome were prone to both, and of fornication made very
ſmal account, it pleaſed the Holy Ghoſt to forbid both ſpecially.
\MNote{The Churches authoritie in making Decrees.}
Concerning the other points of abſteining from bloud & ſtifled meates,
they were things of their owne nature indifferent, in which for a time
the Iewes were to be borne withal, and the Gentils to be a litle
exerciſed to obedience. By which we may ſee the great authoritie of
Gods Church and Councels, which may command for euer, or for a time,
ſuch things as be fit for the ſtate of times and Nations, without any
expreſſe Scriptures at al, and ſo by commandement make things neceſſarie
that were before indifferent.}
fornication, and ſtrangled things, and bloud. \V For Moyſes of old times
hath in euery citie them that preach him in the Synagogues, where he is
read euery Sabboth.

%%% o-2435
\V Then it pleaſed the Apoſtles and Ancients with the whole Church, to
chooſe men out of them, & to ſend to Antioche with Paul and Barnabas,
Iudas, who was ſurnamed Barſabas, & Silas, cheefe men among the
Brethren, \V
\SNote{Other Latin copies and the greeke read thus \Emph{writing by
their hands an epiſtle conteining theſe things.}}
writing by their hands.

The Apoſtles and Ancients, the Brethren, to the Brethren of the Gentils
that are at Antioche and in Syria and Cilicia, greeting. \V Becauſe we
haue heard that certaine
\LNote{Going forth from vs.}{A
\MNote{Going out a marke of heretikes.}
proper deſcription or note of Heretikes, Schiſmatikes, and ſeditious
Teachers to goe out from their ſpiritual Paſtours and Gouernours, & to
teach without their commiſſion and approbation, to diſquiet the
Catholike people with multitude of wordes and ſweet ſpeaches, and
finally to ouerthrow their ſoules.}
going forth from vs haue troubled you with wordes ſubuerting your
ſoules, to whom we gaue no commandement: \V It hath pleaſed vs being
gathered in one, to chooſe out men & to ſend them vnto you with our
deereſt Barnabas and Paul, \V men that haue giuen their liues for the
name of our Lord \Sc{Iesvs Christ}. \V We haue ſent therfore Iudas and
Silas who themſelues alſo wil in wordes report vnto you the ſame
things. \V For it hath ſeemed good
\LNote{To the Holy Ghoſt and to vs.}{By
\MNote{God & our Ladie, and the like ſpeaches.}
this firſt we note, that it is not ſuch a fault as the Heretikes would
make it in the ſight of the ſimple, or any incongruitie at al, to ioyne
God and his creatures, as the principal cauſe and the ſecondarie, in one
ſpeach, & to attribute that to both, which though diuerſly, yet
proceedeth of both. \Emph{God and you}, ſay good people
commonly: \Emph{God and our Ladie, Chriſt and S.~Iohn: We confeſſe to
God & to Peter & Paul}, as
\CNote{\XRef{Gen.~48,~15,~16.}
%%% !!! where do these go?
\XRef{Iud.~7,~18,~20.}
\XRef{Exo.~14,~31.}
\XRef{1.~Tim.~5,~21.}}
\Emph{God & his Angel, To our Lord & Gideon, The ſword of our Lord & of
Gideon, Our Lord and Moyſes, Chriſt and his Angels, Our Lord and al
Saints}.
\XRef{ep. ad Philem.}
\Emph{S.~Paul and our Lord}.
\XRef{2.~Theſ.~1,~6.}
Al theſe ſpeaches being partly Scriptures, partly like vnto the
Scriptures ſpeaches, are warranted alſo by this Councel, which ſaith
boldly, & hath giuen the forme thereof to al other Councels lawfully
called and confirmed, to ſay the like:
\MNote{\Sc{\L{Visvm est spiritvi sancto et nobis}}.}
\Emph{It hath pleaſed the Holy Ghoſt and vs.} S.~Cyprian
\Cite{ep.~54. nu.~2.}
reporting the like of a Synode holden in Afrike, ſaith: \Emph{It hath
pleaſed vs by the ſuggeſtion of the Holy Ghoſt.}

Secondly
%%% !!! This marginal note is split into two parts.
\MNote{The Holy Ghoſt's Aſsiſtance in al lawful
%%% !!! this part goes with the next paragraph?
Councels, to the worlds end, and that by Chriſts promiſe.}
we note, that the holy Councels lawfully kept for determination or
cleering of doubts, or condemning of errours & Hereſies, or appeaſing of
Schiſmes & troubles, or reformation of life, and ſuch like important
matters, haue euer the aſsiſtance of Gods Spirit, & therfore can not
erre in their Sentences & determinations concerning the ſame, becauſe
the Holy Ghoſt can not erre, from whom (as you ſee here) ioyntly with
the Councel the reſolution proceedeth.

Thirdly we learne, that in the Holy Councels ſpecially (though
otherwiſe & in other Tribunals of the Church it be alſo verified)
Chriſtes promiſe is fulfilled,
\CNote{\XRef{Io.~16,~13.}}
that the Holy Ghoſt ſhould ſuggeſt them and teach them al truth, &
that not in the Apoſtles time only, but to the worldes end. For ſo long
ſhal Councels, the Church, & her Paſtours haue this priuiledge of Gods
aſsiſtance, as there be either doubts to reſolue, or Heretikes to
condemne, or truths to be opened, or euil men to be reformed, or
Schiſmes to be appeaſed. For which cauſe
\MNote{S.~Gregories reuerence of General Councels.}
S.~Gregorie
\Cite{li.~1. ep.~24. ſub fin.}
reuerenceth the foure general Councels
(Nicen. Conſtantinop. Epheſ. Chalced.) as the foure books of the holy
Ghoſpel, alluding to the number: and of the fifth alſo he ſaith that he
doth reuerence it alike: and ſo would he haue done moe, if they had been
before his time, who ſaith of them thus: \Emph{Whiles they are concluded
and made by vniuerſal conſent, himſelf doth he deſtroy, and not them,
whoſoeuer preſumeth either to looſe whom they bind, or to bind whom they
looſe.}

S.~Gregorie
\MNote{The Proteſtãts fond diſtinctiõ between the 4.~firſt & the later
Councels.}
therfore reuerencing al fiue alike, it may be marueled whence the
Heretikes haue 
their fond difference betwixt thoſe foure firſt and other later:
attributing much to them, and nothing to the reſt. Whereas indeed the
later can erre no more then the firſt foure, being holden & approued as
they were, and hauing the Holy Ghoſt as they had. But in thoſe firſt alſo
when a man findeth any thing againſt their Hereſies (as there be diuers
things) then they ſay plainly that they alſo may erre, and that the Holy
Ghoſt is not tied to mens voices, nor to the number of ſentences. Which
is directly to reproue this firſt Councel alſo of the Apoſtles, and
Chriſtes promiſe of the Holy Ghoſts aſsiſtance to teach al truth. Yea
that you may know and abhorre theſe Heretikes throughly, heare ye what a
principal Sect Maiſter with his blaſphemous mouth or penne vttereth,
ſaying, 
\MNote{Beza's blaſphemie againſt the firſt general Councels.}
\CNote{\Cite{Beza in Præf. Teſt. No. an.~1565.}}
that \Emph{in the very beſt times ſuch was partly the ambition of
Bishops, partly their foolishnes and ignorance, that the very blind may
eaſily perceiue, Satan verily to haue been preſident of their
aſſemblies.} Good Lord deliuer the people and the world from ſuch
blaſphemous tongues and bookes, and giue men grace to attend to the holy
Scriptures and Doctours that they may ſee
\MNote{What the Fathers attribute to Councels, & namely S.~Auguſtine.}
how much, not only S.~Auguſtine and other Fathers attribute to al
general Councels ſpecially, to which they referre themſelues in al
doubts among themſelues and in al their controuerſies with Heretikes:
but to which euen S.~Paul himſelf (ſo ſpecially taught by God) and
others alſo yealded themſelues. Notorious is the ſaying of S.~Auguſtine
concerning S.~Cyprian, who being a Bleſſed Catholike Bishop and Martyr,
yet erred about the rebaptizing of ſuch as were Chriſtned by
Heretikes. \Emph{If he had liued} ſaith S.~Auguſtine
\Cite{li.~2. de bapt. c.~4.}
\Emph{to haue ſeen the determination of a plenarie Councel, which he ſaw
not in his life time, he would for his great humilitie and charitie
ſtraight way haue yealded, and preferred the general Councel before his
owne iudgement and his fellow Bishops in a Prouincial Councel only.}
\MNote{Prouincial Councels.}
Whereby alſo we learne, that Prouincial Councels may erre, though many
times they doe not, and being conformable to the general Councels, or
confirmed and allowed by them or the See Apoſtolike, their ſolutions be
infallible as the others are.

If
\MNote{Notwithſtanding the Holy Ghoſts aſsiſtance, yet humane meanes
muſt be vſed to ſearch the truth.}
any here aske, what need ſo much diſputing, ſtudy, and trauail in
Councels to find out and determine the truth, if the Holy Ghoſt
infallibly guide them? We anſwer that ſuch is the ordinarie prouidence
of God in this caſe, to aſsiſt them when they doe their endeauour, and
vſe al humane meanes of induſtrie, and not els. And ſo (though ſomewhat
otherwiſe) God aſsiſted the Euangeliſts & other Writers of the holy
Scriptures, that they could not erre in penning the ſame; but yet they
did and ought to vſe al poſsible humane diligence to know and learne out
the hiſtories and truth of matters, as is plaine in the
\CNote{\XRef{Lu.~1,~2.}}
beginning of S.~Lukes Ghoſpel: els the Holy Ghoſt would not haue
aſſiſted them. Euen ſo in this Councel of the Apoſtles, though they had
the Holy Ghoſt Aſsiſtant, yet the text ſaith, \L{cum magna conquisitio
fieret}, when there was great diſputation, ſearch and examination of the
caſe, then Peter ſpake &c.
\MNote{Though the See Apoſtolike itſelf haue the ſame aſsiſtãce, yet
Councels alſo neceſſarie for many cauſes.}
If againe it be demanded, what need is there to expect the Councels
determination, if the Popes or See Apoſtolikes iudgement be infallible
and haue the aſsiſtance of God alſo, as the Catholikes affirme? We
anſwer, that for the Catholike and peaceable obedient children of the
Church it is a comfort to haue ſuch various meanes of determination,
trail, and declaration of the truth; and that it is neceſſarie for the
recouery of Heretikes, and for the contentation of the weake, who not
alwaies giuing ouer to one mans determination, yet wil either yeald to
the iudgement of al the learned men & Bishops of al Nations, or els
remaine deſperate and condemned before God and man for euer. And as I
ſaid before, this aſsiſtance of the Holy Ghoſt promiſed to Peters See,
preſuppoſeth humane meanes of ſearching out the truth, which the Pope
alwaies hath vſed, & wil, & muſt vſe in matters of great importance, by
calling Councels: euen as here you ſee SS.~Peter & Paul themſelues and
al the Apoſtles, though indowed with the Holy Ghoſt, yet thought it
notwithſtanding neceſſarie for further trial & cleering of truth and
maintenance of vnitie, to keepe a Councel.

Laſtly it is to be noted, that as Chriſt and the Holy Ghoſt be preſent
by his promiſe, to al ſuch aſſemblies as gather in the obedience & vnitie of
the Church, with ful mind to obey whatſoeuer shal be determined, whereby
the aſſembled though of diuers iudgements before, doe moſt peaceable
yeald to truth, and agree in one vniforme determination of the ſame:
\MNote{Heretical or Schiſmatical aſſemblies.}
ſo al ſuch as gather out of the Church, without humilitie or intention
to yeald one to
another, or to any Superiour, man or Councel, or what els ſoeuer, but
chalenge to themſelues learning, ſpirit, & we can not tel what; ſuch,
how many meetings ſoeuer they make, being deſtitute of the Holy Ghoſt
the Authour of truth and concord, are further off & further out, then
euer before: as God hath shewed by the ſucceſſe of al Heretical
Colloquies, Synodes, and Aſſemblies in Germanie, France, Poole, and
other places in our daies. Read a notable place in
\CNote{\Cite{de vnit. Ec. nu.~7.}}
Saint Cyprian, that the promiſe of Chriſt, that he would be in the
middes of two or three gathered in his name, pertaineth not to them that
aſſemble out of the Church.}
to the Holy Ghoſt and to vs, to lay no further burden vpon you then theſe
neceſſarie things: \V That you abſteine frõ the things immolated to
Idols, and bloud, and that which is ſtrangled, and fornication, from the
which things keeping your ſelues, you ſhal doe wel. Fare ye wel.

\V They therfore being dimiſſed went downe to Antioche: and gathering
the multitude, deliuered the epiſtle. \V Which when they had read, they
\LNote{Reioyced vpon the conſolation.}{Straight
\MNote{Al good Chriſtians reſt vpon the determination of a general
Councel.}
vpon the intelligence of the Councels determination, not only the
Gentils, but euen the Maiſters of the former troubles and diſſenſion,
were at reſt; & al tooke great comfort that the controuerſie was ſo
ended. And ſo ſhould al Chriſtian men doe, when they ſee the Sectes of
our time condemned by the like authoritie and moſt graue iudgement of
the holy Councel of Trent. Againſt which the Heretikes of our time make
the like friuolous exceptions and falſe cauillations, as did the old
Heretikes heretofore againſt thoſe Councels that ſpecially condemned
their errours.
\MNote{Al Heretikes make exceptiõ againſt the Councels that condemne
them.}
The Pope and Bishops (ſay they) are a partie, and they
ought not to be our iudges: they are partial and come with preindicate
mindes to condemne vs; and we accuſe them al of Idolatrie and other
crimes, and we wil be tried by Gods word only, and we wil expound it
according to another rule, that is to ſay, as we liſt. So they againſt
this Councel, and the like ſaid the Arians againſt the firſt Nicene
Councel, and al ſuch like againſt thoſe Councels namely that condemned
their hereſies. And ſo ſay al theeues againſt their correctours and
punishers, & would both ſay and doe more againſt temporal tribunals,
Iudges, Iuſtices, and iuries, if they had as much licence and libertie
in thoſe matters, as men haue now in religion.}
reioyced vpon the conſolation: \V but Iudas and Silas, themſelues alſo
being Prophets, with many wordes comforted the Brethren, and confirmed
them. \V And hauing ſpent ſome time there, they were with peace dimiſſed
of the Brethren vnto them that had ſent them. \V But it ſeemed good vnto
Silas to remaine there: and Iudas departed alone: \V and Paul and
Barnabas taried at Antioche, teaching and euangelizing with many others
the word of our Lord.

\V And after certaine daies, Paul ſaid to Barnabas, let vs returne and
\SNote{Hereof our Catholike Bishops tooke vp the neceſſarie vſe of oftẽ
viſiting their flockes & cures cõmitted to their charge, for cõfirmation
in faith & vertue, & reformation of mãners both of clergie and laitie.}
viſite our Brethren in al cities wherein we
\CNote{\XRef{Act.~13,~14.}}
haue preached the word of our Lord, how they doe. \V And Barnabas
would haue taken with them Iohn alſo that was ſurnamed Marke. \V But
Paul deſired that he (as who
\CNote{\XRef{Act.~13,~13.}}
had departed from them out of Pamphylia, & had not gone with them to the
worke) might not be receiued. \V And there roſe a
\LNote{Diſſenſion.}{Such occaſion of differences fal out euen among the
perfect men often, without any great offence. And this their departing
fel out to the great increaſe of Chriſtians. And therfore it is very
ridiculouſly applied to excuſe the diſagreeing of the Heretikes among
themſelues in the principal pointes of religion, namely the Sacrament.}
diſſention, ſo that they departed one from another, & that, Barnabas
indeed taking Marke ſailed to
\Fix{Cypers.}{Cypres.}{obvious typo, fixed in other}
\V But
%%% o-2436
Paul chooſing Silas departed, being deliuered of the Brethren to the
grace of God.

\V And he walked through Syria and Cilicia, confirming the Churches
\SNote{Not only the things cõmanded by Chriſtes expreſſe word, or writtẽ
in the Scriptures (as our Heretikes hold) but whatſoeuer the Apoſtles
and Rulers of the Church command, is to be kept & obeied. See theſe
wordes repeated againe
\XRef{c.~16,~4.}
& that ĩ the greeke leſt any man cauil, becauſe here the greeke hath
them not.}
commanding them to keepe the precepts of the Apoſtles & the Ancients.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

