%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 26
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-26


\project douay-rheims


%%% 2615
%%% o-2465
\startChapter[
  title={Chapter 26}
  ]

\Summary{In that honourable Audience being permitted to ſpeake, 2.~he
  declareth to the King what he firſt was, 12.~and how miraculouſly he
  was conuerted, 19.~and that he hath preached ſince, as he was
  commanded from Heauen, and as the Prophets had foretold of
  Chriſt. 24.~Which ſtrange tale Feſtus the Heathen Preſident hearing,
  ſaith that he is mad. 25.~But Paul anſwereth, & exhorteth them al to be
  Chriſtiãs as he is. 30.~They finally pronounce that he might be ſet at
  libertie, but only for his appeale.}

Bvt Agrippa ſaid to Paul: Thou art permitted to ſpeake for thy
ſelf. Then Paul ſtretching forth his hand, began to make his anſwer:

\V Touching al things whereof I am accuſed of the Iewes, King Agrippa, I
account my ſelf happie for that I am to defend my ſelf this day before
thee, \V eſpecially whereas thou knoweſt al things that are among the
Iewes, cuſtomes and queſtions: for the which cauſe I beſeech thee,
heare me patiently. \V And my life truely from my youth, which was from
the beginning in my Nation in Hieruſalem, al the Iewes doe know: \V
knowing me before from the beginning (if they wil giue teſtimonie) that
according to the moſt ſure
\TNote{\G{αἵρεσιν}}
ſect of our religion I liued a Phariſee.
%%% 2616
\V And now for the hope of the promiſe that was made of God to our
Fathers, doe I ſtand ſubiect to iudgement: \V
\TNote{\L{in quã}}
the which, our twelue Tribes
\TNote{\G{λατρεῦον}}
ſeruing night and day, hope to come vnto. Of the which hope, o King, I
am accuſed of the Iewes. \V What incredible thing is it iudged with you,
if God raiſe the dead? \V And my ſelf truely had thought that I ought to
doe againſt the name of \Sc{Iesvs} of Nazareth many contrarie things. \V
Which alſo I
\CNote{\XRef{Act.~8,~3.}}
did at Hieruſalem, and many of the Saints did I ſhut vp in priſons,
hauing receiued authoritie of the cheefe Prieſts: and when they were put
to death,
\TNote{\L{detuli sententiam}}
I brought the ſentence. \V And throughout al the Synagogues oftentimes
puniſhing them, I compelled them to blaſpheme:
%%% o-2466
and yet more mad againſt them I perſecuted them euen vnto forraine
cities. \V Among which things whiles
\CNote{\XRef{Act.~9,~3.}}
I went to Damaſcus with authoritie and permiſſion of the cheefe
Prieſts, \V at mid-day, in the way, I ſaw (o King) from Heauen a light
to haue ſhined round about me and them that were in companie with me,
aboue the brightnes of the ſunne. \V And when al we were fallen downe on
the ground, I heard a voice ſpeaking to me in the Hebrew tongue: Saul,
Saul, why perſecuteſt thou me? It is hard for thee to kicke againſt the
pricke. \V And I ſaid: Who art thou Lord? And our Lord anſwered: I
am \Sc{Iesvs} whom thou doeſt perſecute. \V But riſe vp and ſtand vpon
thy feet: for to this end haue I appeared to thee, that I may ordaine
thee a Miniſter and witnes of thoſe things which thou haſt ſeen, and of
thoſe things wherein I wil appeare to thee, \V deliuering thee out of
the peoples and Nations vnto the which now I ſend thee, \V to open their
eyes, that they may be conuerted from darkenes to light, and from the
power of Satan to God, that they may receiue remiſſion of ſinnes, and
lot among the Saints by the faith that is in me. \V Whereupon, King
Agrippa, I was not incredulous to the heauenly viſion: \V but to them
firſt that are at Damaſcus, and at Hieruſalem, and vnto al the countrie
of Iewrie, and to the Gentils
\CNote{\XRef{Act.~9,~20.}}
did I preach that they ſhould doe
\SNote{Penãce often inculcated, and workes agreable to the ſame.}
penance, & turne to God, doing workes worthie of penance. \V For this
cauſe the Iewes, when I was in the Temple,
\CNote{\XRef{Act.~21,~30.}}
apprehending me, attempted meaning to kil me. \V But aided by the help
of God, I ſtand vntil this day, teſtifying to ſmal & to great, ſaying
nothing beſide thoſe things which the Prophets did ſpeake ſhould come to
paſſe, & Moyſes; \V if \Sc{Christ} were paſſible, if the firſt of the
reſurrection from the dead, he were to ſhew light to the people and to
the Gentils.

\V As he ſpake theſe things and made his anſwer, Feſtus with a loud
voice ſaid: Thou art mad, Paul; much learning turneth thee to
madneſſe. \V And Paul ſaid: I am not mad, moſt excellent Feſtus; but I
ſpeake words of veritie & ſobrietie. \V For the King knoweth of theſe
things, to whom alſo I ſpeake conſtantly. For I thinke none of theſe
things to be vnknowen to him. For neither was any of theſe things done
in a corner. \V Beleeueſt thou the Prophets, King Agrippa? I know that
thou beleeueſt. \V And Agrippa ſaid to Paul:
%%% o-2467
A litle thou perſuadeſt me to become a \Sc{Christian}. \V And Paul ſaid:
I wiſh of God, both in litle,
%%% 2617
and in much, not only thee, but alſo al that heare this day, to become
ſuch as I am alſo, except theſe bands. \V And the King roſe vp, and the
Preſident, and Bernice, and they that ſate by them. \V And going aſide,
they ſpake among themſelues, ſaying: That this man hath done nothing
worthie of death or bands. \V And Agrippa ſaid to Feſtus: This man might
be releaſed, if he had not appealed to Cæſar.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

