%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 2557
%%% o-2398
\startChapter[
  title={Chapter 4}
  ]

\Summary{The Rulers of the Iewes oppoſe themſelues and impriſon Peter &
  Iohn. 4.~But yet thouſands of the people are conuerted: 5.~and to the
  Rulers alſo Peter boldly auoucheth by the foreſaid miracle,
  that \Sc{Iesvs} is Chriſt, telling them of their heinous fault out of
  the Pſalmes, & that without him they can not be ſaued. 13.~They though
  confounded with the miracle, yet proceed in their obſtinacie,
  forbidding them to ſpeake any more of \Sc{Iesvs}, adding alſo
  threates. 23.~Whereupon the Church flyeth to praier, wherein they
  comfort themſelues with the omnipotencie of God, and prediction of
  Dauid, and aske for the guift of boldnes and miracles againſt thoſe
  threates. 31.~And God sheweth miraculouſly that he hath heard their
  praier. 32.~The whole Churches vnitie and communitie of life. 36.~Of
  Barnabas by name.}

And when they
\Fix{where}{were}{obvious typo, fixed in other}
ſpeaking to the people, the Prieſts and Magiſtrates of the Temple & the
Sadducees came vpon them, \V being greeued that they taught the people,
and ſhewed in \Sc{Iesvs} the reſurrection from the dead: \V and they
laid handes vpon them, and put them into ward, vntil the morow, for it
was now euening. \V And
\SNote{Here againe we ſee the proceeding & increaſe of the Church
viſibly.}
many of them that had heard the word, beleeued: and the number of the
men was made fiue thouſand.

%%% o-2399
\V And it came to paſſe on the morow, that their Princes, and Ancients,
and Scribes were gathered into Hieruſalem. \V And Annas the high Prieſt,
and Caiphas and Iohn, and Alexander, and as many as were of the Prieſts
Stocke. \V And ſetting them in the middes, they asked: In what power or
in what name haue you done this? \V Then Peter repleniſhed with the Holy
Ghoſt, ſaid to them: Ye Princes of the people and Ancients: \V If we
this day be examined for a good deed vpon an impotent man, in what he
hath been made whole, \V be it knowen to al you and to al the people of
Iſrael, that 
\MNote{The name of \Sc{Iesvs}. See
\XRef{Annot. Philip.~2,~10.}}
in the name of \Sc{Iesvs Christ} of Nazareth, whom you did crucifie,
whom God hath raiſed from the dead, in this ſame this man ſtandeth
before you whole. \V This is
\CNote{\XRef{Pſ.~117,~22.}}
\Emph{the ſtone that was reiected} of you \Emph{the builders: which is
made into the head of the corner}: \V And there is not ſaluation in any
other. For neither is there any other name vnder Heauen giuen to men,
wherin we muſt be ſaued. \V
\MNote{The Apoſtles conſtãcie, learning, & wiſedõ after the comming of
the Holy Ghoſt, being but idiotes, that is, ſimple vnlettered men &
timorous before.}
And ſeeing Peters conſtancie and Iohns, vnderſtanding that they were men
vnlettered, and of the vulgar ſort, they marueled; and they knew them
that they had been with \Sc{Iesvs}: \V ſeeing the man alſo that had been
cured, ſtanding with them, they could ſay nothing to the contrarie. \V
But they commanded them to goe aſide forth out of the Councel: and they
conferred together, \V ſaying: What ſhal we doe to theſe men? for a
notorious ſigne indeed hath been done by them, to al the inhabitants of
Hieruſalem: it is manifeſt, and we can not denie it. \V But that it be
no further ſpred abrode among the people, let vs threaten them, that
they ſpeake no more in this name to any man. \V And calling them, they
charged them that they ſhould not ſpeake
%%% 2558
at al, nor teach in the name of \Sc{Iesvs}. \V But Peter and Iohn
anſwering, ſaid to them:
\SNote{Marke their conſtancie and courage after their confirmation,
being ſo weake before. And if any Magiſtrate command againſt God, that
is to ſay, forbid Catholike Chriſtiã mẽ to preach or ſerue God, this
ſame muſt be their anſwer; though they be whipped and killed for their
labour. See
\XRef{c.~5. v.~19.}}
If it be iuſt in the ſight of God, to heare you rather then God, iudge
ye. \V For we can not but ſpeake the things which we haue ſeen and
heard. \V But they threatning, diſmiſſed them: not finding how they
might puniſh them, for the people, becauſe al glorified that which had
been done, in that which was
\Fix{chanched.}{chanced.}{obvious typo, fixed in other}
\V For the man was more then fourtie yeares old in whom that ſigne of
health had been wrought.

\V And being diſmiſſed they came to theirs, and ſhewed al that the
cheefe Prieſts and Ancients had ſaid to them. \V Who hauing heard it,
with one accord lifted vp their voice
%%% o-2400
to God, and ſaid: Lord, thou that
didſt make Heauen & earth, the ſea, and al things that are in them, \V
who in the Holy Ghoſt by the mouth of our Father Dauid thy ſeruant haſt
ſaid:
\CNote{\XRef{Pſ.~2,~1.}}
\Emph{Why did the Gentils rage, and the people meditate vaine
thinges: \V the Kings of the earth ſtand vp, and the Princes aſſemble
together againſt our Lord, and againſt his} \Sc{Christ}? \V For there
aſſembled indeed in this citie againſt thy holy Child \Sc{Iesvs} whom
thou haſt anointed, Herod, and Pontius Pilate, with the Gentils and the
people of Iſrael, \V to doe what
\SNote{Chriſtes death, as needful for mãs redemption, was of Gods
determination: but as of the malice of the Iewes, it was not his act
otherwiſe then by permiſsion.}
thy hand and thy counſel decreed to be done. \V And now Lord looke vpon
their threatnings, and giue vnto thy ſeruants with al confidence to
ſpeake thy word, \V in that, that thou extend thy hand to cures & ſignes
and wonders to be done by thy name of thy holy Sonne \Sc{Iesvs}. \V And
when they had praied, the place was moued wherein they were gathered:
and they were al repleniſhed with the Holy Ghoſt, and they ſpake the
word of God with confidence.

\V And the multitude of beleeuers had one hart and one ſoule: neither
did any one ſay that ought was his owne of thoſe things which he poſſeſſed,
but
\CNote{\XRef{Act.~2,~44.}}
al things were cõmon vnto thẽ. \V And with great power did the Apoſtles
giue teſtimonie of the reſurrection of \Sc{Iesvs Christ} our Lord: and
great grace was in al them. \V For neither was there any one needie
among them. For as many as were owners of landes or houſes
\SNote{Note the ardent charitie & contempt of
\Fix{wordly}{worldly}{obvious typo, fixed in other}
things ĩ the firſt Chriſtians: who did not only giue great almes, but
ſold al their land to beſtow on the Apoſtles & the reſt that were in
neceſsitie, according to Chriſtes counſel.
\XRef{Mt.~16,~21.}
Note alſo the great honour & credit giuen to the Apoſtles, in that the
Chriſtian men put al the goods & poſſeſsions they had, to their
diſpoſition.}
ſold and brought the prices of thoſe things which they ſold, \V and laid
it before the feete of the Apoſtles. And to euery one was deuided
according as euery one had need. \V And Ioſeph who was ſurnamed of the
Apoſtles Barnabas (which is by interpretation, the ſonne of conſolation)
a Leuite, a Cyprian borne, \V whereas he had a peece of land, ſold it,
and brought the price, and laid it
\LNote{Before the feete.}{He,
\MNote{Reuerence to holy perſons.}
as the reſt, did not only giue his goods
as in vulgar almes, but in al humble and reuerent maner as things
dedicated to God, he laid them downe at the Apoſtles holy feete, as
S.~Luke alwaies expreſſeth, and gaue them not into their handes. The
\CNote{\XRef{4.~Reg.~4,~27.}}
Sunamite fel downe and embraced Eliſæus feete. Many that asked benefites
of Chriſt (as the
\CNote{\XRef{Luc.~8,~47.}}
woman ſicke of the bloudy fluxe) fel downe at his feete &
\CNote{\XRef{Luc.~7,~38.}}
Marie kiſſed his feet.
\MNote{Kiſsing the feet of holy perſons.}
Such are ſignes of due reuerence done both to Chriſt and to other
ſacred perſons either Prophetes, Apoſtles, Popes, or others repreſenting
his perſon in earth. See in S.~Hierom of Epiphanius Biſhop in Cypres, how
the people of Hieruſalem \Emph{of al ſortes flocked together vnto him,
offering their children} (to take his bleſsing) \Emph{kiſsing his feet,
plucking the hemmes of his garment, ſo that he could not moue for the
throng.}
\Cite{Ep.~61. c. cont. erro. Io. Hieroſol.}}
before the feete of the Apoſtles.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

