%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 2594
%%% o-2439
\startChapter[
  title={Chapter 16}
  ]

\Summary{Paul hauing for his part viſited the Churches of Syria,
  Cilicia, and Lycaonia, deliuering vnto them withal to keep the Decrees
  of the Councel; 6.~beginneth a new iourney, ouer Phrygia, Galatia,
  Myſia: 8.~Yea into Europe alſo he paſſeth, admonished by a viſion, and
  commeth into Macedonia, 12.~and there he beginneth the Church of the
  Philippians, working miracles, and ſuffering perſecution.}

And he came to Derbe and Lyſtra. And behold, there was a certaine
Diſciple there named Timothee, the ſonne of
\Var{a widow}{Iew}
woman that beleeued, of a father a Gentile. \V To this man the Brethren
that were in Lyſtra and Iconium, gaue a good teſtimonie. \V Him Paul
would haue to goe forth with him: and taking him he circumciſed him
becauſe of the Iewes that were in thoſe places. For they al knew that
his father was a Gentile.

\V And when they paſſed through the cities, they deliuered vnto them to
keepe the
\SNote{Here again they take order that the decrees and articles of faith
agreed vpon in the Councel of
\Fix{Ieruſalẽ,}{Hieruſalem,}{likely typo, same in both}
ſhould be executed & obſerued. Whereby we ſee both the great authoritie
of Councels, & the diligence that al Prelates ought to haue to ſee the
Decrees & Canons of the Councels put in executiõ.}
decrees that were decreed of the Apoſtles and Ancients which were at
Hieruſalem. \V And the Churches were confirmed in faith, and did abound
in number daily.

%%% o-2440
\V And paſsing through Phrygia and the countrie of Galatia, they were
\SNote{This people had not the Ghoſpel denied vnto them altogether, but
for a time: becauſe (as Venerable Bede thinketh) God foreſaw they would
not beleeue, & ſo should haue beẽ more grieuouſly damned.}
forbidden by the Holy Ghoſt to preach the word in Aſia. \V And when they
were come into Myſia, they attempted to goe into Bithynia; and the
Spirit of \Sc{Iesvs} permitted them not.

\V And when they had paſſed through Myſia, they went downe to Troas: \V
and a viſion by night was ſhewed to Paul. There was a certaine man of
Macedonia ſtanding and beſeeching him, and ſaying: Paſſe into Macedonia,
and help vs. \V And as ſoone as he had ſeen the viſion, forthwith we
ſought to goe into Macedonia, being aſſured that God had called vs to
euangelize to them. \V And ſailing from Troas, we came with a ſtraight
courſe to Samothracia, and the day following to
\Fix{Neapopolis:}{Neapolis:}{obvious typo, fixed in other}
\V and from thence to Philippi, which is the firſt citie of the part of
Macedonia, a
\SNote{\L{Colonia}, is ſuch a citie where the moſt
\Fix{inhabitans}{inhabitants}{obvious typo, fixed in other}
are ſtrangers, ſent thither from other great cities & States, namely
from the Romanes.}
\L{Colonia}. And we were in this citie certaine daies abiding. \V And
vpon the day of the Sabboths, we went forth without the gate beſide a
riuer, where it ſeemed that there was a praier: & ſitting we ſpake to
the women that were aſſembled. \V And a certaine woman named Lydia, a
ſeller of purple of the citie of the Thyatirians, one that worſhipped
God, did heare: whoſe hart our Lord opened to attend to thoſe things
which were ſaid of Paul. \V And when ſhe was baptized, & her houſe, ſhe
beſought vs, ſaying: If you haue iudged me to be faithful to our Lord,
enter into my houſe & tarie. And ſhe conſtrained vs. \V And it came to
paſſe as we went to praier, a certaine wench hauing a Pythonical ſpirit,
met vs, that brought great gaine to her Maiſters by diuining. \V This
ſame following Paul & vs, cried ſaying:
\SNote{Either the Diuel was compelled by the vertue of Paul's preſence
to ſay truth, or els (as ſuch doe oftentimes) he ſpake truth now, that
they might the more truſt him, and he better beguile them at other
times.}
Theſe mẽ are the ſeruãts of the high God, which preach vnto you the way
of ſaluatiõ. \V And
%%% 2595
this ſhe did many daies. And Paul being ſorie, and turning, ſaid to the
ſpirit: I command thee in the name of \Sc{Iesvs Christ} to goe out from
her. And he went out the ſame houre. \V But her Maiſters ſeeing that the
hope of their gaine was gone, apprehending Paul and Silas, brought them
into the market place to the Princes: \V and preſenting them to the
Magiſtrates, they ſaid: Theſe men trouble our citie, being Iewes: \V and
they preach a faſion which it is not lawful for vs to receiue, nor doe,
being Romanes. \V And the people ranne againſt them: and the Magiſtrates
tearing their coates, commanded
%%% o-2441
them to be beaten with roddes. \V And when they had laid
\CNote{\XRef{2.~Cor.~11.}}
many ſtripes vpon them, they did caſt them into priſon, commanding the
keeper that he ſhould keep them diligently. \V Who when he had receiued
ſuch commandement, caſt them into the inner priſon, and made their feet
faſt in the ſtocks. \V And at mid-night Paul and Silas praying, did
praiſe God. And they that were in priſon, heard them. \V But ſodenly
there was made a great earthquake, ſo that the foundations of the priſon
were ſhaken. And forthwith al the doores were opened: and the bands of
al were looſed. \V And the keeper of the priſon waked out of his ſleep,
and ſeeing the doores of the priſon opened, drawing out his ſword, would
haue killed himſelf, ſuppoſing that the priſonners had been fled. \V But Paul
cried with a loud voice, ſaying: Doe thy ſelf no harme, for we are al
here. \V And calling for light, he went in, and trembling fel downe to
Paul and Silas at their feet: \V and bringing them forth, he ſaid:
Maiſters, what muſt I doe that I may be ſaued? \V But they ſaid:
\SNote{It is no other faith that ſaueth but that which worketh by
Charitie.
\Cite{Aug. Enchirid. c.~67.}}
Beleeue in our Lord \Sc{Iesvs}; and thou ſhalt be ſaued and thy
houſe. \V And they preached the word of our Lord to him with al that
were in his houſe. \V And he taking them in the ſame houre of the night,
\SNote{Happie Gailers that doe mercie toward their godly priſonners, and
receiue againe by them ſuch ſpiritual benefites.}
waſhed their wounds: and himſelf was baptized and al his houſe
incontinent. \V And when he had brought them into his owne houſe, he
laid the table for them, and reioyced with al his houſe, beleeuing
God. \V And when day was come, the Magiſtrates ſent the Sergeants,
ſaying: Let thoſe men goe.  \V And the keeper of the priſon told theſe
words to Paul, That the Magiſtrates haue ſent that you ſhould be let
goe. Now therfore departing, goe ye in peace. \V But Paul ſaid to them:
Being whipped openly, vncondemned, men that are Romanes, they haue caſt
vs into priſon: & now doe they ſend vs out ſecretly? Not ſo, but let
them come, & let vs out themſelues. \V And the Sergeants reported theſe
wordes to the Magiſtrates. And they were afraid hearing that they were
Romanes: \V and comming they beſought them, & bringing thẽ forth they
deſired thẽ to depart out of the citie. \V And going out of the priſon,
they entred in vnto Lydia: and hauing ſeen the Brethren, they comforted
them, and departed.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

