%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-07


\project douay-rheims


%%% 2563
%%% o-2406
\startChapter[
  title={Chapter 7}
  ]

\Summary{Steuen being permitted to anſwer, beginning at Abraham, sheweth
  that God was with their Fathers both in other places, and alſo long
  before the Temple, 48.~and that after it was built, it could not be
  (as they groſly imagined) a houſe for God to dwel in. 51.~Then he
  inueigheth againſt their ſtifneckednes, and telleth them boldly of
  their traiterous murdering of Chriſt, as their Fathers had done his
  Prophets afore him. 54.~Whereat they being wood, he ſeeth Heauen
  open, and \Sc{Iesvs} there in his Diuine Maieſtie. 57.~Whereat they
  become more mad, ſo that they ſtone him to death (Saul conſenting) he
  commending his ſoul to \Sc{Iesvs}, & humbly praying for them.}

And the cheefe Prieſt ſaid: Are theſe things ſo? \V Who ſaid: Ye men,
Brethren and Fathers, heare. The God of glorie appeared to our Father
Abraham when he was in Meſopotamia, before that he abode in Charan, \V &
ſaid to him:
\CNote{\XRef{Gen.~12,~1.}}
\Emph{Goe forth out of thy countrie, and out of thy kinred, & come into
a land that I shal shew thee.} \V Then
%%% 2564
went he forth out of the land of the Chaldees, and dwelt in Charan. And
from thence, after his father was dead, he tranſlated him into this land
wherein you doe now dwel. \V And he gaue him no inheritance in it, no
not the paſe of a foot: and he promiſed to giue it him in poſſeſſion,
and to his ſeed after him, when as he had no child. \V And God ſpake to
him:
\CNote{\XRef{Gen.~15,~13.}}
\Emph{That his ſeed shal be a ſeiourner in a ſtrange countrie, and they
shal ſubdue them to ſeruitude, and shal euil intreat them foure hundred
yeares: \V and the Nation which they shal ſerue, wil I iudge}, ſaid
God. \Emph{And after theſe things they shal goe forth and
\TNote{\G{λατρεύσευσι}}
shal ſerue me in this place.}

\V And he
\CNote{\XRef{Gen.~17.}}
gaue him the teſtament of circumciſion, and ſo he
\CNote{\XRef{Gen.~21.}}
begat Iſaac, and circumciſed him the eight day: and
\CNote{\XRef{Gen.~25.}}
Iſaac, Iacob: and
\CNote{\XRef{Gen.~29.}
\XRef{30.}
\XRef{35.}}
Iacob, the twelue Patriarches. \V And the Patriarches through emulation,
\CNote{\XRef{Gen.~37.}}
ſold Ioſeph into Ægypt. And God was with him: \V and deliuered him out
of al his tribulations. And he
\CNote{\XRef{Gen.~41.}}
gaue him grace and wiſedom in the ſight of Pharao the King of Ægypt, and
he appointed him Gouernour ouer Ægypt and ouer al his houſe. \V And
there came famin vpon al Ægypt and Chanaan, and great tribulation: and
our Fathers found no victuals. \V But when
\CNote{\XRef{Gen.~42.}}
Iacob had heard that there was corne in Ægypt; he ſent our
%%% o-2407
Fathers firſt: \V and at the
\CNote{\XRef{Gen.~45.}}
ſecond time Ioſeph was knowen of his brethren, and his kinred was made
knowen vnto Pharao. \V And Ioſeph ſending, called thither Iacob his
father and al his kinred in ſeuentie fiue ſoules. \V And
\CNote{\XRef{Gen.~46.}}
Iacob deſcended into Ægypt: and
\CNote{\XRef{Gen.~49.}}
he died, and our Fathers. \V And they were
\SNote{Tranſlation of Saints bodies agreable to nature and
Scripture. And the deſire to be buried in one place more then an other
(which the holy Patriarches alſo had 
\XRef{Gen.~49,~29.}
\XRef{Gen.~50,~24.}
\XRef{Hebr.~11,~22.})
hath ſometime great cauſes.
\Cite{Aug. de Cur. pro mort. c.~1. &~vlt.}}
tranſlated into Sichem, and were
\CNote{\XRef{Gen.~50.}}
laid in the ſepulchre that Abraham
\CNote{\XRef{Gen.~23.}
\XRef{Ios.~24.}}
bought for a price of ſiluer of the ſonnes of Hemor the ſonne of
Sichem.

\V And when the time drew neere of the promiſe which God had promiſed to
Abraham, the people
\CNote{\XRef{Exod.~1,~7.}}
increaſed and was multiplied in Ægypt, \V vntil an other King aroſe in
Ægypt, that knew not Ioſeph. \V This ſame circumuenting our ſtocke,
afflicted our Fathers, that they ſhould expoſe their children, to the
end they might not be kept aliue. \V The ſame time was
\CNote{\XRef{Exod.~2,~2.}}
Moyſes borne, and he was acceptable to God; who was nouriſhed three
moneths in his fathers houſe. \V And when he was expoſed, Pharaos
daughter tooke him vp, and nouriſhed him for her owne ſonne. \V And Moyſes
was inſtructed in al the wiſedom of the Ægyptians: and he was mightie in
his wordes and workes. \V And
\CNote{\XRef{Exo.~2,~11.}}
when he was fully of the age of fourtie yeares, it came to his mind to
viſite his brethren the children of Iſraël. \V And when he had ſeen one
ſuffer wrong, he defended him: and ſtriking the Ægyptian, he reuenged
his quarel that ſuſteined the wrong. \V And he thought that his brethren
did vnderſtand that God by his hand would ſaue them: but they vnderſtood
it not. \V And the day following
\CNote{\XRef{Exo.~2,~13.}}
he appeared to them being at ſtrife: and he reconciled them vnto peace,
ſaying: Men, ye are brethren, wherfore hurt you one another? \V But he
that did the iniurie to his neighbour, repelled him, ſaying: \Emph{Who
hath appointed thee prince and iudge ouer vs: \V What, wilt thou kil me,
as thou didſt yeſterday kil the Ægyptian?} \V And Moyſes fled vpon this
word: and he became a ſeiourner in the land of Madian, where he begat
two ſonnes. \V And after fourtie yeares were expired, there
\CNote{\XRef{Exod.~3,~2.}}
appeared to him in the deſert of mount Sina an Angel in the
%%% 2565
fire of the flame of a buſh. \V And Moyſes ſeeing it, marueled at the
viſion. And as he went neere to view it, the voice of our Lord was made
to him: \V \Emph{I am the God of thy fathers; the God of Abraham, the
God of Iſaac, and the God of Iacob.} And Moyſes being made to tremble,
durſt not view it. \V And our Lord
%%% o-2408
ſaid to him: \Emph{Looſe of the shoe of thy feet: for the place wherein
thou ſtandeſt, is
\LNote{Holy ground.}{If that apparition of God himſelf or an Angel,
could make the place and ground holy, and to be vſed of Moyſes with al
ſignes of reuerence and feare: how much more the corporal birth, abode,
and wonders of the Sonne of God in Iewrie, and his perſonal preſence in
the B.~Sacrament, may make that countrie and al Chriſtian Churches and
altars holy? And it is the greateſt blindnes that can be, to thinke it
ſuperſtition to reuerence any things or places in reſpect of Gods
preſence or wõderous operation in the ſame. See
\Cite{S.~Hierom ep.~17. 18. 27. of the holy land.}}
holy ground. \V Seeing I haue ſeen the affliction of my people which is
in Ægypt, and I haue heard their groning and am deſcended to deliuer
them. And now come, and I wil ſend thee into Ægypt.}

\V This Moyſes, whom they
\Fix{dinied,}{denied,}{possible typo, fixed in other}
ſaying: \Emph{VVho hath appointed thee Prince and Captaine?} him God ſent
Prince &
\SNote{Chriſt is our Redeemer, and yet Moyſes is here called redeemer. So
Chriſt is our Mediatour & Aduocat, and yet we may haue Saints our
inferiour mediatours & aduocates alſo. See
\XRef{Annot. 1.~Io.~2,~1.}}
redeemer with the hand of the Angel that appeared to him in the buſh. \V
He
\CNote{\XRef{Exo.~7.}
\XRef{8.}
\XRef{9.}
\XRef{10.}
\XRef{11.}
\XRef{12.}
\XRef{37.}}
brought them forth doing wonders and ſignes in the land of Ægypt, and in
the redde ſea, and in the
\CNote{\XRef{Exo.~16.}}
deſert fourtie yeares. \V This is that Moyſes which ſaid to the children
of Iſraël: 
\CNote{\XRef{Deu.~18.}}
\Emph{A Prophet wil God raiſe vp to you of your owne brethren, as my
ſelf: him you shal heare.} \V This is he that
\CNote{\XRef{Exo.~19, 3--19.}}
was in the aſſemblie in the wilderneſſe, with the Angel that ſpake to
him in Mount Sina, and with our Fathers: who receiued the wordes of life
to giue vnto vs. \V To whom our Fathers would not be obedient: but they
repelled him, and in their harts turned away into Ægypt, \V ſaying to
Aaron:
\CNote{\XRef{Exo.~32,~1.}}
\Emph{Make vs Goddes that may goe before vs. For this, Moyſes that
brought vs out of the land of Ægypt, we know not what is befallen to
him.} \V And they made a calfe in thoſe dayes, and offered Sacrifice to
the Idol, and reioyced in the workes of their owne handes. \V And God
turned, and
\SNote{For a iuſt puniſhment of their former offenſes God gaue them vp
to worke what wickednes they would them ſelues, as it is ſaid of the
Gentils. 
\XRef{Rom.~1.}}
deliuered them vp
\TNote{\G{λατρεύειν}}
to ſerue the hoſt of Heauen, as it is written in the booke of the
Prophets:
\CNote{\XRef{Amos.~5,~25.}}
\Emph{Did you offer victims and hoſtes vnto me fourtie yeares in the
deſert, O houſe of Iſraël? \V And you tooke vnto you the tabernacle of
Moloch, and the ſtarre of your God Rempham, figures which you made, to
adore them. And I wil tranſlate you beyond Babylon.}

\V The tabernacle of teſtimonie was among our Fathers in the deſert, as
God ordained ſpeaking to Moyſes,
\CNote{\XRef{Exo.~25,~40.}}
\Emph{that he should make it according to the forme which he had
ſeen.} \V Which our Fathers
\CNote{\XRef{Ios.~3,~14.}}
with
\SNote{This is Ioſue ſo called in Greeke in type of our Sauiour.}
Ieſus receiuing, brought it in alſo into the poſſeſſion of the
Gentils. Which God expelled from the face of our Fathers, til in the
daies of Dauid, \V who found grace before God, and
\CNote{\XRef{Pſ.~131,~51.}}
deſired that he might find a tabernacle for the God of Iacob. \V And
\CNote{\XRef{Par.~17.}}
Salomon built him a houſe. \V But the Higheſt dwelleth
\LNote{Not in houſes.}{The vulgar Heretikes alleage this place againſt
the corporal being of Chriſt in the B.~Sacrament & in Churches: by which
reaſon they might haue driuen him out of al houſes, Churches, and
corporal places, when he was viſible in earth. But it is meant of the
Diuinitie only, & ſpoken to correct the carnal Iewes:
\MNote{God is not cõteined in place, yet he wil be worſhipped in one
place more then in another.}
who thought God
either ſo to be conteined, compaſſed, and limited to their Temple, that
he could be no where els, or at leaſt that he would not heare or receiue
mens praiers and Sacrifices in the Churches of the Gentils, or elſwhere,
out of the ſaid Temple. And ſo as it maketh nothing for the
Sacramentaries, no more doth it ſerue for ſuch as eſteeme Churches and
places of publike praier no more conuenient nor more holy then any other
profane houſes or chambers. For though his perſon or vertue be not
limited to any place, yet it pleaſeth him condeſcending to our
neceſſitie and profit, to worke his wonders, and to be worshipped of vs
in holy places rather then profane.}
not in houſes
\CNote{\XRef{Act.~17,~25.}}
made by hand, as the Prophet ſaith: \V
\CNote{\XRef{Eſa.~66,~1.}}
\Emph{Heauen is my ſeat: and the earth the foote-ſtole of my feet. VVhat
houſe wil you build me, ſaith our Lord, or what place is there of my
reſting? \V Hath not my hand made al theſe things?}

\V You ſtiffe-necked and of vncircumciſed hartes and eares, you
alwaies reſiſt the holy Ghoſt: as your Fathers, your ſelues alſo. \V
Which of the Prophets did not your Fathers
%%% o-2409
perſecute? And they ſlewe them that foretold of the comming of the Iuſt
one, of whom now \V you haue been betraiers and murderers: who receiued
the Law by the diſpoſition of Angels, and haue not kept it.

\V And hearing theſe things they were cut in their harts, and they
gnaſhed with their teeth at him. \V But he being ful of the holy Ghoſt,
looking ſtedfaſtly vnto Heauen,
\SNote{The comfort of al Martyrs.}
ſaw the glorie of God, and \Sc{Iesvs} ſtanding on the right hand of
God. \V And he ſaid: Behold I ſee the Heauens
%%% 2566
opened, and the Sonne of Man ſtanding on the right hand of God. \V And
they crying out with a loud voice, ſtopped their eares, and with one
accord ranne violently vpon him. \V And caſting him forth without the
citie they
\LNote{They ſtoned him.}{Read
\MNote{Relikes.}
a maruelous narration in S.~Auguſtine of one ſtone, that hitting the
Martyr on the elbow reboũded backe to a faithful man that ſtood neere,
who keeping & carying it with him, was by reuelation warned to leaue it
at Ancona in Italie: whereupon a Church or Memorie of S.~Steuen was
there erected, and many miracles done after the ſaid Martyrs body was
found out, and not before.
\Cite{Aug. to.~10. ſer.~38. de diuerſis. in edit. Pariſ.}}
ſtoned him: and the witneſſes laid off their garments
\CNote{\XRef{Act.~22,~20.}}
beſide the feet of a yong man that was called Saul. \V And they ſtoned
Steuen inuocating, and ſaying: Lord \Sc{Iesvs}, receiue my ſpirit. \V
And falling on his knees, he cried with a loud voice, ſaying:
\SNote{Euſebius Emiſſinus ſaith, whẽ he praieth for his perſecutours, he
promiſeth to his worſhippers his manifeſt interceſſion & ſuffrages.
\Cite{in ho. S.~Steph.}
& S.~Auguſtine, \L{Si Stephanus ſic non oraſſet, Eccleſia Paulum non
haberet.}
\Cite{Serm.~1. de S.~Stephano.}}
Lord, lay not this ſinne vnto them. And when he had ſaid this, he fel
aſleepe. And Saul was conſenting to his death.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

