%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 2599
%%% o-2445
\startChapter[
  title={Chapter 18}
  ]

\Summary{At Corinth in Achaia, he worketh with his owne hands, preaching
\Sc{Iesvs} to be \Sc{Christ}, vnto the Iewes vpon their
Sabboths. 6.~But they being obſtinate and blaſpheming, he in plaine
termes forſaketh them, and turneth to the Gentils, among whom
according to a viſion that he had to embolden him, he planteth the
Church in great numbers, 12.~the obſtinat Iewes in vaine ſoliciting
the Proconſul againſt him. 18.~From thence at length departing he
returneth, 19.~by Epheſus (where he promiſeth the Iewes to returne to
them) 22.~and ſo to Antioch in Syria (from whence he began his iourney
\XRef{Act.~15.})
23.~but not reſting, by and by he goeth againe to viſite the new
Churches that the planted
\XRef{Act.~16.}
in Galatia and Phrygia: 24.~Apollo in his abſence mightily confounding
the Iewes at Epheſus, 27.~and afterward at Corinth.}

After theſe things, departing from Athens, he came to Corinth, \V and
finding a certaine Iew, named Aquila, borne in Pontus, who of late was
come out of Italia, & Priſcilla his wife (becauſe Claudius had commanded
al Iewes to depart from Rome,) he came to them. \V And becauſe he was of
the ſame craft, he remained with them, and wrought, (and they were
tent-makers by their craft.) \V And
%%% o-2446
he diſputed in the Synagogue euery Sabboth, interpoſing the name of our
Lord \Sc{Iesvs}, and he exhorted the Iewes and the Greekes. \V And when
Silas and Timothee were come from Macedonia, Paul was inſtant in
preaching, teſtifying to the Iewes that \Sc{Iesvs} is \Sc{Christ}.
\V But they contradicting and blaſpheming, he ſhaking his garments,
ſaid to them: Your bloud vpon your owne head: I being cleane, from
henceforth wil goe to the Gentils. \V And departing thence, he entred
into the houſe of a certaine man, named Titus Iuſtus, one that ſerued
God, whoſe houſe was adioyning to the Synagogue. \V And Criſpus the
Prince of the Synagogue beleeued our Lord, with al his houſe: & many of
the Corinthians hearing beleeued, and were baptized. \V And our Lord
ſaid in the night by a viſion to Paul: Doe not feare, but ſpeake, and
hold not thy peace, \V for becauſe I am with thee: and no man ſhal ſet
vpon thee to hurt thee: for I haue much people in this citie. \V And he
ſate there a yeare & ſix moneths, teaching among them the word of God.

\V But Gallio being Proconſul of Achaia, the Iewes with one accord roſe
vp againſt Paul, and brought him to the iudgement ſeat, \V ſaying: That
this man contrarie to the Law perſuadeth men to worſhip God.
%%% 2600
\V And Paul beginning to open his mouth, Gallio ſaid to the Iewes: If it
were ſome iniuſt thing, or an heinous fact, O you men Iewes, I ſhould by
reaſon tolerate you. \V But if they be queſtions of word and names, and
of your law, your ſelues looke vnto it: I wil not be iudge of theſe
things. \V And he droue them from the iudgement ſeat. \V And al
apprehending Softhenes the Prince of the Synagogue, ſtrook him before
the iudgement ſeat: and Gallio cared for none of thoſe things.

\V But Paul when he had ſtaied yet many daies, taking his leaue of the
Brethren, ſailed to Syria, (and with him Priſcilla and Aquila,) who had
ſhorne his head in Cenchris. For he had
\CNote{\XRef{Nu.~6,~18.}
\XRef{Act.~21,~24.}}
a vow. \V And he came vnto Epheſus, and them he left there. But himſelf
entring into the Synagogue, diſputed with the Iewes. \V And when they
deſired him, that he would tarie a longer time, he conſented not, \V
but taking his leaue, and ſaying: I wil returne to you againe God
willing, he departed from Epheſus. \V And going downe to Cæſarea, he
went vp, and ſaluted the Church, and came downe to Antioche.

%%% o-2447
\V And hauing taried there a certaine time, he departed, walking in order
through the countrie of Galatia and Phrygia, confirming al the
Diſciples.

\V And a certaine Iew, named
\MNote{Apollo.}
Apollo, borne at Alexandria, an eloquent man, came to Epheſus, mighty in
the Scriptures. \V This man was taught the way of our Lord: and being
ſeruent in ſpirit he ſpake, & taught diligently thoſe things that
pertaine to \Sc{Iesvs}, knowing only the Baptiſme of Iohn. \V This man
therfore began to deale confidently in the Synagogue. Whom when
Priſcilla and Aquila had heard, they tooke him vnto them, and expounded
to him the way of our Lord more diligently. \V And whereas he was
deſirous to goe to Achaia, the Brethren exhorting wrote to the Diſciples
to receiue him. Who, when he was come, profited them much that had
beleeued. \V For he with vehemencie conuinced the Iewes openly, ſhewing
by the Scriptures, that \Sc{Iesvs} is \Sc{Christ}.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

