%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 2582
%%% o-2426
\startChapter[
  title={Chapter 13}
  ]

\Summary{The
\MNote{The 5.~part.

The taking of the Ghoſpel away from the obſtinate Iewes, and giuing of
it to the Gentils by the miniſterie of Paul and Barnabas.}
  Preachers of the Church of Antioche preparing themſelues,
  the Holy Ghoſt out of them al, chooſeth Saul and Barnabas. 3.~They
  being firſt conſecrated Bishops, 4.~goe their appointed circuit ouer
  al the land of Cypres; the Proconſul wherof is alſo conuerted, ſeing
  the miraculous excecation of a Iew by Paul. 13.~Thence, into
  Pamphilia: 14.~and Piſidia, where in Antioche Paul preached to the
  Iewes, shewing that \Sc{Iesvs} is Chriſt, 38.~and that in him is
  ſaluation, and not in their Law of Moyſes: 40.~warning them to beware
  of the reprobation foretold by the Prophets. 44.~But the next Sabboth,
  they blaſpheming, he in plaine termes forſaketh them, and turneth to
  the Gentils. Whereat the Gentils be as glad on the contrarie
  ſide. 50.~Finally the Iewes raiſing perſecution, they forſake them,
  pronouncing them to be obſtinate contemners.}

%%% o-2427
And there were in the Church which was at Antioche, Prophets and
Doctours, among whom was Barnabas, and Simon that was called Niger, and
Lucius of Cyrene, and Manahen who was the foſter-brother of Herod the
Tetrarch, and Saul. \V And
\TNote{\G{λειτουργούντων αὐτῶν}}
as they were
\LNote{As they were miniſtring.}{If
\MNote{The Apoſtles liturgie or Maſſe.}
we should, as our aduerſaries doe, boldly turne what text we liſt, & fly
from one language to another for the aduantage of our cauſe, we might
haue tranſlated for \Emph{miniſtring}, \Emph{ſacrificing}. For ſo
\TNote{\G{λειτουργούντων αὐτῶν}}
the Greeke doth ſignifie, and ſo Eraſmus tranſlated. Yea we might haue
tranſlated, \Emph{Saying Maſſe}, for ſo they did: and the Greeke Fathers
hereof had the name, Liturgie, which Eraſmus tranſlateth \Emph{Maſſe},
ſaying, \L{Miſſa Chriſoſtomi}. But we keepe our text, as the
tranſlatours of the Scriptures should doe moſt religiouſly.}
miniſtring to our Lord, & faſting, the Holy Ghoſt ſaid:
\LNote{Separate me.}{Though
\MNote{Paul & Barnabas are conſecrated by men.}
\CNote{Gal.~1.}
Paul were taught by God himſelf and ſpecially deſigned by Chriſt to be
an Apoſtle, and here choſen by the Holy Ghoſt together with Barnabas,
yet they were to be ordered, conſecrated, and admitted by men. Which
wholy condemneth al theſe new rebellious diſordered ſpirits, that
chalenge and vſurpe the office of preaching and other ſacred actions
from heauen, without the Churches admiſſion.}
Separate me Saul and Barnabas vnto the worke, whereto I haue taken
them. \V Then they
\LNote{Faſting.}{Hereof
\MNote{Imber daies.}
the Church of God vſeth and preſcribeth publike faſtes at the foure
ſolemne times of giuing holy Orders (which are our \Emph{Imber daies})
as a neceſſarie preparatiue to ſo great a worke as
\CNote{\Cite{Leo ep.~81. c.~1. Epipha. in compend.}
\Cite{Leo ſer.~3. &~4. de ienu. 7.~mẽſis.}}
S.~Leo declareth by this place, naming it alſo an Apoſtolical
tradition. See S.~Leo
\Cite{Ser.~9. de ieiunio 7.~menſis.},
&
\Cite{Calixtus ep.~1. to.~1.}
\Cite{Conc. Magunt. c.~34.~35. to.~3.}
\MNote{Præſcript times of faſts.}
And this faſting was not faſting from ſinne, nor moral or Chriſtian
temperance, as the Proteſtants ridiculouſly affirme, for ſuch faſting
they were bound euer to keepe: but it was abſtinence for a time from al
meates or from certaine kindes of meates, which was ioyned with praier
and ſacrifice, and done ſpecially at ſuch ſeaſons as the Church
preſcribed, of al together (as in Lent, the Imber daies, Friday,
Saturday) and not when euery man liſt, as Aërius and ſuch Heretikes did
hold.
\Cite{S.~Auguſt. har.~53.}}
faſting and praying, and
\LNote{Impoſing hands.}{Becauſe
\MNote{Impoſition of handes.}
al bleſſings & conſecrations were done in the Apoſtles times by the
external ceremonie of impoſition of hands, diuers Sacraments were named
of the ſame, ſpecially Confirmation, as is noted before, and
\MNote{Holy orders.}
holy Ordering or conſecrating Biſhops, Prieſts, and Deacons, and
Subdeacons, as we
ſee here and elſwhere. In which though there were many holy wordes and
ceremonies and a very ſolemne action; yet whatſoeuer is done in thoſe
Sacraments, is altogether called
\CNote{\XRef{Act.~6,~6.}}
\Emph{Impoſition of hands}: as whatſoeuer was done in the whole diuine
myſterie of the
\CNote{\XRef{Act.~2,~41.}}
B.~Sacrament, is named \Emph{fraction of bread}. For the Apoſtles (as
S.~Denys
\Cite{Eccl. hier. c.~1. in fine}
writeth) purpoſely kept cloſe in their open ſpeaches and writings which
might come to the hands or eares of Infidels, the ſacred wordes &
actions of the Sacraments. And S.~Ambroſe ſaith,
\Cite{in 1.~Tim. c.~4.}
\Emph{The impoſition of the hands is myſtical wordes, wherewith the
elected is conformed and made apt to his function, receiuing authoritie
(his conſcience bearing witnes) that he may be bold in our Lordes ſteed
to offer Sacrifice to God.} And S.~Hierom:
\CNote{\Cite{Hiero. in c.~58. Eſa.}}
\Emph{The impoſition of hands is the Ordering of Clerkes, which is done
by praier of the voice, & impoſition of the hand.} And this is in ſome
inferiour orders alſo: but Paul and Barnabas were ordered to a higher
function then inferiour Prieſts, euen to be Biſhops throughout al
Nations.}
impoſing hands vpon them, diſmiſſed them.

\V And they being
\LNote{Sent of the Holy Ghoſt.}{Whoſoeuer
\MNote{Spiritual officers of our ſoules.}
be ſent by the Church, are ſent of the Holy Ghoſt, though in ſuch an
extraordinarie ſort it be not done. Whereby we ſee how farre the
Officers of our ſoules in the Church doe paſſe the temporal Magiſtrates,
who though they be of Gods ordinance, yet not of the Holy Ghoſts ſpecial
calling.}
ſent of the Holy Ghoſt, went to Seleucia, and thence ſailed to
Cypres. \V And when they were come to Salamina, they preached the word of
God in the Synagogues of the Iewes. And they had Iohn alſo in their
miniſterie. \V And when they had walked throughout the whole iland as
farre as Paphos, they found a certaine man that was a magician, a
falſe-Prophet, a Iew, whoſe name was
\Var{Bar-ieſu,}{Barieu}
\V who was with the Proconſul Sergius Paulus a wiſe man. He ſending for
Barnabas & Saul, deſired to heare the word of God. \V But Elynias the
magician (for ſo is his name interpreted) reſiſted them, ſeeking to
auert the Proconſul from the faith. \V But Saul, otherwiſe Paul,
repleniſhed with the Holy Ghoſt, looking vpon him, \V ſaid: O ful of
guile, and al deceit, ſonne of the Diuel, enemie of al iuſtice, thou
ceaſeſt not to ſubuert the right waies of our Lord. \V And now behold
the hand of our Lord vpõ thee, & thou ſhalt be blind, not ſeing the
ſunne vntil a time. And forthwith there fel dimneſſe and darkeneſſe vpon
him, and going about he ſought ſome body that would giue him his
hand. \V Then the Proconſul, when he had ſeen that which was done,
beleeued marueling at the doctrine of our Lord. \V And when Paul and
they that were with him had ſailed from Paphos, they came to Perge in
Pamphylia. And Iohn departing from them, returned to Hieruſalem. \V But
they paſſing through Perge, came to Antioche in Piſidia: and entring
into the Synagogue on the day of the Sabboths, they ſate downe. \V And
after the leſſon of the Law and the Prophets, the Princes of the
%%% 2583
Synagogue ſent to them, ſaying: Men Brethren, if there be among you any
ſermon of exhortation to the people, ſpeake.

%%% o-2428
\V And Paul riſing vp, and with his hand beckning for ſilence, ſaid: Ye
men of Iſrael, and you that feare God, harken. \V The God of the people
of Iſrael choſe our Fathers,
\CNote{\XRef{Exod.}}
and exalted the people when they were
ſeiourners in the land of Ægypt, and in a mightie arme brought them out
thereof, \V and for the ſpace of fourtie yeares tolerated their manners
in the deſert. \V
\CNote{\XRef{Ioſue.}}
And deſtroying ſeuen Nations in the land of Chanaan, by lot he deuided
their land among them, \V as it were after foure hundred and fiftie
yeares,
\CNote{\XRef{Iud.}}
and after theſe things he gaue Iudges, vntil Samuel the Prophet. \V
And thenceforth they deſired a King: & he gaue them
\CNote{\XRef{1.~Reg.~8.}}
Saul the ſonne of Cis, a man of the Tribe of Beniamin, fourtie
yeares. \V And remouing him he raiſed them vp
\CNote{\XRef{1.~Reg.~16.}}
Dauid to be King: to whom giuing teſtimonie, he ſaid:
\CNote{\XRef{Pſ.~88,~21.}}
\Emph{I haue found Dauid the ſonne of Ieſſe, a man according to my hart,
who shal doe al my willes.}

\V Of his ſeed God according to his
\CNote{\XRef{Pſ.~131,~11.}}
promiſe hath brought forth to Iſrael a Sauiour \Sc{Iesvs}, \V Iohn
\CNote{\XRef{Lu.~3,~3.}}
preaching before the face of his comming, Baptiſme of penance to al the
people of Iſrael. \V And when Iohn fulfilled his courſe, he ſaid: Whom
doe
\CNote{\XRef{Lu.~3,~15.}}
you thinke me to be? I am not he: but behold there commeth after me,
whoſe ſhoes of his feet I am not worthie to looſe.

\V Men Brethren, children of the ſtocke of Abraham, & they among you
that feare God; to you the word of this ſaluation was ſent. \V For they
that inhabited Hieruſalem, and the Princes thereof, not knowing him, nor
the voices of the Prophets that are read euery Sabboth, iudging haue
fulfilled them, \V and finding no cauſe of death in him,
\CNote{\XRef{Lu.~23,~1.}}
deſired of Pilate, that they might kil him. \V And when they had
conſummated al things that were written of him, taking him downe from
the tree, they put him in a monument. \V But God raiſed him vp from the
dead the third day: \V who was
\CNote{\XRef{Act.~1,~3.}}
ſeen for many daies of them that came vp together with him from Galilee
into Hieruſalem, who vntil this preſent are his witneſſes to the
people. \V And we preach vnto you that promiſe which was made to our
Fathers: \V that God hath fulfilled this ſame
\Var{to our children,}{to vs their childrẽ}
raiſing vp \Sc{Iesvs}, as in the ſecond Pſalme alſo it is written:
\CNote{\XRef{Pſ.~2,~7.}}
\Emph{My Sonne art thou, this day haue I begotten thee.} \V And that he
raiſed him vp from the dead, not to returne now any more into
corruption, thus he ſaid:
\CNote{\XRef{Eſa.~55,~3.}}
\Emph{That I wil giue you the holy things of Dauid faithful.} \V And
%%% o-2429
therfore in another place alſo he ſaith:
\CNote{\XRef{Pſ.~15,~10.}}
\Emph{Thou shalt not giue thy holy one to ſee corruption.} \V For Dauid
in his generation when he had ſerued, according to the wil of God ſlept:
and he was laid to his Fathers & ſaw corruption. \V But he whom God hath
raiſed vp, ſaw no corruption.

\V Be it knowen therfore to you, men Brethren, that through him,
forgiueneſſe of ſinnes is preached to you, from al the things from the
which you could not be iuſtified by the law of Moyſes. \V In him euery
one that beleeueth, is iuſtified. \V Take heed therfore leſt that come
vpon you which is ſpoken in the prophets, \V
\CNote{\XRef{Abac.~1,~5.}}
\Emph{See ye contemners, and wonder, and perish: becauſe I worke a worke
in your daies, a worke which you wil not beleeue, if any man shal tel
it you.}

%%% 2584
\V And they going forth,
\Var{they deſired}{the Gentiles deſired}
them that the Sabboth following they would ſpeake vnto them theſe
wordes. \V And when the Synagogue was diſmiſſed, many of the Iewes, and
of the
\Var{ſtrangers ſeruing God,}{deuout proſelytes}
followed Paul & Barnabas: who ſpeaking exhorted them to continue in the
grace of God. \V But the next Sabboth the whole citie almoſt aſſembled
to heare the word of God. \V And the Iewes ſeing the multitudes, were
repleniſhed with enuy, & contradicted thoſe things which were ſaid of
Paul, blaſpheming. \V Then Paul and Barnabas conſtantly ſaid: To you it
behoued vs firſt to ſpeake the word of God: but becauſe
\SNote{The Iewes of their owne free wil repelling the truth, are
vnworthy of Chriſt & worthily forſakene & the Gentils though they
beleeued ſpecially by Gods grace & preordination, yet they beleeue alſo
by their owne free wil, which ſtandeth wel with Gods prouidence.}
you repel it, and iudge yourſelues vnworthie of eternal life; behold we
turne to the Gentils. \V For ſo our Lord commanded vs:
\CNote{\XRef{Eſ.~46,~6.}}
\Emph{I haue put thee to be the light of the Gentils; that thou maieſt
be ſaluation vnto the vtmoſt of the earth.} \V And the Gentils hearing
it, were glad, and glorified the word of our Lord: and there beleeued as
many as were preordinate to life euerlaſting. \V And the word of our
Lord was ſpred throughout the whole countrie. \V But the Iewes ſtirred
vp religious and honeſt women, and the cheefe of the citie, and raiſed
perſecution againſt Paul and Barnabas: and they did caſt them forth out
of their coaſts. \V But they
\CNote{\XRef{Lu.~9,~5.}}
ſhaking off the duſt of their feet againſt them, came to Iconium. \V The
Diſciples alſo were repleniſhed with ioy and with the holy Ghoſt.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

