%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 28
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-28


\project douay-rheims


%%% 2619
%%% o-2470
\startChapter[
  title={Chapter 28}
  ]

\Summary{After their shipwracke hauing wintered in the Iland (now named
  Malta) where many miracles were wrought by Paul, they take ship
  againe, and ſo by Cicilie they come to Puteoli in Italie, the
  Chriſtian Romans comming a great way to meet him, to his great
  ioy. 16.~Finally being come to Rome, in his lodging he declareth to
  the Iewes his cauſe. 23.~And on a day appointed preacheth \Sc{Iesvs}
  vnto them. 25.~And ſeeing their incredulitie, he sheweth how it was
  foretold by Eſay: 28.~but that the Gentils wil not be
  incredulous. 30.~To whom he there preacheth two whole yeares without
  prohibition.}

And when we had eſcaped, then we knew that the
\SNote{This Iland (now Malta) is the ſeat of the Knights of Rhodes: the
inhabitãts wherof haue a ſpecial deuotion to S.~Paul: to whom both the
cheefe Church (being the Biſhops Seat) is dedicated, and the whole Iland
(as they count it) conſecrated. Where the people shew yet to ſtrãgers,
his priſon and other memories of his miracles.}
Iland was called
\Var{Mitilene.}{Melita}
But the Barbarous ſhewed vs no ſmal courteſie. \V For, kindling a fire
they refreshed vs al, becauſe of the imminent raine and the cold. \V And
when Paul had gathered together ſome number of ſtickes, and had laid
them on the fire, a viper iſſuing out of the heat, inuaded his hand. \V
But as the Barbarous ſaw the beaſt hanging on his hand, they ſaid one to
another: Vndoubtedly this man is a
%%% 2620
murderer, who being eſcaped out of the ſea, 
\TNote{\G{ἡ δίκη}}
vengeance doth not ſuffer him to liue. \V And he indeed
\LNote{Shaking of the beaſt.}{The promiſe of Chriſt
\XRef{(Marc.~16.)}
that venemous ſerpents ſhould not hurt them that beleeue in him, is
fulfilled; not in al beleeuers, but in ſuch as had the guift of
miracles, as S.~Paul had: whom here a viper, by nature ſo venemous that
the people thought he ſhould haue died out of hand, did no whit annoy;
he extinguiſhing by the power of Chriſt al the poiſon of the beaſt.
\CNote{\Cite{The Fazellus de rebus Siculus decad.~1. li.~16. c.~1.}}
\MNote{Malta hath S.~Paules bleſſing and grace vntil this day.}
Yea and (as the Chriſtian people there til this day beleeue) by
S.~Paules prayers the Iland was deliuered for euer from al ſuch venemous
ſerpents, in ſo much that children there play with ſcorpions euer ſince
that time: and Pilgrimes daily carie with them peeces of ſtones out of
the place where S.~Paul abode, by which they affirme that they heale
them which in other countries adioyning are bitten of ſcorpions, the
medicine therfore being called, \Emph{S.~Paules grace}.
\MNote{Gods miraculous vertue in certaine countries and creatures, by
his Saints.}
The Heretikes that know not the power of God, nor the miraculous vertues
giuen to his Saints, maruel and blaſpheme, when they heare ſuch things
as be proper to certaine countries, attributed ſometimes to Gods
miracles done by his Saints: as though that were not poſſible, or were
not as much to Gods honour, and more, then things proceeding only of
natural cauſes. Such profane men would not haue attributed the holſomnes
of the waters of Iericho to Eliſeus his vertue and miracles, ammending
them by caſting ſalt into them, if the
\CNote{\XRef{4.~Reg.~2,~19.}}
Scripture had not expreſly teſtified the ſame. It is the part of al
faithful men to referre ſuch things to God, when any iuſt occaſion is
giuen thereunto, rather then to nature: though the incredulous doe
alwaies contrarie, for feare of ſuperſtition & diſhonouring God. As
though this eſcape of drowning, might better and more to Gods glorie, be
referred to chance and the mariners induſtrie, then to S.~Paules praiers
and extraordinarie working.}
ſhaking off the beaſt into the fire, ſuffred no harme. \V But they
ſuppoſed that he ſhould be turned into a ſwelling, and that he would
ſodenly fal and die. But expecting long and ſeeing that there was no
harme done on him, being changed they ſaid, that he was a God. \V And
in thoſe places were land of the Prince of the Ile, named Publius, who
%%% o-2471
receiuing vs, for three daies intreated vs courteouſly. \V And it
chanced that the father of Publius lay vexed with feuers and the bloudy
flixe. Vnto whom Paul entred: and when he had praied,
\SNote{Miraculous healing by impoſition of the Apoſtles hãds.}
and impoſed hands on him, he healed him. \V Which being done, al in the
Ile alſo that had infirmities, came, and were cured: \V who alſo
honoured vs with many honours, and when we were ſailing away, laded vs
with neceſſaries.

\V And after three moneths, we ſailed in a ſhip of Alexandria, that had
wintered in the Iland, whoſe ſigne was the Caſtors. \V And when we were
come to Syracuſa, we taried there three daies. \V Thence compaſſing by
the ſhore, we came to Rhegium: and after one day the South-wind blowing,
we came the ſecond day to Puteoli, \V where finding Brethren, we
were deſired to tarie with them ſeuen daies: and ſo we came to
Rome. \V And from thence, when the Brethren had heard, they came to meet
vs vnto \L{Apij-forum}, and the Three-tauerns, whom when Paul had ſeen,
giuing thanks to God, he tooke courage. \V And when we were come to
Rome, Paul was permitted to remaine to himſelf with a ſouldiar that kept
him. \V And after the third day he called together the cheefe of the
Iewes. And when they were aſſembled, he ſaid to them:

Men Brethren, I doing nothing againſt the people, or the cuſtome of the
Fathers, was deliuered priſoner from Hieruſalem into the hands of the
Romanes, \V who when they had examined me, would haue releaſed me, for
that there was no cauſe of death in me. \V But the Iewes contradicting
it, I was compelled to appeale vnto Cæſar, not as hauing any thing to
accuſe my Nation. \V For this cauſe therfore I deſired to ſee you and to
ſpeake to you. For, becauſe of the hope of Iſrael, am I compaſſed with
this
\LNote{Chaine.}{I
\MNote{S.~Pauls chaines honoured.}
would wiſh now (ſaith S.~Chryſoſtome) to be for a time in the place
where theſe chaines remaine, and to ſee the fetters which Diuels feare,
and Angels reuerence.
\Cite{hom.~5. ad populum Antiochonum.}
See alſo
\Cite{S.~Gregorie li.~3. epiſt.~30.}
of the miracles done by S.~Paules chaines, and that he ſendeth to the
Empereſſe Conſtantia ſome duſt thereof filed off, for a great Relike and
holy guift.}
chaine.

\V But they ſaid to him: We neither receiued letters concerning thee from
Iewrie, neither did any of the Brethren that came hither, report or
ſpeake any euil of thee. \V But we deſire of thee to heare what thou
thinkeſt: for
\LNote{Concerning this Sect.}{The
\MNote{The name of Sect is wel giuen to al Hereſies, though the
Chriſtian religion at the firſt was falſely ſo called.}
Heretikes of al ſortes comfort
themſelues much, when they find here or els where the Chriſtian faith
called of the Iewes or incredulous perſons, a Sect or an Hereſie, &
ſometimes in contempt of Chriſts perſon the Maiſter of the ſame, the
Sect of the Nazarens: as though the Church of God might as wel erre in
naming their doctrine Hereſie, as the Iewes and Pagans might and did
miſſe in
condemning Chriſtian religion for an Hereſie: or as though the
Proteſtants doctrine were as wel proued & tried to be no Hereſie, by the
Prophets and other Scriptures, miracles, and conſent of al Nations and
Ages, as Chriſts bleſſed doctrine is. Whereas indeed the Proteſtants
doctrine is 
euidently conuinced to be heretical, by the ſame arguments that Chriſts
religion is proued to be the only true doctrine of ſaluation, and not an
Hereſie. And whoſoeuer can deduce the Chriſtian faith from Adam to this
day, throughout al the Fathers, Patriarchs, Prophets, Prieſts, Apoſtles, &
Bishops, by deſcent and ſucceſſion of al Lawes and States of true
worshippers and beleeuers (which is the only or ſpecial way to proue
that the Chriſtian faith is no Hereſie) he shal by the ſame meanes al at
once proue the Proteſtants doctrine to be an Hereſie & a falſe
Sect. That the Iewes therfore and il men in al places contradicted the
Chriſtian religion, calling it an Hereſie or a Sect, as though it had a
beginning of ſome certaine Sect-Maiſter other then God himſelf, they
were deceiued: and the Church of God neuertheleſſe calling the
Proteſtãts doctrine Hereſie in the worſt part that can be, and in the
worſt ſort that euer was, doth right and moſt iuſtly.}
concerning this
\TNote{\G{αἱρέσεως}}
Sect, it is knowen to vs that it is gaineſaid euery where. \V And when
they had appointed him a day, they came to him vnto his lodging very
many: to whom he expounded, teſtifying the Kingdom of God, and vſing
perſuaſion to them
%%% o-2472
of \Sc{Iesvs}, out of the Law of Moyſes and the Prophets, from morning
vntil euening. \V And certaine beleeued thoſe things that were ſaid: but
certaine beleeued not. \V And whereas they did not agree among
themſelues, they departed, Paul ſaying one word: That wel did the Holy
Ghoſt ſpeake by Eſaie the Prophet to our Fathers, \V ſaying:
\CNote{\XRef{Eſa.~6,~9.}
\XRef{Mat.~13,~14.}
\XRef{Mr.~4,~12.}
\XRef{Lu.~8,~10.}
\XRef{Io.~12,~40.}
\XRef{Ro.~11,~8.}}
\Emph{Goe to this people, and ſay to them: With the eare you shal heare,
and shal not vnderſtand: and ſeeing you shal ſee and shal not
perceiue. \V For the hart of this people is waxen groſſe, and
%%% 2621
with their eares haue they heauily heard, and their eyes
\SNote{Here alſo (as
\XRef{Mat.~13.})
it is plaine that they would not ſee, nor heare, and that their
execration is to be attributed to themſelues & not to God. See
\XRef{annot. Io.~12,~40.}}
they haue shut: leſt perhaps they may ſee with their eyes, and heare
with their eares, and vnderſtand with their hart, and be conuerted, and
I heale them.} \V Be it knowen therfore to you, that this Saluation of
God is ſent to the Gentils, and they wil heare.

\V And when he had ſaid theſe things, the Iewes went out from him,
hauing much queſtioning among themſelues. \V And he taried ful two
yeares in his hired lodging: and he receiued al that came in to him, \V
preaching the Kingdom of God, and teaching the things that concerne our
Lord \Sc{Iesvs Christ} with al confidence, without prohibition.

\Sc{The End of the Actes of the Apoſtles.}

Wherevnto we adioyne for the Readers behalfe, two Tables of the two
cheefe Apoſtles, and a note of the reſt, as an abridgement of the ſaid
booke, and a ſupply of ſome things not there mentioned.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

