%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 2559
%%% o-2401
\startChapter[
  title={Chapter 5}
  ]

\Summary{Ananias and his wife Saphira, for their ſacriledge, at Peters
  word fal downe dead, to the great terrour of the reſt. 12.~By the
  Apoſtles miracles, not only the number, but alſo their faith ſo
  increaſeth, that they ſeeke in the ſtreetes to the very shadow of
  Peter, the townes alſo about bringing their diſeaſed to
  Hieruſalem. 17.~The Rulers againe oppoſe themſelues, but in
  vaine. 19.~For out of priſon an Angel deliuereth them, bidding them
  preach openly to al: 27.~& in their Councel Peter is nothing afraid of
  their bigge wordes: 34.~Yea Gamaliel being one of themſelues caſteth a
  doubt among them, leſt the matter be of God, and therfore impoſsible
  to be diſſolued. 40.~Finally, the Apoſtles being ſcourged by them,
  count it an honour, & ceaſe no day from preaching.}

But a certaine man named Ananias, with Saphira his wife ſold a peece of
land, \V and
\LNote{Defrauded.}{In that (ſaith
\CNote{\Cite{Aug. ſer.~9. in append. de diuerſis tom.~10.}}
S.~Auguſtine) he withdrew any part of that which he promiſed, he was
guilty at once, both of ſacrilege, and of fraud:
\MNote{Sacrilege.}
of ſacrilege, becauſe he robbed God of that which was his by promiſe: of
fraud, in that he withheld of the whole guift, a peece. Let now the
Heretikes come, and ſay it was for lying or hypocriſie only that this
fact was condemned, becauſe they be loth to haue ſacrilege counted any
ſuch ſinne, who haue taught men not only to take away from God ſome
peece of that, or al that themſelues gaue, but plainly to ſpoile &
applie to themſelues al that other men gaue.}
defrauded of the price of the land, his wife being priuie thereto: and
bringing a certaine portion, laid it at the feet of the Apoſtles. \V And
\LNote{Peter ſaid.}{S.~Peter (as you ſee here) without mans relation
knew this fraud and the cogitations of Ananias, and as head of the
Colledge and of the whole Church againſt which this robbery was
committed, executed
\MNote{Excommunication ioyned with corporal paines.}
this heauy ſentence of Excommunication both againſt him, & his wife
conſenting to the Sacrilege. For it was excommunication by S.~Auguſtines
iudgement,
\Cite{(li.~3. cont. ep. Parm. c.~2. to.~7.)}
and had this corporal miraculous death ioyned withal, as the
Excommunication that
\CNote{\XRef{1.~Cor.~5.}}
S.~Paul gaue out againſt the inceſtuous and others, had
the corporal vexation of Satan incident vnto it.}
Peter ſaid: Ananias, why hath Satan tempted thy hart, that thou
ſhouldeſt lie to the holy Ghoſt, and defraud of the price of the
land? \V Remaining, did it not remaine to thee: and being ſold, was it
not
\LNote{In thy povver.}{\Emph{If
\MNote{Vow of Chaſtitie, and the breach thereof.}
it diſpleaſed God} (ſaith S.~Auguſtine.) \Emph{to vvithdravv of the
money vvhich they had vovved to God, hovv is he angry vvhen chaſtitie is
vovved and is not performed: for to ſuch may be ſaid that vvhich
S.~Peter ſaid of the money: Thy virginitie remaining did it not remaine
to thee, and before thou didſt vovv, vvas it not in thine ovvne povver?
for, vvhoſoeuer haue vovved ſuch things and haue not paied them, let
them not thinke to be condemned to corporal deaths, but to euerlaſting
fire.}
\Cite{Auguſt. Ser.~10. de diuerſis.}
And S.~Gregorie to the ſame purpoſe vvriteth thus: 
\CNote{\Cite{Greg. li.~1. ep.~33.}}
\Emph{Ananias had vovved
money to God, vvhich aftervvard ouercome vvith diuelish perſuaſion he
vvithdrevv: but vvith vvhat death he vvas punished, thou knovveſt. If
then he vvere vvorthy of that death, vvho tooke avvay the money that
\Fix{the}{he}{obvious typo, fixed in other}
had giuen to God, conſider vvhat great peril in Gods iudgement thou
shalt be vvorthy of, vvhich haſt vvithdravven, not money, but 
\Fix{they}{thy}{obvious typo, fixed in other}
ſelf
from almighty God, to vvhom thou hadſt vovved 
\Fix{they}{thy}{obvious typo, fixed in other}
ſelf vnder the habite or vveede of a Monke.}}
in thy power? Why haſt thou put this thing in thy hart? Thou haſt not
lied
\LNote{Not to men, but.}{To take from the Church or from the Gouernours
therof, things dedicated to their vſe and the ſeruice of God, or to lie
vnto Gods Miniſters, is ſo iudged of before God, as if the lie were
made, and the fraud done to the Holy Ghoſt himſelf, who is the Churches
Preſident and Protectour.}
to men, but to God. \V And Ananias hearing theſe wordes, fel downe, and
gaue vp the ghoſt. And there came great feare vpon al that heard it. \V
And yong men riſing vp, remoued him, & bearing him forth buried him. \V
And it was the ſpace as it were of three houres, and his wife, not
knowing what was chanced, came in. \V And Peter anſwered her: Tel me
woman, whether did you ſel the land for ſo much? But ſhe ſaid: Yea, for
ſo much. \V And Peter vnto her: Why haue you agreed together to tempt
the Spirit of our Lord? Behold, their feete that haue buried thy
huſband, at the doore, and they ſhal beare thee forth. \V Forthwith ſhe
fel before his feete, and gaue vp the ghoſt. And the yong men going in,
found her dead: and caried her forth, & buried her by her huſband. \V
And there fel great
\SNote{Hereupon roſe great reuerence, awe, & feare of the vulgar
Chriſtians toward the holy Apoſtles; for an exãple to al Chriſtian
people how to behaue themſelues toward their Biſhops and Prieſts.}
feare in the whole Church, and vpon al that heard theſe things.

\V And by the handes of the Apoſtles were many ſignes and wonders done
among the people. And they were al with one accord in Salomons porch. \V
But of the reſt
%%% o-2402
none durſt ioyne themſelues vnto them: but the people magnified
them. \V And the multitude of men and women that beleeued in our Lord,
was more increaſed:
%%% 2560
\V ſo that they did bring forth the ſick into the ſtreetes, & laid them
in beddes and couches, that when Peter came,
\MNote{Peters ſhadow.}
\LNote{His shadovv.}{Specially
\MNote{Peters ſhadow & interceſſion.}
they ſought to Peter the cheefe of al,
who not only by touching, as the other, but by his very ſhadow cured al
diſeaſes. Whereupon S.~Auguſtine ſaith: If then the shadow of his body
could help, how much more now the fulnes of power? And if then a
certaine litle wind of him paſſing by did profit them that humbly aſked,
how much more the grace of him now being permanent & remaining?
\Cite{Ser.~39. de Sanctis.}
ſpeaking of the miracles done by the Saints now reigning in Heauen.}
his ſhadow at the leaſt might ouerſhadow any of them, and they al might
be deliuered from their infirmities. \V And there ranne together vnto
Hieruſalem the multitude alſo of the cities adioyning, bringing ſicke
perſons and ſuch
\Fix{vs}{as}{obvious typo, fixed in other}
were vexed of vncleane Spirits; who were al cured.

\V And the high Prieſt riſing vp, and al that were with him, which is
the hereſie of the Sadducees, were repleniſhed with zeale: \V laid hands
vpon the Apoſtles, and put them in the common priſon. \V
\MNote{An Angel leadeth them out of priſon.}
But an Angel of our Lord by night opening the gates of the priſon, &
leading them forth, ſaid: \V Goe; and ſtanding ſpeake in the Temple to
the people al the wordes of this life. \V Who hauing heard this, early
in the morning entred into the Temple, and taught. And the high Prieſt
comming, and they that were with him, called together the Councel & al
the Ancients of the children of Iſraël: and they ſent to the priſon that
they might be brought. \V But when the miniſters were come, and opening
the priſon, found them not; returning they told, ſaying: \V The priſon
truly we found ſhut with al diligence, and the keepers ſtanding before
the gates: but opening it, we found no man within. \V And as ſoone as
the Magiſtrate of the Temple and the cheefe Prieſts heard theſe wordes,
they were in doubt of them, what would befal. \V And there came a
certaine man and told them: That the men, loe, which you did put in
priſon, are in the Temple ſtanding, and teaching the people. \V Then
went the Magiſtrate with the miniſters, & brought them without force, for
they feared the people, leſt they ſhould be ſtoned. \V And when they had
brought them, they ſet them in
\Fix{the the}{the}{obvious typo, fixed in other}
Councel. And the high Prieſt asked
them, \V ſaying:
\CNote{\XRef{Act.~4,~18.}}
Commanding we commanded you that you ſhould not teach in this name: and
behold you haue filled Hieruſalem with your doctrine, and you wil bring
vpon vs the bloud of this man. \V But Peter anſwering and the Apoſtles,
ſaid: God muſt be obeied, rather then men.

\V The God of our Fathers hath raiſed vp \Sc{Iesvs}, whom you did kil,
hanging him vpon a tree. \V This Prince and
%%% o-2403
Sauiour God hath exalted with his right hand, to giue repentance to
Iſraël, and remiſſion of ſinnes. \V And we are witneſſes of theſe
wordes, and the holy Ghoſt, whom God hath giuen to al that obey him. \V
When they had heard theſe things, it cut them to the hart, and they
conſulted to kil them.

\V But one in the Councel riſing vp, a Phariſee named Gamaliel, a
Doctour of law honourable to al the people, commanded the men to be put
forth a while, \V and he ſaid to them: Ye men of Iſraël, take heed to
your ſelues touching theſe men what you meane to doe. \V For before
theſe daies there roſe
\Var{Theodas,}{Theudas}
ſaying he was ſome body, to whom conſented a number of men about foure
hundred, who was ſlaine: and al that beleeued him, were diſperſed, and
brought to nothing. \V After this fellow there roſe Iudas of Galilee in
the daies of the Enroling, and drew away the people after him, and he
periſhed: and as many as euer conſented to him, were diſperſed. \V And
now therfore
%%% 2561
I ſay to you, depart from theſe men and let them alone: for if this
counſel or worke be of men, it wil be diſſolued: \V but if it be
\SNote{Time, & the euidẽt ſucceſſe of Chriſtes Church &
religion, proue it to be of God: no violence of the Iewes, no
perſecution of the Heathen Princes, no endeauour of domeſtical
Aduerſaries, Heretikes, Schiſmatikes, or il liuers, preuailing againſt
it; as on the other ſide, many attempts haue been made by Arius,
Macedonius, Neſtorius, Luther, and the like, who thought thẽſelues ſome
body: but after they had plaied their partes awhile, their memory is
buried, or liueth only in malediction & infamie, and their ſcholars come
to naught. Therfore let no Cath. man be ſcandalized that this hereſie
holdeth vp for a time. For the Arians & ſome others floriſhed much
longer then theſe, & were better ſupported by Princes and learning, and
yet had an end.}
of God, you are not able to diſſolue
\Var{them,}{it}
leſt perhaps you be found to reſiſt God alſo. And they conſented to
him. \V And calling in the Apoſtles, after they had ſcourged them, they
charged them that they ſhould not ſpeake in the name of \Sc{Iesvs}, and
diſmiſſed them. \V And they went from the ſight of the Councel
reioycing, becauſe they were accounted worthy to ſuffer reproch for the
name of \Sc{Iesvs}. \V And euery day they ceaſed not in the Temple, and
from houſe to houſe to teach and euangelize Chriſt \Sc{Iesvs}.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

