%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 2596
%%% o-2442
\startChapter[
  title={Chapter 17}
  ]

\Summary{How in other parts of Macedonia he planted the Church, and
  namely at Theſſalonica, 5.~where the obſtinate Iewes are ſo malitious,
  that they purſue him alſo into Berœa. 14.~From whence being conducted
  into Greece, he preacheth at Athens both to the Iewes and Gentils
  diſputing with the Philoſophers, 19.~and in Areopagus, perſuading them
  from their Idols vnto one God and \Sc{Iesvs Christ} raiſed from the
  dead.}

And when they had walked through Amphipolis and Apollonia, they came to
Theſſalonica, where there was a Synagogue of the Iewes. \V And Paul
according to his cuſtome entred in vnto them, & three Sabboths he
diſcourſed to them out of the Scriptures, \V declaring and inſinuating
that it behoued \Sc{Christ} to ſuffer and to riſe againe from the dead:
& that this is \Sc{Iesvs Christ}, whom I preach to you. \V And certaine
of them beleeued, and were ioyned to Paul and Silas, and of the Gentils
that ſerued God a great multitude, and noble women not a few. \V But the
Iewes
\SNote{\L{Zelantes}. This is the zeale of Heretikes, and a liuely paterne
of their dealing at this day againſt Catholike Prieſts & Preachers, and
the good \Emph{Iaſons} that receiue them.}
enuying, & taking vnto them of the raſcal ſort certaine naughtie men,
and making a tumult, ſtirred the citie: and beſetting Iaſons houſe,
ſought to bring them forth vnto the people. \V And not finding them,
they drew Iaſon and certaine Brethren to the Princes of the citie
crying, That theſe are they that ſtirre vp the world, and are come
hither, \V whom Iaſon hath receiued, and al theſe doe againſt the
decrees of Cæſar, ſaying that there is another King, \Sc{Iesvs}. \V And
they moued the people, and the Princes of the citie hearing theſe
things. \V And taking a ſatisfaction of Iaſon and of the reſt, they
diſmiſſed them. \V But the Brethren forthwith by night ſent away Paul
and Silas vnto Berœa.

Who when they were come, entred into the Synagogue of the Iewes. \V (And
theſe were more noble then they that are at Theſſalonica, who receiued
the word with al greedines, daily
\LNote{Searching the Scriptures.}{The
\MNote{The people may not iudge of the ſenſe of Scriptures.}
Heretikes vſe this place to proue that the
\Fix{heares}{hearers}{obvious typo, fixed in other}
muſt trie & iudge by the Scriptures, whether their Teachers & Preachers
doctrine be true; & ſo reiect that they find not in the Scriptures: As
though here the sheep were made iudges of their Paſtours, the people of
the Prieſts, & men and woman of al ſortes, euen of S.~Paules doctrine it
ſelf: which were the moſt foolish diſorder in the world. And they did
not therfore read the Scriptures of the old Teſtament (for none of the
new were yet extant commonly) to diſpute with the Apoſtle, or to trie
and iudge of his doctrine, or whether they should beleeue him or no: for
they were bound to beleeue him and obey his word, whether he alleaged
Scripture or no, and whether they could read or vnderſtand the
Scriptures or no:
\MNote{The cõfort of Chriſtian men by hearing or reading the
Scriptures.}
but it was a great comfort and confirmation for the Iewes that had the
Scriptures, to find euen as S.~Paul ſaid, that Chriſt was God,
crucified, riſen, and aſcended to Heauen; which by his preaching and
expounding they vnderſtood, and neuer before, though they read them, and
heard them read euery Sabboth: As it is a great comfort to a Catholike
man, to heare the Scriptures declared & alleadged moſt euidently for the
Churches truth againſt Heretikes, in Sermons or otherwiſe. And it doth
the Catholikes good & much confirmeth them, to vew diligently the places
alleadged by the Catholike Preachers. Yet they muſt not be iudges for al
that, ouer their owne Paſtours, whom Chriſt commandeth them to heare and
obey, and by whom they heare the true ſenſe of Scriptures.}
ſearching the ſcriptures, if theſe things were ſo. \V And many ſurely of
them beleeued, and of honeſt women Gentils, and men not a few.) \V And
when the Iewes in Theſſalonica vnderſtood, that at Berœa alſo the word
of God was preached by Paul, they came thither alſo, mouing and
troubling the multitude. \V And then immediatly the
%%% o-2443
Brethren ſent away Paul, to goe vnto the ſea: but Silas and Timothee
remained there. \V And they that conducted Paul, brought him as farre as
Athens, and receiuing commandement of him to Silas and Timothee, that
they ſhould come to him very ſpeedily, they departed.

\V And when Paul expected them at Athens, his ſpirit was incenſed within
him, ſeeing the citie giuen to Idolatrie. \V He diſputed therfore in the
Synagogue with the Iewes, & them that ſerued God, and in the
market-place, euery day with them that were there. \V And certaine
Philoſophers of the Epicures and the Stoikes diſputed with him, and
%%% 2597
certaine ſaid, what is it that this
\TNote{\G{σπερμολόγος}}
Word-ſower would ſay? But others: He ſeemeth to be a Preacher of new
\TNote{\L{dæmoniorum}}
Gods: becauſe he preached to them \Sc{Iesvs} and the reſurrection. \V
And apprehending him, they led him to Areopagus, ſaying: May we know
what this new doctrine is that thou ſpeakeſt of? \V for thou bringeſt in
certaine new things to our eares. We wil know therfore what theſe things
may meane. (\V And al the Athenians, and the ſtrangers ſeiourning there,
emploied them ſelues to nothing els but either to ſpeake, or to heare
ſome newes.) \V But Paul ſtanding in the middes of Areopagus, ſaid:

Ye men of Athens, in al things I perceiue you as it were
%%% !!! Not marked in either
\LNote{Superſtitious.}{S.~Paul
\MNote{The Proteſtãts cal deuotion, ſuperſtition.}
calleth not them ſuperſtitious for
adoring the true and only God with much deuotion or many ceremonies or
in comely preſcribed order, or for doing reuerence to holy Sacraments,
to Saints and their memories, Images, or Monuments: or for keeping the
preſcribed lawes, daies, & faſtes of the Church, or for fulfilling vowes
made to God, or for bleſsing with the ſigne of the Croſſe, or for
capping and kneeling at the name of \Sc{Iesvs}, or for religiouſly vſing
creatures ſanctified in the ſame name, or any other Chriſtian
obſeruation, for which our new Maiſters condemne the Catholike people of
Superſtition (themſelues wholy void of that vice by al wiſe mens
iudgement, becauſe they haue in a manner taken away al religion, and are
become Epicureans and Atheiſts; who are neuer troubled with
ſuperſtition, becauſe it is a vice conſiſting in exceſſe of worship or
religion, whereof they are void) but
\MNote{The Apoſtle ſpeaketh of the Heathens ſuperſtition.}
the Apoſtle calleth them
ſuperſtitious for worshipping the Idols and Gods of the Heathen, and
\TNote{\G{δεισιδαιμονες}}
for the feare that they had, leſt they should leaue out any God that was
vnknowen to them: for thus their Altar was inſcribed: \L{Dijs Aſiæ,
Europæ, & libiæ; Deo ignoto & peregrino}: that is, To the Gods, of Aſia,
Europe, and Lybia; to the vnknowen & ſtrange God. This ſuperſtition
(ſaith
\CNote{\Cite{Aug. de conf. Euang. li.~1. c.~26.}}
S.~Auguſtine) is wholy taken away from the Church by Chriſts
incarnation, and by the Apoſtles preaching, and by Martyrs holy life and
death.
\MNote{The Catholike Church alloweth no ſuperſtition.}
Neither doth the Catholike Church allow this or any other kind of
ſuperſtitious obſeruation. Only we muſt take heed that we beleeue not her
Adſuerſaries definition of ſuperſtition, for they would imply therein al
true religion.}
ſuperſtitious. \V For paſſing by and ſeeing your
\SNote{The Aduerſaries (in the
\Cite{new Teſt.~1580.})
tranſlate, \Emph{your deuotions} moſt coruptly againſt the nature of the
Greeke word
\XRef{(2.~Theſ.~2,~4.)}
and moſt wickedly, againſt the laudable deuotion of good Chriſtians,
calling the Pagans idolatrie and ſuperſtition their deuotiõ.}
\TNote{\G{τὰ σεβάσματα}}
Idols, I found an altar alſo where vpon was written: \Emph{To the
vnknowen God}. That therfore which you worſhip, not knowing it, the ſame
doe I preach to you. \V The God that made the world and al things that
are in it, he being Lord of Heauen & earth dwelleth
\SNote{God is not concluded in Temples, nor needeth them for his
neceſſitie of dwelling or other vſes of indigence. See
\XRef{Annot. c.~7. Act. v.~48.}}
not in
\CNote{\XRef{Act.~7,~48.}}
Temples made with hand, \V neither is he ſerued with mens hands, needing
any thing, whereas himſelf giueth life vnto all, and breathing, and al
things: \V and he made of one, al mankind, to inhabite vpon the whole
face of the earth, aſſigning ſet times, and the limits of their
habitation, \V for to ſeeke God, if happily they may feele or find him;
although he be not farre from euery one of vs: \V For in him we liue and
moue and be, as certaine alſo of your owne
\MNote{Aratus.}
Poetes ſaid, \Emph{For of his kind alſo we are}. \V Being therfore of
Gods kind, we may not ſuppoſe,
\LNote{The Diuinitie to be like.}{Nothing can be made by mans hand of
what forme or ſort ſoeuer, that is like to Gods eſſence, or to the forme
or shape of his Godhead or Diuinitie. Therfore howſoeuer the Heathens
did paint or graue their Idols, they were nothing like to God. And this
alſo is impertinently alleadged by Heretikes againſt the Churches
images: which are not made, either to be adored with godly honour, or to
be any reſemblence of the Diuinitie or any of the three Perſons in
Godhead, but only of Chriſt as he was in forme of man, who in that
reſpect may be truly expreſſed, as other men by their portraictes: and
of the Holy Ghoſt, not as he is in him ſelf, but as he appeared in firy
tongues or in the ſimilitude of a doue, or ſuch like.
\MNote{How there may be Images or reſemblãces of the three Perſons
\Fix{in in}{in}{obvious typo, fixed in other}
Trinitie and of Angels.}
And ſo to paint or graue any of the three Perſons as they appeared
viſibly and corporally, is no more inconuenient or vnlawful, then it was
vndecent for them to appeare in ſuch formes. And therfore to paint or
portraict the Father alſo being the firſt Perſon, as he hath shewed
himſelf in viſion to any of the Prophets of the old or new Teſtament
(namely to
\CNote{\XRef{Dan.~7,~22.}}
Daniel as an old man) or the three Angels repreſenting the three Perſons
to
\CNote{\XRef{Gen.~18,~2.}}
Abraham, or the one Angel that wraſtling with
\CNote{\XRef{Gen.~32,~24.}}
Iacob bare our Lords Perſon, no ſuch thing is any where forbidden, but
is very agreable to the peoples inſtruction. In which ſort the Angels
were commonly pourtraicted (and namely the
\CNote{\XRef{Exo.~37,~7.}}
Cherubins ouer the Propitiatorie) as they be now in the Church, not in
their natural forme, but with corporal wings (as the Seraphins appeared
to
\CNote{\XRef{Eſa.~6,~2.}}
Eſay the Prophet) to expreſſe their qualitie and office of being God's
Angels, that is, \Emph{Meſſengers}: and God the Father with the world in
his hand, to ſignifie his creation and gouernement of the ſame, and ſuch
like: whereof the people being wel inſtructed may take much good, and no
harme in the world, being now through their faith in Chriſt farre from
al fond imagination of the falſe Gods of the Pagans. And therfore
S.~Gregorie ſaith of the Churches Images:
\CNote{\Cite{Greg. li.~9. ep.~9.}}
\MNote{Images are for the peoples inſtruction.}
\Emph{That which ſcripture or writing doth to the readers, the ſame doth
the pictures to the ſimple that looke therevpon, for in it euen the
ignorant ſee what they ought to follow: in it they doe read, that know
no letters.} Where he calleth it a matter of antiquitie and very
conuenient, that in holy places Images were painted to the peoples
inſtruction, ſo they be taught that they may not be adored with diuine
honour. And he in the ſame place sharply rebuketh Serenus the Bishop of
Maſsilia, that of indiſcrete zele he would take away Images, rather then
teach the people how to vſe them.}
the Diuinitie to be like vnto gold or ſiluer, or ſtone, the grauing of
art and deuiſe of man. \V And the times truly of this ignorance whereas
God diſpiſed, now he denounceth vnto men that al euery
%%% o-2444
where doe
penance, \V for that he hath appointed a day wherein he wil iudge the
world in equitie, by a man whom he hath appointed, giuing al men faith,
raiſing him vp from the dead.

\V And whẽ they had heard the reſurrectiõ of the dead, certaine indeed
mocked, but certaine ſaid: We wil heare thee againe concerning this
point. \V So Paul went forth out of the middes of them. \V But certaine
men ioyning vnto him, did beleeue: among whom was alſo
\MNote{Dionyſius Areopagita.}
\LNote{Dionyſius Areopagita.}{This
\MNote{S.~Dionyſius Areopagita is al for the Catholikes.}
is that famous Denys that firſt conuerted France, and wrote thoſe
notable & diuine workes
\Cite{De Eccleſiaſtica & Cæleſti hierarchia},
\Cite{de diuinis nominibus},
& others in which he confirmeth and proueth plainly almoſt al things
that the Church now vſeth in the miniſtratiõ of the holy Sacramẽts, &
affirmeth that he learned them of the Apoſtles, giuing alſo teſtimonie
for the Catholike faith in moſt things now controuerted, ſo plainly,
that our Aduerſaries haue no shift but to deny this Denys to haue been
the authour of them; feyning that they be an others of later age. Which
is an old flight of Heretikes, but moſt proper to theſe of al
others. Who ſeeing al antiquitie againſt them, are forced to be more
bold or rather impudent then others in that point.}
Dionyſius Areopagita, and a woman named Damaris, and others with them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

