%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 2573
%%% o-2417
\startChapter[
  title={Chapter 10}
  ]

\Summary{Becauſe
\MNote{The 4.~part.

The propagation of the Church to the Gentils alſo.}
the Iewes ſo much abhorred the Gentils, for the better warrant of their
Chriſtning, an Angel appeareth to Cornelius the deuout Italian, 9.~and a
viſion is shewed to Peter himſelf (the Cheefe and Paſtour of al) 19.~and
the Spirit ſpeaketh to him, 34.~yea and as he is Catechizing them
about \Sc{Iesvs}, 44.~the
holy Ghoſt commeth viſibly vpon them: and therfore not fearing any
longer the offenſe of the Iewes, he commandeth to baptize them.}

And there was a certaine man in Cæſarea, named Cornelius, Centurion of
that which is called the Italian band, \V religious, & fearing God with
al his houſe,
\LNote{Doing many almes-deedes.}{\Emph{He
%%% !!! I follow other in this note.
\MNote{Good workes before faith, are preparatiues to the ſame, not
properly meritorious.}
knevv God Creatour of al, but that his omnipotent Sonne vvas incarnate,
he knevv not: and in that faith he made praiers and gaue almes vvhich
pleaſed God, and by vvel doing he deſerued to knovv God perfectly, to
beleeue the myſterie of the Incarnation, and to come to the Sacrament of
Baptiſme.} So ſaith
\CNote{\Cite{Bed. in hunc locum.}}
Venerable Bede out of S.~Gregorie. And S.~Auguſtin thus,
\Cite{li.~1. de Bapt. c.~3.}
\Emph{Becauſe vvhatſoeuer goodnes he had in praiers and almes, the ſame
could not profite him vnles he vvere by the band of Chriſtian Societie
and peace, incorporated to the Church, he is bidden to ſend vnto Peter,
that by him he may learne Chriſt, by him he may be baptized, &c.} Whereby it
appeareth that ſuch workes as are done before iuſtification, though they
ſuffice not to ſaluation yet be acceptable preparatiues to the grace of
iuſtification, and ſuch as moue God to mercie, as it might appeare alſo
by Gods like prouident mercifulnes
\CNote{\XRef{Act.~3.}}
to the Eunuche: though al ſuch workes preparitiue come of grace alſo;
otherwiſe they could neuer deſerue Gods hand of congruity or any otherwiſe
toward iuſtification.}
doing many almes-deeds to the people. And alwaies praying to God, \V he
ſaw in a viſion manifeſtly, about the ninth houre of the day, an Angel
of God comming in vnto him and ſaying to him: Cornelius. \V But he
beholding him, taken with feare, ſaid: Who art thou Lord? And he ſaid to
him: Thy prayers and thy almes-deeds are aſcended into remembrance in
the ſight of God. \V And now ſend men
%%% o-2418
vnto Ioppe, and cal hither one Simon that is ſurnamed Peter. \V He
lodgeth with one Simon a tanner, whoſe houſe is by the ſea ſide. He wil
tel thee what thou muſt doe. \V And when the Angel was departed
%%% 2574
that ſpake to him, he called two of his houſhold, and a ſoldiar that
feared our Lord, of them that were vnder him. \V To whom when he had
told al, he ſent them vnto Ioppe.

\V And the next day whiles they were going on their iourney, and drawing
nigh to the citie, Peter went vp into the higher partes,
\LNote{To pray about the ſixt houre.}{The
\MNote{The Canonical houres.}
houre is ſpecified, for that there were certaine appointed times of
prayer vſed in the Law which deuout perſons, according to the publike
ſeruice in the Temple, obſerued alſo priuately: and which the Apoſtles
and holy Church afterwards both kept and increaſed. Whereof thus writeth
S.~Cyprian very notably:
\CNote{\Cite{De Orat. Dom. nu.~15.}}
\Emph{In celebrating their praiers, vve find that the three children
vvith Daniel obſerued the third, ſixt, and ninth houre, as in Sacrament}
(or myſterie) \Emph{of the holy Trinitie, &c.} And a litle after,
\Emph{vvhich ſpaces of houres the vvorshippers of God ſpiritually} (or
myſtically) \Emph{determining long ſince, obſerued fit times to pray: &
afterward the thing became manifeſt, that it was for Sacrament} (or
myſterie) \Emph{that the iuſt ſo praied. For
\CNote{\XRef{Act.~2.}}
at the third houre the Holy
Ghoſt deſcended vpon the Apoſtles, fulfilling the grace of our Lords
promiſe, &
\CNote{\XRef{Act.~10.}}
at the ſixt houre Peter going vp to the higher roome of the houſe, was
both by voice and ſigne from God inſtructed, that al Nations should be
admitted to the grace of ſaluation, whereas of cleanſing the Gentils he
doubted before, and our Lord being crucified
\CNote{\XRef{Luc.~23.}}
at the ſixt houre, at the ninth washed away our ſinnes with his
bloud. But to vs (Dearly beloued) beſide the ſeaſons obſerued of old,
both the times and ſacraments of praying be increaſed, for we muſt
\MNote{Mattins.}
pray in the morning early, that the Reſurrection of our Lord may be
celebrated by morning praier: as of old the Holy Ghoſt deſigned in the
\CNote{\XRef{Pſal.~5.}}
Pſalme, ſaying, In the morning early wil I ſtand vp to thee, early in
the morning wilt thou heare my voice.}
\MNote{Euenſong.}
Toward the euening \Emph{alſo when the ſunne departeth, and the day
endeth, we muſt of neceſsitie pray againe.}

S.~Hierom alſo writing of
\CNote{\XRef{Dan.~6.}}
Daniels praying \Emph{three times in a day}, ſaith: \Emph{There are
three times, wherein we muſt bow our knees to God. The third, the ſixt,
and the ninth houre the Eccleſiaſtical tradition doth wel
vnderſtand. Moreouer at the third houre the Holy Ghoſt deſcended vpon
the Apoſtles; at the ſixt, Peter went vp into a higher chamber to pray;
at the ninth,
\CNote{\XRef{Act.~3.}}
Peter and Iohn went to the Temple.} Againe writing to Euſtochium a
virgin and Nonne
\Cite{ep.~22. c.~16.}
\Emph{Though the Apoſtle bid vs pray alwaies, and to holy perſons their
very ſleepe is praier; yet we muſt haue diſtinct houres of praier, that if
perhaps we be otherwiſe occupied, the very time may admonish vs of our
office or duty. The third, ſixt, ninth houre, morning early, and the
euening, no man can be ignorant of.} And to Demetrias
\Cite{ep.~8. c.~8.}
that in the Pſalmes and praier she muſt keepe \Emph{alwaies the third,
ſixt, ninth houre, euening, mid-night, and morning.} He hath the like
\CNote{ep.~7. c.~5.}
And
\Cite{(ep.~27. c.~10.)}
he telleth how Paula the Holy abbeſſe with her religious
Nonnes \Emph{ſang the Pſalter in order, in the morning, at the third,
ſixt, ninth houre, euening, mid-night}: by mid-night meaning the time of
Matins (therfore called \Emph{Nocturnes} agreably to S.~Cyprian
\Cite{de Orat. Do. num.~15.})
and by the morning, the firſt houre called \Emph{Prime}: al correſpondent
to the times & houres of Chriſts Paſſion, as in S.~Matthew is noted
\XRef{c.~26.~27.}
By al which we ſee, how agreable the vſe of the Churches ſeruice is euen
at this time to the Scriptures and primitiue Church: and how wicked the
\Fix{Purtian-Caluiniſts}{Puritan-Caluiniſts}{obvious typo, fixed in other}
be, that count al ſuch order and ſet ſeaſons of
praier, ſuperſtition: and laſtly, how inſufficient and vnlike the new
pretended Church ſeruice of England is to the primitiue vſe, which hath
no ſuch houres of night or day, ſauing a litle imitation of the old
Matins and Euenſong, and that in Schiſme and Hereſie, and therfore not
only vnprofitable, but alſo damnable.}
to pray about the ſixt houre. \V And being hungrie, he was deſirous to
take ſomewhat. And as they were preparing, there fel vpon him an exceſſe
of mind: \V and he ſaw the Heauen opened, and a certaine veſſel
deſcending, as it were a great linnen ſheete with foure corners let
downe from Heauen to the earth, \V wherein were al foure-footed beaſtes,
and that creepe on the earth, and foules of the aire. \V And there came
a voice to him: Ariſe Peter, kil, and eate. \V But Peter ſaid: God
forbid, Lord; for I did neuer eate any common and vncleane thing. \V And
\SNote{Here God firſt vttered to Peter that the time was come to preach alſo
to the Gentils, and to cõuerſe with them for their ſaluation no leſſe
then with the Iewes with ful freedom to eate al meates without reſpect
of the prohibitiõ of certaine, made in the old Law.}
a voice came to him againe the ſecond time: That which God hath
purified, doe not thou cal common. \V And this was done thriſe, &
forthwith the veſſel was taken vp againe into Heauen. \V And whiles
Peter doubted within himſelf, what the viſion ſhould be that he had
ſeen, behold the men that were ſent from Cornelius, inquiring for Simons
houſe, ſtood at the gate. \V And when they had called, they asked, if
Simon that is ſurnamed Peter were lodged there. \V And as Peter was
thinking of the viſion the Spirit ſaid to him: Behold three men doe
ſeeke thee. \V Ariſe therfore, and get thee downe, and goe with them,
doubting nothing; for I haue ſent them. \V And Peter going downe to the
men, ſaid: Behold I am he whom you ſeeke; what is the cauſe, for the
which you are come? \V Who ſaid: Cornelius the Centurion, a iuſt man and
that feareth God, and hauing teſtimonie of al the Nation of the
Iewes, receiued an anſwer of an holy Angel to ſend for thee into his
houſe, and to heare wordes of thee. \V Therfore bringing them in, he
lodged them.

\V And the day following he aroſe and went with them: and certaine of
the brethren of Ioppe accompanied him. \V And on the morow he entred
into Cæſarea. And Cornelius expected them, hauing called together his
kinne, and ſpecial
%%% o-2419
freinds. \V And it came to paſſe, when Peter was come in, Cornelius came
to meete him, & falling at his feete
\LNote{Adored.}{S.~Chryſoſtom
\MNote{Adoration of creatures.}
\Cite{ho.~21. in Act.}
thinketh Peter refuſed this adoration of humilitie only; becauſe euery
falling downe to the ground for worſhip ſake, is not Diuine worſhip or
due only to God,
\TNote{\G{προσεκύνησεν}}
the word of adoration and proſtration being commonly vſed in the
Scriptures toward men. But S.~Hierom
\Cite{adu. Vigil. c.~2. to.~2.}
rather thinketh that Cornelius by errour of Gentility, and of Peters
perſon, did goe about to adore him with Diuine honour, & therfore was
lifted vp by the Apoſtle, adding that he was but a man.}
adored. \V But Peter lifted him vp ſaying: Ariſe, my ſelf alſo am a
man. \V And talking with him, he went in, and findeth many that were
aſſembled. \V And he ſaid to them: You know how abominable it is for a
man that is a Iewe, to ioyne, or to approche vnto a ſtranger: but God
hath ſhewed to me, to cal no man common or vncleane. \V For the which
cauſe, making no doubt, I came when I was ſent for. I demand therfore,
for what cauſe you haue ſent for me? \V And Cornelius ſaid: Foure daies
ſince, vntil this houre, I was
\TNote{In the Greeke, faſting & praying.}
\SNote{At the time of praier ſpecially God ſendeth men comfortable
viſitations.}
praying the ninth houre in my houſe, and behold
\SNote{Note theſe apparitions & viſions to S.~Peter, Cornelius, &
others, in the Scripture very oftẽ, againſt the incredulitie of our
Heretikes, that wil beleeue neither viſion nor miracle, not expreſſed in
Scripture: theſe being 
\Fix{beleeved}{beleeued}{obvious typo, fixed in other}
of Chriſtian men euen before they were
written.}
a man ſtood before me in white apparel, \V and ſaid: Cornelius, thy
praier is heard, and thy almes-deeds are in memorie in the ſight of
God. \V Send therfore to Ioppe, and cal hither Simon that is ſurnamed
Peter: he lodgeth in the houſe of Simon a tanner by the ſea ſide. \V
Immediatly therfore I ſent to thee: and thou haſt done wel in
comming. Now therfore al we are preſent in thy ſight, to heare al things
whatſoeuer are commanded thee of the Lord.

%%% 2575
\V And Peter opening his mouth, ſaid: In very deed I perceiue that God
is not an accepter of perſons. \V But in euery Nation, he that feareth
him, and
\SNote{Not ſuch as beleeue only, but ſuch as feare God and worke
iuſtice, are acceptable to him.}
worketh iuſtice, is acceptable to him. \V The word did God ſend to the
children of Iſraël, preaching peace by \Sc{Iesvs Christ} (this is Lord
of al.) \V
\Fix{Yow}{You}{obvious typo, fixed in other}
know the word that hath been made through al Iewrie, for
\CNote{\XRef{Mat.~4,~12.}}
beginning from Galilee, after the Baptiſme which Iohn
preached: \V \Sc{Iesvs} of Nazareth how God anointed him with the Holy
Ghoſt and with power, who went throughout doing good and healing al that
were oppreſſed of the Diuel, becauſe God was with him. \V And we are
witneſſes of al things that he did in the countrie of the Iewes and in
Hieruſalem, whom they killed hanging him vpon a tree. \V Him God raiſed
vp the third day and
\LNote{Gaue him.}{Chriſt did not vtter his Reſurection and other
myſteries to al at once, and immediatly to the vulgar: but to a few
choſen men that should be the Gouernours of the reſt: inſtructing vs
thereby to take our faith and al neceſſarie things of ſaluation, at the
hands of our Superiours.}
gaue him to be made manifeſt, \V not to al the people, but to vs, who
did eate and drinke with him after he roſe againe from the dead. \V And
he commanded vs to preach to the people, & to teſtifie that it is he that of
God was appointed iudge of the liuing and of the dead. \V To him al the
Prophets giue teſtimonie, that al receiue remiſsion of ſinnes by his
name, which beleeue in him.

%%% o-2420
\V As Peter was yet ſpeaking theſe wordes, the Holy Ghoſt fel vpon al
that heard the word. \V And the faithful of the Circumciſion that came
with Peter, were aſtoniſhed, for that the grace of the Holy Ghoſt was
powred out vpon the Gentils alſo. \V For they heard them ſpeaking with
tongues, and magnifying God. Then Peter anſwered: \V Can any man forbid
water, that theſe ſhould not be
\LNote{Baptized, which haue receiued.}{Such
\MNote{They that are iuſtified before muſt not omit the Sacramẽts.}
may be the grace of God ſometimes toward men, and their charitie and
contrition ſo great, that they may haue remiſsion, iuſtification, and
ſanctification before the external Sacrament of Baptiſme, Confirmation,
or Penance be receiued; as we ſee in this example, where at Peters
preaching they al receiue the Holy Ghoſt before any Sacrament. But in
the ſame we learne one neceſſarie leſſon, that ſuch notwithſtanding muſt
needes receiue the Sacraments appointed by Chriſt, which whoſoeuer
contemneth, can neuer be iuſtified.
\Cite{Aug. ſuper Leuit. q.~84. to.~4.}}
baptized which haue receiued the Holy Ghoſt as wel as we? \V And he
commanded them to be baptized in the name of our Lord \Sc{Iesvs
Christ}. Then they deſired him that he would tarie with them certaine
daies.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

