%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 2555
%%% o-2396
\startChapter[
  title={Chapter 3}
  ]

\Summary{A miracle, and a Sermon of Peters to the people, shewing that
\Sc{Iesvs} is Chriſt, and exhorting them to faith in him and pennance
  for their ſinnes, and ſo they shal haue by him (in Baptiſme) the
  Benediction which was promiſed to Abraham.}

And Peter and Iohn went vp into the Temple, at the
\SNote{This maketh for diſtinction of Canonical houres and diuerſitie of
appointed times to pray in. See
\XRef{Annot. c.~10,~9.}}
ninth houre of praier. \V And a certaine man that was lame from his
mothers wombe, was caried; whom they laid euery day at the gate of the
Temple, that is called Specious, that he might aske almes of them that
went into the
%%% o-2397
Temple. \V He, when he had ſeen Peter and Iohn about to
enter into the Temple, asked to receiue an almes. \V But Peter with Iohn
looking vpon him, ſaid: Looke vpon vs. \V But he looked earneſtly vpon
them, hoping that he ſhould receiue ſomething of them. \V But Peter
ſaid: Siluer and gold I haue not, but
\LNote{That which I haue.}{This
\MNote{Saints doe miracles and the like, but by the power of God.}
power of working miracles was in Peter, and Peter properly did giue this
man his health, though he receiued that force and vertue of God, & in &
by him executed the ſame. Therfore he ſaith: \Emph{That which I haue, I
giue to thee.} And the Heretikes are ridiculous that note here, \Emph{a
miracle done by Chriſt by the handes of the Apoſtles}, to make the
ſimple beleeue that they had no more to doe then a dead inſtrument in
the workemans hand.}
that which I haue, the ſame I giue to thee: In the name of \Sc{Iesvs
Christ} of Nazareth ariſe, and walke. \V And taking his right hand, he
lifted him vp, and forthwith his feet & ſoles were made ſtrong. \V And
ſpringing he ſtood, & walked; & went in with them into the Temple
walking & leaping & praiſing God. \V And al the people ſaw him walking
and praiſing God. \V And they knew him, that it was he which ſate for
almes at the Specious gate of the Temple: and they were exceedingly
aſtoniſhed and agaſt at that that had chanced to him. \V And as he held
Peter and Iohn, al the people ranne to them vnto the porch which is
called Salomons, wondering.

%%% 2556
\V But Peter ſeeing them, made anſwer to the people: Ye men of Iſrael,
why maruel you at this, or why looke you vpon vs, as though
\LNote{By our power.}{When the Apoſtles remit ſinnes or doe any other
miracles, they doe it not by any humane, proper, or natural power in
themſelues: but of ſupernatural force giuen them from aboue, to proue
that the faith of Chriſt is true, and that he is God whom the Iewes
crucified, in whoſe name and faith they worke, and not in their owne.}
by our power or holines we haue made this man to walke? \V The God of
Abraham, and the God of Iſaac, and the God of Iacob, the God of our
Fathers hath glorified his Sonne \Sc{Iesvs}, whom you indeed deliuered &
denied before the face of Pilate, he iudging him to be releaſed. \V But
you denied the Holy and the Iuſt One,
\CNote{\XRef{Luc.~23,~18.}}
and asked
\Fix{a a}{a}{obvious typo, fixed in other}
mankiller to be giuen vnto you. \V But the Authour of life you killed,
whom God hath raiſed from the dead, of which we are witneſſes. \V And in the
\SNote{This faith was not the faith of the lame man (for he looked only
for almes) nor a ſpecial faith of the Apoſtles own ſaluation: but the
whole beleefe of Chriſtian Religion.}
faith of his name, this man whom you ſee and know, his name hath
ſtrengthned; & the faith which is by him, hath giuen this perfect health
in the ſight of al you.

\V And now (Brethren) I know that you did it through ignorance, as alſo
your Princes. \V But God who foreſhewed by the mouth of al the Prophets
that his \Sc{Christ} ſhould ſuffer, hath ſo fulfilled it. \V Be Penitent
therfore & cõuert, that your ſinnes may be put out. \V That, when the
times ſhal come of refreſhing by the ſight of our Lord, and he ſhal ſend
him that hath been preached vnto you \Sc{Iesvs Christ}. \V
\TNote{\G{ὃν δεῖ οὐρανὸν μὲν δέξασθαι}}
Whom
\SNote{Some Heretikes fouly corrupt this place, thus: \Emph{Who muſt be
conteined in Heauen}, of purpoſe (as they proteſt) to hold Chriſt in Heauen
from the B.~Sacrament.
\Cite{Beza.}
As though his preſence there, drew him out of Heauen. Neither can they
pretend the Greeke, which is word for word as in the vulgar Latin, and as
we tranſlate.}
Heauen truly muſt receiue vntil the times of the reſtitution of al
things, which God ſpake by the mouth of his holy Prophets from the
beginning of the world.
%%% o-2398
\V Moyſes indeed ſaid:
\CNote{\XRef{Deu.~18,~25.}}
\Emph{That a Prophet shal the Lord your God raiſe vp to you of your
brethren, as my ſelf: him you shal heare according to al things
whatſoeuer he shal ſpeake to you. \V And it shal be, euery ſoule that
shal not heare that Prophet, shal be deſtroied out of the people.} \V
And al the Prophets from Samuel and afterward that haue ſpoken, told of
theſe daies. \V You are the children of the Prophets and of the
Teſtament which God made to our Fathers, ſaying to Abraham:
\CNote{\XRef{Gen.~12,~18.}
\XRef{26,~4.}}
\Emph{And in thy ſeed shal al the families of the earth be bleſſed.} \V
To you firſt God raiſing vp his Sonne hath ſent him bleſſing you: that
euery one ſhould conuert himſelf from his naughtines.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

