%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-22


\project douay-rheims


%%% 2608
%%% o-2456
\startChapter[
  title={Chapter 22}
  ]

\Summary{Being licenſed by the Tribune to ſpeake to the people, he
  sheweth them that he was once as earneſt on that ſide as they now be:
  6.~and how ſtrange and miraculous his conuerſion was. 17.~They heare
  him quietly vntil he began to make mention of a viſion that ſent him
  away from them to the Gentils. 22.~Then they crie out vpon him ſo,
  23.~that for their crying the Tribune commandeth him to be
  ſcourged. 25.~Which yet by his wiſedom he eſcapeth.}

Men Brethren and Fathers, heare what account I doe render now vnto
you. \V (And when they had heard that he ſpake to them in the Hebrew
tongue, they did the more keep ſilence. \V And he ſaith:)
\CNote{\XRef{Act.~21,~39.}}
I am a man a Iewe, borne at Tarſus in Cilicia, but brought vp in this
citie, at the feet of Gamaliel inſtructed according to the veritie of
the law of the Fathers,
%%% o-2457
an emulatour of the Law as alſo al you are this day: \V who
\CNote{\XRef{Act.~8,~3.}}
perſecuted this way vnto death, binding & deliuering into cuſtodies men
& women, \V as the high Prieſt doth giue me teſtimonie, and al the
Ancients. \V Of whom
\CNote{\XRef{Act.~9,~1.}}
receiuing letters alſo to the Brethren, I went to Damaſcus, that I might
bring them thence bound to Hieruſalem, to be puniſhed. \V And it came to
paſſe as I was going, and drawing nigh to Damaſcus at mid-day, ſodenly
from Heauen there ſhone round about me much light: \V and falling on the
ground, I heard a voice ſaying to me: Saul, Saul, why perſecuteſt thou
me? \V And I anſwered: Who art thou Lord? And he ſaid to me: I
am \Sc{Iesvs} of Nazareth, whom thou perſecuteſt. \V And they that were
with me, ſaw
%%% 2609
the light indeed, but the voice they heard not of him that ſpake with
me. \V And I ſaid: What ſhal I doe Lord? And our Lord ſaid to me: Ariſe
and goe to Damaſcus; and there it ſhal be told thee of al things that
thou muſt doe. \V And whereas I did not ſee for the brightneſſe of that
light, being led of my companions by the hand, I came to Damaſcus. \V
And one Ananias, a man according to the Law, hauing teſtimonie of al the
Iewes
\Fix{inhabitans,}{inhabitants,}{likely typo, fixed in other}
\V comming to me, and ſtanding by me, ſaid to me: Brother Saul, looke
vp. And I the ſelf-ſame houre looked vpon him. \V But he ſaid: The God
of our Fathers hath preordained thee, that thou ſhouldeſt know his wil,
and ſee the Iuſt one, and heare a voice from his mouth: \V becauſe thou
ſhalt be his witnes to al men, of thoſe things which thou haſt ſeen and
heard. \V And now what tarieſt thou? Riſe vp, and be baptized, and
\SNote{The Sacrament of Baptiſme doth it ſelf
wash away ſinnes as here is plaine, & therfore doth not only ſignifie
(as the Heretikes affirme) that our ſinnes be forgiuen before, or
otherwiſe by faith only remitted. Wherby the Churches doctrine is proued
to be fully agreable to the Scriptures, that the Sacraments giue
grace \L{ex opere operato}, that is, by the force & vertue of the worke
and word, done & ſaid in the Sacrament.}
waſh way thy ſinnes inuocating his name. \V And it befel me returning
into Hieruſalem, and 
praying in the Temple, that I was in a trance, \V and ſaw him ſaying
vnto me: Make haſt, and depart quickely out of Hieruſalem: becauſe they
wil not receiue thy teſtimonie of me. \V And I ſaid: Lord, they know
that I did caſt into priſon and beat in euery Synagogue them that
beleeued in thee. \V And when the bloud of Steuen thy
\TNote{\G{μάρτυρός}}
Witnes was ſhed,
\CNote{\XRef{Act.~7,~38.}}
\SNote{Not only the Principals, but al that conſent to the
death or vexation of Chriſtian men for the Catholike faith, doe highly
offend: which the Apoſtle confeſſeth here, that Gods mercie may be more
notoriouſly glorified in him hereby.}
I ſtood by and conſented, and kept the garments of them that killed
him. \V And he ſaid to me: Goe, for into the Gentils a farre wil I ſend
thee.

\V And they heard him vntil this word, and they lifted vp their voice,
ſaying: Away with ſuch an one from the earth: for it is not meet he
ſhould liue. \V And when they cried
%%% o-2458
out, and threw of their garments, and caſt duſt into the aire, \V the
Tribune commanded him to be caried into the caſtel, and to be beaten
with whips, & that he ſhould be tormented: to know for what cauſe they
did ſo crie at him. \V And when they had bound him very ſtraight with
thongs, Paul ſaith to the Centurion ſtanding by him: Is it lawful for
you to whip a man that is a Roman and vncondemned? \V Which the
Centurion hearing, went to the Tribune, and told him, ſaying: What wilt
thou doe? for this man is a citizen of Rome. \V And the Tribune comming,
ſaid to him: Tel me, art thou a Romane? But he ſaid: Yea. \V And the
Tribune anſwered: I obtained this citie with a great ſumme. And Paul
ſaid: But I was alſo borne to it. \V Immediatly therfore they departed
from him that were to torment him. The Tribune alſo feared after he
vnderſtood that he was a citizen of Rome, and becauſe he had bound
him. \V But the next day meaning to know more diligently for what cauſe
he was accuſed of the Iewes, he looſed him, and commanded the
\TNote{\G{ἀρχιερεῖς}}
Prieſts to come together and al the Councel: & bringing forth Paul, he
ſet him among them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

