%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 2548
%%% o-2389
\startChapter[
  title={Chapter 1}
  ]

\Summary{Chriſt
\MNote{The 1.~part. The expectation of the Holy Ghoſt between the
  Aſcenſion of Chriſt, and the beginning of the Church.}
now ready to aſcend, biddeth the Apoſtles to expect the Holy Ghoſt which
he had promiſed, foretelling where (being 
\Fix{ſtrenthned}{ſtrengthened}{likely typo, fixed in other}
by him) they should begin his Church, and how far they should cary
it. 9.~After his Aſcenſion they are warned by two Angels to ſet their
mindes vpon his ſecond comming. 14.~In the daies of their expectation,
15.~Peter beginneth to execute his vicarship, giuing inſtruction and
order, by which Mathias is elected Apoſtle in the place of Iudas.}

The
\CNote{\XRef{Lu.~1,~3.}}
firſt treatiſe I made of
\SNote{Not al particularly, (for the other Euangeliſts write diuers
things not touched by him) but al the principal and moſt neceſſarie
things.}
al things, O Theophilus, which \Sc{Iesvs} began to doe and to teach, \V
vntil the day wherein
\LNote{Giuing commandement.}{He meaneth the power giuen them to preach,
to baptize, to remit ſinnes, and generally the whole commiſsion and
charge of gouernement of the Church after him, and in his name, ſteed, &
right: the which regiment was giuen them together with the Holy Ghoſt
to aſsiſt them therein for euer.}
giuing commandement by the Holy Ghoſt to the Apoſtles whom he choſe, he
was aſſumpted: \V to whom he ſhewed alſo himſelf aliue after his
paſſion in many arguments, for fourtie daies appearing to them, and
ſpeaking of the Kingdom of God. \V And eating with them,
\CNote{\XRef{Lu.~24,~49.}}
he commanded them, that they ſhould not depart from Hieruſalem, but
ſhould expect the promiſe of the Father, which you
\CNote{\XRef{Io.~14,~26.}}
haue heard (ſaith he) by my mouth: \V for Iohn indeed baptized with
\SNote{Iohns Baptiſme gaue not the Holy Ghoſt.}
water, but
\CNote{\XRef{Lu.~3,~16.}}
you ſhal be
\SNote{The aboundant powring of the Holy Ghoſt vpon them on whitſunday,
he calleth Baptiſme.}
baptized with the Holy Ghoſt after theſe few daies. \V They therfore
that were aſſembled, asked him, ſaying: Lord, whether at this time wilt
thou reſtore the Kingdom of Iſrael? \V but he ſaid to them:
\LNote{It is not for you.}{It
\MNote{The times and monents of things to come pertaine not to vs.}
is not for vs, nor needful for the Church, to know the times & moments
of the world, the comming of Antichriſt, and ſuch other Gods
ſecrets. This is enough in that caſe, to be aſſured that Chriſts faith
shal be preached, and the Church ſpred throughout al Nations, the Holy
Ghoſt concurring continually with the Apoſtles & their Succeſſours for
the ſame.}
It is not for you to know times or moments, which the Father hath put in
his owne power: \V but you ſhal receiue the
\CNote{\XRef{Lu.~24,~49.}
\XRef{Act.~2,~1.}}
vertue of the Holy Ghoſt comming vpon you, and you ſhal be witneſſes
vnto me in Hieruſalem, and in al Iewrie, and Samaria, and euen to the
vtmoſt of the earth. \V And
\CNote{\XRef{Mr.~16,~19.}
\XRef{Lu.~24,~31.}}
when he had ſaid theſe things, in their ſight he was eleuated: & a cloud
receiued him out of their ſight. \V And when they beheld him going into
Heauen, behold two men ſtood beſide them in white garments, \V who alſo
ſaid: Ye men of Galilee, why
%%% o-2390
ſtand you looking into Heauen? This \Sc{Iesvs} which is
\LNote{Aſſumpted from you.}{By
\MNote{Chriſt is aſcended, and yet really in the B.~Sacrament.}
this viſible Aſcending of Chriſt to Heauen and like returne from thence
to iudgement, the Heretikes doe incredulouſly argue him not to be in the
Sacrament. But let the faithful rather giue eare to S.~Chryſoſtome
ſaying thus: \Emph{O miracle! he that ſitteth with the Father in Heauen
aboue, at the very ſame time is handled of men beneath. Chriſt aſcending
to Heauen, both hath his flesh with him, and left it with vs
beneath. Elias being taken vp, left to his Diſciples his cloke only: but
the Sonne of man aſcending, left his owne flesh to vs.}
\Cite{Li.~3. de Sacerd.}
\Cite{Ho.~2. ad Po. Ant. in fine.}
\Cite{Ho. de diuit. & paup. in fine.}}
aſſumpted from you into Heauen, ſhal ſo come as you haue ſeen him going
into Heauen.

%%% 2549
\V Then they returned to Hieruſalem from the mount that is called
Oliuet, which is by Hieruſalem, diſtant a Sabboths iourney. \V And when
they were entred in, they went vp into an vpper chamber, where abode
\SNote{This viſible companie was the true Church of Chriſt, which he
left, & commanded to keepe together til the comming of the Holy Ghoſt,
by him to be further informed & furnished to gaine al Nations to the
ſame Societie.}
Peter & Iohn, Iames and Andrew, Philippe and Thomas, Bartholomew and
Matthew, Iames of Alphæus and Simon Zelotes, and Iude of Iames. \V Al
theſe were perſeuering with one mind in praier with the
\SNote{The Heretikes, ſome in the text, other in the margẽt, tranſlate,
\Emph{wiues}, to wit, of the Apoſtles, moſt impudently; knowing in their
conſciences that he meaneth the Maries & other holy women that followed
Chriſt as
\XRef{Lu.~8,~2.}
\XRef{24,~10.}
See Beza and the
\Cite{Engl. Bible. 1579.}}
women and
\LNote{\Sc{Marie} the mother of \Sc{Iesvs}.}{This
\MNote{Our \Sc{B.~Lady}.}
is the laſt mention that is made in holy Scripture, of our B.~Lady. For
though she were ful of al diuine wiſedom, & opened (no doubt) vnto the
Euangeliſtes and other Writers of holy Scriptures, diuerſe of Chriſts
actions, ſpeaches, and myſteries, wherof she had both experimental and
reuealed knowledge: yet for that she was a woman, and the humbleſt
creature liuing, and the paterne of al order and obedience, it pleaſed not
God that there should be any further note of her life, doings, or death,
in the Scriptures.
\MNote{Her life.}
She liued the reſt of her time with the Chriſtians (as here she is
peculiarly named and noted among them) and ſpecially with S.~Iohn the
Apoſtle,
\CNote{Io.~19,~26.~27.}
to whom our Lord recommended her: Who prouided for her al neceſſaries,
her ſpouſe Ioſeph (as it may be thought) being deceaſed before. The
common opinion is that she liued 63.~yeares in al.
\MNote{Her death.}
At the time of her death, (as
\CNote{\Cite{Dionyſ. ep ad Timotheum.}}
S.~Denys firſt, & after him S.~Damaſcene
\Cite{de dormit. Deipara.}
writeth) al the Apoſtles then diſperſed into diuers Nations to preach
the Ghoſpel, were miraculouſly brought together (ſauing S.~Thomas who
came the third day after) to Hieruſalem, to honour her diuine departure
and funeral, as the ſaid S.~Denys witneſſeth. Who ſaith that himſelf,
S.~Timothee, and S.~Hierotheus were preſent: teſtifying alſo of his owne
hearing, that both before her death and after for three daies, not only
the Apoſtles and other holy men preſent, but the Angels alſo and Powers
of Heauen did ſing moſt melodious Hymnes. They buried her ſacred body in
Gethſemani.
\MNote{Her \Sc{Assvmption}.}
But for S.~Thomas ſake, who deſired to ſee & to reuerence it, they
opened the ſepulcher the third day; and finding it void of the holy
body, but exceedingly fragrant, they returned, aſſuredly deeming that
her body was aſſumpted into Heauen; as the Church of God holdeth: being
moſt agreable to the ſingular priuiledge of the mother of God, &
therfore celebrateth moſt ſolemnly the day of her Aſſumption. And that
is conſonant not only to the ſaid S.~Denys and S.~Damaſcene, but to Holy
Athanaſius alſo, who auoucheth the ſame,
\Cite{Serm. in Euang. de Deipara.}
Of which Aſſumption of her body, S.~Bernard alſo wrote fiue notable
Sermons extent in his workes.

But
\MNote{The
\Fix{Proteſtãs}{Proteſtãts}{obvious typo, fixed in other}
haue no feaſt of her at al, as they haue of other Saints.}
neither theſe holy Fathers, nor the Churches tradition and teſtimonie
doe beare any ſway now a daies with the Proteſtants, that haue abolished
this greateſt feaſt of her Aſſumption, who of reaſon should at the leaſt
celebrate it as the day of her death, as they doe of other Saints. For
though they beleeue not that her body is aſſumpted, yet they wil not (we
trow) deny that she is dead, & her ſoule in glorie: neither can they
aske ſcriptures for that, no more then they require for the deaths of
Peter, Paul, Iohn, & other, which be not mentioned in ſcriptures, & yet
are ſtil celebrated by the Proteſtants. But concerning the
B.~Virgin \Sc{Marie}, they haue blotted out alſo both her Natiuitie, &
her Conception: So as it may be thought the Diuel beareth a ſpecial
malice to this woman whoſe ſeed brake his head. For as for the other two
daies of her Purification & Annunciation, they be not proper to our
Lady, but the one to Chriſts Conception, the other to his
Preſentation. So that she by this meanes shal haue no feſtiuitie at al.

But
\MNote{How the Primitiue Church & ancient Fathers honoured our B.~Lady.}
contrawiſe, to conſider how the ancient Church & Fathers eſteemed,
ſpake, and wrote of this excellent veſſel of grace, may make vs deteſt
theſe mens impietie, that can not abide the praiſes of her
\CNote{\XRef{Lu.~1,~49.}}
whom al Generations should cal \Sc{Blessed}, & that eſteeme her honours
a derogation to her Sonne. Some of their ſpeaches we wil ſet downe, that
al mẽ may ſee, that we neither praiſe her, nor pray to her more amply
then they did.
\MNote{S.~Athanaſius.}
S.~Athanaſius in the place alleaged, after he had declared how al the
Angelical Spirits & euery order of thẽ honoured & praiſed her with
the \Sc{Ave}, wherewith S.~Gabriel ſaluted her: \Emph{We alſo}, ſaith he,
\Emph{of al degrees vpon the earth extol thee with loude voice ſaying:
\L{Aue gratia plena &c.} Haile ful of grace, our Lord is with thee. Pray
for vs, o Maiſtreſſe, and Lady, and Queene, and mother of God.}
\MNote{Saint Ephrem.}
Moſt holy & ancient Ephrem, alſo in a ſpecial oration made in praiſe of
our Lady, ſaith thus in diuerſe places thereof: \Emph{\L{Intemerata
Deipara &c.} Mother of God vndefiled, Queene of al, the hope of them
that deſpaire, my Lady moſt glorious, higher then the Heauenly Spirits,
more honourable then the Cherubins, holier then the Seraphins, &
without compariſon more glorious then the ſupernatural hoſtes, the hope
of the Fathers, the glorie of the Prophets, the praiſe of the Apoſtles.}
And a litle after: \Emph{\L{Virgo ante partum, in partu, & poſt partum};
by thee we are reconciled to Chriſt my God, thy ſonne: thou art the
helper of ſinners, thou the hauen for them that are toſſed with ſtormes,
the ſolace of the world, the deliuerer of the empriſoned, the help of
orphans, the redemption of captiues.} And afterward \Emph{Vouchſafe me
thy ſeruant to praiſe thee. Haile Lady \Sc{Marie} ful of grace, haile
Virgin moſt Bleſſed among women.} And much more in that ſenſe which were
to long too repeate.

S.~Cyril
\MNote{Saint Cyril.}
hath the like vvonderful ſpeaches of her honour,
\Cite{hom.~6. contra Neſtorium.}
\Emph{Praiſe and glorie be to thee, o holy Trinitie: to thee alſo be
praiſe, holy mother of God, for thou art the pretious pearle of the
world, thou the candel of vnquencheable light, the crowne of Virginitie,
the ſcepter of the Catholike faith. By thee the Trinitie is glorified
and adored in al the world: by thee Heauen reioyceth, Angels &
Archangels are glad, Diuels are put to flight, and man is called againe
to Heauen, and euery creature that was held with the errour of Idols, is
turned to the knowledge of the truth. By thee churches are founded
through the world: thou being their helper, the Gentils come to
penance.} and much more which we omit.
\MNote{The Greeke Liturgies of S.~Iames, S.~Baſil, S.~Chryſoſtom.}
Likewiſe the Greeke Liturgies or Maſſes of S.~Iames, S.~Baſil, and
S.~Chryſoſtom, make moſt honourable mention of our B.~Lady, praying vnto
her, ſaluting her with the Angelical hymne, \L{Aue Maria}, and vſing
theſe ſpeaches: \Emph{Moſt holy, vndefiled, bleſſed aboue al, our
Queene, our Lady, the mother of God, \Sc{Marie}, a virgin foreuer, the
ſacred arke of Chriſts Incarnation, broder then the Heauens that didſt
beare thy Creatour; holy mother, of vnſpeakable light, we magnifie thee
with Angelical hymnes. Al things paſſe vnderſtanding, al things are
glorious in thee, o mother of God. By thee are the myſterie before
vnknowen to the Angels is made manifeſt & reuealed to thẽ on the earth,
thou art more honourable then the Cherubins, and more glorious then the
Saraphins. To thee, O ful of grace, al
creatures, both men and Angels doe gratulate and reioyce: glorie be to
thee, Which art a ſanctified temple, a ſpiritual Paradiſe, the glorie of
Virgins, of whom God tooke flesh and made thy womb to be his throne,
&c.}

And
\MNote{S.~Auguſtine.}
S.~Auguſtine
\Cite{Serm.~18. de Sanctis to.~10.}:
or (as ſome thinke) S.~Fulgentius: \Emph{O Bleſſed \Sc{Marie}, who can be
able worthily to praiſe or thanke thee? Receiue our praiers, obtaine vs
our requeſtes for thou art the
\TNote{\L{vnica ſpes.}}
ſpecial hope of ſinners. By thee we hope for pardon of our ſinnes, & in
thee, o moſt Bleſſed, is the expectation of our rewards.} And then
follow theſe wordes now vſed in the Churches ſeruice: 
\MNote{\L{Sancta Maria ſuccurre miſeris, &c.}}
\Emph{\L{Sancta Maria
ſuccurre miſeris, iuua puſillanimes, refoue ſlebiles, ora pro populo,
interueni pro clero, intercede pro deuoto fœmineo ſexu. Sentiant omnes
tuum iuuamen, quicunque celebrant tuam commemorationem.} Pray thou
continually for the people of God which didſt deſerue to beare the
Redeemer of the world, who liueth and reigneth for euer.}
\MNote{S.~Damaſcene.}
S.~Damaſcene alſo
\Cite{ſer. de dormitione Deiparæ.}
\Emph{Let vs crie with Gabriel: \L{Aue gratia plena}, Haile ful of
grace, Haile ſea of ioy that can not be emptied; haile the
\TNote{\L{Vnicũ leuamen.}}
only eaſe of greefes, haile holy Virgin, by whom death was expelled, and
life brought in.}
\MNote{S.~Irenæus.}
See S.~Ireneus
\Cite{li.~3. c.~33.}
&
\Cite{li.~5. circa medium.}
& S.~Aug.
\Cite{de fide}
&
\Cite{Symbolo}
&
\Cite{de agone Chriſtiano.}
\MNote{As Adam and Eue, ſo Chriſt & our Lady.}
Where they declare how both the ſexes concurre to our ſaluation, the man
and the woman, Chriſt and our Lady; as Adam & Eue both were the cauſe of
our fal: though Adam farre more then his wife, and ſo Chriſt farre more
excellently and in another ſort then our Lady: who (though his mother)
yet is but his creature and handmaid, himſelf being truly both God &
man. In al which places alleaged &
\CNote{See S.~Greg. Naziã. in fi. Trag. Chriſtus patiens.}
many other like to theſe, if it pleaſe the Reader to ſee and read, &
make his owne eyes witneſſes, he shal perceiue that there is much more
ſaid of her, & to her, then we haue here recited, and that the very ſame
or the like ſpeaches & termes were vſed then, that the Church vſeth now,
in the honour and inuocation of the B.~Virgin: to the confuſion of al
thoſe that wilfully wil not vnderſtand in what ſenſe al ſuch ſpeaches
are applied vnto her:
\MNote{The meaning of the titles & termes giuen to our B.~Lady.}
to wit, either becauſe of her praier and
interceſsion for vs, whereby she is our hope, our refuge, our aduocate
&c. or becauſe she brought forth the Authour of our redemption &
ſaluation, whereby she is the mother of mercie, and grace, & life, and
whatſoeuer goodnes we receiue by Chriſt.}
\Sc{Marie} the mother of \Sc{Iesvs}, and his brethren. \V In thoſe daies
\LNote{Peter riſing vp.}{Peter
\MNote{Peter beginneth to practiſe his Primacie.}
in the meane time practiſed his Superioritie in the companie or Church,
publishing an election to be made of one to ſupply Iudas roome. Which
Peter did not vpon commandement of Chriſt written, but by ſuggeſtion of
Gods Spirit and by vnderſtanding the Scriptures of the old Teſtament to
that purpoſe: the ſenſe whereof Chriſt had opened to the Apoſtles before
his departure, though in more ful manner afterward at the ſending of the
Holy Ghoſt. And this acte of Peter in preſcribing to the Apoſtles and
the reſt, this election, and the manner thereof, is ſo euident for his
Supremacie, that
\CNote{Beza in
\Cite{no. Teſt. Græcol. an.~1565.}}
the Aduerſaries confeſſe here that he was, \L{Antiſtes}, the cheefe &
Bishop of this whole College and companie.}
Peter riſing vp in the middes of the Brethren, ſaid: (and the multitude
of perſons together, was almoſt an hundred and twentie.) \V You men,
Brethren, the
\CNote{\XRef{Pſ.~40,~9.}
\XRef{Io.~13,~18.}}
ſcripture muſt be fulfilled which the Holy Ghoſt ſpake before by the
mouth of Dauid concerning Iudas, who was the
\CNote{\XRef{Luc.~22,~47.}
\XRef{Io.~18,~3.}}
captaine of them that apprehended \Sc{Iesvs}: \V who was numbred among
vs and obteined the lot of this miniſterie. \V And he indeed hath poſſeſſed
a
\CNote{\XRef{Mt.~27,~5.~7.}}
field of the reward of iniquitie, and being hanged he burſt in the
middes, and al his bowels guſhed out. \V And it was made notorious to al
the inhabitants of Hieruſalem: ſo that the ſame field was called in
their tõgue, \HH{Hacel-dema}, that is to ſay, the field of bloud. \V For
it is written in the booke of Pſalmes:
\CNote{\XRef{Pſa.~68,~26.}}
\Emph{Be their habitation made deſert, & be there none to dwel in it.}
\CNote{\XRef{Pſ.~108,~8.}}
\Emph{And his Bishoprike let another take.} \V Therfore, of theſe men
that haue aſſembled with vs, al the time that our Lord \Sc{Iesvs} went
in and went out among vs, \V beginning from the Baptiſme of Iohn vntil
the day wherein he was aſſumpted from vs,
\SNote{No ſmal myſterie, that the number of the twelue Apoſtles muſt
needes be made vp againe.}
there muſt one of theſe be made a witnes with vs of his reſurrection.

\V And they appointed two, Ioſeph, who was called Barſabas, who was
ſurnamed Iuſtus: and Matthias. \V And praying they ſaid: Thou Lord that
knoweſt the harts of al men, ſhew of theſe two, one, whom thou haſt
choſen, \V to take the place of this miniſterie & Apoſtleſhip, from the
which Iudas hath preuaricated that he might goe to his owne place. And
they gaue them
\LNote{Lots.}{When
\MNote{Caſting of lottes.}
the euent or fal of the lot is not expected of Diuels, nor of the
ſtarres, nor of any force of fortune, but looked and praied for to be
directed by God and his holy Saints: thẽ may lots be vſed lawfully. And
ſometimes to diſcerne betwixt two things mere indifferent, they may be
neceſſarie, as S.~Auguſtine teacheth.
\Cite{ep.~180. ad Honoratium.}}
lots, and the lot fel vpon Matthias, and he was numbred with the eleuen
Apoſtles.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

