%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-06


\project douay-rheims


%%% 2562
%%% o-2404
\startChapter[
  title={Chapter 6}
  ]

\Summary{By occaſion of a murmur in the Church (whoſe number now is ſo
  growen that it can not be numbred) Seuen of them being ordered by the
  Apoſtles in the holy order of Deacons; 8.~one of them, Steuen, worketh
  great miracles: and is by ſuch as he confounded in diſputation,
  falſely accuſed in the Councel of blaſphemie againſt the Temple and
  rites thereof.}

And in thoſe daies the number of Diſciples increaſing, there aroſe a
\LNote{Murmuring.}{It
\MNote{Murmuring & emulation.}
commeth of humane infirmitie, that in euery
Society of men (be it neuer ſo holy) there is ſome cauſe giuen or taken
by the weake, of murmur & difference, which muſt euer be prouided for &
ſtaied in the beginning, leſt it grow to further ſchiſme or
ſedition. And to al ſuch defects, the more the Church increaſeth in
number and diuerſitie of men and Prouinces, the more it is ſubiect. In
al which things the ſpiritual Magiſtrates, by the Apoſtles example and
Authoritie, muſt take order, as time and occaſion ſhal require.}
murmuring of the Greekes againſt the Hebrewes, for that their widowes
were deſpiſed in the daily miniſterie. \V And the Twelue calling
together the multitude of the Diſciples, ſaid: It is not reaſon, that we
leaue the word of God, and ſerue tables. \V Conſider therfore Brethren,
\LNote{Seuen men.}{We
\MNote{The 7.~Deacons.}
may not thinke that theſe ſeuen (here made Deacons) were only choſen to
ſerue profane tables, or diſpoſe of the Churches mere temporals, though
by that occaſion only they may ſeeme to ſome now elected, no expreſſe
mention being made of any other function. For, diuers circumſtances of
this ſame place giue euidence, and ſo doth al antiquitie, that their
Office ſtood not principally about profane things, but about the holy
Altar. The perſons to be elected, muſt be ful of the Holy Ghoſt and
wiſedom, they muſt after publike praier be ordered and conſecrated by
the Apoſtles impoſition of hands, as Biſhops and Prieſts were afterward
ordered.
\XRef{ep. ad Tim.}
where
\CNote{\XRef{1.~Tim.~3.}
\XRef{&~4.}}
S.~Paul alſo requireth in a manner the ſame conditions in them as in
Biſhops. Al which would not haue beene preſcribed for any ſecular
ſtewardſhip. Yea ſtraight vpon their Ordering here (no doubt by
commiſſion of the Apoſtles, which they had not before their election)
they preached, baptized, diſputed, and as it may appeare by 
\CNote{\XRef{Act.~6,~8.}}
the wordes ſpoken of S.~Steuen, that he was ful of grace and fortitude,
they receiued great increaſe of grace by their Deaconſhip.

But
\MNote{The office of Deacons.}
S.~Ignatius
\Cite{ep. ad Tral.}
can beſt witnes of their office and the Apoſtles manner and meaning in
ſuch things, who writeth thus: \Emph{It behoueth alſo to pleaſe by al
meanes the Deacons, vvhich are for the miniſterie of} \Sc{Iesvs
Christ}. \Emph{For they are not ſeruitours of meate and drinke, but
Miniſters of the Church of God. For vvhat are Deacons but imitatours or
follovvers of Chriſt, miniſtring to the Biſhops, as Chriſt to his
Father, & vvorking vnto him a cleane and immaculate vvorke, euen as
S.~Steuen to S.~Iames &c.} S.~Polycarpe hath the like in his
\Cite{epiſtle as Philippenſes.}
And
\CNote{\Cite{Li. Eccl. Hier. c.~3. part.~2.}}
S.~Denys writeth that their office was about the Altar, and putting
the holy bread and chalice vpon the ſame. S.~Clement alſo
\Cite{(Apoſt. Conſt. li.~2. c.~61.)}
that their Office among other things, is to aſſiſt the Biſhops, and read
the Ghoſpel in the Seruice &c. S.~Cyprian in diuers places
\Cite{ep.~65.}
&
\Cite{ep.~49. ad Cornel.}
calleth Deacons, the Churches and the Apoſtles Miniſters, and their
office \L{adminiſtrationem ſacram}, an holy adminiſtration. S.~Hierom
affirmeth, in
\Cite{caput.~7. Michae.}
and in
\Cite{epiſto.~85. ad Euangrium tom.~2.}
where he checketh ſome of them for preferring themſelues before Prieſts,
and putteth them in remembrance of their firſt calling, that they be as
the Leuites were in reſpect of the Prieſts of the old Law. Finally by
S.~Ambroſe
\Cite{li.~1. Offic. c.~41.}
and Prudentius in
\Cite{Hymno de S.~Laurent.}
ſpeaking of S.~Laurence the Deacon, we may ſee their Office was moſt
holy. See S.~Auguſtine alſo of the dignitie of Deacons.
\Cite{ep.~148. ad Valerium Conc. Carthag.~4. can.~37, 28.~39.~41.}}
ſeuen men of you of good teſtimonie, ful of the holy Ghoſt and wiſedom,
whom we may appoint ouer this buſines. \V But we wil be inſtant in
prayer and the miniſterie of the word. \V And the ſaying was liked
before al the multitude.
\MNote{The election of the 7.~firſt Deacons.}
And they choſe Steuen a man ful of faith and of the holy Ghoſt, and
Philippe, and Próchorus, and Nicanor, and Timon, and Pármenas, and
Nicolas a
\TNote{\G{προσήλυτον}}
ſtranger of Antioch. \V Theſe they did ſet in the preſence of the
Apoſtles: and praying
\SNote{Order of Deacõs giuen by impoſition of handes. See
\XRef{Annot. Act.~13,~3.}}
they impoſed handes vpon them. \V And the word of God increaſed, and the
number of the Diſciples was multiplied in Hieruſalem exceedingly: a
great multitude
%%% o-2405
alſo of the
\SNote{Now alſo the Prieſts and they of greater knowledge and eſtimation
began to beleeue.}
Prieſts obeied the faith.

\V And Steuen ful of grace and fortitude did great wonders & ſignes
among the people. \V And there aroſe certaine of that which is called
the Synagogue of the Libertines, and of the Cyrenians, and of the
Alexandrians, and of them that were of Cilicia and Aſia, diſputing with
Steuen: \V and they could not reſiſt the wiſedom and the Spirit that
ſpake. \V Then they ſuborned men, to ſay they had heard him ſpeake
wordes of blaſphemie againſt Moyſes and God. \V They therfore ſtirred vp
the people, and the Ancients, and the Scribes: and running together they
tooke him, and brought him into the Councel, \V and they ſet falſe
witneſſes that ſaid: This man ceaſeth not to ſpeake wordes againſt the
holy place & the Law. \V For we haue heard him ſay, that this
ſame \Sc{Iesvs} of Nazareth ſhal deſtroy this place, and ſhal change the
traditions, which Moyſes deliuered vnto vs. \V And al that ſate in the
Councel beholding him, ſaw his face as it were
\SNote{Such is the face of al conſtant & chereful Martyrs, to their
perſecutours & iudges.}
the face of an Angel.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

