%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-21


\project douay-rheims


%%% 2606
%%% o-2453
\startChapter[
  title={Chapter 21}
  ]

\Summary{From Miletum going on his iourney, 4.~he can not be diſſuaded
  neither at Tyre, 8.~nor at Cæſarea (in both which places the Holy Ghoſt
  reuealed how he should be handled in Hieruſalem, 10.~the Prophet
  Agabus expreſly foretelling that the Iewes there should deliuer him to
  the Gentils) 15.~but to Hieruſalem he commeth: where being welcome to
  the Chriſtians, and namely to Iames the Bishop, and to the Prieſts,
  while he goeth about to ſatisfie the Chriſtian Iewes there, who had
  been miſinformed of him as if he had taught it to be vnlawful for the
  Iewes to keepe Moyſes Law; 27.~he is inuaded by the infidel Iewes, and
  ready to be murdered by them, vntil the Romane ſouldiars doe reſcue
  him.}

And when it came to paſſe that we ſailed, being caried from them, with a
ſtraight courſe we came to Coos, and the day following to Rhodes, and
from thence to Patara. \V And when we had found a ſhip that paſſed ouer
to Phœnice, going vp into it we ſailed. \V And when we were in the ſight
of Cypres, leauing it on the left hand, we ſailed
%%% o-2454
into Syria, and came to Tyre: for there the ſhip was to diſcharge her
load. \V And finding Diſciples, we taried there ſeuen daies: who ſaid to
Paul by the Spirit, that he ſhould not goe vp to Hieruſalem. \V And the
dayes being expired, departing we went forward, al bringing vs on the
way, with their wiues and children, til we were out of the citie: and
falling vpon our knees on the ſhore, we praied. \V And when we had bid
one another farewel, we went vp into the ſhip; and they returned vnto
their owne. \V But we hauing ended the nauigation, from Tyre came downe
to Ptolomais: and ſaluting the Brethren, we taried one day with them. \V
And the next day departing, we came to Cæſarea. And entring into the
houſe of
\CNote{\XRef{Act.~6,~5.}}
Philip the Euangeliſt, who was one of the ſeuen, we taried with him. \V
And he had
\SNote{As S.~Peter had a wife, but vſed her not after his calling, as it
is noted elſwhere out of
\Cite{S.~Hierom, Luc.~4,~38.}
ſo may it be ſaid of S.~Philip being Deacõ.}
foure daughters 
%%% !!! Only marked in other
\LNote{Virgins.}{S.~Luke
\MNote{Virgins.}
noteth ſpecially that his daughters were Virgins, meaning (no doubt)
that they were of the ſtate, profeſsion, or purpoſe of perpetual
virginitie, not only that they were yong maides vnmarried: and that they
were the rather for that, indowed with the guift of prophecie, as
S.~Hierom ſaith
\Cite{li.~2. adu. Iouin. c.~24.}
See
\Cite{Oecum. c.~29. in hunc locum.}}
virgins, that did prophecie.

\V And as we abode there for certaine daies, there came a certaine
Prophet from Iewrie, named Agabus. \V He, when he was come to vs, tooke
Paules girdle: and binding his owne feet & handes, he ſaid: Thus ſaith
the Holy Ghoſt: The man whoſe girdle this is, ſo ſhal the Iewes bind in
Hieruſalem, and ſhal deliuer him into the handes of the Gentils. \V
Which when we had heard, we and they that were of the ſame place,
deſired him that he would not goe vp to Hieruſalem. \V Then Paul
anſwered, and ſaid: What doe you, weeping and afflicting my hart? for I
am ready not only to be bound, but to die alſo in Hieruſalem for the
name of our Lord \Sc{Iesvs}. \V And when we could not perſuade him, we
ceaſed, ſaying: The wil of our Lord be done.

\V And after theſe daies, being prepared, we went vp to Hieruſalem. \V
And there came alſo of the Diſciples from Cæſarea with vs, bringing with
them one Iaſon a Cyprian (with whom we ſhould lodge) an old Diſciple. \V
And when we were come to Hieruſalem, the
%%% 2607
Brethren receiued vs gladly. \V And the day following Paul went in with
vs to Iames, and al the Ancients were aſſembled. \V Whom when he had
ſaluted, he told particularly what God had done among the Gentils by his
miniſterie. \V But they hearing it, magnified God, and ſaid to him: Thou
ſeeſt (Brother) how many thouſands there are among the Iewes that haue
beleeued: and al are Zelatours of the Law. \V But they haue heard of
thee that thou doeſt teach thoſe Iewes
%%% o-2455
that are among the Gentils, to
depart from Moyſes: ſaying that they ought not to circumciſe their
children, nor walke according to the cuſtome. \V What is it then? needes
muſt the multitude aſſemble: for they wil heare that thou art come. \V
Doe this therfore which we tel thee. There are with vs foure men, that
haue a vowe on them. \V Taking theſe vnto thee, ſanctifie thy ſelf with
them; and beſtow on them, that they may
\CNote{\XRef{Nu.~6,~18.}}
ſhaue their heads: and al ſhal know that the things which they heard of
thee, are falſe: but that thy ſelf alſo walkeſt
\LNote{Keeping the Law.}{Al
\MNote{Auoiding of ſcandal in things not vnlawful.}
the obſeruations of the Law were now in themſelues dead & vnprofitable;
yet til further propagation of the Ghoſpel, they were not damnable to
the keepers, nor offenſiue to God, but might be obſerued euen of the
Chriſtian Iewes. And for feare of ſcandalizing the weake of that Nation,
newly conuerted, or prone to receiue the faith, the Apoſtles by Gods
ſuggeſtion did thinke it good to obſerue them as occaſion required.}
keeping the Law. \V But concerning them that beleeue of the Gentils,
\CNote{\XRef{Act.~15,~20.}}
we haue written, decreeing that they ſhould refraine themſelues from the
immolated to Idols, and bloud, and ſuffocated, and fornication. \V Then
Paul taking the men vnto him, the next day being purified with them
entred into the Temple, ſhewing the accompliſhment of the
\CNote{\XRef{Nu.~6,~13.}}
daies of the purification, vntil an oblation was offered for euery one
of them.

\V But whiles the ſeuen daies were a finiſhing, thoſe Iewes that were of
Aſia, when they had ſeen him in the Temple, ſtirred vp al the people,
and laid handes vpon him, \V crying: Ye men of Iſrael, help: this is the
man that againſt the people and the Law and this place teaching al men euery
where, hath alſo moreouer brought in Gentils into the Temple, and hath
violated this holy place. (\V For they had ſeen Trophimus the Epheſian in
the citie with him, whom they ſuppoſed that Paul had brought into the
temple.) \V And the whole citie was in an vproare: and there was made a
concourſe of the people. And apprehending Paul, they drew him forth of
the Temple: and immediatly the doores were ſhut. \V And as they ſought
to kil him, it was told the Tribune of the band, That al Hieruſalem is
in a confuſion.

\V Who forthwith taking vnto him ſouldiars and Centurions, ranne downe
to them. Who, when they had ſeen the Tribune and the ſouldiars, ceaſed
to ſtrike Paul. \V Then the Tribune comming neer apprehended him, and
commanded him to be bound with two chaines: and he demanded who he was,
and what he had done. \V And ſome cried one thing, ſome another, in the
multitude. And whereas he could not know the certaintie for the tumult,
he commanded him to be led
\TNote{\L{In caſtra}. So in the places following.}
into the caſtel. \V And when he was come to the ſtaires, it chanced that
he was caried of the ſouldiars becauſe
%%% o-2456
of the violence of the people. \V For the multitude of the people
followed, crying: Away with him. \V And when Paul began to be brought
into the caſtel, he ſaith to the Tribune: Is it lawful for me to ſpeake
ſome thing to thee? Who ſaid: Canſt thou ſpeake Greeke? \V Art not thou
the Ægyptian that before theſe daies did raiſe a tumult, and didſt lead
forth into the deſert
%%% 2608
foure thouſand men that were murderers? \V And Paul ſaid to him:
\CNote{\XRef{Act.~22,~3.}}
I am a man truly a Iewe of Tarſus, a citizen not of an obſcure citie of
Cilicia. And I deſire thee, permit me to ſpeake to the people. \V And
when he had permitted him, Paul ſtanding on the ſtaires, beckned with
his hand to the people. And a great ſilence being made, he ſpake vnto
them in the Hebrew tongue, ſaying.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

