%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 24
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-24


\project douay-rheims


%%% 2612
%%% o-2461
\startChapter[
  title={Chapter 24}
  ]

\Summary{They proſecute him to Cæſarea, bringing with them an Oratour,
  who before the Preſident Felix accuſeth him. 10.~He anſwereth,
  defending himſelf from the crimes they charged him with, but
  confeſsing his religion plainly. 21.~The Iudge perceiuing his religion
  to be irreprehenſible, yealdeth not to condemne him at their pleaſure:
  24.~yea he oftentimes with his wife heareth his preaching, 27.~but yet
  doth not his dutie to deliuer him out of priſon.}

And after fiue daies the high Prieſt Ananias deſcended, with certaine
Ancients and one Tertullus an Oratour, who went to the Preſident againſt
Paul. \V And Paul being cited, Tertullus began to accuſe, ſaying:

Whereas we liue in much peace by thee, & many things are corrected by
thy prouidence, \V we doe alwaies & in al places receiue it, moſt
excellent Felix, with al thanks-giuing. \V But leſt I hinder thee any
longer, I deſire thee of thy clemencie breifely to heare vs. \V We haue
found this mã peſtiferous, & raiſing ſeditiõs to al the Iewes in the
whole world, and authour of the ſedition
\TNote{\G{τῆς αἱρέσεως}}
of the ſect of the Nazarenes. \V Who alſo hath attempted to violate the
%%% 2613
Temple,
%%% o-2462
whom alſo being apprehended we would haue iudged according to our
law. \V But Lyſias the Tribune comming in, with great force tooke him
away out of our hands, \V commanding his accuſers to come to thee, of
whom thou maieſt thyſelf iudging, vnderſtand of al theſe things, whereof
we accuſe him. \V And the Iewes alſo added, ſaying that theſe things
were ſo.

\V But Paul anſwered: (the Preſident making a ſigne vnto him for to
ſpeake)

Knowing that of many yeares thou art iudge ouer this Nation, I wil with
good courage anſwer for my ſelf. \V For thou maieſt vnderſtand that it
is not aboue twelue daies to me, ſince I went vp to adore in
Hieruſalem. \V And neither in the Temple did they find me diſputing with
any man, or cauſing concourſe of the multitude, neither in the
ſynagogues, nor in the citie: \V neither can they proue vnto thee the
things whereof they now accuſe me. \V But this I confeſſe to thee, that
\TNote{\G{κατὰ τὴν ὁδὸν }}
according to the
\SNote{Becauſe Tertullus the Iewes Oratour called Chriſtian religion
the ſect or (as it is there
\XRef{vers.~5.}
in the Greeke) the hereſie of the Nazarenes; S.~Paul anſwereth and
sheweth that it is no hereſie. And as for the word, \Emph{Sect}, in this
place; it is in the Greeke, \Emph{According to the way, which they cal
hereſie}, as alſo
\XRef{Act.~9,~2.}
\XRef{24,~22.}
And therfore the word \Emph{Sect} here is ſo taken. See
\XRef{Annot. c.~28,~22.}}
ſect, which they cal hereſie, I doe ſerue
\Var{the Father my God,}{the God of my Fathers, (or) the Father and my
God,}
beleeuing al things that are written in the Law & the Prophets: \V
hauing hope in God, the which theſe alſo thẽſelues expect, that there
ſhal be a reſurrection of iuſt and vniuſt. \V In this my ſelf alſo doe
ſtudie to haue a conſcience without offenſe toward God and toward men
alwaies. \V And after many yeares
\CNote{\XRef{Ro.~15,~18.}}
I came to beſtow almes vpon my Nation, & oblations, and vowes. \V In the
which they found me
\CNote{\XRef{Act.~21,~26.}}
purified in the Temple: not with multitude nor with tumult. \V But
certaine Iewes of Aſia, who ought to be preſent before thee and to
accuſe, if they had any thing againſt me: \V or let theſe men thẽſelues
ſay, if they haue found in me any iniquitie, foraſmuch as I ſtand in the
Councel, \V but of this one voice only that I cried ſtanding among them,
That
\CNote{\XRef{Act.~23,~6.}}
of the reſurrection of the dead am I iudged this day of you. \V And
Felix differred them, knowing moſt certainely of this way, ſaying: When
Lyſias the Tribune is come downe, I wil heare you. \V And he commanded
the Centurion to keep him, and that he ſhould haue reſt, neither to
prohibit any of his to miniſter vnto him.

\V And after ſome daies, Felix comming with Druſilla his wife, which was
a Iew, called Paul, and heard of him the faith that is in
Chriſt \Sc{Iesvs}. \V And he diſputing of
\SNote{The Apoſtolike teaching was not of only, or ſpecial faith; but of
iuſtice, & chaſtitie, & iudgement; that is to ſay, of the terrour of Hel
and other God's iudgements in the next life anſwerable to our deeds in
this world: by which the hearers were firſt terrified, and ſo induced to
penance. How ſay Heretikes then that ſuch things make men hypocrites.}
iuſtice and chaſtitie, and of the iudgement to come, Felix
%%% o-2463
being terrified, anſwered: For this time, goe thy way: but in time
conuenient I wil ſend for thee. \V Hoping alſo withal, that money would
be giuen him of Paul; for the which cauſe alſo oftentimes ſending for
him, he ſpake with him. \V But when two yeares were ended, Felix had a
Succeſſour Portius Feſtus. And Felix being willing to ſhew the Iewes a
pleaſure, left Paul in priſon.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

