%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 2571
%%% o-2414
\startChapter[
  title={Chapter 9}
  ]

\Summary{Saul not content to perſecute ſo cruelly in Hieruſalem, 3.~is
  in the way to Damaſcus told by our Lord \Sc{Iesvs} of his vaine
  attempt, and miraculouſly conuerted to be an Apoſtle: and after great
  penance, reſtored to his ſight by Ananias, and baptized. 20.~And
  preſently he dealeth mightily againſt the Iewes, prouing \Sc{Iesvs} to
  be Chriſt to their great admiration. 23.~But ſuch is their
  obſtinancie, that they lay al Damaſcus to kil him. 26.~From thence he
  goeth to Hieruſalem, & there ioyneth with the Apoſtles, and againe by
  the obſtinate Iewes his death is ſought. 31.~The Church being now
  growen ouer al Iewrie, Galilee, and Samaria, Peter viſiteth al: and in
  his viſitation, 33.~healing a lame man, 36.~and raiſing a dead woman
  conuerteth very many.}

%%% o-2415
And
\CNote{\XRef{Act.~22,~4.}
\XRef{Gal.~1,~13.}}
Saul as yet breathing forth threatnings and ſlaughter againſt the
Diſciples of our Lord, came to the high Prieſt, \V and asked letters of
him vnto Damaſcus to the Synagogues that if he had found any men and
women of this way, he might bring them bound vnto Hieruſalem. \V And as
he went on his iourney, it chanced that he drew nigh to Damaſcus: and
\CNote{\XRef{1.~Cor.~15,~8.}}
ſodenly a light from Heauen ſhined round about him. \V And falling on
the ground, he heard a voice ſaying to him
\SNote{The heretikes that conclude Chriſt ſo in heauen that he can be
no where els til the day of Iudgemẽt, ſhal hardly reſolue a mã that would
know where \Sc{Christ} was when he appeared here in the way, & ſpake
theſe words to Saul.}
Saul, Saul why perſecuteſt thou me? \V Who ſaid: Who art thou Lord? And
he: I am \Sc{Iesvs} whom thou doeſt perſecute. It is
%%% 2572
hard for thee to kicke againſt the pricke. \V And trembling and being
aſtoniſhed he ſaid: Lord, what wilt thou haue me to doe? \V And our Lord
to him: Ariſe and goe into the citie, & it ſhal be told thee what thou
muſt doe. But the men that went in companie with him, ſtood amaſed,
hearing the voice, but ſeeing no man. \V And Saul roſe vp from the
ground, and his eyes being opened, he ſaw nothing. And they drawing him
by the handes, brought him into Damaſcus. \V And he was there three
daies not ſeeing, and he did neither eate nor drinke.

\V And there was a certaine Diſciple at Damaſcus, named Ananias: and our
Lord ſaid to him in a viſion, Ananias. But he ſaid, Loe, here I am
Lord. \V And our Lord to him: Ariſe, and goe into the ſtreete that is
called Straight; and ſeeke in the houſe of Iudas, one named Saul of
Tarſus. For behold he praieth. (\V And he ſaw a man named Ananias,
comming in and impoſing handes vpon him for to receiue his ſight.) \V
But Ananias anſwered: Lord, I haue heard by many of this man, how much
euil he hath done to thy Saintes in Hieruſalem: \V and here he hath
authoritie from the cheefe Prieſts to bind al that inuocate thy name. \V
And our Lord ſaid to him: Goe, for a veſſel of election is this man vnto
me, to carie my name before the Gentils, and Kings, and the children of
Iſraël. \V For I wil ſhew him how great things he muſt ſuffer for my
name.

\V And Ananias went, and entred into the houſe: and impoſing hands vpon
him, he ſaid: Brother Saul, our Lord \Sc{Iesvs} hath ſent me, he that
appeared to thee in the way that thou cameſt; that thou maieſt ſee and
be filled with the Holy Ghoſt. \V And forthwith there fel from his eyes
as it
%%% o-2416
were ſcales, and he receiued ſight: and riſing he was
\SNote{Paul alſo himſelf, though with the diuine and heauenly voice
proſtrated and
inſtructed, yet was ſent to a man to receiue the Sacramẽts, & to be
ioyned to the Church.
\Cite{Auguſt. de doct. Chr. lib.~1. in proœm.}}
baptized. \V And when he had taken meate, he was ſtrengthened.

And he was with the Diſciples that were at Damaſcus, for certaine
daies. \V And incontinent
\Var{entring into the Synagogues, he}{in the Synagogues he}
preached \Sc{Iesvs}, that this is the Sonne of God. \V And al that
heard, were aſtoniſhed, and ſaid: Is not this he that expunged in
Hieruſalem thoſe that inuocated this name; and came hither to this
purpoſe that he might bring them bound to the cheefe Prieſts? \V But
Saul waxed mightie much more and confounded the Iewes that dwelt at
Damaſcus, affirming that this is \Sc{Christ}. \V And when many daies
were paſſed, the Iewes conſulted that they might kil him. \V But their
conſpiracie came to Sauls knowledge. And
\CNote{\XRef{2.~Cor.~11,~32.}}
they kept the gates alſo day and night, that they might kil him. \V But
the Diſciples taking him in the night, conueied him away by the wal,
letting him downe in a basket.

\V And when he was come into Hieruſalem, he aſſaied to ioyne himſelf to
the Diſciples, & al feared him, not beleeuing that he was a Diſciple. \V
But Barnabas tooke him & brought him to the Apoſtles, and told them how
in the way he had ſeen our Lord, and that he ſpake vnto him, & how in
Damaſcus he dealt confidently in the name of \Sc{Iesvs}. \V And he was
with them going in & going out in Hieruſalẽ, & dealing confidently in
the name of our Lord. \V He ſpake alſo to the Gentils, & diſputed with
the Greekes: but they ſought to kil him. \V Which when the Brethren had
knowen, they brought him downe to Caſarea, & ſent him away to Tarſus.

\V The
\SNote{The Church viſibly proceedeth ſtil with much cõfort and manifold
increaſe euen by perſecution.}
\Sc{Chvrch} truely through al Iewrie & Galilee and Samaria
%%% 2573
had peace, & was edified, walking in the feare of our Lord and was
repleniſhed with the conſolation of the Holy Ghoſt.

\V And it came to paſſe, that Peter as he paſſed through al, came to the
Saints that dwelt at Lydda. \V And he found there a certaine man named
Æneas, lying in his bed from eight yeares before, who had the palſey. \V
And Peter ſaid to him: Æneas, our Lord \Sc{Iesvs Christ}
\Var{heale}{healeth}
thee: ariſe, and make thy bed. And incontinent he aroſe. \V And al that
dwelt at Lydda and Saróna, ſaw him: who conuerted to our Lord.

%%% o-2417
\V And in Ioppe there was a certaine Diſciple named Tabitha, which by
interpretation is called Dorcas. This woman was ful of
\SNote{Behold good workes and almes-deeds, and the force therof reaching
euen to the next life.}
good workes & almes-deedes which ſhe did. \V And it came to paſſe in
thoſe daies, that ſhe was ſicke and died. Whom when they had waſhed they
laid her in an vpper chamber. \V And whereas Lydda was nigh to Ioppe,
the Diſciples hearing that Peter was in it, they ſent two men vnto him,
deſiring him: Be not loth to come ſo farre as to vs. \V And Peter riſing
vp came with them. And when he was come, they brought him into the vpper
chamber: and al the widowes ſtood about him weeping,
\SNote{The praiers of our Almes folke & beadſmen may doe vs great
good euen after our departure. For if they procured her temporal life,
much more may they help vs to Gods mercie and to releaſe of puniſhment
in Purgatorie.}
and ſhewing him the coates and garments which Dorcas made them. \V And
al being put forth, Peter falling on his knees praied, and turning to
the body he ſaid: Tabitha, ariſe. And ſhe opened her eyes: and ſeeing
Peter, ſhe ſate vp. \V And giuing her his hand, he lifted her vp. And
when he had called the Saints and the widowes, he preſented her
aliue. \V And it was made knowen throughout al Ioppe: and many beleeued
in our Lord. \V And it came to paſſe that he abode many daies in Ioppe,
with one Simon a tanner.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

