%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 2585
%%% o-2430
\startChapter[
  title={Chapter 14}
  ]

\Summary{Next in Iconium they preach, where many being conuerted of both
  ſortes, the obſtinate Iewes raiſe perſecution. 6.~Then in the townes
  of Lycaonia, where the Heathen firſt ſeing that Paul had healed one
  borne lame, are hardly perſuaded but they are Gods. 18.~But afterward,
  by the inſtigation of the malitious Iewes, they ſtone Paul, leauing him
  for dead. 20.~And ſo hauing done their circuit, they returne the ſame
  way confirming the Chriſtians, and making Prieſts for euery
  Church. 25.~And being come home to Antioche in Syria, they report al
  to the Church there.}

%%% o-2431
And it came to paſſe at Iconium that they entred together into the
Synagogue of the Iewes, and ſo ſpake, that a very great multitude of
Iewes and of the Greekes did beleeue. \V But the Iewes that were
incredulous, ſtirred vp and incenſed the hartes of the Gentils to anger
againſt the Brethren. \V A long time therfore they abode, dealing
confidently in our Lord, who gaue teſtimonie to the word of his grace,
granting ſignes and wonders to be done by their hands. \V And the
multitude of the citie was deuided: and certaine of them indeed were
with the Iewes, but certaine with the Apoſtles. \V And when the Gentils
and the Iewes with their Princes had made an aſſault, to vſe them
contumeliouſly, and to ſtone them, \V vnderſtanding it, they fled to the
cities of Lycaonia, Lyſtra and Derbe, and the whole countrie about, and
there they were euangelizing.

\V And a certaine man at Lyſtra impotent of his feet ſate there, lame
from his mothers womb, that neuer had walked. \V This ſame heard Paul
ſpeaking. Who looking vpon him, and ſeeing that he had faith for to be
ſaued, \V he ſaid with a loud voice: Stand vp right on thy feet. And he
leaped & walked. \V And the multitudes when they had ſeen what Paul had
done, lifted vp their voice in the Lycaonian tongue, ſaying: Gods made
like to men, are deſcended to vs. \V And they called Barnabas, Iupiter:
but Paul Mercurie, becauſe he was the cheefe
%%% 2586
ſpeaker. \V The Prieſt alſo of
\TNote{\G{διὸς τοῦ ὄντος}}
Iupiter that was before the citie, bringing oxen and garlands before the
gates, would with the people
\LNote{They would Sacrifice.}{This
\MNote{\L{Latria}.}
loe is the diuine worſhip, conſiſting in external Sacrifice, and in
acknowledging the parties worſhipped to be Gods: which
\CNote{\Cite{Aug. li.~10. de Ciu. Dei c.~1.}}
may be done to no man nor creature. And therfore the Apoſtles refuſe it
with al poſſible diligence, and al the Angels and Saints in Heauen
refuſe that adoration by Sacrifice. The Catholike Church ſuffereth no
Prieſt nor other ſo to worſhip any Saint in Heauen or earth. She hath
but one external Sacrifice, which is in the holy Maſſe, of Chriſts body
and bloud: that ſhe offereth to God alone; & \Emph{neither to Peter nor
to Paul} (ſaith
\CNote{\Cite{Aug. li.~2. de Ciu. Dei c.~27.}}
S.~Auguſtine) \Emph{though the Prieſt that ſacrificeth, ſtandeth ouer
their bodies, & offereth in their memories.} But
\MNote{\L{Dulia.}}
other kindes of honours and dueties, inferiour without al compariſon
(how great ſoeuer they be) to this, we doe, as the Scriptures & Nature
teach vs, to al Superiours in Heauen and earth, according to the degrees
of grace, honour, and bleſſednes that God hath called them vnto, from
our B.~Ladie Chriſts owne mother, to the leaſt ſeruant he hath in the
world. For which the Heretikes would neuer accuſe Chriſtian people of
Idolatrie, if they had either grace, learning, faith, or natural
affection.}
Sacrifice. \V Which thing when the Apoſtles Barnabas and Paul heard,
renting their coates, they leaped forth into the multitudes, crying \V
and ſaying: Ye men, why doe you theſe things? We alſo are mortal, men
like vnto you, preaching to you for to conuert from theſe vaine things,
to the liuing God that made the Heauen, and the earth, and the ſea, and
al things that are in them: \V who in the Generations paſt ſuffred al
the Gentils to goe their owne waies. \V Howbeit he left not himſelf
\SNote{The Heathẽ might by the daily benefites of God haue knowen him at
the leaſt to haue been their Creatour and only Lord, though the myſterie
of our Redemption were not opened to them.}
without teſtimonie, being beneficial from Heauen, giuing raines, and
fruitful ſeaſons, filling our hartes with food and gladnes. \V And
ſpeaking theſe things, they ſcarſe appeaſed
%%% o-2432
the multitudes from ſacrificing to them. \V But there came in
certaine Iewes from Antioche and Iconium: and perſwading the multitudes,
and
\CNote{\XRef{2.~Cor.~11,~25.}}
ſtoning Paul, they drew him out of the citie, thinking him to be
dead. \V But the Diſciples compaſſing him round about, he riſing vp,
entred into the citie, and the next day he went forth with Barnabas vnto
Derbe.

\V And when they had euangelized to that citie, & had taught many, they
returned to Lyſtra and Iconium, and to Antioche: \V confirming the
hartes of the Diſciples, and exhorting them to continue in the faith;
and that by many tribulations we muſt enter into the Kingdom of God. \V
And when
\SNote{
\Fix{Wee ſe}{We ſee}{likely typo, fixed in other}
by this, firſt that Ss.~Paul & Barnabas were Biſhops,
hauing here authoritie to giue holy Orders: ſecõdly that there was euen
thẽ a differẽce betwixt Biſhops and Prieſts, though the name in the
primitiue Church was often vſed indifferently: laſtly, that alwaies
faſting and praying were preparatiues to the giuing of holy Orders.}
they
\LNote{Had ordained.}{The
\MNote{Heretical tranſlation againſt holy orders.}
Heretikes, to make the world beleeue that al Prieſts ought to be choſen
by the voices of the people, and that they need no other Ordering or
Conſecration by Biſhops, preſsing the profane vſe of the
\TNote{\G{χειροτονήσαντες}}
Greeke word more then the very natural ſignification requireth &
Eccleſiaſtical vſe beareth, tranſlate thus, \Emph{Ordained by
election}. Whereas indeed this word in Scripture ſignifieth, Ordering by
impoſition of hands, as is plaine by other wordes equiualent,
\XRef{Act.~6,~13.}
\XRef{1.~Tim.~4,~5.}
\XRef{2.~Tim.~1.}
Where the Ordering of Deacons, Prieſts, & others is called
\TNote{\G{ἐπέθηκαν χεῖρας}}
Impoſitiõ of hãds, not of the people, but of the Apoſtles. And this to
be the Eccleſiaſtical vſe of the word, appeareth by S.~Hierom ſaying (as
is before alleaged) that 
\CNote{\Cite{Hiero. in 38.~Eſa.}}
\G{χειροτονία} \Emph{is the Ordering of Clerkes or Clergie men by
praier of voice and impoſition of hand.}}
had ordained to them
\LNote{Prieſts.}{Euen
\MNote{Heretical tranſlation againſt Prieſthood.}
ſo here alſo, as before, flying from the proper, apt, & knowen word,
which is moſt preciſely correſpondent to the very Greeke in our tongue
and al Nations, they tranſlate for \Emph{Prieſt}, Elder, that is, for a
calling of Office, a word of age: for a terme of art and by conſent of
al the Church and Apoſtolike authoritie and Fathers, appropriated to
holy Order, a vulgar, common and profane terme. With as litle grace as
if they ſhould tranſlate \L{Pontificem}, a bridgemaker, the \L{Maior},
of London, the Bigger of London. And thus you ſee within three wordes
compaſſe they flye guilefully from the Latin to the Greeke, & againe
guilefully from the Greeke to the vulgar Engliſh. Such corruption of
Scriptures their hatred of Prieſthood driueth thẽ vnto. If they had
tranſlated it ſo when the Scriptures were firſt written, (at which time
the word was but newly receiued into the ſpecial and Eccleſiaſtical
ſignification, & when it was yet taken ſometimes in common profane ſort,
as
\XRef{1.~Tim.~5.}
or there only where our ancient Latin verſion turneth \L{Preſbyter} into
\L{Senior}, becauſe the word was not yet wholy and only appropriated to
holy Orders, as afterward by vſe of many hundred yeares it was and is)
their dealing might haue had ſome colour of honeſtie and plaineſſe,
which now can not be but of plaine falſhood and corruption, and that of
further purpoſe then the ſimple can ſee. Which is to take away the
office of Sacrificing and other functions of Prieſts, proper in the new
Teſtament to ſuch as the Apoſtles often, and the poſteritie in manner
altogether cal Prieſts, \L{Presbyteros}. 
\MNote{If \L{Sacerdos} be Prieſt, much more \L{Preſbyter}.}
Which word doth ſo certainely
imply the authoritie of ſacrificing, that it is by vſe made alſo the only
Engliſh of \L{Sacerdos}, the Aduerſaries themſelues as wel as we, ſo
tranſlating it in al the old and new Teſtament: though they can not be
ignorant that \Emph{Prieſt} commeth of \L{Presbyter}, and not
of \L{Sacerdos}: and that antiquitie for no other cauſe applied the
ſignificatiõ of \L{Presbyter} to \L{Sacerdos}, but to ſhew
that \L{Preſbyter} is in the new Law, that which \L{Sacerdos} was in the
old: The Apoſtles abſtaining from this and other like old names at the
firſt; and rather vſing the wordes, Biſhops, Paſtours, and Prieſts,
becauſe they might be diſtinguiſhed from the Gouernours and ſacrificers
of Aarons order, who as yet in the Apoſtles time did their old functions
ſtil in the Temple. And this to be true, and that to be a Prieſt is to
be a man appointed to ſacrifice, the Heretikes themſelues
calling \L{Sacerdos} alwaies a Prieſt, muſt needes be driuen to
confeſſe. Although their folly is therein notorious, to apply willingly
the word \Emph{Prieſt} to \L{Sacerdos}, and to take it from
\MNote{\L{Preſbyter.}}
\L{Preſbyter.}
whereof it is properly deriued, not only in
\MNote{Prieſt.}
English, but in other languages both
\MNote{Prebſtre.}
French &
\MNote{Preti.}
Italian. Which is to take away the name that the Apoſtles and
Fathers gaue to the Prieſts of the Church, and to giue it wholy and only
to the order of Aaron, which neuer had it before our Prieſthood
began. Neuer did there Heretikes ſtand ſo much vpon doubtful deriuations
and deſcant of wordes as theſe Proteſtants doe, and yet neuer men
behaued themſelues more fondly in the ſame: as whoſoeuer marketh the
diſtinction of their Elders, Miniſters, Deacons, and ſuchlike, shal
perceiue.}
Prieſts in euery Church, and had prayed with faſtings, they commended
them to our Lord in whom they beleeued. \V And paſſing through Piſidia,
they came to Pamphylia, \V and ſpeaking the word of our Lord in Perge,
they went downe into Attalia: \V and from thence they ſailed to
Antioche,
\CNote{\XRef{Act.~13,~2.}}
whence they had been deliuered to the grace of God vnto the worke which
they accompliſhed. \V And when they were come, and had aſſembled the
Church, they reported what great things God had done with them, & that
he had opened a doore of faith to the Gentils. \V And they abode no
litle time with the Diſciples.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

