%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 27
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-27


\project douay-rheims


%%% 2617
%%% o-2467
\startChapter[
  title={Chapter 27}
  ]

\Summary{What a dangerous nauigation he had towards Rome: and that by
  his prediction and counſel the ship might haue been ſaued. And for his
  ſake (as God alſo reuealed to him before) al the companie was
  preſerued, being 276.~perſons.}

And after it was decreed that he ſhould ſaile into Italie, and that Paul
with other priſoners ſhould be deliuered to a Centurion named Iulius, of
the band Auguſta, \V we going vp into a ſhip of Adrumetum, beginning to
ſaile about the places of Aſia, looſed from the land, Ariſtarchus the
Macedonian of Theſſalonica continuing with vs. \V And the day following
we came to Sidon. And Iulius intreating Paul courteouſly, permited him
to goe to his freinds, and to take care of himſelf. \V And when we had
looſed thence we ſailed vnder Cypres: becauſe the winds were
contrarie. \V And ſailing the ſea of Cilicia and Pamphilia, we came to
\Var{Lyſtra,}{Myra}
which is in Lycia: \V and there the Centurion finding a ſhip of
Alexandria ſailing into Italie, remoued vs into it. \V And whereas many
daies we ſailed ſlowly, and were ſcarſe come ouer againſt Gnidus, the
wind hindering vs, we ſailed neere Crete by Salmone: \V and with much
adoe ſailing by it, we came into a certaine place that is called
Good-hauens, nigh to the which was a citie Thalaſſa.

\V And when much time was ſpent, and whereas now it was not ſafe
ſailing, becauſe the
\SNote{It may ſignifie the Iewes faſt of the ſeuenth moneth Septẽber,
after which the nauigation was perilous, winter approching.}
faſt now was paſt, Paul comforted them, \V ſaying to them: Ye men, I ſee
that the ſailing beginneth to be with hurt and much damage, not only of
the lading and the ſhip, but alſo of our liues. \V But the
%%% o-2468
Centurion beleeued the Gouernour and Maiſter of the ſhip, more then
thoſe things 
which were ſaid of Paul. \V And whereas it was not a commodious hauen to
winter in, very many taking counſel appointed to ſaile thence, if by any
meanes they might comming to Phœnice, winter there, a hauen of
\Fix{Drete}{Crete}{obvious typo, fixed in other}
looking toward the
%%% !!! TNote ?
\SNote{Names of windes.}
Afrike and the Chore. \V And the ſouth-wind blowing, they thinking that
they had obteined their purpoſe, when they had parted from Aſſon, ſailed
along by Crete. \V But not long after, a tempeſtuous wind that is called
Euro-aquilo, droue againſt it. \V And when the ſhip was caught and could
not make way againſt the wind, giuing vp the ſhip to the winds, we were
driuen. \V And running vpon a certaine Iland, that is called
%%% !!! \TNote{In Greek, Clauda.} ???
\SNote{Græc. Clauda.}
Cauda, we could ſcarſe get the cock-boat. \V Which being taken vp, they
vſed helps, girding the ſhip, and
%%% 2618
fearing leſt they ſhould fal into
\SNote{A place of quick-ſandes.}
the Syrte, letting downe the veſſel, ſo were they caried. \V And when we
were mightily toſſed with the tempeſt, the next day they caſt forth. \V
And the third day with their owne hands they threw forth the tacklings
of the ſhip. \V And neither ſunne, nor ſtarres appearing for many daies,
and no ſmal ſtorme being toward, al hope was now taken away of our
ſauing.

\V And when there had been long faſting, then Paul ſtanding in the
middes of them, ſaid: You ſhould indeed, O ye men, haue heard me, and
not haue parted from Crete, & haue gained this hurt and loſſe. \V And
now I exhort you to be of good cheere. For there ſhal be no loſſe of any
ſoule among you, but of the ſhip. \V For
\LNote{An Angel.}{S.~Paul
\MNote{Gods prouidẽce to the See of \Sc{Rome}.}
had many viſions, ſpecially to aſſure him that he should to Rome & ſtand
before Cæſar, our Lord himſelf before
\XRef{(23,~11.)}
appearing to him, & here an Angel, for that purpoſe. Whereby we plainely
ſee the ſpecial prouidence of God toward that See, where his two
principal Apoſtles were deſigned to preach, plant the faith, liue, die,
be buried, and honoured til the worldes end.}
an Angel of the God, whoſe I am, and
\TNote{\G{ᾧ λατρεύω}}
whom I ſerue, ſtood by me this night, \V ſaying: Feare not Paul, thou
muſt appeare before Cæſar: and behold God hath
\SNote{Paul (ſaith S.~Hierom) had ſo many ſoules in the ship giuen him,
that is, ſo many men ſaued for his ſake: & after he is with Chriſt, shal
he shut his mouth, & not be able once to ſpeake for thẽ that haue
beleeued in his Ghoſpel?
\Cite{Hier. ad. vigil.}
Wherby he proueth that if God doe much for the merits of Saints in
this life, much more at their interceſſion and praier in Heauen.}
giuen thee al that ſaile with thee. \V For the which cauſe be of good
cheere ye men: for I beleeue God, that it ſhal ſo be, as it hath been
ſaid to me. \V And we muſt come vnto a certaine Iland. \V But after the
fourteenth night was come on vs, as we were ſailing in Adria about
mid-night, the ſhip-men deemed that there appeared ſome countrie to
them. \V Who alſo ſounding, found twentie ſadomes: and being parted a
litle from thence, they found fifteene fadomes. \V And fearing leſt we
ſhould fal into rough places, caſting out of the ſterne foure ankers,
they wiſhed
%%% o-2469
that day were come. \V But as the ſhip-men ſought to fly out
of the ſhip, hauing let downe the cock-boat into the ſea, pretending as
if they were about to caſt out ankers out of the fore-part of the
ſhip, \V Paul ſaid to the Centurion and to the ſouldiars:
\LNote{Vnles theſe tarie.}{When
\MNote{Gods predeſtination and appointment taketh not away mans free-wil
& endeauours.}
God reuealeth to vs any thing, or aſſureth vs of any euent to come, he
diſchargeth vs not thereby of our requiſit endeauours and labours for
atcheiuing the ſame; not executing ordinarily his diſignements toward
men otherwiſe then by their owne free-wil and actions. S.~Paul ſaid not
here: Let vs doe what we liſt; worke we or ſit we ſtil, whether the
mariners goe out or tarie within, we are al ſure to be ſaued, for ſo God
hath reuealed to me, and he can not lie, neither can it fal otherwiſe;
but contrariewiſe ſaith he: If theſe mariners leaue the ship, you can
not be ſaued. So ſay al true Catholike Preachers to Chriſtian people:
What prouidence, predeſtination, or foreſight ſoeuer God haue of your
ſaluation you are not thereby conſtrained any way; you haue free-wil
ſtil, and cannot be ſaued (though you be predeſtinate) except you keep
Gods commandments, repent you of your ſinnes, beleeue, liue & die
wel. And if it were reuealed to any man, that he were one of Gods elect,
& that he should finally die in grace and be ſaued, yet he were bound to
worke his ſaluation with feare & trembling, as
\CNote{\XRef{1.~Cor.~9,~17.}
\XRef{Philip.~2,~12.}}
S.~Paul both did, and taught, leſt he become reprobate: no leſſe then
the ſame Apoſtle here and his fellowes, though they had their life
promiſed to them of God, yet were bound to labour and vſe al poſsible
diligence that they might not be drowned.}
Vnles theſe tarie in the ſhip you can not be ſaued. \V Then the
ſouldiars cut of the ropes of the cock-boat; and ſuffered it to fal
away.

\V And when it began to be light, Paul deſired al to take meat, ſaying:
This day is the fourteenth day that you expect and remaine faſting,
taking nothing. \V For the which cauſe I deſire you to take meat for
your health ſake: for there ſhal not an haire of the head perish of any
of you. \V And when he had ſaid theſe things, taking bread he gaue
thankes to God in the ſight of them al: and when he had broken it, he
began to eate. \V And being al made of better cheere, they alſo tooke
meat. \V And we were in al in the ſhip, ſoules two
\Fix{hnndred}{hundred}{obvious typo, fixed in other}
ſeuentie ſix. \V And being filled with meat, they lighted the ſhip,
caſting the wheat into the ſea. \V And when day was come, they knew not
the land: but they ſpied a certaine creeke that had a ſhore, into the
which they minded, if they could, to caſt aland the ſhip. \V And when
they had taken vp the ankers, they committed themſelues to the ſea,
looſing withal the rudder bands: and hoiſting vp the maine ſaile
according as the wind blew, they went on toward the ſhore. \V And when
we were fallen into a place between two ſeas, they graueled the ſhip:
and the fore-part truly ſticking faſt remained vnmoueable: but the
hinder-part was broken by the violence of the ſea. \V And the counſel of
the ſouldiars was, that they ſhould kil the priſoners: leſt any ſwimming
out, might runne away. \V But the Centurion willing
%%% 2619
to ſaue Paul, forbad it to be done: & he commanded them that could
ſwimme, to caſt out themſelues firſt, and eſcape, and goe forth to
land: \V and the reſt, ſome they caried on bordes, & ſome vpon thoſe
things that were of the ſhip. And ſo it came to paſſe, that al the
ſoules eſcaped to land.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

