%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-12


\project douay-rheims


%%% 2580
%%% o-2424
\startChapter[
  title={Chapter 12}
  ]

\Summary{Herod the firſt King that perſecuted the Church, hauing at
  Hieruſalem (when Barnabas and Saul were there with the collation of
  the Antiochians) killed Iames the Apoſtle, 3.~& to pleaſe the Iewes
  impriſoned Peter with the mind to kil him alſo, but fruſtrate by an
  Angel ſent of God at the continual prayers of the Church made for her
  cheefe Paſtour, 19.~being puffed vp with ſuch pride that at Cæſaria he
  refuſeth not to be honoured as God; 23.~is miraculouſly ſtriken of
  Gods Angel. 24.~And ſo after the perſecutours death, the Churches
  preaching proſpereth exceedingly.}

And at the ſame time Herod the King ſet his hands, to afflict certaine
of the Church. \V And he killed Iames the brother of Iohn with the
ſword. \V And ſeing that it pleaſed the Iewes, he added to apprehend
Peter alſo. And it was the daies of the Azymes. \V Whom when he had
apprehended, he caſt into priſon, deliuering him to
\SNote{As Peters perſon was more notorious then others, & therfore
better garded then other, for feare he should eſcape; ſo Gods prouidence
in preſeruing & deliuering him for the longer gouernement of his Church,
is very maruelous.}
foure quaternions of ſouldiars to be kept, meaning after the Paſche to
bring him forth to the people. \V And Peter indeed was kept in
priſon. But
\LNote{Praier was made.}{The Church praied inceſſantly for her cheefe
Paſtour, and was heard of God: and al Chriſtian people are warned
thereby to pray for their Biſhopes and Paſtours in priſon.}
praier was made of the Church without
%%% o-2425
intermiſſion vnto God for him. \V
And when Herod would haue brought him forth, the ſame night Peter was
ſleeping betweene two ſouldiars, bound with
\LNote{Two chaines.}{Theſe
\MNote{S.~Peters chaines.}
chaines are famous for miracles, and were brought from Hieruſalem to
Rome by Eudoxia the Empereſſe, wife to Theodoſius the yonger, where they
were matched & placed with another chaine that the ſame Apoſtle was tied
with by Nero, & a Church founded therevpon, name \L{Petri ad vincula};
where they are religiouſly kept and reuerenced vntil this day: & there
is a Feaſt in the whole Church for the ſame, the firſt of Auguſt, which
we cal \Emph{Lammas day}.}
two chaines: and the keepers before the doore kept the priſon. \V And
behold an Angel of our Lord ſtood in preſence: and light ſhined in the
houſe: and ſtriking Peters ſide, he raiſed him, ſaying: Ariſe
quickely. And the chaines fel from his hands. \V And the Angel ſaid to
him: Gird thee, and put on thy ſhoes. And he did ſo. And he ſaid to him:
Put thy garment about thee, & follow me. \V And going forth he followed
him, & he knew not that it was true which was done by the Angel: but he
thought that he ſaw a viſion. \V And paſſing through the firſt & the ſecond
watch, they came to the iron gate that leadeth to the citie, which of it
ſelf opened to them. And going out, they went forward one ſtreet: and
incontinent the Angel departed from him. \V And Peter returning to
himſelf, ſaid: Now I know in very deed that our Lord hath ſent his
Angel, and deliuered me out of Herods hand, & from al the expectation of
the people of the Iewes.

\V And conſidering, he came to the
\SNote{It is much for the praiſe of theſe good Chriſtiãs that the
aſſemblie to Gods ſeruice & praier was kept in their houſes in the time
of perſecution, & that the Apoſtle came thither ſtraight out of priſon,
as his firſt refuge; as now Chriſtian people doe much to their
commendatiõ, in places where Hereſie doth reigne.}
houſe of Marie the mother of Iohn, who was ſurnamed Marke, where many
were gathered and praying. \V And when he knocked at the doore of the
gate, there came forth a wench to ſee, named Rhode. \V And as ſhe knew
Peters voice, for ioy ſhe opened not the gate, but running in ſhe told
that Peter ſtood before the gate. \V But they ſaid to her: Thou art
mad. But ſhe affirmed that it was ſo. But they ſaid: It is
\LNote{His Angel.}{\Emph{If
\MNote{Patronage of Angels.}
proper Angels} (ſaith S.~Chryſoſtom) \Emph{be deputed by our Lord to
ſuch as haue only charge of their owne life, (as one of the iuſt ſaid,
\CNote{\XRef{Gen.~48,~16.}}
The Angel which hath deliuered me from my youth vpward) much more are
ſupernal Spirits at hand to help them vnto whom the charge and burden of
the world is committed.}
\Cite{Chrys. in laud. Pauli. ho.~7. to.~3.}}
his Angel. \V And Peter continued knocking. And when they had opened,
they ſaw him, & were aſtoniſhed. \V And beckning with his hand to them,
that they ſhould hold their peace, he told how our Lord had brought him
out of priſon, & he ſaid:
\LNote{Tel Iames.}{He
\MNote{Publike praier for S.~Peter the Head.}
willeth them to ſhew this to S.~Iames Biſhop of Hieruſalem and to the
Chriſtians, that they might ſee the effect of their praiers for him, &
giue God thankes. For S.~Iames no doubt publiſhed common praier for
S.~Peter.}
Tel theſe things to Iames & to the Brethren. And going forth
%%% 2581
he went
\SNote{Though God had ſo miraculouſly deliuered him, yet he would not
tẽpt God by tarying among his perſecutours, but according to Chriſtes
commandement fled for a time.}
into another place. \V And when day was come, there was no litle adoe
betweene the ſouldiars, what was become of Peter. \V And Herod, when he
had ſought him, and had not found, making inquiſition of the keepers,
commanded them to be led away: & going downe from Iewrie into Cæſarea,
there he abode. \V And he was angrie with the Tyrians and the
Sidonians. But they with one accord came to him, and perſuading Blaſtus
that was cheefe of the Kings chamber, they deſired peace, for that
their countries were nouriſhed by him.
%%% o-2426
\V And vpon a day appointed, Herod being araied with kingly attire, ſate
in the iudgement ſeate, and made an oration to them. \V And the people
made acclamation, The voices of a God, & not of a man. \V And forthwith
an Angel of our Lord
\SNote{Princes that take delight in the flattery & praiſes of the
people, ſo much that they forget themſelues to be men, & to giue the
honour to God, may be warned by this exãple.}
ſtrook him, becauſe he had not giuen the honour to God: and being
conſumed of wormes, he gaue vp the ghoſt. \V But the word of our Lord
increaſed and multiplied. \V And Barnabas and Saul returned from
Hieruſalem, hauing accompliſhed their
\CNote{\XRef{Act.~11,~19.}}
miniſterie, taking with them Iohn that was ſurnamed Marke.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

