%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 25
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-25


\project douay-rheims


%%% 2614
%%% o-2463
\startChapter[
  title={Chapter 25}
  ]

\Summary{After two yeares empriſonment the Iewes continue their ſute
  againſt him, ſoliciting the new Preſident Feſtus, 6.~firſt at
  Hieruſalem, then at Cæſarea: 9.~where through the Iewes partialitie he
  is faine to appeale vnto the Emperour: 13.~and is in the meane time
  brought forth by Feſtus (giuing him good teſtimonie, notwithſtanding
  the exclamations of the Iewes againſt him) vnto King Agrippa and his
  Queene Bernice.}

Feſtvs therfore when he was come into the prouince, after three daies
went vp to Hieruſalem from Cæſarea. \V And the cheefe Prieſts, and
principal men of the Iewes went vnto him againſt Paul: and they deſired
him, \V requeſting fauour againſt him, that he would command him to be
brought to Hieruſalem, laying wait for to kil him in the way. \V But
Feſtus anſwered, that Paul is kept in Cæſarea: and that he would very
ſhortly goe thither. \V They therfore, ſaith he, that are of abilitie
among you, going downe with me, if there be any crime in the man, let
them accuſe him.

\V And hauing taried among them not aboue eight or ten daies, he went
downe to Cæſarea, and the next day he ſate in the iudgement ſeat: and he
commanded Paul to be brought. \V Who being brought, there ſtood about
him the Iewes that were come downe from Hieruſalem, obiecting many and
greuious
\TNote{Crimes, as v.~27.}
cauſes which they could not proue; \V Paul making anſwer, That neither
againſt the law of the Iewes, nor againſt the Temple, nor againſt Cæſar
haue I any thing offended. \V But Feſtus willing to ſhew the Iewes a
pleaſure, anſwering Paul, ſaid: Wilt thou goe vp to Hieruſalem, and
there be iudged of theſe things before me? \V And Paul ſaid: At Cæſar's
iudgement ſeat doe I ſtand, where I ought to be iudged: the Iewes I haue
not hurt, as thou very
%%% o-2464
wel knoweſt. \V For if I haue hurt them, or done any thing worthie of
death, I refuſe not to die. But if none of thoſe things be, whereof
theſe accuſe me, no man can giue me to them.
\SNote{If S.~Paul both to ſaue himſelf from whipping and from death
ſought by the Iewes, doubted not to cire for ſuccour of the Romane
lawes, and to appeale to Cæſar the Prince of the Roman's not yet
Chriſtned: how much more may we cal for aide of Chriſtiã Princes and their
lawes, for the punishment of Heretikes, and for the Churches defenſe
againſt them?
\Cite{S.~Auguſtine ep.~50.}}
I appeale to Cæſar. \V Then Feſtus hauing conferred with the Councel,
anſwered: Haſt thou appealed to Cæſar? to Cæſar ſhalt thou goe.

\V And when certaine daies were paſſed, King Agrippa & Bernice came
downe to Cæſarea to ſalute Feſtus. \V And as they taried there a good
many daies, Feſtus ſignified to the King, of Paul ſaying: A certaine
perſon was left priſoner by Felix, \V concerning whom, when I was at
Hieruſalem, the cheefe Prieſts and the Ancients of the Iewes came vnto
me, deſiring condemnation againſt him. \V To whom I anſwered: That it is
not the Romanes cuſtome to yeald vp any man before that he which is
accuſed haue his accuſers preſent, and take place to make his anſwer for
to cleere himſelf of the crimes. \V When they therfore were aſſembled
hither, without any delaie, the day following, ſitting in the iudgement
ſeat, I commanded the man to be brought. \V Of
%%% 2615
whom, when the accuſers ſtood vp, they brought no cauſe which I thought
il of: \V but certaine queſtions of their owne ſuperſtition they had
againſt him, and of
\SNote{This whom he termeth by contempt, one \Sc{Iesvs}, hath now made al
the Romane Emperours & Princes of the world to know him, and hath giuen
the ſeat of the Cæſars to his poore ſeruants, Peter & his Succeſſours.}
one \Sc{Iesvs} deceaſed, whom Paul affirmed to liue. \V Doubting
therfore of this kind of queſtion, I ſaid, whether he would goe to
Hieruſalem, & there be iudged of theſe things. \V But Paul appealing to
be kept vnto the knowledge of Auguſtus, I commanded him to be kept, til
I ſend him to Cæſar. \V And Agrippa ſaid to Feſtus: My ſelf alſo would
heare the man. To morow, ſaid he, thou ſhalt heare him.

\V And the next day when Agrippa and Bernice were come with great pompe,
and had entred into the hal of audience with the Tribunes and principal
men of the citie, at Feſtus commandement Paul was brought. \V And Feſtus
ſaith: King Agrippa, and al ye men that are preſent together with vs,
you ſee this man, concerning whom al the multitude of the Iewes called
vpon me at Hieruſalem, requeſting and crying out that he ought not to
liue any longer. \V Yet haue I found nothing that he hath committed
worthie of death. But foraſmuch as he himſelf appealed to Auguſtus, I
haue determined to ſend him. \V Of whom what to write for certaintie to
my Lord, I haue not. For the which cauſe I
%%% o-2465
haue brought him forth to you, and eſpecially to thee, King Agrippa, that
examination being made, I may haue what to write. \V For it ſeemeth to
me without reaſon, to ſend a priſoner, & not to ſignifie
\TNote{\G{τὰς κατ' αὐτοῦ αἰτίας}}
his cauſes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

