%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 2600
%%% o-2447
\startChapter[
  title={Chapter 19}
  ]

\Summary{How Paul began the Church of Epheſus, firſt in twelue that were
  baptized with Iohns Baptiſme, 8.~then preaching three moneths in the
  Synagogue of the Iewes, vntil for their obſtinacie & blaſpheming, he
  forſooke them, diſputing afterward in a certaine ſchoole for two
  yeares ſpace to the maruelous increaſe of the Church, ſpecially
  through his great miracles alſo, in healing diſeaſes with the touch
  of his clothes, and expelling Diuels, 13.~who yet contemned the
  Exorciſts of the Iewes. 18.~How the Chriſtians there confeſſe their
  actes, and burne their vnlawful bookes: 21.~and how he foretold that
  after he had been at Hieruſalem, he muſt ſee Rome. 23.~And what a
  great ſedition was raiſed againſt him at Epheſus, by them that got
  their liuing by working to the idolatrous Temple of Diana.}

And it came to paſſe when Apollo was at Corinth, that Paul hauing gone
through the higher parts came to Epheſus, and found certaine
Diſciples: \V and he ſaid to them: Haue you receiued the Holy Ghoſt,
beleeuing? But they ſaid to him:
%%% 2601
Nay, neither haue we heard whether there be a Holy Ghoſt. \V But he
ſaid: In what then were you baptized? Who ſaid:
\SNote{Iohns Baptiſme not ſufficient.}
In Iohns Baptiſme. \V And Paul ſaid:
\CNote{\XRef{Mt.~3,~11.}
\XRef{Mr.~1,~8.}
\XRef{Lu.~3,~16.}}
Iohn baptized the people with the Baptiſme of penance, ſaying, That they
ſhould beleeue in him that was to come after him, that is to ſay,
\SNote{Chriſts Baptiſme neceſſarie.}
in \Sc{Iesvs}. \V Hearing theſe things, they were baptized in the name
of our Lord \Sc{Iesvs}.
%%% o-2448
\V And when Paul had
\SNote{S.~Paul miniſtred the Sacrament of Cõfirmation. See
\XRef{annot. c.~8,~17.}}
impoſed hands on them, the Holy Ghoſt came vpon them, & they ſpake with
tongues, and prophecied. \V And al the men were about twelue.

\V And entring into the Synagogue, he ſpake confidently for three
moneths, diſputing & exhorting of the Kingdom of God. \V But when
certaine were indurate, and beleeued not, il-ſpeaking the way of our
Lord before the multitude, departing from them, he ſeparated the
Diſciples, daily diſputing in the ſchoole of one Tyrannus. \V And this
was done for the ſpace of two yeares, ſo that al which dwelt in Aſia,
heard the word of our Lord, Iewes and Gentils.

\V And God wrought by the hand of Paul miracles not common: \V ſo that
there were alſo brought from his body
\LNote{Napkins.}{The
\MNote{Touching of Relikes, & miracles done by the ſame.}
napkins that had touched S.~Paules body, wrought miracles, and it was no
ſuperſtition to attribute that vertue to them which God gaue to them
indeed: nor to ſeeke to touch them for health, was any diſhonour to God,
but it much proued Chriſts religion to be true, and him to be the only
God, whoſe ſeruants, yea whoſe ſeruants
\CNote{\XRef{Act.~5,~26.}}
shadows and napkins could doe ſuch wonders, as S.~Chryſoſtom
\Cite{(to.~5. cont. Gentiles, quod Chriſtus ſit Deus, in vita Babylæ.)}
ſheweth in a whole booke to that purpoſe, againſt the Pagans: prouing
hereby and by the like vertue of other Saints and their Relikes, that
Chriſt their Lord and Maiſter is God. For it is al one concerning the
bodies of Saints, relikes, garmẽts, ſtaues, bookes, or any thing that
belonged to thẽ: al which may & haue done & yet doe (when it is
neceſſarie to our edification) the like wonders to Gods great honour;
not only in their life time, but after their death much more. For
S.~Paules napkins had as great force when he was dead, as when he liued,
and ſo much more, as his grace and dignitie with God is greater
\Fix{them}{then}{obvious typo, fixed in other}
before. Which S.~Chryſoſtom in the place alleaged proueth at large by
the ſhrine of S.~Babylas the Martyr: and to thinke the contrarie, is the
Hereſie of Vigilantius, condemned ſo long ſince as S.~Hieroms time, and
by him refuted abundantly.}
napkins or handkerchefs vpon the ſicke, and the diſeaſes departed from
them, and the wicked ſpirits went out. \V And certaine alſo of the
Iudaical Exorciſts that went about, aſſaied to inuocate vpon them that
had euil Spirits, the name of our Lord \Sc{Iesvs}, ſaying: I adiure you
by \Sc{Iesvs} whom Paul preacheth. \V And there were certaine ſonnes of
Sceua a Iewe, cheefe Prieſt, ſeuen, that did this. \V But the wicked
Spirit anſwering, ſaid to them: \Sc{Iesvs} I know, and
\LNote{Paul I know.}{Both
\MNote{The name, or preſence, or Reliques of Saints & holy men, confound
the Diuel.}
the ſaid napkins taken from S.~Paules body, and his name alſo, were
dreadful and able to expel Diuels. Whereby we learne that not only
Chriſtes name, which is the principal, but his ſeruants names alſo
inuocated vpon the poſſeſſed, haue power ouer Diuels: which is a
maruelous honour to Saints, and nothing diminiſheth the glorie of God,
but exceedingly increaſeth the ſame, not only himſelf, but his ſeruãts
alſo being able to doe ſuch things, and to be ſtronger then any Diuel in
Hel. So we read in
\CNote{\Cite{In vit. Hilarionis.}}
S.~Hierom, that many did inuocate the name of S.~Hilarion vpon the
poſſeſſed, and the Diuels ſtraight departed. So did the Diuel know
\CNote{\Cite{Chryſ. loco citato.}}
S.~Babylas & other Saints, euen after they were dead, when they could
not ſpeake for the preſence of their Relikes, and when they were
tormented & expelled by them: whereof al antiquitie is ful of
teſtimonies. But our Heretikes, Luther and Caluin and their Schollers
attempting to caſt out Diuels, ſped much like as theſe good fellowes
did.}
Paul I know: but you, what are ye? \V And the man in whom the wicked
Spirit was leaping vpon them, and maſtring
%%% !!! Var only in other
\Var{both,}{them}
preuailed againſt them, ſo that they fled out of that houſe naked and
wounded. \V And this was made notorious to al the Iewes and the Gentils
that dwelt at Epheſus: and feare fel vpon al them, and the name of our
Lord \Sc{Iesvs} was magnified. \V And many of them that beleeued, came
confeſſing & declaring
\SNote{They made not only a general confeſſiõ wherin al men ſhew
thẽſelues alike to be ſinners, as our Proteſtãts doe, but euery one
confeſſed his owne proper deeds & faults.}
their deeds. \V And many of them that had followed
\LNote{Curious things.}{Curious
\MNote{\Fix{Superſtious,}{Superſtitious,}{obvious typo, fixed in other}
heretical, & al hurtful bookes muſt be made away.}
and vnlawful ſciences, as Witchcraft, Necromancie, and other meanes of
diuination by ſouthſaying, figure-caſting, interpretation of dreames,
or any way not allowed by God and his Church, muſt much more be
abhorred of old Chriſtians, when theſe ſo lately conuerted were ſo
zelous and diligent to leaue them. And by this example al that are newly
reconciled to the Church, are taught, the firſt thing they doe, to burne
their heretical and naughtie bookes.}
curious things, brought together their
\LNote{Bookes.}{A
\MNote{Decrees & penal lawes againſt heretical bookes.}
Chriſtian man is bound to burne or deface al wicked bookes of what ſort
ſo euer, ſpecially Heretical bookes: Which though they infect not him
alwaies that keepeth them, yet being forth-comming, they may be noiſom &
pernicious to others that ſhal haue them & read them after his death, or
otherwiſe. Therfore hath the Church taken order for condemning al ſuch
bookes, & againſt the reading of them, where danger may enſue: & the
Chriſtian Emperours, Conſtantinus Magnus, Valentinian, Theodoſius,
Martian, Iuſtinian, made penal lawes for the burning or defacing of
them. 
%%% !!! How is this supposed to be divided?
\Cite{Sozom. li.~1. c.~20. li.~2. c.~31. conc. Chalc. act.~3. in fine,
cap. Ampla. & in fine totius
Conc. c. Imperatur. Con. Conſtantinop. 2. confeſ. 5. cap. Debitam. &
Act.~2. cap.~1. & cap. Rom.}
See 
\Cite{Euſebius li.~3. de vita Conſtant. c.~61,~62,~63,~64.}
The danger of reading them, as it is manifeſt, ſo it is ſignified by
\Cite{Euſeb. li.~7. c.~6.}
\Cite{Auguſtin li.~3. de bapt. c.~14.}
\Cite{S.~Greg. li.~5. ep.~64.}}
bookes, and burnt them before al: and counting the prices of them, they
found the money to be fiftie thouſand pence. \V So mightily increaſed
the word of God and was confirmed.

\V 
%%% !!! Not quite as marked, but I think this is best.
\MNote{The 6.~part.

Of taking away the Ghoſpel from Hieruſalem the head citie of the Iewes,
and giuing it to Rome the head citie of the Gentils.}
And when theſe things were ended, Paul purpoſed in the Spirit, when
he had paſſed through Macedonia and Achaia, to goe to Hieruſalem,
ſaying: After I ſhal haue been there I muſt ſee Rome alſo. \V And
ſending into Macedonia two of them that miniſtred vnto him, Timothee and
Eraſtus, himſelf remained for a time in Aſia.

\V And at that time there was made no litle trouble about
%%% o-2449
the way of our
Lord. \V For one named Demetrius, a ſiluer-ſmith, that made ſiluer
\SNote{The Proteſtants trãſlate, \Emph{shrines}, in the
\Cite{Bible an.~1577.}
to make the people thinke that it toucheth the holy shrines of Saints;
moſt corruptly, the greeke ſignifying plainly, \Emph{tẽples}, & that of
heathẽ Gods.}
\TNote{\G{ναοὺς}}
Temples of Diana, procured to the artificers no ſmal gaine: \V whom
calling together and them that were the ſame kind of worke-men, he ſaid:
Sirs, you know that our gaine is of this occupation; and you ſee, and
heare that this ſame Paul by perſuaſion hath auerted a great multitude
not only of Epheſus, but almoſt of al Aſia, ſaying: That they are not
Gods which be made by hands. \V And not only vnto
%%% 2602
vs is this part in danger to be reproued, but alſo the Temple of great
Diana ſhal be reputed for nothing, yea & her maieſtie ſhal begin to be
deſtroied, whom al Aſia & the world worſhippeth. \V Hearing theſe things
they were repleniſhed with anger, and cried out ſaying: Great is Diana
of the Epheſians. \V And the whole citie was filled with confuſion, and
they ranne violently with one accord into the theater, catching Gaius &
Ariſtarchus Macedonians, Paules companions. \V And when Paul would haue
entred into the people, the Diſciples did not permit him. \V And
certaine alſo of the Princes of Aſia that were his freinds, ſent vnto
him, deſyring that he would not aduenture himſelf into the theater: \V
and others cried another thing. For the aſſemblie was confuſe, & the
more part knew not for what cauſe they were aſſembled. \V And of the
multitude they drew forth Alexander, the Iewes thruſting him forward. But
Alexander with his hand deſiring ſilence, would haue giuen the people
ſatisfaction. \V Whom as ſoone as they perceiued to be a Iewe, there was
made one voice of al, almoſt for the ſpace of two houres crying out:
Great is Diana of the Epheſians. \V And when the Scribe had appeaſed the
multitudes, he ſaith: Ye men of Epheſus, for what man is there that
knoweth not the citie of the Epheſians to be a worſhipper of great Diana,
and
\SNote{Here the Heretikes adde to the text this word, \Emph{image}, more
then is in the Greeke, to put a ſcruple into the peoples mind concerning
holy images.}
\TNote{\G{τοῦ διοπετοῦς}}
Iupiters child? \V For aſmuch therfore as theſe things can not be
gainſaid, you muſt be quieted, and doe nothing raſhly. \V For you haue
brought theſe men, being neither ſacrilegious, not blaſpheming your
Goddeſſe. \V But if Demetrius & the artificers that are with him, haue
matter to ſay againſt any man, there are Courts kept in the common
place, & there are Proconſuls; let them accuſe one another. \V And if
you aske any other matter, it may be reſolued in a lawful aſſemblie. \V
For we are in danger alſo to be accuſed for this daies ſedition:
%%% o-2450
whereas there is no man guilty by whom we may giue an account of this
concourſe. And when he had ſaid theſe things he dimiſſed the aſſemblie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

