%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Acts
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 2552
%%% o-2393
\startChapter[
  title={Chapter 2}
  ]

\Summary{The
\MNote{The 2.~part. The comming of the Holy Ghoſt and beginning of the
  Church in Hieruſalem.}
Holy Ghoſt comming to the Faithful vpon whitſunday, 5.~Iewes in
Hieruſalem of al Nations doe wonder to heare them ſpeake al
tongues. 14.~And Peter to the deriders declareth, that it is not
drunkennes, but the Holy Ghoſt, which Ioel did prophecie of,
which \Sc{Iesvs} (whom they crucified) being now riſen againe and
aſcended (as he sheweth alſo out of the Scriptures) hath powred out from
Heauen: concluding therfore that he is \Sc{Christ}, and they moſt
horrible murderers. 37.~Whereat they being compunct, and ſubmitting
themſelues, he telleth them that they muſt be baptized, and then they
alſo shal receiue the ſame Holy Ghoſt, as being promiſed to al the
baptized. 41.~And ſo 3000. are baptized that very day. 42.~Whoſe godly
exerciſes are here reported, and alſo their liuing in ſtate of
perfection. The Apoſtles worke many miracles, and God daily increaſeth
the number of the Church.}

And 
\CNote{\XRef{Act.~1,~4.}}
when the daies of
\LNote{The daies of Pentecoſt.}{As
\MNote{The feaſtes of Paſche and Pentecoſt.}
Chriſt our Paſche, for correſpondence to the figure, was offered at the
Iewes great feaſt of Paſche, ſo fifty daies after (in Greeke, Pentecoſt)
for accomplishing the like figure of the Law-giuing in Mount Sinai, he
ſent downe the Holy Ghoſt iuſt on the day of their Pẽtecoſt: which was
alwaies on Sunday, as appeareth
\XRef{Leuit.~23,~15.}
Both which daies the Church keepeth yearely for memorie of Chriſts death
and Reſurrection, and the ſending downe of the Holy Ghoſt; as they did
the like for record of their deliuerie out of Ægypt, & their Law-giuing
aforeſaid: the ſaid Feaſtes with vs conteining, beſides the remembrance
of benefites paſt, great Sacraments alſo of the life to come.
\Cite{Aug. ep.~119. c.~16.}}
Pentecoſt were accompliſhed, they were al together in one place: \V and
ſodenly there was made a ſound from Heauen, as of a vehement wind
comming, & it filled the whole houſe where they were ſitting. \V And
there appeared to them parted tongues as it were
\CNote{\XRef{Mt.~3,~11.}}
of fire, and it ſate vpon euery one of them: \V and they were
\LNote{Al replenished.}{Though
\MNote{The ſending of the Holy Ghoſt on
\Fix{whithſũſay}{whitſunday}{obvious typo, fixed in other}
and the effects thereof.}
the Apoſtles and the reſt were baptized before, and had thereby receiued
the grace of the Holy Ghoſt to ſanctification and remiſſion of ſinnes,
as for diuers other purpoſes alſo: yet as Chriſt
\CNote{\XRef{Luc.~34,~49.}}
promiſed them they should be further indowed with ſtrength and vertue
from aboue, ſo here he fulfilled his promiſe, viſibly powring downe the
Holy Ghoſt vpon al the companie and vpon euery one of them, thereby
replenishing the Apoſtles ſpecially with al truth, wiſedom, and
knowledge neceſſarie for the gouernement of the Church, & giuing both to
them and to al other preſent, the grace and effect of the Sacrament of
Confirmation, accompliſhing, corroborating, and ſtrengthning them in
their faith and the confeſſion of the ſame. And laſtly for a viſible
token of Gods Spirit, he indowed them al with the guift of diuers
ſtrange tongues: al (I ſay) there preſent, as wel
\MNote{Our B.~Ladie.}
our Ladie, as other holy women and Brethren, beſides the
Apoſtles. Though
\CNote{\Cite{Beza Annot. in hoc c. v.~1.}}
the Heretikes fondly argue, for the deſire they haue to diſhonour
Chriſts mother, that neither she nor they were there preſent, nor had
the guift of tongues, contrarie to the plaine text that
ſaith, \Emph{They were al together}, to wit, al the 120.~mentioned
before.
\XRef{c.~1,~15.}}
al repleniſhed with the
\CNote{\XRef{Io.~7,~39.}}
Holy Ghoſt, and they began to ſpeake with diuerſe tongues according as
the Holy Ghoſt gaue them to ſpeake.

\V And there were dwelling at Hieruſalem Iewes, deuout men of euery
Nation that is vnder Heauen. \V And when this voice was made, the
multitude came together, and was aſtoniſhed in mind, becauſe euery man
heard them ſpeake in his owne tongue. \V And they were al amaſed, and
marueled ſaying: Are not, loe, al theſe that ſpeake, Galilæans, \V and
how
%%% Variation is only in other
\Var{haue we heard,}{do we heare}
each man our owne tõgue wherein we were borne? \V
Parthiãs, & Mediãs, & Elamites, & that inhabite Meſopotamia, Iewrie, &
Capadocia, Pontus, and Aſia, \V Phrygia, and Pamphilia, Ægypt & the
partes of Lybia that is about Cyrenee, & ſtrangers of Rome, \V Iewes
alſo, and Proſelytes, Cretenſians, and Arabians: we
%%% Variation is only in other
\Var{haue heard,}{do heare}
them ſpeake in our owne tongues the great workes of God. \V And they
were al aſtoniſhed, and marueled, ſaying one to another: What meaneth
this? \V But others deriding ſaid: That theſe are ful of new wine.

\V But
\SNote{Peter the Head of the reſt and now newly repleniſhed with al
knowledge & fortitude, maketh the firſt Sermon.}
Peter ſtanding with the Eleuen, lifted vp his voice, and ſpake to them:
Ye men, Iewes, and al you that dwel in Hieruſalem, be this knowen to you,
and with your eares receiue my wordes. \V For theſe are not drunke, as
you
%%% o-2394
ſuppoſe, whereas it is the third houre of the day: \V But this is it
that was ſaid by the Prophet Ioel:
\CNote{\XRef{Iol.~2,~28.}}
\Emph{And it shal be, in the laſt daies (ſaith our Lord) of my Spirit I
wil powre out vpon al flesh: and your ſonnes and your daughters shal
prophecie, and your yong men shal ſee viſions, and your Ancients shal
dreame dreames. \V And vpon my ſeruants truely, and vpon my handmaides
wil I power out in thoſe daies of my Spirit, and they shal prophecie: \V
and I wil giue wonders in the Heauen aboue, and ſignes in the earth
beneath, bloud and fire,
%%% 2553
and vapour of ſmoke. \V The ſunne shal be turned into darkenes, and the
moone into bloud before the great and manifeſt day of our Lord doth
come. \V And it shal be, euery one whoſoeuer calleth vpon the name of
our Lord, shal be ſaued.}

\V Ye men of Iſrael heare theſe wordes: \Sc{Iesvs} of Nazareth a man
approued of God among you, by miracles & wonders and ſignes which, God
did by him in the middes of you, as you alſo know; \V this ſame
\LNote{By the determinate counſel of God deliuered.}{God
\MNote{Gods determination that Chriſt should die, excuſeth not the Iewes.}
deliuered him, and he deliuered himſelf, for loue and intention of our
ſaluation; & ſo the act was holy and Gods owne determination. But the
Iewes and others which betraied and crucified him, did it of malice
and wicked purpoſe, & their fact was damnable, and not of Gods counſel
or cauſing; though he tolerated it, for that he could and did turne
their abominable fact to the good of our ſaluation. Therfore abhorre
thoſe new Manichees of our time, both Lutherans and Caluiniſts, that
make God the Authour and cauſe of Iudas betraying of Chriſt, no leſſe
then of Paules conuerſion; beſides the falſe tranſlation of
\MNote{Beza.}
Beza, ſaying for Gods preſcience or foreknowledge (in the Greeke,
\G{πρόγνωσις}) \Emph{Gods prouidence}.}
by the determinate counſel & preſcience of God being deliuered, you by
the handes of wicked men haue crucified and ſlaine. \V Whom God hath
raiſed vp
\LNote{Looſing the ſorrowes.}{Chriſt was not in paines himſelf, but
looſed other men of thoſe dolours of Hel, wherewith it was impoſſible
himſelf should be touched. See
\CNote{\Cite{Li.~12. c.~13. de Gen. ad lit.}}
S.~Auguſt.}
looſing the ſorrowes of Hel, according as it was impoſſible that he
ſhould be holden of it. \V For Dauid ſaith concerning him:
\CNote{\XRef{Pſ.~15,~8.}}
\Emph{I foreſaw the Lord in my ſight alwaies: becauſe he is at my right
hand that I be not moued. \V for this, my hart hath been glad and my
tongue hath reioyced: moreouer my flesh alſo shal reſt in hope. \V
Becauſe thou wilt not leaue
\LNote{My ſoul in Hel.}{Where
\MNote{Corrupt tranſlation againſt the Article of Chriſts deſcẽding into
Hel.}
al the Faithful, according to the Creed, euer haue beleeued that Chriſt
according to his ſoul, went downe to Hel, to deliuer the Patriarches and
al iuſt men there holden in bondage til his death, & the Apoſtle here
citing the Prophets wordes, moſt euidẽtly expreſſeth the ſame,
diſtinguishing his ſoule in Hel, from his body in the graue: yet the
Caluiniſts to defend againſt Gods expreſſe wordes, the blaſphemie of
their Maiſter, that Chriſt ſuffered the paines of Hel, and that no where
but vpon the Croſſe, and that otherwiſe he deſcended not into Hel, moſt
falſely and flatly here corrupt the text, by turning and wreſting both
the
\MNote{\H{שפו} \G{ψυχήν} \\ \H{לואש} \G{ᾅδην}}
Hebrew and Greeke wordes from their moſt proper and vſual ſignifications
of, \Emph{Soule}, and \Emph{Hel}, into, \Emph{body}, and \Emph{graue}:
ſaying for, \Emph{my ſoule in Hel}, thus, \Emph{my body,
\CNote{\Cite{Eng. Bible. 1579.}}
life, perſon}, yea (as Beza in his
\Cite{New Teſtament an.~1556.})
\Emph{my carcas} in the \Emph{graue}. And this later they corrupt almoſt
throughout the Bible for that purpoſe. But for refelling of both
corruptions, it shal be ſufficient in this place: firſt, that al Hebrewes
& Greekes, & al that vnderſtãd theſe tongues, know that the foreſaid
Hebrew & Greeke wordes are as proper, peculiar & vſual to
ſignifie \Emph{ſoul} & \Emph{Hel}, as \L{anima} & \L{infernus} in Latin;
yea as \Emph{ſoul} and \Emph{Hel} in Engliſh doe properly ſignifie the
ſoule of man, and Hel that is oppoſite to Heauen; and that they are as
vnproperly vſed to ſignifie body and graue, as to ſay in Engliſh, ſoul
for body, or Hel for graue. Secondly, it doth ſo miſlike the Heretikes
themſelues, that Caſtaleo one of their fine Tranſlatours refelleth it,
and to make it the more ſure, he for \L{in inferno},
tranſlateth, \Emph{in Orco}; that is \Emph{in Hel}. Thirdly, Beza
himſelf partly recanteth in his later edition, and confeſſeth
that, \Emph{Carcas}, was no fit word for the body of Chriſt, & therfore,
\Emph{I haue}, (ſaith he) \Emph{changed it, but I retaine and keepe the
ſame ſenſe ſtil}, meaning, that he hath now tranſlated
it, \Emph{ſoule}, but that he meaneth thereby as before, Chriſts dead
body. Fourthly,
\CNote{\Cite{No. Teſt. an.~1556. annot in 2.~Act. v.~27.}
&
\Cite{24.}
& in
\Cite{1.~Pet.~3,~19.}}
he ſaith plainely that tranſlating thus: \Emph{Thou shalt not leaue my
carcas in the graue}, he did it of purpoſe againſt \L{Limbus Patrum},
Purgatorie, and Chriſts deſcending into Hel; which he calleth four
\Fix{errous,}{errours,}{obvious typo, fixed in other}
and marueleth, \Emph{that moſt of the ancient Fathers were in that
errour}: namely of Chriſts deſcending into Hel, and deliuering the old
Fathers. What need we more? He oppoſeth himſelf both againſt plaine
ſcriptures and al Ancient Fathers, peruerting the one, and contemning
the other, to ouerthrow that truth which is an Article of
our \Emph{Creed}. Whereby it is euidently falſe which ſome of them ſay
for their defenſe, that none of them did euer of purpoſe tranſlate
falſely. See the
\XRef{Annotation vpon 1.~Pet.~3. v.~19.}}
my ſoul
\SNote{Who but an infidel (ſaith S.~Auguſtine) wil deny
\Fix{Ghriſt}{Chriſt}{obvious typo, fixed in other}
to haue deſcẽded to Hel?
\Cite{ep.~99.}}
in Hel, nor giue thy Holy one to ſee
\SNote{As his ſoul ſuffered no paines in Hel, ſo neither did his body
take any corruption in the graue.}
corruption. \V Thou haſt made knowen to me the waies of life: thou shalt
make me ful of ioyfulnes with thy face.} \V Ye men, Brethren, let me
boldly ſpeake to you of the Patriarch Dauid: that
\CNote{\XRef{3.~Reg.~2,~10.}}
he died, and was buried; and his ſepulchre is with vs vntil this preſent
day. \V Whereas therfore he was a Prophet, and knew that by an othe
\CNote{\XRef{Pſ.~131,~11.}}
\Emph{God had ſworne to him, that of the fruit of his loynes there
should ſit vpon his ſeat}; \V forſeeing he ſpake of the reſurrection of
Chriſt. For neither was he left in Hel, neither did his fleſh ſee
corruption. \V This \Sc{Iesvs} hath God raiſed againe, whereof al we are
witneſſes.

\V Being exalted therfore by the right hand of God, and hauing receiued
of his Father the promiſe of the Holy Ghoſt, he hath powred out
%%% Variation is only in other
\Var{this whom}{this gift which}
you ſee and heare. \V For Dauid aſcended not into Heauen; but he ſaith:
\CNote{\XRef{Pſ.~109,~1.}}
\Emph{Our Lord hath ſaid to my Lord, ſit on my right hand, \V vntil I
make thine enemies the foote-ſtoole of thy feet.} \V Therfore let al the
houſe of Iſrael know moſt certainly that God hath made him both Lord, &
\Sc{Christ}, this \Sc{Iesvs}, whom you haue crucified.

\V
\MNote{Contrition.}
And hearing theſe things, they were compunct in hart, and ſaid to Peter and
to the reſt of the Apoſtles: What ſhal
%%% o-2395
we doe men, Brethren? \V But Peter ſaid to them,
\SNote{Not only amendment of life, but penãce alſo required before
Baptiſme, in ſuch as be of age, though not in that ſort as afterward in
the Sacrament of penance.
\Cite{Aug. de fid. & oper. c.~11.}
&
\Cite{ep.~108.}}
doe pennãce, & be euery one of you baptized in the name of \Sc{Iesvs
Christ} for remiſſion of your ſinnes; and you ſhal receiue the guift
of the Holy Ghoſt. \V For to you is the promiſe, and to your children,
and to al that are farre off, whomſoeuer the Lord our God ſhal cal. \V
With very many other wordes alſo did he teſtifie and exhorted them,
ſaying: Saue your ſelues from this peruerſe Generation. \V They therfore
that receiued his word, were baptized & there were added in that day
about
\SNote{Three thouſand were conuerted at this firſt Sermon, & they were
put to the other viſible companie & Church.}
three thouſand ſoules.

\V And they were perſeuering in the doctrine of the Apoſtles, and in the
communication of
\SNote{This was the B.~Sacrament, which the Apoſtles dayly miniſtred to
the Chriſtians at leaſt in one kinde. See
\XRef{c.~20,~7.}}
the breaking of bread, and praiers. \V And feare came vpon euery ſoul;
many wonders alſo and ſignes were done by the Apoſtles in Hieruſalem,
and there was great feare in al. \V Al they alſo that beleeued, were
together,
\CNote{\XRef{Act.~4,~32.}}
& had
\LNote{Al things common.}{This
\MNote{Liuing in cõmon.}
liuing in common is not a rule or a precept to al Chriſtian men, as the
Anabaptiſtes falſely pretend: but a life of perfection and counſel
followed of our Religious in the Catholike Church. See S.~Aug. in
\Cite{Pſ.~112. in principio.}
&
\Cite{ep.~109.}}
al things common. \V Their poſſeſſions and ſubſtance they ſold, and
deuided them to al, according as euery one had need. \V Daily alſo
continuing with one accord in the Temple, and breaking bread from houſe
to houſe, they tooke their meate with ioy and ſimplicitie of hart: \V
praiſing God,
%%% 2554
and hauing grace with al the people. And our Lord
\LNote{Increaſed.}{Moe
\MNote{The increaſe & perpetuitie of the Viſible Church.}
and moe were added to the Church (as the
\TNote{\G{προσετίθει τῆ ἐκκλησία}}
Greeke more plainely expreſſeth) that we may ſee the viſible propagation
& increaſe of the ſame. From which time a diligent man may deduce the
very ſame viſible Societie of men ioyned in Chriſt, through the whole
booke, and afterward by the Eccleſiaſtical ſtorie, downe til our daies,
againſt the pretenſed inuiſible Church of the Heretikes.}  increaſed
them that ſhould be ſaued, daily together.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

