%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Argument
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent argument


\project douay-rheims


%%% 0536
%%% o-0482
\startArgument[
  title={\Sc{The Argument of the Booke of Ivdges.}},
  marking={Judges.}
  ]

Sainct
\MNote{A rule for reading hiſtorical bookes.}
Hierom geuing this general rule
\Cite{(Epiſt. ad Euſtoch. virg.)}
\Emph{that in reading hiſtorical bookes of holie Scripture, the hiſtorie,
as fundation of veritie, is to be loued, but the ſpiritual vnderſtanding
rather to be folowed}: agreably therto teacheth
\Cite{(Epiſt. ad Paulin.)}
that \Emph{in this booke of Iudges there be as manie figures, as princes
of the people}. Neither doth he meane that there were no more, but for
example ſake affirmeth that theſe Iudges, raiſed vp after Ioſue, and
ſent of God to deliuer the people fallen for their ſinnes into
afflictions;
\MNote{The Iudges of Iſrael figures of Chriſts Apoſtles.}
were types and figures of the Apoſtles and Apoſtolical men, ſent by
Chriſt to propagate and defend his Church of the new Teſtament. For
albeit diuers of theſe Iudges were ſometimes great offenders, yet they
were reclamed by Gods ſpecial grace, and ſo amending their errors did
great thinges, to the ſingular honour of God: and are renowmed among the
holie Patriarches and Prophetes, particularly prayſed in holie
Scripture, ſaying:
\CNote{\XRef{Eccl.~46.}}
\MNote{They were al finally holie men.}
\Emph{And the Iudges, euerie one by his name, whoſe hart was not
corrupt: Who were not auerted from our Lord, that their memorie may be
bleſſed, and their bones ſpring out from their place, and their name
remaine for euer, the glorie of holie men remaining to their children.}
After Ioſue therfore, who it ſemeth guided and ruled the people
32.~yeares, this booke, written (as is moſt probable) by Samuel, shewing
\MNote{The Contents of this booke.}
the famouſe Actes of theſe Iudges of Iſrael, proſecuteth the hiſtorie of
the Church the ſpace of 288.~yeares more.
\MNote{Diuided into three partes.}
And may be diuided into three partes. Firſt, is deſcribed in general the
ſtate of the people, ſometimes wel and ſincerly ſeruing God, other times
falling to great ſinnes. In the two firſt chapters. Secondly, their
offences, afflictions, repentance, and deliuerie from their enemies are
more particularly reported. From the third chap. to the 17. Thirdly,
other ſpecial accidents, which happened within the ſame time, are
recorded. In the laſt fiue chapters.


\stopArgument


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
