%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 0558
%%% o-0502
\startChapter[
  title={Chapter 10}
  ]

\Summary{Thola ruleth in Iſrael twentie three yeares. 3.~Iair twentie
  two. 6.~The people fal againe to idolatrie, are afflicted by the
  Philiſthimes, and Ammonites. 10.~They crie to God for helpe, who
  biddeth them cal for helpe to the goddes whom they haue ſerued,
  16.~but crying ſtil to God, and throwing away their idoles, he hath
  compaſsion of them.}

After Abimelech there aroſe Ruler in Iſrael Thola the ſonne of Phua the
vncle of Abimelech, a man of Iſſachar, which dwelt in Samir of mount
Ephraim: \V and iudged
%%% 0559
Iſrael three and twentie yeares, and died and was buried in Samir. \V
After him ſucceded Iair the Galaadite, who iudged Iſrael for two and
twentie yeares, \V hauing thirtie ſonnes ſitting vpon thirtie aſſe
coltes, & princes of thirtie cities, which of his name were called
Hauoth Iair, that is, the townes of Iair, vntil this preſent day in the
Land of Galaad. \V And Iair died, and was buried in the place, which is
called Camon. \V But the children of Iſrael ioyning new ſinnes to their
old, did euil in the ſight of our Lord, & ſerued the Idols, Baalim and
Aſtaroth, & the goddes of Syria and of Sidon and of Moab and of the
children of Ammon and of the Philiſthimes: and they left our Lord, and
did not ſerue him. \V Againſt whom our Lord being wrath, deliuered them
into the handes of the Philiſthijms and of the children of Ammon. \V And
they were afflicted, and ſore opreſſed for eightene yeares, al that
dwelt beyond Iordan in the Land of the Amorrheite, which is in
Galaad: \V in ſo much, that the children of Ammon paſſing ouer Iordan,
waſted Iudas and Beniamin and Ephraim: and Iſrael was afflicted
exceedingly. \V And crying to our Lord, they ſaid: We haue ſinned to
thee, becauſe we haue forſaken our Lord God & haue ſerued Baalim. \V To
whom our Lord ſpake: Haue not the Ægyptians and the Ammorrheites, and
the children of Ammon and the Philiſthijms, \V the Sidonians alſo and
Amalech and Chanaan oppreſſed you, & you cried to me, and I deliuered
you out of their hand? \V And yet you haue forſaken me, and haue
worſhipped ſtrange goddes: therfore I wil not adde to deliuer you any
more: \V goe and inuocate the goddes which you haue choſen: let them
deliuer you in the time of diſtreſſe. \V And the children of Iſrael ſaid
to our Lord: We haue ſinned, tender to vs whatſoeuer pleaſeth thee: only
now deliuer vs. \V In ſaying which thinges,
\SNote{Not euerie one that ſayeth: \Emph{Lord, Lord}, but he that doth
the wil of God, &c.
\XRef{Mat.~7.}}
they threw away out of their coaſtes al the idols of ſtrange goddes, and
ſerued our Lord God: who ſorowed for their miſeries. \V Therfore the
children of Ammon crying together pitcht their tentes in Galaad: againſt
whom the children of Iſrael being aſſembled camped in
%%% o-0503
Maſpha. \V And the princes of Galaad ſaid euerie one to their
neighbours: Who of vs shal firſt beginne to fight againſt the children
of Ammon, ſhal be the duke of the people of Galaad. 


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
