%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 0566
%%% o-0509
\startChapter[
  title={Chapter 14}
  ]

\Summary{Samſon deſirous to marrie a Philiſthime woman. 5.~By the way
  killeth a lion. 8.~In whoſe mouth after few dayes, finding honey,
  12.~he propoſeth therof a riddle to the Philiſthiims for a wager:
  15.~which reueling to his wife, she telleth it to his
  aduerſaries. 19.~He killeth and ſpoyleth thirtie men, ſo payeth the
  wager: and his wife taketh an other man.}

Samſon therfore went downe into Thamnatha, and ſeeing there a woman of
the daughters of the Philiſthims, \V he went vp, and told his father and
his mother, ſaying: I ſaw a woman in Thamnatha of the daughters of the
Philiſthijms: which I beſech you take for me to wife. \V To whom
%%% 0567
his father and mother ſaid:
\SNote{It was prohibited
\XRef{(Deut.~7. v.~3.)}
to make mariage with the Gentiles, but God ſome times diſpenſed: as here
it appeareth he did.
\XRef{v.~4.}}
Is there not a woman among the daughters of thy brethren, and in al my
people, that thou wilt take a wife of the Philiſthijms, which are
vncircumciſed? And Samſon ſaid to his father: Take this for me: becauſe
she hath pleaſed mine eyes. \V But his parentes knew not that the thing
was done of our Lord, and he ſought an occaſion againſt the
Philiſthims. For at that time the Philiſthiims had dominion ouer
Iſrael. \V Samſon therfore went downe with his father and mother into
Thamnatha. And when they were come to the vineyardes of the towne, there
appeared a lions whelpe cruel, and roaring, and mette him. \V And the
Spirit of our Lord came vpon Samſon, and he tore the lion, as if he
ſhould teare a kidde into peeces, hauing nothing at al in his hand: and
this thing he would not tel to his father and mother. \V And he went
downe and ſpake to the woman, that had pleaſed his eies. \V And after
ſome dayes returning to take her, he went aſide to ſee the carcaſſe of
the lion, and behold there was a ſwarme of bees in the mouth of the lion
and a honie combe. \V Which when he had taken in his handes, he did eate
in the way: and coming to his father and mother, he gaue them part, who
alſo them ſelues did eate: neither would he for al that tel them, that
he had taken the honie from the bodie of the lion. \V His father therfore
went downe to the woman, and made his ſonne Samſon a feaſt, for ſo yong
men were accuſtomed to doe. \V When the citizens therfore of that place
had ſeene him, they gaue him thirtie companions to be with him. \V To
whom Samſon ſpake: I wil propoſe you a riddle, which if you ſhal ſolue
me within the ſeuen dayes of the feaſt, I wil geue you thirtie ſindones,
and as many coates: \V but if you shal not be able to ſolue it, you shal
geue me thirtie ſindones, and cotes of the ſame number. Who
%%% o-0510
anſwered him: Propound the riddle, that we may heare it. \V And he ſaid
to them: Out of the eater came forth meate, and out of the ſtrong iſſued
forth ſweetenes. Neither could they for three dayes ſolue the
propoſition. \V And when the ſeuenth day was come, that ſaid to the wife
of Samſon: Speake to thy husband, and vſe perſwaſion to him, that he tel
thee what the riddle ſignifieth. Which thing
\SNote{By threates they made her betray her huſband: and neuertheles
deſtroyed both her and her father.
\XRef{cha.~15. v.~6.}
So perſecuters of the Church deale with ſuch, as trayterouſly or of
frailtie ſerue their turne.}
if thou wilt not doe, we wil burne thee, and thy fathers houſe: haue you
therfore called vs to the bridal that you might ſpoyle vs?
%%% 0568
\V Who shed teares before Samſon, and complained ſaying: Thou hateſt me,
and loueſt me not: therfore the probleme, which thou haſt propounded to
the ſonnes of my people, thou wilt not expound to me. But he anſwered: I
would not tel it to my father and mother: and can I tel it to thee? \V
The ſeuen daies therfore of the feaſt ſhe wept before him: and at the
length of the ſeuenth day for that she moleſted him he expounded it. Who
immediately told her countrie men. \V And they told it him the ſeuenth
day before the going downe of the ſunne: What is ſweeter then honie, and
what ſtronger then a lyon? Who ſaid to them: If you had not ploughed
with my hayſer, you had not found out my propoſition. \V The Spirit
therfore of our Lord came vpon him, and he went downe to Aſcalon, and
ſtroke there thirtie men, whoſe garmentes being taken away he gaue to
them, that had ſolued the probleme. And being exceding wrath he went vp
into his fathers houſe: \V but his wife tooke a husband one of his
freindes and bridal companions.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
