%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 0544
%%% o-0490
\startChapter[
  title={Chapter 5}
  ]

\Summary{The Canticle of Debbora and Barac geuing thankes after their
  victorie.}

And Debbora and Barac the ſonne of Abinoem ſang in that day, ſaying:

\V You that of Iſrael haue voluntarily offered your liues to peril,
\SNote{The greater bleſſe the leſſe by imparting ſpiritual benefites, ſo
God, and ſuperiors bleſſe their ſubiectes. Men bleſſe God, & the leſſe
their betters, by geuing thankes, and prayſes.}
bleſſe our Lord.

%%% 0545
\V Heare you kinges, and geue eare ye princes: I am,
\SNote{She inculcateth that ſhe muſt ſo much more praiſe God for this
victorie, becauſe he forſhewed it by her, & by her directed the general
captaine Barac, leſt it might be aſcribed either to wiſdome or valure of
anie man.}
I am ſhe, that wil ſing to our Lord, I wil chaunte to our Lord the God
of Iſrael.

\V Lord when thou wentſt out of Seir, and didſt paſſe by the countries
of Edom, the earth was moued, & the heauens and cloudes diſtilled
waters.

\V The mountaines melted before the face of our Lord, and Sinni before
the face of our Lord God of Iſrael.

\V In the daies of Samgar the ſonne of Anath, in the daies of Iahel the
pathes reſted: and they that went by them, walked by bywaies.

\V The valiantes in Iſrael ceaſed, and reſted: vntil Debbora aroſe, a
mother roſe in Iſrael.

\V Our Lord choſe new warres, and the gates of the enemies him ſelfe
ſubuerted: ſhield and ſpeare if there appeared among fourtie thouſand of
Iſrael.

\V My hart loueth the princes of Iſrael: you that of your owne good wil
offered your ſelues to danger, bleſſe our Lord.

\V You that ride vpon your
\SNote{Thiſe that ſubdew their bodies to the ſpirite ride vpon fayre
aſſes.
\Cite{Origen, hom.~6. in c.~5. Iudic.}}
faire aſſes, and ſitte in iudgement, and walke in the way, ſpeake.

\V Where the chariottes were
\Fix{frushed}{crushed}{obvious typo, fixed in other}
together, and the armie of the enemie was ſuffocated, there let the
iuſtices of our Lord be told, and his clemencie toward the valiantes of
Iſrael: then did the people of our Lord goe downe to the gates, and
obteyned the principalitie.

\V Ariſe, ariſe Debbora, ariſe, ariſe, and ſpeake a canticle: Ariſe
Barac, and apprehend thy captiues thou ſonne of Abinoem.

\V The remnant of the people is ſaued, our Lord hath fought in the
valiantes.

\V Out from Ephraim he deſtroyed them into Amalec, and after him out
from Beniamin into thy peoples O Amalec: Out from Machir there deſcended
princes, and out from Zabulon they that led the armie to fight.

\V The captaines of Iſſachar were with Debbora, and folowed
%%% o-0491
the ſteppes of Barac, who as it were into a headlong and bottomeles pitte
gaue himſelf to danger: Ruben being diuided againſt it ſelf, there was
found contention of couragious perſons.

\V Why dwelleſt thou betwen the two boundes, that thou mayeſt heare the
whiſtlinges of the flockes? Ruben being
%%% 0546
diuided againſt it ſelf, there was found contention of couragious men.

\V Galaad reſted beyond Iordan, and Dan gaue him ſelf to ſhippes: Aſer
dwelt in the ſea shore, and abode in hauens.

\V But Zabulon and Nepthali offered their liues to death in their
countrie of Merome.

\V The kinges came and fought, the kinges of Chanaan fought in Thanac
beſides the waters of Mageddo, and yet going a praying they tooke nothing.

\V From heauen they fought againſt them: the ſtarres remayning in their
order and courſe, fought againſt Siſara.

\V The torrent of Ciſon drew their carcaſſes, the torrent of Cadumin,
the torrent of Ciſon: my ſoule tread downe the ſtrong ones.

\V The hoofes of the horſes fel of, the ſtrongeſt of the enemies fleeing
violently, and falling downe headlong.

\V Curſe ye the land of Meroz, ſaid the Angel of our Lord: Curſe the
inhabitantes therof, becauſe they came not to help our Lord, to aide his
moſt mightie ones.

\V
\SNote{Iahel the figure was bleſſed amongſt wemen: much more the moſt
holie virgin mother of God is bleſſed aboue al wemen.}
Bleſſed among wemen be Iahel the wife of Haber the Cineite, and bleſſed
be she in her tabernacle.

\V To him that asked water she gaue milke, and in the phial of princes
she offered butter.

\V Her left hand she put to the naile, and her right hand to the ſmithes
hammer, and ſtroke Siſara, ſeeking in his head a place for the wound,
and piercing valiantly through his temple.

\V Betwen her feete he fel: he failed, and died: he was rowled before
her feete, and he lay without life and miſerable.

\V Looking through a window, his mother howled: & she ſpake out of a
higher chamber: Why lingereth his chariote to come backe? Wherfore are
the feete of his waggans ſlow?

\V One wiſer then the reſt of his wiues, anſwered theſe wordes to her
mother in law:

\V Peraduenture now he diuideth the ſpoyles, and the fayreſt of the
wemen is choſen for him: garmentes of ſundrie colours are deliuered to
Siſara for a praye, and diuerſe furniture is laid together to adorne the
neckes.

\V So perish al thine enemies O Lord: but they that loue thee, as the
Sunne shineth in his ryſing, ſo let them glitter.

\V And the Land reſted for fourtie yeares.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
