%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 0582
%%% o-0522
\startChapter[
  title={Chapter 21}
  ]

\Summary{The tribe of Beniamin is repayred, 8.~by foure hundred virgins
  reſerued in the ſlaughter of Iabes Galaad: 19.~and by other virgins
  taken, that come forth of Silo to daunce.}

The children of Iſrael ſware alſo in Maſpha, and ſaid: None of vs shal
geue of his daughters to the children of Beniamin to wife. \V And they
came al to the houſe of God in Silo, and ſitting in his ſight vntil
euening, lifted vp their voice, and with great wayling beganne to weepe
ſaying: Wherfore ô Lord God of Iſrael is this euil done in thy people,
that this day one tribe ſhould be taken away out of vs? \V And on the
morrow riſing early, they built an altar: and offered there holocauſtes,
and pacifique victimes, and ſaid: \V Who hath not aſcended in the hoſte
of our Lord of al the tribes of Iſrael? For they had bound them ſelues
with a great othe, when they were in Maſpha, that they ſhould be ſlayne
which had beene wanting. \V And the children of Iſrael being
%%% 0583
\SNote{Leſt either iuſtice be ouer ſharpe, or mercie too relaxe, with
great art of diſcretion, gouernours muſt obſerue \Emph{mercie iuſtly
aduiſing, and diſcipline piouſly chaſticing}.
\Cite{S.~Greg. li.~1. Epiſt.~24.}}
moued with repentance vpon their brother Beniamin, beganne to ſay: One
tribe is taken away out of Iſrael, \V whence shal they take wiues? For
we haue al ſworne in common, that we wil not geue our daughters to
them. \V Therfore they ſaid: Who is there of al the tribes of Iſrael,
that went not vp to our Lord into Maſpha? And behold the inhabitantes of
Iabes Galaad were found not to haue bene in that armie. (\V At that
time alſo when they were in Silo, none of them was found there.) \V They
ſent therfore ten thouſand the ſtrongeſt men, and commanded them: Goe,
and ſtrike the inhabitantes of Iabes Galaad in the edge of the ſword, as
wel their wiues as their litle ones. \V And this ſhal
%%% o-0523
be it which you shal obſerue: Al of the male kinde, and wemen, that haue
knowen men, kil ye, but the virgins reſerue. \V And there were found of
Iabes Galaad foure hundred virgins, that knew not mans bedde, and they
brought them to the campe in Silo, into the Land of Chanaan. \V And they
ſent meſſengers to the children of Beniamin, that were in Rocke Remmon,
and commanded them that they should receiue them in peace. \V And the
children of Beniamin came at that time, and there were geuen vnto them
wiues of the daughters of Iabes Galaad: but others they found not, which
they might geue them in like maner. \V And al Iſrael was very ſorie,
and repented for the killing of one tribe out of Iſrael. \V And the
ancientes ſaid: What shal we doe to the reſt, that haue not taken wiues?
For al the wemen in Beniamin are dead. \V And we muſt very carefully,
and with great ſtudie prouide, that one tribe be not deſtroyed out of
Iſrael. \V For our owne daughters we can not geue them, being bound with
an oath and a curſe, wherby we ſaid: Curſed be he that shal geue to
Beniamin anie of his daughters to wife. And they tooke counſail, and
ſaid: Behold there is an anniuerſaire ſolemnitie of our Lord in Silo,
which is ſituate on the North of the citie of Bethel, on the Eaſt ſide
of the way, that goeth from Bethel to Sichem, and on the South of the
towne of Lebona. \V And they commanded the children of Beniamin, and
ſaid: Goe, and lie hidde in the vineyardes. \V And when you shal ſee the
daughters of Silo come forth after the maner to lead daunſes, iſſue
forth ſodenly out of the vineyardes, and catch of them euerie one his
wife, and goe into the Land of Beniamin. \V And when their fathers shal
%%% 0584
come, and their brethren, and shal beginne to complaine againſt you, and
to chide, we wil ſay to them: Haue pittie on them: for they tooke them
not away by the right of warryers and conquerours, but when they deſired
to receiue them, you gaue them not, and on your part the fault was
committed. \V And the children of Beniamin did, as it had beene
commanded them: and according to their number, they tooke away to them
ſelues of thoſe that ledde the daunſes, euerie one his wife: and they
went into their poſſeſſion, building cities, and dwelling in them. \V
The children of Iſrael alſo returned by their tribes, and families into
their tabernacles. In thoſe daies there was not a King in Iſrael: but
euerie one did that
\SNote{In the time of the Iudges the people preſumed more to
do \Emph{that ſemed to them ſelues right}, or good, though it was
nought; which afterwardes the kinges more reſtrained and puniſhed.}
which ſemed right to him ſelfe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
