%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 0576
%%% o-0517
\startChapter[
  title={Chapter 19}
  ]

\Summary{A Leuite bringing homeward his reconciled wife, 15.~at Gabaa in
  the tribe of Beniamin hardly getteth lodging. 25.~His wife is there
  vilanouſly abuſed by wicked men, and in the morning found
  dead. 29.~Whereupon her huſband cutteth her bodie, and ſendeth peeces
  to euerie tribe of Iſrael, requiring them to reuenge the wicked fact.}

There was a certaine man a Leuite, dwelling on the ſide of mount
Ephraim, who tooke a wife of Bethlehem Iuda: \V which left him, and
returned vnto her fathers houſe into Bethlehem, and abode with him foure
monethes. \V And her huſband folowed her, willing to be reconciled
vnto her, and to ſpeake her fayre, and to bring her backe with him,
hauing in his companie a ſeruant and two aſſes: who receiued him, and
brought him into her fathers houſe. Which when his father in law had
heard, and had ſeene him, he mette him ioyful, \V and embraced the
man. And the ſonne in law taried in the houſe of his father in law three
daies, eating with him and drinking familiarly. \V But the fourth day
aryſing before day, he would depart. Whom his father in law held, and
ſaid to him: Taſt firſt a litle bread, & ſtrengthen thy ſtomacke, and ſo
thou shalt depart. \V And they ſate together, and did eate and
drinke. And the father of the yong
%%% 0577
woman ſaid to his ſonne in law: I beſech thee that thou tarie here to
day, and let vs make merie together. \V But he ryſing vp, beganne as if
he would depart. And neuertheles with much adoe his father in law ſtayed
him, and made him to tarie with him. \V But
%%% o-0518
when morning was come, the Leuite prepared to goe his iourney. To whom
his father in law againe: I beſech thee, quoth he, that thou take a
litle meate, and making thy ſelf ſtrong, til the day be farder ſpent,
afterward thou mayeſt depart. They did eate therfore together. \V And
the yong man aroſe, that he might ſette forward with his wife and
ſeruant. To whom his father in law ſpake againe: Conſider that the day
is more declining to the weſt, and draweth nigh to euening: tarie with
me to day alſo, and ſpend the day in mirth, and to morrow thou shalt
depart that thou mayſt goe into thy houſe. \V His ſonne in law would
not condeſcend to his wordes: but forthwith went forward, and came ouer
againſt Iebus, which by an other name is called Ieruſalem, leading with
him two aſſes loden, and his
\SNote{She was his lawful wife and ſo called
\XRef{v.~1. &~9.}
yet is alſo called concubine becauſe ſhe had no dawrie, nor as yet
enioyed the priuiledges of a miſtris in her huſbands houſe.}
concubine. \V And now they were come nigh to Iebus and the day changed
into night: & the ſeruant ſaid to his maiſter: Come, I beſech thee, let
vs turne into the citie of Iebuſeites, and tarie in it. \V To whom his
maiſter anſwered: I wil not enter into the towne of a ſtrange nation,
which is not of the children of Iſrael, but I wil paſſe as farre as
Gabaa: \V and when I shal come thither, we wil lodge in it, or at the
leaſt in the citie of Rama. \V They paſſed therfore by Iebus, and went
on their iourney begone, and the ſonne went downe to them byſide Gabaa,
which is in the tribe of Beniamin: \V and they turned into it, that they
might lodge there. Whither when they were entred, they ſate in the
ſtreate of the citie, and no man would receiue them to lodge. \V And
behold there appeared an old man, returning out of the field and from
his worke in the euening, who him ſelf alſo was of mount Ephraim, and
dwelt as a ſtranger in Gabaa; but the men of that countrie were the
children of Iemini. \V And lifting vp his eies, the old man ſaw the man
ſitting with his fardels in the ſtreate of the citie, and ſaid to him:
Whence comeſt thou? and whither goeſt thou? \V Who anſwered him: We
departed from Bethlehem Iuda, and we goe to our place, which is on the
ſide of mount Ephraim, from whence we went into Bethlehem: and now we
goe to the houſe of God, and none
%%% 0578
wil receiue vs vnder his roofe, \V hauing ſtraw and hay for prouender of
the aſſes, and bread and wine for the vſe of my ſelf and of thy handmaid,
and of the ſeruant that is with me: we lacke nothing but lodging. \V To
whom the old man anſwered: Peace be with thee, I wil geue al thinges
that are neceſſarie: only, I beſech thee, tarie not in the ſtreate. \V
And he brought him into his houſe, and gaue prouender to his aſſes: and
after they had waſhed their feete, he receiued them to a bankette. \V
They making merie, and after the labour of their iourney, refreshing
their bodie with meate and drinke, there came men of that citie, the
children of Belial (that is to ſay, without yoke) and beſetting the old
mans houſe, beganne to knocke at the doores, crying to the maiſter of
the houſe, and ſaying: Bring forth the man, that entred into thy houſe,
that we may abuſe him. \V And the old man went out to them, and ſaid:
Doe not ſo brethren, doe not this
%%% o-0519
euil: becauſe this man is entered to my lodging, and ceaſe from this
folie: \V I haue a daughter that is a virgin, and this man hath a
concubine, I wil bring them forth to you, that you may humble them, &
fulfil your luſt: only, I beſeche you, worke not this wickednes againſt
nature on the man. \V They would not agree to his wordes which the man
ſeing, he brought forth his concubine to them, and he deliuered her to them
to be illuded: whom when they had abuſed al the night, they let her goe
in the morning. \V But the woman, when the darkenes departed, came to
the doore of the houſe, where her lord lodged, and there fel downe. \V
Morning being come, the man aroſe, and opened the doore, that he might
finiſh his iourney begone: and behold his concubine lay before the
doore, her handes ſpredde on the threſhold. \V To whom he, thinking that
ſhe tooke her reſt, ſpake: Ariſe, and let vs walke. Who anſwering
nothing, perceiuing that ſhe was dead; he tooke her, and laid her vpon
his aſſe, & returned into his houſe. \V Which when he was entered vnto,
he tooke a ſword, and cutting the carcaſſe of his wife with her bones
into twelue partes and peeces, he ſent them into al the borders of
Iſrael. \V Which when euerie one had ſeene, they cried together: There
was neuer ſuch a thing done in Iſrael, from that day, when our fathers
aſcended out of Ægypt, vntil this preſent time: geue ſentence, and
decree in common what is needeful to be done.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
