%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 0563
%%% o-0506
\startChapter[
  title={Chapter 12}
  ]

\Summary{Ephraites riſing againſt Iephte, fourtie two thouſand of them
  are ſlaine. 8.~Abeſan is Iudge. 11.~After him Ahialon. 13.~Then Abdon.}

But behold in Ephraim there aroſe a ſedition. For they paſſing againſt
the North, ſaid to Iephte: Going to fight againſt the children of Ammon,
why wouldſt thou not cal vs, that we might goe with thee? Therfore we
wil burne thy houſe. \V To whom he anſwered: I and my people were at
great ſtrife againſt the children of Ammon: and I called you, that you
should ayde me, and you would not doe it. \V Which I ſeeing
\SNote{That is, expoſed my ſelf to danger truſting to Gods helpe & myn
owne handes, when others would not aſſiſt me.}
put my life in myne owne handes, and paſſed to the children of Ammon,
and our Lord deliuered them into my handes. What haue I deſerued, that
you riſe againſt me in battel. \V Therfore al the men of Galaad being
called to him,
%%% 0564
he fought againſt Ephraim: and the men of Galaad ſtroke Ephraim, becauſe
he had ſaid:
\SNote{Iephte being of Manaſſes tribe the Ephraites enuied his glorie
and calumniouſly obiected that he and his followers were fugitiues ſo
raiſed a tumulte to their owne loſſe.}
Galaad is a fugitiue of Ephraim, and dwelleth in the middes of Ephraim
and Manaſſes. \V And the Galaadites tooke the fordes of Iordan, by the
which Ephraim was to returne. And when there had come to the ſame one of
the number of Ephraim, fleeing, and had ſaid: I beſech you let me paſſe:
The Galaadites ſaid to him: Art thou not an Ephraite? Who ſaying: I am
not: \V they asked him: Say then Schibboleth, which is interpreted an
Eare of corne. Who anſwered, Sibboleth, not being able by the ſame
letter to expreſſe, an eare of corne. And immediatly being apprehended
they killed him in the very paſſage of Iordan. And there fel at that
time of Ephraim two and fourtie thouſand. \V Therfore Iephte the
Galaadite iudged Iſrael ſix yeares: and he died, and was buried in his
citie of Galaad. \V After him Abeſan of Bethlehem iudged Iſrael: \V who
had thirtie ſonnes, and as manie daughters, which he ſending
%%% o-0507
abrode,
gaue to huſbandes, and tooke wiues for his ſonnes of the ſame number,
bringing them into his houſe. Who iudged Iſrael ſeuen yeares: \V and
died and was buried in Bethlehem. \V To whom ſucceeded Ahialon a
Zabulonite: and he iudged Iſrael ten yeares: \V and he died and was
buried in Zabulon. \V After him Abdon iudged Iſrael, the ſonne of Illel
a Pharathonite: \V who had fourtie ſonnes, and of them thirtie nephewes,
mounting vpon ſeuentie aſſe coltes, and he iudged Iſrael eight
yeares: \V and he died, and was buried in Pharathon of the Land of
Ephraim, in the mount of Amalec.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
