%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 0549
%%% o-0494
\startChapter[
  title={Chapter 7}
  ]

\Summary{Gedeon marching with thirtie two thouſand men, al that are
  fearful, and that drinke water kneeling are dimiſsed, & onlie three
  hundred that drinke litle remaine. 9.~By a Madianites dreame Gedeon is
  encouraged. 16.~By a ſtratageme the enemie is frighted &
  ouerthrowne. 24.~The Ephraites kil Oreb and Zeb.}

Therfore Ierobaal, which is alſo Gedeon, ryſing in the night, and al the
people with him, came to the fountaine that is called Harad, and the
campe of Madian was
%%% 0550
in the valley on the North ſide of the high hil. \V And our Lord ſaid to
Gedeon: There is much people with thee, neither ſhal Madian be deliuered
into their hand: Leſt Iſrael glorie againſt me, and ſay: By myne owne
force I am deliuered. \V Speake to the people, and proclame in al their
hearing: He that is fearful and timorous, let him returne. And they
departed from mount Galaad, and there returned two and twentie thouſand
men, and onlie ten thouſand remained. \V And our Lord ſaid to Gedeon:
Yet there is a great multitude, leade them to the waters, and there I
wil proue them: and of whom I shal forbid to goe, let him returne. \V
And when the people were come downe to the waters, our Lord ſaid to
Gedeon: They that ſhal lappe the water with their tongues, as dogges are
wont to lappe, thou ſhalt ſeparate them apart: but they that ſhal drinke
bowing downe their knees, ſhal be on the other part. \V The number
therfore of them that had lappe
\Fix{watter,}{water,}{likely typo, fixed in other}
their hand caſting it to their
mouth, was three hundred men: and al the reſt of the multitude had
drunck kneeling. \V And our Lord ſaid to Gedeon: In the three hundred
men, that lapped water, I wil deliuer you, and geue Madian in thy hand:
but let al the reſt of the multitude returne into their place. \V Taking
therfore victuals and trumpettes
%%% o-0495
according to the number, al the reſt of the multitude he commanded to
depart to their tabernacles: and him ſelfe with the three hundred gaue
him ſelfe to the battel. And the campe of Madian was beneath in the
valley. \V The ſame night our Lord ſaid to him: Ariſe, and goe downe
into the campe: becauſe I haue deliuered them into thy hand. \V But if
thou be afraid to goe alone, let Phara thy ſeruant goe downe with
thee. \V And when thou shalt heare what they ſpeake, then shal thy
handes be ſtrengthned, and thou shalt goe downe more ſecure to the
enemies campe. He therfore went downe and Phara his ſeruant into part of
the campe, where the watch was of men in armes. \V But Madian and
Amalec, and al the Eaſt peoples lay ſcattered in the valley, as a
multitude of locuſtes: their camels alſo were innumerable, as the ſand
that lieth in the ſea shore. \V And when Gedeon was come, one told his
neighbour
\SNote{Obſeruation of dreames is generally forbid.
\XRef{Leuit.~19. v.~26.}
\XRef{Deut.~18. v.~10.}
Yet here, and in other places, it is euident, God would haue ſome
obſerued. See
\XRef{Annot. Gen.~40.}}
a dreame: & in this manner he reported that which he had ſeene: I ſaw a
dreame, & there ſemed to me as it were a harth loafe of barlie to roll,
& to come
%%% 0551
into the campe of Madian: and when it was come to the tabernacle, it
ſtroke it, and ouerthrew it, and beate it al flatte with the earth. \V
He to whom he ſpake, anſwered: This is no other thing, but the ſword of
Gedeon the ſonne of Ioas the man of Iſrael. For the Lord hath deliuered
Madian into his handes, and al their campe. \V And when Gedeon had heard
the dreame, and the interpretation therof, he adored: and turned to the
campe of Iſrael, and ſaid: Ariſe ye, for our Lord hath deliuered the
campe of Madian into our handes. \V And he diuided the three hundred men
into three partes, and gaue them trumpettes in their handes, and
\SNote{Theſe thinges were ridiculous (ſaith Venerab. Beda
\Cite{c.~5. qq.~in Iudic.}
if they had not bene terrible to the enemies.}
emptie pitchers, and lampes in the middes of their pitchers. \V And he
ſaid to them: What you shal ſee me doe, that doe ye: I wil enter into
part of the campe, and that which I shal doe folow you. \V When the
trumpet shal ſound in my hand, doe you alſo ſound and crie together
round about the campe: To our Lord
\SNote{It is no derogation to God, that honour is alſo geuen to his
ſeruantes.}
and to Gedeon. \V And Gedeon went in, and the three hundred that were
with him, into part of the campe, the watch of midnight beginning, and
rayſing vp the watch men they began to ſound with their
\SNote{Trumpetes ſignified preachers of Chriſt, pitchers the bodies of
Martyres, & lampes their vertues and miracles.
\Cite{Vener. Beda. qq.~in Iudic. c.~3.}}
trumpettes, and to clappe the pitchers one againſt an other. \V And when
they ſounded in three places round about the campe, and had broken the
pitchers, they held the lampes in the left handes, and with the right
they ſounded the trumpettes, and cried: The ſword of our Lord and of
Gedeon: \V ſtanding euerie one in his place round about the enemies
campe. Therfore the whole campe was trubled, and crying out and howling
they fled: \V and the three hundred men neuertheles perſiſted ſounding
with the trumpettes. And our Lord ſent in the ſword in al the campe, and
they murdered one an other, \V fleeing as farre as Bethſetta, and the
brinke of Abelmehula in Tebbath. But the men of Iſrael of Nepthali, and
Aſer shoutting together, and al Manaſſes purſewed Madian. \V And Gedeon
ſent meſſengers into al mount Ephraim, ſaying: Come downe to meete
Madian, and take the waters before them to Bethbera and Iordan. And al
Ephraim
%%% o-0496
shouted, and tooke the waters before them and Iordan vnto Bethbera. \V
And two men that were apprehended of Madian, Oreb, and Zeb: Oreb he ſlew
in the Rocke of Oreb, Zeb in the Preſſe of Zeb. And they purſewed
Madian, carying the heades of Oreb and Zeb to Gedeon beyond the ſtreames
of Iordan.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
