%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 0543
%%% o-0488
\startChapter[
  title={Chapter 4}
  ]

\Summary{Againe the people ſinning are oppreſsed, and crying to God,
  4.~by direction of Debbora a propheteſſe, Barac fighteth againſt
  Siſara their enemie. 15.~Who ſtricken with feare, and fleeing away,
  17.~Iahel pretending his ſaftie killeth him in her houſe.}

And the children of Iſrael added to doe euil in the ſight of our Lord
after the death of Aod, \V and our Lord deliuered them into the handes
of Iabin the king of Chanaan, which reigned in Aſor: and he had a
captaine of his armie named Siſara, and he dwelt in Haroſeth of the
gentiles. \V And the children of Iſrael cried to our Lord: for he had
nine hundred yron hooked chariottes, and for twentie yeres had
vehemently oppreſſed them. \V And there was one Debbora a propheteſe the
wife of Lapidoth, which
\SNote{Being a propheteſſe ſhe reſolued hard and obſcure thinges, but
exerciſed no iuriſdiction in anie cauſes, for that belonged to the councel
of Prieſtes and of ſeuentie ancientes, where the high prieſt was the
chief Iudge.
\XRef{Num.~11.}
\XRef{Deut.~17.}
Spiritually Debbora ſignified the Church & Barac chriſtian Princes, who
are directed in their warres, and other actions by ſpiritual ſuperiors,
as Origen, and other ancient writers expound this hiſtorie.}
iudged the people at that time. \V And ſhe ſate vnder a palme tree,
which was called by her name, betwen Rama and Bethel in the mount of
Ephraim: and the children of Iſrael went vp to her for al iudgement. \V
Who ſent, and called Barac the ſonne of Abinoem of Cedes in Nepthali:
and ſhe ſaid to him: Our Lord God of Iſrael hath commanded thee, Goe,
and lead an armie into mount Thabor, and thou ſhalt take with thee ten
thouſand fighting men of the children of Nepthali, and of the children
of Zabulon: \V and I wil bring vnto thee in the place of the Torrent
Ciſon, Siſara the prince of the hoſte of Iabin, and his chariotes, and
al the multitude, and wil deliuer them in thy hand. \V And Barac ſaid to
her: If thou come with me, I wil goe: If thou wilt not come with me, I
wil not goe. \V Who ſaid to him: I wil goe in deede with thee, but at
this time the victorie ſhal not be imputed to thee, becauſe Siſara ſhal
be deliuered into the hand of a woman.
%%% o-0489
Debbora therfore aroſe, and went with Barac into Cedes. \V Who calling
vnto him Zabulon and Nepthali, went vp with ten thouſand fighting men,
hauing Debbora in his companie. \V And Haber the Cineite was in time
paſt departed from the reſt of the Cineites his brethren of Hobab, the
coſin of Moyſes: and had pitched his tentes vnto the valley, which is
called Sennim, and was nere Cedes. \V And it was told Siſara, that Barac
the ſonne of Abinoem was gone vp into mount Thabor: \V and he gathered
nine hundred yron hooked chariottes, and al his armie from Haroſeth of
the gentiles
%%% 0544
to the torrent Ciſon. \V And Debbora ſaid to Barac: Ariſe, for this is
the day, wherin our Lord hath deliuered Siſara into thy handes: behold
he is thy leader. And Barac went downe from mount Thabor, and ten
thouſand fighting men with him. \V And our Lord terrified Siſara, and al
his chariotes, and al the multitude in the edge of the ſword at the
ſight of Barac: in ſo much, that Siſara leaping downe from his chariote,
fled a foote, \V and Barac purſewed the chariotes fleeing, and the armie
vnto Haroſeth of the gentiles, and al the multitude of the enemies was
vtterly deſtroyed. \V But Siſara fleeing came to the tent of Iahel the
wife of Haber the Cineite. For there was peace betwixt Iabin the king of
Aſor, and the houſe of Haber the Cineite. \V Iahel therfore going forth
to meete Siſara, ſaid to him: Come in to me my Lord, come in, feare
not. Who being entred into her tabernacle, and couered of her with a
cloke, \V ſaid to her: Geue me, I beſeech thee, a litle water, for I am
very thirſtie. Who opened a bottle of milke, and gaue him to drinke, and
couered him. \V And Siſara ſaid to her: Stand before the doore of the
tabernacle, and when any shal come asking thee, and ſaying: Is there any
man here? Thou shalt ſay: There is none. \V
\SNote{VVho is this woman ful of confidence, piercing the temples of the
enimies head with a naile, but the faith of the Church deſtroying the
diuels kingdomes with the croſſe of Chriſt?
\Cite{S.~Aug. li.~12. c.~32. cont. Fauſt. Manich.}
Iahel alſo prefigured our B.~Ladie who cruſhed the ſerpentes head.}
Iahel therfore the wife of Haber tooke a nayle of the tabernacle, taking
withal a hammer alſo: and going in ſecretely, and with ſilence she put
the naile vpon the temple of his head, and ſtriking it with the hammer,
faſtened it into his braine euen to the ground: who ioyning deepe ſleepe
and death together, faynted, and died. \V And behold Barac folowing
Siſara came: & Iahel going forth to meete him, ſaid to him: Come, and I
wil shew thee the man whom thou ſeekeſt. Who when he was entered in vnto
her, ſaw Siſara lying dead, and the nayle faſtened in his temples. \V
Therfore God that day humbled Iabin the king of Chanaan before the
children of Iſrael: \V which increaſed daily, and with ſtrong hand
oppreſſed Iabin the king of Chanaan, til they deſtroyed him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
