%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 0537
%%% o-0483
\startChapter[
  title={Chapter 1}
  ]

\Summary{Vnder
\MNote{The firſt part.

A general recapitulation of the peoples ſtate.}
a general captaine of the tribe of Iuda, aſsiſted by the tribe of
Simeon, Iſrael ſubdueth diuers cities of the gentiles. (12.~Othoniel
taking Cariath ſepher poſseſseth it, and marieth Calebs daughter,
obtaining alſo addition of her dowrie.) 21.~Iebuſeites yet dwel in
Hieruſalem with Beniamin, 27.~and the Chananeites with diuers of the
tribes.}

After the death of Ioſue the children of Iſrael
\SNote{The maner of conſulting our Lord was by the High prieſt praying
in the tabernacle.
\XRef{Exo.~29. v.~42.}}
conſulted our Lord, ſaying: Who ſhal goe vp before vs againſt the
Chananeite, and ſhal be captaine of the warre? \V And our Lord ſaid:
\SNote{The firſt general captaine after Ioſue, and diuers of the Iudges,
were of the tribe of Iuda, but not al, as appeareth in this booke.}
Iudas ſhal goe vp: behold I haue deliuered the Land into his handes. \V
And Iudas ſaid to Simeon his brother: Come vp with me into my lotte, and
fight againſt the Chananeite, that I alſo may goe forward with thee into
thy lotte. And Simeon went with him. \V And Iudas went vp, and our Lord
deliuered the Chananeite, and the Pherezeite into their handes: and they
ſtroke in Bezec ten thouſand men. \V And they found Adonibezec in Bezec,
and fought againſt him, and ſtroke the Chananeite, and the
Pherezeite. \V And Adonibezec fled: whom purſewing they tooke, cutting
of the extreme partes of his handes and feete. \V And Adonibezec ſaid:
Seuentie kinges hauing the extreme partes of their handes and feete cut
of, gathered vp the reliques of meates vnder my table: as I haue done,
ſo hath God repayed me. And they brought him into Ieruſalem, and there
he died. \V Therfore the children of Iudas aſſaulting Ieruſalem, tooke
it, and ſtroke it in the edge of the ſword, ſetting the whole citie on
fyre. \V And afterward going downe they fought againſt the Chananeite,
which dwelled in the mountaines, and
%%% o-0484
ſouthward, and in the champaine countries. \V And Iudas going forward
againſt the Chananeite, that dwelled in Hebron (the name wherof was
before time Cariatharbe) ſtroke Seſai, and Ahiman, and Tholmai: \V and
departing thence went to the inhabitantes of Dabir, the old name wherof
was Cariath Sepher, that is, a citie of letters. \V And Caleb ſaid: He
%%% 0538
that shal ſtrike Cariath Sepher, and ſpoile it, I wil geue him Axa my
daughter to wife. \V And when Othoniel the ſonne of Cenez, the yonger
brother of Caleb had taken it, he gaue him Axa his daughter to wife. \V
Whom going on her way her husband admoniſhed to aske a field of her
father. To whom, when she had ſighed ſitting on her aſſe, Caleb ſaid:
What aileth thee? \V But she anſwered: Geue me a bleſſing, for a drie
land thou haſt geuen me: geue me alſo a waterie. Caleb therfore gaue her
a waterie ground aboue, & waterie beneath. \V And the children of the
Cineite the coſin of Moyſes went vp from the citie of palmes, with the
children of Iudas into the deſert of his lotte, which is at the ſouth
ſide of Arad, and dwelt with him. \V But Iudas went with Simeon his
brother, and together they ſtroke the Chananeite that dwelt in Sephaath,
and ſlew him. And the name of the citie was called, Horma, that is,
Anathema. \V And Iudas tooke Gaza with the coaſtes therof, & Aſcalon,
and Accaron with their boundes. \V And our Lord was with Iudas, and he
poſſeſſed the mountaines: neither could he deſtroy the inhabitantes of
the valley, becauſe they had manie
\SNote{Strong weapons crooked like ſickles made faſt to the chariottes
which cut in peeces, men, horſes, and other chariottes that came in
their way.}
hooked chariotes. \V And they gaue to Galeb Hebron, as Moyſes had ſaid,
who deſtroyed out of it the three ſonnes of Enac. \V But the Iebuſeite
the inhabiter of Ieruſalem the children of Beniamin deſtroyed not: and
the Iebuſeite dwelt with the children of Beniamin in Ieruſalem vntil this preſent
day. \V The houſe alſo of Ioſeph went vp into Bethel, and our Lord was
with them. \V For when they beſieged the citie, which before was called
Luza, \V they ſaw a man comming out of the citie, and ſaid to him: Shew
vs the entrance of the citie, and we wil shew thee mercie. \V Who when
he had shewed them, they ſtroke the citie in the edge of the ſword: but
that man, and al his kinred they diſmiſſed. \V Who being diſmiſt, went
into the Land of the Hetthims, and built there a citie, and called it
Luza: which is ſo called vntil this preſent day. \V Manaſſes alſo
deſtroyed not Bethſan, and Thanac with their litle townes, and the
inhabitantes of Dor, and Ieblaam, and Mageddo with their litle
townes. And the Chananeite began to dwel with them. \V But after that
Iſrael was waxen ſtrong, he made them tributaries, and would not deſtroy
them. \V Ephraim alſo killed not the Chananeite, that dwelt in Gazer,
but dwelt with him. \V Zabulon deſtroyed not the
%%% 0539
inhabitantes of Cetron, & Naalol: but the Chananeite dwelt in the middes
of him, and was made tributarie to him. \V Aſer alſo deſtroyed not the
inhabitantes of Accho, and of Sidon, of Ahalab, and Achazib, and Helba,
and Aphec, and Rohob: \V and he dwelt in the middes of the Chananeite
the inhabiter of that Land, neither did he kil them. \V Nepthali alſo
deſtroyed not the inhabitantes of Bethſames, & Bethanath: and
%%% o-0485
he dwelt in the middes of the Chananeite the inhabiter of the Land, and
the Bethſamites & Bethanites were tributaries to him. \V And the
Amorrheite ſtraytened the children of Dan in the mountaine, and gaue
them not place to goe downe to the playne: \V and he dwelt in mount
Hares which is interpreted shelles, in Aialon and Salebim. And the hand
of the houſe of Ioſeph was agrauated, and he became tributarie
to him. \V And the 
border of the Amorrheite was from the Aſcent of the Scorpion, the rocke,
and the higher places.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
