%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 0539
%%% o-0485
\startChapter[
  title={Chapter 2}
  ]

\Summary{An Angel reciting manie benefites of God towardes Iſrael, and
  their ingratitude, 4.~they weepe for their faultes. 10.~After the
  death of Ioſue and other ancientes of his time, the people often fal,
  and repenting are deliuered from afflictions, 19.~but ſtil fal againe
  worſe and worſe.}

And the
\SNote{An Angel taking the forme of a man as before to Ioſue,
\XRef{(ch.~5.)}
ſo now appearing to the people, ſpake to them in the name of God, whoſe
meſſenger he was.}
Angel of our Lord went vp from Galgal to the place of weepers, and
ſaid: I brought you out of Ægypt, and haue brought you into the Land,
for the which I ſware to your fathers: and I promiſed that I would not
make fruſtrate my couenant with you for euer: \V onlie ſo that you
should not make a league with the inhabitantes of this Land, but should
ouerthrow their altares: and you would not heare my voice: why haue you
done this? \V For the which cauſe I would not deſtroy them from before
your face: that you may haue enemies, and their goddes may be a ruine
vnto you. \V And when the Angel of our Lord ſpake theſe wordes to al the
children of Iſrael: they lifted vp their voice, and wept. \V And the
name of that place was called, the place of weepers, or of teares: and
\SNote{By ſpecial diſpenſation ſacrifice was ſometimes lawfully offered
in other places though the Tabernacle (and afterward the Temple) was the
onlie place commanded.
\XRef{Deut.~12.}
\XRef{Ioſue.~22.}
\Cite{S.~Aug. q.~36. in Iudic.}}
there they immolated hoſtes to our Lord. \V Ioſue therfore diſmiſſed the
people, and the children of Iſrael went euerie one into his poſſeſſion,
to obtayne it: \V  and they ſerued our Lord al his daies, and the daies
of the ancientes, of them that liued a long time after him, and knew al
the workes of our Lord, which
%%% 0540
he had done with Iſrael. \V And Ioſue the ſonne of Nun, the ſeruant of
our Lord, died, being a hundred and ten yeares old, \V and they buried
him in the borders of his poſſeſſion in Thamnathſare in the mount of
Ephraim, on the North ſide of mount Gaas. \V And al that generation was
gathered
\Fix{to to}{to}{obvious typo, fixed in other}
their fathers: and there roſe others, that knew not our Lord, and the
workes which he had done with Iſrael. \V And the children of Iſrael did
euil in the ſight of our Lord, and ſerued Baalim. \V And they left our
Lord the God of their fathers, that had brought them out of the Land of
Ægypt: and folowed ſtrange goddes, and the goddes of the peoples, that
dwelt round about them,
%%% o-0486
and adored them: and they prouoked our Lord to anger, \V leauing him,
and ſeruing Baal and Aſtaroth. \V And our Lord being wrath againſt
Iſrael, deliuered them into the handes of ranſackers: who tooke them and
ſold them to the enemies, that dwelt round about: neither could they
reſiſt their aduerſaries: \V but whither ſoeuer they had meant to goe,
the hand of our Lord was vpon them, as he ſpake, and ſware to them: and
they were vehemently afflicted. \V And our Lord
\SNote{Theſe Iudges were extraordinarily raiſed vp to deliuer the
people repenting, when they were fallen into afflictiõs for their
ſinnes.}
rayſed vp Iudges, that should deliuer them from the handes of the
waſters: but neither would they heare them, \V fornicating with ſtrange
goddes, and adoring them. They did quickly forſake the way, in the which
their fathers had gone: and hearing the commandementes of our Lord, they
did al thinges contrarie. \V And when our Lord rayſed vp Iudges, in
their daies he was moued with mercie, and heard the groninges of the
afflicted, and deliuered them from the ſlaughter of the waſters. \V But
after the Iudge was dead, they returned, and did much worſe things then
their fathers had done, folowing ſtrange goddes ſeruing them, and
adoring them. They left not their inuentions, and the verie hard way, by
which they were accuſtomed to walke. \V And the furie of our Lord was
angrie againſt Iſrael, & ſaid: Becauſe this nation hath made my couenãt
fruſtrate, which I had made with their fathers, & hath cõtemned my
voice: \V I alſo wil not deſtroy the nations which Ioſue did let alone,
and died: \V that in them I may trie Iſrael, whether they wil keepe the
way of the Lord, and walke in it, as their fathers kept it, or no. \V
Our Lord therfore left al theſe nations, and would not quickly ouerthrow
them, neither deliuered them into the handes of Ioſue.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
