%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 0554
%%% o-0498
\startChapter[
  title={Chapter 9}
  ]

\Summary{Abimelech Gedeons concubins ſonne killeth his brethren, 7.~onlie
  the yongeſt eſcapeth, & by a parable expoſtulateth the iniurie donne
  to his fathers houſe. 23.~Shortly the Sichemites deteſt
  Abimelech. 26.~Gaal conſpireth againſt him, but is
  ouercome. 50.~Finally Abimelech is wounded to death by a woman.}

And
\SNote{This ſonne of Gedeon by his ſeruant prefigured Antichriſt who wil
perſecute the Church and reigne for a while: but in the end ſhal be
deſtroyed. 
\Cite{S.~Beda c.~6. qq. in Iudic.}}
Abimelech the ſonne of Ierobaal went into Sichem to his mothers brethren
and ſpake to them, and to al the kinred of the houſe of his mothers
father, ſaying: \V Speake to al the men of Sichem: whether it is better
for you, that ſeuentie men haue dominion ouer you al the ſonnes of
Ierobaal, or that one man haue dominion ouer you? and withal conſider
that I am your bone, and your fleſh. \V And his mothers brethren ſpake
of him to al the men of Sichem, al theſe wordes, and inclined their
hartes after Abimelech, ſaying: He is our brother. \V And they gaue him
ſeuentie weight of ſiluer out of the temple of Baalberit. Who hyred
therewith vnto him ſelf needie men and vagaboundes, and they folowed
him. \V And he came into his fathers houſe in Ephra, and murdered his
brethren the ſonnes of Ierobaal
%%% 0555
ſeuentie men, vpon one ſtone: and there remayned Ioatham the youngeſt
ſonne of Ierobaal, and was hidde. \V And al the men of Sichem aſſembled
together al the families of the citie of Mello: and they went & made
Abimelech king, beſide the oke, that ſtood in Sichem. \V Which being told
to Ioatham, he went,
\SNote{True paſtores in the time of Antichriſt wil ſtill auouch the
truth and the right of the Church.}
and ſtoode in the toppe of mount Garizim: and lifting vp his voice, he
cried, and ſaid: Heare me ye men of Sichem, ſo as God may heare you. \V
\LNote{The trees went to annoint a king.}{According
\MNote{Vngratful people render iniuries for benefites.}
to the hiſtorie Ioatham Gedeons youngeſt ſonne, by a parable iuſtly
expoſtulateth the iniurie donne by the Sichemites to his fathers houſe,
in preferring a baſe bound womans ſonne, and cruelly murdering the reſt
of his ſonnes: who with much trauel, and manie dangers of his owne life,
had deliuered them from ſeruitude. But in the ſpiritual ſenſe, which (as
the ancient fathers note) is chiefly intended,
\MNote{Infidels promote wicked men to authoritie.}
Idolaters and Heretikes are reproued, who rather accept of vniuſt
vſurpers, that wil ſerue their licentious appetites, and mantaine vice
and wickednes, then to be ruled by iuſt and lawful Superiors, appointed
by Gods ordinance, indued with grace of the Holie Ghoſt (ſignified by
\Emph{the oliue tree};) ſuch as bring forth wholſome ſwete vertues
(ſignified by \Emph{the figge tree};) and are repleniſhed with admirable
fortitude (ſignified by the vine tree) and in their places ſet vp baſe,
ambitious, cruel, and crabbed ſpirites, ſignified by the bramble, or
brere.
\MNote{Abimelech a figure of Antichriſt.}
Thus Nemrod, Abimelech, Mahomet, and innumerable other tyrantes haue
benne aduanced, & eſpecially Antichriſt
\CNote{\XRef{2.~Theſſ.~2.}}
\Emph{shal be extolled aboue al that is called God, or is vvorshipped},
and ſhal moſt cruelly perſecute al Catholiques, that wil not conforme
them ſelues to his procedings. But in fine (as here is prefigured in
Abimelech) \Emph{fire shal riſe againſt this bramble Antichriſt, and shal
deuoure him and al his together}.
\Cite{S.~Beda. qq.~in lib. Iudic. c.~6.}}
The trees went to anointe a king ouer them: and they ſaid to the
\SNote{Oyle ſpiritually ſignifieth the grace of the Holie Ghoſt, making
peace of conſcience in mens ſoules towardes God.}
oliue tree: Reigne ouer vs. \V Which anſwered: Can I forſake my fatnes,
which both goddes do vſe, and men, and come to be promoted among the
trees? \V And the trees ſaid to the
\SNote{The ſwetnes of Gods law producing good workes.}
figge tree: Come, and take the kingdome ouer vs. \V Which anſwered them:
Can I forſake my ſweetnes, and my moſt ſweete fruites, and go to be
promoted among the other trees? \V And the trees ſpake to the
\SNote{Contemtible in outward ſhew, but bringing forth liquour of
meruelous force; which ſorte of workes God is moſt delighted withal: and
men moſt admire.
\XRef{Pſal.~85.}}
vine: Come, and reigne ouer
%%% o-0499
vs. \V Which anſwered them: Can I forſake my wine, that chereth God and
men, & be promoted among the other trees? \V And the trees ſaid to
\SNote{The rhamnus ſignifieth baſe and ambicious men.}
the
\TNote{Brierre, bramble, or thiſtle.}
rhamnus: Come, and reigne ouer vs. \V Who anſwered them: If in deede you
make me your king, Come, and reſt vnder my ſhadow: but if you meane it
not, let there fyre iſſue forth of the rhamnus, and deuoure the ceders
of Libanus. \V Now therfore, if you haue wel, and without ſinne
appointed Abimelech king ouer you, and haue dealt wel with Ierobaal, and
with his houſe, and haue requitted him the like for his benefites, who
fought for you, \V and put his life in dangers, that he might deliuer
you from the hand of Madian, \V who now are riſen againſt my fathers
houſe, and haue killed his ſonnes ſeuentie men vpon one ſtone, and haue
made Abimelech the ſonne of his handmaide king ouer the inhabitantes of
Sichem, becauſe he is your brother. \V If therfore you haue dealt wel,
and without fault with Ierobaal, and his houſe, reioyce this day in
Abimelech, and reioyſe he in you. \V But if vniuſtly: let there fyre
iſſue forth from him, and conſume the inhabitantes of Sichem, and towne
of Mello: and let there fire goe forth from the men of Sichem, and from
the towne of Mello, and deuoure Abimelech. \V Which thinges when he had
ſaid, he fled, and went into Bera: and dwelt there for feare of Abimelech
his brother. \V Abimelech therfore reigned ouer Iſrael three yeares. \V
And our Lord ſent
\SNote{God doth ſuggeſt only good cogitations: as remorſe of conſcience
in the Sechemites, for their ingratitude towardes Gedeon and for ſo
wicked and cruel a murder of his ſonnes. Wherupon they begãne to deteſt
Abimelech and, ſo hatred grew betwen him & them, which is a moſt euil
ſpirit. But their former ſinne, not God, was the cauſe therof.
\Cite{S.~Aug. q.~45. in Iudic.}}
a verie euil ſpirit betwen
%%% 0556
Abimelech and the inhabitantes of Sichem: Who began to deteſt him, \V and
to lay the wickednes of the murdering of the ſeuentie ſonnes of
Ierobaal, and the ſheading of their bloud vpon Abimelech their brother,
and vpon the reſt of the princes of the Sichemites, that had holpen
him. \V And they ſette an ambuſhment againſt him on the toppe of the
mountaines: and whiles they taryed for his coming, they committed
robberies, taking prayes of them that paſſed by: and it was told
Abimelech. \V And Gaal the ſonne of Obed came with his brethren, and
paſſed into Sichem. At whoſe coming the inhabitantes of Sichem taking
courage, \V iſſued forth into the fieldes, waſting the vineyardes, and
treading the grapes: and gathering companies of muſicions went into the
temple of their god, and in the bankettes and cuppes curſed Abimelech. \V
Gaal the ſonne of Obed crying: Who is Abimelech, and what is Sichem, that
we ſhould ſerue him? Is he not the ſonne of Ierobaal, and hath made
Zebul his ſeruant prince ouer the men of Emor the father of Sichem? Why
then shal we ſerue him? \V Would God that ſome man would geue this
people vnder my hand, that I might take Abimelech out of the way. And
ſome ſaid to Abimelech: Gather together a multitude of an armie, and
come. \V For Zebul the prince of the citie, hearing the wordes of Gaal
the ſonne of Obed, was very wrath, \V and ſent meſſengers ſecretely to
Abimelech, ſaying: Behold, Gaal the ſonne of Obed is come into Sichem
with his brethren, and rayſeth the citie againſt thee. \V Ariſe therfore
in the night with the people, that is with thee, and lie hidde in the
field: \V and betimes in the morning at ſunne ryſing, ſette vpon the
citie. And when he iſſueth forth
%%% o-0500
againſt thee with his people, do to him what thou ſhalt be able. \V
Abimelech therfore aroſe with al his armie in the night, and laide
ambuſhementes nere to Sichem in foure places. \V And Gaal the ſonne of
Obed went forth, & ſtoode in the entrance of the gate of the citie. And
Abimelech roſe, and al his armie with him from the place of the
ambuſhmentes. \V And when Gaal had ſeene the people, he ſaid to Zebul:
Behold a multitude cometh downe from the mountaines. To whom he
anſwered: Thou ſeeſt the ſhadowes of the mountaines as it were heades of
men, and with this errour thou art deceiued. \V Againe Gaal ſaid: Behold
there cometh downe people from the nauel of the land, and one trouppe
cometh by the way,
%%% 0557
that looketh to the oke. \V To whom Zebul ſaid: Where is now thy mouth,
wherwith thou didſt ſpeake? Who is Abimelech that we ſhould ſerue him? Is
not this the people, which thou didſt deſpiſe? Goe forth, and fight
againſt him. \V Gaal therfore went, the people of the Sichemites looking
on, and fought againſt Abimelech, \V who purſewed him fleeing, and driue
him into the citie: and there were ſlaine of his part manie, vnto the
gate of the citie: \V and Abimelech ſate in Ruma: but Zebul expelled
Gaal, and his companions out of the citie, neither did he ſuffer them to
abide in it. \V Therfore the day folowing the people went forth into the
field. Which being told to Abimelech, \V he tooke his armie, and diuided
it into three troupes, ſetting ambuſhmentes in the fieldes. And ſeeing
that the people came out of the citie, he aroſe, & ſet vpon them \V with
his owne troupe, oppugning and beſieging the citie: and two troupes
ſcattered through the field purſewed the aduerſaries. \V Moreouer
Abimelec al that day oppugned the citie: which he tooke, & killed the
inhabitantes therof, and deſtroyed it, ſo that he
\SNote{For more reuenge he ſowed the citie with ſalte, which maketh
ground barren.
\Cite{Theod. q.~17. in lib. Iudic.}}
ſowed ſalt in it. \V Which when they had heard that dwelt in the towre
of Sichem, they entered into the temple of their god Berith, where they
had made a couenant with him, and therof the place had taken his name,
which was exceding wel fenſed. \V Abimelech alſo hearing that the men of
the towre of Sichem were gathered together, \V he went vp into mount
Selmon with al his people: and taking an axe, he cut of the bough of a
tree, and laying it on his ſhoulder & carying it, he ſaid to his
companions: That which you ſee me do, doe ye out of hand. \V They
therfore cutting of boughes from the trees, euerie man as faſt as he
could, folowed their captaine. Who compaſſing the forte burnt it: and ſo
it came to paſſe, that with the ſmoke and the fyre a thouſand perſons
were ſlaine, men and wemen together, of the inhabitantes of the towre of
Sichem. \V And Abimelech departing thence came to the towne of Thebes,
which compaſſing he beſieged with his armie. \V And there was in the
middes of the citie an high towre, to the which were fled both men and
wemen together, and al the princes of the citie, the gate being ſhut
very ſtrongly, and they ſtanding vpon the batlementes of the towre by
the bulwarkes. \V And Abimelech coming nere the towre, fought manfully:
%%% 0558
and approching to the doore, endeuoured to put fire vnder it:
%%% o-0501
\V and behold one woman caſting from aboue a peece of a milſtone,
\SNote{Euels ſhal betide the vniuſt man to deſtruction.
\XRef{Pſal.~139.}}
daſhed it againſt the head of Abimelech, and brake his brayne. \V Who
called by and by his eſquire, and ſaid to him: Draw out thy ſword, and
ſtrike me: leſt perhaps it be ſaid that I was ſlaine of a woman. Who
doing as he was commanded ſlew him. \V And when he was dead, al that
were with him of Iſrael, returned into their ſeates: \V and God repayed
the euil, that Abimelech had done againſt his father, killing his
ſeuentie brethren. \V The Sichemites were alſo rewarded for that which
they had wrought, and the curſe of Ioatham the ſonne of Ierobaal came
vpon them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
