%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 0552
%%% o-0496
\startChapter[
  title={Chapter 8}
  ]

\Summary{The Ephraimites quareling becauſe they were not called to the
  warres, are pacified by Gedeon. 4.~The men of Soccoth and Phanuel
  denying victuals for the campe, Gedeon (in the meane time
  10.~ouerthrowing the enemie) 15.~reuengeth their reprochful contempt,
  18.~killeth Sebee and Salmana, 22.~refuſeth dominion, 24.~but
  receiueth as a gift, the iewels taken in the praye: 27.~maketh therof
  an Ephod, which turneth to the ruine of his familie. 30.~Hauing
  ſeuentie ſonnes by his wiues, and one by a concubine, dieth in good
  old age. 33.~And the people fal againe to idolatrie.}

And the men of Ephraim ſaid to him: What is this that thou didſt meane
to doe, that thou wouldeſt not cal vs when thou didſt goe to fight
againſt Madian? chyding bitterly and almoſt offering violence. \V To
whom he anſwered:
\SNote{A ſoft anſwer breaketh anger: hard ſpeach ſtirreth vp furie.
\XRef{Prou.~35.}}
What could I haue done like to that, which you haue done. Is not the
cluſter of Ephraim better then the vintages of Abiezer? \V Into your
handes hath our Lord deliuered the princes of Madian, Oreb and Zeb, what
could I haue done the like as you haue done? Which when he had ſpoken,
their ſpirit reſted, wherwith they did ſwel againſt him. \V And when
Gedeon was come to Iordan, he paſſed ouer it with the three hundred men,
that were with him: and for wearines, they could not purſew them that
fled. \V And he ſaid to the men of Soccoth: Geue, I beſeech you, bread
to the people, that is with me, becauſe they are verie faint: that we
may purſew Zebee, and Salmana the kinges of Madian. \V The princes of
Soccoth anſwered: Peraduenture the palmes of the handes of Zebee and
Salmana are in thy hand, & therfore thou requireſt that we geue bread to
thy armie. \V To whom he ſaid: When our Lord therfore shal haue
deliuered Zebee and Salmana into my handes, I wil teare your flesh with
the thornes, and briers of the deſert. \V And going vp from thence, he
came into Phanuel: and he ſpake to the men of that place the like
thinges. To whom they alſo anſwered, as the men of Soccoth had
anſwered. \V He ſaid therfore to them alſo: When I shal be returned
conquerour in peace, I wil deſtroy this towre. \V But Zebee and Salmana
reſted with al their armie. For fiftene thouſand men were remayning of
al the troupes of the Eaſt peoples, an hundred and twentie thouſand
fighting men and thoſe that drew ſword,
%%% o-0497
being ſlayne. \V And Gedeon going vp by the way of them,
%%% 0553
that dwelt in tabernacles, on the Eaſt ſide of Nobe, and Iegbaa, ſtroke
the campe of the enemies, which were ſecure, and ſuſpected no
miſchance. \V And Zebee and Salmana fled, whom Gedeon purſewing
apprehended, al their hoſte being put out of aray. \V And returning from
the battel before ſunne ryſing, \V he tooke a boy of the men of Soccoth:
and he asked him the names of the princes and ancientes of Soccoth, and
he deſcribed ſeuentie ſeuen men. \V And he came to Soccoth, and ſaid to
them: Behold Zebee, and Salmana concerning whom you vpbraided me,
ſaying: Peraduenture the handes of Zebee and Salmana are in thy handes,
and therfore thou deſireſt that we geue bread to the men that be wearie,
and are fainte. \V He tooke therfore the ancientes of the citie, and
thornes and briers of the deſert, and tore them with the ſame, and cut
the men of Soccoth into pieces. \V The tower alſo of Phanuel he ouerthrew,
killing the inhabitantes of the citie. \V And he ſaid to Zebee and
Salmana: What manner of men were they, whom you ſlew in Thabor? Who
anſwered: Like vnto thee, and one of them as it were the ſonne of a
king. \V To whom he anſwered: They were my brethren, the ſonnes of my
mother. Our Lord liueth, that if you had ſaued them,
\SNote{Zebee and Salmana were not of anie of the ſeuen nations, whom God
commanded to deſtroy and therfore Gedeon might haue ſpared their liues,
if he would.}
I would not kil you. \V And he ſaid to Iether his eldeſt ſonne: Ariſe,
and kil them. Who drew not out his ſword: for he was afraid, becauſe he
was yet a boy. \V And Zebee and Salmana ſaid: Doe thou riſe, and runne
vpon vs: becauſe according to his age is the ſtrength of a man. Gedeon
roſe vp, and ſlew Zebee and Salmana: and he tooke the ornamentes and
boſſes, wherwith the neckes of kinges camels are wont to be adorned. \V
And al the men of Iſrael ſaid to Gedeon:
\SNote{Kinges may do anie thing not contrarie to the law: but Iudges &
Dukes may onlie do according to the law. See
\XRef{1.~Reg.~8.}}
Rule thou ouer vs, and thy ſonne, and thy ſonnes ſonne: becauſe thou
haſt deliuered vs from the hand of Madian. \V To whom he ſaid: I wil not
rule ouer you, neither shal my ſonne rule ouer you, but our Lord shal
rule ouer you. \V And he ſaid to them: One petition I requeſt of you:
Geue me the earlettes of your praye. For the Iſmalites were accuſtomed
to haue golden earlettes. \V Who anſwered: We wil geue them moſt
willingly. And ſpreading a mantel on the ground, they caſt on it the
earlettes of the praye: \V and the weight of the earlettes that he
deſired, was a thouſand fiue hundred ſicles of gold, beſides the
ornamentes, and iewels, and purple veſture, which the kinges of Madian
were
%%% 0554
wont to vſe, and beſides the golden chaines of the camels. \V And Gedeon
made therof an Ephod, and put it in his citie Ephra. And al Iſrael did
fornicate in it, and it became a ruine to Gedeon and to al his houſe. \V
But Madian was humbled before the children of Iſrael, neither could they
any more lift vp their neckes: but the land reſted for fourtie yeares,
wherin Gedeon ruled. \V Ierobaal therfore the ſonne of Ioas went, and
dwelt in his owne houſe: \V and he had ſeuentie ſonnes, which came out
of his thigh, becauſe he had manie wiues. \V And his
\SNote{His handmaide or ſeruant not a harlotte: to wit, ſuch a one as
had not the priuiledge of a wife. As
\XRef{Gen.~25. v.~6.}}
concubine, which he had in Sichem, bare him a ſonne named Abimelec. \V
And
%%% o-0498
Gedeon the ſonne of Ioas died in a good old age, & was buried in the
ſepulchre of his father in Ephra of the familie of Ezri. \V But after
that Gedeon was dead, the children of Iſrael were auerted, and did
fornicate with Baalim. And they made a couenant with Baal, that he
should be their God: \V neither did they remember our Lord their God,
which deliuered them out of the handes of al their enemies round about:
\V neither did they mercie with the houſe of Ierobaal Gedeon, according
to al the benefites that he had done to Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
