%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 0568
%%% o-0510
\startChapter[
  title={Chapter 15}
  ]

\Summary{Samſon tying firebrandes to foxes tayles burneth the
  Philiſthiims corne. 6.~They burne his wife and her father. 8.~He
  beateth them and hideth him ſelfe. 10.~His owne countrimen to get
  peace with the Philiſthimes, take and bind him, ſo meaning to deliuer
  him. 14.~But he breaketh the cordes, and with the iaw bone of an aſſe
  killeth a thouſand of his enemies. 18.~Being exceding drie, is
  refreshed with water, from the tooth of the ſame iaw.}

And after a certaine time, when the dayes of wheate harueſt were at
hand, Samſon came, meaning to viſite his wife, and he brought her a
kidde of goates. And when he would enter into her chamber as he was
wont, her father prohibited him, ſaying: \V I thought that thou hadſt hated
her, and therfore I deliuered her to thy freind: but she hath a ſiſter,
which is yonger & fayrer then she, let this be thy wife in ſteade of
her. \V To whom Samſon anſwered: From this day there shal be no fault in
me againſt the Philiſthimes: for I wil doe you euils. \V And he went,
and
\SNote{Being Iudge of the people he had helpe of others to take ſo manie
foxes with nettes, or otherwiſe, being great ſtore in that countrie.}
caught three hundred foxes, and he coupled them tayle to tayle, and tyed
%%% o-0511
fyrebrandes in the middes: \V which kindling with fyre, he let them goe,
that they might runne abroade hither and thither. Who immediately went
on into the corne of the Philiſthimes.
%%% 0569
Which being ſette on fire, both the corne now caried together, and that
which yet ſtoode in the ſtalke, was al burnt, in ſo much, that the flame
conſumed the vineyardes alſo and the oliuetes. \V And the Philiſthijmes
ſaid: Who hath done this thing? To whom it was ſaid: Samſon the ſonne in
law of the Thamnathate: becauſe he tooke his wife, and gaue her to an
other, he hath wrought theſe thinges. And the Philiſthims went vp, and
burnt both the woman and her father. \V To whom Samſon ſaid: Although
you haue done theſe thinges, notwithſtanding yet wil I requite reuenge
of you, and then I wil reſt. \V And he ſtroke them with a great plague,
ſo that aſtonied they laid the calfe of the leg vpon the thigh. And
going downe he dwelt in the caue of the rocke Etam. \V Therfore the
Philiſthijms going vp into the Land of Iuda camped in the place, which
afterward was called Lechi, that is, the iaw bone, where their armie was
ſpred abrode. \V And they of the tribe of Iuda ſaid to them: Why are you
come vp againſt vs? Who anſwered: That we may bind Samſon, we are come,
and may repay him the thinges that he hath wrought againſt vs. \V There
went downe therfore three thouſand men of Iuda, to the caue of the flint
Etam, and ſaid to Samſon: Knoweſt not thou that the Philiſthijms reigne
ouer vs? Why wouldeſt thou doe this thing? To whom he ſaid: As they did
to me ſo haue I done to them. \V To bind thee, quoth they, we are come,
and to deliuer thee into the handes of the Philiſthijms. To whom Samſon:
Sweare, quoth he, & promiſe me that you kil me not. \V They ſaid: We wil
not kil thee, but wil deliuer
\Fix{the}{thee}{obvious typo, fixed in other}
bound. And they bound him with two new cordes, and tooke him from the
rocke Etam. \V Who when he was come to the place of the Iawebone and the
Philiſthijms shouting were come againſt him, the Spirit of our Lord fel
vpon him: and as flax is wont to be conſumed at the fauour of fyre, ſo
the bandes wherwith he was bound, were diſſipated and looſed. \V And
finding a iawe bone, to witte, the iawe bone of an aſſe, which lay
there, catching it,
\SNote{A notorious miracle to kil ſo manie with ſo meane a weapon
without other helpe of man. And by common reaſon as vncredible, as the
great myſteries of Catholique Religion.}
he ſlewe therwith a thouſand men, \V and ſaid: In the iawe bone of an
aſſe, in the iawe of the colt of ſhe aſſes haue I deſtroyed them,
and haue ſtrooke a thouſand men. \V And when he had ended theſe wordes
ſinging, he threw the iaw bone out of his hand, and called the name of
that place Ramathlechi, which is interpreted the
%%% 0570
lifting vp of the iawbone. \V And being very thirſtie, he cried to our
Lord, and ſaid: Thou haſt geuen in the hand of thy ſeruant this very
great ſaluation and victorie: and behold I die for
\Fix{thriſt,}{thirſt,}{obvious typo, fixed in other}
and shal fal into the handes of the vncircumciſed. \V Our Lord therfore
\SNote{It was a greater miracle to draw water out of a drie bone, then
out of the earth or ſtones: but al things are poſſible to God, which he
pleaſeth to do.}
opened a great tooth in the iawe of the aſſe, & there iſſued out of it
waters. Which being drunke, he refreſhed his ſpirit, and receiued
ſtrength againe. Therfore the name of that place was called: The
fountaine of him that inuocated from the iawbone, vntil this preſent
day. \V And he iudged Iſrael in the daies of the Philiſthijms twentie
yeares.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
