%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 0541
%%% o-0486
\startChapter[
  title={Chapter 3}
  ]

\Summary{The
\MNote{The ſecond part.

Of the common peoples often falling to idolatrie, their repentance, and deliuerie.}
people aſſociating them ſelues with Gentiles, againſt whom they ought to
fight, 8.~are inuaded by forraine kinges: 9.~but repenting are deliuered
by Othoniel. 12.~Falling againe, afflicted, and repenting, 15.~are
deliuered by Aod, 21.~ſecretely killing Eglon their enemie. 31.~After
him Samgar defendeth Iſrael againſt the Philiſthimes.}

Theſe are the Nations, which our Lord left, that in them he might
inſtruct Iſrael, and al that had not knowen the warres of the
Chananeites: \V that afterward their children might lerne to fight with
their enemies, and to be accuſtomed to warre: \V the fiue princes of the
Philiſthimes, and the Chananeite, and Sidonian, and Heueite, that dwelt
in mount Libanus, from mount Baal Hermon to the entring into Emath. \V
And he left them, that in them he might trie Iſrael, whether they would
heare the commandementes of our Lord, which
%%% o-0487
he had commanded their
fathers by the hand of Moyſes, or not. \V Therfore the children of
Iſrael dwelt in the middes of the Chananeite, and Hetheite, and
Amorrheite, and Pherezeite, and Heueite, and Iebuſeite: \V and they
tooke their daughters to wiues, and them ſelues gaue their owne
daughters to their ſonnes, & ſerued their goddes. \V And they did euil
in the ſight of our Lord, and forgat their God, ſeruing Baalim and
Aſtaroth. \V And our Lord being wrath againſt Iſrael, deliuered them
into the handes of Chuſan Raſathaim the king of Meſopotamia, and they
ſerued him eight yeares. \V And they cried to our Lord: who rayſed them
vp
\SNote{In manie places we ſee the worde \Emph{ſauiour}, and like titles
geuen to men, as the ſeruantes and officers of God, who is the proper
and principal Sauiour of al.
\Cite{S.~Aug. q.~18. in Iudic.}}
a ſauiour, and deliuered them, to witte: Othoniel the ſonne of Cenez, the
yonger brother of Caleb: \V and the Spirit of our Lord was in him, and
he iudged Iſrael. And he went forth to fight, and our Lord deliuered
into his handes Chuſan Raſathaim the king of Syria, and oppreſſed
him. \V And the land reſted
\SNote{In theſe 40.~yeares are included the eight yeares of their
ſeruitude,
\XRef{v.~8.}
& ſo in the reſt of this hiſtorie. Otherwiſe the number of yeres agreeth
not with the count
\XRef{3.~Reg.~6. v.~1.}}
fourtie yeares, and Othoniel the ſonne of Cenez died. \V And the
children of Iſrael added to doe euil in the ſight of our Lord: who
ſtrengthned againſt them Eglon the king of Moab: becauſe they did euil
in his ſight. \V And he ioyned to him the children of Ammon, and Amalec:
and he went and ſtroke Iſrael, and poſſeſſed the Citie of palmes. \V And
the children of Iſrael ſerued Eglon the king of Moab eightene yeares: \V
and afterward they cried to our Lord: who rayſed vp vnto them a ſauiour
called Aod, the ſonne of Gera, the ſonne of Iemini, who vſed both
%%% 0542
handes for the right. And the children of Iſrael ſent by him preſentes
to Eglon the king of Moab. \V Who made him ſelf a two edged ſword,
hauing in the middes a hafte in length the palme of a hand, and was
girded therwith vnder his caſſocke on the right thigh. \V And he
preſented the giftes to Eglon the king of Moab. And Eglon was exceeding
groſſe. \V And when he had preſented the giftes vnto him, he brought his
felowes on the way that came with him. \V And returning from Galgal, where
the Idols were, he ſaid to the King: I haue a ſecrete meſſage to thee, ô
king. And he commanded ſilence: and al being gone forth, that were about
him, \V Aod went in to him: and he ſate in a ſommer chamber alone, and
he ſaid:
\SNote{Aod hauing ſpecial inſpiration from God to do this fact, (as
S.~Auguſtin noteth vpon theſe wordes
\Cite{q.~20. in Iudic.)}
is not to be imitated by priuat men. See
\XRef{Num.~25. v.~21.}}
A word from God I haue to thee. Who forthwith roſe out of his throne. \V
And Aod put forth his left hand, and tooke the dagger from his right
thigh, and faſtened it into his bellie \V ſo mightely that the haft
folowed the blade in the wound, and was cloſed vp faſt with the moſt
fatte greaſe. Neither did he plucke out the dagger, but as he had
ſtriken ſo leift it in the bodie: and forthwith by the ſecrete partes of
nature the ordure of the bellie came forth. \V But Aod shutting the
doores of the chamber very diligently, and locking them ſure, \V went
out by a poſterne doore. And the kinges ſeruantes going in, ſaw the
doores of the chamber shut, and they ſaid: Peraduenture he purgeth his
bellie in the ſommer chamber. \V And expecting long til they were
ashamed, and ſeeing that no man did open, they tooke a key: and opening
they found their lord on the earth lying dead. \V But Aod, whiles they
were trubled, eſcaped,
%%% o-0488
and paſſed by the Place of Idols, whence he had returned. And he came
into Seirath: \V and forthwith the trumpet ſounded in the mount of
Ephraim: and the children of Iſrael went downe with him, him ſelf going
in the front. \V Who ſaid to them: Folow me: for our Lord hath deliuered
our enemies the Moabites into our handes. And they went downe after him,
and occupied the fordes of Iordan, which bring ouer into Moab: and they
ſuffered no man to paſſe: \V but they ſtroke the Moabites at that time,
about ten thouſand, al ſtoute and ſtrong men, none of them could
eſcape. \V And Moab was humbled that day vnder the hand of Iſrael: and
the Land reſted eightie yeares. \V After him was Samgar the ſonne of
Anath, who ſtroke, of the Philiſthimes ſix hundred men with the culter
of a ploughe: and he alſo defended Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
