%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 0579
%%% o-0519
\startChapter[
  title={Chapter 20}
  ]

\Summary{Al the other tribes fighting againſt Beniamin, 13.~becauſe they
  wil not punish the malefactors, 21.~haue the worſe, 25.~alſo the
  ſecond time. 29.~But the third time the Beniamites are al ſlaine
  ſauing ſix hundred men.}

Therfore al the children of Iſrael went forth, and were gathered
together, as it were one man, from Dan to Berſabee, and the Land of
Galaad, to our Lord in Maſpha: \V and al the corners of the people, and
al the tribes of Iſrael aſſembled into the church of the people of God
foure hundred thouſand footemen warriers. (\V Neither were the children
of Beniamin ignorant that the children of Iſrael were come vp into
Maſpha.) And the Leuite the huſband of the woman that was killed being
asked, how ſo great wickednes had beene cõmitted, \V anſwered: I came
into Gabaa of Beniamin with my wife, and there I tooke my lodging: \V
and behold the men of that citie by night beſette the houſe wherein I
taried, meaning to kil me, and vexing my wife with incredible furie of
luſt, finally she died. \V Whom being taken I did cutte into peeces, and
ſent the partes into al the borders of your poſſeſſion: becauſe neuer
was there ſo heinous an offenſe, and ſo great an abomination done in
Iſrael. \V You are al preſent the children of Iſrael, determine what
%%% o-0520
you ought to doe. \V And al the people ſtanding, anſwered as it were by
the word of one man: we wil not depart into our tabernacles, neither
shal any man enter into his houſe: \V but this wil we doe in common
againſt Gabaa. \V Let ten men be choſen of an hundred out of al the
tribes of Iſrael, and an hundred of a thouſand, and a thouſand of
ten thouſand, to bring victuals for the armie, and that we may fight
againſt Gabaa of Beniamin, & render to it for the wicked fact, which it
deſerueth. \V And al Iſrael aſſembled to the citie, as it were one man
with one minde, and one counſel: \V and they ſent meſſengers to al the
tribe of Beniamin, which should ſay: Why is there ſoe great abomination
found in you? \V Deliuer the men of Gabaa, that haue committed this
heinous fact,
\SNote{Omiſſion & contempt to puniſh hainous crimes is a iuſt cauſe to
make warre againſt anie people.}
that they may die, and the euil may be taken away out of Iſrael. Who
would not heare the commandment of their brethren the children of
Iſrael: \V but out of al cities, which were of their lotte, they
aſſembled into Gabaa, to ayde them, and to fight againſt al the people
of Iſrael. \V And there were found
%%% 0580
fiue and twentie thouſand of Beniamin of them that drew ſword, beſide
the inhabitantes of Gabaa, \V which were ſeuen hundred moſt valiant men,
ſo fighting with the left hand as with the right: and ſo directly
caſting ſtones with ſlinges, that they could ſtrike a heare alſo, and
the ſtroke of the ſtone ſhould not be caried awry on either part. \V Of
the men of Iſrael alſo, beſide the children of Beniamin, were found
foure hundred thouſande of them that drew ſwordes, & were prepared to
fight. \V Who riſing came into the houſe of God, that is, into Silo: and
they conſulted God, and ſaid: Who shal be in our armie general of the
battel againſt the children of Beniamin? To whom our Lord anſwered: Let
\SNote{One of the tribe of Iuda.}
Iudas be your captayne. \V And forthwith the children of Iſrael aryſing
in the morning, camped beſide Gabaa: \V and thence proceeding to fight
againſt Beniamin, beganne to aſſault the citie. \V And the children of
Beniamin iſſuing out of Gabaa, ſlew of the children of Iſrael that day
two and twentie thouſand men. \V Agayne Iſrael hauing confidence
\SNote{Being farre more in number & hauing the iuſt cauſe, yet had the
worſe, becauſe they truſted in their owne ſtreingth.}
in their ſtrength and number, ſette the armie in aray in the ſame place,
wherein they had fought before: \V yet ſo that they did firſt goe vp and
weepe before our Lord vntil night: and conſulted him, and ſaid: Shal I
procede any more to fight againſt the children of Beniamin my brethren,
or not? To whom he anſwered:
\SNote{God alſo puniſhed al Iſrael by this ciuil warre, for ſuffering
idolatrie in the tribe of Dan.
\XRef{cha.~18. v.~30.}
which they ought to haue puniſhed.
\XRef{Deut.~13. v.~12.}}
Goe vp to them, and enter battel. \V And when the children of Iſrael the
next day had proceded againſt the children of Beniamin to battel, \V the
children of Beniamin brake forth out of the gates of Gabaa: and meeting
them they raged with ſo great a ſlaughter againſt them, that they
ouerthrew eightene thouſand men that drew ſword. \V For the which thing
al the children of Iſrael came into the houſe of God, and ſitting wept
before our Lord: and they faſted that day vntil euening, and offered to
him holocauſtes, and pacifique victimes, \V and asked him concerning
their ſtate. At that time the arke of the couenant of our Lord was
there, \V and
\SNote{By this it appeareth that this hiſtorie happened not long after
the death of Eleazarus.
\XRef{Ioſ.~24. v.~33.}
to whom his ſonne Phinees ſucceded in the ſpiritual Supremacie of the
Church.}
Phinees the ſonne of
%%% o-0521
Eleazarus the ſonne of Aaron prouoſt of the houſe. They therfore
conſulted our Lord, and ſaid: Shal we goe forth any more to fight
againſt the children of Beniamin our brethren, or reſt? To whom our Lord
ſaid: Goe vp, for to morrow I wil deliuer them into your handes. \V And
the children of Iſrael ſette ambuſhmentes round about the citie of
Gabaa: \V and the third time,
%%% 0581
as once and twiſe, they brought forth their armie againſt Beniamin. \V
But the children of Beniamin alſo iſſued forth boldly out of the citie,
and purſewed a long way the aduerſaries fleeing, ſo that they wounded of
them, as the firſt day and the ſecond, and ſlew them turning their
backes by two wayes, wherof the one went into Bethel, and the other into
Gabaa, and ouerthrew about thirtie men: \V for they thought to kil them
after their accuſtomed maner. Who feyning artificially as though they
fled tooke aduiſe to draw them away from the citie, & as it were fleing
to bring them to the pathes aforeſayd. \V Therfore al
the children of Iſrael ryſing out of their ſeates, ſette their armie in
battel aray, in the place which is called Baalthamar. The ambuſhmentes
alſo, which were about the citie, began by litle and litle to open them
ſelues, \V and to proceede from the Weſt part of the citie. Yea and
other ten thouſand men of al Iſrael prouoked the inhabitantes of the
citie to skirmishes. And the battel grew ſore againſt the children of
Beniamin: and they vnderſtoode not that on euerie ſide deſtruction hong
ouer them. \V And our Lord ſtroke them in the ſight of the children of
Iſrael, and they ſlew of them in that day fiue and twentie thouſand, and
an hundred men, al warryers and that drew ſword. \V But the children of
Beniamin when they ſaw them ſelues to be inferiour, beganne to
flee. Which the children of Iſrael ſeing, gaue them place to flee, that
they might come to the ambuſhmentes prepared, which they had ſette neere
the citie. \V Who when they had ſodenly riſen out of their dennes, and
Beniamin turned their backes to the fleaers, they entred the citie, and
ſtroke it in the edge of the ſword. \V And the children of Iſrael had
geuen a ſigne to them, whom they had laid in the ambushementes, that
after they had taken the citie, they ſhould kindle a fire: that the
ſmoke aſcending on high, they might ſhew that the citie was taken. \V
Which when the children of Iſrael ſaw being in the verie fight (for the
children of Beniamin thought that they fled, and purſewed more
inſtantly, hauing ſlaine thirtie men of their armie,) \V and they ſaw as
it were a piller of ſmoke to riſe vp from the citie. Beniamin alſo
looking backe, when he ſaw the citie taken, and the flames caried on
high: \V they that before had feyned as if they fled, turning their face
reſiſted more manfully. Which when the children of Beniamin had ſene,
they were
%%% 0582
turned into flight, \V and beganne to goe the way of the deſert, the
aduerſaries purſewing them thither alſo. But they alſo that had fyred
the citie, mette them. \V And ſo it came to paſſe, that on both ſides
they were ſlaine of the enemies, neither was there any reſt of men
dying. They fel, and were ouerthrowen on the eaſt ſide of the citie of
Gabaa. \V And
%%% o-0522
there were that were ſlaine in the ſame place, eightene thouſand men, al
moſt valiant warryers. \V Which when they had ſeene, that were remayning
of Beniamin, they fled into the wildernes, and went on to the rocke, the
name wherof is Remmon. In that flight alſo ſtragling, and going diuers
waies, they ſlew fiue thouſand men. And whereas they went farder, they
purſewed them, and ſlew alſo other two thouſand. \V And ſo it came to
paſſe, that al which were ſlaine of Beniamin in diuerſe places, were
fiue and twentie thouſand one hundred fighting men, moſt prompt to
warres. \V There remayned therfore of al the number of Beniamin that
could eſcape, and flee into the wildernes, ſix hundred men: and they
abode in the Rocke Remmon foure monethes. \V But the children of Iſrael
retyring, ſtroke al the remaines of the citie with the ſword from men
euen to beaſtes, and al the cities and villages of Beniamin the
deuouring flame did conſume.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
