%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 0547
%%% o-0492
\startChapter[
  title={Chapter 6}
  ]

\Summary{The people falling againe to ſinne, are oppreſsed by the
  Madianites. 12.~An Angel appearing to Gedeon, ſendeth him to deliuer
  Iſrael, 17.~confirming his miſsion by miracle. 25.~So he firſt
  deſtroyeth Baals altar, 34.~then gathereth an armie againſt Idolaters,
  36.~and is aſsured againe of Gods protection by two miracles in a
  fleece of woolle.}

And the children of Iſrael did euil in the ſight of our Lord: who
deliuered them into the hand of Madian ſeuen yeares, \V & they were ſore
oppreſſed of them. And they made them ſelues dennes and caues in the
mountaines, and very wel fenſed places to reſiſt. \V And when Iſrael had
ſowen, Madian came vp and Amalec, and the reſt of the Eaſt nations: \V
and pitching their tentes by them waſted al thinges as they were in the
blade vnto the entring of Gaza: and they left nothing at al in Iſrael
that perteyned to mans life, not ſheepe, not oxen, not aſſes. \V For
they and al their flockes came with their tabernacles, and like vnto
locuſtes filled al places, an inumerable multitude of men, and of
camels, waſting whatſoeuer they touched. \V And Iſrael was ſore humbled
in the ſight of Madian. \V And he cried to our Lord deſiring helpe
againſt the Madianites. \V Who ſent vnto them
\SNote{S.~Auguſtin
\Cite{(q.~31. in Iudic.)}
ſuppoſeth that this meſſenger ſent from God called \Emph{a man & a prophete}
(for the forme, wherin he appeared) was the ſame Angel which ſate vnder
the oke and ſent Gedeon to deliuer Iſrael.
\XRef{v.~11.~12. &c.}}
a man that was a prophet, and he ſpake: Thus ſayeth our Lord the God of
Iſrael: I made you to come vp out of Ægypt, and brought you out of the
houſe of ſeruitude, \V and deliuered you out of the handes of the
Ægyptians, and of al the enemies, that afflicted you: and I caſt them
out at your entring, and deliuered you their Land. \V And I ſaid: I the
Lord your God, feare not the goddes of the Amorrheites, in whoſe land
you dwel. And you would not heare my voice. \V And an Angel of our Lord
came, and ſate vnder an oke, that was in Ephra, and perteyned to Ioas
the father of the familie of Ezri. And when Gedeon his ſonne did thresh
and purge wheate in a winepreſſe, to flee Madian, \V the Angel of our
Lord appeared to him, and ſaid: Our Lord be with thee ô moſt valiant of
men. \V And Gedeon ſaid to him: I beſech thee my Lord, if our Lord be
with vs, why haue theſe euils apprehended vs? where are his meruelous
workes, which our fathers haue told vs, and ſaid: Out of Ægypt did our
Lord bring vs? but now our Lord hath forſaken vs, and deliuered vs into
the hande of Madian. \V And our Lord looked
%%% 0548
toward him, and ſaid: Goe in this thy ſtrength, and thou shalt deliuer
Iſrael out of the hand of Madian: know that I haue ſent thee. \V Who
anſwering ſaid: I beſech thee, my Lord, wherein ſhal I deliuer Iſrael?
behold my familie is the meaneſt in Manaſſes, and I the leaſt in my
fathers houſe. \V And our Lord
%%% o-0493
ſaid to him: I wil be with thee: and thou ſhalt ſtrike Madian as it were one
man. \V And he ſaid: If I haue found, quoth he, grace before thee, geue
me a ſigne that it is thou which ſpeakeſt to me. \V Neither depart thou
hence, til I returne to thee, bringing a ſacrifice, and offering
\SNote{He meant not to offer ſacrifice to the Angel, but that either the
Angel, or himſelfe in preſence of the Angel, ſhould offer it to God and
ſo in dede the Angel partly directed him what to do, partly executed the
office himſelfe, by touching the oblation with his rodde; and
miraculouſly bringing fire to conſume the ſacrifice.}
to thee. Who anſwered: I wil tarie thy coming. \V Gedeon therfore went
in, and boyled a kidde, and of a buſhel of flowre baked vnleuened
loaues: and putting the flesh in a basket, and the broth of the flesh
into a potte, he caried al vnder the oke, and offered to him. \V To whom
the Angel of our Lord ſaid: Take the fleſh and the vnleuened loaues, and
put them vpon that rocke, and powre out the broth theron. And when he
had done, \V the Angel of our Lord ſtretched forth the tippe of the
rodde, which he held in his hand, and touched the flesh and the
vnleuened loaues: and there aroſe a fyre from the rocke, and conſumed
the flesh, and the
\Fix{leuened}{vnleuened}{obvious typo, fixed in other}
loaues: And the Angel of our Lord vaniſhed from his eies. \V And Gedeon
ſeeing that it was the Angel of our Lord, ſaid: Alas my Lord God: that I
haue ſeene the Angel
\Fix{our}{of our}{obvious typo, fixed in other}
Lord face to face. \V And our Lord ſaid to
him: Peace be with thee: feare not, thou ſhalt not die. \V And Gedeon
built there
\SNote{An altar for a monument, not for ſacrifice.}
an altar to our Lord, and called it, our Lords peace, vntil this preſent
day. And when he was yet in Ephra, which is of the familie of Ezri, \V
that night our Lord ſaid to him: Take a bullocke of thy fathers, and an
other bullocke of ſeuen yeares, and thou ſhalt deſtroy the altar of
Baal, which is thy fathers: and cut downe the groue, that is about the
altar: \V and thou ſhalt build an altar to the Lord thy God in the toppe
of this rocke, whereupon thou didſt lay the ſacrifice before: and thou
ſhalt take the ſecond bullocke, and ſhalt offer an holocauſte vpon a
pile of wood, which thou ſhalt cut downe out of the groue. \V Gedeon
therfore taking to him ten men of his ſeruantes, did as our Lord had
commanded him. But fearing his fathers houſe, and the men of that citie,
he would not doe it by day, but accompliſhed al thinges by night. \V And
when the men of that towne were
%%% 0549
riſen in the morning, they ſaw the altar of Baal deſtroyed, and the
groue cut downe, and the other bullocke laid vpon the altar, which then
was built. \V They ſaid one to an other: Who hath done this? And when
they inquired for the author of the fact, it was ſaid: Gedeon the ſonne
of Ioas did al theſe thinges. \V And they ſaid to Ioas: Bring forth thy
ſonne hither, that he may die: becauſe he hath deſtroyed the altar of
Baal, and hath cut downe his groue. \V To whom he anſwered: Why are you
reuengers of Baal, that you fight for him? he that is his aduerſarie,
let him die before to morrow light appeare: if he be God, reuenge he him
ſelf, on him that hath rayſed his altar. \V From that day Gedeon was
called
\SNote{The ſtrength of Baal, or ſtronger then Baal.}
Ierobaal, becauſe Ioas had ſaid: Let Baal reuenge him ſelf on him, that
hath rayſed his altar. \V Therfore al Madian, and Amalec, and the eaſt
peoples were gathered together, and paſſing ouer Iordan, camped in the
vaile Iezrael. \V But the ſpirit of our Lord reueſted Gedeon, who
ſounding with a trumpet called together the houſe
%%% o-0494
of Abiezer, to folow him. \V And he ſent meſſengers into al Manaſſes,
which it ſelfe alſo did folow him: and other meſſengers into Aſer and
Zabulon and Nepthali, which mette him. \V And Gedeon ſaid to God: If
thou ſaue Iſrael by my hand, as thou haſt ſpoken, \V I wil putte this
fleece of wool on the floore: if there ſhal be
\SNote{Dew firſt in the fleece and after on the ground ſignified grace
and true religion firſt in one people, after in al nations.
\Cite{Sainct Amb. Ser.~13. de Natal. Dom.}
\Cite{Venr. Beda qq. in Iudic. c.~4.}
Alſo Chriſts Incarnation without detriment of his mothers virginitie, of
whoſe grace al are repleniſhed.
\Cite{S.~Bernard ho.~2. in Miſſus eſt.}}
dew in the fleece onlie, and on al the ground drienes, I shal know that
by my hand, as thou haſt ſpoken, thou wilt deliuer Iſrael. \V And it
came ſo to paſſe. And ryſing in the night wringing the fleece, he filled
a veſſel with the dew. \V And he ſaid againe to God: Let not thy furie
be angry againſt me if I tempt once againe, ſeeking a ſigne in the
fleece. I deſire that the fleece onlie may be drie, and al the ground
wette with dew. \V And God did that night as he had requeſted: and there
was drienes in the fleece onlie, and dew on al the ground.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
