%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 0564
%%% o-0507
\startChapter[
  title={Chapter 13}
  ]

\Summary{The people fal againe to idolatrie and are afflicted by the
  Philiſthims. 3.~An Angel fortelleth Manue his wife, that she shal haue
  a ſonne, and that he shal be a Nazareite from his birth,
  11.~confirmeth the ſame to Manue. 16.~They offer ſacrifice to
  God. 24.~The childe is borne, called Samſon, and bleſſed of God.}

And againe the children of Iſrael did euil in the ſight of our Lord: who
deliuered them into the handes of the Philiſthimes fourtie yeares. \V
And there was a certaine man of Saraa, and of the ſtocke of Dan, named
Manue, hauing a wife barren. \V To whom an Angel of our Lord appeared,
and ſaid to her: Thou art barren and without children: but thou ſhalt
conceiue & beare a ſonne: \V beware therfore that
%%% 0565
thou
\SNote{Abſtinence not only from thinges vncleane by the law but alſo
from wine and ſider was a preparation to the childe, who ſhould abſtaine
from them al his life.}
drinke not wine & ſicer, nor eate any vncleane thing: \V becauſe thou
ſhalt conceiue and beare a ſonne, whoſe head the raſer ſhal not touch:
for he ſhal be a Nazareite of God,
\SNote{Other Nazarites obſerued a preſcripte rule of abſtinence for a
time only.
\XRef{Num.~6.}
But Samſon al his life; as a more perfect figure of Chriſt.}
from his infancie, and from his mothers wombe, and he ſhal beginne to
deliuer Iſrael from the handes of the Philiſtijmes. \V Who when ſhe was
come to her huſband, ſaid to him: A man of God came to me, hauing an
Angelical contenance, exceeding terrible. Whom when I had asked, who he
was, and whence he came, and by what name he was called, he would not
tel me: \V but this he anſwered: Behold thou shalt conceiue and beare a
ſonne: beware thou drinke not wine, nor ſicer, and that thou eate not
any vncleane thing: for the child shal be the Nazereite of God from his
infancie, and from his mothers wombe vntil the day of his death. \V
Manue therfore prayed to our Lord, and ſaid: I beſech thee ô Lord, that
the man of God, whom thou didſt ſend, may come againe, and teach vs what
we ought to doe concerning the child, that shal be borne. \V And our
Lord heard Manue praying, and the Angel of our Lord appeared againe to
his wife ſitting in the field. But Manue her huſband was not with
her. Who when ſhe had ſeene the Angel, \V haſtened, and ranne to her
huſband: and she told him, ſaying:
%%% o-0508
Behold
\LNote{The man hath appeared.}{Al
\MNote{Proteſtantes either contradict themſelues, or teach Arrianiſme.}
ancient fathers and Catholique writers ſay, this was an Angel, which
appeared in the forme of a man, and it is plaine by the text. Yet ſome
proteſtantes wil haue this perſon to be Chriſt, \Emph{the eternal vvord
of God}. VVho afterwardes became man.
\CNote{\Cite{Bible 1603.}}
And neuertheles where by & by
\XRef{(v.~16.)}
he admoniſheth Manue to offer ſacrifice to God, they note that
\Emph{he ſought not his ovvne honour but Gods, vvhoſe meſſenger he
vvas}: either plainly contradicting themſelues, or els teaching
Arrianiſme, as though the Sonne of God were not God; or inferiour to God
the Father.}
the man hath appeared to me, whom I ſaw before. \V Who roſe, and folowed
his wife: and comming to the man, ſaid to him: Art thou he that didſt
ſpeake to the woman? And he anſwered: I am. \V To whom, Manue, when,
ſayd he: Thy word shal be fulfilled, what wilt thou that the child doe?
or from what shal he keepe him ſelf? \V And the Angel of our Lord ſaid
to Manue: From al thinges, which I haue ſpoken to thy wife, let him
refraine him ſelf: \V and whatſoeuer groweth of the vineyard, let him
not eate: wine and ſicer let him not drinke, let him not eate any
vncleane thing: and whatſoeuer I haue commanded her, let him fulfil and
keepe. \V And Manue ſaid to the Angel of our Lord: I beſech thee that
thou condeſcend to my petitions, and let vs
\SNote{Manue taking the Angel for a holie prophete iuſtly thought he
would not admitte, nor cõmand anie thing but that was lawful. And ſo did
as the Angel appointed him, though he was no prieſt, nor the place
proper to ſacrifice, but by extraordinarie diſpenſation.}
make to thee a kidde of goates. \V To whom the Angel anſwered: If thou
conſtraine me, I wil not eate thy breade: but if thou wilt make
holocauſt, offer it to our Lord. And Manue knew not that it was an
Angel of our Lord. \V And he ſaid to him: What is thy name, that, if thy
word shal be fulfilled, we may honour thee? \V To whom he anſwered: Why
askeſt thou
%%% 0566
my name, which is merueilous? \V Manue therfore tooke a kidde of the
goates, and the libamentes, and put them vpon a rocke, offering to our
Lord, who doeth meruelous thinges: and he and his wife looked on. \V And
when the flame of the altar aſcended into heauen, the Angel of our Lord
aſcended together in the flame. Which when Manue and his wife had ſeene, they fel
flatte on the ground, \V and the Angel of our Lord appeared to them no
more. And forthwith Manue vnderſtood that it was an Angel of our
Lord, \V and he ſaid to his wife: Dying we ſhal die, becauſe we haue
ſeene
\SNote{Though Manue ſaw not God in his owne perſon, yet ſeing him in his
meſſenger feared death.
\Cite{S.~Aug. q.~54. in Iudic.}}
God. \V To whom his wife anſwered: If our Lord would haue killed vs, he
would not haue taken of our handes holocauſtes and libamentes, neither
would he haue shewed vs al theſe thinges, nor haue told vs theſe thinges
that are to come. \V She therfore bare a ſonne, and called his name
Samſon. And the child grewe, and our Lord bleſſed him. \V And the Spirit
of our Lord beganne to be with him in the campe of Dan betwixt Saraa and
Eſthaol.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
