%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-18


\project douay-rheims


%%% 0574
%%% o-0515
\startChapter[
  title={Chapter 18}
  ]

\Summary{Firſt ſending ſpies to diſcouer, 11.~ſix hundred armed men of
  the tribe of Dan goe to ſeeke poſſeſsions. 14.~By the way they take
  the idol, and idolatrical prieſt from Michas, 27.~ſurpriſe the towne
  of Lais, 30.~and there ſet vp idolatrie.}

In thoſe dayes there was not a king in Iſrael, and the tribe of Dan
ſought poſſeſſion for it ſelfe, that it might dwel therein: for vntil
that day it had
\SNote{Their whole portiõ was aſſigned
\XRef{(Ios.~19.)}
but through their owne ſlouth they poſſeſſed litle of it, ſo that
hitherto the greateſt part was not receiued.}
not receiued a lotte among the other tribes. \V Therfore the children of
Dan ſent fiue men of their ſtocke and familie moſt valiant from Saraa
and Eſthaol, that they might view the land, and diligently behold it,
and they ſaid to them: Goe, and conſider the land. Who going forward
when they were come into mount Ephraim, and had entered into the houſe
of Michas, they reſted there: \V and knowing the voice of the yong man
the Leuite, and vſing his lodging, they ſaid to him: Who brought thee
hither? What doeſt thou here? For what cauſe wouldeſt thou come
hither? \V Who anſwered them: Theſe, and theſe thinges hath Michas done
to me, & hath hyred me for wages to be his prieſt. \V And
%%% o-0516
they deſired him that he would conſult
\SNote{They ment the falſe god which the apoſtata Leuite ſerued.}
the Lord, that they might know whether they should goe on a proſperous
iourney, and the thing should haue effect. \V
\SNote{The diuel anſwered, as his maner is obſcurly, ſome times truly &
ſometimes falſly.}
Who anſwered them: Goe in peace: The Lord regardeth your way, and the
iourney that you goe. \V The fiue men therfore going came to Lais, &
they ſaw the people dwelling in it without any feare, according to the
cuſtome of the Sidonians, ſecure and quiet, no man at al reſiſting them,
& of greate riches, and ſeparated farre from Sidon and from al men. \V
And returning to their brethren in Saraa and Eſtaol, and asking what
they had done they anſwered them: \V Ariſe, and let vs goe vp to them:
for we haue ſeene the Land exceding rich and plentiful: neglect not,
ſlacke not: let vs goe, and poſſeſſe it, it wil be no labour.
%%% 0575
\V We shal enter vnto them being ſecure, into a moſt large countrie, and
our Lord wil deliuer to vs the place, wherein is penurie of nothing, of
thoſe thinges that grow on the earth. \V There departed therfore from
the kinred of Dan, that is to ſay, from Saraa and Eſthaol ſix hundred
men, furnished with warlike armour, \V and going vp they taried in
Cariathiarim of Iuda: which place from that time tooke the name of the
Tentes of Dan, and it is at the backe of Cariathiarim. \V Thence they
paſſed into mount Ephraim. And when they were come to the houſe of
Michas, \V the fiue men, that before had been ſent to view the Land of
Lais, ſaid to the reſt of their brethren: You know that in theſe houſes
there is an Ephod, and Theraphim, and a grauen, and molten god: See what
pleaſeth you. \V And when they had turned a litle aſide, they entred
into the houſe of the yong man the Leuite, which was in the houſe of
Michas: and ſaluted him with peacable wordes. \V And the ſix hundred men
ſo as they were armed, ſtoode before the doore. \V But they, that were
entred the houſe of the yong man, endeuoured to take away the grauen,
the Ephod and the idols, and molten god, and the
prieſt ſtoode before the doore, the ſix hundred moſt valiant men
expecting not farre of. \V They therfore that were entred tooke the
grauen, the Ephod, the Theraphim, and molten god. To whom the prieſt
ſaid: What doe you? \V To whom they anſwered: Hold thy peace, and put
thy finger vpon thy mouth and come with vs, that we may haue thee for a
father, and a prieſt. Whether is better for thee, that thou be a prieſt
in the houſe of one man, or in one tribe and familie in Iſrael? \V Which
when he had heard, he agreed to their wordes, and tooke the Ephod, and
idols, and grauen god, and departed with them. \V Who when they went
forward, and had made the children and the cattel to goe before them,
and al that was
\Fix{percious,}{precious,}{obvious typo, fixed in other}
\V and were now farre from the houſe of Michas, the men that dwelt in
the houſe of Michas crying out together folowed, \V and at their backe
began to shoute. Who looking backe, ſaid to Michas: What meaneſt thou?
Why doeſt thou crie? \V Who anſwered: My goddes, which I made me, you
haue taken away, and the prieſt, and al that I haue, and doe you ſay:
What aileth thee? \V And the children of Dan ſaid to him: Beware thou
ſpeake no more vnto vs, and there come vnto
%%% o-0517
thee men prouoked in mind, and thou with al thy houſe
%%% 0576
perish. \V And ſo they went on their iourney begone. But Michas ſeing
that they were ſtronger then he, returned into his houſe. \V And the ſix
hundred men tooke the prieſt, and the thinges which we ſpake of before,
and came into Lais to a people that was quiet and ſecure, and ſtroke
them in the edge of the ſword: and the citie they deliuered to fyre, \V
no man at al bringing them ſuccour, for that they dwelt farre from
Sidon, and had with no men anie ſocietie and affayres. And the citie was
ſituated in the countrie of Rohob: which building agayne they dwelt in
it, \V calling the name of the citie Dan, according to the name of their
father, whom Iſrael had begotten, which before was called Lais. \V And
they ſette vp to them ſelues the
\SNote{\HH{peſel}, \GG{eidolon}, \L{ſculptile}, \Emph{the grauen thing},
falſly called god.
\XRef{c.~17. v.~5.}}
grauen idol, and Ionathan the ſonne of Gerſon the ſonne of Moyſes, and
his ſonnes prieſtes in the tribe of Dan, vntil the day of their
captiuitie. \V And the idol of Michas remayned with them al the time,
that the houſe of God was in Silo. In thoſe daies there was not a king
in Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
