%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 0570
%%% o-0512
\startChapter[
  title={Chapter 16}
  ]

\Summary{Samſon enuironed in a citie taketh away the gates, and carieth
  them on his shoulders into a mountaine. 4.~Is at laſt deceiued by
  Dalila, 21.~his eyes put out, and ſcornfully abuſed. 26.~But finally
  God reſtoring his ſtrength, he ſtriking two pillers the houſe falleth,
  and with himſelf, killeth three thouſand Philiſthiims.}

He went alſo into Gaza, and ſaw there a woman that was
\TNote{or, an Inkeeper.}
an harlot, and went in vnto her. \V Which when the Philiſthiims had
heard, and it was bruted among them, that Samſon was entered into the
citie, they compaſſed him, keepers being ſette in the gate of the citie:
and there al the night wayting with ſilence, that in the morning they
might kil him going out. \V But Samſon ſlept vntil
\Fix{midnigt,}{midnight,}{obvious typo, fixed in other}
and then ariſing he
\SNote{For ſuch admirable ſtreingth the hethniſh people thought Samſon
to be Hercules.
\Cite{S.~Aug. li.~18. c.~19. ciuit.}
But he was in deede farre ſtronger then they feaned of Hercules, who
they ſaid, was not able to fight againſt two: whereas Samſon alone killed
a thouſand with the iaw bone of an aſſe.
\XRef{c.~15. v.~15.}}
tooke both the leaues of the gate, with their poſtes and locke, and
laying them on his shoulders, caried them to the toppe of the mountaine,
which looketh toward Hebron. \V After theſe thinges he loued a woman,
which dwelt in Valley Sorec, and she was called Dalila. \V And the
princes of the Philiſthiims came to her, and ſaid: Deceiue him, and
learne of him, wherein he hath ſo great ſtrength, and how we may be able
to ouercome him, and being bound to afflict him. Which if thou shalt
doe, we wil geue thee euerie one a thouſand and an hundred peeces of
ſiluer. \V Dalila therfore ſpake to Samſon: Tel me, I beſech thee,
wherein thy greateſt ſtrength is, and what it is wherewith being bound
thou canſt not breake forth. \V To whom Samſon anſwered: If I shal be
bound with ſeuen cordes of ſinewes not yet drie, and moyſte as yet, I
shal be weake as other men. \V And the princes of the Philiſthimes
brought
%%% 0571
vnto her ſeuen cordes, as he had ſaid: with the which she bound him, \V
ambushementes lying ſecretely in wayte neere her, and in the chamber
expecting the end of the thing, and she cried to him: The Philiſthijms
vpon thee Samſon. Who brake the bandes, as if a man should breake a
thread of toe twyned with ſpittle, when it hath taken the fauor of fyre:
and it was not knowen wherein his ſtrength was. \V And Dalila ſaid to
him: Behold thou haſt deluded me, and haſt ſpoken falſe: not at the
leaſt tel me wherwith thou mayeſt be bound. \V To whom he anſwered: If I
shal be bound with new cordes, that were neuer occupied, I shal be
weake, and like to other men. \V With the which Dalila againe bound him,
and cried: The Philiſthijmes vpon thee Samſon, ambuſhementes being
prepared in the chamber. Who did ſo breake the bandes as threades of
linnen cloth. \V And Dalila ſaid to
%%% o-0513
him againe: How long deceiueſt thou
me, and ſpeakeſt falſe? Shew wherewith thou mayeſt be bound. To whom
Samſon anſwered: If thou platte ſeuen heares of my head with a heare
laſe, and faſten a nayle tyed round about them in the ground, I shal be
weake. \V Which when Dalila had done, she ſaid to him: The Philiſthijms
vpon thee Samſon. Who ryſing vp from ſleepe drew out the nayle with the
heares and the heare laſe. \V And Dalila ſaid to him: How doeſt thou ſay
that thou loueſt me, whereas thy mind is not with me? Theſe three times
thou haſt lied to me, & wouldeſt not tel wherein thy greateſt ſtrength
is. \V And when she moleſted him, and continually hong vpon him for many
daies, not geuing him ſpace to reſt, his ſoule faynted, and was wearied
euen vnto death. \V Then opening the truth of the thing, he ſaid to her:
There neuer came yron vpon my head, becauſe I am a Nazareite, that is to
ſay, conſecrated to God from my mothers wombe: if my head shal be ſhauen,
\SNote{Supernatural ſtreingth, or grace departeth when any leaue the
rule of their profeſſion.}
my ſtrength shal depart from me, and I shal fayle, and shal be as other
men. \V And she ſeing that he had confeſſed to her al his minde, ſent
to the princes of the Philiſthijms and willed them: Come vp yet once
more, for now he hath opened his hart to me. Who went vp taking with
them the money which they had promiſed. \V But she made him to ſleepe
vpon her knees, and to lay his head in her boſome. And she called a
barber, and shaued his ſeuen heares, and beganne to driue him away, and
thruſt him from her: for immediately the
\Fix{ſtrungth}{ſtrength}{obvious typo, fixed in other}
departed
%%% 0572
from him: \V and she ſaid: The Philiſthijms vpon thee Samſon. Who
aryſing from ſleepe, ſaid in his mind: I wil goe forth as I did before,
and wil ſhake my ſelf, not knowing that our Lord was departed from
him. \V Whom when the Philiſthijmes had apprehended, forthwith they
plucked forth his eies, and led him to Gaza bound with chaynes, and
being shut vp in priſon they made him grinde. \V And now his heares had
begone to grow againe, \V and the princes of the Philiſthijms aſſembled
in one, that they might immolate magnifical hoſtes to Dagon their god,
and might feaſt, ſaying: Our god hath deliuered our enemie Samſon into
our handes. \V Which thing the people alſo ſeing, prayſed their god, and
ſaid the ſame thinges: Our god hath deliuered our aduerſarie into our
handes, who deſtroyed our countrie, and killed verie manie. \V And
reioyſing through out their bankettes, when they had now taken their
good cheere, they commanded that Samſon should be called, and should
play before them. Who being brought out of priſon played before them,
and they made him to ſtand betwen two pillers. \V Who ſaid to the
ſeruant that gouerned his ſteppes: Suffer me to touch the pillers, on
which al the houſe ſtayeth, and let me leane vpon them, and reſt a
litle. \V And the houſe was ful of men and wemen, and there were al the
princes of the Philiſthijms, alſo from the roofe and higher part, about
three thouſand of both ſexe beholding Samſon playing. \V But he
inuocating our Lord, ſaid: Lord God remember me, & reſtore now to me
myne old
%%% o-0514
ſtrength my God, that I may
\SNote{He deſired to be reuenged, not of rancour of mind but of zele of
iuſtice. And ſo al the elect & glorified Sainctes deſire reuẽge.
\XRef{Luc.~18. v.~8.}
\XRef{Apoc.~6. v.~10.}}
reuenge me of myne enemies, and for the loſſe of two eies may receiue
one reuenge. \V And taking both the pillers, on which the houſe reſted,
and holding the one in his right hand, and the other in his left, \V he
ſaid:
\LNote{Let me die with the Philiſthiims.}{Manie
\MNote{Samſon excuſed in killing himſelf with his enemies.}
thinges do iuſtifie Samſons fact in killing himſelf with the
Philiſthiims. Firſt it appeareth by the miracle, that God directly and
extraordinarily concurred by reſtoring in that moment his admirable
ſtreingth, that he could pull downe two ſuch pillers. And conformably we
may gather, that God inſpired his mind to attempt this fact, and ſo he
erred not, but obeyed God herein: as S.~Auguſtin noteth.
\Cite{li.~1. c.~21. &~26. de ciuit Dei.}
Secondly he was moued with zele of Gods honour, hearing the Idolaters
praiſe their falſe god Dagon. Thirdly, he had a good and pure intention
to reuenge him ſelfe for Gods more glorie, praying to him for
reſtauration of ſtreingth. Fourthly he did not directly deſire to kil
himſelfe, but to kil the Philiſthims, though himſelfe muſt alſo die with
them.
\MNote{Samſon a figure of Chriſt.}
And in this act eſpecially he was a figure of Chriſt, who chiefly by his
death conquered his enemies.}
Let me die with the Philiſthijms. And the pillers being ſtrongly shaken,
the houſe fel vpon al the princes, and the reſt of the multitude, that
was there: and he killed manie moe dying, then before he had killed
liuing. \V And his brethren going downe and al his kinred, they tooke
his bodie, and buried it betwixt Saraa and Eſthaol in the ſepulchre of
his father Manue, and he iudged Iſrael twentie yeares.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
