%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judges
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-17


\project douay-rheims


%%% 0573
%%% o-0514
\startChapter[
  title={Chapter 17}
  ]

\Summary{Michas
\MNote{The third part.

Of certaine accidentes which happened in the time of the Iudges.}
and his mother cauſe a grauen, and a molten idol to be made of
ſiluer. 5.~He maketh one of his ſonnes prieſt for the idol, 10.~and for
the ſame purpoſe hyreth alſo a Leuite.}

There was at that time a certaine man of mount Ephraim named Michas, \V
who ſaid to his mother: The thouſand and hundred ſiluer peeces, which
thou hadſt ſeparated to thy ſelfe, and concerning the which thou didſt
ſweare in my hearing, behold I haue, and they are with me. To whom she
ſaid: Bleſſed be my ſonne to the Lord. \V He therfore rendred them to
his mother, who had ſaid to him: I haue conſecrated and vowed this
ſiluer to the Lord, that my ſonne may receiue it of my hand, and make
\SNote{In hebrew \HH{peſel vmaſſecah}, in Latin \L{ſculptile &
conflatile}, \Emph{a grauen & molten thing}, an image or forme made in
mettle for a god, and ſo called,
\XRef{v.~5.}
was in dede an idol of Gentilitie, and nothing at al againſt ſacred Images
of Chriſt and his Sainctes in the Catholique Church. Wherof more is
noted,
\XRef{Gen.~31.}
\XRef{Exo.~20.}}
a grauen and
%%% !!! Same SNote as above.
%%% \SNote{}
a molten (god) and now I deliuer that to thee. \V He rendred them
therfore to his mother: who tooke
%%% o-0515
two hundred ſiluer peeces and gaue them to the ſiluerſmith, that he
might make of them a grauen and a molten (god) which was in the houſe of
Michas. \V Who ſeparated alſo therein a litle houſe to the god, and made
an Ephod, and Theraphim, that is to ſay, a prieſtlie veſtiment, and
idoles: and he
\SNote{Annointed his hãdes with oyle, as was preſcribed.
\XRef{Exo.~29.}
\XRef{Leu.~8.}
But ſuch an apiſh imitation was of no value, where was neither true
vocation in the annointed (for he deſcended not of Aaron, but of Moyſes,
\XRef{chap.~18. v.~30.)}
nor authoritie in him that vſed this ceremony.}
filled the hand of one of his ſonnes, and he became his prieſt. \V In
thoſe daies there was not a king in Iſrael, but euerie one did that,
which ſemed right to him ſelf. \V There was alſo an other yong man of
\Fix{Bethelem}{Bethlehem}{obvious typo, fixed in other}
Iuda, of the kinred therof: and he was a Leuite, and dwelt there. \V And
going forth out of the citie of Bethelehem, he would ſeiourne
whereſoeuer he should find it comodious for him. And when he was come
into mount Ephraim, making his iourney, and had turned aſide a litle
into the houſe of Michas, \V he was demanded of him whence he came. Who
anſwered: I am a Leuite of Bethlehem Iuda, and I goe to dwel where I
shal be able, and shal perceiue it to be profitable for me. \V And
%%% 0574
Michas ſaid: Abide with me, and be to me a father and a prieſt, and I
wil geue thee euerie yeare tenne ſiluer peeces, and duble liuerie, and
the thinges that be neceſſarie for victual. \V He was content, and abode
with the man, and was vnto him as one of his ſonnes. \V And Michas
filled his hand, and had the yong man for a prieſt with him, ſaying: \V
Now I know that God wil do me good
\SNote{An Apoſtata Leuite was accounted more ſufficient then an
idolatrical prieſt to ſerue an idol. So he that is a Prieſt or a Deacon
once catholikly conſecrated, is a ſufficient (yea too ſufficient) a
miniſter with proteſtantes.}
hauing a prieſt of the Leuitical kinde.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
