%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 2320
%%% o-2134
\startChapter[
  title={Chapter 13}
  ]

\Summary{Speaking in parables (as the Scripture foretold of him, and as
  meet was for the reprobate Iewes:) he showeth by the parable of the
  Sower, that in the labours of his Church, three parts of foure do
  perish through the fault of the
\Fix{heares.}{hearers.}{obvious typo, fixed in other}
24.~and yet, by the parable of good ſeed and cockle (as alſo of the Net)
that  his ſeruants muſt not for al that, neuer while the world laſteth,
make any Schiſme or Separation. 31.~And by parables of the litle Muſtard
ſeed and Leauen, that notwithſtanding the three parts perishing, and
ouerſowing of cockle, yet that fourth part of the good ſeed shal ſpread
ouer al the world. 44.~And withal, what a treaſure, and pearle it
is. 53.~After al which, yet his owne countrie wil not honour him.}

The ſame day \Sc{Iesvs} going out of the houſe, ſate by the ſea ſide. \V
And
\CNote{\XRef{Mr.~4,~1.}
\XRef{Lu.~8,~4.}}
great multitudes were gathered togeather vnto him, in ſo much that
he went vp into a boat & ſate: and al the multitude ſtood in the ſhore;
\V and he ſpake to them many things in parables, ſaying:

Behold the Sower went forth to ſow. \V And whiles he ſoweth, ſome fel by
the way ſide, and the foules of the aire did come and eate it. \V Other
ſome alſo fel vpon rockie places, where they had not much earth: and
they ſhot vp incontinent, becauſe they had not deepnes of earth, \V and
after the ſunne was vp, they parched: and becauſe they had not roote,
they withered. \V And other fel among thornes: and the thornes grew and
choked them. \V And other ſome fel vpon good ground: and they yealded
fruit, the
\LNote{one an hundred}{This
\MNote{Difference of merits and rewards.}
differẽce of fruits is the difference of merits in this life, and
rewards for them in the next life, according to the diuerſities of
ſtates, or other differences. Of ſtates, as that the hundred-fold
agreeth to virgins profeſſed, threeſcore-fold to religious widowes,
thirtie-fold to the maried. 
\Cite{Aug. li. de Virginis. c.~44. & ſeq.}
which truth the old Heretike Iouinian denied (as ours doe at this day)
affirming that there is no difference of merits or rewards
\Cite{Hier. li.~2. adu. Iouin.}
\Cite{Ambros. ep.~82.}
\Cite{Aug. har.~82.}}
one an hundredfold, the other threeſcore, and an other thirtie. \V He
that hath eares to heare, let him heare.

\V And his Diſciples came and ſaid to him: Why ſpeakeſt thou to them in
parables? \V Who anſwered and ſaid vnto them: Becauſe
\LNote{To you it is giuen}{To the Apoſtles and ſuch as haue the guiding
  and teaching of others, deeper knowledge of Gods Word and myſteries is
giuen, then to the common People. As alſo to Chriſtiãs generally, that
which was not giuen to the obſtinate Iewes.}
to you it is giuen to know the myſteries of the Kingdom of Heauen: but
to them it is not giuen. \V For 
%%% o-2135
he that hath, to him ſhal be giuen, and
he ſhal abound: but he that hath not, from him shal be taken away that
alſo which he hath. \V Therfore in parables I ſpeake to them: becauſe
ſeeing they ſee not, and
\SNote{When Gods word is preached, they properly haue eares to heare,
  that haue hartes to obey: & they hearing do not heare, which heare by
  ſẽſe of their body, & obey not by conſent of their harts,
\Cite{Auguſt. de dona perſeu. c.~14.}}
hearing they heare not, neither do they vnderſtand; \V and the Prophecie
of Eſay is fulfilled in them, which ſaith: \Emph{With hearing shal you
  heare; and you shal not vnderſtand: and ſeeing shal you ſee, and you
  shal not ſee.} \V \Emph{For the hart of this people is waxed groſſe,
  and with their eares they haue heauily heard, and their eyes}
\LNote{They haue shut}{In ſaying that they ſhut their owne eies,
which S.~Paul alſo repeateth
\XRef{Act.~28.}
he teacheth vs the true vnderſtãding of al other places, where it might
ſeeme by the bare words that God is the very authour and worker of this
induration, & blindnes, and of other ſinnes:
\CNote{\Cite{Iren. apud. Euſeb. li.~5. c.~19.}}
which was an old condemned blaſphemie, & is now the Hereſie of
\CNote{\Cite{Calu.~l.~2. Inſtit. c.~4.}}
Caluin:
\MNote{God is not the authour of euil.}
whereas our Sauiour here teacheth vs, that they ſhut their owne eies,
and are the cauſe of their owne ſinne and damnation; God not doing, but
permitting it, and ſuffering them to fal further becauſe of their former
ſinnes, as S.~Paul declareth of the reprobate Gentiles. 
\XRef{Ro.~1.}}
\Emph{they haue shut: leſt any time they may ſee with their eyes, and
  heare with their eares, and vnderſtand with their hart and be
  conuerted, and I may heale them.} \V But bleſſed are your eyes becauſe
they doe ſee, and
%%% 2321
your eares becauſe they do heare. \V For, Amen I ſay to you, that
\CNote{\XRef{Luc.~10,~23.}}
many
Prophets and iuſt men haue deſired to ſee the things that you ſee, and
haue not ſeen them: and to heare the things that you heare, and haue not
heard them. \V Heare you therfore the parable of the Sower.

\V Euery one that heareth the Word of the Kingdom and vnderſtandeth not,
there cometh the wicked one, and catcheth away that which was ſowen in
his hart: this is he that was ſowen by the way ſide. \V And he that was
ſowen vpõ rockie places: this is he that heareth the Word, and
incontinent receaueth it with ioy, \V yet hath he not root in him ſelf,
but is for a time: and when there falleth tribulation and perſecution
for the Word, he is by and by ſcandalized. \V And he that was ſowen
among thornes, this is he that heareth the Word, and the carefulnes of
this world and the deceitfulnes of riches choketh vp the Word, and he
become fruitles. \V But he that was ſowen vpon good ground: this is he
that heareth the Word, and vnderſtandeth, and bringeth fruit, and
yealdeth ſome an hundred-fold, and other threeſcore, and an other
thirtie.

\V An other parable he propoſed to them, ſaying: The Kingdom of Heauen is
reſembled to a man that ſowed good ſeed in his field. \V But when men
were aſleep, his enemy came and
\LNote{Ouerſowed}{Firſt by Chriſt and his Apoſtles was planted the
  truth, and falſhood came afterward, and was ouerſowed by the enemy the
  Diuel, and not by Chriſt, who is not the authour of euil.
\Cite{Tertul. de praſcrips.}}
ouerſowed cockle amõg the wheat, and went his way. \V And when the blade
was ſhot vp, and had brought forth fruit, then appeared alſo the
cockle. \V And the ſeruants of the Goodman of the houſe comming ſaid to
him: Sir, didſt thou not ſow good ſeed in thy field? whence then hath it
cockle? \V And he ſaid to them: The Enemie-mã hath done this. And the
ſeruãts ſaid to him: Wilt thou we goe and gather it vp? \V And he ſaid:
Noe 
\LNote{Leſt perhaps}{The good muſt tolerate the euil, when it is ſo
  ſtrong that it can not be redreſſed without danger and diſturbance of
  the whole Church; and commit the matter to Gods iudgement in the later
day. Otherwiſe where il men (be they Heretikes or other malefactours)
may be puniſhed or ſuppreſſed without diſturbance and hazard of the
good, they may and ought by publike authority either Spiritual or
temporal to be chaſtiſed or executed.}
leſt perhaps
%%% o-2136
gathering vp the cockle, you may root vp the wheat alſo togeather with
it. \V 
\LNote{Suffer both to grow}{The 
\MNote{Good and euil in the Church.}
good and bad (wee ſee here) are mingled
  togeather in the Church. Which maketh againſt certaine Heretikes and
  Schiſmatikes, which ſeuered themſelues of old from the reſt of the
  whole world, vnder pretence that themſelues only were pure, and al
  others, both Prieſts and People ſinners: and againſt ſome Heretikes of
this time alſo, which ſay that euil men are not of, or in the Church.}
Suffer both to grow vntil the harueſt, and in the time of harueſt
I wil ſay to the reapers: Gather vp firſt the cockle, and bind it into
bundels to burne, but the wheat gather ye into my barne.

\V An other parable he propoſed vnto them, ſaying: 
\CNote{\XRef{Mar.~4,~30.}
\XRef{Luc.~13,~18.}}
The Kingdom of Heauen is
like to a Muſtardſeed, which a man tooke and ſowed in his field. \V
Which is the
\LNote{The leaſt of al ſeeds}{The Church of Chriſt had a ſmal
  beginning, but afterward became the moſt glorious and known
  Common-welth in earth: the greateſt powers and the moſt wiſe of the
  world putting themſelues into the ſame.}
leaſt ſurely of al ſeeds: but when it is growen, it is greater then al
herbs, and is made a tree, ſo that the foules of the aire come, and dwel
in the branches thereof. \V An other parable he ſpake to them: The Kingdom
of Heauen is like to Leauen, which a woman tooke and hid in three
meaſures of meale, vntil the whole was leauened.

\V Al theſe things \Sc{Iesvs} ſpake in parables to the multitudes, and
without parables he did not ſpeake to them: \V that it might be
fulfilled which was ſpoken by the Prophet ſaying:
\CNote{\XRef{Pſa.~77,~2}}
\Emph{I wil open my mouth in parables, I wil vtter things hidden from
the fundation of the world.}

\V Then hauing dimiſſed the multitudes, he came into the houſe, and his
Diſciples came vnto him, ſaying: Expound vs the parable of the cockle of
the field. \V Who made anſwer & ſaid to them: He that ſoweth the good
ſeed, is the Sonne of man. \V And the field is the world. 
\Fix{}{And the good ſeede: theſe are the children of the kingdom.}{Missing text, present in other}% to prevent patagraph break
And the cockle: are the children of the wicked one. \V And the enemie
that 
%%% 2322
ſowed them, is
\SNote{Not God then, but the Diuel is the authour of al euil.}
the Diuel. But the harueſt, is the end of the world. And the reapers,
are the Angels. \V Euen as cockle therfore is gathered vp, and burnt with fire:
ſo ſhal it be in the end of the world. \V The Sonne of man ſhal ſend his
Angels, and they ſhal gather out of his Kingdom al ſcandals, and them
that worke iniquitie: \V and ſhal caſt them into the furnace of fire:
There ſhal be weeping & gnaſhing of teeth. \V Then ſhal the iuſt ſhine
as the ſunne, in the Kingdom of their Father. He that hath eares to
heare, let him heare.

\V The Kingdom of Heauen is like a treaſure hidden in a field. Which a
man hauing found, did hide it, and for ioy thereof goeth, and ſelleth al
that he hath, and buyeth that field. \V Againe the Kingdom of Heauen is
like to a marchant man, ſeeking good pearles. \V And hauing found one
precious pearle, he went his way, and ſold al that he had, and 
%%% o-2137
bought it.

\V Againe the Kingdom of Heauen is like to a net caſt into the ſea, and
gathering togeather of al kind of fiſhes. \V Which, when it was filled,
drawing it forth, and ſitting by the ſhore, they choſe out of the
\SNote{Here alſo are ſignified good and bad in the Church.}
good into veſſels, but the bad they did caſt out. \V So ſhal it be in
the conſummation of the world. The Angels ſhal goe forth, and ſhal
ſeparate the euil from among the iuſt. \V And ſhal caſt them into the furnace
of fire; there ſhal be weeping and gnaſhing of teeth. \V Haue ye
vnderſtood al theſe things? They ſay to him, Yea. \V He ſaid vnto them:
Therfore euery Scribe inſtructed in the Kingdom of Heauen, is like to a
man that is an houſholder, which bringeth forth out of his treaſure new
things and old.

\V And it came to paſſe: when \Sc{Iesvs} had ended theſe parables, he
paſſed from thence. \V And
\CNote{\XRef{Mr.~6,~1.}
\XRef{Luc.~4,~16.}}
coming into his owne countrie, he taught them
in their Synagogues, ſo that they marueled, and ſaid: How came this
fellow by this wiſedom and vertues? \V Is not this the
\LNote{Carpenters Sonne}{Herevpon Iulian the Apoſtate and his flatterer
Libanius tooke their ſcoffe againſt our Sauiour, ſaying (at his going
againſt the Perſians) to the Chriſtians, what doth the Carpenters Sonne
now? and threatning that after his returne, the Carpenters Sonne ſhould
not be able to ſaue them from his furie. Wherevnto a godly man anſwered,
by the Spirit of Prophecie, \Emph{He whom Iulian calleth the Carpenters
  Sonne, is making a wodden coffin for him againſt his death.} And
indeed not long after, there came newes, that in that bataile he dyed
miſerably 
\Cite{Sozo. lib.~6. c.~2.}
\Cite{Theodo. li.~3. c.~18.}
The very like ſcoffe vſe Heretikes that cal the body of Chriſt in the
B.~Sacrament, bakers bread. It ſeemeth indeed to the ſenſes to be ſo, as
Chriſt ſeemed to be Ioſephs natural Sonne, but faith telleth vs the
contrarie, as wel in the one, as in the other.}
carpenters Sonne? Is not his mother called \Sc{Marie}, and his brethren
Iames, and Ioſeph, and Simon, and Iude: \V and his ſiſters, are they not
al with vs? whence therfore hath he al theſe things? \V And they were
ſcandalized in him. But \Sc{Iesvs} ſaid to them: There is not a Prophet
without honour but in his owne countrie, and in his owne houſe. \V And
he wrought not many miracles there becauſe of their incredulity.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
