%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 27
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-27


\project douay-rheims


%%% 2361
%%% o-2180
\startChapter[
  title={Chapter 27}
  ]

\Summary{The cheefe of the Iewes accuſe him to Pilat the Gentil (his
  betrayer, and the Iudge, and the Iudges wife, teſtifying in the meane
  time manifoldly his innocencie:) 20.~and perſwade the common People
  alſo, not only to preferre the murderer Barabbas, but alſo to
  crie, \Sc{Crvcifige}: (Al, to the reprobation of their whole Nation,
  and nothing but fulfilling the Scriptures) 27.~After many illuſions,
  31.~he is crucified by the Gentils. 38.~Which the Iewes ſeeing, doe
  triumph as if they had now the victorie. 45.~But euen then by many
  wonderful works he declareth his might, to their
  confuſion. 57.~Finally being buried, they to make al ſure, ſet
  ſouldiars to keepe his ſepulcher.}

And
\CNote{\XRef{Mr.~15,~1.}
\XRef{Lu.~23,~1.}
\XRef{Io.~18,~28.}}
\MNote{\Sc{Good} friday}
when morning was come, al the cheefe Prieſts and Ancients of the People
conſulted togeather againſt \Sc{Iesvs}, that they might put him to
death. \V And they brought him bound and deliuered him to Ponce Pilate the
Preſident.

\V Then Iudas that betrayed him, ſeeing that he was condemned
\LNote{Repenting him}{Note
\MNote{Horrour of conſcience.}
how ſpeedily the plague of God falleth after ſinne; and ſpecially men
 muſt note what torment of conſcience, and deſperation often foloweth
 the ſheading of innocent bloud.}
repenting him, returned the thirtie ſiluer peeces to the cheefe Prieſts
and Ancients, \V ſaying: I haue ſinned, betraying iuſt bloud. But they
ſaid: What is that to vs? looke thou to it. \V And caſting downe the
ſiluer peeces in the temple, he departed: and went and
\LNote{Hanged himſelf}{If
\MNote{Deſperation.}
he had rightly repented, notwithſtanding his horrible treaſon, he might
 haue obtained mercy: but by hanging himſelf he tooke away al meanes of
 mercy and ſaluation, becauſe he died finally impenitent.}
hanged himſelf with an halter. \V And the cheefe Prieſts hauing taken
the ſiluer peeces, ſaid: It is not lawful to caſt them into the
\SNote{This Corbana was a place about the Tẽple, which receaued the
Peoples guifts, of offerings. See
\XRef{Mar.~12, v.~42.}}
Corbana, becauſe it is the price of bloud. \V And after they had
conſulted togeather, they bought with them the potters field, to be a
burying place for ſtrangers. \V For this cauſe that field was
called \Emph{Haceldama}, that is, \Emph{the field of bloud}, euen to
this preſent day. \V Then was fulfilled that which was ſpoken by
Ieremie the Prophet, ſaying:
\CNote{\XRef{Zac.~11,~12.}}
\Emph{And they tooke the thirtie peeces of
ſiluer, the price of the priced, whom they did price of the children of
Iſrael.} \V \Emph{And they gaue them into the potters field, as our Lord
did appoint to me.}

\V And \Sc{Iesvs} ſtood before the Preſident, and the Preſident asked
him, ſaying: Art thou the King of the Iewes? \Sc{Iesvs} ſaith to him:
Thou ſayeſt. \V And when he was accuſed of the cheefe Prieſts and
Ancients, he anſwered nothing. \V Then Pilate ſaith to him: Doeſt thou
not heare how great teſtimonies they alleage againſt thee? \V And he
anſwered him not to any word: ſo that the Preſident did maruel
excedingly.

\V And vpon the ſolemne day the Preſident had accuſtomed to releaſe vnto
the People one priſoner, whom they
%%% o-2181
would. \V And he had then a notorious priſoner, that was called
Barabbas. \V They
%%% 2362
therfore being gathered togeather, Pilate ſaid: whom wil you that I
releaſe to you, Barabbas, or \Sc{Iesvs} that is called Chriſt? \V For he
knew that for enuie they had deliuered him. \V And as he was ſitting in
place of iudgement, his wife ſent vnto him, ſaying: Haue thou nothing to
doe with that iuſt man. For I haue ſuffred many things this day in my
ſleep for him. \V But the cheefe Prieſts and Ancients perſuaded the
People, that they ſhould aske Barabbas, and make \Sc{Iesvs} away. \V And
the Preſident anſwering, ſaid to them: Whether wil you of the two to be
releaſed vnto you? But they ſaid, Barabbas. \V Pilat ſaith to them: What
ſhal I doe then with \Sc{Iesvs} that is called Chriſt? They ſay al: Let
him be crucified. \V The Preſident ſaid to them: Why what euil hath he
done? But they cried the more, ſaying: Let him be crucified. \V And
Pilat ſeeing that he nothing preuailed, but rather tumult was toward;
taking water he waſhed his hands before the People, ſaying: I am
\LNote{Innocent of his bloud}{Though
\MNote{They that execute lawes againſt their conſcience, are like to
Pilate.}
Pilate was much more innocent then the Iewes, and would haue been free
 frõ the murder of our Sauiour, ſeeking al the meanes that he could
 (without offending the People & the Emperours lawes) to dimiſſe him:
 Yet he is dãned for being the miniſter of the Peoples wicked wil
 againſt his owne cõſcience. Euen as al Officers be, and ſpecially the
 Iudges and Iuries which execute lawes of temporal Princes againſt
 Catholike men: for al ſuch be guilty of innocent bloud, and be nothing
 excuſed by that they execute other mens wil according to the lawes,
 which be vniuſt. For they ſhould rather ſuffer death them ſelues, then
 put an innocent man to death.}
innocent of the bloud of this iuſt man: looke you to it. \V And the whole
People anſwering, ſaid: His bloud be vpon vs, and vpon our children. \V
Then he releaſed to them Barabbas, and hauing ſcourged \Sc{Iesvs},
deliuered him vnto them for to be crucified.

\V Then the Preſidents ſouldiars taking \Sc{Iesvs} into the Palace,
gathered togeather vnto him the whole band: \V
\CNote{\XRef{Io.~19,~2.}}
and ſtripping him, put a
ſcarlet cloke about him, \V and platting a crowne of thornes, put it
vpon his head, and a reed in his right hand. And bowing the knee before
him, they mocked him, ſaying: Haile King of the Iewes. \V And ſpitting
vpon him, they tooke the reed, and ſmote his head. \V And after they had
mocked him, they tooke of the cloke from him, and put on him his owne
garments, and led him away to crucifie him. \V And in going they found a
man of Cyrene, named Simon: him they forced to take vp his Croſſe. \V
And they came into the place that is called Golgatha, which is the place
of Caluarie. \V And they gaue him wine to drinke mingled with gal. And
when he had taſted, he would not drinke.

\V And after they had crucified him, they deuided his garments, caſting
lots; that it might be fulfilled which was ſpoken by the Prophet,
ſaying:
\CNote{\XRef{Pſa.~21,~19.}}
\Emph{They deuided my garments among
%%% o-2182
them; and vpon my veſture they did caſt lots.} \V And they ſate and
watched him. \V And they put ouer his head his cauſe written: \Sc{This
is Iesvs the king of the iewes.} \V Then were crucified with him two
theeues; one on the right hand, and one on the left. \V And they that
paſſed by, blaſphemed him wagging their heads, \V and ſaying: Vah, thou
that deſtroyeſt the Temple of God and in three daies doeſt reedifie
it; ſaue thine owne ſelf:
\LNote{If thou be the Sonne}{Maruel
\MNote{Chriſt derided in the B.~Sacrament, euen as vpõ the Croſſe.}
not, when thou heareſt our Sauiour in the B.~Sacrament mocked at, or
 ſeeſt him abouſed of wicked men that he ſtraight reuengeth not ſuch
 blaſphemies: or he ſheweth not him ſelf there viſibly and to the
 ſenſes, when faithles Heretikes wil ſay: Let me ſee him, taſt him,
 &c. for he ſuffered here the like on the Croſſe, when he might at his
 wil haue come downe with as much eaſe as he roſe when he was dead.}
if thou be the Sonne of God, come downe from the Croſſe. \V In like
manner alſo the cheefe Prieſts with the Scribes and Ancients mocking,
ſaid: \V He ſaued other; himſelf he can not ſaue: if he be the King of
Iſrael, let him now come downe from the Croſſe, and we wil beleeue
him. \V
\CNote{\XRef{Pſ.~21,~9.}
\XRef{Sap.~2,~18.}}
He truſted in God; let him now deliuer him if he wil: for he
ſaid that I am the Sonne of God. \V And the ſelf ſame thing the theeues
alſo that were crucified with him, reproched him withal.

\V And from the ſixt houre there was darkeneſſe made vpon the whole
earth, vntil the ninth houre. \V And about the ninth houre
%%% 2363
\Sc{Iesvs} cried with a mighty voice, ſaying: \Emph{Eli, eli,
lamma-ſabacthani?} that is, \Emph{My God, my God,}
\LNote{Why haſt thou forſaken me}{Beware
\MNote{Deteſtable blaſphemie of Caluin. And other Miniſters of his
ſchoole.}
here of the deteſtable blaſphemie of Caluin and the Caluiniſts,
\CNote{\Cite{Catech. Calu. & Inſtit. l.~2.~16. §.~10.}}
who
thinking not the bodily death of Chriſt ſufficient, ſay, that he was
alſo here ſo forſaken and abandoned of his Father, that he ſuſtained in
ſoule and conſcience, the very feares and torments of the damned. Which
ſtrange doctrine we find alſo auouched in an Engliſh Catechiſme ſet forth by
\Cite{R.~H. Anno.~1583.}
in theſe words: \Emph{Queſt. By what meanes hath
Chriſt appeaſed his Father wrath, and ranſomed vs? Anſwer. By ſuffering
death of the Croſſe, and the torment of Hel in ſoule, and conſcience.}
Quite contrary to the holie Scriptures, which alwayes attribute our
Redemption to Chriſts death, and ſheding of his moſt precious bloud, as
to a moſt ſufficient, & onlie ranſom for al mankind.
\XRef{Ro.~5,~20.}
\XRef{Heb.~9,~14.}
\XRef{1.~Pet.~1,~19.}
\XRef{1.~Io.~1,~7.}
and in innumerable other places. But Caluin and his
Diſciples hold this wicked paradox, to take away the Article of Chriſts
 deſcending into Hel after his death, ſaying  that his deſcending was
 nothing els, but that his ſoule (remaining as yet in his bodie)
 ſuffered the very paines of Hel vpon the Croſſe. Wheras indeed his
 deſcending was in ſoule (parting forth of the bodie) into that place,
 where the Fathers of the old Teſtament were detained, expecting his
 comming to deliuer them (and that with triumph, and not in paine) as is
 proued by the Scriptures, and manie ancient Fathers, in the Annotations
 vpon 
\XRef{S.~Luke. c.~16,~22.}
\XRef{Act.~2,~27.}
\XRef{1.~Pet.~3,~19.}
and in other places. And concerning the true ſenſe
 of theſe words: \Emph{Why haſt thou forſaken me?} as wel by conferring
 them with other holie Scriptures, as by the vniforme conſent of the
 ancient Fathers, our Sauiour would ſignifie hereby, that his paines
 (being now ſo long on the Croſſe, and readie to die) were very great;
 and therfore according to the infirmitie of his humane nature, for very
 anguiſh (as before in the garden he ſweat bloud, when he was but
 toward his Paſſion) he ſaith, he was forſaken, for two cauſes. Firſt
 becauſe it was the wil of God not to deliuer him, but that he ſhould
 die. Secondly, becauſe his diuine nature did ſo repreſſe itſelf for the
 time, that he felt no comfort thereof at al, but was left to die in
 extreme paines, as a mere man. Yea deſtitute of ſuch conſolation, as
 his holie Martyrs commonly haue in their laſt agenie. See
\Cite{Origen. Tract.~35. in Mat.}
\Cite{S.~Hilarie li.~80. de Trinit.}
\Cite{S.~Leo ſer.~17. de Paſſione.}
\Cite{S.~Bede}
and others vpon this place.}
\Emph{why haſt thou forſaken me?} \V And certaine that ſtood there and
heard, ſaid: He calleth Elias. \V And incontinent one of them running,
tooke a ſponge, & filled it with vinegre; and put it on a reed, and gaue
him to drinke. \V And other ſaid: Let be, let vs ſee whether Elias come
to deliuer him. \V And \Sc{Iesvs} againe crying with a mighty voice,
yealded vp the ghoſt. \V And behold the vele of the Temple was rent in
two peeces, from the top euen to the botome, and the earth did quake,
and the rocks were rent, \V and the graues were opened: and many bodies
of the Saints that had ſlept, roſe. \V And they going forth out of the
graues after his Reſurrection, came into the holy citie; and appeared to
many. \V And the Centurion and they that were with him
watching \Sc{Iesvs}, hauing ſeen the earth-quake and the things that
were done, were ſore afraid, ſaying: In deed this was the Sonne of God.

\V And there were there many women a farre of which had folowed \Sc{Iesvs}
from Galilee, miniſtring vnto him: \V among whom was Marie Magdalene,
and Marie the mother of Iames and Ioſeph, and the mother of the Sonnes
of Zebedee. \V And when it was euening, there came a certaine
%%% o-2183
rich man of Arimathæa, named Ioſeph, who alſo him ſelf was Diſciple
to \Sc{Iesvs}. \V He went to Pilate, and asked the body
of \Sc{Iesvs}. Then Pilate cõmanded that the body ſhould be
deliuered. \V And Ioſeph taking the body,
\LNote{Wrapped}{This
\MNote{Reuerent vſing of our Lords Body.}
honour and duty done to Chriſts body being dead, was maruelous grateful
 and meritorious. And this wrapping of it in cleane ſindon may ſignifie
 by
\CNote{\Cite{S.~Hiero. in hunc locum.}}
S.~Hierom, that the body of our Lord is to be wrapped not in gold,
 pretious ſtones, and ſilke, but in pure linnen.
\MNote{Corporals.}
And ſo in the whole Church it is obſerued by
\CNote{\Cite{to.~1. Concil.}}
S.~Silueſters conſtitution,
 that the Corporal whereupon our Lordes body lieth on the Altar, muſt be
 pure and plaine linnen.}
wrapt it in cleane ſindõ, \V & laid it in his owne new monumẽt, which he
had hewed out in a rock. And he roled a great ſtone to the doore of the
monument, & went his way. \V And there was there Marie Magdalene, & the
other Marie ſitting ouer againſt the ſepulchre.

\V 
\MNote{\Sc{Satvrday} called Sabbatum ſanctum.}
And the next day, which is after the Paraſceue, the cheefe Prieſts and the
Phariſees came togeather to Pilate, \V ſaying: Sir, we haue remembred,
that the ſeducer ſaid yet liuing, After three dayes I wil riſe
againe. \V Command therfore the ſepulchre to be kept vntil the third
day; leſt perhaps his Diſciples come, and ſteale him, & ſay to the
People, he is riſen from the dead: and the laſt errour ſhal be worſe
then the firſt. \V Pilat ſaid to them: You haue a gard: goe, gard it as
you know. \V And they departing, made the ſepulchre ſure; ſealing vp the
ſtone, with watchmen.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
