%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 2342
%%% o-2158
\startChapter[
  title={Chapter 21}
  ]

\Summary{Being
\MNote{The fifth part of this Ghoſpel. Of the Holy week of his Paſſion
  in Hieruſalem.}
now come to the place of his paſsion, he entreth with humilitie and
triumph togeather: 12.~Sheweth his zeale for the houſe of God ioyned
with great maruels.  15.~And to the Rulers he boldly defendeth the
acclamations of the children. 18.~He curſeth alſo that fruitles leafie
tree: 23.~auoucheth his power by the witnes of Iohn: 28.~and foretelleth
in two parables their reprobation (with the Gentils vocation) for their
wicked deſerts, 42.~and conſequently their irreparable damnation that
shal enſue therof.}

And
\CNote{\XRef{Mr.~11,~1.}
\XRef{Luc.~19,~20.}
\XRef{Io.~12,~15.}}
\MNote{\Sc{Palme Svnday.}}
when they drew nigh to Hieruſalem, and were come to Beth-phagee vnto
Mount-oliuet, then \Sc{Iesvs} ſent two Diſciples, \V ſaying to them: Goe
ye into the towne that is againſt you, and immediatly
\LNote{You shal find}{Chriſt by diuine power both knew where theſe
  beaſts were, being abſent, and commanded them for his vſe, being an other
mans, and ſuddenly made the colt fit to be ridden on, neuer broken
before.}
you ſhal find an aſſe tied and a colt with her: looſe them & bring them
to me: \V And if any man ſhal ſay ought vnto you, ſay ye, that our Lord
hath need of them: and forthwith he wil let them goe. \V And this was
done that it might be fulfilled which was ſpoken by the Prophet, ſaying:
\V
\CNote{\XRef{Eſa.~62.~11.}
\XRef{Zach.~9,~9.}}
\Emph{Say ye to the daughter of Sion: Behold thy King commeth to
thee, meeke, & ſitting vpon an aſſe and a colt the fole of her that is
vſed to the yoke.} \V And the Diſciples going, did as \Sc{Iesvs}
commanded them. \V And they brought 
\LNote{The aſſe and the colt}{This
\CNote{\Cite{Hiero. in Mat.}
\Cite{Aug. li.~12. cont. Fauſt. c.~41.}}
aſſe vnder yoke ſignifieth the Iewes
vnder the Law and vnder God their Lord, as it were his old and ancient
People: the yong colt now firſt ridden on by Chriſt, ſignifieth the
Gentils, wild hitherto and not broken, now to be called to the faith and
to receaue our Sauiours yoke. And therfore the three laſt Euangeliſts
writing ſpecially to the Gentils, make mention of the colt only.}
the aſſe and the colt: and laid their garments vpon them, and made him
to ſit theron. \V And a very great multitude ſpred their
\LNote{Garments in the way}{Theſe
\MNote{Proceſsion on Palme-ſunday with the B.~Sacrament.}
\MNote{Al deuout offices in that kinde, exceeding grateful.}
offices of honour done to our Sauiour extraordinarily, were very
acceptable: and for a memorie hereof the holy Church maketh a ſolemne
Proceſſion euery yeare vpon this day, ſpecially in our Countrie when it
was Catholike, with the B.~Sacrament reuerently caried, as it were
Chriſt vpon the aſſe, and ſtrawing of ruſhes and floures, bearing of
Palmes, ſetting vp boughes, ſpreading and hanging vp the richeſt
clothes, the quire and queriſters ſinging, as here the children and the
People; al done in a very goodly ceremonie to the honour of Chriſt, and
the memorie of his triumph vpon this day. The like ſeruice and the like
duties done to him in al other ſolemne Proceſsions of the B.~Sacrament,
and otherwiſe, be vndoubtedly no leſſe grateful.}
garments in the way: and others did cut boughs from the trees, and
ſtrawed them in the way: \V and the 
%%% o-2159
multitudes that went before and that folowed, cried, ſaying:
\CNote{\XRef{Ps.~117,~26.}}
\LNote{Hoſanna}{Theſe
\MNote{\Sc{Hosanna}}
very wordes of ioyful crie and triumphant voice of gratulations to our
Sauiour, holy Church vſeth alwaies in the Preface of the Maſſe, as it
were the voice of the Prieſt and al the People (who then ſpecially are
attent and deuout) immediatly before the Conſecration & Eleuation, as it
were expecting, & reioycing at his cõming.}
\Emph{Hoſanna to the Sonne of Dauid: Bleſſed is he that commeth in the
name of our Lord. Hoſanna in the higheſt.}

\V And when he was entred Hieruſalem, the whole citie was moued, ſaying:
who is this? \V And the People ſaid: This is \Sc{Iesvs} the Prophet, of
Nazareth in Galilee. \V And
\CNote{\XRef{Mr.~11,~15.}
\XRef{Lu.~12,~45.}}
\Sc{Iesvs}, entred in the temple of God, and caſt out al that
\SNote{How much the abuſe of Churches by merchãdiſing, walking, or other
profane occupying of them, diſpleaſeth God, here we may ſee.}
ſold and bought in the temple, and the tables of the bankers, & the
chaires of them that ſold pigeons he ouerthrew: \V and he ſaith to them:
It is written,
\CNote{\XRef{Eſa.~56,~7.}
\XRef{Ier.~7,~11.}}
\Emph{My houſe shal be called the}
\LNote{Houſe of prayer}{Note here that he calleth external
Sacrifice (out of the Prophet Eſay) prayer. For he ſpeaketh of the
Temple, which was builded properly and principally for Sacrifice.}
\Emph{houſe of prayer: but you haue made it a denne of theeues.} \V And
there came to him the blind, and the lame in the temple; and he healed
them. \V And the cheefe Prieſts & Scribes ſeeing the maruelous things
that he did, and the children crying in the tẽple, & ſaying,
\Emph{Hoſanna to the Sonne of Dauid}; they had indignation, \V and ſaid
to him: Heareſt thou what theſe ſay? And \Sc{Iesvs} ſaid to them: Very
wel, haue you neuer read:
\CNote{\XRef{Pſ.~8,~3.}}
\Emph{That out of the}
\LNote{Mouth of infants}{Yong 
\MNote{Prayers not vnderſtood of the partie, are acceptable.}
childrens prayers proceeding from
the inſtinct of Gods Spirit, be acceptable: and ſo the voices of the
like, or of other ſimple folke now in the Church, though them ſelues
vnderſtand not particularly what they ſay, be maruelous grateful to
Chriſt.}
\Emph{mouth of infants and ſucklings thou haſt perfited praiſe}? \V And
leauing thẽ, he went forth out of
%%% 2343
the citie into Bethania, and remained there.

\V
\CNote{\XRef{Mr.~11,~13.}}
\MNote{\Sc{Mvnday.}}
And in the morning returning into the citie, he was an hungred. \V And
ſeeing a certaine
\SNote{The Iewes hauing the wordes of the law, and not the deedes, were
the figtree ful of leaues, and void of fruit.
\Cite{Aug. de verb. Do. Serm.~44.}}
figtree by the way ſide, he came to it, and found nothing on it but
leaues only, and he ſaith to it: Neuer grow there fruit of thee for
euer. And incontinent the figtree was withered. \V And the Diſciples
ſeeing it marueled ſaying: How is it withered incontinent? \V
\MNote{\Sc{Tvesday.}}
And \Sc{Iesvs} anſwering ſaid to them: Amen I ſay to you,
\CNote{\XRef{Mt.~17,~20.}}
if you ſhal
haue faith, and ſtagger not, not only that of the figtree ſhal you doe,
but and if you ſhal ſay to this mountaine, Take vp and throw thyſelf into
the ſea, it ſhal be done. \V And al things whatſoeuer you ſhal aske in
prayer 
\LNote{Beleeuing}{In 
reſpect of our own vnworthineſſe, and of the thing
not alwaies expedient for vs, we may wel doubt when we pray, whether we
ſhal obtaine or no: but on Gods part we muſt beleeue, that is, we muſt
haue no diffidence or miſtruſt either of his power or of his wil, if we
be worthie, and the thing expedient. And therfore S.~Marke hath thus:
\CNote{\XRef{Mr.~11,~22.}}
\Emph{Haue ye faith of God.}}
beleeuing, you ſhal receaue.

\V And when he was come into the temple, there came to him as he was
teaching, the cheefe Prieſts and Ancients of the People, ſaying:
\CNote{\XRef{Mr.~11,~28.}
\XRef{lu.~20,~2.}}
\LNote{In what power?}{The Heretikes preſumptuouſly thinke themſelues in
this point like to Chriſt, becauſe they are aſked, in what power they
come, and who ſent them: but when they haue anſwered this queſtion as
fully as Chriſt did here, by that which he inſinuateth of Iohns
teſtimonie for his authority, they ſhal be heard, and til then they
shal be ſtil takẽ for thoſe of whom God ſpeaketh by the Prophet:
\CNote{\XRef{Iere.~23.}}
\MNote{Heretikes runne, not ſent.}
\Emph{They ranne, and I ſent thẽ not.}}
In what power doeſt thou theſe things? and who hath giuen thee this
power? \V \Sc{Iesvs} anſwering ſaid to them: I alſo wil aske you one
word, which if you ſhal tel me, I alſo wil tel you in what power I doe
theſe things. \V The Baptiſme of Iohn whence was it? from Heauen, or
from men? But they thought within themſelues, ſaying: \V If we ſhal ſay
from Heauen, he wil ſay to vs, why then did you not beleeue him? But if
we ſhal ſay from men, we feare the multitude, for al hold Iohn as a
Prophet. \V And anſwering to 
%%% o-2160
\Sc{Iesvs} they ſaid: We know not. He alſo ſaid to them: Neither do I
tel you in what power I doe theſe things. 

\V But what is your opinion? A certaine man had two ſonnes; and comming
to 
\LNote{The firſt}{The firſt ſonne here is the People of the Gentils,
  becauſe Gentilitie was before there was a peculiar and choſen People
  of the Iewes, and therfore the Iewes here as the later, are ſignified
  by the other ſonne.}
the firſt, he ſaid: Sonne, goe worke today in my vineyard. \V And he
anſwering, ſaid: I wil not. But afterward moued with repentance he
went. \V And comming to the other, he ſaid likewiſe. And he anſwering,
ſaid: I goe Lord, and he went not: \V which of the two did the fathers
wil? They ſay to him: The firſt. \Sc{Iesvs} ſaith to them: Amen I ſay to
you, that the Publicans and whoores goe before you into the Kingdom of
God. \V For Iohn came to you in the way of iuſtice, and you did not
beleeue him. But the publicans and whoores did beleeue him: but you
ſeeing it, neither haue ye had repentance afterward, to beleeue him.

\V An other parable heare ye: A man there was an houſholder who
\CNote{\XRef{Pſ.~5,~1.}
\XRef{Mr.~12,~1.}
\XRef{Lu.~20,~9.}}
planted
a vineyard, and made a hedge round about it, and digged in it a preſſe,
and builded a towre, and let it out to husbandmen: and went forth into a
ſtrange countrie. \V And when the time of fruits drew nigh, he ſent his
ſeruants to the husbandmen, to receaue the fruits therof. \V And the
husbandmen apprehending his ſeruants, one they beat, an other they
killed, and an other they ſtoned. \V Againe he ſent other ſeruants more
then the former: and they did to them likewiſe. \V And laſt of al he
ſent to them his ſonne, ſaying: They wil reuerence my ſonne. \V But the
husbandmen ſeeing the ſonne, ſaid within themſelues: This is the Heire,
come, let vs kil him, and we ſhal haue his inheritence. \V And
apprehending him they caſt him forth out of the vineyard, & killed
him. \V When therfore the Lord of the vineyard ſhal come, what wil he
doe to thoſe husbandmen? \V They ſay to him: The
%%% 2344
naughtie men he wil bring to naught: and his vineyard he wil let out to
other huſbandmen, that ſhal render him the fruit in their ſeaſons.

\V \Sc{Iesvs} ſaith to them: Haue you neuer read in the Scriptures:
\CNote{\XRef{Pſa.~117,~21.}}
\Emph{The ſtone which the builders reiected, the ſame is made into the
  head of the corner? By our Lord was this done, and it is maruelous in
  our eyes.} \V Therfore I ſay to you, that the Kingdom of God ſhal be
taken away from you, and ſhal be giuen to a Nation yealding the fruits
therof. \V And
\CNote{\XRef{Eſ.~8,~14.}}
he that falleth vpon this ſtone, ſhal be broken: and on
whom it falleth, it shal al to bruiſe him. \V And when the cheefe
Prieſts and Phariſees had heard his parables, they knew that he ſpake of
them. \V And ſeeking 
%%% o-2161
to lay hands vpon him, they feared the multitudes: becauſe they held him
as a Prophet.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
