%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 24
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-24


\project douay-rheims


%%% 2350
%%% o-2168
\startChapter[
  title={Chapter 24}
  ]

\Summary{To his Diſciples (by occaſion of Hieruſalem and the Temples
  deſtruction) he foretelleth, 4.~what things shal be before the
  conſummation of the world, as ſpecially 14.~the Churches ful preaching
  vnto al Nations: 15.~then, what shal be in the very conſummation, to
  wit, Antichriſt with his paſsing great perſecution and ſeduction, but
  for a short time: 29.~then incontinent, the Day of iudgement, to our
  great comfort in thoſe miſeries vnder Antichriſt. 35.~As for the
  moment, to vs it pertaineth not to know it, 37.~but rather euery man
  to watch, that we be not vnprouided when he commeth to each one
  particularly by death.}

And \Sc{Iesvs} being gone out of the temple, went.
\CNote{\XRef{Mr.~13,~1.}
\XRef{Luc.~21,~5.}}
And his Diſciples
came to ſhew him the buildings of the temple. \V And he anſwering ſaid
to them: Doe you ſee al theſe things? Amen I ſay to you, there ſhal
\LNote{Not left}{This was fulfilled 40. yeares after Chriſtes
  Aſcenſion by Veſpaſian the Emperour and his ſonne Titus.
\Cite{Euſeb. li.~3. c.~6. & ſeq. ex Ioſepho.}
Vpon which words, \Emph{There shal not be left &c.} which threaten the
deſtruction of the Iewes Temple; and thoſe words,
\MNote{The Church cã neuer faile.}
\CNote{\XRef{Mat.~16.}}
 \Emph{vpon this Rock I
wil build my Church}, which promiſe the building of the Catholike Church
of al Nations;
\CNote{\Cite{Li. adu. Gentes quod Chriſtus ſit Deus, prope finum.}}
S.~Chryſoſtome making a long compariſon of theſe two
Prophecies of Chriſt, ſaith thus: Thou ſeeſt in both, his great and
vnſpeakable power, in that that he increaſed and built vp them that
worshipped him, and thoſe that ſtumbled at him, he abaſed, deſtroyed,
and plucked them vp by the root. Doeſt thou ſee how whatſoeuer he hath
built, no man shal deſtroy: and whatſoeuer he hath deſtroyed, no man
shal build? He builded the Church, and no man shal be able to deſtroy
it; he deſtroyed the Temple, and no man is able to build it, and that in
ſo long time: For they haue endeauoured both to deſtroy that, and could
not: and they haue attempted to build vp this, and they could not doe
that neither, &c.}
not be left here a ſtone vpon a ſtone that ſhal not be deſtroied.

\V And when he was ſitting vpon Mount-oliuet, the Diſciples came to him
ſecretly, ſaying: Tel vs when ſhal theſe things be? and what
%%% 2351
ſhal be
\LNote{The ſigne}{Our Maiſter knowing that it was not profitable nor
  ſeemly for them to know theſe ſecrets, gaue them by way of Prophecie
  warning of diuers miſeries, ſignes, and tokens, that should fal, ſome
  further of, and ſome neerer the later day: by which the Faithful might
alwayes prepare them ſelues, but neuer be certaine of the houre, day,
month, nor yeare, when it should fal.
\Cite{Aug. ep.~80.}}
the ſigne of thy comming, and of the conſummation of the world? \V And
\Sc{Iesvs} anſwering, ſaid to thẽ: Beware that no man 
\LNote{Seduce}{The
\MNote{Heretikes ſeduce vnder faire titles.}
firſt and principal warning, needful for the faithful from Chriſtes
Aſcenſion to the very end of the world, is, that they be not deceaued by
Heretikes, which vnder the titles of true teachers, and the name of
Chriſt and his Ghoſpel, wil ſeduce many.}
ſeduce you: \V for many ſhal come in my name ſaying,
\LNote{I am Chriſt}{Not only ſuch as haue named them ſelues Chriſt, as
  Simon, Menander, and ſuch like; but al Arch-heretikes be Chriſts to
  their folowers, Luther to the Lutherans, Caluin to the Caluiniſts,
  becauſe they beleeue them, rather then Chriſt ſpeaking in his Church.}
I am Chriſt; and they shal ſeduce many. \V For you shal heare of warres,
& bruits of warres. See that ye be not troubled. For theſe things muſt
be done, but the end is not yet. \V For Nation shal riſe againſt Nation,
and Kingdom againſt Kingdom; and there shal be peſtilences, and famines,
and earth-quakes in places; \V and al theſe things are the beginnings of
ſorowes. \V Then
\CNote{\Cite{Mt.~10,~17.}}
shal they deliuer you into tribulation, and shal kil
you: and you shal be odious to al Nations for my names ſake. \V And then
many shal be ſcandalized: and they shal deliuer vp one an other: and
they shal hate one an other. \V And many
\SNote{There were in the People falſe Prophets, as among you alſo ſhal be
lying Maiſters, which ſhal bring in Sects of perdition.
\XRef{2.~Pet.~2.}}
Falſe-Prophets shal riſe, and shal ſeduce many. \V And becauſe
\LNote{Iniquity abound}{When Hereſie and falſe teachers raigne in
the world, namely toward the later day, wicked life aboundeth, and
charitie decayeth.}
iniquitie shal abound, the charitie of many shal waxe cold. \V But he
that shal perſeuer to the end, he shal be ſaued. \V And this Ghoſpel of
the Kingdom
\LNote{Shal be preached}{The
\MNote{The Ghoſpel now preached to Infidels.}
Ghoſpel hath been preached of late yeares, and now is, by holy Religious
men of diuers Orders in ſundry great Countries which neuer heard the
Ghoſpel before, as it is thought.}
shal be preached in the whole world, for a teſtimonie to al Nations, and
then shal come the conſummation.

\V Therfore when you shal ſee
\LNote{Abomination of deſolation}{This 
\MNote{The abomination of deſolation.}
abomination of deſolation foretold, was firſt partly fulfilled in
diuerſe prophanations of the Temple of Hieruſalem, when the Sacrifice
and ſeruice of God was taken away, 
\MNote{The abolishing of the holy Sacrifice of the Maſſe by Antichriſt,
  and his miniſters.}
but ſpecially it ſhal be fulfilled by
Antichriſt and his Precurſours, whẽ they ſhal aboliſh the holy Maſſe,
which is the Sacrifice of Chriſtes body and bloud, & the only ſoueraigne
worſhip due to God in his Church: as S.~Hypolitus writeth in theſe
words: The Churches ſhal lament with great lamentations, becauſe there
ſhal neither Oblation be made, nor incenſe, nor worſhip grateful to
God. But the ſacred houſes of Churches shal be like to cottages, and the
pretious body and bloud of Chriſt ſhal not be extant (openly in
Churches) in thoſe dayes, the Liturgie (or Maſſe) shal be extinguished,
the Pſalmodie shal ceaſe, the reciting of the Scriptures shal not be
heard.
\Cite{Hippol. de Antichriſto.}
By which it is plaine, that the Heretikes of theſe daies be the ſpecial
fore-runners of Antichriſt.}
\CNote{\Cite{Dan.~9,~27.}}
\Emph{the abomination of deſolation}, which was ſpoken of by Daniel the
Prophet, ſtanding in the holy place (he that readeth, let him
vnderſtand) \V then they that are in Iewrie, let them flee to the
mountaines: \V and he that is on the houſe-top let him not come downe 
%%% o-2169
to
take any thing out of his houſe: \V and he that is in the field, let him
not goe back to take his coate. \V And woe to them that are with child,
and that giue ſuck in thoſe dayes. \V But pray that your flight be not
in the winter, or on the Sabboth. \V For there shal be then great
tribulatiõ, ſuch as hath not been from the beginning of the world vntil
now, neither shal be. \V And vnles thoſe daies had been shortned, no
flesh should be ſaued: but for the Elect the daies
\LNote{Shal be shortned}{The
\MNote{Antichriſt.}
raigne of Antichriſt shal be short, that is, three yeares and a half.
\XRef{Dan.~7.}
\XRef{Apoc.~11.}
Therfore the Heretikes are blaſphemous and ridiculous, that ſay,
Chriſtes Vicar is Antichriſt, who hath ſitten theſe 1500.\ yeares.}
shal be shortned. \V Then if any man shal ſay vnto you: Loe
\SNote{Whoſoeuer draweth Chriſt or his Church from the Communion &
  felowſhip of al Nations Chriſtened, to one corner, towne, or
  Countrie, beleeue him not.
\Cite{Aug. de vnit. Ec. c.~3.}}
here is Chriſt, or there; doe not beleeue him. \V For there shal riſe
falſe Chriſtes and falſe-Prophets, and shal shew 
\LNote{Great ſignes}{Theſe ſignes and miracles shal be to the outward
  appearance only: for S.~Paul calleth them
\CNote{\Cite{2.~Theſ.~2.}}
lying ſignes, to ſeduce them
only that shal perish. Wherby we ſee that if Heretikes could worke
feyned and forged miracles, yet we ought not to beleeue them, much leſſe
when they can not ſo much as ſeeme to doe any.}
great ſignes and
wonders, ſo that the Elect alſo (if it be poſſible) may be induced into
errour. \V Loe I haue foretold you. \V If therfore they shal ſay vnto
you: Behold he is in the deſert; goe ye not out: Behold
\LNote{In cloſets}{Chriſt
\MNote{The ſecret cõuenticles of Heretikes.}
hauing made the Churches authority bright and cleare to the whole world,
warneth the Faithful to take heed of Heretikes and Schiſmatikes, which
haue their cõuenticles aſide in certaine odd places and obſcure
corners, alluring curious Perſons vnto them.
\Cite{Aug. li.~1. q.~Euang. q.~38.} 
\MNote{Catholike Chriſtians ſecretly aſſembling in time of perſecution.}
For as for the comming togeather of Catholikes to ſerue God in ſecret
places, that is a neceſſarie thing in time of perſecution, and was vſed
of Chriſtians for three hundred yeares togeather after Chriſt;
\CNote{\Cite{Act.~1,~12.}}
& the
Apoſtles alſo and Diſciples came ſo togeather in Hieruſalem for feare of
the Iewes. And Catholikes doe the ſame at this day in our countrie, not
drawing religion into corners from the ſociety of the Catholike Church;
but practiſing ſecretly the ſame faith, that in al Chriſtendom shineth
and appeareth moſt gloriouſly.}
in the cloſets, beleeue it not. \V For as lightning commeth out of the
eaſt, and appeareth euen into the weſt, ſo shal alſo the aduent of the
Sonne of man be. \V Wherſoeuer the body is, thither shal the Egles alſo
be gathered togeather.

\V And 
\LNote{Immediatly}{If 
\MNote{Antichriſt.}
the later day shal immediatly folow the perſecution of Antichriſt, which
is to endure but three yeares and a halfe, as is aforeſaid, then it is
mere blaſphemie to ſay, Gods Vicar is Antichriſt, & that (by their owne
limitation) theſe thouſand yeares almoſt.}
immediatly after the tribulation of thoſe dayes
\CNote{\Cite{Eze.~12,~7.}
\Cite{Ioel.~3,~15.}
\Cite{Dan.~7,~15.}}
the ſunne shal be
darkned, and the moone shal not giue her light, and the ſtarres shal fal
from Heauen, and the powers of Heauen shal be moued: \V and then ſhal
appeare
\SNote{This ſigne of the Sonne of mã, is the holy Croſſe, which thẽ ſhal
appeare to the Iewes to their confuſion. 
\Cite{Chrys. in Mat. hom.~77.}
It ſhal be no leſſe confuſion to Heretikes that can not abide the ſigne
thereof.}
the ſigne of the Sonne of man in Heauen: and then ſhal al Tribes of the
earth bewaile: and they shal ſee the Sonne of man comming in the cloudes
of Heauen with much power and maieſtie. \V And he shal ſend his Angels
with a trumpet, and a great voice: and they shal gather togeather his
Elect from the foure windes, from the furtheſt parts of Heauen euen to
the ends thereof. \V And of the figtree learne a parable: When now the
bough thereof is tender, and the leaues come forth, you know that ſommer
is nigh. \V So you
%%% 2352
alſo, when you ſhal ſee theſe things, know ye that it is nigh euen at
the doores. \V Amen I ſay to you, that this Generation ſhal not paſſe
til al theſe things be done. \V Heauen and earth ſhal paſſe, but my
words ſhal not paſſe.

\V But of that day and houre nobody knoweth, neither the Angels of
Heauen, but the Father alone. \V And as
\CNote{\Cite{Gen.~7,~5.}}
in the dayes of Noe, ſo ſhal
alſo the comming of the Sonne of man be. \V For as they were in the
dayes before the floud, eating and drinking, marying and giuing to
mariage, euen vnto that day in which Noe entred into the arke, \V and
knew not til 
%%% o-2170
the floud came, & tooke thẽ al: ſo alſo ſhal the comming of
the Sonne of mã be. \V Then two ſhal be in the field: one ſhal be taken,
& one ſhal be left. \V Two woman grinding in the mill: one ſhal be
taken, & one ſhal be left. \V Watch therfore becauſe you know not what
houre your Lord wil come. \V But this know ye, that
\CNote{\Cite{1.~Theſ.~5,~1.}}
if the Good-man of
the houſe did know what houre the theefe would come, he would ſurely
watch, and would not ſuffer his houſe to be broken vp. \V Therfore be
you alſo ready, becauſe at what houre you know not, the Sonne of man wil
come. 

\V Who, thinkeſt thou, is a faithful and wiſe ſeruant whom his Lord hath
appointed ouer his familie, to giue them meate in ſeaſon? \V Bleſſed is
that ſeruant, whom when his Lord commeth, he ſhal find ſo doing. \V Amen
I ſay to you, that ouer al his goods ſhal he appoint him. \V But if that
naughtie ſeruant ſhal ſay in his hart: My Lord is long a comming: \V and
ſhal begin to ſtrike his felow-ſeruants, and eateth, & drinketh with
drunkards: \V the Lord of that ſeruant ſhal come in a day that he hopeth
not, & an houre that he knoweth not, \V and ſhal deuide him, & appoint
his portion with the Hypocrites: there ſhal be weeping and gnaſhing of
teeth.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
