%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 2323
%%% o-2138
\startChapter[
  title={Chapter 14}
  ]

\Summary{Hearing the vnworthy Decollation of Iohn Baptiſt by Herod,
  12.~he betaketh him to his vſual ſolitarines in the deſert, and there
  feedeth 5000.\ with fiue loaues. 23.~And then after the night ſpent in
the mountaine in prayer, he walketh vpon the ſea (ſignifying the wide
world) 28.~yea and Peter alſo: wherevpon they adore him as the Sonne of
God. 35.~And with the very touch of his garments hemme he healeth
innumerable.}

At that time
\CNote{\XRef{Mr.~6,~14.}
\XRef{Lu.~9,~7.}
\XRef{3,~19.}}
Herod the Tetrarch heard the fame of \Sc{Iesvs}: \V and
ſaid to his ſeruants: This is Iohn the Baptiſt: he is riſen from the
dead, & therfore vertues worke in him. \V For Herod apprehended Iohn and
bound him, & put him into priſon
\LNote{Becauſe of Herodias}{It is too ordinarie in Princes to put them
  to death that freely tel them ſuch faults: women, whom they fanſie,
  eſpecially inciting them to ſuch miſcheefe.}
becauſe of Herodias, his 
\Var{brothers}{brother Philips}
wife. \V For Iohn ſaid vnto him: It is not lawful for thee to haue
her. \V And willing to put him to death, he feared the People: becauſe
they eſteemed him as a Prophet. \V But on Herods birth-day, the daughter
of Herodias danced before them: and pleaſed Herod. \V Wherevpon he
promiſed with
%%% 2324
an oth, to giue her whatſoeuer ſhe would aske of him. \V But ſhe being
inſtructed before of her mother ſaith: Giue me here in a diſh the head of
Iohn the Baptiſt. \V And the King was ſtroken ſad: yet becauſe of his
\SNote{A wicked & raſh oth, and more wickedly fulfilled: becauſe an
  vnlawful oth bindeth no mã.}
oth, and for them that ſate with him at table, he commanded it to be
giuen. \V And he ſent, and beheaded Iohn in the priſon. \V And his head
was brought in a diſh: 
%%% o-2139
and it was giuen to the damſel, and ſhe brought
it to her mother. \V And his Diſciples came and took the body, and
\LNote{Buried it}{An 
\MNote{Sacrilege againſt holy Relikes.}
example of duty toward the dead bodies of the
  Faithful: wherein ſee the difference of Catholike Chriſtian men, & of
  al Infidels, be they Pagans, Apoſtates, or Heretikes. For whereas the
  Chriſtians had layd the body of this Bleſſed Prophet and Martyr
\CNote{\Cite{Hiero. in Epitaph. Paulæ. c.~6.}}
in Samaria with the Relikes of Elias and Abdias, by vertue wherof
wonderful miracles were wrought in that place, in Iulian the Apoſtates
time, when men might doe al miſcheefe freely againſt Chriſtian religion,
the Pagans opened the tombe of S.~Iohn Baptiſt, burnt his bones,
ſcattered the aſhes about the fields: but certaine religious Monkes
coming thither a pilgrimage at the ſame time, aduentured their life and
ſaued as much of the holy Relikes as they could, and brought them to
their Abbot Philip, a man of God: who eſteeming them too great a
treaſure for him and his to keep for their 
\Fix{primate}{priuate}{obvious typo, fixed in other}
deuotion, ſent them to Athanaſius the B.~of Alexandria; and he with al
reuerence layd them in ſuch a place (as it were by the Spirit of
Prophecie) where afterward by occaſion of them was built a goodly
chappel. 
\Cite{Theod. li.~5. c.~6.}
\Cite{Ruſſ. li.~2. c.~27,~28.}
Marke here that the Heretikes of our time doe as thoſe Pagans, to the
bodies & Relikes of al Bleſſed Saints that they can deſtroy: and
Catholikes contrariwiſe haue the religious deuotion of thoſe old
Chriſtians, as appeareth by the honour done now to his head at Amians in
France.}
buried it
\SNote{S.~Iohns Diſciples at this time had wel learned their duty toward
Chriſt.}
and came and told \Sc{Iesvs}.

\V Which when \Sc{Iesvs} had heard,
\CNote{\XRef{Mr.~6,~31.}
\XRef{Luc.~9,~10.}
\XRef{Io.~6,~2.}}
he
\LNote{Retired}{Chriſt much eſteemed Iohn, and withdrew himſelf aſide,
  to giue example of moderate mourning for the departed, and to ſhew the
horrour of that execrable murder: as in the Primitiue Church many good
men ſeeing the miſerable ſtate of the world in time of perſecution, and
the ſinnes that abounded withal, took an occaſion to forſake thoſe
tumults, and to giue themſelues to contemplation;  
\MNote{Eremites.}
and for that purpoſe
retired into the deſerts of Ægypt, and els where, to doe pennance for
their owne ſinnes, and the ſinnes of the world. Wherevpon partly roſe
that infinite number of Monkes & Eremites, of whom the Fathers and
Eccleſiaſtical hiſtories make mention.
\Cite{Hiero.~2.~10. in vis. Pauli Eremitæ.}
\Cite{Sozo. li.~1. c.~12.~13.}}
retired from thence by boat, into a deſert place apart, and the
multitudes hauing heard of it, folowed him on foot out of the cities. \V
And he cõming forth ſaw a great multitude, and pitied them, and cured
their diſeaſed. \V And when it was euening, his Diſciples came vnto him,
ſaying: It is a deſert place, and the houre is now paſt: dimiſſe the
multitudes, that going into the townes, they may buy themſelues
victuals. \V But \Sc{Iesvs} ſaid to them: They haue no need to goe: giue
ye them to eate. \V They anſwered him: We haue not here, but fiue
loaues, and two fiſhes. \V Who ſaid to them: Bring them hither to me. \V
And when he had commanded the multitude to ſit downe vpon the graſſe, he
took the fiue loaues and the two fiſhes, and looking vp vnto Heauen he
bleſſed, and brake, and gaue the loaues to his Diſciples, and
\LNote{The Siſciples to the multitudes}{A figure of the miniſtrie of
  the Apoſtles; who as they here had the diſtribution and ordering of
  theſe miraculous loaues, ſo had they alſo to beſtow and diſpenſe al
  the foode of our ſoules, in miniſtring of the Word & Sacramẽts;
  neither may laimen chalenge the ſame.}
the Diſciples to the multitudes. \V And they did al eate, and had their
fil. And they took the leauings, twelue ful baſkets of the fragments. \V
And the number of them that did eate was, fiue thouſand men, beſide women
and children.

\V And forthwith \Sc{Iesvs} commanded his Diſciples to goe vp into the
boat, and to goe before him ouer the water, til he dimiſſed the
multitudes. \V And hauing dimiſſed the multitude, he
\CNote{\XRef{Mr.~6,~46.}
\XRef{Io.~2,~26.}}
aſcended into a
mountaine alone to pray. And when it was euening, he was there alone. \V
But the boat in the middes of the ſea was toſſed with waues: for the
wind was contrarie. \V And in the fourth watch of the night, he came
vnto them walking vpon the ſea. \V And ſeeing him vpon the ſea
\LNote{Walking}{When not only Chriſt, but by his power Peter alſo
  walketh vpon the waters, it is euident that he can diſpoſe of his owne
body aboue nature, and contrary to the natural condition therof, as to
goe through a doore.
\XRef{Io.~20.}
to be in the compaſſe of a litle bread.
\Cite{Epiphan. in Anchorato.}}
walking, they were troubled ſaying: That it is a Ghoſt: & for feare
they cried out. \V And immediatly \Sc{Iesvs} ſpake vnto them, ſaying:
Haue confidence it is I, feare ye not. \V And Peter making anſwer ſaid:
Lord if it be thou, bid me come to thee vpon the waters. \V And he ſaid,
Come. And Peter deſcending out of the boat,
\LNote{Walked}{Peter
\MNote{Peters Primacie.}
(ſaith S.~Bernard) walking vpon the waters, as Chriſt did, declared
  himſelf the only Vicar of Chriſt, which ſhould be Ruler not ouer one
  People, but ouer al. For many waters, are many peoples. 
\Cite{Bernard li.~2. de conſid. c.~8.}
See the place, how he deduceth from Peter the like
\Fix{auctoritie}{authoritie}{obvious typo, fixed in other}
and iuriſdiction to his Succeſſour the Biſhop of Rome.}
walked vpon the water to come to \Sc{Iesvs}. \V But ſeeing the wind
rough, he was afraid: and when he began to be drowned, he cried out
ſaying: Lord, ſaue me. \V And incontinent
\SNote{Notwithſtãding the infirmities of them that gouerne the Church,
  yet Chriſt ſuſtaineth them; and holdeth them vp, yea and by them,
  whoſoeuer they are, he vpholdeth and preſerueth his Church.}
\Sc{Iesvs} ſtretching forth his hand took hold of him, and ſaid vnto
him: O thou of litle Faith, why didſt thou doubt? \V And when
%%% o-2140
they were gone vp into the boat, the winde ceaſed. \V And they that were
in the boat, came and adored him, ſaying: Indeed thou art the Sonne of
God.

\V And hauing paſſed the water, they came into the countrie of
Geneſar. \V And when the men of that place vnderſtood of him, they ſent
into al that countrie, and brought vnto him al that were il at eaſe: \V
and they beſought him that they might touch but the
\SNote{See before,
\XRef{Chap.~9,~20.}}
hemme of his garment, and whoſoeuer did touch, were made hole.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
