%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 2314
%%% o-2128
\startChapter[
  title={Chapter 11}
  ]

\Summary{Iohn the Baptiſt in priſon alſo doing his diligence, ſendeth
  ſome of his Diſciples to Chriſt: that as they heard, ſo they might
  alſo ſee his miracles with their eyes. 7.~Afterward Chriſt declareth
  how worthy of credit Iohns teſtimonie was: 16.~& inueigheth againſt
  the Iewes, who with neither of their manners of life could be wonne:
  20.~no nor with Chriſtes infinite miracles: 25.~praying Gods wiſedom
  in this behalfe, 27.~and calling to himſelf al ſuch as feele their
  owne burdens.}

%%% 2315
And it came to paſſe: when \Sc{Iesvs} had done commanding his twelue
Diſciples, he paſſed from thence, to teach & preach in their cities.

\V
\CNote{\XRef{Luc.~7,~18.}}
And when Iohn had heard in priſon the workes of Chriſt; ſending two
of his diſciples, he ſaid to him: \V
\LNote{Art thou he}{Iohn himſelf doubted not, for he baptized him and
gaue great teſtimonie of him before:
\XRef{Io.~1.}
But becauſe his
Diſciples knew him not, nor eſteemed of him ſo much as of Iohn their
owne Maiſter, therfore did he ſend them vnto Chriſt, that by occaſion
of Chriſts anſwer he might the better inſtruct them what he was, & ſo
make them Chriſts Diſciples, preferring them to a better Maiſter.}
Art thou he that art to come, or looke we for an other? \V And
\Sc{Iesvs} making anſwer ſaid to them: Goe and report to Iohn what you
haue heard and ſeen. \V
\CNote{\XRef{Eſa.~35,~5.}
\XRef{61,~1.}}
The blind ſee, the lame walke, the lepers are
made cleane, the deafe heare, the dead riſe againe, to the poore the
Ghoſpel is preached: \V and bleſſed is he that ſhal not be ſcandalized
in me.

\V And when they went their way, \Sc{Iesvs} began to ſay to the
multitudes of Iohn:
\LNote{What went you out}{High
\MNote{Eremitical life.}
commendation of Iohns holines, as wel for his faſting, rough attire,
ſolitary life, and conſtancie, as for the dignitie of his function.}
What went you out
\LNote{Into the deſert}{The faithful people in al ages reſorted of
deuotion into wildernes, to ſee men of ſpecial and rare holines,
Prophets, Eremites, Anchorites &c. to haue their prayers or ghoſtly
counſel. See
\Cite{S.~Hierom. de vita Hilarionis.}}
into the deſert to ſee? a reed ſhaken with the wind? \V But what went
you out to ſee? a man clothed in ſoft garments? Behold they that are
clothed in ſoft garments, are in Kings houſes. \V But what went you out
to ſee? a Prophet? yea I tel you and more then a Prophet. \V For this is
he of whom it is written:
\CNote{\XRef{Mal.~3,~1.}}
\Emph{Behold I ſend myne angel before thy
  face, which shal prepare thy way before thee.}

\V Amen I ſay to you, there hath not riſen among the borne of women a
greater then Iohn the Baptiſt: yet he that is the 
%%% o-2129
leſſer in the Kingdom
of Heauen, is greater then he. \V And
\CNote{\XRef{Luc.~16,~16.}}
from the dayes of Iohn the Baptiſt
vntil now, the Kingdom of Heauen ſuffereth violence, and the violent
beare it away. \V For al the Prophets and the Law prophecied vnto Iohn:
\V and if you wil receaue it, he is
\CNote{\XRef{Mal.~4,~5.}}
\LNote{Elias}{As 
\MNote{Elias.}
Elias ſhal be the meſſenger of Chriſts latter comming, ſo was Iohn his
meſſenger and Precurſour at his former cõming: & therfore is he called
Elias; becauſe of his like office and like ſpirit. 
\XRef{Luc.~1.}
\Cite{Grego. ho.~7. in Euang.}}
Elias that is for to come. \V He that hath eares to heare, let him
heare.

\V And
\CNote{\XRef{Luc.~7,~31.}}
wherevnto ſhal I eſteeme this Generation to be like? It is like
to children ſitting in the market place, \V which crying to their
companions, ſay: we haue piped to you, and you haue not danced: we haue
lamented, and you haue not mourned. \V For
\CNote{\XRef{Mt.~3,~4.}}
Iohn came neither
\LNote{Eating and drinking}{The wicked quarrelers of the world
  miſconſtre eaſely al the actes and life of good men. If they be great
  faſters and auſtere liuers, they are blaſphemed & counted hypocrites:
  if they conuerſe with other men in ordinary manner, then they be
  counted diſſolute.}
eating nor drinking; & they ſay: He hath a Diuel. \V The Sonne of man
came eating and drinking, and they ſay: Behold a man that is a glotton
and a winedrinker, a frend of Publicans and ſinners. And wiſedom is
iuſtified of her children.

\V Then
\CNote{\XRef{Luc.~10,~13.}}
began he to vpbraid the cities, wherin were done the moſt of his
miracles, for that they had not done pennance. \V Woe be to thee
Corozain, woe be to thee Beth-ſaida: for if in Tyre & Sidon had been
wrought the miracles that haue been wrought in you, they had done
\LNote{Penance in ſackcloth}{By
\MNote{Pennance.}
this ſackcloth and aſhes added here, & in other places, wee ſee
euidently that Penance is not only leauing of former ſinnes, and change
or amendment of life paſt, no nor bare ſorowfulnes or recounting of our
offences already committed, but requireth puniſhment and chaſtiſement of
our Perſons by theſe and ſuch other meanes, as the Scriptures do els
where ſet forth. And therfore concerning the word alſo, it is rather to
be called Pennance, as in our tranſlation; then (as the Aduerſaries, of
purpoſe auoyding the word) Repentance or Amendment of life: & that
according to the very vſual ſignification of the 
\TNote{\G{Μετάνοια}}
Greeke word in the moſt ancient Eccleſiaſtical Greeke writers: who for
\Emph{Penitents} (which in the primitiue  Church did publike Pennance)
ſay
\CNote{\Cite{Dionyſ. Eccl. hier. c.~3. in initio.}}
\G{οι ἐν μετανοια οντες} that is, \Emph{Men that are doing
  Pennance}. And concerning 
that part of Pennance which is Confeſſion, the Eccleſiaſtical hiſtorie
calleth it by the ſame Greeke word, and the Penitents comming to
confeſſion, \G{τοὺσ μετανοοῦντας}. 
\Cite{Sozam. li.~76.~16.}
\Cite{Socrat. li.~5. c.~19.}}
pennance in heairecloth & aſhes long agoe. \V But neuertheleſſe, I ſay
to you, it ſhal be more tolerable for Tyre and Sidon in the day of
\Fix{iugdement,}{iudgement,}{obvious typo, fixed in other}
then for you. \V And thou Capharnaum, ſhalt thou be exalted vp to
Heauen? thou ſhalt come downe euen vnto Hel. For if in Sodom had been
wrought the miracles that haue been wrought in thee, perhaps it had
remained vnto this day. \V But notwithſtanding I ſay to you, that it
ſhal be more tolerable for the land of Sodom in the day of iudgement,
then for thee.

\V At that time \Sc{Iesvs} anſwered and ſaid:
\CNote{\XRef{Lu.~10,~21.}}
I confeſſe to thee O Father 
%%% 2316
Lord of Heauen and earth, becauſe thou haſt hid theſe things from the
wiſe and prudent, and haſt reuealed them to
\LNote{Litle ones}{Theſe litle ones do not ſignifie here only the
  vnlearned, as though Coblers, and weauers, and women, & girles had
  this reuelation, & therfore do vnderſtand al Scriptures and are able
  to expound them: but here are ſignified the humble, whether they be
  learned or vnlearned: as when he ſayth,
\CNote{\XRef{Mt.~18,~3.}}
\Emph{vnles you become as
    litle ones, you shal not enter into the Kingdom of Heauen.} And ſo
  alſo the greateſt Doctours (who as they were moſt learned, ſo moſt
  humbled themſelues to the iudgement of the Catholike Church) are theſe
litle ones: and Heretikes, who although vnlearned, yet vant their
knowledge & their ſpirit of vnderſtanding aboue al ancient Fathers and
the whole Church, cannot be of theſe litle and humble ones.}
litle ones. \V Yea Father; for ſo hath it wel pleaſed thee. \V Al things
are deliuered me of my Father. And no man knoweth the Sonne but the Father:
neither doth any know the Father, but the Sonne, and to whom it shal
pleaſe the Sonne to reueale. \V Come ye to me al that labour, and are
burdened, and I wil refresh you. \V Take vp my yoke vpon you, and learne
of me, becauſe I am meeke, and humble of hart: and you shal find reſt to
your ſoules. \V For my
\LNote{Yoke ſweet}{What
\MNote{The commandements poſſible.}
is this light burden and ſweet Yoke, but his commandements, of which
S.~Iohn ſaith,
\XRef{1.~Ep.~5.}
\Emph{His commandments are not heauy}?
cleane contrary to the Aduerſaries that ſay, they are vnpoſſible to be
kept.}
yoke is ſweet, and my burden light.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
