%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 2337
%%% o-2153
\startChapter[
  title={Chapter 19}
  ]

\Summary{He 
\MNote{The fourth part of this Ghoſpel, Chriſts comming into Iurie
  toward his Paſſion.}
anſwereth the tempting Phariſees, that the caſe of a man with his wife
shal be (as in the firſt inſtitution it was) vtterly indiſſoluble,
though for one cauſe he may be diuorced. 10.~And thervpon to his
Diſciples he highly cõmendeth ſingle life for Heauen. 13.~He wil haue
children come vnto him. 16.~He sheweth what is to be done to enter into
life euerlaſting; 20.~what alſo, for a rich man to be perfect; 27.~As
alſo what paſsing reward they shal haue which folow that his counſel of
perfection: 25.~yea though it be but in ſome one peece.}

And
\CNote{\XRef{Mr.~10,~1.}}
it came to paſſe, when \Sc{Iesvs} had ended theſe wordes, he
departed from Galilee, and came into the coaſtes of Iurie beyond
Iordan, \V and great multitudes folowed him; and he cured them there.

\V And there came to him the Phariſees tempting him, and ſaying: Is it
lawful for a man to dimiſſe his wife for euery cauſe? \V Who anſwering,
ſaid to them: Haue ye not read, that he which did
\Var{make}{make man}
from the beginning,
\CNote{\XRef{Gen.~1,~27.}}
\Emph{made them male & femal}? And he ſaid: \V
\CNote{\XRef{Gen.~2,~24.}}
\Emph{For this cauſe, man shal leaue father & mother, & shal cleaue to
  his wife,
%%% 2338
and they two shal be in one flesh.} \V Therfore now they are not two,
but one fleſh. That therfore which God hath ioyned togeather, let
\LNote{Not man ſeparate}{This inſeparability betwixt man and wife
  riſeth of that, that wedlock is a Sacrament.
\Cite{Aug. li.~2. de pec. origine c.~34. to.~7.}
\Cite{De nupt. & concupiſ. li.~1. c.~10.}}
not man ſeparate. \V They ſay to him: Why then
\CNote{\XRef{Deut.~24,~1.}}
did Moyſes command to
giue a bil of diuorce, and to dimiſſe her? \V He ſaith to them: Becauſe
Moyſes for the hardnes of your hart permitted you to dimiſſe your wiues:
%%% o-2154
but from the beginning it was not ſo. \V And I ſay to you, that
\CNote{\XRef{Mt.~5,~32.}
\XRef{Mr.~10,~11.}
\XRef{Luc.~16,~18.}
\XRef{1.~Cor.~7,~11.}}
whoſoeuer ſhal dimiſſe his wife,
\LNote{But fornication}{For
\MNote{Mariage after diuorce vnlawful.}
aduoutrie one may dimiſſe an other.
\XRef{Mat.~5.}
But neither party can mary againe for any cauſe during life. 
\Cite{Aug. li.~21. de adult. coniug. c.~21. 22.~24.}
For the which vnlawful act of marying againe, Fabiola that noble matrone
of Rome albeit she was the Innocent part, did publike pennance, as
\CNote{\Cite{In Epitaph. Fabiola.}}
S.~Hierom writeth in her high commendation therfore. And in 
\XRef{S.~Paul Ro.~7.}
it is plaine that she which is with an other man,
her husband yet liuing, shal be called an aduoutreſſe: contrary to the
doctrine of our Aduerſaries.}
but for fornication, and ſhal mary an other, doth commit aduoutrie: and
he that ſhal mary her that is dimiſſed, committeth aduoutrie. \V His
Diſciples ſay vnto him: If the caſe of a man with his wife be ſo, it is
not expedient to mary. \V Who ſaid to them:
\LNote{Not al take}{Whoſoeuer haue not this guift giuen them it
\Fix{}{is}{missing word, present in other}
  either for that they wil not haue it, or for that they fulfil not that
which they wil; & they that haue this guift or attaine to this word,
haue it of God and their owne free wil. 
\Cite{Aug. li. de grat. & lib. arbit c.~4.}
So that it is euident no man is excluded from this guift, but (as
\CNote{\Cite{Origen. tract.~7. in Mat.}}
Origen here ſaith) it is giuen to al that aſke for it; contrarie to our
Aduerſaries that ſay it is impoſſible, & that for excuſe of breaking
their vowes, wickedly ſay, they haue not the guift.}
Not al 
\TNote{\G{χωροῦσι}. capiunt.}
take this word, but they to whom it is giuen. \V For there are Eunuches
which were borne ſo from their mothers wombe: and there are Eunuchs
which were made by men: and there are Eunuches, which haue
\LNote{Gelded them ſelues}{They
\MNote{Vow of chaſtitie.}
geld themſelues for the Kingdom of Heauen which vow chaſtity. 
\Cite{Aug. de virginitate c.~24.}
Which proueth thoſe kind of
vowes to be both lawful, and alſo meritorious, and more ſure to obtaine
life euerlaſting, then the ſtate of wedlock, contrarie to our Adu. in al
reſpects.}
gelded thẽſelues for the Kingdom of Heauẽ.
\LNote{He that can}{It
\MNote{Counſels not Precepts.}
is not ſaid of the Precepts, keepe them who can, for they be neceſſarie
vnder paine of damnation to be kept; but
\CNote{\Cite{Aug. ſer.~6. de temp.}}
of Counſels only (as of
virginity, abſtaining from fleſh and wine, of giuing al a mans goods
away to the poore) it is ſaid: He that can attaine to it, let him doe
it; which is counſel only, not a commandement. Contrary to our Adu.
that ſay there are no Counſels, but only precepts.}
He that can take, let him take.

\V Then
\CNote{\XRef{Mr.~10,~13.}
\XRef{Luc.~18,~15.}}
were litle children preſented to him, that he ſhould
\LNote{Impoſe}{They
\MNote{Bishops and Religious mẽs bleſſing.}
knew the valour of Chriſts bleſſing, and therfore brought their children
to him; as good Chriſtian people haue at al times brought their children
to Bishops to haue their bleſſing. See
\XRef{Annotation before Chap.~10.~12.}
And of Religious mens bleſſings ſee
\Cite{Ruffin. li.~2. c.~8. hiſt.}
\Cite{s.~Hierom in Epitaph. Paula c.~7,} &
\Cite{in vit. Hilarionis.}
\Cite{Theodoret. in hiſtoria Sanctorum Patrum num.~8.}}
impoſe hands vpon them & pray. And the Diſciples rebuked them. \V But
\Sc{Iesvs} ſaid to them: Suffer the litle children, and ſtay them not
from comming vnto me: for the Kingdom of Heauen is for ſuch. \V And when
he had impoſed hands vpon them, he departed from thence.

\V And
\CNote{\XRef{Mr.~10,~17.}
\XRef{Luc.~18,~18.}}
behold one came and ſaid to him: Good Maiſter, what good ſhal I
doe that I may haue life euerlaſting? \V Who ſaid to him: what askeſt
thou me of good? One is good, God. But
\SNote{I ſee not (ſaith S.~Auguſtine) why Chriſt ſhould ſay, If thou wilt
haue life euerlaſting, keep the commandements, if without obſeruing of thẽ,
by only faith one might be ſaued. 
\Cite{Aug. de Fid. & Op. c.~15.}}
if thou wilt enter into life keep the cõmandements. \V He ſaith to him,
which? And \Sc{Iesvs} ſaid:
\CNote{\XRef{Exo.~20,~13.}}
\Emph{Thou shalt not murder, Thou shalt not
  cõmit aduoutrie, Thou shalt not ſteale, Thou shalt not beare falſe
  witnes.} \V \Emph{Honour thy father & thy mother,}
\CNote{\XRef{Leu.~19,~18.}}
\Emph{Thou shalt loue thy
  neighbour as thyſelf.} \V The yong man ſaith to him: Al theſe haue I kept
frõ my youth; what is yet wãting vnto me? \V \Sc{Iesvs} ſaid to him:
\LNote{If thou wilt be perfect}{Loe, he maketh a plaine difference
  between keeping the commandements, which is neceſſary for euery man:
  and being perfect, which he counſeleth only to them that wil. And this
is the ſtate of great perfection which Religious men doe profeſſe,
according to Chriſtes counſel here, leauing al things and folowing him.}
If thou wilt be perfect, goe, ſel the things that thou haſt, & giue to
the poore, and thou ſhalt haue treaſure in Heauen: and come,
\LNote{Folow me}{Thus
\MNote{The Religious ſtate of perfection.}
to folow Chriſt is to be without wife and care of children, to lack
proprietie, and to liue in common, and this hath great reward in Heauen
aboue other ſtates of life: which S.~Auguſtine ſaith, the Apoſtles
folowed, and himſelf, and that he exhorted others to it as much as lay
in him.
\Cite{Aug. ep.~89. in fine,} &
\Cite{in ps.~103. Conc. 3. poſt med.}}
folow me. \V And when the yong man had heard this word, he went away
ſad: for he had many poſſeſſions. \V And \Sc{Iesvs} ſaid to his
Diſciples: Amen I ſay to you, that a rich man ſhal hardly enter into the
Kingdom of Heauen. \V And againe I ſay to you, it is eaſier for a camel
to paſſe through the eye of a needle, 
\SNote{S.~Marke expoundeth it thus, rich men truſting in their riches. 
\XRef{c.~10,~24.}}
then for a rich man to enter into the Kingdom of Heauen. \V And when
they had heard this, the Diſciples marueled very much, ſaying: who then
can be ſaued? \V And \Sc{Iesvs} beholding, ſaid to them: With men this
is impoſſible: but with God 
\LNote{All things poſsible}{This of the camel through a needels eye,
being poſſible to God, although he neither hath done it, nor by like wil
doe it: maketh againſt the blaſphemous infidelitie of our Aduerſaries
that ſay, God can doe no more then he hath done, or wil doe. We ſee alſo
that God cã bring a camel through a needels eye, & therfore his body
through a doore, and out of the ſepulchre shut, and out of his mother a
virgin, and generally aboue nature doe with his body as he liſt.}
al things are poſſible. \V Then Peter anſwering, ſaid to him: Behold we
haue
\LNote{Left al}{This perfection of leauing al things the Apoſtles
  vowed. 
\Cite{Aug. li.~17. de Ciuit. Dei. c.~4.}}
left al things, & haue folowed thee:
\LNote{What shal we haue}{They
\MNote{Vow of pouertie in reſpect of reward.}
leaue al things in reſpect of reward, and Chriſt doeth wel allow it in
them by his anſwer.}
what therfore ſhal we haue? \V And \Sc{Iesvs} ſaid to them: 
%%% o-2155
Amen I ſay
to you, that you which haue folowed me, in the regeneration, when the
Sonne of man ſhal ſit in the ſeat of his maieſtie,
\LNote{You alſo shal ſit}{Note that not only Chriſt, who is the
  principal and proper Iudge of the liuing and the dead, but with him
  the Apoſtles and al perfect Saints shal iudge; and
\CNote{\Cite{Aug. in Pſa.~121.}}
yet that doeth
  nothing derogate to his prerogatiue, by whom and vnder whom they hold
  this and al other dignities in this life and the next.}
you alſo ſhal ſit vpon twelue ſeats, iudging the twelue Tribes of Iſrael. \V
And euery one that hath left houſe, or brethren, or ſiſters, 
\Fix{of}{or}{obvious typo, fixed in other}
Father, or mother, or
\SNote{Hereof is gathered that the Apoſtles among other things, left
  their wiues alſo to folow Chriſt. 
\Cite{Hier. li.~1. aduerſ. Iouin.}}
wife, or children, or landes for my names ſake: ſhal receaue an
hundred-fold, & ſhal poſſeſſe life euerlaſting. \V And
\CNote{\XRef{Mr.~10,~31.}
\XRef{Lu.~13,~30.}}
many ſhal be firſt, that are laſt; and laſt, that are firſt.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
