%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 2340
%%% o-2156
\startChapter[
  title={Chapter 20}
  ]

\Summary{To shew how through Gods grace the Iewes shal be ouerrunne of
  the Gentils, although they beginne after, he bringeth a parable of men
working ſooner and later in the vinyard, but the later rewarded in the
end euen as the firſt. 17.~He reuealeth more to his Diſciples touching
his paſsion: 20.~Bidding the ambitious two ſuiters to thinke rather of
ſuffering with him: 24.~And teaching vs (in the reſt of his Diſciples)
not to ge greeued at our Eccleſiaſtical Superiours, conſidering they are
(as he was himſelf) to toile for our ſaluation. 29.~Then going out of
Iericho, he giueth ſight vnto two blind.}

The Kingdom of Heauen is like to a man that is an Houſholder which went
forth early
\LNote{In the morning}{God calleth ſome in the morning, that is, in the
  beginning of the world, as Abel, Enoch, Noe, and other the iuſt and
  faithful of the firſt Age; at the 
\Fix{thrid}{third}{obvious typo, fixed in other.}
houre, Abraham, Iſaac, and Iacob, and the reſt of their Age; at the 6.\
houre of the day, Moyſes, Aaron, and the reſt; at the 9.\ houre, the
Prophets; at the eleuenth, that is, at the later end of the world, the
Chriſtian Nations.
\Cite{Aug. de verb. Domini ſer.~59.}
Briefly, this calling at
diuerſe houres ſignifieth the calling of the Iewes from time to time in
the firſt Ages of the world, and of the Gentils in the later Age
thereof. It ſignifieth alſo that God calleth countries to the faith 
ſome ſooner, ſome
later; and particular men to be his ſeruants, ſome yonger, ſome elder,
of diuerſe ages.}
in the morning to hire workemen into his vineyard. \V And hauing made
couenant with the workemen for a penie a day, he ſent them into his
vinyard. \V And going forth about the third houre, he ſaw other ſtãding
in the market place idle, \V and he ſaid to them: Goe you alſo into the
vineyard, and that which ſhal be iuſt, I wil giue you. \V And they went
their way. And againe he went forth about the ſixt & the ninth houre:
and did likewiſe. \V But about the eleuenth houre he went forth and
found other ſtanding, & he ſaith to them: What ſtand you here al the day
idle? \V They ſay to him: Becauſe no man hath hired vs. He ſaith to
them: Goe you alſo into the vineyard.

\V And when euening was come, the Lord of the vineyard ſaith to his
Bailife: Cal the workmen, and pay them their hire, beginning from the
laſt euen to the firſt. \V Therfore when they were come that came about
the eleuenth houre, they receaued euerie one
\LNote{Penie}{The
\MNote{Diuerſitie of glorie in Heauen.}
penie promiſed to al, was life euerlaſting, which is common to al that
ſhal be ſaued. But in the ſame life there be degrees of glorie, as
\CNote{\XRef{1.~Cor.~15.}}
betwixt ſtarre and ſtarre in the element.
\Cite{Aug. li. de virginis. c.~26.}}
a penie. \V But when the firſt alſo came, they thought that they ſhould
receaue more: and they alſo receaued euerie one a penie. \V And
receauing it they
\SNote{The Iewes are noted for enuying the vocatiõ of the Gentils, and
  their reward, equal with thẽ ſelues.}
murmured againſt the Good-man of the houſe, \V ſaying: Theſe laſt haue
continued one houre, and thou haſt made them equal to vs that haue borne
the burdẽ of the day & the heats. \V But he anſwering ſaid to one of thẽ
Freind, I do thee no wrong: didſt thou not couenãt with me for a penie?
\V Take that is thine, and goe: I wil alſo giue to his laſt euen as to
thee alſo. \V Or is it not lawful for me to doe that I wil? is thine eye
naught, becauſe I am good? \V So shal the laſt be firſt, and the firſt,
laſt. For many be called, but
\LNote{Few elect}{Thoſe are elect which deſpiſed not their caller, but
  folowed and beleeued him; for men beleeue not but of their owne free
  wil.
\Cite{Aug. li.~1. ad Simplic. q.~2.}}
few elect.

\V
\CNote{\XRef{Mr.~10,~32.}
\XRef{Luc.~18,~31.}}
And \Sc{Iesvs} going vp to Hieruſalem, tooke the twelue Diſciples
ſecretly, and ſaid to them: \V Behold we goe vp to
%%% o-2157
Hieruſalem, and the Sonne of man ſhal be deliuered to the chiefe Prieſts
and to the Scribes, and they ſhal condemne him to death, \V and ſhal
deliuer him to the Gentils to be mocked, and ſcourged, and crucified,
and the third day he ſhal riſe againe.

\V
\CNote{\XRef{Mr.~10,~35.}
\XRef{Luc.~22,~35.}}
Then came to him the mother of the ſonnes of Zebedee with her
%%% 2341
ſonnes, adoring and deſiring ſome thing of him. \V Who ſaid to her: What
wilt thou? She ſaith to him: Say that theſe my two ſonnes may ſit, one
at thy right hand, & one at thy left hand in the Kingdõ. \V And
\Sc{Iesvs} anſwering, ſaid: You know not what you deſire. Can you drinke
of the cup that I ſhal drinke of? They ſay to him: We can. \V He ſaith
to them: My cup indeed you ſhal drinke of: but to ſit at my right hand
and left, is not mine to giue to you, but
\LNote{To whom it is prepared}{The Kingdom of Heauen is prepared for
  them that are worthie of it and deſerue it by their wel doing; as in
  holy Scripture it is very often: That
\CNote{\XRef{Mat.~16,~27.}}
\Emph{God wil repay euery man according to his workes.} And,
\CNote{\XRef{Ro.~2,~6.}}
\Emph{Come ye bleſſed, poſſeſſe the Kingdom prepared for you.} Why?
\CNote{\XRef{Mt.~25,~34.}}
\Emph{Becauſe I was hungrie, and you gaue me meate; thirſtie, and you
  gaue me drinke &c.} Therfore doth
  Chriſt ſay here: \Emph{It is not mine to giue.} Becauſe he is iuſt and
  wil not giue it to euerie man without reſpect of their deſerts; 
\MNote{Difference of merits and reward.}
yea
  nor alike to euerie one, but diuerſly according to greater or leſſer
  merits; as here
\CNote{\Cite{Chryſ. ho.~66. in Mt.}}
S.~Chryſoſto. maketh it plaine, when our Sauiour
  telleth them, that although they ſuffer martyrdom for his ſake, yet he
  hath not to giue thẽ the two cheefe places. See
\Cite{S.~Hier. vpon this place,} and
\Cite{li.~2. adu. Iouin. c.~15.}
This alſo is a leſſon for
them that haue to beſtow Eccleſiaſtical benefices, that they haue no
carnal reſpect to kindred &c.\ but to the worthines of the perſons.}
to whom it is prepared of my Father.

\V And the ten hearing it, were diſpleaſed at the two brethren. \V And
\Sc{Iesvs} calleth them vnto him, and ſaid:
\CNote{\XRef{Mr.~10,~41.}
\XRef{Luc.~22,~25.}}
You know that the Princes of the Gentils
\SNote{Superioritie is not here forbidden among Chriſtiãs, neither
  Eccleſiaſtical nor temporal; but heathenish tyrannie is forbidden, and
humilitie commended.}
ouerrule them: and they that are the greater, exerciſe power againſt
them. \V It ſhal not be ſo among you, but whoſoeuer wil be the greater
among you, let him be your miniſter: \V And he that wil be firſt among
you, ſhal be your ſeruant. \V Euen 
\LNote{As the Sonne of man}{Chriſt 
\MNote{Purianes.}
himſelf as he was the Sonne of man,
  was their and our Superiour, and
\CNote{\XRef{Io.~13,~13.}}
Lord & Maiſter, notwithſtanding his
  humilitie, and therfore it is pride and haughtines which is forbidden,
and not Superioritie or Lordſhip, as ſome Heretikes would haue it.}
as the Sonne of man is not come to be miniſtred vnto, but to miniſter, and to
giue his life a redemption for many.

\V And
\CNote{\XRef{Mr.~10,~46.}}
when they went out from Iericho, a great multitude folowed
him. \V And behold two blind men ſitting by the way ſide, heard that
\Sc{Iesvs} paſſed by, and they cried out ſaying: Lord, haue mercie vpon
vs, Sonne of Dauid. \V And the multitude rebuked them that they ſhould
hold their peace. But they cried out the more, ſaying: Lord, haue mercie
vpon vs, Sonne of Dauid. \V And \Sc{Iesvs} ſtood, and called them, and
ſaid: What wil ye that I doe to you? \V They ſay to him: Lord, that our
eyes may be opened. \V And \Sc{Iesvs} hauing compaſſion on them, 
\SNote{Our Sauiour who alwaies could, and often did cure the diſeaſed,
  by his only wil, or word, here gaue ſight to theſe blind men, by
  touching their eyes with his holie hand.}
touched their eyes. And immediatly they ſaw, and folowed him.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
