%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 2305
%%% o-2117
\startChapter[
  title={Chapter 7}
  ]

\Summary{In his third and laſt Chapter of his Sermõ, becauſe we know not
mens endes, he biddeth vs beware of iudging: 6.~and neuer the leſſe to
take open dogs and ſwine (ſo he calleth them) as they be. 7.~If theſe
workes of iuſtice ſeeme too hard, we muſt pray inſtantly to him that
giueth them. 12.~In the concluſion, he giueth one short rule of al
iuſtice. 13.~and then he exhorteth with al vehemencie to the ſtraite
way, both of the Catholike faith, 21.~and alſo of good life; becauſe
only faith wil not ſuffice.}

Iudge 
\LNote{Iudge not}{It is no Chriſtian part to iudge il of mens actes,
  which be in them ſelues good, and may proceed of good meaning, or of
  mans inward meanings, and intentions, which we can not ſee, of which
  fault they muſt beware, that are too ſuſpicious, and giuen to deeme
  alwayes the worſt of other men. But to ſay, that Iudas, or an Heretike
euidently known to die obſtinatly in hereſie, is damned, and in al other
playne and manifeſt caſes, to iudge, is not forbidden.}
not,that you be not iudged. \V For
\CNote{\XRef{Luc.~6,~37.}
\XRef{Mr.~4,~24.}}
in what iudgement you iudge, you ſhal be 
iudged: and in what meaſure you mete, it ſhal be meaſured to you
againe. \V And why ſeeſt thou the mote that is in thy brothers eye: and
the beame that is in thine owne eye thou ſeeſt not? \V Or how ſayeſt
thou to thy brother: Let me caſt out the mote of thine eye; and behold a
beame is in thine owne eye? \V Hypocrite, caſt out firſt the beame out
of thyne owne eye; and then shalt thou ſee to caſt out the mote out of
thy brothers eye.

\V Giue not that which is
\LNote{Holy to dogs}{No
\MNote{Worthy receauing.}
holy Sacrament, and ſpecially that of our Sauiours Bleſſed Body muſt be
giuen wittingly to the vnworthie, that is, to them that haue not by
confeſſion of al mortal ſinnes, examined and proued themſelues. See the
\XRef{Annot. 1.~Cor.~11,\ 27. 28.~29.}}
holy to dogs: neither caſt ye your pearles before ſwine, leſt perhaps
they treade them with their feete, & turning, al to teare you.

\V
\CNote{\XRef{Lu.~11,~9.}}
Aske, and it ſhal be giuen you: ſeek, and you ſhal finde, knock, & it
ſhal be opened to you. \V For
\LNote{Euery one that asketh}{Al things that we aske neceſſarie to
  ſaluation, with humilitie, attention, continuance, and other due
  circumſtances, God wil vndoubtedly grant when it is beſt for vs.}
euery one that asketh, receiueth: and that ſeeketh, findeth: and to him
%%% o-2118
that knocketh, it ſhal be opened. \V Or what man is there of you, whom
if his childe ſhal aske bread, wil he reach him a ſtone? \V Or if he
ſhal aske him fish, wil he reach him a ſerpent? \V If you then being
naught, know how to giue good guifts to your children: how much more wil
your Father which is in Heauen, giue
\SNote{Theſe good things are grace and al ſpiritual guifts, and
  whatſoeuer pertayneth to the health of the ſoule.}
good things to them that aske him?

\V
\CNote{\XRef{Luc.~6,~31.}}
Al things therfore whatſoeuer you wil that men doe to you, doe you
alſo to them. For this is the Law and the Prophets.

\V
\CNote{\XRef{Luc.~13,~24.}}
Enter ye by the narrow gate: becauſe brode is the gate, and large is
the way that leadeth to perdition, and many there be that enter by
it. \V How narrow is the gate, and ſtrait is the way, that leadeth to
life: &
%%% 2306
few there are that find it!

\V Take ye great heed of falſe Prophets, which come to you in the
\LNote{Clothing of sheep}{Extraordinarie
\MNote{Heretikes wolues in ſheep ſkinnes.}
apparance of zeale, and holines, is the ſheeps cote, in ſome Heretikes:
but theſe of this time weare not that garment much, being men of
vnſatiable ſinne. This is rather their garment, common to them with al
other Heretikes, to crak much of the word of the Lord, and by pretenſed
allegations, &
\CNote{\XRef{Rom.~16,~18.}}
ſweet words of benediction, and ſpecially by promiſe of
knowledge, light, and libertie of the Ghoſpel, to ſeduce the ſimple and
the ſinful.}
clothing of sheep, but inwardly are rauening wolues. \V
\LNote{Fruits}{Theſe 
\MNote{Heretikes known by their fruits.}
are the fruits which Heretikes are knowen
  by, diuiſion from the whole Church, diuiſion among themſelues, taking
  to themſelues new names and new maiſters, inconſtancie in doctrine,
  diſobedience both to others and namely to ſpiritual officers, loue and
liking of themſelues, pride and intolerable vanting of their owne
knowledge aboue al the holy Doctours, corruption, falſification and
quite denying of the parts of Scriptures that ſpecially make agaynſt
them, and theſe be common to al Heretikes lightly. Otherſome are more
peculiar to theſe of our time, as Inceſtuous mariages of vowed perſons,
Spoile of Churches, Sacrilege and profanation of al holy things, and
many other ſpecial points of doctrine, directly tending to the corruption
of good life in al ſtates.}
By their
\CNote{\XRef{Luc.~6,~44.}}
fruits you shal know thẽ. Do men gather grapes of thornes, or
figs of thiſtels? \V Euen ſo euery good tree yealdeth good fruits, and
the euil tree yealdeth euil fruits. \V A good tree can not yeald euil
fruits, neither an euil tree yeald good fruits. \V Euery tree that
yealdeth not good fruit, shal be cut downe, and shal be caſt into
fire. \V Therfore by their fruits you shal know them.

\V Not euery one that ſayth to me,
\LNote{Lord Lord}{Theſe 
\MNote{Not only faith.}
men haue faith, otherwiſe they could
not inuocate, \Emph{Lord, Lord}
\XRef{Ro.~10.}
But here we ſee that to
beleeue is not enough, and that not only infidelitie is ſinne, as
Luther teacheth. Yea Catholikes alſo that worke true miracles in the
name of our Lord, and by neuer ſo great faith, yet without the workes
of iuſtice ſhal not be ſaued.
\XRef{1.~Cor.~13.}
Againe conſider here
who they are that haue ſo often in their mouth, \Emph{The Lord, the
Lord}, and how litle it ſhal auaile them, that ſet ſo litle by good
workes, and contemne Chriſtian iuſtice.}
\Emph{Lord, Lord}, shal enter into the Kingdom of Heauen: but he that
doth the wil of my Father which is in Heauen, he shal enter into the
Kingdom of Heauen. \V Many shal ſay to me in that day: \Emph{Lord,
  Lord}, haue not we prophecied in thy name, and in thy name caſt out
Diuels, and in thy name wrought many miracles? \V And then I wil
confeſſe vnto them, That I neuer knew you: depart from me you that worke
iniquitie. \V
\CNote{\XRef{Luc.~6,~47.}}
Euery one therfore that heareth theſe my words, and doth
them, shal be likned
to a wiſe man that built his houſe vpon a rock, \V and the rayne fel,
and the fluddes came, and the windes blew, and they beat agaynſt that
houſe, and it fel not, for it was founded vpon a rock. \V And euery one
that heareth theſe my words, and doth them not, shal be like a foolish
man that built his houſe vpon the ſand, \V and the rayne fel, and the
fluddes came, and the windes blew, and they beat againſt that houſe, and
it fel, & the fal therof was great.

\V And it came to paſſe, when \Sc{Iesvs} had fully ended theſe wordes,
the multitude were in
\CNote{\XRef{Mr.~1,~22.}
\XRef{Luc.~4,~32.}}
admiration vpon his 
%%% o-2119
doctrine. \V For he was teaching them as hauing power, and not as their
Scribes and Phariſees.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
