%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-08


\project douay-rheims


%%% 2307
%%% o-2119
\startChapter[
  title={Chapter 8}
  ]

\Summary{Immediatly after his Sermon (to confirme his doctrine with a
  miracle) he cureth a Leper. 5.~But aboue him and al other Iewes, he
  comendeth the faith of the Centurion, who was a Gentil, and
  foretelleth by that occaſion, the vocation of the Gentiles, and
  reprobation of the Iewes. 14.~In Peters houſe he sheweth great
  grace. 18.~In the way to the ſea he ſpeaketh with two, of folowing
  him: 23.~and vpon the ſea commandeth the tempeſt: 28.~and beyond the
  ſea he manifeſteth the Deuils malice againſt man, in an heard of
  ſwine.}

And when he was come downe from the mountaine, great multitudes folowed
him: \V And
\CNote{\XRef{Mr.~1,~40.}
\XRef{Lu.~5,~12.}}
behold a Leper came and adored him ſaying: Lord, if thou
wilt, thou canſt make me cleane. \V And \Sc{Iesvs} ſtretching forth his
hand, touched him ſaying: I wil. Be thou made cleane. And forthwith, his
leproſie was made cleane. \V And \Sc{Iesvs} ſaith to 
%%% o-2120
him: See thou tel no body: but go,
\CNote{\XRef{Leu.~14,~2.}}
ſhew thyſelf to the
\LNote{Prieſt}{The Prieſts of the old law (ſaith S.~Chriſoſtom) had
  authoritie and priuilege only to diſcerne who were healed of leproſie,
and to denounce the ſame to the people: but the Prieſts of the new law,
haue power to purge in very deed, the filth of the ſoule. Therfore
whoſoeuer deſpiſeth them, is more worthie to be puniſhed, then the rebel
Dathan, and his complices.
\Cite{S.~Chryſo. li.~3. de Sacerd.}}
Prieſt, and offer the
\LNote{Guift}{Our Sauiour willeth him to goe, and offer his guift or
  ſacrifice, according as Moyſes preſcribed in that caſe, becauſe the
  other ſacrifice, being the holieſt of al holies, which is his bodie,
  was not yet begune. So ſaith 
\Cite{S.~Aug. li.~2. q.}
\Cite{Euang. q.~3.} &
\Cite{Con. Aduer. leg.}
\Cite{Proph. li.~1. c.~19 20.}}
guift which Moyſes commanded for a teſtimonie to them.

\V And
\CNote{\XRef{Lu.~7,~11.}}
when he was entred into Capharnaum, there came to him a
Centurion, beſeeching him, \V and ſaying: Lord my boy lieth at home ſick
of the palſey, and is ſore tormented. \V And \Sc{Iesvs} ſaith to him: I
wil come, and cure him. \V And the Centurion making anſwer, ſaid: Lord,
\LNote{Not worthie}{
\Cite{Orig. ho.~5. in diuerſ.}
\TNote{\L{\Sc{Domine} nõ ſum dignus.}}
When thou eateſt (ſaith he) and drinkeſt the bodie & blood of our Lord,
he entreth vnder thy roofe. Thou alſo therfore humbling thyſelf, ſay:
Lord I am not worthie, &c. So ſaid
\CNote{\Cite{Liturg. Græc. ſub finẽ.}}
S.~Chryſoſtom in his Maſſe, and ſo
doth the Cath.\ Church vſe at this day in euery Maſſe.  See 
\Cite{S.~Auguſtine ep.~118. ad Ianu.}}
I am not worthie that thou ſhouldeſt enter vnder my roofe: but only ſay
the word, & my boy ſhal be healed. \V For I alſo am a man ſubiect to
authoritie, hauing vnder me ſouldiars; and I ſay to this, goe, and he
goeth, and to an other, come, and he cometh; and to my ſeruant, doe
this, and he doth it. \V And \Sc{Iesvs} hearing this, marueled: and ſaid
to them that folowed him: Amen I ſay to you, I haue not found ſo great
faith in Iſrael. \V And I ſay to you that manie ſhal come from the Eaſt
and Weſt, and ſhal ſit downe with Abraham, and Iſaac, & Iacob in the
Kingdom of Heauen: \V but the children of the Kingdom ſhal be caſt out
into the exteriour darkeneſſe: there shal 
\Fix{he}{be}{obvious typo, fixed in other}
weeping
%%% 2308
and gnashing of teeth. \V And \Sc{Iesvs} ſaid to the Centurion: Goe, and
as thou haſt beleeued, be it done to thee. And the boy was healed in the
ſame houre.

\V And
\CNote{\XRef{Mr.~1,~19.}
\XRef{Lu.~4,~38.}}
when \Sc{Iesvs} was come into Peters houſe, he ſaw
\LNote{His wiues mother}{Of
\MNote{Prieſts mariage.}
Peter ſpecially among the reſt, it is euident, that he had a wife, but
(as S.~Hierom ſaith) after they were called to be Apoſtles, they had no
more carnal companie with their wiues, as he proueth there by the very
words of our Sauiour:
\CNote{\Cite{Lib.~1. adu. Iou. c.~14.}
\XRef{Mt.~19,~29.}
\Cite{Epiph. ho.~59.}}
\Emph{He that hath left wife, &c.} And ſo in the
Latin Church hath been alwayes vſed, that maried men may be, and are
daily made Prieſts, either after the death of the wife, or with her
conſent, to liue in perpetual continencie. And if the Greeks haue
Prieſts that doe otherwiſe, S.~Epiphanius a Greek Doctour telleth them,
that they doe it againſt the ancient Canons; and
\CNote{\Cite{Sozom. l.~1. c.~22.}
\Cite{Socrat. li.~2. c.~8.}}
Paphnutius plainly
ſignifieth the ſame in the firſt Councel of Nice. But this is moſt
plaine, that there was neuer either in the Greek Church or the Latin,
authentical example of any that married after holy Orders.}
his wiues mother layd, and in a fit of feuer: \V and he touched her
hand, and the feuer left her, and ſhe aroſe, and miniſtred to him. \V
And when euening was come, they brought to him manie that had Diuels:
and he caſt out the Spirits with a word: and al that were il at eaſe he
cured: \V that it might be fulfilled, which was ſpoken, by Eſay the
Prophet, ſaying: \Emph{He tooke our infirmities, and bare our diſeaſes.}

\V And \Sc{Iesvs} ſeeing great multitudes about him, commanded to goe
beyond the water. \V And a
\CNote{\XRef{Eſa.~53,~4.}
\XRef{1.~Pet.~2,~24.}
\XRef{Lu.~9,~57.}}
certaine Scribe came, and ſaid to him:
Maſter, I wil folow thee whitherſoeuer thou ſhalt goe. \V And \Sc{Iesvs}
ſaith to him: The 
\Fix{floxes}{foxes}{obvious typo, fixed in other}
haue holes, and the foules of the ayre neſtes: but the Sonne of man hath
not where to lay his head. \V And
\CNote{\XRef{Luc.~9,~59.}}
an other of his Diſciples ſaid to him:
Lord, permit me firſt to goe and burie my Father. \V But \Sc{Iesvs} ſaid
to him: Folow me, and
\LNote{Let the dead}{By this we ſee that not only no 
\Fix{wordly}{worldly}{obvious typo, same in other}
or carnal reſpect, but no other laudable dutie toward our parents, ought
to ſtay vs from folowing Chriſt, and chooſing a life of greater
perfection.}
let the dead burie the dead.

\V And
\CNote{\XRef{Mar.~4,~36.}
\XRef{Lu.~8,~22.}}
when he entered into the boat, his Diſciples folowed him: \V and
loe a great tempeſt aroſe in the ſea, ſo 
%%% o-2121
that the boat was couered with
waues, but he ſlept. \V And they came to him, and raiſed him, ſaying:
Lord, ſaue vs, we periſh. \V And he ſaith to them: Why are you fearful,
O ye of litle faith? Then riſing vp
\LNote{He commanded}{The Church (here ſignified by the boate or ſhip)
  and Catholikes, are often toſſed with ſtormes of perſecution, but
  Chriſt who ſeemed to ſleepe in the meane time, by the Churches prayers
  awaketh, and maketh a calme.}
he commanded the windes, and the ſea, and there enſued a great calme. \V
Moreouer the men marueled, ſaying: What an one is this, for the windes
and the ſea obey him.

\V And
\CNote{\XRef{Mr.~5,~1.}
\XRef{Luc.~8,~26.}}
when he was come beyond the water, into the countrey of the
Geraſens, there met him two that had Diuels, coming forth out of the
ſepulcres, exceeding fierce, ſo that none could paſſe by that way. \V
And behold they cried ſaying: What is betweene vs & thee \Sc{Iesvs} the
Sonne of God? art thou come hither to torment vs before the time. \V And
there was not farre from them an heard of many ſwine feeding. \V And the
Diuels beſought him ſaying: If thou caſt vs out, ſend vs into the heard
of ſwine. \V And he ſaid to them: Goe. But they going forth went into
the ſwine, and behold the whole heard went with a violence, headlong
into the ſea: and they dyed in the waters. \V And the ſwine-heardes
fled: and comming into the citie, told al, and of them that had been
poſſeſſed of Diuels. \V And behold the whole citie went out to meete
\Sc{Iesvs}, and when they ſaw him, they beſought him that he would paſſe
from their quarters.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
