%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 2335
%%% o-2150
\startChapter[
  title={Chapter 18}
  ]

\Summary{To his Diſciples he preacheth againſt ambition the mother of
  Schiſme: 7.~foretelling both the authour whoſoeuer he be, and alſo his
folowers, of their woe to come. 10.~and shewing on the contrary ſide,
how precious Chriſtian ſoules are to their Angels, to the Sonne of man,
and to his Father. 15.~charging vs therfore to forgiue our brethren,
when alſo we haue iuſt cauſe againſt them, be it neuer ſo often, and to
labour their ſaluation by al meanes poſsible.}

At that houre
\CNote{\XRef{Mr.~9,~34.}
\XRef{Lu.~9,~46.}}
the Diſciples came to \Sc{Iesvs}, ſaying: 
\LNote{Who is the greater}{The occaſion of this
queſtion, & of their contention for Superioritie, among the reſt of
their infirmities, which they had before the comming of the Holy
Ghoſt, was (as certaine holy Doctours write) vpon emulation toward
Peter, whom only they ſaw preferred before the reſt, in the payment of
the tribute, by theſe wordes of our Sauiour:
\CNote{\XRef{C.~17. v.~27.}}
Giue it them for me and thee. 
\Cite{Chrys. ho.~59.}
\Cite{Hier. in Matth. Vpon this place.}}
Who, thinkeſt thou, is the greater in the Kingdom of Heauen? \V And
\Sc{Iesvs} calling vnto him a litle child, ſet him in the middes of
them, \V and ſaid: Amen I ſay to you, vnles you be conuerted, and become
as litle children, you ſhal not enter into the Kingdom of Heauen. \V
Whoſoeuer therfore ſhal humble himſelf as this
\SNote{Humility, innocencie, ſimplicity, cõmẽded to vs in the ſtate &
  Perſon of a child.}
litle child, he is the greater in the Kingdom of Heauen. \V And he that
ſhal receaue one ſuch litle child in my name, receaueth me. \V And
\CNote{\XRef{Mr.~9,~42.}
\XRef{Lu.~17,~2.}}
he that ſhal ſcandalize one of theſe litle 
%%% o-2151
ones that beleeue in me, it is expedient for him that a milſtone 
\Fix{he}{be}{obvious typo, fixed in other}
hanged about his neck, and that he be drowned in the depth of the ſea.

\V Woe be to the world for
\LNote{Scandals}{The ſimple be moſt annoyed by taking ſcandal of their
Preachers, Prieſts, and elders il life: and great damnation is to the
guides of the People, whether they be temporal or ſpiritual, but
ſpecially to the ſpiritual, if by their il example and ſcandalous life,
the People be ſcandalized.}
ſcandals. For it is neceſſary that ſcandals do come: but neuertheleſſe
woe to that man by whom the ſcandal commeth. \V And
\CNote{\XRef{Mt.~5,~30.}
\XRef{Mar.~9,~43.}}
if thy
\LNote{Hand, foot, eye}{By theſe parts of the body ſo neceſſarie and
profitable for a man, is ſignified, that whatſoeuer is neereſt and
deereſt to vs, wife, children, friendes, riches, al are to be
contemned and forſaken for to ſaue our ſoule.}
hand, or thy foot ſcandalize thee, cut it of, and caſt it from thee. It
is good for thee to goe into life maimed or lame, rather then hauing two
hands or two feet, to be caſt into euerlaſting fire. \V And if thine eye
ſcandalize thee, pluck him out, and caſt him from thee: It is good for
thee hauing one eye to enter into life, rather then hauing two eyes to
be caſt into the Hel of fire. \V See that you deſpiſe not one of theſe
litle ones: for I ſay to you, that
\LNote{Their Angels}{A 
\MNote{Protection of Angels.}
great dignitie, and a maruelous benefit, that
  eueryone hath from his Natiuitie an Angel for his cuſtodie and
  patronage, againſt the wicked, before the face of God. 
\Cite{Hier. vpon this place.}
And the thing is ſo plaine, that Caluin dare not deny it, and yet he wil
needes doubt of it.
\Cite{lib.~1. Inſt. c.~14. ſect.~7.}}
their Angels in Heauen alwaies do ſee the face of my Father which is in
Heauen. \V For
\CNote{\XRef{Luc.~19,~10.}}
the Sonne of man is come to ſaue that which was
periſhed. \V
\CNote{\XRef{Lu.~15,~4.}}
How thinke you? If a man haue an hundred ſheep, and one of
them ſhal goe aſtray; doth he not leaue ninetie nine in the mountaines,
and goeth to ſeek that which is ſtraied? \V And if it chance that he
find it: Amen I ſay to you, that he reioyceth more for that, then for
the ninetie nine that went not aſtray. \V Euen ſo it is not the wil of
your Father, which is in Heauen, that one periſh of theſe litle ones.

\V But
\CNote{\XRef{Luc.~17,~3.}}
if thy brother ſhal offend againſt thee, goe, and rebuke him
between thee and him alone. If he ſhal heare thee, thou ſhalt gaine thy
brother. \V And if he wil not heare thee, ioyne with thee beſides, 
%%% 2336
one or two: that in the mouth of
\CNote{\XRef{Deu.~19,~15.}}
two or three witneſſes euery word may
ſtand. \V And if he wil not heare them,
\SNote{That is (as S.~Chryſoſtõ here expoũdeth it) Tel the Prelates &
cheefe Paſtours of the Church; for they haue iuriſdiction to bind &
looſe ſuch offenders, by the wordes folowing
\XRef{v.~18.}}
tel the Church. \Emph{And if he wil}
\LNote{Not heare the Church}{Not
\MNote{Diſobedience to the Church.}
only Heretikes, but any other obſtinate offender that wil not be iudged
nor ruled by the Church, may be excommunicated, & ſo made as an Heathen
or Publican was to the Iewes, by the diſcipline of the ſame, caſting him
out of the felowſhip of Catholikes. 
\MNote{Excommunication.}
Which Excommunication is a greater
puniſhment, then if he were executed by ſword, fire, & wild beaſtes. 
\Cite{Aug. cont. Adu. leg. li.~1. c.~17.}
And againe he
ſaith: Man is more sharply & pitifully bound by the Churches Keies, then
with any iron or adamantine manicles or fetters, in the world.
\Cite{Aug. ibidem.}}
\Emph{not heare the Church, let him be to thee as}
\LNote{Heathen}{Heretikes therfore becauſe they wil not heare
the Church, be no better nor no otherwiſe to be eſteemed of Catholikes,
then Heathen men and Publicans were eſteemed among the Iewes.}
\Emph{the Heathen and Publican.} \V Amen I ſay to you, whatſoeuer 
\LNote{You shal bind}{As
\MNote{Power to bind and looſe.}
before
\CNote{\XRef{Mt.~16,~19.}}
he gaue this power of binding and looſing ouer the whole, firſt
of al and principally to Peter, vpon whom he builded his Church; ſo here
not only to Peter, and in him to his ſucceſſours, but alſo to the other
Apoſtles, & in them to their ſucceſſours, euery one in their charge. 
\Cite{Hierom lib.~1\ c.~14. aduers. Iouin.} and
\Cite{Epiſt. ad Heliod.}
\Cite{Cyprian. de vnit. Eccl. nu.~3.}}
you ſhal bind vpon earth, ſhal be bound alſo in Heauen: and whatſoeuer you
\LNote{Shal looſe}{Our Lord giueth no leſſe right and authoritie to the
  Church to looſe, then to bind, as
\CNote{\Cite{li.~1. de pœnit. c.~2.}}
S.~Ambroſe writeth againſt the
  Nouatians, who confeſſed that the Prieſts had power to bind, but not
  to looſe.}
ſhal looſe vpõ earth, ſhal be looſed alſo in Heauen. \V Againe I ſay to
you, that if two of you ſhal
\SNote{Al ioyning togeather ĩ the vnity of Chriſtes Church in Councels,
  and Synods, or publike prayers, is of more force then of any
  particular man.}
conſent vpon earth, concerning euery thing whatſoeuer they ſhal aske,
it ſhal be done to them of my Father which is in Heauen. \V For where
there be two or three gathered in my name, there am I
\LNote{In the middes of them}{Not
\MNote{Catholike Aſſemblies.}
al aſſemblies may chalenge the preſence of Chriſt, but only ſuch as be
gathered togeather in the vnity of the Church, and therfore no
conuenticles of Heretikes directly gathering againſt the Church, are
warranted by this place.
\Cite{Cyp. de vnit. Eccl. nu.~7.~8.}}
in the middes of them.

\V Then came Peter vnto him and ſaid:
\CNote{\XRef{Lu.~17,~4.}}
Lord, how often ſhal my brother offend againſt me, & I forgiue him? vntil 
%%% o-2152
ſeauen times? \V \Sc{Iesvs} ſaid to him: I ſay not to thee
\CNote{\XRef{Lu.~17,~4.}}
vntil ſeauen times but vntil
\LNote{Seauentie times ſeauen}{There muſt be no end of forgiuing them
  that be penitent, either in the Sacrament by abſolution, or one man an
other their offenſes.}
ſeauentie times ſeauen times. \V Therfore is the Kingdom of Heauen
likened to a man being a King, that would make an account with his
ſeruants. \V And when he began to make the account, there was one
preſented vnto him that owed him ten thouſand talents. \V And hauing not
whence to repay it, his Lord commanded that he ſhould be ſold, and his
wife and children, and al that he had, and it to be repaied. \V But that
ſeruãt falling downe, beſought him, ſaying: Haue patience toward me, and
I wil repay thee al, \V And the Lord of that ſeruant moued with pitie,
dimiſſed him, and the debt he forgaue him. \V And when that ſeruant was
gone forth, he found one of his felow-ſeruants that did owe him an
hundred pence: and laying hands vpon him thratled him, ſaying: Repay
that thou oweſt. \V And his felowſeruant falling downe, beſought him,
ſaying: Haue patiẽce toward me, and I wil repay thee al. \V And he would
not: but went his way, and caſt him into priſon, til he repayed the
debt. \V And his felow-ſeruants ſeeing what was done, were very ſorie,
and they came, and told their Lord al that was done. \V Then his Lord
called him; and ſaid vnto him: Thou vngratious ſeruant, I forgaue thee
al the debt, becauſe thou beſoughteſt me: oughteſt not thou therfore
alſo to haue mercie vpon thy felow-ſeruant, euen as I had mercie vpon
thee? \V And his Lord being angrie deliuered him to the tormenters,
vntil he repayed al the debt. \V So alſo ſhal my Heauenly Father doe to
you, if you forgiue not euery one his brother from your harts.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
