%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 2317
%%% o-2130
\startChapter[
  title={Chapter 12}
  ]

\Summary{The blindnes of the Phariſees about the Sabboth he reproueth by
Scriptures, by reaſon, and by a miracle. 14.~and his death being
therfore ſought by them, he meekely goeth out of the way, according as
Eſay had Prophecied of him. 22.~His caſting out of Diuels alſo he
defendeth againſt them. 31.~and ſetteth forth the danger they ſtand in
for their horrible blaſphemie. 38.~And becauſe they aske yet for a
ſigne, he sheweth how worthily they shal be damned. 43.~foretelling how
the Diuel shal poſſeſſe their Nation, 66.~and teſtifying that although
he be of their bloud, yet not they for this, but ſuch as keepe his
commandements, are deare vnto him.}

At that time
\CNote{\XRef{Mr.~2,~23.}
\XRef{Luc.~6,~1.}}
\Sc{Iesvs} went through the corne on the Sabboth: and his
Diſciples being hungrie, began to pluck the eares, and to eate. \V And
the Phariſees ſeeing them, ſaid to him: Loe, thy Diſciples doe that
which is not 
%%% o-2131
lawful for them to doe on the Sabboth-dayes. \V But he ſaid
to them: Haue you not read what
\CNote{\XRef{1.~Re.~21,~4.}}
Dauid did when he was an hungred, and
they that were with him: \V how he entred into the houſe of God, and did
eate the loaues of propoſition, which it was not lawful for him to eate,
nor for them that were with him,
\CNote{\XRef{Leu.~24,~9.}
\XRef{Nu.~28,~9.}}
but for Prieſtes only? \V Or haue ye
not read in the Law, that on Sabboth-dayes the Prieſtes in the temple do
breake the Sabboth, & are without blame? \V But I tel you that there is
here a greater then the temple. \V And if you did know what it is,
\CNote{\XRef{Oſe.~6,~6.}}
\Emph{I wil mercie, and not 
\MNote{See the
\XRef{annotation chap.~9,~13.}}
Sacrifice}: you would neuer haue condemned the innocentes. \V For the
Sonne of man is Lord of the Sabboth alſo. 

\V And when he had paſſed from thence, he came into their Synagogue. \V
And
\CNote{\XRef{Mr.~3,~1.}
\XRef{Lu.~6,~6.}}
behold there was a man which had a withered hand, and they asked him
ſaying: Whether is it lawful to cure on the Sabboths? that they might
accuſe him. \V But he ſaid to them: What man ſhal there be of you, that
ſhal haue one ſheep: and if the ſame fal into a ditch on the Sabboths,
wil he not take hold and lift it vp? \V How much better is a man more
then a ſheep? Therfore it is lawful on the Sabboths to doe a good
deed. \V Then he ſaith to the man: Stretch forth thy hand, and he
ſtretched it forth, and it was reſtored to health euen as the other.

\V And the Phariſees going forth made a conſultation againſt him, how
they might deſtroy him. \V But \Sc{Iesvs} knowing it, retired from
thence: and many folowed him, and he cured them al. \V and he charged
them that they ſhould not diſcloſe him. \V That it might be fulfilled
which was ſpoken by Eſay the Prophet, ſaying: \V
\CNote{\XRef{Eſa.~42,~1.}}
\Emph{Behold my ſeruant
whom I haue choſen, my beloued in whom my ſoul hath wel liked. I wil put
my ſpirit vpon him, and iudgement to the Gentiles shal he shew.} \V
\Emph{He shal not contend, nor crie out, neither shal any man heare in
  the ſtreetes his voyce.} \V \Emph{The reede bruiſed he shal not
  breake, & ſomoking flaxe he shal not extinguish: til he caſt forth 
\Fix{iugdement}{iudgement}{obvious typo, fixed in other}
vnto victorie.} \V \Emph{And in his name the Gentiles shal hope.}

%%% 2318
\V Then
\CNote{\XRef{Luc.~11,~14.} \XRef{Mar.~3,~22.}}
was offered to him one poſſeſſed with a Diuel, blind and dumme:
and he cured him, ſo that he ſpake and ſaw. \V And al the multitudes were
amaſed, and ſaid: Whether this be the Sonne of Dauid? \V But the
Phariſees hearing it, ſayd: This fellow caſteth not out Diuels but 
\LNote{In Beelzebub}{The like blaſphemie againſt the Holy Ghoſt is, to
  attribute the miracles done by Saints either dead or aliue, to the
  Diuel.}
in Beelzebub the Prince of the Diuels. \V And \Sc{Iesvs} knowing their
cogitations, ſaid to them:

Euery Kingdom
\SNote{Therfore the Kingdom of Heretikes can not poſſibly ſtand,
  becauſe it is alwayes ful of diuiſion and diſſenſion.}
deuided againſt itſelf ſhal be made 
%%% o-2132
deſolate: and euery citie or houſe
deuided againſt itſelf, ſhal not ſtand. \V And if Satan caſt out Satan,
he is deuided againſt himſelf: how then ſhal his Kingdom ſtand? \V And
if I in Beelzebub caſt out Diuels, your children in whom do they caſt
out? Therfore they ſhal be your iudges. \V But if I in the Spirit of God
do caſt out Diuels, then is the Kingdom of God come vpon you. \V Or how
can a man enter into the houſe of the ſtrong, and rifle his veſſel, vnles
he firſt binde the ſtrong? and then he wil rifle his houſe. \V He that
is
\LNote{Not with me}{They
\MNote{Neuters in Religion.}
that are indifferent to al religions, commonly and fitly caled Neuters,
ioyning them ſelues to neither part, let them marke theſe words wel, and
they ſhal ſee, that Chriſt accounted al them to be againſt him & his
Church, that are not 
\Fix{planly}{plainely}{obvious typo, fixed in other}
and flatly with him and it.}
not with me, is againſt me: and he that
\LNote{Gathereth not with me}{He ſpeaketh not only of his owne Perſon,
  but of al to whom he hath committed the gouernment of his Church, and
  ſpecially of the chiefe paſtours ſucceeding Peter in the gouernment of
the whole; as 
\CNote{\Cite{Ep.~58.}}
S.~Hierome writing to Damaſus Pope of Rome, applieth theſe
words vnto him, ſaying of al Heretikes: \Emph{He that gathereth not with
  thee, ſcattereth: that is to ſay, He that is not with Chriſt, is with
  Antichriſt.}}
gathereth not with me, ſcattereth. \V Therfore I ſay to you: Euery ſinne
and blaſphemie ſhal be forgiuen men, but
\LNote{The blaſphemie of the Spirit}{He meaneth not that there is any
  ſinne ſo great, which God wil not forgiue, or whereof a man may not
  repent in this life, as ſome Heretikes at this day affirme: but that
  ſome heinous ſinnes (as namely this blaſphemie of the Iewes againſt
  the euident workes of the Holy Ghoſt, and likewiſe Archeheretikes who
  wilfully reſiſt the known truth & workes of the Holy Ghoſt in Gods
  Church) are hardly forgiuen, & ſeldom haue ſuch men grace to
  repent. Otherwiſe among al the ſinnes againſt the Holy Ghoſt (which
  are commonly reckned ſix) one only ſhal neuer be forgiuen, that is,
  dying without repentance wilfully, called
\MNote{Final impenitence.}
Final impenitence; which ſinne he committeth that dieth with contempt of
the Sacrament of Pennance, obſtinatly refuſing abſolution, by the
Churches miniſtrie: as S.~Auguſtine plainly declareth in theſe wordes:
\MNote{Remiſſion of ſinnes in the Church.}
\Emph{Whoſoeuer he be that beleeueth not mans ſinnes to be remitted in
  Gods Church, and therfore deſpiſeth the bountifulnes of God in ſo
  mighty a work, if he in that obſtinate mind continue til his liues
  end, he is guilty of ſinne againſt the Holy Ghoſt}, in which Holy
Ghoſt Chriſt remitteth ſinnes. 
\Cite{Enchir.~83.}
\Cite{Ep.~50. in fine.}}
the blaſphemie of the Spirit ſhal not be forgiuen. \V And whoſoeuer ſhal
ſpeake a word againſt the
\LNote{Sonne of man}{The Iewes in their wordes ſinned againſt the Sonne
of man, when they reprehended thoſe things which he did as man, to wit,
caling him a glutton, a great drinker of wine, a frend of the Publicans,
a Samaritane, and taking offenſe becauſe he kept company with ſinners,
brake the Sabboth, and ſuch like: and this ſinne might more eaſely be
forgiuen them, becauſe they iudged of him, as they would haue don, of
any other man: but they ſinned and blaſphemed againſt the Holy Ghoſt
(caled here the finger of God whereby he wrought miracles) when of
malice they attributed the euident workes of God in caſting out Diuels,
to the Diuel himſelf: & this ſinne ſhal not be remitted, becauſe it shal
hardly be remitted, as we ſee by the plague of their poſteritie vntil
this day.}
Sonne of man, it ſhal be forgiuen him: but he that ſhal ſpeake againſt
the Holie-Ghoſt, it ſhal not be forgiuen him neither in this world,
\LNote{Nor in the world to come}{S.~Auguſtine 
\MNote{Purgatorie.}
& other Holy Doctours gather herevpon, that ſome ſinnes may be remitted
  in the next life, & conſequently proue Purgatorie thereby.
\Cite{De Ciuit. Dei li.~21, c.~13.}
\Cite{D.~Gregor. Dial. li.~4, c.~39.}}
nor in the world to come. \V Either
\SNote{It is a mans owne free wil & election, to be a good tree or an il
tree: to bring forth good fruits or bad. S.~Auguſtine vpon this place. 
\Cite{li.~2. c.~4. de actis cum Felic. Manichæo.}}
make the tree good, and his fruit good: or make the tree euil, and his
fruit euil. For of the fruit the tree is knowen. \V You vipers broods,
how can you ſpeake good things, whereas you are euil? for of the
aboundance of the hart the moth ſpeaketh. \V A good man out of a good
treaſure bringeth forth good things: and an euil man out of an euil
treaſure bringeth forth euil things. \V But I ſay vnto you, that euery 
\LNote{Idle word}{If of euery idle word we muſt make account before God in
  iudgement, and yet ſhal not for euery ſuch word be damned
  euerlaſtingly: then there muſt needs be ſome temporal puniſhment in
  the next life.}
idle word that men ſhal ſpeake, they ſhal render an account for it in
the day of iudgement. \V For of thy wordes thou ſhalt be iuſtified, and
of thy wordes thou ſhalt be condemned.

\V Then anſwered him certaine of the Scribes and Phariſees, ſaying:
Maiſter, we would ſee a ſigne from thee. \V Who anſwered, and ſaid to
them: 

The wicked and aduouterous Generation ſeeketh a ſigne: and a ſigne ſhal
not be giuen it, but the ſigne of Ionas the Prophet. \V For as
\CNote{\XRef{Ion.~2,~2.}}
Ionas was
in the whales belly three dayes and three nightes; ſo ſhal the Sonne of man be
in the hart of the earth three dayes and three nightes. \V The men of
Niniuee ſhal riſe in the iudgement with this Generation, and shal
condemne it: becauſe
\CNote{\XRef{Ion.~3,~5.}}
they did pennance at the preaching of Ionas. And
behold more then Ionas here. \V The
\CNote{\XRef{3.~Reg.~10,~1.}}
Queen of the South shal riſe in the
iudgement with this Generation, and shal condemne it: becauſe she came
from the ends of the earth to heare the wiſedom of Salomon, and behold
more then 
%%% o-2133
Salomon here. \V And
\CNote{\XRef{Luc.~11,~24.}}
when an vncleane Spirit shal goe out of a man, he
walketh through dry places, ſeeking reſt, and findeth not. \V Then he
ſaith: I wil returne into my houſe whence I came out. And coming he
findeth it vacant, ſwept with beſoms, and trimmed. \V Then goeth he, and
taketh with him ſeuen other Spirits more wicked then himſelf, and they
%%% 2319
enter in and dwel there: and
\CNote{\XRef{2.~Pet.~2,~20.}}
the laſt of that man be made worſe then the
firſt. So shal it be alſo to this wicked Generation.

\V As he was yet ſpeaking to the multitudes,
\CNote{\XRef{Mr.~3,~31.}
\XRef{Luc.~8,~20.}}
behold his mother and his
brethren ſtood without, ſeeking to ſpeake to him. \V And one ſaid vnto
him: Behold thy mother and thy brethren ſtand without, ſeeking thee. \V
But he anſwering him that told him, ſaid:
\LNote{Who is my mother}{The 
\MNote{Al Heretikes alleage Scriptures.}
dutiful affection toward our parents and kinsfolke is not blamed, but
the inordinate loue of them, to the hinderance of our ſeruice & duty
toward God. Vpon this place ſome old Heretikes denied Chriſt to haue any
mother. 
\Cite{Aug. li. de Fid. & Symb. c.~4.}
Neither euer, was there any hereſie ſo abſurd, but it would ſeeme to
haue Scripture for it.}
Who is my mother, and who are my brethren? \V And ſtretching forth his
hand vpon his Diſciples, he ſaid: Behold my mother and my brethren. \V
For whoſoeuer shal doe the wil of my Father, that is in Heauen: he is my
brother, and ſiſter, and mother.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
