%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 28
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-28


\project douay-rheims


%%% 2365
%%% o-2184
\startChapter[
  title={Chapter 28}
  ]

\Summary{He riſeth againe the third day, & (the blind moſt obſtinate
Iewes by bribery working to their owne reprobation) he appeareth to
his Diſciples in Galilee (as both before his Paſsion he foretold
them
\XRef{Mat.~26.}
and now after his Reſurrection, firſt the Angel,
then alſo himſelf appointed by the women) 18.~and ſendeth them to al
Nations, to build his Church among the Gentils.}

And
\CNote{\XRef{Mr.~16,~1.}
\XRef{Luc.~24,~1.}
\XRef{Io.~20,~1.}}
\MNote{\Sc{Easter} day.}
in the euening of the Sabboth which dawneth on the firſt of the Sabboth,
came Marie Magdalene, and the other Marie
\LNote{To ſee the Sepulcher}{The
\MNote{The holy Sepulcher, and Pilgrimage therunto.}
deuout women came to viſit our Sauiours ſepulcher, and for their
 deuotion firſt deſerued to know the Reſurrection, and to ſee him
 riſen. The honour of the which Sepulcher, and the Pilgrimage thereunto
 in the Primitiue Church, S.~Hierom declareth in theſe wordes:
\CNote{\Cite{In Ep.~17. Paulæ & Euſtoch. ad Marcel. to.~1.}}
\Emph{The
 Iewes ſometime honoured Sancta Sanctorum, becauſe there were the Cherubs & the
 Propitiatorie, & the Arke of the Teſtament, Manna, Aarons rod, & the
 golden Altar. Doth not the Sepulcher of our Lord ſeeme vnto thee more
 honorable? Which as often as we enter into ſo often doe we ſee our Sauiour
 lie in the ſindon & ſtaying there a while, we ſee the Angel againe ſit
 at his feet, & at his head the napkin wrapped togeather. The glorie of
 whoſe Sepulcher, we know was long prophecied before Ioſeph hewed it
 out, by Eſay ſaying:
\CNote{\XRef{Eſa.~11.}}
And his reſt shal be honour, to wit, becauſe
 the place of our Lordes burial should be honoured of al men.} And at
 this preſent, notwithſtanding the Turkes dominion, yet doe the
 Religious Chriſtian Catholike men by Gods mighty prouidence keepe the
 holy Sepulcher, which is within a goodly Church, and Chriſtians come
 out of al the world in Pilgrimage to it.}
to ſee the ſepulchre. \V And behold there was made a great
earth-quake. For an Angel of our Lord deſcended from Heauen: and
comming, roled back the ſtone, and ſate vpon it: \V and his countenance
was as lightning, and his garment as ſnow. \V And for feare of him, the
watchmen were frighted, and became as dead. \V And the Angel anſwering
ſaid to the women: Feare not you. For I know that you ſeeke \Sc{Iesvs}
that was crucified. \V He is not here, for he is riſen,
\CNote{\XRef{Mt.~24,~32.}}
as he ſaid. Come
and ſee the place where our Lord was laid. \V And going quickly tel ye
his Diſciples that he is riſen: and behold he goeth before you into
Galilee: there you ſhal ſee him. Loe I haue foretold you.

\V And they went forth quickly out of the monument with feare and great
ioy, running to tel his Diſciples. \V And behold \Sc{Iesvs} met them,
ſaying: Al haile. But they came neere and tooke hold of his feet, and
adored him. \V Then \Sc{Iesvs} ſaid to them: Feare not. Goe, tel my
brethren that they goe into Galilee, there they ſhal ſee me.

\V Who when they were departed, behold certaine of the watchmen came
into the citie, and told the chiefe Prieſts al things that had been
done. \V And being aſſembled togeather with the Ancients, taking
counſel, they gaue a great ſumme of money to the ſouldiars, \V ſaying:
Say you, that his Diſciples came by night, and ſtole him away when we
were aſleep. \V And if the Preſident ſhal heare of this, we
\%%% o-2185
wil perſuade
him, and make you ſecure. \V But they taking the money, did as they were
taught. And this word was bruited abrode among the Iewes, euen vnto this
day.

\V And the eleuen Diſciples went into Galilee, vnto the mount
where \Sc{Iesvs} had appointed them. \V And ſeeing him they adored, but
ſome doubted. \V And \Sc{Iesvs} comming neere ſpake vnto them, ſaying:
Al power is giuen to me in Heauen and in earth. \V
\LNote{Going therfore}{Commiſſion
\MNote{The Catholike Church to be gathered of al Nations.}
to baptize and preach to al Nations giuen to the Apoſtles, and grounded
 vpon Chriſts ſoueraigne authority, to whom was giuen al power in Heauen
 and in earth.}
going therfore teach ye al Nations: \Sc{Baptizing them in the name of
the father, and of the sonne, and of the Holy Ghoſt}, \V teaching them
to obſerue al things whatſoeuer I haue cõmanded you: and behold I am
\LNote{With you al daies}{Here
\MNote{Chriſts continual protection of the ſame Church.}
Chriſt doth promiſe his concurrence with his Apoſtles & their
 Succeſſours, as wel in preaching as miniſtring the Sacraments, and his
 protection of the Church neuer to ceaſe til the worlds end: contrary to
 our Aduerſaries, ſaying that the Church hath failed many hundred yeares
 til Luther and Caluin.}
with you al daies, euen to the conſummation of the world.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
