%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-01


\project douay-rheims


%%% 2291
%%% o-2103
\startChapter[
  title={Chapter 1}
  ]

\Summary{The
\MNote{The firſt part of this Ghoſpel, of the Infancie of our Sauiour
  Chriſt.}
pedegree of Ieſus, to shew that he is Chriſt, promiſed to
\CNote{\XRef{Gen.~12.~22.}}
Abraham and
\CNote{\XRef{2.~Reg.~7.}
\XRef{Pſa.~131.}}
Dauid. 18.~That he was conceiued and borne of a Virgin, as Eſay
prophecied of him.}

The booke of the
\CNote{\XRef{Luc.~3.~31.}}
generation of \Sc{Iesvs} Chriſt, the ſonne of Dauid,
the ſonne of Abraham.

\V
\CNote{\XRef{Gen.~21. 25.~29.~38.}}
Abraham begat Iſaac, And Iſaac, begat Iacob. And Iacob begat Iudas
and his brethren: \V And Iudas begat Phares and Zaram of
\LNote{Thamar}{Chriſt abhorred not to take fleſh of ſome that were il,
  as he choſe Iudas among his Apoſtles: Let not vs diſdaine to receaue
  our ſpiritual birth and ſuſtenance of ſuch as be not alwayes good.}
Thamar.
\CNote{\XRef{1.~Par.~2.~5.}
\XRef{Ruth.~4.~18.}}
And Phares begat Efron. And Efron begat Aram. \V And Aram begat
Aminadab. And Aminadab begat Naaſſon. And Naſſon begat Salmon. \V And
Salmon begat Booz of Raab. And Booz begat Obed of Ruth. And Obed begat
Ieſſe. \V And Ieſſe begat Dauid the King.

And
\CNote{\XRef{2.~Re.~12.~24.}}
Dauid the King begat Salomon of her that was the wife of Vrias. \V
And
\CNote{\XRef{1.~Par.~3,~10.}}
Salomon begat Roboam. And Roboam begat Abia. And Abia begat Aſa. \V
And Aſa begat Ioſaphat. And Ioſaphat begat Ioram. And Ioram begat
Ozias. \V And Ozias begat Ioatham. And Ioatham begat Achaz. And Achaz
begat Ezechias. \V And Ezechias begat Manaſſes. And Manaſſes begat
Amon. And Amon begat Ioſias. \V And Ioſias begat Iechonias and his
brethren in the
\CNote{\XRef{4.~Reg.~24.}
\XRef{2.~Par.~36.}}
Tranſmigration of Babylon.

\V And after the Tranſmigration of Babylon,
\CNote{\XRef{1.~Par.~3.}}
\Fix{Iochonias}{Iechonias}{obvious typo, fixed in other}
begat Salathiel.
\CNote{\XRef{1.~Eſd.~3.}}
And Salathiel begat Zorobabel. \V And Zorobabel begat
Abiud. And Abiud begat Eliacim. And Eliacim begat Azor. \V And Azor
begat Sadoc. And Sadoc begat Achim. And Achim begat Eliud. \V And Eliud
begat Eleazar. And Eleazar begat Mathan. And 
%%% 2292
Mathan begat Iacob. \V And Iacob
%%% o-2104
begat
\LNote{Ioſeph}{Ioſeph marying our Lady as neere of kinne (for ſo was the
\CNote{\XRef{Nu.~36.}}
law) by his pedegree ſheweth hers, and conſequently Chriſts pedegree
from Dauid.}
Ioſeph the 
\LNote{Huſband}{True and perfect mariage, and continual liuing in the
  ſame, without carnal copulation. 
\Cite{Aug. lib.~2. Conſen. Euang. c.~1.}}
huſband of \Sc{Marie}:
\TNote{\L{De qua}}
of whom was borne \Sc{Iesvs}, who is called \Sc{Christ}.

\V Therfore al the Generations from Abraham vnto Dauid, fourteen
Generations. And from Dauid to the Tranſmigration of Babylon, fourteen
Generations. And from the Tranſmigration of Babylon vnto \Sc{Christ},
fourteen Generations.

\V And the Generation of \Sc{Christ} was in this wiſe. When his Mother
\Sc{Marie} was ſpouſed to Ioſeph, before they came together, ſhe was
found to be with child, by the Holy Ghoſt. \V Whereupon Ioſeph her
huſband for that he was
\SNote{This word \Emph{Iuſt}, ſounding that a man is iuſt in deede, &
  not only ſo imputed, Proteſtants tranſlate, \Emph{Righteous} in this
  and diuers other places.}
a iuſt man, & would not
\CNote{\XRef{Deu.~14,~1.}}
put her to open ſhame: was minded ſecretly to
dimiſſe her. \V But as he was thus thinking, behold the Angel of our
Lord appeared to him in
\Fix{in }{}{obvious typo, fixed in other}% to prevent a paragraph break
ſleepe ſaying: Ioſeph ſonne of Dauid, feare not to take \Sc{Marie} thy
wife, for that which is
\LNote{Borne in her}{The triple good or perfection of mariage
  accompliſhed in the parents of Chriſt, to wit, Iſſue, Fidelitie,
  Sacrament. 
\Cite{Aug. de nup. & cont. li.~1. c.~13.}}
borne in her, is of the Holy Ghoſt. \V And ſhe ſhal bring forth a Sonne:
and thou ſhalt cal his name
\SNote{\Sc{Iesvs} an Hebrew word, in Engliſh \Sc{Saviovr}.}
\Sc{Iesvs}. For he ſhal ſaue his people from their ſinnes. \V And al
this was done that it might be fulfilled which our Lord ſpake by the
Prophet ſaying. \V
\CNote{\XRef{Eſa.~7,~14.}}
\Emph{Behold}
\LNote{A virgin}{Our 
\MNote{Virginitie preferred.}
Sauiour borne in mariage, but yet of a Virgin, would honour both ſtates:
and withal, teacheth vs againſt Iouinian the old Heretike and theſe of
our time, that virginity and the continent life are preferred before
mariage that hath carnal copulation. See
\Cite{S.~Hierom. adu. Ionin.} &
\Cite{S.~Greg. Nazianz. Ser.~20 de ſtudio in pauperes, in initio.}}
\LNote{A virgin}{As
\MNote{Our B.~Ladies perpetual virginity.}
our Ladie both a virgin and a mother, brought forth Chriſt the head
corporally: ſo the Church a virgin and a mother, bringeth forth the
members of this head ſpiritually. 
\Cite{Aug. li. de virg. ca.~2.}} 
\Emph{a Virgin shal be with childe, &}
\LNote{And bring forth}{The Heretike Iouinian is here refuted,
  holding that her virginity was corrupted in bringing forth Chriſt
\Cite{Aug. har.~28. li.~2. con Iulian c.~2.}}
\Emph{bring forth a ſonne, and they shal cal his name Emmanuel}, which
being interpreted is, \Emph{God with vs.} \V And Ioſeph riſing vp from
ſleepe, did as the Angel of our Lord commaunded him, & tooke his
wife. And he knew her not
\LNote{Til Firſt borne}{Heluidius 
\MNote{Tradition.}
of old abuſed theſe wordes, \Emph{til}, and \Emph{firſt borne}, againſt
the perpetual virginity of our B.~Ladie. 
\Cite{Hiero. cont. Helu.}
which truth though not expreſſed in
Scripture, yet our Aduerſaries alſo do graunt, & Heluidius for denial
therof was condemned for an heretike by tradition only. 
\Cite{Aug. har.~84.}}
til ſhe brought forth her firſt borne Sonne: & called his name \Sc{Iesvs}.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
