%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 25
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-25


\project douay-rheims


%%% 2353
%%% o-2171
\startChapter[
  title={Chapter 25}
  ]

\Summary{Continuing his Sermon, he bringeth two parables, of ten
  Virgins, and of Talents, to shew how it shal be in Domeſday with the
  Faithful that prepare, and that prepare not them ſelues. 31.~Then alſo
  without parables he sheweth that ſuch Faithful as doe workes of mercy,
  shal haue for them life euerlaſting: and ſuch as doe not, euerlaſting
  damnation.}

%%% 2354
Then ſhal the Kingdom of Heauen be like to ten
\LNote{Virgins}{Theſe
\MNote{Good works neceſſarie.}
Virgins fiue wiſe, and fiue fooliſh, ſignifie that in the Church
militant there be good and bad: which bad ſhal be ſhut out at the later
day, although they haue lamps (that is faith) as the other, becauſe
their lamps are out: that is, their faith is dead without charitie and
good works to lighten them.
\Cite{Greg. ho.~12.}}
Virgins, which taking their
\LNote{Lamps}{Theſe lamps lighted, be good works, namely of mercy, and
the laudable conuerſation which ſhineth before men.
\Cite{Aug. ep.~120. c.~23.}}
lamps went forth to meet the Bridegrome and the Bride. \V And fiue of
them were fooliſh, and fiue wiſe. \V But the fiue fooliſh, hauing taken
their lamps, did not take
\LNote{oyle}{This
\MNote{Right intẽtiõ.}
oyle is the right inward intention directing our works to Gods glorie,
and not to the praiſe of our ſelues in the ſight of men.
\Cite{Aug. ep.~120. c.~33.}}
oile with them: \V but the wiſe did take oile in their veſſels with the
lamps. \V And the Bridegrome tarying long they ſlumbered al and
ſlept. \V And at midnight there was a clamour made: Behold the
Bridegrome commeth, goe
%%% o-2172
ye forth to meet him. \V Then aroſe al thoſe Virgins, and they trimmed
their lamps. \V And the fooliſh ſaid to the woſe: Giue vs of
\SNote{If we be not in the fauour of God, and haue not our owne merits,
we ſhal not be holpen by other mens deſerts at the day of iudgement.}
your oile, becauſe our lamps
\Fix{ars}{are}{obvious typo, fixed in other}
going out. \V The wiſe anſwered, ſaying: Leſt peraduenture there ſuffice
not for vs and you, goe rather to them that ſel, and buy for your
ſelues. \V And whiles they went to buy, the Bridegrome was come: and
they that were ready, entred with him to the mariage, and the gate was
ſhut. \V But laſt of al come alſo the other Virgins ſaying: Lord, Lord,
open to vs. \V But he anſwering ſaid: Amen I ſay to you, I know you not. \V
Watch ye therfore, becauſe you know not the day nor the houre.

\V For
\CNote{\XRef{Lu.~10,~12.}}
euen as a man going into a ſtrange countrie, called his ſeruants,
and deliuered them his goods. \V And to one he gaue fiue talents, and to
an other two, and to an other one, to euery one according to his proper
facultie: and immediatly he tooke his iourney. \V And he that had
receaued the fiue talents, went his way, and occupied with the ſame, and
gained other fiue. \V Likewiſe alſo he that had receaued the two, gained
other two. \V But he that had receaued the one, going his way digged
into the earth, and hid his Lords money. \V But after much time the Lord
of thoſe ſeruãts cõmeth, & made acount with thẽ. \V And there came he
that had receaued the fiue talents, & offred other fiue talẽts, ſaying:
Lord fiue talẽts thou didſt deliuer me, behold
\SNote{Free wil with Gods grace doth merit.}
I haue gained other fiue beſides. \V His Lord ſaid vnto him: Wel-fare
thee good & faithful ſeruant, becauſe thou haſt been faithful ouer a few
things I wil place thee ouer many things: enter into the ioy of thy
Lord. \V And there came alſo he that had receaued the two talents, &
ſaid: Lord two talents thou didſt deliuer me: behold I haue gained other
two. \V His Lord ſaid to him: Wel-fare thee good & faithful ſeruãt:
becauſe thou haſt been faithful ouer a few things, I wil place thee ouer
many things, enter into the ioy of thy Lord. \V And he alſo that had
receaued the one talẽt, came forth, and ſaid: Lord, I know that thou art
a hard man; thou reapeſt where thou didſt not ſow, and gathereſt where
thou ſtrawed not: \V and being afraid I went, and hid thy talent in the
earth: behold here thou haſt that which thine is. \V And his Lord
anſwering, ſaid to him:
\SNote{A terrible example for al ſuch as doe not employ the very leaſt
guift of God, to his glorie.}
Naughtie and ſloughful ſeruant, thou didſt know that I reape where
%%% o-2173
I ſow not, & gather where I ſtrawed not: \V thou oughteſt therfore to
haue committed my money to the bankers, and comming I might haue
receaued mine owne
\LNote{With vſurie}{Vſurie is here taken for the lawful gaine that a
man getteth by wel employing his goods. When God giueth vs any talent or
 talents, he looketh for vſurie, that is, for ſpiritual increaſe of the
 ſame by our diligence and induſtrie.}
with vſurie. \V Take ye away therfore the talent
%%% 2355
from him, and giue it him that hath ten talents. \V For to
\CNote{\XRef{Mr.~13,~12.}
\XRef{Luc.~8,~18.}}
euery one
that hath ſhal be giuen, and he ſhal abound: but from him that hath not,
\LNote{That which he ſeemeth to haue}{He
\MNote{We muſt vſe Gods guifts.}
is ſaid to haue Gods guifts, that vſeth them, and to ſuch an one God wil
 increaſe his guifts. He that vſeth them not, ſeemeth to haue, rather
 then hath them, and from him God wil withdraw that which before he
 gaue.}
that alſo which he ſeemeth to haue, ſhal be taken away from him. \V And
 the vnprofitable
ſeruant caſt ye out into the vtter darknes. There ſhal be weeping and
gnaſhing of teeth.

\V And when the Sonne of man ſhal come in his maieſtie, and al the
Angels with him, then ſhal he ſit vpon the ſeat of his maieſtie: \V and
al Nations ſhal be gathered togeather before him, and he ſhal
\LNote{Separate}{Loe
\MNote{Good and bad in the Church.}
here is the ſeparation; for in the Church militant they liued both
 togeather. As for Heretikes, they went out of the Church before, and
 ſeparated them ſelues, and therfore are not to be ſeparated here, as
 being iudged already.}
ſeparate them one from an other, as the paſtour ſeparateth the ſheep
from the goats: \V and ſhal ſet the ſheep at his right hand, but the
goats at his left. \V Then ſhal the King ſay to them that ſhal be at his
right hand:
%%% !!! This one really does not fit into the current setup. The firſt
%%% argument refers to two separate verses!  Thus, even the verse
%%% numbers will need to be hard-coded, not just the text.
\LNote{Come ye. 41.\ Get ye away}{It
\MNote{Heauen is the reward of good workes, and Hel of the contrarie.}
is no incongruitie that God should ſay: Goe into euerlaſting fire, to
them that by their free wil haue repelled his mercie; and to the other:
Come ye Bleſſed of my Father, take the Kingdom prepared for you, that
by their free wil haue receaued faith, and confeſſed their ſinnes, and
done pennance.
\Cite{Aug. li.~2. act cum Fel. Manich. c.~8.}}
Come ye Bleſſed of my Father, poſſeſſe you the Kingdom
\SNote{This Kingdõ then is prepared for thoſe only that doe good works: as
Chriſt alſo ſignifieth els where, ſaying that it is not in his power to
giue it otherwiſe. See the annot.\ 
\XRef{c.~20,~23.}}
prepared for you from the foundation of the
\Fix{world}{world.}{obvious typo, fixed in other}
\V For I was an
hungred, and
\LNote{You gaue me}{Hereby we ſee how much almes-deeds and al works of
mercy preuaile towards life euerlaſting, and to blot out former ſinnes.
\Cite{Aug. in Ps.~49.}}
you gaue me to eate:
\Fix{i}{I}{obvious typo, fixed in other}
was a thirſt, and you gaue me to drinke. \V I was a ſtranger, and you
tooke me in: naked, and you couered me: ſick, and you viſited me. I was
in priſon, and you came to me. \V Then ſhal the iuſt anſwer him, ſaying:
Lord, when did we ſee thee an hungred, and fed thee, a thirſt, and gaue
thee drinke? \V and when did we ſee thee a ſtranger, and tooke thee in?
or naked, and couered thee? \V or when did we ſee thee ſick or in
priſon, and came to thee? \V And the King anſwering, ſhal ſay to them:
Amen I ſay to you, as long as you did it to one of theſe my leaſt
brethren, you did it to me. \V Then he ſhal ſay to them alſo that ſhal
be at his left hand:
%%% !!! See other commented long note above. These seeming two notes are
%%% really one.
%%% \LNote{}{}
Get ye away from me you curſed into fire euerlaſting, which was prepared
for the Diuel and his Angels. \V For I was an hungred, & you
\LNote{Gaue me not}{He chargeth them not here that they beleeued not,
but that they did not good works. For ſuch did beleeue but they cared
not for good works, as though by dead faith they might haue come to
Heauen.
\Cite{Aug. de fid. & op. c.~15.} &
\Cite{Dulcit. q.~2. to 4.}}
gaue me not to eat; I was a thirſt, & you gaue me not to drinke. \V I
was a ſtranger, and you tooke me not in: naked, and you couered me not:
ſick, & in priſon, and you did not viſit me. \V Then they alſo ſhal
anſwer him, ſaying: Lord, when did we ſee thee an hungred, or a thirſt,
or a ſtranger, or naked, or ſick, or in priſon, & did not miniſter to
 thee? \V Then he 
ſhal anſwer them, ſaying: Amen I ſay to you, as long as you did it not
to one of theſe leſſer, neither did you it to me. \V And theſe ſhal goe
into puniſhment euerlaſting: but the iuſt, into life euerlaſting.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
