%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-23


\project douay-rheims


%%% 2347
%%% o-2164
\startChapter[
  title={Chapter 23}
  ]

\Summary{The Scribes and Phariſees after al this, continuing ſtil
  incorrigible, although he wil haue the doctrine of their Chaire obeied,
yet againſt their workes (and namely their ambition) he openly
inueigheth, crying to them eight woes for their eightfold hypocriſie and
blindnes: 34.~and ſo concluding with the moſt worthy reprobation of that
perſecuting Generation and their mother citie Ieruſalem, with her
Temple.}

Then \Sc{Iesvs} ſpake to the multitudes and to his Diſciples, \V ſaying:
Vpon 
\LNote{Chaire of Moyſes}{God
\MNote{The See of Rome preſerued in truth.}
preſerueth the truth of Chriſtian Religion in the Apoſtolike See of
Rome, which is in the new Law anſwerable to the chaire of Moyſes,
notwithſtanding the Biſhops of the ſame were neuer ſo wicked of life:
yea though ſome traitour as il as Iudas were Biſhop thereof, it ſhould
not be preiudicial to the Church and innocent Chriſtians, for whom our
Lord prouiding ſaid: Doe that which they ſay, but doe not as the doe.
\Cite{Auguſt. Epiſt.~165.}}
the chaire of Moyſes haue ſitten 
\Fix{he}{the}{obvious typo, fixed in other}
Scribes & the Phariſees. \V Al things therfore
\LNote{Whatſoeuer they shal ſay}{\Emph{Why} 
\MNote{The dignitie of the See of Rome, notwithſtanding ſome euil
  Bishops therof.}
(ſaith
\CNote{\Cite{Contra lit. Petil. l.~2. c.~51.}}
S.~Auguſtin) \Emph{doeſt thou cal the Apoſtolike Chaire the
  Chaire of peſtilence? If for the men, Why? Did our Lord Ieſus Chriſt
  for the Phariſees, any wrong to the Chaire wherin they ſate? Did he
  not commend that chaire of Moyſes, and preſeruing the honour of the
  Chaire, reproue them? For he ſaith: They ſit vpon the Chaire of
  Moyſes, that which they ſay doe ye. Theſe things if you did wel
  conſider, you would not for the men whom you defame, blaſpheme the See
Apoſtolike wherwith you doe not communicate.} And againe he ſaith:
\CNote{\Cite{Contra lit. Petil. l.~2. c.~6.}}
\Emph{Neither for the Phariſees (to whom you compare vs not of wiſdom
  but of malice) did our Lord command the Chaire of Moyſes to be
  forſaken, in which Chaire verily he figured his owne, for he warned
  the People to doe that which they ſay, and not to doe that which they
  doe, and that the holineſſe of the Chaire be in no caſe forſaken, nor
  the vnity of the flock deuided, for the naughty Paſtours.}}
whatſoeuer they ſhal ſay to you, obſerue ye and doe ye: but according to
their workes doe ye not, for they ſay and doe not. \V For
\CNote{\XRef{Lu.~11,~46.}
\XRef{Act.~15,~10.}}
they bind
heauie burdens & importable: & put them vpon mens ſhoulders:
%%% 2348
%%% o-2165
but with a finger of their owne they wil not moue them. \V But they doe
al their workes for to be ſeen of men. For they make brode their
\SNote{Theſe phylacteries were peeces of parchement, wherein they wrote
the ten commandements, and folded it, and caried it on their forehead
before their eyes, imagining groſly and ſuperſtitiouſly, that ſo they
\Fix{fufilled}{fulfilled}{obvious typo, fixed in other}
that which is ſaid
\XRef{Deu.~6.}
\Emph{They shal be immoueable before thine eyes.}
\Cite{Hiero. in 23.~Mat.}}
phylacteries, and enlarge their
\CNote{\XRef{Deu.~22,~12.}
\XRef{Nu.~15,~38.}}
fringes. \V And they
\LNote{Loue the firſt places}{He condemneth not due places of
Superiority giuen or taken of men according to their degrees, but
ambitious ſeeking for the ſame, and their proud hart and wicked
intention, which he ſaw within them, and therfore might boldly
reprehend them.}
loue the firſt places at ſuppers, and
\CNote{\XRef{Mr.~12,~38.}}
the firſt chaires in the
Synagogues, \V and ſalutations in the market-place, and to be called of
men, Rabbi. \V But be not you called Rabbi. For
\LNote{One is your Maiſter}{In the Catholike Church there is one
  Maiſter, Chriſt our Lord, and vnder him one Vicar, with whom al
  Catholike Doctours and teachers are one, becauſe they teach al one
  thing. 
\MNote{Many Maiſters are many Arch heretikes.}
But in Archheretikes it is not ſo; where euery one of them is a diuerſe
Maiſter, and teacheth contrarie to the other, and wil be called Rabbi &
Maiſter, euery one, of their owne Diſciples: Arius a Rabbi among the
Arians, Luther among the Lutherans, and among the Caluiniſts Caluin.}
one is your Maiſter, and al you are 
\Fix{brethen.}{brethren.}{obvious typo, fixed in other}
\V And cal none Father to yourſelf vpon earth: for one is your Father,
he that is in Heauen. \V Neither
\CNote{\XRef{Ia.~3,~1.}}
be ye called
\LNote{Maiſters}{Wiclefe,
\MNote{Doctours, Maiſters, and ſpiritual Fathers.}
and the like Heretikes of this time, doe herevpon condemne degrees of
Schoole & titles of Doctours and Maiſters where they might as wel
reproue
\CNote{\XRef{2.~Tim.~1.}
\XRef{1.~Tim.~2.}}
S.~Paul for calling himſelf \Emph{Doctour and Maiſter of the
  Gentiles}: and for ſaying
\CNote{\XRef{Eph.~4.}}
that there ſhould be alwayes \Emph{Doctours}
in the Church. And whereas they bring the other words folowing, againſt
Religious men who are called Fathers; as wel might they by this place
take away the name of carnal fathers, and blame
\CNote{\XRef{1.~Cor.~4,~15.}}
S.~Paule for calling
himſelf the only ſpiritual Father of the Corinthians. But indeed nothing
is here forbidden but the contentious diuiſion and partiality of ſuch as
make themſelues Ringleaders of Schiſmes & Sects, as Donatus, Arius,
Luther, Caluin.}
Maiſters: for one is your Maiſter, Chriſt. \V He that is the greater of you,
ſhal be your ſeruitour. \V And he that exalteth himſelf, ſhal be
humbled: and he that humbleth himſelf, ſhal be exalted.

\V But woe to you
\LNote{Scibes and Phariſees}{In
\MNote{The honour of Prieſthood.}
al theſe reprehenſions it is much to be noted, that our Sauiour for the
honour of Prieſthood neuer reprehendeth Prieſts by that name.
\Cite{Cypr. ep.~65.}
whereas our Heretikes vſe this name of purpoſe in reproach and deſpite.}
Scribes & Phariſees, Hypocrites: becauſe you ſhut the Kingdom of Heauen
before men. For your ſelues do not enter in: & thoſe that are going in,
you ſuffer not to enter.

\V Woe to you Scribes and Phariſees, Hypocrites: becauſe you
\CNote{\XRef{Luc.~20,~47.}}
deuoure widowes houſes,
\LNote{Praying long prayers}{They
\MNote{The intention.}
are not reprehended here for the things them ſelues, which for the moſt
part are good, as, long prayer, making Proſelytes, garniſhing the
Prophets ſepulchres, &c. but for their wicked purpoſe and intention, as
before is ſaid of faſting, prayers, almes.
\XRef{Mat.~6.}}
praying long prayers. For this you ſhal receaue the greater iudgement.

\V Woe to you Scribes and Phariſees, Hypocrites: becauſe you goe round
about the ſea and the land to make one Proſelyte: and when he is made,
you make him the child of Hel
\LNote{Double more}{They
\MNote{Not only faith.}
that teach that it is enough to haue only faith, doe make ſuch
Chriſtians, as the Iewes did Proſelytes children of Hel farre more then
before.
\Cite{Aug. lib. de fide & oper. cap.~26.}}
double more then yourſelues.

\V Woe to you blind guides, that ſay, whoſoeuer ſhal ſweare by the
temple, it is nothing: but he that ſhal ſweare by the gold of the
temple, is bound. \V Ye fooliſh and blind: for whether is greater, the
gold, or the temple that ſanctifieth the gold? \V And whoſoeuer ſhal
ſweare by the Altar, it is nothing: but whoſoeuer ſhal ſweare by the
guift that is vpon it, is bound. \V Ye blind; for whether is greater,
the guift, or the Altar that
\LNote{Sanctifieth}{Note that donaries and guifts beſtowed vpon Churches
and Altars, be ſanctified by dedication to God, and by touching the
Altar and other holy things: 
\MNote{The Altar is ſanctified by our Lords body therevpon.}
as now ſpecially the veſſels of the Sacrifice and Sacrament of Chriſtes
body and bloud, by touching the ſame, and the Altar itſelf whervpon it
is conſecrated. Wherof Theophylact writeth thus vpon this place:
\CNote{\Cite{Theoph.~23. in Mat.}}
\Emph{In the
old law Chriſt, permitteth not the guifts to be greater then the Altar;
but with vs, the Altar is ſanctified by the guift: for the hoſtes by the
diuine grace are turned into our Lords body, and therfore is the Altar
alſo ſanctified by them.}}
ſanctifieth the guift? \V He therfore that ſweareth by the Altar,
ſweareth by it, and by al things that are vpon it: \V And whoſoeuer ſhal
ſweare by the temple, ſweareth by it and 
\LNote{By him that dwelleth in it}{By this we ſee that ſwearing by
  creatures, as by the Ghoſpel, by Saints, is al referred to the honour
of God, whoſe Ghoſpel it is, whoſe Saints they are.}
by him that dwelleth in it: \V And he that ſweareth by Heauẽ, ſweareth
by the throne of God, and by him that ſitteth thereon.

\V Woe to you Scribes and Phariſees, Hypocrites: becauſe you tithe mint,
and aniſe, and cummin, and haue left the weightier things of the law,
iudgement, and mercie, and faith. Theſe things you ought to haue done, &
not to haue omitted thoſe. \V Blind guides, that ſtraine a gnat, and
ſwallow a camel.

%%% o-2166
\V Woe to you Scribes and Phariſees, Hypocrites: becauſe you make cleane
that on the outſide of the cup and diſh, but within, 
\Var{you are ful}{they are ful}
of rapine and vncleannes. \V Thou blind Phariſee, firſt make cleane the
inſide of the cup and the diſh that the outſide may become cleane.

\V Woe to you Scribes and Phariſees, Hypocrites: becauſe you are like to
whited ſepulchres, which outwardly appeare vnto men beautiful, but
within are ful of dead mens bones, and al filthines. \V So you alſo
outwardly indeed
\LNote{Appeare to men}{Chriſt might boldly reprehend them ſo often and
  ſo vehemently for hypocriſie, becauſe he knew their harts and
  intentions: but we that can not ſee within men, may not preſume to cal
mens external good doings, hypocriſie: but iudge of men as we ſee, and
know.}
appeare to men iuſt; but inwardly you are ful of hypocriſie and
iniquitie. 

\V Woe to you Scribes and Phariſees, Hypocrites: becauſe you build the
Prophets ſepulchres, and
\LNote{Garnish}{Chriſt blameth not the Iewes for adorning the ſepulchres
of the Prophets, but rebuketh them of their malice toward him, and of
that which by his diuine knowledge he foreſaw, that they would
accomplish the wickednes of their Fathers in sheeding his bloud, as
their Fathers did the bould of the Prophets.
\Cite{Hilar.}}
garniſh the monimẽts of iuſt men, \V and ſay: If we had been in our
Fathers dayes, we had not been their felowes
%%% 2349
in the bloud of the Prophets. \V Therfore you are a teſtimonie to your
owne ſelues, that you are the ſonnes of them that killed the
Prophets. \V And fil you vp the meaſure of your Fathers. \V You
ſerpents, vipers broods, how wil you flee from the iudgemẽt of Hel? \V
Therfore behold I ſend vnto you Prophets, and wiſe men, and Scribes, and
of them you ſhal kil & crucifie, and of them you ſhal ſcourge in your
Synagogues, and perſecute from citie into citie: \V that vpon you may
come al the iuſt bloud that was ſhed vpon the earth, from the bloud of
\CNote{\XRef{Gen.~4,~8.}}
Abel the iuſt, euen vnto the bloud of
\CNote{\XRef{2.~Par.~24,~22.}}
Zacharias the ſonne of Barachias,
whom you murdered between the temple & the Altar. \V Amen I ſay to you,
al theſe things ſhal come vpon this Generation. \V
\CNote{\XRef{Lu.~13,~34.}}
Hieruſalem,
Hieruſalem, which killeſt the Prophets, and ſtoneſt them that were ſent
to thee, how often would I gather togeather thy children as the henne
doth gather togeather her chickens vnder her wings, and thou
\SNote{Free wil.}
wouldeſt not? \V Behold, your houſe ſhal be left deſert to you. \V For I
ſay to you, you ſhal not ſee me from hence forth til you ſay: Bleſſed is
he that commeth in the name of our Lord.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
