%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Matthew
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 2493
%%% o-2105
\startChapter[
  title={Chapter 2}
  ]

\Summary{The Gentils come vnto Chriſt with their offerings, and that ſo
  openly that the Iewes can not pretend ignorance. 3.~The Iewes with
  Herode conſpire againſt him. 13.~He thereupon flyeth from them into
  Ægypt. 16.~They afterward, ſeeing their ſubteltie preuailed not,
  imagined to oppreſſe him by open perſecution. 19.~But they at length
  dyed and he returneth to the land of Iſrael: al according to the
  Scriptures.}

When \Sc{Iesvs} therfore was
\CNote{\XRef{Luc.~2.~7.}}
borne in Bethlehem of Iuda in the dayes of Herod the King,
\LNote{Behold}{Our Lords apparition or Epiphanie to theſe Sages, being
  Gentils, their Pilgrimage to him, and in them the firſt homage of
  Gentilitie done vnto him the twelfth day after his Natiuitie, and
  therfore is \Emph{Twelfth-day} highly celebrated in the Catholike
  Church for ioy of the calling of vs Gentils. His baptiſme alſo and
firſt miracle are celebrated on the ſame day.}
behold,there came Sages from the Eaſt to Hieruſalem, \V ſaying, where is he
that is borne King of the Iewes? For we haue ſeene his
\LNote{Starre}{Chriſts Natiuitie depended not vpon this ſtarre, as the
  Priſcillianiſts falſely ſurmiſed; but the ſtarre vpon his natiuitie,
  for the ſeruice whereof it was created.
\Cite{Grego. Ho.~10.}}
ſtarre in the Eaſt, and are
\LNote{Come to adore}{This
\MNote{Pilgrimage.}
coming ſo farre of deuotion to viſit and adore Chriſt in the place of
his birth, was properly a Pilgrimage to his Perſon; & warranteth the
Faithful in the like kind of external worſhip done to holy Perſons,
places, and things.}
come to adore him. \V And Herod the King hearing this, was troubled,
& al Hieruſalem with him. \V And aſſembling together al the high
Prieſtes and the Scribes of the people, he 
\LNote{Inquired of them}{The high Prieſts were rightly conſulted in
  queſtion of their law and religion; and be they neuer ſo il, are often
forced to ſay the truth by priuilege of their vnction; as here, and
after, they did concerning the true Meſſias.}
enquired of them where Chriſt
ſhould be borne. \V But they ſayd to him: In Bethlehem of Iuda. For ſo it
is written by the Prophet: \V
\CNote{\XRef{Mich.~5,~3.}}
\Emph{And thou Bethlehem the land of Iuda
  art not the leaſt among the Princes of Iuda: for out of thee shal come
forth the Capitaine that shal rule my people Iſrael.} \V Then Herod
ſecretly caling the Sages, learned diligently of them the time of the
ſtarre which appeared to them: \V and ſending them into Bethlehem, ſaid:
Goe, & inquire diligently of the childe; and when you ſhal finde him,
make reporte to me, that I alſo may come and adore him.

\V Who hauing heard the king, went their way; and behold the
\Fix{the }{}{obvious typo, fixed in other}% to prevent a paragraph break
ſtarre which they had ſeen in the Eaſt, went before them, vntil it came
and ſtood ouer, where the childe was. \V And ſeing the ſtarre they
reioyced with exceeding great ioy. \V And entring into the houſe, they found
the childe with \Sc{Marie} his mother, & faling downe
\LNote{Adored him}{This
\MNote{Adoration of the B.~Sacrament.}
body (ſayth S.~Chryſoſtom) the Sages adored in the crib. Let vs at the
leaſt imitate them; thou ſeeſt him not now in the crib, but on the
Altar; not a woman holding him, but the Prieſt preſent, and the
Holy-Ghoſt powred out aboundantly vpon the Sacrifice. 
\Cite{Ho.~24. in 1.~Cor.}
\Cite{Ho.~7. in Mt.}
\Cite{Ho. de ſancto Philogonio.}}
adored him; and opening their
\LNote{Treaſures}{Theſe treaſures are as it were the firſt fruits of
  thoſe riches, and guifts, which (according to the
\CNote{\XRef{Pſal.~75.}
\XRef{Eſa.~60.}}
Prophecies of Dauid
  and Eſay) Gentilitie ſhould offer to Chriſt and his Church, and now
  haue offered, ſpecially from the time of Conſtantine the Great. As
  alſo theſe three Sages, being principal men of their Countrie,
  repreſent the whole ſtate of Princes, Kings, and Emperours, that were
  (according to the ſaid Prophecies) to beleeue in Chriſt, to humble
  themſelues to his croſſe, to foſter, enrich, adorne and defend his
  Church. 
\MNote{The three kings.}
Whereupon it is alſo a very conuenient and agreable tradition
  of antiquitie, and a receiued opinion amõg the Faithful,
\CNote{\Cite{Chrys. Homil.~27. in Mat.}
\Cite{ho.~1. Theo. Cic. de Diuinat.}
\Cite{Plin. li.~20.}}
not lacking
  teſtimonies of ancient writers, and much for the honour of our
  Sauiour, that theſe three alſo were kings: to wit, either according to
the ſtate of thoſe Countries,
\CNote{\XRef{Eſther~1, 13,~14.}}
where the Princes were \Emph{Magi}; and
\Emph{Magi} the greateſt about the Prince; or as we read in the
Scriptures, of Melchiſedech King of Salem, & many other Kings that dwelt
within a ſmal compaſſe: or as
\CNote{\XRef{Tob.~2.}}
Iobes three friends are called
Kings. Theſe are commonly called the three kings of Colen, becauſe their
bodies are there, tranſlated thither from the Eaſt Countrie;
\CNote{\Cite{Amb.~1. off. c.~12.}}
their names
are ſaid to haue been Gaſpar, Melchior, Baltaſar.}
treaſures, they offered to him
\CNote{\XRef{Pſal.~71,~10.}}
\LNote{Guiftes}{The Sages were three, and their guifts three, and each
  one offered euerie of the three, to expreſſe our Faith of the
  Trinitie. The Gold, to ſignifie that he was a King, the frankincenſe,
  that he was God, the myrhh, that he was to be buried as man. 
\Cite{Aug. ſer.~1. de Epiph.}}
guiftes; gold, frankincenſe, and myrrhe. \V And hauing receiued an
anſwer in ſleepe that they ſhould not returne to Herod, they went backe
an other way into their contrey.

\V And after they were departed, Behold an Angel of our Lord appeared in
ſleepe to Ioſeph, ſaying: Ariſe, & take the childe and his mother, and
fly into 
\Fix{Ægipt;}{Ægypt;}{obvious typo, fixed in other}
and be ther vntil I ſhal tel thee. For it wil come to paſſe that Herod
wil ſeeke the childe to deſtroy him. \V Who aroſe, & tooke the childe
%%% o-2106
and his mother by night, and retired into Ægypt: and he was there vntil
the death of Herod: \V that it might be fulfilled which was ſpoken of
our Lord by the Prophet, ſaying:
\CNote{\XRef{Os.~11,~2.}}
\LNote{Out of Ægypt}{This place of the Prophet (and the like in
  the new Teſtament) here applied to Chriſt, wheras in the letter it
  might ſeeme otherwiſe; teacheth vs how to interprete the old
  Teſtament, and that the principal ſence is of Chriſt, and his Church.}
\Emph{Out of Ægypt haue I called my Sonne.}

%%% 2294
\V Then 
\MNote{The Martyrdõ of the holy \Emph{Innocents} whoſe holy-day, is kept
the 28 of December.}
Herod perceauing that he was deluded by the Sages, was exceeding
angrie; and ſending
\LNote{Murdered}{By
\MNote{Canonizing of Saints.}
this example we learne how great credit we owe to the Church in
Canonizing Saints, and celebrating their holy-daies: by whoſe only
warrant, without any word of Scripture, theſe holy 
\Fix{Inuocents}{Innocents}{obvious typo, fixed in other}
 haue been honoured for Martyrs, and their holy-day kept euer ſince the
 Apoſtles time, although they died not voluntarily, nor al perhaps
 circumciſed, and ſome the children of Pagans.
\Cite{Aug. ep.~28.}
\Cite{Orig. ho.~3. in diuerſos.}}
murdered al the men children that were in Bethlehem, & in al the borders
therof, from two yaere old and vnder, according to the time which he had
diligently ſought out of the Sages. \V Then was fulfilled that which was
ſpoken by Ieremie the Prophet ſaying: \V
\CNote{\XRef{Iere.~31,~15.}}
\Emph{A voice in Rama was
  heard, crying out & much wayling: Rachel bewayling her children, &
  would not be comforted, becauſe they are not.}

\V But when Herod was dead, behold an Angel of our Lord appeared in
ſleep to Ioſeph in Ægypt, \V ſaying: Ariſe, and take the child & his
mother, and goe into the land of Iſrael. For they are dead that ſought
the life of the child. \V Who aroſe, & took the child and his mother,
and came into the land of Iſrael. \V But hearing that Archelaus raigned
in Iewrie for Herod his father, he feared to goe thither: and being
warned in ſleep retyred into the quarters of Galilee. \V And coming he
dwelt in a citie called Nazareth: that it might be fulfilled which was
ſayd by the Prophetes: That he ſhal be called a Nazarite.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
