%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 36
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-36


\project douay-rheims


%%% 0416
%%% o-0373
\startChapter[
  title={Chapter 36}
  ]

\Summary{That the inheritances may not be alienated from one tribe to an
  other, al muſt marrie within their owne tribes.}

And the princes alſo of the families of Galaad, the ſonne of Machir, the
ſonne of Manaſſes of the ſtocke of the children of Ioſeph came: and
ſpake to Moyſes before the princes of Iſrael, and ſaid: \V Our Lord hath
commanded thee our lord that thou shouldeſt by lotte diuide the Land to
the children of Iſrael, and that to the daughters of Salphaad our
brother thou shouldeſt geue the poſſeſſion dew to their father: \V Whom
if men of an other tribe take to wiues, their poſſeſſion shal folow, and
being tranſlated to an other tribe, it shal be a diminishing of our
inheritance. \V And ſo it shal come to paſſe, that when the Iubilee,
that is the fiftith yeare of remiſſion is come, the
\LNote{Diſtribution of lottes.}{By
\MNote{Reſtraint in Mariages alſo for a temporal cauſe.}
reaſon of two former lawes, the one
\XRef{(Leuit.~25.)}
prouiding that inheritance of landes ſhould not be ſold, nor otherwiſe
alienated, but vntil the Iubilee yeare, and then returne to him, or his
heyres, to whom it pertained before; the other
\XRef{(Num.~27.)}
ordaining that for lack of a ſonne, daughters ſhould enherite, this
difficultie did riſe; in caſe an enheretrixe did marrie a man of an
other tribe, her landes by that meanes ſhould paſſe from tribe to tribe,
and not be reſtored in the Iubilee yeare. For auoiding of which
inconuenience a further law is made, that none shal marrie out of their
owne tribe.

Neuertheleſſe the tribe of Leui made mariages with the tribe of Iuda:
\CNote{\XRef{Luc.~1. v.~36.}}
as appeareth by that Zacharie the prieſt married Elizabeth coſin to our
B.~Ladie of the tribe of Iuda; though in the old Teſtament there is no
ſuch expreſſe diſpenſation, nor explication of the law, 
\MNote{Tradition.}
but by tradition was holden for lawful and practiſed by ſo holie a man
as Zacharie. And not without myſterie (as S.~Auguſtin noteth
\Cite{li.~2. c.~2. de conſen. Euang.})
\MNote{Chriſt a King and a Prieſt.}
for that Chriſt the Annointed of God, was prefigured by the annointing
of Kings, and Prieſts, and borne of the royal and prieſtlie tribes, being
both a King and a Prieſt.}
diſtribution of the lottes shal be confounded, and the poſſeſſion of one
shal paſſe to others. \V Moyſes anſwered the children of Iſrael, and as
our Lord commanded, ſaid: The tribe of the children of Ioſeph hath
ſpoken rightly. \V And this law is promulgated of our Lord
%%% 0417
touching the daughters of Salphaad: Let them marrie to whom they wil,
onlie that it be to the men of their owne tribe: \V leſt the poſſeſſion
of the children of Iſrael be mingled from tribe into tribe. For
\SNote{Al were not bound by this law to marrie, but al that would
marrie muſt contract within their owne tribe.}
al men shal marrie wiues of their owne tribe and kinred: \V and al wemen
ſhal take huſbandes of the ſame tribe: that the inheritance may remaine
in the families, \V and that the tribes be not mingled among themſelues,
but remaine ſo \V as they were ſeparated by our Lord. And the daughters
of Salphaad did as it had beene commanded: \V and Maala, and Therſa, and
Hegla, and Melcha, and Noa were married to the ſonnes of their vncle by
their father \V of the familie of Manaſſes, who was the ſonne of Ioſeph:
and the poſſeſſion, that had beene allotted to them, remained in the
tribe and familie of their father. \V Theſe are the
%%% o-0374
commandementes and iudgementes, which our Lord commanded by the hand of
Moyſes to the children of Iſrael, in the champion countries of Moab vpon
Iordan againſt Iericho.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
