%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 0348
%%% o-0313
\startChapter[
  title={Chapter 4}
  ]

\Summary{Diſtinct offices are aſsigned to the families of Aaron, 15.~of
  the other ſonnes of Caath, 21.~of Gerſon, 29.~and of Merari. 34.~Who
  are al numbred from the age of thirtie yeares to fiftie, and ſo
  imployed to their offices and burdens.}

%%% 0349
And our Lord ſpake to Moyſes, and Aaron, ſaying: \V Take the ſumme of
the ſonnes of Caath out of the middes of the Leuites, by their houſes
and families, \V from the thirtith yeare & vpward, vnto the fiftith
yeare, of al that goe in to ſtand and to miniſter in the tabernacle of
couenant. \V This is the ſeruice of the ſonnes of Caath: \V when the
campe is to remoue, Aaron and his ſonnes shal enter into the tabernacle
of couenant, and
\SNote{In this caſe it was neceſſarie for the prieſts to enter in, where
otherwiſe none entered but the high Prieſt, and that but once in the
yeare. And al being foulded vp, others alſo entered to carie it away.
\XRef{v.~15.}}
Sanctum ſanctorum, and ſhal take downe the veile that hangeth before the
dore, and shal wrappe in it the arke of teſtimonie,
%%% o-0314
\V and shal couer it agayne with a couer of Ianthine skinnes, and shal
ſpread ouer it a cloth al of hyacinth, and shal draw in the barres. \V
The tabel alſo of propoſition they shal wrappe in a cloth of hiacinth,
and shal put with it the cenſers and litle morters, the goblettes and
cuppes to powre the libaments: the breades shal be alwayes on it: \V and
they shal ſpread ouer it a cloth of ſcarlet, which againe they shal
couer with a veile of ianthine skinnes, and shal put in the barres. \V
They shal take alſo a cloth of hiacinth wherwith they shal couer the
candleſtick with the lampes and tongs therof and ſnuffers and al the
veſſels of oyle, which are neceſſarie for the dreſſing of the lampes: \V
and ouer al they shal put a couer of ianthine skinnes, and put in the
barres. \V Moreouer the golden altar alſo they shal wrappe in a cloth of
hyacinth, and ſhal ſpreade ouer it a couer of ianthine skinnes, and put
in the barres. \V Al the veſſel wherwith the miniſtration is done in the
Sanctuarie, they shal wrappe in a cloth of hiacinth and shal ſpread ouer
it a couer of ianthine skinnes, and put in the barres. \V But the altar
alſo they shal make cleane from the aſhes, and shal wrappe it in a
purple cloth, \V and shal put with it al the veſſel, that they vſe in
the miniſterie therof, that is to ſay, fire pannes, fleshhookes and
forkes, pothookes and shouels. Al the veſſel of the altar together they
shal couer with a veile of ianthine skinnes, and shal put in the
barres. \V And when Aaron & his ſonnes haue wrapped vp the Sanctuarie
and the veſſel therof in the remouing of the campe, then shal the ſonnes
of Caath enter in to carie the thinges wrapped vp: and they shal not
touch the veſſel of the Sanctuarie, leſt they die. Theſe are the burdens
of the ſonnes of Caath in the tabernacle of couenant: \V ouer whom shal
be Eleazar the ſonne of Aaron
%%% 0350
the prieſt, to whoſe charge perteyneth the oyle to dreſſe the lampes,
and the incenſe of compoſition, and the ſacrifice, that is alwayes
offered, and the oyle of vnction, and whatſoeuer perteyneth to the
ſeruice of the tabernacle, and of al the veſſel, that are in the
Sanctuarie. \V And our Lord ſpake vnto Moyſes and Aaron, ſaying: \V
\SNote{See that by your negligence thoſe that are next vnto you incurre
not offence, for ſo none shal be excuſed.}
Deſtroy not the people of Caath out of the middes of the Leuites: \V but
doe this to them, that they may liue, and not die, if they touch Sancta
ſanctorum. Aaron and his ſonnes shal enter, and they shal diſpoſe the
charges of euerie one, and shal diuide what euerie one muſt carie. \V
Let others by no curioſitie ſee the thinges that are in the Sanctuarie
before they be wrapped vp, otherwiſe they shal die. \V And our Lord
ſpake to Moyſes, ſaying: \V Take the ſumme of the ſonnes of Gerſon alſo
by their houſes and families and kinredes, \V from thirtie yeares and
vpward, vnto fiftie yeares. Number them al that goe in and miniſter
in the tabernacle of couenant. \V This is the office of the familie of
the Gerſonites, \V for to carie the curtines of the tabernacle, and the
roofe of the couenant the other couer, and ouer al the ianthine couer,
and the hanging that hangeth in the entrie of the tabernacle of
couenant, \V the curtines of the court, and the veile in the entrie that
is
%%% o-0315
before the tabernacle. Al thinges that pertayne to the altar, the
cordes, and veſſel of the miniſterie, \V shal the ſonnes of Gerſon
carie, by the commandment of Aaron & his ſonnes: and euerie one shal
know to what burden they muſt be aſſigned. \V This is the ſeruice of the
familie of the Gerſonites in the tabernacle of couenant, and they shal
be vnder the hand of Ithamar the ſonne of Aaron the prieſt. \V The
ſonnes of Merari alſo by the families and houſes of their fathers thou
shalt recken, \V from thirtie yeares and vpward, vntil fiftie yeares, al
that enter in to the office of their miniſterie, and to the ſeruice of
the couenant of teſtimonie. \V Theſe are their burdens: They shal carie
the bordes of the tabernacle and the barres therof, the pillers and the
feete of them, \V the pillers alſo of the court round about, with their
feete and pinnes and cordes. Al the veſſel and implementes they shal
receiue by account, and ſo shal carie them. \V This is the office of the
familie of the Merarites, and their miniſterie in the tabernacle of
couenant: and they shal be vnder the hand of Ithamar the ſonne of Aaron
the prieſt.
%%% 0351
Moyſes therfore and Aaron and the princes of the ſynagogue reckened the
ſonnes of Caath, by their kinredes and houſes of their fathers, \V from
thirtie yeares and vpward, vnto the fiftith yeare, al that enter in to
the miniſterie of the tabernacle of couenant: \V and they were found two
thouſand ſeuen hundred fiftie. \V This is the number of the people of
Caath that enter 
into the tabernacle of couenant: theſe did Moyſes and Aaron number
according to the word of our Lord by the hand of Moyſes. \V The ſonnes
of Gerſon alſo were numbered by the kinredes and houſes of their
fathers, \V from thirtie yeares and vpward, vnto the fiftith yeare, al
that enter in to miniſter in the tabernacle of couenant: \V and they
were found two thouſand ſix hundred thirtie. \V This is the people of
the Gerſonites, whom Moyſes and Aaron numbered according to the word of
our Lord. \V The ſonnes of Merari alſo were numbered by the kinredes and
houſes of their fathers, \V from thirtie yeares and vpward, vnto the
fiftith yeare, al that enter in to accomplish the rites of the
tabernacle of couenant: \V and they were found three thouſand two
hundred. \V This is the number of the ſonnes of Merari, whom Moyſes, and
Aaron reckened according to the commandment of our Lord by the hand of
Moyſes. \V Al that were reckened of the Leuites, and whom Moyſes and
Aaron and the princes of Iſrael tooke by name, by the kinredes and
houſes of their fathers, \V from thirtie yeares and vpward, vnto the
fiftith yeare, entring in to the miniſterie of the tabernacle, and to
carie the burdens, \V were in al eight thouſand fiue hundred eightie. \V
According to the word of our Lord did Moyſes recken them, euerie one
according to their office and burdens, as our Lord had commanded him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
