%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 32
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-32


\project douay-rheims


%%% 0408
%%% o-0366
\startChapter[
  title={Chapter 32}
  ]

\Summary{The children of Ruben and Gad, demanding to haue inheritance on
  that ſide Iordan, where they yet are, 7.~Moyſes iuſtly rebuketh them;
  16.~yet vpon condition that they wil march firſt ouer Iordan, and
  remaine the firſt in battle, til the land be ſubdued to the children
  of Iſrael, 28.~he granteth to them, and the halfe tribe of Manaſses,
  that which they requeſt.}

And the children of Ruben and Gad had much cattel, and they had in
beaſtes an infinite ſubſtance. And when they had ſeene Iazer and Galaad
fitte countries for to feede cattel, \V they came to Moyſes and Eleazar
the prieſt, and the princes of the multitude, and ſaid: \V Ataroth, and
Dibon, and Iazer, and Nemra, Heſebon, and Eleale, and Sabam, and Nebo,
and Beon, \V the land, which our Lord ſmote in the ſight of the children
of Iſrael, is of a verie plentiful ſoile for the feeding of beaſtes: and
we thy ſeruantes haue very much cattel: \V and we pray thee if we haue
found grace before thee, that thou geue vs thy ſeruantes the ſame in
poſſeſſion, and
\SNote{A right figure of thoſe that would poſſeſſe heauen without
labour or danger. \Emph{But none shal be crovvned vnles he ſtriue
lavvfully.}
\XRef{2.~Tim.~2.}}
make vs not paſſe ouer Iordan. \V To whom Moyſes anſwered: Why shal your
brethren goe to fight, and wil you ſitte here? \V Why ſubuert you the
mindes of the children of Iſrael, that they may not be bold to paſſe into
the place, which our Lord wil geue them? \V Did not your fathers ſo,
when I ſent from Cadeſbarne to view the Land? \V And when they were
come as farre as the Valley of cluſter, hauing viewed al the countrie,
they ſubuerted the hart of the children of Iſrael, that they should not
enter the
%%% 0409
coaſtes, which our Lord gaue them. \V Who being wrath ſware, ſaying: \V
If theſe men, that came vp out of Ægypt, from twentie yeares and vpward,
shal ſee the land, which vnder othe I promiſed to Abraham, Iſaac, and
Iacob: and they would not folow me, \V except Caleb the ſonne of Iephone
the Cenezeite, and Ioſue the ſonne of Nun: theſe
%%% o-0367
haue fulfilled my wil. \V And our Lord being wrath agaynſt Iſrael, led
him about through the deſert fourtie yeares, vntil the whole generation
was conſumed, that had done euil in his ſight. \V And behold, quoth he,
you are riſen vp in ſteed of your fathers, the offspring and diſciples
of ſinful men, to augment the furie of our Lord agaynſt Iſrael. \V But
if you wil not folow him, he wil leaue the people in the wilderneſſe,
and you shal be the cauſe of the death of al. \V But they coming nere,
ſaid: We wil make shepcottes, and ſtalles for our cattel, for our litle
ones alſo fenſed cities: \V and we our ſelues armed and girded wil march
on to battle before the children of Iſrael, vntil we bring them in vnto
their places. Our litle ones and whatſoeuer we can haue, shal be in
walled cities, becauſe of the lying of wayte of the inhabitantes. \V We
wil not returne into our houſes, vntil the children of Iſrael poſſeſſe
their inheritance: \V neither wil we ſeeke any thing beyound Iordan,
becauſe already we haue our poſſeſſion on the eaſt ſide therof. \V To
whom Moyſes ſaid: If you doe that you promiſſe, march on wel appointed
before our Lord to fight: \V and let euerie man of warre paſſe ouer
Iordan, vntil our Lord ſubuert his enemies, \V and al the Land be
ſubdued to him: then shal you be blameleſſe before our Lord and before
Iſrael, and you shal obteyne the countries that you would in the ſight
of our Lord. \V But if you doe not, that which you ſay, no man can doubt
but you ſinne againſt God: and know ye that you ſinne shal apprehend
you. \V Build therfore cities for your litle ones, and sheepecotes and
ſtalles for your sheepe and cattel: and accomplish that which you haue
promiſed. \V And the children of Gad and Ruben ſaid to Moyſes: We are
thy ſeruantes, we wil do that which our lord commandeth. \V We wil leaue
our litle ones, and our wiues, and sheepe, and cattel in the cities of
Galaad: \V and we thy ſeruantes al wel appoynted wil march on to the
warre, as thou my lord ſpeakeſt. \V Moyſes therfore commanded Eleazar
the prieſt, and Ioſue
%%% 0410
the ſonne of Nun, and the princes of the families by the tribes of
Iſrael, and he ſaid to them: \V If the children of Gad, and the children
of Ruben paſſe with you ouer Iordan al armed to the warre before our
Lord, and the Land be ſubdued to you: geue them Galaad in poſſeſſion. \V
But if they wil not paſſe armed with you into the Land of Chanaan, let
them take places to dwel in among you. \V And the children of Gad, and
the children of Ruben anſwered: As our lord hath ſpoken to his
ſeruantes, ſo wil we doe: \V our ſelues armed wil march before our Lord
into the Land of Chanaan, and we confeſſe that we haue already receiued
our poſſeſſion beyond Iordan. \V Moyſes therfore gaue to the children of
Gad and Ruben, and to the halfe tribe of Manaſſes the ſonne of Ioſeph,
the kingdome of Sehon king of the Amorrheite, and the kingdome of Og
king of Baſan, and their land and the cities therof round about. \V
Therfore the ſonnes of Gad built Dibon, and Ataroth, and Aroer, \V and
Etroth, and Sophan, and Iazar, and Iegbaa, \V and
%%% o-0368
Bethnemra, and Betharan, cities fenſed, and sheepecotes for their
sheepe. \V But the children of Ruben builded Heſebon, and Eleale, and
Cariathaim, \V and Nabo, and Baalmeon changing their names, Sabama alſo:
geuing names to the cities, which they had built. \V Moreouer the
children of Machir, the ſonne of Manaſſes went into Galaad, and waſted
it killing the Amorrheite the inhabites therof. \V Moyſes therfore gaue
the land of Galaad to Machir the ſonne of Manaſſes, who dwelt in it. \V
And Iair the ſonne of Manaſſes went, and tooke the villages therof,
which he called Hauoth Iair, that is to ſay, the Villages of Iair. \V
Nobe alſo went and tooke Canath with the villages therof: and he called
it by his owne name Nobe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
