%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 31
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-31


\project douay-rheims


%%% 0405
%%% o-0364
\startChapter[
  title={Chapter 31}
  ]

\Summary{The Madianites are ſlaine in battle, for that they had drawen
  the people of Iſrael to ſinne. 11.~The praye is brought to Moyſes,
  14.~vvho being angrie that the wemen (which were the occaſion of
  ſinne) are ſaued aliue, commandeth to kil al the male children, and al
  the wemen ſauing only virgins. 19.~The ſouldiars are purified, as alſo
  the praye, 26.~and then diuided among the victours, and other people,
  taking out portions for the Leuites. 48.~The princes of the hoſte geue
  free offerings to our Lord.}

And our Lord ſpake to Moyſes, ſaying: \V Reuenge firſt the children of
Iſrael on the Madianites, and ſo thou ſhalt be gathered to thy
people. \V And Moyſes forthwith ſaid: Arme of you men to fight, which
may take the
%%% 0406
reuenge of our Lord on the Madianites. \V Let a thouſand men of euerie
tribe be choſen of Iſrael to be ſent to the warre. \V And they gaue a
thouſand of euerie tribe, that is to ſay, twelue thouſand wel appointed
to fight: \V whom Moyſes ſent with Phinees the ſonne of Eleazar the
prieſt, and the holie veſſel, and the trumpettes to ſound he deliuered
to him. \V And when they had fought againſt the Madianites and had
ouercome, al the men they ſlew, \V and their kinges Eui, and Recem, and
Sur, and Hur, and Rebe, fiue princes of the Nation:
\SNote{VVith whõ he ſinned, with them he was alſo iuſtly puniſhed.}
Balaam alſo the ſonne of Beor they killed with the ſword. \V And they
tooke their wiues, and litle ones, and al their goodes: whatſoeuer they
had bene able to make, they ſpoiled: \V as wel cities as townes and
caſtels the fire did conſume. \V And they tooke the praye, and al
thinges that they had taken as wel of men as of beaſtes, \V and they
brought them to Moyſes, and Eleazar the prieſt, and to al the multitude
of the children of Iſrael, but the reſt of the implementes they caried to
the campe in the champion countries of Moab, beſide Iordan againſt
Iericho. \V And Moyſes and Eleazar the prieſt went forth, and al the
princes of the ſynagogue to meete them without the campe. \V And Moyſes
being angrie with the princes of the hoſte, the tribunes, and the
centurions that were come from the battle, \V ſaid: Why haue you
reſerued the wemen? \V Are not theſe they, that deceiued the children of
Iſrael at the ſuggeſtion of Balaam, and made you preuaricate againſt our
Lord vpon
\SNote{Theſe wemen bringing
\Fix{imagies}{images}{obvious typo, fixed in other}
of Beelphogor, cauſed the Iſraelites to offer ſacrifice to him, before
they ſhould committe fornication with them.}
the ſinne of Phogor, whereupon alſo the people was ſtroken? \V Therfore
kil al whatſoeuer is of the male ſexe, among
\SNote{In more deteſtation of the parents ſinne, God cõmanded to kil
theſe children & ſo they were preuented, from committing the like
crimes. But ordinarily, wemen & children ought not to be ſlaine after
the victorie.
\XRef{Deut.~20. v.~14.}}
the litle ones alſo: and the wemen, that haue knowen men in carnal
copulation, ſlea ye: \V but the gyrles and al the wemen that are virgins
reſerue to you: \V and
\Fix{tarying}{tarry}{likely typo, fixed in other}
without the campe ſeuen dayes.
%%% o-0365
He that hath killed a man, or touched him that is killed, ſhal be expiated
the third day and the ſeuenth. \V And of al the praye, whether it be
garment, or veſſel,
\Fix{and}{or}{possible typo, same in other}
ſome thing prepared for implementes, of goates
skinnes, and 
\Fix{heare,}{haire,}{likely typo, fixed in other}
and wood shal be expiated. \V Eleazar alſo the
prieſt ſpake to the hoſte, that had fought, in this maner: This is the
precept of the law, which our Lord hath commanded Moyſes: \V Gold, and
ſiluer, and braſſe, and yron, and leade, and tinne, \V and al that may
paſſe through the fyre, shal be purged by fyre, but whatſoeuer can not
%%% 0407
abide the fire, shal be ſanctified with the water of expiation: \V and
you shal wash your garmentes the ſeuenth day, and being purified,
afterward you shal enter into the campe. \V And our Lord ſaid to
Moyſes: \V Take the ſumme of thoſe thinges, that were taken from man
vnto beaſt, thou and Eleazar the prieſt and the princes of the
multitude: \V and thou shalt diuide the praye equally betwene them, that
fought & went forth to the warre, and betwen the reſt of the
multitude: \V and thou shalt ſeparate a portion to the Lord from them,
that fought and were in the battle, one ſoule of fiue hundred as wel of
mankind as of oxen and aſſes and sheepe, \V and thou shalt geue it to
Eleazar the prieſt, becauſe they are the firſt fruites of the Lord. \V
Out of the halfe part alſo of the children of Iſrael thou shalt take the
fiftith head of mankind, and of oxen, and aſſes, and sheepe, and of al
beaſtes, and thou shalt geue them to the Leuites, that watch in the
cuſtodies of the tabernacle of the Lord. \V And Moyſes and Eleazar did,
as our Lord had commanded. \V And the praye which the hoſte had taken,
was of sheepe ſix hundred ſeuentie fiue thouſand, \V of oxen ſeuentieo
two thouſand, \V of aſſes ſixtie one thouſand. \V The ſoules of the
folkes of the female ſexe, that had not knowne men, thirtie two
thouſand. \V And the halfe part was geuen to them, that had bene in the
battle, of sheepe three hundred thirtie ſeuen thouſand fiue hundred: \V
of the which for the portion of our Lord were reckened sheepe ſix
hundred ſeuentie fiue. \V And of the thirtie ſix thouſand oxen, oxen
ſeuentie and two: \V of the thirtie ſix thouſand oxen, oxen ſeuentie and
two: \V of the thirtie thouſand fiue hundred aſſes, aſſes ſixtie one: \V
of the ſoules of mankind ſixtene thouſand, there fel to the portion of
our Lord thirtie two ſoules. \V And Moyſes deliuered the number of the
firſt fruites of our Lord to Eleazar the prieſt, as it had bene
commanded him, \V of the halfe part of the children of Iſrael, which he
had ſeparated to them, that had bene in the battle. \V But of the halfe
part that had chanced to the reſt of the multitude, that is to ſay, of
the three hundred thirtie ſeuen thouſand four hundred ſheepe, \V and of
the thirtie ſix thouſand oxen, \V and of the thirtie thouſand fiue
hundred aſſes, \V and of the ſixtene thouſand of mankind, \V Moyſes
tooke the fiftith head, and gaue it to the Leuites, that watched in the
tabernacle of our Lord, as our Lord had commanded. \V And when the
princes
%%% 0408
of the hoſte were come to Moyſes, and the tribunes, and centurions, they
ſaid: \V We thy ſeruantes haue reckened the
%%% o-0366
number of the warryers, which we had vnder our hand: and not ſo much as
one verily was lacking. \V For this cauſe we offer in the donaries of
our Lord euerie one that which we could finde in the praye of gold,
garters, and tablettes, ringes and braceletes, and cheynes, that thou
maieſt pray to our Lord for vs. \V And Moyſes and Eleazar the prieſt
receiued al the gold in diuerſe formes, \V in weight ſixtene thouſand
ſeuen hundred fiftie ſicles of the tribunes and of the centurions. \V
For that which euerie one had ſpoiled in the praye, was his owne. \V And
that which was receiued they brought into the tabernacle of teſtimonie,
for a monument of the children of Iſrael before our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
