%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 0370
%%% o-0332
\startChapter[
  title={Chapter 14}
  ]

\Summary{The mutinous murmuring people being vnplacable, 11.~God
  expoſtulateth their ingratitude, threatneth to deſtroy them. 13.~Yet
  Moyſes pacifieth his wrath, 22.~but ſo that al which were numbered
  coming from Ægypt, except Caleb and Ioſue, shal die in the wildernes,
  31.~and their children shal poſſeſſe the promiſed land. 40.~Then
  fighting contrarie to Moyſes admonition are beaten, and manie ſlaine
  by their enimies.}

Therfore al the multitude crying out wept that night, \V and al the
children of Iſrael murmured againſt Moyſes and Aaron, ſaying: \V Would
God we had died in Ægypt: and
\SNote{Theſe murmurers had their wiſh, to their owne puniſhmẽt.
\XRef{chap.~14. v.~29.}
\XRef{26. v.~64.}}
in this vaſte wilderneſſe would God we might die, and that our Lord
would not bring vs into this Land, leſt we fal by the ſword, and our
wiues and children be ledde captiue. Is it not better to returne into
Ægypt? \V And one ſaid to an other:
%%% o-0333
\SNote{It is ſo abſolutly neceſſarie in euerie communitie to haue one
Superior of al, that verie mutiners themſelues do euer chooſe ſuch a
one, & cal him, the \L{Electo}.}
Let vs appoint a captaine, and let vs returne into Ægypt. \V Which
Moyſes and Aaron hearing fel
%%% 0371
flatte vpon the ground before al the multitude of the children of
Iſrael. \V But Ioſue the ſonne of Nun, and Caleb the ſonne of Iephone,
who them ſelues alſo had viewed the land, rent their garmentes, \V and ſpake
to al the multitude of the children of Iſrael: The Land, which we haue
circuted, is very good. \V If our Lord be propitious, he ſhal bring vs
into it, and deliuer vs a ground flowing with milke and honie. \V Be not
rebellious againſt our Lord: neither feare ye the people of this land,
for euen as bread ſo we may deuoure them. Al aide is gone from them:
our Lord is with vs, feare ye not. \V And when al the multitude cried,
and would haue ſtoned them, the glorie of our Lord appeared ouer the
roofe of couenant in the ſight of al the children of Iſrael. \V And our
Lord ſaid to Moyſes: How long wil this people detract me? How long wil
they not beleue me in al the ſignes, that I haue done before them? \V I
wil ſtrike them therfore with peſtilence, and wil conſume them: but thee
I wil make prince ouer a great nation, and a ſtronger then this is. \V
And Moyſes ſaid to our Lord: That the Ægyptians, from the middes of whom
thou haſt brought forth this people, \V and the inhabitantes of this
Land, which haue heard that thou Lord art among this people, and art
ſene face to face, and thy cloude protecteth them, and in a piller of
a cloude thou goeſt before them by day, and in a piller of fire by
night, \V may heare that thou haſt killed ſo great a multitude as it
were one man, and may ſay: \V He could not bring in the people into the
Land, for which he had ſworne: therfore did he kil them in the
wilderneſſe. \V Let therfore the ſtrength of our Lord be magnified as
thou haſt ſworne, ſaying: \V The Lord is patient and ful of mercie,
taking away iniquitie and wicked deedes, & leauing no man innocent,
which viſiteſt the ſinnes of the fathers vpon the children vnto the
third and fourth generation. \V Forgeue, I beſech thee, the ſinne of
this thy people, according to the greatnes of thy mercie, as thou haſt
bene propitious to them ſince their going out of Ægypt vnto this
place. \V And our Lord ſaid: I haue forgeuen it according to thy
word. \V Liue I: and the whole earth ſhal be replenished with the glorie
of the Lord. \V
\SNote{After the ſinne is forgeuen, yet puniſhment remaineth to be
inflicted.}
But yet al the men that haue ſene my maieſtie, and the ſignes that I
haue done in Ægypt, and in the wilderneſſe, and haue tempted me now ten
times, neither haue obeied my voice, \V they shal
%%% 0372
not ſee the Land for the which I ſware to their fathers, neither ſhal
any of them that hath detracted me, behold it. \V My ſeruant Caleb, who
\SNote{Although grace be firſt geuẽ without deſert: yet good workes done
by grace do merite reward.
\Cite{S.~Aug. de Grat. & lib. arb. c.~6.}}
being ful of an other ſpirit hath folowed me, wil I bring in vnto this
Land which he hath circuted: and his ſeede shal poſſeſſe it. \V Becauſe
the Amalecite and the Cananeite dwel in the valleis. To morrow remoue
the campe, and returne into the wilderneſſe by the way of the Redde
ſea. \V And our Lord ſpake to Moyſes and Aaron, ſaying: \V How long doth
this vngratious multitude murmur againſt me? I haue heard the
complaintes of the children of Iſrael.
%%% o-0334
\V Say therfore to them: Liue I, ſayeth our Lord: According as you haue
ſpoken I hearing it, ſo wil I do to you. \V In this wilderneſſe shal
your carcaſſes lie. Al you that are numbered from twentie yeares &
vpward, and haue murmured againſt me, \V you shal not enter into the
Land, ouer the which I haue lifted my hand to make you inhabite it,
except Caleb the ſonne of Iephone, and Ioſue the ſonne of Nun. \V But
your litle ones, of whom you ſaid, that they should be a pray to the
enemies, wil I bring in: that they may ſee the Land, that hath miſliked
you. \V Your carcaſſes shal lie in the wilderneſſe. \V Your children
shal wander in the deſert fourtie yeares, and
\SNote{Temporal puniſhment laid vpon the children for their fathers
ſinnes, is for their owne ſpiritual good.
\Cite{S.~Aug. Epiſt.~75. ad Auitum.}}
shal beare your fornication, vntil the carcaſſes of their fathers be
conſumed in the deſert, \V according to the number of the fourtie daies,
wherin you viewed the Land: a yeare shal be reputed for a day. And
fourtie yeares you shal receiue your iniquities, and shal know my
reuenge: \V for as I haue ſpoken, ſo wil I doe to al this wicked
multitude, that hath riſen togeather againſt me: in this wilderneſſe
shal it faile, and die. \V Therfore al the men, whom Moyſes had ſent to
view the Land, and which returning had made al the multitude to murmur
againſt him, detracting from the Land that it was naught, \V died and
were ſtroken in the ſight of our Lord. \V But Ioſue the ſonne of Nun,
and Caleb the ſonne of Iephone liued of al them, that had gone to view
the Land. \V And Moyſes ſpake al theſe wordes to al the children of
Iſrael, and the people mourned excedingly. \V And behold very early in
the morning riſing they went vp to the toppe of the mountaine, and ſaid:
We are readie to goe vp to the place, wherof our Lord hath ſpoken: for
we haue ſinned. \V To whom Moyſes ſaid: Why tranſgreſſe you the word of
our Lord, which shal not ſuceede
%%% 0373
proſperouſly with you? \V Goe not vp, for our Lord is not with you: leſt
you fal before your enemies. \V The Amelecite and the Chananeite are
before you, by whoſe ſword you ſhal fal, for that you would not conſent
to our Lord, neither wil our Lord be with you. \V But they being blinded
went vp to the toppe of the mountaine. But the arke of the teſtament of
our Lord & Moyſes departed not from the campe. \V And the Amalecite came
downe and the Cananeite, that dwelt in the mountaine: and ſtriking and
hewing them, purſewed them as farre as Horma.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
