%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 33
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-33


\project douay-rheims


%%% 0410
%%% o-0368
\startChapter[
  title={Chapter 33}
  ]

\Summary{The 42.~manſions of the children of Iſrael betwen Ægypt and the
  Land of promiſe are recited. 50.~They are commanded to kil al the
  inhabitents, to purge the land of idolatrie, and diuide it among them.}

Theſe are
\SNote{Theſe remouings, and camping places ſignifie by what degrees
chriſtians leauing ſinnes, and folowing Chriſt (our guide) may come to
perfect pietie.
\Cite{S.~Hierom. de manſ.~42. ad Fabiol.}}
the manſions of the children of Iſrael, that went out of Ægypt by their
troupes in the hand of
\LNote{Moyſes and Aaron.}{Moyſes
\MNote{Two handes which worke ſaluation.}
by whom the Law was geuen, and Aaron in whom Prieſthood was eſtabliſhed,
ſignified good workes, and the right worſhip of God. Of which (ſaieth
S.~Hierom) each one needeth the other. For neither doth it profite thee
to exerciſe vertues vnles thou knoweſt thy Creator: nor the worſhiping
of God auaileth thee to ſaluation, vnles thou fulfil the precepts of thy
Maker. By theſe two handes, as with two Seraphims we breake out into
confeſſion of the holie Trinitie, ſaying: Holie, holie, holie Lord God of
hoaſtes.
\Cite{Epiſt. ad Fabiol. de 42.~Manſ.}}
Moyſes and Aaron, \V which Moyſes deſcribed according to the places of
the campe, which by our Lords commandement they changed. \V The children
of Iſrael therfore departing
%%% 0411
from Rameſſes the firſt moneth, the fiftenth day of the firſt moneth, the
morow after they made the Phaſe, in a mightie hand, al the Ægyptians
ſeeing them, \V and burying their firſt borne, which our Lord had
ſtrooken (yea and on their goddes alſo he had exerciſed vengeance) \V
they camped in Soccoth. \V And from Soccoth they came into Etham, which
is in the vttermoſt borders of the wilderneſſe. \V Departing thence they
came ouer againſt Phihahiroth, which looketh toward Beelſephon, and
camped before Magdal. \V And departing from Phihahiroth, they paſſed
through the middes of the ſea into the wilderneſſe: and walking three
dayes through the deſert Etham, they camped in Mara. \V And departing
from Mara, they came into Elim, where there were the twelue fountaines
of waters, and the ſeuentie palme trees: and there they camped. \V But
departing thence alſo, they pitched their tentes vpon the Redde ſea. And
departing from the Redde ſea, \V they camped in the deſert Sin. \V From
whence departing, they came into Daphca. \V And departing from Daphca,
they camped in Alus. \V And departing from Alus, they pitched their
tentes in Raphidim, where the people wanted water to drinke. \V And
departing from Raphidim, they camped in the deſert of Sinai. \V But
departing alſo
%%% o-0369
from the deſert Sinai, they came to the Sepulchres of concupiſcence. \V
And departing from the Sepulchres of concupiſcence, they camped in
Haſeroth. \V And from Haſeroth they came into Rethma. \V And departing
from Rethma, they camped in Remmonphares. \V Whence departing they came
into Lebna. \V From Lebna they camped in Reſſa. \V And departing from
Reſſa, they came into Ceelatha. \V Whence departing, they camped in the
mountayne Sepher. \V Departing from the mountayne Sepher, they came into
Harada. \V Thence departing, they camped in Maceloth. \V And departing
from Maceloth, they came into Thahath. \V From Thahath they camped in
Thare. \V Whence departing, they pitched their tentes in Methca. \V And
from Methca they camped in Heſmona. \V And departing from Heſmona, they
came into Moſeroth. \V And from Moſeroth they camped in Beneiaacan. \V
And departing from Beneiaacan, they came into the mount Gadgad. \V
Whence departing, they camped in Ietebatha. \V And from Ietebatha they
came into Hebrona. \V And departing
%%% 0412
from Hebrona, they camped in Aſiongaber. \V Thence departing, they came
into the deſert Sin, this is Cades. \V And departing from Cades, they
camped in the mount Hor, in the vttermoſt borders of the Land of
Edom. \V And Aaron the prieſt went vp into the mountayne Hor at the
commandement of our Lord: and there he died the fourtith yeare of the
coming forth
\Fix{af}{of}{obvious typo, fixed in other}
the children of Iſrael out of Ægypt, the fifth moneth, the firſt day of
the moneth, \V when he was an hundred three and twentie yeares old. \V
And the Chananeite king of Arad, who dwelt toward the ſouth, heard that
the children of Iſrael were come into the Land of Chanaan. \V And
departing from the mountayne Hor, they camped in Salmona. \V Whence
departing, they came into Phunon. \V And departing from Phunon, they
camped in Oboth. \V And from Oboth they came into Iieabarim, which is in
the borders of the Moabites. \V And departing from Iieabarim they pitched
their tentes in Didongad. \V Whence departing, they camped in
Helmondeblathaim. \V And departing from Helmondeblathaim, they came to
the mountaynes Abarim againſt Nabo. \V And departing from the mountaynes
Abarim, they paſſed to the champion countries of Moab, von Iordan,
againſt Iericho. \V And there they camped from Bethſimoth vnto Abelſatim
in the plainer places of the Moabites, \V where our Lord ſpake to
Moyſes: \V Command the children of Iſrael, and ſay to them: When you
shal haue paſſed Iordan, entring the Land of Chanaan, \V deſtroy al the
inhabitants of that Land: breake their titles, and burſt to ſitters
their ſtatues, and waſt al their excelſes, \V cleanſing the Land, and
dwelling in it. For I haue geuen it you in poſſeſſion, \V which you shal
diuide among you by lotte. To the moe you shal geue the larger, and to
the fewer the ſtraiter. To euerie one as the lotte shal fall, ſo shal
the inheritance be geuen. By the tribes and the families the poſſeſſion
shal be diuided. \V But
\SNote{Danger to thoſe that deſtroy not infidelitie, and al enormous
ſinnes, which dwelt in their ſoules before their conuerſion.}
if you wil not kil the inhabitantes
%%% o-0370
of the Land: they that remaine, shal be vnto you as it were nailes in
your eyes, and ſpeares in your ſides, and they shal be your aduerſaries
in the Land of your habitation: \V and whatſoeuer I had thought to do to
them, I wil do to you.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
