%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 29
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-29


\project douay-rheims


%%% 0402
%%% o-0360
\startChapter[
  title={Chapter 29}
  ]

\Summary{In the ſeuenth moneth are celebrated with particular ſacrifices
  the feaſtes of Trumpets, 7.~of Expiation, 12.~of Tabernacles, 35.~and
  of Aſſemblie and Collection.}

The
\MNote{6.~Feaſt of Trumpets.}
firſt day alſo of the ſeuenth moneth ſhal be venerable and holie vnto
you, no ſeruile worke shal you doe therein, becauſe it is the day of
ſounding and of trumpetes. \V And you shal offer an holocauſt for a moſt
ſweete odour to the Lord, one calfe of the heard, one ramme, and ſeuen
lambes of a yeare old without blemish: \V and in their
%%% o-0361
ſacrifices of floure tempered with oile three tenthes to euerie calfe,
two tenthes to a ramme, \V one tenth to a lambe, which together are
ſeuen lambes: \V and a bucke goate for ſinne, which is offered for the
expiation of the people, \V beſides the holocauſt of the Calendes with
the ſacrifices therof, and the euerlaſting holocauſt with the accuſtomed
libamentes. With the ſame ceremonies shal you offer burnt ſacrifice for
a moſt ſweete odour to the Lord. \V The
\MNote{7.~Feaſt of Expiation.}
tenth day alſo of this ſeuenth moneth shal be holie and venerable vnto
you, and you shal
\SNote{For remiſſiõ of ſinnes the penitents did cooperate by penal
workes of faſting and abſtinence, \Emph{from euen vntil euen}.
\XRef{Leuit.~23. v.~32.}
\Cite{S.~Aug. q.~57. in Num.}
\Cite{Theod. q.~32. in Leuit.}}
afflict your ſoules: no ſeruile worke shal you doe therein. \V And you
shal offer an holocauſt to the Lord for a moſt ſweete odour, one calfe
of the heard, one ramme, ſeuen Lambes of a yeare old without blemish: \V
and in their ſacrifices of floure tempered with oile three tenthes to
euerie calfe, two tenthes to the ramme, \V the tenth part of a tenth to
euerie lambe, which are in al ſeuen lambes: \V and a bucke goate for
ſinne, beſides theſe thinges that are wont to be offered for offence
vnto expiation, and for the euerlaſting holocauſt with their ſacrifice &
libamentes. \V But the
\MNote{8.~Feaſt of Tabernacles.}
fiftenth day of the ſeuenth moneth, which shal be vnto you holie and
venerable, no ſeruile worke shal you doe therein, but you shal celebrate
the ſolemnitie to the Lord ſeuen daies. \V And you shal offer an
holocauſt for a moſt ſweete odour to the Lord, calues of the heard
thirtene, rammes two, lambes of a yeare old without blemish fourtene: \V
& in their libamentes
%%% 0403
of flowre tempered with oile three tenthes to euerie calfe, which are
together thirtene calues: and two tenthes to euerie ramme, that is,
together to two rammes, \V and the tenth part of a tenth to euerie
lambe, which are together fourtene lambes: \V and a bucke goate for
ſinne, beſide the euerlaſting holocauſte, and the ſacrifice, and the
libament therof. \V The next day you shal offer calues of the heard
twelue, two rammes, lambes a yeare old without blemish fourtene: \V
and the ſacrifices and libamentes of euerie one to the calues and the
rammes and the lambes you shal ritely celebrate: \V and a bucke goate
for ſinne, beſide the euerlaſting holocauſte, and the ſacrifice and
libament therof. \V The third day you shal offer eleuen calues, two
rammes, lambes of a yeare old without blemish fourtene: \V and the
ſacrifices and the libamentes of euerie one to the calues and the rammes
and the lambes you shal ritely celebrate: \V and a bucke goate for
ſinne, beſide the euerlaſting holocauſte, and the ſacrifice, and
libament therof. \V The fourth day you ſhal offer ten calues, two
rammes, lambes a yeare old without blemish fourtene: \V and the
ſacrifices and the libamentes of euerie one to the calues and the rammes
and the lambes you ſhal ritely celebrate: \V and a bucke goate for
ſinne, beſide the euerlaſting holocauſte, and the ſacrifice therof and
libament. \V The fifth day you shal offer nine calues, two rammes,
lambes of a yeare old without blemish fourtene: \V and the ſacrifices
and the libamentes of euerie one to the calues and the rammes and the
lambes you shal ritely celebrate: \V and a bucke goate for  ſinne,
beſide the euerlaſting holocauſte, and the ſacrifices
%%% o-0362
therof and libament. \V The ſixt day you shal offer eight calues, two
rammes, lambes of a yeare old without blemish fourtene: \V and the
ſacrifices and the libamentes of euerie one to the calues and the rammes
and the lambes you shal ritely celebrate: \V and a bucke goate for
ſinne, beſide the euerlaſting holocauſte, and the ſacrifice therof and
libament. \V The ſeuenth day you shal offer ſeuen calues, and two
rammes, lambes of a yeare old without blemish fourtene: \V and the
ſacrifices and the libamentes of euerie one to the calues and the rammes
and the lambes you shal ritely celebrate: \V and a bucke goate for
ſinne, beſide the euerlaſting holocauſte and the ſacrifice therof and
libament. \V The
\MNote{9.~Feaſt of Aſſemblie, and Collection.}
eight day, which is moſt ſolemne, no ſeruile
%%% 0404
%%% !!! only partially fixed in other, "shal" missing in other also
\Fix{you}{worke shal you}{obvious typo, fixed in other}
doe therin, \V offering an holocauſte for a moſt ſweete odour to the
Lord, one calfe, one ramme, lambes of a yeare old without blemish
ſeuen: \V and the ſacrifices and the libamentes of euerie one to the
calues and the rammes and the lambes you shal ritely celebrate: \V and a
bucke goate for ſinne, beſides the euerlaſting holocauſte, and the
ſacrifice therof and libament. \V Theſe thinges shal you offer to the
Lord in your
\Fix{ſolemnites:}{ſolemnities:}{obvious typo, fixed in other}
beſides your vowes and voluntarie oblations in holocauſte, in ſacrifice,
in libament, and in pacifique hoſtes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
