%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 25
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-25


\project douay-rheims


%%% 0394
%%% o-0354
\startChapter[
  title={Chapter 25}
  ]

\Summary{By carnal fornication manie are drawen to ſpiritual. For which
  twentie foure thouſand are ſlaine. 10.~Phinees his zele in ſtabbing to
  death two fornicators, is commended by God, and rewarded.}

And Iſrael at that time abode in Settim, and the people
\SNote{Balaam gaue this diueliſh counſel to allure men by bellichere and
lecherie vnto idolatrie.
\XRef{chap.~31. v.~16.}
And euen ſo Heretikes drawe manie in theſe daies to hereſie. See
\XRef{Apoc.~2. v.~14.}}
fornicated with the daughters of Moab, \V who called them to their
ſacrifices. And they did eate and adore their goddes. \V And Iſrael was
profeſſed to Beelphegor, and our Lord being angrie, \V ſaid to Moyſes:
Take al the princes of the people, and hang
\SNote{By aſſiſtance of the princes hang the idolaters.
\XRef{v.~5.}}
them vp againſt the ſunne on gibbettes: that my furie may be auerted
from Iſrael. \V And Moyſes ſaid to the Iudges of Iſrael: Kil euerie man
his neighbours, that are profeſſed to Beelphegor. \V And behold one of
the children of Iſrael entred in before the face of his brethren to a
whore a Madianite in the ſight of Moyſes, and of al the multitude of the
children of Iſrael, who wept before the doores of the tabernacle. \V
Which thing when Phinees had ſene the ſonne of Eleazar the ſonne of
Aaron the prieſt, he aroſe out of the middes of the multitude, and
catching a dagger, \V went in after the man of Iſrael into the brothel
houſe, and thruſt them through both together, to witte, the man and the
woman in the genitalles. And the plague ceaſed from the children of
Iſrael, \V and there were ſlaine fowre and twentie thouſand men. \V And
our Lord ſaid to Moyſes: \V Phinees the ſonne of Eleazar the ſonne of
Aaron the prieſt
\SNote{Either Phinees was one of the Iudges, and ſo an example for ſuch
Magiſtrates to do iuſtice; or he had a particular inſpiration, which is
no warrant for priuate men, to do the like. For by Catholique doctrin,
killing of men is neither allowed, nor excuſed but whẽ it is done by
publique authoritie, or by mere caſualitie, againſt the killers wil.
\Cite{Catechiſ. Rom. p.~3. c.~6. q.~5.}
\Cite{D.~Tho. 2.~2. q.~60. a.~6.}}
hath auerted my wrath from the children of Iſrael: becauſe he was moued
with my zele againſt them, that my ſelf
%%% 0395
might not deſtroy the children of Iſrael in mine owne zele. \V Therfore
ſpeake to him: Behold I geue him the peace of my couenant, \V and there
shal be as wel to him as to his ſeede the couenant of prieſthood for
euer, becauſe he hath bene zelous for his God, and hath expiated the
wicked fact of the children of Iſrael. \V And the name of the man of
Iſrael, that was ſlaine with the woman of Madian, was Zambri the ſonne
of Salu, a prince of the kinred and tribe of Simeon. \V Moreouer the
Madianeſſe, that was ſlaine with him, was called Cozbi the daughter of
Sur a moſt noble prince of the Madianites. \V And our Lord ſpake to
Moyſes, ſaying: \V Let the Madianites finde you their enemies, and
ſtrike you them: \V Becauſe they alſo haue done like enemies againſt
you, and haue guilfully deceiued you by the Idol Phogor, and Cozbi the
daughter of the duke of Madian their ſiſter, who was ſtrooken in the day
of the plague for the ſacrilege of Phogor.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
