%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 0386
%%% o-0347
\startChapter[
  title={Chapter 21}
  ]

\Summary{Iſraelites at the firſt encounter with the Chananeites hauing
  the worſe, after their vow kil the King of Arad, and deſtroy his
  cities. 4.~The people againe murmuring are ſtricken with firie
  ſerpents, 7.~but confeſsing their fault, Moyſes by Gods commandment
  ſetteth vp a braſen ſerpent for a remedie. 10.~They march through
  diuers places: 17.~and ſing a Canticle at a wel which God gaue
  them. 21.~They kil Sebon King of the Amorreites, and conquer his
  land. 33.~Likwiſe Og King of Baſan.}

VVhich when the Chananeite king of Arad, who dwelt toward the ſouth, had
heard, to wit, that Iſrael was come by the way of the ſpies, he fought
againſt them, and being victour, he tooke the pray of them. \V But,
Iſrael binding himſelf, by vow to our Lord, ſaid: If thou wilt deliuer
this people into my hand, I wil deſtroy their cities. \V And our Lord
heard the prayers of Iſrael, and deliuered the Chananeite, whom they
ſlew ouerthrowing their cities: and they called the name of that place
Horma, that is to ſay, Anathema. \V And they marched alſo from the
mountaine Hor, by the way that leadeth to the Redde ſea, that they might
compaſſe the land of Edom. And the people began to be wearie of the
iourney and labour: \V and ſpeaking againſt God and Moyſes, they ſaid:
Why didſt thou bring vs out of Ægypt, to die in the wilderneſſe? There
wanteth bread, waters there are none: our ſoule now lotheth at this moſt
light meate. \V Wherfore our Lord ſent vpon the people firie ſerpentes,
at whoſe plagues and the deathes of verie manie, \V they came to Moyſes,
and ſaid: We haue ſinned, becauſe we haue ſpoken againſt our Lord and
thee: Pray that he take from vs the ſerpentes. And Moyſes prayed for the
people, \V and our Lord ſpake to him:
\SNote{God forbidding images of idols, yet commandeth to make an image
for a good purpoſe.}
Make a braſen ſerpent, and ſette it for a ſigne: he that being ſtriken
looketh on it, ſhal liue. \V Moyſes therfore made
\SNote{Chriſt expoundeth this erecting of the braſen ſerpent, of him
ſelf to be crucified.
\XRef{Ioan.~3.}}
\Sc{a brasen serpent}, and ſet it for a ſigne: whom when they that
%%% 0387
were ſtriken, looked on, they were healed. \V And the children of Iſrael
marching camped in Oboth. \V Whence departing they pitched their tentes
in Ieabarim, in the wilderneſſe, that looketh toward Moab againſt the
eaſt part. \V And remouing from thence, they came to the Torrent
Zared. \V Which they forſaking camped againſt Arnon, which is in the deſert,
& ſtandeth out in the borders of the Amorrheite. For Arnon is the
border of Moab, diuiding the Moabites & the Amorrheites. \V Wherof it
is ſaid in the booke of the
%%% o-0348
warres of our Lord:
\SNote{As the Ægyptians were drowned in the ſea; ſo the Amorrheites were
oppreſſed by the rockes falling vpon them, & the waters caried their
carcaſes into the valley of Moab.}
As he did in the Redde ſea, ſo wil he doe in the ſtreames of Arnon. \V
The rockes of the torrentes were bowed, that they might reſt in Ar, and
lie in the borders of the Moabites. \V From that place appeared the wel,
wherof our Lord ſpake to Moyſes: Gather the people together, and I wil
geue them water. \V Then Iſrael ſang this verſe: Ariſe the wel. The ſang
therto: \V The wel, which the princes digged, and the captaines of the
multitude prepared in the lawgeuer, and in their ſtaues. And they
marched from the wilderneſſe to Mathana. \V From Mathana vnto Nahaliel:
from Nahaliel vnto Bamoth. \V From Bamoth is a valley in the countrie of
Moab, in the toppe of Phaſga, which looketh toward the deſert. \V And
Iſrael ſent meſſengers to Sehon King of the Amorrheites, ſaying: \V I
beſech thee that I may haue licence to paſſe through thy land: we wil
not goe aſide into the fieldes and the vineyardes, we wil not drinke
waters of the welles, we wil goe the kinges high way, til we be paſt thy
borders. \V Who would not grant that Iſrael ſhould paſſe by his borders:
but rather gathering an armie, went forth to meete them in the deſert,
and came vnto Iaſa, and fought againſt them. \V Of whom he was ſtrooken
in the edge of the ſword, and his land was poſſeſſed from Arnon vnto
Ieboc, and to the children of Ammon: for the borders of the Ammonites
were kept with a ſtrong garriſon. \V Iſrael therfore tooke al his
cities, and dwelt in the cities of the Ammorrheite, to wit, in Heſebon,
and the villages therof. \V The citie Heſebon was Sehons the king of the
Amorrheite, who fought againſt the king of Moab: and tooke al the land,
that had bene of his dominion, as farre as Arnon. \V Therfore it is ſaid
in the prouerbe: Come into Heſebon, let the citie of Sehon be built and
erected: \V A fire went forth from Heſebon, a flame from the towne of
Sehon, and
%%% 0388
deuoured Ar of the Moabites, and the inhabitantes of the high places of
Arnon. \V Wo to thee Moab, thou art vndone people of Chamos. He hath
geuen his ſonnes into flight, and his daughters into captiuitie to Sehon
the King of the Amorrheites. \V Their yoke is perished from Heſebon vnto
Dibon, they came wearie into Nophe, and vnto Medaba. \V Iſrael therfore
dwelt in the Land of the Amorrheite. \V And Moyſes ſent ſome to take a
view of Iazer: Whoſe villages they tooke, and poſſeſſed the
inhabitantes. \V And they turned them ſelues, and went vp by the way of
Baſan, and Og the King of Baſan came againſt them with al his people, to
fight in Edrai. \V And our Lord ſaid to Moyſes: Feare him not, for into
thy hand I haue deliuered him, and al his people, and land: and thou
shalt doe to him as thou didſt to Sehon the King of the Amorrheites, the
inhabiter of Heſebon. \V They therfore ſmote him alſo with his ſonnes,
and al his people vnto vtter deſtruction, and they poſſeſſed his land.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
