%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 0368
%%% o-0331
\startChapter[
  title={Chapter 13}
  ]

\Summary{From the deſert of Pharan Moyſes ſendeth twelue men (of euerie
  tribe one) to view the Land of Chanaan: 17.~changeth Oſee his name
  into Ioſue, 18.~inſtructeth them which way to goe, and what to note in
  the land: 22.~which they performe: 26.~and after fourtie dayes
  returne, bringing with them fruites, in token of the landes
  fertilitie. 29.~But in other reſpects (the reſt beſides Caleb and
  Ioſue) diſcouraging the people make them to murmur.}

%%% 0369
And the people marched from Haſeroth pitching their tentes in the deſert
of Pharan. \V And there our Lord ſpake to Moyſes, ſaying: \V Send men,
that may view the Land of Chanaan, which I wil geue to the children of
Iſrael, one of euerie tribe, of the princes. \V Moyſes did that which our
Lord had commanded, from the deſert of Pharan ſending principal men,
whoſe names be theſe. \V Of the tribe of Ruben, Samua the ſonne of
Zechur. \V Of the tribe of Simeon, Saphat the ſonne of Huri. \V Of the
tribe of Iuda, Caleb the ſonne of Iephone. \V Of the tribe of Iſſachar,
Igal the ſonne of Ioſeph. \V Of the tribe of Ephraim, Oſee the ſonne of
Nun. \V Of the tribe of Beniamin, Phalti the ſonne of Raphu. \V Of the
tribe of Zabulon, Geddiel the ſonne of Sodi. \V Of the tribe of Ioſeph,
of the ſcepter of Manaſſes Gaddi, the ſonne of Sufi. \V Of the tribe of
Dan, Ammiel the ſonne of Gemalli. \V Of the tribe of Aſer, Sthur the
ſonne of Michael. \V Of the tribe of Nephthali, Nahabi the ſonne of
Vapſi. \V Of the tribe of Gad, Guel the ſonne of Machi. \V Theſe are the
names of the men, whom Moyſes ſent to view the Land: and he called Oſee
the ſonne of Nun,
\SNote{Changing of his name literally imported the great office of chiefe
Duke vnto which he was deſigned, myſtically prefigured our
Lord \Sc{Iesvs}, for it is the ſame name in Hebrew, and
ſignifieth \Sc{Saviovr}.
\Cite{Theod. q.~25. in Num.}}
Ioſue. \V Moyſes therfore ſent them to view the Land of Chanaan, and
ſaid to them: Goe vp by the ſouth ſide. And when you ſhal come to the
mountaines, \V view the Land, what it is: and the people that are the
inhabitantes therof, whether they be ſtrong or weake: few in number or
manie: \V the land it ſelf, whether it be good or badde: what manner of
cities, walled or without walles: \V the ground, fatte or barren,
wooddie or without trees. Be of good courage, and bring vs of the
fruites of the Land. And it was the time when now the firſt ripe grapes
are to be eaten. \V And when they were gone vp, they viewed the Land
from the deſert of Sin, vnto Rohab as you enter to Emath. \V And they
went vp at the ſouth ſide, and came to Hebron, where were Achiman and
Siſai and Tholmai the ſonnes of Enac, for Hebron was built ſeuen yeares
before Tanis the citie of Ægypt. \V And going forward as farre as the
%%% o-0332
Torrent of cluſter, they cutte of a branch with the grapes therof, which
two men carried vpon a leauer. They tooke of the pomegranates alſo and
of the figges of that place: \V which was called Neheleſcol, that is to
ſay, the Torrent of cluſter, for that thence the children of Iſrael had
%%% 0370
caried a cluſter. \V And the diſcouerers of the Land returning after
fourtie daies, hauing circuted al the countrie, \V came to Moyſes &
Aaron and to al the aſſemblie of the children of Iſrael into the deſert
of Pharan, which is in Cades. And ſpeaking to them & to al the multitude
they ſhewed the fruites of the Land: \V and reported, ſaying: We came
into the Land to which thou didſt ſend vs, which in very deede floweth
with milke and honie, as by theſe fruites may be knowen: \V but it hath
very ſtrong inhabitantes, and cities great and walled. The ſtocke of Enac
we ſaw there. \V Amalec dwelleth in the ſouth, the Hetheite and the
Iebuſeite and the Amorheite in the mountaines: but the Chananeite
abideth beſide the ſea and about the ſtreames of Iordan. \V Among theſe
thinges Caleb appeaſing the murmuring of the people that roſe againſt
Moyſes, ſaid: Let vs goe vp and poſſeſſe the Land, becauſe we may
obtaine it. \V But the others, that had bene with him, ſaid: No, we are
not able to goe vp to this people, becauſe it is ſtronger then we. \V And
they detracted from the Land, which they had viewed, before the children
of Iſrael, ſaying: The Land, which we haue viewed,
\SNote{Pretending falſly that the Land had an vnwholſome
ayre, \Emph{deuouring the inhabitãtes}, & not poſſible to be obtained by
reaſõ of the gyants, couertly they detracted frõ Gods powre, or his good
wil towards them, who had promiſed the ſame. And therfore he gaue it to
their children, but not to theſe ſeducers and murmurers.
\XRef{chap.~14. v.~23,~29.}}
deuoureth her inhabitantes: the people, that we beheld, is of a tall
ſtature. \V There we ſaw certaine monſters of the ſonnes of Enac, of the
gyantes kind: to whom being compared, we ſeemed as it were locuſtes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
