%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 0361
%%% o-0325
\startChapter[
  title={Chapter 9}
  ]

\Summary{The precept of Paſch to be made the fourtenth day of the firſt
  moone, is renewed. 6.~But the vncleane, and trauelers in a iorney the
  fourtenth day of the ſecond moneth. 15.~The campe muſt rest or march,
  as the cloud or piller of fire, remaineth ouer the tabernacle, or
  departeth.}

And our Lord ſpake to Moyſes, in the deſert of Sinai the ſecond yeare,
after they went out of the land of
%%% 0362
Ægypt, the firſt moneth ſaying: \V Let the children of Iſrael make the
Phaſe in his time, \V the fourtenth day of this moneth at euen,
according to al the ceremonies and iuſtifications therof. \V And Moyſes
commanded the children of Iſrael that they should make the Phaſe. \V Who
made it in his time: the fourtenth day of the moneth at euen in mount
Sinai. According to al thinges that our Lord had commanded Moyſes the
children of Iſrael did. \V But behold certaine vncleane
\SNote{By touching the dead.
\Cite{S.~Aug. q.~15. in Num.}}
vpon the ſoule of man, which could not make the Phaſe on that day, coming
to Moyſes and Aaron, \V ſaid to them: We are vncleane vpon the ſoule of
man: Why are we defrauded that we can not offer the oblation to our Lord
in the dew time among the children of Iſrael? \V To whom Moyſes
anſwered: Stand that I may aske counſeil what our Lord wil command
concerning you. \V And our Lord
\SNote{God anſwered by a voice framed by an Angel from the
propitiatorie.
\XRef{chap.~7. v.~89.}}
ſpake to Moyſes, ſaying: \V Speake to the children of Iſrael: The man
that shal be vncleane vpon a ſoule, or in his iourney farre of in your
nation, let him make the Phaſe to the Lord \V in the ſecond moneth, the
fourtenth day of the moneth at euen: with azymes and wilde lettiſe shal
they eate it: \V they shal not leaue anie thing therof vntil morning,
and a bone therof they ſhal not breake, al the rite of the Phaſe they
ſhal obſerue. \V But if anie man both be cleane, and was not in his
iourney, and yet did not make the Phaſe, that ſoule shal be deſtroyed
out from among his peoples, becauſe he offered not ſacrifice to the
Lord in his dew time: he shal beare his ſinne. \V The ſeiourner alſo and
ſtranger if they be with you, shal make the Phaſe to the Lord, according
to the ceremonies and iuſtifications therof. The ſelfe ſame precept shal
be among you aſwel to the ſtranger, as to him that is borne in the
countrie. \V Therfore the day that the tabernacle was erected, a cloude
couered it. And from euening ouer the tabernacle there was as it were
the likeneſſe of fire vntil morning. \V So was it done alwaies: by day
the cloude couered it, and by night as it were the likeneſſe of fire. \V
And when the cloude that protected the tabernacle, had bene taken away,
then the children of Iſrael marched: and in the place where the cloude
had ſtoode, there they camped. \V At the commandment of
%%% o-0326
our Lord they marched, and at his commandment they pitched the
tabernacle. Al the daies that the
%%% 0363
cloude ſtoode ouer the tabernacle, they remained in the ſame place: \V
and if it chanced that it did continew ouer it a long time, the children
of Iſrael were 
in the watches of our Lord, and marched not \V for as manie daies ſoeuer
as the cloude had bene ouer the tabernacle. At the commandment of our
Lord they pitched their tentes, and at his commandement they tooke them
downe. \V If the cloude had bene from euen vntil morning, and immediatly
at day breake had forſaken the tabernacle, they marched: and if it had
departed after a day and a night, they tooke downe their tentes. \V But
if for two daies or one moneth or a longer time it had bene ouer the
tabernacle, the children of Iſrael remained in the ſame place, and
marched not: but immediatly as it had departed, they remoued the
campe. \V By the word of the Lord they pitched their tentes, and by his
word they marched: and were in the watches of our Lord according to his
commandment by the hand of Moyſes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
