%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 0365
%%% o-0328
\startChapter[
  title={Chapter 11}
  ]

\Summary{The people murmuring are punished with fire. 7.~Manna is againe
  deſcribed. 10.~Moyſes being afflicted with ſolicitude of trubleſome
  people, 16.~God adioyneth ſeuentie Ancients to ſuſtaine part of his
  burden. 18.~Promiſeth to geue al the people flesh. 25.~The ancients do
  prophecie. 31.~The people haue their fil of flesh, 33.~but forthwith
  manie die of the plague. Wherof the place is called, The ſepulchres of
  concupiſcence.}

In the meane time there aroſe a murmuring of the people, as it were
repyning for labour, againſt our Lord. Which when our Lord had heard, he
was angrie. And the fire of our Lord being
\Fix{kinled}{kindled}{obvious typo, fixed in other}
againſt them, deuoured the vttermoſt part of the campe. \V And when the
people had cried to Moyſes, Moyſes prayed to our Lord, and the fire was
quenched. \V And he called the name of that place, Kindling: for that
the fire of our Lord had bene kindled againſt them. \V For
\SNote{Theſe were Ægyptians that parted out of their countrie with the
Iſraelites, and now murmuring drawe others by example to the ſame ſinne.}
the common vulgar people, that came vp with them, burned with deſire,
ſitting and weeping, the children of Iſrael being ioyned together with
them, and ſaid: Who ſhal geue vs fleſh to eate? \V We remember the
fishes that we did eate in Ægypt gratis: the cucumbers come vnto our
minde, and the melons, and leekes and onions and garlike. \V Our ſoule
is drie, our eies behold nothing els by Manna. \V And the Manna was as
it were the ſeede of Coriander, of the colour of bdellion. \V And the
people went about, & gathering it, grounde
%%% 0366
it in a milne, or braied it in a morter, boyling it in a potte, and
making cakes therof of the taiſt as it were of oyled bread. \V And when
the dew fel in the night vpon the campe, the Manna alſo fel withal. \V
Moyſes therfore heard the people weeping by their families, euerie one
at the doores of his tent. And the furie of our Lord was exceding wrath:
but to Moyſes alſo it ſemed an intolerable thing. \V And he ſaid to our
Lord: Why haſt thou afflicted thy ſeruant? wherfore do I not finde grace
before thee? and why haſt thou laid the weight of al this people vpon
me? \V Haue I conceaued al this multitude, or begotten them, that thou
shouldeſt ſay to me: Carie them in thy boſome as the nource is wont to
carie the litle infant, and beare them into the land, for the which thou
haſt ſworne to their fathers? \V Whence shal I haue flesh to geue to ſo
great a multitude? they whine againſt me, ſaying: Geue vs flesh that we
may eate. \V I alone can not ſuſtaine al this people, becauſe it
%%% o-0329
is heauie for me. \V But
\SNote{Prayers of holie men are with ſubmiſſion of their willes to Gods
wil, either expreſſed or implied.}
if it ſeme vnto thee otherwiſe, I beſeche thee to kil me, and let me
finde grace in thine eies, that I be not moleſted with ſo great
euils. \V And our Lord ſaid to Moyſes: Gather me ſeuentie men of the
ancientes of Iſrael, whom thou knoweſt to be the ancientes of the people
and maiſters: and thou shalt bring them to the doore of the tabernacle
of couenant, and shalt make them to ſtand there with thee, \V that I may
deſcend and ſpeake to thee: and
\SNote{God
\Fix{imperted}{imparted}{obvious typo, fixed in other}
of the ſame ſpirite to theſe Ancientes, wherof he had geuẽ to Moyſes,
that they might haue ſo much helpe of grace as pleaſed God, and Moyſes
haue neuertheleſſe.
\Cite{S.~Aug. q.~18. in Num.}}
I wil take of thy ſpirit, and wil deliuer to them, that they may
ſuſtaine with thee, the burden of the people, and thou onlie be not
burdened. \V To the people alſo thou shalt ſay: Be ſanctified: to morrow
you ſhal eate fleſh: for I haue heard you ſay: Who ſhal geue vs meates
of flesh? it was wel with vs in Ægypt. That the Lord may geue you
fleſh, and you may eate: \V not one day, nor two, or fiue or ten, no nor
twentie, \V but euen to a moneth of daies, til it goe out
\Fix{a}{at}{likely typo, fixed in other}
your
noſethrils, and be turned to lothſomenes, becauſe you haue reiected the
Lord, who is in the middes of you, and haue whyned before him, ſaying:
Wherfore came we out of Ægypt? \V And Moyſes ſaid: There are ſix hundred
thouſand footemen of this people, & ſayeſt thou: I wil geue them flesh
to eate a whole moneth? \V Why, ſhal a multitude of ſheepe and oxen be
killed, that it may ſuffiſe for meate? or ſhal al the fishes of the ſea
be gathered togeather, for to fil them? \V To whom our Lord anſwered:
Why,
%%% 0367
is the hand of the Lord vnable? Now preſently thou shalt ſee whether my
word shal be accompliſhed in deede. \V Moyſes therfore came, and told
the people the wordes of our Lord, aſſembling ſeuentie men of the
ancientes of Iſrael, whom he cauſed to ſtand about the tabernacle. \V
And our Lord deſcended in a cloude, and ſpake to him, taking away of the
ſpirit that was in Moyſes, and geuing to the ſeuentie men. And when the
ſpirit had reſted on them, they prophecied, neither ceaſed they any
more. \V And there had remained in the campe two men, of the which one
was called Eldad, and the other Medad, vpon whom the ſpirit reſted
\SNote{Gods grace ſometimes preuenteth the ordinarie meanes.
\Cite{Theodoret. q.~21. in Num.}}
for they alſo had bene enrolled, and were not gone forth to the
tabernacle. \V And when they prophecied in the campe, there ran a boy,
and told Moyſes, ſaying: Eldad and Medad do prophecie in the campe. \V
Forthwith Ioſue the ſonne of Nun, the miniſter of Moyſes, and choſen of
manie, ſaid: My Lord Moyſes forbid them. \V But he ſaid: Why haſt thou
emulation for me? O that al the people might prophecie, and that our Lord
would geue them his ſpirit? \V And Moyſes returned, and the ancientes of
Iſrael into the campe. \V And a winde coming forth from our Lord, taking
quailes ouer the ſea brought them, and let them fal into the campe the
ſpace of one daies iourney, on euerie ſide of the campe round about, and
they did flie in the ayre two cubites high ouer the earth. \V The people
therfore riſing vp al that day, and the night, and the next day, gathered
togeather a multitude of quailes, he that did the leaſt, ten cores: and
they dried them round about the campe. \V As yet the flesh was in their
teeth,
%%% o-0330
neither had that kind of meate failed: and behold the furie of
our Lord being prouoked againſt the people, ſtroke them with an exceding
great plague. \V And that place was called,
\SNote{By this example and figure S.~Paul ſheweth that al ſhal not be
ſaued, which are baptiſed, & communicate in the ſame faith & Sacraments,
but thoſe only which alſo pleaſe God in their workes.
\XRef{1.~Cor.~10.}}
The Sepulchres of concupiſcence: for there they buried the people that
had luſted. And departing from the Sepulchres of concupiſcence, they
came vnto Haſeroth, and taried there.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
