%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 26
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-26


\project douay-rheims


%%% 0395
%%% o-0355
\startChapter[
  title={Chapter 26}
  ]

\Summary{Al the men of twelue tribes being againe numbered, from the
  age of twentie yeares and vpward, are found to be ſix hundred one
  thouſand ſeuen hundred and thirtie. 57.~Of the tribe of Leui alſo,
  numbered of the male ſex from the age of one moneth and vpward, are
  found twentie three thouſand: 64.~al being dead in the deſert, which
  were numbered before, except Caleb and Ioſue.}

After
\SNote{Gods wrath being pacified by extirpation of the old bad people,
the new progenie is numbered which ſhal poſſeſſe the promiſed land.
\Cite{S.~Hierom. de Manſ.~42. to.~3.}}
the bloud of the offenders was shed, our Lord ſaid to Moyſes and Eleazar
the ſonne of Aaron, the prieſt: \V Number the whole ſumme of the
children of Iſrael from twentie yeares and vpward, by their houſes and
kinredes, al that can goe forth to warres. \V Moyſes therfore and
Eleazar the prieſt ſpake, in the champion countrie of Moab vpon Iordan
againſt Iericho, to them that were \V from twentie yeares and vpward, as
our Lord had commanded, of whom this is the number: \V Ruben the firſt
borne of Iſrael, his ſonne, Henoch, of whom is the familie of the
Henochites: and Phallu, of whom is the familie of the Phalluites: \V and
Heſron, of whom is the familie of the Heſronites: and
Charmi, of whom is the familie of the Charmites. \V Theſe are the
families of the ſtocke of Ruben: whoſe number was found fourtie three
thouſand, and ſeuen hundred thirtie. \V The ſonne of Phallu, Eliab: \V
his ſonnes, Namuel and Dathan and Abiron. Theſe are Dathan and
%%% 0396
Abiron the princes of the people, that roſe againſt Moyſes and Aaron in
the ſedition of Core, when they rebelled againſt our Lord: \V and the
earth opening her mouth deuoured Core, verie manie dying, when the fire
burnt two hundred fiftie men. And there was a great miracle done, \V
that Core perishing,
\SNote{They were retained in the ayre, til the earth broken vnder them
was cloſed againe.}
his ſonnes perished not. \V The ſonnes of Simeon by their kinredes:
Namuel, of him is the familie of the Namuelites: Iamin, of him is the
familie of the Iaminites: Iachin, of him is the familie of the
Iachinites: \V Zare, of him is the familie of the Zareites: Saul, of him
is the familie of the Saulites. \V Theſe are the families of the ſtocke
of Simeon, of which the whole number was two and twentie thouſand two
hundred. \V The ſonnes of Gad by their kinredes: Sephon, of him is the
familie of the Sephonites: Aggi, of him is the familie of the Aggites:
Suni, of him is the familie of the Sunites: \V Ozni, of him is the
familie of the Oznites: Her, of him is the familie of the Herites: \V
Arod, of him is the familie of the Arodites: Ariel, of him is the
familie of the Arielites. \V Theſe are the families of Gad, of which the
whole number was fourtie thouſand fiue
%%% o-0356
hundred. \V The ſonnes of Iudas, Her, and Onan, who died both in the
land of Chanaan. \V And the ſonnes of Iudas by their kinredes were:
Sela, of whom is the familie of the Selaites: Phares, of whom is the
familie of the Phareſites: Zare, of whom is the familie of the
Zareites. \V Moreouer the ſonnes of Phares: Heſron, of whom is the
familie of the Heſronites: and Hamul, of whom is the familie of the
Hamulites. \V Theſe are the families of Iudas, of which the whole number
was ſeuentie ſix thouſand fiue hundred. \V The ſonnes of Iſſachar, by
their kinredes: Thola, of whom is the familie of the Tholaites: Phua, of
whom is the familie of the Phuaites: \V Iaſub, of whom is the familie of
the Iaſubites: Semran, of whom is the familie of the Semranites. \V
Theſe are the kinredes of Iſſachar, whoſe number ſixtie fowre thouſand
three hundred. \V The ſonnes of Zabulon by their kinredes: Sared, of
whom is the familie of the Saredites: Elon, of whom is the familie of
the Elonites: Ialel, of whom is the familie of the Ialelites. \V Theſe
are the kinredes of Zabulon, whoſe number was ſixtie thouſand fiue
hundred. \V The ſonnes of Ioſeph by their kinredes, Manaſſes and
Ephraim. \V Of Manaſſes was borne Machir, of whom is the
%%% 0397
familie of the Machirites. Machir begat Galaad, of whom is the familie
of the Galaadites. \V Galaad had ſonnes: Iezer, of whom is the familie
of the Iezerites: and Helec, of whom is the familie of the Helecites: \V
and Aſriel, of whom is the familie of the Aſrielites: and Sechem, of
whom is the familie of the Sechemites: \V and Semida, of whom is the
familie of the Semidaites: and Hepher, of whom is the familie of
the Hepherites. \V And Hepher was the father of Salphaad, who had no
ſonnes, but onlie daughters, whoſe names are theſe: Maala, and Noa, and
Hegla, and Melcha, and Terſa. \V Theſe are the families of Manaſſes, and
the number of them is fiftie two thouſand ſeuen hundred. \V And the ſonnes
of Ephraim by their kinredes were theſe: Suthala, of whom is the familie
of the Suthalaites: Becher, of whom is the familie of the Becherites:
Thehen, of whom is the familie of the Thehenites. \V Moreouer the ſonne
of Suthala was Heran, of whom is the familie of the Heranites. \V Theſe
are the kinredes of the ſonnes of Ephraim: whoſe number was thirtie two
thouſand fiue hundred. \V Theſe are the ſonnes of Ioſeph by their
families. The ſonnes of Beniamin in their kinredes: Bela, of whom is the
familie of the Belaites: Aſbel, of whom is the familie of the Asbelites:
Ahiram, of whom is the familie of the Ahiramites: \V Supham, of whom is
the familie of the Suphamites: Hupham, of whom is the familie of the
Huphamites. \V The ſonnes of Bela: Hered, and Noeman. Of Hered, the
familie of the Heredites: of Noeman, the familie of the Noemanites. \V
Theſe are the ſonnes of Beniamin by their kinredes, whoſe number was
fourtie fiue thouſand ſix hundred. \V The ſonnes of Dan by their
kinredes: Suham, of whom is the familie of the Suhamites: theſe are the
kinredes of Dan by their families. \V Al were Suhamites, whoſe number
was ſixtie foure thouſand foure hundred. \V The ſonnes of
%%% o-0357
Aſer by their kinredes: Iemna, of whom is the familie of the Iemnaites:
Ieſſui, of whom is the familie of the Ieſſuites: Brie, of whom is the
familie of the Breites. \V The ſonnes of Brie: Heber, of whom is the
familie of the Heberites: and Melchiel, of whom is the familie of the
Melchietes. \V And the name of the daughter of Aſer, was Sara. \V Theſe
are the kinredes of the ſonnes of Aſer, and their number fiftie three
thouſand foure hundred. \V The ſonnes of Nephthali by their kinredes:
Ieſiel, of
%%% 0398
whom is the familie of the Ieſielites: Guni, of whom is the familie of
the Gunites: \V Ieſer, of whom is the familie of the Ieſerites: Sellem,
of whom is the familie of the Sellemites. \V Theſe are the kinredes of
the ſonnes of Nephthali by their families: whoſe number was fourtie fiue
thouſand foure hundred. \V This is the ſumme of the children of Iſrael,
that were reckened, ſix hundred thouſand, and a thouſand ſeuen hundred
thirtie. \V And our Lord ſpake to Moyſes
ſaying: \V To theſe shal the land be diuided according to the number of
names for their poſſeſſions. \V To the greater number thou shalt geue a
greater portion, and to the fewer a leſſe: to euerie one, as they haue
now beene reckened, ſhal poſſeſſion be deliuered: \V yet ſo that lotte
doe diuide the Land to the tribes and the families. \V Whatſoeuer ſhal
chance by lotte, that let either the more take, or the fewer. \V This
alſo is the number of the ſonnes of Leui by their families: Gerſon, of
whom the familie of the Gerſonites: Caath, of whom the familie of the
Caathites: Merari, of whom the familie of the Merarites. \V Theſe are
the families of Leui: The familie of Lobni, the familie of Hebroni, the
familie of Moholi, the familie of Muſi, the familie of Core. Howbeit
Caath begatte Amram: \V who had to wife Iochabed
\SNote{Amram married his aunt: which ſheweth that the ſecond degree in
conſanguinitie is not forbid by the law of nature; though it was after
prohibited by a poſitiue law.
\XRef{Leuit.~18.}}
the daughter of Leui, who was borne to him in Ægypt. She bare to Amram
her husband ſonnes, Aaron and Moyſes, and Marie their ſiſter. \V Of
Aaron were borne Nadab and Abiu, and Eleazar and Ithamar: \V of the
which Nadab and Abiu died, when they had offered the ſtrange fyre before
our Lord. \V And al that were numbered, were twentie three thouſand of
the male kind from one moneth and vpward: who were not reckened among
the children of Iſrael, neither was their poſſeſſion geuen with the
reſt. \V This is the number of the children of Iſrael, that were
enrolled by Moyſes and Eleazar the prieſt, in the champion countrie of
Moab vpon Iordan againſt Iericho. \V Among whom there was none of them
that were numbered before by Moyſes and Aaron in the deſert of Sinai. \V
For our Lord had foretold that al ſhould die in the wilderneſſe. And
none remained of them, but Caleb the ſonne of Iephone, and Ioſue the
ſonne of Nun.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
