%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 35
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-35


\project douay-rheims


%%% 0414
%%% o-0371
\startChapter[
  title={Chapter 35}
  ]

\Summary{Cities and ſuburbes are appointed for the Leuites, among the
  other tribes. 6.~Of which  ſix shal be for refuge of ſuch as
  vnwittingly kil anie man, 22.~where, keping them ſelues til the death
  of the high Prieſt, they shal be ſafe. 30.~VVilful murtherers
  conuicted (ſo be it by more then one witnes) muſt die forthwith.}

Theſe thinges alſo ſpake our Lord to Moyſes in the champion countries of
Moab vpon Iordan, againſt Iericho: \V Command the children of Iſrael
that they geue vnto the Leuites of their poſſeſſions \V cities to dwel
in, and their ſuburbes round about: that them ſelues may abide in the
townes, and the ſuburbes may be for their cattel and beaſtes:
%%% o-0372
which from the walles of the cities outward,
%%% 0415
round about, ſhal reach the ſpace of a thouſand paces. \V Toward the
Eaſt shal be two thouſand cubites: and toward the South in like manner
ſhal be two thouſand: toward the ſea alſo, which looketh to the Weſt,
ſhal be the ſame meaſure: and the North part ſhal be limitted with equal
boundes. And the cities ſhal be in the middes, and the ſuburbes
without. \V And of the ſame townes, which you ſhal geue to the Leuites,
ſix ſhal be ſeparated for the ayde of fugitiues, that he may flee to
them which hath shed bloud: and beſide theſe other fourtie two
townes, \V that is, in al fourtie eight with their ſuburbes. \V And thoſe
cities, that shal be geuen of the poſſeſſions of the children of Iſrael,
from them that haue more, moe shal be taken away: and that haue leſſe,
fewer. Eache shal geue townes to the Leuites according to the meaſure of
their inheritance. \V Our Lord ſaid to Moyſes: \V Speake to the children
of Iſrael, & thou shalt ſay to them: When you shal haue paſſed ouer
Iordan into the Land of Chanaan, \V determine what cities shal be for
the ſuccour of fugitiues, that haue not voluntaryly shed bloud: \V in
which when the fugitiue shal be, the kinſeman of him that is killed can
not kil him, vntil he ſtand in the ſight of the multitude, and his cauſe
be iudged. \V And of thoſe cities, that are ſeparated for refuge of
the fugitiues, \V three shal be beyond Iordan, and three in the Land of
Chanaan, \V as wel for the children of Iſrael as for ſtrangers and
ſoiourners, that he may flee to them, which hath not voluntaryly shed
bloud. \V If any man ſtrike with yron, and he die that was ſtrooken, he
shal be guiltie of murder, and him ſelf shal die. \V If he caſt a ſtone,
and he that is ſtrooken die: he ſhal be punished in like manner. \V If
he that is ſtrooken with wood die, he ſhal be reuenged by the
ſtrikers bloud. \V The kinſman of him that is ſlaine, shal kil the
murderer,
\SNote{Yet not before ſentence of death.
\XRef{v.~12. &~24.}
\Cite{S.~Aug. q.~65. in Num.}}
as ſowne as he apprehendeth him, he shal kil him. \V If through hatred one
push a man, or caſt any thing at him by ſtratageme: \V or being his
enemie, ſtrike him with his hand, and he die: the ſtriker shal be
guiltie of murder. The kinſeman of him that is ſlaine as ſowne as he
findeth him, shal kil him. \V But if by chance medlie, and without
hatred, \V and enmitie, he do any of theſe thinges, \V and this be
proued in the hearing of the people, and the queſtion debated betwen him
that ſtroke, & the next of bloud: \V the innocent shal be deliuered
%%% 0416
from the hand of the reuenger, & by ſentence shal be brought backe into
the citie, to which he had fled, and he shal tarie there,
\SNote{Myſtically this ſignified, that the way to true life was not open
before Chriſts death.
\Cite{Theodor. q.~51. in Num.}
\Cite{S.~Greg. ho.~6. in Ezechielem.}}
vntil the High prieſt, that is anointed with the holie oile, do die. \V
If the murderer be found without the limites of the cities, that are
deputed to the banished, \V and be ſtrooken of him that is the reuenger
of bloud: he shal be guiltles that killed him. \V For the fugitiue ought
to haue ſtayed in the citie vntil the death of the High prieſt. And
after he is dead, then shal the murderer returne into his
countrie. \V Theſe thinges shal be perpetual, and for an ordinance in al
your habitations. \V The
%%% o-0373
murderer shal be punished by witneſſes: none shal be condemned, at the
teſtimonie of one man. \V You shal not take a price of him, that is
guiltie of bloud, he alſo shal die forthwith. \V The banished and
fugitiues before the death of the High prieſt may by no meanes returne
into their owne cities. \V Pollute not the land of your habitation,
which is polluted with the bloud of innocentes: neither can it
otherwiſe be expiated, but by his bloud, that shedeth an other mans
bloud. \V And ſo shal your poſſeſſion be cleanſed my ſelf abyding with
you. For I am the Lord that dwel among the children of Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
