%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 0351
%%% o-0316
\startChapter[
  title={Chapter 5}
  ]

\Summary{Lepers and al polluted perſons muſt be caſt out of the
  campe. 5.~Confeſſion of ſinne, and ſatisfaction for
  treſpaſse. 9.~Firſt fruites and oblations pertaine to the
  Prieſtes. 11.~The law of ieloſie.}

And our Lord ſpake to Moyſes, ſaying: \V Command the children of Iſrael,
that they caſt out of the campe
\SNote{Stil by the leſſe (ſaieth Theodoret) God inſtructeth in the
greater.
\Cite{q.~8. in Nũ.}
If therfore lepers were caſt out of the campe, how much more iuſtly are
heretiques caſt out of the Church?}
euerie leper, and whoſoeuer hath a fluxe of ſeede, and is polluted vpon
the dead: \V as wel man as woman caſt yee out of the campe, leſt when
they shal dwel with you, they
%%% 0352
contaminate it. \V And the children of Iſrael did ſo, and they did caſt
them forth without the campe, as our Lord had ſpoken to Moyſes. \V And
our Lord ſpake to Moyſes, ſaying: \V Speake to the children of Iſrael:
man, or woman, when they ſhal do any of al the ſinnes, that are wont to
chance to men, and by negligence haue tranſgreſſed the commandement of
the Lord, and haue offended, \V they
\LNote{shal confeſſe their ſinne.}{General
\MNote{Particular confeſſion of ſinnes, & ſatisfaction, required by the
law of God.}
confeſſion (ſuch as the Proteſtantes make) ſufficed not here for purging
ſinnes: but whoſoeuer tranſgreſſed anie of Gods commandments, were bound
by this diuine poſitiue law, to confeſſe expreſly and
diſtinctly \Emph{their ſinne}, which in particular they had
committed. Alſo to make reſtitution, if wrong were donne to anie other,
with a fifth part aboue the principal. And for further ſatisfaction to
God they muſt offer ſacrifice. Al which did plainly prefigure & foreſhew
the neceſſitie of particular confeſſion of ſinnes, and ſatisfaction, in
the Sacrament of Penance, inſtituted by Chriſt.
\XRef{Ioan.~20.}}
shal confeſſe their ſinne, and reſtore the principal it ſelf, and the
fifth part ouer to him, againſt whom they ſinned. \V But if there
be none to receiue it, they ſhal geue it to the Lord, and it ſhal be the
prieſtes, the ramme excepted, that is offered for expiation, to be a
placable hoſte. \V Al the firſt fruites alſo, which the children of
Iſrael doe offer, pertaine to the prieſt: \V and whatſoeuer is offered
into the Sanctuarie of euerie one, and is deliuered to the handes of the
prieſt, it shal be his. \V And our Lord ſpake to Moyſes, ſaying: \V
Speake to the children of Iſrael, and thou shalt ſay to them: The man,
whoſe wife erreth, and contemning her huſband \V hath ſlept with an
other man, and her huſband could not find it, but the adulterie is
ſecrete, and can not be proued  by witneſſes, becauſe ſhe was not found
in the adulterous fact: \V
\SNote{God ordained this law, and miraculouſly concurred therin, to
auoid
\Fix{wiueſſlaughter}{wiueſlaughter,}{obvious typo, fixed in other}
vpõ vehement ieloſie.
\Cite{Theod. q.~10. in Num.}}
if the ſpirit of ielouſie ſtirre vp the husband againſt his wife, which
either is polluted, or is charged with falſe ſuſpition, \V he shal bring
her to the prieſt, and shal offer an oblation for her the
\SNote{about the eight part of our peck.}
tenth part of a ſatum of barley meale: he shal not powre oile theron,
nor put frankincenſe vpon it: becauſe it is a ſacrifice of ielouſie, and
an oblation ſearching out adulterie. \V The prieſt therfore shal offer
it, and ſet it before the Lord. \V And he shal take
\SNote{VVater ſactified by ſpecial rites is called holie water, and
ſerueth to holie vſe.
\XRef{chap.~8. v.~7.}
\XRef{c.~19. v.~9.}}
holie water in an earthen veſſel, and he shal caſt a little grauel of
the pauement of the tabernacle into it. \V And when the woman shal ſtand
in the ſight of the Lord, he shal vncouer her head, and shal put vpon
her handes the ſacrifice of recordation, and the oblation of ielouſie:
and him ſelfe shal hold the moſt bitter waters, wheron he heaped curſes
with execration. \V And he shal adiure her, and shal ſay: If an other
man hath not ſlept
%%% o-0317
with thee, and if thou be not polluted by forſaking thy huſbandes bedde,
theſe moſt bitter waters shal not hurt thee, wherupon I haue heaped
curſes. \V But if thou haſt declined from thy huſband, & art polluted,
and haſt lien with an other man: \V thou shalt be ſubiect to theſe
maledictions: Our Lord geue thee for a
%%% 0353
malediction, and an example of al among his people: make he thy thigh to
rotte, and bellie ſwelling burſt aſunder. \V The
\SNote{The water wheron the prieſt laide curſes to light on the woman if
ſhe were guiltie.}
curſed water enter into thy bellie, and thy wombe being ſwolne let thy
thigh rotte. And the woman shal anſwer, Amen, amen. \V And the prieſt
shal write theſe curſes in a booke, and shal wash them out with the moſt
bitter waters, wherupon he heaped the curſes, \V and he shal geue them
her to drinke. Which when she hath drunke vp, \V the prieſt shal take of
her hand the ſacrifice of ielouſie, and shal eleuate it before the Lord,
and shal put it vpon the altar: yet ſo notwithſtanding that firſt, \V he
take a handful of the ſacrifice of that, which is offered, & burne it
vpon the altar: and ſo geue the moſt bitter waters to the woman to
drinke. \V Which when she hath drunke, if she be polluted, and by
contempt of her huſband guiltie of adulterie, the waters of malediction
shal goe through her, and her bellie being puft vp her thigh shal rotte
withal: and the woman shal be for a malediction, and an example to al
the people. \V But if she be not polluted, she shal be blameleſſe, and
shal beare children. \V This is the law of ielouſie. If the woman
decline from her huſband, and if she be polluted, \V and the huſband
ſtirred with the ſpirit of ielouſie brought her in the ſight of the
Lord, and the prieſt haue done to her according to al thinges that are
written: \V the huſband shal be without fault, and she shal beare her
iniquitie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
