%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 0353
%%% o-0318
\startChapter[
  title={Chapter 6}
  ]

\Summary{Conſecration, 14.~and oblation of Nazareites. 22.~A ſette forme
  how the Prieſt shal bleſſe the people.}

%%% 0354
And our Lord ſpake vnto Moyſes, ſaying: \V Speake to the children of
Iſrael, and thou ſhal ſay to them: Man, or woman, when they ſhal make a
vow to be ſanctified, and
\LNote{VVil conſecrate themſelues.}{To
\MNote{God preſcribed the rule of Nazareites, & the rites of their
conſecration.}
ſuch as of their owne accord, would bind them ſelues by vow to certaine
thinges not cõmanded, God preſcribeth a Rule containing three ſpecial
obſeruations: not to drincke wine, or anie thing that may make drunck; not
to cut their haire; and not to touch a dead corps. He appointeth
likewiſe rites in making this profeſſion, and calleth the
profeſſed \Emph{Nazareites}, that is, \Emph{ſegregated}
or \Emph{ſeparated} from the ordinarie ſtate of people;
\MNote{The ſame was a figure of vowes both temporal and perpetual.}
though it
was, for moſt part, but for a time, to be limited by the parties
them ſelues, and in ſome perpetual, as in Samſon.
\XRef{Iudic.~13.}
%%% ??? S. Aug? S. Gregorie? Does this go here?
\CNote{\Cite{S.~Aug. q.~52. in lib. Iudic.}}
Now \Emph{vvhat do Nazareites ſignifie}, ſaith S.~Gregorie
\Cite{(li.~33. c.~23. Moral.)}
\Emph{but thoſe that abſtaine, and containe voluntarily from things
othervviſe lavvful?} As from eating fleſh, either for certaine daies and
times, or altogether; from Mariage; from proprietie in worldlie goodes;
and the like.}
wil conſecrate them ſelues to the Lord: \V they ſhal abſteine from wine,
& euerie thing, that can make one drunke, vinegar of wine, and of any
other potion, and whatſoeuer is preſſed out of the grape, they shal not
drinke: new grapes and drie they shal not eate \V al the daies wherin
they are by vow cõſecrated to the Lord: whatſoeuer may be of the
vineyard, from the reiſen to the kernel they shal not eate. \V Al the
time of his ſeparation a raſour shal not paſſe ouer his head vntil the
day be expired, that he is conſecrated to the Lord. He shal be holie,
\SNote{VVhen Samſon was depriued of theſe haires he loſt his ſtreingth.
\XRef{Iudic.~16.}}
whiles the bush of haire on his head doth grow. \V Al the time of his
conſecration he shal not enter in to the dead, \V neither shal he be
contaminated, no not on his fathers and mothers and brothers and ſiſters
corps, becauſe the conſecration of his God is vpon his head. \V Al the
daies of his ſeparation he shal be holie to the Lord. \V But if anie man
die ſodenly before him, the head of his conſecration shal be polluted:
which he shal shaue forthwith in the ſame day of his purgation, and
againe the ſeuenth day: \V and in the eight day he shal bring two
turtles, or two young pigeons to the prieſt in the entrie of the
couenant of teſtimonie. \V And the prieſt shal offer one for ſinne, and
the other for an holocauſt, and shal pray for him, becauſe he hath
ſinned by occaſion of the dead: and he shal ſanctifie his head that
day: \V and shal conſecrate to the Lord the daies of his ſeparation,
offering a lambe of a yeare old for ſinne: yet ſo that the former daies
be made fruſtrate, becauſe his ſanctification was polluted. \V This is
the law of conſecration. When the daies, that by vow he had determined,
shal be expired: he shal bring him to the doore of the tabernacle of
couenant, \V and shal offer his oblation to the Lord, a male lambe of a
yeare old without ſpotte for an holocauſt, and an ewe lambe of a yeare
old without ſpotte for ſinne, and a ramme without ſpotte, for a
pacifique hoſte, \V a basket alſo of vnleauened breades that are
tempered with oile, and wafers without leuen anointed with oile, and the
libamentes of euerie one: \V which the prieſt shal offer before the
Lord, and shal offer as wel for ſinne, as for an holocauſt. \V But the
ramme he shal immolate for a pacifique hoſte to the Lord,
%%% 0355
offering withal the baskette of azymes, and the libamentes that by
cuſtome are dew. \V Then shal the Nazareite
%%% o-0319
be shauen before the dore of the tabernacle of couenant, from the bush
of the haire of his conſecration: and he shal take his haires, and lay
them vpon the fire, that is put vnder the ſacrifice of pacifiques. \V
And a shoulder of the ramme boyled, and one cake without leuen out of
the baskette, and one wafer vnleuened, and he shal deliuer them into
the handes of the Nazareite, after that his head be shauen. \V And
receiuing them againe from him, he shal eleuate them in the ſight of the
Lord: and being ſanctified they shal be the prieſtes, as the breaſt,
which was commanded to be ſeparated, and the shoulder. After theſe
thinges the Nazareite may drinke wine. \V This is the law of the
Nazareite, when he shal vow his oblation to the Lord in the time of his
conſecration, beſides thoſe thinges which his hand shal find, according
to that which he had vowed in his minde, ſo shal he do to the fulfilling
of his ſanctification. \V And our Lord ſpake to Moyſes, ſaying: \V Speake
to Aaron and his ſonnes:
\SNote{A ſpecial and determinate forme of bleſſing.}
Thus shal you bleſſe the children of Iſrael, and you shal ſay to
them: \V Our Lord bleſſe thee, and keepe thee. \V Our Lord shew his face
to thee, and haue mercie vpon thee. \V Our Lord turne his countenance
vnto thee, and geue thee peace. \V And
\SNote{VVhen the prieſt vttereth the wordes God geueth the effect.}
they shal inuocate my name vpon the children of Iſrael, and I wil bleſſe
them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
