%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 0375
%%% o-0337
\startChapter[
  title={Chapter 16}
  ]

\Summary{Core and his complices, making ſchiſme againſt Moyſes and
  Aaron, 31.~ſome are ſwalowed in the earth, with their families and
  ſubſtance; 35.~other two hundred and fiftie offering incenſe, 41.~and
  fourtene thouſand ſeuen hundred of the common people, murmuring in
  behalfe of the ſedicious, are conſumed with fire from heauen.}

And behold Core the ſonne of Iſaar, the ſonne of Caath, the ſonne of
Leui, and Dathan and Abiron the ſonnes of Eliab, Hon alſo the ſonne of
Pheleth of the children of Ruben, \V
\LNote{Roſe againſt Moyſes.}{By
\MNote{This hiſtorie & others were writtẽ for our admonition.
\XRef{1.~Cor.~10.}}
this moſt famous Schiſme and terrible puniſhment therof, al are warned
to kepe order, vnitie, and peace within the Church of God; and in nowiſe
to communicate with heretikes, or ſchiſmatikes in the act of hereſie or
ſchiſme. A neceſſarie admonition eſpecially in ſuch times and places as
we liue in, and ſee greater breaches made from ordinarie and lawful
Paſtors, then this was. For as S.~Ignatius wel noteth
\Cite{(Epiſt.~3. ad Magnatianos.)}
\MNote{Core and his cõplices were not heretikes, but only ſchiſmatikes.}
Core Dathan and Abiron impugned not directly the law, but reſiſted
Moyſes and Aaron: Yet were they and manie thouſandes with them, ſeuerly
puniſhed for their conſpiracie. S.~Cyprian
\Cite{(lib.~1. Epiſt.~6.)}
obſerueth the ſame, ſaying: Core Dathan and Abiron acknowledged the ſame
God with Aaron and Moyſes, liuing vnder the ſame law and religion, and
inuocated one true God, yet becauſe, paſſing the appointed miniſterie of
their owne place, oppofite to Aaron (who by Gods fauour and ordinance
had receiued lawful Prieſthood) they tooke vpon them to ſacrifice, they
were forthwith puniſhed by God for their vnlawful attempts:
\MNote{God accepteth not ſacrifice done againſt his ordinance.}
neither could their ſacrifices irreligiouſly and vnlawfully offered
againſt Gods ordinance be ratified, nor profite them at al. Thus
teacheth S.~Cyprian the glorious Martyr. And the text is clere, that
they were neither Idolaters nor Heretikes, but the chiefe of them being
Leuites, of the familie of Caath, (who were nereſt in kinred, and in
office to the prieſts) and other principal men of diuers tribes,
enuying the ſuperioritie of Moyſes and Aaron, and that prieſthood was
eſtabliſhed only in Aarons progenie, arrogated to themſelues the office
of prieſts and offered incenſe, further pretending, for vpholding their
ſchiſme, that there ſhould be no Superior at al aboue the holie people
of God, which albeit they did not beleue, yet therby they drew the
multitude to fauour and folow them.
\MNote{God by ſpeedie puniſhmẽt preuented hereſie, wherto al ſchiſme
tendeth.}
But God deciding this debate, to take away the contradiction, made the
earth to open, and ſwallow vp thoſe that firſt refuſed to obey Moyſes,
with their tabernacles and ſubſtance; and fire from heauen to conſume
two hundred and fiftie which offered incenſe; and fourtene thouſand
ſeuen hundred of the common people, for imputing to their Superiors the
deſtruction of the ſedicious, were alſo conſumed with fire, raging
amongſt the whole multitude, til Aaron ſent by Moyſes, and offering
incenſe appeaſed Gods wrath, and ſaued the reſt.
\MNote{Not only the authores of wickednes, \Emph{but al that conſent},
much more that cooperate, \Emph{are vvorthie of death}.
\XRef{Rom.~1.}} 
And wil anie Chriſtians, (that know they haue immortal ſoules, and that
God is a iuſt Iudge) thincke to eſcape with leſſe damnation, who for
anie worldlie gaine, fauour, or feare, yeld their bodilie and perſonal
preſence at ſeruice or ſermon of heretikes, or anie way communicate with
heretikes in practiſe of hereſie?}
roſe againſt Moyſes, and other of the children of Iſrael two hundred
fiftie men, princes of the ſynagogue, and which in the time of aſſemblie
were called by name. \V And when they had ſtoode vp againſt Moyſes and
Aaron, they ſaid: Let it ſuffice you, that
\SNote{So Luther
\Cite{(li. de abrog. Miſsa.)}
and other enimies of Eccleſiaſtical Hierarchy, wil haue no proper
Prieſthood in the Church of Chriſt, becauſe al Chriſtians are called
\Emph{a holie prieſthood}.
\XRef{1.~Pet.~2.}
and \Emph{Prieſtes}.
\XRef{Apoc.~1.}}
al the multitude conſiſteth of holie ones, and our Lord is among them:
Why lift you vp your ſelues aboue the people of our Lord? \V Which when
Moyſes had heard, he fel flatte on his face: \V and ſpeaking to Core and
al the multitude, he ſaid: In the morning our Lord wil make it knowne
who pertaine to him, and the holie
\Fix{the}{he}{obvious typo, fixed in other}
wil ioyne to him ſelfe: and whom he shal chooſe, they shal
approch to him. \V This do therfore: Take euerie man their cenſars, thou
Core, and al thy councel: \V and taking fire in them to morrow, put vpon
it incenſe before our Lord: and whom ſoeuer he shal chooſe, the ſame
shal be holie: you do much exalt your ſelues ye ſonnes of Leui. \V And he
ſaid againe to Core: Heare ye ſonnes of Leui, \V Is it a ſmal thing vnto
you, that the God of Iſrael hath ſeparated you from al the people, and
ioyned you to him ſelfe, that you should ſerue him in the ſeruice of the
tabernacle, and should
%%% 0376
ſtand before the ful aſſemblie of the people, and should miniſter to
him? \V Did he therfore make thee and al thy brethren the ſonnes of Leui
to approch vnto him, that you should chalenge vnto you the prieſthood
alſo, \V and al thy companie should ſtand againſt our Lord? for what is
Aaron that you murmur againſt him? \V Moyſes therfore ſent to cal Darhan
and Abiron the ſonnes of Eliab. Who anſwered: We come not. \V Why, is it
a ſmal matter to thee that thou haſt brought vs out of a land, that
\Fix{folowed}{flowed}{obviious typo, fixed in other}
with milke and honie, to kil vs in the deſert, vnles thou rule alſo
like a lord ouer vs? \V In deede haſt thou brought vs into a land, that
floweth with riuers of milke and honie, & haſt thou geuen vs poſſeſſions
of fieldes & vineyardes? What, wilt thou plucke out our eies alſo? We
come not. \V Moyſes therfore being very wrath, ſaid to our Lord: Reſpect
not their ſacrifices: thou knoweſt that I haue not taken of them ſo much
as a little
%%% o-0338
aſſe at anie time, neither haue afflicted anie of them. \V And he ſaid
to Core: Thou, and al thy congregation ſtand ye apart before our Lord,
and Aaron to morrow apart. \V Take euerie one your cenſars, and put
incenſe vpon them, offering to our Lord two hundred fiftie cenſars: Let
Aaron alſo hold his cenſar. \V Which when they had done, Moyſes and
Aaron ſtanding, \V and had heaped together al the multitude againſt
them to the dore of the tabernacle, the glorie of our Lord appeared to
them al. \V And our Lord ſpeaking to Moyſes and Aaron, ſaid: \V Separate
your ſelues from the middes of this congregation, that I may ſodenly
deſtroy them. \V Who fel flatte on their face, and ſaid: Moſt mightie
God of the ſpirites of al flesh, when one ſinneth, shal thy wrath rage
againſt al? \V And our Lord ſaid to Moyſes: \V Command the whole people
that they ſeparate them ſelues from the tabernacles of Core and Dathan
and Abiron. \V And Moyſes aroſe, and went to Dathan and Abiron: and the
ancientes of Iſrael folowing him, \V he ſaid to the multitude: Depart
from the tabernacles of the impious men, and touch not the thinges that
pertaine to them,
\SNote{Thoſe that touch things perteining to impietie, or depart not
from the tabernacles of ſchiſmatikes are inwrapped in their ſinnes: much
more to goe vnto heretical Synagogues is condemned. See
\Cite{S.~Cyprian. li. de lapſis Parag.~5.}}
leſt you be wrapped in their ſinnes. \V And when they were departed from
their tentes round about, Dathan and Abiron coming forth ſtood in the
entrie of their pauilions with their wiues and children, and al the
multitude. \V And Moyſes ſaid:
\SNote{Moyſes proued before by miracles,
\XRef{(Exod.~4.)}
that he was ſent of God: and now he proueth againe by miracle, that he
and Aaron, and not theſe ſchiſmatikes were called and ſent by God to
gouerne his people.}
In this you shal know that our Lord hath ſent me to do al thinges that
you ſee, and 
%%% 0377
that I haue not forged them of my owne mind: \V If they die the
accuſtomed death of men, and if the plague, wherwith others alſo are
wont to be viſited, do viſite them, our Lord did not ſend me: \V but if
our Lord do a new thing, that the earth opening her mouth ſwallow them
downe, & al thinges that pertaine to them, and they deſcend quicke into
hel, you shal know that they haue blaſphemed our Lord. \V Immediatly
therfore as he ceaſed to ſpeake, the earth brake inſunder vnder their
feete: \V and opening her mouth, deuoured them with their tabernacles &
al their ſubſtance. \V And they went downe into hel quicke couered with
the ground, and perished out of the middes of the multitude. \V But al
Iſrael, that ſtoode round about, fled at the crie of them that perished,
ſaying: Leſt perhappes the earth ſwallow vs alſo. \V But a fire alſo
coming forth from our Lord, ſlew the two hundred fiftie men, that
offered the incenſe. \V And our Lord ſpake to Moyſes, ſaying: \V Command
\SNote{Aaron being already eſtabliſhed high Prieſt, God againe
confirmeth in \Emph{Eleazar the progenie of prieſtlie ſucceſſion}, in
Aarons ſtock, and not in other Leuites.
\Cite{S.~Aug. q.~30. in Num.}}
\Fix{Elezar}{Eleazar}{obvious typo, fixed in other}
the ſonne of Aaron the prieſt that he take vp the cenſars that
lie in the burning fire, and that he ſprinkle the fire hither and
thither: becauſe they be ſanctified \V in the deathes of the ſinners:
and let him beate them into plates, and faſten them to the altar,
becauſe there hath bene offered incenſe in them to the Lord, and they
are ſanctified, that the children of Iſrael may ſee them for a ſigne and
a monument. \V Eleazar therfore the prieſt tooke the braſen cenſars,
wherin they had offered, whom the burning fire deuoured, and
\Fix{bette}{beate}{obvious typo, fixed in other}
them onto plates, faſtening them to the altar: \V that the children of
Iſrael
%%% o-0339
afterward might haue, wherwith to be admonished, that no ſtranger
approch, and he that is not of the ſeede of Aaron, to offer incenſe to
our Lord, leſt he ſuffer as Core hath ſuffered, and al his congregation,
according as our Lord ſpake to Moyſes. \V And al the multitude of the
children of Iſrael murmured the day folowing againſt Moyſes and Aaron,
ſaying: You haue killed the people of our Lord. \V And when there roſe a
ſedition, and the tumult grew farder, \V Moyſes and Aaron fled to the
tabernacle of couenant. Which after they were entred the cloude couered
it, and the glorie of our Lord appeared. \V And our Lord ſaid to
Moyſes: \V Depart from the middes of this multitude,
\SNote{This multitude did only in wordes fauour ſchiſmatikes, what
iudgemẽt then remaineth to thoſe which in external actes participate
with heretikes?}
euen now wil I deſtroy them. And as they lay vpon the ground, \V Moyſes
ſaid to Aaron: Take the cenſar, and drawing fire from the altar, put
incenſe
%%% 0378
vpon it, going quickly to the people to pray for them: for euen now is
the wrath come forth from our Lord, and the plague rageth. \V Which when
Aaron had done, and had runne to the middes of the multitude, which now
the burning fire did waſte, he offered the incenſe: \V and ſtanding
betwen the deade and the liuing, he prayed for the people, and the
plague ceaſed. \V And there were, that were ſtrooken, fourtene thouſand
and ſeuen hundred men, beſide them that had perished in the ſedition of
Core. \V And Aaron returned to Moyſes vnto the doore of the tabernacle
of couenant after that the deſtruction was ceaſed.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
