%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 28
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-28


\project douay-rheims


%%% 0400
%%% o-0359
\startChapter[
  title={Chapter 28}
  ]

\Summary{Special ſacrifices are appointed for euerie day in the morning
  and euening. 9.~Likewiſe for euerie Sabbath day, 11.~for the firſt day
  of euerie moneth, 16.~for Paſch, 26.~and for Penticoſt.}

Ovr Lord alſo ſaid to Moyſes: \V Command the children of Iſrael, and
thou shalt ſay to them: My oblation and breades, and burnt ſacrifice of
moſt ſweete odour offer ye
\MNote{Varietie of Sacrifices for diuers times.}
in their times. \V Theſe are the ſacrifices which you muſt offer: Two
lambes of a yeare old without blemish daily for
\MNote{1.~Euerie day twiſe.}
the euerlaſting holocauſt: \V one you shal offer in the morning, and the
other at euen: \V the tenth part of an ephi of floure, which shal be
tempered with the pureſt oile, and shal haue the fourth part of an
hin. \V It is the continual holocauſt which you offered in Mount Sinai
for a moſt ſweete odour of the burnt ſacrifice of the Lord. \V And for a
libament you shal offer of wine the fourth part of an hin for euerie
lambe in the Sanctuarie of the Lord. \V And the other lambe in like
manner you shal offer at euen according to al the rite of the morning
ſacrifice, and of the libamentes therof, an oblation of moſt ſweete
odour to the Lord. \V And on the day of
\MNote{2.~On the Sabbath day.}
the Sabbath, you shal offer two lambes of a yeare old without blemish,
and two tenthes of flowre tempered
%%% 0401
with oile in ſacrifice, and the libamentes \V which are ritely powred
euerie Sabbath for an euerlaſting holocauſt. \V And in
\MNote{3.~The Neomenia, or new moone.}
the Calendes you shal offer an holocauſte to the Lord, two calues of the
heard, one ramme, ſeuen lambes of a yeare old without blemish, \V and
three tenthes of flowre tempered with oile in ſacrifice for euerie
calfe: and two tenthes of flowre tempered with oile to euerie ramme: \V
and the tenth part of a tenth of flowre tempered with oile in ſacrifice
to
%%% o-0360
euerie lambe. It is an holocauſt of moſt ſweete odour and burnt
ſacrifice to the Lord. \V And the libamentes of wine, that are to be
powred for euerie victime, shal be theſe: The halfe part of an hin for
euerie calfe, the third to a ramme, the fourth to a lambe. This shal be
the holocauſt through al monethes, that ſuccede one an other as the
yeare turneth about. \V A bucke goate alſo shal be offered to the Lord
for ſinnes an euerlaſting holocauſt with his libamentes. \V And in the
firſt moneth,
\MNote{4.~Paſch, or Phaſe.}
the fourtenth day of the moneth shal be the Phaſe of the Lord, \V and
the fiftenth day the ſolemnitie: ſeuen daies shal they eate azymes. \V
Of the which the firſt day ſhal be venerable and holie: no ſeruile worke
ſhal you doe therein. \V And you ſhal offer burnt ſacrifice an holocauſt
to the Lord, two calues of the heard, one ramme, ſeuen lambes of a yeare
old without blemish: \V and the ſacrifice of euerie one of flowre which
shal be tempered with oile, three tenthes to euerie calfe, and two
tenthes to euerie ramme, \V and the tenth part of a tenth, to euerie
lambe, that is to ſay, to the ſeuen lambes. \V And one bucke goate for
ſinne, that expiation may be made for you, \V beſide the morning
holocauſt which you shal alwaies offer. \V So shal you do euerie day of
the ſeuen daies for a foode of the fire, and for a moſt
\Fix{ſweee}{ſweete}{obvious typo, fixed in other}
odour to the Lord, which shal riſe of the holocauſt, and of the
libamentes of euerie one. \V The ſeuenth day alſo shal be moſt ſolemne
and holie vnto you: no ſeruile worke shal you doe therein. \V The day
alſo of
\MNote{5.~Pentecoſt.}
firſt fruites when you shal offer new fruites to the Lord, after that
the weekes be accomplished, shal be venerable and holie: no ſeruile
worke shal you doe therein. \V And you shal offer an holocauſt for a
moſt ſweete odour to the Lord, two calues of the heard, one ramme, and
ſeuen Lambes of a yeare old without blemish: \V and in the ſacrifices of
them three tenthes of flowre tempered with oile to euerie calfe, to
euerie ramme
%%% 0402
two, \V euerie lambe the tenth part of a tenth, which together are ſeuen
lambes: a goate alſo \V which is ſlaine for expiation: beſide the
euerlaſting holocauſt & the libamentes therof. \V Al shal you offer
without blemish with their libamentes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
