%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 0363
%%% o-0326
\startChapter[
  title={Chapter 10}
  ]

\Summary{Trumpets are ſounded by the prieſtes, diuerſly for diuerſe
  purpoſes. 11.~The campe marcheth from the deſert of Sinai. 19.~Moyſes
  intreateth Hobab the Madianite to remaine with them. 35.~His prayer
  when the arke is taken vp, and ſette downe.}

And
\MNote{The ſecond part.

Of diuers impediments, which happened to the Iſraelites, and renouation
of ſundrie precepts, in their iorney from the deſert of Sinai to the
campe of Moab.}
our Lord ſpake to Moyſes, ſaying: \V Make thee two trumpettes of beaten
ſiluer, wherwith thou maieſt cal together the multitude when the campe
is to be remoued. \V And when thou shalt ſound with the trumpettes, al
the multitude shal gather vnto thee to the doore of the tabernacle of
couenant. \V If thou ſound but once, the princes ſhal come to thee, and
the heades of the multitude of Iſrael. \V But if the trumpeting ſound in
length and with a broken tune, they ſhal moue their campe firſt that are
on the eaſt ſide. \V And at the ſecond ſound and the like noyſe of the
trumpet, they shal take vp their tentes that dwel toward the ſouth, and
after this maner shal the reſt doe, when the trumpettes shal ſound to
the marching. \V But when the people is to be gathered together, the
ſound of the trumpet shal be plaine, and they
\SNote{Before the whole multitude, plaine and neceſſarie pointes of
doctrine muſt only be vttered, but before the lerned and wiſer ſorte, higher
myſteries may be treated and taught.
\Cite{Theod. q.~15. in Num.}}
shal not make a broken ſound. \V And the ſonnes of Aaron the prieſtes
shal ſound with the trumpettes: and this shal be an ordinance for euer
in your generations. \V If you goe forth to warre out of your land
againſt the enimies
%%% 0364
that fight againſt you, you shal ſound with trumpettes in length, and
there shal be a remembrance of you before the Lord your God, that you
may be deliuered out of the handes of your enemies. \V If at anie time
you shal haue a banket, and feſtiual daies, and
%%% o-0327
Calendes, you shal ſound with trumpettes ouer the holocauſtes, and
pacifique victimes, that they may be vnto you for a remembrance of your
God. I the Lord your God. \V The ſecond yeare, in the ſecond moneth, the
twentith day of the moneth was the cloude lifted vp from the tabernacle
of couenant: \V and the children of Iſrael marched by their troupes from
the deſert of Sinai, and the cloud reſted in the wilderneſſe of
Pharan. \V And the firſt moued their campe according to the cõmandement
of our Lord by the hand of Moyſes. \V The ſonnes of Iudas by their
troupes: whoſe prince was Nahaſſon the ſonne of Aminadab. \V In the
tribe of the ſonnes of Iſſachar, the prince was Nathanael the ſonne of
Suar. \V In the tribe of Zabulon, the prince was Eliab the ſonne of
Helon. \V And the tabernacle was taken downe, which the ſonnes of Gerſon
and Merari carying, marched. \V And the ſonnes of Ruben alſo marched, by
their troupes and order, whoſe prince was Heliſur the ſonne of
Sedeur. \V And in the tribe of Ruben alſo marched, by their troupes and
order, whoſe prince was Heliſur the ſonne of Sedeur. \V And in the tribe
of Simeon, the prince was Salamiel the ſonne of Suriſaddai. \V Moreouer
in the tribe of Gad, the prince was Eliaſaph the ſonne of Duel. \V And
the Caathites alſo marched carying the Sanctuarie. So long was the
tabernacle caried, til they came to the place of erecting it. \V
The ſonnes of Ephraim alſo moued their campe by their troupes, in whoſe
hoſte the prince was Eliſama the ſonne of Ammiud. \V And in the tribe of
the ſonnes of Manaſſes, the prince was Gamaliel the ſonne of
Phadaſſur. \V And in the tribe of Beniamin the prince was Abidan the
ſonne of Gedeon. \V The laſt of al the campe marched the ſonnes of Dan
by their troupes, in whoſe hoſte the prince was Ahiezer the ſonne of
Ammiſaddai. \V And in the tribe of the ſonnes of Aſer, the prince was
Phegiel the ſonne of Ochran. \V And in the tribe of the ſonnes of
Nephthali the prince was Ahira the ſonne of Enan. \V Theſe are the
campes, and the marchinges of the children of Iſrael by their troupes
when they marched. \V And Moyſes ſaid to Hobab the ſonne of Raguel the
Madianite,
\SNote{Either this Hobab was otherwiſe called Raguel,
\XRef{Exo.~2.}
and alſo Iethro
\XRef{Exo.~3.}
and was father in law to Moyſes, or els he was ſonne of the ſame Raguel,
and brother in law to Moyſes.}
his
\TNote{\L{cognato} vel \L{affini}}
allied: We march toward the place, which our Lord wil geue vs: come with
vs, that we
%%% 0365
may do thee good: for our Lord hath promiſed good thinges to Iſrael. \V
To whom he anſwered: I wil not goe with thee, but I wil returne to my
countrie, wherein I was borne. \V And he ſaid: Doe not forſake vs: for
thou knoweſt in what places through the deſert we may campe, and
\SNote{Moyſes meaneth, that whẽ by the cloud, and piller of fire (their
ſpecial guides) the people ſhould come to new places, this Madianite his
allied might direct them, where to finde beſt paſture, water and like
commodities, nere to them.}
thou ſhalt be our guide. \V And when thou comeſt with vs, whatſoeuer
ſhal be beſt of the riches, which our Lord ſhal deliuer vs, we wil geue
thee. \V They marched therfore from the Mount of our Lord three daies
iourney, and the arke of the couenant of our Lord went before them, for
three daies prouiding a place for the campe. \V The cloude alſo of our
Lord was ouer them by day when they marched. \V And when the arke was
lifted vp, Moyſes ſaid:
\SNote{Beſides general prayers for al purpoſes, ſome are compoſed, and
applied for ſpecial times and occaſions.}
Ariſe Lord, and by thyne enemies diſperſed, and let them flee that hate
thee, from thy face. \V And when it was ſette downe, he ſaid: Returne Lord
to the multitude of the hoſte of Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
