%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 34
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-34


\project douay-rheims


%%% 0413
%%% o-0370
\startChapter[
  title={Chapter 34}
  ]

\Summary{The ſituation and limites of Chanaan, towards the South, the
\Fix{VVeaſt,}{VVeſt,}{possible typo, fixed in other}
the North, and the Eaſt, 13.~which muſt be diuided by lotte among nine
tribes and a halfe (the other two and a half hauing their partes ouer
Iordan) 16.~with the names of the perſons that shal make the diuiſion.}

And our Lord ſpake to Moyſes, ſaying: \V Command the children of Iſrael,
and thou ſhalt ſay to them: When you are entred into the Land of
Chanaan, and it ſhal be fallen vnto you by lotte in poſſeſſion, it ſhal
be limited by theſe borders: \V The ſouthſide ſhal beginne from the
wilderneſſe Sin, which is beſide Edom: and it ſhal haue toward the Eaſt
\SNote{Otherwiſe called \Emph{the dead ſea}, where Sodom and
\Fix{Gomort}{Gomorre}{likely typo, fixed in other}
had ſtood.}
the moſt ſalt ſea for the limites. \V The which ſhal compaſſe the ſouth
ſide by the aſcending of
\SNote{A mountaine ſo called of the multitude of ſcorpions, which were
in it.}
the Scorpion, ſo that they ſhal paſſe into Senna, and reach toward the
South as farre as Cadeſbarne, from whence the frontiers shal goe forth
to the towne named Adar, and shal reach as farre as Aſemona. \V And the
border shal goe round about from Aſemona to the Torrent of Ægypt, and
shal end in the shore of the
\SNote{
\Fix{Mediterraneal}{Mediterranean}{possible typo, fixed in other}
ſea, called great in reſpect of the lakes in the holie Land.}
great ſea. \V And the weſt ſide shal beginne from the great ſea, and the
ſame shal be the limite therof. \V Moreouer toward the North part the
borders shal beginne from the great ſea, reaching vnto
%%% o-0371
the
\SNote{Mount of Libanus.}
moſt high mountaine, \V from
\Fix{he}{the}{obvious typo, fixed in other}
which they shal come vnto Emath as farre as the borders of Sedada: \V
and the frontiers shal goe as farre as Sephorna, and the towne
Enan. Theſe shal be the borders in the North part. \V From thence they
shal marke out the boundes toward the Eaſt ſide from the village Enan
vnto Sephama, \V and from Sephama the boundes shal goe downe vnto Rebla
againſt the fountaine Daphnim: from thence they shal come eaſtward to
the ſea Cenereth, \V and shal
%%% 0414
reach as farre as Iordan, and at the laſt shal be cloſed in by the moſt
ſalt ſea. This you shal haue for your Land by the boundes therof round
about. \V And Moyſes commanded the children of Iſrael, ſaying: This shal
be the Land, which you shal poſſeſſe by lotte, and which our Lord hath
commanded to be geuen to the nine tribes, and to the halfe tribe. \V For
the tribe of the children of Ruben by their families, and the tribe of
the children of Gad according to the number of their kinredes, the halfe
alſo of the tribe of Manaſſes, \V that is, two tribes and an halfe, haue
taken their portion beyond Iordan againſt Iericho at the eaſt ſide. \V
And our Lord ſaid to Moyſes: \V Theſe are the names of the men, that
ſhal diuide the Land vnto you, Eleazar the prieſt, and Ioſue the ſonne
of Nun, \V and one prince of euerie tribe, \V whoſe names are theſe: Of
the tribe of Iuda, Caleb the ſonne of Iophone. \V Of the tribe of
Simeon, Samuel the ſonne of Ammiud. \V Of the tribe of Beniamin, Elidad
the ſonne of Chaſelon. \V Of the tribe of the children of Dan, Bocci the
ſonne of Iogli. \V Of the children of Ioſeph of the tribe of Manaſſes,
Haniel the ſonne of Ephod. \V Of the tribe of Ephraim, Camuel the ſonne
of Sephthan. \V Of the tribe of Zabulon, Eliſaphan the ſonne of
Pharnach. \V Of the tribe of Iſſachar, duke Phaltiel the ſonne of
Ozan. \V Of the tribe of Aſer, Ahiud the ſonne of Salomi. \V Of the
tribe of Nephthali, Phedael the ſonne of Ammiud. \V Theſe are they whom
our Lord commanded to diuide the Land of Chanaan to the children of
Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
