%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 0380
%%% o-0341
\startChapter[
  title={Chapter 18}
  ]

\Summary{The charge and burden of Prieſtes within the tabernacle, and
  of Leuites about the ſame. 8.~Firſt fruites, and oblations are due to
  the Prieſtes; 21.~the tithes to the Leuites; 26.~who of the ſame pay
  tithes to the Prieſtes.}

And our Lord ſaid to Aaron: Thou, and thy ſonnes, and the houſe of thy
father with thee shal beare the iniquitie of the Sanctuarie: both thou &
thy ſonnes together shal beare the ſinnes of your prieſthood. \V But thy
brethren alſo of the tribe of Leui, and the ſcepter of thy father take
with thee, and let them be readie at hand, and miniſter to thee: but
thou and thy ſonnes shal miniſter in the tabernacle of teſtimonie. \V And
the Leuites shal watch vpon thy preceptes, and vpon al the workes of the
tabernacle: ſo not withſtanding, that they approch not to the
\Fix{veſtel}{veſſels}{obvious typo, fixed in other}
of the Sanctuarie and to the altar, leſt both they die, and you perish
withal. \V But let them be with thee, and watch in the cuſtodies of the
tabernacle, and in al the ceremonies therof. A ſtranger shal not ioyne
with you. \V Watch in the cuſtodie
%%% o-0342
of the Sanctuarie, and in the miniſterie of the altar: leſt indignation
riſe vpon the children of Iſrael. \V I haue geuen you your brethren the
Leuites out of the middes of the children of Iſrael, and haue deliuered
them a gifte to the Lord, to ſerue in the miniſteries of his
tabernacle. \V And thou and thy ſonnes looke to your prieſthood: and al
thinges that perteyne to the ſeruice of the altar, and that are within
the vele, ſhal be executed by the prieſtes. If anie ſtranger approch,
\SNote{Oza was ſlaine for touching the arck.
\XRef{2.~Reg.~6.}}
he ſhal be ſlaine. \V And our Lord ſpake to Aaron: Behold I haue geuen
thee the cuſtodie of my firſt fruites. Al thinges that are ſanctified of
the children of Iſrael, haue I deliuered to thee and to thy ſonnes for
the prieſtlie office, as euerlaſting
%%% 0381
ordinances. \V Theſe thinges therfore ſhalt thou take of thoſe, that are
ſanctified, and are offered to the Lord. Al oblation, and ſacrifice, and
whatſoeuer is rendred to me for ſinne and offence, & becometh Holie of
holies, ſhal be thine, and thy ſonnes. \V In the Sanctuarie ſhalt thou
eate it: males onlie shal eate therof, becauſe it is to thee a
cõſecrated thing. \V But the firſt fruites, which the children of Iſrael
shal vow and offer, I haue geuen thee, and thy ſonnes, and thy daughters
for a perpetual right: he that is cleane in thy houſe, shal eate
them. \V Al the beſt of oile, and wine, and corne, whatſoeuer firſt
fruites they offer to the Lord, I haue geuen them to thee. \V Of fruites
al the firſt, that the ground bringeth forth, and are brought to the
Lord, shal turne to thy vſes: he that is
cleane in thy houſe, shal eate them. \V Euerie thing that the children
of Iſrael render by vow, shal be thine. \V Whatſoeuer firſt breaketh
forth from the matrice of al flesh, which they offer to the Lord, whether
it be of men, or of beaſtes, shal be thy right: yet ſo, that for the
firſt borne of man thou take a price, and euerie beaſt that is vncleane
thou cauſe to be redemed, \V whoſe redemption shal be after one moneth,
for fiue ſicles of ſiluer, by the weight of the Sanctuarie. A ſicle hath
twentie oboles. \V But the firſt borne of beefe and ſheepe and goate
thou ſhalt not cauſe to be redemed, becauſe they are ſanctified to the
Lord. Onlie the bloud of them thou shalt powre vpon the altar, and the
fatte thou shalt burne for a moſt ſweete odour to the Lord. \V But the
flesh shal turne to thy vſe, as the conſecrated breſt, and the right
shoulder, shal be thine. \V Al the firſt fruites of the Sanctuarie which
the children of Iſrael offer to the Lord, haue I geuen thee and thy
ſonnes, and daughters for a perpetual right.
\SNote{This couenant of firſt fruites and other rights geuen to the
Prieſts ſhal be perpetual, as ſalt is in euerie ſacrifice.
\XRef{Leuit.~2.}}
A couenant of ſalt is it for euer before the Lord, to thee and to thy
ſonnes. \V And our Lord ſaid to Aaron: In their land you shal poſſeſſe
nothing, neither shal you haue a portion among them: I am thy portion
and inheritance in the middes of the children of Iſrael. \V And to the
ſonnes of Leui I haue geuen al the tithes of Iſrael in poſſeſſion for the
miniſterie wherwith they ſerue me in the tabernacle of couenant: \V that
the children of Iſrael approch not any more to the tabernacle, nor
committe
\SNote{Puniſhable with death.}
deadlie ſinne, \V onlie the ſonnes of Leui ſeruing me in the tabernacle,
and bearing the ſinnes of the people. It shal be an euerlaſting
ordinance in
%%% 0382
%%% o-0343
your generations. No other thing shal they poſſeſſe, \V being content
with the oblation of tithes, which I haue ſeparated for their vſes and
neceſſaries. \V And our Lord ſpake to Moyſes, ſaying: \V Command the
Leuites, and denounce vnto them: When you ſhal receiue of the children
of Iſrael the tithes, which I haue geuen you, offer firſt fruites of
them to the Lord, that is to ſay, the tenth part of the tenth: \V that
it may be reputed to you for an oblation of firſt fruites, as wel of the
barne floores as of the preſſes, \V and of al thinges wherof you receiue
tithes, the firſt fruites offer to the Lord, and geue them to Aaron the
prieſt. \V Al thinges that you ſhal offer of the tithes, and ſhal
ſeparate for the giftes of the Lord, they shal be of the beſt and choſen
thinges. \V And thou ſhalt ſay to them: If you offer al the goodlie and
the better thinges of the tithes, it ſhal be reputed to you as if you had
geuen firſt fruites of the barne floore and the preſſe: \V and you shal
eate them in al your places, as wel you as your families: becauſe it is
the reward for the miniſterie, wherwith you ſerue in the tabernacle of
teſtimonie. \V And you shal not ſinne in this point, reſeruing the
principal and fatte thinges to your ſelues leſt you pollute the
oblations of the children of Iſrael, and die.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
