%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-22


\project douay-rheims


%%% 0388
%%% o-0349
\startChapter[
  title={Chapter 22}
  ]

\Summary{Balac King of Moab fearing the Iſraelites, ſendeth for Balaam a
  ſouthſayer to curſe them. 8.~VVho conſulting his falſe god, is forbid
  by God almightie to goe, and ſo excuſeth him ſelf. 15.~Balac ſendeth
  againe offering greater reward, 19.~he againe conſulteth, and God
  biddeth him goe. 22.~But ſendeth an Angel to meete him in the way,
  whom his aſse ſeing, shunneth three times, and ſo often he beateth
  her, 28.~then she ſpeaketh, expoſtulating his hard vſage: 31.~he alſo
  ſeeth the Angel, 35.~and is charged to ſpeake nothing but that the
  Angel shal ſuggeſt.}

And marching forward they camped in the champion countrie of Moab, where
Iericho is ſituated beyond Iordan. \V And Balac the ſonne of Sephor
ſeeing al thinges that Iſrael had done to the Amorrheite, \V and that
the Moabites were in greate feare of him, and could not ſuſteyne his
aſſault, \V he ſaid to the elders of Madian: So wil this people deſtroy
al that dwel in our coaſtes, as the oxe is wont to eate the graſſe vnto
the verie rootes. And he was at the ſame time King in Moab. \V He ſent
therfore meſſengers to Balaam the ſonne of Behor a Southſayer, who dwelt
vpon the riuer of the land of the children of Ammon, to cal him, and to
ſay: Behold a people is come out of Ægypt, that hath couered the
%%% 0389
face of the earth, ſitting againſt me. \V Come therfore, and curſe this
people, becauſe it is mightier then I: if by any meanes I may ſtrike
them and caſt them out of my land. For I know that he is bleſſed whom
thou ſhalt bleſſe, and curſed vpon whom thou ſhalt heape curſes. \V And
the ancientes of Moab went, and the elders of Madian, hauing the price
of diuination in their handes. And when they were come to Balaam, and
had told him al the wordes of Balac: \V he anſwered: Tarie here this
night, and I wil anſwer whatſoeuer
\SNote{He conſulted his falſe god, whom he ſerued, & called him the Lord,
not knowing our Lord God almightie.
\Cite{Theodoret. q.~40. Procopius. in Num.}}
the Lord shal ſay to me. And whiles they ſtayed with Balaam,
\SNote{God our Lord anſwered him, not ſuffering the diuel to ſpeake in
this caſe.}
God came, and ſaid to him: \V What would theſe men that are with thee? \V He
anſwered: Balac the ſonne of Sephor king of the Moabites hath ſent to
me, \V ſaying: Behold a people that is come out of Ægypt, hath couered
the face of the land: come, and curſe them, if by any meanes fighting I
may driue them away. \V And God ſaid to Balaam: Goe not with them,
neither doe thou curſe the people: becauſe it is bleſſed. \V Who in the
morning ariſing ſaid to the princes: Goe into your countrie, becauſe the
Lord hath forbid me to come with you. \V The princes returning, ſaid to
Balac: Balaam would not come with vs. \V Againe he ſent many moe and
more noble, then he had ſent before. \V Who when they were come to
Balaam, ſaid: Thus ſayeth Balac the ſonne of
%%% o-0350
Sephor: Slacke not to come to me: \V for I am readie to honour thee, and
whatſoeuer thou wilt I wil geue thee: Come, and curſe this people. \V
Balaam anſwered: If Balac would geue me his houſe ful of ſiluer and
gold, I can not change the word of the Lord my God, to ſpeake either
more, or leſſe. \V I beſech you that you wil alſo tarie here this night,
and I may know what the Lord wil anſwer me
\SNote{Being ſufficiently informed before, for worldlie lucre he
demandeth againe, & God for his puniſhment letteth him goe: but
ſuffereth him not to curſe, nor ſpeake euil of the Iſraelites, but
cõtrariwiſe to prophecie much good, which ſhould come vnto them.}
once more. \V God therfore came to Balaam in the night, and ſaid to him:
If theſe men be come to cal thee, ariſe and goe with them: yet ſo, that
thou doe that which I shal command thee. \V Balaam aroſe in the morning,
and ſadling his aſſe went with them. \V And God was angrie. And an Angel
of our Lord ſtoode in the way againſt Balaam, who ſate vpon the aſſe,
and had two ſeruantes with him. \V The aſſe ſeeing the Angel ſtanding in
the way, with a drawen ſword, turned her ſelf out of the way, and went
by the field. Whom when Balaam had
\Fix{bette,}{beate,}{possible typo, fixed in other}
and had brought her againe to the beaten way, \V the Angel
%%% 0390
ſtoode in the ſtreictes of two walles, wherwith vineyardes were
incloſed. \V Whom the aſſe ſeeing, ioyned her ſelfe cloſe to the wal,
and bruſed the foote of him that ridde. But he beate her againe: \V and
neuertheleſſe the Angel paſſing to a narrow place, where there could be
no going a ſide neither to the right hand nor to the left, ſtoode to
meete him. \V And when the aſſe had ſeene the Angel ſtanding, she fel
vnder the feete of the rider. Who being angrie,
\Fix{bette}{beate}{possible typo, fixed in other}
her ſides with a ſtaffe more vehemently. \V And our Lord opened the
mouth of the aſſe, and she ſpake: What haue I done to thee? why ſtrikeſt
thou me? loe now the third time? \V Balaam
\SNote{Nothing is here to be more merueled (\Emph{ſayeth S.~Auguſtin})
then that he was not afraide whẽ he heard his aſſe ſpeake vnto him. But
being accuſtomed to ſuch monſtrous things he replied familiarly,
nothing therwith aſtooniſhed.}
anſwered: Becauſe thou haſt deſerued, and haſt abuſed me: I would I had
a ſword, that I might kil thee. \V The aſſe ſaid: Am not I thy beaſt, on
which thou haſt beene alwayes accuſtomed to ride vntil this preſent day?
tel me what like thing did I euer to thee. But he ſaid: Neuer. \V
Forthwith our Lord opened the eies of Balaam, and he
\SNote{Yet ſeing an Angel, he fel downe terrified, and adored him.
\Cite{q.~48. in Num.}}
ſaw the Angel ſtanding in the way with a drawen ſword, and he
\SNote{Balaam knowing him to be an Angel that appeared, neither adored
him with diuine honour as God, nor with ciuil, as a mortal man, but with
religious honour, leſſe then diuine, & more then ciuil. See
\XRef{Exod.~20.}}
adored him flatte to the ground. \V To whom the Angel ſaid: Why beateſt
thou thy aſſe the third time? I am come to withſtand thee, becauſe thy
way is peruerſe, and contrarie to me: \V and vnleſſe the aſſe had turned
out of the way, geuing place to me reſiſting thee, I had ſlaine thee,
and she should haue liued. \V Balaam ſaid: I haue ſinned, not knowing
that thou didſt ſtand againſt me: and now if it diſpleaſe thee that I
goe, I wil returne. \V The Angel ſaid:
\SNote{God ſometimes maketh falſe prophetes to vtter a truth, becauſe
preiudicate mindes do rather geue eare to ſuch, thẽ to true prophetes.
\Cite{Theod. q.~45. in Num.}}
Goe with theſe men, and ſee thou ſpeake no other thing then I shal
command thee. He went therfore with the princes. \V Which when Balac
heard, he came forth to meete him in a towne of the Moabites, that is
ſituated in the vttermoſt borders of Arnon. \V And he ſaid to Balaam: I
ſent meſſengers to cal thee, why didſt thou not come immediatly vnto me?
was it becauſe I can not reward thy coming? \V To whom he anſwered: Loe
here I am: Shal I be able to ſpeake any other thing, but that which God
shal put in my mouth? \V They therfore
%%% o-0351
went on together, and came into a citie, that was in the vttermoſt
borders of his kingdome. \V And when Balac had killed oxen, & sheepe, he
ſent therof to Balaam, and to the princes that were with him,
preſentes. \V And when morning was come, he brought him to the excelſes
of Baal, and he beheld the vttermoſt part of the people.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
