%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 0367
%%% o-0330
\startChapter[
  title={Chapter 12}
  ]

\Summary{Marie and Aaron murmur againſt Moyſes. 6.~Whom God praiſeth
  aboue other prophetes. 10.~Marie being ſtricken with leproſie, Aaron
  confeſſeth his fault. 13.~Moyſes prayeth for her, and after ſeuen
  daies ſeparation from the campe, she is reſtored.}

And Marie and Aaron ſpake againſt Moyſes, for his
\CNote{\XRef{Exod.~2.}}
wife the
\SNote{Madianites were alſo called Æthiopians.
\Cite{S.~Aug. q.~20. in Num.}}
Æthiopian, \V and they ſaid: Hath our Lord
%%% 0368
ſpoken by Moyſes onlie? hath he not ſpoken to vs alſo in like manner?
Which when our Lord had heard, (\V
\SNote{The Holie Ghoſt forced Moyſes to vtter his owne praiſe, which of
him ſelf, he deſired not.}
For Moyſes was the mildeſt man aboue al men, that dwelt vpon the
earth) \V immediatly he ſpake to him, and to Aaron and Marie: Goe forth
you three onlie to the tabernacle of couenant. And when they were come
forth, \V our Lord deſcended in the piller of a cloude, and ſtoode in
the entrie of the tabernacle calling Aaron and Marie. Who going vnto
him, \V he ſaid to them: Heare my wordes: If there shal be among you a
prophete of the Lord, in viſion wil I appeare to him, or in ſleepe I wil
ſpeake 
\Fix{ho}{to}{obvious typo, fixed in other}
him. \V But my ſeruant Moyſes is not ſuch an one, who in al my
houſe is moſt faythful: \V for mouth to mouth I ſpeake to him: and
plainely, and not by riddels and figures doth he ſee the Lord. Why
therfore did you not feare to detract from my ſeruant Moyſes? \V And
being wrath againſt them, he went away: \V the cloude alſo departed that
was ouer the tabernacle: and behold Marie appeared white with leproſie
as it were ſnow. And
\SNote{Aaron was not publikly puniſhed, leſt therby he had bene made
cõtemptible to the people, but was otherwiſe
\Fix{chatiſed.}{chaſtiſed.}{obvious typo, fixed in other}}
when Aaron had looked on her, and ſaw her wholy couered with
leproſie, \V he ſaid to Moyſes: I beſech thee my Lord, lay not vpon vs
this ſinne which we haue fooliſhly committed, \V let not this woman be
as it were dead, and as an abortiue that is caſt forth of the mothers
wombe. Lo now the one halfe of her flesh is deuoured with the
leproſie. \V And Moyſes cried to our Lord, ſaying: God, I beſech
thee, heale her. \V To whom our Lord anſwered: If her father had ſpitte
vpon her face, ought she not to haue bene ashamed ſeuen daies at the
leaſt? Let her be ſeparated ſeuen daies without the campe, and
afterwards she shal be called againe. \V Marie therfore was shut forth
without the campe ſeuen daies: and the people moued not from that place,
vntil Marie was called againe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
