%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 0344
%%% o-0309
\startChapter[
  title={Chapter 2}
  ]

\Summary{At the eaſt ſide of the Tabernacle the tribe of Iuda as chiefe,
  with Iſſachar and Zabulon do pitch their tents, and march firſt:
  10.~on the ſouth Ruben, with Simeon and Gad (17.~the Tabernacle is
  caried, and erected by the Leuites, who lodge and march round about it)
  18.~on the weaſt ſide, Ephraim with Manaſſes and Beniamin: 25.~on the
  north, Dan with Aſer and Nephthali.}

And our Lord ſpake to Moyſes and Aaron ſaying: \V Euerie one of the
children of Iſrael by the troupes, enſignes, and ſtanderts, and houſes
of their kinredes shal campe, round about the tabernacle of couenant. \V
On the eaſt  Iudas shal pitche his tents by the troupes of his bande:
and the prince of his ſonnes shal be Nahaſſon the ſonne of Aminadab. \V
And the whole ſumme of the warriers of his ſtocke, ſeuentie foure
thouſand ſix hundred. \V Beſide
%%% o-0310
him camped they of the tribe of Iſſachar, whoſe prince was Nathanael the
ſonne of Suar: \V and al the number of his warriers fiftie foure
thouſand foure hundred. \V In the tribe of Zabulon the prince was Eliab
the ſonne of Helon \V and al the hoſte of warriers of his ſtocke,
fiftie ſeuen thouſand foure hundred. \V Al that were numbered in the
campe of Iudas, were an hundred eightie ſix thouſand foure hundred: and they
by their troupes shal march firſt. \V In the campe of the ſonnes of
Ruben on the ſouth ſide the prince shal be Eliſur the ſonne of
Sedeur: \V and the whole hoſte of his warriers, that were numbred,
fourtie ſix thouſand fiue hundred. \V Beſide him camped they of the
tribe of Simeon: whoſe prince was Salamiel the ſonne of Suriſaddai: \V
and the whole hoſte of his warriers, that were numbred, fiftie nine
thouſand three hundred. \V In the tribe of Gad, the prince was Eliaſaph
the ſonne of Duel: \V and the whole hoſte of his warriers, that were
numbred, fourtie fiue thouſand ſix hundred fiftie. \V Al that were
reckened in the campe of Ruben, an hundred fiftie thouſand and a
thouſand foure hundred fiftie by their troupes: they shal march in the
ſecond place. \V But the tabernacle of teſtimonie shal be lifted vp by
the offices of the Leuites and their troupes. As it shal be ſette vp, ſo
shal it be taken downe. Euerie one shal march by their places, and
orders. \V On the weſt ſide shal be the campe of the ſonnes of Ephraim,
whoſe prince was Eliſama the ſonne of Ammiud: \V the whole
%%% 0345
hoſte of his warriers, that were numbered, fourtie thouſand fiue
hundred. \V And with them the tribes of the ſonnes of Manaſſes, whoſe
prince was Gamaliel the ſonne of Phadaſſur: \V and the whole hoſte of
his warriers, that were numbered, thirtie two thouſand two hundred. \V
In the tribe of the ſonnes of Beniamin the prince was Abidan the ſonne
of Gedeon: \V and the whole hoſte of his warriers, that were reckened,
thirtie fiue thouſand foure hundred. \V Al that were numbered in the
campe of Ephraim, an hundred eight thouſand one hundred by their
troupes: they shal march the third. \V On the north part camped the
ſonnes of Dan: whoſe prince was Ahiezar the ſonne of Ammiſaddai: \V the
whole hoſte of his warriers, that were numbered, ſixtie two thouſand
ſeuen hundred. \V Beſides him they of the tribe of Aſer pitched their
tents: whoſe prince was Phegiel the ſonne of Ochran: \V the whole hoſte
of his warriers, that were numbered, fourtie thouſand and a thouſand
fiue hundred. \V Of the tribe of the ſonnes of Nephthali the prince was
Ahira the ſonne of Enan: \V the whole hoſte of his warriers, fiftie
three thouſand foure hundred. \V Al that were numbered in the campe of
Dan, were an hundred fiftie ſeuen thouſand ſix hundred: and they shal
march laſt. \V This is the number of the children of Iſrael, by the
houſes of their kinredes and troupes of the hoſte being diuided, ſix
hundred three thouſand fiue hundred fiftie. \V And the Leuites were not
numbered among the children of Iſrael: for ſo our Lord had commanded
Moyſes. \V And the children of Iſrael did according to al
%%% o-0311
thinges, that our Lord had commanded. They camped by their troupes, and
marched by their families and houſes of their fathers.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
