%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 24
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-24


\project douay-rheims


%%% 0392
%%% o-0352
\startChapter[
  title={Chapter 24}
  ]

\Summary{Balaam forced by the euidence of truth (though not conuerted in
  wil to ſerue God, whom he confeſseth to be omnipotent) prophecieth
  ſtil more good of Iſrael: 10.~wherfore Balac interrupteth him, and he
  anſwereth plainly that God almightie wil haue it ſo. 15.~And ſo
  proceeding he prophecieth of \Sc{Christ}. 20.~Alſo of Amalacheites,
  Cineites, and Italians.}

And when Balaam had ſeene that it pleaſed our Lord that he should bleſſe
Iſrael, he went not as before he had gone, to ſeeke Southſaying: but
directing his countenance againſt the deſert, \V and lifting vp his
eies, he ſaw Iſrael abiding in their tentes by their tribes: and the
ſpirit of
%%% 0393
God coming
\SNote{The ſpirite of prophecie came vpon him, but not grace iuſtifying:
\Emph{Manie do prophecie, and caſt out diuels}, and worke other
miracles, and yet be dãned.
\XRef{Mat.~7. v.~22.}}
vehemently vpon him, \V taking vp his parable he ſaid: Balaam the ſonne
of Bephor hath ſaid: The man hath ſaid whoſe eye is ſtopped: \V The
hearer of the
%%% o-0353
wordes of God hath ſaid: he that hath beholden the viſion of the
Omnipotent, he that falleth, and ſo his eies are opened: \V How
beautiful are thy tabernacles ô Iacob, and thy tentes ô Iſrael! \V As
woodie valleis, as watered gardens beſide the riuers, as tabernacles
which the Lord hath pitched, as cedres by the waters ſide. \V Water shal
flow out of his bucket, and his ſeede shal be into manie waters. For
Agag shal his king be taken away, and his kingdom shal be taken away. \V
God hath brought him out of Ægypt, whoſe ſtrẽgth is like to the
rhinocerote. They shal deuoure nations his enimies, and breake their
bones, and pearce them with arrowes. \V Lying he hath ſlept as a lion,
and as a lioneſſe, whom none shal be bold to rayſe vp. He that bleſſeth
thee, him ſelf alſo shal be bleſſed: he that curſeth thee, shal be
reputed accurſt. \V And Balac being angrie againſt Balaam, clapping his
handes together ſaid: To curſe mine enemies I called thee, whom thou
contrariewiſe haſt bleſſed the third time: \V returne to thy place. I
was determined verily to honour thee magnifically, but the Lord hath
depriued thee of the honour appointed. \V Balaam made anſwer to Balac:
did I not ſay to thy meſſengers, whom thou didſt ſend to me: \V If Balac
would geue me his houſe ful of ſiluer and gold, I can not paſſe the word
of the Lord my God, to vtter of my owne minde either any good, or euil: but
whatſoeuer the Lord shal ſay, that wil I ſpeake? \V But yet going to my
people, I wil geue thee counſel, what thy people shal doe to this people
in the laſt time. \V Therfore taking vp his parable, againe he ſaid:
Balaam the ſonne of Beor hath ſaid: The man whoſe eye is ſtopped, hath
ſaid: \V the hearer of the wordes of God hath ſaid: who knoweth the
doctrine of the Higheſt, and ſeeth the viſions of the Omnipotent, who
falling hath his eies opened: \V I shal ſee him, but not now: I shal
behold him but not neere.
\SNote{In ſome ſort this prophecie was fulfilled in King Dauid.
\XRef{2.~Reg.~5,~8. &~ſeq.}
but perfectly in Chriſt the bright, and mornĩg ſtarre.
\XRef{Apoc.~22.}
To whom not anie one, or manie, but in general al nations are geuẽ for
inheritãce, and the endes of the world, in poſſeſſion.
\XRef{Pſal.~2.}
\XRef{Act.~2. v.~8.}}
\Sc{A starre shal rise} out of Iacob, and a rodde shal ariſe from
Iſrael: and shal ſtrike the dukes of Moab, and shal waiſt al the
children of Seth. \V And Idumea shal be his poſſeſſion: the inheritance
of Seir shal come to their enemies: but Iſrael shal doe manfully. \V Of
Iacob shal he be that shal rule, and shal
%%% 0394
deſtroy the remaines of the citie. \V And when he had ſeene Amalec,
taking vp his parable, he ſaid: Amalec the beginning of Gentiles, whoſe
latter endes shal be deſtroyed. \V He ſaw alſo the Cineite: and taking
vp his parable, he ſaid: Thy habitation in deede is ſtrong: but if thou
build thy neſt in a rocke, \V and thou be choſen of the ſtocke of Cin,
how long shalt thou be able to continew? For Aſſur shal take thee. \V And
taking vp his parable againe he ſpake: Alas, who shal liue, when God
shal doe theſe thinges? \V They shal come in galleies from Italie, they
shal ouercome the Aſſyrians, and shal waiſt the Hebrewes, and at the
laſt them ſelues alſo shal perish. \V And Balaam roſe, and returned into
\SNote{Not to his countrie, but to his place of abode among the
Madianites: for there he was ſlaine ſhortly after.
\XRef{chap.~31. v.~8.}}
his place: Balac alſo returned the way that he came.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
