%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 0373
%%% o-0335
\startChapter[
  title={Chapter 15}
  ]

\Summary{Certaine precepts concerning Sacrifices, 17.~and firſt fruites,
  are repeted, 22.~alſo touching different punishment of ſinne comitted
  by ignorance and by ſet wilfulnes: 32.~and accordingly one is ſtoned
  to death, for gathering ſtickes on the Sabbath day. 37.~Al are
  commanded to carie a ſigne in their garments, therby to remember the
  commandments of God.}

And our Lord ſpake to Moyſes, ſaying: \V Speake to the children of
Iſrael, and thou ſhalt ſay to them: When you shal be entred into the
Land of your habitation, which I wil geue you, \V and shal make oblation
to the Lord for an holocauſt, or victime, paying vowes, or voluntarily
offering giftes, or in your ſolemnities burning a ſweete ſauour vnto the
Lord, of oxen or of sheepe: \V whoſoeuer immolateth the victime, shal
offer a ſacrifice of flowre, the tenth part of an ephi tempered with
oyle, which shal haue in meaſure the fourth part of an hin: \V and wine
of the ſame meaſure to powre the libamentes shal he geue for the
holocauſt or for the victime. For euerie lambe \V and ramme there shal be
a ſacrifice of flowre of two tenthes, which shal be tempered with oile
the third part of an hin: \V and wine for the libamente, the third part
of the ſame meaſure, shal he offer for a ſweete ſauour to the Lord. \V
But when thou makeſt an holocauſt or hoſte of oxen, to fulfil thy vow or
for pacifique victimes, \V thou shalt geue for euerie oxe three tenthes
of flowre tempered with oile, which shal haue halfe the meaſure of an
hin: \V and wine to powre libamentes of the ſame meaſure for an oblation
of moſt ſweete ſauour to the Lord. \V So shalt thou doe \V for euerie
oxe and ramme and lambe and buckegoate. \V As wel they that are borne in
the countrie as the ſtrangers \V after one rite shal offer ſacrifices.
%%% 0374
\V There shal be al one precept and iudgement as wel to your ſelues as to
the ſtrangers of the land. \V Our Lord ſpake to Moyſes, ſaying: \V
Speake to the children of Iſrael, and thou shalt ſay to them: \V When
you are come into the land, which I wil geue you, \V and shal eate of
the breades of that countrie, you shal ſeparate firſt fruites to the
Lord \V of your meates. As of your barne floores you ſeparate firſt
fruites, \V ſo of your paſtes shal you geue firſt fruites to the
Lord. \V And if through ignorance you omitte any of theſe thinges, which
the Lord hath ſpoken to Moyſes, \V and by him hath commanded you, from
the day that he beganne to command and ſo forward, \V and the multitude
haue forgotten to do it: they shal offer a calfe out of the heard, an
holocauſte for a moſt ſweete ſauour to the Lord, and the ſacrifice and
libamentes therof, as the ceremonies require, and a buckegoate
%%% o-0336
for ſinne: \V and the prieſt shal pray for al the multitude of the
children of Iſrael: and it shal be forgeuen them, becauſe they ſinned
not wittingly, offering notwithſtanding burnt ſacrifice to the Lord for
them ſelues and for their ſinne and errour: \V and it shal be forgeuen
al the people of the children of Iſrael, and the ſtrangers, that
ſeiourne among them, becauſe it is the fault of al the people through
ignorance. \V But if one ſoule shal ſinne vnwitting, he shal offer a she
goate of a yeare old for his ſinne: \V and the prieſt shal pray for
him, becauſe he ſinned vnwitting before the Lord: and he shal obteine
him pardon, and it shal be forgeuen him. \V As wel to them that are
borne in the countrie as to the ſtrangers one law shal be for al, that
ſinne by ignorance. \V But the ſoule, that
\SNote{Though ſinnes wittingly committed, procedĩg of pride and contempt
of Gods commandment, could not be pardoned by the law: yet ſuch may alſo
be remitted through true repentance.
\Cite{S.~Aug. q.~25. in Num.}}
through pride committeth any thing, whether he be borne in the countrie,
or a ſtranger, (becauſe he hath bene rebellious againſt the Lord) shal
perish out of his people: \V for he hath contemned the word of the Lord,
and made his precept of no effect: therfore shal he be deſtroyed, and
shal beare his iniquitie. \V And it came to paſſe, when the children of
Iſrael were in the wilderneſſe, and had found a man gathering ſtickes on
the Sabbath day, \V they preſented him to Moyſes and Aaron and the whole
multitude. \V Who shut him into priſon, not knowing what they should doe
with him. \V And our Lord ſaid to Moyſes:
\SNote{Seueritie is vſed towardes thoſe that knowĩg Gods wil do
contrarie.
\XRef{Luc.~12. v.~47.}}
Dying let this man die, let al the multitude ſtone him without the
campe. \V And when they had brought him out, they ſtoned
%%% 0375
him, and he died as our Lord had commanded. \V Our Lord alſo ſaid to
Moyſes: \V Speake to the children of Iſrael, and thou shalt ſay to them,
that they make them ſelues
\SNote{The Iewes in Chriſts time hypochritically enlarged theſe fringes,
for vaine ſhew of holines.
\XRef{Mat.~23.}}
fringes in the corners of their garmentes, putting in them ribandes of
hyacinth: \V which when they shal ſee, they may remẽber al the
commandements of the Lord, and not folow their owne cogitations and
eies fornicating after diuers thinges, \V but rather mindful of the
preceptes of the Lord may do them, and be holie to their God. \V I the
Lord your God, that brought you out of the Land of Ægypt, that I might be
your God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
