%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 0345
%%% o-0311
\startChapter[
  title={Chapter 3}
  ]

\Summary{The Leuites are aſsumed to the ſeruice of the Tabernacle,
  14.~numbered by their ſeueral families, and their offices
  diſtinguished, 45.~they are taken to God in place of the firſt borne
  of the children of Iſrael. The reſidue of the firſt borne, aboue the
  number of Leuites, are redemed with price.}

Theſe are the generations of Aaron and Moyſes in the day that our Lord
ſpake to Moyſes in Mount Sinai. \V And theſe be the names of the ſonnes
of Aaron: his firſt begotten Nadab, then Abiu, and Eleazar, and Ithamar.
%%% 0346
\V Theſe are the names of the ſonnes of Aaron the prieſts that were
anointed, and whoſe handes were filled and cõſecrated, to doe the
function of prieſthood. \V For Nadab and Abiu died, when they offered
the ſtrange fire in the ſight of our Lord, in the deſert of Sinai, without
children: and Eleazar and Ithamar did the function of prieſthood in the
preſence of Aaron their father. \V And our Lord ſpake to Moyſes,
ſaying: \V Bring the tribe of Leui, and make them ſtand in the ſight of
Aaron the prieſt to miniſter vnto him, and let them watch, \V and
obſerue whatſoeuer pertaineth to the ſeruice of the multitude before the
tabernacle of teſtimonie, \V and let them keepe the veſſel of the
tabernacle, ſeruing in the miniſterie therof. \V And thou ſhalt geue the
Leuites for a gift, \V to Aaron and to his ſonnes, to whom they are
deliuered of the children of Iſrael. But Aaron and his ſonnes thou ſhalt
appoint ouer the ſeruice of prieſthood.
\SNote{As none but Leuites might ſerue in the tabernacle: ſo none but of
Aarons ſtock might do the office of Prieſthood.}
The ſtranger, that approcheth to miniſter, shal die. \V And our Lord
ſpake to Moyſes, ſaying: \V I haue taken the Leuites from the children
of Iſrael, for euerie firſt borne, that openeth the matrice among the
children of Iſrael, and the Leuites ſhal be mine. \V For the firſt borne
is mine: ſince I ſtroke the firſt borne in the Land of Ægypt: I haue
ſanctified to me whatſoeuer is firſt borne in Iſrael from man vnto
beaſt, they are mine: I the Lord. \V And our Lord ſpake to Moyſes in the
deſert of Sinai, ſaying: \V Number the ſonnes of Leui by the houſes of
their fathers and their families, euerie male from one moneth and
vpward. \V Moyſes numbered, as our Lord had commanded, \V and there were
found the ſonnes of Leui by their names, Gerſon and Caath and Merari. \V
The ſonnes of Gerſon: Lebni and Semei. \V The ſonnes of Caath: Amram,
and Ieſaar, Hebron and Oziel. \V The ſonnes of Merari: Moholi and
Muſi. \V Of Gerſon were two families, the Lebnitical, and Semeitical: \V
of whom were numbered the people of male ſexe from one moneth and
vpward, ſeuen thouſand fiue hundred. \V Theſe ſhal pitch behind the
tabernacle on the Weſt, \V vnder their prince Heliaſaph, the ſonne of
Lael. \V And their charge ſhal be in the tabernacle of couenãt, \V the
tabernacle it ſelfe and the couer therof,
%%% o-0312
the hanging that is drawne before the dores of the roofe of couenant,
and the curtines of the court: the hanging alſo that is hanged in the
entrie of the court of the tabernacle, and whatſoeuer
%%% 0347
pertaineth to the rite of the altar, the cordes of the tabernacle, and
al the furniture therof. \V The kinred of Caath ſhal haue the peoples of
the Amramites, and Ieſaarites, and Hebronites, and Ozielites. Theſe are
the families of the Caithites reckened by their names: \V al of the male
ſexe from one moneth and vpward, eight thouſand ſix hundred, they ſhal
haue the gard of the Sanctuarie, \V and ſhal campe on the ſouth ſide. \V
And their prince ſhal be Elizaphan the ſonne of Oziel: \V and they shal
keepe the arke, and table and the candleſticke, the altars, and the
veſſel of the Sanctuarie, wherin the miniſtration is, and the veile, and
al ſuch kind of implementes. \V And
\SNote{One chiefe Monarch in the Church to whom al other Superiors are
ſubordinate.}
the prince of the princes of the Leuites, Eliazar, the ſonne of Aaron
the prieſt, shal be ouer them that watch for the cuſtodie of the
Sanctuarie. \V But of Merari shal be the peoples of the Moholites, and
Muſites, reckened by their names: \V al of the male kind from one moneth
and vpward, ſix thouſand two hundred. \V Their prince Suriel the ſonne
of Abihaiel: they shal campe on the north ſide. \V Vnder their cuſtodie
shal be the bordes of the tabernacle, and the barres, and the pillers
and their feete, and al thinges that pertaine to this kind of
ſeruice: \V and the pillers of the court round about with their feete,
and the pinnes with the cordes. \V Before the tabernacle of couenant,
that is to ſay, on the eaſt ſide, shal
\SNote{Moyſes is ſtil counted, and hath chiefe place and office among
the Prieſtes, which were abſurde, ſaieth S.~Auguſtin
\Cite{(in Pſal.~98.)}
if he were not a prieſt.}
Moyſes and Aaron campe, with
\SNote{The ſonnes of Moyſes were with him ſo long as he liued, but
after his death they ſerued the Prieſts, as other Leuites did, and were
numbred with the Caathites.
\XRef{1.~Paral.~23. v.~12.}}
their ſonnes, hauing the cuſtodie of the Sanctuarie, in the middes of
the children of Iſrael. What ſtranger ſoeuer cometh therto, shal die. \V
Al the Leuites, that Moyſes and Aaron numbered according to the precept
of our Lord by their families in the male kinde from one moneth and
vpward, were 
%%% Not marked in either text
\LNote{Tvventie tvvo thouſand.}{Immediatly
\MNote{VVhy 300. are omitted in the total ſumme of the Leuites.}
before were numbered of Gerſonites
\XRef{(v.~22.)}
7500. of Caathites
\XRef{(v.~28.)}
8600. and of Merarites
\XRef{(v.~34.)}
6200. which make in al 22300. VVhy then are 300. left out of the total
ſumme? expecially ſeing by and by
\XRef{(v.~43.)}
a ſmaler, and other odde numbers of 273. are expreſſed, and therupon
inferred
\XRef{(v.~46.)}
that the firſt borne of the Iſraelites did ſo much exceede the
Leuites. VVheras if the whole number of Leuites had bene alſo expreſſed
in the general ſumme, as it is contained in the three particular ſummes,
the Leuites ſhould exceede the firſt borne of Iſraelites by the number
of 27. For anſwere to this difficultie, ſome ſuppoſe that the firſt
borne of the Leuites were iuſt 300. and therfore ſo manie ſupplying only
their owne places, there remained iuſt twentie two thouſand to ſupplie
the places of the firſt borne of other Iſraelites, and ſo the Leuites
were fewer then thoſe for whom they were taken vnto Gods ſeruice, by the
number of 273. But that the firſt borne of Leuites were neither more nor
fewer then 300. is not euident by the text. Howſoeuer therfore this
doubt be ſolued, ſure it is, by S.~Hieroms iudgement (cited in the
\XRef{argument)}
that theſe numbers are myſtical.
\MNote{Perfect numbers ſignifie perfection.}
And the iuſt number of 22.~thouſand Leuites may ſignifie (ſaieth
\Cite{Origen, homil.~4. in Num.)}
perfection, required in thoſe that are deſigned to the particular
ſeruice of God; as there be alſo iuſt 22.~Hebrew letters; and
22.~Patriarches from Adam to Iacob, from whom the Iſraelites deſcended.}
twentie two thouſand. \V And our Lord ſaid to Moyſes:
Number the firſt borne of the male ſex of the children of Iſrael, from
one moneth and vpward, and thou shalt haue the ſumme of them. \V And
thou ſhalt take the Leuites vnto me for al the firſt borne of the
children of Iſrael, I am the Lord: and their cattel for al the firſt
borne of the cattel of the children of Iſrael. \V Moyſes reckened, as
our Lord had commanded, the firſt borne of the children of Iſrael. \V
And the males by their names, from one moneth and vpward, were twentie
two thouſand two hundred ſeuentie three. \V And our Lord ſpake to
Moyſes, ſaying: \V Take the Leuites for the firſt borne of the children
of Iſrael, and the cattel of the Leuites for their
%%% 0348
cattel, and the Leuites ſhal be mine. I am the Lord. \V But for the
price of two hundred ſeuentie three, of the firſt borne of the children
of Iſrael,
\SNote{This number exceedeth the other by reaſõ that 300.~Leuites are
omitted in the general ſumme, otherwiſe the number of Iſraelites ſhould
exceede the firſt borne by 27. See the
\XRef{Annotation.}}
that exceede the number of the Leuites, \V thou shalt take fiue ſicles
for euerie head, after the meaſure of the Sãctuarie. A ſicle hath
twentie oboles. \V And thou shalt giue the money to Aaron and to his
ſonnes the price of them that are aboue. \V Moyſes therfore tooke their
money, that were aboue, and whom he redemed of the Leuites, \V for the
firſt borne
%%% o-0313
of the children of Iſrael, a thouſand three hundred ſixtie fiue ſicles,
according to the weight of the Sanctuarie, \V and gaue it to Aaron and
his ſonnes, acording to the word that our Lord had commanded him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
