%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 30
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-30


\project douay-rheims


%%% 0404
%%% o-0362
\startChapter[
  title={Chapter 30}
  ]

\Summary{Voluntarie vowes or oathes of men; 14.~of maides in their
  fathers houſes, 7.~or newly maried; 10.~of widowes, or wiues diuorced,
  11.~and of wiues in their husbands houſes: and how they bind, or are
  made fruſtrate.}

And Moyſes told the children of Iſrael al thinges that our Lord had
commanded him: \V and he ſpake to the princes of the tribes of the
children of Iſrael: This is the word that our Lord hath commanded: \V If
any man
\SNote{He that voweth abſtinence from a thing lawful, maketh it vnlawful
to him ſelfe by his vow.
\Cite{S.~Aug. q.~56. in Num.}}
make a vowe to our Lord, or binde him ſelf by an oath: he shal not make
his word fruſtrate, but al that he promiſed he shal fulfil. \V If a
woman vowe any thing, and binde her ſelf with an oath, ſhe that is in
her fathers houſe, and as yet in maydens age: if her father know the vowe
that she promiſed, and the oath wherwith she bound her ſoule, and hold
his peace, ſhe shal be bound to the vowe: \V Whatſoeuer she promiſed and
ſware, she shal fulfil in deede. \V But if immediatly as he heareth it,
her father do gaine ſay it, both her vowes and her othes
\LNote{shal be fruſtrate.}{Vowes,
\MNote{Vowes of things not commanded.}
which fathers and huſbandes coulde
fruſtrate in their daughters and wiues, muſt needes be ment of things
not commanded by God. For it was not in their powre to diſpence in Gods
precepts. As for example, touching the feaſt of
\Fix{Expiration,}{Expiation,}{obvious typo, fixed in other}
it was not \Emph{in the arbitrement} of fathers and huſbandes (as here
it was
\XRef{v.~14.})
whether their daughters and wiues ſhould faſt or no:
\CNote{\Cite{S.~Aug. q.~57. in Num.}}
but was abſolutely ſaied:
\XRef{(Leuit.~23. v.~29.)}
\Emph{Euerie ſoule that is not afflicted} (that is, which faſteth
not) \Emph{this day, shal perish out of his people}. And ſo in other
precepts were neceſſarie obligations. Beſides which it is gratful alſo
to God, that his ſeruantes bind them ſelues to other good workes, wherto
they are not otherwiſe bound.}
shal be fruſtrate, neither shal she be bound to the promiſſe, for that
her father hath gaineſaid it. \V If she haue a husband, and vowe any
thing, and the word once going out of her mouth binde her ſoule by an
oath: \V the day that her husband heareth it, and doth not gaine ſay it,
%%% o-0363
she shal be bound to the vowe, and shal render whatſoeuer she
promiſed. \V But if as ſowne as he heareth he gaine ſay it, and make
her promiſes fruſtrate, and the wordes wherwith she had bound her ſoule:
our Lord wil be
\SNote{In this caſe God wil not impute it to her for a sĩne.
\Cite{S.~Aug. q.~58.}}
propitious to her. \V The widowe, and she that is deuorced whatſoeuer
they vowe, they shal render. \V The wife in the houſe of her husband,
when she hath bound her ſelf by vowe and by oath, \V if her husband
heare, and hold his peace, neither doe
%%% 0405
gayneſay the promiſſe, she shal render whatſoeuer she had promiſſed. \V
But if forthwith he gayneſay it, she shal not be holden bound to the
promiſſe: becauſe her husband gayneſaid it, and our Lord wil be
propitious to her. \V If she vowe and binde her ſelfe by oath, to
\SNote{By afflicting of the ſoule is here vnderſtood, reſtraining
ſenſual delectation.
\Cite{S.~Aug. q.~57. in Num.}}
afflict her ſoule by faſting, or abſtinence from other thinges, it shal
be in the arbitrement of her husband, whether she shal do it, or not do
it. \V And if the husband hearing it hold his peace, and differre
ſentence til an other day: whatſoeuer she had vowed and promiſed, she
shal render: becauſe immediatly as he heard it, he held his peace, \V
but if he gayneſay it after that he knew it,
\SNote{The huſbãd reuoking his conſent, once geuen by word, or by
ſilence to his wiues vow, ſinned, but the wife was bound to obey him,
and ſo was excuſed.}
him ſelfe shal beare her iniquitie. \V Theſe are the lawes, which our
Lord appointed to Moyſes betwen the husband and the wife, betwene the
father and the daughter, that is as yet in maidens age, or that abideth
in her fathers houſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
