%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 0356
%%% o-0320
\startChapter[
  title={Chapter 7}
  ]

\Summary{In dedication of the Tabernacle, the princes of the twelue
  tribes offer iointly ſix waines, and twelue oxen. 11.~Then euerie
  prince ſeueral daies make other oblations. 89.~And God ſpeaketh to
  Moyſes from the Propitiatorie.}

And it came to paſſe in the day that Moyſes finiſhed the tabernacle, and
erected it: he anoynted alſo and ſanctified it with al the veſſel
therof, the altar likewiſe and al the veſſel therof. \V The princes of
Iſrael and the heades of the families, that were in euerie tribe, and
the rulers of them, that had bene numbered, offered \V giftes before our
Lord ſix waines couered, with twelue oxen. Two princes offered one
waine, and euerie man one oxe and they offered them before the
tabernacle. \V And our Lord ſaid to Moyſes: \V Take them of their handes
to ſerue in the miniſterie of the tabernacle, and thou shalt deliuer
them to the Leuites according to the order of their miniſterie. \V When
Moyſes therfore had taken the waines and the oxen, he deliuered them to
the Leuites. \V Two waines and foure oxen he gaue to the ſonnes of
Gerſon, according to that which was neceſſarie for them. \V The other
foure waines, and the eight oxen he gaue to the ſonnes of Merari
according to their offices and ſeruice, vnder the hand of Ithamar the
ſonne of Aaron the prieſt. \V But to the ſonnes of Caath he gaue not
waines and oxen: becauſe they ſerue in the ſanctuarie, and carie their
burdens
\SNote{Of great reuerence they caried the arke, and propitiatorie, and
the holie veſſel ordinarily vpon their ſhoulders, yet the ſame were
ſonetimes caried on waines.
\XRef{2.~Reg.~6.}}
vpon their owne shoulders. \V The princes therfore offered vnto the
dedication of the altar, the daie wherin it was anoynted, their oblation
before the altar. \V And our Lord ſaid to Moyſes: Let the princes one
and one euerie day offer their giftes vnto the dedication of the
altar. \V The firſt day Nahaſſon the ſonne of Aminadab of the tribe of
Iuda offered his oblation: \V and there were in it a ſiluer plate of an
hundred and thirtie ſicles weight, a phial of ſiluer hauing ſeuentie
ſicles according to the weight of the Sanctuarie, both ful of flower
tempered with oyle for a ſacrifice: \V a little morter of ten ſicles of
gold ful of incenſe: \V an oxe out of the hearde, and a ramme, and a
lambe of a yeare old for an holocauſt: \V and a bucke goat for ſinne: \V
and for a ſacrifice of pacifiques, two oxen, fiue rammes, fiue bucke
goates, fiue lambes of a yeare old. This is the oblation of Nahaſſon the
ſonne of Aminidab. \V The
%%% 0357
ſecond day offered Nathanael the ſonne of Suar, prince of the tribe of
Iſſachar, \V a plate of ſiluer weying an hundred thirtie ſicles, a phial
of ſiluer hauing ſeuentie ſicles, according to the weight of the
Sanctuarie, both ful of flowre tempered with oile for a ſacrifice: \V a
%%% o-0321
litle morter of gold hauing ten ſicles ful of incenſe: \V an oxe out of
the heard, and a ramme, and a lambe of a yeare old for an holocauſt: \V
and a bucke goate for ſinne: \V and for a ſacrifice of pacifiques, two
oxen, fiue rammes, fiue bucke goates, fiue lambes of a yeare old. This
was the oblation of Nathanael the ſonne of Suar. \V The third day the
prince of the ſonnes of Zabulon Eliab the ſonne of Helon, \V offered a
plate of ſiluer weighing an hundred thirtie ſicles, a phial of ſiluer
hauing ſeuentie ſicles after the weight of the Sanctuarie, both ful of
flower tempered with oyle for a ſacrifice: \V a litle morter of gold
weighing ten ſicles ful of incenſe: \V an oxe out of the heard, and a
ramme, and a lambe of a yeare old for an holocauſt: \V and a bucke-goate
for ſinne, \V and for a ſacrifice of pacifiques, two oxen, fiue rammes,
fiue bucke goates, fiue lambes of a yeare old. This is the oblation of
Eliab the ſonne of Helon. \V The fourth day the prince of the ſonnes of
Ruben, Eliſur the ſonne of Sedeur, \V offered a plate of ſiluer weighing
an hundred thirtie ſicles, a phial of ſiluer hauing ſeuentie ſicles
after the weight of the Sanctuarie, both ful of flower tempered with
oile for a ſacrifice: \V a litle morter of gold weighing ten ſicles ful
of incenſe: \V an oxe out of the heard, and a ramme, and a lambe of a
yeare old, for an holocauſt: \V and a bucke goate for ſinne: \V and for
pacifiques hoſtes two oxen, fiue rammes, fiue bucke goates, fiue lambes
of a yeare old. This was the oblation of Eliſur the ſonne of Sedeur. \V
The fifth day the prince of the ſonnes of Simeon Salamiel the ſonne of
Suriſaddai, \V offered a plate of ſiluer weighing an hundred thirtie
ſicles, a phial of ſiluer hauing ſeuentie ſicles after the weight of the
Sanctuarie, both ful of flowre tempered with oile for a ſacrifice: \V a
litle morter of gold weighing ten ſicles ful of incenſe: \V an oxe out
of the heard, and a ramme, and a lambe of a yeare old for an
holocauſte: \V and a bucke goate for ſinne: \V and for pacifique hoſtes,
two oxen, fiue rammes, fiue bucke goates, fiue lambes of a yeare
old. This was the oblation of Salamiel the ſonne of Suriſaddai.
%%% 0358
\V The ſixt day the prince of the ſonnes of Gad, Eliaſaph the ſonne of
Duel \V offered a plate of ſiluer weighing a hundred thirtie ſicles, a
phial of ſiluer hauing ſeuentie ſicles after the weight of the
Sanctuarie, both ful of flowre tempered with oyle for a ſacrifice: \V a
litle morter of gold weighing ten ſicles ful of incenſe: \V an oxe out
of the heard, and a ramme, & a lambe of a yeare old for an holocauſt: \V
and a bucke goate for ſinne: \V and for pacifique hoſtes, two oxen, fiue
rammes, fiue bucke goates, fiue lambes of a yeare old. This was the
oblation of Eliaſaph the ſonne of Duel. \V The ſeuenth day the prince of
the ſonnes of Ephraim, Eliſama the ſonne of Ammiud \V offered a plate of
ſiluer weighing a hundred thirtie ſicles, a phial of ſiluer hauing
ſeuentie ſicles after the weight of the Sanctuarie, both ful of flowre
tempered with oile for a ſacrifice: \V a litle morter of gold weighing
ten ſicles ful of
%%% o-0322
incenſe: \V an oxe out of the heard, and a ramme, and a lambe of a yeare
old for an holocauſt: \V and a bucke goate for ſinne: \V and for
pacifique hoſtes, two oxen, fiue rammes, fiue bucke goates, fiue lambes
of a yeare old. This was the oblation of Eliſama the ſonne of Ammiud. \V
The eight day the prince of the ſonnes of Manaſſes, Gamaliel the ſonne
of Phadaſſur, \V offered a plate of ſiluer weighing an hundred thirtie
ſicles, a phial of ſiluer hauing ſeuentie ſicles after the weight of the
Sanctuarie, both ful of flowre tempered with oile for a ſacrifice: \V a
litle morter of gold weighing ten ſicles, ful of incenſe: \V an oxe out
of the heard, and a ramme, and a lambe of a yeare old for an
holocauſt: \V and a bucke goate for ſinne: \V and for pacifique hoſtes,
two oxen, fiue rammes, fiue bucke goates, fiue lambes of a yeare
old. This was the oblation of Gamaliel the ſonne of Phadaſſur. \V The
ninth day the prince of the ſonnes of Beniamin, Abidan the ſonne of
Gedeon, \V offered a plate of ſiluer weighing an hundred thirtie ſicles,
a phial of ſiluer hauing ſeuentie ſicles after the weight of the
Sanctuarie, both ful of flower tempered with oile for a ſacrifice: \V a
litle morter of gold weighing ten ſicles ful of incenſe: \V an oxe out
of the heard, and a ramme, and a lambe of a yeare old for an
holocauſt: \V and a bucke goate for ſinne: \V and for pacifique hoſtes,
two oxen, fiue rammes, fiue bucke goates, fiue lambes of a yeare
old. This was the oblation of Abidan the ſonne of Gedeon. \V The tenth
%%% 0359
day the prince of the ſonnes of Dan, Ahiezer the ſonne of Ammiſaddai \V
offered a plate of ſiluer weighing an hundred thirtie ſicles, a phial of
ſiluer hauing ſeuentie ſicles, after the weight of the Sanctuarie, both
ful of flowre tempered with oile for a ſacrifice: \V a litle morter of
gold weighing ten ſicles ful of incenſe: \V an oxe out of the heard, and
a ramme, and a lambe of a yeare old for an holocauſt: \V and a bucke
goate for ſinne: \V and for pacifique hoſtes, two oxen, fiue rammes,
fiue bucke goates, fiue lambes of a yeare old. This was the oblation of
Ahiezer the ſonne of Ammiſaddai. \V The eleuenth day the prince of the
ſonnes of Aſer, Phegiel the ſonne of Ochran \V offered a plate of ſiluer
weighing an hundred thirtie ſicles, a phial of ſiluer hauing ſeuentie
ſicles after the weight of the Sanctuarie, both ful of flowre tempered
with oile for a ſacrifice: \V a litle morter of gold weighing ten ſicles
ful of incenſe: \V an oxe out of the heard, and a ramme, and a lambe of
a yeare old for an holocauſt: \V and a bucke goate for ſinne: \V and for
pacifique hoſtes, two oxen, fiue rammes, fiue bucke goates, fiue lambes
of a yeare old. This was the oblation of Phegiel the ſonne of Ochran. \V
The twelfth day the prince of the ſonnes of Nephthali, Ahira the ſonne
of Enan \V offered a plate of ſiluer weighing an hundred thirtie ſicles,
a phial of ſiluer hauing ſeuentie ſicles after the weight of the
Sanctuarie, both ful of flowre tempered with oile for a ſacrifice: \V a
litle morter of gold weighing ten ſicles, ful of incenſe: \V an oxe out
of the heard, and a ramme, and a lambe of a yeare old for an holocauſt: \V
and a bucke goate for ſinne: \V and for pacifique hoſtes, two oxen,
%%% o-0323
fiue rammes, fiue bucke goates, fiue lambes of a yeare old. This was the
oblation of Ahira the ſonne of Enan. \V Theſe thinges were offered of
the princes of Iſrael in the dedication of the altar, in the day wherin
it was conſecrated: plates of ſiluer twelue: phials of ſiluer twelue:
litle morters of gold twelue: \V ſo that one plate had an hundred and
thirtie ſicles of ſiluer, and one phial had ſeuentie ſicles: that is, in
the whole of al the veſſel of ſiluer two thouſand foure hundred ſicles,
by the weight of the Sanctuarie: \V litle morters of gold twelue ful of
incenſe weighing ten ſicles a peece, by the weight of the Sanctuarie:
that is, in the whole an hundred twentie ſicles of gold: \V oxen out of
the heard for an holocauſt twelue, rammes twelue, lambes of a yeare old
twelue,
%%% 0360
and their libamentes: twelue bucke goates for ſinne. \V For pacifique
hoſtes, oxen twentie fowre, rammes ſixtie, bucke goates ſixtie, lambes
of a yeare old ſixtie. Theſe thinges were offered in the dedication of
the altar, when it was anointed. \V And when Moyſes entred into the
tabernacle of couenant, to conſult the oracle, he heard the voice of him
that ſpake to him from the propitiatorie, that was ouer the arke betwen
the two Cherubs, from whence alſo he ſpake to him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
