%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 0382
%%% o-0343
\startChapter[
  title={Chapter 19}
  ]

\Summary{A redde cow is offered in burnt victime for ſinne: 9.~whoſe
  ashes are mingled in water, for expiation of diuers legal vncleanes,
  11.~as by touching the dead, 14.~by entring into the tent of the dead,
  alſo the veſſel that is therin, and the veſſel that lacketh a couer,
  22.~and whatſoeuer the vncleane toucheth.}

And our Lord ſpake to Moyſes and Aaron, ſaying: \V This is the religion
of
%%% !!! Can we do these SNotes better?
\SNote{The ſame things ſaith S.~Auguſt.
\Cite{(q.~33.)}
and Theodoret
\Cite{(q.~16. in Num.)}
which were ſhadowed and prefigured in the old Teſtament, are reueled and
manifeſted in the New. And this ſpecial ſacrifice prefigured diuers
particular things in Chriſts Paſſiõ.}
the victime, which the Lord hath appointed. Command the children of
Iſrael, that they bring vnto thee
\SNote{His body of \Emph{Adam} ſignifying \Emph{redde earth}.}
a redde cow of
\SNote{At the age of 33.~yeares.}
ful age, wherin is
\SNote{Alwayes moſt pure from ſinne.}
no blemiſh, and that hath
\SNote{Free from al bondage.}
not caried yoke: \V and you ſhal deliuer her to Eleazar the prieſt, who
bringing her forth
\SNote{Crucified without the gate of Hieruſalem.}
without the campe, shal immolate her in the ſight of al: \V and dipping
his finger in her
\SNote{Al Sacramẽts haue their vertue of Chriſts bloud.}
bloud, shal ſprinkle it againſt the doores of the tabernacle ſeuen
times, \V and ſhal burne her in the ſight of al, committing
\SNote{From the ſole of his feet to the toppe of his head al wounded,
with the whippes, crowne, nailes, &c.}
aſwel her skinne and the flesh as the bloud, and the
\Fix{dong}{dung}{possible typo, fixed in other}
to the fire. \V
\SNote{VVood of the croſſe brought life to the world.}
Wood alſo of the cedar, and
\SNote{Liuely heate of ſo infinite charitie diſſolueth the coldnes of death.}
hyſſope, and ſcarlet
\SNote{Chriſts ſuffering an exãple that we muſt alſo ſuſtaine afflictions.}
twiſe died shal the prieſt
%%% 0383
caſt into the flame, that waſteth the cow. \V And then at length washing
his garmentes, and his bodie, he shal enter into the campe, and
\SNote{Thoſe that procured, or cooperated to Chriſts death were polluted
with ſinne.}
shal be polluted vntil euen. \V But he
%%% o-0344
alſo that burneth her, shal wash his garmentes and his bodie, and shal
be vncleane vntil euen. \V And
\SNote{Ioſeph and Nicodemus buried Chriſt.}
a man that is cleane shal gather the ashes of the cow, and shal powre
them out without the campe in
\SNote{His ſepulcher was glorious.}
a moſt cleane place, that they may be reſerued for the multitude of the
children of Iſrael, and for water of aſperſion: becauſe the cowe was
burnt for ſinne. \V And when he that caried the ashes of the cow, hath
washed his garmentes,
\SNote{Thoſe alſo that buried him, neded cleanſing by his Paſſion, not
for that work, but for their ſinnes.}
he shal be vncleane
\SNote{The old law did not remitte ſinnes but the new.}
vntil euen. The children of Iſrael, and the ſtrangers that dwel among
them, shal haue this for a holie thing by a perpetual ordinance. \V He
that toucheth the dead corps of a man, and is vncleane therfore ſeuen
daies: \V shal be ſprinkled of
\SNote{Baptiſme in the name of the B.~Trinitie}
this water the third day, and the ſeuenth, and ſo shal be cleanſed. If
he were not ſprinkled, the third day
\SNote{without which no other Sacramẽt auaileth.}
the ſeuenth day he can not be clenſed. \V Euerie one that toucheth the dead
corps of mans ſoule, and is not ſprinkled with this commiſtion, shal
pollute the tabernacle of the Lord, and shal perish out of Iſrael:
becauſe he was not ſprinkled with the water of expiation, he shal be
vncleane, and his filthineſſe shal remaine vpon him. \V This is the law of
the man that dieth in a tabernacle: Al that enter into his tent, and al
the veſſel that are there, shal be polluted ſeuen daies. \V The veſſel,
that hath no couer, nor bynding ouer it, shal be vncleane. \V If any man
in the field touch the corps of a man that was ſlaine, or that died of
himſelf, or his bone, or graue, he shal be vncleane ſeuen daies. \V And
they shal take of the ashes of combuſtion and of ſinne, and shal powre
liuing water vpon them into a veſſel: \V in the which when a man that is
cleane hath dipped hyſſope, he shal ſprinkle therwith al the tent, and
al the implementes, and the men polluted with ſuch contagion: \V and in
this maner he that is cleane shal purge the vncleane the third and
ſeuenth day. And being expiated the ſeuenth day, he shal wash both
himſelf and his garmentes, and be vncleane vntil euening. \V If anie man
be not expiated after this rite, his ſoule shal periſh out of the middes
of the Church: becauſe he hath polluted the Lordes Sanctuarie, and is not
ſprinkled with water of luſtration. \V This precept shal be an ordinance
for euer. He alſo that ſprinkleth the waters, shal wash his garmentes:
%%% 0384
Euerie one that toucheth the waters of expiation, ſhal be vncleane
vntil euen. \V Whatſoeuer he toucheth that is vncleane, he ſhal make it
vncleane: and the ſoule that toucheth anie of theſe thinges, ſhal be
vncleane vntil euen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
