%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 0360
%%% o-0323
\startChapter[
  title={Chapter 8}
  ]

\Summary{Seuen lampes are ſo placed on the golden candleſtucke, that
  they may shine towardes the breades of propoſition. 5.~The ordination
  of the Leuites. 24.~And at what age they shal ſerue in the
  tabernacle.}

And our Lord ſpake to Moyſes, ſaying: \V Speake to Aaron, and thou shalt
ſay to him: When thou shalt place the ſeuen lampes, let the candleſticke
be ſette vp in the ſouth part ouer againſt the north, toward the table
of the breades of propoſition, ouer againſt that part toward which the
candleſticke looketh, shal they shine. \V And Aaron did ſo, and he put
the lampes vpon the candleſticke, as our Lord had cõmanded Moyſes. \V
This was the making of the candleſticke, of beaten gold, as wel the
middle shaft, as al thinges that aroſe out of both ſides of the
branches: according to the example which our Lord ſhewed to Moyſes, ſo
wrought he the candleſticke. \V And our Lord ſpake to Moyſes ſaying: \V
Take the Leuites out of the middes of the children of Iſrael, and thou
shalt purifie them, \V according to this rite: Let them be ſprinkled
with
\SNote{This water was mixed with aſhes of a redde cow ſacrificed without
the campe.
\XRef{chap.~19.}}
the water of luſtration, & ſhaue al the haires of their fleſh. And when
they haue washed their garmentes, and are cleanſed, \V they shal take an
oxe out of the heardes, and his libament flowre tempered with oyle: and
an other oxe out of the heard thou shalt take for ſinne: \V and thou
ſhalt bring the Leuites before the tabernacle of couenant, calling
%%% o-0324
together al the multitude of the children of Iſrael. \V And when the
Leuites are before the Lord, the children of Iſrael shal put their
handes vpon them, \V and Aaron shal offer the Leuites, a gift in the
ſight of the Lord from the children of Iſrael, that they may ſerue in
his miniſterie. \V The Leuites alſo shal put their handes vpon the
heades of the oxen, of the which thou shalt make one for ſinne, and the
other for holocauſte of the Lord, to pray for them. \V And thou shalt
ſette
%%% 0361
the Leuites in the ſight of Aaron and of his ſonnes, and being offered
ſhal conſecrate them to the Lord, \V and shal ſeparate them from the
middes of the children of Iſrael, to be mine. \V And afterward they shal
enter into the tabernacle of couenant, to ſerue me. And thou shalt ſo
purifie and conſecrate them for an oblation of the Lord: becauſe they
were geuen me for a gift of the children of Iſrael. \V For the firſt
borne that open euerie matrice in Iſrael, I haue taken them. \V For mine
are al the firſt borne of the children of Iſrael, as wel of men as of
beaſtes. From the day that I ſmote euerie firſt borne in the Land of
Ægypt, haue I ſanctified them to me: \V and I haue taken the Leuites for
al the firſt borne of the children of Iſrael: \V and haue deliuered them
for a gift to Aaron and his ſonnes out of the middes of the people, to
ſerue me for Iſrael in the tabernacle of couenant, and to pray for them
that there be no plague among the people, if they should preſume to
approch vnto my Sanctuarie. \V And Moyſes and Aaron and al the multitude
of the children of Iſrael did concerning the Leuites the thinges that
our Lord had commanded Moyſes: \V and they were purified, and washed
their garmentes. And Aaron
\SNote{Aaron hauing receiued the Leuites, preſẽted them to God, and ſo
addicted them to their deſigned offices.}
eleuated them in the ſight of our Lord, and prayed for them, \V that
being purified they might enter to their offices into the tabernacle of
couenant before Aaron & his ſonnes. Euen as our Lord had cõmanded Moyſes
touching the Leuites, ſo was it done. \V And our Lord ſpake to Moyſes,
ſaying: \V This is the law of the Leuites: From twentie fiue yeares and
vpward, they shal enter in to miniſter in the tabernacle of couenant. \V
And when they shal haue accomplished the fiftith yeare of their age,
they shal ceaſe to ſerue: \V and shal be the miniſters of their brethren
in the tabernacle of couenant, to keepe the thinges that are cõmended to
them, but not to do the verie workes. Thus shalt thou diſpoſe to the
Leuites in their cuſtodies.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
