%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 27
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-27


\project douay-rheims


%%% 0399
%%% o-0358
\startChapter[
  title={Chapter 27}
  ]

\Summary{Salphaads daughters ſuccede to their fathers
  inheritance. 8.~And the law is eſtablished that for lack of ſonnes
  daughters shal inherite, and for lack alſo of daughters the next of
  kinne. 12.~God commandeth Moyſes to aſcend into Mount Abarim, and
  thence view the promiſed Land, but fortelleth him that he shal die,
  and not goe into it. 15.~He then prayeth God to prouide an other to
  lead the people, 18.~and Ioſue is deſigned in preſence of Eleazar and
  the people.}

And there came the daughters of Salphaad, the ſonne of Hepher, the ſonne
of Galaad, the ſonne of Machir, the ſonne of Manaſſes, who was the ſonne
of Ioſeph: whoſe names are, Malaa, and Noa, and Hegla, and Melcha, and
Therſa. \V And they ſtood before Moyſes and Eleazar the prieſt, and al
the princes of the people at the doore of the tabernacle of couenant,
and ſaid: \V Our father died in the deſert, neither was he in the
ſedition, that was raiſed againſt our Lord vnder Core, but he died in
\SNote{For the general murmuring, wherof al the people were guiltie.
\XRef{cha.~14. v.~29.}}
his owne ſinne: he had no men children. Why is his name taken away out
of his familie, becauſe he hath not a ſonne? Geue vs poſſeſſion among
the kinne of our father. \V And Moyſes referred their cauſe to the
iudgement of our Lord. \V Who ſaid to him: \V The daughters of Salphaad
require a iuſt thing: geue them poſſeſſion among their fathers kinne,
and let them ſuccede him in the inheritance. \V And to the children of
Iſrael thou ſhalt ſpeake theſe thinges: \V When a man dieth without a
ſonne, his inheritance ſhal paſſe to his daughter. \V If he haue no
daughter, he ſhal haue his brethren his ſucceſſours. \V And if he haue
no brethren neither, you ſhal geue the inheritance to his fathers
brethren. \V But if he haue no ſuch vncles by the father neither, the
inheritance ſhal be geuen to them that are the next of kinne: and this
ſhal be to the children of Iſrael a holie ordinance by a perpetual law,
as the Lord hath commanded Moyſes. \V Our Lord alſo ſaid to Moyſes: Goe
vp into this mountaine Abarim, and view from thence the Land which I wil
geue to the children of Iſrael: \V and when thou ſhalt haue ſeene it,
thou alſo ſhalt goe to thy people, as thy brother Aaron is gone: \V
becauſe you did offend me in the deſert Sin in the contradiction of the
multitude, neither would you ſanctifie me before them vpon the
waters. Theſe are the waters of contradiction in
%%% 0400
Cades of the deſert Sin. \V To whom Moyſes anſwered: \V Our Lord the God
of the ſpirites of al flesh prouide a man, that may be ouer this
multitude: \V and may goe out and enter in before them, and bring them
out, or bring them in: leſt the
%%% o-0359
people of our Lord be as ſheepe without a
\SNote{Temporal Princes are alſo paſtors, or ſhepheards of the people:
but this made not Ioſue ſupreme in ſpiritual cauſes. For it is clere in
the next lines that he had but \Emph{part of Moyſes his glorie}, or
office: and that was to be temporal prince, Eleazar being chief in
cauſes ſpiritual, \Emph{before vvhom and the multitude} he was ordained
Duke, but \Emph{Eleazar conſulted God for him, and directed} his
principal actions, called here \Emph{his going out, and going in}.}
paſtor. \V And our Lord ſaid to him: Take Ioſue the ſonne of Nun, a man
in whom is the Spirit, and put thy hand vpon him. \V Who ſhal ſtand
before Eleazar the prieſt and al the multitude: \V and thou shalt geue
him preceptes in the ſight of al, and part of thy glorie, that al the
ſynagogue of the children of Iſrael may heare him. \V For him, if anie
thing be to be done, Eleazar the prieſt shal conſult the Lord. At his
word shal he goe out and shal goe in, and al the children of Iſrael with
him, and the reſt of the multitude. \V Moyſes did as our Lord had
commanded. And when he had taken Ioſue, he ſette him before Eleazar the
prieſt, and al the aſſemblie of the people. \V And impoſing his handes
on his head, he repeted al thinges that our Lord had commanded.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
