%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 0379
%%% o-0340
\startChapter[
  title={Chapter 17}
  ]

\Summary{Moyſes taking of the princes of twelue tribes twelue roddes,
  and one of Aaron for the tribe of Leui, layeth them al in the
  tabernacle al night, 8.~where Aarons rodde (and none of the reſt)
  buddeth, bloometh, and bringeth forth fruite. 9.~And al being shewed
  to the people, Aarons is caried back, and kept for a monument in the
  tabernacle.}

And our Lord ſpake to Moyſes, ſaying: \V Speake to the children of
Iſrael, and
\SNote{For more ſatisfaction to the whole people God confirmeth Aarons
authoritie by a new miracle.
\XRef{v.~10.}}
take a rodde of euerie one of their kinredes, of al the princes of their
tribes, twelue roddes, and the name of euerie one thou shalt write vpon
his rodde. \V And the name of Aaron shal be in the tribe of Leui, and
one rodde shal conteine al their families: \V and thou shalt lay them in
the tabernacle of couenant before the teſtimonie, where I wil ſpeake to
thee. \V Whomſoeuer of theſe I shal chooſe, his rodde shal bloſſome: and
I shal ſtay from me the grudginges of the children of Iſrael, wherwith
they murmur againſt you. \V And Moyſes ſpake to the children of Iſrael:
and al the princes gaue him roddes by euerie tribe: and they were twelue
roddes beſides Aarons rodde. \V Which when Moyſes had laid before our
Lord in the tabernacle of teſtimonie: \V returning the day folowing he
found that
\LNote{Aarons rodde.}{This
\MNote{Example of one miracle proueth the poſſibilitie of an other.}
rodde without roote, neither it ſelf planted, nor anie liue branch
ingraffed therin, al drie without iuyce or moiſture, bringing forth
buddes, flowers, and fruite, was a figure that our B.~Ladie ſhould beare
a ſonne, and remaine a Virgin. And this example euidently demonſtrateth
that ſhe could ſo do, it being no more contrarie to nature, that a virgin
ſhould conceiue, and beare a child without loſſe of virginitie, then
that the drie rodde ſhould bud and beare fruite, without ordinarie
concourſe of nature.
\Cite{S.~Aug. ſer.~3. de temp.}
\Cite{S.~Greg. Niſſen. de nattuit. Dom.}
\Cite{S.~Bernard. ho.~2. in Euang. Miſsus eſt.}}
Aarons
\SNote{The rodde ſignified the B.~Virgin mother; and the almondes
Chriſt, to wit, the vtter pill his humanitie, the ſhel his croſſe, the
kernel his diuinitie; VVho \Emph{pacified by the bloud of his croſſe, al
thinges in earth and in heauen.}
\XRef{Colloſ.~1.}
\Cite{S.~Aug. ſer.~3. de temp.}}
rodde in the houſe of Leui was budded: and the buddes therof ſwelling,
the bloſſomes were shotte forth, which ſpredding the leaues, were
\TNote{\L{deformati} fully formed.}
fashioned into
%%% !!! This SNote is really part of the above SNote. How should we
%%% handle this?
%%% \SNote{}
almondes. \V Moyſes therfore brought forth al the roddes from the ſight
of our Lord to al the children of Iſrael: and
%%% o-0341
they ſaw and euerie one receiued their roddes. \V And our Lord ſaid to
Moyſes: Carie backe Aarons rodde into the tabernacle of teſtimonie, that
it may be kept there for a ſigne of the rebellious children of Iſrael,
and let their complaintes ceaſe from me, leſt they die. \V And Moyſes
did as our Lord had commanded. \V And the children of Iſrael ſaid to
Moyſes: Behold we are conſumed, we are al perished. \V Whoſoeuer
approcheth to the tabernacle of our Lord, he dieth. Are we al to be
deſtroyed vnto vtter conſumption?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
