%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 0341
%%% o-0307
\startChapter[
  title={Chapter 1}
  ]

\Summary{Al
\MNote{The firſt part.

Of thoſe which are nũbred of the 12.~Tribes fitte for warre, & of the
Leuites deſigned to ſerue the Tabernacle.}
the men of twelue tribes of Iſrael, of the age of twentie
  yeares and vpwardes (but not vnder, nor wemen,) are numbred: 20.~and
  are found in al ſix hundred thirtie thouſand fiue hundred
  fiftie. 47.~The Leuites not yet numbred, are deſigned to ſerue about
  the Tabernacle.}

And our Lord ſpake to Moyſes in the deſert of Sinai in the tabernacle of
couenant, the firſt day of the ſecond moneth, the ſecond yeare of their
going out of Ægypt, ſaying: \V Take the ſumme of the whole aſſemblie of
the children of Iſrael by their kinredes, & houſes, and the names of
euerie one, whatſoeuer of the male ſexe, \V from the twentith yeare and
vpward, of al the ſtrong men of Iſrael, and you shal number them by
their troupes, thou and Aaron. \V And there shal be with you the princes
of the tribes, and of the houſes in their kinredes, \V whoſe names are
theſe: Of Ruben, Elizur the ſonne of Sedeur. \V Of Simeon, Salamiel the
ſonne of Suriſaddai. \V Of Iuda, Nahaſſon, the ſonne of Aminadab. \V Of
Iſſachar, Nathanael the ſonne of Suar. \V Of Zabulon Eliab the ſonne of
Helon. \V And of the ſonnes of Ioſeph, of Ephraim, Eliſama the ſonne of
Ammiud. Of Manaſſes, Gamaliel, the ſonne of Phadaſſur. \V Of Beniamin,
Abidan the ſonne of Gedeon. \V Of Dan, Ahiezer the ſonne of
Ammiſaddai. \V Of Aſer, Phegiel the ſonne
%%% o-0308
of Ochran. \V Of Gad, Eliazaph the ſonne of Duel. \V Of Nephthali, Ahira
the ſonne of Enan. \V Theſe are the moſt noble princes of the multitude
by their tribes and kinredes,
%%% 0342
and the heades of the hoſte of Iſrael: \V whom Moyſes and Aaron tooke
with al the multitude of the common people: \V and aſſembled them the
firſt day of the ſecond moneth, reckening them by the kinredes, and
houſes, and families, and heades, and names of euerie one from the
twentith yeare and vpwarde, \V as our Lord had comanded Moyſes. And they
were numbered in the deſert of Sinai. \V Of Ruben the firſt begotten of
Iſrael, by their generations and families and houſes, and names of
euerie heade, al that is of the male ſexe, from twentith yeare and
vpwarde, of them that goe forth to warre, \V fourtie ſix thouſand fiue
hundred. \V Of the ſonnes of Simeon by the generations and families, and
houſes of their kinredes were reckened by the names and heades of euerie
one, al that is of the male ſexe, from twentith yeare and vpward, of
them that goe forth to warre, \V fiftie nine thouſand three hundred. \V
Of the ſonnes of Gad, by the generations and families and houſes of
their kinredes were reckned by the names of euerie one from twentie yeares
and vpwarde, al that went forth to warre, \V fourtie fiue thouſand ſix
hundred fiftie. \V Of the ſonnes to Iuda by the generations and families
and houſes of their kinredes, by the names of euerie one from the
twentith yeare and vpward, al that could goe forth to warre, \V were
reckned ſeuentie foure thouſand ſix hundred. \V Of the ſonnes of
Iſſachar, by the generations and families and houſes of their kinredes,
by the names of euerie one from the twentith yeare and vpward, al that
went forth to warre, \V were reckened fiftie foure thouſand foure
hũdred. \V Of the ſonnes of Zabulon, by the generations and families and
houſes of their kinredes, were reckened by names of euerie one from the
twentith yeare and vpward, al that could goe forth to warre, \V fiftie
ſeuen thouſand foure hundred. \V Of the ſonnes of Ioſeph, namely of the
ſonnes of Ephraim by the generations families and houſes of their
kinredes were reckened by the names of euerie one, from the twentith
yeare and vpward, al that could goe forth to warre, \V fourtie thouſand
fiue hundred. \V Moreouer of the ſonnes of Manaſſes, by the generations
and families and houſes of their kinredes, were reckened by the names of
euerie one from twentie yeares and vpward, al that could goe forth to
warre, \V thirtie two thouſand two hundred. \V Of the ſonnes of Beniamin
by the generations
%%% 0343
and families and houſes of their kinredes were reckened by names of
euerie one from the twentith yeare and vpward, al that could goe forth
to warre, \V thirtie fiue thouſand foure hundred. \V Of the ſonnes of
Dan, by the generations and families, and houſes of their kinredes, were
reckened by the names of euerie one from twentie yeares and vpward, al
that could goe forth to warre, \V ſixtie two thouſand ſeuen hundred. \V
Of the ſonnes of Aſer, by the generations and families and houſes of
their kinredes, were reckened by the names of euerie
%%% o-0309
one from the twentith yeare and vpward, al that could goe forth to
warre, \V fourtie thouſand and a thouſand fiue hundred. \V Of the ſonnes
of Nephthali, by the generations, and families, and houſes of their
kinredes were reckened by the names of euerie one from the twentith
yeare and vpward, al that could goe forth to warre, \V fiftie three
thouſand foure hundred. \V Theſe are they, whom Moyſes and Aaron
numbered, and the twelue princes of Iſrael, euerie one by the houſes of
their kinredes. \V And the whole number of the children of Iſrael by
their houſes and families, from the twentith yeare and vpward, that
could goe to warre, \V were
\SNote{Coming into Ægypt they were but 70.
\XRef{Exo.~1.}
increaſed in 216.~yeares vnto 603550. not counting the tribe of Leui,
nor wemen, nor anie vnder 20.~yeares, nor old men vnable to goe to
warre.}
ſix hundred three thouſand men fiue hundred fiftie. \V But the Leuites
in the tribe of their families were not numbered with them. \V And our
Lord ſpake to Moyſes, ſaying: \V Number not the tribe of Leui, neither
shalt thou put the ſumme of them with the children of Iſrael: \V but
appoint them ouer the tabernacle of teſtimonie, and al the veſſel
therof, and whatſoeuer pertaineth to the ceremonies. They shal carie the
tabernacle and al the furniture therof: and they shal be in the
miniſterie, and shal pitch round about the tabernacle. \V When you are
to goe forward, the Leuites shal take downe the tabernacle: when you are
to campe, they shal ſet it vp. What
\SNote{Al other tribes were in reſpect of ſeruing about the tabernacle
called ſtrangers.
\Cite{S.~Aug. q.~3. in Num.}}
ſtranger ſoeuer cometh to it, he shal be ſlaine. \V And the children of
Iſrael shal campe euerie man by his troupes and bands and hoſte. \V
Moreouer the Leuites shal pitch their tents round about the tabernacle,
leſt there come indignation vpon the multitude of the children of Iſrael,
and they shal watch in the cuſtodies of the tabernacle of teſtimonie. \V
The children of Iſrael therfore did according to al thinges which our
Lord had commanded Moyſes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
