%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-23


\project douay-rheims


%%% 0391
%%% o-0351
\startChapter[
  title={Chapter 23}
  ]

\Summary{Balaam endeuoureth to curſe Iſrael, but God forceth him to
  bleſse them. 11.~Againe he vrgeth him to curſe them, but he ſtil
  prophecieth good things of them. 26.~Yet Balac inſiſteth willing him
  to curſe, or not to bleſſe them.}

And Balaam ſaid to Balac:
\SNote{Before the diuel would curſe Gods people he required ſacrifices
which being offered yet he could not worke his malice. For by Gods
commandmẽt (ſaith
\Cite{S.~Hierom. de manſ.~42.})
Iſrael is bleſſed, curſing turned into praiſe and Gods voice is heard
ſounding from a profaine mouth.}
Build me here ſeuen altares, and prepare as manie calues, and rammes of
the ſame number. \V And when he had done according to the word of
Balaam, they together laid a calfe and a ramme vpon an altar. \V And
Balaam ſaid to Balac: Stand a while beſide thy holocauſte, vntil I goe,
if happely the Lord wil meete me, and whatſoeuer he shal command, I wil
ſpeake to thee. \V And when he was gone in haſt, God mette him. And
Balaam ſpeaking to him, ſaid: I haue erected ſeuen altares, and haue
laid thereon a calfe and a ramme. \V And our Lord put a word in his
mouth, and ſaid: Returne to Balac, and thus shalt thou ſpeake. \V
Returning he found Balac ſtanding beſide his holocauſte, and al the
princes of the Moabites: \V and taking vp his parable, he ſaid: From
Aram hath Balac King of the Moabites brought me, from the mountaines of
the Eaſt: Come, quoth he, and curſe Iacob: make haſt, and deteſt
Iſrael. \V How shal I curſe, whom God hath not curſed? By what meanes
may I deteſt, whom the Lord deteſteth not? \V From the higheſt flintes
shal I ſee him, and from the hilles shal I view him. The people shal
dwel alone, and among the Gentiles shal not be accounted. \V Who may be
able to number the duſt of Iacob, and to know the number of the ſtocke
of Iſrael?
\SNote{Heretikes, & other infidels deſire ſometimes to die in ſtate of
Catholiques, though they wil not liue as they do.
\Cite{S.~Greg. li.~33. c.~27. Moral.}}
My ſoule die the death of the iuſt, and my laſt endes be made like to
them. \V And Balac ſaid to Balaam: What is this that thou doeſt? That
thou shouldeſt curſe mine enemies I called thee: and thou contrariewiſe
bleſſeſt them. \V To whom he anſwered: Can I ſpeake ought els, but that
which the Lord commandeth? \V Balac therfore ſaid: Come with me into an
other place whence thou mayeſt ſee part of Iſrael, and canſt not ſee the
whole, from thence curſe them. \V And when he had brought him into a
high place, vpon the toppe of the mountaine Phaſga, Balaam builded
ſeuen altares, & laying thereon calues and rammes, \V he ſaid to Balac:
Stand here beſide thy holocauſt, whiles
%%% o-0352
I goe to meete him. \V Whom when our Lord had mette, and had put the
word in his mouth, he ſaid:
%%% 0392
Returne to Balac and thus thou shalt ſpeake to him. \V Returning he
found him ſtanding beſide his holocauſte, and the princes of the
Moabites with him. To whom Balac ſaid: What hath the Lord ſpoken? \V But
he taking vp his parable, ſaid: Stand Balac, and harken, heare thou
ſonne of Sephor: \V God is not as man, that he may lie, nor as the ſonne
of man, that he may be changed. Hath he ſaid then, and wil he not doe?
hath he ſpoken, and wil he not fulfil? \V I was brought to bleſſe, the
bleſſing I am not able to ſtay. \V There is no Idol in Iacob, neither is
there
\TNote{liknes of idol.}
ſimulachre to be ſeene in Iſrael. The Lord his God is with him, and the
ſound of the victorie of the king in him. \V God hath brought him out
of Ægypt, whoſe ſtrength is like to the vnicorne. \V There is no
Southſaying in Iacob, nor diuination in Iſrael. In their times it ſhal
be ſaid to Iacob and Iſrael what God hath wrought. \V Behold the people
ſhal riſe vp as a lioneſſe, and as a lion shal raiſe it ſelf: It shal
not lie downe til it deuoure the pray, and drinke the bloud of the
ſlaine. \V And Balac ſaid to Balaam:
\SNote{VVhen Infidels can not draw others to falſe worſhip, or curſing,
they are content that men profeſſe no God, nor religion at al.}
neither curſe, nor bleſſe him. \V And he ſaid: Did I not tel thee, that
whatſoeuer God should command me, that would I doe? \V And Balac ſaid to
him: Come, and I wil bring thee to an other place: if happely it pleaſe
God that thence thou mayeſt curſe them. \V And when he had brought him
vpon the toppe of the mountaine Phoſgor, which looketh to the
wilderneſſe, \V Balaam ſaid to him: Build me here ſeuen altares, and
prepare as manie calues, and rammes of the ſame number. \V Balac did as
Balaam had ſaid: and he laide the calues and the rammes on euerie
altar.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
