%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Numbers
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 0384
%%% o-0345
\startChapter[
  title={Chapter 20}
  ]

\Summary{Marie the ſiſter of Moyſes dieth. 2.~The people murmure for
  lack of water, 7.~Moyſes and Aaron being commanded to draw ſome out of
  a rock, do it doutfully: 12.~and for the ſame are foretold that they
  shal die in the deſert. 14.~Not obtaining licence to paſse through
  Edom, 22.~they come into Mount Hor, where Eleazar is ordained hiegh
  Prieſt, Aaron dieth, and is mourned by the people thirtie daies.}

And the children of Iſrael, and al the multitude came into the deſert
Sin, the firſt moneth: and the people abode in Cades. And Marie died
there, and was buried in the ſame place. \V And when the people lacked
water, they came together againſt Moyſes and Aaron: \V and being turned
into ſedition, ſaid: Would God we had perished among our brethren before
our Lord. \V Why haue you brought forth the Church of our Lord into the
wilderneſſe, that both we and our cattel ſhould die? \V Why did you make
vs aſcend out of Ægypt, and haue brought vs into this exceding naughtie
place which can not be ſowed, which bringeth forth neither figge, nor
vines, nor pomegranates, moreouer alſo hath no water for to drinke? \V
And Moyſes and Aaron, the multitude being diſmiſſed, entring into the
tabernacle of couenant, fel flatte vpon the ground, and cried to our
Lord, and ſaid: Lord God heare the crie of this people, and open vnto
them thy treaſure the fountaine of liuing water, that being
ſatisfied, their murmuring may ceaſe. And the glorie of our Lord
appeared ouer them. \V And our Lord ſpake to Moyſes, ſaying: \V Take the
rodde, and aſſemble the people together, thou and Aaron thy brother, and
ſpeake to
\SNote{The rock ſignified Chriſt, the rodde his Croſſe.
\Cite{S.~Aug. q.~35. in Num.}}
the rocke before them, and it shal geue waters. And when thou haſt
brought forth water out of the rocke, al the multitude shal drinke and
their cattel. \V Moyſes therfore tooke the rodde, which was in the ſight
of our Lord, as he commanded him, \V the multitude being aſſembled
before the rocke, and he ſaid to them: Heare ye rebellious and
incredulous: Can we out of this rocke bring you forth water? \V And when
Moyſes had lifted vp his hand ſtricking the rocke,
\SNote{By this the croſſe is more euidently ſignified, which was made of
two peeces of wood.
\Cite{S.~Aug. ibidem.}}
twiſe with
%%% 0385
the rodde, there came forth great plentie of water, ſo that the people
drunke and their cattel. \V And our Lord ſaid to Moyſes and Aaron:
Becauſe
\SNote{Not doubting of Gods powre, but ſuppoſing it vnmete, that God
\Fix{ſhold}{ſhould}{possible typo, fixed in other}
ſtil worke miracles for ſo ſtubborne a people, they did not ſpeake to
the rock, as they were cõmanded, but chiding the incredulous multitude,
ſpake to them ambiguouſly, & ſo by occaſion of others ſinne, alſo
offended, and for the ſame were puniſhed. See
\XRef{Deut.~1, v.~37.}
\XRef{3, v.~16.}
&
\XRef{4, v.~21.}
\Cite{S.~Aug. q.~19.}
\Cite{Theod. q.~38. in Num.}}
you haue not beleued me, to ſanctifie me before the children of Iſrael,
you shal not bring in theſe peoples into the land, which I wil geue
them. \V This is the water of contradiction, where the children of
Iſrael quarelled againſt our Lord, and he was ſanctified in them. \V In
the meane time Moyſes ſent meſſengers from
Cades to
%%% o-0346
the King of Edom, which should ſay: Thus thy brother Iſrael biddeth vs
to ſay: Thou knoweſt al the labour, that hath taken vs, \V in what maner
our fathers went downe into Ægypt, and there we dwelt a great time, and
the Ægyptians afflicted vs, and our fathers: \V and in what maner we
cried to our Lord, and he heard vs, and ſent an Angel, that hath brought
vs out of Ægypt. Loe being preſently in the citie of Cades, which is in
thy vttermoſt borders, \V we beſech thee that we may haue licence to
paſſe through thy countrie. We wil not goe through the fieldes, nor
through the vineyardes, we wil not drinke the waters of thy welles, but
we wil goe the common high way, declining neither to the right hand, not
to the left, til we be paſt thy borders. \V To whom Edom anſwered: Thou
shalt not paſſe by me, otherwiſe I wil come armed againſt thee. \V And
the children of Iſrael ſaid: We wil goe by the beaten way: and if we and
the cattel drinke thy waters, we wil geue thee that which is iuſt: there
shal be no difficultie in the price, only let vs paſſe ſpeedely. \V But
he anſwered: Thou shalt not paſſe: And immediatly he came forth to meete
them with an infinite multitude, and a ſtrong hand, \V neither would he
condeſcend to them deſiring to grant them paſſage through his
borders. For the which cauſe Iſrael turned an other way from him. \V And
when they had remoued the campe from Cades, they came into the mountaine
Hor, which is in the borders of the land of Edom: \V Where our Lord
ſpake to Moyſes: \V Let Aaron, ſayeth he, goe to his people: for he shal
not enter to the Land, which I haue geuen the children of Iſrael, for
that he was incredulous to my mouth, at the Waters of contradiction. \V
Take Aaron and his ſonne with him, and thou shalt bring them into the
mountaine Hor. \V And when thou haſt vnueſted the father of his veſture,
thou shalt reueſt therwith Eleazar his ſonne: Aaron shal be gathered,
and die there. \V Moyſes
%%% 0386
did as our Lord had commanded: and they went vp into the mountaine Hor
before al the multitude. \V And when he had ſpoyled Aaron of his
veſtimentes, he reueſted Eleazar his ſonne with them. \V After that he
was dead in the toppe of the mountaine, he went downe with Eleazar. \V
And al the multitude ſeeing that Aaron was dead,
\SNote{Mouring the dead was long before this obſerued by tradition.
%%% !!! Illegible in Heretics, not present in Other.
%%%\XRef{??? 23.~50.}
}
mourned vpon him thirtie daies throughout al their families.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
