%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 0675
%%% o-0606
\startChapter[
  title={Chapter 12}
  ]

\Summary{Nathan the prophet by a parable induceth Dauid to condemne him
  ſelf of great ſinne, 7.~blameth and threatneth him for the
  ſame. 13.~But vpon his confeſsion denounceth remiſsion of his ſinne,
  with reſeruation of temporal punishment, 15.~the death of the
  childe. 24.~Bethſabee beareth an other ſonne, who is called
  Salomon. 26.~The citie of Rabbath is taken, and a rich crowne with
  other praye.}

Our Lord therfore ſent Nathan to Dauid: Who when he was come to him, he
ſayd vnto him: There were two men in one citie, one riche, and the other
poore. \V The rich man had sheepe, and oxen
\Fix{axceding}{exceding}{obvious typo, fixed in other}
manie. \V But the
poore man had nothing at al, beſide one litle ewe, which he had bought
and nourished, and which had growen in his houſe together with his
children, eating of his bread, and drinking
%%% 0676
of his cuppe, and ſleping in his boſome: and it was to him as a
daughter.
%%% o-0607
\V And when a certayne ſtranger was come to the riche man, he
ſparing to take of his owne ſheepe and oxen, to make a feaſt for that
ſtranger, which was come to him, tooke the poore mans ewe, and made
meates therof for the man that was come to him. \V And Dauid being
excedingly wrath with indignation againſt that man, ſayd to Nathan: Our
Lord liueth, the man that hath done this is the childe of death. \V He
ſhal render the ewe fourefold, becauſe he hath done this thing, and hath
not ſpared. \V And Nathan ſayd to Dauid: Thou art that man. Thus ſayth
our Lord the God of Iſrael: I annointed thee to be king ouer Iſrael, and
I deliuered thee from the hand of Saul, \V and gaue thee the houſe of
thy lord, and the wiues of thy lord in thy boſome, and haue geuen thee
the houſe of Iſrael and Iuda: and if theſe thinges be litle, I wil adde
farre greater thinges vnto thee. \V Why therefore haſt thou contemned
the word of the lord, that thou wouldeſt doe euil in my ſight? Vrias the
Hetheite thou haſt ſmitten with the ſword, & his wife thou haſt taken to
thy wife, and haſt ſlayne him with the ſword of the children of
Ammon. \V For which thing the ſword ſhal not depart from thy houſe
\SNote{Now & then ſome of thy ſeede ſhal be violently ſlaine: ſo were
ſlaine three of his owne ſonnes, Ammon,
\XRef{chap.~13.}
Abſalom,
\XRef{chap.~18.}
Adonias,
\XRef{3.~Reg.~2.}
ſix ſonnes of Ioſaphat, and al Iorams ſonnes ſaue one,
\XRef{2.~Paralip.~21.}
alſo Ochozias, Amaſias, Ioſias,
\XRef{2.~Par.~24.~25.~35.}
and the ſonnes of Sedechias, himſelfe hauing his eyes put out and ſo
brought into Babylon.
\XRef{4.~Reg.~25.}}
for euer, becauſe thou haſt deſpiſed me, and haſt taken the wife of
Vrias the Hetheite, to be thy wife. \V Therfore thus ſayth our Lord:
Behold, I wil rayſe vpon thee euil out of thine owne houſe, and wil take
thy wiues before thine eies, and geue them to thy neighbour, and he ſhal
ſlepe with thy wiues in the ſight of this Sunne. \V For thou haſt done
it ſecretly: but I wil doe this word in the ſight of al Iſrael, and in
the ſight of the Sunne. \V And Dauid ſayd to Nathan: I haue ſinned to
our Lord. And Nathan ſayd to Dauid: Our Lord alſo hath taken away thy
ſinne: thou ſhalt not die. \V Neuertheles, becauſe thou haſt made the
enemies of our Lord to blaſpheme, for this thing, the ſonne that is
borne to thee, dying ſhal dye. \V And Nathan returned into his
houſe. Our Lord alſo ſtroke the child, which the wife of Vrias had borne
to Dauid, and he was paſt hope. \V And Dauid beſought our Lord for the
child: and Dauid faſted a faſt, and going in aſide, lay vpon the
ground. \V And the ancientes of his houſe came, being earneſt with him,
that he would riſe from the ground: who would not, neither did he eate
meate with them. \V And it chanced the ſeuenth day
%%% 0677
that the infant died: and the ſeruantes of Dauid feared to tel him, that
the child was dead. For they ſayd: Behold when the child yet liued, we
ſpake to him, and he heard not our voice: how much more if we ſhal ſay:
The child is dead, wil he afflict him ſelf? \V When Dauid therfore ſawe
his ſeruantes muttering, he vnderſtood that the infant was dead: and he
ſayd to his ſeruantes: Is the child dead? Who anſwered him: He is
dead. \V Dauid therfore roſe from the ground; and was washed and
annoynted: and when he had changed his garment, he entered into the
houſe of our Lord: and adored, and came into his owne houſe, and he
called for bread, and he did eate. \V And his ſeruantes ſayd vnto him:
What thing is this,
%%% o-0608
that thou haſt done? for the infant, when he yet liued, thou didſt faſt
and weepe: but the child being dead, thou didſt riſe vp, and haſt eaten
bread. \V Who ſayd: For the infant, whiles he yet liued, I faſted and
wept: for I ſayd: Who knoweth if perhaps our Lord wil geue him to me,
and the infant may liue? \V But now becauſe he is dead, why do I faſt?
Shal I be able to cal him againe any more? I ſhal go to him rather: but
he ſhal not returne to me. \V And Dauid comforted Bethſabee his wife,
and going in vnto her, ſlept with her: Who bare a ſonne, and he called
his name Salomon, and our Lord loued him. \V And he ſent by the hand of
Nathan the prophete, and called his name, Amiable to our Lord, becauſe
our Lord loued him. \V Ioab therfore fought againſt Rabbath of the
children of Ammon, and wonne the kinges citie. \V And Ioab ſent
meſſengers to Dauid, ſaying: I haue fought againſt Rabbath, and the
Citie of waters is to be taken. \V Now therfore gather the reſt of the
people, and beſiege the citie, & take it: leſt when the citie ſhal be
waſted of me, the victorie be aſcribed to my name. \V Dauid therfore
gathered al the people, and went forth againſt Rabbath: and when he had
fought, he tooke it. \V And he tooke the crowne of their king from his
head, in weight a talent of gold, hauing moſt pretious ſtones, and it was
put vpon Dauids head. Yea & the praye of the citie he caryed away
exceding much: \V bringing forth alſo the people therof ſawed them, and
drewe round about ouer them chariotes ſhod with yron: and he diuided
them with kniues, and drew them through in forme of brikes: ſo did he
to al the cities of the children of Ammon: and Dauid returned, and al
the armie into Ieruſalem.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
