%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 0671
%%% o-0602
\startChapter[
  title={Chapter 9}
  ]

\Summary{Miphiboſeth a lame ſonne of Ionathas is piouſly releeued by
  Dauid: 9.~geuing to his vſe the particular inheritance of Saul.}

And Dauid ſayd: Is there any thinke you that is remaining of the houſe
of Saul, that I may do mercie with him for Ionathas ſake? \V And there
was of the houſe of Saul, a ſeruant named Siba: whom when the king had
called vnto him, he ſaid to him: Art thou Siba? And he anſwered: I am ſo
thy ſeruant. \V And the king ſaid: Is there anie remaining of the houſe
of Saul, that I may doe with him the mercie of
%%% o-0603
God? And Siba ſaid to the king: There is yet liuing a ſonne of Ionathas,
lame of his feete. \V Where is he? quoth he. And Siba ſayd to the king:
Behold he is in the houſe of Machir the ſonne of Ammiel in Lodabar. \V
King Dauid therefore ſent, and tooke him out of the houſe of Machir the
ſonne of Ammiel of Lodabar. \V And when Miphiboſeth the ſonne of
Ionathas the ſonne of Saul was come to Dauid, he fel on his face, and
adored. And Dauid ſaid: Miphiboſeth? Who anſwered: Here I am thy
ſeruant. \V And Dauid ſaid to him: Feare not, becauſe doing I wil do
mercie on thee for Ionathas thy father, & I wil reſtore
\SNote{The particular inheretãce that pertained to Sauls familie.}
the landes of Saul thy father, and thou ſhalt eate bread vpon my table
alwaies. \V Who adoring him, ſaid: Who am I thy ſeruant, that thou haſt
reſpect vpon a dead dogge like vnto me? \V The king therefore called
Siba the ſeruant of Saul, and ſaid to him: Al thinges whatſoeuer were
Sauls, and al his houſe, I haue geuen to thy maſters ſonne. \V Til for
him therefore the land, thou and thy ſonnes, and thy ſeruants: and thou
ſhalt bring in meates for thy maſters ſonne, that he may be maintained:
and Miphiboſeth the ſonne of thy lord ſhal eate alwaies bread
\SNote{Not ſitte at table with the king, but haue his diet of the kinges
prouiſion, beſides the forſaide inheritance.}
vpon
%%% 0672
my table. And Siba had fiftene ſonnes, and twentie ſeruants. \V And Siba
ſaid to the king: As thou my lord king haſt commanded thy ſeruant, ſo
wil thy ſeruant doe: and Miphiboſeth ſhal eate vpon my table, as one of
the ſonnes of the king. \V And Miphiboſeth had a litle ſonne called
Micha: and al the kinred of the houſe of Siba ſerued Miphiboſeth. \V
Moreouer Miphiboſeth dwelt in Ieruſalem: becauſe he did eate alwaies of
the kings table: and he was lame on both feete.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
