%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 24
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-24


\project douay-rheims


%%% 0704
%%% o-0632
\startChapter[
  title={Chapter 24}
  ]

\Summary{For Dauids ſinne in numbring the people, 11.~three ſortes of
  punishments are propoſed to his election: 14.~of which he chooſeth the
  plague, and ſeuentie thouſand die in three dayes. 16.~God sheweth
  mercie. 17.~Dauid prayeth, 18.~buildeth an altar, 25.~and the plague
  ceaſeth.}

And
\SNote{This ſinne & puniſhment happened before, when Dauid had health
and ſtreingth of bodie.}
\SNote{The furie of our Lord, that is, Satan (a furious ſpirite, yet
Gods creature) not our Lord him ſelfe, but by permiſſion only.
\XRef{1.~Par.~2,~1.}
\Emph{Satan aroſe againſt Iſrael, & moued Dauid.}}
the furie of our Lord added to be angrie agaynſt Iſrael, and ſtirred vp
Dauid among them ſaying: Goe, number Iſrael and Iuda. \V And the king
ſayd to Ioab the General of his armie: Walke through al the tribes of
Iſrael from Dan to Berſabee, and number ye the people, that I may know
the number therof. \V And Ioab ſayd to the king: Our Lord thy God
increaſe thy people, as much more as now it is, and agayne multiplie it
an hundred fold in the ſight of my lord the king: but what meaneth my
lord the king by this kind of thing? \V Howbeit the kinges word more
preuailed then the wordes of Ioab, and of the chiefe of the armie: and
Ioab went forth, and the captaynes of the ſouldiars from the face of the
king, to number the people of Iſrael. \V And when they had paſſed
Iordan, they came into Aroer to the right hand of the citie, which is in
the Vale of Gad. \V And by Iazer they paſſed into Galaad, and into the
lower countrie of Hodſi, and they came into the wooddie countrie of
Dan. And going about neere Sidon, \V they paſſed nigh to the walles of
Tyre, and al the land of the Heueite, and the Chananeite, and they came
to the ſouth of Iuda into Berſabee: \V and hauing viewed the whole land,
after nine monethes and twentie dayes, they were come to Ieruſalem. \V
Ioab therfore gaue the number of the deſcription of the people to the
king, and
%%% 0705
there were found of Iſrael eight hundred thouſand ſtrong men, that could
drawe ſword: and of Iuda fiue hundred thouſand fighting men. \V But
Dauids
\SNote{Contrition.}
hart ſtrooke him, after the people was numbred: and Dauid ſayd to our
Lord:
\SNote{Confeſſion.}
I haue ſinned very much in this fact: but I pray thee Lord to transferre
the iniquitie of thy ſeruant, becauſe I haue done exceding foliſhly. \V
Dauid therfore aroſe in the morning, and the word of our Lord was made to
Gad the prophete and Seer of Dauid, ſaying: \V Goe, and ſpeake to Dauid:
Thus ſayth our Lord:
\SNote{Satisfaction.}
Choyſe is geuen thee of three thinges, chooſe one of them which thou
wilt, that I may do it to thee. \V And when Gad was come to Dauid, he
told him, ſaying: Either famine ſhal come to thee ſeuen yeares in thy
land: or three monethes thou ſhalt flee thy aduerſaries, and they ſhal
purſew thee: or certes three dayes the peſtilence ſhal be in thy
land. Now therfore deliberate, and ſee what word I
%%% o-0633
shal anſwer to him that ſent me. \V And Dauid ſayd to Gad: I am
diſtreſſed excedingly: but it is better that I fal into the handes of
our Lord (for his mercies be manie) then into the handes of men. \V And
our Lord ſent the peſtilence in Iſrael, from morning vnto the time
appoynted, and there died of the people from Dan to Berſabee ſeuentie
thouſand men. \V And when the Angel of Our Lord had ſtretched forth his
hand ouer Ieruſalem to deſtroy it, our Lord had pitie vpon the
affliction, and ſayd to the Angel that ſtroke the people:
\SNote{Temporal puniſhment inflicted after the guilt of ſinne was
remitted.}
It is ſufficient: now hold thy hand: and the Angel of our Lord was
beſide the floore of Areuna the Iebuſeite. \V And Dauid ſayd to our Lord
when he ſaw the Angel ſtriking the people: I am he that haue ſinned,
I haue done wickedly: theſe that are the ſheepe, what haue they done?
let thy hand, I beſeche thee be turned agaynſt me,
and agaynſt my fathers houſe. \V And Gad came to Dauid in that day, and
ſayd to him: Goe vp, and build an altar to our Lord in the floore of
Areuna the Iebuſeite. \V And Dauid went vp according to the word of Gad,
which our Lord had cõmanded him. \V And Areuna looking, perceiued the
king and his ſeruantes to come towards him. \V And going forth he adored
the king with his face bowing to the earth, and ſayd: What is the cauſe
that my lord the king cometh to his ſeruant? To whom Dauid ſayd: That I
may bye of thee the floore, and build an altar to our Lord, and the
ſlaughter may ceaſe which rageth among the people.
%%% 0706
\V And Areuna ſayd to Dauid: Let my lord the king take, and offer, as it
pleaſeth him: thou haſt the oxen for holocauſte, and the wayne, and the
yokes of the oxen for prouiſion of wood. \V Areuna gaue al thinges to
the king: and Areuna ſayd to the king: The Lord thy God receiue thy
vowe. \V To whom the king anſwering, ſayd: Not ſo as thou wilt, but I
wil bye it of thee at a price, and I wil not offer to our Lord my God
holocauſtes
\SNote{If ſubiectes had not proprietie in their goodes, but that the
right and dominion of al perteyned to the prince, then could nothing at
al, in anie caſe be geuen gratis by the ſubiect, but only yelded as due,
to his ſouereigne.}
geuen gratis. Dauid therfore bought the floore, and the oxen, for fiftie
ſicles of ſiluer: \V and Dauid built there an altar to our Lord, &
offered holocauſtes and pacifiques: and our Lord became merciful to the
land, and the plague was ſtayed from Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
