%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 0697
%%% o-0626
\startChapter[
  title={Chapter 21}
  ]

\Summary{Famine oppreſsing Iſrael three yeares, for the ſinne of Saul
  agaynſt the Gabaonites, 6.~ſeuen of Sauls race (7.~Miphiboſeth ſaued)
  are crucified. 12.~Their bones with Sauls and Ionathas are buried in
  the Land of Beniamin. 15.~Dauid hath foure great battels and victories
  againſt the Philiſtians.}

And there came a famine in the daies of Dauid three yeares continually:
and Dauid conſulted the oracle of our Lord. And our Lord ſayd: For Saul,
and his bloudy houſe, becauſe he ſlewe the Gabaonites. \V The king
therfore calling the Gabaonites, ſayd to them. (Moreouer the Gabaonites
were not of the children of Iſrael, but the reliques of the Amorrheites:
\CNote{\XRef{Ioſue.~9.}}
For the children of Iſrael had ſworne to them, and Saul would ſtrike
them of zeale, as it were for the children of Iſrael and Iuda.) \V Dauid
therfore ſayd to the Gabaonites: What shal I do for you? And what shal
be the expiation for you, that you may bleſſe the inheritance of our
Lord? \V And the Gabaonites ſayd to him: We haue no
%%% 0698
queſtion vpon ſiluer and gold, but agaynſt Saul, and agaynſt his houſe:
neither wil we that a man be ſlayne of Iſrael. To whom the king ſayd:
What wil you then that I do for you? \V Who ſayd to the king: The man,
that hath waſted vs and oppreſſed vs vniuſtly, we muſt ſo deſtroy, that
there be not ſo much as one leift of his ſtocke in al the coaſtes of
Iſrael. \V Let there be geuen vs ſeuen men of his children, that we may
crucifie them to our Lord in Gabaa of Saul, once the choſen of our
Lord. And the king ſayd: I wil geue them. \V And the king ſpared
Miphiboſeth the ſonne of Ionathas the ſonne of Saul, for the oth of our
Lord, that had beene betwen Dauid, and betwen Ionathas the ſonne of
Saul. \V The king therfore tooke the two ſonnes of Reſpha the daughter
of Aia, whom ſhe bare to Saul, Armoni, and Miphiboſeth: and the fiue
ſonnes of Michol the daughter of Saul, which ſhe bare to Hadriel the
ſonne of Berzellai, that was of Molathi, \V and gaue them into the
handes of the Gabaonites: who crucified them on a hil before our Lord:
and theſe ſeuen dyed together in the firſt dayes of harueſt, when the
reaping of barley began. \V And Reſpha the daughter of Aia taking a
heare cloth, ſpred it vnder her vpon the rocke from the beginning of
harueſt, til water dropped vpon them from heauen: and she ſuffered not
the birdes to teare them by day, nor the beaſtes by night. \V And the
thinges were told Dauid, which Reſpha had done, the daughter of Aia, the
concubine of Saul. \V And Dauid went, and tooke the bones of Saul, and
the bones of Ionathas his ſonne from the men of Iabes Galaad, who
%%% o-0627
had ſtolen them out of the ſtreate of Bethſan, in the which the
Philiſtijms hanged them when they had killed Saul in Gelboe. \V And he
caried thence the bones of Saul, and the bones of Ionathas his ſonne:
and gathering the bones of them, that were crucified, \V they buried
them with the bones of Saul, and the bones of Ionathas his ſonne in the
Land of Beniamin, in the ſide, in the ſepulchre of Cis his father: and
they did al thinges that the king had commanded, and God was made
propitious agayne to the land after theſe thinges. \V And there was a
battel made agayne of the Philiſtianes againſt Iſrael, and Dauid went
downe, and his ſeruantes with him, and fought agaynſt the
Philiſthijms. And Dauid faynting, \V Ieſbibenob, which was of the kinred
of Arapha, the yron of whoſe ſpeare weyed three hundred
%%% 0699
ounces, and he was girded with a new ſword, aſſayed to ſtrike Dauid. \V
And Abiſai the ſonne of Saruia reskewed him, and ſtriking the Philiſtian
killed him. Then ſware Dauids men, ſaying: Thou shalt no more goe forth
with vs into battel, leſt thou put out the lampe of Iſrael. \V There was
alſo a ſecond battel in Gob againſt the Philiſtians: then ſtroke
Sobochai of Huſathi, Saph of the ſtocke of Arapha of the kindred of the
gyantes. \V There was alſo a third battel in Gob agaynſt the
Philiſtians, in the which Adeodatus the ſonne of the Foreſt a broderer
the Bethlehemite ſtroke Goliath the Getheite, the shaft of whoſe ſpeare
was as it were a weauers beame. \V The fourth battel was in Geth: in the
which was a tal man, that had ſix fingers and ſix toes on eche hand and
foote, that is fowre and twentie, and he was of the race of Arapha. \V
And he blaſphemed Iſrael: and Ionathan the ſonne of Samaa the brother of
Dauid ſtroke him. \V Theſe foure were borne of Arapha in Geth, and they
fel by the hand of Dauid, and of his ſeruantes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
