%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 0673
%%% o-0605
\startChapter[
  title={Chapter 11}
  ]

\Summary{Dauid ouercome with concupiſcence committeth adulterie with
  Bethſabee: 6.~not finding other meanes to hide the crime, cauſeth her
  husband Vrias to be ſlaine. 27.~Then marieth her, she beareth a ſonne,
  and God is offended.}

And it came to paſſe the yeare turning about, at ſuch time when kinges
are wont to procede to battels, Dauid ſent Ioab, and his ſeruantes with
him, and al Iſrael, and they ſpoyled the children of Ammon, and beſieged
Rabba: but Dauid remayned in Ieruſalem. \V Whiles theſe thinges were
in doing, it chanced that Dauid aroſe from his bed after
%%% 0674
noone, and walked in the toppe of the kinges houſe: and he ſaw a woman
washing her ſelf, ouer againſt the roofe of his houſe: and the woman was
very beautiful. \V The king therefore ſent, and inquired what woman it
was. And it was told him, that she was Bethſabee the daughter of Eliam,
the wife of Vrias the Hetheite. \V Dauid therefore ſending meſſengers,
\SNote{Theodoſius the Emperour pretending to be excuſed from puniſhment
for his ſinnes, becauſe king Dauid alſo was an adulterer and a
manſlayer, S.~Ambroſe replied, ſaying: \Emph{Thou haſt folovved king Dauid
erring, folovv him repenting.} After which admonition the Emperour moſt
humbly did publique penance inioyned him by the Biſhop.
\Cite{in vita. Theodoſij.}}
tooke her, who when she was entered in to him, he ſlept with her: and
forthwith she was ſanctified from her vnclennes: \V and she returned
into her houſe hauing conceiued a childe. And ſending she told Dauid,
and ſayd: I haue conceiued. \V And Dauid ſent to Ioab, ſaying: Send me
Vrias the Hetheite. And Ioab ſent Vrias to Dauid. \V And Vrias came to
Dauid. And Dauid asked how wel Ioab did, & the people, and how the warre
was ordered. \V And Dauid ſayd to Vrias: Goe into thy houſe, and wash
thy feete. And Vrias went forth out of the kinges houſe, and the kinges
meate folowed him. \V But Vrias ſlept before the gate of the kinges
houſe, with the other ſeruantes of his lord, and went not downe to his
owne houſe. \V And it was told Dauid of them that ſayd: Vrias went not
into his houſe. And Dauid ſayd to Vrias: didſt thou not come from thy
iourney? Why didſt thou not goe downe into thy houſe? \V And Vrias ſayd
to Dauid: The Arke of God and Iſrael and Iuda dwel in pauilions, & my
lord Ioab and the ſeruantes of my lord abide vpon the face of the earth:
and shal I enter into my houſe, to eate and to drinke, and ſleepe with
my wife? by thy health, and by the health of thy ſoule I wil not do this
thing. \V Dauid therefore ſayd to Vrias: Tarie here alſo this day, and
to morow I wil diſmiſſe thee. Vrias taried in Ieruſalem, that day and the
next: \V and Dauid called him to eate before him and to drinke, and he
made him drunke: who going out at euen, ſlept on his couche with the
ſeruantes of his lord, and went not downe into his houſe. \V The morning
therefore was come, and Dauid wrote a letter to Ioab: and ſent it by the
hand of Vrias, \V writing in the letter: Sette ye Vrias in the front of
the battel, where the fight is ſtrongeſt: and leaue him, that being
ſtriken he may die. \V Therefore
%%% o-0606
when Ioab beſieged the citie, he put Vrias in the place where he knew
the ſtrongeſt men were. \V And the men iſſuing out of the citie, fought
againſt Ioab, and there fel of the people of the ſeruantes
of Dauid, and Vrias alſo the Hetheite died. \V Ioab therefore ſent, and
told Dauid
%%% 0675
al the ſtory of the battel: \V and he commanded the meſſenger, ſaying:
When thou haſt told al the ſtory of the battel to the king, \V if thou
ſee him to be angrie, and he ſay: Why approched you to the wal, to
fight? Knew you not that manie weapons are throwen from aboue of the
wal? \V Who ſtroke Abimelec the ſonne of Ierobaal? did not a woman caſt
vpon him a peece of a milſtone from the wal, and ſlew him in Thebes? Why
approched you nere the wal? Thou ſhalt ſay: Alſo thy ſeruant Vrias the
Hetheite is ſlayne. \V The meſſenger therefore departed, and came, and
told Dauid al thinges that Ioab had commanded him. \V And the meſſenger
ſayd to Dauid: The men haue preuailed againſt vs, and they iſſued forth
to vs into the field: and we violently purſewed them euen to the gate of
the citie. \V And the archers ſhot arrowes at thy ſeruantes from of the
wal aboue: and there died of the kinges ſeruantes, yea and thy ſeruant
Vrias the Hetheite is dead. \V And Dauid ſayd to the meſſenger: Thus
ſhalt thou ſay to Ioab: Let not this thing diſcomfort thee: for the euent
of warre is diuerſe: now this man, and now that man the ſword conſumeth:
encourage thy warryers againſt the citie, that thou mayeſt deſtroy it,
and exhort them. \V Alſo the wife of Vrias heard, that Vrias her huſband
was dead, & she mourned for him. \V And the mourning being paſt Dauid
ſent, and brought her in into his houſe, and ſhe became his wife, and
she bare him a ſonne: and this thing which Dauid had done, was
diſpleaſant before our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
