%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 0658
%%% o-0592
\startChapter[
  title={Chapter 2}
  ]

\Summary{Dauid is receiued and annointed King of Iuda. 5.~He commendeth
  thoſe of Iabes Galaad, which buried Saul. 8.~Iſboſeth the ſonne of
  Saul reigneth ouer the reſt of Iſrael. 12.~Whereby riſeth sharpe warre
  betwen Abner and Ioab, chief captaines of the two kinges. 30.~Manie
  more are ſlaine of Abners partie then of Ioabs.}

%%% 0659
Therefore after theſe thinges Dauid conſulted our Lord, ſaying: Shal I
goe vp into one of the cities of Iuda? And our Lord ſaid to him: Goe
vp. And Dauid ſaid: Whither ſhal I goe vp? And he anſwered him: Into
Hebron. \V Dauid therefore went vp, and his two wiues, Achinoam the
Iezrahelite, and Abigail the wife of Nabal of Carmel: \V yea and the men
alſo that were with him, Dauid brought euery one with his houshould: and
they abode in the townes of Hebron. \V And the men of Iuda came, and
\SNote{This ſecond annointing, as alſo the third,
\XRef{(ch.~5.)}
was in confirmation and to put him in poſſeſſion of the firſt made long
before.
\XRef{1.~Reg.~16.}}
annoynted Dauid there, to reigne ouer the houſe of Iuda. And it was told
Dauid, that the men of Iabes Galaad had buried Saul. \V Dauid therefore
ſent meſſengers vnto the men of Iabes Galaad, and ſayd vnto them: Bleſſed
be you to our Lord, which haue done this mercie with your lord Saul, and
haue buried him. \V And now our Lord certes wil render you mercie and
truth: but I alſo wil requite you the good turne, for that you haue done
this thing. \V Let your handes be ſtrengthned, and be yee ſtout men: for
although your lord Saul be dead, yet the houſe of Iuda hath annointed me
to be their king. \V But Abner the ſonne of Ner prince of Sauls armie,
tooke Iſboſeth the ſonne of Saul, & led him about through the campe, \V
and ordained him king ouer Galaad, and ouer Geſſuri, and ouer Iezrahel,
& ouer Ephraim, and ouer Beniamin, and ouer al Iſrael. \V Fourtie yeares
old was Iſboſeth the ſonne of Saul when he began to reigne ouer Iſrael,
and he reigned
\SNote{He reigned two yeares before he beganne to decline, but in al
ſeuen yeares and a half, for ſo long Dauid reigned only in Iuda.
\XRef{v.~11.}}
two yeares: and only the houſe of Iuda folowed Dauid. \V And the number
of the daies, that Dauid abode, reigning in Hebron ouer the houſe of Iuda,
was ſeuen yeares and ſix monethes. \V And Abner the ſonne of Ner went
forth, and the ſeruantes of Iſboſeth the ſonne of Saul, out of the campe
into Gabaon. \V Moreouer Ioab the ſonne of Saruia, and the ſeruants of
Dauid went forth, and mette them beſide the poole of Gabaon, and when
they were come together into one place, they ſate one ouer againſt an
other: theſe on the one ſide of the poole, and they on the other. \V And
Abner ſaid to Ioab: Let the yong men riſe, and
\SNote{Hence perhaps cometh the phraiſe, that \Emph{one armie playeth
vpon an other vvith ſmal and great artilerie}, that is, ſtrike and kil
their enimies with al ſortes of gunnes.
\Cite{Ioſephus li.~7. c.~1. Antiq.}}
play before vs. And Ioab anſwered: Let them riſe. \V There roſe
therefore and paſſed twelue in number, of Beniamin, of Iſboſeths part
the ſonne of Saul, and twelue of the ſeruantes of
%%% o-0593
Dauid. \V And euery one taking the head of his mate, ſticked his ſword
into the ſide of his aduerſarie, and they fel together:
%%% 0660
& the name of the place was called: The field of the valiantes, in
Gabaon. \V And there roſe a very ſore battel in that day: and Abner was
put to flight, & the men of Iſrael, by the ſeruantes of Dauid. \V And
there were there the three ſonnes of Saruia, Ioab, and Abiſai, and
Aſael: moreouer Aſael was a moſt ſwift runner, as it were one of the
roes, that abide in the wooddes. \V And Aſael purſued Abner, and declined
not to the right hand nor to the left omitting to purſue Abner. \V Abner
therefore looked backe behind him, and ſaid: Art thou Aſael? Who
anſwered: I am. \V And Abner ſaid to him: Goe to the right hand, or to
the left, and apprehend one of the young men, and take to thee his
ſpoyles. But Aſael would not leaue but vrged him. \V And againe Abner
ſaid to Aſael: Retyre, and do not folowe me, leſt I be compelled to
ſticke thee to the ground, and I ſhal not be able to lift vp my face to
Ioab thy brother. \V Who contemned to heare, and would not goe aſide:
Abner therefore ſtroke him, with his ſpeare turned backe in the priuy
partes, and thruſt him through, and he dyed in the ſame place: and al
that paſſed by that place, wherein Aſael fel and dyed, did ſtay. \V But
whiles Ioab and Abiſai purſued Abner fleeing, the ſunne went downe: and
they came as farre as the hil of the water conduit, that is ouer againſt
the valley the way of the deſert in Gabaon. \V And the children of
Beniamin were aſſembled together to Abner: and being gathered in a
plumpe into one troupe, they ſtoode in the toppe of one hil. \V And
Abner cried out to Ioab, and ſaid: Shal thy ſword rage vnto vtter
deſtruction? knoweſt thou not that deſparation is dangerous? how long
differreſt thou to ſay to the people, that they leaue to purſue their
brethren? \V And Ioab ſaid: Our Lord liueth, if thou hadeſt ſpoken, in
the morning had the people retyred, from purſuing their brethren. \V
Ioab therefore ſounded the trumpette, and al the armie ſtood, neither
did they purſue Iſrael any further, nor enter into fight. \V And Abner
and his men went through the champaine country, al that night: and they
paſſed Iordan, and hauing viewed al Beth horon, came to the campe. \V
Moreouer Ioab returning, after he had leift Abner, aſſembled al the
people: and there wanted of Dauids ſeruantes nintene men, beſide
Aſael. \V But the ſeruantes of Dauid ſtroke of Beniamin, and of the men,
that were with Abner three hundred three ſcore, who alſo died. \V And
they tooke Aſael, & buried
%%% 0661
him in the ſepulchre of his father in Bethlehem: and Ioab, and the men
that were with him, walked al the night, and in the very twolight they
came into Hebron.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
