%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 0664
%%% o-0597
\startChapter[
  title={Chapter 5}
  ]

\Summary{VVith general conſent Dauid is annointed king of al
  Iſrael. 7.~He taketh the towre of Sion in Ieruſalem, deſtroying the
  Iebuſeites. 9.~Buildeth there a new houſe: 13.~marieth more wiues, and
  hath more children. 17.~The Philiſtijms riſing againſt him are
  ouerthrowen: 22.~alſo the ſecond time.}

And al the tribes of Iſrael came to Dauid in Hebron, ſaying: Behold we
are thy bone and thy fleſh. \V Yea and yeſterday alſo and the day before
when Saul was king ouer vs,
%%% 0665
thou waſt he that didſt leade vs forth and bring backe Iſrael: and our
Lord ſayd to thee: Thou ſhalt feede my people Iſrael, and thou shalt be
prince ouer Iſrael. \V The ancientes alſo of Iſrael came to the king
into Hebron, and king Dauid made a league with them in Hebron before our
Lord: and they
\SNote{They annoint him againe in confirmation of their conſent (as Iuda
had donne
\XRef{chap.~2.)}
acknowledging Gods ordinance.
\XRef{1.~Reg.~16.}}
annointed Dauid to be king ouer Iſrael. \V Thirtie yeares old was Dauid
when he began to reigne, and he reigned fourtie yeares. \V In Hebron he
reigned ouer Iuda ſeuen yeares and ſix monethes: and in Ieruſalem he
reigned three and thirtie yeares ouer al Iſrael & Iuda. \V And the king
went, & al the men that were with him, into Ieruſalem to the Iebuſeite
the inhabiter of the land: & they ſaid to Dauid: Thou shalt not come in
hither, vnleſſe thou take away the blind and the lame, ſaying: Dauid
ſhal not come in hither. \V But
\SNote{King Dauid now atcheued that the tribe of Iuda could not in the time
of Ioſue.
\XRef{Ioſue.~15.}}
Dauid tooke the towre of Sion, this is the citie of Dauid. \V For Dauid
had propoſed in that day a reward to whoſoeuer ſhould ſtrike the
Iebuſeite, and touch the gutters of the houſe toppes, and take away the
blind and the lame that hated the ſoule of Dauid: therefore it is ſayd
in the prouerbe:
\SNote{Idoles that haue eyes and can not ſee, feete and can not goe,
ſhal not enter into the Church of Chriſt.}
The blind and the lame ſhal not enter into the temple. \V And Dauid
dwelt in the towre, and called it, The citie of Dauid: & built round
about from Mello and inwards. \V And he went proſpering and growing vp,
and our Lord the God of hoſtes was with him. \V Hiram alſo the king of
Tyre ſent meſſengers to Dauid, and ceder trees, and carpenters, and
maſons for walles: and they built a houſe for Dauid. \V And Dauid knew
that our Lord had confirmed him king ouer Iſrael, and that he had
exalted his kingdome ouer his people Iſrael. \V Dauid
therefore tooke yet concubines and wiues of
%%% o-0598
Ieruſalem, after he was come from Hebron: and there were borne to Dauid
other ſonnes alſo and daughters: \V And theſe be the names of them, that
were borne to him in Ieruſalem, Samua, and Sobab, and Nathan, and
Salomon, \V and Iebahar, and Eliſua, and Nepheg, \V and Iaphia, and
Eliſama, and Elioda, and Eliphaleth. \V The Philiſtijms therefore heard
that they had annointed Dauid to be king ouer Iſrael: and they went vp
al to ſeeke Dauid: which when Dauid had heard, he went downe into a
hold. \V And the Philiſtijms coming were ſpred in the Vale Raphaim. \V
And Dauid conſulted our Lord, ſaying: Shal I goe vp to the Philiſtijms?
and wilt thou geue them into my hand? And our Lord ſayd to Dauid: Goe
vp, becauſe deliuering I wil geue the
\Fix{Philiſtjims}{Philiſtijms}{pbvious typo, fixed in other}
%%% 0666
in thy hand. \V Dauid therefore came into Baal Pharaſim: and ſtroke them
there, and ſaid: Our Lord hath diuided mine enemies before me, as waters
are diuided, therefore the name of the place was called Baal
Pharaſim. \V And they leift there their
\TNote{\L{ſculptilia.}}
grauen goddes: which Dauid, and his men tooke. \V And the Philiſtians
added yet to go vp, and ſpred them ſelues in the Vale Raphaim. \V And
Dauid conſulted our Lord: Shal I goe vp againſt the Philiſtijms, and
wilt thou deliuer them into my handes? Who anſwered: Goe not vp againſt
them, but fetch a compaſſe behind their backe, and thou ſhalt come to
them ouer againſt the peare trees. \V And when thou ſhalt heare the
ſound of one going in the toppe of the pearetrees, then ſhalt thou enter
battel: becauſe then wil our Lord goe forth before thy face, to ſtrike
the campe of the Philiſtijms. \V Dauid therefore did as our Lord had
commanded him, and he ſtroke the Philiſtijms from Gabaa, vntil thou come
to Gezer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
