%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 0680
%%% o-0611
\startChapter[
  title={Chapter 14}
  ]

\Summary{Ioab ſuborning a woman firſt to propoſe the ſuite by a parable,
  21.~obtayneth pardon for Abſalom. 24.~But ſo that he appeareth not in
  the kinges preſence. 25.~He is exceding fayre, hath three ſonnes and
  one daughter. 29.~Ioab refuſing to deale further for his free releaſe,
  Abſalom burneth his corne. 31.~Then Ioab procureth his acceſſe to the
  king.}

And Ioab the ſonne of Saruia, vnderſtanding that the kinges hart was
turned to Abſalom, \V he ſent to Thecua, and tooke thence a prudent
woman: and he ſayd to her: Feyne that thou mourneſt, and put on a
mourning garment, and be not annoynted with oyle, that thou mayſt be as
a woman now a long tyme mourning for one dead.
%%% 0681
\V And thou ſhalt goe in vnto the king, and ſhalt ſpeake to him theſe
maner of wordes. And Ioab put the wordes in her mouth. \V Therefore when
the woman of Thecua was gone in to the king, ſhe fel before him vpon the
ground, and adored and ſaid: Saue me ô king. \V And the king ſayd to
her: What matter haſt thou? Who anſwered: Alas, I am a wodow woman: for
my huſband is dead. \V And thy handmaide had two ſonnes: who fel at
wordes againſt eche other in the field, and there was none to ſtay them:
and the one ſtroke the other, and ſlew him. \V And behold the whole
kinred ryſing againſt thy handmaide, ſaith: Deliuer him that hath
ſtriken his brother that we may kil him for the life of his brother,
whom he hath ſlayne, and may cleane deſtroy the heire: and they ſeeke to
extinguiſh my ſparkle, which is leift, that there may no name remaine to
my huſband, nor reliques vpon the earth. \V And the king ſaid to the
woman: Goe into thy houſe, and I wil geue commandement for thee. \V And
the woman of Thecua ſaid to the king: Vpon me, my lord, be the
iniquitie, and vpon the houſe of my father: but be the king and his
throne innocent. \V And the king ſaid: He that ſhal gayneſay thee bring
him to me, & he ſhal adde no more to touch thee. \V Who ſayd: Let the
king remember our Lord his God, that the next of bloud be not multiplied
to reuenge, and that they kil not my ſonne. Who ſayd: Our Lord liueth,
there ſhal not fal of the heares of thy ſonne vpon the earth. \V The
woman therefore ſayd: Let thy handmayde ſpeake to my lord the king a
word. \V And he ſayd: Speake. \V And the woman ſayd: Why haſt thou
thought ſuch a thing againſt the people of God, and why hath the king
ſpoken this word, that he would ſinne, and not bring againe his baniſhed
one? \V We doe al dye, and as waters that returne not, we fal downe on
%%% o-0612
the earth: neither wil God haue a ſoule to periſh, but reuoketh,
meaning that he perish not altogether that is caſt of. \V Now therefore
I come, that I may ſpeake to my lord the king this word, the people
being preſent. And thy handmayd ſayd: I wil ſpeake to the king, if by
any meanes the king may doe the word of his handmaide. \V And the king
hath heard, to deliuer his handmaide out of the hand of al, that would
deſtroy me out of the inheritance of our Lord, and my ſonne together. \V
Let thy handmaide therefore ſay, that the word of my lord the king be
made as a ſacrifice. For euen as an
%%% 0682
Angel of God, ſo is my lord the king, that he is moued neither with
bleſſing nor curſing: Wherefore our Lord alſo thy God is with thee. \V
And the king anſwering, ſayd to the woman: Hide not from me the thing
that I aske thee. And the woman ſayd to him: Speake my lord king. \V And
the king ſayd: Is the hand of Ioab with thee in al theſe thinges? The
woman anſwered, and ſayd: By the health of thy ſoule, my lord king, it
is neither on the left hand, nor on the right of al theſe thinges, which
my lord the king hath ſpoken: for thy ſeruant Ioab, he commanded me, and
he put al theſe wordes into the mouth of thy handmayde. \V That I ſhould
change the forme of this ſpeach, thy ſeruant Ioab commanded this: and
thou my lord king, art wiſe, as an Angel of God hath wiſdom, that thou
vnderſtandeſt al thinges vpon the earth. \V And the king ſayd to Ioab:
Behold I being pacified haue done thy word: Goe therefore, and cal
agayne the boy Abſalom. \V And Ioab falling vpon his face vnto the
earth, adored, and
\SNote{Praiſed and thanked the king.}
bleſſed the king: and Ioab ſayd: This day thy ſeruant hath vnderſtood,
that I haue found grace in thy ſight my lord king: for thou haſt done
the word of thy ſeruant. \V Ioab therefore aroſe and went into Geſſur,
and brought Abſalom into Ieruſalem. \V But the king ſayd: Let him
returne into his houſe, and not ſee my face. Abſalom therefore returned into his
houſe, and the kings face he ſaw not. \V Moreouer like
as Abſalom, there was not a man in al Iſrael ſo beautiful, and exceding
comelie: from the ſole of the foote to the crowne there was no blemish
in him. \V And when he powled his heare (once a yeare he was powled,
becauſe his bush did burden him) he weighed the heare of his head at two
hundred ſicles, of the common weight. \V And there was borne to Abſalom
\SNote{Theſe children died before him as appeareth.
\XRef{cha.~18.}}
three ſonnes: and one daughter, named Thamar, of a goodly beautie. \V
And Abſalom abode in Ieruſalem two yeares, and ſaw not the kinges
face. \V He therefore ſent to Ioab, to ſend him to the king: who would
not come to him. And when he had ſent the ſecond time, and he would not
come to him, \V he ſayd to his ſeruantes: You know the filde of Ioab
beſide my filde, that hath barley harueſt: goe therefore and burne it
with fyre. The ſeruantes therefore of Abſalom burnt the corne with
fyre. And Ioabs ſeruantes coming, renting their garmentes, ſayd: The
ſeruantes of Abſalom haue burnt part of the filde with fyre. \V And Ioab
aroſe, & came to Abſalom
%%% 0683
in his houſe, & ſayd: Why haue thy ſeruantes burnt my corne with
fire? \V And Abſalom anſwered Ioab: I ſent to thee
%%% o-0613
beſeching thee that thou wouldeſt come vnto me, and I might ſend thee to
the king, and thou ſhouldeſt ſay to him: Wherefore came I out of Geſſur?
It was better for me to be there: I beſech thee therefore that I may ſee
the face of the king: & if he be mindeful of mine iniquitie, let him kil
me. \V Ioab therefore entring in to the king, told him al thinges: and
Abſalom was called, and he entered in to the king: and adored vpon the
face of the earth before him: and the king kiſſed Abſalom.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
