%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 0692
%%% o-0621
\startChapter[
  title={Chapter 19}
  ]

\Summary{Dauid moued by Ioabs admonition, 8.~ceaſeth mourning for
  Abſalom, and reconcileth the rebels. 19.~Semei is
  pardoned. 24.~Miphiboſeth clereth him ſelf of his ſeruantes falſe
  accuſation, yet recouereth not his whole right. 32.~Berzellai is
  courteouſly intreated. 40.~The other tribes contend with Iuda for
  their affection to the king.}

And it was told Ioab, that the king wept, and mourned for his ſonne: \V
And the victorie was turned into mourning that day to al the people: for
the people heard it ſayd in that day: The king ſoroweth vpon his
ſonne. \V And the people shunned that day to enter into the citie, as a
people turned, & fleing out of battel is wont to shrinke aſide. \V
Moreouer the king couered his head, and cried with a lowd voice: O my
ſonne Abſalom, ô Abſalõ my ſonne, ô my ſonne. \V Ioab therfore entring
in to the king, into his houſe, ſayd: Thou haſt cõfounded this day the
countenances of al thy ſeruantes, that haue ſaued thy life, and the life
of thy ſonnes, and thy daughters, and the life of thy wiues, and the
life of thy concubines. \V Thou loueſt them that hate thee, and thou
hateſt them that loue thee: and thou haſt shewed this day that thou
careſt not for thy nobles, and for thy ſeruantes: and in deede I knowe
now, that if Abſalom liued, and al we had beene ſlayne, then it would
pleaſe thee. \V Now therfore ariſe, and come forth, and ſpeaking vnto
them ſatisfie thy ſeruantes: for I ſweare to thee by our Lord, that if
thou wilt not goe forth, not one verely wil remayne with thee
this night: and this shal be worſe for thee, then al the euils, which
haue come vpon thee from thy youth vntil this preſent. \V The king
%%% 0693
therfore aroſe and ſate in the gate: and it was told al the people that
the king ſate in the gate: and al the multitude came forth before the
king, but Iſrael fled into their tabernacles. \V Al the people alſo
ſtroue in al the tribes of Iſrael, ſaying: The king hath deliuered vs
out of the hand of our enemies, he hath ſaued vs from the head of the
Philiſtianes: and now he fled out of the land for Abſalom. \V But
Abſalom whom we annoynted ouer vs, is dead in the battel: how long are
you ſtil, and reduce not the king? \V But king Dauid ſent to Sadoc, and
Abiathar the prieſtes, ſaying: Speake to the Ancientes of Iuda, ſaying:
Why come you laſt to bring
%%% o-0622
backe the king into his houſe? (And the ſaying of al Iſrael was come to
the king in his houſe.) \V You are my brethren, you my bone, and my
flesh, why do you laſt bring backe the king? \V And ſay ye to Amaſa: Art
not thou my bone, and my flesh? Theſe thinges do God to me, and theſe
adde he, if thou be not the chiefe captayne of warfare before me alwayes
for Ioab. \V And he inclined the hart of al the men of Iuda, as it were
of one man: and they ſent to the king, ſaying: Returne thou, and al thy
ſeruantes. \V And the king returned, and came as far as Iordan, and al
Iuda came as far as Galgal to meete the king, and to bring him ouer
Iordan. \V And Semei the ſonne of Gera the ſonne of Iemini of Bahurim
made haſt, and went downe with the men of Iuda to meete king Dauid \V
with a thouſand men of Benjamin, and Siba the ſeruant of the houſe of
Saul: and his fiftene ſonnes, and twentie ſeruantes were with him: and
rushing into Iordan, \V paſſed the fordes before the king, that they
might helpe ouer the kinges houshould, and doe according to his
commandement. And Semei the ſonne of Gera proſtrate before the king,
when he had now paſſed Iordan, \V ſayd to him: Impute not to me my lord
the iniquitie, nor remember the iniuries of thy ſeruant in the day that
thou my lord king wenteſt out of Ieruſalem, nor put it in thy hart ô
king. \V For I thy ſeruant acknowledge my ſinne: and therefore this day
I am firſt come of al
\SNote{Al the eleuen tribes are called by the name of Ioſeph, being
chiefe after Iuda. So Semei not of the proper tribe of Ioſeph but of
Beniamin, pleading for pardon of his former fault, alleageth that he
came firſt of the eleuen tribes, to ſubmitte him ſelf and ſerue the
king.}
the houſe of Ioſeph, and am deſcended to meete my lord the king. \V But
Abiſai the ſonne of Saruia anſwering, ſayd: What shal Semei for theſe
wordes not be ſlayne, becauſe he reuiled the annoynted of our Lord? \V
And Dauid ſayd: What is to me and you ye ſonnes of Saruia? Why are you
made this day as ſatan to me? Shal there a man be
%%% 0694
killed in Iſrael to day? Doe I not know that this day I am made king
ouer Iſrael? \V And the king ſayd to Semei: Thou shalt not die. And he
ſware to him. \V Miphiboſeth alſo the ſonne of Saul came downe to meete
the king, his feete vnwashed, and his beard not pouled: and he had not
washed his garmentes from the day that the king went forth, vntil the
day of his returne in peace. \V And when he had mette the king at
Ieruſalem, the king ſayd to him: Why cameſt thou not with me
Miphiboſeth? \V And he anſwering, ſayd: My lord king, my ſeruant
contemned me: and I thy ſeruant ſpake to him that he should ſadle me an
aſſe, that getting on I might goe with the king: for I thy ſeruant am
lame. \V Moreouer he hath alſo accuſed me thy ſeruant to thee my lord
king: but thou my lord king art as an Angel of God, doe what pleaſeth
thee. \V For neither was my fathers houſe ought els, but guiltie of
death to my lord king: and thou haſt put me thy ſeruant among the
gueſtes of thy table? What iuſt complaynt therfore haue I? or what can I
further crie out to the king? \V The king therfore ſayd to him: What
ſpeakeſt thou any more? That is determined which I haue ſpoken: Thou,
and Siba diuide the poſſeſſions. \V And Miphiboſeth anſwered
\Fix{the the}{the}{obvious type, fixed in other}
king: Yea let him take al, for ſo much as my lord
%%% o-0623
king is returned peacably into his houſe. \V Berzellai alſo the
Galaadite, coming downe from Rogelim, brought the king ouer Iordan,
being readie alſo to attend on him beyond the riuer. \V And Berzellai
the Galaadite was verie old, that is to ſay, of foure ſcore yeares, and
he gaue the king victuals, when he abode in the Fild: for he was an
exceding rich man. \V The king  therfore ſayd to Berzellai: Come with
me, that thou mayſt reſt ſecure with me in Ieruſalem. \V And Berzellai
ſayd to the king: How manie are the daies of the yeares of my life, that
I ſhould goe vp with the king into Ieruſalem? \V I am this day foure
ſcore yeares old, are my ſenſes quicke to diſcerne ſweete, or ſowre? or
can meate or drinke delight thy ſeruant? or can I heare more the voyce
of ſinging men and ſinging wemen? Why should thy ſeruant be a burden to
my lord the king? \V I thy ſeruant wil goe forward a litle from Iordan
with thee: I neede not this recompenſe, \V but I beſech thee that I thy
ſeruant may returne, and die in my citie, and be buried by the ſepulchre
of my father, and my mother. But there is thy ſeruant Chamaam, let him
goe with thee, my lord
%%% 0695
king, and doe to him whatſoeuer ſemeth good to thee. \V The king
therfore ſayd to him: Let Chamaam paſſe on with me, and I wil doe for
him whatſoeuer ſhal pleaſe thee, and al, that thou shalt aske of me, thou
ſhalt obtayne. \V And when al the people and the king had paſſed Iordan,
the king kiſſed Berzellai, and bleſſed him: and he returned into his
place. \V The king therfore paſſed into Galgal, and Chamaam with him,
and al the people of Iuda had brought ouer the king, and the halfe part
onlie of the people of Iſrael were preſent. \V Therfore al the men of
Iſrael concurring to the king, ſayd to him: Why haue our brethren the
men of Iuda ſtolen thee, and brought the king and his houshould ouer
Iordan, and al the men of Dauid with him? \V And euerie man of Iuda
anſwered the men of Iſrael: Becauſe the king is neerer to me: why art
thou angrie for this matter? haue we eaten any thing of the kinges, or
were there giftes geuen vs? \V And a certayne man of Iſrael anſwered the
men of Iuda, and ſayd: I am greater by ten partes with the king, & to me
pertayneth Dauid more then to thee: Why haſt thou done me wrong, and it
was not told me firſt, that I might bring backe my king? And the men of
Iuda anſwered more sharply then the men of Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
