%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 0672
%%% o-0603
\startChapter[
  title={Chapter 10}
  ]

\Summary{Hanon king of Ammon for euil entreating Dauids men, ſent vnto
  him of curteſie, 7.~is iuſtly plagued, with his confederates. 15.~Alſo
  the ſecond time they are ouerthrowen by Dauid.}

And it came to paſſe after theſe thinges, that the king of the children
of Ammon died, and Hanon his ſonne reigned for him. \V And Dauid ſaid: I
wil doe mercie with Hanon the ſonne of Naas, as
\SNote{This Naas king of Ammon
\Fix{curtouſly}{curteouſly}{likely typo, fixed in other}
intertained Dauids freindes which eſcaped from the king of Moab, killing
moſt of them that were cõmended to him, becauſe Dauid had leift his
countrie, and was returned into Iuda.
\XRef{1.~Reg.~22.}
\Cite{Hiſtor. Eccleſ.}}
his father hath done mercie with me. Dauid therefore ſent, conforting
him by his ſeruants vpon his fathers death. But when the ſeruantes of
Dauid were come into the land of the children of Ammon, \V the princes
of the children of Ammon ſaid to Hanon their lord: Thinkeſt thou that
for the honour of thy father Dauid hath ſent comforters vnto thee, and
not rather that he might ſearch, and ſpy into the citie, and ouerthrow
it, hath Dauid ſent his ſeruants vnto thee? \V Hanon therefore tooke the
ſeruants of Dauid, and ſhaued the one half of their
\Fix{breard,}{beard,}{obvious typo, fixed in other}
%%% o-0604
and cutte away halfe their garments vnto the buttockes, and ſent them
away. \V Which when it was told Dauid, he ſent to meete them: for the
men were confounded very fowly, and Dauid commanded them: Tary in
Iericho, til your beard be growen, and then returne. \V And the children
of Ammon ſeing that they had done iniurie to Dauid, ſent, and hyred for
wages the Syrian of Rohob, and the Syrian of Soba, twentie thouſand
footemen, and of the king Maacha a thouſand men, and of Iſtob twelue
thouſand men. \V Which when Dauid had heard, he ſent Ioab and the whole
armie of warryers. \V The children therefore of Ammon iſſued forth, and
ſette their men in aray before the verie entrance of the gate: but the
Syrian of Soba, and Rohab, and Iſtob, and Maacha were by themſelues in
the fielde. \V Ioab therefore ſeing, that
%%% 0673
there was battel prepared againſt him, both before him and behind him,
he piked out of al the choſen of Iſrael, and directed his armie agaynſt
the Syrian: \V and the reſt of the people he deliuered to Abiſai his
brother, who directed his armie againſt the children of Ammon. \V And
Ioab ſayd: If the Syrian shal preuayle againſt me, thou shalt ayde me:
and if the children of Ammon shal preuayle agaynſt thee, I wil ayde
thee. \V Play the man, and let vs fight for our people, and the citie of
our God: and our Lord wil doe that which is good in his ſight. \V Ioab
therefore and the people that were with him, began to fight againſt the
Syrians: Who immediatly fled from his face. \V And the children of Ammon
ſeing that the Syrians were fled, they alſo fled from the face of
Abiſai, and entred into the citie: and Ioab returned from the children
of Ammon, and came to Ieruſalem. \V Therefore the Syrians ſeing that
they were fallen before Iſrael, they gathered them ſelues together. \V
And Adar-ezer ſent, and fetched out the Syrians, that were beyond the
riuer, and brought their armie: and Sobach the maiſter of Adar-ezers
warre, was their chief captaine. \V Which when it was told Dauid, he
gathered together al Iſrael, and paſſed ouer Iordan, and came into
Helam: & the Syrians put them ſelues in aray againſt Dauid, & fought
againſt him. \V And the Syrians fled from the face of Iſrael, and Dauid
ſlewe of the Syrians ſeuen hundred chariotes, and fourtie thouſand
horſemen: and Sobach the prince of the warre he ſtroke: who forthwith
died. \V And al the kinges, that were to ayde Adarezer, ſeing them
ſelues ouercome of Iſrael, were afrayd and fled eight & fiftie thouſand
before Iſrael. And they made peace with Iſrael: and ſerued them, and the
Syrians were afrayd any more to ayde the children of Ammon.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
