%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-22


\project douay-rheims


%%% 0699
%%% o-0627
\startChapter[
  title={Chapter 22}
  ]

\Summary{King Dauids Canticle of thankeſgeuing, for his deliuerie from
  al enemies: 44.~VVith a prophecie of the reiection of the Iewes, and
  vocation of the Gentiles.}

And Dauid ſpake to our Lord the wordes of this ſong, in
\SNote{After that Dauid was deliuered from the handes of Saul (who firſt
and longeſt & moſt dangerouſly of al men perſecuted him, and therfore is
here ſpecially named) and from al his enimies, corporal & ſpiritual,
when he had good repoſe of mind, his viſible enimies being ſubuerted,
and his ſinnes remitted, acknowledging Gods infinite goodnes, by
inſpiration of the Holie Ghoſt, made this Canticle of thankſgeuing, and
praiſe of God.

It is inſerted amongſt the Pſalmes the 17. in order, al one in ſenſe, ſo
differing in ſome wordes that the one explicateth the other.}
the day that our Lord deliuered him out of the hand of al his enemies,
and out of the hand of
%%% !!! Same Snote as above.
%%% \SNote{}
Saul. \V And he ſayd:

Our Lord is my rocke, and my ſtrength, and my ſauiour.

\V God is my ſtrong one, I wil hope in him: my shilde, and the horne of
my ſaluation: my lifter vp, and my refuge: my ſauiour, from iniquitie
thou wilt deliuer me.

%%% o-0628
\V Our Lord prayſe worthie wil I inuocate: and from mine enemies I ſhal be
ſaued.

\V Becauſe the pangues of death haue compaſſed me: the ſtreames of
Belial haue terrified me.

\V The ropes of hel haue compaſſed me: the ſnares of death haue
preuented me.

\V In my tribulation I wil inuocate our Lord, and I wil crie my God: and
he wil heare my voice out of his holie temple, and my crie shal come to
his eares.

\V The earth quaked and trembled, the fundations of the
%%% 0700
mountaynes were ſtrycken, and ſhaken, becauſe he was angrie with them.

\V A ſmoke aroſe out of his noſethrels, and a fyre from his mouth shal
deuoure: coles were kindled from him.

\V And he bowed the heauens, and deſcended: and miſte vnder his feete.

\V And he aſcended vpon the Cherubins, and flew: and ſlidde ouer the
winges of the winde.

\V He put darkenes round about him a couer: ſtilling waters out of the
clowdes of heauen.

\V By the shyning in his preſence: the coles of fire were kindled.

\V Our Lord wil thunder from heauen: and the high one wil geue his
voice.

\V He shot his arrowes and diſperſed them: lightning, and conſumed them.

\V And the ouerflowinges of the ſea appeared, and the fundations of the
world were diſcouered at the rebuking of our Lord, at the breathing of
the ſpirit of his furie.

\V He ſent from hygh heauen, and tooke me, and drewe me out of manie
waters.

\V He deliuered me from my moſt mightie enemie, and from them that hated
me: becauſe they were ſtronger then I.

\V He preuented me in the day of my affliction, and our Lord became my
ſtay.

\V And he brought me forth into latitude, he deliuered me, becauſe I
wel pleaſed him.

\V Our Lord wil reward me according to my iuſtice: and according to the
cleannes of my handes wil he render to me.

\V Becauſe I haue kept the wayes of our Lord, and haue not done
impiouſly, from my God.

\V For al his iudgementes are in my ſight: and his preceptes I haue not
remoued from me.

\V And I shal be perfect with him: and shal keepe my ſelf from
myne iniquitie.

\V And our Lord wil reſtore vnto me according to my iuſtice: and
according to the cleannes of my handes in the ſight of his eyes.

%%% o-0629
\V With the holie one thou shalt be holie: and with the ſtrong perfect.

\V With the elect thou shalt be elect: and with the peruerſe thou shalt
be peruerted.

%%% 0701
\V And the poore people thou wilt ſaue: and the haughtie in thyne eies
thou wilt humble.

\V Becauſe thou art my lampe ô Lord: and thou wilt illuminate my
darkenes.

\V For in thee I wil runne girded: in my God I wil leape ouer the wal.

\V God, his way immaculate, the word of our Lord is examined by fyre: he
is the ſhield of al that truſt in him.

\V Who is God beſide our Lord: and who is ſtrong beſide our God?

\V God who hath girded me with ſtrength: and made euen my perfect way.

\V Making my feete equal with the hartes, and ſetting me vpon my high
places.

\V Teaching my handes vnto battel: and framing myne armes as it were a
braſen bow.

\V Thou haſt geuen me the ſhield of thy ſaluation: and thy mildenes hath
multiplied me.

\V Thou ſhalt enlarge my ſteppes vnder me: and myne ankles ſhal not
fayle.

\V I wil purſew myne enemies, and bruiſe them: and wil not returne til I
conſume them.

\V I wil conſume and breake them, that they riſe not: they ſhal fal
vnder my feete.

\V Thou haſt girded me with ſtrength to battel: thou haſt bowed vnder me
them that reſiſt me.

\V Myne enemies thou haſt made to turne to me the backe: them that hated
me, and I shal deſtroy them.

\V They shal crie, and there shal not be to ſaue, to our Lord, and he
wil not heare them.

\V I wil deſtroy them as the duſt of the earth: as the myre of the
ſtreates wil I bruiſe and breake them.

\V Thou wil ſaue me from the contradictions of my people: thou wilt
keepe me to be
\SNote{Though ſome few Gentiles were ſubdued by Dauid, and ſome were
conuerted to true religion in the old Teſtamẽt: yet the ful conuerſion
of Gentiles perteineth to the Church of Chriſt, which is here forſhewed
and deſcribed to haue perpetual ſeede for euer.}
the head of the Gentiles: the people which I knowe not, wil ſerue me.

\V The children alienes wil reſiſt me, with the hearing of the eare they
wil obey me.

\V The children alienes are fallen away, and shal be ſtraytened in their
diſtreſſes.

\V Our Lord liueth, and my God is bleſſed: and the ſtrong God of my
ſaluation shal be exalted.

%%% 0702
%%% o-0630
\V God which geueſt me reuenges, and throweſt downe peoples vnder me.

\V Which bringeſt me out from myne enemies, and from them that reſiſt me
doſt lift me vp: from the wicked man thou shalt deliuer me.

\V Therfore wil I confeſſe vnto thee ô Lord among the Gentiles, and wil
ſing to thy name.

\V Magnifying the ſaluations of his king, and doing mercie to his Chriſt
Dauid, and to his ſeede for euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
