%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 0668
%%% o-0600
\startChapter[
  title={Chapter 7}
  ]

\Summary{Dauids good purpoſe to build a Temple is differred by Gods
  appointment. 12.~With promiſe that his ſonne shal performe it, and be
  eſtablished in the kingdom. 18.~For al which he rendereth thankes to
  God.}

And it came to paſſe when the king ſate in his houſe, and our Lord had
geuen him reſt on euery ſide from al his enemies, \V he ſayd to Nathan
the prophet: Doeſt thou ſee that I dwel in a houſe of cedar, and the
arke of God is ſette
\SNote{The tabernacle made by Moyſes was a goodlie thing, but being
couered with skinnes and in manie reſpectes inſufficient for Gods
ſeruice, Dauid deſired to build a glorious Temple. But was not permitted
to do it, for myſterie ſake to ſignifie that Chriſt the true Salomon
ſhould build his Church, that farre excelleth the Synagogue of the
Iewes, and old Teſtament.
\Cite{S.~Auguſtin. li.~17. c.~8. de ciuit.}}
in the middes
\Fix{skinnes?}{of skinnes?}{obvious typo, fixed in other}
\V And Nathan ſayd to the king: Al that is in thy hart, goe doe, becauſe
our Lord is with thee. \V And it came to paſſe in that night: and behold
the word of our Lord to Nathan, ſaying: \V Goe, and ſpeake to my ſeruant
Dauid: Thus ſayth our Lord: Shalt thou build me a houſe to dwel in? \V
For neither haue I dwelt in houſe from the day that I brought the
children of Iſrael out of the Land of Ægypt, vntil this day: but I
walked in tabernacle, and in tent. \V Through out al the places, that I
haue paſſed with al the children of Iſrael, ſpeaking did I ſpeake to one
of the tribes of Iſrael, which I commanded to feede my people Iſrael,
ſaying: Why did you not build me a houſe of cedar? \V And now theſe
thinges ſhalt thou ſay to my ſeruant Dauid: Thus ſayth the Lord of
hoſtes: I tooke thee out of the paſtures folowing the flockes, that thou
ſhouldeſt be prince ouer my people Iſrael: \V and I haue beene with thee
whereſoeuer thou haſt walked, and haue ſlayne al thine enemies from thy
face: and haue made thee a great name, according to the name of the
great ones, that are in the earth. \V And I wil appoint a place for my
people Iſrael, and wil plant it, and they ſhal dwel vnder it, and ſhal
be trubled no more: neither ſhal the children of iniquitie adde to
afflict them as before. \V From the day that I appointed Iudges ouer my
people Iſrael: and I wil geue thee reſt from al thine enemies. And our
Lord foretelleth thee, that our Lord wil make thee a houſe. \V And when
thy daies ſhal be accompliſhed, and thou ſhalt ſleepe with thy fathers,
I wil rayſe vp thy ſeede after thee,
%%% 0669
which ſhal come forth of thy wombe, and
\SNote{He that ſuppoſeth this great promiſe to be fulfilled in Salomon,
erreth much ſayeth S.~Auguſtin.
\Cite{ibedem.}}
I wil eſtablish his kingdom. \V He ſhal build a houſe to my name, and I
wil eſtabliſh the throne of his kingedome for euer. \V
\SNote{S.~Paul expoundeth this of Chriſt.
\XRef{Heb.~1. v.~5.}}
I wil be to him for a father, and he ſhal be to me for a ſonne: who if
\SNote{This can not be ſaied of Chriſt, but of Salomon, and of anie
chriſtian, ſo this place hath manie literal ſenſes.}
he ſhal do any thing vniuſtly, I wil rebuke him in the rod of men, and
in the plagues of the ſonnes of men. \V But my mercie I wil not take
away from him, as I tooke from Saul, whom I remoued from thy face. \V
And thy houſe ſhal be faithful, and
\SNote{The Sea Apoſtolique, & prieſtly powre in the Church of Chriſt, is
this perpetual kingdom.
\Cite{S.~Epiphanius. Hereſi.~29.}}
thy kingdom for euer before thy face, and thy throne shal be firme
continually. \V According
%%% o-0601
to al theſe wordes, and according to al this viſion, ſo did Nathan
ſpeake to Dauid. \V And Dauid went in, and ſate before our Lord, and
ſaid: Who am I ô Lord God, and what is my houſe, that thou haſt brought
me thus farre? \V But this alſo hath ſemed litle in thy ſight ô Lord
God, vnles thou didſt ſpeake alſo of the houſe of thy ſeruant for a long
time: for this is the law of Adam, Lord God. \V What can Dauid therefore
adde yet, to ſpeake vnto thee? for thou knoweſt thy ſeruant ô Lord
God. \V For thy word, and according to thy hart thou haſt done al theſe
great thinges, ſo that thou wouldeſt notifie it to thy ſeruant. \V
Therefore art thou magnified ô Lord God, becauſe there is none like to
thee, neither is there a God beſides thee, in al thinges that we haue
heard with our eares. \V And what nation is there in the earth, as thy
people Iſrael, for the which
\SNote{Here and in manie other places the Hebrew word is of the plural
number, \HH{Elohim}, \Emph{Goddes}, ſignifying more diuine Perſons.}
God hath gone, that he might
\SNote{The worke of mans Redemption is appropriated to God the Sonne.}
redeme it to be his people, and might make him ſelfe a name, & doe for
them great wonders, and horrible thinges vpon the earth, before the face
of thy people, whom thou redemeſt to thy ſelf out of Ægypt, from the
nations and from their goddes. \V For thou haſt confirmed thy people
Iſrael to be an euerlaſting people: and thou Lord God art become their
God. \V Now therefore ô Lord God, raiſe vp for euer the word, that thou
haſt ſpoken vpon thy ſeruant, and vpon his houſe: and doe as thou haſt
ſpoken, \V that thy name may be magnified for euer, and it may be ſaid:
The Lord of hoſtes is God ouer Iſrael. And the houſe of thy ſeruant
Dauid shal be eſtablished before our Lord, \V becauſe thou ô Lord of
hoſtes God of Iſrael haſt reueled the eare of thy ſeruant, ſaying: A
houſe I build thee: therefore hath thy ſeruant found his hart to pray
thee with this prayer. \V Now therefore ô Lord God, thou art God, and
thy wordes shal be true: for thou haſt ſpoken to thy ſeruant
%%% 0670
theſe good thinges. \V Beginne therefore, and bleſſe the houſe of thy
ſeruant, that it may be for euer before thee: becauſe thou Lord God haſt
ſpoken, by thy bleſſing shal the houſe of thy ſeruant be bleſſed for
euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
