%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 0695
%%% o-0624
\startChapter[
  title={Chapter 20}
  ]

\Summary{Seba raiſeth rebellion, is purſued by Ioab, (10.~VVho in the
  way trecherouſly killeth Amaſa,) 13.~Abela is beſieged, becauſe Seba
  ſaueth him ſelf there. 20.~But his head being cut of and caſt ouer the
  wal to Ioab, the armie departeth. 23.~Chief men in office are
  mentioned.}

It chanced alſo that there was a man of Belial, named Seba, the ſonne of
Bochri, a man of Iemini: and he ſounded the trumpet, and ſayd: We haue
no part in Dauid, not inheritance in the ſonne of Iſai: Returne into thy
tabernacles Iſrael. \V And al Iſrael was ſeparated from Dauid, and
folowed Seba the ſonne of Bochri: but the men of Iuda ſtucke to their
king from Iordan vnto Ieruſalem. \V And when the king was come into his
houſe to Ieruſalem, he tooke the ten wemen his concubines, which he had
leift to keepe his houſe, and he deliuered them into cuſtodie, allowing
them victuals: and he went not in vnto them, but they were ſhut vp vntil
the day of their death liuing in widowhood. \V And the king ſayd to
Amaſa: Cal me together al the
%%% 0696
men of Iuda agaynſt the third day, and be thou preſent. \V Amaſa
therfore went to cal together Iuda, and taryed beyond the time appoynted
which the king had aſſigned him. \V And Dauid ſayd to Abiſai: Now wil
Seba the ſonne of Bochri more afflict vs, then Abſalom: take therefore
the ſeruantes of thy Lord, and purſew him, leſt perhaps he finde fenſed
cities, & eſcape vs. \V There went forth therfore with him Ioabs men,
Cerethi alſo and Phelethi: and al the ſtrong men yſſued forth of
Ieruſalem to purſew Seba the ſonne of Bochri. \V And when they were
beſide the great ſtone, which is in Gabaon, Amaſa coming mette
them. Moreouer Ioab was clothed with a ſtrayte cote according to the
meaſure of his ſtature, and vpon it girded with a ſword hanging downe to
the flanke, in a ſcabbarde, which being made for the purpoſe could with
light mouing come forth and ſtrike. \V Ioab therfore ſayd to Amaſa: God
ſaue thee my brother. And he held with his right hand the chinne of
Amaſa, as it were kiſſing him. \V But Amaſa marked not the ſword, which
Ioab had, who ſtrick him in the ſide, and powred out his bowels on the
ground, neither added he the ſecond wound, and he dyed. And Ioab, and
Abiſai his brother purſewed Seba the ſonne of Bochri. \V In the meane
time certayne men, when they ſtoode by the carcaſſe of Amaſa, Ioabs
company, ſayd: Loe he that would haue beene for Ioab the companion of
Dauid. \V And Amaſa embrewed with bloud, lay in the
%%% o-0625
middes of the way. A certayne man ſaw this that al the people ſtayed to
ſee him, and he remoued Amaſa out of the way into the filde, and couered
him with a garment, that they which paſſed might not ſtay becauſe of
him. \V He therefore being remoued out of the way, euery man paſſed
folowing Ioab to purſew Seba the ſonne of Bochri. \V Moreouer he had
paſſed through al the tribes of Iſrael vnto Abela, and Bethmaaca: and al
the choſen men were gathered together vnto him. \V They therfore came,
and aſſaulted him in Abela, and in Bethmaaca, and they compaſſed the
citie with munitions, and the citie was beſieged: and al the multitude,
that was with Ioab, laboured to deſtroy the walles. \V And a wife woman
cryed out from the citie: Heare ye, heare ye, tel Ioab: Approche hither,
and I wil ſpeake with thee. \V Who when he was come to her, ſhe ſayd to
him: Art thou Ioab? And he anſwered: I am. To whom she ſpake thus: Heare
the wordes
%%% 0697
of thy handmayd. Who anſwered: I doe heare. \V And she agayne ſayd: A
ſaying was vſed in the old prouerbe: They that aske, let them aske in
Abela: and ſo they proſpered. \V Am not I she that anſwer truth in
Iſrael, and thou ſeekeſt to ſubuert the citie, & to ouerthrowe a mother
citie in Iſrael? Why throweſt thou downe hedlong the inheritance of our
Lord? \V And Ioab anſwering, ſayd: God forbid, God forbid that I should,
I do not throw downe, nor deſtroy. \V The matter is not ſo, but a man of
mount Ephraim, Seba the ſonne of Bochri by name, hath lifted vp his hand
agaynſt king Dauid: Deliuer him onlie, and we wil depart from the
citie. And the woman ſayd to Ioab: Behold his head shal be throwen to
thee of the wal. \V She therfore went to al the people, and ſpake to
them wiſely: Who threw the head of Seba the ſonne of Bochri being cut
of, to Ioab. And he ſounded the trumpet, and they departed from the
citie, euery one into their tabernacles: and Ioab returned to Ieruſalem
vnto the king. \V Ioab therfore was ouer al the armie of Iſrael: and
Banaias the ſonne of Ioiada ouer the Cheretheites and Pheletheites. \V
But Aduram ouer the tributes: moreouer Ioſaphat the ſonne of Ahilud, was
regiſter. \V And Siua, a ſcribe: and Sadoc and Abiathar, prieſtes. \V
And Ira the Iairite was the
\SNote{Chiefe or great in familiaritie.}
prieſt of Dauid.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
