%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 0689
%%% o-0619
\startChapter[
  title={Chapter 18}
  ]

\Summary{King Dauid diſpoſeth his armie in three partes, geuing ſpecial
  charge to ſaue Abſalom aliue. 9.~VVhom neuertheles (hanging by the
  heares of his head in an oke) Ioab killeth: 16.~and ſaueth the common
  people. 19.~VVhich Dauid vnderſtanding greatly bewayleth Abſalom.}

Dauid therfore hauing viewed his people, appointed ouer them tribunes &
centurions, \V and gaue the third part of the people vnder the hand of
Ioab, and the third part
%%% 0690
vnder the hand of Abiſai the ſonne of Seruia the brother of Ioab, and
the third part vnder the hand of Ethai, who was of Geth: and the king
ſaid to the people: I alſo wil goe forth with you. \V And the people
anſwered: Thou ſhalt not goe forth: for whether we ſhal flee, it wil be
no great importance to them of vs: or whether the halfe part of vs
ſhal fal, they wil not greatly care: becauſe thou alone art accounted
for ten thouſandes: it is better therefore that thou be in the citie to
ayde vs. \V To whom the king ſaid: What ſeemeth good to you, that I wil
doe. The king therfore ſtoode beſide the gate: and al the people went
forth by their troupes, by hundredes and by thouſandes. \V And the King
commanded Ioab, and Abiſai, and Ethai, ſaying:
\SNote{Dauid moued with compaſſion towardes his ſonne Abſalom, being in
actual rebellion againſt him prefigured Chriſts cõpaſſion, towards his
perſecuters, being his creatures, praying for them in his paſſion.
\Cite{S.~Ambroſe in Pſal.~118. v.~108.}}
Saue me the child Abſalom. And al the people heard the King commanding
al the princes for Abſalom. \V The people therfore went out into the
filde againſt Iſrael, & the battel was fought in the foreſt of
Ephraim. \V And the people of Iſrael was ſlayne there of Dauids armie,
and there was made a great ſlaughter in that day, of twentie
thouſand. \V And the battel there was diſperſed vpon the face of al the
earth, and there were manie moe, whom the foreſt had conſumed of the
people, then they whom the ſword deuoured in that day. \V And it chanced
that Abſalom mette the ſeruantes of Dauid, ſitting on a mule: and when
the mule was gone in vnder a thicke oke and a great, his head ſtucke to
the oke: and he hanging betwen heauen and earth, the mule that he rode
vpon paſſed through. \V And one ſawe this & told Ioab, ſaying: I ſaw
Abſolom hang vpon an oke. \V And Ioab ſayd to the man that told him: If
thou ſaweſt him, why didſt thou not nayle him to the earth, and I had
geuen thee ten ſicles of ſiluer, and one belt? \V Who ſayd to Ioab: If
thou wouldeſt pay downe in my handes a thouſand peeces of ſiluer, I
would not lay my handes vpon the kinges ſonne: for in our hearing the
king commanded thee, and Abiſai, and Ethai, ſaying: Keepe me the child
Abſalom. \V Yea and if I had done agaynſt my life boldly, this could not
haue beene
%%% o-0620
hid from the king, and thou wouldeſt haue ſtand agaynſt it? \V And Ioab
ſayd: Not as thou wilt, but I wil ſet vpon him before thee. He tooke
therefore three lances in his hand, and thruſt them in the hart of
Abſalom: and when as yet he panted for life ſticking on the oke, \V
there ranne ten yong men the ſquyers of Ioab, and ſtriking they
killed him.
%%% 0691
\V And Ioab ſounded the trumpet, and ſtayed the people, that they should
not purſew Iſrael fleeing, willing to ſpare the multitude. \V And they
tooke Abſalom, and caſt him in the forreſt; into a great pitte, and they
heaped vpon  him an exceding great heape of ſtones: but al Iſrael fled
into their tabernacles. \V Moreouer Abſalom had erected to himſelf,
whiles he yet liued, a title which is in the kinges Valley: for he ſaid:
I haue
\SNote{Al his ſonnes being dead, for he had once three ſonnes & a daughter.
\XRef{chap.~14. v.~27.}}
no ſonne, and this shal be a monument of my name. And he called the
title by his name, and it is called The hand of Abſalom, vntil this
preſent day. \V And Achimaas the ſonne of Sadoc ſaid: I wil runne, and
tel the king, that our Lord hath done him iudgement of the hand of his
enemies. \V To whom Ioab ſaid: Thou shalt not be meſſenger this day, but
thou shalt carie the meſſage an other day: to day I wil not haue thee
carie the meſſage, for the kinges ſonne is dead. \V And Ioab ſaid to
Chuſai: Goe, and tel the king what thou haſt ſeene. Chuſai adored Ioab,
and ranne. \V And againe Achimaas the ſonne of Sadoc ſaid to Ioab: What
letteth if I alſo runne after Chuſai? And Ioab ſayd to him: Why wilt
thou runne my ſonne? thou shalt not be caryer of good tydinges. \V Who
anſwered: But what if I runne? And he ſaid to him: Runne. Achimaas
therfore running a nere way out went Chuſai. \V And Dauid ſate betwen
the two gates: and the watchman that was in the toppe of the gate vpon
the wal, lifting vp his eies, ſaw a man running alone. \V And crying out
he told the king: and the king ſaid: If he be alone, there are good
tydinges in his mouth. And he making haſt, and coming neerer, \V the
watchman ſaw an other man running, and crying alowde in the toppe, he
ſaid: There appeareth vnto me an other man running alone. And the king
ſaid: And this is a good meſſenger. \V And the watchman: I behold, ſaid
he, the running of the former, as it were the running of Achimaas the
ſonne of Sadoc. And the king ſaid: He is a good man: and cometh bringing
good newes. \V And Achimaas crying, ſayd to the king: God ſaue thee ô
king. And adoring the king before him flatte to the earth, he ſaid:
Bleſſed be our Lord thy God, who hath shut vp the men that haue lifted
vp their handes againſt my lord the king. \V And the king ſaid: Is the
child Abſalom ſafe? And Achimaas ſayd: I ſaw a great tumult, when thy
ſeruant Ioab ſent, ô king, me thy ſeruant: other thing I know not. \V To
whom
%%% 0692
the king: Paſſe, ſayd he, and ſtand here. \V And when he had paſſed, and
ſtood, \V Chuſai appeared: and coming he ſayd: I bring good tydinges my
lord king: for our Lord hath iudged for thee this day of the hand of al
that haue ryſen againſt thee. \V And the king ſayd to Chuſai: Is the
child Abſalom ſafe? To whom Chuſai anſwering, ſayd:
%%% o-0621
Let the enemies of my lord the king become, as the child, and al that
ryſe againſt him vnto euil. \V The king therfore being made ſorie, went
vp into the high chamber of the gate, and wept. And thus he ſpake,
going: My ſonne Abſalom, Abſalom my ſonne: who would graunt me that I
might die for thee, Abſalom my ſonne, my ſonne Abſalom.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
