%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 0685
%%% o-0615
\startChapter[
  title={Chapter 16}
  ]

\Summary{Siba bringing victuals obtaineth (by falſe ſuggeſtion) his
  maiſter Miphiboſeths enheritance. 5.~Semei curſeth, and threweth
  ſtones at the king, who neuertheles forbiddeth to kil him. 15.~Abſalom
  entreth into Hieruſalem, 16.~intertayneth Chuſai, 20.~and by
  Achitophels aduiſe lieth with his fathers concubines.}

%%% 0686
And when Dauid had paſſed a litle the toppe of the Mount, Siba the
ſeruant of Miphiboſeth appeared coming to meete him, with two aſſes,
which were loden with two hundred loaues, and a hundred bunches of
rayſens, an hundred maſſes of figges, and a bottel of wine. \V And the
king ſaid to Siba: What meane theſe thinges? And Siba anſwered: The
aſſes are for the kinges houſhould to ſitte on: and the loaues and the
figges to eate for thy ſeruantes, and the wine to drinke if any man ſhal
fainte in the deſert. \V And the king ſaid: Where is thy maſters ſonne?
And Siba anſwered the king: He hath remained in Ieruſalem, ſaying: This
day wil the houſe of Iſrael reſtore me the kingdom of my father. \V And
the king ſaid to Siba:
\SNote{King Dauid was here abuſed by falſe information: to which he
ought not ſo eaſely to haue geuen credite.
\XRef{chap.~19. v.~24.}}
Let al thinges be thine that were Miphiboſeths. And Siba ſaid: I beſech
thee let me find grace before thee, my lord king. \V King Dauid
therefore came as farre as Bahurim: & behold there came forth thence a
man of the kinred of the houſe of Saul named Semei, the ſonne of Sera,
and he proceded going forth, & curſed. \V And he threw ſtones againſt
Dauid, & againſt al the ſeruantes of king Dauid & the whole people, & al
the warriers went on the right, and the left ſide of the king. \V And
thus ſpake Semei when he curſed the king: Come forth, come forth thou
man of bloud, and man of Belial. \V Our Lord hath repayed thee al the
bloud of the houſe of Saul: becauſe thou haſt inuaded the
%%% o-0616
kingdom for him, and our Lord hath geuen the kingdom into the hand of
Abſalom thy ſonne: and behold thyne euiles preſſe thee, becauſe thou art
a man of bloud. \V And Abiſai the ſonne of Seruia ſaid to the king: Why
curſeth this dead dogge my lord the king? I wil goe, and ſtrike of his
head. \V And the king ſaid: What is it to me and you, ye ſonnes of
Seruia? Let him alone that he may curſe: for our Lord hath
\SNote{God ſuffered Semei, being of his owne freewil malicious for
puniſhment of Dauids ſinnes to curſe him but was not the author of his
malice, for ſo Semei had committed no fault therein and then he could
not lawfully haue benne puniſhed for it, as he was.
\XRef{3.~Reg.~2.}}
commanded him to curſe Dauid: and who is he that dare ſay, why hath he
ſo done? \V And the king ſaid to Abiſai, and to al his ſeruantes: Behold
my ſonne, that came out of my wombe, ſeeketh my life: how much more the
ſonne of Iemini? let him alone that he may curſe according to the
precept of our Lord: \V if perhaps our Lord may reſpect mine affliction,
and our Lord may render me good for this dayes curſing. \V Dauid
therefore walked and his companie in the way with him. And Semei by the
banke on the hils ſide, went ouer againſt him, curſing, and caſting
ſtones againſt him, and ſprinkling
%%% 0687
earth. \V The king therefore came, and al the people with him wearie,
and they were refreſhed there. \V But Abſalom and al his people entered
into Ieruſalem, yea and Achitophel with him. \V And when Chuſai the
Arachite Dauids frend was come to Abſalom, he ſaid to him: God ſaue thee
ô king, God ſaue thee ô king. \V To whom Abſalom: is this, quoth he, thy
kindenes toward thy frend? why wenteſt thou not with thy frend? \V And
Chuſai anſwered Abſalom: Not ſo: becauſe I wil be his, whom our Lord
hath choſen, and al this people, and al Iſrael, and with him wil
tarie. \V Yea that I may adde this alſo, whom ſhal I ſerue? not the
kinges ſonne? as I haue ſerued thy father, ſo wil I ſerue thee alſo. \V
And Abſalom ſaid to Achitophel: Conſult what we ought to doe. \V And
Achitophel ſaid to Abſalom: Goe in to the concubines of thy father,
which he hath left to keepe the houſe:
\SNote{The people doubting leſt Abſalõ might be reconciled to his
father, were not aſſured vnto him til they ſaw ſuch a crime committed as
ſemed to make reconciliatiõ impoſſible. So al rebelles and vſurpers of
others right, ſeeke by ſome enormious fact to make their adherentes and
folowers ſure vnto them: but God plagueth them in the end, as he did
both Achitophel and Abſalom.}
that when al Iſrael ſhal heare that thou haſt defiled thy father, their
handes may be ſtrengthened with thee. \V They pitched therefore a tent
for Abſalom in the houſe toppe, and he went in to his fathers concubines
before al Iſrael. \V And the counſel of Achitophel, which he gaue in
thoſe dayes, as if a man ſhould conſult God: ſo was al the counſel of
Achitophel, both when he was with Dauid, and when he was with Abſalom.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
