%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 0666
%%% o-0598
\startChapter[
  title={Chapter 6}
  ]

\Summary{VVith great ſolemnitie Dauid bringeth the Arke of God from
  Abinadabs houſe. 6.~Oza for touching it is ſodenly
  ſlaine. 9.~Whereupon Dauid fearing to bring it to his owne houſe,
  leaueth it in the houſe of Obededom three monethes, 12.~then fetcheth
  it, dancing before it, though Michol ſcorne his deuotion. 17.~He
  offereth ſacrifices, diſtributeth giftes, bleſſeth the people. 23.~And
  Michol is barren.}

And Dauid againe gathered together al the choſen of Iſrael, thirtie
thouſand. \V And Dauid aroſe, and went, and al the people that was with
him of the men of Iuda, to bring the arke of God, vpon which was
inuocated the name of the Lord of hoſtes, which ſitteth in the Cherubins
vpon it. \V And they layd the arke of God vpon a new wayne: and tooke it
out of the houſe of Abinadab, who was in Gabaa: and Oza, and Ahio the
ſonnes of Abinadab, did driue the new wayne.
%%% o-0599
\V And when they had taken it out of the houſe of Abinadab, who was in
Gabaa, Ahio keeping the arke of God went before the arke. \V But Dauid,
and al Iſrael played before our Lord in al wrought wood, both on harpes,
and lutes and timbrels and ſitterns and cymbals. \V And after they came
to the Floore of Nachon, Oza put forth his hand to the arke of God, and
held it: becauſe the oxen ſpurned, and made it leane aſide. \V And our
Lord was wrath with indignation
%%% 0667
againſt Oza, and ſtroke him for the
\SNote{Either there was no probable danger, & ſo without cauſe Oza
touched the arke, that he might ſeme to be very careful; or els he vſed
not ſo much reuerence, as he ought to haue donne.
\Cite{li.~2. c.~12. de mir. bil. S.~Scrip. apud S.~Aug. tom.~3.}}
raſhenes: who died there before the arke of God. \V And Dauid was
ſtrooken ſad, for that our Lord had ſtrooken Oza, and the name of the
place was called: The ſtriking of Oza vntil this day. \V And Dauid
feared our
Lord in that day, ſaying: How ſhal the arke of our Lord come vnto me? \V
And he would not haue the arke of our Lord turne in to himſelf into the
citie of Dauid: but he cauſed it to turne in vnto the houſe of Obededom
the Getheite. \V And the arke of our Lord dwelt in the houſe of Obededom
the Getheite three monethes: and our Lord bleſſed Obededom, and al his
houſe. \V And it was told king Dauid, that our Lord had bleſſed
Obededom, and al that he had for the arke of God. Dauid therefore went,
and brought away the arke of God out of the houſe of Obededom, into the
citie of Dauid with ioy. \V And when they had paſſed, that caried the
arke of our Lord, ſix paſſes, he immolated an oxe and a ramme. \V And
Dauid danced with al his might before our Lord. Moreouer Dauid was
girded with a linnen ephod. \V And Dauid, and al the houſe of Iſrael
brought the arke of teſtament of our Lord in iubilation, and in ſounde
of trumpet. \V And when the arke of our Lord was entred into the citie
of Dauid, Michol the daughter of Saul looking forth through a window,
ſawe king Dauid leaping, and dancing
\SNote{To dance before the arke is to dance before our Lord.}
before our Lord: and ſhe deſpiſed him in her hart. \V And they brought
in the arke of our Lord, and ſette it in his place in the middes of the
tabernacle, which Dauid had pitched for it: and Dauid offered
holocauſtes, and pacifiques before our Lord. \V And when he had
accompliſhed offering holocauſtes and pacifiques, he bleſſed the people
in the name of the Lord of hoſtes. \V And he diſtributed to al the
multitude of Iſrael as wel man as woman, to euerie one, one cake of
bread, and one roaſted peece of beefe, and flowre fryed with oyle: and
al the people went, euerie man into his houſe. \V And Dauid returned to
bleſſe his owne houſe: and Michol the daughter of Saul coming forth to
meete Dauid, ſayd: How glorious was the king of Iſrael to day vncouering
himſelf before the handmaydes of his ſeruantes, and was naked, as if one
of the ribbaldes ſhould be naked. \V And Dauid ſayd vnto Michol: Before
our Lord, which hath choſen me rather then thy father, and then al his
houſe, and commanded me that I ſhould be duke ouer the people of our
Lord in
%%% 0668
Iſrael, \V both wil I play, & wil become more vile then I haue beene:
and I wil be humble in mine eies, and with the handmaydes, of whom thou
ſpeakeſt, I wil appeare more glorious. \V Therefore vnto Michol the
daughter of Saul was there no child borne vnto the day of her death.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
