%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 0657
%%% o-0590
\startChapter[
  title={Chapter 1}
  ]

\Summary{Dauid hearing that Saul and Ionathas are ſlaine, 11.~mourneth
  with al his familie, weeping and faſting. 13.~Cauſeth him to be ſlaine
  who affirmed that he had killed king Saul. 18.~He traineth vp archers,
  19.~and inuiteth alſo al Iſrael to mourne.}

And it came to paſſe, after that Saul was dead, that Dauid returned from
the ſlaughter of Amalec, and taryed in Siceleg two dayes. \V And in the
third day there appeared a man coming out of Sauls campe, his garments
torne, and ſprinkled on the head with duſt, and as he came to Dauid, he
fel vpon his face, and adored. \V And Dauid ſaid vnto him: Whence comeſt
thou? Who ſaid to him: I fledde out of the campe of Iſrael. \V And Dauid
ſaid vnto him: What is the matter that is done? tel me. Who ſaid: The
people is fled out of the battel, and many of the people are ouerthrowen
and dead: yea Saul alſo and Ionathas his ſonne are dead. \V And Dauid
ſaid to the yong man that told him: How knoweſt thou that Saul is dead,
and Ionathas his ſonne? \V And the young man that told him, ſayd:
\SNote{He fained al this, thincking to get fauoure (for Saul killed
himſelf,
\XRef{li.~1. ch.~31.)}
but Dauid puniſhed  him, as ſuch a crime deſerued.
\XRef{v.~15.}}
By chance I came into mount Gelboe, and Saul leaned vpon his ſpeare:
moreouer the chariots and horſemen approched vnto him, \V and turning
backward, and ſeing me he called. To whom when I had anſwered, here I am: \V
he ſaid to me: Who art thou? And
%%% o-0591
I ſaid to him: I am an Amalecite. \V And he ſaid to me: Stand vpon me,
and kil me: becauſe anguiſhes hold me, and as yet al my life is in
me. \V And ſtanding vpon him, I killed him: for I knewe that he could
not liue after the fal: and I tooke the Diademe that was on his head, &
the bracelette from his arme, and haue brought to thee my lord
hither. \V And Dauid taking his garments rent them, and al the men that
were with him, \V and they
\SNote{Exequies of Saul obſerued with mourning weeping and faſting.}
mourned, and wept, and faſted vntil euening vpon Saul and vpon Ionathas
his ſonne, and vpon the people
%%% 0658
of our Lord, and vpon the houſe of Iſrael, becauſe they were fallen by
the ſword. \V And Dauid ſaid to the yong man that had told him: Whence
art thou? Who anſwered: I am the ſonne of a man a ſtranger of Amalec. \V
Dauid ſaid to him: Why didſt thou not feare to put to thy hand, to kil
the annointed of our Lord? \V And Dauid calling one of his ſeruants,
ſaid: Goe runne vpon him. Who ſtroke him, and he died. \V And Dauid ſaid
to him: Thy bloud be vpon thine owne head: for thyn owne mouth hath
ſpoken againſt thee, ſaying: I haue ſlaine the annointed of our Lord. \V
And Dauid mourned this kind of mourning vpon Saul, and vpon Ionathas his
ſonne, (\V and he commanded that they ſhould
\SNote{The Philiſthijms were ſtrong & cunning archers, therefore Dauid
commanded that his ſubiectes ſhould lerne and exerciſe the ſame maner of
fight.}
teach the children of Iuda the bowe, as it is written in the Booke of
the iuſt.) And he ſaid: Conſider, ô Iſrael for them that be dead wounded
vpon thy high places. \V The Nobles, ô Iſrael, are ſlayne vpon thy
mountaynes: how are the valiants fallen? \V Tel it not in Geth, neither
tel ye it in the high waies of Aſcalon: leſt perhaps the daughters of
the Philiſthijms be gladde, leſt the daughters of the vncircumciſed
reioice. \V Mountaines of Gelboe, let neither dew, nor rayne come vpon
you, neither be they fields of the firſt fruits: becauſe there was the
ſhield of the valiantes caſt away, the ſhield of Saul, as though he were
not annointed with oyle. \V From the bloud of the ſlaine, from the fatte
of the valiants, the arrowe of Ionathas neuer returned backward, and the
ſword of Saul did not returne emptie. \V Saul and Ionathas
amiable, and comely in their life, in death alſo were not diuided:
ſwifter then eagles, ſtronger then lyons. \V Yee daughters of 
Iſrael weepe vpon Saul, who clothed you with ſcarlet in delicaces, who gaue
golden ornaments to your attyre. \V How haue the valiantes falled in
battel? Ionathas has bene ſlayne in thy high places? \V I am ſorie for thee
my brother Ionathas exceeding beautiful, and amiable aboue the loue of
wemen. As the mother loueth her onlie ſonne, ſo did I loue thee. \V How
haue the ſtrong fallen, and the weapons of warre periſhed?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
