%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 0678
%%% o-0608
\startChapter[
  title={Chapter 13}
  ]

\Summary{Amnon rauisheth Thamar. 20.~For which Abſalom killeth him,
  37.~and flyeth into Geſsur.}

And it came to paſſe after theſe thinges, that Amnon the ſonne of Dauid
loued the ſiſter of Abſalom the ſonne of Dauid, being very beautiful,
called Thamar, \V and was fond on her excedingly, ſo that for the loue
of her he was ſicke: becauſe wheras ſhe was a virgin, it ſemed vnto him
\Fix{had hard}{hard}{obvious typo, fixed in other}
to doe any thing vnhoneſtly with her. \V But Amnon
\Fix{a}{had a}{obvious typo, fixed in other}
freind, named Ionadab the ſonne of Semmaa Dauids brother, a very wiſe
man: \V Who ſayd to him: Why art thou ſo worne away with leanenes the
kinges ſonne, day by day? Why doeſt thou not tel me? And Amnon ſayd to
him: I loue Thamar the ſiſter of my brother Abſalom. \V To whom Ionadab
anſwered: Lye vpon thy bed, and fayne ſickenes: and when thy father ſhal
come to viſite thee, ſay to him: Let my ſiſter Thamar, I pray, come to
me, to geue me meate, and
%%% o-0609
to make me broth, that I may eate of her hand. \V Amnon therefore lay
downe, and began as it were to be ſicke: and when the king came to
viſite him, Amnon ſayd to the king: Let Thamar my ſiſter come, I beſech
you, that ſhe may make in my ſight two litle ſuppinges, and I may take
meate of her hand. \V Dauid therfore ſent home to Thamar, ſaying: Come
into the houſe of Amnon thy brother, & make him broth. \V And Thamar
came into the houſe of Amnon her brother: and he lay, who taking meale
tempered it: and reſoluing it in his ſight she made ſuppinges. \V And
taking that which she had boyled, she powred it out, and ſet it before
him, and he would not eate: and Amnon ſayd: Put forth al from me. And
when they had put forth al, \V Amnon ſayd to Thamar: Bring in the meate
into the parler, that I may eate of thy hand. Thamar therfore tooke the
ſuppinges, which she had made, and caryed it in to Amnon her brother in
the parler. \V And when she had offered him the meate, he caught her,
and ſayd: Come, lie with me my ſiſter. \V Who anſwered him: Doe not ſo
my brother, doe not rauish me: for this is not lawful in Iſrael. Doe not
this folie. \V For I shal not be able to beare my reproch, and thou
shalt be as one of the foolish in Iſrael: but rather ſpeake to the king, and he
wil not denie me to thee. \V But he would not reſt at her petitions, but
preuayling by
%%% 0679
force rauiſhed her, and lay with her. \V And Amnon hated her with
exceding great hatred: ſo that the hatred was greater, wherewith he
hated her, then the loue with the which before he loued her. And Amnon
ſaid to her: Ariſe; and goe. \V Who anſwered him: This euil, which now thou
doeſt againſt me expelling me, is greater then that which thou didſt
before. And he would not heare her: \V but calling the ſeruant, that
miniſtred to him, he ſaid: Thruſt this woman out from me: and ſhut the
doore after her. \V Who was clothed with a garment downe to the foote:
for the kinges daughters that were virgins, vſed ſuch kinde of
garmentes. His ſeruant therfore thruſt her out: and ſhut the doore after
her. \V Who ſprinkling ashes on her head, renting her long garment, and
her handes vpon her head, went going on, and crying. \V And Abſalom her
brother ſayd to her: Hath Amnon thy brother lyen with thee? but now
ſiſter hold thy peace, he is thy brother: neither afflict thou thy hart
for this thing. Thamar therefore taryed pyning in the houſe of Abſalom
her brother. \V And when Dauid the king had heard theſe wordes, he was
greeued excedingly. \V Moreouer Abſalom ſpake not to Amnon neither good
nor euil: for Abſalom hated Amnon becauſe he had rauiſhed Thamar his
ſiſter. \V And it came to paſſe after the ſpace of two yeares, that the
ſheepe of Abſalom were ſhorne in Baalhaſor, which is beſide Ephraim: and
Abſalom called al the kinges ſonnes, \V and he came to the king, and
ſaid to him: Behold thy ſeruantes sheepe are to be shorne: Let the king,
I pray, with his ſeruantes come to his ſeruant. \V And the king ſaid to
Abſalom, Doe not ſo my ſonne, requeſt not that we come al, & charge
thee. And when he was earneſt with him, & he would not goe, he
%%% o-0610
bleſſed him. \V And Abſalom ſaid: If thou wilt not come, at the leaſt
let Amnon my brother, I beſech thee, come with vs. And the king ſaid to
him: It is not neceſſary that he goe with thee. \V Abſalom therefore was
earneſt with him, and he let Amnon and al the kinges ſonnes goe with
him. And Abſalom made a feaſt as it were the feaſt of a king. \V And
Abſalom had commanded his ſeruantes, ſaying: Marke when Amnon shal be
drunke with wine, and I shal ſay to you: Strike him, and kil him, feare
not: for it is I that command you: take courage, and play the valiant
men. \V Therefore the ſeruantes of Abſalom did againſt Amnon, as Abſalom
had commanded them.
%%% 0680
And al the kinges ſonnes ryſing gatte vp euery one vpon their mules, and
fled. \V And when they yet went on in their way, a rumour came to Dauid,
ſaying: Abſalom hath ſtricken al the kinges ſonnes, and there is not
leift of them ſo much as one. \V The king therfore roſe vp, and rent his
garmentes: and fel vpon the ground, and al his ſeruantes, that ſtood
about him, rent their garmentes. \V But Ionadab the ſonne of Semmaa
Dauids brother anſwering, ſayd: Let not my lord the king thinke, that al
the kinges ſonnes be ſlayne: Amnon only is dead, becauſe he was put in
the mouth of Abſalom ſince the day that he rauished Thamar his
ſiſter. \V Now therefore let not my lord the king put this word vpon his
hart, ſaying: Al the kinges ſonnes are ſlayne: becauſe Amnon only is
dead. And Abſalom fled: and the ſeruant that was the ſcoutewatch, lifted
vp his eies, and looked: and behold much people came by a byway on the
ſide of the mountayne. \V And Ionadab ſayd to the king: Loe the kinges
ſonnes be come: according to the wordes of thy ſeruant, ſo it is done. \V
And when he had ceaſed to ſpeake, the kinges ſonnes alſo appeared: &
entring in they lifted vp their voice, and wept: yea the king alſo and
al his ſeruantes bewailed with an exceding great weeping. \V Moreouer
Abſalom fleing, went to Tholomai the ſonne of Ammiud the king of Geſſur,
Dauid therefore mourned for his ſonne al daies. \V And Abſalom when he
was fled, and come into Geſſur, was there three yeares. \V And king
Dauid ceaſed to purſew Abſalom, becauſe he was comforted vpon the death
of Amnon.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
