%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 0663
%%% o-0596
\startChapter[
  title={Chapter 4}
  ]

\Summary{Baana and Rechab ſecretly kil Iſboſeth: 8.~bring his head to
  Dauid. 9.~Who condemning their fact, putteth them to death.}

And Iſboſeth the ſonne of Saul heard that Abner was ſlaine in Hebron:
and his handes were weakened, and al Iſrael was trubled. \V And the
ſonne of Saul had two men captaynes of robbers, the name of one Baana,
and the name of the other Rechab, the ſonnes of Rhemmon the Berothite
%%% 0664
of the ſonnes of Beniamin: for Beroth alſo was accounted in Beniamin. \V
And the Berothites fled into Gethaim, and were there ſtrangers vntil
that time. \V And Ionathas the ſonne of Saul had a ſonne lame in his
feete: for he was fiue yeares old, when the tydinges came of Saul and
Ionathas from Iezrahel. His nurſe therefore taking him, fled: and when
ſhe made haſt to flee, he fel, and was made lame: and he was called
Miphiboſeth. \V Therefore the ſonnes of Rhemmon the Berothite, Rechab
and Baana coming, entered into the houſe of Iſboſeth in the heat of the
day: who ſlept vpon his bed at noone. \V And they entered into the houſe
ſecretely taking eares of corne, and Rechab and Baana his brother,
ſtroke him in priuy partes and fled. \V And when they were entred into
the houſe, he ſlept vpon his bed in a parler, and ſtriking they killed
him: and taking away his head they went by the way of the deſert al
night. \V And brought the head of Iſboſeth to Dauid into Hebron: and
they ſaid to the king: Behold the head of Iſboſeth the ſonne of Saul
thine enemie who ſought thy life: and our Lord hath geuen my lord the
king this day reuenge of Saul, and of his ſeede. \V But Dauid anſwering
Rechab, and Baana his brother, the ſonnes of Rhemmon the Berothite, and
ſayd to them: Our Lord liueth, which hath deliuered my ſoule out of al
diſtreſſe, \V for ſo much as him that told me, and ſaid: Saul is dead,
who thought
%%% o-0597
that he told proſperous thinges, I apprehended, and ſlewe him in
Siceleg, to whom I ſhould haue geuen a reward for his tydinges. \V How
much more now when wicked men haue ſlaine an innocent man in his owne
houſe, vpon his bed, ſhal I not require his bloud of your hand, and take
you away from the earth? \V Dauid therefore commanded his ſeruantes, and
they ſlew them, and cutting of their handes and feete, hanged them ouer
the poole in Hebron: but the head of Iſboſeth they tooke, and buried in
the ſepulcher of Abner in Hebron.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
