%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 0670
%%% o-0601
\startChapter[
  title={Chapter 8}
  ]

\Summary{Diuers nations and countries to which Iſrael payed tribute, are
  ſubdued by Dauid, and pay tribute to Iſrael. 16.~Chief officers are
  mentioned.}

And it came to paſſe after theſe thinges Dauid ſtroke the Philiſtijms,
and humbled them, and Dauid tooke the Bridle of tribute out of the hand
of the Philiſtijms. \V And he ſtroke Moab, and meaſured them with
\SNote{For more reproch Dauid compaſſing them with cordes, as cattel are
enuironed, caſt them on the ground, and by lotte killed ſome, and ſpared
ſome aliue.}
a corde, making them euen with the earth: and he meaſured two cordes,
one to kil, and one to ſaue aliue: and Moab became ſeruing Dauid vnder
tribute. \V And Dauid ſtroke Adarezer the ſonne of Rohob king of Soba,
when he went forth to haue dominion ouer the riuer Euphrates. \V And
Dauid hauing
%%% o-0602
taken of his part a thouſand ſeuen hundred horſemen, and twentie
thouſand footemen, hoghſinewed al the chariot horſes: and he leift of
them a hundred chariotes. \V There came alſo Syria of Domaſcus, to bring
ayde vnto Adarezer the king of Soba: and Dauid ſtroke of Syria two and
twentie thouſand men. \V And Dauid put a garriſon in Syria of Damaſcus:
and Syria became ſeruing Dauid vnder tribute: and our Lord preſerued
Dauid in al thinges to whatſoeuer he went forth. \V And Dauid tooke the
golden armour, which the ſeruantes of Adarezer had, and brought them
into Ieruſalem. \V And out of Bete, and out of Beroth the cities of
Adar-ezer king Dauid tooke braſſe exceding much. \V And Tou the king of
Emath heard, that Dauid had ſtriken al the force of Adar-ezer. \V And
Tou ſent Ioram his ſonne to king Dauid, to ſalute him congratulating,
and to geue thankes: for that he had ouerthrowen Adarezer, and ſtriken
him. For Tou was enemie to Adarezer, and in his hand were veſſels of
gold, and veſſels of ſiluer, and veſſels of braſſe: \V which alſo king
Dauid ſanctified to our Lord with the ſiluer and gold, that he had
ſanctified of al the nations, which he had ſubdewed \V of Syria,
and Moab, and the children of Ammon, and the Philiſtijms, and Amalec,
and of the ſpoiles of Adarezer the ſonne of Rohob king of Soba. \V Dauid
alſo
\SNote{Sette vp an Arch in memorie of triumph.}
made himſelfe a name, when he returned hauing taken Syria in the Vaile
of Salt-pittes, eightene thouſand being ſlayne: \V and he put
%%% 0671
ſouldiours in Idumea, and placed a garriſon: and al Idumea was made to
ſerue Dauid: and our Lord preſerued Dauid in al thinges to whatſoeuer he
proceeded. \V And Dauid reigned ouer al Iſrael: Dauid alſo did iudgement
and iuſtice to al his people. \V And Ioab the ſonne of Seruia was ouer
the armie: moreouer Ioſaphat the ſonne of Ahilud was
\TNote{or chancelor.}
recorder: \V and Sadoc the ſonne of Achitob, and Achimelech the ſonne of
Abiathar, were prieſtes: and Saraias, ſcribe: \V And Banaias the ſonne
of Ioiada was ouer
\SNote{Theſe were archers and ſling throwers of the guard.
\Cite{Paraphraſis chald.}}
the Cerethi and Phelethi: and the ſonnes of Dauid
\SNote{Or prieſtes, or chief rulers. See the
\XRef{annotation. Gen.~47. v.~22.}
\XRef{Paral.~18. v.~17.}}
princes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
