%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 0687
%%% o-0617
\startChapter[
  title={Chapter 17}
  ]

\Summary{Achitophel counſelleth Abſolom preſently to aſsault his father
  with forces, 7.~Chuſai perſwadeth the contrarie, 15.~and ſecretly
  aduertiſeth the king therof. 23.~Achitophel hangeth him
  ſelf. 25.~Abſolom appointeth Amaſa general of his armie. 27.~Other
  freindes bring victuals to the kinges campe.}

Achitophel therefore ſaid to Abſolom: I wil chooſe me twelue thouſand
men, and ryſing I wil purſew Dauid this night. \V And falling vpon him
(for as much as he is weary, and of weakened handes) I wil ſtrike him:
and when al the people is fled, that is with him, I ſhal ſtrike the king
being deſolate. \V And I ſhal reduce al the people, as one man is wont
to returne: for thou ſeekeſt one man: and al the people ſhal be in
peace. \V And his ſaying pleaſed Abſalom, and al the ancientes of
Iſrael. \V But Abſalom ſayd: Cal Chuſai the Arachite, and let vs heare
what he alſo ſayeth. \V And when Chuſai was come to Abſalom, Abſalom
ſayd to him: This maner of ſpeache ſpake Achitophel: ſhal we doe it or
no? What counſel geueſt thou? \V And Chuſai ſayd to Abſalom:
%%% 0688
It is not good counſel, that Achitophel hath geuen this time. \V And
agayne Chuſai inferred: Thou knoweſt thy father, and the men that are
with him, to be verie valiant, and of fel courage, as if a beare in the
wood her whelpes being taken away ſhould rage: yea and thy father is a
man of warre, neither wil he abyde with the people. \V Perhaps he lyeth
now hid ſecretely in caues, or in ſome one place where he liſt: and when
any one shal fal in the beginning, there shal one heare whoſoeuer shal
heare it, & ſay: There is made a ſlaughter in the people that folowed
Abſalom. \V And euerie one of the moſt valiant whoſe hart is as it were
a lyons, shal fainte for feare: for al the people of Iſrael knowe thy
father to be a valiant man, and that al be ſtrong which are with him. \V
But this ſemeth vnto me to be good counſel: Let al Iſrael be gathered to
thee, from Dan to Berſabee, as the ſand of the ſea innumerable: and thou
shalt be in the middes of them. \V And we shal ſette vpon them in what
place ſoeuer they shal be found: and we shal couer them, as dew is wont
to fal vpon the earth: and we shal not leaue of the men, that are with
him, not ſo much as one. \V And if he shal enter into any citie, al
Iſrael shal caſt ropes vpon that citie round about, and we wil drawe it
into the torrent, that there be not found therof not ſo much as a litle
ſtone. \V And Abſalom ſayd, and al the children of Iſrael: The counſel
of Chuſai the Arachite is better then the counſel of Achitophel: and by
the wil of our Lord was the profitable counſel of Achitophel defeated,
that our Lord
%%% o-0618
might bring in euil vpon Abſalom. \V And Chuſai ſayd to Sadoc and
Abiathar the prieſtes: In this and this maner gaue Achitophel counſel to
Abſalom, and to the Ancientes of Iſrael: and I gaue ſuch and ſuch
counſel. \V Now therfore ſend quickly, and tel Dauid, ſaying: Tarie not
this night in the champayne of the deſert, but without delay paſſe ouer:
leſt perhaps the king be ſwalowed vp, and al the people that is with
him. \V And Ionathas & Achimaas ſtood by the Fountayne rogel: there went
a maide and told them: and they went forward, to report the meſſage to
king Dauid: for they could not be ſeene, nor enter into the citie. \V
And a certayne boy ſaw them, and told Abſalom: but they making haſt
entered into the houſe of a certayne man in Bahurim, who had a wel in
his court, and they went downe into it. \V And a woman tooke, and ſpred
a couering
%%% 0689
ouer the mouth of the wel, as it were drying ſodde barley: and ſo the
thing was not knowen. \V And when Abſaloms ſeruantes were come into the
houſe, they ſayd to the woman: Where is Achimaas, and Ionathas? And the
woman anſwered them: They paſſed in haſt, hauing taſted a litle
water. But they that ſought, when they had not found, returned into
Ieruſalem. \V And when theſe were gone, they went vp out of the wel, and
going on told king Dauid, and ſayd: Aryſe ye, and paſſe quickly the
riuer: becauſe this maner of counſel hath Achitophel geuen againſt
you. \V Dauid therfore aroſe, and al the people that was with him, and
they paſſed ouer Iordan, vntil it waxed light, and not one at al was
remayning, which did not paſſe the riuer. \V Moreouer Achitophel ſeing
that his counſel was not executed, ſadled his aſſe, and roſe and went
into his houſe and into his citie: and taking order with his houſe,
\SNote{Bad counſail often falleth worſt to the counſeller.}
hanged him ſelf, and was buried in the ſepulchre of his father. \V But
Dauid came into the Campe, and Abſalom paſſed ouer Iordan, he and al the
men of Iſrael with him. \V But Abſalom appoynted Amaſa for Ioab ouer the
armie: and Amaſa was the ſonne of a man, which was called Iethra of
Iezrael, who went in to Abigail the daughter of Naas, the ſiſter of
Saruia which was the mother of Ioab. \V And Iſrael camped with Abſalom
in the Land of Galaad. \V And when Dauid was come into the Campe, Sobi
the ſonne of Naas of Rabbath the ſonnes of Ammon, and Machir the ſonne
of Amihel of Lodabar, & Berzellai the Galaadite of Rogelim, \V preſented
vnto him hanginges, and tapeſtrie, and earthen veſſels, wheate, and
barley, and meale, and polent, and beanes, and
\Fix{rishe,}{riſe,}{obvious typo, fixed in other}
and fryed peaſe, \V and honie, and butter, sheepe, & fatte calues, and
they gaue to Dauid and the people, that was with him, to eate: for they
ſuſpected that the people with hunger and thyrſt was faynte in the
deſert.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
