%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-23


\project douay-rheims


%%% 0702
%%% o-0630
\startChapter[
  title={Chapter 23}
  ]

\Summary{The laſt wordes of Dauid concerning reward of the good, 6.~and
  punishment of the bad. 8.~A catalogue of Dauids valiant men.}

And theſe are Dauides laſt wordes. Dauid the ſonne of Iſai ſayd: The man
ſayd, to whom it was appointed concerning the Chriſt of the God of
Iacob, the excellent Pſalmiſt of Iſrael: \V The Spirit of our Lord hath
ſpoken by me, and his wordes by my tongue. \V The God of Iſrael ſayd to
me, the Strong one of Iſrael hath ſpoken, the Dominatour of men, the
iuſt ruler in the feare of God. \V As the light of the mourning when the
ſunne ryſeth, early without cloudes gliſtereth, and as by rayne graſſe
ſpringeth out of the earth. \V
\SNote{King Dauid in this laſt prophecie plainly diſtinguiſheth betwen
the couenant & promiſe made to him touching his earthlie kingdom and the
kingdom of \Emph{Chriſt}, who ſhould be borne of his ſeede.

In both which he foretelleth the reward of the good and puniſhment of
the bad.}
Neither is my houſe ſo great with God, that he should enter with me an
eternal couenant firme in al thinges & aſſured. For al my ſaluation, and
al my wil: neither is there ought therof that ſpringeth not. \V And
tranſgreſſors shal be plucked vp as thornes euery one: which are not
taken with handes. \V And if a man wil touch them, he shal be armed with
yron and a lance ſtaffe, and kindled with fyre they shal be burnt vnto
nothing. \V Theſe be the names of the valiantes of Dauid. Sitting in his
chaire the wiſeſt prince amongeſt three, he is as it were the moſt tender
litle worme of the wood, which killed eight hundred at one brunt. \V
After him, Eleazar the ſonne of his vncle the Ahohite among the three
valiantes, that were with Dauid when they defyed the Philiſthijms, and
were gathered thither into battel. \V And when the men of Iſrael were
gone vp, he ſtood and ſtroke the Philiſtians til his hand faynted, and
waxed ſtiffe with the ſword: and our Lord made a great victorie that
day: and the people, that was fled, returned to take away of the ſpoyles
of them that were ſlayne. \V And after him Semma the ſonne of
%%% 0703
Age of Arari. And the Philiſthijms were gathered
\Fix{togehter}{together}{obvious typo, fixed in other}
in their ward: for there was there a filde ful of rice. And when the people
was fled from the face of the Philiſthijms, \V he ſtood in the middes
of the filde, and defended it, and
%%% o-0631
ſtroke the Philiſthians: and our Lord gaue great ſaluation. \V Moreouer
alſo before there went downe three which were princes among thirtie, and
came to Dauid in the harueſt time into the caue of Odollam: and the
campe of the Philiſtianes was placed in the Vale of the giantes. \V And
Dauid was in a hold: moreouer the ward of the Philiſthianes was then in
Bethlehem. \V Dauid
\SNote{The king propoſed not this for deſire of that water, but to trie
and exerciſe his mens fortitude.}
therfore deſyred, & ſayd: O that ſome man would geue me drinke of the
water out of the ceſterne, that is in Bethlehem beſide the gate. \V
Three valiantes therfore brake into the campe of the Philiſtianes, and
drew water out of the ceſterne of Bethlehem, that was beſide the gate,
and brought it to Dauid: but he would not drinke, but
\SNote{Precious thinges are moſt mete to be offered to God.}
offered it to our Lord, \V ſaying: Our Lord be merciful to me, that I
doe not this thing: ſhal I drinke the bloud of theſe men that went, &
the peril of their liues? Therfore he would no drinke. Theſe thinges did
the three ſtrongeſt. \V Abiſai alſo the brother of Ioab the ſonne of
Saruia, was prince of three, it is he that lifted vp his ſpeare agaynſt
three hundred, whom he ſlewe, renowmed among three, \V and the nobleſt
of three, and he was the chiefe of them, but to the three firſt he
raught not. \V And Banaias the ſonne of Ioiada the moſt valiant man of
great workes, of Cabſeel: he ſtroke the two lions of Moab, and he went
downe, and ſtroke the lyon in the middes of the ceſterne in the dayes of
ſnow. \V He alſo ſtroke the Ægyptian, a man worthie to be a ſpectacle,
hauing in his hand a ſpeare: therfore when he came downe to him with a
rod, by force he wreſted the ſpeare out of the hand of the Ægyptian, and
ſlewe him with his owne ſpeare. \V Theſe thinges did Banaias the ſonne
of Ioiada. \V And he renowmed among the three valiantes, which were the
nobler among thirtie: but vnto the three he raught not: and Dauid made
him of his ſecrete counſel. \V Aſael the brother of Ioab among the
thirtie, Elahanan the ſonne of his vncle of Bethlehem. \V Semma of
Harodi, Elica of Harori, \V Heles of Phalti, Hira the ſonne of Acces of
Thecua, \V Abiezer of Anathoth, Mobonnai of Huſati, \V Selmon the
Ahohite, Maharai the Netophathite, \V Heled the ſonne of Baana, he alſo
a
%%% 0704
Netophathite, Ithai the ſonne of Ribai of Gabaath of the children of
Benjamin, \V Banaia the Pharathonite, Heddai of the Torrent Gaas, \V
Abialbon the Arbathite, Azmaueth of Beromi, \V Eliaba of Salaboni: The
ſonnes of Iaſſen, Ionathan, \V Semma of Orori, Ahiam the ſonne of Sarar
the Ararite, \V Eliphelet the ſonne of Aaſbai the ſonne of Machati, Eliam
the ſonne of Achitophel the Gelonite, \V Heſrai of Carmel, Pharai of
Arbi, \V Igaal the ſonne of Nathan of Soba, Bonni of Gadi, \V Selec of
Ammoni, Naharai the Berothite the ſquyer of Ioab the ſonne of Saruia, \V
Ira the Iethrite, Gareb he alſo a Iethrite, \V Vrias the Hetheite. Al
thirtie ſeuen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
