%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 0683
%%% o-0613
\startChapter[
  title={Chapter 15}
  ]

\Summary{Abſalom getteth fauour of the people, 7.~and conſpireth in
  Hebron agaynſt his father. 14.~VVho fleeing, 19.~with difficultie
  permitteth Ethas a ſtranger to goe with him: 24.~but ſendeth Sadoc,
  and other prieſtes and leuites with the arke back into the
  citie. 31.~Sendeth Chuſat to defeate Achitophels counſel.}

Therefore after theſe thinges Abſalom made him ſelf chariotes, and
horſemen, and fiftie men, that should goe before him. \V And Abſalom
riſing early, ſtoode beſide the entrance of the gate, and euerie man
that had buſines to the kinges iudgement, did Abſalom cal to him, and
ſayd: Of what citie art thou? Who anſwering ſayd: Of ſuch a tribe of
Iſrael am I thy ſeruant. \V And Abſalom anſwered him: Thy wordes ſeeme
vnto me good and iuſt. But there is none to heare thee appointed of the
king. And Abſalom ſayd: \V Oh who would appoint me iudge ouer the land,
that al might come to me which haue buſines, and I might iudge
iuſtly? \V Yea and when a man came vnto him to ſalute him, he put forth
his hand, and taking him, kyſſed him. \V And this did he to al Iſrael
coming for iudgement, to be heard of the king, and he intiſed the hartes
of the men of Iſrael. \V And after fourtie yeares, Abſalom ſayd to king
Dauid: Let me goe, and pay my vowes which I haue vowed to our Lord in
Hebron. \V For thy ſeruant vowing did vow, when he was in Geſſur of
Syria, ſaying: If our Lord wil bring me againe into Ieruſalem, I wil
ſacrifice to our Lord. \V And king Dauid ſayd to him: Goe in peace. And
he aroſe, and went into Hebron. \V And Abſalom ſent ſpies into al the
tribes of Iſrael, ſaying: Forthwith as you shal heare the ſound of the
trumpet, ſay ye:
%%% 0684
Abſalom reigneth in Hebron. \V Moreouer with Abſalom there went two
hundred men out of Ieruſalem being called, going with a ſimple hart, and
vtterly ignorant of the cauſe. \V Abſalom alſo ſent for Achitophel the
Gilonite, Dauids counſeler, from his citie Gilo. And when he immolated
victimes, there was made a ſtrong conſpiracie, and the people running
together increaſed
%%% o-0614
with Abſalom. \V A meſſenger therefore came to Dauid, ſaying: Al Iſrael
with al their hart foloweth Abſalom. \V And Dauid ſaid to his ſeruantes,
that were with him in Ieruſalem: Ariſe let vs flee: for there wil be no
eſcape for vs from the face of Abſalom: make haſt to goe out, leſt
coming perhaps he ouertake vs, and force ruine vpon vs, and ſtrike the
citie in the edge of the ſword. \V And the kinges ſeruantes ſaid to him:
Al thinges whatſoeuer our lord the king ſhal command, we thy ſeruantes
wil gladly execute. \V The king therefore went forth, and al his houſe
on foote: & the king leift ten wemen his
\SNote{Concubines were lawfully maried but had not al priuilegies as
other wiues. See
\XRef{Gen.~25.}
\XRef{Iudic.~19.}}
concubines to keepe the houſe. \V And the king going forth & al Iſrael
on their feete, ſtoode farre from the houſe: \V and al his ſeruantes
walked by him, and the legions Cerethi, and Phelethi and al the
Getheites, valiant warriers, ſix hundred men which had folowed him from
Geth footemen, went before the king. \V And the king ſaid to Ethai the
Getheite: Why comeſt thou with vs? returne and dwel with the king,
becauſe thou art a ſtranger, and art come forth out of thy place. \V
Yeſterday thou cameſt, and to day ſhalt thou be forced to goe forth with
vs? but I wil goe whither I ſhal goe: returne, and leade backe thy
brethren with thee, and our Lord wil doe with thee mercie, and veritie,
becauſe thou haſt ſhewed grace and fidelitie. \V And Ethai anſwered the
king, ſaying: The Lord liueth, and my lord the king liueth: for that in
what place ſoeuer thou ſhalt be, my lord king, either in death, or in
life, there wil thy ſeruant be. \V And Dauid ſaid to Ethai: Come, and
paſſe. And Ethai the Getheite paſſed, and al the men that were with him,
and the reſt of the multitude. \V And they al wept with a lowd voice,
and al the people paſſed: the king alſo went ouer the Torrent Cedron,
and al the people marched againſt the way, that looketh to the
deſert. \V And Sadoc alſo the prieſt came, and al the Leuites with him
carying the arke of the couenant of God, and they ſette downe the arke
of God: & Abiathar aſcended, til al the people was fully
%%% 0685
paſſed, which was come forth of the citie. \V And the king ſayd to
Sadoc: Cary backe the Arke of God into the citie: if I ſhal finde grace
in the ſight of my Lord, he wil bring me agayne, and wil show me it, and
his tabernacle. \V But if he shal ſay to me: Thou pleaſeſt me not: I am
readie, let him doe that which is good before him. \V And the king ſayd
to Sadoc the prieſt: O ſeer returne into the citie in peace: and
Achimaas thy ſonne, and Ionathas the ſonne of Abiathar, your two ſonnes
let them be with you. \V Behold I wil be hid in the champayne of the
deſert, til there come word from you aduertiſing me. \V Sadoc therefore
and Abiathar caryed backe the Arke of God into Ieruſalem: and they
taried there. \V Moreouer Dauid went vp mount Oliuet, climbing &
weeping, going bare foote, and his head
\SNote{He couered his head that he might not be ſeene to weepe, leſt he
ſhould diſcorege the people. Neuertheles the people alſo wept, and
likewiſe couered their heades.}
couered, yea and al the people which was with him, their head couered
went vp weeping. \V And it was told Dauid that Achitophel alſo was in
the conſpiracie with Abſalom, and Dauid ſayd: Infatuate ô Lord I beſech
thee, the counſel of Achitophel.
%%% o-0615
\V And when Dauid went vp to the toppe of the mount, wherein he would
adore our Lord, behold there mette him Chuſai the Arachite, his garment
rent and his head ful of earth. \V And Dauid ſayd to him: If thou come
with me, thou shalt be a burden to me: \V but if thou returne into the
citie, and wilt ſay to Abſalom: I am thy ſeruant, ô king: as I haue
beene thy fathers ſeruant, ſo I wil be thy ſeruant: thou shalt defeate
the counſel of Achitophel. \V And thou haſt with thee Sadoch, and
Abiathar the prieſts: and euery word whatſoeuer thou shalt heare from
out of the kinges houſe, thou shalt tel Sadoc, and Abiathar the
prieſtes. \V And there are with them their two ſonnes Achimaas the ſonne
of Sadoc, and Ionathas the ſonne of Abiathar: and you shal ſend by them
vnto me euery word whatſoeuer you shal heare. \V Chuſai therefore the
frend of Dauid coming into the citie, Abſalom alſo entred into
Ieruſalem.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
