%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Kings
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 0661
%%% o-0594
\startChapter[
  title={Chapter 3}
  ]

\Summary{Iſboſeths forces daily decay, Dauid increaſeth in powre, and
  hath manie ſonnes. 7.~Abner vpon occaſion of a
  \Fix{ſcalnder}{ſlander}{obvious typo, fixed in other}
  breaketh from Iſboſeth, and ſerueth Dauid, bringing Michol to him, and
  much people of Iſrael. 23.~Ioab of emulation and reuenge killeth
  Abner. 28.~For which Dauid is offended with Ioab, 31.~and mourneth for
  Abner.}

There was made therefore long ſtrife betwen the houſe of Saul, and
betwen the houſe of Dauid: Dauid proſpering and alwaies ſtronger then
himſelf, but the houſe of Saul decreaſing daily. \V And there were
ſonnes borne to Dauid in Hebron: and his firſt begotten was Ammon of
Achinoam the Iezrahelite. \V And after him Cheleab of Abigail the wife
of Nabal of Carmel: moreouer the third Abſalom the ſonne of Maacha the
daughter of Tolmai king of Geſſur. \V And the fourth, Adonias, the ſonne
of Haggith: and the fifth Saphathia, the ſonne of Abital. \V The ſixt
alſo Iethraam of Egla the wife of Dauid. Theſe were borne to Dauid in
Hebron. \V Therefore when there was battel betwen the houſe of Saul and
houſe of Dauid, Abner the ſonne of Ner ruled the houſe of Saul. \V And
Saul had a concubine named Reſpha, the daughter of Aia. And Iſboſeth
ſaid to Abner: \V Why diddeſt thou go in to my fathers concubine? Who
being wrath exceedingly for the wordes of Iſboſeth, ſaid: What
\SNote{Am I contemptible in thy ſight, and yet head of them that oppoſe
againſt Dauid, I that haue donne ſo much for thee, wil not indure to be
reprehended, for a ſmal fault. So God ſuffereth the maintainers of an
euil quarel to fal out among themſelues, wherby the right cauſe is
aduanced.}
am I a dogges head againſt Iuda this day, which haue done mercie vpon
the houſe of Saul thy father, and vpon his brethren and neereſt freinds,
& haue not deliuered thee into the hands of Dauid, & haſt thou ſought
againſt me that thou mighteſt charge me for a woman to day? \V Theſe
things do God to Abner, and theſe things adde he to him, vnleſſe as our
Lord hath ſworne to Dauid, ſo I doe truth to him. \V That the kingdom be
transferred from the houſe of Saul, and the throne of Dauid be exalted
ouer Iſrael, and our Iuda, from Dan to Berſabee. \V And he could not
anſwer him any thing, becauſe he feared him. \V Abner therefore ſent
meſſengers to Dauid for himſelf ſaying: Whoſe is the land? And that they
ſhould ſay: Make amitie with me, & my hand ſhal be with thee: and I wil
reduce vnto thee al Iſrael. \V Who ſaid: Very wel: I wil make amitie
%%% 0662
with thee: but one thing I deſire of thee, ſaying: Thou ſhalt not ſee my
face before thou bring Michol the daughter of Saul: and ſo thou ſhalt
come, and ſee me. \V And Dauid ſent meſſengers to Iſboſeth the ſonne of
Saul, ſaying: Reſtore my wife Michol, which I betrothed to me for an
hundred prepuces of the Philiſthijms. \V Iſboſeth therefore ſent, and
tooke her from her huſband Phaltiel, the ſonne of Lais.
%%% o-0595
\V And her huſband folowed her, weeping as farre as Bahurim: and Abner
ſaid to him: Goe, and returne. Who returned. \V Abner alſo ſpake to the
ancientes of Iſrael, ſaying: Both yeſterday and the day before you
ſought Dauid that he might reigne ouer you. \V Now therefore do ſo:
becauſe our Lord hath ſpoken to Dauid, ſaying: In the hand of my
ſeruant Dauid I wil ſaue my people Iſrael from the hands of the
Philiſtijms, and of al their enemies. \V And Abner ſpake alſo to
Beniamin. And he went to ſpeake vnto Dauid in Hebron al things which
pleaſed Iſrael, and al Beniamin. \V And he came to Dauid into Hebron with twentie
men: and Dauid made a feaſt to Abner, and to his men that came with
him. \V And Abner ſayd to Dauid: I wil riſe, that I may gather vnto thee
my lord king al Iſrael, and may enter a league with thee, and thou
mayeſt reigne ouer al, as thy ſoule deſireth. When Dauid therefore had
brought Abner on the way, and he was gone in peace, \V forthwith Dauids
ſeruantes and Ioab came, hauing ſlayne the robbers, with a praye
exceding great: and Abner was not with Dauid in Hebron, becauſe he had
now diſmiſſed him, and he was departed in peace. \V And Ioab, and al the
armie which were with him, came afterward: it was therefore told Ioab,
that Abner the ſonne of Ner came to the king, and he diſmiſſed him, and
he departed in peace. \V And Ioab went in to the king, and ſaid: What
haſt thou done? Behold Abner came to thee: why didſt thou diſmiſſe him
and he is gone and departed? \V Knoweſt thou not Abner the ſonne of Ner,
that to this end he came to thee, that he might deceiue thee, and might
know thy going out, and thy coming in, and vnderſtand al things that
thou doeſt? \V Ioab therfore being gone from Dauid, ſent meſſengers
after Abner, and brought him backe from the ceſterne Sira, Dauid being
ignorant therof. \V And when Abner was returned into Hebron, Ioab
brought him aſide to the middes of the gate, to ſpake vnto him in guile:
and ſtroke him there in the priuie
%%% 0663
partes, and he died, in reuenge of the bloud of Aſael his brother. \V
Which when Dauid had heard, that the thing was now done, he ſaid: I am
innocent, and my kingdom before God for euer from the bloud of Abner the
ſonne of Ner, \V and come it vpon the head of Ioab, and vpon al his
fathers houſe: neither let there fayle of the houſe of Ioab one hauing a
fluxe of ſeede, and a leper, and houlding the diſtaffe, and falling by
the ſword, lacking bread. \V Ioab therefore and Abiſai his brother ſlewe
Abner, becauſe he had killed Aſael their brother in Gabaon in the
battel. \V And Dauid ſaid to Ioab, and to al the people, that were with
him: Rent your garmentes, and be girded with ſacke clothes, and mourne
before the funeral of Abner. Moreouer king Dauid folowed the beere. \V
And when they had buried Abner in Hebron, king Dauid lifted vp his
voice, and wept vpon the graue of Abner: and al the people alſo wept. \V
And the king mourning and lamenting Abner, ſayd: Not as cowardes are
wont to die, hath Abner died. \V Thy handes were not bound, and thy
feete were not loden with fetters: but as men are wont
%%% o-0596
to fal before the children of iniquitie, ſo art thou fallen. And al the
people dubling it wept vpon him. \V And when al the multitude was come
to take meate with Dauid, when it was yet cleere day Dauid ſware,
ſaying: Theſe thinges do God to me, and theſe adde he, if before
ſunne ſet I ſhal taſt bread or any thing els. \V And al the people
heard, and al thinges pleaſed them which the king did in the ſight of al
the people. \V And al the people knewe, and al Iſrael in that day that
it was not the kinges doing, that Abner the ſonne of Ner was ſlayne. \V
The king alſo ſaid to his ſeruantes: Are you ignorant that a prince and
the greateſt is ſlayne this day in Iſrael? \V But I as yet
\SNote{Weake, being newly receiued king, and not able to puniſh ſtrong
offenders. But Ioab & others were afterward puniſhed.
\XRef{3.~Reg.~2.}}
delicate, and annointed king: moreouer
\Fix{the the}{the}{obvious typo, fixed in other}
ſonnes of Saruia are hard to
me: our Lord reward him that doth euil according to his malice.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
