%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 0905
%%% o-0813
\startChapter[
  title={Chapter 10}
  ]

\Summary{Roboam
\MNote{The ſecond part.

Diuiſion of the kingdom. Dauids iſſue reigning only in two tribes, which
alſo are caried captiue into Babylon.}
requeſted by Ieroboam and the people to lighten their yoke of ſeruice,
6.~leauing the counſel of the ancient, and folowing young counſellers,
threatneth to preſse the people: 16.~wherupon manie reuolt from him.}

And
\CNote{\XRef{3.~Reg.~12.}}
Roboam went forth into Sichem: for thither al Iſrael was aſſembled, to
make him king. \V Which when
\Fix{Ieraboam}{Ieroboam}{obvious typo, fixed in other}
the ſonne of Nabat had heard, who was in Ægypt (for he was fled thither
from Salomon) forthwith he returned. \V And they called him, & he came
with al Iſrael & ſpake to Roboam, ſaying: \V Thy father preſſed vs with
a moſt hard yoke, do thou command lighter thinges then thy father, who
layd vpon vs a heauie ſeruitude, and eaſe thou a litle of the burden,
that we may ſerue thee. \V Who ſayd: After three daies returne ye to
me. And when the people was gone, \V he tooke counſel with the
Ancientes, which ſtoode before his father Salomon, whiles he yet liued,
ſaying: What counſel geue you, that I may anſwer the people? \V Who ſayd
to him: If thou wilt pleaſe this people, and pacifie them with
%%% 0906
wordes of clemencie, they wil ſerue thee at al times. \V But he forſooke the
counſel of the Ancientes, and began to treat with yongmen, that had
beene brought vp with him, and were in his trayne. \V And he ſayd to
them: What ſemeth to you? or what ſhal I anſwer this people, which hath
ſayd to me: Eaſe the yoke which thy father layd vpon vs? \V But they
anſwered as yongmen, and brought vp with him in
\Fix{delicaneſſe,}{delicateneſſe,}{obvious typo, fixed in other}
and ſayd: Thus ſhalt thou ſpeake to the people, that ſayd to thee: Thy
father aggrauated our yoke, doe thou eaſe it: and thus ſhalt thou anſwer
them: My leaſt finger is thicker then the loynes of my father. \V My
father layd vpon
\Fix{yow}{you}{obvious typo, fixed in other}
an heauie yoke, and I wil adde a greater weight: my father bette you
with ſcourges, but I wil beate you with ſcorpions. \V Ieroboam therfore
came, and al the people to Roboam the third day, as he had commanded
them. \V And the king anſwered rough wordes, leauing the counſel of the
Ancientes: \V and he ſpake according to the yongmens wil: My father layd
vpon you a heauie yoke, which I wil make heauier: my father bette you
with ſcourges, but I wil beate you with ſcorpions. \V And he
condeſcended not to the peoples requeſtes: for it was the wil of God,
that his word ſhould be accompliſhed, which he had ſpoken by the hand of
Ahias the Silonite to Ieroboam the ſonne of Nabat. \V And al the
%%% o-0814
people when the king ſpake rough wordes, ſayd thus vnto him: We haue no
part in Dauid, nor inheritance in the ſonne of Iſai. Returne into thy
tabernacles ô Iſrael, and do thou feede thy houſe Dauid. And Iſrael went
into their tabernacles. \V But ouer the children of Iſrael, that dwelt
in the cities of Iuda, Roboam reigned. \V And king Robaom ſent Aduram,
who was ouer the tributes, and the children of Iſrael ſtoned him, and he
died: moreouer king Roboam made haſt to get vp into his chariote, and
fled into Ieruſalem. \V And Iſrael reuolted from the houſe of Dauid
vntil this day.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
