%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 31
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-31


\project douay-rheims


%%% 0940
%%% o-0844
\startChapter[
  title={Chapter 31}
  ]

\Summary{Idoles being deſtroyed in al Iuda and part of Iſrael, Prieſtes
  and Leuites freely execute their functions. 4.~Tithes and firſt
  fruictes are payed in ſuch abundance, 15.~that they are put in ſtore
  houſes, and diſtributed by officers.}

%%% 0941
And when theſe thinges had bene ritely celebrated, al Iſrael that was
found in the cities of Iuda, went forth, and they brake the idols, and
cut downe the groues, ouerthrew the excelſes, and deſtroyed the altars,
not onlie out of al Iuda and Beniamin, but out of Ephraim alſo and
Manaſſes, til they vtterly deſtroyed them: and al the children of Iſrael
returned into their poſſeſſions and cities. \V And Ezechias appoynted
companies of Prieſtes, and of Leuites, by their diuiſions, euerie man in
his owne office, to witte, as wel of the Prieſtes, as of the Leuites,
for the holocauſtes, and pacifiques, that they ſhould miniſter, and
\SNote{Beſides confeſſion of ſinnes there is alſo cõfeſſion of Gods
excellencie & goodnes.}
confeſſe, and ſing in the gates of the campe of our Lord. \V And the
kinges part was, that of his proper ſubſtance holocauſt ſhould be
offered, morning alwaies and euening, in the Sabbathes alſo, and the
Calendes and in other
\Fix{ſolennities,}{ſolemnities,}{obvious typo, fixed in other}
as it is writen in the law of Moyſes. \V He commanded alſo the people
that dwelt in Ieruſalem, to geue portions to the Prieſtes, and the
Leuites, that they might attend the law of our Lord. \V Which when it
was noyſed in the eares of the multitude, the children of Iſrael offered
very manie firſt fruites of corne, of wine, and of oyle, of honie alſo:
and of al thinges, which the ground bringeth forth, they offered
tithes. \V Yea and the children of Iſrael and Iuda, that dwelt in the
cities of Iuda, offered tithes of oxen, and ſheepe, and tithes of
ſanctified thinges, which they had vowed to our Lord their God: and
carying them al, made manie heapes. \V The third moneth they began to
lay the fundations of the heapes, and in the ſeuenth moneth they
finiſhed them. \V And when Ezechias, and his princes came in, they ſaw
the heapes, and bleſſed our Lord, and the people of Iſrael. \V And
Ezechias asked the Prieſtes and the Leuites, why the heapes lay ſo: \V
Azarias the high Prieſt of the ſtocke of Sadoc anſwered him, ſaying:
Since firſt fruites began to be offered in the houſe of our Lord, we
haue eaten, and haue bene ful, and very
%%% o-0845
much hath remayned, becauſe our Lord hath bleſſed his people: and of the
remaynes this is the abundance, which thou ſeeſt. \V
\Fix{Fzechias}{Ezechias}{obvious typo, fixed in other}
therfore commanded that ſtorehouſes ſhould be prepared in the houſe of
our Lord. Which when they had done, \V they brought in as wel the firſt
fruites, as the tithes, and
\Fix{watſoeuer}{whatſouer}{likely typo, fixed in other}
they had vowed, faythfully. And the ouerſeer of them was Chonenias a
Leuite, and Semei his brother
%%% 0942
the ſecond, \V after whom Iahiel, and Azarias, and Nahath, and Aſael,
and Ierimoth, Ioſabad alſo, and Eliel, and Ieſmachias, and Mahath, and
Banaias, ouerſeers vnder the hand of Chonenias, and Semei his brother, by
the commandment of Ezechias the king, and Azarias the high Prieſt of the
houſe of God, to whom al thinges apperteyned. \V But Core the ſonne of
Iemna a Leuite, and porter of the eaſt gate, was ouerſeer of thoſe
thinges, which were voluntarily offered to our Lord, and of the firſt
fruites and the thinges conſecrated for \L{Sancta ſanctorum}. \V And
vnder his charge Eden, and Beniamin, Ieſue, and Semeias, Amarias alſo,
and Sechenias, in the cities of the Prieſtes, that they ſhould
faithfully diſtribute to their brethren portions, to the leſſer and
greater: \V ſauing the men children from three yeares and aboue, to al
that entered into the temple of our Lord, and whatſoeuer day by day was
profitable in the miniſterie, and the obſeruances according to their
diuiſions, \V to the Prieſtes by families, and to the Leuites from the
twentith yeare and vpward, by the orders and companies, \V and to al the
multitude, as wel their wiues, as their children of both ſex, meates
were geuen faithfully of theſe thinges, that had bene ſanctified. \V Yea
and of the children of Aaron by the fildes and the ſuburbes of euerie
citie, there were men ordayned, that ſhould diſtribute portions, to al
the male ſexe, of the Prieſtes and the Leuites. \V Ezechias therfore
did al thinges which we haue ſayd in al Iuda: and wrought good and
right, and truth before our Lord his God, \V in al the ſeruice of the
miniſterie of the houſe of our Lord, according to the law and the
ceremonies, willing to ſeeke his God in al his hart, and he did it and
proſpered.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
