%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 0908
%%% o-0815
\startChapter[
  title={Chapter 12}
  ]

\Summary{For the ſinnes of Roboam, and the people manie ſtrong cities,
  alſo Ieruſalem, are taken and ſpoyled by the king of Ægypt. 8.~They
  repent, and the Ægyptians depart, 9.~but carie away the
  treaſures. 13.~Roboam dieth, and his ſonne Abias reigneth.}

And when the kingdom of Roboam was ſtrengthened and fortified, he
forſooke the law of our Lord, and al Iſrael with him. \V And in the
fifth yeare of the kingdom of Roboam, came vp Seſac the king of Ægypt
into Ieruſalem (becauſe they had ſinned to our Lord) \V with a thouſand
two hundred chariotes, and threeſcore thouſand horſemen. \V Neither was
anie number of the common people, that came with him out of Ægypt, to
witte, Lybians, and Troglodytes, and Æthiopians. \V And he tooke the
moſt fenſed cities in Iuda, and came ouer vnto Ieruſalem. \V And Semeias
the prophete went to Roboam, and to the princes of Iuda, that were
gathered together in Ieruſalem, fleing from Seſac, and he ſayd to them:
Thus ſayth our Lord: You haue leaft me, and I haue leaft you in the hand
of Seſac. \V And the princes of Iſrael, and the king, being aſtonied,
ſayd: Our Lord is iuſt. \V And when our Lord had ſeene that they were
humbled, the word of our Lord came to Semeias, ſaying: Becauſe they are
humbled, I wil not deſtroy them, and I wil geue them a litle ayde, and
my furie shal not droppe vpon Ieruſalem
%%% o-0816
by the hand of Seſac. \V But yet they shal ſerue him, that they may know
the diſtance of my ſeruice, and of the ſeruice of the kingdom of the
earth. \V Therfore Seſac the king of Ægypt retyred from Ieruſalem,
taking away the treaſures of the houſe of our Lord, and of the kinges
houſe, and he tooke al thinges with him, and the golden shieldes that
Salomon had made, \V for the which the king made braſen ones, and
deliuered them to the princes of the shieldbearers, which kept the
%%% 0909
entrance of the palace. \V And when the king entred into the houſe of
our Lord, the shieldbearers came, and tooke them, & brought them backe
agayne to their armorie. \V But yet becauſe they were humbled, the wrath
of our Lord was turned away from them, neither were they vtterly
deſtroyed: for in Iuda there were found good workes. \V King Roboam
therfore was ſtrengthened in Ieruſalem, & reigned: one and fourtie
yeares old was he when he beganne to reigne, and he reigned ſeuentene
yeares in Ieruſalem, the citie, which our Lord choſe, to confirme his
name there, out of al the tribes of Iſrael: and the name of his mother
was Naama an Ammonite. \V And he did euil, and prepared not his hart to
ſeke our Lord. \V But the workes of Roboam the firſt and the laſt are
writen in the Bookes of Semeias the Prophete, and of Addo the Seer, and
diligently expounded: and Roboam and Ieroboam fought one agaynſt the
other al their daies. \V And Roboam ſlept with his fathers, and was
buried in the citie of Dauid. And Abias his ſonne reigned for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
