%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 29
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-29


\project douay-rheims


%%% 0936
%%% o-0840
\startChapter[
  title={Chapter 29}
  ]

\Summary{Ezechias repayreth the Temple, and diuine ſeruice. 5.~Zelouſly
  exhorteth offenders to repentance. 12.~The Temple is purified in
  ſixtene daies. 18.~The king and nobles offer hoſtes, which the
  Prieſtes immolate, 25.~with ſolemne muſick (ordained by king Dauid)
  and great ioy of al the people.}

Therfore
\CNote{\XRef{4.~Reg.~18.}}
Ezechias began to reigne, when he was fiue and twentie yeares old, and
he reigned nine and twentie yeares in Ieruſalem: the name of his mother
was Abia, the daughter of Zacharias. \V And he did that which was
pleaſing in the ſight of our Lord, according to al thinges that Dauid
his father had done. \V He in the firſt yeare and moneth of his reigne
opened the doores of the houſe of our Lord, and repayred them. \V And he
brought the Prieſtes and the Leuites, and aſſembled them in the Eaſt
ſtreate. \V And he ſayd to them: Heare me ye Leuites, and be ſanctified,
cleanſe the houſe of our Lord the God of your fathers, and take away al
vncleannes out of the Sanctuarie. \V Our fathers haue ſinned and done
euil in the ſight of our Lord God, forſaking him: they haue turned away
their faces from the tabernacle of our Lord, and geuen the backe. \V
They haue ſhut the doores, that were in the porch, and put out the
lampes, and haue not burnt incenſe, and haue not offered holocauſtes in
the Sanctuarie of the God of Iſrael. \V Therfore was the furie of our
Lord ſtirred vp vpon Iuda and Ieruſalem, and he hath deliuered them into
commotion, and into deſtruction, & to be hiſſed at, as your ſelues ſee
with your eyes. \V Behold, our fathers haue fallen by the ſwordes, our
ſonnes, and our daughters, and wiues are led captiue for this
wickedneſſe. \V Now therfore it pleaſeth me that we make a couenant with
our Lord the God of Iſrael, and he wil turne away the furie of his wrath
from vs. \V My children be not negligent: our Lord hath choſen you to
ſtand before him, and to miniſter to him, and to worſhipe him, and to burne
incenſe to him. \V The Leuites therfore aroſe: Mahath the ſonne of
Amaſai, and Ioel the ſonne of Azarias, of the children of Caath:
moreouer the children of Merari, Cis the ſonne of Abdi, and Azarias the
ſonne of Ialaleel. And of the children of Gerſon, Ioah the ſonne of
Zemma, and Eden the ſonne of Ioah. \V
\Fix{Bu}{But}{obvious typo, fixed in other}
of the children of Eliſaphan, Samri, and Iahiel. Alſo of the children of
Aſaph, Zacharias, and Mathanias.
%%% 0937
\V Moreouer alſo of the children of Heman, Iahiel, & Semei: yea and of
the children of Idithun, Semeias, and Oziel. \V And they gathered
together their brethren, and were ſanctified, and went in according to
the commandment of the king, and the precept of our Lord, to purge the
houſe of God. \V The Prieſtes
%%% o-0841
alſo going into the Temple of our Lord to ſanctifie it, brought out al
vncleannes, which they found within the entrance of the houſe of our
Lord, which the Leuites tooke, and caried to the Torrent cedron
without. \V And they began to clenſe it the firſt day of the firſt
moneth, and in the eight day of the ſame moneth they entred into the
porche of the temple of our Lord, and they purged the temple in eight
dayes, and in the ſixtenth day of the ſame moneth, they accompliſhed
that which they began. \V They entered in alſo to Ezechias the king, and
ſayd to him: We haue ſanctified al the houſe of our Lord, and the altar
of holocauſt, and the veſſels therof, moreouer alſo the table of
propoſition with al the veſſels therof, \V and al the furniture of the
temple, which king Achaz in his reigne had polluted, after that he
tranſgreſſed; and behold al thinges are ſet forth before the altar of our
Lord. \V And Ezechias the king riſing early, aſſembled al the princes of
the citie, and went vp into the houſe of our Lord: \V and they offered
together ſeuen oxen, and ſeuen rammes, ſeuen lambes, and ſeuen
buckgoates for ſinne, for the kingdom, for the ſanctuarie, for Iuda, and
he ſayd to the Prieſtes the children of Aaron, that they ſhould offer
them vpon the altar of our Lord. \V They killed therfore the oxen, and
the Prieſtes tooke the bloud, and powred it vpon the altar, they killed
alſo the rammes, and their bloud they powred alſo vpon the altar, and
they immolated the lambes, and powred the bloud vpon the altar. \V They
brought the buckgoates for ſinne before the king, and the whole
multitude, and they put their handes vpon them: \V and the Prieſtes
immolated them, and ſprinkled their bloud on the altar for an expiation
of al Iſrael: for the king had commanded for al Iſrael, that holocauſt
ſhould be made, and for ſinne. \V He appoynted alſo the Leuites in the
houſe of our Lord with cymbals, and pſalteries, and harpes according to
the diſpoſition of Dauid the king, and of Gad the Seer, and of Nathan
the Prophete: for it was the precept of our Lord by the hand of his
prophetes. \V And the
%%% 0938
Leuites ſtoode, holding the inſtrumentes of Dauid, and the Prieſtes
trumpettes. \V And Ezechias commanded that they should offer holocauſtes
vpon the altar: and when holocauſtes were offered, they began to ſing
prayſes to our Lord, and to ſound with trumpettes, and on diuerſe
inſtrumentes, which Dauid the king of Iſrael had prepared for to
ſound. \V And al the multitude adoring, the ſinging men, and they that
held the trumpettes, were in their office, whiles the holocauſt was
accomplished. \V And when the oblation was ended, the king bowed, and al
that were with him, and adored. \V And Ezechias, and the princes
commanded the Leuites, that they should prayſe our Lord in the wordes of
Dauid, and Aſaph the Seer: who prayſed him with great ioy, and bowing
the knee adored. \V But Ezechias added theſe wordes alſo: You haue
filled your handes to our Lord, come, and offer victimes, and prayſes in
the houſe of our Lord. Al the multitude therfore offered hoſtes, and
prayſes, and holocauſtes with a deuout minde. \V Moreouer the
%%% o-0842
number of the holocauſtes, which the multitude offered, was
\Fix{thls,}{this,}{obvious typo, fixed in other}
oxen ſeuentie, rammes an hundred, lambes two hundred. \V And they
ſanctified to our Lord ſix
\Fix{hunered,}{hundred,}{obvious typo, fixed in other}
& sheepe three thouſand. \V But the Prieſtes were few, neither could they
ſuffiſe to draw of the ſkinnes of the holocauſtes: wherfore the Leuites
alſo their brethren holpe them, til the worke was accomplished, and the
Prieſtes were ſanctified, for the Leuites are ſanctified with an eaſier
rite, then the Prieſtes. \V There were holocauſtes therfore verie many,
the fatte of pacifiques, and the libamentes of the holocauſtes: and the
ſeruice of the houſe of our Lord was accomplished. \V And Ezechias
reioyſed, and al the people, becauſe the miniſterie of our Lord was
accomplished. For it pleaſed them that the thing should be done of a
ſoden.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
