%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 24
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-24


\project douay-rheims


%%% 0926
%%% o-0832
\startChapter[
  title={Chapter 24}
  ]

\Summary{Ioas reigning piouſly, ſo long as Ioiada liueth, cauſeth the
  Temple to be repayred, 14.~and new ſacred veſsels to be
  made. 15.~Ioiada an hundred thirtie
  %%% 0927
  yeares old dieth. 17.~Ioas falleth to idolatrie: 20.~cauſeth Zacharias
  to be ſlayne in the court of the Temple. 23.~A few Syrians kil the
  chief men about the king, and carie away great prayes. 25.~Ioas is
  ſlayne by his owne men, and his ſonne Amaſias reigneth.}

Seuen yeares old was Ioas when he began to reigne: and he reigned fourty
yeares in Ieruſalem, the name of his mother was Sebia of Berſabee. \V
And he did that which is good before our Lord al the dayes of Ioiada the
Prieſt. \V And Ioiada tooke for him two wiues, of whom he begat ſonnes &
daughters. \V After which thinges it pleaſed Ioas to repayre the houſe
of our Lord. And he aſſembled the Prieſtes, and the Leuites, and ſayd to
them: Goe ye forth to the cities of Iuda, and gather of al Iſrael money
for the reparation of the temple of your God, yeare by yeare, and doe
this in haſt: moreouer the Leuites did negligently. \V And the king
called Ioiada the prince, and ſayd to him: Why haſt thou had no care to
conſtrayne the Leuites to bring in out of Iuda and Ieruſalem the money,
that was
\SNote{By the law euerie one payed yearly halfe a ſicle, towards the
repayring of the tabernacle and ſo afterwardes of the temple.
\XRef{Ex.~30.}}
appointed of Moyſes the ſeruant of our Lord, that al the multitude of
Iſrael ſhould bring it in into the tabernacle of teſtimonie? \V For the
moſt impious Athalia, and her children haue deſtroyed the houſe of God,
and of al thinges that had bene ſanctified in the temple of our Lord,
they adorned the temple of Baalim. \V The king therfore commanded and
they made a cheſt: and ſet it by the gate of our Lord on the out
ſide. \V And it was proclaymed in Iuda and Ieruſalem, that euery man
should bring the price to our Lord, which Moyſes the ſeruant of God
appoynted ouer al Iſrael, in the deſert. \V And al the princes reioyſed,
and al the people: and going in they contributed into the cheſt of our
Lord, and caſt in ſo that it was filled. \V And when it was time that
they should bring the cheſt before the king by the handes of Leuites
(for they ſaw much money) the kinges Scribe went in, and he whom the
high prieſt had appoynted: & they powred out the money that was in the
cheſt, & recaried it to his place: and ſo did they from day to day, and
there was gathered infinite money. \V Which the king and Ioiada gaue to
them, that ouerſaw the workes of the houſe of our Lord: but they hired
with it hewers of ſtones, and artificers of al workes, to repayre the
houſe of our Lord: ſmithes alſo of yron and braſſe, that that
%%% o-0833
which began to fal, might be vpholden. \V And
%%% 0928
they that wrought did induſtriouſly, and the breach of the walles was
cloſed by their handes, and they rayſed the houſe of our Lord into the
old ſtate, and made it ſtand firmely. \V And when they had accompliſhed
al the workes, they brought the reſt of the money before the king and
Ioiada: of the which were made veſſels of the temple to the miniſterie,
and for holocauſtes, phials alſo, and other veſſels of gold and ſiluer:
and holocauſtes were offered in the houſe of our Lord continually al the
daies of Ioiada. \V But Ioiada became old being ful of dayes, and died
when he was an hundred and thirtie yeares old. \V And they buried him in
the citie of Dauid with the kinges, becauſe he had done good with
Iſrael, and with his houſe. \V And after that Ioiada was dead, the
princes of Iuda went in, and adored the king, who being altered by their
ſeruiceableneſſe, agreed to them. \V And they forſooke the temple of our
Lord the God of their fathers, and ſerued groues, and ſculptilles, and
there came wrath agaynſt Iuda, and Ieruſalem for this ſinne. \V And he
ſent them prophetes, that they ſhould returne to our Lord, whom
proteſting they would not heare. \V The ſpirit of God therfore inueſted
Zacharias the ſonne of Ioiada the Prieſt, & he ſtood in the ſight of the
people, and ſayd to them: Thus ſayth our Lord God: Why tranſgreſſe you
the precept of our Lord, which thing ſhal not profit you, & haue
forſaken our Lord, that he ſhould forſake you? \V Who being gathered
agaynſt him, they threw ſtones according to the kinges commandement, in
the court of the houſe of our Lord. \V
\CNote{\XRef{Mat.~23.}}
And Ioas the king did not remember the mercie, that Ioiada his father
had done with him, but he killed his ſonne. Who when he died, ſayd: Our
Lord ſee, and requite it. \V And when a yeare was come about, the armie
of Syria came vp againſt him: & it came into Iuda & Ieruſalem, & ſlewe
al the princes of the people, and al the pray they ſent to the king into
Damaſcus. \V And wheras there was come a very ſmal number of the
Syrians, our Lord deliuered into their handes an infinit multitude, for
that they had forſaken our Lord the God of their fathers: on Ioas alſo
they exerciſed ignomious iudgementes. \V And departing they left him in
great diſeaſes: and his ſeruantes roſe agaynſt him,
\SNote{He that killed his ſpiritual father was ſlaine by his owne
ſeruantes.}
for reuenge of the bloud of the ſonne of Ioiada the prieſt, & they ſlewe
him in his bed, & he dyed: and they buried him in the Citie of Dauid,
but not in the kinges ſepulchres. \V And
%%% 0929
there conſpired againſt him Zabad the ſonne of Semmaath an Ammoniteſſe,
& Iozabad the ſonne of Semarith a Moabiteſſe. \V Moreouer his children,
and the ſumme of money, which was gathered vnder him, & the repayring of
the houſe of God are writen more diligently in the Booke of kinges: and
Amaſias his ſonne reigned for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
