%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 28
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-28


\project douay-rheims


%%% 0933
%%% o-0838
\startChapter[
  title={Chapter 28}
  ]

\Summary{For his great wickednes, Achaz is taken in battel, his countrie
  ſpoyled, and manie ſlayne by the kinges of Syria, and Iſrael. 9.~Yet
  God ſuffereth not the kingdom to be ſubdued. 16.~Then requiring helpe
  of the Aſsyrians, 17.~is ſpoyled by the Idumeans, Philiſthians, and
  Aſsyrians. 22.~After al which plagues he committeth more idolatrie,
  26.~dieth, and his ſonne Ezechias reigneth.}

%%% 0934
Twentie
\CNote{\XRef{4.~Reg.~16.}}
yeares old was Achaz when he began to reigne, & he reigned ſixtene
yeares in Ieruſalem: he did not right in the ſight of our Lord as Dauid
his father: \V but walked in the wayes of the kinges of Iſrael, moreouer
alſo he did caſt ſtatues to Baalim. \V He it is that burnt incenſe to
the Valebenennom, and he conſecrated his ſonnes in fire according to the
rite of the nations, which our Lord ſlewe in the coming of the children
of Iſrael. \V He
\Fix{ſacrified}{ſacrificed}{obvious typo, fixed in other}
alſo, & burnt incenſe in the excelſes, & on
hilles, and vnder euerie tree ful of grene leaues. \V And our Lord his
God deliuered him into the handes of the king of Syria, who ſtroke him,
and tooke a great praye out of his kingdom, & brought into Damaſcus: to
the handes alſo of the king of Iſrael was he deliuered, and ſtriken with
a great plague. \V And Phacee the ſonne of Romelia ſlewe of Iuda an
hundred twentie thouſand in one day, al men of warre: for that they had
forſaken our Lord the God of their fathers. \V At that time Zechri a
mightie man of Ephraim, ſlewe Maaſias the kinges ſonne, and Ezricam the
gouernour of his houſe, Elcana alſo ſecond from the king. \V And the
children of Iſrael tooke of their brethren two hundred thouſand of wemen, of
boyes, and of wenches, and an infinite praye: and they brought it into
Samaria. \V At that time there was a Prophete of our Lord, named Oded:
who going forth to meete the armie coming into Samaria, ſayd to them:
Behold our Lord the God of your fathers being angrie againſt Iuda, hath
deliuered them into your handes, and you haue ſlayne them cruelly, ſo
that your crueltie did reach to heauen. \V Moreouer the children of Iuda
and Ieruſalem you wil ſubdue vnto you for bondmen and bondwemen, which
nedeth not to be done: for you haue ſinned hereupon to our Lord your
God. \V But heare ye my
%%% o-0839
counſel, and carie backe the captiues, that you haue brought of your
brethren, becauſe the great furie of our Lord hangeth ouer you. \V There
ſtood therfore princes of the children of Ephraim, Azarias the ſonne of
Iohanan, Barachias the ſonne of Moſollomoth, Ezechias the ſonne of
Sellum, and Amaſa the ſonne of Hadali, againſt them that came out of the
battel, \V and they ſayd to them: You ſhal not bring in the captiues
hither, leſt we ſinne to our Lord. Why wil you adde vpon our ſinnes, and
heape vp old offences? for it is a great ſinne, and the anger of the
furie of our Lord hangeth ouer Iſrael.
%%% 0935
\V And the men of warre diſmiſt the pray, & al the thinges that they had
taken, before the princes and al the multitude. \V And the men, whom we
mentioned aboue, ſtood and taking the captiues, and al that were naked
they clothed and ſhod them, with the ſpoyles: and when they had clothed
and refreſhed them with meate and drinke, and annoynted them becauſe of
their labour, and had looked carefully to them: as manie as could not
walke, and were of a weake bodie, they ſet on beaſtes, and brought them
to Iericho the Citie of palme trees to their brethren, and themſelues
returned into Samaria. \V At that time king Achaz ſent to the king of
the Aſſyrians asking helpe. And the Idumeans came and ſtroke manie of
Iuda, and tooke a great praye. \V The Philiſthijms alſo were ſpred
abroad by the cities of the champayne, and toward the South of Iuda: and
they tooke Bethſames, & Aialon, and Gaderoth, Socho alſo, & Thamnan, and
Gamzo, with their villages, and dwelt in them. \V For our Lord had
humbled Iuda becauſe of Achaz the king of Iuda, for that he had made it
naked of helpe, and had contemned our Lord. \V And he brought agaynſt
him Theglathphalnaſar the king of the Aſſyrians, who alſo afflicted him,
and ſpoyled him no man reſiſting. \V Therfore Achaz ſpoyling the houſe
of our Lord, and the houſe of the kinges, and of the princes gaue giftes
to the king of the Aſſyrians, and yet it did nothing profite him. \V
Moreouer alſo in the time of his diſtreſſe he increaſed contempte
agaynſt our Lord, king Achaz himſelf by himſelf, \V immolated victimes to the
goddes of Damaſcus that ſtroke him, and ſayd: The goddes of the kinges
of Syria doe helpe them, whom I wil pacifie with hoſtes, and they wil
ayde me,
%%% !!! SNote unmarked in both
\SNote{VVicked policie auaileth nothing, but hurteth much}
wheras on the contrarie part they were his ruine, and al Iſraels. \V
Achaz therfore hauing ſpoyled al the veſſels of the houſe of God, and
broken them ſhut the gates of the temple of God, and made him altars in
al the corners of Ieruſalem. \V In al the cities alſo of Iuda he built
altars to burne frankincenſe, and he prouoked to wrath our Lord the God
of his fathers. \V But the reſt of his wordes, al his workes the firſt
and the laſt are writen in the Booke of the kinges of Iuda and
Iſrael. \V And Achaz ſlept with his fathers, and they buried him in the
Citie of Ieruſalem: for they receiued him not into the ſepulchres of the
kinges of Iſrael. And Ezechias his ſonne reigned for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
