%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 0895
%%% o-0804
\startChapter[
  title={Chapter 5}
  ]

\Summary{Manie giftes are offered. 4.~The Arke is brought with great
  ſolemnitie into the Temple. 6.~Innumerable hoſtes are offered,
  11.~with excellent muſick, & the Temple is replinished with the glorie
  of God.}

Salomon therfore brought in al the thinges, that Dauid his father had
vowed, the ſiluer, and gold, and al the veſſels he put in the treaſures
of the houſe of God. \V
\CNote{\XRef{3.~Reg.~7.}}
After which thinges he gathered together al the ancientes of Iſrael, and
al the princes of the tribes, and the heades of families, of the
%%% 0896
children of Iſrael into Ieruſalem, to bring
\SNote{An other Arke was not made, becauſe that was moſt holie & moſt
excellent, which Moyſes made. Likewiſe other holie thinges of the
tabernacle were brought into the Temple with great ſolemnitie.}
the Arke of the couenant of our Lord from the Citie of Dauid, which is
Sion. \V There came therfore vnto the king al the men of Iſrael in the
ſolemne day of the ſeuenth moneth. \V And when al the ancientes of
Iſrael were come, the Leuites caried the Arke, \V and brought it in, and
al the furniture of the tabernacle. Moreouer the Prieſtes, with the
Leuites did carrie the veſſels of the Sanctuarie, which were in the
tabernacle. \V And king Salomon, and al the aſſemblie of Iſrael, and al
that were gathered before the Arke, immolated rammes, and oxen
%%% o-0805
without anie number: for ſo great was the multitude of victimes. \V And
the prieſtes brought in the Arke of the couenant of our Lord into his
place, that is, to the oracle of the temple, into \L{Sancta ſanctorum}
vnder the winges of the cherubs: \V ſo that the cherubs ſpred their
winges ouer the place, wherin the Arke was ſet, and couered the Arke it
ſelfe with his barres. \V And the heades of the barres, wherwith the
Arke was caried, becauſe they were a litle longer, appeared before the
oracle: but if a man had beene a litle outward, he could not ſee
them. The Arke therfore was there vntil this preſent day. \V And there
was nothing in the Arke, but the two tables, which Moyſes had put in
Horeb, when our Lord gaue the law to the children of Iſrael coming out
of Ægypt. \V And the Prieſtes being gone out of the Sanctuarie (for al
the prieſtes that could be found there, were ſanctified, neither as yet
at that time were the courſes, and the order of the miniſteries deuided among
them) \V as wel the Leuites as the ſinging men, that is, both they which
were vnder Aſaph, and they which were vnder Heman, and they which were
vnder Idithun, their ſonnes, & brethren reueſted with
\Fix{finne}{fine}{likely typo, fixed in other}
linnen clothes, ſounded on cymbals, and pſalteries, and harpes, ſtanding
at the Eaſt ſide of the Altar, and with them Prieſtes an hundred
twentie, ſounding with trumpettes. \V Therfore al ſounding together, both
with trumpettes, and voyce, and cymbals, and organes, and with diuerſe
kind of muſical inſtrumentes, and
\Fix{lifteng}{lifting}{obvious typo, fixed in other}
vp their voice on high: the ſound was heard far of, ſo that when they
began to prayſe our Lord, and to ſay: Confeſſe to our Lord becauſe he is
good, becauſe his mercie is for euer: the houſe of God was filled with a
cloud, \V that the Prieſtes could not ſtand and miniſter for the
darkeneſſe. For the glorie of our Lord had filled the houſe of God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
