%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 0909
%%% o-0816
\startChapter[
  title={Chapter 13}
  ]

\Summary{Abias maketh warre agaynſt Ieroboam, 4.~exhorteth the people of
  Iſrael for iuſtice, and religions ſake to returne to him. 13.~In the
  meane time Ieroboam inuironeth him with forces, but by Gods aſsiſtance
  Abias preuaileth, 21.~and reigneth ſecurely.}

In
\CNote{\XRef{3.~Reg.~15.}}
the eightenth yeare of king Ieroboam, reigned Abias ouer Iuda. \V Three
yeares reigned he in Ieruſalem, and his mothers name was Michaia, the
daughter of Vriel of Gabaa: there was warre betwen Abias and
Ieroboam. \V And when Abias had begun battel, and had moſt warlike
men, & of choſen ones four hundreth thouſand: Ieroboam put his armie in
aray on the contrarie ſide, eight hundreth thouſand men, who them ſelues
alſo were choſen men, and moſt valiant to battels. \V Abias therfore
ſtood vpon mount Semeron, which was in Ephraim, and ſayd: Heare
Ieroboam, and al Iſrael: \V Are you ignorant that our Lord the God of
Iſrael gaue the kingdom to Dauid ouer Iſrael for euer, to him and his
children as a
\SNote{A firme and perpetual couenant.
\XRef{Num.~18. v.~19.}}
couenant of ſalt? \V And there
%%% o-0817
roſe vp Ieroboam the ſonne of Nabat, the ſeruant of Salomon the ſonne of
Dauid: and rebelled agaynſt his lord. \V And there were gathered to him
al the moſt vayne men, and the children of Belial: and they preuayled
agaynſt Roboam the ſonne of
%%% 0910
Salomon: moreouer Roboam was rude, and of a fearful hart, and could not
reſiſt them. \V Now therfore you ſay that you are able to reſiſt the
kingdom of our Lord, which he poſſeſſeth by the children of Dauid, and
you haue a great multitude of people, and golden calues, which Ieroboam
hath made you for goddes. \V And you haue caſt out the Prieſtes
\Fix{of of}{of}{obvious typo, fixed in other}
our Lord, the children of Aaron, and the Leuites: and you haue made you
prieſtes, as al the peoples of the earth: who ſoeuer ſhal come &
conſecrate his hand in a bullock of oxen, and in ſeuen rammes, is made
the prieſt of them that are not goddes. \V But our Lord is God, whom we
forſake not, and the Prieſtes do miniſter to our Lord of the children of
Aaron, and the Leuites are in their order. \V Holocauſtes alſo they do offer
to our Lord, euerie day morning and euening, and incenſe made according
to the preceptes of the law, and the loaues are ſet forth on a moſt
cleane table, and there is with vs the golden candleſticke, and the
lampes therof, that they may be lighted alwaies at euening: for we keepe
the preceptes of the Lord our God, whom you haue forſaken. \V Therfore
in our hoſt God is the prince, and his Prieſtes, which ſound with
trumpettes, and reſound agaynſt you: children of Iſrael fight not
agaynſt our Lord the God of your fathers, becauſe it is not expedient
for you. \V He ſpeaking theſe thinges, Ieroboam endeuoured to entrappe
him behind. And when he ſtood ouer agaynſt the enemies, he compaſſed
Iuda vnwitting with his armie. \V And Iudas looking backe, ſaw the
battel at hand before and behind, and cryed to our Lord: and the
Prieſtes begane to ſound with trumpettes. \V And al the men of Iuda made
a ſhout: and behold they crying, God terrified Ieroboam, and al Iſrael
that ſtood agaynſt Abias and Iuda. \V And the children of Iſrael fled
from Iuda, and our Lord deliuered them into their hand. \V Abias
therfore and his people ſtroke them with a great ſlaughter, and there
fel wounded of Iſrael fiue hundreth thouſand valiant men. \V And the
children of Iſrael were humbled, at that time, and the children of Iuda
excedingly encouraged becauſe they had truſted in our Lord the God of their
fathers. \V And Abias purſewed Ieroboam fleing, and he tooke his cities,
Bethel and her daughters, and Ieſana with her daughters, Ephron alſo and
her daughters. \V Neither was Ieroboam able to reſiſt anie more in the
daies of Abias: whom our Lord ſtroke, and he died.
%%% 0911
\V Therfore Abias, his empire being ſtrengthened, tooke fourtene wiues:
and he begat two and twentie ſonnes, and ſixtene daughters. \V But the
reſt of the wordes of Abias, and of his waies and wordes, are writen
diligently in the Booke of Addo the Prophete.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
