%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 0900
%%% o-0808
\startChapter[
  title={Chapter 7}
  ]

\Summary{Fire from heauen deuoureth the holocauſtes, and glorie
  replenisheth the Temple. 5.~More ſacrifices are offered, 8.~and the
  Temple is dedicated, with ſeuen daies ſolemnitie, the eight day
  collection is made. 12.~God ſignifieth that he hath heard Salomons
  prayer, 17.~conditionally (as it was made) if they ſerue him,
  19.~otherwiſe he wil punish them.}

And when Salomon had finished to poure out his prayers, fyre deſcended
from heauen, and deuoured the holocauſtes and victimes: and the maieſtie
of our Lord filled the houſe. \V Neither could the Prieſtes enter into
the Temple of our Lord, becauſe the maieſtie of our Lord had filled the
temple of our Lord. \V Yea and al the children of Iſrael ſawe the fyre
deſcending, and the glorie of our Lord vpon the houſe: and falling flat
on the earth vpon the pauement paued with ſtone, they adored, and
prayſed our Lord: Becauſe he is good, becauſe his mercie is for euer. \V
And the king and al the people immolated victimes before our Lord. \V
King Salomon therfore killed hoſtes, of oxen twentie two thouſand, of rammes
an hundred twentie thouſand: and the king and al the people dedicated
the houſe of God. \V And the Prieſtes ſtood in their offices: and the
Leuites with the inſtrumentes of the ſongues of our Lord, which Dauid
the king made to prayſe our Lord: Becauſe his mercie is for euer,
ſinging
\SNote{As the Temple was the ſpecial place of prayer, choſen by God:
\XRef{v.~12.}
ſo ſpecial
\Fix{Hymmes}{Hymnes}{likely typo, fixed in other}
Pſalmes and other ſet formes of bleſſing are more gratful to God, and
more effectual to his ſeruantes.}
the hymnes of Dauid by their handes: moreouer the Prieſtes ſounded with
trumpettes before them, and al Iſrael ſtood. \V Salomon alſo ſanctified
the middes of the court before the temple of our Lord: for he had
offered there the holocauſtes, and the fatte of the pacifiques: becauſe
the braſen altar, which he had made, could not ſuſteyne the holocauſtes
and the ſacrifices and the fatte. \V Salomon therfore made a ſolemnitie
at that time ſeuen daies, and al Iſrael with him, an aſſemblie verie
great, from the entrance of Emath to the Torrent of Ægypt. \V And he
made in the eight day an aſſemblie, becauſe he had dedicated the altar
ſeuen daies, and had celebrated the ſolemnitie ſeuen daies. \V Therfore
in the three and twenteth day of the ſeuenth 
moneth, he diſmiſt the people to their tabernacles,
%%% 0901
reioyſing and being glad for the good, that our Lord had done to
%%% o-0809
Dauid, and Salomon, and Iſrael his people. \V And Salomon accomplished
the houſe of our Lord, and the houſe of the king, and al that he had
diſpoſed in his hart to do, in the houſe of our Lord, and in his owne
houſe, and he proſpered. \V And our Lord appeared to him by night, and
ſayd: I haue heard thy prayer, and haue choſen this place to me for a
houſe of ſacrifice. \V If I ſhal shut heauen, and rayne fal not, and
shal bid and command the locuſt to deuoure the land, and ſhal ſend
peſtilence into my people: \V and my people being conuerted, 
vpon whom my name is inuocated, ſhal beſech me, and ſeeke out my face,
and shal doe penance from their moſt wicked waies: I alſo wil heare from
heauen, and wil be propitious to their ſinnes, and wil ſaue their
land. \V Mine eies alſo shal be opened, and mine eares erected to his
prayer, that shal pray in this place. \V For I haue choſen, and haue
ſanctified this place, that my name may be there for euer, and mine eies
and my hart may remaine there al daies. \V Thou alſo if thou walke
before me, as Dauid thy father walked, and shalt doe according to al
thinges, which I haue commanded thee, and shalt keepe my iuſtices and
iudgementes: \V I wil rayſe vp the throne of thy kingdom, as I promiſed
to Dauid thy father, ſaying: There shal not be taken away of thy ſtocke
a man, that shal be prince in Iſrael. \V But if you shal be turned away,
and shal forſake my iuſtices, and my preceptes, which I haue propoſed to
you, and going shal ſerue ſtrange goddes, and adore them, \V I wil
plucke you out of my land, which I haue geuen you: and this houſe, which
I haue ſanctified to my name, I wil caſt away from my face, and wil
deliuer it for a parable, and an exemple to al peoples. \V And this
houſe shal be for a prouerbe to al paſſengers, and they shal ſay being
aſtonyshed: Why hath the Lord done ſo to this land, and to this
houſe? \V And they shal anſwer: Becauſe they forſooke the Lord the God
of their fathers, who brought them out of the Land of Ægypt, and tooke
hold of ſtrange goddes, and adored them, and worshipped them: therfore
are al theſe euils come vpon them.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
