%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 0903
%%% o-0811
\startChapter[
  title={Chapter 9}
  ]

\Summary{The Queene of Saba admireth Salomons wiſdom, 9.~geueth, and
  receiueth preſents. 13.~Of the great ſtore of treaſure which is
  yearely brought in, 13.~king Salomon maketh pretious armour, 17.~a
  throne, 20.~and plate. 24.~Other kinges ſend him giftes, ſo he
  abundeth in glorie, and riches. 29.~After fourtie yeares reigne he
  dieth, and his ſonne Roboam ſuccedeth.}

The
\CNote{\XRef{1.~Reg.~10.}}
queene of Saba alſo, when ſhe had heard the fame of Salomon, came to
proue him in hard propoſitions in Ieruſalem, with great riches, and
camels, which caried ſpices, and very much gold, and pretious
ſtones. And when ſhe was come to Salomon, ſhe ſpake to him what thinges
ſoeuer were in her hart. \V And Salomon expounded to her al thinges that
ſhe propounded: neither was there anie thing, that he made not playne
vnto her. \V Who after she ſaw, to witte, the wiſdom of Salomon, and the
houſe which he had built, \V moreouer alſo the meates of his table, and
the habitations of his ſeruantes, and the offices of his miniſters, and
their garmentes, the cupbearers alſo, and their garmentes, and the
victimes, which he immolated in the houſe of our Lord: there was no
ſpirit in her anie longer, she was ſoe aſtonied. \V And she ſayd to the
king: The word is true, which I heard in my countrie of thy vertues and
wiſdom. \V I did not beleue them that told it, vntil my ſelf was come,
and mine eies had ſeene, and I had proued ſcarce the half part of thy
wiſedom to haue beene told me: thou haſt paſſed the ſame with thy
vertues. \V Bleſſed are thy men, and bleſſed are thy ſeruantes, which
aſſiſt before thee at al time, and heare thy wiſedom. \V Be the Lord thy
God bleſſed, who
%%% 0904
would ordayne thee ouer his throne, king
\SNote{They are called kinges of God, which reigne by his grace and
according to his wil. VVherevpon they vſe this ſtile: \Emph{By the grace
of God, K.~of England, Ieruſalem, &c.}}
of the Lord thy God. Becauſe God loueth Iſrael, and wil preſerue it for
euer: therfore hath he ſette thee king ouer it, to doe iudgementes and
iuſtice. \V And she gaue to the king an hundred twentie talentes of
gold, and ſpices exceding much, and
%%% o-0812
moſt pretious ſtones: there were not ſuch ſpices, as theſe, which the
Queene of Saba gaue to King Salomon. \V But the ſeruantes of Hiram alſo
with the ſeruantes of Salomon brought gold from Ophir, and Thymtrees,
and moſt pretious ſtones: \V wherof the king made, to witte, of the
Thymtrees, ſtayers in the houſe of our Lord, and in the kinges houſe,
harpes alſo and pſalteries for the ſinging men: neuer were there ſeene
ſuch trees in the Land of Iuda. \V And king Salomon gaue to the Queene
of Saba al thinges that ſhe would, and that she asked, and manie moe
thinges then she brought to him: who returning, went into her countrie
with her ſeruantes. \V And the weight of gold, that was brought to
Salomon euerie yeare was ſix hundred ſixtie ſix talentes of gold: \V
beſide that ſumme, which the legates of diuers nations, and the
merchantes were accuſtomed to bring, and al the kinges of Arabia, and
the Dukes of the landes, which brought gold and ſiluer to Salomon. \V
King Salomon therfore made two hundred golden ſpeares, of the ſumme of
ſix hundred peces of gold, which were ſpent in euerie ſpeare: \V alſo
three hundred golden ſhieldes of three hundred peces of gold, with which
euerie shield was couered: and the king put them in the armarie, which
was beſette with a wood. \V The king alſo made a great throne of iuorie,
and couered it with moſt fyne gold. \V Six ſteppes alſo, wherwith the
going vp was to the throne, and a foote ſtoole of gold, and two litle
armes on either ſide, and two lions ſtanding by the litle armes, \V yea
and other twelue litle lions ſtanding vpon the ſteppes on both ſides:
there was not ſuch a throne in al kingdomes. \V Al the veſſels alſo of
the kinges table were of gold, and the veſſels of the houſe of the
foreſt of Libanus of moſt pure gold. For ſiluer in thoſe daies was
reputed for nothing. \V For the kinges shippes went into Tharſis with
the ſeruantes of Hiram, once in three yeares: and they brought from
thence gold and ſiluer, and yuorie, and apes, and pecockes. \V Salomon
therfore was magnified aboue al the kinges of the earth for riches and
glorie.
%%% 0905
\V And al the kinges of the earth deſired to ſee Salomons face, that
they might heare the wiſedom, which God had geuen in his hart. \V And
they brought him giftes, veſſels of ſiluer, and gold, and garmentes, and
armour, and ſpices, horſes, and mules, euerie yeare. \V Salomon alſo had
fourtie thouſand horſes in the ſtables, and of chariotes, and horſemen,
twelue thouſand, and he placed them in the cities of the chariotes, and
where the king was in Ieruſalem. \V He exerciſed alſo authoritie ouer al
the kinges from the riuer Euphrates vnto the land of the Philiſthines,
and vnto the borders of Ægypt. \V And he made ſo great plentie of ſiluer
in Ieruſalem as it were of ſtones: and of cedres ſo great a multitude as
of ſicomores, which grow in the champayne. \V And horſes were brought
him out of Ægypt, and al countries. \V But the reſt
\Fix{of of}{of}{obvious typo, fixed in other}
the workes of
Salomon the firſt and laſt are writen in the wordes of Nathan the
Prophet, and in the bookes of Ahias the Silonite, in the Viſion
%%% o-0813
alſo of Addo the Seer,
\Fix{agayaſt}{agaynſt}{obvious typo, fixed in other}
Ieroboam the ſonne of Nabat. \V And Salomon reigned in Ieruſalem ouer al
Iſrael fourtie yeares. \V And he
\SNote{His fall to luxurie and idolatrie is recorded,
\XRef{3.~Reg.~11.}}
ſlept with his fathers: and they buried him in the citie of Dauid: and
Roboam his ſonne reigned for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
