%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 34
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-34


\project douay-rheims


%%% 0947
%%% o-0850
\startChapter[
  title={Chapter 34}
  ]

\Summary{Ioſias deſtroyeth idolatrie, 8.~repaireth the Temple. 14.~The
  booke of the law found in the Temple is redde before him. 19.~Wherby
  vnderſtanding the ſinnes of the people, 23.~a prophete foresheweth
  their punishment, 26.~but not in his daies. 29.~He reneweth the
  couenant betwen God and the people.}

Eight
\CNote{\XRef{4.~Reg.~22.}}
yeares old was Ioſias when he began to reigne, & he reigned thirtie and
one yeares in Ieruſalem. \V And he did that which was right in the ſight
of our Lord, and walked in the waies of Dauid his father: he declined
not neither to the right hand, nor to the left. \V And in the eight
yeare of his reigne, when he was yet a child, he began to ſeeke the God
of his father Dauid: and the twelfth yeare after he began to reigne, he
cleanſed Iuda and Ieruſalem from excelſes, and groues, and idols, &
ſculptils. \V And they deſtroyed before him the altars of Baalim, and
the idols, that had bene ſet vpon them, they deſtroyed: the groues alſo
and ſculptils he cut downe and brake in peeces: and ouer their tombes,
that were accuſtomed to immolate vnto them, he ſtrawed the
fragmentes. \V
\CNote{\XRef{3.~Reg.~12.}}
Moreouer the bones of the Prieſtes he burnt on the altars of the idols,
and he clenſed Iuda and Ieruſalem. \V Yea and in the cities of
Manaſſes, and of Ephraim, and of Simeon, vnto Nephthali he ouerthrew
al. \V And when he had deſtroyed the
altars, and the groues, and had broken the idols into peces, and had
ouerthrowen al profane temples out of al the land of Iſrael, he returned
into Ieruſalem. \V Therfore in the eightenth yeare of his reigne, the
land now clenſed, and the temple of our Lord, he ſent Saphan the ſonne
of Eſelias, and Maaſias the gouernour of the citie, and Ioha the ſonne
of Ioachaz Commenter, that they ſhould repayre the houſe of our Lord his
God. \V Who came to Helcias the high prieſt: and taking of him the
money, which had bene brought into the houſe of our Lord, and which the
Leuites and porters had gathered together of Manaſſes, and
%%% 0948
Ephraim, and of al the remnant of Iſrael, of al Iuda alſo, and Beniamin,
and the inhabiters of Ieruſalem, \V they deliuered in their handes, that
ouerſaw the workemen in the houſe of our Lord, that they ſhould repayre
the temple, and mend al weake partes. \V But they gaue it to the
artificers, and to the maſons, that they ſhould bye ſtones out of the
quarries, and timber for the ioyntures of the building, and for making
the roofe of the houſe, which the kinges of Iuda had deſtroyed. \V Who
did al thinges faithfully. And the ouerſeers of the workemen were Iahath
& Abdias of the children of Merari, Zacharias and Moſollam of the
children of Caath, which vrged the worke: al Leuites ſkilful to ſing on
%%% o-0851
inſtrumentes. \V But ouer them, that caried burdens to diuerſe vſes, were
ſcribes, and maſters of the Leuites, and porters. \V And when they
caried forth the money, that had bene brought into the temple of our
Lord, Helcias the Prieſt found the Booke of the law of our Lord,
\SNote{Geuen by the hand of Moyſes.}
by the hand of Moyſes. \V And he ſayd to Saphan the Scribe: I haue found
the booke of the law in the houſe of our Lord: and deliuered it to
him. \V But he brought in the volume to the king, and told him, ſaying:
Al thinges, which thou gaueſt into the hand of thy ſeruantes, loe are
accomplished. \V The ſiluer that was found in the houſe of our Lord,
they haue gathered into a maſſe: and it was geuen to the ouerſeers of
the artificers, and of them that make diuerſe workes. \V Moreouer
Helcias the Prieſt deliuered me this booke. Which when he had read in
the kinges preſence, \V and he had heard the wordes of the law, he rent
his garmentes: \V and he commanded Helcias, and Ahicam the ſonne of
Saphan, and Abdon the ſonne of Micha, Saphan alſo the ſcribe, and Aſaa
the kinges ſeruant, ſaying: \V Goe, and pray our Lord for me, and for
the remnant of Iſrael, and Iuda, concerning al the wordes of this Booke,
which is found: for the great furie of our Lord hath diſtilled vpon vs,
for that our fathers haue not kept the wordes of our Lord, to doe al
thinges, that are writen in this volume. \V Helcias therfore went, and
they that were ſent together by the king, to Olda a propheteſſe, the
wife of Sellum the ſonne of Thecuath, the ſonne of Haſra keeper of the
garmentes: who dwelt in Ieruſalem in the ſecond part: and they ſpake to
her the wordes, which we haue aboue mentioned. \V But ſhe anſwered them: Thus
ſayth our Lord the God of Iſrael: Tel
%%% 0949
the man, that ſent you to me: \V Thus ſayth our Lord: Behold I wil bring
euils vpon this place, and vpon the inhabitantes therof, and al the
curſes, that are writen in this Booke, which they redde before the king
of Iuda. \V Becauſe they haue forſaken me, and haue ſacrificed to
ſtrange goddes, that they might prouoke me to wrath in al the workes of
their handes, therfore ſhal my furie droppe vpon this place, & ſhal not
be extinguished. \V But to the king of Iuda that ſent you to beſeche our
Lord, thus ſpeake ye: Thus ſayth our Lord the God of Iſrael: Becauſe
thou haſt heard the wordes of this volume, \V and thy hart is mollified,
and thou art humbled in the ſight of our Lord for theſe thinges, which
are ſpoken againſt this place, and the inhabitantes of Ieruſalem, and
reuerencing my face, haſt rent thy garmentes, and wept before me: I alſo
haue heard thee, ſayth our Lord. \V
\SNote{It is a benefite to be taken out of this world, before general
miſerie come vpon the people.}
For now wil I gather thee to thy fathers, and thou ſhalt be brought into
thy graue in peace: neither shal thyne eies ſee al the euil that I wil
bring in vpon this place, and vpon the inhabitantes therof. They therfore
reported to the king al thinges that ſhe had ſayd. \V But he calling
together al the ancientes of Iuda and Ieruſalem, \V went vp into the
houſe of our Lord, and al the men of Iuda, and the inhabitantes of
Ieruſalem, Prieſtes and Leuites, and al the people from the leaſt to the
greateſt. In whoſe hearing in the houſe of our Lord, the
%%% o-0852
king read al the wordes of the volume. \V And ſtanding on his tribunal
ſeate, he made a couenant before our Lord, that he would walke after
him, and keepe his preceptes, and teſtimonies, and iuſtifications in al
his hart, and in al his ſoule, and would do the thinges that were writen
in that volume, which he had read. \V He adiured alſo vpon this al that
were found in Ieruſalem, and Beniamin: and the inhabitantes of Ieruſalem
did according to the couenant, of our Lord the God of their fathers. \V
Ioſias therfore tooke away al abominations of al the countries of the
children of Iſrael: and made al, that were left in Iſrael, to ſerue our
Lord their God. Al his daies they reuolted not from our Lord the God of
their fathers.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
