%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 36
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-36


\project douay-rheims


%%% 0952
%%% o-0854
\startChapter[
  title={Chapter 36}
  ]

\Summary{Iochaz reigneth three monethes. 4.~His brother Eliakim (named
  Ioakim) eleuen yeares, 9.~his ſonne Ioachin three monethes, 11.~his
  vncle Sedecias eleuen yeares. 14.~Moſt Prieſtes, and people contemning
  the admonitions of Prophetes, 17.~manie are ſlaine by the Chaldees,
  the Temple and Ieruſalem ſpoiled and burnt. 20.~The ſayd kinges
  ſucceſsiuely, and people are caryed captiue into Babylon. 22.~After
  ſeuentie yeares Cyrus king of Perſia, releaſeth the captiuitie, and
  geueth leaue to reedifie the Temple.}

The people therfore of the land tooke Ioachaz the ſonne of Ioſias, and
made him king for his father in Ieruſalem. \V Three and twentie yeares
old was Ioachaz, when he began to reigne, and he reigned three monethes
in Ieruſalem. \V And the king of Ægypt when he came into Ieruſalem,
depoſed him, and condemned the land in an hundred talentes of ſiluer, and
a talent of gold. \V And he made Eliakim
\SNote{Hitherto from K.~Dauids death the ſonne had euer ſucceded to his
father.}
his brother king for him, ouer Iuda and Ieruſalem: and he turned his
name Ioakim: but he tooke Ioachaz himſelf with him, and caried him away
into Ægypt. \V Fiue and twentie yeares old was Ioakim when he began to
reigne, and he reigned eleuen yeares in Ieruſalem: and he did euil
before our Lord his God. \V Agaynſt him came vp Nabuchodonoſor the king
of the Chaldees, and brought him bound in chaynes into Babylon. \V
Whither he tranſported alſo the veſſels of our Lord, and put them in his
temple. \V But the reſt of the wordes of Ioakim, and of his
abominations, which he wrought, and the thinges that were found in him,
are conteyned in the Booke of the kinges of Iuda and Iſrael. And Ioachin
his ſonne reigned for him. \V Eight yeares old was Ioachin when he began
to reigne, and he reigned three monethes and ten dayes in Ieruſalem, and
he did euil in the ſight of our Lord. \V And when the
compaſſe of a yeare was come about, Nabuchodonoſor the king ſent ſome,
that brought him in to Babylon, the moſt precious veſſels of the houſe
of our Lord being caried away withal: But he made Sedecias his vncle
king ouer Iuda and Ieruſalem. \V One & twentie yeares old was Sedecias
when he began to reigne: & he reigned eleuen yeares in Ieruſalem. \V And
he did euil in the eies
%%% 0953
of our Lord his God, neither did he reuerence the face of Ieremie the
prophet ſpeaking to him from the mouth of our Lord. \V He reuolted alſo
from king Nabuchodonoſor, who had adiured him by God: & he hardened his
necke & his hart, that
%%% o-0855
he would not returne to our Lord the God of Iſrael. \V Yea & al the
chiefe of the Prieſtes, and the people tranſgreſſed vnlawfully according
to al the abominations of the Gentiles: and they polluted the houſe of
our Lord, which he had ſanctified to him in Ieruſalem. \V And our Lord
the God of their fathers ſent to them, by the hand of his meſſengers,
ryſing by night, and daily admoniſhing them: for that he ſpared his
people and his habitation. \V But they mocked the meſſengers of God, and
litle eſtemed his wordes, and ſcorned the prophetes, vntil the furie of
our Lord aſcended vpon his people, and there was no amendment. \V For he
brought vpon them the king of the Chaldees, and ſlewe their yong men
with the ſword in the houſe of his ſanctuarie, he pitied not yong man,
and virgin, and old man, no neither him that ſtouped for age, but he
deliured al into his handes. \V And al the veſſels of the houſe of our
Lord, as wel greater as leſſer, and the treaſures of the temple, and of
the king, and the princes he tranſported into Babylon. \V The enemies
ſet fyre on the houſe of God, and deſtroyed the wal of Ieruſalem, al the
towres they burnt, and what ſoeuer was pretious, they deſtroyed. \V If
anie man eſcaped the ſword, being led into Babylon he ſerued the king
and his ſonnes, til the king of the Perſians reigned. \V That the word
of our Lord by the mouth of Ieremie might be accompliſhed, and the land
might celebrate their Sabbathes:
\CNote{\XRef{Iere.~25.}}
For al the daies of the deſolation ſhe kept a Sabbath: til the ſeuentie
yeares were expyred. \V But in
\SNote{It is like that Eſdras added this cõcluſion when he reſtored the
holie Scriptures that were loſt: for he beginneth his owne booke with the
ſame wordes.}
the firſt yeare of Cyrus king of Perſians, to fulfil the word of our
Lord, which he had ſpoken by the mouth of Ieremie, our Lord rayſed vp
the ſpirit of Cyrus king of the Perſians: who commanded to be proclaymed
in al his kingdom, yea by writing, ſaying: \V Thus ſayth Cyrus king of
the Perſians: Al kingdomes of the earth hath the Lord the God of heauen
geuen me, and he hath commanded me that I ſhould build him a houſe in
Ieruſalem, which is in Iewrie: who of you is there in al his people?
\MNote{The end of the fifth age.}
The Lord his God be with him, and let him goe vp.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
