%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 0922
%%% o-0827
\startChapter[
  title={Chapter 21}
  ]

\Summary{Ioſaphat dieth, and Ioram ſucceding killeth his owne brethren,
  and ſome other chiefe men, 6.~reigneth wickedly. 8.~Edom, and Lobna
  reuolt from him. 12.~Elias by letters forwarneth him of plagues,
  16.~which falling vpon him, he dieth after two yeares languishing, and
  horrible diſeaſe.}

And
\CNote{\XRef{4.~Reg.~8.}}
Ioſaphat ſlept with his fathers, & was buried with them in the Citie of
Dauid: and Ioram his ſonne reigned for him. \V Who had brethren the
ſonnes of Ioſaphat, Azarias, and Iahiel, and Zacharias, and Azarias, and
Michael, and Sapharias. Al theſe were the ſonnes of Ioſaphat the king of
Iuda. \V And their father gaue them manie giftes of ſiluer, and of gold,
and penſions, with the moſt fenſed cities in Iuda: but the kingdom he
deliuered to Ioram, becauſe he was the firſt begotten. \V And Ioram roſe
ouer the kingdom of his father: and when he had eſtabliſhed himſelf, he
ſlew al his
%%% o-0828
brethren with the ſword, and certain of the princes of Iſrael. \V Two
and thirtie yeare old was Ioram when he began to reigne: and he reigned
eight yeares in Ieruſalem. \V And he walked in the waies of the kinges
of Iſrael, as the houſe of Achab had done: for Achabs daughter was his
wife, and he did euil in the ſight of our Lord. \V But our Lord would
not deſtroy the houſe of Dauid for the couenant, which he had made with
him: and becauſe he had promiſed that he would geue him a lampe, and to
his ſonnes for euer. \V In thoſe daies Edom rebelled, from being ſubiect
to Iuda, and made them ſelues a king. \V And when Ioram had paſſed with
his princes, and al the horſemen, that were with him, he roſe in the
night,
%%% 0923
and ſtroke Edom, which had compaſſed him, and al the captaynes of his
horſemen. \V But yet Edom rebelled, from being vnder the dominion of
Iuda vntil this day: at that time Lobna alſo reuolted, from being vnder
his hand. For he had forſaken our Lord the God of their fathers. \V
Moreouer he built alſo excelſes in the cities of Iuda, and he made the
inhabitantes of Ieruſalem to fornicate, and Iuda to tranſgreſſe. \V And
there were letters brought him
\SNote{Elias was aſſumpted from ordinarie conuerſation with mortal men
the eightenth yeare of king Ioſaphat
\XRef{4.~Reg.~2,~3.}
who reigned twentie fiue yeares.
\XRef{3.~Reg.~22. v.~42.}
So he ſhewed this ſpecial care of Ioram and his kingdom, after his
aſſumption ſeuen yeares.}
from Elias the prophete, in which was writen: Thus ſayth our Lord the
God of Dauid thy father: Becauſe thou haſt not walked in the waies of
Ioſaphat thy father, & in the waies of Aſa the king of Iuda, \V but haſt
gone by the waies of the kinges of Iſrael, and haſt made Iuda to
fornicate, and the inhabitantes of Ieruſalem, hauing imitated the
fornication of the houſe of Achab, moreouer alſo haſt killed thy
brethren, the houſe of thy father, better men then thou: \V behold our
Lord wil ſtrike thee with a great plague with al thy people, and
children, and thy wiues, and al thy ſubſtance. \V And thou shalt be
ſicke of a very ſore diſeaſe of thy bealie, til thy vital partes come
forth by litle and litle euerie day. \V Our Lord therfore raiſed vp
agaynſt Ioram the ſpirit of the Philiſthianes, and of the Arabians,
which are borderers to the Æthiopians. \V And they went vp into the Land
of Iuda, and waſted it, and they ſpoyled al the ſubſtance, that was
found in the kinges houſe, moreouer alſo his ſonnes, and wiues: neither
was there a ſonne left him but Ioachaz, who was the yongeſt. \V And
beſide al theſe thinges our Lord ſtroke him with an incurable diſeaſe of
the bealie. \V And when day ſucceded day, and the ſpaces of times paſſed
about, the circuite of two yeares was complete: and ſoe being waſted
with a long comſumption, ſoe that he voyded euen his very bowels, he was
ridde of the diſeaſe, and of his life together. And he died in an
exceding vile infirmitie, & the people made him not exequies according
to the maner of burning, as they had done to his
\Fix{anceters.}{anceſters.}{obvious typo, fixed in other}
\V He was two and thirtie yeares old, when he began to reigne, and he
reigned eight yeares in Ieruſalem. And he walked not rightly, and they
buried him in the Citie of Dauid: but yet not in the ſepulchre of the
kinges.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
