%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 32
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-32


\project douay-rheims


%%% 0942
%%% o-0846
\startChapter[
  title={Chapter 32}
  ]

\Summary{Sennacherib king of Aſſyria inuading Iuda, king Ezechias
  encorageth the people, and prouideth to defend the countrie. 9.~The
  Aſſyrians threaten the people, and blaſpheme God. 20.~Ezechias and
  Iſaias pray. 21.~An Angel deſtroyeth the Aſſyrians armie, ſo their
  king retiring home, is ſlayne in his idols temple, by his owne
  ſonnes. 22.~Ezechias reigneth in peace, 24.~falleth into deadlie
  ſicknes, but miraculouſly recouereth, offendeth in pryde, and
  repenteth: 27.~is exceeding rich, which he imprudently sheweth to
  ſtrangers: 32.~dieth, and Manaſſes ſuccedeth.}

After which thinges, and this maner of truth,
\CNote{\XRef{4.~Reg.~18.}
\XRef{Iſaie.~36.}}
came Sennacherib the king of the Aſſyrians, and entering
%%% 0943
into Iuda, beſieged the fenſed cities, deſirous to take them. \V Which
when Ezechias had ſene, to witte, that Sennacherib was come, and the
whole force of the battel to be turned agaynſt Ieruſalem, \V taking
counſel with the princes, and the moſt valiant men, to ſtoppe vp the
heades of the fountaynes, that were without the citie: and the ſentence
of them al decreing this, \V he gathered a very great multitude, & they
ſtopped vp al the fountaynes, and the riuer, that ranne in the middes of
the land, ſaying: Leſt the kinges of the Aſſyrians come, and finde
abundance of waters. \V He built alſo doing induſtriouſly euerie wall
that had bene deſtroyed, and built towers vpon them, and an other wall
without: and he repayred Mello in the citie of Dauid, and made armour
and ſhildes of al ſortes: \V And he appointed princes of warryers in the
armie: and he called them al together in the ſtreate of the gate of the
citie, and ſpake to their hart, ſaying: \V Doe manfully, and take
courage: feare not, neither dread ye the king of the Aſſyrians, and al
the multitude, that is with him: for there are manie moe with vs, then
with him. \V For with him is an arme of fleſh: with vs the Lord our God,
which is our helper, and fighteth for vs. And the people was encouraged
with theſe maner of wordes of Ezechias the king of Iuda. \V Which
thinges after they were done, Sennacherib the king of the Aſſyrians ſent
his ſeruantes to Ieruſalem (for himſelf with al his armie beſeiged
Lachis) to Ezechias the king of Iuda, & to al the people, that was in
the citie, ſaying: \V Thus ſayth Sennacherib the king of the Aſſyrians:
In whom hauing affiance, doe you ſitte beſieged in Ieruſalem? \V Hath
Ezechias deceiued you, to deliuer you to death in hunger and thirſt,
affirming that the Lord your God can deliuer you from the hand of the
king of the Aſſyrians? \V Why, is not this Ezechias, that hath deſtroyed
his excelſes, and altars, and hath commanded Iuda & Ieruſalem, ſaying:
Before
%%% o-0847
one altar you ſhal adore, and on it you ſhal burne incenſe? \V Are you
ignorant what thinges I haue done, and my fathers to al the peoples of
the landes? haue the goddes of nations, and of al landes bene able to
deliuer their countrie out of my hand? \V Who is there of al the goddes
of the nations, which my fathers waſted, that could deliuer his people
out of my hand, that your God alſo can deliuer you out of this hand? \V
Let not therfore Ezechias deceiue you, nor delude you
%%% 0944
with vayne perſuaſion, neither beleue ye him. For if no god of al
nations and kingdomes, could deliuer his people out of my hand, and out
of the hand of my fathers, conſequently neither ſhal your God be able to
deliuer you out of my hand. \V Yea and manie other thinges did his
ſeruantes ſpeake, agaynſt our Lord God, and agaynſt Ezechias his
ſeruant. \V Letters alſo he wrote ful of blaſphemie againſt our Lord the
God of Iſrael, and he ſpake agaynſt him: as the goddes of their nations
could not deliuer their people out of my hand, ſo the God alſo of
Ezechias can not deliuer his people out of this hand. \V Moreouer alſo
with a lowd crie, in the Iewes tongue, he ſounded againſt the people,
that ſate on the walles of Ieruſalem, that he might terrifie them, and
take the citie. \V And he ſpake agaynſt the God of Ieruſalem, as agaynſt
the goddes of the peoples of the earth, the workes of mens handes. \V
Ezechias therfore the king, and Iſaias the prophet the ſonne of Amos,
prayed agaynſt this blaſphemie, and cried out euen to heauen. \V And our
Lord ſent an Angel, which ſtroke euerie ſtrong man, and warryer, and
prince of the armie of the king of the Aſſyrians: and he returned with
ignominie into his countrie. And when he was entered into the houſe of
his god, his ſonnes that were come forth of his wombe, ſlewe him with
the ſword. \V And our Lord ſaued Ezechias and the inhabitantes of
Ieruſalem, out of the hand of Sennacherib the king of the Aſſyrians, and
out of the hand of al, & gaue them reſt round about. \V Manie alſo
brought hoſtes, and ſacrifices to our Lord into Ieruſalem, and giftes to
Ezechias the king of Iuda: who was exalted after theſe thinges before al
nations. \V In thoſe daies Ezechias was ſick euen to death, and he
prayed our Lord: and he heard him, and gaue him a ſigne. \V But not
according to the benefites, which he receiued, did he recompenſe,
\SNote{More danger of pride in proſperitie, then in aduerſitie.}
becauſe his hart was eleuated: and wrath came agaynſt him, and agaynſt
Iuda and Ieruſalem. \V And he was humbled afterward, becauſe his hart
had bene exalted, both he and the inhabitantes of Ieruſalem: and
therfore the wrath of our Lord came not vpon them in the daies of
Ezechias. \V And Ezechias was rich, and glorious excedingly and gathered
to himſelf great treaſures of ſiluer and of gold, and of pretious ſtone,
of ſpices, and of armour of al kinde, and of veſſels of great price. \V
Storehouſes alſo of corne, of wyne, and of oyle, and ſtalles of al
beaſtes, and
%%% 0945
foldes of cattel, \V and ſix cities he built to him ſelf: for he had
flockes of ſheepe, and of heardes innumerable, becauſe our Lord had
geuen him ſubſtance exceding much. \V The ſame is Ezechias, that ſtopped
the vpper fountaine of the waters
%%% o-0848
of Gihon, and turned them away vnderneth toward the Weſt of the Citie of
Dauid: in al his workes he did proſperouſly what he would. \V But yet in
the embaſſie of the princes of Babylon, that were ſent to him, to aske
of the wonder, that had chanced vpon the earth, God leift him that he
might be tempted, and al thinges might be made knowen, that were in his
hart. \V But the reſt of the wordes of Ezechias, and of his mercies are
writen in the viſion of Iſaias the ſonne of Amos the prophete, and in
the booke of the kinges of Iuda & Iſrael. \V And Ezechias ſlept with his
fathers, and they buried him in the chiefe ſepulchres of the children of
Dauid: and al Iuda celebrated his funeralles, & al the inhabitantes of
Ieruſalem: and Manaſſes his ſonne reigned for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
