%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 0897
%%% o-0805
\startChapter[
  title={Chapter 6}
  ]

\Summary{Salomon bleſſeth the people, 4.~prayeth to God, geuing thankes
  for benefites receiued, 16.~and requeſting continual protection, and
  that God wil heare the prayers of the people, 32.~yea alſo of
  ſtrangers that pray in the Temple.}

Then Salomon ſayd: Our Lord promiſed that he would dwel
\SNote{God ſhewed his preſence by the darke cloud which repleniſhed the
Temple.}
in darkeneſſe: \V and I haue built a houſe to his name, that he might
dwel there for euer. \V And the king turned his face, and bleſſed al the
multitude of Iſrael (for al the multitude ſtood attent) and ſayd: \V
Bleſſed be our Lord the God of Iſrael, who in worke hath accompliſhed
that, which he ſpake to Dauid my father, ſaying: \V From the day, that I
brought my people out of the Land of Ægypt, I choſe not a citie of al
the tribes of Iſrael, that a houſe might be built in it to my name:
neither did I
%%% o-0806
chooſe anie other man, to be Duke in my people Iſrael. \V But I choſe
Ieruſalem, that my name may be in it: and I choſe Dauid, that I might
appoynt him ouer my people Iſrael. \V And wheras Dauid my father had
meant to build a houſe to the name of our Lord the God of Iſrael, \V our
Lord ſayd to him: Becauſe thy wil was this, that thou wouldeſt build a
houſe to my name, thou haſt done wel certes to haue ſuch a wil: \V
howbeit thou ſhalt not build the houſe, but thy ſonne, which ſhal come
out of thy loynes, he ſhal build a houſe to my name. \V Our Lord
therfore hath accompliſhed his word, which he ſpake: and I am riſen for
Dauid my father, and ſit vpon the throne of Iſrael, as our Lord hath
ſpoken: and haue built a houſe to the name of our Lord the God of
Iſrael. \V And I haue put in it the Arke, wherin is the couenant of our
Lord, which he made with the children of Iſrael. \V He therfore ſtoode
before the Altar of our Lord, ouer agaynſt al the multitude of Iſrael,
and ſtretched forth his handes. \V For Salomon had made an eminent place
of braſſe, and had put it in the middes of the
\TNote{Baſilica.}
Temple, hauing fiue cubites of length, & bredth, & three of height: &
he ſtoode vpon it: & afterward
\SNote{Kneeling,}
kneeling ouer agaynſt al the multitude of Iſrael, and
\SNote{Lifting vp handes, ſtretching forth armes, and the like external
geſtures do much helpe internal attention in prayer, and alſo ſturre vp
others to godlie imitation: and therfore hath bene much practiſed by
deuout perſons both in the old and new Teſtament.}
his handes lifted vp toward heauen, \V he ſayd: Lord God of Iſrael, there
is not the like God to thee in heauen and in earth: which keepeſt
couenant and mercie with thy ſeruantes, that walke before thee in al
their hart: \V which haſt performed to thy ſeruant Dauid my father what
thinges ſoeuer thou hadſt ſpoken to him: and the thinges that
%%% 0898
by mouth thou hadſt promiſed, in worke thou haſt accompliſhed, as alſo
the preſent time proueth. \V Now therfore Lord God of Iſrael, fulfil to
thy ſeruant my father Dauid, what ſoeuer thou didſt ſpeake to him,
ſaying: There ſhal not fayle of thee a man before me, to ſitte vpon the
throne of Iſrael: yet ſo if thy children keepe their waies, and walke in
my law, as thou alſo haſt walked before me. \V And now Lord God of
Iſrael, be thy word confirmed, which thou haſt ſpoken to thy ſeruant
Dauid. \V Is it credible then that God ſhould dwel with men vpon the
earth? If heauen and the heauens of heauens doe not take thee, how much
more this houſe, which I haue built? \V But to this end only it is made,
that thou ſhouldeſt reſpect the prayer of thy ſeruant, and his
ſupplication Lord my God: and mayſt heare the prayers, which thy ſeruant
powreth out before thee: \V that thou open thine eies vpon this houſe
daies & nightes, vpon the place, wherein thou haſt promiſed that thy
name ſhould be inuocated, \V and wouldeſt heare the prayer; which thy
ſeruant prayeth in it: and mayſt heare the prayers of thy ſeruant, and
of thy people Iſrael. Whoſoeuer ſhal pray in this place, heare out of
thy habitation, that is from the heauens, and be propitious. \V If anie
man ſhal ſinne agaynſt his neighbour, and come readie to ſweare agaynſt
him, and binde himſelf with a curſe before the altar in this houſe: \V
thou ſhalt heare from heauen, and ſhalt doe the iudgement of thy
ſeruantes, ſo that thou render to the vniuſt his way vpon his owne head,
and reuenge the
%%% o-0807
iuſt, rewarding him according to his iuſtice. \V If thy people Iſrael
ſhal be ouercome of their enemies, (for they wil ſinne to thee) and
conuerted ſhal doe penance, and beſech thy name, and pray in this
place, \V thou ſhalt heare from heauen, and be thou
\Fix{propetious}{propitious}{likely typo, fixed in other}
to the ſinne of thy people Iſrael, and reduce them into the land, which
thou gaueſt them, and their fathers. \V If the heauen being ſhut, there
fal no rayne for the ſinnes of the people, and they ſhal beſech thee in
this place, and ſhal confeſſe to thy name, and ſhal be conuerted from
their ſinnes, when thou ſhalt afflict them, \V heare from heauen ô Lord,
and forgeue the ſinnes of thy ſeruantes and of thy people Iſrael, and
teach them the good way, by the which they may goe: and geue rayne to
thy land, which thou haſt geuen thy people to poſſeſſe. \V If famine
ſhal ariſe in the land, and peſtilence, ruſt, and blaſt, and locuſt, and
bruche, and the
%%% 0899
enemies, waſting the countries, ſhal beſiege the gates of the citie, and
al plague and infirmitie ſhal preſſe them: \V if any of thy people
Iſrael ſhal pray, knowing their plague and infirmitie, and ſhal ſpred
forth their handes in
\SNote{In this chapter and often els where it is euident, that places
dedicated to Gods ſeruice are more acceptable to him, then other
places.}
this houſe, \V thou ſhalt heare from heauen, to wit, out of thy high
habitation, and be thou propitious, and render to euerie one according
to his wayes which thou knoweſt him to haue in his hart: (for thou onlie
knoweſt the hartes of the children of men:) \V that they may feare thee,
and walke in thy waies al the daies, that they liue vpon the face of the
land, which thou haſt geuen to our fathers. \V The Forener alſo, which
is not of thy people Iſrael, if he come from a far countrie, for thy
ſtrong hand, & thy ſtreched out arme, and adore in this place: \V thou
ſhalt heare from heauen thy moſt firme habitation, & shalt do al
thinges, for the which that pilgrime shal inuocate thee: that al the
peoples of the earth may know thy name, and may feare thee, as thy
people Iſrael, and may know, that thy name is inuocated vpon this houſe,
which I haue built. \V If thy people ſhal goe forth to warre agaynſt
their aduerſaries, by the way, that thou ſhalt ſend them, ſhal adore
thee agaynſt the way, wherein is this citie, which thou haſt choſen, and
the houſe, which I haue built to thy name: \V thou ſhalt from heauen
heare their prayers, and petition, and doe thou reuenge. \V And if they
ſhal ſinne to thee (for there is no man that ſinneth not) and thou be
angrie with them, and deliuer them to the enemies, and they leade them
captiue into a farre countrie, or which is neere at the leaſt, \V and
being conuerted in their hart in the land, to the which they were led
captiue, ſhal doe penance, and ſhal beſech thee in the land of their
captiuitie, ſaying: We haue ſinned, we haue done wickedly, we haue delt
vniuſtly: \V and ſhal returne to thee in al their hart, and in al their
ſoule, in the land of their captiuitie, to the which they were led, shal
adore thee agaynſt the way of their land, which thou gaueſt their
fathers, and of the citie, which thou haſt choſen, and of the houſe,
which I haue built to thy name: \V thou shalt heare from heauen, that
is, from thy firme habitation their prayers, and doe thou iudgement, and
forgeue thy people, although ſinneful: \V for thou
%%% o-0808
art my God: let thine eies, I beſech thee, be opened, and let thine
eares be attent, to the prayer, that is made in this place. \V Now
therfore ariſe Lord God in thy reſt, thou and the Arke of thy ſtrength:
Let
%%% 0900
thy Prieſtes Lord God put on ſaluation, & thy Saintes reioyſe in good
thinges. \V Lord God turne not away from the face of thy Chriſt:
remember the mercies of Dauid thy ſeruant.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
