%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 27
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-27


\project douay-rheims


%%% 0933
%%% o-0837
\startChapter[
  title={Chapter 27}
  ]

\Summary{Ioathan a godlie king, 5.~preuaileth in battel againſt the
  Ammonites, 7.~dieth, and his ſonne Achaz ſuccedeth.}

Fiue
\CNote{\XRef{4.~Reg.~15.}}
and twentie yeares old was Ioathan when he began to reigne, and he
reigned ſixtene yeares in Ieruſalem: the name of his mother was Ieruſa
the daughter of Sadoc. \V And he did that which was right before our
Lord, according to al thinges, which Ozias his father had done, ſauing
that he entered not into the temple of our Lord, and as yet the people
did ſinne. \V He built the high gate of the houſe of our Lord, and in
the wal of Ophel he built manie thinges. \V Cities alſo he built in the
mountaynes of Iuda, and caſtelles and towres in the forreſtes. \V He
fought agaynſt the king of the children of Ammon, and ouercame them, and
the children of Ammon gaue him at that time an hundred talentes of
ſiluer, and tenne thouſand cores of wheate, and as manie cores of
barley: the children of Ammon gaue him theſe thinges in the ſecond and
third yeare. \V And Ioathan was ſtrengthened,
%%% o-0838
becauſe he had directed his waies before our Lord his God. \V But the
reſt of the wordes of Ioathan, and al his battels, and workes, are
writen in the Booke of the kinges of Iſrael and Iuda. \V He was fiue and
twentie yeares old when he began to reigne, and he reigned ſixtene
yeares in Ieruſalem. \V And Ioathan ſlept with his fathers, and they
buried him in the Citie of Dauid: and Achaz his ſonne reigned for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
