%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 0912
%%% o-0819
\startChapter[
  title={Chapter 15}
  ]

\Summary{Azarias prophecieth that Iſrael shal lack the true God,
  Prieſtes, and the law, a long time. 8.~Which king Aſa hearing moſt
  ſeriouſly deſtroyeth idolatrie, 12.~maketh couenant and oath to ſerue
  God, 16.~and depoſeth his mother for offering ſacrifice to Priapus.}

And Azarias the ſonne of Oded, the Spirit of God coming vpon him, \V
went out to meete Aſa, and ſayd to him: Heare ye me Aſa, and al Iuda and
Beniamin: Our Lord is with you,
\SNote{Cooperation with Gods grace meriteth increaſe of grace.}
becauſe you haue beene with him. If you wil ſeeke him, you ſhal finde:
but if you forſake him, he wil forſake you. \V And manie daies ſhal
paſſe in Iſrael without the true God, and without Prieſt a teacher, and
without the Law. \V And when they ſhal returne in their diſtreſſe to our
Lord the God of Iſrael, and ſhal ſeeke him, they ſhal finde him. \V At
that time there ſhal not be peace to him that goeth out and cometh in,
but terrours on euerie ſide in the inhabiters of the earth. \V For
nation ſhal fight agaynſt nation, and citie agaynſt citie, becauſe our
Lord wil truble them with al diſtreſſe. \V You therfore take courage,
and let not your handes be diſſolued: for there ſhal be reward to your
worke. \V Which when Aſa had heard, to witte, the wordes, and the
prophecie of Azarias the ſonne of Oded the prophete, he tooke courage,
and tooke away the Idols out of al the land of Iuda, and out of
Beniamin, and out of the cities, which he had taken, of mount Ephraim,
and he dedicated the altar of our Lord, which was before the porche of
our Lord. \V And he gathered together al Iuda and Beniamin, and the
ſtrangers with them of Ephraim, and of Manaſſes, and of Simeon: for
manie were fled to him of Iſrael, ſeing that our Lord his God was with
him. \V And when they were come into Ieruſalem the third
%%% 0913
moneth, in the fiftenth yeare of the reigne of Aſa, \V they immolated to
our Lord in that day of the ſpoiles, & the praie, that they had brought,
oxen ſeuen hundred, and rammes ſeuen thouſand. \V And he went in after
the maner to eſtabliſh the couenant, that they ſhould ſeeke our Lord the
God of their fathers in al their hart, and in al their ſoul. \V And if
anie man, quoth he, ſhal not ſeeke our Lord the God of Iſrael, let him
die, from the leaſt to the greateſt, from man vnto woman. \V And they
ſware to our Lord with a lowd voyce in iubilation, and in noyſe of
trumpet, and ſound of ſhaulmes, \V al that were in Iuda with execration:
for in al their hart did they ſweare, and with al their wil did they
ſeeke him, and found him, & our Lord gaue them reſt round about. \V Yea
and Maacha the mother of king Aſa he depoſed from the royal empyre,
becauſe ſhe had made in a groue the idol of Priapus:
%%% o-0820
which he wholy
deſtroyed, and breaking into peeces, burnt it in the Torrent cedron. \V
But the
\SNote{King Aſa deſtroyed the places where Idoles were ſerued,
\XRef{chap.~14. v.~2.}
but tollerated the places where ſome offered ſacrifice to God beſide the
proper altar in Ieruſalem, becauſe this was diſpenſable and not the
other.}
Excelſes were leaft in Iſrael: neuertheleſſe the hart of Aſa was perfect
al his daies. \V And thoſe thinges which his father had vowed, and him
ſelf, he brought into the houſe of our Lord, gold and ſiluer, and of
veſſels diuers furniture. \V And there was no warre vnto the fiue and
thirteth yeare of the reigne of Aſa.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
