%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 0918
%%% o-0824
\startChapter[
  title={Chapter 19}
  ]

\Summary{Ioſaphat being reproued by a prophet for yelding help to Achab,
  5.~appoynteth Iudges in ſeueral cities, admonishing them to doe
  iuſtice: 8.~exhorteth Prieſtes and Leuites to execute their functions
  carefully. 11.~Amarias High Prieſt directing and ruling in thinges
  belonging to God, Zabedias general captayne gouerneth the kinges
  affayres.}

And Ioſaphat the king of Iuda returned into his houſe peaceably, into
Ieruſalem. \V Whom Iehu the ſonne of Hanani the Seer mette, and ſayd to
him:
\SNote{Preciſian Donatiſtes holding it vnlawful to conuerſe with ſinners
amongſt other Scriptures alleaged this reprehenſion of Ioſaphat for his
ſocietie with Achab. To whom \Emph{S.~Auguſtin} anſwereth, that he was
not blamed for other conuerſation with Achab, but for ayding him,
and ioyning with him in the act of ſinne, when he contemning Micheas the
true prophete of God, and beleuing falſe prophetes went to battel wherin
both kinges offended, but with difference; ſo one was ſlaine though he
ſemed to be ſecure, the other was ſaued in great danger, and repenting was
pardoned.
\Cite{li.~2. c.~18. cont. epiſt. Parmen.}}
To the impious man thou geueſt ayde, and to them that hate our Lord thou
art ioyned in frendſhip, and therfore thou didſt deſerue in deed the
wrath of our Lord: \V but good workes are found in thee, for that thou
haſt taken away the groues out of the land of Iuda, and haſt prepared
thy hart to ſeeke our Lord the God of thy fathers. \V Ioſaphat therfore
dwelt in Ieruſalem: and he went forth to the people agayne from Berſabee
vnto mount Ephraim, and recalled them to our Lord the God of their
fathers. \V And he appoynted iudges of the land in al the fenſed cities
of Iuda, in euerie place, \V and commanding the iudges, he ſayd: Take
heede what you doe: for you exerciſe not the iudgement of man, but of
our Lord: and whatſoeuer you ſhal iudge, it ſhal redound to you. \V Let
the feare of our Lord be with you, and with diligence doe al thinges:
for there is no iniquitie with the Lord our God, nor acception
%%% 0919
of perſonnes, nor deſyre of giftes. \V In Ieruſalem alſo Ioſaphat
appoynted Leuites, and Prieſtes, and princes of families of Iſrael, that
they ſhould iudge the iudgement and cauſe of our
%%% o-0825
Lord to the inhabitantes therof. \V And he commanded them, ſaying: Thus
shal you doe in the feare of our Lord faithfully, and with a perfect
hart. \V Euerie cauſe, that shal come to you of your brethren, that dwel
in their cities, betwen kinred and kinred, wherſoeuer there is queſtion
of the law, of the commandement, of ceremonies, of iuſtifications: shew it
them, that they ſinne not agaynſt our Lord, and leſt there come wrath
vpon you and your brethren: ſo doing therfore you shal not ſinne. \V And
\SNote{A moſt plaine diſtinction of ſpiritual and temporal authoritie
and offices, not inſtituted by Ioſaphat, nor anie other king, but by God
himſelf.
\XRef{Deut.~17.}
\XRef{Num.~27.}}
Amarias the prieſt and your Bishop shal be chiefe in theſe thinges,
which pertayn to God: moreouer Zabadias the ſonne of Iſmahel, who is the
prince in the houſe of Iuda, shal be ouer thoſe workes, which pertayne
to the kinges office: and you haue maiſters the Leuites before you, take
courage, and doe diligently, and our Lord wil be with the good.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
