%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 0913
%%% o-0820
\startChapter[
  title={Chapter 16}
  ]

\Summary{Agaynſt the king of Iſrael, king Aſa procureth helpe of the
  Aſſyrians, 7.~which a prophet reprouing is put in fetters. 11.~Aſa
  dieth, with diſeaſe of his feete, and is buried with pompe.}

And
\CNote{\XRef{3.~Reg.~15.}}
in the ſix and thirteth yeare of his reigne, came vp Baaſa the king of
Iſrael into Iuda, and with a wall compaſſed Rama, that none could ſafely
goe out and come in of the kingdom of Aſa. \V Aſa therfore brought forth
ſiluer and gold, out of the treaſures houſe of our Lord, and of the kinges
treaſures, and he ſent to Benadad the king of Syria, who dwelt in
Damaſcus, ſaying: \V There is league betwen me & thee, my father alſo
and thy father had concord, wherfore I haue ſent thee ſiluer and gold,
that breaking the league, which thou haſt with Baaſa the king of Iſrael,
thou make him retire from me. \V Which being knowen, Benadad ſent the
princes of his hoſtes to the cities of Iſrael: who ſtroke Ahion, and
Dan, and Ablemaim, and al the walled cities of Nephthali. \V Which when
Baaſa had heard, he ceaſed to build
%%% 0914
Rama, and intermitted his worke. \V Moreouer Aſa the king tooke al Iuda,
and caried away the ſtones of Rama, and the timber that Baaſa had
prepared for the building: and he built of them Gabaa, & Maſpha. \V At
that time came Hanani the prophete to Aſa the king of Iuda, and ſayd to
him: Becauſe thou haſt had confidence in the king of Syria, and not in
our Lord thy God, therfore hath the armie of the king of Syria eſcaped
out of thy hand. \V Were not the Æthiopians, and Libyians manie moe in
chariotes, and horſemen, and a multitude exceding
great: whom, when thou didſt beleue in our Lord, he deliuered into thy
hand? \V For the eies of our Lord behold al the earth, and geue ſtrength
to them, that with perfect hart beleue in him. Thou therfore haſt done
foolyshly, & for this cauſe from this preſent time ſhal battels ariſe
agaynſt thee. \V And Aſa being angrie agaynſt the Seer, commanded him to
be caſt into fetters: for he tooke indignation excedingly vpon this
thing: and he ſlewe of the people at that time verie manie. \V But the
workes of Aſa the firſt & the laſt are writen in the Booke of the kinges
of Iuda and Iſrael. \V Aſa alſo fel ſicke in the nine and thirteth yeare
%%% o-0821
of his reigne, of a moſt vehement
\SNote{Iuſt puniſhment with paine of his feete, for iniuriouſly putting
Gods prophet in fetters.
\XRef{v.~10.}
So God puniſhed him temporally, for that and other paſſionate ſinnes:
and he died in good ſtate, for his hart was perfect al his dayes,
\XRef{(Chap.~15. v.~17.)}
that is, moſt part of his life, eſpecially in his laſt dayes.}
payne of his feete, and neither in his infirmitie did he ſeeke our Lord,
but rather truſted in the arte of Phiſitians. \V And he ſlept with his
fathers: and he died the one and fourteth yeare of his reigne. \V And
they buried him in his ſepulchre, which he had digged for himſelf in the
Citie of Dauid: and they layd him vpon his bed ful of ſpices and
odoriferous oyntementes, which were made by the arte of apothecaries,
and they burnt it ouer him with exceding ambition.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
