%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 0893
%%% o-0802
\startChapter[
  title={Chapter 3}
  ]

\Summary{The Temple is begunne to be built in mount Moria, the fourth
  yeare of Salomons reigne. 3.~The forme wherof is deſcribed, and the
  precious matter, 8.~eſpecially of \L{Sancta ſanctorum}: 14.~with a
  coſtlie veile, 15.~and two excellent pillers before the gate.}

And Salomon began to build the houſe of our Lord in Ieruſalem, in mount
\SNote{This mount \Emph{Moria}, ſignifying \Emph{viſion}, was ſo named
by Abraham, who was there readie to ſacrifice his ſonne Iſaac
\XRef{Gen.~22.}}
Moria, which had bene ſhewed to Dauid his father, in the place, which
\SNote{And Dauid, by the appointment of an Angel, offered there ſacrifice.
\XRef{2.~Reg.~14.~1.}
\XRef{Par.~21.}}
Dauid had prepared in the floore of Ornan the Iebuſeite. \V
\CNote{\XRef{2.~Reg.~24.}
\XRef{3.~Reg.~6.}}
And he began to build in the ſecond moneth, in the fourth yeare of his
kingdom. \V And theſe be the fundations, which Salomon layd, to build
the houſe of God, of length in the firſt meaſure ſixtie cubites, of
bredth twentie cubites. \V But the porch before the front, which was
extended in length according to the meaſure of the bredth of the houſe,
of twentie cubites: moreouer the height was of an hundred twentie cubites: and
he did
\Fix{guilt}{guild}{obvious typo, fixed in other}
it on the inſide with moſt pure gold. \V Alſo the greater houſe he
couered with wodden bordes of firre tree, and he faſtened on plates of
fine gold throughout: and he graued in it palme trees, and as it were
litle chaines embracing one an other. \V He paued alſo the floore of the
Temple with moſt precious marble, in much beautie. \V Moreouer it was
moſt tried gold, of the plates wherof he couered the houſe, and the
beames therof, and the poſtes, and the walles, and the doores: and he
graued cherubs in the walles. \V He made alſo the houſe of \L{Sanctum
ſanctorum}: the length according to the bredth of the houſe, of twentie
cubites: and the bredth likewiſe of it twentie cubites: and he couered
it with plates of gold, as it were ſix hundred talentes. \V Yea and he
made nailes of gold, ſo that euerie naile weighed fiftie ſicles a peece:
the vpper chambers alſo he couered with gold. \V He made alſo in the
houſe of \L{Sanctum ſanctorum} two Cherubs of ſtatuarie worke: and he
couered them with gold. \V The winges of the cherubs were extended
twentie cubites, ſo that one wing had fiue cubites, and touched the wal
of the houſe: and the other hauing fiue cubites, touched the wing of the
other cherub, \V In like maner the wing of the other cherub, had fiue
cubites, and touched the wal: and his other wing
of fiue cubites, touched the wing of the other cherub. \V Therfore the
winges of both the cherubs
%%% 0894
were ſpred forth, and were extended twentie cubites: and they ſtoode
vpright on their feete, and their faces were turned to the vtter
houſe. \V He made alſo a vele of hyacinth, purple, ſcarlet, and ſilke: &
woue in it cherubs. \V Before the doores alſo of the temple two pillers,
which had fiue and
%%% o-0803
thirtie cubites in height: moreouer their heades, of fiue cubites. \V
Moreouer alſo as it were litle chaynes in the oracle, & he put them to
the heades of the pillers: pomegranates alſo an hundred, which he put
betwen the litle chaynes. \V The pillers alſo them ſelues he put in the
entrance of the temple, one on the right hand, and the other on the
left:
\CNote{\XRef{3.~Reg.~7.}}
that, which was on the right hand, he called Iachin: and that on the
left hand, Boz.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
