%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 0901
%%% o-0810
\startChapter[
  title={Chapter 8}
  ]

\Summary{Salomon buildeth diuers cities, 7.~maketh the reſidue of the
  Chananites tributarie, 12.~offereth the ordinarie hoſtes of ſacrifices
  in the ſolemne feaſtes,
  %%% 0902
  14.~diſpoſeth the Prieſtes and Leuites in their offices, as Dauid had
  ordained: 17.~and ſendeth shippes to fetch gold from Ophir.}

And twentie yeares being complete after that Salomon built the houſe of
our Lord and his owne houſe: \V he built the cities, which Hiram had
geuen to Salomon, and made the children of Iſrael dwel there. \V He went
alſo into Emath Suba, and obteyned it. \V And he built Palmira in the
deſert, and he built other cities very wel fenſed in Emath. \V And he
built Beth horon the vpper, and Beth horon the nether, walled cities
hauing gates and barrigates and lockes. \V Balaath alſo and al the
ſtrongeſt cities that were Salomons, and al the cities of the chariotes,
and the cities of the horſemen. Al thinges whatſoeuer Salomon would, &
diſpoſed, he built in Ieruſalem and in Libanus, and in al the land of
his dominion. \V Al the people that was leaft of the Hetheites, and
Amorrheites, and Pherezeites, and Heueites, and Iebuſeites, which were
not of the ſtocke of Iſrael, \V of their children: and of the
poſteritie, which the children of Iſrael had not ſlaine, Salomon
ſubdewed to be tributaries, vntil this day. \V Moreouer of the children
of Iſrael he ſette not to ſerue the kinges workes: for they were men of
warre, and the chiefe captaines, and princes of his chariotes and
horſemen. \V And al the princes of king Salomons armie were two hundred
fiftie, which taught the people. \V But the daughter of Pharao he
remoued from the citie of Dauid, into the houſe, which he had built for
her. For the king ſayd: My wife shal not dwel in the houſe of Dauid the
king of Iſrael,
\SNote{Salomon was yet ſo farre from communicating with Infidels, that
he ſuffered not his wife (an infidel) to be preſent where the Arke of
God had bene placed.
\XRef{1.~Par.~15.}}
becauſe it is ſanctified: becauſe the Arke of our Lord is entered into
it. \V Then Salomon offered holocauſtes to our Lord vpon the altar of
our Lord, which he had built before the porch, \V that euerie day there
might be offering on it, according to the precept of Moyſes, in the
Sabbathes, and in the Calendes, and in the feſtiual daies thriſe a yere,
that is to ſay, in the Solemnitie of Azymes, and in the Solemnitie of
weekes, and in the Solemnitie of tabernacles. \V And he appoynted
according to the diſpoſition of Dauid his father the offices of the
Prieſtes in their miniſteries: & the Leuites in their order that they
should prayſe, and miniſter before the Prieſtes according to the rite of
euerie day: & the porters in their diuiſions by gate and gate: (for ſo
Dauid the man of God had
%%% 0903
commanded. \V Neither did they tranſgreſſe of the kinges commandementes
as wel the Prieſtes
%%% o-0811
as the Leuites, touching al thinges, that he had commanded, and in the
cuſtodies of the treaſures. \V Salomon had al expenſes prepared, from
the day that he founded the houſe of our Lord, vntil the day wherein he
perfitted it. \V Then went Salomon into Aſiongaber, and into Ailath to
the coaſt of the Red ſea, which is in the Land of Edom. \V And Hiram
ſent vnto him by the handes of his ſeruantes, shippes, and cunning
mariners, and they went with Salomons ſeruantes into Ophir, and they
tooke from thence foure hundred fiftie talentes of gold, and brought it
to king Salomon.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
