%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 0894
%%% o-0803
\startChapter[
  title={Chapter 4}
  ]

\Summary{The formes of the braſen altar, 2.~of the lauatorie (or Sea)
  with figuers of twelue oxen, 6.~of other tenne ſmal lauatories,
  7.~tenne candleſtickes, 8.~tenne tables, and an hundred baſſens, a
  great hal for the Prieſtes, 10.~and other veſſel, and ornamentes of
  the Temple are deſcribed.}

He made alſo an altar of braſſe,
\SNote{As wel the Temple as the Altars, and other appertinances were
made after the forme of Moyſes Tabernacle altares of ſacrifice &
incenſe, and the reſt: but greater in quantitie, of more precious
matter, and more excellent workmanſhip.}
of twentie cubites in length, and of twentie cubites in bredth, and of
ten cubites in height. \V
\CNote{\XRef{3.~Reg.~7.}}
\SNote{A great braſen veſſel, and ſome other thinges that were not
before.}
A Sea alſo caſt, ten cubites from brimme to brimme, round in compaſſe:
it had fiue cubites in height, and a cord of thirtie cubites did
compaſſe it round about. \V There was alſo vnder it the ſimilitude of
oxen, and certaine engrauinges of ten cubites on the outſide compaſſed
the
\Fix{bealie}{belie}{obvious typo, fixed in other}
of the Sea, as it were with two rewes. \V And the oxen were caſt: and
the Sea it ſelf was ſette vpon the twelue oxen, of the which three
looked toward the North, and other three to the Weſt: moreouer other
three to the South, and the three that remayned, to the Eaſt, hauing the
Sea vpon them: and the hinder partes of the oxen were inward vnder the
ſea. \V Moreouer the thickneſſe therof had the meaſure of a palme, and
the brimme therof was as it were the brimme of a chalice, or of a
criſped lilie: and it held three thouſand metretes. \V He made alſo ten
lauatories: and ſet fiue on the right hand, and fiue on the left, that
they might waſh in them al thinges, that they would offer for holocauſt:
moreouer in the Sea the prieſtes were washed. \V And he made alſo
\SNote{Alſo a greatter number of candleſtickes, and other the like.}
ten golden candleſtickes, according to the fashion which they were
commanded to be made by: and he ſet them in the temple, fiue on the
right hand, & fiue on the left. \V Moreouer alſo ten tables: and he ſet
them in the temple, fiue on
%%% 0895
the right hand, and fiue on the left. Phials alſo of gold an hundred. \V
He made alſo the court of the prieſtes, and a great hal: and doores in
the hal, which he couered with braſſe. \V Moreouer he ſet the Sea on the
right ſide agaynſt the Eaſt toward the South. \V And Hiram made
cauldrons, and flesh hookes, and phials: and accomplished al the kinges
worke
%%% o-0804
in the houſe of God: \V that is to ſay, two pillers, and the
\TNote{or pommels.}
chapiters, and the heades, and as it were certayne litle nettes, which
should couer the heades ouer the chapiters. \V Pomegranates alſo foure
hundred, and two litle nettes, ſoe that two rewes of the pomegranates
were ioyned to ech litle nette, which couered the pommels, and the
heades of the pillers. \V He made feete alſo, and lauatories, which he
put vpon the feete: \V one ſea, alſo twelue oxen vnder the ſea. \V And the
cauldrons, and flesh hookes, and phials. Al the veſſels did Hiram his
father make for Salomon in the houſe of our Lord of moſt pure braſſe. \V
In the countrie of Iordan did the king caſt them, in a clay ground
betwen Socot and Saredatha. \V And the multitude of veſſels was
innumerable, ſo that the weight of the braſſe was not knowen. \V And
Salomon made al the veſſels of the houſe of God, and the golden altar,
and the tables, & vpon them the loaues of propoſition, \V the
candleſtickes alſo with their lampes to giue light before the oracle,
according to the rite, of moſt pure gold: \V and certayne florishing
thinges, and lampes, and golden tonges al were made of moſt fine
gold. \V The veſſels alſo of perfume and cenſars, and phials, and litle
mortars, of moſt pure gold. And he graued the doores of the inner
temple, that is, in \L{Sancta ſanctorum}: and the doores of the
temple without of gold. And ſo al the worke was
\Fix{finishd}{finished}{obvious typo, fixed in other}
which Salomon made in the houſe of our Lord.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
