%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 26
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-26


\project douay-rheims


%%% 0931
%%% o-0836
\startChapter[
  title={Chapter 26}
  ]

\Summary{Ozias ſeruing God, 6.~preuaileth in battel againſt the
  Philiſtians, Arabians, and Ammonites: 9.~proſpereth in honour and
  wealth. 16.~Then waxing proud offereth incenſe on the altar, is
  ſtriken with leproſie, expelled out of the Temple, and citie: 20.~and
  his ſonne Ioathan ruleth the kingdom.}

And al the people of Iuda made his ſonne Ozias ſixtene yeares old, king
for Amaſias his father. \V He built Ailath, and reſtored it to the
dominion of Iuda, after that the king ſlept with his fathers. \V Sixtene
yeares old was Ozias when he began to reigne, and he reigned two and
fiftie yeares in Ieruſalem, the name of his mother was Iechelia of
Ieruſalem. \V And he did that which was right in the eies of our Lord,
according to al thinges, which Amaſias his father had done. \V And he
ſought our Lord in the daies of Zacharias that vnderſtood and ſaw God:
and when he ſought our Lord,
\SNote{So long as this king obſerued the ordinance of God to be directed
by the high prieſt
\XRef{Nu.~27. v.~21.}
he proſpered in his affayres.}
he directed him in al thinges. \V Moreouer he went forth, and fought
againſt the Philiſthijms, and deſtroyed the wal of Geth, and the wal of
Iabinia, and the wal of Azotus: he built alſo townes in Azotus, and
among the Philiſthijms. \V And God did helpe him againſt the
Philiſthijms, and againſt the Arabians, that dwelt in Garbaal, and
againſt the Ammonites. \V And the Ammonites gaue giftes to Ozias: and
his name was renowmed vnto the entrance of Ægypt for his often
victories. \V And Ozias built towers in Ieruſalem ouer the gate of the
corner, and ouer the gate of the valley, and the
%%% 0932
reſt, in the ſame ſide of the wal, and ſtrengthened them. \V He built
towers alſo in the wilderneſſe, and digged manie ceſternes, becauſe he
had much cattel as wel in the champaine, as in the vaſtitie of the
deſert: he had alſo vineyardes & dreſſers of vines in the mountaynes,
and in Carmel: for he was a man geuen to huſbandrie. \V And the hoſt of
his warriers, which went forth to battels, was vnder the hand of Iehiel
the ſcribe, & Maaſias the doctor, and vnder the hand of Hananias, who
was of the kinges dukes. \V And al the number of the princes by families
of valiant men, was two thouſand ſix hundred. \V And vnder them al the
hoſt of three hundred & ſeuen thouſand fiue hundred: which were apt to
battels, and fought for the king againſt the aduerſaries. \V Ozias alſo
prepared for them, that is to ſay, for al the armie, ſhildes, and
ſpeares, and helmettes, and coates of mayle, and bowes, and ſlinges to
caſt ſtones. \V And he made in Ieruſalem engines of diuerſe kind, which
he placed in the towers, and in the corners of the walles, to
shoote arrowes, and great ſtones: and his name went forth farre, for that
our Lord did ayde him, and had ſtrengthned him. \V But when he was
ſtrengthned, his hart was eleuated to his deſtruction, and
%%% o-0837
he neglected our Lord his God: and entering into the temple of our Lord:
he would burne
\SNote{For vſurping ſpiritual authority which pertayned not to him, the
high prieſt with his aſſiſtantes oppoſed themſelues againſt the king:
and God confirmed their ſentence, by ſtriking the ſame king with
leproſie. And ſo he was not only caſt out of the temple, but alſo out of
his kingdom, and common conuerſation with other men, & forced to dwel in
a ſeparet houſe without the citie according to the law.
\XRef{Leuit.~13. v.~46.}}
incenſe vpon the altar of incenſe. \V And incontinent Azarias the Prieſt
going in after him, & with him the Prieſtes of our Lord eightie, moſt
valiant men, \V they reſiſted the king, and ſayd: It is not thy office
Ozias, to burne incenſe to our Lord, but of the Prieſtes, that is, of the
children of Aaron, which are conſecrated to this kind of miniſterie: goe
out of the Sanctuarie, contemne not: becauſe this thing shal not be
reputed to thee for glorie of our Lord God. \V And Ozias being angrie,
and holding in his hand the cenſar to burne incenſe, threatned the
Prieſtes. And forthwith there roſe a leproſie in his forehead before the
Prieſtes, in the houſe of our Lord vpon the altar of incenſe. \V And
when Azarias the high Prieſt had beheld him, and al the reſt of the
Prieſtes, they ſaw the leproſie in his forehead, and in haſt they thruſt
him out. Yea and him ſelf being ſore afrayd, made haſt to goe out,
becauſe he felt by and by the plague of our Lord. \V Ozias therfore the
king was a leper vntil the day of his death, and he dwelt in a houſe
apart ful of the leproſie, for the which he had bene caſt out of the
houſe of our Lord.
%%% 0933
Moreouer Ioathan his ſonne gouerned the kinges houſe, and iudged the
people of the land. \V But the reſt of the wordes of Ozias the firſt and
the laſt wrote Iſaias the ſonne of Amos, the prophet. \V And Ozias ſlept with
his fathers, and they buried him in the
\SNote{Neither could he be buried in the proper ſepulchres of the
kinges.}
kings ſepulchres field, becauſe he was a leper: and Ioathan his ſonne
reigned for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
