%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-22


\project douay-rheims


%%% 0923
%%% o-0829
\startChapter[
  title={Chapter 22}
  ]

\Summary{Ochozias reigning one yeare, 3.~is ſlaine together with Ioram
  king of Iſrael, by king Iehu. 10.~Athalia killeth the kinges children
  (onlie Ioas being ſaued by his aunt) and vſurpeth the kingdom ſix
  yeares.}

%%% 0924
And
\CNote{\XRef{4.~Reg.~8. v.~25.}
\XRef{ch.~9.}}
the inhabitantes of Ieruſalem made Ochozias his leaſt ſonne, king for
him: for al the elders, that had bene before him, the rouers of the
Arabians had ſlayne, which inuaded the campe: and Ochozias the ſonne of
Ioram the king of Iuda reigned. \V Two and
\SNote{To wit when he beganne to reigne alone: for he reigned together
with his father at the age of 22.
\XRef{4.~Reg.~8. v.~26.}
And after his fathers death but one yeare.}
fourtie yeares old was
\Fix{Ochozias}{Ochozias}{obvious typo, fixed in other}
when he began to reigne, and he reigned one yeare in Ieruſalem, and the
name of his mother was Athalia the daughter of
\SNote{See
\XRef{4.~Reg.~8. v.~18.}}
Amri. \V But he alſo went by the waies of the houſe of Achab: for his
mother forced him to doe impiouſly. \V He therfore did euil in the ſight
of our Lord, as the houſe of Achab: for they were his counſelers after
the death of his father, to his deſtruction. \V And he walked in their
counſels. And he went forth with Ioram the ſonne of Achab king of
Iſrael, into battel agaynſt Hazael king of Syria, into Ramoth Galaad:
and the Syrians wounded Ioram. \V Who returned to be cured into Iezrael:
for he had taken manie woundes in the foreſayd battel. Therfore Ochozias
the ſonne of Ioram king of Iuda, went downe to viſit Ioram the ſonne of
Achab in Iezrael being ſicke. \V For it was the wil of God agaynſt
Ochozias, that he ſhould come to Ioram: and when he was come he ſhould
goe out alſo againſt Iehu the ſonne of Namſi, whom our Lord anoynted to
deſtroy the houſe of Achab. \V When Iehu therfore ouerthrew the houſe of
Achab, he found the princes of Iuda, and the ſonne of the brethren of
Ochozias, which ſerued him, and he ſlewe them. \V Searching alſo for
Ochozias himſelf, he tooke him lying hid in Samaria: and being brought
vnto him, he killed him, and they buried him: becauſe he was the ſonne
of Ioſaphat, who had ſought our Lord in al his hart
\SNote{Human hope failed, but Gods prouidence vſed meanes to conſerue
ſome of Dauids iſſue to ſitte in his throne. Yea to continue the
\Emph{ſucceſsion til Chriſt}.
\XRef{Mat.~1.}}
neither was there anie more hope that anie ſhould reigne of the ſtocke of
Ochozias. \V For Athalia his mother, ſeing that her ſonne was dead,
aroſe, and
\Fix{ſleu}{ſlew}{likely typo, fixed in other}
al the kinges ſtocke of the houſe of Ioram. \V Sauing that Ioſabeth the
kinges daughter tooke Ioas the ſonne of Ochozias, and ſtole him out of
the middes of the kinges ſonnes, when they were ſlayne, and she hid
\Fix{thim}{him}{obvious typo, fixed in other}
with his nource in the bedde chamber: and Ioſabeth that hid him, was the
daughter of king Ioram, the wife of Ioiada the high prieſt, the ſiſter
of Ochozias, and therfore Athalia did not kil him. \V He therfore was
with them in the houſe of God ſix yeares, in the which Athalia reigned
ouer the Land.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
