%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 25
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-25


\project douay-rheims


%%% 0929
%%% o-0834
\startChapter[
  title={Chapter 25}
  ]

\Summary{Amaſias killeth thoſe that ſlew his father. 5.~Beſides his owne
  people, hyreth ſouldiars of Iſrael, but by aduiſe of a Prophete
  diſmiſſeth them, 11.~and with his owne
  \Fix{owerthroweth}{ouerthroweth}{likley typo, same in both}
  the Idumeans, whoſe idols taken in battel (13.~the diſmiſſed ſouldiars
  in the meane time ſpoyling his countrie) he adoreth. 15.~Contemning
  admonition, 17.~and prouoking the king of Iſrael to warre, 22.~is
  taken in battel and ſpoyled. 27.~Fearing treaſon in Ieruſalem fleeth,
  and is ſlaine in Lachis.}

Fiue
\CNote{\XRef{4.~Reg.~14.}}
and twentie yeares old was Amaſias when he began to reigne, and he
reigned nine and twentie yeares in Ieruſalem, the name of his mother was
Ioaden of Ieruſalem. \V And he did good in the ſight of our Lord: but
yet not in a perfect hart. \V And when he ſaw his kingdom ſtrengthned,
he put to death the ſeruantes, that had ſlayne the king his father, \V
but their children he ſlew not, as it is writen in the Booke of the law
of Moyſes, where our Lord commanded, ſaying:
\CNote{\XRef{Deut.~24.}}
The fathers ſhal not be ſlayne for the children, nor the children for
their fathers, but euerie one ſhal die in his owne ſinne. \V Amaſias
therfore gathered together Iuda, and appoynted them by families, and
tribunes, and centurions in al Iuda, and Beniamin: and he numbred from
twentie yeares vpward, and found three hundred thouſand of yong men that
went forth to battel, and held ſpeare and ſhielde. \V He hyred alſo for
wages of Iſrael an hundred thouſand ſtrong men, for an hundred talentes
of ſiluer. \V But a man of God came to him, and ſayd: O king, let not
the hoſt of Iſrael goe forth with thee, for our Lord is not with Iſrael,
and al the children of Ephraim: \V and if thou thinke that battels
conſiſt in the force of an armie, God wil make thee to be ouercome of
the enemies: for it perteyneth to God both to helpe, and to put to
flight. \V And Amaſias ſayd to the man of God: What ſhal become then of
the hundred talentes, which I haue geuen the ſouldiars of Iſrael? And
the man of God anſwered him: Our Lord hath wherby he is able to
%%% 0930
geue thee much more then this. \V Amaſias therfore ſeperated the hoſt,
that came to him out of Ephraim, that they ſhould returne into their
place: but they being wrath excedingly agaynſt Iuda, returned into their
countrie. \V Moreouer Amaſias brought forth his people confidently, and
went into the Vale of ſalt pittes, and ſtroke the children of Seir, ten
thouſand. \V And other ten thouſand men did the children of Iuda take,
and bring to the ſteepe of a certaine rocke, and caſt them down headlong
from the toppe, who burſt in ſunder euerie one. \V But that armie which
Amaſias had ſent backe, from going with
%%% o-0835
him to battel, was ſpred in the cities of Iuda, from Samaria vnto
Bethhoron, & killing three thouſand tooke away a great praye. \V But
Amaſias after the ſlaughter of the Idumeans, ſette vp the goddes of the
children of Seir, which he had brought thence, for his goddes, and
adored them, and burnt incenſe to them. \V For which thing our Lord
being angrie againſt Amaſias, ſent a prophete vnto him, which ſhould ſay
to him: Why haſt thou adored goddes, that haue not deliuered their owne
people out of thy hand? \V
\CNote{\XRef{4.~Reg.~8.}}
And when he ſpake theſe thinges, he anſwered him: Art thou the kinges
counſeler? be quiet, leſt I kil thee. And the prophet departing, ſayd: I
know that God is minded to kil thee, becauſe thou haſt done this euil,
and beſides haſt not agreed to my counſel. \V Therfore Amaſias the king
of Iuda taking verie il counſel, ſent to Ioas the ſonne of Ioachaz the
ſonne of Iehu, the king of Iſrael, ſaying: Come, let vs ſee one an
other. \V But he ſent backe the meſſengers, ſaying: A thiſtle that is in
Libanus ſent to a cedar of Libanus, ſaying: Geue thy daughter to my
ſonne to wife: & behold the beaſtes that were in the wood of Libanus
paſſed, and trode downe the thiſtle. \V Thou haſt ſayd: I haue ſtroken
Edom, and therfore thy hart is extolled into pryde, ſitte in thy houſe,
why doſt thou prouoke euil againſt thee, that both thou mayeſt fal, and
Iuda with thee. \V Amaſias would not heare, becauſe it was our Lordes
wil, that he should be deliuered into the handes of the enemies
\SNote{Obduration of hart for former ſinne.}
for the goddes of Edom. \V Ioas therfore the king of Iſrael went vp, and
they gaue themſelues one the ſight of the other: and Amaſias the king of
Iuda was in Bethſames of Iuda: \V And Iuda fel before Iſrael, and fled
into their tabernacles.
%%% !!! This verse is garbled?
%%% The latin (according to NewAdvent) is:
%%%
%%% Porro Amasiam regem Juda, filium Joas filii Joachaz, cepit Joas rex
%%% Israël in Bethsames, et adduxit in Jerusalem: destruxitque murum
%%% ejus a porta Ephraim usque ad portam anguli quadringentis cubitis.
%%%
%%% This has
%%%
%%% \V Moreouer Amaſias the king of Iuda, the ſonne of Ioas, the ſonne
%%% of Ioachaz, did take Ioas
%%% 0931
%%% the king of Iſrael in Bethſames, & brought him into Ieruſalem: and
%%% deſtroyed the wal therof from the gate of Ephraim, to the gate of
%%% the corner, foure hundred cubites.
%%%
%%% Other has:
\V Moreouer Ioas the king of Iſrael in Bethſames tooke Amaſias the king
of Iuda, the ſonne of Ioas, the ſonne of Ioachaz, & brought him into
Ieruſalem: and deſtroyed the wal therof from the gate of Ephraim, to the
gate of the corner, four hundred cubites.
%%% Back to normal now:
\V Al the gold alſo, and ſiluer, and al the veſſels, that he found in
the houſe of God, and with Obededom in the treaſures alſo of the kinges
houſe, moreouer he brought backe the ſonnes of the hoſtages into
Samaria. \V And Amaſias the ſonne of Ioas the king of Iuda liued, after
that Ioas died the ſonne of Ioachaz the king of Iſrael, fiftene
yeares. \V But the reſt of the wordes of Amaſias the firſt and the laſt are
writen in the Booke of the kinges of Iuda and Iſrael. \V Who after he
reuolted from our Lord, they lay in waite agaynſt him in Ieruſalem. And
when he had fled into Lachis, they ſent, and ſlew him there. \V And
carying him backe vpon horſes, buried him with his fathers in the Citie
of Dauid.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
