%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 0914
%%% o-0821
\startChapter[
  title={Chapter 17}
  ]

\Summary{Ioſaphat ſucceding in the kingdom preuaileth in battel agaynſt
  the king of Iſrael, 6.~deſtroyeth Idolaters, and ſendeth Prieſtes and
  Leuites to inſtruct the people. 11.~The Philiſtians and Arabians ſend
  preſentes to king Ioſaphat. 13.~The leaders of the armie and ſouldiars
  are numbred.}

And Ioſaphat his ſonne reigned for him, & grew ſtrong agaynſt Iſrael. \V
And he appoynted numbers of ſouldiars in al the cities of Iuda, that were
compaſſed with walles. And he placed garriſons in the land of Iuda, and
in the cities of Ephraim, which Aſa his father had taken. \V And our
Lord was with Ioſaphat, becauſe he walked in the firſt waies
%%% 0915
of Dauid his father: and he truſted not in Baalim, \V but in the God of
his father, and went forward in his preceptes, and not according to the
ſinnes of Iſrael. \V And our Lord confirmed the kingdom in his hand, and
al Iuda gaue giftes to Ioſaphat: and there grew to him infinite riches,
and much glorie. \V And when his hart had taken courage
\SNote{Good workes approued by new benefites from God, as a reward
therof, do geue more hope & confidence to procede from vertue to vertue.
\Cite{S.~Tho. li. de regimine Regum.}}
for the waies of our Lord, he tooke away alſo the Excelſes and groues
out of Iuda. \V And in the third yeare of his kingdom, he ſent of his
princes Benhail, and Abdias, and Zacharias, and Nathanael, and Micheas,
that they ſhould teach in the cities of Iuda: \V and with them Leuites,
Semeias, and Nathanias, and Zabadias, Aſael alſo, and Semiramoth, and
Ionathan, and Adonias and Tobias, and Thobadonias Leuites, and with them
Eliſama, and Ioram Prieſtes. \V And they taught the people in Iuda,
hauing the booke of the law of our Lord: and they went about al the
cities of Iuda, and inſtructed the people. \V Therfore the dread of our
Lord came vpon al the kingdomes of the landes, that were round about
Iuda, neither durſt they make battel agaynſt Ioſaphat. \V Yea and the
Philiſtians brought giftes to Ioſaphat, and tribute of ſiluer, the
Arabians alſo brought cattel, of rammes ſeuen thouſand ſeuen hundred,
and buckegoates as manie. \V Ioſaphat therfore grewe, and was magnified
on high: and he built in Iuda houſes like to toures, and walled
cities. \V And he prepared manie workes in the cities of Iuda: there
were
%%% o-0822
alſo men of warre, and valiant in Ieruſalem, \V of whom this is the
number by the houſes and families of euerie one: In Iuda princes of the
armie, Ednas duke, and with him moſt valiant men three hundred
thouſand. \V After him Iohanan the prince, and with him two hundred
eightie thouſand. \V After him alſo Amaſias the ſonne of Zechri,
conſecrated to our Lord, and with him two hundred thouſand of valiant
men. \V Him followed Eliada valiant to battels, and with him of them
that held bow & ſhield two hundred thouſand. \V After this man alſo
Iozabad, and with him an hundred eightie thouſand readie ſouldiars. \V
Al theſe were at the hand of the king, beſide others, whom he had put in
walled cities, in al Iuda.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
