%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 0911
%%% o-0818
\startChapter[
  title={Chapter 14}
  ]

\Summary{Abias dieth, and his ſonne Aſa reigneth, deſtroyeth idolatrie,
  6.~fortifieth his cities, 9.~and ouercometh, by Gods ſpecial helpe,
  the Æthiopians armie of a million of men.}

And Abias ſlept with his fathers, and they buried him in the Citie of
Dauid: and Aſa his ſonne reigned for him, in whoſe daies the land was
quiet tenne yeares. \V And Aſa did that which was good and pleaſing in
the ſight of his God, and he ouerthrew the altars of
\SNote{It perteyned properly to the king to deſtroy the exterior
practiſe of idolatrie, but to the Prieſtes, & prophetes to informe the
interior mind, & conſciences of euerie one.}
ſtrange ſeruice, and the excelſes, \V and brake the ſtatues, and cut
downe the groues. \V And he commanded Iuda that they ſhould ſeeke our
Lord the God of their fathers, and ſhould doe the law, and al the
commandementes. \V And he tooke away out of al the cities of Iuda the
altars, and temples, & reigned in peace. \V He built alſo fenſed cities
in Iuda, becauſe he was quiet, and there had no battels riſen in his
time, our Lord geuing peace. \V And he ſayd to Iuda: Let vs build theſe
cities, and compaſſe them with walles, and ſtrengthen them with towers,
and gates, and lockes, whiles thinges are quiet from battels, becauſe we
haue ſought our Lord the God of our fathers, and he hath geuen vs peace
round about. They therfore did build, and there was no impediment in
building. \V And Aſa had in his armie of them that caried ſhieldes and
ſpeares, of Iuda three hundred thouſand: and of Beniamin ſhieldbearers
and archers, two hundred eightie thouſand, al theſe were moſt valiant
men. \V And Zara the Æthiopian with his armie iſſued forth agaynſt them,
tenne hundred thouſand, and with three hundred chariotes: and he came as
far as Mareſa. \V Moreouer Aſa went on to meete him, and ſet his armie
in aray to battel in the vale Sephata, which is neere Mareſa. \V And he
inuocated our Lord God, and ſayd: Lord there is no difference with thee,
whether thou helpe in few, or in manie: helpe vs ô Lord our God: for
hauing confidence in thee, and in thy name we are come agaynſt this
multitude. Lord, thou art our God, let not man preuaile agaynſt thee. \V
Our Lord
%%% 0912
therfore terrified the Æthiopians before Aſa and Iuda: and the
Æthiopians fled. \V And Aſa purſewed them, and the people that was with
him, vnto Gerara: and the Æthiopians fel to vtter deſtruction, becauſe
our Lord killing them, and his armie fighting they were deſtroyed. They
tooke therfore manie ſpoiles, \V and they ſtroke al the cities round
about Gerara: for great terrour had inuaded al men: and they ſpoyled the
cities, and caried away much praye. \V Yea and deſtroying the
ſheepecotes of ſheepe, they tooke an infinite multitude of cattel, and
of camels: and returned into Ieruſalem.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
