%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 0919
%%% o-0825
\startChapter[
  title={Chapter 20}
  ]

\Summary{The Ammonites, Moabites, and Syrians ioyning forces agaynſt
  Ioſaphat, 3.~he ſeeketh Gods helpe by publique prayer and
  faſting. 14.~A Prophet fortelleth that God wil fight for them: 20.~ſo
  they ſinging praiſes to God, the enemies kil ech other. 24.~Ioſaphat
  with his men gather verie great ſpoiles, 30.~reigneth in peace,
  35.~but his nauie perisheth, for his ſocietie with wicked Ochozias.}

After theſe thinges were the children of Moab gathered together, and the
children of Ammon, and with them of the Ammonites, to fight agaynſt
Ioſaphat. \V And there came meſſengers, and told Ioſaphat, ſaying: There
cometh agaynſt thee a great multitude from thoſe places, which are
beyond the ſea, and out of Syria, and behold they ſtay in Aſaſonthamar,
which is Engaddi. \V And Ioſaphat being frighted with feare, he tooke
him wholy to beſech our Lord, and he
\SNote{They faſted not only to ſubdue the fleſh to the ſpirite but alſo
for other neceſſities.}
proclamed a faſt to al Iuda. \V And Iudas was gathered together to pray
to our Lord: yea and al
\SNote{Example of pilgremage to holie places, becauſe it pleaſeth
\Fix{Go}{God}{obvious typo, fixed in other}
to heare the prayers of good people rather in one place then in an
other.
\XRef{ch.~6.~7.~&c.}}
came out of their cities to beſech him. \V And when Ioſaphat ſtood in
the middes of the aſſemblie of Iuda, and Ieruſalem in the houſe of our
Lord before the new court, \V he ſayd: Lord God of our fathers, thou art
God in heauen, and ruleſt ouer al the kingdomes of Nations, in thy hand
is ſtrength and might, neither
%%% 0920
can anie man reſiſt thee. \V Didſt not thou our God kil al the
inhabitantes of this land before thy people Iſrael, and gaueſt it to the
ſeed of Abraham thy frend for euer? \V And they dwelt in it, & built in
it a Sanctuarie to thy name, ſaying: \V If euils fal vpon vs, the ſword
of iudgement, peſtilence, & famine, we wil ſtand before this houſe in
thy ſight, wherein thy name is inuocated: & we wil crie to
%%% o-0826
thee in our tribulations, and thou ſhalt heare, and ſaue vs. \V Now
therfore behold the children of Ammon, and mount Seir, by whom thou
didſt not grant Iſrael to paſſe, when they came out of Ægypt, but they
declined from them, & ſlew them not: \V doe the contrarie, and endeuoure
to caſt vs out of the poſſeſſion, which thou haſt deliuered to vs. \V
Our God, wilt not thou therfore iudge them? In vs in deed there is not
ſo great ſtrength, that we can reſiſt this multitude, which cometh
violently vpon vs. But whereas we are ignorant what we ought to doe,
this onlie we haue leaft, that we direct our eies to thee. \V And al
Iuda ſtood before our Lord with their litle ones, and wiues, and their
children. \V And there was Iahaziel the ſonne of Zacharias, the ſonne of
Banaias, the ſonne of Iehiel, the ſonne of Mathanias, a Leuite of the
children of Aſaph, vpon whom the ſpirit of our Lord came in the middes
of the multitude, \V and he ſayd: Attend ye al Iuda, and you that dwel
in Ieruſalem, and thou king Ioſaphat: thus ſayth our Lord to you: Feare
not, neither dread ye this multitude: for it is not your battel, but
Gods. \V To morrow you shal goe downe agaynſt them: for they wil come vp
by the ſteepe named Sis, and you shal find them in the vtmoſt part of
the torrent, which is agaynſt the wilderneſſe of Ieruel. \V It shal not
be you that shal fight, but onlie ſtand confidently, and you shal ſee
the helpe of our Lord ouer you, ô Iuda, and Ieruſalem: feare not,
neither dreade ye: to morrow you shal goe out agaynſt them, and our Lord
wil be with you. \V Ioſaphat therfore, and Iuda, and al the inhabitantes
of Ieruſalem fel flat on the earth before our Lord, and adored him. \V
Moreouer the Leuites of the children of Caath, and of the children of
Core prayſed our Lord the God of Iſrael with a lowd voice, on high. \V
And when they had riſen earely in the morning, they went forth by the
deſert of Thecua: and they being gone forth, Ioſaphat ſtanding in the
middes of them, ſayd: Heare me ye men of Iuda, and al the inhabiters of
Ieruſalem:
\SNote{Faith is the fundation of al good workes,
\XRef{Heb.~11.}
but not ſufficient to ſaluation without other vertues, and therfore they
not only beleued but alſo faſted and prayed, and where nede required
fought with weapons, though at this time it pleaſed God to fight for
them. See
\XRef{Ioſue.~23.}}
beleue in the Lord
%%% 0921
your God, and you ſhal be ſecure: beleue his prophetes, and al thinges
ſhal fal out proſperous. \V He gaue counſel alſo to the people, and
appoynted the ſinging men of our Lord, that they ſhould prayſe him in
their companies, and ſhould goe before the hoſt, and with agreable voice
ſhould ſay:
\CNote{\XRef{Pſal.~135.}}
Confeſſe to our Lord, becauſe his mercie is foreuer. \V And when they
began to ſing prayſes, our Lord turned their embuſhementes vpon
themſelues, to witte, of the children of Ammon, and of Moab, and of
mount Seir, who were gone forth to fight agaynſt Iuda, and were
ſtriken. \V For the children of Ammon, and of Moab, roſe together
agaynſt the inhabitantes of mount Seir, to kil and deſtroy them: and
when they had in worke atcheued this, being turned alſo agaynſt
themſelues, they fel wounded one of an other. \V Moreouer Iuda when they
were come to the watch place, that looketh to the deſert, ſaw a far of
al the countrie abrode ful of dead bodies, and that none remayned aliue
that could eſcape death. \V Ioſaphat therfore came, and al the people
with him to take away the ſpoiles of the dead, and they found among the
%%% o-0827
dead bodies, diuerſe ſtuffe, garments alſo, and moſt pretious veſſels:
and they ſpoiled it, ſoe that they could not carie al thinges, nor in
three dayes take away the ſpoiles for the greatneſſe of the praye. \V
And in the fourth day they were aſſembled in the Vale of bleſſing: for
becauſe there they had bleſſed our Lord, they called that place the Vale
of bleſſing vntil this preſent day. \V And euerie man of Iuda returned,
and the inhabitantes of Ieruſalem, and Ioſaphat before them into
Ieruſalem with great ioy, becauſe our Lord had geuen them ioy of their
enemies. \V And they entered into Ieruſalem with pſalteries, and harpes,
and trumpettes into the houſe of our Lord. \V And the dread of our Lord
fel vpon al the kingdomes of the landes when they heard that our Lord
had fought agaynſt the enemies of Iſrael. \V And the
\Fix{kidgdom}{kingdom}{obvious typo, fixed in other}
of Ioſaphat was quiet and God gaue him peace round about. \V Ioſaphat
therfore reigned ouer Iuda, and he was fiue and thirtie yeares old when
he begane to reigne: and he reigned fiue and twentie yeares in
Ieruſalem: and the name of his mother was Azuba the daughter of
Selahi. \V And he walked in the way of his father Aſa, neither declined
he from it, doing the thinges that were pleaſing before our Lord. \V But
\SNote{He deſtroyed the places where ſacrifice was offered to idoles,
\XRef{ch.~17. v.~6.}
but tolerated other places where the people offered to God our Lord
without the temple, not being able to reduce al to perfection.}
yet the excelſes he tooke not away, and as yet the people had not
directed their hart to
%%% 0922
our Lord the God of their fathers. \V But the reſt of the actes of
Ioſaphat, the firſt and the laſt are writen in the wordes of Iehu the
ſonne of Hanani, which he diſpoſed into the Bookes of the kinges of
Iſrael. \V After theſe thinges Ioſaphat the king of Iuda entered
frendshipe with Ochozias the king of Iſrael, whoſe workes were moſt
impious. \V And he was partaker to make ſhippes, which should goe into
Tharſis: and they made a nauie in Aſiongaber. \V And Eliezer the ſonne
of Dodau of Mareſa prophecied to Ioſaphat, ſaying: Becauſe thou haſt had
a league with Ochozias, our Lord hath ſtrooken thy workes, and the
shippes are broken, neither could they goe into Tharſis.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
