%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 0906
%%% o-0814
\startChapter[
  title={Chapter 11}
  ]

\Summary{Roboam intending by force to reduce al Iſrael to his
  ſubiection, is warned by a Prophet to ceaſe from that
  enterpriſe. 5.~He maketh walles about diuers cities, 11.~furnisheth
  them with victuals, and munition. 13.~Prieſtes, Leuites, and manie
  others repaire to Ieruſalem, becauſe Ieroboam maketh a new religion,
  and new prieſtes. 18.~Roboam taketh manie wiues, and
  concubines. 22.~Preferreth Abias aboue al his other ſonnes.}

%%% 0907
And Robaom came into Ieruſalem, and called together al the houſe of Iuda
and Beniamin, an hundred foureſcore thouſand choſen men and warriers, to
fight agaynſt Iſrael, and to conuert his kingdom vnto him. \V And the
word of our Lord came to Semeias the man of God, ſaying: \V Speake to
Roboam the ſonne of Salomon the king of Iuda, and to al Iſrael, that is
in Iuda and Beniamin: \V Thus ſayth our Lord: You ſhal not goe vp,
neither ſhal you fight agaynſt your brethren: let euerie man returne
into his houſe, becauſe this thing is done by my wil. Who when they had
heard the word of our Lord, returned, neither went they forward agaynſt
Ieroboam. \V And Roboam dwelt in Ieruſalem, and built walled cities in
Iuda. \V And he built Bethlehem, and Etam, and Thecue, \V Bethſur alſo,
and Socho, & Odollam, \V moreouer alſo Geth, and Mareſa, and Ziph, \V
yea and Aduram, and Lachis, and Azeca, \V Saraa alſo, and Aialon, and
Hebron, which were in Iuda and Beniamin, moſt fenſed cities. \V And when
he had incloſed them with walles, he put in them princes, and ſtore
houſes of victuals, that is, of oile and wine. \V Yea and in euerie
citie he made armories of ſhieldes and ſpeares, and he ſtrengthened them
with great diligence, and reigned ouer Iuda and Beniamin. \V And the
Prieſtes and Leuites, that were in al Iſrael, came to him out of al
their ſeates, \V
\SNote{A worthie example to ſuffer temporal damage rather then to
conforme them ſelues to the practiſe of falſe religion.}
leauing their ſuburbes, and their poſſeſſions, and paſſing to Iuda, and
Ieruſalem, becauſe Ieroboam had caſt them of, and their poſteritie: that
they should not execute the prieſthood of our Lord. \V Who made vnto him
ſelfe
%%% o-0815
prieſtes of the excelſes, and of diuels, and of the calues which he had
made. \V Yea and of al the tribes of Iſrael, whoſoeuer had geuen their
hart to ſeeke our Lord the God of Iſrael, came into Ieruſalem to
immolate their victimes before our Lord the God of their fathers. \V And
they ſtrengthened the kingdom of Iuda, and eſtablished Roboam the ſonne
of Salomon for three yeares: for they walked in the waies of Dauid and
Salomon, onlie three yeares. \V And Robaom tooke to wife Mahalath, the
daughter of Ierimoth the ſonne of Dauid: Abihail alſo the daughter of
Eliab the ſonne of Iſai, \V who bare him ſonnes Iehus, and Somorias, and
Zoom. \V After this woman alſo he tooke Maacha the daughter of Abſalom,
who bare him Abia, and Ethai, and Ziza, and Salomith. \V And Roboam
loued Maacha the daughter of Abſalom aboue al his wiues,
%%% 0908
and concubines: for he had maried eightene wiues, and threeſcore
concubines: and he begat eight and twentie ſonnes, and threeſcore
daughters. \V But he appoynted for head Abias the ſonne of Maacha duke
ouer al his brethren: for he meant to make him king, \V becauſe he was
wiſer, and mightier aboue al his ſonnes, and in al the coſtes of Iuda,
and of Beniamin, and in al the walled cities: and he gaue them much
meate, and he deſired manie wiues.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
