%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 0891
%%% o-0800
\startChapter[
  title={Chapter 2}
  ]

\Summary{Other workemen being prouided to build the Temple, 7.~Salomon
  procureth a cunning artificer from the king of Tyre, and ſpecial
  timber, 16.~to be cutte, and ſent from thence.}

And Salomon determined to build a houſe to the name of our Lord, and a
palace for himſelf. \V
\CNote{\XRef{3.~Reg.~5.}}
And he numbred ſeuentie thouſand men that caried
on their ſhoulders, and eightie thouſand that ſhould hew ſtones in the
mountaynes, and ouerſeers of them three thouſand ſix hundred. \V He ſent
alſo to Hiram the king of Tyre, ſaying: As thou didſt with Dauid my
father, and didſt ſend him cedar trees, to build
%%% o-0801
him a houſe, wherin alſo he dwelt: \V ſo doe with me that I may build a
houſe to the name of our Lord my God, that I may conſecrate it to burne
incenſe before him, and to perfume with aromatical ſpices, and to the
euerlaſting propoſition of loaues, and for holocauſtes, morning and
euening, on the ſabbathes alſo, and the newmoones, and the ſolemnities
of our Lord God for euer, which are commanded Iſrael. \V For the houſe
which I deſire to build, is great: for our God is great aboue al
goddes. \V Who then can be able, to build him a worthie houſe? if
heauen, and the heauens of heauens can not conteyne him: how great am I,
that I may build him a
%%% 0892
houſe? but to this end only, that incenſe may be burnt before him. \V
Send me therfore a cunning man, that hath skil to worke in gold, and
ſiluer, braſſe, and yron, purple, ſcarlet, and hyacinth, and that
knoweth to make engraued workes with theſe artificers, which I haue with
me in Iewrie, and Ieruſalem, whom Dauid my father prepared. \V But ſend
me alſo cedar trees, firretrees, and pintrees from Libanus: for I know
that thy ſeruantes haue skil to hew the timber of Libanus, and my
ſeruantes ſhal be with thy ſeruantes, \V that manie trees may be
prepared for me. For the houſe which I deſire to build, is exceding
great, and glorious. \V Moreouer to the workemen, that ſhal hew the
trees, thy ſeruantes, I wil geue for victuals of wheat twentie thouſand
cores, and of barley as manie cores, and of wine twentie thouſand, oyle
alſo twentie thouſand ſates. \V And Hiram the king of Tyre by letters,
which he ſent to Salomon, ſayd: Becauſe the Lord loued his people,
therfore hath he made thee to reigne ouer it. \V And he added, ſaying:
Bleſſed be the Lord the God of Iſrael, that made heauen and earth, who
hath geuen to Dauid the king a ſonne wiſe and learned, and of
vnderſtanding, and prudent, to build a houſe to the Lord, and a palace
for himſelf. \V I therfore haue ſent thee a man wiſe and moſt skilful,
Hiram,
\SNote{It is probable, that this man had inſtructed the king of Tyre in
true religion of one God whom he confeſſeth
\XRef{v.~11.~12.}
and that therfore the king called him his father.}
my father, \V the ſonne of a woman of the daughters of Dan, whoſe father
was a Tyrian, who knoweth to worke in gold, and ſiluer, braſſe, and
yron, and marble, and in timber, in purple alſo, and hiacinth, and
ſilke and ſcarlet: and that knoweth to graue al engrauing, and to deuiſe
wiſely whatſoeuer in the worke is neceſſarie with thy artificers, and
with the artificers of my lord Dauid thy father. \V The wheate therfore,
and barley, and oile, and wine, which thou my lord haſt promiſed, ſend
to thy ſeruantes. \V And we wil cut downe the trees out of Libanus, as
manie as shal be neceſſarie for thee, and wil conuey them in boates by
the ſea vnto Ioppe: and it ſhal be thy part to tranſport them into
Ieruſalem. \V Salomon therfore numbred al the men, that were proſelytes
in the land of Iſrael, after the numbering which Dauid his father
numbered, and they were found an hundred fiftie three thouſand and ſix
hundred. \V And he made of them ſeuentie thouſand, that ſhould carie
burdens on their ſhoulders, and eightie thouſand, that ſhould cut ſtones
out of the mountaynes: and three thouſand and ſix hundred ouerſeers of
the worke of the people.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
