%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 0915
%%% o-0822
\startChapter[
  title={Chapter 18}
  ]

\Summary{Ioſaphat ioyned in affinitie with wicked Achab king of Iſrael,
  goeth with him againſt Ramoth Galaad, four hundred falſe prophetes
  promiſing victorie.
%%% 0916
14.~Micheas prophecying the contrarie, 25.~is put in priſon. 28.~Achab
(notwithſtanding he changeth his attire, and leaueth Ioſaphat in danger)
33.~is ſlaine.}

Ioſaphat therfore was rich and verie glorious, and was ioyned in
affinitie to Achab. \V And he went downe to him after certaine yeares
into Samaria: at whoſe coming Achab killed muttons, and oxen very manie
for him and the people that came with him: and he perſwaded him to goe
vp into Ramoth Galaad. \V And Achab the king of Iſrael ſayd to Ioſaphat
the king of Iuda: Come with me into Ramoth Galaad. To whom he anſwered:
As I am thou alſo: as thy people; ſo my people alſo: and
\SNote{For this he was iuſtly reprehended, but his ſimple intention
diminiſhed his fault, and ſo he was more eaſily pardoned, & for his good
workes otherwiſe donne in Gods ſeruice.
\XRef{ch.~19.~3.}}
we wil be with thee in battel. \V Ioſaphat ſayd to the king of Iſrael:
Conſult I beſech thee preſently the word of our Lord. \V Therfore the
king of Iſrael gathered together of the prophetes four hundred men, and
ſayd to them: Shal we goe into Ramoth Galaad to fight, or ſitte ſtil?
But they ſayd: Goe vp, ſay they, and God wil deliuer it into thy
hand. \V And Ioſaphat ſayd: Is there not here a prophete of our Lord,
that we may enquire alſo of him? \V And the king of Iſrael ſayd to
Ioſaphat: There is one man, of whom we may aske the wil of our Lord:
but I hate him, becauſe he doth not prophecie me good, but euil at al
times: and it is Micheas the ſonne of Iemla. And Ioſaphat ſayd: Speake
not in this maner ô king. \V The king of Iſrael therfore called one of
the Eunuches, and ſayd to him: Cal quickly Micheas the ſonne of
Iemla. \V Moreouer the king of Iſrael,
%%% o-0823
and Ioſaphat the king of Iuda, both ſate in their thrones, clothed with
kinglie attyre, and they ſate in the court beſide the gate of Samaria,
and al the prophetes prophecied before them. \V But Sedecias the ſonne
of Chanaana made him hornes of yron, and ſayd: Thus ſayth our Lord: With
theſe ſhalt thou ſtrike Syria, til thou deſtroy it. \V And al the
prophetes in like maner prophecied, and ſayd: Goe vp into Ramoth Galaad,
and thou ſhalt proſper, and our Lord wil deliuer them into the kinges
hand. \V And the meſſenger that went to cal Micheas, ſayd to him: Behold
the wordes of al the prophetes with one mouth tel the king good thinges:
I beſech thee therfore that thy word alſo diſſent not from them, and
that thou ſpeake proſperous thinges. \V To whom Micheas anſwered: Our
Lord liueth, whatſoeuer my God ſhal ſay to
%%% 0917
me, that I wil ſpeake. \V He therfore came to the king. To whom the king
ſayd: Micheas, ſhal we goe into Ramoth Galaad to fight, or ſit ſtil? To
whom he anſwered:
\SNote{See
\XRef{Annotation 3.~Reg.~22. v.~15.}}
Goe ye vp: for al thinges ſhal fal out proſperous, and the enemies ſhal
be deliuered into your handes. \V And the king ſayd: Agayne, and agayne
I adiure thee, that thou ſpeake not to me, but that which is true in the
name of our Lord. \V But he ſayd: I ſaw al Iſrael diſperſt in the
mountaynes, as ſheepe without a ſhepheard: and our Lord ſayd: Theſe haue
no maiſters: let euerie man returne into his houſe in peace. \V And the
king of Iſrael ſayd to Ioſaphat: Did I not tel thee that this man did
not prophecie me anie good, but theſe thinges that be euil? \V But he
ſayd: Heare ye therfore the word of our Lord: I ſaw our Lord ſitting in
his throne, and al the hoſt of heauen aſſiſting him on the right hand
and on the left. \V And our Lord ſayd: Who ſhal deceiue Achab the king
of Iſrael, that he may goe vp and fal in Ramoth Galaad? And when one
ſayd in this maner, and an other otherwiſe: \V there came forth a
ſpirit, and ſtood before our Lord, and ſayd: I wil deceiue him. To whom
our Lord ſayd: Wherin wilt thou deceiue him? \V But he anſwered: I wil
goe forth, and wil be a lying ſpirit in the mouth of al his
prophetes. And our Lord ſayd: Thou ſhalt deceiue, and ſhalt preuaile:
goe forth, and doe ſo. \V Now therfore, behold our Lord hath geuen the
ſpirit of lying in the mouth of al thy prophetes, and our Lord hath
ſpoken of thee euil thinges. \V And Sedecias the ſonne of Chanaana came,
and ſtroke the cheeke of Micheas, and ſayd: Which way paſſed the ſpirit
of our Lord from me, that it ſhould ſpeake to thee? \V And Micheas ſayd:
Thou thy ſelf ſhalt ſee in that day, when thou ſhalt enter into chamber
out of chamber, to be hid. \V And the king of Iſrael commanded, ſaying:
Take Micheas, and lead him to Amon the gouernour of the citie, and to
Ioab the ſonne of Amelech. \V And you ſhal ſay: Thus ſayth the king:
Caſt this felow into priſon, and giue him a litle bread, and a litle
water, til I returne in peace. \V And Micheas ſayd: If thou returne in
peace, our Lord hath not ſpoken in me, & he ſayd: Heare al ye
peoples. \V Therfore the king of Iſrael, and Ioſaphat the king of Iuda
went vp into Ramoth Galaad. \V And the king of Iſrael ſayd
%%% o-0824
to Ioſaphat: I wil change my habite, and ſo I wil goe to the fight, but
be thou clothed with thine owne garmentes. And
%%% 0918
the king of Iſrael changing his habit came to the battel. \V And the
king of Syria had commanded the captaynes of his horſemen ſaying: Fight
not agaynſt the leaſt, or agaynſt the greateſt, but agaynſt the king of
Iſrael only. \V Therfore when the princes of the horſemen ſaw Ioſaphat,
they ſayd: This is the king of Iſrael. And they compaſſed him fighting:
but he cried to our Lord, and he holpe him, and turned them away from
him. \V For when the captaynes of the horſemen ſaw, that
\Fix{is}{it}{obvious typo, same in other}
was not the king of Iſrael, they leaft him. \V And it chanced that one
of the people ſhot an arrow at aduenture, and ſtroke the king of Iſrael
betwen the necke and the ſhoulders, and he ſayd to his cocher: Turne thy
hand, and carie me out of the battel, becauſe I am wounded. \V And the
fight was ended in that day: moreouer the king of Iſrael ſtood in his
chariote agaynſt the Syrians vntil euening, and died at the ſunne
ſette.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
