%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 30
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-30


\project douay-rheims


%%% 0938
%%% o-0842
\startChapter[
  title={Chapter 30}
  ]

\Summary{Ezechias by meſſengers and letters exhorteth the people both of
  Iuda and Iſrael, to make Paſch in Ieruſalem.  11.~Which ſome of
  Iſrael, and al Iuda performe, 18.~the fourtenth day of the ſecond
  moneth, though al could not be purified according to the law. 23.~They
  make an other feaſt of Azimes ſeuen dayes more, the king and princes
  geuing hoſtes to the people.}

Ezechias alſo ſent to al Iſrael and Iuda: and he wrote letters to
Ephraim and Manaſſes, that they ſhould come
%%% 0939
to the houſe of our Lord in Ieruſalem, and should make a Phaſe to our
Lord the God of Iſrael. \V Counſel therfore being taken of the king and
the princes, and of al the aſſemblie of Ieruſalem, they decreed to make
the Phaſe the ſecond moneth. \V For they could not make it in his time:
becauſe the Prieſtes that might ſuffiſe, had not bene ſanctified, and
the people had not as yet bene gathered into Ieruſalem. \V And the word
pleaſed the king, and al the multitude. \V And they decreed to ſend
meſſengers into al Iſrael from Berſabee vnto Dan, that they should come,
and make the Phaſe of our Lord the God of Iſrael in Ieruſalem: for manie
had not made it as is preſcribed by the law. \V And the poſtes went
forth with letters of commandement from the king, and his princes, into
al Iſrael and Iuda, according to that, which the king had commanded,
proclaiming: Children of Iſrael returne ye to our Lord the God of
Abraham, and Iſaac, and Iſrael: and he wil returne to the remnant, that
hath eſcaped the hand of the king of the Aſſyrians. \V Become not as
your fathers, and brethren, which haue reuolted from our Lord the God of
their fathers, who hath deliuered them into deſtruction, as
%%% o-0843
your ſelues ſee. \V Harden not your neckes, as your fathers: geue handes
to our Lord, and come to his Sanctuarie, which he hath ſanctified for
euer: ſerue our Lord the God of your fathers, and the wrath of his furie
shal be turned away from you. \V For if you shal returne to our Lord:
your brethren, and children shal haue mercie before their Lordes, that
haue led them captiue, and they shal returne into this land: for our
Lord your God is merciful, and wil not turne away his face from you, if
you shal returne to him. \V Therfore the poſtes went forward ſpedely
from citie to citie, through the land of Ephraim, and of Manaſſes, as
farre as Zabulon, they mocking and ſkorning them. \V Neuertheleſſe
certayne men of Aſer, and Manaſſes, and Zabulon, condeſcending to the
counſel, came to Ieruſalem. \V But the hand of God was in Iuda, to geue
them one hart to doe the word of our Lord, according to the precept of
the king and of the princes. \V And much people was gathered into
Ieruſalem to make the ſolemnitie of Azimes in the ſecond moneth: \V And
riſing they deſtroyed the altars that were in Ieruſalem, and
ouerthrowing al thinges wherin incenſe was burnt to idols, they threw it
into the Torrent cedron. \V And they immolated
%%% 0940
the Phaſe the fourtenth day of the ſecond moneth. The Prieſtes alſo and
the Leuites at length being ſanctified offered holocauſtes in the houſe
of our Lord. And they ſtoode in their order according to the
diſpoſition, & law of Moyſes the man of God: but the Prieſtes receiued
the bloud to be powred out of the handes of the Leuites, \V becauſe a
great multitude was not ſanctified: & therfore the Leuites immolated the
Phaſe for them, that came not in time to be ſanctified to our Lord. \V
For a great part of the people of Ephraim, and Manaſſes, and Iſſachar,
and Zabulon, that had not bene ſanctified, did eate the
Phaſe, not according to that which is writen: and Ezechias prayed for
them, ſaying: Our good Lord wil be merciful, \V to al them,
\SNote{Being penitẽt in hart for their ſinnes, Gods diſpenſation might
be ſuppoſed for legal purification in caſe of neceſitie, which otherwiſe
was ſtrictly commanded.
\XRef{Leuit.~5. et~ſeq.}
\XRef{Deut.~27.~&c.}}
that in al their hart ſeeke our Lord the God of their fathers: and wil
not impute it to them that they are not ſanctified. \V Whom our Lord
heard, and was pacified to the people. \V And the children of Iſrael,
that were found in Ieruſalem, made the ſolemnitie of Azimes ſeuen daies
in great ioy, prayſing our Lord euerie day. The Leuites alſo and the
Prieſtes by inſtrumentes, that agreed to their office. \V And Ezechias
ſpake to the hart of al the Leuites, that had good vnderſtanding
concerning our Lord: and they did eate during the ſeuen daies of the
ſolemnitie, immolating victimes of pacifiques, and prayſing our Lord the God of
their fathers. \V And it pleaſed the whole multitude to celebrate it
\SNote{Voluntarie workes of ſupererogation, more then was commanded.}
other ſeuen daies: which alſo they did with great ioy. \V For Ezechias
the king of Iuda had geuen the multitude a thouſand oxen, and ſeuen
thouſand ſheepe: but the princes had geuen the people oxen a thouſand, &
ſheepe ten thouſand: there was ſanctified therfore a verie great
multitude of Prieſtes. \V And al the multitude of Iuda was ful of mirth,
as wel of the Prieſtes and Leuites, as of al the aſſemblie, that came
out of Iſrael; of the
%%% o-0844
proſelites alſo of the land of Iſrael, and them that dwelt in Iuda. \V
And there was made a great ſolemnitie in Ieruſalem, ſuch as had not bene
in that citie from the dayes of Salomon the ſonne of Dauid the king of
Iſrael. \V And the Prieſtes & the Leuites roſe vp bleſſing the people: &
their voice was heard: and their prayer came into the holie habitation
of heauen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
