%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-23


\project douay-rheims


%%% 0925
%%% o-0830
\startChapter[
  title={Chapter 23}
  ]

\Summary{Ioiada the Hiegh prieſt annoynteth and crowneth Ioas king,
  12.~cauſeth Athalia to be ſlaine, 16.~idolatrie to be deſtroyed,
  18.~and Gods ſeruice aduanced.}

And
\CNote{\XRef{4.~Reg.~11.}}
in the ſeuenth yeare Ioiada taking courage, tooke the centurions, to
witte, Azarias the ſonne of Ieroham, and Iſmahel the ſonne of Iohanan,
Azarias alſo the ſonne of Obed, and Maaſias the ſonne of Adaias, and
Eliſaphat the ſonne of Zechri: and made a couenant with them. \V Who
going about Iuda, gathered together the Leuites out of al the cities of
Iuda, and the princes of the families of Iſrael, and they came into
Ieruſalem. \V Therfore al the multitude made a couenant with the king in
the houſe of God: and Ioiada ſayd to them: Behold the kinges ſonne ſhal
reigne, as our Lord hath ſpoken, vpon the ſonnes of Dauid. \V This
\SNote{Gods promiſe being abſolute and certayne, yet humane meanes were
neuertheles required.}
therfore is the thing which you ſhal doe. \V The third part of you that
come to the
\TNote{The
\Fix{vvekelie}{vveekelie}{likely typo, fixed in other}
vvatch.}
Sabbath of the Prieſtes, and of Leuites, and of porters ſhal be in the
gates: and a third part at the kinges houſe: and a third at the gate,
which is called of the Fundation: but let al the reſt of the common
people be in the courtes of the houſe of our Lord. \V Neither let anie
other enter into the houſe of our Lord, but the Prieſtes, and they that
miniſter of the Leuites: let them onlie goe in, becauſe they are
ſanctified and let al the reſt of the multitude obſerue the watches of
our Lord. \V And let the Leuites enuiron the king, hauing euerie one
their weapons: (and if anie other ſhal enter into the temple, let him be
ſlayne) and let them be with the king both coming in, and going out. \V
The Leuites therfore, & al Iuda did according to al thinges, which
\SNote{In caſe of right and neceſſitie we ſee here what the high Prieſt
could do and did by his authoritie: who otherwiſe intermedled not in
the kings affayres.
\XRef{ch.~19. v.~11.}}
Ioiada the high Prieſt had commanded; and they tooke euerie one the men
that were vnder them, and came by the order of the Sabbath, with them
that had fulfilled the Sabbath, and were to goe forth. For Ioiada the
high Prieſt permitted not the companies to depart, which were accuſtomed
to ſuccede one an other euerie weeke. \V And Ioiada the Prieſt gaue to
the centurions the ſpeares, and  the ſhieldes, and targattes of king
Dauid, which he had conſecrated in the houſe of our Lord. \V And he
appoynted al the people of them that held weapons on the right ſide of
the temple, vnto the left ſyde of the temple, before the altar, and the
temple, round about the
%%% 0926
king. \V And they brought forth the kinges ſonne, and put the crowne
vpon him, and the teſtimonie, and gaue the law to be in his hand, & they
made him king: Ioiada alſo the high Prieſt, and his ſonnes annoynted
him: and they wiſhed him wel, and ſaid: God ſaue
%%% o-0831
the king. \V Which thing when Athalia had heard, to witte, the voice of
them that ranne and prayſed the king, ſhe went in vnto the people, into
the temple of our Lord. \V And when ſhe had ſeene the king ſtanding vpon
the ſteppe in the entrance, and the princes, and the companies about
him, and al the people of the land reioyſing, and ſounding with
trumpettes, and playing on inſtrumentes of diuerſe kind, and the voice
of them that prayſed, ſhe rent her garmentes, and ſayd: Treaſon,
treaſon. \V And Ioiada the high Prieſt going forth to the centurions,
and captaines of the armie, ſayd to them: Bring her forth without the
precinct of the temple, and let her be killed with the ſword
without. And the Prieſt commanded that ſhe ſhould not be killed in the
houſe of our Lord. \V And they layd handes vpon her necke: and when ſhe
was entred within the gate of the horſes of the kinges houſe, they
killed her there. \V And Ioiada made a couenant betwen himſelf, and al
the people, and the king, that they would be the people of our Lord. \V
Al the people therfore entred into the houſe of Baal, and deſtroyed it:
and they brake his altars and
\SNote{They are wilfully blind that wil not ſee difference betwen images
of Baal & of Chriſt or of Sainctes.}
his
\TNote{\L{ſimulacra}.}
images: Mathan alſo the prieſt of Baal they ſlewe before the altars. \V
And Ioiada appoynted ouerſeers in the houſe of our Lord, vnder the
handes of the Prieſtes, & the Leuites, which Dauid diſtributed in the
houſe of our Lord: that they ſhould offer holocauſtes to our Lord, as it
is writen in the law of Moyſes, in ioy and ſonges, according to the
diſpoſition of Dauid. \V He appoynted alſo porters in the gates of the
houſe of our Lord, that the vncleane in anie thing ſhould not enter
in. \V And he tooke the centurions, and the moſt valiant men and princes
of the people, and al the common people of the
land, and they made the king to goe downe from the houſe of our Lord, &
to enter by the middes of the vpper gate into the kinges houſe, and
placed him in the royal throne. \V And al the people of the land
reioyced, & the citie was quiet: moreouer Athalia was ſlayne with the
ſword.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
