%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 33
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-33


\project douay-rheims


%%% 0945
%%% o-0848
\startChapter[
  title={Chapter 33}
  ]

\Summary{Manaſſes for his manifold wickednes is led captiue into
  Babylon: 12.~repenteth in priſon, is reſtored to his kingdom, and
  deſtroyeth idolatrie: 18.~dieth, 23.~& Amon ſucceding, 24.~is ſlayne
  by his ſeruantes, and Ioſias reigneth.}

Twelue
\CNote{\XRef{4.~Reg.~21.}}
yeares old was Manaſſes when he began to reigne, and he reigned fiftie
fiue yeares in Ieruſalem. \V And he did euil before our Lord, according
to al the abominations of the nations, which our Lord ouerthrew before
the children of Iſrael: \V and being turned, he reedified the excelſes
which Ezechias his father had deſtroyed: and he built altars to Baalim,
and made groues, and he adored al the hoſt of heauen, and worſhipped it. \V
He built alſo altars in the houſe of our Lord, wherof our Lord had ſayd:
In Ieruſalem ſhal my name be for euer. \V And he built them to al the hoſt
of heauen in the two courtes of the houſe of our Lord. \V And he made his
ſonnes to paſſe through fyre, to the Valebenennom: he obſerued dreames,
folowed ſouthſayings, gaue him ſelf to magike artes, he had with him
magicians, & inchaunters: and he wrought manie euils before our Lord, to
prouoke him. \V Alſo a grauen, and a molten ſigne he put in the houſe of
God, wherof our Lord ſpake to Dauid, and to Salomon his ſonne, ſaying:
In this houſe, and in Ieruſalem, which I haue choſen
\Fix{al}{of al}{obvious typo, fixed in other}
the tribes of Iſrael, wil I put my
%%% 0946
name for euer. \V And I wil not make the foote of Iſrael to moue out of
the land, which I haue deliuered to their fathers: yet ſo if they ſhal
obſerue to doe that I haue commanded them, and al the law, and the
ceremonies, and iudgementes by the hand of Moyſes. \V Manaſſes therfore
ſeduced Iuda, and the inhabitantes of Ieruſalem, that they did euil
aboue al the nations, which our Lord had ouerthrowen before the face of
the children of Iſrael. \V And our Lord ſpake to him, and to his people,
and they would
%%% o-0849
not attend. \V Therfore he brought vpon them the princes of the hoſt of
the king of the Aſſyrians: and they tooke Manaſſes, & led him bound with
chaynes, and fetters into Babylon. \V Who after that he was in
diſtreſſe, prayed our Lord his God: and did penance excedingly before
the God of his fathers. \V And he deſired him, and beſought him
earneſtly:
\SNote{A pregnant example of the effect of harty repentance.}
and he heard his prayer, and brought him agayne to Ieruſalem into his
kingdom, and Manaſſes knew that our Lord he was God. \V After theſe
thinges he built a wal without the citie of Dauid, on the weſt of Gihon
in the valley, from the entrance of the
\Fix{fifh}{fifth}{obvious typo, fixed in other}
gate round about vnto Ophel, and rayſed it excedingly: and he appoynted
princes of the hoſt in al the fenſed cities of Iuda: \V and he tooke
away ſtrange goddes, and the idol out of the houſe of our Lord: the
altars alſo which he had made in the mount of the houſe of our Lord, and
in Ieruſalem, and he threw al without the citie. \V Moreouer he
reedified the altar of our Lord, and immolated vpon it victimes, and
pacifiques, and prayſe: and he commanded Iuda to ſerue our Lord the God
of Iſrael. \V Neuertheleſſe as yet the people immolated in the excelſes
to our Lord their God. \V But the reſt of the actes of Manaſſes: and his
obſecration to his God: the wordes alſo of the Seers, that ſpake to him
in the name of our Lord the God of Iſrael, are conteyned in the wordes
of the kinges of Iſrael. \V His
\SNote{This prayer is not extant in the Hebrew but in Greke & Latin, as
yet neither receiued for canonical, by the Church nor refuſed.}
prayer alſo, and his obtayning, and al his ſinnes, and contempt, the
places alſo wherein he built excelſes, and made groues, and ſtatues
before he did penance, are writen in the wordes of Hozai. \V Manaſſes
therfore ſlept with his fathers, and they buried him in his houſe: & his
ſonne Amon reigned for him. \V Ammon was two and twentie yeares old when
he began to reigne, and he reigned two yeares in Ieruſalem. \V And he
did euil in the ſight of our Lord, as Manaſſes his father had done: and
he immolated to al the idols, which
%%% 0947
Manaſſes his father had made, and ſerued them. \V And he did not
reuerence the face
\Fix{our}{of our}{obvious typo, fixed in other}
Lord, as Manaſſes his father did reuerence it, and he committed farre
greater ſinnes. \V And when his ſeruantes had conſpired agaynſt him,
they ſlewe him in his owne houſe. \V Moreouer the reſt of the multitude
of the people, hauing ſlaine them that ſtroke Amon, made Ioſias his
ſonne king for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
