%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 35
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-35


\project douay-rheims


%%% 0949
%%% o-0852
\startChapter[
  title={Chapter 35}
  ]

\Summary{Ioſias celebrateth a moſt Solemne Paſch. 20.~Is ſlaine by the
  king of Ægypt, al Iuda lamenting him, 25.~moſt ſpecially Ieremias.}

%%% 0950
And
\CNote{\XRef{4.~Reg.~16.}}
Ioſias made in Ieruſalem
\SNote{This Phaſe, or Paſch, made by Ioſias is famous, partly for that
this feaſt had bene omitted ſome yeares, but ſpecially for the great and
extraordinarie ſolemnitie made at this time.}
a Phaſe to our Lord, which was immolated the fourtenth day of the firſt
moneth. \V And he appoynted the Prieſtes in their offices, and exhorted
them that they would miniſter in the houſe of our Lord. \V To the
Leuites alſo, at whoſe inſtruction al Iſrael was ſanctified to our Lord,
he ſpake: Put the Arke in the Sanctuarie of the temple, which Salomon
built the ſonne of Dauid the king of Iſrael, for you ſhal carie it no
more: but now miniſter to our Lord your God, and to his people
Iſrael. \V And prepare your ſelues by your houſes, and kinredes in the
diuiſions of euerie one, as Dauid the king of Iſrael commanded, and
Salomon his ſonne deſcribed. \V And miniſter ye in the Sanctuarie by
families and Leuitical companies, \V and being ſanctified immolate the
Phaſe, prepare alſo your brethren, that they may doe according to the
wordes, which our Lord ſpake in the hand of Moyſes. \V Moreouer Ioſias
gaue to al the people, that was found there in the ſolemnitie of the
Phaſe, lambes and kiddes of the flockes, and of the reſt of the cattel
thirtie thouſand, of oxen alſo three thouſand, al theſe thinges of the
kinges ſubſtance. \V His dukes alſo voluntarily offered that which they
vowed, as wel to the people, as to the Prieſtes and the
Leuites. Moreouer Helcias, and Zacharias, and Iahiel princes of the
houſe of our Lord, gaue to the Prieſtes to make the Phaſe cattel one
with an other two thouſand ſix hundred, and oxen three hundred. \V And
Chonenias, and Semeias, alſo Nathanael, his brethren, moreouer Haſabias,
and Iehiel, and Iozabad princes of the Leuites, gaue to the reſt of the
Leuites to celebrate the Phaſe fiue thouſand sheepe, and oxen fiue
hundred. \V And
%%% o-0853
the miniſterie was prepared, and the Prieſtes ſtood in their office: the
Leuites alſo in companies, according to the kinges commandement. \V And
the Phaſe was immolated: and Prieſtes ſprinkled the blood with their
hand, and the Leuites drew of the skinnes of the holocauſtes: \V and they
ſeperated them to geue them by the houſes and families of euerie one,
and to be offered to our Lord, as it is writen in the Booke of Moyſes,
of oxen alſo they did in like maner. \V And they roſted the Phaſe vpon
fyre, according to that which is writen in the law: but the
pacifique hoſtes they boyled in caudrons, and kettles, and
pottes, and in haſt they diſtributed it to al the people. \V And for
themſelues, and for
%%% 0951
the Prieſtes they prepared afterward: for in oblation of holocauſtes and
of fatte the Prieſtes were occupied vntil night: wherfore the Leuites
prepared for themſelues, and for the Prieſtes the children of Aaron
laſt. \V Moreouer the ſinging men the children of Aſaph ſtood in their
order, according to the precept of Dauid, and Aſaph, and Heman, and
Idithun the prophetes of the king: and the porters watched at euerie
gate, ſo that they departed not a moment from the miniſterie: for the
which cauſe alſo their brethren the Leuites prepared meates for them. \V
Therfore al the ſeruice of our Lord was ritely accompliſhed that day, ſo
that they made the Phaſe, and offered holocauſtes vpon the altar of our
Lord, according to the precept of king Ioſias. \V And the children of
Iſrael that were found there, made the Phaſe at that time, and the
ſolemnitie of Azymes ſeuen daies. \V There was not a Phaſe like to this
in Iſrael, from the daies of Samuel the prophete: neither did anie of al
the kinges of Iſrael make a Phaſe as Ioſias, to the Prieſtes, and the
Leuites, and to al Iuda, and Iſrael that was found, and to the
inhabitantes of Ieruſalem. \V In the eightenth yeare of the kingdom of
Ioſias was this Phaſe celebrated. \V After that Ioſias had repayred the
temple, came vp Nechao the king of Ægypt to fight in Charcamis beſide
Euphrates: and Ioſias went forth to meete him. \V But he ſending
meſſengers vnto him, ſayd: What haue I to doe with thee king of Iuda? I
come not agaynſt thee this day, but I fight agaynſt an other houſe, to
the which God hath commanded me to goe in haſt: leaue to doe agaynſt
God, who is with me, leſt he kil thee. \V
\SNote{Ioſias thought that the king of Ægypt intended to inuade his
kingdome. And it was Gods wil he ſhould be ſlaine, and not ſee the
euiles that ſhould happen to the people.}
Ioſias would not returne, but prepared battel agaynſt him, neither did
he agree to the wordes of Nechao from the mouth of God, but went forward
to fight in the fielde of Mageddo. \V And there being wounded of the
Archers, he ſayd to his ſeruantes: Carie me out of the battel, becauſe I
am ſore wounded. \V Who remoued him from one chariote into an other
chariote, that folowed him after the maner of kinges, and they caried
him away into Ieruſalem, & he died, and was buried in the monument of
his fathers, and al Iuda and Ieruſalem
\SNote{Solemne exequies with lamentations and muſike.}
mourned for him. \V Ieremie moſt of al: whoſe lamentations al the
ſinging men and ſinging wemen repeate vntil this preſent day vpon
Ioſias, and it is growen as it were a law in Iſrael. Behold it is ſayd
to be writen in the lamentations. \V But the reſt of the wordes
%%% 0952
%%% o-0854
of Ioſias & of his mercies: which are commanded by the law of our
Lord: \V his workes alſo the firſt and the laſt, are writen in the Booke
of the kinges of Iuda and Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
