%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Two Paralipomenon
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 0890
%%% o-0799
\startChapter[
  title={Chapter 1}
  ]

\Summary{Salomon
\MNote{The firſt part.

Salomons reigne in the whole kingdome.}
eſtablished in the throne, offereth a thouſand hoſtes of ſacrifice,
7.~asking wiſdome, 12.~it is geuen him; with richeſſe, and temporal
glorie, which he asked not. 14.~He prouideth manie chariottes and
horſemen.}

Salomon therfore the ſonne of Dauid was ſtrengthened in his kingdom, and
our Lord was with him, and magnified him on high. \V And Salomon
commanded al Iſrael, the tribunes, and the centurions, and the dukes,
and iudges of al Iſrael, and the princes of the families: \V and he went
with al the multitude into the Excelſe of Gabaon, where was the
tabernacle of the couenant of our Lord, which Moyſes the ſeruant of God
made, in the wilderneſſe. \V For
\CNote{\XRef{1.~Paral.~13.}}
Dauid had brought the Arke of God from Cariathiarim into the place,
which he had prepared for it, and where he had pitcht a tabernacle for
it, that is, in Ieruſalem. \V The altar alſo of braſſe, which Beſeleel the
ſonne of Vri the ſonne of Hur had made was there before the tabernacle
of our
Lord: which alſo Salomon ſought, and al the aſſemblie. \V And Salomon
went vp to the altar of braſſe, before the tabernacle of the couenant of
our
%%% o-0800
Lord, and offered on it a thouſand hoſtes. \V And behold in that
verie night God appeared to him, ſaying: Aske what thou wilt that I may
geue it thee. \V And Salomon ſayd to God: Thou haſt done great mercie
with my father Dauid: and haſt made me king for him. \V Now therfore
Lord God be thy word fulfilled, which thou haſt promiſed to Dauid my
father: for thou haſt made me king ouer thy people great in number,
which is ſo innumerable as the duſt of the earth. \V Geue me wiſdom and
intelligence, that I may come in & goe out before thy people: for who
can worthely iudge this thy people, which is ſo great? \V And God ſayd
to Salomon: Becauſe this rather hath pleaſed thy hart, and thou haſt not
asked riches, and ſubſtance, and glorie, nor their liues that hate thee,
no nor manie dayes of life:
%%% 0891
but haſt deſired wiſdom and knowledge, that thou mayſt be able to iudge
my people, ouer which I haue made thee king. \V Wiſdom and knowledge are
geuen thee: and riches, and ſubſtance and glorie I wil geue thee, ſoe
that none among the kinges neither before thee, nor after thee ſhal be
like thee. \V Salomon therfore came from the Excelſe of Gabaon into
Ieruſalem before the tabernacle of couenant, & reigned ouer Iſrael. \V
And he gathered to him chariotes and horſemen, and there amounted to him
a thouſand foure hundred chariotes, and twelue thouſand horſemen: and he
cauſed them to be in the cities of the chariotes, and with the king in
Ieruſalem. \V And the king gaue ſiluer and gold in Ieruſalem
\SNote{It is the maner of holie ſcripture to expreſſe thinges exceding
vulgar capacitie by the figure \Emph{Hyperbole}.}
as ſtones, & cedar trees as ſycomores, which grow in the champayne in
great multitude. \V And there were horſes brought him from Ægypt, and
from Coa by the kinges merchantes, which went, and brought by a
price, \V a chariote of foure horſes for ſix hundred peces of ſiluer,
and an horſe for an hundred fiftie: in like maner of al the kingdomes of
Hetheites, and of the kinges of Syria market was made.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
