%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 1516
%%% o-1399
\startChapter[
  title={Chapter 06}
  ]

\Summary{Reproch, enuie, and ferocitie hinder from louing our neighbour,
  5.~ſwetenes norisheth it. 6.~A truſtie freind is much worth. 18.~Seeke
  and kepe wiſdom, with al diligence. 35.~Frequent the companie of the
  wiſe: and meditate in Gods law.}


Be not for a frend made an enemie to thy neighbour: for the euil man
shal inherite reproch and contumelie, and euerie ſinner enuious and
duble tongued. \V Extol not thyſelf in the cogitation of thy ſoule as it
were a bul: leſt perhaps thy ſtrength be quashed, \V and it eate thy
leaues, and deſtroy thy fruites, and thou be leaft as a drie tree in the
wildernes. \V For a wicked ſoule shal deſtroy him that hath it, and it
geueth him to be a ioy to his enemies, and shal lead him into the lot of
the impious. \V
\SNote{So Gedeon pacified the Ephraimites, that were incenſed againſt
him.
\XRef{Iudic.~8.}}
A ſweete word multiplieth frends, & appeaſeth enemies, and a gratious
tongue in a good man aboundeth. \V Let there be manie at peace with
thee, and let one of a thouſand be thy counſeler. \V If thou poſſeſſe a
frend, in tentation poſſeſſe him, and not eaſely credite him. \V For he
is a frend according to his owne time, and wil not abide in the day of
tribulation. \V And there is a frend that is turned to emnitie: and there
is a frend that wil diſcloſe hatred, and reproches. \V And there is a
frend companion at the table, and he wil not abide in the day of
neceſſitie. \V A frend if he continew ſtedfaſt, shal be to thee as an
equal, and in them of thy houshold shal deale confidently: \V if he
humble himſelf
%%% 1517
againſt thee, and hide himſelf from thy face, thou shalt haue
frendshippe of one accord for good. \V Be ſeperated from thine enemies,
& take heede of thy frends. \V A faithful frend, is a ſtrong protection:
and he that hath found him, hath found a treaſure. \V To a faythful
frend there is no compariſon, and there is no poyſe of gold and ſiluer
able to counteruaile the goodnes of his fidelitie. \V A faythful frend,
is the medecine of life & immortalitie: & they that feare our Lord, shal
finde him. \V He that feareth God, shal likewiſe haue good frendshipe:
becauſe according to him shal his frend be. \V Sonne, from thy youth
receiue doctrine, and euen to thy hoare heares thou shalt finde wiſdom. \V
As he that ploweth, and that ſoweth, goe to her, and expect her good
fruites. \V For in her worke thou shalt labour a litle, and shalt quickly
eate of her generation. \V How exceding sharpe is wiſdom to the vnlerned
men, and the vnwiſe wil not continew in her. \V As the vertue of a ſtone
she shal be a probation in them, and they wil not ſtay to caſt her
forth. \V For
\SNote{Manie preferre lerning before pietie, which S.~Auguſtin reproueth,
preferring the godlie before the lerned that are leſſe vertuous: The
vnlerned riſe (ſayth he) and take the kingdom of heauen, and we with our
lernings without hart, loe how we tumble in the durt.
\Cite{li.~8. c.~8. Confeſſ.}}
the wiſdom of doctrine is according to her name, and she is not manifeſt
to manie, but to whom she is knowen, she contineweth euen to the ſight
of God. \V Heare my ſonne, and take counſel of vnderſtanding, and caſt
not away my counſel. \V Thruſt
%%% o-1400
thy foote into her fetters, and thy necke into her cheynes: \V put vnder
thy shoulder, and carie her, and be not wearie of her bands. \V With al
thy minde goe to her, and wil al thy ſtrength keepe her wayes. \V Search
her out, and she shal be made manifeſt to thee, and hauing obteyned her,
forſake her not: \V for in the later end, thou shalt finde reſt in her,
and she shal be turned vnto delectation. \V And her fetters shal be to
thee for a protection of ſtrength, and foundation of powre, & her
cheynes for a ſtole of glorie. \V For the beautie of life is in her, and
her bands are a healthful bynding. \V Thou shalt put on her a ſtole of
glorie, and as a crowne of gratulation thou shalt ſet her vpon thee. \V
Sonne, if thou attend to me, thou shalt learne: and if thou wilt applie
thy minde, thou shalt be wiſe. \V If thou wilt incline thine eare, thou
shalt receiue doctrine: and if thou loue to heare, thou shalt be
wiſe. \V Stand in the multitude of wiſe ancients, and be ioyned to their
wiſdom from thy hart, that thou maiſt heare al the narration of God, and
the prayſe may not eſcape thee. \V And if thou ſee a wiſe man, watch
after him, and let thy foote weare the ſteppes of his
%%% 1518
doores. \V Haue thy cogitation in the precepts of God, and in his
commandements moſt of al be dayly conuerſant: and he wil geue thee hart,
and the deſire of wiſdom shal be geuen thee.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
