%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 1513
%%% o-1396
\startChapter[
  title={Chapter 04}
  ]

\Summary{An exhortation to practiſe workes of mercie, 12.~with
  commendation of wiſdom: 23.~(& obſeruing due times) not to omite for
  anie feare, or
  \Fix{shamfaſtnes,}{shamefaſtnes,}{possible typo, fixed in other}
  to ſay the truth, 34.~alſo to be diligent, meke, and liberal.}


Sonne defraude not the almes of the poore, and turne not away thyne eies
from the poore. \V Deſpiſe not the
%%% 1514
hungrie ſoule: and exaſperate not the poore in his pouertie. \V Afflict
not the hart of the needie, and deferre not the gift to him that is in
diſtreſſe. \V Reiect not the petition of him that is afflicted: and
turne not away thy face from the needie. \V From the poore turne not
away thine eies for anger: and
\SNote{Geue not occaſion by thy hard dealing with the poore, that they
curſe thee.}
leaue not to them that aske of thee, to curſe thee behinde thy backe. \V
For the prayer of him that curſeth thee in the bitternes of his ſoule,
\SNote{For if thou geue cauſe, God who is their protector, wil reuenge
them againſt thee.}
shal be heard: and he that made him, wil heare him. \V Make thyſelf
affable to the congregation of the poore, and to the ancient humble thy
ſoule, and to a great man bow thy head. \V Bow
%%% o-1397
downe thyne eare to the poore without ſadneſſe, and render thy debt, and
anſwer him peaceable wordes in mildenes. \V Deliuer him that ſuffereth
iniurie out of the hand of the proud: and be not faynt harted in thy
ſoule. \V In iudging be merciful to pupils as a father, & as an husband
to their mother: \V and thou shalt be as the obedient ſonne of the
Higheſt, and he wil haue mercie on thee more then a mother. \V Wiſdom
inſpireth life to her children, and receaueth them that ſeeke after her, and
wil goe before them in the way of iuſtice. \V And he that loueth her,
loueth life: and they that shal watch to her, shal embrace her
ſweetnes. \V They that shal hold her, shal inherite life: and whither
ſoeuer he shal enter, God wil bleſſe him. \V They that ſerue her, shal
be ſeruants to the holie: and them that loue her God loueth. \V He that
heareth her, shal iudge nations: and he that beholdeth her, shal
remayne confident. \V If he beleue her, he shal inherite her, and her
\SNote{They that folow wiſdom shal be ſafe.}
creatures shal be in confirmation: \V becauſe in tentation she walketh
with him, and firſt of al she chooſeth him. \V Feare and dread, and
probation she wil bring vpon him: and she wil torment him in the
tribulation of her doctrine, til she trie him in her cogitations, and
credite his ſoule. \V And she wil eſtablish him, and make a direct way
vnto him, and reioyce him, \V and wil diſcloſe her ſecretes to him, and
wil heape vpon him as treaſures knowlege and vnderſtanding of
iuſtice. \V But if he goe amis, she wil forſake him, and deliuer him
into the handes of his enimie. \V Sonne obſerue time, and auoyd from
euil. \V For thy ſoule be not ashamed to ſay the truth. \V For there is
\SNote{Through
\Fix{shamfaſtnes,}{shamefaſtnes,}{possible typo, fixed in other}
to yeld vnto ſinne: or not to reproue ſinne, is vicious.}
shame that bringeth ſinne, & there is
\SNote{But to be aſhamed, and to abhorre ſinne is very good and
neceſſarie.}
shame that bringeth glorie and grace. \V Accept no face againſt thine
owne face, nor againſt thy ſoule a lie. \V Reuerence not thy neighbour
%%% 1515
in his offence: \V nor kepe in a word in time of ſaluation. Hide not thy
wiſdom in the beautie thereof. \V For by the tongue wiſdom is diſcerned:
and vnderſtanding, and knowlege, and doctrine by the word of the wiſe,
and ſteedfaſtnes in the workes of iuſtice. \V
\SNote{Al men are bonde to ſay the truth at conuenient times,
\XRef{v.~23.}
And euer bond to auoide vntruthes.}
Doe not gayneſay the word of truth by any meanes, and be ashamed of the
lie of thyne vnskilfulnes. \V Be not ashamed to confeſſe thy ſinnes, and
ſubmite not thy ſelf to euerie man for ſinne. \V Reſiſt not againſt the
face of the mightie, neither labour againſt the ſtreame of the riuer. \V
For
\SNote{Euerie one is bond rather to loſe his life, then to do againſt
iuſtice, or to denie the truth.}
iuſtice contend for thy ſoule, and vnto death ſtriue for iuſtice, and
God wil ouerthrow thyne enemies for thee. \V Be not haſtie in thy
tongue: and vnprofitable, and remiſſe in thy workes. \V Be not as a lion
in thy houſe, ouerthrowing them of thy houshold, and oppreſſing them
that are ſubiect to thee. \V Let not thine hand be ſtretched out to
receiue, and cloſed to geue.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
