%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 1533
%%% o-1414
\startChapter[
  title={Chapter 18}
  ]

\Summary{Gods wonderful workes excede mans capacitie. 7.~Our weaknes is
  ſtreingthened by grace, 15.~wherto man muſt cooperate, 19.~by purging
  his conſcience, 22.~by prayer, 24.~by meditating Gods iudgements,
  30.~and by mortifying his owne concupiſcence.}


He that liueth for euer, created al thinges
\SNote{Al creatures according to their ſubſtantial forme in general,
were created together, though they were afterwards formed in particular
kindes, as they are diſtinctly recited in Geneſis, with the order &
ornaments of the world.
\Cite{S.~Aug. li.~4. c.~33. &~34. de Gen. ad litter.}
VVhere he expreſly affirmeth that this Scripture was written by
inſpiration of the ſame ſpirite of truth, wherby Geneſis was written.}
together. God onlie shal be iuſtified, and remaineth an inuincible king
for euer. \V Who is ſufficient to declare his workes? \V For who shal
ſearch out his glorious thinges? \V and who shal shew forth the powre of
his greatneſſe? or who shal adde to declare his mercie? \V It is not
poſſible to diminish, nor adde, neither is it poſſible to finde the
glorious workes of God. \V When a man shal haue done, then shal he
beginne: and when he shal reſt, he shal worke. \V What is man, and what
is his grace? and what is his good, or what his euil? \V The number of the
daies of men at the moſt an hundred yeares:
\CNote{\XRef{Pſal.~89. v.~10.}}
as droppes of the water of the ſea they are reputed: and as the grauel
ſtone of the ſand, ſo a few yeares in the day of eternitie. \V For this
cauſe God is patient toward them, and powreth out his mercie vpon
them. \V He hath
\Fix{ſenne}{ſeen}{obvious typo, fixed in other}
the preſumption of their hart that it is naught, and hath knowen their
ſubuerſion that it is euil. \V Therefore hath he fulfilled his
propitiation toward them, and hath shewed them the way of equitie. \V
Mans compaſſion is touching he neighbour: but the mercie of God is vpon
al flesh. \V He that hath mercie, teacheth, and inſtructeth, as a
paſtour his flocke. \V He hath mercie on him that receiueth the doctrine
of compaſſion, and he that haſteneth in his iudgementes. \V Sonne in
good deedes geue no blame, and in euerie gift geue not the ſadnes of an
euil word. \V Shal not the dew coole heate? ſo alſo a word better then a
gift. \V Is not a word aboue a good gift? but both are with a iuſtified
man. \V A foole wil vpbraide bitterly: and the gift of one vntaught
maketh the eies to drie away. \V Before iudgement prepare thee iuſtice,
and before thou ſpeake lerne. \V Before ſickenes take medicine,
\CNote{\XRef{1.~Cor.~11.}}
and before
%%% 1534
iudgement examine thyſelf, and in the ſight of God thou shalt finde
propitiation. \V Before ſickenes humble thy ſelf, and in time of
infirmitie shew thy conuerſation. \V
\CNote{\XRef{Luc.~18.~1.}
\XRef{1.~Theſ.~5.}}
Be not hindered
\SNote{They pray alwayes that pray at certaine conuenient times. And
ſtil haue intention ſo to frequent the ſame exerciſe al their life.
\Cite{S.~Aug. Epiſt.~121. c.~9. ad Probam.}}
to pray alwayes, & feare not to be iuſtified euen to death: becauſe the
reward of God abideth for euer. \V Before praier prepare thy ſoule: and
be not as a man that tempteth God. \V Remember the wrath in the day of
conſummation, and the time of reward in conuerſation of the face. \V
Remember pouertie in the time of abundance, and the neceſſities of
pouertie in the day of riches. \V From morning vnto euening time shal be
changed, & al theſe are ſowne in the eies of God. \V A wiſe man in al
thinges wil feare, & in
%%% o-1415
the daies of offences wil beware of ſloth. \V Euerie ſubtile man knoweth
wiſdom, and to him that findeth her he wil geue prayſe. \V The wiſe in
wordes, and they alſo haue done wiſely: and haue vnderſtood truth and
iuſtice, and haue rayned prouerbes & iudgementes. \V Goe not after thy
concupiſcences, and turne away from thy wil. \V If thou geue to thy
ſoule her concupiſcences, she wil make thee a ioy to
\Fix{the}{thy}{obvious typo, fixed in other}
enemies. \V Be not delighted in multitudes, nor in few: for their
concertation is continual. \V Be not poore in contention of borowing,
and thou haſt nothing in thy purſe: for thou shalt be enuious to thine
owne life.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
