%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 44
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-44


\project douay-rheims


%%% 1572
%%% o-1448
\startChapter[
  title={Chapter 44}
  ]

\Summary{Praiſes
\MNote{The 2.~part.

Examples and praiſes of holie men: with praiſe & thankes to God.}
of holie fathers in general: 16.~and in particular of Enoch, 17.~Noe,
20.~Abraham, 24.~Iſaac, and Iacob.}


Let vs praiſe glorious men, and our fathers in their generation. \V Much
glorie hath our Lord made by his magnifence from the beginning of the
world. \V Ruling in their
\SNote{Vertuous men are rightly called Lordes, and Princes, ſo the
children of Heth ſayde to Abraham: \Emph{My Lord, thou art a prince of
God among vs.}
\XRef{Gen.~23.}}
dominions, men great for force, and endued with their wiſdom, declaring
in the prophetes the dignitie of prophetes, \V and ruling in the people
that was preſent, and by the vertue of wiſdom moſt holie wordes to the
peoples. \V In their
\Fix{kil}{skil}{obvious typo, fixed in other}
ſeeking out muſical melodies, & vttering
\Fix{ſongnes}{ſongs}{obvious typo, fixed in other}
of
%%% o-1449
ſcriptures. \V Rich men in force ſtudying beautifulnes: liuing at peace
in their houſes. \V Al theſe in the generations of their nation haue
obteyned glorie, and in their daies are praiſed. \V They that were borne
of them haue leaft a name to tel their praiſes: \V and there are ſome of
whom there is no memorie: they are perished, as they that neuer were; &
are borne, as not borne at al, & their children with them. \V But they
are men of mercie, whoſe godlie deedes haue not failed: \V good thinges
continew with their ſeede, \V their nephewes are an holie inheritance,
and their ſeede hath ſtoode in the teſtamentes: \V and their children
becauſe of them abide for euer: their ſeede and their glorie shal not be
forſaken. \V Their bodies are buried in peace, and their name liueth
vnto generation and generation. \V
%%% !!! SNote marked here in both:
%%% \SNote{Enoch shal preach penance in the time of Antichriſt.}
Let peoples tel their wiſdom, and the Church declare their praiſe. \V
\CNote{\XRef{Gen.~5.}
\XRef{Gen.~6.}}
Henoch pleaſed God, and was tranſlated into paradiſe, that he may
%%% !!! SNote is alſo marked here in this, but not other. This does ſeem
%%% like the right place though.
\SNote{Enoch shal preach penance in the time of Antichriſt.}
geue repentance to the nations.
%%% 1573
\V
\CNote{\XRef{Gen.~9.}}
Noe was found
\SNote{Noe was perfect.}
perfect, iuſt, and in the time of wrath he was made a reconciliation. \V
Therefore was there a remnant leaft to the earth, when the flood was
made. \V The teſtaments of the world were made with him, that al flesh
should no more be deſtroyed with the flood. \V
\CNote{\XRef{Gen.~12.}}
Abraham the great
\SNote{Abraham father of al the beleuers in Chriſt.}
father of the multitude of the nations, and there was not found the like
to him in glorie, who kept the law of the Higheſt, and was in couenant
with him. \V
\CNote{\XRef{Gen.~22.}
\XRef{Heb.~11.}}
In his flesh he made the couenant to ſtand, and in tentation he was
found faithful. \V Therefore by an oath he gaue him glorie in his
nation, that he should encreaſe as an heape of earth, \V and that he
would exalt his ſeede as the ſtarres, and they should inherite from ſea
to ſea, and from the riuer to the endes of the earth. \V And he did in
like manner in Iſaac for Abraham his father. \V Our Lord gaue him the
\SNote{Iſaac, and Iacob were bleſſed in Abraham.}
bleſſing of al nations, and confirmed his couenant vpon the head of
Iacob. \V He knew him in his bleſſinges, and gaue him an inheritance, &
diuided him his portion in twelue tribes. \V And he preſerued vnto him
men of mercie, and found grace in the eies of al flesh.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
