%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 1531
%%% o-1412
\startChapter[
  title={Chapter 17}
  ]

\Summary{God creating man to his owne image, gaue him giftes, 9.~and
  precepts, 14.~choſe the Iſraelites for his peculiar people. 18.~Workes
  of mercie are commended to al men. 20.~Repentance to
  ſinners. 28.~Mercie is offered to al.}


God created man of the earth, and after his owne image he made him. \V
And againe he turned him into it, and conformable to himſelfe clothed
him
\SNote{Man at firſt receiued original iuſtice, by loſing wherof we al
fel into original ſinne.}
with ſtrength. \V He gaue him a number of daies and time, and gaue him
power of thoſe thinges, that are vpon the earth. \V He put his feare
ouer al flesh, and he had dominion of beaſtes and fowles. \V He created
of him an helper like to himſelf: he gaue them counſel,
%%% 1532
and tongue, and eies, & eares, and
%%% o-1413
hart to deuiſe: and he filled them with the diſcipline of
vnderſtanding. \V He created in them the knowlege of the ſpirit, he
filled their hart with vnderſtanding, and euil and good he shewed
them. \V He ſet his eie vpon their hartes to shew them the great thinges
of his workes: \V that they might praiſe the name of ſanctification: and
glorie in his meruelous workes, that
\Fix{thy}{they}{obvious typo, fixed in other}
might declare the glorious thinges of his workes. \V He added
\SNote{God gaue a precept to man to be obſerued for exerciſe of his
obedience, & ſo to be rewarded, and vnder paine of puniſhment if he
tranſgreſſed.
\XRef{Gen.~2.}}
diſcipline vnto them, and made them inherite the lawe of life. \V He
made an euerlaſting teſtament with them, & he shewed them iuſtice and
his iudgementes. \V And their eie ſaw the glorious thinges of his
honour, and their eares heard the honour of his voice, and he ſaid to
them: Beware of euerie vniuſt thing. \V And he gaue them commandment
euerie one concerning his neighbour. \V Their wayes are before him
alwaies, they are not hid from his eies. \V Ouer euerie nation he
appointed a ruler. \V
\CNote{\XRef{Rom.~13.}}
And Iſrael was made the manifeſt portion of God. \V And al their workes
as the ſunne in the ſight of God: and his eies without intermiſſion
looking on their wayes. \V The teſtamentes were not hid by their
iniquitie, and al their iniquities are in the ſight of God. \V The almes
of a man is as a ſeale with him, and shal preſerue the grace of a man as
the apple of the eie. \V And afterward he shal ariſe, and shal render
them reward, to euerie one vpon their head, and shal turne into the
inner partes of the earth. \V But to the penitent he hath geuen the way
of iuſtice, and he hath confirmed them that faile to ſuſteine, and hath
appointed to them the lot of truth. \V Turne to our Lord, and forſake
thy ſinnes: \V pray before the face of our Lord, and diminish
offences. \V Returne to our Lord, and turne away from thine iniuſtice,
and hate excedingly abomination: \V and know the iniuſtices and
iudgementes of God, and
\SNote{Perſeuerance in vertue to the end is neceſſarie which none can
merite,}
ſtand in the lot of thy purpoſe, and
\SNote{but muſt ſtil pray for it.}
of praier of the moſt high God. \V Goe into the partes of the holie
world, with the liuing, and them that geue praiſe to God. \V Tarie not
in the errour of the impious, before death confeſſe. From the dead as
nothing, confeſſion perisheth. \V Thou shalt confeſſe liuing, aliue and in
health thou shalt confeſſe, and shalt praiſe God, and shalt glorie in his
mercies. \V How great is the mercie of our Lord, and his propitiation to
them that turne to him! \V For al thinges can not be in men, becauſe the
ſonne of man is not immortal, and they haue
%%% 1533
delighted in the vanitie of malice. \V What is brighter then the ſunne?
& it shal faile. Or what more wicked then that which flesh and bloud
hath inuented? and this shal be reproued. \V He beholdeth the powre of
the height of heauen: and al men be earth and ashes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
