%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 26
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-26


\project douay-rheims


%%% 1546
%%% o-1425
\startChapter[
  title={Chapter 26}
  ]

\Summary{The praiſes of a good woman. 5.~The betraying of a citie,
  mutenie of people, and falſe accuſation are terrible, but a ielous
  woman is more greuous. 10.~Diuers il qualities of a bad woman. 16.~More
  commendations of a good woman. 25.~A doleful thing to ſee a valiant
  warier wanting liuelihood, a wiſeman not regarded, and greateſt griefe
  to ſee a iuſt man become wicked. 28.~A man ful of buſines hardly
  careth for his ſoule: and an Inne keper often ſinneth in wordes.}

The husband of a good wife is happie: for the number of his yeares is
\SNote{An vnquiet life depriueth a man of much comforth: and therfore
continual peace is as a duble life.}
duble. \V A ſtrong woman delighteth her husband, and shal accomplish the
yeares of his life in peace. \V A good woman is a good portion, in the
good portion of them that feare God shal she be geuen to a man for good
deedes: \V And the hart of rich and poore is good, at al time their
countenance is merie. \V Of three thinges my hart hath bene afraid, and
at the fourth my face hath trembled: \V The betraying of a citie, and a
gethering together of the people: \V falſe calumnie, al more greuous
then death. \V A ielous woman, is the ſorow and moorning of the hart. \V
A ielous woman is a ſcourge of the tongue, communicating
with al. \V As
\SNote{As when oxen moue, the yoke on their neckes doth moue withal: ſo
a wicked woman (to witte hereſie) can not reſt, nor let others reſt
quiet.}
a yoke of oxen, that is moued, ſo alſo a wicked woman: he that holdeth
her, is as he that taketh hold of a ſcorpion. \V A woman geuen to
drunkennes is great anger: & her contumelie and turpitude shal not be
hid. \V The fornication of a woman shal be knowen in the lifting vp of
her eies, and in her eieliddes. \V On thy daughter that turneth not away
her ſelf, ſet ſure watche: leſt occaſion found she abuſe herſelf. \V
Take heede of the impudencie of her eies, and meruel not if she contemne
thee. \V As a wayfaring man that thirſteth, wil she open
%%% o-1426
her mouth to the fountaine, and wil drinke of euerie water that is next,
and wil ſit againſt euerie hedge, and open her quiuer againſt euerie
arrow, vntil she faile. \V
\SNote{The Catholique faith is the ground of al vertues.}
The grace of a diligent woman shal delight her husband, and shal fatte
his bones. \V Her diſcipline is the gift of God. \V A wiſe and ſtil
woman, there is no exchange for a ſoule inſtructed. \V A holie, and
shamefaſt woman, is grace vpon grace. \V And al weight is not worthie a
continent ſoule. \V As the ſunne ryſing to the world in the higheſt
places of God, ſo is the beautie of a good woman for an ornament of her
houſe.
%%% 1547
\V A lampe shyning vpon the holie candleſticke, & the beautie of the face vpon
ſtayed age. \V Pillars of gold vpon feete of ſiluer, and ſtable feete
vpon the ſoules of a ſtayed woman. \V Eternal fundations vpon a ſound
rocke, and the commandments of God in the hart of a holie woman. \V At
two thinges my hart is greeued, and at the third anger is come vpon
me: \V A man of warre decaying by pouertie: and a wiſe man contemned: \V
and he that tranſgreſſeth from iuſtice to ſinne, God hath prepared him
to the ſword. \V Two ſortes haue appeared vnto me hard and dangerous, a
merchant is hardly rid of negligence: and
\SNote{Thoſe that kepe innes or ailehouſes are alwayes talking, as
willing to pleaſe al, but in much talke wanteth not ſinne.
\XRef{Prou.~10. v.~19.}}
a viteler shal not be iuſtified from the ſinnes of the lippes.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
