%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 38
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-38


\project douay-rheims


%%% 1562
%%% o-1440
\startChapter[
  title={Chapter 38}
  ]

\Summary{God hath ordained corporal, 9.~and ſpiritual
  medicines. 16.~Vſe moderate, not exceſſiue ſorow for the
  dead. 26.~Tradeſmen and artificers are neceſſarie, much more ſpiritual
  paſtors.}


Honour the phiſition
\SNote{Mans bodie naturally needeth ſuſtenãce & ſometimes phiſike, and
ſo doth his ſoule; which as it is more excellent, ſo it ought to be more
regarded.}
for neceſſitie: for the Higheſt hath created him. \V For al medicine is
of God, & it shal receiue gift of the king. \V The knowlege of the
phiſition shal exalt his head, and in the ſight of great men he shal be
praiſed.
%%% 1563
\V The Higheſt hath created medicines of the earth, and a wiſe man wil
not abhorre them. \V
\CNote{\XRef{Exo.~15.}}
Was not bitter water made ſweete by wood? \V The vertue of theſe thinges
is come to the knowlege of men, and the Higheſt hath geuen knowlege to
men, for to be honoured in his meruelous thinges. \V Curing with theſe
thinges he shal mitigate paine, and the apothecarie shal make
confections of ſwetenes, and shal make ointments of health, and his
workes shal not be conſummated. \V
\CNote{\XRef{Iſa.~38.}}
For the peace of God is vpon the face of the earth. \V Sonne
\SNote{Spiritual infirmitie requireth ſpiritual phiſike.}
in thine infirmitie contemne not thy ſelf, but pray our Lord, and he wil
cure thee. \V Turne away from ſinne, and direct thy handes, and from al
offence cleanſe thy hart. \V Geue ſweetenes and a memorial of fine
floure, and make a fat oblation, and geue place to the phiſition. \V For
our Lord created him: and let him not depart from thee, becauſe his
workes be neceſſarie. \V For there is a time when thou maiſt fal into
their handes: \V and they shal beſeech our Lord, that he directeth their
reſt, and healing; for their conuerſation. \V He that ſinneth in his
ſight, that made him, shal fal into the handes of the phiſition. \V
Sonne vpon the dead shed teares, and beginne to weepe as hauing ſuffered
doleful thinges, and according to iudgement couer his bodie, and neglect
not his burial. \V
\CNote{\XRef{Prou.~15.~&~17.}}
But
\SNote{To auoide euil ſpeach, ſhew that thou art truly ſorie for the
death of thy freind. But moderate thy ſorow, leſt it hurt thyſelf.}
for detraction beare bitterly the mourning of him one day, and be
comforted for the heauines, \V and make mourning according
%%% o-1441
to his deſert one day, or two, becauſe of detraction. \V For by heauines
death haſteneth, and it couereth the ſtrength, and ſorow of the hart
boweth the necke. \V In abſtraction ſorow is permanent: and the
ſubſtance of the poore is according to his hart. \V Geue not thine hart
into heauines, but expel it from thee: and remember the latter endes,
\V and forget not: for neither is there returne, and him thou shalt profite
nothing, and thou shalt hurt thy ſelf. \V Be mindful of my iudgement:
for thine alſo muſt be ſo: to me yeſterday, and to thee to day. \V
\CNote{\XRef{2.~Reg.~12.}}
In the repoſe of the dead make the memorie of him to reſt, and comfort
him in the departing of his ſpirit. \V
\SNote{VVhen paſtores haue leyſure from preaching they man then profite
others by writing. So S.~Paul, not only preached, but alſo writte. So
likewiſe the doctors of the Church and other holie Fathers.}
The wiſdom of a ſcribe in the time of vacance: and he that is leſſe in
action, shal rceiue wiſdom. \V With what wiſdom shal he be replenished,
that holdeth the plough, and glorieth in the goade, driueth oxen with
the prickle, and conuerſeth in their workes, and his talke is in the
breede of bulles? \V He wil
%%% 1564
geue his hart to turne vp ſorrowes, and his watching in the feeding of
kine. \V So euerie craftſman and workemaſter that paſſeth the night as
the day, that maketh grauen ſeales, and his continual diligence varieth
the picture: he wil geue his hart to the ſimilitude of the picture, and
his watching wil perfect the worke. \V So the yronſmith ſitting by the
anuil and conſidering the worke of yron. The vapour of the fire wil
parche his flesh, and he ſtriueth in the heate of the fournace. \V The
noyſe of the hammer reneweth his eare, and his eye is againſt the
ſimilitude of the veſſel. \V He wil geue his hart to the finishing of
the workes, and his watching wil polish to perfection. \V So the potter
ſitting at his worke, turning the wheele with his feete, who is alwayes
ſet in carefulnes for his worke, and al his working is in number. \V
With his arme he wil fashion the clay, and before his feete he wil bend
his ſtrength. \V He wil geue his hart to finish the vernishing thereof,
and his watching wil make cleane the fournace. \V Al theſe haue hoped in
their handes, and euerie one is wiſe in his owne art. \V Without theſe a
citie is not built. \V And they shal not inhabite, nor walke therein,
and they shal not leape high into the congregation. \V Vpon the iudges
ſeate they shal not ſitte, and the ordinance of iudgement they shal not
vnderſtand, neither shal they declare diſcipline and iudgement, and in
parables they shal not be found: \V but they shal confirme the creature
of the world, and their prayer shal be in the worke of their art,
applying their ſoule, & ſearching in the law of the Higheſt.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
