%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 28
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-28


\project douay-rheims


%%% 1548
%%% o-1427
\startChapter[
  title={Chapter 28}
  ]

\Summary{Abſtaine from reuenge, 8.~and ſtrife, 15.~from making debate,
  28.~from hearing, and ſpeaking euil.}


He
\SNote{He that ſeeketh reuenge contrarie to the courſe of iuſtice, or of
euil intention, or of rancor of mind, ſinneth greuouſly.}
that wil be reuenged, shal finde reuenge of our Lord, and keepeing he
wil keepe his ſinne. \V
%%% !!! Different set in other. Add those?
\CNote{\XRef{Mat.~6.}
\XRef{Rom.~12.}}
\SNote{Charitie requireth that we remitte iniuries with three
conditions, if the offender be truly penitent, if the remiſſion of
puniſhment be not againſt iuſtice, nor againſt neceſſarie diſcipline.}
Forgeue thy neighbour hurting thee: & then shal thy ſinnes be looſed
%%% 1549
to thee when thou prayeſt. \V Man to man reſerueth anger, and doth he
ſeeke remedie of God? \V He hath not mercie on a man like vnto himſelf,
and doth he intreate for his owne ſinnes? \V Himſelf whereas he is
flesh, reſerueth anger, and doth he aske propitiation of God? Who by
prayer shal obteyne for his ſinne? \V Remember the laſt thinges, & ceaſe
to be at enmitie: \V for conſumption and death are imminent in his
%%% o-1428
commandmentes. \V Remember the feare of God, and be not angrie with thy
neighbour. \V Remember the teſtament of the High, & contemne the
ignorance of thy neighbour. \V Refraine thyſelf from ſtrife, and thou
shalt diminish thy ſinnes: \V for an angrie man kindleth ſtrife, and a
ſinful man wil truble his freindes, and in the middes of them that are
at peace he wil caſt in enmitie. \V
\CNote{\XRef{Pro.~26.}}
For according to the wood of the foreſt, ſo the fire burneth: and
according to the power of a man, ſo shal his anger be, and according to
his ſubſtance he wil increaſe his anger. \V Haſtie contention kindleth a
fire: and haſtie ſtrife shedeth bloud: and an il teſtifying tongue
bringeth death. \V If thou blow vpon a ſparck, it wil burne as a fire:
and if thou ſpitte thereon, it shal be quenched: both procede out of the
mouth. \V The whiſperer & duble
\Fix{tounged}{tongued}{possible typo, same in both}
is accurſt: for he hath trubled manie that were at peace. \V A
\SNote{Vndiſcrete reporte to one what an other hath ſaide, is often
cauſe of much diſcontentment, and of diſſention.}
third tongue hath moued manie, and diſperſed them from nation into
nation. \V It hath deſtroyed the walled citie of the rich, and hath
digged downe the houſes of great men. \V It hath cut the forces of
peoples, and vndone ſtrong nations. \V A third tongue hath caſt out
manlie wemen, and depriued them of their labours. \V He that regardeth
it, shal not haue reſt, neither shal he haue a freind in whom he may
repoſe. \V The ſtroke of a whippe maketh a blew marke: but the ſtroke of
the tongue wil breake the bones. \V Manie haue fallen by the edge of the
ſword, but not ſo as they that haue perished by their tongue. \V Bleſſed
is he that is couered from a wicked tongue, that hath not paſſed into
the anger therof, and that hath not drawen the yoke therof, and hath not
bene tyed in the bandes therof: \V for the yoke of it, is a yoke of
yron: and the band of it is a band of braſſe. \V The death of it, is a
moſt wicked death: and hel is more profitable then it. \V The
continuance of it shal not be permanent, but it shal obteyne the waies
of the vniuſt: and it shal not burne the iuſt in the flame therof. \V
They that forſake God, shal fal into it, and it shal
%%% 1550
burne in them, and shal not be quenched, and it shal be ſent in vpon
them as a lion, and as a leopard it shal hurt them. \V
\SNote{To heare detraction is as bad as to ſpeake it.}
Hedge thine eares with thornes, and heare not a wicked tongue, and make
doores to thy mouth, and locks. \V Lay together thy ſiluer, and make
balance to thy wordes, and right bridles to thy mouth: \V and take heede
leſt perhaps thou ſlippe in thy tongue, & fal in the ſight of the
enemies, that lie in wayte for thee, and thy fal be vncurable vnto
death.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
