%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 43
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-43


\project douay-rheims


%%% 1570
%%% o-1447
\startChapter[
  title={Chapter 43}
  ]

\Summary{Gods incomparable excellencie appeareth in the heauens; 2.~in
  the ſunne, 6.~moone, 9.~other ſtarres, 12.~rainebow, 14.~ſnow,
  15.~cloudes, 16.~hayle, 17.~winde, 18.~thunder, 21.~froſt,
  22.~chriſtal, 24.~dew, 26.~the ſea, and innumerable thinges
  therin. 29.~No man is able to praiſe God ſufficiently.}


The firmament of height is his beautie, the beautie of heauen is in the
viſion of glorie. \V The
\SNote{Of al ſenſles creatures, yea & of ſenſible alſo, that haue not
reaſon, the ſunne is moſt excellent. Of which al corporal creatures
receiue their light: & by whoſe influence al generation of creatures
procedeth: wherof is this Maxime in Philoſophie, that \Emph{the ſunne and
man begette a man}. And Ariſtotel calleth the ſunne the father of men
and of goddes.
\Cite{li.~2. de anima}
But the faithful know it is a creature inferior to man, in reſpect of
his reaſonable ſoule, and in them both & in al other creatures,
acknowlege ſuperexcellent & infinite Maieſtie in God.}
ſunne in ſight declaring at his coming forth, a meruelous inſtrument,
the worke of the Higheſt. \V At noone it burneth the earth, and who can
abide in the preſence of the heate thereof: kepeing a fornace in the
workes of heate: \V the ſunne three times ſo much burning the
mountaines, caſting our fyrie beames, and shining with his beames
blindeth the eyes. \V Great is our
%%% 1571
Lord that made it, and by his wordes it hath haſtened his courſe. \V And
the moone in al her time, is the shewing of ſeaſon and the ſigne of
age. \V By the moone is the ſigne of a feſtiual day, a light that
diminisheth in conſummation. \V The moneth is according to her name,
increaſing meruelouſly in conſummation. \V An inſtrument of the campe
on high, shining gloriouſly in the firmament of heauen. \V The glorie of
the ſtarres is the beautie of heauen, our Lord illuminating the world on
high. \V In the words of the holie one they shal ſtand to iudgement, and
shal not faile in their watches. \V See the bow, and bleſſe him that
made it: it is very beautiful in his brightnes. \V It hath compaſſed
heauen in the circuite of his glorie, the handes of the Higheſt haue
opened it. \V But by his commandement he hath haſtened ſnow, & he
haſteth to ſend forth the lightninges of his iudgement. \V Therefore are
the treaſures opened, and the cloudes flie forth as birdes. \V By his
greatnes he hath ſet the cloudes, and the haileſtones are broken. \V In his
ſight the mountaines shal be moued, and at his wil the ſouth winde hath
blowen. \V The noyſe of his thunder
%%% o-1448
shal beate the earth, the tempeſt of
the northwind, and the gathering together of wind: \V and as the bird
lighting downe to ſitte, he ſcatereth ſnow, and the falling thereof, is
as the locuſt dyuing downeward. \V The eye shal admire the beautie of
the whitenes thereof, and the hart quaketh at the shower thereof. \V He
shal power out froſt vpon the earth as ſalt: and when it freeſeth, it
shal be made as the toppes of a thiſtle. \V The cold north winde blewe,
& of water there froſe chryſtal, vpon al gathering together of waters it
shal reſt, & as a breſt plate it shal put it ſelf vpon the waters. \V
And it shal deuoure the mountaines, & burne the deſert, & extinguish
that which is grene as fire. \V The remedie of al is in the haſtie
coming of a cloude, and a dew meeting it by the heate that cometh, shal
make it quaile. \V At his word the wind was ſtil, and with his thought
he appeaſed the depth, and our Lord planted Ilandes therein. \V They
that ſaile on the ſea, tel the perils therof: & hearing with our eares
we shal meruel. \V 
%%% !!! SNote preſent, but not marked in both.
%%% Where should it go?
\SNote{VVhich alſo appeareth euen in the leaſt creatures whoſe natural
ſubſtances & qualities, with other accidents, the more anie man
conſidereth the more he shal admire God, the onlie Creator of al.}
There are goodly workes, & meruelous: diuers kindes of beaſtes, & of al
cattel, & the creature of mightie beaſtes. \V Through him is the end of
their iourney confirmed, and by his word al thinges are ſet in order. \V
We shal ſay manie thinges, & shal faile in wordes; but the ſumme of our
wordes is, he is in al. \V Glorifying
%%% 1572
him how far shal we be able? for the omnipotent himſelf is aboue al his
workes. \V Our Lord is terrible, & exceeding great, & his might is
meruelous. \V Glorifying our Lord as much as euer you can, he shal yet
ſurpaſſe, and his magnificence is meruelous. \V Bleſſing our Lord, exalt
him as much as you can: for he is greater then al praiſe. \V Exalting
him be ye replenished with ſtrength. Labour not: for you shal not
comprehend. \V Who shal ſee him, and shal declare him? and who shal
magnifie him as he is from the beginning? \V Manie thinges hid are
greater then theſe: for we haue ſeene few of his workes. \V But our Lord
made al thinges, and to them that liue piouſly he hath geuen wiſdom.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
