%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 40
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-40


\project douay-rheims


%%% 1566
%%% o-1443
\startChapter[
  title={Chapter 40}
  ]

\Summary{The firſt matter of ſpiritual meditation may be mans miſerie,
  contracted by original ſinne, 4.~and increaſed by actual,
  \Fix{17.~reliued}{17.~relieued}{obvious typo, fixed in other}
  by Gods grace: 22.~which geueth manie benefites, 27.~man adding his
  voluntarie cooperation.}


Great trauel is created to al men, and an heauie yoke vpon the children
of Adam, from
\SNote{The penalties which al men euen new borne infants ſuffer, do ſhew
that al in general are guiltie of original ſinne, for if they were not
guiltie their puniſhment were not iuſt. That Chriſt alſo would be
ſubiect to the ſame penalties was for the ſinnes of others. And though
his B.~mother vvas preſerued from this ſinne yet ſhe vvas not exempted
from the general penaltie of al mankinde.}
the day of their coming forth of their mothers wombe, vntil the day of
their burying, into the mother of al. \V Their cogitations, and feares
of the hart, imagination of thinges to come, and the day of their
ending: \V from him that ſitteth vpon the glorious ſeate, vnto him that
is humbled in earth & ashes. \V From him that weareth hyacinth, and
beareth the crowne, euen to him, that is couered with rude linen: furie,
enuie, tumult, wauering, and the feare of death, anger perſeuering, and
contention, \V and in the time of repoſe in bed, the ſleepe of night
changeth his knowlege. \V A litle is as nothing in reſt, and afterward
in ſleepe, as in the day of watche. \V He is trubled in the viſion of
his hart, as he that hath eſcaped in the day of battel. In the time of
his ſafetie he roſe vp, and merueleth at noe feare. \V With al flesh,
from man euen to beaſt, and vpon ſinners ſeuenfold.
%%% o-1444
\V Beſide theſe thinges, death, bloud, contention, and ſword,
oppreſſions, famine,
%%% 1567
and contrition, and ſcourges: \V
for the wicked al theſe were created, and
\CNote{\XRef{Gen.~7.}}
for them the floud was made. \V
\CNote{\XRef{Eccle.~2.}}
Al thinges that are of the earth, shal turne into the earth, and al
waters shal returne into the ſea. \V Al bribing, and iniquitie shal be
cleane taken away, and fidelitie shal ſtand for euer. \V The riches of
the vniuſt shal be dried vp as a riuer, and they shal ſound as great
thunder in rayne. \V In opening his handes he shal reioyce: ſo
tranſgreſſors shal pine away in conſumption. \V The nephewes of the
impious shal not multiplie boughes, nore vncleane rootes ſound vpon the
toppe of a rocke. \V Ouer al water
\Fix{grennes,}{greennes,}{obvious typo, fixed in other}
and at the brincke of the riuer it shal be plucked vp before al
graſſe. \V Grace is as paradiſe in bleſſinges, and mercie remayneth for
euer. \V The life of a workeman that is ſufficient for himſelf shal be
ſweete, and in it thou shalt finde a treaſure. \V Children, and building
of a citie shal confirme the
name, and an vnſpotted woman shal be counted aboue this. \V Wine and
muſicke make a ioyful 
hart: and the loue of wiſedom is aboue both. \V Shalmes, and Pſalterie
make ſweete melodie, and a ſweete tongue is aboue both. \V Thine eye wil
deſire grace and beautie, and
\SNote{The ſocietie of Chriſts Church floriſhing in al vertues excelleth
the benefites of the old Teſtament.}
greene ſowen fieldes are aboue this. \V A freind and companion meeting
together in time, and aboue them both is a woman with her husband. \V
Bretheren are an helpe in the time of tribulation, and mercie shal
deliuer more then they. \V Gold and ſiluer are the eſtablishing of the
feete: and counſel is wel accepted aboue them both. \V Riches and
ſtrength exalt the hart, and aboue theſe is the feare of our Lord. \V
There is no diminution in the feare of our Lord, and in it there is no
neede to ſeeke for helpe. \V The feare of our Lord is as a paradiſe of
bleſſing, and they haue couered it aboue al glorie. \V Sonne in thy life
time
\SNote{Euerie one is bound to labour that he want not neceſſaries.}
want not: for it is better to die then to want. \V A man that looketh
toward an other mans table, his life is as no life, thinking how to
liue, for he feedeth his ſoule with an other mans meates. \V But a man
nurtered, and taught wil looke to him ſelfe. \V Pouertie wil be ſweete
in the mouth of the vnwiſe, and in his bellie a fire wil burne.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
