%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 47
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-47


\project douay-rheims


%%% 1576
%%% o-1452
\startChapter[
  title={Chapter 47}
  ]

\Summary{Praiſes of Nathan, 2.~Dauid, 14.~and Salomon; in whoſe progenie
  (21.~notwithſtanding his fal) 27.~the royal ſcepter remained for
  Dauids ſake, though for his and the peoples ſinnes, tenne tribes were
  cut of, and fel into ſchiſme.}


After
\CNote{\XRef{2.~Reg.~11.}}
theſe thinges aroſe Nathan the Prophet in the daies of Dauid. \V And as
the fatte ſeparated from the flesh, ſo was Dauid from the children of
Iſrael. \V
\CNote{\XRef{1.~Reg.~17.}}
He plaied with lyons as it were with lambes: and with beares he did in
like maner as with lambes of sheepe in his youth. \V
\CNote{\XRef{Ibidem.}}
Did not he kil the giant, and tooke away reproch from his nation? \V In
lifting vp his hand, with a ſtone of the ſling he ouerthrew the boaſting
of Goliah: for he inuocated our Lord the omnipotent, and he gaue in his
right hand, to take away the man ſtrong in battel, and to exalt the horne
of his nation. \V
\CNote{\XRef{1.~Reg.~18.}}
So in ten thouſand did he glorifie him, and praiſed him in the
bleſſinges of our Lord, in offering to him a crowne of glorie: \V for he
deſtroyed the enemies on euerie ſide, and rooted out the Philiſthijms
the aduerſaries euen vntil this preſent day: he brake their horne for
euer. \V In euerie worke he gaue confeſſion to the Holie one, and to the
Higheſt, in the word of glorie. \V From
\SNote{Amongſt al the renoumed actes of Dauid his pure and ſincere hart
moſt pleaſed God.}
al his hart he praiſed our Lord, & loued
%%% 1577
God that made him: and gaue him might againſt his enemies: \V
\CNote{\XRef{1.~Par.~25.}}
and he made ſingers to ſtand before the altar, and by their ſound he
made ſweete tunes. \V And in the
\Fix{ſolennities}{ſolemnities}{likely typo, fixed in other}
he gaue honour, and adorned the times euen to the end of his life, that
they should praiſe the holie name of our Lord, and magnifie the holines
of God in the morning. \V
\CNote{\XRef{2.~Reg.~11.}}
Our Lord purged his ſinnes, and exalted his horne for euer: and he gaue
him a teſtament of the kingdom, and the ſeate of glorie in Iſrael. \V
After him aroſe a wiſe ſonne, and for
\SNote{For Dauids ſake God gaue wiſdom to Salomon, and peace in his
kingdom.}
him did he ouerthrowe al the might of the enemies. \V
\CNote{\XRef{3.~Reg.~3.}}
Salomon reigned in dayes of peace, to whom God ſubdewed al his enemies,
that he might build an houſe in his name, and prepare holines for euer:
as
\SNote{By Apoſtrophe the auctor turneth his ſpeach to Salomon.}
thou art inſtructed in thy youth. \V
\CNote{\XRef{3.~Reg.~3.}}
And thou art replenished as a riuer with wiſdom, and thy ſoule
diſcouered the earth. \V And thou didſt multiplie darke ſayinges in
compariſons: thy name was bruited to the ilandes far of, and thou waſt
beloued in thy peace. \V The landes merueled at the ſonges and
prouerbes, and compariſons, and interpretations, \V and at the name of
our Lord God, whoſe name is, God of Iſrael. \V
\CNote{\XRef{3.~Reg.~10.}}
Thou didſt gather gold as copper, and filledſt
%%% o-1453
ſiluer as lead, \V and bowdeſt thy thighes to wemen: thou haſt had power
on thy bodie, \V
\CNote{\XRef{3.~Reg.~11.}}
thou
\SNote{Salomons ſinnes were punished, but Gods mercie continued in
conſeruing his poſteritie.
\XRef{Pſal.~88.}}
haſt made a blotte in thy glorie, and profaned thy ſeede to bring wrath
to thy children, and thy follie to be kindled, \V to
make the kingdom diuided, and a ſtubburne kingdom to reigne of
Ephraim. \V
\CNote{\XRef{Pſal.~88. v.~34.}}
But God wil not leaue his mercie, and he wil not corrupt, nor abolish
his owne workes, neither wil he deſtroy from the ſtocke the nephewes of
his elect: and he wil not corrupt the ſeede of him, that loueth our
Lord. \V But he gaue a remnant to Iacob, and to Dauid of the ſame
ſtocke. \V And Salomon had an end with his fathers. \V And he leaft
after him of his ſeede, the follie of the nation, \V
\CNote{\XRef{3.~Reg.~12.}}
and Roboam hauing litle wiſedom, who turned away the nation by his
counſel, \V and Ieroboam the ſonne of Nabat, who made Iſrael to ſinne,
and made a way of ſinning to Ephraim, and their ſinnes did abound very
manie. \V They remoued them away from their land very far. \V And he
ſought al iniquities, til there came defenſe vnto them, and he rid them
from al ſinnes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
