%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 39
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-39


\project douay-rheims


%%% 1564
%%% o-1442
\startChapter[
  title={Chapter 39}
  ]

\Summary{Godlie knowlege, 16.~puritie of ſoule, 20.~humble conceipt of
  our ſelues, 27.~and conſideration of eternal reward, are good
  diſpoſitions to ſpiritual contemplation.}


The
\SNote{Beſides actiue life commended in the former chap. it is
neceſſarie that vertuous men vſe alſo meditation, & cõtemplation. Vnto
which foure diſpoſitions are requiſite, deſcribed in this chapt.}
wiſe man wil ſearch out the wiſdom of al the ancientes, and wil be
occupied in the prophetes. \V He wil keepe the narration of famous men,
and wil enter withal into the ſubtilities of parables. \V He wil ſearch
out the hidden ſenſes of prouerbes, and wil conuerſe in the ſecretes of
parables. \V In the middes of great men he wil miniſter, and in the
ſight of the preſident he shal appeare. \V He shal paſſe into the land
of ſtrange nations: for he shal trie good and euil in men. \V He wil
geue his hart to watch early vnto our Lord, that
%%% 1565
made him, and he wil pray in the ſight of the Higheſt. \V He wil open
his mouth in prayer, and wil entreate for his ſinnes. \V For if it shal
pleaſe our great Lord, he wil fil him with
\MNote{1.

True knowlege of holie Myſteries: without the which meditation wil be
erronious.}
the ſpirit of vnderſtanding: \V and he wil power forth the wordes of his
wiſdom as showres, and in prayer wil confeſſe to our Lord. \V And he wil
direct his counſel, and diſcipline, and in his ſecretes he wil
conſult. \V He wil open the diſcipline of his doctrine, and wil glorie
in the law of the teſtament of our Lord. \V Manie wil praiſe his wiſdom,
and it shal not be abolished for euer. \V The memorie of him shal not
depart, and his name shal be required from generation to generation. \V
Nations shal declare his wiſdom, and the church wil shew forth his
praiſe. \V If he continew, he shal leaue a name more then a thouſand:
and if he reſt, it shal profite him. \V I wil yet conſult that I may
declare: For as with furie I am replenished. \V In voice he ſaith: Heare
me ye diuine fruites, and as the roſe planted vpon the riuers of waters
fructifie ye. \V As Libanus
\MNote{2.

Puritie of ſoule free frõ greuous ſinnes, and endowed vvith vertues.}
haue ye the odours of ſweetnes. \V Florish ye flowres, as the lilie, and
geue forth an odour, and bring forth leaues in grace, and praiſe with
\Fix{ſoungue,}{ſong,}{possible typo, fixed in other}
and bleſſe our Lord in his workes. \V
\MNote{3.

Humilitie is expecially required in contemplatiue perſons.}
Geue magnificence to his name, and confeſſe vnto him in the voice of your
lippes, and in
\Fix{ſoungues}{ſongs}{possible typo, fixed in other}
of the lippes, and harpes, & thus shal ye ſay in confeſſion: \V
\CNote{\XRef{Gen.~1.}}
Al the workes of our Lord are exceeding good. \V
\CNote{\XRef{Exo.~14.}}
At his word the water ſtood as an heape: and at the word of his mouth as
it were receptacles of waters: \V becauſe in his commandment
placabilitie is made, and there is no diminishing of his ſaluation. \V
The workes of al flesh are before him, and there is nothing hid from his
eyes. \V From world to world he beholdeth, and nothing is meruelous in
his ſight. \V It is not to be ſaied: What is this, or what is that? for
al thinges shal be ſought in their time. \V
\MNote{4.

Hope of eternal reward comforteth, & encoregeth the ſeruants of God.}
His bleſſing hath ouerflowed as a ſtreame. \V
\CNote{\XRef{Gen.~7.}}
And as a flood hath watered the drie land, ſo his wrath shal inherite
the nations, that haue not ſought
%%% o-1443
him: \V euen as he turned waters into
\Fix{drught,}{drought,}{obvious typo, fixed in other}
and the earth was made drie: and his waies are direct to the waies of
them: ſo to ſinners ſtumbling blockes in his wrath. \V Good thinges were
created for the good from the beginning, ſo for the wicked, good thinges
and euil. \V
\SNote{In the meane time God geueth neceſſaries for this life. VVhich
the good vſe rightly to their merite & the wicked vſe to their
damnation.}
The beginning of the thing neceſſarie for the life of men, water, fire,
and yron, ſalt, milke, and bread of flower, and honie, and the cluſter
of grape, and
%%% 1566
oyle, & clothing. \V Al theſe shal be conuerted to ſaintes into good, ſo
alſo to the impious and to ſinners into euil. \V There are ſpirites,
that were created
\SNote{Diuels were created in ſtate of grace, and of their owne wil fel
from God, & conſequently are eternally puniſhed.}
for vengeance, and in their furie they haue confirmed their
tormentes: \V in the time of conſummation they shal power out ſtrength:
and they shal accomplish the furie of him, that made them. \V Fire,
haile, famine, and death, al theſe were created for vengeance: \V the
teeth of beaſtes, and ſcorpions, and ſerpentes, and ſword reuenging the
impious vnto deſtruction. \V In his commandmentes they shal make merrie,
and on the earth they shal be prepared when nede is, and in their times
they shal not pretermitte a word. \V Therefore from the beginning I was
confirmed, and I haue conſulted, and thought, and leaft written. \V Al
the workes of our Lord are good, & he wil geue euerie worke in his
houre. \V It is not to be ſaid: This is worſe then that: for al shal be
approued in their time. \V And now with al hart and mouth praiſe ye, and
bleſſe the name of our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
