%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 46
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-46


\project douay-rheims


%%% 1575
%%% o-1451
\startChapter[
  title={Chapter 46}
  ]

\Summary{Praiſes of Ioſue, 9.~Caleb, 13.~the Iudges of Iſrael,
  16.~namely of Samuel Iudge and Prophet.}


Strong in battel was Ieſus the ſonne of Naue,
\SNote{Ioſue ſucceded in the temporal gouernment: for the ſpiritual
perteyned to the ſucceſſors of Aaron.
\XRef{Num.~27.}}
ſucceſſour of Moyſes among the prophets, who was great according to his
name, \V moſt great in the ſaluation of Gods elect, to ouerthrow the
enemies riſing vp, that he might get the
\Fix{inheritanc}{inheritance}{likely typo, fixed in other}
of Iſrael. \V What glorie obteyned he in lifting vp his handes, and
caſting ſwordes againſt the cities? \V Who before him did ſo reſiſt? for
our Lord himſelf brought the enemies. \V
\CNote{\XRef{Ioſu.~10.}}
Whether was not the ſunne hindered in his anger, and one day was made as
two? \V He inuocated the mightie ſoueraine in aſſaulting of the enemies
on euerie ſide, and the great and holie God heard him in haile ſtones of
exceeding great force. \V He made violent aſſault againſt the
nation of his enemies, and in the goeing downe he deſtroyed the
aduerſaries, \V that the nations might know his might, that it is not
eaſie to fight againſt God. And he folowed at the back of the mightie. \V
\CNote{\XRef{Num.~14.}}
And in the daies of Moyſes did mercie, and
%%% !!! SNote not marked in this, only in other. But note present in
%%% both.
\SNote{Only Ioſue and Caleb remained of thoſe which came out of Ægypt,
al the reſt dyed in the deſert, and their children entered into Chanaan.
\XRef{Num.~14.}}
Caleb the ſonne of Iephone, did ſtand againſt the enemie, and ſtayed the
nation from ſinnes, and appeaſed the murmuring of malice. \V And they
two being appointed, were deliuered out of danger from among the number
of ſix hundred thouſand footemen, to bring them into their inheritance,
into the land that yeldeth milke and honie. \V
\CNote{\XRef{Ioſu.~14.}}
And our Lord gaue ſtrength to Caleb himſelf, and his ſtrength continued
euen vntil old age, ſo that he went vp into the high place of the land,
& his ſeede obteyned inheritance. \V That al the children of Iſrael
might ſee, that it is good to obey the holie God. \V And
\SNote{Though ſome of the Iudges were ſometimes great ſinners, yet they
were finally iuſt & for their good actes much renowmed.}
al the iudges by their name, whoſe hart was not corrupted: which were
not turned away from our Lord, \V
\CNote{\XRef{1.~Reg.~17.}}
that their memorie might be bleſſed, and their bones ſpring out of their
place, \V and their name continew for euer, the glorie of the holie men
remayning vnto their children. \V The beloued of our Lord his God
%%% 1576
Samuel the prophet of our Lord, renewed the empire, and
\SNote{Samuel annointed Saul and Dauid kinges.}
anoynted princes in his nation. \V By the law of our Lord he iudged the
congregation, and the God of Iacob ſaw, and in his fidelitie was proued
a prophet. \V And he was knowen faithful in his wordes, becauſe he ſaw
the God of light: \V and inuocated our Lord omnipotent, in aſſaulting
the enemies beſetting him on euerie ſide, in the oblation of an
immaculate lambe. \V
\CNote{\XRef{1.~Reg.~7.}}
And our Lord thundered from heauen, and in great ſound he made his voice
heard, \V and he deſcomfited the princes of the Tyrians, and al the
dukes of the Philiſthiims: \V
\CNote{\XRef{1.~Reg.~12.}}
and before the time of the end of his life, and the world, he gaue
teſtimonie before
%%% o-1452
our Lord, and his Chriſt, money and whatſoeuer beſides vnto the verie
shoes he tooke not of al flesh, and no man accuſed him. \V
\CNote{\XRef{1.~Reg.~28.}}
And after this he ſlept, and
\SNote{If Samuel himſelf had not appeared, (but ſome other ſpirite) it
could not haue bene noted in his praiſes. See
\XRef{1.~Reg.~28.}}
he notified to the king, and shewed him the end of his life, and he
exalted his voice out of the earth in prophecie to take cleane away the
impietie of the nation.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
