%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 48
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-48


\project douay-rheims


%%% 1577
%%% o-1453
\startChapter[
  title={Chapter 48}
  ]

\Summary{Praiſes of Elias, 13.~Eliſeus, 19.~Ezechias, 23.~and Iſaias.}


%%% 1578
And there aroſe Elias the prophet, as it were fire, and his word burnt
as a litle torche. \V
\CNote{\XRef{3.~Reg.~17.}}
Who brought famine vpon them, and they prouoking him in their enuie,
were made fewe, for they could not abide the preceptes of our Lord. \V
By the word of our Lord he ſtayed heauen, and he brought downe fire from
heauen
\SNote{Elias procured fire from heauen to burne his ſacrifice,
\XRef{3.~Reg.~18.}
and twiſe more to burne an hundred men which perſecuted him.
\XRef{4.~Reg.~1.}}
thriſe. \V So was Elias magnified in his meruelous workes. And who can
ſo glorie like vnto thee? \V Who didſt rayſe vp the dead from hel, from
the lotte of death, in the word of our Lord God. \V Who didſt caſt downe
kinges to deſtruction, and didſt eaſily breake their might, and the
glorious from their bed. \V Who heareſt iudgement in Sina, and in Horeb
iudgementes of defence. \V
\CNote{\XRef{3.~Reg.~9.}}
Who anoynteſt kinges to repentance, and makeſt prophetes ſucceſſoures
after thee. \V
\CNote{\XRef{4.~Reg.~2.}}
Who waſt receiued in a whirlewind of fire, in a chariot of fierie
horſes.
\V
\LNote{VVho art vvritten.}{Amongſt
\MNote{Not only this booke but alſo other holie ſcriptures witnes that
Elias shal returne and preach before the end of the
world. S.~Chryſoſtom, Aretas, and other Doctors teſtifie the ſame. See
\XRef{Annot. Gen.~5.}}
other quarels, Proteſtantes except againſt the authentical auctoritie of
this booke, becauſe the auctor ſaith, that Enoch and Elias shal come
againe, to appeaſe the wrath of our Lord, to reconcile the father to the
ſonne, and to reſtore the tribes of Iacob. But that this is no iuſt
exception, is clere by other holie Scriptures, where the ſame vniforme
doctrin of the whole Church, is no leſſe euident, then in this
booke. For God himſelf ſaith the ſame alſo by the mouth of his prophet
Malachie:
\CNote{\XRef{Mal.~4.}}
Behold I wil ſend you Elias the prophet, before the day of our Lord
come, the great and dreadful. Chriſt alſo ſayth:
\CNote{\XRef{Mat.~17.}}
Elias in deede shal come, and reſtore al thinges. VVherupon
S.~Chryſoſtom, after he hath ſhewed how terrible Antichriſt ſhal be, by
reaſon of his temporal powre, crueltie, and wicked lawes, he addeth:
\CNote{\Cite{In 2.~Theſ.~2.}}
Feare thou not. He shal only haue force in the reprobate, that
perish. For then alſo Elias ſhal come, to fortifie the
faithful. Likewiſe the wordes in the Apocalips,
\CNote{\XRef{Apoc.~11.}}
I wil geue to my two witneſſes, & they shal prophecie a thouſand two
hundred and three ſcore dayes, were euer inuariably vnderſtood by
tradition, from the firſt preachers of Chriſt (as the ancient writer
Aretas teſtifieth) that Enoch and Elias shal come, & admonish al, not to
geue credite to the deceiptful wonders of Antichriſt: and that they shal
trauel in this teſtimonie the ſpace of three yeares and a half. For
1260.~dayes come very nere to that ſpace of time.}
Who art written in the iudgements of times, to appeaſe the wrath of our
Lord, to reconcile the hart of the father to the ſonne, and to reſtore
the tribes of Iacob. \V Bleſſed are they, that ſaw thee, and were
honored in thy freindshipe. \V For we liue by life only, but after death
our name shal not be ſuch. \V Elias was in dede hid in the whirlewind, &
his ſpirit was complete in Eliſeus: in his daies he feared not the
prince, and no man ouercame him by might. \V Neither did any word
ouercome him, and his bodie
\SNote{The miracle wrought by his dead bodie ſhewed that he was an holie
prophet.
\XRef{4.~Reg.~13.}
See the miracles of Elias, and Eliſeus.
%%% ???
\Cite{To.~1. pag.~940.}}
prophecied being dead. \V
\CNote{\XRef{4.~Reg.~13.}}
In his life he did wonders, and in death he wrought meruelous
thinges. \V In al theſe thinges the people repented not, and they
departed not from their ſinnes, til they were caſt out of their land,
and were diſperſed into al
%%% o-1454
the earth. \V And there was leaft a verie ſmal nation, and a prince in
the houſe of Dauid. \V Some of them did that which pleaſed God: but
others committed manie ſinnes. \V Ezechias fenced his citie, and
\Fix{brough}{brought}{obvious typo, fixed in other}
in water into the middes thereof, and digged a rocke with yron, and
built a wel for water. \V
%%% ???
\CNote{\XRef{4.~Reg.~20.}
\XRef{4.~Reg.~18.}}
In his daies came vp Sennacherib, and ſent Rabſaces, and lifted vp his
hand againſt them, and put forth his hand vpon Sion, and became proude
by his mightines. \V Then were their harts, and hands moued: and they
were in ſorow as trauailing wemen. \V And they inuocated our merciful
Lord, and ſpredding their handes, they lifted them vp to heauen: and the
holie Lord God
\SNote{Prayer preuailed, when forces were not ſufficient. See
\XRef{4.~Reg.~19.}}
quickly heard their voice. \V He was not mindful of their ſinnes,
neither did he geue them to their
%%% 1579
enemies, but purged them by the hand of Iſaie the holie prophete. \V
\CNote{\XRef{4.~Reg.~19.}
\XRef{Iſa.~37.}}
He ouerthrew the campe of the Aſſirians, and the Angel of our Lord
deſtroyed them. \V For Ezechias did that which pleaſed God, and went
ſtrongly in the way of Dauid his father, which Iſaie commanded him, the
great prophet, and faithful in the ſight of God. \V
\CNote{\XRef{4.~Reg.~20.}
\XRef{Iſa.~38.}}
In his daies the ſunne returned backward, & added life to the king. \V
By a great ſpirite he ſaw the laſt thinges, and comforted the mourners
in Sion. \V For euer he shewed the thinges to come, & ſecret thinges
before they came to paſſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
