%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 32
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-32


\project douay-rheims


%%% 1554
%%% o-1433
\startChapter[
  title={Chapter 32}
  ]

\Summary{Superiors muſt rule with mekenes, 4.~teaching thoſe wiſdom that
  are capable thereof. 7.~Be moderate in muſike, and in wine. 9.~Let
  yongmen be diligent to heare, and ſparing to ſpeake, 13.~eſpecially
  before their betters. 15.~Be alwayes wel occupied. 17.~Serue, and
  feare God. 21.~Admitte correction. 24.~Do nothing without counſel.}


Haue they made thee Ruler?
\SNote{Humilitie is neceſſarie in al, but moſt eſpecially in men of
auctoritie. The greateſt art in this life is to contemne vaine glorie in
height of auctoritie.
\Cite{S.~Greg. de cura paſterali}}
be not extolled: be among them as one of them. \V Haue care of them, and
ſo ſitte thou ſtil, and al thy care being diſpatched, repoſe. \V That
thou maiſt reioyce for their ſakes, & receiue a crowne as an ornament of
grace, and obteyne the dignitie of the contribution. \V Speake thou that
art elder: for it becometh thee, \V the firſt word to him that loueth
with knowlege, & hinder not muſike. \V
\CNote{\XRef{Eccle.~3.}}
Where there is no hearing, power not out
%%% 1555
ſpeache, and extol not thyſelf out of time in thy wiſdom. \V A litle
pearle of the carbuncle in an ornament of gold, and the compariſon of
muſicians in a banket of wine. \V As a ſignet of the emerauld is in the
working of gold: ſo the melodie of muſike in ioyful and moderate
wine. \V Heare holding thy peace, & for thy reuerence good grace shal
come to thee. \V Yong man ſpeake in thine owne cauſe ſcarſely. \V If
thou be asked twiſe, let thyne anſwer haue an head. \V In manie thinges
be as it were ignorant, and heare holding thy peace and withal
asking. \V In the middes of greate men preſume not: and where ancients
are, ſpeake not much. \V Before haile there shal goe lightning: & grace
shal goe before shamfaſtnes, & for thy reuerence good grace shal come to
thee. \V And at the houre of ryſing ſlacke not thyſelf: but runne before
firſt into thy houſe, and there withdraw thyſelf, and there play, \V and
doe thy conceites, and not in ſinnes and proud word. \V And aboue al
theſe thinges bleſſe our Lord, that made thee, & that doth replenish
thee with al his goodes. \V He that feareth our Lord, shal receiue his
doctrine: and they that wil watch after him, shal finde bleſſing. \V He
that ſeeketh the law, shal be replenished with it: and he that doth
deceitfully, shal be ſcandalized by it. \V They that feare our Lord,
shal finde iuſt iudgement, and shal kindle iuſtices as light. \V A
ſinful man wil flee reprehenſion, and according to his wil, wil finde
excuſe. \V A man of counſel wil not deſtroy vnderſtanding, an aliene and
proud man wil not dread feare. \V Yea after he hath done with feare
without counſel, he shal be controwled euen by his owne purſuites. \V
Sonne
\SNote{God directeth mens reſolutions, and actions, when they vſe the
meanes of conſultation, as he hath appointed: for vvhen two or three
are gathered in his name, he is in the middes of them.
\XRef{Mat.~18.}}
doe nothing without counſel, and after the fact thou shalt not
repent. \V Goe not in the way of ruine, and thou shalt not ſtumble at
ſtones: commite not
%%% o-1434
thyſelf to a laborious way, leſt thou ſet a ſcandal to thy ſoule. \V And
beware of thy children, and take heede of them of thy houſehold. \V In
al thy worke beleue thy ſoule
\SNote{Do not againſt thy cõſcience.}
by faith: for this is the keeping of the commandmentes. \V He that
beleueth God, attendeth to the commandmentes: and he that truſteth in
him, shal not be leſſened.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
