%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 1529
%%% o-1410
\startChapter[
  title={Chapter 15}
  ]

\Summary{The fruictes of fearing God; 7.~which fooles, and liers reape
  not, but the wiſe only. 12.~God is no way auctor of ſinne, 14.~but
  ſinners themſelues are the auctors, abuſing their freewil.}


He that feareth God, shal doe good thinges, and he that hath
\SNote{VVhoſoeuer shal reſolue with himſelf to liue iuſtly, shal be
ſure to haue grace, for God preuenteth our weakenes and ſo continueth to
helpe al that accept his grace.}
iuſtice shal apprehend her, \V and she wil meete him as an honourable
mother, and as a wife from virginitie she wil receiue him. \V She shal
fede him with the bread of life and vnderſtanding, and geue him the
water of wholſome wiſdom to drinke: and she shal be made ſure in him,
and he shal not be bowed: \V and she shal hold him faſt, and he shal not
be confounded: and she shal exalt him before his neighbours, \V and in
the middeſt of the Church she shal open his mouth, and shal fil him with
the ſpirite of wiſdom and vnderſtanding, and shal clothe him with a
ſtole of glorie. \V Ioy & exultation she shal heape vpon him, & shal
make him inherite an euerlaſting name. \V Foolish men shal not apprehend
her, & wiſe men shal meete her, foolish men shal not ſee her: for she is far
from pride and deceite. \V Lying men shal not be myndful of her: and
true men shal be found in her, and shal haue ſucceſſe, euen to the
beholding of God. \V
\SNote{He doth iniurie to God, & to his word, that preacheth wel and
liueth euil.}
Prayſe is not comelie in the mouth of a ſinner: \V becauſe wiſdom
proceded from God. For prayſe shal be with the wiſdom of God, & shal
abound in a faithful mouth, & the dominator wil giue praiſe to
\Fix{yt.}{it.}{obvious typo, fixed in other}
\V Say not:
\SNote{Beza ſayeth, God ordained Adams fall, but to a good end: and that
God iuſtly decreed that which men vniuſtly haue done.
\Cite{in refut. 2.~calumna, ad Caſtel.}
But this holie Scripture teacheth the contrarie, that God gaue man both
freewil, and ſufficient grace, that he might if he would haue kept his
\Fix{preceps.}{precepts.}{obvious typo, fixed in other}
The ſame is alſo clerely taught.
\XRef{Deut.~7. 11.~30.}
and other places.}
It is by God, that she is abſent: for doe not thou the thinges that he
hateth. \V Say not: He hath made me erre: for impious men are not
neceſſarie for
%%% o-1411
him. \V Our Lord hateth al abomination of errour, and it shal not be
amiable to them, that feare him. \V
\CNote{\XRef{Gen.~2.}}
God from the beginning made man, and left him in the hand of his owne
counſel. \V He added his commandmentes and precepts. \V
\CNote{\XRef{Mat.~15. v.~17.}}
If thou wilt keepe the commandment, and keepe acceptable fidelitie for
euer, they shal preſerue thee. \V He hath ſet before thee water and
fire: to which thou wilt, ſtretch forth thine hande. \V
\CNote{\XRef{Ier.~2. v.~8.}}
Before man there is life and death, good and euil: what pleaſeth him
shal be geuen him. \V Becauſe the wiſdom of God is much,
%%% 1530
and he is ſtrong in mighte, ſeing al men without intermiſſion. \V The
eies of our Lord are towards them that feare him, and he knoweth al the
worke of man. \V He hath commanded no man to do impiouſly, and he hath
geuen no man ſpace to ſinne: \V for he deſireth not a multitude of
faithleſſe and vnprofitable children.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
