%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 49
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-49


\project douay-rheims


%%% 1579
%%% o-1455
\startChapter[
  title={Chapter 49}
  ]

\Summary{Praiſes of Ioſias, who (like to Dauid, and Ezechias) tooke away
  occaſions of idolatrie. 8.~Praiſes of Ieremie, 10.~Ezechiel, 12.~and
  the twelue Prophetes. 13.~Alſo of Zorobabel, Ieſus the ſonne of
  Ioſedech, Nehemias, Enoch, Ioſeph, Seth, Sem, and Adam.}


The
\MNote{\XRef{4.~Reg.~22.}}
memorie of Ioſias is according to the confection of perfume made by the
worke of an apothecarie. \V His remembrance shal be ſweete as honie in
euerie mouth, and as muſick in banket of wine. \V
\CNote{\XRef{2.~Par.~34.}}
He was directed by God into the repentance of the nation, and he tooke
away the abominations of impietie. \V
\CNote{\XRef{4.~Reg.~23.}}
And he gouerned his hart toward our Lord, and in the daies of ſinners he
ſtrengthened pietie.
%%% 1580
\V Except Dauid, and Ezechias, and Ioſias,
\SNote{Manie other kinges of Iuda refrained alwayes from committing
idolatrie, but theſe three deſtroyed al places of idolatrie in their
kingdom, which the others did not.}
al committed ſinne. \V For the kinges of Iuda forſooke the law of the
Higheſt, and contemned the feare of God. \V For they gaue their kingdom
to others, and their glorie to a ſtrange nation. \V
\CNote{\XRef{4.~Reg.~25.}
\XRef{Iere.~1.}}
They burnt the choſen citie of holines, and made the waies thereof
deſolate in the hand of Ieremie. \V For they euil intreated him, who was
conſecrated a prophet from his mothers wombe, to ouerthrow, and pluck
vp, and deſtroy, and to build againe, and renewe. \V
\CNote{\XRef{Ezec.~1.}}
Ezechiel who ſaw the ſight of glorie, which he shewed him in the
chariote of Cherubs. \V For he made mention of the enemies in rayne, to
doe good vnto them, that haue shewed right waies. \V And the bones of
the twelue prophets wel may they ſpring out of their place: for they
haue ſtrengthened Iacob, and haue redeemed themſelues in the fidelitie
of power. \V
\CNote{\XRef{Agge.~2.}
\XRef{1.~Eſd.~3.}}
How may we magnifie Zorobabel? for he alſo was as a ſignet on the right
hand, \V and ſo Ieſus the ſonne of Ioſedec: who in their daies built the
houſe, and erected the holie temple to our Lord, prepared to euerlaſting
glorie. \V
\CNote{\XRef{2.~Eſd.~2.}
\XRef{2.~Eſd.~5.}
\XRef{Zach.~3.}}
And Nehemias in the memorie of much time, who erected vs our walles
ouerthrowen, and ſet vp the gates and lockes, who built our houſes. \V
\CNote{\XRef{Gen.~5.}}
No man hath bene borne in the earth like to Henoch: for
\SNote{See the
\XRef{Annotation ch.~38. v.~10.}}
he alſo was taken vp from the earth. \V
\CNote{\XRef{Gen.~39.~40. &c.}}
Neither as Ioſeph who was a man borne prince of his bretheren, the ſtay
of the nation, the ruler of his bretheren, the ſtay of the people: \V
and his bones were viſited, and after death
\SNote{Ioſeph prophecying that the people should depart from Ægypt, willed
them to carie his bones with them.
\XRef{Gen.~50.}
So by carying his bones they profeſſed, that he had truly prophecied.}
they prophecied. \V Seth, and Sem obteyned glorie with men: and aboue
euerie ſoule, in the beginning Adam.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
