%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 1530
%%% o-1411
\startChapter[
  title={Chapter 16}
  ]

\Summary{A few good children, yea none at al, are better then manie
  wicked. 9.~Gods wrath falleth vpon the euil, and his mercie on the
  good. 14.~Workes of mercie merite great reward. 15.~Nothing is hid from
  God, 20.~but manie thinges from men.}


Reioice not in impious children, if they be multiplied: neither be
delighted vpon them, if the feare of God be not in them. \V Credite not
their life, and reſpect not their labours. \V For better is one fearing
God, then a thouſand impious children. \V And it is more profitable to
die without children, rather then to leaue impious children. \V By one
wiſe a countrie shal be inhabited, and the tribe of the impious shal be
made deſolate. \V Manie ſuch thinges hath mine eie ſene, and myne eare
hath heard thinges of more force then theſe. \V In the ſynagogue of
ſinners
\SNote{Fire of concupiſcence, if it be not ouercome in this life, wil
procure the fire of Gods wrath, which ſhal neuer be extinguished.}
a fire shal flame, and in an incredulous nation wrath shal waxe
hotte. \V
\CNote{\XRef{Gen.~6.}}
The old giantes did not obteine for their ſinnes, who were deſtroied
truſting to their owne ſtrength: \V and he ſpared not the peregrination
of Lot, and he deteſted them for the pride of their word. \V He had not
pitie on them, deſtroying the whole nation, and extolling themſelues in
their ſinnes. \V
\CNote{\XRef{Num.~26.}}
And as the ſix hundred thouſand foote men, which were gathered together
in the hardnes of their hart: and if one had bene ſtiffe necked, it is
meruel if he had bene vnpunished. \V For there is mercie and wrath with
him. Mightie exoration, & powring out wrath: \V according to his mercie,
ſo his chaſtiſement iudgeth a man according to his workes. \V The ſinner
shal not eſcape in robberie, and the ſufferance of him that doth mercie
shal not ſlacke.
%%% o-1412
\V
\CNote{\XRef{Rom.~2. v.~6.}}
Al mercie shal make a place to euerie man according to
\SNote{Euen a diſh of colde water geuen in almes shal be rewarded.
\XRef{Mat.~10.}}
the merite of his workes, & according to the vnderſtanding of his
peregrination. \V Say not: I shal be hid from God, and from on high who
shal remember me? \V In a great people I shal not be knowen: for what is
my ſoule among ſo innumerable creatures? \V Behold heauen,
%%% 1531
and the heauens of heauens; the depth, and al the earth, & the thinges
that are in them, in his ſight shal be moued, \V the mountaines
together, and the litle hilles, & the fundations of the earth: & when
God shal behold them, they shal be shaken with trembling. \V And in al
theſe thinges the hart is ſenſeles: and euerie hart is vnderſtood of
him: \V and his wayes who doth vnderſtand, and the ſtorme, which neither
the eie of man shal ſee? \V For manie of his workes are in ſecretes: but
the workes of his iuſtice who shal declare? or who shal ſuſteine? For
the teſtament is far from ſome, and the examination of al is in
conſummation. \V He that is leſſe of hart, thinketh vaine thinges: and
the vnwiſe, and erring man, thinketh
\Fix{folish}{foolish}{obvious typo, fixed in other}
thinges. \V Heare me my ſonne, and lerne the diſcipline of
vnderſtanding, and attend to my wordes in thy hart, \V and I wil ſpeake
diſcipline in equitie, and wil ſearch to declare wiſedom, and to my
wordes attend in thy hart, and I ſpeake in equitie of ſpirit the
vertues, that God hath put vpon his workes from the beginning, and in
truth I shew forth his knowlege. \V In the iudgement of God are his
workes from the beginning, and from the inſtitution of them he
diſtinguished their partes, and their beginninges in their nations. \V
He adorned their workes for euer, neither haue they hungred, nor
laboured, and they haue not ceaſed from their workes. \V Euerie one shal
not vexe his neighbour for euer. \V Be not incredulous to his word. \V
After this God looked vpon the earth, & filled it with his good
thinges. \V And
\SNote{Reaſonable ſoules (if they folow reaſon) and al ſenſible ſoules
doe (in their maner) praiſe the prouidence of God, in vſing al creatures
to that end for which they were created.}
the ſoule of euerie liuing thing shal shew before the face thereof, and
into it againe is their returne.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
