%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 33
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-33


\project douay-rheims


%%% 1555
%%% o-1434
\startChapter[
  title={Chapter 33}
  ]

\Summary{Feare of God defendeth from al aduerſaries. 5.~Follie is
  vnconſtant. 8.~God diſpoſeth al to the beſt. 13.~Man is in Gods hand,
  as clay in the potters. 20.~Superiors muſt keepe their auctoritie: and
  their ſubiectes in diſcipline.}


%%% 1556
To him that feareth our Lord euils shal not happen, but in tentation God
wil keepe him, and deliuer him from euils. \V A wiſe man hateth not the
commandments and iuſtices, and he shal not be shaken as a shippe in a
ſtorme. \V A man of vnderſtanding beleueth the law of God, and the law
is ſure to him. \V He that repeteth an interrogation, shal better
prepare his anſwer, and ſo shal be heard, and shal keepe diſcipline. \V
\SNote{Men ledde with paſſions runne from one vice into an other without
ceaſing. Eſpecially heretikes runne into manie errors. Againſt which
S.~Paul preſcribeth this rule. \Emph{It is beſt that the hart be
eſtablished with grace.}
\XRef{Heb.~13. v.~9.}}
The hart of a foole is as a wheele of a carte: & his cogitation as a
turning axeltree. \V A ſtalion horſe neyeth vnder euerie one that
ſitteth vpon him, ſo a freind that is a ſcorner. \V Why doth one day
excel an other, and one light an other, and one yeare an other yeare of
the ſunne? \V By the knowlege of our Lord they were ſeperated, the ſunne
being made, and keeping the precept. \V And he changed times, and the
feſtiual daies thereof, and in the ſame they celebrated the feſtiual
daies at an houre. \V Of them God exalted and magnified, and of them he
put into the number of daies. And al men are of the ground, and of the
earth, from whence Adam was created. \V In the multitude of the
diſcipline of our Lord he ſeparated them, and changed their waies. \V Of
them he bleſſed, and exalted: and of them he ſanctified, and applied
to himſelf: and of them he curſed and humbled, and conuerted them from
their ſeparation. \V
%%% !!! Rom.~3 in other.
\CNote{\XRef{Rom.~9.}}
As potters clay is in his hand, to fashion and diſpoſe it. \V Al his
wayes according to his diſpoſition: ſo man in the hand of him, and he
wil render to him according to his iudgement. \V Againſt euil is good,
and againſt death life: ſo alſo againſt a iuſt man a ſinner. \V And ſo
looke vpon al the workes of the Higheſt. Two againſt two, and one
againſt one. \V And I awaked laſt, and as he that gathereth bearies
after the grape gatherers. \V In the bleſſing of God I alſo haue hoped:
and as he that gathereth grapes, haue I filled the wine preſſe. \V See
%%% o-1435
that I haue not laboured for myſelf only, but for al that ſeeke out
diſcipline. \V Heare me ye great men, and al peoples, and ye rulers of
the Church
\Fix{karken}{harken}{obvious typo, fixed in other}
with your eares. \V To ſonne and wife, brother and freind, geue not
power ouer thee in thy life: and geue not thy poſſeſſion to an other:
leſt perhaps thou repent thee, and thou entreate for them. \V Whiles
thou are yet aliue and takeſt breath, al flesh shal not change thee. \V
For it is better that thy children aske of thee, then that thou looke
toward the hands of thy children. \V In al thy workes
%%% 1557
be exquiſite. \V Geue no ſtaine to thy glorie. In the day of the
conſummation of the dayes of thy life, and in the time of thy deceaſe
diſtribute thine inheritance. \V Fodder, and wande, and burden for an
aſſe: bread, and diſcipline, and worke for a ſeruant. \V He that worketh
in diſcipline, and ſeeketh to reſt: releaſe him his
handes, and he ſeeketh libertie. \V The yoke and the reyne bend a ſtiff
necke, and continual workes do bowe a ſeruant. \V For a malicious
ſeruant torment & fetters, ſend him into worke, that he be not idle. \V
For idlenes hath taught much naughtines. \V Set him to worke: for ſo it
becometh him. And if he be not obedient, bow him with fetters, and
exceede not ouer al flesh: but without iudgement doe no greuous
thing. \V If thou haue a faithful ſeruant, let him be vnto thee as thy
ſoule: as a brother ſo entreate him: becauſe in the bloud of thy ſoule
thou haſt gotten him. \V If thou hurt him vniuſtly, he wil runne
away: \V if riſing vp he depart: thou knoweſt not whom to aske, and what
way to ſeeke him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
