%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 51
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-51


\project douay-rheims


%%% 1582
%%% o-1457
\startChapter[
  title={Chapter 51}
  ]

\Summary{The auctor rendereth praiſes and thankes to God, 18.~and
  inuiteth others to do the ſame, by his owne example, 31.~and by
  earneſt exhortation.}


The prayer of Ieſus the ſonne of Sirach.
\SNote{VVhere we are not able to render recompence to benefactors;
eſpecially to God, we are the more bond to acknowlege his manie great
benefites, altogether vndeſerued by vs.}
I wil confeſſe to thee ô Lord king, and wil praiſe thee God my
ſauiour. \V I wil confeſſe to thy name: becauſe thou art become my
helper and protectour, \V and haſt deliuered my bodie from perdition,
from the ſnare of an vniuſt tongue, and from the lippes of them that
worke lying, and in the ſight of them that ſtoode vp thou art become my
helper. \V And thou haſt deliuered me according to the multitude of the
mercie of thy name from them that did roare, prepared to deuoure, \V out
of the handes of them that ſeeke my ſoule, and from the gates of
tribulations which haue compaſſed me: \V from the oppreſſion of the
flame, which hath compaſſed me, and in the middes of fire I was not
burnt. \V From the depth of the bellie of hel, and from a defiled
tongue, and from the word of lying, from a wicked king, and from an
vniuſt tongue: \V my ſoule shal praiſe our Lord euen to death, \V and my
life was approching to hel beneth. \V They haue compaſſed me on euerie
ſide, & there was none that would helpe. I looked toward the helpe of
men, & there was none. \V I remembred thy mercie ô Lord, and thy
operation, which are from the beginning of the world. \V Becauſe thou
deliuereſt them that patiently expect thee ô Lord, and ſaueſt them out
of the handes of the nations. \V Thou haſt exalted my habitation vpon
the earth, and I haue prayed for death to paſſe away. \V I haue
inuocated our Lord the father of my Lord, that he
%%% o-1458
leaue me not in the
%%% 1583
day of my tribulation, and in the time of the proude without helpe. \V I
wil praiſe thy name continually, and wil collaude it in confeſſion, and
my prayer was heard. \V And thou haſt deliuered me from perdition, and
haſt reſcued me from the wicked time. \V Therfore wil I confeſſe, & ſay
praiſe to thee, and bleſſe the name of our Lord. \V When I was yet
\SNote{VVhen ſenſes are moſt ripe, and the ſoule moſt free from great
ſinnes, is the apteſt time to ſerue God, & to get al vertues, and true
knowlege.
\XRef{Eccle.~12.}}
young, before I erred, I ſought for wiſdom openly in my prayer. \V
Before the temple I prayed for it, and vnto the later end I wil ſeeke
after it, and it shal flourish as the grape timely ripe, \V my hart hath
reioyced in it, my foote hath walked the right way, from my youth I
ſearched after it. \V I bowed mine eare a litle, and receiued it. \V I
found much wiſdom in myſelf, & I haue much profited therein. \V To him
that geueth me wiſdom, wil I geue glorie. \V For I haue conſulted to doe
it: I haue had a zele to good, and shal not be confounded. \V My ſoule
hath wreſtled in it, and in doing it I was confirmed. \V I ſtretched
forth my handes on high, & I lamented foolishnes. \V I directed my ſoule
to wiſdom, and in knowlege I found it. \V I poſſeſſed with it an hart
from the beginning: for this cauſe I shal not be forſaken. \V My bellie
was trubled in ſeeking it: therefore shal I poſſeſſe a good
poſſeſſion. \V Our Lord hath geuen me a tongue for my reward: and with
the ſame I wil praiſe him. \V Approch vnto me ye vnlerned, and gather
yourſelues together into the houſe of diſcipline. \V Why ſlacke ye yet?
and what ſay you herein? your ſoules are exceeding thirſtie. \V I haue
opened my mouth, and haue ſpoken: Bye it for you without ſiluer, \V and
ſubmit your necke to the yoke, and let your ſoule receiue diſcipline:
for it is very neere to finde it. \V See with your eyes that I haue
laboured a litle, and haue found much reſt to myſelf. \V Take ye
diſcipline
\SNote{Inſtead of riches labour to get wiſdom for it is much better then
al gold & ſiluer.}
in a great ſumme of ſiluer, and poſſeſſe abundance of gold in it. \V Let
your ſoule reioyce in his mercie, and you shal not be confounded in
praiſe. \V Worke your worke
\SNote{Merite is in this life, and reward in the next.}
before the time, and he wil geue you your reward in his time.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
