%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 30
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-30


\project douay-rheims


%%% 1551
%%% o-1430
\startChapter[
  title={Chapter 30}
  ]

\Summary{Chaſtiſment of children is neceſſarie, and indulgence very
  dangerous. 14.~Health is better then riches. 17.~A trubleſome life is
  worſe then death. 21.~Be not penſiue but chereful in mind.}


He
\CNote{\XRef{Pro.~13.}
&
\XRef{23.}}
that loueth his ſonne, doth accuſtome him to ſtripes, that he may reioyce
in his later end, and not grope after the doores of his neighboures. \V
He that teacheth his ſonne, shal be praiſed in him, & in the middes of
them of his houshold he shal glorie in him. \V He that teacheth his
ſonne, doth caſt
%%% 1552
the enemie into emulation, and in the middes of his freindes he shal
glorie in him. \V
\CNote{\XRef{Deut.~6.}}
His father is dead, & he is as it were not dead: for he hath left behind
him the like to himſelf. \V In his life he ſawe and reioyced in him: in
his death he was not made ſorie, neither was he confounded before the
enemies. \V For he left a defender of his houſe againſt the enemies, &
one that should render thanck to his freindes. \V
\SNote{The eldeſt ſonne being heyre to his father, is to haue ſpecial
care of the other children, euen to the aduenturing of his owne hurt for
their liues.}
For the ſoules of his ſonnes he wil binde vp his woundes, & at euerie
voice his bowels shal be trubled. \V An vntamed horſe becometh
ſtubburne, and a diſſolute childe wil become headie. \V Pamper thy
ſonne, and he wil make thee afraid: play with him, and he wil make thee
ſorowful. \V Laugh not with him, leſt thou be ſorie, and at the laſt
\SNote{Teeth on edge or gnaſhing of teeth, is a part of hel paines.
\XRef{Mat.~8.}
&
\XRef{13.}
& ſometimes beginneth in this life.}
thy teeth shal be on edge. \V Geue him not power in his youth, and
contemne not his cogitations. \V Curbe his necke in youth, and knock his
ſides whiles he is a childe, leſt perhaps he be hardned, and beleeue
thee not, and he shal be ſorow of minde to thee. \V Teach thy ſonne, and
worke in him, that thou offend not in his dishoneſtie. \V Better is a
poore man whole, and ſtrong of force, then a rich man weake and
%%% o-1431
ſcourged with miſerie. \V The
\SNote{Pietie is aboue al riches and honour.}
health of the ſoule in holines of iuſtice, is better then al gold and
ſiluer: and a ſound bodie, then infinite reuenewes. \V There is no
riches aboue the riches of the health of the bodie: and there is noe
delight aboue the ioy of the hart. \V Better is death then a bitter
life: and euerlaſting reſt, then continual ſicknes. \V Good thinges hid
in a mouth that is shut, are as meſſes of meates ſet about a graue. \V
\CNote{\XRef{Deut.~14.}}
What shal ſacrifice profite an idol? for neither shal he eate, nor
ſmel: \V ſo he that is chaſed away of our Lord, beareth the rewardes of
iniquitie: \V ſeing with his eies, & groning, as an eunuch embracing a
virgin and ſighing. \V
\CNote{\XRef{Prou.~12.~15.~&~17.}}
\SNote{Suffer not vnneceſſarie penſiuenes to afflict thy mind, through
puſillanimitie.}
Geue not heuines to thy ſoule, & afflict not thyſelf in thy counſel. \V
Ioyfulnes of the hart, this is the life of a man, and a treaſure without
defect of holines: and the ioy of a man is long life. \V Haue mercie on
thine owne ſoule,
\SNote{But relie wholly vpon Gods wil and prouidence, reſigning thy wil
vnto his.}
pleaſing God, and refraine: and comfort thy hart in his holines: and
expel ſorow far from thee. \V For ſorow hath killed manie, and there is
noe profite in it. \V Enuie and anger diminish the daies, and thought
wil bring old age before the time. \V A magnifical hart, is good in
bankettes: for his bankettes are made diligently.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
