%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 34
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-34


\project douay-rheims


%%% 1557
%%% o-1435
\startChapter[
  title={Chapter 34}
  ]

\Summary{Truſt not vaine dreames, ſouthſayinges, nor lies. 9.~Much good
  is got by experience: 14.~and more by fearing God. 21.~God reiecteth
  the oblations of the wicked. 24.~Defrauding the poore is like to
  manſlaughter. 28.~Deſtroy not that an other buildeth. 30.~Repentance
  without amendment is nothing worth.}


Vaine hope, & lying is to a foolish man: &
\SNote{Vaine dreames are not to be regarded: but ſome are good, & from
God.
\XRef{Gen.~37.~40.~41.}
\XRef{Dan.~2.~4.}
\XRef{Mat.~1.}}
dreames extol the vnwiſe. \V As he that apprehendeth a shadow, and
purſueth the winde: ſo is he alſo that attendeth to lying viſions. \V
According to this is the viſion of dreames: as a mans ſimilitude before
the face of a man. \V Of the vncleane what shal be made cleane? and of a
lyer what truth shal be ſaid? \V Diuination of errour, and lying
ſouthſayinges, and the dreames of them that do euil, are vanitie. \V
And as a woman that traueleth thy hart ſuffereth phantaſies: vnleſſe it
be a viſion ſent forth from the Higheſt, ſet not thy hart vpon them. \V
For dreames haue made manie to erre, and they that hoped in them haue
failed. \V
\SNote{VVhatſoeuer is written in holy Scripture is vndoubtedly true: and
no iote of the law ſhal periſh.}
The word of the law shal be fulfilled
%%% o-1436
without lying, and
\SNote{God alſo prouideth that alwayes there be ſome which truly
explicate his lavv.}
wiſedom in the mouth of the faithful shal be made plaine. \V He that
hath not bene proued, what knoweth he? A man expert in manie thinges,
shal thinck manie thinges: and he that hath lerned manie thinges, shal
declare vnderſtanding. \V He that is not tryed, knoweth
%%% 1558
few thinges: and he that hath bene in manie things, multiplyeth
wickednes. \V He that hath not bene proued, what maner of thinges
knoweth he? He that is deceiued, shal abound with wickednes. \V I haue
ſeene manie thinges in wandring to and fro, and very manie fashions of
words. \V Some times I haue bene in danger vnto death for theſe thinges,
and I was deliuered by the grace of God. \V The ſpirit of them that
feare God is ſought, & at his ſight shal be bleſſed. \V For their hope
is on him that ſaueth them, and the eies of God vpon them that loue
him. \V He that feareth our Lord shal tremble at nothing, and shal not
dread: becauſe he is his hope. \V His ſoule is bleſſed that feareth our
Lord. \V To whom doth he looke, and who is his ſtrength? \V
\CNote{\XRef{Pſa.~32.}}
The eies of our Lord are vpon them that feare him, a protectour of
might, a ſtay of ſtrength, a couer from the heate, and shadow for the
noone time, \V a ſauing from offence, and helpe from falling, exalting
the ſoule, and illuminating the eies, geuing health, and life, and
bleſſing. \V
\CNote{\XRef{Prou.~21. v.~27.}}
The oblation of him that immolateth of an vniuſt thing is ſpotted, and
the ſcorninges of the vniuſt are not acceptable. \V Our Lord is onlie
theirs that expect him, in the way of truth and iuſtice. \V
\CNote{\XRef{Prou.~15. v.~8.}}
The Higheſt alloweth not the giftes of the wicked: neither hath he
regard to the oblations of the vniuſt, neither wil he be made propitious
for ſinnes by the multitude of their ſacrifices. \V He that offereth
ſacrifice of poore mens ſubſtance, is as he that ſacrificeth the ſonne
in the preſence of his father. The bread of the needie, is the life
of the poore: he that defraudeth it, is a man of bloud. \V He that
taketh away bread in ſwet, is as he that killeth his neighbour. \V
\CNote{\XRef{Leuit.~19.}}
He that sheddeth bloud, and that defraudeth the hired man, are
bretheren. \V
\SNote{He that deſtroyeth that an other buildeth bringeth tvvo mens
labours to nothing.}
One building, and an other
\Fix{deſtroing:}{deſtroying:}{probable typo, fixed in other}
what profite haue they but the
labour? \V One praying, and an other curſing: whethers voice wil God
heare? \V He that is washed from the dead,
\SNote{Recidiuation into ſinne maketh the former repentãce fruſtrate.
\XRef{Mat.~18. v.~33.}}
and toucheth him againe, what doth his washing profit? \V So a man that
faſteth in his ſinnes: and doing the ſame againe, what doth he profite
in humbling himſelf? who wil heare his prayer?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
