%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 1512
%%% o-1395
\startChapter[
  title={Chapter 03}
  ]

\Summary{Honour of parents procureth Gods bleſſing, 11.~dishonoring them
  his curſe. 19.~Mekenes and modeſtie auaile much: but curioſitie to
  know ſecret myſteries is dangerous. 27.~A charitable, ſincere, and
  docile hart, 33.~with workes of mercie merite reward.}


The children of wiſdom, are the Church of the iuſt: and their
\SNote{The progenie of Gods children bring forth the fruite of obedience
and loue.}
nation, obedience and loue. \V
\SNote{Honour and eſtimation of parents is the chiefeſt obligation that
man hath towardes his neighbour, & the firſt after his dutie to God.}
Heare your fathers iudgement ô children, and ſo doe that you may be
ſaued. \V For God hath honoured the father in the children: and
inquiring of the mothers iudgement, hath confirmed it vpon the
children. \V He that loueth God, praying shal obteyne for ſinnes, and
shal refrayne himſelf from them, and shal be heard in the prayer of
dayes. \V And as he that gathereth treaſure, ſo he alſo that honoureth
his mother. \V He that honoureth his father, shal haue ioy in children,
and in the day of his prayer he shal be heard. \V He that honoureth his
father, shal liue the longer life: & he that obeyeth the father, shal
refresh the mother. \V
\CNote{\XRef{Exo.~20.}}
He that feareth our Lord honoureth his parents, and as his lordes he wil
ſerue them, that begat him. \V 
\CNote{\XRef{Deut.~5.}}
In worke and word, & al patience honour thy father, \V
\CNote{\XRef{Gen.~27.}}
that bleſſing may come vpon thee from him, & his bleſſing may remayne in
the later end. \V
\CNote{\XRef{&~49.}}
The fathers bleſſing eſtablisheth the houſes of the
children: but the mothers curſe rooteth vp the foundation. \V Glorie not
in the contumelie of thy father: for his confuſion is no glorie to
thee. \V For the glorie of a man is by the honour of his father, and the
father without honour is the dishonour of the ſonne. \V Sonne receiue
the old age of thy father, and make him not ſorowful in his life: \V and
if he fayle in vnderſtanding, pardon him, and deſpiſe him not in thy
ſtrength: for the almes to the father shal not be in obliuion. \V For
good shal be reſtored
%%% 1513
thee
\SNote{Almes geuen, or prayer made, or ſacrifice offered for parents,
doth merite reward at Gods hand.}
for the ſinne of thy mother, \V and in iuſtice it shal be builded to
thee, and in the day of tribulation there shal be remembrance of thee:
and as yce in the clere weather shal thy ſinnes melt away. \V Of what an
euil fame is he, that forſaketh his
%%% o-1396
father: and he is curſed of God, that doth exaſperate his mother. \V
Sonne doe thy workes in meekenes, and thou shalt be beloued aboue the
glorie of men. \V
\CNote{\XRef{Philip.~2.}}
The
\SNote{Al greatnes in men proceding from God, bindeth them ſo much the
more to humilitie. VVhich vertue God moſt ſpecially loueth, that ſo they
may ſhew gratitude for his benefites.}
greater thou art, humble thy ſelf in al thinges, and thou shalt finde
grace before God: \V becauſe the might of God onlie is great, and he is
honoured of the humble. \V 
\CNote{\XRef{Pro.~25. v.~2.}}
Seeke not thinges higher then thy ſelf, and ſearch not thinges ſtronger
then thy habilitie: but the thinges that God hath commanded thee,
thincke on them alwayes, and in manie of his workes be not curious. \V
For it is not neceſſarie for thee, to ſee with thyne eies thoſe thinges
that are hid. \V In ſuperfluous thinges ſearch not manie wayes, and in
manie of his workes thou shalt not be curious. \V For verie manie
thinges are shewed to thee aboue the vnderſtanding of men. \V Manie alſo
haue their ſuſpicion ſupplanted, and haue held their ſenſes in
vanitie. \V A hard hart shal fare il in the later end: and he
\SNote{Thoſe that dare liue in ſinful ſtate, tempting God euen to the
laſt houre, commonly periſh therin.}
that loueth danger, shal perish in it. \V A hart that goeth two wayes,
shal not haue ſucceſſe, and the peruerſe of hart shal be ſcandalized in
them. \V A wicked hart shal be laden with ſorowes, and the ſinner wil
adde to commit ſinne. \V To the ſynagogue of the proude there shal be no
health: for the shrubbe of ſinne shal be rooted vp in them, and it shal
not be perceiued. \V The hart of the wiſe is vnderſtood in wiſdom, and a
good eare wil heare wiſdom with al deſire. \V A wiſe hart, and that
which hath vnderſtanding wil abſteyne it ſelf from ſinnes, and in the
workes of iuſtice shal haue ſucceſſe. \V
\CNote{\XRef{Dan.~4. v.~2.}}
Water quencheth burning fyre, and almes reſiſteth ſinnes: \V and God is
the beholder of him that rendreth grace: he remembreth him afterward,
and in the time of his fal he shal finde a ſure ſtay.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
