%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 35
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-35


\project douay-rheims


%%% 1558
%%% o-1437
\startChapter[
  title={Chapter 35}
  ]

\Summary{Obſeruation of the commandments, 4.~and ſacrifice of the iuſt
  pleaſe God, 12.~not the ſacrifice of the wicked. 14.~God protecteth
  the poore and deſolate, 19.~heareth the prayer of the humble, and
  rendereth to al as they deſerue.}


%%% 1559
He
\SNote{Sacrifices of penance,
\XRef{Pſal.~50.}
of iuſtice,
\XRef{Pſal.~4.}
and of praiſe,
\XRef{Pſa.~49.}
are better then external ſacrifices.}
that keepeth the law, multiplieth oblation. \V
\CNote{\XRef{1.~Rom.~15. v.~22.}
\XRef{Ier.~7.}}
It is an holſome ſacrifice to attend to the commandments, and to depart
from al iniquitie. \V To depart from iniquitie is a thing that pleaſeth
our Lord wel: and to depart from iniuſtice is an intreating for
ſinnes. \V
\CNote{\XRef{Exo.~23.~&34.}
\XRef{Deu.~16.}}
\SNote{Neuertheles external ſacrifice is alſo neceſſarie.}
Thou shalt not appeare before the ſight of our Lord emptie. \V For al
theſe thinges are done becauſe of the commandment of God. \V The
oblation of the iuſt maketh a fatte altar, and is an odour of ſweetenes
in the ſight of the Higheſt. \V The ſacrifice of the iuſt is acceptable,
and our Lord wil not forget the memorie thereof. \V Render glorie to God
with a good minde: and diminish not the firſt fruites of thine
handes. \V 
\CNote{\XRef{2.~Cor.~9.}
\XRef{Tob.~4.}}
In euerie gift make thy countenance chereful, and in ioyfulnes ſanctifie
thy tithes. \V  Geue to the Higheſt according to his gift, and with a
good eie, doe according to the abilitie of thine handes: \V becauſe our
Lord is a rewarder, and wil repay thee
\SNote{Often times ſo much. As
\XRef{Prou.~24.}
\Emph{Seuen times shal the iuſt fal, & riſe againe.}}
ſeuen times ſo much. \V
\CNote{\XRef{Leuit.~22.}
\XRef{Deut.~15.}}
Offer not wicked giftes, for he wil not receiue them. \V And looke not
vpon an vniuſt ſacrifice, becauſe our Lord is iudge, and there is not
with him the glorie of perſon. \V 
\CNote{\XRef{2.~Para.~19.}
\XRef{Rom.~2.}
\XRef{Act.~10.}}
Our Lord wil not accept perſon againſt the poore, and he wil heare the
prayer of him that is hurt. \V He wil not deſpiſe the prayers of the
pupil: nor the widow, if she power out ſpeach of mourning. \V Do not
the widows teares runne downe to the cheeke, & her exclamation vpon him
that cauſeth them to runne? \V For from the cheeke they goe vp euen to
heauen, and our Lord the hearer wil not be delighted in them. \V He that
adoreth God in delectation, shal be receiued, & his petition shal
approch euen to the cloudes. \V The prayer of him that humbleth himſelf,
shal penetrate the cloudes: and til it approch he wil not be comforted:
and he wil not depart til the Higheſt behold. \V And our Lord wil not be
long, but wil iudge the iuſt, and wil do iudgement: and the ſtrongeſt
wil not haue patience in them, that he may crush their backe: \V and he
wil repay vengeance to the Gentiles, til he take away the multitude of
the proude: & breake the ſcepters of the vniuſt, \V til he reward men
according to their doings: and according to the workes of man, and
according to his preſumption, \V til he iudge the iudgement of his
people, and shal delight the iuſt with his mercie. \V The mercie of God
is beautiful in the time of tribulation, as a cloude of raine in the
time of
\Fix{drught.}{drought.}{obvious typo, fixed in other}


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
