%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 1520
%%% o-1403
\startChapter[
  title={Chapter 09}
  ]

\Summary{Great prudence is required in conuerſation betwen men and
  wemen. 14.~Eſteme old freindes. 16.~Emulate not ſinners. 18.~Auoide
  the companie of malicious. 21.~Conſult with the prudent, hauing God
  euer before thyne eyes.}


%%% 1521
Be
\SNote{Seing ieloſie betwen man & wife is dangerous, much more
vnneceſſarie conuerſation betwen other men and wemen: eſpecially
probable occaſions of ſinne muſt be auoided.}
not ielous ouer the wife of thy boſome, leſt she shew vpon thee the
malice of wicked doctrine. \V Geue not to a woman the power of thy
ſoule, leſt she goe in thy ſtrength, and thou be confounded. \V Looke not
vpon a woman that is deſirous of manie: leſt perhaps thou fal into her
ſnares. \V With her that is a dauncer be not daily conuerſant, nor heare
her, leſt perhaps thou perish in her efficacie. \V
\CNote{\XRef{Prou.~5.}}
Behold not a virgin, leſt perhaps thou be ſcandalized in her beautie. \V
Geue not thy ſoule to harlottes in any poynt: leſt thou deſtroy thyſelf,
and thine inheritance. \V Looke not round about in the wayes of the
citie, nor wander vp and downe in the ſtreates therof. \V Turne away thy
face from a trimmed woman, and gaze not about vpon an others beautie. \V
By the beautie of a woman manie haue perished: and hereby concupiſcence
is inflamed as a fire. \V Euerie woman, that is an harlot, shal be
troden vpon as dung in the way. \V Manie hauing admired the beautie of
an other mans wife, haue become reprobate, for her communication burneth
as fire. \V Sit not at al with an other mans wife, nor repoſe vpon the
bed with her: \V and ſtriue not with her at the wine, leſt perhapes thy
hart decline toward her, & with thy bloud thou fal into perdition. \V
\SNote{Conſtancie in good thinges namely in freindſhippe is very
neceſſarie.}
Forſake not an old frend: for the new wil not be like to him. \V A new
frend, is as new wine: it shal waxe old, and thou shalt drinke it with
ſweetnes. \V
\SNote{A ſinner that proſpereth is like to a ſomer floure in the filde,
that is quickly cut downe, and withereth.}
Doe not zelouſely deſire the glorie, and the riches of a ſinner: for
thou knoweſt not what his ſubuerſion shal be. \V Let not the iniurie of
the vniuſt pleaſe thee, knowing that euen to hel the impious shal not
pleaſe. \V Be far from the man that hath power to kil, and thou shalt
not ſuſpect the feare of death. \V And if thou come to him, committe
nothing, leſt perhaps he take away thy life. \V Know it to be
communication with death; becauſe thou shalt goe in the middes of
\Fix{ſnarres,}{ſnares,}{obvious typo, fixed in other}
and shalt walke vpon the weapons of the ſorowful. \V According to
thy powre beware thee of thy neighbour; and treate with the wiſe and
prudent. \V Let iuſt men be thy gheſts, and let thy gloriation be in the
feare of God, \V and let the cogitation of God be in thy vnderſtanding,
& al thine enarration in the precepts of the Higheſt. \V Workes shal be
prayſed in the handes of artificers, and the prince of the people in the
wiſdom of his ſpeach, but the word of the ancients in the ſenſe. \V A
man ful of tongue is terrible in his citie, and he that is rash in his
word shal be odious.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
