%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 37
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-37


\project douay-rheims


%%% 1561
%%% o-1439
\startChapter[
  title={Chapter 37}
  ]

\Summary{Beware of a feaned, & loue a ſure freind. 7.~Conſult with the wiſe,
  truſtie, 15.~and vertuous, 19.~eſpecially relying vpon God. 21.~The
  tongue is cauſe of much good, or much euil. 30.~Be temperate in diet.}


Euerie freind wil ſay: I alſo haue ioyned freindshipe:
\SNote{As freindſhippe is a moſt neceſſarie thing in humane life: ſo
fained frẽdſhippe is moſt dangerous.}
but there is a freind, in name only a freind. Doth there not ſorow
remaine euen to death? \V But a companion and freind wil be turned to
enmitie. \V O moſt wicked preſumption, whence waſt thou created to couer
the drie land with malice, and with the deceitfulnes thereof? \V A
companion is pleaſant with his freind in delectations, and in the time of
tribulation he wil be an aduerſarie. \V A companion is ſorie with his
freind for his bellies ſake, and he wil take a shield againſt the
enemie. \V Forgete not thy freind in thy minde, and be not vnmindeful of
him in thy riches. \V Conſult not with him which betraieth, and hide thy
counſel from them that enuie thee. \V Euerie counſeler vttereth counſel,
but there is a counſeler in him ſelfe. \V From ſuch a counſeler keepe
thy ſoule. Firſt know what his neceſſitie is: for he wil deuiſe to his
owne minde: \V leſt perhaps he thruſt a sharpe ſtake into the ground and
ſay to thee: \V Thy way is good, and ſtand ouer againſt thee to ſee what
wil befal thee. \V With an irreligious man treate not of holines, and
with the vniuſt of iuſtice, and with a woman of the thing whereof she is
ielous, with a feareful man of warre, with a marchant of traficke, with
a byer of ſelling, with an enuious man of geuing thankes, \V with the
impious of pietie, with the vnhoneſt of honeſtie, with the field
labourere of al worke, \V with him that worketh by the yeare of the
ending of the yeare, with a ſlothful ſeruant of much working: attend not
to theſe in al counſel. \V But be
%%% 1562
continual with a holie man, whomſoeuer thou shalt know to obſerue the
feare of God, \V whoſe ſoule is according to thine owne ſoule: and who
when thou shalt ſtumble in the darke, wil be ſorie for thee. \V And
eſtablish with thy ſelf an hart of good counſel: for there is none other
thing more worth to thee then it. \V The ſoule of a holie man vttereth
ſometime true thinges, more then ſeuen watchmen that ſitte in a high
place to watch. \V And in al theſe
\SNote{Man is bond to vſe al prudent endeuour & withal moſt eſpecially to
pray for Gods direction.}
beſeech the Higheſt, that he direct thy way in truth. \V Before al
workes let a true word goe before thee, and ſtable counſel before euerie
act. \V A wicked word shal change the hart: out of which riſe foure
partes, good, and euil, life, and death: and the tongue is a continual
ruler of them. There is a ſubtile man teacher of manie, and to his owne
ſoule he is vnprofitable. \V A cunning man hath taught manie, and is
ſwete to his owne ſoule. \V He that ſpeaketh ſophiſtically, is odious:
in euerie thing he shal be defrauded. \V Grace is not geuen him
%%% o-1440
of our Lord: for he is defrauded of al wiſdom. \V There is a wiſe man,
wiſe to his owne ſoule: and the fruite of his vnderſtanding is
laudable. \V A wiſe man teacheth his people, and the fruites of his
vnderſtanding are faithful. \V A wiſe man shal be filled with
bleſſinges, and they that ſee wil praiſe him. \V The life of a man is in
the number of dayes: but the dayes
\SNote{The dayes of the Bleſſed in heauen, who ſee God, are eternal.}
of Iſrael are innumerable. \V A wiſe man in the people shal inherite
honour, and his name shal liue for euer. \V
\CNote{\XRef{1.~Cor.~6.}}
Sonne in thy life
\SNote{The ſame which S.~Paul admoniſheth
\XRef{(1.~Cor.~11.)}
\Emph{let euerie one proue him ſelf}, by examining & rectifying his
conſcience.}
proue thy ſoule: & if it be wicked, geue it not power: \V for al thinges
are not expedient for al, and euerie kinde pleaſeth not euerie ſoule. \V
Be not greedie in al feaſting, and power not out thy ſelf vpon al
meate: \V for in manie meates there shal be infirmitie, and greedines
shal approch euen to choler. \V Becauſe of ſurfet manie haue died: but
he that is abſtinent, shal adde life.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
