%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 1518
%%% o-1400
\startChapter[
  title={Chapter 07}
  ]

\Summary{Flie from al euil thinges, 4.~as ambition, preſumption,
  ſcandal, puſillanimitie, lying, and babling. 16.~Husbandrie of the
  ground, 21.~a good wife, and good ſeruantes are to be
  \Fix{cheriſed,}{cherished,}{likely typo, fixed in other}
  25.~kepe children in diſcipline. 29.~Honour parents, and elders, 36.~&
  pittie the poore. 40.~Memorie of the laſt thinges preſerueth from
  ſinne.}


Doe
\SNote{Whatſoeuer anie ſoweth that he shal reape, and the wicked shal
eate the fruites of his owne workes.}
not euils, and they shal not apprehend thee. \V Depart from the wicked,
& euil shal ſayle from thee. \V
\CNote{\XRef{Job.~9.}}
Sonne, ſow not euils in the furrowes of iuſtice, & thou shalt not reape
them ſeuenfold. \V
\CNote{\XRef{Pſal.~142.}}
Seeke not of the Lord chiefe principalitie, nor of the king the chayre
of honour. \V
\CNote{\XRef{Eccle.~7.}}
Iuſtifie not thyſelf before God, becauſe he is the knower of the hart:
and before the king deſire not to ſeme wiſe. \V
\SNote{Great prudence and fortitude are requited in al Iuſges ſpiritual
and temporal.}
Seeke not to be made a iudge, vnles thou be able by power to breake
iniquities: leſt perhaps thou feare the face of the mightie, and put a
ſcandal in thyne equitie. \V Sinne not againſt the multitude of a citie,
neither thruſt thyſelf into the people, \V nor binde together duble
ſinnes: for neither in one shalt thou be free from punishment. \V Be not
fayntharted in thy minde: \V deſpiſe not to pray, and to geue almes. \V Say
not: In the multitude of my giftes God wil haue reſpect, and when I
offer to
%%% o-1401
God moſt high, he wil receiue my giftes. \V Laugh not a man to ſcorne in
the bitternes of his ſoule: for there is that humbleth and exalteth, God
the ouerſeer of al. \V Plowe not a lie agaynſt thy brother: neither doe
thou likewiſe agaynſt thy frend. \V Be not willing to make any lie: for
the cuſtome therof is not good. \V Be not ful of wordes in a multitude
of ancients, and
%%% !!! This SNote is marked in both, but the next SNote in this is
%%% missing in other, but that one is probably the correct location for
%%% the actual note.
%%% \SNote{}
iterate not a word in thy ſpeach. \V Hate not laborious workes, and
husbandrie created of the Higheſt. \V Counte not thy ſelfe in the
multitude of men without diſcipline. \V Be mindeful of wrath,
becauſe it wil not ſlacke. \V Humble thy ſpirit very much: becauſe the
vengeance of the flesh of the impious, is
\SNote{In hel are two puniſhments: fire burning, and the worme of
conſcience vexing the ſoule, & both are eternal.}
fyre and the worme. \V Doe not preuaricate againſt thy frend differring
money, nor deſpiſe thy deareſt brother for gold. \V Depart not from a
wiſe and good woman, which thou haſt gotten in the feare of our
%%% 1519
Lord: for the grace of her bashfulnes is aboue gold. \V
\CNote{\XRef{Leuit.~19.}}
Hurt not the ſeruant that worketh in truth, nor the hyred man that
geueth his ſoule. \V Let a wiſe ſeruant be beloued of thee as thy ſoule,
defraude him not of libertie, nor leaue him needie. \V Haſt thou cattel?
looke wel to them: and if they be profitable, let them continew with
thee. \V Haſt thou children? inſtruct them, & bowe them from their
childehood. \V Haſt thou daughters? keepe their bodie, and shew not thy
countenance merrie towards them. \V Beſtow thy daughter, and thou shalt
doe a great worke, and geue her to a wiſe man. \V If thou haue a wife
according to thy ſoule, caſt her not of: and to her that is hateful
commit not thyſelf. With thy whole hart \V
\CNote{\XRef{Tob.~4.}}
honour thy father, and forget not the gronings of thy mother: \V
remember that thou hadſt not bene borne but by them: and recompence
them, as they alſo thee. \V In al thy ſoule feare our Lord, and
ſanctifie his prieſtes. \V With al thy ſtrength loue him that made thee:
& forſake not his miniſters. \V
\CNote{\XRef{Deut.~12.}}
Honour God with al thy ſoule, and honour the prieſtes, and purge thyſelf
with the armes. \V
\CNote{\XRef{Leuit.~2.}}
Geue them the portion, as it is commanded thee, of the firſt fruites and
purgation: and of thy negligence purge thy ſelf with few. \V
\CNote{\XRef{Num.~18.}}
The gift of thyne armes and the ſacrifice of ſanctification thou shalt
offer to our Lord, and the firſt of holie thinges: \V and to the poore
ſtretch out thyne hand, that thy propitiation may be perfected, and thy
bleſſing. \V The grace of a gift is in the ſight of al the liuing:
\SNote{VVorkes of mercie are alſo profitable to the dead, as prayer,
almes, and ſacrifice for ſoules in purgatorie.}
and from the dead ſtay not grace. \V
\CNote{\XRef{Rom.~12. v.~15.}}
Want not in conſolation to them that weepe, and walke with them that
moorne. \V Be not loth to viſite the ſicke: for by theſe thinges thou
shalt be confirmed in loue. \V In al thy workes
\SNote{A moſt eſpecial preſeruatiue from ſinne.}
remember thy later ends, and thou wilt not ſinne for euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
