%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 1515
%%% o-1398
\startChapter[
  title={Chapter 05}
  ]

\Summary{Let not riches, youth, nor ſtreingth hold thee in ſinne: 5.~but
  do penance for ſinnes remitted, and adde not ſinne vpon
  ſinne. 6.~Neither preſume to ſinne, becauſe God is merciful. 10.~Be
  not couetous, nor vnconſtant. 13.~Be meeke in lerning, and careful in
  teaching, 16.~not a batemaker, nor duble tongued.}


Attend not to vniuſt poſſeſſions, and ſay not: I haue ſufficient
liuelihood: for it shal nothing profite in the time of vengeance &
affliction. \V Folow not in
\SNote{Though thou be ſo potent, that no man can hinder thy euil
deſignment; yet do it not becauſe God wil puniſh it.}
thy ſtrength the concupiſcence of thy hart: \V and ſay not: How mightie
am I? and who shal bring me vnder for my factes? for God reuenging wil
reuenge. \V Say not: I haue ſinned, and what ſorowful thing hath chanced
to me? For the Higheſt is a patient rewarder. \V
\SNote{After that ſinne is forgeuen there remaineth oftentimes
puniſhment for ſatisfaction.}
Of ſinne forgeuen be not without feare, neither adde thou ſinne vpon
ſinne. \V And ſay not: The mercie of our Lord is great, he wil haue
mercie on the multitude of my ſinnes. \V For mercie and wrath quickly
approch from God, and his wrath looketh vpon ſinners. \V Slacke not to
be conuerted to our Lord, and differre not from day to day. \V For his
wrath shal come ſodainly, and in the time of vengeance he wil deſtroy
thee. \V Be not careful in vniuſt riches: for they shal not profite thee
in the day of
\TNote{Couering, or death.}
obduction and vengeance. \V 
\CNote{\XRef{Prou.~10. v.~4.}}
Toſſe not thy ſelf into euerie winde, and goe
%%% 1516
not into euerie way: for ſo euerie ſinner is proued by a duble
tongue. \V
\SNote{Conſtancie in good purpoſe, meriteth}
Be ſtedfaſt in the way of our Lord, & in the truth of thy vnderſtanding
and in knowlege, and let the
\SNote{the promiſed peace, and iuſt reward.}
word of peace and iuſtice accompanie thee. \V Be milde to heare the
word, thou maiſt vnderſtand: and with wiſdom vtter thou a true
anſwer. \V If thou haue vnderſtanding, anſwer thy neighbour: but if not,
let thine hand be vpon thy mouth, leſt thou be taken in an vnskilful
word, and be confounded. \V Honour and glorie in the word of the wiſe,
but the tongue of the vnwiſe is his ſubuerſion. \V Be not called a
whiſperer and be not taken in thy tongue, & confounded. \V For vpon a
theefe is confuſion & repentance, and a verie euil condemnation vpon the
duble tongued, but to the whiſperer hatred, and emnitie, and
contumelie. \V Iuſtifie thou the litle one, and the great alike.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
