%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 25
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-25


\project douay-rheims


%%% 1544
%%% o-1424
\startChapter[
  title={Chapter 25}
  ]

\Summary{Concord betwen bretheren, neighboures, and man and wife, much
  pleaſeth God. 3.~A poore man proud, a richman a lier, and an old man
  doting in carnal, or worldlie thinges, are very hateful. 9.~He that
  ſeeth his children good; and his enimies ouerthrowne; hath a good
  wife; offendeth not in ſpeach; conſenteth not to ſinne; hath a true
  freind; teacheth good doctrine; hath ſacred; and humane knowlege: hath
  vndoubtedly nine happie thinges: but to feare God conteyneth 14.~and
  excelleth al. 17.~A wicked woman (hereſie) is very deteſtable, 30.~and
  moſt vntolerable, if she haue ſupreme dominion.}


In three thinges my ſpirit is pleaſed, which are approued before God,
and men: \V
\SNote{Three very commendable thinges.}
The concord of bretheren, and the loue of neighboures, and man and wife
wel agreeing together. \V Three ſortes my ſoule hateth, and I am
greatly greeued at their life; \V
\SNote{Other three deteſtable.}
A poore man proud: & a rich man a lyer: an old man a foole, and
doting. \V The thinges that thou haſt not gathered in thy youth, how
shalt thou find them in thy old age? \V How beautiful is iudgement for a
grey head, and for ancientes to know counſel! \V How beautiful is wiſdom
for the aged, & vnderſtanding glorious, and counſel! \V Much cunning is
the croune of old men, and the feare of
%%% 1545
God is their glorie. \V
\SNote{Nine happie thinges in this life.}
Nine thinges not to be imagined of the hart haue I magnified, and the
tenth I wil tel vnto men with my tongue: \V A man that hath ioy in his
children; liuing and ſeeing the ſubuerſion of his enemies. \V
\CNote{\XRef{Iac.~3.}}
Bleſſed is he that dwelleth with a wiſe woman; & that hath not offended
with his tongue; and that hath not ſerued ſuch as are vnworthie of
him. \V Bleſſed is he that findeth a true freind; and that declareth
iuſtice to an eare that heareth. \V How great is he, that findeth
wiſdom; and knowlege; but he is not aboue him
\SNote{Al happines in this life is grounded in the feare of God.}
that feareth our Lord. \V The feare of God hath ſet it ſelf aboue al
thinges: \V bleſſed is the man, to whom is geuen to haue the feare of
God: he that holdeth it, to whom shal he be reſembled? \V The feare of
God is the beginning of his loue: and the beginning of faith is to be
faſt ioyned vnto it. \V The heauines of the hart is al plague: & al
malice,
\SNote{The deſcription of hereſie vnder the figure of a wicked woman:
whoſe malice is ſecretly couered vnder pretence of truth and pietie.}
the wickednes of a woman. \V And he wil ſee al plague, and not the
plague of the hart: \V & al wickednes, & not the wickednes of a
woman: \V and al obduction, and not the obduction of them that hate
him: \V and al reuenge, and not the reuenge of the enemies. \V
\CNote{Prou.~21.}
There is no head worſe then the head of a ſerpent: \V and there is no
anger aboue the anger of woman. It shal be more pleaſant to abide with a
lyon and dragon, then to dwel with a wicked woman. \V The wickednes of a
woman changeth her face: and
%%% o-1425
darkeneth her countenance as a beare: and wil shew it as a ſacke. In the
middes of her neighbours, \V her husband groned, and hearing he ſighed a
litle. \V Al malice is short to the malice of a woman, the lot of
ſinners fal vpon her. \V As the goeing vp a grauelie way in the feete of
the aged, ſo a woman ful of tongue to a quiet man. \V Looke not vpon a
womans beautie, and deſire not a woman for beautie. \V A womans anger,
and impudencie, and confuſion is great. \V A woman
\SNote{Lay headſhipe in ſpiritual cauſes is ſo vnreaſonable and abſurde,
that few heretikes wil indure it.}
if she haue ſuperioritie, is contrarie to her husband. \V An humbled hart, and 
heauie countenance, and plague of hart, is a wicked woman. \V Feeble
handes, and diſiointed knees, a woman that doth not make her husband
happie. \V From woman came the beginning of ſinne, and by her we doe al
die. \V Geue not iſſue to thy water, no not a litle: nor to a wicked
woman leaue, to goe forth. \V If she walke not at thine hand, she wil
confound thee in the ſight of thyn enemies. \V Cut her of from thy
flesh, leſt she alwaies abuſe thee.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
