%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 45
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-45


\project douay-rheims


%%% 1573
%%% o-1449
\startChapter[
  title={Chapter 45}
  ]

\Summary{Praiſes of Moyſes, 7.~Aaron, 16.~and his prieſtlie
  progenie. 22.~Againſt whom Chore with his complices rebelling, were
  deſtroyed.}


Moyſes beloued of God, and men: whoſe memorie is in benediction. \V
\CNote{\XRef{Exo.~3.}}
He made him like in the glorie of ſaintes, and magnified him in the
feare of his enemies. And with his wordes he appeaſed monſters. \V He
glorified him in the ſight of kinges, and gaue him commandment before
his people, & shewed him his glorie. \V
\CNote{\XRef{Num.~21.}}
In his faith and meekenes he made him holie, and choſe him of al
flesh. \V For he heard him, and his voice, and brought him into a
cloude. \V And he gaue him precepts
\SNote{Moyſes ſaw Gods workes more clerely then other Prophets, yet ſaw
not his ſubſtance, as is noted
\XRef{Exo.~33.}}
face to face, and a law of life and diſcipline, to
%%% o-1450
teach Iacob his teſtament, and Iſrael his iudgements. \V He exalted
Aaron his brother high, and like to himſelf of the tribe of Leui. \V
\CNote{\XRef{Exo.~28.}}
He eſtablished vnto him
\SNote{Aarons prieſthood continued ſo long as Moyſes law: that was til
Chriſt. And now the prieſthood according to the order of Melchiſedech
continueth to the end of the world.}
an euerlaſting teſtament, and gaue him the prieſthood of the nation, and
made him bleſſed in glorie, \V and he girded him about with a girdle,
and put vpon him a robe of glorie, and crowned him in furniture of
power. \V Garments to the feete, and breches, and an Ephod he put vpon
him, and compaſſed him with litle
%%% 1574
belles of gold very manie round about, \V to geue a ſound in his going,
to make ſound heard in the temple for a memorie to the children of his
nation. \V An holie robe, of gold, and hyacinthe, and purple, a wouen
worke, of a wiſeman, indued with iudgement and truth: \V Of twiſted
ſcarlet the worke of an artificer, with precious ſtones figured in the
cloſure of gold, and grauen by the worke of a lapidarie for a memorial,
according to the number of the tribes of Iſrael. \V A crowne of gold
vpon his miter grauen with a ſeale of holines, and the glorie of honour:
a worke of power, and the adorned deſires of the eies. \V There were
none ſuch ſo faire before him, euen from the beginning. \V No ſtranger
was clothed with them, but only his children alone, and his nephewes for
euer. \V
\CNote{\XRef{Leuit.~8.}}
His ſacrifices were conſumed with fire euerie day. \V Moyſes filled his
handes, & anoynted him with holie oile. \V It was made vnto him for an
euerlaſting teſtament, and to his ſeede as the daies of heauen, to doe
the function of prieſthood, and to haue praiſe and to glorifie his
people in his name. \V He choſe him of al that liued, to offer ſacrifice
to God, incenſe, and good odour, for a memorial to pacifie for his
people: \V and he gaue them power in his preceptes, in the teſtaments of
his iudgementes, to teach Iacob his teſtimonies, and in his law to geue
light to Iſrael. \V
\CNote{\XRef{Num.~16.}}
Becauſe ſtrangers ſtood againſt him, and for enuie men compaſſed him
about in the deſert, they that were with Dathan and Abiron, and the
congregation of Core in anger. \V Our Lord God ſaw, and it pleaſed him
not, and they were conſumed in the violence of wrath. \V He did
prodigious thinges vnto them, and conſumed them in flame of fire. \V And
he added glorie to Aaron, and gaue him an inheritance, and diuided vnto
him the firſt fruites of the increaſe of the earth. \V He prepared them
bread in the firſt vnto ſatietie: for the ſacrifices alſo of our Lord
they shal eate, which he gaue to him, and to his ſeede. \V But
\SNote{The tribe of Leui had not a portion of inheritance ſeparate from
the reſt, but had tithes, firſt fruites, and oblations for their
temporal prouiſion.}
he shal not inherite the nations in the land, and he hath no part in the
nation: for himſelf is his portion & inheritance. \V
\CNote{\XRef{Num.~25.}}
Phinees the ſonne of Eleazar is the third in glorie, in imitating him in
the feare of our Lord: \V and to ſtand in the reuerence of the nation:
in the goodnes and alacritie of his ſoule he pacified God for Iſrael. \V 
Therefore did he eſtablish vnto him a couenant of peace, to be the
prince of the holies, and of his nation, that the dignitie of prieſthood
should
%%% 1575
be to him and to his ſeede for euer. \V And the teſtament to
\SNote{King Dauid gaue ſpecial aſſiſtance to the Prieſtes, and greatly
aduanced Gods ſeruice.
\XRef{1.~Paral.~23. &c.}}
Dauid king, the ſonne of Ieſſe of the tribe of Iuda, and inheritance to
him and to his ſeede, that he might geue wiſdom into our hart to iudge
his nation in iuſtice, that their good thinges might not be abolished, &
their glorie in their nation he made euerlaſting.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
