%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 36
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-36


\project douay-rheims


%%% 1560
%%% o-1438
\startChapter[
  title={Chapter 36}
  ]

\Summary{A prayer for conuerſion of al nations: 14.~and for conſeruation
  of the Iſraelites. 20.~Diſcretion is neceſſarie in al actions, and
  deſires.}


Haue mercie vpon vs ô God of al, and reſpect vs, and shew vs the light
of thy mercies: \V and ſend in thy feare vpon
\SNote{This prayer implieth alſo a prophecie of the conuerſion of the
Gentiles, as the like very often in the Pſalmes.
\XRef{58.~64.~65.}
&c.}
the nations, that haue not ſought after thee, that they may know that
there is no God but thou, and that they may shew forth thy glorious
thinges. \V Lift vp thy hand ouer the ſtrange Nations, that they may ſee
thy might. \V For as in their ſight thou art ſanctified in vs, ſo in our
ſight thou shalt be magnified in them, \V that they may know thee, as we
alſo haue knowen, that there is no God beſide thee ô Lord. \V Renewe
ſignes, and change meruels. \V Glorifie thy hand, and thy right arme. \V
Raiſe vp furie, and power out wrath. \V Take away the aduerſarie, and
afflict the enemie. \V Haſten the time, and remember the end, that they
may declare thy meruels. \V Let him that is ſaued be deuoured in the
wrath of flame: and let them that euil intreate thy people, finde
perdition. \V Breake the head of princes of the enemies, that ſaie:
There is none other beſide vs. \V Gather together al the tribes of
Iacob: and let them know that there is no God but thou, that they may
declare thy great workes: & thou shalt inherite them as from the
beginning. \V
\CNote{\XRef{Exo.~4.}}
Haue mercie on thy people, vpon which thy name is inuocated: and vpon
Iſrael, whom thou haſt made equal to thy firſt begotten. \V Haue mercie
on the citie of thy ſanctification Ieruſalem, the citie of thy reſt. \V
Replenish Sion with thy wordes that can not be vttered, & thy people
with thy glorie. \V Geue the teſtimonie to them, that are thy creatures
from the beginning, and raiſe vp the prophecies, which the former
prophets ſpake in thy name. \V Geue reward to them that patiently expect
thee, that thy prophets may be found faithful: and heare the prayers of
thy ſeruants, \V
\CNote{\XRef{Num.~6.}}
according to Aarons benediction of thy people, and direct vs into the
way of iuſtice, and let al knowe that inhabite the earth, that thou art
God the beholder of the worldes. \V The
\SNote{As there is difference of meates:}
bellie wil eate al meate, and one meate is better then an other
meate. \V The iawes taiſt veniſon, &
\SNote{ſo there ought to be diſcretion in wordes, in chooſing a wife, &
in al actions, and deſires.}
the wiſe hart lying wordes. \V A peruerſe hart wil geue ſorow, and a
cunning man wil reſiſt it. \V Some woman wil receiue euerie man: and one
daughter is better then an other daughter.
%%% 1561
\V The beautie of a woman chereth the face of her husband, and
increaſeth the deſire aboue al mans concupiſcence. \V If there be a
tongue of curing, there is alſo of mitigating and of mercie: her husband
is not according to the ſonnes of men. \V He that poſſeſſeth a good
woman, beginneth riches: she is an helpe like vnto him, & a piller as
\SNote{In concord ſmal thinges increaſe, by diſcord al thinges goe to
hauocke.}
reſt. \V Where there is no hedge, the poſſeſſion shal be ſpoiled: and
where there is no wife, he mourneth wanting. Who doth credite him that
hath no reſt, and turning aſide whereſoeuer it waxeth darke, as a robber
girded, leaping from citie to citie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
