%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 41
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-41


\project douay-rheims


%%% 1567
%%% o-1444
\startChapter[
  title={Chapter 41}
  ]

\Summary{An other matter of meditation is death, 8.~wherof ſinne is the
  cauſe. 15.~Care of a good fame is neceſſarie. 19.~Let shamfaſtnes be a
  bridle to auoide fornication, 22.~iniquitie, 24.~theft, and other
  ſinnes.}


%%% 1568
O death how bitter is thy memorie to a man that hath peace in his
riches: \V to a man that is at reſt, and whoſe wayes are proſperous in
al thinges, and that is yet able to take meate! \V O death, thy
iudgement is good to a needy man, and him that is diminished in
ſtrength, \V and fayleth in age, and that is careful of al thinges, and
to the incredulous, that loſeth
%%% o-1445
patience! \V Feare not the iudgement of death. Remember what thinges
haue bene before thee, and what come after thee: this is the iudgement
from our Lord to al flesh: \V and what shal come vpon thee by the good
pleaſure of the Higheſt? whether it be ten, or an hundred, or a thouſand
yeares. \V For in hel there is no
\SNote{It is to no purpoſe after death to accuſe the ſhortnes, or length
of life, pretending the ſame to haue bene cauſe of ſinne. For God doth
al iuſtly, yea and for the beſt, if men would ſo vſe his benefites.}
accuſing of life. \V The children of ſinners become children of
abominations, and they that conuerſe neere the houſes of the impious. \V
The inheritance of the children of ſinners shal perish, and with their
ſeede shal be continuance of reproch. \V The children complaine of an
impious father, becauſe for him they are in reproch. \V Woe to you ye
impious men, which haue forſaken the law of our Lord the Higheſt. \V And
if ye be borne, ye shal be borne in malediction: and if ye die, in
malediction shal be your portion. \V Al thinges that are of the earth,
shal returne into the earth: ſo the impious from malediction to
perdition. \V The
\Fix{moorning}{mourning}{likely typo, fixed in other}
of men is in their bodie, but the name of the
impious shal be cleane wyped out. \V Haue care of a good name: for this
shal be more permanent to thee, then a thouſand treaſures precious and
great. \V There is a number of the daies of a good life: but a good name
shal continew for euer. \V Children, keepe ye diſcipline in peace. For
wiſdom hid, and treaſure not ſeene, what profite is there in them
both? \V Better is the man that hideth his follie, then the man that
hideth his wiſdom. \V But yet
\SNote{It behoueth children, and ſcholars to eſteme that which their
elders teach:}
haue reuerence to theſe thinges, which proceede from my mouth. \V For it
is not good to obſerue al shamfaſtnes: &
\SNote{though the ſame doth not ſeme reaſonable in their owne opinion.}
al thinges do not pleaſe al men in opinion. \V Be ashamed before father
& before mother, of fornication: and before the preſident and before the
mightie, of lying: \V before the prince, and before the iudge, of
offence: before the ſinagogue and the people, of iniquitie: before
companion and freind, of iniuſtice: and before the place where thou
dwelleſt, \V of theft, of the truth of God, and his teſtament: of
leaning on the bread, and of reproofe for the thing geuen and taken: \V
before them that
%%% 1569
ſalute thee, of ſilence: of beholding a woman that is an harlot: and of
turning away thy countenance from thy kinſman. \V Turne not away thy
face from thy neighbour, & of taking away part and not reſtoring. \V
\CNote{\XRef{Mat.~5. v.~28.}}
Behold not an other mans wife, and ſearch not his handmayde, neither
ſtand by her bed. \V Before freindes of opprobrious wordes: and when thou
haſt geuen, vpbrayde not.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
