%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 1536
%%% o-1416
\startChapter[
  title={Chapter 20}
  ]

\Summary{Correction ought to be without paſſion of anger. 5.~Much
  diſcretion is required in ſpeach, and in ſilence. 14.~Alſo in geuing,
  20.~& promiſing, 29.~and in vttering knowlege, where and when it
  behoueth.}


How good a thing is it to rebuke, rather then to be angrie, and not to
hinder him that confeſſeth in prayer! \V The concupiſcence of an eunuch
shal defloure a young maide: \V ſo he that by violence doth vniuſt
iudgement. \V
\CNote{\XRef{Eccle.~3.}}
How good a thing is it
\SNote{He that taketh reprehenſion in good part when he is faulty
meriteth pardon, and when he is not faultie he ſatisfieth for his other
ſinnes and meriteth reward.}
being rebuked to shew repentance! for ſo thou shalt auoid wilful ſinne. \V
There is that holdeth his peace, which is found wiſe: and there is that
is odious, he which is malepert to ſpeake. \V There is that holdeth his
peace, knowing the time of fitte opportunitie. \V A wiſe man wil hold
his peace vntil a time: but a wanton, and the vnwiſe wil not obſerue
time. \V He that vſeth manie wordes, shal hurt his owne ſoule: & he
that taketh authoritie to himſelf vniuſtly, shal be hated. \V There is
proceding in euils to a man without diſcipline, and there is finding to
loſſe. \V There is a gift, that is not profitable: and there is a gift,
the reward whereof is duble. \V There is debaſing becauſe of
%%% o-1417
glorie: and there is that from humilitie shal lift vp the head. \V There
is that redemeth manie thinges for a ſmale price, and reſtoreth the ſame
ſeuen fold. \V A man wiſe in wordes shal make him ſelfe beloued: but the
graces of fooles shal be powred out. \V The gift of the vnwiſe shal not
be profitable for thee: for his eies are
\SNote{His intention is ful of guile that flattereth by ſhew of loue and
of praiſe, but he wil detract ſo much the more, & in the end reproch
thee when he may gette probable aduantage againſt thee.}
ſeuenfold. \V He wil geue few thinges, and vpbraide manie: and the
opening of his mouth is an inflammation. \V To day a man lendeth, and to
morow he asketh it againe: ſuch a man is odious. \V A foole shal haue no
freind, and there shal be no thanke to his goodes. \V For they that eate
his bread, are of a falſe tongue. How often, and how manie wil laugh him
to ſcorne? \V For he doth not diſtribute with right vnderſtanding, that
which was to be had: in like manner alſo that which was not to be had. \V
The flipping of a falſe tongue, as he that falleth on the pauement: ſo
the falles of the euil shal come haſtely. \V A man without grace is as a
vaine fable, it shal be continual in the mouth of them that are without
diſcipline. \V A parable out of a fooles
%%% 1537
mouth shal be reiected: for he doth not ſpeake it in his time. \V There
is that is forbid to ſinne for pouertie, and in his reſt he shal be
pricked. \V There is that wil deſtroy his ſoule for shamfaſtnes, and by
an vnwiſe perſon he wil deſtroy it: and by acception of perſon he wil
deſtroy himſelf. \V There is that for shamfaſtnes promiſeth to his
freind, and hath gotten an enemie of him for naught. \V Lying is a
wicked reproch in a man, and in the mouth of men without diſcipline it
shal be continually. \V Better is a theefe, then the continual cuſtome
of a lying man, but both shal inherite perdition. \V The maners of lying
men are without honour: and their confuſion is with them without
intermiſſion. \V A wiſeman in his wordes shal vtter himſelf, and a
prudent man shal pleaſe great perſons. \V
\CNote{\XRef{Prou.~12.}}
He that tilleth his land, shal make an high heape of corne: and who ſo
worketh iuſtice, he shal be exalted: and he that pleaſeth great men,
shal auoide iniquitie. \V
\CNote{\XRef{Exo.~23.}
\XRef{Deut.~16.}}
Preſentes, & giftes blind the eies of iudges, and as one dumbe in the
mouth turneth away their chaſtiſementes. \V
\SNote{VViſdom is to be ſhewed in vvordes and deedes vvhen it may
profite others.}
Wiſdom hid, and treaſure not ſene: what profit is there in both? \V
\SNote{To conceale faultes, ſo they be amended, is moſt conuenient.}
Better is he that concealeth his foolishnes, then the man that hideth
his wiſdom.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
