%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-23


\project douay-rheims


%%% 1540
%%% o-1420
\startChapter[
  title={Chapter 23}
  ]

\Summary{A prayer againſt pride, 6.~gluttonie, and luxurie. 7.~Beware of
  offending in ſpeach, 9.~eſpecially of vnlawful ſwearing,
  15.~blaſphemie, 17.~irreuerent and reprocheful wordes. 21.~Alſo of
  auarice, 24.~fornication, and adultrie, 30.~al which God ſeeth, and
  wil ſeuerly punish. 33.~With other ſinnes that folow therof.}


O Lord father, and dominatour of my life
\SNote{Man being weake and the enimie ſuttle & ſtrong, he is not able to
reſiſt tentations without Gods ſpecial grace, for which the vviſman
therfore prayeth, inſtructing al by his example to do the ſame.}
leaue me not in their counſel: nor ſuffer me to fal in them. \V Who
laieth on ſtripes in my cogitation, and in my hart the doctrine of
wiſdom, and in their ignorances they ſpare me not, and their offences
appeare not, \V and my ignorances increaſe not, and my offences be
multiplied, and my ſinnes abound, and I fal in the ſight of mine
aduerſaries, & mine enemies reioyce. \V O Lord father, and God of my
life, leaue me not in their cogitation. \V Hautines of mine eies geue me
not, and al deſire turne away from me. \V Take from me the
concupiſcences of the bellie, and let not the concupiſcences of
copulation take hold of me, and geue me not ouer to a shameleſſe and
foolish minde. \V O children heare
\SNote{Precepts how to gouerne the mouth, and tongue.}
the doctrine of the mouth: and he that wil keepe it, shal not perish by
his lippes, nor be ſcandalized in moſt wicked workes. \V A ſinner is
taught in his
%%% 1541
vanitie, and the proud and the euil ſpeaker shal be ſcandalized in
them. \V Let not thy mouth be
%%% o-1421
accuſtomed
\SNote{Againſt raſh, vntrue, & vnlawful ſwearing.
\XRef{Iere.~4.}}
to ſwearing: for there be manie falles in it. \V
\CNote{\XRef{Mat.~5.}}
But let not the naming of God be vſual in thy mouth, and meddle not with
the names of Sainctes, becauſe thou shalt not ſcape free from them. \V For as
a ſeruant daily examined, lacketh not the marke therof: ſo euerie one
that ſweareth, and nameth, shal not be wholy purged from ſinne. \V A man
that ſweareth much shal be filled with iniquitie, and plague shal not
depart from his houſe. \V And if he fruſtrate it, his ſinne shal be vpon
him: and if he diſſemble, he offendeth duble: \V and if he ſweare in
vaine, he shal not be iuſtified: for his houſe shal be filled with
retribution. \V There is an other
\SNote{In oathes God is called to witnes as he that can not lie, but
blaſphemie attributeth that to God which perteyneth not
\Fix{him,}{to him,}{obvious typo, fixed in other}
or attributeth to ſome creature that which only belongeth to God, and ſo
is a contrarie ſinne to vnlawful ſwearing.}
contrarie ſpeach, alſo
\SNote{Both are mortal ſinnes.}
to death, be it not found in the inheritance of Iacob. \V For of the
merciful al thoſe thinges shal be taken away, and they wil not wallow in
ſinnes. \V Let not thy mouth be accuſtomed to
\SNote{Admonition againſt ſinnes of the tongue.}
vnnuttered ſpeach: for there is in it a word of ſinne. \V
\SNote{Reproch to parents, and other neighboures.}
Remember thy father and thy mother, for thou ſitteſt in the middeſt of
greate men: \V leſt perhapes God forget thee in their ſight, and being
ſotted with thy dailie cuſtome, thou ſuffer reproch, and hadſt better
not haue bene borne, and curſe the day of thy natiuitie. \V A man
accuſtomed to the wordes of reproch, wil not be inſtructed in al his
daies. \V Two ſortes abound in ſinnes, and the third bringeth wrath and
perdition. \V An
\SNote{Againſt couetouſnes.}
hote ſoule as a burning fire wil not be quenched, til it ſwallow
ſomewhat, \V and a wicked man in the mouth of his flesh wil not ceaſe
til he kindle a fire. \V To a man that is
\SNote{Fornication.}
a fornicatour al bread is ſweete, he wil not be wearie tranſgreſſing
vnto the end. \V Euerie man that paſſeth
\SNote{Adultrie.}
beyond his owne bed, contemning againſt his owne ſoule, and ſaying: Who
ſeeth me? \V
\CNote{\XRef{Iſa.~29.}}
Darkenes compaſſeth me, and the walles couer me, and no man beholdeth
me: whom do I feare? the Higheſt wil not be mindful of my ſinnes. \V And
he vnderſtandeth not that his eye ſeeth al thinges, for that ſuch feare
of man expelleth from him the feare of God, & the eies of men fearing
him: \V and he knoweth not that the eies of our Lord are much more
brighter then the ſunne, beholding round about al the waies of men, and
the botome of the depth, and the hartes of men looking into the hidden
partes. \V For al thinges were knowen to our Lord God, before they were
created: ſo alſo after it is perfected he beholdeth al
%%% 1542
thinges. \V
\CNote{\XRef{Leui.~20.}
\XRef{Deut.~22.}}
\SNote{Seing carnal adultrie ſhal be ſeuerely puniſhed, much more
ſpiritual: as ſchiſme, hereſie, and apoſtaſie from Catholique Religion.}
Reuenge shal be taken on this man in the ſtreates of the cittie, and as
an horſecolt he shal be chaſed: and where he expected not, he shal be
apprehended. \V And he shal be in dishonour with al men, for that he
vnderſtood not the feare of our Lord. \V So euerie woman alſo that
forſaketh her husband, & getteth inheritance by mariage of an other. \V
For firſt she hath bene vnfaithful in the law of the Higheſt: and
ſecondly she hath ſinned againſt her husband: thirdly she hath
fornicated in adulterie, and hath gotten her children of an other
man. \V This woman shal be brought into the Church, and vpon her
children there shal be examination. \V Her children shal not take roote,
and her boughes shal not yeld fruite. \V She shal leaue her memorie to
be curſed, and her dishonour shal not be wiped out. \V And they that are
leaft shal know, that nothing is better then the feare of God: and
nothing ſweter, then to haue regard
%%% o-1422
to the commandmentes of our Lord. \V It is great glorie to folow our
Lord: for length of daies shal be taken of him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
