%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 1523
%%% o-1405
\startChapter[
  title={Chapter 11}
  ]

\Summary{Wiſdom by humilitie meriteth exaltation. 7.~Iudge not before
  examination. 16.~Truſt not in riches. 14.~God ſendeth both proſperitie
  and aduerſitie, for the good of his ſeruants. 31.~Take heede of the
  deceiptful.}


The
\SNote{A wiſman humbling himſelf by penance as Daniel did, or being
vniuſtly humbled by others, as Ioſeph was shal be exalted by God.}
wiſdom of the humble shal exalt his head, & shal make him ſitte in the
middes of great men. \V Praiſe not a man in his beautie, neither deſpiſe
a man by his looke. \V The bee is ſmal among fowles, and her fruite hath
the beginning of ſweetnes. \V In apparel doe not glorie at any time,
nor be extolled in the day of thine honour, becauſe the workes of the
Higheſt onlie be meruelous, and his workes are glorious,
%%% 1524
and ſecrete, and not ſeene. \V Manie tyrantes haue ſitte in the throne,
and he whom no man would thincke hath worne the crowne. \V Manie mightie men
haue bene greatly oppreſſed, and the glorious haue bene deliuered into
the handes of others. \V Before thou enquire, blame no man: and when
thou haſt enquired, chaſten iuſtly. \V
\SNote{Expect the end of an other mans ſpeach, before you beginne to
anſwer.}
Before thou heare, anſwer not a word, and in the middes
\SNote{Expect alſo if anie that is elder, or better able wil anſwer
firſt.}
of ancients adde not to ſpeake. \V
\CNote{\XRef{Deut.~13.}}
Striue not for that thing, which doeth not moleſt thee: and conſiſte not
in the iudgement of ſinners. \V Sonne let not thy doings be in manie
thinges: and if thou be rich, thou shalt not be free from ſinne. For if
thou purſew, thou shalt not attayne: and if thou runne before, thou
shalt not eſcape. \V There is one that laboureth, and haſteneth, and is a
ſorowful impious man, and ſo much the more he shal not abound. \V
\CNote{\XRef{Iob.~42.}}
There is a lither man that wanteth recouerie, more fayling in ſtrength,
and abunding in pouertie: \V and the eie of God hath reſpected him in
good, and hath erected him from his low eſtate,
%%% o-1406
and hath exalted his head: and manie haue merueled at him, and haue
honoured God. \V Good thinges and euil, life and death, pouertie and
honeſtie are of God. \V Wiſdom and diſcipline, and the knowlege of the
law are with God. Loue, and the wayes of good thinges are with him. \V
\SNote{One puniſhment of ſinne is blindnes of hart. Eſpecially where is
no remorſe of conſcience.}
Errour and darkenes are created with ſinners: and they that reioyce in
euils, waxe old in euil. \V The gift of God is permanent to the iuſt,
and his proſpering shal haue ſucceſſe for euer. \V There is that is
enriched by doing ſparingly, and this is the portion of his reward \V in
that he ſayth: I haue found me reſt, and now I wil eate of my goods
alone: \V and he knoweth not that time paſſeth, & death approcheth, and
he muſt leaue al to others, and shal die. \V Stand in thy couenant, and
commen therein, and grow old in the worke of thy commandements. \V Abide
not in the workes of ſinners. But truſt in God, and tarie in thy
place. \V For it is eaſie in the eies of God ſodainly to enrich the
poore man. \V The bleſſing of God haſteth to the reward of the iuſt, and
in a ſwift houre his proſpering fructifieth. \V Say not: What neede I,
and what good shal I haue by this? \V Say not: I am ſufficient for my
ſelf: and what shal I be made worſe by this? \V
\SNote{In proſperitie is feare, and in aduerſitie hope of change.}
In the day of good thinges be not vnmindful of euils: and in the day of
euils be not vnmindful of good thinges: \V becauſe
it is eaſie before God in the day of death to reward
%%% 1525
euerie one according to his wayes. \V The malice of an houre maketh
obliuion of great voluptuouſnes, and in the end of a man is the
diſcloſing of his workes. \V Before death prayſe no man, becauſe a man
is knowen in his children. \V Bring not euerie man into thine houſe: for
there be manie traynes of the deceitful man. \V For as the ſtomakes
belche of ſtinking breathes, and as the partriche is brought in the
cage, and as the doe into the ſnare: ſo alſo the hart of the proude, and
as a watche man that ſeeth the fal of his neighbour. \V For turning good
thinges into euil he lyeth in wayte, and on the elect he wil lay a
blot. \V For of one ſparke fire is increaſed, and of a deceitful man
bloud is increaſed: and a ſinful man lyeth in wayte for bloud. \V Take
heede to thy ſelf of the peſtiferous perſon, for he forgeth euils: leſt
perhaps he bring vpon thee deriſion for euer. \V Admitte a ſtraunger to
thee, and he shal ouerthrow thee in an hurlewind, & shal make thee an
aliene from thine owne.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
