%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 29
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-29


\project douay-rheims


%%% 1550
%%% o-1429
\startChapter[
  title={Chapter 29}
  ]

\Summary{Lend charitably, 3.~and reſtore faithfully. 10.~For the fault
  of il debters, omitte not to helpe the honeſt. 12.~Rather geue almes
  where nede is. 19.~Be thankful for ſuretiship. 28.~Liue
  frugally. 32.~Goe not a gheſtning for delicate chere.}


He that
\SNote{Lending is a worke of mercie, & a kinde of almes.}
lendeth to his neighbour, doth mercie: and he that preuayleth
with hand, keepeth the commandments. \V Lend to thy neighbour in the
time of his neceſſitie, and
%%% !!! Not marked (in the SNote) as a seperate note in both
\SNote{To repay that is borowed is a worke of iuſtice, wittingly not to
reſtore is as bad as theft.}
againe repay thy neighbour in his time. \V Confirme thy word, and doe
faithfully with him: & thou shalt finde at al time, that which is
neceſſarie for thee. \V Manie haue eſteemed a thing lent as a thing
found, and haue geuen moleſtation to them that did helpe them. \V Til
they receiue, they kiſſe the handes of the lender, and in promiſes they
humble their voice: \V and in the time of repaying they wil aske a time,
and wil ſpeake wordes of tediouſnes and murmurings, and wil make the
time an excuſe: \V and if he be able to pay, he wil reſiſt, he wil pay
ſcarſe halfe of the whole, and wil account it as a thing found: \V but
if not, he wil defraude him of his money, and poſſeſſe him an enemie
without cauſe: \V and wil repay him reproches and curſes, and for
honour and benefite wil repay him contumelie. \V Manie haue not lent,
not becauſe of wickednes, but they were afraid to be defrauded without
cauſe. \V But yet vpon the humble be ſtronger of minde, & for almes
differre him not. \V
\CNote{\XRef{Exo.~22.}
\XRef{Deut.~14.}}
Becauſe of the commandment receiue the poore: and becauſe of his
pouertie, ſend him not away emptie. \V Loſe money for thy brother and
thy freind: and hide it not vnder a ſtone vnto perdition. \V
\CNote{\XRef{Tob.~4.}}
Put thy
treaſure in the precepts of the Higheſt,
\SNote{Great fruite of workes of mercie.}
& it shal profite thee more then gold. \V Shut vp almes in the hart of
the poore, and the ſame shal obteyne for thee againſt al euil. \V Aboue
the shilde of the mightie, & aboue the ſpeare, it shal fight againſt
%%% 1551
thyne enemie. \V A good man becometh ſuretie for his neighbour: and he
that hath loſt shame, wil leaue him to himſelf. \V Forget not the
kindnes of a ſuretie: for he hath geuen his life for thee. \V The ſinner
and vncleane perſon fleeth from his ſuretie. \V A ſinner counteth the
goodes of his ſuretie to himſelf: and vnthankful in minde wil forſake him
that deliuered him. \V A man is ſuretie for his neighbour: and when he
hath loſt shame, he shal be forſaken of him. \V Naughtie ſuretieshippe
hath vndone manie, that were in good caſe, and hath toſſed them as a
waue of the ſea. \V Whurling round about, it hath made mightie men to
remoue, and they haue wandred in ſtrange nations. \V A ſinner that
tranſgreſſeth the commandment of our Lord, shal fal into naughtie
%%% o-1430
ſuretieshippe: and he that endeuoureth to doe manie thinges, shal fal
into iudgement. \V Recouer thy neighbour according to thy power, and
\SNote{Prudence requireth that by helping an other, thou doeſt not
ouerthrovv thyſelf.}
take heede to thy ſelf that thou fal not. \V The beginning of mans life
water & bread, and garment, and houſe couering his turpitude. \V
Better is the poore mans fayre vnder a roofe of bordes, then ſumptuous
cheere in a ſtrange place without a houſe. \V Let the leaſt thing pleaſe
thee in ſteede of a great, and thou shalt not heare the reproach of
peregrination. \V It is a naughtie life to change lodging from houſe to
houſe: and where he shal lodge, he shal not deale boldely, nor open his
mouth. \V He shal lodge, and feede, and make the vnthanckful drinke, and
beſide theſe thinges, he shal heare bitter wordes. \V Paſſe thou
ſtranger, & furnish the table, & with the thinges thou haſt in thy hand,
feede the reſt. \V Depart from the preſence of the honour of my
freindes: for the neceſſitie of my houſe my brother is to be lodged with
me. \V Theſe thinges be greuous to a man that hath vnderſtanding: rebuke
for the houſe, and the reproch of the lender.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
