%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 1511
%%% o-1394
\startChapter[
  title={Chapter 02}
  ]

\Summary{Whoſoeuer wil ſerue God muſt haue iuſtice, feare of God, and
  patience, 6.~with confidence in God. 14.~Diſſemblers, incredulous, and
  impatient shal be miſerable, 18.~but the godlie shal receiue more
  grace.}


Sonne, coming to the ſeruice of God, ſtand in iuſtice, and in feare, &
prepare thy ſoule to tentation. \V Repreſſe thy hart, & ſuſteyne:
incline thine eare, & receiue the wordes of vnderſtanding:
\SNote{Deſire not to dye, that thou maiſt therby be couered from
vvorldlie miſeries, for that is a deſperate deſire, but haue patience in
this life that thou maiſt finde reſt in God.}
and make no haſt in the time of
\TNote{Couering.}
obduction. \V Suſteyne the ſuſtentations of God: be ioyned to God, and
ſuſteyne, that thy life may increaſe in the later end. \V Al that shal
be applied to thee, receiue: and in ſorow ſuſteyne, and in thy
humiliation haue patience: \V for gold and ſiluer are tryed in the fyre,
but acceptable men in the fornace of humiliation. \V Beleue God, and he
wil recouer thee: and direct thy way, and hope in him. Keepe his feare,
and grow old therin. \V Ye that feare our Lord expect his mercie: &
decline not from him leſt ye fal. \V Ye that feare our Lord, beleue him,
and your reward shal not be voyde. \V Ye that feare our Lord, hope in
him: and mercie shal come to you for your delectation. \V Ye that feare
our Lord, loue him, & your hartes shal be illuminated. \V Children
behold the nations of men: and know ye that none hath hoped in our Lord,
and hath bene confounded. \V For who hath continewed in his commandment,
and hath bene forſaken? or who hath inuocated him, and he deſpiſed
him? \V Becauſe God is pitiful and merciful, and wil forgeue ſinnes in
the day of tribulation: and he is protector to al that ſeeke him in
truth. \V
\SNote{As vertues are to be revvarded: ſo al ſinnes shal be
punished. Namely external pretence of holines, vvith ſecrete euil
intention. Diſtruſt in God. Impatience, and the like.}
Woe to them of a duble hart, and to wicked lippes, and to the handes
that doe euil, and to the ſinner that goeth on the earth two wayes. \V
Woe to them that be diſſolute of hart, which beleue not God: and
therfore they shal not be protected of him. \V Woe be to them, that haue
loſt patience, and that haue forſaken the right wayes, and haue declined
into peruerſe wayes. \V And what wil they doe, when our Lord
%%% o-1395
shal begynne to looke on them? \V They that feare our Lord, wil not be
incredulous to his word: and they that loue him, wil kepe
%%% 1512
his waye. \V They that feare our Lord, wil ſeeke after the thinges that
are wel pleaſing to him: and they that loue him, shal be filled with his
law. \V They that feare our Lord, wil prepare their hartes, and in his
ſight wil ſanctifie their ſoules. \V They that feare our Lord, kepe his
commandements, and wil haue patience euen vntil his viſitation, \V
ſaying: If we doe not penance, we shal fal into the handes of our Lord,
and not into the handes of men. \V For according to his greatnes, ſo
alſo his mercie is with him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
