%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 42
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-42


\project douay-rheims


%%% 1569
%%% o-1446
\startChapter[
  title={Chapter 42}
  ]

\Summary{Further admonition to auoide ſinnes in wordes, and deedes,
  6.~with care that others offend not by our negligence. 15.~An other
  matter of meditation is Gods excellencie, appearing in his workes.}


Repeate not the word
\SNote{It is not lawful to reueale that which we iuſtly promiſe to
conceale.}
which thou haſt heard, neither reueale thou of a ſecret word, & thou
shalt in deede be without confuſion, and shalt finde grace in the ſight of
al men: be not ashamed for al theſe thinges, and accept not perſon
therby to ſinne. \V
\CNote{\XRef{Leuit.~19.}
\XRef{Deu.~1.~16.}
\XRef{Prou.~24.}
\XRef{Iacob.~2.}}
Of the law of the Higheſt, and his teſtament, and of iudgement to
iuſtifie the impious, \V of the word of companions and wayfaring men,
and of the geuing of the inheritance of freindes, \V of the equalitie of
balance and weightes, of the getting of manie thinges and few, \V of the
corruption of bying, and of marchantes, and of much diſcipline of thy
children, and to make bloudie the ſide of a wicked ſeruant. \V Ouer a
naughtie woman a ſeale is good. \V Where there are manie handes, shut
vp, and whatſoeuer thou shalt deliuer, number, and weigh it: and write
euerie thing geuen and receiued. \V Of the diſcipline of the vnwiſe and
foolish, and of ancientes, that are iudged of young men: and thou shalt
be wel inſtructed in al thinges, and approued in the ſight of al the
liuing. \V A daughter is the ſecret watch of the father, and the care of
her taketh away ſleepe, leſt perhaps in her youth she become paſt age, &
abiding with an husband she become odious: \V leſt at anie time she be
corrupted in her virginitie, and in her fathers houſe she be found with
childe: leſt perhaps abyding with her husband she tranſgreſſe, or at the
leaſt become barren. \V Ouer a diſſolute daughter keepe ſure watch: leſt
at anie time she make thee come into
\Fix{reprooche}{reproch}{likely typo, fixed in other}
with thine enemies, becauſe of detraction in the citie, and the
obiection of the people, and she confound thee in the multitude of the
people. \V Looke not on euerie bodie
%%% 1570
for beautie ſake: & among wemen tarie not. \V For out of garments cometh
forth the moth, and from a woman the iniquitie of a man. \V For
\SNote{There is leſſe danger in conuerſing familiarly with a wicked man,
then with a freindlie woman. In which conuerſation, much prudence is
required, as is before admoniſhed.
\XRef{chap.~9.}}
better is the iniquitie of a man, then a woman doing a good turne, and a
woman shaming vnto reproche. \V I therefore wil be mindeful of the workes
of our Lord, and I wil shew forth which I haue ſeene. By the words of
our Lord are his workes. \V The ſunne illuminating hath looked through out al,
and ful of the glorie of our Lord is his worke. \V Hath not our Lord
made the ſainctes to declare al his meruelous thinges, which our Lord
the omnipotent confirmed to be eſtablished in his glorie? \V He hath
ſearched out the depth, and the hart of men: and in their ſubtilitie he
hath conſidered. \V For our Lord hath knowen al knowlege, and hath
beheld the ſigne of age,
%%% o-1447
declaring what thinges are paſt, and what are to come, reueling
\SNote{The excellencie of God, which can not be ſene with mortal eye
\XRef{(Exo.~33.)}
is propoſed to our meditation in his workes. The like in
\XRef{Iob.~38.}}
the tokens of ſecret thinges. \V No cogitation eſcapeth him, and no word
hideth it ſelf from him. \V The glorious workes of his wiſdom he hath
beautified: who is before the world and world without end, neither is
there added, \V nor
\Fix{diminiſed,}{diminished,}{obvious typo, fixed in other}
and he needeth not anie mans counſel. \V How are al his workes to be
deſired, and which is as it were a ſparke to conſider! \V Al theſe liue,
and remaine for euer, and in al neceſſitie al thinges obey him. \V Al
thinges duble, one againſt one, and he hath made nothing to want. \V He
hath confirmed the good thinges of euerie one. And who shal be filled
ſeeing his glorie?

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
