%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 24
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-24


\project douay-rheims


%%% 1542
%%% o-1422
\startChapter[
  title={Chapter 24}
  ]

\Summary{True and laudable wiſdom 5.~proceding from God, 6.~shineth in
  his workes: 12.~eſpecially in his Church; where she bringeth forth al
  vertues. 26.~She inuiteth al vnto her, 44.~and lighteneth her folowers
  with ſplendore of doctrine.}


VViſdom shal praiſe
\SNote{Diuine wiſdom the Second Perſon of the B.~Trinitie begotten not
created praiſeth it ſelf. According to the phraſe of ſpeach
\XRef{Iere.~51. v.~14.}
\Emph{The Lord of hoſtes hath ſvvorne by his ſoule}, that is, by
himſelf.}
her ſoule, & shal be honoured in God, and shal glorie in the middes of
her people, \V and shal open her mouth in the churches of the Higheſt,
and shal glorie in the ſight of his power, \V and in the middes of her
people she shal be exalted, and in the holie aſſemblie she shal be
admired, \V and in the multitude of the elect she shal haue praiſe, and
among the bleſſed, she shal be bleſſed, ſaying: \V I come forth from the
mouth of the Higheſt, the firſt begotten before al creatures. \V I made
that in the heauens there should riſe light that faileth not, and as a
cloud I couered al the earth. \V I dwelt in the higheſt places, and my
throne is in the pillar of a cloude. \V I alone haue gone round about
the compaſſe of heauen, and haue penetrated into the bottome of the
depth, and haue walked in the waues of the ſea, \V and ſtood in al the
earth: and in al people, \V and in euerie nation I haue had the
primacie: \V and I haue by ſtrength troden
%%% 1543
downe the hartes of al the excellent, and the baſe; and in al theſe
thinges
\SNote{God offereth his grace, but forceth not anie to accept it.}
I ſought reſt, & I shal abide in the inheritance of our Lord. \V Then
the creatour of al commanded, and ſaid to me: & he that
\SNote{Creation is not here taken in the ſtrict ſignification, but for
diuine production, in that God the Father by vnderſtanding begetteth God
the Sonne. As likewiſe the Father and the Sonne by loue produce the
Holie Ghoſt.}
created me, reſted in my tabernacle, \V and he ſaid to me: Inhabite
\SNote{In the Church only is effectual grace.}
in Iacob, and inherite in Iſrael, and take roote in myne elect. \V From
the beginning and before the worlds was I created, and vnto the world to
come I shal not ceaſe, and in the holie habitation I haue miniſtred
before him. \V And ſo in Sion was I eſtablished, and in the ſanctified
cittie likewiſe I reſted, and my power was in Ieruſalem. \V And I tooke
roote in an honorable people, and in the portion of my God his
inheritance, and my abiding is in the ful aſſemblie of ſaintes. \V I am
exalted as a cedar in Libanus, and as a cypres tree in mount Sion. \V As
a palme tree in Cades am I exalted, and as a roſe plant in Iericho. \V
As a faire oliue tree in the fieldes, and as a plane tree by the water
in the ſtreates am I exalted. \V I gaue an odour as cinnamon, &
aromatical balme: as choſen myrrhe haue I geuen the ſweetenes of
odour: \V and as ſtorax, and galbanum, and onyx, and aloes, and as
Libanus not cut, haue I perfumed myne habitation, and myne odour is as
baulme non mingled. \V I haue ſpred out my boughes as the terebinth, and
my boughes are of honour and grace. \V I as a vine haue fructified
ſweetenes of
%%% o-1423
odour: and my flowers are fruite of honour and honeſtie. \V I am the
mother of beautiful loue, and of feare, and of knowlege, and of holie
hope. \V In me is al grace of way and truth, in me al hope of life and
vertue. \V Paſſe to me al ye that deſire me, and be filled of my
generations. \V
\CNote{Pſa.~18.}
For my ſpirit is ſweete aboue honie, and myne inheritance aboue honie and
the honie combe. \V My memorie is vnto generations of worldes. \V They
that
\SNote{The more grace any hath the more he deſireth and receiueth.}
eate me, shal yet hunger: and they that drinke me, shal yet thirſt. \V
He that heareth me, shal not be confounded: and they that worke in me,
shal not ſinne. \V They that explicate me, shal haue life
euerlaſting. \V Al theſe thinges are the booke of life, and the
teſtament of the Higheſt, & the knowlege of truth. \V Moyſes commanded a
law in the preceptes of iuſtices, and an inheritance to the houſe of
Iacob, and the promiſes to Iſrael. \V He appointed to Dauid his ſeruant
for to raiſe vp a king of him moſt ſtrong, and ſitting in the throne of
honour
\SNote{In Dauids progenie God preſerued the kinglie ſtate til the
captiuitie: and the eſtimation of the royal bloud vnto Chriſt. But in al
this Dauid was a figure of Chriſt. VVho ſitteth in the Throne of honour
abſolutly for euer.}
for euer. \V Who filleth wiſdom as Phiſon, and as Tigris in the daies of
new fruites. \V Who
%%% 1544
replenisheth vnderſtanding as Euphrates, who multiplieth it as Iordan in
the time of harueſt. \V
\CNote{\XRef{Ioſu.~3.}}
Who ſendeth diſcipline as the light, and aſſiſting as Gehon in the day
of vintage. \V Who firſt hath perfect knowledge of it, & a weaker shal
not ſearche it out. \V For her cogitation shal abound aboue the ſea, and
her counſels aboue the greate depth. \V I wiſdom haue powred out
riuers. \V I as a ſluſe of a mightie water out of the riuer,
I as the riuer Dioryx, & as a water coundite I came out of paradiſe. \V
I ſaid: I wil water my garden of plantes, and wil inebriate the fruite
of my medow. \V And behold my ſluſe was made aboundant, and my riuer
came neere to a ſea. \V Becauſe I illuminated doctrine to al as the
morning light, & I wil declare it far. \V I wil penetrate al the
inferiour partes of the earth, and wil behold al that ſleepe, and wil
illuminate al that hope in our Lord. \V I wil yet powre out doctrine as
prophecie, and wil leaue it to them that ſeeke wiſdom, and wil not ceaſe
vnto their progenies euen to the holie age. \V See ye that I haue not
laboured for myſelf only, but for al that ſeeke out the truth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
