%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Chapter 31
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-31


\project douay-rheims


%%% 1553
%%% o-1431
\startChapter[
  title={Chapter 31}
  ]

\Summary{By ſeeking vertue, and laboring for neceſſaries, the flesh is
  ſubdued to the ſpirite. 8.~Moderate riches are beſt, 12.~with
  temperance in diette, 30.~eſpecially in drinking.}


VVatching
\SNote{They that imploy al their ſtudie to gette vertues, ſhal be more
free from tentations of the fleſh.}
after honeſtie shal pine the flesh, & the thought thereof
\SNote{And from drawſines of mind: whereupon S.~Ierom admoniſheth:
\Emph{Loue the ſtudies of holie Scriptures, and thou vvilt not loue the
vices of the flesh.}
\Cite{Epiſt. as Ruſticum Monach.}}
taketh away ſleepe. \V The thought of foreknowlege turneth away the
vnderſtanding, & greuous infirmitie maketh a ſober ſoule. \V The riche
man hath laboured in gathering of ſubſtance together, & in his reſt he
shal be replenished with his goodes. \V The poore man hath laboured in
the diminishing of his liuing, and in the end he is made poore. \V He
that loueth gold shal not be iuſtified: & he that foloweth after
corruption, shal be replenished of it. \V Manie haue bene geuen into
falles for gold, and their perdition hath come by the beautie
thereof. \V The gold of them that ſacrifice is a wood of offence: wo to
them, that folow after it, and euerie vnwiſe man shal perish in it. \V
Bleſſed is the rich man that is found without ſpot: and that hath not
gone after gold, nor hoped in money and treaſures. \V Who is this, & we
wil praiſe him, for he hath done meruelous thinges in his life. \V Who
is proued therin, & perfect, shal haue eternal glorie. He that
\SNote{It is in mans freewil to tranſgreſſe, & therfore they are happie,
that through Gods grace do not breake his commandments.}
could tranſgreſſe, and hath not tranſgreſſed: and doe euils, and hath
not done: \V Therfore are his good thinges
\Fix{ſtablished}{eſtablished}{likely typo, fixed in other}
in our Lord, & al the church of ſaintes shal
%%% o-1432
declare his almes. \V Art thou ſet at a great table? open not thy iawe
therevpon firſt. \V Say not this: There be manie thinges which are vpon
it. \V Remember that a naughtie eie is euil. \V What is created worſe
then the eie? therefore shal it weepe at euerie face. When it shal
ſee, \V ſtretch not out thy hand firſt, and ſo contaminated with enuie
thou be ashamed. \V Be not oppreſſed in a feaſt. \V Vnderſtand by
thyſelf what thy neighbours thinges are. \V Vſe as a frugal man thoſe
thinges, that are ſet before thee: leſt thou be hated when thou eateſt
much. \V
\CNote{\XRef{Prou.~23.}}
Leaue of firſt, for maners ſake, and exceede not, leſt thou perhaps
offend. \V And if thou be ſet in the middes of manie, ſtretch not forth
thy hand before them: neither doe thou firſt aske to drinke. \V How
ſufficient is a little wine for a man wel taught, and in ſleeping thou
shalt not be pained with it, and thou shalt feele no griefe. \V
Watching, & choler, & torment to an vnſatiable man: \V ſleepe of health
is in a man of ſpare diet:
%%% 1554
he shal ſleepe vntil morning, and his ſoule with him shal be
delighted. \V And if thou haſt bene forced with eating much, riſe from
the middes, and vomite, and it shal refresh thee, and thou shalt not
bring infirmitie to thy bodie. \V Heare me my ſonne, and deſpiſe me not:
and in the end thou shalt finde my wordes. \V In al thy workes be
quicke, and al infirmitie shal not chance vnto thee. \V
\CNote{\XRef{Prou.~22.}}
The lippes of manie shal bleſſe him that is magnifical in breads, and
the teſtimonie of his truth is faithful. \V In
\SNote{As bread is the chiefe ſuſtenãce of the bodie, ſo doctrine is of
the minde, which being good nouriſheth, & if it be bad corrupteth them
that receiue it.}
naughtie bread the cittie wil murmur, and the teſtimonie of the
naughtines thereof is true. \V Prouoke not them that loue wine: for wine
hath deſtroyed very manie. \V Fire tryeth hard yron: ſo wine dronken in
drunkennes shal rebuke the hartes of the proud. \V Equal life to al men,
wine in ſobrietie: if thou drinke it moderatly, thou shalt be ſober. \V
What is the life that is diminished with wine? \V What defraudeth life?
death. \V Wine was created for ioyfulnes, and not for drunkẽnes from the
beginning.\CNote{\XRef{Pſa.~103.}
\XRef{Prou.~31.}}
Wine drunken moderately is the ioy of the ſoule and bodie. 
\V Sober drinking is health to ſoule and bodie. \V Much wine
drunken maketh prouocation, & wrath, & manie ruines. \V Much wine
drunken is bitternes of the ſoule. \V The couragiouſnes of drunkennes,
is offence of the vnwiſe, leſſening the ſtrength, and making woundes. \V
In a banquet of wine rebuke not thy neighbour: and deſpiſe him not in
his mirth. \V Speake not to him wordes of
\Fix{repoch:}{reproch:}{obvious typo, fixed in other}
and preſſe him not in demanding againe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
