%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ecclesiasticus
%%%% Prologue
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent prologue


\project douay-rheims


%%% 1508
%%% o-1392
\startArgument[
  title={\Sc{The Prologve vpon Ecclesiasticvs of Ieſvs the Sonne of
  Sirach.}},
  marking={Prologue of Ecclesiasticus.}
  ]

The knowlege of manie and great thinges hath bene shewed vs by the Law,
and the Prophetes, and others that folowed them: in which we ought
\SNote{By reaſon of a more perfect law the people of Iſrael were more
renowmed then anie other nation in the world.
\XRef{Deut.~4.}}
to prayſe Iſrael for doctrine & wiſdom: becauſe not onlie they in
ſpeaking muſt nedes be cunning, but ſtrangers alſo both lerning &
writing, may become moſt lerned. My grandfather Ieſus, after he gaue
himſelfe more amply to the diligence of reading the Law, and the
Prophetes, and other Bookes, that were deliuered vs from our fathers:
himſelf alſo would write ſome of thoſe thinges, which perteyne to
doctrine and wiſdom: that ſuch as are deſirous to lerne, and to be made
\Fix{counning}{cunning}{likely typo, fixed in other}
in the ſame thinges, may more and more be attent in minde, and be
confirmed to the life that is according to the law. I exhort you
therfore to come with beneuolence, and to read with attent ſtudie, and
to pardon vs for thoſe thinges, wherein we ſeming to folow the image of
wiſdom, may
%%% 1509
fayle in the compoſition of wordes: for the Hebrew wordes alſo
\SNote{Tranſlations into other languages hardly expreſſe the ſenſe of
the original tongue.}
fayle when they shal be tranſlated to an other tongue. And not onlie
theſe, but the Law alſo itſelf, and the Prophetes, and the reſt of other
bookes, haue no ſmal difference, when they are ſpoken within
themſelues. For in the eight and thirteth yeare, in the time of Ptolomee
Euergetes the king, after I came into Ægypt: and when I had bene there
much time, I found there bookes leaft, of no ſmal nor contemptible
doctrine. Therfore myſelf alſo thought it good, and neceſſarie
\Fix{ta}{to}{obvious typo, fixed in other}
adde ſome diligence and labour to interprete this booke: and with much
watching I brought forth this doctrin in ſpace of time, that men may
lerne thoſe thinges which teach them, that wil applie their minde, how
they ought to order their maners, them that purpoſe to lead their life
according to the Law of our Lord.


\stopArgument


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
