%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Three John
%%%% Chapter
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.




\startcomponent chapter


\project douay-rheims


%%% 2929
%%% o-2790
\startChapter[
  title={}
  ]

\Summary{He commendeth Gaius, for continuing in the truth, & for
  ſuſteining or ſuccouring true Preachers, 9.~noting Diotrepes for the
  contrarie, and praiſing Demetrius.}

The Seniour to Gaius the Deareſt, whom I loue in truth.

\V My Deereſt, concerning al things I make my praier that thou proceed
proſperouſly, and fare wel, as thy ſoule doth proſperouſly. \V I was
exceeding glad when the Brethren came, and gaue 
teſtimonie to thy truth, euen as thou walkeſt in truth. \V Greater
\TNote{\G{χαράν χάριν}}
\Var{thãke}{pleaſure}
haue I not of thẽ, that then I may heare my childrẽ doe walke in
truth. \V My Deareſt, thou doeſt faithfully whatſoeuer thou workeſt on
the Brethren,
\SNote{A great grace to be beneficial to ſtrangers, ſpecially to them
that be of our Catholike faith and ſuffer for the ſame.}
& that vpon ſtrangers. \V They haue rendred teſtimonie to thy charitie
in the ſight of the Church: whom, thou ſhalt doe wel, bringing on their
way in manner worthie of God. \V For, for his name did they depart,
taking nothing
%%% o-2791
of the Gentils. \V We therfore ought to receiue ſuch: that we may be
coadiutours of the truth.

\V I had written perhaps to the Church: but he that loueth to beare
\SNote{It ſeemeth (ſaith
%%% !!! Cite?
S.~Bede) he was an Arch-heretike or proud Sect-maiſter.}
primacie among them Diotrepes, doth not receiue vs. \V For this cauſe,
if I come,
\SNote{That is, \Emph{I wil rebuke them and make them knowen to be
wicked.}
%%% !!! Cite?
Bede.}
\TNote{\L{commonebo}, \G{ὑπομνήσω}}
I wil aduertiſe his workes which he doeth: with malicious words chatting
againſt vs. And as though theſe things ſuffice him not, neither himſelf
doth receiue the Brethren, and them that doe receiue, he prohibiteth,
and caſteth out of the Church. \V My Deareſt, doe not imitate euil, but
that which is good. He that doeth wel, is of God: he that doeth il hath
not ſeen God. \V To Demetrius teſtimonie is giuen of al, and of the
truth it-ſelf, yea and we giue teſtimonie: and thou knoweſt that our
teſtimonie is true.

\V I had many things to write vnto thee: but I would not by inke and
penne write to thee. \V But I hope forth-with to ſee thee, and we wil
ſpeake mouth to mouth. Peace be to thee. The freinds ſalute thee. Salute
the freinds by name.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

