%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Two Thessalonians
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 2792
%%% o-2653
\startChapter[
  title={Chapter 1}
  ]

\Summary{He thanketh God for their increaſe in faith and charitie, and
  conſtancie in perſecution (aſſuring them that they merit thereby the
  Kingdom of God, as their perſecutours doe damnation:) 11.~and alſo
  praieth for their accomplishment.}

Paul and Syluanus and Timothee, to the Church of the Theſſalonians in
God our Father and our Lord \Sc{Iesvs} Chriſt.

\V Grace to you and peace from God our Father and our Lord \Sc{Iesvs}
Chriſt.

\V We ought to giue thankes alwaies to God for you, Brethren, ſo as meet
is, becauſe your faith increaſeth exceedingly, and the charitie of euery
one of you aboundeth towards each other: \V ſo that we our ſelues alſo
glorie in you in the Churches of God, for your patience, and faith in al
your perſecutions and tribulations, which you ſuſtaine \V for an example
of the iuſt iudgement of God, that
\SNote{Note that by conſtant and patient ſuffering of affliction for
Chriſt men are made worthie (ſo the Greek ſignifieth, as the Aduerſaries
themſelues tranſlate
\XRef{v.~11.)}
of the crowne or Kingdom of Heauen: and ſo doe merit and deſerue the
ſame. See
\XRef{Anno. luc.~20,~35.}
And the Apoſtle here ſaith that it is God's iuſtice no leſſe to repay
glorie to the afflicted, then to rẽder punishmẽt to them that afflict,
becauſe of their contrarie deſerts or merits.}
you may be
\TNote{\G{καταξιωθῆναι ὑμᾶς}}
counted worthie of the Kingdom of God, for the which alſo you ſuffer. \V
If yet it be iuſt with God to repay tribulation, to them that vexe
you: \V and to you that are vexed, reſt with vs in the
\CNote{\XRef{1.~Theſs.~4,~16.}}
reuelation of our
Lord \Sc{Iesvs} from Heauen with the Angels of his power, \V in flame of
fire, giuing reuenge to them that know not God, and that obey not the
Ghoſpel of our Lord \Sc{Iesvs} Chriſt. \V Who ſhal ſuffer eternal paines
in deſtruction, from the face of our Lord and from the glorie of his
power: \V when he ſhal come to be
\SNote{Chriſt shal be glorified in his Saints, that is, by the great and
vnſpeakable honour and exaltation of them he shal be honoured, as now
he is: the honour which the Church doth to them, not diminiſhing
Chriſt's glorie (as the Aduerſaries foolishly pretend) but exceedingly
augmenting the ſame.}
glorified in his Saints, and to be made maruelous in al them that haue
beleeued, becauſe our teſtimonie
%%% o-2654
concerning you was credited in that day. \V Wherin alſo we pray alwayes
for you, that our God
\TNote{\G{ἀξιώσῃ}}
make you worthie of his vocation, and accompliſh al the good pleaſure of
his goodneſſe & the worke of faith in power, \V that the name of our Lord
\Sc{Iesvs} Chriſt may be glorified in you, and you in him, according to
the grace of our God, and of our Lord \Sc{Iesvs} Chriſt.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

