%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Tobias
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 1011
%%% o-0908
\startChapter[
  title={Chapter 1}
  ]

\Summary{Tobias
\MNote{The firſt part.

Tobias his holie maner of life.}
of the tribe of Nephthali, neuer communicateth in
\Fix{Ierobams}{Ieroboams}{likely typo, fixed in other}
ſciſme, 9.~teacheth his ſonne to feare God, and flee ſinne. 11.~Being in
captiuitie eateth not forbidden meates, as others do. 13.~Amongſt other
workes of mercie, he lendeth tenne talentes of ſiluer to Gabelus. 21.~Is
perſecuted and ſpoyled. 24.~Shortly the king being ſlaine, he recouereth
libertie and his goodes.}

Tobias of the tribe, and citie of Nepthali (which is in the vpper partes
of Galilee aboue Naaſſon, beyond the way, that leadeth to the
\Fix{weaſt,}{weſt,}{likely typo, fixed in other}
hauing on the right hand the citie Sephet) \V when he was captiue in the
daies of Salmanaſar the king of the Aſſyrians, yet being in captiuitie,
he forſooke not the way of truth, \V ſo that he imparted al things that
he could make, dayly to his brethren captiues with him, which were of
his kinred. \V And whereas he was yonger then al the tribe of Nephthali,
yet he did no childiſh thing in his worke. \V Finally when
\SNote{Not abſolutly al, but very manie: for ſome of the ſame tribe and
kinred did alſo feare God.
\XRef{c.~2. v.~2.}}
al went to the golden calues, which Ieroboam the king of Iſrael had
made, he alone fled the companies of al, \V and went into Ieruſalem to
the temple of our Lord, and there adored our Lord God of Iſrael,
offering faithfully al his firſt fruites, and his tithes, \V ſo that in
the third yeare he miniſtred al the tithing to the proſelytes, and
ſtrangers. \V Theſe thinges and the like to theſe did he obſerue being a
childe according to the law of God. \V But when he was a man, he tooke
to wife Anna of his owne tribe, and he begat a ſonne of her, geuing him
his owne name, \V whom from his infancie he taught to feare God, and to
abſtayne from al ſinne. \V Therfore when by the captiuitie he was
%%% o-0909
come with his wife and ſonne into the citie of Niniue, with al his
tribe, \V (when al did eate of the meates of the Gentiles) he kept his
ſoule, and neuer was contaminated in their meates. \V And becauſe he was
mindful of our Lord in al his hart, God gaue him grace in the ſight of
Salmanaſar the king, \V and he gaue him leaue to goe whitherſoeuer he
would, hauing libertie to doe what thinges ſoeuer he would. \V He
therfore went to al that were in the captiuitie, and gaue them holeſome
admonitions. \V And when he was come into Rages a citie of the Medes,
and had ten talentes of ſiluer of theſe, wherwith he had beene honoured
of the king: \V and
%%% 1012
when in a great multitude of his kinred, he ſaw Gabelus ſtand in nede,
who was of his tribe, vnder a bil of his hand he gaue him the ſayd
weight of ſiluer. \V But after much time, Salmanaſar the king being
dead, when Sennacherib his ſonne reigned for him, and eſtemed the
children of Iſrael odious in his ſight: \V Tobias dayly went through al
his kinred, and conforted them, and diuided to euerie one, as he was
able, of his goods: \V the hungrie he nouriſhed, and to the naked he
gaue clothes, and the dead, and them that were ſlayne, he buryed
carefully. \V Finally when king Sennacherib was returned fleing from
Iewrie the
\Fix{ſhaughter,}{ſlaughter,}{obvious typo, fixed in other}
that God had made about him for his blaſphemie, and being angrie ſlewe
manie of the children of Iſrael, Tobias buryed their bodies. \V But when
it was told the king, he commanded him to be ſlayne, and tooke al his
ſubſtance. \V But Tobias fleing with his ſonne and with his wife,
nakedly lay hid, becauſe manie loued him. \V But after fourtie fiue
dayes the king was ſlayne of his owne ſonnes, \V and Tobias returned
into his houſe, and al his ſubſtance was reſtored to him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
