%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Tobias
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 1020
%%% o-0917
\startChapter[
  title={Chapter 8}
  ]

\Summary{Tobias bruling part of the fishes liuer, Raphael bindeth the
  diuel. 4.~Tobias and Sara pray. 11.~Raguel fearing that Tobias is
  dead, maketh a graue for him, but vnderſtanding that he is wel,
  filleth it vp againe, 21.~prepareth a feaſt, geueth the half of his
  goodes preſently for Saraes dawrie, the other halfe after her parents
  death.}

And after they had ſupped, they brought in the yong man to her. \V
Tobias therfore remembring the Angels word, brought forth out of his
bag, part of
\SNote{See
\XRef{chap.~6. v.~8.}
Myſtically it ſignified Chriſts paſſion whereby the diuel was expelled
out of mens hartes. 
\Cite{S.~Aug. ſer.~28. de Sanctis. Proſper.}
\Cite{li. de promiſſ. p.~2. c.~39.}}
the liuer, and layd it vpon liue coales. \V Then Raphael the Angel tooke
the diuel, and bound him in the deſert of higher Ægypt. \V Then Tobias
exhorted the virgin, & ſayd to her: Sara ariſe, and let vs pray to God
to day, and to morow, and the next morow: becauſe theſe three nights we
are ioyned to God: and when the third night is paſt, we wil be in our
wedlocke. \V For we are the children of holie men, & we may not be
ioyned together as gentiles, that know not God. \V And they riſing
together, prayed both together that health might be geuen them. \V And
Tobias ſayd: Lord God of our fathers, the heauens & the earth, and the
ſea & fountaynes, and riuers, and al thy creatures that are in them,
bleſſe thee. \V Thou madeſt Adam of the ſlime of the earth, & gaueſt him
Eue an helper. \V And now Lord thou knoweſt, that not for fleshlie luſt
doe I take my ſiſter to wife, but only for the loue of poſteritie, in
the which thy name may be bleſſed for euer & euer. \V Sara alſo ſayd:
Haue mercie on vs Lord, haue mercie vpon vs and let vs grow old both
together in health. \V And it came to paſſe about the cock crowing,
Raguel bad his ſeruantes to be called for, & they went with him together to
digge a graue. \V For he ſayd: Leſt perhaps it may
%%% 1021
chance to him, as alſo to the other ſeuen huſbandes, that went in vnto
her. \V And when they had prepared the pitte, Raguel returning to his
wife, ſayd to her: \V Send one of thy handmaydes, and let her ſee if he
be dead, that I may burie him before it be day. \V But ſhe ſent one of
her handmaydes, who going into the chamber, found them ſafe and ſound,
ſleeping both together. \V And returning ſhe brought good tydings: and
they bleſſed our Lord, to witte, Raguel & Anna his wife, \V and ſayd: We
bleſſe thee Lord God of Iſrael, becauſe it hath not chanced as we
thought. \V For thou haſt done thy mercie with vs, & haſt excluded from
vs the enemie, that perſecuted vs. \V And thou haſt taken pitie vpon two
the
\SNote{In the one familie there were no more children but one ſonne, in
the other one onlie daughter.}
only children. Make them Lord bleſſe thee more fully: and to offer vp to
thee a ſacrifice of thy prayſe, and of their health, that al nations may
know, that thou art God onlie in al the earth. \V And forthwith Raguel
commanded his ſeruantes, that
%%% o-0918
they ſhould fil vp the pitte, which they
had made, before it were day. \V And he bad his wife make readie a
feaſt, and prepare al thinges, that for victuals were neceſſarie to them
that goe a iourney. \V He cauſed alſo two fatte kyne, and foure wethers
to be killed, and great chere to be prepared for al his neighbours, and
al his freindes. \V And Raguel adiured Tobias, that he should abide with
him two weekes. \V And of al thinges which Raguel poſſeſſed, he gaue the
halfe part to Tobias, and made this writing, that the halfe part, which
was remayning after their deceaſe, should come to the dominion of
Tobias.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
