%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Tobias
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 1029
%%% o-0925
\startChapter[
  title={Chapter 14}
  ]

\Summary{Old Tobias dieth at the age of an hundred and two yeares,
  5.~exhorteth his ſonne and nephewes to pietie, forshewing that Niniue
  shal be deſtroyed, and Ieruſalem reedified. 14.~Yonger Tobias
  returneth with his familie to Raguel, and dieth happely as he had
  liued.}

And the wordes of Tobias were ended. And after that Tobias was reſtored
to his ſight, he liued two and fourtie yeares, and ſaw the children of
his nephewes. \V Therfore an hundred and two yeares being accompliſhed,
he was buried honorably in Niniue. \V For being ſix and fiftie
yeares old he loſt the ſight of his eies, and being threeſcore he
receiued it agayne. \V And the reſt of his life was in ioy, and with
great increaſe of the feare of God he went forward in peace. \V And at
the houre of his death he called vnto him Tobias his ſonne, and his
ſeuen yong ſonnes, his nephewes, and ſayd to them: \V The deſtruction of
Niniue is neere: for the word of our Lord fayleth not: and our brethren,
which are diſperſed from the land
%%% o-0926
of Iſrael, ſhal returne to it. \V And al the deſert land therof ſhal be
repleniſhed, and the houſe of God which is burnt in it, shal agayne be
reedefied: and thither shal al returne that feare God, \V and the
Gentiles shal forſake their idols, and shal come into Ieruſalem, and
shal inhabite in it, \V and
\SNote{This can not be vnderſtood of the citie of Ieruſalem but of the
Church of Chriſt.}
al the kings of the earth shal reioyce in it, adoring the king of
Iſrael. \V Heare ye therfore my children your father: ſerue our Lord in
truth, and
\SNote{Gods ſeruantes muſt not only expect what ſhal be commanded, but
alſo ſeeke to know what they ought to do.}
ſeeke to doe the thinges that pleaſe him: \V and command your children
that they doe iuſtices and almes deedes, that they be mindeful of God,
and bleſſe him at al time in truth, and in al their power.
%%% 1030
\V Now therfore children heare me, and doe not tarie here: but what day
ſoeuer you shal burie your mother by me in one ſepulchre, from
thenceforth directe your ſteppes to depart hence: \V for I ſee that the
iniquitie therof wil giue it an end. \V And it came to paſſe after the
death of his mother, Tobias departed out of Niniue with his wife, and
children, and childrens children, and returned to his father and mother
in law. \V And he found them in health in good old age: and he tooke
care of them, and he cloſed their eies: and al the inheritance of
Raguels houſe he receiued: & he ſaw the fifth generation, his childrens
children. \V And nintie nine yeares being accomplished in the feare of
our Lord, with ioy they buried him. \V And
\SNote{A notable cõmendation of Tobias familie and poſteritie. And a
plaine teſtimonie that true Religion neuer failed wholy in the tenne
tribes, much leſſe in the kingdom of Iuda, leaſt of al in the Church of
Chriſt, & new Teſtament.}
al his kinred, and al his generation continewed in good life, and in
holie conuerſation, ſo that they were acceptable both to God, and to
men, and to al the inhabitantes
\Fix{in in}{in}{obvious typo, fixed in other}
the land.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
