%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Tobias
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 1016
%%% o-0913
\startChapter[
  title={Chapter 5}
  ]

\Summary{Yong
\MNote{The ſecond part.

The iourney and affayres of yong Tobias aſſiſted by the Angel Raphael.}
Tobias ſeeking a guide for his iourney, Raphael the Angel in shape of a
man preſenteth himſelf, and vndertaketh this office. 23.~The mother
lamenteth the abſcence, and danger of her ſonne.}

Then Tobias anſwered his father, and ſayd: I wil doe al things, father,
whatſoeuer thou haſt commanded me. \V But how I shal require this money,
I can not tel, he knoweth not me, and I know not him: what token shal I
geue him? Yea neither the way which leadeth thither, did I euer know. \V
Then his father anſwered him, and ſayd: I haue the bil of his hand with
me, which when thou shalt shew him he wil forthwith reſtore it. \V But
goe now, and ſeeke thee out ſome faythful man, that may goe with thee being
ſure of his hyre: that thou mayſt receiue it, whiles I yet liue. \V Then
Tobias going forth, found a
\SNote{The Angel Raphael appearing in forme of a man prefigured our
Sauiour who indede became a verie man.
\Cite{S.~Beda.}}
goodlie yongman, ſtanding girded, and as it were readie to walke. \V And
not knowing that it was an Angel of God, he ſaluted him, and ſayd: From
whence haue we thee, good yongman? \V But he anſwered: Of the children
of Iſrael. And Tobias ſayd to him: Knoweſt thou the way, that leadeth
vnto the countrie of the Medes? \V To whom he anſwered: I know it: and
al the wayes therof I haue often walked, and I haue taried with Gabelus
our brother, who abydeth in Rages a citie of the Medes, which is ſituate
in the Mount Ecbatanis. \V To whom Tobias ſayd: Stay for me I beſeech
thee, til I tel theſe ſame things to my father. \V Then Tobias going in
told al theſe things to his father. Whereupon his father marueiling,
requeſted that he would come in vnto him. \V Going in therfore he
ſaluted him, and ſayd: Ioy be to thee alwaies. \V And Tobias ſayd: What
%%% 1017
maner of ioy ſhal be to me, which ſitte in darkenes, and ſee not the
light of heauen? \V To whom the yong man ſayd: Be of good cheere, it is
very neere that thou mayſt be cured of God. \V Tobias therfore ſayd to
him: Canſt thou bring my ſonne to Gabelus into Rages a citie of the
Medes? and when thou ſhalt returne, I wil pay thee thy hyre. \V And the
Angel ſayd to him: I wil
%%% o-0914
conduct him, & bring him to thee againe. \V To whom Tobias anſwered: I
pray thee, tel me, of what houſe, or what tribe art thou? \V To whom
\SNote{\Emph{Raphael} ſignifying \Emph{medicine of God}.
\Cite{S.~Greg. ho.~34.}
calleth him ſelfe Azarias, whoſe ſhape and viſage he tooke vpon him,
which name alſo ſignifieth \Emph{the helpe of God}.}
Raphael the Angel ſayd: Seekeſt thou the kinred of an hyred ſeruant, or
an hyred ſeruant himſelfe, that may goe with thy ſonne? \V But leſt
perhaps I make thee careful, I am Azarias the ſonne of Ananias the
great. \V And Tobias anſwered: Thou art of a great kinred. But I pray
\Fix{the}{thee}{obvious typo, fixed in other}
be not angrie that I would know thy kinred. \V And the Angel ſayd to
him: I wil lead thy ſonne ſafe, and bring him to thee agayne ſafe. \V
And Tobias anſwering, ſayd: Wel may you walke, and God be in your
iourney, and his Angel accompanie you. \V Then al things being readie,
that were to be caried in the way, Tobias bid his father and his mother,
fare wel, and they walked both together. \V And when they were departed,
his mother began to weepe, & to ſay: Thou haſt taken the ſtaffe of our
old age, and ſent him away from vs. \V I would the money had neuer bene,
for the which thou haſt ſent him. \V For our pouertie ſufficed vs, that
we might account this thing riches, that we ſay our ſonne. \V And Tobias
ſayd to her: Weepe not, our ſonne ſhal come thither ſafe, and ſhal
returne ſafe to vs, and thyne eies ſhal ſee him. \V For I beleue that
%%% !!! SNote missing in this, marked in both
\SNote{Angel Guardian.}
the good Angel of God doth accompanie him, & doth wel diſpoſe al things,
that are done about him, ſo that he ſhal returne to vs with ioy. \V At
this voyce his mother left weeping, and held her peace.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
