%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Tobias
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 1021
%%% o-0918
\startChapter[
  title={Chapter 9}
  ]

\Summary{The Angel Raphael goeth to Gabelus, receiueth the money, and
  bringeth him to the mariage. 8.~They ſalute ech other, and Gabelus
  wisheth al proſperitie to yong Tobias, and his ſpouſe.}

Then Tobias called the Angel to him, whom he thought to be a man, and he
ſayd to him: Brother Azarias, I pray thee harken to my wordes. \V If I
should deliuer my ſelf to be thy ſeruant I shal not deſerue thy
prouidence. \V Howbeit I beſech thee, that thou take vnto thee beaſtes
and ſeruantes, and goe to Gabelus into
\SNote{The Angel went to the citie it ſelfe called Rages, Tobias
remaining in the territorie or prouince therof, with Raguel, which place
is alſo called Rages.
\XRef{c.~3. v.~7.}}
Rages the citie of Medes: & render him his handwriting, and receiue of
him the money, and deſire him to come to my mariage. \V For thyſelf
knoweſt that my father numbreth the dayes: and if I ſlacke one day more,
his ſoule is made ſorowful. \V And ſurely thou ſeeſt how
%%% 1022
Raguel hath adiured me, whoſe adiuring I can not diſpiſe. \V Then
Raphael taking foure of Raguels ſeruantes, & two camels, went into Rages
the citie of Medes: & finding Gabelus gaue him his handwriting, and
receiued of him al the money. \V And he told him of Tobias the ſonne of
Tobias, al thinges that were done: and made him come with him to the
mariage. \V And when he was entered into Raguels houſe, he found Tobias
ſitting at the table: and he leaping vp, they kiſſed ech other: and
Gabelus wept, and bleſſed God, \V and ſayd: The God of Iſrael bleſſe
thee, becauſe thou art the ſonne of a verie good man, and iuſt, and that
feareth God, and doth almes deedes: \V and bleſſing be geuen vpon thy
wife, and vpon your parentes: \V & that you may ſee your children, and
your childrens children, vnto the third & fourth generation: and your
ſeede be bleſſed of the God of Iſrael, who reigneth for euer and
euer. \V And when al had ſayd, Amen; they went to the feaſt: but with
the feare of our Lord alſo did they celebrate the feaſt of the mariage.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
