%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Tobias
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 1024
%%% o-0921
\startChapter[
  title={Chapter 11}
  ]

\Summary{Leauing Sara with the reſt of the companie, and the flocke to
  folow, the Angel Raphael and Tobias goe before, 5.~are ioyfully
  receiued. 8.~Tobias annointeth his fathers eyes with the fishes gal,
  and he ſeeth. 18.~Sara arriueth ſeuen dayes after with her familie and
  cattle.}

And when they returned they came to Charan, which is in the middeway
agaynſt Niniue, the eleuenth day. \V And the Angel ſaid: Brother Tobias
thou knoweſt how thou dideſt leaue thy father. \V If it pleaſe thee
therfore, let vs goe before, and let the families folow ſoftly after vs,
together with thy wife, and with the beaſtes. \V And when this pleaſed
him that they should goe, Raphael ſaid to Tobias: Take with thee of the
gal of the fish: for it shal be neceſſarie. Tobias therfore tooke of that
gal and they departed. \V But Anna ſate beſide the way dayly, in the
toppe of a hil, from whence she might ſee afar of. \V And whiles she
watched his coming out of that place, she ſaw afar of, and by and by
perceiued her ſonne coming: and running she told her huſband ſaying:
Behold thy ſonne cometh. \V And Raphael
%%% 1025
ſaid to Tobias: But when thou art entred into thy houſe forthwith adore
our Lord thy God: and geuing thankes to him goe to thy father, and kiſſe
him. \V And immediatly annointe vpon his eies of this gal of the fish,
which thou carieſt with thee. For know thou that forthwith his eies ſhal
be opened, and thy father ſhal ſee the light of heauen, and ſhal reioice
in the ſight of thee. \V Then ranne
\SNote{It nothing diſgraceth the ſacred hiſtorie that a ſmal matter
being alſo true is recorded with the reſt. \Emph{As not one letter nor
one title of the lavv may be omitted}.
\XRef{Mat.~5.}
S.~Beda alſo expoundeth myſtically of Gods preachers. S.~Ierom doth the
like,
\Cite{in Iſaiæ.~56.}
and S.~Auguſtin
\Cite{li.~22. c.~56. cont. Fauſtum. Manich.}}
the dogge before, which had beene with them in the way, and coming as it
were a meſſenger with the fawning of his tayle reioyſed. \V And his
father that was blind ryſing vp, began to runne ſtumbling with his
feete: and geuing a ſeruant his hand, went to meete his ſonne. \V And
receiuing him kiſſed him, with his wife, and they began both to weepe
for ioy. \V And when they had adored God, and geuen thankes, they ſate
downe together. \V Then Tobias taking of the gal of the fiſh, annoynted
his fathers eies. \V And he ſtayed as it were almoſt halfe an houre: &
the white blemiſh began to come out of his eies, as it were the skine
of an egge. \V Which Tobias taking drewe from his eies, and immediatly
%%% o-0922
\SNote{God vſed this gal of a fiſh in curing Tobias eyes, in like ſorte
as the liuer in driuing away the diuel.
\XRef{c.~6. v.~8.}}
he receiued ſight. \V And they glorified God, to wit, himſelfe and his
wife, and al that knewe him. \V And Tobias ſaid: I bleſſe thee Lord God
of Iſrael, becauſe thou haſt chaſtiſed me, and thou haſt ſaued me: and
behold I ſee Tobias my ſonne. \V After ſeuen dayes alſo came in Sara his
ſonnes wife, and al the familie ſafe, and cattel, and the camels, and
much money of his wiues: and that money alſo, which he had receiued of
Gabelus: \V and he told his parentes al the benefites of God, which he
had done to him by the man, that conducted him. \V And Achior and Nabath
Tobias ſiſters ſonnes came, reioyſing at Tobias, and congratulating him
for al good thinges, that God had ſhewed towards him. \V And for ſeuen
daies making good chere, they reioyſed al with great ioy.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
