%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Tobias
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 1022
%%% o-0919
\startChapter[
  title={Chapter 10}
  ]

\Summary{The parentes lament the long abſence of their ſonne
  Tobias. 8.~Whom when Raguel can not perſwade to ſtay longer, 11.~he
  wisheth much good vnto him, and his wife, admonishing her to be
  dutiful in al thinges.}

But when Tobias taried long
\LNote{Taried long becauſe of the Mariage.}{Out
\MNote{Inſtructions to maried perſons out of the example of Tobias and
Saras Mariage.}
of this example of a holie Mariage betwen yong Tobias and Sara,
deſcribed in the fiue precedent chapters, ſome ſpecial good leſſons may
be briefly gathered, for the inſtruction & conſolation of ſuch as are to
marie, or be already maried. Not that either al, or onlie the ſame
obſeruations perteyne now to chriſtian Mariage, but that al Chriſtianes
may ſee, how orderly godlie perſons proceded in making, and performing
\MNote{True Mariage alwayes a holie contract.}
this holie contract in the old Teſtament, and therby lerne, that more
perfection is required in Chriſtian Matrimonie, being
\MNote{Now a Sacrament.}
now (as then it was not) a holie Sacrament,
\CNote{\XRef{Epheſ.~5.}}
ſignifying the Vnion betwen Chriſt & his Church:
\CNote{\XRef{Mat.~19.}}
and geuing peculiar Grace to the parties (if themſelues hinder it not)
to beare more eaſily the incident burdens, and rightly to fulfil the
proper duties of man and wife, according to the godlie inſtruction of
the
\CNote{\Cite{part.~2. de Matrim. q.~22.~23.}}
Roman Catechiſme. Al that we ſhal here note, may be reduced to two
heades, or principal partes.
\MNote{Proper inſtructions for man and wife.}
For in this Mariage there were certaine neceſſarie, and as it were
eſſential pointes, without which it had not benne a lawful nor right
contract: Other acceſſarie ceremonies, and Rites were alſo very
conuenient, for the more ſolemnitie and better performance therof. As
the like of both ſortes are now, but more perfect in Chriſtian
Mariages. In the former kind,
\MNote{Three neceſſarie pointes in Matrimonie.}
firſt of al, \Emph{the end of Mariage} was propoſed by the Angel, and
deſired by the parties,
\XRef{chap.~6. v.~22.}
\XRef{c.~8. v.~9.}
\Emph{not for pleaſure but for children, and poſteritie, by vvhich God
might be bleſſed and ſerued}: Secondly Mariage muſt be \Emph{betvven
lavvful perſonnes}. For the ſeruantes of God could not lawfully match,
neither with
\CNote{\XRef{Exod.~34.}}
Infideles, nor with 
\CNote{\XRef{Deu.~7.}}
ouer nere kinred, but 
\CNote{\XRef{Leuit.~18.}}
in conuenient degrees, and that
\CNote{\XRef{Nu.~36. v.~7.}}
ordinarily in the ſame Tribe.
\XRef{ch.~6. v.~11.}
\XRef{ch.~7. v.~14.}
as the law of Moyſes preſcribed. Thirdly, the \Emph{parties} muſt geue,
and \Emph{expreſse their mutual conſent}, and the good wil of parentes
was alſo requiſite.
\XRef{ch.~7. v.~10. 15.~20.}
Theſe principal pointes premiſed;
\MNote{Tenne godly Rites obſerued in the Mariage of Tobias and Sara.}
for the due ſolemnization other Rites were alſo adioyned.
\MNote{1.}
Firſt Raguel, the maides \Emph{father gaue his daughter} taking her by
the right hand, and ſo deliuered her into the right hand of Tobias,
\XRef{ch.~7. v.~15.}
as with vs the father, or nere freind of the woman, geueth her to be
maried to ſuch a man.
\MNote{2.}
Secondly her father prayed that it might wel ſuccede,
\XRef{ibidem.}
and ſo do al freindes now, eſpecially the Prieſt by the publique prayers
of the Church.
\MNote{3.}
The third rite, the \Emph{couenantes of Mariage} were made, with
aſſigment of dawrie, alſo \Emph{vvritten},
\XRef{ch.~7. v.~16.}
& ſealed, as the
\CNote{\Cite{Apud. Munſterum.}}
Hebrew and Greke text witneſſe.
\MNote{4.}
The fourth, Inuitation of freindes to the Mariage.
\XRef{ch.~8. v.~2.}
\XRef{ch.~9. v.~3.}
\MNote{5.}
The fifth. They made \Emph{a feaſt}, which
\CNote{\XRef{Gen.~29. v.~27.}
\XRef{Iudic.~14. v.~17.}}
commonly dured ſeuen dayes, but here fourtene
\XRef{ch.~8. v.~23.}
for the duble ioy of the mariage, and expulſion of the diuel. But though
\Emph{the feaſt} continued long, yet was \Emph{it moſt moderate} and
ſober, \Emph{vvith feare of our Lord they celebrated the feaſt of the
Mariage.} 
\XRef{ch.~9. v.~12.}
Plato a heathen Philoſopher,
\Cite{li.~6. de legibus,}
preſcribeth what temperance in meate & drinke, and what modeſtie of
behauiour, are required in mariage feaſtes. VVhoſe eccellent ſentence,
\Cite{F.~Serarius in Tob.~10.}
reciteth. 
\MNote{6.}
The ſixth rite, the bringing of the bride to her chamber,
\XRef{chap.~7. v.~19.}
the Catholique Church hath a particular forme of bleſſing the bridal
chamber.
\MNote{7.}
The ſeuenth, Tobias & Sara obſerued three dayes continencie after the
Mariage,
\XRef{ch.~6. v.~18.}
\XRef{ch.~8. v.~4.}
which now is commended by way of counſel, not of precept.
\MNote{8.}
The eight, Tobias obſerued that which the Angel aduiſed him, in burning
the liuer of the fiſh vpon coales in their chamber.
\XRef{ch.~6. v.~19.}
\XRef{ch.~8. v.~2.}
So deuout perſons commonly obſerue that which ſpiritual men exhort them
vnto, though it be not a commandment.
\MNote{9.}
The ninth, Priuate prayer of the new maried perſons,
\XRef{ch.~6. v.~18.}
\XRef{c.~8. v.~4.}
\MNote{10.}
The tenth, the ſending away of the ſpouſe to her huſbands houſe, or
dwelling place, with good wiſhes, due payment of the promiſed dawrie,
and godlie admonitions by her parentes.
\XRef{ch.~10. v.~10. 11.~12.~13.}}
becauſe of the mariage, Tobias his father was careful, ſaying: Why
thinkeſt thou doth my ſonne tarie, or why is he held there? \V Is
Gabelus dead thinkeſt thou, and no man wil reſtore him the money? \V And
he began to be ſorowful exceedingly him ſelfe, and Anna his wife with
him: and
\SNote{Such of the Iewes as beleue in Chriſt, hartely lament that he
tarieth ſo
lõg from their nation. Some more aſſuredly with old Tobias, others more
doubtfully with his wife, expect his returne.
\Cite{S.~Beda in Tobiam}
Euen ſo the remnant of Catholiques in countries fallen to hereſie
\Emph{haue great ſadnes and continual ſorovv in their hart}
\XRef{(Rom.~9.)}
\Emph{vvishing} (with what temporal loſſe ſoeuer) \Emph{the ſaluation of
their brethren, kinſmen}, and countriemen: ſome hoping more confidently
and comforting others, that Chriſt wil againe illuminate our whole
nation, as ſight was reſtored to old Tobias.}
they began both to weepe together: becauſe their ſonne did not returne
to them the day appointed. \V His mother therfore wept with
diſcomfortable teares, and ſayd: Woe, woe is me, my ſonne, why ſent we
thee to goe to a ſtrange countrie, the light of our eies, the ſtaffe of
our old age, the comforte of our life, the hope of our poſteritie? \V We
hauing al things together in thee onlie, ought not to haue let thee goe
from vs. \V To whom Tobias ſaid: Hold thy peace, and be not trubled, our
ſonne is ſafe, that man with whom we ſent him is faithful ynough. \V
Howbeit she could by no meanes be comforted, but dayly running out
looked about, and went about al waies by which there ſeemed hope he
would returne, that ſhe might ſee him a far
%%% 1023
of, if it were poſſible, coming. \V But Raguel ſayd to his ſonne in law:
Tarie here, and I wil ſend a meſſenger to Tobias thy father, that thou
art in health. \V To whom Tobias ſayd: I know that my father & my mother
do now count the dayes, and their ſpirite is tormented in them. \V And
when Raguel deſired Tobias in manie wordes, and he by no meanes would
heare him, he deliuered Sara vnto him, and the halfe part of al his
ſubſtance in men ſeruantes, & wemen ſeruantes, in cattel, in camels,
and in kine, and in much money and diſmiſt him ſafe and ioyful from
him, \V ſaying: The holie Angel of our Lord be in your iourney, and
bring you through ſafe, and that you may finde al thinges wel about your
parentes, and myn eies may ſee your children before I die. \V And the
parentes taking their daughter, kiſſed her, and let her goe: \V
admoniſhing her to honour her father and mother inlaw, to loue her
huſband, to rule the houſhold, to gouerne the houſe, and to shewe her ſelfe
irreprehenſible.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
