%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Tobias
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 1028
%%% o-0924
\startChapter[
  title={Chapter 13}
  ]

\Summary{Tobias
\MNote{The third part.

Old Tobias prayſeth God, prophecieth better ſtate of Gods people.}
the father prayſeth God, exhorting al Iſrael to do the ſame,
11.~prophecieth the reſtauration and better ſtate of Ieruſalem.}

And Tobias the elder opening his mouth, bleſſed our Lord, and ſayd: Thou
art great Lord for euer, and thy kingdom world without end: \V becauſe
thou ſcourgeſt, and ſaueſt: leadeſt downe to hel, and bringeſt backe
agayne: and there is none that may eſcape thy hand. \V Cõfeſſe to our
Lord ye children of Iſrael, and in the ſight of the Gentiles prayſe
him: \V becauſe he hath therfore diſperſed you among the gentiles, which know not
him, that you may declare his maruelous workes, and
\SNote{Temporal affliction for the ſpiritual good of others.}
make them know, that there is no other God omnipotent beſides him. \V He
hath chaſtiſed vs for our iniquities: and he wil ſaue vs for his mercie. \V
Behold therfore what he hath done with vs, and with feare and trembling
confeſſe ye to him: and extol the king of the worldes in your workes. \V
And I
\SNote{His bodie being in captiuitie, yet his ſpirit was free to praiſe
and thank God.
\Cite{S.~Cypr. li. de mortalitate}}
in the land of my captiuitie wil confeſſe to him: becauſe he hath ſhewed
his maieſtie toward a ſinful nation. \V Conuert therfore ye ſinners, &
do iuſtice
%%% o-0925
before God, beleuing that he wil doe his mercie with you. \V And I, and
my ſoule wil reioyce in him. \V Bleſſe ye our Lord al his elect,
celebrate daies of gladnes, and confeſſe to him. \V Ieruſalem the citie
of God, our Lord hath chaſtiſed thee in the workes of thy handes. \V
Confeſſe to our Lord in thy good thinges, and bleſſe the God of the
worldes, that he may
\SNote{He prophecieth the reedifying of the temple, and citie of Ieruſalem.}
reedefie his tabernacle in thee, and may cal backe al the captiues to
thee, & thou mayſt reioyce for euer and euer. \V Thou shalt shine with a
glorious light: and al the coaſtes of the earth shal adore thee. \V
Nations from far shal come to thee: and bringing giftes, they shal adore
our Lord in thee, and shal eſteeme thy land for ſanctification. \V For
they shal inuocate the great name in thee. \V Curſed shal they be that
shal contemne thee: and damned shal they be that shal blaſpheme thee:
and bleſſed shal they be that shal build thee. \V And thou shalt reioyce
in thy children, becauſe they shal
%%% 1029
al be bleſſed, & shal be gathered together to our Lord. \V Bleſſed are
al that loue thee, and that reioyce vpon thy peace. \V My ſoule, bleſſe
thou our Lord, becauſe he hath deliuered Ieruſalem his citie
\SNote{Onlie celeſtial Ieruſalem ſhal be free from al tribulations.}
from al her tribulations, the Lord our God. \V Bleſſed ſhal I be if
there ſhal remayne of my ſeede, to ſee the glorie of Ieruſalem. \V The
gates of Ieruſalem ſhal be built of Saphire and the Emerauld: and al the
compaſſe of the walles therof of pretious ſtone. \V With white and
cleane ſtone ſhal al the ſtreates therof be paued: and in the ſtreates
therof
\SNote{No Chriſtian is ignorant (ſayth S.~Auguſtin
\Cite{Epiſt.~86.)}
that \Emph{Alleluia} is a voice of praiſe. In Engliſh it is \Emph{Praiſe
ye the Lord vvith ioy.}}
\Emph{Alleluia} ſhal be ſong. \V Bleſſed be our Lord, which hath exalted
it, and his kingdom be for euer and euer ouer it. Amen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
