%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Tobias
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 1013
%%% o-0911
\startChapter[
  title={Chapter 3}
  ]

\Summary{The prayer of Tobias, 7.~and Sara, in their ſeueral
  afflictions, 24.~are heard by God, and the Angel Raphael is ſent to
  releeue them.}

Then Tobias lamented, and began to pray with teares, \V ſaying: Thou art
iuſt ô Lord, & al thy iudgementes are iuſt, and al thy waies, mercie, &
truth, and iudgement. \V And now Lord be mindful of me, and take not
reuenge of my ſinnes, neither remember the ſinnes of me, or of my
parentes.
%%% 1014
\V Becauſe we haue not obeyed thy commandmentes, therfore we are
deliuered in ſpoile, and captiuitie, and death, and into a fable, and
into reproch to al nations, in which thou haſt diſperſed vs. \V And now
Lord great are thy iudgementes, becauſe we haue not done according to
thy preceptes, & haue not walked ſincerely before thee: \V and now Lord
according to thy wil do with me, & command my ſpirite to be receiued in
peace: for it is expedient for me to die, rather then to liue. \V The
verie ſame day therfore it chanced that Sara the daughter of Raguel in
\SNote{In a prouince of the Medes wherof Rages was the head citie, for
when they came where Raguel dwelt, Tobias ſtayed there, and the Angel
went to the citie of Rages where Gabelus dwelt.
\XRef{c.~9.}
As one may ſay: ſuch a one dwelleth in Rome, that dwelleth in anie part
of Romania: in Yorck, Lincolne, or Mummoth, that dwelleth in one of
thoſe ſhires.}
Rages a citie of the Medes, ſhe alſo heard reproch of one of her fathers
handmaydes, \V becauſe ſhe had beene deliuered to ſeuen huſbandes, & the
diuel named
\SNote{\Emph{Aſmodeus}, ſignifying \Emph{Deſtroyer} is a captaine or
king of thoſe diuels which ſpecially deſtroy ſoules by the ſinnes of the
fleſh, & afterward tormenteth both ſoules and bodies for the ſame
ſinnes.}
Aſmodeus had killed them, forthwith as they were entred in vnto her. \V
Therfore when she rebuked the wench for her fault, she anſwered her,
ſaying: Let vs no more ſee ſonne of thee, or daughter vpon the earth,
thou murderer of thy huſbandes. \V What wilt thou kil me alſo, as thou
haſt now killed ſeuen husbandes? At this voice ſhe went into an higher
chamber of her houſe: and three dayes, and three nightes did not eate,
nor drinke: \V but continewing in prayer with teares beſought God, that
he would deliuer her from this reproch. \V And it came to paſſe the
third day, whiles ſhe accomplished her prayer, bleſſing our Lord, \V ſhe
ſayd: Bleſſed is thy name ô God of our fathers: who when thou haſt bene
angrie, wilt doe mercie, & in the time of tribulation forgeueſt them
their ſinnes, that inuocate thee. \V To thee ô Lord I turne my face, to
thee I direct myne eyes. \V I deſire Lord that thou looſe me from the
bond of this reproch, or els take me away from the earth. \V Thou
knoweſt Lord that I neuer coueted a huſband, and haue kept my ſoule
cleane from al concupiſcence. \V Neuer haue I companied my ſelf with
ſporters: neither haue I made my ſelfe partaker with them that walke in
lightneſſe. \V But a husband I conſented to take, with thy feare, not
with my luſt. \V And either I was vnworthie of them, or they perhaps
were not worthie for me: becauſe perhaps thou haſt kept me for an other
man. \V For thy counſel is not in mans power. \V But this hath euerie
one for certayne, that worſhippeth thee, that his life, if it be in
probation, shal be
%%% o-0912
crowned: and if it be in tribulation, it shal be deliuered: and if it be
in correction, it shal be lawful to come to thy mercie. \V For thou art
not delighted in our perditions: becauſe after
%%% 1015
a tempeſt thou makeſt a calme, and after teares and weeping thou powreſt
in ioyfulneſſe. \V Be thy name God of Iſrael bleſſed foreuer. \V At that
time the prayers of both were heard in the ſight of the glorie of the
high God: \V and the holie Angel of our Lord Raphael was ſent, to cure
them both, whoſe prayers at one time were
\SNote{\XRef{Act.~10.}
An Angel ſhewed Cornelius that his prayers were heard.
\XRef{Apoc.~5.}
Prayers of the faithful are offered to God by Angels, & other Sainctes.}
recited in the ſight of our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
