%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Tobias
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 1019
%%% o-0915
\startChapter[
  title={Chapter 7}
  ]

\Summary{They are kindly entertained by Raguel. 10.~Tobias demandeth
  Sara to wife, which Raguel, encoredged by the Angel, granteth, 15.~and
  the mariage is made.}

And they went to Raguel, and Raguel receiued them with ioy. \V And
Raguel beholding Tobias, ſayd to Anna his wife: How like is this yongman
to my ſiſters ſonne! \V And when he had ſpoken theſe wordes, he ſayd:
Whence are you ye yongman our brethren? \V But they ſayd: We are of
the tribe of Nephthali, of the captiuitie of Niniue.
%%% o-0916
\V And Raguel ſayd to them: Know you Tobias my brother? Who ſayd: We
know him. \V And when he ſpake much good of him, the Angel ſayd to
Raguel: Tobias, of whom thou askeſt is this mans father. \V And Raguel
put forth him ſelfe, and with teares kiſſed him, and weeping vpon his
necke, ſayd: Bleſſing haue thou my ſonne, becauſe thou art the ſonne of
a good and moſt vertuous man. \V And Anna his wife, and Sara theyr daughter
wept. \V And after they had talked, Raguel commanded a wether to be
killed, and a banket to be prepared. And when he deſired them to ſitte
downe to dinner, \V Tobias ſaid: I wil not eate nor drinke here this
day, vnleſſe thou firſt aſſure my petition, and promiſe to geue me Sara
thy daughter. \V Which word Raguel hearing, was ſore afrayd, knowing what had
chanced to thoſe ſeuen husbands, which went in vnto her: and he began to
feare leſt perhaps it might chance to him alſo in like maner: and when
he doubted, and
\SNote{A iuſt man (ſayth S.~Ambroſe
\Cite{lib.~3. Off. c.~14.)}
feared other mens harmes, and would rather his daughter ſhould not be
maried then others ſhould be in danger, preferring honeſtie before
profite.}
gaue no anſwer to him demanding, \V the Angel ſayd to him: Feare not to
geue her to this man, for to him fearing God is thy daughter dewe to be
his wife: therfore an other could not haue her. \V Then ſayd Raguel: I
doubt not but God hath admitted my prayers and teares in his ſight. \V
And I beleue that therfore he hath made you come to me, that this mayde
might be ioyned to her kinred, according to the law of Moyſes: and now
haue no doubt but I wil deliuer her
%%% 1020
to thee. \V And taking his daughter by the right hand, gaue it into the
right hand of Tobias, ſaying: The God of Abraham, and the God of Iſaac,
and the God of Iacob be with you, and he ioyne you together, and fulfil
his bleſſing in you. \V And taking paper they made a writing of the
mariage. \V And after theſe things they made merie, bleſſing God. \V And
Raguel called to him Anna his wife, and commanded her to prepare an
other chamber. \V And ſhe brought Sara her daughter in thither, and ſhe
wept. \V And ſhe ſayd to her: Be of good cheere my daughter, our Lord of
heauen geue thee ioy for the tediouſneſſe which thou haſt ſuffered.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
