%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Tobias
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 1017
%%% o-0914
\startChapter[
  title={Chapter 6}
  ]

\Summary{By the Angels aduiſe yong Tobias apprehendeth a fish, that
  aſsaulteth him, 5.~reſerueth the hart, gal, and liuer for
  medicines. 10.~They lodge at the houſe of Raguel, whoſe daughter Sara
  Tobias is to marie. 14.~A diuel hath heretofore ſlaine her ſeuen
  husbandes, 16.~but shal not hurt him.}

And Tobias went forward, & a dogge folowed him, and he made his firſt
abode by the riuer of Tigris. \V And he went out to waſh his feete, and
behold an huge fiſh came forth to deuoure him. \V Of whom Tobias being
afrayd, cryed
%%% 1018
out with a loud voice, ſaying: Sir, he inuadeth me. \V And the Angel
ſaid to him: Take him by the gille, & draw him to thee. Which when he
had done, he drew him on the drie land, and he began to ſtrugle before
his feete. \V Then ſayd the Angel to him: Take out the entralles of this
fiſhe, and his hart, and gal, and liuer, keepe to thee: for theſe are
neceſſarie and profitable for medecines. \V Which when he had done, he
roſted the
\SNote{S.~Paul alſo calleth fleſh of fiſh
\XRef{1.~Cor.~15.}
and Plinie
\Cite{lib.~9. c.~15.}}
fleſh therof, and they tooke it with them in the way: the reſt they
ſalted which might ſuffiſe them, til they came to Rages the citie of
Medes. \V Then Tobias asked the Angel, and ſayd to him: I beſech thee
brother Azarias, tel me what remedies ſhal theſe things haue, which thou
haſt
%%% o-0915
bid me keepe of the fiſh? \V And the Angel anſwering, ſayd to him: If
thou put a litle peece of his
\TNote{and liuer,
\XRef{v.~19.}}
hart vpon coales, the
\SNote{Diuels who exalted them ſelues as equal with God, are iuſtly made
ſubiect to corporal creatures. God cõcurring with natural cauſes, whoſe
good pleaſure is ſometimes to vſe inſtrumẽts naturally vnapt, as when
Chriſt gaue ſight to the blinde by putting clay on his eyes,
\XRef{Ioan.~9.}
ſometimes more apt, as when he fed manie with few loaues.
\XRef{Ioan.~6.}
So the Angel by Gods appointment vſed this meanes to expel the diuel.}
ſmoke therof driueth out al kinde of diuels, either from man or from
woman, ſo that it cometh no more vnto them. \V And the gall is auailable
to annoynt the eies, in which there ſhal be white blemiſh, and they ſhal
be healed. \V And Tobias ſayd to him: Where wilt thou that we tarie? \V
And the Angel anſwering, ſayd: Here is one named Raguel, a nerekinſman
of thy tribe, and he hath a daughter named Sara, and he hath neither man
child, nor anie woman child beſide her. \V Al his ſubſtance is dewe to
thee, and thou mayſt take her to wife. \V Aske her therfore of her
father, and he wil giue her thee to wife. \V Then Tobias anſwered, and
ſayd: I heare that she hath beene deliuered to ſeuen husbandes, and they
are dead: yea and I haue heard, that a diuel killed them. \V I am
afrayd therfore, leſt theſe things may happen to me alſo: & whereas I am
the onlie childe of my parentes, I may bring downe their old age with
ſorow vnto
\SNote{Into the place where good ſoules reſted, none then hauing acceſſe
into heauen. See
\XRef{Annotations Gen.~37.}}
hel. \V Then the Angel Raphael ſayd to him: Heare me, and I wil shewe
thee who they are, on whom the diuel can preuaile. \V For they that ſo
receiue matrimonie, that they exclude God from them ſelues, and from
their mind, and ſo geue them ſelues to their luſt, as horſe and mule,
which haue not vnderſtanding, ouer them the diuel hath power. \V But
thou when thou shalt take her, entring into the chamber, for three daies
be continent from her, and thou shalt geue thy ſelf to nothing els but
to prayers with her. \V And the ſame night, the
\TNote{and hart
\XRef{v.~8.}}
liuer of the fish ſette on the fyre, the diuel shal be driuen away. \V
But the
\SNote{The ſecond night he asked and obtayned this grace; for he knew
not his wife vntil the fourth night.
\XRef{v.~22.}}
ſecond night thou shalt be admitted in the copulation of
%%% 1019
the holie Patriarches. \V And the third night thou shalt obteyne a
bleſſing that ſound children may be procreated of you. \V And when the
third night is paſt, thou shalt take the virgin with the feare of our
Lord, moued rather for loue of children then for luſt, that in the ſeede
of Abraham thou mayſt obteyne bleſſing in children.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
