%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Tobias
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 1012
%%% o-0909
\startChapter[
  title={Chapter 2}
  ]

\Summary{Tobias to burie an Iſraelite that is ſlaine in the ſtreete,
  leaueth his dinner and gheſtes. 10.~Is made blind by Gods permisſion,
  for manifeſtation of his patience. 19.~His wife getteth her liuing by
  worke, 22.~and for a ſcrupulous word, reprocheth his ſinceritie.}

But after theſe thinges, when there was a feſtiual day of our Lord, and
a good dinner was made in Tobias houſe, \V he ſayd to his ſonne: Goe,
and bring ſome of our tribe,
\SNote{Al the people of the tenne tribes did not ſerue Ieroboams golden
calues, but ſome \Emph{feared God}  & conſequently \Emph{refrayned from
euil},
\XRef{Prou.~3.}
at leaſt from idolatrie.}
that feare God, to make merie with vs. \V And when he had gone,
returning he tolde him, that one of the children of Iſrael lay ſlayne in
the ſtreate. And he forthwith leaping vp from his place at the table,
leauing his dinner, came faſting to the bodie: \V and taking it vp
caried it to his houſe ſecretely,
%%% o-0910
that when the ſunne should be downe, he might warely burie him. \V And
whẽ he had hid the bodie, he eate bread with mourning and trẽbling, \V
remembring that word, which our Lord ſayd by Amos the Prophet:
\CNote{\XRef{Amos.~8. v.~10.}}
Your feſtiual dayes shal be turned into lamentation and mourning. \V But
when the ſunne was downe, he went, and buried him. \V And al his
neighbours rebuked him, ſaying: Euen now thou waſt commanded to be
ſlayne becauſe of this matter, and thou dideſt ſcarce
%%% 1013
eſcape the commandment of death, and
\SNote{True zele is not hindred from workes of mercie by feare of death:
\Emph{becauſe perfect charitie caſteth out feare}.
\XRef{1.~Ioan.~4.}}
doeſt thou burie the dead agayne? \V But Tobias more fearing God, then
the king, in haſt tooke the bodies of them that were ſlayne, and hid
them in his houſe, and at midnight buried them. \V And it happened that
on a certayne day, being wearied with burying, coming into his houſe, he
had caſt himſelf downe by the wal, and ſlept, \V and as he was ſleeping,
hote dung out of the ſwallowes neſt fel vpon his eyes, and he was made
blinde. \V And this tentation therfore our Lord permitted to chance vnto
him, that an example might be geuen to poſteritie of his patience, alſo
of holie Iob. \V For whereas he feared God alwaies from his infancie,
and kept his commandmentes, he grudged not agaynſt God for that the
plague of blindnes had chanced to him, \V but continewed immoueable in
the feare of God, geuing thankes to God al the dayes of his life. \V For
as the kinges inſulted againſt bleſſed Iob: ſo his
\SNote{Both elder and yonger ſorte of his kinred derided him, not his
proper parentes, for he was depriued of father and mother when he was a
child, as it ſemeth.
\XRef{c.~1. v.~4.}}
parentes and coſins derided his life, ſaying: \V Where is thy hope, for
the which thou dideſt beſtowe almes and burials? \V But Tobias rebuked
them, ſaying: Speake not ſo: \V becauſe we are the children of holie
ones, and looke for that life, which God wil geue to them, that neuer
change their faith from him. \V But Anna his wife went dayly to weauing
worke, and she brought the gaines of her handie labour, which ſhe could
get. \V Whereby it came to paſſe, that she receiuing a kid of goates had
brought it home: \V the voice whereof bleating when her huſband had
heard, he ſayd: Take heed, leſt perhaps it be ſtollen, reſtore ye it to
his owners, becauſe it is not lawful for vs either to eate anie thing of
theft or to touch it. \V To theſe wordes his wife being angrie anſwered:
Thy hope is become vayne manifeſtly, and thine almes now haue
appeared. \V And with theſe, and other ſuch like wordes she vpbrayded
him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
