%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Tobias
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 1025
%%% o-0922
\startChapter[
  title={Chapter 12}
  ]

\Summary{Old Tobias and his ſonne offer the half of al the goodes which
  they had newly receiued to Raphael, for his wages. 6.~VVho then
  declareth to them that he was ſent from God to helpe them, 15.~and
  that he is an Angel. 20.~He parteth away, and they render thankes to
  God.}

Then Tobias called to him his ſonne, and ſayd to him: What may we geue to
this holie man, that is come with thee. \V Tobias anſwering, ſayd to his
father: Father what reward ſhal we geue him? or what can be worthie of
his
%%% 1026
benefittes? \V He hath
\MNote{1.}
led me and brought me agayne ſafe, he
\MNote{2.}
receiued the money of Gabelus, he
\MNote{3.}
cauſed me to haue my wife, and the
\MNote{4.}
euil ſpirit he chaſed from her, he
\MNote{5.}
made ioy vnto her parentes, my ſelf he
\MNote{6.}
deliuered from being deuoured of the fish,
\MNote{7.}
thee alſo he hath made to ſee the light of heauen, and we are
replenished with
\SNote{Gratful Tobias recounteth ſeuen benefites receiued by the
companion of his iourney, & that they \Emph{haue receiued} (not manie
but) \Emph{al good things by him}.}
al good things by him. What can we geue him worthie for theſe things? \V
But I beſeech thee my father, that thou deſire him, if perhaps he wil
voutſafe to take vnto him the one halfe of al thinges, which are
brought. \V And they calling him, to wit the father and the ſonne, tooke
him aſide: & began to deſire him that he would voutſafe to accept the
halfe part of al thinges, that they had brought. \V Then he ſayd to them
ſecretely: Bleſſe ye the God of heauen, and before al that liue confeſſe
to him, becauſe he hath done mercie with you. \V For to hide the ſecrete
of a king is good: but to reueale & confeſſe the workes of God is an
honorable thing. \V
\SNote{Faſting and almes are as two
\Fix{wingues}{wings}{possible typo, fixed in other}
with which prayer flyeth into heauen.}
Prayer is good with faſting and almes, rather then to lay vp treaſures
of gold: \V becauſe almes deliuereth from death, and that is it which
purgeth ſinnes, and maketh to finde mercie and life euerlaſting. \V But
they that committe ſinne and iniquitie, are enemies to their owne
ſoule. \V I open therfore vnto you the truth, and I wil
%%% o-0923
not hide from you the ſecrete word. \V When thou didſt pray with teares,
and didſt burie the dead and
leaft thy dinner, and dideſt hide the dead by day in thy houſe, and by
night didſt burie them,
\LNote{I offered thy prayers to our Lord.}{Here
\MNote{Offices of Angels towardes men.}
the Angel Raphael reporteth certaine good offices, which he had done for
Tobias. He did other like for his ſonne, and for Raguel, and his
daughter, which are likewiſe recorded in this booke. And the whole
world, eſpecially Gods ſeruantes, receiue continual great benefites by
holie Angeles, as partly may be gathered in this holie hiſtorie, and
more elſwhere.
\MNote{Readie to helpe al.}
For firſt the office of Angeles is, to aſſiſt, or be alwayes readie, as
moſt diligent ſeruitoures of God, expecting what his diuine goodnes wil
appoint them, whither to goe, and what to doe for the benefite of men:
as holie Raphael was ſent, when yong Tobias wanted a guide.
\MNote{Offer mens prayers and good workes.}
Secondly, Angels offer the prayers of the faithful, or as the Greke text
readeth,
\XRef{v.~15.}
Angels preſent the prayers of Sainctes, that is, of godlie men and wemen
to God: ſo Raphael teſtifieth here himſelf, that he offered Tobias
prayers to our Lord.
\MNote{Ayde the godlie.}
Thirdly, Angeles ayde and aſſiſt thoſe that loue puritie of life, ſincere
ſeruice of God, hate vice, embrace vertue, & do workes of mercie; ſo
Raphael aſſiſted Tobias, when he traueled to burie the dead, fleing from
the kings furie, and hiding himſelf.
\XRef{ch.~1. v.~21.~23.}
\MNote{Exhort to good.}
Fourthly Angels echort to good workes, as in this
\XRef{12.~chapter v.~6. 8.~9.~10.~18.}
\MNote{Inſtruct.}
Fiftly they ſuggeſt and inſtruct what to do,
\XRef{ch.~6. v.~4.~5.}
Raphael taught yong Tobias to take the fiſh, vnbowel him, reſerue partes
therof,
\XRef{v.~11. 12.~13.}
aduiſed him to lodge at Raguels houſe, to demand Sara to wife, and
\XRef{v.~16.~17.}
inſtructed him, againſt whom diuels haue powre.
\MNote{Expel euil ſpirites.}
Sixthly, they expel diuels from perſons, and places,
\XRef{ch.~8. v.~3.}
Raphael tooke and bound the diuel \Emph{Aſmodeus} in the deſert of
hiegher Ægypt.
\MNote{Deliuer from euils and dangers.}
Seuenthly, they deliuer men from dangers and euiles,
\XRef{c.~6. v.~3.}
as when the great fiſh aſſaulted Tobias, and Sara from moleſtation, and
ſlaunder, and old Tobias from blindnes.
\XRef{chap.~3. v.~10.}
\XRef{ch.~11. v.~8.}
\XRef{ch.~12. v.~14.}
\MNote{Procure temporal commodities for the ſoules good.}
Eightly vvhen it redoundeth to the honour and more ſeruice of God, and
good of the ſoule, Angeles procure riches, and worldly commodities.
\XRef{ch.~12. v.~3.}
Yong Tobias gratfully confeſſed the great benefites receiued by his
guide, concluding generally, \Emph{by him vve are replenished} (ſayth
he) \Emph{vvith al good thinges}.
\MNote{Proue the good by tribulations.}
Ninthly, good Angels alſo proue men, for their more merite, ſo the Angel
witneſſeth
\XRef{ch.~12. v.~13.}
\Emph{Becauſe thou vvaſt acceptable to God, it vvas neceſſarie, that
tentation should proue thee.}
\MNote{Comforte the patient and al vertuous.}
Tenthly, and finally (for we remite the reader to larger documentes of
others) after proofe of patience, fortitude, and other vertues, holie
Angeles comforte good men: ſo Raphael encoraged old Tobias, ſaying,
\XRef{ch.~5. v.~13.}
his blindenes ſhould ſhortly be cured,
\XRef{ch.~12. v.~12.}
ſhewed him how gratful his prayers with teares and
\Fix{woorkes}{workes}{obvious typo, fixed in other}
of mercie were in Gods ſight.
\MNote{Comforte the patient and al vertuous.}
He comforted Raguel and his familie by bringing yong Tobias to their
houſe.
\XRef{ch.~7. v.~5.}
Much more,
\XRef{ch.~8. v.~16.}
both them and al Tobias his familie, by driuing away the diuel; and
laſtly by reuealing himſelfe vnto them.
\MNote{Angeles guardians haue ſpecial care of ſoules committed to their
charge.}
So holie Angeles, eſpecially the
proper gardian patrones of euerie one, are alwayes readie to helpe men;
guard them, exhort them to good, do inſtruct them, do expel euil
ſpirites, and deliuer men from many euils & dangers, do procure them
temporal commodities, proue their vertues, offer their prayers and good
workes, aſſiſt them al their liues, and at their deathes, then alſo
bring their ſoules to the Iudgement ſeate, and (if they die in good
ſtate) to eternal ioy and glorie, wherof the ancient fathers writinges
are ful.
S.~Gregorie the great in his
\Cite{Dialogue, li.~4. c.~58.}
S.~Athanaſius
\Cite{li. de communi eſſentia.}
S.~Chryſoſtom
\Cite{ho.~3. in Epiſt. ad Colloſſ.}
\Cite{li.~6. de Sacerdota.}
Gregorius Turonen
\Cite{li. de gloria Martyrum & Confeſſorum.}
S.~Auguſtin
\Cite{Epiſt. ad probam. c.~9.}
\Cite{Epiſt.~68. ad fratres in ererro.}
\Cite{li.~11. c.~31. ciuit.}
&
\Cite{octoginta trium. qq.~q.~79.}
Our Sauiour himſelf teſtifieth that Angels reioyce at the conuerſion of
a ſinner,
\XRef{Luc.~15.}
and therfore, they know and haue care of mens ſtates in this life, and
finally Angeles caried the ſoule of poore Lazarus into Abrahams boſome.
\XRef{Luc.~16.}}
I
\SNote{O how ſwete (or excellent) a thing is it (ſaith S.~Auguſtin
\Cite{ſer.~3. de Natiuit.)}
when Angeles guardianes of our life offer our vowes (or reſolution to
flee vices and embrace vertues) before the ſight of Gods Maieſty?}
offered thy prayer to our Lord. \V And becauſe thou waſt acceptable to
God, it was neceſſarie that tentation should proue thee. \V And now our
Lord ſent me to cure thee, and to deliuer Sara thy ſonnes wife from the
diuel. \V For I am Raphael an Angel, one of the ſeuen, which aſſiſt
before our Lord. \V And when they had heard theſe thinges, they were
trubled, & trembling fel vpon the ground on their face. \V And the Angel
ſayd to them: Peace be to you, feare not. \V For when I was with you, by
the wil of God I was ſo: bleſſe ye him, and ſing to him. \V I ſeemed
indeede to eate with you, and to drinke: but I vſe an inuiſible meate and
drinke, which can not be ſeene of men. \V It is time therfore that I
returne to him, that ſent me: but bleſſe ye God, and tel al his
maruelous workes. \V And when he had ſayd theſe things, he was taken
from their ſight, and they could ſee him no more. \V Then proſtrate for
three houres vpon their face, they bleſſed God: and ryſing vp they told
al his maruelous workes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
