%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ruth
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 0588
%%% o-0528
\startChapter[
  title={Chapter 3}
  ]

\Summary{Ruth inſtructed by her mother in law ſleepeth at Booz feete,
  8.~and ſignifying that she perteyneth to him by the law of affinitie,
  receiueth a good anſwer, 14.~and ſix meaſures of barley.}

But after that she was returned to her mother in law, she heard of her:
My daughter, I wil ſeeke thee reſt, and wil prouide that it may be wel
with thee. \V This Booz, to whoſe maides thou art ioyned in the field,
is our nigh kinſman, and this night he wynoweth the barne floore of the
barley. \V Wash therfore and annoynte thy ſelf, and put on thy better
garmentes, and goe downe into the barne floore, let no man ſee thee, til
he shal haue ended eating & drinking. \V And when he shal goe to ſleepe,
marke the place wherein he ſleepeth: and thou
%%% o-0529
shalt come, and diſcouer the mantel wherwith he is couered toward his
feete, and shal caſt thy
%%% 0589
ſelf downe and lie there:
\SNote{The euent ſhewed that Noemi was inſpired by God to geue ſuch
direction to Ruth, & to foretel what Booz would doe.}
and he wil tel thee what thou muſt doe. \V Who anſwered: Whatſoeuer thou
shalt command, that wil I doe. \V And she went downe into the barne
floore, and did al the thinges which her mother in law had commanded
her. \V And when Booz had eaten, & drunken, and was made pleaſant, and
was gone to ſleepe by the heape of sheaues, she came cloſely, and
diſcouering the mantel, at his feete, layd her ſelf downe. \V And
behold, when it was now midnight the man was afrayd, and trubled: and he
ſaw a woman lying at his feete, \V and ſaid to her: Who art thou? And
she anſwered: I am Ruth thy handmaide: ſpred thy mantel vpon thy
ſeruant, becauſe thou art nigh of kinne. \V And he ſaid: Bleſſed art
thou of our Lord my daughter, and the former mercie thou haſt paſſed
with the later: becauſe thou haſt
\SNote{It was very cõmendable that ſhe loued her firſt huſband and
mother in law: but more vertue in fleing occaſion of ſinne with young
men, and ſeeking to marie according to the law of God with her former
huſbandes kinſman.
\XRef{Deut.~25.}}
not folowed yong men either poore or rich. \V Feare not therfore, but
whatſoeuer thou shalt ſay to me, I wil doe to thee. For al the people
that dwelleth within the gates of my citie, knowe, that thou art a
woman of vertue. \V Neither doe I denie my ſelf nigh of kinne, but there
is an other neerer then I. \V Reſt this night: and when morning is come,
if he wil retayne thee by the right of nigh of kindred, the thing is wel
done, but if he wil not, I wil take thee without al doubt, our Lord
liueth, ſleepe vntil morning. \V She ſlept therfore at his feete til the
night was gone. Therfore she aroſe before men could know one an other,
and Booz ſaid: Beware leſt any man know that thou cameſt hither. \V And
agayne: Spred, quoth he, thy mantel, wherwith thou art couered, and hold
it with both handes. Who ſpredding and holding it, he meaſured ſix
meaſures of barley, and put it vpon her. Who carying it entred into the
citie, \V and came to her mother in law. Who ſaid to her: What haſt thou
done daughter? And she told her al thinges, that the man had done to
her. \V And she ſaid: Behold ſix meaſures of barley hath he geuen me,
and he ſaid: I wil not haue thee returne emptie to thy mother in law. \V
And Noemi ſaid: Expect daughter til we ſee what end the thing wil
haue. For the man wil not ceaſe vntil he haue accomplished that which he
hath ſpoken.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
