%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ruth
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 0589
%%% o-0530
\startChapter[
  title={Chapter 4}
  ]

\Summary{Booz before the ancientes of the citie (the neerer kinſman
  refuſing) poſſeſſeth the inheritance Elimelech, 10.~and marieth
  Ruth. 13.~Hath by her a
%%% 0590
ſonne, the grandfather of Dauid. 18.~VVhoſe
  genealogie by this occaſion is recited, from Phares the ſonne of Iudas
  the patriarch.}

Booz therfore went vp to the gate, and ſate there. And when he had ſeene
the nigh kinſman paſſe by, of whom the talke was had before, he ſaid to
him: Turne in a litle while, and ſitte here: calling him by his
name. Who turned in, and ſate. \V And Booz taking ten men of the citie,
ſaid to them: Sitte ye here. \V Who ſitting downe, he ſpake to the nigh
kinſeman: Noemi, who is returned from the countrie of Moab, wil ſel the
part of the field belonging to
\SNote{Booz calleth his kinſman brother, as Abraham called Lot his
brother,
\XRef{Gen.~13.}
being \Emph{his Nephevv}.}
our brother Elimelech. \V Which I would thee to vnderſtand, and would
tel thee before al that ſitte, and the ancientes of my people. If thou
wilt poſſeſſe it by the right of nigh kindred: bye, and poſſeſſe it. But
if it pleaſe thee not, tel me the ſame, that I may know what I ought to
doe. For there is no nigh kinſeman ſauing thee, which art firſt, and me,
who am ſecond. But he anſwered: I wil bye the field. \V To whom Booz
ſaid: When thou shalt bye the field at the womans hand, thou muſt take
alſo Ruth the Moabite, which was the wife of the deceaſed: that thou
mayeſt rayſe vp the name of thy kinſman in his inheritance. \V Who
anſwered: I yeld my right of nigh kindred: for I may not aboliſh the
poſteritie of myne owne familie. Doe thou vſe my priuiledge, which I
profeſſe that I doe willingly forgoe. \V And
\SNote{See
\XRef{Deut.~25.}
noting here withal, that the penaltie was leſſe, when an other kinſman
vndertaking the mariage, the woman was preuẽted from complayning before
the iudge.}
this in old time was the maner in Iſrael betwen kinſemen, that if at any
time one yelded to an other his right: that the graunt may be ſure, the
man put of his ſhoe, and gaue it to his neighbour. This was a teſtimonie
of yelding in Iſrael. \V Booz therfore ſaid to his kinſeman: Take of thy
ſhoe. Which immediatly he looſed from his foote. \V But to the
ancientes, and the whole people he ſaid: You are witneſſes this day, that
I haue purchaſed al thinges which were Elimelechs; and Chelions and
Mahalons, Noemi deliuering them: \V and haue taken in mariage Ruth the
Moabite, the wife of Mahalon, that I may rayſe vp the name of the
deceaſed in his inheritance, leſt his name be aboliſhed out of his
familie and brethren and people. You, I ſay, are witneſſes of this
thing. \V Al the people that was in the gate anſwered, and the
ancientes: We are witneſſes: Our Lord make this woman, which entereth
into thy houſe, as Rachel, and Lia, which
%%% 0591
%%% o-0531
builded the houſe of Iſrael: that ſhe may be an example of vertue in
Ephrata, and may haue a famous name in Bethlehem: \V and that thy houſe
may be, as the houſe of Phares, whom Thamar bare to Iudas, of the ſeede
which our Lord ſhal geue thee of this yong woman. \V Booz therfore tooke
Ruth, and had her to wife: and went in vnto her, and our Lord gaue her
to conceiue, and to beare a ſonne. \V And the wemen ſaid to Noemi:
Bleſſed be our Lord, which hath not ſuffered that there should fayle a
ſucceſſor of thy familie: that his name ſhould be called in Iſrael. \V
And thou ſhouldeſt haue one that may comfort thy ſoule, and cheriſh thy
old age. For of thy daughter in law is he borne, which wil loue thee:
and much better is ſhe to thee, then if thou hadſt ſeuen ſonnes. \V And
Noemi taking the child put it in her boſome, and did the office of
a nource and of one that ſhould carie him. \V And the wemen her neighbours
congratulating her, and ſaying: There is a ſonne borne to Noemi: called
his name Obed: this is
\SNote{Here appeareth the final cauſe of writing this hiſtorie, to ſhew
the Genealogie of King Dauid from Iudas the Patriarch, of whom Chriſt
ſhould deſcend, ſo prophecied:
\XRef{Gen.~49.}
and ſhewed to be performed:
\XRef{Mat.~1.}}
the father of Iſai, the father of Dauid. \V Theſe are the generations of
Phares: Phares begat Eſron, \V Eſron begat Aram, Aram begat Aminadab, \V
Aminadab begat Nahaſſon, Nahaſſon begat Salmon, \V Salmon begat Booz,
Booz begat Obed, \V Obed begat Iſai, Iſai begat Dauid.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
