%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ruth
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 0585
%%% o-0525
\startChapter[
  title={Chapter 1}
  ]

\Summary{By occaſion of famine Elimelech of Bethleem going with his wife
  Noemi, and two ſonnes, into the Land of Moab, there dieth. 4.~His
  ſonnes marrie wiues of that countrie, and die without iſſue. 6.~Noemi
  returning homewardes hardly perſwadeth one of her daughters in law, to
  part from her. 15.~The other, called Ruth, wil needes goe with her,
  profeſsing the ſame God and Religion. 19.~So theſe two arriue in
  Betheleem.}

In the dayes of one Iudge, when the Iudges ruled, there came a famine in
the Land. And there went a man of Bethleem Iuda, to ſeiourne in the land
of Moab with his wife, and two children. \V Himſelf was called
Elimelech, and his wife, Noemi: and his two ſonnes, the one Mahalon, and
the other Chelion, Ephraites of Bethleem Iuda. And entring into the
countrie of Moab, they abode there. \V And Elimelech the huſband of
Noemi died: and ſhe remained with her ſonnes. \V Who tooke wiues of
the Moabites, of the which one was called Orpha, and the other Ruth. And
they abode there ten yeares, \V and both died, to witte, Mahalon and
Chelion: and the woman remayned deſtitute of her two children & her
huſband. \V And ſhe aroſe to goe into her countrie with both her
daughters in law from the countrie of Moab: for ſhe had heard that our
Lord had reſpected his people, & had geuen them victuals. \V She
therfore went forth from the place of her peregrination, with both her
daughters in
%%% o-0526
law: and being now ſette in the way to returne into the Land of Iuda, \V
ſhe ſaid to them: Goe into your mothers houſe, our Lord doe mercie with
you, as you haue done with the dead and with me. \V Grant he vnto you to
find reſt in the houſes of your huſbandes, which you ſhal take. And she
kiſſed them. Who lifting vp their voice beganne to weepe, \V & to ſay:
We wil goe on with thee to thy people. \V To whom ſhe anſwered: Returne
my daughters, why come you with me? shal I haue ſonnes any more in my
wombe, that you may hope for huſbandes of me? \V Returne my daughters,
and goe your wayes: for I am now ſpent with old age, and not fitte for
wedlocke. Although I might conceiue this
%%% 0586
night, and beare children, \V if you would expect til they grow, and be
of mans age, you ſhal be old women before you marrie. Doe not ſo my
daughters, I beſech you: for your diſtreſſe doth the more greue me, and
the hand of our Lord is come forth againſt me. \V Therfore lifting vp
their voice, they beganne to weepe agayne, Orpha kiſſed her mother in
law, and returned: Ruth cleaued to her mother in law, \V to whom Noemi
ſaid: Behold thy kinſewoman is returned to her people, and
\SNote{Noemi perſwaded not to idolatrie, but inſinuated that if Ruth
would not returne to her countrie, ſhe muſt alſo leaue the falſe
goddes. And ſo ſhe anſwered, that ſhe would ſerue the ſame true God of
Iſrael.}
to her goddes, goe with her. \V Who anſwered: Be not againſt me, to the
end that I should leaue thee and depart: for whitherſoeuer thou shalt
goe, I wil goe: and where thou ſhalt abide, I alſo wil abide. Thy people
my people, and thy God my God. \V The land that ſhal receiue thee dying,
in the ſame wil I die: and there wil I take a place for my burial. Theſe
thinges doe God to me, & theſe thinges adde he, if death onlie ſhal not
ſeparate me and thee. \V Noemi therfore ſeing, that Ruth with a ſtid
faſt mind had determined to goe forward with her, would not be againſt
it, nor perſwade her any more to returne to her frendes: \V and they
went forth together, and came into Bethlehem. Who being entered into the
citie, a brute was quickly ſpred among them: and the wemen ſaid: This is
that Noemi. \V To whom ſhe ſaid: Cal me not Noemi (that is to ſay,
beautiful) but cal me Mara (that is to ſay, bitter) becauſe with
bitternes hath the Almightie very much repleniſhed me. \V I went forth
\SNote{She had a huſband and two ſonnes, and ſufficient prouiſion, but
now was bereued of them al.}
ful, and our Lord hath brought be backe emptie. Why therfore doe you cal
me Noemi whom our Lord hath humbled, and the Almightie hath
afflicted? \V Noemi therfore came with Ruth the Moabite her daughter in
law, from the Land of her peregrination: and returned into Bethlehem,
when barley was firſt reaped.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
