%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Ruth
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 0586
%%% o-0527
\startChapter[
  title={Chapter 2}
  ]

\Summary{Ruth gathering eares of corne in Booz field, 8.~he kindly
  biddeth her tarie with his ſeruantes. 17.~At night she returneth
  carying good quantitie of corne, and part of the meate, which they
  gaue her, to her mother in law.}

And Elimelech her huſband had a coſin, a mightie man, and of great
riches, named Booz. \V And Ruth the Moabite ſaid to her mother in law:
If thou command, I wil goe into the field, and gather the eares of
corne, that shal eſcape the handes of the reapers, whereſoeuer I ſhal
find the
%%% 0587
grace of the father of the houſe fauorable towardes me. To whom ſhe
anſwered: Goe my daughter. \V She went therfore and gathered the eares
of corne after the backes of the reapers. And it chanced that the owner
of the ſame field was Booz, who was of the kinred of Elimelech. \V And
behold, he came out of Bethlehem, and ſaid to the reapers:
\SNote{The Church vſeth this ſalutation in the holie ſacrifice and other
diuine office.}
Our Lord be with you. Who anſwered him: Our Lord bleſſe thee. \V And
Booz ſaid to the yongman, that was ouerſeer of the reapers: Whoſe maide
is this? \V To whom he anſwered: This is that Moabite, which came with
Noemi, from the countrie of Moab, \V and she deſired that she might
gather the eares of corne that remayne, folowing the ſteppes of the
reapers: and from morning vntil now she ſtayeth in the field, and not ſo
much as for a very moment hath she returned home. \V And Booz ſaid to
Ruth: Heare me daughter, goe not into an other field to gather, neither
depart thou from this place: but ioyne thy ſelfe to my maides, \V and
where they haue reaped, folow. For I haue commanded my ſeruantes, that
no man moleſt thee: but if thou shalt thirſt alſo, goe to the fardels,
and drinke the waters, wherof the ſeruantes alſo doe drinke. \V Who
falling on her face and adoring vpon the ground, ſaid to him: Whence
cometh this to me, that I should find grace before thine eies, and that
thou wouldeſt voutſafe to know me a ſtrange woman? \V To whom he
anſwered: Al thinges haue beene told me, which thou haſt done to thy
mother in law after the death of thy huſband: and that thou haſt leift
thy parentes, and the land wherein thou waſt borne, and art come to a
people, which before thou kneweſt not. \V Our Lord
\SNote{Booz doubted not but reward was due to good workes.}
render vnto thee for thy worke, and God grant thou mayeſt receiue
\SNote{Yea a ful reward, anſwerable to Ruth pietie: VVhich muſt be
ſpiritual and eternal.}
a ful reward of our Lord the God of Iſrael, to whom thou art come, &
vnder whoſe winges thou art fled. \V Who ſaid: I haue found grace in
thine eies my lord, which haſt comforted me, and haſt ſpoken to the hart
of thy handmaide, which am not like to one of thy maides. \V And Booz
ſaid to her: When the houre ſhal come to eate, come hither, and eate
bread, and dippe thy morſel in the vinagre. She therfore ſate at the
ſide of the
%%% o-0528
reapers, and she heaped to her ſelfe polent, and did eate and was
filled, and tooke the leauinges. \V And from thence she aroſe, to gleane
the eares of corne after her maner. And Booz commanded his ſeruantes,
ſaying: Yea and if she wil reape with
%%% 0588
you, forbid her not: \V and of your owne handfuls alſo caſt forth of
purpoſe, and let them remaine, that she may gather them without
bashefulnes, and gathering let no man controwle her. \V She gleaned
therfore in the field vntil euening: & that which she had gathered
beating with a rodde & threſhing she found of barley as it were the
meaſure of an ephi, that is, three bushels. \V Which carying she
returned into the citie, and shewed to her mother in law: moreouer she
brought forth, and gaue her of the remaynes of her meate, wherwith she
had beene filled. \V And her mother in law ſaid to her: Where haſt thou
gathered to day, and where haſt thou wrought? bleſſed be he that hath
had mercie on thee. And she told her with whom she had wrought: and she
told the mans name, that he was called Booz. \V To whom Noemi anſwered:
Be he bleſſed of our Lord: becauſe the ſame grace, which he had shewed
to the liuing, he hath kept alſo to the dead. And agayne she ſaid: The
man is our nigh coſin. \V And Ruth, This alſo, quoth she, he commanded
me, that ſo long I should ioyne my ſelf to the reapers, til al the corne
were reaped. \V To whom her mother in law ſaid: It is better my
daughter, that thou goe forth with his maides to reape, leſt in an other
mans field ſome may reſiſt thee. \V She therfore ioyned her ſelf to the
maides of Booz: and ſo long reaped with them, til the barley and the
wheate were layd vp in the barnes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
