%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% James
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 2886
%%% o-2747
\startChapter[
  title={Chapter 03}
  ]

\Summary{Againſt proud Maiſters and Authours of Sects. 5.~Of the
  manifold ſinnes of the vnbridled tongue. 12.~The difference betwixt
  proud, contentious, and worldly wiſedom, and that wiſedom which is
  heauenly, peaceable, modeſt, and ſo-forth.}

Be
\CNote{\XRef{Mt.~23,~8.}}
yee not
\LNote{Many Maiſters.}{He
\MNote{Many Maiſters are many proud Sect-Maiſters.}
meaneth principally Sect-maiſters that make themſelues ſeueral
Ring-leaders in ſundry ſorts of new deuiſed doctrines: euery one
arrogating to himſelf to be Maiſter, and none ſo humble as to be a
ſcholer, either to God's Church and true Paſtours, or to other guides
and Authours of the ſaid Sects. So did Zuinglius diſdaine to be Luthers
ſcholer, and Caluin to be the follower of Zuinglius.}
many Maiſters, my Brethren, knowing that you receiue the greater
iudgement. \V For in many things we offend al.
\CNote{\XRef{Eccl.~14,~1.}
\XRef{19,~16.}}
If any man offend not in word; this is a perfect man. He is able alſo
with bridle to turne about the whole body. \V And if we put bits into the
mouths of horſes that they may obey vs, we turne about al their body
alſo. \V And behold, the ſhips, whereas they be great, and are driuen of
ſtrong winds: yet
%%% o-2748
are they turned about with a litle ſterne whither the violence of the
directour wil. \V So the tongue alſo is certes a litle member, and
\TNote{\G{μεγαλαυχεῖ}}
vanteth great things. Behold how
\Var{much}{litle}
fire what a great wood it kindleth? \V And the tongue, is fire, a whole
world of iniquitie. The tongue is ſet amõg our mẽbers, which defileth
the whole bodie, and inflameth the wheele of our natiuitie, inflamed of
hel. \V For al nature of beaſts & foules & ſerpẽts & of the reſt is
tamed & hath been tamed by the nature of mã. \V But the tongue no man
can tame, an vnquiet euil, ful of deadly poiſon. \V By it we bleſſe God
and the Father; & by it we curſe men which are made after the ſimilitude
of God. \V Out of the ſelf-ſame mouth proceedeth bleſſing &
curſing. Theſe things muſt not be ſo done, my Brethren. \V Doth the
fountaine giue forth out of one hole ſweet and ſoure water? \V Can, my
Brethren, the figge-tree yeald grapes, or the vine, figges? So neither
\Var{can the ſalt water yeald ſweet.}{can it yeald ſalt & ſweet water.}

\V Who  is wiſe and hath knowledge among you? Let him ſhew by good
conuerſation his working in mildneſſe of wiſedom. \V But if you haue
bitter zeale, and there be contentions in your harts; glorie not and be
not liers againſt the truth. \V For this is not
\SNote{The difference betwixt the humane wiſedom, ſpecially of
heretikes; & the wiſedom of the Catholike Church & her children.}
wiſedom deſcending from aboue: but earthly, ſenſual, diuelish. \V For
where zeale and contention is, there is inconſtancie, and euery peruerſe
worke. \V But the wiſedom that is from aboue, firſt certes is chaſt;
then peaceable, modeſt, ſuaſible, conſenting to the good, ful of mercie
and good fruits not iudging, without ſimulation. \V And the fruit of
iuſtice, in peace is ſowed, to them that make peace.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

