%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% James
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 2887
%%% o-2748
\startChapter[
  title={Chapter 04}
  ]

\Summary{By concupiſcence and loue of this world, we are made enemies to
  God: but we should rather humble vs to him, punishing our ſelues for
  our ſinnes. 11.~Againſt detraction & rash iudging. 13.~To remember
  alwaies the vncertaintie of our life.}

%%% o-2749
From whence are warres and contentions among you? Are they not hereof?
of you cõcupiſcences which warre in your members? \V You couet, & haue
not. You kil, and enuie; and can not obtaine. You contend and warre: and
you haue not: becauſe you aske not. \V You aske, and receiue not:
becauſe you aske amiſſe: that you may cõſume it on your
concupiſcences. \V Aduouterers, know you not that the
\CNote{1.~Io.~2,~15}
freindſhip of this world, is the enemie of God? Whoſoeuer therfore wil
be a freind of this world, is made an enemie of God. \V Or doe you
thinke that the Scripture ſaith in vaine:
\CNote{\XRef{Prou.~3,~35.}}
\Emph{To enuie doth the ſpirit couet which dwelleth in you?} \V And
\SNote{The boldnes of Heretikes adding here the word, \Emph{Scripture},
to the text thus, \Emph{And the Scripture giueth greater grace.}}
giueth greater grace. For the which cauſe it ſaith,
\CNote{\XRef{1.~Pet.~5,~5.}}
\Emph{God reſiſteth the proud and giueth grace to the humble.}

\V Be ſubiect therfore to God, but reſiſt the Diuel, and he wil fly from
you. \V
\SNote{Free-wil and man's owne endeauour neceſſarie in comming to God.}
Approch to God, & he wil approch to you. Cleanſe your hands, ye ſinners:
and
\LNote{Purifie your harts.}{Man
\MNote{Man's working with God's grace, is no derogatiõ thereunto.}
(we ſee here) maketh himſelf cleane and purgeth his owne hart. Which
derogateth nothing to the grace of God being the principal cauſe of the
ſame. Yet Proteſtants thinke we derogate from Chriſt's Paſſion, when we
attribute ſuch effects to our owne workes, or to other ſecundarie helps
and cauſes.}
purifie your harts, ye double of mind. \V Be miſerable, and mourne, &
weep: let your laughter be turned into mourning; and ioy, into
ſorrow. \V
\CNote{\XRef{1.~Pet.~5,~6.}}
Be humbled in the ſight of our Lord, and he wil exalt you. \V
\SNote{He forbiddeth detractiò, euil ſpeaking, ſlandering.}
\TNote{\G{Μὴ καταλαλεῖτε ἀλλήλων}}
Detract not one from another, my Brethren. He that detracteth from his
Brother, or he that iudgeth his Brother, detracteth from the Law, and
iudgeth the Law. But if thou iudge the Law, thou art not a doer of the
Law, but a Iudge. \V For there is one Law-maker, and Iudge that can
deſtroy and deliuer. \V But thou,
\CNote{\XRef{Ro.~14,~4.}}
what art thou that iudgeſt thy neighbour?

Behold now you that ſay, To day or to morrow we wil goe into that citie,
and there certes wil ſpend a yeare, and wil traffike, and make our gaine
(\V who are ignorant what ſhal be on the morrow. For what is your life?
It is a vapour appearing for a litle while, and afterward it ſhal vanish
away) \V for that you ſhould ſay,
\SNote{Al promiſes and purpoſes of our worldly affaires are to be made
vnder condition of God's good liking & pleaſure: and it be cõmeth a
Chriſtiã man to haue vſually this forme of ſpeach in that caſe, \Emph{If
God wil, if God otherwiſe diſpoſe not.}}
If our Lord wil: and, if
%%% 2888
we ſhal liue, we wil doe this or that. \V But now you reioyce in your
arrogancies. Al ſuch reioycing is wicked. \V To one therfore knowing to
doe good, and not doing it: to him it is ſinne.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

