%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Romans
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 2667
%%% o-2519
\startChapter[
  title={Chapter 15}
  ]

\Summary{He proceedeth to make peace between the Chriſtian Gentils and
  Iewes, 8.~with this reſolution, that the Iewes vocation is of promiſe
  indeed, but the Gentils alſo of mercie & foretold by the
  Scriptures. 14.~Then drawing to an end, he excuſeth himſelf to the
  Romanes for writing thus vnto them, 21.~hoping now at length to ſee
  them, after that he hath been at Hieruſalem, 29.~whervnto alſo he
  requeſteth their praires.}

And we that are the ſtronger, muſt ſuſtaine the infirmities of the weak, &
not pleaſe our ſelues. \V Let euery one of you pleaſe his neighbour vnto
good, to edification. \V For, Chriſt did not pleaſe himſelf, but as it
is written:
\CNote{\XRef{Pſa.~68,~10.}}
\Emph{The reproches of them that reproched thee, fel vpon me.} \V For
\SNote{He meaneth al that is written in the old Teſtament: much more al
things written in the new Teſtamẽt, are for our learning & comfort.}
what things ſoeuer haue been written, to our learning they are written:
that by the patience and conſolation of the Scriptures, we may haue
hope. \V And the God of patience and of comfort giue you to be of one
mind one toward another according to \Sc{Iesvs Christ}: \V that of
\SNote{Vnitie in religion commended.}
one mind, with one mouth you may glorifie God & the Father of our
Lord \Sc{Iesvs Christ}. \V For the which cauſe receiue one another; as
Chriſt alſo hath receiued you vnto the honour of God. \V For I
ſay \Sc{Christ Iesvs} to haue been
\SNote{Chriſt did execute his office and miniſterie only toward the
people of Circumciſion, that is, the Iewes.}
miniſter of the circumciſion for the veritie of God to confirme the
promiſes of the Fathers. \V But the Gentils to honour God for his
mercie, as it is written:
\CNote{\XRef{Pſa.~17,~50.}}
\Emph{Therfore wil I confeſſe to thee in the Gentils, O Lord, and wil
ſing to thy name.} \V And againe he ſaid:
\CNote{\XRef{Deu.~32,~41.}}
\Emph{Reioyce ye Gentils with his people.} \V And againe:
\CNote{\XRef{Pſ.~116,~1.}}
\Emph{Praiſe al ye Gentils our Lord; and magnifie him al ye peoples.} \V
And againe Eſaie ſaith:
\CNote{\XRef{Eſa.~11,~10.}}
\Emph{There shal be the root of Ieſſe; and he that shal riſe vp to rule
the Gentils, in him the Gentils shal hope.} \V And the God of hope
repleniſh you with al ioy and peace in beleeuing; that you may abound in
hope, and in the vertue of the Holy Ghoſt.

\V And I my ſelf alſo, my Brethren, am aſſured of you, that you alſo are ful
of loue, repleniſhed with al knowledge, ſo that you are able to admoniſh
one another. \V But I haue written to you (Brethren) more boldly in
part, as it were putting you in remembrance; for the grace which is
giuen me of God, \V to be the miniſter of \Sc{Christ Iesvs} in the
Gentils: ſanctifying the Ghoſpel of God, that the oblation of the
Gentils may be made acceptable and ſanctified in the Holy
%%% o-2520
Ghoſt. \V I haue therfore glorie in \Sc{Christ Iesvs} toward God. \V For
I dare not ſpeake any of thoſe things which Chriſt worketh not by me for
the obedience of the Gentils, by word and deed, \V in the vertue of
ſignes and wonders, in the vertue of the Holy Ghoſt: ſo that from
Hieruſalem round about vnto Illyricum I haue repleniſhed the Ghoſpel of
Chriſt. \V And I haue ſo preached this Ghoſpel, not where Chriſt was
named, leſt I ſhould build vpon another man's foundation; \V but as it
is written:
\CNote{\XRef{Eſ.~52,~13.}}
\Emph{They to whom it hath not been preached of him, shal ſee, and they
that haue not heard, shal vnderſtand.} \V For the which cauſe alſo I was
hindred very much from comming vnto you. \V But now hauing no longer
place in theſe countries, and hauing a deſire to come vnto you theſe
%%% 2668
many yeares now paſſed: \V when I ſhal begin to take my iourney into
Spaine, I hope that as I paſſe, I ſhal ſee you, and be brought thither
of you, if firſt in part I ſhal haue enioyed you. \V Now therfore I wil
goe vnto Hieruſalem to miniſter to the
\SNote{He meaneth the holy perſons that hauing forſaken al their goods
for Chriſt, were wholy conuerted to ſerue our Lord with al their mind.
\Cite{S.~Hiero. againſt Vigilantius.}
the Heretike reprehẽding the almes-giuen to ſuch, as doe the Heretikes
alſo of our time.}
Saints. \V For Macedonia and Achaia haue liked wel to make ſome
contribution vpon the poore Saints that are in Hieruſalem. \V For it
hath pleaſed them: and they are their debters. For if the Gentils be
made partakers of their ſpiritual things; they ought alſo in carnal
things to miniſter vnto them. \V This therfore when I ſhal haue
accompliſhed, and ſigned them this fruit, I wil goe by you into
Spaine. \V And I know that comming to you, I ſhal come in aboundance of
the bleſſing of Chriſt. \V I beſeech you therfore Brethren by our
Lord \Sc{Iesvs Christ}, and by the charitie of the Holy Ghoſt, that you
\SNote{In that the Apoſtle deſired to be praied for we may be moued to
ſeeke the ſame as a great benefit.}
help me in your praiers for me to God, \V that I may be deliuered from
the Infidels that are in Iewrie, and the oblation of my ſeruice may
become acceptable in Hieruſalem to the Saints, \V that I may come to you
in ioy by the wil of God, that I may be refreſhed with you. \V And the
God of peace be with you al. Amen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

