%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Romans
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-11


\project douay-rheims


%%% 2658
%%% o-2509
\startChapter[
  title={Chapter 11}
  ]

\Summary{Not al the Iewes were reprobate, but ſome elect: and they by
  grace obtained iuſtice, the reſt (according to the Prophets) being
  execrated. 11.~Againſt whom notwithſtanding the Chriſtian Gentils (to
  whom by that occaſion Chriſt is come) muſt not inſult; but rather
  feare euery man himſelf to be likewiſe cut of the tree (which is the
  Catholike Church) 25.~and know that when al the Gentils are brought
  into the Church, then (about the end of the world) shal the multitude
  of the Iewes alſo come in: 33.~according to the diſpoſition of the
  wonderful wiſedom of God.}

I ſay then: Hath God reiected his people? God forbid. For I alſo am an
Iſraelite, of the ſeed of Abraham, of the Tribe of Ben-iamin. \V God
hath not reiected his people which he foreknew. Or know you not in Elias
what the Scripture ſaith; how he requeſteth God againſt Iſrael? \V
\CNote{\XRef{3.~Reg.~19,~10.}}
\Emph{Lord, they haue ſlaine thy Prophets, they haue digged downe thine
Altares: and I am left alone, & they ſeeke my life.} \V But what ſaith
the diuine
%%% o-2510
anſwer vnto him? \Emph{I haue left me
\LNote{Seuen thouſand.}{The
\MNote{God's anſwer to Elias of 7000, maketh nothing for the Proteſtants
inuiſible Church.}
Heretikes alleage this place and example very impertinently to proue
that the Church may be wholy ſecret, hid, or vnknowen. For though the
faithful were forced to keep cloſe in that perſecution of Achab and
Iezebel, which was only in the Kingdom of the ten Tribes, that is, of
Iſrael; yet at the very ſame time, in Hieruſalem and al the Kingdom of
Iuda, the external worship and profeſsion of faith was open to al the
world, and wel knowen to Elias & the faithful, ſo many, that
\CNote{\XRef{2.~Par.~17.}}
the very ſouldiars only were numbred aboue ten hundred thouſand. Beſides
that there is a great differẽce between the Chriſtian Church & the
Iewes; ours reſting vpon better promiſes then theirs. And we wil not put
the Proteſtãts to proue that there were 7000 of their Sect when their
new Elias Luther began; but let thẽ proue that there were ſeuen, or any
one, either then or in al Ages before him, that was in al points of his
beleefe. Heretikes there were before him, as Iouinian, Vigilantius,
Heluidius, Wicleffe &c. and with him, Zuinglius, Caluin, &c. who
beleeued as he did in ſome things, but not in al.}
ſeuen thouſand men, that haue not bowed their knees to
\SNote{The Heretikes adde here alſo to the text, \Emph{Image}, as
\XRef{Act.~19,~35.}}
Baal}? \V So therfore at this time alſo, there are remaines ſaued
according to the election of grace. \V And if by grace,
\LNote{Not now of workes.}{If
\MNote{What workes are not, & what are the cauſe of ſaluation.}
ſaluation be attributed to good workes done of nature without faith &
God's help, the ſame can not be of grace. For ſuch workes exclude grace,
fauour, and mercie: and chalenge only of debt, and not of
guift. Therfore take heed here of the Heretikes expoſition, that
vntruely exclude Chriſtian mens workes from neceſsity or merit of
Saluation, which are done with and by God's grace, and therfore
euidently conſiſt with the ſame, and be ioyned with God's grace as
cauſes of our ſaluation. Our Aduerſaries are like il Potecaries, euer
taking \L{quid pro quo}, either of ignorance, or of intent to deceiue
the ſimple.}
not now of workes. Otherwiſe grace now is not grace.

\V What then? that which Iſrael ſought, the ſame he hath not obtained;
but the election hath obtained, and the reſt were blinded: \V as it is
written:
\CNote{\XRef{Eſ.~6,~9.}}
\LNote{God hath giuen.}{It
\MNote{God is not Authour of ſinne.}
doth not ſignifie his working or action, but his permiſsion.
\Cite{Chryſ. ho.~19. in ep.~Ro.}
And
\CNote{\Cite{Aug. Ep.~105. ad Sixtum.}}
S.~Auguſtin ſaith, not by putting malice into them, but by not imparting
his grace vnto them, and that through their owne deſerts alwaies, and
their owne willes euer properly working the ſame. See
\XRef{Annot. Mat. c.~13,~14.}
\XRef{Ro. c.~3,~24.}}
\Emph{God hath giuen them the ſpirit of compunction; eyes, that they may
not ſee, and eares, that they may not heare; vntil this preſent day.} \V
And Dauid ſaith:
\CNote{\XRef{Pſ.~68,~23.}}
\Emph{Be their table made for a ſnare and for a trap & for a ſcandal &
for a retribution vnto them. \V Be their eyes darkned, that they may not
ſee: & their backe make thou alwaies crooked.} \V I ſay then, haue they
ſo ſtumbled,
\SNote{The Iewes are not reiected wholy & incurably for euer: but for a
part, and for a time ſuffred to fal. Which God did turne to the Gentils
general good.}
that they ſhould fal? God forbid. But by their offence, ſaluation is to
the Gentils, that they may emulate them. \V And if the offence of them
be the riches of the world, and the diminution of them
\SNote{If God could and did turne their fal and ſinne into the good of
the Gentils, much more wil he worke good of their general conuerſion,
which shal be at length the accomplishment of the Church conſiſting of
both the Nations.}
the riches of the Gentils; how much more the fulneſſe of them?

\V For to you Gentils I ſay, as long verily as I am the Apoſtle of the
Gẽtils, I wil honour my miniſterie, \V if by any meanes I may prouoke my
fleſh to emulation, and may ſaue ſome of them. \V For if the loſſe of
them be the reconciliation of the world; what ſhal the receiuing be, but
life from the dead? \V And if the firſt fruit be holy, the maſſe alſo:
and if the root be holy, the boughes alſo. \V And if ſome
%%% 2659
of the boughes be broken, and thou whereas thou waſt a wild oliue, art
graſſed in them, and art made partaker of the root and of the fatneſſe
of the oliue, \V glorie not againſt the boughes. And if thou glorie; not
thou beareſt the root, but the root thee. \V Thou ſaiſt them: The
boughes were broken, that I might be graſſed in. \V Wel:
\LNote{Becauſe of incredulity.}{He
\MNote{A paraphraſtical explicatiõ of the text, concerning the Iewes and
Gentils, their ſtãding, falling, riſing againe, &c.}
repreſſeth the pride of the Gentils vanting themſelues of their
receiuing, and of the Iewes reiection, namely in that they thought the
Iewes to be forſaken for no other cauſe, but that they might come into
their roomes: declaring that the direct and proper cauſe of their
forſaking, was their incredulity, exhorting the Gentils to beware of the
ſame, becauſe they may fal as wel as the other, and that God is as like
to execute iuſtice againſt them as againſt the Iewes; as he hath done in
many Nations falling to hereſie.}
becauſe of incredulitie they were broken, but thou by faith doeſt ſtand:
be not too highly wiſe; but
\SNote{We ſee that he which ſtandeth by faith, may fal from it, and
therfore muſt liue in feare, and not in the vaine preſumption and
ſecuritie of the Heretikes.}
feare. \V For if God hath not ſpared the natural boughes; leſt perhaps
he wil not ſpare thee neither. \V See then the goodnes and the ſeueritie
of God: vpon them ſurely that are fallen, the ſeueritie; but vpon thee
the goodnes of God, if thou abide in his goodnes; otherwiſe thou alſo
ſhalt be cut off. \V But they alſo, if they doe not abide in
incredulitie, ſhal be graſſed in. For God is able to graſſe them in
againe. \V For if thou waſt cut out of the natural wild oliue, and contrarie
to nature waſt graſſed into the good oliue; how much more they that are
according to
%%% o-2511
nature ſhal be graſſed into their owne oliue? \V For I wil
not haue you ignorant, Brethren, of this myſterie (that you be not wiſe
in your ſelues) that blindnes in part hath chanced in Iſrael, vntil the
fulnes of the Gentils might enter: \V and ſo al Iſrael, might be ſaued,
as it is written:
\CNote{\XRef{Eſa.~56,~20.}}
\Emph{There shal come out of Sion, he that shal deliuer, and shal auert
impietie from Iacob. \V And this to them the Teſtament from me}: when I
ſhal haue taken away their ſinnes. \V
\LNote{According to the Ghoſpel.}{In reſpect, or, as concerning beleefe
in Chriſt and receiuing the Ghoſpel, they are God's enemies: by occaſion
of which their incredulity, the Gentils found mercie: otherwiſe in
reſpect of his ſpecial election of that Nation, and the promiſes made to
the Patriarches, the Iewes are deere to him ſtil. For God neuer promiſeth
but he performeth, nor repenteth himſelf of the priuiledges giuen to
that Nation.}
According to the Ghoſpel indeed enemies for you: but according to the
election, moſt deere for the Fathers. \V For without repentance are the
guifts & the vocation of God. \V For
\LNote{For as you.}{As the Gentils which before beleeued not, found
mercie and came to faith, when the Iewes did fal: ſo the Iewes not now
beleeuing, when al the Gentils haue obtained mercie, shal in the end of
the world by God's diſpoſition obtaine grace and pardon as the Gentils
haue done.}
as you alſo ſometime did not beleeue God, but now haue obteined mercie
becauſe of their incredulitie; \V ſo theſe alſo now haue not beleeued,
for your mercie, that they alſo may obteine mercie. \V For God hath
\LNote{Concluded al.}{That ſo God taking al Nations and al men in ſinne
(which they fel into, not by his drift or cauſing, but of their owne
free-wil) may of his mercie cal & conuert whom and in what order he
wil; and the parties haue no cauſe to brag of their deſeruings: but both
Countries and particular men may referre their eternal election and
their firſt calling and conuerſion, to Chriſt, and to his mercie only:
no workes which they had before in their incredulity deſeruing any ſuch
thing, though their workes afterward proceeding of faith and grace doe
merit Heauen.}
concluded al into incredulitie, that he may haue mercie on al. \V
\LNote{O depth.}{The
\MNote{How farre to deale and to know, in the doctrine of
predeſtination.}
Apoſtle concludeth that no man ought to ſearch further into God's ſecret
and vnſearchable counſels of the vocation of the Gentils, & reiecting the
Iewes, otherwiſe then this; that al which be reiected, for their ſinnes
be iuſtly reiected, & al that be ſaued, by God's great mercie and
Chriſt's grace be ſaued. And whoſoeuer ſeeketh among the people to
ſpread contagion of curioſity by ſeeking further after things paſt man's
& Angel's reach, they ouer-reach and ouerthrew themſelues. If thou
wilt 
\Fix{the}{be}{likely typo, fixed in other}
ſaued, beleeue, obey the Church, feare God and keep his
commandements: that is thy part and euery man's els. Thou maiſt not
examine whether thou be predeſtinate or reprobate, nor ſeeke to know the
waies of God's ſecret iudgement toward thy ſelf or other men.
\MNote{The Heretikes writings of  predeſtination.}
It is the
common enemy of our ſoules, that in this vnhappy time hath opened
blaſphemous tongues, and directed the proud pennes of Caluin, Beza,
Verone, and ſuch reprobates, to the diſcuſsing of ſuch particulars, to
the perdition of many a ſimple man, and ſpecially of yong Scholers in
Vniuerſities, which with leſſe ſtudie may learne to be prowd and curious,
then to be humble, wiſe, and obedient.}
O depth of the riches of the wiſedom and of the knowledge of God! How
incomprehenſible are his iudgements, and his waies vnſearcheable? \V For
\CNote{\XRef{Eſa.~40,~13.}}
who hath knowen the mind of our Lord? or who hath been his Counſeler? \V
Or who hath firſt giuen to him, and retribution ſhal be made him? \V
For of him, and by him, and in him are al things: to him be glorie for
euer. Amen.
 

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

