%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 31
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-31


\project douay-rheims


%%% 0655
%%% o-0588
\startChapter[
  title={Chapter 31}
  ]

\Summary{Saul with his ſonnes are ſlaine in battel. 7.~The Philiſtijms
  poſseſse the place, and hang the dead bodies on a wal, 11.~but valiant
  men of Iabes Galaad take them away, and burne them, burie their bones,
  and faſt ſeuen dayes.}

And the Philiſtijms fought againſt Iſrael, and the men of Iſrael fled
before the face of the Philiſtijms, and fel being ſlaine in mount
Gelboe. \V And the Philiſtijms fel vpon Saul, and vpon his ſonnes, and
they ſtroke Ionathas, and Abinadab, and Melchiſua the ſonnes of Saul, \V
and the whole weight of the battel was turned vpon Saul: and the archers
ouertooke him, and he was ſore wounded of the archers. \V And Saul ſayd
to his eſquier: Drawe out thy ſword, and ſtrike me: leſt perhaps theſe
vncircumciſed come, and kil me, mocking me. And his eſquier would not:
for he was frighted with exceding feare.
\SNote{Saul killing himſelf after that he was wounded by his enimies
ſignifieth thoſe that being ouercome by tentations deſperatly perſiſt, &
wilfully die in their ſinne.
\Cite{S.~Greg. ho.~10.}}
Saul therefore caught his ſword, and fel
%%% 0656
vpon it. \V Which when his eſquier had ſeene, to wit that Saul was dead,
himſelfe alſo fel vpon his ſword and died with him. \V Saul therefore
died, and his three ſonnes, & his eſquier, and al his men in that day
together. \V And the children of Iſrael, that were beyond the valley,
and beyond Iordan, ſeing that the men of Iſrael were fled, and that Saul
was dead, and his ſonnes, they left their cities, and fled: and the
Philiſtijms came, and dwelt there. \V And when the next day was come,
the Philiſtijms came to ſpoile them that were ſlaine, and they found
Saul and his three ſonnes, lying in mount Gelboe. \V And they did cut of
Saules head, and ſpoiled him of his armour, and ſent into the land of
the Philiſtians round about, that it ſhould be declared in the temple of
their Idols, and among their people. \V And they did put his armour in
the temple of Aſtaroth, but his bodie they hung on the wal of
Bethſam. \V Which when
\SNote{Theſe men are cõmendable for \Emph{gratitude} towards Saul, who
had deliuered them.
\XRef{ch.~11.}
for a \Emph{vvorke of mercie} in burying the dead: for \Emph{pietie}
towards their king and princes, and for \Emph{fortitude} in atchiuing ſo
heroical an act.}
the inhabiters of Iabes Galaad had heard whatſoeuer the Philiſtijms had
done to Saul, \V al the moſt valiant men aroſe, & walked al the night,
and tooke the bodie of Saul, and the bodies of his ſonnes, from the wal
of Bethſam: and they came to Iabes Galaad, and burnt them there: \V and
they tooke their bones, and buried them in the wood of Iabes, and faſted
ſeuen daies.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
