%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 29
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-29


\project douay-rheims


%%% 0652
%%% o-0585
\startChapter[
  title={Chapter 29}
  ]

\Summary{Dauid going with the Philiſtijms towards the warre, 4.~the
  princes vrge and force the king to ſend him back.}

Therefore al the companies of the Philiſtijms were gathered together
into Aphec: and Iſrael alſo camped vpon the fountaine, which was in
Iezrahel. \V And the princes in dede of the Philiſtijms marched in
hundreds and thouſandes: but Dauid and his men were in the laſt companie
with Achis. \V And the princes of the Philiſtijms ſayd to Achis: What
meane theſe Hebrewes? And Achis ſayd to the princes of the Philiſtijms:
Doe you not knowe Dauid, which was the ſeruant of Saul the king of
Iſrael, and is with me manie daies, or
\SNote{He ſpeaketh by amplification to make his fact ſeme more
reaſonable: whereas the time of Dauids abode with him was but foure
monethes.
\XRef{ch.~27. v.~7.}}
yeares, and I haue not found any thing in him, ſince the day that he
fled to me, vntil this day? \V But the princes of the Philiſtijms were
angrie againſt him, and ſayd to him: Let this man returne, and abide in
his place, wherein thou haſt appointed him, and let him not goe downe
with vs into battel, leſt he become an aduerſarie to vs, when we ſhal
beginne to fight: for how can he otherwiſe pacifie his lord, but in our
heades? \V Is not this Dauid, to whom they ſang in dances, ſaying: Saul
hath ſtrooke his thouſandes, & Dauid his ten thouſandes? \V Achis
therefore called Dauid, and ſayd to him: The Lord liueth, thou art iuſt,
and good in my ſight: and thy going out, & thy coming in is with me in
the campe: and I haue not found in thee anie euil, ſince the day that
thou cameſt to me vntil this day: but thou pleaſeſt not the nobles.
%%% 0653
\V Returne therefore, and goe in peace, and offend not the eies of the
princes of the Philiſtijms. \V And Dauid ſayd to Achis: For what haue I
done, and what haſt thou found in me thy ſeruant, ſince the day that I
haue beene in thy ſight, vntil this day, that I may not come, and fight
againſt the enemies of my lord the King? \V And Achis anſwering ſpake to
Dauid: I know that thou art good in my ſight, as an Angel of God: but
the princes of the Philiſtijms haue ſayd: He shal not goe vp with vs
into battel. \V Therefore ariſe in the morning, thou, and the ſeruantes
of thy lord, which came with thee: and when you are riſen in the night,
and it shal beginne to waxe light, goe foreward. \V Dauid therefore aroſe
in the night, he and his men, that they might ſette forward in the
morning, and returne to the land of the Philiſtijms: and the Philiſtijms
went vp into Iezrahel.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
