%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 0602
%%% o-0541
\startChapter[
  title={Chapter 6}
  ]

\Summary{The Arke is ſent backe with fiue emeroids and fiue miſe of
  gold, vpon a new wayne drawne by two milch kyne. 13.~Which coming
  directly to Bethſames are ſacrificed, the wayne ſeruing for fire, the
  Leuites kepe the Arke. 19.~Many others are ſlaine looking of
  curioſitie into it.}

Therfore the arke of God was in the country of the Philiſthijms ſeuen
monethes. \V And the Philiſthijms called the prieſtes and ſoothſaiers,
ſaying: What ſhal we doe with the arke of the Lord? tel vs how we may
ſend it backe into his place. Who ſaid: \V If you ſend back the arke of
the God of Iſrael, ſent it not away emptie, but that which you owe
render vnto it for ſinne, and they you ſhal be cured: and you ſhal know
why his hand departeth not from you. \V Who anſwered: What is that which
we ought to render vnto it for ſinne? And they anſwered: \V According to
the number of the prouinces of the Philiſthijms you ſhal make fiue
golden emroides, and fiue golden miſe: becauſe there hath bene one
plague to you, and to your princes. And you shal make the ſimilitudes of
your emeroides, and the ſimilitudes of the miſe, that haue deſtroied the
land, and you ſhal geue glorie to the God of Iſrael: if perhaps he wil
lighten his hand from you, and from your goddes and from your land. \V
Why doe you harden your hartes, as
\SNote{Obſtinate ſinners doe harden their owne hartes, not God, but by
ſuffering them ſo to do. See
\XRef{Annot. Exod.~7.}}
Ægypt and Pharao did harden their hart? did not he after he was ſtriken
then dimiſſe them, and they departed? \V Now therfore take and make one
new wayne: and two kine hauing calued, on which there hath no yoke beene
put, couple in the wayne, and shut vp their calues at home. \V And you
%%% 0603
shal take the arke of the Lord, and put it in the wayne, and the veſſels
of gold, which you haue payed him for ſinne, you shal put into a litle
casket at the ſide thereof: and dimiſſe it that it may goe. \V And you
ſhal looke: and if ſo be that it shal goe vp by the way of his coaſtes
againſt Bethſames, he hath donne vs this great euil: but if not: we shal
know that his hand hath not touched vs, but it hath happened by
chance. \V They therfore did in this maner: and taking two kine, that
had ſucking calues, yoked them to the wayne, and shut vp their calues at
home. \V And they layd the arke of God vpon the wayne, and the litle
casket, that had the golden miſe and the ſimilitudes of emeroides. \V
And the kine went directly by the way, that leadeth to Bethſames, and
they went one way, going forward and lowing: and they declined not
neither to the right hand nor to the left: but the princes alſo of the
Philiſtijms folowed vnto the borders of Bethſames. \V Moreouer the
Bethſamites reaped wheat in the valley: and lifting vp their eies, they
ſaw the arke, and
%%% o-0542
were gladde when they had ſeene it. \V And the wayne came into the field
of Ioſue the Bethſamite, and ſtoode there. And there was a great ſtone,
and they did cut the wood of the wayne, and layed the kine vpon it an
holocauſt to our Lord. \V And the Leuites tooke downe the arke of God,
and the litle casket, that was at the ſide of it, wherin were the
veſſels of gold, and they put it vpon the great ſtone. The men alſo of
Bethſames offered holocauſtes, and immolated victimes that day to our
Lord. \V And the fiue princes of the Philiſtijms ſaw, and returned into
Accaron that day. \V And theſe are the golden emeroides, which the
Philiſtijms rendred for ſinne to our Lord: Azorus one, Gaza one, Aſcalon
one, Geth one, Accaron one: \V and the golden miſe according to the
number of the cities of the Philiſtijms, of the fiue prouinces, from
walled citie vnto towne that was without wal, and vnto Abel the great,
wherupon they put the arke of our Lord, which was vntil that day in the
field of Ioſue the Bethſamite. \V But he ſtroke of the men of Bethſames,
for that they had
\SNote{As the arke was terrible to the infideles,
\XRef{(chap.~5.)}
ſo alſo to thoſe that beleued right but vſed it not reuerently.}
ſeene the arke of our Lord: and he ſtroke of the people ſeuentie men,
and fiftie thouſand of the common people. And the people mourned,
becauſe our Lord had ſtriken the common people with a great plague. \V
And the men of Bethſames ſayd: Who ſhal be able to ſtand in the ſight of
%%% 0604
our Lord God this holie one? and to whom shal he goe vp from vs? \V And
they ſent meſſengers to the inhabitantes of Caria Thiarim, ſaying: The
Philiſtijms haue brought backe the arke of our Lord, come downe & fetch
it backe vnto you.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
