%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 0601
%%% o-0540
\startChapter[
  title={Chapter 5}
  ]

\Summary{Dagon falleth downe twiſe in preſence of the Arke, his head and
  handes broken of. 6.~The Philiſthijms being ſore plagued in al their
  cities where the arke cometh, 11.~determine to ſend it backe to the
  Iſraelites.}

And the Philiſtijms tooke the arke of God, and caried it from the Stone
of helpe into Azotus. \V And the Philiſtijms tooke the arke of God, and
brought it into the temple of Dagon, and ſette it beſide Dagon. \V And
when the Azotians had riſen early the next day, behold
\SNote{So ſowne as Chriſts Goſpel or Teſtament came among the Gentiles,
al falſe goddes & idolatry fel downe.
\Cite{S.~Beda. qq. in 1.~Reg. c.~3.}}
Dagon lay flatte on the ground before the arke of our Lord: and they
tooke Dagon, and reſtored him into his place. \V And agayne early the
next day riſing vp, they found Dagon lying vpon his face on the earth
before the arke of our Lord: and the head of Dagon, and the two palmes
of his handes were cutte of vpon the threshold: \V moreouer the bodie
only of Dagon was remayning in his place. For this cauſe the prieſtes of
Dagon, and al that enter into his temple, tread not vpon the threshold
of Dagon in Azotus vntil this day. \V And the hand of our Lord was
heauie vpon the Azotians, and he plagued them, and ſtroke Azotus and the
coaſtes thereof in the ſecrete part of the fundament. And the townes and
fieldes bubbled forth in the middes of that country, and there came
forth miſe, and there was confuſion of great death in the citie. \V And
the men of Azotus ſeing this maner of plague, ſaid: Let not the arke of
the God of Iſrael tarie with vs: becauſe his hand is ſore vpon vs, and
\SNote{The arke being a holie thing, as Reliques are, was terrible to
their falſe god, the diuel, ſo the Reliques of S.~Babilas ouerthrew the
falſe god Apollo, as S.~Chryſoſt. teſtifieth at large.
\Cite{li. cont. Gentiles, to.~5.}}
vpon Dagon our God. \V And ſending they gathered together al the princes
of the Philiſtijms to them, and ſaid: What shal we doe with the arke of
the God of Iſrael? And the Getheites anſwered: Let the arke of the God
of Iſrael be caried about, and they caried about the arke of the God of
Iſrael. \V And they carying it about, the hand of our Lord was made
through euerie citie by an exceding great ſlaughter: and it ſtrake the
men of euery city, from litle vnto great, & they had emeroides
%%% 0602
in their ſecrete partes. And the Getheites tooke counſeil, and made
themſelues ſtooles of skinnes. \V They ſent therfore the arke of God
into Accaron. And when the arke of God was come into Accaron, the
Accaronites cryed out, ſaying: They haue brought vnto vs the arke of the
God of Iſrael, to kil vs & our people. \V They ſent therfore & gathered
together al the princes of the Philiſtijms: who ſaid: Dimiſſe the arke
of the God of Iſrael, & let it returne into his place, & not kil vs with
our people. \V For there was made the feare of death in euery citie, &
the hand of God exceding greuous. The men alſo that had not died, were
ſtriken in the ſecrete part of the buttockes: and the howling of euery
citie went vp into heauen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
