%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 0598
%%% o-0537
\startChapter[
  title={Chapter 3}
  ]

\Summary{Samuel thriſe called vpon in ſleepe by viſion from God,
  repaireth to Heli. 10.~The fourth time our Lord reueleth to him the
  euil, that shal fall to Heli, and his houſe. 16.~Which he, being
  requeſted, declareth to Heli.}

And the childe Samuel miniſtred to our Lord before Heli, and the word of
our Lord was
\SNote{Rare thinges are called precious, and ſo the gift of prophecie is
here termed, which was then granted to few.}
precious in thoſe daies, there was no viſion manifeſt. \V It came to
paſſe therfore on a certayne day Heli lay in his place, and his eies
were become dimme, neither could he ſee. \V
\SNote{This viſion happened early in the morning, before the time of
dreſſing the lampes, when ſome were put out and others light.}
Before the lampe of God was extinguished, Samuel ſlept in the temple of
our Lord, where the arke of God was. \V And our Lord called Samuel. Who
anſwering, ſaid: Loe here I am. \V And he ranne to Heli and ſaid: Loe
here I am: for thou didſt cal me. Who ſaied: I did not cal thee: returne
and ſleepe. And he went and ſlept. \V And our Lord added againe to cal
Samuel. And Samuel ryſing vp went to Heli, and ſaid: Loe here I am:
becauſe thou didſt cal me. Who anſwered: I did not cal thee my ſonne:
returne and ſleepe. \V Moreouer Samuel did not yet know our Lord,
neither had the word of our Lord beene reueled to him. \V And our Lord
added, and called Samuel yet the third time. Who ryſing vp went to
Heli, \V and ſaid: Loe here I am: becauſe thou didſt cal me. Heli
therfore vnderſtood that our Lord called the childe, and ſaid to Samuel:
Goe, & ſleepe: & if he shal cal
\Fix{the}{thee}{obvious typo, fixed in other}
hereafter, thou shalt ſaie: Speake Lord, for thy ſeruant heareth. Samuel
therfore went & ſlept in his place. \V And our Lord came, and ſtoode:
and he called, as he had called twiſe: Samuel, Samuel. And Samuel ſayd: Speake
Lord for thy ſeruant heareth. \V And our Lord ſaid to Samuel:
%%% o-0538
Behold I doe a thing in Iſrael: which whoſoeuer shal heare, both his
eares shal tingle. \V In that day wil I rayſe vp againſt Heli al thinges
which I haue ſpoken touching his houſe: I wil beginne, and accomplish
it. \V For I haue foretold him that I would iudge his houſe for euer,
becauſe of iniquitie, for that he knewe that his ſonnes did wickedly, and
hath not corrected them. \V Therfore haue I ſworne to the houſe of Heli,
that the iniquitie of his houſe can not be expiated with victimes and
giftes for euer. \V And Samuel ſlept vntil morning, and opened the
doores of the houſe of our Lord. And Samuel feared to tel the viſion
vnto Heli. \V Heli therfore called Samuel, and ſaid: Samuel my ſonne:
Who
%%% 0599
anſwering, ſaid: Here I am. \V And he asked him: What is the word, that
our Lord hath ſpoken to thee? I beſech thee conceale it not from
me. Theſe thinges doe God to thee, and theſe doe he adde, if thou shalt
hide from me a word of al the wordes, which were ſaid to thee. \V Samuel
therfore told him al the wordes, & did not hide them from him. And he
anſwered: It is our Lord: let him doe that which is good in his eies. \V
And Samuel grewe, and our Lord was with him, and there fel not of his
wordes vpon the ground. \V And al Iſrael knewe from Dan to Berſabee,
that faithful Samuel was the prophet of our Lord. \V And our Lord added
to appeare in Silo, becauſe our Lord had bene reueled to Samuel in Silo,
according to the word of our Lord. And the word of Samuel came to paſſe
to al Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
