%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 0636
%%% o-0570
\startChapter[
  title={Chapter 21}
  ]

\Summary{In caſe of neceſsitie Achimelech the prieſt geueth halowed
  bread to Dauid, 8.~alſo the ſword which he had taken from
  Goliath. 10.~Then Dauid going to Achis king of Geth, is forced to faine
  himſelf madde.}

And Dauid came into Nobe to Achimelech the prieſt: & Achimelech was
aſtonyed, becauſe Dauid was come. And he ſaid to him: Why thou alone,
and none is with thee? \V And Dauid ſaid to Achimelech the prieſt: The
king hath commanded me a word and ſaid: Let no man know the thing, for
which thou art ſent by me, and what maner precepts I haue geuen thee,
for my ſeruantes alſo I haue appointed into ſuch and ſuch a place. \V
Now therfore if thou haue any thing at hand, yea if but fiue loaues,
geue me, or whatſoeuer thou shalt finde. \V And the prieſt anſwered
Dauid, ſaying: I haue no
\SNote{Diſtinction of common & holie bread. Alſo an example of
diſpenſation in caſe of neceſſitie.}
lay breads at hand, but only holy bread, if the ſeruants be cleane,
eſpecially from wemen? \V And Dauid anſwered the prieſt, and ſaid to
him: And truly, if the matter be concerning wemen, we haue refrained our
ſelues from yeſterday and the day before, when we came forth, and the
veſſels of the ſeruants were holie. Moreouer this way is polluted, but
it alſo shal be ſanctified this day in the veſſels. \V The prieſt therefore gaue
him halowed bread; for neither was anie bread there, but only the loaues
of propoſition, which had bene taken away from the face of our Lord,
that hoate loaues might be ſette downe. \V And there was there a
certaine man of the ſeruantes of Saul that day, within the tabernacle of
our Lord: and his name was Doeg an Idumeite, the mightieſt of Saules
paſtours. \V And Dauid ſaid to Achimelech: Haſt thou here at hand a
ſpeare, or a ſword? becauſe myn owne ſword, and myne owne weapons I tooke
not with me. For the kings word haſtened forward. \V And the prieſt
%%% 0637
ſaid: Loe here the ſword of Goliath the Philiſtian, whom thou ſleweſt in
the Valley of terebinth, is wrapped vp in a mantel behind the Ephod: if
thou wilt take this, take it. For neither is here any other beſide
that. And Dauid ſaid: There is none other like to that, geue me it. \V
Dauid therefore aroſe, and fled that day from the face of Saul: and came
to Achis the king of Geth, \V and the ſeruantes of Achis ſaid to him,
when they had ſeene Dauid: Is not this Dauid the king of the land? Did
they not ſing in dances to this man ſaying: Saul ſtroke a thouſand, and
Dauid ten thouſand? \V But Dauid
%%% o-0571
put theſe wordes in his hart, and feared excedingly at the face of Achis
the king of Geth. \V And he
\SNote{Dauid moſt wiſely in ſuch diſtreſſe fained himſelf to be a
foole. By which the Holie Ghoſt myſtically ſignified that Chriſt ſhould
do ſuch things not of feare but of diuine wiſdom, as he ſhould be
coũted a foole.
\XRef{Mar.~3. v.~21.}
\XRef{Luc.~23. v.~11.}
\XRef{1.~Cor.~1. v.~23.}
\Cite{S.~Beda. Queſt. in 1.~Reg. c.~14.}}
changed his countenance before them, and ſlipt downe betwen their
handes: and he ſtumbled at the doores of the gate, & his ſpittle ranne
downe vpon his bearde. \V And Achis ſaid to his ſeruantes: You haue
ſeene the man madde: why haue you brought him to me? \V Doe we lack
madde men, that you haue brought in this felowe, to play the madde man
in my preſence? ſhal this man enter into my houſe?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
