%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 0629
%%% o-0564
\startChapter[
  title={Chapter 18}
  ]

\Summary{Dauid and Ionathas enter league of frendship. 6.~Saul hearing
  Dauid praiſed aboue himſelf is offended: 10.~and vexed with an euil
  ſpirit, attempteth twoſe to kil him: 17.~promiſeth to geue him his
  eldeſt daughter in mariage, but geueth her to an other, 20.~and geueth
  him the yonger, thereby to ouerthrow him: 25.~putting him alſo in more
  danger, by requiring of him an hundred prepuces of
  Philiſtijms. 27.~Dauid bringeth him two hundred, and his fame
  encreaſeth.}

And it came to paſſe, when he had finished to ſpeake vnto Saul, the
ſoule of Ionathas was ioyned faſt to the ſoule of Dauid, & Ionathas
loued him as his ſoule. \V And Saul tooke him in that day, and did not
grant vnto him to returne into his fathers houſe. \V And Dauid and
Ionathas entered a league, for he loued him as his ſoule. \V For
Ionathas ſtripped himſelf of the cote wherwith he was clothed, and gaue
it to Dauid, and the reſt of his garments, vnto his ſword, & bowe, &
vnto his belt. \V Dauid alſo went forth to al thinges whereſoeuer Saul
ſent him, & he behaued himſelf wiſely: and Saul placed him ouer the men
of warre, and he was accepted in the eies of al the people, and
ſpecially in the eies of Saules ſeruantes. \V Moreouer
\SNote{This happened not immediatly after the victorie againſt Goliath
but when Dauid had dwelt ſome time in Sauls houſe, and was very gratful
to him and to al the court and people.}
when Dauid returned, after he ſtroke the Philiſtian, the wemen came
forth from al the tribes of Iſrael, ſinging and dancing to Saul the
King, in timbrels of ioy, and in cornettes. \V And the wemen ſang,
playing, and ſaying: Saul ſtroke a thouſand, and Dauid ten thouſand. \V
And Saul was
\SNote{VVhen the praiſes of one importe diminiſhing of an other, he that
is proud, as Saul now was, is moued to enuie and malice, more and more
as the vertues of the other and his praiſes do increaſe.}
exceding angrie, and this word was diſpleaſant in his eies: and he ſayd:
They haue geuen Dauid ten thouſand,
%%% 0630
%%% o-0565
and to me they haue geuen a thouſand: what remayneth for him but only
the kingdom? \V Therefore Saul did not looke vpon Dauid with right eies
from that day and afterward. \V And a day after, the euil ſpirit of God
inuaded Saul, and he prophecied in the middes of his houſe. And Dauid
played with his hand as euery day. And Saul held a ſpeare, \V and threw
it, thinking that he could naile Dauid to the wal, and Dauid declined
from his face the ſecond time. \V And Saul feared Dauid becauſe our Lord
was with him, and was departed from himſelf. \V Saul therefore remoued
him from him, and made him a tribune ouer a thouand men, and he went out
and came in before al the people. \V Alſo in al his wayes, Dauid delt
wiſely, and our Lord was with him. \V Saul therefore ſaw that he was
exceding wiſe, and he beganne to beware of him. \V But al Iſrael, and
Iuda loued Dauid, for he came in and went out before them. \V And Saul
ſayd to Dauid: Behold my elder daughter Merob, her wil I geue thee to
wife, only be thou a valyant man, and fight the battels of our Lord. And
Saul thought ſaying: Be not my hand vpon him, but let the handes of the
Philiſtians be vpon him. \V And Dauid ſayd to Saul: What am I, or what
is my life, or the kindred of my father in Iſrael, that I ſhould be made
the ſonne in lawe of the king? \V And it came to paſſe, at what time,
Merob the daughter of Saul ſhould haue beene geuen to Dauid, ſhe was
geuen to Hadriel the Molathite to wife. \V But Dauid loued Michol the
other daughter of Saul. And it was told Saul, and it pleaſed him. \V And
Saul ſayd: I wil geue her to him, that ſhe may be a ſcandal vnto
him, and that the hand of the Philiſtians may be vpon him. And Saul ſayd
to Dauid: In two thinges thou ſhalt be my ſonne in 
lawe this day. \V And Saul commanded his ſeruantes: Speake to Dauid
ſecretly out of my preſence, ſaying: Behold thou pleaſeſt the King, and
al his ſeruantes loue thee. Now therefore be thou the kinges ſonne in
lawe. \V And the ſeruantes of Saul ſpake al theſe wordes in the eares of
Dauid. And Dauid ſayd: Doth it ſeme vnto you a ſmal matter to be the
ſonne in lawe of a King? But I am a poore man, and of ſmal ability. \V
And the ſeruantes of Saul reported, ſaying: Theſe maner of wordes hath
Dauid ſpoken. \V And Saul ſayd: Speake thus to Dauid: The king nedeth no
dowrie, but only an hundred prepuces of the Philiſtians, that reuenge
may be made of the kinges
%%% 0631
enemies. Moreouer Saul thought to deliuer Dauid into the handes of the
Philiſtians. \V And when his ſeruantes had reported to Dauid the wordes
that Saul had ſayd, the word was liked in the eies of Dauid, to be made
the kings ſonne in lawe. \V And after fewe days Dauid riſing vp, went
with the men that were vnder him, and he ſtroke of the Philiſtijms two
hundred men, and brought their prepuces, and numbered them to the King,
that he might be his ſonne in law. Saul therefore gaue him Michol his
daughter to wife. \V And Saul ſaw, and vnderſtood that our Lord was with
Dauid. And Michol the daughter of Saul loued him. \V And Saul began more
to feare Dauid: and Saul became enemie to Dauid al daies. \V And the
princes of the Philiſtians
%%% o-0566
went forth: and from the beginning of their going forth, Dauid behaued
him ſelf more wiſely, then al the ſeruantes of Saul, and his name was
made renowmed excedingly.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
