%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 0610
%%% o-0548
\startChapter[
  title={Chapter 10}
  ]

\Summary{Saul is annointed king, and confirmed by ſignes that his
  ordinance is of God. 10.~He prophecieth, which the people doth
  admire. 17.~Samuel calleth the people together, for appointing a king,
  the lotte falleth on Saul, 25.~and the law of the king is againe
  mentioned.}

And Samuel tooke
\SNote{S.~Gregorie here noteth, that ſuch as are placed in height of
gouernement, are annointed with oyle, which ſignifieth mercie, light,
and curing of others.}
a litle veſſel of
\SNote{But the \Emph{litle} veſſel foreſhewed that Saul not perſeuering
in grace, ſhould be depoſed from his kingdom.
\Cite{li.~4. c.~5. in 1.~Reg.~10.}}
oyle, and powred vpon his head, and kiſſed him, and ſayd: Behold, our
Lord hath annointed thee vpon his inheritance to be prince, and thou
ſhalt deliuer his people out of the handes of their enemies, that are
round about them. And this ſhal be a ſigne vnto thee, that God hath
annointed thee to be prince. \V When thou ſhalt be departed from me this
day, thou ſhalt finde two men beſide the ſepulchre of Rachel in the
borders of Beniamin, in the South, and they ſhal ſay to thee: The aſſes
are found, which thou didſt goe to ſeeke: and thy father letting goe
the aſſes, is careful for you, and ſayeth: What ſhal I doe concerning
my ſonne? \V And when thou ſhalt depart thence, and paſſe farder, and
ſhalt come to the oke Thabor, three men going vp to God into Bethel ſhal
finde thee there, one carying three kiddes, and an other three
manchettes of bread, and an other carying a flagon of wine. \V And when
they haue ſaluted thee, they wil geue thee two loaues, and thou ſhalt
take them of their hand. \V After theſe thinges thou ſhalt come into the
hil of God, where the garriſon of the Philiſthiimes is: and when thou
shalt be entered there into the citie, thou shalt meete there a flocke of
prophetes coming downe from the excelſe, and before them: pſalterie and
tymbrel, and ſhalme, and harpe, and themſelues prophecying. \V And the
Spirit of our Lord shal ſeaſe vpon thee, and thou shalt prophecie with
them, and shalt be changed into an other man. \V Therfore when al theſe
ſignes shal chance to thee, doe whatſoeuer thy hand shal finde, becauſe
our Lord is with thee. \V And thou shalt goe downe before me into
Galgala (for I wil come downe to thee) that thou
%%% 0611
mayeſt offer oblation, and immolate pacifique victimes:
\SNote{Samuel enioyneth obedience to Saul to trie his humilitie.
\Cite{S.~Greg. li.~4. c.~5. in 1.~Reg.~10.}}
ſeuen daies shalt thou expect, til I come to thee, and I wil shew thee
what thou muſt doe. \V Therfore when he had turned away his shoulder to
depart from Samuel, God
\SNote{God gaue him peculiar grace for executing the office of a King.}
changed vnto
%%% o-0549
him another hart, and al theſe thinges came in that day. \V
And they came to the foreſaid hil, and behold a troupe of prophetes
meeting him: & the
\SNote{By and by alſo the gift of prophecie.}
Spirit of our Lord ſeated vpon him, and he prophecied in the middes of
them. \V And al that had knowen him yeſterday and the day before, ſeing
that he was with the prophetes, & did prophecie, ſaid to ech other: What
thing hath happened to the ſonne of Cis? what is Saul alſo among the
prophetes? \V And one anſwered an other, ſaying: And who is
\SNote{Their ſuperiour.}
their father? therefore it was turned into a prouerbe: What is Saul alſo
among the prophetes? \V And he ceaſed to prophecie, and came to the
excelſe. \V And Sauls vncle ſayd to him, and to his ſeruant: Whither
went you? who anſwered: To ſeeke the aſſes: which when we had not found,
we came to Samuel. \V And his vncle ſayd to him: Tel me what Samuel ſayd
to thee. \V And Saul ſayd to his vncle: He told vs that the aſſes were
found. But concerning the word of the kingdom which Samuel had ſpoken to
him, he told him not. \V And Samuel called together the people to our
Lord in Maſpha: \V and ſayd to the children of Iſrael: Thus ſayeth our
Lord the God of Iſrael: I brought Iſrael out of Ægypt, and deliuered you
from the hand of the Ægyptians, and from the hand of al the kinges which
afflicted you. \V But you this day haue reiected your God, who only hath
ſaued you out of al your euils and tribulations: and you haue ſaid: Not
ſo: but appoint a king ouer vs. Now therfore ſtand before our Lord by your
tribes, and by your families. \V And Samuel brought al the tribes of
Iſrael, and the
\SNote{By lotte the people were aſſured that the election was of God.
\Cite{S.~Greg. ibidem.}}
lotte fel on the tribe of Beniamin. \V And he brought the tribe of
Beniamin and the kinreds thereof, and it fel vpon the kindred of Metri,
and it came vnto Saul the ſonne of Cis. They therfore ſought him, and he
was not found. \V And after theſe thinges they conſulted our Lord
whether he would come thither. And our Lord anſwered: Behold he is hid
at home. \V They ranne therfore and tooke him from thence: and he ſtood
in the middes of the people, and he was higher then al the people from
the shoulder and vpward. \V And Samuel ſaid to al the people: Certes you
ſee whom our Lord hath choſen, that
%%% 0612
there is not the like to him in al the people. And al the people cried,
and ſayd: God ſaue the King. \V And Samuel ſpake to the people the law
of the kingdom, and wrote it in a booke, and layd it vp before our Lord:
and Samuel diſmiſſed al the people, euerie one into his owne houſe. \V
But Saul alſo departed vnto his houſe into Gabaa: and there went with
him part of the armie, they whoſe hartes God had touched. \V But the
children of Belial ſayd: What shal this fellow be able to ſaue vs? and
they deſpiſed him, and brought him not preſentes: but he diſſembled as
though he heard not.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
