%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 0599
%%% o-0538
\startChapter[
  title={Chapter 4}
  ]

\Summary{The Iſraelites are beaten in batle by the Philiſthijms. 3.~VVho
  for their better protection and comfort, fetch the Arke of God into
  the campe: 10.~but are beaten againe, the Arke taken, and with manie
  others the two ſonnes of Heli are ſlaine. 13.~Al which Heli
  vnderſtanding falleth from his ſeate, and breaketh his neck: 19.~alſo
  his daughter in law preſently traueling of childe is deliuered of a
  ſonne.}

And it came to paſſe in thoſe daies, the Philiſthijms aſſembled together
to fight: and Iſrael went forth to meete the Philiſthims into battle, &
camped beſide the Stone of helpe. Moreouer the Philiſthijms came into
Aphec, \V and put their armie in aray againſt Iſrael. And after they had
ioyned battle, Iſrael turned their backes to the Philiſthims: and there
were ſlaine in the fight here and there through the fieldes, as it were
foure thouſand men. \V And the people returned to the campe: and the
ancientes of Iſrael ſaid: Why hath our Lord ſtricken vs to day before
the Philiſthijms?
\SNote{Their confidence of helpe from God, by preſence of the arke was
good and commendable, but their ſinnes deſerued to be puniſhed.}
Let vs fetch vnto vs the arke of the couenant of our Lord from Silo, and
let it come into the middes of vs, that it may ſaue vs from
%%% o-0539
the hand of our enemies. \V The people therfore ſent into Silo, and they
tooke from thence the arke of the couenant of the Lord of hoſtes ſitting
vpon the Cherubims: and the two ſonnes of Heli were with the arke of the
couenant of God, Ophni and Phinees. \V And when the arke of the couenant
of our Lord was come into the campe, al Iſrael made a ſhoute with a
great crie, and the earth ſounded. \V And
%%% 0600
the Philiſthims heard the voice of the crie, and ſaid: What is this
voice of a great crie in the campe of the Hebrewes? And they knewe that
the arke of our Lord was come into the campe. \V And the Philiſthijms
were afrayd, ſaing: God is come into the campe. And they mourned,
ſaing: \V Woe to vs: for there was not ſo great reioyſing yeſterday and
the day before: woe to vs. Who ſhal keepe vs from the hand of theſe high
Goddes? theſe be the Goddes, that ſtricke Ægypt with al plague, in the
deſert. \V Take courage, and be men, ye Philiſthijms: leſt you be
ſeruantes to the Hebrewes, as they alſo haue ſerued you: take courage
and fight. \V The Philiſthijms therfore fought, and Iſrael was ſlaine,
and euerie man fled into his tabernacle: and there was made an exceding
great plague: and there fel of Iſrael thirtie thouſand footemen. \V And
the arke of God was taken: the two ſonnes alſo of Heli died, Ophni and
Phinees. \V And a man of Beniamin running out of the battle aray, came
into Silo that day; his garment rent, and ſprinkled on his head with
duſt. \V And when he was come, Heli ſate vpon a ſtoole ouer againſt the
way looking. For his hart was fearful for the arke of God. And that man
after he was entred in, told it to the citie: and al the citie
howled. \V And Heli heard the ſound of the crie, and ſaid: What is this
ſound of this ſame tumult? But he haſtened, and came, and told Heli. \V
And Heli was nintie and eight yeares old, and his eyes were dimme, and
he could not ſee. \V And he ſaid to Heli: I am he that came from the
battle, and I he that fled out of the field this day. To whom he ſaid:
What is done my ſonne? \V And he brought the newes anſwering: Iſrael,
quoth he, is fled before the Philiſthijms, and a great ruine is made in
the people: moreouer alſo thy two ſonnes are dead, Ophni and Phinees:
and the arke of God is taken. \V And when he had
\SNote{This zele of religion in Heli towards the arke, is a great ſigne
that he died in good ſtate though he was temporally puniſhed for not
correcting his ſonnes.}
named the arke of God, he fel from his ſtoole backward beſide the doore,
& his necke being broken he died. For he was an old man, and of a great
age: and he iudged Iſrael fourtie yeares. \V And his daughter in law,
the wife of Phinees was great with childe, and nigh to be deliuered: and
hearing the reporte that the arke of God was taken, and her father in
law was dead, and her huſband, she bowed her ſelf and was deliuered: for
ſudden paynes were fallen vpon her. \V And in the very moment of her
death, they ſaid to her that ſtoode about her: Feare not
%%% 0601
becauſe thou haſt borne a ſonne. Who anſwered them not, nor gaue heede
to it. \V And she called the childe Ichabod, ſaing: The glorie is
tranſlated from Iſrael, becauſe the arke of God is taken, and for her
father in law, and for her huſband; \V and she ſaid: The glorie is
tranſlated from Iſrael, for that the arke of God was taken.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
