%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 0612
%%% o-0550
\startChapter[
  title={Chapter 11}
  ]

\Summary{Ammonites fighting againſt Iabes Galaad, and the citie readie
  to yelde, 5.~Saul gathereth an armie, 11.~ouerthroweth the enimie,
  14.~and is eſtablished King.}

And it came to paſſe as it were a moneth after, Naas the Ammonite
aſcended, and began to fight againſt Iabes of Galaad. And al the men of
Iabes ſayd to Naas: Make a league with vs, and we wil ſerue thee. \V And
Naas the Ammonite anſwered them: In this wil I make a league with you,
that I may plucke out the right eyes of you al, and may make you a
reproch in al Iſrael. \V And the ancientes of Iabes ſayd to him: Graunt
vnto vs ſeuen daies, that we may ſend meſſengers vnto al the coaſtes of
Iſrael: and if there shal not be that may defend vs, we wil come forth
to thee. \V The meſſengers therefore came into Gabaa of Saul: and they
ſpake theſe wordes, in the hearing of the people: and al the people
lifted vp their voice, and wept. \V And behold Saul came, folowing oxen
out of the field, and ſayd: What ayleth the people that they weepe? And
they told him the wordes of the men of Iabes. \V And the Spirit of our
Lord ſeaſed on Saul, when he had heard theſe wordes, and his furie was
exceding wrath. \V And taking both the oxen, he cutte them into peeces,
and ſent them into al the coaſtes of Iſrael by meſſengers, ſaying:
Whoſoeuer shal not goe forth, and folow Saul and Samuel, ſo shal it be
done to his oxen. Therefore the feare of our Lord inuaded the people,
and they went forth as it were one man. \V And he numbered them in
Bezec: and there were of the children of Iſrael three hundred thouſand:
and of the men of Iuda thirtie thouſand. \V And they ſayd to the
meſſengers that came: Thus shal you ſay to the
%%% 0613
men, that are in Iabes Galaad: To morow, when the ſunne ſhal waxe hote,
you ſhal haue relife. The meſſengers therefore came, and told the men of
Iabes: Who were glad. \V And they ſaid: In the morning
\SNote{As the men of Iabes deluded their enimies by equiuocation, ſo
ſpeaking that they were otherwiſe vnderſtood then they ment: ſo the
ſeruantes of God being tempted with concupiſcence of glutonie (ſignified
by \Emph{Naas}) muſt deceiue their carnal appetite, by promiſing to
ſatisfie the deſire of the fleſh, but in deede kepe ſuch temperance, as
they may kil the cõcupiſcence, and not be killed by it.
\Cite{S.~Greg. li.~5. c.~1. in 1.~Reg.~11.}}
we wil come forth to you: and you ſhal doe to vs whatſoeuer ſhal pleaſe
you. \V And it came to paſſe, when the morow was come, Saul ſette the
people into three partes: and entered into the middes of the campe in
the morning watch, and ſtroke Ammon vntil the day waxed hote, and the
reſt were diſperſed, ſo that there were not left among them two
together. \V And the people ſayd to Samuel: Who is this that ſaid: What
ſhal Saul reigne ouer vs? Geue vs the men and we wil kil them. \V And
Saul ſayd: No man ſhal be killed this day, becauſe our Lord this day
hath releued Iſrael. \V And Samuel ſaid to the people: Come and let vs
goe into Galgal, and let vs renewe there a kingdome. \V And al the
people went into Galgal, and there they made Saul
%%% o-0551
king before our Lord in Galgal, & they immolated there pacifique
victimes before our Lord. And Saul reioyſed there, and al the men of
Iſrael excedingly.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
