%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 0615
%%% o-0552
\startChapter[
  title={Chapter 13}
  ]

\Summary{Saul and Ionathas preuaile in battel againſt the
  Philiſthijms. 5.~Who increaſing their forces, the Iſraelites for feare
  flee away and hid themſelues. 8.~Samuel not coming to the campe, Saul
  preſumeth to offer ſacrifice, 11.~for which Samuel reproueth him, and
  declareth that his kingdom shal be tranſlated to an other. 17.~The
  Philiſthijms oppreſſe the Iſraelites, and depriue them of armour.}

A child of
\SNote{Saul beginning to reigne was innocent and humble as a child of
one yeare.}
one yeare was Saul when he began to reigne, and
\SNote{And in that ſtate reigned the firſt two yeares.
\Cite{S.~Greg. in hunc locum.}}
two yeares he reigned ouer Iſrael. \V And Saul choſe to him ſelfe three
thouſand of Iſrael: and there were with Saul two thouſand in Machmas, and
in the mount of Bethel: and a thouſand with Ionathas in Gabaa of
Beniamin, moreouer the reſt of the people he ſent backe euerie
%%% o-0553
man into their tabernacles. \V And Ionathas ſtroke the garriſon of the
Philiſthijms, which was in Gabaa. Which when the Philiſthijms had heard,
Saul ſounded with the trumpet in al the land, ſaying: Let the Hebrewes
heare. \V And al Iſrael heard this maner of bruite: Saul hath ſtriken
the garriſon of the Philiſthijms: and Iſrael tooke courage againſt the
Philiſthijms. The people therfore cried after Saul in Galgal. \V And the
Philiſthims were gathered together to fight againſt Iſrael, thirtie
thouſand chariotes, and ſix thouſand horſemen, and the reſt of the
common people, as the land which is in the ſea shore very much. And
going vp they camped in Machmas at the Eaſt of Bethauen. \V Which when
the men of Iſrael had ſeene them ſelues put in a ſtreict (for the people
was afflicted) they hid them ſelues in caues, and in ſecrete places, in
rockes alſo, and in dennes, and in ceſternes. \V And the
%%% 0616
Hebrewes paſſed Iordan into the Land of Gad and Galaad. And when Saul
was yet in Galgal, al the people was ſore afrayd, which folowed him. \V
And he expected ſeuen daies according to the appointment of Samuel, and
Samuel came not into Galgal, and the people ſlipt away from him. \V Saul
therfore ſaid: Bring me the holocauſte, and the pacifiques. And he
offered the holocauſte. \V And when he had finiſhed offering the
holocauſte, behold Samuel came: and Saul went forth to mete him & ſalute
him. \V And Samuel ſpake to him: What haſt thou done? Saul anſwered:
Becauſe I ſawe that the people ſlipt from me, and thou waſt not come
according to the dayes appointed, moreouer the Philiſthijms were
gathered together into Machmas, \V I ſaid: Now wil the Philiſthijms come
downe to me into Galgal, & I haue not pacified the face of our
Lord. Compelled by neceſſitie, I offered the holocauſte. \V And Samuel
ſaid to Saul: Thou haſt
\SNote{He offended in offering ſacrifice being neither a prieſt, nor
extraordinarily allowed to do that office, and for this and other
faultes was depoſed.}
done foliſhly, neither haſt thou kept the commandementes of our Lord thy
God, which he commanded thee. Which
\SNote{Gods foreſight of ſinne, and perordination to puniſh it, taketh
not away freewil, nor poſſibilitie of wel doing, nor of reward.
\Cite{S.~Aug. li.~3. c.~4. de lib. arbit.}}
if thou hadſt not done, euen now had our Lord prepared thy kingdom ouer
Iſrael for euer, \V but thy kingdom shal no farder ariſe. Our Lord hath
ſought him a man according to his hart: and him hath our Lord commanded
to be prince ouer his people, becauſe thou haſt not obſerued the thinges
which our Lord commanded. \V And Samuel aroſe and went vp from Galgal
into Gabaa of Beniamin. And numbered the people, which were found with
him, as it were ſix hundred men. \V And Saul and Ionathas his ſonne, and
the people that were found with them, was in Gabaa of Beniamin: moreouer
the Philiſthijms had pitched in Machmas. \V And there iſſued forth to
praye from the campe of the Philiſthians three companies. One companie
went on againſt the way of Ephra to the Land of Saul. \V Moreouer an
other went by the way of Bethoron, & the third had turned it ſelf to the
way of the border, in the valley Seboim againſt the deſert. \V Moreouer
there was not found an yron ſmith in al the Land of Iſrael. For the
Philiſthijms had ſo prouided, leſt perhaps the Hebrewes should make
ſword or ſpeare. \V Al Iſrael therefore went downe to the Philiſthijms,
that euerie man might
%%% o-0554
whette his plough culter, and ſpade, & axe, and rake. \V Therefore the
edges of the shares, and ſpades, & forkes with three teeth, and axes,
were blunt, euen to the godeprick, which
%%% 0617
was to be mended. \V And when the day was come to fight, there was not
found ſword and ſpeare in the hand of al the people, that was with Saul
and Ionathas, except Saul and Ionathas his ſonne. \V And the ſtation of
the Philiſtijms went forth, to paſſe vp into Machmas.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
