%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 26
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-26


\project douay-rheims


%%% 0646
%%% o-0579
\startChapter[
  title={Chapter 26}
  ]

\Summary{The Ziphians bewtraying the place, Saul beſiegeth Dauid. 5.~Who
  by night goeth where Saul and his men are a ſleepe, 9.~hurteth him
  not: 12.~but taketh from him his ſpeare and bottle of water:
  14.~sheweth what he hath done. 21.~Saul againe confeſſeth his fault,
  and promiſeth peace.}

And there came Zepheites vnto Saul in Gabaa, ſaying: Behold Dauid is hid
in the hil Hachila, which is ouer againſt the wildernes. \V And Saul
aroſe, and went downe into the deſert Ziph, and with him three thouſand
men of the choſen of Iſrael, to ſeeke Dauid in the deſert Ziph. \V And
Saul camped in Gabaa Hachila, which was ouer againſt the wildernes in
the way. And Dauid dwelt in the deſert. And ſeing that Saul was come
after him into the deſert, \V he ſent diſcouerers, and lerned that he
was come thither moſt certainly. \V And Dauid aroſe ſecretly, and came
to the place where Saul was: and when he had ſeene the place, wherein
Saul ſlept, and Abner the ſonne of Ner, the prince of his warre, and
Saul ſleeping in the tent, and the reſt of the multitude round about
him, \V Dauid ſpake to Achimelech the Hetheit, and Abiſai the ſonne of
Seruia the brother of Ioab, ſaying: Who wil goe downe with me to Saul
into the campe? And Abiſai ſaid: I wil goe with thee. \V Dauid therefore
and Abiſai came to the people by night, and found Saul lying and
ſleeping in the tent, and his ſpeare fixed in the ground at his head:
and Abner, and the people ſleeping round about him. \V And Abiſai ſaid
to Dauid: God hath shut vp thine enemie this day into thy handes: now
therefore I wil thruſt
%%% 0647
him through with my ſpeare in the earth once, and twiſe shal not
neede. \V And Dauid ſaid to Abiſai: Kil him not: for
\SNote{Dauid is reſolute, and often repeteth, that it is not lawful for
priuate ſubiectes to kil their prince, no although him ſelfe was
annointed to ſuccede.}
who ſhal extend his hand vpon the annointed of our Lord, & shal be
innocent? \V And Dauid ſaid: Our Lord liueth, vnles our Lord shal ſtrike
him, or his day come to die, or deſcending into battel he perish. \V Our
Lord be merciful vnto me, that I extend not my hand
vpon the annointed of our Lord. Now therefore take the ſpeare, which is
at his head, & cuppe of water, and let vs goe. \V Dauid therefore tooke
the ſpeare, and cuppe of water, which was at Saules head, and they went
away: and there was none that ſawe, or vnderſtood, or awaked, but al
ſlept, becauſe the dead ſleepe
\SNote{Gods prouidence ſent this extraordinary ſleepe and inſpired
Dauid, to doe this fact, for more iuſtification of his innocencie.}
of our Lord had fallen vpon them. \V And when Dauid had paſſed ouer
%%% o-0580
againſt, and ſtood in the toppe of the mountaine farre of, and a good
ſpace betwen them, \V Dauid cried to the people, and to Abner the ſonne
of Ner, ſaying: Wilt thou not anſwer Abner? And Abner anſwering ſayd:
Who art thou, that crieſt and diſquieteſt the king? \V And Dauid ſayd to
Abner: Art not thou a man? And who is like thee in Iſrael? why therefore
haſt thou not kept thy lord the king? for one of the multitude hath
entered in to kil the king thy lord. \V This thing is not good, which you
haue done: Our Lord liueth, you are the children of death, which haue
not kept your lord, the annointed of our Lord. Now therefore behold
where the kinges ſpeare is, & where the cup of water is, which was at
his head. \V And Saul knew Dauids voice, and ſayd: Is this thy voice, my
ſonne Dauid? And Dauid ſayd: My voice, my lord king, \V and he ſaid: For
what cauſe doth my lord perſecute his ſeruant? What haue I done? or what
euil is there in my hand? \V Now therefore heare, I pray, my lord King,
the wordes of thy ſeruant: If our Lord ſtirre thee vp againſt me, let
there be odoure of ſacrifice: but if the ſonnes of men, they are curſed
in the ſight of our Lord, which haue caſt me out this day, that I ſhould
not dwel in the inheritance of our Lord, ſaying: Goe, ſerue ſtrange
goddes. \V And now let not my blood be ſhed vpon the earth before our
Lord, for the king of Iſrael, is come forth to ſeeke one flea, as the
perdix is purſued in the mountaines. \V And Saul ſayd: I haue ſinned,
returne my ſonne Dauid, for I wil no more doe thee euil, for that my
life hath bene precious in thyne eies to day: for it appeareth that I
haue done foolishly, and haue bene ignorant of very many
%%% 0648
thinges. \V And Dauid anſwering, ſayd: Behold the kings ſpeare, let one
of the kings ſeruants paſſe, and take it. \V And our Lord wil reward
euerie one according to his iuſtice, and fidelitie: for our Lord hath
deliuered thee this day into my hand, & I would not extend my hand vpon
the annointed of our Lord. \V And as thy life hath bene magnified to day
in myne eies, ſo be my life magnified in the eies of our Lord, and
deliuer he me from al diſtreſſe. \V Saul therefore ſayd to Dauid:
Bleſſed art thou my ſonne Dauid: and truly doing thou shalt doe, and
preuayling thou shalt preuaile. And Dauid went into his way, and Saul
returned into his place.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
