%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 0631
%%% o-0566
\startChapter[
  title={Chapter 19}
  ]

\Summary{Saul intending to kil Dauid is pacified by
  Ionathas. 9.~Neuertheles attempteth agayn to kil him, and miſsing his
  purpoſe, 11.~ſendeth ſouldiars to take and bring him backe that he may
  be ſlaine, but Michol his wife helpeth him away, and excuſeth her ſelf
  to her father, as if she had done it for feare. 18.~Dauid and Samuel
  flee into Naioth. 20.~Againe Saul ſendeth ſouldiars after them three
  times, and they al doe prophecie. 22.~Then him ſelf purſueth Dauid,
  and alſo prophecieth.}

And Saul ſpake to Ionathas his ſonne, and to al his ſeruantes, that they
ſhould kil Dauid. Moreouer Ionathas the ſonne of Saul, loued Dauid
excedingly. \V And Ionathas told Dauid, ſaying: Saul my father ſeeketh
to kil thee: wherefore looke to thy ſelf I beſech thee in the morning,
and thou ſhalt abide ſecretly, and ſhalt be hid. \V But I going forth
wil ſtand beſide my father, in the field whereſoeuer he ſhal be: and I
wil ſpeake of thee to my father, and whatſoeuer I ſhal ſee, I wil tel
thee. \V Ionathas therefore ſpake good wordes of Dauid to Saul his
father: and ſayd to him: Sinne not, ô King againſt thy ſeruant Dauid,
becauſe he hath not ſinned toward thee, and his workes are very good for
thee. \V And he put his life in his hand, and ſtroke the Philiſtian, and
our Lord made great ſaluation to al Iſrael. Thou haſt ſeene & didſt
reioice, why therefore ſinneſt thou in innocent blood killing Dauid, who
is without fault? \V Which when
%%% 0632
Saul had heard, being pacified with the voice of Ionathas, he ſware: Our
Lord liueth, he shal not be ſlaine. \V Ionathas therefore called Dauid
and shewed him al theſe wordes, and Ionathas brought in Dauid to Saul,
and he was before him, as he had bene yeſterday and the day before. \V
And there was battel rayſed againe, and Dauid going forth, fought
againſt the Philiſtijms, and ſtroke them with a great ſlaughter, and
they fled from his face. \V And the euil ſpirit of our Lord came vpon
Saul, and he ſate in his houſe, and held a ſpeare: moreouer Dauid played
with his hand. \V And Saul endeuoured to naile Dauid to the wal with his
ſpeare. And Dauid declined from the face of Saul: and the ſpeare without
making wound, pearced the wal, and Dauid fled, and was ſaued that
night. \V Saul therefore ſent of his guarde into Dauids houſe, that they
ſhould keepe him, & that he might be killed in the morning. Which when
Michol his wife had told Dauid, ſaying:
%%% o-0567
Vnles thou ſaue thy ſelf this night, to morowe thou shalt die: \V she
let him downe through a windowe. Moreouer he went and fled away, and was
ſaued. \V And Michol tooke
\SNote{This ſtatua, or image can not importe an idol, as the ſame word
\HH{terephim} doth,
\XRef{Gen.~31.}
for here no idolatrie at al was cõmitted, but a right office done by the
wife towards her huſband. The ſame word alſo ſignifieth religious
thinges belonging to Gods ſeruice.
\XRef{Oſee.~3.}}
a ſtatua, and put it vpon the bed, and a hearie skinne of goates she
layd at the head thereof, and couered it with garmentes. \V And Saul
ſayd to Michol: Why haſt thou mocked me, and let goe myn enemie that he 
might flee? And Michol anſwered Saul: Becauſe he ſayd to me: Let me goe,
otherwiſe I wil kil thee. \V But Dauid fleing was ſaued, and came to
Samuel in Ramatha, and told him al thinges that Saul had done to him:
and he & Samuel went & abode in Naioth. \V And it was told Saul by ſome
ſaying: Behold Dauid is in Naioth in Ramatha. \V Saul therefore ſent
ſerientes, to take away
Dauid: who when they had ſeene a troupe of prophetes prophecying, &
Samuel ſtanding ouer them, the ſpirit of our Lord came alſo on them, and
they alſo began to prophecie. \V Which when it was told Saul, he ſent
other meſſengers: but they alſo did prophecie. And againe Saul ſent the
third meſſengers: who alſo prophecied. And Saul being wrath for
anger, \V went alſo him ſelfe into Ramatha, and came as
%%% 0633
farre as the great ceſterne, which is in Socho, and asked, and ſaid: In
what place are Samuel and Dauid? And it was told him: Loe they are in
Naioth in Ramatha. \V And he went into Naioth in Ramatha, and the Spirit
of our Lord came vpon him, and he walked going, and he
\SNote{Saul and his meſſengers had not the gift of prophecie, as godlie
prophetes had by an inherent habite, but tranſitorie, as Baalams aſſe
had facultie to ſpeake, at one time, but did not afterward ſpeake anie
more.
\Cite{S.~Aug. li.~2. q.~1. ad Simplician.}}
prophecied til he came into Naioth in Ramatha. \V And he ſtripped
himſelfe of his garments, and prophecied with the reſt before Samuel,
and ſang naked al that day and night. Wherupon there went out alſo a
prouerb: What is Saul alſo among the prophetes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
