%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 0623
%%% o-0559
\startChapter[
  title={Chapter 16}
  ]

\Summary{Samuel
\MNote{The third part.

Dauids annointing, his vertues, and perſecution.}
by Gods commandment annointeth Dauid King. 14.~Gods ſpirite parteth from
Saul, and a wicked ſpirite vexeth him. 16.~The vexation is mitigated by
Dauids playing on a harpe.}

And our Lord ſaid to Samuel: How long doeſt thou mourne Saul, whom I
haue reiected that he rule not ouer Iſrael? Fil thy horne with oile, and
come, that I may ſend thee to Iſai the Bethlehemite: for I haue prouided
me a king among his ſonnes. \V And Samuel ſaid: How ſhal I goe? for Saul
wil heare of it, and wil kil me. And our Lord ſaid: A calfe of the heard
ſhalt thou take in thy hand, and shalt ſay: I am come to immolate vnto
our Lord. \V And thou shalt cal Iſai to the victime, and I wil ſhew thee
what thou muſt doe, and thou shalt annointe whomſoeuer I ſhal ſhew to
thee. \V Samuel therefore did as our Lord ſpake to him. And he came into
Bethlehem, and the ancientes of the citie merueled, meeting him, and
they ſaid: Is thy entrance peaceable? \V And he ſaid: Peaceable: I am
come to immolate vnto our Lord, be ye ſanctified, and come with me that
I may immolate. He therefore ſanctified Iſai and his ſonnes, and called
them to the ſacrifice. \V And when they were entered in, he ſaw Eliab,
and ſaid: Is there before our Lord his Chriſt? \V And our Lord ſaid to
Samuel: Reſpect not his countenãce, nor the talnes of his ſtature:
%%% 0624
becauſe I haue reiected him, neither doe I iudge according to the looke
of man: for man ſeeth thoſe thinges which appeare, but our Lord
\LNote{Beholdeth the hart.}{It
\MNote{How Saintes and Prophetes know mens thoughtes.}
is proper to God to ſee the ſecrete cogitations of mens hartes, of
himſelfe and by his owne powre. And glorified Sainctes know our
cogitations by ſeing God, in whom al things appeare, that perteine to
their ſtate, and for the profite of others.
\Cite{S.~Aug. li. de cura pro mortuis, c.~15.}
\Cite{S.~Greg. li.~12. c.~13. Moral.}
Prophetes being yet mortal, by inſpiration do alſo ſee ſecrete
cogitations.
\XRef{1.~Reg.~9.}
\XRef{3.~Reg.~14. &c.}}
beholdeth the hart. \V And Iſai called Aminadab, and brought him before
Samuel. Who ſaid:
%%% o-0560
Neither this hath our Lord choſen. \V And Iſai brought Samma, of whom he
ſayd: This alſo hath not our Lord choſen. \V Iſai therefore brought his ſeauen
ſonnes before Samuel: and Samuel ſayd to Iſai: Our Lord hath not choſen
of theſe. \V And Samuel ſayd to Iſai: Are al thy ſonnes now fully come?
Who anſwered: Yet there is left a litle one, and he feedeth sheepe. And
Samuel ſayd to Iſai: Send, and bring him: for neither wil we ſitte downe
til he come hither. \V He ſent therefore, and brought him. And he was
reade and beautiful to behold, and of a comelie face. And our Lord ſaid:
Ariſe, and annoint him, for he it is. \V Samuel therefore tooke the
horne of oile, and annointed him in the middes of his brethren: and the
Spirit of our Lord from that day, and ſo forward was directed vpon
Dauid: and Samuel riſing went into Ramatha. \V And the Spirit of our
Lord departed from Saul, and a wicked ſpirit vexed him,
\SNote{By Gods permiſſion.
\Cite{S.~Aug. li.~2. q.~1. ad Simplician.}
\Cite{S.~Beda. qq. in 1.~Reg. c.~5.}}
from our Lord. \V And the ſeruantes of Saul ſaid to him: Behold an euil
ſpirit of God vexeth thee. \V Let our Lord command, and thy ſeruantes
which are before thee, wil ſeeke a man skilful to
\SNote{Naturally (as theſe men truly iudged) muſique helpeth ſome il
diſpoſition of humores, and draweth alſo the mind from ſo vehement
apprehenſion of afflictions: but here it ſemeth more probable, that God
ſupernaturally reliued Saul by Dauids playing on the harpe, together
with his ſincere deuotion, for more manifeſt condemnation of the one, and
iuſtification of the other.
\Cite{Gloſ. ordin. S.~Greg.}}
play on the harpe, that when the euil ſpirit of our Lord ſhal take thee,
he may play with his hand, and thou beare
\Fix{it more it more}{it more}{obvious typo, fixed in other}
eaſily. \V And Saul ſayd to his ſeruantes: Prouid me therefore ſome man
that playeth wel, and bring him to me. \V And one of the ſeruantes
anſwering, ſayd: Behold I haue ſeene the ſonne of Iſai the Bethlehemite
skilful to play, and very valiant in ſtrength, and a warlike man, and
wiſe in his wordes, and a beautiful man: and our Lord is with him. \V
Saul therfore ſent meſſengers to Iſai, ſaying: Send vnto me Dauid thy
ſonne, which is in the paſtures. \V Iſai therefore tooke an aſſe loaden
with loaues, and a flagon of wine, and one kidde of the goates, and ſent
it by the hand of Dauid his ſonne to Saul. \V And Dauid came to Saul,
and ſtoode before him: but he loued him excedingly, & was made his
eſquier. \V And Saul ſent to Iſai, ſaying: Let Dauid ſtand in my ſight:
for he hath found grace in myn eies. \V Therefore whenſoeuer the euil
ſpirit of our Lord caught Saul, Dauid tooke his harpe, & ſtrooke with
his hand, and Saul was refreshed, and waxed better. For the euil ſpirit
departed from him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
