%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 0633
%%% o-0567
\startChapter[
  title={Chapter 20}
  ]

\Summary{Ionathas comforteth Dauid, 3.~confirmeth their former
  league. 18.~By an appointed ſigne (24.~endeuoring firſt, but in vaine,
  to pacifie his father) 35.~certifieth Dauid of his fathers malice
  againſt him. 41.~They meete againe ſecretly, and ſorowfully part ech
  from other.}

But Dauid alſo fled from Naioth, which is in Ramatha, and coming ſpake
before Ionathas: What haue I done? what is myn iniquitie, and what ſinne
of myn againſt thy father, that he ſeeketh my life? \V Who ſayd to him:
God forbid, thou ſhalt not die: for neither wil my father doe any thing
great or litle, vnles he firſt tel me: this word therefore only hath my
father concealed from me? no this ſhal not be. \V And he
%%% o-0568
ſware againe to Dauid. And Dauid ſayd: Thy father ſurely knoweth, that I
haue found grace in thy ſight, and wil ſay: Let not Ionathas know this,
leſt perhaps he be ſad. Yea more our Lord liueth, and thy ſoule liueth,
by one degree only (as I may ſo ſay) I and death are diuided. \V And
Ionathas ſaid to Dauid: Whatſoeuer thy ſoule shal ſay to me, I wil doe
for thee. \V And Dauid ſayd to Ionathas: Behold the calendes are to
morowe, & I after the maner am wont to ſitte beſide the king to
eate: diſmiſſe me therefore that I may be hid in the field vntil the
euening of the third day. \V If thy father looing inquire for me, thou
ſhalt anſwer him: Dauid deſired me, that he might goe quickely into
Bethlehem his citie: becauſe there be ſolemne victimes to al of his
tribe. \V If he ſhal ſay: Wel; peace ſhal be to thy ſeruant: but if he
be angrie, know that his malice is complete. \V Doe mercie therefore
toward thy ſeruant: becauſe thou haſt cauſed me thy ſeruant to enter the
league of our Lord with thee. But if there be any iniquitie in me, do
thou kil me, and bring me not in to thy father. \V And Ionathas ſayd: Be
this farre from thee, for
%%% 0634
neither can it be, that I ſhould not tel thee, if I ſhal certainly know
that my fathers malice is complete againſt thee. \V And Dauid anſwered
Ionathas: Who ſhal bring me word, if thy father anſwer thee perhaps any
thing ſharpely of me? \V And Ionathas ſayd to Dauid: Come, let vs goe
forth abroad into the field, and when they were both gone forth into the
field, \V Ionathas ſaid to Dauid: Lord God of Iſrael, if I ſhal ſearch
out my fathers meaning, to morowe or the day after, and ſome good thing
be vpon Dauid, and I ſend not immediatly vnto thee, and make thee know
thereof, \V theſe thinges doe our Lord to Ionathas, and theſe thinges
adde he. But if my fathers malice ſhal perſeuer againſt thee, I wil
reuele thyn eare, and wil diſmiſſe thee, that thou mayeſt goe in peace,
and our Lord be with thee, as he hath beene with my father. \V And if I
liue, thou ſhalt doe me the mercie of our Lord; but if I die, \V thou
ſhalt not take away thy mercie from my houſe for euer, when our Lord
ſhal haue rooted out the enemies of Dauid, euerie one out of the land,
take he away Ionathas from his houſe, and our Lord require it of the
handes of Dauides enemies. \V Ionathas therefore made a league with the
houſe of Dauid: and our Lord required it of the handes of Dauids
enemies. \V And Ionathas added to ſweare vnto Dauid, becauſe he loued
him, for as his owne ſoule, ſo he loued him. \V And Ionathas ſayd to
him: To morowe are the calendes, and thou ſhalt be asked for: \V for thy
ſitting wil be inquired of til after to morowe. Thou ſhalt therefore goe
downe in haſt, and ſhalt come to the place, where thou muſt be hid in
the day, when it is lawful to worke, and thou ſhalt ſit beſide the
ſtone, which is named Ezel. \V And I wil ſhoote three arrowes nere it,
and wil shoote as it were exerſiſing my ſelf at a marke. \V I wil ſend
alſo a boy ſaying to him: Goe, and fetch me the arrowes. \V If I ſhal
ſay to the boy: Loe the arrowes are on this ſide thee, take them vp:
come thou to
%%% o-0569
me, becauſe there is peace to thee, and there is no euil, our Lord
liueth. But if I shal ſpeake thus to the boy: Loe the arrowes are beyond
thee: Goe in peace, becauſe our Lord hath dimiſſed thee. \V And
concerning the word which I and thou haue ſpoken, our Lord be betwen
thee and me for euer. \V Dauid therefore was hidde in the fielde, and
the calendes came, and the king ſate downe to eate bread. \V And when
the king was ſette vpon his chaire (according to the cuſtome)
%%% 0635
which was beſide the wal, Ionathas aroſe, and Abner ſate at the ſide of
Saul, and Dauids place appeared voide. \V And Saul ſayd nothing that
day, for he thought it had chanced perhaps vnto him, that he was not
cleane, nor purified. \V And when the ſecond day was come after the
calendes, againe Dauids place appeared emptie. And Saul ſaid to Ionathas
his ſonne: Why came not the ſonne of Iſai neither yeſterday, nor to day
to eate? \V Ionathas anſwered Saul: He deſired me inſtantly, that he
might goe into Bethlehem, \V and he ſaid: Let me goe, becauſe there is a
ſolemne ſacrifice in the citie, one of my brethren hath ſent for me: now
therefore if I haue found grace in thy ſight, I wil goe quickly, and ſee
my brethren. For this cauſe he came not to the kings table. \V But Saul
being wrath againſt Ionathas, ſaid to him: Thou ſonne of a woman which
of her owne accord rauisheth a man, am I ignorant that thou loueſt the
ſonne of Iſai vnto thyne owne confuſion, and to the confuſion of thyne
ignominious mother? \V For al the dayes, that the ſonne of Iſai shal
liue vpon the earth, thou shalt not be eſtablished, nor thy
kingdom. Therfore now preſently ſend, and bring him to me: Becauſe he is
the ſonne of death. \V And Ionathas anſwering Saul his father, ſaid: Why
shal he dye? what hath he done? \V And Saul caught a ſpeare to ſtrike
him. And Ionathas vnderſtood that it was determined of his father, that
he would kil Dauid. \V Ionathas therefore roſe from the table in anger
of furie, and did not eate bread the ſecond day of the calendes. For he
was ſtroken heauie vpon Dauid, becauſe his father had confounded him. \V
And when the morning appeared, Ionathas came into the field according to
the appointment with Dauid, and a litle boy with him. \V And ſaid to his
boy: Goe, and fetch me the arrowes, which I shoote. And when the boy had
runne, he shotte an other arrowe beyond the boy. \V The boy therefore
came to the place of the arrowe, which Ionathas had shotte: and Ionathas
cried behind the back of the boy, and ſaid: Loe the arrowe is there
further beyond thee. \V And Ionathas cried againe behind the back of the
boy, ſaying: Make haſt ſpedely, ſtand not. And Ionathas his boy gathered
vp his arrowes, and brought them to his maſter: \V and he was altogether
ignorant, what was done: for only Ionathas and Dauid knew the matter. \V
Ionathas therefore gaue his armoure to the boy, and ſaid to him: Goe,
and cary
%%% 0636
them into the citie. \V And when the boy was gone, Dauid roſe out of his
place, which did bend to the South, and falling flatt on the ground,
adored thriſe: and kiſſing one an other, they wept together, but Dauid
%%% o-0570
more. \V Ionathas therefore ſaid to Dauid: Goe in peace: whatſoeuer we
haue ſworne both of vs in the name of our Lord, ſaying: Our Lord be
betwen me and thee, and betwen my ſeede and thy ſeede for euer. \V And
Dauid aroſe, and departed: but Ionathas alſo entred into the Citie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
