%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 0617
%%% o-0554
\startChapter[
  title={Chapter 14}
  ]

\Summary{Ionathas truſting in God,
\Fix{accompained}{accompanied}{obvious typo, fixed in other}
with one man, his father not knowing, goeth into the Philiſthijms campe;
killeth twentie men and trubleth their whole armie. 16.~Saul
vnderſtanding the ſame, approcheth with his armie, and they gette a
great victorie. 24.~But Saul hauing commanded vnder paine of death, that
none should eate til night, Ionathas for taiſting a litle honie (though
ignorant of the prohibition) is iudged to die. 45.~But the people oppoſe
themſelues, and deliuer him from death. 47.~Saul proſpereth in his
kingdom, with his familie.}

And it chanced on a certeine day that Ionathas the ſonne of Saul ſayd to
the yong man that bare his armour: Come, & let vs paſſe to the gariſon
of the Philiſthijms, which is beyond yonder place. But to his father he
told not this ſame thing. \V Moreouer Saul abode in the vtmoſt part of
Gabaa vnder the pomegranate tree, which was in Magron: and the people
with him was about ſix hundred men. \V And Achias the ſonne of Achitob
the brother of Ichabod the ſonne of Phinees, which was borne of Heli the
prieſt of our Lord in Silo, bare the ephod. But the people alſo was
ignorant whither Ionathas was gone. \V And there were betwen the
aſcentes, by the which Ionathas endeuoured to paſſe vnto the gariſon of
the Philiſtijms, rockes ſtanding vp on both ſides, and as it were in
maner of teeth ſtiepe broken rockes on either ſide, the name of one
Boſes, and the name of the other Sene: \V one rocke ſtanding out toward
the North ouer againſt Machams, and the other to the South, againſt
Gabaa. \V And Ionathas ſayd to the yong man that bare his armour: Come,
let vs paſſe to the ſtation of theſe vncircumciſed, if haply our Lord wil
make for vs: becauſe it is not hard for our Lord to ſaue either in
manie, or in fewe. \V And his eſquier ſayd to him: Doe al thinges which
pleaſe thy minde: goe whither thou deſireſt, and I wil be with thee
whereſoeuer thou wilt. \V And Ionathas ſayd: Behold we paſſe to theſe
men. And when we ſhal appeare to them, \V If they ſhal ſpeake to vs in
this maner: Tarie til we come to
%%% 0618
you: let vs ſtand in our place, and not goe vp to them. \V But if they
ſhal ſay: Come vp to vs: let vs goe vp, becauſe our Lord hath
%%% o-0555
deliuered them into our handes,
\SNote{Ominous ſpeaches are proued by this and ſome other examples to be
ſometimes of God, though ſometimes this kind of obſeruation is
ſuperſticious. As before is noted.
\XRef{Gen.~24.}}
this ſhal be a ſigne vnto vs. \V Both of them therefore appeared to the
ſtation of the Philiſthims: & the Philiſtijms ſayd: Behold the Hebrewes
come out of the caues, wherein they were hid. \V And the men of the
\Fix{garniſon}{garriſon}{obvious typo, fixed in other}
ſpake to Ionathas and to his eſquier, and ſayd: Come vp to vs, and we
wil ſhew you a thing. And Ionathas ſayd to his eſquier: Let vs goe vp,
folow me: for our Lord hath deliuered them into the handes of Iſrael. \V
And Ionathas went vp on his handes & feete creeping, and his eſquier
after him. Therefore ſome fel before Ionathas, other ſome his eſquier
folowing ſlewe. \V And the firſt ſlaughter, with which Ionathas & his
eſquier made, was as it were of twentie men in the halfe part of an
aker, which a yoke of oxen is wont to plough in a day. \V And there was
made a miracle in the campe, through the fieldes: yea and al the people
of their garriſon, which had gone to take prayes, was aſtonyed, and the
land was trubled: and it happened as a miracle from God. \V And the
watchemen of Saul, which were in Gabaa, of Beniamin looked, & loe a
multitude ouerthrowen, & fleeing hither and thither. \V And Saul ſayd to
the people, which was with him: Enquire, and ſee who is gone from
vs. And when they had ſought, it was found that Ionathas was not preſent
and his eſquier. \V And Saul ſayd to Achias: Bring the arke of our
Lord. (For the arke of God was there that day with the children of
Iſrael.) \V And when Saul ſpake to the prieſt, there aroſe a great
tumult in the campe of the Philiſtijmes: and it grewe by litle and
litle, and ſounded more cleerely. And Saul ſayd to the prieſt:
\SNote{Pray no more nor expect no longer. So they proceded to battel
\Fix{withou}{without}{obvious typo, fixed in other}
further warrant.}
Draw together thy hand. \V Saul therefore and al the people that was
with him, ſhouted together, and they came to the place of the fight: and
behold euerie mans ſword had beene turned to his neighbour, and a
ſlaughter exceding great. \V But the Hebrewes alſo which had bene with
the Philiſtijms yeſterday and the day before, and went vp with them in
the campe, returned to be with Iſrael, which were with Saul and
Ionathas. \V Al the Iſraelites alſo which had hid themſelues in mount
Ephraim, hearing that the Philiſtijms were fled, ioyned them ſelues with
their fellowes in battel. And there were with Saul as it were ten
thouſand men. \V And our Lord in that day ſaued Iſrael.
%%% 0619
And the fight reached as farre as Bethauen. \V And the men of Iſrael
were ioyned among themſelues in that day: and Saul adiured the people,
ſaying: Curſed be the man, that ſhal eate bread vntil euening, til I be
reuenged of myne enemies. And the whole people did eate no bread: \V and
al the common people of the land came into a foreſt, wherein was honie
vpon the face of the field. \V The people therefore entred into the
foreſt, and there appeared dropping honie, and no man put his hand to
his mouth, for the people feared the oath. \V But Ionathas had not heard
when his father adiured the people: and he put forth the tippe of the
rod, which he had in his hand, and dipped it into a honie combe: and he
turned his hand to his mouth, and his eies were illuminated. \V And
%%% o-0556
one of the people anſwering, ſayd: Thy father hath bound the people with
an oath, ſaying: Curſed be the man that ſhal eate bread this day. (And
the people was faynt.) \V And Ionathas ſayd:
\SNote{Ionathas was excuſed by ignorance, & by neceſſitie; and therfore
was iuſtly deliuered by the people; and his father offended, in not
excepting the caſe of neceſſitie, & through more eggre zele of reuẽge,
then he had warrant from God, whoſe anſwer he would not expect.
\XRef{v.~19.}}
My father hath trubled the land: your ſelues haue ſeene that myn eies
are illuminated, becauſe I haue taſted a litle of this honie: \V how
much more if the people had eaten of the praye of their enemies, which
they found? had there not beene made a greater plague in the Philiſtijms? \V
They ſtroke therefore in that day the Philiſtijms from Machmas vnto
Ailon. And the people was wearied excedingly: \V and being turned to the
praye tooke ſheepe, and oxen, and calues, & ſlew them on the ground: and
the people did eate
\SNote{VVherof folowed alſo an other ſinne, that the people fainting for
lacke of meate did eate fleſh with the blood, contrarie to the law.}
with bloud. \V And they told Saul ſaying that the people had ſinned to
our Lord, eating with bloud. Who ſayd: You haue tranſgreſſed: Roule to
me euen now a great ſtone. \V And Saul ſayd: Diſperſe your ſelues among
the common people, and tel them that euerie man bring me his oxe and
ramme, and kil ye them vpon this ſame, and eate, and you shal not ſinne
to our Lord eating with bloud. Al the people therefore brought euerie
man his oxe in his hand vntil night: and ſlewe them there. \V And Saul
built an altar to our Lord; and then firſt did he beginne to
build an altar to our Lord. \V And Saul ſayd: Let vs fal vpon the
Philiſtijms by night, and let vs ſpoyle them til it waxe light in the
morning, neither let vs leaue a man of them. And the people ſayd: Doe al
that ſemeth good in thyne eies. And the prieſt ſayd: Let vs approch
hither to God. \V And Saul conſulted our Lord: Shal I purſew the
Philiſtijms? wilt thou deliuer them into the handes
%%% 0620
of Iſrael? And
\SNote{Before,
\XRef{v.~19.}
Saul would not expect Gods anſwer: now therfore God wil not anſwer him.}
he anſwered him not in that day. \V And Saul ſayd: Bring hither al the
corners of the people: and know, and ſee by whom this ſinne hath chanced
to day. \V Our Lord the ſauiour of Iſrael liueth, that if it were done
by Ionathas my ſonne, he shal die without reuoking. Whereunto none of
the people gayneſayed him. \V And he ſayd to al Iſrael: Be you ſeperated
into one ſide, and I with Ionathas my ſonne wil be on the other
ſide. And the people anſwered Saul: Doe what ſemeth good in thyn
eies. \V And Saul ſayd to our Lord: Lord God of Iſrael, geue me a ſigne:
and Ionathas was caught and Saul, and the people went forth. \V And Saul
ſayd: Caſt ye lotte betwen me, and Ionathas my ſonne. And
\SNote{Ionathas was found to haue tranſgreſſed the vnaduiſed cõmandment,
but Saul him ſelf was in a greatter fault of raſh proceding, and
vndiſcrete commanding.}
Ionathas was taken. \V And Saul ſayd to Ionathas: Tel me what thou haſt
done. And Ionathas told him, and ſayd: Taſting I taſted in the tippe of
the rod, which was in myn hand a litle honie, and behold I die. \V And
Saul ſayd: Theſe thinges doe God to me, and theſe thinges adde he, that
dying thou shalt die Ionathas. \V And the people ſaid to Saul: Shal
Ionathas then die, which hath made this great ſaluation in Iſrael? this
is vnlawful: our Lord liueth, if there shal fal a heare from his head
vpon the ground, becauſe with God hath he wrought to day. The people
therefore deliuered Ionathas, that he should not die. \V And Saul
retyred, neither did he purſew the
\Fix{Philiſtjims:}{Philiſtijms:}{obvious typo, fixed in other}
moreouer the Philiſtijms departed into their places.
%%% o-0557
\V And Saul, his kingdom being eſtablished ouer Iſrael, fought round about
againſt al his enemies, againſt Moab, and the children of Ammon, and
Edom, and the kinges of Soba, and the Philiſtians: and whither ſoeuer he
turned him ſelf, he ouercame. \V And gathering together an armie, he
ſtroke Amalec, and deliuered Iſrael from the hand of the ſpoylers
thereof. \V And the ſonnes of Saul, were Ionathas and Ieſui, and
Melchiſua: and the names of his two daughters, the name of the firſt
borne Merob, and the name of the yonger Michol. \V And the name of Sauls
wife, Achinoam the daughter of Achimaas: and the name of the prince of
his hoſt Abner, the ſonne of Ner, the coſin german of Saul by the
father. \V Moreouer Cis was the father of Saul, and Ner the father of
Abner, the ſonne of Abiel. \V And there was mightie battel agaynſt the
Philiſthians al the dayes of Saul. For whomſoeuer Saul had ſeene a
valiant man, and fitte for battel, he ioyned him to him ſelf.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
