%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 0608
%%% o-0546
\startChapter[
  title={Chapter 9}
  ]

\Summary{Saul
\MNote{The ſecond part.

The election, annointing, & gouernement of King Saul.}
by occaſion of ſeeking his fathers aſses cometh to
  Samuel. 15.~Who had a reuelation of his coming, and a commandment to
  annoint him. 22.~He is entertained and lodged with Samuel.}

And there was a man of Beniamin named Cis, the ſonne of Abiel, the ſonne
of Seor, the ſonne of Bechorath, the ſonne of Aphia, the ſonne of a man
of Iemini, valiant in ſtrength. \V And he had a ſonne called Saul,
choſen & good: and there was not a man of the children of Iſrael better
then he: from the ſhoulder and vpward he appeared aboue al the
people. \V And the aſſes of Cis the father of Saul were loſt: and Cis
ſaid to Saul his ſonne: Take one of the ſeruants with thee, and riſing
goe, and ſeeke the aſſes. Who when they had paſſed by mount Ephraim, \V
and by the land of Salifa, and had not found, they paſſed alſo through
the land of Salim, and they were not: yea and by the Land of Iemini, and
found them not. \V And when they were come into the Land of Suph, Saul
ſaide to the ſeruant that was with him: Come let vs returne, leſt
perhaps my father hath let alone the aſſes, and be careful for vs. \V
Who ſayd to him: Behold a man of God is in this citie, a famous man: al
that he ſpeaketh, cometh to paſſe without doubt. Now therfore let vs goe
thither, if perhaps he may tel vs of our way, for which we are come. \V
And Saul ſayd to his ſeruant: Loe we wil goe: what ſhal we carie to the
man of God? The bread is ſpent in our males: and preſent we haue none to
geue vnto the man of God, nor anything els. \V Agayne the ſeruant
anſwered
%%% o-0547
Saul, and ſayd: Behold there is found in my hand the fourth part of a
ſicle of ſiluer, let vs geue it to the man of God, that he may tel vs
our way. \V (For in time paſt in Iſrael ſo euery man ſpake, going to
conſult God: Come, and let vs goe to the Seer. For he that at this day
is called a Prophete, in time paſt was called
\SNote{One that by diuine inſpiration foretelleth thinges too come.}
a Seer.) \V And Saul ſayd to his ſeruant: Thy word is very good, come
let vs goe. And they went into the citie, wherein the man of God was. \V
And when they went vp the aſcent of the citie, they found maides coming
forth to draw water, and ſayd to them: Is the Seer here? \V Who
anſwering ſayd to them: Here he is. Loe before thee, make haſt now: for
this day he came into the citie, becauſe this day there is a ſacrifice
of the people in the excelſe. \V Entring into the citie immediatly
%%% 0609
you ſhal find him, before he goe vp into the excelſe to eate. For the
people wil not eate til he come: becauſe he wil bleſſe the Hoſte, and
afterward they ſhal eate that are inuited. Now therfore goe vp, becauſe
this day you ſhal finde him. \V And they went vp into the citie. And
when they walked in the middes of the citie, Samuel appeared coming
forth againſt them, to goe vp into the excelſe. \V And our Lord had
reueled the eare of Samuel one day before Saul came, ſaying: \V This
very houre, that now is, to morrow wil I ſend to thee a man of the Land
of Beniamin, and thou ſhalt annoint him ruler ouer my people of Iſrael:
and he shal ſaue my people from the hand of the Philiſtijms: becauſe I
haue reſpected my people, for
\SNote{Oppreſſion of innocentes crieth to heauen.}
their crie is come to me. \V And when Samuel had beheld Saul, our Lord
ſayde to him: Behold the man, of whom I told thee, this man shal rule
ouer my people. \V And Saul came to Samuel in the middes of the gate,
and ſayd: Shew me, I pray thee, where is the houſe of the Seer? \V And
Samuel anſwered Saul, ſaying: I am the Seer, goe vp before me into the
excelſe, that you may eate with me to day, and I wil dimiſſe thee in the
morning: and al thinges that are in thy hart, wil I tel thee. \V And
concerning the aſſes, which thou didſt loſe three dayes agone, be not
careful, becauſe they are found. And whoſe shal be al the beſt thinges
of Iſrael? not to thee and to al thy fathers houſe? \V And Saul
anſwering, ſayd: Am not I the ſonne of Iemini of the leaſt tribe of
Iſrael, and my kindred the laſt among al the families of the tribe of
Beniamin? Why therfore haſt thou ſpoken this word to me? \V Samuel
therfore taking Saul and his ſeruant, brought them into the parlour, and
gaue them a place in the chiefe rowme of them that were inuited, for
there were about thirtie men. \V And Samuel ſayd to the cooke: Geue the
portion, which I gaue thee, and commanded that thou shouldeſt lay it vp
apart with thee. \V And the cooke lifted vp a shoulder, and ſette it
before Saul. And Samuel ſaid: Behold that which hath remayned, ſette it
before thee, and eate: becauſe of purpoſe it was kept for thee, when I
called the people. And Saul did eate with Samuel that day. \V And they
deſcended from the excelſe into the towne, and he ſpake with Saul in the
toppe of the houſe: and he prepared a bed for Saul in the higheſt rowme,
& he ſlept. \V And when they were
%%% o-0548
riſen in the morning, and it beganne now to be light, Samuel
%%% 0610
called Saul in the high chãber, ſaying: Ariſe that I may diſmiſſe
thee. And Saul aroſe: and they went both forth: to witte, he and
Samuel. \V And when they came downe in the vttermoſt part of the citie,
Samuel ſaid to Saul: Speake to the ſeruant that he goe before vs, and
paſſe: but ſtay thou a litle while, that I may tel thee the word of our
Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
