%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 0613
%%% o-0551
\startChapter[
  title={Chapter 12}
  ]

\Summary{Samuel being iuſtified by the people for his good behauiour,
  6.~chargeth them with ingratitude towardes God, 14.~admonishing them,
  and shewing by a ſigne, that they offended in demanding a
  king. 20.~Exhorteth them now to ſerue God, promiſeth to pray for them,
  and forwarneth that they shal receiue as they deſerue.}

And Samuel ſayd to al Iſrael: Behold I haue heard your voice according
to al thinges which you haue ſpoken to me, and I haue appointed a king
ouer you. \V And now the king goeth before you: and I am waxen old and
haue gray heares: moreouer my ſonnes are with you: therfore hauing
conuerſed with you from my youth vntil this day, loe I am readie. \V
Speake of me before our Lord, and before his Chriſt, whether I haue
taken any mans oxe, or aſſe: If I haue calumniated any man, if I haue
oppreſſed any man, if I haue taken gift of any mans hand: and I wil
contemne that ſame this day, and wil reſtore it to you. \V And they ſaid
thou haſt not calumniated vs, nor oppreſſed vs, nor taken ought of any
mans hand. \V And he ſayd to them: Witnes is our Lord againſt you, and
witnes is his
\SNote{The annointed king.}
Chriſt in this day, that you haue not found any thing in my hand. And
%%% 0614
they ſaid: Witnes. \V And Samuel ſaid to the people: Our Lord who made
Moyſes and Aaron, and brought our fathers out of the Land of Ægypt is
preſent. \V Now therefore ſtand, that I may contend in iudgement againſt
you before our Lord, concerning al the mercies of our Lord, which he
hath done with you, and with your fathers: \V how Iacob entred into
Ægypt, and your fathers cried to our Lord: and our Lord ſent Moyſes and
Aaron, and brought your fathers out of Ægypt: and placed them in this
place. \V Who forgat our Lord their God, and he deliuered them in the
hand of Siſara maſter of the hoſte of Haſor, and in the hand of the
Philiſthijmes, and in the hand of the king of Moab, and they fought
againſt them. \V But afterward they cried to our Lord, and ſaid: We haue
ſinned, becauſe we haue forſaken our Lord, and haue ſerued Baalim and
Aſtaroth: now therfore deliuer vs from the hand of our enemies, and we
wil ſerue thee. \V And our Lord ſent Ierobaal, and
\TNote{mightie Samſon.}
Badan, and Iepte, and Samuel, and deliuered you from the hand of your
enemies round about, and you dwelt ſecurely. \V But you ſeing that Naas,
king of the children of Ammon was come againſt you, you ſaid to me:
%%% o-0552
\SNote{They ment that they would not be ruled and protected as hitherto
they had bene, by Dukes & Iudges ordained and ſent immediatly by God,
but would haue a king to reigne ouer them: imagening that ſo they ſhould
be better protected, and defended from forraine enemies wherin they
preferred their owne conceite and iudgement before Gods diſpoſition. And
therfore this ſinne is often here inculcated, and much reprehended.
\Cite{S.~Greg. li.~5. c.~2. in 1.~Reg.~12.}}
Not ſo, but a king shal reigne ouer vs: whereas our Lord your God did
reigne among you. \V Now therfore your king is readie, whom you haue
choſen and deſired: behold our Lord hath geuen you a king. \V If you
shal feare our Lord, and ſerue him, and heare his voice, and not
exaſperat the mouth of our Lord: both you, and the king which reigneth
our you, shal be folowers of our Lord your God. \V But if you wil not
heare the voice of our Lord, but shal exaſperat his wordes, the hand of
our Lord shal be vpon you, and vpon your
\TNote{gouernoures.
\Cite{S.~Greg.}}
fathers. \V But now alſo ſtand, and ſee this great thing which our Lord
wil doe in your ſight. \V Is it not wheate harueſt to day? I wil cal
vpon our Lord, & he wil geue
\TNote{thunder.}
noyſes and rayne: and you shal know, and ſee that you haue done great
euil to your ſelues in the ſight of our Lord, deſiring a king ouer
you. \V And Samuel cried to our Lord, and our Lord gaue noyſes and raine
in that day. \V And al the people feared excedingly our Lord and
\SNote{They feared God ſo much the more, becauſe they feared alſo his
ſeruant.}
Samuel. And al the people ſaid to Samuel: Pray for thy ſeruantes to our
Lord thy God, that we die not. For we haue added euil to al our ſinnes,
that we deſired vnto vs a king. \V And Samuel ſayd to the people: Feare
not, you haue done al this euil: but yet depart not from the
\SNote{For ſinners to come before Gods face were preſumption, not to
come to his back or to depart from thence, were deſperation. So Marie
Magdalen with feare and hope approched & taried at Chriſts back.
\XRef{Luc.~7.}
\Cite{S.~Greg. loco citate.}}
backe of our Lord,
%%% 0615
but ſerue our Lord in al your hart. \V And decline not after vaine
thinges, which shal not profite you, nor deliuer you, becauſe they are
vaine. \V And our Lord wil not forſake his people for his great name:
becauſe our Lord hath ſworne to make you a people to him ſelf. \V And
farre from me be this ſinne in our Lord, that I should ceaſe to pray for
you, and I wil teach you the good and right way. \V Therefore feare our
Lord, and ſerue him in truth and from your whole hart. For you haue
ſeene the great workes which he hath done among you. \V But if you shal
perſeuer in malice: both you and your king shal perish together.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
