%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 25
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-25


\project douay-rheims


%%% 0643
%%% o-0576
\startChapter[
  title={Chapter 25}
  ]

\Summary{Samuel dieth and is mourned. 3.~Dauid requeſting, 10.~and not
  obtaining victuals of Nabal, 13.~threatneth to kil him. 14.~But his
  wife Abigail prudently preuenteth the reuenge, 18.~by ſending
  victuals, 23.~and geuing good worde. 37.~Al which when Nabal
  vnderſtandeth, he fainteth, and after tenne dayes dieth. 39.~Dauid
  marieth Abigail: 43.~alſo Achinoam: 44.~and his wife Michol is geuen
  to an other.}

And Samuel died, and al Iſrael was gathered together, and they mourned
for him, and buried him in his houſe in Ramatha. And Dauid riſing went
downe into the deſert of Pharan. \V And there was a certaine man in the
wildernes of Maon, and his poſſeſſion in Carmel, and that man was
exceding great: and he had three thouſand sheepe, & a thouſand goates:
and it chanced that his flocke was shorne in Carmel. \V And the name of
that man was Nabal: and the name of his wife Abigail. And that woman was
very wiſe and beutiful: moreouer her huſband hard, and very il, and
malicious: and he was of the kindred of Caleb. \V When Dauid therefore
heard in the deſert, that Nabal sheared his flocke, \V he ſent tenne
yong men, and ſayd to them: Goe vp into Carmel, and you shal come to
Nabal, and shal ſalute him in my name peaceably. \V And you shal ſay:
Peace be to my brethren, and to thee, and peace to thy houſe, and to al
whatſoeuer thou haſt be peace. \V I haue
%%% o-0577
heard, that thy shephards which were with vs in the deſert did
sheare. We haue neuer moleſted them, neither hath ought beene wanting to
them at any time of the flocke, al the time that they were with vs in
Carmel. \V Aske thy ſeruantes, and they wil tel thee. Now therefore let
thy ſeruantes finde grace in thyne eies: for we are come in a good day,
whatſoeuer thy hand shal finde, geue thy ſeruantes, and thy ſonne
Dauid. \V And when Dauids ſeruantes were come, they ſpoke to Nabal al
theſe wordes in Dauids name: and ſo held their peace. \V But Nabal
anſwering the ſeruantes of Dauid, ſaid: Who is Dauid? and what is the
ſonne of Iſai? There are ſeruantes multiplied now a daies which flee
from their maſters. \V Shal I then take my breades, and my waters, and
the flesh of my cattel, which I haue killed for my
\Fix{sheares,}{shearers,}{obvious typo, fixed in other}
and geue to men whom I know not whence they are? \V Therefore the
ſeruantes of Dauid returned by their way, and returning came and told
him al the
%%% 0644
wordes that he had ſaid. \V Then ſayd Dauid to his ſeruantes: Euery man
gird him with his ſworde. And they were euery one girded with their
ſwordes. And Dauid alſo was girded with his ſworde: and there folowed
Dauid about foure hundred men: moreouer two hundred remayned at the
baggage. \V But to Abigail the wife of Nabal one of their ſeruantes
told, ſaying: Behold Dauid hath ſent meſſengers from the deſert, to
bleſſe our maiſter, & he diſdayned them: \V Theſe men were good ynough
to vs, and not trubleſome: neither did euer any thing perish al the
time, that we haue conuerſt with them in the deſert: \V they were in
ſteed of a wal to vs both in the day and in the night, al the dayes
that we fed the flockes with them. \V Wherefore conſider, & thinke what
thou haſt to doe, for
\SNote{Not a malicious mind, ſuch as Saul bore vniuſtly againſt Dauid;
but puniſhmẽt is decreed againſt Nabal, for not only denying a
reaſonable requeſt, but alſo for ſo reprochful an anſwer.}
malice is accompliſhed againſt thy huſband, and againſt thy houſe, and
he is the ſonne of Belial, ſo that no man can ſpeake to him. \V Abigail
therefore made haſt, and tooke two hundred loaues, and two bottels of
wine, and fiue muttons ready dreſt, and fiue meaſures of polent, and a
hundred branches of rayſens, and two hundred maſes of drie 
figges, and laid them vpon aſſes: \V and ſaid to her ſeruantes: Goe
before me: loe, I wil folow you at your backe: but ſhe told not her
huſband Nabal. \V When ſhe therefore had gotten vpon an aſſe, and came
downe to the foote of the mountaine, Dauid and his men came downe
meeting her, whom ſhe alſo mette. \V And Dauid ſaid: In vaine veryly
haue I preſerued al thinges that were this mans in the deſert, and there
periſhed nothing of al that perteined to him: and he hath rendred me
\SNote{Euil wordes for curteous vſage heretofore, and for late gentil
intreating by meſſengers.}
euil for good. \V Theſe thinges doe God to the enemies of Dauid, and
theſe thinges adde he, if I ſhal leaue of al thinges that perteine to
him vntil morning, any thing piſſing againſt the wal. \V And when
Abigail had ſeene Dauid, ſhe made haſt, and light from her aſſe, and fel
downe before Dauid vpon her face, and adored vpon the earth, \V and fel
at his feete, and ſayd: In me my lord be this iniquitie: let thy
handmayde ſpeake I beſech thee in thine eares: and heare the wordes of
thy ſeruant. \V Let not my lord the king I pray thee, ſet his hart vpon
%%% o-0578
this naughtie man Nabal: becauſe according to his name, he is a foole,
and follie is with him: but I thy handmaide ſawe not thy ſeruantes my
lord, whom thou didſt ſend. \V Now therefore my lord, our Lord liueth,
and thy ſoule liueth, who hath ſtayed thee that thou ſhouldeſt
%%% 0645
not come vnto blood, and hath ſaued thy hand to thee: and now be thine
enemies as Nabal, and they that ſeeke euil to my lord. \V Wherefore
receiue this benediction, which thy handmaide hath brought to thee my
lord: and geue to thy ſeruantes that folow thee my lord. \V Take away
the iniquitie of thy handmaide: for our Lord making wil make to my lord
a faithful houſe, becauſe thou my lord doeſt fight the battels of our
Lord: let not malice therefore be found in thee al the dayes of thy
life. \V For if a man ſhal riſe, perſecuting thee, and ſeeking thy life,
the life of my lord ſhal be kept, as in the
\SNote{Things tyed in bundels are ſtronger and more ſecure, then ſingle
and looſe.}
bundel of the liuing, with our Lord thy God: Moreouer the life of thine
enemies ſhal be whurled, and in the violence, and whurle of a ſling. \V
When our Lord therefore ſhal doe to thee, my lord, al good thinges,
which he hath ſpoken concerning thee, and ſhal conſtitute thee prince
ouer Iſrael, \V this ſhal not be an occaſion of ſobbing to thee, and a
ſcruple of hart to my lord, that thou haſt ſhed innocent blood, or thy
ſelfe haſt reuenged thy ſelfe: and when our Lord hath beſtowed theſe
benefites vpon my lord, thou ſhalt remember thy handmaide. \V And Dauid
ſayd to Abigail: Bleſſed be our Lord the God of Iſrael, who hath ſent
thee this day to meete me, and bleſſed be thy ſpeach, \V and bleſſed by
thou, which haſt ſtayed me to day, that I might not goe to blood, and
reuenge me with myn owne hand. \V Otherwiſe our Lord liueth the God of
Iſrael, who hath ſtaied me, that I ſhould not doe thee any euil: vnles
thou hadſt quickely come to meete me, there had not remained to Nabal
vntil morning light, any piſſing againſt a wal. \V And Dauid tooke of
her hand al thinges which ſhe had brought him, and ſaid to her: Goe
peaceable into thy houſe, behold I haue heard thy voice, and haue
honoured thy face. \V And Abigail came to Nabal: and behold he had a
feaſt in his houſe, as it were the feaſt of a king, and Nabals hart was
pleaſant: for he was drunke excedingly: and ſhe told him not a word
litle or great vntil morning. \V But early when Nabal had digeſted his
wine, his wife told him theſe wordes, and his hart was dead inwardly, and
he became as a ſtone. \V And when ten dayes had paſſed, our Lord ſtroke
Nabal, and he died. \V Which when Dauid had heard that Nabal was dead,
he ſayd: Bleſſed be our Lord, who hath iudged the cauſe of my reproch at
the hand of Nabal, and hath kept his ſeruant from euil, and the
%%% 0646
malice of Nabal hath our Lord rendred vpon his head. Dauid therefore
ſent, and ſpake to Abigail, that he might take her to himſelfe to
wife. \V And Dauids ſeruantes came to Abigail into Carmel, and ſpake to
her, ſaing: Dauid hath ſent vs vnto thee, to take thee to his wife. \V
Who ariſing adored flat toward the earth, and ſayd: Loe let thy ſeruant
be as an handmaide, to waſh the feete of the ſeruantes of my lord. \V And
%%% o-0579
Abigail aroſe, and made haſt, and gatte vpon an aſſe, and fiue wemen
went with her waiting maides, and folowed the meſſengers of Dauid, and
became his wife. \V Yea and Achinoam alſo did Dauid take of Iezrahel:
and both were his wiues. \V But Saul gaue Michol his daughter Dauids
wife to Phalti, the ſonne of Lais, who was of Gallim.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
