%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 0625
%%% o-0561
\startChapter[
  title={Chapter 17}
  ]

\Summary{Goliath a Philiſtian chalengeth anie one of Iſrael to
  combate. 12.~Dauid being ſent by his father to viſit his brethren,
  23.~and hearing al Iſrael ſo reprochfully prouoked, offereth to
  vndertake the chalenge: 34.~shewing by former actes that he dare
  accept it. 37.~And ſo by Gods ſpecial helpe wherein he truſteth,
  49.~ouerthroweth the chalenger with a ſtone of his ſling, and cutteth
  of his head with his owne ſword. 51.~The Philiſtijms fleeing are
  ſlaine, and Dauid bringeth the mans head to Saul.}

And the Philiſthijms gathering together their companies vnto battel,
aſſembled into Socho of Iuda: and camped betwen Socho & Azeca in the
borders of Dommim. \V Moreouer Saul and the children of Iſrael being
gathered together came into the Valley of terebinth, and they put the
armie in aray to fight againſt the Philiſtijms. \V And the Philiſtijms
ſtoode vpon the mountaine on this ſide, and Iſrael ſtoode vpon the
mountaine on the other ſide: and the valley was betwen them. \V And
there came forth a man that was a baſtard from the campe of the
Philiſtians named
\SNote{This bold and impudent chalenger ſignifieth the diuel, or anie
arch heretike, that prouoketh the Church of God. But is ouercome by the
humble of hart and confident in God, & ſlaine with his owne weapon.}
Goliath, of Geth, in height ſix cubites and a palme: \V and a helmet of
braſſe vpon his head, and he was clothed with a cote of mayle linked,
moreouer the weight of his cote of mayle was fiue thouſand ſicles of
braſſe: \V and he had braſſen bootes on his thighes, and a target of
braſſe couered his shoulders. \V And the ſhaft of his ſpeare had ſix
hundred ſicles of yron: and his eſquier went before him. \V And ſtanding
he cried againſt the bandes of Iſrael, and ſayd to them: Why came you
prepared to fight? Am not I a Philiſtian, and you the ſeruantes of Saul?
Chooſe out a man of you and let him deſcend to fight hand to hand. \V If
he ſhal be able to fight
%%% 0626
with me, and ſtrike me, we wil be ſeruantes to you: but if I ſhal
preuaile, and ſhal beate him, you ſhal be ſeruantes, and shal ſerue
vs. \V And the Philiſtian ſayd: I haue defyed the bandes of Iſrael this
day: Geue me a man, and let him fight with me hand to hand. \V And Saul
and al the Iſraelites hearing ſuch wordes of the Philiſtian were
aſtonied, and feared excedingly. \V And there was Dauid the ſonne of a
man that was an Ephratheite, of whom there was mention before, of
Bethlehem Iuda, whoſe name was Iſai, who had eight ſonnes, and he was in
the dayes of Saul an old man, and aged among men. \V And his three elder
ſonnes went after Saul into battel: and the names of his three ſonnes,
which went to battel, were Eliab the firſt begotten, and the ſecond
Abinadab, the third alſo
%%% o-0562
Samma: \V and Dauid was the yongeſt. The three elder therefore hauing
folowed Saul, \V Dauid went, and returned from Saul, to feede his
fathers flocke in Bethlehem. \V But the Philiſtian came forth morning
and euening, and ſtoode fourtie dayes. \V And Iſai ſayd to Dauid his
ſonne: Take for thy brethren an ephi of polent, and theſe ten loaues,
and runne into the campe to thy brethren, \V and theſe ten litle cheeſes
thou shalt carie to the tribune: and shalt viſite thy brethren, if they
doe wel: and learne with whom they are placed. \V And Saul, and they,
and al the children of Iſrael fought in the Valley of terebinth againſt
the Philiſtijms. \V Dauid therefore aroſe in the morning, and commended
the flocke to the keeper: and he went loaded as Iſai had commanded
him. And he came to the place Magala, and to the hoſt, which iſſuing out
to fight had made a shoute in the battel. \V For Iſrael had put them
ſelues in aray, and the Philiſtijms on the contrarie ſaid were
prepared. \V Dauid therefore leauing the veſſels which he had brought,
vnder the hand of him, that was keeper at the baggage, ranne to the
place of the battel and asked if al thinges went wel with his
brethren. \V And when he yet ſpake to them, that man the baſtard
appeared coming vp, named Goliath, the Philiſtian of Geth, coming vp
from the campe of the Philiſtians: and he ſpeaking theſe ſelf ſame
wordes, Dauid heard them. \V And al the Iſraelites when they had ſene
the man, fled from his face, fearing him excedingly. \V And ſome one of
Iſrael ſayd: Haue you ſeene this man that came vp, to defye Iſrael he
came vp? The man therefore
\SNote{Chriſt ouercoming the diuel receiued the Church to his ſpouſe.
\Cite{S.~Greg.}}
that shal ſtrike him, the King
%%% 0627
wil geue him his daughter, and he wil make his fathers houſe without
tribute in Iſrael. \V And Dauid ſpake to the men that ſtood with him,
ſaying: What shal be geuen to the man that shal beate this Philiſtian,
and shal take away the reproch from Iſrael? For who is this
vncircumciſed Philiſtian, which hath vpbraided the armies of the liuing
God? \V And the people reported vnto him the ſelf ſame worde, ſaying:
Theſe thinges shal be geuen to the man, that shal ſtrike him. \V Which
when Eliab his eldeſt brother had heard, when he ſpake with others, he
was angrie againſt Dauid, and ſayd: Wherefore cameſt thou? and why haſt
thou left thoſe few sheepe in the deſert? I know thy pride, and the
wickednes of thy hart: that to ſee the battel thou art come downe. \V
And Dauid ſayd: What haue I done? is there not cauſe to ſpeake? \V And
he went a litle aſide from him to an other: and ſayd the ſelf ſame word. And the
people anſwered him as before. \V And the wordes which Dauid ſpake were
heard, and told in the ſight of Saul. \V To whom when he was brought, he
ſpake vnto him: Let not any mans hart be diſcouraged in him: I thy
ſeruant wil goe, and wil fight againſt the Philiſtian. \V And Saul ſayd
to Dauid: Thou art not able to reſiſt this Philiſtian, nor to fight
againſt him, becauſe thou art a childe, but he is a man of warre from
his youth. \V And Dauid ſayd to Saul: Thy ſeruant did feede his fathers
flock, and there came a lyon,
\TNote{or, for and.
\XRef{v.~36.~37.}}
or a beare, and tooke a ramme out of
%%% o-0563
the middes of the flocke: \V and I purſued them, and ſtroke them, and
plucked them out of their mouth, and they aroſe vp againſt me, and I caught
their chinne, and I ſtrangled and ſlew them. \V For
\SNote{He that hath ouercome the ſpirit of pride, and of carnalitie
(ſignified by a lion, and a beare) is able alſo to ouercome the diuel.}
both the lyon and the beare did I thy ſeruant kil: therefore this
vncircumciſed Philiſtian alſo, shal be as it were one of them. Now wil I goe and
take away the reproch of the people: for who is this vncircumciſed
Philiſtain, which hath beene ſo hardie to curſe the hoſt of the liuing
God? \V And Dauid ſayd: Our Lord which hath deliuered me from the hand
of the lion, and of the beare, he wil deliuer me from the hand of this
Philiſtian. And Saul ſayd to Dauid: Goe, and our Lord be with thee. \V
And Saul clothed Dauid with his rayments, and put an helmet of braſſe
vpon his head, and veſted him with a coate of maile. \V Dauid therefore
being girded with his ſword ouer his rayment, beganne to proue if he
could goe armed: for he was not accuſtomed.
%%% 0628
And Dauid ſayd to Saul: I can not goe ſo, becauſe I am not vſed, and he
layd them of. \V And he tooke his ſtaffe, which he had alwaies in his
handes: & he choſe him fiue moſt bright ſtones out of the torrent, and
caſt them into the shepherds skrippe, which he had with him, and he
tooke a ſling in his hand, and went forth againſt the Philiſtian. \V And
the Philiſtian went, going, and approching againſt Dauid, and his eſquier before
him. \V And when the Philiſtian had ſeene, and beheld Dauid, he deſpiſed
him. And he was a yong man redde, and beautiful to behold. \V And the
Philiſtian ſayd to Dauid: Why am I a dogge, that thou comeſt to me with
a ſtaffe? And the Philiſtian curſed Dauid in his goddes. \V And ſayd to
Dauid: Come to me, and I wil geue thy flesh to the foules of the ayre
and the beaſtes of the earth. \V And Dauid ſayd to the Philiſtian: Thou
comeſt to me with a ſword, and ſpeare, and sheeld, but I come to thee in
the name of the Lord of hoſts, the God of the bandes of Iſrael, whom
thou haſt defied \V this day, and our Lord shal geue thee in my hand,
and I shal ſtrike thee, and take away thy head from thee: and I shal
geue the carcaſſes of the campe of the Philiſtijms this day, to the
fowles of the ayre, and to the beaſtes of the earth: that al the earth
may knowe that there is a God in Iſrael. \V And al this aſſemblie shal
know, that not in ſword, nor in ſpeare, doth our Lord ſaue, for it is
his battel, and he wil deliuer you into our handes. \V When the
Philiſtian therefore was riſen vp, and came and approched againſt Dauid,
Dauid made haſt, & ranne to the battel againſt the Philiſtian. \V And he
put his hand into his skrippe, and tooke one ſtone, and caſt it with the
ſling, and fetching it about ſtroke the Philiſtian in
\SNote{Pride hauing impudencie of the forhead, is ouerthrowne by
humilitie of Chriſts croſſe. Whoſe ſigne therfore we carie in our
forhead.
\Cite{S.~Aug. li.~50. homil. ho.~31.}}
the forehead, and he fel on his face vpon the earth. \V And Dauid
preuailed agaynſt the Philiſtian with ſling and ſtone, and he ſtroke,
and ſlew the Philiſtian. And whereas
Dauid had no ſword in his hand, \V he ranne, and ſtood vpon the
Philiſtian, and tooke his ſword, and drew it out of the ſcabard, and
ſlew him, and cut of his head. And the Philiſtijms ſeing, that the
ſtrongeſt of them was dead, did flee. \V And the men of Iſrael and Iuda
riſing vp shouted, and purſued the Philiſtians til they came into a
valley to the gates
%%% o-0564
of Accaron, and there fel wounded of the Philiſtijms in the way of
Saraim, as farre as Geth, & as farre as Accaron. \V And the children of
Iſrael returning, after they had purſued the
%%% 0629
Philiſtians, inuaded their campe. \V And Dauid taking the head of the
Philiſtian brought it into Ieruſalem: but his armour he layd in his
tabernacle. \V And at the ſame time that Saul ſaw Dauid going forth
againſt the Philiſtian, he ſayd to Abner the prince of the armie:
\SNote{Saul knew not Dauid, being perhaps in a ſhepwards habite: though
he had not long before ſerued & pleaſed him wel.
\XRef{ch.~16. v.~21.}}
Of what ſtocke is this yong man deſcended, Abner? And Abner ſayd: Thy
ſoule liueth ô king, if I know. \V And the king ſayd: Aske thou, whoſe
ſonne this yong man is. \V And when Dauid was returned, after the
Philiſthian was ſlaine, Abner tooke him, and brought him in before Saul,
hauing the head of the Philiſtian in his hand. \V And Saul ſayd to him:
O yong man of what progenie art thou? And Dauid ſayd: I am the ſonne of
thy ſeruant Iſai the Bethlemite.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
