%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 30
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-30


\project douay-rheims


%%% 0653
%%% o-0586
\startChapter[
  title={Chapter 30}
  ]

\Summary{Dauid returning to Siceleg, findeth it burned and ſpoiled, and
  himſelf in danger of the people: 7.~By our Lords warrant he purſueth
  the enemie, 11.~taketh a guide: 17.~recouereth al that was taken away:
  22.~and rewardeth the ſouldiars, alſo thoſe that ſtayed with the
  baggage; 25.~making it a lawe for the time to come, that the keepers
  of the baggage, shal haue like share with thoſe that fight in battel.}

And when Dauid and his men were come to Siceleg the third day, the
Amalecites had made an inuaſion on the ſouth ſide into Siceleg, and had
ſtrooken Siceleg, and burnt it with fire. \V And had led away wemen
captiue out of it, from the leſſe vnto the great: and had not killed any
man, but had led them with them, and went on their iourney. \V When
Dauid therefore and his men were come to the citie, and had found it
burnt with fire, and their wiues, and their ſonnes, and their daughters
to be led away captiue, \V Dauid and the people that was with him,
lifted vp their voices, and mourned til teares fayled them. \V For the
two wiues alſo of Dauid were led away captiue, Achinoam the Iezraelite,
and Abigail the wife of Nabal of Carmel. \V And Dauid was ſtrooken very
ſad: for the people would haue ſtoned him, becauſe the ſoule of euerie
man was bitterly affected vpon their ſonnes, and daughters: but Dauid
was ſtrengthened in our Lord his God. \V And he ſayd to Abiathar the
prieſt the ſonne of Achimelech: Applie
\SNote{Conſult our Lord for me: ſo Dauid by the prieſts mediation, was
inſtructed what to do.}
vnto me the Ephod. And
%%% 0654
Abiathar applied the Ephod to Dauid, \V and Dauid conſulted our Lord,
ſaying: Shal I purſue theſe theeues, and shal I take them or no? And our
Lord ſayd to him: Purſue them: for thou shalt take them without doubt, &
take from them the praye. \V Dauid therefore went himſelfe, and the ſix
hundred men that were with him, and they came vnto the Torrent Beſor: &
ſome being wearie ſtayed. \V But Dauid himſelf, and foure hundred men
purſued: for two hundred ſtayed, who being wearie could not paſſe the
Torrent Beſor. \V And they found an Ægyptian in the field, and brought
him to Dauid: and they gaue him bread to eate, and water to drink, \V
and alſo a peece of a maſſe of drie figges, and two bunches of
reſinnes. Who when he had eaten, his ſpirite returned, and he was
refreshed: for he had not eaten bread, nor drunken water three daies,
and three nightes. \V Dauid therefore ſayd to him: Whoſe art thou? or
whence? and whither goeſt thou? Who ſayd: I am a yong man of Ægypt, the
ſeruant of an Amalacite: and my maiſter hath leift me, becauſe I began to be
ſicke three daies agone. \V For we brake forth at the ſouth ſide of
Cerethi, and againſt Iuda, and at the ſouth of Caleb, and burnt
Siceleg
%%% o-0587
with fire. \V And Dauid ſayd to him: Canſt thou bring me to this
companie? Who ſayd: Sweare to me by God, that thou wilt not kil me, nor
deliuer me into the handes of my lord, and I wil bring thee to this
companie. And Dauid ſware to him. \V Who when he had brought them,
behold they ſate vpon the face of al the earth, eating and drinking, and
as it were keping feſtiual day, for al the praye, and ſpoiles which they
had taken out of the Land of the Philiſtijms, and out of the Land of
Iuda. \V And Dauid ſtroke them from euening vntil the euening of the
next day, and there eſcaped not anie of them, but foure hundred yong
men, which had gotten vpon camels, and were fled. \V Dauid therefore
reskewed al thinges, that the Amalecites had taken, and he reskewed his
two wiues. \V Neither was there anie wanting from litle to great as wel
of their ſonnes as of their daughters, and of the ſpoiles, and what
thinges ſoeuer they had taken, Dauid brought al againe. \V And he tooke
al the flockes & heardes, and draue them before his face: and they ſayd:
This is the praye of Dauid. \V And Dauid came to the two hundred men,
which being wearie had ſtaied, neither were able to folow Dauid, and he
had bidden them to reſt in the Torrent Beſor:
%%% 0655
who came forth to meete Dauid, and the people that were with him. And
Dauid coming to the people, ſaluted them peaceably. \V And euerie
naughtie, and wicked felow of the men, that had gone with Dauid
anſwering, ſayd: Becauſe they came not with vs, we wil not geue them
anie thing of the praye, which we haue recouered: but let their wife and
children ſuffice euerie man, whom when they haue receiued, let them
depart. \V But Dauid ſayd: You ſhal not doe ſo my brethren, of theſe
thinges, which our Lord hath deliuered to vs, and hath kept vs, and hath
geuen the theeues, that were broken out againſt vs, into our handes: \V
neither ſhal anie man heare you vpon this talke. For there ſhal be equal
portion of him that went downe into battel, and of him that abode at the
baggage, & they shal diuide alike. \V And this hath beene downe from
that day, and euer after
\SNote{It is not againſt Gods cõmandment,
\XRef{Deut.~4.}
&
\XRef{12.}
to make new lawes, ſo they be conformable, & not contrarie to Gods
former lawes.}
it was decreed, and ordained as a law in Iſrael. \V Dauid therefore came
into Siceleg, and ſent giftes of the praye to the ancientes of Iuda his
neighboures, ſaying: Take a benediction of the praye of our Lords
enemies. \V To them, that were in Bethel, and that were in Ramoth toward
the South, and them that were in Gether, \V and them in Aroer, and them
in Sephamoth, and them in Eſthamo, \V and them in Rachal, and them in
the cities of Ierameel, and them in the cities of Semi, \V and them in
Harama, and them in the lake of Aſan, and them in Athach, \V and them in
Hebron, and to the reſt that were in thoſe places, in the which Dauid
had taried, and his men.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
