%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 0621
%%% o-0557
\startChapter[
  title={Chapter 15}
  ]

\Summary{Saul is commanded vtterly to deſtroy the Amalecites. 8.~But he
  taking Agag their king ſpareth his life, & chiefe of the
  praye. 10.~For which diſobedience (20.~though pretending that the
  beſt thinges were reſerued for ſacrifice) he is depoſed from his
  kingdom. 24.~Then acknowledgeth his fault. 32.~Samuel cutteth Agag in
  peeces, 35.~and mourneth for Saul.}

And Samuel ſaid to Saul: Our Lord ſent me to annointe thee king ouer his
people Iſrael; now therefore heare the voice of our Lord: \V Thus ſayeth
the Lord of hoſtes: I haue recounted whatſoeuer Amalec hath done to
Iſrael: how he reſiſted them in the way when they came vp out of
Ægypt. \V Now therefore goe, and
\SNote{Amalec is ſtricken when the fleſh is chaſticed by abſtinence.}
ſtrike Amalec, and
\SNote{Deſtroyed when the mind is reſtrained from vncleane cogitations.
\Cite{S.~Greg. li.~6. c.~1. in 1.~Reg.~15.}}
deſtroy al that he hath: ſpare him not, and couet not ought of his
thinges: but kil from man vnto woman, both childe & ſuckling, oxe and
ſheepe, camel and aſſe. \V Saul therefore commanded the people, and
numbered them as it were lambes: two hundred thouſand footemen, & ten
thouſand of the men of Iuda. \V And when Saul was come vnto the citie of
Amalec, he laid ambuſhementes in the torrent. \V And Saul ſaid to the
Cineite: Goe ye, retyre and depart from Amalec: leſt perhaps I wrappe
thee in with him. For thou haſt done mercie with al the children of
Iſrael, when they deſcended out of Ægypt. And the Cineite departed out
of the middes of Amalec. \V And Saul ſtroke Amalec from Heuila,
vntil thou come to Sur, which is ouer againſt Ægypt. \V And he
apprehended Agag the king of Amalec aliue: but al the comon people he
ſlewe in the edge of the ſword. \V And Saul and the people ſpared Agag,
and the beſt flockes of ſheepe and heardes, and the garmentes and
rammes, and al thinges,
%%% o-0558
that were fayre, neither would they deſtroy them: but whatſoeuer was
vile and refuſe, that they deſtroyed. \V And the word of our Lord was
made to Samuel, ſaying: \V It repenteth me that I haue made Saul king:
becauſe he hath forſaken me, & hath not fulfilled my wordes in
worke. And Samuel was ſtrooken ſadde, and cried to our Lord al the
night. \V And when Samuel had riſen in the night, to goe to Saul in the
morning, it was told Samuel, that Saul was come into Carmelus, and had
erected to him ſelfe a triumphant arch, and returning was paſſed, and
gone into Galgal. Samuel therefore came to Saul, and Saul offered an
holocauſt to our Lord of the firſt of the prayes,
%%% 0622
which he had brought from Amalec. \V And when Samuel was come to Saul,
Saul ſayd to him: Bleſſed be thou to our Lord, I haue fulfilled the word
of our Lord. \V And Samuel ſayd: And what is this voice of flockes,
which ſoundeth in myne eares, and of heardes, which I heare? \V And
Saul ſaid: They haue brought them from Amalec: for the people hath
ſpared the better ſheepe and heardes that they might be immolated to our
Lord thy God, but the reſt we haue ſlaine. \V And Samuel ſaid to Saul:
Suffer me, and I wil ſhew thee what our Lord hath ſpoken to me this
night. And he ſayd to him: Speake. \V And Samuel ſaid: When thou waſt a
little one in thyne owne eyes,
\SNote{VVhen Saul was humble he was exalted, now being proud he is
reiected.}
was thou not made chief in the tribes of Iſrael? And our Lord annointed
thee to be king ouer Iſrael, \V and our Lord ſent thee on the way, and
ſayd: Goe, and kil the ſinners of Amalec, and thou shalt fight againſt
them vntil the vtter deſtruction of them. \V Why therefore haſt thou not
heard the voice of our Lord: but art turned to the praye, and haſt done
euil in the eies of our Lord? \V And Saul ſaid to Samuel: Yea I haue
heard the voice of our Lord, and haue walked in the way by which our
Lord ſent me, and haue brought Agag the king of Amalec, and Amalec I
haue ſlaine. \V But the people tooke of the praye ſheepe and oxen, the
principal of thoſe thinges which were ſlaine, to immolate to our Lord
their God in Galgal. \V And Samuel ſaid: Why wil our Lord haue
holocauſtes and victimes, and not rather that the voice of our Lord be
obeyed? For
\SNote{By ordinarie ſacrifices, we geue our external goodes to God, by
obedience we geue our ſelues.
\Cite{S.~Greg. li.~6. c.~2. in 1.~Reg.~15.}}
\Sc{Better is obedience then victimes}: and to harken rather then to
offer the fatte of rammes. \V Becauſe it is as it were the ſinne of
inchantment, to reſiſt: and as it were the wickednes of idolatrie, to
refuſe to obey. For as much therefore as thou haſt reiected the word of
our Lord, our Lord hath reiected thee that thou shalt not be king. \V And
Saul ſaid to Samuel: I haue ſinned, becauſe I haue tranſgreſſed the
ſaying of our Lord, and thy wordes, fearing the people, and obeying
their voice. \V But now beare I beſech thee my ſinne, and returne with
me, that I may adore our Lord. \V And Samuel ſayd to Saul: I wil not
returne with thee, becauſe thou haſt reiected the word of our Lord, and
our Lord hath reiected thee that thou ſhalt not be king ouer Iſrael. \V
And Samuel turned him ſelfe to depart: but he caught the hemme of his
cloke, which alſo did rent. \V And Samuel ſaid to him: Our Lord hath
rent the
%%% 0623
kingdom of Iſrael from thee this day, and hath deliuered it
%%% o-0559
to thy neighbour better then thou. \V Moreouer the Triumpher in Iſrael
wil not ſpare, and he wil not be turned with repentance: for neither is
he a man that he may repent. \V But he ſaid: I haue ſinned: howbeit now
honour me before the ancientes of my people, and before Iſrael, and
returne with me, that I may adore our Lord thy God. \V Samuel therefore
returning folowed Saul: and Saul adored our Lord. \V And Samuel ſaid:
Bring vnto me Agag the king of Amalec. And Agag was preſented to him
very fatte, trembling. And Agag ſayd: Doth bitter death thus ſeparat? \V
And Samuel ſaid: As thy ſword hath made wemen without children, ſo ſhal
thy mother among wemen be without children. And Samuel hewed him into
peeces before our Lord in Galgal. \V And Samuel departeth into Ramatha:
but Saul aſcended vnto his houſe into Gabaa. \V And Samuel ſaw Saul no
more vnto the day of his death: but yet Samuel lamented Saul becauſe it
\SNote{God is ſaid improperly to repent, when he altereth the thing,
that he did before.
\Cite{S.~Amb. li. de Noe ca.~4.}}
repented our Lord that he had appointed him king ouer Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
