%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 24
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-24


\project douay-rheims


%%% 0641
%%% o-0575
\startChapter[
  title={Chapter 24}
  ]

\Summary{Saul chancing to come into a caue, Dauid being hid with others
  in the ſame place, only cutteth the skirt of his cloke, 5.~which after
  shewing, Saul acknowledgeth his fault, and ceaſeth for a time to
  perſecute him.}

Dauid therefore went vp from thence, and dwelt in the ſafeſt places of
Engaddi. \V And when Saul was returned, after he purſued the
Philiſtians, they told him, ſaying: Behold, Dauid is in the deſert
Engaddi. \V Saul therefore taking vnto him three thouſand choſen men of
al Iſrael, went forth to ſearch out Dauid, and his men, yea ouer the
ſteepe broken rockes, which are acceſſible only to wilde goates. \V And
he came to the shepcotes, which fel in his way as he went. And there was
a caue, which Saul entred into, to doe his eaſement: moreouer Dauid and
his men lay hid in the inner part of the caue. \V And the ſeruantes of
Dauid ſayd to him: Behold the day, whereof our Lord ſayd to thee: I wil
deliuer thee thyne enemie, that thou mayeſt doe to him as it ſhal ſeeme
good in thyne eies. Dauid therefore aroſe, and cut of the hemme of
Saules cloke ſoftly. \V After this
\SNote{A tender cõſcience had ſcruple of a ſmal doubt which was no
ſinne: a large cõſcience ſticketh at nothing.}
Dauids hart ſtroke him, for that he had cut of the hemme of Saules
cloke. \V And he ſayd to his men: Our Lord be merciful vnto me,
\SNote{Saul being annointed king by Gods appointment, could not lawfully
be ſlaine by his ſubiectes without like ordinance from God. For though
Dauid was alſo already annointed, yet that was not to reigne preſently,
but when Saul ſhould die, or otherwiſe be taken away.}
that I doe not this thing to my Maiſter the annointed of our Lord, that
I ſhould lay my hand vpon him, becauſe he is the annointed of our
Lord. \V And Dauid perſwaded
%%% 0642
his men with wordes, and permitted them not to riſe againſt Saul:
moreouer Saul riſing out of the caue, went on his iourney begunne. \V
And Dauid roſe vp after him: and going out of the caue, cried behind
Sauls backe, ſaying: My Lord King. And Saul looked backe behind him: and
Dauid bowing him ſelfe flat toward the ground adored, \V and
ſayd to Saul: Why doeſt thou heare the wordes of men that ſay: Dauid
ſeeketh euil againſt thee? \V Loe this day thyne eies haue ſeene, that
our Lord deliuered thee in my hand, in the caue, and I had a cogitation
to kil thee, but myn eie hath ſpared thee. For I ſayd: I wil not extend
myn hand vpon my lord, becauſe he is the annointed of our Lord. \V But
rather ſee and knowe, ô my father, the hemme of thy cloke in my hand,
that when I did cut of the hemme of thy cloke. I would not extend my
hand vpon thee. Marke, and ſee, that there is no euil in my hand, nor
iniquitie, neither haue I ſinned againſt thee: but thou lyeſt in waite
for my life, to take it away. \V Our Lord iudge betwen me and thee, and
our Lord reuenge me of thee, but be not my hand vpon thee. \V As alſo it
is ſayd in the old
%%% o-0576
prouerbe: \Sc{From the Impiovs} ſhal impietie procede: be not therefore
my hand vpon thee: Whom doeſt thou perſecute Ô King of Iſrael? \V Whom
doeſt thou perſecute? thou perſecuteſt a dead dog, and a flea. \V Our
Lord be iudge, & iudge betwen me and thee, and he ſee, and iudge my
cauſe, and deliuer me out of thy hand. \V And when Dauid had fully ended
ſpeaking ſuch wordes to Saul, Saul ſayd: Is this thy voice my ſonne
Dauid? and Saul lifted vp his voice, and wept: \V and ſayd to Dauid:
Thou art iuſter then I: for thou haſt donne me good turnes, and I haue
rendred thee euil. \V And thou haſt ſhewed this day what good thinges
thou haſt done to me: how our Lord deliuered me into thy hande and thou
haſt not killed me. \V For who when he hath found his enemie, wil let
him goe in a good way? But our Lord render thee this good turne, for
that which thou haſt wrought toward me this day. \V And now becauſe I
knowe that thou moſt certeinly ſhalt reigne, and haue the kingdome of
Iſrael in thy hand: \V ſweare to me in our Lord, not to deſtroy my ſeede
after me, nor to take away my name from the houſe of my father. \V And
Dauid ſware to Saul. Saul therefore went into his houſe: and Dauid and
his men went vp into ſafer places.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
