%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 0595
%%% o-0535
\startChapter[
  title={Chapter 2}
  ]

\Summary{Anna geueth thankes in a Canticle. 11.~The ſonnes of Heli
  greuouſly ſinning are reprehended, but not duly corected, by their
  father. 21.~Anna beareth three ſonnes more, and two
  daughters. 27.~Heli is threatned, 34.~and the death of his two ſonnes
  fortold.}

My hart hath reioyced in our Lord, and my horne is exalted in my God: my
mouth is dilated vpon myn enemies: becauſe I haue ioyed in thy
ſaluation.

\V There is none holie as our Lord is: for neither is there an other
beſide thee, and there is none ſo ſtrong as our God.

\V Doe not multiplie to ſpeake high thinges, boaſting:
\SNote{Leaue of to praiſe idoles, as ye haue accuſtomed to doe.}
let old matters depart from your mouth: becauſe our Lord is a God of al
knowlege, and to him cogitations are prepared.

\V The bow of the ſtrong men is ouercome, and the weake are girded with
ſtrength.

\V They that before were filled haue hyred out them ſelues for bread:
and the hungrie are filled, vntil
\SNote{The Church of Gentiles.}
the barren woman bare verie manie: and
\SNote{The Synagogue of the Iewes.
\Cite{S.~Aug. li.~17. c.~4. ciuit.}}
she that had manie children, was weakened.

\V Our Lord mortifieth and quickeneth, bringeth downe to hel and
fetcheth backe agayne.

\V Our Lord maketh poore and enricheth, humbleth and lifteth vp.

\V He rayſeth the needie man from the duſt, and from the dung he lifteth
vp the poore: that he may ſitte with princes, and hold the throne of
glorie. For the poles of the earth are our Lords, and vpon them he hath
ſette the world.

\V The feete of his Saintes he wil keepe, and the impious shal be ſilent
in darknes: becauſe in his owne force man ſhal
%%% 0596
not be ſtrengthned.

%%% !!! New paragraph in other only
\V Our Lord shal his aduerſaries feare: and vpon them shal he thunder in
the heauens: our Lord shal iudge the endes of
\SNote{Neither Dauid nor Salomon, much leſſe anie other King, poſſeſſed
or iudged the endes of the earth: but Chriſts enheritance reacheth to
the endes of the earth.
\XRef{Pſal.~2. v.~18.}}
the Earth, and shal geue empire to his king, and shal exalt the horne of
his Chriſt.

\V And Elcana went into Ramatha, vnto his houſe: but the childe
miniſtered in the ſight of our Lord before the face of Heli the
prieſt. \V Moreouer the ſonnes of Heli, were the ſonnes of Belial, not
knowing
%%% o-0536
our Lord, \V nor the office of prieſtes to the people: but whoſoeuer had
immolated a victime, the ſeruant of the prieſt came, whiles the flesh
was in boyling, and had a flesh hooke with three teeth in his hand, \V
and thruſt it into the kettle, or into the caudron, or into the potte,
or into the panne: and al, that the flesh hooke brought vp, the prieſt
tooke to him ſelfe. So did they to al Iſrael that came into Silo. \V Yea
before they burnt the fatte, the ſeruant of the prieſt came, and ſayd to
him that immolated: Geue me flesh, that I may boyle it for the prieſt:
for I wil not take flesh of thee ſodde, but raw. \V And he that
immolated ſayd to him: Let the fatte firſt be burnt to day according to
the maner, and take vnto thee how much ſoeuer thy ſoule deſireth. Who
anſwering ſaid to him: Not ſo: for thou shalt geue it now, or els I wil
take it away by force. \V Therfore the ſinne of the yong men was exceding
great before our Lord: becauſe men detracted from the ſacrifice of our
Lord. \V But Samuel miniſtred before the face of our Lord: a child,
girded with an ephod of linnen. \V And his mother made him a litle
tunike, which she brought vpon the ordinarie daies, going vp with her
husband, to immolate the ſolemne hoſt. \V And Heli bleſſed Elcana and
his wife: and he ſaied to him: Our Lord render thee ſeede of this woman,
for the vſurie that thou haſt geuen our Lord. And they went into their
place. \V Our Lord therfore viſited Anna, and she conceiued, and bare
three ſonnes, and two daughters: and the childe Samuel was magnified
before our Lord. \V And Heli was very old, and heard al thinges which
his ſonnes did to al Iſrael: and how they ſlept with the wemen that
wayted at the doore of the tabernacle: \V and he ſayd to them: Why doe
you theſe kinde of thinges, which I heare, very naughtie thinges, of al
the people? \V Doe not ſo my ſonnes: for it is not a good report, which
I doe heare, that you make the people of our Lord to tranſgreſſe. \V If
man shal ſinne againſt
%%% 0597
man, God may be pacified toward him: but if a man ſhal ſinne againſt our
Lord
\SNote{Sinnes directly againſt God, and that hinder his ſeruice, are
more hardly remitted: but none at al are irremiſſible before death,
becauſe during life euerie one may truly repent, if he wil: and to al
true penitẽtes God promiſſeth remiſſion of ſinnes.
\XRef{Ezech.~33.}}
who ſhal pray for him? And they heard not the voice of their father,
\SNote{Gods determination to puniſh tooke not away their freewil, but
for their obſtinacie he leift them to them ſelues, without his grace,
and ſo iuſtly puniſhed them. See
\Cite{S.~Aug. li.~5. cont. Iulian. c.~3.}}
becauſe our Lord would kil them. \V But the childe Samuel proſpered, and
grew, and pleaſed both our Lord and men. \V And there came a man of God
to Heli, and ſaid to him: Thus ſayth our Lord: Was not I openly reueled
to thy fathers houſe, when they were in Ægypt in the houſe of Pharao? \V
And I choſe him of al the tribes of Iſrael for my prieſt, that he might
aſcend to my altar, and burne to me incenſe, and might carie the ephod
before me: and I gaue to thy fathers houſe al thinges of the ſacrifices
of the children of Iſrael. \V Why haue you with your heele reiected my
victime, and my giftes which I commanded to be offered in the temple:
and haſt rather honoured thy ſonnes then me, that you would eate the
firſt fruites of euerie ſacrifice of Iſrael my people? \V Therfore
ſayeth our Lord the God of Iſrael: Speaking I ſpeake that thy houſe, and
the houſe of thy father should miniſter in my ſight, for euer. But now
ſayeth our Lord: Be this farre from me: but whoſoeuer ſhal glorifie me I
wil glorifie him:
%%% o-0537
and they that contemne me, ſhal be baſe. \V Behold the daies come: and I
wil cut of thy arme, and the arme of thy fathers houſe, that there may
not be an old man in thy houſe. \V And thou ſhalt ſee
\SNote{This was fulfilled as in the figure in Samuel, not wholly, for
prieſthood ſtil remained in the line of Aaron, as appeareth in Achias,
Abiathar, & Sadoc.
\XRef{ch.~14.}
\XRef{22.}
\XRef{2.~Reg.~8.}
but perfectly in Chriſt.
\Cite{S.~Beda. qq. in 1.~Reg. c.~2.}}
whom thou enuieſt in the temple, in al proſperities of Iſrael and there
ſhal not be an old man in thy houſe for euer. \V Notwithſtanding I wil
not altogether take away a man of thee from myn altar: but that thyne
eyes may fayle, and thy ſoule melt: and a great part of thy houſe ſhal
die when it is come to mans age. \V And this shal be a ſigne to thee,
which shal come vpon thy two ſonnes, Ophni, and Phinees: In one day they
shal both die. \V And I wil rayſe vp vnto me a faithful prieſt, which
ſhal doe according to my hart, and my ſoule: and I wil build him a
faythful houſe, and the ſame shal walke before my Chriſt al daies. \V
And it shal come to paſſe, that whoſoeuer shal remayne in thy houſe,
shal come that he may be prayed for, and shal offer a peece of ſiluer,
and a manchet of bread, and shal ſay: Leaue me I beſech thee to one
prieſtly part, that I may eate a morſel of bread.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
