%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 0604
%%% o-0542
\startChapter[
  title={Chapter 7}
  ]

\Summary{The Arke is brought to the houſe of Abinadab in Gabaa. 3.~By
  Samuels exhortation, the people caſt away the idols and ſerue only
  God. 10.~Samuel offering ſacrifice and praying, Iſrael preuaileth
  againſt the Philiſthijms.}

Therfore the men of Caria Thiarim came, and
\SNote{Theſe men knowing that the preſence of the arke was good for them
(though the Bethſamites had benne puniſhed for their irreuerence
towardes it) feared not to receiue and kepe it.}
brought backe the arke of our Lord, and caried it into the houſe of
Abinadab in
\Fix{Gaaba:}{Gabaa:}{obvioius typo, fixed in other}
And Eleazar his ſonne they ſanctified, that he might keepe the arke of
our Lord. \V And it came to paſſe, from the day that the arke of our
Lord abode in Caria Thiarim the dayes were multiplied (for it was now
the twentith yeare) and al the houſe of Iſrael reſted after our Lord. \V
And Samuel ſpake to al the houſe of Iſrael, ſaying: If you turne to our
Lord in al your hart, take away the ſtrange goddes out of the middes of
you, Baalim, and Aſtaroth: and prepare your hartes to our Lord, and
ſerue him only, and he wil deliuer you from the hand of the
Philiſthijms. \V Therfore the children of Iſrael tooke away Baalim and
Aſtaroth, and ſerued our Lord only. \V And Samuel ſayd: Gather together
al
%%% o-0543
Iſrael into Maſphath, that I may pray our Lord for you. \V And they
aſſembled into Maſphath: and they drew water, and powred it out in the
ſight of our Lord, and they faſted that day, and ſayd there: We haue
ſinned to our Lord. And Samuel iudged the children of Iſrael in
Maſphath. \V And the Philiſthijms heard that the children of Iſrael were
gathered together into Maſphath, and the princes of the Philiſthijms went
vp to Iſrael. Which when the children of Iſrael had heard, they were
afrayde at the face of the Philiſthijms. \V And they ſaid to Samuel:
Ceaſe not to crie to our Lord God for vs, that he ſaue vs from the hand
of the Philiſthims. \V And Samuel tooke one ſucking lambe, and offered
it a whole holocauſte to our Lord: and Samuel cried to our Lord for
Iſrael, and our Lord heard him. \V And it came to paſſe, when Samuel
offered the holocauſte, the Philiſthijms beganne battel againſt Iſrael:
but our Lord thundered with a great noiſe in that day vpon the
Philiſthijms, and terrified them, and they were ſlaine before the face
of Iſrael. \V And
%%% 0605
the men of Iſrael iſſuing out of Maſphath purſued the Philiſthijms, and
ſtroke them vnto the place, that was vnder Bethcar. \V And Samuel tooke
one ſtone, and layd it betwen Maſphath and Sen: and he called the name
of that place, The ſtone of helpe. And he ſayd: Thus farre hath our Lord
holpen vs. \V And the Philiſtijms were humbled, neither added they any
more to come into the borders of Iſrael. Therfore the hand of our Lord
was made vpon the Philiſtijms, al the dayes of Samuel. \V And the
cities, which the Philiſtijms had taken from Iſrael, were rendred to
Iſrael, from Accaron vnto Geth, and their borders: and he deliuered
Iſrael from the hand of the Philiſtijms, and there was peace betwen
Iſrael and the
\SNote{That is to ſay, the Philiſthijms, who were one of the ſeuen
nations of Chanaan, which God commanded his people to deſtroy,
cõmonly called the Amorrheites.}
Amorrheite. \V Samuel alſo iudged Iſrael al the daies of his life: \V
and he went euerie yeare circuting Bethel and Galgala and Maſphath, and
iudged Iſrael in the foreſaid places. \V And he returned into Ramatha:
for there was his houſe, and there he iudged Iſrael: he built alſo there
an altar to our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
