%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-23


\project douay-rheims


%%% 0639
%%% o-0573
\startChapter[
  title={Chapter 23}
  ]

\Summary{The citie of Ceila oppugned by the Philiſtijms is releeued by
  Dauid. 7.~Who fearing to be there betrayed, 13.~fleeth into the deſert
  of Ziph. 16.~Ionathas repayreth ſecretly to him, and they confirme
  againe their former league. 19.~The Ziphians promiſe to betray Dauid:
  27.~but Saul leaueth for a while to perſecute him, being forced to
  defend the land from the Philiſtians inuading it.}

And they told Dauid, ſaying: Behold the Philiſtijms oppugne Ceila, and
ſpoyle the barnes. \V Dauid therefore conſulted our Lord, ſaying: Shal I
goe, and ſtrike theſe Philiſtians? And our Lord ſayd to Dauid: Goe, and
thou ſhalt ſtrike the Philiſtians, and shalt ſaue Ceila. \V And the men
that were with Dauid, ſayd to him: Behold we reſting here in Iurie are
afrayd, how much more if we shal goe into Ceila againſt the bandes of
the Philiſtians? \V Againe therefore Dauid conſulted our Lord. Who
anſwering ſayd to him: Ariſe, and goe into Ceila: for I wil deliuer the
Philiſtians in thy hand. \V Dauid therefore, and his men, went into
Ceila, and fought againſt the Philiſtians, and droue away their beaſtes,
and ſtroke them with a great ſlaughter, and Dauid ſaued the inhabitantes
of Ceila. \V Moreouer at that time, when Abiathar the ſonne of
Achimelech fled to Dauid into Ceila, he went downe hauing with him an
ephod. \V And it was told Saul that Dauid was come into Ceila: and Saul
ſayd: Our Lord hath deliuered him into my handes, and he is shut vp
being entered the citie, wherein are gates and lockes. \V And Saul
commanded al the people, that they should goe downe into Ceila to fight,
and beſeige Dauid, and his men. \V Which when
%%% 0640
Dauid vnderſtood, that Saul ſecretly prepared euil againſt him, he ſayd
to Abiathar the prieſt:
\SNote{By this it appeareth that Dauid conſulted God by the high prieſt.
And the high prieſt for this purpoſe ioyned the prieſtlie ornament
called \Emph{Ephod} to the \Emph{Rationale}, and ſo had reuelatiõ from
God.
\XRef{Exo.~28.}
\XRef{Leuit.~8.}}
Applie the Ephod. \V And Dauid ſayd: Lord God of Iſrael, thy ſeruant
hath heard a bruite, that Saul determineth to come into Ceila, to
deſtroy the citie for me: \V Wil the men of Ceila deliuer me into his
handes? and wil Saul come downe, as thy ſeruant hath heard? Lord God of
Iſrael tel thy ſeruant. And our Lord ſayd: He wil come downe. \V And
Dauid ſaid: Wil the men of Ceila deliuer me, and the men that are with
me, into the handes of Saul? And our Lord ſayd:
\SNote{Conditionally: If thou tarie here.}
They wil deliuer thee. \V Dauid therefore aroſe and his men about ſix
hundred, and going out of Ceila, wandered hither and thither vncertaine:
and it was told Saul that Dauid was fled from Ceila, and was ſaued: for
which cauſe he diſſembled to goe forth. \V But Dauid abode in the deſert
in moſt ſtrong places, and he taried in the mount of the deſert Ziph, in
a ſhadowed hil. Saul notwithſtanding ſought him alwayes: and our Lord
deliuered him not
%%% o-0574
into his handes. \V And Dauid ſaw that Saul was gone forth to ſeeke his
life. Moreouer Dauid was in the deſert Ziph, in a wood. \V And Ionathas
the ſonne of Saul aroſe, and went to Dauid into the wood, and ſtrengthened his
handes in God: and ſayd to him: \V Feare not: for neither ſhal the hand
of Saul my father finde thee, and thou shalt reigne ouer Iſrael, and I
ſhal be ſecond to thee, yea and my father knoweth this. \V Both
therefore made a league before our Lord: and Dauid abode in the wood:
but Ionathas returned into his houſe. \V And the Zeipheites went vp vnto
Saul in Gabaa, ſaying: Loe doth not Dauid lye hid with vs in the moſt
ſafe places of the wood, in the Hil Hachila, which is on the right hand
of the deſert? \V Now therefore, as thy ſoule hath deſired, come downe:
& it ſhal be our charge to deliuer him into the kinges handes. \V And
Saul ſayd: Bleſſed be ye of our Lord, becauſe you haue pitied my
caſe. \V Goe therefore I pray you, and prepare diligently, and deale
curiouſly, and conſider the place, where his foote is, and who hath
ſeene him there, for he thinketh of me, that I craftely lye in waite for
him. \V Conſider and ſee al his lurking holes, wherein he is hid, and
returne to me with the certeintie of the thing, that I may goe with
you. Yea and if he shal ſtoppe vp himſelfe into the earth, I wil ſearch
him out among al the thouſandes of Iuda. \V But they riſing went into
Ziph before Saul: and Dauid and his men
%%% 0641
were in the deſert Maon, in the champaine country at the right hand of
Ieſimon. \V Saul therefore and his companie went to ſeeke him: and it
was told Dauid, and forthwith he went downe to the rocke, and abode in
the deſert Maon, which when Saul had heard, he purſued Dauid in the
deſert Maon. \V And Saul went at the ſide of the mountaine on the one
part: and Dauid and his men were in the ſide of the mountaine on the
other part: moreouer Dauid deſpayred that he could eſcape, from the face
of Saul: Saul therefore and his men in maner of a ring, compaſſed Dauid
and his men, to take them. \V And a meſſenger came to Saul, ſaying: Make
haſt, and come, becauſe the Philiſtijms haue powred in themſelues vpon
the land. \V Saul therefore returned leauing of to purſew Dauid, and he
went to meete the Philiſtians. For this cauſe, they called that place,
the Rocke diuiding.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
