%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 27
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-27


\project douay-rheims


%%% 0648
%%% o-0581
\startChapter[
  title={Chapter 27}
  ]

\Summary{Dauid for more ſecuritie goeth againe to Achis king of Geth,
  5.~obtaineth of him the citie of Siceleg, (6.~by which meanes it
  becometh ſubiect to the kinges of Iuda) 8.~and maketh prayes vpon the
  enemies of King Achis.}

And Dauid ſayd in his hart: at length I shal fal one day into the handes
of Saul: is it not better that I flee, and be ſaued in the Land of the
Philiſtians, that Saul may deſpaire, and ceaſe to ſeeke me in al the
coaſtes of Iſrael? I wil flee therefore his handes. \V And Dauid aroſe,
and went himſelfe, and the ſix hundred men with him, to Achis the ſonne
of Maoch, the King of Geth. \V And Dauid dwelt with Achis in Geth, he
and his men; euerie man & his houſe, and his two wiues, Achinoam the
Iezrahelite, and Abigail the wife of Nabal of Carmel. \V And it was told
Saul that Dauid was fled into Geth, and he added no more to ſeeke
him. \V And Dauid ſayd to Achis: If I haue grace in thy ſight, let there
a place be geuen me in one of thy cities of this countrie, that I may
dwel there: for why abideth thy ſeruant in the citie of the king with
thee? \V Achis therefore gaue him in that day Siceleg; for which cauſe
Siceleg became the kinges of Iuda vntil this day. \V And the number of
the daies, that Dauid dwelt in the country of the Philiſtians, was foure
monethes. \V And Dauid went vp, and his men, and draue prayes out of
\SNote{Theſe countries were neither ſubiect to the Philiſtijms nor to
the Iſraelites, and were alſo of thoſe nations whom God had cõmanded to
deſtroy, dwelling within the land of Chanaan.
\XRef{Deut.~25.}}
Geſſuri, and Gerzi, and from the Amalecites: for theſe villages were
inhabited in the land in old time, as men goe to Sur, as farre as the
Land of Ægypt. \V And Dauid ſtroke al the land, neither left he anie man
or woman: and taking the ſheepe & oxen, and aſſes, & camels, and
garments, he returned & came to Achis. \V And Achis ſayd to him: Whom
haſt thou
%%% 0649
ſet vpon to day? Dauid anſwered: Againſt the ſouth of Iuda, and againſt
the ſouth of Ieremiel, and againſt the ſouth of Ceni. \V Dauid gaue life
neither to man nor woman, neither brought them into Geth, ſaying: Leſt
perhaps they ſpeake againſt vs. Theſe thinges did Dauid, and this was
decreed of him al the daies that he dwelt in the countrie of the
Philiſtians. \V Achis therefore did credite Dauid, ſaying: Manie euils
hath he wrought againſt his people Iſrael: Therefore he ſhal be my
ſeruant for euer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
