%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-22


\project douay-rheims


%%% 0637
%%% o-0571
\startChapter[
  title={Chapter 22}
  ]

\Summary{Dauid with a great retinew goeth to the king of Moab. 5.~But by
  aduiſe of Gad the prophet, returneth into Iuda. 6.~Saul lamenting that
  many conſpire againſt him. 9.~Doeg accuſeth Achimelech, 14.~who
  iuſtifieth both Dauid and himſelf. 16.~He and al the Prieſtes with
  much people in Nobe are ſlaine by Sauls commandment, 20.~onlie
  Abiathar eſcaping fleeth to Dauid.}

Dauid therefore went from thence, and fled to the caue of Odellam. Which
when his brethren had heard, and
\Fix{a}{al}{obvious typo, fixed in other}
his fathers houſe, they went downe to him thither. \V And there were
gathered vnto him, al that were in diſtreſſe, and oppreſſed with debt,
and of a penſiue hart, and he was made their prince, and there were with
him about foure hundred men. \V And Dauid departed from thence into
Maſpha, which is Moab: and he ſaid to the king of Moab: Let my father
and my mother tary with you, I beſech thee, til I know what God wil doe
to me. \V And he left them before the face of the king of Moab, and they
abode with him al the dayes, that Dauid was in garriſon. \V And Gad the
prophet ſaid to Dauid: Tary not in garriſon, depart, and goe into the
Land of Iuda. And Dauid departed, and came into the foreſt of Haret.
%%% 0638
\V And Saul heard that Dauid had appeared, and the men that were with
him. And Saul when he abode in Gabaa, and was in the wood, which is in
Rama, holding a ſpeare in his hand, and al his ſeruantes that ſtood
about him, \V he ſayd to his ſeruantes that ſtoode about him: Heare me
now ye children of Iemini: wil the ſonne of Iſai geue to al you fieldes,
and vineyardes, and make al you tribunes, & centurions: \V becauſe you
haue al conſpired againſt me, and there is none that telleth me,
eſpecially where my ſonne alſo hath entred league with the ſonne of
Iſai? There is none of you, that pitieth my caſe, neither is there that
telleth me: for that my ſonne hath raiſed vp my ſeruant againſt me,
lying in waite for me vntil
%%% o-0572
this day. \V And Doeg the Idumeite which ſtood by, and was the chief
among the ſeruantes of Saul, anſwering: I ſawe, quoth he, the ſonne of
Iſai, in Nobe with Achimelech the ſonne of Achitob the prieſt. \V Who
conſulted our Lord for him; and gaue him victuals, yea and the ſword of
Goliath the Philiſtian he gaue to him. \V The king therefore ſent to cal
for Achimelech the prieſt the ſonne of Achitob, and al his fathers
houſe, the prieſtes that were in Nobe, who came al to the king. \V And
Saul ſaid to Achimelec: Heare thou ſonne of Achitob. Who anſwered: I am
readie, my Lord. \V And Saul ſaid to him: Why haue you conſpired againſt
me, thou, and the ſonne of Iſai, and haſt geuen him bread and a ſword,
and haſt conſulted our Lord for him, that he might riſe vp againſt me,
continuing a traitour vntil this day? \V And Achimelech anſwering the
king, ſaid: And who amongſt al thy ſeruantes faithful as Dauid, and the
kings ſonne in lawe, and going fourth at thy commandmẽt, and glorious in
thy houſe? \V Did I beginne this day to conſult our Lord for him? farre
be this from me: let not the king ſuſpect ſuch a thing againſt his
ſeruant, in al the houſe of my father: for thy ſeruant knew not any
thing concerning this buſines, either litle or great. \V And the king
\Fix{ſaidy: Ding}{ſaid: Dying}{obvious typo, fixed in other}
thou shalt dye Achimelec, thou, and al thy fathers houſe. \V And the
king ſaid to the curtiers, that ſtood about him: Turne your ſelues, and
kil the prieſts of our Lord,
\SNote{Saul vniuſtly condemning Dauid, condemned alſo al thoſe that
iuſtified him.}
for their hand is with Dauid, knowing that he was fled, and they told me
not. And
\SNote{Seing ſo manifeſt iniuſtice and crueltie they obeyed God, rather
then man.}
the kings ſeruantes would not extend their handes vpon the prieſts of
our Lord. \V And the king ſaid to Doeg: Turne thou, and runne vpon the
prieſts. And Doeg the Idumeite being turned, ranne vpon
%%% 0639
the prieſts, and murdered in that day eightie fiue men reueſted with an
ephod of linnen. \V And Nobe the citie of the prieſts, he ſtroke in the
edge of the ſword, men and wemen, and children, and ſucklings, and oxe
and aſſe, and ſhepe in the edge of the ſword. \V But one ſonne of
Achimelech the ſonne of Achitob, eſcaping, whoſe name was Abiathar, fled
to Dauid, \V and told him that Saul had ſlaine the prieſtes of our
Lord. \V And Dauid ſaid to Abiathar: I knew in that day when Doeg the
Idumeit was there, without doubt, he would tel Saul, I am giltie of al
the ſoules of thy father. \V Abide with me, feare not: if any man ſhal
ſeeke my life, he ſhal ſeeke thy life alſo, and with me thou ſhal be
preſerued.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
