%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Kings
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 0593
%%% o-0533
\startChapter[
  title={Chapter 1}
  ]

\Summary{Elcana
\MNote{The firſt part.

Of the gouernments of Heli and Samuel: and of changing the ſtate into a kingdome.}
hauing two wiues, the one called Anna, is barren, and
  for the ſame is reproched by the other, called Phenenna. 9.~Anna
  voweth, and prayeth for a man child, 19.~conceiueth and beareth a
  ſonne, calleth him Samuel: 24.~and preſenteth him to the ſeruice of
  God in Silo.}

There was a man of Ramathaimſophim, of mount Ephraim, & his name Elcana,
the ſonne of Ieroham, the ſonne of Eliu, the ſonne of Thohu, the ſonne
of Suph, an Ephraite: \V and he had two wiues, the name of one was Anna,
and the name of the ſecond Phenenna. Phenenna had children: but Anna had
not children. \V And that man went vp from his citie vpon ordinarie
dayes, to adore and ſacrifice vnto the Lord of hoſtes in Silo. And there
were the two ſonnes of Heli, Ophni and Phinees, prieſtes of our Lord. \V
The day came therfore, & Elcana immolated, and gaue to Phenenna his
wife, & to al her ſonnes and daughters partes: \V but to Anna he gaue
one part with heauie cheere, becauſe he loued Anna. And our Lord had
ſhut her matrice. \V Her aduerſarie alſo afflicted her, and vexed her
ſore, in ſo much
%%% o-0534
that ſhe vpbrayded her, that our Lord had ſhut her matrice: \V and ſo
did she euerie yeare, when the time returned, that they went vp to the
temple of our Lord: and ſo ſhe prouoked her: moreouer she wept, and
tooke not meat. \V Elcana therfore her huſband ſaid to her: Anna, why
weepeſt thou? and why doeſt thou not eate? and wherfore doeſt thou
afflict thy hart? Am not I better to thee, then
%%% 0594
tenne children? \V And Anna aroſe after she had eaten and drunke in
Silo. And Heli the prieſt ſitting vpon a Stoole before the poſtes of the
houſe of our Lord, \V wheras Anna had a heauie hart, she praied to our
Lord, weeping aboundantly, \V and she vowed a vowe, ſaying: O Lord of
hoſtes, if regarding thou wilt behold the affliction of thy ſeruant, and
wilt be mindeful of me, and not forgette thy handmaide, and wilt geue
vnto thy ſeruant a man childe: I
\SNote{This childe being of the tribe of Leui, though not of Aarons
ſtock, was lawfully vowed to the ſeruice of the tabernacle, by his
parentes during his childhood, but coming to yeares of diſcretion he was
at  his owne electiõ to continew, or to depart. If he had bene of anie
other tribe, he muſt haue bene redemed.
\XRef{Leuit.~27.}}
wil geue him to our Lord al the daies of his life, & the raſour shal not
come vpon his head. \V And it came to paſſe, when she multiplied praiers
before our Lord, that Heli obſerued her mouth. \V Moreouer Anna ſpake in
her hart, and onlie her lippes moued, and voice there was not heard at
al. Heli therfore thought her to be drunke, \V and ſayd to her: How long
wilt thou be drunke? digeſt a litle the wyne, wherwith thou art
wette. \V Anna anſwering: Not ſo, quoth she, my lord: for I am an
exceding vnhappie woman, and wine and whatſoeuer may inebriate, I haue
not drunke, but I haue powred out my ſoule in the ſight of our Lord. \V
Account not thy handmaide as one of the daughters of Belial: for of the
multitude of my ſorrow and heauines haue I ſpoken vntil this preſent. \V
Then Heli ſaied to her: Goe in peace: and the God of Iſrael geue thee
thy petition, which thou haſt asked him. \V But she ſayd: Would God thy
handmaide may find grace in thyne eyes. And the woman went on her way,
and did eate, and her countenance was no more changed otherwiſe. \V And
they roſe in the morning, and adored before our Lord: and they returned,
& came into their houſe to Ramatha. And Elcana knew Anna his wife: and
our Lord remembred her. \V And it came to paſſe after a certaine
compaſſe of dayes, Anna conceiued & bare a ſonne, and called his name
Samuel: becauſe she asked him of our Lord. \V And Elcana her husband
went vp, and al her houſe, to
\Fix{immolat}{immolate}{likely typo, fixed in other}
vnto our Lord the ſolemne hoſte, and his vowe, \V and Anna went not vp:
for she ſayd to her huſband: I wil not goe til the infant be weaned, and
til I may bring him, that he may appeare before the ſight of our Lord,
and may remayne there continually. \V And Elcana her huſband ſayd to
her: Doe that which ſeemeth good to thee, and tarie til thou weane him:
and I pray that our Lord fulfil his word. The woman therfore taried, and
gaue her ſonne ſucke, til she remoued him from the milke. \V And she
brought him
%%% 0595
with her, after she had weaned him, with three calues, & three buſhels
of meale, and a flagon of wine, and she brought him to the houſe of our
Lord in Silo. But the childe was yet a litle infant: \V and they
immolated a calfe, and offered the childe to Heli. \V And Anna ſaid: I
beſech thee my
%%% o-0535
lord, thy ſoule liueth my lord: I am that woman, which ſtoode before
thee here praying our Lord. \V For this childe did I pray, and our Lord
hath geuen me my petition, which I asked him. \V Therfore I alſo haue
geuen him to our Lord al the daies, which he shal liue, that he may be
applied to our Lord. And they adored our Lord there. And Anna prayed,
and ſayd:


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
