%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 2450
%%% o-2280
\startChapter[
  title={Chapter 14}
  ]

\Summary{By occaſion of dining with a Phariſee, 2.~after that he hath
  againe confounded them for maligning him for his miraculous good-doing on the
  Sabboth, 7.~he teacheth them humilitie, ſeing their ambition, 12.~and
  in their workes to ſeeke retribution not of men in this world, but of
  God in the world to come: 16.~foretelling alſo that the Iewes for
  their worldly excuſes shal not taſt of the Supper, but the Gentils in
  their place. 25.~Yea that ſo far muſt men be from al worldlines, that
  they muſt earneſtly bethinke them before they enter into his Church,
  and be ready to forgoe al 34.~ſpecially conſidering they muſt be the
  ſalt of others alſo.}

And it came to paſſe when \Sc{Iesvs} entred into the houſe of a certaine
Prince of the Phariſees vpon the Sabboth to eate bread, and they watched
him. \V And behold there was a certaine man before him that had the
dropſie. \V And \Sc{Iesvs} anſwering, ſpake to the Lawyers and
Phariſees, ſaying: Is it Lawful to cure on the Sabboth? \V But they
held their peace. But he taking him, healed him, and ſent him away. \V
And anſwering them he ſaid: Which of you ſhal haue an aſſe or an oxe
fallen into a pit; and wil not incontinent draw him out on the Sabboth
day? \V And they could not anſwer him to theſe things.

\V And he ſpake to them alſo that were inuited a parable, marking how
they choſe the firſt ſeats at the table, ſaying to them: \V When thou
art inuited to a mariage, ſit not downe in the firſt place, leſt perhaps
a more honourable then thou be inuited of him; \V and he that bade thee
and him, come and ſay to thee, Giue this man place: and then thou begin
with ſhame to take the laſt place. \V But when thou art bidden, goe, ſit
downe in the loweſt place: that when he that inuited thee commeth, he
may ſay to thee, Friend ſit vp higher: then ſhalt thou haue glorie
before them that ſit at table with thee. \V becauſe euery one that
exalteth him ſelf, ſhal be humbled: and he that humbleth himſelf, ſhal
be exalted.

%%% o-2281
\V And he ſaid to him alſo that had inuited him: When thou makeſt a dinner or
a ſupper, cal not thy friendes, nor thy brethren, nor kinſmen, nor thy
neighbours that are rich: leſt perhaps they alſo inuite thee againe, and
recompence be made to thee. \V But when thou makeſt a feaſt, cal the
poore, feeble, lame, and blind, \V and thou ſhalt be bleſſed, becauſe
they haue not to recompence thee: for
\SNote{Reward for charitable deedes: and that they may be done for
reward againſt our Aduerſaries.}
recompence ſhal be made thee in the reſurrection of the iuſt. \V When
one of them that ſate at the table with him, had heard theſe things, he
ſaid to him: Bleſſed is he that ſhal eate bread in the Kingdom of God.

\V But he ſaid to him: A certaine man made a great ſupper, and called
many. \V And he ſent his ſeruant at the houre of ſupper to ſay to
the inuited, that they ſhould come, becauſe now al things are ready. \V
And they began al at once to make excuſe. The firſt ſaid to him,
\SNote{Worldlines, wealth, and voluptuouſnes, are the things that
ſpecially hinder men from God.}
I haue
%%% 2451
bought a farme, and I muſt needs goe forth and ſee it; I pray thee hold
me excuſed. \V And an other ſaid, I haue bought fiue yoke of oxen, and I
goe to proue them; I pray thee, hold me excuſed. \V And an other ſaid, I
haue maried a wife, and therfore I can not come. \V And the ſeruant
returning told theſe things to his Lord. Then the Maiſter of the houſe
being angrie, ſaid to his ſeruant: Goe forth quickly into the ſtreetes
and lanes of the citie, and the poore and feeble and blind and lame
bring in hither. \V And the ſeruant ſaid: Lord, it is done as thou didſt
command, and yet there is place. \V And the Lord ſaid to the ſeruant:
Goe forth into the waies and hedges; and
\LNote{Compel them}{The vehement perſuaſion that God vſeth both
externally by force of his word and miracles, and internaly by his
grace, to bring vs vnto him, is called compelling:
\MNote{Free-wil.}
not that he forceth any to come to him againſt their owne willes, but
that he can alter & mollifie an hard hart, and make him willing that
before would not.
\CNote{\Cite{ep.~50. paulo poſt princip.}
&
\Cite{ep.~204.}
&
\Cite{li.~1. cont. ep. Gaudent. c.~25.}}
\MNote{Heretikes may by penal lawes be compelled to the Catholike
faith.}
S.~Auguſtine alſo referreth this compelling to the
penal lawes which Catholike Princes doe iuſtly vſe againſt Heretikes and
Schiſmatikes, prouing that they who are by their former profeſſiõ in
Baptiſme ſubiect to the Catholike Church, & are departed frõ the ſame
after Sectes, may and ought to be compelled into the vnitie and ſocietie
of the vniuerſal Church againe. And therfore in this ſenſe, by the two
former partes of the Parable, the Iewes firſt, and ſecondly the Gentils,
that neuer beleeued before in Chriſt, were inuited by faire meanes only:
but by the third, ſuch as are inuited as the Church of God hath power
ouer, becauſe they promiſed in Baptiſme, and therfore are to be reuoked
not only by gentle meanes, but by iuſt punishment alſo.}
compel them to enter, that my houſe may be filled. \V But I ſay to you,
that none of thoſe men that were called, ſhal taſt my ſupper.

\V And great multitudes went with him; and turning, he ſaid to them: \V
\CNote{\XRef{Mt.~10,~37.}}
If any man come to me and hateth not his
\SNote{No creature ſo deere vnto vs, which we muſt not hate or forſake,
if it hinder vs, and in that reſpect that it hindereth vs from Chriſt,
or his Church, & our Saluation.}
father and mother, and wife and children, and brethren and ſiſters, yea
and his owne life beſides; he can not be my Diſciple. \V And he that
doth not beare his croſſe and come after me; can not be my Diſciple. \V
For, which of you minding to build a toure, doth not firſt ſit downe and
reckon the charges that are neceſſarie, whether he haue to finiſh it: \V
leſt, after that he hath laid the foundation, and is not able to finish
it, al that ſee it, begin to mocke him, \V ſaying, That this man began
to build, and he could not finiſh it? \V Or
%%% o-2282
what King about to goe to make warre againſt an other King, doth not
firſt ſit downe and thinke whether he be able with ten thouſands to
meete him that with twentie thouſands commeth againſt him? \V Otherwiſe
whiles he is yet farre off, ſending a legacie he asketh thoſe things
that belong to peace. \V So therfore euery one of you that doth not
\SNote{He that is a right Chriſtian man, muſt make his account that if
he be put to it (as he often may be in times of perſecution) he muſt
renounce al that euer he hath, rather then forſake the Catholike faith.}
renounce al that he poſſeſſeth, can not
\Fix{by}{be}{obvious typo, fixed in other}
my
\Fix{diſciple.}{Diſciple.}{likely typo, same in other}

\V
\CNote{\XRef{Mt.~5,~13.}
\XRef{Mar.~9,~50.}}
Salt is good. But if the ſalt leeſe his vertue, wherewith ſhal it be
ſeaſoned? \V It is profitable neither for the ground, nor for the
dunghil, but it ſhal be caſt forth. He that hath eares to heare, let him
heare. 




\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
