%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 2420
%%% o-2245
\startChapter[
  title={Chapter 4}
  ]

\Summary{Chriſt going into the Deſert to prepare himſelf before his
  manifeſtation, ouercometh the tentations of the Diuel. 24.~then
  beginning gloriouſly in Galilee. 16.~he sheweth to them of Nazareth
  his commiſsion out of Eſay the Prophet. 23.~inſinuating by occaſion
  the Iewes his countriemens reprobation. 31.~In Capharnaum his doctrine
  is admired, 33.~ſpecially for his miracle in the Synagogue. 38.~from
  which going to Peters houſe, he sheweth there much more
  power. 42.~Then retiring into the wilderneſſe, he preacheth afterward
  to the other cities of Galilee.}

And
\CNote{\XRef{Mr.~4,~1.}
\XRef{Mt.~1,~12.}}
\Sc{Iesvs} ful of the Holy Ghoſt, returned from Iordan, & was driuen
in the ſpirit into the deſert. \V
\SNote{The Churches faſt of 40.\ daies (called Lent) commeth of this, &
is an Apoſtolical Tradition.
\Cite{Clem. Conſtit. Apoſt. li.~5. c.~13.}
\Cite{Hier. ep. ad Marcel. adu. erro. Mötani.}
\Cite{Leo ſer.~6. et.~9. de Quadrageſ.}}
fourtie daies, and was tempted of the Diuel. And he did eate nothing in
thoſe daies; and when they were ended, he was an hungred. \V And the
Diuel ſaid to him: If thou be the Sonne of God, ſay to this ſtone that
it be made bread. \V And \Sc{Iesvs} made anſwer vnto him: It is
written,
\CNote{\XRef{Deu.~8,~3.}}
\Emph{That not in
%%% 2421
bread alone shal man liue, but in euery word of God.} \V And the Diuel
brought him into an high mountaine, and shewed him al the Kingdoms of
the whole world in a moment of time; \V and he ſaid to him: To thee wil
I giue this whole power, and the glorie of them; for to me they are
deliuered, and to whom I wil, I doe giue them. \V Thou therfore if thou
wilt adore before me, they ſhal al be thine. \V And \Sc{Iesvs} anſwering
ſaid to him: It is written,
\CNote{\XRef{Deu.~6,~13.}
\XRef{10,~20.}}
\SNote{See the Annot.\ in
\XRef{S.~Matthew c.~4,~11.}}
\Emph{Thou shalt adore the Lord thy God & him only shalt thou
\TNote{\G{λατρεύσεις}}
ſerue.} \V And he brought him into Hieruſalem, & ſet him vpon the
pinnacle of the Temple; and he ſaid to him: If thou be the Sonne of God,
caſt thy ſelf from hence downward. \V For
\SNote{If the Diuel himſelf alleage Scripture againſt Chriſt, no maruel
that Heretikes doe ſo againſt Chriſts Church.}
it is written, that
\CNote{\XRef{Pſa.~90,~11.}}
\Emph{He hath giuen his Angels
%%% o-2246
charge of thee, that they preſerue thee:} \V and that \Emph{in their
hands they shal beare thee vp, leſt perhaps thou knock thy foote againſt
a ſtone.} \V And \Sc{Iesvs} anſwering ſaid to him: It is ſaid,
\CNote{\XRef{Deut.~6,~16.}}
\Emph{Thou shalt not tempt the Lord thy God.} \V And al the
tentation being ended, the Diuel
\LNote{Departed vntil a time}{No
\MNote{The Diuels tentations.}
maruel if the Diuel be often or alwaies buſie with Chriſtian men, ſeeing
after he was plainely ouercome by Chriſt, yet did he not giue him ouer
altogether, but for a time.}
departed from him vntil a time.

\V
\CNote{\XRef{Mt.~4,~12.}
\XRef{Mr.~1,~14.}}
\MNote{The third part: of Chriſtes manifeſting himſelf by preaching and
miracles, ſpecially in Galilee.}
And \Sc{Iesvs} returned in the force of the Spirit into Galilee, and
the fame went forth through the whole countrie of him. \V And he taught
in their Synagogues, and was magnified of al.

\V
\CNote{\XRef{Mt.~13,~54.}
\XRef{Mr.~6,~1.}
\XRef{Io.~4,~43.}}
And he came to Nazareth where he was brought vp: and he entred
\SNote{Our Sauiour vſed to preach in their Synagogues.}
according to his cuſtom on the Sabboth day into the Synagogue; and he
roſe vp to read. \V And the booke of Eſay the Prophet was deliuered vnto
him. And as he vnfolded the booke, he found the place where it was
written: \V
\CNote{\XRef{Eſa.~61.~1.}}
\Emph{The Spirit of the Lord vpon me, for which he anointed
me, to euangelize vnto the poore he ſent me, to heale the contrite of
hart, \V to preach to the captiues remiſsion, and ſight to the blind, to
dimiſſe the bruiſed vnto remiſsion, to preach the acceptable yeare of
the Lord, and the day of retribution.} \V And when he had folded the
booke, he rendred it to the miniſter, and ſate downe. And the eyes of al
in the Synagogue were bent vpon him. \V And he began to ſay vnto them:
That this day is fulfilled this ſcripture in your eares. \V And al gaue
teſtimonie to him; and they
\SNote{He had a maruelous grace, and an extraordinarie force in mouing
the harts of his hearers.}
marueled in the words of grace that proceeded from his mouth, and they
ſaid: Is not this Ioſephs ſonne? \V And he ſaid to them: Certes you wil
ſay to me this ſimilitude, \Emph{Phyſicion, cure thy ſelf}: as great
things as we haue heard
\LNote{Done in Capharnaum}{God
\MNote{Miracles at one place and not at another.}
maketh choiſe of perſons and places where he worketh miracles or doeth
benefits, though he might doe the ſame elſwhere, if it liked his
wiſedom. So doth he in doing miracles by Saints, not in al places, nor
towards al perſons, but as it pleaſeth him.
\Cite{Aug. ep.~137.}}
done in Capharnaum, doe alſo here in thy countrie. \V And he ſaid: Amen
I ſay to you, that no Prophet is accepted in his owne countrie. \V In
truth I ſay to you,
\CNote{\XRef{3.~Reg.~7,~9.}}
there were many widowes in the daies of Elias in
Iſrael, when the Heauen was ſhut three yeares and ſix moneths, when
there was a great famine made in the whole earth: \V And to none of them
was Elias ſent, but into Sarepta of Sidon, to a widow woman. \V
\CNote{\XRef{4.~Reg.~5,~14.}}
And
there were many lepers in Iſrael vnder Eliſæus the Prophet; and none of
them was made cleane by Naaman the Syrian. \V And al in the Synagogue
were filled with anger, hearing theſe things. \V And they roſe, and caſt
him out of the citie: and they brought him to the edge of the hil, where
vpon their citie was built, that they might throw him downe
headlong. But he
\LNote{Paſsing through the middes of them}{Either
\MNote{Chriſts body conteined in place aboue nature.}
by making himſelf inuiſible, or alſo more wonderfully, penetrating the
multitude and paſsing through them, as he did through the doore, his
body either being without ſpace of place, or with other bodies in one
place. By al which & the like doings mentioned in the Ghoſpel, it is
euident that he can alter and order his body as he liſt, aboue the
natural conditions of a body.}
paſsing through the middes of them, went his way.

%%% 2422
%%% o-2247
\V
\CNote{\XRef{Mt.~4,~13.}
\XRef{7,~28.}
\XRef{Mt.~1,~21.}}
And he went downe into Capharnaum a citie of Galilee; and there he
taught thẽ on the Sabboths. \V And they were aſtoniſhed at his doctrine;
becauſe his talke was in power. \V And in the Synagogue there was a man
hauing an vncleane Diuel, and he cried out with a loud voice. \V Saying:
Let be, what to vs and thee \Sc{Iesvs} of Nazareth? art thou come to
deſtroy vs? I know thee who thou art, \Emph{the Sainct of God}. \V
And \Sc{Iesvs}, rebuked him, ſaying: Hold thy peace, & goe out of
him. And when the Diuel had throwen him into the middes, he went out of
him, and hurt him nothing. \V And there came feare vpon al, and they
talked togeather one with another, ſaying: What word is this, that in
power and vertue he commandeth the vncleane Spirits, and they goe
out? \V And the fame of him was published into euery place of the
countrie.

\V And \Sc{Iesvs} riſing vp out of the Synagogue, entred into Simons
houſe.
\CNote{\XRef{Mat.~8,~14.}
\XRef{Mr.~1,~30.}}
And
\LNote{Simons wiues mother}{It
\MNote{The Apoſtles left their wiues.}
is euident that Peter had a wife, but after his calling to be an
Apoſtle, he left her, as S.~Hierom writeth in many places
\Cite{ep.~34. c.~2. ad Iulianum.}
\Cite{Li.~1. adu. Ionin.}
See the 
\XRef{Annot. Matth.~19,~29.}}
Simons wiues mother was holden with a great feuer; and they beſought him
for her. \V And ſtanding ouer her, he commanded the feuer, and it left
her. And incontinent riſing, she miniſtred to them. \V And when the
ſunne was downe, al that had diſeaſed of ſundrie maladies, brought them
to him. But he
\SNote{See
\XRef{Mar.~8,~22. in margine.}}
impoſing hands vpon euery one, cured them. \V And Diuels went out from
many, crying and ſaying: That thou art the Sonne of God. And rebuking
them he ſuffred them not to ſpeake, that they knew he was Chriſt.

\V And when it was day, going forth he went into a deſert place; & the
multitudes ſought him, and came euen vnto him; and they held him that he
should not depart from them. \V To whom he ſaid: That to other cities
alſo muſt I euangelize the Kingdom of God; becauſe therfore I was
ſent. \V And he was preaching in the Synagogues of Galilee.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
