%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 2423
%%% o-2248
\startChapter[
  title={Chapter 5}
  ]

\Summary{Hauing taught the People out of Peters ship, 4.~he sheweth in a
  miraculous taking of fishes how he wil make him the fisher of
  men. 12.~He cureth a leper by touching him, and ſendeth him to the
  Prieſt in witneſſe that he is not againſt Moyſes. 15.~The People
  flocking vnto him, he retireth into the wilderneſſe. 17.~To the
  Phariſees in a ſolemne aſſembly he proueth by a miracle his power to
  remit ſinnes in earth. 27.~He defendeth his eating with ſinners, as
  being the Phyſicion of ſoules. 33.~and his not preſcribing as yet of any
  faſtes to his Diſciples.}

And it came to paſſe, when the multitudes preſſed vpon him to heare the
word of God, and him ſelf ſtood beſide the lake of Geneſareth. \V
\CNote{\XRef{Mt.~4,~18.}
\XRef{Mr.~1,~16.}}
And he ſaw two ſhips ſtanding by the lake: and the fiſhers were gone
downe, and waſhed their nets. \V And he going vp into
\LNote{One ship Simons}{It
\MNote{Peters ship.}
is purpoſely expreſſed that there were two ships, and that one of them
was Peters, and that Chriſt went into that one, and ſate downe in it,
and that ſitting he taught out of that ſhip: no doubt to ſignifie the
Church reſembled by Peters ſhip, and that in it is the chaire of Chriſt,
and only true preaching.}
one ſhip that was Simons, deſired him to bring it back a litle from the
land. And ſitting, he taught the multitudes out of the ship.

\V And as he ceaſed to ſpeake, he ſaid to Simon: Launch forth into the
deep, and let looſe your nets to make a draught. \V And Simon anſwering,
ſaid to him: Maiſter, labouring al the night, we haue taken nothing, but
in thy word I wil let looſe the net. \V And when they had done this,
they incloſed
\LNote{A great multitude of fishes}{Likewiſe
\MNote{Peters fishing.}
by this ſignificatiue miracle wrought about Peters fiſhing, is euidently
forſhewed what wonderful ſucceſſe Peter ſhould haue in conuerting men to
Chriſt, both Iewes and Gentils: as when at one draught, that is to ſay,
\CNote{\XRef{Act.~2,~41.}
\XRef{4,~4.}}
at one Sermon he drew into his ſhip, which is Chriſtes Church, a great
number of men, as he did now fiſhes: and ſo continually by himſelf and
his Succeſſours vnto the worlds end.}
a very great multitude of fiſhes, and their net was broken. \V And they
\LNote{Beckened to their fellowes}{Peter
\MNote{Peters coadiutours.}
had ſo much worke that he called for help & ioyned vnto him the other
ſhip repreſenting to vs his Copartners in the preaching of the Ghoſpel,
and the coniunction of the Synagogue and the People of Gentilitie vnto
Peters ſhip, that is, to the Church of Chriſt.
\Cite{Ambro. li.~4. in Luc. c.~vlt.}}
beckned to their fellowes that were in the other ſhip, that they ſhould
come and help them. And they came and filled both ſhips, ſo that they
did ſinke. \V Which when Simon Peter did ſee, he fel downe at \Sc{Iesvs}
knees, ſaying: Goe forth from me, becauſe I am a ſinful man, O Lord. \V
For he was wholy aſtoniſhed and al that were with him, at the draught of
fiſhes which they had taken. \V In like manner alſo Iames and Iohn the
ſonnes of Zebedee, who were Simons fellowes. And \Sc{Iesvs} ſaid to
Simon: Feare not; from this time now,
\LNote{Thou ſhalt be taking men}{That
\MNote{Peters preeminẽce in fishing for mẽs ſoules.}
al this aforeſaid did properly meane Peters trauailles to come, in the
cõuerſion of the world to Chriſt, and his prerogatiue before al men
therin, it is euident by Chriſts ſpecial promiſe made to him ſeuerally
and apart in this place, that he should be made the taker of men, though
to other he giueth alſo, as to Peters cooperatours and coadiutours the
like office.
\XRef{Mat.~4,~19.}}
thou ſhalt be taking men. \V And hauing brought their ſhips to land,
leauing al things they folowed him.

\V
\CNote{\XRef{Mt.~8,~2.}
\XRef{Mr.~1,~40.}}
And it came to paſſe, when he was in one of the
%%% o-2249
cities, & behold a man ful of leproſie, and ſeeing \Sc{Iesvs}, and
falling on his face, beſought him ſaying: Lord, if thou wilt, thou canſt
make me cleane. \V And ſtretching forth the hand he
\SNote{Our Sauiour often
\Fix{health}{healeth}{likely typo, but note missing in other}
by touching.}
touched him, ſaying: I wil. Be thou made cleane. And immediatly the
leproſie departed from him. \V And he commanded him that he should tel
no body, but, Goe,
\SNote{See
\XRef{S.~Mat. Annot. c.~8,~4.}}
shew thy ſelf to the Prieſt, and offer for thy cleanſing
\CNote{\XRef{Leu.~14,~2.}}
as Moyſes commanded, for a teſtimonie to them.

\V But the bruit of him went abrode the more, and great multitudes came
togeather to heare, and to be cured of their infirmities. \V And he
retired into the deſert, and praied.

\V
\CNote{\XRef{Mt.~9,~2.}
\XRef{Mr.~2,~3.}}
And it came to paſſe one day, and he ſate teaching. And there were
Phariſees ſitting and Doctours of Law that were come out of euery towne
of Galilee and Iewrie and Hieruſalem; and the vertue of our
%%% 2424
Lord was to heale them. \V And behold men carying in a bed a man that
had the palſey: and they ſought to bring him in, and to lay him before
him. \V And not finding on which ſide they might bring him in for the
multitude, they
\LNote{Went vp vpon the roofe}{A
\MNote{Zeale of ſoules.}
ſtrange diligence in procuring corporal health of & by Chriſt: and an
example for vs of the like or greater, to obtaine ſaluation of him
either for our ſelues or our friends, and to ſeeke to his Church &
Sacraments with what extraordinarie paine ſoeuer.}
went vp vpon the roofe, and through the tiles let him downe with the bed
into the middes, before \Sc{Iesvs}. \V
\LNote{Whoſe faith}{Great
\MNote{The interceſſiõ of others.}
is God (ſaith S.~Ambroſe) and pardoneth one ſort through the merites of
others. Therfore if thou doubt to obtaine forgiueneſſe of thy great
offenſes, ioyne vnto thy ſelf interceſſours, vſe the Churches help,
which may pray for thee and obtaine for thee that which our Lord might
denie to thyſelf.
\Cite{Amb. li.~5. in Luc.}}
Whoſe faith when he ſaw, he ſaid: Man, thy ſinnes are forgiuen thee. \V
And the Scribes and Phariſees began to thinke, ſaying: Who is this that
ſpeaketh blaſphemies? Who can forgiue ſinnes, but only God? \V And
when \Sc{Iesvs} knew their cogitations, anſwering he ſaid to them: What
doe you thinke in your hartes? \V Which is eaſier to ſay, Thy ſinnes are
forgiuen thee; or to ſay, Ariſe, and walke? \V But that you may know
that
\LNote{The Sonne of man in earth}{By
\MNote{Prieſts doe remit ſinnes.}
which act
\CNote{\Cite{In catena S.~Thomæ.}}
(ſaith S.~Cyril) it is cleere that the Sonne of man hath power
in earth to remit ſinnes; which he ſaid both for himſelf and vs. For he,
as God being made man & Lord of the Law, forgiueth ſinnes. And we alſo
haue obtained by him that wõderful grace. For it is ſaid to his
Diſciples:
\CNote{\XRef{Io.~20,~23.}}
\Emph{Whoſe ſinnes you shal remit, they are remitted to
them.} And how should not he be able to remit ſinnes, who gaue others
power to doe the ſame?}
the Sonne of man hath power in earth to forgiue ſinnes (he ſaid to the
ſick of the palſey) I ſay to thee, Ariſe, take vp thy bed, and goe into
thy houſe. \V And forthwith riſing vp before them, he tooke that wherein
he lay; and he went into his houſe, magnifying God. \V And al were
aſtonished: and they magnified God. And they were repleniſhed with
feare, ſaying: That we haue ſeen maruelous things today.

\V
\CNote{\XRef{Mt.~9,~9.}
\XRef{Mr.~2,~14.}}
And after theſe things he went forth, and ſaw a Publican called Leui,
ſitting at the Cuſtome-houſe, and he ſaid to him: Follow me. \V And
\LNote{Leauing al followed him}{The
\CNote{\Cite{Hiero. in Mt.~9.}}
\MNote{Forſaking al, and following Chriſt.}
profane Iulian charged Matthew of too much lightnes, to leaue al and
follow a ſtranger, at one word. But indeed hereby is ſeen the maruelous
efficacie of Chriſts Word and internal working, that in a moment can
alter the hart of a man, and cauſe him nothing to eſteeme the things
moſt deere vnto him. Which he did not only then in preſence, but alſo
daily in the Church. For ſo
\CNote{\Cite{Athan. in Vit. S.~Antonij.}
\Cite{Auguſt. confeſ. li.~2. c.~12.}}
S.~Antonie,
\CNote{\Cite{Bonau. in vit. S.~Frãciſci.}}
S.~Francis, and others, by
hearing only the Word of our Sauiour read in the Church forſooke al, and
followed him.}
leauing al things, he roſe and followed him. \V And Leui made him a
great feaſt in his houſe; and there was a great multitude of Publicans;
and of others that were ſitting at the table with them. \V And their
Phariſees and
%%% o-2250
Scribes murmured, ſaying to his diſciples: Why doe you eate and drinke
with Publicans and ſinners? \V And \Sc{Iesvs} anſwering ſaid to them:
They that are whole, need not the Phyſicion: but they that are il at
eaſe. \V
\SNote{Chriſt came not to cal thoſe who preſume of their owne iuſtice,
and that count themſelues to haue no need of Chriſt.}
I came not to cal the iuſt, but ſinners to pennance.

\V But they ſaid to him:
\CNote{\XRef{Mat.~9,~14.}
\XRef{Mar.~2,~18.}
\XRef{Lu.~5,~33.}}
Why doe the Diſciples of Iohn
\SNote{See
\XRef{S.~Mat. Annot. c.~9,~14.}}
faſt often, and make obſecrations, and of the Phariſees in like manner;
but thine doe eate and drinke? \V To whom he ſaid: Why, can you make the
children of the bridegroome faſt whiles the bridegroome is with them? \V
But the daies wil come; and when the bridegroome ſhal be taken away from
them, then they ſhal faſt in thoſe daies. \V And he ſaid a ſimilitude
alſo vnto them: That no man putteth a peece from a new garment into an
old garment; otherwiſe both he breaketh the new, and the peece from the
new agreeth not with the old. \V And no bodie putteth new wine into old
bottels; otherwiſe the new wine wil breake the bottels, and it ſelf wil
be ſhed, and the bottels wil be loſt. \V But new wine is to be put into
new bottels: and both are preſerued togeather. \V And no man drinking
old, wil new by and by, for he ſaith, The old is better.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
