%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 2435
%%% o-2262
\startChapter[
  title={Chapter 9}
  ]

\Summary{His Twelue alſo now preaching euery where and working
  miracles. 6.~Herod and al doe wonder much. 10.~After which, he taketh
  them and goeth into the wilderneſſe: where he cureth and teacheth,
  feeding 5000.\ with fiue loaues. 18.~Peter confeſsing him to be
  Chriſt, 22.~he on the other ſide foretelleth his Paſsion, and that al
  muſt in time of perſecution follow him therein. 27.~whereunto to
  encourage vs the more, 27.~he giueth in his Transfiguration a ſight of
  the glorie, which is the reward of ſuffering.
  \Fix{17.~The}{37.~The}{Obvious typo, same in other}
  next day he caſteth out a Diuel which his Diſciples could not.
  43.~whom amiddes theſe wonders he forewarneth againe of his ſcandalous
  Paſsion. 49.~And to cure their ambition, he telleth them, that the
  moſt humble he eſteemeth moſt: 49.~bidding them alſo not to prohibit
  any that is not againſt them. 51.~Yea and toward ſuch as be againſt
  them Schiſmatically, to shew mildnes for al that. 57.~Of following
  him, three examples.}

And
\CNote{\XRef{Mt.~10,~1.}
\XRef{Mr.~3,~13.}
\XRef{6,~8.}}
calling together the twelue Apoſtles, he gaue them
\SNote{To command Diuels and diſeaſes either of body or ſoule, is by
nature proper to God only: but by Gods guift, men alſo may haue the
ſame: euen ſo to forgiue ſinnes.}
vertue and power ouer al Diuels, and to cure maladies. \V And he ſent them
to preach the Kingdom of God; and to heale the ſicke. \V And he ſaid to
them: Take nothing for the way, neither rod, nor skrip, nor bread, nor
money, neither haue two coates. \V And into whatſoeuer houſe you enter,
tarie there, and thence doe not depart. \V And whoſoeuer shal not
receiue you, going forth out of that citie,
%%% o-2263
shake off the duſt alſo of your feet
\SNote{A great fault to reiect the true Preachers, or not to admit them
into houſe for needful harbour and ſuſtenace.}
for a teſtimonie vpon them. \V And going forth they went a circuit from
towne to towne euangelizing and curing euery where.

\V And
\CNote{\XRef{Mt.~14,~1.}
\XRef{Mar.~6,~14.}}
Herod the Tetrach heard al things that were done by him; and he
ſtaggered becauſe it was ſaid of ſome, That Iohn was riſen from the
dead. \V But of other ſome, That Elias hath appeared; and of others,
that a Prophet one of the old ones was riſen. \V And Herod ſaid: Iohn I
haue beheaded; but who is this of whom I heare ſuch things? And he
ſought for to ſee him.

\V And
\CNote{\XRef{Mt.~14,~13.}
\XRef{Mar.~6,~31.}
\XRef{Io.~6,~5.}}
the Apoſtles being returned, reported to him whatſoeuer they did:
and taking them he retired apart into a deſert place, which belongeth to
Beth-ſaida. \V Which the multitudes vnderſtãding, followed him & he
receiued them, and ſpake to them of the Kingdom of God, and them that
had need of cure he healed. \V And the day began to draw towards an
end. And the Twelue comming neere, ſaid to him: Dimiſſe the multitudes,
that going into townes and villages here about, they may haue lodging,
and find meates; becauſe here we are in a deſert place. \V And he ſaid
to them: Giue you them to eate. But they ſaid: We
%%% 2436
haue no more but fiue loaues and two fiſhes; vnles perhaps we should goe
& buie meates for al this multitude. \V And there were men almoſt fiue
thouſand. And he ſaid to his Diſciples: Make them ſit downe by companies
fiftie and fiftie. \V And ſo they did. And they made al ſit downe. \V
And taking the fiue loaues and the two fishes, he looked vp vnto Heauen,
and 
\TNote{\G{εὐλόγησεν αὐτοὺς}}
\SNote{Here you ſee that he bleſſed the things, and not only gaue thanks
to God. See
\XRef{Annot. Marci. c.~8,~7.}}
bleſſed them; and he brake, and diſtributed to his Diſciples, for to ſet
before the multitudes. \V And
\SNote{The miraculous prouidence of God toward ſuch as follow Chriſt
into deſerts, priſone, banishment, or whitherſoeuer.}
they did al eate, and had their fil. And there was taken vp that which
remained to them, twelue baskets of fragments.

\V
\CNote{\XRef{Mt.~16,~13.}
\XRef{Mar.~8,~27.}}
And it came to paſſe; when he was alone praying, his Diſciples alſo
were with him; and he asked them ſaying: Whom doe the multitudes ſay
that I am? 
\Fix{}{\V }{Missing verse number, fixed in other}% to prevent a paragraph break
But they anſwered, and ſaid: Iohn the Baptiſt; and ſome,
Elias; but ſome, that one of the Prophets before time is riſen. \V And
he ſaid to them: But whom ſay ye that I am? Simon Peter anſwering, ſaid:
The Chriſt of God. \V But he rebuking them, commanded that they should
tel this to no man, \V ſaying: That the Sonne of man muſt ſuffer many
things, and be reiected of the Ancients and cheefe Prieſts and Scribes,
and be killed, and the
%%% o-2264
third day riſe againe.

\V And he ſaid to al: If any man wil come after me, let him denie
himſelf, and take vp his croſſe daily, and follow me. \V For he that wil
ſaue his life, shal loſe it; for he that shal loſe his life for my ſake,
shal ſaue it. \V For what profit hath a man if he gaine the whole world,
and loſe himſelf, and caſt away himſelf? \V For he that shal be ashamed
of me and of my wordes, him the Sonne of man shal be ashamed of, when he
shal come in his maieſtie, and his Fathers, and of the holy Angels. \V
And I ſay to you aſſuredly: There be ſome ſtanding here that shal not
taſt death,
\LNote{Til they ſee}{To
\MNote{The Transfiguration.}
the Apoſtles, that had to preach the Kingdom of God and to ſuffer ſo
much miſerie for the ſame in this world, he would shew his glorie, and
giue them a taſt of his owne ioyful ſtate and of his Saints in Heauen,
calling thither Moyſes and Elias, that the Law & Prophets might be
witneſſes of the ſame. See the
\XRef{annotation vpon S.~Matthew c.~17,~2.}}
til they ſee the Kingdom of God.

\V
\CNote{\XRef{Mt.~17,~1.}
\XRef{Mr.~9,~1.}}
\MNote{The \Sc{Transfigvration.}}
And it came to paſſe after theſe wordes almoſt eight daies, and he tooke
Peter and Iames and Iohn, and went into a mountaine to pray. \V And
whiles he prayed, the shape of his countenance was altered: and his
raiment white and gliſtering. \V And behold two men talked with him. And
they were Moyſes and Elias, \V appearing in maieſtie. And they told his
deceaſe that he should accomplish in Hieruſalem. \V But Peter and they
that were with him, were heauie with ſleepe. And awaking, they ſaw his
maieſtie, and the two men that ſtood with him. \V And it came to paſſe,
when they departed from him, Peter ſaid to \Sc{Iesvs}: Maiſter, it is
good for vs to be here; and let vs make three tabernacles, one for thee,
and one for Moyſes, and one for Elias: not knowing what he ſaid. \V And
as he ſpake theſe things, there came a cloud, and ouerſhadowed them: and
they feared, when they entered into the cloud. \V
\CNote{\XRef{2.~Pet.~1,~17.}}
And a voice was made
out of the cloud, ſaying: This is my beloued Sonne, heare him. \V And
whiles the voice was made, \Sc{Iesvs} was found alone. And they held
their peace, and told no man in thoſe daies any of theſe things which
they had ſeen.

\V
\CNote{\XRef{Mt.~17,~14.}
\XRef{Mr.~9,~17.}}
And it came to paſſe the day folowing, when they came downe from the
mountaine, there met him a great multitude. \V And behold a man of the
multitude cried out, ſaying: Maiſter, I beſeech thee, looke vpon my
ſonne becauſe he is mine only one. \V And loe, the ſpirit
%%% 2437
taketh him, and he ſodenly crieth, and he dasheth him, and teareth him
that he fometh, and with much adoe departeth renting him. \V And I
deſired thy Diſciples to caſt him out, and they could not. \V
And \Sc{Iesvs} anſwering ſaid:
\SNote{Increduitie hindereth the effect of Exorciſmes, and other
miraculous power giuen to the Church.}
O faithles and peruerſe
%%% o-2265
Generation, how long shal I be with you and ſuffer you? bring hither thy
ſonne. \V And when he came to him, the Diuel dashed, and tore
him. And \Sc{Iesvs} rebuked the vncleane Spirit, and healed the boy, and
rendred him to his father. \V And al were aſtoniſhed at the might of
God; and al merueling at al things that he did, he ſaid to his Diſciples: \V
Lay you in your hartes theſe wordes, for it shal come to paſſe that the
Sonne of man shal be deliuered into the hands of men. \V But they did
not know this word, and it was couered before them, that they perceiued
it not. And they were afraid to aske him of this word.

\V
\CNote{\XRef{Mt.~18,~1.}
\XRef{Mar.~9,~34.}}
And there entred
\SNote{Deſire of preeminence is an humane infirmitie often euen amõg the
good. Againſt which, Chriſt teacheth humility, but forbiddeth not
Superiority.}
a cogitation into them, which of them should be greater. \V
But \Sc{Iesvs} ſeeing the cogitations of their hart, tooke a child and
ſet him by him, \V and ſaid to them, whoſoeuer receiueth this childe in
my name, receiueth me, and whoſoeuer receiueth me, receiueth
him that ſent me. For he that is the leſſer among you al, he is the
greater.

\V
\CNote{\XRef{Mar.~9,~39.}}
And Iohn anſwering ſaid: Maiſter, we ſaw a certaine man caſting out
Diuels in thy name, and we prohibited him, becauſe he followeth not with
vs. \V And \Sc{Iesvs} ſaid to him:
\SNote{There be ſome that follow not Chriſt preciſely in life and
doctrine, of whom we may make our aduantage to the propogation of
Chriſtes honour and religion, when they doe any thing for the
aduancement thereof, of what intention ſoeuer they doe it.
\XRef{Philip. 1,~25.}}
Prohibit not. For he that is not againſt you, is for you.

\V And it came to paſſe, whiles the daies of his aſſumption were
accompliſhing, and he fixed his face to goe into Hieruſalem. \V And he
ſent meſſengers before his face; and going they entred into a citie of
the Samaritans to prepare for him. \V And they receiued him not, becauſe
his
\LNote{Face to go to Hieruſalem}{The
\CNote{\XRef{Io.~4,~9.}
\XRef{3.~Reg.~12,~17.}
\XRef{Tob.~1.}}
\MNote{Schiſmatikes.}
Samaritans were Schiſmatikes from the Iewes, and had a Schiſmatical
Temple in mount Garizim, of purpoſe to draw men thither from Gods Temple
in Hieruſalem, where only was the true and as it were the Catholike ſeruice
and Sacrifice vnto God. Therfore they did not gladly receiue our
Sauiour, becauſe they perceiued he was going to Hieruſalem.}
face was to goe to Hieruſalem. \V And when his Diſciples Iames and Iohn
had ſeen it, they ſaid: Lord wilt thou we ſay that fire come downe from
Heauen and conſume
\Var{them?}{them as Elias alſo did?}
\V And turning,
\LNote{He rebuked them}{Not
\MNote{Deſire of reuenge.}
iuſtice nor al rigorous punishment of ſinners is here forbidden, nor
Elias fact
reprehended, nor the Church or Chriſtian Princes blamed for putting
Heretikes to death: but that none of theſe should be done for deſire of
our particular reuenge, or without diſcretion, & regard of their
amendement, and example to others.
\CNote{\XRef{Act.~5.}}
\MNote{The Churches ſeueritie.}
Therfore S.~Peter vſed his power vpon Ananias and Saphira, when he
ſtroke them both downe to death for defrauding the Church.}
he rebuked them, ſaying: You know not of what ſpirit you are. \V The
Sonne of man came not to deſtroy ſoules, but to ſaue. And they went into
another towne.

\V And it came to paſſe as they walked in the way, a certaine man ſaid
to him:
\CNote{\XRef{Mt.~8,~19.}}
I wil follow thee whitherſoeuer thou goeſt. \V \Sc{Iesvs} ſaid to him:
\SNote{This man would haue followed him for temporal commodities, and
therfore was not ſuffered.}
The foxes haue holes, and the foules of the aire neſtes; but the Sonne
of man hath not where to repoſe his head. \V But he ſaid to another:
Follow me. And he ſaid: Lord, permit me firſt to goe, and to burie my
father. \V And \Sc{Iesvs} ſaid to him: Let the dead burie their dead;
but goe thou, ſet forth the Kingdom of God. \V And another ſaid: I wil
follow thee Lord, but permit me
%%% o-2266
firſt to take my leaue of them that are at home. \V \Sc{Iesvs} ſaid to
him:
\LNote{No man looking backe}{It
\MNote{Looking backe.}
is a dangerous temptation for a man that hath loſt or left his goods for
Chriſt, to looke much backe at them, and to remember with delight the
pleaſures & eaſes of this world. For it breedeth in him diſcontentment
of the troubles and croſſes that are incident to the ſtate of ſuch as
fully follow Chriſt. In which caſe a man should euer looke forward
towards Heauen, and neuer backward to the world.}
No man putting his hand to the plough, and looking backe, is apt for the
Kingdom of God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
