%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 2454
%%% o-2284
\startChapter[
  title={Chapter 16}
  ]

\Summary{He teacheth the rich to procure Heauen with their
  riches. 14.~And being therfore derided of the couetous Phariſees (who
  ſaw temporal riches promiſed in the letter of the Law) he sheweth that
  now is come the preaching of the Kingdom of God, howbeit the Law for
  al that in no iot shal be fruſtrat. 19.~foretelling them alſo, that
  the couetous Iewes shal be denied of their father Abraham, when poore
  Lazarus (the penitent Gentil) shal reſt in his boſome.}

%%% o-2285
And he ſaid alſo to his Diſciples: There was a certaine rich man that
had a
\TNote{\G{οἰκονόμον}}
bailife: & he was il reported of vnto him, as he that had waſted his
goods. \V And he called him, and ſaid to him: What heare I this of thee?
render account of thy
\TNote{\G{οἰκονομίας }}
bailiſhip: for now thou canſt no more be bailife. \V And the bailife
ſaid within him ſelf: What ſhal I doe, becauſe my Lord taketh away from
me the bailiſhip? dig I am not able, to beg I am aſhamed. \V I know what
I wil doe, that when I ſhal be remoued from the bailiſhip, they may
receiue me into their houſes. \V Therfore calling together euery one of
his Lords debters, he ſaid to the firſt: How much doeſt thou owe my
Lord? \V But he ſaith: An hundred pipes of oile. And he ſaid to him:
Take thy bil and ſit downe, quickly write fiftie. \V After that he ſaid
to an other: But thou, how much doeſt thou owe? Who ſaid: An hundred
quarters of wheat. He ſaid to him: Take thy bil, and write eightie. \V
And
\LNote{The Lord praiſed}{This mans deceiuing his maiſter is not praiſed,
nor we
warranted by his fact to gaine vniuſtly for to haue wherewith to giue
almes: but his prudence, in that he prouided ſo ſubſtantially for him
ſelfe whileſt his maiſters goods were in his handes, is commended, not
for a vertue, but for a worldly pollicie:
\MNote{Good workes.}
and propoſed as an example of
the careful prouiſion that rich men (who are Gods ſtewards in earth)
should make for their ſoules, againſt they be put out of their bailiship
and be called to account, which is the day of their death: and for a
condemnation of faithful mens folly and negligence, that being aſſured
they shal out of their offices, and wel knowing they might gaine
ſaluation by their money, haue ſo litle regard thereof.}
the Lord praiſed the bailife of iniquitie, becauſe he had done wiſely:
for the children of this world, are wiſer then the children of light in
their generation. \V And I ſay to you: Make vnto you freinds of the
\SNote{Mammon (ſaith S.~Hierom
\Cite{q.~6. ad Algaſ.})
in the Syriake tongue ſignifieth riches. \Emph{Mammon of iniquitie},
becauſe they are often il gotten, or il beſtowed, or occaſion of euil, or
at the leaſt worldly & falſe, & not the true heauenly riches.}
mammon of iniquitie that when you faile,
\LNote{They may receiue}{A
\MNote{Almes meritorious.}
great comfort to al great almes-men, & a wonderful force and vertue in
almes, which beſide the merit of the worke of mercie, which
\CNote{\XRef{Tob.~12,~9.}
\XRef{Mt.~25.}}
(as in other
places of Scripture is ſaid) purgeth ſinne and gaineth Heauen,
\Fix{pocureth}{procureth}{possible typo, same in other}
alſo not only the praiers of their beadſmen in earth, but their
patronage in Heauen alſo.
\MNote{The Saints doe pray for vs.}
Whereby alſo the praiers of Saints for the
liuing, and namely for them to whom they were beholding in their life,
are proued. Yea and that they be in ſuch fauour with God, that they may
and doe receiue their freinds which were once their benefactours, into
their manſions in Heauen, no leſſe then the farmers whom the il ſteward
pleaſured, might receiue their freind into their earthly houſes. Which
alſo inſinuateth to vs, that almes beſtowed ſpecially vpon holy men, who
by their merites and praiers are great in Gods grace, may much more
helpe vs then our charitable deedes done vpon vulgar men in neceſſitie,
though that be of exceeding great merit alſo. See al this in theſe
Doctours following.
\Cite{Hierom. quæſt.~6. ad Algaſ. tom.~3.}
\Cite{Ambroſ. in Luc.}
\Cite{Auguſt. ſerm.~35. de verb. Do. c.~3.}
\Cite{Gregor. moral. li.~21. c.~24.}
\Cite{Auguſt. li.~2. qu. Euang. qu.~34.}
\Cite{Chryſ. ho.~33. ad Po. Antioch. to.~5.}}
they may receiue you into the eternal tabernacles. \V He that is
faithful in the leaſt, is faithful in the greater alſo: and he that is
vniuſt in litle, is vniuſt in the greater alſo. \V If then you haue not
been faithful in the vniuſt mammon, with that which is the true who may
credit you? \V And if you haue not been faithful in other mens, that
which is yours, who wil giue you? \V
\CNote{\XRef{Mt.~6,~24.}}
No ſeruant can ſerue two maiſters;
for either he ſhal hate the one, and loue the other; or cleaue to one,
and contemne the other. You can not ſerue God and mammon.

\V And the Phariſees which were couetous, heard al theſe things: and
they derided him. \V And he ſaid to them: You are they that iuſtifie
your ſelues before men, but God knoweth your hartes, becauſe that which
is high to men, is abomination before God. \V
\CNote{\XRef{Mt.~11,~12.}}
The Law and the Prophets,
vnto Iohn. From that time the Kingdom of God is euangelized, and euery
one doth force toward it. \V
\CNote{\XRef{Mat.~5,~18.}}
%%% !!! ??? \CNote{Mat.~5,~18. Mat.~5,~31. 19,~9.}
And it is eaſier for Heauen and earth to paſſe, then one tittle of the
Law to fal. \V
\CNote{\XRef{Mr.~10,~11.}
\XRef{1.~Cor.~7,~11.}}
Euery one that dimiſſeth his wife,
\LNote{And marieth}{The
\MNote{Mariage after diuorce vnlawful.}
good of Mariage throughout al Nations and men, is in iſſue and fidelitie
of chaſtitie, but among the people of God it conſiſteth alſo in holines
of Sacrament: whereby it commeth to paſſe that it is a heinous crime to
mary againe, though there be a diuorce made, ſo long as the parties
liue.
\Cite{Aug. de bono coniug. c.~24. to.~6.}
See the
\XRef{Annotations vpon Marke~10,~11.}}
and marieth another, committeth aduoutrie: and he that marieth her that
is dimiſſed from her
%%% o-2286
huſband, commiteth aduoutrie.

%%% 2455
\V There was a certaine rich man, & he was clothed with purple and
ſilke: and he fared euery day magnifically. \V And there was a certaine
begger called Lazarus, that lay at his gate, ful of ſores: \V deſiring
to be filled of the crummes, that fel from the rich mans table,
\Var{}{and no man did giue him.}% to prevent paragraph break
but the dogges alſo came, and licked his ſores. \V And it came to paſſe
that the begger died, and was caried
\LNote{Of the Angels}{Angels
\MNote{Vnmerciful rich men.}
carie good mens ſoules to Heauen now, as they did then his to Abrahams
boſome. See the reward of pouerty, affliction, and patience: and on the
contrarie, the end and reward of wealth ioyned with vnmercifulnes. Note
alſo here that at the day of euery mans death there is a particular
iudgement, and therfore the ſoule ſleepeth not, nor hangeth in ſuſpenſe
til the general iudgement.}
of the Angels into
\LNote{Abrahams Boſome}{The
\MNote{Abrahams boſome.}
Boſome of Abraham is the reſting place of al them that died in perfect
ſtate of grace before Chriſts time, Heauen before being ſhut from
men. It is called in Zacharie,
\CNote{\XRef{Zach.~9,~11.}}
\Emph{a lake without water}, and ſometimes
\CNote{\XRef{Eſa.~42,~7.}}
\Emph{a priſon}, but moſt commonly of the Diuines 
\MNote{\L{Limbus patrũ}.}
\L{Limbus
patrum}, for that it is thought to haue been the higher part or brimme
of Hel, the places of puniſhment being far lower then the ſame, which
therfore be called \L{Infernum inferius}
\CNote{\Cite{Aug. in Pſ.~85.}}
\Emph{the lower hel}. Where
this manſion of the Fathers ſtood, or whether it be any part of Hel,
S.~Auguſtine doubted: but that there was ſuch a place,  
\CNote{\Cite{Ep.~99.}}
he nor no Catholike man euer doubted:
\MNote{Chriſt deſcended into Hel, & deliuered the Fathers.}
as al the Fathers make it moſt certaine,
that our Sauiour deſcending to Hel, went thither ſpecially, and
deliuered the ſaid Fathers out of that manſion.
\Cite{Iren. li.~4. c.~19.}
\Cite{Euſeb. Demonſt. Euang. li.~10. c.~8. ſub finem.}
\Cite{Nazian. orat.~2. de Paſch.}
ſit Deus,
\Cite{Epipha. in hereſ.~26. Tatiani}
\Cite{Ambros. de myſt. Paſch. c.~4.}
\Cite{Hiero. in 9.~Zachariæ.}
\Cite{Auguſt. ep.~99. & li.~20. de Ciuit. c.~15.}
\Cite{Paulinus in Panegyrico Celſo.}
illiud, Inclinato
\Cite{Gregor. li.~6. ep.~179.}
which truth and place though of al the ancient writers confeſſed and
proued by this and other Scriptures; yet the Aduerſaries deny it (as
they doe Purgatorie) moſt impudently.}
Abrahams boſome. And the rich man alſo died: and he was buried in
Hel. \V And lifting vp his eyes, when he was in torments, he ſaw Abraham
a farre off,
\SNote{Lazarus in Abrahams boſome, and reſt, but both in
\Fix{hel,}{Hel,}{likely typo, same in other}
and not in the Kingdom of Heauen before Chriſt.
\Cite{Hiero. ep.~3.
\Fix{Epithoph. }{Epitaph. }{Possible typo, fixed in other}
Nepot.}}
and Lazarus in his boſome: \V and he crying ſaid: Father Abraham, haue
mercie on me, and ſend Lazarus that
\Fix{the}{he}{obvious typo, fixed in other}
may dip the tip of his finger into water for to coole my tongue, becauſe
I am tormented in this flame. \V And Abraham ſaid to him: Sonne,
remember that thou didſt receiue
\SNote{To be in continual pleaſures, eaſe, wealth, peace, and proſperity
in this world, is perilous, & a ſigne of paines in the next.
\Cite{S.~Hiero.}}
good things in thy life time, and Lazarus likewiſe euil: but now he is
comforted, and thou art tormented. \V And beſide al theſe things,
between vs and you there is fixed
\LNote{A great chaos}{A
\MNote{Purgatorie.}
great diſtance betwixt Abrahams boſome and the inferiour Hel. Some iudge
Purgatorie to be placed there, from whence (no doubt) Chriſt alſo
deliuered ſome at his deſcending to Hel. For, theſe in Abrahams boſome
were not in paines: and  
\CNote{\Cite{Aug. Ep.~99.}}
S.~Auguſtine ſaith the Scriptures be plaine that
he tooke ſome out of the places of puniſhment, and yet none out of the
Hel of the damned. What other place then can that be but Purgatorie?}
a great
\TNote{\G{χάσμα}, a horrible diſtance.}
chaos: that they which wil paſſe from hence to you, may not, neither goe
from thence hither. \V And he ſaid: Then, father, I beſeech thee that
thou wouldeſt ſend him vnto my fathers houſe, for I haue fiue
brethren, \V for to teſtifie vnto them,
\LNote{Leſt they alſo}{If 
\MNote{Saints doe heare our praiers & haue care of vs.}
the damned had care of their freindes aliue, how & for what
cauſe ſoeuer, much more haue the Saints and ſaued perſons. And if thoſe
in Hel haue meanes to expreſſe their cogitations and deſires, and to be
vnderſtood of Abraham ſo far diſtant both by place and condition, much
rather may the liuing pray to the Saints and be heard of them: betwixt
earth (that is to ſay) the Church militant and Heauen, being continual
paſſage of ſoules, and
\CNote{\XRef{Gen.~32.}}
Angels aſcending and deſcending by Iacobs
ladder. Men muſt not for al that be curious to ſearch how the ſoules of
the deceaſed expreſſe their mindes, and be heard one of another,
\MNote{Caluins blaſphemie.}
and ſo fal to blaſphemie, as
\CNote{\Cite{Cal. li.~2. Inſtit. c.~22. ſect.~24.}}
Caluin doth, aſking whether their eares be
ſo lõg to heare ſo far off, & wickedly meaſuring al things by mortal
mens corporal groſſe manner of vttering conceits one to another. Which
was not here done by this damned  nor by Abraham, with corporal
inſtruments of tongue, teeth, & eares: though for the better expreſſing
of the damneds caſe, Chriſt vouchſafed to vtter it in termes agreeing to
our capacitie.}
leſt they alſo come into this place of torments. \V And Abraham ſaid to
him:
\SNote{Abraham had knowledge of things in earth which were not in his
time; as that they had Moyſes and the Prophets bookes which he neuer
ſaw.
\Cite{Auguſt. de cura pro mor. c.~14.}}
They haue Moyſes and the Prophets: let them heare them. \V But he ſaid:
No, father Abraham, but if ſome man ſhal goe from the dead to them, they
wil doe penance. \V And he ſaid to him: If they heare not Moyſes and the
Prophets, neither if one ſhal riſe againe from the dead, wil they
beleeue.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
