%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 2441
%%% o-2269
\startChapter[
  title={Chapter 11}
  ]

\Summary{He teacheth a forme of prayer, 5.~and exhorteth to pray
  inſtantly, 11.~aſſuring that ſo God wil giue vs good things. 14.~The
  Iewes blaſpheming his caſting out of Diuels, and asking for a miracle
  from Heauen, 17.~he defendeth his doing: 22.~foretelling alſo the
  Diuels expulſion by him out of the world (that is, the vocation of the
  Gentils) 24.~and his reentrie into their Nation, 27.~with their
  reprobation though he be of their flesh, 29.~and alſo their final moſt
  worthy damnation. 37.~Againe, to the Phariſees and Scribes he crieth
  woe, as authours of the ſaid reprobation now at hand.}

%%% o-2270
And it came to paſſe, when he was in a certaine place praying, as he
ceaſed, one of his Diſciples ſaid to him: Lord teach vs to pray, as
Iohn alſo taught his Diſciples. \V And he ſaid to them:
\CNote{\XRef{Mat.~6,~9.}}
When you pray,
ſay, \Sc{Father}, \Emph{ſanctified by thy name. Thy Kingdom come, \V Our
daily bread giue vs this day, \V and forgiue vs our ſinnes, for becauſe
our ſelues alſo doe forgiue euery one that is in debt to vs. And lead vs
not into temptation.} \V And he ſaid to them: Which of you ſhal haue a
freind, and ſhal goe to him at midnight, and ſhal ſay to him, Freind,
lend me three loaues, \V becauſe a freind of mine is come out of his way
to me, and I haue not what to ſet before him: \V and he from within
anſwering ſaith: Trouble me not, now the doore is ſhut, and my children
are with me in bed; I can not riſe and giue thee. \V And if he ſhal
perſeuer knocking, I ſay to you, although he wil not riſe and giue him
becauſe he is his friend, yet for his importunitie he wil riſe, and giue
him as many as he needeth. \V
\CNote{\XRef{Mat.~7,~7.}}
And I ſay to you, Aske, and it ſhal be
giuen you: ſeeke, and you shal find: knock, and it shal be opened to
you. \V For euery one that asketh, receiueth: and he that ſeeketh,
findeth: and to him that knocketh, it shal be opened. \V And which of
you if he aske his father bread, wil he giue him a ſtone? or a fiſh, wil
he for a fiſh giue him a ſerpent? \V Or if he aske an egge, wil he
reach him a ſcorpion? \V If you then being naught, know how to giue good
guiftes to your children, how much more wil your father from Heauen giue
the good Spirit to them that aske him?

\V
\CNote{\XRef{Mt.~12,~12.}}
And he was caſting out a Diuel, and that was dumme. And when he had
caſt out the Diuel, the dumme ſpake: and the multitudes marueled. \V
\CNote{\XRef{Mr.~3,~22.}}
And certaine of them ſaid: In Beel-zebub the prince of
%%% 2442
Diuels he caſteth out Diuels. \V And other tempting, asked of him a ſigne
from Heauen. \V But he ſeeing their cogitations, ſaid to them: Euery
Kingdom deuided againſt itſelf, ſhal be made deſolate, and
\TNote{\G{οἶκος ἐπὶ οἶκον}}
houſe vpon houſe, ſhal fal. \V And if Satan alſo be deuided againſt
himſelf, how ſhal his Kingdom ſtand? becauſe you ſay that in Beel-zebub
I doe caſt out Diuels. \V And if I in Beel-zebub caſt out Diuels: your
children, in whom doe they caſt out? therfore they ſhal be your
iudges. \V But if I in the
\SNote{This finger, is the Spirit of God,
\XRef{Mt.~12,~13.}}
finger of
%%% o-2271
God doe caſt out Diuels; ſurely the Kingdom of God is come
vpon you. \V When the ſtrong armed keepeth his court, thoſe things are
in peace that he poſſeſſeth. \V But if a ſtronger then he come vpon him
and ouercome him; he wil take away his whole armour wherein he truſted,
and wil diſtribute his ſpoiles. \V He that is not with me, is againſt
me: and he that gathereth not with me, ſcattereth. \V When the vncleane
ſpirit ſhal depart out of a man, he wandereth through places without
water, ſeeking reſt. And not finding, he ſaith, I wil returne into my
houſe whence I departed. \V And when he is come, he findeth it ſwept
with a beſome, and trimmed. \V Then he goeth and taketh ſeuen other
ſpirits worſe then himſelf, and entring in they dwel there. And the
\TNote{\L{fiunt nouiſsima}}
laſt of that man be made worſe then the firſt.

\V And it came to paſſe, when he ſaid theſe things, a certaine woman
lifting vp her voice out of the multitude ſaid to him:
\LNote{Bleſſed is the wombe}{Let
\MNote{Our B.~Lady.}
vs alſo (ſaith Venerable Bede) lift vp our voice with the Catholike
Church, of which this woman was a figure; let vs lift vp our hartes
among the People, and ſay to our Sauiour: Bleſſed be the wombe that bare
thee, and the paps which thou didſt ſucke. For Bleſſed indeed is the
mother which bare the King that ruleth Heauen & earth for euer.}
Bleſſed is the wombe that bare thee, and the paps that thou didſt
ſucke. \V But he ſaid:
\SNote{The ſaid mother of God, in that alſo was Bleſſed that she was the
tẽporal meanes & miniſter of the Incarnatiõ, but much more Bleſſed, in
that she cõtinueth the perpetual keeper of his word.
\Cite{Beda},
\Cite{Aug. tract.~19. in Ioan.}}
Yea rather, bleſſed are they that heare the word of God, and keepe it.

\V And the multitudes running together,
\CNote{\XRef{Mt.~1,~29.}}
he began to ſay: This
Generation, is a wicked Generation: it asketh a ſigne, and a ſigne ſhal
not be giuen it but
\LNote{The ſigne of Ionas}{Of
\MNote{The ſigne of Ionas.}
al miracles, his Reſurrection, after he had been according to his body,
in the graue, according to his ſoule, in Hel three daies, was the
greateſt, and moſt conuinceth the incredulous Iewes: and therfore a
greater or more euident then that, he ſaith he wil not giue them.}
the ſigne of Ionas the Prophet. \V
\CNote{\XRef{Io.~2,~2.}}
For as Ionas was a ſigne to the
Niniuites; ſo ſhal the Sonne of man alſo be to this Generation. \V
\CNote{\XRef{3.~Reg.~10,~1.}}
The
Queene of the South ſhal riſe in the iudgement with the men of this
Generation, and ſhal condemne them: becauſe ſhe came from the endes of
the earth to heare the wiſedom of Salomon. And behold, more then Salomon
here. \V The men of Niniuee ſhal riſe in the iudgement with this
Generation, and shal condemne it,
\CNote{\XRef{Io.~3,~5.}}
becauſe they
\SNote{\G{μετενόησαν}. Marke that the great pennance of the Niniuites
\XRef{(Ionæ~3.)}
is here expreſſed by this Greeke word. See
\XRef{Annot. Mt.~3,~2.}}
did pennance at the preaching of Ionas. And behold, more then Ionas
here.

\V
\CNote{\XRef{Mt.~5,~15.}
\XRef{Mr.~4,~21.}}
No man lighteth a candle, and putteth it in ſecret, neither vnder a
bushel: but vpon a candleſticke, that they that goe in may ſee the
light. \V
\CNote{\XRef{Mt.~6,~22.}}
The cãdle of thy body is thine eye. If thine eye be ſimple,
thy whole body shal be lightſome: but if it be naught, thy body alſo
shal be darkeſome. \V See therfore that the light which is in thee, be
not darkeneſſe. \V If then thy whole body be lightſome, hauing no part
of darkeneſſe: it ſhal be lightſome wholy, and as a
%%% o-2272
bright candel it ſhal lighten thee.

\V And when he was ſpeaking, a certaine Phariſee deſired him that he
would dine with him. And he going in ſate downe to eate. \V And the
Phariſee began to thinke within himſelf and to ſay: Why he was not waſhed
before dinner. \V And our Lord ſaid to him: Now you
%%% 2443
Phariſees doe make cleane that on the out ſide of the cup and of the
platter; but that of yours which is within, is ful of rapine and
iniquitie. \V Fooles, did not he that made that on the out ſide, make
that alſo that is on the inſide? \V But yet
\TNote{\L{quod superest}, \G{τὰ ἐνόντα}.}
that that remaineth,
\LNote{Giue almes}{The
%%% !!! Compare this with other. Cites confused.
\CNote{\XRef{Eccl.~3,~33.}
\XRef{Dan.~4,~24.}
\XRef{Tob.~4,~12.}
\XRef{Mat.~25,~35.}}
\MNote{The force of the almes.}
great force of almes is here and in diuers places of holy writ
ſignified. In one place, they
\Fix{extingusih}{extinguiſh}{obvious typo, fixed in other}
ſinne: in another, they redeeme
ſinnes: in another, they deliuer from death: in another, to thẽ giuen or
omitted, our iudgement to Heauen or Hel is attributed: & here they make
cleane and ſatisfie for the Iewes former offenſes. For (as S.~Auguſtine
ſaith
\Cite{c.~70. Enchiridij)}
almes-deedes profit not a man that hath a wil to cõtinue in his ſinnes,
but they are to be done for a propitiation to God of former
offenſes. Now how wel the Proteſtants like this doctrine ſo euidently
ſet forth in Scripture, let the indifferent iudge, and how wel it
agreeth with their only faith.}
giue almes, & behold al things are cleane vnto you. \V But woe to you
Phariſees, becauſe you tithe mint and rew and euery herbe: and paſſe
ouer iudgement and the charitie of God. But theſe things you ought to
haue done, and not to omit thoſe. \V Woe to you Phariſees, becauſe you
loue the firſt chaires in the Synagogues, and ſalutations in the market
place. \V Woe to you, becauſe you are as monuments that appeare not, and
men walking ouer, are not ware.

\V And one of the Lawyers anſwering ſaith to him: Maiſter, in ſaying
theſe things, thou ſpeakeſt to our reproch alſo. \V But he ſaid:
\LNote{Woe to you Lawyers}{Theſe
\MNote{The Lawyers and Prieſts of the old Teſtament.}
were Doctours of Moyſes Law, otherwiſe called Scribes. Shal we therfore
crie out againſt al Lawyers now, or ought the name of Lawyer be odious
with vs, becauſe of theſe naughty Lawyers among the Iewes? much leſſe
ought the name of Prieſts to be odious (as Heretikes would haue it)
becauſe of the Iewes Prieſts that were ſo buſy againſt our Sauiour.}
Woe to you Lawyers alſo: becauſe you load men with burdens which they
can not beare, and your ſelues touch not the packes with one of your
fingers. \V Woe to you that
\SNote{Not the building of the Prophets monumẽts is condemned, but their
imitation of their fathers that ſlew the Prophets.
\Cite{Ambroſ.}}
build the monuments of the Prophets: and your fathers did kil them. \V
Surely you doe teſtifie that you cõſent to the workes of your fathers:
becauſe they indeed did kil them, and you build their ſepulchres. \V For
this cauſe the wiſedom alſo of God ſaid, I wil ſend to them Prophets and
Apoſtles, and of them they wil kil and perſecute. \V That the bloud of
al the Prophets that was ſhed from the making of the world, may be
required of this Generation, \V
\CNote{\XRef{Gen.~4,~8.}}
from the bloud of Abel vnto the
\CNote{\XRef{2.~Par.~24,~22.}}
bloud of
Zacharie that was ſlaine between the Altar and the Temple. Yea I ſay to
you, it ſhal be required of this Generation. \V Woe to you Lawyers,
becauſe you haue taken away the key of knowledge: your ſelues haue not
entred, & thoſe that did enter you haue prohibited. \V And whẽ he ſaid
theſe things to them, the Phariſees & the Lawyers began vehemently to
vrge him, & to ſtop his mouth about many things, \V lying in waite for
him, and ſeeking to catch ſomething of his mouth, that they might accuſe
him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
