%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-07


\project douay-rheims


%%% 2429
%%% o-2255
\startChapter[
  title={Chapter 7}
  ]

\Summary{He teſtifieth, the faith of the Centurion who was a Gentil, to
  be greater then he found among al the Iewes, and cureth his ſeruant
  abſent. 11.~The widowes ſonne he reuiueth and reſtoreth to her, & is
  renowned therevpon. 18.~To Iohns meſſengers he anſwereth with
  miracles, leauing to Iohn to preach thereby vnto them that he is
  Chriſt. 24.~And afterward he declareth how worthy credit was Iohns
  teſtimonie, 29.~inueighing againſt the Phariſees, 31.~who with neither
  of their mãners of liuing could be wonne, 36.~shewing alſo vnto them
  by occaſion of Marie Magdalen, how he is a freind to ſinners, not to
  maintaine them in ſinne, but to forgiue them their ſinnes vpon their
  faith and pennance.}

And
\CNote{\XRef{Mt.~8,~5.}}
when he had fully ſaid al his words into the eares of the People, he
entred into Capharnaum. \V And the ſeruant of a certaine Centurion being
ſick, was readie to die: who was deare vnto him. \V And when he had
heard of \Sc{Iesvs}, he ſent vnto him the Ancients of the Iewes,
deſiring him to come and heale his ſeruant. \V But they being come
to \Sc{Iesvs}, beſought him earneſtly, ſaying to him, That he is worthie
that thou ſhouldeſt doe this for him. \V For he loueth our Nation; and
he hath
\LNote{Built a Synagogue}{As
\MNote{Building of Churches, Monaſteries, &c.}
at that time to found a Synagogue, was acceptable to God,
and procured the praiers of the faithful People for whoſe vſe it was
made: ſo now much more in the new Teſtament, to build a Church,
Monaſterie, College, or any like worke for the honour and ſeruice of
God, is grateful to him and procureth the praiers of the good People for
whoſe vſe ſuch things be founded.}
built a Synagogue for vs. \V And \Sc{Iesvs} went with them. And when he
was now not farre from the houſe, the Centurion ſent his freinds vnto
him, ſaying: Lord, trouble not thyſelf. For
\SNote{See the
\XRef{Annotations vpon S.~Matth. c.~8,~8.}}
\Emph{I am not worthie that thou shouldeſt enter vnder my roofe.} \V For
the which cauſe neither did I thinke my ſelf worthie to come to thee;
but ſay the word, and my ſeruant ſhal be made whole. \V For I alſo am a
man ſubiect to authoritie, hauing vnder me ſouldiars: and I ſay to this,
goe, and he goeth; and to an other, come, and he commeth; and to my
ſeruant, doe this, and he doeth it. \V Which \Sc{Iesvs} hearing,
marueled; and turning to the multitude that followed him he ſaid: Amen I
ſay to you, neither in Iſrael haue I found ſo great faith. \V And they
that were ſent, being returned home, found the ſeruant that had been
ſick, whole.

%%% o-2256
\V And it came to paſſe, afterward he went into a citie that is called
Naim; and there went with him his Diſciples and a very great
multitude. \V And when he came nigh to the gate of the citie, behold a
dead man was caried forth, the only ſonne of his mother; and ſhe was a
widow: and a great multitude of the citie with her. \V Whom when our
Lord had ſeen, being moued with mercie vpon her, he ſaid to her: Weep
not. \V And he came neere and touched the coffin. And they that caried
it ſtood ſtil; and he ſaid: Yong man, I ſay to thee, ariſe. \V And he
that was dead, ſate vp, and began to ſpeake. And he gaue him to his
mother. \V And feare tooke them al; and they magnified God, ſaying, That
a great Prophet is riſen among vs: and, that God hath viſited his
People. \V And this ſaying went forth into al Iewrie of him, & into al
the countrie about.

\V And Iohns Diſciples ſhewed him of al theſe things. \V
\CNote{\XRef{Mt.~11,~2.}}
And Iohn
%%% 2430
called two of his Diſciples, and ſent them to \Sc{Iesvs}, ſaying: Art
thou he that art to come; or expect we another? \V And when the men were
come vnto him, they ſaid: Iohn the Baptiſt hath ſent vs to thee, ſaying:
Art thou he that art to come; or expect we another? \V (And the ſelf
ſame houre, he cured many of maladies, and hurts, and euil Spirits: and
to many blind he
\TNote{\G{ἐχαρίσατο}}
gaue ſight.) \V And
\Fix{Anſwering,}{anſwering,}{obvious typo, fixed in other}
he ſaid to them: Goe & report to Iohn what you haue heard and ſeen:
\CNote{\XRef{Eſa.~35,~5.}
\XRef{61,~1.}}
That the blind ſee, the lame walke, the lepers are made cleane, the
deafe heare, the dead riſe againe, \V
\SNote{\L{pauperes evangelizantur}, that is, to the poore the Ghoſpel is
preached, and they receaue it.}
the poore are euangelized: and bleſſed is he whoſoeuer ſhal not be
ſcandalized in me.

\V
\CNote{\XRef{Mt.~11,~7.}}
And when Iohns meſſengers were departed, he began to ſay of Iohn to
the multitudes: What went you out into the deſert to ſee? a reed moued
with the wind? \V But what went you forth to ſee? a man clothed in
\SNote{Marke this wel cõcerning Iohns apparel and diet. See the
\XRef{Annotatiõs vpon S.~Matth. c.~3,~4.}}
ſoft garments? Behold they that are in coſtly apparel and delicacies,
are in the houſes of Kings. \V But what went you out for to ſee? a
Prophet? Certes I ſay to you, and more then a Prophet. \V This is he of
whom it is written:
\CNote{\XRef{Mal.~3,~1.}}
\Emph{Behold I ſend mine Angel before thy face,
which shal prepare thy way before thee.} \V For I ſay to you: A greater
Prophet among the children of women then Iohn the Baptiſt, there is no
man. But he that is the leſſer in the Kingdom of God, is greater then
he. \V And al the People hearing and the Publicans, iuſtified God, being
baptized with
%%% o-2257
Iohns Baptiſme. \V But the Phariſees and the lawyers
\SNote{As they that cõtẽned Iohns Baptiſme, deſpiſed Gods counſel
& wiſedom; ſo much more they that make no accoũt of the Sacramẽts of the
Church, deſpiſe Gods counſel & ordinance touching their ſaluation,
to their owne damnation.}
deſpiſed the counſel of God againſt them ſelues, being not baptized of
him. \V And our Lord ſaid: Wherevnto then ſhal I liken the men of this
Generation, and wherevnto are they like? \V They are like to children
ſitting in the market-place, and ſpeaking one to another, and ſaying: We
haue piped to you, and you haue not danced; we haue lamented, and you
haue not wept. \V For Iohn Baptiſt came
\CNote{\XRef{Mt.~3,~4.}}
neither eating bread nor
drinking wine; and you ſay: He hath a Diuel. \V The Sonne of man came
eating and drinking; and you ſay: Behold a man that is a gurmander and a
drinker of wine, a freind of Publicans and ſinners. \V And wiſedom is
iuſtified of al her children.

\V And one of the Phariſees deſired him to eate with him. And being
entred into the houſe of the Phariſee, he ſate downe to meate. \V And
behold a woman that was in the citie, a ſinner, as ſhe knew that he was
ſet downe in the Phariſees houſe, ſhe brought an alabaſter boxe of
ointment; \V and ſtanding behind beſide his feet, ſhe began to
\SNote{A perfect paterne of true pennãce in this woman, who ſought of
Chriſt with open teares & other ſtrange works of ſatisfactiõ & deuotion,
remiſsion of her ſinnes.}
water his feet with teares, & wiped them with the haires of her head,
and kiſſed his feet, & anointed them with the ointment. \V And the
Phariſee that had bid him ſeeing it, ſpake within himſelf, ſaying: This
man if he were a Prophet, would know certes who and what manner of woman
ſhe is which toucheth him, that ſhe is a ſinner. \V And \Sc{Iesvs}
anſwering ſaid to him: Simon, I haue ſomewhat to ſay vnto thee. But he
ſaid: Maiſter, ſay. \V A certaine creditour had two debters; one did owe
fiue hundred pence, & the other fiftie. \V They hauing not wherewith
to pay, he forgaue both. Whether therfore doth loue him more? \V Simon
anſwering ſaid: I ſuppoſe that he to whom he forgaue more. But he ſaid
to him: Thou haſt iudged rightly. \V And turning to the woman, he ſaid
%%% 2431
vnto Simon: Doeſt thou ſee this woman? I entred
\LNote{Into thy houſe}{An
\MNote{Exteriour ſignes of more then common deuotion.}
exceeding approbation of the extraordinary workes and ſignes of external
deuotion, which ſeeme to carnal men (though otherwiſe faithful) to be
often ſuperfluous or not acceptable. This Simon was perhaps of a good
wil, and therfore (as diuers other did elſwhere) inuited Chriſt to his
houſe, not of curioſitie or captiouſnes, as ſome other did; but of
affectiõ, as it may ſeeme by Chriſts familiar talke with
him. Notwithſtanding his duties towards him were but ordinary. But the
anointing, waſhing, kiſsing, wiping of his feet in ſuch ſort as the
woman did, were further ſignes of more then vulgar loue: ſuch as is in
deuout men or women that goe on pilgrimage and kiſſe deuoutly the holy
memories of Chriſts and his Saints. Which is no more but an exteriour
expreſsing of their affection, and that they loue much, as euery vulgar
Chriſtian man doth not.}
into thy houſe, water to my feet thou didſt not giue; but ſhe with
teares hath watered my feet, and with her haires hath wiped them. \V
Kiſſe thou gaueſt me not; but ſhe ſince I came in, hath not ceaſed to
kiſſe my feet. \V With oile thou didſt not anoint my head; but ſhe with
ointment hath anointed my feet. \V For the which I ſay to thee: Many
ſinnes are forgiuen her, becauſe ſhe hath
\SNote{Not only faith (as you may perceiue) but loue or charitie
obtaineth remiſsion of ſinnes.}
loued much. But to whom leſſe is forgiuen, he loueth leſſe. \V And he
ſaid to her: Thy ſinnes are
%%% o-2258
forgiuen thee. \V And they that ſate togeather at the table began to ſay
within them ſelues,
\SNote{As the Phariſees did al wayes carpe Chriſt for remiſsion of
ſinnes in earth, ſo the Heretike reprehend his Church that remitteth
ſinnes by his authoritie.}
Who is this that alſo forgiueth ſinnes? \V And he ſaid to the woman:
\LNote{Thy faith}{The
\MNote{Iuſtification attributed not to faith only.}
remiſsion  of her ſinnes being attributed before to charitie, is now
alſo ſaid to come of her faith. Whereby you may know that it commonly
proceedeth of both, and of hope alſo though but one named. Becauſe when
there be diuers cauſes concurring to one effect, the ſcriptures commonly
name but one, and that eſpecially which is moſt proper to the purpoſe
and time, not excluding the other. And therfore his working miracles vpon
any perſon, is attributed to the faith of them on whom or at whoſe
deſire they be done. Becauſe he wrought his miracles to induce al men to
beleeue in him; and therfore ſpecially required faith at their hands,
and namely before other things, whether they did beleeue that he was
able to doe that which they aſked at his hands: without which it had
been rather a mockerie and tentation of him, then a true deſire of
benefit at his hands.}
Thy faith hath made thee ſafe, goe in peace.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
