%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 2416
%%% o-2239
\startChapter[
  title={Chapter 2}
  ]

\Summary{The Natiuitie of Chriſt, 8.~and manifeſtation thereof to the
  Shepheards by an Angel & by them to others. 21.~His
  Circumciſion. 22.~His Preſentation, togeather with Simeons (as alſo
  Annes) atteſtation and prophecying of his Paſsion, of the Iewes
  reprobation, and of the Gentils illumination. 41.~His annual aſcending
  to Hieruſalem with his parents, to whom he was ſubiect, and his fulnes
  of wiſedom shewed among the Doctours at twelue of his age.}

And it came to paſſe,
\SNote{In the yeare, frõ the creatiõ of the world 5199.\ frõ Noës
flud, 2957.\ from the Natiuitie of Abrahã, 2015.\ frõ Moyſes & the
comming forth of the People of Iſrael out of Ægypt, 1510.\ frõ Dauid
anointed King, 1032.\ from the firſt Olympias, 800.\ from the building
of Rome, 752.\ hebdomada 63 according to the prophecie of Daniel
\XRef{(c.~9)}
that is, in the yeare 440.\ or there about: in the ſixt Age of the world,
whẽ there was vniuerſal peace in al the world: the Eternal God the Sonne
of the Eternal Father, meaning to conſecrate & ſanctifie the world with
his moſt bleſſed comming, being conceaued of the Holy Ghoſt nine moneths
after his cõception, \Sc{Iesvs Christ} the ſonne of God is borne in
Bethlehem of Iuda, in the yeare of Ceſar Auguſtus 42.
\Cite{Vſuard in Martyrol. Decẽb.~25.}
\Emph{according to the common ancient ſupputation.}}
in thoſe daies there came forth an Edict from Cæſar Auguſtus, that
\Fix{he}{the}{obvious typo, fixed in other}
whole world ſhould be enroled. \V This firſt enroling was made by the
Preſident of Syria Cyrinus. \V And al went to be enroled, euery one into
his owne citie. \V And Ioſeph alſo went vp from Galilee out of the citie
of Nazareth into Iewrie, to the citie of Dauid that is called
Beth-lehem: for becauſe he was of the houſe and familie of Dauid, \V to
be enroled with \Sc{Marie} his deſpouſed wife that was with child. \V
And it came to paſſe, when they were there, her daies were fully come
that ſhe ſhould be deliuered. \V And ſhe brought forth her firſt
begotten Sonne,
%%% o-2240
and ſwadled him in clothes, and laid him downe in a
manger; becauſe there was not place for them in the inne.

\V And there were in the ſame countrie ſhepheards watching, and keeping
the night watches ouer their flock. \V And behold, an Angel of our Lord
ſtood beſide them, and the brightnes of God did ſhine round about them,
and they feared with a great feare. \V And the Angel ſaid to them: Feare
not, for behold I euangelize to you great ioy, that shal be to al the
People: \V becauſe this day is borne to you a \Sc{Saviovr} which is
Chriſt our Lord, in the citie of Dauid. \V And this shal be a ſigne to
you; You shal find the infant ſwadled in clothes, and laid in a
manger. \V And ſodenly there was with the Angel a multitude of the
heauenly armie, praiſing God, and ſaying, \V 
\MNote{Gloria in excelſis at Maſſe.}
\Emph{Glorie in the higheſt to God: and in earth peace to
\LNote{Men of good wil}{The
\MNote{Free wil.}
birth of Chriſt giueth not peace of mind or ſaluation but to ſuch as be
of good wil, becauſe he worketh not our good againſt one willes, but our
willes concurring.
\Cite{Aug. quæſt. ad Simplic. li.~1. q.~2. to.~4.}}
men of good wil.} \V And it came to paſſe, after the Angels departed frõ
them into Heauen, the shepheards ſpake one to an other: Let vs goe ouer
to Bethlehem, and let vs ſee this Word that is done, which our Lord hath
shewed to vs. \V And they came with ſpeed; and they found \Sc{Marie} and
Ioſeph, and the infant laid in the manger. \V And ſeeing it, they
vnderſtood of the Word that had been ſpoken to them concerning this
child. \V And al that heard, did maruel; and concerning thoſe things that
were reported to them by the shepheards. \V But \Sc{Marie}
\LNote{Kept al}{Our
\MNote{Our B.~Lady ful of deep contemplations.}
Lady though litle be ſpoken of her concerning ſuch matters in the
Scriptures, becauſe she was a womã and not admitted to teach or diſpute
in publick of high myſteries: yet she knew al theſe myſteries, and
wiſely noted and contemplated of thoſe things that were done and ſaid
about Chriſt, from the firſt houre of his Conception til the end of his
life and his Aſcenſion.}
kept al theſe words, conferring them in her hart. \V And the shepheards
returned, glorifying and prayſing God in al things that they had heard,
and ſeen, as it was ſaid to them.

\V And
\CNote{\XRef{Gen.~17,~12.}}
after eight daies were expired, that the child should be
circumciſed; his name was called \Sc{Iesvs}, which was
\CNote{\XRef{Lu.~1,~31.}}
called by the Angel, before that he was conceaued in the womb.

%%% 2417
\V And after the daies were fully ended of her purification
\CNote{\XRef{Leu.~12,~6.}}
according to
the law of Moyſes, they caried him into Hieruſalem, to preſent him to
our Lord (\V as it is written in the law of our Lord,
\CNote{\XRef{Exo.~13,~2.}}
\Emph{That euery
male opening the matrice, shal be called holy to the Lord.}) \V and to
giue a ſacrifice
Nu.~8,~16. Leu.~12,~8.
according as it is written in the law of our Lord, a
paire of turtles, or two yong pigeons. \V And behold there was a man in
Hieruſalem named Simeon, and this man was iuſt and religious, expecting
the conſolation of Iſrael: and the Holy Ghoſt was in him. \V And he had
receaued an anſwer of the Holy Ghoſt, that he
%%% o-2241
ſhould not ſee death vnles he ſaw firſt the
\SNote{See
\XRef{Iohn.~1,~22. and 41.}}
\Sc{Christ} of our Lord. \V And he came in Spirit into the Temple. And
when his parents brought in the child \Sc{Iesvs}, to doe according to
the cuſtome of the Law for him, \V he alſo tooke him into his armes, and
bleſſed God, and ſaid:

\V \Sc{Now thov}
\MNote{Nunc dimittis at Complin.}
\Emph{doeſt dimiſſe thy ſeruant O Lord, according to thy word in peace.}

\V \Emph{Becauſe mine eyes haue ſeen thy} \Sc{Salvation},

\V \Emph{Which thou haſt prepared before the face of al Peoples:}

\V \Emph{A light to the reuelation of the Gentils, & the glorie of thy
People Iſrael.}

\V And his father and mother were marueling vpon thoſe things which were
ſpoken concerning him. \V And Simeon bleſſed them, and ſaid
to \Sc{Marie} his mother: Behold this is ſet
\LNote{To the ruine}{Therfore
\MNote{Mens ruine and damnation is of them ſelues.}
to the ruine of ſome, becauſe they would not beleeue in him, and ſo were
the cauſe of their owne ruine, as he is elſwhere called
\CNote{\XRef{1.~Pet.~2,~1.}}
\Emph{A ſtumbling
ſtone}, becauſe many would ſtumble at him and ſo fal by their owne
fault. Otherſome he raiſed by grace from ſinne to iuſtice, and ſo he was
the reſurrection of many. The Apoſtle vſeth the like ſpeach,
ſaying:
\CNote{\XRef{2.~Cor.~2,~16.}}
\Emph{We are to ſome the odour of life, vnto life; to others, the odour of
death vnto death.} Not that their preaching was to cauſe death, but
becauſe they that would not beleeue their preaching, wilfully incurred
deadly ſinne and damnation.}
vnto the ruine, and vnto the reſurrection of many in Iſrael, & for a
ſigne which shal be contradicted, \V and
\SNote{Simeon prophecied not only of Chriſt but alſo our B.~Lady:
ſpecially of her ſorowes, wherein she was alwaies partaker with our
\Fix{ſauiour,}{Sauiour,}{obvious typo, fixed in other}
from his flight into Ægypt euen to his death.}
thine owne ſoule shal a ſword pearce, that out of many harts cogitations
may be reuealed. \V And there was Anne a prophetiſſe, the daughter of
Phanuel, of the Tribe of Aſer: she was farre ſtriken in daies, and had
liued with her husband ſeauen yeares from her virginitie. \V And she was
\LNote{A widow}{Marke
\MNote{Holy widowhood.}
that widowhood is here mentioned to the commandation therof euen in the
old Teſtament alſo, and the fruit, and as it were the profeſſion
thereof, is here commended, to wit, faſting, praying, being continually
in the Temple euen as
\CNote{\XRef{1.~Cor.~7.}
\XRef{1.~Tim.~5.}}
S.~Paul more at large for the ſtate of the new
Teſtament ſpeaketh of widowhood & virginitie, as being profeſsions more
apt and commodious for the ſeruice of God.}
a widdow vntil eightie and foure yeares; who departed not from the
Temple,
\LNote{By faſtings and praiers ſeruing}{Seruing,
\MNote{Faſting an act of Religion.}
in the Greeke is \G{λατρεύουσα}, that is, doing diuine worship vnto God, as by
praier, ſo alſo by faſting: ſo that faſting is \G{λατρεία},
\Fix{that, is}{that is,}{obvious typo, fixed in other}
an act
of Religion whereby we doe worship God, as we doe by praier, and not
vſed only to ſubdue our flesh, much leſſe (as Heretikes would haue it)
as a matter of pollicie.}
by faſtings and praiers
\SNote{\G{λατρεύουσα}}
ſeruing night and day. \V And she at the ſame houre ſodenly comming in,
confeſſed to our Lord; and ſpake of him to al that expected the
redemption of Iſrael. \V And after they had wholy done al things
according to the law of our Lord, they returned into Galilee, into their
citie Nazareth.

\V And the child grew, and waxed
\Var{ſtrong}{ſtrong in ſpirit}:
ful of wiſedom, and the grace of God was in him. \V And his parents went
euery yeare vnto Hieruſalem
\CNote{\XRef{Exo.~23,~15.}
\XRef{34,~17.}
\XRef{Deu.~16,~1.}}
at the ſolemne day of Paſche. \V And when he
was twelue yeares old, they going vp into Hieruſalem according to the
cuſtome of the feſtiual day; \V and hauing ended the daies, when they
returned, the child \Sc{Iesvs} remained in Hieruſalem: and his parents
knew it not. \V And thinking that he was in the companie, they came a
daies iourney, and ſought him among their kinsfolke and acquaintance. \V
And not finding him, they returned into Hieruſalem, ſeeking him. \V And
it came to paſſe, after three daies they found him in the Temple ſitting
in the middes of the Doctours, hearing them and asking them. \V And al
were aſtoniſhed that heard him, vpon his wiſedom & anſwers.

%%% o-2242
\V And ſeeing him, they wondered. And his mother ſaid to him: Sonne, why
haſt thou ſo done to vs? behold thy father and I ſorowing did ſeeke
thee. \V And he ſaid to them: What is it that you ſought me? did you not
know, that I muſt be about thoſe things which are my
%%% 2418
Fathers? \V And they vnderſtood not the word
\Fix{thath}{that he}{obvious typo, fixed in other}
ſpake vnto them. \V And he went downe with them, and came to Nazareth;
and was
\LNote{Subiect to them}{Al
\MNote{Dutiful obedience to parents.}
children may learne hereby, that great ought to be their ſubiection and
obedience to their Parents, when Chriſt himſelf, being God, would be
ſubiect to his parents being but his creatures.}
ſubiect to them. And his mother kept al theſe words in her hart. \V
And \Sc{Iesvs} proceeded in wiſedom and age, and grace with God and
men.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
