%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 2452
%%% o-2282
\startChapter[
  title={Chapter 15}
  ]

\Summary{By occaſion of the Phariſees murmuring at him for receiuing
  penitent ſinners, he sheweth what ioy shal be in Heauen for the conuerſion
  of one ſinner, 11.~and for the yonger ſonne, which is the Gentils:
  25.~the elder (to wit the Iewes) in the meane time diſdaining thereat,
  and refuſing to come into his Church.}

And there approched Publicans and ſinners vnto him for to heare him. \V
And the Phariſees and the Scribes murmured ſaying: That this man
receiueth ſinners, and eateth with them. \V And he ſpake to them this
parable, ſaying: \V
\CNote{\XRef{Mt.~18,~12.}}
What
\SNote{This man, is our Sauiour Chriſt: whoſe care & trauaile in
ſearching & reducing ſinners to repentance, al ſpiritual men ſpecially
should follow.}
man of you hauing an hundred ſheep: and if he hath loſt one of them,
doth he not leaue the ninetie nine in the deſert, and goeth after that
which was loſt vntil he find it? \V And when he hath found
%%% o-2283
it, laieth it vpon his ſhoulders reioycing: \V and comming home calleth
together his freindes and neighbours, ſaying to them: Reioyce with me,
becauſe I haue found my sheepe that was loſt? \V I ſay to you, that euen
ſo there shal be ioy in Heauen vpon one ſinner that doth penance,
\LNote{Then vpon ninety nine iuſt}{Neither
\MNote{Ioy in Heauen for euery penitent.}
God, nor the Saints in Heauen, nor men in earth doe for al that eſteeme
more of penitẽt ſinners, then they doe of them that continue iuſt and
godly: though by the ſoden motion and preſent affection of ioy that man
taketh and expreſſeth in ſuch alteration and new fallen good, it be here
ſignified that the conuerſion of euery ſinner is exceeding acceptable to
God, and giueth his Saints new cauſe of ioy and thankes-giuing to God in
another kind then for the continuance of the iuſt.}
then vpon ninetie nine iuſt that neede not penance. \V Or what
\SNote{This woman is the
\Fix{catholike}{Catholike}{likely typo, same in other}
Church, who alſo cõtinually ſeeketh her loſt children.}
woman hauing ten grotes: if she leeſe one grote, doth she not light a
candle, and ſweepe the houſe, and ſeeke diligently, vntil she finde? \V
And when she hath found, calleth together her friendes and neighbours,
ſaying: Reioyce with me, becauſe I haue found the grote which I had
loſt? \V So I ſay to you, there shal be ioy
\LNote{Before the Angels}{The
\MNote{The Angels & Saints know our hartes.}
Angels and other celeſtial Spirits in Heauen, doe reioyce at euery
ſinners conuerſion: they know then and haue care of vs, yea our hartes
and inward repentance be open to them: how then can they not heare our
prayers? And betwixt Angels and the bleſſed ſoules of Saints there is
no difference in this caſe,
\CNote{\XRef{Mt.~22,~30.}}
the one being as highly exalted as the
other, and as neere God, in whom and by whom only they ſee and know our
affaires, as the other.}
before the Angels of God vpon one ſinner that doth penance.

\V
%%% !!! unmarked note!
\SNote{The prodigal ſonne is a parable, both of the Gentils conuerſion,
& alſo euery diſſolute ſinner penitently returning to God.}
And he ſaid: A certaine man had two ſonnes; \V and the yonger of them
ſaid to his father: Father, giue me the portion of ſubſtance that
belongeth to me. And he deuided vnto them the ſubſtance. \V And not many
daies after the yonger ſonne gathering al his things together went from
home into a farre countrie: and there he waſted his ſubſtance, liuing
riotouſly. \V And after he had ſpent al, there fel a ſore famine in that
countrie, and he began to be in need. \V And he went, and cleaued to one
of the citizens of that countrie. And he ſent him into his farme to feed
ſwine. \V And he would faine haue filled his bellie of the huſkes that
the ſwine did eate; and no body gaue vnto him. \V And returning to him
ſelf he ſaid: How many of my fathers hirelings haue aboundance of bread,
and I here perish for famine? I wil ariſe, and wil goe to my father, and
ſay to him: Father, I haue ſinned againſt Heauen and before thee: \V I
am not now worthie to be called thy ſonne: make me as one of thy
hirelings. \V And riſing vp he came to his father. And
\SNote{Gods wonderful and tender mercie toward penitent ſinners.}
when he was yet farre off, his father ſaw him, and was moued with
mercie, and running to him fel vpon his necke, and kiſſed him. \V And
his ſonne ſaid to him: Father, I haue ſinned againſt Heauen & before
thee, I am not now worthie to be called thy ſonne. \V And the father
%%% 2453
ſaid to his ſeruants: Quickly bring forth the firſt ſtole, and doe it on
him, and put a ring vpon his hand, and shoes vpon his feet: \V and bring
\LNote{The fatted calfe}{This
\MNote{The B.~Sacrament & Sacrifice of the Altar.}
feaſting and feſtiuitie (ſaith S.~Auguſtine
\Cite{li.~2. qu. Euang. c.~33. to.~4.})
are now celebrated throughout the whole world, the Church being dilated
and ſpred: for, that calfe, in the body and bloud of our Lord, is both
offered to the Father, and alſo feedeth the whole houſe. And as the calfe
ſignifieth the B.~Sacrament of the body and bloud of Chriſt, ſo the
firſt ſtole may ſignifie our innocencie reſtored in Baptiſme: and the
reſt, other graces and guifts giuen vs in the other Sacraments.}
the fatted calfe, and kil it, and let vs eate, and make merie: \V
becauſe this my ſonne was dead, and is reuiued: was loſt, and is
found. And they began to make merie. \V But his elder ſonne was in the
field,
%%% o-2284
and when he came and drew nigh to the houſe, he heard muſicke and
dancing: \V and he called one of the ſeruants, and asked what theſe
things should be. \V And he ſaid to him: Thy brother is come, and thy
father hath killed the fatted calfe, becauſe he hath receiued him
ſafe. \V But he had indignation, and would not goe in. His father
therfore going forth began to deſire him. \V But he anſwering ſaid to
his father: Behold, ſo many yeares doe I ſerue thee, and I neuer
tranſgreſſed thy commandement, and thou didſt neuer giue me a kidde to
make merie with my freindes: \V but after that this thy ſonne, that hath
deuoured his ſubſtance with whoores, is come, thou haſt killed for him
the fatted calfe. \V But he ſaid to him: Sonne, thou art alwaies with
me, and al my things are thine. \V But it behoued vs to make merie and
be glad becauſe this thy brother was dead, and is reuiued, was loſt, and
is found. 


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
