%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 2447
%%% o-2277
\startChapter[
  title={Chapter 13}
  ]

\Summary{He threatneth the Iewes to be ſoone forſaken vnles they doe
  pennance, 10.~and confoundeth them for maligning him for his
  miraculous good-doing on the Sabboths. 18.~but his Kingdom (the
  Church) as contemptible as it ſeemeth to them now in the beginning,
  shal ſpread ouer al the world, 20.~and conuert al, 23.~and what an
  hart-ſore it shal be to them at the laſt day to ſee them ſelues
  excluded from the glorie of this Kingdom, and the Gentils admitted in
  their place. 31.~He foretelleth that it is not Galilee that he feareth,
  but that obſtinate & reprobate Hieruſalem wil needes murder him, as
  alſo his meſſengers afore and after him.}

And there were certaine preſent at that very time telling him of the
Galilæans, whoſe bloud Pilate mingled with their Sacrifices. \V And he
anſwering ſaid to them: Thinke you that
\LNote{Theſe Galilæans}{It
\MNote{Some punished for example.}
is Gods mercie that he ſtraight punisheth not al offenders, but ſome few
for a warning to al: as that for Schiſme he ſtriketh not al ſuch as haue
forſaken the Church and the lawful Prieſts,
\CNote{\XRef{Nu.~16,~2.}}
as he did Core and his
complices: that for ſpoile of Churches he reuengeth not al,
\CNote{\XRef{Mac.~3.}}
as he did
Heliodorus: and al that vow and reuoke their guifts to God,
\CNote{\XRef{Act.~5.}}
as Ananias
and Sapphira. Some few therfore for their iuſt deſerts, be ſo handled
for example, to prouoke al others guilty of the ſame crimes to doe
penance. Which if they doe not in this life, they ſhal al aſſuredly
periſh in the next world.
\Cite{Optatus li.~1. cont. Parmen. ſub finem.}}
theſe Galilæans were ſinners more thẽ al the Galilæans, that they
ſuffred ſuch things? \V No, I ſay to you: but vnleſs you
\TNote{\L{nisi pœnitentiam habueritis}}
\SNote{Or as it is vttered in other places, \Emph{do pennance}, which in
the new Teſtament ſignifieth perfect repentance. See
\XRef{Annot. S.~Mar. 3,~2: 11,~21.}}
haue pennance, you ſhal al likewiſe periſh. \V As thoſe eighteene vpon
whom the toure fel in Siloe; and ſlew them: thinke you that they alſo
were debters aboue al the men that dwel in Hieruſalem? \V No, I ſay to
you: but if you haue not pennance you ſhal al likewiſe periſh.

%%% o-2278
\V And he ſaid this ſimilitude: A certaine man had
\SNote{The figtree with only leaues & no fruit is the Iewes Synagogue, &
euery other people or Perſon which hath faith and faire wordes, & no
good workes.}
a figtree planted in his vineyard, and he came ſeeking for fruit on it,
and found not,
%%% 2448
\V And he ſaid to the dreſſer of the vineyard, Loe it is three yeares
ſince I come ſeeking for fruit vpon this figtree, and I find not. Cut it
downe therfore; whereto doth it alſo occupie the ground? \V But he
anſwering ſaid to him: Lord, let it alone this yeare alſo, vntil I digge
about it, and dung it. \V And if happily it yeald fruit; but if not,
hereafter thou ſhalt cut it downe.

\V And he was teaching in their Synagogue on the Sabboths. \V And behold
a woman that had a ſpirit of infirmitie eighteene yeares: and ſhe was
crooked, neither could ſhe looke vpward at al. \V Whom when \Sc{Iesvs}
ſaw, he called her vnto him, and ſaid to her: Woman, thou art deliuered
from thy infirmitie. \V And he impoſed hands vpon her, and forthwith ſhe
was made ſtraight and glorified God. \V And the Archſynagogue anſwering
(becauſe he had indignation that \Sc{Iesvs} had cured on the Sabboth)
ſaid to the multitude: Six daies there are wherein you ought to
worke. In them therfore come, and be cured; and not in the Sabboth
day. \V And our Lord anſwering to him, ſaid: Hypocrite, doth not euery
one of you vpon the Sabboth looſe his oxe or his aſſe from the manger,
and leadeth them to water? \V But
\LNote{This daughter}{We
\MNote{Diſeaſes not natural.}
may ſee that many diſeaſes which ſeeme natural, doe proceed of the Diuel
by Gods permiſſion, either for ſinne, or for probation: and both thoſe
kindes Chriſt ſpecially cured, for that no natural medecins could cure
them, and ſpecially becauſe he came to diſſolue the workes of Satan both
in body and ſoul.}
this daughter of Abraham whom Satan hath bound, loe, theſe eighteene
yeares, ought not ſhe to be looſed from this bond on the Sabboth day? \V
And when he ſaid theſe things, al his aduerſaries were aſhamed: and al
the People reioyced in al things that were gloriouſly done of him.

\V He ſaid therfore:
\CNote{\XRef{Mt.~13,~31.}
\XRef{Mr.~4,~30.}}
Whereunto is the Kingdom of God like, and whereunto
ſhal I eſteeme it like? \V It is like to a
\SNote{See
\XRef{Annot. Matth.~13,~31.}}
muſtard ſeed, which a man tooke and caſt into his garden and it grew;
and became a great tree, & the foules of the aire reſted in the boughes
therof. \V And againe he ſaid:
\CNote{\XRef{Mt.~13,~33.}}
Like to what ſhal I eſteeme the Kingdom
of God? \V It is like to leauen, which a woman tooke and hid in three
meaſures of meale, til the whole was leauened. \V And he went by cities
and townes teaching, and making his iourney vnto Hieruſalem.

\V And a certaine man ſaid to him: Lord, be they few that are ſaued?
\CNote{\XRef{Mt.~7,~13.}}
But he ſaid to them: \V
\SNote{Chriſtians in their liues muſt ſeeke the ſtrait way, but in
religion the ancient common way.}
Striue to enter
\LNote{By the narrow}{Our Lord is not contrarie to himſelf in that he
anſwereth, the
gate to be ſtraite, and few to be ſaued, whereas els where he ſaid, that
many should come from the Eaſt & Weſt &c.\ & ioy with Abraham in the
Kingdom of Heauen.
\XRef{Mat.~8,~11.}
For though they be few in reſpect of the wicked of al ſortes, yet they
be many in them ſelues and in the ſocietie of Angels. The wheate cornes
are ſcarce ſeen at the threſhing, when they are medled with the chaffe;
but when the il are remoued, the whole barne of Heauen shal be
filled. So ſaith
\Cite{S.~Aug. Ser.~32. de verbis Do.}}
by the narrow gate: becauſe many, I ſay to you,
\LNote{Shal ſeeke}{Many
\MNote{Penance.}
would be ſaued and looke to be ſaued, but can not, becauſe they wil not
take paines to enter in at ſo ſtraite a paſſage, that is to ſay, to faſt
much, pray often, doe great penance for their ſinnes, liue in holy
Churches diſcipline, abſtaine from the pleaſures of this world, and
ſuffer perſecution and loſſe of their goods and liues for Chriſtes
ſake.}
ſhal ſeeke to enter,
%%% o-2279
and ſhal not be able. \V But when the good-man of
the houſe ſhal enter in, and ſhut the doore, and you ſhal begin to ſtand
without, and knocke at the doore, ſaying, Lord open to vs: and he
anſwering ſhal ſay to you, I know you not whence you are: \V then you
ſhal begin to ſay: We did
\LNote{Eate before thee}{It
\MNote{Schiſme.}
is not enough to feed with Chriſt in his Sacraments, or to heare his
word in the Church, to chalenge Heauen thereby, vnleſſe we liue in
vnitie of the Catholike Church. So
\CNote{\Cite{Con. lit. Petil. li.~2. c.~35.}}
S.~Auguſtine applieth this againſt the 
\Fix{Danatiſtes,}{Donatiſtes,}{obvious typo, fixed in other}
that had the very ſame ſeruice and Sacraments which the Catholike Church
had, but yet ſeuered themſelues from other Chriſtian countries by
Schiſme.}
eate before thee and drinke, and in our ſtreetes didſt thou teach. \V
And he ſhal ſay to you, I know you not whence you are, depart from me al ye
workers of iniquitie. \V There ſhal be weeping and gnaſhing of teeth:
when you ſhal ſee Abraham and Iſaac and Iacob, and al the Prophets in
the Kingdom of God, and you to be thruſt out. \V And there ſhal come
from the Eaſt and Weſt & the North and the South; and ſhal ſit downe in
the Kingdom of God. \V And behold, they are
\SNote{The Gentils comming into Gods fauour later, are preferred before
the Iewes which were firſt.}
laſt that ſhal be firſt, and they be firſt that shal be laſt.

\V The ſame day there came certaine of the Phariſees, ſaying to him:
Depart and get thee hence, becauſe Herod wil kil thee. \V And he ſaid to
them: Goe, and tel that foxe, Behold I caſt out Diuels, and perfit cures
%%% 2449
this day and to morow, and the third day I am conſummate. \V But yet I
muſt walke this day and to morow and the day following, becauſe it
\TNote{\L{non capit}, \G{οὐκ ἐνδέχεται}}
cannot be that a Prophet periſh out of Hieruſalem. \V
\CNote{\XRef{Mt.~23,~37.}}
Hieruſalem,
Hieruſalem which killeſt the Prophets, and ſtoneſt them that are ſent to
thee, how often would I gather thy children as the bird doth her brood
vnder her wings, and
\SNote{The Iewes loſt their preeminence, by their own free wil, & not
by Gods cauſing: who ceaſed not to cal and crie vpon them, and they
would not heare: whereby free wil is plainly proued.}
thou wouldeſt not? \V Behold your houſe ſhal be left deſert to you. And
I ſay to you, that you ſhal not ſee me til it come when you ſhal ſay:
Bleſſed is he that commeth in the name of our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
