%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 2432
%%% o-2258
\startChapter[
  title={Chapter 8}
  ]

\Summary{Going ouer al Galilee with his traine, 4.~he preacheth to the
  Iewes in parables becauſe of their reprobation, 9.~but to his
  Diſciples manifeſtly; becauſe he wil not for the Iewes incredulity
  haue his cõming fruſtrate: 19.~ſignifying alſo that we are his kinne
  (though we be Gentils) and not his carnal brethren the Iewes. 22.~To
  whom alſo (ſignified by the Geraſens) after the tempeſt in his ſleep
  (that is, in his death) & calme in his reſurrection, he commeth: but
  they preferring their temporals before his preſence, he leaueth them
  againe. 41.~Likewiſe comming to cure the Iewes (who were borne when
  the Gentils ſickned about Abrahams time) he is preuented with the
  faith of the Gentils, and then the Iewes die, but them alſo in the end
  he wil reſtore.}

And it came to paſſe afterward, and he made his iourney by cities and
townes preaching and euangelizing the Kingdom of God; and the Twelue
with him, \V and ſome women that had been cured of wicked Spirits and
infirmities;
\CNote{\XRef{Mr.~16,~9.}}
Marie which is called Magdalene, out of whom ſeuen Diuels were gone
%%% o-2259
forth, \V and Ioane the wife of Chuſa Herods Procuratour, and Suſan, and
\TNote{\L{aliæ multæ}}
many others that
\LNote{That did miniſter}{It
\MNote{Holy women that followed Chriſt.}
was the cuſtome of the Iewes that women of their ſubſtance did miniſter
meate drinke and cloth to their teachers, going about with them. Which
becauſe it might haue been ſcandalous among the Gentils,
\CNote{\XRef{1.~Cor.~9,~5--12.}}
S.~Paul maketh
mention that he vſed it not. And they miniſtred to our Lord of their
ſubſtance for this cauſe, that he whoſe ſpiritual benefites they reaped,
might reape their carnal things.}
did miniſter vnto him of their ſubſtance.

\V And
\CNote{\XRef{Mt.~13,~1.}
\XRef{Mr.~4,~1.}}
when a very great multitude aſſembled and haſtned out of the
cities vnto him, he ſaid by a ſimilitude. \V The ſower went forth to ſow
his ſeed. And whiles he ſoweth, ſome fel by the way ſide, and was troden
vpon, and the foules of the aire did eate it. \V And otherſome fel vpon
the rock; and being ſhot vp, it withered, becauſe it had not
moiſture. \V And otherſome fel among thornes, and the thornes growing vp
withal, choked it. \V And otherſome fel vpon good ground; and being ſhot
vp, yealded fruit an hundred fold. Saying theſe things he cried: He that
hath eares to heare, let him heare.

\V And his Diſciples asked him what this parable was. \V To whom he
ſaid: To you it is giuen to know the myſterie of the Kingdom of God; but
to the reſt in parables,
\SNote{See the
\XRef{Annotations vpon S.~Matthew. c.~13,~14.}}
that
\CNote{\XRef{Eſ.~6,~9.}}
ſeeing they may not ſee, and hearing may not vnderſtand. \V And the
parable is this: The ſeed, is the Word of God. \V And they beſides the
way, are thoſe that heare, then the Diuel commeth, and taketh the Word
out of their hart, leſt beleeuing they be ſaued. \V For they vpon the
rock; ſuch as when they heare, with ioy receaue the Word: and theſe haue
no roots; becauſe
\SNote{Againſt the Heretikes that ſay, faith once had can not be loſt,
and that he which now hath not faith, neuer had.}
for a time they beleeue, and in time of tentation they reuolt. \V And
that which fel into thornes, are they that haue heard, and going their
waies, are choked with cares and riches and pleaſures of this life, and
render not fruit. \V And that vpon good ground, are they which in a good
and verie good hart, hearing the Word, doe retaine it, and yeald fruit
in patience.

\V And no man lighting a candel doth couer it with a veſſel, or put it
vnder a bed; but ſetteth it vpon a candelſticke, that they that enter
in, may ſee the light. \V For there is not any thing ſecret, that ſhal
not be made manifeſt; nor hid, that ſhal not be knowen, & come
abrode. \V See
%%% 2433
therfore how you heare. For he that hath, to him ſhal be giuen; and
whoſoeuer hath not, that alſo which he thinketh he hath, ſhal be taken
away from him.

\V And
\CNote{\XRef{Mt.~12,~46.}
\XRef{Mr.~3,~31.}}
his mother and brethren came vnto him; and they could not come at
him for the multitude. \V And it was told him: Thy mother and
\LNote{Thy brethren}{Theſe
\MNote{The brethren of Chriſt.}
brethren of our Lord, were not the ſonnes of the B.~Virgin \Sc{Marie}
the mother of God, as Heluidius wickedly taught: neither are they to be
thought (as ſome others ſay) the ſonnes of Ioſeph by an other wife: for
\CNote{\Cite{Hiero. cont. Heluid. c.~9.}}
(as S.~Hierom writeth) not only our Lady was a virgin, but by reaſon of
her, Ioſeph alſo: that our Sauiour might be borne of a virginal
matrimonie. But they are called his brethren
Ibidẽ. c.~8.
(according to the vſual
ſpeach of the Scriptures) becauſe they were his coſins, either the
ſonnes of Ioſephs brother, or (as the more receiued opinion is) the
ſonnes of our Ladies ſiſter called Marie of Iames, which Iames therfore
is alſo called the brother of our Lord.}
thy brethren ſtand without, deſirous to ſee thee. \V Who anſwering ſaid
to them: My
\SNote{He did not here diſdainfully ſpeake of his mother, but teacheth
that our ſpiritual kinred is to be preferred before carnal cognation.
\Cite{Hilar. in 12.~Mat.}}
mother and
%%% o-2260
my brethren, are they that heare the Word of God and doe it.

\V
\CNote{\XRef{Mat.~8,~23.}
\XRef{Mar.~4,~26.}}
And it came to paſſe one day; and he went vp into a boat, and his
Diſciples, and he ſaid to them: Let vs ſtrike ouer the lake. And they
launched forth. \V And when they were ſailing, he ſlept; and there fel a
ſtorme of wind into the lake, and they
\TNote{\L{complebantur}}
were filled, and were in danger. \V And
\SNote{See the
\XRef{Annotations vpon S.~Matthew. c.~8,~24.}}
they came and raiſed him, ſaying: Maiſter, we periſh. But he riſing,
rebuked the wind and the tempeſt of water; and it ceaſed, and there was
made a calme. \V And he ſaid to them: Where is your faith? Who fearing,
marueled one to an other, ſaying: Who is this (trow ye) that he
commandeth both the winds and the ſea, and they obey him? \V
\CNote{\XRef{Mt.~8,~28.}
\XRef{Mr.~5,~1.}}
And they
ſailed to the countrie of the Geraſens which is ouer againſt Galilee.

\V And when he was come forth to the land, there met him a certaine man
that had a Diuel now a very long time, and he did weare no clothes,
neither did he tarie in houſe, but in the monuments. \V And as he
ſaw \Sc{Iesvs}, he fel downe before him; and crying out with a great
voice, he ſaid: What is it to me & thee \Sc{Iesvs} Sonne of God moſt
high? I beſeech thee doe not tormẽt me. \V For he commanded the vncleane
Spirit to goe forth out of the man. For many times he caught him, and he
was bound with chaines, and kept with fetters; and breaking the bonds
was driuen of the Diuel into the deſerts. \V And \Sc{Iesvs} asked him
ſaying: What is thy name? But he ſaid: Legion; becauſe many Diuels were
entred into him. \V And they beſought him that he would not command them
to goe into the depth. \V And there was there a heard of many ſwine
feeding on the mountaine; & they deſired him that he would permit thẽ to
enter into them. And he permitted them. \V The Diuels therfore went
forth out of the man, and entred into the ſwine; and the heard with
violence went headlong into the lake, and was ſtifled. \V Which when the
ſwineheards ſaw done, they fled, ãd told into the citie and into the
townes. \V And they went forth to ſee that which was done; and they came
to \Sc{Iesvs}, and found the man, out of whom the Diuels were gone
forth, ſitting at his feet, clothed, and wel in his wits, and they were
afraid. \V And they alſo that had ſeen, told them how he had been made
whole from the legion. \V And al the multitude of the countrie of the
Geraſens beſought him to depart from
%%% o-2261
them; for they were taken with great feare. And he going vp into the
boat, returned. \V And the man out of whom the Diuels were departed,
deſired him that he might be with him. But \Sc{Iesvs} dimiſſed him,
ſaying: \V Returne into thy houſe, and tel how great things God hath
done to thee. And he went through the whole citie, preaching how great
things \Sc{Iesvs} had done to him.

%%% 2434
\V And it came to paſſe; when \Sc{Iesvs} was returned, the multitude
receiued him. And al were expecting him. \V And
\CNote{\XRef{Mt.~9,~18.}
\XRef{Mr.~5,~22.}}
behold there came a man
whoſe name was Iairus, and he was Prince of the Synagogue: and he
fel at the feet of \Sc{Iesvs}, deſiring him that he would enter into his
houſe, \V becauſe he had an only daughter almoſt twelue yeares old, and
she was a dying. And it chanced, whiles he went, he was thronged of the
multitudes.

\V And there was
\SNote{See the
\XRef{Annotations vpon S.~Mat. c.~9,~19.}}
a certaine woman in a fluxe of bloud from twelue yeares paſt, which had
beſtowed al her ſubſtance vpon Phyſicions, neither could she be cured of
any: \V she came 
behind him, and touched the hemme of his garment; and forthwith the
fluxe of her bloud ſtinted. \V And \Sc{Iesvs} ſaid: Who is it that
touched me? And al denying,
\SNote{It is an euidẽt ſigne of prerogatiue, that Peter only is named ſo
often as cheefe of the company.
\XRef{Mar.~1,~36.}
\XRef{Act.~5,29.}
\XRef{Luc.~9,~22.}
\XRef{Mar.~16,~7.}
\XRef{1.~Cor. 15,~5.}}
Peter ſaid, and they that were with him: Maiſter, the multitudes throng
and preſſe thee, and doeſt thou ſay, Who touched me? \V And \Sc{Iesvs}
ſaid: Some bodie hath touched me; for I know that there is vertue
proceeded from me. \V And the woman ſeeing, that she was not hid, came
trembling, and fel downe before his feet; and for what cauſe she had
touched him, she shewed before al the People, and how forthwith she was
made whole. \V But he ſaid to her: Daughter, thy faith hath made thee
ſafe, goe thy way in peace.

\V As he was yet ſpeaking, there commeth one to the Prince of the
Synagogue, ſaying to him, That thy
\Fix{daugther}{daughter}{obvious typo, fixed in other}
is dead, trouble him not. \V
And \Sc{Iesvs} hearing this word, anſwered the father of the maide:
Feare not;
\SNote{See the
\XRef{Annotations vpon S.~Marke c.~5,~56.}}
beleeue only, and she shal be ſafe. \V And when he was come to the
houſe, he permitted not any man to enter in with him, but Peter, and
Iames, and Iohn, and the father and mother of the maide. \V And al wept,
and mourned for her. But he ſaid: Weep not, the maide is not dead, but
ſleepeth. \V And they derided him, knowing that ſhe was dead. \V But he
holding her hand cried ſaying: Maide ariſe. \V And
\LNote{Her ſpirit returned}{This
\MNote{A third place after this life.}
returning of the ſoules againe into the bodies of them whom \Sc{Christ}
and his Apoſtles raiſed from death (ſpecially Lazarus who had been dead
foure
\Fix{daies doth)}{daies) doth}{obvious typo, fixed in other}
euidently proue a third place againſt our aduerſaries, that ſay, euery
one goeth ſtraight to Heauen or Hel. For it can not be thought that they
were called from the one or the other, and therfore from ſome third
place.}
her ſpirit returned, and she roſe
%%% o-2262
incontinent. And he bade them giue her
to eate. \V And her parentes were aſtoniſhed, whom he commanded to tel
no man that which was done.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
