%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 2463
%%% o-2295
\startChapter[
  title={Chapter 20}
  ]

\Summary{To the Iewes he auoucheth his power by the witnes of Iohn who
  was a man ſent of God. 9.~& foretelleth in a parable their reprobation
  moſt worthy (with the vocation of the Gentils in their place) 17.~and
  conſequently their irreparable damnation that shal enſue
  thereof. 20.~He defeateth their ſnare about paying tribute to Cæſar:
  27.~he anſwereth alſo the inuention of the Sadducees againſt the
  Reſurrection. 40.~And ſo hauing put them al to ſilence, 41.~he turneth
  and poſeth them, becauſe they imagined that Chriſt should be no more
  but a man: 45.~bidding al to beware of the Scribes (authours of the
  Iewes ſchiſme from him) being ambitious & hypocrites.}

%%% o-2296
And
\CNote{\XRef{Mt.~21,~23.}
\XRef{Mr.~11,~27.}}
\MNote{\Sc{Tveſday}.}
it came to paſſe in one of the daies, when he was teaching the people in
the Temple & euangelizing, the cheefe Prieſts & the Scribes with the
Ancients aſſembled, \V and ſpake ſaying to him,
\SNote{See
\XRef{Annot. Mat. c.~21,~23.}}
Tel vs, in what power doeſt thou theſe things? or who is he
%%% 2464
that hath giuen thee this power? \V And \Sc{Iesvs} anſwering, ſaid to
them: I wil alſo aske you one word: Anſwer me. \V The Baptiſme of Iohn
was it from Heauen, or of men? \V But they thought within themſelues,
ſaying: That if we ſay, From Heauen, he wil ſay: Why then did you not
beleeue him? \V But if we ſay, of men, the whole people wil ſtone vs:
for they are certaine that Iohn is a Prophet. \V And they anſwered that
they knew not whence it was. \V And \Sc{Iesvs} ſaid to them: Neither doe
I tel you in what power I doe theſe things.

\V And he began to ſay to the people this parable:
\CNote{\XRef{Eſa.~5,~1.}
\XRef{Mt.~21,~33.}
\XRef{Mr.~12,~1.}}
A certaine
\SNote{See the
\XRef{marginal annotations Marc.~12.}}
man planted a vineyard, & let it out to husbandmen: and he was from home
a long time. \V And in time he ſent to the husbandmen a ſeruant, that
they ſhould giue him of the fruit of the vinyard. Who beating him ſent
him away emptie. \V And againe he ſent another ſeruant. But they beating
him alſo and reprochfully abuſing him, ſent him away enptie. \V And
againe he ſent the third: who wounding him alſo, caſt him out. \V And
the Lord of the vineyard ſaid: What ſhal I doe? I wil ſend my beloued
ſonne: perhaps when they ſhal ſee him, they wil reuerence him. \V Whom
when the husbandmen ſaw, they thought within themſelues, ſaying: This is
the heire, let vs kil him, that the heritage may be ours. \V And caſting
him forth out of the vineyard, they killed him. What therfore wil the
Lord of the vineyard doe to them? \V He wil come, and wil deſtroy theſe
husbandmen, and wil giue the vinyard to others. Which they hearing, ſaid
to him: God forbid. \V But he beholding them ſaid: What is this then
that is written,
\CNote{\XRef{Ps.~117,~22.}}
\Emph{The ſtone which the builders reiected, the ſame
is become into the head of the corner}? \V Euery one that falleth vpon
this ſtone, ſhal be quaſhed: and vpon whom it ſhal fal, it ſhal breake
him to pouder. \V And the cheefe Prieſts and Scribes ſought to lay hands
vpon him that houre: and they feared the people, for they knew that he
ſpake this ſimilitude to them.

%%% o-2297
\V
\CNote{\XRef{Mt.~22,~15.}
\XRef{Mr.~12,~13.}}
And watching, they ſent ſpies which ſhould feine themſelues iuſt:
that they might take him in his talke, and deliuer him to the
principaltie and power of the Preſident. \V And they asked him, ſaying:
Maiſter, we know that thou ſpeakeſt and teacheſt rightly; and thou doeſt
not accept perſon, but teacheth the way of God in truth. \V Is it lawful
for vs to giue tribute to Cæſar, or no? \V But conſidering their guile,
he ſaid to them: Why tempt you me? \V Shew me a penie. Whoſe image hath
it and inſcription? They anſwering ſaid: Cæſars. \V And he ſaid to them:
Render therfore the things that are Cæſars,
\SNote{So duties muſt be done to Princes, that our dutie to God be not
neglected: See
\XRef{Annot. Mat. c.~22,~21.}}
to Cæſar: and the things that are Gods, to God. \V And they could not
reprehend his word before the people: and marueling at his anſwer, they
held their peace. \V
\CNote{\XRef{Mt.~22,~23.}
\XRef{Mr.~12,~18.}}
And there came certaine of the Sadducees, which
denie that there is a reſurrection, and they asked him, \V ſaying:
Maiſter, Moyſes gaue vs in writing:
\CNote{\XRef{Deu.~25,~5.}}
If a mans brother die hauing a wife,
and he haue no children, that his brother take her to wife, and raiſe vp
ſeed to his brother. \V There were therfore ſeuen brethren: and the
firſt tooke a wife, and died without children. \V And the next tooke
her, & he died without child. \V And the third tooke her. In like manner
alſo al the ſeuen, and they left no ſeed, and died. \V Laſt of al the
woman
%%% 2465
died alſo. \V In the reſurrection therfore, whoſe wife ſhal ſhe be of
them? ſithens the ſeuen had her to wife. \V And \Sc{Iesvs} ſaid to them:
The children of this world marrie, and are giuen in marriage: \V But
they that
\LNote{Shal be counted worthie}{This
\MNote{To be worthie of
\Fix{heauen,}{Heauen}{likely typo, same in other}
or to deſerue & merite it.}
truth and
\Fix{Speach}{ſpeach}{likely typo, fixed in other}
that good men be worthy of Heauen, is according to the Scriptures, and
ſignifieth that mans workes done by Chriſts grace doe condignely or
worthily deſerue eternal ioy: As
\XRef{Sap.~3.}
\Emph{God proued them, and found them worthy of himſelf}: and
\XRef{Mat.~10.}
\Emph{He that loueth his father more then me, is not worthy of me}: and
\XRef{Coloſ.~1.}
\Emph{That you may walke worthy of God}: and moſt plainly
\XRef{Apoc.~3.}
\Emph{They shal walke with me in white, becauſe they are worthy}: as of
Chriſt 
\XRef{(c.~3)}
\Emph{Thou art worthie, O Lord, to receiue glorie &c.} And that, to be
counted worthie, & to be worthie, is here al one, it is plaine, by the
Greeke word, which S.~Paul vſeth ſo, as the aduerſaries owne
\MNote{The new Teſtament. an.~1580.}
English Teſtament doe teſtifie, reading thus
\XRef{Hebre.~10.}
\CNote{\Cite{The new Teſtament. an.~1580.}}
\Emph{Of how much ſorer punishment shal he be
\TNote{\G{καταξιωθέντες}. \G{ἀξιωθήσεται.}}
worthie, which &c.} And it muſt needes ſignifie, becauſe men for ſinnes
are not only counted, but are indeed worthie of punishment, as
themſelues doe grant. They doe greatly therfore forget themſelues, and are
ignorant in the Scriptures, and know not the force nor the valure of the
grace of God, which doth not only make our labours grateful to God, but
worthie of the reward which he hath prouided for ſuch as loue him. See
the
\XRef{Annot. 2.~Theſſ. 1,~5.}}
ſhal be
\SNote{The greeke \G{οἱ καταξιωθέντες} importeth alſo this
much \Emph{They that are made worthie}, to wit, by the grace of God; and
ſo they are indeed worthie: as alſo in the
\XRef{next chapter verſe~36.}
&
\XRef{2.~Theſs. 1,~5.}}
counted worthie of that world and the reſurrection from the dead,
neither marrie, nor take wiues; \V neither can they die any more, for
they are
\LNote{Equal to Angels}{Saints
\MNote{The dignitie of Saints.}
of our kind, now in their ſoules, and after their reſurrection in body
and ſoule together, shal be in al things equal to Angels: and for degree
of bliſſe, many Saints of greater merit shal be aboue diuers Angels: as
S.~Iohn Baptiſt, the Apoſtles, and others, and our B.~Lady aboue al the
orders of holy Spirits in dignitie and bliſſe: & no maruel, our nature
by Chriſt being ſo highly exalted aboue al Angels.}
equal to Angels: and they are the ſonnes of God, ſeeing they are the
ſonnes of the reſurrection. \V But that the dead riſe againe, Moyſes
alſo ſhewed, beſide the buſh, as he calleth the Lord:
\CNote{\XRef{Exo.~3,~6.}}
\Emph{The God of
Abraham, and the God of Iſaac, and the God of Iacob.} \V For God is not
of the dead, but of the liuing: for al liue to him. \V And certaine of
the Scribes anſwering, ſaid to him: Maiſter, thou haſt ſaid wel. \V And
further they durſt not aske him any thing.

\V But he ſaid to them:
\CNote{\XRef{Mt.~22,~44.}
\XRef{Mr.~12,~36.}}
How ſay they that Chriſt is the ſonne of
Dauid, \V and Dauid himſelf ſaith in the booke of Pſalmes:
\CNote{\XRef{Pſ.~109,~1.}}
\Emph{The
Lord ſaid to my Lord, ſit on my right hand, \V til I put thine enemies,
the footſtoole of thy feet}? \V Dauid then calleth him Lord: and how is
he his ſonne?

%%% o-2298
\V And al the people hearing him, he ſaid to his Diſciples: \V
\CNote{\XRef{Mt.~23,~6.}
\XRef{Mr.~12,~38.}}
Beware of
the Scribes, that wil walke in robes, and loue ſalutations in the
market-place, and the firſt chaires in the ſynagogues, and the cheefe
roomes in feaſtes. \V Which deuoure widowes houſes: feining long
praier. Theſe ſhal receiue greater damnation.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
