%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 2418
%%% o-2242
\startChapter[
  title={Chapter 3}
  ]

\Summary{Iohn, to prepare al to Chriſt (as Eſay had prophecied of him)
  baptizeth them to penance, 7.~inſinuating their reprobation, and the
  Gentils vocation. 10.~teaching alſo & exhorting each ſort to doe their
  dutie. 15.~That himſelf is not Chriſt, he sheweth by the difference of
  their two Baptiſmes: 17.~and ſaith that Chriſt wil alſo iudge his
  baptized. 19.~Iohns impriſonment. 21.~Chriſt being him ſelfe alſo
  baptized of Iohn, hath 
  teſtimonie from Heauen, 23.~as he whoſe Generation reduceth vs againe
  to God.}

%%% o-2243
And
\MNote{The ſecond part: The preparation that was made to the
manifeſtation of Chriſt.}
in the fifteenth yeare of the empire of Tiberius Cæſar, Pontius Pilate
being Gouernour of Iewrie, and Herod being Tetrarch of Galilee, and
Philip his brother Tetrarch of Ituréa and the countrie Trachonitis, and
Lyſanias Tetrarch of Abilina,
%%% 2419
\V vnder the high Prieſts Annas and Caiphas: the Word of our Lord was
made vpon Iohn the ſonne of Zacharie, in the deſert. \V And
\CNote{\XRef{Mt.~3,~1.}
\XRef{Mr.~1,~1.}}
he came into al the countrie of Iordan, preaching the Baptiſme of
\SNote{Pennance prepareth the way to Chriſt.}
penance vnto remiſsion of ſinnes; as it is written in the booke of the
ſayings of Eſay the Prophet: \V
\CNote{\XRef{Eſa.~40,~3.}}
\Emph{A voice of one crying in the
deſert; prepare the way of our Lord, make ſtraight his paths. \V Euery
valley shal be filled; and euery mountaine and hil shal be made low, and
crooked things shal become ſtraight; and rough waies, plaine: \V And al
flesh shal ſee the} \Sc{salvation} \Emph{of God.}

\V He ſaid therfore to the multitudes that went forth to be baptized of
him:
\CNote{\XRef{Mt.~3,~7.}}
Ye vipers broods, who hath shewed you to flee from the wrath to
come? \V Yeald therfore
\SNote{Fruits of pennance be works ſatisfactorie.}
fruits worthie of penance; and doe ye not begin to ſay, we haue Abraham,
to our father. For I tel you, that God is able of theſe ſtones to raiſe
vp children to Abraham. \V For now the axe is put to the roote of the
trees.
\SNote{A man without good works is vnfruitful, and shal be caſt into
euerlaſting fire.}
Euery tree therfore that yealdeth not good fruit,
\Var{shal be}{is}
cut downe, and caſt into fire. \V And the multitudes asked him, ſaying:
What shal we doe then? \V And he anſwering, ſaid vnto them:
\SNote{Almes counſeled or enioyned for ſinnes and to auoid damnation.}
He that hath two coats, let him giue to him that hath not; and he that
hath meat, let him doe likewiſe. \V And the Publicans alſo came to be
baptized, and ſaid to him: Maiſter, what shal we doe? \V But he ſaid to
them: Doe nothing more then that which is appointed you. \V And the
ſouldiars alſo asked him, ſaying: What shal we alſo doe? And he ſaid to
them: Vexe not, neither calumniate any man; and be content with your
ſtipends.

\V And
\SNote{Iohn was ſo holy that many might by errour eaſily thinke he was
Chriſt.}
the People imagining, and al men thinking in their harts of Iohn, leſt
perhaps he were Chriſt: \V Iohn anſwered, ſaying vnto al:
\CNote{\XRef{Mt.~3,~11.}
\XRef{Mr.~1,~8.}}
I indeed baptize you with water;
\SNote{How ſay then the Heretikes that the Baptiſme of Chriſt is of no
greater vertue then Iohns?}
but there
\Var{shal come}{commeth}
a mightier then I, whoſe latchet of his shoes I am not worthie to looſe;
\CNote{\XRef{Io.~1,~26.}
\XRef{Act.~1,~5.}
\XRef{11,~16.}
\XRef{19,~4.}}
he shal baptize you in the Holy Ghoſt and fire: \V
whoſe fanne is in his
hand, and he wil purge his floore; and wil gather the wheat into his
barne, but the chaffe he wil burne with vnquencheable fire.

\V Many other things alſo exhorting did he euangelize to the
%%% o-2244
People.

\V
\CNote{\XRef{Mt.~14,~3.}
\XRef{Mar.~6,~17.}}
And Herod the Tetrarch, when he was rebuked of him for Herodias his
\Var{brothers}{brother Philips}
wife, and for al the euils which Herod did: \V
\LNote{He added this aboue al}{The fault of Princes and other great men,
that can not only 
\Fix{}{not}{obvious typo, fixed in other}
abide to heare their faults, but alſo punish by death or impriſonment
ſuch as reprehend them for the ſame (ſpecially if they warne them, as
Prophets & Prieſts doe, from God) is exceeding great.}
He added this alſo aboue al, and shut vp Iohn into priſon. \V
\CNote{\XRef{Mt.~3,~13.}
\XRef{Mr.~1,~9.}
\XRef{Io.~1,~32.}}
And it
came to paſſe when al the People was baptized, \Sc{Iesvs} alſo being
baptized and praying, Heauen was opened: \V And the Holy Ghoſt deſcended
in corporal shape as a doue vpon him: and a voice from Heauen was made:
Thou art my beloued Sonne, in thee I am wel pleaſed. \V And \Sc{Iesvs}
him ſelf was beginning to be about thirtie yeares old: as it was
thought, the ſonne of Ioſeph, who was
\LNote{Of Heli}{Whereas
\MNote{The reconciliation of Matthew and Luke in our Sauiours pedegree.}
in S.~Matthew, Iacob is father to Ioſeph, and here Heli, the caſe was
thus. Mathan (named in S.~Matthew) of his wife called Eſcha begat Iacob:
and after his death, Melchi (named here in S.~Luke) of the ſame woman
begat Heli: ſo that Iacob and Heli were brethren of one mother. This
Heli therfore marying and dying without iſſue: Iacob his brother,
according to the Law maried his wife, and begat Ioſeph, and ſo raiſed vp
ſeed to his brother Heli. Whereby it came to paſſe, that Iacob was the
natural father of Ioſeph which was
\Fix{as (S.~Matthew}{(as S.~Matthew}{likely typo, same in other}
ſaith) begat him: and Heli was his legal father according to the Law, as
S.~Luke ſignifieth.
\Cite{Euſeb. li.~1. Ec. Hiſt. c.~7.}
\Cite{Hiero. in c.~1. Mat.}
\Cite{Aug. li.~2. c.~2. &~3. de conſ. Euang.}}
of Heli, \V who was of Matthat, who was of Leui, who was of Melchi, who
was of Ianne, who was of Ioſeph, \V who was of Matthathias, who was of
Amos, who was of Naum, who was of Heſli, who was of Nagge, \V who was of
Mahath, who was of Matthathias, who was of Semei, who was of Ioſeph, who
was of Iuda, \V who was of Iohanna, who was of Reſa, who was of
Zorobabel, who was of Salathiel, who was of Neri, \V who was of Melchi,
who was of Addi, who was of Coſam,
%%% 2420
who was of Elmadan, who was of Her, \V who was of Ieſus, who was of
Eliézer, who was of Iorim, who was of Matthat, who was of Leui, \V who
was of Simeon, who was of Iudas, who was of Ioſeph, who was of Iona, who
was of Eliacim, \V who was of
\Var{Melcha,}{Melca}
who was of Menna, who was of Matthatha, who was of Nathan, who was of
Dauid, \V
\CNote{\XRef{Mt.~1,~5.}}
who was of Ieffe, who was of Obed, who was of Booz, who was of
Salmon, who was of Naaſſon, \V who was of Aminadab, who was of Aram, who
was of Eſron, who was of Phares, who was of Iudas, \V who was of Iacob,
who was of Iſaac, who was of Abraham, who was of Thare, who was of
Nachor, \V who was of Sarug, who was of Ragau, who was of Phaleg, who
was of Heber, who was of Sale, \V
\SNote{Beza boldly wipeth out of this Ghoſpel, theſe wordes, \Emph{who
was of Cainan}: though al the Greeke copies both of the old Teſtament &
of the new, haue them with ful conſent. Whereby we learne the intolerable
ſaucines of the Caluiniſts, and their contempt of holy ſcripture, that
dare ſo deale with the very Ghoſpel it ſelf.}
who was of Cainan, who was of Arphaxad, who was of Sem, who was of Noë,
who was of Lamech, \V who was of Mathuſale, who was of Henoch, who was
of Iared, who was of Malaleel, who was of Cainan, \V who was of Henos,
who was of Seth, who was of Adam, who was of God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
