%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 2444
%%% o-2273
\startChapter[
  title={Chapter 12}
  ]

\Summary{He prepareth his Diſciples againſt perſecutions to come vpon
  them at their publishing of his Doctrine. 13.~with deuiding the
  brethrens inheritãce he wil not meddle, but exhorteth them againſt
  auarice, 22.~and his Diſciples (by this occaſion) againſt ſolicitude
  ſo much as of neceſſaries, 32.~yea counſeling them to giue al in
  almes, 35.~and to be ready at a knocke: 41.~namely admonishing Peter
  and other Prelats to ſee to their charge: 49.~and al, not to looke
  but for perſecution. 54.~The Iewes he reprehendeth for that they wil
  not ſee this time of grace, 58.~whereas it is ſo horrible to die
  without reconciliation.}

And when great multitudes ſtood about him, ſo that they trode one
another, he began to ſay to his Diſciples: Take good heed of the leauen
of the Phariſees, which is hypocriſie. \V
\CNote{\XRef{Mt.~10,~26.}
\XRef{Mr.~4,~22.}}
For nothing is hid, that ſhal
not be reuealed: nor ſecret, that ſhal not be knowen. \V For the things
that you haue ſaid in darkneſſe, ſhal be ſaid in the light: and that
which you haue ſpoken into the eare in the chambers, ſhal be preached in
the houſe-tops. \V And I ſay to you my freindes, Be not afraid of them
that kil the body, and after this haue no more to doe. \V But I wil ſhew
you whom ye ſhal feare:
\SNote{The feare of Hel alſo is profitable: contrarie to the
proteſtants, teaching ſecuritie of ſaluation & that feare of Hel maketh
mẽ hypocrites.}
feare him who after he hath killed, hath power to caſt into Hel. Yea I
ſay to you, feare him. \V Are not fiue ſparowes ſold for two farthings,
and one of them is not forgotten before God? \V Yea the haires alſo of
your head are
%%% o-2274
al numbred. Feare not therfore: you are more worth then many
ſparowes. \V
\CNote{\XRef{Mt.~10,~32.}
\XRef{Mr.~8,~38.}}
And I ſay to you,
\LNote{Euery one that confeſſeth}{A
\MNote{Open confeſſiõ of our faith.}
Catholike man is bound to confeſſe his faith, being called to account or
examined by Iew, Heathen, or Heretike, cõcerning the ſame. Neither is it
enough to keepe Chriſt in his hart, but he muſt alſo acknowledge him in
his wordes & deedes. And to deny Chriſt, or any Article of the Catholike
faith, for shame or feare of any worldly creature, hath no leſſe
punishment, then to be denied, refuted, and forſaken by Chriſt at the
houre of his death before al his Angels: which is another manner of
preſence and Conſiſtorie, then any Court or Seſſion that men can be
called to for their faith, in this world.}
Euery one that confeſſeth me before men, the Sonne of man alſo wil
confeſſe him before the Angels of God. \V But he that denieth me before
men, ſhal be denied before the Angels of God. \V
\CNote{\XRef{Mt.~12,~33.}
\XRef{Mar.~3,~28.}}
And
\SNote{See
\XRef{Annot. Mt.~12,~32.}}
euery one that ſpeaketh a word againſt the Sonne of man it ſhal be
forgiuen him: but he that ſhal blaſpheme againſt the Holy Ghoſt, to him
it ſhal not be forgiuen. \V
\CNote{\XRef{Mt.~10,~19.}
\XRef{Mr.~13,~11.}}
And when they ſhal bring you into the
Synagogues and to Magiſtrates and Poteſtates,
\LNote{Be not careful}{That
\MNote{The Holy Ghoſt teacheth euery vnlearned Catholike to giue
ſufficiẽt reaſon of his faith.}
the poore vnlearned Catholike should not be diſcouraged, or make his
excuſe that he is a ſimple man, not able to anſwer cunning Heretikes,
nor to giue a reaſon of his beleefe, and therfore muſt ſuffer or ſay any
thing rather then come before them: our Maiſter giueth them comfort
promiſing that the Holy Ghoſt shal euer put into their hartes at the
time of their appearance, that which shal be ſufficient for the
purpoſe: not that euery one which is conuented before the Aduerſaries of
faith, should alwaies be indowed with extraordinary knowledge to diſpute
and confute, as the Apoſtles and others in the primitiue Church were:
but that God wil euer giue to the ſimple that truſteth in him,
ſufficient courage and wordes to confeſſe his beleefe. For ſuch an one
called before the Commiſſioners, ſaith enough and defendeth himſelf
ſufficiently, when he anſwereth that he is a Catholike man, & that he
wil liue and die in that faith which the Catholike Church through out al
Chriſtian countries hath and doth teach, and that this Church can giue
them a reaſon of al the things which they demand of him, &c.}
be not careful in what manner, or what you ſhal anſwer, or what you ſhal
ſay. \V For the Holy Ghoſt ſhal teach you in the very houre what you
muſt ſay.

\V And one of the multitude ſaid to him: Maiſter, ſpeake to my brother
that he deuide the inheritance with me. \V But he ſaid to him: Man,
\LNote{Who hath appointed?}{Chriſt
\MNote{The dealing of Clergie men in worldly affaires.}
refuſed to medle in this temporal matter, partly becauſe the demand
proceeded of couetouſnes & il intention, partly to giue an example to
Clergie men, that they should not be withdrawen by ſecular affaires and
controuerſies from their principal function of praying, preaching, and
ſpiritual regiment: but not wholy to forbid them al actions pertaining
to worldly buſines, ſpecially where and when the honour of God, the
increaſe of religion, the peace of the people, and the ſpiritual benefit
of the parties doe require. In which caſes S.~Auguſtin
\CNote{\Cite{In vita c.~19.}}
(as Paſsidonius
writeth) was occupied often whole daies in ending worldly
controuerſies: and ſo
\CNote{\Cite{Li. de op. Monac. c.~29.}}
he writeth of himſelf alſo, not doubting but to
haue reward therfore in Heauen.}
who hath appointed me iudge or deuider ouer you? \V And he ſaid to them:
See and beware of al auarice: for not in any mans aboundance doth his
life conſiſt, of thoſe things which he poſſeſſeth. \V And
%%% 2445
he ſpake a ſimilitude to them, ſaying: A certaine rich mans field
yealded plentie of fruits. \V And he thought within himſelf, ſaying:
\SNote{Giue it to the poore, that ſhouldſt thou doe, ſaith S.~Baſil.}
What ſhal I doe, becauſe I haue not whither to gather my fruits? \V And
he ſaid: This wil I doe, I wil deſtroy my barnes, and wil make greater,
and thither wil I gather al things that are growen to me, and my
goods. \V And I wil ſay to my ſoule; Soule, thou haſt much goods laid vp
for many yeares, take thy reſt, eate drinke, make good cheere. \V But
God ſaid to him,
\SNote{A goodly warning for al rich men.}
Thou foole, this night they require thy ſoule of thee; and the things
that thou haſt prouided, whoſe shal they be? \V So is he that laieth vp
treaſure to himſelf, and is not
\LNote{Rich to God-ward}{He
\MNote{Meritorious workes.}
is rich towards God, that by his goods beſtowed vpon the poore, hath
ſtore of merits, and many almes-mens prayers procuring mercie for him at
the day of his death and iudgement, which is here therfore called
treaſure laid vp in Heauen, where the barnes be large enough. The
neceſsitie of which almes is by Chriſt himſelf here shewed to be ſo
great, and ſo acceptable to God, that rather then they should lacke the
fruit thereof, they should ſel al they haue and giue to the poore.}
rich to God-ward.

\V And he ſaid to his Diſciples:
\CNote{\XRef{Mt.~6,~25.}}
Therfore I ſay to you,
\SNote{He forbiddeth not cõpetẽt prouidẽce, but too much carefulnes. See
\XRef{Annot. vpõ S.~Matth. c.~6,~25.}}
Be not careful for your life, what you ſhal eate; nor for your body,
what you ſhal doe on. \V The life is more then the meate, and the body
is more then the raiment. \V Conſider the rauens, for they ſow not,
neither doe they reape, which neither haue ſtorehouſe not barne, and God
feedeth them. How much more are you of greater price then they? \V And
which of you by caring can adde to his ſtature one cubite? \V If then
you be not able to doe ſo much as the leaſt thing, for the reſt why are
you careful? \V Conſider the lillies how
%%% o-2275
they grow: they labour not, neither doe they ſpinne. But I ſay to you,
Neither Salomon in al his glorie was araied as one of theſe. \V And if
the graſſe that to day is in the field, and to morow is caſt into the
ouen, God ſo clotheth; how much more you, O ye of litle faith?
\Fix{}{\V}{missing verse number, fixed in other.}
And you, doe not ſeeke what you ſhal eate, or what you ſhal drinke: and
\TNote{\G{μὴ μετεωρίζεσθε:}}
be not lifted vp on high. \V For al theſe things the Nations of the
world doe ſeeke. But your Father knoweth that you haue need of theſe
things. \V But ſeeke firſt the Kingdom of God, and al theſe things ſhal
be giuen you beſides. \V Feare not
\SNote{It was litle at the beginning, & is ſtil in cõpariſõ of al the
reprobate: but in itſelf very great, as in the parable of the great tree
that grew of the litle muſtard-ſeed.
\XRef{Mat.~13.}}
litle flocke, for it hath pleaſed your Father to giue you a Kingdom. \V
Sel the things that you poſſeſſe, and giue almes.
\CNote{\XRef{Mt.~6,~20.}}
Make to you purſes
that weare not, treaſure that waſteth not, in Heauen: whither the theefe
approcheth not, neither doth the mothe corrupt. \V For
\LNote{Where your treaſure is}{If
\MNote{Almes.}
the rich man withdrawen by his worldly treaſure, can not ſet his hart
vpon Heauen, let him ſend his money thither before him, by giuing it in
almes vpon ſuch as wil pray for him, and his hart wil follow his purſe
thither.}
where your treaſure is, there wil your hart be alſo. \V Let your
\SNote{To girde our loines, is to keepe chaſtitie & continencie.
\Cite{Grego. ho.~33.}}
loynes be girded, and candles burning in your handes, \V and you like to
men expecting their Lord, when he ſhal returne from the marriage: that
when he doth come and knocke, forthwith they may open vnto him. \V
Bleſſed are thoſe ſeruants, whom when the Lord commeth, he ſhal find
watching. Amen I ſay to you, that he wil gird himſelf, and make them ſit
downe, and paſſing wil miniſter vnto them. \V And if
he come in the ſecond watch, and if in the third watch he come, and ſo
find, bleſſed are thoſe ſeruants. \V
\CNote{\XRef{Mt.~24,~34.}}
And this know ye, that if the
houſholder did know what houre the theefe would come, he would watch
verily, and would not ſuffer his houſe to be broken vp. \V Be you alſo
ready: for at what hour you thinke not, the Sonne of man wil come.

\V And Peter ſaid to him: Lord, doeſt thou
\Fix{ſpake}{ſpeake}{obvious typo, fixed in other}
this parable to vs, or likewiſe to al? \V And our Lord ſaid: Who
(thinkeſt thou) is a faithful ſteward and wife, whom the Lord appointeth
ouer his
%%% 2446
familie, to giue them in ſeaſon their meaſure of wheate? \V Bleſſed is
that ſeruant, whom when the Lord commeth, he ſhal find ſo doing. \V
Verily I ſay to you, that ouer al things which he poſſeſſeth, he ſhal
appoint him. \V But if that ſeruant ſay in his hart, My Lord is long a
comming; and ſhal begin to ſtrike the ſeruants and handmaides, and eate
and drinke, and be drunke: \V the Lord of that ſeruant ſhal come in a day
that he hopeth not, and at an houre that he knoweth not, and ſhal deuide
him, and
%%% o-2276
ſhal appoint his portion with the infidels. \V And that ſeruant that
knew the wil of his Lord, and prepared not himſelf, & did not according
to his wil, ſhal be beaten with many ſtripes. \V But he that knew not,
and did things worthie of ſtripes, ſhal be beaten with few. And euery
one to whom much was giuen, much ſhal be required of him: and to whom
they committed much, more wil they demand of him. \V I came to caſt fire
on the earth; & what wil I but that it be kindled? \V But I haue to be
baptized with a Baptiſme: and how am I ſtraitned vntil it be
diſpatched. \V
\CNote{\XRef{Mt.~10,~34.}}
Thinke you that I came to giue
\SNote{He meaneth the naughtie peace that is betweene worldlings &
ſinnes, the agreement that is in Infidelity, in Hereſie, or in any other
wickednes. He came to breake this peace. See
\XRef{Annot. Matth. c.~10,~34.}}
peace on the earth? No, I tel you, but ſeparation. \V For there ſhal be from
this time fiue in one houſe deuided: three againſt two, and two, againſt
three. \V There ſhal be deuided, the father againſt the
\Fix{Sonne,}{ſonne,}{obvious typo, fixed in other}
and the ſonne againſt his father, the mother againſt the daughter, and
the daughter againſt the mother, the mother in law againſt her daughter
in law, and the daughter in law againſt her mother in law.

\V
\CNote{\XRef{Mt.~16,~2.}}
And he ſaid alſo to the multitudes, when you ſee a cloud riſing from
the weſt, by and by you ſay, A ſhoure commeth, and ſo it commeth to
paſſe: \V and when the ſouth wind blowing, you ſay, That there wil be
heate: & it commeth to paſſe. \V Hypocrites, the face of the Heauen and
of the earth you haue skil to diſcerne: but this time how doe you not
diſcerne? \V And why of your ſelues alſo iudge you not that which is
iuſt? \V
\CNote{\XRef{Mt.~5,~25.}}
And
\SNote{See
\XRef{Annot. Mt.~5,~25.}}
when thou goeſt with thy aduerſarie to the Prince, in the way endeauour
to be deliuered from him: leſt perhaps he draw thee to the iudge, and
the iudge deliuer thee to the exactour, and the exactour caſt thee into
priſon. \V I ſay to thee, thou ſhalt not goe out thence, vntil thou pay
the very laſt mite.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
