%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 24
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-24


\project douay-rheims


%%% 2477
%%% o-2310
\startChapter[
  title={Chapter 24}
  ]

\Summary{Deuout women not finding his body in the Sepulchre, 4.~Angels
  tel them that he is riſen according to his owne prediction: 9.~yet the
  Apoſtles wil not beleeue it. 12.~But neither Peter findeth his body
  there. 13.~He walketh with two Diſciples, declaring al this vnto them
  out of the Scriptures, and is knowen of them by breaking of
  bread. 36.~The ſame day he appeareth to the Apoſtles being together,
  is felt of them, and eateth with them. Finally teaching them out of
  the Scriptures not only of his Paſsion and Reſurrection, 47.~but alſo of
  his Catholike Church, 49.~he promiſeth the Holy Ghoſt to confirme
  them, 50.~and ſo aſcendeth into Heauen.}

And
\CNote{\XRef{Mat.~28,~1.}
\XRef{Mr.~16,~1.}
\XRef{Io.~20,~1.}}
\MNote{\Sc{Eaſter} day.}
in the
\SNote{That is, firſt after the Sabboth, which is (ſaith
\Cite{S.~Hiero. q.~4. ad Hedib.})
\L{dies Dominica} our Lords day, wherein he aroſe. For the weeke is
deuided into the Sabboth, & the 1. 2. 3. 4. 5. 6. of the
Sabboth. And the Apoſtle
\XRef{(1.~Cor. 16,~2.)}
commanded a collection of money to be made on the firſt of the
Sabboth. Whereby we learne, both the keeping of Sunday, & the Churches
coũt of daies by the 2. 3. 4. of the Sabboth, to be
Apoſtolical. Which S.~Sylueſter afterward named, 2. 3. 4. feriam,
&c.
\Cite{Breuiar. Roman. Decemb.~31.}}
firſt of the Sabboth very early they came to the monument, carying the
ſpices which they had prepared. \V And they found the ſtone roled backe
from the monument. \V And going in, they found not the body of our
Lord \Sc{Iesvs}. \V And it came to paſſe, as they were aſtoniſhed in
their mind at this, behold two men ſtood beſide them in gliſtering
apparel. \V And when they feared and caſt downe their countenance toward
the ground, they ſaid vnto them: Why ſeeke you the liuing with the
dead? \V he is not here, but is riſen. Remember how he ſpake to you, when
he yet was in Galilee, \V ſaying:
\CNote{\XRef{Luc.~9,~22.}}
That the Sonne of man muſt be
deliuered into the handes of ſinners and be crucified, and the third day
riſe againe. \V And they remembred his wordes. \V And going
%%% o-2311
backe from the monument, they told al theſe things to thoſe eleuen, and
to al the reſt. \V And it was Marie Magdalene, and Ioane, and Marie of
Iames, and the reſt that were with them, which ſaid theſe things to the
Apoſtles. \V And theſe wordes ſeemed before them as dotage, and they did
not beleeue them.

\V
\CNote{\XRef{Io.~20,~3.}}
But Peter riſing vp ranne to the monument; and ſtouping downe he ſaw
the linnen clothes lying alone, and went away maruelling with himſelf at
that which was done.

\V
\CNote{\XRef{Mr.~16,~12.}}
And behold, two of them went the ſame day into a towne which was the
ſpace of ſixtie furlongs from Hieruſalem, named Emmaus. \V And they
talked betwixt themſelues of al thoſe things that had chanced. \V And it
came to paſſe, while they talked and reaſoned with
themſelues, \Sc{Iesvs} alſo himſelf approching went with them. \V But
their eyes were held that they might not know him. \V And he ſaid to
them: What are theſe communications that you conferre one with another
walking, and are ſad? \V And one whoſe name was Cleophas, anſwering,
ſaid to him: Art thou only a ſtranger in Hieruſalem, and haſt not knowen
the things that haue been done in it, theſe daies? \V To whom he ſaid:
What things? And they ſaid, concerning \Sc{Iesvs} of Nazareth, who was a
man a Prophet, mightie in worke and word before God and al the
People. \V And how our cheefe Prieſts and Princes deliuered him into
condemnation of death, and crucified him. \V But we hoped that it was he
that ſhould redeeme Iſrael: and now beſides al this, to day is
%%% 2478
the third day ſince theſe things were done. \V But certaine women alſo
of ours, made vs afraid: who before it was light, were at the
monument, \V and not finding his body, came, ſaying that they ſaw a
viſion alſo of Angels, who ſay that he is aliue. \V And certaine men of
ours went to the monument: and they found it ſo as the woman ſaid, but
him they found not. \V And he ſaid to them: O fooliſh, and ſlow of hart
to beleeue in al things which the Prophets haue ſpoken. \V Ought not
Chriſt to haue ſuffred theſe things, and ſo to enter into his glorie? \V And
beginning from Moyſes and al the Prophets, he did interpret to them in
al the ſcriptures the things that were concerning him. \V And they drew
nigh to the towne whither they went: and he made ſemblance to goe
further. \V And they forced
%%% o-2312
him, ſaying: Tarie with vs, becauſe it is toward night, and the day is
now farre ſpent. And he went in with them. \V And it came to paſſe,
whiles he ſate at the table with them, he
\LNote{Tooke bread}{The Fathers in diuers places take this to be meant
of the B.~Sacrament.
\Cite{Author operis imperf. ho.~17.}
\Cite{A.~Auguſtine li.~19. de ſonſenſu. Euang. c.~25.}
&
\Cite{ſer.~140. de temp.}
&
\Cite{ep.~59. ad Paulinum q.~8.}
\Cite{Paulinus himſelf in the next epiſtle before that, among S.~Auguſtines.} 
\Cite{Venerable Bede alſo vpon this place.}
\Cite{Theophylact vpon this place.}
And that it ſhould be meant of the holy Sacrament; the forme of ſolemne
taking the bread into his handes, bleſsing it, breaking it, and reaching
it to his Diſciples (exceeding proper to the conſecration, and common to
none other vulgar benediction, nor any where vſed but in Chriſts
miraculous multiplying the loaues) and the ſingular effect in
notifying Chriſt vnto them, doe proue.
\MNote{The B.~Sacrament in one kind.}
And if it be the Sacrament (as it is moſt probable) then it is an
euident example and warrant of miniſtration in one kind.}
tooke bread, and bleſſed and brake, and did reach to them. \V And their
eyes were opened, and they knew him: and he vaniſhed out of their
ſight. \V And they ſaid one to the other: Was not our hart burning in
vs, whiles he ſpake in the way, and opened vnto vs the ſcriptures? \V
And riſing vp the ſame houre they went backe into Hieruſalem: and they
found the eleuen gathered together, and thoſe that were with them, \V
ſaying: That our Lord is riſen indeed, and hath appeared to Simon. \V
And they told the things that were done in the way: and how they knew
him in the breaking of bread.

\V
\CNote{\XRef{Mr.~16,~14.}
\XRef{Io.~20,~19.}}
And whiles they ſpeake theſe things, \Sc{Iesvs} ſtood in the middes
of them, and he ſaith to them: Peace be to you; it is I, feare not. \V
But they being troubled and frighted, imagined that they ſaw a
Spirit. \V And he ſaid to them: Why are you troubled, and cogitations
ariſe into your harts? \V See my handes, and feet, that it is I my ſelf:
handle, and ſee: for a Spirit hath not fleſh and bones, as you ſee me to
haue. \V And when he had ſaid this, he ſhewed them his handes and
feet. \V But they yet not beleeuing and maruelling for ioy, he ſaid:
Haue you here any thing to be eaten? \V But they offred him a peece of
fiſh broiled, and a honie combe. \V And when he had eaten before them,
taking the remaines he gaue to them.

\V And he ſaid to them: Theſe are the wordes which I ſpake to you, when
I was yet with you, that al things muſt needes be fulfilled, which are
written in the law of Moyſes, and the Prophets, and the Pſalmes, of
me. \V Then he opened their vnderſtanding, that they might vnderſtand
the Scriptures. \V And he ſaid to them: That ſo it is written, and ſo it
behoued Chriſt to ſuffer, and to riſe againe from the dead the third
day: \V
\SNote{As he ſhal be Anathema (ſaith S.~Aug.) which preacheth that
Chriſt neither ſuffered nor roſe againe, becauſe we learne by the
Ghoſpel, \Emph{That it behoued Chriſt to ſuffer and riſe againe the
third day}: ſo he ſhal alſo be Anathema, whoſoeuer preacheth the Church to be
elſwhere thẽ in the cõmuniõ of al Nations: becauſe by the ſelf ſame
Ghoſpel we learne in the wordes next following, \Emph{and pennance to be
preached in his name & remiſsion of ſinnes throughout al Nations}.
\Cite{Auguſt. ep.~45.}}
and
\LNote{Pennance to be preached}{He
\MNote{The Catholike or vniuerſal Church.}
ſhewed vnto them out of the Scriptures, not only the things that were
now accompliſhed in himſelf, but alſo that were yet to come about his
Church: as, where it should begin, to wit, at Hieruſalem & how farre it
ſhould goe, to wit, to al Nations: that he might not ſuffer vs
\CNote{\Cite{De. vnit. Ec. c.~10.}}
(ſaith
S.~Auguſtine) to erre neither in the bridegrome nor in the bride. For
this maketh manifeſtly againſt al Heretikes and Schiſmatikes, that ſet
vp new Churches in particular countries, drawing the people from the
foreſaid only true Church which frõ Hieruſalem ſo groweth
\CNote{\XRef{Mt.~24,~14.}}
ouer al Nations til the end of the world come.}
pennance to be preached in his name and remiſſion of ſinnes vnto al
Nations, beginning from Hieruſalem. \V And you are witneſſes of theſe
things. \V
\CNote{\XRef{Act.~1,~8.}
\XRef{2,~1.}
\XRef{10,~41.}}
And I ſend the promiſe of my Father vpon you: but you, tarie
in the citie, til you be indowed with power from high.

\V 
\MNote{\Sc{Ascension} day.}
And he brought them forth abrode into Bethania: and lifting vp his
handes
\LNote{Bleſſed them}{Chriſt
\MNote{Chriſt bleſſed diuers waies.}
our high Prieſt,
%%% !!! Do all these go here? rearrange them to match biblical order?
\CNote{\XRef{Heb.~7,~6.}
\XRef{Io.~20,~21.~26.}
\XRef{Mr.~10,~16.}
\XRef{Gen.~48,~14.}}
prefigured ſpecially therein by Melchiſedech, often
gaue his bleſſing to his: ſomtimes by wordes, as, \Emph{Peace be to
you}: ſomtimes by impoſing his handes: and now hereby lifting vp his
hands ouer his Diſciples as it were for his farewel. In what forme, the
Scripture doth not expreſſe, but very like it is that in forme of the
croſſe, as Iacob the Patriarch bleſſed his nephewes for ſignification of
Chriſts benediction: for now the croſſe began to be glorious among the
faithful,
\MNote{Bleſsing with the ſigne of the croſſe.}
and the Apoſtles (as it is moſt certaine
\CNote{\Cite{Tertu. de coron. milit. nu.~3.}
\Cite{Baſil de Sp. ſancti. 1,~27.}}
by the Fathers which
cal it an ancient tradition) vſed that ſigne for an external note of
benediction. Yea S.~Auguſtine ſaith (in
\Cite{Ps.~30. con.~3.})
that Chriſt himſelf not without cauſe would haue his ſigne to be fixed
in our foreheads as in the ſeat of ſhamefaſtnes, that a Chriſtian man
ſhould not be aſhamed of the reproch of Chriſt. And what forme can a
Chriſtian man vſe rather to bleſſe himſelf or others, then that which
was dedicated in Chriſts death, & is a conuenient memorial of the ſame?
Howſoeuer it be, that the Biſhops and Prieſts of Gods Church bleſſe with
an external ſigne, no man can reprehend, being warranted by Chriſts owne
example and action.}
he bleſſed them. \V
\CNote{\XRef{Mr.~16,~19.}
\XRef{Act.~1,~9.}}
And it came to paſſe whiles he bleſſed
%%% 2479
them, he departed from them, and
%%% o-2313
was caried into Heauen. \V And they adoring went backe into Hieruſalem
with great ioy: \V and they were alwaies in the
\Fix{temple}{Temple}{likely typo, same in other}
praiſing and bleſſing God. Amen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
