%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 2466
%%% o-2298
\startChapter[
  title={Chapter 21}
  ]

\Summary{He commendeth the poore widow for her two mites, aboue
  al. 5.~Hauing ſaid that the Temple shal be quite deſtroied, 7.~he
  foretelleth firſt many things that shal goe before, 20.~then a ſigne
  alſo when it is neere, after which shal come the deſtruction it ſelf
  in moſt horrible manner, without hope of reſtitution, vntil al Nations
  of the Gentils be gathered into his Church in the very end of the
  world. 25.~And then what ſignes shal come of the laſt day, terrible to
  the world, 28.~but comfortable to vs of his Church, 34.~ſo that we be
  alwaies watchful.}

And
\CNote{\XRef{Mr.~12,~14.}}
beholding, he ſaw them that did caſt their guifts into the
treaſurie, rich perſons. \V And he ſaw alſo a certaine poore widow
caſting two braſſe mites. \V And he ſaid: Verily I ſay to you, that this
poore widow hath caſt more then al. \V For al theſe of their aboundance
haue caſt into the guifts of God: but ſhe
\LNote{Of her penurie}{To offer or giue almes of our ſuperfluities, is
not ſo acceptable nor meritorious, as to beſtow ſome of that which is of
our neceſſarie prouiſion and which we may hardly ſpare from our ſelues:
for, that proceedeth of greater zeale, wil, and intention, which be more
reſpected of God then the ſubſtance of the guift.}
of her penurie, hath caſt in al her liuing that ſhe had.

\V And certaine ſaying of the Temple that it was adorned
%%% o-2299
with goodly ſtones and donaries, he ſaid: \V Theſe things which you ſee,
\CNote{\XRef{Mt.~24,~1.}
\XRef{Mar.~13,~1.}}
the daies wil come wherein
\SNote{This was fulfilled 40.~yeares after the death of Chriſt, the
19.~of Auguſt, being the very moneth and day wherein the Babyloniãs
burnt it: from the firſt building thereof by Salomon 1130.~yeares: from
the re-edifying thereof vnder Cyrus 639.~yeares.
\Cite{Ioſep. de bel. Iud. li.~7. cap.~10.}}
there ſhal not be left a ſtone vpon a ſtone that ſhal not be
deſtroied. \V And they asked him, ſaying: Maiſter, when ſhal theſe
things be: and what ſhal be the ſigne when they ſhal begin to come to
paſſe? \V Who ſaid: See you be not ſeduced. For
\SNote{Many falſe prophets & Heretikes. See
\XRef{An. Mt.~24. Mar.~13.}}
many wil come in my name, ſaying that I am he: and the time is at hand,
goe not therfore after them. \V And when you ſhal heare of warres and
ſeditions, be not terrified: theſe things muſt firſt come to paſſe, but
the end is not yet by and by. \V Then he ſaid to them: Nation ſhal riſe
againſt Nation, and Kingdom againſt Kingdom. \V And there ſhal be greate
earth-quakes in places, and peſtilences and famines, and terrours from
Heauen, and there ſhal be great ſignes. \V But before al theſe things
they wil
\SNote{Great perſecution of Catholike men.}
lay their hands vpon you: and perſecute you deliuering you into
Synagogues and priſons, drawing you to Kings and Preſidents for my
name. \V And it ſhal happen vnto you for teſtimonie. \V Lay vp this
therfore in your hartes, not to premeditate how you ſhal anſwer. \V For
I wil giue you mouth and wiſdom, which al your aduerſaries ſhal not be
able to reſiſt and gaineſay. \V And you ſhal be deliuered vp of your
parents and brethren, and kinſemen and freinds: and they wil put to
death of you. \V And you ſhal be odious to al men for my name: \V and a
haire of your head ſhal not periſh. \V In your patience you ſhal
poſſeſſe your ſoules.

\V And when you ſhal ſee Hieruſalem compaſſed about with an armie, then
know that the deſolation thereof is at hand. \V then they that are in
Iewrie, let them flee to the mountaines: and they in the middes thereof,
let them depart: and they in the countries, let them not enter into
it. \V For theſe are the daies of vengeance, that al things may be
fulfilled that are written. \V But woe to them that are with child &
that
%%% 2467
giue ſucke in thoſe daies. For there ſhal be great affliction vpon the
land, and wrath on this people. \V And they ſhal fal by the edge of the
ſword; and ſhal be led captiue into al Nations: and Hieruſalem ſhal be
troden of the Gentils, til the times of Nations be fulfilled.

\V
\CNote{\XRef{Eze.~32,~7.}
\XRef{Ioel.~3,~15.}}
And there ſhal be ſignes in the ſunne and the moone and the ſtarres:
and vpon earth diſtreſſe of Nations, for the confuſion of the ſound of
ſea and waues, \V men withering
%%% o-2300
for feare and expectation, what ſhal come vpon the whole world. For the
powers of Heauen ſhal be moued: \V and then they ſhal ſee the Sonne of
man comming in a cloude with great power and maieſtie. \V But when theſe
things begin to come to paſſe, looke vp and lift vp your heades: becauſe
your redemption is at hand. \V And he ſpake to them a ſimilitude. See
the figtree and al trees: \V when they now bud forth fruit out of
them ſelues, you know that ſummer is nigh. \V So you alſo when you ſhal
ſee theſe things come to paſſe, know that the Kingdom of God is nigh. \V
Amen I ſay to you, that this Generation ſhal not paſſe, til al be done. \V
Heauen and earth ſhal paſſe; but my wordes ſhal not paſſe.

\V And looke wel to your ſelues, leſt perhaps your hartes be ouercharged
with ſurfetting and drunkenneſſe and cares of this life: and that day
come vpon you ſodenly. \V For as a ſnare ſhal it come vpon al that ſit
vpon the face of al the earth. \V Watch therfore, praying at al times,
that you may be
\TNote{\G{ἵνα καταξιωθῆτε} See
\XRef{Annot. c.~20,~35.}}
accounted worthie to eſcape al theſe things that are to come, and to
ſtand before the Sonne of man.

\V And the daies he was teaching in the Temple: but the nightes going
forth, he abode in the
\SNote{Solitarineſſe or eremitage (as S.~Gregorie Nazian.\ ſaith) is a
goodly thing: this doth the mount Garmel of Elias teach, Iohns deſert, &
that mount vnto which \Sc{Iesvs} often retired, & was quietly alone with
him ſelf.
\Cite{Ser.~26. de amore pauperum.}}
mount that is called Oliuet. \V And al the people in the morning went
vnto him in the Temple to heare him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
