%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 2457
%%% o-2287
\startChapter[
  title={Chapter 17}
  ]

\Summary{So damnable it is to be authour of a Schiſme, 3.~that we muſt
  rather forgiue be it neuer ſo often. 5.~We muſt be feruant in faith,
  and humble withal, knowing that we are bound to God, and not he to
  vs. 11.~The nine Iewes are vngrateful after that he hath cured their
  leproſie: but the one Samaritane (the one Catholike Church of the
  Gentils) far otherwiſe. 20.~The Phariſees asking, when commeth this
  Kingdom of God (of whoſe approching they had now heard ſo much) he
  teacheth that God muſt reigne within vs: 22.~and warneth vs after his
  Paſsion neuer to goe out of his Catholike Church for any new ſecret
  comming of Chriſt that Heretikes shal pretend, but only to expect his
  ſecond comming in glorie, 26.~preparing our ſelues vnto it, becauſe it
  shal come vpon many vnprouided, 31.~ſpecially through the perſecution
  of Antichriſt a litle before it.}

%%% o-2288
And
\CNote{\XRef{Mt.~18,~7.}
\XRef{Mar.~9,~42.}}
he ſaid to his Diſciples: It is
\SNote{Not of mere neceſsitie, for then it were no fault: but
preſuppoſing the great wickednes of men, it is impoſsible but there ſhal
be ſcandals, and therfore it followeth, \Emph{woe to him by whom they
come.}}
impoſſible that ſcandal ſhould not come: but woe to him by whom they
come. \V It is more profitable for him, if a mil-ſtone be put about his
necke, and he be caſt into the ſea, then that he ſcandalize one of theſe
litle ones. \V Looke wel to your ſelues.
\CNote{\XRef{Mt.~18,~21.}}
If thy brother ſinne againſt
thee, rebuke him: and if he doe pennance, forgiue him. \V And if he
ſinne againſt thee ſeuen times in a day, and ſeuen times in a day be
conuerted vnto thee, ſaying, It repenteth me, forgiue him.

\V And the Apoſtles ſaid to our Lord: Increaſe faith in vs. \V And our
Lord ſaid:
\CNote{\XRef{Mt.~18,~21.}}
If you had faith like to a muſtard-ſeed, you might ſay to
this mulberie tree, be thou rooted vp, and be tranſplanted into the ſea:
and it would obey you. \V And which of you hauing a ſeruant plowing or
keeping cattle, that wil ſay to him returning out of the field: Paſſe
quickly, ſit downe: \V and ſaith not to him: Make ready ſupper, and gird
thy ſelf, and ſerue me whiles I eate and drinke, and afterward thou
ſhalt eate and drinke? \V Doth he giue that ſeruant thankes, for doing
the things which he commanded him? \V I trow not: So you alſo, when you
shal haue done al things that are commanded you, ſay: We are
\LNote{Vnprofitable ſeruants}{If
\MNote{How we are vnprofitable & profitable ſeruants.}
our Sauiour had ſaid that the keeping of Gods commandements had bene
vnprofitable & not auailable to our ſelues, then might the Proteſtant
haue truly argued therby that our workes deſerne not Heauen or any
reward at Gods hand: but ſo he ſaid not, but that our ſeruice is to God
vnprofitable, who calleth for it as duty, and not as a thing needful or
profitable to himſelf. And though here our Maiſter teach vs ſo humbly to
conceiue of our owne doings toward him, yet him ſelf elswhere calleth
not his ſeruants vnprofitable when they haue done their labour, but
ſpeaketh thus:
\CNote{\XRef{Mt.~25,~21.}}
\Emph{Good & faithful ſeruants, becauſe thou waſt
faithful in a litle, I wil place thee ouer much: enter into the ioy of
thy Lord.} Yea of ſuch as ſerue him in the grace of the new Teſtament,
he affirmeth that he wil not now
\CNote{\XRef{Io.~15,~15.}}
\Emph{name them ſeruants but freinds},
yea & take them for his owne children, & as his freinds and ſonnes he
counteth of vs and our workes towards
\Fix{heauen,}{Heauen,}{likely typo, same in other}
though we in humilitie and truth muſt confeſſe alwaies that we be to him
vnprofitable ſeruants. Yea and S.~Paul ſaith plainly, that by cleanſing
our ſelues from ſinful workes we ſhal be profitable veſſels to our
Lord.
\XRef{2.~Tim. 2,~21.}}
vnprofitable ſeruants; we haue done that which we ought to doe.

\V
\MNote{The fourth part of this Ghoſpel. The comming of Chriſt into
Iewrie, towardes his Paſsiõ.}
And it came to paſſe, as he went vnto Hieruſalem, he paſſed through the
middes of Samaria and Galilee. \V And when he entred into a certaine
towne, there met him ten men that were lepers, who ſtood a farre off; \V
and they lifted vp their voice, ſaying: \Sc{Iesvs} Maiſter, haue mercie
on vs. \V Whom as he ſaw, he ſaid: Goe,
\CNote{\XRef{Leu.~14,~2.}}
shew your ſelues
\LNote{To the Prieſts}{This
\MNote{Confeſſion to the Prieſt.}
leproſie ſignifieth ſinne, which though God may & can heale without any
mans meanes, yet he doth it not ordinarily but by the Prieſts miniſterie:
therfore let no man deſpiſe Gods ordinance, nor ſay that it is enough to
confeſſe to God though he neuer come at the Prieſt.
\Cite{\Emph{Li. de viſit. infirm. apud Auguſt.}}}
to the Prieſts. And it came to paſſe,
\LNote{As they went}{A man may ſometimes be ſo contrite and penitent,
that his ſinne
is forgiuen before he come to the Prieſt, but then alſo he muſt
notwithſtanding goe to the Prieſt, as theſe lepers did: ſpecially
whereas we are neuer ſure how contrite we are, and becauſe there is no
true contrition, but with deſire alſo of the Sacrament in time and
place.}
as they went, they were made cleane. \V And one of them as he ſaw that
he was made cleane, went backe with a loud voice magnifying God, \V and
he fel on his face before his feet, giuing thankes: and this was a
Samaritane. \V And \Sc{Iesvs} anſwering ſaid: Were not ten made cleane?
and where are the nine? \V There was not found that returned, and gaue
glorie to God, but
%%% 2458
this ſtranger. \V And he ſaid to him: Ariſe, goe thy waies; becauſe
\SNote{And yet we ſee here it was not only faith, but alſo his
thankfulnes & returning to giue glorie to God.}
thy faith hath made thee ſafe.

\V And being asked of the Phariſees: when commeth the Kingdom of God? he
anſwered them and ſaid: The
\Fix{kindom}{Kingdom}{obvious typo, fixed in other}
%%% o-2289
of God commeth not with obſeruation: \V neither ſhal they ſay: Loe here,
or loe there. For loe
\SNote{Whiles they aſke and looke for a temporal Kingdom in pompe &
glorie, loe their Kĩg & Meſsias was now amõg thẽ: whoſe ſpiritual Kingdõ
is within al the faithful that haue dominiõ ouer ſinne.}
the Kingdom of God is within you.

\V And he ſaid to his Diſciples: The daies wil come when you ſhal deſire
to ſee one day of the Sonne of man; and you ſhal not ſee. \V
\CNote{\XRef{Mt.~24,~23.}
\XRef{Mr.~13,~21.}}
And they wil ſay to you: Loe here and loe there.
\SNote{No man muſt rũne out of the Church after Schiſmatikes to heare
them preach Chriſt in corners, Chriſts doctrine being open in al the
world. See
\XRef{annot. Mt.~24,~23.}}
Goe not, neither doe ye follow after. \V For euen as the lightning that
lightneth from vnder Heauen, vnto thoſe partes that are vnder Heauen,
ſhineth: ſo ſhal the Sonne of man be in his day. \V But firſt he muſt
ſuffer many things and be reiected of this Generation. \V And as it came
to paſſe in the
\CNote{\XRef{Gen.~7,~5.}}
daies of Noe, ſo ſhal it be alſo in the dayes of the
Sonne of man. \V They did eate and drinke, they did marie wiues and were
giuen to mariage euen vntil the day that Noe entred into the arke: and
the floud came, and deſtroyed them al. \V Likewiſe as it came to paſſe
in the
\CNote{\XRef{Gen.~19,~24.}}
daies of Lot: They did eate and drinke, bought and ſould,
planted, and builded: \V and in the day that Lot went out frõ Sodome, it
rained fire &
\Fix{brimſtome}{brimſtone}{obvious typo, fixed in other}
frõ Heauen, & deſtroyed them al: \V according to theſe things it ſhal be
in the day that the Sõne of man ſhal be reuealed. \V In that houre he
that ſhal be in the houſe-top, and his veſſel in the houſe, let him not
goe downe to take them vp: & he that is in the field, in like manner let
him not returne backe. \V Be mindful of
\CNote{\XRef{Gen.~19,~26.}}
Lots wife. \V Whoſoeuer ſeeketh
to ſaue his life, ſhal loſe it: and whoſoeuer doth loſe the ſame, ſhal
quicken it. \V I ſay to you, in that night there ſhal be two in one bed:
the one ſhal be taken, and the other ſhal be left: \V two women ſhal be
grinding together: the one ſhal be taken, and the other ſhal be left:
two in the field: the one ſhal be taken, and the other shal be left. \V
They anſwering ſay to him: Where Lord? \V Who ſaid to them: Whereſoeuer
the body shal be, thither wil the eagles alſo be gathered together.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
