%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-18


\project douay-rheims


%%% 2459
%%% o-2290
\startChapter[
  title={Chapter 18}
  ]

\Summary{The Church is taught to commit the reuenge of her perſecutions
  to God, and 
  to pray inceſſantly, for he no doubt (though in the perſecution of
  Antichriſt few wil ſo thinke) wil at length come. 9.~We muſt alſo pray
  with humility, becauſe we know not with the Phariſee if we be iuſt,
  but we know with the Publican that we be ſinners. 15.~He wil haue
  children to be brought to him, and al to be as children. 18.~What is
  to be done to get life euerlaſting. 22.~What alſo, to get perfection;
  28.~and what reward they shal haue that leaue al, yea or any part, for
  his ſake. 31.~He foretelleth of his Paſsion moſt particularly, 33.~and
  entring into Iericho, cureth one blind man.}

And he ſpake alſo a parable to them that it behoueth
\SNote{We ſhould pray alwaies by faith, hope & charitie and by working
the things that be acceptable to God: though ſpecial times of vocal
praiers in the Canonical houres be aſsigned for the ſturring of vs vp to
God through external ſignes of deuotion.}
alwaies to pray, & not to be weary, \V ſaying: There was a certaine
iudge in a certaine citie, which feared not God, and of man made no
account. \V And there was a certaine widow in that citie, and ſhe came
to him, ſaying: Reuenge me of mine aduerſarie. \V And he would not of a
long time. But afterward he ſaid within himſelf: Although I feare not
God, nor make account of man, \V yet becauſe this widow is importune
vpon me, I wil reuenge her, leſt at the laſt she come and defame me. \V
And our Lord ſaid: Heare what the iudge of iniquitie ſayeth. \V And wil
not God reuenge his elect that crie to him day and night: and wil he
haue patiẽce in them? \V I ſay to you that he wil quickly reuenge
them. But yet the Sonne of man comming,
\LNote{Shal he find faith}{The
\MNote{The Church erreth not in faith.}
Luciferians and Donatiſts vſed this place to excuſe their fal from the
Church, as our Aduerſaries now doe, ſaying that it was decaied in faith,
when they forſooke it. To whom we anſwer as
\CNote{\Cite{Hier. cont. Lucif. c.~6.}}
S.~Hierom and
\CNote{\Cite{Aug.~de vnit. Ec. c.~15.}
&
\Cite{de verb. Do. Ser.~36.}}
S.~Auguſtin
anſwered them, that Chriſt ſaith not that there should be no faith left
in earth: but by this manner of ſpeach inſinuateth, that at the later
day in the great perſecution of Antichriſt faith ſhould be more rare,
and the faithful among ſo many wicked not ſo notorious: ſpecially that
perfect faith containing deuotion, truſt, and affection toward God,
which our Maiſter ſo praiſed in certaine vpon whom he wrought miracles,
and by force whereof mountaines might be moued, which is rare euen when
the Church flourisheth moſt.}
shal he find trow you, faith in the earth?

\V And he ſaid alſo to certaine that truſted in them ſelues as iuſt, and
deſpiſed others, this parable: \V Two men went vp into the Temple to
pray: the one a Phariſee, and the other a Publican. \V The Phariſee
ſtanding, praied thus with
%%% o-2291
him ſelf: God, I giue thee thankes that I am not as the reſt of men,
extorcioners, vniuſt, aduouterers, as alſo this Publican. \V
\SNote{To take pride of faſting, tithing, or any good worke, is naught,
though the workes thẽ ſelues be very good.}
I faſt twiſe in a weeke: I giue tithes of al that I poſſeſſe. \V And the
Publican ſtanding a farre off would not ſo much as lift vp his eyes
toward Heauen; but he
\SNote{So doe the Prieſts & people at the holy Altar knocke their
breaſts, & ſay with the hũble Publican, \L{Deus propitius}.
\Cite{Aug. Pſ.~31. conc.~3.}}
knocked his breaſt, ſaying: God be merciful to me a ſinner. \V I ſay to
you, this man went downe into his houſe iuſtified more then he: becauſe
euery one that exalteth himſelf, shal be humbled; and he that humbleth
himſelf, shal be exalted.

\V
\CNote{\XRef{Mt.~19,~13.}
\XRef{Mr.~10,~13.}}
And they brought vnto him infants alſo, that he might touch
them. Which thing when the Diſciples ſaw, they rebuked them. \V
But \Sc{Iesvs} calling them together, ſaid: Suffer children to come vnto
me, and forbid them not, for the Kingdom of Heauen is for ſuch.
%%% 2460
\V Amen I ſay to you: Whoſoeuer receiueth not the Kingdom of God
\SNote{In matters of faith & religiõ we muſt be as humble & obedient to
the Catholike Church as yõg children to their parents.}
as a child, shal not enter into it.

\V
\CNote{\XRef{Mt.~19,~16.}
\XRef{Mr.~10,~17.}}
And a certaine Prince asked him, ſaying: Good Maiſter, by doing what,
shal I poſſeſſe euerlaſting life? \V And \Sc{Iesvs} ſaid to him: Why
doeſt thou cal me good? None is good but only God. \V Thou knoweſt
\SNote{Not faith only but alſo keeping the cõmandements purchaſe life
euerlaſting. See
\XRef{annot. Mat.~19,~36.}}
the commandements:
\CNote{\XRef{Exo.~20,~13.}}
\Emph{Thou shalt not kil, Thou shalt not commit
aduoutrie, Thou shalt not ſteale, Thou shalt not beare falſe witnes,
Honour thy father & mother.} \V Who ſaid: Al theſe things haue I kept
from my youth. \V Which \Sc{Iesvs} hearing, ſaid to him: Yet one thing
thou lackeſt.
\SNote{This is not a cõmandement or precept, but counſel: which the
religious doe follow. See
\XRef{Annot. Mat.~19.}}
Sel al that euer thou haſt, & giue to the poore, and thou shalt haue
treaſure in Heauen: and come, follow me. \V He hearing theſe things, was
ſtroken ſad: becauſe he was very rich. \V And \Sc{Iesvs} ſeeing him
ſtroken ſad, ſaid: How hardly shal they that haue money enter into the
Kingdom of God? \V For it is eaſier for a camel to paſſe through the eye
of a nedle, then for a rich man to enter into the Kingdom of God. \V And
they that heard, ſaid: And who can be ſaued? \V He ſaid to them: The
things that are impoſſible with men, are poſſible with God. \V And Peter
ſaid: Loe, we haue left al things, and haue followed thee. \V Who ſaid
to them: Amen I ſay to you, there is no man that hath left houſe, or
parents, or brethren, or
\SNote{The Apoſtles among other things left their wiues alſo, as
S.~Hierõ noteth out of this place.
\Cite{Li.~1. adu. Iouin.}}
wife, or children for the Kingdom of God, \V and shal not receiue much
more in this time, and in the world to come
\SNote{Life euerlaſting the reward for leauing or loſing willingly our
goods for Chriſts ſake.}
life euerlaſting.

\V
\CNote{\XRef{Mt.~20,~17.}
\XRef{Mr.~10,~32.}}
And \Sc{Iesvs} tooke the Twelue, and ſaid to them: Behold we goe vp
to Hieruſalem, and al things shal be
%%% o-2292
conſummate which were written by the Prophets of the Sonne of man. \V
For he shal be deliuered to the Gentils, and shal be mocked, and
ſcourged, and ſpit vpon: \V and after they haue ſcourged him, they wil
kil him, and the third day he shal riſe againe. \V And they vnderſtood
none of theſe things, and this word was hid from them, and they
vnſerſtood not the things that were ſaid.

\V And it came to paſſe, when he drew nigh to Iericho, a certaine blind
man ſate by the way, begging. \V And when he heard the multitude paſſing
by, he asked what this should be. \V And they told him that \Sc{Iesvs}
of Nazareth paſſed by. \V And he cried ſaying: \Sc{Iesvs} ſonne of
Dauid, haue mercie vpon me. \V And they that went before, rebuked him,
that he should hold his peace. But he cried much more, Sonne of Dauid
haue mercie vpon me. \V And \Sc{Iesvs} ſtanding commanded him to be
brought vnto him. And when he was come neere, he asked him, \V ſaying:
What wilt thou that I doe to thee? but he ſaid: Lord, that I may ſee. \V
And \Sc{Iesvs} ſaid to him: Doe thou ſee; thy faith hath made thee
whole. \V And forthwith he ſaw, and followed him, magnifying God. And al
the people as they ſaw it, gaue praiſe to God.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
