%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Luke
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-23


\project douay-rheims


%%% 2474
%%% o-2307
\startChapter[
  title={Chapter 23}
  ]

\Summary{The Iewes accuſe him to Pilate the Gentil: 4.~who ſeeking
  earneſtly to deliuer him, ſpecially after that Herod ſent him backe,
  17.~they not only preferre the murderer Barabbas, but alſo
  crie, \Sc{Crvcifige}: 26.~In the way to Caluarie he foretelleth the
  women that lamented vpon him, the horrible deſtruction of their
  Hieruſalem. 32.~Vpon the croſſe he is between two theeues, 35.~ſcorned
  of the Iewes, 36.~of the ſouldiares, 39.~and of one of the theeues,
  40.~but euen there confeſſed of the other theefe, 44.~and after his
  death (becauſe of the great miracles concurring) alſo of the
  Centurion, 48.~yea and of the whole multitude. 50.~And finally he is
  buried honourably.}

And
\CNote{\XRef{Mt.~27,~1.~11.}
\XRef{Mr.~15,~1.}
\XRef{Io.~18,~17.}}
\MNote{\Sc{Good} friday.}
al the multitude of them riſing vp, led him to Pilate. \V And they began
to accuſe him, ſaying: We haue found this man ſubuerting our Nation, &
prohibiting to giue tributes to Cæſar, and ſaying that he is Chriſt the
King. \V And Pilate asked him, ſaying: Art thou the King of the Iewes?
But he anſwering ſaid: Thou ſayeſt. \V And Pilate ſaid to the cheefe
Prieſts and multitudes: I find no cauſe in this man. \V But they were
more earneſt, ſaying: He ſtirreth the people teaching throughout al Iewrie,
beginning from Galilee euen hither. \V But Pilate hearing Galilee, asked
if the man were of Galilee. \V And when he vnderſtood that he was of
Herods iuriſdiction, he ſent him backe to Herod, who was alſo himſelf at
Hieruſalem in thoſe daies.

\V And Herod ſeeing \Sc{Iesvs}, was very glad, for he was deſirous of a
long time to ſee him, for becauſe he heard many things of him: and he
hoped to ſee ſome ſigne wrought by him. \V And he asked him in many
wordes. But he anſwered him nothing. \V And there ſtood the cheefe
Prieſts and the Scribes conſtantly accuſing him. \V And Herod with his
armie ſet him at naught: and he mocked him, putting on him a white
garment, and ſent him back to Pilate. \V And Herod and Pilate were made
freindes that day. For before they were enemies one to another.

\V And Pilate calling together the cheefe Prieſts and Magiſtrates, and
the people, \V ſaid to them: You haue preſented vnto me this man, as
auerting the People, and behold I examining him before you, haue found
no cauſe in this man of thoſe things, wherein you accuſe him. \V No, nor
Herod neither. For I ſent you to him, and behold, nothing worthie of
death is done to him. \V I wil chaſten him therfore and dimiſſe him.

%%% o-2308
\V And he of neceſſitie had to releaſe vnto them vpon the feaſt day,
one. \V But the whole multitude together cried out, ſaying: Diſpatch
him, and releaſe vs Barabbas. \V Who was for a certaine ſedition made in
the citie and murder, caſt into priſon. \V And
\Fix{Pilat}{Pilate}{obvious typo, fixed in other}
againe ſpake to them, deſirous to releaſe \Sc{Iesvs}. \V But they cried
againe, ſaying: Crucifie, crucifie him. \V And he the third time ſaid
to them: Why, what euil hath this man done? I find no cauſe of death in
him. I wil correct him therfore & let him goe. \V But they were inſtant
with loud voices
%%% 2475
requiring that he might be crucified. And their voices preuailed. \V And
Pilate adiudged their petition to be done. \V And he releaſed vnto them
him that for murder and ſedition had been caſt into priſon, whom they
demanded: but \Sc{Iesvs} he deliuered to their pleaſure.

\V And when they led him, they tooke one Simon of Cyrene cõming from the
countrie: and they laid the Croſſe vpon him to carie
after \Sc{Iesvs}. \V And there followed him a great multitude of people,
and of women which bewailed and lamented him. \V But \Sc{Iesvs} turning
to them, ſaid: Daughters of Hieruſalem, weepe not vpon me, but weepe
vpon your ſelues, and vpon your children. \V For behold the daies ſhal
come, wherein they wil ſay: Bleſſed are the barren, and the wombes that
haue not borne, & the pappes that haue not giuen ſuck. \V
\CNote{\XRef{Oſ.~10,~8.}}
\Emph{Then
shal they begin to ſay to the mountaines, Fal vpon vs: and to the
hilles, Couer vs.} \V For if in the greene wood they doe theſe things,
in the drie what ſhal be done?

\V And there were led alſo other two malefactours with him, to be
executed. \V And after they came to the place which is called Caluarie,
there they crucified him: and the theeues, one on the right hand and the
other on the left. \V And \Sc{Iesvs} ſaid: Father,
\LNote{Forgiue them}{A perfect example of charitie in our Sauiour
praying for his crucifiers, which the firſt martyr S.~Steuen did follow,
\XRef{Act.~7}:
and the prayers of both were heard: Chriſts praier taking effect in the
Centurion and others, Steuens in Paul.}
forgiue them, for they know not what they doe. But they deuiding his
garments, did caſt lots.

\V And the people ſtood
\Var{beholding,}{expecting,}
and the Princes with them derided him, ſaying: Others he hath ſaued, let
him ſaue himſelf, if this be Chriſt, the elect of God. \V And the
ſouldiars alſo mocked him cõming to him, and offering him vinegre, \V
ſaying: If thou be the King of the Iewes, ſaue thy ſelf. \V And there
was alſo a ſuperſcription written ouer him in Greeke, and Latine, and
Hebrew letters: \Sc{This
%%% o-2309
is the king of the iewes}. \V And one of thoſe
theeues that were hanged, blaſphemed him, ſaying: If thou be Chriſt,
ſaue thy ſelf, and vs. \V But the other anſwering, rebuked him, ſaying:
Neither doeſt thou feare God, where as thou art in the ſame
damnation? \V And we indeed iuſtly, for we receiue worthie of our
doings: but this man hath done no euil. \V And he ſaid to \Sc{Iesvs}:
Lord, remember me when thou ſhalt come into thy Kingdom. \V
And \Sc{Iesvs} ſaid to him: Amen I ſay to thee; this day thou ſhalt be
with me
\LNote{In Paradiſe}{Thou
\MNote{The good theefe.}
maiſt not hereupon differre thy conuerſion or ammendement, preſuming of
grace at the laſt houre of thy life, nor looke to haue ſaluation by
faith and confeſſion of Chriſt without good workes, not to goe ſtraight
to Heauen without ſatisfaction pennance, or puniſhment for thy former
ſinnes & life il ſpent; nor chalenge ſecuritie and certaine knowledge of
thy ſaluation. For this good theeues caſe is not common but a rare
exãple of mercie & prerogatiue. But for the firſt point, learne only not
to deſpaire, though thou haſt been il to the laſt moment of thy
life. For the ſecond, that faith, hope, and charitie, repentance, and
good wil be ſufficient, and good workes not required, where for want of
time and opportunitie they cannot be had.
\MNote{Pardon of due pennance and ſatisfaction.}
For the third, that Chriſt gaue to this happy man for his zelous
confeſsion of him & reprehending his fellow not only remiſsion of his
ſinnes, but alſo by extraordinarie grace, a pardon of al pennance and
ſatisfaction due either in this life or the next for the ſame: euen as
the holy Church by his example and commiſsion giueth pardons alſo to ſome
of her zelous children of al puniſhment due for their offenſes, & ſuch
goe ſtraight to Heauen. Laſtly, that euery one hath not a reuelation of
his ſaluation as this man had, and therfore can not be ſo ſure as he
was.}
in Paradiſe.

\V And it was almoſt the ſixt houre: and there was made darkeneſſe vpon
the whole earth vntil the ninth houre. \V And the
\SNote{This eclipſe was ſeene and wondred at as a thing aboue nature, of
Dionyſius Areopagita at Thebes, when he was yet a Pagan.
\Cite{Dionyſ. ep. ad Polycarp}
et
\Cite{ep. ad Apollophanem.}}
ſunne was darkened; and the veile of the Temple was rent in the
middes. \V And \Sc{Iesvs} crying with a loud voice, ſaid:
\CNote{\XRef{Pſ.~30,~6.}}
\Emph{Father, into thy handes I commend my ſpirit.} And ſaying this, he
gaue vp the ghoſt.

\V And the Centurion ſeeing that which was done, glorified God, ſaying:
Indeed this man was iuſt. \V And al the multitude of them that were
preſent together at that ſight, and ſaw the things that were done,
returned knocking their breaſts. \V And al his acquaintance ſtood a far
off, and the women that had followed him from Galilee, ſeeing theſe
things.

\V
\CNote{\XRef{Mat.~27,~57.}
\XRef{Mr.~15,~43.}
\XRef{Io.~19,~38.}}
And behold a man named Ioſeph, which was a Senatour, a good man and a
iuſt. \V He had not conſented to their counſel and
%%% 2476
doings, of Arimathæa a citie of Iewrie, who alſo himſelf expected the
Kingdom of God. \V This man came to Pilate, and asked the body
of \Sc{Iesvs}. \V And taking it downe, wrapped it in ſindon, and laid
him in a monument hewed of ſtone,
\LNote{Wherein neuer}{As in the wombe of
\Fix{Marie}{\Sc{Marie}}{obvious typo, fixed in other}
none was conceiued before nor after him, ſo in this monument none was
laid before nor after him: which prerogatiue (no doubt) was of Gods
prouidence; this Ioſeph no leſſe abſtaining afterward to be buried in
it, then the other Ioſeph from copulation with the mother of our Lord.
\Cite{S.~Auguſtine.}}
wherein neuer yet any man had been laid. \V And it was the day of
Paraſceue, and the Sabboth drew neere. \V And the women that were come
with him from Galilee, following after
\LNote{Saw the monument}{Theſe
\MNote{Viſiting the Sepulcher, or Sepulchres.}
good women of great deuotion obſerued the ſepulchre for the honour of
the holy body. Whereupon the deuotion of faithful folke watching &
viſiting on Good-Friday and
\Fix{eaſter}{Eaſter}{obvious typo, fixed in other}
eue the ſepulchre made in euery Church for memorie of our Lords burial,
is exceeding good & godly, ſpecially the B.~Sacrament for more
ſignification ſake being preſent in the ſame Sepulcher.}
ſaw the monument, and how his body was laid. \V And returning they
prepared ſpices and ointments: and on the Sabboth they reſted according
to the commandement.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
