%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Galatians
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 2745
%%% o-2602
\startChapter[
  title={Chapter 3}
  ]

\Summary{By their owne conuerſion at the firſt, 6.~and by the example of
  Abraham, and promiſe made to him, he sheweth that the way to obtaine the
  benediction, is to ſeeke vnto God by faith in Chriſt. 10.~Seeing alſo
  that the Law curſeth euery one that hath not euermore kept the
  Law. 15.~And, that the Law was not giuen to alter God's Teſtament,
  19.~but to conuince the Iewes of ſinne, 23.~and ſo to be their
  pedagogue or leader vnto Chriſt, 25.~and then to ceaſe.}

O
\Fix{ſenſes}{Senſles}{obvious typo, fixed in other}
Galatians, who hath
\SNote{For any people or perſon to forſake the faith of their firſt
Apoſtles & conuerſion, at the voice of a few nouellaries, ſeemeth to
wiſe men a very bewitching and ſenſles brutishnes. Such is the caſe of
our poore countrie, Germanie, and others.}
bewitched you, not to obey the truth, before whoſe eyes \Sc{Iesvs}
Chriſt was proſcribed, being crucified among you? \V This only I would
learne of you: By the workes of the Law, did you receiue the Spirit, or
by the hearing of the faith? \V Are you ſo fooliſh, that wheras you
began with the Spirit, now you wil be conſummate with the fleſh? \V Haue
you ſuffered ſo great things without cauſe? if yet without cauſe. \V He
therfore that giueth you the Spirit, and worketh miracles among you; by
the workes of the Law, or by the hearing of the faith doeth he it? \V
As,
\CNote{\XRef{Gen.~15,~6.}
\XRef{Ro.~4,~3.}}
\Emph{Abraham beleeued God, and it was reputed to him vnto iuſtice.}

\V Know ye therfore that they that are
\SNote{This faith whereby Abraham, was iuſtified, and his children
the Gentils beleeuing in Chriſt, implieth al Chriſtian vertues; of which
the firſt is faith the ground and foundatiõ of al the reſt, and therfore
here and elſwhere often named of the Apoſtle.}
of faith, the ſame are the children of Abraham. \V And the Scripture
fore-ſeeing that God iuſtifieth the Gentils by faith, ſhewed vnto
Abraham before,
\CNote{\XRef{Gen.~12,~3.}}
\Emph{That in thee shal al Nations be bleſſed.} \V Therfore they that
are of faith, ſhal be bleſſed with the faithful Abraham. \V For
whoſoeuer are of the workes of the Law, are vnder curſe. For it is
written:
\CNote{\XRef{Deut.~27,~16.}}
\LNote{Curſed be.}{By
\MNote{Notwithſtanding venial ſinnes, men are truly iuſt, and may keep
the cõmandemẽts.}
this place the Heretikes would proue that no man is iuſt truely before
God, al being guiltie of damnation and God's curſe, becauſe they keep not
euery iot of the Law. Where indeed the Apoſtle meaneth not ſuch as
offend venially (as it is plaine by the place of Deuteronomie whence he
reciteth this text) but only ſuch as commit great & damnable crimes: and
ſo by grieuous and mortal tranſgreſſions wholy breake God's precepts,
and thereby incurre the curſe of the Law; from which the ſaid Law could
not deliuer them of it-ſelf, nor by any other meanes, but by the faith
and grace of \Sc{Christ Iesvs}.}
\Emph{Curſed be euery one that abideth not in al things that be written
in the booke of the Law, to doe them.} \V But that in the Law no man is
iuſtified with God, it is manifeſt, becauſe
%%% o-2603
\CNote{\XRef{Aba.~2.}}
\Emph{the iuſt
\LNote{Liueth by faith.}{It
\MNote{Not only by faith.}
is neither the Heretikes ſpecial preſumption and confidence, nor the
faith of Diuels, nor faith without workes, which is dead in it-ſelf as
\CNote{\XRef{Iac.~2.}}
S.~Iames ſaith, that can giue life to the iuſt: For that which is dead,
can not be the cauſe of life: But it is the Catholike faith, as
S.~Auguſtin writeth, which worketh by charitie (according to the
Apoſtles owne explication of this whole paſſage) by which the iuſt
liueth. 
\Cite{Li.~3. c.~5. cont. duas ep. Pelag.}
See
\XRef{the Annotation vpon the ſame words Rom.~1.}}
liueth by faith.} \V But the Law is not by faith: but,
\CNote{\XRef{Rom.~1.}}
\Emph{He that doeth thoſe things, shal liue in them.} \V Chriſt hath
redemed vs from the curſe of the Law, being made a curſe for vs (becauſe
it is written:
\CNote{\XRef{Leu.~18.}
\XRef{Deu.~21.}}
\Emph{Curſed is euery one that hangeth on a tree}) \V that on the
Gentils the bleſſing of Abraham might be made in Chriſt \Sc{Iesvs}: that
we may receiue the promiſe of the Spirit by faith.

\V Brethren (I ſpeake according to man) yet a man's teſtament being
confirmed no man deſpiſeth, or further diſpoſeth. \V To Abraham were
the promiſes ſaid, and to his ſeed. He ſaith not, \Emph{And to ſeeds},
as in many: but as in one:
\CNote{\XRef{Gen.~22,~18.}}
\Emph{And to thy ſeed}, which is Chriſt. \V And this I ſay, the
%%% 2746
teſtament being confirmed of God, the Law which was made after foure
hundred and thirtie yeares, maketh not void to fruſtrate the promiſe. \V
For if the inheritance be of the Law, now not of promiſe. But God gaue
it to Abraham by promiſe. \V Why was the Law then? It was put for
trãſgreſſions, vntil the ſeed came to whõ he had promiſed: ordeined by
Angels in the hand of a Mediatour. \V And a Mediatour is not of one: but
God is one. \V Was the Law then againſt the promiſes of God? God
forbid. For if there had been a Law giuen that could iuſtifie,
vndoubtedly iuſtice ſhould be of the Law. \V But the Scripture
\CNote{\XRef{Ro.~2,~9.}
\XRef{11,~32.}}
hath concluded al things vnder ſinne: that the promiſe by the faith of
\Sc{Iesvs} Chriſt might be giuen to them that beleeue. \V But before the
faith came, vnder the Law we were kept ſhut vp, vnto that faith which
was to be reuealed. \V Therfore the Law was our Pedagogue in Chriſt:
that we may be iuſtified by faith. \V But when the faith came, now we
are not vnder a Pedagogue. \V For you are al the children of God by
faith in Chriſt \Sc{Iesvs}. \V For as many of you as are baptized in
Chriſt,
\LNote{Haue put on Chriſt.}{Here
\MNote{Baptiſme giueth grace and iuſtification, not faith only.}
the Aduerſaries might haue ſeen, if they were not blinded by contentious
ſtriuing againſt God's Church, that when iuſtification is attributed to
faith without mention of good workes or other Chriſtian vertues &
Sacraments, it is not meant to exclude any of the ſame from the working
of iuſtice or ſaluation. For here we learne that by the Sacrament of
Baptiſme alſo we put on Chriſt, which is to put on faith, hope,
charitie, & al Chriſtian iuſtice. By the ſame we proue alſo that the
Sacramẽts of the new law gaue grace, for that the receiuers therof put
on Chriſt. And the Aduerſaries euaſion, that it is faith which worketh
in the Sacrament, and not the Sacrament it-ſelf, is plainely falſe:
Baptiſme giuing grace and faith it-ſelf to the infant that had none
before.}
haue put on Chriſt. \V There is not Iew nor Greek, there is not bond nor
free, there is not male nor femal.  For al you are one in
Chriſt \Sc{Iesvs}. \V And if you be Chriſts, then are you the ſeed of
Abraham, heires according to promiſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

