%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Two Timothy
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 2828
%%% o-2688
\startChapter[
  title={Chapter 2}
  ]

\Summary{He exhorteth him to labour diligently in his office,
  conſidering the reward in Chriſt, & his denial of them that deny him:
  14.~Not to contend, but to shun heretikes: neither to be moued to ſee
  ſome ſubuerted, conſidering that the elect continue Catholikes, and
  that in the Church be of al ſorts. 24.~Yet with al ſweetenes to reclaime
  the deceiued.}

Thou therfore, my ſonne, be ſtrong in the grace which is in Chriſt
\Sc{Iesvs}: \V and the things which thou haſt heard of me by many
witneſſes, theſe commend to faithful men, which ſhal be fit to teach
others alſo. \V Labour thou as a good ſouldiar of Chriſt \Sc{Iesvs}. \V
\LNote{No man being a ſouldiar.}{Firſt
\MNote{What ſecular affaires doe not agree nor conſiſt with ſpiritual
mens function.}
of al, the Apoſtle
\XRef{(1.~Cor.~7.)}
maketh marriage and the needful cares, ſolicitude, and diſtractions
therevpon euer depending, ſpecial impediments of al ſuch as ſhould
employ themſelues wholy to God's ſeruice, as Biſhops and Prieſts are
bound to doe. \Emph{He that is with a wife} (ſaith he) \Emph{is careful
for the world, how to pleaſe his wife, and is diſtracted or deuided.}
\XRef{1.~Cor.~7.}

Secondly, the practiſe of phyſicke, merchandiſe, or any other profane
facultie and trade of life to gather riches, and much more to be giuen
to hunting, hawking, gamning, ſhewes, enterludes, or the like paſtimes,
is here forbidden.

Thirdly, the ſeruice of Princes and manifold baſe offices done to them
for to obtaine dignities and promotions, are diſagreable to Prieſtly
functions.
\MNote{How ſpiritual men may ſerue ſecular Princes, & deale in ciuil
cauſes.}
Not ſo, to be their Chaplens for this purpoſe to preach vnto them, to
heare their confeſſions, to miniſter the
\Fix{Sacramens}{Sacraments}{obvious typo, fixed in other}
vnto them, to ſay Diuine ſeruice before them, and ſuch other ſpiritual
dueties. For, al ſuch ſeruices done to principal perſons both of the
Clergie and Laitie, be godly and conſonant to Prieſtly vocation. As alſo
ſeruing of Princes and Commonweales in ciuil cauſes and matters of
ſtate, in making peace and quietnes among the people, by deciding or
compounding their controuerſies, al ſuch like affaires tending to the
honour of God and good of men, and to the vpholding of true religion,
when they may be done without notorious damage or hinderance of their
ſpiritual charge, or when the hurts thereof be aboundantly recompenſed by
the neceſſarie dueties done for the general good of Kingdom or Countrie:
al ſuch things (I ſay) be lawful and often very requiſit. And
\CNote{\Cite{In vit. S.~Ambroſ. & Bern.}}
S.~Auguſtin, S.~Ambroſe, S.~Bernard, and other holy Biſhops of old were
much occupied therein, as we ſee in S.~Auguſtines booke
\Cite{de opere Monachorum c.~20.}
&
\Cite{Poſsid in vit. c.~19.}}
No man being a ſouldiar to God, intangleth himſelf with ſecular
buſineſſes; that he may pleaſe him to whom he hath approued himſelf. \V
For he alſo that ſtriueth for the maiſterie, is not crowned vnleſſe he
ſtriue lawfully. \V The huſband-man that laboureth, muſt firſt take of
the fruits. \V Vnderſtand what I ſay: for our Lord wil giue thee in al
things vnderſtanding. \V Be mindful that our Lord \Sc{Iesvs} Chriſt is
riſen againe from the dead, of the ſeed of Dauid, according to my
Ghoſpel, \V wherin I labour euen vnto bands, as a malefactour: but the
word of God is not tied. \V Therfore
\SNote{Marke here that the elect (though ſure of ſaluation) yet are
ſaued by meanes of their Preachers and Teachers, as alſo by their own
endeauours.}
I ſuſtaine al things for the elect, that they alſo may obtaine the
ſaluation, which is in Chriſt \Sc{Iesvs}, with heauenly glorie. \V A
faithful ſaying. For is we be dead with him, we ſhal liue alſo
together. \V If we ſhal ſuſtaine, we ſhal alſo reigne together.
\CNote{\XRef{Mat.~10.}}
If we ſhal deny, he alſo wil denie vs. \V
\CNote{\XRef{Ro.~3,~3.}}
If we beleeue not; he continueth faithful, he can not denie himſelf. \V
Theſe things admoniſh: teſtifying before our Lord.

%%% o-2689
Contend not in word, for it is profitable for nothing, but for the
ſubuerſion of them that heare. \V Carefully prouide to preſent thy ſelf
approued to God, a worke-man not to be confounded,
\LNote{Rightly.}{The
\MNote{Catholikes only, right handlers of the Scripture.}
Scriptures or chalenge of the word of God is common to Catholikes and
Heretikes, but al is in the handling of them. Theſe later handle them
guilefully, adultering the word of God, as
\CNote{\XRef{2.~Cor.~2.}
&
\XRef{4.}}
els-where the Apoſtle ſpeaketh: the other ſincerely after the manner of
the Apoſtles and Doctours of God's Church. Which the Greek expreſſeth by
a ſignificant word of cutting a thing ſtraight by a
line, \G{ὀρθοτομοῦντα}.}
rightly handling the word of truth. \V But
\SNote{See the
\XRef{Annotations before 1.~Tim.~6. v.~20.}}
prophane and vaine ſpeaches auoid: for they doe much grow to
impietie: \V and
\LNote{Their ſpeache.}{The
\MNote{Heretical books and ſermons are to be auoided.}
ſpeaches, preachings, and writings of Heretikes be peſtiferous,
contagious, and creeping like a canker. Therfore Chriſtian men muſt
neuer heare their ſermons nor read their books. For ſuch men haue a
popular way of talke whereby the vnlearned, and ſpecially women loden
with ſinne, are eaſily beguiled. \Emph{Nothing is ſo eaſy} (ſaith
S.~Hierom) \Emph{as with voluble and rolling ſong to deceiue the rude
people, which admire whatſoeuer they vnderſtand not.}
\Cite{ep.~2. ad Nepot. c.~10.}}
their ſpeach ſpreadeth as a canker: of whõ is Hymenæus
%%% 2829
and Philetus: \V who haue erred from the truth, ſaying that the
reſurrection is done already, and haue ſubuerted the faith of ſome.

\V But the ſure foundation of God ſtandeth, hauing this ſeale, Our Lord
knoweth who be his, and let euery one depart from iniquitie that nameth
the name of our Lord. \V But
\LNote{In a great houſe.}{He meaneth not that Hymenæus and Philetus (of
whom he ſpake immediately before) or other heretikes, be properly within
the Church, as Catholike men are, though grieuous ſinners:
\MNote{Who are out of the Church or within it.}
but that euil men who for the puniſhment of their ſinnes become
heretikes, were before they fel from their faith as veſſels of
contumelie, within the Church. Yea and often alſo after they be ſeuered
in hart and in the ſight of God, ſo long as they ſtand in external
profeſſion and vſe of the ſame Sacraments, and in the outward fellowſhip
of Catholikes, not yet either ſeparated of themſelues, nor caſt out by
the Gouernours of the Church, ſo long (we ſay) they be after a ſort in
the Church: though properly and indeed they be out of the compaſſe of
God's houſe. Mary of thoſe that are openly ſeuered in Sacraments,
Seruice, and communion, there is no queſtion but they are out of the
Church.}
in a great houſe there are not only veſſels of gold and of ſiluer, but
alſo of wood and of earth: and certaine indeed vnto honour, but certaine
vnto contumelie. \V If any man therfore ſhal
\LNote{Cleanſe himſelf.}{Man
\MNote{Free-wil.}
then hath free-wil to make himſelf a veſſel of ſaluation or damnation:
though ſaluation be attributed to God's mercie principally, the other to
his iuſt iudgement: neither of both being repugnant to our free-wil, but
working with and by the ſame, al ſuch effects in vs as to his prouidence
and our deſerts be agreable.}
cleanſe himſelf from theſe, he ſhal be a veſſel vnto honour, ſanctified
& profitable to our Lord, prepared to euery good worke.

\V But youthful deſires fly, and purſue iuſtice, faith, charitie, and
peace with them that inuocate our Lord from a pure hart. \V And
\CNote{\XRef{Tit.~3,~9.}}
fooliſh and vnlearned queſtions auoid, knowing that they ingender
brauls. \V But the ſeruant of our Lord muſt not wrangle: but be mild
toward al men, apt to teach, patient, \V with modeſtie admonishing them
that reſiſt the truth: leſt ſometime
\SNote{Conuerſion from ſinne & hereſie is the guift of God and of his
ſpecial grace: yet here we ſee, good exhortations and prayer and ſuch
other helps of man be profitable thereunto. Which could not be if we had
not free-wil.}
God giue them repentance to know the truth: \V and they recouer
themſelues from the ſnares of the diuel, of whom they are held captiue
at his wil.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

