%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Two Timothy
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 2826
%%% o-2686
\startChapter[
  title={Chapter 1}
  ]

\Summary{With his praiſes he couertly exhorteth him not to be diſmaied
  for his trouble, 6.~(hauing grace giuen in Orders to help him, 8.~and
  knowing for what cauſe he is perſecuted) and namely with the example
  of Oneſiphorus.}

Paul an Apoſtle of \Sc{Iesvs} Chriſt by the wil of God, according to the
promiſe of the life which is in Chriſt \Sc{Iesvs}: \V to Timothee my
deareſt ſonne, grace, mercie, peace from God the Father, and Chriſt
\Sc{Iesvs} our Lord.

\V I giue thankes to God, whom I ſerue from my Progenitours in a pure
conſcience, that without intermiſſion I haue a memorie of thee in my
praiers, night and day \V deſiring to ſee thee, mindful of thy teares,
that I may be filled with ioy, \V calling to mind that faith which is in
thee not feined, which alſo dwelt firſt
\LNote{In thy grandmother.}{Though
\MNote{A great bleſſing to haue Catholike progenitours: and very
commendable to cleaue faſt to their faith.}
God ſhew mercie to many that be of incredulous, heretical or il parents,
yet it is a goodly benediction of God to haue good education & to haue
good faithful progenitours and Catholike parents. And it is a great
ſinne to forſake the faith of our fathers that be Catholikes, or
contrarie to our education in the Church to follow ſtrange doctrines,
abandoning not only our next natural parents faith, but the
\Fix{antient}{ancient}{likely typo, fixed in other}
faith and beleefe of al our progenitours for many hundred yeares
together. And if to follow the faith of mother and grandmother only, the
Chriſtian religion being then but newly planted, was ſo commendable euen
in a Biſhop, how much more it is now laudable to cleaue faſt to the
faith of ſo many our progenitours and Ages that continued in the ſame
Chriſtian religion which they firſt receiued.

Our
\MNote{The peoples ſpeaches of their fathers faith, is very Chriſtian
and laudable.}
Proteſtants in their great wiſedom laugh at good ſimple men when they
talke of their fathers faith. But S.~Hierom,
\CNote{\Cite{Apol. cont. Ruff. li.~1. c.~8.}}
\Emph{I am a Chriſtian}, ſaid he, \Emph{and borne of Chriſtian parents,
and carie the ſigne of the croſſe in my forehead.} And againe
\Cite{ep.~65. c.~3,}
\Emph{Vntil this day the Chriſtian world hath been without this
doctrine, that faith wil I hold faſt being an old man, wherein I was
borne a child.} And the holy Scriptures ſet vs often to ſchole to our
fathers.
%%% !!! Split this?
\CNote{\XRef{Deut.~32.}
&
\XRef{Pſal.~43.}}
\Emph{Aske thy fathers, and they wil shew thee, thy anceſtours, and they
wil tel thee.} And againe, \Emph{Our fathers haue shewed vnto vs.} And
commonly the true God is called the God of the faithful and of their
forefathers,
\XRef{Dan.~2,~3.}
And falſe Gods and new doctrines or opinions be named, \Emph{New and
fresh, ſuch as their fathers worshipped not.}
\XRef{Deut.~32.}
Finally S.~Paul both here and
\CNote{\XRef{Act.~14.}
\XRef{2.~Cor.~11.}}
often els alleageth for his defenſe and commendation, that he was of
faithful progenitours. And it is a caſe that Heretikes can not lightly
bragge of, no one ſect commonly during ſo long without intermiſſion,
that they can haue many progenitours of the ſaid ſect. Which is a
demonſtration that their faith is not true, and that it is impoſſible
our Catholike faith to be falſe, ſuppoſing the Chriſtian religion to be
true.}
in thy grand-mother Lois, and thy mother Eunice, and I am ſure that in
thee alſo. \V For the which cauſe I admoniſh thee that thou reſuſcitate
the
\SNote{Here againe it is plaine that holy Orders giue grace, and that
euen by and in the external ceremonie of impoſing the Biſhops hands. And
it is a manner of ſpeach ſpecially vſed in this Apoſtle, and S.~Luke,
that Orders giue grace to the ordred, and that to take orders or
authoritie to miniſter Sacramẽts or preach, is, to be giuen or
deliuered to God's grace.
\XRef{Act.~14,~25.}}
grace of God, which is in thee by the impoſitiõ of my hands. \V For God
hath not giuen vs the ſpirit of feare: but of power, and loue, and
ſobrietie. \V Be not therfore aſhamed of the teſtimonie of our Lord, nor
of me his priſoner: but trauail with the Ghoſpel according to the power
of God, \V who hath deliuered and called vs by his holy calling,
\CNote{\XRef{Tit.~3,~5.}}
not according to our workes, but according to his purpoſe and grace,
which was giuen to vs in Chriſt \Sc{Iesvs}
\CNote{\XRef{Tit.~1,~3.}}
before the ſecular times. \V But it is manifeſted now by the
illumination
%%% o-2687
of our Sauiour \Sc{Iesvs} Chriſt, who hath deſtroied death, and
illuminated life and incorruption by the Ghoſpel: \V Wherin
\CNote{\XRef{1.~Timo.~2,~7.}}
I am appointed a preacher and Apoſtle and Maiſter of the Gentils. \V For
the which cauſe alſo I ſuffer theſe things: but I am not confounded. For
I know
%%% 2827
whom I haue beleeued, and I am ſure that he is able to keep my
\LNote{Depoſitum.}{A
\MNote{Al our good deedes are laid vp with God, to be rewarded.}
great comfort to al Chriſtians, that euery of their good deedes and
ſufferings for Chriſt, and al the worldly loſſes ſuſteined for defenſe or
confeſſion of their faith, be extant with God, and kept
as \L{depoſitum}, to be repaied or receiued againe in heauen. Which if
the worldlings beleeued or conſidered, they would not ſo much maruel to
ſee Catholike men ſo willingly to loſe land, libertie, credit, life and
al for Chriſtes ſake and the Churches faith.}
\L{depoſitum} vnto that day.

\V Haue thou
\LNote{A forme.}{The
\MNote{We muſt ſpeake in Catholike termes, after a certaine rule of
faith, and forme of wordes.}
Apoſtles did ſet downe a platforme of faith,
doctrine, and phraſe of Catholike ſpeach and preaching, & that not ſo
much by writing (as here we ſee) as by word of mouth: to which he
referreth Timothee ouer and aboue his Epiſtles vnto him. And how
preciſely Chriſtian Doctours ought to keep the forme of words anciently
appropriated to the myſteries & matters of our religion, S.~Auguſtin
expreſſeth in theſe wordes
\Cite{li.~10. de ciuit. c.~13.}
\Emph{Philoſophers ſpeake with freedom of words &c. but we muſt ſpeake
according to a certaine rule, leſt licentious libertie of words breed an
impious opinion of the things alſo that are ſignified by the ſame.}
Trinitie, perſon, eſſence, Conſubſtantial, Tranſſubſtantiation, Maſſe,
Sacrament, and ſuch like, be \L{verba ſana} (as the Apoſtle ſpeaketh)
\Emph{ſound words}, giuen to expreſſe certaine high truths in religion,
partly by the Apoſtles and firſt Founders of our religion vnder Chriſt,
and partly very aptly inuented by holy Councels & Fathers, to expreſſe
as neere as could be the high ineffable or vnſpeakable veritie of ſome
points, and to ſtop the Heretikes audacitie and inuention of new words
and prophane ſpeaches in ſuch things, which the Apoſtle warneth Timothee
to auoid.
\XRef{1.~ep. c.~6,~20.}
and
\XRef{2.~ep.~2,~16.}
See the
%%% 1 Tim. 6, 20
\XRef{Annotation there.}}
a forme of ſound words, which thou haſt heard of me in faith and
\SNote{Faith and loue copled cõmonly together in this Apoſtles
writings.}
\TNote{\G{ἀγάπῃ τῇ ἐν Χριστῷ}}
in the loue in Chriſt \Sc{Iesvs}. \V Keep the good \L{depoſitum} by the
Holy Ghoſt, which dwelleth in vs.

\V Thou knoweſt this, that al which are in Aſia, be auerted from me: of
whom is Phigelus and Hermogenes.

\V Our Lord giue mercie to
\CNote{\XRef{2.~Tim.~4,~19.}}
the houſe of Oneſiphorus: becauſe he hath often refreſhed me, and hath
\SNote{What a happie and meritorious thing it is to relieue the
afflicted for religion, and not to be aſhamed of their diſgrace, yrons,
or what miſeries ſo euer.}
not been aſhamed of my chaine: \V but when he was come to Rome, he
ſought me carefully, and found me. \V
\LNote{Our Lord.}{To
\MNote{Relieuers of Cath. priſoners.}
haue this praier of an Apoſtle, or any Prieſt or poore Cath. man ſo
relieued, giueth the greateſt hope at the day of our death or general
iudgement, that can be: and it is worth al the lauds, honours, and
riches of the world.}
Our Lord grant him to find mercie of our Lord in that day. And how many
things he miniſtred to me at Epheſus, thou knoweſt better.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

