%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Hebrews
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-11


\project douay-rheims


%%% 2869
%%% o-2730
\startChapter[
  title={Chapter 11}
  ]

\Summary{He exhorteth them by the definition of faith, to ſticke vnto
  God, though they ſee not yet his reward: shewing that al the Saints
  afore-time did the like, being al conſtant in faith, though not one of
  them receiued the promiſe, that is, the inheritance in heauen: but
  they and we now after the comming of Chriſt receiue it together.}

And
\LNote{Faith is.}{By
\MNote{Not only or a ſpecial faith.}
this deſcription of faith, and by al the commendation thereof through
the whole chapter, you may wel perceiue that the Apoſtle knew not the
forged ſpecial faith of the Proteſtants, whereby euery one of theſe new
Sect-Maiſters and their followers beleeue their ſinnes are remitted,
and that themſelues shal be ſaued, though their ſectes be cleane
contrarie one to another.}
faith is,
\SNote{By this word ſubſtãce is meant, that faith is the ground of our
hope.}
the ſubſtance of things to be hoped for,
\TNote{\G{ἔλεγχος}}
the argument of things
\LNote{Not appearing.}{\Emph{This
\MNote{Faith is of things not ſeen: as in the B.~Sacrament.}
is the praiſe of faith}, ſaith S.~Auguſtin, \Emph{if that which is
beleeued, be not ſeen. For what great thing is it, if that be beleeued,
which is ſeen? according to that ſentence of our Lord when he rebuked
his diſciple, ſaying: Becauſe thou haſt ſeen me Thomas, thou haſt
beleeued: bleſſed are they that haue not ſeen and haue beleeued.}
\Cite{Aug. in euang. Io. tract.~79.}
Which may be a rebuke alſo and a check to al thoſe faithles ſpeaches, I
would ſee him, taſt him, touch him and feele his very flesh in the
Sacrament, otherwiſe I wil not beleeue.}
not appearing. \V For in this the old men obtained teſtimonie.

\V
\CNote{\XRef{Gen.~1,~13.}}
By faith, we vnderſtand that the worlds were framed by the word of God:
that of inuiſible things viſible things might be made.

\V
\CNote{\XRef{Gen.~4,~4.}}
By faith, Abel offered a greater hoſt to God then Cain:
\CNote{\XRef{Mt.~23,~35.}}
by which he obtained teſtimonie that he was iuſt, God giuĩg teſtimonie
to his guifts, and by it, he being dead, yet ſpeaketh. \V
\CNote{\XRef{Gen.~5,~24.}
\XRef{Ec.~44,~16.}}
By faith
\SNote{Here it appeareth that Henoch yet liueth and is not dead: againſt
the Caluiniſts. See the
\XRef{annot. Apocal. chap.~11.}}
Henoch was tranſlated, that he ſhould not ſee death, and he was not
found: becauſe God tranſlated him. For before his tranſlation he had
teſtimonie that he pleaſed God. \V But without faith it is impoſſible to
pleaſe God. For
\LNote{He that commeth.}{Faith
\MNote{Nothing profitable or meritorious without faith.}
is the foundation and ground of al other vertues and worship of God,
without which no man can pleaſe God. Therfore if one be a Iewe, a
Heathen, or an heretike, that is to ſay, be without the Catholike faith,
al his workes shal profit him no whit to ſaluation.}
he that commeth to God, muſt beleeue that he is, and is a
\SNote{We muſt beleeue that God wil reward al our good workes: for he is
a rewarder of true iuſtice, not an accepter or imputer of that that is
not.}
rewarder to them that ſeeke him.

\V
\CNote{\XRef{Gen.~6,~13.}
\XRef{Eccl.~44,~17.}}
By faith, Noe hauing receieued an anſwer concerning thoſe things which
as yet were not ſeen, fearing, framed the arke for the ſauing of his
houſe, by the which he condemned the world: and was inſtituted heire of
the iuſtice which is by faith.

\V
\CNote{\XRef{Gen.~12,~4.}
\XRef{13,~1.}}
By faith, he that is called, Abraham, obeied to goe forth into the place
which he was to receiue for inheritance: and he went forth, not knowing
whither he went. \V By faith, he abode in the land of promiſe, as in a
ſtrange land,
%%% o-2731
dwelling in cottages with Iſaac & Iacob the
%%% !!! pages out of order here
%%% 2871
coheires of the ſame promiſe. \V For he expected the citie that hath
foundations: whoſe artificer and maker is God.

\V
\CNote{\XRef{Gen.~17,~19.}
\XRef{18,~10.}
\XRef{21,~2.}
\XRef{Eccl.~44,~22.}}
By faith, Sara alſo her ſelf being barren, receiued vertue in conceauing
of ſeed, yea paſt the time of age: becauſe ſhe beleeued that he was
faithful which had promiſed. \V For the which cauſe euen of one (and him
quite dead) there roſe as the ſtarres of heauen in multitude, and as the
ſand that is by the ſea ſhore innumerable.

\V According to faith died al theſe, not hauing receiued the promiſes,
but beholding them a farre off, and ſaluting them, and confeſſing that
they are pilgrimes & ſtrangers vpon the earth. \V For they that ſay
theſe things, doe ſignifie that they ſeeke a countrie. \V And indeed if
they had been mindful of the ſame from whence they came forth, they had
time verily to returne. \V But now they deſire a better, that is to ſay,
a heauenly. Therfore God is not confounded to be called their God. For
he hath prepared them a citie.

\V
\CNote{\XRef{Gen.~22,~9.}}
By faith, Abraham offered Iſaac, when he was tempted: and his
onlie-begotten did he offer who had receiued the promiſes: (\V to whom
it was ſaid,
\CNote{\XRef{Gen.~21,~12.}
\XRef{Rom.~9,~7.}}
\Emph{That in Iſaac shal ſeed be called to thee.}) \V accounting that God
is able to raiſe vp euen from the dead. Whereupon he receiued him alſo
\SNote{That is, in figure and myſterie of Chriſt dead, & aliue againe.}
for a parable. \V
\CNote{\XRef{Gen.~27,~27.~36.}}
By faith, alſo of things to come, Iſaac bleſſed Iacob and Eſau.

\V
\CNote{\XRef{Gen.~48,~15.}}
By faith, Iacob dying, bleſſed euery one of the ſonnes of Ioſeph:
\CNote{Gen.~47,~31.}
and
\LNote{Adored the top of his rod.}{The
\MNote{The citations in the new Teſtament, not only according to the
Hebrew, but to the Septuagint.}
learned may ſee here that the Apoſtle doth not tye himſelf to the Hebrew
in the place of Geneſis
\CNote{\XRef{Gen.~47. v.~31.}}
whence it is alleaged, but followeth the
Septuaginta, though it differ from the Hebrew, as alſo the other Apoſtles
and Euangeliſts & our Sauiour himſelf did: neither were they curious (as
men now adaies) to examine al by the Hebrew only, becauſe they
writing and ſpeaking by the Holy Ghoſt, knew very wel that this
tranſlation
\CNote{\Cite{Aug. de ciu. Dei li.~15. c.~14.}}
is the ſenſe of the Holy Ghoſt alſo, and as true, and as directly
intended as the other: and therfore alſo that tranſlation continued
alwaies authentical in the Greek Church, notwithſtanding the diuerſitie
thereof from the Hebrew.
\MNote{The vulgar Latin tranſlation.}
Euen ſo we that be Catholikes, follow with al the Latin fathers the
authentical Latin tranſlatiõ, though it be not alwaies agreable to the
Hebrew or Greek that now is. But Caluin is not only very ſaucie, but
very ignorant, when he ſaith that the Septuaginta were deceiued, and yet
that the Apoſtle without curioſity was content to follow them: becauſe
it is euident, that
%%% !!! Marked in both, but only in other
\TNote{\H{מטה}}
the Hebrew being then without points,
\TNote{\G{ῥάβδος}, rod
\G{κλίνη}, bed}
might be tranſlated the one way as wel as the other. Which they
vnderſtood ſo wel (and therfore were not deceiued) that within three
lines after, in the beginning of the next chapter, they tranſlate the
ſame word, as he would haue it in this place.

Againe
\MNote{Adoration of creatures, and namely of holy things.}
obſerue in thoſe words, \Emph{He adored the top of his rod}, that
adoration (as the Scripture vſeth this word) may be done to creatures,
or to God at and before a creature: as, at or
\CNote{\XRef{Ioſ.~7,~6.}}
before the Arke of the Teſtament in old time, now at or before the
crucifixe, relikes, images: and in the Pſalmes
\XRef{98.}
\XRef{131.}
\Emph{Adore ye his foot-ſtoole. Adore ye toward his holy mount. We wil
adore toward the place where his feet ſtood}: or (which by the Hebrew
phraſe is al one) \Emph{Adore ye his holy mount. We wil adore the place
where his feet ſtood}; as alſo
\CNote{\Cite{Oecũ in collect.}}
the Greek Fathers, S.~Damaſcene
\Cite{li.~1. de imaginibus,}
& Leontius cited of him, yea S.~Chryſoſtom alſo doe handel
%%% !!! Cite ?
theſe places, and namely that of the Apoſtle which we not ſpeake of,
interpreting the Greek as our Latin hath, and as we doe, He adored the
rod or the top of his rod, that is, the ſcepter of Ioſeph now Prince of
Ægypt, ſo fulfilling Ioſephs dreames which foretold the ſame
\XRef{Gen.~17.}
and withal ſignifying as it were by this prophetical fact
\CNote{\XRef{3.~Reg.~11,~12.}}
the Kingdom of Iſrael or of the ten Tribes that was to come of Ioſeph by
Ephraim his yonger ſonne in the firſt King Ieroboam. Thus the Greek
Fathers. Whereunto may be added, that al this was done in type and
figure of Chriſtes ſcepter & Kingdom, whom he adored by and in his
Croſſe, as he did Ioſeph by or in his rod and ſcepter: and therfore the
Apoſtle ſaith, he did it by faith, as hauing reſpect toward things to
come.
\MNote{Corrupt tranſlation againſt \L{Dulia}.}
By al which it is euident, that it is falſe which the Caluiniſts teach,
that we may not adore image, crucifixe, or any viſible creature, that
is, we may not adore God at or by ſuch creatures, nor kneel before them:
and therfore their corrupt tranſlation of this place for the ſame
purpoſe is intolerable, ſaying thus, (\Sc{Leaning}) \Emph{vpon his ſeate
he adored} (\Sc{God},) adding no leſſe then two words more then is in
the Greek. Which though it might be the ſenſe of the place and
S.~Auguſtin ſo expoundeth it, yet they should not make his expoſition
the text of holy Scripture, ſpecially whereas he only of al the ancient
Fathers (as Beza confeſſeth) ſo expoundeth.}
adored the top of his rod.

\V
\CNote{\XRef{Gen.~50,~24,~25.}}
By faith, Ioſeph dying, made mention of the going forth of the children
of Iſrael: and gaue commandement
\SNote{The Tranſlation of Relikes or Saints bodies, & the due regard and
honour we ought to haue to the ſame, are proued hereby.}
concerning his bones.

\V
\CNote{\XRef{Exo.~2,~2.}}
By faith, Moyſes being borne, was hid three moneths by his parents:
becauſe they ſaw him a proper infant, and they feared not
\CNote{\XRef{Exo.~1,~16.}}
the Kings edict.

\V
\CNote{\XRef{Exo.~2,~11.}}
By faith, Moyſes being made great, denied himſelf to be the ſonne of
Pharao's daughter: \V rather choſing to be afflicted with the people of
God, then to haue the pleaſure of temporal ſinne, \V eſteeming the
reproche of Chriſt, greater riches then the treaſure of the
Ægyptians. For
\SNote{The Proteſtants that deny we may or ought to doe good in reſpect
or for reward in heauen, are confuted.}
he looked vnto the remuneration. \V
\CNote{\XRef{Exo.~12,~37.}}
By faith, he left Ægypt: not fearing the fiercenes of the King. For him
that is inuiſible he ſuſteined as if he had ſeen him. \V By faith, he
celebrated
%%% o-2732
the Paſche, & the ſheading of the bloud: that he which deſtroyed the
firſt-borne, might not touch them. \V
\CNote{\XRef{Gen.~14,~22.}}
By faith they paſſed the red ſea as it were by the drie land: which the
Ægyptians aſſaying, were deuoured.

\V
\CNote{\XRef{Ios.~6,~20.}}
By faith the walles of Iericho fel downe, by the circuiting of ſeuen
daies.

\V
\CNote{\XRef{Ios.~6,~23.}
\XRef{23,~2.~3.}}
By faith, Rahab the harlot periſhed not with the incredulous, receiuing
the ſpies with peace.

\V And what ſhal I yet ſay? For the time wil faile me telling of Gedeon,
Barac, Sampſon, Iephte, Dauid, Samuel, & the Prophets: \V who by faith
ouercame Kingdoms,
\LNote{Wrought iuſtice.}{Men
\MNote{Not faith only.}
are not iuſt by beleefe only, as the Proteſtants affirme, but by working
iuſtice. And we may note that in al this long commendation of faith in
the Fathers and holy perſons, their good workes are alſo ſpecially
recounted, as Rahabs harbouring the ſpies,
\CNote{\XRef{Iac.~2.}}
Abrahams offering his ſonne (which their workes S.~Iames doth inculcate)
Noes making the Arke
\XRef{Gen.~6.}
Abels better oblatiõ then Cains
\XRef{Gen.~4.}
&
\XRef{Heb.~11. v.~4.}
and ſo-forth. Therfore S.~Clement Alexandrinus ſaith,
\CNote{\Cite{Li.~4. Stro. pag.~240.}}
that the ſaid perſons & others were iuſt by faith and obedience, by
faith and hoſpitality, by faith and patience, by faith and humility.

The
\MNote{No workes of the Patriarkes or any other profitable, but by
their faith in Chriſt: Which is alwaies the Apoſtles meaning in
commending faith.}
Apoſtles purpoſe then is nothing els, but to proue to the Hebrewes (who
made ſo great account of their Patriarches and forefathers and their
famous actes) that al theſe glorious perſonages and their workes were
commendable and acceptable only through the faith they had of Chriſt,
without which faith none of al their liues & workes should haue profited
them any whit: the Gentils doing many noble actes (as Heretikes may alſo
doe) which are of no eſtimation before God, becauſe they lacke
faith. And that is the ſcope of S.~Paules Epiſtle to the Romanes, and of
al other paſſages where he commendeth faith: further prouing ſpecially
in this Epiſtle to the Hebrewes, that al their Sacrifices were nothing
els but figures and atteſtations of the Chriſtian faith in Chriſt and his
death. Al which high reſolution & concluſion againſt the Iewes and
Gentils, that the Chriſtian faith is the true faith & religion, the
Heretikes of our time ignorantly and brutishly abuſe againſt Chriſtian
workes, Sacrifice and Sacraments, which the Apoſtle meant ſpecially to
commend and eſtablish by his high commendation of the faith in Chriſt.}
wrought iuſtice, obteined promiſes, ſtopped the mouths of lions, \V
extinguiſhed the force of fire, repelled the edge of the ſword,
recouered of their infirmitie, were made
%%% !!! pages out of order here
%%% 2870
ſtrong in battel, turned away the camp of forrainers: \V women receiued
of reſurrection their dead, and others were racked, not accepting
redemption, that they might find a better reſurrection. \V And others had
trail of mokeries and ſtripes, moreouer alſo of bands and priſons: \V
they were ſtoned, they were hewed, they were tempted, they died in the
ſlaughter of the ſword, they went about in ſheep-skinnes, in goats
skinnes, needy, in diſtreſſe, afflicted: \V of whom the world was not
worthie; wandering in deſertes, in mountaines & dẽnes, and in caues of
the earth. \V And al theſe being approued by the teſtimonie of faith,
receiued not the promiſe, \V God for vs prouiding ſome better thing,
that they
%%% !!! LNote not marked in either
\LNote{Without vs should not.}{The
\MNote{The Patriarches and other iuſt not in heauen before Chriſt.}
Fathers before Chriſt should not be accomplished, that is, not admitted
to the heauenly ioyes, viſion, and fruition of God, til the Apoſtles and
other of the new law were aſſociate to them, and the way to euerlaſting
glorie opened by our Lordes death and Aſcenſion. Neither shal either
they or we be fully perfected in glorie both of body and ſoul, til the
general reſurrection: God's prouidence being ſo, that we should not one
be conſummated without another, al being of one faith, and redeemed by
one Lord Chriſt.}
without vs ſhould not be conſummate.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

