%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Hebrewes
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 2845
%%% o-2706
\startChapter[
  title={Chapter 03}
  ]

\Summary{By example of Chriſt (who is incomparably more excellent then
  Moyſes alſo) he exhorteth them to be faithful vnto God. 7.~Their reward
  shal be, to enter into euerlaſting reſt, if they perſeuer: as
  contrariewiſe to be excluded (as was shadowed in their forefathers in
  the wildernes) if they ſinne and become incredulous.}

Wherefore, holy Brethren, partakers of the heauenly vocation, conſider
the Apoſtle, and high Prieſt of our confeſſion \Sc{Iesvs}: \V who is
faithful to him that made him, as alſo 
\CNote{\XRef{Nu.~12,~7.}}
Moyſes in al his houſe. \V 
%%% !!! SNote ?
\MNote{The excellencie of Chriſt aboue Moyſes.}
For, this man is eſteemed worthie of more ample glorie aboue Moyſes, by
ſo much as more ample glorie then the houſe, hath he that framed it. \V
For euery houſe is framed of ſome man. But he that created al things, is
God. \V And Moyſes indeed was faithful in al his houſe as a ſeruant, for
a teſtimonie of thoſe things which were to be ſaid: \V but Chriſt as the
Sonne in his owne houſe: which houſe are we, if we keep firme the
confidence and glorie of hope vnto the end.

\V Wherfore, as the Holy Ghoſt ſaith:
\CNote{\XRef{Pſ.~94,~8.}}
\Emph{To day if you shal heare his voice, \V harden not your harts as in
the exacerbation according to the day of tentation in the deſert, \V
where your fathers tempted me: proued and ſaw my workes \V fourtie
yeares. For the which cauſe I was offended with this Generation, and
ſaid, They doe alwaies erre in hart. And they haue not knowen my
waies. \V
%%% !!! Only in this
\Var{To whom}{as}
I ſware in my wrath, If they shal enter into my reſt.}

\V Beware Brethren, leſt perhaps there be in ſome of you an euil hart of
incredulitie, to depart from the liuing God. \V But exhort your ſelues
euery
%%% 2846
day, whiles \Emph{to day} is named, that none of you be obdurate with
the fallacie of ſinne. \V For we be made partakers of Chriſt: yet ſo if
we keep the
\SNote{Faith is the groundworke of our creation in Chriſt, which if we
hold not faſt, al the building is loſt.}
beginning of his ſubſtance firme vnto the end. \V While it is ſaid,
\Emph{to day if you shal heare his voice, doe not obdurate your harts as
in that exacerbation}. \V For ſome hearing did exaſperate: but not al
they that went out of Ægypt by Moyſes. \V And with whom was he offended
fourtie yeares? was it not with them that ſinned,
\CNote{\XRef{Nu.~14,~37.}
\XRef{21,~23.}}
whoſe carcaſſes were ouerthrowen in the deſert? \V And to whom did he
ſweare that they ſhould not enter into his reſt: but to them that were
incredulous? \V And we ſee that they could not enter in, becauſe of
incredulitie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

