%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Hebrewes
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 2873
%%% o-2734
\startChapter[
  title={Chapter 12}
  ]

\Summary{By the foreſaid examples he exhorteth them to patience, and by
  example of Chriſt him ſelf crucified, 5.~and becauſe this diſcipline
  is an argument that they be God's children, 9.~with whoſe rod they
  should be much more content then with that of their carnal Fathers:
  and becauſe it bringeth iuſtification. 12.~Exhorting them therfore to
  plucke vp their harts, and to take faſter footing: 18.~conſidering
  that al being now ſo ſweet, and not terrible as in the old Teſtament,
  their damnation, if they refuſe to heare, wil be ſo much the greater.}

And therfore we alſo hauing ſo great a cloud of witneſſes put vpon vs:
\CNote{\XRef{Col.~3,~8.}
\XRef{1.~Pet.~2,~1.}}
laying away al weight and ſinne that compaſſeth vs, by patience let vs
run to the fight propoſed vnto vs, \V looking on the author of faith,
and the conſummatour \Sc{Iesvs}, who, ioy being propoſed vnto him,
ſuſtained the Croſſe, contemning confuſion, and ſitteth on the right
hand of the ſeat of God.

\V For, thinke diligently vpon him which ſuſtained of ſinners ſuch
contradiction againſt himſelf: that you be not wearied, fainting in your
minds. \V For you haue not yet reſiſted vnto bloud, repugning againſt
ſinne: \V and you haue forgotten the conſolation, which ſpeaketh to you,
as it were to children, ſaying,
\CNote{\XRef{Prou.~3,~11.}}
\Emph{My ſonne neglect not the diſcipline of our Lord: neither be thou
wearied whiles thou art rebuked of him.} \V
\CNote{\XRef{Apoc.~3,~19.}}
\Emph{For whom our Lord loueth, he chaſteneth, &
\LNote{He ſcourgeth.}{By
\MNote{Temporal punishment after remiſsiõ of ſinnes, either here or in
Purgatorie.}
this we proue that God often punisheth the ſinnes euen of his louing
children, though not with eternal damnation, yet with temporal
chaſtiſement and correction; & that he doth not alwaies together with
the remiſſion of deadly ſinnes and eternal punishment, exempt the
offender receiued to his grace, from al fatherly correction either in
this life or in the next. Neither haue the Heretikes of this time any
reaſon or Scripture in the world, why they should take away Gods
chaſtiſement of his children in the next life, more then in this world.}
he ſcourgeth euery child that he receiueth.}

\V Perſeuer ye in diſcipline. As vnto children doth God offer himſelf to
you. For what ſonne is there, whom the father doth not correct? \V But
if you be without diſcipline, whereof al be made partakers; then are you
baſtards, and not children. \V Moreouer the fathers indeed of our fleſh
we had for inſtructours, and we did reuerẽce them: ſhal we not much more
obey the Father of ſpirits, and liue? \V And they indeed for a time of
few daies, according to their wil inſtructed vs: but he, to that which
is profitable in receiuing of his ſanctification. \V And al diſcipline
for the preſent certes ſeemeth not to be of ioy, but of ſorrow: but
afterward it wil render to them that are exerciſed by it, moſt peaceable
fruit of iuſtice.

\V For the which cauſe ſtretch vp the ſlacked hands & the looſe knees \V
and make ſtraight ſteps to your feet: that no man halting erre, but
rather be healed. \V
\CNote{\XRef{Rom.~12,~18.}}
Follow peace with al men, and holineſſe: without which no man
%%% o-2735
ſhal ſee God: \V looking diligently leſt any man
\SNote{That we be not good, there is noe lacke on Gods part, who
offereth his grace to vs: but the defect is in our ſelues that are not
anſwerable to Gods calling of vs and grace towards vs.}
be wanting the grace of God: leſt any root of bitternes ſpringing vp doe
hinder and by it many be polluted. \V Leſt there be any fornicatour or
prophane perſon
\SNote{Such as forſake their ſaluation and religion to ſaue their lands
and goods are like Eſau.}
as Eſau:
\CNote{\XRef{Gen.~25,~33.}}
who for one diſh of meat ſold his firſt-birth-rights. \V For know ye
that afterward alſo deſiring to inherit the benediction, he was
reprobated:
\CNote{\XRef{Gen.~27,~38.}}
for
\LNote{He found not.}{It is not meant, that Eſau could not find
remiſsion of his ſinne at Gods hand: but that hauing once ſold and
yealded vp the right of his firſt-birth to his yonger brother, it was
too late to be ſorie for his vnaduiſed bargaine.}
he found not place of repentance, although with teares he had ſought it.

%%% 2874
\V For you are not come to
\CNote{\XRef{Exo.~19,~20.}}
a palpable mount, and an
\Var{acceſſible}{kindled (or) burning}
fire and whirle-wind, and darkenes, and ſtorme, \V and the ſound of
trompet, and voice of words, which they that heard excuſed themſelues,
that the word might not be ſpoken to them,
%%% !!! There is an opening '(' here in both, but no closing ')'
\V for they did not beare that which was ſaid:
\CNote{\XRef{Exo.~19,~12.}}
\Emph{And if a beaſt shal
touch the mount, it shal be ſtoned.} \V And ſo terrible was it which was
ſeen, Moyſes ſaid: \Emph{I am frighted and tremble.} \V But
\SNote{The faithful are made fellowes of Angels & of al the perfect
ſoules departed ſince the beginning of the world, and of Chriſt
himſelf.}
you are come to mount Sion, and the citie of the liuing God, heauenly
Hieruſalem, and the aſſemblie of many thouſand Angels, \V & the Church
of the firſt-borne, which are written in the Heauens, and the iudge of
al, God: and the ſpirits of the iuſt
\TNote{\G{τετελειωμένων}}
made perfect, \V and the mediatour of the new Teſtament \Sc{Iesvs}, and
the ſprinkling of bloud ſpeaking better then
\CNote{\XRef{Gen.~4,~10.}}
Abel.

\V See that you refuſe him not ſpeaking. For if they eſcaped not
refuſing him that ſpake vpon the earth, much more we, that turne away
from him ſpeaking to vs from Heauen. \V Whoſe voice moued the earth
then: but now he promiſeth, ſaying,
\CNote{\XRef{Ag.~2,~7.}}
\Emph{Yet once; and I wil moue not only the earth, but heauen alſo.} \V
And in that he ſaieth, \Emph{Yet once}, he declareth the tranſlation of
moueable things as being made, that thoſe things may remaine which are
vnmoueable. \V Therfore receiuing an vnmoueable Kingdom, we haue grace:
by the which
\TNote{\G{λατρεύωμεν}}
let vs ſerue pleaſing God, with feare & reuerence. \V For
\CNote{\XRef{Deut.~4,~24.}}
our God is a conſuming fire.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

