%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Hebrewes
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 2846
%%% o-2706
\startChapter[
  title={Chapter 04}
  ]

\Summary{That they muſt feare to be excluded out of the foreſaid reſt
  (which he proueth out of the pſalme) 12.~conſidering that Chriſt ſeeth
  their moſt inward ſecrets. 14.~And that he (as their Prieſt who alſo
  himſelf ſuffered) is able and ready to ſtrengthen them in confeſsion
  of their faith.}

%%% o-2707
Let vs feare therfore leſt perhaps forſaking the promiſe of entring into
his reſt, ſome of you be thought to be wanting. \V For to vs alſo it
hath been denounced, as alſo to them. But the word of hearing did not
profit them, not mixt with faith of thoſe things which they heard. \V
For we that haue beleeued, shal enter into the reſt: as he ſaid,
\CNote{\XRef{Pſ.~94,~11.}}
\Emph{As I ſware in my wrath, if they shal enter into my reſt}: & truely
the workes from the foundation of the world being perfited. \V For he
ſaid in a certaine place of the ſeuenth
\SNote{If the Apoſtle had not euidẽtly here ſhewed that the Sabboths
reſt was a figure of the eternal repoſe in heauen, who durſt to haue
applied that Scripture of Gods reſt the ſeuenth day, to that purpoſe? Or
how can our Aduerſaries now reprehend the like application manyfoldly
vſed in al holy Writers to the like end?}
day thus:
\CNote{\XRef{Gen.~2,~2.}}
\Emph{And God reſted the ſeuenth day from al his workes.} \V And againe
in this, \Emph{If they shal enter into my reſt.} \V Becauſe then it
remaineth that certaine enter into it, and they to whom firſt it was
preached, did not enter becauſe of incredulitie: \V againe he limiteth a
certaine day:
\CNote{\XRef{Heb.~3,~7.}}
\Emph{to day}, in Dauids ſaying, after ſo long time, as is aboue ſaid,
\Emph{to day if you shal heare his voice; doe not obdurate your
harts.} \V For if
\TNote{So Ioſue is called in Greeke.}
Ieſus had giuen them reſt: he would neuer ſpeake of another day
afterward. \V Therfore there is left a ſabbatiſme for the people of
God. \V For he that is entred into his reſt, the ſame alſo hath reſted
from his workes, as God from his.

\V Let vs haſten therfore to enter into that reſt: that no man fal into
the ſame example of incredulitie. \V For
\SNote{Whatſoeuer God threatneth by his word concerning the puniſhmẽt of
ſinne and incredulitie, ſhal be executed, be the offence neuer ſo
ſecret, deepe, or hidden in our harts becauſe Gods ſpeach paſſeth eaſily
and ſearcheth throughly euery part, power, and facultie of mans ſoul.}
the word of God is liuely and forcible, and more perſing then any
two-edged ſword: & reaching vnto the diuiſion of the ſoule and the
ſpirit, of the ioynts alſo and the marowes, and a diſcerner of the
cogitations and intents of the hart. \V And there is no creature
inuiſible in his ſight. But al things are naked and open to his eyes, to
whom our ſpeach is.

\V Hauing therfore a great high Prieſt that hath entred the heauens,
\Sc{Iesvs} the Sonne of God, let vs hold the confeſſion. \V For we haue
not a high Prieſt that can not haue compaſſion on our infirmities: but
tempted in al things by ſimilitude, except ſinne. \V
\LNote{Let vs goe with confidence.}{The
\MNote{Scripture abuſed againſt inuocatiõ of Saints.}
Aduerſaries goe about to proue by theſe words that we need no help of
Saints to obteine any thing, Chriſt himſelf being ſo readie, and we
being admoniſhed to come to him with confidence as to a moſt merciful
Mediatour and Biſhop. But by that argument they may as wel take away the
helps and praiers of the liuing one for another. And we doe not require
the help either of the Saints in heauen, or of our brethren in earth,
for any miſtruſt for God's mercie but of our owne vnworthines: being
aſſured that the praier of a iuſt man auaileth more with him, then the
deſire of a grieuous ſinner: and of a number making interceſsion
together, rather then of a man alone. Which the Heretikes can not deny
except they reproue the plaine Scriptures. Neither doe we come leſſe to
him, or with leſſe confidence, when we come accompanied with the praiers
of Angels, Saints, Prieſts, or iuſt men ioyning with vs, as they fondly
imagine and pretend: but with much more affiance in his grace, mercie,
and merits, then if we praied our ſelues alone.}
Let vs goe therfore with confidence to the throne of grace: that we may
obteine mercie and find grace in ſeaſonable aid.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

