%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Hebrewes
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 2844
%%% o-2704
\startChapter[
  title={Chapter 02}
  ]

\Summary{He inferreth of the foreſaid, that it shal be incomparably more
  damnable for them to neglect the new Teſtament then the old,
  3.~conſidering the irrefragable authoritie of the Apoſtles
  alſo. 5.~Then he proſecuteth the excellencie of Chriſt aboue the
  Angels, 9.~who neuertheles, was made leſſer then Angels, to ſuffer and
  die for men to deſtroy the dominion of the Diuel, 15.~to deliuer men
  from feare of death, 17.~and be a fit Prieſt for men.}

Therfore more aboundantly ought we to obſerue thoſe things which we haue
heard:
\SNote{As that which runneth out of a broken veſſel, or that runneth by,
is loſt.}
\TNote{\G{μήποτε παραρυῶμεν}}
leſt perhaps we runne out. \V For if the word that was ſpoken by Angels,
became ſure, and al preuarication and diſobedience hath receiued a iuſt
retribution of reward: \V how ſhal we
%%% o-2705
eſcape if we neglect ſo great ſaluation? which when it was begun to
be declared by our Lord, of them that heard was confirmed on vs, \V
\CNote{\XRef{Mar.~16,~20.}}
God withal teſtifying by ſignes, and
\Fix{vonders,}{vvonders,}{obvious typo, fixed in other}
and diuers miracles, and diſtributions of the Holy Ghoſt according to
his wil. \V For not to Angels hath God made ſubiect the world to come,
whereof we ſpeake. \V But one hath teſtified in a certaine place,
ſaying:
\CNote{\XRef{Pſ.~8,~5.}}
\Emph{What is man, that thou art mindful of him: or the ſonne of man,
that thou viſiteſt him? \V Thou didſt minish him litle leſſe then
Angels: with glorie and honour thou haſt crowned him, and conſtituted
him ouer the workes of thy hands.} \V
\CNote{\XRef{1.~Cor.~15.}
\XRef{Eph.~1.}}
\Emph{Al things haſt thou made ſubiect vnder his feet.} For in that he
ſubiected al things to him, he left nothing not ſubiect to him. But now
we ſee not as yet al things ſubiect to him. \V But
\CNote{\XRef{Philip.~2,~8.}}
him that was
%%% 2845
a litle leſſened vnder the Angels, we ſee \Sc{Iesvs},
\SNote{This proueth againſt the Caluiniſts that Chriſt by his Paſſion
merited his owne glorification: which they would not for ſhame deny of
Chriſt, 
but that they are at a point to deny al meritorious workes, yea euen in
Chriſt alſo. And therfore they trãſlate alſo this ſentence heretically,
by tranſpoſing the words.
\Cite{In the Bible printed the yeare~1579.}}
becauſe of the paſſion of death, crowned with glorie and honour: that
through the grace of God he might taſt death for al. \V For it became
him for whom al things, and by whom al things, that had brought many
children into glorie, to conſummate the Authour of their ſaluation, by
his paſſion. \V For he that ſanctifieth, and they that be ſanctified; al
of one. For the which cauſe he is not aſhamed to cal them Brethren, \V
ſaying:
\CNote{\XRef{Pſ.~21,~23.}}
\Emph{I wil declare thy name to my Brethren: in the middes of the
Church wil I praiſe thee.} \V And againe:
\CNote{\XRef{Pſ.~17,~3.}}
\Emph{I wil haue affiance in him.} And againe:
\CNote{\XRef{Eſ.~8,~18.}}
\Emph{Behold here am I & my children: whom God hath giuen me.} \V
Therfore becauſe the children haue communicated with fleſh and bloud,
himſelf alſo in like manner hath been partaker of the ſame: that
\CNote{\XRef{Oſe.~13,~14.}
\XRef{1.~Cor.~15,~54.}}
by death he might deſtroy him that had the empire of death, that is to
ſay, the Diuel: \V and might deliuer them that by the feare of death
through al their life were ſubiect to ſeruitude.

\V For no where doth he take Angels: but
\SNote{The dignitie of man, in that Chriſt tooke our nature vnto his
Perſon in Deitie, and not the nature of Angels.}
the ſeed of Abraham he taketh. \V Wherevpon he ought in al things to be
like vnto his brethren: that he might become a merciful and faithful
high Prieſt before God, that he might repropitiate the ſinnes of the
people. \V For in that wherein himſelf ſuffered and was tempted: he is
able to help them alſo that are tempted.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

