%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 1048
%%% o-0943
\startChapter[
  title={Chapter 13}
  ]

\Summary{Holofernes lying in a drunken ſleepe, Iudith cutteth of his
  head, 12.~ſo returneth with her maide through the gard, as it were to
  pray, cometh to Bethulia, 16.~exhorteth al to thanke God, and sheweth
  them the head. 22.~She is bleſſed of al. 27.~Achior alſo praiſeth God
  and Iudith.}

And when it was waxen late, his ſeruantes made haſt to their lodgings,
and Vagao shut the chamber doores, and went his way. \V And they were al
ouerladen with wine.
%%% 1049
\V And Iudith was alone in the chamber. \V Moreouer Holofernes lay in
his bed, faſt a ſlepe with very much
\SNote{The faſting of one woman ouerthrew an innumerable hoſt of
drunkards.
\Cite{S.~Amb. li. de ieiunio & Elia c.~9.}}
drunkennes. \V And Iudith ſaid to her mayd that ſhe ſhould ſtand without
before the chamber, and watch. \V And Iudith ſtood before the bed,
praying with teares, and with mouing of her lippes in ſilence, \V
ſaying: Confirme me ô Lord God of Iſrael,
%%% o-0944
and in this houre haue reſpect to the workes of my handes, that as thou
haſt promiſed, thou mayſt aduance
\SNote{If Bethulia had benne taken Ieruſalem alſo had bene in extreme
danger: for ſaftie wherof God had made more ſpecial promiſe to Dauid and
others.
\XRef{Iſaia.~37. 38.~&c.}}
Ieruſalem thy citie: and I may bring to paſſe that which I beleuing that
it may be done by thee, haue purpoſed. \V And when ſhe had ſayd theſe
thinges, she went to the piller, that was at his beds head, and his
ſword that hong tyed on it, she looſed. \V And when she had drawen it
out, she tooke him by the heare of his head, and ſayd: Confirme me ô
Lord God in this houre, \V and she ſtroke twiſe vpon his necke, and cut
of his head, and tooke his canopie from the pillers, and rolled aſide
his bodie a truncke. \V And after a while she went out, and deliuered
the head of Holofernes to her mayde, and bad her put it into her
wallet. \V And they two went forth, according to their cuſtome, as it
were to prayer, and they paſſed the campe, and compaſſing the valley,
they came to the gate of the citie. \V And Iudith a far of ſaid to
the keepers of the walles: Open the gates, becauſe God is with vs, which
hath wrought power in Iſrael. \V And it came to paſſe, when the men had
heard her voyce, they called the ancientes of the citie. \V And they ran
al to meete her, from the leaſt to the greateſt: becauſe they hoped not
that now she would come. \V And they lighting lightes gathered round
about her euerie one: and she going vp into a higher place, commanded
ſilence to be made. And when al had held their peace, \V Iudith ſaid:
Prayſe yee the Lord our God, who hath not forſaken them that hope in
him: \V and in me his handmayde he hath fulfilled his mercie, which he
promiſed to the houſe of Iſrael: and he hath killed by my hand the
enemie of his people this night. \V And bringing forth the head of
Holofernes out of the wallet, she shewed it them, ſaying: Loe the head
of Holofernes the general of the armie of the Aſſyrians, and behold his
canopie, wherein he lay in his drunkennes, where the Lord our God ſtroke
him by the hand of a woman. \V But the ſame our Lord liueth, that his
\SNote{Iudiths proper Angel ſo defended her, as Iacobs Angel deliuered
him from all euiles.
\XRef{Gen.~48.}}
Angel hath kept me, both going hence, and abyding there, and from thence
returning hither: and our
%%% 1050
Lord hath not ſuffered me his handmayde to be defyled, but without
pollution of ſinne he hath called me backe to you, reioycing in this
victorie, in my eſcape, and in your deliuerie. \V Confeſſe ye al to him,
becauſe he is good, becauſe his mercie is for euer. \V And they al
adoring our Lord, ſaid to 
her: Our Lord hath bleſſed thee in his power, becauſe by thee he hath
brought our enemies to nothing. \V Moreouer Ozias the prince of the
people of Iſrael, ſaid to her: Bleſſed art thou daughter of our Lord the
high God aboue al wemen vpon the earth. \V Bleſſed be our Lord, which
made heauen and earth, which hath directed thee vnto the woundes of the
head of the prince of our enemies. \V Becauſe this day he hath ſo
magnified thy name, that thy prayſe shal not depart out of the mouth of
men, which shal be mindeful of the power of our Lord for euer, for that
thou haſt not spared thy life for the diſtreſſes and tribulation of thy
kinred, but haſt holpen the ruine before the preſence of our God. \V And
al the people ſayd: So be it, ſo be it.
%%% o-0945
\V Moreouer Achior being called came, and Iudith ſaid to him: The God of
Iſrael, to whom thou gaueſt teſtimonie, that he reuengeth him ſelf of his
enemies, he hath cut of the head of al the vnfaithful this night by my
hand. \V And that thou mayſt proue that it is ſo, loe the head of
Holofernes, who in the contempt of his pride contemned the God of
Iſrael: and threatened thee death, ſaying: When the people of Iſrael
shal be taken, I wil command thy ſides to be pearſed with a ſword. \V
But Achior ſeing the head of Holofernes, being in anguish for feare, fel
on his face vpon the earth, and his ſoule was ſore trubled. \V But after
taking spirit agayne he was refreshed, & fel downe at her feete, and
adored her, and ſayd: \V Bleſſed art thou of thy God in euerie
tabernacle of Iacob, becauſe in euerie nation, which shal heare thy
name, the God of Iſrael shal be magnified in thee.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
