%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 1035
%%% o-0932
\startChapter[
  title={Chapter 5}
  ]

\Summary{Holofernes hearing that the Iſraelites prepare to reſiſt him,
  in great rage demandeth diuers particulars concerning that
  people. 5.~Achior an Ammonite telleth the meruelous workes of God
  towards them, 22.~aduiſeth him not to fight againſt them. 26.~The
  chief capitanies are offended, & threaten Achior.}

And it was told Holofernes the General of the warres of the Aſſyrians,
that the children of Iſrael prepared them ſelues to reſiſt, and had ſhut
vp the wayes of the mountaynes, \V and with exceding furie he chaſed in
great anger, and called al the princes of Noab and dukes of Ammon, \V
and he ſayd to them: Tel me
\SNote{Holofernes could not be altogether ignorant of ſo famous a people
as the Iewes, but in his rage marueling at their attempt demãdeth more
particularly of their habilitie to reſiſt him. See
\XRef{v.~27.}}
what is this people which beſetteth
%%% 1036
the mountaynes: or what, and of what ſorte, and how great their cities
are: alſo what their power is, or what is their multitude: or who is the
king of their warfare: \V and why aboue al that dwel in the Eaſt, haue
theſe contemned vs, and haue not come forth to meete vs, that they might
receiue vs with peace? \V Then Achior captaine of al the children of
Ammon anſwering, ſaid: If thou
\Fix{woutſafe}{voutſafe}{obvious typo, fixed in other}
my Lord to heare, I wil tel the truth in thy ſight of this people, which
dwelleth in the mountaynes, and there shal not a falſe word come forth
of my mouth. \V This people is of the progenie of the Chaldees. \V
\CNote{\XRef{Gen.~11.}}
The ſame dwelt firſt in Meſopotamia, becauſe they would not folow the
goddes of their fathers, which were in the land of the Chaldees. \V
Forſaking therfore the ceremonies of their fathers, which were in
multitude of goddes, \V they worshipped one God of heauen, who alſo
commanded them that they should depart from thence, and should dwel in
Charan.
\CNote{\XRef{Gen.~12.}}
And when there was famine ouer al the land, they went downe into Ægypt,
and there for foure hundred yeares were ſo multiplied, that the hoſt of
them could not be numbred. \V
\CNote{\XRef{Exod.~1.}}
And when the king of Ægypt oppreſſed them, and in the buildinges of his
cities had ſubdewed them in brick and claye, they cried to their Lord,
and he ſtroke the whole Land of Ægypt with diuers plagues. \V
\CNote{\XRef{Exo.~7. &c.}}
And when the Ægyptians had caſt them out from them, and the plague
%%% o-0933
had ceaſed from them, and they would take them againe, and cal them
backe to their ſeruice, \V theſe fleeing away, the God of heauen opened
the ſea, ſo that the waters were
\CNote{\XRef{Exo.~14.}}
\TNote{Made ſolid or firme.}
conſolidated as a wal, on either ſide, and they walking through the
botome of the ſea paſſed drie foote. \V In which place whiles an
innumerable armie of the Ægyptians purſewed them, they were ſo
ouerwhelmed with the waters, that there was not one remayning, to tel
the fact to poſteritie. \V Alſo being paſt the read ſea, they poſſeſſed
the deſertes of Mount Sinai, in which neuer man could dwel, or ſonne of
man reſted. \V
\CNote{\XRef{Exod.~15. 16.~17.}}
There bitter fountaynes were made ſweete for them to drinke, and for
fourtie yeares they receiued meate from heauen. \V Whereſoeuer they
entered without bow and arrow, and without shielde and ſword, their God
fought for them, and ouercame. \V
\CNote{\XRef{Iudic.~2. 3.~4.~&c.}}
And there was not that did inſult againſt this people, but when they
departed from the worſhip of the Lord their God. \V
\CNote{\XRef{Nu.~21.}}
But as often as
%%% 1037
beſide their owne God, they worshiped an other, they were geuen to
praye, and into the ſword, and to reproch. \V And as often as they were
penitent for that they reuolted from the worship of their God, the God
of heauen gaue them power to reſiſt. \V
\CNote{\XRef{Ioſ.~12.}}
Finally the king of the Chananeites, and of Iebuſeites, and of the
Pherezeites, and of the Hetheites, and of the Heueites, and of the
Amorrheites, and al the mightie in Heſebon they ouerthrew, and they
poſſeſſed their landes, & their cities: \V And as long as they ſinned
not in the ſight of their God, it was wel with them, for their God
hateth iniquitie. \V
\CNote{\XRef{4.~Reg.~21.}}
For
\SNote{He ſemeth to ſpeake of the captiuitie of king Manaſſes.
\XRef{Paral.~33.}
See the
\XRef{argument of this booke}.}
theſe yeares alſo paſt when they had reuolted from the way, which God
had geuen them, that they ſhould walke in it, they were deſtroyed in
battels by manie nations, and verie manie of them were led captiue into
a ſtrange land. \V But of late returning to the Lord their God, from
the diſperſion wherein they were diſperſed, they are vnited & are come
vp into al theſe mountaynes, and poſſeſſe Ieruſalem againe where their
Holies are. \V Now therfore my Lord, ſearch if there be any iniquitie of
theirs in the ſight of their God: and let vs goe vp to them, becauſe
their God deliuering wil deliuer them to thee, and they ſhal be ſubdewed
vnder the yoke of thy power: \V but if there be no offence of this
people before their God, we can not reſiſt them, becauſe their God wil
defend them: and we ſhal be a reproche to the whole earth. \V And it
came to paſſe, when Achior had ceaſed to ſpeake theſe wordes, al the
great men of Holofernes were angrie, and they thought to kil him, ſaying
to each other: \V
\SNote{As Holofernes
\XRef{v.~3.}
demanded that which he in great part knew, ſo his men, (knowing Achior
wel ynough) in rage demanded \Emph{vvho} he is that dare ſay ſuch thinges?}
Who is this, that ſayth the children of Iſrael can reſiſt king
Nabuchodonoſor, and his armies, men vnarmed, and without force, and
without skil of the feates of warre? \V That Achior therfore may know
that he deceiueth vs, let vs goe vp into the mountaynes: and when their
mightie ones ſhal be taken, then ſhal he with them be ſtrooken through
with the ſword: \V that euerie nation may know that Nabuchodonoſor is
god of the earth, and beſides him there is none other.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
