%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 1041
%%% o-0937
\startChapter[
  title={Chapter 8}
  ]

\Summary{Iudith
\MNote{The third part.

Iudith deliuereth her whole nation from perſecution.}
a moſt vertuous, rich, fayre, & renowmed widow, 9.~rebuketh the high
prieſt and ancientes, for their determination to deliuer the citie, if
ayde come not in fiue dayes, 14.~exhorteth al to penance. 28.~They al
agree to her godlie aduiſe, 30.~praying for good ſuceſſe of her
intention, which they yet know not.}

And it came to paſſe, when Iudith a widow had heard theſe wordes, which
was the daughter of Merari, the ſonne of Idox, the ſonne of Ioſeph, the
ſonne of Ozias, the ſonne of Elai, the ſonne of Iamnor, the ſonne of
Gedeon, the ſonne of Raphaim, the ſonne of Achitob, the ſonne of
Melchias, the ſonne of Enan, the ſonne of Nathanias, the ſonne of
Salathiel, the ſonne of Simeon, the ſonne of Iſrael: \V and her huſband
was Manaſſes, who died in the daies of barley harueſt: \V for he was
occupied with them that bound sheues in the field, and the heate came
vpon his head, and he died in Bethulia his citie, and was buried there
with his father. \V And Iudith was leaft his widow now three yeares and
ſix monethes. \V And in the higher partes of her houſe she made her ſelf
\SNote{In three ſpecial obſeruances this holie widow led a moſt
religious life: In much priuate prayer,}
a ſecrete chamber, in which she abode shut vp with her maides, \V and
hauing
\SNote{in wearing hearecloth,}
cloth of heare vpon her loynes, she
\SNote{and in much faſting: as it were a perpetual lent, \Emph{al dayes
but Sabbathes}, & feaſtes excepted.}
faſted al the daies of her life but Sabbathes, and
\TNote{The firſt day of the moneth.}
new
%%% 1042
moones, and the feaſtes of the houſe of Iſrael. \V And she was of an
exceding beautiful countenance, to whom her husband had leaft much
richeſſe, and a great familie, and poſſeſſions ful of heardes of oxen,
and flockes of sheepe. \V And she was among al moſt famous, becauſe she
feared our Lord very much, neither was there that spake an il word of
her. \V When she therfore had heard that Ozias had promiſed, that after
the fifth day were paſt he would yeld the citie, she ſent to the
ancientes Chabri and Charmi. \V And they came to her, and she ſaid
to them: What is this word, wherein Ozias hath conſented, to yeld the
citie to the Aſſyrians, if withing fiue daies there come no ayde to
vs? \V And what are you, that tempt our Lord? \V This is no word that
may prouoke mercie, but rather that may rayſe vp wrath, and inflame
furie. \V You
\SNote{In this eſpecially they offended, that they prefixed a time,
meaning then to yeld the citie, for they
\Fix{ſhuld}{ſhould}{obvious typo, fixed in other}
rather haue fought againſt the enemie, then deliuer Gods people to a
tyranne, holie thinges into prophane handes, & by their example put
Ieruſalem and the temple in like danger.
\Cite{Rabanus}.}
haue ſet a time for the mercie of our Lord, and according to your
pleaſure, you haue appointed him a day. \V But becauſe our Lord is
patient, let vs be penitent for this ſame thing, and sheding teares let
vs deſire his pardon:
%%% o-0938
\V for not as man, ſo wil God threaten, neither as the ſonne of man wil
he be inflamed to anger. \V And therfore let vs humble our ſoules to
him, and being ſetled in an humble ſpirit, ſeruing him: \V let vs ſay
weeping to our Lord, that according to his wil ſo he doe his mercie with
vs: that as our hart is trubled in their pride, ſo alſo we may glorie in
our humilitie: \V becauſe we haue not folowed the ſinnes of our fathers,
which forſooke their God, and adored ſtrange goddes, \V for which
abomination they were geuen into the ſword, and into confuſion to their
enemies: but we know no other God but him. \V Let vs humbly expect his
conſolation, and he wil require our bloud of the afflictions of our
enemies, and he wil humble al Nations, what ſoeuer ſhal riſe vp againſt
vs, and the Lord our God wil make them without honour. \V And now
bretheren, becauſe you are ancientes in the people of God, and their
ſoules depende of you: by your ſpeach comforte their hartes, that they
be mindful, that our fathers were tempted to be proued, whether they did
worſhippe their God truly. \V They muſt be mindful how our father
Abraham was tempted, and by many tribulations proued, was made the
freind of God. \V So Iſaac, ſo Iacob, ſo Moyſes, & al that haue pleaſed
God, through manie tribulations haue paſſed faithful. \V But they that
did not receiue the tentations with the feare
%%% 1043
of our Lord, and vttered their impatience and reproch of their murmuring
againſt our Lord, \V were deſtroyed of
\SNote{S.~Paul
\XRef{1.~Cor.~10. v.~10.}
alleaging that happened to the people of Iſrael in the deſert, addeth
this word \Emph{deſtroyer}, which is only in this place, and not
\XRef{Num.~11.}
nor
\XRef{14.}
nor els in the old Teſtamẽt. Which is an other argumẽt that this
\Fix{boke}{booke}{obvious typo, fixed in other}
is Canonical.}
the deſtroyer, and periſhed by ſerpents. \V And we therfore let vs not
reuenge ourſelues for theſe thinges, which we ſuffer, \V but reputing
theſe verie puniſhments to be the ſcourges of our Lord leſſe then our
ſinnes, wherwith as ſeruantes we are chaſtiſed, let vs thinke them to
haue chanced to our amendement, and not to our deſtruction. \V And Ozias
and the ancientes ſayd to her: Al thinges which thou haſt spoken, be
true, and there is no reprehenſion in thy wordes. \V Now therfore pray
for vs, becauſe thou art a holie woman, and fearing God. \V And Iudith
ſaid to them: As you know, that which I could ſpeake to be of God: \V ſo
that which I haue diſpoſed to doe, proue if it be of God, and pray that
God eſtabliſh my counſel. \V You ſhal ſtand at the gate this night, and
I wil goe out with myne
\SNote{Some thinke Abra was a proper name, but whether it was or no, it
ſignifieth an handmaide of more honour as we ſpeake: \Emph{A maide of
honour}.}
abra: and pray ye, that as you haue ſaid, in fiue dayes our Lord reſpect
his people Iſrael. \V But I wil not that you ſearch my doing, and til I
bring you word, let nothing els be done, but pray for me to our Lord
God. \V And Ozias the prince of Iuda ſaid to her: Goe in peace, and our
Lord be with thee to the reuenge of our enemies. And returning they
departed.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
