%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 1047
%%% o-0942
\startChapter[
  title={Chapter 12}
  ]

\Summary{Iudith is brought into Holofernes treaſurehouſe, 2.~hath leaue
  to abſtaine from the Gentils meates, 5.~and to goe forth in the night
  to pray. 10.~The fourth day she cometh to Holofernes banket, 16.~he is
  taken with concupiſcence, and drinketh very much wine.}

Then he bad her goe in where his treaſures were layd vp, and bad her
tarie there, and he appoynted what should be geuen her of his owne
banket. \V To whom Iudith anſwered, and ſaid: Now,
\SNote{See
\XRef{chap.~10. v.~5.}}
I can not eate of theſe thinges, which thou commandeſt to be geuen me,
leſt there come diſpleaſure vpon me: but I wil eate of theſe thinges,
which I haue brought. \V To whom Holofernes ſaid: If theſe thinges which
thou haſt brought with thee shal fayle thee, what shal we doe to
thee. \V And Iudith ſaid: Thy ſoule liueth my lord, that thy handmayde
shal not ſpend al theſe thinges, til God doe by my hand theſe thinges,
which I haue purpoſed. And his ſeruantes brought her into the
tabernacle, which he had commanded. \V And whiles she went in, she
deſired that
%%% 1048
ſhe might haue licence to goe forth, in the night and before
%%% o-0943
day to prayer, and to beſech our Lord. \V And he commanded his
chamberlaynes, that as it pleaſed her, she should goe out & come in to
adore her God for three dayes. \V And she went forth in the nightes into
the vale of Bethulia, and washed her ſelf in a fountaine of water. \V
And as she came vp, she prayed our Lord the God of Iſrael, that he would
direct her way to the deliuerie of his people. \V And going in, she
remayned
\SNote{This puritie conſiſted in abſtaining from meates ſorbid by the
Law, or offered to Idoles.}
pure in the tabernacle, vntil she tooke her owne meate in the
euening. \V And it came to paſſe in the fourth day, Holofernes made a
ſupper to his ſeruantes, and ſayd to Vagao his eunuch: Goe, and perſwade
that Hebrew woman, that ſhe conſent of her owne accord to dwel with
me. \V For it is a foule thing with the Aſſyrians, if a woman mocke a
man in doing, that ſhe paſſe free from him. \V Then Vagao went in to
Iudith, & ſaid: Let not the good
\TNote{\L{Puella}.}
yong maide feare to goe in to my Lord, that she may be honoured before
his face, that she may eate with him and drinke wine in ioyfulnes. \V To
whom Iudith anſwered: Who am I, that I should gayneſay my lord? \V Al
that ſhal be good and beſt before his eies, wil I doe. And whatſoeuer
shal pleaſe him, that shal be beſt to me al the dayes of my life. \V And
she aroſe, and decked herſelf with her garmentes, and going in she ſtood
before his face. \V And the hart of Holofernes was ſtrooken: for he
burnt in the concupiſcence of her. \V And Holofernes ſaid to her: Drinke
now, and ſit downe in ioyfulnes, becauſe thou haſt found grace before
me. \V And Iudith ſaid: I wil drinke my
lord, becauſe my ſoule is magnified this day aboue al my dayes. \V And
she tooke, and did eate, and dranke before him thoſe thinges, which her
handmayde had prepared her. \V And Holofernes was made pleaſant toward
her, and dranke wine exceeding much, ſo much as he had neuer drunke in
his life?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
