%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 1039
%%% o-0935
\startChapter[
  title={Chapter 7}
  ]

\Summary{Holofernes beſiegeth Bethulia, 6.~cutteth their conduite of
  water, 9.~kepeth their fountaines. 12.~The people murmure, and
  mutenie, 18.~yet they pray to God, 23.~and the high prieſt
  determineth, if aide come not within fiue dayes, to deliuer the citie
  to the Aſſyrians.}

But Holofernes the next day commanded his armies, that they should goe
vp againſt Bethulia. \V And there were of warre foote men an hundred
twentie thouſand, and horſe men two and twentie thouſand, beſides the
preparations of thoſe men, whom the captiuitie had taken, and had beene
led away out of the prouinces and cities, of al youth. \V Al prepared
them ſelues together to the fight agaynſt the children of Iſrael, and
they came by the hil ſide vnto the toppe, which looketh toward Dorhaim,
from the place which is called Belma vnto Chelmon, which is againſt
Eſdrelon. \V But
\SNote{Gods ſeruantes firſt of al before they fight, do workes of
penãce, cõmend their cauſe to God, then take weapons in hand, & expect
oppurtinitie to encounter with the enimies.}
the children of Iſrael, as they ſaw the multitude of them, laid them
ſelues proſtrate vpon the earth, caſting ashes vpon
%%% 1040
their heades, praying with one accord, that the God of Iſrael
%%% o-0936
would ſhew his mercie vpon his people. \V And taking their weapons of
warre, they ſate at the places, which by a narrow path way lead directly
betwen the mountaynes, and they were keping them al day and night. \V
Moreouer Holofernes, whiles he compaſſeth round about, found that the
foũtayne, which ranne in, went directly to their conduite on the ſouth
ſide without the citie: and he commanded their conduite to be cut
aſunder. \V Neuertheleſſe there were fountaynes not far from the walles,
out of which ſecretly they ſemed to draw water to refreſh them ſelues,
rather then to drinke. \V But the children of Ammon, & Moab came to
Holofernes, ſaying: The children of Iſrael truſt not in ſpeare, nor in
arrow, but the mountaynes defend them, & the hilles ſtanding meruelous
ſtipe gard them. \V Therfore that thou mayſt ouercome them without
ioyning battel, ſet kepers of the fountaynes, that they may not draw
water of them, and without ſword thou ſhalt kil them, or at the leaſt
being wearied they wil yeld their citie, which they thinck being ſet in
the mountaynes, can not be ouercome. \V And theſe wordes pleaſed
Holofernes, and his ſouldiars, and he placed round about an hundred men
at euerie fountayne. \V And when that watch had beene fully kept for
twentie dayes, the ceſternes fayled, and the collections of waters to al
the inhabitantes of Bethulia, ſo that there was not within the citie,
wherof they might be ſatisfied no not for one day, becauſe water was
dayly geuen to the people by meaſure. \V Then al the men, and wemen,
yong men, and children, being gathered together to
\SNote{A common frailtie of people in diſtreſſe to impute error to their
Superiour: though in deede Ozias erred rather in yelding at laſt to
their ſuggeſtion
\XRef{v.~25.}
then in withſtanding thus long.}
Ozias, al together with one voice, \V ſaid: God iudge betwen vs and
thee, becauſe thou haſt done euil agaynſt vs, in that thou wouldeſt not
ſpeake peaceably with the Aſſyrians, and for this cauſe God hath ſold vs
into their handes. \V And there is none to helpe, wheras we lye
proſtrate before their eies in thirſt, & great deſtruction. \V And now
aſſemble ye al, that are in the citie, that we may of our owne accord
yeld vs al to the people of Holofernes. \V For it is better, that
captiues we bleſſe our Lord liuing, then we ſhould die, and be a reproch
to al fleſh, when we ſhal ſee our wiues, and our infantes die before our
eies. \V We cal to witnes this day heauen and earth, and the God of our
fathers, which taketh vengeance of vs according to our ſinnes, that now
you deliuer
%%% 1041
the citie into the hand of Holofernes armie, that our end may be ſhort
in the edge of the ſword, which is made longer in the
\Fix{drught}{drought}{obvious typo, fixed in other}
of thirſt. \V And when they had ſaid theſe thinges, there was made great
weeping and howling of al in the aſſemble, and for manie houres with one
voice they cried to God, ſaying: \V We haue ſinned with our fathers, we
haue done vniuſtly, we haue committed iniquitie. \V Thou becauſe thou
art gracious, haue mercie vpon vs, or in thy ſcourge reuenge our
iniquities, and deliuer not them that truſt in thee to a people, that
knoweth not thee, \V that they ſay not among the Gentiles: Where is
their God? \V And when they wearied with theſe cries, and tyred with
theſe weepings, had held their peace, \V Ozias
%%% o-0937
ryſing vp embrued with teares, ſaid: Be of good chere bretheren, and
theſe fiue daies let vs expect mercie of our Lord. \V For peraduenture
he wil cut of his indignation, and wil geue glorie to his name. \V But
if after fiue daies be paſt there come no aide, we wil doe theſe wordes,
which you haue spoken.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
