%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 1037
%%% o-0934
\startChapter[
  title={Chapter 6}
  ]

\Summary{Holofernes in great rage ſendeth Achior to Bethulia, that he
  may there be ſlaine with the Iſraelites. 8.~He is leift bound to a
  tree, 10.~from whence the Iſraelites taking him, he telleth them the
  cauſe. 14.~They entertaine him courteouſly and earneſtly pray to God
  for helpe.}

%%% 1038
And it came to paſſe when they had ceaſed to ſpeake, Holofernes being
ſore offended, ſaid to Achior: \V Becauſe thou haſt prophecied vnto vs
ſaying, that the nation of Iſrael is defended of their God, that I may
\Fix{ſheu}{ſhew}{obvious typo, fixed in other}
thee
\SNote{Nabuchodonoſors vaſſals would haue no other god but him. And God
almightie is alſo ielous and wil haue no God but him ſelfe.}
that there is no god, but Nabuchodonoſor: \V when we ſhal haue ſtrookẽ
them al as one man, then thy ſelf with them ſhalt die by the ſword of
the Aſſyrians, and al Iſrael with thee ſhal periſh by deſtruction: \V
and thou ſhalt proue that Nabuchodonoſor is lord of the whole earth: and
then the ſword of my warfare ſhal paſſe through thy ſides, & pearſed
thou ſhalt fal among the wounded of Iſrael, and thou ſhalt no more fetch
breath, til thou be deſtroyed with them. \V But if thou thinke thy
prophecie true, let not thy countenance quaile, and the paleneſſe that
is in thy face, let it depart from thee, if thou thinke theſe my wordes
can not be accompliſhed. \V And that thou mayſt know that thou shalt
proue theſe thinges together with them, behold from this houre thou shalt
be aſſociate to their people, that whiles they shal receiue worthie
punishment of my ſword, thou withal may be ſubiect to the vengeance. \V
Then Holofernes commanded his ſeruantes that they should take Achior,
and lead him into Bethula, and should deliuer him into the handes of the
children of Iſrael. \V And the ſeruantes of Holofernes taking him, went
through the champaine: but when they came neere the mountaynes, the
ſlingers came forth againſt them. \V And they turning out of the way by
the ſide of the mountayne, tyed Achior to a tree hand and foote, and ſo
left him bound with withes, and returned to their lord. \V Moreouer the
children of Iſrael deſcending from Bethulia, came to him. Whom
\Fix{looſnig}{looſing}{obvous typo, fixed in other}
they brought to Bethulia, and ſetting him in the middes
of the people, demanded what was the matter, that the Aſſyrians had left
him bound. \V In thoſe dayes the princes there, were Ozias the ſonne of
Micha of the tribe of Simeon, and Charmi, who alſo is Gothoniel. \V In
the middes therefore of the ancientes, and in the ſight of al, Achior
told al thinges that he had ſpoken being aſked of Holofernes: and how
the people of Holofernes would haue killed him for this word, \V and how
Holofernes him ſelfe being angrie cõmanded him to be deliuered for this
cauſe to the Iſraelites: that when he should ouercome the children of
Iſrael, then he
%%% o-0935
might command Achior alſo himſelf to die by diuerſe tormentes,
%%% 1039
for this that he had ſaid: The God of heauen is their defender. \V And
when Achior had declared al theſe thinges, al the people fel on their
face, adoring our Lord, and with common lamentation and weeping they
powred out their prayers with one accord to our Lord, \V ſaying: Lord
God of heauen and earth, behold
\SNote{God regardeth the pride of ſinners,}
their pride, and haue regard to
\SNote{the humiliation of penitentes,}
our humilitie, and attend
\SNote{the ſanctitie of his Church in general, as wel holie perſons, as
holie thinges,}
the face of thy ſainctes, and ſhew that
\SNote{and true cõfidence in him, without preſumĩg of mens poure.}
thou forſakeſt not them that preſume of thee and, that
\Fix{thow}{thou}{likely typo, same in both}
humbleſt them that preſume of them ſelues, and glorie of their power. \V
Their weeping therfore being ended, & the peoples prayer by the whole
day being finiſhed, they comforted Achior, \V ſaying: The God of our
fathers, whoſe power thou haſt ſet forth, he wil geue thee this
recompence, that thou rather ſhalt ſee their deſtruction. \V And when
the Lord our God shal geue this libertie to his ſeruantes, be God with
thee alſo in the middes of vs: that as it shal pleaſe thee, ſo thou with
al thine mayſt conuerſe with vs. \V Then Ozias, the counſel being ended,
receiued him into his houſe, and made him a greate ſupper. \V And al the
ancientes being called, they refreshed them ſelues together after the
faſting was ended. \V But afterwarde al the people was called together,
and al the night long within the church, they prayed deſiring helpe of
the God of Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
