%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 1052
%%% o-0946
\startChapter[
  title={Chapter 15}
  ]

\Summary{The Aſſyrians flee for feare, 3.~the Hebrewes purſue them,
  7.~and are enriched by the spoyles. 9.~The high prieſt with others
  come from Ieruſalem and prayſe Iudith. 14.~The goodes of Holofernes
  are geuen to her.}

And when al the armie heard that Holofernes was beheaded, courage and
counſel fel from them, and being shaken with trembling onlie and feare,
they helpe them ſelues by flight, \V ſo that none spake with his
neighbour, but hanging the head, al thinges leaft behind, they made haſt
to eſcape the Hebrewes, whom they heard to come armed vpon them, fleing
by the waies of the fieldes, and the pathes of the hilles. \V The
children of Iſrael therfore ſeing them fleing, folowed them. And they
went downe ſounding with trumpettes, and shouting after them. \V And
becauſe the Aſſyrians were not vnited together, they went headlong into
flight: but the children of Iſrael purſewing in one companie,
diſcomfited al that they could find. \V Ozias therfore ſent meſſengers
through al the cities and countries of Iſrael. \V Euerie countrie
therfore, and euerie citie ſent choſen youth armed after them, and
they purſewed them in the edge of the ſword, vntil they came to the
extremetie of their borders. \V And the reſt that were in Bethulia,
entered into the campe of the Aſſyrians, and tooke away the praye, which
the Aſſyrians fleing had leaft, and were loden excedingly. \V But they
that were returned conquerers to Bethulia, tooke away with them
%%% o-0947
al thinges that were theirs, ſo that there was no number in cattel, and
beaſtes, and al their moueables, that from the leaſt vnto the greateſt
al were made rich of their prayes.

\V
\MNote{The fourt part.

The prayſes of Iudith, who with al the people praiſe God.}
And Ioachim the high Prieſt came from Ieruſalem into Bethulia with al
his ancientes to ſee Iudith. \V Who when ſhe was come out to him, they
al bleſſed her with one voyce, ſaying:
\SNote{Iudith was a ſpecial figure of the B.~Virgine Marie, to whom
theſe praiſes perteyne in more eminent ſorte, then to anie other
creature.
\Cite{S. Iulbertus Carnotenſis.}}
Thou art the glorie of Ieruſalem, thou the ioy of Iſrael, thou the
honour of our people: \V becauſe thou haſt done manfully, and thy hart
was ſtrengthened, for that thou haſt loued chaſtitie, and after thy
husband not knowne any other: therfore alſo the hand of our Lord hath
ſtrengthened thee, and therfore shalt thou be bleſſed for euer. \V And
al the people ſayd: So be it, ſo be it. \V And for thirtie daies, ſcarſe
were the ſpoyles of the Aſſyrians gathered of the people of Iſrael. \V
Moreouer al thinges, that were proued to be the
%%% 1053
peculiar goodes of Holofernes, they gaue to Iudith in gold, and ſiluer,
and garmentes, and pretious ſtones, and al ſtuffe, and al the thinges were
deliuered her of the people. \V And al the people reioyſed with wemen,
and virgins, and yongmen, on inſtrumentes and harpes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
