%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 1050
%%% o-0945
\startChapter[
  title={Chapter 14}
  ]

\Summary{Holofernes head is hanged on the wall. 6.~Achior is
  circumciſed. 7.~The Iſraelites aſſault the Aſſyrians, 8.~who going
  to awake their General, 14.~finde him ſlaine, 17.~and are al
  confounded with feare.}

And Iudith ſaid to al the people: Heare me brethren, hang ye this head
vpon our walles: \V and it shal be, when the ſunne shal riſe, let euerie
man take his armour, and yſſue ye forth with violence, not that you goe
downe
%%% 1051
beneath, but as it were inuading violently. \V Then the watchmen muſt of
neceſſitie runne to rayſe vp their prince to battel. \V And when the
captaynes of them shal runne to the tabernacle of Holofernes, and shal
finde him headles rowled in bloud, feare wil fal vpon them. \V And when
you shal know that they flee, goe after them ſecurely, becauſe our Lord
wil deſtroy them vnder your feete. \V Then
\SNote{Achior an Ammonite
\XRef{(ch.~5. v.~5.)}
being truly conuerted and beleeuing in God, was admitted into the
Church, notwithſtanding that Ammonites and Moabites were excluded by
name,
\XRef{Deut.~23. v.~3.}
which is therfore ment only of thoſe that perſiſt impenitent and
obſtinate.}
Achior ſeeing the power, that God of Iſrael wrought, forſaking the rite
of gentilitie, beleued God, and
\Fix{circuncided}{circumciſed}{obvious typo, fixed in other}
the flesh of his prepuce, and was ioyned to the people of Iſrael, and al
the ſucceſſion of his kinred vntil this preſent day. \V And immediatly
as day brake, they hong the head of Holofernes vpon the walles, and
euerie man tooke his armour, and they went forth with great noyſe and
shouting. \V Which the watchmen ſeing, ranne to the tabernacle of
Holofernes. \V Moreouer they that were in the tabernacle, coming, and
before the dore of the tabernacle making a noyſe, to rayſe him, they
endeuored by art to diſquiet him, that Holofernes might awake not by
them rayſing him, but by them making a noyſe. \V For no man durſt by
knocking, or entring, to open the chamber of the chiefe of the
Aſſyrians. \V But when his dukes and tribunes were come, and al the
chiefe of the armie of the
%%% o-0946
king of the Aſſyrians, they ſaid to the chamberlayns: \V Goe in, and
awake him, becauſe the mice yſſuing out of their holes, haue preſumed to
prouoke vs to battel. \V Then Vagao entring into his chamber, ſtoode
before the cortine, and made a clapping with his handes: for he thought
that he ſlept with Iudith. \V But when with the ſenſe of his eares he
percieued no motion of perſon lying, he came neere to the cortine, and
lifting it vp, and ſeing the bodie without the head of Holofernes
weltred in his bloud lye vpon the ground, cried out in a lowd voyce with
weeping, and rent his garmentes. \V And going into the tabernacle of
Iudith, he found her not, and he lept forth to the people, \V and ſayd:
One Hebrewe woman hath made confuſion in the houſe of king
Nabuchodonoſor: for behold Holofernes lyeth vpon the ground, and his
head is not vpon him. \V Which when the princes of the power of the
Aſſyrians had heard, they al rent their garmentes, and intollerable
feare and dread fel vpon them, and their mindes were trubled
excedingly. \V And there was made an incomparable crie in the middes of
their campe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
