%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 1053
%%% o-0947
\startChapter[
  title={Chapter 16}
  ]

\Summary{Iudith ſingeth a canticle of thanksgeuing to our Lord. 22.~The
  people goe to Ieruſalem and offer ſacrifices. 25.~She in great honour
  liueth a widow til her death in good old age. 30.~The people haue long
  peace, and a feaſtiual day is inſtituted in perpetual memorie of her
  fact.}

Then ſang Iudith this ſong to our Lord, ſaying:

%%% !!! I follow other's paragraph breaks for this canticle.
\V Beginne ye to our Lord in timbrels, ſing ye to our Lord in cymbals,
tune to him a new pſalme, reioyſe, and inuocate his name.

\V
\SNote{S.~Ephrem
\Cite{ſer. de 2.~Aduentu,}
citeth this place as holie ſcripture ſo teſtifying this booke to be
canonical.}
Our Lord confoundeth battels, Lord is his name.

\V Who hath ſet his campe in the middes of his people, that he might
deliuer vs from the hand of al our enemies.

\V Aſſur came out of the mountaynes from the North in the multitude of
his ſtrength: whoſe multitude ſtopped vp the torrentes, and their horſes
couered the valles.

\V He ſaid that he would ſet my borders on fyre, and kil my yongmen with
the ſword, to geue my infantes into praye, and virgins into
captiuitie.

\V But our Lord omnipotent hath hurt him, and hath deliuered
him into the hands of a woman, and hath pearſed him.

\V For, not by yongmen is their mightie one fallen, neither haue the
ſonnes of
\SNote{Such giantes as were before Noes floud,
\XRef{Gen.~6.}}
Titan ſtrooken him, neither did the high
\SNote{Nor ſuch as were after,
\XRef{Num.~13.}
\XRef{Deu.~3.}}
giantes ſet them ſelues vpon him, but Iudith the daughter of Merari in
the beautie of her face diſſolued him.

%%% o-0948
\V For ſhe put from her the garmentes of widowhood, and put on her the
garmentes of ioy, in the reioyſing of the children of Iſrael.

\V She anoynted her face with oyntment, and tyed together her lockes
with a crowne, ſhe tooke a new ſtole to deceiue him.

\V Her ſandals rauiſhed his eies, her beautie made his ſoule captiue,
ſhe with a ſword cut of his head.

\V The Perſians did quake at her conſtancie, and the Medes at her
boldneſſe.

\V Then did the campe of the Aſſyrians howle, when my humble ones
appeared, withering in thirſt.

\V The ſonnes of yongwemen haue pearſed them, and they haue killed them
as boyes fleing away: they haue periſhed in battel before the face of
the Lord my God.

\V Let vs ſing an hymne to our Lord, let vs ſing a new hymne to our
%%% 1054
God.

\V Adonai ô Lord great art thou, and noble in thy power, and whom no man
can ouercome.

\V Let euerie creature of thyne ſerue thee: becauſe thou ſaydſt, and
they were made: thou didſt ſend thy ſpirit, and they were created, and
there is none that can reſiſt thy voyce.

\V The mountaynes with the waters ſhal be moued with the foundations:
the rockes ſhal melt as waxe before thy face.

\V But they that feare thee, ſhal be great with thee in al thinges.

\V Wo be to the nation, that ryſeth vp vpon my kinred: for our Lord
omnipotent wil be reuenged on them, in the day of iudgement he wil
viſite them.

\V For he wil geue
\SNote{Euerlaſting torments of fire & wormes perteyne to the damned
bodies,
\Cite{S.~Aug. li.~21. c.~4. &c. ciuit.}
and great paines to the damned ſoules, eſpecially the loſſe of Gods
viſion.}
fyre, and wormes into their fleſh, that they may be burnt, and may feele
for euer.

\V And it came to paſſe after theſe thinges, al the people after the
victorie came into Ieruſalem to adore our Lord: and forthwith as they
were purified, they al offered holocauſtes, and vowes, and their
promiſes. \V Moreouer Iudith offered for an anathema of obliuion al the
inſtrumentes of warre of Holofernes, which the people gaue her, and the
canopie that her ſelf had taken away out of his chamber. \V And the
people was pleaſant according to the face of ſainctes, and for three
monethes the ioy of this victorie was celebrated with Iudith. \V And after
thoſe daies euerie man returned into his houſe, and Iudith was made
great in Bethulia, and ſhe was more glorious to al the land of
Iſrael. \V There was alſo chaſtitie ioyned to her vertue,
\SNote{In the Greke, v.~22. \Emph{Manie deſired her to mariage}.}
ſo that ſhe knew not man al the daies of her life,
\LNote{After that her huſband vvas dead.}{As
\MNote{Iudith an example of holie widowhood.}
yong Tobias and Sara were notable patterns to maried perſons: ſo Iudith
is a like good example to deuout widowes, excelling moſt part in manie
reſpectes.
\MNote{Manie incitementes concurring to the contrarie made her widowhood
more excellent.}
For firſt ſhe profeſſed this holie ſtate of life in the old
Teſtament, when it was moſt rare,
\CNote{\XRef{Deut.~25. v.~5.}
\XRef{Ruth.~3. v.~12.}}
the law prouiding that the brother, or next kinſman, ſhould marie the
widow of him, that died without children, as it ſemeth she had none, the
Greke text affirming that ſhe gaue al her goodes before her death to
other kinred.
\XRef{ch.~16. v.~24.}
Secondly, ſhe was only once maried,
\XRef{ch.~15. v.~13.}
\XRef{ch.~16. v.~26.}
wheras, it is alſo commendable after twiſe or oftener mariage at laſt to
abſtaine. Thirdly, ſhe was yong, about 36.~yeares: for three yeares and
a half after that her husband was dead, ſhe was called a yong maide.
\XRef{ch.~12. v.~12.}
Fourtly, ſhe was of excellent beautie.
\XRef{ch.~8. v.~7.}
Fiftly exceeding rich.
\XRef{ibidem.}
\Fix{Siztly,}{Sixtly,}{obvious typo, fixed in other}
very noble, eſpecially after the deliuerie of the people from
ſuch diſtreſſe.
\XRef{ch.~15. v.~10.}
Seuently, for this renowmed fact, and for her other great vertue
\XRef{(ch.~8. v.~8.)}
manie principal men deſired to marie her.
\XRef{ch.~16. v.~22.}
Eightly, al the people wished much iſſue of ſo noble a ſtock.
\XRef{ch.~16. v.~25.}
Nintly, she liued long in the ſtate of widowhood, about threeſcore and
nine yeares, from 36.~to~105.
\XRef{ch.~16. v.~28.}
Tently, there was great and long peace in al Iſrael, after that she had
releeued Bethulia.
\XRef{ch.~16. v.~30.}
Al which might eaſely haue inuited an other to haue maried:
\MNote{Wodowhood an Euangelical counſell.}
but her great deuotion, and feruent deſire to ſerue God in a retired
auſtere life, faſting & praying,
\XRef{ch.~8. v.~6.}
cutte of al incitements to mariage, and made her before the Goſpel, to
embrace Euangelical counſel, not commanded, but for better attaining to
perfection counſeled by our Sauior and S.~Paul.
\XRef{Mat.~19.}
\XRef{1.~Cor.~7.}}
after that Manaſſes her huſband was dead. \V And on feſtiual daies ſhe
came forth with great glorie. \V And ſhe abode in her husbands houſe,
\SNote{Liuing in al 105.~yeares she was widow about ſixtie nine: for
when she ſlew Holofernes she was about fourtie yeares old,
\XRef{ch.~10. v.~18.}
and her huſband was then dead three yeares and a halfe before.
\XRef{ch.~8. v.~4.}}
an hundred fiue yeares, and diſmiſt her abra free, and ſhe died and was
buried with her huſband in Bethulia. \V And al the people mourned for
her ſeuen dayes. \V And in al the space of her life there was not that
trubled Iſrael, and after her death
%%% o-0949
manie yeares. \V But the day of the
feſtiuitie of this victorie is receiued of the Hebrewes in the number of
holie daies, and is worſhiped of the Iewes from that time vntil this
preſent day.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
