%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 1044
%%% o-0940
\startChapter[
  title={Chapter 10}
  ]

\Summary{Iudith excellently adorned, and very beautiful goeth with her
  handmaide towards the campe, 11.~is taken by the watch, 16.~and
  brought to Holofernes.}

And it came to paſſe, when she had ceaſed to crie to our Lord, she roſe
from the place, wherein she lay proſtrate to our Lord. \V And she called
her
\SNote{See
\XRef{ch.~8. v.~32.}}
abra, and going downe into her houſe she tooke from her the hearecloth,
and put of the garmentes of her widowhood, \V and washed her bodie, and
anoynted her ſelf with ointement, and plaited the heare of her head, and
put a crowne vpon her head, and clothed her
%%% 1045
ſelf with the garmentes of her ioyfulnes, and put
\TNote{\L{Sandalia}.}
pantoſles on her feete, and tooke braceletes, and Lilies, and eareletes,
and ringes, and with al her ornamentes she adorned her ſelf. \V To whom
alſo
\SNote{God by this teſtified her holie intention in adorning herſelfe.}
our Lord gaue beautie: becauſe al this trimming did not depend of
ſenſualitie, but of vertue: & therfore our Lord amplified this beautie
on her, that she might appeare to al mens eies of incomparable
comlines. \V She therfore layd vpon her abra
\SNote{Not only certaine meates were forbid by the law
\XRef{Leuit.~11.}
but in maner al the meates of the Gentiles were vnlawful being offered
to idoles, for which cauſe Daniel and the three children would not eate
of the kings meates in Babylon.
\XRef{Dan.~1.}}
a bottel of wyne, and a veſſel of oyle, and
%%%!!!  \L{Polentam} in other
\TNote{Stiped barley.}
polent, and drie figges, and bread, and cheeſe, and went forward. \V And
when they were come to the gate of the citie, they found Ozias
expecting, and the ancientes of the citie. \V Who when they ſaw her,
being aſtonished maruailed at her beautie, exceedingly. \V Yet asking her
no queſtion, they let her paſſe, ſaying: The God of our fathers geue
thee grace, and ſtrengthen al the counſel of thy hart, with his powre,
that Ieruſalem may glorie vpon thee, and thy name may be in the number
of the holie and iuſt. \V And they that were there ſaid, al with one
voyce: So be it, ſo be it. \V But Iudith praying our Lord, paſſed
through the gate she and her Abra. \V And it came to paſſe, when she
went downe the hil, about breake of day the watchmen of the Aſſyrians
mette her, and held her, ſaying: Whence comeſt thou? or whither goeſt
thou? \V Who anſwered: I am a daughter of the Hebrews, therfore am I fled
from their face,
\SNote{She told manie things euidently true, other things conditionally,
as it was like to come to paſſe, if God ſent not better meanes.}
becauſe I knew it should come to paſſe, that they should be geuen you
vnto spoyle, becauſe that contemning you, they would not of their owne
accord yeld themſelues, that they might find mercie in your ſight. \V
For this cauſe I thought with my ſelf, ſaying: I wil goe to the preſence
of the prince Holofernes, that I may tel him their ſecretes, and shew
him by what entrance he may winne them, ſo that there shal not fal one
man of his armie. \V And when thoſe men had heard her wordes, they
conſidered
%%% o-0941
her face, and she was a wonder in their eies, for they marueiled at her
beautie excedingly. \V And they ſaid to her: Thou haſt ſaued thy life,
in that thou haſt found ſuch counſel, that thou wouldeſt come downe to
our lord. \V And this know thou, that when thou shalt ſtand in his
ſight, he wil deale wel with thee, and thou shalt be moſt gratious in
his hart. And they brought her to the tabernacle of Holofernes, telling
him of her. \V And when she was entered before his face, forthwith
Holofernes was caught in his eies. \V And his
%%% 1046
guard ſaid to him: Who can contemne the people of the Hebrewes, which
haue
\SNote{S.~Fulgentius probably iudgeth that Iudith was now about fourtie
yeares old.
\Cite{Epiſt.~2. cap.~6.}}
ſo beautiful wemẽ, that we should not worthely fight againſt them for
theſe? \V Iudith therfore ſeeing Holofernes ſitting in a canopie, which
was wouen of purple, and gold, and emerauld, and pretious ſtones: \V and
when she had looked in his face, she adored him, falling proſtrate vpon
the ground. And the ſeruantes of Holofernes lifted her vp, their lord
commanding it.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
