%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 1034
%%% o-0931
\startChapter[
  title={Chapter 4}
  ]

\Summary{The
\MNote{The ſecond part.

Gods people are in great feare and diſtreſſe.}
children of Iſrael excedingly fearing Holofernes forces, 3.~prouide to
reſiſt him, by the exhortation of the High Prieſt, vſing both humaine,
8.~and diuine meanes.}

Then the children of Iſrael, which dwelt in the Land of Iuda, hearing
theſe thinges, were ſore afrayd of his preſence. \V Trembling alſo, and
horrour inuaded their ſenſes, leſt he would doe that to Ieruſalem and to
the temple of our Lord, which he had done to other cities, and their
temples. \V And they ſent into al Samaria round about, as
\Fix{for}{far}{obvious typo, fixed in other}
as Iericho, and preoccupated al the toppes of mountaynes: \V and they
compaſſed their townes with walles, and gathered together corne for
prouiſion of battel. \V Eliachim
\SNote{The high prieſt managed alſo the temporal affayres of the
cõmonwelth, at this time, by conſent & commiſſion of king Manaſſes, as
is moſt probable theſe thinges happening ſhortly after his repentãce.
\XRef{2.~Paral.~33.}}
the prieſt wrote to al, that were againſt Eſdrelon, which is againſt the
face of the great fielde beſide Dothain, and to al, by whom there might
be paſſage of way, that they should take the aſcentes
%%% 1035
of the mountaynes, by which there might be way to Ieruſalem, and should
keepe watch, where the way was narrow betwen the mountaynes. \V And the
children of Iſrael did according as the Prieſt of our Lord, Eliachim had
appointed them. \V And al the people cried to our Lord with great
inſtance, and they humbled their ſoules in faſtings, and prayers, they
& their wiues. \V And the Prieſtes put on heareclothes, and they laide
the infantes proſtrate againſt the face of the temple of our Lord, and
the altar of our Lord they couered with hearecloth: \V and they cried to
our Lord the God of Iſrael with one accord, that their infantes might
not be geuẽ into praye, & their wiues into ſpoile, and their cities into
deſtruction, and their holie thinges into pollution, and they be made a
reproch to the Gentiles. \V Then Eliachim the high Prieſt of our Lord
went about al Iſrael and ſpake to them, \V ſaying: Know ye, that our
Lord wil heare your prayers, if continewing you continew in faſtinges
and prayers in the ſight of our Lord. \V Be mindful of Moyſes the
ſeruant of our Lord,
\CNote{\XRef{Exo.~17.}}
who ouerthrew Amalec truſting in his power, and in his might, and in his
armie, & in his shieldes, & in his chariotes, & in his horſemen, not by
fighting, but with holie
%%% o-0932
prayers: \V ſo shal al the enemies of Iſrael be; if you perſeuere in
this worke, which you haue begunne. \V They therfore at this exhortation
of his, beſeching our Lord, continewed in the ſight of our Lord, \V ſo
that they alſo, which offered the holocauſtes to our
\Fix{Lod,}{Lord,}{obvious typo, fixed in other}
did offer the ſacrifices to our Lord girded with heareclothes, and there
were ashes vpon their head. \V And they al prayed God with al their hart,
that he would viſite his people Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
