%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 1032
%%% o-0929
\startChapter[
  title={Chapter 2}
  ]

\Summary{Nabuchodonoſor ſendeth Holofernes his General to waiſt al
  countries of the weſt, 7.~with a great armie, and aboundant
  munition. 11.~They ſubdue manie places, and others are ſtrooken with
  great feare.}

In the thirtenth yeare of king Nabuchodonoſor, the two and twenteth day
of the firſt moneth, the word was geuen out in the houſe of
Nabuchodonoſor the king of the Aſſyrians, that he would reuenge him
ſelfe. \V And he called al the ancientes, and al the captaynes, and his
men of warre, and communicated with them the ſecrete of his counſel: \V
and he ſaid that his cogitation was vpon that, to ſubdew al the earth to
his empire. \V Which ſaying when it had pleaſed them al, Nabuchodonoſor
the king called Holofernes the General of his warres, \V and ſaid to
him: Goe forth againſt euerie kingdom of the weſt, & againſt them
eſpecially, that contemned my commandment. \V
\SNote{This cruel cõmandment was as cruelly put in execution by
Holofernes.
\XRef{ch.~3. v.~11.}}
Thyne eie shal ſpare no kingdom, and euerie fenſed citie thou shalt
ſubdew to me. \V Then Holofernes called the captaynes, & magiſtrates of
the powre of the Aſſyrians: and he muſtered men for the expedition, as
the king commanded him, an hundred twentie thouſand fighting
%%% 1033
men on foote, and twelue thouſand archers horſemen. \V And he made al
his expedition to goe before in a multitude of inumerable camels, with
thoſe thinges that might ſuffice the armies abundantly, heardes of oxen
alſo, and flockes of sheepe, which had no number. \V He appoynted corne
to be prepared out of al Syria in his paſſage. \V But gold and ſiluer he
tooke out of the kings houſe exceding much. \V And he went forth and al
his armie with the chariotes, & horſemen, and the archers, which couered
the face of the earth, as locuſtes. \V And when he had paſſed through
the coaſtes of the Aſſyrians, he came to the great mountaines of Ange,
which are on the left hand of Cilicia: and he went vp into al their
caſteles, and wonne euerie fortreſſe. \V And he brake downe the renowmed
citie of Melothus, and ſpoyled al the children of Therſis, and the
children of Iſmael, which were againſt the face of the deſert, and
%%% o-0930
on the ſouth of the land of Cellon. \V And he paſſed ouer Euphrates, and
came into Meſopotamia: and he brake al the high cities, that were there,
from the torrent of Membre, til ye come to the ſea: \V and he tooke the
borders therof, from Cilicia vnto the coaſtes of Iapheth, which are
toward the ſouth. \V And he caried away al the children of Madian, and
ſpoyled al their riches, and al that reſiſted him he ſlew in the edge of
the ſword. \V And after theſe thinges he went downe into the fieldes of
Damaſcus in the daies of harueſt, and he ſet al the corne on fire, and
he made al the trees and vineyardes to be cut downe; \V and the feare of
him fel vpon al the inhabitantes of the land.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
