%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 1033
%%% o-0930
\startChapter[
  title={Chapter 3}
  ]

\Summary{Manie kinges and other princes ſubmitte them ſelues to
  Holofernes. 8.~He receiueth them, and taketh of their chief men to
  reinforce his armie, 11.~neuertheles deſtroyeth their cities, and
  their goddes, that Nabuchodonoſor only might be called god.}

Then the kinges and princes of al cities and prouinces, namely of Syria
and Meſopotamia, and Syria Sobal, and Lybia, and Cilicia ſent their
embaſſadours, which coming to Holofernes, ſaid: \V Let thy indignation
towarde vs ceaſe: For it is better that liuing we feare Nabuchodonoſor
the great king, and be ſubiect to thee, then dying, we should with our
deſtruction ſuffer the damages of our ſeruitude. \V Euerie citie of
ours, and al our poſſeſſion, al mountaynes,
%%% 1034
and hilles, and fieldes, and heardes of oxen, and flockes of sheepe, and
goates, and of horſes, and camels, and al our goddes, and families are
in thy ſight: \V let al our thinges be vnder thy law. \V We alſo, and
our children are thy ſeruantes. \V Come to vs a peaceable Lord, and vſe
our ſeruice, as it shal pleaſe thee. \V Then went he downe from the
mountaynes with horſemen in a great powre, and tooke euerie citie, and
euerie inhabiter of the land. \V And of al the cities he tooke to helpe
him valiant men, and choſen for battel. \V And ſo great feare lay vpon
al thoſe prouinces, that the inhabitantes of al cities, princes and
honorable perſons, together with the people went out to meete him
coming, \V receyuing him with garlandes, and torches, dauncing with
timbrels, & shaulmes. \V Neither doing theſe thinges, could they for al
that mitigate the firceneſſe of his ſtomacke: \V for he did both deſtroy
their cities, and cut downe their groues. \V For Nabuchodonoſor the king
had commanded him, that he should deſtroy
%%% o-0931
al the goddes of the earth, that
\SNote{An expreſſe figure of Antichriſt
\XRef{2.~Theſ.~2.}
for whom al heretikes make way as precurſers, for the ſingular
\Emph{man of ſinne} wil confeſſe no God but himſelf.}
he only might be called god of thoſe nations, which could be ſubdewed
with the might of Holofernes. \V And paſſing through al Syria Sobal, and
al Apamea, & al Meſopotamia he came to the Idumeians into the land of
Gabaa, \V and tooke their cities, and ſate there for thirtie dayes, in
which daies he commanded al the armie of his powre to be vnited.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
