%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 1043
%%% o-0939
\startChapter[
  title={Chapter 9}
  ]

\Summary{Iudith in hearecloth & aſhes prayeth for the deliuerie of the
  people, remembering like benefites of God, 12.~craueth powre to
  ouerthrow Holofernes.}

VVho going away, Iudith
\SNote{Of ſuch priuate oratories as this our Sauiour ſemeth to ſpeake
\XRef{Mat.~6.}
ſaying: \Emph{Enter into thy chamber &c.} Of like oratories among
Chriſtians read
\Cite{Card. Baronius An. D.~293.}
\Cite{Cyp. ep.~56.}}
went into her oratorie: and clothing her ſelf with hearecloth, put aſhes
vpon her head: and falling proſtrate to our Lord, she cried to our Lord,
ſaying: \V Lord God of
\Fix{may}{my}{obvious typo, fixed in other}
father Simeon, who gaueſt him a ſword for defence againſt ſtrangers,
which were rauishers in their coinquination, and diſcouered the virgine
vnto confuſion: \V and gaueſt their wemen into praye, and their
daughters into captiuitie: and al the praye to be diuided to thy
ſeruantes, which were zelous in
\SNote{The zele of Simeon and Leui was commendable, but for diuers euil
circumſtances the fact was reproued by Iacob.
\XRef{Gen.~34.}
&
\XRef{49.}}
thy zele: helpe I
\Fix{be-beſech}{beſech}{obvious typo, fixed in other}
thee ô Lord God me a widow. \V For thou haſt done the thinges
heretofore, and haſt purpoſed one thing after an other: and that hath
bene done which thou wouldeſt. \V For al thy wayes are readie, and thy
iudgementes thou haſt put in thy prouidence. \V Looke vpon the campe of
the Aſſyrians now, as thou didſt voutſafe to ſee the campe of the
Ægyptians,
%%% 1044
\CNote{\XRef{Exo.~14.}}
when they ran armed after thy ſeruantes, truſting in chariotes, and in
their horſemen, and in a multitude of men of warre. \V But thou didſt
looke ouer their campe, and darkeneſſe wearied them. \V The depth held
their feete, and the waters ouerwhelmed them. \V So let theſe alſo be
made ô Lord, which truſt in their multitude, and in their chariotes, and
in pikes, and in shieldes, and in their arrowes, and glorie in their
ſpeares, \V and know not thy ſelf art our God, which deſtroyeſt warres from
the beginning, and Lord is thy name. \V Lift vp thyne arme as from the
beginning, and daſh their power in thy power: let their power fal in thy
wrath, which promiſe that they wil violate thy holie thinges, and
pollute the tabernacle of thy holie name, & with their ſword throw downe
the horne of thyne altar. \V Bring to paſſe Lord, that his pride be cut
of with his owne ſword, \V
\SNote{She prayed, that Holofernes ſhould be moued with her beautie, &
ſwete ſpeach: which he might be and not ſinne, but he by his free wil
ſinning God turned his ſinne to the good of others. See
\XRef{Annot. Exod.~7.}
\Cite{S.~Aug. Ser.~228.}}
let him be caught with the ſnares of his eies in me, and thou shalt
ſtrike him from the lippes of my charitie. \V Geue me conſtancie in
mind, that I may contemne him: and powre that I may ouerthrow him. \V
For this shal be a memorial of thy name, when the hand of a woman shal
ouerthrow him. \V For not in multitude is thy power ô Lord, neither in
ſtrength of horſes is thy wil, neither haue the proud pleaſed thee from
the beginninge: but the prayer of the humble and meeke hath alwayes
pleaſed thee. \V God of the heauens, creatour of the waters, and Lord of
euerie creature, heare me wretch, praying and preſuming of thy
mercie. \V Remember Lord thy teſtament, and geue a
%%% o-0940
word in my mouth, and ſtrengthen counſel in my hart, that thy houſe may
continew in thy ſanctification: \V and al Nations may acknowledge, that
thou art God, and there is no other beſides thee.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
