%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 1031
%%% o-0928
\startChapter[
  title={Chapter 1}
  ]

\Summary{Nabuchodonoſor
\MNote{The firſt part.

The occaſion of the Iewes perſecution at this time.}
king of Aſsyrians ouercometh Arphaxad king of the Medes: 7.~ſummoneth
manie other nations to ſubmitte themſelues to his Empyre: 11.~which they
refuſing he threatneth reuenge.}

Arphaxad
\SNote{The author being reſolued to write this hiſtorie, ioyneth his
narration to his internal purpoſe, ſaying: \Emph{Arphaxad therfore &c.}
\Cite{S.~Greg. ho.~2, in Ezech.}}
therfore king of the Medes had ſubdued manie nations to his empire, & he
built a moſt mightie citie, which he called Ecbatanis, \V of ſtone
ſquared and hewed: he made walles therof in height ſeuentie cubites, and
in breadth thirtie cubites, and the towers therof he made in height an
hundred cubites. \V But each ſide of them was in foure ſquare twentie
foote long, and he made the gates therof according
%%% 1032
to the height of the towers: \V and he
\SNote{Thou foole this night they ſhal take thy life from thee, & theſe
things whoſe ſhal they be?
\XRef{Luc.~12. v.~20.}}
gloried as mightie in the force of his armie, and in the glorie of his
chariotes. \V In the twelfth yeare of his reigne: Nabuchodonoſor the
king of the Aſſyrians, who reigned in Niniue the greate citie, fought
againſt Arphaxad, and ouercame him \V in the great field, which is
called Ragau, about Euphrates, and Tigris, and Iadaſon in the field of
Erioch the king of the Elicians. \V Then was the kingdom of
Nabuchodonoſor exalted, and his hart was eleuated: and he ſent to al,
that dwelt in Cilicia and Damaſcus, and Libanus, \V and to the nations,
that are in Carmelus, and Cedar, and the inhabitantes of Galilee in the
great field of Eſdrelon, \V and to al that were in Samaria, and beyond
the riuer Iordan euen to Ieruſalem, and al the land of Ieſſe til you
come to the
%%% o-0929
borders of Æthiopia. \V To al theſe Nabuchodonoſor king of the Aſſyrians
ſent meſſengers: \V who al with one minde ſaid nay, & ſent them backe
emptie, and reiected them without honour. \V Then Nabuchodonoſor the
king taking indignation againſt al that land, ſwore by his throne and
kingdom that he would
\TNote{\L{defenderet ſe}.}
reuenge him ſelfe of al thoſe countries.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
