%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Judith
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 1046
%%% o-0941
\startChapter[
  title={Chapter 11}
  ]

\Summary{Holofernes demanding Iudith the cauſe of her coming: 4.~she
  deceiueth him with a probable narration.}

Then Holofernes ſaid to her: Be of good chere, and feare not in thy
hart: becauſe I haue neuer hurt man, that would ſerue Nabuchodonoſor the
king. \V And thy people, if they had not contemned me, I would neuer
haue lifted vp my ſpeare ouer them. \V But now tel me, for what cauſe
haſt thou departed from them, and it hath pleaſed thee to come to vs? \V
And Iudith ſaid to him:
\SNote{Al that Iudith ſayth is true in her ſenſe as she meant it, but
not in the vnderſtanding of Holofernes and his ſouldiars, yet was no
leſſe lawful then that Ioſue did in deceuing the citizenes of Hay.
\XRef{Ioſue.~8.}
\Cite{S.~Aug. q.~10. in Ioſue.}}
Take the wordes of thy handmayde, for if thou wilt folow the wordes of
thy handmayde, our Lord wil do with thee a perfect thing. \V For
Nabuchodonoſor the king of the earth liueth, and his power liueth which
is in thee to the chaſtiſing of al ſtraying ſoules: that not onlie men
ſerue him by thee, but alſo the beaſtes of the field obey him. \V For
the induſtrie of thy mind is reported to al nations, and it is declared
to al the world, that thou onlie art good, and mightie in al his
kingdom, and thy diſcipline is bruted to al prouinces. \V Neither is that
vnknowne, which Achior ſpake, neither are we ignorant of that thou haſt
commanded to come vpon him. \V For it is certaine that our God is ſo
offended with ſinnes, that he hath ſent word by his prophetes to the
people, that he wil deliuer them for their ſinnes. \V And becauſe the
children of Iſrael know they offended their God, thy dread is vpon
them. \V Moreouer alſo famine hath inuaded them, and for drought of
water they are now eſtemed among the dead. \V Finally they ordayne this,
to kil their cattel, and to drinke the bloud of them, \V and the
%%% o-0942
holie thinges of our Lord their God which God commanded not to be
touched, in corne, wine, and oile, theſe haue they purpoſed to beſtow,
and they wil conſume the thinges which they ought not to touch with
their handes: therfore
%%% 1047
becauſe they do theſe thinges, it is ſure that they ſhal be geuen into
perdition. \V Which I thy handmayde knowing, am fled from them, and our
Lord hath ſent me to tel thee theſe verie thinges. \V For I thy
handmayde worſhippe God, euen now being with thee, and thy handmayde wil
goe forth, and I wil pray God, \V and he wil tel me when he wil repay
them their ſinnes, and I coming wil tel thee, ſo that I may bring thee
through the middes of Ieruſalem, and thou ſhalt haue al the people of
Iſrael, as ſheepe, that haue no paſtor, and there ſhal not ſo much as
one dog barke againſt thee: \V becauſe theſe thinges are told me by the
prouidence of God. \V And becauſe God is angrie with them, I am ſent to
tel theſe verie thinges to thee. \V And al theſe wordes pleaſed
Holofernes, and his ſeruantes, and they maruailed at her wiſedom, and
one ſaid to an other: \V There is not ſuch a woman vpon the earth in
looke, in beautie, and in ſenſe of wordes. \V And Holofernes ſayd to
her: God hath done wel, which ſent thee before the people, that thou
mighteſt geue them into our handes: \V and becauſe thy promiſe is good,
if thy God shal doe this for me, he shal alſo be my God, and thou shalt be
great in the houſe of Nabuchodonoſor, and thy name shal be renowmed in
al the earth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
