%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Canticle
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 1475
%%% o-1362
\startChapter[
  title={Chapter 07}
  ]

\Summary{}


VVhat
\SNote{Chriſt interpoſeth his commendation of the Ievves, vvho at laſt
shal returne to him vvith great feruoure of faith and deuotion.}
shalt thou ſee in the Sulamiteſſe but the companies of campes? How
beautiful are thy paſes in shoes, ô princes daughter!
\SNote{And ſo iointly praiſeth his Church conſiſting of both peoples.}
the ioyntes of thy thighes are as iewels, that are made by the hand of
the artificer. \V Thy nauel as a round bowle, neuer wanting cuppes. Thy
bellie as an heape of wheate, compaſſed about with lilies. \V Thy two
breaſts, as two fawnes the twinnes of a roe. \V Thy necke as a towre of
yuorie. Thine eies as the fishpooles in Heſebon, which are in the gate
of the daughter of the
%%% 1476
multitude. Thy noſe as the towre of Libanus, that looketh againſt
Damaſcus. \V Thy head as Carmelus: and the heares of thy head as a kings
purple tyed to cundite pipes. \V How beautiful art thou, and how comely
my deareſt, in delightes! \V Thy ſtature is like to a palmetree, & thy
breaſtes to cluſters of grapes. \V I ſayd: I wil goe vp into the
palmetree, and wil take hold of the fruites therof: and thy breaſts shal
be as the cluſters of a vineyard: and the odour of thy mouth as it were
of apples. \V Thy throate as the beſt wine,
\SNote{The Church, as it vvere taking the vvord out of Chriſts mouth
vvhiles he praiſed her, she returneth al the praiſe to him.}
worthie for my beloued to drinke, & for his lippes and his teeth to
ruminate. \V I to my beloued, and his turning is toward me. \V
\SNote{Praying him to come and ſtil remaine vvith her,}
Come my beloued, let vs goe forth in to the filde, let vs abide in the
villages. \V Let vs riſe earely to the vineyards, let vs ſee if the
vineyard florishe, if the flowers be readie to bring forth fruites, if
the pomegranates florish: there wil I giue thee my breaſts. \V The
Mandragoraes haue geuen a ſmel. In our gates al fruites:
\SNote{acknovvleging him to be the only Sauiour of both old and nevv
teſtament.}
the new and the old, my beloued, I haue kept for thee.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
