%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Canticle
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 1470
%%% o-1357
\startChapter[
  title={Chapter 01}
  ]

\Summary{}


Let
\SNote{The Church of the old teſtament deſireth Chriſts coming in fleſh:
and the Chriſtian Church prayeth for his coming in glorie.}
him kiſſe me with the kiſſe of his mouth: becauſe thy breſtes are better
then wine, \V ſmelling fragrantly of the beſt ointments. Oile powred out
is thy name: therfore haue yongmaydes loued thee. \V Draw me: we wil
runne after thee in the odour of thine ointments. The king hath brought
me into his cellars: we wil reioyce & be glad in thee, mindful of thy
breſts aboue wine: the righteous loue thee. \V
\SNote{The Church outwardly afflicted, is inwardly fayre.}
I am blacke but
\Fix{beutiful,}{beautiful,}{possible typo, fixed in other}
ô ye daughters of Ieruſalem, as the tabernacles of Cedar, as the skinnes
of Salomon. \V Doe not conſider me that I am browne, becauſe the ſunne
hath altered my colour: the ſonnes of my mother haue fought againſt me,
they haue made me a
%%% 1471
keeper in the vinyards: my vinyard I haue not kept. \V Shew me ô thou,
whom my ſoule loueth, where thou feedeſt, where thou lyeſt in the
midday, leſt I beginne to wander after the flockes of thy companyons. \V
\SNote{Chriſt encorageth his ſpouſe the Church.}
If thou know not thyſelfe, ô moſt fayreſt among wemen, goe forth, and
folow after the ſteppes of the flockes, and feede thy kiddes byſide the
tabernacles of the paſtours. \V To my companie of horſemen, in the
chariotes of Pharao, haue I likened thee, ô my loue. \V Thy cheekes are
beautiful as the turtledoues, thy necke as iewels. \V We wil make thee
cheynes of gold, enamoled with ſiluer. \V
\SNote{She meditateth of his Paſſion, and Reſurrection.}
Whiles the king was at his repoſe, my ſpikenard gaue the odour
thereof. \V A bundle of myrrhe my beloued is to me, he shal abide betwen
my breſtes. \V A cluſtre of cypre my loue is to me, in the vineyardes of
Engaddi. \V
\SNote{Chriſt praiſeth his ſpouſe.}
Behold thou art fayre, ô my loue, behold thou art fayre, thyne eyes are
as of doues. \V
\SNote{She againe praiſeth him,}
Behold thou art fayre my beloued, & comlie:
\SNote{vvith thankes for her repoſe, and preſent conſolation.}
our litle bed is florishing. \V The beames of our houſes are of cedar,
our rafters of cypreſſe trees.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
