%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Canticle
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 1476
%%% o-1362
\startChapter[
  title={Chapter 08}
  ]

\Summary{}


VVho
\SNote{The Synagogue proſecuteth her prayer, deſiring Chriſts
Incarnation.}
shal giue to me thee my brother, ſucking the breaſts of my mother, that
I may finde thee without, and kiſſe thee, and now no man deſpiſe me? \V I
wil take hold of thee, and wil bring thee into my mothers houſe: there
thou shalt teach me, and I wil giue thee a cuppe of ſpiced wine, and new
wine of my pomegranats. \V His left hand vnder my head, and his right
hand shal embrace me. \V
\SNote{Chriſt admonisheth vvorldlie men not to moleſt thoſe that ſerue
him in contemplation, & other ſpiritual vertues.}
I adiure you ô daughters of Ieruſalem, that you rayſe not vp, nor make
the beloued to awake til herſelfe wil. \V
\SNote{Angels and other Sainctes of the triumphant Church admire the
beautie of the Gentiles conuerted; which is alſo vnderſtood of euerie
holie ſoule aſcending from this world into heauen. And more ſingularly of
the moſt glorious virgin mother of God.}
Who is this, that cometh vp from the deſert, flowing with delightes,
leaning vpon her beloued? Vnder the appletree I rayſed thee vp:
\SNote{The Synagogue of the Iewes was corrupt vnder the tree of Chriſts
Croſſe, when they cried: Crucifie him, Crucifie him. And againe: His
bloud be vpon vs, and vpon our children. Againe: VVe haue no King but
Cæſar. &c.}
there thy mother was corrupted, there she was defloured that bare
thee. \V Put me as a ſeale vpon thy hart, as a ſeale vpon thyne arme:
becauſe loue is ſtrong as death: ielouſie is hard as hel, the lampes
thereof lampes
%%% o-1363
of fyre and flames. \V Manie waters can not quench charitie, neither
shal floudes ouerwhelme it: if a man shal giue al the ſubſtance of his
houſe for loue, as nothing he shal deſpiſe it. \V
\SNote{Chriſt againe ſheweth his affection towards his Church of the
Gentiles: calling her his owne ſiſter, and the Synagogs ſiſter,
promiſing and beſtowing on her manie excellent benefites.}
Our ſiſter is litle, and hath no breaſts. What shal we doe to our ſiſter
in the day when she is to be ſpoken vnto? \V If she be a wal, let vs
build vpon
%%% 1477
it bulwarkes of ſiluer: if she be a doore, let vs ioyne it together with
bordes of ceder. \V
\SNote{The Church of Gentiles reioyceth in the ſtrong defence,
vvhervvith her Sauiour hath eſtablished her.}
I am a wal: and my breaſts are as a towre, ſince I was made before him
as one finding peace. \V The peacemaker had a vinyard, in that which
hath peoples: he deliuered the ſame
\SNote{Keepers of this vinyard, vvere the Prophetes and Apoſtles, and
their Succeſſors are ſtil the kepers therof.}
to keepers, a man bringeth for the fruite thereof a thouſand peeces of
ſiluer. \V
\SNote{Chriſt ſhevveth that together vvith the paſtors, himſelf
eſpecially hath care of his Church, alvvayes aſſiſting the viſible
gouerners therof vvith his inuiſible grace.}
My vineyard is before me. A thouſand are thy peacemakers, and two
\Fix{hundreth}{hundred}{likely typo, fixed in other}
for them, that keepe the fruites thereof. \V Thou that dwelleſt
in the gardens, the frends doe harken: make me heare thy voice. \V
\SNote{The vvhole Church militant vvel contented, yea deſiring Chriſts
Aſcenſion into heauen, for the good of al that here ſerue him, prayeth
him from thence to ſend abundance of his grace, that vve may aſcend the
high mountaines of perfect charitie, and zele of Gods honour, that he
vvil make our ſoules ſuch hilles, the garden of al vertues, & ſo
voutſafe to dwel therin. Amen.}
Flee, ô my beloued, and be like to the roe, and to the fawne of harts
vpon the mountaines of aromatical ſpices.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
