%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Canticle
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 1473
%%% o-1359
\startChapter[
  title={Chapter 04}
  ]

\Summary{}


How
\SNote{Chriſt againe praiſeth the beautie of his Church.}
beautiful art thou my loue, how beautiful art thou! thine
\SNote{Sincere and ſimple intention.}
eies as it were of doues, beſides that, which lyeth hid within. Thy
\SNote{Al her temporal occupations directed to Gods glorie.}
heares as the flockes of goates, which haue come vp from mount
Galaad. \V Thy
\SNote{Paſtors who like nurces geue bread of good doctrine to litle
ones.}
teeth as flockes of them that are shorne, which haue come vp from the
lauatorie, al with
\SNote{Faith and good workes.}
twinnes, and there is no barren among them. \V Thy
\SNote{Preaching Chriſts paſſion.}
lippes as a ſcarlet lace: and thy ſpeach ſweete. As a peece of a
pomegranate, ſo are alſo thy
\SNote{And not aſhamed to profeſſe Chriſt Crucified.}
cheekes, beſides that which lyeth hid within. \V Thy
\SNote{Adminiſtration of Sacraments wherby the Church, Chriſts myſtical
bodie, is ioyned to him her head,}
necke is as the
\SNote{which is an inexpugnable fortreſſe.}
towre of Dauid, which is built with bulworkes: a thouſand targattes hand
on it, al the armour of the valiants.
%%% o-1360
\V Thy
\SNote{Both Iewes and Gentiles are fed with the principles of Chriſtian
doctrin.}
two breaſtes as two fawnes the twinnes of a roe, which feede among the
lilies, \V til the day aſpire, and the shadowes decline.
\SNote{Chriſt dwelleth in mortified, and deuout mindes.}
I wil goe to the mount of myrrhe, and to the litle hil of
frankencenſe. \V Thou art al fayre ô my loue, and there is
\SNote{The Church triumphant is without ſpotte, and euerie particular
ſoule entring into heauen, the B.~Virgin mother was alſo in this life
alwayes immaculate.}
not a ſpotte in thee. \V Come from Libanus my ſpouſe, come from Libanus,
come: thou shalt be crowned from the head of Amena, from the toppe of
Sanit & Hermon, from the dennes of lions, from the mountaynes of
leopardes. \V Thou haſt wounded my heart, my ſiſter ſpouſe, thou haſt
wounded my heart in one of thine eies, and in one heare of thy necke. \V
How beautiful are thy breaſtes my ſiſter ſpouſe! thy breaſtes are more
beautiful then wine, and the odour of thine ointmentes aboue al
aromatical ſpices. \V Thy lippes my ſpouſe are as an honie combe
diſtilling, honie and
\Fix{mile kare}{milk are}{obvious typo, fixed in other}
vnder thy tongue: and the odour of thy garments as the odour of
frankincenſe. \V My ſiſter ſpouſe is a garden incloſed, a garden
incloſed, a fountaine ſealed vp. \V Thy offsprings a paradiſe of
pomegranats with orchard fruites. Cypres with ſpiknard, \V ſpiknard, and
ſafren, ſweete cane and cinnamon, with al the trees of Libanus, myrrhe
and aloes with al the chiefe ointmentes. \V The fountaine of gardens:
the wel of liuing waters, which runne with violence from Libanus. \V
\SNote{Al tentations, whether they be in manifeſt crueltie, or in
flatering ſuteltie, make conſtant ſoules more gratful to God.}
Ariſe Northwinde, & come Southwinde, blow through my garden, and let the
aromatical ſpices therof flowe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
