%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Canticle
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 1474
%%% o-1360
\startChapter[
  title={Chapter 05}
  ]

\Summary{}


Let
\SNote{The ſpouſe condeſcending to Gods vvil, is vvel content to ſuffer
perſecution.}
my beloued come into his garden, and eate the fruite of his appletrees.
\SNote{Chriſt againe ſhevveth his good liking in his ſpouſes patience:}
I am come into my garden ô my ſiſter ſpouſe, I haue reaped my myrrhe,
with myne aromatical ſpices: I haue eaten the honiecombe with mine
honie, I haue drunke my wine with my milke:
\SNote{and vvilleth the glorious Sainctes to congratulate vvith the
patient.}
eate ô frendes, and drinke, and be inebriated my deareſt.
%%% !!! Verse not marked in this
\V
\SNote{The ſpouſe deſireth to reſt in meditation,}
I ſleepe, and my hart watcheth:
\SNote{but is called vpon to helpe others:}
the voice of my beloued knocking: Open to me my ſiſter, my loue, my
doue, mine immaculate:
\SNote{and vrged by Chriſts owne example working for al mankind.}
becauſe my head is ful of dew, and my lockes of the droppes of the
nightes. \V I haue ſpoyled myſelfe of my robe, how shal I be clothed
with it? I haue washed my feete, how shal I defile them? \V My beloued
put his hand through the hole, and my bellie trembled at his touch. \V
\SNote{And ſo ſhe imployeth herſelf alſo in actiue life.}
I aroſe, that I might open to my beloued: my handes haue diſtilled
myrrhe, and my fingers are ful of moſt approued myrrhe. \V I opened the
bolt of my dore to my beloued: but he had turned aſide, and was
paſſed. My ſoule melted, as he ſpake: I ſought, and found him not: I
called, and he did not anſwer me. \V The keepers that goe about the
citie found me: they
%%% o-1361
ſtroke me, and wounded me: the keepers of the
walles tooke away my cloke. \V
\SNote{Stil conſeruing a deſire to returne vnto contemplation.}
I adiure you ô daughters of Hieruſalem, if you shal finde my beloued,
that you tel him, that I languishe with loue. \V
\SNote{The deuout confer together deſcribing the excellencies of
Chriſt.}
What maner of one is thy beloued of the beloued, ô moſt beautiful of
wemen? What maner of one is thy beloued of the beloued, that thou haſt ſo
adiured vs? \V My beloued is white and ruddie, choſen of thouſands. \V
His head is as the beſt gold: his heares, as the branches of palmetrees,
blacke as a rauen. \V His eies as doues vpon the litle riuers of waters,
which are washed with milke, and ſitte beſide the moſt ful ſtreames. \V
His cheekes are as litle beddes of aromatical ſpices ſet of the
pigmentaries. His lippes are as lilies diſtilling principal myrrhe. \V
His handes wrought round of gold, ful of hyacinthes. His
bellie of iuorie, diſtinguished with ſapphires. \V His thighes as
pillers of marble, that are vpon feete of gold. His forme as of Libanus,
elect as the cedars. \V His throte moſt ſweete, and he whole to be
deſired: ſuch an one is my beloued, and he is my frend, ô daughters of
Hieruſalem. \V
\SNote{And reſolue to ſeke him, wherſoeuer he be.}
Whither is thy beloued gone ô moſt beautiful of wemen? whither is thy
beloued turned aſide, and we wil ſeeke him with thee?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
