%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Canticle
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 1475
%%% o-1361
\startChapter[
  title={Chapter 06}
  ]

\Summary{}


My
\SNote{The Church teacheth her children that Chriſt is delighted with
the godlie deſires, and fructful vvorkes of the faithful.}
beloued is gone downe into his garden, to the bed of aromatical ſpices,
to feede in the gardens, and to gather lilies. \V I to my beloued, and
my beloued to me, who feedeth among the lilies. \V
\SNote{Chriſt againe commendeth his Church, wel compoſed of diſtinct
orders (ſome gouerning, ſome retired in cloiſters from this world, the
reſt alſo exerciſing vvorkes of mercie, in the trubles of this life) al
together making a complete armie, terrible to al enimies.}
Thou art fayre ô my loue, ſweete, and comelie as Hieruſalem: terrible as
the armie of a campe ſet in aray. \V
\SNote{The more anie contemplate Gods Maieſtie, the better they perceiue
that he is
\Fix{incomprenſenſible.}{incomprehenſible.}{obvious typo, fixed in other}}
Turne away thine eies from me, becauſe they haue made me flee away. Thy
heares as a flocke of goates, which haue appeared from Galaad. \V Thy
teeth as a flocke of sheepe, which haue come vp from the lauatorie, al
with twinnes, and there is no barren among them. \V As the barke of a
pomegranate, ſo are thy cheekes beſide thy hidden. \V There are
\SNote{Manie true paſtores,}
threeſcore queenes, &
\SNote{more hyrelinges, that alſo preach truth, but for temporal
commoditie,}
foureſcore concubines, and of
\SNote{and innumerable faithful ſoules in the Church.}
yongmaydes there is no number. \V My doue is
\SNote{Al vvhich are but one bodie in vnitie of faith.}
one, my perfect one, she is the only to her mother, elect to her that
bare her. The daughters haue ſeene her, and declared her to be moſt
bleſſed: the queenes and concubines, and haue prayſed her. \V
\SNote{The voice of the old ſynagogue, admiring the beautie of Chriſts
Church.}
What is she, that cometh forth as the morning ryſing, fayre as the moone,
elect as the ſunne, terrible as the armie of a campe ſet in aray? \V I
came downe into the garden of nuttes, to ſee the fruites of the valles,
and to looke if the vineyarde had florished, and the pomegranats
budded. \V I knew not: my ſoule trubled me for the chariotes of
Aminadab. \V
\SNote{The Church of Chriſt exhorteth the Synagogue of the Ievves to
returne to Chriſt.}
Returne, returne ô Sulamiteſſe: returne, returne that we may behold
thee.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
