%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Canticle
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 1471
%%% o-1358
\startChapter[
  title={Chapter 02}
  ]

\Summary{}


I am
\SNote{Chriſt profeſſeth himſelf the floure of mankinde: yea Lord of al
creatures.}
the flower of the filde, and the lilie of the valley. \V
\SNote{The Church excelleth al other ſocieties: In the Church the godlie
excel ſinners, among the innocent and holie, the virgin Marie ſurpaſſeth
al.}
As the lilie among the thornes, ſo is my loue among the daughters. \V
\SNote{The Church praiſing Chriſt reſteth vnder his protectiõ.}
As the apletree among trees of the
\Fix{woddes,}{woodes,}{obvious typo, fixed in other}
ſo is my beloued among
the ſonnes. Vnder his shadow, whom I deſired, I ſate: and his fruite was
ſweete vnto my throte. \V He brought me into the wineceller, he hath
ordered in me charitie. \V Stay me vp with flowers, compaſſe me about
with apples: becauſe I languish with loue. \V His lefthand vnder my
head, and his righthand shal embrace me. \V I adiure you ô daughters of
Ieruſalem, by the roes, and the hartes of the fildes,
\SNote{He for the weakes ſake permitteth her not to be moleſtes, til ſhe
be prepared to ſuffer vvith patience.}
that you rayſe not, nor make the beloued to awake, vntil herſelfe
wil. \V
\SNote{She feeling Chriſts aſſiſtance, confeſſeth, & preacheth boldly
his Goſpel, & truth againſt al Paganes, and Heretikes.}
The voice of my beloued, behold he cometh leaping in the mountaines,
leaping ouer the little hilles: \V my beloued is like vnto a roe, and to
a fawne of hartes. Behold he ſtandeth behind our walle,
\SNote{VVho though he ſhew not himſelf viſibly,}
looking through the windowes, looking forth by the grates. \V Behold my
beloued ſpeaketh to me:
\SNote{yet encorageth her to approch vnto him:}
Ariſe, make haſt my loue, my doue, beautiful one, and come. \V For
winter is now paſt, the rayne is gone, and departed. \V The flowers haue
appeared in our land, the time of pruning is come: the voice of the
turtle doue is heard in our land: \V the figgetree hath brought
%%% 1472
forth her greene figges: the florishing vineyards haue geuen their
ſauour. Ariſe my loue, my beautiful one, & come. \V My doue in the holes
of the rocke, in the holow places of the wal, shew me thy face, let thy
voice ſound in mine eares: for thy voice is ſweete, and thy face
comely. \V
\SNote{commandeth his paſtours to deſtroy hereſies.}
Catch vs the litle foxes, that deſtroy the vineyards: for our vineyard
hath florished. \V
\SNote{And ſo she repoſeth in him.}
My beloued to me, and I to him, who feedeth among the lilies, \V til the
day breake, and the shadowes decline. Returne: be like, my beloued, to a
roe, and to the fawne of hartes vpon the mountaynes of Bether.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
