%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Canticle
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 1472
%%% o-1359
\startChapter[
  title={Chapter 03}
  ]

\Summary{}


In
\SNote{The Church finding Chriſt not in darke ignorance, nor in
philoſophie but by his reueiling him ſelfe to her, holdeth him for
euer:}
my litle bed in the nightes I haue ſought him, whom my ſoule loueth, I
haue ſought him; and haue not found. \V I wil riſe, and wil goe about
the citie: by the ſtreates and high waies, I wil ſeeke him whom my ſoule
loueth: I haue ſought him, and haue not found. \V The watchmen which
kepe the citie found me: Haue you ſeene him, whom my ſoule
loueth. \V When I had a litle paſſed by them, I found him whom my ſoule
loueth: I held him: neither wil I let him goe, til I bring him into
\SNote{euen til the Iewes ſhal at laſt alſo find him.}
my mothers houſe, and into the chamber of her that bare me. \V
\SNote{Chriſt ſpeaketh as before
\XRef{ch.~2. v.~7.}}
I adiure you ô daughters of Ieruſalem by the roes, and the hartes of the
fildes, that you rayſe not vp, nor make the beloued to awake, til
herſelfe wil. \V
\SNote{The Church of Chriſt admireth her owne conuerſion from
Gentilitie,}
What is she, that aſcendeth by the deſert, as a litle rod of ſmoke of
\SNote{now ful of good workes.}
the aromatical ſpices of
\Fix{myrrthe,}{myrrhe,}{obvious typo, fixed in other}
and frankincenſe, & of al powder of the apothecharie? \V
\SNote{She alſo profeſſeth that the aſcending to eternal reſt, is by
fighting manfully, in obſeruing the ten commandments, in the fix dayes
of this life:}
Behold threeſcore valiants of the moſt valiant of Iſrael, compaſſe the
litle bed of Salomon: \V al holding ſwordes, and moſt cunning to
battels: euerie mans ſword vpon his thigh for feares by night. \V King
Salomon hath made him a portable throne of the wood of Libanus: \V the
pillers therof he hath made of ſiluer, the ſeate of gold, the going vp
\SNote{euen to bloud, if nede be,}
of purple: the middes he hath paued with
\SNote{which is the higheſt degree of charitie.}
charitie for the daughters of Ieruſalem. \V
\SNote{And inuiteth al others to come vnto Chriſt,}
Goe forth ye daughters of Sion, and ſee king Salomon in
\SNote{who in the fleſh which he tooke of his mother, was crowned in
heauen after his Paſſion.}
the diademe, wherewith his mother hath crowned him in the day of his
deſpouſing, and in the day of the ioy of his heart.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
