%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 2503
%%% o-2339
\startChapter[
  title={Chapter 7}
  ]

\Summary{The Iewes (of Hieruſalem) ſeeking his death, he walketh in
  Galilee: where he ſignifieth to his Brethren, that not in this feaſt
  Scenopegia, but in another (to wit, Paſche following) the Iewes should
kil him: that is, not when they would, but when he wil. 10.~In ſo much
that at this feaſt he teacheth openly in the Temple, and conuerteth many,
14.~both in the middle day 37.~and the laſt day thereof, without any
hurt, though alſo the Rulers ſend to apprehend him.}

After theſe things \Sc{Iesvs} walked \Var{into Galilee}{in Galilee}, for
he would not walke \Var{into Iewrie}{in Iewrie}: becauſe the Iewes
ſought to kil him. \V And the feſtiual day of the Iewes,
\CNote{\XRef{Leu.~23,~34.}}
\SNote{\Emph{Scenopegia} 
\XRef{(leu.~23.}
\G{σκηνῶν ἑορτή}) is the
\Emph{feaſt of Tabernacles}, which the Iewes kept from the 7.\ of
octo.\ for eight daies together by Gods commandement, for a memorie
that their Fathers dwelt by God's protection fourtie yeares in
tabernacles or tentes, and not in houſes, comming out of
Ægypt. \Emph{See}
\XRef{Leuit~23,~34.}}
Scenopégia, was at hand. \V And his Brethren ſaid to him: Paſſe from
hence, and goe into Iewrie; that thy Diſciples alſo may ſee thy workes
which thou doeſt. \V For no man doth any thing in ſecret, and ſeeketh
himſelf to be in publike. If thou doe theſe things, manifeſt thy ſelf to
the world. \V For neither did his Brethren 
%%% 2504
beleeue in him. \V \Sc{Iesvs} therfore ſaith to them: My time is not yet
come; but your time is alwaies readie. \V The world can not hate you,
but me it hateth: becauſe I giue teſtimonie of it, that the workes
thereof are euil. \V Goe you vp to this feſtiual day: \Var{I goe not
  vp}{I wil not yet goe vp} to this feſtiual day; becauſe my time is not
yet accompliſhed.

\V When he had ſaid theſe things, himſelf taried in Galilee. \V But
after his Brethren were gone vp, then he alſo went vp to the feſtiual
day, not openly, but as it were in ſecret. \V The Iewes therfore ſought
him in the feſtiual day, and ſaid: Where is he? \V And there was much
murmuring in the multitude of him. For certaine ſaid: That he is
good. And others ſaid: No, but he ſeduceth the multitudes. \V Yet no man
ſpake openly of him for feare of the Iewes.

\V And when the feſtiuitie was now half done, \Sc{Iesvs} went vp into
the Temple, and taught. \V And the Iewes 
%%% o-2340
maruailed, ſaying: How doth
this man know letters, whereas he hath not learned? \V
\Fix{Ieſus}{\Sc{Iesvs}}{obvious typo, fixed in other}
anſwered them, and ſaid: My doctrine is not mine, but is that ſent
me. \V If any man
\SNote{The way to come to know the truth, is to liue wel.}
wil doe the wil of him, he ſhal vnderſtand of the doctrine whether it be
of God, or I ſpeake of my ſelf. \V 
\SNote{It is ſpoken of Antichriſt ſpecially, and it is true in al
  Heretikes
\Cite{Auguſt tract.~29. in Euang. Ioan.}}
He that ſpeaketh of himſelf, ſeeketh his owne glorie. But he that
ſeeketh the glorie of him that ſent him, he is true, and iniuſtice in
him there is not. \V Did not Moyſes giue you the Law, and none of you
doth the law? \V
\CNote{\XRef{Io.~5,~18.}}
Why ſeeke you to kil me? The multitude anſwered, and ſaid, 
\SNote{No maruel, when theſe ſpeake thus to Chriſt himſelf, if Heretikes
cal his Vicar Antichriſt.}
Thou haſt a Diuel, who ſeeketh to kil thee? \V \Sc{Iesvs} anſwered and
ſaid to them: One worke I haue done; and you doe al marueil. \V Therfore
\CNote{\XRef{Leu.~12,~3.}}
Moyſes gaue you circumciſion: not that it is of Moyſes, but
\CNote{\XRef{Gen.~17,~10.}}
of the
Fathers, and in the Sabboth you circumciſe a man. \V If a man receiue
circumciſion in the Sabboth, that the Law of Moyſes be not broken; are
you angry at me becauſe I haue healed a man wholy in the Sabboth? \V
Iudge not according to the face, but iudge iuſt iudgement.

\V Certaine therfore of Hieruſalem ſaid: Is not this he whom they ſeeke to
kil? \V And behold, he ſpeaketh openly, and they ſay nothing to
him. Haue the Princes knowen indeed that this is \Sc{Christ}? \V But
this man we know whence he is. But when \Sc{Christ} commeth, no man
knoweth whence he is. \V \Sc{Iesvs} therfore cried in the
Temple teaching and ſaying: Both me you doe know, and whence I am you
know. And of my ſelf I am not come, but he is true that ſent me, whom
you know not. \V I know him, becauſe I am of him, and he ſent me. \V
They ſought therfore to apprehend him: and no man laid hands vpon him,
becauſe his houre was not yet come. \V But of the multitude many
beleeued in him, and ſaid: \Sc{Christ}, when he commeth ſhal he doe more
ſignes then theſe which this man doth? \V The Phariſees heard the
multitude murmuring theſe things touching him: and the Princes and
Phariſees ſent Miniſters to apprehend him. \V \Sc{Iesvs} therfore ſaid
to them: Yet a litle time I am with you: and I goe to him that ſent
me. \V
\CNote{\XRef{Io.~13,~33.}}
You ſeeke me, and ſhal not find: and where I am, you can not
come. \V The Iewes therfore ſaid among themſelues, whither wil this man
goe, that we ſhal not find him? wil he goe into the diſperſion of the
Gentils, and teach the Gentils? \V What is this ſaying that he hath
ſaid: You ſhal ſeek
%%% 2505
me, and ſhal 
%%% o-2341
not find; and where I am, you can not come.

\V And in the laſt, the
\CNote{\XRef{Leuit.~23,~26.}}
great day of the feſtiuitie \Sc{Iesvs} ſtood and
cried, ſaying: If any man thirſt, let him come to me, and drinke. \V He
that beleeueth in me, as the ſcripture ſaith,
\CNote{\XRef{Ioel.~2,~28.}
\XRef{Leu.~23,~36.}}
\Emph{Out of his belly
  shal flow riuers of liuing water.} \V (And this he ſaid
\CNote{\XRef{Act.~2,~1.}}
of the Spirit
that they ſhould receiue which beleeued in him. 
\SNote{This was fulfilled on whitſunday
\XRef{Act.~2.}
& afterward alwaies by impoſition of hands in the Sacrament of
  Confirmation: viſibly in the primitiue Church, and inuiſibly to the
  end of the world.}
For as yet the Spirit was not giuen: becauſe \Sc{Iesvs} was not yet
glorified.)

\V Of that multitude therfore, when they had heard theſe words of his,
ſome ſaid: This is the Prophet indeed. \V Others ſaid: This is
\Fix{Chriſt.}{\Sc{Christ.}}{obvious typo, fixed in other}
But certaine ſaid: Why, doth \Sc{Christ} come from Galilee? \V Doth not
the
\CNote{\XRef{Pſ.~131,~11.}
\XRef{Mich.~5,~2.}}
ſcripture ſay: That of the ſeed of Dauid, and from Bethlehem the
towne where Dauid was, \Sc{Christ} doth come? \V Therfore there aroſe
diſſenſiõ in the multitude for him. \V And certaine of them would haue
apprehended him: but no man laid hands vpon him. \V The Miniſters
therfore came to the cheefe Prieſts and the Phariſees. And they ſaid to
them: Why haue you not brought him? \V The Miniſters anſwered: Neuer did
there man ſo ſpeake, as this man. \V The Phariſees therfore anſwered
them: Why, are you alſo ſeduced? \V Hath any of the Princes beleeued in
him, or of the Phariſees? \V but this multitude that knoweth not the
Law, are accurſed. \V Nicodemus ſaid to them
\SNote{Chriſt hath ſome good alwaies euen amõg the wicked, which
  ſecretly ſerue him and by wiſe delaies auert the execution of uniuſt
  lawes againſt him and his people, as Nicodemus & Gamaliel.}
he
\CNote{\XRef{Io.~3,~2.}}
that came to him by night, who was one of them: \V Doth our Law iudge
a man, vnles it firſt heare him, and know what he doth? \V They anſwered
& ſaid to him: Why, art thou alſo a Galilæan? \Var{Search}{Search the
  ſcriptures}, & ſee that from Galilee a Prophet riſeth not. \V And
euery man returned to his houſe.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
