%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 2518
%%% o-2361
\startChapter[
  title={Chapter 13}
  ]

\Summary{At his laſt ſupper, to giue his farewel, and that in moſt
  wonderful louing manner, 4.~he washeth his Diſciples feet, 6.~begining
with Peter, 8.~(shewing how neceſſarie it is for vs to be washed of him
in Baptiſme; and needful alſo after Baptiſme) 12.~and by this example
teaching them al humilitie one toward another. 21.~Then he foretelleth,
that (notwithſtanding his exceeding loue toward them) one euen of them
wil betray him, meaning Iudas, 22.~as to Iohn he ſecretly sheweth. After
whoſe going out, he reioyceth and ſaith, that euen now the houre is
come, 34.~commendeth vnto them to loue together, as a new commandement,
36.~and foretelleth Peter, who preſumed too much of his owne ſtrength,
that euen this night he wil deny him thriſe.}

And
\CNote{\XRef{Mt.~26,~1.}
\XRef{Mr.~14,~1.}
\XRef{Luc.~22,~1.}}
before the feſtiual day of Paſche, \Sc{Iesvs} knowing that his houre
was come that he ſhould paſſe out of this world to his Father: whereas
he had loued his that were in the world, vnto the end he loued them. \V
And when
\SNote{By ſupper, he meaneth the eating of the Paſchal lambe. for, the
  inſtitution of the B.~Sacrament was after this.}
ſupper was done, whereas the Diuel now had put into the hart of Iudas
Iſcariote the ſonne of Simon, to betray him: \V knowing that the Father
gaue him al things into his handes, & that he came from God, and goeth
to God: \V he riſeth from 
%%% o-2357
ſupper, and laieth aſide his garments, and
hauing taken a towel, 
\Fix{grided}{girded}{obvious typo, fixed in other}
himſelf. \V After that, he put water into a baſon, and
\LNote{Began to wash}{This
\MNote{Puritie required to the receiuing of the B.~Sacrament.}
lotion was not only of curteſy, ſuch as the Iewes vſed toward their
gheſts, nor only for example of humilitie: but for myſterie and
ſignification of the great puritie that is required before we come to
receiue the holy Sacrament, which ſtraight after this washing was to be
\Fix{iuſtituted}{inſtituted}{obvious typo, fixed in other}
and giuen to the Apoſtles.
\Cite{Ambr. li.~3. de Sacra. c.~1.}
\Cite{Bernard de cœna Domini Sermon.~1.}}
began to wash the feete of the Diſciples, and to wipe them with the
towel where with he was girded. \V He commeth therfore to Simon
Peter. And Peter ſaith to him: Lord: doeſt thou wash my feet? \V
\Sc{Iesvs} anſwered and ſaid to him: That which I doe, thou knoweſt not
now, hereafter thou ſhalt know. \V Peter ſaith to him: Thou ſhalt not
wash my feet for euer. \Sc{Iesvs} anſwered him: If I waſh thee not, thou
ſhalt not haue part with me. \V Simon Peter ſaith to him: Lord, not only
my feet, but alſo handes and head. \V \Sc{Iesvs} ſaith to him: He that
is waſhed, needeth not but
\LNote{To wash his feete}{The
\MNote{Venial ſinnes taken away by ſacred ceremonies.}
foulnes of the feete, when al the reſt is cleane, ſignifieth the earthly
affections and relikes of former ſinnes remitted: which are to be
cleanſed by deuout actes of charitie and humilitie, as
\Cite{S.~Ambroſe li.~3. de Sacr c.~1.}
and
\Cite{S.~Auguſtin ep.~108. & tract.~56. in Io.}
doe note. And
becauſe this was only a ceremonie, & yet had ſuch force, both now, and
afterward vſed of the Apoſtles, that it purged ſmaller offenſes and
filthines of the ſoule, as
\CNote{\Cite{Ambr.}
&
\Cite{Bern. locis citatis.}}
\Emph{S.~Ambroſe} and \Emph{S.~Bernard} gather,
it may not ſeeme ſtrange that holy-water and ſuch ceremonies may remit
venial ſinnes.}
to wash his feet, but is cleane wholy. And you are cleane, but not
al. \V For he knew who he was that would betray him. therfore he ſaid:
You are not cleane al.

\V Therfore, after he had washed their feete, and taken his garments,
being ſet downe, againe he ſaid to them: Know you what I haue done to
you? \V You cal me, Maiſter, and Lord: and you ſay wel, for I am ſo. \V
If then I haue washed your feete, Lord and Maiſter;
\LNote{You alſo ought}{Our Maiſter neuer ſpake plainer, nor ſeemed to
  command more preciſely, either of Baptiſme or the Euchariſt or any
  other Sacramẽt: and yet 
\MNote{The Church defineth which are Sacraments and which not &c.}
by the Churches iudgement directed by the Holy
  Ghoſt we know this to be no Sacrament nor neceſſarie ceremonie, and
  the other to be. And why doe they beleeue the Church in this, and doe
  not credit her affirming the chalice not to be neceſſarie for the
  communicants?}
you alſo ought to wash one anothers feete. \V For I haue giuen you an
example, that as I haue done to you, ſo you doe alſo. \V Amen, amen I
ſay to you, a ſeruant is not greater then his Lord, neither is an
Apoſtle greater then he that ſent him. \V If you know theſe things, you
shal be bleſſed if you doe them. \V I ſpeake not of you al: I know whom
I haue choſen. But that the ſcripture may be fulfilled,
\CNote{\XRef{Pſ.~40,~10.}}
\Emph{He that
  eateth bread with me shal lift vp his heele againſt me.} \V From this
time I tel you, before it come to paſſe: that when it shal come to
paſſe, you may beleeue, that I am he. \V Amẽ, amẽ, I ſay to you, he that
receiueth any that I ſend, receiueth me:
%%% 2519
& he that receiueth me, receiueth him that ſent me. \V When \Sc{Iesvs}
had ſaid theſe things, he was troubled in ſpirit: and he proteſted, and
ſaid:
\CNote{\XRef{Mt.~26,~18.}
\XRef{Mr.~14,~16.}
\XRef{Luc.~22,~21.}}
Amen, amen I ſay to you that one of you shal betray me. \V The
Diſciples therfore looked one vpon another, doubting of whom he
ſpake. \V There was therfore one of his Diſciples leaning in the boſome
of \Sc{Iesvs}, he whom \Sc{Iesvs} loued. \V Therfore Simon Peter
beckneth to him, and ſaid to him: Who is it of whom he ſpeaketh? \V He
therfore leaning vpon the breaſt of \Sc{Iesvs} ſaith to him: Lord, who
is he? \V \Sc{Iesvs} anſwered: He it is 
%%% o-2358
to whom I ſhal reach the dipped bread. And when he had dipped the bread,
he gaue it to Iudas Iſcariote Simons ſonne. \V And after the morſel,
then Satan entred into him. And \Sc{Iesvs} ſaith to him: That which thou
doeſt, doe it quickly. \V But no man knew of thoſe that ſate at table to
what purpoſe he ſaid this vnto him. \V For certaine thought, becauſe
Iudas had the
\SNote{Chriſt had ſome prouiſion before hand giuen him by the
  Collections of the faithful; which was vſed both in his owne
  neceſſities, & beſtowed vpon the poore.}
purſe, that \Sc{Iesvs} had ſaid to him: Buie thoſe things which are
needful for vs to the feſtiual day: or that he ſhould giue ſome thing to
the Poore. \V He therfore hauing receiued the morſel, incontinent went
forth. And it was night.

\V When he therfore was gone forth, \Sc{Iesvs} ſaid: Now the Sonne of
man is glorified, and God is glorified in him. \V If God be glorified in
him, God alſo wil glorifie him in himſelf, and incontinent wil he
glorifie him. \V Litle children, yet a litle while I am with you. You
ſhal ſeeke me, &
\CNote{\XRef{Io.~7,~34.}}
as I ſaid to the Iewes, whither I goe, you can not
come: to you alſo I ſay now. \V
\CNote{\XRef{1.~Io.~3,~23.}}
\LNote{A new commandement}{The commandement of mutual loue was giuen
  before, but manifoldly miſconſtrued, and abridged by the Iewes to
  freinds only, to this life only, for earthly reſpects only: but Chriſt
  reneweth it and enlargeth it after the forme of his owne loue towards
  vs, and giueth grace to fulfil it.}
A new cõmandement I giue to you, That you loue one another; as I haue
loued you, that you alſo loue one another. \V In this al men ſhal know
that you are my Diſciples, if you haue loue one to another. \V Simon
Peter ſaith to him: Lord, whither goeſt thou? \Sc{Iesvs} anſwered,
Whither I goe, thou canſt not now follow me, but hereafter thou ſhalt
follow. \V Peter ſaith to him: Why can not I floow thee now?
\CNote{\XRef{Mt.~26,~35.}
\XRef{Mr.~14,~29.}
\XRef{Lu.~22,~33.}}
I wil yeald
my life for thee. \V \Sc{Iesvs} anſwered him: Thy life wilt thou yeald
for me? Amen, amen I ſay to thee, the cocke ſhal not crow, vntil thou
denie me thriſe.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
