%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-10


\project douay-rheims


%%% 2510
%%% o-2347
\startChapter[
  title={Chapter 10}
  ]

\Summary{He continueth his talke to the Phariſees, shewing that they
  and al other that wil not enter in by him, are wolues: and that they
  which heare them, are not the true sheep. 11.~But that himſelf is the
  good Paſtour, and therfore to ſaue the sheep from theſe wolues, he wil
yeald his life, which otherwiſe no might of theirs could take from him,
foretelling alſo his Reſurrection, and vocation of the
Gentils. 22.~Againe another time, he telleth theſe Iewes openly, that
they are not of his sheep, and that no might of theirs shal take from
him his true sheep, becauſe he is God, euen as his Father is
God. 31.~Which by his miracles and by Scripture he sheweth to be no
blaſphemie: and they in vaine ſeeking to ſtone and to apprehend
him, 40.~he goeth out to the place where Iohn Baptiſt had giuen open
witnes of him.}

%%% o-2348
Amen, amen I ſay to you, he that entreth not by the doore into the fold
of the ſheep, but
\LNote{Climeth another way}{Whoſoeuer
\MNote{Arch-heretikes ſpecially, are the theeues that clime in another
  way, not by the doore.}
taketh vpon him to preach without lawful ſending, to miniſter
Sacraments, & is not Canonically ordered of a true Catholike Bishop, to
be a Curate of ſoules, Perſon, Bishop, or what other ſpiritual Paſtour
ſoeuer, and commeth not in by lawful election and holy Churches
ordinance to that dignity, but breaketh in againſt order by force or
fauour of men, and by humane lawes, he is a theefe & a murderer. So came
in Arius, Caluin, Luther, & al Heretikes: & al that ſucceed them in
roome and doctrine. And generally euery one that deſcendeth not by
Lawful ſucceſsion in the knowen ordinarie line of Catholike Bishops and
Paſtours that haue been in al Countries ſince their conuerſions. And
according to this rule
\Cite{S.~Irenæus li.~3. c.~3.}
trieth the true shepheards from the theeues and Heretikes. So doe 
\Cite{Tertul. de Praſe. nu.~11.}
\Cite{S.~Cypr. de vnit. Ec. nu.~7.}
\Cite{S.~Auguſt. ep.~165. & cont. ep.}
\Cite{Manich. c.~4.}
and
\Cite{Lirinenſis.}}
climeth vp another way; he is
\SNote{The theefe, is the Heretike ſpecially, & any other that
  vnlawfully breake in vpon the ſheepe to kil & deſtroy thẽ by falſe
  doctrine & otherwiſe.}
a theefe and a robber. \V But he that entreth by the doore, is the
Paſtour of the ſheep. \V To this man the porter openeth; and the
%%% 2511
ſheep heare his voice: and he calleth his owne ſheep by name, and
leadeth thẽ forth. \V And when he hath let forth his owne ſheep, he 
\SNote{That is the faſhion of Iewrie & other coũtries, ſignifying that
  the ſhepheard or Paſtour muſt teach the ſheepe, & not they him.}
goeth before them: and the ſheep follow him, becauſe they know his
voice. \V But a ſtranger they follow not, but fly from him becauſe they
know not the voice of ſtrangers. \V This prouerb \Sc{Iesvs} ſaid to
them. But they knew not what he ſpake to them.

\V \Sc{Iesvs} therfore ſaid to them againe: Amen, amen I ſay to you,
that I am the doore of the ſheep. \V And how many ſoeuer haue come, are
theeues and robbers: but the ſheep heard them not. \V I am the doore. By
me if any enter, he ſhal be ſaued: & he ſhal goe in and ſhal goe out, &
ſhal find paſtures. \V The theefe cõmeth not but to ſteale and kil &
deſtroy. I came that they may haue life, & may haue more aboundãtly. \V
I am the good Paſtour.
\V
\CNote{\XRef{Eſa.~40,~11.}
\XRef{Ezec.~34,~23.}}
The
\LNote{Good Paſtour}{The good Paſtour, is he whoſe ſpecial care is not
  of his owne aduantage, but of the ſafty of the flocke. The hireling,
  is he that reſpecteth not the profit and good of the flocke, but
  his owne lucre. The woulfe, is the Heretike, or any perſecutour of the
  Church, which is Chriſts flocke.}
good Paſtour giueth his life for his ſheep. But the hireling & he
that is not the Paſtour, whoſe owne the ſheep are not, ſeeth the wulfe
comming, and leaueth the ſheep, and flieth: and the wulfe raueneth, and
diſperſeth the ſheep. \V And the hireling
\LNote{Flieth}{Euery
\MNote{When the Paſtour muſt tarie, or may flie.}
Bishop and Paſtour is bound to abide with his flocke in times of danger
& perſecution euen to death, except himſelf be perſonally ſought for,
rather then the flocke, or the flocke it ſelf forſake him. For in ſuch
caſes the Paſtour may flye, as the Apoſtles did, and S.~Athanaſius, and
others.
\Cite{S.~Athan. Apol. de ſua fuga.}
\Cite{Auguſt. ep.~180.}}
flyeth becauſe he is a hireling; and he hath no care of the ſheep. \V I
am the good Paſtour; and I know mine, and mine know me. \V As the Father
knoweth me, and I know the Father: and
\SNote{Chriſtes death was ſo neceſſarie for the flocke, that when he
  might haue eſcaped, he voluntarily offered himſelf to death for his
  flocke.}
I yeald my life for my ſheep. \V And
\SNote{He meaneth the Church of the Gentils.}
other ſheep I haue that are not of this fold: them alſo I muſt bring,
and they ſhal heare my voice, and there ſhal be made
\CNote{\XRef{Eze.~37,~24.}}
one fold and one
Paſtour. \V Therfore the Father loueth me: becauſe I yeald my life, that
I may take it againe. \V No man taketh it away from me: but
\CNote{\XRef{Eſa.~53,~7.}}
I yeald it of myſelf. And I haue the power to yeald it: and I haue power
to take it againe. This commandement I receiued of my Father.

\V A diſſenſion roſe againe among the Iewes for theſe wordes. \V And
many of them ſaid: He hath a Diuel and is mad; why heare you him? \V
Others ſaid: Theſe are not the wordes of one that hath a Diuel. Can a
Diuel open the eyes 
%%% o-2349
of blind men?

\V
\CNote{\XRef{1.~Mac.~4,~56.~59.}}
And
\LNote{The dedication}{This is the feaſt of Dedication inſtituted by
  Iudas Machabæus
\MNote{Iudas Machabeus}
\Cite{li.~1. Mach. c.~4.}
Chriſt vouchſafed to honour and keepe that feaſt inſtituted by him: &
our Heretikes vouchſafe not to pray and ſacrifice for the dead,
\CNote{\XRef{2.~Macha.~12.}}
vſed and approued by him. 
\MNote{Dedication of Churches.}
The Dedication alſo of Chriſtian Churches is warranted
thereby, with the annual memories thereof. And it proueth that ſuch
things may be inſtituted without any expreſſe commandement in
Scripture.}
the Dedication was in Hieruſalem: and it was winter. \V And \Sc{Iesvs}
walked in the Temple, in Salomons porch. \V The Iewes therfore compaſſed
him round about, and ſaid to him: How long doeſt thou hold our ſoule in
ſuſpenſe? if thou be \Sc{Christ}, tel vs openly. \V \Sc{Iesvs} anſwered
them: I ſpeake to you; and you beleeue not; the workes that I doe in the
name of my Father, they giue teſtimonie of me. \V But you doe not
beleeue, becauſe you are not of my ſheep. \V My ſheep heare my voice;
and I know them, and they follow me. \V And I giue them life
euerlaſting; and they ſhal not periſh for euer, and no man ſhal plucke
them out of my hand. \V 
\SNote{Another reading is, \Emph{My Father that hath giuen me, &c.}}
My Father,
\LNote{That which he gaue me}{Thus
\MNote{Chriſts eſſence & diuine nature of the Father.}
read alſo diuers of the
Fathers, namely
\Cite{S.~Hilar. Trin. li.~7. poſt medium.}
\Cite{S.~Ambr. de Sp. S. li.~3. c.~18.}
\Cite{S.~Auguſt. in Io. tract~48.}
\Cite{S.~Cyril. li.~7. in Io. c.~10.}
and vſe it to proue that
Chriſt had his eſſence and nature of the Father. And therfore ſome
Heretikes of our time wickedly accuſe the Councel of Laterane for
falſifying this place & applying it to the ſame purpoſe. Which they leſſe
can abide, for that it is againſt 
\MNote{Caluins autotheiſme}
Caluins Autotheiſme, holding that Chriſt tooke his perſon of the Father,
  but not his ſubſtance. \Emph{See the}
\XRef{2.~Annot. in 1.~Io. v.~1.}}
that which he hath giuen me, is greater then al: and no man can plucke
them out of the hand of my Father. \V I and the Father are
\TNote{\L{vnum.}}
one.

\V The Iewes tooke vp ſtones, to ſtone him. \V \Sc{Iesvs} anſwered them:
Many good workes I haue ſhewed you from my Father, for which of thoſe
workes doe you ſtone me? \V The Iewes anſwered him: For a 
\Fix{a }{}{obvious typo, fixed in other}% to prevent a paragraph break
good
worke we ſtone thee not, but for blaſphemie, and becauſe thou being a
man, makeſt thyſelf God. \V \Sc{Iesvs} anſwered them: Is it not written
in your law, that
\CNote{\XRef{Pſ.~81,~6.}}
\Emph{I ſaid, you are Goddes}? \V If he called them
Goddes, to whom the word of God was made, and the ſcripture can not be
%%% 2512
broken; \V whom the Father hath ſanctified and ſent into the world, ſay
you, That thou blaſphemeſt, becauſe I ſaid I am the Sonne of God? \V If
I doe not the workes of my Father, beleeue me not. \V But if I doe, and
if you wil not beleeue me, beleeue the workes: that you may know and
beleeue that the Father is in me, and I in the Father. \V They ſought
therfore to apprehend him; and he went forth out of their handes.

\V And he went againe beyond Iordan into that place 
\Fix{were}{where}{obvious typo, fixed in other}
Iohn was baptizing firſt; and he taried there. \V and many came to
him. and they ſaid: That Iohn indeed did no ſigne. But al things
whatſoeuer Iohn ſaid of this man, were true. \V And many beleeued in
him.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
