%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-19


\project douay-rheims


%%% 2529
%%% o-2370
\startChapter[
  title={Chapter 19}
  ]

\Summary{The Iewes are not ſatiſfied with his ſcourging &
  irriſion. 8.~Pilate hearing them ſay that he made himſelf the Sonne of
God, is more afraid. 12.~Yet, they vrging him with his loialty toward Cæſar,
& profeſsing that themſelues wil no King but Cæſar, he yealdeth vnto
them. 17.~And ſo Chriſt carying his owne Croſſe, is crucified between
two theeues, 19.~Pilate writing notoriouſly the only cauſe of his death
to be, for that he is their King or Chriſt. 23.~His garments be ſo vſed,
euen as the Scriptures foretold. 25.~He hath ſpecial care of his mother
to the end. 28.~He ſignifieth al that was written of his Paſsion, to be
fulfilled and ſo yealdeth vp his ghoſt. 31.~Then by the Iewes meanes
alſo other Scriptures about his legges and ſide, are fulfilled. 38.~And
finally, he is honourably buried.}

Then
\CNote{\XRef{Mt.~27,~27.}
\XRef{Mr.~15,~16.}}
therfore Pilate tooke \Sc{Iesvs}, and ſcourged him. \V And the
ſouldiars platting a crowne of thornes, put it vpon his head: and they
put about him a purple garment. \V And they came to him, and ſaid: Haile
King of the Iewes; and they gaue him blowes. \V Pilate went forth
againe, and ſaith to them: Behold I bring him forth vnto you, that you
may know that I find no cauſe in him. \V \Sc{Iesvs} therfore went forth
carying the crowne of thornes, and the purple veſtiment. And he ſaith to
them: Loe the man. \V When the cheefe Prieſts therfore and the miniſters
had ſeen him, 
%%% o-2371
they cried, ſaying: Crucifie, crucifie him. Pilate ſaith
to them: Take him you, and crucifie him. For I find no cauſe in him. \V
The Iewes anſwered him: We haue a Law: and according to the Law he ought
to die; becauſe he hath made himſelf the Sonne of God.

%%% 2530
\V When Pilate therfore had heard this ſaying, he feared more. \V And he
entred into the Palace againe; and he ſaith to \Sc{Iesvs}: Whence art
thou? But \Sc{Iesvs} gaue him no anſwer. \V Pilate therfore ſaith to
him: Speakeſt thou not to me? knoweſt thou not that I haue power to
crucifie thee, and I haue power to releaſe thee? \V \Sc{Iesvs} anſwered:
Thou ſhouldeſt not haue any power againſt me, vnles it were giuen thee
from aboue. Therfore he that hath betraied me to thee, hath the greater
ſinne. 

\V From thence forth Pilate ſought to releaſe him. But the Iewes cried,
ſaying: If thou releaſe this man, thou art not Cæſars freind. Euery one
that maketh himſelf a King, ſpeaketh againſt Cæſar. \V But Pilate when
he had heard theſe wordes, brought forth \Sc{Iesvs}: and he ſate  in the
iudgement ſeate, in the place that is called Lithoſtrotos, and in Hebrew
Gabbatha. \V And it was the Paraſceue of Paſche, about the
\SNote{He meaneth mid-day, counting from ſun-riſing. For ſo doth the
Scripture coũt the houres of the day.
\XRef{Mt.~20.}
\XRef{Mr.~15.}
\XRef{Lu.~21.}
\XRef{Io.~4.}
\XRef{Act.~13. & 10}}
ſixt houre, and he ſaith to the Iewes: Loe your King. \V But they cried:
Away, away with him, crucifie him. Pilate ſaith to them: Shal I crucifie
your King? The cheefe Prieſts anſwered: We haue no King, but Cæſar. \V
Then therfore he deliuered him vnto them for to be crucified.

And they tooke \Sc{Iesvs}, and led him forth. \V
\CNote{\XRef{Mt.~27,~33.}
\XRef{Mr.~15,~20.}
\XRef{Lu.~23,~33.}}
And bearing
\LNote{His owne croſſe}{This
\MNote{The \Sc{Holy Crosse}}
croſſe, for that it was the inſtrument of our redemption, and as it were
the Altar of the ſupreme Sacrifice, highly ſanctified by the touching,
bearing, & oblation of the ſacred body and bloud of our Lord, is truly
called the \Sc{Holy Crosse}: & hath been indowed with vertue of
miracles, both the whole & euery litle peece thereof. For the which
cauſes, & ſpecially for the moſt neere memorial of Chriſts Paſsion, it
hath been viſited in Pilgrimages, honoured by feſtiual daies &
otherwiſe, reſerued & reuerenced of the ancient Fathers with al
deuotion: as contrariwiſe it hath been abuſed of Pagans, Apoſtates, and
Heretikes, ſeeking in vaine to deface and deſtroy it. See
\Cite{S.~Cyril.  li.~6. cont. Iulian} 
\Cite{S.~Hierom. ep.~17.}
\Cite{S.~Paulinus ep.~11.}
\Cite{Ruffinus li.~1. c.~7.~&~8.}
\Cite{Euagr. hiſto. li.~4. c.~25.}
\Cite{S.~Leo ep.~72. and Ser.~8. de Paſsione.}
\Cite{Paulus Diac. li.~18.}}
his owne croſſe he went forth into that which is called the place of
Caluarie, in 
Hebrew Golgotha. \V Where they crucified him, and with him two others,
on the one ſide & on the other, and in the middes \Sc{Iesvs}. \V And
Pilate wrote a title alſo: and he put it vpon the croſſe. And it was
written, \Sc{Iesvs of nazareth the king of the Iewes}. \V This title
therfore many of the Iewes did read: becauſe the place where \Sc{Iesvs}
was crucified, was nigh to the citie: and it was written in 
\SNote{Theſe three tongues being for other cauſes moſt famous before in
  al the world, are now alſo dedicated to God in the triumphant title of
the Croſſe of Chriſt, and in them the holy Scriptures are more
conueniently written, taught, & preſerued.}
Hebrew, in Greeke, and in Latin. \V The cheefe Prieſts therfore of the
Iewes ſaid to Pilate: Write not, \Emph{The King of the Iewes}; but that
he ſaid, I am King of the Iewes. \V Pilate anſwered: That which I haue
written, I haue written.

\V The
\CNote{\XRef{Mt.~27,~35.}
\XRef{Mr.~15,~24.}
\XRef{Luc.~23,~34.}}
ſouldiars therfore when they had crucified 
%%% o-2372
him, tooke his
garments (and they made foure partes, to euery ſouldiar a part) & his
coate. And his
\SNote{This coate without ſeame is a figure of the vnitie of the
Church.
\Cite{Cypr. de vnit. Ec.}
And Euthymius and other write that our Lady made it.}
coate was without ſeame, wrought from the top through out. \V They ſaid
therfore one to another: Let vs not cut it, but let vs caſt lots for it
whoſe it ſhal be. That the ſcripture might be fulfilled ſaying:
\CNote{\XRef{Pſ.~21,~19.}}
\Emph{They haue parted my garments among them: and vpon my veſture they
  haue caſt lotte.} And the ſouldiars did theſe things.

\V And there ſtood beſide the croſſe of \Sc{Iesvs},
\SNote{The great loue, faith, courage, compaſsion, and ſorrowes, that
our Lady had who forſooke not the Croſſe and her Sonne when ſo many
were fled from him, and his cheefe Apoſtles denied him.}
his mother, and his mothers ſiſter, Marie of Cleophas, and Marie
Magdalene. \V When \Sc{Iesvs} therfore had ſeen his mother and the
Diſciple ſtanding whom he loued, he ſaith
\LNote{To his mother}{The
\MNote{By this you ſee why in Catholike Churches Marie and Iohn ſtand by
the Rood.}
maruelous reſpect that Chriſt had to his mother, vouchſafing to ſpeake
to her, and to take order for her euen from the croſſe in the middes of
his infinite anguishes and myſteries a working for mankind.}
to his mother: Woman, behold thy ſonne. \V After that, he ſaith
\LNote{To the Diſciple}{A great honour to Iohn and charge, to haue that
bleſſed iewel in keeping: & an vnſpeakable comfort, that from that day
forward the one was to the other mother and ſonne,
\CNote{\Cite{li.~2. adu. Iouinian.}}
\L{Virginem virgini commendauit}, ſaith S.~Hierom, \Emph{He commended
the virgin to a virgin.}}
to the Diſciple: Behold thy mother. And from that houre the Diſciple
tooke her to his owne.

\V Afterward \Sc{Iesvs} knowing that al things were now conſummate, that
the
\CNote{\XRef{Pſ.~68,~22.}}
ſcripture might be fulfilled, he ſaith, I thirſt. \V A veſſel
therfore ſtood there ful of vinegre. And they putting a ſponge ful of
vinegre about hiſſope, offered it to his mouth. \V \Sc{Iesvs} therfore
when he had taken the vinegre, ſaid: It is conſummate. And bowing his
head, he gaue vp the ghoſt.

%%% 2531
\V The Iewes therfore (becauſe it was the Paraſceue) that the bodies
might not remaine vpon the croſſe on the Sabboth (for that was a great
Sabboth day) they deſired Pilate that their legges might be broken, and
they might be taken away. \V The ſouldiars therfore came: and of the
firſt indeed they brake the legges, and of the other that was crucified
with him. \V But after they were come to \Sc{Iesvs}, when they ſaw that
he was dead, they did not breake his legges, \V but one of the ſouldiars
with a ſpeare opened his ſide, & incontinent there came forth 
\LNote{Bloud and water}{This
\MNote{The Sacramẽts iſſued out of Chriſts ſide, & thence haue their
  vertue.}
pearſing of Chriſts ſide, though on the ſouldiars part it was done
blindly and inſolently, yet by Gods ordinance it conteined great
myſteries, and was prefigured by
\CNote{\XRef{Exo.~18.}
\XRef{Num.~30.}}
Moyſes ſtriking the rocke with his
rodde: as this ſtreame of bloud and water drawen miraculouſly out of his
dead body, running in the Sacraments of the Church after the people of
God, was ſignified by the water of the ſame rocke following the
Iſraelites in the deſert.
\CNote{\Cite{Tract.~9. in Ioan.}}
\Emph{Out of this ſide} (ſaith S.~Auguſtin) \Emph{iſſued the Sacraments.}
\CNote{\Cite{Hom.~24. in Ioan.}}
\Emph{Hence} (ſaith S.~Chryſoſtom)
\Emph{the great myſteries haue their beginning.} Who warneth vs, that
when we come to drinke of the holy chalice, we should ſo approch, as
though we drew the bloud out of Chriſts ſide. And both bloud and water
apart did flow forth, to shew vs the fountaine of the two principal Sacraments
and their ſeueral matters, Baptiſme and the Euchariſt, ſpringing to life
euerlaſting in the Church. 
\MNote{The Church builded of Chriſts ſide, as Eue of Adams.}
The fathers alſo ſay that the Church who is
Chriſts ſpouſe and his coadiutrice in applying the bloud and water to
the benefits of the Faithful, was here formed, builded, and taken out of
this holy ſide of Chriſt ſleeping on the Croſſe, as
\CNote{\XRef{Gen.~2,~21.}}
Eue was of Adams ſide, when he was caſt aſleep in Paradiſe.}
bloud and water. \V And he that ſaw it, hath giuen teſtimonie: and his
teſtimonie is true. And he knoweth that he ſaith true, that you alſo may
beleeue. \V For theſe things were done that the ſcripture might be
fulfilled:
\CNote{\XRef{Exo.~12,~46.}}
\Emph{You shal not breake a bone of him.} \V And againe another
ſcripture ſaith:
\CNote{\XRef{Zac.~12,~10.}}
\Emph{They shal looke on him whom they pearſed.}

\V And after theſe things
\CNote{\XRef{Mt.~27,~57.}
\XRef{Mr.~15,~42.}
\XRef{Lu.~23,~50.}}
Ioſeph of Arimathæa (becauſe he was a Diſciple
of \Sc{Iesvs}, but ſecret for feare of the Iewes) deſired Pilate that he
might take away the body of \Sc{Iesvs}. And Pilate permitted. He came
therfore, & tooke away the body of \Sc{Iesvs}. \V
\CNote{\XRef{Io.~3,~1.}
\XRef{7,~50.}}
Nicodemus came alſo, he that at the 
%%% o-2373
firſt came to \Sc{Iesvs} by night, bringing a mixture of
myrrhe and aloes, about an hundred poundes. \V They tooke therfore the
body of \Sc{Iesvs}, and bound it in linnen clothes with the ſpices, as
the manner is with the Iewes to burie. \V And there was in the place
where he was crucified, a garden: and in the garden a new monument,
wherein no man yet had been laid. \V There therfore becauſe of the
paraſceue of the Iewes, they laid \Sc{Iesvs}, becauſe the monument was
hard by.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
