%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 2515
%%% o-2353
\startChapter[
  title={Chapter 12}
  ]

\Summary{The
\MNote{The 4.~part. The 4.~Paſche, & holy weeke of his Paſsion in
  Hieruſalem.}
Rulers dealing as if he hid himſelf, 1.~he commeth to Bethania. 3.~Where
by occaſion of Iudas the theefe murmuring at Marie Magdalens coſtly
deuotion, he foretelleth his death. 12.~From thence, though they did now
intend to kil Lazarus alſo, he ride openly into Hieruſalem, the people
(becauſe he had raiſed Lazarus) confeſsing with their acclamations that
he is Chriſt. 20.~Where certaine Gentils deſiring to ſee him, 23.~he
foretelleth the conuerſion of the whole world from the Diuel to him, to
be now inſtant, as the effect of his death vpon the Croſſe. 28.~The
Father alſo anſwering from Heauen to his prayer made to that purpoſe,
37.~yet after al this, the Iewes 
\Fix{continve}{continue}{obvious typo, fixed in other}
incredulous as Eſay prophecied of them: 42.~though many beleeued, but
were ashamed to confeſſe him. 44.~Whereupon he sheweth that it is
glorious before God, and ſaluation to themſelues to beleeue in him, and
confeſſe him: and damnable to deſpiſe him.}

\Sc{Iesvs} therfore
\CNote{\XRef{Mt.~26,~6.}
\XRef{Mar.~14,~3.}}
ſixe daies before the Paſche came to Bethania, where
Lazarus was, that had been dead, whom \Sc{Iesvs} raiſed. \V And they
made him a ſupper there: and Martha miniſtred, but Lazarus was one of
them that ſate at the table with him. \V 
\SNote{Of this womans extraordinarie offices of deuotion, & how
acceptable they were to Chriſt, ſee the
\XRef{Annot. Mt.~26.}}
Marie therfore tooke a pownd of ointement of right ſpikenard, pretious,
and 
%%% o-2354
anointed the feete of \Sc{Iesvs}, and wiped his feete with her
haire: and the houſe was filled of the odour of the ointement. \V One
therfore of his Diſciples, Iudas Iſcariote, he that was to betray him,
ſaid: \V
\LNote{Why was}{So
\MNote{Church ornaments.}
wicked, couetous, and ſacrilegious perſons reprehend good men for
beſtowing their goods vpon Church ornaments &c. vnder pretence of better
beſtowing them on the poore: ſuch prouide for the poore as Iudas did.}
Why was not this ointment ſold for three-hundred pence, and giuen to the
poore? \V And he ſaid this, not becauſe he cared for the poore; but
becauſe he was
\LNote{A theefe}{Iudas
\MNote{Toleration of euil.}
did not then firſt perish when he ſould our Lord, for he was a theefe
before: and being loſt he yet followed Chriſt, not in hart, but in body
only. Which our Maiſter tolerated, to giue vs a leſſon to tolerate the
il, rather then deuide the body.
\Cite{Aug. tract.~50. in Io.}}
a theefe, and hauing the purſe, caried the things that were put in. \V
\Sc{Iesvs} therfore ſaid: Let her alone that she
%%% 2516
may keepe it for the day of my 
\SNote{The deuout offices of balming & anointing the dead bodies of the
  faithful are here alſo allowed.}
burial. \V For the poore you haue alwaies
with you; but
\SNote{Not ĩ viſible & 
\Fix{moral}{mortal}{obvious typo, fixed in other.}
condition, to receiue almes of you or ſuch like offices for ſupply of my
neceſsities.}
me you shal not haue alwaies. \V A great multitude therfore of the Iewes
knew that he was there; and they came, not for \Sc{Iesvs} only, but that
they might ſee Lazarus, whom he raiſed from the dead. \V But the cheefe
Prieſts deuiſed for to kil Lazarus alſo: \V becauſe many for him of the
Iewes went away, and beleeued in \Sc{Iesvs}.

\V And on the morow a great multitude that was come to the feſtiual day
when they had heard that \Sc{Iesvs} commeth to Hieruſalem, \V they tooke
the
\CNote{\XRef{Mt.~21,~1.}
\XRef{Mr.~11,~7.}
\XRef{Lu.~19,~35.}}
boughes of palmes, and went forth to meet him, & cried:
\CNote{\XRef{Pſ.~117.~26.}}
\Emph{Hoſanna, Bleſſed is he that commeth in the name of our Lord, the
  King of Iſrael.} \V And \Sc{Iesvs} found a yong aſſe, and ſate vpon
it, as it is written: \V
\CNote{\XRef{Zach.~9,~9.}}
\Emph{Feare not daughter of Sion: behold, thy
  King commeth ſitting vpon an aſſes colt.} \V Theſe things his
Diſciples did not know at the firſt: but when \Sc{Iesvs} was glorified,
then they remembred that theſe things had been written of him, and theſe
things they did to him. \V The multitude therfore gaue teſtimonie, which
was with him when he called Lazarus out of the graue, and raiſed him
from the dead. \V For therfore alſo the multitude came to meete him,
becauſe they heard that he had done this ſigne. \V The Phariſees
therfore ſaid among themſelues: Doe you ſee that we preuaile nothing?
Behold the whole world is gone after him.

\V And there were certaine Gentiles of them that
\SNote{We may ſee there is a great difference where a man pray or adore,
at home or in the Church & holy places: when the Gentils alſo came of
deuotion a pilgrimage to the Temple in Hieruſalem.}
came vp to adore in the feſtiual day. \V Theſe therfore came to Philippe
who was of Bethſaida of Galilee, and deſired him, ſaying: Sir, we are
deſirous to ſee \Sc{Iesvs}. \V Philippe commeth and telleth
Andrew. Againe Andrew & Philippe told \Sc{Iesvs}. \V But \Sc{Iesvs}
anſwered them, ſaying: The houre is come, that the Sonne of man ſhal be
glorified. \V Amen, amen I ſay to you, vnles the graine of wheate
falling into the ground, 
%%% o-2355
die: it ſelf remaineth alone. but if it die, it
bringeth much fruit. \V He that loueth his life, ſhal loſe it: and he
that hateth his life in this world, doth keep it to life euerlaſting. \V
If any man miniſter to me, let him follow me: and where I am, there alſo
ſhal my miniſter be. If any man miniſter to me, my Father wil honour
him. \V Now my ſoule is troubled. And what ſhal I ſay? Father, ſaue me
from this houre. But therfore came I into this houre. \V Father,
glorifie thy name. A voice therfore came from Heauen: Both I haue
glorified it, and againe I wil glorifie it. \V The multitude therfore
that ſtood and heard, ſaid that it thundered. Others ſaid, An Angel
ſpake to him. \V \Sc{Iesvs} anſwered and ſaid: This voice came not for
me, but for your ſake. \V Now is the iudgement of the world: now the
Prince of this world ſhal be caſt forth. \V And I,
\CNote{\XRef{Io.~3,~14.}}
if I be exalted from
the earth, wil draw al things to my ſelf. \V (And this he ſaid,
ſignifying what death he ſhould die) \V The multitude anſwered him: We
haue heard out of the law, that \Sc{Christ} abideth for euer; and how
ſaieſt thou: The Sonne of man muſt be exalted? Who is this Sonne of man?
\V \Sc{Iesvs} therfore ſaid to them: Yet a litle while, the light is
among you. Walke whiles you haue the light, that the darkeneſſe ouertake
you not. And he that
%%% 2517
walketh in darkeneſſe, knoweth not whither he goeth. \V Whiles you haue
the light, beleeue in the light, that you may be the children of
light. Theſe things \Sc{Iesvs} ſpake and he went away, and hid himſelf
from them.

\V And whereas he had done ſo many ſignes before them, they beleeued not
in him: \V that the ſaying of Eſay the Prophet might be fulfilled, which
he ſaid:
\CNote{\XRef{Eſ.~53,~1.}}
\Emph{Lord, who hath beleeued the hearing of vs?} \V \Emph{and
  the arme of our Lord to whom hath it beene reuealed?} \V Therfore they
\SNote{If any man aske (ſaith S.~Auguſtine) why they could not beleeue:
I anſwer roundly, becauſe they would not
\Cite{Tract.~33 in Io.}
See the meaning of this ſpeach  
\XRef{Annot. Mat.~13,~15.}
\XRef{Mar.~4,~12.}}
could not beleeue, becauſe Eſay ſaid againe:
\CNote{\XRef{Eſa.~6,~9.}
\XRef{Mt.~13,~14.}
\XRef{Mr.~4,~12.}
\XRef{Lu.~8,~10.}
\XRef{Act.~28,~27.}}
\Emph{He hath blinded their
  eyes, and indurated their hart: that they may not ſee with their eyes,
  nor vnderſtand with their hart, and be conuerted, and I heale them.}
\V Theſe things ſaid Eſaie, when he ſaw his glorie, & ſpake of him. \V
But yet of the Princes alſo many beleeued in him: but
\SNote{This is the caſe of many principal men in ſuch countries where
  hereſie hath the vpper hand, who know & beleeue the Catholike faith;
  but making choiſe rather to keepe mans fauour then Gods, they dare not
confeſſe the ſame. Such may pray that God and the world agree together:
for els it is ſeen whoſe part they wil take.}
for the Phariſees they did not confeſſe, that they might not be caſt out
of the Synagogue. \V For they loued the glorie of men more, then the
glorie of God.

\V But \Sc{Iesvs} cried, and ſaid: He that beleeueth in me, doth not
beleeue in me, but in him that ſent me. \V And he that 
%%% o-2356
ſeeth me, ſeeth
him that ſent me \V I a light am come into this world: that euery one
which beleeueth in me, may not remaine in the darkeneſſe. \V And if any
man heare my wordes, and keepe them not: I doe not iudge him. For I came
not to iudge the world, but to ſaue the world. \V He that deſpiſeth me,
& receiueth not my wordes, hath that iudgeth him. the word that I haue
ſpoken, that ſhal iudge him in the laſt day. \V Becauſe of my ſelf I
haue not ſpoken, but the Father that ſent me, he gaue me commandement
what I ſhould ſay, and what I ſhould ſpeake. \V And I know that his
commandement is life euerlaſting. The things therfore that I ſpeake: as
the Father ſaid to me, ſo doe I ſpeake.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
