%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 2482
%%% o-2316
\startChapter[
  title={Chapter 1}
  ]

\Summary{The
\MNote{The 1.~parte: \Sc{the acts} of Chriſt before his
  manifeſtation, whiles Iohn Baptiſt was yet baptizing}
preface of the Euangeliſt, commending Chriſt (as being God the Sonne
incarnate) to the Gentils, and ſetting out the blindnes of the Iewes in
not receiuing him.  19.~Then, the teſtimonies of Iohn Baptiſt, firſt to
the ſolemne legacie of the Iewes: 29.~ſecondly, when he ſaw 
\Fix{\Sc{iesvs}}{\Sc{Iesvs}}{obvious typo, fixed in other}
come to him: 35.~thirdly, to his owne Diſciples alſo putting them ouer
from himſelf to 
\Fix{\Sc{iesvs}}{\Sc{Iesvs}}{obvious typo, fixed in other}
who made it plainer to them that he is
Chriſt, 40.~and ſo began he to haue Diſciples.}

In the beginning
\LNote{Was the Word}{The
\MNote{How God the 
\Fix{the }{}{obvious typo, fixed in other}% to prevent paragraph break
Sonne is called the \Sc{word}.}
ſecond Perſon in Trinitie which is the natural, only, and eternal Sonne
of God the Father, is called the \Sc{word}: not as the holy Scriptures
or ſpeaches of the Prophets and Apoſtles (written and ſpoken by Gods
commandement for the vttering of his diuine wil towards man) be called
his word, but in a more diuine, eminent and ineffable ſort, to expreſſe
vnto vs in a ſort, by a terme agreable to our capacitie, that he Sonne
of God ſo is and ſo from euerlaſting is borne of God the Father, as our
prime concept (which is our internal and mental word) is & iſſueth out
of our intelligence & mind.  This \Sc{word} then, Sonne, or ſecond
Perſon in the holy Trinitie, was & had his being then already, when
other creatures (of what ſort ſoeuer) had but their beginning, and
therfore cannot be a creature, as many Heretikes before the writing of
this Ghoſpel thought, and as the Arians after taught.  And this firſt
ſentence of the Ghoſpel not only the faithful, but the Platonikes
\MNote{The Platonikes.}
did ſo admire
\CNote{\Cite{Augu. de Ciu. Dei. li.~10. c.~19.}}
(as S.~Auguſtine writeth) that they wished it to be
written in gold.}
was the \Sc{word}, and the \Sc{word} was 
\LNote{With God}{Becauſe
\MNote{The \Sc{word} coeternal with the Father, diſtinct in Perſon
and of the Father.}
a man might ſay: If the \Sc{word} were before any thing was created,
where or how could he be?  the Euangeliſt preuenting that carnal
concept, ſaith firſt, that he was with God, whoſe being dependeth not
vpon time, place, ſpace, or any other creatures, al which 
\Fix{where}{were}{obvious typo, fixed in other}
made by him.  Secondly, he giueth vs to vnderſtand, that the
\Sc{word} hath his proper ſubſiſtence or perſonalitie diſtinct from God
the Father, whereby Sabellius the old Heretike is refuted.  Thirdly,
here is inſinuated the order of theſe two perſons, one towardes the
other, to wit, that this Sonne is with and of the Father, and not the
Father of the Sonne.  Fourthly, you may confute here the blaſphemie of
\CNote{\Cite{Calu. inſt. li.~1. c.~13. Sect.~23.}
&
\Cite{25.}}
Caluin, holding the ſecond Perſon to be God, not as of the Father, but
as of himſelf.  And yet ſuch are the bookes that our youth now read
commonly in England, and that by commandement.}
with God, and
\LNote{God was the Word}{Leſt
\MNote{The \Sc{word} true God by nature.}
 any man vpon the premiſſes,
which ſet forth the relation, & diſtinction of the ſecond Perſon from
the firſt, might thinke that the Father only were God, the Euangeliſt
expreſly teacheth, the \Sc{word} to be God.  For though the wordes ſeeme
to lie otherwiſe (becauſe we haue of purpoſe followed the elegancie
which the Euangeliſt himſelf obſerued in placing them ſo, and therfore
they ſtand ſo both in Greeke and Latin) yet indeed the conſtruction is
thus: \Emph{The \Sc{word} was God}, and
\CNote{\XRef{1.~Io. c.~5,~20.}}
(as
\Fix{is}{in}{obvious typo, fixed in other.}
his firſt Epiſtle the ſame Apoſtle writeth) \Emph{true God}: leſt any
might ſay (as the Arians did) that he was God indeed, but not truely &
naturally, but by common adoption or calling, as good men in the Church
be called the ſonnes of God.  What wonderful wrangling and tergiuerſation
the Arians vſed to auoid the euidence of this place, we ſee in
\Cite{S.~Auguſtine. li.~3. de Doct. Chriſt. c.~2.}
euen ſuch
\MNote{The Proteſtants are like the wrangling Arians.}
 as the Proteſtants doe, to auoid the like wordes, \Emph{This
  is my body}, concerning the B.~Sacrament.}
God was the \Sc{word}. \V This was in the beginning with God.  \V Al
things were made 
\LNote{By him}{Againe,
\MNote{The \Sc{word} not a creature but the Creatour.}
 by this he ſignifieth the eternitie, diuinitie, omnipotencie, and
 equalitie of the \Sc{word} or
\Fix{Sone,}{Sonne,}{obvious typo, fixed in other.}
with God the Father, becauſe by him al things were created.  Al things
he ſaith, both viſible of this world, and inuiſible, as Angles and al
Spiritual creatures.  Wherevpon it is euident alſo, that himſelf is no
creature, being the Creatour of al: neither is ſinne of his creation,
being a defect of a thing, rather then a thing it ſelf, and therfore
neither of not by him.}
by him: and without him was made
\Var{nothing. That which was made,}{nothing that was made.}
\V in him was life, and the
life was the light of men: \V and the light ſhineth in darkeneſſe, and
the darkeneſſe did not comprehend it.  \V There was a man ſent from
God, whoſe name was Iohn.  \V This man came for teſtimonie; to giue
teſtimonie of the light, that al might beleeue through him.  \V He was
not the light, but to giue teſtimonie of the light.  \V It was the true
light, which lighteneth euery man that commeth into this world.  \V He
was in the world, and the world was made by
%%% 2483
him, and the world knew him not.  \V He came into his owne, and his owne
receiued him not.  \V But as many as receiued him, 
\LNote{He gaue them power}{Free
\MNote{Free-wil.}
wil to receiue or acknowledge Chriſt, & power giuen to men, if they
wil, to be made by Chriſt the ſonnes of God: but not forced or drawen
therevnto by any neceſſitie.}
he gaue them
\SNote{Beza falſly tranſlated \L{dignitatem} for \L{poteſtatem}.
  \G{ἐξουσίαν}}
power to be made the ſonnes of God, to thoſe that beleeue in his name.
\V Who, not of bloud, nor of the wil of fleſh, nor of the wil of man,
but of God are borne.  \V \Sc{And}
\TNote{\L{\Sc{et verbvm caro factvm est.}}}
\Sc{the} 
\LNote{The Word made flesh}{This is the high and diuine teſtimonie
  of Chriſts incarnation and that he vouchſafed to become man.  For the
  acknowledging of which inexplicable benefit & giuing humble thankes
  for the ſame, al
\MNote{Humble kneeling at the ſolemne wordes of Chriſts
  incarnation}
Chriſtian people in the world by tradition of the Fathers proſtrate
themſelues or kneele downe, when they heare it ſung or ſaid at the holy
Maſſe, either in this Ghoſpel, or in the Creede by theſe wordes: \L{\Sc
  et homo factvs est}.}
\Sc{word was made flesh}, and dwelt in
vs (and we ſaw the glorie of him, glorie as it were of the Only-begotten
of the Father) ful of grace and 
%%% o-2317
veritie.  \V Iohn giueth teſtimonie of
him, and crieth ſaying: This was he of whom I ſpake, He that shal come
after me,
\SNote{He is preferred & made of more dignitie and excellencie then
  I, becauſe he was before me & al things Eternal God.}
is made before me: becauſe he was before me.  \V And of his fulnes we al
haue receiued, and grace for grace.  \V For the law was giuen by Moyſes,
grace and veritie was made by \Sc{Iesvs} Chriſt.  \V God 
\LNote{No man hath ſeen}{Neuer
\MNote{How mortal men ſee God.}
man in this mortalitie ſaw God in the very shape and natural forme of
the diuine eſſence, but men ſee him only in the shape of viſible
creatures, in or by which it pleaſeth him to ſhew himſelf vnto many
diuerſly in this world: but neuer in ſuch ſort as when he ſhewed him
ſelf in the Perſon
of the Sonne of God, being made truly man and conuerſing with men.}
no man hath ſeen at any time: the only-begotten Sonne which is in the
boſome of the Father, he hath declared.

\V And this is Iohns teſtimonie, when the Iewes ſent from Hieruſalem
Prieſts and Leuites to him, that they should aske him, who art thou?  \V
And he confeſſed, and did not denie: and he confeſſed, That I am not
\Sc{Christ}.  \V And they asked him: What then? Art thou
\CNote{\XRef{Mal.~4,~5.}}
Elias? And he
ſaid: I am not.  Art thou
\CNote{\XRef{Deu.~18,~15.}}
\SNote{By like the Iewes ignorãtly vnderſtood not the place in
  Deuteronomie, of Chriſt, and therfore they aske alſo whether he be the
Prophet there ſpoken of.  See alſo
\XRef{c.~7,~40.}}
the Prophet?  And he anſwered: No.  \V They ſaid therefore vnto him: Who
art thou, that we may giue an anſwer to them that ſent vs? what ſaieſt
thou of thy ſelf?  \V He ſaid:
\CNote{\XRef{Eſa.~40,~3.}}
\Emph{I am the voyce of one crying in the
deſert, make ſtraight the way of our Lord}, as Eſaie the prophet
ſaid. \V And they that were ſent were of the Phariſees.  \V And they
asked him, and ſaid to him: why then doeſt thou baptize, if thou be not
Chriſt, not Elias, nor the Prophet  \V Iohn anſwered them, ſaying:
\CNote{\XRef{Mt.~3,~11.}
\Cite{Mr.~1,~8.}
\Cite{Lu.~3,~16.}}
\SNote{He doth oftẽ here ſignifie the great difference of his
Baptiſme & of Chriſts, as of his perſon & Chriſts.  See
\XRef{Annot. Mat.~3.}}
I baptize in water; but there hath ſtood in the middes of you whom you
know not.  \V The ſame is he that ſhal come after me, that is made
before me; whoſe latchet of his ſhoe I am not worthie to looſe.  \V
Theſe things were done in Bethania beyond Iordan, where Iohn was
baptizing. 

\V The next day Iohn ſaw \Sc{Iesvs} comming to him, and he ſaith:
\Emph{Behold
\MNote{Agnus Dei at Maſſe.}
the Lamb of God, behold him that taketh away
\Var{the ſinne}{ſinnes}
of the world.} \V This is he of whom I ſaid: After me
there commeth a man, which is made before me: becauſe he was before
me. \V And I knew him not, but that he may be manifeſted in Iſrael,
therefore came I baptizing in water.  \V And Iohn gaue teſtimonie,
ſaying: That I ſaw 
\LNote{The Spirit}{Here
\MNote{The B.~Trinitie.}
is an euident teſtimonie of the third Perſon in Trinitie, which is the
Holy Ghoſt: ſo that in this one Chapter we finde expreſly againſt al
Heretikes, Iewes, & Pagans, ſet forth the truth of the Churches
doctrine concerning the whole Trinitie.}
the Spirit deſcending as a doue from Heauen, and he
remained vpon him.  \V And I knew him not: but he that ſent me to
baptize in water, he ſaid to me: He vpon whom thou shalt ſee the Spirit
deſcending and remaining vpon him, he it is that baptizeth in the Holy
Ghoſt.  \V And I ſaw; and I gaue teſtimonie that this is the Sonne of
God.

\V The next day againe Iohn ſtood, and two of his Diſciples. \V And
beholding \Sc{Iesvs} walking, he ſaith: Behold
%%% o-2318
the Lamb of God. \V And the two Diſciples heard him ſpeaking, and they
followed \Sc{Iesvs}.  \V And \Sc{Iesvs} turning, and ſeeing them
following him, ſaith to them: What ſeeke you?  Who ſaid to him: Rabbi
(which is called by interpretation, Maiſter) where dwelleſt thou?  \V He
ſaith to them:
%%% 2484
Come and ſee.  They came, and ſaw where he abode, and they taried with
him that day: and it was about the tenth houre. \V And Andrew the
brother of Simon Peter was one of the two that had heard of Iohn, and
followed him. \V He findeth firſt his brother Simon, and ſaith to him:
We haue found \Sc{Messias},
\SNote{Meſsias in Hebrew, in Greeke Chriſt, in English Anointed, to
wit, with the ſpiritual oile of grace aboue his brethren.
\XRef{Ps.~44}}
which is being interpreted, \Sc{Christ}.  \V And he brought him to
\Sc{Iesvs}.  And \Sc{Iesvs} 
\LNote{Looking vpon him}{This
\MNote{Peter by his new name deſigned to be the Rocke of the
  Church.} 
beholding of Simon, inſinuateth Chriſts deſignement: and preferring of
him to be the cheefe Apoſtle, the Rocke of the Church and his Vicar, and
therfore vpon that Diuine prouidence & intention he accordingly
changeth his name, calling him for Simon, \Emph{Cephas}, which is a
Syriake word, as much to ſay as Rocke or ſtone.  And S.~Paul commonly
calleth him by this name Cephas, whereas other both Greekes and Latines
cal him altogether by the Greeke word, \Emph{Peter}, which ſignifieth
the ſelf ſame thing.  Whereof
\CNote{\Cite{Li.~2. c.~12. in Iohn.}}
S.~Cyril ſaith, that our Sauiour by
foretelling that his name ſhould no more now be Simon, but \Emph{Peter},
did by the word itſelf aptly ſignifie, that on him, as on a rocke and
ſtone moſt firme, he would build his Church.}
looking vpon him, ſaid: Thou art Simon the
ſonne of Iona: thou ſhalt be called \Emph{Cephas},
\SNote{Cephas in Syriake, & Peter in Greeke, in Engliſh Rocke.  See
\XRef{Mat.~16,~18.}}
which is interpreted, \Emph{Peter}.  \V On the morow he would goe forth
into Galilee, and he findeth Philippe.  And \Sc{Iesvs} ſaith to him:
Follow me.  \V And Philippe was of Bethſaida, the citie of
Andrew and Peter.  \V Philippe findeth Nathanael, and ſaith to him: Him whom
Moyſes in the law, and the Prophetes wrote of, we haue found, \Sc{Iesvs}
the ſonne of Ioſeph, of Nazareth.  \V And Nathanael ſaid to him: From
Nazareth can there be any good?  Philippe ſaith to him: Come and ſee.
\V \Sc{Iesvs} ſaw Nathanael comming to him, and he ſaith of him: Behold
an Iſraelite in very deed, in whom there is no guile.  \V Nathanael
ſaith to him: How knoweſt thou me?  \Sc{Iesvs} anſwered and ſaid to him:
Before that Philippe did cal thee, when thou waſt vnder the figtree, I
ſaw thee.  \V Nathanael anſwered him and ſaith: Rabbi, thou art the
Sonne of God, thou art the King of Iſrael. \V \Sc{Iesvs} anſwered, and
ſaid to him: Becauſe I ſaid vnto thee, I ſaw thee vnder the figtree,
thou beleeueſt; greater then theſe things ſhalt thou ſee.  \V And he
ſaith to him: Amen Amen I ſay to you, You shal ſee the Heauen opened,
and the
\CNote{\XRef{Gen.~28,~12.}}
Angels of God aſcending and deſcending, vpon the Sonne of man.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

