%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 2526
%%% o-2366
\startChapter[
  title={Chapter 17}
  ]

\Summary{After his Sermon of farewel, he prayeth to his Father, that
  ſeeing he hath now finished his worke, he wil giue him his appointed
  glorie, for the conuerſion of al Nations, 6.~and preſerue his
  Apoſtles, and his Church after them in vnitie and veritie (that is,
  from Schiſme and Hereſie:) 14.~finally alſo glorifie them with him in
  Heauen.}

Theſe things ſpake \Sc{Iesvs}: and lifting vp his eyes into Heauen, he
ſaid: Father, the houre is come,
\SNote{The Father glorifieth the Sonne by rayſing him from death,
  exalting him vp to his right hand, making al creatures to bow downe at
his name, and giuing him al power and iudgement. The Sonne againe
glorifieth the Father, by making his honour, which only in a manner was
in Iewrie before, now knowen to al Nations.}
glorifie thy Sonne, that they Sonne may glorifie thee. \V As thou haſt
giuen him power ouer al fleſh, that al which thou haſt giuen him, to
them he may giue life euerlaſting. \V And this is
\LNote{Life euerlaſting}{Both
\MNote{True knowledge of God.}
the life of glorie in Heauen, and of grace here in the Church,
conſiſteth in the knowledge of God: that, in perfect viſion: this, in
faith working by charitie. For knowledge of God without keeping his
commandements, is not true knowledge, that is to ſay, it is an
vnprofitable knowledge.
\XRef{1.~Io.~2.}}
life euerlaſting that they know thee, the only true God, and whom thou
haſt ſent \Sc{Iesvs Christ}. \V I haue glorified thee vpon the earth: I
haue conſummated the worke which thou gaueſt me to doe: \V and now
glorifie thou me O Father with thyſelf, with the glorie which I had
before the world was, with thee. \V I haue manifeſted thy name to the
men whom thou gaueſt me out of the world. Thine they were, and to me
thou gaueſt them: and they haue kept thy word. \V Now they haue knowen
that al things which thou gaueſt me are from thee: \V becauſe the wordes
which thou gaueſt me, I haue giuen them: and they haue receiued, and
knowen in very deed that I came forth from thee, and haue beleeued that
thou didſt ſend me. \V For them doe I pray: Not for the world doe I
pray, but for them whom thou haſt giuen me; \V becauſe they be thine:
and al my things be thine, and thine be mine: and I am glorified in
them. And now I am not in the world, and theſe are in the world, & I
come to thee. \V Holy Father,
\SNote{His petition is ſpecially to keepe the Apoſtles and his Church
  in vnitie and from Schiſmes.}
keep them in thy name, whom thou haſt giuen me; that they may be one, as
alſo we. \V When I was with them, I kept them in thy name. Thoſe
\CNote{\XRef{Io.~18,~9.}}
whom
thou gaueſt me, haue I kept and none of the periſhed, but the ſonne of
perdition, that the
\CNote{\XRef{Pſ.~40,~10.}
\XRef{108,~8.}}
ſcripture may be fulfilled. \V And now I come to
thee: and theſe things I ſpeake in the world, that they may haue my ioy
filled in themſelues. 
%%% o-2367
\V I haue giuen them thy word, and the world hath
hated them, becauſe they are not of the world: as I alſo am not of the
world. \V I pray not that thou take them away out of the world, but that
thou preſerue them from euil. \V Of the world they are not: as I alſo am
not of the world. \V
\LNote{Sanctifie them}{Chriſt
\MNote{The Church can not erre.}
prayeth that the Apoſtles, their Succeſſours, & al that ſhal be of their
beleefe, may be ſanctified in truth. Which is as much to ſay, as to
deſire that the Church may euer haue the Spirit of truth, and be free
from errour. Which praier of Chriſt had not been heard, if the Church
might erre.}
Sanctifie them in truth. Thy word is truth. \V As thou didſt ſend me
into the world, I alſo haue ſent them into the world. \V And for them I
doe 
\SNote{To ſanctifie himſelf, is to ſacrifice himſelf by dedicating his
  holy body & bloud to his Father, both vpon the Croſſe, and in the holy
Sacrament.}
ſanctifie myſelf: that they alſo may be ſanctified in truth. \V And not
for thẽ only doe I pray,
\LNote{But for them}{He
\MNote{The Canon of the Maſſe.}
expreſſeth (and it is a great comfort) that he praieth not only for the
Apoſtles, but for the whole Church after them, that is, for al
beleeuers. And al this profound & diuine praier is reſembled in the holy
Canon of the Maſſe before the conſecration, as here it was made before
his viſible Sacrifice on the Croſſe.}
but for them alſo that by their word ſhal beleeue in me: \V that they al
may be one, as thou (Father) in me & I in thee; that they alſo in vs may
be one: that the world may beleeue that thou haſt ſent me. \V And the
glorie that thou haſt giuen me, haue I giuen to them; that they may be
one as we alſo are one. \V I in them, and thou in me: that they may be
conſummate in one: and the world may know that 
%%% 2527
thou haſt ſent me, and haſt loued them, as me alſo thou haſt loued. \V
Father, whom thou haſt giuen me, I wil, that where I am, they alſo may
be with me: that they may ſee my glorie which thou haſt giuen me,
becauſe thou haſt loued me before the creation of the world. \V Iuſt
Father, the world hath not knowen thee. But I haue knowen thee: and
theſe haue knowen, that thou didſt ſend me. \V And I haue notified thy
name to them, and wil notifie it: that the loue wherwith thou haſt
loued me may be in them, and I in them.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
