%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-04


\project douay-rheims


%%% 2490
%%% o-2325
\startChapter[
  title={Chapter 4}
  ]

\Summary{Leauing Iewrie becauſe of the Phariſees, in the way to Galilee
  he talketh with a Samaritane woman, telling her that he wil giue water
of euerlaſting life, 16.\ ſhewing himſelf to know mens ſecrets, 19.\
preferring the Iewes religion before the Samaritanes, but ours (the
Chriſtian Catholike religion) before them both, 25.\ and vttering vnto
her that he is Chriſt. 28.\ which by her teſtimonie and his preaching
very many Samaritanes doe beleeue: he in the meane  time foretelling his
Diſciples, of the harueſt he wil ſend them into. 45.\ The Galileans alſo
receiue him, where againe he worketh his ſecond miracle.}

When \Sc{Iesvs} therfore vnderſtood that the Phariſees heard that
\Sc{Iesvs} maketh moe Diſciples, and baptizeth, then Iohn, \V (howbeit
\Sc{Iesvs} did not baptize, but his Diſciples)
\SNote{He did not
baptize ordinarily: yet that he baptized his Apoſtles, S.~Aug.\
thinketh it very probable,
\Cite{ep.~108.}}
 \V he left Iewrie,
and went againe into Galilee. \V And he had of neceſſitie to paſſe
through Samaria. \V He commeth therfore into a citie of Samaria which is
called Sichar;
\CNote{\XRef{Gen.~48,~22.}}
beſide the Manour that Iacob gaue to Ioſeph his ſonne. \V
And there was there the fountaine of Iacob. \Sc{Iesvs} therfore wearied
of his iourney, ſate ſo vpon the fountaine. It was about the ſixt houre.

\V There cometh a woman of Samaria
\SNote{This woman is a figure of
the Church, not yet iuſtified, but now to be
iuſtified.
\Cite{Aug. tract.~15. in Ioan.}}
to draw
water. \Sc{Iesvs} ſaid to her: Giue me to drinke. \V For his Diſciples
were gone into the citie, to buie meates. \V Therfore that Samaritane
woman ſaith to him: How doeſt thou being a Iewe, aske of me to drinke,
which am a Samaritane woman? For the Iewes doe not communicate with the
Samaritanes.
\SNote{There were many other cauſes why the faithful
  Iewes could not abide the Samaritans, but their preciſe abſtaning from
their companie & cõuerſation, was their Schiſmatical Tẽple and Seruice
in moũt Garizim.}
\V \Sc{Iesvs} anſwered, and ſaid to her: If thou didſt know
the guift of God, and who he is that ſaith vnto thee, Giue me to drinke;
thou perhaps wouldſt haue asked of him, and he would haue giuen the
liuing water.
\SNote{He ſpeaketh of his baptizing in the Holy
Ghoſt. See
\XRef{Io. c.~7,~39.}}
\V The woman ſaith to him: Sir,
neither haſt thou wherein to draw, and the well is deep; whence haſt
thou liuing water? \V art thou greater then our Father Iacob, who gaue
vs the wel, & himſelf dranke of it, and his children, and his cattel?
\V \Sc{Iesvs} anſwered,
%%% 2491
and ſaid to her: Euery one that drinketh of this water, ſhal thirſt
againe; but he that ſhal drinke of the water that I wil giue him, ſhal
not thirſt for euer, \V but the water that I wil giue him, ſhal become
in him a fountaine of water ſpringing vp vnto life euerlaſting. \V The
woman ſaith to him: Lord giue me this water, that I may not thirſt, nor
come hither to draw.

\V \Sc{Iesvs} ſaith to her: Goe, cal thy husband and come hither. \V The
woman anſwered and ſaid: I haue no huſband. \Sc{Iesvs} ſaid to her: Thou
haſt ſaid wel, that I haue no 
%%% o-2326
husband. \V For thou haſt had fiue
husbands; and he whom thou now haſt, is not thy huſband. This thou haſt
ſaid truely.

\V The woman ſaith to him: Lord, I perceiue that thou art a Prophet. \V
\LNote{Our Fathers adored}{By
\MNote{The Schiſmatical Temple contendeth againſt the true Temple.}
adoration is meant doing of Sacrifice. For other offices of Religion
might be done in any place. The Samaritanes to defend their adoring in
Garizim, pretẽded their worſhiping there to be more ancient then the
Iewes in Hieruſalem, referring it to Iacob: whereas indeed that
Patriarch adoring there before the Temple was appointed, or the Law
giuen, made nothing for their Schiſme:
\CNote{\Cite{Ioſeph. l. antiq.~11. c.~8.}}
which was begun by Manaſſes a
fugitive Prieſt, only to hold his vnlawful wife thereby, and to obtaine
Superioritie in Schiſme: which he could not doe in the vnitie of his
brethren; long after the Temple of Hieruſalem from which reuolt was made.
Therfore Chriſt giueth ſentence for the Iewes & the Temple of
Hieruſalem, affirming that they had a good ground thereof, but the
Samaritans none at al.

Ioſephus alſo recordeth how the Samaritans demanded of Alexander the
Great, the like priuiledges & immunities as he had granted to the high
Prieſt and Temple of Hieruſalem, pretending their Temple to be as great
and as worthy, and themſelues to be Iewes as the other, and to worſhip
the ſame God. But
\MNote{The true Temple preuaileth.}
 their Schiſmatical hypocriſie was eaſily ſpied and
dimiſſed with nothing. Another time the Iewes and Samaritanes
\CNote{\Cite{Ioſeph. l.~12. antiq. c.~6.}}
(as the
ſame writer teſtifieth) made a great ſturre in Alexandria about the
truth and antiquitie of the Schiſmatical Temple & ſeruice in Garizim,
and the other true Temple of Salomon: in ſo much that the matter was put
to arbitrement by Ptolomæus the Kings commandement, only to trie whether
of the two was firſt. And the Schiſmatikes (as their cuſtome is) \L{per
  ſaltum} can make their Church or ſeruice as old as they like,
referring it to the Patriarches, as our Schiſmatikes doe now to Chriſt
and the Apoſtles. But
\MNote{The true Temple is proued by continual ſucceſsion.}
 when the trial was made, only they of Hieruſalem
did inuincibly proue by continual ſucceſsion of their Prieſts, and by
the iuſt note of the time when the Schiſmatikes went out from them, that
theirs was the lawful, and the other the falſe Temple and falſe
adoration. And ſo it was iudged, and the Samaritanes put to
ſilence. Afterward the ſaid Schiſmatikes (which is lightly the end of al
Schiſmes) reuolted quite from the Iewes religion, and dedicated their
Temple in Garizim to Iupiter Olympius, as Caluins ſupper and his bread
and wine is like at length to come to the ſacrifice of Ceres &
Bacchus.}
Our Fathers adored in this mountaine, and you ſay
\CNote{\XRef{Deu.~12,~6.}
\XRef{Pſ.~121,~13.}}
that at
Hieruſalem is the place where men muſt adore. \V \Sc{Iesvs} ſaith to
her: Woman beleeue me, that the houre ſhal come, when you ſhal neither
in this mountaine, nor in Hieruſalem adore the Father. \V
\CNote{\XRef{4.~Reg.~17,~20.}
\XRef{18,~36.}}
You adore that
you know not: we adore that we know; for ſaluatiõ is of the Iewes. \V
But the houre commeth, and now is, when the true adorers ſhal adore the
Father
\LNote{In Spirit and veritie}{Our
\MNote{Chriſtian adoratiõ throughout al Nations in euery place, in
  ſpirit & veritie: that is in the Sacraments & Seruice of the new
  law, ful of ſpirit & grace: in the veritie of things before
  prefigured, ſpecially the true Sacrifice of Chriſts body and bloud.}
Sauiour foretelleth her that the end & ceaſing of their Sacrifice &
adoration in both the Temples should shortly be, and euen then was begun
to be fulfilled: inſtructiong her in three things concerning that
point. Firſt, that the true Sacrifice should be tied no more to that one
place or Nation, but that true adoration should be throughout al Nations
according to the
\CNote{\XRef{Mal.~1.~11.}}
Prophecie of Malachie. Secondly, that the groſſe and
carnal adoration by the flesh and bloud of beaſtes and other external
terrene creatures not hauing in them grace, ſpirit, and life, should be
taken away, & another Sacrifice ſucceed, which should be in itſelf
inuiſible, celeſtial, diuine, ful of life, Spirit & grace. And thirdly,
that this adoration & Sacrifice should be the veritie itſelf whereof al
the former Sacrifices and Hoſtes were but ſhadowes & figures. And he
calleth that here ſpirit and truth, which in the
\CNote{\XRef{Io.~1,~17.}}
firſt Chapter is called
grace and truth. Al which is no more but a prophecie and deſcription of
the Sacrifice of the faithful Gẽtils in the bodie & bloud of Chriſt:
not that it is not by external meanes giuen to vs (for otherwiſe we
being men conſiſting of flesh & bloud could not be capable thereof) but
that it is ſpirit and life in it ſelf, being the flesh of the \Sc{word}
of God. And if a man enlarge the word of Adoration, (which here as is
ſaid, ſignifieth properly the worship of God by Sacrifice) to al the
Sacraments of the new Law, they al likewiſe be ſpirit and grace, the
Holy Ghoſt working inuiſibly and internally vpon our ſoules by euery one
of thẽ. Whereupon our Baptiſme, is water & the Holy Ghoſt: our Penance,
the word of abſolutiõ & the Holy Ghoſt: our Confirmatiõ, oile & the
Holy Ghoſt by impoſition of handes: finally, al the adoration of the
Catholike Church, is properly ſpiritual, though certaine external
creatures for our natures, ſtate and neceſſitie, be ioyned
thereunto. Take heed therfore thou gather not of Chriſts wordes, that
Chriſtian men should haue no vſe of external office towards God: for that would
take away al Sacrifice, Sacraments, Praiers, Churches, and ſocietie of
men in his Seruice.}
in ſpirit and veritie.For the Father alſo ſeeketh
ſuch to adore him. \V God is a Spirit, and they that adore him, muſt
adore in ſpirit and veritie. \V The woman ſaith to him: I know that
\Sc{Messias} commeth, (which is called \Sc{Christ}) therfore when he
commeth, he wil ſhew vs al things. \V \Sc{Iesvs} ſaith to her: I am he,
that ſpake with thee.

\V And incontinent his Diſciples came: and they maruelled that he talked
with a woman. No man for al that ſaid: What ſeekeſt thou, or why talkeſt
thou with her?

\V The woman therfore left her water-pot: and ſhe went into the citie,
and ſaith to thoſe men: \V Come, and ſee a man that hath told me al
things whatſoeuer I haue done. Is not he \Sc{Christ}? \V They went forth
therfore out of the citie, and came to him.

\V In the meane time the Diſciples deſired him, ſaying: Rabbi eate. \V
But he ſaid to them: I haue meate to eate which you know not. \V The
Diſciples therfore ſaid one to another: Hath any man brought him for to
eate? \V \Sc{Iesvs} ſaith to them: My meate is to doe the wil of him
that ſent me, to perfit his worke. \V Doe not you ſay that yet there are
foure moneths, and harueſt commeth? Behold I ſay to you, lift vp your
eyes, and ſee the Countries, that they are white already to harueſt. \V
And he that reapeth, receiueth hire, and gathereth fruit vnto life
euerlaſting: that both he that ſoweth, and he that reapeth, may reioyce
together. \V For in this is the ſaying true: that it is one man that
ſoweth, and it is another that reapeth. \V I haue ſent you to
reape that which you laboured not: others haue laboured, & you 
%%% o-2327
haue entred into their labours.

\V And of that citie many beleeued in him of the Samaritans, for the
word of the
\SNote{This woman myſtically beĩg the Church, it is here
  ſignified that they which at the firſt beleeue becauſe the Church
  teacheth ſo, afterward be much confirmed, finding it in the Scripture
alſo, and by other inſtructions.}
woman giuing teſtimonie, that he told
me al things whatſoeuer I haue done. \V Therfore when the Samaritans
were come to him, they deſired him that he would tarie there. And he
taried 
%%% 2492
there two daies. \V And many moe beleeued for his owne word. \V And they ſaid
to the woman, That now not for thy ſaying doe we beleeue; for our ſelues
haue heard, and doe know that this is the Sauiour of the world indeed.

\V And after two daies he departed thence; and went into Galilee. \V For
\Sc{Iesvs} himſelf gaue teſtimonie that a Prophet hath not homour in his
owne countrie: Therfore
\CNote{\XRef{Mt.~4,~12.}
\XRef{Mr.~1,~14.}
\XRef{Luc.~4,~14. }}
when he was come into Galilee, the Galilæans
receiued him, whereas they had ſeen al things that he had done at
Hieruſalem in the feſtiual day: for themſelues alſo came to the feſtiual
day. 

\V He came againe therfore into Cana of Galilee,
\CNote{\XRef{Io.~2,~9.}}
where he made water
wine.  And there was a certaine Lord whoſe ſonne was ſicke at
Capharnaum. \V He hauing heard that \Sc{Iesvs} came from Iewrie into
Galilee, went to him, and deſired him that he would come downe and heale
his ſonne, For he began to die. \V \Sc{Iesvs} therfore ſaid to him:
Vnleſſe you ſee ſignes and wonders, you beleeue not. \V The Lord ſaith
to him: Lord, come downe before that my ſonne die. \V \Sc{Iesvs} ſaith to
him: Goe, thy ſonne liueth. The man beleeued the word that \Sc{Iesvs}
ſaid to him, and went. \V And as he was now going downe, his ſeruants
met him: and they brought word, ſaying, That his ſonne liued. \V He
asked therfore of them the houre, wherein he was amended. And they ſaid
to him, That yeſterday at the ſeauẽth houre the feuer left him. \V The
father therfore knew that it was in the ſame houre wherein \Sc{Iesvs}
ſaid to him: Thy ſonne liueth, and himſelf beleeued and his whole
houſe. \V This againe the
\CNote{\XRef{Io.~2,~11.}}
ſecond ſigne did \Sc{Iesvs}, when he was come from Iewrie into Galilee.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
