%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 2527
%%% o-2368
\startChapter[
  title={Chapter 18}
  ]

\Summary{Being gone to the place that Iudas the Traitour did know, 4.~he
offereth himſelf to the band of his enemies, sheweth his Diuine might in
ouerthrowing them al with a word, and in ſauing his Apoſtles from them
alſo with a word: 10.~rebuketh Peter that would defend him from them
12.~and ſo being apprehended, is brought bound to Annas and Caiphas,
where he is ſtriken by a ſeruant, and thriſe denied of
Peter. 28.~Againe in the morning he is by them brought to
Pilate. 29.~Who demanding their accuſation, whereas they would oppreſſe
him with their authoritie, 33.~& examining the point of his Kingdom,
pronounceth him innocent: yet they crie rather to haue a theeues life
ſaued.}

When
\CNote{\XRef{Mt.~26,~36.}
\XRef{Mr.~14,~32.}
\XRef{Lu.~22,~39.}}
\Sc{Iesvs} had ſaid theſe things, he went forth with his Diſciples
beyond the Torrent-Cedron, where was a garden, into the which he entred
and his Diſciples. \V And Iudas alſo, that betraied him, knew the place:
becauſe \Sc{Iesvs} had often reſorted thither together with his
Diſciples. \V
\CNote{\XRef{Mt.~26,~47.}
\XRef{Mr.~14,~43.}
\XRef{Luc.~22,~47.}}
Iudas therfore hauing receiued the band of men, and of the
cheefe Prieſtes and the Phariſees, miniſters, commeth thither with
lanternes and torces and weapons. \V \Sc{Iesvs} therfore knowing al
things that ſhould come vpon him, went forth, and ſaid to them: Whom
ſeeke ye? \V They anſwered him, \Sc{Iesvs} of Nazareth. \Sc{Iesvs} ſaith
to them: I am he. And Iudas alſo that betraied him, ſtood with
%%% 2528
them. \V As ſoone therfore as he ſaid to them, I am he; they went
backward, and fel to the ground. \V Againe therfore he asked them: Whom
ſeeke ye? And they ſaid: \Sc{Iesvs} of Nazareth. \V \Sc{Iesvs} anſwered,
I haue told you, that I am he. If therfore you ſeeke me, let theſe goe
their waies. \V That the word might be fulfilled which he ſaid,
\CNote{\XRef{Io.~17,~12.}}
That of
them whom thou haſt giuen me, I haue not loſt any. \V Simon Peter
therfore hauing a ſword, drew it out; and ſmote the ſeruant of the high
Prieſt, & cut off his right eare. And the name of the ſeruant was
Malchus. \V \Sc{Iesvs} therfore ſaid to Peter: Put vp thy ſword into the
ſcabbard. The chalice which my Father hath giuen me, shal not I drinke
it? \V The band therfore and the Tribune & the miniſters of the Iewes
apprehended \Sc{Iesvs}, and bound him: \V and they brought him to Annas
firſt, for he was father in law to Caiphas, who was the high Prieſt of
that yeare. \V And
\CNote{\XRef{Io.~11,~49.}}
Caiphas was he that had giuen the counſel to the
Iewes, That it is expedient that one man die for the people.

\V
\CNote{\XRef{Mt.~26,~58.}
\XRef{Mr.~14,~54.}
\XRef{Lu.~22,~54.}}
And Simon Peter followed \Sc{Iesvs}, and another Diſciple. And that
Diſciple was knowen to the high Prieſt, and 
%%% o-2369
went in with \Sc{Iesvs} into
the court of the high Prieſt. \V But Peter ſtood at the doore
without. The other Diſciple therfore that was knowẽ to the high Prieſt,
went forth, and ſpake to the portreſſe, and brought in Peter. \V The
wench therfore that was portreſſe, ſaith to Peter: Art not thou alſo
of this mans Diſciples? He ſaith to her, 
\SNote{It is al one for a man to deny Chriſt, & that he is a Diſciple of
  Chriſt, or a Catholike, or a Chriſtian man, when he is demanded. 
\Cite{Aug tract.~123. in Io.}
for ſo Peter here denieth Chriſt,
in denying himſelf to be his Diſciple.}
I am not. \V And the ſeruants & miniſters ſtood at a fire of coles,
becauſe it was cold, and warmed themſelues. And with them was Peter alſo
ſtanding, and warming himſelf.

\V The high Prieſt therfore asked \Sc{Iesvs} of 
\Fix{this}{his}{obvious typo, fixed in other}
Diſciples, and of his doctrine. \V \Sc{Iesvs} anſwered him: I haue
openly ſpoken to the world: I haue alwaies taught in the Synagogue, and
in the Temple whither al the Iewes reſort together: and in ſecret I haue
ſpoken nothing. \V Why askeſt thou me? aske them that haue heard what I
haue ſpoken vnto thẽ: behold they know what things I haue ſaid. \V When
he had ſaid theſe things, one of the miniſters ſtanding by, gaue
\Sc{Iesvs} a blow, ſaying: Anſwereſt thou the high Prieſt ſo? \V
\Sc{Iesvs} anſwered him: If I haue ſpoken il, giue teſtimonie of euil:
but if wel, why ſtrikeſt thou me?

\V And Annas ſent him bound to Caiphas the high Prieſt. \V And Simon
Peter was ſtanding, and warming himſelf. They ſaid therfore to him: Art
not thou alſo of his Diſciples: He denied and ſaid: I am not. \V One of
the ſeruants of the high Prieſt ſaith to him, his coſin whoſe eare Peter
did cut off: Did not I ſee thee in the garden with him? Againe therfore
Peter denied: and forthwith the cocke crew.

\V
\CNote{\XRef{Mt.~27,~1.}
\XRef{Mr.~15,~1.}
\XRef{Luc.~23,~1.}}
They therfore bring \Sc{Iesvs} from Caiphas into the Palace. And it
was morning: and they went not in into the Palace, that they might not
be contaminated, but that they might eate the Paſche. \V Pilate therfore
went forth to them without, and ſaid: What accuſation bring you againſt
this man? \V They anſwered and ſaid to him: If he were not a
malefactour, we would not haue deliuered him vp to thee. \V Pilate
therfore ſaid to them: Take him you, and according to your law iudge
him. The Iewes therfore ſaid to
%%% 2529
him: It is not lawful for vs to kil any man. \V
\CNote{\XRef{Io.~12,~33.}
\XRef{Mt.~20,~19.}}
That the word of \Sc{Iesvs} might be fulfilled which he ſaid, ſignifying
what death he ſhould die.

\V
\CNote{\XRef{Mt.~27,~11.}
\XRef{Mr.~15,~2.}
\XRef{Lu.~23,~4.}}
Pilate therfore went into the Palace againe, and called \Sc{Iesvs},
and ſaid to him: Art thou the King of the Iewes? 
%%% o-2370
\V \Sc{Iesvs} anſwered:
Saieſt thou this of thy ſelf, or haue others told it thee of me? \V
Pilate anſwered: Why, am I a Iew?
\SNote{It pleaſed God, that Chriſt who was to die both for the Iewes &
  the Gentils, should be betraied of the one, and put to death by the
  other.}
Thy Nation, and the cheefe Prieſts haue deliuered thee vp to me: what
haſt thou done? \V \Sc{Iesvs} anſwered: My Kingdom is not of this
world. If my Kingdom were of this world, my miniſters verily would
ſtriue that I ſhould not be deliuered to the Iewes. But now my Kingdom
is not from hence. \V Pilate therfore ſaid to him: Art thou a King then?
\Sc{Iesvs} anſwered: Thou ſaieſt, that I am a King. For this was I
borne, and for this came I into the world: that I ſhould giue teſtimonie
to the truth. Euery one that is of the truth, heareth my voice. \V
Pilate ſaith to him: What is truth?

And when he had ſaid this, he went forth againe to the Iewes, and ſaith
to them: I find no cauſe in him. \V
\CNote{\XRef{Mt.~27,~15.}
\XRef{Mr.~15,~6.}
\XRef{Lu.~23,~17.}}
But you haue a cuſtome that I ſhould
releaſe one to you in the Paſche: wil you therfore that I releaſe vnto
you the King of the Iewes? \V They al therfore cried againe, ſaying: Not
him but Barabbas. And Barabbas was a theefe.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
