%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-05


\project douay-rheims


%%% 2493
%%% o-2328
\startChapter[
  title={Chapter 5}
  ]

\Summary{Curing
\MNote{The 2.~part: The Actes of Chriſt in Iewrie (hauing already begun
his ſolẽne Manifeſtation in Galilee.
\XRef{Mt.~4,~12})
the ſecond Paſche of his preaching.}
 a bedred man at the pond of miracle, becauſe he doth it
  on the Sabboth, the blind Iewes doe perſecute  him. 17.~And againe
  becauſe he ſaith that God is his natural Father. 19.~He thereupon
  continueth ſaying, the Fathers operation and his to be in euery thing
  al one, and that he shal doe greater things then theſe miraculous
  cures, to wit, 21.~quicken the dead in ſoule by ſinne, as being
  appointed Iudge of al, 28.~yea and quicken the dead in bodies alſo,
  incontinent iudging al vprightly. 31.~And that theſe are not bragges
  of his owne but his witneſſes to be 33.~Iohn Baptiſt, 36.~his owne
  miraculous workes, 37.~his Fathers voice at his Baptiſme, 39.~the
  Scriptures alſo, namely of Moyſes.}

%%% o-2329
After theſe things there was a feſtiual day of the Iewes, and \Sc{Iesvs}
%%% !!! This \Var's ſecond arg shpuld be empty !!!
went vp to Hieruſalem. \V And there is at Hieruſalem \Var{vpon}{.}
\SNote{By our latin text and the Greeke, this miraculous pond was in or
  vpon \Emph{Probatica}, that is, a place where the sheep to be
  ſacrificed, were kept. But by other latin copies, S.~Hierom, and ſome
  Greeke Fathers, \Emph{Probatica} is the very põd itſelf: ſo called
  becauſe the ſheepe of ſacrifice were there waſhed.}
Probatica
\LNote{A pond}{This is as great a wonder and worke as was in the old
  Law, yet neuer recorded in the Scripture before: the conditions &
  circumſtances of the ſame much to be diſtinctly weighed againſt the
  miſcreants of this time for many cauſes. Firſt,
\MNote{1.~Vertue of miracles giuen to creatures.}
that God without derogatiõ to his honour, yea to the great cõmendation
of it, doth giue vertue of miracles and cure to water or other
creatures. Secondly,
\MNote{2.~The ſame giuen ſpecially to ſanctified creatures.}
that he giueth ſuch vertues to theſe creatures ſpecially which be by vſe
& occupying in ſacred functions or otherwiſe, as it were ſanctified:
for this pond was it wherein the carcaſſes of ſheep (therfore called
Probatica) & other beaſts to be ſacrificed, were firſt waſhed, to which
being alwayes red
\CNote{\Cite{Hiero. de locus Heb. poſt med.}}
(as S.~Hierom ſaith) with the bloud of hoſtes, this
force was giuen, for the commendation of the Sacrifice of the Law there
offered. How much more may we acknowledge ſuch workes of God
miraculouſly done in or about the Sacrifice or Sacraments of the new
Teſtament, which faithleſſe men wholy reiect and condemne for
fables, becauſe they know not the Scriptures nor the power of
God. Thirdly, 
\MNote{Miracles done at one time more thẽ at another, ſpecially in
  greater ſolemnities.}
that this operation was giuen at one time more then another & rather on
great feſtiual daies then other vulgar times (for this was the
feaſt of Paſche or Pentecoſt) as daies more ſanctified, & when the
people made greater concurſe: which ſheweth that we ſhould not wonder to
ſee miracles done at the Memories and feaſtes of Martyrs or other great
Feſtiuities, more then at other places and times. Fourthly,
\MNote{4.~Angles and Saintes patrones & workers in places of miracles.}
that the Angles or ſome ſpecial Saints are Preſidents or Patrons of ſuch
places of miracle, and workers alſo vnder God of the effects that there
extraordinarily be done. Which ought to make Chriſtians leſſe doubt,
that the force of diuers waters in the world is iuſtly attributed by our
forefathers and good ſtories to the prayers and preſence of Saints,
which profane incredulous men referre only to nature, vntruly pretending
that God is more glorified by the workes of nature, which be of his
ordinarie prouidence, then by the graces of Miracle giuen to his Saints
or Angels by his extraordinarie prouidence. Fiftly,
\MNote{5.~Miracles in certaine places, & wrought vpon them that corporally
viſite the ſame. See
\Cite{S.~Auguſtine ep.~137.}}
that miracles be  not wrought on men by their faith only and as wel by
their preſence in ſpirit as in body, or vpon the parties deſire or
deuotion only, according to the Heretikes pretext that God is alike
preſent by his power & grace to euery man & place: & therfore that men
need not to goe from their owne houſes or countries to ſeeke holines or
health at the places of Chriſts or his Saints birth, death, memories:
for none could haue benefit of this water but he that could touch it,
and be in it corporally, and at that iuſt time when the water was in
motion by the Angel. Yea ſixtly,
\MNote{6.~Al reaſoning in theſe matters muſt yeald to Gods pleaſure.}
we may conſider that in ſuch caſes to make the matter more maruelous,
rare, and more earneſtly to be ſought for, and to ſignifie to vs that
God hath al ſuch extraordinary operations in his owne wil &
commandement, without al rules of our reaſons and queſtioning thereon,
none could be healed but that perſon who firſt could get into the pond
after the Angel came & ſtirred the ſame.  Seuenthly,
\MNote{7.~This water is a figure of Baptiſme.}
that theſe graces of corporal cures giuen to this water,
\CNote{\Cite{Hiero. cont. Lucifer. c.~3. to.~2.}}
prefigure the
like force of the Sacrament of Baptiſme for the cure of ſoules, though
we need not ſeeke the correſpondence thereof to the figure in euery
point. Laſtly,
\MNote{8.~Chriſt extraordinarily healeth and ſaueth without creatures.}
Chriſt by his power of excellencie and prerogatiue could and did heale
this poore man that could get no body to help him into the water,
becauſe he earneſtly and long deſired the remedie by God appointed, but
was excluded by neceſsitie: as our Lord ſaueth al ſuch as die without
Baptiſme, if they in their owne perſons earneſtly intended, deſired, and
ſought for the ſame.}
a pond which in hebrew is ſurnamed \Var{Bethſaida}{Betheſda}, hauing fiue
porches. \V In theſe lay a great multitude of ſicke perſons, of blind,
lame, withered, expecting the ſtirring of the water. \V And an Angel of
our Lord deſcended at a certaine time into the pond, and the water was
ſtirred. And he that had gone downe firſt into
%%% 2494
the pond after the ſtirring of the water, was made whole of
\Fix{whatſouer}{whatſoeuer}{obvious typo, fixed in other}
infirmitie he was holden. \V And there was a certaine man there that had
been eight and thirtie yeares in his infirmitie. \V Him when \Sc{Iesvs}
had ſeen lying, & knew that 
\TNote{\L{multũ tempus haberet,}}
he had now a long time, he ſaith to him: Wilt thou be made whole? \V The
ſicke man anſwered him: Lord, I haue no man, when the water is troubled,
to put me into the pond. For whiles I come, another goeth downe before
me. \V \Sc{Iesvs} ſaith to him: Ariſe, take vp thy bed, and walke. \V
And forthwith he was made whole: and he tooke vp his bed, and
walked. And it was the Sabboth that day. \V The Iewes therfore ſaid to
him that was healed: It is the Sabboth, thou maiſt not take vp thy
bed. \V He anſwered them: He that made me whole, he ſaid to me, take vp
thy bed, and walke. \V They asked him therfore, what is that man that
ſaid to thee, take vp thy bed, and walke? \V But he that was made whole,
knew not who it was. For \Sc{Iesvs} ſhronke aſide from the multitude
ſtanding in the place. \V Afterward \Sc{Iesvs} findeth him in the
Temple, & ſaid to him: Behold thou art made whole; 
\LNote{Sinne no more}{We
\MNote{Sinne the cauſe of ſicknes and infirmities.}
may gather hereby that this mans long infirmitie was for puniſhment of ſinnes,
and that men often attribute their ſicknes to other natural defects, and
ſeeke for remedies of the world in vaine, when the ſinne for which it
was ſent, remaineth, or is not repented of: and therfore that in al
infirmities men should firſt turne to God and goe to their Ghoſtly
father, and then cal for the worldly Phiſicions afterward.}
ſinne no more,leſt ſome worſe thing chance to thee. \V That man went his
way, and told 
the Iewes that it was \Sc{Iesvs} that made him whole. \V Thereupon the
Iewes perſecuted \Sc{Iesvs}, becauſe he did theſe things on the Sabboth.

\V But \Sc{Iesvs} anſwered them: My Father worketh vntil now; and I doe
worke. \V Thereupon therefore the Iewes ſought the more to kil him:
becauſe he did not only breake the Sabboth, but alſo he ſaid God was his
Father, making himſelf equal to God. \V \Sc{Iesvs} therfore anſwered,
and ſaid to them: Amen, amen, I ſay to you, the Sonne can not doe any
thing of himſelf, but that which he ſeeth the Father doing. 
%%% o-2330
For what things ſoeuer he doeth, theſe the Sonne alſo doeth in like
manner. \V For the Father loueth the Sonne, and ſheweth him al things
that himſelf doeth, and greater workes then theſe wil he ſhew him, that
you may maruel. \V For as the Father doth raiſe the dead and quickeneth:
ſo the Sonne alſo quickeneth whom he wil. \V For neither doth the Father
iudge any man: but al iudgement he hath giuen to the Sonne, \V that al
may honour the Sonne, as they doe honour the Father, who ſent him. \V
Amen, amen I ſay to you, that he which heareth my word, and beleeueth
him that ſent me, hath life euerlaſting: and he commeth not into
iudgement, but \Var{ſhal paſſe}{is paſſed} from death into life. \V
Amen, amen I ſay to you, that the houre commeth, and now it is when the
dead ſhal heare the voice of the Sonne of God, and they that haue heard,
ſhal liue. \V For as the Father hath life in himſelf; ſo he hath giuen
to the Sonne alſo to haue life in himſelf: \V and he hath giuen him
power to doe iudgement alſo becauſe he is the Sonne of man. \V Maruel
not at this, becauſe the houre commeth wherein al that are in the
graues, ſhal heare his voice, \V and they that haue
\MNote{Not faith only, but good & il deedes ſhal be counted, &
  accordingly rewarded at the day of iudgement.}
done good things, ſhal come forth into the reſurrection of life: but
they that haue done euil into the reſurrection of iudgement. \V I can
not of myſelf doe any thing. As I heare ſo I iudge: and my
%%% 2495
iudgement is iuſt, becauſe I ſeeke not my wil, but the wil of him that
ſent me. \V If I giue teſtimonie of myſelf, my teſtimonie is not
true. There is another that giueth teſtimonie of me: and 
\Fix{a}{}{obvious typo, fixed in other}% to prevent paragraph break
know that the teſtimonie is true which he giueth of me.

\V
\CNote{\XRef{Io.~1,~19.}}
You ſent to Iohn; and he gaue teſtimonie to the truth, \V But
\LNote{I receiue not}{Our Maiſter meaneth that mans teſtimonie is not
  neceſſarie to him, nor that the truth of his Diuinitie dependeth on
  worldly witneſſes, or mens commendations: though to vs ſuch
  teſtimonies be agreable and neceſſarie. And ſo for our inſtruction he
  vouchſafed to take the teſtimonies of Iohn the Baptiſt and Moyſes the
  Prophets: and departing out of this world, to ſend forth al his
  Apoſtles, and in them al Bishops and Lawful Paſtours, to be his
  witneſſes from Hieruſalem to the end of the world.}
I receiue not teſtimonie of man: but I ſay theſe things that you may be
  ſaued. \V He 
was the lampe burning and ſhining. And you would for a time reioyce in
his light. \V But I haue a greater teſtimonie then Iohn. For the workes
which the Father hath giuen me to perfit them: the very workes
themſelues which I doe, giue teſtimonie of me, that the Father hath ſent
me. \V And the Father that ſent me, himſelf hath
\CNote{\XRef{Mt.~3,~17.}}
giuen teſtimonie of me:
neither haue you heard his voice at any time, nor ſeen his ſhape, \V and
his word you haue not remaining in you: becauſe whõ he hath ſẽt, him you
beleeue not. \V 
\LNote{Search the Scriptures}{He
\MNote{Neither Iewes nor Heretikes find the truth, becauſe they ſearch
  not the Scriptures deeply, but read ſuperficially.}
reprehendeth the Iewes, that reading daily the Scriptures and
acknowledging that in them they should find life and ſaluation, they yet
looked ouer them ſo ſuperficially that they could not find therein him
to be \Sc{christ} their King, Lord, Life, and Sauiour. For the ſpecial
maiſters & Scribes of the Iewes then, were like vnto our Heretikes now,
who be euer talking and turning and shuffling the Scriptures, but are of
al men moſt ignorant in the deep knowledge thereof. And therfore our
Maiſter referreth them not to the reading only, learning them without
booke, or hauing the ſentences thereof gloriouſly painted or written in
their Temple, houſes, or coates: but to the deepe ſearch of the meaning
and myſteries of the Scriptures, which are not ſo eaſily to be ſeene in
the letter.}
Search the
\SNote{Catholikes ſearch the ſcriptures, and find
    there, Peters & his Succeſſours Primacie, the Real preſence, the
    Prieſts power to forgiue ſinnes, Iuſtificatiõ by faith and good
    workes, Virginitie preferred before matrimonie, breach of the vow of
    continencie damnable, Volontarie pouertie, Penãce, almes, & good
    deeds meritorious, diuers rewardes in heauen according to diuers
    merites, & ſuch like.}
ſcriptures, for you
%%% o-2331
 thinke in thẽ to haue life euerlaſting: & the ſame are they
that giue teſtimonie of me: \V and you wil not come to me that you may
haue life. \V Glorie of men I receiue not. \V But I haue knowen you,
that the loue of God you haue not in you. \V I am come in the name of my
Father, and you receiue me not: if
\SNote{He meaneth ſpecially Anti-Chriſt. How thẽ can the Pope be he,
ſeing the Iewes receiue him not.}
another ſhal come in his owne name, him you wil receiue. \V How can you
beleeue, that receiue glorie one of another: and the glorie which is of
God only, you ſeeke not? \V Thinke not that I wil accuſe you to the
Father. There is that accuſeth you, Moyſes, in whom you truſt. \V For if
you did beleeue Moyſes: you would perhaps beleeue me alſo. For of me he
hath written. \V And if you doe not beleeue his writings: how wil you
beleeue my wordes.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
