%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 2508
%%% o-2345
\startChapter[
  title={Chapter 9}
  ]

\Summary{To shew that by his Baptiſme (being the Sacrament of illumination
  or faith) he wil take away the blindnes of the world, he giueth with
  ſtrange ceremonies ſight to one borne blind. 8.~By which wonderful
  miracle (the atteſtation of the partie himſelf and of his parents
  concurring) firſt the neighbours, then alſo the Phariſees themſelues
  are plainely confounded. Yet ſo obſtinate they are, that becauſe it
  was the Sabboth when he wrought it, they inferre that he is not of
  God: yea and throw out of their Synagogue the partie for confeſsing
  him. 35.~But our Lord, receiueth him; 39.~and foretelleth by this
  occaſion, the excecation, of the Iewes (becauſe of their wilful
  obſtinacie) and illumination of the Gentils who confeſſe their owne
  blindnes.}

And \Sc{Iesvs} paſſing by, ſaw a man blind: from his natiuitie; \V and
his Diſciples asked him: Rabbi, who hath ſinned, this man, or his
parents, that he ſhould be borne blind? \Sc{Iesvs} anſwered: \V 
\SNote{Though many infirmities fal for ſinne, yet not al: ſome comming
  for probation, and ſome ſent that God by the cure thereof may be
  glorified.}
Neither hath this man ſinned, nor his parents; but that the workes of
God may be manifeſted in him. \V I muſt worke the workes of him that
ſent me, whiles it is day. The night commeth,
\SNote{The time of working by our deedes, and meriting, is in this life:
  after death 
  we can deſerue no more by our deedes , but muſt only receiue good or
  il, according to 
  the difference of workes here.}
when no man can worke. \V As long as I am in the world, I am the light
of the world. \V When he had ſaid theſe things, he ſpit on the ground,
and 
\LNote{Made clay}{Chriſt
\MNote{External ceremonies.}
that could haue cured this man by his only wil or word, yet vſed
certaine creatures as his inſtruments in working, and diuers
circumſtances and ceremonies, clay, water, annoynting, waſhing, &c. No
maruel then that he and his Church vſe ſuch diuerſities of Sacraments
and ceremonies external in curing our ſoules.}
made clay of the ſpittle, and ſpred the clay vpon his eyes, \V and ſaid
to him: Goe,
\SNote{This was a figure of Baptiſme, to which al men borne in ſinne &
blindnes are ſent for health & ſight.
\Cite{Ambr. li.~3. c.~2, de Sacramentu.}}
waſh in the poole of Siloe, which is interpreted, \Emph{Sent}. He went
therfore, and waſhed; and he came ſeeing.

\V Therfore the neighbours, and they which had ſeen him before, that he
was a beggar, ſaid: Is not this he that ſate, and begged? Others ſaid:
That this is he. \V But others: No, not ſo, but he is like him. But he
ſaid: That I am he. \V They ſaid therfore to him: How were thine eyes
opened? \V He anſwered: That man that is called \Sc{Iesvs}, made clay,
and anointed mine eyes, and ſaid to me: Goe to the poole of Siloe, and
waſh. And I went, and waſhed, and ſaw. \V And they ſaid to him: Where is
he? He ſaith: I know not. \V They bring him that had been
%%% 2509
blind, to the Phariſees. \V And it was the Sabboth when \Sc{Iesvs} made
the clay, and opened his eyes.

\V Againe therfore the Phariſees asked him, how he ſaw. But he ſaid to
them: He put clay vpon mine eyes, & I waſhed; and I ſee. \V Certaine
therfore of the Phariſees ſaid: This mã is not of God, that keepeth not
the Sabboth. But others ſaid: How can a man that is a ſinner doe theſe
ſignes? And there 
%%% o-2346
was a ſchiſme among them. \V They ſay therfore to the blind againe: Thou,
what ſaieſt thou of him that opened thine eyes? And he ſaid: That he is
a Prophet. \V The Iewes therfore did not beleeue of him, that he had
been blind and ſaw, vntil they called the parents of him that ſaw, \V
and asked them, ſaying: Is this your ſonne, whom you ſay that he was
borne blind? how then doth he now ſee? \V His parents anſwered them, and
ſaid: We know that this is our ſonne, and that he was borne blind; \V
but how he now ſeeth, we know not, or who hath opened his eyes, we know
not, aske himſelf; he is of age, let him ſpeake of himſelf. \V Theſe
things his parents ſaid, becauſe they feared the Iewes. For the Iewes
had now conſpired, that if any man ſhould confeſſe him to be
\Sc{Christ}, he ſhould be
\LNote{Put out of the Synagogue}{The
\MNote{Heretical trãſlation.}
Heretikes vntruly tranſlate here (&
\XRef{v.~35})
\Emph{Excommunicate}:
to make the ſimple conceaue the Churches Excommunication to be no other,
or no better, or no more rightly vſed againſt them, then this 
\MNote{Caſting out of the Synagogue.}
caſting
out of the Synagogue of ſuch as confeſſed our Sauiour. They might as wel
haue Tranſlated for Synagogue, Church: for the old Teſtament, the new:
for Law, grace: for fleſh, ſpirit: for Moyſes, Chriſt. For no leſſe
difference is there between caſting out of the Synagogue, and
Excommunication. Beſides that, not euery one which was not of the Iewes
Synagogue, was therfore out of communion of the Faithful, many true
beleeuers being in other partes of the world not ſubiect to the Iewes
Synagogue, Law, nor Sacraments. And therfore it was not al one to be out
of the Synagogue, and 
\MNote{The Churches Excommunication.}
to be excommunicated, as now, whoſoeuer is out of
the Churches communion, either by his owne wil, or for his iuſt deſerts
thruſt out of it by the ſpiritual Magiſtrate, 
\MNote{See in the
\XRef{Annot. Mt.~18,~17.}}
he is quite abandoned out
of al the ſocietie of Saints in Heauen and earth, ſo long as he ſo
continueth. 

As for the cauſe of thruſting this poore man & ſuch other out of the
Synagogue, & excommunicating Heretikes, there is as great oddes as
betwixt Heauen & hel: he being vſed ſo for following Chriſt and his
Church, theſe for forſaking Chriſt and his Church. Some more agreement
there is between that corrupt ſentence of the Iewes againſt the
followers of Chriſt, & the pretended excõmunication executed againſt
Catholike men by our Heretikes: although in truth there is no great
reſemblance. For, the Iewes though they abuſed their power ſometimes,
yet had they authoritie indeed by Gods law ſo to puniſh contemners of
their Law, & therfore it was feared and reſpected euen of good men. 
\MNote{The Heretikes ridiculous Excõmunication.}
But
the excommunication vſed by Heretikes againſt Catholikes or any offenders,
is not to be reſpected at al, being no more but a ridiculous vſurpation
of the Churches right and faſhion of the ſame. For, out of their
Synagogues al faithful men ought to flye, and not tarie to be thruſt
out: according to the warning giuen againſt Core & Dathan:
\CNote{\XRef{Num.~16.}}
\Emph{Be ye
  ſeparated from their tabernacles, leſt you be wrapped in their
  ſinnes.}}
put out of the Synagogue. \V Therfore did his parents ſay: That he is of
age, aske himſelf. \V They therfore againe called the man that had been
blind, and ſaid to him: 
\SNote{So ſay the Heretikes whẽ they derogate from miracles done by
Saints or their Relikes, phariſaically pretending the glorie of
God. As though it were not Gods glorie, whẽ his Saints doe it by his
power & vertue: yea his greater glorie, that doth ſuch things by his
ſeruants, & by the meaneſt things belonging to thẽ, as Peters ſhadow
\XRef{Act.~5.}
& Paules napkins
\XRef{Act.~19.}}
Giue glorie to God. We know that this man is a ſinner. \V He therfore
ſaid to them: Whether he be a ſinner, I know not: one thing I know, that
whereas I was blind, now I ſee. \V They ſaid therfore to him: What did
he to thee? how did he open thine eyes? \V He anſwered them: I haue now
told you, and you haue heard; why wil you heare it againe? wil you alſo
become his Diſciples? \V They reuiled him therfore, & ſaid: Be thou his
Diſciple: but we are the Diſciples of Moyſes. \V We know that to Moyſes
God did ſpeake; but this man we know not whence he is. \V The man
anſwered and ſaid to them: For in this it is maruelous that you know not
whence he is, and he hath opened mine eyes. \V And we know that ſinners
God doth not heare. But if a man be a ſeruer of God, and doe the wil of
him, him he heareth. \V From the beginning of the world it hath not been
heard that any man hath opened the eyes of one borne blind. \V Vnles
this man were of God, he could not doe any thing. \V They anſwered and
ſaid to him: Thou waſt wholy borne in ſinnes, and doeſt thou teach vs?
And they caſt him forth.

\V \Sc{Iesvs} heard that they caſt him forth; & when he had found him,
he ſayd to him: Doeſt thou beleeue in the Sonne of God? \V He anſwered,
and ſaid: Who is he Lord, that I may beleeue in him? \V And \Sc{Iesvs}
ſaid to him: Both thou haſt ſeen him; and he that talketh with thee, he
it is. \V But he 
%%% o-2347
ſaid, I beleeue Lord. And falling downe he adored
him. \V And \Sc{Iesvs} ſaid to him: For iudgement came I into this world, 
\SNote{By this we ſee that this miracle was not only maruelous &
  beneficial to the blind, but alſo ſignificatiue of taking away
  ſpiritual blindneſſe.}
that they that ſee not, may ſee; and they that ſee, may become blind. \V
And certaine of the Phariſees that were with him, heard; and they ſaid
to him: Why, are we alſo blind? \V \Sc{Iesvs} ſaid to them: If you were
blind, you ſhould not haue ſinne, but now you ſay, That we ſee. Your
ſinne remaineth.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
