%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 2522
%%% o-2361
\startChapter[
  title={Chapter 15}
  ]

\Summary{He exhorteth them to abide in him (that is his Church, being
  the true vine, and not the Synagogue of the Iewes any more) 9.~and in
  his loue, louing one another, and keeping his commandements:
  13.~shewing how much he accounteth of them, by this that he dieth for
  them, 15.~and reuealeth vnto them the ſecrets of Heauen, 17.~and
  appointeth their fruits to be perpetual: 18.~confirming them alſo
  againſt the perſecutions & hatred of the obſtinate Iewes.}

%%% o-2362
I am the true vine; and my Father is the husband-man. \V
\LNote{Euerie branch in me}{By this it appeareth that Chriſt hath ſome
  branches in his bodie myſtical that be fruitles: therfore il liuers
  alſo may be members of Chriſt and the Church, though none can be ſaued
  but fruitful branches.}
Euery branch in me, not bearing fruit, he wil take it away: & euery one
that beareth fruit, he wil purge it, that it may bring
\LNote{More fruit}{This proueth that a iuſt man may continually
  increaſe in iuſtice & ſanctification ſo long as he liueth.}
more fruit. \V Now you are cleane for the
\SNote{S.~Auguſtin expoundeth it of the Sacramental word of Baptiſme and
not as Heretikes doe, of preaching only.
\Cite{Tract.~80. in Io.}}
word which I haue ſpoken to you. \V Abide in me: and I in you. As the
brãch cã not beare fruit of it ſelf,
\LNote{Vnles it abide}{Whoſoeuer
\MNote{No ſaluation out of the Church.}
by Hereſie, or Schiſme, or for anie other cauſe, is cut off, or
ſeparated from the Church, he can doe no meritorious worke to
ſaluation. Neither can be heard, pray he neuer ſo much in Schiſme,
becauſe he is not in the bodie of Chriſt, which is a condition
neceſſarily required in prayer.
\XRef{ver.~7.}}
vnles it abide in the vine, ſo you neither
\LNote{Vnles you abide}{Theſe
\MNote{No man ſure of perſeuerance.}
conditional ſpeaches, \Emph{If you remaine in the vine}, \Emph{If you
  keepe my commandements}, and ſuch like, giue vs to wit that we be not
ſure to perſiſt, perſeuer, nor to be ſaued, but vnder condition to be
fulfilled by vs.
\Cite{Aug. de corrupt. & gra. c.~13.}}
vnles you abide in me. \V I am the vine: you are the brãches. He that
abideth in me, and I in him, the ſame beareth much fruit: for without me
you can doe nothing. \V If any abide not in me: he ſhal be caſt forth as
the branch, and ſhal wither, and they ſhal gather him vp, and caſt him
into the fire, and he burneth. \V If you
%%% !!! Extra note mark? Notes rearranged? See other.
%\SNote{}
abide in me, & my wordes abide in you: you ſhal aske what thing ſoeuer
you wil, and it ſhal be done to you. \V In this my Father is glorified;
that you bring very much fruit, and become my Diſciples. \V As my Father
hath loued me, I alſo haue loued you. Abide in my loue. \V If you
\LNote{Keepe my precepts}{This
\MNote{Onlie faith ſuficeth not to ſaluation.}
careful and often admonition of keeping his commandements, proueth that
a Chriſtian mans life is not only or principally in faith, but in good
workes.}
keepe my precepts, you ſhal abide in my loue: as I alſo haue kept my
Fathers precepts, and doe abide in his loue. \V Theſe things I haue
ſpoken to you, that my ioy may be in you, and your ioy may be filled. \V
\CNote{\XRef{Io.~13,~34.}}
This is
%%% 2523
my precept, that you loue one another, as I haue loued you. \V Greater
loue then this no man hath, that a man yeald his life for his
freinds. \V You are my freinds, if you doe the things that I cõmand
you. \V Now I cal you not ſeruants: for the ſeruãt knoweth not what his
Lord doeth. But you I haue called freinds; becauſe al things whatſoeuer
I heard of my Father, I haue notified vnto you. \V You choſe not me, but
I choſe you; and haue appointed you: that you goe, & bring fruit; and
your fruit abide: that whatſoeuer you aske the Father in my name, he may
giue it you. \V Theſe things I command you, that you loue one another.

\V If the world hate you, know ye that it hath hated me before you. \V
If you had been of the world, the world would loue his owne. But becauſe
you are not of the world, but I haue choſen you out of the world,
therfore the world hateth you. \V Remember my word that I ſaid to you:
\CNote{\XRef{Io.~13,~16.}
\XRef{Mt.~10,~24.}
\XRef{Luc.~6,~40.}}
The ſeruant is not greater then his Maiſter. If they haue perſecuted me,
you alſo wil they perſecute. If they haue kept my Word, 
\SNote{He foreſheweth that many wil not obey the Churches wordes; & no
  maruel, becauſe they cõtemned Chriſts owne precepts.}
yours alſo wil they keepe. \V But al theſe things they wil 
%%% o-2363
doe to you
for my name ſake; becauſe they know not him that ſent me. \V If I had
not come, and ſpoken to them, they ſhould not haue ſinne: but now they
haue no excuſe of their ſinne. \V He that hateth me, hateth my Father
alſo. \V
\LNote{If I had}{If the Iewes had not ſinned by refuſing Chriſt, in caſe
he had not done greater miracles then any other: then were it a great
folly of Catholikes to beleeue Luthers or Caluins new opinions without any
miracles at al.}
If I had not done among them workes that no other man hath done, they
ſhould not haue ſinne: but now both they haue ſeen, and they doe hate
both me and my Father. \V But that the word may be fulfilled, which is
written in their law:
\CNote{\XRef{Pſ.~24,~19.}}
\Emph{That they hated me gratis.} \V But when the Paraclete commeth
\LNote{Whom I wil ſend}{The Holy Ghoſt is ſent by the Sonne, therfore he
\Fix{preceedeth}{proceedeth}{obvious typo, fixed in other}
from him alſo, as from the Father: though the late Schiſmatical Greekes
thinke otherwiſe.}
whom I
\CNote{\XRef{Act.~2,~1.}}
wil ſend you from the Father, the Spirit of truth, which
proceedeth from the Father, he ſhal giue teſtimonie of me: and
\LNote{You shal giue}{He vouchſafeth to ioyne together the teſtimonie of
the Holy Ghoſt, and of the Apoſtles: that we may ſee the teſtimonie of
truth ioyntly to conſiſt in the Holy Ghoſt and in the Prelates of the
Church. See
\XRef{Annot. Act.~25. v.~28.}}
you ſhal giue teſtimonie, becauſe you are with me from the beginning.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
