%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-11


\project douay-rheims


%%% 2513
%%% o-2350
\startChapter[
  title={Chapter 11}
  ]

\Summary{He commeth once againe into Iewrie boldly (the time that he
  would be killed of them, being not yet come) and raiſeth Lazarus foure
daies buried. 47.~At which miracle the blind malice of the Rulers ſo
increaſeth, that in Councel they conclude to make him away. Howbeit the
high Prieſt prophecieth vnawares, of the ſaluation of the world by his
death. 54.~He thereupon goeth againe out of the way.}

And there was a certaine ſicke man, Lazarus of Bethania, of the towne of
Marie & Martha her ſiſter. \V (And Marie was ſhe
\CNote{\XRef{Lu.~7,~37.}
\XRef{Mat.~26,~7.}
\XRef{Mar.~14,~3.}
\XRef{Io.~12,~3.}}
that anointed our Lord
with ointemẽt, & wiped his feete with her haire: whoſe brother Lazarus
was ſicke.) \V His ſiſters therfore ſent to him ſaying: Lord, behold, he
whom thou loueſt, is ſicke. \V And \Sc{Iesvs} hearing, ſaid to them:
This ſickneſſe is not to death, but for the glorie of God: that the
Sonne of God may be glorified by it. \V And \Sc{Iesvs} loued Martha, and
her ſiſter Marie, and Lazarus. \V As he heard therfore that he was
ſicke, then he taried in the ſame place two daies: \V then after this he
ſaith to his Diſciples: Let vs goe into Iewrie againe. \V The Diſciples
ſay to him: Rabbi, now the Iewes ſought to ſtone thee; and 
%%% o-2351
goeſt thou
thither againe? \V \Sc{Iesvs} anſwered: Are there not twelue houres of
the day? If a man walk in the day, he ſtumbleth not: becauſe he ſeeth
the light of this world: \V but if he walke in the night, he ſtumbleth,
becauſe the light is not in him. \V Theſe things he ſaid; and after this
he ſaith to them: Lazarus our freind ſleepeth; but I goe that I may
raiſe him from ſleep. \V His Diſciples therfore ſaid: Lord, if he ſleep,
he ſhal be ſafe. \V But \Sc{Iesvs} ſpake of his death; & they thought
that he ſpake 
\TNote{\L{de dormitione ſomni}}
of the ſleeping of ſleep. \V Then therfore \Sc{Iesvs} ſaid
to them plainely: Lazarus is dead; \V and I am glad for your ſake, that
you may beleeue, becauſe I was not there: but let vs goe to him. \V
Thomas therfore, who is called Didymus, ſaid to his Condiſciples: Let
vs alſo goe, to die with him.

\V \Sc{Iesvs} therfore came, and found him now hauing been foure daies
in the graue. \V 
\Fix{}{(}{obvious typo, fixed in other}% to prevent a paragraph break
And Bethania was nigh to Hieruſalem about fifteene furlonges.) \V And
many of the Iewes were come to Martha and Marie, to comfort them
concerning their brother. \V Martha therfore when ſhe heard that
\Sc{Iesvs} was come, went to meet him; but Marie ſate at home. \V Martha
therfore ſaid to \Sc{Iesvs}: Lord if thou hadſt been here, my brother
had not died. \V But now alſo I know that what things ſoeuer thou ſhalt
aske of God, God wil giue thee. \V \Sc{Iesvs} ſaith to her: Thy brother
ſhal riſe againe. \V Martha ſaith to him: I know that he ſhal riſe
againe in the reſurrection, in the laſt day. \V \Sc{Iesvs} ſaid to her:
I am the reſurrection and the life; he that beleeueth in me, although he
be dead, ſhal liue. \V And euery one that liueth, and beleeueth in me,
ſhal not die for euer. Beleeueſt thou this? \V She ſaith to him: Yea
Lord, I haue beleeued that thou art \Sc{Christ} the Sonne of God that
art come into this world.

\V And when she had ſaid theſe things, ſhe went, and called Marie her
ſiſter ſecretly, ſaying: The Maiſter is come, & calleth thee. \V She,
%%% 2514
when ſhe heard, riſeth quickly & cõmeth to him. \V For \Sc{Iesvs} was
not yet come into the towne: but he was yet in that place where Martha
had met him. \V The Iewes therfore that were with her in the houſe and
did comfort her, when they ſaw Marie that she roſe quickly and went
forth, followed her, ſaying: That ſhe goeth to the graue, to weepe
there. \V Marie therfore when 
%%% o-2352
she was come where \Sc{Iesvs} was, ſeeing
him, fel at his feete, and ſaith to him: Lord, if thou hadſt been here,
my brother had not died. \V \Sc{Iesvs} therfore when he ſaw her,
weeping, and the Iewes that were come with her, weeping, he groned in
ſpirit, and troubled himſelf, \V and ſaid: Where haue you laid him? They
ſay to him: Lord, come and ſee. \V And \Sc{Iesvs} wept. \V The Iewes
therfore ſaid: Behold how he loued him. \V But certaine of them ſaid:
Could not he that
\CNote{\XRef{Io.~9,~6.}}
opened the eyes of the blind man, make that this man
should not die? \V \Sc{Iesvs} therfore againe groning in him ſelf,
commeth to the graue: and it was a caue; and a ſtone was laid ouer
it. \V \Sc{Iesvs} ſaith: Take away the ſtone. Martha the ſiſter of him
that was dead, ſaid to him: Lord, now he ſtinketh, for he is now of
foure daies. \V \Sc{Iesvs} ſaith to her: Did not I ſay to thee, that if
thou beleeue, thou shalt ſee the glorie of God? \V They tooke therfore
the ſtone away. And \Sc{Iesvs} lifting his eyes vpward, ſaid: Father, I
giue thee thankes that thou haſt heard me. \V And I did know that thou
doeſt alwaies heare me, but for the people that ſtandeth about, haue I
ſaid it, that they may beleeue that thou haſt ſent me. \V When he had
ſaid theſe things, he cried with a loud voice: Lazarus, come forth. \V
And forthwith he came forth that had been dead, bound feete and handes
with winding bandes, and his face was tied with a napkin. \Sc{Iesvs}
ſaid to them:
\SNote{
\Cite{S.~Cyril. li.~7. c.~vlt. in Io.}
and
\Cite{S.~Auguſt. Tract.~49. in Io.}
apply this to the Apoſtles & Prieſts authoritie of abſoluing ſinners:
affirming Chriſt to reuiue none from ſinne, but in the Church and by the
Prieſts miniſterie.} 
Looſe him, and let him goe.

\V Many therfore of the Iewes that were come to Marie & Martha, & had
ſeẽ the things that \Sc{Iesvs} did, beleeued in him. \V And certaine of
thẽ went to the Phariſees, & told thẽ the things that \Sc{Iesvs} did. \V
The cheefe Prieſts therfore & the phariſees gathered a Councel, & ſaid:
What doe we, for this man doeth many ſignes. \V If we let him alone ſo,
al wil beleeue in him: & the Romanes wil come, & 
\SNote{Al men, but ſpecially Nations muſt take heed, that whiles to ſaue
their temporal ſtate, they forſake God, they loſe not both, as the Iewes
did.
\Cite{Aug. tract.~47. in Io.}}
take away our place & Nation. \V But one of them named Caiphas, being
the high Prieſt of that yeare, ſaid to them: You know nothing, \V
neither doe you cõſider: that it is expedient for vs that one man die
for the people, & the whole Nation periſh not. \V And this he ſaid not
of himſelf: but
\LNote{Being the high Prieſt}{Maruel
\MNote{The priuiledge of the office & order, though in a wicked perſon.}
not that Chriſt preſerueth his truth in the Church as wel by the
vnworthy as the worthy Prelates thereof: the guifts of the Holy Ghoſt
following their Order and office, as we ſee here in Caiphas, & not their
merites or perſon. And if this man being many waies wicked, and in part
an vſurper, and the Law & Prieſthood being to decline & giue place to
Chriſts new ordinance, had yet ſome aſsiſtance of God for vtterance of
truth which himſelf meant not, nor knew not: how much more may we be
aſſured, that Chriſt wil not leaue Peters Seat,
\CNote{\XRef{Luc.~22,~23.}}
whoſe faith he promiſed
should neuer faile, though the perſons which occupie the ſame, were as
il as the blaſphemous & malitious mouthes of Heretikes doe affirme.}
being the high Prieſt of that yeare, he prophecied that \Sc{Iesvs}
should die for the Nation. \V And not only for the Nation, but to gather
into one the children of God that were diſperſed. \V From that day
therfore they deuiſed to kil him. \V \Sc{Iesvs} 
%%% o-2353
therfore walked no more
openly among the Iewes, but he went into the countrie beſide the deſert
vnto a citie that is called Ephrem, and there he abode with his
Diſciples.

\V And the Paſche of the Iewes was at hand: & many of the countrie went
vp to Hieruſalem before the Paſche to ſanctifie themſelues. \V They
ſought \Sc{Iesvs} therfore; and they communed one with another, ſtanding
in the Temple: What
%%% 2515
thinke you, in that he is not come to the feſtiual day? And the cheefe
Prieſts & Phariſees had giuen commandement, that if any man ſhould know
where he was, he ſhould tel, that they might apprehend him.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
