%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-03


\project douay-rheims


%%% 2488
%%% o-2322
\startChapter[
  title={Chapter 3}
  ]

\Summary{He teacheth Nicodemus, that to come to the Kingdom of God,
  Baptiſme is neceſsarie, as being our Regeneration.  10.~Which point
  Nicodemus as then not vnderſtanding.  11.~he sheweth that they muſt
  beleeue him, and what good cauſe there is for them ſo to doe.
  23.~After this he alſo baptized, & Iohn likewiſe at the ſame time,
  25.~Wherupon a queſtion being moued, whether Baptiſme is better, 
  \Fix{15}{27}{obvious typo, fixed in other}
  Iohn anſwereth it by ſaying, that he is ſo farre inferiour to Chriſt,
  as a mere man to God moſt high.}

And there was a man of the Phariſees, named Nicodemus, a Prince of the
Iewes.  \V This man came to \Sc{Iesvs} by night, and ſaid to him: Rabbi,
we know that thou art come from God a Maiſter; for no man can doe theſe
ſignes which thou doeſt, vnles God be with him. \V \Sc{Iesvs} anſwered,
and ſaid to  him: Amen, Amen I ſay to thee, vnles a man be borne againe,
he can not ſee the Kingdom of God. \V Nicodemus ſaid to him: How can a
man be borne, when he is old? can he enter into his mothers wombe againe
& be borne? \V \Sc{Iesvs} anſwered: Amen, Amen I ſay to thee, vnles a
man be 
\LNote{Borne againe of water}{As
\MNote{Baptiſme in water neceſſarie to ſaluatiõ.}
no man can enter into this world nor haue his life & being in the ſame,
except he be borne of his carnal parents: no more can a man enter into
the life & ſtate of grace which is in Chriſt, or attaine to life
euerlaſting, vnles he be borne and baptized of water and the Holy
Ghoſt.  Whereby we ſee firſt, this Sacrament to be called our
regeneration or ſecond birth, in reſpect of our natural and carnal which
was before. Secondly, that this Sacrament conſiſteth of an external
element of water, and internal vertue of the Holy Spirit: wherein it
excelleth Iohns Baptiſme, which had the external elemẽt, but not the
ſpiritual grace.  Thirdly, that no mã can enter into the Kingdom of God,
nor into the fellowſhip of Holy Church, without it.  Whereby the
\CNote{\Cite{Aug. hares.~88.}}
Pelagians, & Caluiniſts be condemned, that promiſe life euerlaſting to
yong children that die without Baptiſme, and al other that thinke only
faith to ſerue, or the external element of water ſuperfluous or not
neceſſarie: our Sauiours wordes being plaine & general.  Though
\MNote{Baptiſme in water in two caſes otherwiſe ſupplied.}
 in this
caſe, God which hath not bound his grace, in reſpect of his owne
freedom, to any Sacrament, may and doth accept them as baptized, which
either are martyred before they could be baptized, or els depart this
life with vow and deſire to haue that Sacrament, but by ſome remedileſſe
neceſſitie could not obtaine it.  Laſtly, it is proued that this
Sacrament giueth grace \L{ex opere operato}, that is, of the worke
itſelf (which al Proteſtants denie) becauſe it ſo breedeth our ſpiritual
life in God, as our carnal birth giueth the life of the world.}
borne againe of water and the Spirit, he can not enter into the
Kingdom of God. \V That which is borne of the fleſh, is fleſh: & that
which is borne of the Spirit, is ſpirit. \V Maruel not, that I ſaid to
thee, You muſt be borne againe. \V The Spirit breatheth where he wil;
\SNote{We follow rather S.~Aug.\ & thoſe anciẽt Fathers, which
  moſt cõmonly vnderſtand this place of the Holy Ghoſt, & not of the winde:
  although both ſenſes be good.}
and thou heareſt his voice, but thou knoweſt not whence he commeth and
whither he goeth: ſo is euery one that is borne of the Spirit. \V
Nicodemus anſwered, & ſaid to him: How can theſe things be done? \V
\Sc{Iesvs} anſwered, and ſaid to him: Thou art a 
%%% o-2323
Maiſter in Iſrael, and
art thou ignorant of theſe things? \V Amen, Amen I ſay to thee, that we
ſpeeke that which we know, and that which we haue ſeen we teſtifie, and
our teſtimonie you receiue not. \V If I haue ſpoken to you earthly things,
and you beleeue not: how if I ſhal ſpeake to you heauenly things, wil
you beleeue? \V And no man hath aſcended into Heauen, but he that
deſcended from Heauen, the Sonne of man which is in Heauen. \V And as
\CNote{\XRef{Nu.~21,~9.}}
Moyſes exalted the ſerpent in the deſert, ſo muſt the Sonne of man be
exalted: \V that euery one which beleeueth in him, periſh not, but may
haue life euerlaſting. \V For ſo God loued the world, that he gaue his
only-begotten Sonne, that euery one that beleeueth in him, periſh not,
but may haue life euerlaſting.  \V For God ſent not his Sonne into the
world, to iudge the world, but that the world may be ſaued by him. \V He
that beleeueth in him, is not iudged. But he that doth not beleeue, 
\LNote{Is iudged already}{He
\MNote{Euery infidel, & namely Heretikes, are iudged already.}
that beleeueth in Chriſt with faith which worketh by charitie
\CNote{\XRef{Gal.~5,~6.}}
(as the
Apoſtle ſpeaketh) ſhal not be condemned at the later day nor at the
houre of his death: but the Infidel, be he Iew, Pagan, or Heretike, is
already (if he die in his incredulitie) by his owne profeſſion and
ſentence condemned, and ſhal not come to iudgement either particular or
general, to be diſcuſſed according to his workes of mercie done or
omitted. In which ſenſe
\CNote{\XRef{Tit.~3,~11.}}
S.~Paul ſaith that the obſtinate Heretike is
condemned by his owne iudgement, preuenting in himſelf, of his owne free
wil, the ſentence both of Chriſt and of the Church.}
is already iudged: becauſe he hath not beleeued in the name of the
only-begotten Sõne of
%%% 2489
God. \V And this is the iudgment: becauſe the light is come into the
world, and men haue loued the darkeneſſe rather then the light: for
their workes were euil.  \V For euery one that doth il, hateth the
light, and commeth not to the light, that his workes may not be
controuled. \V But he that doth veritie, commeth to the light, that his
workes may be made manifeſt, becauſe they were done in God.

\V After theſe things \Sc{Iesvs} came and his Diſciples into the
countrie of Iewrie; & there he abode with them, and baptized. \V And
Iohn alſo was baptizing in Ænon beſide Salim; becauſe there was much
water there, and they came, and were baptized. \V For Iohn was not yet
caſt into priſon. \V And there roſe a queſtion of Iohns Diſciples with
the Iewes concerning purification. \V And they came to Iohn, and ſaid to
him: Rabbi, he that was with thee beyond Iordan, to whom
\CNote{\XRef{Io.~1,~19.}}
thou didſt giue
teſtimonie, behold he baptizeth, and al come to him. \V Iohn anſwered
and ſaid: A man can not receiue any thing, vnleſſe it be giuen him from
Heauen. \V Your ſelues doe beare me witneſſe,
\CNote{\XRef{Io.~1,~20.}}
that I ſaid, I am not
\Sc{Christ}; but that I am ſent before him.  \V He that hath the bride,
is the bridegrome: but the freind of the bridegrome that ſtandeth and
heareth him, reioyceth with ioy for the voice of the bridegrome.  This
my ioy therfore is 
%%% o-2324
filled. \V He muſt increaſe, and I diminiſh. \V
\LNote{He that commeth from aboue}{As
\MNote{The excellẽcie of Chriſts power & graces.}
though he ſhould ſay: No maruel that men reſort to Chriſt ſo faſt and
make leſſe account of me.  For, his Baptiſme and his preaching and his
Perſon are al from Heauẽ immediatly.  He bringeth al from the very
boſome, mouth, and ſubſtance of God his Father.  Whatſoeuer is in me, is
but a litle drop of his grace.  His ſpirit & graces are aboue al
meaſures or mẽs guifts,
 euen according to his Manhood: and al power temporal & ſpiritual, the
 Kingdom & the Prieſthood, and al ſoueraigntie in Heauen and earth are
 beſtowed vpon him as he is man alſo.}
He that commeth from aboue, is aboue al. He that is of the earth, of the
earth he is, and of the earth he ſpeaketh. He that commeth from Heauen,
is aboue al.  \V And what he hath ſeen and heard, that he teſtifieth:
and his teſtimonie no men receiueth. \V He that hath receiued his
teſtimonie, hath ſigned that God is true. \V For he whom God hath ſent,
ſpeaketh the wordes of God. For God doth not giue the ſpirit by
meaſure. \V The Father loueth the Sonne: & he hath giuen al things in
his hand. \V He that beleeueth in the Sonne: hath life euerlaſting: but
he that is incredulous to the Sonne, ſhal not ſee life, but the wrath of
God remaineth vpon him.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
