%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-16


\project douay-rheims


%%% 2524
%%% o-2363
\startChapter[
  title={Chapter 16}
  ]

\Summary{The cauſe why he fortelleth them their perſecution by the
  Iewes, is, that they be not afterward ſcandalized thereat. 6.~Though
  they thinke this heauie newes, it is for their vantage that he
  departeth, becauſe of the great benefits that they shal receiue by
  the comming then of the Holy Ghoſt, who shal alſo be his witnes
  againſt his enemies. 16.~Although in this world they shal ſo be
  perſecuted, yet to his heauenly Father they and their praiers made in
  his name, shal be moſt acceptable, and at length the child (that is,
  Chriſt in al his members) being borne, their ioy shal be ſuch as no
  perſecutour can take from them. 31.~Howbeit at this inſtant of his
  apprehenſion, they wil al forſake him.}

%%% o-2364
Theſe things haue I ſpoken to you, that you be not ſcandalized. \V Out
of the Synagogues they wil
\SNote{The Heretikes tranſlate, \Emph{Excommunicate you}. See what
corruptiõ this is, and the reaſon thereof,
\XRef{Annot. c.~9.~22.}}
caſt you: but the houre commeth, that euery one which killeth you, ſhal
thinke that he doeth ſeruice to God. \V And theſe things they wil doe to
you, becauſe they haue not knowen the Father, nor me. \V But theſe
things I haue ſpoken to you: that when the houre ſhal come, you may
remember them, that I told you.

\V But I told you not theſe things from the beginning, becauſe I was with
you. And now I goe to him that ſent me, and none of you asketh me:
Whither goeſt thou? \V But becauſe I haue ſpoken theſe things to you,
ſorrow hath filled your hart. \V But I tel you the truth, it is expedient for
you that I goe. For if I goe not, the Paraclete ſhal not come to you:
but if I goe, I wil ſend him to you. \V And when he is come, he ſhal
argue the world of ſinne, and of iuſtice, and of iudgement. \V Of ſinne:
becauſe they beleeued not in me. \V But of iuſtice: becauſe I goe to the
Father; and now you ſhal not ſee me. \V And of iudgement: becauſe the
Prince of this world is now iudged. \V
\LNote{Yet many things}{This
\MNote{Chriſt left many things to be taught by the Church.}
place conuinceth that the Apoſtles and the faithful be taught many
things, which Chriſt omitted to teach them for their weaknes: and that it
was the prouidence of God that Chriſt in preſence ſhould not teach and
order al things, that we might be no leſſe aſſured of the things that
the Church teacheth by the Holy Ghoſt, then of the things that himſelf
deliuered.}
Yet many things I haue to ſay to you: but you can not beare them now. \V
But when he,
\LNote{The Spirit of truth}{Euer
\MNote{The Spirit of truth.}
note that the Holy Ghoſt in that he is promiſed to the Church, is called
the Spirit of truth. Which Holy Spirit for many other cauſes is giuen to
diuers priuate men and to al good men, to ſanctification: but to teach al truth
& preſerue in truth and from errour, he is promiſed and performed only
to the Church and the cheefe Gouerner and general Councels thereof.}
the Spirit of truth, commeth,
\SNote{If he shal teach al truth, & that for euer, (as before
\XRef{c.~14,~16:})
how is it poſsible, that the Church cã erre or hath
erred at any time or ĩ any point?}
he ſhal teach you al truth. For he ſhal not ſpeake of himſelf; but what
things ſoeuer he ſhal heare, he ſhal ſpeake: and the things that are to
come he ſhal ſhew you. \V He ſhal glorifie me; becauſe he ſhal receiue
of mine, and ſhal ſhew to you. \V Al things whatſoeuer the Father hath,
be mine. Therfore I ſaid, that he ſhal receiue of mine, and ſhal ſhew to
you. \V A litle while, and now you ſhal not ſee me; & againe a litle
while, & you ſhal ſee me: becauſe I goe to the Father.

\V Some therfore of his Diſciples ſaid one to another: What is this that
he ſaith to vs: A litle while, and you ſhal not ſee me; and againe a
%%% 2525
litle while and you ſhal ſee me, and, becauſe I goe to the Father? \V
They ſaid therfore: What is this that he ſaith, A litle while? we know
not what he ſpeaketh. \V And \Sc{Iesvs} knew, that they would aske him;
& he ſaid to them: Of this you doe queſtion among you ſelues, becauſe I
ſaid to you: A litle while, and you shal not ſee me; and againe a litle
while, and you shal ſee me. \V Amen,
%%% o-2365
amen I ſay to you, that you shal weepe and lament, but the world shal
reioyce: and you shal be made ſorrowful, but you ſorrow shal be turned
into ioy. \V A woman when she trauaileth, hath ſorrow, becauſe her houre
is come: but when she hath brought forth the child, now she remembreth
not the anguish for ioy, that a man is borne into the world. \V And you
therfore, now indeed you haue ſorrow, but I wil ſee you againe, and your
hart shal reioyce; and your ioy no man shal take from you. \V And in
that day me you shal not aske any thing. Amen, amen I ſay to you, if you
aske the Father any thing
\SNote{Vpon this the Church cõcludeth al her praiers, \L{Per Chriſtũ
    Dominum noſtrum}, euen thoſe alſo that be made to Saints.}
in my name, he wil giue it you. \V Vntil now you haue not asked any
thing in my name. Aske and you shal receiue; that your ioy may be
ful. \V Theſe things in prouerbs I haue ſpoken to you. The houre commeth
when in prouerbs I wil no more ſpeake to you, but plainely of the Father
I wil ſhew you. \V In that day you shal aske in my name: and I ſay not
to you, that I wil aske the Father for you. \V For the Father himſelf
loueth you, becauſe you haue loued me, and haue beleeued that I came
forth from God. \V I came forth from the Father, and came into the
world: againe I leaue the world, and I goe to the Father.

\V His Diſciples ſay to him: Behold now thou ſpeakeſt plainely, and
ſaieſt no prouerb. \V Now we know that thou knoweſt al things, and thou
needeſt not that any man aske thee. In this we beleeue that thou cameſt
forth from God. \V \Sc{Iesvs} anſwered them: Now doe you beleeue? \V
\CNote{\XRef{Mt.~27,~31.}
\XRef{Mr.~14,~27.}}
Behold the houre commeth, and it is now come, that you ſhal 
\Fix{de}{be}{obvious typo, fixed in other}
ſcattered euery man into his owne, and me you shal leaue alone: and I am
not alone becauſe the Father is with me. \V Theſe things I haue ſpoken
to you, that in me you may haue peace. In the world you shal haue
diſtreſſe, but haue confidence, I haue ouercome the world.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
