%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 2505
%%% o-2341
\startChapter[
  title={Chapter 8}
  ]

\Summary{Againe in the Temple (abſoluing an aduoutreſſe after his
merciful manner, & yet withal declaring againſt his enemies that he
is not a fauourer of ſinne, no more then Moyſes) 12.~he teacheth
openly, and is not for al that apprehended: telling them both of his
Godhead, 21.~and of their reprobation, 28.~of his
exaltation alſo by their Crucifying of him: 31.~exhorting the
beleeuers to perſeuer, 33.~and shewing them that ſeeke his death,
that they are neither free, 39.~nor of Abraham, 41.~nor 
of God, 44.~but of the Diuel. 45.~But that himſelf is of God,
52.~and greater and ancienter then Abraham. 59.~For 
the which they goe about to ſtone him, but in vaine.}

%%% o-2342
And \Sc{Iesvs} went into the Mount-oliuet: \V and early in the morning
againe he came into the Temple, and the people came to him, and ſitting
he taught them.

\V And the Scribes and Phariſees bring a woman taken in aduoutrie; and
they did ſet her in the middes, \V And ſaid to him: Maiſter, this woman
was euen now taken in aduoutrie. \V And
\CNote{\XRef{Leu.~20,~10.}}
in the Law Moyſes commanded vs
to ſtone ſuch. What ſaieſt thou therfore? \V And this they ſaid tempting
him, that they might accuſe him. But \Sc{Iesvs} bowing himſelf downe,
with his finger
%%% 2506
wrote in the earth. \V When they therfore continued asking him, he
lifted vp himſelf, and ſaid to them:
\SNote{We can not conueniently reprehend or cõdemne other mens faults,
  if our ſelues be guilty of the ſame or other greater.
\Cite{Cyril. in Io.}
\Emph{See}
\XRef{Annot. Mt.~7.~1.}}
He that is without ſinne of you, let him firſt throw the ſtone at
her. \V And againe bowing himſelf, he wrote in the earth. \V And they
hearing, went out one by one, beginning at the Seniours: and \Sc{Iesvs}
alone remained, and the woman ſtanding in the middes. \V And \Sc{Iesvs}
lifting vp himſelf, ſaid to her: Woman, where are they that accuſed
thee: hath no man condemned thee? \V Who ſaid: No man, Lord. And
\Sc{Iesvs} ſaid:
\SNote{S.~Auguſtine by this example of our Maiſter proueth that Clergie
  men ſpecially ſhould be giuẽ much to mercie and that they ought often,
as the cauſe and time require to get pardon of the ſecular Magiſtrates
for offenders that be penitent.
\Cite{Ep.~54.}}
Neither wil I condemne thee. Goe, and now ſinne no more.

\V Againe therfore \Sc{Iesvs} ſpake to them, ſaying: I am the light of
the world: he that followeth me, walketh not in darkeneſſe, but ſhal
haue the light of life. \V The Phariſees therfore ſaid to him: Thou
giueſt teſtimonie of thy ſelf; thy teſtimonie is not true. \V \Sc{Iesvs}
anſwered, and ſaid to them: Although I doe giue teſtimonie of my ſelf,
my teſtimonie is true: becauſe I know whence I came, & whither I goe:
but you know not whence I come, or whither I goe. \V You iudge according
to the fleſh: I doe not iudge any man. \V And if I doe iudge, my
iudgement is true: becauſe I am not alone, but I and he that ſent me,
the Father. \V And in your law it is written, that
\CNote{\XRef{Deu.~17,~6.}
\XRef{19,~15.}}
the teſtimonie of two
men is true.  \V I am he that giue teſtimonie of my ſelf: and he that
ſent me, the Father, giueth teſtimonie of me. \V They ſaid therfore to
him: Where is thy Father? \Sc{Iesvs} anſwered: Neither me doe you know,
nor my Father. If you did know me, perhaps you might know my Father
alſo. \V Theſe words \Sc{Iesvs} ſpake in the Treaſurie, teaching in the
Temple, and no man apprehended him, becauſe his houre was not yet come.

%%% o-2343
\V Againe therfore \Sc{Iesvs} ſaid to them: I goe, and you ſhal ſeeke
me, and ſhal die in your ſinne. Whither I goe, you can not come. \V The
Iewes therfore ſaid: Why, wil he kil himſelf, becauſe he ſaith: Whither
I goe, you can not come? \V And he ſaid to them: You are from beneath, I
am from aboue. You are of this world, I am not of this world. \V
Therfore I ſaid to you that you ſhal die in your ſinnes. For if you
beleeue not that I am he, you ſhal die in your ſinne. \V They ſaid
therfore to him: Who art thou? \Sc{Iesvs} ſaid to them:
\SNote{So read S.~Cyril, & S.~Ambroſe, & S.~Auguſtine: expoũding it of
  Chriſtes Perſon, that he is the beginning or cauſe of al creatures.}
The Beginning who alſo ſpeake to you \V Many things I haue to ſpeake and
iudge of you. But he that ſent me, is true: and what I haue heard of
him, theſe things I ſpeake in the world. \V And they knew not that he
ſaid to them that his Father was God. \V \Sc{Iesvs} therfore ſaid to
them: When you ſhal haue exalted the Sonne of man, then you ſhal know
that I am he, and of my ſelf I doe nothing, but as the Father hath
taught me, theſe things I ſpeake: \V and he that ſent me, is with me:
and he hath not left me alone, becauſe the things that pleaſe him I doe
alwaies \V When he ſpake theſe things many beleeued in him.

\V \Sc{Iesvs} therfore ſaid to them that beleeued him, the Iewes: If you
\SNote{Only faith is not ſufficient without perſeuerance or abiding in
  the keeping of his cõmandements.}
abide in my word, you ſhal be my Diſciples indeed. \V And you ſhal know
the truth, and the truth ſhal make you free. \V They anſwered him: We
are the ſeed of Abraham, and we neuer ſerued any man: how ſaieſt thou,
You ſhal be free? \V \Sc{Iesvs} anſwered them:
%%% 2507
\LNote{Amen, amen}{\Emph{What is it}
\MNote{Why \Emph{Amen, amen}, is not tranſlated.}
(ſaith S.~Auguſt. vpon this place) \Emph{when our Lord ſaid,
  \Emph{Amen, amen}? He doth much commend and vrge the thing that he ſo
  vttereth, doubling it. It is a certaine othe of his, if a man may ſo
  ſay: for \Emph{Amen} in Hebrew ſignifieth \L{verum}, \Emph{a
    truth}. Yet it is not tranſlated, whereas it might haue been ſaid,
  \L{verum verum dice vobis}, but neither the
\TNote{Αμήν, ἀμήν.}
Greeke interpreter durſt
  doe it, nor the Latin, the Hebrew word hath remained ſtil, that ſo it
  might be the more eſteemed.}
\Cite{Tract.~41. in Ioan.}
By which wordes & the like
\SNote{See
\XRef{the preface,}
&
\XRef{annot. in Apoc. c.~19,~4.}}
recorded in other places of this new Teſtament, the Reader may ſee great
reaſon, why we alſo ſay, \Emph{Amen, amen}, and durſt not tranſlate it
and ſuch like wordes into our Engliſh tongue.}
Amen, amen I ſay to you, that
\CNote{\XRef{Ro.~6,~16.}
\XRef{2.~Pet,~2,~19.}}
euery one which committeth ſinne, is the
ſeruant of ſinne. \V And the ſeruant abideth not in the houſe for euer:
the ſonne abideth for euer. \V If therfore the Sonne make you free, you
ſhal be
\SNote{Man was neuer without free wil: but hauing the grace of Chriſt,
  his wil is truly made free (as S.~Auguſtine ſaith) from ſeruitude of
  ſinne alſo.
\Cite{tract~41. in Euang.~Io.}}
free indeed. \V I know that you are the children of Abraham: but you
ſeeke to kil me, becauſe my word taketh not in you. \V I ſpeake that
which I haue ſeen with my Father: and you doe the things that you haue
ſeen with your Father. \V They anſwered, and ſaid to him: Our Father is
Abraham. \Sc{Iesvs} ſaith to them: If you be the children of Abraham, 
\SNote{Not only faith but good workes alſo make men the children of
  Abraham according as S.~Iames alſo ſpeaketh of Abrahams
  workes.
\XRef{c.~2.}}
doe the workes of Abraham. \V But now, you ſeeke to kil me, a man that
haue ſpoken the truth to you, which I haue heard of God. This did not
Abraham. \V You doe the workes of your Father. They ſaid therfore to
him: We were not borne of fornication. We haue one Father, God. \V
\Sc{Iesvs} therfore ſaid to them: If 
%%% o-2344
God were your Father, verily you
would loue me. For from God I proceeded, and came: for I came not of my
ſelf, but he ſent me: \V Why doe you not know my ſpeach? Becauſe you cã
not heare my word. \V You are of your Father the Diuel, & the deſires of
your father you wil doe. He was a
\SNote{S.~Auguſtin compareth Heretikes in their ſpiritual murder by
  driuing Chriſtian men out of the Church to the Diuel that droue our
  parents out of Paradiſe
\Cite{Cont. lit. Petil. li.~2. c.~13.}}
man-killer from the beginning, & he ſtood not in the veritie; becauſe
veritie is not in him. When he ſpeaketh a lie, he ſpeaketh of his owne,
becauſe he is a lyer, & the father thereof. \V But becauſe I ſay the
veritie, you beleeue me not. \V Which of you ſhal argue me of ſinne? If
I ſay the veritie, why doe you not beleeue me? \V He that is of God,
heareth the wordes of God. Therfore you heare not, becauſe you are not
of God. \V The Iewes therfore anſwered, and ſaid to him: Doe not we ſay
wel that thou art a Samaritane, and haſt a Diuel? \V \Sc{Iesvs}
anſwered:
\SNote{He denieth not that he is a Samaritane, becauſe he is our keeper
  or protectour, as the word ſignifieth; & becauſe he is indeed that
  merciful Samaritane in the parable of the wounded man
\XRef{Luc~10,~33}
\Cite{Aug. tract.~43. in Ioan.}}
I haue no Diuel: but I doe honour my Father, and you haue diſhonoured
me. \V but I ſeeke not mine owne glorie. There is that ſeeketh and
iudgeth.

\V Amen, amen I ſay to you: If any man keep my word, he ſhal not ſee
death for euer. \V The Iewes therfore ſaid: Now we haue knowen that thou
haſt a Diuel. Abraham is dead, and the Prophets; and thou ſaieſt: If any
man keep my word, he ſhal not taſt death for euer. \V Why, art thou
greater then our father Abraham, who is dead? and the Prophets are
dead. Whom doeſt thou make thy ſelf: \V \Sc{Iesvs} anſwered: If I doe
glorifie my ſelf, my glorie is nothing. It is my Father that glorifieth
me, whom you ſay that he is your God. \V And you haue not knowen him,
but I know him. And if I ſhal ſay that I know him not, I ſhal be like to
you, a lyer. But I doe know him, and doe keep his word. \V Abraham your
father reioyced that he might ſee my day: and he ſaw, and was glad. \V
The Iewes therfore ſaid to him: Thou haſt not yet fiftie yeares and haſt
thou ſeen Abraham? \V \Sc{Iesvs} ſaid to them: Amen, amen I ſay to you,
before that Abraham was made I am. \V They tooke ſtones therfore to caſt
at him. But
\Fix{Ieſus}{\Sc{Iesvs}}{obvious typo, fixed in other}
hid himſelf, and went out of the temple.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
