%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% John
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 2520
%%% o-2359
\startChapter[
  title={Chapter 14}
  ]

\Summary{They being ſad, becauſe he ſaid that he muſt goe from them, he
  comforteth them many waies as putting them in hope to follow him vnto
  the ſame place, ſo that they keepe his commandements. Where he telleth
them, that himſelf is the way thither according to his Humanitie, and alſo
the end according to his Diuinitie, no leſſe then his Father, becauſe he
is conſubſtantial, 15.~promiſing alſo to ſend vnto them (that is, to his
Church) the Holy Ghoſt to be after his departure with them for
euer. 18.~And ſaying that it is his promotion (according to his
Humanitie) to goe to the Father, for whoſe obedience this his death shal
be, & not for any guilt of his owne.}

Let not your hart be troubled. You beleeue in God, beleeue in me
alſo. \V In my Fathers houſe there be
\SNote{Theſe manſiõs ſignifie differẽces of glorie in
Heauen.
\Cite{Hiero. li.~2. adu. Iouin.}}
many manſions. If not, I would haue told you, Becauſe I goe to prepare
you a place. \V And if I goe, and prepare you a place: I come againe and
wil take you to my ſelf, that where I am,
you alſo may be. \V And whither I goe you know, and the way you know.

\V Thomas ſaith to him: Lord, we know not whither thou goeſt; and how
can we know the way? \V \Sc{Iesvs} ſaith to him: I am the way, and the
veritie, and the life. No man commeth to the Father, but by me. \V If
you had knowen me, my Father alſo certes you had knowen: and from hence
forth you
\Var{ſhal know}{doe know}
him, and you haue ſeen him.

\V Philippe ſaith to him: Lord ſhew vs the Father, and it ſufficeth
vs. \V \Sc{Iesvs} ſaith to him: So lõg time I am with you; & haue you
not knowẽ me? Philippe, he that ſeeth me, ſeeth the Father alſo. How
ſaieſt thou, Shew vs the Father? \V
\Var{Doeſt thou not beleeue}{doe ye not beleeue.}
that I am in
the Father, & the Father in me? The wordes that I ſpeake to you, of my
ſelf I ſpeake not. But my Father that abideth in me, he doeth the
workes. \V Beleeue you not, that I am in the Father and the Father in
me? Otherwiſe for the workes themſelues beleeue. \V Amen, amen I ſay to
you, he that beleeueth in me, the workes that I doe, he alſo ſhal doe,
and 
\LNote{Greater then theſe}{S.~Chryſoſtom
\MNote{We may and muſt eaſily beleeue the miracles of Saints and of
  their relikes, when Chriſt himſelf foretelleth they shal doe ſuch
  wonderful things.}
in a whole
\CNote{\Cite{Li. de Babyla Mart. to.~5.}}
booke againſt the Pagans proueth that this was fulfilled not
only in Peters ſhadow, and Paules garments, which as we read in the
Actes, healed infirmities: but alſo by the Relikes and monuments of
Saints, namely of S.~Babylas, of whom he there treateth: thereby
inferring that Chriſt is God, who could and did performe theſe wonderful
wordes, by the very aſhes of his ſeruants, The Proteſtants cleane
contrarie, as patrones of the Pagans infidelitie, as though our Sauiour
had promiſed theſe & the like miraculous workes in vaine, either not
meaning or not able to fulfil them, ſo doe they diſcredit al the
approued hiſtories of the Church concerning miracles wrought by Saints:
namely that \Emph{S.~Gregorie Thaumaturgus}
\CNote{See
\XRef{Annot. Mat.~17,~20.}}
remoued a mountaine, the miracles of \Emph{S.~Paul} the 
\Fix{Eremitie}{Eremite}{obvious typo, fixed in other}
and \Emph{S.~Hilarion} written by \Emph{S.~Hierom}, the miracles of
\Emph{S.~Martin} written by \Emph{Seuerus Sulpitius}, the miracles
teſtified by \Emph{S.~Auguſtin de Ciuit.\ Dei}, the miracles approued by
\Emph{S.~Gregorie} in his Dialogues, the miracles reported by
\Emph{S.~Bede} in his Eccleſiaſtical ſtorie and liues of Saints: & al
other miracles neuer ſo faithfuly recorded in Eccleſiaſtical writers. 
\MNote{The Heretikes as faithles in this point, as the old Pagãs.}
In al which
things aboue their reach of reaſon and nature, they are as litle
perſuaded and haue no more faith then had the Pagans, againſt whom
\Emph{S.~Chryſoſtom} in the foreſaid booke, &
\Cite{S.~Auguſtin de Ciu. Dei li.~22. c.~3.}
and other Fathers heretofore haue
written. No man therfore needeth to maruel that the very Image of our
Lady, & the like, doe miracles, euen as Peters ſhadow did: nor wonder,
if ſuch things ſeeme ſtranger and greater then thoſe which Chriſt
himſelf did: whereas our Sauiour to put vs out of doubt, ſaith expreſly,
that his Saints ſhal doe greater things then himſelf did.}
greater then theſe ſhal he doe, \V becauſe I goe to the Father, &
whatſoeuer you ſhal aske in my 
%%% o-2360
name, that wil I doe: that the Father may be glorified in the Sonne. \V
If you ask me any thing in my name, that wil I doe. \V If you
\SNote{It is thẽ poſſible both to loue Chriſt & to keepe his
  commandements.}
loue me, keepe my commandemẽts. \V And I wil aske the Father, and he wil
giue you another
\SNote{\Emph{Paraclete} by interpretation is either a comforter, or an
  aduocate: and therfore to tranſlate it by any one of thẽ only, is
  perhaps to abridge the ſenſe of this place.}
Paraclete, that he may abide with you
\LNote{For euer}{If
\MNote{The Holy Ghoſt is promiſed to the Church for euer.}
the Holy Ghoſt had been promiſed only to the Apoſtles, their Succeſſours
& the Church after them could not haue chalenged it but it was promiſed
them for euer. Whereby we may learne, both that the priuiledges and
promiſes made to the Apoſtles were not perſonal, but pertaining to their
offices perpetually: and alſo that the Church and Paſtours in al Ages
had and haue the ſame Holy Ghoſt to gouerne them, that the Apoſtles and
primitiue Church had.}
for euer, \V
\LNote{The Spirit of truth}{They
\MNote{The Spirit of truth ſhal aſsiſt the Church alwaies.}
had many particular guifts and graces of the Holy Ghoſt before, & many
vertues by the ſame, as al holy men haue at al times: but the Holy Ghoſt
here promiſed to the Apoſtles and their Succeſſours for euer, is to this
vſe ſpecially promiſed, to direct them in al truth and veritie: & is
contrarie to the ſpirit of errour, hereſie, & falſhood. And therfore the
Church can not fall to Apoſtaſie or Hereſie, or to nothing, as the
Aduerſaries ſay.}
the Spirit of truth, whom the world can not receiue, becauſe it
%%% 2521
ſeeth him not, neither knoweth him, but you
\Var{know}{shal know}
him: becauſe he ſhal abide with you, and ſhal be in you. \V I wil not
leaue you orphanes: I wil come to you. \V Yet a litle while: and the
world ſeeth me no more. But you ſee me: becauſe I liue, and you ſhal
liue. \V In that day you ſhal know that I am in my Father, and you in
me; and I in you. \V He that hath my cõmandements, & keepeth them: he
it is that loueth me. And he that loueth me, ſhal be loued of my Father:
and I wil loue him, and wil manifeſt my ſelf to him.

\V Iudas ſaith to him, not that Iſcariote: Lord, what is done, that thou
wilt manifeſt thy ſelf to vs, and not to the world? \V \Sc{Iesvs}
anſwered, and ſaid to him: If any loue me, he wil keepe my word, and my
Father wil loue him, and we wil come to him, and wil make abode with
him. \V He that loueth me not, keepeth not my wordes. And the word which
you haue heard, is not mine; but his that ſent me, the Fathers. \V Theſe
things haue I ſpoken to you abiding with you. \V But the Paraclete the
Holy Ghoſt, whom the Father wil ſend in my name, he ſhal
\SNote{See the
\XRef{Annot. vpon the 16.~Chap. vers. 12.~&~13.}}
teach you al things, & ſuggeſt vnto you al things whatſoeuer I ſhal ſay
to you. \V Peace I leaue to you, my peace I giue to you; not as the
world giueth, doe I giue to you. Let not your hart be troubled, nor
feare. \V You haue heard that I ſaid to you, I goe, and I come to you. If
you loued me, you would be glad verily, that I goe to the Father:
becauſe the
\LNote{Father greater then I}{There 
\MNote{The Arians alleage as plaine Scriptures as the Proteſtãts.}
is no place of Scripture that
  ſeemeth any thing ſo much to make for the Sacramentaries, as this and
  other in outward ſhew of wordes ſeemed to make for the Arians, who
  denied the equalitie of the Sonne with the Father. Which wordes yet
  indeed rightly vnderſtood after the Churches ſenſe, make nothing for
  their falſe Sect, but only ſignifie that Chriſt according to his Manhood was
  inferiour indeed, and that according to his Diuinitie he came of the
  Father. And if the Hereſie or diſeaſe of this time were Arianiſme, we
  ſhould ſtand vpon theſe places and the like againſt the Arians, as we
  now doe vpon others againſt the Proteſtants, whoſe Sect, is the diſeaſe
  and bane of this time.}
Father is greater then I. \V And now I haue told you before it come to
paſſe: that when it ſhal come to paſſe, you may beleeue. \V Now I wil
not ſpeake many things with you. For the Prince of this world commeth,
and in me he hath not any thing. \V But that the world may know that I
loue the Father: and as the Father hath giuen me commandement, ſo doe I:
Ariſe, let vs goe hence.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
