%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 1433
%%% o-1321
\startChapter[
  title={Chapter 20}
  ]

\Summary{}


VVine
\MNote{Sobrietie.}
is a luxurious thing, & drunkenes tumultuous: whoſoeuer is delighted
therwith shal not be wiſe.

\V
\MNote{Obedience to Princes.}
As the roaring of a lyon, ſo alſo the terrour of a king: he that
prouoketh him, ſinneth alſo againſt his owne ſoule.

\V
\MNote{Concord.}
It is honour to a man, that ſeparateth himſelf from contentions: but
al fooles medle with contumelies.

\V
\MNote{Induſtrie.}
Becauſe of cold the ſlothful would not plowe: he shal begge therfore in
the ſummer, and it shal not be geuen him.

\V
\MNote{Profound counſel.}
As deepe water, ſo counſel in the hart of a man: but a wiſe man shal
draw it out.

\V
\MNote{Fidelitie.}
Manie men are called merciful: but a faithful man who shal fynd?

\V
\MNote{Sinceritie.}
The iuſt that walketh in his ſimplicitie, shal leaue bleſſed children.

%%% o-1322
\V
\MNote{Execution of Iuſtice.}
The king, that ſitteth in the throne of iudgement, diſſipateth al euil
with his looke.

\V
\MNote{Neceſſitie of grace.}
Who can ſay: My hart is cleane, I am pure from ſinne?

\V
\MNote{Equitie.}
Weight and weight, meaſure and meaſure: both are abominable before God.

\V
\MNote{Good toward liues.}
By his conuerſation a child is perceiued, if his workes be cleane and
right.

\V
\MNote{Right vſe of ſenſes.}
The eare hearing, and the eie ſeing, our Lord made both.

\V
\MNote{Diligent trauel.}
Loue not ſleepe, leſt pouertie oppreſſe thee: open thyne eies and be
filled with breades.

\V
\MNote{Prudence.}
It is naught, it is naught, ſayth euerie byer: and when he is departed
he wil boaſt.

\V
\MNote{Knowlege is a iewel.}
There is gold, and multitude of pearles: but a precious veſſel the
lippes of knowlege.

\V
\MNote{Securitie in ſuertiſhipe.}
Take his garment, that was the ſuretie of a ſtranger, and for ſtrangers
take a pledge from him.

\V
\MNote{Truth.}
The bread of lying is ſwete to a man: and afterward his mouth shal be
filled with the grauelſtone.

\V
\MNote{Counſel in warre.}
Cogitations are ſtrengthened by counſels: and battels are to be handled
by gouernmentes.

%%% 1434
\V
\MNote{Conſult with the faithful.}
Medle not with him that reuealeth myſteries, and walketh fraudulently,
and dilateth his lippes.

\V
\MNote{Honour of parents.}
He that curſeth his father, and mother, his lampe shal be extinguished
in the middes of darkenes.

\V
\MNote{Lawful gaine.}
The inheritance wherunto haſte is made in the beginning, in the later
end shal lacke bleſſing.

\V
\MNote{Meeknes.}
Say not: I wil requit euil: expect our Lord, and he wil deliuer thee.

\V
\MNote{Equitie.}
Weight and weight are abomination with our Lord: a deceitful balance is
not good.

\V
\MNote{Truſt in Gods prouidence.}
The ſteppes of man are directed of our Lord: but who of men can
vnderſtand his owne way?

\V
\MNote{Performance of vowes.}
It is ruine to a man to deuoure ſaintes, and afterward to retracte the
vowes.

\V
\MNote{Publique iuſtice.}
A wiſe king diſſipateth the impious, and bendeth ouer them a triumphant
arch.

\V
\MNote{Puritie of minde.}
The lampe of our Lord, the breath of a man, which ſearcheth al the
ſecretes of the bellie.

\V
\MNote{Three kinglie vertues.}
Mercie & truth kepe the king, and his throne is ſtrengthened by
clemencie.

\V
\MNote{Prudence with fortitude.}
The ioy of yongmen their ſtrength: and the dignitie of oldmen a gray
head.

\V
\MNote{Puniſhment of ſinnes.}
The blewneſſe of the wound shal wipe away euils: and ſtripes in the more
ſecrete place of the bellie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
