%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 1428
%%% o-1317
\startChapter[
  title={Chapter 17}
  ]

\Summary{}


Better
\MNote{Peace.}
is a drie morſel with ioy, then a houſe ful of victimes with brawling.

%%% 1429
\V
\MNote{VViſdom in general.}
A wiſe ſeruant shal rule ouer foolish children, and diuide inheritance
among bretheren.

\V
\MNote{Purification.}
As ſiluer is tried by fyre, and gold in the fournace: ſo our Lord
proueth
\Fix{the the}{the}{obvious typo, fixed in other}
harts.

\V
\MNote{%
\Fix{Feloſhipe}{Felowshipe}{obvious typo, fixed in other}
with the good.}
The euil man obeyeth an vniuſt tongue: and the deceitful obeyeth lying
lippes.

\V
\MNote{Commiſeration.}
He that deſpiſeth the poore, vpbraydeth his maker: and he that reioyceth
at an other mans ruine, shal not be vnpunished.

\V
\MNote{Succeſſion.}
The crowne of oldmen the childrens children: and the glorie of children
their fathers.

\V
\MNote{Grauitie in ſpeach.}
Eloquent wordes become not a foole, nor lying lippes a prince.

\V
\MNote{Deſire of eternal life.}
A moſt gratful pearle, the expectation of him that expecteth: whither
ſoeuer he turneth himſelf, he vnderſtandeth wiſely.

%%% o-1318
\V
\MNote{Remitting offences.}
He that concealeth offence, ſeeketh frendshipes: he that in other word
repeteth it, ſeperateth the confederate.

\V
\MNote{Diſcrete correction.}
Reprehenſion doth more profit with a wiſeman, then an hundred ſtripes
with a foole.

\V
\MNote{Concord.}
An euil man alwayes ſeeketh brawles: but a cruel angel shal be ſent
againſt him.

\V
\MNote{Gentlenes.}
It is better to meete a beare when her yong are taken away, then a foole
truſting to himſelfe in his owne follie.

\V
\MNote{Gratitude.}
He that rendereth euil thinges for good, euil shal not depart from his
houſe.

\V
\MNote{Conſeruation of peace.}
He that letteth water goe, is the head of brawles: & before he ſuffer
contumelie, he forſaketh iudgement.

\V
\MNote{Iuſt cenſure.}
He that iuſtifieth the impious, and he that condemneth the iuſt, both
are abominable before God.

\V
\MNote{Spiritual riches.}
What doth it profit a foole to haue riches, wheras he can not buy
wiſdom?

He
\MNote{Humilitie.}
that maketh his houſe high, ſeeketh ruine: and he that refuſeth to
learne, shal fal into euils.

\V
\MNote{Conſtancie.}
He loueth at al time that is a frend: and a brother is proued in
diſtreſſes.

\V
\MNote{Carefulnes in promiſing.}
A foolish man wil clappe the handes, when he is ſuretie for his freind.

\V
\MNote{Concord.}
He that meditateth diſcordes, loueth brawles: and he that exalteth the
doore, ſeeketh ruine.

\V
\MNote{Sinceritie.}
He that is of a peruerſe hart, shal not finde good: and he that turneth
his tongue, shal fal into euil.

\V
\MNote{Education of youth.}
A foole is borne to his owne ignominie: but neither shal the father
reioyce in a foole.

%%% 1430
\V
\MNote{Alacritie of hart.}
A ioyful minde maketh a florishing age: a ſorowful ſpirit dryeth vp the
bones.

\V
\MNote{Iuſt iudgement.}
The impious receiueth giftes out of the boſome, that he may peruert the
pathes of iudgement.

\V
\MNote{Contempt of the world.}
In the face of the prudent wiſdom shineth: the eies of fooles are in the
endes of the earth.

\V
\MNote{Filial obedience.}
A foolish ſonne is the anger of the father: and the ſorow of the mother
that bare him.

\V
\MNote{Obedience to Magiſtrates.}
It is not good, to doe hurt to the iuſt: nor to ſtrike the prince, which
iudgeth right.

\V
\MNote{Diſcretion in ſpeaking.}
He that moderateth his wordes, is lerned and prudent: and the lerned
man is of a precious ſpirit.

\V
\MNote{Moderate ſilence.}
The foole alſo if he hold his peace, shal be reputed wiſe: and if he
cloſe his lippes, a man of vnderſtanding.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
