%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 1415
%%% o-1304
\startChapter[
  title={Chapter 09}
  ]

\Summary{Wiſdom hauing built her houſe with ſeuen pillers, inuiteth al
  to a prepared banquette, 11.~promiſing to multiplie ioyful
  dayes. 13.~Folie inciteth to her contrarie banquet of ſtollen water
  and hidde bread.}


VViſdome
\LNote{VViſdom hath built herſelf a houſe.}{According
\MNote{God built his Church with ſpiritual Paſtors & Rites of Religion
in the old Teſtament.}
to the literal ſenſe, wherin the myſtical is grounded (both intended by
the Holie Ghoſt) VViſdom which is God himſelf, Creator & Conſeruer of
al thinges,
\CNote{\XRef{Prou.~8. v.~31.}}
whoſe ſpecial good pleaſure and delight is to be with men,
built his houſe, the Church, firſt in the Patriarches, Prieſtes,
Prophetes and his other faithful ſeruantes in the old Teſtament; partly
before; but more conſpicouſly in the people of Iſrael, eſtabliſhing the
ſame with ſeuen, that is (according to the frequent phraſe of holie
Scripture) with manie pillers, Paſtors and chief gouerners, by whom the
whole people were directed in al ſpiritual cauſes; as the Pſalmiſt
likevviſe induceth God, ſaying:
\CNote{\XRef{Pſal.~74.}}
I haue confirmed the pillers therof. And as
\CNote{\XRef{Galat.~2.}}
S.~Paul aftervvard calleth S.~Peter, S.~Iames, and S.~Iohn, pillers.
In this houſe vviſdom alſo prepared
a banquette, appointed victimes of diuers ſortes, as gratful Sacrifices
to God, & therto inuited al men in much better order, and to their more
profite, then the adultreſſe vvoman, follie and vvicked concupiſcence,
inciteth to her carnal and vvorldlie pleaſures, vvhich bring to eternal
ruine. And for this purpoſe, God neuer ceaſed to ſend Prieſtes and
Prophetes to inuite the people of Iſrael, to this ſtrong tovvre fenſed
vvith vvalles.
\MNote{The ſame are more excellent in the Church of Chriſt.}
In the Allegorical ſenſe, the ſame diuine increated VViſdom, the ſecond
Perſon in the B.~Trinitie, the Diuine \Sc{VVord} coeternal to the
Father, built himſelf a houſe, his humane bodie in the virgins vvombe,
and therunto, as to the head, adioyned the members, his myſtical bodie
the Church, immolated victimes of Martyrs, prepared the Table in breade
and vvine, vvhere alſo appeareth his Prieſthood, according to the Order
of Melchiſedec, and called therto ſuch as before vvere vnvviſe, and of
ſmal vnderſtanding: becauſe as the Apoſtle ſaith:
\CNote{\XRef{1.~Cor.~1.}}
God choſe the vveake of this vvorld to confound the ſtrong. As
S.~Auguſtin expoundeth this paſſage
\Cite{li.~17. c.~10. de ciuit.}
&
\Cite{q.~51. veter. & noui. Teſt. to.~4.}}
hath built herſelf an houſe, she hath cut out ſeuen pillers. \V She
hath immolated her victimes, mingled her wine, and ſet forth her
table. \V She hath ſent her handmaides to cal to the towre, and to the
walles of the citie: \V If any be a litle one, let him come to me. And
to the vnwiſe she ſpake: \V Come,
\SNote{S.~Cyprian
\Cite{li.~2. ep.~3.}
citeth this whole paſſage of Chriſts Sacrifice in the formes of bread and
vvine.}
eate ye my bread, & drinke the wine which I haue mingled for you. \V
Leaue infancie, and liue, and walke by the wayes of prudence. \V He that
teacheth a ſcorner doth iniurie to himſelf; and he that rebuketh the
impious, purchaſeth a blotte to himſelf. \V Rebuke
\SNote{VVhere is no hope of amendment prudence directeth vs not to
admoniſh, nor rebuke ſinners, leſt without anie fruite we procure
emnitie. Charitie alſo requireth rather to expect better opportunitie,
leſt the offender become worſe by our admonition.}
not the ſcorner leſt
\Fix{hee}{he}{obvious typo, fixed in other}
hate thee.
\SNote{But when there is hope of good, euerie one is bond, eſpecially
ſuperiors to correct offenders.
\Cite{S.~Aug. li.~1. c.~9. de ciuit.}
\Cite{S.~Baſil. regulis fuſe diſput.~158.}}
Rebuke a wiſe man, and he wil loue thee. \V Geue
%%% o-1305
occaſion to a wiſe man, and wiſdom shal be added to him. Teach the iuſt,
and he shal make haſte to take it. \V The beginning of wiſdom, the feare
of our Lord: and the knowlege of the holie, prudence. \V For my me shal
thy dayes be multiplied, and yeares of life shal be added to thee. \V If
thou be wiſe, to thyſelf thou shalt be: and if a ſcorner, thou alone
shalt beare the euil. \V A foolish woman and clamorous, and ful of
alurementes, and knowing nothing at al, \V ſate in the doores of her
houſe vpon a ſeate in a high place of the citie, \V to cal them that
paſſe by the way, and goe on their iourney: \V He that is a litle one,
let him turne to me. And to the foole she ſpake: \V Stolen waters are
ſweeter, and hidden bread more pleaſant. \V And he was ignorant that
giantes are there, and her gueſtes in the depthes of hel.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
