%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 1420
%%% o-1309
\startChapter[
  title={Chapter 12}
  ]

\Summary{}


He
\MNote{Loue of diſcipline.}
that loueth diſcipline, loueth knowlege: but he that hateth
reprehenſions is vnwiſe.

\V
\MNote{Progreſſe in vertue.}
He that is good, shal draw grace from our Lord: but he that truſteth in
his owne cogitation, doth impiouſly.

\V
\MNote{Pietie.}
Man shal not be ſtrengthened by impietie: and the roote of the iuſt shal
not be moued.

\V
\MNote{Diligent induſtrie.}
A diligent woman is a crowne to her husband: and putrefaction in his
bones, she that doth thinges worthie of confuſion.

\V
\MNote{Deſire of iuſtice.}
The cogitations of the iuſt are iudgements: & the counſels of the
impious are fraudulent.

\V
\MNote{Sinceritie in vvordes.}
The wordes of the impious lie in wayte for bloud: the mouth of the iuſt
shal deliuer them.

\V
\MNote{Iuſtice.}
Turne the impious, and they shal not be: but the houſe of the iuſt shal
be permanent.

\V
\MNote{Sound doctrine.}
A man shal be knowen by his doctrine: but he that is vaine and foolish,
shal lye open to contempt. 

\V
\MNote{Contentment of minde.}
Better is the poore and ſufficient to himſelf, then he that is glorious
and wanteth bread.

\V
\MNote{Compaſſion.}
The iuſt knoweth the liues of his beaſtes: but the bowels of the impious
are cruel.

\V
\MNote{Diligent trauel.}
He that tilleth his land, shal be filled with breads: but he that
purſeweth idlenes is a verie foole.

%%% o-1310
He
\MNote{Temperance.}
that is delighted in much quaffing of wine, leaueth contumelie in his
munitions.

\V
\MNote{Iuſt puniſhment of the wicked.}
The deſire of the impious is the
\TNote{defence}
muniment of the moſt wicked: but the roote of the iuſt shal proſper.

\V
\MNote{Guard of the tongue.}
For the ſinnes of the lippes ruine approcheth to the euil man: but the
iuſt eſcapeth out of diſtreſſe.

\V
\MNote{Godlie inſtructions and good dedes.}
Of the fruite of his owne mouth shal euerie man be replenished with
good thinges, and according to the workes of his handes it shal be
repayed him.

\V
\MNote{Yelding to good counſel.}
The way of a foole is right in his eies: but he that is wiſe heareth
counſels.

%%% 1421
\V
\MNote{Toleration of iniuries.}
A foole by & by sheweth his anger: but he that diſſembleth iniuries is
wiſe.

\V
\MNote{True teſtimonie.}
He that ſpeaketh that which he knoweth, is an vtterer of iuſtice: but he
that lyeth, is a fraudulent witneſſe.

\V
\MNote{Aduiſed promiſing.}
There is that promiſeth, and is pricked as it were with the ſword of
conſcience: but the tongue of the wiſe is health.

\V
\MNote{Conſiderate teſtimonie.}
The lippe of truth shal be ſtable for euer: but he that is an haſtie
witneſſe, frameth a tongue of lying.

\V
\MNote{Pacification.}
Guile is in the hart of them that thinke euil thinges: but ioy foloweth
them that geue counſels of peace.

\V
\MNote{Iuſtice in general.}
It shal not make the iuſt ſorie whatſoeuer shal fal to him: but the
impious shal be replenished with euil.

\V
\MNote{Fidelitie.}
Lying lippes are an abomination to our Lord: but they that doe
faithfully pleaſe him.

\V
\MNote{Prudent ſilence.}
A circumſpect man concealeth knowlege: and the hart of the vnwiſe
prouoketh folie.

\V
\MNote{Reſiſtance of tentations.}
The hand of the ſtrong shal rule: but that which is ſlothful, shal ſerue
vnder tributes.

\V
\MNote{Alacritie in good workes.}
Penſifneſſe in the hart of a man shal humble him, & with a good word he
shal be made glad.

\V
\MNote{Freindſhipe.}
He that neglecteth damage for a freind, is iuſt: but the way of the
impious shal deceiue them.

\V
\MNote{Honeſt induſtrie.}
The fraudulent man shal not finde gayne: and the ſubſtance of a
\TNote{a iuſt man}
man shal be the price of gold.

\V
\MNote{Good life.}
In the path of iuſtice, life: but the byway leadeth to death.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
