%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 27
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-27


\project douay-rheims


%%% 1443
%%% o-1331
\startChapter[
  title={Chapter 27}
  ]

\Summary{}


Boaſt
\MNote{Truſt not to future vncertainties.}
not for to morow, being ignorant what the day to come may bring forth.

%%% 1444
\V
\MNote{Praiſe not thyſelf.}
Let an other prayſe thee, and not thine owne mouth: a ſtranger and not
thine owne lippes.

\V
\MNote{Beware of a fooles wrath:}
A ſtone is heauie, and ſand weightie: but the anger of a foole is
heauier then both.

\V
\MNote{for it hath no mercie.}
Anger hath no mercie, nor furie breaking forth: and the violence of a
moued ſpirit who can ſuſteyne?

%%% o-1332
\V
\MNote{Loue freindlie ſeueritie:}
Better is manifeſt correption, then loue hidden.

\V
\MNote{not enemies flaterie.}
Better are the woundes of him that loueth, then the fraudulent kiſſes of
him that hateth.

\V
\MNote{Temperance in meate.}
A ſoule that is ful shal treade vpon the honiecombe: and a ſoule that is
hungrie shal take bitter alſo for ſweete.

\V
\MNote{Iuſt care of thy familie.}
As a bird fleeting from her neſt, ſo a man that forſaketh his place.

\V
\MNote{Eſteme freindlie counſel.}
The hart is delighted with oyntement and diuers odours: and with the good
counſels of a frend the ſoule is ſweetned.

\V
\MNote{Old freindes are better, & ſurer then a kinſman.}
The frend, and thy fathers frend do not leaue: and goe not into thy
brothers houſe in the day of thyne affliction.

Better is a neighbour neere
\Fix{thene}{then}{obvious typo, fixed in other}
a brother far of.

\V
\MNote{Be wiſe euen for thy fathers ſake.}
Studie wiſedom my ſonne, & make my hart ioyful, that thou maiſt make
anſwer to the vpbrayder.

\V
\MNote{Foreſee and preuent dangers.}
The ſubtel man ſeeing euil, hideth himſelf: litle ones paſſing through
haue ſuſteyned euil detriments.

\V
\MNote{Truſt not enimies.}
Take his garment, that hath bene ſuretie for a ſtranger: and for alienes
take from him a pledge.

\V
\MNote{Suſpect extraordinarie curteſie.}
He that bleſſeth his neighbour with a loud voice, riſing in the night,
he shal be like him that curſeth.

\V
\MNote{Tolerate incorrigible brawlers with patience; leſt they become
worſe, by thy ſtriuing to amend them.}
Dropping through in the day of cold, and a brawling woman are compared
together: \V He that reteyneth her, as he that should hold the winde, and
shal cal in the oyle of his right hand.

\V Iron is sharpened with iron, and a man sharpeneth the face of his
frend.

\V
\MNote{Conſerue that is good.}
He that kepeth the feegtree, shal eate the frute therof: and he that is
the keper of his maſter, shal be glorified.

\V
\MNote{Conſcience is beſt witnes.}
As in waters the countenance of them that looke therin shyneth, ſo the
hartes of men are manifeſt to the prudent.

\V
\MNote{Contentment with our ſtate.}
Hel and perdition are neuer filled: in like maner alſo the eies of men
are vnſatiable.

\V
\MNote{Humilitie deſireth not praiſe.}
As ſiluer is tried in the forge, and gold in the fornace: ſo a man is
proued by the mouth of him that praiſeth.

%%% 1445
The
\MNote{Right of hart.}
hart of the wicked ſeeketh after euils, but the righteous hart
ſeeketh after knowlege.

\V
\MNote{Striue not to teach a foole.}
If thou shalt bray a foole in a morter, as when a peſtle ſtriketh vpon
ptiſane, his follie shal not be taken from him.

\V
\MNote{Diſcharge thy dutie:}
Know diligently the countenance of thy cattel, and conſider thy
flockes: \V
\MNote{vvhiles thou art in this life.}
For thou shalt not haue power alwayes: but a crowne shal be geuen into
generation and generation.

\V
\MNote{Be content with thinges neceſſarie, and they ſhal be geuen to
thee.
\XRef{Mat.~6. v.~33.}}
The medowes are open, and the grene herbes haue appeared, and the graſſe
is gathered out of the mountaines.

\V Lambes for thy garment: and kiddes the price of the filde.

\V Let the milke of the goates ſuffice thee for thy meates, and for the
neceſſities of thy houſe: and for victual to thy handmaides.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
