%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 1430
%%% o-1319
\startChapter[
  title={Chapter 18}
  ]

\Summary{}


He
\MNote{Conſtant amitie.}
ſeeketh occaſions that wil depart from a frend: he shal euer be ſubiect
to reproch.

\V
\MNote{Harken to good counſel.}
A foole receiueth not the wordes of prudence: vnleſſe thou ſay thoſe
thinges which are in his hart.

\V
\MNote{Speedie repentance.}
The impious when he shal come into the depth of ſinnes, contemneth: but
ignominie and reproch folow him.

\V
\MNote{VViſe diſcourſe.}
Deepe water wordes from the mouth of a
\TNote{a vviſeman}
man: and a ſtreame ouerflowing the fountaine of wiſdom.

\V
\MNote{Equitie.}
To accept the perſon of the impious in iudgement is not good, that thou
decline from the truth of iudgement.

\V
\MNote{Studie of peace.}
The lippes of a foole mingle him with ſtrife: and his mouth prouoketh
brawles.

\V
\MNote{Guard of the tongue.}
The mouth of a foole is his deſtruction: and his lippes are the ruine of
his ſoule.

\V
\MNote{Auoid ſecrete bate makers.}
The wordes of the duble tongued, as it were ſimple: and the ſame come
euen to the inner part of the bellie.

Feare
\MNote{Fortitude.}
caſteth downe the ſlothful: and the ſoules of the effeminate shal be
hungrie.

\V
\MNote{Conſtant labour.}
He that is ſoft and diſſolute in his worke, is the brother of him that
deſtroyeth his owne workes.

\V
\MNote{Confidence in God.}
A moſt ſtrong towre, the name of our Lord: the iuſt runneth to it, and
shal be exalted.

\V
\MNote{Spiritual riches.}
The ſubſtance of the rich man a citie of his ſtrength: and as a ſtrong
wal compaſſing him about.

%%% 1431
\V
\MNote{Humilitie.}
Before he be broken, the hart of a man is exalted: and before he be
glorified, it is humbled.

\V
\MNote{Conſideration in anſwering.}
He that anſwereth before he heare, sheweth him ſelf to be a foole, and
worthie of confuſion.

\V
\MNote{Quietnes of ſpirite.}
The ſpirit of a man vpholdeth his imbecilitie: but a ſpirit that is
eaſie to be angrie who can ſuſteyne?

\V
\MNote{Deſire to lerne.}
A wiſe hart shal poſſeſſe knowlege: and the eare of the wiſe ſeeketh
doctrine.

\V
\MNote{Almes dedes.}
The gift of a man enlargeth his way, & maketh him roome before princes.

\V
\MNote{Acknowleging of faultes.}
The iuſt is firſt accuſer of himſelf: his frend commeth, and shal ſearch
him.

\V
\MNote{Vſe of lote to make agreement.}
Lotte ſuppreſſeth contradictions, and betwen the mightie alſo it
determineth.

\V
\MNote{Fraternal concord.}
Brother that is holpen of brother, is as a ſtrong citie: and iudgements
are as the barres of cities.

\V
\MNote{Godlie inſtructions.}
Of the fruite of mans mouth his bellie shal be filled: and the
%%% o-1320
offsprings of his lippes shal fil him.

\V
\MNote{Right vſe of the tongue.}
Death and life in the hand of the tongue: they that loue it, shal eate
the fruites therof.

\V
\MNote{A wiues vertue is her beſt dowrie.}
He that hath found a good wife, hath found a good thing, and hath
receiued a pleaſure of our Lord.

\V
\MNote{Meeknes.}
The poore ſpeaketh with ſupplications, and the rich wil ſpeake roughly.

\V
\MNote{Amitie with affabilitie.}
A man amiable to ſocietie, shal be more frendlie, then a brother.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
