%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 1413
%%% o-1303
\startChapter[
  title={Chapter 08}
  ]

\Summary{Wiſdom is preached in conſpicious and moſt frequented places,
  that none may pretend wante of admonition, 7.~her doctrin is true,
  godlie, profitable, & neceſſarie to al ſortes of men. 12.~Wiſdom
  (increated which is God himſelfe) is eternal, 32.~and bringeth eternal
  happines.}


Doth not wiſdom crie, and prudence geue her voice? \V Standing
\SNote{True vviſdom directing to good life & ſo to eternal ſaluation, is
only found in the viſible Church ſtanding vpon a mountaine, not hidde in
corners or obſcure places.}
in the high & loftie toppes ouer the way, in the middes of the
pathes, \V beſide the gates of the citie in
%%% 1414
the verie doores she ſpeaketh, ſaying: \V O men, to you I crie, and my
voice is to the children of men. \V O litle ones vnderſtand ſubtiltie,
and ye vnwiſe marke. \V Heare ye, becauſe I wil ſpeake of great thinges:
and my lippes shal be opened to preach right thinges. \V My throte shal
meditate truth, and my lippes shal deteſt the impious. \V Al my wordes
are iuſt, there is no wicked, nor peruerſe thing in them. \V They are
right to them that vnderſtand, and iuſt to them that finde knowlege. \V
Receiue ye diſcipline, & not money: chooſe doctrine rather then gold. \V
For wiſdom is better then al moſt precious riches: and whatſoeuer is to
be deſired can not be compared to it. \V I
\SNote{Theſe ſingular praiſes perteine to the increated wiſdom, God
himſelf, from whom procedeth wiſdom geuen to men by the Holie Ghoſt. See
\XRef{Annot. ch.~1. v.~2.}}
wiſdome dwel in counſel, and am preſent in lerned cogitations. \V The
feare of our Lord hateth euil: arrogancie, and pride, and wicked way,
and a duble toungued mouth I doe deteſt. \V Myne is counſel & equitie,
prudence is mine, ſtrength is mine. \V By me kinges doe reigne, and the
makers of law decree iuſt thinges. \V By me princes rule, and the
mightie decree iuſtice. \V I loue them that loue me: and they that watch
toward me shal finde me. \V With me are riches, and glorie, glorious
riches, and iuſtice. \V For my fruite is better then gold, and precious
ſtone, and my bloſomes then
%%% o-1304
choſen ſiluer. \V I walke in the wayes of iuſtice, in the middes of the
pathes of iudgement, \V that I may enriche them that loue me, and may
replenish their treaſures. \V Our Lord poſſeſſed me in the beginning of
his wayes, before he had made anything from the beginning. \V From
eternitie I was ordayned & of old before the earth was made. \V The
depthes were not as yet, and I was now conceiued, neither had the
fountaines of waters as yet gushed forth: \V the mountaines with heauie
hugeneſſe ſtoode not as yet: before the litle hilles I was brought
forth: \V he had not yet made the earth, and the riuers, and the poles
of the round world. \V When he prepared the heauens, I was preſent: when
with a certaine law and circuite he compaſſed the depthes: \V when he
eſtablished the firmament aboue, & poyſed the fountaines of waters: \V
when he compaſſed the ſea with her limites, and ſet a law to the waters
that they should not paſſe their boundes: when he hanged the foundations
of the earth. \V I was with him, framing al thinges: and was delighted
euerie day, playing before him at al time, \V playing in the world: &
\SNote{God much preferreth man before al other corporal creatures.}
my delights to be with the children of men. \V Now therfore children
heare me.
%%% 1415
Bleſſed are they, that keepe my wayes. \V Heare ye diſcipline, and be
wiſe, and reiect it not. \V Bleſſed is the man that heareth me, & that
watcheth at my doores dayly, & wayteth at the poſtes of my doore. \V He
that shal finde me, shal finde life, and shal draw ſaluation of our
Lord. \V But he that shal ſinne againſt me, shal hurt his owne ſoule. Al
that hate me, loue death.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
