%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-23


\project douay-rheims


%%% 1437
%%% o-1325
\startChapter[
  title={Chapter 23}
  ]

\Summary{}


VVhen
\MNote{Modeſtie and diſcretion at the table of great perſons.}
thou shalt ſitte to eate with a prince, attend diligently what thinges
are ſet before thy face: \V and ſet a knife in thy throte, if
notwithſtanding thou haue thy ſoule in thine owne power. \V Deſire not
his meates, in which is the bread of lying.

\V
\MNote{Mediocritie of riches, and contentment with meane ſtate.}
Labour not to be rich: but ſet a meane to thy prudence.

\V Lift not vp thine eies to the riches, which thou canſt not haue:
%%% o-1326
becauſe they shal make to themſelues winges as of an eagle, and shal
flie into heauen.

\V
\MNote{Eate not with a niggard, leſt he be ſecretly greued, outwardly
diſſembling, and thyſelf repent it.}
Eate not with an enuious man, and deſire not his meates.

\V Becauſe after the ſimilitude of a ſouthſayer, and diuiner, he
eſteemeth that which he knoweth not. Eate and drinke, wil he ſay to
thee: and his minde is not with thee.

%%% 1438
\V The meates which thou hadſt eaten thou shalt vomite vp: and shalt loſe
thy beautiful wordes.

\V
\MNote{Aduiſe not the vncapable.}
Speake not in the eares of the vnwiſe: becauſe they wil deſpiſe the
doctrine of thy ſpeach.

\V
\MNote{Hurt not orphanes: for they are deare to God.}
Touch not the boundes of litle ones: and into the filde of pupils enter
not.

\V For their nerekinsman is ſtrong: and he wil iudge their cauſe againſt
thee.

\V
\MNote{Diligent ſtudie of wiſdom.}
Let thy hart enter into doctrin: and thyne eares to wordes of knowlege.

\V
\MNote{Moderate
\Fix{chatiſment}{chaſtiſment}{likely typo, same in both}
profiteth youth.}
Withdrawe not diſcipline from a childe: for if thou shalt ſtrike him
with the rod, he shal not die.

\V Thou shalt ſtrike him with the rod: and deliuer his ſoule from hel.

\V
\MNote{Vertues of children are ioyful to the parentes.}
My ſonne, if thy minde shal be wiſe, my hart shal be glad with thee: \V
And my reines shal reioyce, when thy lippes shal ſpeake right thinges.

\V
\MNote{Contentment with Gods wil. Hope of reward.}
Let not thy hart enuie ſinners: but in the feare of our Lord be thou al
the day: \V becauſe thou shalt haue hope in the later end, and thyn
exaltation shal not be taken away.

\V
\MNote{Lerne and practiſe vviſdom.}
Heare my ſonne, & be wiſe: & direct thy minde in the way.

\V
\MNote{Sobrietie and Temperance.}
Be not in the feaſtes of great drinkers, nor in their comeſſations,
which contribute flesh together to eate: \V becauſe they that are geuen
to drinking, and that pay shottes, shal be conſumed, and drouſines shal
be clothed with ragges.

\V
\MNote{Honour of parents.}
Heare thy father, that begot thee: and contemne not thy mother when she
is old.

\V
\MNote{VViſdom moſt precious.}
Bye truth, and ſel not wiſedom, and doctrine, and intelligence.

\V
\MNote{See
\XRef{v.~15 and~16.}}
The father of the iuſt reioyceth with gladnes: he that hath begotten a
wiſeman, shal reioyce in him.

\V Let thy father be glad, and thy mother, and let her reioyce that bare
thee.

\V
\MNote{Bodelie and ſpiritual puritie.}
My ſonne geue me thy hart: & let thyne eies kepe my wayes.

\V For an harlot is a deepe dich: & a ſtrange woman a narrow pitte.

\V She lyeth in wayt in the way as a robber, and whom she shal ſee not
circumſpect, she wil kil.

%%% o-1327
\V
\MNote{Sobrietie alvvayes neceſſarie.}
To whom is woe? to whoſe father woe? to whom brawles? to whom diches? to
whom woundes without cauſe? to whom bloud sheeding eies? \V
\MNote{Deteſt drunkennes.}
Is it not to them that paſſe their time in wine, and ſtudie to drinke
out their cuppes?

%%% 1439
\V
\MNote{Drunkennes is deceiptful,}
Behold not wine when it waxeth yelow, when the colour therof shal shine
in the glaſſe: it goeth in pleaſantly, \V
\MNote{dangerous,}
but in the end, it wil bite like a ſnake, and as a baſiliske it wil
powre abrode poyſones.

\V
\MNote{beaſtlie,}
Thine eies shal ſee ſtrange wemen, and thy hart shal ſpeake peruerſe
thinges.

\V
\MNote{hurtful to others,}
And thou shalt be as one ſleeping in the middes of the ſea, and as the
gouernour faſt aſleepe, the ſterne being loſt.

\V
\MNote{ſenſeleſſe and vnſatiable.}
And thou shalt ſay: They haue beaten me, but I was not greeued: they drew
me, and I felt not: When shal I awake, and finde wines againe?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
