%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 1412
%%% o-1302
\startChapter[
  title={Chapter 07}
  ]

\Summary{He further exhorteth youngmen to ſeke wiſdom, 5.~eſpecially to
  flee from the intiſements of harlots, largely deſcribing the ſame,
  22.~and the ruine of them that are ſo deluded.}


My ſonne,
\SNote{Becauſe tentations ſtil occurre in this life, and man is fraile,
good exhortations, and earneſt admonitions muſt alſo be continually
inculcated, as here the vviſman often repeteth and much vrgeth the ſame
good and neceſſarie aduiſes to embrace vviſdom and to vvalke ſtil in the vvay
of vertue, eſpecially to flee from vices, and dangers of ſinne.}
keepe my wordes, and my preceptes hide with thee. Sonne, \V keepe my
commandmentes, and thou shalt liue: and my law as the apple of thine
eie: \V binde it on thy fingers, write it in the tables of thy hart. \V
Say to wiſdom, thou art my ſiſter: & cal prudence thy freind, \V that
she may keepe thee from the ſtrange woman, and from the
\Fix{forenrer}{forener}{likely typo, same in both}
which maketh her wordes ſweete. \V For out of the window of my houſe I
looked out through the lattiſe, \V and
%%% 1413
I ſee litle ones, I behold a foolish youngman, \V which paſſeth through
the ſtreates by the corner, and goeth nigh the way of her houſe, \V in
the darke the day being toward euening, in the darkenes of the night,
and dimnes. \V And behold the woman meeteth him in harlotes atyre,
prepared to deceiue ſoules: babling and wandering, \V impatient of
reſt, nor able to conſiſt in the houſe on her feete, \V now abrode, now
in the ſtreates, now lying in wayte neere the corners. \V And taking the
youngman she kiſſeth him, and with malepert countenance ſpeaketh fayre,
ſaying: \V I vowed victimes for welfare, this day I haue payed my
vowes. \V Therfore I am come forth to meete thee, deſirous to ſee thee,
and I haue found thee. \V I haue wouen my bed with cordes, I haue
adorned it with tapeſtrie pictures out of Ægypt. \V I haue ſprinkled my
bed with myrrhe, aloes, and cinamone. \V Come let vs be inebriated with
breſtes, and let vs enioy deſired embracings, til the day appeare. \V
For my husband is not at home, he is gone a very long iourney. \V
He caried with him a bagge of money:
%%% o-1303
in the day of the ful moone he wil returne to his houſe. \V She
intangled him with many wordes, and with flatterie of lippes drewe
him. \V Immediatly he foloweth her
\SNote{Sinners after conſent geuen to tentations, are as inconſiderate
of their ovvne ſtate, & of their perile and ruine, as an oxe, vvhen he
is ledde to the ſhamles, or a bird allured vvith a bate, that flieth
into the ſnare or nette.}
as an oxe led to be a victime, & as a lambe playing the wanton, and not
knowing that he is drawen as a foole to bondes, \V til the arow pearſe
his liuer: as if a birde should make haſte to the ſnare, and knoweth not
that his life is in danger. \V Now therfore my ſonne, heare me, and
attend to the wordes of my mouth. \V Let not thy mind be drawen away in
her wayes: neither be thou deceiued with her pathes. \V For she hath
caſt downe manie wounded, and al the moſt ſtrong are ſlaine by her. \V
Her houſe the wayes of hel, penetrating to the inner partes of death.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
