%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 31
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-31


\project douay-rheims


%%% 1450
%%% o-1338
\startChapter[
  title={Chapter 31}
  ]

\Summary{An exhortation to chaſtitie, temperance, 8.~and to workes of
  mercie, 10.~with praiſe of a valiant wiſe woman.}


The
\MNote{VVordes perteining to Salomon, ſpoken to him by his mother: who
here calleth him Lamuel, ſignifying, \Emph{God vvith him}.}
wordes of Lamuel the king. The viſion wherwith his mother inſtructed
him. \V What ô my beloued, what ô the beloued of my wombe, what ô
beloued of my vowes? \V Geue not thy ſubſtance to wemen, & thy riches to
deſtroy kinges. \V Geue not to kinges, ô Lamuel, geue not wine to
kinges: becauſe there is no ſecrete where drunknes reigneth, \V & leſt
perhaps they drinke, & forget iudgements, & change the cauſe of the
children of the poore. \V Geue ſtrong drinke to them that be ſad, and
wine vnto them, that are of a penſiue minde: \V let them drinke, and
forget their pouertie, and not remember their ſorow any more. \V Open
thy mouth to
\SNote{Doctrine is moſt profitable to thoſe, that are modeſtly dumme,
more vvilling to heare then to ſpeake.}
the dumme, & to the cauſes of al the children that paſſe: \V open thy
mouth,
\SNote{The chief & moſt proper office of a king is to do iuſtice.}
decree that which is iuſt, & iudge the needie & poore. \V
%%% !!! LNote not marked in both
\LNote{A valiant vvoman vvho shal finde?}{Vpon
\MNote{The praiſe of a right vviſe woman vvritten in verſe, & in order
of the Alphabet.}
occaſion of his mothers moſt prudent admonition, the wiſeman ſingularly
praiſeth a perfect vertuous woman. And that in an exquiſite kind of
ſtile: in Tetramical Iambike verſe, with perfect order and number of the
Alphabet letters. Signifying, as S.~Ierom
\CNote{\Cite{Proem. lament.}}
teacheth, that as none can reade, or ſpeale wordes, vnles they firſt
lerne to know the letters: ſo we can not attaine to know the greater
Myſteries in holie Scriptures, except we beginne with moral good life,
according to that the Prophet ſayth:
\CNote{\XRef{Pſal.~118. v.~104.}}
By thy commandments I haue vnderſtood.
\MNote{The Church hath al good properties requiſite.}
And therfore wiſe Salomon, by inſtinct of the Holie Ghoſt, as wel by the
maner of ſtile, as by the doctrine conteined, concludeth his Booke of
Parables, with praiſe both of the Church in general, which hath al the
vertues, and good properties here mentioned;
\MNote{A faithful ſoule hath them at leaſt in deſire.}
and of euerie faithful ſoule, ſincerely ſeruing God, which either in
dede, or in deſire of mind, hath ſuch part of them, as may ſuffice to
the attaining of eternal life. For concerning the whole Church,
S.~Auguſtin in two Sermons
\Cite{(217.~and~218. de temp.)}
S.~Beda
\Cite{vpon this place,}
and other Fathers ſhew it euidently. Touching alſo particular ſoules,
not only of holie men, but alſo of vvemen, the frailer ſexe, holie
Scriptures, and Eccleſiaſtical monuments yelde manie examples, beſides
the moſt Excellent and immaculate virgin Mother of God;
\MNote{Rare and excellent wemen both in the old and nevv teſtament.}
as of Sara, Rebecca, Lia, Rachael, Elizabeth, Marie Magdalen, Martha,
and innumerable others, moſt precious pearles, deare ſpouſes of Chriſt,
and ſingular ornaments of his Church.}
A valiant woman
\SNote{A vvoman of ſuch perfectiõ as is here deſcribed, is in dede rare,
yet poſſible to be found.}
who shal finde? far, and from the vtmoſt borders is the price of her. \V
The hart of her husband truſteth in her, and he shal not neede
ſpoyles. \V She shal render good, and not euil, al the dayes of her
life. \V She hath ſought wool and flaxe, and hath wrought by the
counſel of her handes. \V She is become as a marchants shippe, bringing
her bread from farre. \V And she hath riſen in the night, and geuen pray
to her houshold, and meates to her handmaides. \V She hath vewed a
filde, and bought it, of the fruite of her handes she hath planted a
vineyard. \V She hath gyrded her loines with ſtrength, and hath
ſtrengthened her arme. \V She hath taſted, and ſene that her traficke is
good: her lampe shal not be extinguished in the night. \V She hath put
her hand to
%%% 1451
ſtrong thinges, and her fingers haue taken hold of
\SNote{Flaxe on the diſtafe ſignifieth purpoſe to do good workes, yarne
on the ſpindle the vvorke vvel begune, which geueth confidence in God
that the ſame ſhal be perfect and haue due revvard.}
the ſpindle. \V She hath opened her hand to the needie, and ſtretched
out her palmes to the poore. \V She shal not feare for her houſe in the
coldes of ſnow: for al her houshould are clothed with duble. \V Tapeſtrie
clothing she hath made to herſelf: ſilke, and purple is her garment. \V
Her husband is noble in the gates, when he shal ſitte with the ſenatours
of the land. \V She
\SNote{They make and ſel cloth, which lerning and obſeruing Gods lavv,
do teach it others.}
made ſindon, and ſold it, and deliuered a girdle to the Chananeite. \V
Strength and beautie is her garment, and she shal laugh in the later
day. \V She hath opened her mouth to wiſedom, and the law of clemencie
is in her tongue. \V She hath conſidered the pathes of her houſe, and
hath not eaten her bread idle. \V Her children aroſe, and commended her
to be moſt bleſſed: her husband, and he prayſed her. \V Manie daughters
haue gathered together riches: thou haſt paſſed them al. \V
\SNote{External comelines is not durable: but the feare of God is more
vvorthie of praiſe.}
Grace is deceitful, and beautie is vayne: the woman that feareth our
Lord shal be prayſed. \V Geue ye to her
\SNote{Good vvorkes ſhal be revvarded,}
of the fruite of her handes: and let her workes praiſe her
\SNote{at the tribunal ſeate of iudgement.}
in the gates.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
