%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 1434
%%% o-1323
\startChapter[
  title={Chapter 21}
  ]

\Summary{}


As
\MNote{Gods prouidence.}
diuiſions of waters, ſo the hart of the king is in the hand of our Lord:
whither ſoeuer he wil, he shal incline it.

\V
\MNote{%
\XRef{ch.~14. v.~12.}
&
\XRef{16. v.~2.}}
Euerie way of a man ſemeth to himſelf right: but our Lord weigheth the
hartes.

\V
\MNote{Mercie and iudgement.}
To doe mercie and iudgement, doth more pleaſe our Lord, then victimes.

\V
\MNote{Humilitie.}
Exaltation of the eies is the dilation of the hart: the lampe of the
impious ſinne.

\V
\MNote{Fortitude.}
The cogitations of the ſtrong are alwayes in abundance: but euerie
ſluggard is alwayes in pouertie.

\V
\MNote{Truth.}
He that gathereth treaſures with a lying tongue, is vaine and witles,
and shal ſtumble at the ſnares of death.

\V
\MNote{Equitie.}
The robberies of the impious, shal draw them downe, becauſe they would
not doe iudgement.

%%% 1435
\V
\MNote{Right in hart.}
The peruerſe way of a man is ſtrange: but he that is cleane, his worke
is right.

\V
\MNote{Domeſtical peace.}
It is better to ſitte in a corner of the houſe toppe, then with a
brawling woman, and in a common houſe.

\V
\MNote{Commiſeration.}
The ſoule of the impious deſireth euil, he wil not haue pitie on his
neighbour.

\V
\MNote{Chaſtiſment of ſinne.}
The peſtilent man being punished, the litle one wil be wiſer: and if he
folow the wiſeman, he wil take knowlege.

\V
\MNote{Charitie.}
The iuſt deuiſeth concerning the houſe of the impious, that he may draw
the impious from euil.

\V
\MNote{Almes dedes.}
He that ſtoppeth his eare at the crie of the poore, himſelf alſo shal
crie, and shal not be heard.

\V
\MNote{Almes in ſecrete.}
A gift hid
\Fix{quencketh}{quencheth}{obvious typo, fixed in other}
angers: and a gift in the boſome the greateſt indignation.

\V
\MNote{Iuſt puniſhment.}
It is a ioy to doe iudgement: and dread to them that worke iniquitie.

\V
\MNote{Obſeruation of orders.}
A man, that shal erre from the way of doctrine, shal abyde in the
aſſemblie of giantes.

\V
\MNote{Temperance & parcimonie.}
He that loueth good cheere, shal be in pouertie: he that loueth wine,
and fatte thinges, shal not be rich.

\V
\MNote{Pietie.}
The impious shal be geuen for the iuſt: and the vniuſt for the
righteous.

\V
\MNote{Domeſtical peace.}
It is better to dwel in a deſert land, then with a brawling and angrie
woman.

\V
\MNote{Frugalitie.}
Treaſure to be deſired, and oyle in the habitation of the iuſt: and the
vnwiſe man shal diſſipate it.

%%% o-1324
\V
\MNote{Iuſtice and mercie.}
He that foloweth iuſtice and mercie, shal finde life, iuſtice, and
glorie.

\V
\MNote{Fortitude.}
The wiſe hath ſealed the citie of the ſtrong, and hath deſtroyed the
confidence therof.

\V
\MNote{Guard of the tongue.}
He that kepeth his mouth, and his tongue, kepeth his ſoule from
diſtreſſes.

\V
\MNote{Modeſtie.}
The proude and arrogant is called vnlerned, which in anger worketh
pride.

\V
\MNote{Diligent trauel.}
Deſires kil the ſlothful: for his handes would not worke anything: \V
\V
\MNote{Liberalitie.}
al the day he longeth and deſireth: but he that is iuſt, wil geue, and
wil not ceaſe.

\V
\MNote{Sinceritie in religion.}
The hoſtes of the impious abominable, becauſe they are offered of
wickednes.

\V
\MNote{True teſtimonie and ſound doctrine.}
A lying witnes shal perish: an obedient man shal ſpeake victorie.

%%% 1436
\V
\MNote{Amendment of faultes.}
The impious man malepertly hardeneth his countenance: but he that is
righteous, correcteth his way.

\V
\MNote{Conformitie to Gods wil.}
There is no wiſdom, there is no prudence, there is no counſel againſt
our Lord.

\V
\MNote{Gods grace neceſſarie.}
The horſe is prepared to the day of battel: but our Lord geueth
ſaluation.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
