%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 1418
%%% o-1308
\startChapter[
  title={Chapter 11}
  ]

\Summary{}


A
\MNote{Equitie.}
deceitful balance, is abomination before God: and an equal weight is his
wil.

\V
\MNote{Humilitie.}
Where pride shal be, there shal be contumelie alſo: but where humilitie,
there alſo wiſdom.

\V
\MNote{Sinceritie.}
The ſimplicitie of the iuſt shal direct them: and the ſupplanting of the
peruers shal waſte them.

\V
\MNote{Spiritual riches.}
Riches shal not profit in the day of reuenge: but iuſtice shal deliuer
from death.

\V
\MNote{Pietie.}
The iuſtice of the ſimple shal direct his way: and the impious shal fal
in his impietie.

\V
\MNote{Iuſtice.}
The iuſtice of the righteous shal deliuer them: and the vniuſt shal be
caught in their deceitful practiſes.

\V
\MNote{Care of the other life.}
The impious man being dead, there shal be no hope any more: and the
expectation of the careful shal perish.

\V
\MNote{Future ioy.}
The iuſt is deliuered from diſtreſſe: and the impious shal be geuen for
him.

%%% 1419
\V
\MNote{Truth in wordes.}
The diſſembler with his mouth deceiueth his freind: but the iuſt shal be
deliuered by knowlege.

\V
\MNote{Common good.}
In the good thinges of the iuſt the citie shal reioyce: and in the
deſtruction of the impious there shal be prayſe.

\V
\MNote{Publique ioy.}
With the benediction of the iuſt the citie shal be exalted: and by the
mouth of the impious it shal be ſubuerted.

\V
\MNote{True freindſhipe.}
He that deſpiseth his freind, lacketh hart: but the wiſe man wil hold
his peace.

\V
\MNote{Fidelitie.}
He that walketh fraudulently, reueleth ſecretes: but he that is
faithful, concealeth the thing comitted of his freind.

\V
\MNote{Good gouernment.}
Where there is no gouernour, the people shal fal: but there is health
where is much counſel.

\V
\MNote{Prudence in ſuretiſhipe.}
He shal be afflicted with euil, that is ſuretie for a ſtranger: but he
that is ware of the ſnares, shal be ſecure.

\V
\MNote{Diligence.}
A gracious woman shal finde glorie: and the ſtrong shal haue riches.

\V
\MNote{Mercie.}
A merciful man doth good to his kinred alſo.

\V
\MNote{Iuſtice.}
The impious maketh an vnſtable worke: but to him that ſoweth iuſtice, is
a faythful reward.

\V
\MNote{Clemencie.}
Clemencie prepareth life: & the
\Fix{pureſewing}{purſewing}{obvious typo, fixed in other}
of euil thinges death.

\V
\MNote{Sinceritie.}
A peruers hart is abominable to our Lord: and his wil is in them that
walke ſimply.

\V
\MNote{Iuſt dealing.}
Hand in hand the euil man shal not be innocent: but the ſeede of the
iuſt shal be ſaued.

\V
\MNote{Internal vertue.}
A ring of gold in a ſwines ſnoute, a fayre woman & a foole.

%%% o-1309
\V
\MNote{Expectation of glorie.}
The deſire of the iuſt is al good: the expectation of the impious
furie.

\V
\MNote{Almes dedes.}
Some diuide their owne goodes, and are made richer: others take
violently not their owne, and are alwayes in pouertie.

\V
\MNote{Beneuolence.}
The ſoule which bleſſeth, shal be made fatte: and he that inebriateth
himſelf shal alſo be inebriated.

\V
\MNote{Care of the poore.}
He that hydeth corne, shal be curſed among the peoples: but bleſſing
vpon the head of them that ſel.

\V
\MNote{Good deſires.}
Wel riſeth he early who ſeeketh good thinges: but he that is a ſearcher
after euil thinges, shal be oppreſſed of them.

\V
\MNote{Truſt in Gods goodnes.}
He that truſteth in his riches shal fal: but the iuſt shal ſpring as a
greene leafe.

\V
\MNote{Frugalitie.}
He that trubleth his houſe, shal poſſeſſe the windes: and he that is a
foole shal ſerue the wiſe.

%%% 1420
\V
\MNote{Spiritual inſtruction.}
The fruite of the iuſt man a tree of life: and he that gaineth ſoules,
is wiſe.

\V
\MNote{Straict way of ſaluation.}
If the iuſt man receiue in the earth, how much more the impious and
ſinner?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
