%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 1407
%%% o-1297
\startChapter[
  title={Chapter 03}
  ]

\Summary{Wiſdom exhorteth to kepe Gods law (geuing long life) to obſerue
  mercie, and truth: 5.~to confide in God: 7.~to feare: 9.~and honour
  him: 11.~to take his correction gladly (13.~for al good thinges folow
  wiſdom) 27.~to releue the needie without delay: 30.~not to contend
  with the wicked, nor to imitate them. 33.~The euil shal faile, and the
  godlie shal proſper.}


My ſonne,
\SNote{It auaileth litle to heare good inſtructions, except we kepe them
in memorie,}
forget not my law, and let thy
\SNote{not in books only but in the hart:}
hart keepe my precepts. \V For they shal adde to thee length of
%%% 1408
dayes, and yeares of life, and peace. \V Let not
\SNote{and put them in execution.}
mercie and truth leaue thee, put them about thy throte; and write them
in the tables of thy hart:
%%% o-1298
\V and thou shalt finde grace, and good diſcipline before God and
men. \V
\SNote{Knovv alſo that al thy ſtreingth is in God, in whom thou maiſt
ſecurely truſt, not in thyn ovvne prudence.}
Haue confidence in our Lord with al thy hart, and leane not vpon thyne
owne prudence. \V In al thy wayes thinke on him, and he wil direct thy
ſteppes. \V Be not wiſe in thyne owne conceipte: feare God, and depart
from euil: \V for it shal be health to thy nauil, and watering of thy
bones. \V Honour our Lord with thy ſubſtance, and geue to him of the
firſt of al thy fruites: \V and thy barnes shal be replenished with
fulnes, and thy preſſes shal run ouer with wine. \V My ſonne, caſt not
away the diſcipline of our Lord: neither doe thou faint when thou art
\Fix{chaſteced}{chaſticed}{likely typo, fixed in other}
of him: \V
\SNote{Chaſtiſment and tribulation in thoſe that endeuour to ſerue God,
is a ſigne of his fauour tovvards them; and therfore his other promiſes
vvhich ſeme to be temporal, are to be vnderſtood of the next life.}
For whom our Lord loueth he chaſticeth: and as a father in the ſonne he
pleaſeth himſelf. \V Bleſſed is the man that findeth wiſdom, and floweth
with prudence: \V better is the purchaſing therof then marchandiſe of
ſiluer, and her fruite then chiefe and the pureſt gold: \V she is more
precious then al riches: and al thinges that are deſired, are not able
to be compared with her. \V
\SNote{God revvardeth as it vvere vvith both handes promiſing eternal life,}
Length of dayes in her right hand, & in her left hand
\SNote{and competent meanes in this life.}
riches and glorie. \V Her wayes are beautiful wayes, and al her pathes
peaceable. \V She is a tree of life to them that shal apprehend her: and
he that shal hold her is bleſſed. \V Our Lord by wiſdom founded the
earth, eſtablished the heauens by prudence. \V By his wiſdom the depthes
haue broken forth, and the cloudes waxe thicke with dew. \V My ſonne,
let not theſe thinges depart from thyne eies: kepe the law & counſel: \V
and there shal be life to thy ſoule, and
\SNote{Merite for the wordes of thy mouth.}
grace to thy iawes. \V Then shalt thou walke confidently in thy way, and
thy foote shal not ſtumble: \V if thou ſleepe, thou shalt not feare:
thou shalt reſt, and thy ſleepe shal be ſweete. \V Dread not at ſoden
terrour, and the power of the impious falling vpon thee. \V For our Lord
wil be at thy ſide, and wil keepe thy foote that thou be not taken. \V
Doe not prohibite him to doe good, that is able: if thou be able, thy
ſelfe alſo doe good. \V Say not to thy frend: Goe, and returne; and
\SNote{Almes in ſeaſon is duble vvorth to that vvhich is differred long.}
tomorow I wil geue to thee: wheras thou mayeſt geue forthwith. \V
Practiſe not euil againſt thy freind, when he hath affiance in thee. \V
Contend not againſt man without cauſe, wheras he hath done thee no
euil. \V Doe not enuie an vniuſt man, nor imitate his waies: \V becauſe
euerie
%%% 1409
deluder is an abomination of our Lord, and his communication is with the
ſimple. \V There is pouertie from our Lord in the houſe of the impious:
but the habitations of the iuſt shal be bleſſed. \V He shal delude the
deluders, and to the milde he wil geue grace. \V The wiſe shal poſſeſſe
glorie: the exaltation of fooles ignominie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
