%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 1421
%%% o-1310
\startChapter[
  title={Chapter 13}
  ]

\Summary{}


A
\MNote{Filial feare.}
wiſe ſonne, is the doctrine of the father: but he that is a ſcorner,
heareth not when he is rebuked.

\V
\MNote{Sound doctrine.}
Of the fruite of his owne mouth man shal be filled with good thinges:
but the ſoule of the preuaricatours is wicked.

\V
\MNote{Diſcretion in ſpeach.}
He that kepeth his mouth, kepeth his ſoule: but he that is vnaduiſed to
ſpeake shal feele euils.

%%% o-1311
\V
\MNote{Conſtant endeuour.}
The ſluggard wil and wil not: but the ſoule of them that worke, shal be
made fatte.

\V
\MNote{Truth in wordes and dedes.}
The iuſt shal deteſt a lying word: but the impious confoundeth, and shal
be confounded.

\V
\MNote{Iuſtice.}
Iuſtice kepeth the way of the innocent: but impietie ſupplanteth the
ſinner.

\V
\MNote{True acknowlegement of temporal ſtate.}
There is one as it were with riches wheras he hath nothing: and there is
as it were poore, wheras he is in much riches.

%%% 1422
\V
\MNote{Mediocritie of riches.}
The redemption of a mans life, his riches: but he that is poore, beareth
not reprehenſion.

\V
\MNote{Sincere conſcience.}
The light of the iuſt maketh glad: but the candle of the impious shal be
extinguished.

\V
\MNote{Humilitie.}
Among the prowde there are alwayes brawles: but they that doe al thinges
with counſel, are ruled by wiſdom.

\V
\MNote{Iuſt gaine.}
Subſtance haſtened shal be diminished: but that which by litle and litle
is gathered with the hand, shal be multiplied.

\V
\MNote{Speedie performance of promiſe.}
Hope that is differred afflicteth the ſoule: a tree of life the deſire
of
\Fix{coning.}{cunning.}{likely typo, fixed in other}

\V
\MNote{Good reporte.}
Who ſo detracteth from any thing, he byndeth himſelfe for the time to
come: but he that feareth the precept, shal conuers in peace.

Guilful
\MNote{Plaine dealing.}
ſoules erre in ſinnes: the iuſt are merciful & haue pitie.

\V
\MNote{Good inſtruction.}
The law of a wiſe man a fountaine of life, that he may decline from the
ruine of death.

\V
\MNote{True doctrine.}
Good doctrine shal geue grace: in the way of contemners a whirlepoole.

\V
\MNote{Prudence.}
The ſubtel man doth al thinges with counſel: but he that is a foole
openeth folie.

\V
\MNote{Faithful meſſage.}
The meſſenger of the impious shal fal into euil: but a faithful legate,
is health.

\V
\MNote{Deſire to lerne.}
Pouertie, and ignominie, to him that forſaketh diſcipline: but he that
yeldeth to him that rebuketh, shal be glorified.

\V
\MNote{Hate of ſinne.}
Deſire if it be accomplished, delighteth the ſoule: fooles deteſt them
that flee euil thinges.

\V
\MNote{Conuerſation with the good.}
He that walketh with the wiſe, shal be wiſe: a freind of fooles shal be
made like.

\V
\MNote{Iuſtice.}
Euil purſeweth ſinners, and to the iuſt good thinges shal be repayed.

\V
\MNote{Succeſſion.}
The good man leaueth heyres, ſonnes and nephewes: and the ſubſtance of
the ſinner is kept for the iuſt.

\V
\MNote{Frugalitie.}
Much meate in the tilled growndes of the fathers: and to others they are
gathered without iudgement.

\V
\MNote{Chaſtiſment.}
He that ſpareth the rod, hateth his childe: but he that loueth him doth
inſtantly nurture him.

\V
\MNote{Contentment of ſtate.}
The iuſt eateth and filleth his ſoule: but the bellie of the impious
vnſatiable.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
