%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 1409
%%% o-1299
\startChapter[
  title={Chapter 04}
  ]

\Summary{The wiſeman exhorteth others by his owne example to ſeeke
  wiſdom, 14.~to decline from the wicked, and to imitate the iuſt,
  23.~to guide wel the hart, mouth, and feete.}


Children heare ye the fathers diſcipline, and attend that you may knowe
prudence. \V I wil geue you a good gift, forſake not my law. \V For
\SNote{As Salomon was inſtructed by his father king Dauid ſo he teacheth
others, the right order hovve to lerne vviſdom.}
I alſo was the ſonne of my father, tender and as onlie begotten in my
mothers ſight: \V and he taught me, & ſayd: Let thy hart receiue my
wordes, kepe my preceptes, and thou shalt liue. \V Poſſeſſe wiſedom,
poſſeſſe prudence: forget not, neither decline from the wordes of my
mouth. \V Leaue her not, and she wil keepe thee: loue her, and she wil
preſerue thee. \V
\SNote{The firſt part of wiſdom is to deſire it. For nothing hindereth
from being iuſt but that iuſtice is not deſired.
\Cite{S.~Aug. in Pſal.~118. v.~20.}}
The beginning of wiſdom, poſſeſſe wiſdom, and in al thy poſſeſſion
purchaſe prudence: \V take quickly, and she wil exalt thee: thou shalt
be glorified of her, when thou shalt embrace her. \V She wil geue to thy
head increaſe of graces, and with a noble crowne she wil protect
thee. \V Heare my ſonne, and receiue my wordes, that yeares of life may
be multiplied to thee. \V The way of wiſdom I wil shew to thee, I wil
leade thee by the pathes of equitie. \V Which when thou shalt haue
entered, thy ſteppes shal not be ſtraytened, and running thou shalt not
haue a ſtumbling blocke. \V Hold diſcipline, leaue it not: kepe it,
becauſe the ſame is thy life. \V Be not delighted in the pathes of the
impious, neither let the way of the euil pleaſe thee. \V Flee from it,
neither paſſe thou by it: goe aſide, and forſake it. \V For they ſleepe
not vnleſſe they haue done il: and they take no ſleepe vnleſſe they
ſupplant. \V They eate the bread of impietie, and drinke the wine of
iniquitie. \V But the path of the iuſt, as shyning light proceedeth euen
to perfect day. \V The way of the impious is darkeſom: they know not
where they fal. \V My ſonne, heare my wordes, and incline thyne eare to
my ſayinges. \V Let them not depart from thyne eies, kepe them in the
middes of thy hart. \V For they are
%%% 1410
life to thoſe that finde them, and health to al flesh. \V With al garde
keepe thy
\SNote{As the hart is the principal part of the bodie, ſo the vvil is
the chiefeſt powre of the ſoule: from vvhich good or euil procedeth.}
hart, becauſe life proceedeth from it. \V Remoue from thee a froward
mouth, and let detracting lippes be far from thee. \V Let thyne eies ſee
right thinges, & let thine eieliddes goe before thy ſteppes. \V Direct
the path to thy feete, and al thy wayes shal be eſtablished. \V Decline
not to the right hand, nor to the left: turne away thy foote from
euil. \V For our Lord knoweth the wayes that are on the right hand: but
thoſe are peruers, which are on the left hand. \V But he wil make thy
courſes right, and thy wayes he wil bring forward in peace.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
