%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Proverbs
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 1410
%%% o-1300
\startChapter[
  title={Chapter 05}
  ]

\Summary{Againe wiſdom dehorteth from fornication (carnal and ſpiritual)
  6.~shewing that in the end ſinners shal ſee and feele the effect of
  their follie: 20.~which God ſeeth and wil punish.}


My ſonne, attend to my wiſdom, and to my prudence incline thyne eare, \V
that thou mayſt keepe
\SNote{To auoide al impietie it is firſt of al neceſſarie, not to
thinke, ſpeake, nor heare vnlawful thinges.}
cogitations, and thy lippes preſerue diſcipline. \V Attend not to the
deceipt of a
\SNote{By woman is generally vnderſtood concupiſcence of vvhat ſinne
ſoeuer, as
\XRef{ch.~1. v.~10.}
&
\XRef{ch.~3. v.~33.}
\XRef{ch.~4. v.~14.}}
woman: for the lippes of an harlot are as a hony combe diſtiling, and
her throte netter then oyle. \V But her later endes are bitter as
wormewood, and sharpe as a two edged ſword. \V Her feete goe downe into
death, and her steppes penetrate vnto hel. \V They walke not by the path
of life, her steppes are wandering, and vnſearcheable. \V Now therfore
my ſonne heare me, and depart not from the wordes of my mouth. \V Make
thy way far from her, and aproch not to the doores of her houſe. \V Geue
not thy honour
\SNote{The vvorld the flesh and the diuel are ſtrangers,}
to ſtrangers, and thy yeres to the
\SNote{and cruel enemies that render for revvard eternal damnation.}
cruel. \V Leſt perhaps ſtrangers be filled with thy ſtrength, and thy
labours be in an other mans houſe, \V and thou mourne in the end, when
thou shalt haue ſpent thy flesh and thy bodie, and ſay: \V Why haue I
deteſted diſcipline, and my hart conſented not to reprehenſions, \V nor
I heard the voice of them that taught me, and haue not inclined mine
eare to maſters? \V I haue almoſt bene in al euil, in the middes of the
church and of the ſynagoge. \V Drinke water of thyne owne ceſterne, and
the ſtremes of thy wel: \V
\SNote{Good doctrine is to be imperted to men of ſincere intention,}
Let thy fountaines be deriued abrode, and in the ſtreates diuide thy
waters. \V
\SNote{not to contemners and obſtinate infidels.}
Haue them alone, neither let ſtrangers be partakers with thee. \V Let
thy vayne be bleſſed, and reioyce with the woman of thy youth: \V a
hinde moſt deare, and
%%% 1411
a moſt grateful fawne: let her breaſtes inebriate thee at al time, in
her loue be thou delighted continually. \V Why art thou ſeduced my ſonne
of a ſtrange woman, and art cherished in the boſome of an other? \V Our
Lord beholdeth the wayes of a man, and conſidereth al his ſteppes. \V
His owne iniquities take the impious, and he is faſt bonde with the
ropes of his ſinnes. \V He shal die becauſe he hath not had diſcipline,
and in the multitude of his folie he shal be deceiued.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:


  
