%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 24
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-24


\project douay-rheims


%%% 0533
%%% o-0479
\startChapter[
  title={Chapter 24}
  ]

\Summary{In conſideration of diuers principal benefites here recited,
  14.~Ioſue exorteth the people to ſerue God ſincerly, ſeing it is in
  their choiſe to do wel or euil. 16.~They promiſe al true ſeruice and
  obedience to God. 25.~VVherupon he reneweth the pact betwen God and
  them, writing it in the volume of the law, and erecting a great ſtone
  in teſtimonie. 29.~He dieth and is buried in mount
  Ephraim. 32.~Ioſephs bones are buried in Sichem. 33.~Eleazar the high
  prieſt alſo dieth and is buried in Ephraim.}

And Ioſue gathered together al the tribes of Iſrael into Sichem, and
called the ancientes, and princes, and iudges, and maiſters: and they
ſtoode in the ſight of our Lord: \V and to the people he ſpake in this
maner: Thus ſaith our Lord the God of Iſrael: Beyond the riuer did your
fathers dwel from the beginning, Thare the father of Abraham, and
\SNote{The Iſraelites deſcended alſo of Nachor, by Rebecca, his ſonnes
daughter, the wife of Iſaac.
\XRef{Gen.~24.}}
of Nachor: and
\LNote{They ſerued falſe goddes.}{It
\MNote{Thare ſometime ſerued falſe goddes, but Abraham neuer.}
is euident by this place, that Thare, and ſome other progenitors of
Iſrael ſometimes ſerued falſe goddes, from which they were reduced: but
Abraham was euer preſerued in true religion; and the whole familie of
Thare was therfore perſecuted in Chaldea, as S.~Auguſtin ſheweth
\Cite{li.~16. c.~13. de ciuit.}
Likewiſe Theodoret,
\Cite{q.~18. in Ioſue,}
and other both ancient and late writers teach the ſame, as is already
noted
%%% !!! Fix XRef
\XRef{pag.~203}}
they ſerued ſtrange goddes. \V I tooke therfore your father Abraham from
the coaſtes of Meſopotamia: and brought him into the Land of Chanaan:
and multiplied his ſeede, \V and gaue him Iſaac: and againe to him I
gaue Iacob and Eſau. Of whom, to Eſau I gaue mount Seir to poſſeſſe: but
Iacob, and his children went downe into Ægypt. \V And I ſent Moyſes and
Aaron, and ſtroke Ægypt with many ſignes and wonders. \V And I brought
you and your fathers out of Ægypt, and you came to the ſea: and the
Ægyptians purſewed your fathers with chariotes and horſemen, as farre as
the Read ſea. \V And the children of Iſrael cried to the Lord: who did
put darkenes betwen you and the Ægyptians, and brought the ſea vpon
them, & ouerwhelmed
%%% o-0480
them. Your eies ſaw al thinges that I did in Ægypt, and you dwelt in the
wildernes a great time: \V and I brought you into the Land of the
Amorrheite, which dwelt beyond Iordan. And when they fought againſt you,
I deliuered them
%%% 0534
into your handes, and you poſſeſſed their Land, and ſlew them. \V And
there roſe Balac the ſonne of Sephor king of Moab, and fought againſt
Iſrael. And he ſent and called Balaam the ſonne of Beor, that he might
curſe you: \V and I would not heare him, but contrariwiſe by him I
bleſſed you, and deliuered you out of his hand. \V And you paſſed
Iordan, and came to Iericho. And the men of that citie fought againſt
you, the Amorrheite, and Pherezeite and Chananeite, and the Hetheite,
and Gergeſeite, and the Heueite, and Iebuſeite: and I deliuered them
into your handes. \V And I ſent before you hornettes: and I caſt them
forth out of their places, the two Kinges of the Amorrheites, not in thy
ſword and bow. \V And I gaue you the Land, wherein you laboured not, and
the cities which you built not, to dwel in them: vineyardes and oliue
trees, which you planted not. \V Now therfore feare our Lord and ſerue
him with a perfect and verie true hart: and
\SNote{Ioſue being a prophet ſaw ſome of their hartes inclined to
idoles, though exteriorly they then had none among them.
\Cite{S.~Aug. q.~29. in Ioſue.}}
take away the goddes, which your fathers ſerued in Meſopotamia and in
Ægypt, and ſerue our Lord. \V But if it like you not to ſerue our Lord,
choiſe is geuen you: chooſe this day that which pleaſeth you, whom you
ought eſpecially to ſerue, whether the goddes, which your fathers ſerued
in Meſopotamia, or the goddes of the Amorrheites, in whoſe Land you
dwel: but I and my houſe wil ſerue our Lord. \V And the people anſwered,
and ſaid: God forbid we should leaue our Lord, and ſerue ſtrange
goddes. \V Our Lord God he brought vs, and our fathers out of the Land
of Ægypt, out of the houſe of ſeruitude: and did in our ſight great
ſignes, and kept vs in al the way, by the which we walked, and among al
the peoples, through which we paſſed. \V And he hath caſt out al
the nations, the Amorrheite inhabiter of the Land, which we haue entred. We
therfore wil ſerue our Lord, becauſe he is our God. \V And Ioſue ſaid to
the people: You can not ſerue our Lord: for God is holie, and a mightie
æmulator, neither wil he pardon your wickednes and ſinnes. \V If you
leaue our Lord, and ſerue ſtrange goddes, he wil turne him ſelf, and wil
afflict you, and ouerthrow you after he hath geuen you good thinges. \V
And the people ſaid to Ioſue: No, it shal not be ſo as thou ſpeakeſt,
but we wil ſerue our Lord. \V And Ioſue ſaid to the people: You are
witneſſes, that your ſelues haue choſen to you our Lord for to ſerue
him. And they anſwered: Witneſſes.
%%% 0535
\V Now therfore, quoth he, take away ſtrange goddes out of the middes of
you, and incline your hartes to our Lord the God of Iſrael. \V And the
people ſaid to Ioſue: We wil ſerue our Lord God, and wil be obedient to
his preceptes. \V Ioſue therfore in that day
\SNote{This renouation of the ſame couenãt prefigured the law of the new
Teſtament.
\Cite{S.~Aug. q.~30. in Ioſue.}}
made a couenant, and propoſed to the people preceptes and iudgementes in
Sichem. \V He wrote alſo al theſe wordes in the volume of the law of our
Lord: and he tooke a very great ſtone, and put it vnder the oke, that
was in the Sanctuarie of our Lord:
%%% o-0481
\V and ſaid to al the people: Behold this ſtone shal be a teſtimonie for
you, that 
\SNote{To the more confuſion of reaſonable creatures wilfully
offending, vnſenſible things are made witneſſes, becauſe they euer obey
Gods wil, which is the beſt maner of hearing.
\Cite{Theod. q.~19. in Ioſue.}}
it hath heard al the wordes of our Lord, which he hath ſpoken to you:
leſt perhaps hereafter you wil denie, and lye to our Lord your God. \V
And he diſmiſt the people, euerie one into their poſſeſſion. \V
\SNote{If Ioſue writ the reſt of this booke, then Samuel added theſe
laſt verſes.
\Cite{Hiſt. Schol.}}
And after theſe thinges Ioſue the ſonne of Nun the ſeruant of our Lord
died, being a hundred and ten yeares old: \V and
\LNote{They buried.}{In
\MNote{Before Chriſt none entered into heauen.}
that no mention is made of mourning for Ioſue, S.~Hierom noteth a
myſterie, and a ſpecial point of Chriſtian doctrin: It ſemeth to me
(ſaieth he
\Cite{Epiſt. de 42.~Mauſ. manſ.~33.}
that in Marie prophecie is dead, in Moyſes and Aaron, an end is put to
the law and prieſthood of the Iewes. For ſo much as they could neither
paſſe into the land of promiſe, nor bring the beleuing people out of the
wildernes of this world. And
\Cite{(Mauſ.~34.)}
Aaron (ſayeth he) was mourned, (and ſo was Moyſes) Ieſus is not
mourned. That is, in the law was deſcent into hel
(\Emph{called} \L{limbus}) in the Goſpel is paſſage to paradiſe.}
they buried him in the coaſtes of his poſſeſſion in Thamnathſare, which
is ſituated in the mountaine of Ephraim, on the North part of mount
Gaas. \V And Iſrael ſerued our Lord al the daies of Ioſue, and of the
ancientes, that liued a long time after Ioſue, and that had knowen al
the workes of our Lord which he had done in Iſrael. \V The
\SNote{Ioſephs \Emph{Mauſoleum} (or famous ſepulchre) remained in Sichem
in S.~Hieroms time, as he witneſſeth.
\Cite{Tradit, Hebra. in Gen. prope finem.}}
bones alſo of Ioſeph which the children of Iſrael had taken out of
Ægypt, they buried in Sichem, in part of the field, which
\CNote{\XRef{Gen.~50.}
\XRef{Exo.~13.}}
Iacob had bought of the ſonnes of Hemor the father of Sichem, for a
hundred yong ewes, and it was in the poſſeſſion of the ſonnes of
Ioſeph. \V  Eleazar alſo the ſonne of Aaron died: and they buried him in
Gabaath of Phinees his ſonne, which was geuen him in mount Ephraim.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
