%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 0501
%%% o-0451
\startChapter[
  title={Chapter 7}
  ]

\Summary{For the ſinne of Achan, reſeruing ſecretly to himſelf certaine
  money, and other precious thinges, the Iſraelites are beaten in
  battel. 13.~But the offender being found out, and ſtoned to death,
  Gods wrath is turned from them.}

But the children of Iſrael tranſgreſſed the commandment, & vſurped of
the anathema. For Achan the ſonne of Charmi, the ſonne of Zabdi, the
ſonne of Zare of the tribe of Iuda, tooke ſomewhat of the anathema: and
our Lord was angrie againſt the children of Iſrael. \V And when Ioſue
ſent from Iericho men againſt Hai, which is beſide Bethauen, at the Eaſt
ſide of the towne of Bethel, he ſaid to them: Goe vp, and view the Land:
who accomplishing his commandmentes, viewed Hai. \V And returning they
ſaid to him: Let not al the people goe vp, but let two or three thouſand
men goe, and deſtroy the citie: why ſhal al the people be vexed in vaine
againſt verie few enemies? \V There went vp therfore three thouſand
fighting men. Who immediatly turning their backes, \V were ſtrooken of
the men of the citie of Hai, and there fel of them
\SNote{So God tempered his puniſhment that but few were ſlaine, and
afterwards gaue the towne to the Iſraelites without loſſe of any of
their men.}
ſix and thirtie men: and the aduerſaries purſewed them from the gate as
farre as Sabarim, and they ſtricke them flying away by the deſcent: and
the hart of the people was much afrayd, and melted like vnto water. \V
But Ioſue rent his garmentes, and fel flatte on the ground before the
arke of our Lord vntil euening, as wel he as al the ancientes of Iſrael:
and they caſt duſt vpon their heades, \V and Ioſue ſaid: Alas ô Lord
God, why wouldeſt thou bring this people ouer the riuer of Iordan, to
deliuer vs into the handes of the Amorrheite, and to deſtroy vs? Would
God as we beganne, we had taried beyond Iordan. \V My Lord God what shal
I ſay, ſeeing Iſrael turning their backes to their enemies? \V The
Chananeites shal heare of it, and al the inhabitantes of the Land, and
being gathered together in a plumpe shal compaſſe vs about, & shal
deſtroy our name from the earth:
%%% o-0452
and what wilt thou doe to thy greate name? \V And our Lord ſaid to
Ioſue: Ariſe,
\SNote{Prayer wil not auaile til iuſtice be firſt donne.}
why lieſt thou flatte on the ground? \V Iſrael hath ſinned, and
tranſgreſſed my couenant: and they haue taken of the anathema, and haue
ſtolen and lyed, and haue hid it among their veſſel. \V Neither can
Iſrael ſtand before his enemies, and he shal flee them: becauſe he is
polluted with the anathema. I wil
%%% 0502
be no more with you, til you diſpatch him, that is guiltie of this
wicked fact. \V Ariſe, ſanctifie the people, and ſay to them: Be
ſanctified againſt to morrow: for thus ſaith our Lord God of Iſrael:
There is anathema in the middes of thee ô Iſrael: thou canſt not ſtand
before thyne enemies, til he be deſtroyed out of thee that is
contaminated with this wicked fact. \V And you shal come in the morning
euerie one by your tribes: and what tribe ſoeuer the lote shal finde, it
shal come by the kindredes therof, the kinred by the houſes, and the
houſe by the men. \V And whoſoeuer he be that shal be taken in this
fact, he ſhal be burnt in the fyre with al his ſubſtance, becauſe he
hath tranſgreſſed the couenãt of our Lord, and hath done abomination in
Iſrael. \V Ioſue therfore ryſing in the morning, made Iſrael to come by
their tribes, and it was found the tribe of Iuda. \V Which being
preſented by the families therof, it was found the familie of
Zare. Preſenting that alſo by the houſes, he found it Zabdi: \V whoſe
houſe diuiding into euerie man, he found Achan the ſonne of Charmi, the
ſonne of Zabdi: the ſonne of Zare of the tribe of Iuda. \V And Ioſue
ſaid to Achan: My ſonne, geue glorie to our Lord God of Iſrael, and
confeſſe, and tel me what thou haſt done, hide it not. \V And Achan
anſwered Ioſue, and ſaid to him: In deede I haue ſinned to our Lord the
God of Iſrael, and thus and thus haue I done. \V For I ſaw among the
ſpoiles a cloke of ſcarlet verie good, and two hundred ſicles of ſiluer,
and a golden rule of fiftie ſicles: and coueting I tooke it away, and
hid it in the ground againſt the middes of my tabernacle, and the ſiluer
I couered with the earth digged vp. \V Ioſue therfore ſent miniſters:
who running to his tabernacle, found al thinges hid in the ſame place,
and the ſiluer withal. \V And taking it away out of the tent, brought it
to Ioſue, and to al the children of Iſrael, and threw it before our
Lord. \V Ioſue therfore taking Achan the ſonne of Zare, and the ſiluer
and the cloke, and the golden rule, his ſonnes alſo and daughters, his
oxen and aſſes, and ſheepe, and the tabernacle it ſelf, and al the
ſtuffe: (and al Iſrael with him) they brought them to the valley of
Achor: \V where Ioſue ſaid: Becauſe thou haſt diſturbed vs, our Lord
diſturbe thee in this day. And al Iſrael ſtoned him: and al thinges that
were his, were conſumed with fyre. \V And gathered together vpon him a
greate heape of ſtones, which
%%% 0503
remaineth vntil this preſent day. And the furie of our Lord was auerted
from them. And the name of that place was called, The valley of Achor,
vntil this day.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
