%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 0520
%%% o-0468
\startChapter[
  title={Chapter 17}
  ]

\Summary{The half tribe of Manaſſes (eldeſt ſonne of Ioſeph) receiue
  their lotte, 3.~including the daughters of Salphaad: 14.~with an
  enlargement of inheritance to the ſame tribes of Ephraim and
  Manaſſes.}

And this lotte fel to the tribe of Manaſſes (for he is the firſt borne
of Ioſeph) to Machir the firſt borne of Manaſſes the father of Galaad,
who was a warlike man, and had for poſſeſſion Galaad and Baſan: \V and
to the reſt of the children of Manaſſes according to their families, to
the children of Abiezer, and to the children of Helec, and to the
children of Eſriel, and to the children of Sechem, and to the children
of Hepher, and to the children of Semida. Theſe are the children of
Manaſſes the ſonne of Ioſeph, males by their kinredes. \V But Salphaad
the ſonne of Hepher the ſonne of Galaad the ſonne of Machir the ſonne of
Manaſſes had no ſonnes, but onlie daughters: whoſe names be theſe,
Maala, and Noa and Hegla and Melcha and Therſa. \V And they came in the
preſence of Eleazar the prieſt, and of Ioſue the ſonne of Nun, and of
the princes, ſaying: Our Lord commanded by the hand of Moyſes, that a
poſſeſſion should be geuen vs in the middes of our brethren. And he gaue
them according to the commandement of our Lord a poſſeſſion in the
middes of their fathers brethren. \V And the cordes fel to Manaſſes ten,
beſide the Land of Galaad and Baſan beyond Iordan. \V For the daughters
of Manaſſes poſſeſſed inheritance in the middes of his ſonnes. And the
Land of Galaad fel to the lotte of the children of Manaſſes that
remayned. \V And the border of Manaſſes from Aſer, was Machmathath
%%% 0521
which looketh to Sichem: and goeth out on the right hand beſide the
inhabitantes of the Fountaine of Taphua. \V For in the lotte of Manaſſes
was fallen the Land of Taphua,
\SNote{The land or territorie of Taphua fallĩg to the lotte of Manaſſes,
yet Taphua (that is the citie it ſelfe) was the childrens of Ephraim.}
which is beſide the borders of Manaſſes, the childrens of Ephraim. \V
And the border of the Reede valley went downe into the ſouth of the
torrent of the cities of Ephraim, which are in the middes of the cities
of Manaſſes: the border of Manaſſes on the North of the torrent, and the
iſſue therof goeth to the ſea: \V ſo that the poſſeſſion of Ephraim is
on the South, and on the North of Manaſſes, and the ſea incloſeth both,
and they be ioyned one to an other in the tribe of Aſer on the
\Fix{Noth,}{North,}{obvious typo, fixed in other}
and in the tribe of Iſſachar on the Eaſt. \V And the inheritance of
Manaſſes in Iſſachar and in Aſer was Bethſan and the villages therof,
and Ieblaam with the villages therof, and the inhabitantes of Dor, with
the townes therof, the inhabitantes alſo of Endor with the townes
therof. And in like maner the inhabitantes of Thenac with the
townes therof, and the inhabitantes of
%%% o-0469
Mageddo with the townes therof, and the third part of the citie of
Nopheth. \V Neither could the children of Manaſſes ouerthrow the cities,
but the Chananeite began to dwel in
\SNote{The Chananeite remained in the Land of Manaſſes, for a time.}
his Land. \V But after that the children of Iſrael grew to be ſtrong,
they ſubdewed the Chananeites, and made them their tributaries, neither
did they kil them. \V And the children of Ioſeph ſpake to Ioſue, and
ſaid: Why haſt thou geuen me the poſſeſſion of one lotte and corde,
wheras I am of ſo great a multitude, and our Lord hath bleſſed me? \V To
whom Ioſue ſaid: If thou be a great people, goe vp into the wood, and
cutte thee roome in the Land of the Pherezeite and Raphaims: becauſe the
poſſeſſion of mount Ephraim is narrow for thee. \V To whom the children
of Ioſeph anſwered: We can not goe vp to the mountaines, wheras the
Chananeite that dwel in the champayne countrie, wherein are ſituated
Bethſan with the townes therof, and Iezrael poſſeſſing the middes of the
valley, vſe yron chariottes. \V And Ioſue ſaid to the houſe of Ioſeph,
of Ephraim and Manaſſes: Thou art a great people, and of great ſtrength,
thou shalt not haue one lotte, \V but thou shalt paſſe to the mountaine,
and shalt cutte and make glades for thee to inhabite: and mayeſt procede
farder, when thou haſt ſubuerted the Chananeite, whom thou ſayeſt to
haue yron chariotes, and to be very ſtrong.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
