%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 0513
%%% o-0462
\startChapter[
  title={Chapter 13}
  ]

\Summary{God
\MNote{The third part.

Partitiõ of the land among nine tribes & a half.}
commandeth Ioſue to diuide the land (deſcribing the limites therof)
amongſt nine tribes and a half. 8.~With a recapitulation of the partes
already geuen, on the other ſide Iordan, to the other two tribes and a
half. The tribe of Leui
\XRef{(v.~14. &~33.)}
hath their prouiſion in other maner.}

Ioſue was old, and ſtriken in age, and our Lord ſaid to him: Thou art
old, and of a great age, and there is a verie large countrie left, which
is not yet diuided by lotte: \V to witte, al Galilee, Philiſthijm, and
al Geſturi. \V From the trubled riuer, that watereth Ægypt, vnto the
borders of Accaron againſt the North: the Land of Chanaan, which is
diuided vnto fiue Lordes of the Philiſthiimes, the Gazeites, the
%%% 0514
Azotians, the Aſcalonites, the Getheites, and the Accaronites. \V But on
the South ſide are the Heueites, al the Land of Chanaan, and Maara of
the Sidonians as farre as Apheca, and the borders of the Amorrheite, \V
and his confines. The countrie alſo of Libanus againſt the Eaſt from
Baalgad vnder mount Hermon, til thou enter into Emath. \V Of al that
dwel in the mountaine from Libanus, vnto the Waters Maſerephoth, and al
the Sidonians. I am he that wil deſtroy them from the face of the
children of Iſrael.
\SNote{Theſe partes are deſigned though not yet conquered.}
Let it come therfore into a portion of the inheritance of Iſrael, as I
haue commanded thee. \V And now diuide the Land in poſſeſſion to the
nine tribes, and to the half tribe of Manaſſes, \V with the which Ruben
& Gad haue poſſeſſed the Land, which Moyſes the ſeruant of our Lord
deliuered to them beyond the ſtreames of Iordan, on the eaſt ſide. \V
From Aroer, which is ſituate on the banke of the torrent Arnon, and in
the middes of the valley, and al the champayne of Medaba, as farre as
Dibon: \V and al the cities of Cehon, the King of the Amorrheite, which
reigned in Heſebon vnto the borders of the children of Ammon. \V And
Galaad, and the border of Geſſuri and
\Fix{Maccati,}{Machati,}{obvious typo, fixed in other}
and al mount Hermon, and al Baſan, as farre as Salecha, \V al the
Kingdome of Og in Baſan, which reigned in Aſtaroth and Edrai, he was of
the reliques of the Raphaims: and Moyſes ſtroke, and deſtroyed them. \V
And the children of Iſrael would not deſtroy Geſſuri and Machati: and
they haue dwelt in the middes of Iſrael vntil this preſent day. \V But
to the tribe of Leui he gaue no poſſeſſion: but the ſacrifices and
victimes of our Lord the God of Iſrael, that is his inheritance, as he
ſpake to him. \V Moyſes therfore
%%% o-0463
gaue poſſeſſion to the children of Ruben according to their kinredes. \V
And their border was from Aroer, which is ſituate on the banke of the
torrent Arnon, and in the middes of the valley of the ſame torrent: al
the plaine, that leadeth to Medaba, \V and Heſebon, and al their
villages, which are in the champayne. Dibon alſo, and Bamothbaal, and
the towne Baalmaon, \V and Iaffa, and Cedimoth, and Mephaath, \V and
Cariathaim, and Sabama, and Sarathaſar in the mountaine of the
Valley. \V Bethfogor and Aſedoth, Phaſga and Bethieſimoth, \V and al the
champayne cities, and al the Kingdomes of Sehon the King of the
Amorrheite, that reigned in Heſebon, whom Moyſes ſtroke with the princes
of
%%% 0515
Madian: the Heueite, & Recem, and Sur, and Hur, and Rebe Dukes of Sehon
inhabitantes of the Land. \V And Balaam the ſonne of Beor the
ſoothſayer, did the children of Iſrael kil by the ſword with the reſt
that were ſlayne. \V And the riuer of Iordan was made the border of the
children of Ruben. This is the poſſeſſion of the Rubenites by their
kinredes of cities and villages. \V And Moyſes gaue to the tribe of Gad
and to his children poſſeſſion by their kinredes, the diuiſion wherof is
this. \V The border of Iaſer, and al the cities of Galaad, and the half
part of the Land of
\SNote{This part the Amorreithes had taken and poſſeſſed, otherwiſe the
Iſraelits were prohibited to take anie thĩg from the Ammonites.
\XRef{Deut.~2.}}
the children of Ammon: as farre as Aroer, which is againſt Rabba: \V and
from Heſebon vnto Ramoth, Maſphe and Betonim: and from Manaim vnto the
borders of Dabir. \V In the valley
\Fix{alo}{alſo}{obvious typo, fixed in other}
Bethhara, and Bethnemra, and Socoth, and Saphon the other part of the
Kingdom of Sehon the King of Heſebon: the end of this alſo is Iordan,
vnto the vttermoſt part of the ſea Cenereth beyond Iordan on the eaſt
ſide. \V This is the poſſeſſion of the children of Gad by their
families, their cities, and villages. \V He gaue alſo to the half tribe
of Manaſſes, and their children poſſeſſion according to their kinredes,
\V the beginning wherof is this: from Manaim al Baſan, and al the
kingdoms of Og the King of Baſan, and al the villages of Iair, which are
in Baſan, threeſcore townes. \V And the half part of Galaad, and
Aſtaroth, and Edrai, cities of the kingdom of Og in Baſan: to the
children of Machir, the ſonne of Manaſſes to the half part of the
children of Machir according to their kinredes. \V This poſſeſſiõ
diuided Moyſes in the champayne countries of Moab, beyond Iordan,
againſt Iericho on the Eaſt ſide. \V But to the tribe of Leui he gaue no
poſſeſſion: becauſe our Lord the God of Iſrael him ſelf is their
poſſeſſion, as he ſpake to them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
