%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 0507
%%% o-0457
\startChapter[
  title={Chapter 10}
  ]
  
\Summary{Fiue kinges of the Amorrheites beſieging Gabaon, becauſe it is
  confederate with Iſrael, 6.~Ioſue with his armie defeateth theirs,
  killing and purſewing them. 11.~Manie alſo are ſlaine with haile
  ſtones. 12.~At the prayer of Ioſue the ſunne and moone ſtand ſtil the
  ſpace of one day. 22.~The fiue kinges are hanged on gibbetes. 28.~He
  taketh alſo and ſubdueth diuers cities 40.~and countries.}

VVhich thinges when Adoniſedec king of Ieruſalem had heard, to witte,
that Ioſue had taken Hai, and had ſubuerted it (for as he had done to
Iericho & the king therof, ſo did he to Hai, & their king) and that the
Gabaonites were fled to Iſrael, and were their confederates, \V he was
ſore afrayd. For Gabaon was a great citie, and one of
%%% 0508
the kinglie cities, and greater then the towne of Hai, and al their men
of warre moſt valiant. \V Therfore Adoniſedec king of Ieruſalem
ſent to Oham king of Hebron, and to Pharam king of Ierimoth, to Iaphia
alſo king of Lachis, and to Dabir king of Eglon, ſaying: \V Come vp to
me, and bring ayde, that we may ouercome Gabaon, becauſe it reuolted to
Ioſue, & to the children of Iſrael. \V Therfore the fiue kinges of the
Amorrheites being aſſembled went vp: the king of Ieruſalem, the king of
Hebron, the king of Ierimoth, the king of Lachis, the king of Eglon,
together with their hoſtes, & camped about Gabaon, aſſaulting it. \V But
the inhabitantes of the citie Gabaon which was beſieged, ſent to Ioſue,
who then abode in the campe at Galgal, & ſaid to him: withdraw not thy
handes from the helpe of thy ſeruantes: come vp quickly and deliuer vs,
and bring ayde: for there are aſſembled againſt vs al the kinges of the
Ammorrheites, which dwel in the mountaines. \V And Ioſue went vp from
Galgal, and al the hoſte of the men of warre with him moſt valiant men.
\V And our Lord ſaid to Ioſue: Feare them not: for I haue deliuered them
into thy handes: none of them shal be able to reſiſt thee. \V Ioſue
therfore came in vpon them ſodenly, going vp al the night from
Galgal. \V And our Lord trubled them at the ſight of Iſrael: and
deſtroyed them with a greate ſlaughter in Gabaon, and purſewed them by
the way of the aſcent to Bethoron, and ſtroke them vnto Azeca and
Maceda. \V And when they fled the children of Iſrael, and were in the
deſcent of Bethhoron, our Lord ſent vpon them greate ſtones from heauen
as farre as Azeca: and there died farre more with the ſtones of haile,
then whom the children of Iſrael had ſtrooken with the ſword. \V Then
ſpake Ioſue to our Lord in the day, that he deliuered the Amorreite in
the ſight of Iſrael, and ſaid before them: Thou Sunne
%%% o-0458
againſt Gabaon moue not, and thou
\SNote{Ioſue did thinke if the moone moued the ſunne alſo muſt
neceſſarily moue ſo he obtained the ſtay of both.}
Moone againſt the valley of Aialon. \V And the Sunne and Moone ſtoode
ſtil, til the people reuenged themſelues of their enemies. Is not this
written in the booke of the iuſt? The Sunne therfore ſtood ſtil in the
middes of heauen, and haſtened not to goe downe the ſpace of one day.
\V There was not before nor
\SNote{Til after the time this booke was written.}
after ſo long a day, our Lord
\SNote{God condeſcending to worke ſo great a miracle at the inſtance of
his ſeruant.}
obeying the voice of a man, and fighting for Iſrael. \V And Ioſue
returned with al Iſrael into the campe of Galgal. \V For the fiue kinges
were fled, and had hid themſelues
%%% 0509
in a caue of the citie of Maceda. \V And it was told Ioſue that the fiue
kinges were found lying hid in a caue of the citie of Maceda. \V Who
commanded them ſaying: Rowle great ſtones into the mouth of the caue,
and ſette induſtrious men, which may keepe them ſhut in: \V and ſtand
not you ſtil, but purſew the enemies, and kil al the hindermoſt of them
that flee, neither let them enter into the fortes of their cities, whom
our Lord God hath deliuered into your handes. \V The aduerſaries
therfore being ſlaine with a great ſlaughter, and almoſt conſumed to
vtter deſtruction, they that could eſcape from Iſrael,
\SNote{God ſo diſpoſed, that they ſhould not cõquer al in one yeare:
\Emph{leſt the land had benne brought into a vvildernes, and beaſtes
increaſed againſt them}.
\XRef{Exod.~23. v.~29.}
Morally it ſignifieth, that the children of God muſt be exerciſed in
tribulations and mortification leſt vices grow in them.
\Cite{Procop. in Exod.}}
entered into fenſed cities. \V And al the hoſte returned to Ioſue in
Maceda, where then the campe was, ſafe and the ful number: and no man
durſt once mutter againſt the children of Iſrael. \V And Ioſue commanded,
ſaying: Open the mouth of the caue, and bring forth to me the fiue
kinges, that lie hid therin. \V And the miniſters did as it was
commanded them: and they brought to him the fiue kinges out of the caue,
the king of Ieruſalem, the king of Hebron, the king of Ierimoth, the
king of Lachis, the king of Eglon. \V And when they were brought forth
to him, he called al the men of Iſrael, and ſaid to the princes of the
hoſte that were with him: Goe, and ſette your feete vpon the neckes of
theſe kinges. Who when they had gone, and troden with their feete the
neckes of them lying vnder, \V againe he ſaid to them: Feare ye not,
neither dread, take courage and be ſtrong: for ſo wil our Lord doe to al
your enemies, againſt whom you fight. \V And Ioſue ſtroke, and ſlew
them, and hanged them vpon fiue gibbettes: & they hung vntil euening.
\V And when the ſunne was downe, he commanded the ſoldiars to take them
downe from the gibbettes. Who caſt them being taken downe into the caue,
wherin they had lyen hid, & put on the mouth therof great ſtones, which
continew vntil this preſent. \V The ſame day Ioſue tooke Maceda and
ſtroke it in the edge of the ſword, & killed the king & al the
inhabitantes therof: he left not in it ſo much as ſmale reliques. And he
did to the king of Maceda, as he had done to the king of Iericho. \V
And he paſſed with al Iſrael from Maceda vnto Lebna, and fought againſt
it: \V which our Lord deliuered with the king therof into the handes of
Iſrael: and they ſtroke the citie in the edge of the ſword, and al the
inhabitantes therof. They left not in it anie remaines. And
%%% 0510
they did to the king of Lebna, as they had done to the king of Iericho.
\V From Lebna he paſſed vnto Lachis with al Iſrael: and placing the
hoſte round about aſſaulted it.
%%% o-0459
\V And our Lord deliuered Lachis into the handes of Iſrael, and he tooke it
the day folowing, and ſtroke it in the edge of the ſword, and euerie
ſoule, that was in it, as he had done to Lebna. \V At that time went vp
Horam the king of Gazer, to ayde Lachis: whom Ioſue ſtroke with al his
people to vtter deſtruction. \V And he paſſed from Lachis vnto Eglon,
and compaſſed it, \V and wonne it the ſame day: & ſtroke in the edge of
the ſword al the ſoules, that were in it according to al thinges that
he had done to Lachis. \V He went vp alſo with al Iſrael from Eglon vnto
Hebron, and fought againſt it: \V tooke it, and ſtroke it in the edge of
the ſword, the king alſo therof, and al townes of that countrie, & al
the ſoules, that remained in it: he left not therein anie remaynes: as
he had done to Eglon, ſo did he alſo to Hebron, al thinges that he found
in it conſumning with the ſword. \V Thence returning vnto Dabir, \V he
tooke it, and waſted it: the king alſo therof and al the townes round
about he ſtroke in the edge of the ſword: he left not in it anie
remaines: as he had done to Hebron and Lebna and to their kinges, ſo did
he to Dabir and the king therof. \V Ioſue therfore ſtroke al the hillie
countrie and ſouth and champaine, and Aſedoth with their kinges: he left
not in it anie reliques, but euerie thing that could breath he ſlew, as
our Lord the God of Iſrael had commanded him, \V from Cadeſbarne vnto
Gaza. Al the Land of Goſen vnto Gabaon, \V and al their kinges, and
countries he tooke and waſted at one aſſault: for our Lord the God of
Iſrael fought for him. \V And he returned with al Iſrael to the place of
the campe in Galgal.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
