%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 0523
%%% o-0471
\startChapter[
  title={Chapter 19}
  ]

\Summary{The ſecond lotte falleth to the tribe of Simeon, the ſituation
  of whoſe inheritance is deſcribed, with the names of their principal
  cities. 10.~The third
  %%% 0524
  in like ſorte, to Zabulon. 17.~The fourth to Iſſachar. 24.~The fifth
  to Aſer. 32.~The ſixt to Nepthali. 40.~And the ſeuenth to Dan. 49.~Al
  the Land being diſtributed among the tribes, with common conſent they
  geue a ſpecial citie to Ioſue, in the middes of them.}

And the ſecond lotte came forth of the children of Simeon by their
kinreddes: and their inheritance was, \V in the middes of the poſſeſſion
of the children of Iudas: Berſabee and Sabee and Molada, \V and
Haſerſual, Bala and Aſem, \V and Eltholad, Bethul, and Harma, \V and
Siceleg and Bethmarchaboth and Haſerſuſa, \V and Bethlebaoth and
Sarohen: thirtene cities, and their townes. \V Ain and Remmon and Athor
and Aſan: foure cities, and their townes: \V al the litle townes round
about theſe cities vnto Baalath Beer Ramath againſt the ſouth
quarter. This is the inheritance of the children of Simeon according to
their kinredes, \V in the poſſeſſion and corde of the children of Iudas:
becauſe it was greater, and therfore the children of Simeon poſſeſſed in
the middes of their inheritance. \V And the third lotte fel of the
children of Zabulon by their kinredes: and the border of their
poſſeſſion was made as farre as Sarid. \V And it went vp from the ſea
and Merala, and came into Debbaſeth: as farre as the torrent, which is
againſt Ieconam. \V And it returneth from Sared againſt the Eaſt into
the endes of Ceſeleth thabor: and it goeth out to Dabereth, and ryſeth
vp againſt Iaphie. \V And thence it paſſeth along to the eaſt ſide of
Geth hepher and Thacaſim: and goeth out into Remmon, Amthar and Noa. \V
And it compaſſeth to the North of Hanathon: and the iſſues therof are the
valley Iephtahel, \V and Cateth and Naalol and Semeron and Ierala and
Bethlehem: twelue cities, and their townes. \V This is the inheritance
of the tribe of the children of Zabulon by their kinredes, the cities
and their litle townes. \V The fourth lotte came forth to Iſſachar by
their kinredes. \V And his inheritance was Iezrael and Caſaloth and
Sunem, \V and Hapharaim and Sehon, and Anaharath \V and Rabboth and
Ceſion, Abes, \V and Rameth, and Engannim, and Enhadda and
Bethpheſes. \V And the border therof cometh to Thabor and Seheſema and
Bethſames: and their iſſues were Iordan: ſixtene cities, and their
townes. \V This is the poſſeſſion of
%%% o-0472
Iſſachar by their kinredes, the cities, and their litle townes.
%%% 0525
\V And the fifth lotte fel to the tribe of the children of Aſer by their
kinredes: \V and their border was Halcath and Chali and Beten and
Axaph, \V and Elmelec and Amaad and Meſſal: and it reacheth to Carmel of
the ſea and Sihor and Labanath. \V And it returneth againſt the eaſt of
Bethdagon: and paſſeth along to Zabulon and the Valley Iephthael againſt
the North into Bethemec and Nehiel. And it goeth out to the left ſide of
Cabul, \V and Abran and Rohob and Hamon and Cana, as farre as great
Sidon. \V And it returneth into Horma vnto the verie wel fenſed citie
Tyre, and vnto Hoſa: and the iſſues therof shal be into the ſea from the
corde of Achziba: \V and Amma and Aphec and Rohob. Cities twentie two,
and their townes. \V This is the poſſeſſion of the children of Aſer by
their kinredes, and the cities and their townes. \V Of the ſonnes of
Nepthali fel the ſixt lotte by their families: \V and the border beganne
from Heleph and Elon into Saanaim, and Adami, which is Neceb, and
Iebnael vnto Lecum: and their iſſues vnto Iordan: \V and the border
returneth againſt the Weſt into
\Fix{Azanoththabor,}{Azanotthabor,}{likely typo, fixed in other}
and thence goeth out into Hucuca, and paſſeth along into Zabulon againſt
the South, and into Azer againſt the Weſt, and into Iuda vnto Iordan
againſt the riſing of the ſunne. \V Cities very wel fenſed, Aſſedim,
Ser, and Emath, and Reccath and Cenereth, \V and Edema and Arama,
Aſor, \V and Cedes and Edrai, Enhaſor \V and Ieron and Magdalel, Horem
and Bethanath and Bethſames: ninetene cities, and their townes. \V This
is the poſſeſſion of the tribe of the children of Nephthali by their
kinredes, the cities and their townes. \V To the tribe of the children
of Dan by their families came forth the ſeuenth lotte: \V and the border
of their poſſeſſion was Sara and Eſthaol, and Hiriſemes, that is the
ſunne. \V Selebin and Aialon and Iethela, \V Elon and Themna and
Acron, \V Elthece, Gebbethon and Balaath, \V and Iud and Bane and Barac
and Gethremmon: \V and Meiarcon & Arecon, with the border that looketh
toward Ioppe, \V and is ſhut vp with the ſame end. And the children of
Dan went vp, and fought againſt Leſem, and they tooke it: and they
ſtroke it in the edge of the ſword, and poſſeſſed, and dwelt in it,
calling the name of it Leſem Dan, by the name of Dan the father
therof. \V This is the poſſeſſion of the tribe of the ſonnes of Dan, by
their kinredes, the cities and their townes. \V And when he had made
%%% 0526
an end of diuiding the Land by lotte to euerie one by their tribes,
\SNote{Of modeſtie Ioſue would not aſſigne to him ſelf anie place, but
the whole people freely granted his requeſt.}
the children of Iſrael gaue poſſeſſion to Ioſue the ſonne of Nun in the
middes of them, \V according to the commandement of our Lord, the citie
which he requeſted, Thamnath Saraa in mount Ephraim: and he built the
citie, and dwelt in it. \V Theſe are the poſſeſſions, which Eleazar the
prieſt, and Ioſue the ſonne of Nun, and the princes of the families, and
of the tribes of the children of Iſrael, diuided by lotte in Silo,
before our Lord at the doore of the tabernacle of teſtimonie, and they
parted the Land.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
