%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 0522
%%% o-0469
\startChapter[
  title={Chapter 18}
  ]

\Summary{From the campe of Iſrael in Silo, ſurueyers are ſent to diuide
  the reſt of the Land into ſeuen partes, for the ſeuen tribes yet
  without portions. 10.~VVhich being donne, Ioſue caſteth lottes for
  them, 11.~and the firſt lotte falleth to Beniamin, 12.~whoſe part is
  deſcribed by the limites, 21.~with the names of the principal cities.}

And al the children of Iſrael were aſſembled in Silo, and there they
pitched the tabernacle of the teſtimony, and the Land was ſubdewed to
them. \V But there remained ſeuen tribes of the children of Iſrael,
which as yet had not receiued their poſſeſſions. \V To whom Ioſue ſaid:
How long are you ſlack with cowardenes, and enter not to poſſeſſe the
Land, which our Lord the God of your fathers hath geuen you? \V Chooſe
of euerie tribe three men, that I may ſend them, and they may goe and
circuite the Land, and marke it out according to the number
\SNote{Not equal but proportionable partes were aſſigned, for ſo it was
commanded.
\XRef{Num.~26. v.~54.}
\Emph{To the greater number to geue a greater portion, and to the fewer
a leſſe}.}
of euerie multitude: and report vnto me that which they haue marked
out. \V Diuide vnto you the Land into ſeuen partes: let Iudas be in his
boundes on the ſouth quarter, and the houſe of Ioſeph on the North. \V
The Land in the middes betwen theſe marke out into ſeuen partes: and you
shal come
%%% o-0470
hither to me, that
\SNote{After the portions were appointed by mens induſtrie & diſcretion,
God confirmed the ſame by lotte, to take away al occaſiõ of
diſcontentment.}
before our Lord your God I may caſt the lotte for you: \V for the
Leuites part is not among you, but the prieſthood of our Lord is
their inheritance. And Gad and Ruben, and the half tribe of Manaſſes had
now receiued their poſſeſſions beyond Iordan at the Eaſt ſide: which
Moyſes the ſeruant of our Lord gaue them. \V And when the men were ryſen
vp, that they might goe to marke out the land, Ioſue commanded them,
ſaying: Circuite the Land and marke it out, and returne to me: that here
before our Lord, in Silo I may caſt the lotte for you. \V They therfore
went on: and going ouer it, diuided it into ſeuen partes, writing it in
a volume. And they returned to Ioſue into the campe in Silo. \V Who did
caſt lottes before our Lord in Silo, and diuided the Land to the
children of Iſrael into ſeuen partes. \V And firſt came vp the lotte of
the children of Beniamin by their families, to poſſeſſe the Land betwen
the children of Iudas and the children of Ioſeph. \V And their border
was againſt the North from Iordan: going forward by the ſide of Iericho
on the north quarter, and thence Weſtward ryſing vp vnto the
%%% 0523
mountaines, and reaching to the wildernes of Bethauen, \V & paſſing
through by Luza to the South, the ſame is Bethel: and goeth downe into
Ataroth addar vnto the mountaine, that is on the South of Bethhoron the
lower. \V And it bendeth compaſſing againſt the ſea, Southward of the
mountaine that looketh to Bethhoron againſt the South: and the iſſues
therof are into Cariathbaal,
\Fix{which is called alſo Cariathbaal which is called alſo
Cariathiarim,}{which is called alſo Cariathiarim,}{obvious typo, fixed
in other}
a citie of the children of Iudas. This is their coaſt againſt the ſea,
toward the Weſt. \V But on the South from part of Cariathiarim the
border iſſueth forth againſt the ſea, and cometh to the fountaine of the
waters of Nephtoa. \V And it goeth downe into part of the mountaine that
looketh toward the Valley of the children of Ennom: and is againſt the
north quarter in the vttermoſt part of the Valley raphaim. And it goeth
downe into Gehennom (that is, the valley of Ennom) by the ſide of the
Iebuſeite to the South: and cometh to the Fountaine of Rogel, \V paſſing
to the north, and going forth to Enſemes, that is to ſay, the fountaine
of the ſunne: \V and it paſſeth vnto the litle hilles, that are againſt
the aſcent of Adommim: and goeth downe to Abenboen, that is, the ſtone
of Boen the ſonne of Ruben: and it paſſeth on the north ſide to the
champaine countries: and goeth downe into the playne, \V and paſſeth by
againſt the North of Bethagla: and the iſſues therof are againſt the
brincke of the moſt ſalt ſea on the North in the end of Iordan to the
ſouth quarter: \V which is the border therof on the Eaſt. This is the
poſſeſſion of the children of Beniamin by their borders round about, and
their families. \V And their cities were, Iericho and Bethhagla and
Vallis Caſis, \V Beth Araba and Samaraim and Bethel, \V and Auim and
Aphara and Ophera, \V Towne Emona and Ophni and Gabee: twelue cities, &
their townes. \V Gabaon and Rama and Beroth, \V and Meſphe, and Caphara,
and Amoſa, \V and Recem, Iarephel and Tharela, \V and Sela, Eleph, and
Iebus, which is Ieruſalem, Gabaath and Cariath: fourteenne cities, and
their townes. This is the poſſeſſion of the children of Beniamin by
their families.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
