%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 0491
%%% o-0442
\startChapter[
  title={Chapter 2}
  ]

\Summary{Two
\Fix{diſcouers}{diſcouerers}{obvious typo, fixed in other}
ſent into Hiericho are hid, and concealed by Rahab: 8.~and vpon promiſe
of like ſafetie to her whole familie, 21.~she helpeth them ſecretly
away.}

Therfore Ioſue the ſonne of Nun ſent from Setim two men, to ſpie in
ſecrete: and ſaid to them: Goe, and view the Land, and the citie of
Iericho. Who going entred into the houſe of a woman a harlot, named
Rahab, and reſted with her. \V And it was told the king of Iericho, and
ſaid: Behold there are men come in hither by night of the children
%%% 0492
of Iſrael, to ſpie the Land. \V And the king of Iericho ſent to Rahab,
ſaying: Bring forth the men, that came to thee, and are entred into thy
houſe: for they be ſpies, and are come to view al the Land. \V And the
woman taking the men, hid them, and ſaid: I confeſſe they came to me,
%%% o-0443
but
\SNote{Notwithſtanding this officious lie, (which is a venial ſinne)
S.~Paul,
\XRef{Heb.~11.}
and S.~Iames
\XRef{(c.~2.)}
teſtifie, that ſhe was iuſtified by her faith in God, and by good workes
towards theſe men.
\Cite{S.~Aug. cont. Mendac. c.~17.}
See
\XRef{Annot. Iac.~2. v.~25.}}
I knew not whence they were: \V and when the gate was a shutting in the
darke, and they withal went out, I know not whither they be gone: purſew
quickly, and you shal ouertake them. \V But she made the men to goe vp
into the roofe of her houſe, and couered them with the ſtalke of flaxe,
which was there. \V And they that were ſent, folowed them, the way that
leadeth to the ford of Iordan: and they being gone out the gate
forwith was shutte. \V Neither were they yet a ſleepe that lay hidde,
and behold the woman went vp to them, and ſaid: \V I know that the Lord
hath geuen this Land to you: for your terrour is fallen vpon vs, and al
the inhabitantes of the Land are become fainte. \V We haue heard that
the Lord dried vp the water of the Redſea at your entring, when you came
out of Ægypt: and what thinges you did to the two kinges of the
Amorrheites, that were beyond Iordan: Sehon and Og, whom you ſlew. \V And
hearing theſe thinges we greatly feared, and our hart fainted, neither
did there remaine ſpirite in vs at your entring in: for the Lord your
God he is God in heauen aboue, & in the earth beneth. \V Now therfore
ſweare to me by the Lord, that as I haue done mercie with you, ſo you
alſo doe with my fathers houſe: and you geue me a true ſigne, \V that
you ſaue my father and mother, my brethren and ſiſters, and al thinges
that
\Fix{he}{be}{obvious typo, fixed in other}
theirs, and deliuer our ſoules from death. \V Who anſwered her: Be our
liues for you vnto death,  only if thou betray vs not. And when our Lord
shal haue deliuered vs the land, we wil doe in thee mercie and truth.
\V She therfore did let them downe by a corde out of a window: for her
houſe ioyned faſt to the wall. \V And she ſaid to them: Goe vp to the
mountaines, leſt perhaps they meete you returning: and there lie hid
three daies, til they returne, and ſo you shal goe on your way. \V Who
ſaid to her: We shal be quitte from this oath, wherwith thou haſt ſworne
vs: \V if we entring the Land, there be this purple corde a ſigne, and
thou tie it in the window, by the which thou haſt let vs downe: and
gather thy father and mother, and brethren and al thy kindred into thy
houſe. \V He that ſhal goe forth
%%% 0493
of the doore of thy houſe, his bloud ſhal be vpon his head, and we shal
be quitte. But the bloud of al, that shal be with thee in the houſe,
shal redound vpon our head, if any man touch them. \V But if thou wilt betray
vs, and vtter this talke abroade, we shal be quitte from this oath,
wherwith thou haſt adiured vs. \V And ſhe anſwered: As you haue ſpoken,
ſo be it done, and diſmiſſing them to goe, she hong the purple corde in
the window. \V But they walking came to the mountaines, and taried there
three dayes, til they returned that purſewed them: for ſeeking euerie
way, they found them not. \V Who being entred into the citie, the
diſcouerers returned, and came downe from the mountaine: and paſſing
ouer Iordan, they came to Ioſue the ſonne of Nun, and told him al
thinges that had chanced to them, \V and ſaid: Our Lord hath deliuered
al this land into our handes, and al the inhabitantes therof are
ouerthrowen with feare.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
