%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 0527
%%% o-0474
\startChapter[
  title={Chapter 21}
  ]

\Summary{Cities with ſuburbes are aſsigned to the tribe of Leui. 4.~To
  the ſonnes of Caath by the line of Aaron being prieſtes, thirtenne,
  5.~to the reſt of Caaths progenie, being Leuites, tenne. 6.~To the
  ſonnes of Gerſon Leuites thirtenne. 7.~To the ſonnes of Merari Leuites
  (34.~of a lower degree) twelue. 9.~With the names of al the cities:
  39.~in al fourtie eight. 41.~So Gods promiſe is fully performed hauing
  geuen the whole Land to Iſrael in peaceable poſſeſsion.}

And the princes of the families of the Leui came to Eleazar the prieſt,
and Ioſue the ſonne of Nun, and to the chief of the kinredes in euerie
tribe of the children of Iſrael: \V and they ſpake to them in Silo of
the Land of Chanaan, and ſaid: Our Lord commanded by the hand of Moyſes,
that cities should be geuen vs to inhabite, and their ſuburbes to feede
cattel. \V And the children of Iſrael gaue of their poſſeſſions
according to the commandement of our Lord, cities and their ſuburbes. \V
And the lotte came forth vnto the familie of Caath of the children of
Aaron the prieſt out of the tribe of Iudas, and Simeon, and Beniamin,
thirtene cities. \V And to the reſt of the children of Caath, that is to
the Leuites, which remayned, out of the tribes of Ephraim, and Dan, and
the halfe tribe of Manaſſes, ten cities. \V Moreouer to the children of
Gerſon came forth a lotte, that they should take of the tribes of
Iſſachar and Aſer and Nephthali, and the halfe tribe of Manaſſes in
Baſan, cities in number thirtene. \V And to the ſonnes of Merari by
their kinredes, of the tribe of Ruben and Gad and Zabulon, twelue
cities. \V And the children of Iſrael gaue to the Leuites cities and
their ſuburbes, as our Lord commanded by the hand of Moyſes, geuing to
euerie one by lotte. \V Of the tribes of the children of Iudas and
Simeon Ioſue gaue cities: whoſe names be theſe, \V to the children of
Aaron by the families of Caath of the Leuitical ſtocke (for the firſt
lotte came forth to them) \V Cariatharbe the father of Enac, which is
called Hebron, in the mountaine of Iudas, and the ſuburbes therof round
about. \V But the fieldes and the townes therof he had geuen
%%% 0528
to Caleb the ſonne of Iephone to poſſeſſe. \V He gaue therfore to the
children of Aaron the prieſt Hebron a citie of refuge, and the ſuburbes
therof: & Lobna with the ſuburbes therof: \V and Iether and Eſtemo, \V
and Holon, and Dabir, \V and Ain, and Ieta, and Bethſames, with the
ſuburbes therof: nine cities of two tribes, as hath bene ſaid. \V And of
the tribe of the children of Beniamin, Gabaon, and Gabae, \V and
Anathoth and Almon, with their ſuburbes: foure cities. \V Al the cities
together of the children of Aaron the prieſt, thirtene, with their
%%% o-0475
ſuburbes. \V But to the reſt by the families of the children of Caath of
the Leuitical ſtocke was geuen this poſſeſſion. \V Of the tribe of
Ephraim, the cities of refuge, Sichem with the ſuburbes therof in the
mountayne of Ephraim, and Gazer \V and Cibſaim, and Beth horon, with the
ſuburbes therof, foure cities. \V Of the tribe of Dan alſo, Eltheco and
Gabathon, \V and Aialon and Gethremmon, with the ſuburbes therof, foure
cities. \V Moreouer of the half tribe of Manaſſes, Thanac and
Gethremmon, with their ſuburbes two cities. \V Al the cities ten, and
their ſuburbes, were geuen to the children of Caath of the inferiour
degree. \V To the children of Gerſon alſo of the Leuitical ſtocke he
gaue of the half tribe of Manaſſes the cities of refuge, Gaulon in
Baſan, and Boſram, with their ſuburbes, two cities. \V Moreouer of the
tribe of Iſſachar, Ceſion, and Dabereth, \V and Iaramoth, and Engannim,
with their ſuburbes, foure cities. \V And of the tribe of Aſer, Maſal
and Abdon, \V and Helcath, and Rohob, with their ſuburbes, foure
cities. \V Of the tribe alſo of Nephthali the cities of refuge, Cedes in
Galilee: and Hammoth Dor, and Carthan, with their ſuburbes, three
cities. \V Al the cities of the families of Gerſon, thirtene, with their
ſuburbes. \V And to the children of Merari Leuites of the inferiour
degree by their families was geuen of the tribe of Zabulon, Iecnam and
Cartha \V and Damna and Naalol, foure cities with their ſuburbes. \V Of
the tribe of Ruben beyond Iordan againſt Iericho the cities of refuge,
Boſor in the wildernes, Miſor and Iaſer and Iethſon and Mephaath, foure
cities with their ſuburbes. \V Of the tribe of Gad the cities of refuge,
Ramoth in Galaad, and Manaim and Heſebon and Iaſer, foure cities with
their ſuburbes. \V Al the cities of the children of Merari by their
families and kinredes, twelue. \V Therfore al the cities of the
%%% 0529
Leuites in the middes of the poſſeſſion of the children of Iſrael were
fourtie eight \V with their ſuburbes, euerie one diſtributed by the
families. \V And our Lord God gaue to Iſrael
\SNote{God gaue al the Land in due time, but not al at once, for the
cauſes expreſſed.
\XRef{Exo.~23. v.~29.}
\XRef{Deut.~7. v.~22.}}
al the Land, that he had ſworne he would geue to their fathers: and they
poſſeſſed it, and dwelt in it. \V And peace was geuen by him on al
nations round about: and none of their enemies durſt reſiſt them, but al
were brought into their dominion. \V Not ſo much certes as one word,
which he had promiſed, that he would performe vnto them, was fruſtrate,
but al thinges were accompliſhed in deedes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
