%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 0515
%%% o-0464
\startChapter[
  title={Chapter 14}
  ]

\Summary{Caleb of the tribe of Iuda (ſeing ſome knew their lottes
  already, and that the whole land was now to be diuided) 6.~demandeth,
  according to Gods promiſe made by Moyſes (for his true and good report
  of the ſame land, when he with others viewed it) that Hebron be geuen
  him, and his ſeede to inherite, 13.~which Ioſue confirmeth vnto him.}

This is it, which the children of Iſrael poſſeſſed in the Land of
Chanaan, which Eleazar the prieſt, and Ioſue the ſonne of Nun, & the
princes of the families by the tribes
%%% 0516
of Iſrael geue to them: \V diuiding al thinges by lotte, as our Lord had
commanded in the hand of Moyſes, to the nine tribes, and the half
tribe. \V For to two tribes and a half, Moyſes had geuen poſſeſſion
beyond Iordan: beſides the Leuites, which receiued no land among their
brethren: \V but
\SNote{The Leuites hauing their portions in other maner, yet there were
twelue tribes to receiue portiõs by reaſon that Iacob adopted Ioſephs
two ſonnes.
\XRef{Gen.~48.}}
into their place ſucceded the children of Ioſeph diuided into two
tribes, of Manaſſes and Ephraim: neither did the Leuites receiue other
portion in the Land, but cities to inhabite, and their ſuburbes to feede
their beaſtes and cattel. \V As our Lord had commanded Moyſes, ſo did
the children of Iſrael, and they diuided the Land. \V Therfore the
children of Iuda came to Ioſue in Galgal, and Caleb the ſonne of Iephone
the Cenezeite  ſpake to him: Thou knoweſt what our Lord ſpake to Moyſes
the man of God concerning me & thee in Cadesbarne. \V Fourtie yeares old
was I when Moyſes the ſeruant of our Lord ſent me from Cadesbarne, to
view the Land, and I reported to him that which to me ſemed true. \V But
my brethren, that had gone vp with me, diſcouraged the hart of the
people: and I neuertheles folowed our Lord my God. \V And Moyſes ſware
in that day, ſaying: The Land, which thy foote hath troden, shal be thy
poſſeſſion, and thy childrens for euer, becauſe thou haſt folowed our
Lord my God. \V Our Lord therfore hath granted me life, as he promiſed
vntil this preſent day. It is
\SNote{After the viewing of the land
\XRef{Num.~13.}
the Iſraelites remained in the deſert 38.~yeares: ſo they were nere
ſeuen yeares in warres.}
fourtie and fiue yeares, ſince our Lord ſpake this worde to Moyſes, when
Iſrael walked through the wildernes: this day am I eightie and fiue
yeares old \V ſo luſtie, as I was at that time when I was ſent to view:
the ſtrength of that time contineweth in me vntil this day, as wel to
fight as to goe. \V Geue me therfore this mountaine, which our Lord
promiſed, thy ſelfe alſo hearing it, wherein are the Enacims, and great
cities and fenſed:
\SNote{Gods promiſe is euer ſure on his part, but becauſe it is
conditional, if we ſerue him ſincerly, our wil being free, he ſaith:
\Emph{if perhaps our Lord be vvith me.}}
if perhaps our Lord be with me, and I ſhal be able to deſtroy them,
as he promiſed me. \V And Ioſue bleſſed him, and deliuered to him Hebron
in poſſeſſion. \V And from thence forth
\SNote{Onlie the countrie of Hebron was geuen to Caleb, for the citie it
ſelf and ſuburbes belonged to the Prieſtes.
\XRef{chap.~21. v.~11.}}
Hebron belonged to Caleb the ſonne of Iephone the Cenezeite, vntil this
preſent day: becauſe he folowed our
%%% o-0465
Lord the God of Iſrael. \V The name of Hebron before was called Cariath
Arbe: Adam the greateſt among the Enacims was layd there: and
\SNote{See before
\XRef{chap.~11. v.~23.}}
the Land ceaſed from battels.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
