%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 0499
%%% o-0449
\startChapter[
  title={Chapter 6}
  ]

\Summary{Some
\MNote{The ſecond part.

Of conquering the Land of promiſe.}
prieſtes carying the arke, others ſounding Iubilee trumpetes, armed men
going before, and the reſt of the people folowing, goe euerie day once,
ſix dayes together, and the ſeuenth day, ſeuen times, round about
Iericho. 16.~At laſt al making a great shoote, the walles fal downe, and
they entering in kil and deſtroy al (22.~ſauing Rahab and her kinred.)
The gold, ſiluer, braſse, and iron are brought into the
treaſurie. 26.~And he is curſed that shal build the citie againe.}

And Iericho was ſhut and fenſed, for feare of the children of Iſrael,
and no man durſt goe out or come in. \V And our Lord ſaid to Ioſue:
Behold I haue geuen into thy handes Iericho, and the king therof, and al
the valiant men. \V
\SNote{God appointed this long and ſolemne proceſſion to the end it might
appeare, that the walles of Iericho fel not by chance, nor by force of
mans induſtrie, but by the mightie hand of God.}
Goe round about the citie al you that be men of warre once a day: ſo
shal you doe ſix daies. \V And the ſeuenth day the prieſtes shal take vp
the ſeuen trumpettes, which are vſed in the Iubilee, and shal goe before
the arke of the couenant: and you shal goe about the citie ſeuen times,
and the prieſtes shal ſound with trumpetes. \V And when the voice of
the trumpet shal ſound in length and with a broken tune, and shal ſound
in your eares, al the people shal crie together with a verie greate
ſhoote, and the walles of the citie shal fall to the ground, and they
shal enter in euerie one at the place againſt which they shal ſtãd.
%%% o-0450
\V Ioſue therfore the ſonne of Nun called the prieſtes, and ſaid to
them: Take vp the arke of the couenant: and let ſeuen other prieſtes
take vp the ſeuen trũpettes of the iubilees, and march before the arke
of our Lord. \V To the people alſo he ſaid: Goe, and cõpaſſe the citie,
the armed going before the arke of our Lord. \V And when Ioſue had ended
his wordes, and the ſeuen prieſtes ſounded with ſeuen trumpettes before
the arke of the couenant of our Lord, \V and al the armed hoſte went
before, the reſt of the commõ people folowed the arke, and al places
ſounded with the trumpettes. \V But Ioſue had commanded the people,
ſaying: You shal not crie, neither shal your voice be heard, nor any
word goe out of your mouth: vntil the day come wherin I shal ſay to you:
Crie, and shoote. \V Therfore the arke of our Lord went about the citie
once a day, and returning into the campe, abode there. \V Ioſue therfore
ryſing in the night, the prieſtes tooke the arke of our Lord, \V and
ſeuen of them ſeuen trumpettes, which are vſed in the iubilee: and they
went before the arke of our Lord walking and ſounding: and the armed
people
%%% 0500
went before them, and the reſt of the common people folowed the arke,
and they ſounded with trumpettes. \V And they went round about the citie
the ſecond day once, and returned into the campe. So did they ſix
daies. \V But the ſeuenth day, ryſing vp early, they went about the
citie, as it was ordained, ſeuen times. \V And when in the ſeuenth going
about the prieſtes ſounded with the trumpettes, Ioſue ſaid to al Iſrael:
Make a shoote: for our Lord hath deliuered to you the citie: \V and let
this citie be anathema: and al thinges that are in it, to our Lord. Onlie
Rahab the harlot let her liue, with al that be with her in the houſe:
for ſhe hidde the meſſengers whom we ſent. \V But you beware you touch
not ought of thoſe thinges, that are commanded, and be guiltie of
preuarication, and
\SNote{Manie are held guiltie of ſinne, and are iuſtly puniſhed for the
fact of one or few, either becauſe they cõſented, or concealed, or
neglected to puniſh the offenders: or els they ſuffer temporal
affliction for their warning to abhorre ſinne, & for increaſe of their
merite.
\Cite{S.~Aug. q.~8. &~9. in Ioſue.}}
al the campe of Iſrael be vnder ſinne, and be trubled. \V But whatſoeuer
gold or ſiluer there shal be, and of braſen veſſels and yron, let it be
conſecrated to our Lord, layd vp in his treaſures. \V Therfore al the
people making a shoote, and the trumpettes ſounding, after that the
voice and the ſound thundred in the eares of the multitude, the walles
forthwith fell: and euerie man went vp by the place, that was againſt
him: and they tooke the citie, \V and killed al thinges that were in it,
from man to woman, from the infant to the old man. The oxen alſo and sheepe,
and the aſſes they ſtroke in the edge of the ſword. \V But to the two
men that had bene ſent for ſpies, Ioſue ſaid: Goe into the houſe of the
woman the harlotte, and bring her forth, and al thinges that be hers, as
you aſſured her by oath. \V And the young men going in, brought out
Rahab, and her parentes, her brethren alſo and al her ſtuffe and kinred,
and made them to tarie without the campe. \V But the citie, and al
thinges, that were found therein they burnt; except the gold and ſiluer,
and braſen veſſels, and yron, which they conſecrated vnto the treaſurie
of our Lord. \V But Rahab the
%%% o-0451
harlotte and the houſe of her father, and al that ſhe had, Ioſue cauſed
to liue, and they dwelt in the middes of Iſrael vntil this preſent day:
for that she hidde the meſſengers, which he had ſent to view Iericho. At
that time, Ioſue pronounced a curſe, ſaying: \V
\SNote{Hiel fel into this curſe for reparing Iericho.
\XRef{3.~Reg.~16. v.~34.}}
Curſed be the man before our Lord, that shal rayſe vp and build the citie
of Iericho. In his firſt borne lay he the fundations therof, and in the
laſt of his children ſette he vp the gates therof. \V Our Lord therfore
was with Ioſue, and his name was bruited in al the earth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
