%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-23


\project douay-rheims


%%% 0532
%%% o-0478
\startChapter[
  title={Chapter 23}
  ]

\Summary{Ioſue being old admonisheth the people to keepe Gods
  commandementes, 12.~and to auoide mariages, and al ſocietie with
  gentiles, leſt falling to idolatrie God as certainly punish them, as
  he hath benne benificial, and bountiful to them.}

And when much time was paſſed, after that our Lord had geuen peace to
Iſrael, al the nations being ſubdewed round about, and Ioſue now verie
ancient, and of a great age: \V Ioſue called al Iſrael, and the elders,
and the princes and the iudges, and the maiſters, and ſaid to them: I am
old, and farre gone in age: \V and you ſee al thinges, that our Lord
your God hath done to al the nations round about,
\SNote{God fought for the Iſraelites three
\Fix{ma}{mãner}{obvious typo, fixed in other}
of wayes: ſometimes alone, they not fighting at al, as when the
Ægyptians were drowned in the read ſea: ſometimes they doing his
cõmandmentes he apparently aſſiſted them, as in the ſiege of Iericho,
the walles miraculouſly fel downe
\XRef{(c.~6.)}
& haile ſtones killed their enimies
\XRef{(chap.~10.)}
but moſt times inuiſibly, as wel by geuing them courege, as by ſtriking
their enimies with terrour. And al theſe wayes God alſo fighteth for his
ſeruantes in ſpiritual warres againſt the diuel, the fleſh, & the
world.}
how him ſelfe hath fought for you: \V and now becauſe he hath by lotte
diuided to you al the Land, from the Eaſt part of Iordan vnto the great
ſea, and manie nations yet remaine: \V Our Lord your God wil deſtroy
them, and take them away from your face, and you shal poſſeſſe the Land,
as he hath promiſed you. \V Onlie take courage, and be careful, that you
keepe al thinges which be written in the volume of the law of Moyſes:
and decline not from them neither to the right hand nor to the left: \V
leſt after that you are entred in to the Gentiles, which ſhal be among
you, you ſweare in the name of their goddes, and ſerue them, and adore
them: \V but cleaue to our Lord your God: which you haue done vntil this
day. \V And then our Lord God wil take away in your ſight the great
nations and very ſtrong, and no man shal be able to reſiſt you. \V One
of you ſhal purſew a thouſand men of the enemies: becauſe our Lord your
God him ſelf wil fight for you, as he hath promiſed. \V This onlie
beware very diligently before hand, that you loue our Lord your God. \V
But if you wil cleaue to the errour of theſe nations, that dwel among
you, and make mariages with them, and ioyne amitie: \V euen now know
%%% o-0479
ye that our Lord your God wil not deſtroy them before your face, but
they ſhal be a pitte and a ſnare for you, and a ſtumbling blocke at your
ſide, and ſtakes in your eies, til he take you away and deſtroy you from
this excellent Land, which he hath deliuered to you. \V Behold I this
day enter into the way of al fleſh, and you shal know with al your
minde that al the wordes, which our Lord promiſed that he would performe
to you, one is not eſcaped without effect.
%%% 0533
\V Therfore as he hath fulfilled in deede that which he promiſed, and al
thinges proſperous haue come: ſo wil he bring vpon you what euils ſoeuer
he hath threatened, til he take you away and deſtroy you from this
excellent Land, which he hath deliuered to you, \V becauſe you haue
tranſgreſſed the couenant of our Lord your God, which he hath made with
you, and haue ſerued ſtrange goddes, and adored them: quickly and in
haſt shal the furie of our Lord ryſe againſt you, and you shal be taken
away from this excellent Land, which he hath deliuered to you.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
