%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 0512
%%% o-0461
\startChapter[
  title={Chapter 12}
  ]
  
\Summary{Beſides Sehon and Og kinges of Heſebon and Baſan ſlaine by
  Moyſes, 7.~are reckened thirtie one kinges ſlaine by Ioſue.}

Theſe are the kinges, which the children of Iſrael ſtroke, and
poſſeſſed, their Land beyond Iordan toward the ryſing of the ſunne, from
the torrent Arnon vnto mount Hermon, and al the Eaſt part, that looketh
toward the wildernes. \V
\CNote{\XRef{Num.~21.}
\XRef{Deut.~3.}}
Sehon the king of the Amorrheites, which dwelt in Heſebon, had dominion from
Aroer, which is ſituated vpon the banke of the torrent Arnon, and of the
middel part in the valley and of halfe Galaad, as farre as the torrent
Iaboc, which is the border of the children of Ammon. \V And from the
wildernes vnto the ſea of Ceneroth againſt the Eaſt, and vnto the Sea of
the wildernes, which is the moſt ſalt ſea, on the Eaſt ſide by the way
that leadeth to Beſimoth: and on the South ſide, which lyeth vnder
Aſedoth, as farre as Phaſga. \V The border of Og the king of Baſan, of
the remnant of the Raphaims who dwelt in Aſtaroth, and in Edrai, and had
dominion in mount Hermon, and in Salecha, and in al Baſan, vnto the
borders \V of Geſſuri, and Machati, and of the halfe part of Galaad: the
borders of Sehon the king of Heſebon.
%%% 0513
\V Moyſes the ſeruant of our Lord, and the children of Iſrael ſtroke
them, and Moyſes deliuered their Land in poſſeſſion to the Rubenites,
and Gadites, and the half tribe of Manaſſes. \V Theſe are the Kinges of
the Land, whom Ioſue ſtroke and the children of Iſrael beyond Iordan on
the Weſt ſide, from Balaalgad in the field of Libanus, vnto the mount,
part wherof goeth vp into Seir: and Ioſue deliuered it in poſſeſſion to
the tribes of Iſrael, to euerie one their portions, \V as wel in the
mountaines as in the plaine and champaine countries. In Aſedoth, and in
the wildernes, and in the ſouth was the Hetheite and the Amorrheite, and
in the ſouth was the Hetheite and the Amorrheite, the
Chananeite, and the Pherezeite, the Heueite, and the Iebuſeite. \V The
King of Iericho one: the King of Hai, which is on the ſide of Bethel,
one: \V the King of Ieruſalem one, the King of Hebron one, \V the King
of Ierimoth one, the King of Lachis one, \V the King of Eglon one, the
King of Gazer one, \V the King of Dabir one, the King of Gader one,
\V the King of Herma one, the King of Hered one, \V the King of Lebna
one, the King of Adullam one, \V the King of Maceda one, the King of
Bethel one, \V the King of Taphua one, the King of Opher one, \V the
King of Aphec one, the King of Saron one, \V the King of Madonone, the
King of Aſor one, \V the King of Semeron one, the
King of Acſaph one, \V the King of Thenac one, the King of Mageddo one,
\V the King of Cades one, the King of Iachanan one,
%%% o-0462
\V the King of Carmel one, the King of Dor, and of the prouince of Dor
one, the King of the Nations of Galgal one, \V the King of Therſa one:
al the Kinges 
\SNote{Moyſes ſlew two kinges, & Ioſue thirtie one.}
thirtie one.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
