%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 0526
%%% o-0473
\startChapter[
  title={Chapter 20}
  ]

\Summary{Six cities of refuge for ſuch as commit caſual manſlaughter are
  named, 6.~in which remayning til the death of the high prieſt, they may
  then returne to their proper dwelling place, and be ſafe.}

And our Lord ſpake to Ioſue, ſaying: Speake to the children of Iſrael,
and ſay to them: \V Separate the cities of the fugitiues, of the which I
ſpake to you by the hand of Moyſes; \V that he may flee to them
whoſoeuer ſhal ſtrike a ſoule vnwitting: and may eſcape the wrath of
the nigh kinſeman, which is the reuenger of bloud: \V when he shal be
fled to one of theſe cities: he ſhal ſtand before the gate of the citie,
and ſhal ſpeake to the ancientes of that citie
\SNote{At firſt entering it ſufficed to allege in general his
innocencie, but after he muſt be tried in particular.}
thoſe thinges, that may proue him ſelfe innocent: and ſo they shal
receiue him, and geue him place to inhabite. \V And when the reuenger of
the bloud ſhal purſew him, they ſhal not deliuer him into his handes:
becauſe he ſtroke his neighbour by ignorance, neither is he proued to be
his enemie two or three dayes before. \V And he shal dwel in that citie,
til he ſtand before iudgement rendring a cauſe of his fact, and the high
prieſt die, which shal be at that time: then shal the manſlaer returne,
and enter into the citie and his houſe out of the which he had fled. \V
And they appointed Cedes in Galilee of the mount of Nepthali, and Sichem
in the mount of Ephraim, and Cariatharbee, the ſame is Hebron in the
mount of Iuda. \V And beyond Iordan againſt the Eaſt quarter of Iericho,
they appointed Boſor, which is ſituated in the champaine wildernes of
the tribe of Ruben, and Ramoth in Galaad of the tribe of Gad, and Gaulon
in Baſan of the tribe of Manaſſes. \V Theſe cities were appointed to al
the children of Iſrael, and to the
\Fix{ſtrangesrs,}{ſtrangers,}{obvious typo, fixed in other}
that dwelt among
%%% 0527
them: that he might flee to them which vnwittingly had ſtriken a ſoule,
and might not die in the hand of the kinſeman, coueting to reuenge the
bloud shed, vntil he might ſtand before the people to declare his cauſe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
