%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-22


\project douay-rheims


%%% 0529
%%% o-0476
\startChapter[
  title={Chapter 22}
  ]

\Summary{The
\MNote{The fourth part.

Two tribes & a halfe returne to their poſſeſſions; Ioſues godly
admonitions; & his, and Eleazars death.}
tribes of Ruben and Gad, and half Manaſses returne to their
poſſeſsions. 10.~VVho building an altar by the ſide of Iordan, the other
tribes ſuſpect that they wil make a ſchiſme, and therfore purpoſe to
fight againſt them. 13.~But firſt ſending an ambaſſage to admonish them,
21.~they anſwer that they made not an altar, for ſacrifice, but only for
a monument, that notwithſtãding they dwel on the other ſide of Iordan,
yet they are of the ſame people of God, 30.~wherwith al Iſrael is
ſatisfied.}

The ſame time Ioſue called the Rubenites, and Gadites, and the halfe
tribe of Manaſſes, \V and ſaid to them: You haue done al thinges that
Moyſes the ſeruant of our Lord cõmanded you: me alſo haue you obeyed in
al thinges, \V neither haue you left your brethren a long time, vntil
this preſent day, keeping the commandement of our Lord your God. \V
Therfore becauſe our Lord your God hath geuen your brethren quietnes and
peace, as he promiſed: returne, and goe into your tabernacles, and to
the land of your poſſeſſion, which Moyſes the ſeruant of our Lord
deliuered to you beyond Iordan: \V
\SNote{It perteineth to al magiſtrates, and other ſuperiors to admoniſh
their ſubiectes of their dutie towardes God, before al other thinges.}
alwayes ſo that you keep attentiuely, and in worke fulfil the
commandement, and the law which Moyſes the ſeruant of our Lord commanded
you, that you loue our Lord your God, and walke in al his wayes, and
obſerue al his commandementes, & cleaue to him, and ſerue him in al your
hart, and in al your ſoule. \V And Ioſue
\SNote{Al ſuperiors may impert bleſſing to their ſubiectes.}
bleſſed them, and diſmiſſed them. Who returned into their
tabernacles. \V And to the halfe tribe of Manaſſes Moyſes had geuen
poſſeſſion in Baſan: and therfore to the halfe that remayned, Ioſue gaue
a lotte among the reſt of their brethren beyond Iordan at the Weſt
ſide. And when he diſmiſſed them into their tabernacles, & had
bleſſed them, \V he ſaid to them:
%%% 0530
In much ſubſtance and riches returne to your ſeates, with ſiluer and
gold, braſſe and yron, and varietie of rayment: diuide the praye of your
enemies with your brethren. \V And the children of Ruben, and the
children of Gad, and the halfe tribe of Manaſſes returned, and went from
the children of Iſrael in Silo, which is ſituated in Chanaan, to enter
into Galaad the Land of their poſſeſſion, which they had obteyned
according to the commandement of our Lord in the hand of Moyſes. \V And
when they were come to the litle banckes of Iordan, into the Land of
Chanaan, they built beſide Iordan an altar of an infinite greatnes. \V
\SNote{So it behoueth al the ſeruantes of God to ſee in time, that no
ſchiſme be made. And therfore orderly to inquire of euerie ſhow of euil.
\XRef{1.~Theſ.~5.}}
Which thing when the children of Iſrael had heard, and certaine
meſſengers had reported to them that the children of Ruben, and Gad, and
the halfe tribe of Manaſſes had builded an altar in the Land of Chanaan,
vpon the litle banckes of Iordan, againſt the
%%% o-0477
children of Iſrael: \V
they aſſembled al in Silo, that they might goe vp, and fight againſt
them. \V And in the meane time they ſent to them into the Land of
Galaad, Phinees the ſonne of Eleazar the prieſt, \V and ten princes with
him, one of euerie tribe. \V Who came to the children of Ruben, and Gad,
and the halfe tribe of Manaſſes into the Land of Galaad, and ſaid to
them: \V This meſſage doth al the people of our Lord ſend to you: What
is this tranſgreſſion? Why haue you forſaken our Lord the God of Iſrael,
building a ſacrilegious altar, and reuolting from the worſhippe of
him? \V Is it a ſmale thing to you that
\SNote{Sinnes paſt are imputed to ſuch as fal againe as agrauating their
new ſinnes by reaſon of more ingratitude.}
you ſinned in Beelphegor, and vntil this preſent day the ſpotte of this
abomination abideth in vs? and many of the people fel dead. \V And you
to day haue forſaken our Lord, and to morrow his wrath wil rage
\SNote{As before
\XRef{chap.~7.}
ſo if theſe had bene culpable al Iſrael might feare to be puniſhed,
except iuſtice were donne vpon the offenders.}
againſt al Iſrael. \V But if you thinke the land of your poſſeſſion to
be vncleane, paſſe to the Land, wherein is the tabernacle of our Lord,
and dwel among vs: only that you depart not from our Lord, and from our
companie, an altar being built beſide the altar of our Lord God. \V Did
not
\CNote{\XRef{chap.~7.}}
Achan the ſonne of Zare tranſgreſs the commandement of our Lord, and his
wrath lay vpon al the people of Iſrael? And he was one man, and I would
he alone had periſhed in his wicked fact. \V And the children of Ruben,
and Gad, and of the halfe tribe of Manaſſes anſwered the princes of the
legacie of Iſrael: \V The moſt mightie God our Lord, The moſt mightie
God our Lord, him ſelfe knoweth, and Iſrael together shal vnderſtand: If
%%% 0531
with the mind of preuarication we haue erected this altar, let him not
keepe vs, but puniſh vs preſently: \V and if we did it with that minde,
that we might lay vpon it holocauſtes, and ſacrifice, and pacifique
victimes, let him ſelfe examine it and iudge: \V and not rather with
that meaning and deliberation, that we ſaid: To morrow your children wil
ſay to our children: What haue you to doe with our Lord the God of
Iſrael? \V Our Lord hath put a border betwen vs and you, O ye children
of Ruben, and children of Gad, the riuer Iordan: and therfore you haue
no part in our Lord. And by this occaſion your children ſhal auert our
children from the feare of our Lord. We therfore thought it better, \V
and ſaid: Let vs build vs an altar,
\SNote{In the law of Moyſes was one onlie altar for ſacrifice, for the
whole people of God, to auoide ſchiſme and idolatrie. Now in the Church
(being in al nations) are manie altares, & but one onlie Sacrifice
prefigured by al the former, as S.~Auguſtin teacheth
\Cite{li.~17. c.~20. de ciuit.}
\Cite{S.~Leo Ser.~8. de Paſsione.}}
not for holocauſtes, nor to offer victimes, \V but for a teſtimonie
betwen vs and you, and our iſſue and your progenie, that we may ſerue
our Lord, and it may be our right to offer both holocauſtes, and
victimes, and pacifique hoſtes: and that your children to morrow ſay not
to our children: You haue no part in our Lord. \V And if they wil ſay
ſo, they ſhal anſwer them: Behold the altar of our Lord, which our
fathers made, not for holocauſtes, nor for Sacrifice, but for our
teſtimonie and yours. \V God ſaue vs from this abomination that we
should reuolt from our Lord, and leaue his ſteppes, erecting an altar to
offer holocauſtes, and ſacrifices, and victimes, beſide the altar of our
Lord God, which is erected before his tabernacle. \V Which thinges being
heard, Phinees the prieſt, and the princes of the legacie, which were
with him, were pacified: and they
%%% o-0478
admitted moſt willingly the wordes of the children of Ruben, & Gad, and
of the halfe tribe of Manaſſes. \V And Phinees the prieſt the ſonne of
Eleazar ſaid to them: Now we know that our Lord is with vs, becauſe you
are not culpable of this preuarication, and haue deliuered the children
of Iſrael from the hand of our Lord. \V And her returned with the
princes from the children of Ruben and Gad, out of the Land of Galaad,
into the Land of Chanaan, to the children of Iſrael, and reported to
them. \V And the ſaying pleaſed al that heard it. And the children of
Iſrael prayſed God, and they did no more ſay, that they would goe vp
againſt them, and fight, and deſtroy the Land of their poſſeſſion. \V
And the children of Ruben, & the children of Gad called the altar, which
they had built, Our teſtimonie, that our Lord he is God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
