%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 0505
%%% o-0455
\startChapter[
  title={Chapter 9}
  ]
  
\Summary{Other nations fearing Iſrael ioyne their forces to fight
  againſt them, 3.~but the Gabaonites ſend embaſſadors, feyning craftily
  to come farre of, 14.~with whom Ioſue and the ancientes of Iſrael make
  league binding it by oath. 16.~Within three dayes they are detected to
  be nere inhabitantes. VVherupon the people murmur againſt their
  princes. VVho yet for their oath, let the Gabaonites liue, 20.~only
  making them and their progenie perpetual ſeruantes, to cutte wood, and
  carie water.}

VVhich thinges being heard, al the kinges beyond Iordan, that dwelt by
the ſea ſide and the ſhore of the great ſea, they alſo that dwelt beſide
Libanus, the Hetheite and Amorreite, the Chananeite, the Pherezeite, and
%%% 0506
the Heueite, and the Iebuſeite, \V were gathered, to fight againſt Ioſue
and Iſrael with one minde, and one ſentence. \V But they that dwelt in
Gabaon, hearing al thinges that Ioſue had done to Iericho and Hai: \V
ſubtelly deuiſing tooke prouiſion for them ſelues, laying old ſackes
vpon their aſſes, and bottels of wine rent and ſowed againe, \V and
ſhoes very old which for a ſhew of oldeneſſe were clouted with ſpeckes,
putting on them old garmentes: the loaues alſo, which they caried for
prouiſion by the way, were heard, and broken into peeces: \V and they
went on to Ioſue, who then abode in the campe at Galgal, and ſaid to
him, and withal to al Iſrael: We are come from a farre countrie,
deſirous to make peace with you. And the children of Iſrael anſwered
them, and ſaid: \V Leſt perhaps you dwel in the Land which is dew to vs
by lotte, and we can not enter a league with you. \V But they ſaid to
Ioſue: We are thy ſeruantes. To whom Ioſue ſaid: Who are you? and whence
came you? \V They anſwered: From a very farre countrie are thy ſeruantes
come in the name of the Lord thy God. For we haue heard the fame of
his might, al thinges that he did in Ægypt, \V and to the two kinges of
the Amorrheites that were beyond Iordan, Sehon the king of Heſebon, and
Og the king of Baſan, that were in Aſtaroth:
%%% o-0456
\V and our ancientes, and
al the inhabitantes of our Land ſaid to vs: Take in your handes victuals
for a very long way, and goe meete them, and ſay: We are your ſeruantes,
enter a league with vs. \V Behold, theſe loaues we tooke hotte, when we
departed from our houſes to come to you, now they are become drie, and
broken for ouer much oldeneſſe. \V The bottels of wine we filled being
new, now they are burſt & diſſolued. The garmentes and ſhoes that we
haue vpon vs, and which we haue on our feete, for the length of the long
way are worne, and almoſt conſumed. \V They tooke therfore of their
victuals, and
\SNote{In ſuch a caſe they ought to haue conſulted God, to witte, the
high Prieſt, putting on the Ephod & Rationale, ſhould haue prayed at the
dore of the tabernacle, where our Lord appointed to ſpeake vnto him.
\XRef{Exo.~19. v.~42.}
It was alſo ordained
\XRef{Num.~17. v.~21.}
that Eleazar ſhould cõſult our Lord, when anie thing was to be donne by
Ioſue, which omittĩg to do they were deceiued by theſe Gabaonites.}
asked not the mouth of our Lord. \V And Ioſue made peace with them, and
entring a league promiſed that they ſhould not be ſlaine: the princes
alſo of the multitude ſware to them. \V But three daies after that the
league was made, they heard that they dwelt nigh, and they ſhould be
among them. \V And the children of Iſrael remoued the campe, and came
into their cities the third day, whoſe names are theſe, Gabaon, and
Chaphira, and Beroth, and Chariathiarim. \V And they ſtroke them not,
becauſe
%%% 0507
the princes of the multitude had ſworne in the name of our Lord the God
of Iſrael. Therfore al the common people murmured againſt the princes.
\V Who anſwered them: We haue ſworne to them in the name of our Lord the
God of Iſrael, and therfore we may not touch them. \V But this we wil
doe to them: Let them be reſerued in dede aliue, leſt the wrath of God
be ſtirred againſt vs, if we shal be forſworne: \V but ſo let them liue,
that for the vſes of the whole multitude they hew wood, and carie in
water. Who ſpeaking theſe thinges, \V Ioſue called the Gabaonites, and
ſaid to them: Why would you deceiue vs by fraude to ſay: We dwel very
farre of from you, wheras you are in the middes of vs? \V Therfore you
shal be vnder a curſe, and they shal not faile of your ſtocke a hewer of
wood, and a carier of water into the houſe of my God. \V Who anſwered:
It was told vs thy ſeruantes, that the Lord thy God had promiſed Moyſes
his ſeruant, that he would deliuer you al the Land, and would deſtroy al
the inhabitantes therof. Therfore we feared excedingly and prouided for
our liues, cõpelled by your terrour, and we tooke this counſel. \V And
now we are in thy hand: that which ſeemeth vnto thee good and right, doe
to vs. \V Ioſue therfore did as he had ſaid, and deliuered them from the
hand of the children of Iſrael, that they should not be ſlaine. \V And
he decreed in that day, that
\SNote{In theſe Gabaonites of Chanaans race was fulfilled Noes prophecie
\XRef{Gen.~9.}
that \Emph{Chanaan should be a ſeruant to his brethren}.}
they should be in the miniſterie of al the people, and of the altar of
our Lord, hewing wood, and carying water, vntil this preſent time, in
the place which our Lord hath choſen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
