%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 0503
%%% o-0453
\startChapter[
  title={Chapter 8}
  ]

\Summary{By ſtratageme of an ambushment the citie of Hai is taken and
  burned, and al the inhabitantes ſlaine. 29.~The king hanged. 30.~An
  Altar built, Sacrifice offered, 32.~the law written in ſtones, the
  people bleſſed, and the bleſſinges and curſinges read before them al.}

And our Lord ſaid to Ioſue: Feare not, neither doe thou dread: take with
thee al the multitude of fighting men, and ryſing goe vp vnto the towne
of Hai. Behold I haue deliuered into thy hand the King therof, & the
people, and the citie and the land. \V And thou shalt do to the citie of
Hai, and to the King therof, as thou haſt done to Iericho, and to the
King therof: but the praye and al the cattel you shal ſpoyle for your
ſelues:
\SNote{Deceiptes & ſtratagemes are lawful in iuſt warre; but not
falſhood, nor breach of promiſe.
\Cite{S.~Aug. q.~10. in Ioſue.}}
lay ambushmentes to the citie behind it. \V And Ioſue aroſe, and al the
hoſte of the men of warre with him, to goe vp into Hai: and thirtie
thouſand choſen valiant men he ſent in the night, \V and commanded them,
ſaying: Lay ambushmentes behinde the citie: neither retyre you farre of:
and you shal al be readie. \V But I and the reſt of the multitude, which
is with me, wil goe vp on the contrarie ſide againſt the citie. And when
they shal iſſue out againſt vs, as we did before, we wil flee, and turne
our backes: \V til purſewing they be drawen forward farre from the
citie: for they wil thinck that we flee as before. \V We therfore
fleeing, and they purſewing, you shal riſe out of the ambushmentes, and
shal waiſt the citie: and our Lord your God wil deliuer it into your
handes. \V And when you shal take it, burne it, and you shal doe al
thinges ſo, as I haue commanded. \V And he diſmiſſed them away, and they
went on to the place of the ambushment, and ſate betwen Bethel and Hai,
at the Weſt ſide of the citie of Hai. But Ioſue that night ſtayed in the
middes of the people, \V and ryſing early he muſtered his ſoldiars, and
went vp with the ancientes in the fronte of the hoſte, enuironed with
the ayde of the fighting men. \V And when they were come, and were gone
vp directly againſt the citie, they ſtoode on the North ſide of the
citie, betwen which and them was a valley in the middes. \V And
\SNote{Theſe fiue thouſãd were of the thirtie thouſand, which were firſt
ſent.
\XRef{v.~3.}
The other 25000. ioyned with Ioſues troupe, & entered into the citie.}
fiue thouſand men had he choſen, and ſette in the
%%% 0504
embushmentes betwen Bethel and Hai on the Weſt ſide of the ſame
citie: \V but al the reſt of the hoſte went in battel aray on the North
ſide, ſo that the laſt of the multitude did reach to the Weſt ſide of
the citie. Ioſue therfore went that night, and ſtoode in the middes of
the valley. \V Which when the King of Hai had ſeene, he made haſt in the
mourning, and iſſued forth with al the hoſt of the citie, and bent his
armie toward
%%% o-0454
the deſert, being ignorant that there lay embushmentes ſecretly behind
his backe. \V But Ioſue, and al Iſrael gaue backe, feyning feare, and
fleeing by the way of the wildernes. \V But they cried alowde together,
and encouraging one an other, purſewed them. And when they were gone
from the citie, \V and
\SNote{Not one fitte to beare armes was left.}
not one remained in the citie of Hai and Bether, that purſewed not
Iſrael (euen as they had rushed out leauing the townes open,) \V our
Lord ſaid to Ioſue:
\SNote{He lift his ſhield vpon a long pike or lance, that it might be
ſene farre of.}
Lift vp the shield, that is in thy hand, againſt the citie of Hai, for I
wil deliuer it to thee. \V And when he had lifted vp his shield againſt
the citie, the embushmentes, that lay hidde, roſe vp immediatly: and
going to the citie, tooke and burnt it. \V And the men of the citie,
that purſewed Ioſue, looking backe and ſeeing the ſmoke of the citie
riſe vp euen to heauen, they could no more flee hither and thither:
eſpecially whereas they, that had feyned running away, and went toward
the wildernes, moſt valiantly reſiſted againſt the purſewers. \V And
Ioſue and al Iſrael ſeeing that the citie was taken, and the ſmoke of
the citie roſe vp, returning he ſtroke the men of Hai. \V For they alſo
that had taken and burnt the citie, iſſuing out of the citie againſt
their owne men, beganne to ſtrike the enemies in the middes of
them. When the aduerſaries therfore were ſlaine on both ſides, ſo that
none of ſo great a multitude was ſaued, \V they tooke the King of the
citie of Hai aliue, and preſented him to Ioſue. \V Therfore al being
ſlaine, that had purſewed Iſrael fleeing to the deſertes, and falling by
the ſword in the ſame place, the children of Iſrael returning ſtroke the
citie. \V And there were that fel that ſame day from man vnto woman,
twelue thouſand men, al of the citie of Hai. \V But Ioſue plucked not in
his hand, which he had
\Fix{ſtrethched}{ſtretched}{obvious typo, fixed in other}
forth on high, holding the ſhield til
al the inhabitantes of Hai were ſlaine. \V And the cattel and the praye
of the citie the children of Iſrael diuided among them, as our Lord had
commanded
%%% 0505
Ioſue. \V Who burnt the citie, and made it a heape for euer: \V the king
alſo therof he hong on a gibbet vntil euening and the going downe of the
ſunne. And he commanded, and they tooke downe his corps from the gibbet:
and threw it in the verie entring of the citie, heaping vpon it a great
heape of ſtones, which remaineth vntil this preſent day. \V Then Ioſue
built an altar to our Lord the God of Iſrael in mount Hebal, \V as
Moyſes the ſeruant of our Lord had commanded the children of Iſrael, and
it is written in the volume of the law of Moyſes: an Altar of vnhewed
ſtones which yron hath not touched; and he offered vpon it holocauſtes
to our Lord, and immolated pacifique victimes. \V And he wrote vpon
ſtones the Deuteronomie of the law of Moyſes, which he had ordered
before the children of Iſrael. \V And al the people, and the ancientes,
and the princes and iudges ſtoode on both ſides of the arke, in the
ſight of the prieſtes that caried the arke of the couenant of our Lord,
as wel the ſtranger as alſo the man of the ſame countrie. The half part
of them beſide mount Garizim, and halfe beſide mount Hebal,
\CNote{\Cite{Deut.~27.}}
as Moyſes the ſeruant of our Lord had commanded.
%%% o-0455
And firſt in deede he
\SNote{Al ſuperiors may bleſſe their ſubiectes, Princes their peoples and
parentes their children.}
bleſſed the people of Iſrael. \V After this he reade al the wordes of
the bleſſing & the curſing, and al thinges that were written in the
volume of the law. \V Nothing of thoſe thinges, which Moyſes had
commanded, did he leaue vntouched, but he repeated al thinges before al
the multitude of Iſrael, the wemen and children and ſtrangers, that
dwelt among them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
