%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 0490
%%% o-0441
\startChapter[
  title={Chapter 1}
  ]

\Summary{Ioſue
\MNote{The firſt part.

Of the paſſage of Iſrael ouer Iordan.}
encoreged by our Lord, 10.~admonisheth the people to prepare themſelues
to paſſe ouer Iordan; 12.~and al the able men of the tribes of Ruben,
Gad, and halfe Manaſſes to march armed before the reſt. 16.~Al promiſe
to doe whatſoeuer he commandeth.}

And it came to paſſe after the death of Moyſes the ſeruant of our Lord,
that our Lord ſpake to Ioſue the ſonne of Nun, the miniſter of Moyſes,
and ſaid to him: \V Moyſes my ſeruant is deade: ariſe, and paſſe ouer
this Iordan thou and al the people with thee, into the Land, which I wil
geue to the children of Iſrael. \V Euerie place, the ſteppe of your
foote ſhal treade, wil I deliuer to you, as I haue ſpoken to Moyſes. \V
From the deſert and Libanus vnto the great riuer Euphrates, al the land
of the Hetheites vnto the great ſea againſt the going downe of the
ſunne, shal be your border. \V No man shal be able to reſiſt you al the
daies of thy life: as I haue beene with Moyſes, ſo wil I be with thee: I
wil not leaue, nor forſake thee. \V Take courage, and be ſtrong: for
thou shalt by lotte diuide to this people the Land, for the which I
ſware to their fathers, that I would deliuer it to them. \V Take courage
therfore, and be very ſtrong: that thou keepe and doe al the Law, which
Moyſes my ſeruant hath commanded thee: decline not from it to the right
hand or to the left, that thou mayeſt vnderſtand al thinges which thou
doeſt. \V Let
%%% o-0442
not the volume of this law depart from thy mouth: but thou shalt
meditate in it daies and nightes, that
%%% 0491
thou maieſt keepe and doe al thinges that be written in it: then shalt
thou direct thy way, and vnderſtand it. \V Behold I command thee, take
courage, and be ſtrong. Feare not, and dreade not: becauſe the Lord thy
God is with thee in al the thinges to whatſoeuer thou shalt goe. \V And
Ioſue commanded the princes of the people, ſaying: Paſſe through the
middes of the campe, and command the people, and ſay: \V Prepare for
your ſelues
\SNote{Beſides Manna, which yet ceaſed not, they might if they would
prouide other meate: prefiguring that in the primitiue Church, it ſhould
be lawful to vſe legal ceremonies, with euangelical rites for a time,
til the old law were buried with honour.}
victuals: for after the third day you shal paſſe ouer Iordan, and shal
enter to poſſeſſe the Land, which our Lord your God wil geue you. \V To
the Rubenites alſo and Gaddites, and halfe tribe of Manaſſes he ſaid:
\V Remember the word, which Moyſes the ſeruant of our Lord commanded
you, ſaying: Our Lord your God hath geuen you reſt, and al this Land.
\V Your wiues, and children, and cattel shal tarie in the Land, which
Moyſes deliuered to you beyond Iordan: but paſſe you ouer armed before
your brethren, al that are ſtrong of hand, & fight for them, \V vntil
our Lord geue reſt to your brethren as to you alſo he hath geuen, and
they alſo poſſeſſe the Land which our Lord your God wil geue them: and
ſo returne into the Land of your poſſeſſion, and you shal dwel in it,
which Moyſes the ſeruant of our Lord gaue you beyond Iordan, againſt the
ryſing of the ſunne. \V And they made anſwer to Ioſue, and ſaid: Al
thinges, that thou haſt commanded vs, we wil doe: and whither ſoeuer
thou shalt ſend vs, we wil goe. \V As we obeyed Moyſes in al thinges, ſo
wil we obey thee alſo: only be our Lord thy God with thee, as he was
with Moyſes. \V He that shal gaineſay thy mouth, and not obey al thy
wordes, that thou shalt command him, let him die. Thou only take
courage, and doe manfully.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
