%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-05


\project douay-rheims


%%% 0497
%%% o-0448
\startChapter[
  title={Chapter 5}
  ]

\Summary{The kinges of Chanaan are ſore frighted with the newes of
  Iſraels paſsage ouer Iordan. 2.~Circumciſion is againe commanded, and
  obſerued, which had bene ommitted in the deſert fourtie
  yeares. 10.~They make their Paſch. 12.~Manna ceaſeth. 13.~And an Angel
  appeareth to Ioſue.}

Therfore after that al the kinges of the Ammorrheites, which dwelt
beyond Iordan at the weſt ſide, and al the kinges of Chanaan, which
poſſeſſed the places nigh to the great ſea, had heard that our Lord had
dried the ſtreames of Iordan before the children of Iſrael, til they
paſſed ouer, their hart failed, and there remained no ſpirit in them,
fearing the entring of the children of Iſrael. \V At that time our Lord
ſaid
\SNote{See
\XRef{Annotations c.~3. v.~8.}}
to Ioſue: Make thee kniues of ſtone, and circumciſe
\SNote{Circũciſion had bene omitted fourtie yeares, whiles they were in
the deſerte, alwayes vncertaine when to march, & ſo it is now commanded
the ſecond time.}
the ſecond time the children of Iſrael. \V He did that which our Lord
had commanded, and he circumciſed the children of Iſrael in the hil of
the prepuces. \V And this is the cauſe of the ſecond circumciſion: Al the
people, that
%%% 0498
came out of Ægypt of the malekinde, al the fighting men, died in the
deſert by the long circuits of the way, \V who were al circumciſed. But
the people that was borne in the deſert, \V during the fourtie yeares of
the iourney in the wide wildernes was incircumciſed: til they were
conſumed that had not heard the voice of our Lord, and to whom he had
ſworne before, that he would not shew them a land flowing with milke and
honie. \V The children of theſe ſucceded in the place of the fathers,
and were circumciſed of Ioſue: for they were in the prepuce euen as they
were borne, neither had any circumciſed them in the way. \V And after
that they were al circumciſed, they abode in the ſame place of the
campe, vntil they were whole. \V And our Lord ſaid to Ioſue: This day
haue I taken away the reproch of Ægypt from you. And the name of that
place was called Galgal, vntil this preſent day. \V And the children of
Iſrael abode in Galgal, & they made the Phaſe, the fourtenth day of the
moneth at euen in the champion of Iericho: \V and they did eate of the
corne of the Land the next day, azyme loaues and
\TNote{\Fix{\L{Furmentis.}}{\L{Frumentis.}}{obvious typo, same in both}}
polẽt of the ſame yeare. \V And Manna failed after they did eate of the
corne of the Land, neither did the children of Iſrael vſe that meate any
more, but they did eate of the corne of the preſent Land of Chanaan. \V
And when Ioſue was in the field of the citie of Iericho, he lifted vp
his eies, and ſaw a man ſtanding againſt him, holding a drawen ſword,
and he went to him, and ſaid: Art thou ours, or our aduerſaries? \V Who
anſwered: No: but I am a
\SNote{Not God, but of Gods hoſte.}
prince of the hoſt of our Lord, and now I come. \V Ioſue fel
%%% o-0449
flatte on the ground. And
\LNote{Adoring.}{Ioſue
\MNote{Religious honour due to Angels.}
knowing that the perſon which appeared, was an Angel and not God, nor a
man, neither adored him with godlie honour, for that had bene idolatrie,
nor with ciuil, for that perteineth to
\Fix{wordlie}{worldlie}{obvious typo, same in both}
and temporal excellencie, and is not competent to ſacred thinges,
eſpecially to immortal and glorious ſpirites;
\CNote{See
\XRef{Annot. Exo.~20.}}
and therfore the honour he did to this Angel, was religious honour
infinitly inferior to diuine, and yet much greater then ciuil.}
adoring he ſaid: What ſpeaketh my Lord to his ſeruant? \V
\LNote{Looſe thy shoe.}{The
\MNote{Holie places.}
Angel did not only accept of the honour donne vnto him, but alſo
required more, ſhewing that the verie place was holie for his preſence,
being otherwiſe the common field of Iericho.}
Looſe, ſayth he, thy ſhoe from thy feete: for the place wherein thou
doſt ſtand, is holie. And Ioſue did as it was commanded him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
