%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 0496
%%% o-0446
\startChapter[
  title={Chapter 4}
  ]

\Summary{In memorie of their miraculous paſſage, twelue chief men, of
  the twelue tribes, take ſo manie great ſtones from the middes of
  Iordan, 9.~and put other twelue, where the prieſtes ſtood with the
  arke. 18.~The waters returne to their former courſe. And the twelue
  ſtones are erected for a monument.}

VVho being paſſed ouer, our Lord ſaid to Ioſue: \V Chooſe twelue men one
in euerie tribe: \V and command them that they take vp out of the middes
of the chanel of Iordã, where the feete of the prieſtes ſtoode, twelue
moſt hard ſtones, which you ſhal put in the place of the campe, where
you ſhal pitch tentes this night. \V And Ioſue called twelue men, whom
he had choſen out of the children of Iſrael, one of euerie tribe, \V and
he ſaid to them: Goe before the arke of our Lord your God to the middes
of Iordan, and carrie from thence euerie man a ſtone on your ſhoulders,
according to the number of the children of Iſrael, \V that it may be
\SNote{Is not the forme of a croſſe as conueniẽt a ſigne, to put
chriſtians in mind, how our Sauiour redemed vs, as theſe ſtones were to
the Iewes, how God brought their fathers ouer Iordan?}
a ſigne among you: and when your children ſhal aske you to morrow,
ſaying: What meane theſe ſtones? \V You ſhal anſwer
%%% o-0447
them: The waters of Iordan decayed before the arke of the couenant of
our Lord, when it paſſed ouer the ſame: therfore were theſe ſtones ſette
for a monument of the children of Iſrael for euer. \V The children of
Iſrael therfore did as Ioſue commanded them, carying out of the chanel
of Iordan twelue ſtones, as our Lord had commanded him, according to the
number of the children of Iſrael, vnto the place, wherein they camped,
and there they ſette them. \V Other twelue ſtones alſo Ioſue put in the
middes of the chanel of Iordan, where the prieſtes ſtoode, that caried
the arke of the couenant: and they be there vntil this preſent day. \V
But the prieſtes that caried the arke, ſtoode in the middes of Iordan,
til al thinges were accompliſhed, which our Lord had commanded Ioſue, to
ſpeake to the people, and Moyſes had ſaid to him. And the people made
haſt, and paſſed ouer. \V And when they had al paſſed ouer, the arke
alſo of our Lord paſſed ouer, the prieſtes alſo went before the people.
\V The children of Ruben alſo and Gad, and the half tribe of Manaſſes,
went armed before the children of Iſrael, as Moyſes had commanded them:
\V and fourtie thouſand fighting men by troupes, and bandes,
%%% 0497
marched through the plaine and champion countrie of the citie of
Iericho. \V In that day our Lord magnified Ioſue before al Iſrael, that
they ſhould feare him, as they had feared Moyſes, whiles he yet
liued. \V And he ſaid to him: \V Command the prieſtes, that carie the
arke of the couenant, that they comme vp out of Iordan. \V Who commanded
them, ſaying: Come ye vp out of Iordan. \V And when they that caried the
arke of the couenant of our Lord, were come vp, and begane to treade on
the drie ground, the waters returned into their chanel, and ranne as
they were wont before. \V And the people came vp out of Iordan, the
tenth day of the firſt moneth, and camped in Galgal againſt the Eaſt
ſide of the citie of Iericho. \V The twelue ſtones alſo, which they had
taken out of the chanel of Iordan, Ioſue ſette in Galgal, \V and ſaid to
the children of Iſrael: When your children shal aske their fathers to
morrow, and ſhal ſay to them: What meane theſe ſtones: \V You ſhal teach
them, and ſay: By the drie chanel did Iſrael paſſe ouer this Iordan, \V
your Lord God drying the waters therof in your ſight, vntil you paſſed
ouer: \V as he had done before in the readſea, which he dried til we
paſſed through: \V that al the people of the earth may learne the moſt
ſtrong hand of our Lord, that you alſo may feare our Lord your God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
