%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 0510
%%% o-0459
\startChapter[
  title={Chapter 11}
  ]
  
\Summary{Iabin a more principal king ſomoneth other kinges to ioyne with
  him againſt Iſrael. 6.~Ioſue animated with Gods promiſe of victorie,
  7.~ouerthroweth them al. 16.~Subdueth their countries: 21.~killeth
  alſo manie of the giantes ſtocke.}

VVhich thinges when Iabin the king of Aſor had heard, he ſent to Iobab
the king of Madon, and to the king of Semeron, and to the king of
Achſaph: \V to the kinges alſo of the North, that dwelt in the
mountaines
%%% 0511
and in the plaine againſt the ſouth ſide of Ceneroth, in the champaine
alſo and countries of Dor by the ſea ſide: \V the Chananeite alſo on the
Eaſt and Weſt, and the Amorrheite and Hetheite and Pherezeite &
Iebuſeite in the mountaines: the Heueite alſo which dwelt at the foote
of Hermon in the Land of Maſpha. \V And they iſſued forth al with their
troupes, a people exceeding manie as the ſand, that is in the ſhore of
the ſea, their horſes alſo and chariottes of paſſing great
multitude. \V And al theſe kinges aſſembled together
%%% o-0460
in one at the Waters of Merom, to fight againſt Iſrael. \V And our Lord
ſaid to Ioſue: Feare them not: for to morrow this ſelfe ſame houre wil I
deliuer al theſe to be wounded in the ſight of Iſrael: their horſes
thou ſhalt hoghſinew, and their chariottes thou ſhalt burne with
fire. \V And Ioſue came, and al the hoſte with him againſt them to the
Waters of Merom ſodenly, and ranne in vpon them, \V and our Lord
deliuered them into the handes of Iſrael. Who ſtroke them, and purſewed
them as farre as great Sidon, and the Waters of Maſerephoth, and the
field of Maſphe, which is on the Eaſt ſide therof. Therfore he ſtroke al,
ſo that he left no reliques of them: \V and he did as our Lord had
commanded him, their horſes he hoghſinewed, and their chariottes he
burnt. \V And returning immediatly he tooke Aſor: and the king therof he
ſtroke with the ſword. For Aſor in old time among al theſe kingdomes held
the principalitie. \V And he ſtroke al the ſoules that abode there: he
left not in it anie remaines, but to vtter deſtruction he
waſted al thinges, and the citie it ſelfe he deſtroyed with fyre. \V And
al the cities round about, their kinges alſo he tooke, ſtroke &
deſtroyed, as Moyſes the ſeruant of God had commanded him. \V Except the
cities, that were ſituated on hilles and higher ground, the reſt Iſrael
burnt: one onlie Aſor verie wel fenſed he conſumed with fyre. \V And al
the praye of theſe cities and the cattel the children of Iſrael diuided
among them ſelues, al the men being ſlaine. \V As our Lord had commanded
Moyſes his ſeruant, ſo did Moyſes command Ioſue, and he accompliſhed al
thinges: he ommitted not of al the commandementes, not ſo much as one
word, which our Lord had commanded Moyſes. \V Ioſue therfore tooke al
the hillie countrie, and ſouth, and the land of Goſen, and plaine, and
the Weſt quarter, and the mountaine of Iſrael, and the champaine
%%% 0512
countrie therof: \V and the part of the mountaine, that goeth vp to Seir
as farre as Baalgad by the plaine of Libanus vnder mount Hermon: al
their kinges he tooke, ſtroke, and ſlew. \V
\SNote{Theſe warres cõtinued nere ſeuen yeares, as appeareth
\XRef{chap.~14. v.~10.}}
A great time did Ioſue fight againſt theſe kinges. \V There was not a
citie that did deliuer it ſelfe to the children of Iſrael, except the
Heueite, which dwelt in Gabaon: for he tooke al by fight. \V For it was
\SNote{For their enormious ſinnes God left them in their owne reprobate
ſenſe, not impoſing neceſſitie of ſinning, but permitting them to
indurate their owne hartes. See
\XRef{Annot. Exod.~7.}}
the ſentence of our Lord, that their harts ſhould be indurate, and they
should fight againſt Iſrael, and fall, and should not deſerue anie
clemencie, and should perish, as our Lord had commanded Moyſes. \V At
that time Ioſue came, and ſlew the Enacimes of the mountaines, of
Hebron, and Dabir, and Anab, and from al the mountaine of Iuda and
Iſrael, and deſtroyed their cities. \V He leift not any of the ſtocke of
Enacimes, in the Land of the children of Iſrael: ſauing the cities of
Gaza, and Geth, and Azotus, in the which onlie they were leift. \V Ioſue
therfore tooke al the Land, as our Lord ſpake to Moyſes, and deliuered
it in poſſeſſion to the children of Iſrael, according to their partes
and tribes. And
\SNote{Hence forth they had not general warres but diuers tribes had
particular as appeareth
\XRef{chap.~15. &c.}}
the Land reſted from battels.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
