%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 0517
%%% o-0465
\startChapter[
  title={Chapter 15}
  ]

\Summary{The borders of the lotte of Iuda, 13.~including Calebs
  particular inheritance (16.~out of which he geueth Cariath Sepher, and
  his daughter to Othoniel, for winning it:) 21.~with the names of the
  cities therof. 63.~The Iebuſeite yet dwelling with Iuda in
  Hieruſalem.}

Therfore the lotte of the children of Iudas by their kinredes was this:
From the border of Edom, vnto the deſert of Sin againſt the South, and
vnto the vttermoſt part of the ſouth coaſte. \V The beginning therof was
from the toppe of the moſt ſalt ſea, and from the brinke therof, that
looketh to the South. \V And it goeth forth againſt the Aſcent of the
Scorpion, and paſſeth through into Sina: and ryſeth vp into Cadeſbarne,
& reacheth into Eſron, aſcending to Addar, and cõpaſſing Carcaa, \V and
thence paſſing through into Aſemona, and reaching to the Torrent of
Ægypt: and the borders therof shal be the great ſea. This ſhal be the
end of the ſouth coaſt. \V But on the Eaſt ſide the beginning ſhal be
the moſt ſalt ſea vnto the vtmoſt partes of Iordan: and thoſe places
that looke to the North from the brinke of the ſea vnto the ſame riuer
of Iordan. \V And the border goeth vp into Beth hagla, and paſſeth from
the north into Beth Araba: aſcending to the ſtone of Bohen the ſonne of
Ruben. \V And reaching as farre as the borders of Debera from the Valley
of Achor, againſt the North looking toward Galgal, which is oppoſite to
the Aſcent of Adommim, on the ſouth ſide of the torrent: and paſſeth the
waters, that are called The Fountayne of the ſunne: and the endes therof
ſhal be to the Fountayne rogel. \V And it aſcendeth by the valley of the
ſunne of Ennom on the ſide of the Iebuſeite toward the South, this is
Ieruſalem: and thence rearing it ſelf to the toppe of the mountayne,
which is againſt Geennom toward the Weſt in the toppe of the Valley of
Raphaim againſt the North. \V And it paſſeth through from the toppe of
the mountaine to the fountaine of the water Nephtoa: and reacheth to the
townes of mount Ephron: and bendeth into Baala, which is Cariathiarim,
that is to ſay, a citie of wooddes. \V And it compaſſeth from Baala
againſt the Weſt, vnto mount Seir: and paſſeth by the ſide of mount
Iarim toward the North into Cheſlon: and goeth downe into Bethſames, and
paſſeth into Thamna. \V And it reacheth
%%% 0518
toward the North coaſt of a part of Accaron at the ſide: and bendeth to
Sechrona, and paſſeth mount Baala: and cometh into
%%% o-0466
Iebneel, and is ſhut vp with the end of the great ſea toward the
Weſt. \V Theſe are the borders of the children of Iudas in circuite in
their kinredes. \V But to Caleb the ſonne of Iephone he gaue a portion
in the middes of the children of Iudas, as our Lord had commanded him:
Cariath Arbe the father of Enac, that is Hebron. \V And Caleb deſtroyed
out of it the three ſonnes of Enac, Seſai and Ahiman & Tholmai of the
ſtocke of Enac. \V And from thence going vp he came to the inhabitantes
of Dabir, which before was called Cariath ſepher, that is to ſay, a
citie of letters. \V And Caleb ſaid: He that ſhal ſtrike Cariath-ſepher,
and take it, I wil geue him Axa my daughter to wife. \V And Othoniel the
ſonne of Cenez, the yonger brother of Caleb tooke it: and he gaue him
\SNote{Though it was prohibited that the nephew ſhould marie his aunt,
yet the vncle was not expreſly forbid to marie his neece by Moyſes law.
\XRef{Leuit.~18.}
And albeit there is the ſame degree of conſanguinitie, yet not the ſame
incõuenience, by reaſon the ſame perſon remaineth ſubiect, that was
inferior before mariage.}
Axa his daughter to wife. \V Who going together, she was moued by her
husband to aske a field of her father, and ſhe ſighed as ſhe ſate on her
aſſe. To whom Caleb ſaid: What aileth thee? \V But ſhe anſwered: Geue me
a bleſſing:
\SNote{Becauſe ordinarily the ſouth part of the world is more drie and
barren then other partes, a barren place is called a ſouth land, in
reſpect of more fertile ground.}
a South and drie Land thou haſt geuen me, ioyne alſo a waterie. Caleb
therfore gaue her a waterie ground aboue & beneath. \V This is the
poſſeſſion of the tribe of the children of Iudas by their kinredes. \V
And the cities from the vttermoſt partes of the children of Iudas by the
borders of Edom on the South: were Gabſeel and Eder and Iagur, \V and
Cina and Dimona and Adada, \V and Cades, and Aſor, and Iethnan, \V Ziph
and Telem and Baloth, \V Aſor the new and Carioth, Heſron, this is
Aſor. \V Amam, Sama, and Molada, \V and Aſergadda and Haſſemon and
Bethphelet, \V and Haſerſual and Berſabee and Baziothia \V and Baala and
Iim
\Fix{an}{and}{obvious typo, fixed in other}
Eſem, \V and Eltholad and Ceſil and Harma \V and Siceleg, and Medemena
and Senſenna, \V Labaoth and Selim and Aen and Remon. Al the cities
twentie nine, and their villages. \V But in the champayne countries:
Eſtaol and Sarea and Aſena, \V and Sanoe & Engannim and Taphua and
Enaim, \V and Ierimoth, and Adullam, Socho and Azeca, \V and Saraim and
Adithaim and Gedera and Gederothaim: fourteene cities, and their
villages. \V Sanan and Hadaſta and Magdalgad, \V Delean and Maſepha and
Iecthel, \V Lachis and Baſcath and Eglon, \V Chebbon and Leheman and
Cethlis \V and Gideroth and Bethdagon and Naama and Maceda: ſixtene
cities, and their
%%% 0519
villages. \V Labana and Ether and Aſan, \V Iephtha and Eſna and Neſib, \V
and Ceila and Achzib and Mareza: nine cities, and their villages. \V
Accaron with the townes and villages therof. \V From Accaron vnto the
ſea: al places that bend toward Azotus and the villages therof. \V
Azotus with the townes and villages therof. Gaza with the townes and
villages therof, vnto the torrent of Ægypt, and the great ſea is the
border therof. \V And in the mountayne: Samir and Iether and Socoth \V
and Danna and Cariath ſenna, this is Dabir: \V Anab and Iſtemo and Anim,
\V Goſen and Olon and Gilo: eleuen cities and their villages. \V Arab
and Ruma and Eſaan, \V and Ianum and Beth thaphua and Apheca, \V
Athmatha and Cariath-arbe, this is Hebron, and Sior:
%%% o-0467
nine cities and their villages. \V Maon and Carmel and Ziph and Iota, \V
Iezrael and Iucadam and Zanoe, \V Accain, Gabaa and Thamna: ten cities
and their villages. \V Halhul, and Bethſur, and Gedor, \V Mareth, and
Beth-anoth, & Eltecon: ſix cities and their villages. \V Cariathbaal,
this is Cariathiarim the citie of wooddes, and Arebba: two cities and
their villages. \V In the deſert Beth-araba, Meddin, and Sachacha, \V
and Nebſan, and the citie of ſalt, and Engaddi: ſix cities, and their
villages. \V But the Iebuſeite the inhabitant of Ieruſalem the children
of Iudas
\SNote{The Iebuſeites kept a ſtrong caſtle in Ieruſalem, til king Dauid
tooke it from them.
\XRef{2.~Reg.~5.}}
could not deſtroy: and the Iebuſeite dwelt with the children of Iudas
in Ieruſalem vntil this preſent day.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
