%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Joshua
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 0519
%%% o-0467
\startChapter[
  title={Chapter 16}
  ]

\Summary{The tribe of Ephraim, (younger ſonne of Ioſeph) receiue their
  lotte. 10.~The Chananeite yet dwelleth with them paying tribute.}

The lotte alſo of
\SNote{Ruben for the ſinne of inceſt loſing the
\Fix{priuelegies}{priueleges}{obvious typo, fixed in other}
of birthright
\XRef{(Gen.~49.)}
Prieſthood was geuen to Leui, the kingdome to Iuda, and duble portiõ to
Ioſeph.
\Cite{Paraphr. Chald.}}
the children of Ioſeph fel from Iordan againſt Iericho and the Waters
therof, on the eaſt: the wildernes which goeth vp from Iericho to the
mountaine of Bethel: \V and goeth out from Bethel to Luza: and paſſeth
the border of Archia, to Atharoth. \V And deſcendeth Weſtward, by the
border of Iephleti, vnto the borders of Beth horon the lower, and to
Gazer: and their countries are ended by the great ſea: \V and Manaſſes
and Ephraim the children of Ioſeph poſſeſſed it. \V And the border of
the children of Ephraim was made according to their kinredes: and their
poſſeſſion toward the Eaſt was Ataroth addar vnto Beth-horon the
higher. \V And the confines goe out vnto the
%%% 0520
ſea: but Machmethath looketh to the North, and it compaſſeth the borders
againſt the Eaſt into Thanath-ſelo: and paſſeth through on the Eaſt ſide
to Ianoe. \V And it goeth downe from Ianoe into Ataroth & Naaratha: and
it cometh into Iericho, and goeth out to Iordan. \V From Taphua it
paſſeth through againſt the ſea into the Valley of reedes, and the
iſſues therof are into the moſt ſalt ſea. This is the poſſeſſion of the
tribe of the children of Ephraim by their families. \V And cities with
their villages were ſeparated to the children of Ephraim in the middes
of the poſſeſſion of the children of Manaſſes, \V and the children of
Ephraim ſlew not the Chananeite, which dwelt in Gazer: and the
Chananeite dwelt in the middes of Ephraim vntil this day tributarie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
