%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Colossians
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 2782
%%% o-2642
\startChapter[
  title={Chapter 3}
  ]

\Summary{He exhorteth to mortifie & put off al corrupt manners of the
  old man, & to put on ſuch vertues as are for the new man. 18.~In
  particular alſo, wiues, and husbands, children and parents and
  maiſters, each ſort to doe their duetie.}

Therfore if you be riſen with Chriſt, ſeeke the things that are aboue;
where Chriſt is ſitting on the right hand of God. \V Mind the things
that are aboue, not the things that are vpon the earth.

\V For you are dead; and your life is hidde with Chriſt in God. \V When
Chriſt ſhal appeare, your life; then you alſo ſhal appeare
%%% 2783
with him in glorie.

\V
\CNote{\XRef{Eph.~5,~3.}}
Mortifie therfore your members that are vpon the earth, fornication,
vncleaneſſe, luſt, euil concupiſcence, and
\LNote{Auarice, which is the ſeruice of Idols.}{Here
\MNote{Heretical and foolish tranſlation.}
is a maruelous impudent and foolish corruptiõ in the vulgar
\Cite{English Bible printed the yere 1577}
and (as it ſeemeth) moſt authoriſed. Where for their errour againſt the
Images of Chriſt and his Saints, and to make image and Idol, alone; the
tranſlatour, for that which the Apoſtle ſaith in
Greek, \Emph{Couetouſnes in idolatrie}, maketh him ſay in English 
\Emph{Couetouſnes in worshipping of Images}: as alſo
\XRef{Eph.~5,~4.}
he tranſlateth thus, \Emph{The couetous perſon is a worshipper of
Images}: for that which the Apoſtle ſaith: \Emph{The couetous man is an
idolater}, meaning ſpiritual idolatrie, becauſe he maketh money his
God. In which ſenſe to cal this ſpiritual idolatrie, worshipping of
Images, is too ridiculous, and muſt needs proceed of blind hereſie.} 
auarice, which is
%%% !!! marked in both, present only in other
\TNote{\G{εἰδωλολατρία}}
the ſeruice of Idols. \V For which things the wrath of God commeth vpon
the children of incredulitie. \V In which you alſo walked ſometime, when
you liued in them. \V But now lay you alſo away anger, indignation,
malice, blaſphemie, filthie talke out of your mouth. \V Lie not one to
another:
\CNote{\XRef{Eph.~4,~22.}}
ſpoiling your ſelues of the old man with his actes, \V and
\LNote{Doing on the new.}{By
\MNote{Iuſtice an inherent qualitie in vs.}
this and the whole diſcourſe of this chapter conteining as exhortation
to good life and to put on the habit of the new man with al vertues we
may ſee, our iuſtice in Chriſt to be a very qualitie and forme inherent
in our ſoul, adorning the ſame, and not an imputation only of Chriſtes
righteouſnes, or a hiding only of our ſinnes and wickednes, which the
Heretikes falſely affirme to remaine in vs after Baptiſme and alwaies
during life. See S.~Auguſtin
\Cite{de pec. mer. & remiſ. li.~2. c.~7.}
&
\Cite{cont. Iulian. .ib.~6. c.~7.}}
doing on the new, him that is renewed vnto knowledge,
\CNote{\XRef{Gen.~1,~26.}}
according to the image of him that created him. \V Where there is not,
Gentile & Iew, circumciſion and prepuce, Barbarous and Scythian, bond
and free: but al, and in al Chriſt.

\V Put ye on therfore as the elect of God, holy, and beloued,
\CNote{\XRef{Eph.~4,~32.}}
the bowels of mercie, benignitie, humilitie, modeſtie, patience, \V
ſupporting one another, & pardoning one another, if any haue a quarel
againſt any man. As alſo our Lord hath pardoned vs: ſo you alſo. \V But
aboue al theſe things haue charitie, which is the band of perfection: \V
and let the peace of Chriſt
\SNote{\G{βραβευέτω} triumph and haue the victorie.}
exult in your harts, wherin alſo you are called in one body: and be
thankeful. \V Let the word of Chriſt dwel in you aboũdantly, in al
wiſedom: teaching and admoniſhing your owne ſelues, with pſalmes,
hymnes, and ſpiritual canticles, in grace ſinging in your harts to
God. \V Al whatſoeuer you doe in word or in worke, al things in the name
of our Lord \Sc{Iesvs} Chriſt, giuing thankes to God and the Father by
him.

%%% o-2643
\V
\CNote{\XRef{Ep.~5,~22.}}
Women be ſubiect to your husbands, as it behoueth in our Lord. \V
\CNote{\XRef{1.~Pet.~3,~7.}}
Men, loue your wiues and be not bitter toward them. \V
\CNote{\XRef{Eph.~6,~1.}}
Children obey your parents in al things: for that is wel pleaſing to our
Lord. \V Fathers prouoke not your children to indignation; that they
become not diſcouraged. \V
\CNote{\XRef{Eph.~6,~5.}
\XRef{Tit.~2,~9.}
\XRef{1.~Pet.~2,~11.}}
Seruants, obey in al things your Maiſters according to the fleſh, not
ſeruing the eye, as pleaſing men, but in ſimplicitie of hart, fearing
God. \V Whatſoeuer you doe, worke it from the hart as to our Lord, and
not to men: \V knowing that you ſhal receiue of our Lord the
\SNote{Retribution or reward for good workes: \G{ἀνταπόδοσις}, which
ſignifieth rendring one for another.}
retributiõ of inheritance. Serue our Lord Chriſt. \V For he that doeth
iniurie, ſhal receiue that which he hath done vniuſtly: and
\CNote{\XRef{Deu.~10,~17.}
\XRef{Ro.~2,~11.}
\XRef{Gal.~2,~6.}}
there is not acceptation of perſons with God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

