%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Colossians
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 2784
%%% o-2643
\startChapter[
  title={Chapter 4}
  ]

\Summary{He exhorteth to inſtance in praier, 5.~and to wiſedom in
  behauiour. 7.~He ſendeth Tychicus. 10.~He doeth commendations, 15.~and
  inioyneth to be done.}

You Maiſters, that which is iuſt and equal, doe to your ſeruants:
knowing that you alſo haue a Maiſter in heauen. \V
\CNote{\XRef{Luc.~18,~1.}
\XRef{Eph.~6,~18.}}
Be inſtant in praier; watching in it in thanks-giuing, \V
\CNote{\XRef{2.~Theſ.~18,~2.}}
praying withal
\SNote{S.~Paul euer much deſireth the praiers of the faithful: wherby we
learne the great efficacie of them.}
for vs alſo, that God may open vnto vs the doore of ſpeach to ſpeake the
myſterie of Chriſt (for the which alſo I am bound) \V that I may
manifeſt it, ſo as I ought to ſpeake. \V
\CNote{\XRef{Eph.~5,~25.}}
Walke with wiſedom toward them that be without;
%%% o-2644
redeeming the time. \V Your talke alwaies, in grace let it be ſeaſoned
with ſalt: that you may know how you ought to anſwer euery man.

\V The things that are about me, Tychicus, our deareſt Brother, and
faithful Miniſter, & fellow-ſeruant in our Lord, wil make you vnderſtand
al, \V whom I haue ſent to you for this ſame purpoſe, that he may know
the things that concerne you, and may comfort your harts, \V with
\CNote{\XRef{Phile.~10.}}
Oneſimus, the moſt deare and faithful Brother who is of you. Al things
that are done here, shal they doe you to vnderſtand.

\V Ariſtarchus my fellow-priſoner ſaluteth you, & Marke the coſin-german
of Barnabas (concerning whom you haue receiued commandements, If he come
to you, receiue him) \V and Ieſus that is called Iuſtus: who are of the
Circumciſion. Theſe only are my coadiutours in the Kingdom of God: which
haue been a
\Fix{confort}{comfort}{likely typo, fixed in other}
to me. \V Epaphras ſaluteth you who is of you, the ſeruant of Chriſt
\Sc{Iesvs}, alwaies careful for you in prayers, that you may ſtand
perfect and ful in al the wil of God. \V For I giue him teſtimonie that
he hath much
\SNote{He did not only pray, but tooke other great paines to procure
God's grace for the Coloſſians: perhaps by watching, faſting, & doing
other pennance of body; that God would not ſuffer them to fal from their
receiued faith to the Sect of Simon Magus or the Iudaizing Chriſtians.}
labour for you, & for them that be at Laodicia, and that are at
Hierapolis. \V
\CNote{\XRef{2.~Tim.~4,~10.~11.}}
Luke, the moſt deare phyſicion, ſaluteth you; and Demas. \V Salute the
Brethren that are at Laodicia; and Nymphas, and the Church that is in
his houſe. \V And when the epiſtle shal be read with you, make that it
be read alſo in the Church of the Laodicians: and that you read that
which is of the Laodicians. \V And ſay to Archippus: See the Miniſterie
which thou haſt receiued of our Lord, that thou fulfil it. \V The
ſalutation: with mine owne hand, Paules. Be mindful of my bands. Grace
be with you. Amen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

