%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 0454
%%% o-0408
\startChapter[
  title={Chapter 18}
  ]

\Summary{In ſteed of other inheritance Prieſtes and Leuites haue
  prouiſion by Sacrifices and oblations. 9.~Al ſuperſtition to be
  auoided. 15.~Perpetuitie of prophetes, and finally one
  ſpecial \Sc{Prophet} (to wit, \Sc{Christ}) is promiſed. 20.~Falſe
  prophetes muſt be ſlaine.}

The prieſtes and Leuites, and al that are of the ſame tribe, shal haue
no part nor inheritance with the reſt of Iſrael, becauſe they shal eate
the ſacrifices of our Lord, and his oblations, \V and nothing els shal
they receiue of the poſſeſſion of their brethren: for our Lord him ſelfe
is their inheritance, as he hath ſpoken to them. \V This shal be the
right of the prieſtes from the people, and from them that offer
victimes: whether they immolate oxe, or sheepe, they shal geue to the
prieſt the shoulder and the mawe: \V the firſt fruites of corne, of
wine, and oile, and a part of the woole of their sheepe shearing. \V For
him hath our Lord choſen of al thy tribes, that he might ſtand, and
miniſter to our Lord, he and his ſonnes for euer. \V If a Leuite goe out
of one of thy cities of al Iſrael in the which he dwelleth, and would
come deſiring the place which our Lord shal chooſe, \V he shal
miniſter in the name of our Lord his God, as al his brethren the
Leuites, that shal ſtand at that time before our Lord. \V He shal
receiue the ſame portion of meates, that the reſt doe: beſide that,
which in his owne citie is dew to him by ſucceſſion from his fathers. \V
When thou art entred the Land, which our Lord thy God shal geue thee,
beware
%%% 0455
thou be not willing to imitate the abominations of thoſe nations. \V
Neither let there be found in thee any that shal expiate his ſonne, or
daughter, making them to paſſe through the fyre: or that demandeth of
ſouthſayers, and obſerueth dreames and diuinations, neither let there be
a ſorcerer, \V nor inchanter, nor that conſulteth with pithone, or
diuiners, and ſeeketh the truth of the dead. \V For al theſe thinges our
Lord abhorreth, and for theſe abominations wil he deſtroy them at thy
entring in. \V Thou shalt be perfect; and without ſpotte with our Lord
thy God. \V Theſe nations, whoſe land thou shalt poſſeſſe, heare
ſouthſayers and diuiners: but thou art otherwiſe inſtructed of our Lord
thy God. \V
\LNote{A \Sc{Prophet} of thy nation.}{Amongſt
\MNote{The ſame wordes may haue diuers literal ſenſes.}
other places this plainly proueth that the ſame wordes in holie
Scripture may haue diuers literal ſenſes. For firſt the coherence of the
text ſheweth, that God here promiſed to geue his people an other
extraordinarie prophet, after Moyſes death, of their owne nation: as wel
to take away occaſion of ſeeking to ſouthſayers, diuiners, and other
prophane prophetes of falſe goddes, ſtrictly forbidden in the wordes
going immediatly before; as in approbation of their conuenient deſire,
mentioned in the wordes folowing, to heare Gods wil, not by him ſelf,
nor by an Angel, but by Moyſes, who was now ſhortly to be taken from
them. And ſo this promiſe was firſt performed in Ioſue, ſucceding next
after Moyſes in gouernment. And as neede required God ceaſed not to ſend
more prophetes beſides their ordinarie Prieſts. Againe this place is
alſo vnderſtood of Chriſt our Sauiour, chief Prophete, and maſter of al
prophetes; S.~Peter ſo expounding it.
\XRef{Act.~3. v.~22. &~23.}}
A \Sc{prophet} of thy nation and of thy brethren like vnto me, wil our
Lord thy God raiſe vp to thee: him thou shalt heare, \V as thou didſt
requeſt of our Lord thy God in Horeb, when the aſſemblie was gathered,
and ſaidſt: I wil no more heare the voice of our Lord my God, and this
exceding great fire I wil ſee no more, leſt I die. \V And our Lord ſaid to
\Fix{mie:}{me:}{obvious typo, fixed in other}
They haue ſpoken al
%%% o-0409
thinges wel. \V A prophete wil I rayſe vp to them out of the middes of
their brethren like to thee: and I wil put my wordes in his mouth, and
he shal ſpeake al thinges that I shal command him. \V But he that wil
not heare his wordes, which he shal ſpeake in my name, I wil be the
reuenger. \V And the prophet that being depraued with arrogancie wil
ſpeake in my name, the thinges
\SNote{This ſorte of falſe prophets ſignified Heretikes, that preach
falſe things in Chriſts name.}
that I did not command him to ſay,
\SNote{Theſe prefigured Apoſtates, which renouncing Chriſt, expreſly
profeſſe falſe goddes.}
or in the name of ſtrange goddes, shal be ſlaine. \V And if in ſecrete
cogitation thou anſwer: How shal I vnderſtand the word, that our Lord
ſpake not? \V This ſigne thou shalt haue: That which the ſame prophete
foretelleth, in the name of the Lord and cometh not to paſſe: that our
Lord hath not ſpoken, but by the arrogancie of his minde the prophet
hath forged it: and therfore thou shalt not feare him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
