%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 0445
%%% o-0399
\startChapter[
  title={Chapter 13}
  ]

\Summary{Falſe Prophets muſt be ſlaine, 6.~how nere ſoeuer they be in
  kinred, or freindshipe. 12.~The whole citie that shal permite falſe
  doctrin muſt be vtterly deſtroyed, men, beaſtes, and al moueables, and
  neuer be built againe.}

If there riſe in the middes of thee a prophete, or one that ſaieth he
hath ſeene a dreame, and foretel a ſigne and a wonder, \V and it come to
paſſe which he ſpake, and he ſay to thee:
\SNote{Noueltie in Religion is a marke of idolatrie or hereſie.}
Let vs goe, and folow ſtrange goddes, which thou knoweſt not, and let vs
ſerue them: \V thou shalt not heare
%%% 0446
the wordes of that prophete or dreamer: for the Lord your God tempteth
you, that it may appeare whether you loue him or no, with al your hart,
and with al your ſoule. \V Folow the Lord your God, and feare him, and
keepe his commandementes, and heare his voice: him you shal ſerue, and
to him
%%% o-0400
you shal cleaue. \V And that prophete or forger of dreames shal be
ſlaine: becauſe he ſpake that he might auert you from the Lord your God,
which brought you out of the Land of Ægypt, and redemed you from the
houſe of ſeruitude: that he might make thee to erre from the way, that
the Lord thy God commanded thee: and thou shalt take away the euil out
of the middes of thee. \V If thy brother the ſonne of thy mother, or thy
ſonne or daughter, or thy wife that is in thy boſome, or thy freind,
whom thou loueſt as thy ſoule, wil perſwade thee ſecretly, ſaying: Let
vs goe, and ſerue ſtrange goddes, which thou knoweſt not, nor thy
fathers, \V of al nations round about, that be nigh or farre, from the
beginning vnto the end of the earth, \V conſent not to him, nor heare
him, neither let thyne eie ſpare him to pitie and hide him, \V but
\SNote{Euerie priuat man is not commanded, nor warrented by this to kil: but
euerie one is bound to informe the Magiſtrate, and ſo by order of
iuſtice to procede againſt the wicked.}
forthwith thou shalt kil him. Let thy hand be firſt vpon him, and after
thee al the people lay hand on him. \V With ſtones shal he be ſtoned to
death: becauſe he would haue withdrawen thee from the Lord thy God, which
brought thee out of the Land of Ægypt, from the houſe of ſeruitude: \V
that al Iſrael hearing may feare, and may doe no more any thing like to
this. \V If in one of thy cities, which the Lord thy God shal geue thee
to inhabite, thou heare ſome ſay: \V There are gone forth
\SNote{Such as wil not indure diſcipline are called \Emph{children of
Belial}, that is \Emph{vvithout yoke}.}
children of Belial out of the middes of thee, and haue auerted the
inhabitants of their citie, and haue ſaid: Let vs goe, and ſerue ſtrange
goddes which you know not: \V inquire carefully, and diligently, the
truth of the thing being looked into, if thou finde it certaine that is
ſaid, and that this abomination is in act committed, \V thou shalt
forthwith ſtrike the inhabitantes of that citie in the edge of the
ſworde, and shalt deſtroy it and al thinges that are in it, vnto the
very beaſtes. \V What ſtuffe alſo ſoeuer there is, thou shalt gather
together in the middes of the ſtreates therof, and shalt burne it
with the citie it ſelfe, ſo that thou conſume al thinges to the Lord thy
God, and it be a heape for euer: it shal be built no more, \V and there
shal nothing ſticke in thy hand of that anathema:
%%% 0447
%%% !!! Extra SNote marker? No note, missing in other
%%% \SNote{}
that our Lord may be turned from the wrath of his furie, and may haue
mercie on thee, and multiplie thee as he ſware to thy fathers, \V when
thou shalt heare the voice of the Lord thy God, keeping al his
preceptes, which I command thee this day, that thou mayeſt doe that
which is pleaſing in the ſight of the Lord thy God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
