%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 0450
%%% o-0404
\startChapter[
  title={Chapter 16}
  ]

\Summary{Three
%%% !!! Slight rearangement of this SNote and the next three MNotes
%%% !!! Is this right?
\SNote{The reſt of the feaſtes are mentioned
\XRef{Leuit.~23.}
\XRef{Num.~28.}
\XRef{&~29.}
Here only three of the principal.}
more ſolemne feaſtes to be kept euerie yeare, Paſch, 9.~Pentecoſt,
13.~and the feaſt of tabernacles. 18.~Iuſt Iudges to be appointed in
euerie citie. 21.~Al occaſions of Idolatrie to be auoyded.}

Obſerue the moneth of new corne, and
\MNote{1.~Paſch.}
the firſt of the ſpring time, that thou mayeſt make the Phaſe to our
Lord thy God: becauſe in this moneth our Lord thy God, brought thee out
of Ægypt by night. \V And thou shalt immolate the Phaſe to our Lord thy
God, of sheepe, and of oxen in the place, which our Lord thy God shal
chooſe, that his name may dwel there. \V Thou shalt not eate in it
leuened bread: Seuen daies shalt thou eate without leuen, the bread of
affliction, becauſe in feare didſt thou come out of Ægypt: that thou
mayeſt remember the day of thy comming out of Ægypt, al the dayes of thy
life. \V Leuened shal not appeare in al thy coaſtes for ſeuen daies, and
there shal not remayne of the flesh of that which was immolated at euen
the firſt day vntil morning. \V Thou canſt not immolate the Phaſe in
euerie one of thy cities, which our Lord thy God wil geue thee; \V but
in the place, which our Lord thy God shal chooſe, that his name may dwel
there: thou shalt immolate the Phaſe at euen at the going downe of the
ſunne, when thou cameſt out of Ægypt. \V And thou shalt boyle, and eate
it in the place, which our Lord thy God shal chooſe, and in the morning
ryſing vp thou shalt goe into thy tentes. \V Six
%%% 0451
daies shalt thou eate azymes: and in the ſeuenth day, becauſe it is the
collection of our Lord thy God, thou shalt doe no worke. \V
\MNote{2.~Pentecoſt.}
Seuen weekes shalt thou number from that day wherein thou didſt put the
ſickle to the corne, \V and thou shalt celebrate the feſtiual day of
weekes to our Lord thy God, a voluntarie oblation of thy hand, which
thou shalt offer according to the bleſſing of our Lord thy God: \V and
thou shalt feaſt before our Lord thy God, thou, & thy ſonne, and thy
daughter, and thy man ſeruant, and thy woman ſeruant, and the Leuite
that is within thy gates, and the ſtranger and pupil and widow, which
abide with you: in the place which our Lord thy God shal chooſe, that his
name may dwel there: \V and thou shalt remember that thou waſt a ſeruant
in Ægypt: and thou shalt keepe and doe the thinges that are
commanded. \V
\MNote{3.~Feaſt of Tabernacles.}
The ſolemnitie alſo of Tabernacles thou ſhalt celebrate ſeuen daies,
when thou haſt gathered thy fruite of the barne floore and the
preſſe: \V and thou shalt feaſt in the feſtiuitie, thou, thy ſonne, and
thy daughter,
%%% o-0405
thy man ſeruant and woman ſeruant, the Leuite alſo and ſtranger, and
pupil and widow that are within thy gates. \V Seuen daies shalt thou
celebrate the feaſtes to our Lord thy God in the place, which our Lord
shal chooſe: and our Lord thy God wil bleſſe thee in al thy fruites, and
in euerie worke of thy handes, and thou shalt be in ioye. \V Three times
in a yeare shal al thy male appeare in the ſight of our Lord thy God in
the place which he shal chooſe: in the ſolemnitie of Azymes, in the
ſolemnitie of weekes, and in the ſolemnitie of Tabernacles. There shal
not appeare before our Lord any emptie: \V but euerie one shal offer
according to that he hath, according to the bleſſing of our Lord his
God, which he shal geue him. \V Iudges and maiſters shalt thou appoynt
in al thy gates, which our Lord thy God shal geue thee, in euerie of thy
tribes: that they may iudge the people with iuſt iudgement, \V and not
decline to either part. Thou shalt not accept perſon, nor giftes:
becauſe that giftes blinde the eies of the wiſe, and change the wordes
of the iuſt. \V
\SNote{It is not ynough to doe that is iuſt except it be donne iuſtly,
to a good end for loue of iuſtice.}
Iuſtly shalt thou purſew that which is iuſt: that thou mayeſt liue and
poſſeſſe the Land, which our Lord thy God shal geue thee. \V
Thou shalt plante no groue, nor any tree neere the altar of our Lord thy
God. \V Neither ſhalt thou make nor ſette to thy ſelf a ſtatue: which
thing our Lord thy God hateth.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
