%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 0441
%%% o-0395
\startChapter[
  title={Chapter 11}
  ]

\Summary{For the benefites of God (wherof ſome are repeted, and others
  promiſed) the Iſraelites are bound to loue him. 16.~But if they
  forſake him he threatneth punishmentes, 26.~propoſing benediction and
  malediction as they shal deſerue.}

Loue therfore the Lord thy God, and obſerue his preceptes and
ceremonies, his iudgementes and commandmentes at al time. \V Know this
day the thinges that your children know not, who ſaw not the diſcipline
of the Lord your God, his great doinges and ſtrong hand, and ſtretched
out arme, \V the ſignes and workes which he did in the middes of Ægypt
to Pharao the king, and to al his land, \V and to al the hoſte of the
Ægyptians, and to their horſes and charriottes: how the waters of the
red ſea couered them, when they purſewed you, and how our Lord deſtroyed
them vntil this preſent day: \V and to you what thinges he hath done in
the
%%% o-0396
wildernes, til you came to this place: \V and to Dathan and Abiron the
ſonnes of Eliab, which was the ſonne of Ruben: whom the earth opening
her mouth ſwalowed vp with their houſes and tabernacles, and al their
ſubſtance, which they had in the middes of Iſrael. \V Your eies haue
ſeene al the great workes of our Lord, that he hath done, \V that you
may keepe al his commandementes, which I command you this day, and may
enter in, and poſſeſſe the Land, to the which you enter, \V and may liue
in it a great time: which our Lord by oath promiſed to your fathers, and
to their ſeede, flowing with milke and honie. \V For the Land, which
thou goeſt to poſſeſſe, is not as the Land of Ægypt, which thou cameſt
out of, where when the ſeede is ſowen, waters are brought in to water it
after the maner of gardens: \V but it is hilly and champion, expecting
raine from heauen, \V which the Lord thy God doth alwaies viſite, and
his eies are on it from the beginning of the yeare vnto the end
therof. \V If then you obey my commandementes, which I command you this
day, that you loue the Lord your God, and ſerue him
%%% 0442
with al your hart, and with al your ſoule: \V he wil geue rayne to your
Land
\SNote{Raine after ſeeding and before harueſt ſignifieth Gods grace
firſt ſtirring vp the ſoule, and aſſiſting the ſame to the end.}
the timely and the lateward, that you may gather your corne, and wine,
and oile, \V and haye out of the fieldes to feede your cattel, and that
your ſelues may eate and be filled. \V Beware leſt perhaps your hart be
deceiued, and you depart from our Lord, and ſerue ſtrange goddes, and
adore them: \V and our Lord being wrath shutte vp heauen, and the raine
come not downe, nor the earth geue her ſpring, and you perish quickly
from the excellent Land, which our Lord wil geue you. \V Put theſe my
wordes in your hartes and mindes, and hang them for a ſigne on your
handes, and place them betwen your eies. \V Teach your children that
they meditate them, when thou ſitteſt in thy houſe, & walkeſt on the
way, and lieſt downe and ryſeſt vp. \V Thou shalt write them vpon the
poſtes and gates of thy houſe: \V that thy daies may be multiplied, and
the dayes of thy children in the Land, which our Lord ſware to thy
fathers, that he would geue it them as long as the heauen hangeth ouer
the earth. \V For if you keepe the commandementes which I command you,
and doe them, that you loue the Lord your God, and walke in al his
wayes, cleauing to him, \V Our Lord wil deſtroy al theſe nations before
your face, and you ſhal poſſeſſe them, which are greater and ſtronger
then you. \V Euerie place, that your foote shal treade, shal be
yours. From the deſert, and from Libanus, from the great riuer Euphrates
vnto the weſt ſea shal be your borders. \V None shal ſtand againſt you:
your terrour and feare shal the Lord your God geue vpon al the land that
you shal treade, as he hath ſpoken to you. \V Behold I ſette forth in
your ſight this day
\SNote{God worketh, and we cooperate, for he taketh not away, but
helpeth freewil.
\Cite{S.~Aug. q.~15. in Deut.}}
benediction and malediction: \V benediction, if you obey the
commandementes of the Lord your God, which I command you this day: \V
malediction, if you obey not the commandementes of the Lord your God, but
reuolt from the way, which now I doe shew you, and walke after ſtrange
goddes,
%%% o-0397
which you know not. \V And when the Lord thy God shal haue brought thee
into the Land, to the which thou goeſt to inhabite, thou shalt put the
benediction vpon mounte Garizim, the malediction vpon mounte Hebal: \V
which are beyond Iordan behinde the way that bendeth to the going downe
of the ſunne in the Land of the Chananeite, which dwelleth in the
champion countrie againſt Galgala, which is
%%% 0443
beſide the valle that reacheth and entreth farre. \V For you shal paſſe
ouer Iordan, to poſſeſſe the Land, which the Lord your God wil geue you,
that you may haue and poſſeſſe it. \V See therfore that you fulfil the
ceremonies and iudgementes, which I shal ſette this day in your ſight.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
