%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 34
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-34


\project douay-rheims


%%% 0488
%%% o-0438
\startChapter[
  title={Chapter 34}
  ]

\Summary{Moyſes ſeeth the promiſed land, but is not ſuffered to goe into
  it. 5.~He dieth at the age of 120.~yeares. God burieth his bodie
  ſecretly, and al Iſrael mourne for him thirtie dayes. 9.~Ioſue
  replenished (by impoſition of Moyſes handes) with the ſpirite of God
  ſuccedeth. 10.~But Moyſes for his ſpecial familiaritie with God, and
  for moſt wonderful miracles is commended aboue al other Prophetes.}

Moyſes
\MNote{The fifth part.

The death, burial and ſingular praiſe of Moyſes.}
therfore went vp from the champion of Moab vpon mount Nebo, into the
toppe of Phaſga againſt Iericho: and our Lord shewed him
\SNote{God eleuated his viſiue powre aboue nature to ſee ſo farre.}
al the land of Galaad as farre as Dan, \V and al Nephthali, and the land
of Ephraim and Manaſſes, and al the Land of Iuda vnto the vtmoſt ſea, \V
and the ſouth part, and the bredth of the plaine of Iericho a citie of
plametrees as farre as Segor. \V And our Lord ſaid to him: This is the
Land, for the which I ſware to Abraham, Iſaac, and Iacob, ſaying: To thy
ſeede wil I geue it. Thou haſt ſeene it with thyne eies, and shalt not
paſſe ouer to it. \V And Moyſes the ſeruant of our Lord died there, in
the land of Moab, our Lord commanding it: \V and he buried him in the
%%% o-0439
valley of the Land of Moab, againſt Phogor: and
\SNote{Onlie Angels (whoſe miniſterie God vſed herein) knew the place of
his burial: leſt the Iewes prone to idolatrie might haue honored him for
God.}
no man hath knowne his ſepulchre vntil this preſent day.

\V Moyſes was an hundred and twentie yeares old when he died: his eie
was not dimme, neither were his teeth moued. \V And the children of
Iſrael mourned him in the champion countrie of Moab thirtie daies: and
the daies of their mourning that mourned for Moyſes were
accomplished. \V And Ioſue the ſonne of Nun was replenished with the
ſpirit of wiſedome, becauſe Moyſes did put his handes vpon him. And the
children of Iſrael obeied him, and did as our Lord commanded Moyſes. \V
And there roſe no more a prophete in Iſrael as Moyſes, whom our Lord had
knowen face to face, \V in al ſignes and wonders, which he ſent by him,
to doe in the Land of Ægypt to Pharao, and to al his ſeruantes, and to
his whole Land, \V and al the ſtrong hand, and great meruailes, which
Moyſes did before al Iſrael.

\Emph{The end of the fiue bookes of Moyſes, conteining the Law.}


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
