%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 0437
%%% o-0392
\startChapter[
  title={Chapter 9}
  ]

\Summary{Leſt they should impute the victories (which they shal haue) to
  them ſelues, 6.~they are put in mind of their often prouoking Gods
  wrath, 12.~by idolatrie, 22.~by murmuring, by concupiſcence, by
  contempt, and other ſinnes, 25.~for which they should haue bene
  deſtroyed, but God ſpared them for his promiſe made to Abraham Iſaac
  and Iacob.}

Heare Iſrael: Thou shalt goe ouer Iordan this day; to poſſeſſe verie
greate nations and ſtronger then thy ſelfe, huge cities, and walled
\SNote{Holie Scripture vſeth the figure \Emph{Hyperbola}, folowing the
vulgar maner of ſpeakĩg as wel to helpe the vnderſtanding, as to moue
affectiõ in great and extraordinarie things.}
euen vnto heauen, \V a great people and tal, the ſonnes of the Enacims,
whom thou haſt ſeene, and heard, againſt whom no man is able to
reſiſt. \V Thou shalt know therfore this day that the Lord thy God him
ſelfe wil paſſe ouer before thee, a deuouring and conſuming fyre, who
shal deſtroy and abolish and bring them to nothing before thy face
quickly, as he hath ſpoken to thee. \V Say not in thy hart, when the
Lord thy God shal haue deſtroyed them in thy ſight: For my iuſtice hath
our Lord brought me in to poſſeſſe this land, wheras theſe nations were
deſtroyed for their impieties. \V For not becauſe of thy iuſtices, and
equitie of thy hart doeſt thou enter in to poſſeſſe their landes: but
becauſe they haue done impiouſly, at thy
%%% 0438
entring in they are deſtroyed: and that our Lord might accomplish his
word, which by oath he promiſed to thy fathers Abraham, Iſaac, and
Iacob. \V Know therfore that not for thy iuſtices hath the Lord thy God
geuen thee this excellent land in poſſeſſion, wheras thou art a verie
ſtiffe necked people. \V Remember, and forget not how thou didſt prouoke
%%% o-0393
the Lord thy God to wrath in the wilderneſſe. From the ſame day, that
thou cameſt out of Ægypt vnto this place, thou haſt alwayes contended
againſt our Lord. \V For in Horeb alſo thou didſt prouoke him, and being
wrath he would haue deſtroyed thee, \V when I went vp into the mounte,
to receiue the tables of ſtone, of the couenant which our Lord made with
you: and I continewed in the mounte fourtie daies and nightes, not
eating bread, nor drinking water. \V And our Lord gaue me two tables of
ſtone written with the finger of God, and conteyning al the wordes that
he ſpake to you in the mounte from the middes of the fyre, when the
aſſemblie of the people was gathered. \V And when fourtie dayes were
paſſed, and as manie nightes, our Lord gaue me the two tables of ſtone,
the tables of couenant, \V and he ſaid to me: Ariſe, and goe downe from
hence quickly: for thy people, which thou didſt bring out of Ægypt, haue
quickly forſaken the way, that thou haſt ſhewed them, and haue made them
\SNote{The ſimilitude of a calfe and called it their god.
\XRef{Exod.~32.}}
a molten idol. \V And againe our Lord ſaid to me: I ſee that this people
is ſtiffe necked: \V ſuffer me that I may deſtroy them, and abolish
their name from vnder heauen, and may ſet thee ouer a Nation, that is
greater and ſtronger then this. \V And when I came downe from the
burning mounte, and held the two tables of couenant with both handes, \V
and ſaw that you had ſinned to the Lord your God, and had made you a
molten calfe, and had quickly forſaken his way, which he had shewed
you: \V I caſt the tables out of my handes, and brake them in your
ſight. \V And I fel downe before our Lord as before, fourtie dayes and
nightes not eating bread, nor drinking water, for al your ſinnes, which
you committed againſt our Lord, and prouoked him to wrath: \V for I
feared his indignation and anger, wherwith being moued agaynſt you, he
would haue deſtroyed you. And our Lord heard me this time alſo. \V
Againſt Aaron alſo being exceeding angrie, he would haue deſtroyed him,
and for him, in like maner did I
%%% 0439
pray. \V And your ſinne that you had committed, that is, the calfe, I
tooke, and burnt it with fyre, and breaking it into peeces, and bringing
it wholy into duſt, I threw it into the torrent, that deſcendeth from the
mount. \V In the burning alſo and in the tentation, and in the
Sepulchres of concupiſcence you prouoked our Lord: \V and when he ſent
you from Cadeſbarne, ſaying: Goe vp, and poſſeſſe the Land, that I haue
geuen you, and you contemned the commandement of your Lord God, and did
not beleue him, neither would you heare his voice: \V but were alwaies
rebellious from the day that I beganne to know you. \V And I lay before
our Lord fourtie dayes and nightes, in the which I humbly beſought him,
that he would not deſtroy you as he had threatened: \V and praying I
ſaid: Lord God, deſtroy not thy people, and thyne inheritance, which
thou haſt redemed in thy greatnes, whom thou didſt bring out of Ægypt in
a ſtrong hand. \V Remember thy ſeruantes Abraham, Iſaac, and Iacob:
regard not the
%%% o-0394
ſtubbournes of this people, and his impeitie and ſinne: \V leſt
perhappes the inhabitantes of the land, out of which thou haſt brought
vs, ſay: The Lord could not bring them in vnto the Land, that he
promiſed them, and he hated them: therfore did he bring them forth, that
he might kil them in the wildernes. \V Which are thy people and thyne
inheritance, whom thou didſt bring forth in thy great ſtrength, and in
thy ſtretched out arme.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
