%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 0419
%%% o-0376
\startChapter[
  title={Chapter 1}
  ]

\Summary{Moyſes beginneth, the firſt day of the eleuenth moneth and
  fourtith yeare after the children of Iſrael parted from Ægypt, to
  repete and explicate the Law; 6.~firſt putting them in mind of Gods
  munifence, his owne and other ſuperiors care ouer them, their
  ingratitude, incredulitie, murmuring, 34.~and punishment for the ſame.}


These
\MNote{The firſt part.

A repetition of Gods benefites, the peoples ingratitude, and
puniſhment.}
are the wordes, which Moyſes ſpake to al Iſrael beyond Iordan, in the
champion wilderneſſe, againſt the Read ſea, betwen Pharan and Thophel and
Laban and Haſeroth, where there is verie much gold: \V eleuen daies from
Horeb by the way of mount Seir to Cadeſbarne. \V The fourtith yeare, the
eleuenth moneth, the firſt day of the moneth Moyſes ſpake to the
children of Iſrael al thinges that our Lord had commanded him to ſay
vnto them: \V after that he had ſtroke Sehon king of the Amorrheites,
which dwelt in Heſebon: and Og the king of Baſan which abode in Aſeroth,
and in Edrai, \V beyond Iordan in the Land of Moab. And Moyſes began to
expound the law, and to ſay: \V The Lord our God ſpake to vs in Horeb,
ſaying: It is ſufficient for you that you haue ſtayed in this
mountaine: \V returne, and come to the mountaine of the Amorrheites, and
to the reſt that are next to it champion and hillie and lower places
againſt the South, and beſide the shore
%%% o-0377
of the ſea, the Land of the Chananeites, and of Libanus vnto the greate
riuer Euphrates. \V Behold (quoth he) I haue deliuered it to you: enter
in and poſſeſſe it, vpon the which our Lord ſware to
%%% 0420
your fathers Abraham, Iſaac, and Iacob, that he would geue it to them,
and to their ſeede after them. \V And I ſaid to you at that time: \V I
alone can not ſuſteyne you: becauſe the Lord your God hath multiplied
you, and you are this day as the ſtarres of heauen, verie manie. \V (The
Lord God of your fathers adde to this number manie thouſandes, and
bleſſe you as he hath ſpoken.) \V I alone am not able to ſuſteyne your
buſineſſes, and the charge of you and your quareles. \V Geue from among
you wiſe and ſkilful men, and ſuch whoſe conuerſation is approued in
your tribes, that I may appoint them your princes. \V Then you anſwered
me: The thing is good which thou meaneſt to do. \V And I tooke of your
tribes men wiſe and noble, and appointed them princes, tribunes, and
centurions, and quinquagenarians, and deanes, that might teach you al
thinges. \V And I commanded them, ſaying: Heare them, and iudge that
which is iuſt: whether he be the ſame countrie man, or a ſtranger. \V
There shal be no difference of perſons, ſo shal you heare the litle as
the great: neither shal you accept any mans perſon, becauſe it is the
iudgement of God. And if any thing ſeme hard to you, referre it to me,
and I wil heare it. \V And I commanded al thinges that you ought to
do. \V And departing from Horeb, we paſſed through the terrible and huge
wilderneſſe, which you ſaw, by the way of the mountaine of the
Amorrheite, as the Lord our God had commanded vs. And when we were come
into Cadeſbarne, \V I ſaid to you: You are come to the mountaine of the
Amorrheite, which the Lord our God wil geue to vs. \V See the Land which
the Lord thy God geueth thee: goe vp and poſſeſſe it, as the Lord our
God hath ſpoken to thy fathers: feare not, neither dread you any
thing. \V And you came al vnto me, and ſaid: Let vs ſend men that may
view the Land: and may bring vs word what way we shal aſcend, and to
what cities to goe. \V And becauſe the ſaying pleaſed me, I ſent of you
twelue men, one of euerie tribe. \V Who when they had gone, and were
aſcended into the mountaines, they came as farre as the Valley of
cluſter: and the Land being viewed, \V taking of the fruites therof, to
shew the fruitfulneſſe, they brought vnto vs, and ſaid: The Land is
good, which the Lord our God wil geue vs. \V And you would not goe vp,
but being incredulous at the word of the Lord our God, \V you murmured
in your
%%% 0421
tabernacles, and ſaid: Our Lord hateth vs, and therfore hath brought vs
out of the Land of Ægypt, that he might deliuer vs into the hand of the
Amorrheite, and deſtroy vs. \V Whither ſhal we goe vp? the meſſengers
haue feared our hart, ſaying: The multitude is verie great, and taller
of ſtature then we: the cities greate, and fenſed euen vnto heauen, the
ſonnes of the Enacims we haue ſeene there. \V And I ſaid to you:
\SNote{God ſo helpeth his ſeruantes, that they alſo muſt cooperate.
\Cite{S.~Aug. q.~1. in Deut.}}
Feare not, neither be ye afrayed of them: \V Our Lord
%%% o-0378
God, which is your conductour, him ſelf wil fight for you, as he did in
Ægypt in the ſight of al. \V And in the wilderneſſe (thy ſelfe haſte
ſeene) the Lord thy God hath caried thee, as a man is wont to beare his
litle ſonne, al the way, that you haue walked, vntil you came to this
place. \V And neither ſo did you beleue the Lord your God, \V who went
before you in the way, and marked out the place wherein you ſhould pitch
your tentes, in the night ſhewing you the way by fyre, and in the day by
the piller of a clowde. \V And when our Lord had heard the voice of your
wordes, being wrath he ſware and ſaid: \V There shal not any of the men
of this wicked generation ſee the good Land, which by oath I promiſed to
your fathers: \V beſide Caleb the ſonne of Iephone. For he shal ſee it,
and to him I wil geue the Land, that he hath troden, and to his
children, becauſe he hath folowed the Lord. \V Neither is
\SNote{Difference of ſinnes.}
his indignation againſt the people to be merueiled at, wheras our Lord
being
\SNote{God is alſo angrie with his good ſeruantes, and puniſheth thẽ
temporally, for ſmal ſinnes.
\Cite{S.~Aug. q.~1. in Ioſue.}}
angrie with me alſo for you, ſaid: Neither shalt thou enter in
thither. \V But Ioſue the ſonne of Nun thy miniſter, he shal enter for
thee: exhort and
\Fix{ſtrenghen}{ſtrengthen}{obvious typo, fixed in other}
him, and he shal by lotte diuide the Land to Iſrael. \V Your litle ones,
of whom you ſaid that they should be ledde captiues, and your ſonnes
that this day know not the difference of good and euil, they shal enter
in: and to them I wil geue the Land, and they shal poſſeſſe it. \V But
returne you and goe into the wilderneſſe by the way of the Redde ſea. \V
And you anſwered me: We haue ſinned to our Lord: we wil goe vp and
fight, as the Lord our God hath commanded. \V And when you readie armed
went vnto the mountaine, \V our Lord ſaid to me: Say to them: Goe not
vp, and fight not, for I am not with you: leſt you fal before your
enemies. \V I ſpake, and you heard not: but reſiſting the commandement
of our Lord, and ſwelling with pride you went vp into the mountaine. \V
Therfore the
%%% 0422
Amorrheite that dwelt in the mountaines iſſuing forth, and coming to
meete you, purſewed you, as bees are wont to purſew: and ſmote you from
Seir as farre as Horma. \V And when returning you wept before our Lord,
he heard you not, neither would he condeſcend to your voice. \V You
abode therfore in Cadesbarne a great time.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
