%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 0422
%%% o-0379
\startChapter[
  title={Chapter 2}
  ]

\Summary{VVith commemoration of Gods continual protection of the
  Iſraelites, they are forbid to fight againſt the Idumeans, 9.~the
  Moabites, or Ammonites. 24.~But againſt Schon King of Heſebon they
  should fight, kil him and al his, and poſſeſse his land.}

And departing thence we came into the wilderneſſe, that leadeth to the
Redde ſea, as our Lord had ſaid to me: and we compaſſed the mountaine
Seir a long time. \V And our Lord ſaid to me: \V It is ſufficient for you
to haue compaſſed this mountaine: goe toward the North. \V And command thou
the people, ſaying: You shal paſſe by the borders of your brethren the
children of Eſau, which dwel in Seir, and they wil be affraid of you. \V
Looke diligently therfore that you ſturre not againſt them. For I wil
not geue you of their land ſo much as the ſteppe of one foote can
treade, becauſe I haue geuen the mountaine Seir to be the poſſeſſion of
Eſau. \V Meates you ſhal bie of them with money, and shal eate: bought
water shal you draw, and drinke. \V The Lord thy God hath bleſſed thee
in euerie worke of thy handes: he knoweth thy iourney, how thou haſt
paſſed this great wilderneſſe, for fourtie yeares the Lord thy God
dwelling with thee, & thou haſt wanted nothing. \V And when we had
paſſed by our brethren the children of Eſau, that dwelt in Seir, by the
champion way from Elath & from Aſiongaber, we came to the way, that
leadeth into the deſert of Moab. \V And our Lord ſaid to me: Fight not
againſt the Moabites, neither make battel againſt them: for I wil not
geue thee any of their land, becauſe I haue geuen Ar to the children of
Lot in poſſeſſion. \V Emim firſt were the inhabiters therof, a great
people, and valiant, and ſo tall that
\SNote{Theſe were men of very great ſtature, but not equal to the
giantes before the floud.}
they were thought, \V as it were giantes, of the Enacims ſtocke, & were
like the children of the Enacims. Moreouer the Moabites cal them
Emim. \V But in Seir before dwelt  the Horrins: who being expelled and
deſtroyed, the children of Eſau did inhabite
%%% 0423
it, as Iſrael did in the land of his poſſeſſion, which our Lord gaue
him. \V Ryſing vp therfore to paſſe the Torrent Zared, we came to it. \V
And the time, that we walked from Cadeſbarne vnto the paſſage of the
torrent Zared, was thirtie and eight yeares: vntil al the generation of
the men that were warriers was conſumed out of the campe, as our Lord
had ſworne: \V whoſe hand was againſt them, that they should perish from
among the campe. \V And after al the warryers were dead, \V our Lord
ſpake to me, ſaying: \V Thou shalt paſſe this day the borders of Moab, the
citie named Ar: \V and
%%% o-0380
approching vnto the frontiers of the children of Ammon, beware thou
fight not againſt them, neither once moue to battel: for I wil not geue
thee of the land of the children of Ammon, becauſe I haue geuen it to
the children of Lot in poſſeſſion. \V It was reputed the land of
giantes: and giantes in old time dwelt in it, whom the Ammonites cal
Zomzommim, \V a great and huge people, and of long ſtature, as the
Enacims whom our Lord deſtroyed before their face: and he made them to
dwel in their ſteede, \V as he had done to the children of Eſau, that
dwelt in Seir, deſtroying the Horrheites, and deliuering their land to
them, which they poſſeſſe vntil this preſent. \V The Heueites alſo, that
dwelt in Haſerim as farre as Gaza, the Capadocians expelled: who iſſuing
out of Capadocia, deſtroyed them, and dwelt in their ſteede. \V Ariſe
ye, and paſſe the torrent Arnon: behold I haue deliuered in thy hand
Sehon king of Heſebon the Amorrheite, and beginne to poſſeſſe his land,
and
\SNote{By this we are inſtructed to fight againſt infidels, but not
without ſpecial cauſe againſt chriſtians, ſignified by the childrẽ of
Lot and Eſau.}
make warre againſt him. \V This day wil I beginne to ſend thy terrour
and feare vpon the peoples, that dwel vnder the whole heauen: that
hearing thy name they may quake, and tremble after the manner of wemen
in trauel, and be pinched with ſorow. \V I ſent therfore meſſengers from
the wildernes of Cademoth to Sehon the king of Heſebon with peacable
wordes, ſaying: \V We wil paſſe through thy land, we wil goe the common
high way: we wil not decline neither to the right hand, nor to the
left. \V Sel vs meates for money, that we may eate: Geue vs water for
money, and ſo we wil drinke. Onlie this that thou wilt graunt vs
paſſage, \V as
\SNote{The Idumeans once denied them paſſage
\XRef{Num.~20. v.~20.}
but afterward granted therto.}
the children of Eſau haue done, that dwel in Seir, and the Moabites,
that abide in Ar: vntil we come to Iordan, and paſſe to the Land, which
the Lord our God wil geue vs.
%%% 0424
\V And Sehon the king of Heſebon would not geue vs paſſage: becauſe the
Lord thy God had
\SNote{God permitted him for his ſinnes, to indurate him ſelfe. See
\XRef{Exod.~7. v.~3.}}
indurated his ſpirit, and hardened his hart, that he might be deliuered
into thy handes, as now
thou ſeeſt. \V And our Lord ſaid to me: Behold I haue begunne to
deliuer Sehon vnto thee, and his land, beginne to poſſeſſe it. \V And
Sehon came forth to meete vs with al his people to battel in Iaſa. \V
And the Lord our God deliuered him to vs: and we ſmote him with his
ſonnes and al his people. \V And al his cities we tooke at that time,
killing the inhabiters therof, men and wemen and litle ones. We left
nothing among them. \V Except the cattel, which came to their portion
that tooke prayes: and the ſpoyles of the cities, which we tooke \V from
Aroer, which is ſituated vpon the banke of the torrent Arnon, a towne
that is ſituated in a valley, as farre as Galaad. There was not a
village or citie, that eſcaped our handes: the Lord our God deliuered al
vnto vs. \V Except the land of the children of Ammon, to the which we
approched not: and al that adioyne to the torrent Ieboc, and the cities
on the mountaine, and al the places from which the Lord our God
prohibited vs.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
