%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 25
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-25


\project douay-rheims


%%% 0466
%%% o-0419
\startChapter[
  title={Chapter 25}
  ]

\Summary{Punishment afflicted according to the fault, but ſo that he
  which is beaten haue not aboue fourtie ſtripes. 4.~The oxes mouth not
  be mooſeled that treadeth corne. 5.~A maried man dying without iſsue,
  his brother muſt marie the widow. 11.~The wife that taketh her
  huſbands aduerſarie by priuates muſt loſe her hand. 13.~No falſe
  weightes, nor meaſures to be kept. 17.~Amelicites muſt be vtterly
  deſtroyed.}

If there be a controuerſie betwen ſome, and they cal vpon the iudges:
whom they shal perceiue to be iuſt, to him they shal geue the price of
iuſtice: whom impious, him they shal condemne of impietie. \V And if
they ſee that the offendeer be worthie of ſtripes: they ſhal caſt him
downe, & shal cauſe him to be beaten before them. According to the
meaſure of the ſinne shal the meaſure alſo of the ſtripes be: \V yet ſo,
that they exceede not the number of fourtie: leſt thy brother depart
ſowly torne before thyne eies. \V Thou shalt
\SNote{S.~Paul expoundeth this of the ſpiritual laborer in Gods Church;
that he muſt haue his maintenãce for his trauel.
\XRef{1.~Cor.~9.}
&
\XRef{1.~Tim.~5.}
It was alſo ment of oxen, ſo it hath two literal ſẽſes.
\Cite{Theod. q.~31. in Deut.}}
not mooſel the
%%% 0467
mouth of the oxe that treadeth out thy corne in the flore. \V When
brethren shal dwel together, & one of them die without children, the
wife of the deceaſed ſhal not marie to an other: but
\LNote{His brother shal take her.}{This
\MNote{Mariage with the brothers wife, he dying without iſſue.}
proueth euidently that the prohibition, not to marie the brothers wife
\XRef{(Leuit.~18.)}
was a poſitiue law, binding only when the firſt brother dying left
iſſue. For dying without iſſue, his brother was bound by this law to
marie the widow. In default of the brother, the next of kinne was to
marie her: and for default of neerer, the more remote. So
\CNote{Ruth.~3. &.~4.}
Booz maried Ruth. Neither was it contrarie, but agreable to the law of
nature, to marie the brothers wife when he was dead without iſſue, as is
before noted.
\XRef{Gen.~38.}}
his brother ſhal take her, and rayſe vp the ſeede of his brother: \V and
the firſt borne of her he shal cal by his name, that his name be not
abolished out of Iſrael. \V But if he wil not take his brothers wife,
that by law is dew to him, the woman shal goe to the gate of the citie,
and cal vpon the ancientes, and ſay: My husbandes brother wil not rayſe
vp his ſeede in Iſrael: nor take me to his wife. \V And forthwith they
shal cauſe him to be ſent for, and shal aske him. If he anſwer: I wil
not take her to wife: \V the woman shal come to him before the
ancientes, and shal take of his shoe from his foote, and
\SNote{He that diſdaineth to honour his brother is iuſtly deſpiſed.}
ſpitte in his face, and ſay: So shal it be done to the man, that
buildeth not his brothers houſe. \V And his name shal be called in
Iſrael
%%% !!! Too long? Make into an LNote?
\SNote{A laſie familie & vnprofitable to the cõmõwealth. Myſtically,
Paſtors and Doctors muſt beget ſpiritual children to Chriſt, not to them
ſelues; & ſo they are called Chriſtians, not Paulians, whom S.~Paul
conuerted. And he that is elected by the church to ſpiritual functiõ, &
neglecteth his dutie, is worthie of reproch and infamie.
\Cite{S.~Aug. li.~32. c.~10. cont. Fauſt. Manich.}}
The houſe of the vnshodde. \V If two men fal at wordes betwixt them
ſelues, and one beginne to brawle againſt the other, and the wife of the
one willing to deliuer her husband out of the hand of the
%%% o-0420
ſtronger, put forth her hand, and take his priuities: \V thou shalt
cutte of her hand, neither shalt thou be moued with any pitie vpon
her. \V Thou shalt not haue diuerſe weightes in thy bagge, a greater and
a leſſe: \V neither shal there be in thy houſe a greater bushel and a
leſſe. \V Thou shalt haue a weight iuſt and true, and thy bushel shal be
equal and true: that thou mayeſt liue a long time vpon the Land, which
our Lord thy God shal geue thee. \V For thy Lord abhorreth him, that
doth theſe thinges, and deteſteth al iniuſtice. \V Remember what
\SNote{Amalec firſt impugned Iſrael after they had paſſed the red ſea.
\XRef{Exo.~17.}}
Amalec did to thee in the way when thou cameſt out of Ægypt: \V how he
meete thee in the way when thou cameſt out of Ægypt: \V how he mette
thee: & ſtroke the hindemoſt of thy armie, which being wearie reſted
them ſelues, when thou waſt ſpent with famine and labour, and he feared
not God. \V Therfore when our Lord thy God shal geue thee reſt, and
ſubdewe al nations round about in the Land, which he hath promiſed thee:
thou shalt deſtroy his name vnder heauen. Beware thou forget it not.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
