%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 0439
%%% o-0394
\startChapter[
  title={Chapter 10}
  ]

\Summary{Moyſes receiuing the ſecond tables of the tenne commandments,
  and making an arke put them therin. 6.~With mention of certaine places
  where the children of Iſrael had camped, of Aarons death, and to the
  Leuites offices, and poſseſsions, 12.~he inculcateth the feare and
  loue of God, and the keping of his precepts. 16.~Namely to circumciſe
  the hart, 19.~to loue ſtrangers 20.~and not to ſerue, nor ſweare by
  falſe goddes.}

At that time our Lord ſaid to me: Hewe thee two tables of ſtone, as the
former were, and come vp to me into the mounte: and thou shalt make an
arke of wood, \V and I wil write in the tables the wordes that were in
them, which before thou didſt breake, and thou shalt put them in the
arke. \V I made therfore an arke of the wood Settim. And when I had
hewed two tables of ſtone like to the former, I went
vp into the mount, hauing them in my handes. \V And
%%% 0440
he wrote in the tables, according as he had written before, the ten
wordes, which our Lord ſpake to you in the mount from the middes of the
fyre, when the people was gathered: and he gaue them to me. \V And
returning from the mount, I came downe, and put the tables into the
arke, that I had made, which are there til this preſent, as our Lord
commanded me. \V And the children of Iſrael remoued their campe from
Beroth of the children of Iacan into
\SNote{This Moſera where Aaron died, is more commonly
called \Emph{Hor}.
\XRef{Num.~10.}
&
\XRef{33.}}
Moſera, where Aaron died and was buried, for whom, Eleazar his
\Fix{ſoone}{ſonne}{likely typo, same in both}
did the function of prieſthood. \V Thence they came into Gadgad: from
the which place departing, they camped in Ietebatha, in a Land of waters
and torrentes. \V At that time he ſeparated the tribe of Leui, to carie
the arke of the couenant of our Lord, and to ſtand before him in the
miniſterie, and to bleſſe in his name vntil this preſent day. \V For the
which cauſe Leui had no part, nor poſſeſſion with his brethren: becauſe
our Lord him ſelf is his poſſeſſion, as the Lord thy God promiſed
him. \V And I ſtoode in the mount, as before, fourtie daies and nightes:
and our Lord heard me this time alſo, and would not deſtroy thee. \V And
he ſaid to me: Goe, and march before the people, that they may enter,
and poſſeſſe the Land,
%%% o-0395
which I ſware to their fathers that I would deliuer to them. \V And now
Iſrael, what doth the Lord thy God require of thee, but that thou feare
the Lord thy God, and walke in his waies, and loue him, and ſerue the
Lord thy God with al thy hart, and with al thy ſoule: \V and keepe the
commandementes of our Lord, and his ceremonies, which I command thee
this day, that it may be wel with thee? \V Behold heauen is the Lords
thy God, and the heauen of heauen, the earth and al thinges that are in
it. \V And yet to thy fathers was our Lord ioyned, and he loued them,
and choſe their ſeede after them, that is to ſay you, from al Nations,
as this day it is proued. \V Circumciſe therfore the prepuce of your
hart, and your necke indurate no more: \V becauſe the Lord your God he
is the God of goddes, and the Lord of lordes, a great God and mightie,
and terrible, that accepteth not perſon nor giftes. \V He doth iudgement
to the pupil and the widowe, loueth the ſtranger, and geueth him victual
& rayment. \V And do you therfore loue ſtrangers becauſe you alſo were
ſtrangers in the Land of Ægypt. \V Thou ſhalt feare the Lord thy God,
and ſerue him only: to him thou
%%% 0441
ſhalt cleaue, and
\SNote{VVhen iuſt cauſe requireth an oath, it muſt be made in the name
of God, not of falſe goddes.}
shalt ſweare in his name. \V He is thy praiſe, and thy God, that hath
done for thee theſe greate and terrible thinges, which thyne eies haue
ſeene. \V In ſeuentie ſoules did thy fathers goe downe into Ægypt: and
behold now the Lord thy God hath multiplied thee as the ſtarres of
heauen. 


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
