%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-22


\project douay-rheims


%%% 0460
%%% o-0414
\startChapter[
  title={Chapter 22}
  ]

\Summary{Pietie towardes neighboures. 5.~Neither ſexe may vſe the
  apparel of the other. 6.~Crueltie to be auoided euen towardes
  birdes. 8.~Batlement about the roofe of a houſe. 9.~Things of diuers
  kindes not to be mixed. 12.~Cordes in the
  \Fix{hemes}{hemmes}{possible typo, fixed in other}
  of a cloke. 13.~Trail
  and punishment of adulterie and of deflowring virgines. 30.~The ſonne
  may not marie his ſtepmother.}

%%% 0461
Thou ſhalt not ſee thy brothers oxe, or ſheepe ſtraying, and paſſe by:
but ſhalt bring it backe to thy brother, \V although thy brother be not
nigh, and thou know him not: thou shalt bring them vnto thy houſe, and
they ſhal be with thee vntil thy brother ſeeke them, and receiue
them. \V In like manner ſhalt thou doe with his aſſe, and with his
rayment, and with euerie thing of thy brothers, that shal be loſt: if
thou finde it, neglect it not as perteyning to an other. \V If thou ſee
thy brothers aſſe or oxe to be fallen in the way, thou shalt not
contemne it, but shalt lift it vp with him. \V A woman ſhal not be
clothed with mans apparel, neither ſhal a man vſe womans apparel: for he
is abominable before God that doeth theſe thinges. \V If walking by the
way thou finde a birdes neſt in a tree or on the ground, and the damme
ſitting vpon the young or the egges: thou shalt not hold her with her
young, \V but shalt let her goe, taking the young and holding them: that
it may be wel with thee, and thou mayeſt liue a long time. \V When thou
buildeſt a new houſe, thou shalt make
\SNote{Their houſes had flat roofes as manie of our churches, pallaces,
and caſtles, where battlements are neceſſarie for danger of falling
when, anie walke theron.}
a batlement to the roofe round about: leſt bloud be ſhed in thy houſe,
and thou be guiltie an other ſlipping, and falling headlong. \V Thou
shalt not ſowe thy vineyard with diuerſe ſeede: leſt both the ſeede
which thou didſt ſow, and the thinges that grow of the vineyard,
\SNote{For correction of ſo couetous a mind the whole fruicte muſt be
offered to pious vſes.
\Cite{Theodoret. q.~23. in Deut.}}
be ſanctified together. \V Thou ſhalt not plough with an oxe and aſſe
together. \V Thou ſhalt not weare a garment that is wouen of wolle and
linnen. \V Thou shalt make litle cordes in the hemme at the foure
corners of thy cloke, wherwith thou shalt be couered. \V If a man marrie
a wife, and afterward hate her, \V and ſeeke occaſions to put her away,
obiecting vnto her a verie il name, and ſay: I tooke this wife, and
compayning with her: I found her not a virgin: \V her father and mother
shal take her, and shal carie with them the ſignes of her virginitie to
the ancientes of the citie that are in the gate: \V and the father shal
ſay: I gaue my daughter vnto this man to wife: whom becauſe he
hateth, \V he layeth vnto her a verie il name, ſo that he ſayeth: I
found not thy daughter a virgin: and behold theſe are the ſignes of my
daughters virginitie. They shal ſpread the veſture
%%% o-0415
before the ancientes of the citie: \V and the ancientes of that citie
shal take the man, and beate him, \V condemning him beſides in a hundred
ſicles of ſiluer, which he shal geue to
%%% 0462
the wenches father, becauſe he hath infamouſly ſpred a verie il name vpon
a virgin of Iſrael: and he shal haue her to wife, and can not put her
away al the daies of his life. \V But if it be true which he obiected,
and virginitie be not found in the wench: \V they shal caſt her forth
without the doores of her fathers houſe, and the men of her citie shal
ſtone her to death, and she ſhal die: becauſe ſhe hath done wickednes in
Iſrael, to fornicate in her fathers houſe: and thou shalt take away the
euil out of the middes of thee. \V If a man lie with an
\Fix{others}{other}{possible typo, same in both}
mans wife, both shal die, that is to ſay, the aduouterer and the
aduoutereſſe: and thou shalt take away the euil out of Iſrael. \V If a
man haue deſpouſed a maide that is a virgine, and ſome man finde her in
the citie, and lie with her, \V thou shalt bring
\Fix{forh}{forth}{obvious typo, fixed in other}
both of them to the gate of that citie, and they shal be ſtoned: the
maide, becauſe she cried not, being in the citie: the man, becauſe he
hath humbled his neighbours wife, and thou shalt take away the euil from
the middes of thee. \V But if the man finde the maide that is deſpouſed,
in the field, and taking her, lie with her, he alone shal die: \V the
maide shal ſuffer nothing, neither is she guiltie of death: for as a
theefe ryſeth againſt his brother, and taketh away his life, ſo alſo did
the maide ſuffer. \V She was alone in the field: she cried, and there
was no man to deliuer her. \V If a man finde a maide that is a virgin,
which hath not a ſpouſe, and taking her lie with her, and the matter
come into iudgement: \V he that lay with her, shal geue to the father of
the maide fiftie ſicles of ſiluer, and shal haue her to wife, becauſe he
hath humbled her: he can not put her away al the daies of his life. \V
No man shal take his fathers wife, nor reuele his couering.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
