%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 0443
%%% o-0397
\startChapter[
  title={Chapter 12}
  ]

\Summary{Al idolatrie, and whatſoeuer apperteineth therto muſt be
  deſtroyed. 5.~Sacrifices, tithes, and donaries muſt be offered in the
  ſpecial place. 15.~Eating flesh they muſt not eate the bloud. 29.~In
  no caſe to imitate the idolatrie of gentiles.}

These are the preceptes and iudgementes, that you muſt do in the Land,
which the Lord God of thy fathers wil geue thee, to poſſeſſe it al the
daies, that thou shalt goe vpon the earth. \V Subuert al places, wherein
the nations, which you shal poſſeſſe, worshipped their goddes vpon the
high mountaines, and hilles, and vnder euerie tree ful of leaues. \V
Ouerthrow their altares, and breake their ſtatues, their groues burne
with fire, and their Idols hewe al to peeces: deſtroy their names out of
thoſe places. \V You shal not doe ſo to the Lord your God: \V but
\SNote{Peculiar place appropriate to Gods ſeruice.}
to the place, which the Lord your God hath choſen of al your tribes, to
put his name there and to dwel in it, shal you come: \V and shal offer
in that place your holocauſtes and victimes, the tithes and firſt
fruites of your handes, and your vowes and donaries, the firſt borne of
your oxen and sheepe. \V And you shal eate there in the ſight of the
Lord your God: and you shal reioyce in al thinges, whereunto you shal
put your hand, you and your houſe, wherein the Lord your God hath
bleſſed you. \V You shal not doe there the thinges, that we doe here
this day
\SNote{In the deſert they could not obſerue the ceremonies of the Law:
but comming to reſt they were bound to kepe al one ſette forme of holie
rites.}
euerie man that which ſeemeth good to him ſelf. \V For vntil this
preſent time you are not come to reſt, and to the poſſeſſion, which the
Lord your God wil geue you. \V You shal paſſe ouer Iordan, and shal dwel
in the Land, which the Lord your God wil geue you, that you may haue
reſt from al enemies round about: and may dwel without al feare, \V in
the place, which the Lord your God shal chooſe, that his name may be
therin. Thither shal you bring al the thinges, that I command you,
holocauſtes,
%%% o-0398
and hoſtes, and tithes, and the firſt fruites of your handes: and whatſoeuer
is the principal in the giftes, that you shal vowe to our Lord. \V There
%%% 0444
shal you feaſte before the Lord your God, you and your ſonnes and
daughters, men ſeruantes and wemen ſeruantes, and the Leuite, that
dwelleth in your cities, for he hath no other part and poſſeſſion among
you. \V Beware thou offer not thy holocauſtes in euerie place, that thou
shalt ſee: \V but in that, which our Lord shal chooſe, in one of thy
tribes shalt thou offer hoſtes, and shalt doe what thinges ſoeuer I
command thee. \V But if thou wilt eate, and the eating of flesh delight
thee, kil, and eate according to the bleſſing of the Lord thy God, which
he hath geuen thee in thy cities: whether it be vncleane, that is to
ſay, blemished and feeble: or cleane, that is to ſay, ſound and without
blemish, ſuch as is lawful to be offered, as the doa and the hart, shalt
thou eate it, \V only without eating of the bloud, which thou shalt
power out vpon the earth as water. \V Thou canſt not eate in thy townes
the tithe of thy corne, and wine, and oyle, the firſt borne of thy
heardes and cattel, and al thinges that thou voweſt, and that thou wilt
offer voluntarily, and the firſt fruites of thy handes: \V but before
the Lord thy God shalt thou eate them in the place, which the Lord thy
God shal chooſe, thou and thy ſonne and thy daughter, and man ſeruant,
and woman ſeruant, and the Leuite, that dwelleth in thy cities: and thou
shalt reioyce and be refreshed before the Lord thy God in al thinges,
whereunto thou shalt extend thy hand. \V Take heede thou forſake not the
Leuite al the time that thou liueſt in the land. \V When the Lord thy
God shal haue dilated thy borders, as he hath ſpoken to thee, and thou
wilt eate the flesh, that thy ſoule deſireth: \V and if the place be
farre of, which the Lord thy God shal chooſe, that his name may be
there, thou shalt kil of the heardes and cattel, which thou haſt as I
haue commanded thee, and shalt eate in thy townes, as it pleaſeth
thee. \V As the doa is eaten and the hart, ſo shalt thou eate them: both
the cleane and vncleane shal eate in common. \V This onlie beware, that
thou eate not the bloud, for their bloud is for the ſoule: and therfore
thou muſt not eate the ſoule with the flesh: \V but vpon the earth thou
shalt power it as water, \V that it may be wel with thee and thy
children after thee, when thou shalt doe that which pleaſeth in the
ſight of our Lord. \V But the thinges which thou haſt ſanctified, and
vowed to our Lord, thou shalt take vp, and shalt come to the place,
which
%%% 0445
our Lord shal chooſe: \V and shalt offer thy oblations the flesh and the
bloud vpon the altar of the Lord thy God: the bloud of thy hoſtes thou
shalt power on the altar: and the flesh thy ſelf shalt eate. \V Obſerue
and heare al thinges that I command thee, that it may be wel with thee
and thy children after thee for euer, when thou shalt doe that which is
good and pleaſing in the ſight of the Lord thy God. \V When the Lord thy
God shal haue deſtroyed before thy face the nations, that thou entreſt
in to poſſeſſe, and thou shalt poſſeſſe them, and dwel in their land: \V
beware leſt thou imitate them,
%%% o-0399
after they be ſubuerted at thy entring in, and thou require their
ceremonies, ſaying: As theſe nations haue worshipped their goddes, ſo
wil I alſo worshippe. \V Thou shalt not doe in like maner to the Lord
thy God. For al the abominations, that our Lord doeth abhorre, haue they
done to their goddes, offering their ſonnes and daughters, and burning
them with fyre. \V What I command thee,
\LNote{That only do to our Lord.}{VVheras
\MNote{No hoſtes lawful in ſacrifice but ſuch as the law appointed.}
the Gentiles offered their ſonnes and daughters
\XRef{(v.~31.)}
and other abominable ſacrifices to Idols, God commandeth his people to
offer thoſe things only, which are preſcribed by the law, and neither to
imolate anie other thing, nor exclude anie thing appointed by the ſame
law for ſacrifice. As for other preceptes, it is likewiſe forbid to adde
or diminiſh anie thing that may corrupt the law:
\MNote{New precepts may be added, not contrarie to the former.}
but was euer lawful for Superiours, to adde more preceptes agreable, and
not contrarie to the former. So King Dauid eſtabliſhed a new law that
ſuch as ſtaued with the baggage, ſhould haue like portion of the praye,
with thoſe that fought in battel.
\XRef{1.~Reg.~30.}
And our Sauiour by his preſence
\XRef{(Ioan~10.)}
approued the feaſt of dedication, inſtituted long after Moyſes law.
\XRef{1.~Machab.~4.}}
that onlie doe to our Lord: neither adde any thing, nor diminish.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
