%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 0459
%%% o-0412
\startChapter[
  title={Chapter 21}
  ]

\Summary{How to ſeke out a ſecrete murtherer. 10.~Wemen taken in battel
  may be maried, and afterwardes can not be ſold nor made
  bondwemen. 15.~The eldeſt ſonne may not be depriued of his birthright
  for hatred of his mother. 18.~A ſtubburne ſonne muſt be ſtoned to
  death. 22.~VVhen one is hanged on a gibbet, he muſt be taken downe the
  ſame day, and buried.}

VVhen there shal be found in the Land, which our Lord thy God wil geue
thee, the corps of a man ſlaine, and he that is guiltie of the murder is
not knowne \V thy ancientes, and iudges shal goe forth, and meaſure
from the place of the corps the diſtance of euerie citie round about: \V
and which they ſhal perceiue to be neerer then the reſt,
%%% o-0413
the ancientes of that citie shal take an heifer out of the heard, that
hath not drawen yoke, nor ploughed the ground, \V and shal bring her to
a rough and ſtonie valley, that neuer was ploughed, nor receiued ſeede:
and in it they shal ſtrike of the necke of the heifer: \V and the
prieſtes the ſonnes of Leui shal come, whom our Lord thy God hath choſen
to miniſter to him, and to bleſſe in his name, and at their word euerie
matter dependeth, and whatſoeuer is cleane or vncleane muſt be
iudged. \V And the ancientes of that citie shal come to the ſlaine
perſon, and shal wash their handes ouer the heifer, that was ſtrooken in
the valley, \V and shal ſay: Our handes did not ſheede this bloud,
\SNote{By this ceremonie and abiuration they purged them ſelues, that
they were not negligent in doing iuſtice.}
nor our eies
\Fix{ſeee}{ſee}{obvious typo, fixed in other}
it. \V Be merciful to thy people Iſrael, whom thou haſt redemed O Lord,
and impute not innocent bloud in the middes of thy people Iſrael. And
the guilte of bloud shal be taken from them: \V and thou shalt be free
from the innocents bloud, that was ſhed, when thou shalt haue done that
which our Lord hath commanded thee. \V If thou goe forth to fight
againſt thyne enemies, and our Lord thy God deliuer them in thy hande and
thou leade them away captiue, \V and ſeeſt in
%%% 0460
the number of the captiues a beautiful woman, and loueſt her, and wilt
haue her to wife, \V thou ſhalt bring her into thy houſe: who ſhal shaue
of her hare, and payre her nailes, \V and put of the rayment, wherein
she was taken: and ſitting in thy houſe, shal mourne her father and
mother one moneth: and afterward thou shalt enter vnto her, and ſhalt
ſleepe with her, and ſhe ſhal be thy wife. \V But if afterward she
content not thy mynde, thou shalt let her goe free, neither canſt thou
ſel her for money, nor oppreſſe her by might: becauſe thou haſt humbled
her. \V If a man haue two wiues, one beloued, and the other hated, and
they haue begotten children by him, and the ſonne of the hated be the
firſtborne, \V and he meaneth to diuide his ſubſtance among his ſonnes:
he can not make the ſonne of the beloued the firſt borne, and preferre
him before the ſonne of the hated, \V but the ſonne of the hated he shal
acknowledge for the firſt borne, and shal geue to him of thoſe thinges,
which he hath, al duble: for this is the beginning of his children, & to
this are dew the firſt
\Fix{brith}{birth}{obvious typo, same in both}
rightes. \V If a man begette a ſtubbourne and froward ſonne, that wil
not heare the commandementes of his father and mother, and being
chaſtened, contemneth to be obedient: \V they shal take him, and bring
him to the ancientes of his citie, and to the gate of iudgement, \V and
shal ſay to them: This our ſonne is froward and ſtubborne, he contemneth
to heare our admonitions, he geueth himſelf to comeſſation, and to
ryote and bankeringes: \V the people of the citie shal ſtone him: and he
shal die, that you may take away the euil out of the middes of you, and
al Iſrael hearing it may be afrayde. \V When a man hath offended ſo that
he is to be punished by death, and being condemned to die is hanged on a
gybbet: \V his bodie shal not remaine vpon the tree, but the ſame day
shal be buried: becauſe he is
\SNote{Myſtically, he is curſed that perſiſteth in ſinne, as it were
hanging on the tree, by which our firſt parents ſinned.}
accurſed of God that hangeth on a tree: and thou shalt not contaminate
thy Land, which our Lord thy God geueth thee in poſſeſſion.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
