%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 0456
%%% o-0410
\startChapter[
  title={Chapter 19}
  ]

\Summary{Certaine cities of refuge muſt be aſsigned for caſual
  manſlaughter. 11.~Wilful murther punished by death without remiſsion,
  15.~ſo it be conuinced by two or three witneſſes. 16.~Falſe witneſſes
  punished with the paine, which the crime obiected deſerueth.}

VVhen our Lord thy God hath deſtroyed the nations, whoſe land he wil
deliuer to thee, and thou doeſt poſſeſſe it, and dwelleſt in the cities and
houſes therof: \V three cities shalt thou ſeparate to thee in the middes
of the Land, which our Lord thy God wil geue thee in poſſeſſion, \V
\SNote{The way to the cities of refuge were paued, and markes ſet for
direction, that he which fled might not erre in his way.}
preparing diligently the way: and thou ſhalt diuide the whole prouince
of thy Land equally into three partes: that he which for murder is a
fugitiue, may haue neere at hand, whither to eſcape. \V This shal be the
law of the murderer that fleeth, whoſe life is to be ſaued: He that
ſtriketh his neighbour vnwitting, and that is proued yeſterday and the
day before to haue had no hatred againſt him: \V but to haue gone with
him ſimply vnto the wood to cutte wood, and in cutting
\Fix{the wood of}{of wood the}{obvious typo, fixed in other}
axe ſlipped out of his hand, and the yron falling from the handle ſtrooke
his freind, and killed him: he shal flee to one of the cities aforeſaid,
and liue: \V leſt perhaps the next kinſeman of him, whoſe bloud was
shed, pricked with ſorow, purſew, and apprehend him if the way be too
long, and ſtricke his life, that is not guiltie of death, becauſe he is
proued to haue had no hatred before, againſt him that was ſlaine. \V
Therfore I command thee, that thou ſeparat three cities of equal
diſtance one from an other. \V And when our Lord thy God shal haue
dilated thy borders, as he ſware to thy fathers, and shal geue thee al
the Land, that he promiſed them, (\V yet ſo, if thou keepe his
commandementes, and do the thinges which I command thee this day, that
thou loue our Lord thy God, and walke in his wayes at al time) thou
shalt added to thee other three cities, and shalt duble the number of
the three cities aforeſaid: \V that innocent
%%% 0457
bloud be not shed in the middes of the Land, which our Lord thy God wil
geue thee to poſſeſſe, leſt thou be guiltie of bloud. \V But if any man
hating his neighbour, lie in wayte for his life, and riſe and ſtrike
him, and he die, and he flee to one of the cities aforeſaid, \V the
ancientes of his citie shal ſend,
\Fix{acd}{and}{obvious typo, fixed in other}
take him out of the place of refuge, and shal deliuer him into the hand
of the kinſeman of him, whoſe bloud was ſhed, and he shal die. \V
\SNote{This was ſayd to the whole people who muſt not intreate for the
murderers pardon: but the kinſmen of him that was ſlaine, might remitte
the puniſhment.}
Thou shalt not pitie him, and thou ſhalt take away the guiltie bloud out
of Iſrael, that it may be
%%% o-0411
wel with thee. \V Thou shalt not take, and
transferre thy neighbours boundes, which thy predeceſſours did ſette in
thy poſſeſſion, which our Lord thy God wil geue thee in the Land, that
thou shalt receiue to poſſeſſe. \V One witneſſe shal not ſtand againſt
any man, whatſoeuer ſinne, or wickedneſſe it be: but in the mouth of two
or three witneſſes shal euerie word ſtand. \V If a lying witneſſe ſtand
againſt a man, accuſing him of preuarication, \V both of them, whoſe the
cauſe is, shal ſtand before our Lord in the ſight of the prieſtes and
the iudges that shal be in thoſe daies. \V And when ſearching moſt
diligently, they shal finde that the falſe witnes hath ſaid a lie
againſt his brother: \V they shal render to him as he meant to doe to
his brother, and thou shalt take away the euil out of the middes of
thee: \V that others hearing may haue feare, and may not be bold to doe
ſuch things. \V
\SNote{This pertained to the Iudge: who without partialitie muſt do
iuſtice.}
Thou shalt not pitie him, but life for life, eie for eie, tooth for
tooth, hand for hand, foote for foote shalt thou exacte.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
