%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 26
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-26


\project douay-rheims


%%% 0468
%%% o-0420
\startChapter[
  title={Chapter 26}
  ]

\Summary{Firſt fruites muſt be offered in ſpecial place aſsigned to Gods
  ſeruice, profeſſing of gratitude for the land poſseſſed according to
  Gods promiſe. 12.~Likewiſe tithes of the third yeare, 16.~with
  concluſion, that the people promiſe to obſerue al the preceptes of
  God, and ſo doing he wil protect and proſper them.}

And when thou art entred into the Land, which our Lord thy God wil geue
thee to poſſeſſe, and haſt obteyned it, and dwelleſt in it: \V thou
ſhalt take firſt of al thy fruites, and put them in a maunde, and ſhalt
goe to the place, which our Lord thy God ſhal chooſe, that his name may
be inuocated there: \V and thou shalt goe to the prieſt, that shal
%%% o-0421
be in thoſe daies, and ſay to him: I profeſſe this day before our Lord
thy God, that I am entred into the Land, for the which he ſware to our
fathers, that he would geue it vs. \V And the prieſt taking the maund at
his hand, shal ſette it before the altar of our Lord thy God: \V and
thou shalt ſpeake in the ſight of our Lord God:
\SNote{Laban purſued Iacob, when he parted from Meſopothamia of Syria.
\XRef{Gen.~27.}}
The Syrian perſecuted my father, who deſcended into Ægypt, and ſeiourned
there in a verie ſmale number and grew into a nation greate and ſtrong
and of an infinite multitude. \V And the Ægyptians afflicted vs, and
perſecuted vs laying on moſt grieuous burdens: \V and we cried to our
Lord the God of our fathers: who heard vs, and reſpected our affliction,
and labour, and diſtreſſe: \V and brought vs out of Ægypt in a ſtrong
hand, a ſtretched out arme, in great terrour, in ſignes and wonders: \V
and brought vs into this place, and deliuered to vs this Land flowing
with milke and honie. \V And therfore now I offer firſt fruites of the
Land, which our Lord hath geuen me. And thou shalt leaue them in the
ſight of our Lord thy God, adoring our Lord thy God. \V And thou shalt
feaſt in al the good thinges, which our Lord thy God hath geuen to thee,
and thy houſe, thou and the Leuite, and the ſtranger that is with
thee. \V When thou haſt finiſhed the tithe of al thy fruites, in
\SNote{The people payed euerie yeare two tithes: firſt to the Leuites:
the ſecond for entertaining trauelers to & from Ieruſalẽ & euerie third
yeare, a third tith for relief of the poore inhabitantes.}
the third yeare of tithes thou shalt geue to the Leuite, and the
%%% 0469
ſtranger, and the pupil, and the widow, that they may eate within thy
gates, and be filled: \V and thou ſhalt ſpeake in the ſight of our Lord
thy God: I haue brought that which is ſanctified out of my houſe, and
haue geuen it to the Leuite and the ſtranger, and the pupil and the
widow, as thou haſt commanded me: I haue not tranſgreſſed thy
commandements nor forgotten thy preceptes. \V I haue not eaten of them
in my mourning, nor ſeparated them in any vncleannes, nor ſpent of them
any thing in funerals. I haue obeyed the voice of our Lord my God, and
haue done al thinges as thou didſt command me. \V Looke from thy
ſanctuarie, and thy high habitation of heauen, and bleſſe thy people
Iſrael, and the Land, which thou haſt geuen vs, as thou ſwareſt to our
fathers, a land flowing with milke and honie. \V This day our Lord thy
God hath commanded thee to doe theſe commandmentes and iudgementes: that
thou keepe and fulfil them with al thy hart, and with al thy ſoule. \V
\SNote{Mutual pact betwen God & his people; that they ſeruing him, he
wil reward them.}
Thou haſt choſen our Lord this day, to be thy God, and to walke in his
waies, and keepe his ceremonies, and preceptes and iudgementes, and obey
his commandement. \V And our Lord hath choſen thee this day, that thou
shouldeſt be his peculiar people, as he hath ſpoken to thee, and thou
shouldeſt keepe al his commandementes: \V and make thee higher then al
nations which he created, to his prayſe, and name, and glorie: that thou
mayeſt be a holie people of our Lord thy God, as he hath ſpoken.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
