%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 33
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-33


\project douay-rheims


%%% 0485
%%% o-0436
\startChapter[
  title={Chapter 33}
  ]

\Summary{Moyſes bleſsing the tribes of Iſrael (Simeon omitted)
  prophecyeth particularly of euerie one. 26.~Againe exhorteth them,
  that as God hath choſen them his peculiar people, ſo they loue and
  honour him their onlie God.}

This is
\SNote{The ancient fathers expound theſe bleſſinges rather of the Church
of Chriſt, then of the Iewes Synagogue.
\Cite{S.~Aug. q.~56.}
\Cite{Theod. q.~44. in Deut.}}
the bleſſing, wherwith Moyſes the man of God bleſſed the children of
Iſrael, before his death.

\V And he ſaid: Our Lord
\LNote{Came from Sinai.}{According to the hiſtorie Moyſes recounteth
here three benefites. Firſt that God gaue the Law in Sinai.
\XRef{(Exod.~20.)}
Secondly, he cured thoſe which were bitten with ſerpentes nere to Seir.
\XRef{(Num.~21.)}
Thirdly in mount Pharan he appointed Seuentie ancientes to aſſiſt Moyſes
in iudgementes.
\XRef{(Num.~11.)}
\MNote{The prophetical ſenſe of theſe bleſsĩges is more certaine, & more
euident, then the hiſtorical.}
But according to the Myſterie, which ſpecially is intended, S.~Auguſtin
\Cite{(q.~56. in Deut.)}
ſaieth this prophecie is not to be negligently paſſed ouer. For it
euidently appeareth that this benediction perteineth to a new people,
whom Chriſt our Lord hath ſanctified, in whoſe perſon Moyſes ſpake and
not in his owne. So in this prophetical and proper ſenſe (ſaieth this
Doctor) our Lord and Sauiour cometh from \Emph{Sinai}, which is
interpreted \Emph{tentation}, when he paſſed the tentation of his
paſſion and death,
\XRef{Heb.~2. v.~18.}
Chriſt riſeth from \Emph{Seir}, interpreted
\Fix{\Emph{hearie},}{\Emph{hairie},}{possible typo, fixed in other}
for that in the ſimilitude of the fleſh of ſinne,
\Fix{enen}{euen}{obvious typo, fixed in other}
of ſinne, he damned ſinne in the fleſh.
\XRef{Rom.~8. v.~3.}
He appeareth from mount \Emph{Pharan} interpreted \Emph{fruitful
mountaine}, in that he geueth abundance of grace in his Church of the
new Teſtament, which is a citie ſet vpon a hil.
\XRef{Mat.~5.}}
came from Sinai, and from Seir is he riſen to vs: he hath appeared from
mount Pharan, and with him thouſãdes of Sainctes. In his right hand a
fyrie law.

\V He hath loued the peoples, al the ſaintes are in his hand: and they
that approch to his feete, shal receiue of his doctrine.

\V Moyſes commanded vs a law, the inheritance of the multitude of Iacob.

%%% 0486
\V He shal be king with the moſt right, the princes of the people being
aſſembled with the tribes of Iſrael.

\V Liue Ruben, and die he not, and be he litle in number.

\V This is the bleſſing of Iudas: Heare Lord the voice of Iudas, and
bring him in vnto his people: his handes shal fight for him, and he shal
be his helper againſt his aduerſaries.

\V To Leui alſo he ſaid: Thy perfection, and thy doctrine be to thy
holie man, whom thou haſt proued in tentation, and iudged at the Waters
of contradiction. \V He that
\SNote{The prieſtly tribe muſt eſpecially preferre Gods ſeruice before
their neereſt kinred.}
ſaid to his father, and to his mother: I know you not; and to his
brethren: I know you not: & they knew not their children. Theſe kept thy
word, and obſerued thy couenant, \V thy iudgementes O Iacob, and thy law
O Iſrael: they shal put incenſe in thy furie, and holocauſt vpon thyne
altar. \V Bleſſe Lord his ſtrength, and receiue the workes of his
handes. Strike the backes of his enemies, and they that hate him, let
them not riſe vp.

\V And to Beniamin he ſaid: The beſt beloued of our Lord
\SNote{The Temple was built in the tribe of Beniamin, which God more
ſpecially protected, and ſo they dwelt more ſecurely.
\Cite{Theod. q.~45. in Deut.}}
shal dwel confidently in him: as in a bride chamber al the day shal he
abide, and betwen his shoulders shal he reſt.

%%% o-0437
\V To Ioſeph alſo he ſaid: Of the bleſſing of our Lord be his land, of
the fruites of heauen, and the dew, & the depth lying vnderneth. \V Of
the pomes of the fruites of the ſunne and moone, \V of the toppes of the
old mountaynes, of the pomes of the eternal hilles: \V and of the
fruites of the earth, and of the fulnes therof. The bleſſing of him,
that appeared in the bush, come vpon the head of Ioſeph, and vpon the
crowne of the nazarite among his brethren. \V His beautie as of the
firſt borne of an oxe, his hornes the hornes of an vnicorne: in them
shal he winow the Nations euen to the endes of the earth. Theſe are the
multitudes of
\SNote{Ephraim is preferred before his elder brother, agreable to their
granfathers prophetical bleſſing.
\XRef{Gen.~48.}}
Ephraim, and theſe the thouſandes of Manaſſes.

\V And to Zabulon he ſaid: Reioyſe Zabulon in thy going out, and
Iſſachar in thy tabernacles. \V They shal cal the peoples to the
mountaine: there shal they immolate the victimes of iuſtice. Who shal
ſucke the inundation of the ſea as milke, and the hidden treaſures of
the ſandes.

\V And to Gad he ſaid: Bleſſed be Gad in breadth: as a lion hath he
reſted, and taken the arme and the toppe of the head. \V And he ſaw his
principalitie, that in his part the doctor was repoſed: which was with
the princes of the people,
%%% 0487
and did the iuſtices of our Lord, and his iudgement with Iſrael.

\V To Dan alſo he ſaid: Dan a lions whelpe, he ſhal flow largely
from Baſan.

\V And to Nephthali he ſaid: Nephthali shal enioy abundance, and ſhal be
ful of the bleſſinges of our Lord: the ſea and the ſouth he shal
poſſeſſe.

\V To Aſer alſo he ſaid: Bleſſed be Aſer in children, be he acceptable
to his brethren, and dippe he in oile his foote. \V His ſhoe yron and
braſſe. As the daies of thy youth, ſo alſo thy old age.

\V There is no other God as the God of the righteſt: the mounter of
heauen is thy helper. By his magnificence the cloudes runne hither and
thither, \V his habitation is aboue, and vnder the euerlaſting armes: he
shal caſt out the enemie from thy face, and shal ſay: Be deſtroyed.

\V
\SNote{The ſinne of Zãbri a prince of Simeons tribe, in freſh memorie
\XRef{(Nu.~25.)}
ſemeth to be the cauſe, why this tribe is not
\Fix{partilary}{particularly}{obvious typo, fixed in other}
bleſſed, but only in general with al Iſrael.}
Iſrael shal dwel confidently, and alone. The eie of Iacob in the land of
corne and wine, and the heauens shal be miſtie with dew.

\V Bleſſed art thou Iſrael: who is like to thee O people, that art ſaued
in our Lord? the ſhield of thy helpe, and the ſword of thy glorie: thy
enemies shal denie thee, and thou shalt treade their neckes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
