%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 0448
%%% o-0402
\startChapter[
  title={Chapter 15}
  ]

\Summary{Remiſsion of debtes in the ſeuenth yeare to the Iſraelites, but
  not to ſtrangers. 4.~Albeit there wil alwayes be ſome poore, yet they
  muſt ſo lend to their needie bretheren, that none be forced to
  begge. 12.~A bought ſeruant that is an hebrew muſt be ſet free in the
  ſeuenth yeare, 16.~except he deſire to ſerue ſtil. 19.~The firſtborne
  in al cattel muſt be conſecrated to God, without making priuate
  profite therof.}

In the ſeuenth yeare thou shalt make a remiſſion, \V which shal be
celebrated in this order. He to whom any thing is owing of his freind or
neighbour and brother, can not aske it againe, becauſe it is the
yeare of remiſſion of our Lord. \V Of the ſeiourner and ſtranger thou
shalt exact: of thy
%%% 0449
countrie man and neighbour thou shalt not haue power to require it. \V
And
\SNote{The Iſraelites were boũd to do their endeuour that none ſhould be
needie among them: notwithſtanding for exerciſe of loue & charitie Gods
prouidence ſuffered ſome to be poore.
\XRef{v.~7. &~11.}}
needie perſon and begger there shal be none among you: that our Lord thy
God may bleſſe thee in the land, which he wil geue thee in
poſſeſſion. \V Yet ſo if thou heare the voice of our Lord thy God, and
keepe al thinges that he hath bid, and which I command thee this day, he
wil bleſſe thee, as he hath promiſed. \V Thou shalt lend to manie
nations, and thy ſelfe shalt borrow of
%%% o-0403
no man. Thou shalt haue dominion ouer verie manie nations, and no man
shal haue dominion ouer thee. \V If one of thy brethren that abideth
within the gates of thy citie in the land, which our Lord thy God wil
geue thee, come to pouertie: thou shalt not harden thy hart, nor cloſe
thy hand, \V but shalt open it to the poore man, and shalt lend him,
that which thou perceiueſt he hath neede of. \V Beware leſt perhaps an
impious cogitation ſteale in vpon thee, and thou ſay in thy hart: The
ſeuenth yeare of remiſſion draweth nigh, & turne away thy eies from thy
poore brother denying to lend him that which he asketh: leſt he crie
againſt thee to our Lord, and
\SNote{He that can and wil not feede his neighboure in extremitie,
killeth him.
\Cite{S.~Amb. li.~2. de Offic. c.~7.}}
it become a ſinne vnto thee. \V But thou shalt geue to him: neither
shalt thou doe any thing craftely in releuing his neceſſities: that our
Lord thy God may bleſſe thee at al times, and in al thinges whereunto
thou shalt put thy hand. \V There shal not want poore in the land of thy
habitation: therfore I command thee that thou open thy hand to thy
needie and poore brother, that liueth in the Land. \V When thy brother
an Hebrew man, or Hebrew woman is ſold to thee, and hath ſerued thee ſix
yeares, in the ſeuenth yeare, thou shalt let him goe free: \V and to
whom thou geueſt freedom, thou ſhalt in no caſe ſuffer him to depart
emptie: \V but geue him his wayfare of thy flockes, and of thy barne
floore, and thy preſſe, wherwith our Lord thy God ſhal bleſſe thee. \V
Remember that thy ſelf alſo didſt ſerue in the Land of Ægypt, and our
Lord thy God made thee free, and therfore doe I now command thee. \V But
if he ſay: I wil not depart: becauſe he loueth thee, and thy houſe, and
feeleth that he is wel with thee: \V thou ſhalt take an awle, and bore
through his eare in the dore of thy houſe, and he ſhal ſerue thee for
euer. To thy woman ſeruant alſo thou ſhalt doe in like maner. \V Turne
not away thine eies from them, when thou makeſt them free: becauſe he
hath ſerued thee ſix yeares after the wages of an
%%% 0450
hireling: that our Lord thy God may bleſſe thee in al the workes that
thou doeſt. \V Of the firſt borne, that come forth in thy heardes and
ſheepe, whatſoeuer is of the male ſexe, thou ſhalt ſanctifie to our Lord
thy God. Thou shalt not worke with the firſt borne of an oxe, and thou
shalt not sheare the firſt borne of thy ſheepe. \V In the ſight of our
Lord thy God shalt thou eate them euerie yeare in the place, that our
Lord shal chooſe, thou and thy houſe. \V But if it haue blemish, and be
either lame, or blind, or in any part disfigured or feeble, it shal not
be immolated to our Lord thy God: \V but within the gates of thy citie
shalt thou eate it: as wel the cleane as the vncleane in like maner shal
eate them as the doa, and the hart. \V This onlie shalt thou
obſerue, that their bloud thou eate not, but power it out on the earth
as water.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
