%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 30
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-30


\project douay-rheims


%%% 0477
%%% o-0428
\startChapter[
  title={Chapter 30}
  ]

\Summary{If the children of Iſrael, offending and falling into the
  forſaide curſes shal repent, God wil reſtore them to his bleſsings
  againe: 11.~leauing it in their powre to ſerue him if they wil,
  17.~and therfore warneth them that the impenitent shal aſſuredly
  perish, becauſe hauing life and death, bleſsing and curſing propoſed,
  they chooſe the worſe.}

Therfore when al theſe wordes shal be come vpon thee, the bleſſing or
curſing, which I haue ſette forth before thee: & thou be touched with
repentance of thy hart in al nations, into which our Lord thy God
diſperſed thee, \V and ſhalt returne to him, and obey his commandmentes,
as I this day command thee, with thy children, in al thy hart,
%%% o-0429
and in al thy ſoule: \V our Lord thy God wil bring thee againe from thy
captiuitie, and haue mercie vpon thee, and gather thee againe out of al
the peoples, into which he diſperſed thee before. \V If thou be
diſperſed as farre as the poles of heauen, thence wil our Lord thy God
draw thee backe, \V and wil take thee to him, and bring thee into the
Land, which thy fathers poſſeſſed, and thou shalt obteyne it: and
bleſſing thee, wil make thee to be
\SNote{Some ſinners through great repentance become more vertuous, and
are more rewarded then ſome that offended leſſe.}
of a greater number, then were thy fathers. \V Our Lord thy
\LNote{God vvil circumciſe thy hart.}{Moſt
\MNote{By grace men are made able to kepe Gods cõmandments.}
true it is, that of our ſelues, without Gods grace none can kepe or
fulfil the commandmentes. But he, whoſe hart God doth circumciſe, is
therby made able to loue God with al his hart, and with al his
ſoule. And except ſome hartes were thus circumciſed, and ſo made able to
loue God aboue al, and conſequently their neighboures, God ſhould not
performe his promiſe, that he wil circumciſe the hart of ſome.}
God wil circumciſe thy hart, and the hart of thy ſeede: that thou mayeſt
loue
\Fix{oure}{our}{likely typo, fixed in other}
Lord thy God in al thy hart, and in al thy ſoule, that thou mayeſt
liue. \V And al theſe curſes he wil turne vpon thine enemies, and them
that hate and perſecute thee. \V But thou ſhalt returne, and heare the
voice of our Lord thy God, and shalt doe al the commandmentes which I
command thee this day: \V and our Lord wil make thee abound in al
%%% 0478
the workes of thy handes, in the iſſue of thy wombe, and in the fruite
of thy cattel, in the fertilitie of thy ground, and in the plentie of al
thinges. For our Lord wil returne to reioyce vpon thee in al riches, as
he reioyced in thy fathers: \V yet ſo, if thou heare the voice of our
Lord thy God, and keepe his preceptes and ceremonies, which are written
in this law: and returne to our Lord thy God in al thy hart, and in al
thy ſoule. \V This commandment, that I command thee this daie,
\LNote{Is not aboue thee.}{VVhen
%%% !!! Where do these CNotes really go? They spread along all four
%%% LNotes.
\CNote{S.~Aug. donat. et grat. c.~69. et q.~54 in Deut.}
\CNote{Theod. q.~38. in Deut.}
\CNote{S.~Cypri. li.~3. c.~52. ad Quir.}
\CNote{S.~Amb. in Pſal.~40.}
\MNote{So the commandmentes are not impoſſible.}
thou art ſtirred vp, aſſiſted, and indued with Gods grace, the
commandment of God \Emph{is not} (then) \Emph{aboue thee}, nor farre of
from thee, but very nere thee, in thy mouth, (to confeſſe God, and his
truth) and in thy hart, to do it. But you wil aske: How then cometh it
to paſſe, that manie hauing receiued ſufficient grace, yet do not kepe
Gods commandmentes? God him ſelf anſwereth:
%%% !!! See next LNote
That he hath}
is not aboue thee, nor ſo farre of, \V nor ſituated in heauen that thou
maieſt ſay: Which of vs is able to aſcend vnto heauen to bring it to vs,
that we may heare and fulfil it in worke? \V Nor placed beyond the ſea:
that thou mayeſt pretend, and ſay: Which of vs can paſſe ouer the ſea,
and bring it euen vnto vs: that we
\Fix{way}{may}{obvious typo, fixed in other}
heare, and doe that which is commanded? \V But the word is very neere
thee, in thy mouth and in thy hart, to doe it. \V Conſider that I haue
%%% !!! The words "That he hath" that are now part of the previous
%%% LNote, really belong at the beginning of this one, but then they
%%% should be in upright, not italic. How should we handle this? (Plus,
%%% they then mess up the fact that every other LNote starts with a
%%% quote from the text to help with indexing.)
\LNote{ſet before thee life and good, and contrarivviſe death and
euil;}{he
\MNote{Freewil.}
inuiteth and helpeth, yet forceth thee not: he geueth thee powre &
abilitie, helping and not deſtroying thy freewil, that thou maieſt loue
our Lord thy God, walke in his wayes, and kepe his commandements. But if
thy hart be auerted and
\XRef{(v.~17.)}
\Emph{thou vvilt not heare}, thou ſhalt periſh. Againe God inculcateth:}
ſet before thee this day life and good, and contrariewiſe death and
euil: \V that thou mayeſt loue our Lord thy God, and walke in his waies,
and keepe his commandmentes and ceremonies and iudgementes: and thou
mayeſt liue, and he multiplie thee, and bleſſe thee in the Land, which
thou ſhalt enter to poſſeſſe. \V But if thy hart be auerted, and thou
wilt not heare, and deceaued with errour thou adore ſtrange goddes, and
ſerue them: \V I foretel thee this day that thou ſhalt perish, and abide
a litle time in the Land, which paſſing ouer Iordan, thou shalt enter to
poſſeſſe. \V
\LNote{I cal for vvitneſſes heauen and earth,}{that I
haue \Emph{propoſed} to you life and death, bleſſing and
curſing. \Emph{Chooſe} therfore life &c. VVhat Doctor can teach more
plainly the poſſibilitie of keeping Gods commandmentes: and frewil in
man, then this text of holie Scripture?}
I cal for witneſſes this day heauen and earth, that I haue propoſed to
you life and death, bleſſing and curſing.
\SNote{God gaue man libertie to chooſe, what he would folow.
\Cite{S.~Amb. in Pſal.~40. v.~10.}}
Chooſe therfore life, that both thou mayeſt liue, and thy ſeede: \V and
mayeſt loue our Lord thy God, and obey his voice, and cleaue to him (for
he is thy life, and the length of thy daies) that thou mayeſt liue in
the Land, for the which our Lord ſware to thy fathers, Abraham, Iſaac,
and Iacob, that he would geue it them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
