%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 27
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-27


\project douay-rheims


%%% 0469
%%% o-0422
\startChapter[
  title={Chapter 27}
  ]

\Summary{Gods commandmentes muſt be written in plaſtred ſtones. An
  Altar erected, and Sacrifices offered. 12.~Obſeruers of the
  commandments muſt be bleſsed, and tranſgreſſors curſed. 14.~With the
  forme of curſing idolaters, and diuers other enormious ſinners.}

And
\MNote{The third part.

Gods promiſes & threates, for keeping or breaking his commandments.}
Moyſes and the ancientes of Iſrael commanded the people, ſaying: Keepe
euerie commandment that I command you this day. \V And when you are
paſſed ouer Iordan into the Land, which our Lord thy God wil geue thee,
thou ſhalt erect great ſtones, and ſhalt polish them with plaſter, \V
that thou mayeſt write on them al the wordes of this law, when thou haſt
paſſed ouer Iordan: that thou mayeſt enter into the Land, which our Lord
thy God wil geue thee, a land flowing with milke and honie, as he ſware
to thy
%%% 0470
fathers. \V When therfore you are paſſed Iordan, erect the ſtones which
I command you this day in mount Hebal, and thou ſhalt poliſh them with
plaſter: \V and thou ſhalt build there an altar to our Lord thy God of
ſtones, which yron hath not touched, \V and of ſtones not
fashioned nor polished: and thou shalt put vpon it holocauſtes to our
Lord thy God, \V and shalt immolate pacifique hoſtes, and eate there,
and feaſte there before our Lord thy God. \V And thou shalt write vpon the
ſtones al the wordes of this law plainely and cleerly. \V And Moyſes and
the prieſtes of the Leuitical ſtocke ſaid to al Iſrael: Attend, and
heare Iſrael: This day thou art made the people of our Lord thy God: \V
thou shalt heare his voice, and doe the commandementes and iuſtices,
which I command thee. \V And Moyſes commanded the people in that day,
ſaying: \V
\SNote{The ancientes of euerie tribe.}
Theſe shal ſtand to bleſſe the people, vpon mount Garizim, when you are
paſt Iordan: Simeon,
\SNote{The Leuites proper office was to bleſſe.}
Leui, Iudas, Iſſachar, Ioſeph, and Beniamin. \V And ouer againſt them
theſe shal ſtand to curſe on mount Hebal. Ruben, Gad, and Aſer, and
Zabulon, Dan and Nephthah. \V And the
\SNote{But by occaſion of ſinne their office was alſo to pronounce
curſes.}
Leuites shal pronounce, and ſay to al the men of Iſrael with a high
voice: \V Curſed be the man that maketh a grauen and molten thing, the
abomination of our Lord, the worke of the handes of artificers, and shal
put it
\SNote{Though the ſinnes were ſecrete, yet the offenders were curſed:
publique ſinnes were alſo publikely puniſhed.}
in ſecrete: and al the people shal anſwer, and ſay: Amen. \V Curſed be
he that honoureth not his father and mother: and al the people shal ſay:
Amen. \V Curſed be he that remoueth his neighbours boundes: and al the
people shal ſay: Amen. \V Curſed be he that maketh the blinde to goe
amiſſe in his iourney: and al the people shal ſay: Amen. \V Curſed be he
that peruerteth the iudgement of the ſtranger, of the
%%% o-0423
pupil and the widow: and al the people shal ſay: Amen. \V Curſed be he
that ſleepeth with his fathers wife, and reuealeth the couer of his
bedde: and al the people shal ſay: Amen. \V Curſed be he that lyeth with
any beaſt: and al the people shal ſay: Amen. \V Curſed be he that
ſleepeth with his ſiſter, the daughter of his father, or of his
mother: and al the people shal ſay: Amen. \V Curſed be he that ſleepeth
with his mother in law: and al the people shal ſay: Amen. \V Curſed be
he that ſecretely ſtriketh his neighbour: and al the people shal ſay:
Amen. \V Curſed be he that taketh giftes, to kil the ſoule of innocent
bloud: and al the people shal ſay:
%%% 0471
Amen. \V Curſed be he that obydeth not in the wordes of this law, and
fulfilleth them not in worke: and al the people shal ſay: Amen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
