%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 0432
%%% o-0388
\startChapter[
  title={Chapter 6}
  ]

\Summary{God is diligently to be ſerued, and loued with thy whole hart,
  thy whole ſoule, and whole ſtrength. Al his precepts, ceremonies and
  iudgements muſt be carefully kept, and commended to poſteritie.}

Theſe are the preceptes, and ceremonies, and iudgementes, which the Lord
your God commanded that I should teach you, and you should doe them in
the Land, whereunto you paſſe ouer to poſſeſſe it: \V that thou mayeſt
feare the Lord thy God, and keepe his commandementes and preceptes,
which I command thee, and thy ſonnes, and nephewes, al the dayes of thy
life, that thy dayes may be prolonged. \V Heare Iſrael, and obſerue that
thou doe the thinges which our Lord hath commanded thee, and it may be
wel with thee, and thou mayeſt be greately multiplied, as the Lord God
of thy fathers hath promiſed thee a land
\Fix{flowig}{flowing}{obvious typo, fixed in other}
with milke and honie. \V Heare Iſrael, The Lord our God, is one Lord. \V
Thou shalt loue the Lord thy God with thy whole hart, and with thy whole
ſoule, and with thy whole
%%% 0433
ſtrength. \V And theſe wordes, which I command thee this day, shal be in
thy hart: \V and thou shalt tel them to thy children, and thou shalt
meditate ſitting in thy houſe, and walking on thy iourney, ſleeping,
and ryſing. \V And thou shalt bind them as a ſigne on thy hand, and they
shal be & shal moue betwen thine eies, \V and thou shalt write them in
the entrie, and on the doores of thy houſe. \V And when the Lord thy God
shal haue brought thee into the Land, for the which he ſware to thy
fathers Abraham, Iſaac, and Iacob: and shal haue geuen thee great and
goodlie cities, which thou didſt not build, \V houſes ful of al riches,
which thou didſt not erect, ceſternes which thou didſt not digge,
vineyardes and oliueyardes, which thou didſt not plant, \V and thou
shalt haue eaten and be ful: \V take heede diligently leſt thou forget
our Lord, that brought thee out of the Land of Ægypt, out of the houſe
of ſeruitude. Thou shalt feare the Lord thy God, and
\SNote{Some adoration agreeth to creatures, but ſeruice of Latria to God
onlie.
\Cite{S.~Aug. q.~61. in Gen.}}
him onlie shalt thou ſerue, and by his name shalt thou ſweare. \V You
shal not goe after the ſtrange goddes of al Nations, that are round
about you: \V becauſe the Lord thy God is a Iealouſe God in the middes of
thee: leſt ſometime the furie of the Lord thy God be wrath againſt thee,
and take thee away from the face of the earth. \V Thou shalt not tempt
the Lord thy God, as thou didſt tempt him in the place of tentation. \V
Keepe the preceptes of the Lord thy God, and the teſtimonies and
ceremonies, which he hath commanded thee: \V And doe that which is
pleaſant
%%% o-0389
and good in the ſight of our Lord, that it may be wel with thee: and
entring in thou mayeſt poſſeſſe the goodlie Land, wherof our Lord ſware
to thy fathers, \V that he would deſtroy al thy enemies before thee, as he
hath ſpoken. \V And when thy ſonne shal aske thee to morrow, ſaying:
What meane theſe teſtimonies, and ceremonies, and iudgementes, which the
Lord our God hath commanded vs? \V thou shalt ſay to him: We were the
bondmen of Pharao in Ægypt, and our Lord brought vs out of Ægypt in a
ſtrong hand: \V and he did ſignes & wonders great and verie ſore in
Ægypt againſt Pharao, and al his houſe, in our ſight, \V and he brought
vs out from thence, that being brought in he might geue vs the Land,
wherupon he ſware to our fathers. \V And our Lord commanded that we
should doe al theſe ordinances, and should feare the Lord our God, that
it might be wel with vs
%%% 0434
al the daies of our life, as it is at this day. \V And he wil be
merciful to vs, if we keepe and doe al his preceptes before the Lord our
God, as he commanded vs.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
