%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 32
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-32


\project douay-rheims


%%% 0481
%%% o-0432
\startChapter[
  title={Chapter 32}
  ]

\Summary{A Canticle of the Law, wherin the people are exhorted to ſerue
  God, for his perfect goodnes, for his ſingular benefites, for their
  former ingratitude, and for his mercie ſtil mixed with his
  punishmentes. 44.~Al which being earneſtly commended to them to
  remember and teach their children, 48.~Moyſes is commanded to goe into
  a mountaine, whence he shal ſee the promiſed land, but not enter into
  it.}

Heare ye
\SNote{Al thinges in heauen and in earth teſtifie, that God dealeth wel
with his people.}
heauens what thinges I ſpeake, the earth heare the wordes of my mouth.

\V My
\SNote{Doctrine doth fructifie in good ſoules as raine & dew in the
ground.}
doctrine grow together as raine, my ſpeach flow as the dew, as it were a
ſhower vpon the herbe, and as it were droppes vpon the graſſe.

%%% 0482
\V
\SNote{Mans firſt dutie is to praiſe God.}
Becauſe I wil inuocate the name of our Lord: geue magnifence to our God.

\V The workes of God be perfect, and al his waies iudgementes: God is
faithful, and without any iniquitie, iuſt and right.

%%% o-0433
\V They
\SNote{The next, to acknowledge his owne ſinnes & defectes.}
haue ſinned to him, and not his children in filthines: a froward and
peruerſe generation.

\V Theſe thinges doeſt thou render to our Lord thou fooliſh and vnwiſe
people? Is not he thy father, that hath poſſeſſed thee, and made, and
created thee?

\V Remember the old daies, thinke vpon euerie generation: aske thy
father, and he wil declare to thee: thy elders, and they wil tel thee.

\V When the higheſt
\SNote{At the towre of Babel.}
diuided the nations: when he ſeparated the ſonnes of Adam, he appointed
the limites of people according to
\SNote{Iſrael being but one people poſſeſſed the inheritance of ſeuen
other nations.}
the number of the children of Iſrael.

\V But our Lords part, is his people: Iacob the corde of his
inheritance.

\V He
\SNote{God
\Fix{choiſe}{choſe}{obvious typo, fixed in other}
Iſrael to be his peculiar people of mere grace and protected them.}
found him in a deſert land, in a place of horrour, and of waſt
wildernes: he ledde him about, and taught him: and kept him as the apple
of his eye.

\V As the eagle prouoking her young to flie, and houering ouer them,
hath he ſpred his winges, and he hath taken him, and caried him on his
shoulders.

\V Our Lord onlie was his guide: and there was not with him a ſtrange
God.

\V He placed him ouer an high land: that he might eate the fruites of
the fieldes, that he might ſucke
\SNote{Bees without mens induſtrie made honie in the rockes.}
honie out of the rocke, and
\SNote{Oliue trees proſpered in ſtonie places.}
oile out of the hardeſt ſtone.

\V Butter from the heard, and milke of the ſheepe with the fatte of
lambes, and of rammes the ſonnes of Baſan: and bucke goates with the
marow of wheate, and might drinke the bloud of the grape moſt pure.

\V The
\SNote{Temporal proſperitie occaſion of the Iewes reuolting from God.}
beloued was made groſſe, and ſpurned: made groſſe, fatted, dilated, he
left God his maker, and departed from God his ſaluation.

\V They prouoked him in ſtrange goddes, and in abominations ſtirred him
to anger.

\V They immolated to diuels and not to God, to goddes, which they knew
not: there came
\SNote{Noueltie allureth carnal people to idolatrie and hereſie.}
new & freſh ones, whom their fathers worshipped not.

%%% 0483
\V God that begatte thee thou haſt forſaken, and haſt
\LNote{Forgotten our Lord.}{Caluin
\MNote{Caluin contradicteth the holie Scripture.}
\Cite{(li.~1. Inſtit. c.~11. parag.~9.)}
contending that it is idolatrie, to worſhip Chriſt in his picture,
ſaieth the idolatrie committed about the molten calfe
\XRef{(Exod.~32.)}
conſiſted in worſhipping the true God in that image of a calfe:
affirming that \Emph{the Ievves vvere not ſo inconſiderate, as not to
remember, that is vvas God vvhich had brought them out of the Land of
Ægypt.} Quite contrarie to this text, which ſaieth: God that begate thee
thou haſt forſaken, and \Emph{haſt forgotten} our Lord thy
creator. Caluin therfore was either ignorant or forgetful, that the
Holie Ghoſt here chargeth them, to haue forſaken and forgotten God the
Creator; or els (which is worſe) knowing and remembring it, he was moſt
impudent in auouching the contrarie.}
forgotten our Lord thy creatour.

\V Our Lord ſaw, and was moued to wrath: becauſe his ſonnes and
daughters prouoked him.

\V And he ſaid:
\SNote{For their peruerſnes God withdrew his helpe from them.}
I wil hide my face from them, and wil conſider their laſt: for it is a
peruerſe generation, and vnfaithful children.

\V They
\SNote{God firſt loueth, before anie man loueth him, but men firſt
forſake God, before he forſake them.}
haue prouoked me in that, which was no God, and haue angred me in their
vanities: and I wil prouoke them in that, which is no people, and in
\SNote{The Iewes reputed moſt Gẽtiles fooliſh yet now they are inferior
to al.}
a foolish nation wil I anger them.

%%% o-0434
\V A fyre is kindled in my wrath, and shal burne euen to the loweſt
partes of hel: and shal deuoure the earth with her ſpring, and shal
burne the foundations of mountaines.

\V I wil heape euils vpon them, and myne arrowes I wil ſpend in them.

\V They shal be conſumed with famine, and birdes shal deuoure them with
moſt bitter biting: the teeth of beaſtes wil I ſend vpon them, with the
furie of thoſe that traile vpon the ground, and creepe.

\V Without shal the ſword deuoure them, & within fearefulnes, the young
man and the virgin together, the ſucking child with the old man.

\V I ſaid: Where are they? I wil make their memorie to ceaſe from among
men.

\V But
\SNote{For iuſt cauſes God ſome times differreth puniſhment.}
for the wrath of the enemies I haue differred: leſt perhaps their
enemies might be proude, and would ſay: Our mightie hand, and not the
Lord, hath done al theſe thinges.

\V A nation without counſel is it, and without wiſedome.

\V O that they were
\SNote{True wiſdome \Emph{conſidereth} thinges
paſt, \Emph{vnderſtandeth} things preſent, and \Emph{prouideth} for
things to come.}
wiſe, and vnderſtoode, and would prouide for their laſt.

\V How should one purſew a thouſand, and two put ten thouſand to flight?
was it not therfore, becauſe their God ſold them, and our Lord incloſed
them?

\V For our Lord is not as their goddes:
\SNote{Al infidels confeſſe more Maieſty in the true God and in his
Religiõ, then in their owne.}
our enemies alſo are iudges.

\V Of the vineyard of Sodom, is their vineyard, and of the ſuburbes of
Gomorrha: their grape the grape of gall, and the cluſters moſt bitter.

\V The gall of dragons their wine, and the venime of Aſpes vncurable.

%%% 0484
\V Are not theſe thinges laid vp with me, and ſigned in my treaſures?

\V Reuenge is myne, and I wil repay them in time, that their foote may
ſlide: the day of perdition is at hand, and the times make haſt to be
preſent.

\V Our Lord wil iudge his people, and wil haue mercie on his ſeruantes:
he shal ſee that their hand is weakened, and
\SNote{Euen ſuch offenders as thinck them ſelues ſecure, eſcape not.}
the shut vp alſo haue fayled, and the reſidew be conſumed.

\V And he shal ſay: Where are their goddes, in whom they had confidence.

\V Of whoſe victimes they did eate fatte, and dranke the wine of their
libamentes:
\SNote{It is vnpoſſible that falſe goddes ſhould helpe their folowers in
neceſſitie.}
let them ariſe, and help you, and protect you in neceſſitie.

\V See ye that I am onlie, and there is no other God beſides me: I wil
kil, and I wil make to liue: I wil ſtrike, and I wil heale, and there is
none that can deliuer out of my hand.

%%% o-0435
\V I wil lift vp my hand to heauen, and wil ſay: I liue for euer.

\V If I shal whette my ſword as the lightening, and my hand take
iudgement: I wil repay vengeance to myne enemies, and them that hate me
wil I requite.

\V I wil embrew my arrowes with bloud, and my ſword ſhal deuoure flesh,
of the bloud of the ſlaine and of captiuitie, of
\SNote{The vaine counſel of the wicked being detected ſhal be puniſhed.}
the bare head of the enemies.

\V You gentiles
\LNote{Praiſe his people.}{In
\MNote{Alwaies ſome good in the Church of the old Teſtamẽt.}
the people of the Iewes were alwayes ſome good, & worthie of praiſe,
that ſerued God; and ſometimes ſuffered perſecution for iuſtice; whoſe
bloud God promiſed here to reuenge, and for their ſakes to be merciful
to others. Al which we ſee was performed, in that God reduced them from
captiuitie, and conſerued them after in their countrie til Chriſts time.
For ſo he was \Emph{propitious to the land of his people}.}
prayſe his people, becauſe he wil reuenge the bloud of his ſeruantes:
and wil repay vengeance vpon their enemies, and wil be propitious to the
land of his people.

\V Moyſes therfore came and ſpake al the wordes of this canticle in the
eares of the people, he and Ioſue the ſonne of Nun. \V And he finished
al theſe wordes, ſpeaking to the children of Iſrael. \V And he ſaid to
them: Sette your hartes on al the wordes, which I teſtifie to you this
day: that you command them to your children to keepe and to doe, and to
fulfil al thinges of this law that are written: \V for not in vaine are
they commanded you, but that euerie one should liue in them: which doing
you may continew a long time in the Land, which paſſing ouer Iordan you
enter to poſſeſſe. \V And our Lord ſpake to Moyſes in the ſame day,
ſaying: \V Goe vp into this mount Abirim, that is to ſay, of paſſages,
into mount Nebo, which is in the Land of Moab againſt Iericho: and ſee
the Land of Chanaan, which I wil deliuer to
%%% 0485
the children of Iſrael to obteyne, and die thou in the mount. \V Which
going vp vnto thou shalt be ioyned to thy peoples, as Aaron thy brother
died in mount Hor, and was layd to his people: \V becauſe you did
preuaricate againſt me in the middes of the children of Iſrael at the
Waters of contradiction in Cades of the deſert of Sin: and you did not
ſanctifie me among the children of Iſrael. \V Ouer againſt shalt thou
ſee the Land, and shalt not enter into it, which I wil geue to the
children of Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
