%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 31
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-31


\project douay-rheims


%%% 0479
%%% o-0430
\startChapter[
  title={Chapter 31}
  ]

\Summary{Moyſes ſubſtituteth Ioſue his ſucceſſour in temporal
  gouernment, 9.~deliuereth the law to the Prieſtes. 16.~God fortelleth
  that the people wil often  forſake him, and that he wil punish
  them. 19.~Commandeth Moyſes to write a canticle, (an abrigement of the
  Law) eaſie to be remembred, 25.~and in further teſtimonie againſt
  them, the Leuites muſt put this booke in the arke of couenant.}

Moyſes
\MNote{The fourth part.

An exhortation to ſerue God, with predictiõ of their often ſinnes and
puniſhmentes.}
therfore went, and ſpake al theſe wordes to al Iſrael, \V and
ſaid to them: I am this day a hundred and twentie yeares old, I can not
\SNote{He meaneth that he cã not exerciſe the office of a captaine
general, and bring the people into the promiſed land.}
goe out and come in any longer, eſpecially wheras our Lord alſo hath
ſaid to me: Thou ſhalt not paſſe ouer this Iordan. \V Our Lord therfore
thy God wil paſſe ouer before thee: he wil deſtroy al theſe nations in
thy ſight, and thou ſhalt poſſeſſe them: and
%%% o-0431
this Ioſue ſhal paſſe ouer before thee, as our Lord hath ſpoken. \V And
our Lord ſhal doe to them as he did to Sehon and Og the kinges of the
Ammorheites, and to their land, and ſhal deſtroy them. \V Therfore when
our Lord ſhal haue deliuered theſe alſo to you, you shal doe in like
manner to them as I haue commanded you. \V Doe manfully, and be
ſtrengthned: feare not, neither tremble ye at their ſight: becauſe our
Lord thy God him ſelfe is thy conductor, and wil not leaue, nor forſake
thee. \V And Moyſes called Ioſue, and ſaid to him before al Iſrael: Take
courage, and be ſtrong: for thou ſhalt bring in this people into the
Land, which our Lord ſware that he would geue to their fathers, and thou
ſhalt diuide it by lotte. \V And our Lord that is your conductor, him
ſelfe
%%% 0480
wil be with thee: he wil not leaue, nor forſake thee: feare not, neither
dread thou. \V Moyſes therfore wrote this law, and deliuered it to the
prieſtes the ſonnes of Leui, which caried the arke of the couenant of
our Lord, and to al the ancientes of Iſrael. \V And he commanded them,
ſaying: After ſeuen yeares, in the yeare of remiſſion, in the ſolemnitie
of tabernacles, \V when al come together out of Iſrael, to appeare in
the ſight of our Lord thy God in the place, which our Lord ſhal chooſe,
thou shalt read the wordes of this law before al Iſrael, they
hearing, \V and the people being aſſembled together, as wel men as
wemen, children, and ſtrangers, that are within thy gates: that hearing
they may learne, and feare our Lord your God, and keepe, and fulfil al
the wordes of this law. \V Their children alſo who now are ignorant,
that they may heare, and feare our Lord their God, al the daies that
they liue in the Land, which paſſing ouer Iordan you goe to obteyne. \V
And our Lord ſaid to Moyſes: Behold the daies of thy death are nigh: cal
Ioſue, and ſtand ye in the tabernacle of teſtimonie, that I may command
him. Moyſes therfore and Ioſue went, and ſtoode in the tabernacle of
teſtimonie, \V and our Lord appeared there in the piller of a cloude:
which ſtood in the entring of the tabernacle. \V And our Lord ſaid to
Moyſes: Behold thou ſhalt ſleepe with thy fathers, and this people
ryſing vp wil fornicate after ſtrange goddes in the Land, to the which
it entreth to dwel therein: there wil they forſake me, and wil make the
couenant, which I haue made with them, of none effect. \V And my furie
shal be wrath againſt them in that day: and I wil forſake them, and wil
hide my face from them, and they shal be deuoured: al euils and
afflictions shal finde them, ſo that they shal ſay in that day: In truth
becauſe God is not with me, theſe euils haue found me. \V But I wil
hide, and keepe cloſe my face in that day, for al the euils, which they
haue done, becauſe they haue folowed ſtrange goddes. \V Now therfore
write vnto you
\SNote{Meeter is more eaſily kept in memorie then proſe.}
this canticle, and teach the children of Iſrael: that they know it by
hart, and ſing it by mouth, and this ſong be vnto me for
\SNote{And ſo by this Canticle they are conuinced that they were
abundantly forwarned, not to breake couenãt with God.}
a teſtimonie among the children of Iſrael. \V For I wil bring them into
the Land, for the which I ſware to their fathers, flowing with milke and
honie. And when they haue eaten, and are ful, and fatte, they wil
turne away to ſtrange goddes, and ſerue them: and wil detract
%%% 0481
%%% o-0432
from me, and make my couenant of none effect. \V After that manie euils
and afflictions shal haue found them, this canticle shal anſwer them for
a teſtimonie, which no obliuion shal take away out of the mouth of their
ſeede. For I know their cogitations, what thinges they are about to doe
this day, before that I bring them into the Land, which I haue promiſed
them. \V Moyſes therfore wrote the canticle, and taught it the children
of Iſrael. \V And our Lord commanded Ioſue the ſonne of Nun, and ſaid:
Take courage, and be ſtrong: for thou shalt bring the children of Iſrael
into the Land, which I haue promiſed, and I wil be with thee. \V
Therfore after that Moyſes wrote the wordes of this law in a volume, and
finished it: \V he commanded the Leuites, that caried the arke of the
couenant of our Lord, ſaying: \V Take this booke, and put it in the ſide
of the arke of the couenant of our Lord your God: that it may be for a
teſtimonie againſt thee. \V For I know thy contention, and thy moſt
ſtiffe necke. Whiles I yet liue and goe in with you, you haue done
alwayes contenciouſly againſt our Lord: how much more when I shal be
dead? \V Gather to me al the ancientes by your tribes, and your doctors,
and I wil ſpeake theſe wordes in their hearing, and wil inuocate againſt
them heauen and earth. \V For I know that after my death you wil doe
wickedly, and wil decline quickly from the way, that I haue commanded
you: and euils ſhal come vpon you in the later times, when you ſhal doe
euil in the ſight of our Lord, to prouoke him by the workes of your
handes. \V Moyſes therfore ſpake, in the hearing of the whole aſſemblie
of Iſrael, the wordes of this ſong, and finiſhed it euen to the end.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
