%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 29
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-29


\project douay-rheims


%%% 0475
%%% o-0427
\startChapter[
  title={Chapter 29}
  ]

\Summary{A couenant and oath is made betwen God and his people (with
  commemoration of ſundrie benefites by them receiued) that keping his
  law they shal be more bleſſed: and breaking the ſame shal ſuſteine
  the threatned punishmentes.}

Theſe are the wordes of the couenant which our Lord commanded Moyſes to
make with the children of Iſrael in the Land of Moab: beſide that
couenant which he made with them in Horeb. \V And Moyſes called al
Iſrael, and ſaid to them: You ſaw al thinges, that our Lord did before
you in the Land of Ægypt to Pharao, and to al his ſeruantes, and to his
whole land, \V the great tentations, which thine eies haue ſeene, thoſe
mightie ſignes, and wonders, \V and our Lord
\SNote{For ſinnes paſt God letteth ſome runne into reprobate ſenſe,
permitting them to their owne freewil, who being voide of grace wilfully
obdurate them ſelues.
\Cite{Theod. q.~37. in Deut.}}
hath not geuen you a hart to vnderſtand, and eies to ſee, and eares that
can heare, vnto this preſent day. \V He hath brought you fourtie yeares
by the deſert: your garmentes are not worne out, neither are the shoes
of your feete conſumed with age. \V Breade you haue not eaten, wine and
ſicer you haue not drunke: that you might know that I am the Lord your
God. \V And you came to this place: and there came forth Sehon the King
of Heſebon, and Og the King of Baſan, meeting vs to fight. And we ſtroke
them, \V and tooke their land, and deliuered it in poſſeſſion to Ruben
and Gad, and the half tribe of Manaſſes. \V Keepe
%%% 0476
therfore the wordes of this couenant, and fulfil them: that you may
vnderſtand al thinges that you doe. \V You ſtand this day al before our
Lord your God, your princes, and tribes, and ancientes, and doctors, al
the people of Iſrael, \V your children and your wiues, and the ſtrangers
that abide with thee in the campe, beſides the cutters of wood, and
them, that carie water: \V that thou mayeſt paſſe in the couenant of our
Lord thy God, and in the oath which in this day our Lord thy God maketh
with thee: \V that he may rayſe thee vp a people to him ſelfe, and he be
thy God as he hath ſpoken to thee, and as he ſware to thy fathers,
Abraham, Iſaac, and Iacob. \V Neither with you onlie doe I make this
couenant, and confirme theſe oathes, \V but with al that be preſent and
abſent. \V For you know how we dwelt in the Land of Ægypt, and how we
haue paſſed through the middes of nations, which paſſing through \V you
haue ſeene their abominations and filth, that is to ſay, their Idols,
wood and ſtone, ſiluer and gold, which they worshipped. \V Leſt perhaps
there be among you man or woman, familie or tribe, whoſe hart is turned
away this day from our Lord God, to goe and ſerue the goddes
%%% o-0428
of thoſe Nations: and there be among you
\SNote{A mind ſecretly infected with idolatrie.}
a roote bringing forth gal and bitternes. \V And when he shal heare the
wordes of this oath, he bleſſe him ſelfe in his hart, ſaying: I shal
haue place, and walke in the prauitie of my hart: and the
\SNote{The appetite drunken with pleaſures thirſteth ſtil more.}
drunken take to her the thirſtie, \V and our Lord forgeue him not: but
then his furie moſt ſpecially fume, and his zeale againſt that man, and
al the curſes ſitte vpon him, that be written in this volume: and our
Lord aboliſh his name vnder heauen, \V and conſume him vnto perdition
out of al the tribes of Iſrael, according to the curſes, that are
conteyned in the Booke of this law and couenant. \V And the generation
folowing shal ſay, and the children that shal be borne from thence
forth, and the ſtrangers, that shal come from a farre, ſeeing the
plagues of that Land, and the infirmities, wherwith our Lord hath
afflicted it, \V burning it with brimſtone, and heate of the ſalt, ſo
that it can no more be ſowen, nor any grene thing ſpring therof, after
the example of the ſubuerſion of ſodom and Gomorrha, Adama, and Seboim,
which our Lord ſubuerted in his wrath and furie. \V And al the Nations
shal ſay: Why hath the Lord done thus to this Land? what is this
exceding wrath of his furie? \V And they
%%% 0477
ſhal anſwer: Becauſe they forſooke the couenant of the Lord, which he
made with their fathers, when he brought them out of the Land of
Ægypt: \V and they haue ſerued ſtrange goddes, and adored them, whom
they knew not, and to whom they had not beene deſigned: \V therfore the
furie of the Lord was wrath againſt this Land, to bring vpon it al the
curſes, that are written in this volume: \V and he hath caſt them out of
their land, in wrath and furie, and in verie great indignation, and hath
throwen them into a ſtrange land, as this day it is proued. \V Thinges
hidden,
\SNote{Secrete thinges are knowne to God, manifeſt thĩges to men.
\Cite{Theod. q.~38. in Deut.}}
to our Lord God: which are manifeſt, to vs and to our children for euer,
that we may doe al the wordes of this Law.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
