%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 0426
%%% o-0382
\startChapter[
  title={Chapter 4}
  ]

\Summary{Moyſes exhorteth the people to kepe Gods
  commandments. 15.~Namely that they make no ſimilitude nor image of
  man, nor of beaſt, bird, fish, ſunne, moone, nor of anie creature to
  ſerue the ſame for the Creator. He foretelleth his owne death,
  23.~threatneth them if they forſake God, 41.~and appointeth three
  cities of refuge, on the ſame ſide Iordan.}

And
\MNote{The ſecond part.

A repetition & explication of the law.}
now Iſrael heare the preceptes and iudgementes, which I teach thee: that
doing them, thou mayeſt liue, and entring in mayeſt poſſeſſe the Land,
which the Lord the God of your fathers wil geue you. \V You
\LNote{You shal not adde.}{Moyſes can not meane, that no more ſhould be
written, nor commanded; for then the laſt chapter of this booke, and the
reſt of the Bible ſhould not haue benne written after his death; neither
ought the Prieſtes or Prophetes to haue commanded anie thing not
expreſſed in the law.
\MNote{As other Scriptures are included in the law, ſo alſo Traditiõs
are conteined in the Scriptures.}
And wheras
\CNote{Brentius, Kẽniſius, Caluin.}
Proteſtantes ſay that al other Scriptures are included in the lawe, or
pertaine to the explicatiõ or performance therof: we alſo anſwer that
vnwritten Traditiõs both in the old and new Teſtament are likewiſe
implied, included, or perteine to the explication or performance of the
law. For euen as the written doctrin of the Prophetes, yea and of
Chriſt, and his Apoſtles, in general is conteined in the law of Moyſes,
ſo alſo are certaine faſtes, feaſtes, rites, ceremonies and other
traditions proued and confirmed by general ſpeaches and axiomes written
in holie Scriptures, as by our Sauiours wordes to his Apoſtles.
\XRef{Luc.~10.}
\Emph{He that heareth you, heareth me.} S.~Paules to other Chriſtians
\XRef{(1.~Cor.~10.)}
\Emph{other things vvhen I come I vvil diſpoſe.}
\XRef{(2.~Theſſ.~2.)}
\Emph{Hold the traditions vvhich you haue lerned}: and the like.
\MNote{The Church, commended by Scriptures, approueth Traditions.}
VVherupon S.~Auguſtin
\Cite{Li.~1. cont. Creſcon. c.~33.}
geueth this rule, that \Emph{albeit an euident example can not be
produced of holie Scripture, yet the truth of the ſame Scriptures is
holden by vs, vvhen vve do that pleaſeth the vvhole Church, vvhich the
authoritie of Scriptures commendeth.} The ſame he teacheth
\Cite{Epiſt.~80.}
and in manie other places. So ſdo S.~Epiphanius in
\Cite{compend. fides Cathol.}
S.~Hierom
\Cite{Dialog. cont. Lucifer. c.~4.}
S.~Chryſoſt.
\Cite{ho.~4. in 1.~Theſſal.~4.}
S.~Baſil
\Cite{de Spiritu Sancto, c.~39.}
S.~Irenius
\Cite{li.~3. c.~4.}}
shal not adde to the word, that I ſpeake to you, neither shal you take
away from it: keepe the commandment of the Lord your God
%%% o-0383
which I command you. \V Your eyes haue ſeene al thinges that our Lord
hath done againſt Beelphegor, how he hath deſtroyed al his worshippers
out of the middes of you. \V But you that cleaue to the Lord your God,
liue al vntil this preſent day. \V You know that I haue taught you
preceptes and iuſtices, as the Lord my God hath commanded me: ſo shal
you do them in the Land, which you shal poſſeſſe: \V and you shal
obſerue, and fulfil them in worke. For
\SNote{To kepe Gods cõmandments is counted by al nations the moſt
excellent wiſdome.}
this is your wiſedome, and vnderſtanding before peoples, that hearing al
theſe preceptes, may ſay: Behold a people ful of wiſedome and
vnderſtanding, a great nation. \V Neither is there other
%%% 0427
nation ſo great, that hath goddes approching vnto them, as our God is
preſent at al our petitions. \V For what other nation is there ſo
renowmed that hath the ceremonies, and iuſt iudgementes, and the whole
law, which I wil ſette forth this day before your eyes. \V Keepe thy
ſelfe therfore, and thy ſoule carefully. Forget not the wordes, that
thyne eyes haue ſeene, and let them not fal out of thy hart al the daies
of thy life. Thou shalt teach them thy ſonnes and thy nephewes, \V the
day wherin thou didſt ſtand before the Lord thy God in Horeb, when our
Lord ſpake to me, ſaying: Aſſemble vnto me the people, that they may
heare my wordes, and may learne to feare me al the time that they liue
on the earth, and may teach their children. \V And you came to the foote
of the mount, which burned euen vnto heauen: and there was in it
darkenes, and a
\Fix{clould}{cloud}{obvious typo, fixed in other}
and miſt. \V And our Lord ſpake to you from the middes of the fyre. The
voice of his wordes you heard, and forme you ſaw not at al. \V And he
shewed you his couenant, which he commanded you to do, and the
\SNote{Here and in other places it is manifeſt that the commandments,
called the \Emph{Decalogue}, are iuſt tenne.}
tenne wordes, that he wrote in two tables of ſtone. \V And he cõmanded
me at that time that I should teach you the ceremonies and iudgementes,
which you should doe in the Land, that you shal poſſeſſe. \V Keepe
therfore your ſoules carefully. You ſaw not any ſimilitude in the day,
that our Lord ſpake to you in Horeb from the middes of the fire: \V leſt
perhaps deceiued you might make you a grauen ſimilitude, or image of
male or female, \V the ſimilitude of al cattel, that are vpon the earth,
or of birdes, that flie vnder heauen, \V and of creeping beaſtes, that
moue on the earth, or of fishes, that vnder the earth abide in the
waters: \V leſt perhapes lifting vp thyne eies to heauen, thou ſee the
Sunne and the Moone, and al the ſtarres of heauen, and deceiued by
errour thou adore and ſerue them, which the Lord thy God created to
ſerue al nations, that are vnder heauen. \V But you our Lord hath taken,
and brought out of the yron furnace of Ægypt, to haue you his people by
inheritance, as it is this preſent day. \V And our Lord was angrie with
me for your wordes, and he ſware
\SNote{Venial and leaſt ſinnes paſſe not with out temporal puniſhment.}
that I ſhould not paſſe ouer Iordan, nor enter into the excellent Land,
which he wil geue you. \V Behold I die
\SNote{This was alſo a Myſterie, that the old law, ſignified by Moyſes,
could not bring to heauen, the true land of promiſe, but the law of
Chriſt, ſignified by Ioſue.
\Cite{Theodoret. q.~43. in Deut.}}
in this ground, I shal not paſſe ouer Iordan: you shal paſſe, and
poſſeſſe the goodlie Land. \V Beware leſt at any time thou forget the
%%% 0428
couenant of the Lord thy God, which he hath made with
%%% o-0384
thee: and make to thee a grauen ſimilitude of thoſe thinges, which our
Lord hath prohibited to be made: \V becauſe the Lord thy God is a
conſuming fyre,
\Fix{ielouſe}{a ielouſe}{speak English dudes, same in both}
God. \V If you shal begette ſonnes and nephewes, and abide in the Land,
and being deceiued make to you ſome ſimilitude, committing euil before
the Lord your God, to prouoke him to wrath: \V I cal this day heauen and
earth witneſſes, that you shal quikly perish from out of the Land,
which being paſſed ouer Iordan you shal poſſeſſe. You shal not dwel
therin long time, but our Lord wil deſtroy you, \V and diſperſe you into
al nations, and you shal remaine a few among the nations, to the which
our Lord wil lead you. \V And there you shal ſerue goddes, that were
framed with mens hand, wood and ſtone that ſee not, nor heare, nor eate,
nor ſmel. \V And when thou shalt ſeeke there the Lord thy God, thou shalt
finde him: yet ſo, if thou ſeeke him with al thy hart, and al
tribulation of thy ſoule. \V After that al the thinges aforeſaid shal
finde thee, and in
\SNote{Conuerſion of the Iewes in the end of the world.}
the latter time thou shalt returne to the Lord thy God, and shalt heare
his voice. \V Becauſe the Lord thy God is a merciful God: he wil not
leaue thee, nor altogether deſtroy thee, nor forget the couenant,
wherein he ſware to thy fathers. \V Aske of the dayes of old, that haue
bene before thy time from the day that God created man vpon the earth,
from one end of heauen to the other end therof, if euer there was done
the like thing, or it hath beene knowen at any time, \V that a people
should heare the voice of God ſpeaking out of the middes of fyre, as
thou haſt heard, and liued: \V if God ſo did that he went in, and tooke
vnto him a Nation out of the middes of nations, by temptations, ſignes,
and wonders, by fight and ſtrong hand, and ſtretched out arme, and
horrible viſions according to al thinges, that the Lord your God did for
you in Ægypt, thine eies ſeeing it: \V that thou mighteſt know that our
Lord, he is God, and there is none other beſide him, \V from heauen he
made thee to heare his voice, that he might teach thee. And in earth he
shewed thee his fyre, verie greate, and thou didſt heare his wordes out
of the middes of the fyre, \V becauſe he loued thy fathers, and choſe
their ſeede after them. And he brought thee out of Ægypt, going before
thee in his great power, \V to deſtroy verie great nations and ſtronger
then
%%% 0429
thou at thy entring in, and to bring thee in, and geue thee their land
in poſſeſſion, as thou ſeeſt this preſent day. \V Know therfore this
day, and thinke in thy hart that our Lord he is God in heauen aboue, and
in the earth beneth, and there is none other. \V Keepe his preceptes and
commandementes, which I command thee: that it may be wel with thee, and
thy children after thee, and thou mayeſt remayne a long time vpon the
Land, which the Lord thy God wil geue thee. \V Then Moyſes ſeparated
three cities beyond Iordan at the eaſt ſide, \V that he might flee to
them which ſhould kil his neighbour not voluntarily, neither was his
enemie a day or two before, and he might ſcape to ſome of theſe
cities: \V Bofor in the wilderneſſe, which is ſituated in the champion
%%% o-0385
countrie of the tribe of Ruben: and Ramoth in Galaad, which is in the
tribe of Gad: and Golan in Baſan, which is in the tribe of Manaſſes. \V
This is the law, that Moyſes ſette before the children of Iſrael, \V and
theſe are the teſtimonies and ceremonies and iudgementes, which he ſpake
to the children of Iſrael, when they came out of Ægypt, \V beyond Iordan
in the valley againſt the temple of Phogor in the land of Sehon king of
the Ammorrheite, that dwelt in Heſebon, whom Moyſes ſtroke. The children of
Iſrael alſo comming out of Ægypt, \V poſſeſſed his land, and the land
of Og the king of Baſan, the two kinges of the Amorrheites, which were
beyond Iordan toward the ryſing of the ſunne: \V from Aroer, which is
ſituated vpon the banke of the torrent Arnon, vnto the mountaine Sion,
which is alſo Hermon, \V al the plaine beyond Iordan at the eaſt ſide,
vnto the ſea of the wilderneſſe, and vnto the foote of mount Phaſga.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
