%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 0447
%%% o-0401
\startChapter[
  title={Chapter 14}
  ]

\Summary{Gentiles maner of mourning for the dead is
  prohibited. 3.~Likewiſe to eate things vncleane, with mention of
  certaine cleane and vncleane beaſtes, 9.~fishes, 11.~and
  birdes. 21.~Alſo preceptes of pietie, clemencie, paying tithes, firſt
  fruites, 27.~nourishing of Leuites, ſtrangers, orphanes, and widowes.}

Be ye the children of the Lord your God: you shal not cutte your ſelues,
nor make bauldnes for the dead, \V becauſe thou art a holie people to
the Lord thy God: and he choſe thee to be his peculiar people of al
\Fix{nattions,}{nations,}{obvious typo, fixed in other}
that are vpon the earth. \V Eate not the thinges that are vncleane. \V
This is the beaſt, that you ought to eate: The oxe, and the sheepe, and
the goate, \V the hart and the doa, the buffle, the chamois, the
pygargue, the wilde beefe, the cameloparde. \V Euerie beaſt, that
diuideth the hoofe in two partes, and cheweth the cudde, shal you
eate. \V But of them, that chew the cudde, and diuide not the hoofe,
theſe you shal not eate, as the camel, the hare, the cherogril: becauſe
they chew the cudde, and diuide not the hoofe, they shal be vncleane to
you. \V The ſwine alſo becauſe it diuideth the hoof, and cheweth not the
cudde, shal be vncleane. Their flesh you shal not eate, and their
carcaſſes you shal not touche. \V Theſe shal you eate of al that abide
in the waters: Such as haue finnes and ſcales, eate: \V them that are
without finnes and ſcales, eate not, becauſe they are vncleane. \V Al
birdes that are cleane eate. \V The vncleane eate not: to witte, the
eagle, and the grype, and the oſprey, \V the ringtaile, and the vulture
and kite according to their kinde: \V and al of the rauens kinde, \V
and the oſtriche, and the owle, and the ſterne, and the hawke according
to his kinde: \V the herodian and the ſwanne, and the ſtorke, \V and the
diuer, the porphyrion, and the nightcrow, \V the onocratal, and the
charadrion, euerie one in their kinde: the lapwing alſo and the
batte. \V And al that creepeth and hath litle winges, shal be vncleane,
and shal not be eaten. \V Al that is cleane, eate. \V But
%%% 0448
whatſoeuer is dead of it ſelfe, eate not therof.
\SNote{If theſe thĩgs were vncleane by nature they were not lawful for
anie nation to eate, but being only forbid to the Iewes ſheweth, that
this prohibition was ceremonial, only for that time and people.}
To the ſtranger, that is within thy gates, geue it to eate, or ſel it to
him: becauſe thou art the holie people of our Lord thy God. Thou shalt
\SNote{Al ſhew of crueltie to be auoided. Myſtically this prefigured,
that Chriſt (for the ſimilitude of ſinful fleſh ſignified by a kidde)
ſhould not be ſlaine in his infancie.
\Cite{S.~Tho. 1.~2. q.~102. a.~6. ad~4.}}
not boyle a kidde in the milke of his damme. \V The tenth part thou
shalt ſeperate of al thy fruites that ſpring in the earth euerie
yeare, \V and thou shalt eate in the ſight of our Lord thy God in the
place, which he shal chooſe, that his name may be inuocated therin, the
tithe of thy corne,
%%% o-0402
and wine, and oile, and the firſt borne of thy heardes and sheepe: that
thou mayeſt lerne to feare our Lord thy God at al time. \V But when the
way, and the place which our Lord thy God shal chooſe, are farre, and he
hath bleſſed thee, and thou canſt not carie al theſe thinges thither, \V
thou shalt ſel, and bring al into a price, and shalt carie it in thy
hand, and shalt goe to the place, which our Lord thy God shal chooſe: \V
and thou shalt buy with the ſame money whatſoeuer pleaſeth thee, either
of heardes, or of sheepe, wine alſo and ſicere, and al that thy ſoule
deſireth: and thou shalt eate before our Lord thy God, and shalt feaſt,
thou and thy houſe: \V and the Leuite that is within thy gates, beware
thou forſake him not, becauſe he hath no other part in thy
poſſeſſion. \V The third yeare thou shalt ſeparate an other tenth of al
thinges, that growe to thee at that time: and shalt lay it vp within thy
gates. \V And the Leuite shal come that hath no other part nor
poſſeſſion with thee, and the ſtranger and pupil and widow, that are
within thy gates, and shal eate and be filled: that our Lord thy God may
bleſſe thee in al the workes of thy handes that thou shalt doe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
