%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 0457
%%% o-0411
\startChapter[
  title={Chapter 20}
  ]

\Summary{Lawful warres are to be vndertaken with corege and
  confidence. 5.~Such as for ſpecial cauſes may be diſcoreged, muſt be
  diſmiſſed from the field. 10.~VVhat to be obſerued towardes the
  enimie. 19.~VVhat trees may not be cut downe, and what ſorte may be,
  for the vſe of warres.}

If thou goe forth to warre againſt thyne enemies, and ſee the horſe men
and chariottes, and the multitude of the aduerſaries hoſte greater then
thou haſt, thou shalt not feare them: becauſe the Lord thy God is with
thee, which brought thee out of the Land of Ægypt. \V And when the
battel is now at hand, the prieſt shal ſtand before the armie, and shal
ſpeake to the people thus: \V Heare Iſrael, you this day ioyne battel
againſt your enemies, let not your hart feare, be not
%%% 0458
affrayed, retire not, neither dread them: \V becauſe our Lord your God
is in the middes of you, and wil fight for you againſt your aduerſaries,
to deliuer you from danger. \V The Captaines alſo through euerie bande
in the hearing of the hoſte ſhal proclaime:
\SNote{Men poſſeſſed with ſuch deſires haue not like valure to good
ſouldiars. And by word or example often diſcorege others. So in
ſpiritual warfare, we muſt not be addicted to worldlie profites, or
pleaſures.}
what man is there, that hath built a new houſe, and hath not dedicated
it? let him goe, and returne into his houſe, leſt perhaps he die in the
battel, and an other dedicate it. \V What man is there that hath planted
a vineyard, and hath not as yet made it to be common, wherof al men may
lawfully eate? let him goe, and returne into his houſe: leſt perhaps he
die in the battel, and an other man execute his office. \V What man is
there, that hath deſpouſed a wife, and not taken her? let him goe, and
returne into his
%%% o-0412
houſe, leſt perhaps he die in the warre, and an other take her. \V Theſe
thinges being ſaid they shal adde the reſt, and shal ſpeake to the
people: What man is there feareful, and fainte harted? let him goe, and
returne into his houſe, leſt he make the hartes of his brethren to
feare, as him ſelfe is frighted with feare. \V And when the Captaines of
the hoſt ſhal hold their peace, and make an end of ſpeaking, euerie man
shal prepare their bandes to fight. \V If at anie time thou come to
winne a citie, thou shalt firſt offer peace. \V If they receiue it, and
open the gates to thee, al the people that is therein, shal be ſaued,
and shal ſerue thee vnder tribute. \V But if they wil not make peace,
and shal beginne battel againſt thee, thou shalt aſſalt it. \V And when
our Lord thy God shal deliuer it into thy hand, thou shalt ſtrike al,
that is therein of the male ſexe, in the edge of the ſword, \V excepting
wemen and children, the cattel and other thinges, that are in the
citie. Al the praye thou shalt diuide to the armie, and thou shalt eate
of the ſpoiles of thine enemies, which our Lord thy God shal geue
thee. \V So shalt thou doe to al cities, that be verie farre from thee,
and be not of theſe cities, which thou shalt receiue in poſſeſſion. \V
But of thoſe cities, that shal be geuen thee, thou shalt ſuffer none at
al to liue: \V but shalt kil them in the edge of the ſword, to witte,
the Hetheite, and Amorrheite, and Chananeite, the Pherezeite, and
Heueite, and Iebuſeite, as our Lord thy God hath commanded thee: \V leſt
perhaps they teach you to doe al the abominations, which them ſelues did
worke to their goddes: and you ſinne againſt our Lord your God. \V When
thou haſt
%%% 0459
beſieged a citie a long time, and haſt compaſſed it with munition to
winne it, thou shalt not cutte downe the trees, that may be eaten of,
neither ſhalt thou ſpoile the countrie round about with axes: becauſe it
is a tree, and not a man, neither can it increaſe the number of warryers
againſt thee. \V But if there be any trees not fruitful, but wilde, and
apt for other vſes, cutte them downe, and make engines, vntil thou take
the citie, which fighteth againſt thee.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
