%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-23


\project douay-rheims


%%% 0462
%%% o-0415
\startChapter[
  title={Chapter 23}
  ]

\Summary{Eunuches, baſtardes, Moabites, & Ammonites may not enter into
  the Church. 7.~Idumeans, and Ægyptians may be admitted. 9.~Obſeruation
  of ſpiritual and corporal cleannes. 15.~Other preceptes concerning
  fugitiues, 17.~fornication, 19.~vſurie, 21.~vowes, 24.~and eating
  other mens grapes or corne.}

An
\SNote{Such as are barrẽ in good workes can not enter into Gods houſe.
\Cite{Theod. q.~25. in Deut.}}
eunuch that hath his ſtones broken, or cutte of, & his yeard cutte away,
shal not enter into the church of our Lord. \V Mamzer, that is to ſay,
one borne of a common woman, shal not enter into the church of our Lord,
vntil the tenth generation. \V The
\SNote{Theſe natiõs not able to hurt the children of Iſrael, neither by
denying ordinarie curteſies, nor by force, nor by hyring Balaã to curſe
them, yet inuegling them with carnal ſinnes, ſignified obſtinate
peruerſe ſinners, that neuer amending can neuer be rightly receiued into
the Church of God.}
Ammonite, and the
%%% 0463
Moabite yea after the tenth generation shal not enter into
%%% o-0416
the church of our Lord, for euer: \V becauſe they would not meete you
with bread and water in the way, when you came out of Ægypt: and becauſe
they hyred againſt thee Balaam, the ſonne of Beor of Meſopotamia in
Syria, to curſe thee: \V and our Lord thy God would not heare Balaam,
and he turned his curſing into thy bleſſing, for that he loued thee. \V
Thou shalt not make peace with them, neither doe thou ſeeke their good
al the daies of thy life for euer. \V Thou shalt not abhorre the
Idumeite, becauſe he is thy brother: nor the Ægyptian, becauſe thou waſt
a ſtranger in his land. \V They that are borne of them, in the third
generation shal enter into the church of our Lord. \V When thou goeſt
forth againſt thyne enemies to battel, thou shalt keepe thy ſelf from al
euil thing. \V If there be among you a man, that is polluted in a dreame
by night, he shal goe forth without the campe, \V and shal not returne,
before he be washed with water at euen: and after ſunne ſette he shal
returne into the campe. \V Thou shalt haue a place without the campe,
whither thou mayeſt goe to the neceſſities of nature, \V carying on thy
girdle a piked inſtrument, & when thou ſitteſt downe, thou shalt digge
round about, and with the earth that is digged vp shal couer \V that
which thou art eaſed of (for our Lord thy God walketh in the middes of
thy campe, to deliuer thee, and to geue thyne enemies vnto thee) and let
thy campe be holie, and let no filthines appeare therein, leſt he
forſake thee. \V Thou shalt not deliuer the ſeruant to his Maiſter, that
is fled to thee. \V He shal dwel with thee in the place, that shal
pleaſe him, and in one of thy cities shal he reſt: vexe him not. \V
There shal be no whoore of the daughters of Iſrael, nor whooremonger of
the ſonnes of Iſrael. \V Thou shalt not offer the hire of a ſtrompet,
nor the price of a dogge, in the houſe of our Lord thy God, whatſoeuer
it be that thou haſt vowed: becauſe both is abomination before our Lord
thy God. \V Thou shalt not lend to thy brother money to vſurie, nor
coine, nor any other thing: \V but
\SNote{Onlie lawful enemies are here called ſtrangers: where therfore is
iuſt cauſe of warre, there only it is lawful to exerciſe vſurie.
\Cite{S.~Amb. li. de Tobia. c.~15.}}
to the ſtranger. And to thy brother thou shalt lend, that which he
needeth without vſurie: that our Lord thy God may bleſſe thee in al thy
worke in the Land, which thou shalt enter to poſſeſſe. \V When thou haſt
vowed a vow to our Lord thy God, thou shalt not ſlacke to pay it:
becauſe our Lord thy God wil require it
%%% 0464
and if thou delay, it shal be reputed to thee for ſinne. \V
\SNote{Vowes binde where otherwiſe was no obligation.}
If thou wilt not promiſe, thou shalt be without ſinne. \V But that which
is once gone out of thy lippes, thou shalt obſerue, and shalt doe as
thou haſt promiſed to our Lord thy God, and haſt ſpoken with thy proper
wil and thyne owne mouth. \V Entring into thy neighbours vineyarde, eate
grapes as much as shal pleaſe thee: but carrie none out with thee. \V If
thou enter into thy freindes corne, thou shalt breake the eares, and
rubbe them in thy hand: but with a
\Fix{ſikcle}{ſickle}{likely typo, same in other}
thou shalt not reape.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
