%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 0424
%%% o-0381
\startChapter[
  title={Chapter 3}
  ]

\Summary{The victorie againſt Og king of Baſan of the giants ſtock is
  repeted, 12.~Ruben Gad and halfe tribe of Manaſses haue poſſeſsion on
  the other ſide Iordan from their brethren. 23.~Moyſes praying that he
  may goe ouer Iordan, for the ſinnes of the people is denied.}

Therfore turning we went vp by the way of Baſan: and Og the king of
Baſan came forth to meete vs with his people to fight in Edrai. \V And
our Lord ſaid to me: Feare him not: becauſe he is deliuered into thy
hand with al his people and his land: and thou ſhalt doe to him as thou
haſt done to Sehon king of the Amorrheites, that dwelt in Heſebon. \V
Therfore the Lord our God deliuered into our handes Og alſo the king of
Baſan, and al his people: and we ſtroke them to vtter deſtruction, \V
waiſting al his cities at one time. There was not a towne that eſcaped
vs: ſixtie cities, al the countrie of Argob the kingdome of Og in
Baſan. \V Al the cities were fenſed with verie high walles, and with
gates and barres, beſide innumerable townes that had no walles. \V And
we deſtroyed them, as we had done to Sehon the king of Heſebon,
deſtroying euerie citie, and men and wemen
%%% 0425
and children: \V but the cattel, and the ſpoyles of the cities we tooke
for our praye. \V And we tooke at that time the land out of the hand of
two kinges of the Amorrheites, that were beyond Iordan: from the torrent
Arnon vnto the mountaine Hermon, \V which the Sidonians cal Sarion, and
the Amorrheites Sanir: \V al the cities, that are ſituated in the
plaine, and al the Land of Galaad and Baſan as farre as Selcha, and
Edrai cities of the kingdome of Og in Baſan. \V For onlie Og the king of
Baſan remayned of the ſtocke of giantes. His bed of yron is ſhewed,
which is in Rabbath of the children of Ammon, hauing nine cubites
in length, and foure in breadth after the meaſure of
\SNote{Longer ſorte of cubites are a foote and 9.~inches: ſo this bed
was 15.~foote and nine inches long, and 7.~foote brode.
\Cite{Vitruuius Agricola.}}
the cubite of a mans hand. \V And we poſſeſſed the Land at that time
from Aroer, which is vpon the banke of the torrent Arnon, vnto the halfe
part of mount Galaad: and the cities therof I gaue to Ruben and Gad. \V
And the other part of Galaad, and al Baſan of the kingdome of Og, I
deliuered to the halfe tribe of Manaſſes, al the countrie of Argob: and
al Baſan is called the Land of giantes. \V Iair the ſonne of Manaſſes
poſſeſſed al the countrie of Argob vnto the borders of Geſſuri, and
Machati. And he called Baſan by his owne name, Hauoth Iair, that is to
ſay, the townes of Iair,
\SNote{Eſdras adding theſe wordes, and often times the like, did not
againſt the law, becauſe ſuch additions are agreable and not contrarie
to that which was written before.}
vntil this preſent day. \V To Machir alſo I gue Galad. \V And to the
tribes of Ruben and Gad I gaue of the Land of Galaad as farre as
%%% o-0382
the Torrent Arnon, halfe of the torrent, and the confines vnto the
torrent Ieboc, which is the border of the childrẽ of Ammon: \V and the
plaine of the wilderneſſe, and Iordan, and the borders of Cenereth vnto
the ſea of the deſert, which is moſt ſalt, at the foote of mount Phaſga
againſt the eaſt. \V And I commãded you at that time, ſaying: The Lord
your God geueth you this land for an inheritance, goe wel appointed
before your brethren the children of Iſrael al you ſtrong men: \V except
your wiues, and litle ones and your cattel. For I know you haue much
cattel, & they muſt remaine in the cities, which I haue deliuered
you, \V vntil our Lord geue reſt to your brethren, as he hath geuen to
you: and they alſo poſſeſſe the Land, which he wil geue them beyond
Iordan: then shal euerie man returne to his poſſeſſion, which I haue
geuen you. \V Ioſue alſo at that time I commanded, ſaying: Thyne eyes
haue ſeene what the Lord your God hath done to theſe two kinges: ſo wil
he doe to al the kingdomes, to the which thou
%%% 0426
ſhalt paſſe. \V Feare them not: for the Lord your God wil fight for
you. \V And I prayed our Lord at that time, ſaying: \V Lord God thou haſt
begonne to ſhew vnto thy ſeruant thy greatnes, and moſt mightie
hand. For neither is there other God either in heauen, or in earth, that
is able to doe thy workes, and to be compared to thy ſtrength. \V I wil
paſſe ouer therfore, and wil ſee this excellent Land beyond Iordan, and
this goodlie mountaine, and Libanus. \V And our Lord was angrie with me
\SNote{See
\XRef{Num.~20. v.~12.}}
for you, and heard me not, but ſaid to me: It ſufficeth thee: ſpeake no
more to me of this matter. \V Goe vp to the toppe of Phaſga, and caſt
thine eies round about to the weſt, and to the north, and the ſouth,
and the eaſt, and behold it. For thou ſhalt not paſſe this Iordan. \V
Command Ioſue, and encourage and
\Fix{ſtrenghen }{ſtrengthen}{obvious typo, fixed in other}
him: for he shal goe before this people, and shal diuide vnto them the
Land, which thou shalt ſee. \V And we abode in the valley againſt the
temple of Phogor.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
