%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 0434
%%% o-0389
\startChapter[
  title={Chapter 7}
  ]

\Summary{No league nor felowship to be had with the Gentiles: 5.~but
  their altares, groues, and al their idoles to be deſtroyed. 17.~God
  promiſeth victories to his people, willing them to truſt in him, and
  ſerue him.}

VVhen the Lord thy God shal haue brought thee into the land, which thou
doeſt enter in to poſſeſſe, and shal haue deſtroyed manie Nations before
thee, the Hetheite, and the Gergezeite, and the Amorrheite, and the
Chananeite, and the Pherezeite, and the Heueite, and the Iebuſeite,
ſeuen nations of much greater number then thou art, and ſtronger then
thou: \V and the Lord thy God shal haue deliuered them to thee, thou
ſhalt ſtrike them vnto vtter deſtruction. Thou shalt not make league with
them, nor pitie them, \V nor make mariages with them. Thy daughter thou
ſhalt not geue to his ſonne, nor take his daughter for thy ſonne: \V for
he wil ſeduce thy ſonne, that he folow not me, and that he rather ſerue
ſtrange goddes, and the furie of our Lord wil be wrath, and shal quickly
deſtroy thee. \V But theſe thinges rather you shal doe to them:
Ouerthrow their altares, and breake their
\Fix{ſtatuees,}{ſtatues,}{likely typo, fixed in other}
and cutte downe their groues, and burne their ſculptiles. \V Becauſe
thou art a holie people to the Lord thy God. The Lord thy God hath
choſen thee, to be his peculiar people of al peoples, that are vpon the
earth. \V Not becauſe you paſſed al nations in number, is our Lord
ioyned vnto you, and hath choſen you, wheras you are fewer then al
peoples: \V but becauſe
%%% o-0390
our Lord hath loued you, and hath kept the oath, which he ſware to your
fathers: and hath brought you forth in a ſtrong hand, and redemed you
from the houſe of ſeruitude, out of the hand of Pharao the king of
Ægypt. \V And thou shalt know that the Lord thy God, he is a ſtrong and
faithful God, keping his couenant and mercie to them that loue him, and
to them that keepe his preceptes, vnto a thouſand generations: \V and
rendring forthwith to them that hate him, ſo that he deſtroyeth them,
and differeth no longer,
\SNote{Not withſtanding this commination, God oftentimes differeth
puniſhment, expecting the ſinners repentance.}
immediatly rendering to them that they deſerue. \V Keepe therfore the
preceptes and ceremonies and iudgementes, which I command
%%% 0435
thee this day to doe them. \V
\SNote{Gods promiſes cõditional, if his people ſerue him.}
If after thou haſt heard theſe iudgementes, thou keepe and doe them, the
Lord alſo thy God wil keepe the couenant vnto thee, and the mercie which
he ſware to thy fathers: \V and he wil loue and multiplie thee, and wil
bleſſe the fruite of thy wombe, and the fruite of thy land, thy corne,
and vintage, oile, and heardes, the flockes of thy ſheepe vpon the Land,
for the which he ſware to thy fathers that he would geue it thee. \V
Bleſſed ſhalt thou be among al peoples. There ſhal be none barren with
thee of neither ſexe, as wel in men as in thy flockes. \V Our Lord wil
take away from thee al diſeaſe: and the ſore infirmities of Ægypt, which
thou knoweſt, he wil not bring vpon thee, but vpon al thyne enemies. \V
Thou ſhalt deuoure al the peoples, which the Lord thy God wil geue
thee. Thyn eye ſhal not ſpare them, neither shalt thou ſerue their
goddes, leſt they be the ruine of thee. \V If thou ſay in thy hart:
Theſe nations are moe then I, how ſhal I be able to deſtroy them? \V
Feare not, but remember what the Lord thy God did to Pharao and to al
the Ægyptians, \V the exceding great plagues, which thyne eies ſaw, and
the ſignes and wonders, and the ſtrong hand, and the ſtretched out arme,
that the Lord thy God might bring thee forth: ſo wil he doe to al
peoples, whom thou feareſt. \V Moreouer hornettes alſo wil the Lord thy
God ſend vpon them, vntil he deſtroy and conſume al that eſcaped thee,
and can hide them ſelues. \V Thou shalt not feare them, becauſe the Lord
thy God is in the middes of thee, a mightie God and terrible: \V he wil
conſume theſe nations in thy ſight by litle and litle and by
partes. Thou mayeſt not deſtroy them al together: leſt perhappes the
beaſtes of the earth multiplie againſt thee. \V And the Lord thy God wil
geue them in thy ſight: and wil kil them vntil they be vtterly
deſtroyed. \V And he wil deliuer their kinges into thy handes, and thou
shalt deſtroy their names vnder heauen: no man shal be able to reſiſt
thee, vntil thou deſtroy them. \V Their ſculptiles thou shalt burne with
fyre: thou shalt not couet the ſiluer and gold, wherof they were made,
neither shalt thou take to thee any thing therof, leſt thou offend,
becauſe it is the abomination of the Lord thy God. \V Neither shalt thou
bring in ought of the Idol into thy houſe, leſt thou become anathema, as
alſo that is. As filthines thou shalt deteſt it, and as vncleannes and
filth thou shalt account it abominable, becauſe it is anathema.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
