%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 28
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-28


\project douay-rheims


%%% 0471
%%% o-0423
\startChapter[
  title={Chapter 28}
  ]

\Summary{Diuers bleſsings are promiſed to the obſeruers of Gods
  commandmentes, 15.~and curſes threatned to tranſgreſſors.}

But if thou wilt heare the voice of our Lord thy God, that thou doe and
keepe al his commandementes, which I command thee this day, our Lord thy
God wil make thee higher then al nations, that be on the earth. \V And
\SNote{Temporal bleſſings belonged to ſenſual people of the old
teſtament: now the poore in ſpirite are bleſſed, that mourne, and ſuffer
perſecution for truth and iuſtice.}
al theſe bleſſinges shal come vpon thee, and ouertake thee: yet ſo if
thou heare his preceptes. \V Bleſſed shalt thou be in the citie, and
bleſſed in the field. \V Bleſſed shal be the fruite of thy wombe, and
the fruite of thy ground, and the fruite of thy cattel, the troupes of
thy heardes, & the foldes of thy sheepe. \V Bleſſed shal thy barnes be,
and bleſſed
\SNote{The poore beĩg releeued of thy ſuperfluitie shal bleſſe thee.}
thy remaines. \V Bleſſed shalt thou be comming in and going out. \V
Our Lord wil geue thyne enemies, that riſe vp againſt thee, to fal downe
in thy ſight: one way they shal comme againſt thee, and ſeuen wayes they
shal flee from thy face. \V Our Lord wil ſend forth bleſſing vpon thy
cellars, and vpon al the workes of thy handes: and wil bleſſe thee in
the land, that thou shalt receiue. \V Our Lord wil rayſe thee vp vnto
him ſelf to be a holie people, as he ſware to thee: If thou keepe the
commandmentes of our Lord thy God, and walke in his waies. \V And al the
people of the earth shal ſee that the name of our Lord is inuocated vpon
thee, and they shal feare thee. \V Our Lord wil make thee abound with al
goodes, with the fruite of thy wombe, and the fruite of thy cattel, with
the fruite of thy land, which our Lord ſware to thy fathers that he
would geue thee. \V Our Lord wil open his moſt
%%% o-0424
excellent treaſure, the heauen, that it may geue rayne to thy land in
due ſeaſon: and wil bleſſe al the workes of thy handes. And thou shalt
lend to many nations, and thy ſelf shalt take lone of no man. \V And our
Lord ſhal make thee
\SNote{Thou ſhalt rule ouer others & none ouer thee.}
the heade, and not
\Fix{thr}{the}{obvious typo, fixed in other}
taile: and thou shalt be alwaies aboue, and not vnder:
\SNote{Yet alwaies with this condition: if thou ſerue God.}
yet ſo, if thou wilt heare the commandmentes of our Lord thy God which I
command thee this day, and keepe and doe them, \V and decline not from
them neither to the right hand, nor to the left, nor folow ſtrange
goddes, nor
%%% 0472
ſerue them. \V But if thou wilt not heare the voice of our Lord thy God,
to keepe, and doe al his commandmentes and ceremonies, which I command
thee this day,
\SNote{Thus moſt cõmonly ſinners were curſed in the old Teſtament, but
ſuch as now ſerue not God rightly, and yet proſper in this world, shal
in a moment deſcend into hel.
\XRef{Iob.~21.}}
al theſe curſes shal come vpon thee, and ouertake thee. \V Curſed shalt
thou be in the citie, curſed in the field. \V Curſed shal thy barne be,
and curſed thy remaines. \V Curſed shal be the fruite of thy wombe, and
the fruite of thy ground, the heardes of thy oxen, and the
flockes of thy sheepe. \V Curſed shalt thou be comming in, and
curſed going out. \V Our Lord shal ſend vpon thee famine & hunger, and
rebuke vpon al the workes, which thou shalt doe: vntil he conſume, and
deſtroy thee quickly, for thy moſt wicked inuentions, wherein thou haſt
forſaken me. \V Our Lord ſette the peſtilence vpon thee, vntil he
conſume thee out of the land, which thou shalt enter in to poſſeſſe. \V
Our Lord ſtrike thee with pouertie, with the feuer and
cold, with burning and heate, and with corrupt ayre and blaſting, and
purſew thee til thou perish. \V Be the heauen, that is ouer thee, of
braſſe: and the ground, that thou treadeſt, of yron. \V Our Lord geue
duſt for rayne vpon thy land, and ashes deſcend from heauen vpon thee,
til thou be conſumed. \V Our Lord deliuer thee to fal downe before thyne
enemies. One way goe thou forth againſt them, and flee ſeuen, and be
thou diſperſed through out al the kingdomes of the earth. \V And be thy
carcaſſe meate to al the fowles of the ayre, and beaſtes of the earth,
and be there none to driue them away. \V Our Lord ſtrike thee with the
boile of Ægypt, and the part of thy bodie, by the which dung is caſt out,
with ſcabbe alſo and itche: ſo that thou canſt not be cured. \V Our Lord
ſtrike thee with madnes & blindnes and furie of minde, \V and grope thou
at midday as the blined is wont to grope in the darke, and direct not thy
wayes. And at al times ſuſteyne thou wrong, and be thou oppreſſed with
violence, neither haue thou any to deliuer thee. \V Take thou a wife,
and an other ſleepe with her. Build thou a houſe, and dwel not
therin. Plant thou a vineyard, and take not the vintage there of. \V Be
thy oxe immolated
\Fix{beforre}{before}{obvious typo, fixed in other}
thee, and thou not eate therof. Be thy aſſe taken away in thy ſight, and
not reſtored to thee. Be thy sheepe geuen to thyne enemies, and be there
none to helpe thee. \V Be thy ſonnes and thy daughters deliuered to an other
people, thyne eies ſeing, and daſeling at the ſight of them al the day,
and be there no ſtrength in
%%% 0473
thy hand. \V The fruites of thy land, and al thy laboures let a people
eate, which thou knoweſt not: and be thou alwaies ſuſteyning calumnie,
and oppreſſed al dayes, \V and aſtonished at the terrour of thoſe
%%% o-0425
thinges, which thyne eies shal ſee. \V Our Lord ſtrike thee with a verie
ſore botche in the knees and shankes, and be thou vncurable from the
ſole of the foote vnto the toppe of thy head. \V Our Lord shal bring
thee, and thy King, whom thou shalt appoint ouer thee, vnto a nation,
which thou and thy fathers know not: and there thou shalt ſerue ſtrange
goddes, wood and ſtone. \V And thou shalt be deſtroyed for a prouerbe
and fable to al peoples, vnto whom our Lord shal bring thee in. \V Thou
shalt caſt much ſeede into the ground, and gather litle: becauſe the
locuſtes shal deuoure al thinges. \V Thou shalt plant a vineyard, and
digge, and the wine thou shalt not drinke, nor gather any thing therof:
becauſe it shal be waſted with wormes. \V Thou shalt haue oliues in al
thy borders, and shalt not be anointed with the oyle: becauſe they shal
droppe away & perish. \V Thou shalt begette ſonnes and daughters, and
shalt not enioy them: becauſe they shal be ledde into captiuitie. \V Al
thy trees and the fruites of thy ground the blaſting shal conſume. \V
\SNote{After manie other plagues and puniſhments, at laſt the Iewes
refuſing and perſecutĩg Chriſt, were reiected, and Gentiles called into
the Church, and aduanced aboue them.
\Cite{Theod. q.~34. in Deut.}}
The ſtranger that liueth with thee in the Land, shal aſcend ouer thee,
and shal be higher: and thou shalt deſcend downeward, and be
inferior. \V He shal lend thee, and thou shalt not lend him. He shal be
as the head, and thou shalt be the tayle. \V And al theſe
curſes shal come vpon thee, and purſewing shal ouertake thee, til thou
perish: becauſe thou heard not the voice of our Lord thy God, nor kept
his commandmentes and ceremonies which he commanded thee. \V And they
shal be in thee as ſignes and wonders, and in thy ſeede for euer: \V
becauſe thou didſt not ſerue our Lord thy God in ioy, and gladnes of
hart, for the abundance of al thinges. \V Thou shalt ſerue thine enemie,
whom our Lord wil ſend vpon thee, in hunger, and thirſt, and nakednes,
and al penurie: and he shal put an yron yoke vpon thy necke, til he
conſume thee. \V Our Lord wil bring vpon thee a Nation from a farre, and
from the vttermoſt endes of the earth, in likenes of an eagle that
flieth with vehemencie: whoſe tongue thou canſt not vnderſtand: \V a
verie malapert Nation, that wil attribute nothing to the ancient, nor
haue pitie on the litle one, \V and wil deuoure
%%% 0474
the fruite of thy cattel, and the fruites of thy Land: vntil thou
perish, and wil not leaue thee wheate, wine, and oile, heardes of oxen,
and flockes of sheepe vntil it deſtroy thee, \V and conſume thee in al
thy cities, and thy ſtrong and highe walles be deſtroyed, wherin that
hadſt confidence in al thy Land. Thou shalt be beſieged within thy gates
in al thy Land, which our Lord thy God wil geue thee: \V and thou shalt
eate the fruite of thy wombe, and the flesh of thy ſonnes and of thy
daughters, which our Lord thy God ſhal geue thee, in the diſtreſſe and
vaſtation wherwith thyne enemie shal oppreſſe thee. \V The man that is
delicate in thee, and very riotious, shal much enuie his owne brother,
and his wife, that lieth in his boſome, \V ſo that he shal not geue them
of the flesh of his children, which he wil eate: becauſe he hath nothing
els in the ſiege and penurie, wherwith thine enemies shal waſte thee
%%% o-0426
within al thy gates. \V The tender and delicate woman, that could not
goe vpon the ground, nor ſette downe her foote for ouer much nicenes and
tendernes, wil enuie her husband, that lyeth in her boſome, vpon the
flesh of her ſonne, and daughter, \V and the filthines of the after
birthes, that come forth from the middes of her thighes, and vpon the
children that are borne of the ſame howre. For they shal eate them
ſecretely becauſe of the penurie of al thinges, in the ſiege and
vaſtation, wherwith thine enemies shal oppreſſe thee within thy
gates. \V Vnleſſe thou keepe, and doe al the wordes of this law, that be
written in this volume, and feare his name glorious and terrible, that
is. Our Lord thy God: \V our Lord shal increaſe thy plagues, and the
plagues of thy ſeede, greate plagues and continuing, ſore infirmities
and perpetual. \V And he shal turne vpon thee al the afflictions of
Ægypt, which thou didſt feare, and they shal cleaue to thee. \V Moreouer
alſo al the diſeaſes, and plagues, that be not written in the volume of
this law, our Lord wil bring vpon thee, til he conſume thee: \V and you
shal remaine few in number, which before was as the ſtarres of heauen
for multitude, becauſe thou heardſt not the voice of our Lord thy
God. \V And as before our Lord reioyced vpon you, doing good to you, and
multiplying you: ſo he shal reioyce deſtroying and ſubuerting you, ſo
that you may be taken away from the Land, which thou shalt enter to
poſſeſſe. \V Our Lord shal diſperſe thee into al peoples, from the
%%% 0475
fartheſt partes of the earth to the endes therof: and there thou shalt
ſerue ſtrange goddes, which thou art ignorant of and thy fathers, woode
and ſtone. \V In thoſe nations alſo thou shalt not be quiet, neither
shal there be reſting for the ſteppe of thy foote. For our Lord wil geue
thee a feareful hart, and daſeling eies, and a ſoule conſumed with
penſifenes: \V and thy life shal be as it were hanging before thee. Thou
shalt feare night and day, and thou shalt not truſt in thy life. \V In
the morning thou shalt ſay: Who wil graunt me euening? and at euening:
Who wil grant me morning? for the fearefulnes of thy hart, wherwith thou
shalt be terrified, and for thoſe thinges, which thou shalt ſee with
thine eies. \V Our Lord shal bring thee againe with shippes into Ægypt
by the way, wherof he ſaid to thee that thou shouldeſt ſee it no
more. There shalt thou be ſould to thine enemies for bondmen and
\Fix{bondweomen,}{bondwomen,}{likely typo, fixed in other}
and no man shal bye you.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
