%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 0436
%%% o-0391
\startChapter[
  title={Chapter 8}
  ]

\Summary{The people is put in mind of afflictions which happened in the
  deſerte, and of benefites as wel paſt, as promiſed, 11.~to the end
  they loue and ſerue God more effectually.}

Everie commandement, that I command thee this day, take diligent heede
that thou doe it: that you may liue, and be multiplied, and entring in
may poſſeſſe the Land, for the which our Lord ſware to your fathers. \V
And thou shalt remember al the iourney, through the which the Lord thy
God hath brought thee fourtie yeares by the deſert, that he might
afflict and proue thee, and that the thinges that were in thy hart might
be made knowen, whether thou wouldeſt keepe his commandementes or
not. \V He afflicted thee with penurie, and gue thee for meate Manna,
which thou kneweſt not nor thy fathers: for to shew vnto thee that
\SNote{God is able to make foode of what he pleaſe, or to ſuſtaine men
without meate.}
not in bread onlie a man liue, but in euerie word that procedeth from
the mouth of God. \V Thy rayment, wherwith thou waſt couered, hath not
decayed for age, and thy foote is not worne, lie this is the fourtith
yeare. \V That thou mayeſt recount in thy hart, that as a man
diſciplineth his ſonne, ſo the Lord thy God hath diſciplined thee, \V
that thou shouldeſt keepe the cõmandementes of the Lord thy God, & walke
in his wayes, and feare him. \V For the Lord thy God wil bring thee in
vnto a good land, a land of riuers & waters and of fountaynes: in the
plains wherof and mountaynes deepe floudes gush out: \V a land of
wheate, of barley & vineyardes, wherein figge trees and pomegranates,
and oliueyardes doe grow: a land of oyle and honie. \V Where without any
penurie thou shalt eate thy bread, and enioy abundance of al thinges:
whoſe ſtones are yron, and out of the mountaynes therof are digged
metalles of braſſe: \V that when thou haſt eaten, and art ful, thou
mayeſt bleſſe the Lord thy God for the excellent land, which he hath
geuen thee. \V Obſerue, and beware leſt at any time thou forget the Lord
thy God, and neglect his commandementes and iudgementes and ceremonies,
which I command thee this day: \V leſt after thou haſt eaten and art
filled, haſt built goodlie houſes, and dwelled in them, \V and ſhalt
haue heardes of oxen and flockes of sheepe, of gold and ſiluer, and of
al thinges plentie, \V thy hart be lifted vp, and thou remember not the
Lord
%%% 0437
thy God, that brought thee out of the Land of Ægypt, out of the houſe of
ſeruitude: \V and was thy conductor in the huge and terrible
wilderneſſe, wherein was the ſerpent
%%% o-0392
burning with his breath, and the ſcorpion and
\SNote{A ſerpent leſſe then a ſcorpion, making thoſe whom he byteth to
die of thirſt.
\Cite{Solinus in polyhiſt. cap. de Africa.}}
the dipſas, and no waters at al: who brought forth riuers out of the
hardeſt rocke, \V and fed thee with Manna in the wilderneſſe, which thy
fathers knew not. And after he had afflicted and proued thee, at the
laſt he had mercie vpon thee, \V leſt thou ſhouldeſt ſay in thy hart:
Myne owne force, and the ſtrength of myne owne hand, haue atchieued al
theſe thinges for me. \V But remember the Lord thy God, that he hath
geuen thee ſtrength, that he might fulfil his couenant, concerning which
he ſware to thy fathers, as this preſent day ſheweth. \V But if
forgetting the Lord thy God, thou ſhalt folow ſtrange goddes, and ſhalt
ſerue and adore them: behold now I foretel thee that thou shalt periſh
vtterly. \V As the Nations, which our Lord deſtroyed at thyne entrie, ſo
ſhal you alſo perish, if you be diſobedient to the voice of the Lord
your God. 


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
