%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Deuteronomy
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 0452
%%% o-0405
\startChapter[
  title={Chapter 17}
  ]

\Summary{Perfect hoſtes, not mamed nor defectiue, muſt be offered to
  God, Idolaters ſtoned to death. 8.~VVhen inferior iudges differ, the
  cauſe muſt be decided by the High Prieſt in conſiſtorie. Who is
  warranted not to erre therin, and al are bound to obey his
  ſentence. 14.~The dutie alſo of a king (whom in future time God wil
  condeſcend to geue them) is deſcribed, with ſpecial charge to receiue
  the law of God at the Prieſtes handes.}

Thou shalt not immolate to our Lord thy God a sheepe, and an oxe,
wherein there is blemish, or any fault: becauſe it is abomination to
our Lord thy God. \V When there shal be found with thee within one of
thy gates, which our Lord thy God shal geue thee, man or woman that do
euil in the ſight of our Lord thy God, and tranſgreſſe his couenant, \V
that they goe and ſerue ſtrange goddes, and adore them, the ſunne and
the moone, and al the hoſte of heauen, which thinges I commanded not: \V
and this is told thee, and hearing it thou haſt inquired diligently, and
found it to be true, and the abomination is committed in Iſrael: \V thou
ſhalt bring forth the man and the woman, that
%%% o-0406
haue committed that moſt heynous thing, to the gates of thy citie, and
they shal be ſtoned. \V At the mouth of two, or three witneſſes shal he
perish that is to be ſlaine. Let no man be killed, one onlie geuing
teſtimonie againſt him. \V The hand of the witneſſes shal be firſt to
kil him, and the hand of the reſt of the people shal be layd on laſt:
that thou mayeſt take away the euil out of the middes of thee. \V
\LNote{If the iudgement be hard.}{For
\MNote{Supreme
\Fix{Iuge}{Iudge}{likely typo, same in other}
of controuerſies.}
a ful and aſſured deciſion of al controuerſies, God here inſtituted to
his people a ſupreme Tribunal, that in caſe inferior Iudges varied in
iudgement, recourſe might be had to the Councel of Prieſtes, where one
chiefe Iudge, the High Prieſt, was appointed to geue ſentence, and al
others commanded to receiue and obey the ſame.}
If thou perceiue that the iudgement with thee be hard and doubtful
betwen bloud and bloud, cauſe and cauſe, leproſie and not leproſie: and
thou ſee that the wordes of the iudges within thy gates doe varye: ariſe,
and goe vp the place, which our Lord thy God shal chooſe. \V And
thou shalt come to the prieſtes of the Leuitical ſtocke, and to
\SNote{In the councel of Prieſtes one ſupreme Iudge, which was the High
Prieſt.
\XRef{v.~12.}}
the iudge, that shal be at that time: and thou shalt aske of them,
\LNote{vvho shal shevv the truth.}{God
\MNote{Sentence of the Iewes conſiſtorie infallible.}
ſo aſſiſted this conſiſtorie with his ſpirite of truth, that their
ſentence was infallible: though otherwiſe they might erre, either in
life, or in priuate opinion. VVherfore, our Sauiour diſtinguiſhing
betwen their publique doctrin, and their workes, taught the people, that
for ſomuch as the Scribes and Phariſees ſate in Moyſes chaire, and yet
tranſgreſſed Gods commandmentes, euerie one should \Emph{obſerue and doe
as they ſaide: but not doe according to their vvorkes.}
\XRef{Mat.~23.}
And S.~Iohn aſcribeth the true ſentence geuen by Caiphas in the councel,
to his office of High Prieſt, ſaying:
\XRef{(Ioan.~11.)}
\MNote{The high prieſt was chief Iudge.}
\Emph{He ſaid not this of him ſelfe, but being the high prieſt of that
yeare, he prophecied that} \Sc{Iesvs} \Emph{should die for the nation
and to gather into one the children of God.} VVhere the high prieſt by
vertue and priuilege of his office, vttered the truth, which him ſelfe
neither meant nor vnderſtood. And this happened when the Law and
Prieſthood of the Iewes was to decline & geue place to Chriſts new
ordinance, and therfore no doubt God euer directed the ſentence of the
high Prieſt: and moſt ſpecially now Chriſt preſerueth the Apoſtolique
See from error in faith, and in general decrees touching manners: yea
though the chiefe viſible Iudge were as wicked as Caiphas.
\MNote{Proteſtantes friuolous euaſion.}
And therfore the Proteſtantes euaſion is friuolous, limiting the prieſts
ſentence to binde the ſubiectes,
\CNote{English Bible. 1603.}
\Emph{ſo long as he is the true miniſter of God, and pronounceth
according to his vvord.} For except God aſſiſted him, that he should
pronounce according to his word, and ſo al men reſt ſatisfied,
ſubmitting them ſelues to his ſentence, the controuerſie ſhould be
endles, and this conſiſtorie nothing worth: but ſtil be new examinations
and new iudgementes, whether the former were according to Gods word or
no.}
who shal shew thee the truth of the iudgement. \V And thou ſhalt do
whatſoeuer they, that are
\SNote{There were not manie preſidentes at once, but in ſucceſſion, one
after an other.}
preſidentes of the place, which our Lord ſhal chooſe, ſhal ſay and teach
thee, \V according to his law; and thou shalt folow their ſentence:
neither shalt thou decline to the right hand nor to the left hand. \V
But
\LNote{He that shal be proude.}{This
\MNote{Pride in priuate opinion puniſhed with death.}
alſo conuinceth, that al were bound to accept of the high prieſts
ſentence, the law condemning him of pride, that \Emph{refuſed to obey
the commandment of the Prieſt, vvhich at that time miniſtred to our
Lord}, and for his diſobenience puniſhing him with death.}
he that shal be proude, refuſing to obey the commandement of the Prieſt,
which at that time
%%% 0453
miniſtreth to our Lord thy God, and the decree of the iudge, that man
shal die, and thou shalt take away the euil out of Iſrael: \V and the
whole people hearing shal feare, that none afterward ſwel in pride. \V
When thou art entred the Land, which our Lord thy God wil geue thee, and
doeſt poſſeſſe it, and dwelleſt in it, and ſayeſt: I wil ſette a king
ouer me, as al nations haue round about: \V him shalt thou ſette, whom
our Lord thy God shal chooſe of the number of thy brethren. A man of an
other nation that is not thy brother, thou canſt not make king. \V And
when he is made, he shal not multiplie to him ſelfe horſes, nor lead
backe the people into Ægypt, taking high courage for the number of his
horſemen, eſpecially whereas our Lord hath commanded you that in no caſe
you returne any more the ſame way. \V He
\SNote{Pluralitie of wiue is not here forbid; for king Dauid
tranſgreſſed not this precept hauing more then one or two: but Salomon
offended in multiplying manie wiues.
\Cite{S.~Aug. q.~17. in Deut.}}
shal not haue manie wiues, that may allure his minde, nor huge weightes
of ſiluer and gold. \V And after he shal ſitte in the throne of his
kingdome, he shal copie to him ſelfe the Deuteronomie of this Law in a
volume,
\SNote{Temporal good Princes take the law, and word of God, at the
Prieſts handes.}
taking the copie of the prieſtes of the Leuitical tribe, \V and he shal
haue it with him, and shal reade it al the dayes of his life, that he
may learne to feare our Lord his God, and keepe his wordes and
ceremonies, that are commanded in the law. \V And that his hart be not
lifted vp into pride ouer his brethren, nor decline to the right ſide or
the left ſide, that he may reigne a long time, and his ſonnes ouer
Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
