%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Mark
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 2393
%%% o-2216
\startChapter[
  title={Chapter 11}
  ]

\Summary{Being now come to the place of his Paſsion, he entreth with
  triumph as their Chriſt. 12.~He curſeth that fruitleſſe leafie
  tree. 15.~He sheweth his zeale for the houſe of God: for which the
  Rulers ſeeke his deſtructiõ. 24.~He exhorteth his Diſciples to
  ſtedfaſtnes of faith, and to forgiue their enemies. 27.~He auoucheth
  his power by the witnes of Iohn who was a man ſent of God.}

And
\CNote{\XRef{Mat.~12,~1.}
\XRef{Lu.~19,~20.}
\XRef{Io.~12,~15.}}
\MNote{The fourth part of this Ghoſpel. The Holy weeke of his Paſsion in
Hieruſalem.}
when they came nigh vnto Hieruſalem and Bethania to Mount-oliuet,
\Fix{ſe}{he}{obvious typo, fixed in other}
ſendeth two of his Diſciples, \V and ſaith to them: Goe into the towne
that is againſt you, and immediatly entring in thither, you ſhal find a
colt tied, vpon which no man yet hath ſitten: looſe him, and bring
him. \V And if any man ſhal ſay to you, what doe you? ſay that he is needful
for our Lord: and incontinent he wil
%%% 2394
ſend him hither. \V And going their waies, they found the colt tied
before the gate without in the meeting of two waies: and they looſe
him. \V And certaine of them that ſtood there, ſaid to them: What doe
you looſing the colt? \V Who ſaid to them as \Sc{Iesvs} had commanded
them; and they did let him goe with them. \V And they brought the colt
to \Sc{Iesvs}; and they lay their garments vpon him, and he ſate vpon
him. \V And
\SNote{Al theſe voluntary duties were grateful to our Sauiour: and ſo be
the like done to him in the B.~Sacrament.}
many ſpred their garments in the way:
%%% o-2217
and others did cut boughes from the trees, and ſtrawed them in the
way. \V And they that went before and they that folowed, cried
ſaying:
\CNote{\XRef{Pſ.~117.~26.}}
\Emph{Hoſanna, bleſſed is he that commeth in the name of our
Lord.} \V \Emph{Bleſſed is the Kingdom of our father Dauid that commeth,
Hoſanna in the higheſt.} \V And he entred Hieruſalem into the temple;
and hauing vewed al things round about, when now the euening houre was
come, he went forth into Bethania with the Twelue.

\V And the next day when they departed from Bethania, he was an
hungred. \V And
\CNote{\XRef{Mt.~21,~19.}}
when he had ſeen a farre off a figtree hauing leaues, he
came if happily he could find any thing on it. And when he was come to
it, he found nothing but leaues. For it was not the time for figges. \V
And anſwering he ſaid to it: Now no man eate fruit of thee any more for
euer. And his Diſciples heard it. \V And they come to Hieruſalem.

And
\CNote{\XRef{Mt.~21,~12.}
\XRef{Luc.~19,~45.}}
when he was entred into the temple, he began to caſt out them
that ſold and bought in the temple, and the tables of the bankers, and
the chaires of them that ſold pigeons he ouerthrew. \V And he ſuffred
not that any man should carie a
\LNote{Veſſel through the Temple}{He
\MNote{Profaning of Gods Church.}
could not abide to ſee the Temple of God profaned, no nor ſuffred thoſe
things to be done in it, which otherwiſe were not vnlawful but honeſt,
if they had been done in due place. How much leſſe can he abide the
profaning of Churches now with Heretical ſeruice and preaching of hereſie
and blaſphemie?}
veſſel through the temple; \V And he taught, ſaying to them: Is it not
written,
\CNote{\XRef{Eſa.~56,~7.}
\XRef{Iere.~7,~11.}}
\Emph{That my houſe shal be called the houſe of prayer to al
Nations? But you haue made it
\LNote{Denne of theeues}{If the Temple was then a denne of theeues,
becauſe of profane &
ſecular marchandiſe: how much more now, when the houſe appointed for the
Holy Sacrifice & Sacrament of the Bodie of Chriſt, is made a denne of
the Miniſters of Caluins bread?}
a denne of theeues.} \V Which when the cheefe Prieſtes and the Scribes
had heard, they ſought how they might deſtroy him. For they were afraid
of him, becauſe the whole multitude was in admiration vpon his
doctrine. \V And when euening was come, he went forth out of the citie.

\V And
\CNote{\XRef{Mt.~21,~20.}}
when they paſſed by in the morning, they ſaw the figtree withered
from the rootes. \V And Peter remembring, ſaid to him: Rabbi, behold the
figtree that thou didſt curſe, is withered. \V And \Sc{Iesvs} anſwering
ſaith to them: Haue
\SNote{Faith of God is to beleeue that he is able, and that he wil doe
it, if it be expedient, and no impediment on our part.}
faith of God. \V Amen I ſay to you, that whoſoeuer shal ſay to this
mountaine, Be taken vp and be caſt into the ſea, and shal not ſtagger in
his hart, but beleeue that whatſoeuer he ſaith shal be done; it ſhal be
done vnto him. \V Therfore I ſay to you, al things, whatſoeuer you aske,
praying, beleeue that you ſhal receaue, and they ſhal come vnto you. \V
And when you ſhal ſtand to pray, forgiue if you haue ought againſt any
man; that alſo your Father which is in Heauen, may forgiue you your
ſinnes. \V
\SNote{God neuer forgiueth ſinne to him that pardoneth not his enemies
frõ his hart whereby it is euidẽt that more is required then only
faith.}
If ſo be that
%%% o-2218
you wil not forgiue, neither wil your Father that is in Heauen, forgiue
you your ſinnes. \V And they come againe to Hieruſalem.

\V And
\CNote{\XRef{Mt.~21,~23.}
\XRef{Luc.~20,~2.}}
when he walked in the temple, there come to him the cheefe
Prieſts & the Scribes and the Ancients, \V and they ſay to him: In what
power doeſt thou theſe things? and who hath giuen thee this power, that
thou ſhouldeſt doe theſe things? \V And \Sc{Iesvs} anſwering ſaid to
them: I alſo wil aske you one word, and anſwer you me, and I
%%% 2395
wil tel you in what power I doe theſe things. \V The Baptiſme of Iohn
was it from Heauen, or from men? anſwer me. \V But they thought with
them ſelues, ſaying: If we ſay, from Heauen; he wil ſay, why then did
you not beleeue him? \V If we ſay, from men, we feare the people. For al
accounted Iohn that he was indeed a Prophet. \V And they anſwering ſay
to \Sc{Iesvs}: We know not. And \Sc{Iesvs} anſwering ſaith to them:
Neither doe I tel you in what power I doe theſe things.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
