%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Mark
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 2369
%%% o-2187
\startChapter[
  title={Chapter 1}
  ]

\Summary{Iohn (the Eremite, of whom the Prophets) preaching pennance,
  and liuing himſelf accordingly, baptizeth the People to prepare them
  to Chriſt, 
  7.~telling them that it is not his, but Chriſts Baptiſme, in which
  they shal receaue the Holy Ghoſt. 9.~\Sc{Iesvs} there is manifeſted
  from Heauen: 12.~and by and by he alſo goeth into the
  wildernes. 14.~Beginning in Galilee, 16.~after that he hath called
  foure Diſciples, 21.~he preacheth firſt in Capharnaum, confirming his
  doctrine with beneficial Miracles, to the great admiration of al:
  35.~then alſo (but firſt retiring into the wildernes) in al the reſt
  of Galilee, with like miracles.}

The
\MNote{The firſt part of this Ghoſpel: of the preparatiõ to Chriſts
manifeſtation.}
beginning of the Ghoſpel of \Sc{Iesvs Christ} the Sonne of God. \V As it
is written in Eſay the Prophet:
\CNote{\XRef{Mal.~3,~1.}
\XRef{Eſa.~40,~3.}}
(\Emph{Behold I ſend mine Angel before
thy face, who shal prepare thy way before thee.}) \V \Emph{A voice of
one crying in the deſert. Prepare ye the way of our Lord, make ſtraight
his pathes.} \V
\CNote{\XRef{Mt.~3,~1.}
\XRef{Lu.~3,~4.}
\Cite{Io.~1,~15.}}
Iohn was in the deſert baptizing, and preaching the Baptiſme of pennance
\SNote{Iohns Baptiſme put them in hope only of remiſſiõ of ſinnes as a
preparatiue to Chriſtes Sacrament by which ſinnes were indeed to be
remitted 
\Cite{Aug. li.~5. de Bapt. c.~10.}}
vnto remiſſion of ſinnes. \V And there went forth to him al the countrie
of Iewrie, and al they of Hieruſalem, and were baptized of him in the
riuer of Iordan,
\LNote{Confeſsing their ſinnes}{A
\MNote{Confeſſion.}
certaine confeſſion of ſinnes there was euen in that pennance which
Iohn preached, and which was made before men were baptized. Wherby it is
cleare that Iohn made a preparation to the Sacramẽt of Pennance which
afterward was inſtituted by Chriſt, as wel as he did by baptizing
prepare the way to Chriſts Baptiſme.}
confeſſing
\LNote{Their ſinnes}{He
\MNote{Particular confeſſion.}
doth not ſay that they confeſſed them ſelues to be ſinners, which may be
 done by a general confeſſion: but that they confeſſed their ſinnes,
 which is a particular confeſſion.}
their ſinnes. \V And Iohn was
\LNote{Clothed}{The
\MNote{Iohns example of pennance.}
Holy Ghoſt thought it worthy of ſpecial reporting how ſtraitly this
Prophet liued, and how he abſtained from delicate meates and
apparel. See
\XRef{Mat.~6.~3.}}
clothed with camels haire, and a girdle of a ſkinne about his loines:
and he did eate locuſtes and wild honie. \V And he preached, ſaying:
There commeth a ſtronger then I after me, whoſe latchet of his ſhoes I
am not worthie ſtouping downe to looſe. \V I haue baptized you
\LNote{With water}{Iohn
\MNote{Baptiſme in water.}
with water only, Chriſt with the Holy Ghoſt, not only, as the
\CNote{\Cite{Calu. li.~4. inſt. c.~15.}}
Heretikes hold, that ſay water is not neceſſary, but with water and the
Holy Ghoſt, as it is plaine
\XRef{Io.~3.}
\Emph{vnles a man be borne againe of water and the Holy Ghoſt, he shal
not enter into the Kingdom of Heauen.}}
with water; but he ſhal baptize you with the Holy Ghoſt.

\V And it came to paſſe, in thoſe daies came \Sc{Iesvs} from Nazareth of
Galilee; and was
\LNote{Baptized of Iohn}{The humility of Chriſt not diſdaining his
ſeruants Baptiſme. Which is an example for al faithful not to diſdaine
Chriſts Sacraments of any Prieſt, be he neuer ſo ſimple, being by the
Catholike Church lawfully called. 
\Cite{Aug. li.~5. de bapt. c.~9.}}
baptized of Iohn in Iordan. \V And forthwith comming vp out of the water, he ſaw
the Heauens opened, and
\LNote{The Spirit}{Expreſſe
\MNote{The B.~Trinitie.}
mention of the B.~Trinitie; the Father ſpeaketh from Heauen, the Holy
 Ghoſt appeareth in the likeneſſe of a doue, the Sonne alſo is
 recommended vnto vs.
 \Cite{Ambroſ. li.~1. de Sacramen. c.~9.}}
the Spirit as a doue deſcending, and remaining on him. \V And a voice
was made
%%% 2370
from Heauen: Thou art my beloued Sonne, in thee I am wel pleaſed.

\V And forthwith
\CNote{\XRef{Mat.~4,~1.}
\XRef{Lu.~4,~1.}}
the Spirit droue him out into
\LNote{Deſert}{Chriſt
\MNote{Chriſts exãple of pennance.}
doing pennance by long faſting, ſolitarineſſe, and conuerſing with wilde
beaſtes, gaue example and inſtruction to the Church for Lent faſt, and to
holy Eremites of retiring them ſelues to the wilderneſſe and prayer.}
the deſert.
%%% o-2188
\V And he was in the deſert fourtie daies, and fourtie nights: and was
tempted of Sathan, and he was with beaſts, and the Angels miniſtred to
him.

\V
\MNote{The ſecond part of this Ghoſpel: of Chriſts manifeſtation.}
And
\CNote{\XRef{Mt.~4,~1.}
\XRef{Lu.~4,~14.}}
after that Iohn was deliuered vp, \Sc{Iesvs} came into Galilee,
preaching the Ghoſpel of the Kingdom of God, \V and ſaying: That the
time is fulfilled, and the Kingdom of God is at hand:
\SNote{He doth not preach beleefe or faith only, but pennance alſo.}
be penitent, and beleeue the Ghoſpel.

\V
\CNote{\XRef{Mt.~4,~18.}
\XRef{Lu.~5,~2.}}
And paſſing by the ſea of Galilee, he ſaw Simon and Andrew his brother,
caſting nets into the ſea (for they were fiſhers) \V and \Sc{Iesvs} ſaid
to them: Come after me, and I wil make you to become fiſhers of men. \V
And immediatly leauing their nets, they folowed him. \V And being gone
thence a litle further, he ſaw Iames of Zebedee, and Iohn his brother,
and them repairing their nets in the ſhip: \V and forthwith he called
them. And leauing their father Zebedee in the ſhip with his hired men,
they folowed him.

\V And
\CNote{\XRef{Mt.~4,~13.}
\XRef{Lu.~4,~31.}}
they enter into Capharnaum, and he forthwith vpon the Sabboths
going into the Synagogue, taught them. \V And they were aſtoniſhed at his
doctrine. For he was teaching them as hauing power, and not as the
Scribes. \V And
\CNote{\XRef{Lu.~4,~32.}}
there was in their Synagogue a man in an vncleane
Spirit; and he cried out, \V ſaying: What to vs and to thee \Sc{Iesvs}
of Nazareth? art thou come to deſtroy vs? I know who thou art, the
Sainct of God. \V And \Sc{Iesvs} threatned him, ſaying: Hold thy peace,
and goe out of the man. \V And the vncleane Spirit tearing him, and
crying out with a great voice, went out of him. \V And they marueled al,
in ſo much that they queſtioned among them ſelues, ſaying: What thing is
this? what is this new doctrine? for with power he commandeth the
vncleane Spirits alſo, and they obey him. \V And the bruit of him went
forth incontinent into al the countrie of Galilee.

\V And immediatly
\CNote{\XRef{Mt.~8,~14.}
\XRef{Lu.~4,~42.}}
going forth out of the Synagogue, they came into the
houſe of Simon and Andrew, with Iames and Iohn. \V And Simons wiues
mother lay in a fit of a feuer: and forthwith they tel him of her. \V And
comming neere he lifted her vp taking her by the hand: and incontinent
the ague left her, and ſhe miniſtred vnto them. \V And when it was
euening after ſunne ſet, they brought to him al that were il at eaſe and
that had Diuels. \V And al the citie was
%%% o-2189
gathered togeather at the doore. \V And he cured many that were vexed
with diuerſe diſeaſes; and he caſt out many Diuels, and he ſuffred not
them to ſpeake that they knew him.

\V And riſing very early, and going forth he went into
\LNote{Deſert place}{Chriſt
\MNote{Solitarie contemplation.}
vſed very often to retire into ſolitary places, no doubt for our example
 to teach vs that ſuch places are beſt for prayer and contemplation, and
 that we ſhould often retire our ſelues from worldly matters to ſolitary
 meditation of heauenly things.}
a deſert place: and there he prayed. \V  And Simon ſought after him, and
they that were with him. \V And when they had found him, they ſaid to
him, That al ſeeke for thee. \V And he ſaith to them: Let vs goe into
the next townes and cities, that I may preach there alſo; for to this
porpoſe am I come.

\V And he was preaching in their Synagogues, and in al Galilee, and
caſting out Diuels. \V And a
\CNote{\XRef{Mt.~8,~2.}
\XRef{Lu.~5,~12.}}
leper commeth to him beſeeching him;
%%% 2371
and kneeling downe ſaith to him: If thou wilt, thou canſt make me
cleane. \V And \Sc{Iesvs} hauing compaſſion of him, ſtretched forth his
hand; and touching him, he ſaith vnto him: I wil. Be thou made
cleane. \V And when he had ſpoken, immediatly the leproſie departed from
him, and he was made cleane. \V And he threatned him, and forthwith caſt
him forth. \V And he ſaith to him: See thou tel no body, but goe, ſhew
thyſelfe
\SNote{Our Sauiour euen when he healed the leper by extraordinarie
miraculous power, would yet for order ſake ſend the partie to the
Prieſt.}
to the high Prieſt, and offer for thy cleanſing the things that Moyſes
\CNote{\XRef{Leu.~14,~3.}}
commanded, for a teſtimonie to them. \V But he being gone forth, began
to publiſh, and to blaſe abrode the word: ſo that now he could not
openly goe into the citie, but was abrode in deſert places, and they
came togeather vnto him from al ſides.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
