%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Mark
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 2388
%%% o-2209
\startChapter[
  title={Chapter 9}
  ]

\Summary{The more to confirme them, he giueth them in his
  Transfiguration a ſight of his glorie, wherevnto ſuffering doth bring,
  9.~and then againe doth inculcate his Paſsion. 14.~A Diuel alſo he
  caſteth out, which his Diſciples (vpon whom therefore the peruerſe
  Scribes triumphed in his abſence) could not, for lack of faſting and
  praying. 30.~Being yet in Galilee, he reuealeth more about his
  Paſsion. 33.~And (becauſe in the way to Capharnaum they contended for
  the Primacie) he teacheth them that humility is the way to Primacie
  before God: 38.~bidding them alſo, not to prohibit ſuch as be not
  againſt them: nor to giue ſcandal to any one of the faithful: and on
  the other ſide, the faithful to auoid them by whom they may be
  ſcandalized and fal, be they neuer ſo neere vnto them.}

%%% o-2210
And
\CNote{\XRef{Mat.~16,~27.}
\XRef{17,~1.}
\XRef{Lu.~9,~27.}}
he ſaid to them: Amen I ſay to you, that there be ſome of them that
ſtand here, which ſhal not taſt of death, vntil they ſee the Kingdom of
God comming in power. \V
\SNote{See the
\XRef{Annotations vpon the 17.~of S.~Matthew.}}
And after ſix daies \Sc{Iesvs} taketh Peter and Iames and Iohn, and
bringeth them alone into a high mountaine apart, and was transfigured
before them. \V And his garments were made gliſtering and white
exceedingly as ſnow, the like wherof a fuller can not make white vpon
the earth. \V And there appeared to them
%%% !!! This note is not marked in either version, only the short note
%%% is marked.  How should we handle this double note at one location?
\LNote{Elias with Moyſes}{Moyſes
\MNote{Elias. Moyſes.}
repreſenteth the Perſons of al Saints that shal be departed this life
when Chriſt commeth in his Maieſtie to iudgement: And Elias (who was
then liuing) figureth the holy men that ſhal then be found aliue when he
commeth in glorie. Who both ſhal then begin to reigne
\Fix{which}{with}{obvious typo, fixed in other}
Chriſt in glorie.
\Cite{Beda in 9.~Marc.}}
\SNote{The law and the Prophets ioyne with Chriſt and his Ghoſpel: the
one ſignified by Moyſes,
\Fix{he}{the}{obvious typo, fixed in other}
other by Elias. By whoſe apparitions here we alſo learne that ſometime
there may be perſonal entercourſe betwixt the liuing & the dead, though
not ordinarily.}
Elias with Moyſes; and they were talking with \Sc{Iesvs}. \V And Peter
anſwering, ſaid to \Sc{Iesvs}: Rabbi, it is good for vs to be here; and
let vs make three tabernacles, one for thee, and one for Moyſes, and
one for Elias. \V For he knew not what he ſaid; for they were frighted
with feare. \V And there was a cloud ouerſhadowing them, and a voice
came out of the cloud, ſaying: This is my Sonne moſt deare: heare ye
him. \V And immediatly looking about, they ſaw no man any more
but \Sc{Iesvs} only with them. \V And as they deſcended from the
mountaine, he commanded them that they ſhould not tel any man what
%%% 2389
things they had ſeene, but when the Sonne of man ſhal be riſen againe
from the dead. \V And they kept in the word with them ſelues;
queſtioning togeather what that ſhould be, \Emph{when he shal be riſen
from the dead.}

\V And they asked him, ſaying: What ſay the Phariſees then and the Scribes,
\CNote{\XRef{Mal.~4,~5.}}
that Elias muſt come firſt? \V Who anſwering ſaid to them: Elias when he
commeth firſt, ſhal reſtore al things: and
\Var{how}{as}
it is written of the Sonne of man, that he ſhal ſuffer many things and
be contemned. \V But I ſay to you that
\LNote{Elias alſo is come}{Elias
\MNote{Elias and Iohn Baptiſt Eremites.}
was Zelous for Gods Law, a great reprehender of ſinne, and an Eremite,
and ſhal be the Precurſour of Chriſt in his ſecond Aduent: So was Iohn
before his firſt Aduent, a Zelatour, a Corrector, an Eremite, and his
Precurſour.
\Cite{Theod. in Caten. Thoma ſuper hunc locum.}
See S.~Hierom in the life of Paul the Eremite, that both Elias and Iohn
Baptiſt were counted principal profeſſours of that life.}
Elias alſo is come (and they haue done to him whatſoeuer they would) as
it is written of him.

\V And
\CNote{\XRef{Mt.~17,~14.}
\XRef{Lu.~9,~21.}}
comming to his Diſciples, he ſaw a great multitude about them,
and the Scribes queſtioning with them. \V And forthwith al the People
ſeeing \Sc{Iesvs}, was aſtoniſhed, and much afraid: and running to him,
ſaluted him. \V And he asked them, what doe you queſtion of among
you? \V And one of the multitude anſwering, ſaid: Maiſter, I haue
brought my ſõne to thee, hauing a dumme Spirit, \V Who whereſoeuer he
taketh him, daſheth him, and he fometh, and gnaſheth with the teeth, and
withereth: and I ſpake to thy Diſciples to caſt him out, and they could
not. \V Who anſwering them,
%%% o-2211
ſaid: O incredulous Generation, how long ſhal I be with you? how long
ſhal I ſuffer you? bring him vnto me. \V And they brought him. And when
he had ſeen him, immediatly the Spirit troubled him; and being
throwen vpon the ground, he tumbled foming. \V And he asked his father:
How long time is it ſince this hath chanced vnto him? But he ſaid: From
his infancie: \V And oftentimes hath he caſt him into fire and into waters,
to deſtroy him. But if thou canſt any thing, help vs, hauing compaſſion
on vs. \V And
\Fix{Ieſus}{\Sc{Iesvs}}{obvious typo, same in other}
ſaid to him: If thou canſt beleeue, al things are poſſible to him that
beleeueth. \V And incontinent the father of the boy crying out, with
teares ſaid: I doe beleeue Lord; help my incredulity. \V And
when \Sc{Iesvs} ſaw the People running togeather, he threatned the
vncleane Spirit, ſaying to him: Deafe and dumme Spirit, I command thee,
goe out of him, and enter not any more into him. \V And crying out, and
greatly tearing him, he went out of him, and he became as dead, ſo that
many ſaid, that he is dead. \V But \Sc{Iesvs} holding his hand, lifted
him vp; and he roſe. \V And when he was entred into the houſe, his
Diſciples ſecretly asked him, why could not we caſt him out? \V And he
ſaid to them: This kind can goe out by nothing but
\SNote{Note the great force of prayer, and faſting.}
by prayer and faſting.

\V And
\CNote{\XRef{Mt.~18,~1.}
\XRef{Lu.~9,~46.}}
departing thence they paſſed by Galilee, neither would he that
any man ſhould know. \V And he taught his Diſciples, and ſaid to them:
That the Sonne of man shal be betrayed into the hands of men, and they
ſhal kil him, and being killed the third day he shal riſe againe. \V But
they knew not the word: and they were afraid to aske him.

\V And they came to Capharnaum. Who, when he was in the houſe, asked
them, what did you treat of in the way? \V But they held their peace,
for in the way they had diſputed among them ſelues, which of them should
be the greater. \V And ſitting downe, he called the Twelue, and ſaith to
them: If any man wil be firſt, he shal be laſt of al, and the miniſter
of al. \V And taking a child, he ſet him in the middes of them. Whom
when he had embraced, he ſaid to them: \V
%%% 2390
Whoſoeuer ſhal receaue one of ſuch children in my name, receaueth
me. And whoſoeuer ſhal receaue me, receaueth not me, but him that ſent
me.

\V
\CNote{\XRef{Luc.~9,~49.}}
Iohn anſwered him, ſaying: Maiſter we ſaw one caſting out Diuels
\LNote{In thy name}{Miracles
\MNote{The name of \Sc{Iesvs} worketh miracles.}
are wrought ſometime by the name of \Sc{Iesvs}, whatſoeuer the men be,
when it is for the proofe of a truth or for the glorie of God. In ſo
much that Iulian the Apoſtata himſelfe did driue away Diuels with the
ſigne of the Croſſe: as \Emph{S.~Gregorie Nazianzene} writeth
\Cite{Orat.~1. in Iulian.}
\Cite{Theodoret. li.~3. c.~3. hiſt.}
And ſo alſo Heretikes may doe miracles among the Heathen, to proue any
article of the Chriſtian faith; but they neuer did nor euer ſhal worke
any miracle to proue any of their erroneous opinions; as, to proue that
Chriſt is not really in the B.~Sacrament.}
in thy name, who foloweth not vs, and we
%%% o-2212
prohibited him. \V But \Sc{Iesvs} ſaid: Doe not prohibit him. For there
is no man that doth a miracle in my name, and can ſoone ſpeake il of
me. \V For he that is not againſt you, is for you. \V For whoſoeuer shal
giue you to drinke a cup of water in my name, becauſe you are Chriſts;
amen I ſay to you, he shal not loſe his
\SNote{Reward for almes-deeds, whereby it is euidẽt they be
meritorious.}
reward. \V And whoſoeuer ſhal
\SNote{To giue ſcãdal by our life to the weake in faith, is a great
ſinne, ſpecially in
\Fix{Preſts}{Prieſts,}{obvious typo, fixed in other}
Preachers, & Princes.}
ſcandalize one of theſe litle ones beleeuing in me; it is good for him
rather if a milſtone were put about his necke, and he were caſt into the
ſea. \V And if thy hand ſcandalize thee, cut it off. It is good for thee
to enter into life, maimed, then hauing two hands to goe into Hel, into
the fire vnquencheable, \V where their worme dieth not, and the fire
quencheth not. \V And if thy foot ſcandalize thee, chop it off. It is
good for thee to enter into life euerlaſting, lame, rather then hauing
two feet, to be caſt into the Hel of vnquencheable fire, \V where their
worme dieth not, and the fire quencheth not. \V And if thine eye
ſcandalize thee, caſt it out. It is good for thee with one eye to enter
into the Kingdom of God, rather then hauing two eyes, to be caſt into
the Hel of fire, \V where their worme dieth not, and the fire quencheth
not. \V For euery one shal be ſalted with fire: and
\CNote{\XRef{Leu.~2,~13.}}
euery victime ſhal
be ſalted with ſalt. \V Salt is good. But if the ſalt ſhal be
vnſeaſoned; wherwith wil you ſeaſon it? Haue ſalt in you, and haue peace
among you.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
