%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Mark
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 2395
%%% o-2218
\startChapter[
  title={Chapter 12}
  ]

\Summary{He foretelleth to the Iewes in a parable their reprobation moſt
  worthy, and the vocation of the Church of the Gentils in their place:
  10.~himſelfe being the headſtone thereof. 13.~He defeateth the ſnare
  of the Phariſees and Herodians, about paying tribute to Cæſar:
  18.~anſwereth alſo the inuention of the Sadducees againſt the
  Reſurrection: 28.~alſo the oppoſition of a Scribe. 35.~And ſo hauing
  put al the buſie Sectes to ſilence, he turneth and poſeth them on the
  other ſide, becauſe they imagined Chriſt should be no more but a
  man. 38.~Bidding the people to beware of the Scribes, being ambitious
  and Hypocrites. 41.~He commendeth the poore widow for her two mites,
  aboue al.}

And
\CNote{\XRef{Eſa.~5,~1.}
\XRef{Mt.~21,~33.}
\XRef{Lu.~20,~9.}}
he began to ſpeake to them in parables:
\SNote{This man, is God the Father, this vineyard, is (as Eſay ſaith
5.~1.) the houſe of Iſrael. The ſeruants ſent, are Moyſes and the
Prophets, whõ the Iewes did diuerſely afflict and perſecute.}
A man planted a vineyard and made a hedge about it, and digged a trough,
and built a towre, and lett it out to husbandmen; and went forth into a
ſtrange countrie. \V And ſent to the husbandmen in ſeaſon a ſeruant, to
receaue
%%% o-2219
of the husbandmen, of the fruit of the vineyard. \V Who
apprehending him, beatt him; and ſent him away emptie. \V And againe he
ſent to them another ſeruant; and him they wounded in the head, and vſed
him reprochfully. \V And againe he ſent another, & him they killed: and
many other, beating certaine, and killing others. \V Therefore hauing
yet one
\SNote{His Sonne is Chriſt our Sauiour, whom the Iewes crucified out of
the citie of Ieruſalem, as it were caſting him out of the vineyard.}
ſonne moſt deare; him alſo he ſent vnto them laſt ſaying: That they wil
reuerence my ſonne. \V But the husbandmen ſaid one to an other: This is
the heire; come, let vs kil him; and the inheritance shal be ours. \V And
Apprehending him, they killed him, and caſt
%%% 2396
him forth out of the vineyard. \V What therfore wil the Lord of the
vineyard doe?
\SNote{The Iewes and their guides to whom the vineyard was ſet,
deſtroied: and Gods vineyard giuen to the Apoſtles and their Succeſſours
in the Gentils.}
He wil come and deſtroy the husbandmen; and wil giue the vineyard to
others. \V Neither haue you read this ſcripture,
\CNote{\XRef{Pſ.~117,~22.}}
\Emph{The
\SNote{Chriſt is become the corner ſtone of the Synagogue and the Church
in which the faithful both of the Iewes and Gentils are conteined.}
ſtone which the builders reiected; the ſame is made the head of the
corner: \V By our Lord was this done, and it is maruelous in our
eyes}? \V And they ſought to lay hãds on him, and they feared the
multitude. For they knew that he ſpake this parable to them. And leauing
him they went their way.

\V
\CNote{\XRef{Mt.~22,~15.}
\XRef{Lu.~26,~20.}}
And they ſend to him certaine of the Phariſees and of the Herodians;
that they ſhould entrap him in his word. \V Who comming, ſay to him:
Maiſter, we know that thou art a true ſpeaker, and careſt not for any
man; for thou doeſt not looke vpon the perſon of men, but teacheſt the
way of God in truth. Is it lawful to giue tribute to Cæſar; or ſhal we
not giue it? \V Who knowing their ſubteltie, ſaid to them: Why tempt you
me? bring me a penie that I may ſee it. \V But they brought it him. And
he ſaith to them: Whoſe is this image, and inſcription; They ſay to him,
Cæſars. \V And \Sc{Iesvs} anſwering, ſaid to them: Render therfore the
things that are Cæſars, to Cæſar; and that are Gods,
\LNote{To Gods}{Theſe
\MNote{God firſt to be ſerued & then the Prince.}
men were very circumſpect and wary to doe al duties to Cæſar, but of
their dutie to God they had no regard. So Heretikes, to flatter temporal
Princes, and by them to vphold their Hereſies, doe not only inculcate
mens dutie to the Prince, diſſembling that which is due to God; but alſo
giue to the Prince more then due, and take from God his right and
dutie. But Chriſt alowing Cæſar his right, warneth them alſo of their
dutie toward God. And that is it which Catholikes inculcate: Obey God,
doe as he commandeth; Serue him firſt, and then the Prince.}
to God. And they marueled at him.

\V And
\CNote{\XRef{Mt.~22,~23.}
\XRef{Luc.~20,~27.}
\XRef{Act.~23,~6.}
\XRef{Deu.~25,~5.}}
there came to him the Saducees that ſay there is no reſurrection;
and they asked him ſaying: Maiſter, \V Moyſes wrote vnto vs, that if any
mans brother die, and leaue his wife, and leaue no children,
\LNote{His brother shal take}{Marke
\MNote{Marying the brothers wife.}
wel here, that the Law which ſaith,
\CNote{\XRef{Leu.~18,~16.}}
Thou shalt not marie thy brothers wife,
\Fix{it}{is}{obvious typo, fixed in other}
not ſuch as admitteth not diſpenſation, as though this mariage
were againſt nature. For here the ſame Law ſaith, that in ſome caſe,
\CNote{\XRef{Deu.~25,~5.}}
the
brother not only might, but then was bound to marie his brothers wife.}
his brother ſhal take his wife and raiſe vp ſeed to his brother. \V
There were therfore ſeauen brethren; and the firſt took a wife, and died
leauing no iſſue. \V And the ſecond took her, and died: and neither this
left iſſue. And the third in like manner. \V And the ſeauen took her in
like ſort; and did not leaue iſſue. Laſt of al the woman alſo
%%% o-2220
died. \V
In the reſurrection therfore when they ſhal riſe againe, whoſe wife ſhal
ſhe be of theſe? for the ſeauẽ had her to wife. \V And \Sc{Iesvs}
anſwering, ſaid to them: Doe ye not therefore erre,
\LNote{Not knowing the Scriptures}{Who
\MNote{Many truths deduced out of ſcripture which Heretikes ſee not.}
would haue thought that by this place of Scripture alleaged by Chriſt,
the Reſurrection were proued? and yet we ſee that Chriſt doth hereby
deduce it, and chargeth theſe great Doctours and Maiſters (which arrogated to
them ſelues the knowledge of Scriptures) that it is their ignorance,
that they knew not ſo to deduce it. No maruel then if the Holy Doctours
and Catholike Church make the like deduction ſometime and proofes, where
the Heretike doth not or wil not ſee ſo much, therfore no doubt, becauſe
he knoweth not the Scriptures, whereof he boaſteth ſo much, nor the
ſenſe of the
\Fix{holy}{Holy}{obvious typo, fixed in other}
Ghoſt in them. For example, when of that place, \Emph{It shal not be
forgiuen in this world not in the world to come}, anciẽt Fathers deduce,
that there are ſinnes remitted after this life in Purgatorie. See
\XRef{Mat. 12.~32.}}
not knowing the ſcriptures, not
\LNote{The power of God}{Euen
\MNote{Heretikes ignorant and without faith.}
ſo doe heretikes erre two waies: becauſe they know not the Scriptures,
which they interpret contrarie to the ſenſe of the whole Church and of
al the ancient Fathers, and becauſe they know not the power of God, that
as he is able to raiſe the ſelfe ſame bodies againe, ſo he can make his
body preſent in many places: but being altogether faithleſſe and not beleeuing
his power, they diſpute of al ſuch matters, only by reaſon and their
owne imagination.}
the power of God? \V For when they ſhal riſe againe frõ the dead, they
shal neither marie, nor be maried, but are as the Angels in Heauen. \V
And as concerning the dead, that they doe riſe againe, haue you not read
in the booke of Moyſes, how in the buſh God ſpake to him, ſaying:
\CNote{\XRef{Exo.~3,~6.}}
\Emph{I
am the God of Abraham, and the God of Iſaac, and the God of Iacob}? \V
He is not the God of the dead, but of the liuing. You therfore are much
deceaued.

\V And
\CNote{\XRef{Mt.~22,~34.}}
there came one of the Scribes that had heard them queſtioning,
and ſeeing that he had wel anſwered them, asked him which was the firſt
cõmandement of al. \V And \Sc{Iesvs} anſwered him: That the firſt
cõmandemẽt of al is,
\CNote{\XRef{Deu.~6,~5.}}
\Emph{Heare Iſrael: the Lord thy God is one God. \V
And thou shalt loue the Lord thy God from thy whole hart, and with thy
whole ſoul, and with thy whole mind, and with thy whole power.} This is
the firſt cõmandement. \V And the ſecond is like to it:
\CNote{\XRef{Leu.~19,~18.}}
\Emph{Thou shalt
loue thy neighbour as thy ſelf.} And other commandement greater then
theſe there is not. \V And the Scribe ſaid to him: Wel Maiſter, thou
haſt ſaid in truth, that there is one God, and there is none other
beſides him. \V And that he be loued from the whole hart, and with the
whole vnderſtanding, and with the whole ſoule, and with the whole
ſtrength: and
\SNote{This excellencie of Charitie teacheth vs that faith only is not
ſufficient.}
to loue his
%%% 2397
neighbour as himſelf, is a greater thing then al Holocauſts and
Sacrifices. \V And \Sc{Iesvs} ſeeing that he had anſwered wiſely, ſaid
to him: Thou art not farre from the Kingdom of God. And no man now durſt
aske him.

\V And
\CNote{\XRef{Mt.~12,~41.}
\XRef{Lu.~20,~41.}}
\Sc{Iesvs} anſwering, ſaid, teaching in the temple: How doe the
Scribes ſay, that Chriſt is the ſonne of Dauid? \V For Dauid himſelf
ſaith in the Holy Ghoſt:
\CNote{\XRef{Pſ.~109,~1.}}
\Emph{Our Lord ſaid to my Lord, ſit on my right
hand, vntil I put thine enemies the foote ſtoole of thy feete.} \V Dauid
therfore himſelf calleth him Lord, and whence is he his ſonne? and a great
multitude heard him gladly. \V And he ſaid to them in his doctrine:
%%% !!! Missing CNote? Mark exists, but no marginal note
Take
heed of the Scribes that wil walke in long robes, and be ſaluted in the
market-place, \V And ſit in the firſt chaires in the Synagogues, and
loue the higheſt places at ſuppers: \V Which deuoure widowes houſes
vnder the pretence of long prayer: theſe
%%% o-2221
shal receaue larger iudgement.

\V
\CNote{\XRef{Luc.~21,~2.}}
And \Sc{Iesvs} ſitting ouer againſt the treaſurie, beheld how the
multitude did caſt money into the treaſurie, and many rich men did caſt
in much. \V And when there came a certaine poore widow, she caſt in two
mites, which is a farthing. \V And calling his Diſciples togeather, he
ſaith to them: Amen I ſay to you, that this poore widow hath caſt in
\SNote{God doth accept almes that are correſpondent to euery mans
abilitie: and the more able, the more muſt a man giue.}
more then al that haue caſt into the treaſurie. \V For al they of their
aboundance haue caſt in; but she, of her penurie hath caſt in al that
she had, her whole liuing.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
