%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Mark
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 2376
%%% o-2195
\startChapter[
  title={Chapter 4}
  ]

\Summary{The parables (in which he ſpeaketh to the Iewes becauſe they
  were reprobate) he expoundeth to his Diſciples, shewing that in his
  ſowing, three parts of foure shal perish, through the fault of the
  hearers. 21.~And that his ſeruants muſt confeſſe their faith, 24.~and
  vſe their guifts (contrary to thoſe ſtony and thorny
  \Fix{heares.)}{hearers.)}{obvious typo, fixed in other}
  And that his Church (notwithſtanding the loſing of thoſe three parts
  of the ſeed) shal be brought by his prouidence to the harueſt, that
  is, to the end of the world: 30.~growing ouer al in time, though in
  the beginning it be as the litle muſtard-ſeed, 35.~and though ſuch
  tempeſts of perſecution in the ſea of this world doe riſe againſt it.}

And
\CNote{\XRef{Mt.~13,~1.}
\XRef{Lu.~8,~4.}}
againe he began to teach at the ſea ſide; and a great multitude was
gathered togeather vnto him ſo that he went vp into a boat, and ſate in
the ſea, and al the multitude about the ſea was vpon the land. \V And he
taught them in parables many things, and ſaid to them in his
doctrine: \V Heare ye; Behold, the ſower went forth to ſow. \V And
whiles he ſoweth, ſome fel by the way ſide, and the foules of the aire
came, and did eate it. \V And otherſome fel vpon rocky places where it
had not much earth: and it ſhot vp immediatly, becauſe it had not
deepnes of earth: \V and when the ſunne was riſen, it parched, and
becauſe it had not roote, it withered. \V And ſome fel among thornes:
and the thornes grew vp, and choked it, and it yealded not fruit. \V And
ſome fel vpon good ground; and it yealded fruit that grew vp &
increaſed, and it brought forth, one thirtie, one ſixtie, and one an
hundred.

\V And he ſaid: He that hath eares to heare, let him heare. \V And when
he was alone, the Twelue that were with him, asked him the parable. \V
And he ſaid to them: To you it is giuen to know the myſterie of the
Kingdom of God; but to
\SNote{Such as be out of the Church, though they heare and read neuer
ſo much, they cannot vnderſtand.
\Cite{Bed. in 4.~Marc.}}
them that are without, al things are done in parables: \V that
\CNote{\XRef{Eſa.~6,~5.}}
ſeeing they may ſee, and not ſee; and hearing they may heare, and not
vnderſtand:
\LNote{Leſt they should be conuerted}{Theſe
\MNote{God is not authour of ſinne, but mans owne wil.}
ſpeaches here and els where, we muſt not ſo vnderſtand as though he ſpake
 in parables of purpoſe & to this end, that the
\Fix{heares}{hearers}{obvious typo, fixed in other}
might not vnderſtand, leſt they ſhould be conuerted; which were as much
 to ſay as that he would not haue them vnderſtand, nor be conuerted: but we
 muſt learne the true ſenſe of this very place in S.~Matthew and in the
 Actes, where our Sauiour and S.~Paul ſpeake thus:
\CNote{\XRef{Mt.~13.}
\XRef{Act.~28.}}
\Emph{They haue heard
 heauily, and haue shut their eyes, leſt perhaps they may ſee, and
 vnderſtand, and be cõuerted, and I heale them.} Wherby it is euidẽt,
 that the ſpeaking in parables was not the cauſe (for many beſide the
 Apoſtles heard and vnderſtood) but them ſelues would not heare, and
 vnderſtand, and be conuerted: and ſo were the cauſe of their owne
 wilful & obſtinate infidelity. And therfore alſo he ſpake inparables,
 becauſe they were not worthy to vnderſtand, as the other to whom he
 expounded them.}
leſt at any time they ſhould be conuerted and their ſinnes be forgiuen
them. \V And he ſaith to them: Doe you not know this parable? And how
ſhal you know al parables? \V He that ſoweth, ſoweth the Word. \V And
they by the way ſide, are theſe: where the Word is ſowen, and when they
ſhal haue heard, immediatly commeth Satan, and taketh away the Word that
was ſowen in their harts. \V And they likewiſe that are ſowen vpon the
rocky places, are theſe: who when they heare the Word, immediatly with
ioy receaue it; \V And they haue not root
%%% o-2196
in them ſelues, but are
temporal: afterward when tribulation is riſen and perſecution for the
Word, forthwith they are ſcandalized. \V And other there be that are
ſowen among thornes: theſe are they that heare the Word, \V and the
cares of the World and the deceitfulnes of riches, and concupiſciences
about other things entring in choke the Word, and it is made
fruitles. \V And theſe are they that were ſowen vpon the good ground, which
heare the Word, and receaue it, and yeald fruit one thirtie, one ſixtie,
and
%%% 2377
one an hundred.

\V And he ſaid to them:
\CNote{\XRef{Luc.~8,~16.}}
Cõmeth a candle to be put
\SNote{Chriſt came not to teach his doctrine in corners and hucker
mucker as Heretikes doe, but to lighten the whole world therewith.}
vnder a buſhel, or vnder a bed? and not to be put vpon the
candleſtick? \V For there is nothing hid, which ſhal not be made
manifeſt: neither was any thing made ſecret, but that it ſhal come to
light. \V If any man haue eares to heare, let him heare.

\V And he ſaid to them: See what you heare.
\CNote{\XRef{Luc.~8,~18.}}
In what meaſure you mete, it
ſhal be meaſured to you againe, and more ſhal be giuen
\Var{to you.}{to you that heare.}
\V For he that hath, to him ſhal be giuen: and he that hath not, that
alſo which he hath, ſhal be taken away from him.

\V
\CNote{\XRef{Mt.~13,~31.}
\XRef{Lu.~13,~19.}}
And he ſaid: So is the Kingdom of God, as if a man caſt ſeed into the
earth, \V
\LNote{And ſleep}{The Church, and Chriſts doctrine, (ſleep we, wake we)
increaſeth by the great prouidence of God, only the Preachers muſt ſow,
 and plant, and water, and
\CNote{\XRef{1.~Cor.~3.}}
God wil giue the increaſe, nouriſhing the
 ſeed in mens harts. And therfore we may not giue ouer, or be impatient
 and ſolicitous, if we haue not alwaies good ſucceſſe: but doing our
 duty, commit the reſt to God.}
and ſleep, and riſe vp night and day, and the ſeed ſpring, and grow vp
whiles he knoweth not, \V For the earth of it ſelf bringeth forth fruit,
firſt the blade, then the eare, afterward the ful corne in the eare. \V
And when the fruit hath brought out it ſelf, immediatly he putteth in
the ſickle, becauſe harueſt is come.

\V And he ſaid: To what ſhal we liken the Kingdom of God? or to what
parable ſhal we compare it? \V As a
\LNote{Muſtard ſeed}{If
\MNote{The Church viſibly increaſing.}
the Church and Truth had more and more decaied and been obſcured after
the Apoſtles time vnto ours, as the Heretikes hold: thẽ had it been
great in the beginning, & ſmal afterward: where this Parable ſaith
contrary, that it was a muſtard ſeed firſt, and afterward a great
tree. vide
\Cite{Chryſ. to.~5. contra Gentils in vita S.~Babyla Mart.}}
muſtard ſeed; which when it is ſowen in the earth, is leſſe then al the
ſeeds that are in the earth: \V And when it is ſowen, it riſeth vp, and
becometh greater then al herbes, and maketh great boughes, ſo that
\LNote{The birds}{Of
\MNote{Chriſtian religion wonderfully ſpreading.}
al Sectes or doctrine, Chriſts religion at the beginning was the
 ſmalleſt, and moſt contemptible; but the ſucceſſe therof farre paſſed
 al mans doctrine: in ſo much that afterward al the wiſeſt and greateſt
 of the world made their reſidence and reſt therein.}
the birdes of the aire may dwel vnder the ſhadow thereof.

\V And with many ſuch parables he ſpake to them the Word, according as
they were able to heare: \V and without parable he did not ſpeake vnto
them; but apart, he explicated al things to his Diſciples.

\V And he ſaith to them in that day, when euening was come:
\CNote{\XRef{Mat.~8,~23.}
\XRef{Luc.~8,~22.}}
Let vs paſſe
ouer to the other ſide. \V And dimiſſing the multitude, they take him ſo
as he was in the boat: and there
%%% o-2197
were other boats with him. \V And there aroſe a great ſtorme of wind,
and the waues beatt into the boat, ſo that the boat was filled. \V And
he was in the hinder part of the boat ſleeping vpon a pillow; and they
raiſe him, and ſay to him: Maiſter, doth it not pertaine to thee that we
perish? \V And riſing vp he threatned the wind, and ſaid to the ſea:
Peace, be ſtil. And the wind ceaſed; and there was made a great
calme. \V And he ſaid to them: Why are you feareful? neither yet haue
you faith? And they feared with great feare: and they ſaid one to an
other, who is this (thinkeſt thou) that both wind and ſea obey him?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
