%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Mark
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 2403
%%% o-2228
\startChapter[
  title={Chapter 15}
  ]

\Summary{The cheefe of the Iewes accuſe him to Pilate the Gentil. 6.~And
  (he ſeeking to deliuer him) they perſwade the common People (who
  hitherto were alwaies ready to defend him) not only to preferre the
  murderer Barabbas, but alſo to crie Crucifige (to the reprobation of
  the whole Nation,) 16.~After many illuſions, 20.~he is crucified by
  the Gentils. 29.~which the Iewes ſeing, doe triumph as if they had now
  the victorie. 33.~But euen then by many wonderful workes he declareth
  his might, 42.~and finally is buried honourably.}

And
\CNote{\XRef{Mt.~27,~1.}
\XRef{Lu.~23,~1.}
\XRef{Io.~18,~28.}}
forthwith in the morning the cheefe Prieſts with the Ancients and
the Scribes & the whole Councel, conſulting togeather,
binding \Sc{Iesvs} led & deliuered him to Pilate. \V And Pilate asked
him: Art thou the King of the Iewes? but he anſwering, ſaid to him: Thou
ſaieſt. \V And the cheefe Prieſts accuſed him in many things. \V And
Pilate againe asked him, ſaying: Anſwereſt thou nothing? ſee in how
great things they accuſe thee. \V But \Sc{Iesvs} anſwered nothing more;
ſo that Pilate marueled.

\V And vpon the feſtiual day he was wont to releaſe vnto them one of the
priſoners whomſoeuer they had demanded. \V And there was one called
Barabbas, which was put in priſon with ſeditious perſons, who in a
ſedition had committed murder. \V And when the multitude was come vp,
they
%%% o-2229
began to require, according as alwayes he did
%%% 2404
vnto them. \V And Pilate anſwered them, and ſaid: Wil you that I releaſe
to you the King of the Iewes? \V For he knew that the cheefe Prieſts for
enuy had deliuered him. \V But the
\LNote{Cheefe Prieſts}{Heretikes
\MNote{The Prieſts of the old Teſtament.}
abuſe the ignorant People with theſe naughtie Prieſts of the old
Teſtament, to make that name odious, & to diſcredit the Prieſts of
Chriſt in the new Teſtament. But for theſe Prieſts, thou maiſt not
maruel that they are ſo buſy againſt Chriſt,
\CNote{\Cite{Euſeb. Ec. Hiſt. li.~1. c.~6. ex Ioſeph.}}
partly becauſe they were
ſuch as were intruded by the ſecular power of the Roman Emperour, and
from yeare to yeare by bribery and frendſhip, not by ſucceſſion
according to the Law of Moyſes; partly becauſe the time was now come
when the old Prieſthood of Aaron ſhould ceaſe, and the new begin
according to the order of Melchiſedech: and for theſe cauſes God
ſuffered their former priuileges of wiſedom and iudgement and diſcretion
to decay in theſe later vſurpers, and that according to the prophet
ſaying:
\CNote{\XRef{Eze.~7,~26.}}
\Emph{The Law shal perish from the Prieſt and counſul from the Ancients.}
\MNote{The Prieſthood of the new Teſtamẽt.}
\CNote{\XRef{Io.~16,~13.}
\XRef{Luc.~22.}
and
\XRef{~10.}}
But the Prieſthood of the new Teſtament is to continue vnto the end of
the world, and hath (as being the principal part of the Church) the
aſſiſtance of the Holy Ghoſt for euer promiſed, to teach it al truth;
and for Peter the cheefe Prieſt thereof vnder Chriſt, our Sauiour praied,
That his faith should not faile: and to the reſt alſo he ſaid: \Emph{He
that heareth you, heareth me.}}
cheefe Prieſts moued the People, that he ſhould releaſe Barabbas rather
to them. \V And Pilate againe anſwering, ſaid to them: What wil you then
that I doe to the King of the Iewes? \V But they againe cried: Crucifie
him. \V And Pilate ſaith to thẽ: Why, what euil hath he done? But they
cried the more: Crucifie him. \V And Pilate willing
\LNote{To ſatisfie the People}{Pilate
\MNote{Executing lawes againſt innocents.}
should haue ſuffered death, rather then by other mens prouocation or
commandement haue executed an innocent: as a Chriſtian iudge should
rather ſuffer al extremitie, then giue ſentence of death againſt a
Catholike man for his faith.}
to ſatisfie the People, releaſed to them Barabbas, and
deliuered \Sc{Iesvs}, hauing whipped him, for to be crucified.

\V And
\CNote{\XRef{Io.~19,~2.}}
the ſouldiars led him into the court of the Palace, and they cal
togeather the whole band: \V and they clothe him in purple, and platting
a crowne of thornes, they put it vpon him. \V And they began to ſalute
him: Haile King of the Iewes. \V And they ſmote his head with a reed:
and they did ſpit on him. And bowing the knees, they adored
him. \V And after they had mocked him, they ſtripped him of the purple,
and put on him his owne garments, and they lead him forth to crucifie
him. \V And they forced a certaine man that paſſed by, Simon a Cyrenéan
comming from the countrie, the father of Alexander and Rufus, to take vp
his croſſe. \V And they bring him into the place Golgotha, which being
interpreted is, \Emph{The place of Caluarie}. \V And they gaue him to
drinke wine mixed with myrrhe; and he tooke it not.

\V And crucifying him, they deuided his garments, caſting lots vpon them,
who ſhould take which. \V And it was the third houre, and they crucified
him. \V And the title of his cauſe was ſuperſcribed, \Sc{King of the
iewes}. \V And with him they crucifie two theeues; one on the right
hand, and an other on his left. \V And the Scripture was fulfilled that
ſaith:
\CNote{\XRef{Eſa.~53,~12.}}
\Emph{And with the wicked he was reputed.} \V And they that
paſſed by, blaſphemed him, wagging their heads, and ſaying: Vah,
\Var{he that deſtroieth}{thou that deſtroieſt}
the Temple, and in three daies
\Var{buildeth}{buildeſt}
it: \V
\SNote{So Heretikes ſay of the B.~Sacrament, If it be Chriſt, let him
ſaue him ſelf from al iniuries.}
ſaue thy ſelf, comming downe from the croſſe. \V In like manner alſo the
cheefe Prieſts mocking ſaid with the Scribes one to an other: He ſaued
others, himſelf he can not ſaue. \V Let Chriſt the King of Iſrael come
downe now from the croſſe that we may ſee and beleeue. And they that
were crucified with him, railed at him.

%%% o-2230
\V And when it was the ſixt houre, there was made darknes vpon the whole
earth vntil the ninth houre. \V And at the ninth houre \Sc{Iesvs} cried
out with a mightie voice, ſaying:
\CNote{\XRef{Pſ.~21,~1.}}
\HH{Eloi, Eloi, lamma-ſabacthani?}
Which is being interpreted, \Emph{My God, my God, why haſt thou
\SNote{See
\XRef{(Mat. c.~27,~46.)}
the blaſphemous expoſition of Caluin and his folowers, and take heed
therof.}
forſaken me?} \V And certaine of the ſtanders about hearing, ſaid:
Behold, he calleth Elias. \V And one running and filling a ſpunge with
vinegre, and putting it about a reed, gaue him drinke, ſaying: Let be,
let vs ſee if Elias come to take him downe. \V And \Sc{Iesvs} putting
forth a mightie voice, gaue vp the ghoſt. \V And the vele of the temple
was rent in two, from the top to the bottome. \V And the Centurion that
ſtood ouer againſt him, ſeeing that ſo crying he had giuen vp the ghoſt,
ſaid: In deed this man was the Sonne of God.

\V And there were alſo women looking on a farre off: among whom was
Marie Magdalene, and Marie the mother of Iames the leſſe, and
%%% 2405
of Ioſeph, and Salôme; \V and when he was in Galilee, they followed him,
and miniſtred to him, and many other women that came vp togeather with
him to Hieruſalem. \V And when euening was now come (becauſe it was the
Paraſceue, which is the Sabboth-eue) \V came Ioſeph of Arimathæa a noble
Senatour, who himſelf alſo was expecting the Kingdom of God; and he went
in boldly to Pilate, and asked the body of \Sc{Iesvs}. \V But Pilate
marueled if he were now dead. And ſending for the Centurion, asked him
if he were now dead. \V And when he vnderſtood by the Centurion, he gaue
the body to Ioſeph. \V And Ioſeph
\LNote{Buying ſindon}{This
\MNote{Religious duty toward the bodies of Chriſt and his Saints.}
dutie done to Chriſtes body after his departure, was exceeding
meritorious, and is therfore by holy writ ſo often commended for an
example to faithful men, to vſe al honour and deuotion towards the
bodies of Saints and holy Perſons.}
buying ſindon, and taking him downe, wrapped him in the ſindon, and laid
him in a monument, that was hewed out of a rock. And he roled a ſtone to
the doore of the monument. \V And Marie Magdalene and Marie of Ioſeph
beheld where he was laid.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
