%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Mark
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 2380
%%% o-2200
\startChapter[
  title={Chapter 6}
  ]

\Summary{In his owne countrey (ſignifying the reprobate Iewes) he is
  contemned and therfore worketh litle in reſpect. 7.~His Apoſtles
  preach euery where and worke miracles, ſo that King Herod (who
  ſhamefully killed Iohn Baptiſt) and others are ſtriken with great
  admiration: 30.~After Iohns death he goeth into the Deſert, where
  great concurſe being vnto him, he feedeth 5000 with fiue
  loaues. 46.~And after he hath praied long in the mountaine he walketh
  vpon the ſea. 53.~And with the very touch of his garments hemme he
  healeth innumerable.}

%%% 2381
And going out from thence, he went into his countrie; and his Diſciples
folowed him. \V
\CNote{\XRef{Mt.~13,~54.}
\XRef{Lu.~4,~16.}}
And when the Sabboth was come, he began to teach in the
Synagogue: and many hearing him were in admiration at his doctrine,
ſaying: How came this felow by al theſe things?
%%% o-2201
& what wiſedom is this
that is giuen to him, and ſuch vertues as are wrought by his hands? \V
Is not this
\LNote{The Carpenter}{As his countrie folkes ſeeing him not only to be a
poore man, 
but alſo knowing (as they thought) his whole parentage to be but vulgar,
not reaching to his Godhead and Diuine Generation, did take offence or
ſcandal of him: ſo doe the Heretikes take like offence at his Perſon in
the B.~Sacrament, ſaying: Why, this is not God; for it is bread made of
corne by ſuch a baker, of the ſame mould that ſuch a lofe is; not
marking that it was not made Chriſt by baking, but by Conſecration, and
the vertue of Chriſts words.}
the Carpenter, the Sonne of \Sc{Marie}, the brother of Iames, and
Ioſeph, and Iude, and Simon? why are not alſo his ſiſters here with vs?
And they
\LNote{They were ſcandalized}{This
\CNote{\Cite{Chryſ. ho.~49. in Mt.}}
ſcandal roſe partly of enuy of his
equals by birth, who 
reputing thẽ ſelues as good as he, tooke ſkorne to be taught of
him. Wherevpon Chriſt ſaith: \Emph{A Prophet is not without honour but
in his owne countrie}; ſignifying (as it is plaine in
\CNote{\XRef{Luc.~4,~25.}}
Luke) the malice
and enuy of the Iewes his countrie men in refuſing him
\XRef{(Io.~1.)}
and that the Gentils would more eſteeme of him.}
were ſcandalized in him. \V And \Sc{Iesvs} ſaid to them: That there is
not a Prophet without honour, but in his owne countrie, and in his owne
houſe, and in his owne kinred. \V And
\LNote{He could not}{It is ſaid that he could not worke miracles there,
not meaning
that he was not able, but that on their part there wanted apt
diſpoſition to receaue them. And therfore he would not of congruity
worke there, where their incredulity was ſo great that it would not haue
profited them. And for this cauſe he ſaith
\CNote{\XRef{Mr.~5,~36.}
\XRef{Mt.~10.}}
els where to them, that wil
ſee and enioy his miracles, \Emph{Only beleeue}.}
he could not doe any miracle there, but only cured a few that were ſick,
impoſing his hands. \V And he marueled becauſe of their incredulity, and
he went about the townes in circuit teaching.

\V
\CNote{\XRef{Mt.~10,~1.}
\XRef{Lu.~9,~1.}}
And he called the Twelue; and began to ſend them two and two, and gaue
them power ouer vncleane Spirits. \V And he commanded them that they
ſhould take nothing for the way, but a rod only: not skrip, not bread,
nor money in their purſe, \V but ſhod with ſandals, and that they ſhould
not put on
\LNote{Not two coats}{He forbiddeth ſuperfluities, and too careful
prouiſion of
bodily things, when they are about Gods ſeruice in gaining ſoules. And
for the contrariety that ſeemeth here and in S.~Matthew, vnderſtand that
there he forbiddeth them to carie rod or ſtaffe to defend them ſelues,
here he permitteth a walking rod or ſtaffe to leane and ſtay vpon: there
he forbiddeth ſhooes to couer al the foote, ſuch as we weare: here he
permitteth ſandals, that is ſuch as had ſoles only, which the poore
commonly ware in Iewry, & now ſome religious men. See S.~Auguſtins
opinion
\Cite{li.~2. c.~30. de conſenſu Euang. to.~4.}}
two coats. \V And he ſaid to them: Witherſoeuer you ſhal enter into an
houſe, there tarie til you depart thence. \V And whoſoeuer ſhal not
receaue you, nor heare you; going forth from thence ſhake of the duſt
from your feet for a teſtimonie to them. \V And going forth they preached
that they ſhould doe pennance: \V And they caſt out many Diuels, and
\CNote{\XRef{Ia.~5,~14.}}
anointed
\LNote{With oile}{In the words of the commiſſion oile is not mentioned,
and yet
it is certaine by this their vſing of oile, that either Chriſt did then
appoint them to vſe it, or they might take it vp of them ſelues by
vertue of the general commiſſion.}
%%% !!! Why is this two separate notes? Also, the second is noted as
%%% being for verse 23, corrected to 13 in other.
\LNote{With oile}{By this it is cleere that not only the Apoſtles or
other many
haue power to worke miracles, by their only word and inuocation of
Chriſts name, but alſo by application of creatures: which creatures alſo
haue a miraculous medicinal vertue to heale diſeaſes.}
with
\SNote{A preparatiue to the Sacrament of extreme vnction.
\XRef{Iac.~5.}}
oile many ſick, and healed them.

\V And
\CNote{\XRef{Mt.~14,~1.}
\XRef{Lu.~9,~7.}}
King Herod heard (for his name was made manifeſt) and he ſaid: That
Iohn the Baptiſt is riſen againe from the dead, and therfore vertues
worke in him. \V And others ſaid: That it is Elias. But others ſaid:
That it is a Prophet, as one of the Prophets. \V Which Herod hearing,
ſaid: Iohn whom I beheaded, he is riſen againe from the dead.

\V For the ſaid Herod ſent and apprehended Iohn, and bound him in priſon
for Herodias the wife of
\SNote{He might & ſhould by Moyſes law haue maried his brothers wife if
he had been dead without iſſue: but this Philip was yet aliue, and had
alſo this daughter that danced.}
Philippe his brother, becauſe he had maried her. \V For Iohn ſaid to
Herod:
\CNote{\XRef{Leu.~18,~16.}
\XRef{20,~21.}}
It is not lawful for thee to haue thy brothers wife, \V And
Herodias lay in waite for him: and was deſirous to kil him, and could
not. \V For Herod feared Iohn, knowing him to be a iuſt and holy man:
and he kept him, and by hearing him did many things: and he heard him
gladly. \V And when a conuenient day was fallen, Herod made the ſupper
of his birth-day to the Princes, and the Tribunes, and the cheefe of
Galilee. \V And when the daughter of the ſame
%%% o-2202
Herodias came in, and had danced, and pleaſed Herod, and them that ſate with
him at the table, the King ſaid to the damſel: Aske of me what thou
wilt, and I wil giue it thee. \V And he ſware to her: That whatſoeuer
thou ſhalt aske I wil giue thee, though the halfe of my Kingdom. \V Who
when ſhe was gone forth, ſaid to her mother, what ſhal I aske? But ſhe
ſaid: The head of Iohn the Baptiſt. \V And when ſhe was gone in by and
by with haſt to the King, ſhe asketh ſaying: I wil that forthwith thou
giue me in a platter the head of Iohn the Baptiſt.
%%% 2382
\V And the King was ſtroken ſad. Becauſe of his othe and for them that
ſate togeather at table he would not diſpleaſe her: \V but ſending the
hangman, commanded that his head ſhould be brought in a platter. \V And
he beheaded him in the priſon, and brought his head in a platter: and
gaue it to the damſel, and the damſel gaue it to her mother. \V Which
his Diſciples hearing came, and tooke his body; and they put it in a
monument.

\V And
\CNote{\XRef{Lu.~9,~10.}}
the Apoſtles gathering togeather vnto \Sc{Iesvs}, made relation
to him of al things that they had done and taught. \V And he ſaid to
them: Come apart into the deſert place, and reſt a litle. For there were
that came and went, many: and they had not ſo much as ſpace to eate. \V
And
\CNote{\XRef{Mt.~14,~13.}}
going vp into the boat, they went into a deſert place apart. \V And
they ſaw them going away, and many knew: and they ranne flocking thither
on foot from al cities, and preuented them.

\V And going forth, \Sc{Iesvs} ſaw a great multitude; and he had
compaſſion on them, becauſe they were as ſheep not hauing a ſhepheard,
and he began to teach them many things. \V And
\CNote{\XRef{Mt.~14,~15.}
\XRef{Luc.~9,~11.}
\XRef{Io.~6,~5.}}
when the day was now
farre ſpent, his Diſciples came to him, ſaying: This is a deſert place,
and the houre is now paſt: \V dimiſſe them, that going out into the next
villages and townes, they may buy them ſelues meats to eate. \V And he
anſwering ſaid: Giue ye them to eate. And they ſaid to him: Let vs goe
and buy bread for two hundred pence, and we wil giue them to eate. \V
And he ſaith to them: How many loaues  haue you? goe and ſee. And when
they knew, they ſay: Fiue, and two fiſhes. \V And he commanded them that
they ſhould make al ſit downe, by companies vpon the greene graſſe. \V
And they ſate downe in ranks by hundreds and fifties. \V And when he
had taken the fiue loaues, and the two fiſhes: looking vp vnto Heauen,
%%% o-2203
he bleſſed, and brake the loaues, and gaue to his Diſciples to ſet
before them: and the two fiſhes he deuided to al. \V And al did eate,
and had their fill. \V And they tooke vp the leauings, twelue ful baskets of
fragments, and of the fiſhes. \V And they that did eate, were fiue
thouſand men. \V And immediatly he compelled his Diſciples to goe vp
into the boat, that they might goe before him beyond the
\SNote{The narrow ſea, or water.}
ſtrait to Bethſaida: whiles himſelf did dimiſſe the People.

\V And
\CNote{\XRef{Mt.~14,~23.}
\XRef{Io.~6,~16.}}
when he had dimiſſed them, he went into the mountaine to pray. \V
And when it was late, the boat was in the middes of the ſea, and himſelf
alone on the land. \V And ſeeing them labouring in rowing (for the wind
was againſt them) and about the fourth watch of the night he commeth to
them walking vpon the ſea, and he would haue paſſed by them. \V But they
ſeeing him walking vpon the ſea, thought it was a ghoſt, and cried
out. \V For al ſaw him, and were troubled. And immediatly he talked with
them, and ſaid to them: Haue confidence, it is I, feare ye not. \V And
he went vp to them into the ſhip, and the wind ceaſed: and they were
farre more aſtoniſhed within them ſelues: \V for they vnderſtood not
concerning the loaues; for their hart was blinded.

%%% 2383
\V And
\CNote{\XRef{Mt.~14,~34.}}
when they had paſſed ouer, they came into the land of Genezareth,
and ſet to the ſhore. \V And when they were gone out of the boat,
incontinent they knew him: \V and running through that whole countrie
they began to carie about in couches thoſe that were il at eaſe, where
they heard he was. \V And whitherſoeuer he entred into townes or into
villages or cities, they laid the ſicke in the ſtreets, and beſought him
that they might touch but the hemme of his garment: and as many as
touched him, were made whole.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
