%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Mark
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 2391
%%% o-2213
\startChapter[
  title={Chapter 10}
  ]

\Summary{He anſwereth the tempting Phariſees (and againe his Diſciples
  afterward) that the caſe of a man with his wife shal be (as in the
  firſt inſtitution) vtterly indiſſoluble. 13.~He bleſſeth
  children. 17.~He sheweth what is to be done to get life euerlaſting:
  21.~what alſo for a rich man to be perfect; 28.~as alſo what paſsing
  reward they shal haue that doe ſo in time of perſecution. 32.~He
  reuealeth more to his Diſciples, touching his Paſsion: 35.~bidding the
  two ambitious ſuiters to thinke rather of ſuffring with him: 41.~and
  teaching vs in the reſt of his Diſciples, not to be greeued at our
  Eccleſiaſtical Superiours, conſidering they are (as he was himſelf) to
  toile for our ſaluation. 46.~Then going out of Iericho, he giueth
  ſight to a blind man.}

And
\CNote{\XRef{Mt.~19,~1.}}
\MNote{The third part of this Ghoſpel Chriſtes comming into Iewrie
toward his Paſſion.}
riſing vp thence, he commeth into the coaſts of Iewrie beyond Iordan:
and the multitudes aſſemble againe vnto him. And as he was accuſtomed,
againe he taught them. \V And the Phariſees comming neere, asked him: Is
it lawful for a man to dimiſſe his wife? tẽpting him. \V But he
anſwering, ſaid to them: What did Moyſes command you? \V Who ſaid:
\CNote{\XRef{Deu.~24,~1.}}
Moyſes
\LNote{Permitted}{Some
\MNote{Toleration & permiſſion of euil.}
things are permitted, though not approued or allowed, to auoid greater
inconueniences. No man may doe euil for any cauſe, but he may permit
other mens euil for diuerſe cauſes: as God himſelf doeth, who can doe no
euil. So doth the Prince and Common-wealth permit leſſer euils to eſchew
greater, and ſo may the Holy Church much more
\CNote{\Cite{Aug. ep.~119. c.~19.}}
(as S.~Auguſtine ſaith ſhe
doeth) being placed among much chaffe and much cockle, tolerate many
things: and yet whatſoeuer is againſt faith & good life, ſhe neither
approueth, nor diſſembleth with ſilence, not committeth.}
permitted to write a bil of diuorce, and to dimiſſe her. \V To
whom \Sc{Iesvs} anſwering, ſaid: For the hardnes of your hart he wrote
you this precept. \V But from the beginning of the creation
\CNote{\XRef{Gen.~1,~26.}}
God made thẽ
male & femal. \V For this cauſe,
\CNote{\XRef{Gen.~2,~24.}}
man shal leaue his father & mother; and
shal cleaue to his wife. \V And they two shal be in one flesh. Therfore
now they are not two, but one fleſh. \V That therfore which God hath
ioyned togeather,
\SNote{The obligation betwixt man and wife is ſo great, that during life
it can not be broken.}
let no man ſeparate.

\V And
\CNote{\XRef{Mt.~19,~9.}
\XRef{Mr.~5,~32.}
\XRef{Lu.~16,~18.}
\XRef{1.~Cor.~7,~11.}}
in the houſe againe his Diſciples asked him of the ſame thing. \V
And he ſaith to them: Whoſoeuer dimiſſeth his wife and
\LNote{And marieth another}{That
\MNote{Mariage after diuorce vnlawful.}
which S.~Matthew vttered more obſcurely, and is miſtaken of ſome, as
though he meant that for fornication a man might put away his wife and
mary another, is here by this Euangeliſt
\CNote{\XRef{Luc.~16,~18.}}
(as alſo by S.~Luke) put out of
doubt, generally auouching, that whoſoeuer putteth away his wife and
marieth another committeth aduoutrie.
\Cite{Aug. li.~1. de adult. coniug. c.~11. & ſequentibus.}}
marieth another, committeth aduoutrie vpon her. \V And if the wife
dimiſſe her husband, and mary another, she committeth aduoutrie.

\V And
\CNote{\XRef{Mt.~19,~13.}
\XRef{Lu.~18,~15.}}
they offered to him yong children, that he might touch them. And
the Diſciples threatned thoſe that offered them. \V Whom when \Sc{Iesvs}
ſaw, he took it il, and ſaid to them: Suffer the litle children to come
vnto me, and prohibit them not. For the Kingdom of God is for ſuch. \V
Amen I ſay to you, whoſoeuer receaueth not the Kingdom of God as a litle
child, shal not enter into it. \V And embracing them, and
\SNote{Our Sauiour gaue the children his bleſſing, impoſing his hands
vpon them.}
impoſing hands vpon them, he bleſſed them. \V And when he was gone forth
in the way, a certaine man running forth & kneeling before him asked
him,
%%% o-2214
\CNote{\XRef{Mt.~19,~16.}
\XRef{Lu.~18,~18.}}
Good Maiſter, what shal I doe that I may receaue life euerlaſting? \V
And \Sc{Iesvs} ſaid to him, why calleſt thou me good?
\LNote{None is good}{None is entirely, ſubſtantially, and of himſelf
good, but God:
though by participation of Gods goodnes, men are truly alſo called
good.}
None is good but one, God. \V Thou knoweſt the
\SNote{Note that the keeping of Gods commandements procureth life
euerlaſting.}
commandements,
\CNote{\XRef{Exo.~20,~13.}}
\Emph{Commit not aduoutrie}, \Emph{Kil not}, \Emph{Steale
not}, \Emph{Beare not falſe witneſſe}, \Emph{doe not
fraud}, \Emph{Honour thy father and mother}. \V But he anſwering, ſaid
to him: Maiſter al theſe things I haue obſerued from my youth. \V
And \Sc{Iesvs} beholding him, loued him, and ſaid to him: One thing is
wanting vnto thee:
\SNote{This is counſel of perfection (not a precept) which the Religious
profeſsing and keeping voluntarie pouertie, doe folow.}
goe, ſel whatſoeuer thou haſt, and giue to the poore, and thou shalt
haue treaſure in Heauen; and come, folow me. \V Who being ſtroken ſad at
the word, went away ſorowful for he had many poſſeſsions. \V
And \Sc{Iesvs} looking about, ſaith to his Diſciples: How hardly shal
they that haue
%%% 2392
money, enter into the Kingdom of God! \V And the Diſciples were
aſtoniſhed at his wordes. But \Sc{Iesvs} againe anſwering, ſaith to thẽ:
Children, how hard is it for them that truſt in money, to enter into the
Kingdom of God! \V It is eaſier for a camel to paſſe through a needels
eye, then for
\LNote{A rich man}{He is here called a rich man that hath his confidence
(as here
is expreſſed) in his treaſure, and had rather forſake his faith and duty
to God, then loſe them, as al they which liue in Schiſme or Hereſie to
ſaue their goods.}
a rich man to enter into the Kingdom of God. \V Who marueled more, ſaying
to themſelues: And who can be ſaued? \V And \Sc{Iesvs} beholding them
ſaith: With men it is impoſſible; but not with God. For al things are
poſſible with God. \V And Peter began to ſay vnto him: Behold, we haue
left al things, and haue folowed thee. \V \Sc{Iesvs} anſwering, ſaid:
\SNote{Exceeding happie be they that can forſake their temporal things
for religion.}
Amen I ſay to you, there is no man which hath left houſe, or brethren,
or ſiſters, or father, or mother, or children, or lands for me and for
the Ghoſpel, \V that ſhal not receaue
\LNote{An hundred times ſo much}{Sometime
\MNote{Their reward in this life that forſake ought for Gods ſake.}
God doth ſo bleſſe men alſo in worldly benefits that haue forſaken al
for him, as S.~Greg. S.~Auguſt. and S.~Paulinus doe note: but the
principal meaning is,
\CNote{\Cite{Hiero. in 19.~Mat.}}
that he wil giue to ſuch men in this life
aboundance of grace and ſpiritual comfort and contentation and ioy of
conſcience (as they feele which haue experience) the which ſpiritual
guifts exceed the temporal commodities more then an hundredfold. In ſo
much that he that hath fully forſaken but ſmal things for religion,
would not forſake religion to haue al the world.}
an hundred times ſo much now in this time; houſes, and brethren, and
ſiſters, and mothers and children, and lands, with perſecutions: and in
the world to come life euerlaſting. \V But
\CNote{\XRef{Mat.~19,~30.}
\XRef{Luc.~13,~30.}}
many that are firſt, ſhal be laſt; and the laſt, firſt.

\V And they were in the way going vp to Hieruſalem: and \Sc{Iesvs} went
before them, and they were aſtoniſhed; and folowing were afraid. And
\CNote{\XRef{Mt.~20,~17.}
\XRef{Lu.~18,~31.}}
taking againe the Twelue, he began to tel them the things that ſhould
befal him. \V That, behold we goe vp to Hieruſalem, and the Sonne of man
ſhal be betrayed to the cheefe Prieſts, and to the Scribes and Ancients,
and they ſhal condemne him to death, and shal deliuer him to the
Gentils. \V and they ſhal mock him, and ſpit on him, and ſcourge him,
and kil him, and the third day he ſhal riſe againe.

%%% o-2215
\V And
\CNote{\XRef{Mt.~20,~20.}}
there came to him Iames and Iohn the ſonnes of Zebedee, ſaying:
Maiſter, we wil that what things ſoeuer we ſhal aske, thou doe it to
vs. \V But he ſaid to them: What wil you that I doe to you? \V And they
ſaid: Grant to vs, that we may ſit, one on thy right hãd & the other on
thy left hand, in thy glorie. \V And \Sc{Iesvs} ſaid to them: You wot
not what you aske. Can you drinke the cup that I drinke; or be baptized
with the Baptiſme wherwith I am baptized? \V But they ſaid to him, we
can. And \Sc{Iesvs} ſaid to them: The cup indeed which I drinke, you
ſhal drinke; and with the Baptiſme wherwith I am baptized, ſhal you be
baptized: \V but to ſit on my right hand or on my left, is not mine to
giue vnto you, but to whõ it is prepared. \V And the ten hearing, began
to be diſpleaſed at Iames and Iohn. \V And \Sc{Iesvs} calling them,
ſaith to them:
\CNote{\XRef{Mt.~20,~25.}
\XRef{Lu.~22,~25.}}
You know that they which ſeeme to rule ouer the Gentils,
ouerrule them: and their Princes haue power ouer them. \V But it is not
ſo in you. But whoſoeuer wil be greater, ſhal be your miniſter; \V and
whoſoeuer wil be firſt among you, ſhal be the ſeruant of al. \V For the
Sonne of man alſo is not come to be miniſtred vnto, but to miniſter, and
to giue his life a redemption for many.

\V And
\CNote{\XRef{Mt.~20,~29.}
\XRef{Lu.~13,~35.}}
they come to Iericho: and when he departed from Iericho, and his
Diſciples, and a very great multitude, the ſonne of Timæus, Bartimæus
the blind man, ſate by the way ſide begging. \V Who when he had heard,
that it is \Sc{Iesvs} of Nazareth, he began to crie, and to
ſay: \Sc{Iesvs}, ſonne of Dauid, haue mercie vpon me. \V And many
threatned him, to hold his peace. But he cried much more, Sonne of
Dauid, haue mercie vpon me. \V And \Sc{Iesvs} ſtanding ſtil commanded
him to be called. And
%%% 2393
they cal the blind mã, ſaying to him: Be of better cõfort, ariſe, he
calleth thee. \V Who caſting of his garment leapt vp, and came to
him. \V And \Sc{Iesvs} anſwering, ſaid to  him: What wilt thou that I
doe vnto thee? And the blind man ſaid to him: Rabbôni, that I may
ſee. \V And \Sc{Iesvs} ſaid to him: Goe thy wayes, thy faith hath made
thee ſafe. And forthwith he ſaw, and folowed him in the way.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
