%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Mark
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 2386
%%% o-2207
\startChapter[
  title={Chapter 8}
  ]

\Summary{Of compaſſion he feedeth the People, 4000.\ with ſeauen
  loaues. 10.~After al which miracles as though they were yet
  vnſufficient to proue him to be Chriſt, the obſtinate Phariſees doe
  require ſome miracle from Heauen. 13.~Wherevpon forſaking them, he
  warneth his Diſciples to beware of the leauen of their doctrine,
  neither to feare want of neceſſaries. 22.~He healeth a blind man by
  degrees and with ceremonies. 27.~Peter confeſſeth him (though men al
  this while had not learned ſo farre) to be Chriſt. 31.~And by and by
  he reuealeth to them his Paſſion, 32.~rebuking alſo Peter for
  diſſuading it, 34.~and shewing that it is a thing wherein al that wil
  be ſaued (namely in time of perſecution) muſt folow him.}

In thoſe daies againe
\CNote{\XRef{Mt.~15,~32.}}
when there was a great multitude, and had not what
to eate: calling his Diſciples togeather, he ſaith to them: \V I haue
compaſſion vpon the multitude, becauſe loe
\SNote{Great feruour and deuotion in the good People, & exceeding force
in our Maiſters preaching, that made them abide faſting ſo long to heare
his diuine ſermons.}
three daies they now endure with me, neither haue what to eate. \V And
if I dimiſſe them faſting into their home, they wil faint in the way;
for ſome of them came farre off. \V And his Diſciples anſwered him:
Whence may a man fil them here with bread in the wildernes? \V And he
asked them: How many loaues haue ye? Who ſaid: Seauen. \V And he
commanded the multitude to ſit downe vpon the ground. And taking the
ſeauen loaues, giuing thankes he brake, and
\LNote{Gaue to his Diſciples}{He ſerueth the People not immediatly him
ſelf, but by the
Apoſtles miniſterie to teach vs that we muſt receiue Chriſtes Sacraments
and doctrine, not at our owne hand, but of his Prieſts and our
Paſtours.}
gaue to his Diſciples for to ſet before them, and they did ſet them
before the multitude. \V And they had a few litle fiſhes; and
\LNote{Bleſſed them}{So
\TNote{\G{εὐλογήσας αὐτὰ}}
is it in ſome ancient Greek copies, agreable to our Latin, and in
S.~Luke expreſly in the common Greek text, that
\CNote{\XRef{Luc.~9,~16.}}
he bleſſed the fiue loaues and the two fishes:
\MNote{Bleſsing of creatures worketh an effect in them.}
which muſt be alwayes marked againſt the Heretikes, which denie this
bleſsing to pertaine to the creatures, but feine it alwayes to be
referred to God for thanks giuing. For if it were ſo, he would haue ſaid
grace but once for that whole refection: but he did ſeuerally bleſſe
both the bread firſt, & afterward the fishes alſo, multiplying them by
his ſaid bleſsing (as
\CNote{\XRef{Gen.~1,~22,~28.}}
he did mankind and other creatures in the
beginning by bleſſing them) and ſo working effectually ſome change or
alteration in the very creatures them ſelues.}
he bleſſed them, and commanded them to be ſet before them. \V And they
did eate and were filled, and they tooke vp that which was left of the
fragments, ſeauen maundes. \V And they that had eaten were about four
thouſand: and he dimiſſed them.

\V And
\CNote{\XRef{Mt.~15,~39.}
\XRef{16.~1.}}
immediatly going vp into the boat with his Diſciples, he came
into the quarters of Dalmanutha. \V And the Phariſees went forth, and
began to queſtion with him, asking of him a ſigne from Heauen, tempting
him. \V And groning in Spirit, he ſaith: Why doth this Generation aske a
ſigne? \V Amen I ſay to you, If a ſigne ſhal be giuen to this
Generation. \V And leauing them, he went vp againe into the boat, and paſſed
beyond the ſtrait.

%%% o-2208
\V And they forgot to take bread: and they had but one loaf with them in
the boat. \V And he commanded them, ſaying: Looke wel and beware of the
leauen of the Phariſees, and the leauen of Herod.
%%% 2387
\V And they reaſoned among them ſelues ſaying: Becauſe we haue not
bread. \V Which \Sc{Iesvs} knowing, ſaith to them: Why doe you reaſon,
becauſe you haue not bread? doe you not yet know nor vnderſtand? yet
haue ye your hart blinded? \V hauing eyes ſee you not? and hauing eares
heare you not? Neither doe you remember? \V When
\CNote{\XRef{Mr.~6,~38.}}
I brake fiue loaues
among fiue thouſand; and how many baſkets ful of fragments tooke you vp?
They ſay to him, Twelue. \V
\CNote{\XRef{Mr.~8,~5.}}
When alſo ſeauen loaues among four thouſand,
how many maundes of fragments tooke you vp? And they ſay to him
Seauen. \V And he ſaid to them: How doe you not yet vnderſtand?

\V And they come to Bethſaida; and they bring to him one blind, and
deſired him that he would
\SNote{Our Sauiour vſed to worke much by touching, & by impoſition of
his hands: that we may learne not to cõtemne the corporal and external
application of holy things, nor to chalẽge by the ſpirit & faith only,
as Heretikes doe.}
touch him. \V And taking the hand of the blind, he led him forth out of
the towne; and ſpitting into his eyes, impoſing his hands, he asked him
if he ſaw anything. \V And looking vp, he ſaid: I ſee men as it were
trees, walking. \V After that againe
%%% \SNote{} ???? Missing note?
he impoſed his hands vpon his eyes, and he began to ſee, and was
reſtored, ſo that he ſaw al things clerely. \V And he ſent him into his
houſe, ſaying: Goe into thy houſe; and if thou enter into the towne, tel
no body.

\V And
\CNote{\XRef{Mt.~16,~13.}
\XRef{Lu.~9,~18.}}
\Sc{Iesvs} went forth and his Diſciples into the townes of
Cæſarea-Philippi; and in the way he asked his Diſciples, ſaying to them:
Whom doe men ſay that I am? \V Who anſwered him, ſaying: Iohn the
Baptiſt, ſome Elias, & other ſome as it were one of the Prophets. \V
Then he ſaith to them: But whom doe you ſay that I am? Peter anſwering
ſaid to him: Thou art Chriſt. \V And he threatned them that they ſhould
not tel any man of him.

\V And he began to teach them, that the Sonne of man muſt ſuffer many
things, and be reiected of the Ancients and of the high Prieſts & the
Scribes, and be killed: and after three daies riſe againe. \V And he
ſpake the Word openly. And Peter taking him, began to rebuke him. \V Who
turning, and ſeeing his Diſciples, threatned Peter, ſaying: Goe behind
me Satan, becauſe thou fauoureſt not the things that are of God, but
that are of mẽ. \V And calling the multitude
%%% o-2209
togeather with his Diſciples, he ſaid to them: If any man wil folow me,
let him deny himſelf, and take vp his croſſe, and folow me. \V For he
that wil ſaue his life, ſhal loſe it; and he that ſhal loſe his life
\LNote{For me and the Ghoſpel}{By
\MNote{What is to ſuffer for the Ghoſpel.}
the Ghoſpel is ſignified, not only the foure Euangeliſts, but al
Scriptures, and whatſoeuer Chriſt ſaid that is not in Scriptures; for
he ſaith in this very place:
\CNote{\XRef{Luc.~9.}}
\Emph{He that shal be ashamed of my wordes,
the Sonne of man wil be ashamed of him, &c.} Neither his owne wordes
only, but whatſoeuer the Apoſtles taught in word or writing; for our
Sauiour ſaith: \Emph{He that deſpiſeth you, deſpiſeth me}. For defenſe
of any of al theſe, and of euery Article of the Catholike faith, we
ought to die, and this is to loſe our life for Chriſt and his Ghoſpel.}
for me and the Ghoſpel, ſhal ſaue it. \V For what ſhal it profit a man,
if he
\LNote{Gaine the whole world}{Let 
\MNote{Doing againſt our owne conſcience.}
ſuch note this, that for feare or flattery of the world
condeſcend to obey the vniuſt lawes of men touching religion, againſt
their owne conſciences, and be content for the reſt of a few daies of
this life, and for ſauing their temporal goods, to loſe their ſoule and
the ioyes of heauen.}
gaine the whole world, & ſuffer damage of his ſoule? \V For he that ſhal
be aſhamed of me, and of my wordes in this aduouterous and ſinful
Generation, the Sonne of man alſo wil be aſhamed of him, when he ſhal
come in the glorie of his Father with the holy Angels.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
