%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Mark
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 2406
%%% o-2231
\startChapter[
  title={Chapter 16}
  ]

\Summary{The third day, to three women at his Sepulcher, an Angel
telleth that he is riſen, and wil (as he promiſed
\XRef{Mar.~14.~28.})
shew himſelf in Galilee. 9.~The ſame day he appeareth to Marie
Magdalene, afterward to two Diſciples: yet the Eleuen wil not beleeue
it, vntil to them alſo he appeareth. 15.~To whom hauing giuen
commiſſion into al Nations, with power alſo of Miracles, he aſcendeth,
and they plant his Church euery where.}

And
\CNote{\XRef{Mt.~28,~1.}
\XRef{Luc.~24,~1.}
\XRef{Io.~20,~1.}}
when the Sabboth was paſt, Marie Magdalene and Marie of Iames, and
Salôme
\LNote{Bought ſpices.}{As
\MNote{The womens deuotion toward Chriſts body being dead.}
ſhe did beſtow and conſume a coſtly ointment vpon his body being yet
aliue
\XRef{(c.~14.~3.)}
Chriſt himſelf defending and highly commending the
\Fix{fact}{act}{probable typo, same in other}
againſt Iudas and other who accounted it to be ſuperfluous and better to
be beſtowed otherwiſe: So not without great deuotion and merit, ſhe and
theſe other women ſeeke to anoint his body dead (though Heretikes or
other ſimple perſons may pretend ſuch things to be better beſtowed vpon
the poore) and therfore,
\CNote{\XRef{Mt.~16,~9.}}
ſhe firſt before the other,
\CNote{\XRef{Mt.~28,~9.}}
and they next, ſaw him after his Reſurrection.}
bought ſpices, that comming they might anoint \Sc{Iesvs}. \V And very
early the firſt of the Sabboths, they come to the monument, the ſunne
being now riſen. \V And they ſaid one to an other: Who ſhal role vs back
the ſtone from the doore of the monument? \V And looking, they ſaw the
ſtone roled back. For it was very great. \V And entring into the
monument, they ſaw a yong man ſitting on the right hand couered with a
white robe: and they were aſtoniſhed. \V Who ſaith to thẽ: Be not
diſmaied; you ſeeke \Sc{Iesvs} of Nazareth, that was crucified; he is
riſen, he is not here, behold the place where they laid him. \V But goe,
tel his Diſciples and
\SNote{Peter is named in ſpecial (as often els where) for prerogatiue.}
Peter that he goeth before you into Galilee; there you ſhal ſee him,
\CNote{\XRef{Mr.~14,~28.}}
as he told you. \V But they going forth, fled from the monument. For
trembling and feare had inuaded them: and they ſaid nothing to any body,
For they were afraid.

\V And he riſing early the firſt of the Sabboth,
\CNote{\XRef{Io.~20,~16.}}
appeared firſt to Marie Magdalen,
\CNote{\XRef{Luc.~8,~2.}}
out of whom he had caſt ſeauen Diuels. \V She went and told
them that had been with him, that were mourning & weeping. \V And they
hearing that he was aliue and had been ſeen of her, did not beleeue.

\V And
\CNote{\XRef{Lu.~24,~13.}}
after this he appeared in an
\LNote{In an other shape.}{Chriſt
\MNote{Chriſts body vnder diuers formes.}
though he haue but one corporal ſhape, natural to his Perſon, yet by his
omnipotencie he may be in whatſoeuer forme, & appeare in the likeneſſe of
any other man or creature, as he liſt. Therfore let no man thinke it
ſtrange, that he may be vnder the forme of bread in the B.~Sacrament.}
other ſhape to two
%%% o-2232
of them walking, as they were going into the countrie. \V And they going
told the reſt: neither them did they beleeue.

\V Laſt
\CNote{\XRef{Lu.~24,~36.}
\XRef{Io.~20,~26.}}
he appeared to thoſe Eleuen as they ſate at the table; and he
exprobated their incredulity and hardnes of hart, becauſe they did not
beleeue them that had ſeen him riſen againe. \V And he ſaid to them:
\CNote{\XRef{Mt.~28,~19.}}
Going into the whole world preach the Ghoſpel to al creatures. \V He
that
\LNote{He that beleeueth.}{Note wel, that whereas this Euangeliſt
mentioneth only faith &
Baptiſme, as though to beleeue & to be baptized were enough,
\MNote{Not only faith.}
S.~Matthew addeth theſe wordes alſo of our Sauiour,
\CNote{\XRef{Mt.~28,~20.}}
\Emph{teaching them
to obſerue al things whatſoeuer I haue commanded you}, which containeth
al good workes and the whole iuſtice of a Chriſtian man.}
beleeueth and is baptized, ſhal be ſaued; but he that beleeueth not,
ſhal be condemned. \V And them that beleeue
\LNote{Theſe ſignes shal folowe.}{It
\MNote{The guift of miracles.}
is not meant, that al Chriſtians or true beleeuers ſhould
doe miracles; but that ſome for the proofe of the faith of al, ſhould
haue that guift. The which is the grace or guift of the whole Church,
executed by certaine for the edification and profit of the whole.}
theſe ſignes ſhal follow: In my name ſhal they caſt out Diuels; They
ſhal ſpeake with new tongues; \V Serpents ſhal they take away; And if
they drinke any deadly things, it ſhal not hurt them; They ſhal
\SNote{Miracles in the Church by impoſition of hands.}
impoſe hands vpon the ſick and they ſhal be whole.

\V 
\MNote{The Aſcẽſion.}
And ſo our Lord \Sc{Iesvs} after he ſpake vnto them,
\CNote{\XRef{Luc.~24,~51.}}
was aſſumpted
into Heauen, and ſate on the right hand of God. \V But they going forth
preached euery where: our Lord working withal, and confirming the Word
with ſignes that folowed.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
