%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Mark
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 2373
%%% o-2192
\startChapter[
  title={Chapter 3}
  ]

\Summary{The blind Phariſees ſeeking his death for doing good vpon the
  Sabboths, he meekly goeth out of the way: where the People that flock
  vnto him, and his Miracles, are innumerable. 13.~Yea to his Twelue
  alſo (hauing need of moe workmen) he giueth power to worke
  Miracles. 20.~He ſo occupieth himſelf for ſoules, that his kinne
  thinke him mad. 22.~The Scribes of Hieruſalem come ſo farre, and yet
  haue nothing but abſurdly to blaſpheme his caſting out Diuels, to
  their owne damnation. 31.~That
  %%% 2374
  the Iewes should not (after their manner) thinke it enough, that he
  is of their bloud; he telleth that ſuch rather are deare to him, as
  keepe Gods commandements.}

And
\CNote{\XRef{Mt.~12,~9.}
\XRef{Lu.~6,~6.}}
he entred againe into the Synagogue, and there was a man there that
had a withered hand. \V And they watched him whether he would cure on
the Sabboths; that they might accuſe him. \V And he ſaith to the man
that had the withered hand: Riſe vp into the middes. \V And he ſaith to
them: Is it lawful on the Sabboths to doe wel or il? to ſaue a ſoule, or
to deſtroy? but they held their peace \V And looking round about vpon
them, with anger, being ſorrowful for the blindnes of their hart, he
ſaith to the man: Stretch forth thy hãd. And he ſtretched it
\Fix{fort;hand}{forth; and}{obvious typo, fixed in other}
his hand was reſtored vnto him.

%%% o-2193
\V And the Phariſees going forth, immediatly made a conſultation with
the Herodians againſt him, how they might deſtroy him. \V But \Sc{Iesvs}
with his Diſciples retired to the ſea: and a great multitude from
Galilee & Iewrie followed him, \V and from Hieruſalẽ, & from Idumæa, and
beyond Iordan. And they about Tyre and Sidon, a great multitude, hearing
the things which he did, came to him. \V And he ſpake to his Diſciples
that a boat might attend on him becauſe of the multitude, leſt they
ſhould throng him. \V For he healed many, ſo that there preaſſed in vpon
him for
\SNote{The only touching of Chriſts holy Perſon, or any part of his
clothes, or whatſoeuer belonged to him, did heale al diſeaſes.}
to touch him, as many as had hurts. \V And the vncleane Spirits, when
they ſaw him, fel downe vnto him: and they cryed ſaying: \V
\LNote{Thou art the Sonne}{The confeſſion of the truth is not
grateful to God, proceeding from euery perſon. The Diuel acknowledging
our Sauiour to be the Sonne of God, was bidden hold his peace: Peters
confeſsion of the ſame was highly allowed and rewarded. 
\Cite{Aug. tract 10. in ep. Ioan.}
\Cite{Ser.~30. 31. de verb. Apoſt.}
Therfore
\MNote{Sermons, Seruice, & praier of Heretikes.}
neither Heretikes ſermons muſt be heard, no not though they preach the
truth. So is it of their prayer and ſeruice, which being neuer ſo good
in it ſelf, is not acceptable to God out of their mouthes, yea it is no
better then the howling of wolues.
\Cite{Hier. in 7.~Oſee.}}
Thou art the Sonne of God. And he vehemently charged them that they
ſhould not diſcloſe him.

\V And
\CNote{\XRef{Mt.~5,~1.}
\XRef{Io.~1.}
\XRef{Lu.~6,~12.}
\XRef{9,~1.}}
aſcending into a mountaine, he called vnto him whom he would him
ſelf, and they came to him. \V And he made that
\LNote{Twelue}{This
\MNote{The number of Twelue, myſtical.}
number of twelue Apoſtles is myſtical and of great importance (as
 appeareth
\CNote{\XRef{Act.~1.}}
by the chooſing of Mathias into Iudas place to make vp againe
this number) prefigured in the 12.~Patriarkes,
\XRef{Gen.~49.}
the 12.~Princes of the children of Iſrael,
\XRef{Num.~1.}
\Fix{}{the }{Not really a typo, but following other is clearer}% to prevent a paragraph break
12.~fountaines found in Elim,
\XRef{Exod.~15.}
the 12.~pretious ſtones in the Rational of Aaron,
\XRef{Exod.~39.}
the 12.~Spies ſent by Moyſes,
\XRef{Num.~13.}
the 12.~ſtones taken out of Iordan whereof the Altar was made,
\XRef{Ioſu.~4.}
the 12.~loues of Propoſition,
\XRef{Leuit.~24.} &c.
\Cite{Anſelm. in Mt. c.~10.}
And theſe are the 12.~foundations of heauenly Hieruſalem,
\XRef{Apoc.~23.}}
twelue ſhould be with him, and that he might ſend them to preach. \V And
he gaue them power to cure infirmities, and to caſt out Diuels. \V And
he gaue to Simon the name
\LNote{Peter}{Peter
\MNote{Peters preeminence.}
in numbring the Twelue is alwayes the firſt, and his name is ſo giuen
 him for ſignification of his calling to be the
\CNote{\XRef{Mt.~16.}}
Rock or Foundation of
 the Church vnder Chriſt: as here alſo the name \Sc{Boanerges} is giuen
 to other two Apoſtles for ſignification, and ſo diuers names elſwhere
 in the old Teſtament and in the new.}
Peter. \V And Iames of Zebedee, and Iohn the brother of Iames; and he
called their names, \Emph{Boanerges}, which is, \Emph{the Sonnes of
thunder}. \V And Andrew and Philippe, and Bartlemew and Matthew, and
Thomas and Iames of Alphæus, and Thaddæus and Simon Cananæus, \V and
Iudas Iſcariote, who alſo betrayed him.

\V And they come to a houſe; and the multitude reſorteth togeather
againe, ſo that they could not ſo much as eate bread. \V And when his
had heard of it, they went forth to lay hands on him. For they ſaid that
he was become
\SNote{See here the conceit of worldly friendes, who thinke the Zeale of
Religion, madnes: and therfore count thẽ mad, that are Zelous in Gods
cauſe, and for the Catholike faith: and the more Zelous, the more mad.}
mad. \V And the Scribes which were come downe from Hieruſalem, ſaid:
\CNote{\XRef{Mat.~12,~24.}
\XRef{Luc.~11,~15.}}
That he hath Beelzebub; and that in the Prince of Diuels he caſteth out
Diuels. \V And after he had called them togeather, he ſaid to them in
parables: How can Satan caſt out Satan? \V And if a
\LNote{Kingdom againſt Kingdom}{As
\MNote{Diſſenſion of Heretikes.}
this is true in al Kingdoms and Commonweales where Ciuil diſſenſion
 raigneth,
\CNote{\Cite{Greg. li.~9. ep.~3.}}
ſo is it ſpecially verified in hereſies & Heretikes, which
 haue alwayes diuiſions among themſelues as the plague of God, for
 diuiding them ſelues and others from the Church.}
Kingdom be deuided againſt it ſelf, that Kingdom can not ſtand. \V And
if a houſe be deuided againſt it ſelf, that houſe can not ſtand. \V And
if Satan be riſen againſt himſelf, he is deuided, and can not ſtand, but
hath an end. \V No body can rifle the veſſel of the ſtrong, being entred
into his houſe, vnles he firſt bind the ſtrong, & then ſhal he rifle his
houſe. \V Amen I ſay to you, that al ſinnes ſhal be forgiuen the ſõnes
of men, & the blaſphemies wherewith
%%% o-2194
they ſhal blaſpheme: \V But
%%% 2375
he that ſhal blaſpheme againſt the Holy Ghoſt, he hath not forgiueneſſe
for euer, but ſhal be guilty of an
\LNote{Eternal ſinnes}{That
\MNote{Venial ſinnes forgiuen after death.}
which is here called eternal, is
\CNote{\XRef{Mt.~12,~32.}}
(as S.~Matthew expreſſeth it) that
 which ſhal neither be remitted in this life, nor in the life to
 come. Where we learne by S.~Marke, that there are alſo ſinnes not
 eternal; and by S.~Matthew, that they are ſuch, as ſhal be forgiuen
 either here, or in the life to come.}
eternal ſinne. Becauſe they ſaid: He hath an vncleane Spirit.

\V And
\CNote{\XRef{Mt.~12,~46.}
\XRef{Lu.~8,~19.}}
there come his mother and brethren; and ſtanding without they
ſent vnto him calling him, \V and the multitude ſate about him; and they
ſay to him: \V Behold thy mother and thy brethren without ſeeke thee. \V
And anſwering them, he ſaid:
\LNote{Who is my mother}{Neither
\MNote{Spiritual kinred and buſines preferred before carnal and
worldly.}
is it here ſaid, that he had no mother, as ſome vpon theſe wordes falſly
gather; nor ingratitude to our parents is taught vs by this anſwer: but
we be hereby admoniſhed to preferre the ſpiritual mother of the
Faithful, which is the Church Catholike, and our brethren in her, and
their ſpiritual good, aboue our carnal parents or kinne. For ſo our
Maiſter being occupied here about heauenly things, accounted al them his
mother and brethren, which did the wil of his Father, in which number
our Lady his mother was alſo included, for ſhe did his Fathers wil.
\Cite{Aug. ep.~38.}
\MNote{The B.~Virgin without ſinne.}
Yea and aboue al others, becauſe ſhe had ſo much grace giuen her that
ſhe neuer ſinned not ſo much as venially in al her life.
\Cite{Aug. de nat. & grat. c.~36.}}
Who is my mother and my brethren? \V And
looking about vpon them which ſate round about him, he ſaith: Behold my
mother and my brethren. \V For whoſoeuer ſhal doe the wil of God, he is
my brother and my ſiſter and mother.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
