%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Mark
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 2398
%%% o-2221
\startChapter[
  title={Chapter 13}
  ]

\Summary{To his Diſciples (by occaſion of Hieruſalem and the Temples
  deſtruction) he foretelleth, 5.~what things shal be before the
  conſummation of the world, as ſpecially the Churches ful preaching
  vnto al Nations. 14.~Then, what shal be in the very conſummation, to
  wit, Antichriſt with his paſsing great perſecution and ſeduction, but
  for a short time. 24.~then incontinent the day of Iudgement, 28.~to
  our great comfort in thoſe miſeries vnder Antichriſt. 32.~As for the
  moment, to vs it perteineth not to knowe it. 33.~but rather euery man
  to watch, that we be not vnprouided when he cometh to each one
  particularly by death.}

%%% o-2222
And
\CNote{\XRef{Mt.~24,~1.}
\XRef{Luc.~21,~5.}}
%%% Not marked in either version
\SNote{See the
\XRef{Annotatiõs vpon S.~Matthew. c.~24.}}
when he went out of the Temple, one of his Diſciples ſaith to him:
Maiſter, behold what manner of ſtones, and what kind of buildings. \V
And \Sc{Iesvs} anſwering, ſaid to him: Seeſt thou al theſe great
buildings? There ſhal not be left a ſtone vpon a ſtone, that ſhal not be
deſtroied.

\V And when he ſate in Mount-oliuet againſt the Temple, Peter and Iames
and Iohn and Andrew asked him apart: \V Tel vs,
\LNote{When ſhal theſe things be?}{The
\MNote{Arch-heretikes be falſe-Chriſts and falſe-Prophets.}
miſeries which did fal before the deſtruction of the Temple and citie of
Hieruſalem, were a reſemblance of the extreme calamitie that ſhal befal
before the later day at the time of Antichriſt: whereupon Chriſt
ſpeaketh indifferently of both.}
when ſhal theſe things be? and what ſhal be the ſigne when al theſe
things ſhal begin to be conſumate? \V And \Sc{Iesvs} anſwering began to ſay
to them, See, that no man ſeduce you. \V For many ſhal come in my name
ſaying that
\LNote{I am he}{As before the deſtruction of Hieruſalem, diuers Seducers
aroſe,
and called themſelues Chriſtes, promiſing the People deliuerance from
the feares and dangers, they were in of forraine ſouldiars; ſo ſhal
there come many towards the end of the world and make themſelues Chriſts
& Authours of Sects, and ſhal gaine many Diſciples; as in plaine words
foloweth in this chapter v.~22.\ \Emph{There shal riſe vp Falſe Chriſts,
and Falſe Prophets &c.}}
I am he; and they ſhal ſeduce many. \V
%%% Not marked in either version, should these be marginal notes?
\SNote{Great warres toward the later end.}
And when you ſhal heare of warres and bruits of warres, feare not. For
theſe things muſt be, but the end is not yet. \V For Nation shal riſe
againſt Nation and Kingdom againſt Kingdom, and there ſhal be
earth-quakes in places, & famines.
%%% Not marked in either version, should these be marginal notes?
\SNote{Much perſecutiõ of the faithful and Catholike men in the later
end.}
Theſe things are the beginning of ſorowes. \V But looke to your
ſelues. For they ſhal deliuer you vp in Councels, and in Synagogues ſhal
you be beaten, and you ſhal ſtand before Preſidents and Kings for my
ſake, for a teſtimonie vnto them. \V And into al Nations firſt the
Ghoſpel muſt be preached. \V And when they ſhal lead you and deliuer
you, be not careful before hand what you ſhal ſpeake; but that which
ſhal be giuen you in that houre, that ſpeake ye. For it is not you that
ſpeake, but the Holy Ghoſt. \V
%%% Not marked in either version, should these be marginal notes?
\SNote{Great treacherie and many falſe brethren at the ſame time.}
And brother ſhal deliuer brother vnto death, and the father his
\Fix{Sonne;}{ſonne;}{obvious typo, fixed in other}
and the children ſhal ariſe againſt the parents, and ſhal worke their
death. \V And you ſhal be odious to al men for my name. But he that ſhal
endure vnto the end, he ſhal be ſaued.

\V
%%% Not marked in either version, should these be marginal notes?
\SNote{Conſtancie & perſeuerance neceſſarie in time of perſecution.}
And when you ſhal ſee
\CNote{\XRef{Dan.~9,~27.}
\XRef{2.~Theſſ.~2,~4.}}
\LNote{The abomination of deſolation}{No
\MNote{Caluiniſme tendeth to the abomination of deſolation.}
hereſie doth ſo properly and purpoſely tend to this abomination of
deſolation
\CNote{\Cite{Hyppolyt. li. de Antichriſto.}}
which by Antichriſt ſhal be atchieued, as this Caluiniſme,
which taketh away with other Sacraments and external worship of God, the
very ſacrifice of Chriſtes Body and bloud, which being taken away
\CNote{\Cite{Cypr. de Cœn. Do. nu.~2.}}
(as S.~Cyprian ſaith) no religion can remaine.}
\Emph{the abomination of deſolation}, ſtanding where it ought not (he
that readeth, let him vnderſtand) then they that are in Iewrie, let them
flee vnto the mountaines: \V and he that is on the houſe-top, let him
not goe downe into the houſe, not enter in to take any thing out of his
houſe: \V and he that ſhal be in the field, let him not returne back to
take his garment. \V And woe to them that are with-child and that giue
ſuck in thoſe daies. \V But pray that the things chance not in the
winter. \V For thoſe daies ſhal be ſuch tribulations as were not from
the beginning of the creation that God
%%% 2399
created vntil now, neither ſhal
%%% o-2223
be. \V And vnles the Lord had ſhortned the dayes, no fleſh ſhould be
ſaued: but for the elect which he hath elected, he hath
\SNote{Antichriſts reigne shal be but three yeres and a halfe,
\XRef{Dan.~6.}
\XRef{Apoc.~13.}}
ſhortned the daies. \V And then if any man ſhal ſay vnto you, Loe, here
is Chriſt, loe, there: doe not beleeue. \V For there ſhal riſe vp
falſe-Chriſts and falſe-Prophets, and they shal ſhew
\LNote{Signes and wonders}{Falſe-Chriſtes and Falſe-Prophets be
ſeducers, who in the
later day by the power of the Diuel shal ſeeme to worke wonders, and yet
men muſt not beleeue them; Much leſſe theſe, which for their falſe faith
can not shew ſo much as one falſe miracle.}
ſignes and wonders, to ſeduce (if it be poſſible) the elect alſo. \V You
therfore take heed; behold I haue foretold you al things.

\V But in thoſe daies after that tribulation
\CNote{\XRef{Ez.~31,~7.}
\XRef{Ioel.~3,~15.}}
the ſunne shal be darkned,
and the moone shal not giue her light, \V and the ſtarres of Heauen shal
be falling downe, and the powers that are in Heauen, shal be moued. \V
And then they shal ſee the
\CNote{\XRef{Dan.~7,~13.}}
Sonne of man comming in the clouds, with much
power and glorie. \V And then shal he ſend his Angels, and shal gather
togeather his elect from the foure winds, from the vttermoſt part of the
earth to the vttermoſt part of Heauen. \V And of the figtree learne ye a
parable. When now the bough thereof is tender, and the leaues come
forth, you know that ſommer is very nigh. \V So you alſo when you shal
ſee theſe things come to paſſe, know ye that it is very nigh, at the
doores. \V Amen I ſay to you, that this Generation shal not paſſe, vntil
al theſe things be done. \V Heauẽ & earth shal paſſe, but my words shal
not paſſe.

\V But of that day or houre no man knoweth, neither the Angels in
Heauen, nor the
\SNote{Not as though himſelf knew not, as the Heretikes Agnoitæ held;
but becauſe he knew it not for to teach it others, as being not
expedient.
\Cite{Amb. de fide li.~5. c.~8.}}
Sonne, but the Father. \V Take heed, watch, and pray. For you know not
when the time is. \V Euen as a man who being gone into a ſtrange
countrie, left his houſe; and gaue his ſeruants authoritie ouer each
worke, and commanded the porter to watch. \V Watch ye therfore (for you
know not when the Lord of the houſe commeth: at euen, or at midnight,
or at the cock crowing, or in the morning) \V leſt comming vpon a ſoden,
he find you ſleeping. \V And that which I ſay to you, I ſay to al,
Watch.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
