%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Mark
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 2400
%%% o-2224
\startChapter[
  title={Chapter 14}
  ]

\Summary{Iudas by occaſion of Marie Magdalens ointment, doth ſel him to
the Councel of the Iewes. 12.~After the Paſchal lambe he giueth them
the bread of life
\XRef{(Io.~6.)}
in a myſtical Sacrifice or ſeparation of his bodie and bloud, 27.~and
that night, is after his prayer, 43.~taken of the Iewes men, Iudas
being their Captaine: is forſaken of the other eleuen for feare:
53.~is falſly accuſed and impiouſly condemned of the Iewes Councel:
65.~and shamefully abuſed of them: 66.~and thriſe denied of Peter. Al
euen as the Scriptures and himſelf had often foretold.}

And
\CNote{\XRef{Mat.~26,~1.}
\XRef{Lu.~22,~1.}}
the Paſche was, and the Azymes after two daies: and the cheefe
Prieſts and the Scribes ſought how they might by ſome wile lay hands on
him, and kil him. \V But they ſaid: Not on the feſtiual day, leſt there
might be a tumult of the People.

\V And
\CNote{\XRef{Mat.~26,~6.}
\XRef{Ioan.~12,~3.}}
when he was at Bethania in the houſe of Simon the Leper, and ſate
at meate, there came a woman hauing an alabaſter boxe of ointment, of
pretious ſpike-narde; and breaking the alabaſter-boxe, she powred it out
vpon his head. \V But there were certaine that had indignation within
them ſelues, and ſaid: Whereto is
\LNote{This waſt}{Religious
\MNote{Coſt religiouſly beſtowed vpon Chriſt & his Church.}
offices done to Chriſt for ſignification, deuotion, or honour ſake, both
then in his life, and now in the Holy Sacrament, be of ſome (vnder
pretence of better beſtowing ſuch things vpon the poore) condemned
vniuſtly.}
this waſt of the ointment made? \V For this ointment might haue been
ſold for more then three hundred pence, and giuen to the poore. And they
murmured againſt her. \V But \Sc{Iesvs} ſaid:
\LNote{Let her alone}{Chriſt anſwereth for the deuout woman, and for
defence of her
\Fix{fact,}{act,}{probable typo, same in other}
as we muſt anſwer againſt the ignorant and il men, when they blame good
men for giuing their goods to the Church.}
Let her alone, why doe you moleſt her? ſhe hath wrought a good worke
vpon me. \V For the poore you haue alwaies with you; and when you wil,
you may doe them good; but
\SNote{We haue not Chriſt here needing our almes, as when he conuerſed
vpon the earth. See
\XRef{Mat.~26,~11.}}
me you haue not alwaies. \V That which ſhe had, she hath done; she hath
preuented to anoint my body to the burial. \V Amen I ſay to you,
whereſoeuer this Ghoſpel shal be preached in the whole world, that alſo
which she hath done, shal be told for a memorie of her.

\V And
\CNote{\XRef{Mat.~26,~17.}
\XRef{Lu.~22,~7.}}
Iudas Iſcariote one of the Twelue went his way to the cheefe
Prieſts, for to betray him to them. \V Who
%%% o-2225
hearing were glad; and they promiſed him that they would giue him
money. And he ſought how he might betray him conueniently.

\V And the firſt day of the Azymes when they ſacrificed the Paſche, the
Diſciples ſay to him: Whither wilt thou that we goe, and prepare for
thee to eate the Paſche? \V And he ſendeth two of his Diſciples, and
ſaith to them: Goe ye into the citie; and there shal meet you a man
carying a pitcher of water, folow him; \V and whitherſoeuer he entreth,
ſay to the Maiſter of the houſe, that the Maiſter ſaith, where is my
refectorie, where I may eate the Paſche with my Diſciples? \V And he wil
shew you a great chamber, adorned; and there prepare for vs. \V And his
Diſciples went their waies, and came into the citie; and they found as
he had told them, and they prepared the Paſche. \V And
\CNote{\XRef{Io.~13,~21.}}
when euen was
come, he commeth with the Twelue. \V And when they were ſitting at the
table and eating, \Sc{Iesvs} ſaid: Amen I ſay to you, that one of you
shal betray me, he that eateth with me. \V But
%%% 2401
they began to be ſad, and to ſay to him ſeuerally, Is it I? \V Who ſaid
to them: One of the Twelue, he that dippeth with me his hand in the
dish. \V And the Sonne of man in deed goeth,
\CNote{\XRef{Pſ.~40,~10.}}
as it is written of him,
but woe to that man by whom the Sonne of man shal be betrayed. It were
good for him, if that man had not been borne. \V And
\CNote{\XRef{Mt.~26,~26.}
\XRef{Luc.~21,~19.}
\XRef{1.~Cor.~11,~24.}}
whiles they were eating, \Sc{Iesvs} tooke
\LNote{Bread}{\Emph{This
\MNote{The real preſence by Conſecration.}
is bread before the Sacramental wordes, but the
Conſecration once done, of bread is made the flesh of Chriſt.}
\Cite{Ambroſ. li.~4. c.~4. de Sacramentis.}}
bread: and bleſſing brake, and gaue to them, and ſaid: Take, \Sc{This is
%%% !!! The note for verse 24, also applies to this one. How should we
%%% handle this?  Note the use of ibedem in the Citation for the note to
%%% verse 23.
%%% \LNote{}{}
my body.} \V And taking the
\LNote{Chalice}{Wine
\MNote{Very bloud vnder the forme of wine.}
and water is put into the Chalice, but is made bloud by Conſecration of
the heauenly Word: though to auoid the lothſomneſſe which would be in
the ſight of bloud, thou receaueſt that which hath the likenes and
reſemblance thereof.
\Cite{Ambr. ibedem.}}
chalice, giueing thankes he gaue to them. And they
\SNote{Al dranke, to wit, al the twelue; for moe were not
preſent. Whereby is euident that the words in
\XRef{S.~Mat. (26,~27)}
\Emph{Drinke ye al of this}, were ſpoken to al the Apoſtles only, which
here are ſaid that they al did drinke. And ſo it is no general
commandement to al men.}
al dranke of it. \V And he ſaid to them: \Sc{This is
\LNote{My bloud}{\Emph{Whoſoeuer
\MNote{Faith neceſſarie in the B.~Sacrament.}
beleeueth it not to be true that is ſaid, he falleth from grace and
ſaluation.}
\Cite{Epiph. in Ancorato.}
\Emph{Let vs euer giue credit to God & neuer reſiſt him, though the
thing that he ſaith ſeeme neuer ſo abſurd in our imagination, or farre
paſſe al our ſenſe and vnderſtanding. For his words can not beguile vs,
but our ſenſe may eaſily be deceaued. Seeing therfore that he ſaid, This
is my body, let vs neuer doubt of the matter.}
\Cite{Chryſoſt. ho.~83. in Mat. ſub finem.}}
my blovd of the new Teſtament, that shal be shed for many.} \V Amen I
ſay to you, that now I wil not drinke of the fruit
\SNote{See
\XRef{Annot. in Mt. c.~26.~29.}}
of the vine vntil that day when I shal drinke it new in the Kingdom of
God. \V And an hymne being ſaid, they went forth into Mount-oliuet.

\V And \Sc{Iesvs} ſaith to them: You shal al be ſcandalized in me in
this night; for it is written,
\CNote{\XRef{Zac.~13,~7.}}
\Emph{I wil ſtrike the Paſtour, & the
sheepe shal be diſperſed.} \V But after that I shal be riſen againe, I
wil goe before you into Galilee. \V And Peter ſaid to him: Although al
shal be ſcandalized, yet not I. \V And \Sc{Iesvs} ſaith to him: Amen I
ſay to thee, that thou this day in this night, before the cock crow
twiſe, shal thriſe deny me.
%%% o-2226
\V But he ſpake more vehemently: Although I should die togeather with
thee, I wil not denie thee. And in like manner alſo ſaid they al.

\V And they come into a farme-place called Gethſemani. And he ſaith to
his Diſciples: Sit you here, vntil I pray. \V And he taketh Peter and
Iames and Iohn with him; and he began to feare and to be heauy. \V And
he ſaith to them: My ſoul is ſorowful euen vnto death; ſtay here, and
watch. \V And when he was gone forward
\Fix{at}{a}{obvious typo, fixed in other}
litle, he fel flat vpon the ground; and he prayed that if it might be,
the houre might paſſe from him; \V and he ſaid: Abba, Father, al things
are poſſible to thee, transferre this chalice from me, but not that
which I wil, but that which thou. \V And he commeth, and findeth them
ſleeping. And he ſaith to Peter: Simon, ſleepeſt thou? could thou not
watch one houre? \V Watch ye, and pray that you enter not into
tentation. The ſpirit in deed is prompt, but the flesh infirme. \V And
going away againe, he prayed, ſaying the ſelf ſame word. \V And
returning, againe he found them aſleep (for their eyes were heauy) and
they wiſt not what they ſhould anſwer him. \V And he commeth the third
time, and ſaith to thẽ: Sleep ye now, and take reſt. It ſufficeth: the
houre is come; behold the Sonne of man ſhal be betrayed into the hands
of ſinners. \V Ariſe, let vs goe. Behold he that ſhal betray me, is at
hand.

\V And
\CNote{\XRef{Mt.~26,~47.}
\XRef{Luc.~22,~47.}
\XRef{Io.~18,~3.}}
as he was yet ſpeaking, commeth Iudas Iſcariote one of the
Twelue, and with him a great multitude with ſwords and clubs, from the
cheefe Prieſts and the Scribes and the Ancients. \V And the betrayer of
him had giuen them a ſigne, ſaying: Whomſoeuer I ſhal kiſſe, it is he, lay
hold on him, and lead him warily. \V And when he was come, immediatly
going to him, he ſaith: Rabbi, and he kiſſed him. \V But they laid hands
vpon him, and held him. \V And one certaine man of the ſtanders about
drawing out a ſword, ſmote the ſeruant of the cheefe Prieſt, and cut of
his eare. \V And \Sc{Iesvs} anſwering,
%%% 2402
ſaid to them: As to a theefe are you come out with ſwords and clubs to
apprehend me? \V I was daily with you in the temple teaching, and you
did not lay hands on me. But, that the Scriptures may be fulfilled. \V
Then his Diſciples leauing him, al fled. \V And a certaine yong man
folowed him clothed with ſindon vpon the bare; and they tooke him. \V
But he caſting off the ſindon, fled from them naked.

%%% o-2227
\V And they brought \Sc{Iesvs} to the cheefe Prieſt; and al the Prieſts
and the Scribes and the Ancients aſſembled togeather. \V And Peter
folowed him a farre off euen vnto the court of the high Prieſt; and he
ſate with the ſeruants at the fire, and warmed himſelf. \V And the
cheefe Prieſts and al the Councel ſought teſtimonie againſt \Sc{Iesvs},
that they might put him to death, neither found they. \V For many ſpake
falſe witnes againſt him; and the teſtimonies were not conuenient. \V
And certaine riſing vp, bare falſe witnes againſt him, ſaying: \V That
we heard him ſay,
\CNote{\XRef{Io.~2,~19.}}
I wil diſſolue this temple made with hand, and in
three daies wil I build an other not made with hand. \V And their
teſtimonie was not conuenient. \V And the high Prieſt riſing vp into the
middes, asked \Sc{Iesvs}, ſaying: Anſwereſt thou nothing to theſe things
that are obiected to thee of theſe? \V But he held his peace and
anſwered nothing. Againe the high Prieſt asked him, and ſaid to him: Art
thou Chriſt the Sonne of the Bleſſed God? \V And \Sc{Iesvs} ſaith to
him: I am. And you ſhal ſee the
\CNote{\XRef{Dan.~7,~13.}}
Sonne of man ſitting on the right hand
of power, and comming with the clouds of Heauen. \V And the high Prieſt
renting his garments, ſaith: What need we witneſſes any further? \V You
haue heard blaſphemie. How thinke you? Who al
\SNote{Here we may ſee that they, were worthily reprobated and forſakẽ,
according to our Sauiours prediction by the parable
\XRef{(Mat.~12.)}
\Emph{The kingdom of God shal be taken frõ you &c.}}
condemned him to be guilty of death. \V And certaine began to ſpit vpon
him, and to couer his face, and to beat him with buffets, and to ſay
vnto him, Prophecie; and the ſeruants gaue him blowes.

\V And when Peter was in the court beneath, there commeth one of the
\SNote{He feareth not afterward Rome the Lady of the world, that in the
houſe of Caiphas was afraid of the high Prieſtes wẽch.
\Cite{Leo in Natiu. Pet. & Pau. ſer.~2.}}
woman-ſeruants of the high Prieſt. \V And when ſhe had ſeen Peter
warming him ſelf, beholding him ſhe ſaith: And thou waſt with \Sc{Iesvs}
of Nazareth. \V But he denied, ſaying: Neither know I, neither wot I
what thou ſaieſt. And he went forth before the court; and the cock
crew. \V And againe a wench ſeeing him, began to ſay to the ſtanders
about, That this fellow is of them. \V But he denied againe. And after a
while againe they that ſtood by, ſaid to Peter: Verily thou art of them;
for thou art alſo a Galilæan. \V But he
\LNote{He began to curſe}{In
\MNote{Peter.}
this one Apoſtle Peter, the firſt and cheefe in the order of
Apoſtles, in whom the Church was figured, both ſortes were to be
ſignified, to wit, the ſtrong and the weake, becauſe without both the
Church is not.
\Cite{Aug. ſer.~13. de verb. Do.}
Againe, Our Sauiour would ſhew by the example of the cheefe Apoſtle,
that no man ought to preſume of him ſelf, when neither S.~Peter could
auoid the danger of mutability.
\Cite{Aug. tract.~26. in Euang. Io.}
\Cite{Leo Ser.~9. de Paſs. Do.}}
began to curſe and to ſweare, That I know not this man whom you ſpeake
of. And immediatly the cock crew againe. And Peter remembred the word
that \Sc{Iesvs} had ſaid vnto him: Before the cock crow twiſe, thou ſhalt
thriſe deny me. And he began to weep.

\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
