%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Nehemias
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 0995
%%% o-0894
\startChapter[
  title={Chapter 8}
  ]

\Summary{Eſdras readeth the law before the people. 9.~Nehemias
  comforteth them. 13.~They celebrate the feaſt of tabernacles ſeuen
  dayes: 13.~& of collection the eight day.}

And al the people was gathered togethor as it were one man to the
ſtreate, which is before the water and they ſayd to Eſdras the ſcribe,
that he ſhould bring the booke of the law of Moyſes, which our Lord had
commanded Iſrael. \V Eſdras therfore the prieſt
\SNote{The people requeſted Eſdras, to \Emph{bring the booke} of the
law, and he \Emph{brought it}, neither is there anie mention that he
writte the whole law out of his memorie or by miracle which maketh it
probable that al copies were not burned or loſt, but ſome reſerued by
Ieremias, Ezechiel, Daniel, Aggeus, Zacharias, or by him ſelfe or others
out of which he collected one intire volume correcting faultes committed
by ſcribes, & adding ſome thinges for explication ſake & ſupplement of
the hiſtories: and that either by tradition or reuelation.}
brought the law before the multitude of men and wemen, and al that could
vnderſtand, in the firſt day of the ſeuenth moneth. \V And he read it
playnly in the ſtreate that was before the water gate, from morning
vntil midday, in the preſence of the men, and wemen, and of
\Fix{theſe}{thoſe}{obvious typo, fixed in other}
that vnderſtood: and the eares of al the people were attent to the
booke. \V And
\Fix{Eſdaas}{Eſdras}{obvious typo, fixed in other}
the ſcribe ſtood vpon a ſteppe of wood, which he made to ſpeake vpon:
and there ſtood by him Mathathias, and Semeia, and Ania, and Vria, and
Helcia, and Maaſia, on his right hand: and on the left: Phadaia, Miſael,
and Melchia, and Haſum, and Haſbadana, Zacharia, and Moſollam. \V And
Eſdras
%%% o-0895
opened the booke before al the people: for he appeared aboue al
the people: and when he had opened it, al the people ſtood. \V And
Eſdras bleſſed our Lord the great God: and al the people
\Fix{anſwred:}{anſwered:}{likely typo, fixed in other}
Amen, amen: lifting vp their handes, and they bowed, and adored God
flatte on the earth. \V
\Fix{Moerouer}{Moreouer}{obvious typo, fixed in other}
Ioſue, & Bani, and Serebia, Iamin, Accub, Septhai, Odia,
%%% 0996
Maaſia, Celita, Azarias, Iozabed, Hanan, Phalaia: Leuites made ſilence
in the people to heare the law: and the people ſtoode in their
degree. \V And they read in the booke of the law of God diſtinctly and
playnly, for to vnderſtand: and they vnderſtood when it was read. \V And
Nehemias ſaid (the ſame is
\SNote{\Emph{Atherſatha priuileged}, by reaſon of his fauour with king
Artaxerxes.
\XRef{chap.~2.}}
Atherſatha) and Eſdras the Prieſt and ſcribe, and the Leuites
interpreting to al the people: It is a day ſanctified to the Lord our
God, mourne ye not, and weepe not.
\CNote{\XRef{1.~Eſd.~2.}}
For al the people wept, when they heard the wordes of the law. \V And he
ſaid to them: Goe, eate fat thinges, and drinck the ſweete wine, and
ſend portions to them, that haue not prepared for them ſelues: becauſe
it is the holie day of our Lord, and be not ſadde: for the ioy of our
Lord is our ſtrength. \V And the Leuites made ſilence in al the people,
ſaying: Hold your peace, becauſe the day is holie, and be not
ſorowful. \V Therefore al the people went to eate and drinck, and to
ſend portions, and to make great ioy: becauſe they vnderſtood the
wordes, that he had taught them. \V And in the ſecond day were gathered
the princes of the families of al the people, the Prieſts and Leuites to
Eſdras the ſcribe, that he ſhould interpret vnto them the wordes of the
law. \V And they found written in the law, that our Lord commanded in
the hand of Moyſes, that the children of Iſrael ſhould dwel in
tabernacles, on the ſolemne day, the ſeuenth moneth: \V and that they
ſhould proclame and publiſh a voice in al their cities, and in
Ieruſalem, ſaying: Goe ye forth into the mount, and fetch branches of
the oliue tree, and branches of the moſt fayre tree, branches of the
mirtle tree, and boughes of the palme trees, and branches of the thicke
leaued tree, that tabernacles may be made, as it is written. \V And the
people went forth, and brought. And they made them ſelues tabernacles
euery man in his houſe toppe, and in his courtes, and in the courtes of
the houſe of God, and in the ſtreate of the water gate, and in the
ſtreate of the gate of Ephraim. \V Therfore al the church of them, that
were returned from the captiuitie, made tabernacles, and dwelt in
tabernacles. For from the daies of Ioſue the ſonne of Nun the children
of Iſrael had not done it in ſuch ſort, vntil that day: and there was
exceeding great ioy. \V And he reade in the booke of the law of God day
by day, from the firſt day til the laſt, and they made the ſolemnitie
ſeuen dayes, & in the eight day a collect according to the rite.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
