%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Nehemias
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 0984
%%% o-0884
\startChapter[
  title={Chapter 2}
  ]

\Summary{Nehemias obtaining comiſsion from king Artaxerxes cometh to
  Ieruſalem: 11.~ſecretly vieweth the broken walles, and ruines of the
  citie, 17.~and exhorteth al the Iewes to the reedifying therof.}

And it came to paſſe in the moneth of Niſan, the twentith yeare of
Artaxerxes the king: and there was wine before him, and I lifted vp the
wine, and gaue to the king: and I was as it were languiſhing before his
face. \V And the king ſayd to me: Why is thy countenance ſad, whereas I
doe not ſee thee ſicke? this is not without cauſe, but ſome euil I know
not what is in thy hart. And I was very much excedingly afrayd: \V and I
ſayd to the king: O king for euer mayſt thou liue: why ſhould not my
countenance be heauie, becauſe the citie of the houſe of the ſepulchres
of my fathers is deſolate, and the gates therof are burnt with fire? \V
And the king ſayd to me: For what thing makeſt thou requeſt? And I
prayed the God of heauen, \V and I ſayd to the king: If it ſeme good to
the king, and if thy ſeruant do pleaſe before thy face, thou ſend me
into Iewrie to the citie of the ſepulchre of my father, and I wil build
it. \V And the king ſayd to me, and the Queene that ſate by him: Vnto
what time wil thy iourney be, and when wilt thou returne? And it pleaſed
before the king, and he ſent me: and I appoynted him a time. \V And I ſayd to
the king: If it ſeme good to the king, let him geue me letters to the
gouernours of the country beyond the Riuer, that they conduct me, til I
come into Iewrie: \V and a letter to Aſaph the keeper of the kings
foreſt, to geue me timber that I may couer the gates of the towre of the
houſe, and the walles of the citie, and the houſe that I ſhal enter
into. And the king gaue according to the good hand of my God with me. \V
And I came to the dukes of the countrie
%%% o-0885
beyond the Riuer, and gaue them the kings letters. And the king had ſent
with me captaynes of ſoldiers, and horſemen. \V And Sanaballat an
Horonite, and Tobias a ſeruant an Ammonite heard it, and
\SNote{Infideles & Heretikes are greued, that others endeuoure to
repayre the ruines of the Church in any countrie.}
were greiued with great affliction, that a man was come, which ſought
the proſperitie of the children of Iſrael. \V And I came to Ieruſalem
and was there
%%% 0985
three dayes. \V And I aroſe in the night, I and a few men with me, and I
told not anie man what God had put in my hart to doe in Ieruſalem, and
there was no beaſt with me, but the beaſt wheron I ſate. \V And I went
out by the gate of the valley by night, and before the fountayne of the
dragon, and to the gate of the dung, and I viewed the wal of Ieruſalem
broken downe, and the gates therof conſumed with fire. \V And I paſſed
to the gate of the fountayne, and to the kinges conduite, and there was
no place for the beaſt whereon I ſat, to paſſe. \V And I went vp by the
torrent in the night, and viewed the wal, and going backe I came to the
gate of the valley, and returned. \V But the magiſtrates knew not
whither I went, or what I did: yea and to the Iewes, and the Prieſtes,
and the nobles, and the magiſtrates, and the reſt that did the worke
vntil then I had ſhewed nothing. \V And I ſayd to them: You know the
affliction wherin we are, becauſe Ieruſalem is made deſolate, and the
gates thereof are conſumed with fire: Come, and let vs build the walles
of Ieruſalem, and let vs be no longer a reproch. \V And I ſhewed them
\SNote{Gods hand was clearly ſhewed in the effect of obtaining the kings
fauourable letters.}
the hand of my God, that it was good with me, and the kings wordes,
which he had ſpoken to me, and I ſayd: Let vs riſe, and let vs
build. And their handes were incouraged in good. \V But Sanaballat the
Horonite, and Tobias the ſeruant an Ammanite, and Goſem an Arabian heard
of it, and they ſcorned vs, and deſpyſed vs, and ſayd: What is this
thing, that you doe? Why do you rebel againſt the king? \V And I
anſwered them, and ſayd to them: The God of heauen he helpeth vs, and
we are his ſeruants: let vs riſe and build: but you haue no part, nor
iuſtice, nor memorie in Ieruſalem.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
