%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Nehemias
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 0987
%%% o-0888
\startChapter[
  title={Chapter 4}
  ]

\Summary{Notwithſtanding the enimies rage and oppoſe againſt the
  building of walles, 4.~the Iewes build with one hand, and hold their
  ſwordes in the other. Nehemias encorageth them, and the worke
  proceedeth.}

And it came to paſſe, when Sanaballat had heard that we builded the wal,
he was very angrie: and being
%%% 0988
moued excedingly he ſcorned the Iewes, \V and ſayd before his bretheren,
and the multitude of the Samaritanes: What doe
\SNote{It was Gods prouidence that the enemies mocking at the reparation
of Ieruſalem did not ſo furiouſly reſiſt, til the worke was performed:
ſo ſometimes heretikes ſcoffe at the endeuours of poore prieſtes
laboring to reſtore Catholique religion: but whether they ſcoffe or rage,
Gods holie worke procedeth and proſpereth.}
the ſillie Iewes? Wil the gentiles let them alone? Shal they ſacrifice
and accompliſh in one day? Why, are they able to build vp the ſtones out
of the heapes of duſt, that are burnt? \V Yea Tobias alſo the Ammanite
his neighbour ſayd: Let them build: if a fox come vp, he ſhal leape ouer
their ſtone wal. \V Heare our God, becauſe we are had in diſpiſe: turne
the reproch vpon their head, and giue them into diſpite in a Land of
captiuitie. \V Couer not their iniquitie, and let not their ſinne be put
out before thy face, becauſe they haue mocked thy builders. \V Therfore
we built the wal, and ioyned together the whole vnto the halfe part: and
the hart of the people was prouoked to worke. \V And it came to paſſe,
when Sanaballat had heard, and Tobias, and the Arabians, and the
Ammanites, and the Azotians, that the breach of the wal of Ieruſalem was
cloſed vp, and the gates began to be ſtopped, they were wrath
excedingly. \V And they were al aſſembled together, to come, and to
fight againſt Ieruſalem, and to prepare embuſhments. \V And we prayed
our God, and ſet watchmen vpon the wal day and night againſt them. \V
And Iudas ſayd: The ſtrength of him that carrieth is weakened, & the
morter is very much, and we ſhal not be able to build the wal. \V And
our enimies ſayd: Let them not know, and let them be ignorant, til we
come into the mideſt of them, and kil them, and make their worke to
ceaſe. \V And it came to paſſe, the Iewes that dwelt beſide them, coming
and telling vs ten times, out of al places from whence they came to
vs. \V I ſette the people in a place behind the wal round about in
order, with their ſwordes, and ſpeares, and bowes. \V And I looked and
roſe vp: & I ſayd to the princes and magiſtrates, and to the reſt of the
common people: Be not afrayd at their face. Remember our Lord is great
and terrible, and fight for your bretheren, your ſonnes, and your
daughters, and your wiues, and your houſes. \V And it came to paſſe,
when our enimies had heard that it was told vs, God defeted their
counſel. And we returned al to the walles, euery man to his worke. \V
And it came to paſſe from that day, the halfe part of their yongmen did
the worke, and halfe was ready to
%%% o-0889
battel, and ſpeares, &
\Fix{ſheilds,}{ſhields,}{obvious typo, fixed in other}
and bowes, and cotes of mayle, and the princes behind them in al the
houſe of Iuda, \V of
%%% 0989
them that built in the wal, and that carried burdens, and that laded
them: with one of his hands he did the worke, & with the other he held a
ſword: \V for euery one of the builders was girded with a ſword about
his reynes. And they builded, and ſounded with the trumpet nere me. \V
And I ſayd to the princes, and to the magiſtrates, & to the reſt of the
common people: The worke is great and broad, and we are ſeperated in the
wal one far from an other: \V in what place ſoeuer you ſhal heare the
ſound of the trumpet, thither runne together vnto vs: our God wil fight
for vs. \V And let our ſelues doe the worke: and let the halfe part of
vs hold ſpeares from the ryſing of the morning, til the ſtarres
appeare. \V At that time alſo I ſayd to the people: Let euery man with
his ſeruant tarry in the middes of Ieruſalem, and let there be courſes
among vs by day and night to worke. \V And I and my bretheren, and my
ſeruants, and the watchmen, that were behinde me, did not put of our
clothes: euery man was made bare only to waſhing.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
