%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Nehemias
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 0989
%%% o-0889
\startChapter[
  title={Chapter 5}
  ]

\Summary{Nehemias blameth couetous richmen, for oppreſsing the poore:
  14.~himſelfe geueth largely of his owne to the needie.}

And there was
\SNote{S.~Beda in his time lamented that ſome ſpiritual Superiors
neglected to feede their flock, either ſpiritually or temporally, and
yet exacted temporal reuenewes, and oppreſſed the poore people.
\Cite{li.~3. c.~21. in Eſd.}
How much more may we with him, with an other Nehemias, that is a
\Emph{comforter from our Lord} to correct this fault, & to releeue poore
Catholiques diſtreſſed?}
a great crie of the people, and of their wiues againſt their brethren
the Iewes. \V And there were that ſaid: Our ſonnes and our daughters are
verie manie: let vs take corne for the price of them, and let vs eate,
and liue. \V And there were that ſayd: Let vs pledge our fieldes, and
vineyardes, and our houſes, and let vs take corne in famine. \V And
others ſayd: Let vs borow money for the kings tributes, and let vs giue
our fields and vineyardes, \V and now as the fleſh of our bretheren, ſo
is our fleſh: and as their children ſo alſo our children. Behold we
ſubdew our ſonnes, and our daughters into bondage, and our daughters
there are bondwemen, neither haue we wherewith they may be redemed, and
our fieldes and our vineyardes other men doe poſſeſſe. \V And I was
exceding angrie, when I had heard their crie according to theſe
wordes. \V And my hart thought with my ſelfe: and I rebuked the princes
and magiſtrates, and ſayd to them: Do you euery one exact vſuries of
your
\Fix{brethen?}{brethren?}{obvious typo, fixed in other}
and I gathered againſt them a great aſſemblie, \V and I ſayd to them: We as
you know, haue redemed our bretheren the
%%% 0990
Iewes, that were ſold to the gentiles, according to our abilitie: and
wil you therfore ſel your bretheren, and ſhal we redeme them? And they
held their peace, neither did they
%%% o-0890
find what to anſwer. \V And I ſayd to them: The thing is not good, which
you doe: why walke
%%% !!! SNote not marked in this
\SNote{A right propertie of a true paſtor, to do that good which he
preacheth to others.
\Cite{S.~Beda ibidem.}}
you not in the feare of our God, leſt we be vpbrayded with al the
gentiles our enimies? \V I alſo and my bretheren, and my ſeruantes, haue
lent money and corne to very manie: let vs not aske this againe in
common, let vs remitte them the debt, that is dewe to vs. \V Reſtore ye
to them their fieldes this day, and their vineyards, and their oliuets,
and their houſes: yea the hundreth part alſo of money, corne, wine, and
oyle, which you were wont to exact of them, giue it rather for them. \V
And they ſayd: We wil reſtore, and we wil ſeeke nothing of them: and we
wil do ſo as thou ſpeakeſt. And I called the Prieſtes, and adiured them,
to do according to that which I had ſayd. \V Moreouer I ſhooke my
boſome, and ſayd: God ſo ſhake euery man, that ſhal not accomplish this
word, out of his houſe, and out of his labours, ſo be he shaken out, and
made emptie. And al the multitude ſayd: Amen. And they prayſed God. And
the people did as it was ſayd. \V And from that day, wherin the king
commended me to the duke in the land of Iuda, from the twenteth yeare
til the two & thirteth yeare of Artaxerxes the king for twelue
\Fix{yearees,}{yeares,}{obvious typo, fixed in other}
I and my bretheren did not eate the yearly allowance, that was dewe to
the dukes. \V But the former dukes, that had bene before me, burdened
the people, and tooke of them in bread, and wine, and money euery day
fourtie ſicles: yea and their officers oppreſſed the people. But I did
not ſo for the feare of God. \V Yea I builded rather in the worke of the
wal, and field I bought not, and al my ſeruantes were gathered together
to the worke. \V
\Fix{Tke}{The}{obvious typo, fixed in other}
Iewes alſo and the magiſtrates an hundred fiftie
men, and they that came to vs out of the nations, that were round about
vs were at the table. \V And there was prepared for me day by day one
oxe, ſix choſen rammes, beſide foules, and within ten dayes diuers
wines, & many other things did I giue. Moreouer alſo the yearly allowance of
my dukedome I ſought not, for the people was very much empoueriſhed. \V
\SNote{A good conſcience hath great confidence in God and iuſtly hopeth
for reward.}
Remember me my God to good, according to al things, which I haue done to
this people.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
