%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Nehemias
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 1000
%%% o-0899
\startChapter[
  title={Chapter 10}
  ]

\Summary{Manie, in name of al, ſubſcribe to the couenant made with
  God. 30.~Namely not to marrie with ſtrangers, 31.~to keepe the Sabbath
  day, and the ſeuenth yeare. 32.~To pay oblations, 35.~Firſt fruites,
  38.~and Tithes.}

And the ſubſcribers were Nehemias, Atherſatha the ſonne of Hachelai, and
Sedecias, \V Saraias, Azarias, Ieremias, \V Pheſhur, Amarias,
Melchias, \V Hattus, Sebenia, Melluch, \V Harem, Merimuth, Obdias, \V
Daniel, Genthon, Baruch, \V Moſollam, Abia, Miamin, \V Maazia, Belgai,
Semeia: theſe were the Prieſtes. \V Moreouer Leuites, Ioſue the ſonne of
Azanias, Bennui of the children of Henadad, Cedmihel. \V And their
bretheren, Sebenia, Odaia, Celita, Phalaia, Hanan, \V Micha, Rohob,
Haſebia, \V Zachur, Serebia, Sabania, \V Odaia, Bani, Baninu. \V The
heades of the people, Pharos, Phahathmoab, Aelam, Zethu, Bani, \V Bonni,
Azgad, Bebai, \V Adonia, Begoai, Adin, \V Ater, Hezecia, Azur, \V Odaia,
Haſum, Beſai, \V Hareph, Anathoth, Nebai, \V Megphias, Moſollam,
Hazir, \V Mezſibel, Sadoc, Iedua, \V Pheltia, Hanan, Anai, \V Oſee,
Hanania, Haſub, \V Alohes, Phalea, Sobec, \V Rehum, Haſebna, Maaſia, \V
Echaia, Hanan, Anan, \V Melluch, Haran, Baana. \V And the reſt of the
people, Prieſtes, Leuites, Porteres, and ſinging men, Nathineites, and
al that ſeperated them ſelues from the peoples of the
landes to the law of God, their wiues, their ſonnes, and their
daughters. \V Al that could vnderſtand promiſing for their bretheren,
their princes, and they that came to promiſe, and ſweare that they would
walke in the law of God, which he gaue in the hand of Moyſes the ſeruant
of God, that they would do & kepe al the
%%% 1001
cõmandments of the Lord our God, and his iudgementes and his
ceremonies. \V And that we would not geue
\SNote{In al leagues & couenantes of peace thoſe articles are ſpecially
mentioned wherin breach hath bene made in former times.}
our daughters to the people of the land, and their daughters we would
not take to our ſõnes. \V The peoples of the land, which bring in things
to ſel, & al thinges to be vſed, to ſel them on the Sabbath day, we wil
not take it of them in the Sabbath, and in the ſanctified day. And we
wil let paſſe the ſeuenth yeare, and the exaction of euerie hand. \V And
we wil ordayne preceptes vpon our ſelues, to giue the third part of a
ſicle euerie yeare to the worke of the houſe of our God, \V to the
loaues of propoſition, and to the continual ſacrifice, and for a
continual holocauſt in the Sabbathes, in the Calendes, in the
Solemnities, and in the ſanctified, and for ſinne: that propitiation may
be made for Iſrael, and vnto al vſe of the houſe of our God. \V We
therfore did caſt lotes concerning the oblation of
%%% o-0900
wood betwen the Prieſtes, and the Leuites, and the people, that it
ſhould be brought into the houſe of our God by the houſes of our fathers
at ſet times, from yeare to yeare: that it might burne vpon the altar of
the Lord our God, as it is written in the law of Moyſes. \V And that we
would bring the firſt borne of our land, and the firſt fruites of al the
fruite of euerie tree, from yeare to yeare, in the houſe of our Lord, \V
and the firſt fruites of our ſonnes, and of our cattel, as it is written
in the law, and the firſt fruites of our oxen, and of our ſheepe, that they
might be offered in the houſe of our God, to the Prieſtes which miniſter
in the houſe of our God. \V And the firſt fruites of our meates, and of
our libamentes, and the fruites of euerie tree, of vintage alſo and of
oyle we wil bring to the Prieſtes, vnto the treaſurie of our God, and
the tenth part of our land to the Leuites. The Leuites them ſelues ſhal
receiue the tithes out of al the cities of our works. \V And the Prieſt
the ſonne of Aaron ſhal be with the Leuites in the tythes of the
Leuites, and the Leuites ſhal offer the tenth part of their tythe in the
houſe of our God, to the treaſurie in the treaſure houſe. \V For the
children of Iſrael and the children of Leui ſhal carie to the treaſurie
the firſt fruites of corne, of wine, and of oyle: and the ſanctified
veſſels ſhal be there, and the Prieſtes, and ſinging men, and
porters, and miniſters, and we wil not leaue the houſe of our God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
