%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Nehemias
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 0983
%%% o-0883
\startChapter[
  title={Chapter 1}
  ]

\Summary{Nehemias
\MNote{The firſt part.

Nehemias his cõmiſeration of his countrie.}
hearing the miſerable ſtate of his countrie men in Iurie, 4.~lamenteth,
faſteth, and prayeth God for their relief.}

The wordes of
\SNote{Nehemias by \Emph{his legacie}, being ſent from a king; by
\Emph{his name}, which ſignifieth comforter from our Lord; and by
\Emph{his building} againe the walles of Ieruſalem; prefigured our
Sauiour, who was ſent from God the Father, himſelfe being the comforter
of mankind, and the ſender of an other comforter the Holie Ghoſt, to
remaine with his Church.
\Cite{S.~Beda. prolog. in Nehemiam.}}
Nehemias the ſonne of Helchias. And it came to paſſe in the moneth of
Caſleu, the twenteth yeare, and I was in Suſis the caſtel. \V And Hanani
one of my brethren came, him ſelfe and men of Iuda, and I aſked them of
the Iewes, that remayned, and were left aliue of the captiuitie, and of
Ieruſalem. \V And they ſayd to me: They that remayned, and are left of
the captiuitie there in the prouince, are in great affliction, and in
reproche: and the wal of Ieruſalem is broken downe, and the gates therof
are burnt with fire. \V And when I had heard theſe maner of wordes, I ſate,
and wept, and mourned many dayes: and faſted, and prayed before the face
of the God of heauen. \V And I ſayd: I beſech thee Lord God of heauen,
ſtrong, great, and terrible, which keepeſt couenant & mercie with them
that loue thee, and keepe thy cõmandmentes: \V let thine eares be
harkning, and thine eyes open, to heare the prayer of thy ſeruant,
which I pray before thee this day, night & day for the children of
Iſrael thy ſeruantes: and I confeſſe for the ſinnes of the children of
Iſrael, in which they haue ſinned to thee: I & my fathers houſe haue
ſinned, \V we haue bene ſeduced with vanitie, and haue not kept thy
commandments, and
%%% o-0884
cerimonies and iudgement, which thou haſt commanded to Moyſes thy
ſeruant. \V Remember the word, that thou didſt command vnto Moyſes thy
ſeruant, ſaying: When you ſhal tranſgreſſe, I will depreſſe you into
peoples: \V and if you returne to me, and keepe my precepts, and doe
them, although you ſhal be led away to the vttermoſt partes of heauen,
thence wil I gather you, and bring you backe into the place, which I
haue choſen that my name ſhould dwel there. \V And they are thy
ſeruantes, and thy people: whom thou haſt redemed in thy great ſtrength,
and in thy mighty hand. \V I beſech thee Lord, let thine eare be attent
to the prayer of thy ſeruant, and
%%% 0984
to the prayer of thy ſeruants, which will feare thy name: and direct thy
ſeruant this day, and giue him mercy before this man, for I was the
kings cupbearer.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
