%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Nehemias
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 0997
%%% o-0896
\startChapter[
  title={Chapter 9}
  ]

\Summary{The people repenting in faſting and
\Fix{ſakcloth,}{ſackcloth,}{likely typo, fixed in other}
put away their wiues of ſtrange nations. 5.~Eſdras confeſſeth Gods
benefites, and the peoples ingratitude, 32.~prayeth for them, and maketh
league with God.}

And in the foure and twenteth day of the moneth the children of Iſrael
came together in faſting and ſacke-clothes, and earth vpon them. \V And
the ſeede of the children of Iſrael was
\SNote{True repentance requireth workes of penance, & eſpecially the
remouing of occaſions of ſinne: as ſeperation from euil companie,
abandoning of euil cogitations, and of much worldly pleaſure.}
ſeperated from euerie ſtrange childe: and they ſtood, and confeſſed
their ſinnes, and the iniquities of their fathers. \V And they roſe vp
to ſtand: and they read in the volume of the law of our Lord their God,
foure times in a day, and foure times they confeſſed, and adored our
Lord their God. \V And there aroſe vpon the ſteppe of the Leuites Ioſue,
and Bani, Cedmihel, Sabania, Bonni, Sarebias, Bani, and Chanani: and
they cried with a lowde voice to our Lord their God. \V And the Leuites
Ioſue and Cedmihel, Bonni, Haſebnia, Serebia, Odaia, Sebnia, and
Phathahia, ſaid: Ariſe. Bleſſe our Lord your God from eternitie to
eternitie: and let them bleſſe the high name of thy glorie in al
bleſſing & praiſe. \V Thou the ſame ô Lord, alone thou haſt made heauen,
& al the hoſt therof: the earth & al thinges that are in it: the ſeas
and al thinges that are therin: and thou doſt giue life to al theſe
thinges, and the hoſt of heauẽ adoreth thee. \V Thou the ſame ô Lord
God, which didſt chooſe Abram, & broughteſt him out of the fire of the
Chaldees, and gaueſt him the
\SNote{Changing of names importeth ſome beneficial myſterie.
\XRef{Gen.~17.}}
name Abraham. \V And thou didſt finde his hart faithful before thee: and
thou madeſt a couenante with him, that thou wouldeſt geue him the land
of the Chananite, of the Hetheite, and of the Amorrheite, and of the
Pherezeite, and of the Iebuſeite, & of the Gergeſeite, to geue vnto his
ſeede: and thou haſt fulfilled thy wordes, becauſe thou art iuſt. \V And
thou ſaweſt the affliction of our fathers in Ægypt: & their crie thou
didſt heare vpon the Read ſea. \V And thou gaueſt ſignes & wonders in
Pharao, and in al his ſeruants, and in al the people of his land: for
thou didſt know that they had done proudly againſt them: and thou madeſt
thyſelf a name, as alſo at this day. \V And thou didſt diuide the ſea
before them, and they paſſed through the midſt of the ſea in drie land:
but their perſecutors thou threweſt into the depth, as a ſtone into the
rough waters. \V And in a piller of a cloude thou waſt their leader by
day, and in a piller of fire
%%% 0998
by night, that the way might appeare to them, by the which they went. \V
To mount Sinai alſo thou didſt deſcend, and ſpakeſt with them from
heauen, and thou gaueſt them right iudgements, and the law of truth,
ceremonies, and good preceptes.
%%% o-0897
\V Thy ſanctified Sabbath thou didſt ſhew them, and the commandements,
and ceremonies, and the law thou didſt command them in the hand of
Moyſes thy ſeruant. \V Bread alſo from heauen thou gaueſt them in their
hunger, and water out of the rocke thou didſt bring forth to them
thirſting, and thou ſaideſt to them that they ſhould enter in, and
poſſeſſe the land, vpon which thou didſt lift vp thy hand to diliuer it
them. \V But they and our fathers did proudly, and hardned their neckes
and heard not thy cõmandements. \V And they
\SNote{Free wil in ſinners.}
would not heare, and they remembred not thy merueylous workes which thou
haſt done to them. And they hardned their neckes,
\CNote{\XRef{Num.~17.}}
and gaue the head to
returne to their ſeruitude, as it were by contention. But thou a
propitious God, and gratious, and merciful, long ſuffering and of much
compaſſion, didſt not forſake them. \V Yea and when they had made to
them ſelues a molten calfe, and had ſaid: This is thy god, which brought
thee out of Ægypt: and they did great blaſphemies. \V But thou in thy
manie mercies didſt not leaue them in the deſert: the piller of the
cloude departed not from them by day to lead them into the way, and the
piller of fire by night to ſhew them the way by which they ſhould
goe. \V And thou gaueſt them thy good ſpirite, which ſhould teach them,
and thy Manna thou didſt not withhold from their mouth, and thou gaueſt
them water in thirſt. \V Fourtie yeares didſt thou feede them in the
deſert, and nothing was wanting to them, their garmentes waxed not old,
and their feete not worne. \V And thou gaueſt them kingdomes, and
peoples, and didſt part lottes vnto them: and they poſſeſſed the land of
Sehon, and the land of the king Heſebon, and the land of Og the king of
Baſan. \V And thou didſt multiplie their children as the ſtarres of
heauen, & brought them to the land wherof thou hadſt ſaid to their
fathers, that they ſhould enter and poſſeſſe it. \V And the children
came, and poſſeſſed the land, and thou didſt humble before them the
inhabiters of the land, the Chananeites, and gaueſt them into their
hand, and their kinges, and the peoples of the land, that they might doe
to them as it pleaſed them. \V They therfore tooke the
%%% 0999
fenſed cities and fatte ground, and poſſeſſed houſes ful of al goodes:
ceſternes made by others, vineyardes, and oliuetes, & manie trees that
bare fruite: and they did eate, and were filled, and became fatte, and
abounded with delicious thinges in thy great goodnes. \V But they
prouoked thee to wrath, & departed from thee, and threw thy law behind
their backes: & they killed thy prophetes, which admoniſhed them
erneſtly to returne to thee: and they did great blaſphemies. \V And thou
gaueſt them into the handes of their enemies, and they afflicted
them. And in the time of their tribulation they cried to thee, & thou
from heauen didſt heare, and according to thy manie cõpaſſions gaueſt
them ſauiours, that ſhould ſaue them from the hand of their enemies. \V
And when they had reſted, they returned to do euil in thy ſight: & thou
didſt leaue them in the hand of their enemies, and they poſſeſſed
them. And
%%% o-0898
they returned, & cried to thee: & thou heardeſt from heauen, and
deliueredſt thẽ in thy mercies, manie times. \V And thou didſt admoniſh
them that they ſhould returne to thy law. But they did proudly, & heard
not thy cõmandmentes, and ſinned in thy iudgementes, which a man ſhould
do, and ſhal liue in them, and they gaue the reuolting ſhoulder, and
hardned their necke, neither did they heare. \V And thou didſt prolong
manie yeares ouer them, and didſt teſtifie to charge them in thy ſpirit
by the hand of thy Prophetes: and they heard not, and thou didſt deliuer
them into the hand of the peoples of the nations. \V But in thy very
many mercies thou madeſt them not into conſumption, neither didſt thou
forſake them: becauſe a God of compaſſions and gratious art thou. \V Now
therfore O our God, great, ſtrong, and terrible, keeping couenant and
mercie, turne not away from thy face al the labour, which hath found vs,
our Kinges, and our Princes, and our Prieſts, and our Prophetes, and our
fathers, and al the people from the daies of the king of Aſſur, vntil
this day. \V And thou art iuſt in al thinges, that haue come vpon vs:
becauſe thou haſt done truth, but we haue done wickedly. \V Our Kinges,
our Princes, our Prieſtes, and our fathers haue not done thy law, and
haue not attended thy commandmentes, and thy teſtimonies which thou haſt
teſtified among them. \V And they in their reignes, and in thy manifold
goodnes, which thou gaueſt them, and in the land moſt large and fatte,
which thou didſt deliuer in their
%%% 1000
ſight, ſerued not thee, nor returned from their moſt wicked deuiſes. \V
Behold we our ſelues this day are bondmen: and the land, which thou
gaueſt our fathers, that they ſhould eate the bread therof, and the good
thinges that are therof, and our ſelues are ſeruantes in it. \V And the
fruites therof are multiplied to the kinges, whom thou haſt ſette ouer
vs for our ſinnes, and they haue dominion ouer our bodies, and ouer our
beaſtes, according to their wil, and we are in great tribulation. \V
Therfore vpon al theſe thinges we our ſelues make a couenant, and write,
and our Princes, our Leuites, and our Prieſtes ſigne it.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
