%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Nehemias
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 0992
%%% o-0892
\startChapter[
  title={Chapter 7}
  ]

\Summary{Nehemias appointeth watchmen in Ieruſalem, 5.~and calling the
  people together reciteth the number of thoſe, which came firſt from
  Babylon. 68.~Likewiſe of their cattel, 70.~and the giftes of certaine
  chiefe men towards the reparations.}

And after the wal was built, & I had put on the doores, and numbred the
porters, and ſinging men, and Leuites: \V I commanded Hanani my brother,
and Hananias prince of the houſe of Ieruſalem (for he ſemed as it were a
true man, and one that feared God aboue the reſt) \V and I ſayd to them:
Let not the gates of Ieruſalem be opened vntil the heate of the
ſunne. And when they yet ſtood by, the gates were ſhut, and barred: and
I ſette
\SNote{Three ſpecial defences of a citie, are the ſtrength of walles,
ſhutting and opening the gates in due time, & diligẽt watchmen: ſo to
the cuſtodie of faithful ſoules three thinges are neceſſarie; the grace
of God, due regard of the outward ſenſes, and continual watch againſt
our inuiſible enemies.}
watchmen of the inhabitants of Ieruſalem, euery one by their courſes,
and euery man againſt his houſe. \V And the citie was exceding large and
great, and the people few in the middes therof, & there were no houſes
built. \V But God gaue me in my hart, and I aſſembled the princes and
magiſtrates, and common people, that I might number them: and I found a
booke of the number of them, that came vp firſt, and there was found
written in it: \V Theſe are the children of the prouince, which came vp
from the captiuitie of them that were tranſported, whom Nabuchodonoſor
the king of Babylon had tranſported, and returned into
Iurie, euery one into his owne citie. \V They
%%% 0993
that came with Zorobabel, Ioſue, Nehemias, Azarias, Rahamias, Nahamani,
Mardochæus, Belſam, Meſpharath, Begoai, Nahum, Baana. The number of the
men of the people of Iſrael. \V The children of Pharos, two thouſand an
hundred ſeuenty two. \V The children of Saphatia, three hundred ſeuentie
two. \V The children of Area, ſix hundred fiftie two. \V The children of
Phahathmoab of the children of Ioſue and Ioab, two thouſand eight
hundred eightene. \V The children of Aelam, a thouſand two hundred
fiftie foure. \V The children Zethua, eight hundred fourtie fiue. \V The
childeren of Zachai, ſeuen hundred ſixtie. \V The childeren of Bannui,
ſix hundred fourtie eight. \V The children of Bebai, ſix hundred twentie
eight. \V The children of Azgad, two thouſand three hundred twentie
two. \V The childeren of Adonicam, ſix hundred ſixtie ſeuen. \V The
children of Beguai, two thouſand ſixtie ſeuen. \V The childeren of Adin,
ſix hundred fiftie fiue. \V The childeren of Ater, childeren of
Hezecias, ninetie eight. \V The childeren of Haſem, three hundred
twentie eight. \V The children of Beſai, three hundred twentie foure. \V
The children of Hareph, an hundred twelue. \V The children of Gabaon,
nyntie fiue. \V The childeren of
\Fix{Betlehem,}{Bethlehem,}{obvious typo, fixed in other}
%%% o-0893
and Nerupha, an hundred eightie eight. \V The men of Anathoth, an
hundred twentie eight. \V The men of Bethazmoth, fourtie two. \V The men
of Cariathiacim, Cephira, and Beroth, ſeuen hundred fourtie three. \V
The men of Rama and Geba, ſix hundred twentie one. \V The men of
Machmas, an hundred twentie two. \V The men of Bethel and Hai, an
hundred twentie three. \V The men of an other Nebo, fiftie two. \V The
men of an other Aelam, a thouſand two hundred fiftie foure. \V The
childeren of Harem, three hundred twentie. The childeren of Iericho,
three hundred fourtie fiue. \V The childeren of Lod Hadid and Ono, ſeuen
hundred twentie one. \V The childeren of Senaa, three thouſand nine
hundred
\Fix{thirtitie.}{thirtie.}{obvious typo, fixed in other}
\V The Prieſts: The children of Idaia in the houſe of Ioſue, nine
hundred three. \V The childeren of Emmer, a thouſand fiftie two. \V The
childeren of Phaſhur, a thouſand two hundred fourtie ſeuen. \V The
children of Arem, a thouſand ſeuentene. The Leuites: \V The children of
Ioſue & Cedmiel, the children \V of Cenia, ſeuentie foure. The ſinging
men: \V The childeren of Aſaph, an hundred fourtie eight. \V The
porters: The childeren of Sellum, the
%%% 0994
childeren of Ater, the childeren of Telmon, the childeren of Accub, the
childeren of Hatita, the childeren of Sobai: an hundred thirtie
eight. \V The Nathineites: The childeren of Soha, the childeren of
Haſupha, the childeren of Tebbaath, \V the childeren of Ceros, the
childeren of Siaa, the childeren of Phadon, the childeren of Lebana, the
childeren of Haguba, the childeren of Seimai, \V the childeren of Hanan,
the childeren of Geddel, the childeren of Gaher, \V the childeren of Raaia,
the childeren of Raſin, the childeren of Necoda, \V the childeren of
Geſem, the childeren of Aza, the childeren of Phaſea, \V the childeren
of Beſai, the childeren of Munim, the childeren of Nephuſſim, \V the
childeren of Bacbuc, the childeren of Hacupha, the childeren of
Harur, \V the childeren of Beſloth, the childeren of Mahida, the
childeren of Harſa, \V the childeren of Bercos, the
childeren of Siſara, the childeren of Thema, \V the childeren of Naſia,
the childeren of Hatipha, \V the childeren of the ſeruantes of Salomon, the
childeren of Sotai, the childeren of Sophereth, the childeren of
Pharida, \V the childeren of Iahala, the childeren of Darcon, the
childeren of Ieddel, \V the childeren of Saphia, the childeren of Hatil, the
\Fix{chideren}{childeren}{obvious typo, fixed in other}
of Phohereth, who was borne vnto Sabaim, the ſonne of Amon. \V Al
Natheneites, and the childeren of the ſeruantes of Salomon, three
hundred nyntie two. \V But theſe be they which came vp from Thelmela,
Thelharſa, Cherub, Addon, and Emmer: and could not ſhew the houſe of
their fathers, and their ſeed, whether they were of Iſrael. \V The
children of Dalaia, the childeren of Tobia, the childeren of Necoda, ſix
hundred fourtie two. \V And of the Prieſts, the childeren of Habia, the
childeren of Accos, the childeren of Berzellai, who tooke a wife of the
daughters of Berzellai a Galeadite, and he was called by their name. \V
Theſe ſought their writing in the regiſter, and
%%% o-0894
found it not: & they were caſt out of the Prieſthood. \V And Atherſatha
ſaid to them, that they ſhould not eate of the Holies of holies, vntil
there ſtood vp a Prieſt learned and cunning. \V Al the multitude as it
were one man fourtie two thouſand three hundred ſixtie, \V beſide their
men ſeruants and wemen ſeruantes, which were ſeuen thouſand three
hundred thirtie ſeuen: and among them ſinging men, & ſinging wemen, two
hundred fourtie fiue. \V Their horſes, two hundred fourtie ſix, their
mules, two hundred fourtie fiue, \V their camels, foure
%%% 0995
hundred thirtie fiue, aſſes ſix thouſand ſeuẽ hundred twentie.

\Emph{Hitherto is reported what was written in the Regiſter. From this
place forward goeth on in order the hiſtorie of Nehemias.}

\V
\SNote{S.~Ierom here noteth whẽce he receiued eech part of this booke,
which is al Canonical Scripture being al alike ſo declared by the
Church.}
And certaine of the princes of families gaue vnto the worke. Atherſatha
gaue into the treaſure of gold a thouſand drachmas, phials fiftie,
tunikes for Prieſtes fiue hundred thirtie. \V And of the princes of
families there gaue into the treaſure of the worke of gold, twentie
thouſand drachmas, and of ſiluer two thouſand two hundred pound. \V And
that which the reſt of the people gaue, of gold twentie thouſand
drachmas, and of ſiluer two thouſand pound, and tunikes for Prieſtes
ſixtie ſeuen. \V And the Prieſtes, and Leuites, and porters, and ſinging
men, and the reſt of the common people, and the Nathineites, and al
Iſrael dwelt in their cities. And the ſeuenth moneth was come: and the
childeren of Iſrael were in their cities.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
