%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Nehemias
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 1001
%%% o-0900
\startChapter[
  title={Chapter 11}
  ]

\Summary{New inhabitantes of Ieruſalem are recited. 20.~Likwiſe who
  dwelt in other cities of Iuda.}

%%% 1002
And the princes of the people dwelt in Ieruſalem: but the reſt of the
people caſt lottes, to take
\SNote{Becauſe Ieruſalem was moſt impugned by enimies fewe were willing
to dwel there,}
one part of ten that ſhould dwel in Ieruſalem the holie citie, and nine
partes in the cities. \V And the people bleſſed al the men that had
\SNote{yet many valiant men of the tribes of Iuda Beniamin and Leui
offered them ſelues: of other tribes the tenth part were choſen by
lottes. Wherby is gathered that many of the tenne tribes returned alſo
into Chanaan, though holie Scripture doth not ſo expreſly record what
became of them as of the  other tenne tribes, becauſe Ieruſalem pertained to the
lote of Beniamin, Iuda was the Kinglie tribe, and Leui the Prieſtlie.
\Cite{S.~Beda lib.~3. cap.~31. in Eſdr.}}
voluntarily offered them ſelues to dwel in Ieruſalem. \V Theſe therfore
are the princes of the prouince, which dwelt in Ieruſalem, and in the
cities of Iuda. And euerie one dwelt in his poſſeſſion, in their cities,
Iſrael, the Prieſtes, the Leuites, the Nathineites, and the children of
the ſeruants of Salomon. \V And in Ieruſalem there dwelt of the children
of Iuda, and of the children of Beniamin: of the children of Iuda,
Athaias the ſonne of Aziam, the ſonne of Zacharias, the ſonne of
Amarias, the ſonne of Saphatias, the ſonne of Malaleel: of the children
of Phares, \V Maaſia the ſonne of Baruch, the ſonne of Cholhoza, the
ſonne of Hazia, the ſonne of Adaia, the ſonne of Ioiarib, the ſonne of
Zacharias, the ſonne of a Silonite.
%%% o-0901
\V Al theſe the childeren of Phares, which dwelt in Ieruſalem, foure
hundred ſixtie eight, valiant men. \V And theſe are the childeren of
Beniamin: Sellum the ſonne of Moſollam, the ſonne of Ioed, the ſonne of
Phadaia, the ſonne of Colaia, the ſonne of Maſia, the ſonne of Etheel,
the ſonne of Iſaia, \V and after him Gebbai, Sellai, nine hundred
twentie eight, \V and Ioel the ſonne of Zechri the ouerſeer of them, and
Iudas the ſonne of Senua ſecond ouer the citie. \V And of the Prieſtes,
Idaia the ſonne of Ioarib, Iachim, \V Saraia the ſonne of Helcias, the
ſonne of Moſollam, the ſonne of Sadoc, the ſonne of Meraioth, the ſonne
of Achitob the prince of the houſe of God, \V and their bretheren that
doe the workes of the temple: eight hundred twentie two. And Adaia the
ſonne of Ieroham, the ſonne of Phelelia, the ſonne of Amſi, the ſonne of
Zacharias, the ſonne of Pheſhur, the ſonne of Melchias, \V and his
bretheren the princes of the fathers: two hundred fourtie two. And
Amaſſai the ſonne of Azreel, the ſonne of Ahazi, the ſonne of
Moſolamoth, the ſonne of Emmer, \V and their bretheren exceding mightie:
an hundred twentie eight, and their ouerſeer Zabdiel ſonne of the
mightie ones, \V And of the Leuites Semeia the ſonne of Haſub, the ſonne
of Azaricam, the ſonne of Haſabia, the ſonne of Boni, \V and Sabathai
and Iozabed, ouer al the workes, that were without the houſe of God, of
the princes of the Leuites. \V And Mathania the ſonne of Micha, the
ſonne of
%%% 1003
Zebedei, the ſonne of Aſaph prince to prayſe, and to confeſſe in prayer,
and Becbecia ſecond of his bretheren, and Abda the ſonne of Samua, the
ſonne of Galal, the ſonne of Idithum. \V Al the Leuites in the holie
citie two hundred eightie foure. \V And the porters, Accub, Telmon, and
their bretheren, which kept the dores: an hundred ſeuentie two. \V And
the reſt of Iſrael the Prieſtes and the Leuites in al the cities of
Iuda, euerie man in his poſſeſſion. \V And the Nathineites, that dwelt
in Ophel, and Siaha, and Gaſpha of the Nathineites. \V And the ouerſeer
of the Leuites in Ieruſalem, Azzi the ſonne of Bani, the ſonne of
Haſabia, the ſonne of Mathania, the ſonne of Micha. Of the childeren of
Aſaph, the ſinging men in the miniſterie of the houſe of God. \V For the
kings commandment was vpon them, and an order among the ſinging men day
by day. \V And Phathahia the ſonne of Meſezebel of the childeren of Zara
the ſonne of Iuda in the hand of the king, according to euerie word of
the people, \V and in the houſes through al their countries. Of the
children of Iuda there dwelt in Cariatharbe, & in her daughters: and in
Dibon, and in her daughters, and in Cabſeel, and in the villages
thereof, \V and in Ieſue, and in Molada, and in Bethphaleth, \V and in
Haſerſual, and in Berſabee, & in her daughters, \V and in Siceleg, and
in Mochona, and in her daughters, and in Remmon, and in Saraa, and in
Ierimuth, \V Zanoa, Odellam, and in their townes, Lachis and in her
countries, Azeca, and in her daughters. And they abode in Berſabee vnto
the vale of Ennom. \V And the childeren of Beniamin, of Geba, Mecmas,
and Hai, and Bethhel, and her daughters,
%%% o-0902
\V in Anathoth, Nob, Anania, \V Aſor, Rama, Gethaim, \V Hadid, Seboim,
and Neballac, Lod, \V and Ono the valley of artificers. \V And of the
Leuites were portions of Iuda and Beniamin.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
