%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Nehemias
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 0985
%%% o-0886
\startChapter[
  title={Chapter 3}
  ]

\Summary{The
\MNote{The ſecond part.

How the citie was repayred with walles & people.}
High Prieſt and others beginne to repaire Ieruſalem ſtreingthning it
with an vttermoſt, 19.~middle, 28.~and innermoſt wall with towres and
gates.}

And Eliaſib the high Prieſt aroſe, and his bretheren the Prieſtes: and
they built the gate of the flocke: they ſanctified it, and ſet the
doores therof, and vnto the towre of an hundred cubites they
\SNote{Finiſhing the gate they dedicated it to Gods ſeruice, being for
defence of his holie citie: and ſo ſanctified it.}
ſanctified it, vnto the towre of Hananeel. \V And beſides him did the
men of Iericho build, and beſides him built Zachur the ſonne of Amri.
\V But the fishgate the children of Aſnaa did builde: they couered it,
%%% 0986
and ſet vp the doores therof, and the lockes, and barres. And beſides
them built Marimuth the ſonne of Vrias the ſonne of Accas. \V And
beſides him built Moſollam the ſonne of Bazachias, the ſonne of
Merezebel. And beſides them built Sadoc the ſonne of Baana. \V And
beſide them builded the Thecuenes: but their great men did not put vnder
their neckes in the worke of their Lord. \V And Ioiada the ſonne of
Phaſea, and Moſollam the ſonne of Beſodia built the oldgate, couered it,
and ſet vp the doores therof, and the lockes, and barres. \V And beſide
them built Meltias a Gabaonite, and Iadon a Meronathite, men of Gabaon
and Maſpha, for the duke that was in the country beyond the Riuer. \V
And beſide him built Eziel the ſonne of Araia a goldſmith. And beſide
him built Ananias the ſonne of a Pigmentarie: and they leift Ieruſalem
vnto the wal of the broder ſtreate. \V And beſide him built Raphaia the
ſonne of Hur, prince of the ſtreate of Ieruſalem. \V And beſide them
Iedaia the ſonne of Haromaph againſt his houſe. And beſide him built
Hattus the ſonne of Haſebomia. \V The halfe part of the ſtreate built
Melchias the ſonne of Herem, and Haſub the ſonne of Phahath Moab, and
the towre of the fornaces. \V And beſide him built Sellum the ſonne of
Alohes prince of the halfe part of the ſtreate of Ieruſalem, he and his
daughters. \V And the gate of the valley built Hanun, and the
inhabitants of Zanoe: they built it, and ſette vp the dores therof, and
the lockes, and barres, and a thouſand cubites in the wal vnto the gate
of the dunghil. \V And the gate of the dunghil built Melchias the ſonne
of Rechab, the prince of the ſtreate of Bethacaram: he built it, and
ſette vp the dores therof, and the lockes, and barres. \V And the gate
of the fountayne Sellum builded the ſonne of Colhoza, prince of the
village of Maſpha: he built it, and couered it, and ſet vp the doores
therof, & the lockes, & barres, and the walles of the poole of Siloe
vnto the kings gardẽ, and vnto the ſteppes that goe downe from the citie
of Dauid. \V After him built Nehemias the ſonne of Azboc prince of the
halfe
%%% o-0887
part of the ſtreate of Bethſur, as far as ouer againſt the ſepulchre of
Dauid, and vnto the poole, that is built with great worke, and vnto the
houſe of the valiants. \V After him the Leuites builded, Rehum the
ſonnes of Benni. After him built Haſebias prince of the halfe part of
the ſtreate of Ceila in his ſtreate. \V After him built their bretheren
Bauai the
%%% 0987
ſonne of Enadad, prince of the halfe part of Ceila. \V And beſide him
built Azer the ſonne of Ioſue, the prince of Maſpha the ſecond meaſure,
againſt the viſiting of the moſt ſure corner. \V After him in the mount
built Baruch the ſonne of Zachai the ſecond meaſure, from the corner
vnto the gate of the houſe of Eliaſib the high Prieſt. \V After him
built Merimuth the ſonne of Vrias the ſonne of Haccus, the ſecõd
meaſure, from the gate of the houſe of Eliaſib, as far as the houſe of
Eliaſib extended. \V And after him built the Prieſts, men of the
champayne of Iordan. \V After him built Beniamin & Haſub againſt their
houſe. And after him built Azarias the ſonne of Maſias the ſonne of
Ananias againſt his houſe. \V After him built Bennui the ſonne of
Hanadad the ſecond meaſure, from the houſe of Azarias vnto the bending,
and vnto the corner. \V Phalel the ſonne of Ozi againſt the bending and
the towre, which appeareth from out of the kings high houſe, that is, in
the court of the priſon. After him Phadaia the ſonne of Pharos. \V And
the Nathineites dwelt in Ophel, as far ouer againſt the gate of waters toward the
Eaſt, and the towre that ſtoode out. \V After him the Thecuenes builded
the ſecond meaſure ouer againſt, from the great and eminent towre vnto
the wall of the temple. \V And vpward to the gate of horſes the Prieſtes
built, euery man againſt his houſe. \V After them built Sadoc the ſonne
of Emmer againſt his houſe. And after him built Semaia the ſonne of
Sechenias, keeper of the Eaſt gate. \V After him built Hanania the ſonne
of Selemia, and Hanun the ſonne of Seleph the ſixth, the ſecond
meaſure. After him built Moſollam the ſonne of Barachias, againſt this
treaſurie. After him Melchias a goldſmith ſonne vnto the houſe of the
Natheneites, and of them that ſold old ſtuffe againſt the iudgement
gate, and vnto the vpper chamber of the corner. \V And within the vpper
chamber of the corner in the gate of the flocke, the goldſmithes and
marchants builded.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
