%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Nehemias
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 0991
%%% o-0890
\startChapter[
  title={Chapter 6}
  ]

\Summary{The enemies guilfully offer to make league with the Iewes,
  3.~but Nehemias procedeth in building the walles: 16.~and al bordering
  nations feare them.}

And it came to paſſe, when Sanaballat had heard, and Tobias, and Goſſem
the Arabian, and the reſt of our enimies, that I did build the wal, and
there was no breach remayning in it (howbeit at that time I had not put
the doores in the gates) \V Sanaballat and Goſſem ſent to me, ſaying:
Come, and
%%% o-0891
\SNote{VVhen heretikes & other aduerſaries of the Church finde them
ſelues not able to ſuppreſſe Catholiques, they offer conditions of
peace, and libertie to al, ſo when proteſtãtes beganne & where they are
weake, they would haue none perſecuted for anie opinions in religion;
but where they are ſtrõg they hardly grant toleration to Catholiques.}
let vs make a league together in the villages, in the field of Ono. But
they thought to doe me euil. \V I ſent therfore meſſengers to them,
ſaying: I am doing a great worke, and I can not goe downe, leſt perhaps
it be neglected when I ſhal come, and deſcend to you. \V But they ſent
to me according to this word, foure times: and I anſwered them according
to the former word. \V And Sanaballat ſent his ſeruant to me the fifth
time according to the former word, and he had a letter in his hand
written in this maner: Among the Gentiles it is heard, and Goſſem hath
ſayd, that thou and the Iewes meane to rebel, and therfore thou buildeſt
the wal, and wil aduance thy ſelfe king ouer them: for which cauſe \V
thou haſt ſette vp prophettes alſo, which should preach of thee in
Ieruſalem, ſaying: There is a king in Iurie. The king wil heare of theſe
thinges: therfore come now, that we may take counſel together. \V And I
ſent to them, ſaying: It is not done according to theſe words, which
thou ſpeakeſt: for thou frameſt theſe things of thine owne hart. \V For
al theſe terrified vs, thinking that our hands would ceaſe from the
worke, and we would leaue of. For which cauſe I did the more ſtrengthen
my hands: \V and I entred into the houſe of Samaia the ſonne of Dalaia
the ſonne of Metabeel ſecretly, who ſayd: Let vs conſult with our ſelues
in the houſe of God in the middes of the temple: and let vs ſhutte the
doores of the temple becauſe they wil come to kil thee, and in the night
they wil come to ſlea thee. \V And I ſayd: Doth any man that is like
vnto me flee? and who being as I am, wil goe into the temple, and liue?
I wil not goe in. \V And I vnderſtood that God had not ſent him, but as
it were prophicying he had ſpoken to me, and Tobias, and Sanaballat had
hyred him. \V For he had taken a price, that I being terrified ſhould do
it, and ſinne, and they might haue
%%% 0992
ſome euil to vpbraid me withal. \V Remember me Lord for Tobias and
Sanaballat, according to ſuch their workes. Yea and Noadias the
prophete, and the reſt of the prophetes that terrified me. \V But the
wal was finiſhed the fiue and twenteth day of the moneth of Elul, in two
and fiftie dayes. \V It came to paſſe therfore when al our enimies had
heard it, that al nations which were round about vs, feared, & were
diſmayed within them ſelues, and knew that this worke was done of
God. \V But in thoſe dayes alſo many letters of the principal Iewes were
ſent to Tobias, and from Tobias there came to them. \V For there were
many in Iurie ſworne vnto him, becauſe he was the ſonne in law of
Sechenias the ſonne of Area, and Iohanan his ſonne had taken the
daughter of Moſollam the ſonne of Barachias. \V Yea and they prayſed him
before me, and they reported my words vnto him: and Tobias ſent letters to
terrifie me.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
