%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Nehemias
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 1006
%%% o-0904
\startChapter[
  title={Chapter 13}
  ]

\Summary{The
\MNote{The third part.

Correction of faultes.}
law is read, 3.~ſtrange wemen are diſmiſſed. 5.~Faultes in diſtribution
of the treaſure are amended, 10.~and due portions geuen to the
Leuites. 15.~Breakers of the Sabbath are corrected, 23.~and thoſe which
maried wemen of ſtrange nations.}

And in that day there was read in the volume of Moyſes the people
hearing it: and there was found written in it,
\CNote{\XRef{Deut.~23.}}
that the Ammonites and the Moabites ſhould not enter into the Church of
God for euer: \V for that they mette not the children of Iſrael with
bread and water:
\CNote{\XRef{Num.~22.}}
and they hyred againſt them Balaam, to curſe them: and our God turned
the curſing into bleſſing. \V And it came to paſſe, when they had heard
the law, they ſeperated euerie ſtranger from Iſrael. \V And ouer this
thing was Eliaſib the Prieſt, who had bene made ouerſeer in the
treaſurie of the houſe of our God, and neere akinne to
\SNote{This Tobias was an Ammonite & a perſecuter
\XRef{c.~14.}
to whom Eliaſib being akinne (by reaſon of vnlawful mariages) ioyned
felowſhipe with him for wicked lucre which therfore Nehemias corrected,
prefiguring therin our Sauiours zele, who threw byers and ſellers out of
the temple.
\XRef{Mat.~21.}
And theſe perſecuters prefigured heretikes in their wordes and actes, as
venerable Beda expoundeth.
\Cite{li.~3. in Eſd. c.~19.}}
Tobias. \V He therfore made to himſelfe a great treaſurie, and they were
there before him laying vp giftes, and frankincenſe, and veſſeles, and
the tythe of corne, of wine, and of oyle, the portions of the Leuites,
and of the ſinging men, and of the porters, and the firſt fruites of the
Prieſtes. \V But in al theſe things I was not in Ieruſalem, becauſe in
the two and thirteth yeare of Artaxerxes the king of Babylon I came to
the king, and in the end of certaine dayes I deſired the king. \V And I
came to Ieruſalem, to make him a treaſure in the entrances of the houſe
of God. \V And it ſemed to me exceding euil. And I threw forth the
veſſels of the houſe of Tobias out of the treaſurie: \V and I commanded
and they clenſed the treaſurie: and I brought thither againe the veſſels
of the houſe of God, the ſacrifice, and the frankincenſe. \V And I knew
that the portion of the Leuites had not bene geuen: and that euerie man
was fled into his countrie of the Leuites, and the ſinging men, and of
them that miniſtred: \V and I pleaded the matter againſt the
magiſtrates,
%%% 1007
and ſaid: Why haue we forſaken the houſe of God? And I aſſembled them,
and I made them to ſtand in their ſtandings. \V And al Iuda caried the
tythe of the corne, wine, and oyle into the ſtore houſes. \V And we
appoynted ouer the ſtorehouſes Selemias Prieſt, and Sadoc ſcribe, and
Phadaia of the Leuites, and next to them, Hanan the ſonne of Zachur, the
ſonne of Mathania: becauſe they were approued faithful, and to them were
committed the portions of the bretheren. \V Remember me my God for this
thing, and wipe not out my mercies, which I haue done in the houſe of my
God, and in his ceremonies. \V In thoſe daies I ſaw them in Iuda
treading the preſſes on
%%% o-0905
the Sabbath, carying heapes, and loding vpon aſſes wine, and grapes, and
figges, and al maner of burthen, and bringing it into Ieruſalem on the
Sabbath day. And I charged them, that they ſhould ſel on a day that it
was lawful to ſel in. \V And the Tyrians dwelt in it, bringing fiſhes,
and al thinges to ſel: and they ſold on the Sabbathes to the children of
Iuda in Ieruſalem. \V And I rebuked the Princes of Iuda, and ſaid to
them: What is this euil thing, that you doe, and prophane the day of the
Sabbath? \V Why did not our fathers theſe thinges, and our God brought
vpon vs al this euil, and vpon this citie? And you adde wrath vpon
Iſrael in violating the Sabbath. \V And it came to paſſe, when the gates
of Ieruſalem had reſted on the Sabbath day, I ſpake: and they ſhut the
gates, and I commanded them that they ſhould not open them til after the
Sabbath: and of my ſeruantes I appoynted ouer the gates, that none
ſhould bring in burdens in the Sabbath day. \V And the merchantes, and
they that ſold al merchandiſe, taried without Ieruſalem once and
againe. \V And I charged them, and I ſaid to them: Why tarie you ouer
againſt the wal? if you doe ſo the ſecond time, I wil lay my hand vpon
you. Therfore from that time they came not on the Sabbath. \V I ſpake
alſo to the Leuites that they ſhould be cleanſed, and ſhould come to
keepe the gates, and to ſanctifie the day of the Sabbath: therfore for
this alſo remember me my God, & ſpare me according to the multitude of
thy mercies. \V But in thoſe daies alſo I ſaw the Iewes marying wiues
wemen of Azotus, and of Ammon, and of Moab. \V And their children ſpake,
the halfe part the Azotian tongue, and they could not ſpeake the Iewes
language, and they ſpake according to the language of the people
%%% 1008
and people. \V And I rebuked them, and curſed them. And I beate of them
ſome men, and shaued them bald, and adiured them by God, that they
should not geue their daughters to their ſonnes, nor take their
daughters for their ſonnes, & for themſelues, ſaying: \V Did not Salomon
the king of Iſrael ſinne in this kind of thing? and ſurely in manie
nations, there was not a king like to him, & he was beloued of his God,
and God ſette him king ouer al Iſrael: him therfore alſo foren wemen
brought to ſinne. \V And ſhal we alſo being diſobedient perſons doe al
this great euil, to tranſgreſſe againſt our God, and to marie foren
wemen? \V And Sanaballat the Horonite was ſonne in law to one of the
ſonnes of Ioiada, the ſonne of Eliaſib the high Prieſt, whom I draue
from me. \V Remember ô Lord my God againſt them, that pollute the
Prieſthood, and the right of Prieſtes and Leuites. \V Therfore I
ſeparated from them al ſtrangers, and I appoynted the courſes of the
Prieſtes and Leuites, euerie man in his miniſterie: \V and in the
oblation of woode at times appoynted, and in the firſt fruites:
\SNote{A iuſt man that hath merited by good workes may pray with great
confidence for reward.}
Remember me my God vnto good. Amen.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
