%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Nehemias
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 1003
%%% o-0902
\startChapter[
  title={Chapter 12}
  ]

\Summary{The names and offices of Prieſtes, and
\Fix{Ieuites,}{Leuites,}{obvious typo, fixed in other}
which came with Zorobabel and Ioſue to Ieruſalem 27.~with great
ſolemnitie of thankſgeuing. 31.~Watchmen are oppoynted on the new
walles, 45.~and Kepers of the holy treaſure.}

And theſe are the Prieſtes and Leuites, that came vp with Zorobabel the
ſonne of Salathiel, and Ioſue: Saraia, Ieremias,
\SNote{Eſdras went againe to Babylon, and obtayning a fauorable
commiſſion of the King, brought manie with him into Ieruſalem.
\XRef{1.~Eſd.~7.}}
Eſdras, \V Amaria, Melluch, Hattus, \V Sebenias,
%%% 1004
Rheum, Merimuth, \V Addo Genthon, Abia, \V Miamin, Madia, Belga, \V
Semeia, and Ioiarib, Idaia, Sellum, Amoc, Helcias, \V Idaia. Theſe are
the Princes of the Prieſtes, and their bretheren in the daies of
Ioſue. \V Moreouer the Leuites, Ieſua, Bennui, Cedmihel, Sarebia, Iuda,
Mathanias, ouer the hymnes they & their bretheren. \V And Becbecia, and
Hannia and their bretheren euerie one in his office. \V And
\SNote{The genealogie of high Prieſts from Ioſue to Ieddoa otherwiſe
called Iaddus.}
Ioſue begatte Ioacim, and Ioacim begate Eliaſib, and Eliaſib begate
Ioiada, \V and Ioiada begate Ionathan, and Ionathan begate Ieddoa. \V
And in the daies of Ioacim the Prieſtes and Princes of the families
were: Of Saraia, Maraia: Of Ieremias, Hanania: \V Of Eſdras, Moſellam:
and of Amaria, Iohanan: \V Of Milicho, Ionathan: Of Sebenia, Ioſeph: \V
Of Haram, Edna: Of Maraioth, Helci: \V Of Adaia, Zacharia: Of Genthon,
Moſollam: \V Of Abia, Zechri: Of Miamin and Moadia, Phelti: \V Of Belga,
Sammua: Of Semaia, Ionathan: \V Of Ioiarib, Mathanai: Of Iodaia,
Azzi: \V Of Sellai, Sellai: Of Amoc, Heber: \V Of Helcias, Haſebia: Of
Idaia, Nathanael. \V The Leuites in the daies of Eliaſib, and Ioaiada,
and Iohanan, and Ieddoa, written Princes of the families, and the
Prieſtes in the reigne of Darius the Perſian. \V The children of Leui,
Princes of the families, written in the booke of Cronicles of daies, and
vnto the daies of Ionathan the ſonne of
\Fix{Eilaſib.}{Eliaſib.}{obvious typo, fixed in other}
\V And the Princes of the Leuites, Haſebia, Serebia, and Ioſue the ſonne
of Cedmihel: & their bretheren by their courſes, to praiſe and confeſſe
according to the precept of Dauid the man of God, and to waite equally
in order. \V Mathania, and Becbecia, Obedia, and Moſollam, Telmon,
Accub, keepers of the gates and of the entrances before the gates. \V
Theſe were in the daies of Ioacim the ſonne of Ioſue, the ſonne Ioſedec,
and in the daies of Nehemias the duke, and of Eſdras the Prieſt, and
Scribe. \V And in the dedication of the wal of Ieruſalem they ſought
Leuites out of al other places, to bring them into Ieruſalem, and to
make the dedication and ioy in geuing of thankes, and ſongue, and in
cimbales, pſalteries
%%% o-0903
and harpes. \V And the children of the ſinging men were gathered
together out of the champaine about Ieruſalem, and out of the townes
Nethuphati, \V and from the houſe of Galgal, and from the countries of
Geba and Azmaueth: becauſe the ſinging men did build them ſelues
villages round about Ieruſalem. \V And the Prieſtes and Leuites were
cleanſed, and
%%% 1005
they cleanſed the people, and the gates, and the wall. \V And I made the
Princes of Iuda, goe vp vpon the wal, and I ſette two great quyers of them that
ſhould praiſe. And they went on the right hand vpon the wal to the gate
of the dunghil. \V And after them went Oſaias, and the halfe part of the
princes of Iuda, \V and Azarias, Eſdras, and Moſollam, Iudas, and
\Fix{Beiamin,}{Beniamin,}{obvious typo, fixed in other}
and Semeia, and Ieremias. \V And of the childeren of the Prieſtes with
trumpetes, Zacharias the ſonne of Ionathan, the ſonne of Semeia, the
ſonne of Nathania, the ſonne of Michaia, the ſonne Zechur, the ſonne
Aſaph, \V and his bretheren Semeia, and Azareel, Malalai, Galalai, Maai,
Nathanael, and Iudas, and Hanani, with the inſtrumentes of the ſonge of
Dauid the man of God: and Eſdras the ſcribe before them in the gate of
the fountayne. \V And againſt them there went vp in the ſtayers of the
citie of Dauid, in the riſing of the wal aboue the houſe of Dauid, and
vnto the gate of waters toward the Eaſt. \V And the ſecond quyer of
thankes geuers went on the contrarie ſide, and I after it, and the halfe
part of the people vpon the wal, and aboue the tower of the fornaces, &
vnto the brodeſt wal, \V and aboue the gate of Ephraim, and aboue the
old gate, and aboue the gate of fiſhes and the towre of
Hananeel, and the towre of Emath, and vnto the gate of the flocke: and
they ſtood in the gate of the watch, \V and there ſtood two quyers of
them that prayſed in the houſe of God, and I and the halfe part of the
magiſtrates with me. \V And the Prieſtes, Eliachim, Maaſia, Miamin,
Michea, Elioenai, Zacharia, Hanania with trumpettes, \V & Maaſia, and Semeia,
and Eleazar, and Azzi, and Iohanan, and Melchia, and
Aelam, and Ezer. And the ſinging men ſang alowd, and Iezraia the
ouerſeer: \V and they immolated in that day great victimes, and
reioyced: for God had made them ioyful with great ioy: yea their wiues
alſo and childeren reioyced, and the ioy of Ieruſalem was heard far
of. \V They numbered alſo in that day, men ouer the ſtorehouſes of the
treaſure, for the libamentes, and for the firſt fruites, and for the
tythes, that the princes of the citie might bring in by them in the
honour of thankeſgiuing, Prieſtes and Leuites: becauſe Iuda was made
ioyful, in the Prieſtes & Leuites ſtanding by. \V And they kept the
watch of their God, and the obſeruance of expiation, and the ſinging
men, and the porters, according to the precept of Dauid, and of Salomon
his ſonne, \V becauſe in
%%% 1006
the daies of Dauid, and Aſaph from the beginning there were princes
appoynted of the ſinging men in ſong prayſing, and confeſſing to God. \V
And al Iſrael, in the daies of Zorobabel, and in the daies of Nehemias,
gaue portions to the ſinging men, and to the porters day by day, and
they
\SNote{As others gaue tithes to the Leuites: ſo they gaue to Prieſtes.
\XRef{Num.~18. v.~21.}
\XRef{28.}}
ſanctified the Leuites, and the Leuites ſanctified the children of
Aaron.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
