%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% One Peter
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 2894
%%% o-2755
\startChapter[
  title={Chapter 01}
  ]

\Summary{He comforteth them in their perſecutions (being now by Baptiſme
  made the children of God) with the hope of their heauenly inheritance:
  6.~shewing how meritorious it is for them to be ſo conſtant in
  faith, 10.~and confirming them therein with the authoritie of the
  Prophets and of the Holy Ghoſt. 15.~Exhorting thẽ to liue alſo
  accordingly in al holines, 16.~conſidering the holines of God, the
  vprightnes of his iudgement, the price of their redemption by Chriſt,
  22.~and the vertue of the ſeed in them (which is grace regeneratiue in
  Baptiſme) fore-told by the Prophets alſo.}

Peter an Apoſtle of \Sc{Iesvs} Chriſt, to the elect ſtrangers of the
diſperſion of Pontus, Galatia, Cappadocia, Aſia, and Bithynia, \V
according to the preſcience of God the Father, into ſanctification of
the Spirit, vnto the obedience and ſprinkling of the bloud of
\Sc{Iesvs} Chriſt: Grace to you and peace be multiplied.

\V
\CNote{\XRef{2.~Cor.~1,~3.}
\XRef{Eph.~1,~3.}}
Bleſſed be God and the Father of our Lord \Sc{Iesvs} Chriſt, who
according to his great mercie hath regenerated vs vnto a liuely hope, by
the reſurrection of \Sc{Iesvs} Chriſt from the dead, \V vnto an
inheritance incorruptible, and incontaminate, and that can not fade,
conſerued in the heauens in you, \V (who in the vertue of God are kept
by faith vnto ſaluation) ready to be reuealed in the laſt time. \V
Wherin you ſhal reioyce, a litle now if you muſt be made heauy in diuers
tentations: \V that the probation of your faith much more pretious then
gold (which is proued by the fire) may be found vnto praiſe and glorie
and honour in the reuelation of \Sc{Iesvs} Chriſt: \V whom hauing not
ſeen, you loue: in whom now alſo not ſeeing you beleeue: and beleeuing
you reioyce with ioy vnſpeakable and glorified, \V receiuing the end of
your faith, the ſaluation of your
%%% o-2756
ſoules.

\V Of the which ſaluation the Prophets inquired & ſearched, which
prophecied of the grace to come in you, \V ſearching vnto which or
%%% 2895
what manner of time the Spirit of Chriſt in them did ſignifie:
fore-telling thoſe paſſions that are in Chriſt, and the glories
following: \V to whom it was reuealed, that not to themſelues, but to
you they miniſtred thoſe things which now are told you by them that haue
euãgelized to you, the Holy Ghoſt being ſent from Heauen, on whom the
Angels deſire to looke.

\V For the which cauſe hauing the loynes of your
\SNote{Chaſtitie not only of body but alſo of mind, is required.
\Cite{S.~Bede vpon this place.}}
mind girded, ſober, truſt perfectly in that grace which is offered you,
in the reuelation of \Sc{Iesvs} Chriſt, \V as children of obedience,
not configurated to the former deſires of your ignorance: \V but
according to him that hath called you, the Holy one, be you alſo in al
conuerſation holy: \V becauſe it is written:
\CNote{\XRef{Leu.~11,~19.}
\XRef{20,~7.}}
\Emph{You shal be holy, becauſe I am holy.} \V And if you inuocate the
Father, him which
\CNote{\XRef{Deu.~10.}
\XRef{Ro.~2.}
\XRef{Gal.~2.}}
without acception of perſons iudgeth according to euery ones
\SNote{God wil iudge men according to euery ones workes, & not by faith
only.}
worke: in feare conuerſe ye the time of your ſeiourning. \V Knowing that
\CNote{\XRef{1.~Cor.~6,~20.}
\XRef{7,~23.}}
not with corruptible things, gold or ſiluer, you are redeemed from your
vaine conuerſation of your Fathers
\SNote{He meaneth the errours of Gentility. Or if he write to the Iewes
diſperſed, he meaneth the yoke of the Law with the fond and heauy
additions of their later Maiſters called \Emph{Deuteroſes}. The
Heretikes, to make it ſound to the ſimple againſt the traditions of the
Church, corrupt the text thus, \Emph{which you haue receiued by
tradition of the Fathers.}}
tradition: \V but with the pretious bloud as it were of an immaculate
and vnſpotted Lamb, Chriſt, \V
\CNote{\XRef{Ro.~16.~25.}
\XRef{Col.~1,~26.}
\XRef{Tit.~1,~2.}}
fore-knowen indeed before the conſtitution of the world, but manifeſted
in the laſt times for you, \V which by him are faithful in God who
raiſed him frõ the dead, & hath giuẽ him glorie, that your faith and
hope might be in God. \V Making your ſoules chaſt in obedience of
charitie, in the ſincere loue of the Fraternitie from the hart loue ye
one another earneſtly: \V borne againe not of corruptible ſeed, but
incorruptible by the word of God who liueth & remaineth for euer. \V For
\CNote{\XRef{Eſa.~40,~6.}}
\Emph{al flesh is as graſſe: & al the glorie therof as the floure of
graſſe. The graſſe is withered, and the floure therof is fallen away.}
\V But the word of our Lord remaineth for euer, and this is the word
that is euangelized among you.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

