%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 0816
%%% o-0732
\startChapter[
  title={Chapter 17}
  ]

\Summary{Salmanazar king of Aſsyrians maketh Oſee king of Iſrael
  tributarie, and perceuing his endeuour to be deliuered therof,
  impriſoneth him; after three yeares ſiege taketh Samaria, and carieth
  the people captiue into Aſsyria. 7.~Al which God permitteth for diuers
  great ſinnes here recited. 25.~The new inhabitantes of the countrie
  not knowing God, are deuoured by lions. VVherupon a true prieſt is
  ſent to inſtruct them. 29.~But they lerning the rites of true religion
  do
  \Fix{mixt}{mixe}{obvious typo, fixed in other}
  the ſame with idolatrie.}

In the twelfth yeare of Achaz king of Iuda, reigned Oſee the ſonne of
Ela in Samaria ouer Iſrael nine yeares. \V And he did euil before our
Lord: but not as the kinges of Iſrael that had bene before him. \V
Againſt him came vp Salmanaſar king of the Aſſyrians, and Oſee was made
ſeruant to him, and payd him tributes. \V And when the king of the
Aſſyrians had found, that Oſee endeuoring to rebel had ſent meſſengers
to Sua the king of Ægypt, that he might not pay tributes to the king of
the Aſſyrians, as euery yeare he was accuſtomed, he beſieged him, and
caſt him
\Fix{blound}{bound}{obvious typo, fixed in other}
into priſon. \V And he ranged through al the land: and going vp to
Samaria, he beſieged it three yeares. \V And in the ninth yeare of Oſee,
the king of Aſſyrians tooke Samaria, and transferred Iſrael vnto the
Aſſyrians: and he put them in Hala, and in Habor beſide the riuer of
Gozan, in the cities of the Medes. \V For it came to paſſe, when the
children of Iſrael had ſinned to our Lord their God, which brought them
out of the land of Ægypt, and out of the hand of Pharao the king of
Ægypt, they worſhipped ſtrange goddes. \V And they walked according to
the rite of the Gentiles, which our Lord had conſumed in the ſight of
%%% 0817
the children of Iſrael, and of the kings of Iſrael: becauſe they had
done in like maner. \V And the children of Iſrael offended our Lord
their God with wordes not right: & built them excelſes in al their
cities from the Towre of watchmen vnto the fenſed citie. \V And they
made them ſtatues & groues on euerie high hil, and vnder euerie thicke
\Fix{woddie}{wooddie}{likely typo, fixed in other}
tree: \V and burnt there incenſe vpon the altars after the maner of the
Gentiles, which our Lord remoued from their face: and they did wicked
things, prouoking our Lord. \V And they worſhipped the filthes, wherof
our Lord commanded them, that
\Fix{thou}{they}{obvious typo, fixed in other}
ſhould not doe this thing. \V And our
Lord teſtified in Iſrael and in Iuda by the hand of al the Prophetes and
Seers, ſaying: Returne from your moſt wicked wayes, and keepe my
precepts, and ceremonies according to al the law, which I commanded your
fathers: and as I haue ſent to you in the hand of my ſeruantes the
Prophetes. \V Who heard not, but hardened
%%% o-0733
their necke according to the necke of their fathers, who would not obey
our Lord their God. \V And they caſt away his ordinances, and the
couenant that he made with their fathers, and the teſtifications,
wherwith he conteſted them: and they folowed vanities, and did vaynly:
and they folowed the Gentiles, that were round about them, concerning
which our Lord had commanded them, that they ſhould not doe as they
did. \V And they forſooke al the preceptes of our Lord their God: and
made to them ſelues two molten calues, and groues, and adored al the
hoſte of heauen: and they ſerued Baal, \V and conſecrated their ſonnes,
and their daughters through fyre: and they gaue themſelues to
deuinations, and ſoothſayings: and they deliuered vp themſelues to doe
euil before our Lord,
\SNote{That is, conſequently they did prouoke him. As
\XRef{3.~Reg.~14. v.~9.}}
that they might prouoke him. \V And our Lord was wrath with Iſrael
vehemently, and tooke them away from his ſight, and there remayned but
the tribe of Iuda onlie. \V But neither Iuda it ſelf kept the
commandementes of our Lord their God: but walked in the errours of
Iſrael, which it had wrought. \V And our Lord reiected al the ſeede of
Iſrael, and afflicted them, & deliuered them into the hand of the
ſpoylers, til he threwe them away from his face: \V euen now from that
time, when Iſrael was rent from the houſe of Dauid, and made Ieroboam
the ſonne of Nabat their king: for Ieroboam ſeperated Iſrael from our
Lord, and made them ſinne a great ſinne. \V And the children of Iſrael
walked in
%%% 0818
al the ſinnes of Ieroboam, which he had done: and they departed not from
them, \V vntil our Lord tooke away Iſrael from his face, as he had
ſpoken in the hand of al his ſeruantes the Prophetes: and Iſrael was
tranſported out of their land vnto the Aſſyrians, vntil this day. \V And
the king of the Aſſyrians brought from Babylon, and from Cutha, and from
Auah, and from Emath, and from Sepharuaim: and placed them in the cities
of Samaria for the children of Iſrael: who poſſeſſed Samaria, and dwelt
in the cities therof. \V And when they began to dwel there, they feared
not our Lord: and our Lord ſent lions vpon them, which killed them. \V
And it was told the king of the Aſſyrians, and ſayd: The nations, which
thou haſt transferred, and made to dwel in the cities of Samaria, know
not the ordinances of the God of the land: and the Lord hath ſent lions
vpon them: and behold they kil them, for that they know not the rite of
the God of the land. \V And the king of the Aſſyrians commanded, ſaying:
Bring thither one of the prieſtes, which  you brought thence captiue,
and let him goe, and dwel with them: and let him teach them the
ordinances of the God of the land. \V Therfore when one of thoſe
prieſts, which were led captiue from Samaria, was come, he dwelt in
Bethel, and taught them how they ſhould worship our Lord. \V And euerie
Nation framed their owne god, and put them in the highe temples, which
the Samaritanes had made, Nation and Nation in their cities, where they
dwelt. \V For the men of Babylon made Socothbenoth: and the Cutheites
made Nergel: and the men
%%% o-0734
of Emath made Aſima. \V Moreouer the Heueites made Nebahaz &
Tharthac. And they that were of Sepharuaim burnt their children in fyre,
to Adramelech, and Anamelech the goddes of Sepharuaim. \V And
neuertheles they
\SNote{Not truly worſhipped but made ſhew to worſhippe. For true worſhip
of God admitteth not worſhip of anie falſe god.
\XRef{v.~34.}}
worſhipped our Lord. And they made to themſelues of the vileſt perſones
prieſtes of the excelſes, and they placed them in the highe temples. \V
And when they worſhipped our Lord, they ſerued alſo their owne goddes
according to the cuſtome of the Nations out of the which they were
tranſported to Samaria: \V vntil this preſent day they folow the old
maner: they feare not our Lord, neither keepe they his ceremonies, and
iudgements, and law, and the
\Fix{commandemet,}{commandment,}{obvious typo, fixed in other}
which our Lord commanded the children of Iacob, whom he ſurnamed
Iſrael: \V and he had made a couenant with them, & had commanded
%%% 0819
them, ſaying: Feare not ſtrange goddes, and adore them not, neither
worſhip them, and immolate not to them. \V But the Lord your God, which
brought you out of the Land of Ægypt in great ſtrength, and a ſtretched
out arme, him feare ye, and him adore, and to him doe ye immolate. \V
The ceremonies alſo, and iudgementes, and law, and the commandment, that
he wrote you, kepe ye, that you may doe them alwaies: and feare not
ſtrange goddes. \V And the couenant, that he made with you, forget not:
neither doe ye worſhip ſtrange goddes, \V but feare our Lord your God,
and he wil deliuer you out of the hand of al your enemies. \V But they
heard not, but did according to their old cuſtome. \V Theſe Nations
therfore were fearing of our Lord but neuertheleſſe ſeruing their idols
alſo: for both their children and nephewes, as their fathers did, ſoe
doe they vntil this preſent day.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
