%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 0796
%%% o-0715
\startChapter[
  title={Chapter 8}
  ]

\Summary{After ſeuen yeares famine fortold by Eliſeus, the Sunamite
  woman returning home, recouereth her landes and reuenewes. 7.~Eliſeus
  forsheweth the death of Benadad king of Syria, and cruel reigne of
  Hazael. 16.~Ioram reigning in Iuda, the Idumeans reuolt from him. Alſo
  Lobna. 23.~He dieth, and his ſonne Ochozias ſucceedeth.}

%%% 0797
And Eliſeus ſpake to the woman, whoſe ſonne he reſtored to life, ſaying:
Ariſe, goe thou and thy houſe, and ſoiourne wherſoeuer thou ſhalt finde:
for our Lord hath called a famine, and it ſhal come vpon the land ſeuen
yeares. \V Who aroſe, & did according to the word of the man of God: &
going with her houſhold, ſhe ſoiourned in the land of the Philiſtijms
many dayes. \V And when the ſeuen yeares were ended, the woman returned
out of the Land of the Philiſthijms: and ſhe went forth to ſpeake to the
king for her houſe, and for her landes. \V And the king ſpake with Giezi
the ſeruant of the man of God, ſaying: Tel me al the meruelous thinges
that Eliſeus hath done. \V And when he had told the king how he had
rayſed a dead man, the woman appeared, whoſe ſonne he had reuiued,
crying to the king for her houſe, and her landes. And Giezi ſayd: My
lord king, this is the woman, and this is her ſonne, whom Eliſeus
rayſed. \V And the king asked the woman: who told him. And the king gaue
her an eunuch, ſaying: Reſtore her al thinges that are hers, and al the
reuenewes of the landes, from the day, that ſhe leaft the land vntil
this preſent. \V Eliſeus alſo came to Damaſcus, and Benadad the king of
Syria was ſicke: and they told him, ſaying: The man of God cometh
hither. \V And the king ſayd to Hazael: Take with thee preſentes, and
goe to meete the man of God, and conſult the Lord by him, ſaying: Can I
eſcape of this myne infirmitie? \V Hazael therfore went to meete him,
hauing with him preſentes, and al good thinges of Damaſcus, the lodes of
fourtie camels. And when he ſtood before him, he ſayd: Thy ſonne Benadad
the king of Syria hath ſent me to thee, ſaying: Can I recouer of this
mine infirmitie? \V And Eliſeus ſayd to him: Goe, tel him:
\SNote{This was true in ſome ſenſe, ſicknes ending when death came.}
Thou ſhalt be healed: but our Lord hath ſhewed me that dying he shal
die. \V And he ſtood with him, and was trubled ſo far that he blushed:
and the man of God wept. \V To whom Hazael ſayd: Why doeth my lord
weepe? But he ſayd: Becauſe I know what euils thou wilt doe to the
children of Iſrael. Their fenſed cities thou wilt burne with fyre, and
their yongmen thou wilt kil with the ſword, and their litle ones thou
wilt dash in peeces, and wemen with childe thou wilt diuide. \V And
Hazael ſayd: What am I thy ſeruant a dog, that I should doe this
%%% o-0716
great thing? And Eliſeus ſayd: Our Lord hath shewed me that thou shalt
be king of Syria. \V Who when he was departed
%%% 0798
from Eliſeus, came to his maiſter, who ſayd to him: What ſayd Eliſeus to
thee? But he anſwered: He told me: Thou shalt recouer health. \V And
when the next day was come, he tooke a couerlette, and powred water
theron, and ſpred it vpon his face: who being dead, Hazael reigned for
him. \V In the fifth yeare of Ioram the ſonne of Achab the king of
Iſrael, and of Ioſaphat the king of Iuda, reigned Ioram the ſonne of
Ioſaphat the king of Iuda. \V He was two and thirtie yeares old when he
began to reigne, and he reigned eight yeares in Ieruſalem. \V And he
walked in the waies of the kinges of Iſrael, as the houſe of Achab had
walked: for
\SNote{Athalia
\XRef{v.~26.}
is called the daughter of Amri. VVherfore it ſemeth that either ſhe was
the adopted daughter of her brother Achab, or is there called the
daughter of her grandfather.}
the daughter of Achab was his wife: and he did that which is euil in the
ſight of our Lord. \V But our Lord would not deſtroy Iuda, for Dauid his
ſeruant, as he had promiſed him, to geue him a lampe to him, and to his
children alwaies. \V In his daies reuolted Edom, from being vnder Iuda,
and made to it ſelf a king. \V And Ioram came to Seira, and al the
chariotes with him: and he aroſe in the night, and ſtroke the Idumeans,
that had beſet him, and the captaynes of the chariotes, and the people
fled into their tabernacles. \V Edom therfore reuolted from being vnder
Iuda, vntil this day. Then Lobna alſo reuolted at that time. \V But the
reſt of the wordes of Ioram, and al that he did, are not theſe thinges
written in the
%%% !!! XRef ?
Booke of the wordes of the kinges of Iuda? \V And Ioram ſlept with his
fathers, and was buried with them in the Citie of Dauid, and Ochozias
his ſonne reigned for him. \V In the twelfth yeare of Ioram the ſonne of
Achab the king of Iſrael, reigned Ochozias the ſonne of Ioram the king
of Iuda. \V Two and twentie yeares old was Ochozias when he began to
reigne, and he reigned one yeare in Ieruſalem: the name of his mother
was Athalia the daughter of Amri the king of Iſrael. \V And he walked in
the waies of the houſe of Achab: and he did that which is euil before
our Lord, as the houſe of Achab: for he was the ſonne in lawe of the
houſe of Achab. \V He went alſo with Ioram the ſonne of Achab, to fight
agaynſt Hazael the king of Syria in Ramoth Galaad, and the Syrians
wounded Ioram. \V Who returned to be cured, in Iezrahel: becauſe the
Syrians wounded him in Ramoth fighting agaynſt Hazael the king of
Syria. Moreouer Ochozias the ſonne of Ioram the king of Iuda, went downe
to viſite Ioram the ſonne of Achab into Iezrahel, becauſe he was ſicke
there.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
