%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 24
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-24


\project douay-rheims


%%% 0833
%%% o-0747
\startChapter[
  title={Chapter 24}
  ]

\Summary{Ioakim ſerueth the king of Babylon three yeares. 2.~Manie
  rouers infeſt his countrie. 5.~He dieth, and his ſonne Ioachin
  reigneth. 10.~The king of Babylon carieth king Ioachin, and al the
  chief perſones, and treaſures into Babylon, 17.~appointing Matthanias,
  whom he nameth Sedecias, king of Iuda. 20.~VVho reuolteth from the
  king of Babylon.}

In his daies came vp Nabuchodonoſor the king of Babylon, and Ioakim was
made his ſeruant three yeares: and he rebelled agaynſt him againe. \V
And our Lord ſent in vpon him the rouers of the Chaldees, and the rouers
of Syria, and the rouers of Moab, & the rouers of the children of Ammon:
and he ſent them into Iuda, to deſtroy it, according to the word of our
Lord, which he had ſpoken by his ſeruantes the prophetes. \V And this by
the word of our Lord was done againſt Iuda, to take it away before
him for al the ſinnes of Manaſſes which he did, \V and for the innocent
bloud, that he had ſhed, & filled Ieruſalem with the bloud of
innocentes: & for this thing God would not be made propitious. \V But
the reſt of the wordes of Ioakim, and al that he did, are not theſe
thinges writen in the
%%% !!! XRef ?
Booke of the wordes of the daies of
%%% 0834
the kinges of Iuda? And Ioakim
\SNote{Not dying in peace, for he was ſlaine by Nabuchodonoſor.
\Cite{Ioſephus li.~10. c.~8. Ant.}
And his bodie was caſt out of the citie, according as Ieremie prophecied
\XRef{c.~22.}
\Emph{vvith the burial of an aſse shal he be buried. &c.}}
ſlept with his fathers: \V and Ioachin his ſonne reigned for him. \V And
the king of Ægypt added no more to come out of his countrie: for the
king of Babylon had taken al that had beene the kinges of Ægypt, from
the riuer of Ægypt, vnto the riuer Euphrates. \V Eightene yeares old was
Ioachin when he began to reigne, and he reigned three monethes in
Ieruſalem: the name of his mother was Noheſta the daughter of Elnathan
of Ieruſalem. \V And he did euil before our Lord, according to al
thinges which his father had done. \V At that time came vp the ſeruantes
of Nabuchodonoſor the king of Babylon into Ieruſalem, and the citie was
compaſſed with fortes. \V And Nabuchodonoſor the king of Babylon came to
the citie with his ſeruantes to aſſault it. \V And Ioachin the king of
Iuda went forth to the king of Babylon, he and his mother, and his
ſeruantes, and his nobles, and his eunuches: and the king of Babylon
receiued him the eight yeare of his reigne. \V And he brought forth from
thence al the treaſures of the houſe of our Lord, and the treaſures of
the kinges houſe: and he cut in peces al the golden veſſel, which
Salomon the king of Iſrael had made in the temple of our Lord, according
to the word of our Lord. \V And he tranſported al Ieruſalem, and al the
princes, and al the ſtrong men of the armie, ten thouſand into
captiuitie:
%%% o-0748
and euerie artificer and incloſer: and nothing was left, ſauing the poore
ſorte of the people of the land. \V He tranſported alſo Ioachin into
Babylon, and the kinges mother, and the kinges wiues, and his eunuches:
and the iudges of the land he led into captiuitie from Ieruſalem into
Babylon. \V And al the ſtrong men, ſeuen thouſand, and the artificers,
and incloſers a thouſand, al valiant men and
\Fix{warries:}{warriers:}{obvious typo, same in both}
and the king of Babylon led them captiues into Babylon. \V And he
appointed Matthanias his vncle for him: and called his name Sedecias. \V
One and twentie yeares old was Sedecias when he began to reigne, and he
reigned eleuen yeares in Ieruſalem: the name of his mother was Amital,
the daughter of Ieremie of Lobna. \V And he did euil before our Lord,
according to al thinges which Ioakim had done. \V For our Lord was wrath
againſt Ieruſalem and againſt Iuda, til he caſt them away from his face:
and Sedecias
\SNote{In this he greuoſly offended hauing ſworne to ſerue him. And
therfore Ezechiel
\XRef{c.~17.}
fortelleth the miſerie that wil fal vpon him,
\Emph{shal he that broke couenant eſcape? &c.}}
reuolted from the king of Babylon.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
