%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 0826
%%% o-0741
\startChapter[
  title={Chapter 21}
  ]

\Summary{For the enormious impietie of Manaſſes, 10.~God threatneth
  deſtruction of the kingdom. 16.~He ſpilleth innocent bloud, 18.~dieth,
  and his ſonne Amon reigneth alſo wickedly: 23.~is ſlaine by his
  ſeruantes, and his ſonne Ioſias reigneth.}

Twelue yeares old was Manaſſes, when he began to reigne, & he reigned
fiue and fiftie yeares in Ieruſalem: the name of his mother was
Haphſiba. \V And he did euil in the ſight of our Lord, according to the
idols of the Nations, which our Lord deſtroyed from the face of the
children of Iſrael. \V And he was turned, and built the excelſes, which
Ezechias his father had deſtroyed: and he ſette vp altars to Baal, and
made groues, as Achab the king of Iſrael had done: and he adored al the
hoſt of heauen, and worshipped it.
%%% 0827
\V And he built altars in the houſe of our Lord, of the which our Lord ſayd: In
Ieruſalem I wil put my name. \V And he built altars to al the hoſt of
heauen in the two courtes of the temple of our Lord. \V And he made his
ſonne paſſe through fyre: and he vſed ſoothſaying, and obſerued
diuinations, and made pithones, and multiplied inchanters, to doe euil
before our Lord, and to prouoke him. \V He ſette alſo the idol of the
groue, which he had made in the temple of our Lord: concerning the which
our Lord ſpake to Dauid, and to Salomon his ſonne: In this temple, and
in Ieruſalem, which I haue choſen out of al the tribes of Iſrael, I wil
put my name for euer. \V And I wil noe more make the foote of Iſrael to
be moued out of the land, which I gaue to their fathers: yet ſo if in
worke they ſhal keepe al things, that I haue commanded them, al the law
which my
\Fix{ſeruantes}{ſeruant}{obvious typo, fixed in other}
Moyſes commanded them. \V But they heard not: but were ſeduced by
Manaſſes, to doe euil
\SNote{The Iewes ſinned more greuouſly reuolting from the Law of God and
contemning the admonitions of holie prophetes, then the nations that had
neither law nor prophetes to inſtruct them.}
aboue the Nations, which our Lord deſtroyed before the face of the
children of Iſrael. \V And our Lord ſpake in the hand of his ſeruantes,
the prophetes, ſaying: \V Becauſe Manaſſes the king of Iuda hath done
theſe moſt wicked abominations, paſſing al thinges that the Amorrheites
did before him, and hath made Iuda alſo to ſinne in his filthes: \V
therfore thus ſayth our Lord the God of Iſrael: Behold I wil bring in
euils vpon Ieruſalem and Iuda: that whoſoeuer shal heare it, both his
eares shal tingle. \V And I wil ſtretch out vpon Ieruſalem the corde of
Samaria, and the weight of the houſe of Achab: and I wil wipe out
Ieruſalem, as
\Fix{tables}{tablets}{not really a typo, same in both}
are wont to be wiped out, and wyping out I wil turne it, and draw often
the pencil vpon the face therof. \V
\SNote{God ſtil preſerued ſome in true religion, though they alſo
ſuffered tribulations with the wicked for the general ſinnes of the king
and people.
\XRef{Pſal.~88. v.~35.}
Yea this king Manaſſes in captiuitie became vertuous and recouered his
kingdom.
\XRef{2.~Paralip.~33.}}
But I wil leaue remnantes of mine inheritance, and wil deliuer them into
the handes of their enemies: and they ſhal be vnto waſte, and vnto
ſpoile to
%%% o-0742
al their aduerſaries: \V becauſe they haue done euil before me, and haue
continewed prouoking me, from the day that their fathers came out of
Ægypt, vntil this day. \V Moreouer Manaſſes shed alſo innocent bloud
exceding much til he filled Ieruſalem euen to the mouth: beſide his
ſinnes, wherin he made Iuda to ſinne, to doe euil before our Lord. \V
But the reſt of the wordes of Manaſſes, and al that he did, and his
ſinne, which he ſinned, are not theſe thinges written in the
%%% !!! XRef ?
Booke of the wordes of the dayes of the kinges of Iuda? \V And Manaſſes
ſlept with
%%% 0828
his fathers, and was buried in the garden of his houſe, in the garden of
Oza: and Amon his ſonne reigned for him. \V Two and twentie yeares old
was Amon when he began to reigne: two yeares alſo he reigned in
Ieruſalem: the name of his mother was Meſſalemeth the daughter of
Ieteba. \V And he did euil in the ſight of our Lord, as Manaſſes his
father had done. \V And he walked in al the way, by the which his father
had walked: and he ſerued the filthes, which his father had ſerued, and
he adored them, \V and forſooke our Lord the God of his fathers, and
walked not in the way of our Lord. \V And his ſeruantes lay in wayte
agaynſt him, and ſlewe the king in his houſe. \V But the people of the
land ſtroke al them, that had conſpired agaynſt king Amon: and made
Ioſias his ſonne their king for him. \V But the reſt of the wordes of
Amon which he did, are not theſe writen in the
%%% !!! XRef ?
Booke of the wordes of the dayes of the kinges of Iuda? \V And they
buried him in his ſepulchre, in the garden of Oza: and Ioſias his ſonne
reigned for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
