%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-23


\project douay-rheims


%%% 0830
%%% o-0744
\startChapter[
  title={Chapter 23}
  ]

\Summary{Ioſias reading the law before the people, they al promiſe to
  ſerue God. 4.~Al thinges belonging to Idolatrie are caſt forth of the
  temple, 8.~and other places of Iuda. 15.~He alſo deſtroyeth the altar
  in Bethel (not remouing the bones of a prophet) 19.~and other altares
  in Samaria: killeth the falſe prieſtes. 21.~And maketh a notorious
  great Paſch. 26.~Yet God threatneth the captiuitie of Iuda for their
  ſinnes. 29.~Ioſias is ſlaine in battle by the king of Ægypt: and his
  ſonne Ioachaz reigneth. VVho after three monethes is depoſed by the
  king of Ægypt, and his brother Eliacim, henceforth called Ioakim, made
  king.}

And they reported to the king that which ſhe had ſayd. Who ſent: and al
the ancientes of Iuda and Ieruſalem were aſſembled to him. \V And the
king went vp to the temple of our Lord, and al the men of Iuda, and al
that dwelt in Ieruſalem with him prieſtes and prophetes, and al the
people from litle to great: and in hearing of al he read al the wordes
of the Booke of the couenant, which was found in the houſe of
our Lord. \V And the king ſtood vpon the ſteppe: and made a couenant
with our Lord, to walke after our Lord, & kepe his preceptes, and
teſtimonies and ceremonies, with al their hart, and with al their ſoule,
and to performe the wordes of this couenant, which were writen in that
booke: & the people agreed to the couenant. \V And the king willed
Helcias the high prieſt, & the prieſtes of the ſecond order, & al the
porters, that they ſhould caſt out of the temple of our Lord al the
veſſels, that had bene made to Baal, and in the groue, and to al the
hoſt of heauen: & he burnt them without Ieruſalem in the Valley cedron,
and caried the duſt of them into Bethel. \V And he deſtroyed the
Southſayers, which the kinges of Iuda had appointed to ſacrifice in the
excelſes in the cities of Iuda, and round about Ieruſalem: and them that
burnt incenſe to Baal, and to the Sunne, and to the Moone, and to the
twelue ſignes, and to al the hoſt of heauen. \V And he cauſed the groue
to be caried forth out of the houſe of our Lord without Ieruſalem in the
Valley cedron, & he burnt it there, and brought it into duſt, and threw
it vpon the ſepulchres of the common people. \V He deſtroyed alſo the
litle houſes of the effeminates,
%%% 0831
which were in the houſe of our Lord, for the which the wemen woue as it
were litle houſes of the groue. \V And he gathered together al the
prieſtes of the cities of Iuda: and he contaminated the excelſes, where
the prieſtes did ſacrifice from Gabaa vnto Berſabee: and he deſtroyed the
altars of the gates in the entrance of the doore of Ioſue chief of the
citie, which was on the left hand of the gate
%%% o-0745
of the citie. \V Howbeit the prieſtes
\SNote{Becauſe they had offered ſacrifice to falſe goddes and in
vnlawful places they were ſuſpended from offering anie more ſacrifice at
al.}
of the excelſes went not vp to the altar of our Lord in Ieruſalem: but
only they did eate azimes in the middes of their brethren. \V He
contaminated alſo Topheth, which is in the Valley of the ſonne of Ennom:
that no man ſhould conſecrate his ſonne or daughter by fyre to
Moloch. \V He tooke away alſo the horſes, which the kinges of Iuda had
geuen to the Sunne, in the entrance of the temple of our Lord, beſide
the chamber of Nathanmelech the eunuch, who was in Pharurim: and the
chariotes of the Sunne he burnt with fire. \V The altars alſo that were
vpon the roofes of the vpper chamber of Achaz, which the kinges of Iuda
had made, and the altars which Manaſſes had made in the two cortes of
the temple of our Lord, the king
deſtroyed: and he ranne from thence, and ſprinkled the aſhes of them
into the Torrent cedron. \V The excelſes alſo that were in Ieruſalem on
the right ſide of the Mount of offence,
\CNote{\XRef{3.~Reg.~11.}}
which Salomon the king of Iſrael had built to Aſtaroth the idol of the
Sidonians, and to Chamos the ſcandal of Moab, and to Melchom the
abomination of the children of Ammon, the king deſtroyed. \V And he
brake in peces the ſtatues, and cut downe the groues: and he filled
their places with the bones of dead men. \V Moreouer the altar alſo,
that was in Bethel, and the excelſe, which Ieroboam the ſonne of Nabat
had made, who made Iſrael to ſinne: and that altar, and excelſe he
deſtroyed, and burnt, and brake into powder, and the groue alſo he
burnt. \V And Ioſias turning, ſaw there ſepulchres, that were in the
mount: and he ſent and tooke the bones out of the ſepulchres, and burnt
them vpon the altar, and polluted it according to the word of our Lord,
\CNote{\XRef{3.~Reg.~15.}}
which the man of God ſpake, who had foretold theſe thinges. \V And he
ſayd: What title is that, which I ſee? And the citizens of that citie
anſwered: It is the ſepulchre of the man of God, which came from Iuda,
and foretold theſe thinges which thou haſt done vpon the altar of
Bethel. \V And he ſayd:
%%% 0832
Let him alone, let no man moue his bones. And his bones remayned
vntouched with the bones of the prophet, that came out of Samaria. \V
Moreouer al the temples of the excelſes, which were in the cities of
Samaria, which the kinges of Iſrael had made to prouoke our Lord, Ioſias
tooke away: and he did to them according to al the workes, which he had
done in Bethel. \V And he ſlew al the prieſtes of the excelſes, that
were there vpon the altars: and he burnt mens bones vpon them: & turned
into Ieruſalem. \V And he commanded al the people, ſaying: Make a Phaſe
to our Lord your God, according as it is writen in the booke of this
couenant. \V For
\SNote{Ioſephus writeth that this godlie king gaue thirtie thouſand
lambes and kiddes to the poore people for their Paſch & three thouſand
oxen for Holocauſtes. The prieſtes alſo & Leuites added more of their
owne.
\Cite{li.~10. Antiq. c.~5.}}
there was not ſuch a Phaſe made from the daies of the Iudges, which
iudged Iſrael, and of al the daies of the kinges of Iſrael, and of the
kinges of Iuda, \V as in the eightenth yeare of king Ioſias this Phaſe
was made to our Lord in Ieruſalem. \V Yea and the Pythones, and
Southſayers, and the images of idols, and the filthes, and the
abominations, that had bene in the land of Iuda and Ieruſalem, Ioſias
tooke away: that he might eſtabliſh the
%%% o-0746
wordes of the law, that were writen in the Booke, which Helcias the
prieſt found in the temple of our Lord. \V There was no king before him
like to him, that returned to our Lord in al his hart, & in al his
ſoule, and in al his powre according to al the law of Moyſes: neither
after him did there ariſe the like to him. \V But yet our Lord was not
auerted from the wrath of his great furie, wherwith his furie was wrath
agaynſt Iuda: for the prouocations, wherwith
\SNote{Albeit Manaſſes repented, and was reſtored to God fauour, & to
his kingdome.
\XRef{2.~Paral.~3.}
Yet his ſinnes were temporally puniſhed, both in himſelf, and his
poſteritie.}
Manaſſes had prouoked him. \V Our Lord therfore ſayd Iuda alſo wil I
take away from my face, as I haue taken away Iſrael: and I wil reiect
this citie, which I choſe Ieruſalem, and the houſe, wherof I ſayd: My
name ſhal be there. \V But the reſt of the wordes of Ioſias, and al that
he did, are not theſe thinges writen in the
%%% !!! XRef ?
Booke of the wordes of the
daies of the kinges of Iuda? \V In his dayes came vp Pharao Nechao the
king of Ægypt, agaynſt the king of Aſſyrians to the riuer Euphrates: and
Ioſias the king went to meete him: and was ſlaine in Mageddo, when he
had ſeene him. \V And his ſeruantes caried him dead from Mageddo: & they
brought him into Ieruſalem, and buried him in his ſepulchre. And the
people of the land tooke Ioachaz the ſonne of Ioſias: and they anoynted
him, and made him king for his father. \V Three and twentie yeares old
was Ioachaz when he began to reigne, and
%%% 0833
he reigned three monethes in Ieruſalem: the name of his mother was
Amital, the daughter of Ieremie of Lobna. \V And he did euil before our
Lord, according to al thinges which his fathers had done. \V And Pharao
Nechao bound him in Rebla, which is in the land Emath, that he ſhould
not reigne in Ieruſalem: and he ſette a penaltie vpon the land, an
hundred talentes of ſiluer, and a talent of gold. \V And Pharao Nechao
made Eliacim king the ſonne of Ioſias, for Ioſias his father: and turned
his name Ioakim. Moreouer he tooke Ioachaz and brought him into Ægypt,
and he died there. \V And Ioakim gaue the ſiluer and the gold to Pharao,
when he had taxed the land vpon euerie man, that it might be payd
according to the precept of Pharao: and he exacted of euerie man
according to his abilitie, as wel ſiluer as gold of the people of the
land: to geue vnto Pharao Nechao. \V Fiue and twentie yeares old was
Ioakim, when he began to reigne: and he reigned eleuen yeares in
Ieruſalem: the name of his mother was Zebida the daughter of Phadaia of
Ruma. \V And he did euil before our Lord according to al thinges, which
his fathers had done.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
