%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 0806
%%% o-0723
\startChapter[
  title={Chapter 12}
  ]

\Summary{Ioas vvilleth the prieſtes to repaire vvhat needeth in the
  temple, receiuing al the money offered there. 6.~VVhich they not
  performing, the high prieſt prouideth that al is amended, 16.~the
  prieſtes receiuing only the money due to themſelues. 17.~Hazael king
  of Syria taking Geth, and threatning Ieruſalem is pacified vvith
  money. 19.~Ioas is traitarouſly ſlaine by his ſeruantes, and his ſonne
  Amaſias reigneth.}

In the ſeuenth yeare of Iehu reigned Ioas: and he reigned fourtie yeares
in Ieruſalem. The name of his mother was Sebia of Berſabee. \V And Ioas
did right before our Lord al the daies, that Ioiada the prieſt taught
him. \V But yet the excelſes he tooke not away: for the people immolated
& burnt incenſe in the excelſes. \V And Ioas ſayd to the prieſtes: Al
the money of the ſanctified thinges, which is brought into the temple of
our Lord by the paſſengers, which is offered for the
\SNote{That is, the ordinarie oblation for ech particular perſon.
\XRef{Exod.~30.}}
price of a ſoule, and which of their owne accord, and of their owne free
hart they bring into the temple of our Lord: \V let the prieſtes take it
according to their order, and maynteine reparations of the houſe, if
they shal ſee any thing that needeth reparation. \V Therfore vntil the
three and twenteth yeare of king Ioas, the prieſtes did not make
reparations of the temple. \V And king Ioas called Ioiada the high
prieſt and the prieſtes, ſaying to them: Why make you not the
reparations of the temple? Take you therfore money no more according to
your order, but render it to the reparation of the temple. \V And the
prieſtes were prohibited to take money anie more of the people, and to
make the reparations of the temple. \V And Ioiada the high prieſt tooke
\TNote{A cheſt or almes boxe.}
a treaſurie, and opened a hole in the toppe, and ſette it by the altar
at the right hand of them that goe into the houſe of our Lord, and the
prieſtes that kept the doores, did caſt into it al the money, that was
brought to the temple of our Lord. \V And when they ſaw that there was
very much money in the treaſurie, the
%%% 0807
kinges ſcribe, and the high prieſt went vp, and powred it out, and
counted the money, that was found in the houſe of our Lord: \V and they
gaue it according to number and meaſure into their hand, which did
ouerſee the maſons of the houſe of our Lord: who beſtowed it on
carpenters, and on maſons, ſuch as wrought in the houſe of our Lord, \V
and made reparation: and on them that hewed ſtones, and that they should
bye trees, and ſtones, that were hewed, ſo that the reparation of the
houſe of our Lord was accomplished in al thinges, which had neede of coſt
to vphold the houſe. \V But yet there
%%% o-0724
were not made of the ſame money the water pottes of the temple of our
Lord, and the fleshhookes, and cenſars, and trumpets, and euerie veſſel
of gold and ſiluer, of the money, that was brought into the temple of
our Lord. \V For it was geuen them that did the worke, that the temple
of our Lord might be repayred: \V and there was no account made with
thoſe men, that receiued the money to diſtribute it to the craftes men,
but vpon their fidelitie they beſtowed it. \V But the money for offence,
& the money for ſinnes, they brought not into the temple of our Lord,
becauſe it was the prieſtes. \V Then Hazael the king of Syria went vp,
and fought agaynſt Geth, and tooke it: and directed his face to goe vp
to Ieruſalem. \V For which cauſe Ioas the king of Iuda tooke al the
ſanctified thinges, which Ioſaphat, and Ioram, and Ochozias his fathers
the kinges of Iuda had
\SNote{Dedicated to holie vſe.}
conſecrated, and which him ſelf had offered: and al the ſiluer, that
could be found in the treaſures of the temple of our Lord, and in the
kinges palace: and ſent it to Hazael the king of Syria, and he retyred
from Ieruſalem. \V But the reſt of the wordes of Ioas, and al that he
did, are not theſe thinges writen in the
%%% !!! XRef ?
Booke of the wordes of the daies of the kinges of Iuda? \V And his
ſeruantes aroſe, and conſpired among them ſelues, and ſtroke Ioas in the
houſe of Mello in the deſcent of Sella. \V For Ioſachar the ſonne of
Semaath, and Iozabad the ſonne of Somer his ſeruantes, ſtroke him, and
he died: and they buried him with his fathers in
\SNote{He was buried in the citie but not in the ſepulcher of the kinges.
\XRef{2.~Paral.~24.}
for his impietie in the latter part of his life.}
the Citie of Dauid, and Amaſias his ſonne reigned for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
