%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 0804
%%% o-0721
\startChapter[
  title={Chapter 11}
  ]

\Summary{Athalia killing al the kinges progenie (except Ioas, who is
  ſaued by his aunt) vſurpeth the kingdome. 4.~But the ſeuenth yeare
  Ioiada the high prieſt crowneth Ioas king, 13.~cauſeth Athalia to be
  ſlaine, 17.~and maketh couenant betwen God, the king, and the people.}

But Athalia the mother of Ochozias ſeing her ſonne dead, aroſe, and
\SNote{Ambition cauſe of much crueltie.}
ſlew al the kinges ſeede. \V But Ioſaba king Iorams daughter, the ſiſter
of Ochozias, taking Ioas the ſonne of Ochozias, ſtole him out of the
middes of the kinges children, that were ſlaine, and his nurce out of
the bedchamber: and hid him from the face of Athalia, that he should not
be ſlaine. \V And he was with her ſixe yeares ſecretly in the houſe of
our Lord. Moreouer Athalia reigned ouer the land. \V And in the ſeuenth
yeare
\SNote{Our Sauiour calleth this high prieſt Zacharias, which ſignifieth
\Emph{bleſſed of our Lord}, for the iuſtice which he did towards Athalia
and Ioas.
\Cite{Hierom li.~4. in Math. c.~23.}}
Ioiada ſent and taking the centurions, and the ſouldiars brought them in
to
%%% o-0722
him into the temple of our Lord, and made a couenant with them: and
adiuring them in the houſe of our Lord, ſhewed them the kinges ſonne: \V
and commanded them, ſaying: This is the thing, which you muſt doe. \V
Let the third part
%%% 0805
of you goe in on the Sabbath, and keep the watch of the kinges
houſe. And let a third part be at the gate Sur: & let a third part be at
the gate behind the dwelling of the shildebearers: and you shal keepe
the watch of the houſe of Meſſa. \V But let two partes of you al that
goe forth on the Sabbath, keepe the watch of the houſe of our Lord about
the king. \V And you shal gard him round about, hauing weapons in your
hands: & if anie man shal enter the precinct of the temple, let him be
ſlaine: and you shal be with the king coming in & going out. \V And the
centurions did according to al thinges, that Ioiada the prieſt had
commanded them: & euerie one taking their men, that went in on the
Sabbath, with them that went out in the Sabbath, came to Ioiada the
prieſt. \V Who gaue them the ſpeares, & the weapons of king Dauid, which
were in the houſe of our Lord. \V And they ſtood euerie one hauing their
weapons in their hand, on the right ſide of the temple, vnto the leaft
ſide of the altar, & of the temple, about the king. \V And he brought
forth the kinges ſonne, and put vpon him the diademe, and the couenant:
and they made him king, and anoynted him: and clapping with the hand,
ſayd: God ſaue the king. \V And Athalia heard the voice of people
running: and she going in to the multitudes into the temple of our
Lord, \V ſaw the king ſtanding vpon the tribunal ſeate according to the
maner, and the ſingers, and trumpettes nere him, and al the people of
the land reioyſing, and ſounding the trumpettes: and she rent her
garmentes, and cried: A conſpiracie, a conſpiracie. \V But Ioiada
commanded  the centurions, that were ouer the armie, and ſayd to them:
Lead her forth without the precinct of the temple, and whoſoeuer shal
folow her, let him be ſtriken with the ſword. For the prieſt had ſayd:
Let her
\SNote{Great reſpect is to be had of holie places. VVherof cometh the
priuilege of Sanctuaries.}
not be ſlaine in the temple of our Lord. \V And they layd handes on her:
and threw her by the way of the entrance of the horſes, beſide the
palace, and she was ſlaine there. \V Ioiada therfore made a couenant
betwen our Lord, and the king, and betwen the people, that it should be
the people of our Lord, and betwen the king and the people. \V And al
the people of the land entered into the temple of Baal, and deſtroyed
his altars, and his images they brake in peces ſtoutly: Mathan alſo the
prieſt of Baal they ſlew before the altar. And the prieſt ſette gardes
in the houſe of our Lord. \V And he tooke the centurions, and the
legions of Cerethi
%%% 0806
and Phelethi, and al the people of the land, and they brought the king
from the houſe of our Lord: and they came by the way of the gate of the
shildebearers into the palace, and he ſate vpon the throne of the
kinges. \V And al the people of the land reioyced, and the citie was
quiet: but Athalia was ſlaine with the ſword in the kinges houſe. \V And
Ioas was ſeuen yeares old, when he began to reigne.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
