%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 0822
%%% o-0737
\startChapter[
  title={Chapter 19}
  ]

\Summary{Ezechias in affliction requeſteth the prayers of Iſaias the
  prophet, 6.~who aſſureth him of Gods helpe. 8.~The king of the
  Aſsyrians ſtil threatneth and blaſphemeth. 15.~Ezechias praieth,
  20.~and God hearing the prayers of the one, and blaſphemies of the
  other, 28.~promiſeth to protect Ieruſalem. 35.~An Angel in one night
  killeth an hundred fourſcore and fiue thouſand of the Aſsyrians campe,
  their king returneth to Niniue, is there ſlaine by two of his owne
  ſonnes, and an other ſonne reigneth in his place.}

VVhich thinges when Ezechias the king had heard, he rent his garmentes,
and was couered with ſackcloth, and entered into the houſe of our
Lord. \V And he ſent Eliacim the gouernour of the houſe, and Sobna the
ſcribe, and the ancientes of the prieſtes couered with ſackclothes, to
Iſaias the prophete the ſonne of Amos. \V Who ſayd to him: Thus ſaith
Ezechias: This day is a day of tribulation, and rebuke, and of
blaſphemie: the children are come to the birth, and the woman in trauel
hath not ſtrength. \V If perhaps our Lord thy God wil heare al the
wordes of Rabſaces, whom the king of the Aſſyrians his maiſter hath ſent
to vpbrayd the liuing God, and reproue with wordes, which our Lord thy
God hath heard: and make thou prayer for the remnantes that are
found. \V The ſeruantes therfore of king Ezechias came to Iſaie. \V And
Iſaie ſayd to them: Thus shal you ſay to your maiſter: Thus ſayth our
Lord: Feare not for the wordes which thou haſt heard, with which the
ſeruantes of the king of the Aſſyrians haue blaſphemed me. \V Behold I
wil ſend into him a ſpirit, and he shal here a meſſage, and ſhal returne
into his country, and I wil ouerthrow him with the ſword in his
country. \V Rabſaces therfore returned, and found the king of the
Aſſyrians expugning Lobna: for he heard that he was departed from
Lachis. \V And when he had heard of Tharaca the king of Æthiopia,
ſaying:
%%% 0823
Behold, he is come forth to fight agaynſt thee: and went agaynſt him, he
ſent meſſengers to Ezechias, ſaying: \V Say this to Ezechias the king of
Iuda: Let not thy God ſeduce thee, in whom thou haſt confidence: neither
ſay thou: Ieruſalem ſhal not be deliuered into the handes of the king of
the Aſſyrians. \V For thou thyſelf haſt heard what the kinges of the
Aſſyrians haue done to al the countries, how they haue ſpoyled them:
canſt thou therfore onlie be deliuered? \V Why haue the goddes of the
Nations deliuered al thoſe, whom my fathers haue deſtroyed, to witte,
Gozan, & Haran, and Reſeph, and the children of Eden, which were in
Thelaſſar? \V Where is the king of Emath, and
%%% o-0738
the king of Arphad, and the king of the citie of Sepharuaim, of Ana and
Aua? \V Therfore when Ezechias had receiued the letters of the hand of
meſſengers, & had read them, he went vp into the houſe of our Lord, &
layd them open
\SNote{Before the Arke & Propitiatorie being the ſpecial place of
prayer.}
before our Lord, \V & prayd in his ſight, ſaying: Lord God of Iſrael,
which ſitteſt vpon the cherubins, thou art the only God of al the kinges
of the earth: thou madeſt heauen and earth: \V Incline thine eare, and
heare: open Lord thine eies, and ſee: and heare al the wordes of
Sennacherib, who hath ſent to vpbrayd vnto vs the liuing God. \V In
verie deede Lord, the kinges of the Aſſyrians haue deſtroyed Nations,
and the countries of al. \V And they haue caſt their goddes into fire:
for they were not goddes, but the workes of mens handes of wood and
ſtone, and they deſtroyed them. \V Now therfore O Lord our God, ſaue vs
from his hand, that al the kingdomes of the earth may know, that thou
art the Lord the onlie God. \V And Iſaie the ſonne of Amos ſent to
Ezechias, ſaying: Thus ſaith our Lord the God of Iſrael: That which thou
haſt beſought me concerning Sennacherib the king of the Aſſyrians, I
haue heard. \V This is the word, that our Lord hath ſpoken of him: The
\SNote{Though manie in the kingdom of Iuda fel to idolatrie, yet in
reſpect of the reſt publikly profeſſing true faith and religion the
Church is ſtil called a virgin and deſpiceth
\Fix{and idolaters,}{idolaters,}{obvious typo, fixed in other}
blaſphemers, and falſe goddes.}
virgine daughter of Sion hath diſpiſed thee, and ſcorned thee: Behinde
thy backe hath the daughter of Ieruſalem wagged her head. \V Whom haſt
thou vpbrayded, and whom haſt thou blaſphemed? againſt whom haſt thou
exalted thy voice, and lifted vp thine eies in high? againſt the holie
one of Iſrael. \V By the hand of thy ſeruantes thou haſt vpbrayded our
Lord, and haſt ſayd: In the multitude of my chariotes haue I climed the
height of mountaynes, in the toppe of Libanus, and haue cut downe high
cedars therof, and the choſen firretrees of it. And I haue
%%% 0824
entered into the borders therof, & the forreſt of the
\SNote{A pleaſant hill in the foreſt.}
carmel therof \V haue I cut downe. And I haue drunke ſtrange waters, and
haue dried vp with the ſteppes of my feete al the waters incloſed. \V
Why, haſt thou not heard what I haue done from the begining? From
ancient daies I haue made that thing, and now I haue brought it to
effect: and fenſed cities ſhal be into ruine of litle hilles bickering
together. \V And they that ſitte in them, weake of hand, they haue
trembled and are confounded, they became as the graſſe of the field, and
the greene herbe of the roofes of houſes, which withered before it came
to ripeneſſe. \V Thy habitation, and thy going out, and thy coming in,
and thy waye I haue knowen before, & thy furie againſt me. \V Thou haſt
bene madde againſt me, & thy pride hath aſcended into myn eares: I
therfore wil put a ring in thy noſtrels, and a bitte in thy lippes, and
wil bring thee back into the way, by the which thou cameſt. \V And to
thee Ezechias this ſhal be a ſigne: Eate this yeare what thou ſhalt
finde: & in the ſecond yeare, the things that growe of themſelues:
moreouer in the third yeare ſawe ye and reape: plant ye vineyardes, and
eate the fruite of them. \V And whatſoeuer ſhal be leaft of the houſe of
Iuda, ſhal take roote downeward, and beare fruit vpward. \V For out of
Ieruſalem there ſhal remnantes goe forth, and that
%%% o-0739
which is to be ſaued from the mounte of Sion: the zeale of the Lord of
hoſtes ſhal doe this. \V Wherfore thus ſayth our Lord of the king of the
Aſſyrians: He ſhal not enter into this citie, nor ſhoote arrowe into it,
neither ſhal ſhield occupie it, nor munition compaſſe it. \V By the way
that he came, he ſhal returne: and into this citie he ſhal
\Fix{nos}{not}{obvious typo, fixed in other}
enter, ſayth our Lord. \V And I wil protect this citie, and wil ſaue it
for my ſelf, and for Dauid my ſeruant. \V It came to paſſe therfore in
that night, an Angel of our Lord came, and ſtroke in the campe of the
Aſſyrians an hundred eightie fiue thouſand. And when he was riſen early,
he ſawe al the bodies of the dead, \V and Sennacherib the king of the
Aſſyrians departing went away, and taried in Niniue. \V And when he
adored in the temple of Neſroch his god, Adramelech and Saraſar his
ſonnes ſtroke him with the ſword, and they fled into the land of the
Armenians, and Aſarhaddon his ſonne reigned for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
