%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 0786
%%% o-0705
\startChapter[
  title={Chapter 4}
  ]

\Summary{Eliſeus ſo multiplieth a poore widowes oyle, that she payeth
  her debtes and liueth of the reſt. 8.~By his prayers a Sunamite woman
  hath a ſonne. 18.~Which dying he rayſeth to life. 38.~He taketh away
  the bitternes of coloquintida, which by chance was put in the potte,
  42.~and feedeth manie with few loaues.}

And a certayne woman of the wiues of the prophetes cried to Eliſeus,
ſaying: Thy ſeruant my huſband is dead, and thou knoweſt that thy
ſeruant was one that feared God, & behold the creditour is come to take
away my two ſonnes to ſerue him. \V To whom Eliſeus ſayd: What wilt thou
that I doe for thee? Tel me, what haſt thou in thy houſe? But she
anſwered: I thy handmaid haue nothing in my houſe, but a litle oyle, to
anoynt me withal. \V To whom he ſayd: Goe, borrow of al thy neighbours
emptie veſſels not a few. \V And goe in, and shut thy doore, when thou
art within, thou and thy ſonnes: and powre therof into al theſe veſſels:
and when they shal be ful, thou ſhalt take them away. \V The woman
therfore went, and shut the doore vpon her, and vpon her ſonnes: they
brought her veſſels, and she powred in. \V And when the veſſels were
ful, she ſayd to her ſonne: Bring me yet a veſſel. And he anſwered: I
haue none. And the oile ſtood. \V And she came, and told the man of
God. And he ſayde: Goe, ſel the oyle, and pay thy creditour: and thou
and thy ſonnes liue of the reſt.
%%% o-0706
\V And there came a certayne day, and Eliſeus paſſed by Sunam: and there
was there a great woman, which held him to eate bread: and when he
paſſed often that way, he turned in to her to eate bread. \V Who
%%% 0787
ſayd to her huſband: I perceiue that this is a holie man of God, which
paſſeth by vs often. \V Let vs therfore make him a litle chamber, and
ſette him a litle bed in it, and a table, and a ſtoole, and a
candleſticke, that when he cometh to vs he may tarie there. \V There
came therfore a certayne day, and coming he turned in to the chamber,
and reſted there. \V And he ſayd to Giezi his ſeruant: Call this
Sunamite. Who when he had called her, and she ſtoode before him, \V he
ſayd to his ſeruant: Speake to her: Behold thou haſt diligently
miniſtred to vs in al thinges, what wilt thou that I doe for thee? haſt
thou any buſines, and wilt thou that I ſpeake to the king, or the
general of the warrefare? Who anſwered: I dwel in the middes of myne owne
people. \V And he ſayd: What wil she then that I doe for her? And Giezi
ſayd: Aske not: for she hath no ſonne, and her huſband is old. \V He
therfore cõmanded him to cal her: who when she was called, and ſtoode
before the doore, \V he ſayd to her: At this tyme, this ſelf ſame houre,
if life accompanie, thou shalt haue a ſonne in thy wombe. But she
anſwered: Doe not I beſech thee my lord, man of God, doe not lie to thy
handmayd. \V And the woman conceiued, and brought forth a ſonne in that
time, and in the ſelf ſame houre, that Eliſeus had ſayd. \V And the
childe grewe. And vpon a certaine day, when going forth he went to his
father, vnto the reapers, \V he ſayd to his father: My head aketh, my
head aketh. But he ſaid to his ſeruant: Take him, and bring him to his
mother, \V who when he had taken him, and brought him to his mother, she
ſette him vpon her knees vntil noone, and he dyed. \V And she went vp,
and laid him vpon the bed of the man of God, and shut the doore: and
going forth \V called her husband, and ſayd: Send with me, I beſech
thee, one of the ſeruantes, and an aſſe, that I may runne to the man of
God, and returne. \V Who ſayd to her: For what cauſe doeſt thou goe to
him? The Calendes be not to day, nor the Sabbath. Who anſwered: I wil
goe. \V And she ſadled the aſſe, and commanded her ſeruant: driue, and
make haſt, make no ſtay in going. And doe that which I command thee. \V
She therfore went forward, and came to the man of God into mount Carmel:
and when the man of God ſaw her ouer againſt him, he ſaid to Giezi his
ſeruant: Behold that Sunamite. \V Goe therfore to meete her, and ſay to
her: Is al wel about thee, and about thy husband,
%%% 0788
and about thy ſonne? Who anſwered: Wel. \V And when she was come to the
man of God into the mount, she caught his feet: and Giezi came to remoue
her. And the man of God ſaid: Let her alone, for her ſoule is in
anguish, and our Lord hath hid it from me, and hath not told me. \V Who
ſaid to him: Did I aske a ſonne of my Lord? Did I not ſay to thee: Mocke
me not? \V And he ſaid to Giezi: Girde thy loynes, and take my ſtaffe in
thy hand, & goe. If a man
%%% o-0707
meete thee, ſalute him not: and if any man ſalute thee, anſwer him not:
and thou ſhalt put my ſtaffe vpon the face of the childe. \V Moreouer
the mother of the childe ſaid: Our Lord liueth, and thy ſoul liueth, I
wil not leaue thee. He aroſe therfore, and folowed her. \V But Giezi was
gone before them, and had put the ſtaffe vpon the face of the childe, &
there was not voice, nor ſenſe: and he returned to meete him, and told
him, ſaying:
\SNote{God ſent his law by his ſeruant, and it auailed not to mankinde
dead in ſinnes. But he coming & accomodating himſelf to our infirmitie,
participating our death we are
\Fix{quicned.}{quickned.}{likely typo, fixed in other}
So
\Fix{S.~Auguſting}{S.~Auguſtin}{obvious typo, fixed in other}
myſtically applieth this figuratiue miracle to Chriſt and his Church.
\Cite{li.~12. c.~35. cont. Faſt. Manich.}
and manie like thinges of the old Teſtament he ſheweth to be figures of
the new.}
The childe is not riſen. \V Eliſeus therfore went into the houſe, and
behold the childe lay dead on his bed: \V and going in he shut the doore
vpon him, & vpon the childe & prayed to our Lord. \V And he went vp, and
lay vpon the childe: and he put his mouth vpon his mouth, & his eies
vpon his eies & his handes vpon his handes: and he bowed him ſelf ouer
him, and the childes fleſh was warmed. \V But he returning, walked vp
and downe in the houſe, once hither & thither: & he went vp, and lay
vpon him: and the childe gaped ſeuen tymes, and opened his eies. \V And
he called Giezi, & ſaid to him: Call this Sunamite. Who being called,
went in vnto him: Who ſaid: Take thy ſonne. \V She came, and fel at his
feete, and adored vpon the ground: and tooke her ſonne, and went out, \V
and Eliſeus returned into Galgal. And there was a famine in the land,
and the children of the prophetes dwelt before him; and he ſayd to one
of his ſeruantes: Sette on a great potte, and ſeeth broth for the
children of the prophetes. \V And one went out into the field to gather
wilde herbes: and he found as it were a wilde vine, and gathered of it
the colocynthides of the field, and filled his mantel, and returning cut
it into the potte of broth, for he knew not what it was. \V They
therfore powred it to their felowes, to eate: and when they had taſted
of the broth, they cried, ſaying: Death is in the potte, man of God. And
they could not eate. \V But he ſayd: Bring meale, and when they had
brought it, he caſt it into the potte, and ſaid: Powere in for the
multitude, that they may eate. And
%%% 0789
there was no more anie bitterneſſe in the potte. \V And a certaine man
came from Baalſaliſa bringing to the man of God loaues of the firſt
fruites, twentie barley loaues, and new wheate in his ſcrippe. But he
ſaid: Geue to the people, that they may eate. \V And his ſeruant
anſwered him: How much is this, that I ſhould ſette it before an hundred
men? Agayne he ſaid: Geue to the people, that they may eate: for thus
ſaith our Lord: They ſhal eate, and there shal be leaft. \V He therfore
ſette it before them: who did eate, and there was leaft according to the
word of our Lord.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
