%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 0780
%%% o-0700
\startChapter[
  title={Chapter 1}
  ]

\Summary{Ochozias
\MNote{The firſt part.

Of thinges donne in both kingdomes, with the declination and ruine of
Iſrael.}
King of Iſrael conſulting Beelzebub, for his ſicknes,
 is blamed by Elias, and fortold, that he shal die. 9.~Fire from heauen
 deuoureth two capitaines with ech of them fiftie men. 13.~The third by
 his more modeſtie eſcapeth the like danger. 15.~With him Elias cometh
 to the King. 17.~The ſame King dieth, and his brother Ioram ſuccedeth.}

And Moab moued warre agaynſt Iſrael, after that Achab was dead. \V And
Ochozias fel through the ſtanchions of his vpper chamber which he had in
Samaria, and was ſicke: and he ſent meſſengers, ſaying to them: Goe,
conſult Beelzebub the god of Accaron, whether I may liue of this my
infirmitie. \V And
%%% 0781
an Angel of our Lord ſpake to Elias the Theſbite, ſaying: Ariſe, and goe
vp to meete the meſſengers of the king of Samaria, and thou shal ſay to
them: What is there not a God in Iſrael, that ye goe to conſult
Beelzebub the god of Accaron? \V Wherfore thus ſaith our Lord: From the
bed on which thou art aſcended, thou shalt not goe downe, but dying thou
ſhalt die. And Elias went away. \V And the meſſengers returned to
Ochozias. Who ſaid to them: Why are you returned? \V But they anſwered
him: A man mette vs: and ſayd to vs: Goe, and returne to the king, that
ſent you, and you ſhal ſay to him: Thus ſaith our Lord: Doeſt thou
therfore ſent to conſult Beelzebub the god of Accaron, becauſe
%%% o-0701
there was not God in Iſrael? Therfore from the bed, which thou art vpon,
thou ſhalt not goe downe, but dying thou shalt dye. \V Who ſaid to them:
What shape and habite had that man, which mette you, and ſpake theſe
wordes? \V But they ſaid: A hearie man, and girded about his raines with
a girdle of lether: Who ſaid:
\SNote{Elias was knowen by his much hayre, and diſtinct habite, from
ordinary men.}
It is Elias the Theſbite. \V And he ſent vnto him a captaine of fiftie
men, and the fiftie that were vnder him. Who went vp and ſayd to him
ſitting in the toppe of the mount: Man of God, the king hath commanded
that thou come downe. \V And Elias anſwering, ſayd to the captaine of
fiftie men: If I be a man of God,
\SNote{In zele of iuſtice Elias procured fire from heauen to burne theſe
proud capitaines and their men, as he procured fire to burne the
holocauſt, and then ſlew the falſprophetes.
\XRef{3.~Reg.~18.}
\Cite{apud. Aug. li.~2. c.~20. di mirabil. S.~Script.}}
let fyre come downe from heauen, and deuoure thee, and thy fiftie. Fyre
therfore came downe from heauen, and deuoured him and the fiftie men
that were with him. \V And he ſent againe vnto him an other captaine of
fiftie men, and his fiftie with him. Who ſpake to him: Man of God, Thus
ſaith the king: Make haſt, come downe. \V Elias anſwering ſaid: If I be
a man of God, let fyre come downe from heauen, and deuoure thee, and thy
fiftie. Fyre therfore came downe from heauen, and deuoured him, and his
fiftie. \V Agayne he ſent a third captaine of fiftie men, and the fiftie
that were with him. Who when he was come, bowed his knees toward Elias,
and prayed him and ſayd: Man of God deſpiſe not my life and the liues of
thy ſeruantes that are with me. \V Behold fyre came downe from heauen,
and hath deuoured the two firſt captaynes of fiftie men, and the
fifties, that were with them: but now I beſech
\Fix{the}{thee}{obvious typo, fixed in other}
that thou haue mercie on my life. \V And an Angel of our Lord ſpake to
Elias, ſaying: Goe downe with him, feare not. He therfore aroſe, and
went
%%% 0782
downe with him to the king, \V and ſpake to him: Thus ſaith our Lord:
Becauſe thou haſt ſent meſſengers to conſult Beelzebub the god of
Accaron, as though there were not a God in Iſrael, of whom thou mighteſt
aske the word, therfore from the bed, which thou art aſcended vpon,
thou shalt not deſcend, but dying thou shalt die. \V He died therfore
according to the word of our Lord which Elias ſpake, & Ioram his brother
reigned for him, in the ſecond yeare of Ioram the ſonne of Ioſaphat the
king of Iuda: for he had no ſonne. \V But the reſt of the wordes of
Ochozias, which he wrought, are not theſe written in the
%%% !!! XRef ?
Booke of the wordes of the daies of the kinges of Iſrael?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
