%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 0814
%%% o-0730
\startChapter[
  title={Chapter 16}
  ]

\Summary{Achaz king of Iuda beſides other idolatrie, ſacrificeth his
  owne ſonne to idoles. 5.~The kinges of Iſrael and Syria beſiegeing
  Ieruſalem, the Aſsyrians are hyred to releue it. 10.~Achaz cauſeth
  Vrias to make an altar like to one in Damaſcus, 14.~in place of Gods
  altar: 17.~remoueth diuers other things perteyning to diuine ſeruice,
  19.~dieth, and his ſonne Ezechias reigneth.}

In the ſeuenteth yeare of Phacee the ſonne of Romelia reigned Achaz the
ſonne of Ioatham king of Iuda. \V Twentie yeares old was Achaz when he
began to reigne, and he reigned ſixtene yeares in Ieruſalem: he did not
that which was pleaſing in the ſight of our Lord his God, as Dauid his
father. \V But he walked in the way of the kings of Iſrael: moreouer
alſo he conſecrated his ſonne, making him paſſe through fire according
to the idols of the Gentils: which our Lord deſtroyed before the
children of Iſrael. \V He immolated
%%% 0815
alſo victimes, and burnt incenſe in the excelſes, and on the hilles, and
vnder euery tree full of greene leaues. \V Then went vp Raſin the king
of Syria, and Phacee the ſonne of Romelia the king of Iſrael into
Ieruſalem to fight: & when they beſieged Achaz, they were not able to
ouercome him. \V At that time Raſin the king of Syria reſtored Aila to
Syria, and threw out the Iewes out of Aila: and the Idumeians came into
Aila, and dwelt there vnto this day. \V And Achaz ſent meſſengers to
Theglathphalaſar the king of the Aſſyrians, ſaying: I am thy ſeruant,
and thy ſonne: come vp, and ſaue me out of the hand of the king of
Syria, and from the hand of the king of Iſrael, which are riſen together
againſt me. \V And when he had gathered together the ſiluer and gold,
that could be founde in the houſe of our Lord, and in the kings
treaſures, he ſent giftes to the king of the Aſſyrians. \V Who alſo
agreed to his wil: for the king of the
%%% o-0731
Aſſyrians went vp into Damaſcus, and waſted it: and he transferred the
inhabitants therof to Cyrene, but Raſin he ſlew. \V And king Achaz went
forth to meete Theglathphalaſar the king of the Aſſyrians into Damaſcus,
and when he had ſeene the altar of Damaſcus, king Achaz ſent to Vrias
the prieſt a paterne of it, and a ſimilitude according to al the worke
therof. \V And Vrias the prieſt built an altar, according to al things,
which king Achaz had commanded, from Damaſcus, ſo did the prieſt Vrias,
whileſt king Achaz came from Damaſcus. \V And when the king was come
from Damaſcus, he ſaw the altar and worſhipped it: and went vp and
immolated holocauſtes, and his ſacrifice, \V and offered libamentes, and
powred the bloud of the pacifiques, which he had offered vpon the
altar. \V Moreouer the altar of braſſe that was before our Lord, he
remoued from the face of the temple, and from the place of the altar,
and from the place of the temple of our Lord: and he ſett it at the ſide
of the altar toward the North. \V King Achaz alſo commanded Vrias the
prieſt, ſaying: Vpon the greater altar offer the morning holocauſt, and
the euening ſacrifice, and the kings holocauſt, and his ſacrifice, and
the holocauſt of the whole people of the land, and their ſacrifices, and
their libamentes: and al the bloud of the holocauſt, and al the bloud of
the victime thou ſhalt powre out vpon it: but the altar of braſſe ſhal
be prepared readie at my pleaſure. \V Vrias therfore the prieſt did
according to al
%%% 0816
things, which king Achaz had commanded him. \V And king Achaz tooke the
engrauen feete, and the lauatorie that was vpon them: and he tooke downe
the
\TNote{The great veſſel.}
ſea from the oxen of braſſe, that held it vp, and put it vpon the
pauement paued with ſtone. \V The
\TNote{The place vvhere the king offered.}
Muſach alſo of the Sabbath, which he had built in the temple: and the
entrie of the king outward, he turned into the temple of our Lord becauſe
of the king of the Aſſyrians. \V But the reſt of the wordes of Achaz,
which he did, are not theſe writen in the
%%% !!! XRef ?
Booke of the wordes of the dayes of the kinges of Iuda? \V And Achaz
ſlept with his fathers, and was buried with them in the Citie of Dauid,
and Ezechias his ſonne reigned for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
