%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 0799
%%% o-0717
\startChapter[
  title={Chapter 9}
  ]

\Summary{Iehu is anointed king of Iſrael, 7.~to deſtroy the houſe of
  Achab and Iezabel. 14.~He preſently killeth Ioram king of
  Iſrael. 27.~Likewiſe Ochozias king of Iuda. 30.~Alſo Iezabel, who is
  eaten by dogges.}

And Eliſeus the prophete called one of the children of the prophetes, and
ſayd to him: Girde thy loynes, and take this litle boxe of oyle in thy
hand, and goe into Ramoth Galaad. \V And when thou ſhalt come thither,
thou ſhalt ſee Iehu the ſonne of Ioſaphat the ſonne of Namſi: and going
in thou ſhalt rayſe him out of the middes of his brethren, and shalt
bring him into an inner chamber. \V And holding the litle boxe of oyle,
thou shalt powre vpon his head, and shalt ſay: Thus ſayth our Lord: I
haue anoynted thee king ouer Iſrael. And thou shalt open the doore, and
flee, and shalt not ſtay there. \V The yongman therfore the childe of a
prophete went into Ramoth Galaad, \V and entered in thither: and behold
the captaynes of the armie ſate, and he ſayd: I haue a word to thee ô
prince. And Iehu ſayd: To whom of vs al? But he ſayd: To thee ô
prince. \V And he aroſe, and went into the chamber: but he powred oyle
vpon his head, and ſayd: Thus ſayth our Lord the God of Iſrael: I haue
anoynted thee king ouer the people of the Lord of Iſrael, \V and thou
shalt ſtrike the houſe of Achab thy maiſter, and I wil reuenge the bloud
of my ſeruantes the prophetes, and the bloud of al the ſeruantes of the
Lord of the hand of Iezabel. \V And I wil 
deſtroy al the houſe of Achab, and wil kil of Achab him that piſſeth
agaynſt a wal, and the shut vp, and the meaneſt in Iſrael. \V And I wil
make the houſe of Achab, as the houſe of Ieroboam the ſonne of Nabat,
and as the houſe of Baaſa the ſonne of Ahias. \V Iezabel alſo the dogges
ſhal eate in the field of Iezrahel, neither shal there be anie to burie
her. And he opened the doore, and fled. \V And Iehu went forth to his
maiſters ſeruantes. Who ſaid to him: Are al thinges wel? Why came this
madde man to thee? Who ſayd to them: You know the man, and what he
ſpake. \V But they anſwered: It is falſe, but rather doe thou tel
vs. Who ſayd to them: Thus and thus he ſpake to me: and he ſayd: Thus
ſayth our Lord, I haue anoynted thee king ouer Iſrael. \V They therfore
made haſt, and euerie man taking his mantel layd it vnder his feete,
after the ſimilitude of a iudgement ſeate, & they ſounded the
%%% 0800
trumpet, and ſayd: Iehu hath the kingdom. \V Iehu therfore the ſonne of
Ioſaphat the ſonne of Namſi conſpired agaynſt Ioram. Moreouer Ioram had
beſieged Ramoth Galaad, he, and al Iſrael agaynſt Hazael the king of
Syria: \V and was returned to be cured in Iezrael of the woundes, for
the Syrians had wounded him, fighting
%%% o-0718
againſt Hazael the king of Syria. And Iehu ſayd: If it pleaſe you, let
no man goe forth fugitiue out of the citie, leſt he goe, and tel in
Iezrahel. \V And he mounted, and went into Iezrahel: for Ioram was ſicke
there, and Ochozias the king of Iuda was come downe to viſite Ioram. \V
The watchman therfore, that ſtood vpon the toure of Iezrahel, ſaw the
troupe of Iehu coming, & ſaid: I ſee a troupe. And Ioram ſaid: Take a
Chariote, and ſend to meete them, and let him that goeth ſay: Are al
thinges wel? \V He went therfore, that was gotten vp into the chariote,
to meete him, and ſayd: Thus ſaith the king: Are al thinges peaceable?
And Iehu ſayd: What haſt thou to doe with peace? paſſe, and folow
me. The watchman alſo told, ſaying: The meſſenger came to them, &
returneth not. \V He ſent alſo the ſecond chariote of horſes: and he
came to them, and ſaid: Thus ſayth the king: Is there peace? And Iehu
ſayd: What haſt thou to doe with peace? paſſe, and folow me. \V And the
watchman told, ſaying: He is come as far as they, & returneth not: and
it is the pace as it were the pace of Iehu the ſonne of Namſi, for he
goeth amayne. \V And Ioram ſayd: Make readie the chariote. And they made
readie his chariote, and Ioram the king of Iſrael went forth, and
Ochozias the king of Iuda, ech in their chariotes, and they went forth
to meete Iehu, and found him in the field of Naboth the Iezrahelite. \V
And when Ioram had ſeene Iehu, he ſayd: is there peace Iehu? But he
anſwered: What peace? The fornications of Iezabel thy mother, and her
manie ſorceries are in their vigour. \V And Ioram turned his hand, and
fleing ſayd to Ochozias: Treaſon Ochozias. \V Moreouer Iehu bent  his
bow with his hand, and ſtroke Ioram betwen the shoulders: and the arrow
went out through his hart, and immediatly he fel in his chariote. \V And
Iehu ſayd to captaine Badacer: Take him, throwe him forth in the field of
Naboth the Iezrahelite: for I remember when I and thou ſitting in a
chariote did folow Achab this mans father, that our Lord lifted vp this
burden vpon him, ſaying: \V If not for the bloud of Naboth, and
\SNote{VVhen Naboth was falſly accuſed & vniuſtly ſtoned to death, as if
he had blaſphemed God, and curſed the king, for his pretended crimes,
his ſonnes were alſo ſlaine, and his landes and goodes confiſcate: which
appeareth by the kings preſent going to poſſeſſe the vinyard.
\XRef{3.~Reg.~21.}}
for
%%% 0801
the bloud of his children, which I ſaw yeſterday, ſayth our Lord, I
requite thee not in this field ſayth our Lord. Now therfore take him,
and throw him into the field according to the word of our Lord. \V But
Ochozias the king of Iuda ſeing this, fled by the way of the houſe of
the garden: and Iehu purſewed him, and ſaid: This man alſo ſtrike ye in
his chariote. And they ſtroke him in the going vp of Gauer, which is
beſide Ieblaam: who fled into Mageddo, and died there. \V And his
ſeruantes layd him vpon his chariote, and caried him into Ieruſalem: and
they buried him there in his ſepulchre with his fathers in the Citie of
Dauid. \V In the eleuenth yeare of Ioram the ſonne of Achab, reigned
Ochozias ouer Iuda, \V and Iehu came into Iezrahel. Moreouer Iezabel
hearing of his entrance, paynted her face with ſtibicke ſtone, and
decked her head, and beheld through the window \V Iehu coming at the
gate, and ſayd: Can there be peace to Zambri, that killed his
maiſter? \V And Iehu lifted vp his face to the window, and ſayd: What is
she? And
%%% o-0719
two or three eunuches bowed them ſelues to him. \V But he ſayd to them:
Caſt her downe headlong, & they threw her downe and the wal was
ſprinkled with the bloud, and the hoofes of the horſes trode her. \V And
when he was entered in, to eate, and to drinke, he ſayd: Goe, and ſee
that curſed woman, and burie her: becauſe she is a kinges daughter. \V
And when they went to burie her, they found nothing but the skul, and
the feete, and the extreme partes of the handes. \V And returning they
told him. And Iehu ſayd: It is the word of our Lord, which he ſpake by
his ſeruant Elias the Theſbite, ſaying: In the field of Iezrahel shal
the dogges eate the flesh of Iezabel, \V and the flesh of Iezabel shal be
as dung vpon the face of the earth in the field of Iezrahel, ſo that
they which paſſe by shal ſay: Is this that ſame Iezabel?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
