%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 0801
%%% o-0719
\startChapter[
  title={Chapter 10}
  ]

\Summary{The Samaritanes fearing the force of Iehu, chooſe no other
  king, but offer him their ſeruice: 6.~and by his commandment kil the
  late kings ſeuentie ſonnes. 12.~Fourtie two brothers of Ochozias late
  king of Iuda are ſlaine. 15.~Iehu making league with Ionadab, vtterly
  deſtroyeth Achabs houſe. 18.~By a ſtratagem killeth al the worshippers
  of Baal, 26.~burneth his ſtatua, & turneth his temple into a iakes:
  28.~but maintaineth Ieroboams golden calues. 32.~The Aſyrians afflict
  Iſrael. 34.~Iehu dieth, and his ſonne Ioachaz reigneth.}

%%% 0802
And Achab had ſeuentie ſonnes in Samaria: Iehu therfore wrote letters,
and ſent into Samaria to the chiefe of the citie, and to the ancientes,
and to them that brought vp Achabs children, ſaying: \V As ſowne as you
shal receiue theſe letters, ye that haue your maiſters ſonnes, and
chariotes, and horſes, and fenſed cities, and armour, \V chooſe the
better, and him that shal pleaſe you of your maiſters ſonnes, and ſet
him vpon his fathers throne, and fight for the houſe of your lord. \V
They were ſore afrayd, and ſaid: Behold two kinges could not ſtand
before him, and how ſhal we be able to reſiſt? \V The ouerſeers therfore
of the houſe, and the rulers of the citie, and the ancientes, and the
tutors ſent to Iehu, ſaying: We are thy ſeruantes, whatſoeuer thou shalt
command we wil doe, neither wil we make vs a king: Doe thou whatſoeuer
pleaſeth thee. \V And he wrote letters to them agayne, the ſecond tyme,
ſaying: If you be mine, and obey me, take the heades of your maiſters
ſonnes, and come to me this verie houre to morow into Iezrahel. Moreouer
the kinges ſonnes, ſeuentie men were brought vp with the chiefe of the
citie. \V And when the letters were come to them, they tooke the kinges
ſonnes, and ſlew ſeuentie men, and put their heades in baskets, and ſent
them to him into Iezrahel. \V And a meſſenger
%%% o-0720
came, & told him, ſaying: They haue brought the heades of the kinges
ſonnes. Who anſwered: Lay them in two heapes by the entrance of the gate
vntil morning. \V And when it was light, he went forth, and ſtanding
ſayd to al the people: You are iuſt: If I haue conſpired agaynſt my
maiſter, and haue ſlayne him, who hath ſtrooken al theſe? \V See
therfore now there hath not fallen of the wordes of our Lord on the
ground, which our Lord ſpake vpon the houſe of Achab, and our Lord hath
done that which he ſpake in the hand of his ſeruant Elias. \V Iehu
therfore ſmote al that were leaft of the houſe of Achab in Iezrahel, and
al his nobles, and familiars, and prieſtes, til there remayned no
reliques of him. \V And he aroſe, and came into Samaria: and when he was
come to the cabbin of the ſhepeheardes in the way, \V he found the
brethren of Ochozias the king of Iuda, and he ſayd to them: What are
you? Who anſwered: We are the brethren of Ochozias, and are come downe to
ſalute the kinges ſonnes, and the queenes ſonnes. \V Who ſayd: Take them
aliue. Whom when they had taken aliue, they killed them in a ceſterne
beſide
%%% 0803
the cabbin, two and fourtie men, and he leaft not any of them. \V And
when he was gone thence, he found
\SNote{This Ionadab inſtituted a peculiar rule of religious abſtinence:
which his poſterity duly obſerued.
\XRef{Ierem.~35.}}
Ionadab the ſonne of Rechab coming to meete him, and he bleſſed him. And
he ſayd to him: Is thy hart right as my hart with thy hart? And Ionadab
ſayd: It is. If it be ſo, quoth he, geue me thy hand. Who gaue him his
hand. But he lifted him vp to him into the chariote, \V and ſayd to him:
Come with me, and ſee my zele for our Lord. And being ſette in his
chariote, \V he brought him into Samaria. And he ſtroke al that were
leaft of Achab in Samaria, til there was not one, according to the word
of our Lord, which he ſpake by Elias. \V Iehu therfore aſſembled al the
people, and ſayd to them: Achab worshipped Baal a litle, but
\SNote{Iehu ſinned in feaning, and cauſing others to ſacrifice to Baal,
his zele wanting both diſcretion and equitie for \Emph{euil muſt not be
done that good may come therof}.
\XRef{Rom.~3.}}
I wil worshipe him more. \V Now therfore cal to me al the prophetes of
Baal, and al his ſeruantes, and al his prieſtes: let there be none but
that he come, for I haue a great ſacrifice to Baal: He that ſhal be
wanting shal not liue. Moreouer Iehu did this craftely, that he might
deſtroy the worshippers of Baal. \V And he ſayd: Sanctifie a ſolemne day
to Baal. And he called \V and ſent into al the borders of Iſrael, and al
the ſeruantes of Baal came: there was leaft not one that came not. And
they entered into the temple of Baal: and the houſe of Baal was filled,
from one end to the other. \V And he ſayd to them that were ouer the
garmentes: Bring forth garmentes for al the ſeruantes of Baal. And they
brought them forth garmentes. \V And Iehu going in, and Ionadab the
ſonne of Rechab into the temple of Baal, ſayd to the worshippers of
Baal: Search, and ſee leſt perhaps there be any with you of the
ſeruantes of the Lord, but that there be the ſeruantes of Baal onlie. \V
They therfore went in to make victimes and holocauſtes: but Iehu had
prepared him without foureſcore men, and ſayd to them: Whoſoeuer shal
eſcape of theſe men, whom I wil bring into your handes, his life shal be
for the life of him. \V And it came to paſſe, when the holocauſte was
ended, Iehu commanded
%%% o-0721
his ſouldiars and captaynes: Goe in, and ſtrike them, let none
eſcape. And the ſouldiars and captaynes ſtroke them in the edge of the
ſword, and caſt them forth: and they went into the citie of the temple
of Baal, \V and brought forth the ſtatue out of Baals temple, and burnt
it, \V and brake it in peeces. They deſtroyed alſo the temple of Baal,
and made a iakes for it vntil this day. \V Iehu therfore deſtroyed Baal
out of Iſrael:
%%% 0804
\V but yet from the ſinnes of Ieroboam the ſonne of Nabat, who made Iſrael
to ſinne, he departed not, neither forſooke he the golden calues, that
were in Bethel, and Dan. \V And our Lord ſayd to Iehu:
\SNote{\Emph{Moral good vvorkes} done in ſtate of mortal ſinne not
meriting eternal life, are often rewarded temporally.
\Cite{S.~Aug. cont. mend. c.~2.}}
Becauſe thou haſt diligently done that which was right, and that pleaſed
in mine eies, and haſt done al thinges that were in my hart, againſt the
houſe of Achab: thy children ſhal ſitte vpon the throne of Iſrael to the
fourth generation. \V Moreouer Iehu obſerued not to walke in the law of
our Lord the God of Iſrael in al his hart: for he departed not from the
ſinnes of Ieroboam, who had made Iſrael to ſinne. \V In thoſe daies our
Lord began to be wearie of Iſrael: and Hazael ſmote them in al the coſtes
of Iſrael, \V from Iordan agaynſt the Eaſt quarter, al the land of
Galaad, and Gad, and Ruben, and Manaſſes, from Aroer, which is vpon the
Torrent Arnon, and Galaad, and Baſan. \V But the reſt of the wordes of
Iehu, & al that he did, and his ſtrength, are not theſe thinges writen
in the
%%% !!! XRef ?
Booke of the wordes of the dayes of the kinges of Iſrael? \V And Iehu
ſlept with his fathers, and they buried him in Samaria: and Ioachaz his
ſonne reigned for him. \V And the dayes which Iehu reigned ouer Iſrael,
be eight and twentie yeares in Samaria.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
