%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-22


\project douay-rheims


%%% 0828
%%% o-0742
\startChapter[
  title={Chapter 22}
  ]

\Summary{Ioſias repayreth the temple, and Diuine ſeruice. 8.~The booke
  of law is found, 11.~wherupon they conſult our Lord, 15.~and are
  foretold that much euil shal fal vpon them, 18.~but the good king shal
  dye in peace.}

Eight yeares old was Ioſias when he began to reigne, he reigned one and
thirtie yeares in Ieruſalem: the name of his mother was Idida, the
daughter of Hadaia of Beſecath. \V And he did that which was liked
before our Lord, and walked in al the waies of Dauid his father: he
declined not to the right hand, or to the lefte. \V And in the eightenth
yeare of king Ioſias, the king ſent Saphan the ſonne of Aſlia, the ſonne
of Meſſulam, the ſcribe of the temple of our Lord, ſaying to him: \V Goe
to Helcias the high prieſt, that the money may be gathered into a ſumme,
which hath beene brought into the temple of our Lord, which the porters
of the temple haue gathered of the people, \V and let it be geuen to the
%%% o-0743
workemen by the ouerſeers of the houſe of our Lord: who alſo ſhal
deſtribute it to them that worke in the temple of our Lord, to make
reparations of the temple: \V that is, to the carpenters and maſons, and
to them that mend broken places: and that timber may be bought, and
ſtones out of the quarries to repayre the temple of our Lord. \V Yet let
not
%%% 0829
the money which they receiue be accounted to them, but let them haue it
in their powre, and vpon their fidelitie. \V And Helcias the highe
prieſt ſayd to Saphan the ſcribe: I haue found the Booke of the law in
the houſe of our Lord: and Helcias gaue the volume to Saphan, who alſo
did reade it. \V Saphan alſo the ſcribe came to the king, and reported
vnto him that which he had commanded, and ſayd: Thy ſeruantes haue
gathered into a ſumme the money, which is found in the houſe of our
Lord: and they haue geuen it to be deſtributed to the workemen, by the
ouerſeers of the workes of the temple of our Lord. \V Saphan alſo the
ſcribe told the king, ſaying: Helcias the prieſt hath geuen me a
Booke. Which when Saphan had read before the king, \V and the king had
heard the wordes of the law of our Lord, he rent his garmentes. \V And
he willed Helcias the prieſt, and Ahicam the ſonne of Saphan, and
Achobor the ſonne of Micha, and Saphan the Scribe, and Aſaia the kinges
ſeruant, ſaying: \V Goe and conſult our Lord for me, and for the people,
and for al Iuda, concerning the wordes of this volume, which is found:
for the great wrath of our Lord is kindled agaynſt vs: becauſe our
fathers haue not heard the wordes of this Booke, to doe al that is
writen for vs. \V Helcias therfore the prieſt, and Ahicam, and Achabor,
and Saphan, and Aſaia went to Holda a propheteſſe the wife of Sellum the
ſonne of Thecua, the ſonne of Araas keeper of the garmentes, who dwelt
in Ieruſalem in the
\SNote{VVithin the ſecond wall the citie hauing three wals.
\XRef{3.~Reg.~3.}}
ſecond: and they ſpake to her. \V And ſhe anſwered them: Thus ſayth our
Lord the God of Iſrael: Tel the man, that ſent you to me: \V Thus ſayth
our Lord: Behold, I wil bring euils vpon this place, and vpon the
inhabitantes therof, al the wordes of the law which the king of Iuda
hath read: \V becauſe they haue forſaken me, and haue ſacrificed to
ſtrange goddes, prouoking me in al the workes of their handes: and my
indignation ſhal be kindled in this place, and ſhal not be quenched. \V
But to the king of Iuda, that ſent you to conſult our Lord, thus you
ſhal ſay: Thus ſayth our Lord the God of Iſrael: For that thou haſt
heard the wordes of the volume, \V and thy hart is ſtriken with feare,
and thou art humbled before the Lord, hearing the wordes agaynſt this
place, and the inhabitantes therof, to witte that they ſhould become a
wonder and a curſe: and haſt rent thy garmentes, and wept before me, and
I haue
%%% 0830
heard it, ſayth our Lord: \V therfore I wil gather thee to thy fathers,
and thou shalt be gathered to thy ſepulchre in peace, that thy eies may
not ſee al the euils, which I wil bring in vpon this place.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
