%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 25
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-25


\project douay-rheims


%%% 0835
%%% o-0748
\startChapter[
  title={Chapter 25}
  ]

\Summary{Nabuchodonoſor king of Babylon beſeiging Ieruſalem two yeares,
  the ſouldiars preſsed with famine flee away. King Sedecias alſo
  fleeth, 5.~is taken and brought before Nabuchodonoſor. 7.~Al his
  ſonnes are ſlaine in his ſight: his eyes are put out, and he is caried
  into Babylon: the Temple, the kinges palace, and other houſes burned;
  the walles deſtroyed, the people caried captiue 12.~except poore
  husbandmen. 13.~Al veſsel of braſse, ſiluer, and gold broken and
  tranſported. 18.~Saraias high prieſt, and other principal men are
  ſlaine. 22.~Godolias made
  \Fix{gouernerour,}{gouernour,}{obvious typo, fixed in other}
  25.~is ſlaine by Iſmael. 27.~Ioachin is deliuered from priſon, and exalted
  by a new king of Babylon.}

And it came to paſſe in the ninthe yeare of his reigne, the tenth
moneth, the tenth day of the moneth, came Nabuchodonoſor the king of
Babylon, himſelfe and al his armie vnto Ieruſalem, and they beſette it:
and built rampires
\Fix{rond}{round}{obvious typo, fixed in other}
about it. \V And the citie was ſhut vp & trenched about vntil the
eleuenth yeare of king Sedecias, \V the ninth day of the moneth: and
there was a ſore famine in the citie, neither was there bread for the
people of the land. \V And a breach was made into the citie: and al the
men of warre fled in the night by the way of the gate, which is betwen
the duble wal toward the kinges garden (moreouer the Chaldees beſieged
the citie round about.) Sedecias therfore fled by the way, that leadeth
to the champayne of the wildernes. \V And the armie of the Chaldees
purſewed the king, and ouertooke him in the plaine of Iericho: and al
the warriers, that were with him, were diſperſed, and forſooke him. \V
They therfore hauing taken the king, brought him to the king of Babylon
into
%%% o-0749
Reblatha, who ſpake iudgement with him. \V And he ſlewe the ſonnes of
Sedecias before his face, and he
\SNote{Certaine falſe prophetes perſwaded the king and the people not to
beleue the prophetes, which forwarned them of theſe calamities, becauſe
(ſayd they) they contradicte one an other, Ieremie ſaying
\XRef{(ch.~32.}
&
\XRef{34.)}
\Emph{the eyes of Sedecias should ſee the eyes of Nabuchodonoſor, and
should be led into Babylon}: and Ezechiel ſaying
\XRef{(c.~12. v.~13.)}
\Emph{that he should not ſee Babylon}.}
put out his eies, and bound him with cheynes, and
\SNote{Both ſaying moſt truly: for he was caried thither after his eyes
were put out.
\Cite{Ioſephus li.~10. Antiq. c.~10.}}
brought him into Babylon. \V The fifth moneth, the ſeuenth day of the
moneth, that is, the ninetenth yeare of the king of Babylon, came
Nabuzardan the General of the armie, the ſeruant of the king of Babylon
into Ieruſalem. \V And he burnt the houſe of our Lord, the kings houſe,
and the houſes of Ieruſalem, & euerie houſe he burnt with fyre. \V And
al the armie of the Chaldees, which was with the General of the
ſouldiars, deſtroyed the walles of Ieruſalem round about. \V And
Nabuzardan the General of the armie, tranſported the remnant of the
people that remained in the citie, and the fugitiues, that were fled to
the
%%% 0836
king of Babylon, and the reſt of the comon people. \V And of the poore
of the land he leift dreſſers of vineyardes and huſbandmen. \V And the
braſen pillers, that were in the temple of our Lord, and the feete, and
the ſea of braſſe, which was in the houſe of our Lord, the Chaldees
brake, and tranſported al the braſſe into Babylon. \V The pottes alſo of
braſſe, and maſars, and forkes with three teeth, and cuppes, and litle
mortars, and al the veſſel of braſſe, in the which they miniſtred, they
tooke away. \V Moreouer alſo the cenſars, and phials: thoſe that were of
gold: and that were of ſiluer the General of the warfare tooke, \V that
is, two pillers, one ſea, and the feete which Salomon had made in the
temple of our Lord: there was
\SNote{There was ſo exceding much, that they wel could not, or did not
weigh it.}
no weight of the braſſe of al the veſſels. \V One piller had eightene
cubites in height: and the litle head of braſſe vpon it was three
cubites in height: and the nette, and the pomegranates vpon the litle
head of the piller, al of braſſe: the ſecond alſo had the like
adorning. \V Alſo the general of the armie tooke Saraias the chiefe
prieſt, and Sophonias the ſecond prieſt, and three porters. \V And of
the citie one Eunuch, which was captayne ouer the men of warre: and fiue
men of them, that wayted before the king, whom he found in the citie,
and Sopher the captaine of the armie who proued the yong ſouldiars of
the people of the land: and threeſcore men of the common people, which
were found in the citie. \V Whom Nabuzardan the General of the armie
taking, brought them to the king of Babylon into Reblatha. \V And the
king of Babylon ſtroke them, and ſlew them in Reblatha in the land of
Emath: and Iuda was tranſported out of their land. \V And he made
Godolias the ſonne of Ahicam the ſonne of Saphan Gouernour ouer the
people, that was leift in the land of Iuda, which Nabuchodonoſor the
king of Babylon had leift. \V Which when al the captaynes of the
ſouldiars had heard, they and the men that were with them, to witte,
that the king of Babylon had appoynted Godolias, there came to Godolias
into Maſpha Iſmael the ſonne of Mathanias, and Iohanan the ſonne of
Caree, and
\Fix{the Saraia}{Saraia}{obvious typo, fixed in other}
the ſonne of Thanehumeth a Netophathite, and Iezonias
the ſonne of Maachathi, they and their felowes. \V And Godolias ſware to
them and to their
%%% o-0750
companions, ſaying: Be not afrayd to ſerue the Caldees: tarie in the
countrie, and ſerue the king of Babylon, and it shal be wel with you. \V
But
%%% 0837
it came to paſſe in the ſeuenth moneth, there came Iſmahel the ſonne of
Nathanias, the ſonne of Eliſama of the kinges ſeede, and ten men with
him: and ſtroke Godolias, who alſo died: yea the Iewes alſo & the
Chaldees, that were with him in Maſpha. \V And al the people ryſing vp
from litle to great, & the captaynes of the ſouldiars, came into Ægypt
fearing the Caldees. \V But it came to paſſe in the ſeuen and thirteth
yeare of the Tranſmigration of Ioachin the king of Iuda, in the twelfth
moneth, the ſeuen and twenteth day of the moneth: Euilmerodach the king
of Babylon, in the yeare, that he began to reigne,
\SNote{By Gods ſpecial prouidence king Ioachin (otherwiſe called
Iechonias
\XRef{2.~Par.~3.}
&
\XRef{Mat.~1.)}
is exalted, and ſet ouer al the Iewes; vnto whom others ſuccede in like
authoritie, and ſo is fulfilled the prophecie of Iacob.
\XRef{Gen.~49.}
\Emph{The ſcepter shal not be taken from Iuda, nor a duke of his thiegh
til he come that is to be ſent.}}
lifted vp the head of Ioachin the king of Iuda out of priſon. \V And he
ſpake vnto him courteouſly: and he ſette his throne aboue the throne of
the kinges, that were with him in Babylon. \V And he changed his
garmentes, which he had in the priſon, and he did eate bread alwayes in
his ſight, al the dayes of his life. \V A certayne prouiſion alſo he
appointed for him without intermiſſion, which was alſo geuen him of the
king day by day, al the daies of his life.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
