%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 0809
%%% o-0726
\startChapter[
  title={Chapter 14}
  ]

\Summary{Amaſias king of Iuda killeth thoſe that had ſlaine his father,
  7.~and ſtriketh Edom. 8.~Prouoking Ioas king of Iſrael to warre,
  receiueth a contemtible anſwer: 11.~prouoking againe is beaten in
  battle, and Ieruſalem is ranſact. 15.~Ioas dieth and his ſonne
  Ieroboam foloweth the bad ſteppes of the firſt Ieroboam; yet
  recouereth manie places loſt before. And dying his ſonne Zacharias
  reigneth.}

%%% 0810
In the ſecond yeare of Ioas the ſonne of Ioachaz the king of Iſrael
reigned Amaſias the ſonne of Ioas the king of Iuda. \V Fiue and twentie
yeares old was he when he began to reigne: and nine and twentie yeares
he reigned in Ieruſalem, the name of his mother was Ioadan of
Ieruſalem. \V And he did right before our Lord, but yet not as Dauid his
father. He did according to al thinges, which Ioas his father did: \V
but this onlie that he tooke not away the excelſes: for yet the people
immolated, and burnt incenſe in the excelſes. \V And when he obtayned
the kingdom, he ſmote his ſeruantes, which had ſlaine the king his
father: \V but their children that killed him, he did not put to death,
according to that which is written in the booke of the law of Moyſes, as
our Lord commanded, ſaying:
\CNote{\XRef{Deut.~14.}}
The fathers ſhal not die for the children, neither shal the children die
for the fathers: but euerie one shal die in his owne ſinne. \V He ſmote
Edom in the Vail of Salt pittes ten thouſand, and tooke the rocke in
battel, and called the name therof, Iectehel vntil this preſent daye. \V
Then Amaſias ſent meſſengers to Ioas the ſonne of Ioachaz, the ſonne of
Iehu the king of Iſrael, ſaying: Come,
\SNote{Amongſt kinges being at variance, \Emph{Seing one an other}
importeth, as much as to fight a battle.}
let vs ſee one an other. \V And Ioas the king of Iſrael ſent agayne to
Amaſias the king of Iuda ſaying: A thiſtle of Libanus ſent to a cedar
tree, which is in Libanus, ſaying: Geue thy daughter to my ſonne to
wife. And the beaſtes of the foreſt, that are in Libanus, paſſed and
trode the thiſtle. \V Thou ſtriking haſt preuayled ouer Edom, and thy
hart hath puffed thee vp: be content with the glorie, and ſit in thy
houſe: Why prouokeſt thou euil, that thou mayſt fal and Iuda with
thee. \V And Amaſias agreed not. And Ioas the king of Iſrael went vp,
and they ſaw
%%% o-0727
eche other, he and Amaſias the king of Iuda in Bethſames a towne of
Iuda. \V And Iuda was ſtrooken before Iſrael, and euerie man fled into
their tabernacles. \V But Ioas the king of Iſrael did take Amaſias the
king of Iuda the ſonne of Ioas, the ſonne of Ochozias, in Bethſames, and
brought him into Ieruſalem: and he brake downe the wal of Ieruſalem,
from the gate of Ephraim vnto the gate of the corner, fowre hundred
cubites. \V And he tooke al the gold, and ſiluer, and al the veſſel,
that were found in the houſe of our Lord, and in the kinges treaſures,
and hoſtages, and returned into Samaria. \V But the reſt of the wordes
of Ioas which he did, and his ſtrength, wherwith he fought againſt
%%% 0811
Amaſias the king of Iuda, are not theſe thinges writen in the
%%% !!! XRef ?
Booke of the wordes of the daies of the kinges of Iſrael? \V And Ioas
ſlept with his fathers, and was buried in Samaria with the kinges of
Iſrael: & Ieroboam his ſonne reigned for him. \V And Amaſias the ſonne
of Ioas the king of Iuda liued, after that Ioas the ſonne of Ioachaz the
king of Iſrael was dead, fifetene yeares. \V But the reſt of the wordes
of Amaſias, are not theſe thinges writen in the
%%% !!! XRef ?
Booke of the wordes of the daies of the kinges of Iuda? \V And there was
a conſpiracie made againſt him in Ieruſalem: but he fled into
Lachis. And they ſent after him into Lachis, and killed him there. \V
And they caried him away vpon horſes, and he was buried in Ieruſalem
with his fathers in the Citie of Dauid. \V And al the people of Iuda
tooke Azarias ſixtene yeares old, and made him king for his father
Amaſias. \V He built Ælath, and reſtored it to Iuda, after that the king
ſlept with his fathers. \V In the fiftenth yeare of Amaſias the ſonne of
Ioas the king of Iuda, reigned Ieroboam the ſonne of Ioas the king of
Iſrael in Samaria, one and fourtie yeares: \V and he did that which is
euil before our Lord. He departed not from al the ſinnes of Ieroboam the
ſonne of Nabat, who made Iſrael to ſinne. \V He reſtored the borders of
Iſrael from the entrance of Emath, vnto the Sea of the wilderneſſe,
according to the word of our Lord the God of Iſrael, which
\CNote{\XRef{Iona.~1.}}
he ſpake by his ſeruãt Ionas the ſonne of Amathi, the prophete, who was
of Geth, which is in Opher. \V For our Lord ſaw the affliction of Iſrael
exceding bitter, and that they were conſumed vnto the impriſoned &
meaneſt perſons, and that there was none to helpe Iſrael. \V Neither did
our Lord determine that he would deſtroy the name of Iſrael from vnder
heauen, but he ſaued them in the hand of Ieroboam the ſonne of Ioas. \V
But the reſt of the wordes of Ieroboam, and al that he did, and his
ſtrength, wherwith he fought, and how he reſtored Damaſcus, and Emath to
Iuda in Iſrael, are not theſe thinges writen in the
%%% !!! XRef ?
Booke of the wordes of the daies of the kinges of Iſrael? \V And
Ieroboam ſlept with his fathers the kinges of Iſrael, and Zacharias his
ſonne, reigned for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
