%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 0792
%%% o-0711
\startChapter[
  title={Chapter 6}
  ]

\Summary{Eliſeus maketh iron to ſwimme vpon the water: 8.~leadeth the
  king of Syria his men (ſent to apprehend him) blindly into
  Samaria. 20.~Where their eyes being opened, they are curtouſly
  intertained, and freely diſmiſsed. 24.~The Syrians beſiege
  Samaria. 26.~For extreme famine a mother eateth her child. 31.~And the
  king commandeth to kil Eliſeus.}

And the children of the prophetes ſayd to Eliſeus: Behold the place,
wherin we dwel before thee is ſtreite for vs. \V Let vs goe as far as
Iordan, and take out of the wood euerie man ſome timber, that we may
build there a place to dwel in. Who ſaid: Goe. \V And one of them ſayd:
Come therfore thou alſo with thy ſeruantes. He anſwered: I wil
%%% 0793
come. \V And he went with them. And when they were come to Iordan they
cut wood. \V And it chanced, that when one had cut downe timber, the
head of the axe fel into the water: and he cried out, and ſaid: Alas
alas alas my lord, this ſame alſo I
\SNote{His grief was great, becauſe he had not meanes to recompence the
loſſe to the owner.}
did borow. \V And the man of God ſayd: Where fel it? and he ſhewed him
the place: he therfore cut of a pece of wood, and caſt it thither: and
the yron did ſwimme, \V and he ſaid: Take it vp. Who ſtretched forth his
hand, and tooke it. \V And the king of Syria fought againſt Iſrael, and
tooke counſel with his ſeruantes, ſaying: In that and that place let vs
lay embuſhementes. \V Therfore the man of God ſent to the king of
Iſrael, ſaying: Beware thou paſſe not into ſuch a place: becauſe the
Syrians are there in embushementes. \V The king of Iſrael therfore ſent
to the place, which the man of God had told him, and preuented him, and
looked wel to him ſelf there not once or twiſe. \V And the hart of the
king of Syria was trubled for this thing. And calling together his
ſeruantes, he ſayd: Why doe you not tel me who is the betraier of me to
the king of Iſrael? \V And one of his ſeruantes ſayd: Not ſo my lord
king, but Eliſeus the prophet, which is in
%%% o-0712
Iſrael, telleth the king of Iſrael al wordes, whatſoeuer thou ſhalt
ſpeake in thy priuy chamber. \V And he ſayd to them: Goe, and ſee where
he is: that I may ſend, and take him. And they told him, ſaying: Behold
in Dothan. \V He therfore ſent thither horſes and chariotes, and the
force of his armie: who when they were come in the night, they beſette
the citie. \V And the ſeruant of the man of God riſing early, went out,
and ſaw an armie round about the citie, and horſes and chariotes: and he
told him, ſaying: Alas alas alas my lord, what shal we doe? \V But he
anſwered: Feare not: for there are more with vs then with them. \V And
when Eliſeus had prayed, he ſayd: Lord open the eies of this man, that
he may ſee. And our Lord opened the eies of the ſeruant, and
\SNote{A huſbandman, in yorſhire called Ketle had the gift to ſee euil
ſpirites, wherby he often detected & hindered their bad purpoſes.
\Cite{Neubrig. li.~2. c.~21. Rer. Anglic.}}
he beheld: and loe the mountaine ful of horſes, and of firie chariotes
round about Eliſeus. \V But the enemies went downe to him: moreouer
Eliſeus prayed to our Lord, ſaying: Strike, I beſech thee, this people
with blindeneſſe. And our Lord ſmote them, that they ſaw not, according
to the word of Eliſeus. \V And Eliſeus ſayd to them: This is not the way
neither is this the citie: folow me, & I wil shew you the man, whom you
ſeeke. He therfore led them into Samaria: \V and when they
%%% 0794
were entered into Samaria, Eliſeus ſaid: Lord open the eies of theſe
men, that they may ſee. And our Lord opened their eies, and they ſaw
them ſelues to be in the middes of Samaria. \V And the king of Iſrael
ſaid to Eliſeus, when he had ſene them: Shal I ſtrike them, my
father? \V And he ſaid: Thou shalt not ſtrike them: for thou didſt not
take them with thy ſword, and thy bow, that thou mayſt ſtrike them: but
\SNote{By bread and water is vnderſtood ordinarie meate and drinke.
\XRef{v.~23.}}
ſet bread and water before them, that they may eate and drinke, and goe
to their maiſter. \V And a great preparation of meates was ſette before
them, and they did eate and drinke, and he diſmiſſed them, and they went
away to their maiſter, and the robbers of Syria came no more into the
Land of Iſrael. \V And it came to paſſe after theſe thinges, Benadad the
king of Syria gathered together al his armie, and went vp, and beſieged
Samaria. \V And there was a great famine in Samaria: and ſo long it was
beſieged, til the head of an aſſe was ſold for foure ſcore ſiluer
peeces, & the fourth part of a cabe of pigeons dung, for fiue ſiluer
peeces. \V And when the king of Iſrael paſſed by the wall, a certayne
woman cried out to him, ſaying: Saue me my lord king. \V Who ſayd: No,
our Lord ſaue thee: how can I ſaue thee? of the floore, or of the
preſſe? And the king ſayd to her: What ayleth thee? Who anſwered: \V
This woman ſayd to me: Geue thy ſonne, that we may eate him to day, & my
ſonne we wil eate to morrow. \V We therfore boyled my ſonne, and did
eate him. And I ſayd to her the next day: Geue thy ſonne, that we may
eate him. Who hath hid her ſonne. \V Which when the king had heard, he
rent his garmentes, and paſſed by the wal. And al the people ſaw the
hearecloth, which he ware next vpon his flesh. \V And the king ſayd:
Theſe thinges doe God to me, and theſe adde he, if the head of Eliſeus
the ſonne of Saphat shal ſtand vpon him
%%% o-0713
this day. \V But Eliſeus ſate in his houſe, and the ancientes ſate with
him. He therfore ſent a man before: and before that meſſenger came, he
ſayd to the ancientes: Doe you know that this murderers ſonne hath ſent
to cut of my head? See therfore, when the meſſenger shal come, ſhut the
doore, and ſuffer him not to enter in: for behold the ſound of his
maiſters feete is behinde him. \V Whiles he was yet ſpeaking to them,
the meſſenger appeared, which came to him. And he ſayd: Behold, this ſo
great euil is of our Lord: what shal I looke for more of our Lord?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
