%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 0784
%%% o-0703
\startChapter[
  title={Chapter 3}
  ]

\Summary{Ioram king of Iſrael
\Fix{accompianed}{accompanied}{obvious typo, fixed in other}
with the kinges of Iuda & Edom, fighteth againſt the king of Moab, for
not paying tribute according to their league. 9.~Wanting waters,
16.~Eliſeus procureth ſufficient without raine: and prophecieth victorie.
21.~The king of Moab deceiued by a viſion is ouerthrowne in the fielde,
26.~then being beſiged immolateth his firſt begotten ſonne: and the
Iſraelites leaue the ſiege.}

And Ioram the ſonne of Achab reigned ouer Iſrael in Samaria
%%% !!! SNote marked only in this. Note missing in both.
%%% \SNote{}
the eightenth yeare of Ioſaphat the king of Iuda. And he reigned twelue
yeares. \V And he did euil before our Lord, but not as his father and
mother: for he tooke away
%%% o-0704
the ſtatues of Baal, which his father had made. \V Neuertheles in the
ſinnes of Ieroboam the ſonne of Nabat, who made Iſrael to ſinne, he
ſtucke faſt, and departed not from them. \V Moreouer Meſa the king of Moab,
nouriſhed manie sheepe, and he payed to the king of Iſrael an hundred
thouſand lambes, and a hundred thouſand rammes with their fleeces. \V
And when Achab was dead, he brake the league, which he had made with the
king of Iſrael. \V Therfore king Ioram went forth that day out of
Samaria, and muſtered al Iſrael. \V And he ſent to Ioſaphat the king of
Iuda, ſaying: The king of Moab is reuolted from me, come with me againſt
him to battel. Who anſwered: I wil come vp: he that is mine, is thine: my
people, thy people: & my horſes thy horſes. \V And he ſaid: What way
shal we goe vp? but he anſwered: By the deſert of Idumea. \V Therfore
the king of Iſrael, and the king of Iuda, and the king of Edom went
forward, and compaſſed ſeuen daies iourney, neither was there water for
the armie, and the beaſtes, that folowed them. \V And the king of Iſrael
ſaid: Alas alas alas, our Lord hath gathered vs three kinges
%%% 0785
together, that he might deliuer vs into the handes of Moab. \V And
Ioſaphat ſaid: Is there not here a prophet of our Lord, that we may
beſech our Lord by him? And one of the ſeruantes of the king of Iſrael
anſwered: Here is Eliſeus the ſonne of Saphat, which powred water vpon
the handes of Elias. \V And Ioſaphat ſayd: The word of our Lord is with
him. And the king of Iſrael went downe to him, and Ioſaphat the king of
Iuda, and the king of Edom. \V And Eliſeus ſaid to the king of Iſrael:
What is to me and thee? goe to the prophetes of thy father, and thy
mother. And the king of Iſrael ſaid to him: Why hath our Lord gathered
together theſe three kinges, that he might deliuer them into the handes
of Moab? \V And Eliſeus ſaid to him: The Lord of hoſtes liueth, in whoſe
ſight I ſtand, If I did not reuerence the face of Ioſaphat the king of
Iuda, I would not certes haue harkened to thee, nor looked on thee. \V
But now bring me hither a plaier on inſtrumentes. And when the player
ſang, the hand of our Lord came vpon him, and he ſaid: \V Thus ſaith our
Lord: Make the chanel of this torrent diches and diches. \V For thus
ſaith our Lord: You shal not ſee winde, nor raine: and this chanel shal
be filled with waters, & you shal drinke, and your families, & your
beaſtes. \V And this is a ſmal thing in the ſight of our Lord: moreouer
he wil deliuer alſo Moab into your handes. \V And you ſhal ſtrike euerie
fenſed citie, and euerie principal citie, and
\SNote{It was forbid
\XRef{Deut.~20.}
to cut downe fruict trees in the land of Chanaan, which the Iſraelite
ſhould poſſeſſe, but the land of Moab perteyned not to them, and ſo it
was not prohibited to anoy that countrie vpon iuſt offence.}
shal cut downe al fruiteful trees, and shal ſtoppe vp al fountaines of
waters, and euerie goodlie field you shal couer with ſtones. \V It came
to paſſe therfore in the morning, when they vſed to offer the ſacrifice,
and behold, water came by the way of Edom, and the ground was filled
with waters. \V But al the Moabites hearing that the kinges were come vp
to fight againſt them, they called together al that were girded with a
belt vpon them, and they ſtoode in the borders. \V And ryſing early in
the morning, and the ſunne being now riſen ouer againſt the waters, the
%%% o-0705
Moabites ſaw the waters ouer againſt them red, as it were bloud, \V and
ſaid: It is the bloud of the ſword: the kinges haue fought among them
ſelues, and are ſlaine one of an other: now goe on forward to the pray,
Moab. \V And they went forward into the campe of Iſrael: Moreouer Iſrael
ryſing vp, ſtroke Moab: but they fled before them. They therfore that
had ouercome, came and ſtroke Moab, \V and
%%% 0786
deſtroyed the cities: & euerie principal field they filled euerie man
caſting ſtones: and they ſtopt vp al the fountaines of waters: and cut
downe al trees that bare fruite, ſo that there remained onlie brike
walles: and the citie was beſette of the ſlingers, and for a great part
therof was ſtrooken. \V Which when the king of Moab had ſeene, to witte,
that the enemies had preuayled, he tooke with him ſeuen hundred men that
\Fix{crew}{drew}{obvious typo, fixed in other}
ſword, to breake in vpon the king of Edom: and they could not. \V And
taking his firſt begotten ſonne, that should haue reigned for him, he
offered him an holocauſt vpon the wal: and there was great indignation
in Iſrael, and forthwith they retyred from him, and returned into their
countrie.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
