%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 0819
%%% o-0734
\startChapter[
  title={Chapter 18}
  ]

\Summary{Ezechias
\MNote{The ſecond part.

Actes of other kinges til the captiuitie of Iuda.}
deſtroyeth al places of idolatrie in Iuda, breaking
  alſo the braſen ſerpent, made by Moyſes, becauſe the people offered
  incenſe to it. 9.~The captiuitie of the ten tribes is
  repeted. 13.~Ezechias not able to reſiſt the Aſſyrians payeth much
  money to them. 17.~They neuertheles ſend forces againſt Ieruſalem,
  reproch the king, blaſpheme God, and terrifie the people.}

In the third yeare of Oſee the ſonne of Ela king of Iſrael, reigned
Ezechias the ſonne of Achaz king of Iuda. \V Fiue and twentie yeares old
was he when he began to reigne: and he reigned nine and twentie yeares
in Ieruſalem: the name of his mother was Abi the daughter of
Zacharias. \V And he did that which was good before our Lord, according
to al thinges which Dauid his father had done. \V He deſtroyed the
excelſes, and brake the
\Fix{ſtatuees}{ſtatues}{likely typo, fixed in other}
in peeces, and cut downe the groues, and brake the braſen ſerpent, which
%%% o-0735
Moyſes had made: for vntil that time the children of Iſrael
\SNote{This image of a ſerpent not only when it wrought miraculous
health, but alſo long after was worthely reſerued in memorie of the
benefite: but when the people offered ſacrifice vnto it, which is proper to
God only, good Ezechias did laudably breake it. And to ſhew that there
was no deitie in it: called it \HH{Noheſtan}, that is, a peece of
braſſe. And ſo in the Catholique Church when anie holie Relique or Image
is abuſed, it is taken away, or the errour otherwiſe corrected. See
\Cite{S.~Aug. li.~10. c.~8. ciuit.}
\Cite{Ser.~14. de verb. Apoſt.}
&
\Cite{101.~de temp.}}
burnt incenſe to it: and he called the name therof
\TNote{A peece of braſſe.}
\HH{Noheſtan}. \V He truſted in our Lord the God of Iſrael: therfore after
him there was not the like to him in al the kings of Iuda, yea neither
among them that were before him: \V and he cleaued to our Lord, and
departed not from his ſteppes, and he did his commandmentes, which our
Lord commanded Moyſes. \V Wherfore our Lord alſo was with him, and in al
things, to the which he proceded, he behaued himſelfe wiſely. He
rebelled alſo agaynſt the king of the
%%% 0820
Aſſyrians, and ſerued him not. \V He ſtroke the Philiſthians as farre as
Gaza, & al the borders, from the Towre of watchmen vnto the fenſed
citie. \V In the fourth yeare of king Ezechias, which was the ſeuenth
yeare of Oſee the ſonne of Ela the king of Iſrael, came vp Salmanaſar
the king of the Aſſyrians into Samaria, and aſſaulted it, \V and tooke
it. For after three yeares, in the ſixt yeare of Ezechias, that is, the
ninth yeare of Oſee the king of Iſrael, Samaria was taken: \V and the
king of the Aſſyrians tranſported Iſrael vnto the Aſſyrians, and placed
them in Hala, and in Habor riuers of Gozan in the cities of the Medes: \V
becauſe they heard not the voice of our Lord their God, but tranſgreſſed
his couenant: al things, that Moyſes the ſeruant of our Lord commanded,
they heard not, neither did they it. \V In the fourtenth yeare of king
Ezechias, came vp Sennacherib the king of Aſſyrians to al the fenſed
cities of Iuda: and tooke them. \V Then ſent Ezechias the king of Iuda
meſſengers to the king of the Aſſyrians into Lachis, ſaying: I haue
ſinned, retyre from me: and al that thou ſhalt put vpon me, I will
beare. Therfore the king of the Aſſyrians put a taxe vpon Ezechias the
king of Iuda, three hundred talents of ſiluer, and thirtie talents of
gold. \V And Ezechias gaue al the ſiluer that was found in the houſe of
our Lord, and in the kinges treaſures. \V At that time Ezechias brake
the doores of the temple of our Lord, and the plates of gold, which he
had faſtened on them, and gaue them to the king of the Aſſyrians. \V But
the king of the Aſſyrians ſent Tharthan, and Rabſaris, and Rabſaces from
Lachis to king Ezechias with a ſtrong powre to Ieruſalem: who when they
were come vp they came to Ieruſalem, and ſtoode beſide the conduite of
the vpper poole, which is in the way of the fullers field. \V And they
called the king: and there went out to them Eliacim the ſonne of Helcias
gouernour of the houſe and Sobna the Scribe and Ioahe the ſonne of
Aſaph, the
\TNote{Or recorder.}
regiſter. \V And Rabſaces ſayd to them: Speake to Ezechias: Thus ſayth
the great king, the king of Aſſyrians: What is this confidence, that
thou doeſt ſtay vpon? \V Perhaps thou haſt taken counſel, to prepare thy
ſelfe to battle. Wherin haſt thou confidence, that thou dareſt to
rebel? \V Doſt thou hope in Ægypt a ſtaffe of reede and broken, vpon which
if a man leane, broken into ſplinters it wil enter into his hand, and
pearce it? So is Pharao the king
%%% 0821
of Ægypt, to al that haue confidence in him. \V But if you wil ſay to
me: We haue confidence in our Lord God: is not
%%% o-0736
this he, whoſe excelſes and altars Ezechias hath taken away: and he
commanded Iuda and Ieruſalem: Before this altar ſhal you adore in
Ieruſalem? \V Now therfore paſſe to my lord the king of the Aſſyrians,
and I wil geue you two thouſand horſes, and ſee whether you be able to
haue ryders for them. \V And how can you reſiſte before one prince of
the leaſt ſeruantes of my lord? Haſt thou confidence in Ægypt for the
chariotes and horſemen? \V Why am I come vp without the wil of the Lord
to deſtroy it? The Lord ſayd to me: Goe vp to this land,
\SNote{He falſly addeth of his owne that \Emph{he should deſtroy
it}. For Iſaias prophecied the contrarie, that the Aſſyrians campe
ſhould be deſtroyed,
\XRef{Iſaie.~37.}
and ſo it came to paſſe.
\XRef{ch.~19. v.~35.}
\XRef{2.~Paral.~32.}}
and deſtroy it. \V And Eliacim the ſonne of Helcias, and Sobna, and
Ioahe ſayd to Rabſaces: We pray thee that thou ſpeake to vs thy
ſeruantes in Syryake: for we vnderſtand this tongue: and ſpeake not to
vs in the Iewes language, the people hearing it, which is vpon the
wal. \V And Rabſaces anſwered them, ſaying: What did my lord ſend me to
thee, that I ſhould ſpeake theſe wordes, and not rather to the men that
ſit vpon the wal, that they may eate their owne dung, and drinke their
vrine with you? \V Rabſaces therfore ſtood, and cryed out with a lowd
voyce in the Iewes language, and ſayd: Heare ye the
wordes of the great king, the king of the Aſſyrians. \V Thus ſayth the
king: Let not Ezechias ſeduce you: for he shal not be able to deliuer
you out of my hand. \V Neither let him geue you confidence vpon the
Lord, ſaying: Our Lord deliuering wil deliuer vs, and this citie shal
not be geuen into the hand of the king of the Aſſyrians. \V Doe not
heare Ezechias. For thus ſayth the king of the Aſſyrians: Doe with me
that which is profitable for you, and come forth to me: and euery man
ſhal eate of his vineyard, and of his figge tree: and you shal drinke
waters of your owne ceſternes, \V til I come, and tranſporte you into a
land, that is like to your land, into a fruiteful land, and plentiful of
wyne, a land of bread and of vineyardes, a land of oliuetes, and of oyle
and honie, and you shal liue, and shal not die. Heare not Ezechias, who
deceiueth you, ſaying: Our Lord wil deliuer vs. \V
\SNote{Paganes and Heretikes are fooliſh & impudent to compare their
falſe goddes and phancies with God almightie and Catholique Religion.}
Did the goddes of Nations deliuer their land from the hand of the king
of Aſſyrians? \V Where is the god of Emath & Arphad? Where is the god of
Sepharuaim, of Ana, and Aua? did they deliuer Samaria out of my hand? \V
What are they among al the goddes of nations, which
%%% 0822
haue deliuered their countrey out of my hand, that the Lord can deliuer
Ieruſalem out of my hand? \V The people therfore held their peace, and
did not anſwer him any thing: for they had receiued the kings
commandement that they should not anſwer him. \V And Eliacim the ſonne
of Helcias, gouernour of the houſe, and Sobna the ſcribe, and Ioahe the
ſonne of Aſaph regiſter came to Ezechias, their garments rent, and told
him the wordes of Rabſaces.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
