%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 0811
%%% o-0728
\startChapter[
  title={Chapter 15}
  ]

\Summary{Azarias beginneth wel, 4.~afterward (for offering incenſe on
the altar
\XRef{2.~Paralip.~26.)}
is ſtrooken with leproſie, caſt out of the temple, and from
conuerſation with the people, his ſonne Ioathan ruling the
kingdom. 8.~Sellum
%%% 0812
killeth Zacharias king of Iſrael and reigneth in his place. 14.~After
one moneth Manahem killeth Sellum & reigneth alſo wickedly. 19.~Maketh
league with the Syrians, 22.~dieth and his ſonne Phaceia
reigneth. 25.~Phacee killeth him, and reigneth. 29.~The Aſsyrians
ſpoyle the countrie, and carie away manie captiues. 30.~Oſee killeth
Phacee and reigneth. 32.~In the meane time Ioathan reigning in Iuda is
infeſted with enemies, 38.~dieth, and his ſonne Achaz reigneth.}

In the ſeuen and twenteth yeare of Ieroboam the king of Iſrael reigned
\SNote{Otherwiſe called Ozias.
\XRef{2.~Paral.~26.}
\XRef{Mat.~1.}}
Azarias the ſonne of Amaſias the king of Iuda. \V He was ſixtene yeares
old, when he began to reigne, and two and fiftie yeares he reigned in
Ieruſalem: the name of his mother was Iechelia of Ieruſalem. \V And he
did that which was liked before our Lord, according to al thinges that
his father Amaſias did. \V But the excelſes he deſtroyed not: as yet the
people ſacrificed, and burnt incenſe in the excelſes. \V And our Lord
ſtroke the king, and he was
\SNote{This puniſhment was inflicted vpon him, for his preſumption to
offer incẽſe on the altar.
\XRef{2.~Paralip.~26.}}
a leper vntil the day of his death, and he dwelt in a free houſe a part:
but Ioathan the kinges ſonne gouerned the palace, and iudged the people
of the land. \V But the reſt of the wordes of Azarias, and al that he
did, are not theſe thinges writen in the
%%% !!! XRef ?
Booke of the wordes of the daies of the kinges of Iuda? \V And Azarias
ſlept with his fathers: and they buried him with his anceſtors
\SNote{He was buried honorably \Emph{in the Citie of Dauid}, that is,
nere to the walles, but in the filde, becauſe he was a leper euen to his
death.
\XRef{2.~Paral.~26.}}
in the Citie of Dauid, and Ioathan his ſonne reigned for him. \V In the
eight and thirteth yeare of Azarias the king of Iuda, reigned Zacharias
the ſonne of Ieroboam ouer Iſrael in Samaria ſixe monethes: \V and he
did that which is euil before our Lord, as his fathers had done: he
departed not from the ſinnes of Ieroboam the ſonne of Nabat, who made
Iſrael to ſinne. \V And Sellum the ſonne of Iabes conſpired againſt him:
and ſmote him openly, killed him, and reigned for him. \V But the reſt
of the wordes of Zacharias, are not theſe writen in the
%%% !!! XRef ?
Booke of the wordes of the dayes of the kinges of Iſrael? \V This is the
word of our Lord, which he ſpake to Iehu, ſaying: Thy children ſhal ſit
vnto the fourth generation vpon the throne of Iſrael. And ſo it came to
paſſe. \V Sellum the
\Fix{ſone}{ſonne}{obvious typo, fixed in other}
of Iabes reigned the nine and thirteth yeare of Azarias the king of
Iuda: and reigned one moneth in Samaria. \V And Manahem the ſonne of
Gadi went vp from Therſa: and he came into Samaria,
%%% o-0729
and ſmote Sellum the ſonne of Iabes in Samaria,
%%% 0813
and ſlew him, & reigned for him. \V But the reſt of the wordes of
Sellum, & his conſpiracie, wherwith he wrought treaſon, are not theſe
thinges written in the
%%% !!! XRef ?
Booke of the wordes of the dayes of the kinges of Iſrael? \V Then
Manahem ſtroke Thapſa and al that were in it and the borders therof from
Therſa, for they would not open vnto him: and he ſlew al wemen therof
that were withchilde, and ripped them. \V In the nine and thirteth yeare
of Azarias the king of Iuda, reigned Manahem the ſonne of Gadi ouer
Iſrael ten yeares in Samaria. \V And he did that which was euil before
our Lord: he departed not from the ſinnes of Ieroboam the ſonne of
Nabat, who made Iſrael to ſinne al his dayes. \V Phul alſo the king of
the Aſſyrians came into the land, and Manahem gaue to Phul a thouſand
talentes of ſiluer, that he ſhould ayde him, & eſtabliſh his kingdom. \V
And Manahem put a taxe of ſiluer vpon Iſrael, on them that were mightie
and riche to geue the king of the Aſſyrians, fiftie ſicles of ſiluer euerie man:
and the king of the Aſſyrians returned, and taried not in the land. \V
But the reſt of the wordes of Manahem, and al that he did, are not theſe
thinges writen in the
%%% !!! XRef ?
Booke of the wordes of the daies of the kinges of Iſrael? \V And Manahem
ſlept with his fathers: and Phaceia his ſonne reigned for him. \V In the
fifteth yeare of Azarias the king of Iuda reigned Phaceia the ſonne of
Manahem ouer Iſrael in
\Fix{Samria}{Samaria}{obvious typo, fixed in other}
two yeares: \V and he did that which was euil before our Lord: he
departed not from the ſinnes of Ieroboam the ſonne of Nabat, who made
Iſrael to ſinne. \V And Phacee the ſonne of Romelia, his captaine
conſpired againſt him, and ſtroke him in Samaria, in the towre of the
kinges houſe beſide Argob and beſide Arie, and with him fiftie men of
the children of the Galaadites, and he ſlewe him, and reigned for
him. \V But the reſt of the wordes of Phaceia & al that he did, are not
theſe things written in the
%%% !!! XRef ?
Booke of the wordes of the daies of the kings of Iſrael? \V In the two
and fifteth
\Fix{yeares}{yeare}{obvious typo, fixed in other}
of Azarias the king of Iuda reigned Phacee the
\Fix{ſonne}{ſonne of}{obvious typo, fixed in other}
Romelia ouer Iſrael in Samaria twentie yeares. \V And he did that which
was euil before our Lord: he departed not from the ſinnes of Ieroboam
the ſonne of Nabat, who made Iſrael to ſinne. \V In the daies of Phacee
the king of Iſrael came Theglathphalaſar the king of Aſſur, &
\SNote{Some men of Ruben, Gad, Manaſſes, and Nepthali, were caried
captiues into Aſſyria.}
tooke Aion, and Abel the houſe of Maachar and Ianoe, and Cedes, and
Aſor, and Galaad, and Galilee, and al
%%% 0814
the land of Nepthali: and tranſported them vnto the Aſſyrians. \V And
Oſee the ſonne of Ela conſpired, and lay in wayte againſt Phacee the
ſonne of Romelia, and ſtroke him, and ſlew him: and he reigned for him
the twentith yeare of Ioatham the ſonne of
\SNote{Otherwiſe called Azarias.
\XRef{v.~1.}}
Ozias. \V But the reſt of the wordes of Phacee, and al that he did, are
not theſe things written in the
%%% !!! XRef ?
Booke of the wordes of the dayes of the kings of Iſrael? \V In the
ſecond yeare of Phacee the ſonne of Romelia the king of Iſrael, reigned
Ioatham the ſonne of Ozias the king of Iuda. \V Fiue and twentie yeares
%%% o-0730
old was he when he began to reigne, and ſixtene yeares he reigned in
Ieruſalem: the name of his mother was Ieruſa, the daughter of Sadoc. \V
And he did that which was liked before our Lord: according to al things,
which Ozias his father had done, did he worke. \V Howbeit the excelſes
he tooke not away: as yet the people immolated, and burnt incenſe in the
excelſes. He built the higheſt gate of the houſe of our Lord. \V But the
reſt of the wordes of Ioatham, and al that he did, are not theſe things
written in the
%%% !!! XRef ?
Booke of the words of the dayes of the kings of Iuda? \V In thoſe dayes
our Lord began to ſend into Iuda Raſin the king of Syria, and Phacee the
ſonne of Romelia. \V And Ioatham ſlept with his fathers, and was buried
with them in the Citie of Dauid his father, and Achaz his ſonne reigned
for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
