%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 0795
%%% o-0713
\startChapter[
  title={Chapter 7}
  ]

\Summary{Eliſeus prophecieth plentie of corne the next day, and death to
  a chief man that wil not beleue it. 3.~Foure Lepers going to yeld
  themſelues to the Syrians, 6.~who by Gods prouidence are frighted and
  fled away, 9.~bring newes therof to Samaria, 12.~which by trial is
  found true. 16.~And ſo there is plentie of corne, and the incredulous
  nobleman is trod to death, with preſſe of multitude in the gate, as
  the prophet fortold.}

And Eliſeus ſayd: Heare ye the word of our Lord: Thus ſayth our Lord: At
this time to morow a bushel of floure shal be at one ſtater, and two
bushels of barley at one ſtater, in the gate of Samaria. \V One of the
Dukes, vpon whoſe hand the king leaned,
\Fix{anſwearing}{anſwering}{likely typo, fixed in other}
the man of God, ſayd: If our Lord shal make fludgates in heauen,
\SNote{Diſcourſe of mans reaſon can not reach to the powre of God, who
can do al that he wil, and wil doe al that he ſaith: therfore the
incredulous are iuſtly puniſhed.
\XRef{v.~20.}}
can that poſſibly be which thou ſpeakeſt? Who ſayd: Thou ſhalt ſee it
with thine eies, and shalt not eate therof. \V There were therfore foure
men lepers, beſide the entrance of the gate: who ſayd one to an other:
What meane we to be here til we die? \V Whether we enter into the citie,
we shal die for famine: or whether we tarie here, die we muſt: come
therfore, and let vs runne away to the campe of Syria. If they ſpare vs,
we shal liue: but if they wil kil vs, we shal die neuertheleſſe. \V They
aroſe therfore in the euening, to come to the campe of Syria. And when
they were come to the beginning of the campe of Syria, they found no man
there. \V For our Lord had made them in the campe of Syria to heare the
ſound of chariotes, and horſes, and of a verie great armie: and they
ſayd one to an other: Behold the king of Iſrael hath for wages hyred
agaynſt vs the kinges of the Hetheites, and of the Ægyptians, and they
are come vpon vs. \V They aroſe therfore, and fled in the darke, and
leaft their tentes, and their horſes and aſſes in the campe, and fled
deſirous to ſaue their liues only. \V Therfore when theſe lepers were
come to the beginning of the campe, they entered into one tabernacle,
and did eate and drinke: and they tooke thence ſiluer, and gold, and
rayment, and went, and hid it: againe they returned to an other
%%% o-0714
tabernacle, and from thence likewiſe taking away they hid it. \V And
they ſaid one to an other: We doe not wel: for this is a day of good
tydinges. If we ſhal hold our peace, and wil not tel vntil morning, we
ſhal be blamed of a heinous offence: Come, let vs goe, and make report in
the kinges
%%% 0796
court. \V And when they were come to the gate of the citie, they told
them ſaying: We went to the campe of Syria, and found no man there, but
horſes, and aſſes tyed, and the tentes pitched. \V The porters therfore
went, and told it to the king within his palace. \V Who aroſe in the
night, & ſayd to his ſeruants: I tel you what the Syrians haue done to
vs: They know that we ſuffer great famine, and therfore they are gone
out of the campe and lie hid in the fieldes, ſaying: When they ſhal come
forth out of the citie, we wil take them aliue, and then we may enter
into the citie. \V But one of his ſeruantes anſwered: Let vs take fiue
horſes that are remaining in the citie (becauſe they onlie are in the
whole multitude of Iſrael for the other are conſumed) and ſending, we
may trie. \V They brought therfore two horſes, and the king ſent into
the campe of the Syrians, ſaying: Goe ye, and ſee. \V Who went after
them as far as Iordan: and behold al the way was ful of rayment, and
veſſels, which the Syrians had caſt away, when they were amaſed, and the
meſſengers returning told the king. \V And the people going forth
ſpoyled the campe of Syria: and a buſhel of floure became at one ſtater,
and two buſhels of barley at one ſtater, according to the word of our
Lord. \V Moreouer the king appoynted that duke on whoſe hand he leaned,
to ſtand at the gate: whom the multitude trode in the entrance of the
gate, & he died, according as the man of God had ſpoken, when the king
came downe to him. \V And it came to paſſe according to the word of the
man of God, which he ſpake to the king, when he ſayd: Two bushels of
barley shal be at one ſtater, and a bushel of floure at one ſtater, this
verie tyme to morow
\Fix{it}{in}{obvious typo, fixed in other}
the gate of Samaria: \V when that duke anſwered the man of God, and ſayd:
Although our Lord would make fludgates in the heauen, can this be done
which thou ſpeakeſt? And he ſaid to him: Thou shalt ſee with thine eies,
and shal not eate therof. \V It chanced therfore to him as it was
foretold, and the people trode him in the gate, and he died.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
