%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 0825
%%% o-0739
\startChapter[
  title={Chapter 20}
  ]

\Summary{Ezechias being ſick is told by Iſaias that he shal die: but
  praying to God obtayneth fiftenne yeares longer life, 8.~and in
  confirmation therof receiueth a ſigne in Achaz dial, returning back
  tenne lines. 12.~To the Aſſyrians bringing him preſents, he sheweth
  al his treaſures. 16.~Which Iſaias reprouing prophecieth the
  captiuitie of Iuda. 20.~Ezechias dieth, and his ſonne Manaſses
  reigneth.}

In thoſe dayes Ezechias was ſick euen to death: and Iſaie the ſonne of
Amos the prophete, came and ſayd to him: Thus ſayth our Lord God: Take
order with thy houſe, for thou ſhalt die, and ſhalt not liue. \V Who
turned his face to the wal, and prayed our Lord, ſaying: \V I beſech
thee Lord, remember I pray thee how I haue walked before thee in truth,
and in a perfect hart, and haue done that which is liked before
thee. Ezechias therfore wept with great weeping. \V And before Iſaie was
gone out of the middes of the court, the word of our Lord came to him,
ſaying: \V Returne, and tel Ezechias the prince of my people: Thus ſayth
our Lord the God of Dauid thy father: I haue heard thy prayer, and ſeene
thy teares: and behold I haue healed thee, the third day thou shalt goe
vp to the temple of the Lord. \V And I wil adde to thy daies fiftene
yeares: yea & out of the hand of the king of Aſſyrians I wil deliuer
thee, and this citie, & I wil protect this citie for my ſake, and for
Dauid my ſeruant. \V And Iſaie ſaid: Fetch me a bunch of figges. Which
when they had brought, and had layd it vpon his ſore, he was cured. \V And
Ezechias had ſayd to Iſaie: What ſigne shal there be, that our Lord wil
heale me, and that I shal
%%% o-0740
goe vp the third day to the temple of our Lord? \V To whom Iſaie ſayd:
This shal be the ſigne from our Lord, that our Lord wil doe the word,
which he hath ſpoken: Wilt thou that the shadow goe forward ten lines,
or that it goe backe ſoe many degrees. \V And Ezechias ſayd: It is an
eaſie matter for the shadow to goe forward ten lines, neither wil I that
this be done, but that it returne back ten degrees. \V Iſaie therfore
the prophet inuocated our Lord, and brought backe the ſhadow by
\SNote{If theſe tenne lines importe ſo manie houres, then the dial going
forwardes againe, by like degrees, this day was increaſed by twentie
houres, and ſoe was longer then that in which Ioſue procured ſtay of the
ſunne the ſpace of one day, to witte of twelue houres.
\XRef{Ioſue.~10.}
as S.~Dyoniſe thinketh.
\Cite{Epiſt. ad Policarp.}
See
\Cite{Gloſſa ord. in Ioſue.}}
the lines, by the which it was now gone downe in the dial of Achaz,
backward ten degrees. \V In that time Berodach Baladan, the ſonne of
Baladan, the king of the Babilonians ſent letters and giftes to
Ezechias: for he had heard that Ezechias had bene ſicke.
%%% 0826
\V And Ezechias reioyſed in their coming, and he shewed them the houſe
of aromatical ſpices, and gold and ſiluer, and diuerſe precious, odours,
oyntementes alſo, and the houſe of his veſſels, and al that he had in
his treaſures. There was not any thing which Ezechias shewed them not in
his houſe, and in al his powre. \V And Iſaie the prophete came to king
Ezechias, and ſayd to him: What ſayd theſe men? or from whence came they
to thee? To whom Ezechias ſaid: From a far countrie they came to me out
of Babylon. \V But he anſwered: What ſaw they in thy houſe? Ezechias
ſayd: They ſaw al things whatſoeuer are in my houſe: there is nothing
that I haue not shewed them in my treaſures. \V Iſaie therfore ſaid to
Ezechias, heare the word of our Lord: \V Behold the daies shal come, &
al things shal be taken away, that are in thy houſe, and that thy
fathers haue layd vp vntil this day, into Babylon: there shal not anie
thing remayne, ſayth our Lord. \V Yea of the children alſo that come
forth of thee, whom thou shalt beget shal be taken away, and they shal be
eunuches in the palace of the king of Babylon. \V Ezechias ſaid to
Iſaie: The word of our Lord which thou haſt ſpoken is good: be there
peace and truth in my daies. \V But the reſt of the wordes of Ezechias,
and al his ſtrength, and how he made a poole, and a conduite, and
brought waters in to the citie, are not theſe things writen in the Booke
of the wordes of the daies of the kinges of Iuda? \V And Ezechias ſlept
with his fathers, and Manaſſes his ſonne reigned for him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
