%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 0807
%%% o-0724
\startChapter[
  title={Chapter 13}
  ]

\Summary{Ioachaz King of Iſrael is afflicted by the Syrians, 4.~and
  deliuered. 6.~Yet deſtroyeth not idolatrie: 8.~dieth, and his ſonne
  Ioas foloweth his euil ſteppes. 14.~Eliſeus being ſicke, willeth the
  king to shoote, 18.~and to ſtrike the
  %%% 0808
  earth, who ſtriking thriſe, is told that he shal thriſe ſtrike the
  Syrians. 20.~Eliſeus dieth, and a dead man is reuiued touching his
  bones. 22.~Ioas recouereth that which the Syrians had taken from
  Iſrael.}

In the three and twenteth yeare of Ioas the ſonne of Ochozias the king
of Iuda, reigned Ioachaz the ſonne of Iehu ouer Iſrael in Samaria,
ſeuentene yeares. \V And he did euil before our Lord, and folowed the
ſinnes of Ieroboam the ſonne of Nabat, who made Iſrael to ſinne, and
declined not from them. \V And the furie of our Lord was wrath agaynſt
Iſrael, and he deliuered them into the hand of Hazael the king of Syria,
and into the hand of Benadad the ſonne of Hazael, alwaies. \V But
Ioachaz beſought the face of our Lord, and our Lord heard him: for he
ſaw the diſtreſſe of Iſrael, that
%%% !!! following is missing in other
the king of Syria had broken
%%% o-0725
them: \V and our Lord gaue a ſauiour to Iſrael, and they were deliuered
out of the hand of the king of Syria: and the children of Iſrael dwelt
in their tabernacles as yeſterday and the day before. \V But yet they
departed not from the ſinnes of Ieroboam who made Iſrael to ſinne, but
they walked in them: for the groue alſo remayned in Samaria. \V And
there were leaft to Ioachaz of the people but fiftie horſemen, and ten
chariotes, and ten thouſand footmen: for the king of Syria had ſlaine
them, and had brought them as duſt by threshing in the barne floore. \V
But the reſt of the wordes of Ioachaz, and al that he did, and his
ſtrength, are not theſe thinges writen in the
%%% !!! XRef ?
Booke of the wordes of the daies of the kinges of Iſrael? \V And Ioachaz
ſlept with his fathers, & they buried him in Samaria: and Ioas his ſonne
reigned for him. \V In the ſeuen and thirteth yeare of Ioas the king of
Iuda reigned Ioas the ſonne of Ioachaz ouer Iſrael in Samaria ſixtene
yeares, \V and he did that which is euil in the ſight of our Lord, he
declined not from al the ſinnes of Ieroboam the ſonne of Nabat, who made
Iſrael to ſinne, but he walked in them. \V But the reſt of the wordes of
Ioas, and al that he did, and his ſtrength, how he fought agaynſt
Amaſias the king of Iuda, are not theſe thinges writen in the
%%% !!! XRef ?
Booke of the wordes of the dayes of the kinges of Iſrael? \V And Ioas
ſlept with his fathers: but Ieroboam ſate vpon his throne. Moreouer
Ioas, was buried in Samaria with the kinges of Iſrael. \V And Eliſeus
was ſicke of an infirmitie, wherof alſo he died: and Ioas the king of
Iſrael went downe
%%% 0809
to him, and wept before him, and ſayd: My father, my father, the
chariote of Iſrael and the garder therof. \V And Eliſeus ſayd to him:
fetch a bow and arrowes. And when he had brought him a bow, and
arrowes, \V he ſayd to the king of Iſrael: Put thy hand vpon the
bow. And when he had put his hand, Eliſeus put his handes ouer the
kinges handes, \V and ſayd: Open the eaſt window. And when he had opened
it, Eliſeus ſayd: Shote an arrow. And he shot. And Eliſeus ſayd: The
arrow of the ſaluation of our Lord, and the arrow of ſaluation agaynſt
Syria: and thou shalt ſtrike Syria in Aphec, til thou conſume it. \V And
he ſayd: Take vp the arrowes. Who when he had taken them agayne, he ſayd
to him: Strike the earth with a iauelin. And when he had ſtriken three
times, and ſtood ſtil, \V the man of God was angrie with him, & ſayd:
\SNote{It was reueiled to the prophet that ſo often as the king ſhould
ſtrike the earth, ſo often he ſhould haue victories againſt the Syrians,
but not how often he would ſtrike on the earth.}
If thou hadſt ſtriken fiue or ſix or ſeuen times, thou hadſt ſtriken
Syria euen to deſtruction: but now three times ſhalt thou ſtrike it. \V
Eliſeus therfore died, and they buried him. And the rouers of Moab came
into the land the ſame yeare. \V And certayne perſons burying a man, ſaw
the rouers, and threw the bodie in the ſepulchre of Eliſeus. Which when
it had touched the bones of Eliſeus, the man reuiued, and ſtood vpon his
feete. \V Hazael therfore the king of Syria afflicted Iſrael al the
daies of Ioachaz: \V and our Lord had mercie on them, and returned to
them for his couenant, which he had with Abraham, and Iſaac, and
Iacob: and he would not deſtroy them, nor vtterly caſt them away, vntil
this preſent time.
%%% o-0726
\V And Hazael the king of Syria died, and Benadad his ſonne reigned for
him. \V Moreouer Ioas the ſonne of Ioachaz, tooke the 
cities out of the hand of Benadad the ſonne of Hazael, which he had
taken out of the hand of Ioachaz his father by the right of warre, three
times did Ioas ſtrike him, and he deliuered the cities to Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
