%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Four Kings
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 0782
%%% o-0702
\startChapter[
  title={Chapter 2}
  ]

\Summary{Eliſeus wil not part from Elias. 7.~Fiftie diſciples folow them
  to Iordan. 8.~The water is diuided by Elias cloke, and they two paſse
  the drie chanel. 9.~Elias is aſsumpted in a firie chariote, and his
  duble ſpirite is geuen to Eliſeus, 13.~who returning by like
  miraculous meanes our Iordan, the diſciples receiue and honour him as
  their religious Superiour. 16.~They ſeeke Elias, but find him
  not. 19.~Eliſeus amendeth the waters by caſting in ſalt. 23.~Boyes are
  torne by beares for mocking Eliſeus.}

And it came to paſſe, when our Lord would take
%%% !!! SNote marked only in other
\SNote{VVither Elias was carried being taken vp into the ayre, is
vncertain: but certaine that he yet liueth, and muſt dye. As S.~Auguſtin
teacheth.
\Cite{li.~9. c.~6. de Gen. ad lit.}
See
\XRef{Annot. in Gen.~6.}
&
\XRef{Apocalip.~11.}}
vp Elias by a hurle winde into
\TNote{ayre, or lovver heauen.}
heauen, Elias and Eliſeus went from Galgal. \V And Elias ſayd to
Eliſeus: Sitte here, becauſe our Lord hath ſent me as
\Fix{fur}{far}{likely typo, fixed in other}
as Bethel. To whom Eliſeus ſayd: Our Lord liueth, and thy ſoule liueth,
I wil not leaue thee. And when he was come downe to Bethel, \V the
children of the prophetes, that were in Bethel, went forth to Eliſeus,
and ſayd to him: Doeſt thou know, that this day our Lord wil take thy
maiſter from thee? Who anſwered: I alſo know it: hold your peace. \V And
Elias ſayd to Eliſeus: Sitte here, becauſe our Lord hath ſent me into
Iericho. And he ſayd: Our Lord liueth, & thy ſoule liueth, I wil not
leaue thee. And when they were come to Iericho, \V the children of the
prophetes, that were in Iericho, came to Eliſeus, and ſayd to him: Doeſt
thou know that this day our Lord wil take away thy maiſter from thee?
And he ſayd: I alſo know it, hold your peace. \V And Elias ſayd to him:
Sitte here, becauſe our Lord hath ſent me as farre as Iordan.
Who ſayd: Our Lord liueth, and thy ſoule liueth, I wil
not leaue thee: They went therfore both together, \V & fiftie men of the
children of the prophetes
%%% 0783
folowed them, who alſo ſtood ouer agaynſt them, farre of: but they two
ſtood vpon Iordan. \V And Elias tooke his mantel, and folded it
together, and ſmote the waters, which were diuided into two partes, and
they both paſſed ouer by the drie land. \V And when they were ouer,
Elias ſayd to Eliſeus: Aske what thou wilt haue me to doe for thee,
before I be taken from thee. And Eliſeus ſayd: I beſech thee that in me
may be
\SNote{He deſired not duble ſpirite to his maſter, but the duble ſpirite
of prophecying, and of working miracles, which Elias had.
\XRef{v.~15.}}
thy duble ſpirit. \V Who anſwered: Thou haſt asked a hard thing:
Neuertheleſſe if thou ſee me, when I ſhal be taken from thee, thou ſhalt
haue that thou haſt asked: but if thou ſee me not, thou shalt not haue
it. \V And when they went forward, and going talked together, behold a
fyrie chariote, and fyrie horſes parted them two aſunder: and Elias
aſcended by a hurle winde into heauen. \V And Eliſeus ſaw him, and
cried: My father, my father, the chariote of Iſrael, and the guider
therof. And he ſaw him no more: and he tooke his garmentes, and rent
them in two peeces. \V And he tooke
%%% o-0703
vp the mantel of Elias, that was fallen to him: and returning he ſtood
vpon the banke of Iordan, \V and with the mantel of Elias, that fel
downe to him, he ſmote the waters, and they were not diuided. And he
ſayd: Where is the God of Elias, now alſo? and he ſmote the waters, and
they were diuided this way and that way, and Eliſeus paſſed ouer. \V And
the children of the prophetes, that were in Iericho, ouer agaynſt him
ſeing him, ſayd: The ſpirit of Elias hath reſted vpon Eliſeus. And
coming to meete him,
\SNote{They adored him for his holines, and becauſe God had geuen him
the ſpirit of ſo great a prophete, not for
\Fix{wordlie,}{worldlie,}{obvious typo, same in both}
but ſpiritual excellencie, & therfore not with ciuil but religious
honour.}
adored him flatte to the ground, \V and they ſayd to him: Behold, there
are with thy ſeruantes fiftie ſtrong men, that can goe, and ſeeke thy
maiſter, leſt perhaps the ſpirit of our Lord hath taken him, and caſt
him vpon one of the mountaines, or into one of the
\Fix{valleis.}{valleyes.}{likely typo, fixed in other}
Who ſayd:
Send not. \V And they forced him, til he agreed, and ſayd: Send. And
they ſent fiftie men. Who when they had ſought three daies, found
not. \V And they returned to him: but he dwelt in Iericho, and he ſayd
to them: Did I not ſay to you: Send not? \V The men alſo of the citie
ſayd to Eliſeus: Behold the habitation of this citie is very good, as
thy ſelfe my lord perceiueſt: but the waters are very il, and the ground
barren. \V But he ſayd: Fetch me a new veſſel, and put ſalt into
it. Which when they had brought, \V going out to the fountayne of the
waters, he caſt ſalt into it, and ſayd: Thus ſayth our Lord: I haue
amended theſe waters, and
%%% 0784
death ſhal no more be in them, nor barrenneſſe. \V The waters therfore
were amended vntil this day, according to the word of Eliſeus, which he
ſpake. \V And from thence he went vp into Bethel: & when he went vp by
the way, little laddes came forth out of the citie, and mocked him,
ſaying: Come vp baldhead, come vp baldhead. \V Who when he had looked
backe, he ſaw them, & curſed them in the name of our Lord: and two
beares came forth out of the foreſt, & tore of them two and fourtie
boyes. \V And from thence he went into the mount of Carmel, and from
thence he turned into Samaria.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
