%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 22
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-22


\project douay-rheims


%%% 0095
%%% o-0089
\startChapter[
  title={Chapter 22}
  ]

\Summary{The faith and obedience of Abraham is proued in his redines to
  ſacrifice his ſonne Iſaac. 11.~He is ſtayed from the act by an Angel.
  16.~Former promiſes are renewed to him. 20.~And his brother Nachor
  hath alſo much iſſue.}

VVhich things being done, God
\CNote{Iudith.~8. Heb.~11.}
\LNote{Tempted.}{God
\MNote{God tempteth not to euil, but by experience maketh knowen what
vertue is in men.}
tempteth none to euil (as
\CNote{Iac.~1,~13.}
S.~Iames teacheth) but by trial and experiment
maketh men knowen to the world, as here Abrahams faith and ſingular
obedience were manifeſted, when he doubted not
\CNote{Heb.~11,~19.}
\Emph{To offer his only
begotten in Sacrifice, accounting that God is able to raiſe vp euen from
the dead.}
\CNote{S.~Aug. li.~16. c.~32. ciuit. Theod. q.~72. in Gen.}
VVherupon he receiued his ſonne againe \Emph{in parable},
that is, in figure and myſterie of Chriſt dead and aliue againe.
\MNote{Iſaac figured
\Fix{Chiſts}{Chriſts}{obvious typo, fixed in other}
diuinitie, the Rãme his humanitie.}
Iſaac alſo in this action ſignified the diuinitie of Chriſt which
ſuffered not, and the ramme among the briers figured his humanitie, that
ſhould be offered in Sacrifice to God.}
tempted Abraham, and ſaid to him: Abraham, Abraham. But he anſwered:
Here I am. \V He ſaid to him:
\SNote{This hiſtorie is clere, and a moſt notorious example of perfect
obedience.}
Take thy only begotten ſonne, whom thou loueſt, Iſaac, and goe into the
Land of viſion: and there thou ſhalt offer him for an holocauſt vpon one
of the mountaines which I wil ſhew thee. \V Therfore Abraham riſing vp
in the night, ſadled his aſſe: taking with him two young men, and Iſaac
his ſonne: and when he had cut wood for the holocauſt, he went his way
to the place which God had commanded him. \V And the third day, lifting
vp his eyes, he ſawe the place afarre of: \V and he ſaid to his young
men: Tarie you here with the aſſe: I and the boy going with ſpede as
farre as yonder, after we haue adored, wil returne to you. \V He tooke
alſo the wood of the holocauſt, and laid it vpon Iſaac his ſonne: and
him ſelfe caried in his hands fire and a ſword. And as they went on
together, \V Iſaac ſaid to his father: My father. And he anſwered: What
wilt thou ſonne? Behold, ſaith he, fire and wood: where is the victime
of the holocauſt? \V And Abraham ſaid: God wil prouide vnto himſelfe the
victime of the holocauſt, my ſonne. The went on therfore together: \V
and they came to the place which God had ſhewed him, wherin he builed
an altar, and laid the wood in order vpon it: and when he had bound
Iſaac his ſonne, he laid him on the altar vpon the pile of wood. \V And
he ſtretched forth his hand, and caught the ſword, for to ſacrifice his
ſonne. \V And behold an Angel of our Lord from heauen cried, ſaying:
Abraham, Abraham. Who anſwered: Here I am. \V And he ſaid to him:
Stretch not forth thy hand vpon the boy, neither doe thou any thing to
him: now haue I knowen that thou feareſt God, and haſt not ſpared thine
onlie begotten ſonne for my ſake. \V Abraham lifted vp his eyes, and ſaw
behind his backe a ramme amongſt the briers
%%% 0096
ſticking faſt by the hornes, which he tooke
%%% o-0090
and offered an holocauſt in ſtede of his ſonne. \V And he called the
name of that place,
\SNote{It is a gratful and religious thing, by naming of places, to
cõſerue the memorie of Gods benefites, that poſteritie may know them.
\Cite{S.~Chriſoſt. ho.~48. in Gen.}}
Our Lord ſeeth. Wherupon euen to this day it is ſaid, In the mountaine
our Lord wil ſee. \V And the Angel of our Lord called Abraham the ſecond
time from heauen, ſaying: \V By my owne ſelfe haue I ſworne, ſaith the
Lord: becauſe thou haſt done this thing, and haſt not ſpared thine onlie
begotten ſonne for my ſake: \V I wil bleſſe thee, and I wil multiplie
thy ſede as the ſtarres of heauen, and as the ſand that is by the ſea
ſhore: thy ſede ſhal poſſeſſe the gates of his enemies, \V and \Sc{in
thy sede shal be blessed al the nations of the earth}, becauſe thou haſt
obeyed my voice. \V Abraham returned to his young men, and they went to
Berſabee together, and he dwelt there. \V Theſe things ſo being done, it
was
\SNote{Nachors progenie is here mentioned to ſhew whence Rebecca came,
whom Iſaac maried.}
told Abraham that Melca alſo had borne children to Nachor his brother,
\V Hus the firſt begotten, & Buz his brother, and Camuel the father of
the Sirians, \V and Caſed, and Azau. Pheldas 
alſo & Iedlaph, \V and Bathuel, of whom was borne Rebecca: theſe eight
did Melcha beare, to Nachor Abrahams brother. \V And his concubine, named
Roma, bare Tabee, and Gaham, and Tahas, and Maacha.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
