%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 0037
%%% o-0034
\startChapter[
  title={Chapter 5}
  ]

\Summary{The progenie of Adam, & number of their yeares (vvith the death
  of the reſt, & tranſlation of Enoch) in the line of Seth, to Noe & his
  three ſonnes.}

This is the booke of the generation of Adam. In the day, when God
created man,
\CNote{Sap.~2,~24. Eccli.~17,~1.}
to the likenes of God made he him. \V Male and female
created he them; and bleſſed them: and called their name Adam, in the
day when they were created. \V And Adam liued a hundred and thirtie
yeares; and begat to his owne image and likenes, and called his name
Seth. \V And the dayes of Adam, after he begat Seth, came to eight
hundred yeares: and he
\LNote{Begate ſonnes and daughters.}{Moyſes
\CNote{S.~Aug. li.~15. ciuit. c.~20.}
\MNote{The cõtinual ſucceſſion of Gods Church, and interruption of other
communities.}
in this genealogie reciteth not alwayes the firſt begotten, nor the
whole progenie by their names (for then he ſhould haue repeated Cain and
Abel, and haue named many others) but thoſe onlie by whom the Church of
God continued, ſignifying the reſt in general, whoſe ſucceſſion was cut
of by the floud.}
begat ſonnes
%%% 0038
and daughters. \V And al the time that Adam liued, came to nine hundred
and thirtie yeares,
\LNote{And he died.}{By this Gods word is verified ſaying, that Adam
ſhould dye, if he ſhould eate of the forbidden tree. And the diuel is
proued a lyer, ſaying, they ſhould not dye.
\MNote{How mã died the day that he ſinned.}
It is alſo moſt true that Adam dyed that day in which he did eate. For
he began that very day to decline to death; and ſo doth al mankind euer
ſince, as truly
\CNote{2.~Reg.~14.}
ſaid the woman of Thecua to
\Fix{king}{King}{likely typo, fixed in other}
Dauid: \Emph{vve doe al die, and as vvaters that returne not, vve fal
dovvne on the earth.} \Emph{And vvhat els} (ſaith
\CNote{ho.~37. in Euangel.}
S.~Gregorie) \Emph{is
this daylie decaying of our corruption, but a lingering death?} And none
of al theſe that liued longeſt reaching to a thouſand yeares (which
\CNote{Pſal.~89.}
with God is as one day) man dyed in that day in which he tranſgreſſed.

Morally
\CNote{S.~Ireneus li.~5. aduer. Heret.}
\MNote{Al time is ſhort in reſpect of eternitie.}
ancient Fathers here note, that albeit the life of the Patriarkes
ſeemeth long to vs, yet if we cõpare the ſame to eternitie it is
nothing. Neither by the iudgement of Philoſophers may aniething be
counted long, that hath an end: as
\CNote{Cicero li. de Senec. & q.~1. Tuſcul.}
Tullie bringing Cato wiſely
diſputing, ſheweth the longeſt life to be but a ſhort moment. VVhereby
againe we may ſee what loſſe we ſuſtaine by ſinne: ſeeing if ſinne had
not benne, we ſhould al haue benne tranſlated from earth to heauen, and
neuer haue dyed.}
and he died. \V Seth alſo liued a hundred fiue yeares, and begat
Enos. \V And Seth liued after he begat Enos, eight hundred and ſeuen
yeares, and begat ſonnes and daughters. \V And al the dayes of Seth came
to nine hundred & twelue yeares, and he died. \V And Enos liued nintie
yeares, and begat Cainan. \V After whoſe birth he liued eight hundred &
fiftene yeares, and begat ſonnes and daughters. \V And al the dayes of
Enos came to nine hundred and fiue yeares, and he died. \V Cainan alſo
liued ſuentie yeares, & begat Malaleel. \V And Cainan liued after he
begat Malaleel, eight hundred & fourtie yeares, and begat ſonnes &
daughters. \V And al the dayes of Cainan came to nine hundred and ten
yeares, and he died. \V And Malaleel liued ſixtie fiue yeares, and begat
Iared. \V And Malaleel liued after he begat Iared, eight hundred and
thirtie yeares, and begat ſonnes and daughters. \V And al the dayes of
Malaleel came to eight hundred nyntie fiue yeares, & he died. \V And
Iared liued a hundred ſixtie two yeares, and begat Enoch. \V And Iared
liued after he begat Enoch eight hundred yeares, and begat ſonnes and
daughters. \V And al the dayes of Iared came to nine hundred ſixtie two
yeares, & he died. \V Moreouer Enoch liued ſixtie fiue yeares, & begat
Mathuſala. \V And Enoch
\SNote{This Hebrew phraſe \Emph{vvalked vvith God}, ſignifieth that he
liued wel & pleaſed God.}
walked with God: & liued after he begat Mathuſala, three hundred yeares,
and begat ſonnes and daughters. \V And al the dayes of Enoch came to
three hundred ſixtie fiue yeares. \V And he walked with God, and
\LNote{VVas ſeene no more.}{That
\MNote{Enoch & Elias yet liuing in bodie.}
Enoch and Elias are yet aliue is a conſtant knowne truth, in the hartes and
mouthes of the faithful, ſaith S.~Auguſtin in his firſt booke,
\Cite{de peccat. merit & remiſſ. c.~3.}
and confirmeth the ſame in
\CNote{li.~20. ciuit. c.~29. lib.~2. de gratia Chriſti
c.~23. tract.~4. in Ioã.}
diuers other places. And it is teſtified by very many both Greeke and
Latin Doctors. 
\Cite{S.~Ireneus li.~5.}
\Cite{S.~Iuſtinus Martyr, q.~85. ad Orthodoxos.}
\Cite{S.~Hippolitus li. de Antichriſto.}
\Cite{S.~Damaſcen, li.~4. de Orthodoxa fide.}
\Cite{S.~Hierom. epiſt.~61. ad Pamach. c.~11.}
\Cite{S.~Ambroſe in Pſalm.~45.}
\Cite{S.~Chryſoſtom. ho.~21. in Gen. }
\Cite{ho.~58. in Mat.}
\Cite{ho.~4. in epiſt.~2. ad Theſſ.}
\Cite{ho.~22. in ep. ad Heb.}
\Cite{S.~Greg. li.~14. Moral. c.~11}
\Cite{ho.~12. in Ezech.}
\Cite{S.~Proſp. li.~vlt. de promiſ.}
\Cite{S.~Bede in c.~9. Marc.}
\Cite{Theophilact and Oecumenius in cap.~17. Mat.}
and others innumerable.
\MNote{Manifeſt Scripture that Elias yet liueth, and ſhal be ſlaine & an
other with him.}
Touching Elias it is manifeſt in Scriptures, that
\CNote{Malac.~4.}
he ſhal come, & preach, &
\CNote{Apo.~11.}
be ſlaine with an other witnes of Chriſt, before the terrible day of
Iudgement.
\MNote{Likewiſe that Enoch did not ſee death.}
Of Enoch Moyſes here maketh the matter more then probable, ſaying of
euerie one of the reſt, \Emph{he dyed}, onlie of Enoch ſaith not ſo, but
that \Emph{he appeared, or vvas ſeene no more}. For which the ſeuentie
two interpreters ſay,
\TNote{\G{καὶ ὀυχ ἑυρισκετο ὅτι μετατέθηκεν ἀυτὸν ὁ Θεὸς.}}
\Emph{And he vvas not found, for God tranſlated
him.} VVhich can not ſignifie death, but tranſporting, or remouing to an
other place. VVhereto agreeth the author of Eccleſiaſticus,
ſaying:
\CNote{Eccli.~44.}
\Emph{Enoch pleaſed God, and vvas tranſlated.} But moſt clearly
S.~Paul ſaith:
\CNote{Heb.~11.}
\Emph{Enoch vvas tranſlated, that he should not ſee
death, and he vvas not found for God tranſlated him.} VVith what plainer
wordes can any man declare, that a ſpecial perſon were not dead, then to
ſay: \Emph{He vvas tranſlated}, or cõueyed away, \Emph{that he should
not ſee death}?
\MNote{Theſe Scriptures ſpeake of temporal not of ſpiritual death.}
Neither is it a reaſonable euaſion to interprete this of ſpiritual
death. For ſo Adam being eternally ſaued (as 
\Cite{S.~Ireneus li.~3. c.~34. Epiphan. con. hereſim.~46.}
\Cite{S.~Auguſtin epiſt.~99. ad Euodium},
and others teach, and the whole Church beleeueth) was preſerued from
that death, and ſo vndoubtedly were Seth, and Enos being moſt holie, and
the reſt here recounted, as is moſt probable. 
\MNote{The Fathers proue by the Scriptures that Enoch is not dead.}
Neuertheles for further confutation of the contrarie opinion of
Proteſtants, the reader may alſo obſerue the iudgement of
\CNote{ho.~21. in Gen.}
S.~Chriſoſtom,
who affirmeth that \Emph{Though it be not a matter of faith, vvhether
Enoch be novv in Paradiſe from vvhence Adam and Eue vvere expelled, or
in ſome other pleaſant place}: \L{Dicunt tamen ſactæ Scripturæ quod Deus
tranſtulit eum, & quod viuentem tranſtulit eum, quod mortem ipſe not ſit
expertus} \Emph{The holie Scriptures ſay that God tranſlated him, and
that he tranſlated him aliue, that he felt not (or hath not experienced)
death.} And
\CNote{lib.~1. de pec. mer. c.~2. &~3.}
S.~Auguſtin as expreſly ſaith: \L{Non mortuus, ſed viuus
tranſlatus eſt.} \Emph{He (to vvit Enoch) is tranſlated, not dead but
aliue.} Yea he teacheth how his life is ſuſtayned thus many thouſand
yeares vpon earth. And ſheweth moreouer that both Enoch and Elias ſhal
dye. \Emph{For ſeing Enoch and Elias} (ſaith
%%% !!! Are these right. Where does S. Thomas go?
\CNote{l.~9. de Gẽ. ad lit. c.~6. S.~Tho. in c.~11. ad Hebreos.}
he) \Emph{are dead in Adam,
and carying the ofspring of death in their flesh, to pay that debt, are
to returne to this life} (of common conuerſation) \Emph{and to pay this
debt vvhich ſo long is deferred.}

Diuers
%%% !!! Do all these go here?
\CNote{S.~Chriſ. ho.~21. in Gen. et in 2.~Theſ.~2. Theodor. q.~45. in
Gen. Aretus. in 11.~Apoc. S.~Greg. lib.~14. Moral. c.~vlt. & ho.~12. in
Ezech.}
\MNote{Cauſes why Enoch & Elias are reſerued aliue.}
reaſons are alſo alleaged, why God would reſerue theſe two aliue.
\MNote{1.}
Firſt to ſhew by example, that as their mortal bodies are long conſerued
from corrupting or decaying, in like ſorte Adam and Eue and al others
not ſinning, ſhould haue bene conſerued, and according to Gods promiſe,
neuer haue died, but after ſome good time tranſlated to heauen, and
indued with immortalitie.
\MNote{2.}
Secondly to giue vs an argument of immortalitie, which is promiſed after
the general Reſurrection. For ſeing God doth preſerue ſome mortal, ſo
long from al infirmitie, we may aſſuredly beleue that he wil geue
immortal & eternal life of
\Fix{bobie}{bodie}{obvious typo, fixed in other}
and ſoule to his Sainctes, after they haue payed the debt of death, and
are riſen againe.
\MNote{3.}
Thirdly 
\MNote{See
\Cite{D.~Sand lib.~8. c.~35. de Monar. Eccl.}
And
\Cite{F.~Pererius in c.~12. Danielis.}}
theſe two (one of the law of nature, the other of the law of
Moyſes) are preſerued aliue, to come amongſt men againe towards the end
of the world, to teach, teſtifie, and defend the true faith and doctrin
of Chriſt, againſt Antichriſt, when he ſhal moſt violently oppugne &
perſecute the Church. Of Enoch it is ſaid in the booke of
\CNote{Eccl.~44.}
Eccleſiaſticus, that he was tranſlated, \L{vt det gentibus
pœnitentiam}, \Emph{that he geue repentance to the nations}, by his
preaching, & reducing the deceiued from Antichriſt. And of Elias
\CNote{Mala.~4. Eccl.~48.}
Malachie prophecieth, that he \Emph{shal come before the great and
terrible day of our Lord, and shal turne the hart of the fathers} (that
is the people of the Iewes) \Emph{to the ſonnes} (the
Chriſtians) \Emph{and of the ſonnes} (the deceiued Chriſtians) \Emph{to
the fathers}, the ancient true Catholiques.}
was ſeene no more: becauſe God
\SNote{The ſeuentie two Interpreters ſay, God tranſlated him and ſo doth
S.~Paul.
\XRef{Heb.~11.}}
tooke him. \V Mathuſala alſo liued a
hundred eightie ſeuen yeares, & begat Lamech.
%%% o-0035
\V And Mathuſala liued,
after he begat Lamech, ſeuen hundred eightie two yeares, and begat
ſonnes and daughters. \V And al the dayes of Mathuſala came to
\SNote{This is the longeſt life of al here recited. But if we conſider
that Adam was as ſtrong of bodie, the firſt day he was created, as theſe
others were at the age of 60.~yeares (before which, none are ſaid to
haue begot children) and ſo ſubtract 60.~yeares from Mathuſala, then
Adam liued in mans ſtate longer then he by 21.~yeares.}
nyne hundred ſixtie nine yeares, & he died. \V
And Lamech liued a hundred eightie two yeares, and begat a ſonne: \V and
he called his name Noe, ſaying: This ſonne ſhal comfort vs from the
workes & labours of our handes on the earth, which our Lord curſed. \V
And Lamech liued, after he begat Noe, fiue hundred nintie fiue yeares,
and begat ſonnes and daughters. \V And al the dayes of Lamech came to
ſeuen hundred ſeuentie ſeuen yeares, and he died. And Noe when he was
fiue hundred yeares olde, begat Sem, Cham, and Iaphat.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
