%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 32
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-32


\project douay-rheims


%%% 0124
%%% o-0115
\startChapter[
  title={Chapter 32}
  ]

\Summary{Angels mete Iacob by the way. 3.~He ſendeth meſſengers and
  giftes to pacifie his brother Eſau. 24.~Wreſtling with an Angel is not
  ouercome, in fine the Angel benummeth his thiegh, bleſſeth him, and
  fortelleth that he shal be called Iſrael.}

Iacob alſo went on his iourney that he had begunne: and the Angels of
God met him. \V Whom when he had ſeene, he ſaid: Theſe are the Campes
of God, and he called the name of that place Mahanaim, that is,
Campes. \V And he ſent alſo meſſengers before him to Eſau his brother
into the land of Seir, into the countrie of Edom: \V and he commanded
them, ſaying: Thus ſpeake ye vnto my lord Eſau: This ſaith thy brother
%%% o-0116
Iacob: I haue ſoiourned, and haue bene with Laban vntil this preſent
day. \V I haue oxen, and aſſes, and ſheepe, and men ſeruants, and wemen
ſeruants: and now I ſend a legacie to my lord, that I may finde grace
in thy ſight. \V And the meſſengers returned to Iacob, ſaying: We came
to Eſau thy brother, and behold he cometh with ſpede to mete thee with
foure hundred men. \V Iacob
\LNote{Feared exceedingly.}{Iuſtly
\MNote{Iacobs feare was iuſt, and without fault.}
may we meruel, why Iacob ſo often aſſured by Gods promiſes, confirmed by
his manie bleſſings, protected in al former dangers,
\Fix{accompained}{accompanied}{likely typo, fixed in other}
the night before with armies of Angels, indued alſo with al vertues, and
namely with
\CNote{1.~Ioan.~4.}
perfect charitie (which expelleth feare) was for al this ſo
vehemently afeard!
\CNote{q.~102. in Gen.}
S.~Auguſtin anſwereth, that he neither diſtruſted in
God, nor did anie vnlawful thing: but did his owne endeuour wiſely and
confidently, leſt by preſuming or deſparing he ſhould rather haue
tempted God, then truſted in him.
\MNote{The cauſes of his feare.}
The cauſes of his feare were in reſpect of him ſelfe and his
brother.
\MNote{The humble conceipt of him ſelfe.}
For conſidering Gods former promiſes, benefites & protections
were not to be preſumed as abſolute ſignes of his perpetual loue but
conditional, if him ſelfe perſeuered ſincerly in Gods ſeruice. And ſeing
\CNote{Eccle.~9.}
\Emph{The iuſt man knovveth not vvhether he be vvorthie of loue, or of
hatred}, he might doubt, leſt by his twentie yeares conuerſation among
Infidels in Meſopotamia, he had contracted ſome ſinnes, for which God
might ſuffer him to fall into calamitie and affliction. And though he
was in dede ſtil more and more vertuous, and conſequently in Gods more
fauour and protection: yea ſo much the more, by how much leſſe he
preſumed of his owne good ſtate and merites:
\MNote{Eſaus inclination & meanes to reuenge.}
yet by the vehement apprehending of his brothers inclination to reuenge,
the greatnes of the occaſion by procuring the firſt-birth-right, and his
fathers bleſſing from him, the newes of his ſpeedie coming towards him
with foure hundred men, the natural ſituation of the place, where Eſau
might
\Fix{eſealy}{eaſely}{likely typo, fixed in other}
inuiron him, and (as he humbly thought) his owne vnworthines, he was
poſſeſſed with natural feare (ſuch as happeneth to conſtant men) and was
ſore afflicted for the tender care of his familie. But reflecting vpon
Gods goodnes, he prudently diſpoſed of his people and flockes, and
beſought God to protect him and his,
\MNote{Iacobs prayer qualified with Humilitie.}
by prayer qualified with requiſite conditions, to wit, with humilitie, not
asking for his owne but \Emph{for Abraham and Iſaacs} ſake, and for Gods
owne promiſe, acknowledging himſelfe to be \Emph{leſſe then Gods
mercies} towards him,
\MNote{Gratitude.}
with gratitude recounting great benefites receiued, ſaying, \Emph{VVith
my ſtaffe I paſſed ouer this Iordan, and novv vvith tvvo troupes I do
returne},
\MNote{Confidence.}
with confidence in that God had ſaid, he \Emph{VVould dilate his ſede as
the ſand of the ſea},
\MNote{Meeknes.}
and with meeknes in \Emph{ſending giftes and good vvordes} to Eſau. Thus
finally he pacified him, and ſo his owne feare was turned into ioy.}
feared exceedingly: & being ſore affraid diuided the people that was
with him, the flockes alſo and the ſhepe and the oxen, and the camels,
into two troupes, \V ſaying: If Eſau come to one troupe, and ſtrike it,
the other troupe that remaineth, ſhal be ſaued. \V And Iacob ſaid: O God
of my father Abraham, and God of my father Iſaac: O Lord that dideſt ſay
to me: Returne into thy land, and into the place of thy natiuitie, and I
wil doe thee good. \V I am inferiour to al thy mercies, and thy truth
that thou haſt fulfilled to thy ſeruant. With my ſtaffe I paſſed ouer
this Iordain: and now with two troupes I doe returne. \V Deliuer me from
the hand of my brother Eſau, becauſe I am ſore affraid of him: leſt
perhaps he come, and ſtrike the mother with the children. \V Thou dideſt
ſay that thou wouldeſt do good to me,
%%% 0125
and dilate my ſeed as the ſand of the ſea, which for multitude can not
be numbred. \V And when he had ſlept there that night, he ſeparated of
thoſe things which he had, giftes to his brother Eſau, \V ſhe goates
two hundred, he goates twentie, ewes two hundred, and rammes twentie, \V
thirtie milch camels with their coltes, fourtie kine, and twentie
bulles, twentie ſhe aſſes, and their foles ten. \V And he ſent by the
handes of his ſeruants, euerie flocke by it ſelfe, and he ſaid to his
ſeruants: Goe before me, and let there be a ſpace betwen flocke and
flocke. \V And he commanded the former, ſaying: If thou mete my brother
Eſau, and he aske thee, whoſe art thou? or whither goeſt thou? or whoſe
are theſe that thou doeſt folowe? \V thou ſhalt anſwere: Iacobes thy
ſeruant, he hath ſent them for gifts to my lord Eſau: himſelfe alſo
cometh after vs. \V In like maner he gaue commandements to the ſecond,
and the third, and to al that folowed the flocks, ſaying: With the ſelfe
ſame words ſpeake ye to Eſau, when you ſhal finde him. \V And ye ſhal
adde: Iacob alſo thy ſeruant himſelfe foloweth on after vs; for he
ſaid: I wil pacifie him with the gifts that goe before, and afterward I
wil ſee him, perhaps he wil be gracious vnto me. \V The giftes therfore
went before him, but himſelfe lodged that night in the campe. \V And
when he was riſen early he tooke his two wiues, and his hand-maides as
manie, with his eleuen ſonnes, and paſſed ouer the ford Iaboc. \V And
when he had ſet ouer al things that appertained to him, \V he taried
alone: and behold
\LNote{A man vvraſtled.}{This
\MNote{Iacob wreſtled with an Angel corporally & ſpiritually.}
wreſtling with an Angel aſſumpting a bodie in forme of a man was
corporal, as the effect ſhewed in Iacobs ſinow ſhrunck vp, which made
him to halt.
\XRef{v.~25. &.~31.}
\CNote{Oſee.~12.}
It was alſo ſpiritual, as appeareth by his earneſt prayer, vrging and at
laſt obtayning the Angels bleſſing.
\Cite{S.~Dionuſ. c.~4. cel. Hierer.}
\Cite{S.~Greg. prefat. in Iob.}
\Cite{Theodoret, q.~91. in Gen.}}
a man wraſteled with him til morning. \V Who when he ſaw that he could
not ouercome him, he touched the ſinowe of his thighe, and forthwith it
ſhranke. \V And he ſaid to him: Let me goe for it
%%% o-0117
is breake of day. He
anſwered: I wil not let thee goe, vnleſſe thou bleſſe me. \V He therfore
ſaid: What is thy name? He anſwered: Iacob. \V But he, no, thy name,
quoth he,
\SNote{The changing of his name here promiſed, is performed
\XRef{chap.~35.}
\Cite{S.~Hierom. Tradit. Heb.}}
ſhal not be called Iacob, but Iſrael: for if thou haſt bene ſtrong
againſt God, how much more shalt thou preuaile againſt men? \V Iacob
asked him: Tel me by what name art thou called? He anſwered: Why doeſt
thou aske my name? and bleſſed him in the ſame place. \V And Iacob
called the name of the place Phanuel, ſaying: I haue ſene God face to
face, and my ſoule was made ſafe. \V And immediatly the ſunne roſe to
him, after that he was paſt Phanuel; but he
%%% 0126
halted on his foote. \V For which cauſe the children of Iſrael eate not
the ſinowe, that ſhrunke in Iacobs thighe, vnto
\Fix{his}{this}{obvious typo, fixed in other}
preſent day: becauſe he touched the ſinowe of his thighe, and it
ſhrunke.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
