%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 40
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-40


\project douay-rheims


%%% 0142
%%% o-0132
\startChapter[
  title={Chapter 40}
  ]

\Summary{Ioſeph interpreteth the dreames of two Eunuches priſoners
  12.~that the one should be reſtored to his office, 16.~the other be
  hanged. 20.~The third day the euent declareth the interpretations to
  be true, but Ioſeph is forgotten.}

Theſe thinges being ſo done, it chanced that two Eunuches, the cupbearer
of the king of Ægypt, and his baker, offended againſt their lord. \V And
Pharao being
%%% 0143
wrath againſt them (for the one was chiefe of the cupbearers, the other
chiefe baker) \V he ſent them into the priſon of the captaine of the
ſouldiers, in the which Ioſeph alſo was priſoner. \V But the keeper of
the priſon deliuered them to Ioſeph, who alſo miniſtred to them: ſome
litle time was paſſed, and they were kept in cuſtodie. \V And they ſawe
ech of them both a dreame in one night, according to an interpretation
agreing to them ſelues: \V to whom when Ioſeph was entred in the
morning, and ſaw them ſad, \V he asked them, ſaying: Why is your
countenance ſadder to day then it was
\Fix{woont?}{wont?}{possible typo, fixed in other}
\V Who anſwered: We haue ſeene a dreame, & there is no bodie to
interprete it to vs. And Ioſeph ſaid to them: Why
\LNote{Doth not interpretation belong to God?}{Dreames do come of diuers
cauſes.
\MNote{Some dreames are natural.}
Some of natural complexion, or diſpoſition, wherby Philoſophers or
Phiſitions may probably iudge of the ſtate of mans bodie. Some are
rather effects of things paſt, then ſignes of anie thing to come. Of
which ſorte the wiſe man ſaith: \Emph{Dreames do folovv manie cares.}
\XRef{(Eccle.~5.)}
\MNote{Some are illuſions of euil ſpirites.}
Some are ſuggeſted by euil ſpirites, either to flatter worldlings with
great pretenſes, or to terrifie weake mindes with dangers and
afflictions, or to vexe and truble thoſe in ſleepe, whom they can not
eaſely moue waking, as S.~Gregorie diſcourſeth
\Cite{(li.~8. Moral in cap.~7. Iob)}
\MNote{Some are from God.}
Some dreames are of God, as in Iacob, Ioſeph, theſe Eunuches, Pharao,
\CNote{Dan.~4.}
Nabuchodonoſor, and others both good and euil men. But to diſcerne and
aſſuredly to iudge of ſome dreames, whether they be from God, by holie
Angels, or illuſions of euil ſpirites, is a ſpecial gift of God, as alſo
the interpretation therof belongeth to God, as Ioſeph here teſtifieth.
\MNote{Holie Scripture and the Church are iudges of doubtful dreames.}
VVhoſoeuer therfore wil be ſecure muſt relie either vpon expreſſe
Scripture, or iudgement of the Church, as in ominous ſpeaches was noted
before
\XRef{(chap.~24.)}
Otherwiſe the general rule is, not to obſerue dreames.
\XRef{Deut.~18.}}
doth not interpretation belong to God? Tel me what you haue ſeene. \V
The chiefe of the cupbearers firſt told his dreame: I ſaw before me a
vine, \V wherin were three branches, growing by litle and litle into
buddes, and after the bloſſomes the grapes waxed ripe: \V and the cup of
Pharao in my hand: and I tooke the grapes, and wrong them into the cup
which I held, and I gaue the cup to Pharao. \V Ioſeph anſwered: This is
the interpretation of the dreame: The three branches, are yet three
dayes: \V after the which Pharao wil remember thy ſeruice, and wil
reſtore thee to thy old degree: and thou ſhalt geue him the cup
according to thyne office, as before thou haddeſt wont to doe. \V Only
remember me,
%%% o-0133
when it ſhal be wel with thee, and doe me this mercie: to put Pharao in
mind that he take me out of this priſon: \V becauſe I was taken away by
ſtealth, out of the land of the 
\Fix{hebrewes,}{Hebrewes,}{likely typo, fixed in other}
and here an innocent was I caſt
into the lake. \V The maiſter of the bakers ſeing that he had wiſely
reſolued the dreame, he ſaid: And I alſo ſaw a dreame, That I had three
baskettes of meale vpon my head: \V and that in one basket that was the
higher, I caried al meates that are made by the art of baking, and that
the birds did eate out of it. \V Ioſeph anſwered: This is the
interpretation of the dreame: The three basketts, are yet three
dayes: \V after the which Pharao wil take thy head from thee, and hang
thee
\SNote{Death on the croſſe was moſt cruel, & moſt ignominious.
\Cite{Cicero 7.~Ver.}
yet ſuffered by Chriſt, and by him made glorious.
\XRef{Sap.~2.}
\XRef{Philip.~2.}}
on the croſſe, and the foules ſhal teare thy fleſh. \V The third day
after this was the birth day of Pharao: who making a great feaſt to his
ſeruantes, at the banket he remembred the maiſter of the cupbearers, and
the chiefe of the bakers. \V And he reſtored the one into
%%% 0144
his place, to reach him the cuppe, \V the other he hanged on a gibbet,
that the truth of the interpreter might be approued. \V And yet
notwithſtanding the chiefe of the cupbearers, proſperous thinges
ſucceeding, forgat his interpreter.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
