%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 27
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-27


\project douay-rheims


%%% 0108
%%% o-0101
\startChapter[
  title={Chapter 27}
  ]

\Summary{Iacob by his mothers counſail getteth his fathers bleſsing in
  place of Eſau. 42.~And by her is aduiſed (for auoiding Eſaus wrath,
  who threatned to kil him) to flie to his vncle Laban, in Haran of
  Meſopotamia.}

And Iſaac was old, and his eyes were dimme, and he could not ſee: and he
called Eſau his elder ſonne, and ſaid to him: My ſonne? Who anſwered:
Here I am. \V To whom his father: Thou ſeeſt, quoth he, that I am old,
and know not the day of my death. \V Take thy inſtruments, thy quiver,
and bowe, and goe abrode: and when thou haſt taken any thing by
hunting, \V make me broth therof, as thou knoweſt I like, and bring that
I may eate: and my ſoule may bleſſe thee before I dye. \V Which when
Rebecca had heard, and he was gone into the field to fulfil his fathers
commandement, \V ſhe ſaid to her ſonne Iacob: I heard thy father talking
with Eſau thy brother, and ſaying to him: \V Bring me of thy hunting,
and make me meates that I may eate, and bleſſe thee in the ſight of our
Lord before I dye. \V Now therfore my ſonne, aſſent to my counſel: \V
and go thy way to the flocke, bring me two kiddes of the beſt, that I
may make of them meat for thy father, ſuch as he gladly eateth: \V which
when thou haſt brought in, and he hath eaten, he may bleſſe thee before
he dye. \V To whom he anſwered: Thou knoweſt that Eſau my brother is an
hearie man, and I am ſmooth: \V if my father ſhal feele me, and perceaue
it,
\SNote{Iacob ſecure in conſcience that the right of firſt-birth belonged
to him, yet feared to geue occaſiõ of offence to his father.}
I feare leſt he wil thinke I would
%%% 0109
haue deluded him, and I ſhal bring vpon me a curſe for a bleſſing. \V To whom
his mother ſaid: This curſe, my ſonne, light vpon me: only heare thou my
voice, and go, fetch me the things which I haue ſaid. \V He went, and
brought, and gaue them to his mother. She dreſſed meats, euen as ſhe
knew his father liked. \V And ſhe did on him the garments of Eſau verie
good, which ſhe had at home with her: \V and the litle skinnes of the
kidds ſhe put about his hands, and couered the bare of his necke. \V And
ſhe gaue him the broth, and deliuered him bread that ſhe had baked. \V
Which when he had caried in, he ſaid: My father? But he anſwered: I
heare. Who art thou my ſonne? \V And Iacob ſaid:
\LNote{I am thy firſt begotten Eſau.}{Iacob
\MNote{Iacob did not lie, nor ſinne in ſaying he was Eſau &c.}
was not by nature the firſt begotten; but by Gods ordinance, & by
couenant made with Eſau, had right to the preeminence, and priuilegies
belonging to the firſt borne. So he did not lie, but ſpake a truth,
meaning that he was that ſonne, to whom by diuine election the
firſt-birth-right was dew, which his father ſuppoſed to pertaine to
Eſau.

But becauſe ſome ſcorners of Chriſtian doctrin (like to the old
Manichees) vſe to ſay, that Catholique Doctors, and Schoolmen excuſe,
and condemne whom they liſt by ſuch gloſſes, let ſuch reprouers
vnderſtand,
\MNote{He is proued innocent by the text.}
that both moderne and ancient Catholique writers auow this
defence of the holie Patriarch Iacob, not by priuate ſpirite, but by the
moſt true and proper ſenſe of holie Scripture itſelfe. VVhere it may
appeare, if they wil
\Fix{axamine}{examine}{likely typo, fixed in other}
the text, that Iacob in al this procurement of his fathers bleſſing,
neither did aniething vniuſtly, nor ſaid anie thing falſly.
\MNote{1.}
Firſt it was reueled to his mother
\XRef{(chap.~25. v.~23.)}
\Emph{That the elder} (of her twinnes) \Emph{should ſerue the younger}.
\MNote{2.}
Secondly, holie Scripture teſtifieth in the ſame chapter
\XRef{(v.~27.)}
\Emph{That Iacob vvas a plaine} (or ſincere) \Emph{Man}, void of vniuſt
dealing. 
\MNote{3.}
Thirdly, for more quiet enioying that right, which God had ordained for
him, he procured his brothers conſent and confirmation
\XRef{(v.~33.)}
\MNote{4.}
Fourtly though he was ſecure in conſcience that the bleſſing was dew
to him, yet he feared
\XRef{(v.~12.)}
leſt he might geue occaſion of offence to his father, to whom this
myſterie was not yet reueled.
\MNote{5.}
Fiftly Iſaac perceiuing at laſt Gods wil, that Iacob ſhould be
preferred, was neither offended with him, nor reuoked his bleſſing, as
vnlawfully ſurpriſed, but condeſcending therto, ratified that he had
done, ſaying
\XRef{(v.~33.)}
\Emph{I haue bleſſed him, and he shal be bleſſed.}
\MNote{6.}
Sixtly, God himſelfe from this time forwards often appeared to Iacob,
and with great promiſes, and manie temporal and ſpiritual benefites,
declared his ſingular loue to him.
\MNote{7.}
Seuently, theſe three Patriarches Abraham, Iſaac, and Iacob are ſpecial
renowmed Sainctes of the old Teſtament: yea the Lord and Creator of al
would peculiarly be called
\XRef{(Exodi.~3.)}
the God of Abraham, Iſaac, and Iacob. Moyſes praying inſtantly for Gods
mercie and clemencie towards the people
\XRef{(Exodi.~32.)}
beſought him to remember Abraham, Iſaac, and Iacob his ſeruants: and ſo
in both old and new Teſtament theſe three are often mentioned as chiefe
Princes in the Kingdom of Heauen. Al which ſhewe the great vertues and
holines of them al.

And
\MNote{The Fathers proue his innocencie in this fact.}
touching this fact of Iacob, where (if euer aniewhere) might ſeme to be
ſome great ſinne, S.~Auguſtin at large proueth that he did not herein
ſinne at al: \Emph{That vvhich Iacob did} (ſaith he,
\Cite{li. cont. mendacium c.~10.)}
\Emph{By his mothers inſtruction to deceiue his father, if it be
diligently conſidered, vvas no lie, but a myſterie, and therfore for the
familiar counſail of the Holie Ghoſt, vvhich his mother had receiued, he
is excuſed from ſinne.} The ſame he confirmeth.
\Cite{q.~74. in Gen.}
\Cite{li.~16. c.~37. de ciuit.}
&
\Cite{li.~22. c.~34. cont. Fauſt.}
The ſame alſo teach S.~Chriſoſtom
\Cite{ho.~53. in Gen.}
S.~Hierom
\Cite{Epiſt.~125.}
S.~Theodoret
\Cite{q.~79. &~80. in Gen.}
S.~Gregorie,
\Cite{ho.~6. in Ezechielem.}
S.~Bede, Iſidorus, Innocentius 3, Rupertus and others vpon this place,
\MNote{Euerie lie is a ſinne.}
al agreing abſolutly that euerie lie is a ſinne, declare that Iacob lied
not, but ſtil ſpake the truth, confirming their expoſition by other like
places of Scripture. As when our Sauiour ſaid of S.~Iohn Baptiſt
\XRef{(Math.~11.)}
\Emph{He is Elias} meaning that he was Elias in ſpirite not in
perſon. So Iacob ſaid truly that he was Eſau, not meaning in perſon but
in right of the firſt borne, by Gods ordinance: Eſau alſo hauing
condeſcended therto by couenant and oath. In that alſo he deceiued his
father, was no ſinne.
\MNote{Some deceipt good.}
For it was a lawful and good deceipt, ſuch a one (ſaith
\CNote{Ho.~6. ad Col.}
S.~Chriſoſtom)
as Hieremie ſpeaketh of, \Emph{Lord thou haſt deceiued me, and I am
deceiued}, ſo Iſaac was deceiued, not as we commonly cal deceipt, but to
his owne and others good, by Gods diſpoſition.}
I am thy firſt begotten Eſau: I haue done as thou dideſt command me:
ariſe, ſit, and eate of my hunting, that thy ſoule may bleſſe me. \V And
againe Iſaac to his ſonne: How couldeſt thou, ſaid he, find it ſo
quickly, my ſonne? Who anſwered:
\SNote{It was truly Gods wil, but not in that ſẽſe as Iſaac vnderſtood
it.}
It was the wil of God that that which I would came quickly in my way. \V
And Iſaac ſaid: Come hither, that I may feele thee my ſonne, and
%%% o-0102
may proue whether thou be my ſonne Eſau, or
no. \V He came nere to his father, and when he had felt him, Iſaac ſaid:
The voice verely, is the voice of Iacob: but the hands, are the hands of
Eſau. \V And he
\LNote{Knevv him not.}{S.~Damaſus demanding of
\CNote{Hier.~20. Epiſt.~125.}
S.~Hierom, what might be
the reaſon why God would ſuffer his holie ſeruant Iſaac not to know
Iacob, but to be deceiued, and through ignorance to bleſſe whom he would
not, declareth that it happened not only to
\Fix{Iacob,}{Iſaac,}{likely typo, same in both}
but alſo to manie
other like holie men, to be ignorant of manie things, and to be deceiued
in error of opinion:
\MNote{It was good that Iſaac knew not Iacob when he bleſſed him.}
and that this error was profitable to Iſaac and his houſe. For if he had
geuen this bleſſing (which was a ſpiritual Iuriſdiction) to Eſau, as he
purpoſed, he had committed a noxious error in dede, by preferring a
bloudie man, one that was readie, if he could, to haue killed his
brother
\XRef{(v.~41.)}
omitting him, that was ſincere and very vertuous, and had done his
owne wil, not Gods wil therin.

But why would not God reuel his wil to Iſaac (as he had commanded a
farre greater thing to Abraham to ſacrifice the ſame Iſaac) that he
might wittingly haue bleſſed Iacob by Gods commandment? The Fathers do
probably alleage this for one reaſon,
\MNote{Good in reſpect of Eſau.}
that if Eſau, being a fierce and
cruel man, had perceiued that his father had willingly preferred Iacob,
he would haue bene incenſed againſt his father, conceiued and attempted
euil againſt him. An other reaſon
\CNote{Ho.~53.}
S.~Chriſoſtom and
\CNote{q.~79. in Gen.}
Theodoret do yeld,
\MNote{More to Gods glorie, and Iacobs commendation.}
that by this ſtrange maner of imparting this bleſſing, it might more
manifeſtly appeare to be Gods wil and ordinance, and not to procede from
mans affection, that Iacob ſhould be preferred.}
knew him not, becauſe his hearie hands had made him like vnto the
elder. Bleſſing him therfore, \V he ſaid: Art thou my ſonne Eſau? He
anſwered: I am. \V But he ſaid: Bring me the meats of thy hunting, my
ſonne, that my ſoule may bleſſe thee. Which when they were brought and
he had eaten, he offred him wine alſo, which after he had drunke, \V he
ſaid to him: Come nere me, and geue me a kiſſe, my ſonne. \V He came
nere, and kiſſed him. And immediatly as he felt the fragrant ſauoure of
his garments, bleſſing him, he ſaid: Behold the ſauoure of my ſonne is
as the ſauoure of a plentiful field, which our Lord hath bleſſed. \V God
geue thee of the deaw of heauen, and of the fatnes of the earth
abundance of corne and wine. \V And let peoples ſerue thee, and tribes
adore thee: be thou lord of thy brethren, and thy mothers children bowe
they before thee. He that ſhal curſe thee, be he curſed: and he that
ſhal bleſſe thee, be he replenished with bleſſings. \V Iſaac had ſcarce
ended his wordes, and Iacob now gone forth abroad, but Eſau came, \V and
brought in to his father meates made
%%% 0110
of his hunting, ſaying: Ariſe my father, and eate of thy ſonnes hunting:
that thy ſoule may bleſſe me. \V And Iſaac ſaid to him: Why! who
\Fix{arth}{art}{likely typo, fixed in other}
thou? Who anſwered: I am thy firſt begotten ſonne Eſau. \V Iſaac was
amazed and aſtonied exceadingly: and marueling more then a man can
beleue, ſaid: Who is he then that euen now brought me veniſon that he
had taken, and I did eate of al thinges before thou cameſt? and I haue
bleſſed him,
\SNote{Iſaac now knowing it to be Gods wil, ratified that he had done.}
and he ſhal be bleſſed. \V Eſau hauing heard his fathers wordes, roared
out with a great crye: and being diſmaied, ſaid: Bleſſe me alſo, my
father. \V Who ſaid: Thy brother came deceiptfully and tooke thy
bleſſing. \V But he ſaid again: Rightly is his name called Iacob: for he
hath ſupplanted me loe the ſecond time: my firſt-birth-right he tooke
before, and now the ſecond time he hath ſtolen my bleſſing. And againe
to his father he ſaid: Haſt thou not reſerued me alſo a bleſſing? \V
Iſaac anſwered: I haue appointed him thy Lord, and al his brethren I
haue made ſubiect to his ſeruice: with corne and wine I haue eſtabliſhed
him, and for thee, my ſonne, what ſhal I doe more after this? \V To whom
Eſau ſaid: Haſt thou one only bleſſing, father? I beſech thee bleſſe me
alſo. And when he wept that he howled againe, \V Iſaac being moued, ſaid
to him: In
\SNote{Worldlings bleſſing conſiſteth in tranſitorie welth.}
the fat of the earth, and in the deaw of heauen from aboue \V ſhal thy
bleſſing be. Thou ſhalt liue by the ſworde, and ſhalt ſerue thy brother:
and
\SNote{The Idumeans being ſubdued by king Dauid
\XRef{(2.~Reg.~8.)}
reuolted from king Ioram, and had a king of their owne
\XRef{(4.~Reg.~8.)}
they were againe ſubdued by Hircanus
\Cite{(teſte Ioſepho. li.~13. Antiq.)}
but againe Herod an Idumean raigned in Iewrie.
\XRef{Math.~2.}
\XRef{Luc.~1.}}
the time ſhal come, when thou ſhalt shake of, and looſe his yoake from
thy necke. \V Eſau therfore alwaies hated Iacob for the bleſſing
wherwith his father had bleſſed him: and he ſaid in his hart: The daies
wil come of the mourning of my father, and I wil kil Iacob my
brother. \V Theſe things were told to Rebecca:
%%% o-0103
who ſending & calling Iacob her ſonne, ſaid to him: Behold Eſau thy
brother threatneth to kil thee. \V Now therfore, my ſonne, heare my
voice, and get thee vp and flye to Laban my brother into Haran: \V and
thou ſhalt dwel with him a few daies, til the furie of thy brother be
aſſwaged, \V and his indignation ceaſe, and he forget thoſe things,
which thou haſt done to him: afterward I wil ſend, and bring thee from
thence hither. Why ſhal I be depriued of both ſonnes in one day? \V And
Rebecca ſaid to Iſaac: I am wearie of my life for the daughters of Heth:
If Iacob take a wife of the ſtocke of this land, I liſt not liue.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
