%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 50
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-50


\project douay-rheims


%%% 0171
%%% o-0157
\startChapter[
  title={Chapter 50}
  ]

\Summary{Ioſeph cauſeth his fathers bodie to be embawmed, 3.~the dayes
  of mourning being expired, 6.~with Pharaos leaue, Ioſeph with the
  ancients of Ægypt, al his brethren, and elder ſorte of Iſraelites goe,
  and ſolemnly burie the bodie in Chanaan. 14.~After their returne, his
  brethren fearing, leſt Ioſeph wil now reuenge former iniuries, he
  freely forgeueth al. 22.~At the age of 110.~yeares, adiuring the
  poſteritie to carie his bones into Chanaan he dieth, and is put in a
  coffin.}

VVhich Ioſeph ſeeing, fel vpon his fathers face weeping and kiſſing
him. \V And he commanded his ſeruantes the phyſitians, that they should
embawme his father with ſpices. \V Who fulfilling his commandements,
there paſſed fourtie dayes: for this was the maner of corſes embawmed,
and Ægypt mourned him ſeuentie daies. \V And the mourning time being
expired, Ioſeph ſpake to the familie of Pharao: If I haue found grace in
your ſight, ſpeake in the eares of Pharao: \V for ſo much as my father
did adiure me, ſaing: Behold I die, in my ſepulchre
\SNote{Iacob digged a ſepulchre for himſelfe, though it be not hertofore
mentioned when he did it.
\Cite{S.~Aug. q.~170. in Gen.}}
which I digged for my ſelfe in the land of Chanaan, thou shalt burie
me. I wil goe vp therfore, and burie my father, and returne. \V And
Pharao ſaid to him: Goe vp and burie thy father as thou waſt adiured. \V
Who going vp, there went with him al the ancients of Pharaos houſe, and
al the elders of the Land of Ægypt: \V the houſe of Ioſeph with his
brethren, ſauing their litle ones, and the flockes and heards, which
they had left in the Land of Geſſen. \V He had alſo in his traine
chariotts and horſemen: and it became no ſmal multitude. \V And they came
to the floore of Atad, which is ſituate beyond Iordaine: where
celebrating the exequies with great and vehement mourning, they ſpent
ful ſeuen dayes. \V Which when the Inhabiters of the Land of Chanaan had
ſeene, they ſaid: This is a great mourning vnto the Ægyptians. And
therfore the name of that place was called,
%%% 0172
The mourning of Ægypt. \V Therfore the ſonnes of Iacob did as he
commanded them:
%%% o-0158
\V and carying him into the Land of Chanaan, they buried
him in the duble caue, which Abraham had bought with the field for a
poſſeſſion to burie in of Ephron the Hethite againſt Mambre. \V And
Ioſeph returned into Ægypt with his brethren, and with al the traine,
his father being buried. \V After whoſe death, his brethren fearing, and
talking one with an other: Leſt perhaps he be mindful of the iniurie
which he ſuffered, and requite vs al the euil that we haue done, \V they
aduertiſed him ſaying: Thy father commanded vs before he died, \V that
we ſhould ſay thus much to thee in his wordes: I beſech that thou forget
the wicked fact of thy brethren, and the ſinne & malice which they haue
exerciſed againſt thee: we alſo deſire thee, that to the ſeruants of the
God of thy father thou remit this iniquitie. Whom when Ioſeph had heard
he wept. \V And his brethren came to him, and
\SNote{This word \Emph{adoring} often in holie Scripture ſignifieth
ciuil honour: as here it can haue no other ſenſe.}
adoring proſtrate on the ground they ſaid: We are thy ſeruantes. \V To
whom he anſwered: Feare not: can we reſiſt the wil of God? \V
\LNote{You thought euil.}{This
\MNote{Mans wil, not God the cauſe of ſinne.}
plaine diſtinction ſheweth that ſinne is wholly of the ſinner; and that
God hath no part therin, but turneth it to good. For thoſe things which
Ioſephs brethren did againſt him, were occaſions of his aduancement in
Ægypt, through the omnipotent wiſdome of God. VVhoſe propertie is, out
of euerie euil to draw good.
\Cite{S.~Chriſoſt. ho.~67. in Gen.}
\Cite{S.~Aug. Enchirid. c.~11.}
&
\Cite{li.~14. c.~27. de ciuit.}}
You thought euil againſt me: but God turned that into good, that he
might exalt me, as preſently you ſee, and might ſaue many peoples. \V
Feare not: I wil feed you & your litle ones, and he comforted them, and
ſpake gently & mildly. \V And he dwelt in Ægypt, with al his fathers
houſe: and liued an hundred and tenne yeares. And he ſawe the children of
Ephraim vnto the third generation. Alſo the children of Machir the ſonne
of Manaſſes were borne in Ioſephs knees. \V Which thinges being done, he
ſpake to his brethren: After my death God wil viſite you, and wil make
you goe vp out of this land, to the land which he ſware to Abraham,
Iſaac, and Iacob. \V And when he had adiured them and ſaid: God wil
viſite you,
\LNote{Carie my bones vvith you.}{For
\MNote{Ioſeph, for his brethrẽs ſake, differred his deſired burial in
Chanaan.}
the ſame reaſons Ioſeph would be finally buried in Chanaan, for which
Iacob deſired to be there buried
\XRef{(chap.~47.)}
but Ioſeph would not preſently be caried thither, leſt it might haue
geuen offence to the Ægyptians, or at leaſt haue diminiſhed their
fauoure towardes his brethren: and withal he would confirme his brethren
in their hope of returning, ſeing he was content, that his bodie ſhould
expect in Ægypt, til the whole Nation ſhould returne into Chanaan.}
carie my bones with you out of this place: \V he died, being an hundred
and tenne yeares old. And being embawmed with ſpices, was put in a
coffin in Ægypt.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
