%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 0058
%%% o-0053
\startChapter[
  title={Chapter 9}
  ]

\Summary{God reneweth the bleſsing of multiplication, 3.~alloweth the
  eating of flesh, but not of bloud, 8.~promiſeth neuer againe to
  deſtroy the world by water. 22.~Chem ſaw, and reported his fathers
  nakednes, which Sem and Iapheth couered. 24.~For which he his curſed,
  and they are bleſſed.}

And God bleſſed Noe and his ſonnes. And he ſaid to them:
\SNote{Of this commandment, or rather bleſſing ſee the
\XRef{Annotations. chap.~1. v.~28.}}
Increaſe, & multiplie, and replenish the earth. \V And your terror and
dread be it vpon al the beaſts of the earth, and vpon al the foules of
the ayre, with al that moue vpon the earth: al the fishes of the ſea are
deliuered to your hand. \V And
\LNote{Al that moueth.}{S.~Iuſtinus
\CNote{Ad. q.~119.}
Martyr,
\MNote{Voluntarie abſtinence without cõmandment.}
\CNote{Ho.~17. in Gen.}
S.~Chriſoſtom, and other ancient Doctors proue, that fleſh was lawful to
be eaten before the floud: but being not neceſſarie, becauſe men
were ſtronger, and other things alſo of more force, the better ſorte
which were of Seths race abſtained from it. But after the floud fleſh
being more neceſſarie, God altereth that cuſtome of abſtinence, with
this limitation and commandment, that they ſhal not eate bloud.}
al that moueth, and liueth ſhal be yours for meat: euen as the grene
herbes haue I deliuered al to you. \V Sauing that
\LNote{Flesh vvith bloud.}{Though
\MNote{Abſtinence from bloud ſome times commanded not alwayes.}
this poſitiue precept, of not eating bloud, ſerued wel to make men more
abhorre manſlaughter (which is forbid by the law of nature, and the
reuenge therof here and in other places ſeuerly threatned) yet it was
ſpecially geuen both immediatly after the floud, and in the
\CNote{Leuit.~17.}
law of
Moyſes (with manie the like) to exerciſe men in obedience. And the ſame
was renewed, for a time, by the Apoſtles, to appeaſe a controuerſie in
the primitiue Church.
\CNote{Act.~15.}
For that the Iewes conuerted to Chriſt, hauing
benne long accuſtomed to this obſeruation, could not indure to ſee it
broken, by themſelues, or other Chriſtians, and being no great burden,
for the Gentiles, it was decreed that al ſhould kepe it. And ſo peace
was made. Neuertheles it was aboliſhed when the cauſe ceaſed. As
\CNote{li.~32. c.~13.}
S.~Auguſtin declareth againſt Fauſtus the Manichee. Such is the
authoritie of the Church to decree, and againe to diſanul an obſeruation
of a thing of itſelfe indifferent.}
flesh with bloud you ſhal not eate. \V
\CNote{Leuit.~17.}
For I wil require the bloud of your ſoules at the
%%% o-0054
hands of al beaſts: and at the hand of man, at the
hand of eech man, and of his brother, wil I require the ſoule of man. \V
Who ſoeuer ſhal ſheed mans bloud, his bloud ſhal be ſhed: for to the
image of God man was made. \V But increaſe you and multiplie, and goe vpon
the earth, and fil it.

\V Thus alſo ſaid God to Noe, and to his ſonnes with him: \V Behold I
wil eſtablish my couenant with you, and with your ſeede after you: \V
and with euerie liuing ſoule, that is with you, as wel in al foules as
in cattle & beaſts of the earth that are come forth out of the arke, and
in al beaſts of the earth. \V I wil eſtablish my couenant with
you, and al flesh  ſhal be no more deſtroyed with the waters of a floud,
neither ſhal there be from henceforth a floud to waſt the earth. \V And
God ſaid: This is the ſigne of the couenant which I geue betwen me and
you, and betwen euerie liuing ſoule, that is with you, for perpetual
generations: \V
\SNote{The rainbow was before, but was not a ſigne, as God ſaith
henceforth it ſhould be, for men to remember his promiſe.
\Cite{Seuerianus in Gen.}
\Cite{S.~Tho. Quodlib.~3. a.~30.}}
my bowe wil I ſet in the clouds, and it ſhal be the ſigne of a couenant
betwen me and betwen the earth. \V And when I ſhal couer the element
with cloudes, my bowe ſhal appeare in the cloudes: \V and I ſhal
remember my couenant with you, and with euerie liuing ſoule that beareth
flesh: and there ſhal no more be waters of a floud, to diſtroy al
flesh. \V And my bowe ſhal be in the cloudes, and I ſhal ſee it, and I
ſhal remember the euerlaſting couenant, that was made betwen
%%% 0059
God and euerie liuing ſoule of al flesh which is vpon the earth. \V And
God ſaid to Noe: This ſhal be the ſigne of the couenant, which I
eſtablished, betwen me & al flesh of the earth. \V The ſonnes therfore of
Noe, that came out of the arke, were Sem, Cham, and
\Fix{Iaphet:}{Iapheth:}{obvious typo, fixed in other}
and Cham he is the father of Chanaan. \V Theſe three are the ſonnes of
Noe: and
\SNote{By this it is clere that Noe had no more children after the
floud.
\Cite{S.~Chriſoſt. ho.~29. in Gen.}}
of theſe was al mankind ſpred ouer the whole earth.

\V And Noe a huſbandman began to til the grounde, and planted a
vineyard. \V And drinking of the wine was made
\LNote{Drunke.}{Noe
\CNote{S.~Chriſoſt. ho.~29. S.~Theod. q.~65. S.~Amb. c.~30. de Noe &
Arca}
\MNote{Noe ſinned not in drunkennes.}
ſinned not, by the common iudgement of ancient Fathers, in that he was
ouercome with wine, becauſe he knew not the force therof, hauing drunck
only water al his life before. But this external rebuke, and worldlie
diſgrace happened to Noe,
\MNote{Noe a figure of Chriſt.}
in figure of Chriſt naked on the Croſſe. As
\Cite{S.~Cyprian (Epiſt.~63. ad Cœcilium.)}
\Cite{S.~Auguſtin (li.~16. de ciuit. c.~2.}
&
\Cite{li.~12 contra Fauſtum Manicheum. c.~23. &~24.)}
Eucherius, Rupertus and others (vpon this place) do teach.
\MNote{Sem and Iapheth a figure of the Church, Chem of Infidels.}
And likewiſe that Sem and Iapheth were a figure of the Church conſiſting
of Iewes and Gentiles, and Cham of Heretikes, and other Infidels, that
deride the infirmities, which our Sauiour ſuſtayned. Goe to now (yee
Manichees, ſaith
\CNote{li.~12. c.~24. cont. Fauſt.}
S.~Auguſtin) obiect calumnies to the ancient holie
Scriptures, doe ſo, yee children of Cham, to whom naked fleſh ſemeth
vile, by which your ſelues were begotten. For neither could yee by anie
meanes haue benne called Chriſtians, except
\Fix{Chriſts}{Chriſt,}{obvious typo, fixed in other}
as he was fortold by the Prophetes, had come into the world, had drunck
that cuppe of his vinyard, which could not paſſe from him, had ſlept in
his paſſion, as in drunkennes of folie (which is wiſer then men) and ſo
the infirmitie of mortal fleſh (which is ſtronger then men) had become
naked, by the ſecret counſel of God, which infirmitie vnles the \Sc{Word
of God} had taken vpon him, the very name Chriſtian, wherof yee alſo
glorie, had not benne at al in the earth.}
drunke, and naked in his tabernacle. \V Which when Cham the father of
Chanaan, had ſeene, to wit that his fathers priuities were bare, he told
it to his two bretheren abroad. \V But in dede Sem and Iapheth put a
cloake vpon their ſhoulders, and going backward, couered the priuities
of their father: and their faces were turned away, and they ſawe not
their fathers priuities. \V And Noe awaking from the wine, when he had
learned what his younger ſonne had done to him, \V he ſaid:
\LNote{Curſed be Chanaan.}{VVhy
\MNote{VVhy Chanaan is curſed rather then Cham.}
Chanaan the ſonne is curſed, and not his father Cham, diuers yeld
diuers reaſons.
\CNote{q.~57. in Gen.}
S.~Theodoretus reporteth out of the Hebrew Doctors, that
Chanaan a boy firſt ſaw his grandfathers nakednes, and told the ſame to
his father, and ſo they both derided that they ſhould haue couered,
Chams other ſonnes not offending, and therfore not his whole progenie
but only Chanaan & his poſteritie were here curſed by
Noe.
\CNote{ho.~29.}
S.~Chriſoſtome ſuppoſeth that for ſo much as God had bleſſed Noe &
his three ſonnes coming forth of the Arke, he could not preſume to curſe
anie of thoſe, whom God had bleſſed, therfore curſed Chanaan who in
wickednes was like to his father.
\MNote{Sinnes puniſhed in the poſteritie.}
S.~Gregorie bringeth this, for example of wicked men, eſcaping
puniſhment in this life, and are puniſhed in the next, and in their
poſteritie, following their vices. \Emph{VVhat meaneth it} (ſaith
\CNote{Li.~26. Moral. c.~18.}
he) \Emph{that Cham ſinning, Chanaan his ſonne had ſentence of reuenge,
but that the ſinnes of the reprobate proſper here in this life
vnreuenged, and are punished aftervvards?}
\MNote{The effect of bleſſing and curſing.}
And clere it is that Chanaans moſt wicked poſteritie were ſubdued in the
end, and moſt of them deſtroied by the children of Iſrael (who were of
Sem) vnder the conduct of Ioſue, according to Noes prophetical bleſſing
of Sem, and curſing of Chanaan.}
Curſed be Chanaan, a ſeruant of ſeruantes ſhal he be vnto his
bretheren. \V And he ſaid:
\CNote{Apoc.~13,~8.}
Bleſſed be the Lord God of Sem, Chanaan be his ſeruant. \V
\LNote{God enlarge.}{This
\MNote{Iapheths bleſſing.}
bleſſing of \Emph{Iapheth} was
\CNote{Act.~13. Rom.~11. Epheſ.2.}
\MNote{Literal.}
literaly fulfilled, when (according to his name, which
ſignifieth \Emph{Latitude} or enlargement) his plentiful iſſue poſſeſſed
moſt ample countries, both in the Continent, and Ilandes.
\MNote{Myſtical.}
But myſtically it had effect (as
\CNote{Tradit. Hebra.}
S.~Hierom,
\CNote{Li.~16. ciuit. c.~2. Li.~12. c.~24. cont. Fauſt.}
S.~Auguſtin,
\CNote{Li.~4. c.~38. in Gen.}
Rupertus and
others expound it) when the Apoſtles, being Iewes of the race of Sem,
firſt builded the Chriſtian Church, wherin the moſt part of that nation
refuſed to dwel, contemning Chriſts Goſpel & grace, and the fulnes of
Gentiles entred in, and were made inheriters. Finally Chanaan is
ſeruant to both Sem and Iapheth,
\MNote{How Heretikes ſerue Catholiques.}
in that Heretikes being vnder the Iuriſdiction of the Church, geathered
of Iewes and Gentiles, ſerue to ſtur vp Catholiques diligence to more
exact knowledge of al truth; and their patience to more merite and
glorie.}
God enlarge Iapheth, and dwel he in the tabernacles of Sem, and Chanaan
be his ſeruant. \V And Noe liued after the floud three hundred fiftie
yeares. \V
\CNote{Heb.~11.}
And al his dayes were in the whole nyne hundred fiftie yeares: and he
died. 


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
