%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 0046
%%% o-0042
\startChapter[
  title={Chapter 7}
  ]

\Summary{Noe vvith his familie, and paires of al kindes of beaſtes and
  foules, being entred into the arke, 12.~it raineth fourtie daies and
  fourtie nights. 21.~Al men and other liuing creatures on the earth,
  without the arke, are deſtroyed.}

And our Lord ſaid to him: Get thee in, thou and al thy houſe into the
arke: for I haue ſene thee iuſt
\SNote{Noe was iuſt not only by the eſtimation of men, but in deede and
before God.}
in my ſight in this generation. \V Of al beaſts that are
\SNote{Obſeruation of cleane and vncleane beaſtes by tradition, before
the law of Moyſes.}
cleane, thou ſhalt take ſeauen, and ſeauen, male & female: \V but of the
beaſts that are vncleane two and two, male & female. Yea and of the
foules alſo of the ayre ſeauen & ſeauen, male and female: that ſeede may
be ſaued vpon the face of the whole earth. \V For yet a while, and after
ſeauen dayes, I wil rayne vpon the earth fourtie dayes and fourtie
nights: and I wil cleane deſtroy al ſubſtance, that I haue made, from
the face of the earth. \V Noe therfore did al thinges, which our Lord
had commanded him. \V And he was ſix hundred yeares old, when the waters
of the floud flowed ouer the earth. \V
\CNote{Mat.~24. Luc.~17.}
And Noe entred and his ſonnes, his wife and the wiues
\Fix{of of}{of}{obvious typo, fixed in other}
his ſonnes with him into the arke, becauſe of the waters of the
floud. \V Of beaſts alſo the cleane and the vncleane, & of foules, and
of al that moueth vpon the earth, \V two & two went to Noe into the
arke, male and female, as our Lord had commanded Noe. \V And after the
ſeauen dayes were paſſed, the waters of the floud flowed ouer the
earth. \V In the ſix hundred yeare of the life of Noe, in the ſecond
moneth, in the ſeauententh day of the moneth, al the fountaines of the
\SNote{The Hebrew word \HH{Thehom} ſignifieth a gulfe of water, from
whence new fountaynes ſprang, more abundantly thẽ euer ſince or before.}
greate deapth were broken vp, and
\SNote{\HH{Arubbah} ſignifieth great pipes or windowes, by which water
fell downe in great abũdance from the ayre, here called heauen. 
\Cite{S.~Hier. queſt. Heb.}
\Cite{S.~Epiph. ad Io. Hieroſolom.}
\Cite{S.~Chr. ho.~25. in Gen.}}
the floud gates of heauen were opened: \V and the raine fel vpon the
earth fourtie dayes and fourtie nights. \V In the verie point of that
day entred Noe, and Sem, and Cham, & Iapheth his ſonnes: and his wife,
and the three wiues of his ſonnes with them into the arke: \V they and
euerie beaſt according to
%%% 0047
their kind, and al cattle in their kinde, and al that moueth vpon the
earth according to their kind, and al foule according to their kind, al
birds, and al that fly \V went to Noe into the arke, two and two of al
fleſh, wherin there was breath of life. \V And ſuch as entred in, male
and female of al fleſh did enter in, as God had commanded him: and our
Lord
\LNote{Shut him in.}{God who by his only wil could in a moment haue
drowned al the reſt of the world, ſauing whom he pleaſed, not needing in
any thing the helpe of his creatures,
\MNote{God vſeth both natural and ſupernatural meanes, as ſecundarie
cauſes, in producing, conſeruing, gouerning, puniſhing, & in rewarding
his creatures.}
yet would vſe both natural, &
ſupernatural meanes, as the labour of Noe to build the arke, new
fountaines ſpringing, and the heauens powring downe water fourtie dayes
togeather, afterwards the winde to dry vp the
\Fix{earh,}{earth,}{obvious typo, fixed in other}
and becauſe the dore being great (for
\Fix{Elephantsr}{Elephants}{obvious typo, fixed in other}
to enter in) and was to be firmed without (as
\CNote{li. de Noe & arca. c.~15.}
S.~Ambroſe noteth) for
better induring the forcible waters, could not commodiouſly be cloſed by
Noe, \Emph{our Lord} (by the miniſtrie of Angels) \Emph{shut him in on
the out ſide}, to teach vs by al this, and the like diſpoſition of
things, that albeit his Diuine omnipotencie can do what he wil al alone,
yet he wil haue his creatures to concurre and cooperate as ſecundarie
cauſes, ſometimes naturally, ſometimes ſupernaturally, or miraculouſly,
as it pleaſeth his goodnes to impert to them powre and vertue.}
ſhut him in on the out ſide. \V And the floud grew fourtie daies vpon
the earth: and the waters increaſed, and lifted vp the arke on high from
the earth. \V For they ouerflowed excedingly: and filled al on the face
of the earth: moreouer the arke fleeted vpon the waters. \V And the
waters preuailed out of meaſure vpon the earth: and al the hiegh
mountaines vnder the whole heauen were couered. \V Fiftene cubites
higher was the water aboue the mountaines, which it couered. \V And al
flesh was conſumed that
%%% o-0043
moued vpon the earth, of foule, of cattle, of
beaſts, and of al creepers, that creepe vpon the earth: al men, \V and al
things, wherin there is breath of life on the earth, died. \V
\CNote{Sap.~10. Ecl.~39.}
And he
cleane deſtroied al ſubſtance, that was vpon the earth, from man euen to
beaſt, as wel it that creepeth, as the foules of the ayre: and they were
deſtroied from of the earth:
\LNote{But only Noe.}{As there is not anie thing in al the old Teſtament,
from the creation of the world til the coming of Chriſt, more notable,
more admirable, or of greater importance, then this hiſtorie of the
general floud;
\MNote{Al or moſt thinges in the old Teſtament be in figure of the new,
& no figure more exact then the floud of Noe.}
\CNote{1.~Cor.~10.}
ſo was there nothing (though al, or moſt chanced to them in figure) that
euer more aptly, more liuely, or more exactly prefigured Chriſt and his
Church, with the reſt of al mankind, then did Noe and the arke, & the
drowning of the reſt of the world in that deluge. VVhich
\CNote{Ep.~99. li.~5. de Baptiſ. c.~28. de vnitate Eccleſ. ca.~5. in
Pſal.~103. &~131. Ser.~69. de tẽpore.}
S.~Auguſtin
declareth in many places, but moſt eſpecially and of purpoſe in his
\Cite{twelfth booke againſt Fauſtus the Manichee, from the 14.~chapter
to the 22.}
and in his
\Cite{fifteenth booke of the citie of God, in the two laſt chapters}:
where he ſheweth at large both the certaintie of the hiſtorie, and that as
certainely it was a figure of things in the new Teſtament, and withal
the great congruitie betwen the figure & the things figured.
\MNote{How the Doctors applie the figure to the things figured.}
The ſame did Origen explicate
\Cite{(homil.~2. in Gen.)}
\Cite{S.~Gregorie (homil.~12. in Ezech.)}
\Cite{Rupertus (li.~4. comment. in Gen. c.~71. & ſequent.)}
and diuers other ancient Doctors, confirming their expoſitions by
\CNote{1.~Pet.~3.}
S.~Peters teſtimonie, ſaying: In the arke \Emph{a fevv, that is eight
ſoules} (or perſons) \Emph{vvere ſaued} (from drowning) \Emph{by vvater,
vvherunto Baptiſme being of the like forme novv ſaueth you alſo.} And by
our Sauiours wordes ſaying:
\CNote{Mat.~24. Luc.~17.}
\Emph{As in the dayes of Noe, ſo shal alſo the comming of the Sonne of
man be.} 
\MNote{Noe ſignifed Chriſt.}
In ſumme the Doctors teach, that Noe ſignifying \Emph{reſt} was a figure
of Chriſt, the very reſt of mans ſoule. VVhom who ſoeuer foloweth ſhal
find reſt for their ſoules.
\MNote{The arke the Church.}
The arke ſignifieth the Church, the forme therof being ſix times ſo long
as broad, and tenne times ſo long as hiegh, reſembleth the proportion of
mans bodie, lying prone or proſtrate. The dore in the ſide repreſenteth
the wound in Chriſts ſide, 
\MNote{Entrance into the Church by Baptiſme.}
from whence flowed the holie Sacraments, by which the faithful enter
into the Church, and are ſanctified. The timber wherof the arke is made,
& the water bearing it vp, ſignified the Croſſe of Chriſt and Baptiſme.
\MNote{Vertue of Sacraments cometh from Chriſts Paſſion.}
\Emph{For as Noe} (ſaith
\CNote{li.~22. de Baptiſ. c.~14.}
S.~Auguſtin) \Emph{vvith his, vvas deliuered by
the vvater and the vvode, ſo the familie of Chriſt, by Baptiſme ſigned
vvith Chriſts Paſſion on the Croſſe.}
\MNote{Doctors and Paſtores in the Church.}
Likewiſe the ſquarnes of the
timber which both ſuſtayned the burden of al contayned in the arke, and
reſiſted the boyſtrous waues of the floud beating without, did ſignifie
ſuch men in the Church, as be conſtant & ſtand firmly in al ſortes of
tentations: eſpecially godlie & lerned Doctors and Paſtors, who by worde
and example vphold and cõfirme the faithful people in al afflictions
within, and withſtand and conuince al Heretikes, and other Infidels that
oppugne the Church without.
\MNote{Varietie of ſtates and orders in the Church.}
Againe the hiegher & lower rowmes with the midle chambers & third
loftes, & other diſtinctions of cabinets, and partitions, and al ſortes
of liuing creatures cleane and vncleane, receiued therin, did ſignifie
the varieties of al ſtates & functions, and diuerſitie of manners and
merites in the Church, in which are perſons of al degrees, Clergie and
Laitie, Potentates, Princes, ſubiects,
\MNote{Good and euil in the Church.}
good and euil.
\MNote{Perpetuitie
\Fix{o}{of}{obvious typo, fixed in other}
the Church.}
The moſt ſtrong kind of glew called \Emph{bitumen}, ſignified the
permanent or euerlaſting ſtabilitie, and vnſeparable cõnexion of the
Church, by the grace and continual aſſiſtance of the Holie Ghoſt
conſeruing the ſame.
\MNote{Vnitie of the Church.}
The conſummation of the arke in one cubite ſignified the vnitie of the
ſame Church, which is one in al times, and places. Neither would God
almightie haue manie arkes, for Noe and his ſonnes or other creatures,
nor manie chiefe rulers (though he would that of them ſhould come manie
Nations) but one only arke, 
\MNote{One chiefe gouernour in the Church.}
and one chiefe gouernour therof, 
\MNote{No ſaluation out of the Church.}
and that al
without the ſame ſhould corporally dye, to ſignifie that al which dye
without the Church do perish, and are eternally damned. Wherupon
\CNote{Epiſt.~57. ad Damaſum.}
S.~Hierom, amongſt other Fathers, ſheweth that al within the Church,
that communicate with the Sea Apoſtolique (wherin S.~Damaſus ſate then
gouernour) are as thoſe in the arke of Noe, and al Schiſmatikes,
Heretikes and other Infidels are in like caſe, with the reſt of the
world, that were drowned with the floud.}
but onlie Noe remained, and they that were with him in the arke. \V
\CNote{1.~Pet.~3.}
And the waters held on aboue the earth an hundred fiftie dayes.

%%% !!! This formatting could be improved.
%%% In particular, it should follow the annotations.
\vskip 0.25in

\startalignment[middle]
  The end of the firſt age.
\stopalignment


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
