%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 0061
%%% o-0056
\startChapter[
  title={Chapter 10}
  ]

\Summary{The genealogie of Noes children, by whom the world was
  increaſed againe, after the floud.}

Theſe are the generations of the ſonnes of Noe, Sem, Cham, and Iapheth:
and children were borne to them after the floud. \V The children of
Iapheth: Gomer, and Magog, and Madai, and Iauan, and Thubal, and Moſoch,
and Thiras. \V Moreouer the children of Gomer: Aſcenez and Riphath and
Thogorma. \V And the ſonnes of Iauan: Eliſa and Tharſis, the Cetims and
the Dodanims. \V Of theſe were diuided the Iles of Nations in their
countries, ech one according to his tongue and their families in their
nations. \V And the ſonnes of Cham: Chus, and Meſraim, and Phut, and
Chanaan. \V And the ſonnes of Chus: Saba, and Heuila, & Sabatha, and
Regma, & Sabathaca. The ſonnes of Regma: Saba, and Dadan. \V Moreouer
Chus begat
\LNote{Nemrod.}{To
\MNote{Nemrod king of Babylon, a ſuttle & cruel giant.}
this Nemrod the ſonne of Chus, firſt builder and king of Babylon,
\CNote{Li.~1. c.~4. Antiqit.}
Ioſephus,
\CNote{l.~Antiq. Chalda.}
S.~Epiphanius,
\CNote{Li.~1. c.~2. in Oſee.}
S.~Hierom,
\CNote{li.~16. c.~3. 4.~&~17. ciuit.}
S.~Auguſtin, and generally al ancient
writers aſcribe the firſt tyrannie, and firſt ſetting vp of an earthlie
citie oppoſite to the Citie of God after the floud. He was
a \Emph{Valiant}, or rather a \Emph{Violent hunter} a giant hunter
(ſaith S.~Auguſtin, according to the ſeuentie Interpreters) who by
falſhood and force brought manie vnder his dominion.
\MNote{An Arch-heretike.}
For he ſturred vp pride (ſaith Ioſephus) and contempt of God in men,
auoching that they were not beholding to God for preſent felicitie, but
to their owne vertue; and ſo ſuppoſing men would fal from God to him, if
he offered him ſelfe a leader, and helper againſt a new floud, by litle
and litle drew al to tyrannie. He was otherwiſe called Saturnus, and was
at length amongſt others accounted a god.
\CNote{Beroſus Annian. li.~4. c.~5. Euſebius in chron. & alij.}
After him his ſonne Belus
Iupiter (as moſt authors affirme) raigned 65.~yeares. And then ſucceded
Ninus the firſt king of the Aſſirians.}
Nemrod: he began to be mightie in the earth, \V and he was a valiaunt
hunter
\SNote{That is in his ſight who can not be deceiued.}
before our Lord. Therof roſe a prouerbe: As it were Nemrod the valiaunt
hunter before our Lord. \V And the beginning of his kingdome was
Babylon, and Arach, and Achad, and Chalanne in the land of Sennaar. \V
Out of that land came forth
\LNote{Aſſur builded Niniue.}{Here is great difficultie, and much
diſpute amongſt writers, who this Aſſur is.
\MNote{Aſſur, ſonne of Sem, or Ninus king of Aſſirians.}
Briefly, we may either ſay with
\CNote{li.~1. Antiq. c.~4.}
Ioſephus and
\CNote{Li.~16. c.~3. ciuit.}
S.~Auguſtin, that Aſſur the
ſonne of Sem built a citie, which afterwards
%%% !!! Where does this go?
\CNote{Euſeb. in chron.}
Ninus of Chams race
enlarged, enriched, and changing the name called it Ninum: or els that
this word \Emph{Aſſur} here ſignifieth (as
\XRef{4.~Reg.~15.}
\XRef{Pſal.~82.}
\XRef{Eſaiæ~10. &~31.})
the king of Aſſirians, to wit Ninus the ſonne of Belus, who as al
hiſtories greeke and barbarous reporte (ſaith S.~Hierom) was the firſt
that raigned ouer al Aſia, & among the Aſſirians built Ninum a citie of
his owne name, which the Hebrewes cal Niniue. He ſet vp the Monarchie of
the Aſſirians, called the golden kingdom, which ſtood 1240.~yeares. And
made his father Belus to be honored for a god.
\CNote{Iud.~19. Num.~25. 4.~Reg.~1. 3.~Reg.~16.}
\MNote{Firſt falſe goddes.}
To whom the Babilonians,
as Plinie teſtifieth, firſt erected ſtatues, altares, & temples. Of this
Belus or Bel of Babylon were alſo deriued other falſe goddes, as Belial
the god of Libertines, or without yoke, Beelphegor god of the Moabites,
Beelzebub of the Acharonites, Baal in Samaria, Baalim amongſt the
Philiſtims: and the like in other nations.}
Aſſur, and builded Niniue, and the ſtreets of the citie, and Chale. \V
Reſen alſo betwen Niniue and Chale: this is the great citie. \V But
Meſraim alſo begat the Ludims, & the Anamims, & the Laabims, the
Nephthuims, \V and the
%%% o-0057
Phetruſims, and the Caſluims: of whom came forth
the Philiſtims & the Caphtorims. \V And Canaan begat Sidon his firſt
begotten, Hethæus, \V and Iebuſæus, and Amorrhæus, and Gergeſæus, \V
Heuæus and Aracæus: Sinæus, \V and Aradius, Samaræus, and Hamathæus: and
afterwardes were ſpred the people of the Chananits. \V And the limitts
of Chanaan were from Sidon as we come to Gerara euen to Gaza, vntil thou
enter to Sodoma & Gomorrha, and Adama, & Seboim euen to Leſa. \V Theſe
are the children of Cham in their kinreds, and tongues, and generations
& lands, and nations. \V Of Sem alſo
\SNote{Hence S.~Auguſtin geathereth that the people of Iſrael were
called Hebrewes of this Heber.
\Cite{l.~16. c.~3. ciuit.}}
father of al the children of
\Fix{Herber,}{Heber,}{likely typo, fixed in other}
the elder brother of Iapheth were
borne. \V The children of Sem: Ælam and Aſſur, and
%%% 0062
Arphaxad, and Lud, and Aram. \V The children of Aram: Vs, and Hul, and
Gether, and Mes. \V And Arphaxad alſo begat Sale, of whom was borne
Heber. \V And to Heber were borne two ſonnes: the name of the one was
Pharleg,
\SNote{Heber hauing a ſonne borne when the tongues were diuided called him
Phaleg, which ſignifieth diuiſion.
\Cite{S.~Aug. li.~16. c.~11. ciuit.}}
becauſe that in his dayes was the earth diuided: and his brothers name
was Iectan. \V The which Iectan begat Elmodad, and Saleph, and Aſarmoth,
Iare, \V and Aduram, and Vzal, and Decla, \V and Ebal, and Abimael,
Saba, \V and Ophir, and Heuila, and Iobab. Al theſe were the children of
Iectan. \V And their dwelling was from Meſſa as we goe on as far as
Sephar a mountaine in the eaſt. \V Theſe are the children of Sem
according to their kinred and tongues, and countries in their
nations. \V Theſe are the families of Noe, according to their peoples &
nations. Of theſe were
\LNote{The Nations.}{How manie Nations and tongues were in the world
immediatly after the towre of Babylon, is more cõmonly ſuppoſed, then
clerly ſhewed by old or late writers. Only it ſemeth certaine and
euident, that there were iuſt as manie tongues as Nations.
\MNote{The common opinion of 72.~tongues is not clere in Scripture.}
But to finde preciſly (as the common opinion holdeth) 72, is hard. For
in this chapter (where they would count this number) are not mentioned
ſo manie.
\MNote{Yet this number of nations and tongues may be probably geathered
in this and other places of Geneſis.}
Neuertheles if we adde certaine that begane diſtinct Nations afterwards,
\CNote{S.~Aug. 16. c.~3. ciuit.}
til Iacob with his children went into Ægypt (in whom only the Hebrew
Nation and Tongue continued to Chriſts time) the number wil come
right. Of Iapheth were borne 7.~ſonnes chiefe of Nations. Againe of
Gomer (beſides his ſuppoſed ſucceſſor, who can not be counted beginner
of an other diſtinct nation) came 2.~other heades. Likwiſe of Iauan
(beſides his firſt ſonne) roſe 3.~more nations. Of Cham by his firſt
ſonne Chus were 6.~nephewes princes of nations. Againe of Regma (beſides
his ſucceſſor) came one more. And Nemrod beſides his kingdom of Babylon,
rayſed vp other 6. By his ſecond ſonne Meſraim came 8.~nations, Chams
third ſonne Phut made only one nation. And Chanaan his ſonnes made
11.~more. Of Sem (laſt mentioned for better connecting the maine
Hiſtorie, and ſucceſſion of the Church) came 
\MNote{The Hebrews chiefe of theſe nations.}
the
\Fix{chife}{chiefe}{likely typo, fixed in other}
and principal Nation the Hebrewes, deſcending from him by Arphaxad,
Sale, Heber, and ſo directly to Iacob. Of Sem alſo were borne 4.~other
ſonnes beginners of nations. Againe of Aram (beſides his firſt ſonne)
were 3.~fathers of nations. Likwiſe of Heber (beſides the Hebrewes
deſcending by 
Phaleg) were borne to his other ſonne Iectan 13.~heades of
nations.
\CNote{Gen.~24. 19. 17. 25,~26. 25,~23.}
Theſe are al that are named in this place: to wit, of Iapheth,
12. of Cham~33. and of Sem~21. which make in al 66.~wherunto if we
adioyne Nachor (Abrahams brother) Moab, and Ammon (Lots ſonnes) alſo
Iſmael (Abrahams eldeſt ſonne) and his iſſue by Cetura, and finally Eſau
(Iacobs brother) who made 6.~more diſtinct nations, the whole number is
72. This probable collection, with the reſt, we ſubmit to better
iudgement.}
the nations diuided on the earth after the floud.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
