%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 19
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-19


\project douay-rheims


%%% 0088
%%% o-0084
\startChapter[
  title={Chapter 19}
  ]

\Summary{Lot receiuing Angels in his houſe, is abuſed by the
  Sodomites. 12.~He with his wife (26.~who for looking back is turned
  into a ſtatua of ſalt) and his two daughters are deliuered. 24.~Sodome
  and Gomorre are burned. 31.~Lot lieth vnwitting with both his
  daughters, begat of them Moab and Ammon, of whom came the Moabites and
  Ammonites.}

%%% 0089
And the two
%%% not marked in text, but in other
\SNote{Abraham & Lot by hoſpitalitie merited to receiue Angels in ſteed
of mẽ.
\XRef{Heb.~13.}}
Angels came to Sodome at euen; and Lot ſitting in the gates
of the citie. Who when he had ſene them, roſe vp and went to meete them:
and adored proſtrate vnto the ground, \V and ſaid: I beſech you, my
Lords, turne into the houſe of your ſeruant, and lodge there: wash your
feet, and in the morning you shal go forth on your way. Who ſaid: No,
but we wil abide in the ſtreat. \V He compelled them earneſtly to turne
in vnto him: and when they were entred into his houſe, he made them a
banquet, and baked vnleauened bread, and they did eate. \V And before
they went to bed, the men of the citie beſet the houſe from young to
old, al the people togeather. \V And they called Lot, and ſaid to him:
Where are the men that came in to thee at night? bring them forth hither
that we may know them. \V Lot going forth to them, and ſhutting the dore
after him, ſaid: \V Doe not ſo, I beſech you, my brethren, doe not
commit this euil. \V I haue two daughters, which as yet haue not knowen
man: I wil bring them forth to you, and abuſe you them as it ſhal pleaſe
you, ſo that you do no euil to theſe men, becauſe they are entred vnder
the ſhadowe of my roofe. \V But they ſaid: Get thee backe thither. And
againe: Thou cameſt in, ſaid they, as a ſtranger, what to be a iudge?
Thy ſelfe therfore we wil afflict more then theſe. And they did violence
to Lot exceadingly: and it was euen nowe at the point that they would
break the dores. \V And behold the men put forth their hand, and drew in
Lot vnto them, and ſhut the dore: \V and them, that were without, they
ſtroke with blyndnes from the leaſt to the greateſt, ſo that they could
not find the dore. \V And they ſaid to Lot: Haſt thou here anie of
thine? ſonne in law, or ſonnes, or daughters, al that are thine, bring
them out of this citie: \V for we wil deſtroy this place, for that
\SNote{This ſinne crieth to heauen for reuenge.}
their crye is waxen lowde before our Lord, who hath ſent vs to deſtroy
them. \V Therfore Lot went forth, and ſpake to his ſonnes in lawe that
were to take his daughters, and ſaid: Ariſe: get you forth out of this
place, becauſe our Lord wil deſtroy this citie. And he ſemed vnto them
to ſpeake as it were in ieſt. \V And when it was morning, the Angels
vrged him, ſaying: Ariſe, take thy wife, and the two
%%% o-0085
daughters which
thou haſt: leaſt thou alſo periſh withal in the wickednes of the
citie. \V He lingering, they tooke his hand, and the hand of his wife,
and of his two
%%% 0090
daughters, becauſe our Lord ſpared him. \V And they led him forth, and
ſet him without the citie: and there they ſpake to him, ſaying: Saue thy
life: looke not backe, neither ſtay thou in al the countrie about: but
ſaue thy ſelfe in the mountaine, leſt thou alſo periſh withal. \V And
Lot ſaid to them: I beſeech thee my Lord, \V becauſe thy ſeruant hath
found grace before thee, and thou haſt magnified thy mercie, which thou
haſt wrought with me, in that thou wouldeſt ſaue my life, and ſafe I can
not be in the mountaine, leſt perhaps the euil catch me, and I dye. \V
There is this citie hereby at hand, wherunto I may flee, a litle one,
and I ſhal be ſafe in it: is it not a litle one, and my life ſhal be
ſaued? \V And he ſaid to him: Behold alſo in this point I haue heard
thy prayers, not to ouerthrow the citie for which thou haſt ſpoken. \V
Make haſt, and be ſaued there: becauſe I can not doe any thing til thou
enter in thither. Therfore the name of that citie was called
\TNote{a litle one}
Segor. \V The ſunne was riſen vpon the earth, & Lot entred into
Segor. \V Therfore our Lord rained vpon Sodome and Gomorre brimſtone &
fire from our Lord out of heauen: \V and he ſubuerted theſe cities, and
al the countrie about, al the inhabitants of the cities, and al things
that ſpring of the earth. \V And his wife
\SNote{Lots wife turned into ſalt, admoniſheth the ſeruants of God to
procede in vertue, & not to looke back to vice.
\XRef{Luc.~17.}
\Cite{S.~Aug. li.~16. c.~30. ciuit.}}
looking behind her, was turned into a ſtatua of ſalt. \V And Abraham
getting vp early in the morning, there where before he had ſtood with
our Lord, \V beheld Sodome & Gomorre, and the whole land of that
countrie: and he ſaw the cinders riſe vp from the earth as it were the
ſmoke of a fornace. \V For when God ſubuerted the citties of that
countrie, he
\SNote{Lot neither perfect, nor very wicked was deliuered for Abrahams
ſake.
\Cite{S.~Aug. q.~45. in Gen.}}
\Fix{remenbring}{remembring}{possible typo, fixed in other}
Abraham, deliuered Lot out of the ſubuerſion of the cities wherein he
had dwelt. \V And Lot aſcended out of Segor, and abode in the mountaine,
his two daughters alſo with him (for he was afraid to abide in Segor)
and he abode in a caue him ſelfe, and his two daughters with him. \V And
the elder ſaid to the younger: Our father is old, and there is no man
left on the earth, that may companie with vs after the maner of the
whole earth. \V Come, let vs make him drunke with wine, and let vs lie
with him, that we may preſerue ſeed of our father. \V They therfore made
their father to drinke wine that night: and the elder went in, and lay
with her father: but he perceaued not, neyther when his daughter lay
downe, nor when ſhe roſe vp. \V The next day alſo the elder
%%% 0091
ſaid to the younger: Behold I lay yeſternight with my father, let vs
make him drinke wine alſo this night, and thou ſhalt lye with him, that
we may ſaue ſeed of our father. \V They made their father drinke wine
that night alſo, and the younger daughter went in, and lay with him: and
neyther then truly did he perceaue when ſhe lay downe, or when ſhe roſe
vp. \V The two daughters therfore of Lot were with child by their
father. \V And the elder bare a ſonne and ſhe called his name
%%% o-0086
Moab: he is, the
\SNote{Moabites & Ammonites were two diſtinct nations, perhaps of the
72. See
%%% !!! Fix this reference.
\XRef{p.~x}}
father of the Moabites euen to this preſent day. \V The younger alſo
bare a ſonne, and ſhe called his name Ammon, that is the ſonne of my
people: he is the father of the Ammonites euen to this day.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
