%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 44
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-44


\project douay-rheims


%%% 0154
%%% o-0142
\startChapter[
  title={Chapter 44}
  ]

\Summary{Ioſeph cauſeth their ſackes to be filled with corne, and their
  money to be put againe therin, and in Beniamins ſack alſo a ſiluer
  cuppe, 4.~and when they were parted, ſending after them, chargeth them
  with theft. 12.~And the cuppe being fond in Beniamins ſack, they al
  much afflicted returne to Ioſeph. 17.~Who threatning to kepe Beniamin,
  18.~Iudas intreateth, 32.~and finally offereth himſelfe to ſeruitude
  for him.}

And Ioſeph commanded the ſteward of his houſe, ſaing: Fil their ſackes
with corne, as much as they can hold: and put the money of euerie one in
the top of his ſack. \V And in the ſackes mouth
\SNote{By this Ioſeph tried his bretherens affection, whether they would
intrete for Beniamin, or ſuffer him to be captiue: as they had before
ſold himſelfe to captiuitie.
\Cite{Theod. q.~105. in Gen.}}
of the younger put my ſiluer cup, and the price which he gaue for the
wheat. And it was ſo done. \V And when the morning roſe, they were
diſmiſt with their aſſes. \V And they were now departed out of the
citie, and had gone forward a litle way: then Ioſeph ſending for the
ſteward of his houſe, ſaid: Ariſe, quoth he, and purſew the men: and
ouertaking them ſay to them: Why haue you rendred euil for good? \V The
cup which you haue ſtollen, is that wherin my lord doth drinke, and
wherin he is wont to diuine: you haue done a verie euil thing. \V He did
as he had commanded. And hauing ouertaken them, he ſpake to them in the
ſame order. \V Who anſwered: Why doth our lord ſpeake ſo, as though thy
\Fix{ſeruautes}{ſeruantes}{obvious typo, fixed in other}
had committed ſo haynous a fact? \V The money, that we found
in the top of our ſackes, we recaried to thee from the land of Chanaan:
and how foloweth it, that we haue ſtollen out of thy lordes houſe, gold
or ſiluer? \V With whom ſoeuer of thy ſeruantes that shal be found,
which thou ſeekeſt, let
%%% 0155
him die, and we wil be the bondmen of our lord. \V Who ſaid to them: Be
it done according to your ſentence: with whom ſoeuer it shal be found,
be he my bondman, and you shal be guiltles. \V Therfore in haſte taking
downe their ſackes vpon the ground, euerie man opened. \V Which when he
had
%%% o-0143
ſearched, beginning from the elder euen to the youngeſt, he found the
cup in Beniamins ſack. \V But they
\SNote{O torments of mercie! he vexeth whom he loueth.
\Cite{S.~Greg. ho.~22. in Ezech.}}
renting their garments, and loading their aſſes againe, returned into
the towne. \V And Iudas foremoſt with his brethren entred in to Ioſeph
(for he was not yet gone out of the place) and they fel togeather before
him on the ground. \V To whom he ſaid: Why would you doe ſoe? know you
not that there is not the like to me in the ſcience
\SNote{Ioſeph being in deede a prophet, knowing more then al ſorcerers
in Ægypt ſpoke of himſelfe as he was eſtemed in that place.
\Cite{S.~Aug. q.~145. ſuper Gen.}}
of diuining. \V To whom Iudas ſaid: What shal we anſwere, my lord? or
what shal we ſpeake, or be able iuſtly to pretend? God hath found the
iniquitie of thy ſeruantes: loe we are al bondmen to my lord, both we, &
he, with whom the cup was found. \V Ioſeph anſwered: God forbid that I
should ſo doe: he that ſtole the cup, the ſame be my bondman: and goe
you free vnto your father. \V And Iudas approching nearer, ſaid boldly:
I beſeech thee my lord, let thy ſeruant ſpeake a word in thine eares,
and be not angrie with thy ſeruant: for after Pharao thou art, \V my
lord. Thou dideſt aske thy ſeruantes the firſt time: Haue you a father
or a brother? \V and we anſwered thee my lord: We haue a father an old
man, and a little boy, that was borne in his old age; whoſe brother by
the mother is dead: and his mother hath him only, and his father loueth
him tenderly. \V And thou ſaidſt to thy ſeruants: Bring him hither to
me, and I wil ſet myn eyes on him. \V We ſuggeſted to my lord: The boy
can not leaue his father: for if he leaue him, he wil die. \V And thou
ſaideſt to thy ſeruants: Vnleſſe your youngeſt brother come with you,
you shal no more ſee my face. \V Therfore when we were gone vp to thy
ſeruant our father, we told him al things that my lord did ſpeake. \V
And our father ſaid: Goe againe, and bye vs a litle wheate. \V To whom
we ſaid: We can not goe: if our youngeſt brother shal goe downe with vs,
we wil ſet forward together: otherwiſe he being abſent, we dare not ſee
the mans face. \V Wherunto he anſwered: You know that my wife bare me
two. \V One went forth, and you ſaid: A beaſt did deuour him: and
hitherto he appeareth
%%% 0156
not. \V If you take this alſo, and ought befal him in the way, you shal
bring downe
\SNote{See
\XRef{note for c.~42. v.~38}}
my hoare hayres with ſorow vnto hel. \V Therfore if I shal enter to thy
ſeruant our father, and the boy be wanting (wheras his life dependeth
vpon the life of him) \V and he shal ſee that he is not with vs, he wil
dye, & thy ſeruants shal bring downe his hoare hayres with ſorow vnto
hel. \V Let me be thy proper ſeruant, that did take him into my
protection and promiſed ſaying: Vnleſſe I bring him againe I wil be
guilty of ſinne againſt my father for euer. \V I therfore thy ſeruant
wil tary in ſteed of the childe in the ſeruice of my lord, and let the
childe goe vp with his brethren. \V For I can not returne to my father,
the childe being abſent; leſt I ſtand by a witnes of the calamitie, that
ſhal oppreſſe my father.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
