%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 21
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-21


\project douay-rheims


%%% 0092
%%% o-0087
\startChapter[
  title={Chapter 21}
  ]

\Summary{Iſaac
\MNote{The ſixt part of this booke.}
is borne, 4.~circumciſed, 8.~and weaned. 9.~Agar and Iſmael are put
forth of Abrahams houſe, 15.~but after deſolation are nourished, and
proſper in the deſert. 22.~King Abimelech and Abraham make a league
confirming it with oath.}

And
\MNote{Of the progenie & other bleſſings of Abraham, Iſaac, and Iacob.}
our Lord viſited Sara, as he had promiſed: and fulfilled the things
which he ſpake. \V And she conceaued and bare a ſonne in her old age, at
the time that God had foretold her. \V And Abraham called the name of
his ſonne, which Sara bare him,
\SNote{Abraham & Sara laughed (he admiring, ſhe doubting) at the ioyful
promiſe of a ſonne, & therfore he is called Iſaac, which ſignifieth
laughter.
\Cite{S.~Aug. li.~16. c.~31. ciuit.}}
Iſaac: \V and he circumciſed him the eight day, as God had commanded
him, \V when he was an hundred yeares old: for at this age of his father,
was Iſaac borne. \V And Sara ſaid: Laughter God hath
%%% 0093
made to me: whoſoeuer shal heare of it, wil laugh with me. \V And again
she ſaid: Who that shal heare of it would beleue Abraham, that Sara gaue
ſucke to a ſonne, which she bare him now being an old man? \V The child
therfore grewe, and was weined: and Abraham made a great feaſt in the
day of his weining. \V And when Sara had
\Fix{ſenne}{ſeen}{likely typo, fixed in other}
the ſonne of Agar the Ægyptian playing with Iſaac her ſonne, she ſaid to
Abraham: \V Caſt out this handmaid, and her ſonne: for the ſonne of the
handmaid shal not be heire with my ſonne Iſaac. \V Abraham tooke this
greuouſly for his ſonne. \V To whom God ſaid:
\LNote{Let it not.}{It
\MNote{Separation to be made in families, when iuſt cauſe requireth.}
ſemed ſo hard, inhumaine, and vniuſt to caſt Agar and Iſmael out of the
houſe, that Abraham would not haue done it, onles God had commanded him
in this to heare the voice of Sara. But when he knew Gods wil therin,
not reſpecting fleſh and bloud, nor diſputing further of the lawfulnes
of the fact, ſent them both away, commending them to Gods protection:
and ſo by this ſeparation the familie was made quiet.
\MNote{Myſtical ſenſe.}
It repreſented alſo a notable Myſterie of the Synagog of the Iewes &
Church of Chriſt. As
\CNote{li.~15. c.~2. ciuit.}
S.~Auguſtin doth excellently expound the ſame,
according to S.~Paul
\XRef{(Gal.~4.)}
That Iſmaels playing with Iſaac was perſecution, becauſe it tended to
peruert him, ſhewing it to be a greater iniurie, to delude and deceiue
anie, in drawing them to new and particular companies, then corporally
to perſecute them:
\CNote{Epiſt.~48. &~50.}
\MNote{The true Church doth not perſecute but iuſtly puniſheth
offenders.}
and that the Church, and Catholique Princes puniſh
heretikes for their good, to make them returne to the truth, or finally
to ceaſe from ſeducing others.
\CNote{Tract.~11. in Ioan.}
\L{Ecce libera afflixit ancillam, & non
illam vocat perſecutionem Apoſtolus, ludit ſeruus cum domino, &
perſecutionem vocat. Afflictio iſta non vocatur perſecutio, & luſio illa
vocatur perſecutio.}
\MNote{Heretikes and other infidels do perſecute when either by word, or
ſword they impugne the truth.}
Behold the freewoman afflicted the handmaide, and the Apoſtle calleth it
not perſecution, the ſeruant playeth with the maiſter, and he calleth it
perſecution. This affliction is not called perſecution, and that play is
called perſecution. Againe, he ſayeth of Sara,
\CNote{Tract.~12.}
\L{Illuſionem vidit in
illo luſu, animaduertit ſerui ſuperbiam, diſplicuit illi, eicit foras.}
She ſaw foule play in that play, ſhe noted the ſeruants pride, it
diſpleaſed her, ſhe caſt him forth of the dores.
\CNote{in 4.~Gal.}
S.~Hierom alſo teacheth
that the ſpiritual neuer perſecuteth the carnal: but ſpareth him as his
ruſtical brother, knowing that he may in time be profitable.}
Let it not ſeme greuous to thee for the boy, and for thy handmaid: al
things that Sara shal ſay to thee, heare her voice: becauſe
\CNote{Rom.~9. Heb.~11.}
\SNote{See
\XRef{chap.~17. v.~21.}}
in Iſaac shal ſeed
\Fix{de}{be}{obvious typo, fixed in other}
called to thee. \V But the ſonne alſo of the handmaid I wil make into a
great nation, becauſe he is thy ſeede. \V Abraham therfore roſe vp in
the morning, and taking bread and a bottle of water, put it vpon her
shoulder, and deliuered the boy and diſmiſt her. Who went away, and
wandred in the wilderneſſe of Berſabee. \V And when the water in the
bottle was ſpent, ſhe caſt the boy vnder one of the trees, that were
there. \V And ſhe went her way, and ſate ouer againſt a great way of as
farre as a bowe can caſt. For she ſaid: I wil no ſee the child dying:
and ſitting ouer againſt, she lifted vp her voice and wept. \V And God
heard the voice of the boy: and an Angel of God called Agar from heauen,
ſaying: What doeſt thou Agar? feare not: for God hath heard the voice of
the boy, from the place wherin he is. \V Ariſe, take vp the boy, and
hold his hand: for into
%%% o-0088
a great nation wil I make him. \V And God opened her eyes: who ſeing a
wel of water, went, and filled the bottle, and gaue the boy to
drinke. \V And God was with him: who grew, and abode in the wildernes,
and became a young man archer. \V And he dwelt in the wildernes of
Pharan, and his mother tooke a wife for him out of the land of Ægypt. \V
The ſame time ſaid Abimelech, and Phicol the general of his armie to
Abraham: God is with thee in al things which thou doeſt. \V Sweare
therfore by God, not to hurt me, and my poſteritie, and my ſtocke: but
according to the mercie, that I haue done thee, thou ſhalt doe to me,
and to the land wherin thou haſt liued a ſtranger. \V And Abraham ſaid:
I wil ſweare. \V And he rebuked Abimelech for the
%%% 0094
wel of water, which his ſeruants had taken away by force. \V And
Abimelech anſwered: I knew not who did this thing: yea and thy ſelfe
dideſt not tel me, and I heard not of it but to day. \V Abraham therfore
tooke ſheepe and oxen, and gaue to Abimelech: and both of them made a
league. \V And Abraham ſet ſeuen ewe lambes of the flocke apart. \V To
whom Abimelech ſaid: What meane theſe ſeuen ewe lambes, which thou haſt
made to ſtand apart? \V But he ſaid: Seuen ewe lambes ſhalt thou take at
my hand: that they may be a teſtimonie for me, that I digged this
wel. \V Therfor was that place called
\TNote{vvel of oath}
Berſabee: becauſe there both did ſweare. \V And they made a league for
the wel of oath. \V And Abimelech aroſe, and Phicol the general of his
armie, and they returned to the land of the Paleſtines. But Abraham
planted a groue in Berſabee, and called therupon the name of our Lord
God eternal. \V And he was a ſeiorner in the land of the Paleſtines
manie dayes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
