%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 29
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-29


\project douay-rheims


%%% 0115
%%% o-0108
\startChapter[
  title={Chapter 29}
  ]

\Summary{Iacob intertained by Laban, 15.~ſerueth him ſeuen yeares for
  Rachel, 23.~but firſt receiuing Lia, 27.~ſeuen dayes after receiueth
  alſo Rachel, and ſerueth for her ſeuen yeares more. 31.~She remaining
  barren, Lia beareth foure ſonnes.}

Iacob therfore going on his iourney, came into the Eaſt countrie. \V And
he ſawe a wel in the field, and three flockes of sheepe lying beſide it:
for of it the beaſts were watered, and the mouth therof was cloſed with
a great ſtone. \V And the maner was when al the sheepe were geathered
togeather they did rowle of the ſtone, and after the sheepe were
refreshed they put it on the mouth of the wel againe. \V And he ſaid to
the sheperds: Brethren, Whence are you? Who anſwered: Of Haran. \V And
he asked them, ſaying: Know you Laban the ſonne of Nachor? They ſaid: We
do knowe him. \V Is he in health? quoth he: He is in health, ſay they:
And behold Rachel his daughter cometh
%%% 0116
with his flocke. \V And Iacob ſaid: There is yet much day remaining,
neither is it time to bring the flockes into the foulds againe: firſt
geue the sheep drinke, and ſo bring them backe to feede. \V Who
anſwered: We can not, til al the cattel be gethered together: and we
remoue the ſtone from the wels mouth, that we may water the flocks. \V
They were yet ſpeaking, and behold Rachel came with her fathers ſheepe:
for ſhe fed the flocke. \V Whom when Iacob had ſeene, and knew her to be
his coſen germaine, and that they were the ſheepe of Laban his vncle: he
remoued the ſtone, wherwith the wel was cloſed. \V And hauing watered
the flocke, he
\SNote{S.~Auguſtin
\Cite{(q.~87. in Gen.)}
cõmendeth familiar kiſſing of kinsfolke and frendes as a laudable
cuſtome in ſome countries. It is no where more ciuil & modeſt then in
England.}
kiſſed her: and lifting vp his voice wept, \V and he told her that he
was her fathers brother, and the ſonne of Rebecca: but ſhe in haſt went
and told her father. \V Who when he heard that Iacob his ſiſters
ſonne was come, he ranne forth to mete him: and embracing him, and
hartely kiſſing him, brought him into his houſe. And when he had heard
the cauſes of his iourney, \V he anſwered: Thou art my bone and
my fleſh. And after the dayes of one moneth were expired, \V he ſaid to
him: becauſe thou art my brother, ſhalt thou ſerue me
\TNote{VVithout vvages?}
gratis? Tel me what wages wilt thou take. \V He had in dede two
daughters, the name of the elder was Lia: and the younger was called
Rachel. \V But Lia was bleare eyed: Rachel wel fauored, and of a
bewtiful countenance. \V Whom Iacob louing, ſaid: I wil ſerue thee for
Rachel thy younger daughter, ſeauen yeares. \V Laban anſwered: It is
better that I geue her to thee then to an other
%%% o-0109
man, tary with me. \V Iacob therfore ſerued for Rachel ſeuen yeares: and
they ſemed a few dayes becauſe of the greatnes of his loue. \V And he
ſaid to Laban: geue me my wife: becauſe now the time is complete, that I
may companie with her. \V Who hauing bid a great number of his freinds
to the feaſt, made the mariage. \V And at night
\SNote{Laban greuouſly offended, neither could Lia be excuſed, but Iacob
was innocent, in this fact.}
he brought in Lia his daughter to him, \V geuing his daughter a
handmaid, named Zelpha. With whom when Iacob had
\Fix{compained}{companied}{obvious typo, fixed in other}
after the maner, when morning was come he ſaw Lia: \V and he ſaid to his
father in lawe: What is it that thou dideſt meane to doe? did not I
ſerue thee for Rachel? why haſt thou deceiued me? \V Laban anſwered: It
is not the cuſtome in this place, that we beſtowe the younger in mariage
firſt. \V Make vp the
\SNote{After ſeuen daies he had Rachel who was his firſt ſpouſe.
\Cite{S.~Hierom Tradit. Heb.}
\Cite{S.~Aug. q.~89. in Gen.}}
weeke of dayes
%%% 0117
of this match: and I wil geue thee this ſame alſo, for the worke that
thou shalt ſerue me other ſeauen yeares. \V He yelded to his pleaſure:
and after the weeke was paſt, he maried Rachel to wife: \V to whom her
father had deliuered Bala for to be her ſeruant. \V And hauing at length
obteined the mariage that he wished, he preferred the loue of the later
before the former, ſeruing with him other ſeauen yeares. \V And our Lord
ſeing that he deſpiſed Lia, opned her wombe, her ſiſter remaining
barren. \V Who conceaued and bare a ſonne, and called his name Ruben,
ſaying: Our Lord ſaw mine affliction: now my huſband wil loue me. \V And
againe she conceaued and bare a ſonne, and ſaid: For becauſe our Lord
heard that I was contemned, he hath geuen this alſo to me: and she
called his name Simeon. \V And she conceaued the third time, and bare an
other ſonne: and ſaid: Now alſo my huſband wil be ioyned to me, for
becauſe I haue borne him three ſonnes: and therfore she called his name,
Leui. \V The fourth time she conceaued and bare a ſonne, and ſaid: Now
wil I confeſſe to our Lord. And for this she called him Iuda: and she
left bearing.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
