%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 0074
%%% o-0070
\startChapter[
  title={Chapter 14}
  ]

\Summary{The king of Sodom with other foure kings are ouercome in
  battle, by foure others: 12.~where Lot is taken 14.~but Abram with
  318.~perſons proſecuting and ouercoming the victorers, 16.~reſcued Lot,
  with al the captiues and pray. 18.~Melchiſedech King and Prieſt
  bleſsed Abram, 20.~Abram payed tithes to him, 21.~and rendered the
  ſpoile to the king of Sodome.}

And it came to paſſe in that time, that Amraphel the king of Sennaar,
and Arioch the king of Pontus, and Chodorlahomor king of the Elamyts,
and Thadal the king of nacions \V made warre againſt Barra the king of
Sodome, and againſt Berſa the king of Gomorra, and againſt Sennaab the
king of Adama, and againſt Semebar the king of Seboim, and againſt the
king of Bala, the ſame is Segor. \V Al theſe came together into the
Woodland vale, which now is the ſalt ſea. \V For they had ſerued
\Fix{Chodorlamor}{Chodorlahomor}{obvious typo, fixed in other}
twelue yeares, and the thirtenth yeare they reuolted from him. \V
Therfore in the fourtenth yeare came Chodorlahomor, and the king that
were with him: and they ſtroke Raphaim in Aſtarothcarnaim, and Sufim with
them, and Emim in Sauee of Cariathaim, \V and the Corrheans in the
mountains of Seir, euen to the Champion countrie of Pharan, which is in
the wildernes. \V And they returned, and came as farre as the fountaine
of Miſphat, the ſame is Cades: and they ſtroke al the countrie of the
Ameleichites, and of the Amorheans, that dwel in Aſſaſonthamar. \V And
they went forth the king of Sodome, and the king of Gomorra, and the
king of Adama, and the king of Seboim, moreouer alſo the king of Bala,
which is Segor: and they ſet themſelues againſt them in battaile aray in
the Woodland vale: \V to wit againſt Chodorlahomor king of the Elamites,
and Chadal king of nacions, and Amraphel king of Sennaar, and Arioch king
of Pontus: four kings againſt fiue. \V But the Woodland vale had many
pitts of bitume. Therfore the king of Sodome, and of Gomorra turned
their backes, and were ouerthrowne there: and they that remained fled to
the mountaine. \V And they tooke al the ſubſtance of the Sodomites, and
Gomorrheans, and tooke al kind of victuales, and went their way: \V and
Lot alſo and his ſubſtance, the ſonne of Abrams brother, who
%%% 0075
dwelled in Sodom. \V And behold one, that had eſcaped, told Abram the
Hebrew, that dwelt in the vale of Mambre, the Amorrean brother of
Eſchol, and the brother of Aner: for theſe had made a league with
Abram. \V Which when
%%% o-0071
Abram had heard, to witt, that his brother Lot was taken, he numbred of
the ſeruantes borne in his houſe, wel appointed three hundred and
eightene: and purſued them vnto Dan. \V And diuiding his companie, he
ranne vpon them in the night: and ſtroke them, and purſued them vnto
Hoba, which is on the left hand of Damaſcus. \V And he brought backe al
the ſubſtance, and Lot his brother with his ſubſtance, the wemen alſo
and the people. \V And the king of Sodom went forth to meete them, after
he returned from the ſlaughter of Chodorlahomor, and of the kinges that
were with him in the vale Sauee, which is the kings vale. \V But
\LNote{Melchiſedech.}{S.~Hierom
\CNote{Tomo~3. prope initium.}
\MNote{Old hereſies concerning Melchiſedech.}
being earneſtly requeſted by Euagrius, to geue his iudgement touching
Melchiſedech, whom a nameles author had endeuored to proue to be the
Holie Ghoſt, plainly confuteth that error: as alſo an other error of
Origen and Didymus ſaying, he was an Angel. Likewiſe S.~Epiphanius
\Cite{(hereſi.~55. &~76.)}
relateth and condemneth a third error, of ſome that thought him to be
the Sonne of God. Theſe two Fathers, and S.~Auguſtin
\Cite{(li. de hereſibus her.~34.)}
and diuers others, whom S.~Hierom alleageth proue euidently, that he was
a very man, a Prieſt and a king, yea the hiegh Prieſt (at leaſt of that
countrie) Superior to Abraham, and a figure of Chriſt. Beſides theſe
hereſies the ſame S.~Hierom relateth two probable opinions.
\MNote{A probable opinion that Sem and Melchiſedech was al one.}
The Iewes Rabins generally hold that Melchiſedech was Sem the ſonne of
Noe from whom Abraham and al the Hebrewes deſcended. VVhich they ſeme
rather to affirme, as loath to confeſſe that anie man, of other nation
then their owne, ſhould haue bene greater and more excellent then
Abraham in ſpiritual cauſes, then for anie reaſon they do, or can
alleage. Yet manie eſpecially latter writers as Liranus, Toſtatus,
Genebtardus and others do embrace this opinion as moſt probable. Though
\CNote{Ep. ad Euang.}
S.~Hierom ſemeth only to haue added the Hebrewes opinion (as he ſaith)
becauſe he would intimate al to his freind, when he had firſt cited
grauer authores, S.~S. Ireneus, Hypolitus, Euſebius, Ceſarienſis,
Euſebius Emiſſenus, Apolinarius, and Euſtathius,
\MNote{More probable that Melchiſedech was a Chananite.}
al agreably affirming that Melchiſedech was a Chananite, king of Salem,
which was afterwards called Ieruſalem. To this opinion agreeth Philo
Iudæus continually ſpeaking of him, as of a ſtranger to the Iewes
nation. Ioſephus alſo a Iew writeth plainly
\Cite{(li.~7. de bello Iudaico. c.~18.)}
that he was of Chanaan, and Prince of the Chananites. Alſo S.~Dionyſius
Ariopagita,
\Cite{(Cæleſt. Hier. c.~9.)}
S.~Epiphanius
\Cite{(her.~55. &~67.)}
Theoderetus
\Cite{(q.~63. in Gen.)}
and Suidas, are of the ſame mind, and manie other chriſtian
Doctors. 
\MNote{Proued by S.~Pauls wordes.}
VVho confirme their aſſertion by that
\CNote{Heb.~7. v.~6.}
S.~Paul ſaith to the
Hebrewes: \Emph{He vvhoſe generation is not numbred among them, tooke
tithes of Abraham.} For what els can S.~Paul meane, but that
Melchiſedechs kinred and people, was diuers from the kinred, and people
of the Iewes? which he could not ſay of Sem, from whom Abraham & al
Iewes deſcended: as it can not be ſaid, that Adam and Noe are of diuers
generation from anie people that now liueth, becauſe we al come of
them. Of this difficultie (not pertaining to anie controuerſie of our
time) the ſtudious may ſee more in
\Cite{F.~Pererius his commentaries vpon this 14.~chap of Geneſis, diſp.~3.}}
\Fix{Melcheſidech}{Melchiſedech}{obvious typo, fixed in other}
the king of Salem,
\LNote{Bringing forth.}{Seing
\MNote{Chriſt is a Prieſt according to the order of Melchiſedech.}
the
\CNote{Pſal.~109.}
Royal Prophet Dauid, and
\CNote{Heb.~7.}
S.~Paul ſay Chriſt is a Prieſt for euer
according to the order of Melchiſedech, we demand of Proteſtants, if
Chriſt fulfilled not Melchiſedechs figuratiue Sacrifice offered in bread
and wine, by offering his owne bodie and bloud at his laſt ſupper in
formes of bread & wine, and by inſtituting the ſame to be offered by his
Prieſts til the end of the world, what other figuratiue ſacrifice of
Melchiſedech they can find performed by Chriſt, by which it may appeare
that he is a Prieſt for euer according to that order? Caluin
\Cite{(li.~4. Inſtit. c.~18. para.~2.)}
Kemniſius
\Cite{(par.~2. exam. pag.~740. &~747.)}
Peter martyr
\XRef{(in 1.~Cor.~5.)}
and moſt Engliſh Proteſtants grant that Melchiſedech was a Prieſt,
\MNote{VVhy the Proteſtants haue miniſters and no Prieſts.}
and
that the peculiar function of a Prieſt is to offer Sacrifice, wherfore
they hauing no ſacrifice wil haue only miniſters and no Prieſts, but
they denie that Melchiſedech offered Sacrifice in bread and
wine. VVherupon we ioyne iſſue with them to proue that he did, and that
by this place amongſt others of holie Scripture.

Kemniſius complayneth that the Latin text hath \L{Obtulit} for \L{Protulit},
\Emph{Offered}, for \Emph{Brought forth}. And to diſproue the ſame, he
alleageth the Hebrew, Chaldee, Greke, and S.~Cyprian. But Catholiques
more iuſtly complaine of him, for lying.
\MNote{The ſtate of the controuerſie.}
For al Latin Editions haue \L{Proferens}, \Emph{bringing forth}. The queſtion
therfore in controuerſie is, to what end and vſe Melchiſedech brought
forth bread and wine? Caluin and Kemniſius ſay it was only to refreſh or
feede Abraham and his men, and not for ſacrifice. Which their bare
ſaying is without reaſon, for that there was ſtore of victuals in the
pray
\XRef{(v.~11.)}
and
\CNote{v.~24.}
they had eaten therof.
\CNote{S.~Chriſ. ho.~36. in Gen.}
Moreouer the Hebrew word \HH{Hotſi}, \Emph{brought forth}, is a word
pertayning to Sacrifice, as in the
\XRef{6.~chap. of Iudges. (v.~18. and~19.)}
and importeth that the bread and wine were firſt offered in Sacrifice,
and then doubtles they did participat: though they wanted not other
ſufficient corporal foode. Againe the wordes folowing \Emph{For he vvas
the Prieſt of God moſt hiegh}, can haue no other ſenſe, but that
\MNote{Melchiſedech offered ſacrifice in bread and wine.}
he did
the function of a Prieſt in the bread & wine which he brought, otherwiſe
if the only cauſe of bringing that prouiſion had benne to releue the
campe with victuals, the reaſon would rather haue benne yelded, becauſe
he was a bountiful King, a liberal Prince, a ſpecial freind to Abraham,
as in deede he was, but none of theſe reaſons, or the like fitted this
purpoſe ſo wel, nor touched the cauſe of bringing forth bread and wine,
as to ſignifie that he was a Prieſt, whoſe office is to offer
Sacrifice.

Here
\MNote{The latin text iuſtified by other places, yea by Proteſtants
tranſlations.}
againe ſome Proteſtants take exception againſt the Latin text, that the
cauſual coniunction \L{Enim}, \Emph{for}, is not agreable to the Hebrew,
but ſhould be the copulatiue \L{Et}, \Emph{and}, which is a meere
wrangling. For the lerned know wel ynough, that the Hebrew particle is
better expreſſed in ſuch places, by \L{Enim} or, \L{quia}, \Emph{for} or
\Emph{becauſe}, then by \L{&}. And ſo the Engliſh Bible printed in the
yeare of our Lord, 1552. readeth: \Emph{Melchiſedech king of Salem
brought forth bread and vvine: for he vvas the Prieſt of the moſt
hiegheſt God.} The latter Editions alſo in like places haue not the
copulatiue \Emph{And}, but ſome other word as the ſenſe requireth.
\XRef{Gen.~20. v.~3.}
\Emph{Thou art but a dead man, for the vvomans ſake vvhich thou haſt
taken: for she is a mans vvife.} where the Hebrew phraſe is, \Emph{And
she is maried to a husband.}
\XRef{Gen.~30. v.~27.}
they read thus: \Emph{For I haue proued that the Lord hath bleſſed me
for thy ſake}, where preciſly conſtruing the Hebrew they ſhould
ſay, \Emph{I diuined} (or coniectured) \Emph{and the Lord bleſſed me for
thy ſake.} Likewiſe
\XRef{Eſaie,~64. v.~5.}
they read: \Emph{But Loe, thou haſt bene angrie, for vve offended}, the
Hebrew is thus: 
\Emph{Loe thou art angrie, and vve haue ſinned.} In the ſame place, they
tranſlate, \Emph{Yet shal vve be ſaued}, vvhich the Hebrew expreſſeth by
the copulatiue, \Emph{And} we ſhal be ſaued. So when they thinke it
conuenient, they tranſlate the Hebrew particle, \Emph{For, that, yet},
which ſtrictly ſignifieth \Emph{And}.

Now let vs alſo ſee the original in this place.
\MNote{The Hebrew text ſheweth that Melchiſedech did the office of a
Prieſt in bread and wine.}
In the Hebrew it is thus: \HH{\Sc{Vmalchi tsedec melec chalem hotsi
lechem vaiaiin. Vehv cohen leel elion. Vaie varechehv}}, &c. In Engliſh
word for word thus: \Emph{And Melchiſedech king of Salem brought forth
bread and vvine. And he the prieſt to God moſt hiegh. And he bleſsed
him}, &c. where albeit the cauſual word \Emph{For}, is not expreſſed,
yet theſe wordes, \Emph{And he the prieſt}, further declaring that
beſides the office and dignitie of a King (which was ſaid before)
Melchiſedech was alſo a Prieſt, muſt nedes ſignifie that he did
ſomething about the bread and wine belonging to a Prieſts office.
\MNote{The proper office of a Prieſt is to offer Sacrifice.}
And what that ſomething was, perhaps the Vniuerſitie of Cambridge wil
teſtifie, whoſe late
\CNote{Guiliel vvhitak contra Gregor. Martin.}
profeſſor of Diuinitie teacheth plainly, that
Melchiſedech offered Sacrifice, and was therin a figure of Chriſt
\Cite{(Pag.~6. Reprehen)}
\L{Sacerdotes} (ſaith he) \L{ij vere & proprie ſunt, qui ſacrifitia
faciunt; qualis fuit Aaron, & Aaronis filij, & Melchiſedechus, & quem
illi adumbrabant, Chriſtus.} \Emph{Prieſtes truly and properly are they,
that offer ſacrifices, ſuch as vvas Aaron, and the ſonnes of Aaron, and
Melchiſedech, and Chriſt, vvhom they prefigured.} If then both Aaron &
Melchiſedech were truly and properly Prieſts, becauſe they offered
ſacrifices (according to this Profeſſors definition) and both were
figures of Chriſt, it muſt needes be granted that as Chriſt fulfilled
the figure of Aarons bloudie ſacrifices, in offering him ſelfe vpon the
Croſſe: ſo he alſo fulfilled the figure as wel of vnbloudie ſacrifices
of Aaron, as expecially of Melchiſedechs Sacrifice in ſome other beſides
that on the Croſſe, ſeeing the prophet Dauid and S.~Paul ſay, Chriſt is
a Prieſt (not according to Aarons order, for that was to haue an end,
but)
\MNote{Chriſt ſtil exerciceth the office of Prieſthood according to the
order of Melchiſedech by the miniſterie of Prieſts.}
\Emph{For euer} according to the order of Melchiſedech. And what
other Sacrifice did our Sauiour offer to remaine perpetual, but of his
owne bodie & bloud in vnbloudie maner, vnder the formes of bread and
wine, with commandment to his Apoſtles and Prieſts to do the ſame til
the end of the world? Let the indifferent reader weigh it wel. And
whoſoeuer is not very proud wil for his better inſtruction, or
confirmation, eſteme the vniforme iudgement of manie, ancient, godlie,
and lerned Fathers writing vpon this place. VVe wil only recite their
wordes, without other deduction for breuitie ſake.

S.~Clemens
\MNote{The ancient fathers proue that Melchiſedech offered Sacrifice in
bread & wine in figure of Chriſt, and of other Prieſts of the new
Teſtament.}
Alexandrinus
\Cite{(li.~4. Strom. verſus finem.)}
writeth thus: Melchiſedech king of Salem, Prieſt of God moſt hiegh, gaue
wine & bread ſantified nutriment in type of the Euchariſt.

S.~Cyprian
\Cite{(Epiſt.~63. ad Cæcilium)}
Chriſt is Prieſt for euer according to the order of Melchiſedech, which
order is this coming from that Sacrifice, and thence deſcending, that
Melchiſedech was Prieſt of God moſt hiegh, that he offered bread & wine,
that he bleſſed Abraham. For who is more a Prieſt of God moſt hiegh,
then our Lord Ieſus Chriſt, who offered Sacrifice to God the Father, and
offered the ſame, which Melchiſedech had offered, bread and wine, to
wit, his owne bodie and bloud. And a litle after: That therfore in
Geneſis, the bleſſing might be rightly celebrated, about Abraham, by
Melchiſedech the Prieſt, the image of Chriſts Sacrifice conſiſting in
bread and wine went before, which thing our Lord perfecting and
performing, offered bread and chalice mixt with wine, and he that is the
plenitude, fulfilled the veritie of the prefigured image.

Euſebius Ceſarienſis
\Cite{(li.~5. Demonſt. Euang. c.~3.)}
Euen as he who was Prieſt of Nations was neuer ſene to offer corporal
ſacrifices, but only bread & wine, when he bleſſed Abraham: ſo firſt our
Lord & Sauiour him ſelfe, then prieſts that come from him, exerciſing
the ſpiritual office of Prieſthood in al nations after the
Eccleſiaſtical ordinances, do repreſent the myſteries of his bodie, and
healthful bloud in bread & wine, which myſteries Melchiſedech knew ſo
long before by diuine ſpirite, and vſed as repreſentations of things to
come.

S.~Ambroſe
\Cite{(li.~5. de Sacramen. c.~1.)}
VVe know the figure of the Sacraments went before, in Abrahams times,
when Melchiſedech offered Sacrifice.

Idem. in
\XRef{cap.~5. Heb.}
It is clere that oblations of cattle are vaniſhed, which were in Aarons
order, but Melchiſedechs inſtitution remayneth, which is celebrated al
the world ouer in adminiſtration of the Sacraments.

S.~Hierom.
\Cite{(Epiſt. ad Marcellamo vt migret Bethleem.)}
Haue recourſe to Geneſis, and you ſhal find Melchiſedech king of Salem,
prince of this citie, who euen then in figure of Chriſt offered bread
and wine, and dedicated the Chriſtian myſterie in our Sauiours bodie and
bloud.
\Cite{Idem Epiſt. ad Euagrium}:
Melchiſedech offered not bloudie victims, but dedicated the Sacrament of
Chriſt in bread and wine, ſimple and pure ſacrifice.
\Cite{Idem. Queſt. in Gen. c.~14.}:
Our myſterie is ſignified in the word of order, not by Aaron in
imolating brute victims, but in offering bread and wine, that is the
bodie and bloud of our Lord Ieſus.
\Cite{Idem in c.~26. Matthei}: Melchiſedech the Prieſt of God moſt
hiegh, by offering bread and wine, prefigured the Myſterie of the
Euchariſt.

S.~Chriſoſtom
\Cite{(ho.~35. in Gen.)}
Seeing the figure, thinck alſo I pray thee, of the veritie.
\Cite{ho.~36.}
After that Melchiſedech king of Salem brought bread and wine (for he was
Prieſt of God moſt hiegh) Abraham receiued his oblations.

S.~Auguſtin
\Cite{(Epiſt.~95.)}
Melchiſedech bringing forth the ſacrament (or myſterie) of our Lords
table, knew how to figurate his eternal prieſthood.
\Cite{Idem li.~16. c.~22. ciuit.}
There firſt appeared that ſacrifice, which is now offered to God by
Chriſtians in the whole world.
\Cite{Idem li.~17. c.~17.}
\Cite{li.~18. c.~35.}
and
\Cite{vpon the Pſalme.~109.}
\Cite{li.~1. contra aduers. Leg. & prophet. c.~20.}
\Cite{Ser.~4. de Sanctis Innocentibus.}
But it is bootles or nedles to cite more places, or more authors.
\MNote{Caluin cõtemneth al the ancient fathers. in
\XRef{ca.~9. Heb.}}
For
whoſoeuer wil not ſubmit their iudgements to theſe, would not, it is
like, beleue, if their owne maiſters, ſhould riſe agane and warne them,
leſt they be damned for their incredulitie.}
bringing forth bread and wine, for he was the Prieſt of God moſt
highe, \V
\LNote{Bleſſed him.}{Caluin
\XRef{(in cap.~7. v.~9. Heb.)}
Muſculus
\Cite{(locis com. c. de Miſſa Papiſt)}
and ſome other Proteſtants to auoid the connexion of
Melchiſedechs \Emph{Prieſthood} and bringing forth of bread and wine,
wil nedes haue theſe wordes, \Emph{He vvas a Prieſt}, referred only to
that which foloweth, \Emph{he bleſſed Abraham.}
\MNote{Heretical tranſlations.}
And
\CNote{Bible 1579.}
ſome Engliſh tranſlaters for this purpoſe haue corrupted the text,
by changing, \Emph{And} into \Emph{Therfore}, ſaying thus: \Emph{And he
vvas a Prieſt of the moſt hiegh God, therfore he bleſſed him.} which is
alſo a falſe gloſſe. For Melchiſedech did not bleſſe Abram becauſe he
was a Prieſt, for Abram was alſo a Prieſt, but becauſe he was a greater
Prieſt then Abram. Which
\CNote{Heb.~7.}
S.~Paul vrgeth ſaying:
\MNote{The greater bleſſeth the leſſe.}
\Emph{VVithout al contradiction that vvhich is leſſe, is bleſſed of the
better}, concluding therupon that Melchiſedech was greater then
Abraham. Agane other Superiors that are not Prieſts may bleſſe their
inferiors. As
\CNote{Ioſue~8. &~22.}
Ioſue and
\CNote{3.~Reg.~8.}
Salomon bleſſed the people, and parents bleſſe their children.}
bleſſed him, and ſaid: Bleſſed be Abram to God the higheſt, which
created heauen and earth: \V and bleſſed be God the higheſt, by whoſe
protection, the enemyes are in thy hands. And
\LNote{Gaue tithes.}{This
\MNote{Paying of tithes in the law of nature.}
is an other prerogatiue of Melchiſedech, that Abram payed tithes to him,
which S.~Paul likwiſe explicateth
\XRef{(Heb.~7.)}
and proueth therby that Chriſts Prieſthood is greater then the
Leuitical. Moreouer this paying of tithes by Abraham ſheweth the
antiquitie of this tradition, being practiſed in Abrahams time, that the
ſpiritual Superiors receiued tithes of their inferiors.}
he gaue him the tythes of al. \V And the king of Sodom ſaid to Abram:
Geue me the ſoules, and the reſt take to thee. \V Who anſwered him: I
lift vp my hand to my Lord God moſt hiegh poſſeſſor of heauen and
earth, \V that from the very woofe-thread vnto the ſhoe latchet, I wil
not take of al that are thine:
\SNote{Abraham enriched by God, would take no more of man, but his
ſoldiars ſuſtenance. The proper hire of ſpiritual work men,
\Cite{S.~Chriſoſt. in Gen.}}
leſt thou ſay: I haue enriched Abram: \V except ſuch thinges, as the
young men haue eaten, and the ſhares of the men, that came with me,
Aner, Eſchol, and Mambre: theſe ſhal take their ſhares.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
