%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 17
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-17


\project douay-rheims


%%% 0083
%%% o-0079
\startChapter[
  title={Chapter 17}
  ]

\Summary{God renewing his promiſes to Abram, 5.~changeth his name,
  10.~and commandeth Circumciſion, 15.~changeth alſo his wiues name,
  promiſeth a ſonne of her. 20.~Likewiſe that Iſmael shal proſper,
  23.~and the ſame day Abraham circumciſed him ſelfe, and Iſmael, and al
  the men of his houſe.}

And after that he beganne to be nyntie and nyne yeares old, our Lord
appeared vnto him: and ſaid vnto him: I am the God almightie: walke
before me, and be
\SNote{He is perfect in this life, that ſincerly & diligently tendeth
towards perfection of the next life. And this God here commanded to
Abraham, & Chriſt to al Chriſtians.
\XRef{Math.~5.}
\Cite{S.~Aug. li. de perfect. cont. Cæleſt.}
See
\XRef{Gen.~6. v.~9.}}
perfect. \V And I wil make my couenant betwen me and thee: and I wil
multiplie thee exceadingly. \V Abram fel flat on his face. \V And God
ſaid to him: I am, and my couenant is with thee, & thou shalt be a
father of manie nations. \V Neyther ſhal thy name be called any more
Abram: but thou shalt be called Abraham: becauſe a father of
\LNote{Manie nations.}{Abram
\MNote{Changing of names is myſtical.
\Cite{S.~Aug. li.~16. c.~28. &~39. ciuit.}
\Cite{S.~Hierom. Tradit. Hebraicis.}}
ſignifying \Emph{Hiegh} or \Emph{Noble father}, changed into Abraham,
which ſignifieth \Emph{Father of manie} (nations) and Sarai
ſignifying \Emph{My Princes}, changed into Sara, which abſolutly
ſignifieth \Emph{Princeſſe}, import great Myſteries intended by God. For
by this changing of names was declared and confirmed, that Abraham by
the iſſue that he ſhould haue of Sara, ſhould be the father of manie
nations, which S.~Paul expreſly applieth to his ſpiritual children,
thoſe eſpecially that ſhould beleue in Chriſt, of the Gentiles
prefigured by
\CNote{Gal.~4.}
\Emph{Iſaac borne of the freevvoman, by the promiſe}, as
the Iewes were prefigured by \Emph{Iſmael borne of the bond vvoman
according to the flesh}, concluding with the prophet Eſai,
that
\CNote{Iſa.~54.}
\Emph{Manie are the children of the deſolate, that before vvas barren.}
\MNote{Abraham natural father of foure nations: ſpiritual father of al
that do beleue in Chriſt.
\XRef{Rom.~4,~11.}}
For albeit Abraham was natural father of foure nations, The Iſmaelites,
Madianites, Idumeans, and the Iſraelites, yet he was ſpiritual father of
manie more, to wit, of al nations that beleue in Chriſt from his owne
time to the end of the world, of which ſome are Iewes, but the greateſt
part Gentiles, as the ſame Apoſtle ſheweth.
\XRef{Rom.~4.}
\XRef{Gal.~4.}
\XRef{Epheſ.~3.}
\XRef{Colloſ.~1.}
and in other places.
\MNote{The Church of Chriſt euer conſiſteth of many nations.}
Wherby is clere, that the Church of Chriſt doth alwayes conſiſt of manie
nations, not as Donatiſtes and Proteſtants abſurdly ſay, ſometimes of few
inuiſible or vnknowen perſons, for ſo Abraham ſhould ſometimes be father
of few or 
no nations, which derogateth either from Gods omnipotencie if he could
not, or from his fidelitie if he ſhould not maintaine his promiſe, made
to Abraham for euer.}
many nations I haue made thee. \V And I wil make thee encreaſe
excedingly, and I wil make thee into nations, and kinges shal come forth
of thee. \V And I wil eſtablish my couenant betwen me and thee, and
betwen thy ſeede after thee in their generations by a perpetual
couenant: to be thy God, and thy ſedes after thee. \V And I wil geue to
thee, and to thy ſeed the land of thy peregrination, al the land of
Chanaan for a perpetual poſſeſſion, and I wil be their God. \V Againe
God ſaid to Abraham: And thou therfore shalt keepe my couenant, and thy
ſeed after thee in their generations. \V This is my couenant which you
shal obſerue betwen me and you,
%%% 0084
and thy ſeede after thee: Al the malekind of you
\LNote{Shal be circumciſed.}{The
\MNote{Circumciſion a figure of Baptiſme.}
Sacrament of Circumciſion here firſt inſtituted, about 400.~yeares
before the Law of Moyſes, is a perfect figure of Baptiſme, reſembling it
in foure things. Firſt by both theſe Sacraments the faithful are
diſtinguiſhed from infidels. Secondly profeſſion of faith is made in them
both, either by thoſe that receiue the ſame, if they be of diſcretion,
or by others for them, if they be infants. Thirdly by both theſe
Sacraments entrance is made into the Church, and to the participation of
other Sacraments and ſpiritual rites. Fourthly both theſe Sacraments
induce ſubiection to the iuriſdiction, and lawes of the Church.
\MNote{Baptiſme excelleth Circumciſion.}
But Baptiſme doth farre excel Circumciſion in that it is more eaſie, or
leſſe painful, more vniuerſal, for it pertaineth to al nations, and both
ſexes; and eſpecially in vertue and efficacie, for Baptiſme as an
inſtrumental cauſe remitteth ſinne and iuſtifieth, Circumciſion was only
a ſigne that grace was geuen, & ſinne remitted. Againe Baptiſme
imprinteth a character in the ſoule, the other leaueth a marck only in
the fleſh. Finally Baptiſme openeth the gate of heauen, in vertue of
Chriſts paſſion now paſt, which circumciſion could not before Chriſt
ſuffered death. Of which both reſemblance and difference S.~Auguſtin
treateth in manie places, eſpecially
\Cite{li.~3. de doct. chriſt. c.~9.}
\Cite{Epiſt.~118.}
and
\Cite{Ianuarium. li.~19. c.~13. cont. Fauſt.}
& in
\XRef{Pſal.~73.}
where he alſo maketh like compariſon betwen other Sacraments of the old
and new Teſtament.}
shal be circumciſed: \V and you shal circumciſe the flesh of your
prepuce, that it may be for a ſigne of the couenant betwen me and
you. \V And infant of
\SNote{Circumciſion and name receiued the eight day ſignified the
aſſociation of Saints in heauen after the ſeuen dayes trauel of this
world.
\Cite{Ser. de circum. apud S.~Cyprian.}}
eight daies shal be circumciſed among you, al malekind in your
generations: aſwel the homebred shal be circumciſed, as the bought
ſeruant of whoſoeuer he is, not of your ſtocke: \V and my couenant shal
be in your flesh for a perpetual couenant. \V The male, whoſe flesh of
his prepuce shal not be circumciſed, that ſoule
\LNote{Shal be deſtroyed.}{Here
\MNote{Two difficulties.}
occurre two difficulties about the true ſenſe of this hard place. Firſt
whether this puniſhment belonged to them only, by whoſe fault
circumciſion was omitted: or to infants alſo that ſhould be circumciſed
and were not. Secondly whether temporal puniſhment, or eternal was here
threatned for tranſgreſſion of this precept.
\MNote{To whom the puniſhment pertained whẽ circumciſion was omitted.}
Touching the firſt doubt, it ſemeth probable, that this puniſhment
pertained only to thoſe, by whoſe negligence themſelues, or others of
their charge, were not circumciſed: for that the reaſon why ſuch a one
is puniſhed is aleaged, \Emph{Becauſe he hath broken Gods couenant.}
VVhich is only in their powre to kepe or breake, that are of diſcretion,
and not in the powre of infants. The indifferencie alſo of the Hebrew and
Greeke text fauoreth this ſenſe, for where according to the Latin we
read, \Emph{The male vvhoſe flesh of prepuce is not circumciſed}, the
Hebrew and Greke may be tranſlated, \Emph{The male that doth not
circumciſe the flesh of his prepuce}, which can not be ment of an infant
of eight dayes old. This expoſition is likewiſe confirmed by example,
\XRef{Exodi.~4.}
where Moyſes was in danger to be ſlaine by an Angel, becauſe his ſonne
was not circumciſed. Neuertheles S.~Auguſtin
\Cite{(li.~3. c.~18. cont. Iulian.}
\Cite{li.~16. c.~27. ciuit.}
and in other places) proueth that this commination pertained alſo to
infants. Whoſe iudgement is confirmed by the 70.~Interpreters adding for
explication theſe wordes, \Emph{the eight day}: which neceſſarily
include infants, as ſubiect to this puniſhment; not for that they could
tranſgreſſe this precept, or commit a new ſinne, but for lack of
circumciſion. The reaſon wherof is, for that God now determined this
only remedie for original ſinne, in the male ſex of Abrahams ſeede, in
place of ſacrifice, or other profeſſion of faith vſed before, and that
in caſe it could be conueniently be applied for otherwiſe the former
remedies were ſtil aualable.

Concerning
\MNote{What puniſhment was threatned.}
the other difficultie manie ancient fathers expound theſe
wordes, \Emph{shal be deſtroyed out of his people}, of only temporal
puniſhment; either death, as the like phraſe ſignifieth,
\XRef{Exodi.~12.}
\Emph{VVhoſoeuer shal eate leauened bread, his ſoul shal perish out of
Iſrael}; or temporal ſeparation from the people of God, as
\XRef{Num.~19.}
\Emph{Euerie one that toucheth the dead corſe, shal perish out of
Iſrael.} But S.~Auguſtin
\Cite{(li.~3. c.~18. cont. Iulian}
\Cite{li.~5. hypog. & alibi.)}
S.~Gregorie
\Cite{(li.~4. Moral. c.~2. &.~3.)}
S.~Proſper
\Cite{(li. de promiſſ. Dei p.~1. c.~14.)}
S.~Beda
\Cite{(li.~2. in Lucam. c.~8.)}
and S.~Bernard
\Cite{(Ser.~3. de circum.)}
vnderſtand this threatning not only of temporal puniſhment, but alſo of
eternal ſeparation from God, and the ſocietie of Saints. And that alſo
infants are ſo ſecluded only for lack of this remedie, as now children
which dye without Baptiſme, cõmitting no new fault are depriued of the
viſion of God, for their original ſinne not remitted.
\MNote{Circumciſion inſtituted to diſtinguiſh the people of God and for
remedie of original ſinne in ſome perſons, but not in al.}
VVherupon is
geathered that albeit Circumciſion was principally inſtituted to
diſtinguiſh the people of God, which ſhould come of Abrahams ſeede, from
other nations, yet it was alſo for remiſſion of original ſinne: not in al,
but in thoſe to whom God appointed this particular remedie.}
shal be deſtroied out of his people: becauſe he hath broken my
couenant. \V God ſaid alſo to Abraham: Sarai thy wife thou shalt not cal
Sarai, but Sara. \V And I wil bleſſe her, and of her I wil giue thee a
ſonne, whom I wil bleſſe, and he shal be into nations, and kings of
peoples shal ſpring of him. \V Abraham fel vpon his face, &
\SNote{Abraham laughed not doubting but reioycing.
\Cite{S.~Ambro. li. de Abraham c.~4.}
\Cite{S.~Aug. li.~16. c.~26. de ciuit.}}
laughed, ſaying in his hart: Shal troweſt thou to him that is an hundred
yeare old a ſonne be borne? and Sara that is nyntie yeares old shal she
beare? \V And he ſaid to God: I would that Iſmael may liue before
thee. \V And God ſaid to Abraham: Sara thy wife shal beare thee a
%%% o-0080
ſonne, and thou shalt cal his name Iſaac, and I wil eſtablish my
couenant to him for a perpetual couenant, and to his ſeed after him. \V
Concerning Iſmael alſo I haue heard thee. Behold, I wil bleſſe him, and
encreaſe, and multiplie him exceadingly: twelue dukes shal he beget, and
I wil make him into a great nation. \V But my couenant I wil eſtablish
with
\SNote{Temporal bleſſings were common to Iſmael, but ſpiritual pertained
properly to Iſaac and Iſrael, & their
\Fix{ſucceſſeſſors.}{ſucceſſors.}{obvious typo, fixed in other}}
Iſaac, whom Sara shal bring forth to thee at this time an other
yeare. \V And when he had leaft of ſpeaking with him, God aſcended from
Abraham. \V And Abraham tooke Iſmael his ſonne, and al the homebred of
his houſe: and al whom he had bought, al the males of al the men of his
houſe: and he circumciſed the flesh of their prepuce forthwith the very
ſame day, as God had commanded him. \V Abraham was nyntie and nyne
yeares old, when he circumciſed the flesh of his prepuce. \V And Iſmael
his ſonne was ful thirtene yeares old at the time of his
circumciſion. \V The ſelfsame day was Abraham circumciſed and Iſmael his
ſonne. \V And
\SNote{Conformitie in Religion conſerueth peace in euery familie.
\Cite{Tho. Anglus in hunc locum.}}
al the men of his houſe, as wel the homebred, as the bought ſeruantes
and ſtrangers were circumciſed togeather.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
