%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 42
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-42


\project douay-rheims


%%% 0148
%%% o-0137
\startChapter[
  title={Chapter 42}
  ]

\Summary{Iacob ſendeth tenne of his ſonnes to bye corne in
  Ægypt. 7.~Where Ioſeph knowing them, they not knowing him, with hard
  ſpeaches putteth them in priſon. 18.~At laſt Simeon remayning in
  cuſtodie, til Beniamin be brought, the reſt are diſmiſſed, 25.~with
  their money, vnknowen to them, in their ſackes.}

And Iacob hearing, that victuales were ſold in Ægypt, he ſaid to his
ſonnes: Why neglect ye? \V I haue heard that wheat is ſold in Ægypt: Goe
ye downe, and bye vs neceſſaries, that we may liue, and not be conſumed
with lacke. \V Therfore the tenne brethren of Ioſeph going downe, to bye
corne in Ægypt, \V Beniamin being kept at home by Iacob, who ſaid to his
brethren: Leſt perhaps he take any harme in the iourney: \V entred into
the land of Ægypt with others that went to bye. For the famine was in
the land of Chanaan. \V And Ioſeph was prince in the land of Ægypt, and
at his pleaſure corne was ſold to the people. And when his brethren had
adored him, \V and he knewe them, he ſpake as it were to ſtrangers
\LNote{Some vvhat roughly.}{Ioſeph
\MNote{Contrition neceſſarie for the remiſſion of ſinnes.}
afflicted his brethren to bring them into conſideration of their former
faultes, and to true contrition. VVithout which, though iniuries be
remitted by men, yet the offenders are not abſolued before God. Therfore
he louing them, and hating their ſinne, by afflictions brought them to
vnderſtanding. VVho being at laſt truly penitent, he acknowledged and
moſt louingly embraced them, and prouided for them in their neceſſitie.
\Cite{S.~Aug. ſer.~82. de temp.}
And this example S.~Gregorie
\Cite{(ho.~22. in Ezech.)}
\Fix{applithe}{applieth}{obvious typo, fixed in other}
to the inſtruction of paſtors of ſoules: that they procure true
repentance before abſolution of ſinnes.
\MNote{Diſorderlie remiſſion is hurtful.}
\L{Ne ſi inordinate culpa dimittitur, is qui eſt culpabilis, in reatu
grauius aſtringatur.} \Emph{Leſt, if the fault be diſorderly remitted, he
that is faultie be more greuouſly intangled in guilt of conſcience}:
therfore with much diſcretion ſeueritie in ſhew, & clemencie in mind are
required.}
ſomewhat roughly, asking them: From whence came you? Who anſwered: From
the 
\Fix{Land of chanaan,}{land of Chanaan,}{obvious typo, fixed in other}
that we may bye neceſſaries to liue. \V And yet
himſelfe knowing his brethren, was not knowen of them. \V And remembring
the dreames, which ſome times he had ſeene, he ſaid to them:
\LNote{You are ſpies.}{To
\MNote{Ioſeph calleth his brothers ſpies for their good.}
the ſame purpoſe he 
\Fix{calleh}{called}{obvious typo, fixed in other}
them ſpies. After the maner of examiners
calling ſuſpected perſons, as they may ſeme to be: therby to trie what
they would anſwer. Al for their holſome penance, and withal to procure
them afterwards more compaſſion among the Ægyptians.
\Cite{S.~Aug. q.~139. ſuper Gen.}}
You are ſpies: to view the
%%% 0149
weaker parts of the land you are come. \V Who ſaid: It is not ſo, my
lord, but thy ſeruantes are come to bye victuals. \V We are al the
ſonnes of one man: we are come as men of peace, neither do thy ſeruantes
goe about anie euil. \V To whom he anſwered: It is otherwiſe: you
came to conſider the vndefenſed partes of this land. \V But they: We
thy ſeruantes, ſay they, are twelue brethren, the ſonnes of one man in
the Land of Chanaan: the youngeſt is with our father, the other is not
liuing. \V This is it, quoth he, that I ſaid: You are ſpies. \V Now
preſently I wil take a trail of you:
\LNote{By the health of Pharao.}{Ioſeph
\MNote{It is lawful to ſweare by creatures.}
in ſwearing by Pharaos health, honored God, the geuer and conſeruer of
health, life, power, dignitie and al that was in Pharao. Wherby we ſee
that this maner of ſwearing was lawful. As now likewiſe Chriſtians
lawfully ſweare not only by God, but alſo by his creatures, ſaying: As
God ſhal helpe them, & his holie Goſpels: ſo it be with due
circumſtances, and requiſite conditions, namely with thoſe which the
\CNote{Hier.~4.}
Prophet Hieremie mentioneth: \Emph{in truth, in iudgement, and in
iuſtice}: not falſe, raſh, nor of an vniuſt thing: nor in frequent and
cõmon talke (a moſt bad cuſtome) for ſo thou muſt neither ſweare by God
himſelf, nor (as
\CNote{Mat.~5.}
our Sauiour teacheth) by heauen, nor by the earth, nor by
Hieruſalem, nor by thy head, nor anie thing els:
\MNote{In ſome caſe more conuenient then to name God expreſly.}
ſignifying that in ſome caſe, and due maner we may ſweare by any of
theſe. And in this particular Ioſeph did rather name Pharao in his oath
then God, as wel to conceale himſelf as yet from his brethren, as to
ſtrike more terrour in their hartes, by naming his maſter the King, in
whom he had more intereſt then they.}
by the health of Pharao you ſhal not depart hence, vntil your youngeſt
brother do come. \V Send you one of you to bring him: and you ſhal be in
priſon, til the thinges be proued which you haue ſaid, whether they be
true or falſe:
\SNote{If theſe things which ye ſay be proued falſe, ye are to be held
as ſpies, for your lying.
\Cite{S.~Aug. q.~139. ſuper Gen.}}
otherwiſe by the health of Pharao you are ſpies. \V Therfore he put them
in priſon three dayes. \V And the third day bringing them out of priſon,
he ſaid: Doe as I haue ſaid, and you ſhal liue: for I feare God. \V If
you be men of peace, let one of your brethren be bounde in priſon: and
goe ye your wayes, and carie the corne that you haue bought, vnto your
houſes, \V and bring your youngeſt brother to me, that I may
%%% o-0138
proue your
ſayings to be true, and that you die not. They did as he had ſaid, \V
and they talked one to an other: Worthely do we ſuffer theſe thinges,
becauſe we haue ſinned againſt our brother, ſeing the diſtreſſe of his
ſoule, whileſt he beſought vs, and we heard not: therfore is this
tribulation come vpon vs. \V Among whom Ruben one of them, ſaid: Why,
did not I ſay to you: Sinne not againſt the boy: and you heard me not?
Loe his blood is required. \V And they knew not that Ioſeph vnderſtood:
becauſe he ſpake to them by an interpreter. \V And he turned away himſelfe a
litle while, and wept: and returning he ſpake to them. \V And taking
Simeon, and binding him in their preſence, he commanded his ſeruantes
that they ſhould fil vp their ſackes with wheat, and put euerie mans
money againe in their bagges, geuing them beſides for to eate on the way:
who did ſo. \V But they carying their corne vpon their aſſes, tooke
their iourney. \V And one of them opening his ſacke, to geue his beaſt
prouender in the inne, beholding the money in the ſackes mouth, \V he
ſaid to his brethren: My money is geuen me againe, behold it is in the
%%% 0150
ſacke. And being aſtonyed, and trubled amongſt themſelues, they ſaid:
What is this, that God hath done vnto vs? \V And they came to Iacob
their father into the land of Chanaan, and they told him al thinges that
had chanced vnto them, ſaying: \V The lord of the land ſpake to vs
roughly, and thought vs to be ſpies of the prouince, \V to whom we
anſwered: We are men of peace, neither do we attempt any treachery. \V
We are twelue brethren borne of one father: one is not liuing, the
youngeſt is with our father in the Land of Chanaan. \V Who ſaid to vs:
Thus ſhal I trie that you be men of peace: Leaue one of your brethren
with me, and take ye prouiſion neceſſarie for your houſes, and goe your
wayes, \V and bring your youngeſt brother to me, that I may know you are
not ſpies: and you may receiue this felowe againe, that is kept in
priſon: and afterwardes may haue licence to bye what thinges you wil. \V
This being ſaid, when they powred out their corne, euerie man found his
money tied in the mouth of the ſackes: and al being aſtonyed
together, \V their father Iacob ſaid: You haue made me to be without
children, Ioſeph is not liuing, Simeon is kept in bondes, and Beniamin
you wil take away: al theſe euils are fallen vpon me. \V To whom Ruben
anſwered: Kil my two ſonnes, if I bring him not againe to thee: deliuer
him into my hand, and I wil reſtore him vnto thee. \V But he ſaid: My
ſonne ſhal not go downe with you: his brother is dead, and he alone is
remayning: if any aduerſitie shal chance to him in the land to the which
you goe, you shal bring downe
\SNote{Myn old age or, me an old man.
\Cite{S.~Aug. q.~142.}}
my hoare
\Fix{heares}{haires}{likely typo, fixed in other}
with ſorowe
\LNote{Vnto hel.}{To that place where ſoules remaine, as before.
\XRef{chap.~37. v.~35.}
\MNote{Diuers manſions in hel.}
For this phraſe, of \Emph{bringing vnto hel}, and \Emph{deſcending into
hel}, vſually ſpoken in the old Teſtament, of al ſortes of ſoules both
iuſt and wicked, ſignifieth that al went that time to hel, that is, to a
lowe place, farre diſtant from heauen. But ſome to reſt, and ſome to
paine. VVherupon S.~Hierom ſaith: Hel is a place, wherin ſoules are
included, either in reſt, or in paines, according to the qualitie of
their deſerts.}
vnto hel.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
