%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 38
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-38


\project douay-rheims


%%% 0139
%%% o-0129
\startChapter[
  title={Chapter 38}
  ]

\Summary{Iudas hauing three ſonnes by a Chananite, 6.~marieth the firſt,
  and after his death, the ſecond to Thamar. 10.~Who alſo dying, he
  delayeth to match the third with her. 15.~But him ſelfe begetteth of
  her (taking her for a harlote) two ſonnes twinnes, Phares and Zara.}

The
\SNote{Moyſes inſerteth here this hiſtorie, becauſe Chriſt ſhould be
borne of the genealogie of Iudas & Phares.
\XRef{Mat.~1.}}
ſame time Iudas going downe from his brethren, turned in to a man an
Odollamite, named Hiras. \V And he ſawe there the daughter of a man of
Chanaan, called Sue: and taking her to wife, he did companie with
her. \V Who conceaued, and bare a ſonne, and called his name Her. \V And
conceauing a childe againe, ſhe called her ſonne after he was borne,
Onan. \V She bare alſo the third: whom ſhe called Sela: after whoſe
birth, ſhe ceaſed to beare any more. \V And Iudas gaue a wife to Her his
firſt begotten, named Thamar. \V Alſo Her the firſt begotten of Iudas,
was wicked in the ſight of our Lord: and was ſlaine of him. \V Iudas
therfore ſaid to Onan his ſonne: companie with thy brothers wife, and be
ioyned to her, that thou mayeſt
\LNote{Raiſe ſede to thy brother.}{By
\MNote{How a man might marie his brothers wife in the law of nature.}
this it appeareth, that in the law of nature, when a maried man died
without iſſue, his brother might lawfully marie the widow: whoſe firſt
ſonne ſhould be counted and called the ſonne and heyre of his vncle dead
before. The ſame was eſtabliſhed by the law of Moyſes
\XRef{(Deut.~25.)}
\MNote{The Churches decree is now our rule.}
VVhich being now abrogated, it remaineth in the Churches powre to
conſtitute a law in this behalf, and conſequently to diſpence in the
ſame, ſo farre as it is agreable with the law of nature. VVherof ſee
more.
\XRef{Leuit.~18.}}
rayſe ſeede to thy brother. \V He knowing that the children ſhould not
be borne to himſelfe, companying with his brothers wife, ſhed his ſeede
vpon the ground, leſt children might be borne in his brothers name. \V
And therfore our Lord ſtroke him, becauſe he did a deteſtable thing. \V
For the which cauſe Iudas ſaid to Thamar his daughter in lawe: Be a
widowe in thy fathers houſe, til Sela my ſonne growe vp: for he feared
leſt he alſo might dye, as his brethren. Who went her way and dwelt in
her fathers houſe. \V And after many dayes were come and gone: the
daughter of Sue the wife of Iudas died: who after his mourning hauing
receiued conſolation, went vp to the ſhearers of his ſheepe, himſelfe
and Hiras his shepheard of his flocke, the Odolamite, into Thamnas. \V
And it was told
\SNote{Thamar ſinned deſiring to be a mother without lawful mariage: and
Iudas ſinned lying with a ſuppoſed harlot.
\Cite{S.~Aug. li.~22. c.~61.~62. &~63. cont. Fauſt.}}
Thamar that her father in law came vp into Thamnas to sheare his
sheepe. \V Who putting of the garments of her widowhood, tooke a veile:
and changing her habite, ſate in the croſſe way, that leadeth to
Thamnas: becauſe Sela was growne, and she had not taken him to
huſband. \V Whom when Iudas had
%%% 0140
ſeene, he ſuppoſed her to be an harlot: for she had couered her face,
leſt she should be knowen. \V And going vnto her, he ſaid: Suffer me to
lye with thee: for he knew her not to be his daughter in law. Who
anſwering: What wilt thou geue me that thou maieſt enioy my companie? \V
He ſaid: I wil ſend thee a kid out of the flockes. And when ſhe ſaid
againe: I wil ſuffer that thou wilt, if thou geue me a pledge, til thou
%%% o-0130
ſend that which thou doeſt promiſe; \V Iudas ſaid: What wilt thou to be
geuen thee for a pledge? She anſwered: Thy ring, and bracelet, and
ſtaffe which thou holdeſt in thy hand. The woman therfore by once
companying conceaued, \V and riſing ſhe went her way: and putting of the
apparel which ſhe had taken, put on the garments of her widowhood. \V
And Iudas ſent a kid by his shephard the Odolamite, that he might
receiue the pledge againe, which he had geuen to the woman: who when he
had not found her, \V he asked the men of that place: Where is the woman
that ſate in the croſſe way? Al making anſwere: There was no harlot in
this place. \V He returned to Iudas, and ſaid to him: I haue not found
her: yea the men alſo of that place ſaid vnto me, that there neuer ſate
harlot there. \V Iudas ſaid: Let her take it to her, ſurely she can not
charge vs with a lye, I ſent the kid which I promiſed: and thou dideſt
not find her. \V And behold after three moneths they told Iudas, ſaying:
Thamar thy daughter in law hath played the harlot, and her bellie ſemeth
to ſwel. And Iudas ſaid: Bring her forth
\SNote{Adultrie
\Fix{puniſable}{puniſhable}{likely typo, fixed in other}
by death, in the law of nature.}
that she may be burnt. \V Who when she was led to execution, she ſent to
her father in law, ſaying: By that man, whoſe theſe things are, haue I
conceaued: looke whoſe the ring is, and the bracelet, and the ſtaffe. \V
Who acknowledging the giftes, ſaid: She is iuſter then I: becauſe I did
not geue her to Sela my ſonne. But he knew her no more. \V And when she
was readie to be brought to bed, there appeared twinnes in her bellie:
and in the verie deliuerie of the infants, one put forth the hand,
wherin the midwife tyed a ſkarlet ſtring, ſaying: \V This shal come
forth the former. \V But he drawing backe his hand, the other came
forth: and the woman ſaid: Why is the partition diuided for thee? and
for this cauſe called his name Phares. \V Afterward his brother came
forth, in whoſe hand was the ſkarlet ſtring: whom she called Zara.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
