%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 43
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-43


\project douay-rheims


%%% 0151
%%% o-0140
\startChapter[
  title={Chapter 43}
  ]

\Summary{The famine preſsing the land, Iacob willeth his ſonnes to goe
  againe into Ægypt to bye more corne: 3.~who refuſing to goe without
  Beniamin, 11.~he is ſent with them, and preſents and duble money, leſt
  the former were brought back by error. 16.~Ioſeph ſeing Beniamin,
  23.~deliuereth Simeon, and intertaineth them al at dinner.}

In the meane time the famine did oppreſſe al the land very ſore. \V And
the prouiſion being ſpent, which they had brought out of Ægypt, Iacob
ſaid to his ſonnes: Goe againe, and by vs a litle victual. \V Iudas
anſwered: That ſame man denounced vnto vs vnder atteſtation of an oath,
ſaing: You shal not ſee my face, vnleſſe you bring your
%%% 0152
youngeſt brother with you. \V If therfore thou wilt ſend him with vs, we
wil goe forward together, and wil bye neceſſaries for thee: \V but if
thou wilt not, we wil not goe: for the man, as we haue often ſaid,
denounced vnto vs ſaing: You shal not ſee my face without your youngeſt
brother. \V Iſrael ſaid to them: You haue done this to my miſerie, in
that you told him that you had an other brother alſo. \V But they
anſwered: The man asked vs in order our progenie: if our father liued:
if we had a brother: and we anſwered him conſequently to that which he
demanded: could we know that he would ſay: Bring hither your brother
with you? \V Iudas alſo ſaid to his father: Send the boy with me, that
we may ſet forwarde, and may liue: leſt we and our litle ones die. \V I
take vpon me the boy: require him of my hand, vnleſſe I bring him
againe, and reſtore him to thee, I wil be
\SNote{Guilt of ſĩne is a greater bond then the life of Rubens ſonnes;
which he offered, & Iacob yelded not therto: & yet granted to this offer
of Iudas.}
guiltie of ſinne againſt thee for euer. \V If delay had not bene made,
we had bene come now the ſecond time. \V Therfore Iſrael their father
ſaid to them: If it muſt nedes be ſo, do that you wil: take of the beſt
fruites of the land in your veſſels, and carie to the man for preſents,
a courteſie of roſen, and of honey, and of incenſe, of mirhe, of
terebinth, and of almondes. \V Duble money alſo carie with you: and
recarie that you founde in your ſackes, leſt perhaps it was done by an
errour: \V but take alſo your brother, and goe to the man. \V And my
God almightie make him fauorable vnto you: and ſend backe with you your
brother, whom he keepeth, and this Beniamin: as for me I ſhal be
deſolate without children. \V The men therfore tooke the preſentes, and
the duble money, and Beniamin: and went downe into Ægypt, and ſtood
before Ioſeph. \V Whom when he had ſeene, and Beniamin withal,
%%% o-0141
he commanded the ſteward of his houſe, ſaing: Bring in the men into the
houſe, and kil victimes, and prepare a feaſt: becauſe they ſhal eate
with me at noone. \V He did that which was commanded him, and brought
the men into the houſe. \V And there being ſore afraied, they ſaid one
to an other: Becauſe of the money, which we caried backe the firſt time
in our ſackes, we are brought in: that he may turne vpon vs
\TNote{\L{Calumniam.}}
a falſe accuſation, and forcibly bring both vs, and our aſſes into
bondage. \V Wherfore in the verie dore ſtepping to the ſteward of the
houſe, \V they ſpake: Sir we deſire thee to
%%% 0153
heare vs. Now once before we came downe to bye prouiſion: \V which being
bought, when we were come to the Inne, we opened our ſackes, and found
our money in the mouthes of the ſackes: which we haue now brought againe
in the ſame weight. \V But we haue brought other money beſides, to bye
the thinges that are neceſſarie for vs: our conſcience is not priuie,
who put it in our bagges. \V But he anſwered: Peace be with you, feare
you not: your God, and the God of your father hath geuen you treaſures
in your ſackes. For the money, which you gaue me, I haue for good. And he
brought forth Simeon vnto them. \V And being brought into the houſe, he
fetched water, and they waſhed their feete, and he gaue prouender to
their aſſes. \V But they made readie the preſentes, til Ioſeph ſhould
come in at noone: for they had heard that they ſhould eate bread
there. \V Therfore Ioſeph came in to his houſe, and they offered him
preſentes holding them in their handes, and
\SNote{They now adore him, whom they ſold, leſt they ſhould adore him.
\Cite{S.~Greg. ho.~22. in Ezech.}}
they adored proſtrate to the ground. \V But he curteouſly reſaluting
them, asked them ſaying: Is the old man your father in health, of whom
you told me? Is he yet liuing? \V Who anſwered: Thy ſeruant our father
is in health, he is yet liuing. And bowing themſelues, they adored
him. \V And Ioſeph lifting vp his eyes, ſawe Beniamin his brother of the
ſame mother, and ſaid: Is this your young brother, of whom you told me?
And againe: God, ſaith he, be merciful vnto thee my ſonne. \V And he
made haſt becauſe his hart was moued vpon his brother, and teares brake
forth: and entring into his chamber he wept. \V And when he had waſhed
his face, comming forth againe, he refrained himſelfe, and ſaid: Set
bread on the table. \V Which being ſet downe, to Ioſeph apart, and to
his brethren apart, to the Ægyptians alſo that did eate with him apart
(for it is vnlawful for the Ægyptians to eate with the Hebrewes, and
\SNote{See
\XRef{Exodi.~8. v.~26.}}
they thinke ſuch a feaſt prophane) \V they ſate before him, the firſt
begotten according to his firſt-birth, and the youngeſt according to his
age. And they maruailed out of meaſure, \V taking the portions that they
receiued of him: and the greater portion came to Beniamin, ſo that it
exceeded
\SNote{Euerie one hauing fiue portions, Beniamin had duble.
\Cite{Ioſephus li.~1. Antiq.}}
by fiue partes. And they dranke and were
\LNote{Inebriated.}{Not
\MNote{Moderation to be vſed in feaſting.}
that they did eate or drinck to much, or fel into exceſſe, but
competently. As the earth is ſaid to be inebriated with raine
\XRef{(Pſal.~64.)}
being ſufficiently watered, and ſo made fructful, not drowned, nor
ouerflowed, for ſo it is vnfructful.
\Cite{S.~Hierom. Tradit. Heb.}
\Cite{S.~Aug. q.~144. ſuper Gen.}}
inebriated with him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
