%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 0079
%%% o-0075
\startChapter[
  title={Chapter 15}
  ]

\Summary{Abram doubting and lamenting that he should haue no childe, God
  promiſeth him much iſsue, 6.~who beleuing is iuſtified, 9.~offereth
  Sacrifice preſcribed by God, 13.~and is forewarned that his ſeede shal
  be in ſtrange land 400.~yeares, 14.~shal be deliuered from ſeruitude
  18.~and poſseſse Chanaan.}

VVhen theſe things therfore were done, the word of our Lord was made to
Abram by a viſion ſaying: Feare not Abram, I am thy protector, & thy
reward exceding great. \V And Abram ſaid: Lord God, what wilt thou geue
me? I ſhal goe without children: and the ſonne of the ſtuard of my houſe
is this Damaſcus Eliezer. \V And Abram added: But to me thou haſt not
geuen ſeede: and loe my ſeruant borne in my houſe, ſhal be myn heire. \V
And immediatly the word of our Lord came to him ſaying: He ſhal not be
thy heyre: but he that ſhal come out of thy wombe, him ſhalt thou haue
thin heire. \V And he brought him forth abroad, and ſaid to him: Looke
vp to heauen, and number the ſtarres, if thou canſt. And he ſaid to him:
So ſhal thy ſeed be.
%%% 0080
\V
\CNote{Ro.~4. Gal.~3. Iac.~2.}
Abram
\LNote{Beleued God.}{S.~Hilarie
\MNote{To beleue Gods word without ſtaggering is an act of iuſtice.}
\Cite{(li.~9. de Trin.)}
and S.~Ambroſe
\Cite{(li.~1. de Abraham. c.~3.)}
by this example teach vs, what maner of faith is reputed to iuſtice, to
wit, ſuch a faith as without tergiuerſation, or requiring of proofe or
reaſon, doth ſimply beleue that which God once ſaith, becauſe he is
omnipotent, how farſoeuer the thing that is ſaid ſurmounteth our
vnderſtanding. For ſo heroical was the act of Abrahams faith, promptly
beleuing Gods word in a matter moſt hard to his former conceipt, that
for the ſame he receiued ſingular prayſe; and for the like afterwards
was called \Emph{The father of manie nations}
\XRef{(c.~17. v.~5.)}
and by S.~Paul, \Emph{The father of al that beleue}
\XRef{(Ro.~4. v.~11.)}
where the Apoſtle teacheth that
\MNote{Not workes before faith but ioyned with faith are meritorious.}
Abraham had no iuſtice, nor eſtimation of iuſtice before God, vntil he
beleued in Chriſt 
\XRef{(v.~18, 19,~20.)}
becauſe al workes before that faith are inſufficient. Neyther was this a
ſole faith but had other neceſſarie vertues of hope, and charitie,
humilitie, reuerence, obedience, & the like ioyned with it. Wherfore
S.~Iames teſtifieth that Abraham was iuſtified by workes (that is by
workes folowing faith, not going before faith.)
\MNote{Onlie faith doth not iuſtifie.}
For \Emph{Faith} (ſaith he) \Emph{if it haue not vvorkes, is dead in it
ſelf.}
\XRef{(c.~2. v.~17.)}
\Emph{And by vvorkes Abrahams faith vvas conſumat.}
\XRef{v.~22.}
And concludeth thus: Do yee ſee, that by workes a man is iuſtified and
not by faith only.
\XRef{v.~24.}}
beleued God, and it was reputed to him vnto iuſtice. \V And he ſaid to
him: I am the Lord that brought thee out from Vr of the Chaldees for to
giue thee this land, and that thou mighteſt poſſeſſe it. \V But he ſaid:
Lord God, how may I know that I ſhal poſſeſſe it? \V And our Lord
anſwered, and ſaid:
\SNote{Theſe three kindes of beaſtes and two of birdes ſignifie that the
Iſraelites ſhould be three generations in a ſtrange land, the fourth in
the deſert, the fifth in poſſeſſion of Chanaan,
\Cite{Theod. q.~65. in Gen.}}
Take me a cowe of three yeares old, and a shee goat of three yeares, and
a ramme of three yeares, a turtle alſo, and a pigeon. \V Who taking
al theſe, diuided them by the mydes, and laid ech two peeces arowe one
againſt the other: but the birdes he diuided not. \V And the foules
lighted vpon the carcaſſes, and Abram droue them away. \V And when the
ſunne was ſetting, a deepe ſleepe fel vpon Abram, and a great and
darkeſome horrour inuaded him. \V And it was ſaid vnto him:
\CNote{Act.~17.}
Know and
foreknow that a pilgrime shal thy ſeede be in a land not their owne (and
they shal bring them vnder bondage, and afflict them)
\SNote{Abraham and his ſeed were in ſtrange land 400.~and odde yeares,
but in ſeruitude and affliction about 140.}
foure hundred yeares. \V
\CNote{Exo.~12.}
But the nation, whom they shal ſerue, I wil
iudge: and after this they shal goe forth with great ſubſtance. \V And
thou shalt goe to thy fathers in peace, buried in a good old age. \V But
in the fourth generation they shal returne hither: for
\SNote{God deferreth to puniſh, either that the wicked may amend, or the
good be exerciſed by them.
\Cite{S.~Aug. Pſal.~54.}
or becauſe the iniquity is not come to that great meaſure which his
wiſdome forſeeth, and wil puniſh in the end, to his owne more glorie,
and more good of others.
\Cite{S.~Greg. ho.~11. in 3.~Ezech.}}
as yet the iniquities of the Amorrheanes are not at the ful vntil this
preſent time.

%%% o-0076
\V Therfore when the ſunne was ſet, there aroſe a darke miſt, and there
appeared a fornace ſmoking, and a flake of fire paſſing betwene thoſe
diuiſions. \V That day God made a couenant with Abram, ſaying: To thy
ſeede wil I geue this land from the riuer of Ægypt euen to the great
riuer Euphrates, \V the Cineans, & Cenezites, the Cedmonites, \V and the
Hethits, and the Pherezits, the Raphaims alſo, \V and the Amorreans & the
Cananites, and the Gargaſites, and the Iebuſites.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
