%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 33
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-33


\project douay-rheims


%%% 0127
%%% o-0118
\startChapter[
  title={Chapter 33}
  ]

\Summary{Iacob ſeing Eſau come with a great troupe of men, feareth
  harme, but is moſt curteouſly entertained by him. 10.~He hardly
  perſwadeth Eſau to take giftes, 13.~and to returne home. 17.~So Iacob
  coming by Socoth to Salem, there byeth a field, pitcheth his tents,
  and erecteth an Altar.}

And Iacob lifting vp his eyes, ſaw Eſau coming, and with him foure
hundred men: and he diuided the children of Lia and of Rachel, and of
the two handmaides: \V and he put both the handmaids & their children
foremoſt: and Lia, and her children in the ſecond place: and Rachel, and
Ioſeph laſt. \V And himſelfe going foreward adored proſtrate to the
grownd ſeuen times, vntil his brother came nere. \V Eſau therfore
running to mete his brother, embraced him: and claſping him faſt about
the necke, and kiſſing him wept. \V And caſting vp his eyes, he ſaw the
wemen and their litle ones, and ſaid: What meane theſe? And do they
perteyne to thee? He anſwered: They are the litle ones which God hath
geuen to me thy ſeruant. \V And the handmaides and their children coming
nere, bowed themſelues. \V Lia alſo with her children came nere: and
when they had adored in like maner, laſt Ioſeph and Rachel adored. \V
And Eſau ſaid: What are the troupes that I did mete? He anſwered: That I
might find grace before my lord. \V But he ſaid: I haue plentie, my
brother, be thy things to thy ſelfe. \V And Iacob ſaid: Do not ſo I
beſech thee, but if I haue found grace in thin eyes, take a litle
preſent at my hands: for ſo haue I ſeene thy face, as if I ſhould haue
ſeene
\SNote{Iacob ſeing Gods hand in this change of his brothers mind, not of
flaterie, but ſincerly acknowledged his benignitie, as Gods
countenance towards him.}
the countenance of God: be gracious to me, \V and take the bleſſing,
which I haue brought thee, and which God hath geuen me, who geueth al
thinges. Scarſe at his brothers great inſtance, taking it, \V he ſaid:
Let vs march on together, and I wil accompanie thee in thy iourney. \V
And Iacob ſaid: My lord thou knoweſt that I haue with me litle ones, and
%%% o-0119
ſheepe, and kine with young: which if I cauſe to ouerlaboure themſelues
in going, in one day al the flockes wil die. \V It may pleaſe my lord to
goe before his ſeruant: and I wil folow ſoftly after him, as I ſhal ſee
my litle ones
%%% 0128
to be able, vntil I come to my lord in Seir. \V Eſau anſwered: I beſech
thee, that of my people at the leaſtwiſe, which is with me, there may
remaine ſome to accompanie thee in the way. It is not needful, ſaid he,
this only I haue nede of, that I may finde grace (my lord) in thy
ſight. \V Eſau therfore returned that day the ſame way, that he came
into Seir. \V And Iacob cometh into Socoth: where hauing built a houſe,
and pitched his tents, he called the name of that place Socoth, that is,
Tabernacles. \V And he paſſed into Salem a citie of the Sichimites,
which is in the land of Chanaan, after he returned from Meſopotamia of
Siria: and he dwelt beſide the towne. \V And
he bought that part of the field, wherin he had pitched his tents, of
the children of Hemor, the father
\Fix{fo}{of}{obvious typo, fixed in other}
Sichem for an hundred lambes. \V And erecting an altar there, on it he
called vpon the moſt mightie God of Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
