%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 0071
%%% o-0067
\startChapter[
  title={Chapter 12}
  ]

\Summary{Abram
\MNote{The begining of the third age.}
commanded by God to leaue his countrie, with promiſe to be bleſſed in
his Seede, 5.~taking his wife Sarai, and his nephew Lot, 6.~wandreth in
the land of Chanaan, 7.~erecteth an Altar in Sichem, 8.~an other in
Bethel. 10.~Thence by occaſion of famine paſseth into Ægipt. 14.~Where
his wife (called his ſiſter) is taken into the kings houſe, 19.~but
vntouched is reſtored to him.}

And
\CNote{Act.~7.}
\MNote{The fifth part of this booke. Of Abrahams leauing his coũtrie,
Gods bleſſing of his ſeede, & commandment of Circumciſion.}
our Lord ſaid to Abram: Goe forth of thy countrie, and out of thy
kinred, and out of thy fathers houſe, and come into a land, which I wil
ſhew thee. \V And I wil make thee into a great nation, and I wil bleſſe
thee, and magnifie thy name, and thou ſhalt be bleſſed. \V I wil bleſſe
them
\Fix{and}{that}{obvious typo, fixed in other}
bleſſe thee, and curſſe them that curſſe thee, and
\SNote{In Chriſt borne of Abrahames ſeede al nations are bleſſed.
\XRef{Gal.~3.}}
\Sc{In thee} ſhal al the kindreds of the earth be bleſſed. \V
\CNote{Heb.~11.}
Abram
therfore went out as our Lord had commanded him, and with him went Lot:
ſeauentie fiue yeares old was Abram when he went forth out of Haran. \V
And he tooke Sarai his wife, and Lot his brothers ſonne, and al the
ſubſtance which they had poſſeſſed, and the ſoules which they had gotten
in Haran: and went forth to goe into the land of Chanaan. And when they
were come into it, \V Abram paſſed through the countrie vnto the place
Sichem, as farre as the noble vale: and the Cananite was at that time
in the countrie. \V And our Lord appeared to Abram, and ſaid to him: To
thy ſeed wil I giue this land. Who builded there
\SNote{Abram dedicated Altars to God eſpecially in thoſe places where he
receiued promiſes, or benefites.
\Cite{S.~Chriſoſt. ho.~31. in Gen.}}
an altar to our Lord, that had appeared to him. \V And marching on from
thence to a mountaine, that was on the eaſt ſide of Bethel, there he
pitched his tent, hauing Bethel on the weſt, and Hay on the eaſt: he
builded there alſo an altar to our Lord, and called vpon his name. \V
And Abram went forward going, and proceding on to the ſouth. \V And
there came a famine in the countrie: and Abram deſcended into Ægypt, to
be as a pilgrime there: for the famine was very ſore in the land. \V And
when he was nere to enter into Ægypt, he ſaid to Sarai his wife: I know
that thou art a fayre woman: \V and that when the Ægyptians ſhal ſee
thee, they wil ſay: She is his wife: and they wil
\Fix{til}{kil}{obvious typo, fixed in other}
me, and reſerue
%%% 0072
thee. \V
\LNote{Say therfore.}{Abraham
\CNote{Gen.~15.}
\MNote{Men are bond to do their lawful endeuoures, and to commit the
reſt to God.}
concealed that Sarai was his wife, and lyed not in ſaying, ſhe was his
ſiſter: as he alſo called Lot his brother, being his brothers ſonne, and
ſhe his brothers daughter. VVherby he preuented danger of his owne life,
vſing ſuch lawful meanes as lay in him, committing his wiues chaſtitie
to Gods protection, which him ſelf could not prouide for. In which caſe
if he had not donne his owne endeuour, he had rather tempted God (ſaith
\CNote{li.~16. c.~19. ciuit. & q.~26. in Gen.}
S.~Auguſtin) then truſted in God. And ſo God preſerued her, though ſhe
was in Pharao his houſe.
\XRef{v.~17.}}
Say therfore, I pray thee, that thou art my ſiſter: that I may be wel
vſed for thee, and that my ſoule may liue for thy ſake.

\V When Abram therfore was entred into Ægypt, the Ægiptians
%%% o-0068
ſawe the woman that ſhe was paſſing beautiful. \V And the princes told
Pharao, and praiſed her to him: and the woman was taken into the houſe
of Pharao. \V And they vſed Abram wel for her ſake. And he had ſheepe
and oxen and he aſſes, and men ſeruants, and maid ſeruants, and ſhee
aſſes, and Cammels. \V But our Lord
\SNote{God by corporal affliction hindered Pharao and his men from doing
violence to Sarai.
\Cite{S.~Chriſoſt. ho.~32. in Gen.}}
ſcourged Pharao with very ſore plagues, and his houſe for Sarai Abrams
wife. \V And Pharao called Abram, and ſaid to him: What is this that
thou haſt done to me? why didſt thou not tel me that ſhe was thy
wife? \V For what cauſe didſt thou ſay, ſhe was thy ſiſter, that I might
take her to my wife? Now therfore there is thy wife, take her, and goe
thy ways. \V And Pharao gaue certaine men commandment in the behalfe of
Abram: and they conducted him, and his wife, and al that he had.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
