%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 18
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-18


\project douay-rheims


%%% 0086
%%% o-0082
\startChapter[
  title={Chapter 18}
  ]

\Summary{Angels intertained as gheſts by Abraham 10.~tel when Sara shal
  beare a ſonne, wherat she laughing, they cõfirme that they had
  ſaid. 16.~They alſo fortel the deſtruction of Sodom, 22.~for which
  Abraham prayed ſix times.}

And God appeared to him in the vale of mambre as he ſat in the dore of
his tent, in the verie heat of the day. \V And when he had lifted vp his
eyes, there appeared to him three men ſtanding nere vnto him: whom after
he had ſene,
%%% 0087
he ranne to meete them from the dore of his tent, and adored to the
ground. \V And he ſaid:
\CNote{Heb.~13.}
\SNote{Abraham ſaw three, and adored one, profeſſing three diuine
perſons and one God.
\Cite{S.~Aug. li.~16. c.~29. ciuit.}}
Lord, if I haue found grace in thy ſight, goe not paſt thy ſeruant: \V
but I wil fetch a little water, and waſh ye your feete, and reſt ye vnder
the tree. \V And I wil ſet a morſel of bread, and ſtrengthen your hart,
afterward you ſhal paſſe: for therfore are you come aſide to your
ſeruãt. Who ſaid: Do as thou haſt ſpoken: \V Abraham made haſt into the
tent to Sara, and ſaid to her: Make haſt, temper togeather three
meaſures of floure, and make harth cakes. \V But him ſelfe ranne to the
heard, and tooke from thence a calfe verie tender and verie good, and
gaue it to a young man: who made haſt and boiled it. \V He tooke alſo
butter and mylke, and the calfe which he had boyled, and ſet before
them: but him ſelfe did ſtand beſide them vnder the tree. \V And when
they had eaten, they ſaid to him: Where is Sara thy wife? He anſwered:
Loe ſhe is in the tent. \V To whom he ſaid: Returning I wil come to thee
at this time, life accompaning, and
\CNote{Rom.~9.}
Sara thy wife ſhal haue a
ſonne. Which when Sara heard, ſhe laughed behind the dore of the
tent. \V And they were both aged, and farre entred in yeares, and it
ceaſed to be with Sara after the maner of wemen. \V Who laughed
ſecreatly ſaying: After I am waxen old, & my Lord is an old one, ſhal I
geue my ſelfe to pleaſure? \V And our Lord ſaid to Abraham: Why did Sara
laugh, ſaying: Shal I an old woman beare a child in deed? \V Is there
any thing hard to God? According to appointment I wil returne to thee
this verie ſelfe ſame time, life
\Fix{accompaynig,}{accompanying,}{obvious typo, fixed in other}
and Sara ſhal haue a ſonne. \V Sara denied, ſaying: I laughed not: being
much afraid. But our Lord:
\SNote{Abraham laughing with admiration for ioy was not reprehended, but
Sara laughing of diffidence was reprehended, by him that ſeeth the
hart.
\Cite{S.~Aug. q.~36. in Gen.}
&
\Cite{li.~16. c.~31. ciuit.}}
It is not ſo, ſaith he: but thou didſt laugh. \V When the men therfore
were riſen vp from thence, they turned their eyes againſt Sodome: and
Abraham did goe with them, bringing them on the way. \V And our Lord
ſaid: Can I conceale from Abraham the things which I wil doe: \V wheras
he ſhal be into a nation great, and verie ſtrong, and \Sc{in him
%%% o-0083
are to be blessed al the nations of the earth}? \V For I know that he
wil commande his children, and his houſe after him that they kepe the
way of the Lord, and doe iudgement and iuſtice: that for Abrahams ſake
the Lord may bring to effect al the things that he hath ſpoken vnto him.

Therfore ſaid our Lord: \V The crye of Sodome, and
%%% 0088
Gomorre is multiplied, and their ſinne is aggrauated excedingly. \V I
wil deſcend, and ſee whether they haue in acte accompliſhed the crye
that is come to me: or whether it be not ſo, that I may know. \V And they
turned them ſelues from thence, and went their way to Sodome: but
Abraham as yet ſtood before our Lord. \V And approching he ſaid: What!
wilt thou deſtroy the iuſt with the wicked? \V
\SNote{What a wall are iuſt men to their countrie? their faith ſaueth
vs, their iuſtice defendeth vs from deſtruction.
\Cite{S.~Amb. li.~1. de Abraham. c.~6.}}
If there ſhal be fiftie iuſt perſons in the citie, ſhal they periſh
withal? and wilt thou ſpare that place for fiftie iuſt, if they be
therin? \V Be it farre from thee, that thou doe this thing, and that
thou kil the iuſt with the wicked, and that the iuſt be in like caſe as
the wicked, this is not beſeeming thee: which iudgeſt al the earth, no
thou wilt not do this iudgement. \V And our Lord ſaid to him: If I ſhal
find in Sodome fiftie iuſt perſons within the citie, I wil ſpare the
whole place for their ſake. \V And Abraham anſwered, and ſaid: Becauſe I
haue once begunne, I wil ſpeake to my Lord, wheras I am duſt and
aſhes. \V What if there shal be fiue leſſe then fiftie iuſt perſons?
wilt thou for fortie fiue deſtroy the whole citie? And he ſaid: I wil
not deſtroy it, if I shal finde fiue and fourtie. \V And againe he ſaid
vnto him: But if fourtie shal be found there, what wilt thou doe? He
ſaid: I wil not ſtrike it for fourties ſake. \V Lord, ſaith he, be not
angrie I beſech thee, if I ſpeake: What if thirtie shal be founde there?
He anſwered: I wil not doe it, if I shal find thirtie there. \V Becauſe,
ſaith he, I haue once begunne, I wil ſpeake to my Lord: What if twentie
shal be founde there? He ſaid: I wil not deſtroy it for twenties
ſake. \V I beſeech thee, ſaith he, be not angrie Lord, if I ſpeake yet
once more: What if tenne shal be found there? And he ſaid: I wil not
deſtroy it for tennes ſake. \V And our Lord departed after that he
ceaſed to ſpeake vnto Abraham: and Abraham returned into his place.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
