%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 31
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-31


\project douay-rheims


%%% 0120
%%% o-0112
\startChapter[
  title={Chapter 31}
  ]

\Summary{Iacob by Gods commandment parteth ſecretly with al he hath
  towards his father. 21.~Laban purſueth him, 26.~expoſtulating why he
  went in ſecrete maner, 30.~eſpecially chargeth him with ſtelling his
  goddes. 31.~Iacob excuſeth himſelfe, not knowing that Rachel had taken
  away the Idols, 34.~and she deludeth his diligent ſearching for
  them. 36.~Then Iacob expoſtulateth with Laban for his
  vnkindnes. 43.~Finally they make a league and depart ech to his owne
  countrie.}

After he heard the wordes of Labans ſonnes ſaying: Iacob hath taken al
that was our fathers, and being enriched of his ſubſtance, is become
great: \V and perceauing alſo Labans countenance, that it was not
towards him as yeſterday and the other day, \V eſpecially our Lord
ſaying to him: Returne into the land of thy fathers, and to thy kinred,
and I wil be with thee. \V He ſent, and called Rachel and Lia into the
field, where he fed the flockes, \V and ſaid to them: I ſee your fathers
countenance that it is not towards me as yeſterday and the other day: and
the God of my father hath bene with me. \V And your ſelues knowe that I
haue ſerued your father to the vttermoſt of my power. \V Yea your father
alſo hath circumuented me, and hath changed my wages tenne times: and
yet God hath not ſuffred him to hurt me. \V If at any time he ſaid: They
of diuers colours ſhal be thy wages: al the ſheepe brought forth young
of diuers colours, but when he ſaid contrarie: Thou shalt take al the
white ones for thy wages: al the flockes brought forth white ones. \V
And God hath taken your fathers ſubſtance, and geuen it to me. \V For
after the time came of the ewes conceauing, I lifted vp mine eyes, and
ſawe in my ſleepe the males aſcending vpon the females of diuers
colours, and the ſpotted, and the ſpeckled. \V And the Angel of God ſaid
to me in ſleepe:
%%% 0121
Iacob? And I anſwered: Here I am. \V Who ſaid: Lyft vp thine eyes, and
ſee al the males aſcending vpon the females, them of diuers colours, the
ſpotted and the ſpeckled. For I haue ſeene al things that Laban hath
done to thee. \V I am the God of
\Fix{Bathel,}{Bethel,}{likely typo, fixed in other}
where thou dideſt
\SNote{Annointing of Altars, and free vowes are gratful offices to
God. See
\XRef{chap.~28.}}
annointe the ſtone, and dideſt vowe the vowe vnto me. Now therfore
ariſe, and goe out of this land, returning into the land of thy
natiuitie. \V And Rachel and Lia anſwered: Haue we any thing left in the
goods, and heritage of our fathers houſe? \V Hath he not reputed vs as
ſtrangers, and ſould vs, and eaten vp the price of vs? \V But God hath
taken our fathers riches, and deliuered them to vs, and to our children:
wherfore do al things, that God hath commanded thee.
%%% o-0113
\V Iacob therfore roſe vp, and ſetting his children, and wiues vpon
camels, went his way. \V And he tooke al his ſubſtance, and flockes, and
whatſoeuer he had gotten in Meſopotamia, and went forward to Iſaac his
father into the land of Chanaan. \V At that time Laban was gone to
sheare his sheepe, and Rachel ſtole the
\LNote{Idols.}{Images
\MNote{Images of falſe goddes are idols.}
of falſe goddes (as theſe were) are moſt properly called idols. And ſo
the hebrew word \HH{Teraphim} is here rightly tranſlated idols, which in
other places ſignifieth other things.
\CNote{1.~Reg.~19.}
\MNote{Some images are neither religious nor ſuperſticious.}
As the ſtatua which Michol put in Dauids bed, couering the head therof
with a hearie goates ſkinne, to deceiue Saules ſeriants who ſought
Dauids death, is called \HH{Teraphim}, and may there be tranſlated a
ſtatua, image, or ſimilitude, but not an idol. Againe,
\CNote{Oſee.~3.}
Oſee the Prophet
fortelling the lamentable ſtate of the Iſraelites, ſayeth, they ſhal be
long without King, prince, ſacrifice, altar, ephod, and \HH{Teraphim},
which laſt word in the Proteſtants Engliſh Bibles remaineth
vntranſlated.
\MNote{Some are religious.}
VVhere if they had tranſlated \Emph{Images} (as here they doe) it would
proue, that ſome images pertaine to true religion, the want wherof is
lamented among other principal things.

Theſe
\MNote{Rachel tooke away her fathers Idols, for his good.}
idols Rachel ſtole from her father, to withdraw him from idolatrie, as
S.~Baſil
\Cite{(in lib. Prouerb.)}
S.~Gregorie Nazianzen
\Cite{(orat. de Paſchate)}
and Theodoret
\Cite{(q.~89. in Gen.)}
expound it. And in this, ſaith Theodoret, ſhe was a right figure of the
Catholique Church, which depriueth idolaters of their idols. It is
probable alſo by her baſe vſing of them, that ſhe held them not for
goddes, when ſhe put them vnder the camels litter, and ſate vpon
them. Finally that ſhe reſerued them, and did not caſt them away, nor
burne, nor burie them, argueth that they were perhaps of precious mettal,
or other matter, which ſhe might turne to profite: and that lawfully in
part of
\MNote{She kept them in recompẽce of wrongs.}
recompence, that ſhe and her ſiſter, had no other dowrie, but
rather were ſold to Iacob. VVho alſo had ſuffered much iniurie at their
fathers handes.}
\TNote{\HH{Teraphim.}}
idols of her father. \V And Iacob would not confeſſe to his father in
lawe that he fled. \V And when he was gone aſwel him ſelfe as al things
that were his right, and hauing paſſed the riuer was marching on to
Mount Galaad, \V it was told Laban the third day that Iacob fled. \V
Who, taking his brethren vnto him, purſued him ſeuen dayes: and he
ouertoke him in the Mount Galaad. \V And he ſaw in his ſleepe God ſaying
vnto him: Take hede thou ſpeake not roughly anie thing againſt Iacob. \V
And Iacob had now pitched his tent in the mountaine: and when he with
his brethren had ouertaken him, he pitched his tent in the ſame Mount
Galaad. \V And he ſaid to Iacob: Why dideſt thou ſo, that vnwitting to
me thou wouldeſt carie away my daughters as captiues with the ſword? \V
Why wouldeſt thou flee without my knowledge, and not tel me, that I
might haue brought thee on the way with ioy, and ſongues, and timbrels,
and cithernes? \V Thou haſt not ſuffred me to kiſſe my ſonnes and
daughters: thou haſt donne fooliſhly: now alſo in dede, \V my hand is
able to requite thee euil: but the God of your father ſaid vnto me
yeſterday: Take hede thou ſpeake not any thing againſt Iacob roughly. \V
Suppoſe, thou diddeſt deſire to goe to thy freinds, and hadeſt a longing
to thy fathers houſe: why dideſt thou ſteale
%%% 0122
my
\Fix{goods?}{goddes?}{obvious typo, fixed in other}
\V Iacob anſwered: In that I departed vnwitting to thee, I feared leſt thou
wouldeſt take away thy daughters by force. \V But wheras thou chargeſt
me with theft: with whom ſoeuer thou ſhalt find thy goddes, let him be
ſlaine before our brethren. Search, what ſoeuer of thy things thou ſhalt
finde with me, and take away. Saying this, he knew not that Rachel had
ſtollen the idols. \V Laban therfore hauing gone into the tent of Iacob,
and of Lia, and of both the hand-maides, found them not. And when he was
entred into Rachels tent, \V ſhe in haſt hid the idols vnder the camels
litter, and ſatte therupon: and when he had ſought al the tent, and
found nothing, \V she ſaid: Let not my lord be angrie that I can not
riſe vp before thee, becauſe according to the cuſtome of wemen it is now
chanced to me. So his carefulnes in ſeeking was deluded. \V And Iacob
being
\SNote{Iacob in this iuſt expoſtulation was angrie & ſinned not.
\XRef{Pſal.~4.}}
angrie ſaid in chiding maner: For what fault of myne, and for what
offence of my part haſt thou ſo chaſſed after me, \V and ſearched al my
houshould ſtuffe? What haſt thou found of al the ſubſtance of thy houſe?
lay it here before my brethren, and thy brethren, and let them iudge
betwen me & thee. \V Haue I therfore bene with thee twentie yeares? thy
ewes and goates were not barren, the wethers of thy flocke I did not
eate: \V neyther that which the beaſt had caught did I shew to thee, I
made good al the damage:
%%% o-0114
whatſoeuer periſhed by theft, thou dideſt exact it of me: \V day and
night was I parched with heate, and with froſt, and ſleepe did flye from
myne eyes. \V And in this ſorte haue I ſerued thee in thy houſe twentie
yeares, fourtene for thy daughters, and ſix for thy flockes: thou haſt
changed alſo my wages tenne times. \V Vnles the God of my father
Abraham, and the feare of Iſaac had holpe me, peraduenture now thou
haddeſt ſent me away naked: God beheld my affliction and the laboure of
my hands, and rebuked thee yeſterday. \V Laban anſwered him: The
daughters are mine and the children, and thy flockes, and al things that
thou ſeeſt are mine: what can I do to my daughters, and nephews? \V Come
therfore, let vs enter in league: that it may be for a teſtimonie betwen
me and thee. \V Iacob therfore tooke a ſtone, and erected it for a
title: \V and he ſaid to his brethren: Bring hither ſtones. Who
gethering them together made a heape, and they did eate vpon it: \V
which
%%% 0123
Laban called The witneſſe heape: and Iacob called The hillock of
teſtimonie, either of them according to the proprietie of his
language. \V And Laban ſaid: This heape ſhal be a witnes betwen me and
thee this day, and therfore the name therof was called Galaad, that is,
The witnes heape. \V Our Lord behold and iudge betwen vs when we ſhal be
departed one from the other, \V if thou ſhalt afflict my daughters, and
if thou bring in other wiues ouer them: none is witnes of our talke but
God, who is preſent and beholdeth. \V And he ſaid againe to Iacob:
Behold this heape, and the ſtone which I haue erected betwen me and
thee, \V ſhal be a witnes: this heape, I ſay, and the ſtone be they for
a teſtimonie, if either I ſhal paſſe beyond it, going towards thee, or
thou ſhalt paſſe beyond it, thinking harme to me. \V The God of Abraham, and
the God of Nachor iudge betwen vs, the God of their father. Iacob
therfore ſware by the feare of his father Iſaac: \V and after he had
offred victimes in the mountaine, he called his brethren to eate
bread. Who when they had eaten, lodged there: \V but Laban ariſing in
the night, kiſſed his ſonnes, and daughters, and bleſſed them: and
returned vnto his place.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
