%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 48
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-48


\project douay-rheims


%%% 0164
%%% o-0151
\startChapter[
  title={Chapter 48}
  ]

\Summary{Ioſeph viſiteth his father being ſick. 5.~Who adopteth his two
  ſonnes Manaſses and Ephraim, 13.~and bleſſeth them, preferring the
  younger before the elder, contrarie to Ioſephs mind. 22.~And geueth a
  portion of land to Ioſeph, aboue his brethren.}

Theſe things being ſo done, it was told Ioſeph that his father was
ſicke: who, taking his two ſonnes Manaſſes and Ephraim, went forward. \V
And it was told the old man: Behold thy ſonne Ioſeph cometh to thee. Who
being ſtrengthned ſate on his bed. \V And Ioſeph being entred in to him,
he ſaid: God almightie appeared to me in Luza, which is in the Land of
Chanaan: and He bleſſed me, \V and ſayd: I wil increaſe, and multiplie
thee, and make thee into multitudes
%%% 0165
of peoples: and I wil geue thee this land, and to thy ſede after thee
for an euerlaſting poſſeſſion. \V Thy two ſonnes therfore, which were
borne to thee in the Land of Ægypt before I came hither to thee, ſhal be
myne: Ephraim and Manaſſes, as Ruben and Simeon ſhal be reputed to
me. \V But the reſt begotten of thee after them, ſhal be thyne, and ſhal
be called by the name of their brethren in their poſſeſſions. \V For
vnto me, when I came out of Meſopotamia, Rachel dyed in the land of
Chanaan in the very iourney, and it was ſpring time:
\CNote{Mat.~2.}
and I entered into
Ephrata, and buried her by the way ſide to Ephrata, which by an other
name is called Bethleem. \V And ſeeing his ſonnes he ſaid to him: Who
are theiſe? \V He anſwered: They are my ſonnes, whom God hath geuen me
in this place. Bring them, quoth he, to me, that I may bleſſe them. \V
For Iſraels eyes were
%%% o-0152
dymme by reaſon of very great age, and he could not ſee clearly. And
when they were ſet beſide him, kiſſing and embracing them, \V he ſaid to
his ſonne: I am not defrauded of thy ſight: moreouer God hath ſhewed me
thy ſede. \V And when Ioſeph had taken them from his fathers lappe, he
adored proſtrate vnto the ground. \V And he ſet Ephraim on his right
hand, that is, on the left hand of Iſrael: but Manaſſes on his owne left
hand, to wit, on his fathers right hand, and put them nere to him. \V
Who
\LNote{Streatching forth his right hand.}{As
\CNote{Ariſt. li.~2. de cælo. textu.~8. & de inceſ. animal. ca.~4.}
\MNote{The right hãd alſo in ſpiritual things, preferred before the
left.}
nature hath made the right hand readier to moue, ſtronger to worke and
reſiſt, and apter to frame and faſhion anie thing, ſo generally we vſe
it more then the left. And when we vſe both handes at once, we
ordinarily applie the right hand to the greater, and more excellent
effect, both in ſpiritual and corporal things. As in confirmation of
fidelitie or freindſhip, in bleſſing, writing, fieghting, playing, and
in moſt other things, we vſe the right hand, either only or chiefly. So
the Patriarch Iacob laide his right hand vpon Ephraim, knowing by
prophetical ſpirite, that he ſhould be preferred, before his elder
brother, Manaſſes. Literally fulfilled in Ioſue, Ieroboam, and other
chief Princes of Ephraims iſſue.
\MNote{The younger brother preferred ſignified the Gentiles before the
Iewes.}
And myſtically in the Gentils, being later called of God, and yet
preferred before the Iewes.
\Cite{S.~Cypri. li.~1. c.~21. aduer. Iudeos}:
\Cite{S.~Amb. li. de Benedict. Patriarch. c.~1.}
\Cite{S.~Aug. li.~16. c.~42. de ciuit.}
&c.}
ſtretching forth his right hand, put it vpon the head of Ephraim the
younger brother: and
\SNote{By this he made a croſſe prefiguring the Croſſe of Chriſt.
\Cite{Iſidor in hunc locum.}}
the left vpon the head of Manaſſes, that was the elder,
\LNote{Changing handes.}{The myſterie, of the Gentils excelling the
Iewes in time of grace,
\CNote{Procop. Iſidor. in Gen.}
often prefigured by preferring the younger
brother before the elder (Abel before Cain; Abraham before Nachor; Iſaac
before Iſmael; Iacob himſelf before Eſau; and now Ephraim before
Manaſſes) is here further repreſented by
\MNote{The forme of the Croſſe prefigured by Iacob croſſing his armes.}
Iacobs forming of a croſſe,
with his armes laied one ouer the other, when he bleſſed his two
nephewes: who otherwiſe might haue laied his right hand firſt vpon one,
and then vpon the other, or haue cauſed them to change places; but he
wittingly croſſed his armes, and \Emph{changed his handes}; or according
to the Hebrew, \Emph{made his handes vnderſtand}, that is, by his handes made it
to be vnderſtood, not only that the younger ſhould be in place of the
elder, Ephraim before Manaſſes, and much more the Gentiles before the
Iewes, but alſo that this greater Myſterie ſhould be effected by Chriſt
dying on a Croſſe. For what els could the verie croſſing of his armes,
ſo wittingly and purpoſly done, ſignifie, but the forme and figure of
Chriſts Croſſe? As els where the wood, which young Iſaac caried on his
back vnto the mountaine prefigured the matter or ſubſtance of the ſame
Croſſe. Al accompliſhed when Chriſt was crucified: wherby the
Iewes were ſcandalized, and the Gentiles called and ſaued. Our Sauiour
himſelfe fortelling, that
\CNote{Ioan.~12.}
\Emph{he being exalted} (to wit vpon the
Croſſe) \Emph{vvould dravv al vnto himſelf.} And
\CNote{Colloſ.~2.}
S.~Paul teaching that
Chriſt \Emph{faſtned the hand vvriting, that vvas againſt vs, vpon the
Croſſe.}}
changing handes. \V And Iacob bleſſed the ſonnes of Ioſeph, and ſaid:
God, in whoſe ſight my fathers Abraham and Iſaac haue walked, God that
feedeth me from my youth vntil this preſent day: \V
\LNote{The Angel that deliuereth me.}{It
\MNote{Protection & Inuocation of Angels.}
is euident by this plaine text, that Iacob was deliuered from euiles by
an Angel, and that he inuocated the ſame Angel to bleſſe his nephewes,
S.~Baſil
\Cite{(li.~3. cont. Eunom. in initio.)}
ſheweth by this place amongſt others, that an Angel is preſent with
euerie one, \Emph{as a pedagogue, and paſtour, directing his
life}. S.~Chryſoſtom alſo
\Cite{(ho.~7. in laud. S.~Pauli)}
citeth this place in teſtimonie, that proper Angels are deputed to
protect men. Yet Proteſtants ſay, that
\CNote{Bible.~1603.}
\Emph{this Angel muſt be
vnderſtood of Chriſt}: remitting their gloſſe to the
\XRef{31.~ch. v.~13. and 32.~ch. v.~1. of Geneſis},
where it can not be proued.
\MNote{Proued by ancient Fathers.}
But the ancient Fathers teach the patronage & Inuocation of Angels
grounded in holie Scripture. Namely in this place, and manie other
places in the old Teſtament. Alſo
\XRef{Mat.~18.}
\XRef{Act.~12.}
\XRef{2.~Cor.~11.}
& the like. For example, S.~Iuſtinus Martyr in
\Cite{explic. qq. neceſſ. q.~30.}
affirming it for a knowen truth, declareth that thoſe Angels, which
receiue the charge of guarding men, cõtinew the ſame office either to
both ſoule and bodie, or to the ſoule after it is parted from the
bodie. S.~Cyril of Alexandria
\Cite{(lib.~4. cont. Iulian. prope. init.)}
ſhewing how God vſeth the miniſterie of holie Angels, for mens
ſaluation, ſaith: \L{Hi noxias a nobis abigunt feras &c.} Theſe (Angels)
driue away noyſome wildbeaſts from vs: and reſkew thoſe that are caught,
from their crueltie, and teach what ſoeuer is laudable, to make our
paſſage free, and not peſtered, when with vs they glorifie one
ſoueraigne God. S.~Chryſoſtom
\Cite{(ho.~60. in Math.~18.)}
S.~Hierome
\Cite{vpon the ſame place},
S.~Ambroſe, in
\XRef{Pſal.~38.}
S.~Auguſtin
\Cite{li.~83. qq. q.~79.}
&
\Cite{li. Soliloq. c.~27.}
S.~Gregorie
\Cite{li.~4. c.~31. in 3.~Iob.}
S.~Bernard
\Cite{ſer.~5. Dedicat. Eccleſ.}
&
\Cite{ſer.~12. in Pſal.~90.}
& others, ſo vniuerſally teach the ſame, that Caluin
\Cite{(li. Inſtit. c.~14. ſect.~38.)}
dare not denie it, and yet wil nedes doubt of it.}
The Angel that deliuereth me from al euils, bleſſe theſe children: and
%%% !!! Unmarked note.
\LNote{Be my name called vpon them.}{This
\MNote{God for his Saints ſake ſheweth fauour to their frends.}
place hath two good literal ſenſes. For firſt it importeth, that Ephraim
and Manaſſes were made participant among the Tribes, of the bleſſings of
Abraham, Iſaac, and Iacob. Secondly that God would bleſſe them, for
Abraham, Iſaac, and Iacobs ſake: ſo Moyſes praying for the whole people
\XRef{(Exodi.~32.)}
beſought God to remember, Abraham, Iſaac, and Iſrael, and God was
therwith pacified.}
be my name called vpon them, the names alſo of my fathers Abraham, and
Iſaac, and grow they into a multitude vpon the earth. \V And Ioſeph
ſeing that his father had put his right hand vpon the head of Ephraim,
tooke it heauily: and taking his fathers hand he went about to lift it
from Ephraims head, and to remoue it vpon the head of Manaſſes. \V And
he ſaid to his father: It is not conuenient father ſo to be: becauſe
this is the firſt begotten put thy right hand vpon his head. \V Who
refuſing, ſaid: I know my ſonne, I know: and this ſame in dede ſhal be
into peoples, and ſhal be multiplied:
%%% 0166
but his younger brother ſhal be greater then he: and his ſede ſhal growe
into nations. \V And he bleſſed them at that time, ſaying: In thee ſhal
Iſrael be bleſſed, and it ſhal be ſaid: God do vnto thee as vnto
Ephraim, and as vnto Manaſſes. And he ſette Ephraim before Manaſſes. \V And
he ſaid to Ioſeph his ſonne: Behold I dye, and God wil be with you, and
wil bring you backe into the land of your fathers. \V I doe geue thee
one portion aboue thy brethren, which I tooke out of the hand of the
Amorrhean with my ſword and bowe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
