%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


%%% Latin checked by KK.



\startcomponent chapter-11


\project douay-rheims


%%% 0063
%%% o-0059
\startChapter[
  title={Chapter 11}
  ]

\Summary{God
\MNote{The fourth part of this booke. Of the diuiſion of tõgues and
  nations.}
hindereth the vaine purpoſe of building a hiegh towre,
  7.~by confounding mens tongues. 9.~Wherof it is called Babel. 10.~The
  genealogie of Sem to Abram.}

And the earth was of one tongue, and al one ſpeach. \V And when they
remoued from the eaſt, they found a plaine in the land of Sennaar, and
dwelt in it. \V And eech one ſaid to his neighboure: Come, let vs make
bricke, and bake them with fire. And they had bricke in ſteed of ſtone,
and bitume in ſteed of morter: \V and they ſaid: Come,
\LNote{Let vs make.}{Here
\MNote{Pride is cauſe of ſchiſme and hereſie.}
we may ſee in Nemrod the common cauſes of hereſies, and the maner of
Heretikes proceeding. For he hauing a ſutle proud, and aſpiring mind,
firſt detracted from God,
\CNote{S.~Chriſ. ho.~30. in Gen. Caſſian. collat.~4. c.~12.}
perſwading men (as is noted before) not to
depend vpon Gods prouidence, and finding ſome others of like humour,
they conſpired together, and drew more folowers,
\MNote{Falſe pretences deceiue the ſimple.}
by bearing the ſimpler
ſorte in hand (for it was vnpoſſible wiſe men ſhould beleue it) that
they would make a towre of defence againſt a new floud, if God ſhould
thinck to drowne the world againe. But their principal intention was to
make themſelues great and ſtrong for the preſent, and famous to
poſteritie.
\MNote{Heretikes proſper for a time, but are confounded in the end.}
Al which God ouerthrew neither ſuffering them to build vp their imagined
caſtle of ſtrength nor to be praiſed for their worke, but made them
infamous to the worlds end.}
let vs make vs a citie and a towre, the toppe wherof may reach to
heauen: and let vs renowne our name before we be diſperſed into al
lands. \V And our Lord deſcended to ſee the citie and the towre, which
the children of Adam builded, \V and he ſaid: Behold, it is one people,
and one tongue is to al: and
%%% 0064
they haue begunne to doe this, neyther wil they leaue of from their
determinations, til they accomplish them indede. \V Come ye therfore,
\LNote{Let vs confound.}{God
\MNote{Miniſtrie of Angels.}
in diſſipating this vaine worke of men, would vſe the miniſterie of
Angels. As not only
\CNote{Li. de cõfuſ. ling.}
Philo Iudeus, and
\CNote{ho.~11. in Numer.}
Origen, but alſo
\CNote{li.~16. c.~6. ciuit.}
S.~Auguſtin,
\CNote{li.~2. Moral. in c.~1. Iob.}
S.~Gregorie, and other fathers expound theſe wordes, \Emph{Come, let vs
deſcend, and confound their tongue.}
\MNote{God turneth the offence of men to good.}
Where they alſo note Gods ſingular wiſdome, mercie, and iuſtice, fo
puniſhing the offence, that he turneth it to his owne glorie, and
the profite of al men: ſhewing his powre and ſoueraigne Maieſtie by
\MNote{Two miracles in priuation of one tongue and geuing a new.}
two great miracles. Firſt by ſo ſuddainly and vtterly depriuing al thoſe
builders, of their vſual tongue, that preſently they could neither
ſpeake it, nor vnderſtand it. Secondly, by geuing diuers diſtinct
languages to ſeueral ſortes or families, which they immediatly
vnderſtood, and ſpoke moſt promptly, as if they had long before lerned
and vſed the ſame. But to no man was geuen more then one language. And
ſo, to the more commoditie of al mankind, they were forced to part into
ſundrie coaſtes of the earth, which they inhabited and repleniſhed with
diſtinct Nations, hauing the ſame Angels their ſpiritual Patrones and
Protectors, which had ſeuerally changed their language.
\MNote{Diuiſion among euil men is profitable.}
In particular, it was profitable to the good who being before oppreſſed
by the vnited powre of manie wicked, were releeued (as
\CNote{Li.~34. Moral. in c.~41. Iob.}
S.~Gregorie
teacheth) when their perſecutors were diuided. Theſe good were the
familie of Heber, as
\CNote{ho.~11. in Gen.}
S.~Chriſoſtom, and
\CNote{l.~16. c.~11. & li.~18. c.~39. ci.}
S.~Auguſtin proue.
\MNote{The member offending is puniſhed.}
For ſeing the change of tongues was inflicted for puniſhment,
\MNote{Heber and his familie cõſented not to the building of Babel.}
it appeareth that Heber and his familie were innocent of the vaine
attempt, whoſe tongue was not changed, but remained the ſame, and of him
was called the Hebrew tongue for diſtinction ſake after there were
manie tongues, which before had no diſtinct name being the only tongue
of al men.
\CNote{S.~Aug. li.~6. c.~4. ciuit.}
Againe touching the offenders (who were puniſhed in their
tongues, that they could not be vnderſtood commanding one an other,
becauſe they would not vnderſtand God iuſtly commanding them al) they
alſo reaped this profite, that they were forced to leaue of that bad
worke, and withal to ſeeke more ample habitations, who \Emph{If they had
there more increaſed in number and ſtreingth, vvould vvithout doubt}
(ſaith
\CNote{Ho.~30. in Gen.}
S.~Chriſoſtom) \Emph{haue attempted vvorſe things.} And infinite
manſlaughter would haue benne committed, amongſt ſo manie for poſſeſſion
of that one citie & towre. Finally the
\CNote{S.~Greg. ho.~30. in Euan.}
Fathers note that as God wrought
here much good by diuiſion of tongues:
\MNote{Gift of tõgues moſt profitable to the Church.}
ſo he wrought much more by communion of tongues, geuen to the Apoſtles,
therby inabling them to gether one Church of al Tongues and Nations.}
let vs goe downe, and there confound their tongue, that none may heare
\Fix{is}{his}{obvious typo, fixed in other}
neighbours voice. \V And ſo our Lord diſperſed them from that place into
al lands, and they ceaſed to build the citie. \V And therfore the name
therof was called Babel,
\SNote{He that ſpeaketh ſo confuſedly that he is not vnderſtood is ſaid
to bable.}
becauſe there the tongue of the whole earth was confounded: and from
thence our Lord diſperſed them vpon the face of al countries.

\V Theſe are
\SNote{Moyſes here ſheweth the ſucceſſion of Patriarches, from Sem to
Abraham, as he did before from Adam to Noe.
\Cite{S.~Aug. li.~16. c.~10. ci.}}
the generations of Sem: Sem was an hundred yeares old when he begat
Arphaxad, two yeares after the floud. \V And Sem liued after he begat
Arphaxad, fiue hundred yeares: and begat ſonnes and daughters. \V
Moreouer Arphaxad liued thirtie fiue yeares, and
\LNote{Begat Sale.}{Here
\MNote{Scriptures hard.}
is an intricate difficultie. For the Hebrew and Latin text, both here and
in
\CNote{1.~Par.~1.}
\Fix{Palalippomenon,}{Paralippomenon,}{obvious typo, fixed in other}
ſaying Arphaxad begat Sale, the 72.~Interpreters and
\CNote{Luc.~3. v.~36.}
S.~Luke place Cainan betwen them, as ſonne of Arphaxad, and father of
Sale. Euſebius alſo in his Chronicle, with moſt Greeke Doctors, and
\CNote{Li.~16. c.~10. ciuit.}
S.~Auguſtin, count Cainan in this Genealogie of Sem.
\MNote{Some think Moyſes omitted Cainan for a myſterie.}
VVherupon manie do number him in this ranck, and ſuppoſe that Moyſes
omitted him for ſome Myſterie, and yet writeth truly, that Arphaxad
begat Sale, not his proper ſonne, but his ſonnes ſonne: as
\CNote{Mat.~1.}
S.~Mathew ſayth, Ioram begat Ozias, who was his nephewes nephew. 
\MNote{Refutation.}
But againſt this
ſolution it is replied, that then Arphaxad ſhould haue bene a
grandfather at 35.~yeares of age: which were ſtrange in thoſe daies, how
ſoeuer it is now. And a greater difficultie, or rather abſurditie muſt
alſo be granted, that Arphaxad begat both Cainan at the age of
35.~yeares, according to the 72.~Interpreters, and that Sale was alſo
begotten the ſame yeare, according to the Hebrew, being both
true. VVhich inconuenience is not in the Genealogie written by
S.~Matthew.
\MNote{Others coniecture Cainan ſhould not be in the text of the 70.}
Others therfore according to the Hebrew and Latin text, with moſt Latin
Doctors, omit Cainan in this place, and Paralipomenon, namely with
\CNote{Queſt. Hebræ.}
S.~Hierom, who diligently examining and reconciling varieties, betwen
the Hebrew and the Greeke, maketh no mention at al of this
difference. VVhich maketh ſome to coniecture, that in S.~Hieroms time
Cainan was not in the Greeke copies, at leaſt not in thoſe that he had,
and held for the beſt. And at this day
\CNote{Editio quaedam vaticanæ.}
ſome haue him not. VVhich may be
admitted for a probable anſwere touchhing the Hebrew and Greeke of the old
Teſtament. 
\MNote{But neuer anie Catholique (nor heretike before Beza) put Cainan
out of S.~Lukes Goſpel.}
But for ſo much as al copies, both Greeke & Latin, alſo
S.~Hieroms Edition of S.~Lukes Goſpel haue Cainan, the difficultie ſtil
remaineth betwen Moyſes and S.~Luke. How then ſhal this doubt be ſolued?
we can not ſolue it. And no maruel. For
\CNote{Præfat. cõment. in Acta Apoſt.}
venerable Bede could not. VVhoſe
wordes are theſe:
\MNote{A memorable ſentence of S.~Beda.}
S.~Luke vſeth rather the Greeke teſtimonies then the
Hebrew: wherof hapeneth that I much maruel at, and for dulnes of wit,
being ſtriken with great
\Fix{admirrtion,}{admiration,}{obvious typo, fixed in other}
I can not throughly ſcan, ſeeing in the Hebrew veritie are founed only
tenne generations from the floud vnto Abraham, by what meanes S.~Luke,
who (the Holie Ghoſt gouerning his penne) could in no ſorte write falſe,
would rather ſet downe eleuen generations in the Goſpel, Cainan adioyned
according to the ſeuentie Interpreters. Thus writeth S.~Beda, reuerently
admiting that he could not vnderſtand. For being aſſured that the Holie
Ghoſt gouerned the pennes both of Moyſes and S.~Luke, and that he is not
contrarie to him ſelfe, it muſt nedes be true which ech of them writeth,
though other learned men can not reach the profoundnes of ſome
difficulties that occure.
\MNote{Beza ſacrilegiouſly proud.}
And therfore Beza was extreme ſaucie to daſh Cainan out of S.~Lukes
Goſpel, and that wittingly and moſt impudently ſaying, in his
Annotations: \L{Non dubitauimus expungere}: \Emph{VVe doubted not to put
it out.}
\MNote{The heretical Engliſh Editions differ in this point.}
The former Engliſh Editions, otherwiſe corrupt in manie places, haue
Cainan in the text of S.~Lukes Goſpel, but their latter tranſlaters are
in this point pure Bezites.}
begat Sale. \V And Arphaxad liued after he begat Sale, three hundred
three yeares: and begat ſonnes and daughters. \V Sale alſo liued thirtie
yeares, and begat Heber. \V And Sale liued after he begat Heber, foure
hundred three yeares: and begat ſonnes and daughters. \V And Heber liued
thirtie foure yeares, and begat Phaleg. \V And Heber liued after he
begat Phaleg, foure hundred thirtie yeares: and begat ſonnes and
daughters. \V Phaleg alſo liued thirtie yeares, and begat Reu. \V And
Phaleg liued after he begat Reu, two hundred nine yeares, and begat
ſonnes and
%%% o-0060
daughters. \V And Reu liued thirtie two yeares, and begat
Sarug. \V Reu liued alſo after he begat Sarug, two hundred ſeauen
yeares: and begat ſonnes & daughters. \V And Sarug liued thirtie yeares,
and begat Nachor. \V And Sarug liued after he begat Nachor, two hundred
yeares: and begat ſonnes and daughters. \V And Nachor liued nine and
twentie yeares, and begat Thare. \V And Nachor liued after he begat
Thare, an hundred and nintene yeares: and begat ſonnes and daughters. \V
And Thare liued ſeauentie yeares, and begat Abram, and Nachor, and
Aran. \V And theſe are the generations of Thare: Thare begat Abram, Nachor,
& Aran. Moreouer Aran begat Lot. \V And Aran died before Thare his
father, in the land of his natiuitie in
\SNote{Vr a citie, or territorie of Chaldea.
\Cite{70.~Interpret.}
and
\Cite{Ioſephus li.~1. Antiq.}}
Vr of the Chaldees. \V And Abram & Nachor maried wiues: the name of
Abram his wife was Sarai: and the name of Nachor his wife, Melcha the
daughter of Aran the father of Melcha, and the father of Ieſcha. \V And
Sarai was barren, neither had ſhe
%%% 0065
children. \V Thare therfore
\SNote{Abram was commanded to goe forth of Chaldea, as appeareth
\XRef{Act.~7. v.~4.}
Though this iourney is here aſcribed to Thare as the principal perſon.}
tooke Abram his ſonne, and Lot the ſonne of Aran, his ſonnes ſonne, &
Sarai his daughter in law, the wife of Abram his ſonne, and brought them
out of Vr of the Chaldees, for to goe into the land of Chanaan: and they
came as farre as Haran, and dwelled there. \V And the dayes of Thare
came to two hundred fiue yeares, and died in Haran.



%%% !!! This formatting could be improved.
%%% In particular, it should follow the annotations.
\vskip 0.25in

\startalignment[middle]
  The end of the ſecond age.
\stopalignment


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
