%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 23
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-23


\project douay-rheims


%%% 0096
%%% o-0090
\startChapter[
  title={Chapter 23}
  ]

\Summary{Sara dying Abraham ſolemnizeth her funeral, 4.~byeth a field
  with a duble caue of Ephron, 15.~for hour hundreth ſicles, 19.~and
  there burieth her.}

And Sara liued an hundred twentie ſeuen yeares. \V And ſhe died in the
citie of Arbee which is
%%% 0097
Hebron, in the land of Chanaan: And Abraham came to mourne, and weepe
for her. \V And after that he was riſen vp from
\SNote{A cleare example of religious office in burying the dead, See
\XRef{2.~Reg.~1.}
and
\XRef{2.~Paral.~35.}}
the funeral obſequies, he ſpake to the children of Heth, ſaying: \V I am a
ſtranger and pilgrime among you: geue
%%% o-0091
me the right of a ſepulchre with you, that I may burie my dead. \V The
children of Heth anſwered, ſaying: \V My lord heare vs, thou art a
prince of God among vs: in our principal ſepulchres burie thou thy dead:
and no man can let thee but that in his owne monument thou mayeſt burie
thy dead. \V Abraham roſe vp, and
\SNote{Adoration vſed for reuerence done to men. See alſo
\XRef{c.~27. v.~29.}
\XRef{c.~33. v.~3.~7.}
and
\Cite{S.~Aug. q.~61. in Gen.}}
adored the people of the land, to wit the children of Heth: \V and he
ſaid to them: If it pleaſe your ſoule that I burie my dead, heare me,
and be interceſſors for me to Ephron the ſonne of Seor: \V that he geue
me the duble caue, which he hath in the vttermoſt part of his field: for
money to the worth therof let him deliuer it to me before you for
poſſeſſion of a ſepulchre. \V And Ephron dwelt in the middeſt of the
children of Heth. And Ephron made anſwer to Abraham in the hearing of al
that went in at the gate of the citie, ſaying: \V No, it ſhal not be ſo,
my lord, but thou rather harken to that which I doe ſay: The field I
deliuer to thee, and the caue that is therin, in the preſence of the
children of my people, burie thy dead. \V Abraham adored before the
people of the land. \V And he ſpoke to Ephron, his people ſtanding round
about: I beſeech thee to heare me: I wil geue money for the field: take
it, and ſo I wil burie my dead in it. \V And Ephron anſwered: \V My
lord, heare me. The ground which thou deſireſt, is worth foure hundred
ſicles of ſiluer: this is the price betwen me and thee: but how much is
this? burie thy dead. \V Which when Abraham had heard, he weyed the
money, that Ephron had asked, in the hearing of the children of Heth,
foure hundred ſicles of ſiluer of common currant money. \V And the field
that before time was Ephrons, wherin was the duble caue, looking towards
Mambre, aſwel it, as the caue, and al the trees therof in al the lymits
therof round about: \V was made ſure to Abraham for a poſſeſſion, in the
ſight of the children of Heth, and of al that went in at the gate of his
citie. \V And ſo Abraham buried Sara his wife in the duble caue of the
field, that looked towards Mambre, this is Hebron in the land of
Chanaan.
%%% 0098
\V And the field was made ſure to Abraham, and the caue, that was in it,
for a poſſeſſion to burie in, of the Children of Heth.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
