%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 0072
%%% o-0068
\startChapter[
  title={Chapter 13}
  ]

\Summary{Abram & Lot returne from Ægypt into Chanaan, 6.~and being rich
  ſeparate themſelues, 10.~Lot chooſing the countrie about Iordain,
  Abram dwelleth in Chanaan. 14.~Where againe God promiſeth him that
  land, and multiplication of his ſeede. 18.~And he erected an other
  Altar to God.}

Abram therfore aſcended out of Ægypt, he and his wife, and al that he
had, and Lot with him to the ſouth
%%% 0073
coaſt. \V And he was very rich in poſſeſſion of gold and ſiluer. \V And
he returned by the way, that he came, from the ſouth vnto Bethel, euen
to the place where before he had pitched a tabernacle betwen Bethel and
Hay: \V in the place of the altar which he had made before, and there he
called vpon the name of our Lord.

%%% o-0069
\V But Lot alſo that was with Abram, had flocks of ſheepe, and heards of
beaſts, and tents. \V Neyther was the land able to receiue them, for to
dwel togeather: for their ſubſtance was much, and they could not dwel
togeather. \V Wherupon alſo there aroſe ſtrife amongſt the heardſmen
of Abram and of Lot. And that time the Chananite and the Pheriſite
dwelled in that countrie. \V Abram therfore ſaid to Lot: Let there be no
brawle I beſeech thee betwen me and thee, and betwen my heardſmen, and
thy heardſmen: for we be
\SNote{Foure ſortes of bretheren in holie writte: german brothers, as
Iacob and Eſau: of kinred, as Abraham and Lot: of the ſame nation, as
the Iewes and Samaritanes: in Religion as al Catholiques.}
brethren. \V Behold the whole land is before thee: goe aparte from me, I
pray thee: if thou wilt goe to the left hand, I wil take the right: if
thou chooſe the right hand, I wil paſſe to the left.

\V Lot therfore lifting vp his eyes, ſawe al the countrie about
Iordaine, which was watered through out before that our Lord ſubuerted Sodome
and Gomorre, as the paradiſe of our Lord, and like as Ægypt as men come
vnto Segor. \V And Lot choſe vnto him the countrie about Iordaine, and
he departed from the Eaſt: and they were ſeperated either brother from
the other. \V Abram dwelt in the land of Chanaan: and Lot abode in the
townes, that were about Iordaine, and dwelt in Sodome. \V And the men of
Sodome were verie wicked, and ſinners before the face of our Lord out of
meaſure. \V And our Lord ſaid to Abram, after that Lot was ſeperated
from him: Lyft vp thyn eyes, and looke from the place, wherin thou now
art, to the north and ſouth, to the eaſt and weſt. \V Al the land, which
thou ſeeſt, wil I geue to thee, & to thy ſeed for euer. \V And I wil
make
\SNote{Not the children of the fleſh but the children of the promiſſe
are the ſeede.
\XRef{Rom.~9.}
and are innumerable.
\XRef{Apoc.~7. v.~9.}}
thy ſeede as the duſt of the earth: if any man be able to number the
duſt of the earth, thy ſeede alſo ſhal he be able to number. \V Ariſe
and walke through the land in the length, and in the breadth therof: for
I wil geue it to thee. \V Abram therfore remouing his tent, came, and
dwelt beſide the vale of Mambre, which is in Hebron: and he builded
there an altar to our Lord.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
