%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 35
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-35


\project douay-rheims


%%% 0130
%%% o-0121
\startChapter[
  title={Chapter 35}
  ]

\Summary{Iacob purging his whole familie of idols, goeth by Gods
  commandment into Bethel. 7.~There buildeth an Altar. 8.~Debora
  dieth. 9.~God appearing againe to Iacob bleſſeth him, and changeth his
  name into Iſrael. 16.~Rachel bearing Beniamin dieth, and is buried in
  Bethleem. 22.~Ruben lyeth with Bala. 23.~Iſraels twelue ſonnes are
  recited. 28.~Iſaac dieth at the age of 180.~yeares, and his ſonnes
  Eſau and Iacob burie him.}

In the meane time God ſpake to Iacob: Ariſe, and goe vp to Bethel, and
dwel there, and make an altar to God that appeared to thee when thou
diddeſt flie from Eſau thy brother. \V And Iacob hauing called together
al his houſe, ſaid:
\LNote{Caſt avvay the ſtrange goddes.}{Iacob
\MNote{Clenſing from ſinne is the firſt office of the ſeruants of God.}
preparing to performe his vow of building a houſe & altar to God, firſt
extirpateth al Idolatrie from amongſt his people; and then by Sacrifice
appeaſeth Gods wrath prouoked how ſoeuer and ſpecially by Simeon and
Leui killing the Sichemites. Duly conſidering that what people or perſon
deſireth Gods protection & bleſſings, muſt firſt be pure in Religion,
and clenſed from ſinnes: \L{Quia ei nulla nocebit aduersitas, si nulla ei
dominetur iniquitas}: becauſe no aduerſitie ſhal hurt him, if no
iniquitie haue dominion ouer him.
\Cite{orat. ſer.~6. poſt cineres.}}
Caſt away the ſtrange goddes that are among you, and be clenſed and
change your garments. \V Ariſe, and let vs goe vp into Bethel, that we
may make there an altar vnto God: who heard me in the day of my
tribulation, and
\Fix{accompained}{accompanied}{possible typo, fixed in other}
me in my iourney. \V They gaue to him
therfore al the ſtrange goddes that they had, and the earelets which
were in their eares: but he buried them vnder the
\TNote{An execrable tree.}
terebinth, that is behind the citie of Sichem. \V And when they were
departed,
\SNote{God (when it pleaſeth him) maketh the weake ſtronger then the
mightie; and few more terrible then manie.
\Cite{S.~Chriſoſt. ho.~59.}
\Cite{S.~Aug. q.~112. in Gen.}}
the terror of God inuaded al the cities rounde about, and they durſt not
purſew them going away. \V And Iacob came to Luza, which is in the land
of Chanaan, ſurnamed Bethel: he and al the people that was with him. \V
And he builded there an altar, and called the name of that place, The
houſe of God: for there God appeared to him when he fled from his
brother. \V The ſame time died Debora the nurſe of Rebecca, and was
buried at the foote of Bethel vnder an oke: and the name of that place
was called, The oke of weeping. \V And God appeared again
%%% 0131
to Iacob after he returned from Meſopotamia of Siria, and he bleſſed
him, \V ſaying: Thou
\SNote{The name of ſupplanter not ſufficiently expreſſing his valure he
is alſo called Iſrael. See the Annotation.}
ſhalt not be called any more Iacob, but Iſrael ſhal be thy name. And he
\LNote{Called him Iſrael.}{As
\MNote{The name \Sc{Israel} ſignifieth ſpecial prerogatiues in the
Patriarch Iacob.}
the Patriarch now performeth his vow to God: ſo God fulfilleth his
promiſe, geuing him a new name. For \Emph{Iacob} a \Emph{ſupplanter},
ſignifying too ſmal force & praiſe for ſuch a champion, God therfore
honored him with the name of \Emph{Iſrael}. That is, \Emph{One that
ſeeth and contemplateth God}, as moſt ancient writers expound
it. Alſo \Emph{A prince, or valient vvith God}, as S.~Hierom ſheweth
\Cite{Tradit. Heb.}
For \HH{Iſra} in Hebrew ſignifieth \Emph{To dominier, or rule ouer},
and \HH{El} ſignifieth God. And ſo this name geuen to him teſtifieth that
he, by Gods gift and grace, was valient euen againſt an Angel
repreſenting God, much more againſt men, and other aduerſaries. Others
interprete it, \Emph{The right one of God}: as witneſſeth the ſame
S.~Hierom both here, and in his
\Cite{commentaries in 44.~Iſaie.}
Al do importe a great excellencie in this Patriarch. And the ſucceſſe of
things confirmeth the ſame. Particularly in that not onlie ſome one of
his ſonnes (as in the iſſue of Abraham and Iſaac, the reſt being
excluded) but
\MNote{Al his twelue ſonnes in their poſteritie were heires of the
promiſed land.}
his whole progenie of twelue ſonnes, making twelue Tribes, were
participant of the peculiar bleſſings, in their ofspring poſſeſſing the
promiſed land, and excedingly increaſing became the moſt principal
nation in the world, the ſelected people of God, called by the name and
title \Emph{Of the children of Iſrael}. Of whom not only Moyſes, but al
the old Teſtament moſt ſpecially treateth, and of whom and in whom the
promiſed Meſſias, the Redemer of mankind ſhould be borne.}
called him Iſrael, \V and ſaid to him: I am God almightie, encreaſe thou
and multiplie: Of thee ſhal be nations and peoples of nations, kinges
ſhal come forth of thy loynes. \V And the land which I gaue to Abraham
and Iſaac, I wil geue to thee, and to thy ſeede after thee. \V And he
departed from him. \V But he erected a title of ſtone, in the place
where God had ſpoken vnto him: offering vpon it liquide offeringes, and
powring oile on it: \V and calling the name of that place, Bethel. \V
And being gone forth from thence, he came in the ſpring time to the land
which leadeth to
%%% o-0122
Ephrata: wherin when Rachel was in trauaile, \V becauſe of difficultie
in her trauaile, ſhe beganne to be in danger, and the midwife ſaid vnto
her: Feare not, for thou shalt haue alſo this a ſonne. \V And her ſoule
departing for paine, and death now at hand, she called the name of her
ſonne Benoni, that is the ſonne of my paine: but his father called him
Beniamin, that is the ſonne of the right hand. \V Rachel therfore died,
and
\CNote{Math.~2.}
was buried in the hye way that leadeth to Ephrata, this ſame is
Bethleem. \V And Iacob erected a title ouer her ſepulchre: This is the
\Fix{litle}{title}{obvious typo, fixed in other}
of Rachels monument, vntil this preſent day. \V Departing thence, he
pitched his tent beyond the Flocke tower. \V And when he dwelt in that
countrie: Ruben went, and
\SNote{For this fact Ruben was excluded from the chiefe dignitie among
his bretheren.
\XRef{Gen.~49.}}
ſlept with Bala his fathers concubine: which thing he was not ignorant
of. And the ſonnes of Iacob were twelue. \V The ſonnes of Lia: Ruben the
firſt begotten, and Simeon, and Leui, and Iudas, and Iſſachar, and
Zabulon. \V The ſonnes of Rachel: Ioſeph and Beniamin. \V The ſonnes of
Bala Rachels handmaid: Dan and Nepthali. \V The ſonnes of Zelpha Lias
handmaid: Gad and Aſer: theſe are the ſonnes of Iacob, that were borne
to him in Meſopotamia of Siria. \V He came alſo to Iſaac his father in
Mambre, the citie of Arbee, this is Hebron: wherin Abraham and Iſaac
ſoiourned. \V And the dayes of Iſaac were complete an hundred eyghtie
yeares. \V And ſpent with age he died, and was put to his people,
being old and ful of dayes: and Eſau and Iacob his ſonnes buryed him.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
