%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 24
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-24


\project douay-rheims


%%% 0098
%%% o-0092
\startChapter[
  title={Chapter 24}
  ]

\Summary{Abrahams ſeruant adiured and ſent by him into Meſopotamia, to
  ſeke a wife for Iſaac, 12.~prayeth to God for a ſigne, findeth
  Rebecca, 34.~and demanding her for this purpoſe, 50.~with her parents,
  58.~and her owne conſent, she goeth with him, 67.~is maried to Iſaac:
  who therby is comforted for the death of his mother.}

And Abraham was old, and of manie dayes: and our Lord had bleſſed him in
al things. \V And he ſaid to the elder ſeruant of his houſe, which was
ruler ouer al that he had: Put thy hand vnder my thighe, \V that I may
adiure thee by our Lord, God of heauen and earth, that thou
\SNote{In chooſing a wife a vertuous ſtock and familie eſpecially true
faith and religion, are before al other things to be conſidered &
preferred.
\Cite{S.~Amb. li.~1. c.~9. de Abrah.}
\Cite{S.~Chriſoſt. ho.~48. in Gen.}}
take not a wife for my ſonne, of the daughters of the Cananites, among
whom I dwel: \V but that thou goe vnto myne owne countrie and kindred,
and thence take a wife for my ſonne Iſaac. \V The ſeruant anſwered: If
the woman wil not come with me into this land, whether muſt I bring thy
ſonne backe againe to the place, from whence thou dideſt come forth? \V
And Abraham ſaid: Beware thou neuer bring my ſonne backe againe
thither. \V Our Lord God of heauen, which tooke me out of my fathers
houſe, and out of my natiue countrie, which ſpake to me, and ſware to
me, ſaying: To thy ſede wil I geue this land: he ſhal ſend his Angel
before thee, and thou ſhalt take from thence a wife for my ſonne: \V but
if the woman wil not folow thee, thou ſhalt not be bound by the oath:
only bring not my ſonne thither againe. \V The ſeruant therfore put his
hand vnder the thigh of Abraham his lord, and ſware to him vpon this
worde. \V And he tooke tenne camels of his lords heard, and departed, of
al his goods carying ſomething with him, and ſetting forward went on
into Meſopotamia to the citie of Nachor. \V And when he had made the
camels lye downe without the towne beſide a wel of water at euen, at the
time when wemen are wont to come forth to drawe water, he ſaid: \V O
Lord God of my lord Abraham, mete me to day,
%%% 0099
I beſeech thee, and doe mercifully with my maiſter Abraham. \V Behold I
ſtand nigh to the fountaine of water, and the daughters of the
inhabiters of this citie, wil come forth to drawe water. \V Therfore
\LNote{The maide to vvhom I shal ſay.}{Obſeruation
\MNote{Ominous ſpeaches ſometimes ſuperſticious.}
of ſpeaches called ominous, which are interpreted to ſignifie good or
euil luck, are ſometimes ſuperſticious, & ſuggeſted by euil ſpirites,
who now & then telling, or inſinuating ſome truth, get credite, and ſo
allure men to attend to ſuch vaine, vncertaine, and vnlawful ſignes, as
S.~Auguſtin teſtifieth
\Cite{(li.~2. de Gen. ad lit. c.~17. & li.~12. c.~22.)}
\MNote{Sometimes lawful.}
Neuertheles ſuch ſignes are ſometimes lawfully obſerued, & deſired from
God, as the ſame Doctor diſputeth
\Cite{(li. queſt. ſuper Gen. q.~53.)}
and S.~Chriſoſtome teacheth more clerly
\Cite{(ho.~45. in Gen.)}
likewiſe S.~Theodoret
\Cite{(q.~73. in Gen.)}
\CNote{Ioan.~14.}
\MNote{Holie Scripture and the Church are iudges of doubtful
obſeruations.}
But whoſoeuer wil not erre in particular caſes, muſt folow either
expreſſe Scripture, or the iudgement of the Church, which is alwayes
directed by the ſpirite of truth.
\MNote{Eliezers prayer, for a particular ſigne, was lawful, deuout, and
diſcrete.}
And touching this prayer of Abrahams
ſeruant, and his deſire of this determinate ſigne, to know the maide,
whom God had prouided to be
\Fix{Iſaachs}{Iſaacs}{obvious typo, fixed in other}
wife, the fathers generally hold that it was religious, deuout, and
diſcrete. For he being careful of his maſters buſines, and not truſting
his owne iudgement, but relying vpon that Abraham had ſaid, \Emph{Our
Lord shal ſend his Angel before thee}, commended ſo weightie a cauſe to
God by prayer, the Angel ſuggeſting both to him to deſire, and to the
maide to performe (as the euent ſheweth) ſuch qualities and vertues in
her, as were moſt agreable to the great charitie and hoſpitalitie dayly
practiſed in Abrahams houſe, moſt conuenient and neceſſarie (as he
diſcretly conſidered) for that familie, and good of manie.
\CNote{Iudic.~7. 1.~Reg.~14.}
The like
obſeruations were approued in Gedeon, and Ionathas. And to pray for ſuch
ſignes in ſome caſe, or for manifeſt miracles is alſo approued by the
\CNote{Act.~1. &~4.}
Apoſtles example praying God \Emph{To shevv by lote vvhich of the tvvo
he had choſen to the Apoſtleship, in place of Iudas.} And that he would
extend his hand to cures, & ſignes, & wonders, to be done by the name of
his holie ſonne \Sc{Iesvs}.}
the maid, to whom I shal ſay: Bowe downe thy tankard that I may drinke:
and she shal anſwere: Drinke, yea to thy camels alſo wil I geue drinke:
she it is, whom thou haſt prouided for thy ſeruant Iſaac: and by this I
ſhal vnderſtand, that thou haſt delt mercifully
%%% o-0093
with my maiſter. \V Neyther had he yet ended theſe wordes within him
ſelfe, & behold Rebecca came forth, the daughter of Bathuel, the
ſonne of Melcha wife to Nachor the brother of Abraham, hauing a
tankard on her shoulder: \V a paſſing comely maide, & moſt beautiful
virgin, & not knowen to man: and she was gone downe to the fountaine,
and had filled her tankard, and came backe. \V And the ſeruant ranne to
mete her, and ſaid: Geue me a litle water to drinke of thy tankard. \V
Who anſwered: Drinke my lord. And quickly she let downe the tankard vpon
her arme, and gaue him drinke. \V And when he had druncke, she added:
but for thy camels alſo I wil drawe water, til al doe drinke. \V And
powring out the tankard into the troughes, she ranne backe to the wel to
drawe water: and being drawen gaue it to al the camels. \V But he muſing
beheld her with ſilence, deſirous to know whether our Lord had made his
iourney proſperous, or not. \V And after that the camels had drunck, the
man plucked forth golden earelettes, weying two ſicles: and as manie
braceletts of tenne ſicles weight. \V And he ſaid to her: Whoſe daughter
art thou? shew me: is there anie place in thy fathers houſe to lodge? \V
Who anſwered: I am the daughter of Bathuel, the ſonne of Melcha, whom
she bare to Nachor. \V And she added, ſaying: Of ſtrawe alſo and hay we
haue good ſtore, and a large place to lodge in. \V The man bowed him
ſelfe, and adored our Lord, \V ſaying: Bleſſed be the Lord God of my
lord Abraham, that hath not taken away his mercie & truth from my lord,
and hath brought me the ſtreight way into the houſe of my
lords brother. \V The maide therfore ranne, and reported into
\SNote{Her father hauing perhaps manie wiues and euerie one a ſeueral
houſe ſhe went to her
\Fix{mothes}{mothers}{obvious typo, fixed in other.}
houſe.}
her mothers houſe al things that ſhe had heard.
\Fix{}{\V }{obvious typo, fixed in other}%
And Rebecca had a
brother named Laban, who in al haiſt went forth to the man, where the
fountaine was. \V And when he had ſeene the eareletts and braceletts in
his ſiſters
%%% 0100
hands, and had heard al her words reporting: Theſe words ſpake the man
vnto me: he came to the man which ſtoode beſide the camels, and nighe to
the fountaine of water: \V and ſaid to him: Come in, thou bleſſed of our
Lord: Why ſtandeſt thou without? I haue prepared the houſe, and a place
for the camels. \V And he brought him in into his lodging: and he
vnharneſſed the camels, and gaue ſtrawe and hay, and water to wash his
feet, and of the men that were come with him. \V And bread was ſet before
him. Who ſaid: I wil not eate, til I ſpeake my meſſage. He anſwered him:
Speake. \V And he ſaid: I am the ſeruant of Abraham: \V and our Lord
hath bleſſed my lord wonderfully, and he is magnified: and he hath geuen
him sheepe, and oxen, ſiluer and gold, men ſeruants and wemen ſeruants,
camels, and aſſes. \V And Sara my lordes wife hath borne my lord a ſonne
in her old age, and he hath geuen him al things that he had. \V And my
lord adiured me ſaying: Thou ſhalt not take a wife for my ſonne of the
Chananites, in whoſe land I dwel: \V but thou ſhalt goe to my fathers
houſe, and of mine owne kinred ſhalt thou take a wife for my ſonne: \V
but I anſwered my Lord: What if the woman
%%% o-0094
wil not come with me? \V Our Lord, ſaith he, in whoſe ſight I walke, wil
ſend his Angel with thee, and wil direct thy way: and thou shalt take a
wife for my ſonne of myne owne kinred, and of my fathers houſe. \V Thou
shalt be innocent from my curſe, when thou shalt come to my kinne, and
they wil not geue her thee. \V I came therfore to day to the wel of
water, and ſaid: O Lord God of my lord Abraham, if thou haſt directed my
way, wherin I now walke, \V behold I ſtand beſides the wel of water, and
the virgin, that shal come forth to drawe water, when ſhe ſhal heare me
ſay: Geue me a litle water to drinke of thy tankard: \V and ſhe ſhal ſay
to me: Drinke both thou and for thy camels I wil drawe alſo: that is the
woman, which our Lord hath prepared for my maiſters ſonne. \V And
whileſt I pondered theſe things ſecretly with my ſelfe, Rebecca appeared
comming with a tankard, which she caried vpon her shoulder: and she
went downe to the fountaine, & drew water. And I ſaid to her: Geue me a
litle to drinke. \V Who ſpedelie let downe the tankard from her
shoulder, and ſaid to me: Drinke both thou, and to thy camels I wil geue
drinke.
%%% 0101
I dranke, and ſhe watered the cammels. \V And I asked her, and ſaid:
Whoſe daughter art thou? who anſwered: I am the daughter of Bathuel, the
ſonne of Nachor, whom Melcha bare him. I hoong therfore earelettes to
adorne her face, and I put braceletts vpon her hands. \V And proſtrate I
adored our Lord, bleſſing the Lord God of my lord Abraham, who hath
brought me the ſtraight way to take the daughter of my lords brother for
his ſonne. \V Wherfore if you doe according to mercie and truth with my
lord, ſhew me: but if it pleaſe you otherwiſe, that alſo tel me, that I
may goe to the right hand, or to the left. \V And Laban and Bathuel
anſwered: From our Lord the word hath proceded: we can not ſpeake any
other thing with thee beſides his pleaſure. \V Behold Rebecca is before
thee, take her and goe thy waies, and let her be the wife of thy lords
ſonne, as our Lord hath ſpoken. \V Which when Abrahams ſeruant heard,
falling downe he adored our Lord to the grounde. \V And taking forth
veſſel of ſiluer, and gold, and garments, gaue them to Rebecca for a
preſent. To her brothers alſo, and to her mother he offred giftes. \V A
banket was made, and eating and drinking togeather thy lodged there. And
in the morning, the ſeruant aroſe, and ſaid: Diſmiſſe me, that I may goe
to my lord. \V And her brother, and mother anſwered: Let the maide tarie
at the leaſt tenne days with vs, and after ſhe ſhal depart. \V Stay me
not, ſaid he, becauſe our Lord hath directed my way: diſmiſſe me that I
may goe on to my lord. \V And they ſaid: Let vs cal the maid, and
\SNote{As children ought not to mary without their parents good liking:
ſo the parties owne conſent is moſt neceſſarie.
\Cite{S.~Amb. Epiſt.~43.}}
aske her wil. \V And being called, when ſhe was come, the asked: Wilt
thou goe with this man? who ſaid: I wil goe. \V They diſmiſſed her
therfore, and her nurce, and Abrahams ſeruant, and his companie, \V
wiſhing proſperitie to their ſiſter, and ſaying: Thou art our ſiſter,
encreaſe thou into thouſand thouſands, and thy ſeed poſſeſſe the gates
of their enemies. \V Therfore Rebecca, and her maides being ſet vpon
%%% o-0095
camels, folowed the man: who with ſpeed returned vnto his lord, and \V
the ſame time Iſaac walked along the way, that leadeth to the wel of the
Liuing and the ſeing, ſo called: for he dwelt in the ſouth countrie; \V
and he was gone forth to
\SNote{\HH{Suach} ſignifieth to ſpeake conſideratly with hart or
mouth. Here S.~Ambroſe
\Cite{(li.~1. c.~1. de Iſaac)}
and S.~Aug.
\Cite{(q.~69. in Gen.)}
vnderſtand it of mental prayer.}
meditate in the field, the day nowe being wel ſpent: and when he had
caſt vp his eyes, he ſaw camels coming a farre of. \V Rebecca
%%% 0102
alſo, when ſhe ſaw Iſaac, lighted of the camel, \V and ſaid to the
ſeruant: Who is that man which cometh towards vs along the field? And he
ſaid to her: The ſame is my lord. But she quickly taking her cloake,
couered her ſelfe. \V And the ſeruant told Iſaac al things that he hed
done. \V Who brought her into the tent of Sara his mother, and tooke her
to wife: and he loued her ſo much, that it did moderate the ſorrowe,
which was chanced of his mothers death.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
