%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 34
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-34


\project douay-rheims


%%% 0128
%%% o-0119
\startChapter[
  title={Chapter 34}
  ]

\Summary{For rauishing Dina, the Sichimites (being firſt circumciſed)
  are ſlaine by Simeon and Leui her brothers. 27.~The reſt of Iacobs
  ſonnes ſpoile the citie. 30.~Iacob blameth them, fearing harme may
  come by this fact.}

And Dina the daughter of Lia went forth
\SNote{O Dina (ſaith S.~Bernard) what nede was there to ſee wemen of a
ſtrange countrie?
\Cite{Tract. de gradib. humilitatis.}}
to ſee the wemen of that countrie. \V Whom when Sichem had ſeene the
ſonne of Hemor the Heuite, the prince of that land, he was in loue with
her: and he tooke her away, and lay with her, by force rauiſhing the
virgin. \V And his ſoule was faſt
\Fix{kint}{knit}{obvious typo, fixed in other}
vnto her, and wheras ſhe was ſad, he comforted her with ſweete
wordes. \V And going to Hemor his father, he ſaid: Take me this wench to
be my wife. \V Which when Iacob had heard, his ſonnes being abſent, and
occupied in feeding of the cattle, he held his peace til they
returned. \V And when Hemor Sichems father was come forth to ſpeake vnto
Iacob, \V behold his ſonnes came out of the field: and hearing what had
paſſed, they were paſſing wrath, becauſe he had done a foule thing in
Iſrael, and committed an vnlawful fact, in rauiſhing Iacob's
daughter. \V Hemor therfore ſpake to them: The ſoule of my ſonne Sichem
is faſtned to your daughter; Geue her vnto him to
%%% 0129
wife: \V and let vs contract mariages one
%%% o-0120
with an other: geue vs your daughters, and take you our daughters. \V
And dwel with vs: the land is at your commandement, tille, occupie, and
poſſeſſe it. \V Yea and Sichem alſo ſaid to her father and to her
brethren: Let me finde grace in your ſight: and what ſoeuer you ſhal
appointe I wil geue: \V raiſe the dowrie, and require giftes, and I ſhal
gladly geue, what you ſhal demande: only geue me this wench to wife. \V
Iacobs ſonnes anſwered Sichem & his father
\SNote{They offended by falſly pretending religion, and by exceſſe in
reuenge, & therfore are reproued by their father.
\XRef{v.~30.}
&
\XRef{chap.~49. v.~5.}
Otherwiſe their zeale was iuſt to puniſh ſo foule a fault.
\XRef{Iudith.~9.}}
in guile, being wrath for the deflouring of their ſiſter: \V We can not
doe that which you demande, nor geue our ſiſter to an vncircumciſed
perſon: which with vs is an vnlawful & abhominable thing. \V But in this
order we may be confederate, if you wil be like to vs, and al the man
ſex among you be circumciſed: \V then wil we geue and take mutually your
daughters, and ours: and we wil dwel with you, and wil be one people: \V
but if you wil not be circumciſed, we wil take our daughter, and
depart. \V The offer pleaſed Hemor, and Sichem his ſonne: \V neither did
the young man make delay, but forthwith fulfilled that which was
demanded: for he loued the wench exceedingly, and he was the greateſt
man in al his fathers houſe. \V And going into the gate of the citie,
they ſpake to the people: \V Theſe men are men of peace, and are willing
to dwel with vs: let them occupie in the land, and til it, which being
large and wide doth lacke men to tille it: their daughters we ſhal take
to wife, and ours we wil geue to them. \V One thing there is for the
which ſo great a good is differred: If we circumciſe our men ſexe,
folowing the rite of the nation. \V And their ſubſtance, and cattle, and
al things that they poſſeſſe, ſhal be ours: only in this let vs
condeſcend, and dwelling togeather, we ſhal make one people. \V And they
al aſſented, and circumciſed al the man ſex. \V And behold the third
day, when the griefe of the woundes is moſt paineful: Iacobs two ſonnes,
Simeon and Leui the brothers of Dina, taking their ſwordes, entred into
the citie boldly: and killing al the man ſex, \V murdred withal Hemor
and Sichem, taking away Dina their ſiſter out of Sichems houſe. \V When
they were gone forth, the other ſonnes of Iacob ranne in vpon them that
were ſlaine: and ſpoiled the citie in reuenge of the rape. \V And
waſting al things that were in their houſes, and fildes, their ſheepe
and
%%% 0130
heardes, and aſſes, \V their litle ones alſo, and their wiues they led
away captiue. \V Which things when they had boldly atcheiued, Iacob ſaid
to Simeon and Leui: You haue trubled me, and made me odious to the
Chananites, and Pherezites the inhabiters of this land. We are few: they
being gethered together wil ſtrike me; and I, and my houſe ſhal be
deſtroyed. \V They anſwered: What ſhould they abuſe our ſiſter as a
ſtrumpet?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
