%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 25
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-25


\project douay-rheims


%%% 0102
%%% o-0096
\startChapter[
  title={Chapter 25}
  ]

\Summary{Abraham hauing manie children by his wife Cetura, died at the
  age of 175.~yeares: 12.~Iſmael alſo hauing twelue ſonnes dukes,
  died. 19.~Iſaac praying for his barren wife, she hath Eſau and Iacob
  twinnes. 30.~Eſau ſelleth his firſt birth right to Iacob for a meſſe
  of potage.}

%%% 0103
And Abraham maried an other wife named
\Fix{Cethura:}{Cetura:}{likely typo, fixed in other}
\V which bare him Zamran, and Iecſan, and Madan, and Madian, and Ieſboc,
and Sue. \V Iecſan alſo begat Saba and Dadan. The Children of Dadan were
Aſſurim, and Latuſim, and Loomim. \V But alſo of Madian was borne Epha,
and Opher, and Henoch, and Abida, and Eldaa: al theſe were the children
of Cetura. \V And Abraham gaue al his poſſeſſions to Iſaac: \V and
\LNote{To the children of his concubines.}{S.~Auguſtin
\MNote{VVhy Agar & Cetura being lawful wiues are called cõcubines.}
\Cite{(li.~16. c.~34. de ciuit)}
ſheweth that both Agar and Cetura, being Abrahams lawful wiues (for ſo
they are called in holie Scripture) are alſo called concubines, becauſe
they had not like priuileges to Sara, whoſe ſonne was ſole heyre to his
father, and the children of the others had only giftes (or mouable
goods) not attayning to the promiſed kingdom. And al this for myſterie
ſake.
\MNote{Their childrẽ ſignified Pagaines & Heretikes.}
For Iſmael ſignified the carnal people before Chriſt, the children of
Cetura prefigured Heretikes, who ſuppoſe themſelues to pertaine to the
new Teſtament, but are ſeparated no leſſe then the Iewes from Chriſts
Kingdom. And albeit there was alſo an other particular reaſon, why Agar
was called concubine, becauſe ſhe was a ſecond wife, the firſt then
liuing, yet this lerned father ſaith, he did not ſee, why Cetura being
maried after the death of Sara, ſhould be called concubine, but only for
this Myſterie.}
to the children of his concubines he gaue gifts, and ſeparated them from
Iſaac his ſonne, whileſt himſelfe yet liued, to the eaſt countrie. \V
And the days of Abrahams life were a hundred ſeuentie and fiue
yeares. \V And decaying dyed in
\SNote{The life of the iuſt hath fulnes of dayes though it be otherwiſe
ſhort; the daies of the wicked are void of fruict, be they manie or few.
\Cite{S.~Ambroſe li. de Abraham.}}
a good old age, and hauing liued a great time, and being ful of days:
and was gethered to his people. \V And there buried him Iſaac and Iſmael
his ſonnes in the duble caue, which was ſituated in the field of Ephron
the ſonne of Seor the Hethite, ouer againſt Mambre, \V which he had
bought of the children of Heth: there was he buried, and Sara his
wife. \V And after his death God bleſſed Iſaac his ſonne, who dwelled
beſide the wel of the Liuing and ſeing ſo named. \V Theſe are the
generations of Iſmael the ſonne of Abraham, whom Agar the Ægyptian bare
him, Saraes ſeruant: and \V theſe are the names of his children
according to their calling and generations. The firſt begotten of Iſmael
Nabaioth, then Cedar, and Adbeel, and Mabſam, \V Maſma alſo, and Duma,
and Maſſa, \V Hadar, and Thema, and Iethur, and Naphis, and Cedma. \V
Theſe are the ſonnes of Iſmael: and theſe are their names by their caſtles
and townes, twelue princes of their tribes. \V And the yeares of Iſmaels
life came to an hundred thirtie ſeauen, and decaying died, and was put
vnto his people. \V And he dwelt from Heuila euen to Sur, which looketh
towards Ægypt, as they enter to the Aſſirians, before the face of al
his bretheren died he. \V Theſe alſo are the generations of Iſaac the
ſonne of Abraham: Abraham begat Iſaac: \V who when he was fortie yeares
old, tooke to wife Rebecca the daughter of Bathuel the Syrian of
Meſopotamia, ſiſter to Laban. \V And Iſaac beſought our Lord for his
wife, becauſe ſhe was barren: who
\LNote{Heard him.}{Notwithſtanding
\MNote{Gods predeſtination and forſeing include, & not exclude the
meanes by which his wil is done.}
Gods aſſured promiſe, that
\Fix{Iſaach}{Iſaac}{likely typo, fixed in other}
ſhould haue iſſue
\XRef{(Gen.~21. v.~12.)}
yet he prayeth inſtantly for the ſame. And Moyſes here attributeth
Rebeccas conceiuing to 
\Fix{Iſaachs}{Iſaacs}{likely typo, fixed in other}
prayer. Wherby we ſee that Gods forſeing, predeſtinating, and promiſing
exclude not, but in dede include ſecondarie cauſes, and ordinarie
meanes, by which his eternal wil and pleaſure is fulfilled. For as God
did forſee that Rebecca ſhould haue children, ſo he did forſee, that
Iſaac ſhould pray for it, and obtayne it; and the one was as ſure to
come to paſſe as the other. And the ſame conſequence is true concerning
eternal life as S.~Gregorie teacheth.
\Cite{li.~1. c.~8. Dialog.}}
heard him, and made
%%% o-0097
Rebecca to conceaue. \V But the little ones ſtrugled in her wombe; who
ſaid: If it ſhould be ſo with me, what nede was there to conceaue? And
ſhe
%%% 0104
\SNote{S.~Auguſtin
\Cite{(q.~72. in Gen.)}
diſputeth but could not decide, whether Rebecca went to ſome Prieſt, or
Prophet, or Altar, or whither els, or only retyred to priuate prayer.}
went to conſult our Lord. \V Who anſwering ſaid: Two nations are in thy
wombe, and two peoples ſhal be diuided out of thy wombe, and one people
ſhal ouercome the other, and
\LNote{The elder shal ſerue the younger.}{As
\MNote{The couenant made to Abraham pertained only to Iſaac, and Iacob,
not to the reſt of his iſſue.}
before
\XRef{(c.~17. v.~21.}
&
\XRef{c.~21. v.~12.)}
the couenant and great promiſes made to Abrahams ſede, are declared to
pertaine only to Iſaac, and not to Iſmael, nor to the other brothers: ſo
the ſame belong not to Eſau, the elder, but only to Iacob the younger
ſonne of Iſaac, the Holie Ghoſt ſaying, \Emph{The elder shal ſerue the
younger.} And withal ſignifieth, (ſaith
\CNote{li.~16. c.~35. ciuit.}
S.~Auguſtin) that the elder
people of the Iewes ſhal ſerue the younger Chriſtian people. For
although it may be vnderſtood literally to be fulfilled, in that the
\CNote{2.~Reg.~8.}
Idumeans coming of Eſau, were
\CNote{Pſal.~59.}
ſubdued by King Dauid coming of Iacob; yet
it is more conueniently beleeued, that this prophecie tended to a
greater thing. And what is this, but that which is euidently fulfilled
in the Iewes and Chriſtians?

An
\MNote{Gods mere mercie in electing anie, his iuſtice to the reprobate.}
other great document of grace
\CNote{Rom.~9.}
S.~Paul geathereth vpon this Myſterie:
that the twinnes being not yet borne, nor hauing done any thing good or
euil, without anie good merites, the younger is elected, the elder
reprobate. For doubtles (ſaith S.~Auguſtin) touching original ſinne they
were both equal, and concerning proper ſinne, neither of them had anie
at al. By which example he ſheweth Gods mere mercie in the elect, and
iuſtice in the reprobate, as is more largely noted in the
\Cite{Engliſh New Teſtament, vpon the ninth chapter to the Romanes.}}
the elder ſhal ſerue the younger. \V Now her time was come to be
deliuered, and behold twinnes were found in her wombe. \V He that came
forth firſt, was read, and al hearie in manner of a skinne: and his name was
called Eſau. Immediatly the other coming forth, held his brothers plant
in his hand: and therfore he called him Iacob. \V Threeſcore yeares old
was Iſaac, when the litle ones were borne vnto him. \V Who being growne
vp, Eſau became a man cunning in hunting, and a huſband man: but Iacob
\SNote{Holie Scripture premoniſheth Iacobs ſinceritie, leſt in the
Myſteries folowing he might be ſuſpected of falſe dealing.
\Cite{S.~Aug. li.~16. c.~37. ciuit.}}
a plaine man dwelled in tents. \V Iſaac loued Eſau, becauſe he did eate
of his hunting: and Rebecca loued Iacob. \V And Iacob boyled broth: to
whom Eſau being come faynt out of the field, \V ſaid: Geue me of this
read broth, becauſe I am exceding faint. For which cauſe his name was
called Edom. \V To whom Iacob ſaid:
\LNote{Sel me thy firſt birth right.}{Iacob
\MNote{Iacob lawfully bought but Eſau ſinned in ſelling the
firſt-birth-right.}
inſtructed by his mother, that God had choſen him in place of his
brother Eſau (for to her God had reueled that the elder ſhould ſerue
the younger) did lawfully vſe this oportunitie to get Eſaus grant of the
right pertaining to the firſt borne, but Eſau in ſelling it ſinned,
ſhewing himſelfe an intemperate prophane man.
\XRef{Heb.~12.}}
Sel me thy firſt-birth-right. \V He anſwered, Loe I dye, what wil the
firſt birth right auaile me? \V Iacob ſaid: Sweare therfore to me. Eſau
ſware to him, and ſould his firſt-birth-right. \V And ſo taking bread
and the rice broth, did eate, and drinke, and went his way; little
eſteeming that he had ſold his firſt birth right.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
