%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 0056
%%% o-0051
\startChapter[
  title={Chapter 8}
  ]

\Summary{The
\MNote{The ſecond age of the world.}
waters diminishing by litle and litle, 6.~Noe ſendeth forth a crow, 8.~after
him a doue, thriſe: 18.~laſtly goeth forth with al that were with him in
the arke, 20.~erecteth an Altar, and offereth Sacrifice.}

And
\MNote{The third part of this booke. Of the new increaſe &
multiplication of the world.}
God remembred Noe, and al the beaſts, and al the cattle, which were with
him in the arke, and brought a winde vpon the earth, and the waters
decreaſed. \V And the fountaines of the depth, and the floud gates of
heauen, were ſhut vp: and the rayne from heauen was ſtayed. \V And the
waters returned from the earth going & comming:
%%% o-0052
and they begane to decreaſe after a hundred fiftie dayes. \V And the
arke reſted the ſeauenth moneth, the ſeauen & twentith day of the moneth
vpon the mountaines of Armenia. \V But the waters for al that were going
and decreaſing vntil the tenth moneth: for in the tenth moneth, the
firſt day of the moneth, the topps of the mountaines appeared. \V And
after that fourtie dayes were paſſed, Noe opening the windowe of the
arke, which he had made, let forth a crowe: \V which went forth, and did
\SNote{The crowe returned not into the arke, but (as appeareth by the
Hebrew text) going and returning reſted vpon the arke.}
not returne, til the waters were dried vpon the earth. \V He ſent forth
alſo a doue after him, to ſee if the waters were ceaſed yet vpon the
face of the earth. \V Which finding not where her foote might reſt,
returned to him into the arke: for the waters were vpon the whole earth:
and he ſtretched forth his hand, and caught her and brought her into the
arke. \V And hauing expected yet ſeauen moe dayes, againe he let forth a
doue out of the arke. \V But ſhe came to him at euentide, carrying a
bough of an oliue tree, that had greene leaues in her mouth. Noe
therfore vnderſtood that the waters were ceaſed vpon the earth. \V And
he expected yet neuertheles other ſeauen dayes: and he ſent forth a
doue, which returned not any more vnto him. \V Therfore in the ſixt
hundred and one yeare, the firſt moneth, the firſt day of the moneth the
waters were cleane diminished vpon the earth: and Noe opening the roofe
of the arke, looked, and ſawe that the face of the earth was dried. \V
In
\SNote{They entred into the arke the 17.~day, the ſecõd moneth of the
other yeare: ſo they remained there 12.~monethes and tenne dayes.}
the ſecond moneth, the ſeuen & twentyth day of the moneth the earth was
dried. \V And God ſpake to Noe, ſaying:
%%% 0057
Goe forth of the arke, thou & thy wife, thy ſonnes and the wiues of thy
ſonnes with thee. \V Al cattle, that are with thee of al flesh, as wel
in foules, as in beaſtes, & al creepers, that creepe vpon the earth,
bring out with thee, & goe yee vpon the earth: increaſe and multiplie
vpon it. \V Noe therfore went forth, and his ſonnes: his wife, and the
wiues of his ſonnes with him. \V Yea and al cattle, beaſtes, and
creepers that crepe vpon the earth, according to their kinde, went forth
out of the arke. \V And Noe
\LNote{Built an Altar.}{Noe
\MNote{Noes ſacrifice manie wayes cõmendable.}
without expreſſe commandment, and without delay, offereth Sacrifice to
God, for the benefite receiued, in his, and his families conſeruation,
with the other liuing creatures, in that general deluge of the
world, \Emph{VVel knovving} (ſaith
\CNote{Li. de Noe & Arca. c.~22.}
S.~Ambroſe) \Emph{that to be true
thankeſgeuing,
\MNote{1.~Voluntarie.}
vvhich is preſented, not commanded:
\MNote{2.~Speedie.}
therfore he made no delay. For the vertue of a gratful mind excludeth
doutful deliberation, and he that expecteth, til the debt of thankes be
exacted, is an vngratful perſon.}
\MNote{3.~Solemne.}
For more ſolemnitie, he dedicated an apt and permanent place, for this
peculiar diuine ſeruice, \Emph{Building an Altar to our Lord.} The Hebrew
word \HH{Mizbeach} (of the verbe \HH{Zabach}, to kil, or make
ſacrifice) and the Greeke
\CNote{Heb.~13.~10.}
\GG{Thyſiaſterion}, ſignifie an Altar to ſacrifice on, not a common
table for meate.
\MNote{4.~Pure.}
He offered of the cleane and beſt things becauſe pure and deuout
Sacrifice is due to God.
\MNote{5.~Bountiful.}
Moreouer it was large and bountiful, for he offered of al the kindes of
cleane beaſtes and foules.
\MNote{6.~Holocauſt.}
Finally he offered them in \Emph{Holocauſtes}, where al was burned and
conſumed in the honour of God. How gratful al this was to God, Moyſes
ſignifieth ſaying:
\CNote{Leuit.~1.}
\Emph{Our Lord ſmelled a ſvveete ſauour.}
\CNote{Pſal.~50. Iſaie.~1. Amos.~5. Mal.~1.}
Not that
either anie ſweete corporal ſauour could of it ſelfe delight God, who is
the moſt ſpiritual ſubſtance, or that the burning of fleſh, bones, and
bowels of beaſtes could yeld ſweete ſauour;
\MNote{Sacrifice is pleaſant to God: not for the external things but for
the ſincere mind.}
but the deuout mind declared by ſuch external dutie greatly pleaſed
God. For God requireth both, but ſpecially a ſincere hart.
\CNote{Iuſt. Mar. li.~3. q.~4. S.~Hiero. Q.~Heb.}
As not only
diuine Scriptures, and holie Fathers, but alſo moral Philoſophers teach
vs. \Emph{It vvere a
greuous thing} (ſaith Plate writing of ſacrifices) \Emph{if God had
reſpect rather to the giftes, and ſacrifices of
men, then to their mind.}
\Cite{lib. peri tes proſeuches.}}
built an Altar to our Lord: and taking of al cattle and foules that were
cleane, offered Holocauſts vpon the Altar. \V And our Lord ſmelled a
ſweete ſauour, and ſaid: I wil no more curſe the earth for men: for the
ſenſe and cogitation of mans hart are prone to euil from their youth: I
wil no more therfore ſtrike euerie liuing ſoule as I haue done. \V Al
the dayes of the earth,
\SNote{In the whole yeare of the floud was no ſowing nor reaping, nor
pleaſant varietie of times, but al deſolate & miſerable: henceforth God
promiſeth more ſeaſonable times.
\Cite{S.~Amb. li. de Noe & Arca. c.~23.}}
ſeedtime and harueſt, cold and heate, ſommer and winter, night and day
ſhal not reſt.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
