%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 45
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-45


\project douay-rheims


%%% 0156
%%% o-0144
\startChapter[
  title={Chapter 45}
  ]

\Summary{Ioſeph manifeſteth himſelf to his brethren. Who being much
  terified he comforteth them, and weeping embraceth euerie one. 16.~The
  bruite wherof coming to Pharao, he congratulating commandeth Ioſeph to
  cal his father with al his familie into Ægypt. 21.~So the eleuen
  brothers are ſent away with giftes and prouiſion for Iacobs
  iourney. 26.~Al which the father vnderſtanding is reuiued in ſpirite.}

Ioſeph could no longer refrayne manie ſtanding by in preſence: wherupon
he commanded that al ſhould goe forth, and no ſtranger ſhould be preſent
at their acknowledging one of another. \V And he lifted vp his voice
with weeping, which the Ægyptians heard, and al the houſe of Pharao. \V
And he ſaid to his Brethren: I am Ioſeph: is my father yet liuing? His
brethren could not anſwer him being teribly aſtonyed out of meaſure. \V
To whom gently he ſaid: Come hither to me. And when they were come nere
him, I am, quoth he, Ioſeph, your brother, whom you ſolde into Ægypt. \V
Be not affraid, neither let it ſeeme to you a hard caſe that you did ſel
me into theſe countries: for
\SNote{Gods prouidence turned their euil dealing to the good of the
whole familie.
\XRef{chap.~50. v.~20.}}
God ſent me before you into Ægypt for your preſeruation. \V For it is
two yeares ſince the famine begain to be vpon the earth, and yet fiue
yeares remaine, wherin there can be neither earing nor reaping. \V And
God ſent me before, that you may be
%%% 0157
preſerued vpon the earth, and may
haue victuals to liue. \V Not by your counſail, but by the wil of God
was I ſent hither: who hath made me as it were a father to Pharao, and
lord of his whole houſe, and prince in al the land of Ægypt. \V Make
haſt and get ye vp to my father, and you shal ſay to him: Thy ſonne
Ioſeph willeth thus: God hath made me lord of the whole land of Ægypt:
come downe to me, tary not. \V And thou ſhalt dwel in the land of
Geſſen: and thou ſhalt be nere me thou and thy ſonnes, and thy ſonnes
children, thy ſhepe, and thy heardes, and al things that thou doſt
poſſeſſe. \V And there I wil fede thee (for yet there are fiue yeares of
famine remayning) leſt both thou perish, and thy houſe, & al things that
thou doſt poſſeſſe. \V Behold, your eyes, and the eyes of my brother
Beniamin doe ſee, that it is my mouth that ſpeaketh vnto you. \V Report
to my father my whole glorie and al things that you haue ſeene in Ægypt:
make haſt, and bring him to me. \V And falling vpon the neck of his
brother Beniamin, embracing him he wept: he alſo in like maner weeping
vpon his neck. \V And Ioſeph kiſſed al his brethren, and wept vpon
euerie one: after which things they were
%%% o-0145
bold to ſpeake vnto him. \V And it was heard of, and very famouſly
reported abrode in the kings courte: The brethren of Ioſeph are come:
and
\SNote{Ioſephs prudẽt proceding before he made him ſelfe knowen to his
brethrẽ and them to Pharao, procured al this ioy & fauour towards them
in Ægypt.}
Pharao was glad, and al his familie. \V And he ſpake to Ioſeph that he
ſhould command his brethren, ſaying: Loading your beaſts goe into the
Land of Chanaan, \V and take thence your father and kinne, and come to
me: and I wil geue you al the good things of Ægypt, that you may eate
the marow of the land. \V Geue commandment alſo that they take waynes
out of the land of Ægypt, for the carage of their litle ones and wyues:
and ſay: Take vp your father, and make haſt to come with al ſpede. \V
Neither doe you leaue any thing of your houshould ſtuffe: for al
the riches of Ægypt, shal be yours. \V And the ſonnes of Iſrael did as
it was commanded them. To whom Ioſeph gaue waynes according to Pharaos
commandment: and victuals for the way. \V He bad alſo to be brought vp
for euery one two robes: but to Beniamin he gaue three hundred peeces of
ſiluer with fiue robes of the beſt: \V ſending to his father as much
money and rayment, adding beſides them he aſſes that should carie of al
the riches of Ægypt, and as many shee aſſes, carying wheat for the
iourney and bread. \V Therfore
%%% 0158
he diſmiſſed his brethren, and when they were departing he ſaid to them:
\LNote{Be not angrie in the vvay.}{Trauelers
\MNote{Occaſion of ſinne to be auoided.}
in iourney are eaſily prouoked to anger and brawling: eſpecially if they
auoide not probable occaſions. Therfore Ioſeph admoniſheth his brothers
to beware therof, leſt in talking of him, and how they had ſold him to
ſtrangers, ſome of them might accuſe others, and excuſe themſelues, and
ſo fal in to new offences.
\Cite{S.~Chris. ho.~64. in Gen.}}
Be not angrie in the way. \V Who going vp out of Ægypt, came into the
land of Chanaan to their father Iacob. \V And they told him ſaying:
Ioſeph thy ſonne is liuing: and he ruleth in al the Land of Ægypt. Which
when Iacob heard, awaking as it were out of a heauie ſleepe,
notwithſtanding did not beleeue them. \V They on the contrarie ſide
reported the whole order of the thing. And when he ſawe the waynes and
al things that he had ſent, his ſpirit reuiued, \V and he ſaid: It
ſufficeth me if Ioſeph my ſonne be liuing yet: I wil goe, and ſee him
before I dye.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
