%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 46
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-46


\project douay-rheims


%%% 0158
%%% o-0146
\startChapter[
  title={Chapter 46}
  ]

\Summary{Iſrael
\MNote{The eight and laſt part of this booke.}
warranted in a viſion from God, goeth into Ægypt with al
  his 
\Fix{famimlie,}{familie,}{obvious typo, fixed in other}
  8.~who are here recited. 28.~Ioſeph meeting him in
  Geſſen, aduiſeth him to tel Pharao, that they are shepheards by their
  trade of life.}

And
\MNote{Of Iacob and his progenies going into Ægypt; Of his and Ioſephs
death.}
Iſrael taking his iourney, with al things that he had
\LNote{Came to the vvel of oath.}{In
\MNote{God reueleth his wil in holie places.}
this holie place (called \HH{Berſebee}, that is, \Emph{vvel of oath},
where Abraham and Iſaac had confirmed by oath, their league with the
Kings of the countrie, and erected Altares, Iacob alſo conſulted God,
about his going into Ægypt, and was commanded to goe with al that he
had.}
came to the wel of the oath, and killing there victimes to the God of
his father Iſaac. \V He heard him by a viſion of the night calling him,
and ſaying vnto him: Iacob, Iacob. To whom he anſwered: Loe here I
am. \V God ſaid to him: I am the moſt mightie God of thy father: feare
not, goe downe into
\Fix{Ægipt,}{Ægypt,}{possible typo, fixed in other}
for into a great nation wil I make thee
there. \V I wil goe downe with thee thither, and thence wil I bring thee
returning: Ioſeph alſo ſhal put his handes vpon thine eyes. \V And Iacob
roſe vp from the wel of the oath: and his ſonnes tooke him vp, with
their little ones and wiues in the waynes, which Pharao had ſent to carie
the old man, \V and al that he had poſſeſſed
%%% 0159
in the Land of Chanaan: and he came into Ægypt with al his ſede, \V his
ſonnes, and nephewes, daughters, and al his progenie together. \V And
theſe are the names of the children of Iſrael, that entred into Ægypt,
him ſelfe with his children. His firſt-begotten Ruben. \V The ſonnes of
Ruben: Henoch and Phallu and Heſron and Charmi. \V The ſonnes of Simeon:
Iamuel and Iamin and Ahod, and Iachin and Sohar, and Saul the ſonne of
Chananiteſſe. \V The ſonnes of Leui: Gerſon and Caath and Merari. \V The
ſonnes of Iuda: Her and Onan & Sela and Phares and Zara. And Her and
Onan died in the land of Chanaan. And there were ſonnes borne to Phares:
Heſron and Hamul. \V The ſonnes of Iſſachar: Thola and Phua and Iob and
Semron. \V The ſonnes of Zabulon: Sared & Elon and Iahelel. \V Theſe are
the ſonnes of Lia, which ſhe bare
\SNote{That is, she bare their fathers in Meſopotamia.
\Cite{S.~Aug. q.~151. in Gen.}}
in Meſopotamia of Syria with Dina his daughter. Al the ſoules of his
ſonnes and daughters are thirtie three. \V The ſonnes of Gad: Sephion
and Haggi and Siuni and Eſebon and Heri and Arodi and Areli. \V The
ſonnes of Aſer: Iamne and Ieſua and Ieſſui and Beria, Sara alſo their
ſiſter. The ſonnes of Beria: Heber and Melchiel. \V Theſe be the ſonnes
of Zelpha, whom Laban gaue to Lia his daughter, and theſe ſhe bare to
Iacob ſixtene ſoules. \V The ſonnes of Rachel Iacobs wife: Ioſeph and
Beniamin. \V And theſe were ſonnes borne to Ioſeph, in the Land of
Ægypt, which Aſeneth the daughter of Putiphar prieſt of Heliopolis
bare to him: Manaſſes and Ephraim. \V The ſonnes of Beniamin: Bela and
Bechor
%%% o-0147
and Aſbel and Gera and Naaman and Echi and Ros and Mophim and
Ophim and Ared. \V Theſe be the ſonnes of Rachel, which ſhe bare to
Iacob: al the ſoules, fourtene. \V The ſonnes of Dan: Huſim. \V The
ſonnes of Nepthali: Iaziel and Guni and Ieſer and Sallem. \V Theſe be
the ſonnes of Bala, whom Laban gaue to Rachel his daughter: and theſe
ſhe bare to Iacob: al the ſoules, ſeuen. \V Al the ſoules, that entred
with Iacob into Ægypt, and that came out of his thighe, beſides his
ſonnes wiues,
%%% !!! This note includes the next one also
\LNote{Sixtie ſix.}{The difficultie in theſe two verſes, concerning the
number of Iſraelites, that were at firſt in Ægypt with Iacob, is eaſily
explicated, that iuſt ſixtie ſix, of his owne iſſue, came in with him;
and himſelfe being counted maketh ſixtie ſeuen, adding alſo Ioſeph (who
was there before) and his two ſonnes Manaſſes and Ephraim (borne there)
they were in al ſeuentie.
\MNote{A difficultie how manie Iſraelites came at firſt into Ægypt.}
But a far greater difficultie remaineth: for beſides theſe ſeuentie
perſons, the Septuagint Greeke Interpreters number and name fiue more;
to wit, a ſonne and a nephew of Manaſſes (the firſt
called \Emph{Machir}, the other \Emph{Galaad}) and two ſonnes of Ephraim
(called \Emph{Sutalaam}, and \Emph{Taam}) and one nephew
(named \Emph{Edem}) which number of ſeuentie fiue.
\CNote{Act.~7.}
S.~Steuen alſo
citeth, folowing the vulgar knowen Scripture of the Septuagint, rather
then the Hebrew text. Now in what ſenſe, theſe fiue could be ſaid to
haue entred into Ægypt with Iacob, being not then borne; may ſome of
them not borne during Iacobs life, for Iacob liued but ſeuentene yeares
in Ægypt
\XRef{(chap.~47. v.~28.)}
and Ioſeph being maried but nine yeares before (for this was the ſecond
deare yeare, after the ſeuen plentiful yeares) his ſonnes could not
excede ſeuen or eight yeares, when Iacob came to Ægypt, and ſo being but
24.~or 25.~yeares old at his death, could not then be grandfathers: how
therfore theſe fiue named by the Septuagint, and ſome others, not then
borne, of the lines of Phares and Beniamin, recited here by Moyſes,
could be ſaied to come with Iacob into Ægypt,
\MNote{Numbers myſtical, ſometimes not explicable in the literal ſenſe.}
S.~Auguſtin findeth ſo
inſoluble, that he doubteth not to affirme, ſome great hidden myſterie
to be vnderſtood by the Septuagint Interpreters, in theſe numbers, not
otherwiſe perhaps explicable according to the letter.}
ſixtie ſix. \V And the ſonnes of Ioſeph, that were borne to him in the
land of Ægypt, two ſoules. Al the ſoules of the houſe of Iacob, that
entred into Ægypt, were
%%% !!! This note is part of the previous one.
%\LNote{}{}
ſeuentie. \V And he ſent Iudas before him to Ioſeph, that he ſhould tel
him, and he ſhould come into Geſſen to
%%% 0160
meete him. \V Whither when he was come, Ioſeph addreſſing his chariot
went vp to mete his father, vnto the ſame place: and ſeing him fel vpon
his neck, and as they embraced he wept. \V And his father ſaid to
Ioſeph: Now wil I die with ioy, becauſe I haue ſeene thy face, and do
leaue thee aliue. \V But he ſpake to his brethren, and to al his fathers
houſe: I wil goe vp, and wil tel Pharao, and wil ſay to him: My
brethren, and my fathers houſe, that were in the Land of Chanaan, are
come to me: \V and the men are paſtours of ſheepe, and their trade is to
feede flockes: their cattel, and heardes, and al that they could haue,
they haue brought with them. \V And when he ſhal cal you, and ſhal ſay:
What is your trade? \V You ſhal anſwer: We thy ſeruantes are paſtours,
from our infancie vntil this preſent, both we and our fathers. And this
you ſhal ſay, that you may dwel in the Land of Geſſen, becauſe
\Fix{he}{the}{obvious typo, fixed in other}
Ægyptians
\SNote{Ægyptians honoring ſhepe, goates, and kyne for goddes, deteſted
them that did gouerne, kil, or eate thoſe cattel.}
deteſt al paſtours of ſheepe.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
