%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 36
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-36


\project douay-rheims


%%% 0132
%%% o-0123
\startChapter[
  title={Chapter 36}
  ]

\Summary{Eſau with his wiues and children parteth from Iacob. 9.~His
  genealogie is recited, with their habitations.}

And
\LNote{The generations of Eſau.}{As before
\CNote{chp.~4,~10, 12,~25.}
Moyſes deſcribed the
genealogies of Cain, of Iapheth and Cham, of Nachor, of Iſmael, and
other ſonnes of Abraham, ſo here he recordeth an other collateral
progenie of Eſau, that the difference and diſtinction of them, and the ſelected
people of God might be more conſpicuous, becauſe contraries oppoſed are
ſene more clerly.
\MNote{By compariſon of interupted companies the cõtinual ſucceſſion of
the Church is more glorious.}
And ſo the Churches ſucceſſion and perpetual light, compared with the
interrupted and obſcure companies, ſhineth the brighter. For albeit in
thoſe other generations there might be manie faithful and iuſt perſons,
among the infidels and wicked, and of ſome we are aſſured (as of Lot and
Iob) yet faith and religion decayed, and was extinguiſhed in their
carnal children, and only continued in the right line from Adam to
Iacob, whoſe twelue ſonnes were fathers and beginners of twelue Tribes,
and in them the ſame true Religion was ſtil conſerued and publikly
profeſſed, as in the onlie knowne viſible Church, til the cõming of
Chriſt, as S.~Auguſtin clerly ſheweth in his excellent worke of the
\Cite{Citie of God}:
eſpecially in the
\Cite{15.~and 16.~bookes},
in manie chapters.} 
theſe are the generations of Eſau, the ſame is Edom. \V Eſau tooke wiues
of the daughters of Chanaan:
\LNote{Ada the daughter of Elon.}{In the
\MNote{One place of Scripture ſeemeth contrarie to an other but is not.}
\XRef{26.~chap. (v.~34.)}
Eſaus two wiues, which he tooke in Chanaan are called Iudith the
daughter of Beeri the Hetheite, and Baſemath the daughter of Elon of the
ſame place, and here the ſame two wiues are named Ada the daughter of
Elon the Hetheite, and Oolibama the daughter of Sebeon the
Heueite. VVhich neither agree in names nor countire. Againe his third
wife Iſmaels daughter, here named Baſemath, in
\XRef{28.~chap. v.~9.}
is called Maheleth. For reconciliation of which and other like
difficulties, or ſeeming contradictions, albeit
\Fix{he }{}{obvious typo, fixed in other}%
lerned expoſitors ſay, that either theſe perſons had diuers names, or
one was true and natural father, an other legal, or adoptiue, for there
were ſuch alſo before the law of Moyſes, as appeareth in the hiſtorie of
Thamar:
\MNote{Holie Scriptures not eaſie to be vnderſtood.}
yet it were hard to geue a determinate ſolution of this
difficultie. VVhich example, with manie others (by vs omitted in theſe
briefe annotations) conuince the Proteſtants preſumptuous error, holding
that Scriptures are eaſie to be vnderſtood.}
Ada the daughter of Elon the Hetheite, and Oolibama the daughter of Ana
daughter of Sebeon the Heuite: \V Baſemath alſo the daughter of
Iſmael ſiſter of Nabaioth. \V And Ada bare Eliphaz: Baſemath bare
Rahuel:
%%% 0133
\V Oolibama bare Iehus and Ihelon and Coree, theſe are the ſonnes of Eſau,
that were borne to him in the land of Chanaan. \V And
\SNote{The ſeparation of Eſau from Iacob.}
Eſau tooke his wiues and ſonnes and daughters, and euerie ſoule of his
houſe, and his ſubſtance, and catle, and al that he could haue in the
land of Chanaan: and he went into an other countrie, and
%%% !!! This annotation has two referances in the text. How should
%%% this be handled?
\LNote{Departed from his brother.}{Here is an other difficultie (though
not ſo intricate as the former) how
\MNote{Eſau laſt parting from Iacob.}
Eſau now parted into Mount Seir,
ſeing he dwelled there, when his brother Iacob came from Meſopotamia.
\XRef{chap.~22. v.~3.}
VVhich S.~Auguſtin
\Cite{(q.~119. in Gen.)}
ſolueth ſaying: Eſau firſt dwelt in Seir after he was diſapointed of his
fathers bleſſing, but dwelt againe with his father, after Iacobs returne
from Meſopotamia, and now went to Seir againe after his fathers death.}
departed from his brother Iacob. \V For they were exceding rich, and
could not dwel together: neither was the land of their peregrination
able to beare them, for the multitude of flockes. \V And Eſau
%%% !!! this annotation mark goes with the annotation previous.
%\LNote{}{}
dwelt in Mount Seir, he is Edom. \V And theſe are the generations of
Eſau the father of Edom in mount Seir, \V and theſe are the names of his
ſonnes: Eliphaz the ſonne of Ada the wife of Eſau: Rahuel alſo the ſonne
of Baſemath his wife. \V And Eliphas had ſonnes: Theman, Omar, Sepho,
and Gathan, and Cenes. \V And Thamna was the concubine of Eliphas the
ſonne of Eſau: which bare to him Amalech. Theſe are the ſonnes of Ada
the wife of Eſau. \V And the ſonnes of Rahuel: were Nahath & Zara, Samma
and Meza. Theſe were the ſonnes of Baſemath the wife of Eſau. \V Theſe
alſo were the ſonnes of Oolibama, the daughter of Ana, the daughter of
Sebeon, the wife of Eſau, which ſhe bare to him, Iehus, and Ihelon, and
Coree. \V Theſe were dukes of the ſonnes of Eſau: the ſonnes of Eliphaz
the firſt-begotten of Eſau: duke Theman, duke Omar, duke Sepho, duke
Cenes, \V duke Coree, duke Gatham, duke Amalech, theſe are the ſonnes of
Eliphaz in the land of Edom, and theſe are the ſonnes of Ada. \V Theſe
alſo were the
%%% o-0124
ſonnes of Rahuel, the ſonne of Eſau: duke Nahath, duke Zara, duke Zamma,
duke Meza, and theſe
\Fix{are be}{be}{obvious typo, fixed in other}
the dukes of Rahuel, in the Land of Edom: theſe be the ſonnes of
Baſemath the wife of Eſau. \V And theſe were the ſonnes of
Oolibama the wife of Eſau: duke Iehus, duke Ihelon,
\Fix{euke}{duke}{obvious typo, fixed in other}
Coree, theſe be the dukes of Oolibama, the daughter of Ana, and wife of
Eſau. \V Theſe are the ſonnes of Eſau, and theſe are the dukes of them:
the ſame is Edom. \V Theſe are the ſonnes of Seir the horreite, the
inhabiters of the land: Lotan, and Sobal, and Sebeon, and Ana, \V and
Diſon, and Eſer, and Diſan. Theſe are dukes of the Horreite, the ſonnes
of Seir in the Land of Edom. \V And Lotan had ſonnes: Hori and Heman,
and the ſiſter of Lotan, was Thamna. \V And theſe were the
%%% 0134
ſonnes of Sobal: Aluan and Manahat, and Ebal, and Sepho and Onam. \V And
theſe were the ſonnes of Sebeon: Aia and Ana. This is Ana that found the
hot waters in the wildernes, when he fed the aſſes of Sebeon his
father: \V and he had a ſonne Diſon, and a daughter Oolibama. \V And
theſe were the ſonnes of Diſon: Hamdan, and Eſaban, and Iethram, and
Charan. \V Theſe alſo were the ſonnes of Eſer: Balaan, and Zauan, and
Acan. \V And Diſan had ſonnes: Hus, and Aram. \V Theſe were dukes of the
Horreites: duke Loran, duke Sobal, duke Sebeon, duke Ana, \V duke Diſon,
duke Eſer, duke Diſan: theſe were the dukes of
the Horreites that ruled in the Land Seir. \V And the Kinges that ruled
in the land of Edom, before that the children of Iſrael had a king, were
theſe: \V Bela the ſonne of Beor, and the name of his citie Denaba. \V
And Bela died, and
\SNote{By the common opinion of Latin and Greke fathers, this was holie
Iob as we ſhal diſcuſſe when we come to his booke againſt the hebrew
doctors, who ſay Iob was of Nachors race.}
Iobab the ſonne of Zara of Boſra reigned in his ſteed. \V And when Iobab
was dead, Huſam of the land of the Themans reigned in his ſteed. \V He
alſo being dead, there reigned in his ſteed Adad the ſonne of Balad,
that ſtroke Madian in the countrie of Moab: and the name of his citie
was Auith. \V And when Adad was dead, there reigned for him Semla of
Maſreca. \V He alſo being dead, Saul of the riuer Rohoboth, reigned in
his ſteed. \V And when he alſo was dead, Balanan the ſonne of Achobor
ſucceeded into the kingdome. \V This man alſo being dead Adar reigned in
his place, and the name of his citie was Phau: and his wife was called
Mectabel, the daughter of Matred, daughter of Mezaab. \V Theſe therfore
be the names of the dukes of Eſau in their kinreds, and places, and
callings: duke Thamna, duke Alua, duke Ietheth, \V duke Oolibama, duke
Ela, duke Phinon, \V duke Cenez, duke Theman, duke Mabſer, \V duke
Magdiel, duke Hiram: theſe are the dukes of Edom dwelling in the land of
their empire, the ſame is Eſau the father of the Idumeians.



\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
