%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 30
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-30


\project douay-rheims


%%% 0117
%%% o-0109
\startChapter[
  title={Chapter 30}
  ]

\Summary{Rachel yet barren, deliuereth her handmaide to Iacob, who
  beareth two ſonnes. 9.~Lia ceaſing to beare geueth her handmaid alſo,
  and she beareth two more. 17.~Then Lia beareth other two ſonnes and
  one daughter. 22.~Rachel beareth Ioſeph. 25.~Iacob deſirous to returne
  home, is hyred to ſtay for a certaine part of the ſtockes
  increaſe. 43.~Wherby he becometh exceeding rich.}

And Rachel ſeing ſhe was vnfruitful,
\SNote{Not 
\Fix{propely}{properly}{obvious typo, fixed in other}
enuie, but griefe & lawful emulation.
\Cite{S.~Aug. li.~22. c.~54. cont. Fauſt.}}
enuied her ſiſter, and ſaid to her huſband: Geue me children, otherwiſe I
ſhal dye. \V With whom Iacob being angrie anſwered: Am I as God, who
hath depriued thee of the fruite of thy wombe? \V But ſhe ſaid: I haue
here my ſeruant Bala: Companie with her, that she may beare vpon my
knees,
%%% o-0110
and I may haue children of her. \V And she gaue him Bala vnto
\SNote{Of pluralitie of wiues ſee
%%% !!! Fix this citation
\XRef{pag.~xxx}}
mariage: who, \V when her huſband had companied with her, conceaued and
bare a ſonne. \V And Rachel ſaid: Our Lord hath iudged for me, and hath
heard my voice, geuing
%%% 0118
me a ſonne, and therfore she called his name, Dan. \V And againe Bala
conceauing bare an other, \V for whom Rachel ſaid: God hath compared me
with my ſiſter, and I haue preuailed: and she called him Nepthali. \V
Lia perceauing that she had left bearing, deliuered Zelpha her handmaid
to her huſband. \V Who conceauing and bringing forth a ſonne, \V she
ſaid: Happely. And therfore called his name Gad. \V Zelpha alſo bare an
other. \V And Lia ſaid: This is for my bleſſednes: for wemen wil cal me
bleſſed. Therfore she called him, Aſer. \V And Ruben going forth in the
time of wheat harueſt into the field, found mandragores: which he
brought to his mother Lia. And Rachel ſaid: Geue me part of thy ſonnes
\SNote{Holie Scripture (ſaith S.~Auguſtine) would neuer haue mentioned
ſuch womanlie deſires, but to admoniſh vs to ſeke great miſteries
therin.
\Cite{li.~22. c.~56. cont. Fauſtum.}}
mandragores. \V She anſwered: Doeſt thou thinke it a ſmal matter, that
thou haſt taken my huſband from me, vnleſſe thou take alſo my ſonnes
mandragores? Rachel ſaid: For thy ſonnes mandragores let him ſleepe with
thee this night. \V And when Iacob returned at euen from the field, Lia
went out to meete him, and ſaid: Companie with me, becauſe with wages I
haue hired thee for my ſonnes mandragores. And he ſlept with her that
night. \V And God heard her prayers: and she conceaued and bare the
fifth ſonne, \V and ſaid: God hath geuen me a reward, becauſe I gaue my
handmaid to my huſband. And she called his name Iſſachar. \V Againe Lia
conceauing, bare the ſixt ſonne, \V and ſaid: God hath endowed me with a
good dowrie: this turne alſo my huſband wil be with me, for becauſe I
haue borne him ſix ſonnes, and therfore ſhe called his name, Zabulon. \V
After whom ſhe bare a daughter, name Dina. \V Our Lord alſo remembring
Rachel, heard her, and opened her wombe. \V Who conceaued, and bare a
ſonne, ſaying: God hath taken away my reproch. \V And she called his
name, Ioſeph, ſaying: Our Lord adde to me an other ſonne. \V And when
Ioſeph was borne, Iacob ſaid to his father in lawe: Diſmiſſe me that I
may returne into my countrie, and to my land. \V Geue me my wiues, and
my children, for whom I haue ſerued thee, that I may depart: thou
knoweſt the ſeruice that I haue ſerued thee. \V Laban ſaid to him: Let
me finde grace in thy ſight: I haue learned by experience, that God hath
bleſſed me for thy ſake: \V appoint thee wages which I shal geue
thee. \V But he anſwered: Thou knoweſt how I
%%% 0119
haue ſerued thee, and how great thy poſſeſſion hath benne in my
hands. \V Thou hadeſt a ſmal thing before I came to thee, and now thou
art made rich: and our Lord hath bleſſed thee at my comming in. It is
reaſon therfore that once I prouide alſo for mine owne houſe. \V And
Laban ſaid: What ſhal I geue thee? But he ſaid: I wil nothing: but if
thou wilt doe that which I demande, I wil fede, and kepe thy sheepe
again. \V Goe round about al thy flockes, and ſeparate al the ſhepe
%%% o-0111
of diuers colours, of ſpeckled flyſe: and whatſoeuer shal be ruſſet and
ſpotted, and of diuers colours, aſwel in the shepe, as in the goates,
shal be my wages. \V And my iuſtice shal anſwer for me to morowe, before
thee when the time of the bargaine shal come: and al that shal not be of
diuers colours, and ſpotted, and ruſſet, aſwel in the shepe as in the
goates, shal accuſe me of theft. \V And Laban ſaid: I like wel that thou
demandeſt. \V And he ſeparated the ſame day the shee goates, and the
shepe, and the he goates, and the rammes of diuers colours, and ſpotted:
and al the flocke of one coloure, that is of white and blacke flyſe, he
deliuered in the hand of his ſonnes. \V And he put a ſpace of three
dayes iourney betwixt him and his ſonne in lawe, who fed the reſt of his
flocke. \V Iacob therfore
\SNote{Iacob did iuſtly vſe this meanes to recouer that which Laban
withheld frõ him, being due for the dowrie of his wiues, and recompence
for his ſeruice.
\Cite{Rupert. li.~7. c.~39. in Gen.}}
taking grene roddes of the poplare, and of the almond, and of the plaine
trees, in part pilled them: and when the barkes were taken of, in the
parts that were pilled, there appeared whitnes: but the parts that were
whole, remayned grene: and by this meanes the colour was made diuers. \V
And he put them in the troughes, where the water was poured out: that
when the flockes should come to drinke, they might haue the roddes
before their eyes, and in the ſight of them conceaue. \V And it came to
paſſe that in the verie heate of the ramming, the shepe beheld the
roddes, and brought forth ſpotted, and of diuers colours, and
ſpeckled. \V And Iacob diuided the flocke, and put the roddes in the
troughes before the eyes of the rammes: and al the white and the blacke
were Labans: and the reſt, Iacobs, when the flockes were ſeparated one
from the other. \V Therfore when the ewes went to ramme, in the prime
time, Iacob put the roddes in the troughes of water before the eyes of
the rammes, and of the ewes, that in looking vpon them they might
conceaue: \V but when the
%%% 0120
later comming was, and the laſt conceauing, he did not put them. And
thoſe that were late warde, became Labans: and they of the prime time,
Iacobs. \V And the man was enriched beyond meaſure, and he had manie
flockes, wemen ſeruantes and men ſeruants, camels and aſſes.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
