%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 26
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-26


\project douay-rheims


%%% 0105
%%% o-0098
\startChapter[
  title={Chapter 25}
  ]

\Summary{Iſaac by reaſon of famine goeth into Gerara, 3.~where God
  reneweth to him the promiſes made to Abraham. 9.~Abimelech blameth him
  for calling his wife his ſiſter. 15.~The people enuying his wealth,
  quareleth for his welles. 26.~At laſt Abimelech maketh league with
  Iſaac.}


And when a famine was riſen in the land, after that ſterilitie, that had
chanced in the dayes of Abraham, Iſaac went to Abimelech king of the
Paleſtines into Gerara. \V And our Lord appeared to him, and ſaid:
\SNote{God by Abrahams exãplar life inuited the Ægyptians to true
religion: now commandeth Iſaac to ſtay in Gerara to the like end.
\Cite{S.~Theod. q.~76. in Gen.}}
Goe not downe into Egypt, but reſt in the land which I ſhal tel thee. \V
And ſeiourne in it, and I wil be with thee, and will bleſſe thee: for to
thee and thy ſeed, I wil geue al theſe countries, accompliſhing the oath
which I ſware to Abraham thy father.
%%% 0106
\V And I wil multiplie thy ſeed as the ſtarres of heauen: and I wil geue
to thy poſteritie al theſe countries: and in thy ſeed \Sc{Shal be
blessed
%%% o-0099
al the nations of the earth}, for becauſe Abraham obeyed my voice, and
kept my preceptes and commandements, and obſerued
\LNote{My ceremonies.}{Theſe
\SNote{External ceremonies in the law of nature.}
were not the ſame ceremonies and lawes
which were afterwards preſcribed by God, and deliuered by Moyſes, but
other obſeruances by which Abraham and other holie Patriarches before
him, ſerued God with certaine external worſhip, differing from the rites
of the Gentiles, eſpecially from Enos time
\XRef{Gen.~4.}
and ſo forward.}
my ceremonies & lawes. \V Therfore Iſaac abode in Gerara. \V Who when he
was asked by the men of that place, concerning his wife, anſwered: She
is
\SNote{See the Annot. to c.~12 v.~13.}
my ſiſter; for he was afraid to confeſſe that ſhe was married to him,
thinking leſt peraduenture that they would kil him becauſe of her
beautie. \V And when verie manie days were paſſed, and he abode there,
Abimelech the king of the Paleſtines looking forth through a windowe,
ſawe him ſporting with Rebecca his wife. \V And calling for him, he
ſaid: It is euident that ſhe is thy wife: why dideſt thou faine her to
be thy ſiſter? He anſwered: I feared leſt I ſhould die for her. \V And
Abimelech ſaid: Why haſt thou deceaued vs? ſome man of the people might
haue lyen with thy wife, & thou haddeſt brought vpon vs
\SNote{Adulterie a great ſinne alſo among Painims.}
a great ſinne. And he commanded al the people, ſaying: \V He that ſhal
touch this mans wife, dying ſhal dye. \V And Iſaac ſowed in that land,
and he found that ſame year
\LNote{An hundredfold}{For
\SNote{Chriſtian fortitude preuaileth more by ſuffering, then by
forcible reſiſting.}
this increaſe of wealth the King and people at firſt enuied Iſaac, but
afterwards perceiuing that God almightie, whom he ſerued, ſo bleſſed
him, the reſt of the land remaining barren, they ſought to make league
with him
\CNote{(v.~28.)}
Euen ſo the Kings and nations of the world, firſt enuying and
perſecuting Chriſts Church, at leingth ſeing it ſtil proſperous became
with al humilitie children of the ſame Church, and ſeruants of Chriſt,
being ouercome not by force of armes but by patience, & peacable
endeuours of thoſe whom they moſt hated. VVherof excellently ſaith
S.~Leo
\Cite{(Ser.~1. in Natali. Apoſt.)}
Although Rome renowmed by manie victories, dilated her Empyre by land
and by ſea, yet was it leſſe, that martial trauel ſubdued, then that
which Chriſtian peace hath obtained. The Biſhops of Rome hauing larger
Iuriſdiction ſpiritual, then euer the Roman
\Fix{Cœſars}{Cæſars}{likely typo, same in both}
had temporal Dominion.}
an hundred fold: and our Lord bleſſed him. \V And the man was made rich,
and he went proſpering and encreaſing, til he was made exceeding
great: \V and he had alſo poſſeſſions of ſheep and of heards, and a
verie great familie. For this the Paleſtines enuying him, \V ſtopped at
that time al the welles, that the ſeruants of his father Abraham had
digged; filling them vp with water: \V in ſo much that Abimelech
himſelfe ſaid to Iſaac: Depart from vs, becauſe thou art become mightier
then we a great deale. \V And departing, to come to the Torrent of
Gerara, and to dwel there: \V againe he digged other welles, which the
ſeruants of his father Abraham had digged, and which after his death,
the Philiſtines had ſtopped vp of old: and the called them by the ſame
names, which his father before had called them. \V And they digged in
the
\SNote{The chanel where ſometimes a vehement ſtreame rũneth, ſometimes
none at al.}
Torrent, and found liuing water: \V but there alſo the paſtors of Gerara
made a brawle againſt the paſtors of Iſaac, ſaying: It is our water. For
which cauſe he called the name of the wel, by occaſion of that which had
hapned,
\TNote{VVrangling.}
Calumne. \V And they digged alſo an other: & for that they brawled
likewiſe, and he called the name of it, Enmitie. Going
%%% 0107
foreward from thence he digged an other wel, for which they contended
not: therfore he called the name therof, Latitude, ſaying: Now hath the
Loed dilated vs, and made vs to encreaſe vpon the earth. \V And he went
vp from that place vnto Berſabee, \V where our Lord appeared to him that
ſame night, ſaying: I am the God of Abraham thy father, do not feare,
becauſe I am with thee I wil bleſſe thee, and multiplie thy ſeed for my
ſeruant Abrahams ſake. \V Therfore he builded there an altar: and hauing
called vpon the name of our Lord, he pitched his tent: and commanded his
ſeruants that they ſhould digge a wel. \V To the which place when there
were come from Gerara Abimelech, and Ocazath his freind, and Phicol
chieffe captaine of his ſouldiers, \V Iſaac ſpake to them: Why are you
come to me a
%%% o-0100
man whom you hated, and haue thruſt from you? \V Who anſwered: We ſaw
that the Lord is with thee, and therfore we ſaid: Let there be an oath
betwen vs, and
\SNote{So nations of the world firſt enuyed the Church of Chriſt, but
after made peace with it.}
let vs make a league, \V that thou du vs no harme, as we alſo haue
touched nothing of thine, neither haue we done that which might hurt
thee: but with peace haue we diſmiſt thee encreaſed with the bleſſing of
the Lord. \V Therfore he made them a feaſt, and after they had eaten and
drunken \V ariſing in the morning, they ſware one to an other: and Iſaac
diſmiſſed them peaceably into their place. \V And behold the ſame day
came the ſeruants of Iſaac telling him of a wel, which they had digged,
and ſaying: We haue found water. \V Wherupon he called it Abundance: and
the name of the citie was geuen Berſabee, euen vnto this preſent day. \V
But Eſau being fourtie yeares old married wiues, Iudith the daughter of
Beeri the Hethite, and Baſemath the daughter of Elon of the ſame
place: \V both which had
\SNote{Eſau by marying againſt his parents wil, made breach from them.}
offended the mind of Iſaac and Rebecca.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
