%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 20
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-20


\project douay-rheims


%%% 0091
%%% o-0086
\startChapter[
  title={Chapter 20}
  ]

\Summary{Abraham ſeiorning in Geraris, his wife is taken into King
  Abimelechs houſe, but by Gods commandement is reſtored vntouched,
  14.~with great giftes, 17.~and Abraham praying Abimelechs houſe is
  cured.}

Abraham remoued from thence into the ſouth countrie, and dwelt betwene
Cades, and Sur: and he liued as a pilgrime in Gerara. \V And he ſaid of
Sara his wife: She is
\SNote{See the Annot. to c.~12 v.~13.}
my ſiſter. Abimelech therfore the king of Gerara ſent, and tooke her. \V
And God came to Abimelech in a dreame by night, and ſaid to him: Loe
thou ſhalt dye for the woman that thou haſt taken: for ſhe hath a
huſband. \V But Abimelech had not touched her, and he ſaid: Lord wilt
thou kil a nation that is ignorant and iuſt? \V Did not he ſay to me:
She is my ſiſter: and ſhe ſay, He is my brother? in ſimplicitie of my
hart, and cleanenes of my hands haue I done this. \V And God ſaid to
him: And I do know that thou dideſt it with a ſincere hart: and therfore
I kept thee that thou ſhouldeſt not ſinne againſt me, and I permitted
not that thou ſhouldeſt touch her. \V Now therfore reſtore the wife to
her huſband, becauſe he is a prophet: and he ſhal pray for thee, and
thou ſhalt liue: but if thou wilt not reſtore her, know thou that dying thou
ſhalt dye, thou and al things that are thine. \V And Abimelech forthwith
riſing vp in the night, called al his ſeruantes: and he ſpake al theſe
words to their eares, and al the men were ſore afraid. \V And Abimelech
%%% 0092
called alſo for Abraham, and ſaid to him: What haſt thou done to vs?
what haue we offended againſt thee, that thou haſt brought vpon me and
vpon my kingdom a great ſinne? that which thou oughteſt not to doe, thou
haſt done to vs. \V And again expoſtulating, he ſaid: What ſaweſt thou,
that thou dideſt this? \V Abraham anſwered: I thought with my ſelfe,
ſaying: Perhaps there is not the feare of God in this place: and they wil
kil me for my wife: \V howbeit otherwiſe alſo in verie deed ſhe is my
ſiſter,
\SNote{Not his fathers owne daughter, but of his progenie.
\Cite{S.~Aug. li.~22. cont. Fauſt. c.~35.}
&
\Cite{li.~16. c.~19. &~30. de ciuit.}}
the daughter of my father, and not the daughter of my mother, and I
tooke her to wife. \V And after that God brought me out of my fathers
houſe, I ſaid to her: This mercie thou ſhalt doe with me: In euerie
place, which we shal come vnto, thou ſhalt ſay that I am thy brother. \V
Therfore Abimelech
\SNote{Kinges of the earth eſteming the Church only for a laudable people of
God, ſeeke to ſubiect her to them ſelues: but knowing her to be the
inuiolable ſpouſe of Chriſt, ſubiect them ſelues to her, and offer to
her moſt honorable giftes.
\Cite{S.~Aug. li.~22. cont. Fauſt. c.~38.}}
tooke ſhepe & oxen, and ſeruants, and handmayds, and gaue to Abraham:
and reſtored to him Sara his wife, \V and ſaid: The land is before you,
dwel wherſoeuer it ſhal pleaſe thee. \V And to Sara he ſaid: Behold I
haue geuen thy brother a thouſand pecces of ſiluer, this ſhal
%%% o-0087
ſerue thee for a veile of thine eyes to al that are with thee, and
whither ſoeuer thou shalt goe, 
\Fix{remenber}{remember}{possible typo, fixed in other}
alſo thou waſt taken. \V And Abraham praying, God healed Abimelech and
his wife, and his handmaids, and they bare children: for our Lord had
cloſed vp euerie matrice of the houſe of Abimelech for Sara Abrahams
wife.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
