%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% Chapter 49
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-49


\project douay-rheims


%%% 0167
%%% o-0154
\startChapter[
  title={Chapter 49}
  ]

\Summary{Iacob replenished with the ſpirite of prophecie, a litle before
  his death, fortelleth his ſonnes manie things, that shal happen to
  their poſteritie. Chargeth ſome of them with faultes paſt, bleſſeth
  euerie one. 29.~Appointeth where to burie him, 32.~and dyeth.}

And Iacob called his ſonnes, and ſaid to them: Come together, that I
\SNote{Theſe are predictiõs, not al bleſſings.
\Cite{S.~Amb. li. de Benedict Patriar.}}
may ſhew you the things that ſhal come to you in the laſt dayes. \V Come
together, and heare you ſonnes of Iacob, heare ye Iſrael your father:

\V
%%% !!! Unmarked Note, where does it belong?
\SNote{That theſe are moſt profound Myſteries, is eaſie to conceiue, but
moſt hard to vnderſtand them. In ſome the Patriarch recounteth things
paſt in his life, fortelling the effects therof to come. Other things he
forſheweth pertayning to the diuiſion of the Land of Chanaan, others to the
times of the Iudges, of the Kings, of the Captiuitie, of Deliuerie from
thence, of Chriſt, of Antichriſt, and of the end of this world. Of al
which diuers anciẽt fathers haue written large cõmentaries, & godlie
treatiſes.}
Ruben my firſt begotten, thou art my ſtrength, and the begining of my
ſorow: former in giftes, greater in empyre. \V Thou art poured out as
water,
\SNote{A prophecie not an imprecation.
\Cite{S.~Aug. li.~16. c.~33. cont. Fauſt.}}
encreaſe thou not:
\LNote{Becauſe thou dideſt aſcend thy fathers bed.}{For
\MNote{Ruben for his ſinne was put from the prerogatiues of the firſt
borne.}
this crime of inceſt Ruben was depriued of his firſt-birth-right. VVho
being by order of birth \Emph{former in guiftes}, wherby he ſhould haue
had duble portion, \Emph{and greater in Empire}, wherby he ſhould haue
bene Prince of Lord ouer his bretheren, the former prerogatiue was geuen
to Ioſeph, whoſe two ſonnes were heades of two Tribes, the other was
geuen to the Tribe of Iuda, in Dauid and his poſteritie. He was alſo
depriued of his prerogatiue in Prieſthood, which was after annexed to
the Tribe of Leui, wherupon the Chaldee paraphraſis ſpeaketh thus to
Ruben: It belonged to thee to haue receaued three better
\Fix{lettes}{letters}{likely typo, fixed in other}
then thy brethren, Prieſthood, Beſt portion, and the Kingdome. But
becauſe thou haſt ſinned, the duble portion is geuen to Ioſeph, the
Kingdome to Iudas, and Prieſthood to Leui.}
becauſe thou diddeſt aſcend thy fathers bed, and diddeſt defile his
coutch.

%%% 0168
\V Simeon and Leui brethren:
\LNote{Veſelles of iniquitie.}{Albeit
\CNote{Iudith.~9.}
\MNote{The zeale of Simeon & Leui was cõmendable: but their maner of
reuenge, was manie wayes faultie.}
Simeon and Leui were moued with iuſt zeale to puniſh the foule crime
committed by Sichem, againſt their ſiſter and whole familie: yet in
their maner of reuenging were manie ſinnes worthely condemned by Iacob,
both immediatly after the fact, and here at his death. For before the
ſlaughter they committed there greate ſinnes; in that they raſhly did it
vnknowne to their father therby putting him & themſelues in extreme
danger, if God had not maruelouſly protected them; in falſly pretending
agreement and league with the Sichamites, which they ment not to
performe; and in ſacrilegiouſly abuſing the Sacrament of Circumciſion,
making it a cloke to deceiue their ennemies. In the fact alſo they
committed other foure groſſe crimes: cruelly killing thoſe, that offered
other abundant ſatisfaction; murthering others that were altogether
innocent; ſacking and deſtroying the citie, and carying away wemen
and children captiue.
\MNote{The Prieſts & Scribes furie, obſtinacie, & hard harts againſt
Chriſt.}
Myſtically S.~Ambroſe, Ruffinus, Iſidorus, and others vnderſtand this
prophecie, of the Scribes and the Prieſtes, deſcending of Simeon and
Leui, who were moſt eagre againſt our Sauiour, as himſelf more plainely
fortold, ſaing:
\CNote{Mat.~20.}
\Emph{The ſonne of man shal be betrayed to the chiefe
Prieſtes, and to the Scribes, and they shal condemne him to death.}
Their furie was curſed aboue al furie, \Emph{Becauſe} (as Iacob here
ſaith) \Emph{it was ſtubborne}, or obſtinate, for
\CNote{Ioan.~11.}
they did not only
condemne Chriſt to death in their wicked councel, but alſo vrged and
preſſed Pilat, endeuoring to ſaue him, & ſturred vp the people to
crye:
\CNote{Ioan.~18.}
\Emph{Take him avvay, Crucifie him.} Yea their \Emph{indignation}
was ſo \Emph{hard}, that they preferred Barabbas before Chriſt.}
veſſels of iniquitie warring. \V Into their counſel come not my ſoule,
and in their congregation be not my glorie: becauſe in their furie they
ſlew a man, and in their wilfulnes they vndermined a wall. \V Curſed be
their furie, becauſe it is ſtubborne: and their indignation, becauſe it
is hard: I wil diuide them in Iacob, and wil diſperſe them in Iſrael.

\V Iudas, thee thy brethren shal praiſe: thy hand shal be in the neck of
thyne enemies: thy fathers children shal adore thee. \V A lions whelp
Iudas: to the pray my ſonne thou didſt aſcend: taking thy reſt thou
didſt lye as a lion, and as it were a lyoneſſe, who shal raiſe him
vp? \V
\LNote{The ſcepter shal not be taken avvay.}{Here
\MNote{Taking away of the regal ſcepter from the Iewes a figure of
Chriſts coming.}
the Patriarch Iacob fortelleth the time, when the promiſed Meſſias
ſhould come into the world, by this ſigne that the ſcepter ſhould not be
taken from Iuda, til the ſame Redeemer of mankind were at hand. Not
that the regal ſcepter should remaine in the Tribe of Iuda, from Iacobs
death til Chriſtes comming: for that Kingdome beganne firſt in Dauid,
aboue ſix hundred yeares after Iacobs death, and after the captiuitie of
Babylon the highe Prieſts of the Tribe of Leui did gouerne alſo the
ſtate, & not only the Church, other ſix hundred yeares. But the ſenſe
is, that of the Tribe of Iuda ſhould riſe moſt gloriouſe Kinges, whoſe
crowne and Kingdome ſhould remaine with the Iewiſh Nation, vntil their
expected Meſſias ſhould drawe nere, and then be taken from them by the
Gentiles. VVhich was downe by Herod, whoſe father was an Idumean, his
mother an Arabique. Thus the Ancient fathers with one accorde vnderſtood
this prophecie. 
\Cite{S.~Iuſtinus Martyr. Ser. cum Triphone.}
\Cite{Euſebius Ceſarien. lib. Hiſt. Eccles. cap. 6.}
\Cite{Athanaſius. lib de Incarnat.}
\Cite{S.~Ambroſe lib. de Benedict. Patriarch c. 4.}
\Cite{S.~Chryſiſtom. Ho. 67. in Gen.}
\Cite{S.~Auguſtin lib. 18. c. 45. de ciuit.}
\Cite{Theodoretus. q. vlt. in Gen.}}
\Sc{The scepter shal not be taken away from Ivdas, and a dvke ovt of his
thigh, til he doe come that is to be sent, and the same shal be the
expectation of the gentiles.} \V Tying to the vineyard his colt, and to
the vine, o my ſonne, his the aſſe.
\LNote{He shal vvash his ſtole in vvyne.}{By
\MNote{Our Lords real preſence in the B.~Sacrament prophecied.}
wine, and bloud of the grape, what other thing is ſhewed (ſaith
S.~Cyprian
%%% !!! Does this really go here?
\CNote{li.~2. Ep.~3.}
\Cite{Epiſt. ad Cœcil.~63.})
but the wine of the Chalice of our Lordes bloud? Likwiſe Tertullian
\Cite{(lib.~4. contra Marcionem)}
expoundeth the ſtole to ſignifie Chriſts fleſh, and the wine his
bloud. In al which booke his drift is to ſhewe that Chriſt did not
deſtroy the olde Teſtament, but fulfilled the figures and prophecies
therof. And not that Chriſt gaue his bodie in figure only, as our
aduerſaries alleadge him.}
He ſhal waſh his ſtole in wine, and in the bloud of the grape his
cloke. \V His eyes are more beautiful then wine, and his teeth whiter
then milke.

\V Zabulon shal dwel in the shore of the ſea, and in the road of shippes
reaching as farre as Sidon.

\V Iſſachar a ſtrong aſſe lying at reſt betwen the borders. \V He ſaw
reſt that it was good: and the earth that it was very good: and he put
vnder his shoulder to cary, and became ſeruing vnder tributes.

\V Dan shal iudge his owne people as alſo an other tribe in Iſrael. \V
Be Dan
\LNote{Be Dan a ſnake in the vvay.}{This
\MNote{Ancient Fathers ſuppoſe that Antichriſt ſhal be of the tribe of
Dan.}
prophecie moſt ancient Fathers vnderſtood of AntiChriſt, namely 
\Cite{S.~Irenæus, lib. 5. aduerſ. Hæreſ.}
\Cite{S.~Hyppolitus Martyr Orat. de condummat. ſæculi.}
\Cite{S.~Ambroſe. c. 7. de Benedict. Patriarch.}
\Cite{S.~Auguſtin. q. 12. in Ioſue.}
\Cite{Proſper, lib. de promiſi. & Pradict. Dei. P. 4.}
\Cite{Theodoret. q. vlt. in Gen.}
\Cite{S.~Gregorie, lib. 30. Moral. c. 18.}
and many others vpon the
\XRef{7.~chap of the Apocalips},
where they ſuppoſe S.~Ioan did omitt Dan from amongſt the Elect of the
Iſraelitical Tribes, in deteſtation of Antichriſt, to be borne of that
Tribe.
\CNote{Ioan.~5.}
And certayne it is, that the Iewes wil receiue, and folowe him
for their Meſſias, as our Sauiour himſelfe ſaith. VVhich maketh it very
probable, that he ſhal be a Iewe borne, elſe they would not ſo eaſily
admitt him.}
a ſnake in the way, a ſerpent in the path, byting the hooffes of the
horſe, that his ryder may fal backward. \V \Sc{I wil expect thy
salvation O Lord.}

%%% o-0155
\V Gad, the gyrded shal fight before him: and himſelf shal be gyrded
backward.

\V Aſer, his bread is fat, and he shal geue daynties to kinges.

\V Nepthali, a hart let forth, & geuing ſpeaches of beautie.

\V
\LNote{Ioſeph a childe encreaſing.}{Ioſeph
\MNote{Ioſeph in manie things prefigured Chriſt.}
was in manie reſpectes a figure of Chriſt, eſpecially in that he was
loued of his father before al his bretheren, ſolde by his brethren to
the Gentiles of enuie, and for money, aduanced to dignitie and
authoritie, the deliuerer of Ægypt from famine, and called Sauiour of
the world; al performed in Chriſt, the true Childe encreaſing.}
Ioſeph a child encreaſing, encreaſing and comelie to behold: the
daughters courſed to and
\Fix{for}{fro}{obvious typo, fixed in other}
vpon the wall. \V But the dart men did exaſperate him, & brawled, and
enuied him. \V His bowe ſate vpon the ſtrong, & the bands of his armes
and his handes were diſſolued, by the hands of the mightie of Iacob:
thence came forth a paſtour, the ſtone of Iſrael. \V The God of thy
father shal be thy helper, &
%%% 0169
the Almightie shal bleſſe thee with the bleſſinges of heauen from aboue,
with the bleſſinges of the deapth, that lieth beneath, with the
bleſſings of the pappes and of the wombe. \V The bleſſinges of thy
father were ſtrengthned with the bleſſings of his fathers: vntil the
deſire of the eternal hilles came: be they vpon the head of Ioſeph, &
vpon the crowne of the Nazarite among his brethren.

\V Beniamin
\SNote{This prophecie S.~Auguſtin vnderſtandeth of S.~Paul, of the tribe
of Beniamin; who was firſt a perſecutor, and after an Apoſtle of
Chriſt.
\Cite{Ser.~14. de Sanctis.}}
a rauening woolfe, in the morning shal eate the pray, and in the euening
shal diuide the ſpoile.

\V Al theſe in the tribes of Iſrael twelue: theſe things ſpake their
father to them, and he bleſſed euerie one, with their proper
bleſſings. \V And he commanded them, ſaying: I am geathered vnto my
people: burie you me with my fathers in the duble caue, which is in the
field of Ephron the Hethite, \V againſt Mambre in the Land of Chanaan,
which Abraham bought with the field of Ephron the Hethite for a
poſſeſſion to burie in. \V There they buried him, & Sara his wife: there
was Iſaac buried with Rebecca his wife: there alſo Lia doth lie
buried. \V And when he had finished the precepts wherwith he inſtructed
his ſonnes, he plucked vp his feet vpon the bed, and died: & he was put
vnto his people.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
