%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% First Age
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent first-age


\project douay-rheims


%%% 0049
%%% o-0045
\startArgument[
  title={\Sc{A briefe remonstrance of the state of the Chvrch, and face
  of Religion, in the first age of the world. From the creation to Noes
  floud: the space of 1656.~yeares.}},
  marking={The First Age}
  ]

Here according to our purpoſe mentioned before, we wil briefly recite
certaine principal points of Religion, taught and obſerued in the firſt
age. In which the foundations of the true maner of ſeruing God (that
should be continued to the end of the world) were laid, and proſpered in
ſome, as appeareth in theſe ſeuen firſt chapters of Geneſis. But firſt
of al, we shal in two words repete (as it is clerly geathered in the
ſame holie Scripture) the ſtate of man before, and immediatly after his
fall, being the ſubiect to whom al this pertayneth.

After
\MNote{Man made to Gods image, and in happie ſtate.}
therfore that God had created other things, both in heauen and earth,
laſt of al he made Man, to his owne image and likenes, with
vnderſtanding and freewil, therin like to Angels, and ſuperior to al
other creatures, and ſo made him Lord and maiſter of al earthlie
things. Neither were theſe the greateſt benefits which God beſtowed on
man: for his diuine goodnes indued alſo this his reaſonable creature,
with innocencie & original iuſtice, whereby al things were moſt rightly
ordered within him, and about him.
\MNote{Man obeyed God, and al earthlie creatures obeyed man.}
His mind, wil, and reaſon were obedient to God; his ſenſes & inferior
part of his ſoule were ſubiect to reaſon; his flesh and bodie obeyed
the ſpirite; and al earthlie creatures obeyed him. God alſo adorned man
with excellent knowledge, both natural and ſupernatural. And albeit his
bodie was of corruptible ſubſtance, yet the ſame, and al his poſteritie,
if they had not ſinned, should haue benne conſerued, and without dying,
haue benne tranſlated to euerlaſting life.
\MNote{Man placed in Paradiſe.}
Thus man was placed in Paradiſe, and Eue there made of a ribbe of his
ſide, to be his mate and vnſeparable companion, as man and wife ioyned
in Mariage, with Gods bleſsing,
%%% o-0046
for increaſe and multiplication. As appeareth in the two firſt chapters
of this booke.

But
\CNote{Eccle.~7,~30. Sap.~2,~24.}
\MNote{Man fel by yelding to tentations.}
\Emph{God hauing made man right, he intangled him ſelfe} (as holie
Scripture ſpeaketh) \Emph{with infinite queſtions.} For the diuel
enuying mans felicitie inuegled our mother Eue with queſtions and lies,
and then by her, firſt ſeduced and deceiued, allured alſo Adam to the
tranſgreſsion of Gods commandment. And ſo they loſt original iuſtice,
which Adam had receiued for him ſelfe and al mankind: and al proceeding
from them by natural
%%% 0050
propagation are borne the children of wrath,
\MNote{Original ſinne.}
in original ſinne contracted from Adam, ſlaues of the diuel, not only
ſubiect to temporal death, but alſo are excluded for euer from heauenly
bliſse and glorie: except by Chriſts redemption particularly applied,
they be reſtored to grace & iuſtice in this life.

And
\MNote{Adam and Eue, were penitent.}
touching Adam and Eue, whoſe ſinne was not original but actual, directly
committed by them ſelues, Gods mercie ſo reclamed them by new grace,
that they deſpared not (as Cain, and ſome others did afterwards) but
with hope of remiſsion were ſorie and penitent, and accordingly receiued
penance, and redemption.
\CNote{Sap.~10.}
For God \Emph{brought Adam from his ſinne} (as
holie writte teſtifieth) and the ſame is collected of Eue, God shewing
the like ſignes of his prouident mercie towards them both, of which we
shal by and by note ſome for example.

Now let vs ſee the more principal points of faith and Religion profeſſed
and obſerued by the Church of God before Noes floud.
\MNote{Faith in one God.}
Firſt they beleeued in one Eternal and Omnipotent God, who made the
whole world and al things therin of nothing. Which is eaſely confeſſed
of al that are not plaine Atheiſts, and may be proued againſt them by
reaſon. And therfore Adam and other Patriarches could not erre in this
Article, nor others be ignorant therof, except they were very wicked.

The
\MNote{The bleſſed Trinitie.}
Myſterie alſo of the Bleſſed Trinitie, three Diuine Perſons in one God,
though farre aboue the reach of mans reaſon, yet was beleued more
expreſly by ſome, more implied by others, and conſerued from age to age
by tradition, at leaſt amongſt the chiefe heades and leaders. Wherupon
Moyſes afterwardes inſinuated the ſame great Myſterie, by diuers wordes
and phraſes, writing of God and his workes. The two wordes \Emph{God
created} if they be rightly conſidered importe ſo much. For the
word \HH{Elohim}, \Emph{God}, in the plural number, ſignifyeth pluralitie
of Perſons (for manie Gods it can not ſignifie, ſeeing there is but one
God) and the verbe \HH{bara}, \Emph{created}, in the ſingular number
ſignifyeth one God in nature and ſubſtance, albeit three Perſons.
\MNote{Though the B.~Trinitie worke ioyntly in al creatures, yet diuers
workes are attributed to diſtinct Perſons.}
For whatſoeuer God doth in creatures, is the worke of the whole
Trinitie: though holie Scriptures do oftentimes appropriate ſome worke
to one Diuine Perſon, ſome to an other. Which alſo proueth diſtinction
of Perſons in God. So the wordes
\CNote{Gen.~1.}
\Emph{God created heauen and earth}
ſignifie the \Emph{Father}, to whom \Emph{powre} is attributed. \Emph{In
the beginning}, ſignifie \Emph{the Sonne}, to whom \Emph{wiſdome} is
appropriated, and the words, \Emph{The
\Fix{Sprite}{Spirite}{likely typo, fixed in other}
of God moued ouer the waters}, ſignifie \Emph{the Holie Ghoſt}, by whoſe
bountiful \Emph{goodnes}, the waters were made fruictful.
%%% o-0047
Likewiſe Gods owne wordes: \Emph{Let vs make man} ſignifie the
pluralitie of Perſons, and \Emph{Image and likenes} in the ſingular
number, ſignifie one God.

Men
\MNote{By faith the ſtate of man paſt & preſent was knowne.}
alſo knew by faith manie things perteyning to them ſelues. As that
\Fix{eht}{the}{obvious typo, fixed in other}
bodie was made of the ſlime of the earth: the ſoule not produced of anie
%%% 0051
thing formerly exiſting, but created immediatly of nothing and naturally
immortal: that the ſoule of Adam was indued with grace and iuſtice: that
he fel from that happie ſtate, by yelding to tentation, and breaking
Gods commandment of abſtinence: that for the ſame ſinne Adam and Eue
were caſt forth of Paradiſe, and al mankind ſubiect to death, and other
calamities.

For
\MNote{Beleefe in Chriſt to come.}
remedie againſt ſinne, & reſtauration to grace, they beleeued
in \Emph{Chriſt} promiſed to be borne of the womans ſeede, who by his
death should conquer the wicked ſerpent, deliuer man from captiuitie,
and reſtore him to ſpiritual life. And this is the cauſe of the
perpetual enmitie betwen the woman (eſpecially the moſt bleſſed Virgin
Mother, of whom Chriſt tooke flesh) and the ſerpent, and betwen her
ſeede, the ſpiritual children of Chriſt, and the ſerpents ſeede, the
whole companie of the wicked. Of this battle and conqueſt Targhum
Hieroſolimitanum thus ſpeaketh.
\MNote{Remedie for men but none for Angels that fel.
\XRef{Heb.~2,~16.}}
\Emph{There ſhal be remedie and health to the children of women, but
to thee, o ſerpent, there ſhal be no medicine, yea they ſhal tread thee
vnder their feete, in the latter dayes, by the powre of Chriſt their
King.} Likewiſe Gods familiar conuerſation with diuers men in mans shape
\XRef{(Gen.~2. 3. 4. 6. and~7.)}
was a ſigne of Chriſts incarnation. And the Sacrifices immolated did
prefigurate his death, in reſpect wherof it is ſaid in the Apocalips, 
\CNote{Apoc.~13,~3.}
\Emph{The Lambe was ſlaine from the
\Fix{beginniing}{beginning}{obvious typo, fixed in other}
of the world.} But more expreſly S.~Paul teſtifieth, that Abel, Enoch,
and Noe beleeued in Chriſt, naming them for example of the firſt age,
and others of other times, and in the end concludeth, that manie
more
\CNote{Heb.~11.}
\Emph{being approued by the ſame faith, receiued not the promiſe}
(to wit in their lifetime)
\MNote{None admitted into heauen before Chriſt.}
\Emph{God prouiding that they without others}
(of the new Law) \Emph{ſhould not be conſummate}, that is, not admitted
into heauenlie ioyes & fruition of God, vntil the way of eternal glorie
were opened by our Lords Paſsion and Aſcenſion.

Neither
\MNote{External Sacrifice.}
did the true ſeruants of God, in thoſe firſt dayes, only beleeue
in hart, but they alſo profeſsed their faith, & Religion by external
Rites,
namely in offering of Sacrifice (the moſt ſpecial homage & ſeruice to
God) which is clerly teſtified,
\XRef{cha.~4.}
as wel bloudie in figure of Chriſts Paſsion, as vnbloudie in figure of
the holie Euchariſt. Alſo the accepting of the one rightly offered by
Abel, & reiecting the other not donne ſincerly by Cain, was declared by
external ſignes, which Cain diſdayning and enuying his brothers good
worke, knowing his owne to be naught, of mere malice killed his
brother.

Beſides
\MNote{Publique prayer with other Rites.}
Sacrifice they had alſo other Rites in publique Aſſemblies, praying and
inuocating the name of our Lord, in more ſolemne maner, from Enos time
and ſo forward, according to that is recorded of him, in the
\XRef{end of the fourth chapter}.
For douteles Adam, Abel, and Seth did alſo pray and call vpon God, and
therfore it was ſome addition or increaſe of ſolemnitie in the ſeruice
of God, which is referred to Enos.

%%% 0052
%%% o-0048
They
\MNote{Ceremonial obſeruations.}
had moreouer other ceremonies:
\CNote{Gen.~2,~3.}
\MNote{Feaſtes.}
of the ſeuenth day particularly bleſſed and ſanctified by God, kept
holie by Adam and other Patriarches, as Abben Ezra witneſſeth in his
commentaries vpon the tenne commandements.
\CNote{Gen.~29.}
\MNote{Abſtinence.}
Of abſtayning from meates, for it ſemeth the more godlie ſorte did eate
no flesh, before the floud,
\CNote{Gen.~9.}
which was after permitted.
\CNote{Gen.~7,~2.}
\MNote{Cleane & vncleane.}
Obſeruation of cleane and vncleane beaſtes for Sacrifice.
\CNote{Gen.~4,~26.}
\MNote{Places dedicated to prayer.}
Of peculiar places dedicated to religious vſes where people mette
together to pray.
\MNote{Figures of Chriſts Sacraments.}
Likewiſe diuers other things in the firſt age were figures of Chriſts
Sacraments:
\CNote{Gen.~1.}
\MNote{Baptiſme.}
the Spirite of God geuing powre to the waters, (as Tertullian S.~Hierom
and others expound it) and the floud of Noe, by S.~Peters teſtimonie,
were figures of Baptiſme.
\CNote{1.~Pet.~5.}
\MNote{Mariage.}
Mariage inſtituted in Paradiſe, is the very paterne of holie Matrimonie,
a Sacrament in the Church of Chriſt, where one man and one wife are
onlie lawful, and not more at once in anie wiſe,
\CNote{Mat.~19.}
Chriſt reforming that
which in Moyſes law was tolerated (for hardnes of mens hartes, and for
auoyding murther, to put away one wife, and take an other) to this firſt
inſtitution as it was in the beginning,
\CNote{Gẽ. 2,~14.}
\Emph{two in one flesh}, not three nor more.
\MNote{Penance.}
The repentance of Adam and Eue was a perfect and examplare figure of the
Sacrament of Penance.
\CNote{Gen.~3.}
\MNote{Contrition.}
Firſt they were ashamed, couering their nakednes, and hiding them
ſelues, which shewed their griefe and ſorow for the ſinne committed.
\MNote{Confeſſion.}
Secondly they confeſsed their fault, and by what meanes it happened. For
God examining Adam, he anſwered truly and ſimply ſaing: \Emph{The woman
which thou gaueſt me, to be my companion, gaue me of the tree and I did
eate.} Likewiſe Eue confeſsed ſincerly, ſaying: \Emph{The ſerpent
deceiued me, and I did eate.}
\MNote{Satiſfaction.}
Thirdly God gaue them penance (beſides death before threatned and other
penalties annexed) that Eue \Emph{ſhould in paine and trauel bring forth
her children}; and \Emph{Adam ſhould eate his bread, in the ſweate of
his face.} And withal caſt them forth of Paradiſe. But not forth of his
fauoure, as appeared by his making them garments of skinnes, granting
them and their poſteritie, the reſt of the earth to liue and labour in,
eſpecially to ſerue him, and do penance,
\MNote{From hence is taken the ceremonie of ashes, on Ashweneſday.}
with admonition to remember,
that of duſt man was made, and into duſt he shal returne. Al which were
ſignes of loue, and that finally he would bring them, and manie more to
eternal ſaluation.

The
\MNote{Prieſthood.}
firſt borne and heades of families were Prieſts at the time of the
law of nature,
vntil the law being changed, God tooke Prieſts only of the ſtock of
Aaron, and the reſt of the Leuites to aſsiſt them in that
function,
\CNote{Num.~3,~10,~12,~45.}
\Emph{Aaron & his ſonnes thou ſhalt appoint, ſaith our Lord,
ouer the ſeruice of Prieſthood, for I haue taken the Leuites of the
children of Iſraël for euerie firſt borne.}
\MNote{Prieſthood & Law ſtand & change together.}
And S.~Paul teacheth, that changing of Prieſthood and changing of the
law goe alwayes together, shewing euidently that euerie lawful
communitie or commonwealth vnder God, hath external Prieſthood. So that
if there had benne no diſtinct order of external Prieſthood in the law
%%% 0053
of nature, or now were none in the law of grace
\CNote{Luther li. de abroganda Miſſa.}
(as Proteſtantes ſay
there is not) there were no law at al. See more of this point in the
\XRef{Annotations, chap.~7. ad Hebre.}
Here we only obſerue that Abel, Seth, Enos, and other Patriarches were
Prieſts, and exerciſed prieſtlie functions: yea Cain alſo was a Prieſt
(though a bad one) and offered Sacrifice.

%%% o-0049
But
\MNote{Good workes neceſſarie.}
external offices or miniſterie, without a wel diſpoſed mind, and ſincere
vertues producing \Emph{Good workes}, did neuer iuſtifie anie man. And
therfore
\CNote{Gen.~4.}
Cains Sacrifice, offered with a peruerſe mind, was not
reſpected by God, as Abels was: wherupon he becoming worſe, and more
malicious, God sharply reproued his anger and enuie, conceiued without
iuſt cauſe, ſaying: \Emph{If thou doeſt wel, ſhalt thou not receiue
againe: but if thou doeſt il, ſhal not thy ſinne forthwith be preſent at
the dore?} clerly shewing that euerie one shal receiue according to his
workes.

This
\MNote{Freewil.}
place alſo euidently sheweth \Emph{Freewil}, yea in a wicked man. For
this expoſtulation had neuer benne vttered, by our moſt reaſonable Lord,
and Maiſter, if Cain had benne depriued of freewil. For he might haue
excuſed himſelfe, and muſt needes haue benne holden excuſed, if he had
benne forced to do as he did. But God charged him as inexcuſable, and as
one that knew, or ought to know, that he had freewil. And doth further
inculcate, that he had, and should haue powre, and freewil ouer his
concupiſcence, to correct the ſame, if he would, ſaying: \Emph{The luſt
therof ſhal be vnder thee, and thou ſhalt haue
\Fix{domion}{dominion}{obvious typo, fixed in other}
ouer it.} So that no ſinner, be he neuer ſo wicked, much leſſe a iuſt
man, lacketh freewil. Yet
\CNote{Li. de ſeruo arbit.}
Luther abhorreth the very word, and
\CNote{li.~2. inſtit. ca.~2. para.~8.}
Caluin wisheth it out of the world.

\Emph{Temporal punishment}
\CNote{Gen.~3.}
\MNote{Temporal paine due for ſinne remitted.}
is proued to be due for ſinne remitted, by that both death,
and other penalties are inflicted, by Gods iuſtice vpon men, after
iuſtification, and by the particular punishments laid vpon Adam and Eue,
confeſsing their faultes.

\Emph{Purgatorie}
\MNote{Purgatorie.}
is alſo proued by the ſame iuſtice of God. For when anie dieth penitent,
and yet haue not made ful ſatisfaction, they muſt ſuffer for that
remaineth after death, and be purged, before they can enter into
reſt. Which remnant of debt our B.~Sauiour calleth
\CNote{Mat.~5.}
\Emph{The laſt
farthing}, and ſaith, it muſt be payed. The Iewes alſo at this day hold
the doctrin of Purgatorie by tradition. And conſequently
\MNote{Prayer for the dead.}
they \Emph{Pray for ſoules departed}, not only to God,
\MNote{And to Saincts.}
but alſo to the ancient Patriarches (which likewiſe
sheweth \Emph{Inuocation of Saincts}) in theſe wordes:
\CNote{Officio pro defunctis.}
\Emph{Yee fathers
which ſleepe in Hebron, open to him the gates of Eden.} That is of
Paradiſe, which was planted in Eden. And
\CNote{Ioſue.~14.}
Hebron is the place where Adam was buried, and
\MNote{Sepulchers of Patriarches religiouſly conſerued.}
his ſepulcher religiouſly conſerued in the time of Ioſue, aboue
1500.~yeares after his death.
\CNote{Gen.~23.}
The ſame is the place which
Abraham bought, and there buried Sara: where alſo himſelfe, and Iſaac,
and Iacob were buried: and to which finally
%%% 0054
the bodies of the twelue ſonnes of Iacob were tranſlated from Sichem. As
\CNote{li.~2. Antiquit.}
Ioſephus writeth. And Sichem alſo was ſpecially honored, becauſe ſuch
perſons had benne buried there, as
\CNote{Epiſt. ad Pamach.}
S.~Hierom witneſseth, of his owne
knowledge in his time.

Againe
\MNote{Enoch tranſlated aliue.}
by religious care of burying the dead in this firſt age, \Emph{Enoch}
was more certainly knowen to be \Emph{Tranſlated} aliue, and not to be
dead. For the
\CNote{Gen.~5.}
ſeuentie Interpreters, and
\CNote{Heb.~11.}
S.~Paul ſay \Emph{He was not
found}, which importeth that they ſought diligently for him, and that
his bodie could not be found, for God tranſlated him.

%%% o-0050
By
\MNote{Communion of Saincts.}
al which we ſee mutual offices, and communion of good workes amongſt
good men aliue and dead, which is called \Emph{Cõmunion of Saincts}.
\MNote{Miniſterie of Angels.}
And herein \Emph{Angels} lacked not their offices. For
\CNote{Gen.~3,~24.}
God ſet Cherubins
to kepe the gate of Paradiſe, that neither man should enter, being
iuſtly expelled for ſinne, nor diuels, as
\CNote{li.~11. de Gen. ad lit. c.~40.}
S.~Auguſtin noteth, leſt they
should take fruite of the tree of life, and geuing it to men, allure
them to more ſinne.
\MNote{Honour of Saincts.}
And now \Emph{Saincts} being exalted to Angels glorie, haue like
honorable offices towards other men, as Angels haue.
\CNote{Gen.~4.}
Yea the bloud of
Abel vniuſtly shed by Cain, and iuſtly to be reuenged by God, sheweth
the peculiar honour, which God beſtoweth vpon his Saints, for their
vertues and merites in this life, & eſpecially in their
death. For
\CNote{Pſal.~115.}
\Emph{Precious in the ſight of our Lord, is the death of his Saincts.}

Hence
\MNote{General Iudgement.}
alſo is proued, that ſeeing in this life the good are afflicted, and the
bad oftentimes proſper temporally, there muſt nedes be an
other \Emph{Court} of exact Iuſtice, and an other \Emph{Reaconing day},
wherin euerie one shal receiue, according as they haue donne good or
euil, which was ſufficiently intimated by Gods diſcuſsing, and
manifeſting Abels and Cains deſerts, which were hidden before, and in
part rewarding them accordingly, yet reſeruing the ful reward of the
one, and the punishment of the other to the next world.
\MNote{Iudge of the world.}
Of the \Emph{Iudge} and his ſentence Enoch (alleadged by S.~Iude the
Apoſtle) proficied clerly, ſaying:
\CNote{Epiſt. Iuda. v.~14.}
\Emph{Behold our Lord cometh in his
holie thouſands, to doe iudgement againſt al, and to reproue al the
impious, of al the workes of their impietie, wherby they haue donne
impiouſly, and of al the hard things which impious ſinners haue ſpoken
againſt him.} Thus holie Enoch preached touching the wicked, which
thought there was no Iudgement to come, nor Iudge to be feared!

At
\MNote{Reſurrection.}
this Iudgement al shal appeare in bodie and ſoule returning to life. For
that \Emph{Al men shal riſe} from death is proued, by the immortalitie
of mans ſoule, which God did not make nor produce of corruptible matter,
but immediatly
\CNote{Gen.~2,~7.}
\Emph{Breathed into his face the breath of life, and man
became a liuing ſoule.} So the ſoule being immortal, and hauing a
natural
%%% 0055
inclination to the bodie, mans natural perfection requireth the
coniunction of bodie and ſoule. For neither ſoule nor bodie ſeparated is
a man, but both ioyned in one ſubſiſtence are a man, in ſo much that
mankind should perish, except the bodies shal riſe againe, and liue with
the ſoules. And then shal the bodies be qualified according to the ſtate
of the ſoules, happie or miſerable for euer.

Of
\CNote{Gen.~5.}
\MNote{Euerlaſting life.}
\Emph{Eternal life} the tranſlation of Enoch is a figure. For ſeeing God
preſerueth his corruptible bodie ſo long, from death and infirmitie, it
is a token and manifeſt ſigne, that by the ſame powre of God, the bodies
of men shal at laſt day, after that al men are once dead, riſe againe, and
remaine with the ſoules for euer.
\MNote{The bleſſed in eternal ioy.}
The good in \Emph{Eternal ioy}:
\MNote{The wicked in endles paine.}
the wicked in \Emph{Eternal paine}. Both ſignified by the
\CNote{Gen.~3,~24.}
cuſtodie of
the gate of Paradiſe by Angels: who for euer kepe out thoſe, that are
ſtil defiled with ſinne, and ſo they depart into fire euerlaſting, and
%%% o-0051
admit the innocent and iuſt into the kingdome of heauen, which is
euerlaſting ioy and perfect felicitie.

Thus we ſee the face and briefe ſumme of Religion, in the beginning of
the world, til the floud:
\MNote{Church euer viſible.}
and the ſtate of the \Emph{Church}, which was alwayes \Emph{Viſible},
conſiſting of men good and bad,
\MNote{Succeſſion of Patriarches.}
with a continual \Emph{Succeſſion} of Rulers, as wel ſpiritual as
temporal. For the firſt borne were both Prieſtes and Princes in euerie
familie.
\MNote{One ſupreme head of the Church.}
And amongſt the ſame one euer chief of al. From which ranke Cain was
excluded, or rather excluded him ſelfe, by
\CNote{Gen.~4,~16.}
\Emph{Going forth from the
face of our Lord.} Wherupon holie Moyſes reciteth this
\CNote{Gen.~5.}
Monarchical
ſucceſsion of one chiefe, and \Emph{Supreme Head}, from \Emph{Adam} by
the line of \Emph{Seth, Enos, Cainan, Malaleel, Iared, Enoch, Mathuſala,
Lamech}, and \Emph{Noe}. Neuertheles he ſetteth downe alſo the
\CNote{Gen.~4,~17.}
progenie
of Cain, the firſt beginner of a worldlie, ſchiſmatical, and heretical
conuenticle, oppoſite to the Citie of God.
\CNote{S.~Aug. li. de paſtore. c.~8. &~20.}
He denied Gods prouidence (as
Thargum Hieroſolomitanum teſtifieth) proteſting to Abel,
\MNote{Cains negatiue doctrin.}
\Emph{That there was no Iuſtice nor Iudge, nor other world then
this, no reward for vertue, nor punishment for ſinne, and ſo deſperatly
he killed Abel.} Of theſe negatiue principles proceeded other like
deteſtable opinions, and moſt wicked life, ſauage and barbarous
crueltie, and al kind of impietie.
\MNote{True faith ſtil remained in manie.}
And in proceſse of time albeit manie remained in true faith, and vnitie
of the Church, yet by conuerſation with ſuch miſcreantes, especially by
occaſion of
\CNote{Gen.~6.}
Mariages betwen the faithful and infidels, almoſt the whole
world was corrupted in maners.
\MNote{Some alſo iuſt and perfect.}
But Noe was iuſt and perfect.
\MNote{Interruption of heretical Sinagogues.}
In punishment therefore of ſo great and
enormous ſinnes, God ſent the general floud, wherby al Cains progenie,
and al other infidels were wholly deſtroyed and extinguished, and the
true Church notibly purged; onlie iuſt Noe and his familie reſerued.
\MNote{Cõtinuance of the Church.}
By whom the ſame true Church was continued, and the world againe
replenished with men.


\stopArgument


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
