%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Genesis
%%%% second age
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent second-age


\project douay-rheims


%%% 0067
%%% o-0063
\startArgument[
  title={\Sc{The continvance of the Chvrch and Religion, in the Second
  age of the World.} From Noes floud to Abrahams going forth of his
  countrie. The ſpace of 368.~yeares.},
  marking={The Second Age}
  ]

No man can wel doubt nor wil denie, that the ſame Church continued al
the ſecond age which was in the firſt,
\MNote{Noe and Sem liued in Abrahams time.}
conſidering that Noe liued aboue 50.~yeares after the birth of Abraham,
and Sem 150.~more: and that theſe three, and ſome others of that time
are renowmed in holie Scripture, for ſincere profeſsors of true
Religion. But for more manifeſtation of their faith, and that the Church
was then very conſpicuous, we shal repete certaine
\MNote{Articles of Religion profeſſed in the ſecond age.}
principal pointes of
Religion profeſsed and practiſed al that time, by a continual knowne
viſible companie vnited in one myſtical bodie: though in the meane
while, the wicked ſprong and grew in number and worldlie force, much
oppreſsing yet neuer ſuppreſsing the good.

Firſt
\CNote{Gen.~8.}
iuſt Noe coming forth of the arke with his familie, profeſſed his
religious mind to
\MNote{One God.}
\Emph{One God} almightie, ſupreme Lord of al, by
\MNote{External Sacrifice.}
\Emph{Offering external} voluntarie, ſpeedie, pure, ſolemne, and
bountiful \Emph{Sacrifice} of Holocauſtes, as a
\MNote{Prieſthood.}
\Emph{Prieſt} vpon an
\CNote{Gen.~9.}
\MNote{Altares.}
\Emph{Altar.} After which moſt gratful office, God making a couenant
with him and his ſeede, neuer againe to deſtroy the world by water,
confirmed the ſame by the ſigne of the rainbow, which repreſented the
ſecond Perſon of the
\MNote{The B.~Trinitie.}
\Emph{B.~Trinitie}, the 
\MNote{Chriſt.}
\Emph{Sonne of God, Chriſt} our Lord to be borne \Emph{The Sõne of man},
& extended vpon the
\CNote{Heb.~11.}
\MNote{Croſſe.}
\Emph{Croſſe: in whom Noe beleeuing was inſtituted heyre of the iuſtice,
which is by faith in our
\MNote{Redemer.}
Redemer.}

VVe haue here againe
\CNote{Gen.~9.}
\MNote{Gods bleſſing operatiue.}
\Emph{Gods operatiue bleſſing}, with the effect of increaſe and
multiplication, the iſſue of Noe by his three ſonnes, in short time
making manie Nations. By the way alſo we haue an example of
\CNote{c.~9. v.~25.}
\MNote{Fathers bleſſing & curſing.}
\Emph{Fathers} ſolemne \Emph{Bleſſing} and \Emph{Curſing} their
children. The effectes wherof ſucceded afterwardes accordingly. Likwiſe
in this age was geuen a particular precept
\CNote{c.~9. v.~4.}
\MNote{Abſtinence from bloud.}
\Emph{Not to eate bloud}. And Noe obſerued diſtinction of
\CNote{c.~9. v.~20.}
\MNote{Cleane, and vncleane.}
\Emph{Cleane} beaſtes, offering Sacrifice in them only, as before the
floud, he was commanded to take moe of them into the arke, then of
the \Emph{Vncleane}.

%%% o-0064
In
\CNote{c.~10,~10.}
that ſo ample mention is made of ſinne and wickednes, there is no
doubt, but Noe, the
\CNote{c.~11.}
\MNote{Penance preached and inflicted.}
\Emph{preacher of iuſtice}, admonished and exhorted ſinners
to \Emph{Repentance}:
\CNote{2.~Pet.~2.}
yea he \Emph{Puniſhed} Cham & Chanaan, by his curſe in their poſteritie.
\CNote{c.~9,~5.}
And God him ſelfe \Emph{Threatned} to exacte
the bloud of man vniuſtly shed.
\CNote{c.~11.}
In the meane time \Emph{Inflicted} alſo
ſome \Emph{puniſhment} vpon the builders of Babel, by confounding their
tongues. And that by the
\MNote{Miniſterie of Angels.}
\Emph{Miniſterie of Angels.}

Which punishment in part, and threates of more importe a
\MNote{Reſurection.}
\Emph{General
%%% 0068
Reſurrection}, and
\MNote{Iudgement.}
\Emph{Iudgement}, where al things \Emph{shal be} exactly diſcuſſed
and \Emph{Iudged}. And then wil folow
\MNote{Eternal ioy, and paine.}
\Emph{Eternal life} to the good: and \Emph{Euerlaſting paine} to the
damned.

Al theſe points of Religion (and others mentioned in the former age, and
no doubt taught by Noe & his ſonnes) shew clerly a
\MNote{Church viſible.}
\Emph{Viſible Church}, conſiſting of
\MNote{Good and bad in the Church.}
good and bad.
\CNote{Cap.~8 &~9.}
\Emph{Noe} remaining the ſame man as before the floud;
\CNote{c.~10,~11.}
\Emph{Sem} and \Emph{Iapheth} are commended and bleſsed for wel
doing; \Emph{Cham} blamed, & curſed in his poſteritie: yet neither he
nor anie of his ſonnes or daughters fel into hereſie, or other
infidelitie, for anie thing that appeareth in Scripture, or other
authentical teſtimonie. \Emph{Heber alſo and his familie} are
particularly commended by Moyſes, as the right folowers, and the
ſpiritual children of Sem (who had innumerable other carnal children) as
thoſe that were innocent touching the preſumptuous building of others,
who for the ſame fault loſt their old tongue, which the familie of Heber
kept. As
\CNote{Ho.~30. in Gen.}
S.~Chriſoſtom and
\CNote{li.~16. c.~11. & li.~18. c.~19. ciuit.}
S.~Auguſtin do proue. Againe, diuers of this
familie falling afterwards by litle and litle to other nations, the
\MNote{Alwaies ſome good.}
\Emph{Familie of Thare}, ſaith the ſame S.~Auguſtin
\Cite{(li.~16. c.~12.)}
albeit not al, or not alwayes, yet euer ſome of them, and Abraham
continually, with Sem, Heber, Phaleg, & manie others, not mentioned by
Moyſes in his briefe deſcription (as
\CNote{L.~5. c.~35. Moral in 3.~Iob.}
S.~Gregorie doubteth not to
ſuppoſe) were iuſt, and kept the true faith, and vndefiled Religion.

But
\CNote{Ioſeph. li.~1. c.~4. Antiq.}
\MNote{Nemrod an Arch-heretike.}
\Emph{Nemrod} Chams nephew, and ſonne of Chus deſcribed for a valiant
hunter, a violent giant, and tyrant, was an Archheretike, a deuiſer and
teacher of falſe doctrin, againſt God and true faith. By ſutletie and
tyrannie he induced manie of liking or of feare to folow him, and ſo in
ſchiſme he maintained hereſie,
\MNote{His proud hereſie.}
\Emph{That men were not beholding to God, but to them ſelues, for
temporal proſperitie.}

VVherof begane a new & cruel confederacie, againſt the Citie of God, &
the ſecond great Sect of Infidels.
\MNote{The firſt Sect of Infidels was Barbariſme before the floud.}
\Emph{For Barbariſme} being the firſt, begun by Cain and ended by the
floud:
\MNote{The ſecond Scythiſme.}
\Emph{The ſecond mother} of al Sectes beginning after the floud (as
\CNote{Li. de hereſibus.}
S.~Epiphanius writeth) was \Emph{Scythiſme}: ſo called of the Scythians
a moſt cruel people. Who according to Nemrods hereſie (not thincking
themſelues beholding to God for temporal happines, but to their owne
forces) 
\MNote{Crueltie.}
tyrannized ouer the weaker, and manie wicked banning together extremly
oppreſsed the more peacable, eſpecially the Church and true ſeruants of
God. And this was one ſpecial cauſe of building Babylon, beſides their
ambicious deſire of perpetual fame, and their bearing the ſimple in hand
of a defence againſt a new floud, to make it in dede, a ſtrong hold for
tyrants to offend others, and to
%%% o-0065
defend themſelues. Wherfore God (who before deſtroyed al Infidels by the
floud) confounded theſe builders by diuiding their tongues, and ſo
forced them to breake, and part into manie countries.

%%% 0069
Thus
\MNote{Vntruthes are vnconſtant.}
mankind being diuided vpon the earth, opinions alſo were multiplied
concerning Religion. For shortly the perſwation of mens truſting in
themſelues, and in other mortal men appeared abſurde, euen the ſtrongeſt
feeling aduerſities, or fayling ſometimes of their purpoſes, ſaw there
was neede of ſupernal helpe, and that earthlie things depended much vpon
diuine wil and powre. But hauing forſaken God Almightie, the onlie maker
and conſeruer of al, they begane to imagine and ſerue falſe goddes, both
famous dead men, which had proſpered in this world, and diuers other
things, by which they receiued commoditie, or feared damage.

Hence
\MNote{The third Sect was Greciſme.}
therfore roſe the third principal Sect called \Emph{Greciſme}, beginning
alſo in 
this ſecond age, as the ſame S.~Epiphanius writeth. For Ninus the firſt
king of the Aſsirian great Monarchie, brought to paſſe that his
father
\MNote{Idolatrie.}
\Emph{Belus Iupiter} was eſtemed and worshipped for the onlie great God
by the Aſsirians. To him the Babylonians erected firſt Temples Altares
and Statues. Nemrod alſo by the name of
\TNote{\L{Satur deorum.}}
\Emph{Saturnus}, as the progenitor of Belus, and firſt great King or
Tyrant of Babylon, was accounted a god, and the father of goddes.
\SNote{Dinaſtæ were thoſe that raigned in Ægipt by force and policie:
& after were great goddes, and litle goddes.}
About this time likwiſe begane the Dinaſtæ among the Ægyptians, and not
ſowner, as they vainly brag to haue benne before the floud: yea much
longer then in deede the world hath benne. Moreouer the Chaldees
worshipped the fire. Others the ſunne, the moone, and innumerable other
feaned goddes. Againſt al which (and likwiſe againſt al hereſies) are
two ſpecial arguments.
\MNote{Idolatrie and Hereſie are cõfuted, by that they beginne
diſorderly; and are at diſſection in their imagined Religions.}
Firſt that they were not from the beginning, as the true God, and al
truthes are knowne and receiued by continual Tradition, but brought in
afterwards by men, and comonly by il men. Secondly they are not accepted
and eſteemed for goddes, or truthes, in al places, but with great
diuerſitie and diſsention, one ſort allowing that others deſpice, as
holie. Athanaſius notably writeth in his oration againſt Idols in theſe
wordes. \L{Quot ſunt gentes totidem
deorum genera confinguntur &c.} \Emph{How manie nations} (ſaith
he) \Emph{ſo manie kindes of goddes
are feaned. Alſo the ſame countrie, the ſame citie diſſenteth within it
ſelfe in ſuperſtition of Idols. The Phenicians certes acknowledge not
the Egiptians goddes, neither doe the Egiptians adore the ſame Idols
with the Phenicians: Nor the Scithians receiue the goddes of the
Perſians, nor the Perſians of the Scithians: The Pelaſgies refuſe the
Thraſian goddes, the Thraſians know not the Thebians. The Indians are
againſt the Arabians, the Arabians againſt the Æthiopians: and in like
ſorte the Æthiopians differ in their religious affaires from the
Arabians. The Syrians worship not the goddes of the Cilicians, and the
nations of Capadocia beſides al theſe haue goddes of other names. The
Bithinians alſo feaned
diuers goddes, the Armenians againe diuers from them. What nede we manie
wordes? Thoſe that are in the continent honour other goddes from the
Iland people. In briefe ech ditie and eech village not knowing the
goddes of their
%%% 0070
neighboures, ſetteth forth their owne, & eſtemeth them only in place of
goddes.} Thus farre
%%% o-0066
S.~Athanaſius.
\MNote{Luthers progenies differ as much in opinions of Religion, as
Painimes do in their falſe goddes.}
Name we like countries, prouinces,
cities, and townes in theſe partes of Europe, where Luthers ſcholars haue
ſet their feete, conſider the forme of Religion, and opinions which they
hold, and we shal ſee as vnorderlie beginnings, and as horrible
diſſentions in hereſies (which
\CNote{in c.~11. Oſee.}
S.~Hierom calleth the Idols of the New
Teſtament) as the ancient fathers haue diſcryed in Paganiſme. For
Lutherans or Proteſtants hauing no lawful generation, but proceding of
baſtards race, vpſtartes of vnknowne progenie, are no leſse at diſcordes
among them ſelues, only al agreing againſt Catholiques,
like \Emph{ſyncretiſantes} againſt their cõmon enimies, or Herod, Pilate,
& the Iewes againſt Chriſt.
\MNote{Sectes in England diuers from Luther, and eech one from the
reſt.}
And in England alone are diuers Sectes without poſsible meanes to agree
in one. For albeit the ciuil ſtate endeuoreth prudently and ſeriouſly
to bring al to vniformitie, at leaſt in publique shew, yet they are but
like manie faces vnder one hood, euerie ſorte keping their owne
opinions; yea almoſt euerie preacher and meane ſcholar (to ſay nothing
of artificers and common miniſters) arrogating to be his owne Iudge,
contemneth to ſtand to Luther or Caluin, to Geneua or Parlement, to
Conuocation or Synod of their owne, but to his owne only vnderſtanding,
and interpretation of holie Scripture.
\MNote{Proud and cõtentious ſpirites are hardly perſwaded to the truth.}
Nor yet to that alwayes; for when he is preſſed with that he once ſaid,
he wil forgete it, or eate his owne word, if he haue not written it, or
that you haue readie witnes againſt him, ſo hard it is to make a
deceiued Proteſtant or Puritan confeſse that he is conuinced, except by
very pregnant meanes you can firſt caſt out of him, or bind faſt the
ſpirite of preſumption, diſsention and contention:
\MNote{Vnlerned Catholiques beleue the ſame faith in al points with the
lerned.}
wheras the ſimplieſt Catholique in the world hath the ſelfe ſame faith
in al points, with the whole Church, in which he remaineth, and vpon
whoſe iudgement he dependeth.

To returne therfore, from whence we are not vnneceſsarily digreſsed, we
conclude with
\CNote{li.~16. c.~10. ciuit.}
S.~Auguſtin; VVhen Moyſes had shewed the beginning and
progreſſe of Nemrods earthlie citie, leauing it in Babylon, that
is \Emph{confuſion}, as needles to proſecute it further, he returneth to
declare the 
\Fix{pertual}{perpetual}{likely typo, fixed in other}
\Emph{ſucceſſion} of the Citie of God, the Church, as before the floud
from Adam to Noe, by the line of Seth, ſo after the floud from the ſame
Noe, by the line of
\MNote{The ſucceſſion of Patriarches from Noe to Abraham.}
\Emph{Sem, Arphaxad, Sale, Heber, Phalag, Reu,
Sarug, Nachor, Thare}, and \Emph{Abraham}. The reſt of Sems children,
and al the progenies of Iapheth and Cham, as not pertaining to this
purpoſe, omitted, ſo connecting thoſe in order of generations, by whom
the ſucceſsion is directly brought to
\MNote{Abraham a principal Patriarch.}
Abraham, Prince of the elected
people, a moſt ſpecial Patriarch, to whom new and great promiſses are
made of multiplication of his ſeede, and poſseſsion of the land of
Chanaan, but eſpecially of Chriſt our Redemer, and the ſame manie waies
confirmed, as wil appeare in the next age.


\stopArgument


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
