%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Epistles
%%%% Titus
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.




\startcomponent chapter-02


\project douay-rheims


%%% 2837
%%% o-2697
\startChapter[
  title={Chapter 2}
  ]

\Summary{What to preach both to old and yong (not only with word but
  with example alſo) and to ſeruants. 11.~For there are of al ſorts in
  the Church, and they muſt be inſtructed accordingly.}

But doe thou ſpeake the things that become ſound doctrine. \V Old men
that they be ſober,
\TNote{\G{σεμνούς} \L{pudicos}}
chaſt, wiſe, ſound in the faith, in loue, in patience. \V Old women in
like manner in holy attire, not il ſpeakers, not giuen to much wine:
teaching wel, \V that they may teach the yong women wiſedom, to loue
their huſbands, to loue their children, \V wiſe, chaſt, ſober, hauing a
care of the houſe, gentle, ſubiect to their huſbands, that the word of
God be not blaſphemed. \V Yõg men in like manner exhort that they be
ſober. \V In al things ſhew thy ſelf an example of good workes, in
doctrine, in integritie, in grauitie, \V the word ſound,
irreprehenſible: that he which is on the contrarie part, may be afraid,
hauing no euil to ſay of vs. \V
\CNote{\XRef{Eph.~6,~5.}
\XRef{Col.~3,~22.}
\XRef{1.~Pet.~2,~18.}}
Seruants to be ſubiect to their Maiſters, in al things pleaſing, not
gainſaying: \V not defrauding, but in al things ſhewing good faith, that
they may adorne the doctrine of our Sauiour God in al things.

\V For the grace of God our Sauiour hath appeared to al men: \V
inſtructing vs that denying impietie and worldly deſires, we liue
ſoberly, and iuſtly, and godly in this world, \V expecting the bleſſed
hope and
\TNote{\G{ἐπιφάνειαν}}
aduent of the glorie of the great God and our Sauiour \Sc{Iesvs} Chriſt,
\V who gaue himſelf for vs, that he might redeeme vs from al iniquitie,
and might cleanſe to himſelf a people acceptable, a purſuer of
%%% o-2698
good workes. \V Theſe things ſpeake, and exhort and rebuke
\SNote{Bishops muſt be ſtout & command in Gods cauſe and the people muſt
in no wiſe diſobey or contemne them.}
with al authoritie. Let no man contemne thee.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

