%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% New Testament
%%%% Titus
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Latin checked by KK.



\startcomponent chapter-01


\project douay-rheims


%%% 2835
%%% o-2695
\startChapter[
  title={Chapter 1}
  ]

\Summary{Of what qualitie the Prieſts and Bishops muſt be: 9.~namely
  learned, conſidering the Iudaical ſeducers of that time. 12.~That the
  Cretenſians muſt be roughly vſed, to haue them continue ſound in
  faith.}

Paul the ſeruant of God, and an Apoſtle of \Sc{Iesvs} Chriſt according
to the faith of the elect of God and knowledge of the truth: which is
according to pietie. \V Into the hope of life euerlaſting, which he
promiſed that lieth not, God,
\CNote{2.~Tim.~1,~9.}
before the ſecular times: \V but hath manifeſted in due times his word
in preaching, which is committed to me according to the precept of our
Sauiour God: \V to Titus my beloued ſonne according to the common faith,
grace and peace from God the Father, and Chriſt \Sc{Iesvs} our Sauiour.

\V For this cauſe left I thee in Crete, that thou ſhouldeſt reforme the
things that are wanting, and ſhouldeſt
\LNote{Ordaine Prieſts.}{Though
\MNote{Prieſts muſt be conſecrated by Biſhops only.}
Prieſts or Biſhops may be nominated and elected by the Princes, people,
or Patrons of places, according to the vſe of the time and diuerſitie of
Countries and faſhions, yet they can not be ordered and conſecrated but
by a Biſhop who was himſelf rightly ordered or conſecrated before, as
this Titus was by S.~Paul. And here it ſeemeth that he did not only
conſecrate them whom the people had elected before, but himſelf alſo
made choiſe of the perſons, no mention being here made of any other
election popular.
\MNote{The popular election of the Clergie taken away.}
Which though it were long vſed in the primitiue Church, yet for diuers
cauſes and ſpecially for continual tumultes, partialities, and diſorders
which S.~Auguſtin much complaineth of in his time, was iuſtly taken
away, and other better meanes of their deſignement appointed. See
\Cite{Conc. Laodic. cap.~12.~13.}
S.~Auguſt.
\Cite{de adult. coniug. li.~2. c.~20.}
\Cite{Ep.~110.}
and
\Cite{Poſsid in vita Aug. c.~8.}

And
\MNote{The preeminence of a Bishop aboue a Prieſt.}
that the ordering of Prieſts or impoſition of hands to that purpoſe,
belongeth only to Biſhops, and to no inferiour Prieſts or other perſons,
it is plaine by the Apoſtolike practiſe ſet downe in the Scriptures,
namely in the Actes, and in the Epiſtles to Timothee and Titus. And
S.~Hierom, who ſeemeth ſomtimes to ſay that in the primitiue Church
there was no great difference betwixt a Biſhop and Prieſt, yet he euer
excepteth giuing holy Orders which preeminence he attributeth to Biſhops
only.
\Cite{ep.~83.}
as he doth alſo Confirming the Baptized by giuing them the holy Ghoſt
through impoſition of hand and holy Chriſme.
\Cite{Dial. cont. Lucifer. c.~4.}
\MNote{To put no difference between them is Aërius hereſie.}
Note alſo that Aërius was of old condemned of hereſie, for holding that
there was no difference betwixt a Prieſt and a Biſhop.
\Cite{Epiph. hær.~75.}
\Cite{Auguſt. hær.~53.}
\MNote{Heret. tranſlation.}
Note laſtly the fraudulent tranſlation of the Heretikes, alwaies turning
for \Emph{Prieſts} (which here is euident to be a calling of Order and
office) \Emph{Elders}, ſaying, \Emph{That thou ordaine Elders}. Which in
our vulgar tongue ſignifieth the age, and not the Office properly: and
al this for hatred of Prieſts.}
ordaine Prieſts by cities, as I alſo appointed thee: \V
\CNote{\XRef{1.~Tim.~3,~2.}}
If any be without crime, the huſband
\LNote{Of one wife.}{To
\MNote{\L{Bigami} excluded from holy Orders, and the cauſes thereof.}
that which is ſaid vpon the like words
\XRef{1.~Tim.~3.}
adde this teſtimonie of S.~Epiphanius
\Cite{li.~3. to.~1. cont hæreſes in fine.}
\Emph{Holy Prieſthood}, ſaith he, \Emph{for the moſt part proceedeth of
virgins: and if not of virgins, yet of them that liue a ſole or ſingle
life: but and if the ſingle and ſole perſons ſuffice not to the
Miniſterie, of ſuch as conteine from their wiues, or after once marrying
remaine widowes. For, him that hath been married twiſe, it is not lawful
to take to Prieſthood, &c.} If you liſt to ſee the cauſes why bigamie is
forbidden them that are to be Prieſts, and continencie required of the
Clergie, ſee the ſame Authour
\Cite{li.~2. to.~1. hæreſ.~59.}
S.~Ambroſe
\Cite{li.~1. Offic. c.~50.}
and
\Cite{vpon 1.~Tim.~3.}
S.~Auguſtin
\Cite{de bono Coniugal. c.~18.}
S.~Hierom
\Cite{ep.~50. c.~5. ad Pammachium,}
and
\Cite{againſt Iouinian li.~2. c.~19.}
S.~Leo
\Cite{Ep.~87.}
and other ancient Authours.

And
\MNote{The notable men of both Teſtaments, that liued continently from
wiues.}
if the ſtudious Reader peruſe al antiquitie, he ſhal find al notable
Biſhops and Prieſts of Gods Church to haue been ſingle, or continent
from their wiues, if any were married before they came to the
Clergie. So was S.~Paul, and exhorteth al men to the like
\XRef{1.~Cor.~7,~7.}
So were al the Apoſtles after they followed Chriſt, as S.~Hierom
witneſſeth, affirming that our Lord loued Iohn ſpecially for his
virginitie.
\Cite{Apol. ad Pammach. c.~8.}
&
\Cite{li.~1. cont Iouin. c.~4.}
S.~Ignatius
\Cite{ep.~6. ad Philadelph.}
ſaith of the ſaid Iohn, and of Timothee, Titus, Euodius, Clement, that
they liued and died in chaſtitie, reckning vp of the old Teſtament
diuers notable perſonages that did the ſame: as Elias, Ieſus Naue
(otherwiſe called Ioſue) Melchiſedech, Eliſæus, Hieremie, Iohn
Baptiſt. No man is ignorant that al the notable Fathers of the Greek and
Latin Church liued chaſt: Athanaſius, Baſil, Nazianzen, Chryſoſtom,
Cyprian, Hilarie, (who entred into holy Orders after his wiues death)
Ambroſe, Hierom, Auguſtine, Leo, Gregorie the Great. Certaine other
notable Fathers had once wiues, but no holy men euer vſed them; much
leſſe married after they were in holy Orders.
\MNote{Only the Proteſtãts complaine that they haue not the guift of
chaſtitie.}
A maruelous thing, that ſo many heretofore ſhould haue the guift of
chaſtitie then, and now ſo few, if the Proteſtants ſay true, that ſkarſe
one among them in our Age of al their ſectes, euen of their principal
Superintendents, hath had it.}
of one wife, hauing faithful children, not in the accuſation of riot, or
not obedient. \V For a Biſhop muſt be without crime, as the ſteward of
God: not proud, not angrie, not giuen to wine, no ſtriker, not couetous
of filthy lucre: \V but giuen to hoſpitalitie, gentle, ſober, iuſt,
holy, continent: \V embracing that faithful word which is according to
doctrine, that he may be able to exhort in ſound doctrine, and to
reproue them that gaineſay it.

\V For there be many diſobedient, vaine-ſpeakers, and ſeducers,
eſpecially they that are of the Circumciſion. \V Who
%%% o-2696
muſt be controuled. Who ſubuert whole houſes, teaching the things they
ought not, for filthie lucre. \V One of them ſaid, their owne Prophet,
%%% !!! CNote or SNote ?
\MNote{Epimenides.}
\Emph{The Cretenſians alwaies liers, naughtie beaſts, ſlouthful
bellies.} \V This teſtimonie is true. For the which cauſe rebuke them
sharply, that they may be ſound in the faith, \V not attending to Iewish
fables, and commandments of men, auerting themſelues from the truth.

\V \CNote{Rom.~14,~20.}
Al things are
\SNote{He ſpeaketh not of the Churches abſtaining from meates ſometimes,
which is not for any vncleãnes in the creatures, but for chaſtening
their bodies: but he meaneth the Iewiſh ſuperſtition, who now being
Chriſtians, would not ceaſe to put difference of cleane and
vncleane according to their old law. See S.~Auguſtin.
\Cite{Cont. Fauſt. li.~31. c.~4.}}
cleane to the cleane: but to the polluted and to infidels nothing is
cleane: but polluted are both their mind and conſcience. \V They
confeſſe that they know God: but in their workes they deny, whereas they
be abominable and incredulous and to euery good worke reprobate.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:

