%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Esther
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 1060
%%% o-0955
\startChapter[
  title={Chapter 3}
  ]

\Summary{Aman aduanced by the king, is much offended that Mardocheus
  doth not adore him, 6.~and therfore procureth the kings decree, to
  deſtroy the whole nation of Iewes.}

After
\MNote{\Large D}
theſe things king Aſſuerus aduanced Aman the ſonne of Amadathi, which
was of the ſtocke of Agag: & he put his throne aboue al the princes
which he had. \V And al the kings ſeruantes, that were in the doores of
his pallace, bowed their knees, and adored Aman: for ſo the emperour had
commanded them,
\SNote{Aman exacted ſuch honour as heathen people gaue to their leſſe
goddes, which was alſo idolatrie, for Mardocheus was willing to giue him
ciuil honour, \Emph{Euen to kiſse the ſteppes of his feete, but feared}
(and refuſed) \Emph{to geue Godlie honour to him.}
\XRef{ch.~13. v.~13.~14.}
\Cite{S.~Tho. 2.~2. q.~84. a.~1.}}
onlie Mardocheus did not bowe his knee, nor adore him. \V To whom the
kings ſeruantes, that were chiefe at the doores of the pallace, ſaid:
Why doeſt thou aboue the reſt not obſerue the kings commandment? \V And
when they ſaid this often, and he would not heare, they told Aman,
deſirous to know whether he would perſeuere in his ſentence: for he had
told them that he was a Iew. \V Which when Aman had heard, and had
proued by experience that
%%% 1061
Mardocheus bowed not his knee vnto him, nor adored him, he was angrie
excedingly. \V And he counted it a matter of nothing to lay his handes
vpon Mardocheus alone: for he had heard that he was of the nation of the
Iewes, and he would rather deſtroy al the nation of the Iewes, that
were in the kingdom of Aſſuerus. \V The firſt moneth (which is called
Niſan) in the twelfth yeare of the reigne of Aſſuerus, the lot was caſt
into a potte which in Hebrew is called Phur, before Aman,
\SNote{Such is the prepoſterous furie of rancour, to appoint the day of
executing his malice, before he had got the kings decree.}
in what day and what moneth the nation of the Iewes should be ſlayne:
and there came forth the twelfth moneth, which is called Adar. \V And
Aman ſaid to king Aſſuerus: There is a people diſperſed through al the
prouinces of thy kingdom, and ſeparated one from an other, vſing new
lawes and ceremonies, moreouer alſo contemning the kings ordinances. And
thou knoweſt very wel that it is not expedient for thy kingdom that they
waxe inſolent by libertie. \V If it pleaſe thee, decree that they may
perish, and I wil pay ten thouſand talentes to the cofferers of thy
treaſure. \V
\MNote{The ſecond part.

The Iewes danger to be maſſacred.}
The king therfore tooke the ring that he vſed, from his hand, and gaue
it to Aman the ſonne of Amadathi of the progenie of Agag, the enemie of
the Iewes, \V and he ſaid to him: The ſiluer which thou doeſt promiſe,
be it thine: but concerning the people doe that which pleaſeth thee. \V
And the kings ſcribes were called in the firſt moneth, Niſan, the
thirtenth day of the ſame moneth: and there was written, as Aman had
commanded, to al the kings lieuftenantes, and iudges of the prouinces,
and of diuers nations, as euerie nation could read,
%%% o-0956
and heare according to the varietie of languages, in the name of king
Aſſuerus: and the letters ſigned with his ring, \V were ſent by the
kings poſtes to al prouinces, that they should kil and deſtroy al the
Iewes, from boy vnto old man, children, and wemen, in one day, that is,
in the thirtenth of the twelth moneth, which is called Adar, and should
ſpoyle them of their
\MNote{{\Large E} ch.~13. v.~1.}
goodes.

\V And
\MNote{\Large F}
\SNote{The letter at large is in the
\XRef{ſeuen firſt verſes of the 13.~ch.}
of which this is the ſumme.}
the content of the letters was this, that al prouinces might know, &
prepare them ſelues againſt the day aforeſayd. \V The poſtes that were
ſent, made haſt to fulfil the kings cõmandment. And immediatly the edict
hong in Suſan, the king and Aman feaſting, and al the Iewes that were in
the citie, weeping.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
