%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Esther
%%%% Chapter 13
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-13


\project douay-rheims


%%% 1074
%%% o-0967
\startChapter[
  title={Chapter 13}
  ]

\Summary{A copie of the epiſtle which Aman ſent to al prouinces to
  deſtroy the Iewes. 8.~And Mardocheus his prayer for the people.}

And
\MNote{\Large E}
this was the copie of the letter. The moſt great king Artaxerxes from
India vnto Æthiopia, to the princes of an hundred and ſeuen and twentie
prouinces, and to the captaynes, that are ſubiect to his empire,
greeting. \V Wheras I reigned ouer manie nations, and had ſubdewed al
the world to my dominion, I would not abuſe the greatnes of my might,
but with clemencie and lenitie gouerne my ſubiectes, that paſſing their
life quietly without any terrour, they might enioy peace wiſhed of al
men. \V But when I demanded of my counſelers how this might be
accompliſhed,
\SNote{Great hurt to a king, that is ruled by one counſeler. Salomon
teacheth, that \Emph{VVhere are manie counſelers, cogitations are
confirmed}.
\XRef{Prou.~15. v.~22.}}
one that excelled the reſt in wiſedom and fidelitie, and was ſecond
after the king, Aman by name, \V told me that there was a people
diſperſed through the whole world, which vſed
%%% o-0968
new lawes, and doing againſt the cuſtomes of al Nations, contemned the
cõmandmentes of kings, and brake the concord of al nations by their
diſſention. \V Which when we had learned, ſeing one nation rebellious
againſt al kind of men to vſe peruerſe lawes, and to goe againſt our
commandmentes, and to diſturbe the peace and concord of the prouinces
ſubiect to vs, \V we haue commanded that whomſoeuer Aman ſhal ſhew, who
is chiefe ouer al the prouinces, and ſecond after the king, and whom we
honour in ſteed of a father, they with their wiues and children be
deſtroyed of their enemies, and that none haue pitie on them the
fourtenth day of the twelfth moneth Adar of this preſent yeare: \V that
the wicked men going downe to hel in one day, may reſtore the peace to
our empire, which they had
\MNote{{\Large F} ch.~3. v.~14.}
diſturbed.

%%% !!! Formatting!
Hitherto the copie of the letter.

That which foloweth I found written after that place, where it is read,
\CNote{\XRef{ch.~4. v.~17.}}
\Emph{And Mardocheus going forth, did al thinges, that Eſther had
commanded him.} Yet it in neither in the Hebrew, nor extant with any of
\SNote{Except the 72.}
the interpreters.

\V
\MNote{\Large G}
But Mardocheus beſought our Lord, mindeful of al his workes, \V and
ſaid: Lord Lord king omnipotent, for in thy dominion are al thinges ſet,
and there is none that can reſiſt thy wil, if thou determine to ſaue
Iſrael. \V Thou madſt heauen and earth, and whatſoeuer is conteyned in
the compaſſe of
%%% 1075
heauen. \V Thou art Lord of al, neither is there that can reſiſt thy
maieſtie. \V Thou vnderſtandeſt al things, and knoweſt that I haue not
done this for pride and contumelie, and any deſire of glorie, that I
adored not the proud Aman, \V (for gladly would I be readie for the
ſaluation of Iſrael to kiſſe euen the ſteppes of his feete,) \V but I
feared leſt I ſhould transferre the honour of my God to a man, and leſt
I ſhould adore any except my God. \V And now ô Lord king, the God of
Abraham haue mercie vpon thy people, becauſe our enemies wil deſtroy vs,
and extinguiſh thyne inheritance. \V Deſpiſe not thy portion, which thou
haſt redemed to thyſelf out of Ægypt. \V Heare my prayer, and be
propitious to thy lot and corde, and turne our mourning into ioy, that
liuing we may prayſe thy name ô Lord, and doe not ſhut the mouthes of
them that ſing to thee. \V Al Iſrael alſo with like mind and
ſupplication cried to our Lord, becauſe certayne death did hang ouer
them.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
