%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Esther
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 1065
%%% o-0959
\startChapter[
  title={Chapter 7}
  ]

\Summary{Eſther entertayning the king and Aman againe at a feaſt, is
  promiſed to obtaine, whatſoeuer she wil aske. 3.~She demandeth ſaftie
  for her ſelf, and al the people, againſt Amans crueltie, 6.~and Aman
  is hanged on the gallowes, which he had prepared for Mardocheus.}

The king therfore and Aman went in, to drinke with the queene. \V And
the king ſaid to her alſo the ſecond day, after he was warme with wine:
What is thy petition Eſther, that it may be geuen thee? and what wilt
thou haue done? although thou shalt aske the half part of my kingdome,
thou shalt obteyne. \V To whom she anſwered:
\SNote{After three dayes faſt of al the Iewes with prayers and other
workes of penance, Eſther aſked and obtained the ſaftie of the whole
people more precious to her then halfe of the kingdom.}
If I haue found grace in thyne eies ô king, and if it pleaſe thee,
%%% 1066
geue me my life for the which I make requeſt, and my people for the
which I beſech. \V For we are deliuered I and my people, to be
deſtroyed, murdered, and to periſh. And would God we were ſold for
bondmen and bondwemen: it were a tolerable euil, and mourning I would hold
my peace: but now it is our enemie, whoſe crueltie redoundeth vpon the
king. \V And king Aſſuerus anſwering, ſaid: Who is this, and
%%% o-0960
of what might, that he dare doe theſe things? \V And Eſther ſaid: It is
this Aman our aduerſarie and moſt wicked enemie. Which he hearing
\SNote{Horrour of a guiltie conſcience is commonly the firſt torment of a ſinner.
\Cite{S.~Chriſſocons~4. de Lagaro.}}
forthwith was aſtoniſhed, not enduring to beare the countenance of the
king and of the queene. \V But the king being wrath roſe vp, and from
the place of the banket went into the garden ſet with trees. Aman alſo
roſe vp, to intreate Eſther the queene for his life, for he vnderſtood
that there was euil prepared him of the king. \V Who when he was
returned out of the garden ſet with trees, and had entered into the
place of the banket, he found Aman to haue fallen vpon the bed, wherin
Eſther lay, and he ſaid: The queene alſo he wil force in my preſence, in
my houſe. Neither was the word yet paſſed from the kings mouth, and
immediatly they couered his face. \V And Harbona,
\SNote{VVhat litle truſt in falſe freindshipe, when bad men once fal
into diſgrace.}
one of the eunuches which ſtood wayting on the king, ſayd: Behold the
gibbet, which he had prepared for Mardocheus, that ſpake for the king,
ſtandeth in Amans houſe, hauing in height fiftie cubites. To whom the
king ſaid: Hang him vpon it. \V Aman therfore was hanged on the gibbet,
which he had prepared for Mardocheus: and the kings wrath ceaſed.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
