%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Esther
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 1062
%%% o-0956
\startChapter[
  title={Chapter 4}
  ]

\Summary{Al the Iewes lament their imminent danger. 5.~Mardocheus
  willeth Eſther to intreate the king for their ſaftie. 11.~She fearing
  to goe to the king contrarie to his law vncalled, 15.~yet they al
  faſting and praying for her, she doth it.}

VVhich
\MNote{The third part.

The deliuerie of the Iewes from danger.}
things when Mardocheus had heard, he rent his garmentes, and was clothed
with ſackcloth, ſprinkling ashes on his head: and in the ſtreat of the
middes of the citie he cried with a lowed voyce, shewing the anguish of
his mind, \V and with this wayling going euen to the doores of the
pallace: for it was not lawful for one clothed with ſackcloth to enter
the kings court. \V In al prouinces alſo, townes, and places, to which
the kings cruel decree was come, there was
\SNote{The firſt and beſt remedie in diſtreſſe is to do workes of
penance for ſinnes cõmitted.
\XRef{1.~Cor.~11. v.~31.}}
great mourning with the Iewes, faſting, howling, and weeping, manie
vſing ſackcloth and ashes for their couch. \V And Eſthers maides and
the eunuches went in, and told her. Which she hearing was aſtonyed: and she
ſent a garment, that the ſackcloth being taken away, they should put it
on him: which he would not take. \V And calling for Athach the eunuch,
whom the king had geuen her for a ſeruant to Mardocheus, commanded him
to goe and to learne of him, why he did this. \V And Athach going
forth went to Mardocheus ſtanding in the ſtreate of the citie, before
the palace doore: \V who told him al thinges that had chanced, how Aman
had promiſed, that he would bring ſiluer into the kings treaſures for the
ſlaughter of the Iewes. \V He gaue him alſo a copie of the edict which
hong in Suſan, that he should shew it to the queene, and should admonish
her to enter in to the king, and to intreate him for her people. \V Athach
returning told
\Fix{Eſter}{Eſther}{obvious typo, fixed in other}
al things, that Mardocheus had ſayd. \V Who anſwered him, and bad that
he should ſay to Mardocheus:
%%% o-0957
\V Al the kings ſeruantes, and al the prouinces that are vnder his
dominion, know, that whether man, or woman, not called, shal enter in to
the kings inner court, he muſt immediatly be ſlayne without al delay:
vnleſſe perhaps the king ſtretch forth a golden rod vnto him, for a
ſigne of clemencie, and ſo he may liue. I therfore how can I enter in to
the king, which now theſe thirtie daies haue not bene called vnto
him? \V Which when Mardocheus had heard, \V he ſent word to Eſther
agayne, ſaying: Thincke not that thou mayſt deliuer
%%% 1063
thy owne life only, becauſe thou art in the kings houſe, aboue al the
Iewes: \V for if thou wilt now hold thy peace,
\SNote{Great confidence in Gods helpe when mans helpe fayleth: and in
dede this meanes by mouing the kings hart to grant Eſthers petition was
Gods ſpecial worke.}
the Iewes shal be deliuered by an other occaſion: and thou, and thy
fathers houſe shal perish. And who knoweth whether thou cameſt to the
kingdom therfore, that in ſuch a time thou mightſt be readie? \V And
agayne Eſther ſent to Mardocheus in theſe wordes: \V Goe, and gather
together al the Iewes, whom thou shalt find in Suſan, and pray ye for
me. Eate ye not, and drinke not in three dayes, and three nightes: and I
with my handmaydes in like maner wil faſt, and then wil I goe in to the
king, doing agaynſt the law, not called, and deliuering my ſelf to death
and to peril. \V Mardocheus therfore went, and did al things that Eſther
had commanded
\MNote{{\Large G} ch.~15 v.~1.}
him.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
