%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Esther
%%%% Chapter 12
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-12


\project douay-rheims


%%% 1073
%%% o-0967
\startChapter[
  title={Chapter 12}
  ]

\Summary{The conſpiracie of two eunuches, detected by Mardocheus, is
  repeted, 6.~and Amans malice againſt him for the ſame.}

And
\MNote{\Large C}
he abode that time in the kings court with Bagatha and Thara the kings
eunuches, which were porters of the palace. \V And when he vnderſtood
their cogitations, and had throughly ſeene their cares he learned that
they went about to lay hands on king Artaxerxes, and he told the king
therof. \V Who hauing them both in examination, when they had
\Fix{confeſſsd}{confeſſed}{obvious typo, fixed in other}
he commanded them to be led to death. \V But the king wrote that which
was done, in the comentaries: and Mardocheus alſo committed the memorie
of the thing to writing. \V And the king commanded him, that he ſhould
abide in the court of the palace, geuing him giftes for the delation. \V
But Aman the ſonne of Amadathi the Bugeite was moſt glorious before the
king, and would hurt Mardocheus, and his people,
\SNote{Here it appeareth that Aman was a fauourer of traitors or perhaps
of the ſame conſpiracie.}
for the two eunuches of the king, which were put to
\MNote{{Large D} ch.~3. v.~1.}
death.

%%% !!! Formatting!
Hitherto the proeme. That which foloweth was ſet in that place, where it
is written in the
\Fix{volum:}{volume:}{likely typo, fixed in other}
\CNote{ch.~3. v.~3.}
\Emph{And should ſpoyle their goodes.} VVhich we found in the common
edition only.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
