%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Esther
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 1058
%%% o-0953
\startChapter[
  title={Chapter 2}
  ]

\Summary{Inquirie being made of the beſt and faireſt virgines, 5.~Eſther
  the vnknowne neece of Mardocheus a Iew is preferred, 18.~and made
  Queene in place of Vaſthi, a mariage feaſt made, and preſentes
  geuen. 21.~Mardocheus detecteth traitors, and his ſeruice is recorded
  in the regeſter.}

Theſe thinges ſo done, after the indignation of king Aſſuerus was
aſſwaged, he remembred Vaſthi, and what ſhe had done, or what ſhe had
ſuffered: \V and the kings ſeruantes and his miniſters ſaid: Let there
be maydens ſought for the king, virgins and beautiful, \V and let there
be ſent that may viewe through al prouinces beautiful maydens and
virgins: and let them bring them to the citie of Suſan, and deliuer them
into the houſe of wemen vnder the hand of Egeus the eunuch, who is
ouerſeer of the kings wemen: and let them receiue wemens ornamentes, and
other things neceſſarie to be vſed. \V And which ſo euer among al ſhal
pleaſe the kings eies, let her reigne for Vaſthi. The word pleaſed the
king: and ſo as they had ſuggeſted, he commanded to be done. \V There
was a man in the citie of Suſan a Iew, named Mardocheus, the ſonne of
Iair, the ſonne of Semei, the ſonne of Cis, of the ſtocke of Iemini, \V
who had beene tranſported from Ieruſalem the ſame time, that
\CNote{\XRef{4.~Reg.~24.}}
Nabuchodonoſor the king of Babylon tranſported Iechonias the king of
Iuda, \V who was the foſter father of his brothers daughter Ediſſa,
which by an other name was called Eſther: and ſhe had loſt both her
parentes: exceeding beautiful, and of comely face. And her
%%% 1059
father and mother being dead, Mardocheus adopted her for his
daughter. \V And when the kings commandment was bruited abrode, and
according to his commandement many fayre virgins were brought to Suſan,
and were deliuered to Egeus the eunuch:
\SNote{\XRef{Deut.~7. v.~8.}
Iſraelites are forbid to marie with Gentiles, yet when there was no
danger to be peruerted, and great hope of good, Eſther by Mardocheus
counſel conſented to marie king Aſſuerus, though otherwiſe she
\Emph{deteſted the bed of the incircumciſed, and of euerie ſtranger}.
\XRef{ch.~14. v.~15.}
K.~Dauid alſo maried the kings daughter of Geſſur.
\XRef{1.~Reg.~3. v.~3.}}
Eſther alſo among the reſt of the maydens was deliuered to him, that ſhe
might be kept in the number of the wemen. \V Who pleaſed him, and found
grace in his ſight. And he commanded the eunuch, that he should haſten
the wemens ornamentes, and should deliuer her, her partes, & ſeuen the
moſt beautiful maydens of the kings houſe, and should adorne and decke
both her and her wayting maydes. \V Who would not tel him her people and
countrie. For Mardocheus had commanded her, that of this thing she
should altogether keepe ſilence: \V who walked daily before the entrance
of the houſe, wherin the choſen virgins were kept, taking care of
Eſthers welfare, and deſirous to know what should chance vnto her. \V
And when the time of euerie virgin in order was come, that they should
goe in to the king, al things accomplished that perteyned to wemens
%%% o-0954
ornamentes, it was the twelfth moneth: yet ſo, that for ſix monethes
they were anoynted with oyle of myrtle, and other ſix monethes they vſed
certayne payntings and ſweete ſpices. \V And going in to the 
king, what ſoeuer they asked that perteyned to adorning they receiued:
and being trimmed, as it pleaſed them, they paſſed from the chamber of
the wemen to the kings chamber. \V And she that went in at euening, came
out in the morning, and from thence she was brought to the ſecond houſe,
that was vnder the hand of Suſagazus the eunuch, who was chiefe ouer the
kings concubines: neither had she power to returne any more to the king,
vnleſſe the king had willed, and had commanded her to come by name. \V
And the time by order coming about, the day was at hand, that Eſther the
daughter of Abihail the brother of Mardocheus, whom he had adopted for
his daughter, should goe in to the king. Who ſought not wemens
ornamentes, but whatſoeuer Egeus the eunuch the keper of the virgins
would, thoſe things he gaue her to her adorning. For she was exceding
fayre, and of incredible beautie, she ſemed to al mens eies gratious and
amiable. \V She therfore was brought to the chamber of king Aſſuerus the
tenth moneth, which is called Tebeth, in the ſeuenth yeare of his
reigne. \V And the
%%% 1060
king loued her more then al the wemen, and she had grace and mercie
before him aboue al the wemen, and he put the crowne of the kingdom on
her head, and made her reigne in ſteede of Vaſthi. \V And he commanded a
verie magnifical feaſt to be prepared to al the princes, and to his
ſeruantes, for the coniunction, and mariage of Eſther. And he gaue reſt
to al the prouinces, and beſtowed giftes according to princely
magnificence. \V And when virgins were ſought the ſecond time and gathered
together, Mardocheus taried at the kings gate, \V neither had Eſther as
yet vttered her countrie, & people, according to his cõmandment. For
what ſoeuer he commanded, Eſther obſerued: and she did al thinges ſo as
she was wont at that time, when he nourished her a litle one. \V At that
time therfore, when Mardocheus abode at the kings gate, Bagathan, and
Thares were angrie, two of the kings eunuches, that were porters, and
were chiefe in the firſt entrie of the palace: and they would make
inſurrection againſt the king, and kil him. \V Wherof Mardocheus
\SNote{Mardocheus hearing ſuſpicious wordes and obſeruing their actions
ſawe that they intended euel and informed the king, wherby their treſon
was diſcouered.
%%% !!! Two cites?
\Cite{Rab.~Calom. & Ioſep.}}
gotte notice, and immediatly he told it to queene Eſther: and ſhe to the
king in the name of Mardocheus, who had reported the thing vnto her. \V
It was examined, and found: and they were both hanged on a gibbet. And
it was put in the hiſtories, and regiſtred in the cronicles before the
\MNote{{Large C} ch.~12. v.~1.}
king. 


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
