%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Esther
%%%% Chapter 14
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-14


\project douay-rheims


%%% 1075
%%% o-0969
\startChapter[
  title={Chapter 14}
  ]

\Summary{Eſther prayeth in humilitie of ſpirite, that God wil deliuer
  the people from the crueltie of Aman.}

Eſther alſo the queene fled to our Lord, fearing the peril that was at
hand. \V And when ſhe had laid of her royal garmentes, ſhe tooke clothes
meete for weeping and mourning, and for diuers oyntmentes, filled her
head with aſhes and dung, and her bodie ſhe humbled with faſtes: and al
the places, in which before ſhe was accuſtomed to reioyſe, ſhe filled
with tearing of her heares. \V And ſhe beſought our Lord the God of
Iſrael, ſaying: My Lord, which onlie art our king, helpe me ſolitarie
woman, and which haue no other helper beſide thee. \V My peril is in my
handes. \V I haue heard of my father that thou Lord didſt take Iſrael
out from al Gentiles, and our fathers out of al their predeceſſours
before, that thou mighteſt poſſeſſe an euerlaſting inheritance, and thou
haſt done to them as thou haſt ſpoken. \V We haue ſinned in thy ſight,
and therfore thou haſt deliuered vs into the handes of our enemies: \V
for we haue
\Fix{worſhiped}{worſhipped}{likely typo, fixed in other}
their goddes. Thou art iuſt ô Lord. \V And now it ſuffiſeth them not,
that they oppreſſe vs with moſt hard ſeruitude, but imputing the force
of their handes to the might of their idols, \V they wil change thy
promiſſes, and deſtroy thine inheritance, and ſhut the
%%% 1076
mouthes of them that prayſe thee, and extinguiſh the glorie of thy
temple and altar, \V that they may open the mouthes of Gentiles, and
prayſe the ſtrength of idols, and magnifie a carnal king for euer. \V
Deliuer not ô Lord thy ſcepter to them
\SNote{\Emph{Idoles are nothing in the vvorld,}
\XRef{1.~Cor.~8.}
but bare imaginations of men.}
that are not, leſt they laugh at our ruine: but turne their counſel vpon
them, and deſtroy him, that hath begune to do cruelly againſt vs. \V
Remember ô Lord: and ſhew thy ſelf to vs in the time of our tribulation,
and geue me confidence Lord king of goddes, and of al power: \V geue me
ſpeach wel framed in my mouth in the preſence of the lion, and turne his
hart into the hatred of our enemie, that both him ſelf may periſh, and
the reſt that conſent vnto him. \V But deliuer vs in thy hand, and helpe
me, hauing no other helpe, but thee ô Lord which haſt the knowlege of al
things, \V and knoweſt that I hate the glorie of the wicked, and
\SNote{Of herſelfe, nor for
\Fix{wordlie}{worldlie}{obvious typo, fixed in other}
glorie ſhe deſired not to marie a heathen king but for the cõmon good
ſhe maried him, God diſpenſing with the law in this behalf. See
\XRef{ch.~2.}}
deteſt the bed of the vncircumciſed, and of euerie ſtranger. \V Thou
knoweſt my neceſſitie, that I abhorre the ſigne of my pride and glorie,
which is vpon my head in the daies of my oſtentation, and deteſt it as
the cloth of a woman in her monthlie floores, and weare it not in the
daies of my ſilence, \V and that I haue not eaten at Amans table,
neither the kings banket hath pleaſed me, and that I haue not drunke the
wine of
%%% o-0970
libamentes: \V and that thy handmayde did neuer reioyſe, ſince I was
tranſported hither vnto this day, but in thee ô Lord the God of
Abraham. \V O God ſtrong aboue al, heare the voyce of them, that haue no
other hope, and deliuer vs from the hand of the wicked, and deliuer me
from my feare.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
