%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Esther
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 1063
%%% o-0957
\startChapter[
  title={Chapter 5}
  ]

\Summary{Eſther ſtanding in the kings ſight he calleth her, 4.~she
  requeſteth that he and Aman wil dine with her. 7.~Againe she inuiteth
  them. 9.~Aman is more incenſed againſt Mardocheus, prepareth high
  gallowes to hang him on.}

And
\MNote{\Large H}
the third day Eſther put on royal garmentes, and ſtood in the court of
the kings houſe, which was the inner agaynſt the kings hal: but he ſate
vpon his throne in the conſiſtorie of the palace, againſt the doore of
the houſe.
\MNote{{\Large I} ch.~15. v.~4.}
\V 
\MNote{\Large K}
And when he had ſeene Eſther the queene ſtanding, she pleaſed his
eies, and he put forth toward her the golden rod, which he held in his
hand: who going neere, kiſſed the top of his rod. \V And the king ſaid
to her: What wilt thou queene Eſther? what is thy requeſt? yea if thou
wilt aske the halfe part of the kingdom, it shal be geuen thee. \V But
she anſwered: If it pleaſe the king, I beſeech thee that thou come to me
this day, and Aman with thee to a banket, which I haue prepared. \V And
the king forthwith, cal ye Aman, quoth he, quickly, that he may obey
Eſthers wil. The king therfore and Aman came to the banket, which the
queene had prepared for them. \V And the king ſaid to her, after that he
had drunke wine aboundantly: What doeſt thou deſire to be geuen thee?
and for what thing askeſt thou? although thou aske the halfpart of my
kingdom, thou
%%% o-0958
shalt obteyne. \V To whom Eſther anſwered:
\SNote{This prudence in delaying to propoſe her petition, increaſed the
kings deſire to know it, moued him to promiſe more aſſuredly, & ſo bond
him the more to accomplish it.}
My petition, and requeſtes are theſe: \V If I haue found grace in the
kings ſight, and if it pleaſe the king to geue that which I aske, and to
fulfil my petition: let the
%%% 1064
king and Aman come to the banket, which I haue prepared them, and to
morrow I wil open my wil to the king. \V Aman therfore went forth that
day ioyful and merie. And when he had ſeene Mardocheus ſitting before
the doores of the palace, and not onlie not to haue riſen vp to him, but
not ſo much as to haue moued from the place where he ſate, he was wrath
excedingly: \V and diſſembling his anger, and returning into his houſe,
he called together vnto him his freindes, and Zares his wife: \V and he
declared to them the greatneſſe of his riches, and the multitude of his
children, and with how great glorie the king had aduanced him aboue al
his princes and ſeruantes. \V And after theſe things he ſaid: Queene
Eſther alſo hath called none other to the banket with the king, but me:
with whom to morow alſo I shal dine with the king. \V And wheras I haue
al theſe things, I thinke I haue nothing ſo long as I shal ſee
Mardocheus the Iew ſitting before the kings doores. \V And Zares his
wife, & the reſt of his frendes anſwered him: Cõmand a great beame to be
prepared, hauing fiftie cubites in height, and ſpeake in the morning to
the king, that Mardocheus may be hanged vpon it, and ſo thou shalt
goe ioyful with the king to the banket. The counſel pleaſed him, and he
commanded an high gallowes to be prepared.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
