%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Esther
%%%% Chapter 11
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-11


\project douay-rheims


%%% 1072
%%% o-0966
\startChapter[
  title={Chapter 11}
  ]

\Summary{An Appendix and concluſion of this hiſtorie. 2.~The dreame of
  Mardocheus.}

In the fourth yeare when Ptolomee and Cleopatra reigned, Doſitheus,
which named himſelf a Prieſt, and of the Leuitical kinred, and Ptolomee
his ſonne brought this epiſtle of Phurim, which they ſayd Lyſimachus the
ſonne of Ptolomee did interprete in Ieruſalem.

%%% !!! Formatting!
This beginning was in the common edition, which is neither extant in
Hebrew, nor with any of
\SNote{Except the 72.}
the interpreters.

\V
\MNote{\Large A}
\MNote{The firſt part.

The preſage, and occaſion of Eſthers aduancement and Amans perſecuting
the Iewes.}
In the ſecond yeare, when Artaxerxes the greateſt reigned, in the
firſt day of the moneth Niſan, Mardocheus the ſonne of Iairi, the ſonne
of Semei, the ſonne of Cis, of the tribe of Beniamin. \V A Iew which
dwelt in the citie of Suſan, a great
%%% 1073
man and among the firſt of the kings court,
\SNote{Mardocheus had this dreame before Amans aduancement or his owne
and the peoples danger.}
ſaw a dreame. \V And he was of that number of captiues, whom
\CNote{\XRef{4.~Reg.~24.}}
Nabuchodonoſor the king of Babylon had tranſported from Ieruſalem with
Ieconias the king of Iuda: \V and this was his dreame: There appeared
voices, and tumultes, and thunders and earth quakes, and pertubation
vpon the earth: \V and behold
\SNote{Mardocheus was a good dragon, and Aman a bad one.}
two great dragons, prepared one againſt an other into battel. \V At
whoſe crie al nations were rayſed vppe, to fight againſt the nation of
the iuſt. \V And that was a day of darkeneſſe and danger, of tribulation
and diſtreſſe, and great feare vpon the earth. \V And the nation of the
iuſt fearing their euils, was trubled, and prepared to death. \V And they
cried to God: and they crying,
\SNote{Eſther by her fountaine of teares and humble ſupplication,
quenched a great flame of Gentiles againſt the Iewes.
\XRef{ch.~10.}}
a litle fountayne grew into a verie great riuer, and abounded into verie
manie waters. \V Light and ſunne aroſe, and the humble were exalted, and
they deuoured the glorious. \V Which when Mardocheus had ſeene, and riſen out
of his bed, he muſed what God would doe: and he had it fixed in his
mind, deſirous to know what the dreame ſhould
\MNote{{Large B} chap.~1. v.~1.}
ſignifie?


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
