%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Esther
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 1066
%%% o-0960
\startChapter[
  title={Chapter 8}
  ]

\Summary{Eſther informeth the king that Mardocheus is her vncle, he is
  aduanced in authoritie, 3.~and contrarie letters are ſent that the
  Iewes be ſaued, 11.~and their enemies ſlayne. 15.~Mardocheus is in
  high honour, and his whole nation is eſtemed and feared by other
  people.}

That day king Aſſuerus gaue vnto Eſther the queene the houſe of Aman the
Iewes aduerſarie, and Mardocheus went in before the kings preſence. For
Eſther confeſſed to him that he was her vncle. \V And the king tooke the
ring, which he had commanded to be taken agayne from Aman, and deliuered
it to Mardocheus. And Eſther appointed Mardocheus alſo ouer her
houſe. \V Neither content with theſe things, she
\SNote{Such honour is due to kinges, yea though they be Infideles,
becauſe they haue terreſtrial maieſtie and authoritie from God. And the
contrarie opinion and behauiour of heretikes, \Emph{vvhich deſpice
Dominion and blaſpheme Maieſtie}, is condemned by S.~Iude, in
\Cite{his Epiſtle. v.~8.}}
fel downe at the kings feete, and she
%%% 1067
wept and ſpeaking to him prayed him, that he would command that the
malice of Aman the Agagite, and his moſt wicked deuiſes, which he had
inuented agaynſt the Iewes, ſhould be of none effect. \V But he after
the maner put forth the golden ſcepter with his hand, by the which ſigne
of clemencie was shewed: and ſhe ryſing vp ſtood before him, \V and
ſaid: If it pleaſe the king, and if I haue found grace in his eies, and
my requeſt ſeme not contrarie to him, I beſech thee, that the old
letters of Aman the traytour and enemie of the Iewes, wherein he
commanded that in al the kings prouinces they ſhould perish, may by new
letters be corrected. \V For how can I abide
%%% o-0961
the murder and ſlaughter of my people? \V And king Aſſuerus anſwered
Eſther the queene, and Mardocheus the Iewe: Amans houſe I haue geuen to
Eſther, and him ſelf I haue commanded to be hanged on the gallowes,
becauſe he durſt lay handes on the Iewes. \V Write ye therfore to the
Iewes, as pleaſeth you, in the kings name, ſigning the letters with my
ring. For this was the cuſtome, that no man durſt ſpeake againſt the
\Fix{lettes,}{letters,}{obvious typo, fixed in other}
which were ſent in the kings name, and were ſigned with his ring. \V And
the kings ſcribes and ſecretaries being cald for (and it was the time of
the third moneth, which is called Siban) the three and twenteth day
therof letters were written, as Mardocheus would, to the Iewes, and to
the princes, and the lieuftenantes and iudges, which were rulers ouer
the hundred and ſeuen and twentie prouinces, from India euen to Æthopia:
to prouince and prouince, to people and people, according to their
languages and characters, and to the Iewes, according as they could
read, and heare. \V And the ſaid letters, which were ſent in the kings
name, were ſigned with his ring, and ſent by ryding poſtes: which
running through al the prouinces, should preuent the old letters with
the new meſſages. \V To whom the king gaue commandment, that they ſhould
ſpeake to the Iewes in euerie citie, and ſhould command them to be
gathered together in one, that they might ſtand for their liues, and
might kil and deſtroy al their enemies with their wiues and children and
al their houſes, and to take the ſpoyle of them. \V And there was
appointed through al the prouinces one day of reuenge, that is, the
thirtenth of the twelfth moneth Adar. \V And this was the content of the
letter, that in al landes and peoples, which were ſubiect to
%%% 1068
the empire of king Aſſuerus, it should be notified, the Iewes to be
readie to be reuenged of their enemies. \V And there went forth ſwift
poſtes
\Fix{cariyng}{carying}{obvious typo, fixed in other}
the
\Fix{maſſages,}{meſſages,}{likely typo, fixed in other}
and the kinges edict hong in Suſan. \V But Mardocheus going forth out of
the palace, and from the kinges preſence, shining in royal garmentes, to
wit, hyacinthine and skie colour, bearing a golden crowne on his head,
and clothed with a ſilke and purple cloke. And the citie reioyſed, and
was glad. \V But to the Iewes there ſemed a new light to riſe, ioye,
honour, and dauncing. \V With al peoples, cities, and prouinces, whither
ſoeuer the kinges commandmentes came, meruelous reioyſing, feaſtes, and
banketes, and holie day: in ſo much that manie of the other nation and
ſect, were ioyned to their religion and ceremonies. For great terrour of
the name of the Iewes had inuaded them
\MNote{{Large L} ch.~16. v.~1.}
al.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
