%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Esther
%%%% Chapter 15
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-15


\project douay-rheims


%%% 1076
%%% o-0970
\startChapter[
  title={Chapter 15}
  ]

\Summary{Mardocheus cauſeth Eſther to goe to the king and intreate for
  the people. 10.~She is terrified by
  \Fix{this}{his}{obvious typo, fixed in other}
  countenance, but God turneth his hart, and he calleth her curteouſly
  to him.}

%%% !!! Format!
This alſo I found added in the common edition.

And
\MNote{\Large G}
he commanded her (\Emph{no doubt but Mardocheus commanded Eſther}) that
ſhe ſhould goe vnto the king, and make petition for her people and for
her countrie. \V Remember (quoth he) the daies of thy humilitie, how
thou waſt brought vp in my hand, becauſe Aman the ſecond after the king
hath spoken againſt vs vnto death: \V and doe thou inuocate our Lord,
and ſpeake to the king for vs, & deliuer vs from
\MNote{{\Large H} ch.~13. v.~8.}
death.

%%% !!! Format!
Moreouer this alſo which foloweth I found in the common Greke.

%%% 1077
\V And in the third day ſhe layd away
\SNote{\L{Veſtimenta oratus} not \L{ornatus} her praying weede.}
the garmentes which ſhe ware, & was clothed in her glorie. \V And when
ſhe gliſtered in royal apparel, and had inuocated God the ruler and
ſauiour of al, ſhe tooke two wayting maydes, \V and vpon the one in
deede ſhe leaned, as it were for dilicatnes and verie much tendernes not
able to beare vp her bodie: \V but the other of her maydes folowed her
ladie, bearing vp her garmentes trayling on the ground. \V But her
ſelf in al her countenance of colour like the roſe, alſo with gratious
and ſhining eies
\SNote{A notable example for Noble wemen, how to kepe their ſtate, &
remaine neuertheles humble in hart.}
hid her mind ful of ſorow, and exceding feare. \V She therfore entring
through al the doores in order, ſtood before the king, where he ſate
vpon the throne of his kingdom, clothed with royal garmentes, and
glittering in gold, and pretious ſtones, and he was terrible to
behold. \V And when he had lifted vp his face, and with burning eies had
ſhewed the furie of his breaſt, the Queene fel downe, and her colour
being changed into palenes, ſhe reſted her wearie head vpon her
handmayde. \V And God turned the kings ſpirit in to mildnes, and in haſt
and fearing he lept out of the throne, and holding her vp in his armes,
til ſhe came to her ſelf, ſpake her fayre with theſe wordes: \V What
ayleth thee Eſther? I am thy brother, feare not. \V Thou shalt not die:
for this law is not made for thee, but for the common ſorte. \V Come
neere therfore, and touch the ſcepter. \V And when ſhe held her peace, he
tooke the golden rod, & put it vpon her necke, and kiſſed her, and ſayd:
Why ſpeakeſt thou not to me? \V Who anſwered:
%%% o-0971
I ſaw thee, my Lord,
\SNote{By this compariſon of angels she ment that the king excelled al
or moſt part of men.}
as an angel of God, and my hart was trubled for the feare of thy
glorie. \V For thou, my lord, art exceding maruelous, and thy face is
ful of graces. \V And when she ſpake, she fel downe agayne, and was
almoſt in a
\Fix{ſowne.}{ſwoone.}{possible typo, same in both}
\V But the king was trubled, and al his ſeruantes did comfort
\MNote{{\Large K} ch.~5. v.~2.}
her.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
