%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Esther
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 1071
%%% o-0965
\startChapter[
  title={Chapter 10}
  ]

\Summary{Aſſuerus ſubdueth manie countries. 4.~Mardocheus vnderſtandeth,
  and declareth his dreame. 9.~God ſeparateth his people from other
  nations by a better lotte.}

But king Aſſuerus made al the earth, and al the ilandes of the ſea
tributaries. \V Whoſe ſtrength and empire, and the dignitie and
highneſſe, wherewith he exalted Mardocheus, are written in the bookes of
the Medes, and of the Perſians: \V and how Mardocheus of the Iewes
kinred was ſecond after king Aſſuerus: and great with the Iewes, and
acceptable to the people of his bretheren, ſeking good to his people,
and ſpeaking thoſe things, which pertayned to the good of his ſeede.

%%% 1072
%%% !!! Formatting!
That
\MNote{S.~Ierome here aduertiſeth the reader that he found al hitherto in
the Hebrew. And the parcels which folow only in the Septuaginta Greke
Edition which either they tranſlated out of the Hebrew, or added by
inſpiration of the Holie Ghoſt.}
which is in the Hebrew I haue expreſſed moſt faithfully. And theſe
things that folow, I found written in
\CNote{The Greke of the 72.~Interpreters.}
the common edition, which are conteyned in the greeke tongue and
letters: and in the meane time this chapter was extant after the end of
the booke: which according to our cuſtome, we haue marked with an Obelus
before it, that is to ſay, a broch.

\V And Mardocheus ſaid: Theſe things are done of God. \V I remember the
dreame that I ſaw, ſignifying theſe verie things: neither was any of
them fruſtrate. \V The litle fountayne which grew into a riuer, and was
turned into light, and into the ſunne, and abounded into manie waters,
is Eſther, whom the king tooke to wife, and made her to be queene. \V
But the two dragons: I am, and Aman. \V The nations, that were
aſſembled: are they that endeuoured to deſtroy the name of the Iewes. \V
And my nation: is Iſrael, which cried to our Lord, and our Lord hath
ſaued his people: and he hath deliuered vs from al euils, and hath done
great ſignes and wonders among the nations: \V and he commanded that
there ſhould be two lottes, one of the people of God, and the other of
al Nations. \V And both lottes are come to the day appointed euen now
from that time before God to al nations: \V and our Lord hath remembred
his people, and hath had mercie on his inheritance. \V And theſe daies
ſhal be obſerued in the moneth of Adar the fourtenth, and fiftenth day
of the ſame moneth, with al diligence, and ioy of the people gathered
into one aſſemblie, through out al the generations hereafter of the
people of Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
