%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Esther
%%%% Chapter 16
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-16


\project douay-rheims


%%% 1077
%%% o-0971
\startChapter[
  title={Chapter 16}
  ]

\Summary{A copie of king Artaxerxes letters, which he ſent for the ſaftie of
  the Iewes, 10.~declaring the inſolencie of Aman and dutiful ſeruice of
  Mardocheus and Eſther, 20.~commanding al his prince to aſſiſt the
  Iewes, and with them to celebrate the day which is changed from ſorow
  to ioy.}

%%% !!! Format!
The copie of king Artaxerxes letter, which he ſent for the Iewes to al
the prouinces of his kingdom: the which alſo is not found in the Hebrew
volume.

%%% 1078
The
\MNote{\Large L}
great king Artaxerxes, from India to Æthiopia to the dukes and princes
of an hundred twentie ſeuen prouinces, which obey our commandment,
ſendeth greeting. \V Manie haue abuſed vnto pride the goodnes of
princes, and the honour, that hath beene beſtowed vpon them: \V and they
do not only endeuour to oppreſſe the kings ſubiectes, but not bearing the
glorie that is geuen them, they worke treaſon againſt them that gaue
it. \V Neither are they content not to geue thankes for the benefittes,
and to violate in them ſelues the lawes of humanitie, but they thincke
they can eſcape the ſentence of God alſo who ſeeth al things. \V And
they haue burſt forth into ſo great madnes, that ſuch as obſerue
diligently the offices committed vnto them, and do al thinges ſo that
they are worthie of al mens prayſe, them they endeuour to ouerthrow by
ſubtil meanes of lies, \V whiles with craftie fraude they deceiue the
ſimple eares of princes, and eſteming others by their owne nature. \V
Which thing is proued both out of old hiſtories, and by theſe thinges,
which are done daily, how
\SNote{At laſt this king ſaw his owne errour in beleuing one falſe
counſeller ouer much.
\XRef{ch.~13. v.~3.}}
the enterpriſes of kinges are depraued by the euil ſuggeſtions of
certaine men. \V Wherfore we muſt prouide for the peace of al
prouinces. \V Neither muſt you thincke, if we command diuers things,
that it cometh of the lightneſſe of our mind, but that we geue ſentence
according to the qualitie & neceſſitie of times, as the profite of the
common wealth requireth. \V And that you may more plainly vnderſtand,
that which we ſay, Aman the ſonne of Amadathi, both in mind and countrie
a Macedonian, and an aliene from the Perſians bloud, and with his
crueltie ſtayning our pietie, was receiued
%%% o-0972
of vs a ſtranger: \V and found our humanitie ſo great towardes him, that
he was called
\CNote{\XRef{Ch.~13. v.~6.}}
our father, and was adored of al ſecond after the king: \V who was puft
vp with ſo great ſwelling of arrogancie, that he went about to depriue
vs of kingdom and life. \V For by certaine new deuiſes and not heard of
before he hath ſought to kil Mardocheus, by whoſe fidelitie and
benefittes we liue, and Eſther the felow of our kingdom with al their
nation: \V thincking this that they being ſlayne, he might worke treaſon
to our ſolitarines, and might transferre the kingdom of the Perſians to
the Macedonians. \V But we haue found the Iewes, which were by that moſt
wicked of men oppoynted to be ſlayne, in no fault at al, but
contrariwiſe vſing iuſt lawes, \V and the children of the higheſt and
%%% 1079
the greateſt, and alwayes louing God, by whoſe benefite the kingdom was
geuen both to our fathers and to vs, and is kept vnto this day. \V
Wherfore know ye thoſe letters, which he directed in our name, to be of
none effect. \V For the which heinous fact, before the gates of this
citie, that is, of Suſan, both him ſelf that deuiſed it, and al his
kinred hang on gibbettes: not we, but God repaying him that which he
hath deſerued. \V But this edict, which we now ſend, let it be ſet forth
in al cities, that it be lawful for the Iewes to vſe their owne
lawes. \V Whom you muſt helpe, that thoſe, which had prepared them
ſelues to their ſlaughter, they may kil the thirtenth day of the twelfth
moneth, which is called Adar. \V For God omnipotent hath turned this day
of ſadnes and mourning into ioy to them. \V Wherfore
\SNote{As Mardocheus ſuggeſted, the king confirmed the obſeruation of a
feaſt in memorie of Gods benefice and ſo both Iewes and Gentiles kept
it.}
count you alſo this day among other feſtiual daies, and celebrate it
with al ioy, that hereafter alſo it may be knowne, \V that al, which
faythfully obey the Perſians, receiue worthie reward for their
fidelitie: and they that lie in wayte againſt their kingdom, periſh for
their wicked fact. \V And euerie prouince and citie, that wil not
be partaker of this ſolemnitie, let it periſh by ſword and fyre, and let it
ſo be deſtroyed, that not onlie men, but alſo beaſtes haue no acceſſe
therto for euer, for an example of contempt, and
\MNote{{\Large M} chap.~9. v.~1.}
diſobedience.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
