%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% Esther
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 1064
%%% o-0958
\startChapter[
  title={Chapter 6}
  ]

\Summary{The king hearing the good ſeruice of Mardocheus in detecting
  traitors, read in the chronicle (for which he had yet no rewarde)
  4.~commandeth Aman to honour him next to the king, 11.~which he
  performeth.}

That night the king
\SNote{Gods eye which neuer ſlepeth ſaw
\Fix{whath}{what}{obvious typo, fixed in other}
Aman intended: and for execution of his owne diuine iuſtice, vſed this
meanes
\Fix{bey}{by}{obvious typo, fixed in other}
ſubtracting ſlepe that time from the king and inſpiring his mind to
heare the hiſtories read, and to reward good ſeruice donne.
\Cite{Ioſeph.}}
paſſed without ſleepe, and he commanded the hiſtories and chronicles of
former times to be brought him. Which when they were read in his
preſence, \V they came to that place where it was written, how
Mardocheus had vttered the treaſon of Bagathan, and Thares the eunuches,
coueting to kil king Aſſuerus. \V Which when the king had heard, he
ſayd: What honour and reward hath Mardocheus receiued for this
fidelitie? His ſeruantes and miniſters ſaid to him: He hath receiued no
reward at al. \V And the king by and by, who is, quoth he, in the court?
For Aman had entered the inner court of the kings houſe, that he might
ſuggeſt to the king, and he might command Mardocheus to be hanged faſt
on the gibbet, which was prepared for him. \V The ſeruantes anſwered:
Aman ſtandeth in the
%%% 1065
court. And the king ſaid: Let him come in. \V And when he was come in,
he ſaid to him: What ought to be done to the man, whom the king is
%%% o-0959
deſirous to honour? But Aman
\SNote{Ambicious men are moſt blinde in conceite of their owne deſerts,
and fortunes.}
thinking in his hart, and ſuppoſing that the king would honour no other,
but him ſelf, \V anſwered: The man, whom the king deſireth to honour, \V
ought to be clothed with the kings garmentes, and to be ſet vpon the
horſe, that is for the kings ſaddle, and to take the kings crowne vpon
his head, \V and let the chiefe of the kings princes and nobles hold his
horſe, and going
\Fix{throuh}{through}{obvious typo, fixed in other}
the ſtreat of the citie, crie, and ſay: So shal
he be honoured, whomſoeuer the king wil honour. \V And the king ſaid to
him: Make haſt, and taking a robe and a horſe, do that thou haſt ſpoken,
to Mardocheus the Iewe, which ſitteth before the doores of the palace:
Beware thou pretermitte nothing of thoſe things, which thou haſt
ſpoken. \V Aman therfore tooke a robe and a horſe, and putting it on
Mardocheus, and ſetting him on the horſe, in the ſtreat of the citie
went before him, and cried: This honour is he worthie of, whom ſoeuer
the king is willing to honour. \V And Mardocheus returned to the palace
gate: and Aman made haſt to goe into his houſe, mourning and his head
couered: \V and he told Zares his wife, and his frendes al things, that
had chaunced him. To whom the wiſe men whom he had in counſel, and his
wife anſwered:
\SNote{Either they had read Gods promiſe to the Iewes
\XRef{Gen.~13.~15. &c.}
or coniectured this by humaine prudence.}
If Mardocheus be of the Iewes ſeede, before whom thou haſt begune to fal,
thou canſt not reſiſt him, but thou shalt fal in his ſight. \V As they
were yet ſpeaking, the kings eunuches came, and compelled him to goe
quickly to the banket, which the queene had prepared.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
