%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Esdras
%%%% Chapter 09
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-09


\project douay-rheims


%%% 0978
%%% o-0879
\startChapter[
  title={Chapter 9}
  ]

\Summary{For mariages and other aſſociation which the Iewes had with
  Gentiles, Eſdras lamenteth, 5.~and confeſsing the peoples iniquitie,
  prayeth God to conſerue them from vtter deſtruction.}

And after theſe thinges were accompliſhed, the princes came to me,
ſaying: The people of Iſrael, and the Prieſts and Leuites
\SNote{Malachias the Prophet complaineth alſo of this fault
\XRef{c.~2. v.~11.}
threatning Gods puniſhment both to ſuperiors and ſubiectes for not
correcting it.
\XRef{v.~12.}}
are not ſeperated from the people of the landes, and from the
abhominations of them, to witte, of the Chananeite, and Hetheite, and
Pherezeite, and Iebuſeite, and of the Ammonites, and Moabites, and the
Ægyptians, and Amorrheites. \V For they haue taken of their daughters to
them and to their ſonnes, and haue mingled the holie ſeede with the
people of the landes. The hand alſo of the princes and magiſtrates hath
bene firſt in this tranſgreſſion. \V And when I had heard this word, I
rent my cloke and my coate, and plucked of the heares of my head and
bearde, and
%%% 0979
ſate mourning. \V And there aſſembled to me al that feared the God of
Iſrael, for the tranſgreſſion of them, that were come from the
captiuitie, and I ſate ſorowful, vntil the euening ſacrifice. \V And at
the euening ſacrifice I roſe out of my affliction, and my cloke and
coate being rent I fell vpon my knees, and ſpred forth my hands to our
Lord my God, \V and ſaid: My God I am confounded and aſhamed to lifte vp
my face to thee: becauſe our iniquities are multiplied ouer our head,
and our ſinnes haue growen euen vnto heauen, \V from the dayes of our
fathers: yea and we our ſelues alſo haue ſinned greuouſly vnto this day,
and in our iniquities haue we bene deliuered our ſelues, and our Kinges,
and our Prieſtes into the hand of the kinges of the landes, and into the
ſword, and into captiuitie, and into ſpoyle, and into confuſion of
countenance, as alſo at this day. \V And now as it were a litle, and for
a moment was our prayer made before the Lord our God, that a remnant
might be left vs, and
\SNote{In reſpect of their great iniquities Eſdras preſumeth not to aske
the conſeruation of the whole people, but ſome reliques or ſmalleſt
part, as it were a litle poſt, or a naile of a whole houſe towards the
reedifying therof.}
a nayle might be geuen vs in his holie place, and that our God would
illuminate our eies, and would geue vs a litle life in our bondage. \V
Becauſe we are bondmen, and in our bondage our God did not forſake vs, &
he inclined mercie vpon vs before the king of the Perſians, to geue vs
life, and to aduance the houſe of our God, and to build the deſolations
therof and to geue vs a hedge in Iuda and Ieruſalem. \V And now what
ſhal we ſay, ô our God, after theſe thinges? becauſe we haue forſaken
thy cõmandmẽts, \V which thou haſt cõmanded in the hand of thy ſeruantes
the prophetes, ſaying: The land, to the which you enter to poſſeſſe it,
is an vncleane land, according to the vncleanneſſe of peoples, and of
other landes, by the abhominations of them that haue
%%% o-0880
filled it from mouth vnto mouth in
\Fix{theier}{their}{obvious typo, fixed in other}
coinquination. \V Now therfore geue not your daughters to their ſonnes,
and their daughters take not for your ſonnes, and doe ye not ſeeke their
peace, and their proſperity for euer: that you may be ſtrengthned, and
may eate the goodes that are of the land, and may haue your children
heyres for euer. \V And after al thinges that come vpon vs in our
moſt wicked workes, and our moſt great ſinne, becauſe thou our God haſt
deliuered vs from our iniquitie, and haſt geuen vs health as it is at
this day, \V that we ſhal not turne away, and make fruſtrate thy
commandementes, neither ſhould ioyne matrimonies with the peoples of
theſe abominations
%%% 0980
Why art thou angrie with vs vnto vtter deſtruction, not to leaue vs a
remnant vnto ſaluation? \V Lord God of Iſrael thou art iuſt: becauſe we
are leift, which ſhould be ſaued as at this day. Behold we are before
thee in our ſinne, for there can be no ſtanding before thee vpon this.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
