%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Esdras
%%%% Chapter 05
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-05


\project douay-rheims


%%% 0970
%%% o-0872
\startChapter[
  title={Chapter 5}
  ]

\Summary{By the exhortation of Aggæus, and Zacharias, the people procede
  in building the Temple. 3.~VVhich their enimies ſtriuing to hinder,
  5.~for deciſion of the cauſe, both parties write to king Darius.}

%%% 0971
And there prophecied Aggeus the Prophete, & Zacharias the ſonne of Addo,
prophecying to the Iewes, that were in Iewrie and Ieruſalem, in the name
of the God of Iſrael. \V Then roſe vp Zorobabel the ſonne of Salathiel,
and Ioſue the ſonne of Ioſedec, and began to build the temple of God in
Ieruſalem, and with them the prophetes of God helping them. \V But at
the ſame time there came to them Thathanai, who was prince beyond the
Riuer, and Stharbuzanai, and their counſelers: and ſayd thus to them:
Who hath geuen you counſel to build this houſe, and to repayre the
walles? \V Wherto we anſwered them, what the names were of the men that
were authors of that building. \V And
\SNote{God geuing corege to his ſeruantes, ſtricke their enemies with
terrour, and ſo made them ceaſe from
\Fix{hinderuig.}{hindering}{obvious typo, fixed in other}
his worke as they before intended.}
the eye of theyr God was ſet vpon the ancientes of the Iewes, and they
could not inhibite them. And it pleaſed them that the matter ſhould be
referred to Darius, and then they would ſatisfie agaynſt that
accuſation. \V The copie of the epiſtle, which Thathanai prince of the
countrie beyond the Riuer ſent, and Stharbuzanai, and his counſelers the
Arphaſacheites, which were beyond the Riuer, to Darius the king. \V The
word, which they ſent him, was writen thus: To Darius the king al
peace. \V Be it knowen to the king, that we went to Iurie, the prouince,
to the houſe of the great God, which is in building with ſtone
vnpoliſhed, and timber is put in the walles: and that worke is in
building diligently, and groweth in their handes. \V We therfore
demanded of thoſe ancientes, and thus we ſayd to them: Who hath geuen
you authoritie to build this houſe, & to repaire theſe walles? \V Yea
and their names we asked of them, that we might certifie thee: and we
writte the names of thoſe men, that are the chiefe among them. \V And
they anſwered vs theſe wordes, ſaying: We are the ſeruantes of the God
of heauen & earth, and we do build a temple, that was
%%% o-0873
built theſe manie yeares before, and which a great king of Iſrael built
and ſet vp. \V But after that our fathers prouoked the God of heauen to
wrath, he deliuered them into the handes of Nabuchodonoſor the king of
Babylon the Chaldee: this houſe alſo he deſtroyed, and his people he
tranſported into Babylon. \V But in the firſt yeare of Cyrus the king of
Babylon, Cyrus the king put forth an edict, that this houſe of God
ſhould be built. \V For the veſſeles alſo of the temple of God, of gold
and of ſiluer, which Nabuchodonoſor had taken out of
%%% 0972
the temple, that was in Ieruſalem, and had caried them into the temple
of Babylon, Cyrus the king brought forth out of the temple of Babylon,
and they were geuen to Saſſabaſar ſo called, whom alſo he appointed the
chiefe, \V and ſayd to him: Take theſe veſſels, and goe, and put them in
the temple, that is in Ieruſalem, and let the houſe of God be built in
his place. \V Then therfore came this Saſſabaſar, and layd the
foundations of the temple of God in Ieruſalem, and from that time vntil
now it is in building, and is not yet finiſhed. \V Now therfore if it
ſeeme good to the king, let him ſearch in the kings librarie, which is
in Babylon, whether it hath beene commanded by Cyrus the king, that the
houſe of God in Ieruſalem ſhould be built, and let him ſend the kings
pleaſure concerning this thing vnto vs.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
