%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Esdras
%%%% Chapter 06
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-06


\project douay-rheims


%%% 0972
%%% o-0873
\startChapter[
  title={Chapter 6}
  ]

\Summary{Darius finding in the regiſter, that Cyrus gaue licence to
  build the Temple, commandeth that none hinder it: 8.~geueth alſo money
  towardes the charges, and hoſtes for ſacrifice.}

Then Darius the king commanded, and they ſearched in the librarie of the
bookes, that were layd vp in Babylon, \V and there was found in
Ecbatanis, which is a caſtle in the prouince Medena, one volume, and
there was ſuch a cõmentarie writen therein: \V In the firſt yeare of
Cyrus the king: Cyrus the king decreed, that the houſe of God ſhould be
built, which is in Ieruſalem, in the place where they immolate hoſtes,
and that they lay the foundations ſupporting the height of threeſcore
cubites, & the bredth of threeſcore cubites, \V three rewes of ſtones
vnpoliſhed, & ſo rewes of new timber: and the coſtes ſhal be geuen out
of the kings houſe. \V Yea & the veſſels of the temple of God, of gold &
of ſiluer which Nabuchodonoſor had taken out of the Temple of Ieruſalem,
and had brought them into Babylon, let them be reſtored, & brought backe
into the temple of Ieruſalem vnto their place, which alſo were put in
the temple of God. \V Now therfore Thathanai prince of the countrie,
that is beyond the Riuer, Stharubazanai, and your counſelers the
%%% o-0874
Apharſacheites, which are beyond the Riuer, depart farre from them, \V
and ſuffer that temple of God to be made of the duke of the Iewes, and
of their ancientes, that they may build that houſe of God in his
place. \V Alſo there is cõmandment
%%% 0973
geuen from me what muſt be done of thoſe ancientes of the Iewes, that
the houſe of God may be built, to witte, that of the kings coffer, that
is, of the tributes, that are geuen out of the countrie beyond the
Riuer, the charges be diligently geuen to thoſe men, leſt the worke be
hindred. \V And if it ſhal be neceſſarie, calues alſo, and lambes, and
kiddes, for holocauſt to the God of heauen, wheate, ſalt, wyne, and oyle,
according to the rite of the Prieſtes that are in Ieruſalem, let there
be geuen them day
\Fix{dy}{by}{obvious typo, fixed in other}
day, that there be no complaynte in any thing. \V And let them offer
oblations to the God of heauen, and pray for the life of the king, and
of his children. \V By me therfore there is a decree made. That euerie
man, which ſhal alter this commandement, there be a beame taken of
his houſe, and ſet vp, and he be faſt hanged vpon it, and his houſe be
confiſcate. \V And the God, that hath made his name to dwel there,
deſtroy al kingdomes, and the people that ſhal extend theyr hand to
reſiſt, & to deſtroy the houſe of God, that is in Ieruſalẽ. I Darius
haue made the decree, which I wil to be diligently accõpliſhed. \V
Therfore Thathanai the prince of the countrie beyond the Riuer, and
Stharbuzani, and his counſelers, according to that which Darius the king
had commanded, ſo did execute it diligently. \V And the ancientes of the
Iewes built, and proſpered according to the prophecie of Aggeus the
prophet, and of Zacharias the ſonne of Addo: and they built and ſet vp,
the God of Iſrael commanding, and
\SNote{The hart of the king is in the hand of our Lord.
\XRef{Prouerb.~21.}}
Cyrus commanding, and Darius, and Artaxerxes the kings of the
Perſians. \V And they were finiſhing this houſe of God, vntil the third
day of the moneth of Adar, which is the ſixth yeare of Darius the
king. \V And the children of Iſrael, the Prieſtes and the Leuites, and
the reſt of the tranſmigration, made the dedication of the houſe of God
in ioy. \V And they offered in the dedication of the houſe of God,
calues an hundred, rammes two hundred, lambes foure hundred, buckgoates
for the ſinne of al Iſrael twelue, according to the number of the tribes
of Iſrael. \V And they ſet the Prieſtes in theyr orders, and the Leuites
in theyr courſes ouer the workes of God in Ieruſalem, as it is writen in
the booke of Moyſes. \V And the children of Iſrael of the tranſmigration
made the Phaſe, the fourtenth day of the firſt moneth. \V For al the
Prieſtes and the Leuites were purified as it were one man: al cleane to
%%% 0974
immolate the Phaſe for al the children of the tranſmigration, and for
theyr brethren the Prieſtes, and them ſelues. \V And the children of
Iſrael that were returned from the tranſmigration, did eate, and al that
had ſeparated them ſelues from the coinquination of the Gentiles of the
earth vnto them, to ſeeke our Lord the God of Iſrael. \V And they made
the ſolemnitie of Azymes ſeuen dayes in ioy, becauſe our Lord had made
them ioyful, and had turned the hart of the king of Aſſur to them,
%%% o-0875
that he ſhould helpe theyr handes in the worke of the houſe of our Lord
the God of Iſrael.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
