%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Esdras
%%%% Chapter 02
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-02


\project douay-rheims


%%% 0965
%%% o-0867
\startChapter[
  title={Chapter 2}
  ]

\Summary{The names and number of ſpecial men, which returned vnder the
  conduct of Zorobabel into Ieruſalem: 66.~their ſubſtance of cattel,
  68.~and their oblations for the reedifying of the Temple.}

And
\SNote{This enumeration of the Iſraelites, which aſcended into
Ieruſalem, ſignifieth the Elect which aſcend from the militant Church to
the triumphant.}
theſe are the children of the prouince, that went vp from the
captiuitie, which Nabuchodonoſor the king of Babylon had tranſported
into Babylon, and returned into Ieruſalem and Iuda, euerie man into his
citie. \V They that came with Zorobabel, Ioſue, Nehemia, Saraia,
Rahelaia, Mardochai, Belſan, Meſphar, Beguai, Rehum, Baana. The number
of the men of the people of Iſrael: \V The children of Pharos, two
thouſand an hundred ſeuentie two. \V The children of Sephatia, three
hundred ſeuentie two. \V The children of Area, ſeuen hundred ſeuentie
fiue. \V The children of Phahath Moab, of the children of Ioſue: Iohab,
two thouſand eight hundred fiftie foure. \V The children of Aelam, a
thouſand two hundred fiftie foure. \V The children of Zethua, nine
hundred fourtie fiue. \V The children of Zachai, ſeuen hundred
ſixtie. \V The children of Bani, ſix hundred fourtie two. \V The
children of Bebai, ſix hundred twentie three.
\V The children of Azgad, a thouſand two hundred twentie two.
\V The children of Adonicam, ſix hundred ſixtie ſix. \V The
children of Beguai, two thouſand fiftie ſix. \V The children of Adin,
foure hundred fiftie foure. \V The children of Ather, which were of
Ezechias, ninetie eight. \V The children of Beſai, three hundred and
twentie three. \V The children of Iora, an hundred twelue. \V The
\Fix{childre}{children}{obvious typo, fixed in other}
of Haſum, two hundred twentie three. \V The children of Gebbar, ninetie
fiue. \V The children of Bethlehem, an hundred twentie three. \V The men
of Netupha, fiftie ſix. \V The men of Anathoth, an hundred twentie
eight. \V The children 
%%% o-0868
of Azmaueth, fourtie two. \V The children of Cariathiarim, Cephira, and
Beroth, ſeuen hundred fourtie three. \V The children of Rama and Gabaa,
ſix hundred twentie one. \V The men of Machmas, an hundred twentie two.
%%% 0966
\V The men of Bethel and Hai, two hundred twentie three. \V The children
of Nebo, fiftie two. \V The children of Megbis, an hundred fiftie
ſix. \V The children of an other Ælam, a thouſand two hundred fiftie
foure. \V The children of Harim, three hundred twentie. \V The children
of Lod Hadad and Ono, ſeuen hundred twentie fiue. \V The children of
Iericho, three hundred fourtie fiue. \V The children of Senaa, three
thouſand ſix hundred thirtie. \V The Prieſtes: The children of Iadaia in
the houſe of Ioſue, nine hundred ſeuentie three. \V The children of
Emmer, a thouſand fiftie two. \V The children of Pheſhur, a thouſand two
hundred fourtie ſeuen. \V The children of Harim, a thouſand and
ſeuentie. \V The Leuites: The children of Ioſue and Cedmiel, the
children of Odouia, ſeuentie foure. \V The ſinging men: The children of
Aſaph, an hundred twentie eight. \V The children of the Porters: the
children of Sellum, the children of Ater, the children of Telmon, the
children of Accub, the children of Hatita, the children of Sobai: al an
hundred thirtie nine. \V The Natheneites: The children of Siha, the
children of Haſupha, the children of Tabbaoth, \V the children of Ceros,
the children of Siaa, the children of Phadon, \V the 
children of Lebana, the children of Hagaba, the children of Accub, \V
the children of Hagab, the children of Semlai, the children of Hanan, \V
the children of Gaddel, the children of Gaher, the children of Raaia, \V
the children of Raſin, the children of Necoda, the children of Gazam, \V
the children of Aza, the children of Phaſea, the children of Beſee, \V
the children of Aſena, the children of Munim, the
\Fix{children}{children of}{obvious typo, fixed in other}
Nephuſim, \V the children of Bacbuc, the children of Hacupha, the
children of Harhur, \V the children of Beſluth, the
children of Mahida, the children of Harſa, \V the children of Bercos,
the children of Siſara, the children of Thema, \V the children of Naſia,
the children of Hatipha, \V the children of the ſeruants of Salomon, the
children of Sotai, the children of Sophereth, the children of Pharuda,
\V the children of Iala, the children of Dercon, the childrẽ of
Geddel, \V the children of Saphatia, the children of Hatil, the children
of Phochereth, which were of Aſebaim, the children of Ami. \V Al the
Nathineites, and the children of the ſeruantes of Salomon, three hundred
ninetie two. \V And theſe are they that came vp from Thelmela,
Thelharſa, Cherub,
%%% 0967
and Adon, and Emer. And they could not ſhew the houſe of their fathers
and their ſeede, whether they were of Iſrael. \V The children of Dalaia,
the children of Tobia, the children of Necoda, ſix hundred fiftie
two. \V And of the children of the Prieſtes: The children of Hobia, the
children of Accos, the children of Berzellai, who tooke a wife of the
daughters of Berzellai the Galaadite, and was called by their name: \V
theſe ſought the writing of their genealogie, and
%%% o-0869
\SNote{Such as ſay they are prieſtes and can not ſhew their vocatiõ,
muſt not exerciſe that functon.}
found it not, and they were caſt out of the Prieſthood. \V And
Atherſatha ſaid to them, that they ſhould not eate of the Holie of
holies, til there roſe a prieſt learned and perfect. \V Al the
multitude as it were one man,
\SNote{Al aboue numbred of the tribes of Iuda, Beniamin, & Leui, do not
amoũt to 30.~thouſand three hundred. So in this general number are
contained aboue twelue thouſand of other tribes, not recited among the
reſt, as Rabbi Salomoa explicateth the difficultie.}
fourtie two thouſand three hundred ſixtie: \V beſide their men
ſeruantes, & wemen ſeruantes, which were ſeuen thouſand three hundred
thirtie ſeuen: and among them ſinging men, and ſinging wemen two
hundred. \V Their horſes ſeuen hundred thirtie ſix, their mules two
hundred fourtie fiue, \V their cameles, foure hundred thirtie fiue, their
aſſes ſix thouſand ſeuen hundred twentie. \V And of the princes of the
fathers, when they entred into the temple of our Lord, which is in
Ieruſalem, they offered voluntarily vnto the houſe of our Lord to build
it in his place. \V According to their abilities, they gaue the expenſes
of the worke, of gold ſixtie one thouſand ſoldes, of ſiluer fiue
thouſand poundes, & garmentes for the prieſtes an hundred. \V The
Prieſtes therfore & the Leuites, and they of the people, and the ſinging
men, and the porters, and the Nathinaites dwelt in their cities, and al
Iſrael in their cities.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
