%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Esdras
%%%% Chapter 08
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-08


\project douay-rheims


%%% 0976
%%% o-0877
\startChapter[
  title={Chapter 8}
  ]

\Summary{Eſdras reciteth thoſe that came with him from Babylon, 21.~the
  faſt which he appointed, 33.~and how they brought the holie veſſel
  into the Temple.}

Theſe therfore are the princes of the families, and the genealogie of
them, that came vp with me in the reigne of Artaxerxes the king
\SNote{This great number which by Eſdras perſwaſion came from Babylon,
ſignified the great fruict of ſoules conuerted from ſinne by the
exhortation of holie preachers.
\Cite{S.~Beda li.~2. in Eſdras c.~10.}}
out of Babylon. \V Of the children of Phinees, Gerſom. Of the children
of Ithamar, Daniel. Of the children of Dauid, Hattus. \V Of the children
of Sechenias, the children of Pharos, Zacharias: and with him were
numbered an hundred fiftie men. \V Of the children of Phahath Moab,
Eleoenai the ſonne of Zareha, and with him two hundred men. \V Of the
children of Sechenias, the ſonne of Ezechiel, and with him three hundred
men. \V Of the children of Adan, Abed the ſonne of Ionathan, and with
him fiftie men. \V Of the children of Alam, Iſaias the ſonne of Athalia,
and with him ſeuentie men. \V Of the children of Saphatia, Zebedia the
ſonne of Michael, and with him eightie men. \V Of the children of Ioab,
Obedia the ſonne of Iahiel, and with him two hundred and eightene
men. \V Of the children of Selomith, the ſonne of Ioſphia, and with him
an hundred ſixtie men. \V Of the children of Bebai, Zacharias the ſonne
of Bebai, and with him twentie eight men. \V Of the children of Azgad,
Ioanan the ſonne of Eccetan, and with him an hundred and ten men. \V Of
the children of Adonicam, which were the laſt: and theſe are theyr
names: Elipheleth, and Iehiel, and Samaias, and with them ſixtie men. \V
Of the children of Begui, Vthai and Zachur, and with them ſeuentie
men. \V And I gathered them to
%%% 0977
the riuer, which runneth downe to Ahaua, and we taried there three
daies: and I ſought among the people and among the Prieſtes for the
children of Leui, and found none there. \V Therfore I ſent Eliezer, and
Ariel, and Semeias, and Elnathan, and Iarib, and other Elnathan, and
Nathan, and Zacharias, and Moſollam princes: and Ioiarib, and Elnathan
wiſe men. \V And I ſent them to Eddo, which is chiefe in the place
of Chaſphia, and I did put in theyr mouth the wordes that they ſhould
ſpeake to Eddo, and his brethren the Nathineites in the place of
Chaſphia, that they ſhould bring vs miniſters of the houſe of our God. \V
And they brought vs by the good hand of our God vpon vs a moſt learned
man of the children of Mocholi the ſonne of Leui the ſonne of Iſrael,
and Sarabias and his ſonnes, and his brethren eightene, \V and Haſabias,
and with him Iſaias of the children of Merari, and his brethren, and his
ſonnes twentie. \V And al the Nathineites, which Dauid gaue, and the
princes for the miniſteries of the Leuites,
\Fix{Natineites}{Nathineites}{likely typo, fixed in other}
two hundred twentie: al theſe
%%% o-0878
were called by theyr names. \V And I proclaymed there
\SNote{It ſufficeth not to part from Babylon (that is, from ſinne) but
we muſt alſo doe workes of ſatisfaction, and therfore Eſdras here
proclamed an extraordinarie faſt to thoſe that were come from
captiuitie.}
a faſt beſide the Riuer of Ahaua, that we might be afflicted before the
Lord our God, and might deſire of him a right way for vs and our
children, and al our ſubſtance. \V For I was aſhamed to aſke the king
ayde and horſemen, that might defend vs from the enemie in the way:
becauſe we had ſayd to the king: The hand of our God is vpon al them,
that ſeeke him in goodneſſe: and his empire and ſtrength, and furie vpon
al them that forſake him. \V And we faſted, and beſought our God hereby:
and it fel out proſperouſely vnto vs. \V And I ſeparated twelue of the
chief Prieſtes, Sarabias, and Haſabias, and with them ten of theyr
brethren. \V And I weyed vnto them the ſiluer and gold, and the
conſecrated veſſels of the houſe of our God, which the king had offered
and his counſelers, and his princes, and al Iſrael of them, that were
found: \V and I weyed in theyr handes of ſiluer ſix hundred fiftie
talentes, and veſſels of ſiluer an hundred, of gold an hundred
talentes: \V and cuppes of gold twentie, which had a thouſand ſolides,
and two veſſels of the beſt ſhyning braſſe, fayre, as gold. \V And I
ſayd to them: You are the holie of our Lord, and the veſſels are holie,
and the ſiluer and gold, that is voluntarely offered to our Lord the God
of our fathers. \V Watch & keepe it, vntil you deliuer it by weight
before the princes
%%% 0978
of the Prieſtes, and of the Leuites, and the princes of the families of
Iſrael in Ieruſalem, into the treaſure of the houſe of our Lord. \V And
the Prieſtes and the Leuites receiued the weight of the ſiluer and gold,
and of the veſſels, to carie it to Ieruſalem into the houſe of our
God. \V We therfore did ſette forward from the riuer of Ahaua the
twelfth day of the firſt moneth to goe on to Ieruſalem: and the hand of
our God was vpon vs, and deliuered vs from the hand of the enemie, and
the lyer in wayte by the way. \V And we came to Ieruſalem, and we taried
there three dayes. \V And in the fourth day the ſiluer was weyed, and
the gold, and the veſſels in the houſe of our God by the hand of
Meremoth the ſonne of Vrias the Prieſt, and with him Eleazar the ſonne
of
\Fix{Phimees,}{Phinees,}{obvious typo, fixed in other}
and with them Iozabed the ſonne of Ioſue, and Noadaia the ſonne of Bemai
Leuites, \V according to the number and weight of al: and al the weight
was deſcribed at that tyme. \V Yea and the children of the
tranſmigration that were come from the captiuitie, offered holocauſtes
to the God of Iſrael, calues twelue for al the people of Iſrael, rammes
nyntie ſix, lambes ſeuentie ſeuen, buckegoates for ſinne twelue: al for
holocauſt to our Lord. \V And they gaue the kings edictes to the
princes, that were ouerſeers for the king, and the dukes beyond the
Riuer, and they aduanced the people and the houſe of God.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
