%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Esdras
%%%% Chapter 04
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-04


\project douay-rheims


%%% 0969
%%% o-0870
\startChapter[
  title={Chapter 4}
  ]

\Summary{The Schiſmatical Samaritanes, becauſe they are not admitted to
  communicate with the Iewes, endeuour to hinder the building of the
  Temple. 5.~Which neuertheles procedeth, al the dayes of king Cyrus:
  7.~but is hindered by Artexerxes, til the ſecond yeare of Darius.}

But the enemies of Iuda and Beniamin heard that the children of the
captiuitie built a temple to our Lord the God of Iſrael. \V And coming
to Zorobabel, and the princes of the fathers, they ſayd to them: Let vs
build with you, becauſe
\SNote{Schiſmatikes and Heretikes may not be admitted to communicate in
ſacrifice with Catholiques.}
euen as you, ſo do we ſeeke your God: Behold we haue immolated victimes
from the dayes of Aſor Haddan the king of Aſtur, which brought vs
hither. \V And Zorobabel ſayd to them, and Ioſue, and the reſt of the
princes of the fathers of Iſrael: It is not for you and vs to build a
houſe to our God,
%%% o-0871
but we our ſelues alone wil build to the Lord our God, as Cyrus the king
of the Perſians hath cõmanded vs. \V It came to paſſe therfore that the
people of the land hindred the handes of the people of Iuda, and trubled
them in building. \V And they hyred counſelers agaynſt them, to deſtroy
theyr counſel al the dayes of Cyrus the king of the Perſians, & vntil
the reigne of Darius the king of the Perſians. \V And in the reigne of
Aſſuerus in the begining of his reigne, they wrote an accuſation againſt
the inhabitantes of Iuda and Ieruſalem. \V And in the dayes of Artaxerxes
Beſelam, Mitridates, & Thabeel, & the reſt that were in theyr counſel
writ to Artaxerxes king of the Perſians: and the epiſtle of the
accuſation was writen in Syriake, and was read in the Sirian language.
\V Reum Beelteem, and Samſai ſcribe wrote one epiſtle from Ieruſalem to
Artaxerxes the king, of this tenure: \V Reum Beelteem, and Samſai
ſcribe, and the reſt of their counſelers, the Dineites, and the
Apharſathaceites, the Terphaleites,
\Fix{the the}{the}{obvious typo, fixed in other}
Apharſeites, the Erchueites, the Babylonians, the Suſanecheites, the
Dieuites, and the Aclamites, \V and the reſt of the Gentiles, which
Aſenaphar the great and glorious tranſported: and made them dwel in the
cities of Samaria, and in the reſt of the countries beyond the Riuer in
peace: (\V this is the copie of the epiſtle, which they ſent to him) To
Artaxerxes the king, thy ſeruantes, the men that are beyond the Riuer,
ſend greeting. \V Be it knowen to the king, that the Iewes, which came
vp from thee to vs, are come into 
%%% 0970
Ieruſalem a rebellious and naughtie citie, which they build making the
rampyres thereof, and repayring the walles. \V Now therfore be it knowen
to the king, that if that citie ſhal be built, and the walles therof
repayred, they will not geue tribute, and tolle, & yearly rentes, and
this damage wil come euen to the kinges. \V But we mindful of the ſalt
that we haue eaten in the palace, and becauſe we count it heynous to ſee
the kings harmes, therfore we haue ſent and certified the king, \V that
thou recount in the bookes of the hiſtories of thy fathers, and thou
ſhalt finde writen in the commentaries: and ſhalt know that that citie is a
rebellious citie, and hurtful to the kinges, and prouinces, and battels
are rayſed in it of old time: for the which cauſe alſo the citie it
ſelfe was deſtroyed. \V We certifie the king, that if that citie ſhal be
built, and the walles therof repayred, thou ſhalt haue no poſſeſſion
beyond the Riuer. \V The king ſent word to Reum Beelteem, and Samſai
ſcribe, and to the reſt that were in their counſel inhabitantes of
Samaria, and to the reſt beyond the Riuer, ſending greeting and
peace. \V The accuſation, which you haue ſent to vs, was openly read
before me, \V and I gaue commandment: and they recounted, and haue found,
that that citie of old time rebelleth agaynſt the kinges, and ſeditions
and battels are rayſed in it. \V For there haue beene alſo moſt valiant
kinges in Ieruſalem, which alſo had dominion ouer al the countrie, that
is beyond the Riuer. They tooke alſo tribute, and tolle, and rentes. \V
Now therfore heare the ſentence: Prohibite ye thoſe men, that that citie
be not
%%% o-0872
built, til it perhaps ſhal be commanded by me. \V See that you doe not
negligently accompliſh this thing, and by litle there grow euil agaynſt
the kinges. \V Therfore the copie of the edict of Artaxerxes the king
was read before Reum Beelteem, and Samſai the ſcribe, and their
counſelers: and they went in haſt into Ieruſalem to the Iewes, &
prohibited them with arme and ſtrength. \V Then was the worke of the houſe
of our Lord in Ieruſalem intermitted, and was not done vntil the ſecond
yeare of the reigne of Darius the king of the Perſians.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
