%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Esdras
%%%% Chapter 03
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-03


\project douay-rheims


%%% 0967
%%% o-0869
\startChapter[
  title={Chapter 3}
  ]

\Summary{An Altar is built for ſacrifice. 4.~The feaſt of Tabernacles
  ſolemnly celebrated. 8.~And in the ſecond yeare (after their returne)
  the Temple is founded with great ioy of the people, and mourning of
  ſome.}

And now the ſeuenth moneth was come, and the children of Iſrael were in
their cities: the people therfore was gathered together as it were one
man into Ieruſalem. \V And Ioſue the ſonne of Ioſedec roſe vp, and his
brethren the Prieſtes, and Zorobabel the ſonne of Salathiel, and his
brethren, and they built the altar of the God of Iſrael, that they might
offer on it holocauſtes, as it is writen in the law of Moyſes the man of
God. \V And they placed the altar of God vpon his feete, the people of
the lands round about putting
%%% 0968
them
\SNote{Notwithſtanding the terrour of infideles, Gods ſeruantes tooke
corage to offer ſacrifice.}
in feare, & they offered vpon it holocauſt to our Lord morning and
euening. \V And they made the ſolemnitie of tabernacles, as it is
writen, and holocauſt euery day by order according to the præcept, the
worke of the day in his day. \V And after theſe thinges the continual
holocauſt, as wel in the Calendes, as in al the ſolemnities of our Lord,
that were conſecrated, and in al wherein there was offered voluntarily a
gift to our Lord. \V From the firſt day of the ſeuenth moneth they began
to offer holocauſt to our Lord: moreouer the temple of God was not yet
founded. \V And they gaue money to hewers of ſtones and to maſons: meate
alſo and drinke, and oyle to the Sidonians and Tyrians, that they ſhould
bring ceder trees from Libanus
%%% o-0870
to the ſea vnto Ioppe, according to that which Cyrus the king of the
Perſians had commanded them. \V And in the ſecond yeare of their coming
to the temple of God in Ieruſalem, the ſecond moneth, began Zorobabel
the ſonne of Salathiel, and Ioſue the ſonne of Ioſedec, and the reſt of
their brethren the Prieſtes, and the Leuites, and al that were come from
the captiuitie into Ieruſalem, and they appoynted Leuites from twentie
yeares and vpward, that they ſhould haſten forward the worke of our
Lord. \V And Ioſue ſtood and his ſonnes, and his brethren, Cedmihel, and
his ſonnes, and the children of Iuda, as it were one man, that they
might be inſtant vpon them, that did the worke in the temple of God: the
ſonnes of Henadad, and their ſonnes, and their brethren Leuites. \V The
temple therfore of our Lord being founded by the maſons, the Prieſtes
ſtood in their attyre with trumpettes: and the Leuites the children of
Aſaph in cymbals, to prayſe God
\SNote{By the ordinance of Dauid.
\XRef{1.~Par.~25.}}
by the handes of Dauid the king of Iſrael. \V And they ſang together in
hymnes, and confeſſion to our Lord: Becauſe he is good, becauſe his
mercie is for euer vpon Iſrael. Al the people alſo made a ſhoute with a
lowde crie in prayſing our Lord, becauſe the temple of our Lord was
founded. \V Verie manie alſo of the Prieſtes and the Leuites, and the
princes of the fathers, & the ancients, that had ſeene the former
temple, when they ſaw this temple founded,
\SNote{By their weping they teſtified that the new temple was not ſo
excellent as the former. And therfore Aggeus prophecie
\XRef{(c.~2.)}
can not be vnderſtood of this temple, but of the Church of Chriſt.
\Cite{S.~Aug. li.~18. ciuit, c.~45.}}
they wept with a lowd voyce: and manie ſhouting in ioy, lifted vp their
voyce. \V Neither could anie man diſcerne the voice of the crie of them
that reioyced, and the voice of the weeping of the people: for one with
an other the people ſhowted with a lowd crie, and the voice was heard
far of.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
