%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Esdras
%%%% Chapter 07
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-07


\project douay-rheims


%%% 0974
%%% o-0875
\startChapter[
  title={Chapter 7}
  ]

\Summary{Eſdras,
\MNote{The ſecond part.

Eſdras inſtructeth the people.}
with manie other Prieſtes and Leuites, aſcendeth to Ieruſalem to teach,
and aſſiſt the people: 11.~bringing Artaxerxes Edict, declareth it to
the people, 27.~and geueth thankes to God.}

And after theſe thinges in the reigne of Artaxerxes king of Perſians,
Eſdras the ſonne of Saraias, the ſonne of Azarias, the ſonne of
Helcias, \V the ſonne of Sellum, the ſonne of Sadoc, the ſonne of
Achitob, \V the ſonne of Amarias, the ſonne of Azarias, the ſonne of
Maraioth, \V the ſonne of Zarahias, the ſonne of Ozi, the ſonne of
Bocci, \V the ſonne of Abiſue, the ſonne of Phinees, the ſonne of
Eleazar, the ſonne of Aaron the Prieſt from the begynning. \V The ſame
Eſdras
\SNote{Eſdras came with the firſt from Babylon
\XRef{(2.~Eſd.~12.)}
but returned thither and now aſcended the ſecond time to Ieruſalem.}
came vp from Babylon, and he was a quicke ſcribe in the law of
Moyſes, which our Lord God gaue to Iſrael: and the king gaue him
according to the hand of our Lord his God vpon him al his petition. \V And
there came vp of the children of Iſrael, and of the children of the
Prieſtes, and of the children of the Leuites, and of the ſinging men,
and of the porters, and of the Nathineites into Ieruſalem in the ſeuenth
yeare of Artaxerxes the king. \V And they came into Ieruſalem the fifth
moneth, that is the ſeuenth yeare of the king. \V For in the firſt day
of the firſt moneth he began to goe vp from Babylon, and in the firſt
day of the fifth moneth he came into Ieruſalem according to the good
hand of his God vpon him. \V For Eſdras prepared his hart to ſearch the
law of our Lord, and to doe and to teach in Iſrael preceptes and
iudgement. \V And this is the copie of the epiſtle of the edict, which
king Artaxerxes gaue to Eſdras the Prieſt, the learned ſcribe, in the
wordes and preceptes of our Lord, and his ceremonies in Iſrael. \V
Artaxerxes the king of kings to
%%% 0975
Eſdras the Prieſt, the moſt learned ſcribe of the law of God of heauẽ,
greeting. \V It is decreed by me that whõſoeuer it ſhal pleaſe in my
kingdom of the people of Iſrael, and of the Prieſtes and Leuites, to goe
into Ieruſalẽ, let him goe with thee. \V For thou art ſent from the face
of the king, and of his ſeuen counſelers, that thou mayſt viſite Iewrie
and Ieruſalem in the law of thy God, which is in thy hand. \V And that
thou maiſt carie the ſiluer & gold, which the king & his counſelers haue
voluntarily offered to the God of Iſrael, whoſe tabernacle is in
Ieruſalem. \V And al the ſiluer and gold whatſoeuer thou ſhalt finde in
al the prouince of Babylon, and the people wil offer, and of the
Prieſtes that ſhal voluntarely offer to the houſe of theyr God, which is
in Ieruſalem, \V take freely, and
%%% o-0876
bye diligently of this money calues, rammes, lambes, and the ſacrifices
and libamentes of them, and offer them vpon the altar of the temple of
your God, that is in Ieruſalem. \V Yea and if it ſhal pleaſe thee, and
thy brethren to doe any thing with the reſt of the ſiluer and gold, doe
ye according to the wil of your God. \V The veſſels alſo, which are
geuen thee for the miniſterie of the houſe of thy God, deliuer thou in
the ſight of God in Ieruſalem. \V Yea and other thinges wherof neede
shal be for the houſe of thy God, how much ſoeuer is neceſſarie for thee
to ſpend, thou shalt geue it out of the treaſure, and excheker of the
king, and from me. \V I Artaxerxes the king haue appointed and decreed
to al the keepers of the common coffer, that are beyond the Riuer, that
whatſoeuer Eſdras the Prieſt, the ſcribe of the law of God of heauen,
shal aske of you, you geue it without delay, \V vnto an hundred talentes
of ſiluer, and vnto an hundred cores of wheat, and vnto an hundred bates
of wyne, and vnto an hundred bates of oyle, but ſalt without meaſure. \V
Al that pertayneth to the rite of the God of heauen, let it be geuen
diligantly in the houſe of the God of heauen: leſt perhaps he be angrie
agaynſt the kingdom of the king, and of his ſonnes. \V We doe you alſo
to vnderſtand concerning al the Prieſtes, and Leuites, and the ſingers,
and the porters, the Nathineites, and miniſters of the houſe of this
God, that you haue no authoritie to put tolle and tribute, and yearlie
rentes vpon them. \V And thou Eſdras according to the wiſedom of thy
God, which is in thy hand, appoynt iudges and preſidentes, that they may
iudge for al the people, that is beyond the
%%% 0976
Riuer, that is for them which know the law of thy God, yea and the
ignorant teach ye frely. \V And euerie one that ſhal not doe the law of
thy God, and the law of thy king diligently, there ſhal be iudgement of
him, either vnto death, or into baniſhment, or to the confiſcation of
his ſubſtance, or at the leaſt into priſon. \V Bleſſed be our Lord, the
God of our fathers, which hath put this in the kinges hart, that he
would glorifie the houſe of our Lord, which is in Ieruſalem, \V and hath
inclined his mercie toward me before the king and his counſelers, and al
the mightie princes of the king: and I taking courage by the hand of our
Lord my God, which was on me, gathered together out of Iſrael princes
that ſhould goe vp with me.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
