%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Esdras
%%%% Chapter 10
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-10


\project douay-rheims


%%% 0980
%%% o-0880
\startChapter[
  title={Chapter 10}
  ]

\Summary{Eſdras calling the people together commandeth them to diſmiſse
  the ſtrange wemen, which they haue married: 14.~appointing officers to
  ſee it executed: 18.~and reciteth thoſe which had married ſuch wemen.}

Eſdras therfore thus praying, and beſeeching, and weeping, and lying
before the temple of God, there was gathered to him of Iſrael an
exceeding great companie of men and wemen and children, and the people
wept with much lamentation. \V And Sechenias the ſonne of Iehiel of the
children of Aelam anſwered, and ſaid to Eſdras: We haue tranſgreſſed
againſt our God, and haue taken to wiues ſtrange wemen of the people of
the land: and now if there be penance in Iſrael vpon this, \V let vs
make a couenant with the Lord our God, to put away al the wiues, and
them that are borne of them, according to the wil of our Lord, and of
them that feare the precept of the Lord our God: be it done according
to the law. \V Ariſe, it is thy part to diſcerne, and we wil be with
thee: take courage, and doe it. \V Eſdras therefore roſe vp, and adiured
the Princes of the Prieſtes and of the Leuites, and al Iſrael, that they
ſhould doe according to this word, and they ſware. \V And Eſdras roſe vp
before the houſe of God, and went to the chamber of
\SNote{Eſdras being extraordinarily ſent by God, to correct the people,
repayreth to the high Prieſts ſonne, by his authoritie calleth the
people together, and ſo procedeth to make reformation. So S.~Paul
conferred with other Apoſtles.
\XRef{Gal.~2.}}
Iohanan the ſonne of Eliaſib, and entered into it: he did eate no bread,
and dranke no water: for he mourned for the tranſgreſſion of them, that
were come out of the captiuitie. \V And there was a proclamation ſent in
Iuda and Ieruſalem, to al the children of the tranſmigration, that they
ſhould aſſemble together into Ieruſalem.
%%% o-0881
\V And euerie one that ſhal not come within three dayes, according to
the counſel of the princes and ancientes, al his ſubſtance ſhal be taken
away, and him ſelfe ſhal be caſt out of the companie of the
tranſmigration. \V There aſſembled therfore al the men of Iuda, and
Beniamin into Ieruſalem withing three dayes, that is the ninth moneth,
the twenteth day of the moneth: and al the people ſate in the ſtreate of
the houſe of God, trembling for the ſinne, and
\SNote{Their ſinne was puniſhed by ouer much rayne.
\XRef{v.~13.}
And ſo affliction gaue them more feeling of their faultes.}
the rayne. \V And
%%% 0981
Eſdras the Prieſt aroſe, and ſayd to them: You haue tranſgreſſed, and
taken ſtrange wemen to wife, to adde vpon the ſinne of Iſrael. \V And
now geue confeſſion to our Lord the God of your fathers, and doe his
pleaſure, and be ſeparated from the people of the land, and from your
wiues the ſtrangers. \V And al the multitude anſwered, and ſayd with a
lowde voyce: According to thy word vnto vs, ſo be it done. \V
Neuertheleſſe becauſe there is much people, and a tyme of rayne, and we
can not abyde to ſtand without, and it is not a worke of one day or two
(for we haue excedingly ſinned in this thing) \V let there be princes
appoynted in al the multitude: and let al in our cities that haue taken
ſtrangers to wife, come at ſette tymes, and with them the ancientes by
citie and citie, and the iudges therof, vntil the wrath of our God be
turned away from vs for this ſinne. \V Therfore Ionathan the ſonne of
Azahel, and Iaazia the ſonne of Thecua were appoynted ouer this, and
Meſollam and Sebethai Leuites did helpe them: \V and the children of the
tranſmigration did ſo. And Eſdras the Prieſt, and the men, princes of
the families went into the houſes of theyr fathers, and al by theyr
names, and they ſate in the firſt day of the tenth moneth to ſearch out
the matter. \V And al the men were fully counted that had taken ſtrangers
to wife, vnto the firſt day of the firſt
\Fix{mosneth.}{moneth.}{obvious typo, fixed in other}
\V And there were found of the ſonnes of the Prieſtes that had taken
ſtrangers to wife. Of the children of Ioſue the ſonne of Ioſedec, and
his brethren, Maaſia, and Eliezer, and Iarib, and Godolia. \V And they
gaue theyr handes to put away theyr wiues, and to offer for theyr
offence a ramme of the flocke. \V And of the children of Emmer, Hanani,
and Zebedia. \V And of the children of Harim, Maaſia, and Elia, and
Semeia, and Iehiel, and Ozias. \V And of the children of Pheſhur,
Elionai, Maaſia, Iſmael, Nathanael, Iozabed, and Elaſa. \V And of the
children of the Leuites, Iozabed, and Semei, and Celaia, the ſame is
Calita, Phataia, Iuda, and Eliezer. \V And of the ſinging men Eliaſib:
and of the porters, Sellum, and Thelem, and Vri. \V And of Iſrael, of
the children of Pharos, Remeia, and Iezia, and Melchia, and Miamin, and
Eliezer, and Melchia, and Banea. \V And of the children of Aelam,
Mathania, Zacharias, and Iehiel, and Abdi, and Ierimoth, and Elia. \V
And of the children of Zethua, Elioenai, Eliaſib, Mathania, Ierimuth,
and Zabad, and Aziza.
%%% 0982
\V And of the children of Bebai, Iohanan, Hanamia, Zabbai, Athalai. \V
And of the children of Bani, Moſollam, and Melluch, and Adaia, Iaſub,
and Saal, and Ramoth. \V And of the children of
%%% o-0882
Phahath Moab, Edna, and Chalal, Bananias, and Maaſias, Mathanias,
Beſeleel, Bennui, and Manaſſe. \V And of the children of Herem, Eliezer,
Ioſue, Melchias, Semeias, Simeon, \V Beniamin, Maloch, Samarias. \V And
of the children of Haſom, Mathanai, Mathatha, Zabad, Eliphelet, Iermai,
Manaſſe, Semei. \V Of the children of Bani, Maaddi, Amram, and Vel, \V
Baneas, and Badaias, Cheliau, \V Vania, Marimuth, and Eliaſib, \V
Mathanias, Mathanai, and Iaſi, \V and Bani, and Bennui, Semei, \V and
Salmias, and Nathan, and Adaias, \V and Mechnedebai, Siſai, Sarai, \V
Ezrel, and Selemiau, Semeria, \V Sellum, Amaria, Ioſeph. \V Of the
children of Nebo, Iehiel, Mathathias, Zabad, Zabina, Ieddu, and Ioel,
and Banaia. \V Al theſe had taken ſtrangers to wife, and there were of
them that had borne
\SNote{Amongſt other inconuenciences of vnlawful mariages one is that
children are borne illegitimate.}
children.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
