%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% The (original) Douay Rheims Bible 
%%%%
%%%% Old Testament
%%%% One Esdras
%%%% Chapter 01
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%




\startcomponent chapter-01


\project douay-rheims


%%% 0964
%%% o-0866
\startChapter[
  title={Chapter 1}
  ]

\Summary{Cyrus
\MNote{The firſt part.

The returne of Gods people from Babylon.}
king of Perſia moued by diuine inſpiration, releaſeth Gods people from
captiuitie, with licenſe to returne and build the Temple in Ieruſalem:
7.~reſtoring the holie veſſel, which Nabuchodonoſor had taken from
thence.}

In 
\CNote{\XRef{Iſaiæ~44.~45.}
\XRef{Ierem.~25.–~29.}}
the firſt yeare of Cyrus king of the Perſians, that the word of our
Lord by the mouth of Ieremie might be accompliſhd, our Lord rayſed vp
the ſpirit of Cyrus king of Perſians: and he made proclamation in al his
kingdom, yea by wryting, ſaying: \V Thus ſayth Cyrus king of the
Perſians: Al the kingdomes of the earth hath the Lord the God of heauen
geuen me, & he hath commanded me that I ſhould build him a houſe in
Ieruſalem, which is in Iewrie. \V Who is there among you of al his
people? His God be with him. Let him goe vp into Ieruſalem, which is in
Iewrie, and build the houſe of the Lord the God of Iſrael, he is the God
that is in Ieruſalem. \V And let al the reſt in al places wherſoeuer
they dwel, let euery man of his place helpe him, with ſiluer and gold,
and ſubſtance, and cattel, beſides that which they offer voluntarily to
the temple of God, which is in Ieruſalem. \V And there roſe vp the
princes of the fathers of Iuda and Beniamin, & the Prieſtes, and
Leuites, and euerie one, whoſe ſpirit God rayſed vp, to goe vp to build
the temple of our Lord, which was in Ieruſalem. \V And al that were
round about,
%%% o-0867
\SNote{Liberally gaue ſuch thinges into their handes.}
did helpe their handes in veſſels of ſiluer, and of gold, in ſubſtance,
and beaſtes, in furniture, beſides thoſe thinges which they had offered
voluntarily. \V King Cyrus alſo brought forth the veſſels of the temple
of our Lord, which Nabuchodonoſor had taken of Ieruſalem, and had put
them in the temple of his god. \V But Cyrus the king of Perſians brought
them forth by the hand of Mithridates the ſonne of Gazabar, & numbred
them to Saſſabaſar the prince of Iuda. \V And this is the number of
them; Phials of gold thirtie, phials of ſiluer a thouſand, kniues
twentie nine,
%%% 0965
goblettes of gold thirtie, \V goblettes of ſiluer of the ſecond order,
foure hundred tenne: other veſſels a thouſand. \V Al the veſſels of gold
and ſiluer, fiue thouſand foure hundred: Saſſabaſar tooke al with them,
that went vp from the tranſmigration of Babylon into Ieruſalem.


\stopChapter


\stopcomponent


%%% Local Variables:
%%% mode: TeX
%%% eval: (long-s-mode)
%%% eval: (set-input-method "TeX")
%%% fill-column: 72
%%% eval: (auto-fill-mode)
%%% coding: utf-8-unix
%%% End:
